### 2014-06-16 poster.tex
First draft at a poster for GAeL XXII (using an old picture)

\documentclass[table]{beamer}

\title{Using the internal language of toposes in algebraic geometry}
\author{Ingo Blechschmidt}
\institute{University of Augsburg}
\date{June 23th, 2014}

\usepackage[english]{babel}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{tikz}
\usepackage{booktabs}
\usepackage{ragged2e}
\usepackage[orientation=landscape,size=a2,scale=0.8]{beamerposter}

\usefonttheme{professionalfonts}
\usepackage{pxfonts}

\definecolor{lgreen} {RGB}{180,210,100}
\definecolor{dblue}  {RGB}{20,66,129}
\definecolor{ddblue} {RGB}{11,36,69}
\definecolor{lred}   {RGB}{220,0,0}
\definecolor{nred}   {RGB}{224,0,0}
\definecolor{norange}{RGB}{230,120,20}
\definecolor{nyellow}{RGB}{255,221,0}
\definecolor{ngreen} {RGB}{98,158,31}
\definecolor{dgreen} {RGB}{78,138,21}
\definecolor{nblue}  {RGB}{28,130,185}
\definecolor{jblue}  {RGB}{20,50,100}

\setbeamercolor{palette primary}   {fg=black,bg=white}
\setbeamercolor{palette secondary} {fg=black,bg=white}
\setbeamercolor{palette tertiary}  {bg=jblue,fg=white}
\setbeamercolor{palette quaternary}{fg=black,bg=white}
\setbeamercolor{structure}{fg=jblue}
\setbeamercolor{titlelike}         {bg=jblue,fg=white}
\setbeamercolor{frametitle}        {bg=jblue!10,fg=jblue}
\setbeamercolor{cboxb}{fg=black,bg=jblue}
\setbeamercolor{cboxr}{fg=black,bg=red}

\setbeamercolor{item}{fg=ngreen}
\setbeamercolor{item projected}{fg=white,bg=ngreen}

\setbeamercolor{block title}{fg=nred,bg=white}
\setbeamercolor{block body}{fg=black,bg=white}

\setbeamercolor{block alerted title}{fg=white,bg=jblue}
\setbeamercolor{block alerted body}{fg=black,bg=jblue!10}

\setbeamerfont{section in head/foot}{series=\bfseries}
\setbeamerfont{block title}{series=\bfseries}
\setbeamerfont{block alerted title}{series=\bfseries}
\setbeamerfont{frametitle}{series=\bfseries}
\setbeamerfont{frametitle}{size=\Large}
\setbeamerfont{block body}{series=\rmfamily}

\setbeamertemplate{items}[circle]
\setbeamertemplate{blocks}[width=0.0]
\beamertemplatenavigationsymbolsempty

\setbeamertemplate{block begin}
{
  \par\vskip\medskipamount
  \begin{beamercolorbox}[colsep*=0ex,dp={2ex},center]{block title}
    \vskip-0.25cm
    \usebeamerfont{block title}\large\insertblocktitle
    \begin{flushleft}
    \vskip-0.8em
    \begin{tikzpicture}[remember picture,overlay]
      \shade [inner color=gray,outer color=white]
      (0,0) rectangle (\textwidth,0.3cm);
    \end{tikzpicture}
    \end{flushleft}
  \end{beamercolorbox}
  {\parskip0pt\par}
  \ifbeamercolorempty[bg]{block title}
  {}
  {\ifbeamercolorempty[bg]{block body}{}{\nointerlineskip\vskip-0.5pt}}
  \usebeamerfont{block body}
  \vskip-0.5cm
  \begin{beamercolorbox}[colsep*=0ex,vmode]{block body}
  \justifying
}

\setbeamertemplate{block end}
{
  \end{beamercolorbox}
  \vskip\smallskipamount
}

\newlength{\inboxwd}
\newlength{\iinboxwd}
\newlength{\inboxrule}
\setbeamertemplate{block alerted begin}
{
  \par\vskip\medskipamount
  \begin{beamercolorbox}[sep=0ex,rounded=true,center,dp={2.2ex}]{block alerted title}
    \vskip0cm
    \usebeamerfont{block title}\large\insertblocktitle
  \end{beamercolorbox}
  {\parskip0pt\par}
  \usebeamerfont{block body}
  \vskip-0.8cm
  \begin{beamercolorbox}[sep=0.5cm, rounded=true,center]{block alerted title}
  \setlength{\inboxwd}{\linewidth}
  \addtolength{\inboxwd}{-1cm}
  \begin{beamercolorbox}[rounded=true,wd={\inboxwd},center]{block alerted body}
  \setlength{\iinboxwd}{\inboxwd}
  \setlength{\inboxrule}{\inboxwd}
  \addtolength{\iinboxwd}{-0.5cm}
  \addtolength{\inboxrule}{0.5cm}
  \begin{center}
  \begin{minipage}{\iinboxwd}
  \justifying
}

\setbeamertemplate{block alerted end}
{
  \end{minipage}
  \end{center}
  \end{beamercolorbox}
  \end{beamercolorbox}
  \vskip\smallskipamount
}

\setbeamertemplate{headline}{
 \leavevmode
  \begin{columns}
   \begin{column}{\linewidth}
    \vskip1cm
    \centering
    \usebeamercolor{title in headline}{\color{jblue}\huge{\textbf{\inserttitle}}\\[0.5ex]}
    \usebeamercolor{author in headline}{\color{fg}\large{\insertauthor},}
    \usebeamercolor{institute in headline}{\color{fg}\large{\insertinstitute}\\[1ex]}
    \vskip1cm
   \end{column}
   \vspace{1cm}
  \end{columns}
 %\hspace{0.5in}\begin{beamercolorbox}[wd=\linewidth,colsep=0.15cm]{cboxb}\end{beamercolorbox}
 %\vspace{0.1in}
}

\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\G}{\mathcal{G}}
\renewcommand{\H}{\mathcal{H}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\Sh}{\mathrm{Sh}}

\begin{document}

\begin{frame}[t]\begin{columns}[t]

\begin{column}{0.3\textwidth}
  \begin{alertblock}{Summary}
    With the internal language of toposes, we can:
    \begin{itemize}\justifying
    \item express sheaf-theoretic
    concepts in a simple, ele\-ment-ba\-sed language and thus understand them
    in a more conceptual way,
    \item mechanically obtain
    corresponding sheaf-theo\-re\-tic theorems for any (intuitionistic) theorem of
    linear or commutative algebra, and
    \item understand which properties spread from
    points to neighbourhoods.
    \end{itemize}
  \end{alertblock}
  \bigskip

  \begin{block}{What is a topos?}
    A \emph{topos} is a category which has finite limits, is cartesian closed and
    has a subobject classifier. More simply, a topos is a category which has
    similar properties as the category of sets.\medskip

    Important examples of toposes are
    the category of sets and
    the category of set-valued sheaves on a topological space.
  \end{block}
  \bigskip

  \begin{block}{What is the internal language?}
    The internal language of a topos~$\E$ allows us to
    %{\addtocounter{enumi}{1}\usebeamertemplate{enumerate item}}
    construct objects and morphisms of the topos,
    %{\addtocounter{enumi}{1}\usebeamertemplate{enumerate item}}
    formulate statements about them, and
    %{\addtocounter{enumi}{1}\usebeamertemplate{enumerate item}}
    prove such statements
    in a \emph{naive element-based} language.
    \emph{Special case:} The language of the topos of sets is the usual
    formal mathematical language.

    \begin{center}
      \begin{tabular}{ll}
        \toprule
        external point of view & internal point of view \\
        \midrule
        objects of~$\E$ & sets \\
        morphisms of~$\E$ & maps of sets \\
        \bottomrule
      \end{tabular}
    \end{center}
  \end{block}

  \vspace{3cm}
  \textbf{GAeL XXII, Trieste, 2014}
\end{column}

\begin{column}{0.3\textwidth}
  \begin{block}{The small Zariski topos}
    Let~$X$ be a scheme. Let~$\Sh(X)$ be the small Zariski topos, i.\,e.\@ the
    topos of set-valued sheaves on~$X$. From the point of view of~$\Sh(X)$,
    the structure sheaf~$\O_X$ looks like an \emph{ordinary ring} (instead of a
    sheaf of rings), and sheaves of~$\O_X$-modules look like \emph{ordinary
    modules} on that ring.\bigskip

    \includegraphics[width=\columnwidth]{bohr-topos}
  \end{block}
  \bigskip

  \begin{block}{Basic example}
    Let~$0 \to \F' \to \F \to \F'' \to 0$ be a short exact sequence of sheaves
    of~$\O_X$-modules. Assume that~$\F'$ and~$\F''$ are of finite type. Then it is
    well-known that~$\F$ is of finite type as well.\medskip

    A sheaf is of finite type if and only if, internally, it is a
    finitely generated module. Therefore the proposition follows \emph{at once}
    from interpreting the analogous statement of intuitionistic linear algebra
    in the little Zariski topos:
    Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence of modules.
    Assume that~$M$ and~$M''$ are finitely generated. Then~$M$ is as well.
    \medskip

    \emph{Caveat:} Proofs by contradiction can not be interpreted with the
    internal language.
  \end{block}
\end{column}

\begin{column}{0.3\textwidth}
  \begin{block}{Locally free sheaves}
    Let~$X$ be a reduced scheme. The structure sheaf~$\O_X$ looks like a \emph{field} from the
    internal point of view. This is notable; recall that neither the rings of
    local sections nor the stalks are in general fields.\medskip

    Let~$\F$ be a finite type sheaf of~$\O_X$-modules. 
    Then it is well-known that~$\F$ is locally free on a dense open subset
    of~$X$.\medskip

    This follows \emph{at once} from the following statement of intuitionistic
    linear algebra: Let~$M$ be a finitely generated vector space. Then~$M$ is
    \emph{not not} finite free.
  \end{block}
  \bigskip

  \begin{block}{Rational functions}
    The sheaf~$\K_X$ of rational functions can internally simply be defined as
    the total quotient ring of~$\O_X$.
  \end{block}
  \bigskip

  \begin{block}{Spreading of properties}
    The following meta-theorem covers a wide range of cases:
    Let~$\varphi$ be a property which can be formulated without
    using~``$\Rightarrow$'',~``$\neg$'', and~``$\forall$''. Then~$\varphi$
    holds at a point if and only if it holds on some open neighbourhood of the
    point.\medskip

    For instance, a sheaf of modules~$\F$ is zero if and only if, from the
    internal perspective, ``$\forall x \in \F{:}\ x = 0$''. Because of
    the~``$\forall$'', a stalk may be zero without the sheaf being zero on a
    neighbourhood.\medskip

    But if~$\F$ is of finite type, the condition can be reformulated using
    generators as ``$x_1 = 0 \wedge \cdots \wedge x_n = 0$''. The meta-theorem
    is applicable to this statement, thus a stalk is zero if and only if~$\F$
    is zero on a neighbourhood.
  \end{block}

  \vspace{1cm}
  \begin{alertblock}{Dictionary external vs. internal notions}
    Detailed notes are available at \url{http://tiny.cc/topos} (work in
    progress).
  \end{alertblock}
\end{column}

\end{columns}\end{frame}

\end{document}

### 2014-06-17 poster.tex
typo
Kripke-Joyal semantics and conceptual interpretation

@@
   \par\vskip\medskipamount
-  \begin{beamercolorbox}[sep=0ex,rounded=true,center,dp={2.2ex}]{block alerted title}
+  \begin{beamercolorbox}[sep=0ex,rounded=true,center,dp={2.4ex}]{block alerted title}
     \vskip0cm
@@
   \begin{alertblock}{Summary}
-    With the internal language of toposes, we can:
+    With the internal language of toposes, we can
     \begin{itemize}\justifying
@@
     corresponding sheaf-theo\-re\-tic theorems for any (intuitionistic) theorem of
-    linear or commutative algebra, and
+    linear and commutative algebra, and
     \item understand which properties spread from
@@
     has a subobject classifier. More simply, a topos is a category which has
-    similar properties as the category of sets.\medskip
+    similar properties to the category of sets.\medskip
 
@@
     in a \emph{naive element-based} language.
+    The translation of internal statements and proofs into external ones is
+    facilitated by an easy mechanical procedure, the \emph{Kripke--Joyal
+    semantics}.
     \emph{Special case:} The language of the topos of sets is the usual
@@
     finitely generated module. Therefore the proposition follows \emph{at once}
-    from interpreting the analogous statement of intuitionistic linear algebra
+    by interpreting the analogous statement of intuitionistic linear algebra
     in the little Zariski topos:
     Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence of modules.
-    Assume that~$M$ and~$M''$ are finitely generated. Then~$M$ is as well.
-    \medskip
+    If~$M'$ and~$M''$ are finitely generated, so is~$M$.\medskip
 
+    We can thus recognize notions and statements of scheme theory as notions and
+    statements of non-sheafy linear algebra.
     \emph{Caveat:} Proofs by contradiction can not be interpreted with the
@@
     Then it is well-known that~$\F$ is locally free on a dense open subset
-    of~$X$.\medskip
+    of~$X$. (Ravi Vakil, important hard exercise.) \medskip
 
-    This follows \emph{at once} from the following statement of intuitionistic
+    This follows \emph{at once} from the following easy proposition of intuitionistic
     linear algebra: Let~$M$ be a finitely generated vector space. Then~$M$ is
@@
   \begin{block}{Spreading of properties}
-    The following meta-theorem covers a wide range of cases:
+    The following metatheorem covers a wide range of cases:
     Let~$\varphi$ be a property which can be formulated without
@@
     generators as ``$x_1 = 0 \wedge \cdots \wedge x_n = 0$''. The meta-theorem
-    is applicable to this statement, thus a stalk is zero if and only if~$\F$
+    is applicable to this statement and thus a stalk is zero if and only if~$\F$
     is zero on a neighbourhood.
@@
   \vspace{1cm}
-  \begin{alertblock}{Dictionary external vs. internal notions}
-    Detailed notes are available at \url{http://tiny.cc/topos} (work in
+  \begin{alertblock}{Dictionary of external vs. internal notions}
+    Expository notes are available at \url{http://tiny.cc/topos} (work in
     progress).
@@
 \end{document}
+
+* term "intuitionistic"
+* more examples, other toposes, ...
+* further work
+* own contributions
+* "at once"
+* orphans
+* blue borders in alertboxes

### 2014-06-18 poster.tex
Height of the block captions

@@
 
+\makeatletter
+\newenvironment{beamercolorboxA}[2][]{%
+  \begingroup%
+    \def\beamer@colbox@coladd{0pt}%
+    \def\beamer@vmode{\leavevmode}%
+    \setkeys{beamercolbox}{%
+      wd=\textwidth,ht={},dp={},%
+      leftskip=0pt,rightskip=0pt plus1fil,%
+      sep=0pt,colsep=0pt,colsep*=0pt,%
+      shadow=false,rounded=false,ignorebg=false}%
+    \setkeys{beamercolbox}{#1}%
+    \ifbeamercolorempty[bg]{#2}{\@tempswafalse}{\@tempswatrue}%
+    \ifbeamer@colbox@ignorebg\@tempswafalse\fi%
+    \def\beamer@colbox@color{#2}%
+    \hsize=\beamer@colbox@wd%
+    \setbox\beamer@tempbox=\hbox\bgroup\vbox\bgroup%
+      \leftskip=\beamer@colbox@ls%
+      \advance\leftskip by\beamer@colbox@sep%
+      \rightskip=\beamer@colbox@rs%
+      \advance\rightskip by\beamer@colbox@sep%
+      \ifbeamer@colbox@ignorebg%
+        \colorlet{beamer@temp@color}{bg}%
+        \usebeamercolor[fg]{#2}%
+        \colorlet{bg}{beamer@temp@color}%
+      \else%
+        \usebeamercolor[fg]{#2}%
+      \fi%
+      \if@tempswa%
+        \advance\leftskip by\beamer@colbox@colsep%
+        \advance\rightskip by\beamer@colbox@colsep%
+        \ifdim\beamer@colbox@colsep=0pt\else\vskip\beamer@colbox@colsep\fi%
+        \ifdim\beamer@colbox@colseps=0pt\else\vskip\beamer@colbox@colseps\fi%
+      \fi%
+      \ifdim\beamer@colbox@sep=0pt\else\vskip-0.3cm\fi%
+      \beamer@vmode\ignorespaces}{%
+      \ifdim\beamer@colbox@sep=0pt\else\vskip\beamer@colbox@sep\fi%
+      \if@tempswa\ifdim\beamer@colbox@colsep=0pt\else\vskip\beamer@colbox@colsep\fi\fi%
+      \if@tempswa\ifdim\beamer@colbox@colseps=0pt\else\vskip\beamer@colbox@colseps\fi\fi%
+    \egroup\egroup%
+    \wd\beamer@tempbox=\hsize%
+    \@tempdima=\wd\beamer@tempbox%
+    \ifx\beamer@colbox@ht\@empty%
+    \else%
+      \ht\beamer@tempbox=\beamer@colbox@ht%
+    \fi%
+    \ifx\beamer@colbox@dp\@empty%
+    \else%
+      \dp\beamer@tempbox=\beamer@colbox@dp%
+    \fi%
+    \ifbeamer@colbox@rounded%
+      \if@tempswa%
+        \begin{beamerboxesrounded}[%
+          shadow=\beamer@colbox@shadow,%
+          lower=\beamer@colbox@color,%
+          upper=normal text,%
+          width=\beamer@colbox@wd]{}%
+          \box\beamer@tempbox%
+        \end{beamerboxesrounded}%
+      \else%
+        \ifdim\@tempdima>\textwidth%
+          \setbox\beamer@tempbox=\hbox to\textwidth{\hss\box\beamer@tempbox\hss}%
+        \fi%
+        \box\beamer@tempbox%
+      \fi%
+    \else%
+      \if@tempswa\setbox\beamer@tempbox=\hbox{\vbox{%
+        \usebeamercolor{\beamer@colbox@color}%
+        \advance\hsize by \beamer@colbox@colseps\relax%
+        \advance\hsize by \beamer@colbox@colseps\relax%
+        \hskip-\beamer@colbox@colseps%
+        \fboxsep=0pt\colorbox{bg}{%
+          \hskip\beamer@colbox@colseps%
+          \hbox{\box\beamer@tempbox}%
+          \hskip\beamer@colbox@colseps%
+        }%
+        \hskip-\beamer@colbox@colseps%
+      }}\fi%
+      \ifdim\@tempdima>\textwidth%
+        \setbox\beamer@tempbox=\hbox to\textwidth{\hskip0pt minus\beamer@leftmargin\relax\box\beamer@tempbox\hskip0pt minus\beamer@rightmargin\relax}%
+      \fi%
+      \box\beamer@tempbox%
+    \fi%
+  \endgroup%
+}
+\makeatother
+
 \setbeamertemplate{block begin}
@@
 \newlength{\iinboxwd}
-\newlength{\inboxrule}
 \setbeamertemplate{block alerted begin}
@@
   \par\vskip\medskipamount
-  \begin{beamercolorbox}[sep=0ex,rounded=true,center,dp={2.4ex}]{block alerted title}
-    \vskip0cm
+  \begin{beamercolorbox}[sep=0ex,rounded=true,center,dp={2.0ex}]{block alerted title}
+    \vskip-0.2cm
     \usebeamerfont{block title}\large\insertblocktitle
+    \vskip0.05cm
   \end{beamercolorbox}
@@
   \usebeamerfont{block body}
-  \vskip-0.8cm
-  \begin{beamercolorbox}[sep=0.5cm, rounded=true,center]{block alerted title}
+  \vskip-0.5cm
+  \begin{beamercolorboxA}[sep=0.5cm, rounded=true,center]{block alerted title}
   \setlength{\inboxwd}{\linewidth}
@@
   \setlength{\iinboxwd}{\inboxwd}
-  \setlength{\inboxrule}{\inboxwd}
   \addtolength{\iinboxwd}{-0.5cm}
-  \addtolength{\inboxrule}{0.5cm}
   \begin{center}
@@
   \end{beamercolorbox}
-  \end{beamercolorbox}
+  \end{beamercolorboxA}
   \vskip\smallskipamount

### 2014-06-18 poster.tex
Further keywords

@@
   \end{alertblock}
+  \tiny\rmfamily\justifying
+  Tensor product of sheaves $=$ internal ordinary tensor product,
+  internal Cartier divisors,
+  quasicoherent sheaves $=$ internal ordinary modules satisfying an interesting
+  condition,
+  more metatheorems about spreading of properties,
+  pullback along immersions $=$ internal sheafification,
+  relative spectrum $=$ internal spectrum,
+  scheme dimension $=$ Krull dimension of~$\O_X$,
+  dense $=$ not not,
+  further modal operators,
+  other toposes,
+  group schemes $=$ groups,
+  \ldots
 \end{column}
@@
 
-* term "intuitionistic"
-* more examples, other toposes, ...
-* further work
 * own contributions
 * "at once"
-* orphans
-* blue borders in alertboxes

### 2014-06-18 poster.tex
typo

@@
     the category of sets and
-    the category of set-valued sheaves on a topological space.
+    the category of sheaves on a topological space.
   \end{block}
@@
         morphisms of~$\E$ & maps of sets \\
+        monomorphisms in~$\E$ & surjective maps \\
+        epimorphisms in~$\E$ & injective maps \\
         \bottomrule
@@
 
-  \vspace{3cm}
+  \vspace{0.7cm}
   \textbf{GAeL XXII, Trieste, 2014}
@@
 
-    \includegraphics[width=\columnwidth]{bohr-topos}
+    \includegraphics[width=\columnwidth]{external-internal}
   \end{block}
@@
     Let~$0 \to \F' \to \F \to \F'' \to 0$ be a short exact sequence of sheaves
-    of~$\O_X$-modules. Assume that~$\F'$ and~$\F''$ are of finite type. Then it is
-    well-known that~$\F$ is of finite type as well.\medskip
+    of~$\O_X$-modules. It is well-known that if~$\F'$ and~$\F''$ are of finite type,
+    then~$\F$ is as well.\medskip
 
@@
     statements of non-sheafy linear algebra.
-    \emph{Caveat:} Proofs by contradiction can not be interpreted with the
+    \emph{Caveat:} Non-intuitionistic proofs by contradiction can not be interpreted with the
     internal language.
@@
     Let~$X$ be a reduced scheme. The structure sheaf~$\O_X$ looks like a \emph{field} from the
-    internal point of view. This is notable; recall that neither the rings of
-    local sections nor the stalks are in general fields.\medskip
+    internal point of view. Recall that neither the rings of
+    local sections nor the stalks are fields.\medskip
 
@@
     Then it is well-known that~$\F$ is locally free on a dense open subset
-    of~$X$. (Ravi Vakil, important hard exercise.) \medskip
+    of~$X$. (Important hard exercise in Ravi Vakil's notes.) \medskip
 
@@
     Let~$\varphi$ be a property which can be formulated without
-    using~``$\Rightarrow$'',~``$\neg$'', and~``$\forall$''. Then~$\varphi$
+    using~$\Rightarrow$,~$\neg$,~$\forall$. Then~$\varphi$
     holds at a point if and only if it holds on some open neighbourhood of the

### 2014-06-18 poster.tex
Image credits to Carina
Minor details

@@
     \usebeamercolor{title in headline}{\color{jblue}\huge{\textbf{\inserttitle}}\\[0.5ex]}
-    \usebeamercolor{author in headline}{\color{fg}\large{\insertauthor},}
+    \usebeamercolor{author in headline}{\color{fg}\large{\insertauthor}, }
     \usebeamercolor{institute in headline}{\color{fg}\large{\insertinstitute}\\[1ex]}
@@
     A \emph{topos} is a category which has finite limits, is cartesian closed and
-    has a subobject classifier. More simply, a topos is a category which has
+    has a subobject classifier. Intuitively, a topos is a category which has
     similar properties to the category of sets.\medskip
@@
 
+    \rotatebox{90}{\tiny Illustration: Carina Willbold}\hspace{-0.25cm}%
     \includegraphics[width=\columnwidth]{external-internal}
@@
     A sheaf is of finite type if and only if, internally, it is a
-    finitely generated module. Therefore the proposition follows \emph{at once}
-    by interpreting the analogous statement of intuitionistic linear algebra
+    finitely generated module. Therefore the proposition follows
+    \emph{immediately}
+    by interpreting the analogous statement of linear algebra
     in the little Zariski topos:
@@
 
-    This follows \emph{at once} from the following easy proposition of intuitionistic
+    This is an \emph{immediate} application of the following easy lemma of intuitionistic
     linear algebra: Let~$M$ be a finitely generated vector space. Then~$M$ is
@@
     But if~$\F$ is of finite type, the condition can be reformulated using
-    generators as ``$x_1 = 0 \wedge \cdots \wedge x_n = 0$''. The meta-theorem
+    generators as ``$x_1 = 0 \wedge \cdots \wedge x_n = 0$''. The metatheorem
     is applicable to this statement and thus a stalk is zero if and only if~$\F$
@@
   \tiny\rmfamily\justifying
+  Contents:
   Tensor product of sheaves $=$ internal ordinary tensor product,
-  internal Cartier divisors,
   quasicoherent sheaves $=$ internal ordinary modules satisfying an interesting
   condition,
+  internal Cartier divisors,
   more metatheorems about spreading of properties,
@@
   relative spectrum $=$ internal spectrum,
-  scheme dimension $=$ Krull dimension of~$\O_X$,
+  scheme dimension $=$ internal Krull dimension of~$\O_X$,
   dense $=$ not not,
-  further modal operators,
+  modal operators,
   other toposes,
@@
 \end{document}
-
-* own contributions
-* "at once"

### 2014-06-20 poster.tex
typo

@@
         morphisms of~$\E$ & maps of sets \\
-        monomorphisms in~$\E$ & surjective maps \\
-        epimorphisms in~$\E$ & injective maps \\
+        monomorphisms in~$\E$ & injective maps \\
+        epimorphisms in~$\E$ & surjective maps \\
         \bottomrule

### 2015-09-21 abstract-ihes2015.tex
Abstract for a talk proposal at IHES

\documentclass[10pt]{amsart}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{etex}
\usepackage{amsmath,amsthm,amssymb,stmaryrd,color,graphicx,multirow}
\usepackage{mathtools}
\usepackage{setspace}
\usepackage{bussproofs}
\usepackage{xspace}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{array}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage[bookmarksdepth=2,pdfencoding=auto]{hyperref}
\usepackage[all]{xy}

\usepackage{tikz}
\usetikzlibrary{calc,shapes.callouts,shapes.arrows}
\newcommand{\hcancel}[5]{%
    \tikz[baseline=(tocancel.base)]{
        \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
        \draw[red, line width=0.3mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
    }%
}

\usepackage[natbib=true,style=numeric]{biblatex}
\usepackage[babel]{csquotes}
\bibliography{bibliography}

\theoremstyle{definition}
\newtheorem{defn}{Definition}[section]
\newtheorem{ex}[defn]{Example}

\theoremstyle{plain}

\newtheorem{prop}[defn]{Proposition}
\newtheorem{cor}[defn]{Corollary}
\newtheorem{lemma}[defn]{Lemma}
\newtheorem{thm}[defn]{Theorem}

\theoremstyle{remark}
\newtheorem{rem}[defn]{Remark}
\newtheorem{question}[defn]{Question}

\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\FF}{\mathbb{F}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\A}{\mathcal{A}}
\renewcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\renewcommand{\H}{\mathcal{H}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\L}{\mathcal{L}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\GG}{\mathbb{G}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\qqq}{\mathfrak{q}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\nnn}{\mathfrak{n}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\HOM}{\mathcal{H}\mathrm{om}}
\newcommand{\id}{\mathrm{id}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\placeholder}{\underline{\quad}}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Grp}{\mathrm{Grp}}
\newcommand{\Vect}{\mathrm{Vect}}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\PSh}{\mathrm{PSh}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\Et}{\mathrm{\acute{E}t}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Mod}{\mathrm{Mod}}
\newcommand{\Alg}{\mathrm{Alg}}
\newcommand{\Ring}{\mathrm{Ring}}
\newcommand{\LRL}{\mathrm{LRL}}
\newcommand{\pt}{\mathrm{pt}}
\newcommand{\tors}{\mathrm{tors}}
\DeclareMathOperator{\Spec}{Spec}
\newcommand{\QcohSpec}[2]{\mathrm{Spec}^{\mathrm{qcoh}}_{#1}{#2}}
\newcommand{\RelSpec}[2]{\mathrm{RelSpec}_{#1}{#2}}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\colim}{colim}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\Ann}{Ann}
\DeclareMathOperator{\Int}{int}
\DeclareMathOperator{\Clos}{cl}
\DeclareMathOperator{\Kernel}{ker}
\DeclareMathOperator{\supp}{supp}
\newcommand{\Ass}{\mathrm{Ass}}
\newcommand{\Open}{\T}
\newcommand{\?}{\,{:}\,}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\Ll}{\vcentcolon\Longleftrightarrow}
\newcommand{\notat}[1]{{!#1}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\lhra}{\ensuremath{\lhook\joinrel\relbar\joinrel\rightarrow}}
\newcommand{\hra}{\hookrightarrow}
\newcommand{\brak}[1]{{\llbracket{#1}\rrbracket}}
\newcommand{\sdense}{{\widehat\Box}}
\newcommand{\sdenseother}{\Box}
\newcommand{\ie}{i.\,e.\@\xspace}
\newcommand{\eg}{e.\,g.\@\xspace}
\newcommand{\vs}{vs.\@\xspace}
\newcommand{\resp}{resp.\@\xspace}
\newcommand{\inv}{inv.\@}
\newcommand{\notnot}{\emph{not not}\xspace}
\newcommand{\affl}{\ensuremath{{\ul{\AA}^1_S}}\xspace}
\newcommand{\afflx}{\ensuremath{{\ul{\AA}^1_X}}\xspace}
\newcommand{\affla}{\ensuremath{{\ul{\AA}^1_{\Spec A}}}\xspace}
\newcommand{\xra}{\xrightarrow}

\newcommand{\XXX}[1]{\textbf{XXX: #1}}

\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defequiv}{\vcentcolon\equiv}
\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}

\definecolor{gray}{rgb}{0.7,0.7,0.7}

\title{Using the internal language of toposes in algebraic geometry}
\author{Ingo Blechschmidt}
\email{ingo.blechschmidt@math.uni-augsburg.de}
\pagestyle{empty}

\begin{document}

\begin{abstract}
  In algebraic geometry, one often studies schemes from a local or relative
  point of view. We describe how the internal language of certain
  toposes, the associated petit and gros Zariski toposes of a scheme, can be
  used to give simpler definitions and more conceptual proofs of the basic
  notions and observations in algebraic geometry.

  The starting point is that sheaves of rings and sheaves of modules look
  just like plain rings respectively modules from the internal point of view.
  In this way, some concepts and statements of scheme theory can be reduced to
  concepts and statements of intuitionistic linear algebra.

  Furthermore, modal operators can be used to model phrases such as ``on a
  dense open subset it holds that`` or ``on an open neighbourhood of a given
  point it holds that``. These operators define certain subtoposes; a
  generalization of the double-negation translation is useful in order to
  understand the internal universe of those subtoposes from the internal point
  of view of the ambient topos.

  A particularly interesting task is to internalize the ``relative spectrum''
  construction of algebraic geometry, which given a quasicoherent sheaf of algebras
  on a scheme~$X$ yields a scheme over~$X$. From the internal point of
  view, this construction should simply reduce to a intuitionistically sensible
  variant of the ordinary construction of the spectrum of a ring, but it turns
  out that this expectation is too naive and that a refined approach is
  necessary.
\end{abstract}

\maketitle

\end{document}

* Krull dimension
* Cartier divisors?
* Box translation
* internal spectrum
* big Zariski topos: criterion for affineness; étale topology

### 2015-09-21 abstract-ihes2015.tex
Final version

@@
 \email{ingo.blechschmidt@math.uni-augsburg.de}
-\pagestyle{empty}
 
@@
 \begin{abstract}
-  In algebraic geometry, one often studies schemes from a local or relative
-  point of view. We describe how the internal language of certain
+  We describe how the internal language of certain
   toposes, the associated petit and gros Zariski toposes of a scheme, can be
@@
   notions and observations in algebraic geometry.
+  % This is useful for studying schemes from a local or relative point of view.
 
-  The starting point is that sheaves of rings and sheaves of modules look
-  just like plain rings respectively modules from the internal point of view.
+  The starting point is that, from the internal point of view, sheaves of rings
+  and sheaves of modules look just like plain rings and plain modules.
   In this way, some concepts and statements of scheme theory can be reduced to
@@
 
-  A particularly interesting task is to internalize the ``relative spectrum''
-  construction of algebraic geometry, which given a quasicoherent sheaf of algebras
-  on a scheme~$X$ yields a scheme over~$X$. From the internal point of
-  view, this construction should simply reduce to a intuitionistically sensible
+  A particularly interesting task is to internalize the
+  construction of the relative spectrum, which, given a quasicoherent sheaf of algebras
+  on a scheme~$X$, yields a scheme over~$X$. From the internal point of
+  view, this construction should simply reduce to an intuitionistically sensible
   variant of the ordinary construction of the spectrum of a ring, but it turns
@@
 \maketitle
+\thispagestyle{empty}
 

### 2015-11-02 abstract-ihes2015.tex
typo

@@
   Furthermore, modal operators can be used to model phrases such as ``on a
-  dense open subset it holds that`` or ``on an open neighbourhood of a given
-  point it holds that``. These operators define certain subtoposes; a
+  dense open subset it holds that'' or ``on an open neighbourhood of a given
+  point it holds that''. These operators define certain subtoposes; a
   generalization of the double-negation translation is useful in order to

### 2015-11-20 slides-ihes2015.tex
Beginning of the slides for IHES 2015

\documentclass[12pt,utf8,notheorems,compress,t]{beamer}
\usepackage{etex}

\usepackage[english]{babel}

\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{mathtools}
\usepackage{tikz}
\usetikzlibrary{calc,shapes.callouts,shapes.arrows}

\usepackage[protrusion=true,expansion=true]{microtype}

\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}

\setlength\parskip{\medskipamount}
\setlength\parindent{0pt}

\title{Using the internal language of toposes in algebraic geometry}
\author{Ingo Blechschmidt}
\date{November 27th, 2015}

\usetheme{Warsaw}
\usecolortheme{seahorse}
%\usefonttheme{default}?
%\usepackage{kurier}?
\usefonttheme{serif}
%\usepackage{libertine}?
\usepackage{mathpazo}
\useinnertheme{rectangles}

\setbeamertemplate{blocks}[rounded][shadow=false]

\newenvironment{changemargin}[2]{%
  \begin{list}{}{%
    \setlength{\topsep}{0pt}%
    \setlength{\leftmargin}{#1}%
    \setlength{\rightmargin}{#2}%
    \setlength{\listparindent}{\parindent}%
    \setlength{\itemindent}{\parindent}%
    \setlength{\parsep}{\parskip}%
  }%
  \item[]}{\end{list}}

\newcommand{\pointthis}[2]{%
        \tikz[remember picture,baseline]{\node[anchor=base,inner sep=0,outer sep=0]%
        (#1) {#1};\node[overlay,rectangle callout,%
        callout relative pointer={(-0.2cm,0.8cm)},fill=blue!20] at ($(#1.north)+(1.8cm,-1.4cm)$) {#2};}%
        }%

\setbeamertemplate{frametitle}[default][colsep=-2bp,rounded=false,shadow=false,center]

\setbeamertemplate{headline}{}
\setbeamertemplate{navigation symbols}{}

\newcommand{\floatbox}[3]{%
  \raisebox{0pt}[0pt][0pt]{%
    \begin{picture}(0,0)(#1,#2)#3\end{picture}\leavevmode%
  }%
}

\newcommand{\backupstart}{
  \newcounter{framenumberpreappendix}
  \setcounter{framenumberpreappendix}{\value{framenumber}}
}
\newcommand{\backupend}{
  \addtocounter{framenumberpreappendix}{-\value{framenumber}}
  \addtocounter{framenumber}{\value{framenumberpreappendix}} 
}

\setbeamertemplate{footline}{%
  \begin{beamercolorbox}[wd=\paperwidth,ht=2.5ex,dp=1.25ex,right,rightskip=1mm,leftskip=1mm]{frametitle right}
    {\quad} \inserttitle \hfill \insertauthor \quad
    \insertframenumber\,/\,\inserttotalframenumber {\quad}
  \end{beamercolorbox}}

\newcommand{\hil}[1]{{\usebeamercolor[fg]{item}{\textbf{#1}}}}

\IfSubStr{\jobname}{\detokenize{nonotes}}{
  \setbeameroption{hide notes}
}{
  \setbeameroption{show notes}
}
\setbeamertemplate{note page}[plain]

\begin{document}

\begin{frame}[c]
  \centering
  \includegraphics[scale=0.3]{external-internal-small.png}
  \medskip

  \hil{Using the internal language of toposes in \\ algebraic geometry}
  \medskip

  \scriptsize
  Ingo Blechschmidt \\
  University of Augsburg
  \medskip

  Topos à l'IHES \\
  November 27th, 2015
  \par
\end{frame}

\frame[t]{\frametitle{Outline}\scriptsize\begin{itemize}\item[]\tableofcontents\end{itemize}}

\note{\justifying\fontsize{8pt}{9.6}\selectfont
  \begin{center}\large\textbf{Abstract}\end{center}

  \begin{changemargin}{2.5em}{2.5em}
    We describe how the internal language of certain
    toposes, the associated petit and gros Zariski toposes of a scheme, can be
    used to give simpler definitions and more conceptual proofs of the basic
    notions and observations in algebraic geometry.
    % This is useful for studying schemes from a local or relative point of view.

    The starting point is that, from the internal point of view, sheaves of rings
    and sheaves of modules look just like plain rings and plain modules.
    In this way, some concepts and statements of scheme theory can be reduced to
    concepts and statements of intuitionistic linear algebra.

    Furthermore, modal operators can be used to model phrases such as ``on a
    dense open subset it holds that'' or ``on an open neighbourhood of a given
    point it holds that''. These operators define certain subtoposes; a
    generalization of the double-negation translation is useful in order to
    understand the internal universe of those subtoposes from the internal point
    of view of the ambient topos.

    A particularly interesting task is to internalize the
    construction of the relative spectrum, which, given a quasicoherent sheaf of algebras
    on a scheme~$X$, yields a scheme over~$X$. From the internal point of
    view, this construction should simply reduce to an intuitionistically sensible
    variant of the ordinary construction of the spectrum of a ring, but it turns
    out that this expectation is too naive and that a refined approach is
    necessary.
  \end{changemargin}
}

\section{Basic examples}

\begin{frame}[c]\frametitle{Exact sequences}
  \centering
  \begin{minipage}{0.75\textwidth}
    \begin{exampleblock}{}
      \justifying
      Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence of
      modules. If~$M'$ and~$M''$ are finitely generated, so is~$M$.
    \end{exampleblock}
  \end{minipage}
  \medskip

  \scalebox{3}{$\Downarrow$}

  \begin{minipage}{0.75\textwidth}
    \begin{exampleblock}{}
      \justifying
      Let $0 \to \F' \to \F \to \F'' \to 0$ be a short exact sequence
      of~$\O_X$-modules. If~$\F'$ and~$\F''$ are locally of finite type, so
      is~$\F$.
    \end{exampleblock}
  \end{minipage}
  \par
\end{frame}

\begin{frame}[c]\frametitle{Local freeness}
  \centering
  \begin{minipage}{0.70\textwidth}
    \begin{exampleblock}{}
      \justifying
      Any finitely generated vector space does \emph{not not} possess a basis.
    \end{exampleblock}
  \end{minipage}
  \medskip

  \scalebox{3}{$\Downarrow$}

  \begin{minipage}{0.70\textwidth}
    \begin{exampleblock}{}
      \justifying
      Any sheaf of modules of finite type on a reduced scheme is locally free
      \emph{on a dense open subset}.
    \end{exampleblock}
  \end{minipage}
  \par
\end{frame}

\begin{frame}\frametitle{Further examples}
  Let~$X$ be a scheme. Internally to~$\Sh(X)$,
  \begin{itemize}
    \item any non-invertible element of~$\O_X$ is nilpotent,
    \item $\K_X$ is simply the total quotient ring of~$\O_X$,
    \item $\dim \O_X \leq n$ holds if and only if~$\dim X \leq n$.
  \end{itemize}
\end{frame}

\section{The $\Diamond$-translation}

\newcommand{\idiamond}{{\usebeamercolor[fg]{item}{\boldsymbol{\Diamond}}}}
\newcommand{\gdiamond}[1]{\textcolor{gray}{\boldsymbol{\Diamond}(}#1\textcolor{gray}{)}}

\begin{frame}\frametitle{The $\Diamond$-translation}
  Let~$\E_\Diamond \hookrightarrow \E$ be a subtopos given by a \pointthis{local
  operator}{$\Diamond : \Omega_\E \to \Omega_\E$}~$\Diamond$. Then
  \[ \E_\Diamond \models \varphi \qquad\text{iff}\qquad
    \E \models \varphi^\Diamond\only<2->{.}\only<1>{,}\]
  \only<1>{where the translation~$\varphi \mapsto \varphi^\Diamond$ is given by:
  \begin{align*}
    (s = t)^\Diamond &\defeqv \idiamond(s=t) \\
    (\varphi \wedge \psi)^\Diamond &\defeqv \gdiamond{\varphi^\Diamond \wedge \psi^\Diamond} \\
    (\varphi \vee \psi)^\Diamond &\defeqv \idiamond(\varphi^\Diamond \vee \psi^\Diamond) \\
    (\varphi \Rightarrow \psi)^\Diamond &\defeqv \gdiamond{\varphi^\Diamond \Rightarrow \psi^\Diamond} \\
    (\forall x\?X\_ \varphi(x))^\Diamond &\defeqv \gdiamond{\forall x\?X\_ \varphi^\Diamond(x)} \\
    (\exists x\?X\_ \varphi(x))^\Diamond &\defeqv \idiamond(\exists x\?X\_ \varphi^\Diamond(x))
  \end{align*}}

  \only<2->{
    Let~$X$ be a scheme. There are choices for~$\Diamond$ such that~$\Sh(X)
    \models \Diamond\varphi$ means that~$\varphi$ holds on \ldots
    \begin{itemize}
      \item \ldots{} a dense open subset.
      \item \ldots{} a given open subset~$U$.
      \item \ldots{} an open subset containing a given closed subset~$A$.
      \item \ldots{} a neighbourhood of a given point~$x \in X$.
    \end{itemize}

    \pause
    Can tackle the question~``$\varphi^\Diamond \stackrel{?}{\Rightarrow} \Diamond\varphi$'' logically.
  }
\end{frame}

\section{Quasicoherence}

\begin{frame}\frametitle{Quasicoherence}
  Let~$X$ be a scheme. Let~$\F$ be an~$\O_X$-module.

  Then~$\F$ is quasicoherent
  if and only if, internally to~$\Sh(X)$,
  \begin{quote}\textnormal{$\F[f^{-1}]$ is a $\Diamond_f$-sheaf for any~$f : \O_X$, \\[0.3em]
  \qquad\qquad where~$\Diamond_f\varphi \defeqv (\text{$f$ invertible} \Rightarrow \varphi)$.}
  \end{quote}
  \pause

  In particular: If~$\F$ is quasicoherent, then internally
  \[ (\text{$f$ invertible} \Rightarrow s = 0) \Longrightarrow
    \bigvee_{n \geq 0} f^n s = 0 \]
  \vspace*{-1.5em}\par%
  for any~$f : \O_X$ and~$s : \F$.
\end{frame}

\section{The relative and internal spectrum}

\begin{frame}\frametitle{}
\end{frame}

\end{document}

### 2015-11-21 slides-ihes2015.tex
Further work on the slides

@@
 \usepackage{mathtools}
+\usepackage{soul}\setul{0.2ex}{}
+\usepackage[all]{xy}
 \usepackage{tikz}
@@
 
+\newcommand{\A}{\mathcal{A}}
 \newcommand{\E}{\mathcal{E}}
 \newcommand{\F}{\mathcal{F}}
+\renewcommand{\G}{\mathcal{G}}
 \renewcommand{\O}{\mathcal{O}}
 \newcommand{\K}{\mathcal{K}}
+\newcommand{\ppp}{\mathfrak{p}}
 \newcommand{\defeq}{\vcentcolon=}
@@
 \newcommand{\Sh}{\mathrm{Sh}}
+\newcommand{\Set}{\mathrm{Set}}
+\newcommand{\Sch}{\mathrm{Sch}}
+\newcommand{\Hom}{\mathrm{Hom}}
+\DeclareMathOperator{\Spec}{Spec}
+\newcommand{\RelSpec}{\operatorname{\text{\ul{$\mathrm{Spec}$}}}}
 \renewcommand{\_}{\mathpunct{.}}
 \newcommand{\?}{\,{:}\,}
+\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
 
@@
         }%
+
+\newcommand{\hcancel}[5]{%
+    \tikz[baseline=(tocancel.base)]{
+        \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
+        \draw[red, line width=0.4mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
+    }%
+}
 
@@
 
-\section{Basic examples}
+\section{The internal language}
 
-\begin{frame}[c]\frametitle{Exact sequences}
-  \centering
+\begin{frame}\frametitle{The internal language of a topos}
+  Let~$\E$ be a topos. Then we can define the meaning of
+  \[ \hil{$\E \models \varphi \qquad\textnormal{(``$\varphi$ holds in $\E$'')}$} \]
+  for formulas~$\varphi$ over~$\E$.
+
+  \begin{center}
+    \small
+    \begin{tabular}{ll}
+      \toprule
+      externally & internally \\
+      \midrule
+      object & set/type \\
+      morphism & map of sets \\
+      monomorphism & injective map \\
+      epimorphism & surjective map \\
+      \bottomrule
+    \end{tabular}
+  \end{center}
+  \pause
+
+  \mbox{If~$\varphi$ implies~$\psi$ \hil{intuitionistically}, then~$\E \models \varphi$
+  implies~$\E \models \psi$.}
+\end{frame}
+
+\note{
+  Let~$X$ be a topological space (or locale) and let~$\alpha : \F \to \G$ be a
+  morphism of sheaves on~$X$. Then:
+  \allowdisplaybreaks
+  \begin{align*}
+    & \Sh(X) \models \speak{$\alpha$ is injective} \\[0.5em]
+    \Longleftrightarrow\
+    & \Sh(X) \models \forall s\?\F\_ \forall t\?\F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{for all open~$U \subseteq X$, sections $s \in \F(U)$:} \\
+    &\qquad
+      \text{for all open~$V \subseteq U$, sections $t \in \F(V)$:} \\
+    &\qquad\qquad
+        \text{for all open~$W \subseteq V$:} \\
+    &\qquad\qquad\qquad
+          \text{$\alpha_W(s|_W) = \alpha_W(t|_W)$ implies $s|_W = t|_W$} \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
+    &\qquad
+          \text{$\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{$\alpha$ is a monomorphism of sheaves}
+  \end{align*}
+}
+
+\begin{frame}\frametitle{Building a dictionary}
+  A \hil{scheme} is a locally ringed space~$(X,\O_X)$ which is
+  locally isomorphic to the \hil{spectrum of a commutative ring}:
+  \[ \Spec A \defeq \{ \ppp \subseteq A \,|\, \text{$\ppp$ is a prime ideal} \} \]
+  \pause
+
+  \begin{center}
+    \small
+    \begin{tabular}{ll}
+      \toprule
+      externally to $\Sh(X)$ & internally \\
+      \midrule
+      sheaf of rings & ring \\
+      sheaf of modules & module \\
+      sheaf of finite type & finitely generated module \\
+      finite locally free sheaf & finite free sheaf \\
+      \bottomrule
+    \end{tabular}
+  \end{center}
+
+  \hil{Motto:} \tabto{1.41cm} We can understand notions of algebraic geometry \\
+  \tabto{1.41cm} as notions of linear algebra internal to~$\Sh(X)$.
+\end{frame}
+
+\begin{frame}[c]\frametitle{Using the dictionary}
+  \begin{center}
     \begin{minipage}{0.75\textwidth}
@@
     \end{minipage}
-  \par
+  \end{center}
 \end{frame}
 
-\begin{frame}[c]\frametitle{Local freeness}
-  \centering
+\begin{frame}[c]\frametitle{Using the dictionary}
+  \begin{center}
+    \visible<2->{
       \begin{minipage}{0.70\textwidth}
@@
       \scalebox{3}{$\Downarrow$}
+    }
 
@@
     \end{minipage}
-  \par
+  \end{center}
+  \pause
+  \pause
+
+  \hil{Motto:} \tabto{1.41cm} \mbox{We can understand statements of algebraic geometry} \\
+  \tabto{1.41cm} as statements of linear algebra internal to~$\Sh(X)$.
 \end{frame}
 
-\begin{frame}\frametitle{Further examples}
+\begin{frame}\frametitle{Exploring the internal world}
   Let~$X$ be a scheme. Internally to~$\Sh(X)$,
@@
   \end{itemize}
+
+  \centering
+  \includegraphics[scale=0.3]{tierney-on-the-spectrum-of-a-ringed-topos} \\
+  \tiny
+  %Miles Tierney. ``On the spectrum of a ringed topos''. \\ In: \emph{Algebra,
+  %Topology, and Category Theory. A Collection of Papers in Honor of Samuel
+  %Eilenberg}. Ed.\@ by A.~Heller and M.~Tierney. Academic Press, 1976,
+  %pp.~189--210.
+  Miles Tierney. On the spectrum of a ringed topos. 1976.
+  \par
 \end{frame}
@@
   \only<2->{
-    Let~$X$ be a scheme. There are choices for~$\Diamond$ such that~$\Sh(X)
+    Let~$X$ be a scheme. Depending on~$\Diamond$,~$\Sh(X)
     \models \Diamond\varphi$ means that~$\varphi$ holds on \ldots
@@
 
-\section{Quasicoherence}
+\section{Quasicoherence of sheaves of modules}
 
@@
 
-\begin{frame}\frametitle{}
+\begin{frame}\frametitle{The absolute spectrum}
+  Let~$A$ be a commutative ring (in~$\Set$).
+
+  Is there a \hil{free local ring}~$A \to A'$ over~$A$?
+  \[ \xymatrix{
+    A \ar[rd] \ar[rrr] &&& R \\
+    & A' \ar@{-->}[rru]
+  } \]
+
+  \hil{No,} if we restrict to~$\Set$.
+
+  \hil{Yes,} if we allow a change of topos:
+  Then $A \to \O_{\Spec A}$ is the universal localization.
+\end{frame}
+
+\newcommand{\defspeca}{\text{topological space of the prime ideals of $A$}}
+\newcommand{\defspecb}{\text{topological space of the filters of $A$}}
+\newcommand{\defspecc}{\text{locale of the filters of $A$}}
+
+\begin{frame}\frametitle{The absolute spectrum, internalized}
+  Let~$A$ be a commutative ring in a topos~$\E$.
+
+  To construct the \hil{free local ring} over~$A$, give a constructive account
+  of the spectrum:
+  \begin{align*}
+    \only<1>{\Spec A &\defeq \defspeca}
+    \only<2->{\Spec A &\defeq \hcancel{$\defspeca$}{0pt}{3pt}{0pt}{-2pt}}
+    \\
+    \only<3>{&\defeq \defspecb}
+    \only<4->{&\defeq \hcancel{$\defspecb$}{0pt}{3pt}{0pt}{-2pt}}
+    \\
+    \only<5->{&\defeq \defspecc}
+  \end{align*}
+  \pause
+  \pause
+  \pause
+  \pause
+  Define the frame of opens of~$\Spec A$ to be the frame of radical ideals
+  in~$A$.
+  \pause
+
+  This gives an internal description of
+  Monique Hakim's spectrum functor~$\mathrm{RT} \to \mathrm{LRT}$.
+\end{frame}
+
+\begin{frame}\frametitle{The relative spectrum}
+  Let~$X$ be a scheme. Let~$\A$ be a quasicoherent~$\O_X$-algebra.
+  Goal: Describe~$\RelSpec_X \A$, a scheme over~$X$, internally.
+
+  Desired universal property:
+  \[ \Hom_{\Sch/X}(T, \RelSpec_X \A) \cong \Hom_{\mathrm{Alg}(\O_X)}(\A,
+  \mu_*\O_T) \]
+  for all~$X$-schemes~$T \xrightarrow{\mu} X$.
 \end{frame}
@@
 \end{document}
+
+* Big toposes!
+
+  \scriptsize
+  \begin{itemize}
+    \item $\Sh(X) \models f = g$ iff~$f = g \in \F(X)$.
+    \item $\Sh(X) \models f = g \vee f = h$ iff on a cover~$X = \bigcup_i U_i$,
+    $f|_{U_i} = g|_{U_i}$ or~$f|_{U_i} = h|_{U_i}$.
+    \item $\Sh(X) \models \neg\neg(f = g)$ iff~$f = g$ on a dense open subset
+    of~$X$.
+  \end{itemize}

### 2015-11-22 slides-ihes2015.tex
The relative spectrum from the internal point of view

@@
 \newcommand{\ppp}{\mathfrak{p}}
+\newcommand{\fff}{\mathfrak{f}}
 \newcommand{\defeq}{\vcentcolon=}
@@
   \[ \hil{$\E \models \varphi \qquad\textnormal{(``$\varphi$ holds in $\E$'')}$} \]
-  for formulas~$\varphi$ over~$\E$.
+  for formulas~$\varphi$ over~$\E$ using the \hil{Kripke--Joyal semantics}.
 
@@
 
-\section{The $\Diamond$-translation}
+\section{\texorpdfstring{The $\Diamond$-translation}{The ♦-translation}}  % XXX
 
@@
 \begin{frame}\frametitle{The relative spectrum}
-  Let~$X$ be a scheme. Let~$\A$ be a quasicoherent~$\O_X$-algebra.
-  Goal: Describe~$\RelSpec_X \A$, a scheme over~$X$, internally.
+  Let~$X$ be a scheme and~$\O_X \xrightarrow{\varphi} \A$ be a quasicoherent algebra.
+  Can we describe~$\RelSpec_X \A$, a scheme over~$X$, internally?
 
@@
   for all~$X$-schemes~$T \xrightarrow{\mu} X$.
+  \pause
+
+  Solution: Define internally the frame of~$\RelSpec_X \A$ to be the frame of
+  those radical ideals~$I \subseteq \A$ such that
+  \[ \forall f\?\O_X\_ \forall s\?\A\_
+    (\text{$f$ invertible in~$\O_X$} \Rightarrow s \in I) \Longrightarrow
+      fs \in I. \]
+  \pause
+  Its points are those filters~$\fff$ of~$\A$ such that
+  \[ \forall f\?\O_X\_ \varphi(f) \in \fff \Rightarrow \text{$f$ invertible in~$\O_X$}. \]
 \end{frame}
+
+\begin{frame}\frametitle{The relative spectrum, reformulated}
+  Let~$B \to A$ be an algebra in topos.
+
+  Is there a \hil{free local and local-over-$B$ ring}~$A \to A'$ over~$A$?
+
+  \[ \xymatrix{
+    B \ar[r]\ar@/^1.5pc/[rrrr]\ar@/_/[rrd] & A \ar[rd] \ar[rrr] &&& R \\
+    && A' \ar@{-->}[rru]
+  } \]
+\end{frame}
+% Better decoration (local rings, local morphisms)
+
+\begin{frame}\frametitle{Limits of locally ringed locales}
+  Let~$(X_i)_i$ be a diagram of locally ringed locales.
+
+  Its limit is given by \hil{relocalizing} the corresponding limit
+  in~$\mathrm{RL}$.
+\end{frame}
+% Mention only orally?
+% Phrase more categorically.
+% Extend to toposes.
 

### 2015-11-22 slides-ihes2015.tex
typo

@@
       sheaf of finite type & finitely generated module \\
-      finite locally free sheaf & finite free sheaf \\
+      finite locally free sheaf & finite free module \\
       \bottomrule
@@
       \item \ldots{} an open subset containing a given closed subset~$A$.
-      \item \ldots{} a neighbourhood of a given point~$x \in X$.
+      \item \ldots{} an open neighbourhood of a given point~$x \in X$.
     \end{itemize}

### 2015-11-25 slides-ihes2015.tex
Further work on the slides
Directory reorganization

@@
   \centering
-  \includegraphics[scale=0.3]{external-internal-small.png}
+  \includegraphics[scale=0.3]{images/external-internal-small}
   \medskip
@@
 }
+
+\begin{frame}[plain,c]
+  \centering
+  \includegraphics[scale=0.37]{images/math-overflow-steven-gubkin}
+  \par
+\end{frame}
 
@@
   \centering
-  \includegraphics[scale=0.3]{tierney-on-the-spectrum-of-a-ringed-topos} \\
+  \includegraphics[scale=0.3]{images/tierney-on-the-spectrum-of-a-ringed-topos} \\
   \tiny
@@
 
-* Big toposes!
+1. Tell story; begin with MO question
+2. Basics (include praise for Mike Shulman)
+3. Curious field property and more advanced examples
+4. Relative spectrum (also mention equivalence to Hakim's spectrum in dimension zero)
+5. Outlook: synthetic algebraic geometry, relative geometry, big Zariski topos
 

### 2015-11-26 slides-ihes2015.tex
Ending the talk with a "funny slide"
Slides almost finished

@@
 \usepackage[all]{xy}
+\xyoption{rotate}
 \usepackage{tikz}
@@
 
+\newcommand{\slogan}[1]{%
+  \begin{center}%
+    \setlength{\fboxrule}{0pt}%
+    \setlength{\fboxsep}{-14pt}%
+    {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal
+    text}\parbox{0.9\textwidth}{\begin{center}#1\end{center}}}}%
+  \end{center}%
+}
+
 \setbeamertemplate{frametitle}[default][colsep=-2bp,rounded=false,shadow=false,center]
@@
 \setbeamertemplate{navigation symbols}{}
-
-\newcommand{\floatbox}[3]{%
-  \raisebox{0pt}[0pt][0pt]{%
-    \begin{picture}(0,0)(#1,#2)#3\end{picture}\leavevmode%
-  }%
-}
 
@@
 
-\note{\justifying\fontsize{8pt}{9.6}\selectfont
-  \begin{center}\large\textbf{Abstract}\end{center}
-
-  \begin{changemargin}{2.5em}{2.5em}
-    We describe how the internal language of certain
-    toposes, the associated petit and gros Zariski toposes of a scheme, can be
-    used to give simpler definitions and more conceptual proofs of the basic
-    notions and observations in algebraic geometry.
-    % This is useful for studying schemes from a local or relative point of view.
-
-    The starting point is that, from the internal point of view, sheaves of rings
-    and sheaves of modules look just like plain rings and plain modules.
-    In this way, some concepts and statements of scheme theory can be reduced to
-    concepts and statements of intuitionistic linear algebra.
-
-    Furthermore, modal operators can be used to model phrases such as ``on a
-    dense open subset it holds that'' or ``on an open neighbourhood of a given
-    point it holds that''. These operators define certain subtoposes; a
-    generalization of the double-negation translation is useful in order to
-    understand the internal universe of those subtoposes from the internal point
-    of view of the ambient topos.
-
-    A particularly interesting task is to internalize the
-    construction of the relative spectrum, which, given a quasicoherent sheaf of algebras
-    on a scheme~$X$, yields a scheme over~$X$. From the internal point of
-    view, this construction should simply reduce to an intuitionistically sensible
-    variant of the ordinary construction of the spectrum of a ring, but it turns
-    out that this expectation is too naive and that a refined approach is
-    necessary.
-  \end{changemargin}
-}
-
 \begin{frame}[plain,c]
@@
 
-\begin{frame}\frametitle{The internal language of a topos}
-  Let~$\E$ be a topos. Then we can define the meaning of
-  \[ \hil{$\E \models \varphi \qquad\textnormal{(``$\varphi$ holds in $\E$'')}$} \]
-  for formulas~$\varphi$ over~$\E$ using the \hil{Kripke--Joyal semantics}.
+\begin{frame}\frametitle{Exploiting the internal language}
+  A \hil{scheme} is a locally ringed space~$(X,\O_X)$ which is
+  locally isomorphic to the \hil{spectrum of a commutative ring}:
+  \[ \Spec A \defeq \{ \ppp \subseteq A \,|\, \text{$\ppp$ is a prime ideal} \} \]
+  \pause
+  The topos~$\Sh(X)$ is the \hil{petit Zariski topos} of~$X$.
 
@@
       \toprule
-      externally & internally \\
+      externally & internally to $\Sh(X)$ \\
       \midrule
-      object & set/type \\
-      morphism & map of sets \\
+      sheaf of sets & set/type \\
+      morphism of sheaves & map of sets \\
       monomorphism & injective map \\
       epimorphism & surjective map \\
+      sheaf of rings & ring \\
+      sheaf of modules & module \\
       \bottomrule
@@
   \end{center}
-  \pause
-
-  \mbox{If~$\varphi$ implies~$\psi$ \hil{intuitionistically}, then~$\E \models \varphi$
-  implies~$\E \models \psi$.}
 \end{frame}
 
-\note{
-  Let~$X$ be a topological space (or locale) and let~$\alpha : \F \to \G$ be a
-  morphism of sheaves on~$X$. Then:
-  \allowdisplaybreaks
-  \begin{align*}
-    & \Sh(X) \models \speak{$\alpha$ is injective} \\[0.5em]
-    \Longleftrightarrow\
-    & \Sh(X) \models \forall s\?\F\_ \forall t\?\F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
-    \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$, sections $s \in \F(U)$:} \\
-    &\qquad
-      \text{for all open~$V \subseteq U$, sections $t \in \F(V)$:} \\
-    &\qquad\qquad
-        \text{for all open~$W \subseteq V$:} \\
-    &\qquad\qquad\qquad
-          \text{$\alpha_W(s|_W) = \alpha_W(t|_W)$ implies $s|_W = t|_W$} \\[0.5em]
-    \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
-    &\qquad
-          \text{$\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.5em]
-    \Longleftrightarrow\ &
-      \text{$\alpha$ is a monomorphism of sheaves}
-  \end{align*}
-}
-
 \begin{frame}\frametitle{Building a dictionary}
-  A \hil{scheme} is a locally ringed space~$(X,\O_X)$ which is
-  locally isomorphic to the \hil{spectrum of a commutative ring}:
-  \[ \Spec A \defeq \{ \ppp \subseteq A \,|\, \text{$\ppp$ is a prime ideal} \} \]
-  \pause
-
+  \slogan{\hil{Understand notions of algebraic geometry
+  as notions of algebra internal to~$\boldsymbol{\Sh(X)}$.}}
   \begin{center}
@@
       \toprule
-      externally to $\Sh(X)$ & internally \\
+      externally & internally to $\Sh(X)$ \\
+      \midrule
+      sheaf of sets & set/type \\
+      morphism of sheaves & map of sets \\
+      monomorphism & injective map \\
+      epimorphism & surjective map \\
       \midrule
@@
       sheaf of finite type & finitely generated module \\
-      finite locally free sheaf & finite free module \\
+      locally free sheaf & free module \\
+      tensor product of sheaves & tensor product of modules \\
+      rank function & minimal number of generators \\
+      sheaf of rational functions & total quotient ring of~$\O_X$ \\
       \bottomrule
@@
 
-  \hil{Motto:} \tabto{1.41cm} We can understand notions of algebraic geometry \\
-  \tabto{1.41cm} as notions of linear algebra internal to~$\Sh(X)$.
+  \visible<2>{\begin{tikzpicture}[overlay]
+    \draw[fill=white, draw=white, opacity=0.9] (0,0) rectangle (\paperwidth,7.1);
+    \node[anchor=south west,inner sep=0] (image) at (1.8,1.3) {
+      \includegraphics[width=0.7\textwidth]{images/steven-kleiman-misconceptions-about-kx}
+    };
+  \end{tikzpicture}}
 \end{frame}
 
-\begin{frame}[c]\frametitle{Using the dictionary}
+\begin{frame}[c]\frametitle{Praise for Mike Shulman}
+  \centering
+  \includegraphics[scale=0.4]{images/mike-shulman-stack-semantics}
+  \par
+\end{frame}
+
+\begin{frame}\frametitle{Using the dictionary}
   \begin{center}
@@
   \begin{center}
-    \visible<2->{
     \begin{minipage}{0.70\textwidth}
@@
     \scalebox{3}{$\Downarrow$}
-    }
 
@@
   \end{center}
-  \pause
-  \pause
-
-  \hil{Motto:} \tabto{1.41cm} \mbox{We can understand statements of algebraic geometry} \\
-  \tabto{1.41cm} as statements of linear algebra internal to~$\Sh(X)$.
 \end{frame}
 
-\begin{frame}\frametitle{Exploring the internal world}
+\begin{frame}\frametitle{A curious property}
   Let~$X$ be a scheme. Internally to~$\Sh(X)$,
-  \begin{itemize}
-    \item any non-invertible element of~$\O_X$ is nilpotent,
-    \item $\K_X$ is simply the total quotient ring of~$\O_X$,
-    \item $\dim \O_X \leq n$ holds if and only if~$\dim X \leq n$.
-  \end{itemize}
+  \begin{center}
+    \hil{any non-invertible element of~$\boldsymbol{\O_X}$ is nilpotent.}
+  \end{center}
 
@@
 
-\section{\texorpdfstring{The $\Diamond$-translation}{The ♦-translation}}  % XXX
+\section{\texorpdfstring{The $\Diamond$-translation}{The ♢-translation}}
 
@@
 
-  Is there a \hil{free local and local-over-$B$ ring}~$A \to A'$ over~$A$?
+  Is there a \hil{free local and local-over-$\boldsymbol{B}$ ring}~$A \to A'$ over~$A$?
 
   \[ \xymatrix{
-    B \ar[r]\ar@/^1.5pc/[rrrr]\ar@/_/[rrd] & A \ar[rd] \ar[rrr] &&& R \\
-    && A' \ar@{-->}[rru]
+    B \ar[r]\ar@/^2pc/[rrrr]^{\text{local}}\ar@/_/[rrd]_[@!-33]{\text{local}} &
+      A \ar[rd] \ar[rrr] &&&
+      {\substack{\text{local}\\\text{\normalsize$R$}\\\phantom{\text{local}}}} \\
+    && {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}[rru]_[@!35]{\text{local}}
   } \]
+
+  \medskip
+  Form limits in the category of \hil{locally ringed locales}
+  by \hil{relocalizing} the corresponding limit in ringed locales.
 \end{frame}
-% Better decoration (local rings, local morphisms)
+
+\begin{frame}
+  \slogan{\hil{Understand notions and statements of algebraic geometry
+  as notions and statements of algebra internal to appropriate toposes.}}
+
+  {\vspace{-0.5em}\centering
+  \includegraphics[scale=0.3]{images/external-internal-small}
+  \par\medskip}
+
+  \begin{itemize}
+    \item Simplify routine proofs.
+    \item Gain conceptual clarity.
+    \item Develop a synthetic account of scheme theory.
+  \end{itemize}
+\end{frame}
+% Fun facts:
+% Take O_X, pullback along X_{negneg} --> X, pushforward: result is K_X.
+% Very naive definition of P^n works internal to the big Zariski topos.
+
+\backupstart
+
+\begin{frame}[plain,c]
+  \centering
+  \includegraphics[scale=0.5]{images/sheafification-man}
+  \sffamily
+
+  You should totally look up:
+
+  \hil{The Adventures of Sheafification Man}
+  \par
+\end{frame}
+
+\begin{frame}\frametitle{Translating internal statements}
+  Let~$X$ be a topological space (or locale) and let~$\alpha : \F \to \G$ be a
+  morphism of sheaves on~$X$. Then:
+  \allowdisplaybreaks
+  \begin{align*}
+    & \Sh(X) \models \speak{$\alpha$ is injective} \\[0.5em]
+    \Longleftrightarrow\
+    & \Sh(X) \models \forall s\?\F\_ \forall t\?\F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{for all open~$U \subseteq X$, sections $s \in \F(U)$:} \\
+    &\qquad
+      \text{for all open~$V \subseteq U$, sections $t \in \F(V)$:} \\
+    &\qquad\qquad
+        \text{for all open~$W \subseteq V$:} \\
+    &\qquad\qquad\qquad
+          \text{$\alpha_W(s|_W) = \alpha_W(t|_W)$ implies $s|_W = t|_W$} \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
+    &\qquad
+          \text{$\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{$\alpha$ is a monomorphism of sheaves}
+  \end{align*}
+\end{frame}
+
+\backupend
+
+\end{document}
+
+1. Tell story; begin with MO question
+2. Basics (include praise for Mike Shulman)
+3. Curious field property and more advanced examples
+4. Relative spectrum (also mention equivalence to Hakim's spectrum in dimension zero)
+5. Outlook: synthetic algebraic geometry, relative geometry, big Zariski topos
+
+\note{\justifying\fontsize{8pt}{9.6}\selectfont
+  \begin{center}\large\textbf{Abstract}\end{center}
+
+  \begin{changemargin}{2.5em}{2.5em}
+    We describe how the internal language of certain
+    toposes, the associated petit and gros Zariski toposes of a scheme, can be
+    used to give simpler definitions and more conceptual proofs of the basic
+    notions and observations in algebraic geometry.
+    % This is useful for studying schemes from a local or relative point of view.
+
+    The starting point is that, from the internal point of view, sheaves of rings
+    and sheaves of modules look just like plain rings and plain modules.
+    In this way, some concepts and statements of scheme theory can be reduced to
+    concepts and statements of intuitionistic linear algebra.
+
+    Furthermore, modal operators can be used to model phrases such as ``on a
+    dense open subset it holds that'' or ``on an open neighbourhood of a given
+    point it holds that''. These operators define certain subtoposes; a
+    generalization of the double-negation translation is useful in order to
+    understand the internal universe of those subtoposes from the internal point
+    of view of the ambient topos.
+
+    A particularly interesting task is to internalize the
+    construction of the relative spectrum, which, given a quasicoherent sheaf of algebras
+    on a scheme~$X$, yields a scheme over~$X$. From the internal point of
+    view, this construction should simply reduce to an intuitionistically sensible
+    variant of the ordinary construction of the spectrum of a ring, but it turns
+    out that this expectation is too naive and that a refined approach is
+    necessary.
+  \end{changemargin}
+}
+
+  \scriptsize
+  \begin{itemize}
+    \item $\Sh(X) \models f = g$ iff~$f = g \in \F(X)$.
+    \item $\Sh(X) \models f = g \vee f = h$ iff on a cover~$X = \bigcup_i U_i$,
+    $f|_{U_i} = g|_{U_i}$ or~$f|_{U_i} = h|_{U_i}$.
+    \item $\Sh(X) \models \neg\neg(f = g)$ iff~$f = g$ on a dense open subset
+    of~$X$.
+  \end{itemize}
+
+\begin{frame}\frametitle{The internal language of a topos}
+  Let~$\E$ be a topos. Then we can define the meaning of
+  \[ \hil{$\E \models \varphi \qquad\textnormal{(``$\varphi$ holds in $\E$'')}$} \]
+  for formulas~$\varphi$ over~$\E$ using the \hil{Kripke--Joyal semantics}.
+
+  \begin{center}
+    \small
+    \begin{tabular}{ll}
+      \toprule
+      externally & internally \\
+      \midrule
+      object & set/type \\
+      morphism & map of sets \\
+      monomorphism & injective map \\
+      epimorphism & surjective map \\
+      \bottomrule
+    \end{tabular}
+  \end{center}
+  \pause
+
+  \mbox{If~$\varphi$ implies~$\psi$ \hil{intuitionistically}, then~$\E \models \varphi$
+  implies~$\E \models \psi$.}
+\end{frame}
 
@@
 % Extend to toposes.
-
-\end{document}
-
-1. Tell story; begin with MO question
-2. Basics (include praise for Mike Shulman)
-3. Curious field property and more advanced examples
-4. Relative spectrum (also mention equivalence to Hakim's spectrum in dimension zero)
-5. Outlook: synthetic algebraic geometry, relative geometry, big Zariski topos
-
-  \scriptsize
-  \begin{itemize}
-    \item $\Sh(X) \models f = g$ iff~$f = g \in \F(X)$.
-    \item $\Sh(X) \models f = g \vee f = h$ iff on a cover~$X = \bigcup_i U_i$,
-    $f|_{U_i} = g|_{U_i}$ or~$f|_{U_i} = h|_{U_i}$.
-    \item $\Sh(X) \models \neg\neg(f = g)$ iff~$f = g$ on a dense open subset
-    of~$X$.
-  \end{itemize}

### 2015-11-26 slides-ihes2015.tex
Credit where credit is due
Slide on the gros Zariski topos

@@
 
+\backupstart
+
 \begin{frame}
@@
   {\vspace{-0.5em}\centering
+  \rotatebox{90}{\tiny\scalebox{0.5}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
   \includegraphics[scale=0.3]{images/external-internal-small}
@@
   \begin{itemize}
-    \item Simplify routine proofs.
+    \item Simplify proofs.
     \item Gain conceptual clarity.
@@
 % Very naive definition of P^n works internal to the big Zariski topos.
-
-\backupstart
 
@@
   \par
+\end{frame}
+
+\begin{frame}\frametitle{The gros Zariski topos}
+  Let~$X$ be a scheme. The \hil{gros Zariski topos} is the topos of sheaves
+  on~$\Sch/X$ with respect to the Zariski topology. From its point of view,
+  \ldots
+
+  \begin{itemize}
+    \item \ldots{} $X$-schemes look just like sets,
+    \item \ldots{} $\mathbb{P}^n_X$ is given by the naive expression
+    \[ \{ (x_0,\ldots,x_n) \,|\, x_1 \neq 0 \vee \cdots \vee x_n \neq 0
+    \}/\text{(rescaling)}, \]
+    \item \ldots{} affinity is a ``double dual condition'', and
+    \item \ldots{} the étale topology is the coarsest topology~$\Diamond$ such
+    that
+    \[ \forall f : \mathbb{A}^1_X[T]\_
+      \text{$f$ is separable} \Rightarrow
+      \Diamond(\exists t : \mathbb{A}^1\_ f(t) = 0). \]
+  \end{itemize}
 \end{frame}

### 2015-11-27 slides-ihes2015.tex
Final version of the slides

@@
 
+\newcounter{framenumberpreappendix}
 \newcommand{\backupstart}{
-  \newcounter{framenumberpreappendix}
   \setcounter{framenumberpreappendix}{\value{framenumber}}
@@
 
+\backupstart
 \frame[t]{\frametitle{Outline}\scriptsize\begin{itemize}\item[]\tableofcontents\end{itemize}}
@@
 \end{frame}
+\backupend
 
-\section{The internal language}
+\section{Basic applications of the internal language}
 
@@
     \small
-    \begin{tabular}{ll}
+    \scalebox{0.93}{\begin{tabular}{ll}
       \toprule
@@
       sheaf of finite type & finitely generated module \\
-      locally free sheaf & free module \\
+      finite locally free sheaf & finite free module \\
+      coherent sheaf & coherent module \\
       tensor product of sheaves & tensor product of modules \\
@@
       \bottomrule
-    \end{tabular}
+    \end{tabular}}
   \end{center}
@@
         Let $0 \to \F' \to \F \to \F'' \to 0$ be a short exact sequence
-        of~$\O_X$-modules. If~$\F'$ and~$\F''$ are locally of finite type, so
+        of~$\O_X$-modules. If~$\F'$ and~$\F''$ are of finite type, so
         is~$\F$.
@@
       \end{exampleblock}
+      \centering
+      \tiny Ravi Vakil: ``Important hard exercise'' (13.7.K).
+      \par
     \end{minipage}
@@
       \item \ldots{} a dense open subset.
+      \item \ldots{} a schematically dense open subset.
       \item \ldots{} a given open subset~$U$.
@@
 \begin{frame}\frametitle{Quasicoherence}
-  Let~$X$ be a scheme. Let~$\F$ be an~$\O_X$-module.
+  Let~$X$ be a scheme. Let~$\E$ be an~$\O_X$-module.
 
-  Then~$\F$ is quasicoherent
+  Then~$\E$ is quasicoherent
   if and only if, internally to~$\Sh(X)$,
-  \begin{quote}\textnormal{$\F[f^{-1}]$ is a $\Diamond_f$-sheaf for any~$f : \O_X$, \\[0.3em]
+  \begin{quote}\textnormal{$\E[f^{-1}]$ is a $\Diamond_f$-sheaf for any~$f : \O_X$, \\[0.3em]
   \qquad\qquad where~$\Diamond_f\varphi \defeqv (\text{$f$ invertible} \Rightarrow \varphi)$.}
@@
 
-  In particular: If~$\F$ is quasicoherent, then internally
+  In particular: If~$\E$ is quasicoherent, then internally
   \[ (\text{$f$ invertible} \Rightarrow s = 0) \Longrightarrow
@@
   \vspace*{-1.5em}\par%
-  for any~$f : \O_X$ and~$s : \F$.
+  for any~$f : \O_X$ and~$s : \E$.
 \end{frame}
@@
 \newcommand{\defspeca}{\text{topological space of the prime ideals of $A$}}
-\newcommand{\defspecb}{\text{topological space of the filters of $A$}}
-\newcommand{\defspecc}{\text{locale of the filters of $A$}}
+\newcommand{\defspecb}{\text{topological space of the prime filters of $A$}}
+\newcommand{\defspecc}{\text{locale of the prime filters of $A$}}
 
@@
   Let~$X$ be a scheme and~$\O_X \xrightarrow{\varphi} \A$ be a quasicoherent algebra.
-  Can we describe~$\RelSpec_X \A$, a scheme over~$X$, internally?
+  Can we describe~\hil{$\boldsymbol{\RelSpec_X \A}$}, a scheme over~$X$, internally?
 
@@
 
-  Solution: Define internally the frame of~$\RelSpec_X \A$ to be the frame of
+  \hil{Solution:} Define internally the frame of~$\RelSpec_X \A$ to be the frame of
   those radical ideals~$I \subseteq \A$ such that
@@
   \pause
-  Its points are those filters~$\fff$ of~$\A$ such that
-  \[ \forall f\?\O_X\_ \varphi(f) \in \fff \Rightarrow \text{$f$ invertible in~$\O_X$}. \]
+  Its \hil{points} are those prime filters~$G$ of~$\A$ such that
+  \[ \forall f\?\O_X\_ \varphi(f) \in G \Rightarrow \text{$f$ invertible in~$\O_X$}. \]
 \end{frame}
@@
   \rotatebox{90}{\tiny\scalebox{0.5}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
-  \includegraphics[scale=0.3]{images/external-internal-small}
+  \includegraphics[scale=0.23]{images/external-internal-small}
   \par\medskip}
@@
     \item Simplify proofs.
-    \item Gain conceptual clarity.
+    \item Gain conceptual understanding.
     \item Develop a synthetic account of scheme theory.
   \end{itemize}
+
+  \centering
+  \hil{\href{http://tiny.cc/topos-notes}{http://tiny.cc/topos-notes}} \\
+  \begin{minipage}{1.0\textwidth}\justifying
+    \scriptsize
+    spreading of properties,
+    general transfer principles,
+    applications to constructive algebra,
+    quasicoherence,
+    internal Cartier divisors,
+    pullback along immersions $=$ internal sheafification,
+    scheme dimension $=$ internal Krull dimension of~$\O_X$,
+    dense $=$ not not,
+    modal operators,
+    relative spectrum,
+    other toposes,
+    étale topology,
+    group schemes $=$ groups,
+    \ldots
+  \end{minipage}
+  \par
 \end{frame}
-% Fun facts:
-% Take O_X, pullback along X_{negneg} --> X, pushforward: result is K_X.
+% Fun fact:
 % Very naive definition of P^n works internal to the big Zariski topos.
@@
 
+\begin{frame}\frametitle{Spreading from points to neighbourhoods}
+  All of the following lemmas have a short, sometimes trivial proof.
+  Let~$\F$ be a sheaf of finite type on a ringed space~$X$.
+  Let~$x \in X$. Let~$A \subseteq X$ be a closed subset. Then:
+  \small
+  \begin{itemize}
+    \item $\F_x = 0$ iff~$\F|_U = 0$ for some open neighbourhood of~$x$.
+    \item $\F|_A = 0$ iff~$\F|_U = 0$ for some open set containing~$A$.
+    \item $\F_x$ can be generated by~$n$ elements iff this is true on some open
+    neighbourhood of~$x$.
+%   \item $\alpha_x$ is surjective iff~$\alpha$ is an epimorphism on some open
+%   set containing~$x$, where~$\alpha : \G \to \F$ is any morphism.
+    \item $\mathcal{H}\mathrm{om}_{\O_X}(\F,\G)_x \cong
+    \Hom_{\O_{X,x}}(\F_x,\G_x)$ if~$\F$ is of finite presentation around~$x$.
+    \item $\F$ is torsion iff~$\F_\xi$ vanishes (assume~$X$ integral and~$\F$ quasicoherent).
+    \item $\F$ is torsion iff~$\F|_{\mathrm{Ass}(\O_X)}$ vanishes (assume~$X$
+    locally Noetherian and~$\F$ quasicoherent).
+  \end{itemize}
+\end{frame}
+
+\begin{frame}\frametitle{The smallest dense sublocale}
+  Let~$X$ be a reduced scheme satisfying a technical condition.
+  Let~$i : X_{\neg\neg} \to X$ be the inclusion of the smallest dense sublocale
+  of~$X$.
+
+  Then~$i_* i^{-1} \O_X \cong \K_X$.
+
+%  \begin{itemize}
+%    \item This is a highbrow way of saying ``rational functions are regular
+%    functions which are defined on a dense open subset''.
+%    \item There is a generalization to nonreduced schemes.
+%  \end{itemize}
+\end{frame}
+
+\begin{frame}\frametitle{Transfer principles}
+  Let~$M$ be an~$A$-module. How do~$M$ and the sheaf~$M^{\sim}$ on~$\Spec A$
+  relate?
+
+  Observe that $M^{\sim} \cong \underline{M}[F^{-1}]$ is the localization of~$M$ at
+  the \hil{generic filter}. Therefore:
+  \begin{center}
+    $M^{\sim}$ inherits all those properties of~$M$ which are \\
+    \hil{stable under localization}.
+  \end{center}
+  Examples: finitely generated, free, flat, \ldots
+
+  A converse holds as well, suitably formulated.
+\end{frame}
+
 \begin{frame}\frametitle{The gros Zariski topos}
@@
     \item \ldots{} affinity is a ``double dual condition'', and
-    \item \ldots{} the étale topology is the coarsest topology~$\Diamond$ such
-    that
-    \[ \forall f : \mathbb{A}^1_X[T]\_
-      \text{$f$ is separable} \Rightarrow
+    \item \ldots{} the étale topology is the coarsest topology~$\Diamond$ s.\,th.
+    \[ \forall f : \mathbb{A}^1_X[T]\_\
+      \text{$f$ is monic separable} \Rightarrow
       \Diamond(\exists t : \mathbb{A}^1\_ f(t) = 0). \]

### 2015-11-27 slides-ihes2015.tex
List IHES slides in README.md

@@
 
-  {\vspace{-0.5em}\centering
+  {\vspace{-0.1em}\centering
   \rotatebox{90}{\tiny\scalebox{0.5}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
-  \includegraphics[scale=0.23]{images/external-internal-small}
-  \par\medskip}
+  \includegraphics[scale=0.20]{images/external-internal-small}
+  \par\medskip\vspace{-0.1em}}
 
   \begin{itemize}
-    \item Simplify proofs.
-    \item Gain conceptual understanding.
+    \item Simplify proofs and gain conceptual understanding.
+    \item Understand relative geometry as absolute geometry.
     \item Develop a synthetic account of scheme theory.
+    \item Contribute to constructive algebra.
   \end{itemize}
@@
   A converse holds as well, suitably formulated.
+\end{frame}
+
+\begin{frame}\frametitle{Applications in algebra}
+  Let~$A$ be a commutative ring.
+  The internal language of~$\Sh(\Spec A)$ allows you to say ``without loss of
+  generality, we may assume that~$A$ is local'', even constructively.
+
+  \begin{center}
+    \begin{minipage}{0.75\textwidth}
+      \begin{exampleblock}{}
+        \justifying
+        The kernel of any matrix over a principial ideal domain is finitely
+        generated.
+      \end{exampleblock}
+    \end{minipage}
+    \medskip
+
+    \scalebox{3}{$\Downarrow$}
+
+    \begin{minipage}{0.75\textwidth}
+      \begin{exampleblock}{}
+        \justifying
+        The kernel of any matrix over a Prüfer domain is finitely generated.
+      \end{exampleblock}
+    \end{minipage}
+  \end{center}
 \end{frame}

### 2015-11-29 slides-ihes2015.tex
Begin of note-slides

@@
 \newcommand{\K}{\mathcal{K}}
+\newcommand{\NN}{\mathbb{N}}
 \newcommand{\ppp}{\mathfrak{p}}
@@
 \end{frame}
+
+\note{\justifying
+  The internal language of a topos supports
+  \begin{itemize}
+  \item first-order logic,
+  \item higher-order logic (for instance quantification over subsets),
+  \item dependent types, and
+  \item unbounded quantification.
+  \end{itemize}
+
+  The first three items are standard. The fourth is due to Mike Shulman.
+  Combined, it's possible to interpret ``essentially all of constructive
+  mathematics'' internal to a topos.
+
+  Restrictions persist for operations with a ``set-theoretical flavor'' like
+  building an infinite union of iterated powersets, for example~$\bigcup_{n \in
+  \NN} P^n(\NN)$.
+  \par
+}
 
@@
     \}/\text{(rescaling)}, \]
+    \item \ldots{} the cotangent ``bundle'' of an~$X$-scheme~$T$ is
+    \[ \text{the set of maps $\Delta \to T$,} \]
+    where~$\Delta = \{ \varepsilon \in \mathbb{A}^1_X \,|\, \varepsilon^2 = 0
+    \}$.
     \item \ldots{} affinity is a ``double dual condition'', and
@@
 \end{frame}
+
+\note{\justifying
+  \begin{itemize}
+    \item The functor of points of~$\mathbb{A}^1_X$, that is
+    \[ (T/X) \longmapsto \O_T(T), \]
+    looks like a local ring and indeed like a field from the internal point of
+    view, in the sense that
+    \[ \forall f\?\mathbb{A}^1_X\_ \neg(f = 0) \Rightarrow \text{$f$
+    invertible}. \]
+    \item Describing the functor of points of the projective space was
+    suggested by Zhen Lin Low.
+    \item The statement on the étale topology follows from Gavin Wraith's
+    article \emph{Generic Galois theory of local rings}.
+  \end{itemize}
+  \par
+}
 

### 2015-11-30 slides-ihes2015.tex
More note-slides

@@
 \usepackage{mathtools}
-\usepackage{soul}\setul{0.2ex}{}
+\usepackage{soul}\setul{0.3ex}{}
 \usepackage[all]{xy}
@@
 \newcommand{\A}{\mathcal{A}}
+\renewcommand{\AA}{\mathbb{A}}
 \newcommand{\E}{\mathcal{E}}
@@
 \newcommand{\NN}{\mathbb{N}}
+\newcommand{\ZZ}{\mathbb{Z}}
 \newcommand{\ppp}{\mathfrak{p}}
-\newcommand{\fff}{\mathfrak{f}}
 \newcommand{\defeq}{\vcentcolon=}
@@
 \newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
+\newcommand{\ull}[1]{\underline{#1}}
 
@@
 
+\note{\justifying\fontsize{8pt}{9.6}\selectfont
+  \begin{center}\large\textbf{Abstract}\end{center}
+
+  \begin{changemargin}{2.5em}{2.5em}
+    We describe how the internal language of certain
+    toposes, the associated petit and gros Zariski toposes of a scheme, can be
+    used to give simpler definitions and more conceptual proofs of the basic
+    notions and observations in algebraic geometry.
+    % This is useful for studying schemes from a local or relative point of view.
+
+    The starting point is that, from the internal point of view, sheaves of rings
+    and sheaves of modules look just like plain rings and plain modules.
+    In this way, some concepts and statements of scheme theory can be reduced to
+    concepts and statements of intuitionistic linear algebra.
+
+    Furthermore, modal operators can be used to model phrases such as ``on a
+    dense open subset it holds that'' or ``on an open neighbourhood of a given
+    point it holds that''. These operators define certain subtoposes; a
+    generalization of the double-negation translation is useful in order to
+    understand the internal universe of those subtoposes from the internal point
+    of view of the ambient topos.
+
+    A particularly interesting task is to internalize the
+    construction of the relative spectrum, which, given a quasicoherent sheaf of algebras
+    on a scheme~$X$, yields a scheme over~$X$. From the internal point of
+    view, this construction should simply reduce to an intuitionistically sensible
+    variant of the ordinary construction of the spectrum of a ring, but it turns
+    out that this expectation is too naive and that a refined approach is
+    necessary.
+  \end{changemargin}
+}
+
 \begin{frame}[plain,c]
@@
 \end{frame}
+
+\note{\justifying
+  See the \href{https://github.com/iblech/internal-methods/raw/master/notes.pdf}{notes} for more dictionary entries.
+
+  The simple definition of~$\K_X$ allows to give an internal account of the
+  basics of the theory of Cartier divisors, for instance giving an easy
+  description of the line bundle associated to a Cartier divisor.
+  \par
+}
 
@@
 
+\note{\justifying
+  The~$\Diamond$-translation is a generalization of the \emph{double negation
+  translation}, which is well-known in logic. The double negation translation
+  has the following curious property: A formula~$\varphi$ admits a classical
+  proof if and only if the translated formula~$\varphi^{\neg\neg}$ admits an
+  intuitionistic proof.
+
+  The~$\Diamond$-translation has been studied before (see for instance Aczel:
+  \emph{The Russell--Prawitz modality}, and Escardó, Oliva: \emph{The Peirce
+  translation and the double negation shift}), but to the best of my know\-ledge,
+  this application -- expressing the internal language of
+  subtoposes in the internal language of the ambient topos -- is new.
+  \par
+}
+
+\note{\justifying
+  For ease of exposition, assume that~$X$ is irreducible with generic
+  point~$\xi$. Let~$\Diamond \defeqv \neg\neg$.
+
+  Then~$\Sh(X) \models \Diamond \varphi$ means that~$\varphi$ holds on a dense
+  open subset of~$X$, while~$\Sh(X) \models \varphi^\Diamond$ means
+  that~$\varphi$ holds at the generic point (taking stalks of all involved
+  sheaves).
+
+  The question~``does~$\varphi^\Diamond$ imply~$\Diamond\varphi$?'' therefore
+  means: Does~$\varphi$ spread from the generic point to a dense open subset?
+
+  For the special case of the double negation translation, a general answer to
+  this purely logical question has long been known: This holds if~$\varphi$ is
+  a \emph{geometric formula} (doesn't contain~$\Rightarrow$ and~$\forall$).
+  \par
+}
+
+\note{\justifying
+  Let~$\F$ be a sheaf of modules on a locally ringed space~$X$.
+  Assume that the stalk~$\F_x$ at some point~$x \in X$ vanishes.
+  Then in general it does \emph{not} follow that~$\F$ vanishes on some open
+  neighbourhood of~$x$.
+
+  This can be understood in logical terms: The statement that~$\F$ vanishes,
+  \[ \forall s : \F\_\ s = 0, \]
+  is not a geometric formula.
+
+  However, if~$\F$ is additionally supposed to be of finite type, then it
+  \emph{does} follow that~$\F$ vanishes on an open neighbourhood. This too can
+  be understood in logical terms: If~$\F$ is of finite type, then internally
+  there are generators~$s_1,\ldots,s_n$ of~$\F$. Thus the vanishing of~$\F$ can
+  be reformulated as
+  \[ s_1 = 0 \wedge \cdots \wedge s_n = 0, \]
+  and this condition is manifestly geometric.
+  \par
+}
+
 \section{Quasicoherence of sheaves of modules}
@@
 
+\note{\justifying
+  The sheaf condition and the sheafification functor can be described purely
+  internally. An object~$M$ is \emph{separated} with respect to~$\Diamond$ if
+  and only if, from the internal point of view,
+  \[ \forall x,y : M\_\ \Diamond(x = y) \Rightarrow x = y. \]
+  It is a \emph{sheaf} with respect to~$\Diamond$, if furthermore
+  \[ \forall K \subseteq M\_\
+    \Diamond(\exists x : M\_\ K = \{ x \}) \Longrightarrow
+    \exists x : M\_\ \Diamond(x \in K). \]
+
+  The second condition displayed on the previous slide is equivalent to the
+  separatedness condition. In the special case~$\E = \O_X$, $s = 1$ it reduces
+  to Mulvey's ``somewhat obscure formula''. We now understand this condition in
+  its proper context.
+  \par
+}
+
 \section{The relative and internal spectrum}
@@
   \[ \xymatrix{
-    A \ar[rd] \ar[rrr] &&& R \\
-    & A' \ar@{-->}[rru]
+    A \ar[rd] \ar[rrr] &&& {\substack{\text{local}\\\text{\normalsize$R$}\\\phantom{\text{local}}}} \\
+    & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!34]{\text{local}}[rru]
   } \]
@@
 \end{frame}
+
+\note{\justifying
+  Details on this point of view can be found in one of Peter Arndt's very nice
+  answers on MathOverflow:
+  \begin{center}\url{http://mathoverflow.net/a/14334/31233}\end{center}
+  \par
+}
 
@@
 
+\note{\justifying
+  Monique Hakim constructed in her thesis a very general spectrum functor,
+  taking a ringed topos to a locally ringed one, using explicit calculations
+  with sites.
+
+  Using the internal language allows to reduce these calculations to a minimum.
+  One constructs the spectrum as the sheaf topos over an internal
+  locale and then uses the general theorem that toposes over the base~$\E$ are
+  the same as toposes internal to~$\E$.
+
+  As a byproduct one obtains that Hakim's spectrum is \emph{localic} over the base.
+  \par
+}
+
 \begin{frame}\frametitle{The relative spectrum}
@@
   Its \hil{points} are those prime filters~$G$ of~$\A$ such that
-  \[ \forall f\?\O_X\_ \varphi(f) \in G \Rightarrow \text{$f$ invertible in~$\O_X$}. \]
+  \[ \forall f\?\O_X\_ \varphi(f) \in G \Longrightarrow \text{$f$ invertible in~$\O_X$}. \]
 \end{frame}
+
+\note{\justifying
+  The stated condition on~$I$ is, under the assumption that~$\A$ is
+  quasicoherent, equivalent to the condition that~$I$ is quasicoherent (as
+  an~$\O_X$-module).
+
+  The relative spectrum is thus constructed as a certain sublocale of the
+  absolute one. The two constructions coincide if and only if the dimension of
+  the base scheme is~$\leq 0$.
+  \par
+}
 
@@
 \end{frame}
+
+\note{\justifying
+  One might wonder whether the absolute spectrum or the relative one is ``more
+  fundamental''. The absolute spectrum can be expressed using the relative one,
+  since
+  \[ \Spec A = \RelSpec_{\Spec \ZZ} A^{\sim}, \]
+  but the other way is not in general possible: The absolute spectrum is always
+  (quasi-)compact, while the relative one is not in general.
+  \par
+}
+% XXX: Give details.
 
@@
 
+\appendix
+
+\section{Spreading from points to neighbourhoods}
+
 \begin{frame}\frametitle{Spreading from points to neighbourhoods}
@@
 
+\section{The meromorphic functions revisited}
+
 \begin{frame}\frametitle{The smallest dense sublocale}
@@
 
-%  \begin{itemize}
-%    \item This is a highbrow way of saying ``rational functions are regular
-%    functions which are defined on a dense open subset''.
-%    \item There is a generalization to nonreduced schemes.
-%  \end{itemize}
+  \begin{itemize}
+    \item This is a highbrow way of saying ``rational functions are regular
+    functions which are defined on a dense open subset''.
+    \item Another reformulation is that~$\K_X$ is the sheafification of~$\O_X$
+    with respect to the~$\neg\neg$-modality.
+    \item There is a generalization to nonreduced schemes.
+  \end{itemize}
 \end{frame}
+
+\section{Transfer principles}
 
@@
 
-  Observe that $M^{\sim} \cong \underline{M}[F^{-1}]$ is the localization of~$M$ at
-  the \hil{generic filter}. Therefore:
+  Observe that $M^{\sim} \cong \underline{M}[\F^{-1}]$ is the localization of~$M$ at
+  the \hil{generic prime filter} and that~$M$ and the constant sheaf of
+  modules~$\underline{M}$ share all first-order properties. Therefore:
   \begin{center}
@@
 \end{frame}
+
+\section{Applications in algebra}
 
@@
 
+\begin{frame}\frametitle{Hilbert's program in algebra}
+  \scriptsize\justifying
+  There is a way to combine some of the powerful tools of classical ring theory
+  with the advantages that constructive reasoning provides, for instance
+  exhibiting explicit witnesses. Namely we can devise
+  a language in which we can usefully talk about prime ideals, but which
+  substitutes non-constructive arguments by constructive arguments ``behind
+  the scenes''. The key idea is to substitute the phrase ``for all prime ideals''
+  (or equivalently ``for all prime filters'') by ``for the generic prime filter''.
+
+  More specifically, simply interpret a given proof using prime filters
+  in~$\Sh(\Spec A)$ and let it refer to~$\F \hookrightarrow \underline{A}$.
+
+  \hspace*{-0.75cm}%
+  \begin{tabular}{lll}
+    \toprule
+    Statement & constructive substitution & meaning \\\midrule
+    $x \in \ppp$ for all~$\ppp$. &
+    $x \not\in \F$. &
+    $x$ is nilpotent. \\
+    $x \in \ppp$ for all~$\ppp$ such that~$y \in \ppp$. &
+    $x \in \F \Rightarrow y \in \F$. &
+    $x \in \sqrt{(y)}$. \\
+    $x$ is regular in all stalks~$A_\ppp$. &
+    $x$ is regular in~$\ull{A}[\F^{-1}]$. &
+    $x$ is regular in~$A$. \\
+    The stalks~$A_\ppp$ are reduced. &
+    $\ull{A}[\F^{-1}]$ is reduced. &
+    $A$ is reduced. \\
+    The stalks~$M_\ppp$ vanish. &
+    $\ull{M}[\F^{-1}] = 0$. &
+    $M = 0$. \\
+%   The stalks~$M_\ppp$ are fin.\@ gen.\@ over~$A_\ppp$. &
+%   $\ull{M}[\F^{-1}]$ is fin.\@ gen.\@ over
+%   $\ull{A}[\F^{-1}]$. &
+%   $M$ is fin.\@ gen.\@ over~$A$. \\
+    The stalks~$M_\ppp$ are flat over~$A_\ppp$. &
+    $\ull{M}[\F^{-1}]$ is flat over~$\ull{A}[\F^{-1}]$. &
+    $M$ is flat over~$A$. \\
+    The maps~$M_\ppp \to N_\ppp$ are injective. &
+    $\ull{M}[\F^{-1}] \to \ull{N}[\F^{-1}]$ is injective. &
+    $M \to N$ is injective. \\
+    The maps~$M_\ppp \to N_\ppp$ are surjective. &
+    $\ull{M}[\F^{-1}] \to \ull{N}[\F^{-1}]$ is surjective. &
+    $M \to N$ is surjective. \\
+    \bottomrule
+  \end{tabular}
+
+  This is related (in a few cases equivalent) to the \emph{dynamical methods in
+  algebra} explored by Coquand, Coste, Lombardi, Roy, and others. Their
+  approach applies more generally.
+  \par
+\end{frame}
+
+\section{The gros Zariski topos}
+
 \begin{frame}\frametitle{The gros Zariski topos}
@@
     \item \ldots{} the cotangent ``bundle'' of an~$X$-scheme~$T$ is
-    \[ \text{the set of maps $\Delta \to T$,} \]
-    where~$\Delta = \{ \varepsilon \in \mathbb{A}^1_X \,|\, \varepsilon^2 = 0
+    \[ \text{the set of maps $\Delta \to \ull{T}$,} \]
+    where~$\Delta = \{ \varepsilon \in \ull{\AA}^1_X \,|\, \varepsilon^2 = 0
     \}$.
@@
     \item \ldots{} the étale topology is the coarsest topology~$\Diamond$ s.\,th.
-    \[ \forall f : \mathbb{A}^1_X[T]\_\
+    \[ \forall f : \ull{\AA}^1_X[T]\_\
       \text{$f$ is monic separable} \Rightarrow
-      \Diamond(\exists t : \mathbb{A}^1\_ f(t) = 0). \]
+      \Diamond(\exists t : \ull{\AA}^1\_ f(t) = 0). \]
   \end{itemize}
@@
 
-\note{\justifying
-  \begin{itemize}
-    \item The functor of points of~$\mathbb{A}^1_X$, that is
-    \[ (T/X) \longmapsto \O_T(T), \]
+\note{
+  \begin{itemize}\justifying
+    \item The functor of points of~$\AA^1_X$, that is
+    \[ \ull{\AA}^1_X : (T/X) \longmapsto \O_T(T), \]
     looks like a local ring and indeed like a field from the internal point of
     view, in the sense that
-    \[ \forall f\?\mathbb{A}^1_X\_ \neg(f = 0) \Rightarrow \text{$f$
+    \[ \forall f\?\ull{\AA}^1_X\_ \neg(f = 0) \Rightarrow \text{$f$
     invertible}. \]
+    \item Let~$\A$ be a quasicoherent~$\O_X$-algebra. Let~$\E$ be the
+    induced~$\ull{\AA}^1_X$-algebra given by~$\E(T \xrightarrow{\mu} X) \defeq
+    (\mu^* \A)(T)$. Then the internal Hom set~$[\E,
+    \ull{\AA}^1_X]_{\ull{\AA}^1_X}$ of~$\ull{\AA}^1_X$-algebra morphisms is the
+    functor of points of~$\RelSpec_X(\A)$.
+    \item Let~$\mu : T \to X$ be quasicompact and quasiseparated.
+    Then~$\mu$ is affine iff, from the internal point of view, the map
+    \[ \ull{T} \longrightarrow [ [\ull{T},\ull{\AA}^1_X]^\flat, \ull{\AA}^1_X ]_{\ull{\AA}^1_X},\ x \longmapsto
+    \ull{\ \ }(x) \]
+    into the ``double dual'' is bijective.
+  \end{itemize}
+}
+
+\note{
+  \begin{itemize}\justifying
     \item Describing the functor of points of the projective space was
@@
   \end{itemize}
-  \par
 }
 
-\begin{frame}\frametitle{Translating internal statements}
-  Let~$X$ be a topological space (or locale) and let~$\alpha : \F \to \G$ be a
-  morphism of sheaves on~$X$. Then:
-  \allowdisplaybreaks
-  \begin{align*}
-    & \Sh(X) \models \speak{$\alpha$ is injective} \\[0.5em]
-    \Longleftrightarrow\
-    & \Sh(X) \models \forall s\?\F\_ \forall t\?\F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
-    \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$, sections $s \in \F(U)$:} \\
-    &\qquad
-      \text{for all open~$V \subseteq U$, sections $t \in \F(V)$:} \\
-    &\qquad\qquad
-        \text{for all open~$W \subseteq V$:} \\
-    &\qquad\qquad\qquad
-          \text{$\alpha_W(s|_W) = \alpha_W(t|_W)$ implies $s|_W = t|_W$} \\[0.5em]
-    \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
-    &\qquad
-          \text{$\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.5em]
-    \Longleftrightarrow\ &
-      \text{$\alpha$ is a monomorphism of sheaves}
-  \end{align*}
-\end{frame}
-
-\backupend
-
-\end{document}
-
-1. Tell story; begin with MO question
-2. Basics (include praise for Mike Shulman)
-3. Curious field property and more advanced examples
-4. Relative spectrum (also mention equivalence to Hakim's spectrum in dimension zero)
-5. Outlook: synthetic algebraic geometry, relative geometry, big Zariski topos
-
-\note{\justifying\fontsize{8pt}{9.6}\selectfont
-  \begin{center}\large\textbf{Abstract}\end{center}
-
-  \begin{changemargin}{2.5em}{2.5em}
-    We describe how the internal language of certain
-    toposes, the associated petit and gros Zariski toposes of a scheme, can be
-    used to give simpler definitions and more conceptual proofs of the basic
-    notions and observations in algebraic geometry.
-    % This is useful for studying schemes from a local or relative point of view.
-
-    The starting point is that, from the internal point of view, sheaves of rings
-    and sheaves of modules look just like plain rings and plain modules.
-    In this way, some concepts and statements of scheme theory can be reduced to
-    concepts and statements of intuitionistic linear algebra.
-
-    Furthermore, modal operators can be used to model phrases such as ``on a
-    dense open subset it holds that'' or ``on an open neighbourhood of a given
-    point it holds that''. These operators define certain subtoposes; a
-    generalization of the double-negation translation is useful in order to
-    understand the internal universe of those subtoposes from the internal point
-    of view of the ambient topos.
-
-    A particularly interesting task is to internalize the
-    construction of the relative spectrum, which, given a quasicoherent sheaf of algebras
-    on a scheme~$X$, yields a scheme over~$X$. From the internal point of
-    view, this construction should simply reduce to an intuitionistically sensible
-    variant of the ordinary construction of the spectrum of a ring, but it turns
-    out that this expectation is too naive and that a refined approach is
-    necessary.
-  \end{changemargin}
-}
-
-  \scriptsize
-  \begin{itemize}
-    \item $\Sh(X) \models f = g$ iff~$f = g \in \F(X)$.
-    \item $\Sh(X) \models f = g \vee f = h$ iff on a cover~$X = \bigcup_i U_i$,
-    $f|_{U_i} = g|_{U_i}$ or~$f|_{U_i} = h|_{U_i}$.
-    \item $\Sh(X) \models \neg\neg(f = g)$ iff~$f = g$ on a dense open subset
-    of~$X$.
-  \end{itemize}
+\section{Basics about the internal language}
 
@@
 
-\begin{frame}\frametitle{Limits of locally ringed locales}
-  Let~$(X_i)_i$ be a diagram of locally ringed locales.
+\note{\justifying
+  More generally, for an object~$U$ of a topos~$\E$, we define the meaning of
+  \[ U \models \varphi \qquad\text{($\varphi$ holds on~$U$)}. \]
+  Writing~``$\E \models \varphi$'' is then an abbreviation for~``$1 \models
+  \varphi$'', where~``$1$'' denotes the terminal object of~$\E$.
 
-  Its limit is given by \hil{relocalizing} the corresponding limit
-  in~$\mathrm{RL}$.
+  In addition to soundness with respect to intuitionistic logic, the internal
+  language has the following two important properties:
+  \begin{itemize}
+    \item \hil{Monotonicity:} If~$p : V \to U$ is an arbitrary morphism and~$U
+    \models \varphi$, then also~$V \models \varphi$.
+    \item \hil{Locality:} If~$p : V \to U$ is an epimorphism and~$V \models
+    \varphi$, then also~$U \models \varphi$.
+  \end{itemize}
+}
+
+\note{\justifying
+  In the special case that~$\E = \Sh(X)$ is the topos of sheaves on a
+  topological space (or locale)~$X$, the rules of the Kripke--Joyal semantics
+  look as follows. We tersely write~``$U \models \varphi$'' instead
+  of~``$\Hom(\ull{\ \ }, U) \models \varphi$ for open subsets~$U \subseteq X$.
+  \scriptsize
+  \newcommand{\Ll}{:\Longleftrightarrow}
+  \[ \renewcommand{\arraystretch}{1.3}\begin{array}{@{}lcl@{}}
+    U \models f = g \? \F &\Ll& f|_U = g|_U \in \F(U) \\
+    U \models \varphi \wedge \psi &\Ll&
+      \text{$U \models \varphi$ and $U \models \psi$} \\
+    U \models \varphi \vee \psi &\Ll&
+      \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt} \\
+    && \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
+    && \quad\quad \text{$U_i \models \varphi$ or $U_i \models \psi$} \\
+    U \models \varphi \Rightarrow \psi &\Ll&
+      \text{for all open~$V \subseteq U$: }
+    \text{$V \models \varphi$ implies $V \models \psi$} \\
+    U \models \forall f \? \F\_ \varphi(f) &\Ll&
+      \text{for all sections~$f \in \F(V), V \subseteq U$: $V \models
+      \varphi(f)$} \\
+    U \models \exists f \? \F\_ \varphi(f) &\Ll&
+      \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
+    && \quad\quad \text{there exists~$f_i \in \F(U_i)$ s.\,th.
+    $U_i \models \varphi(f_i)$}
+  \end{array} \]
+}
+
+\begin{frame}\frametitle{Translating internal statements I}
+  Let~$X$ be a topological space (or locale) and let~$\alpha : \F \to \G$ be a
+  morphism of sheaves on~$X$. Then:
+  \allowdisplaybreaks
+  \begin{align*}
+    & \Sh(X) \models \speak{$\alpha$ is injective} \\[0.5em]
+    \Longleftrightarrow\
+    & \Sh(X) \models \forall s\?\F\_ \forall t\?\F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{for all open~$U \subseteq X$, sections $s \in \F(U)$:} \\
+    &\qquad
+      \text{for all open~$V \subseteq U$, sections $t \in \F(V)$:} \\
+    &\qquad\qquad
+        \text{for all open~$W \subseteq V$:} \\
+    &\qquad\qquad\qquad
+          \text{$\alpha_W(s|_W) = \alpha_W(t|_W)$ implies $s|_W = t|_W$} \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
+    &\qquad
+          \text{$\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{$\alpha$ is a monomorphism of sheaves}
+  \end{align*}
 \end{frame}
-% Mention only orally?
-% Phrase more categorically.
-% Extend to toposes.
+
+\begin{frame}\frametitle{Translating internal statements II}
+  Let~$X$ be a topological space (or locale) and let~$\alpha : \F \to \G$ be a
+  morphism of sheaves on~$X$. Then:
+  \allowdisplaybreaks
+  \begin{align*}
+    & \Sh(X) \models \speak{$\alpha$ is surjective} \\[0.5em]
+    \Longleftrightarrow\
+    & \Sh(X) \models \forall t\?\G\_ \exists s\?\F\_ \alpha(s) = t \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{for all open~$U \subseteq X$, sections $t \in \G(U)$:} \\
+    &\qquad
+      \text{there exists an open covering~$U = \textstyle\bigcup_i U_i$ and} \\
+    &\qquad
+      \text{sections~$s_i \in \F(U_i)$ such that:} \\
+    &\qquad\qquad
+        \alpha|_{U_i}(s_i) = t|_{U_i} \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{$\alpha$ is an epimorphism of sheaves}
+  \end{align*}
+\end{frame}
+
+\begin{frame}\frametitle{Translating internal statements III}
+  Let~$X$ be a topological space (or locale) and let~$s, t \in \F(X)$ be global
+  sections of a sheaf~$\F$ on~$X$. Then:
+  \allowdisplaybreaks
+  \begin{align*}
+    & \Sh(X) \models \neg\neg(s = t) \\[0.5em]
+    \Longleftrightarrow\
+    & \Sh(X) \models ((s = t) \Rightarrow \bot) \Rightarrow \bot \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{for all open~$U \subseteq X$ such that} \\
+    &\qquad
+      \text{for all open~$V \subseteq U$ such that} \\
+    &\qquad\qquad
+      s|_V = t|_V, \\
+    &\qquad
+      \text{it holds that~$V = \emptyset$,} \\
+    &
+      \text{it holds that~$U = \emptyset$} \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{there exists a dense open set~$W \subseteq X$ such that $s|_W = t|_W$} \\
+  \end{align*}
+\end{frame}
+
+\backupend
+
+\end{document}
+
+1. Tell story; begin with MO question
+2. Basics (include praise for Mike Shulman)
+3. Curious field property and more advanced examples
+4. Relative spectrum (also mention equivalence to Hakim's spectrum in dimension zero)
+5. Outlook: synthetic algebraic geometry, relative geometry, big Zariski topos
+
+  \scriptsize
+  \begin{itemize}
+    \item $\Sh(X) \models f = g$ iff~$f = g \in \F(X)$.
+    \item $\Sh(X) \models f = g \vee f = h$ iff on a cover~$X = \bigcup_i U_i$,
+    $f|_{U_i} = g|_{U_i}$ or~$f|_{U_i} = h|_{U_i}$.
+    \item $\Sh(X) \models \neg\neg(f = g)$ iff~$f = g$ on a dense open subset
+    of~$X$.
+  \end{itemize}

### 2015-12-01 slides-ihes2015.tex
Further notes

@@
   the base scheme is~$\leq 0$.
+
+  If~$X$ is not a scheme or~$\A$ is not quasicoherent, the construction still
+  gives rise to a locally ringed locale over~$X$ which satisfies the universal
+  property
+  \[ \Hom_{\mathrm{LRL}/X}(T, \RelSpec_X \A) \cong \Hom_{\mathrm{Alg}(\O_X)}(\A,
+  \mu_*\O_T) \]
+  for all locally ringed locales~$T \xrightarrow{\mu} X$ over~$X$.
   \par
@@
   \small
-  \begin{itemize}
+  \begin{enumerate}
     \item $\F_x = 0$ iff~$\F|_U = 0$ for some open neighbourhood of~$x$.
@@
     locally Noetherian and~$\F$ quasicoherent).
-  \end{itemize}
-\end{frame}
+  \end{enumerate}
+\end{frame}
+
+\note{\justifying
+  Statements 1 and 2 follow from \emph{one} proof in the internal language,
+  applied to two different modal operators.
+
+  Similarly with statements~5 and~6.
+  \par
+}
 
@@
   Observe that $M^{\sim} \cong \underline{M}[\F^{-1}]$ is the localization of~$M$ at
-  the \hil{generic prime filter} and that~$M$ and the constant sheaf of
-  modules~$\underline{M}$ share all first-order properties. Therefore:
+  the \hil{generic prime filter} and that~$M$ shares all first-order properties
+  with the constant sheaf of modules~$\underline{M}$. Therefore:
   \begin{center}

### 2016-03-08 slides-pssl99.tex
Beginning of slides for PSSL 99

--- 2015-12-01 slides-ihes2015.tex
@@
 \author{Ingo Blechschmidt}
-\date{November 27th, 2015}
+\date{March 13th, 2016}
 
@@
 
-  Topos à l'IHES \\
-  November 27th, 2015
+  Special PSSL 99 \\
+  in honor of Jirka Adamek's retirement \\
+  March 13th, 2016
   \par
@@
 
-\backupstart
-
 \begin{frame}
@@
 
-\begin{frame}[plain,c]
-  \centering
-  \includegraphics[scale=0.5]{images/sheafification-man}
-  \sffamily
-
-  You should totally look up:
-
-  \hil{The Adventures of Sheafification Man}
-  \par
-\end{frame}
+\backupstart
 
@@
   algebra} explored by Coquand, Coste, Lombardi, Roy, and others. Their
-  approach applies more generally.
+  approach is more versatile.
   \par
@@
 
-1. Tell story; begin with MO question
-2. Basics (include praise for Mike Shulman)
-3. Curious field property and more advanced examples
-4. Relative spectrum (also mention equivalence to Hakim's spectrum in dimension zero)
-5. Outlook: synthetic algebraic geometry, relative geometry, big Zariski topos
-
-  \scriptsize
-  \begin{itemize}
-    \item $\Sh(X) \models f = g$ iff~$f = g \in \F(X)$.
-    \item $\Sh(X) \models f = g \vee f = h$ iff on a cover~$X = \bigcup_i U_i$,
-    $f|_{U_i} = g|_{U_i}$ or~$f|_{U_i} = h|_{U_i}$.
-    \item $\Sh(X) \models \neg\neg(f = g)$ iff~$f = g$ on a dense open subset
-    of~$X$.
-  \end{itemize}
+1. Begin with toposes and their internal logic.
+   Gauge the audience's familiarity with sheaves!
+
+2. State the MO question by Steven Gubkin.
+
+3. Discuss basics (include praise for Mike Shulman).
+
+4. Explain the Diamond-translation.
+
+5. Finish with outlook: more complex internal statements,
+   synthetic algebraic geometry, relative algebraic geometry.
+
+Do not include:
+* Quasicoherence condition
+* Relative spectrum

### 2016-03-13 slides-pssl99.tex
to do
Slides for PSSL 99 mostly finished

@@
 \newcommand{\NN}{\mathbb{N}}
+\newcommand{\RR}{\mathbb{R}}
 \newcommand{\ZZ}{\mathbb{Z}}
@@
   \begin{center}%
+    \setlength{\fboxrule}{2pt}%
+    \setlength{\fboxsep}{8pt}%
+    {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal
+    text}\parbox{0.87\textwidth}{#1}}}%
+  \end{center}%
+}
+
+\newcommand{\sloganwithoutborder}[1]{%
+  \begin{center}%
     \setlength{\fboxrule}{0pt}%
@@
 \setbeamertemplate{note page}[plain]
+\setbeameroption{hide notes}
 
@@
   Special PSSL 99 \\
-  in honor of Jirka Adamek's retirement \\
+  in honor of Jirka Adámek's retirement \\
+  \ \\
   March 13th, 2016
@@
 
-\backupstart
 \frame[t]{\frametitle{Outline}\scriptsize\begin{itemize}\item[]\tableofcontents\end{itemize}}
@@
     We describe how the internal language of certain
-    toposes, the associated petit and gros Zariski toposes of a scheme, can be
+    toposes, the associated small and big Zariski toposes of a scheme, can be
     used to give simpler definitions and more conceptual proofs of the basic
@@
 
-\begin{frame}[plain,c]
-  \centering
-  \includegraphics[scale=0.37]{images/math-overflow-steven-gubkin}
-  \par
-\end{frame}
-\backupend
-
-\section{Basic applications of the internal language}
-
-\begin{frame}\frametitle{Exploiting the internal language}
-  A \hil{scheme} is a locally ringed space~$(X,\O_X)$ which is
-  locally isomorphic to the \hil{spectrum of a commutative ring}:
+\section{Basics}
+
+\begin{frame}\frametitle{What is a scheme?}
+  \begin{itemize}
+  \item A \hil{manifold} is a space which is \hil{locally isomorphic} to some open
+  subset of some~$\RR^n$.
+  \medskip
+  \item A \hil{scheme} is a space which is \hil{locally isomorphic} to the
+  \hil{spectrum of some (commutative) ring}:
   \[ \Spec A \defeq \{ \ppp \subseteq A \,|\, \text{$\ppp$ is a prime ideal} \} \]
-  \pause
-  The topos~$\Sh(X)$ is the \hil{petit Zariski topos} of~$X$.
+  \item By \hil{space} we mean: topological space~$X$ equipped with a local
+  sheaf~$\O_X$ of rings.
+  \end{itemize}
+\end{frame}
+
+\begin{frame}\frametitle{What is a topos?}
+  \begin{block}{Formal definition}A \hil{topos} is a category which has finite limits,
+  is cartesian closed and has a subobject classifier.
+  \end{block}
+
+  \begin{block}{Motto}A topos is a category sufficiently rich to support
+  an \hil{internal language}.
+  \end{block}
+
+  \begin{block}{Examples}\begin{itemize}
+    \item \hil{$\Set$:} \tabto{1.4cm} category of sets
+    \item \hil{$\Sh(X)$:} \tabto{1.4cm} category of set-valued sheaves on a space~$X$
+  \end{itemize}\end{block}
+\end{frame}
+
+\begin{frame}\frametitle{What is the internal language?}
+  Let~$\E$ be a topos. Then we can define the meaning of
+  \[ \hil{$\E \models \varphi \qquad\textnormal{(``$\varphi$ holds in $\E$'')}$} \]
+  for formulas~$\varphi$ over~$\E$ using the \hil{Kripke--Joyal semantics}.
+
+%  The internal language of a topos~$\E$ allows to
+%  \begin{enumerate}
+%    \item construct objects and morphisms of the topos,
+%    \item formulate statements about them and
+%    \item prove such statements
+%  \end{enumerate}
+%  in a \hil{naive element-based} language:
 
@@
       \toprule
-      externally & internally to $\Sh(X)$ \\
+      externally & internally to $\E$ \\
       \midrule
-      sheaf of sets & set/type \\
-      morphism of sheaves & map of sets \\
+      object of~$\E$ & set/type \\
+      morphism in~$\E$ & map of sets \\
       monomorphism & injective map \\
       epimorphism & surjective map \\
-      sheaf of rings & ring \\
-      sheaf of modules & module \\
+      ring object & ring \\
+      module object & module \\
       \bottomrule
@@
   \end{center}
+
+  \small
+  \mbox{If~$\varphi$ implies~$\psi$ \hil{intuitionistically}, then~$\E \models \varphi$
+  implies~$\E \models \psi$.}
 \end{frame}
@@
 \begin{frame}\frametitle{Building a dictionary}
-  \slogan{\hil{Understand notions of algebraic geometry
+  \sloganwithoutborder{\hil{Understand notions of algebraic geometry
   as notions of algebra internal to~$\boldsymbol{\Sh(X)}$.}}
@@
 
-\begin{frame}[c]\frametitle{Praise for Mike Shulman}
-  \centering
-  \includegraphics[scale=0.4]{images/mike-shulman-stack-semantics}
-  \par
-\end{frame}
-
-\note{\justifying
-  The internal language of a topos supports
-  \begin{itemize}
-  \item first-order logic,
-  \item higher-order logic (for instance quantification over subsets),
-  \item dependent types, and
-  \item unbounded quantification.
-  \end{itemize}
-
-  The first three items are standard. The fourth is due to Mike Shulman.
-  Combined, it's possible to interpret ``essentially all of constructive
-  mathematics'' internal to a topos.
-
-  Restrictions persist for operations with a ``set-theoretical flavor'' like
-  building an infinite union of iterated powersets, for example~$\bigcup_{n \in
-  \NN} P^n(\NN)$.
-  \par
-}
-
 \begin{frame}\frametitle{Using the dictionary}
+  Let~$X$ be a scheme. Employ its \hil{small Zariski topos}:~$\Sh(X)$.
+
   \begin{center}
-    \begin{minipage}{0.75\textwidth}
+    \begin{minipage}{0.70\textwidth}
       \begin{exampleblock}{}
@@
 
-    \begin{minipage}{0.75\textwidth}
+    \begin{minipage}{0.70\textwidth}
       \begin{exampleblock}{}
@@
         Let $0 \to \F' \to \F \to \F'' \to 0$ be a short exact sequence
-        of~$\O_X$-modules. If~$\F'$ and~$\F''$ are of finite type, so
-        is~$\F$.
+        of sheaves of~$\O_X$-modules. If~$\F'$ and~$\F''$ are of finite type,
+        so is~$\F$.
       \end{exampleblock}
@@
 
+\begin{frame}\frametitle{The objective}
+  \slogan{\justifying Understand notions and statements of \hil{algebraic geometry} as
+  notions and statements of (intuitionistic) \hil{commutative
+  algebra} internal to suitable \hil{toposes}.}
+
+%  In order to:
+%  \begin{itemize}
+%    \item Gain conceptual understanding.
+%    \item Simplify proofs.
+%    \item Develop a synthetic account of scheme theory.
+%    \item Contribute to constructive algebra.
+%  \end{itemize}
+
+  Further examples:
+  \begin{itemize}
+    \item Characterizing quasicoherence internally
+    \item Understanding spreading of properties in a logical way
+    \item Constructing the relative spectrum internally
+  \end{itemize}
+\end{frame}
+
+\begin{frame}[c]\frametitle{Praise for Mike Shulman}
+  \centering
+  \includegraphics[scale=0.4]{images/mike-shulman-stack-semantics}
+  \par
+\end{frame}
+
+\note{\justifying
+  The internal language of a topos supports
+  \begin{itemize}
+  \item first-order logic,
+  \item higher-order logic (for instance quantification over subsets),
+  \item dependent types, and
+  \item unbounded quantification.
+  \end{itemize}
+
+  The first three items are standard. The fourth is due to Mike Shulman.
+  Combined, it's possible to interpret ``essentially all of constructive
+  mathematics'' internal to a topos.
+
+  Restrictions persist for operations with a ``set-theoretical flavor'' like
+  building an infinite union of iterated powersets, for example~$\bigcup_{n \in
+  \NN} P^n(\NN)$.
+  \par
+}
+
+\section{Quasicoherence of sheaves of modules}
+
 \begin{frame}\frametitle{A curious property}
@@
 
-\section{\texorpdfstring{The $\Diamond$-translation}{The ♢-translation}}
+\begin{frame}\frametitle{Quasicoherence}
+  Let~$X$ be a scheme. Let~$\E$ be an~$\O_X$-module.
+
+  Then~$\E$ is quasicoherent
+  if and only if, internally to~$\Sh(X)$,
+  \begin{quote}\textnormal{$\E[f^{-1}]$ is a $\Diamond_f$-sheaf for any~$f : \O_X$, \\[0.3em]
+  \qquad\qquad where~$\Diamond_f\varphi \defeqv (\text{$f$ invertible} \Rightarrow \varphi)$.}
+  \end{quote}
+  \pause
+
+  In particular: If~$\E$ is quasicoherent, then internally
+  \[ (\text{$f$ invertible} \Rightarrow s = 0) \Longrightarrow
+    \bigvee_{n \geq 0} f^n s = 0 \]
+  \vspace*{-1.5em}\par%
+  for any~$f : \O_X$ and~$s : \E$.
+\end{frame}
+
+\note{\justifying
+  The sheaf condition and the sheafification functor can be described purely
+  internally. An object~$M$ is \emph{separated} with respect to~$\Diamond$ if
+  and only if, from the internal point of view,
+  \[ \forall x,y : M\_\ \Diamond(x = y) \Rightarrow x = y. \]
+  It is a \emph{sheaf} with respect to~$\Diamond$, if furthermore
+  \[ \forall K \subseteq M\_\
+    \Diamond(\exists x : M\_\ K = \{ x \}) \Longrightarrow
+    \exists x : M\_\ \Diamond(x \in K). \]
+
+  The second condition displayed on the previous slide is equivalent to the
+  separatedness condition. In the special case~$\E = \O_X$, $s = 1$ it reduces
+  to Mulvey's ``somewhat obscure formula''. We now understand this condition in
+  its proper context.
+  \par
+}
+
+\section{Spreading of properties}
 
@@
   and this condition is manifestly geometric.
-  \par
-}
-
-\section{Quasicoherence of sheaves of modules}
-
-\begin{frame}\frametitle{Quasicoherence}
-  Let~$X$ be a scheme. Let~$\E$ be an~$\O_X$-module.
-
-  Then~$\E$ is quasicoherent
-  if and only if, internally to~$\Sh(X)$,
-  \begin{quote}\textnormal{$\E[f^{-1}]$ is a $\Diamond_f$-sheaf for any~$f : \O_X$, \\[0.3em]
-  \qquad\qquad where~$\Diamond_f\varphi \defeqv (\text{$f$ invertible} \Rightarrow \varphi)$.}
-  \end{quote}
-  \pause
-
-  In particular: If~$\E$ is quasicoherent, then internally
-  \[ (\text{$f$ invertible} \Rightarrow s = 0) \Longrightarrow
-    \bigvee_{n \geq 0} f^n s = 0 \]
-  \vspace*{-1.5em}\par%
-  for any~$f : \O_X$ and~$s : \E$.
-\end{frame}
-
-\note{\justifying
-  The sheaf condition and the sheafification functor can be described purely
-  internally. An object~$M$ is \emph{separated} with respect to~$\Diamond$ if
-  and only if, from the internal point of view,
-  \[ \forall x,y : M\_\ \Diamond(x = y) \Rightarrow x = y. \]
-  It is a \emph{sheaf} with respect to~$\Diamond$, if furthermore
-  \[ \forall K \subseteq M\_\
-    \Diamond(\exists x : M\_\ K = \{ x \}) \Longrightarrow
-    \exists x : M\_\ \Diamond(x \in K). \]
-
-  The second condition displayed on the previous slide is equivalent to the
-  separatedness condition. In the special case~$\E = \O_X$, $s = 1$ it reduces
-  to Mulvey's ``somewhat obscure formula''. We now understand this condition in
-  its proper context.
   \par
@@
 \begin{frame}
-  \slogan{\hil{Understand notions and statements of algebraic geometry
+  \sloganwithoutborder{\hil{Understand notions and statements of algebraic geometry
   as notions and statements of algebra internal to appropriate toposes.}}
@@
 
+\begin{frame}[plain,c]
+  \centering
+  \includegraphics[scale=0.5]{images/sheafification-man}
+  \sffamily
+
+  You should totally look up:
+
+  \hil{The Adventures of Sheafification Man}
+  \par
+\end{frame}
+
 \appendix
+
+\section{Details on the internal language}
+
+\begin{frame}\frametitle{More on the internal language}
+  More generally, for an object~$U$ of a topos~$\E$, we define the meaning of
+  \[ U \models \varphi \qquad\text{($\varphi$ holds on~$U$)}. \]
+  Writing~``$\E \models \varphi$'' is then an abbreviation for~``$1 \models
+  \varphi$'', where~``$1$'' denotes the terminal object of~$\E$.
+
+  In addition to soundness with respect to intuitionistic logic, the internal
+  language has the following two important properties:
+  \begin{itemize}
+    \item \hil{Monotonicity:} If~$p : V \to U$ is an arbitrary morphism and~$U
+    \models \varphi$, then also~$V \models \varphi$.
+    \item \hil{Locality:} If~$p : V \to U$ is an epimorphism and~$V \models
+    \varphi$, then also~$U \models \varphi$.
+  \end{itemize}
+\end{frame}
+
+\begin{frame}\frametitle{The rules of the Kripke--Joyal semantics}
+  In the special case that~$\E = \Sh(X)$ is the topos of sheaves on a
+  topological space (or locale)~$X$, the rules of the Kripke--Joyal semantics
+  look as follows. We tersely write~``$U \models \varphi$'' instead
+  of~``$\Hom(\ull{\ \ }, U) \models \varphi$ for open subsets~$U \subseteq X$.
+  \scriptsize
+  \newcommand{\Ll}{:\Longleftrightarrow}
+  \[ \renewcommand{\arraystretch}{1.3}\begin{array}{@{}lcl@{}}
+    U \models f = g \? \F &\Ll& f|_U = g|_U \in \F(U) \\
+    U \models \varphi \wedge \psi &\Ll&
+      \text{$U \models \varphi$ and $U \models \psi$} \\
+    U \models \varphi \vee \psi &\Ll&
+      \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt} \\
+    && \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
+    && \quad\quad \text{$U_i \models \varphi$ or $U_i \models \psi$} \\
+    U \models \varphi \Rightarrow \psi &\Ll&
+      \text{for all open~$V \subseteq U$: }
+    \text{$V \models \varphi$ implies $V \models \psi$} \\
+    U \models \forall f \? \F\_ \varphi(f) &\Ll&
+      \text{for all sections~$f \in \F(V), V \subseteq U$: $V \models
+      \varphi(f)$} \\
+    U \models \exists f \? \F\_ \varphi(f) &\Ll&
+      \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
+    && \quad\quad \text{there exists~$f_i \in \F(U_i)$ s.\,th.
+    $U_i \models \varphi(f_i)$}
+  \end{array} \]
+\end{frame}
+
+\begin{frame}\frametitle{Translating internal statements I}
+  Let~$X$ be a topological space (or locale) and let~$\alpha : \F \to \G$ be a
+  morphism of sheaves on~$X$. Then:
+  \allowdisplaybreaks
+  \begin{align*}
+    & \Sh(X) \models \speak{$\alpha$ is injective} \\[0.5em]
+    \Longleftrightarrow\
+    & \Sh(X) \models \forall s\?\F\_ \forall t\?\F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{for all open~$U \subseteq X$, sections $s \in \F(U)$:} \\
+    &\qquad
+      \text{for all open~$V \subseteq U$, sections $t \in \F(V)$:} \\
+    &\qquad\qquad
+        \text{for all open~$W \subseteq V$:} \\
+    &\qquad\qquad\qquad
+          \text{$\alpha_W(s|_W) = \alpha_W(t|_W)$ implies $s|_W = t|_W$} \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
+    &\qquad
+          \text{$\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{$\alpha$ is a monomorphism of sheaves}
+  \end{align*}
+\end{frame}
+
+\begin{frame}\frametitle{Translating internal statements II}
+  Let~$X$ be a topological space (or locale) and let~$\alpha : \F \to \G$ be a
+  morphism of sheaves on~$X$. Then:
+  \allowdisplaybreaks
+  \begin{align*}
+    & \Sh(X) \models \speak{$\alpha$ is surjective} \\[0.5em]
+    \Longleftrightarrow\
+    & \Sh(X) \models \forall t\?\G\_ \exists s\?\F\_ \alpha(s) = t \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{for all open~$U \subseteq X$, sections $t \in \G(U)$:} \\
+    &\qquad
+      \text{there exists an open covering~$U = \textstyle\bigcup_i U_i$ and} \\
+    &\qquad
+      \text{sections~$s_i \in \F(U_i)$ such that:} \\
+    &\qquad\qquad
+        \alpha|_{U_i}(s_i) = t|_{U_i} \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{$\alpha$ is an epimorphism of sheaves}
+  \end{align*}
+\end{frame}
+
+\begin{frame}\frametitle{Translating internal statements III}
+  Let~$X$ be a topological space (or locale) and let~$s, t \in \F(X)$ be global
+  sections of a sheaf~$\F$ on~$X$. Then:
+  \allowdisplaybreaks
+  \begin{align*}
+    & \Sh(X) \models \neg\neg(s = t) \\[0.5em]
+    \Longleftrightarrow\
+    & \Sh(X) \models ((s = t) \Rightarrow \bot) \Rightarrow \bot \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{for all open~$U \subseteq X$ such that} \\
+    &\qquad
+      \text{for all open~$V \subseteq U$ such that} \\
+    &\qquad\qquad
+      s|_V = t|_V, \\
+    &\qquad
+      \text{it holds that~$V = \emptyset$,} \\
+    &
+      \text{it holds that~$U = \emptyset$} \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{there exists a dense open set~$W \subseteq X$ such that $s|_W = t|_W$} \\
+  \end{align*}
+\end{frame}
 
@@
 
-\section{The gros Zariski topos}
-
-\begin{frame}\frametitle{The gros Zariski topos}
-  Let~$X$ be a scheme. The \hil{gros Zariski topos} is the topos of sheaves
+\section{The big Zariski topos}
+
+\begin{frame}\frametitle{The big Zariski topos}
+  Let~$X$ be a scheme. The \hil{big Zariski topos} is the topos of sheaves
   on~$\Sch/X$ with respect to the Zariski topology. From its point of view,
@@
 
-\section{Basics about the internal language}
-
-\begin{frame}\frametitle{The internal language of a topos}
-  Let~$\E$ be a topos. Then we can define the meaning of
-  \[ \hil{$\E \models \varphi \qquad\textnormal{(``$\varphi$ holds in $\E$'')}$} \]
-  for formulas~$\varphi$ over~$\E$ using the \hil{Kripke--Joyal semantics}.
-
-  \begin{center}
-    \small
-    \begin{tabular}{ll}
-      \toprule
-      externally & internally \\
-      \midrule
-      object & set/type \\
-      morphism & map of sets \\
-      monomorphism & injective map \\
-      epimorphism & surjective map \\
-      \bottomrule
-    \end{tabular}
-  \end{center}
-  \pause
-
-  \mbox{If~$\varphi$ implies~$\psi$ \hil{intuitionistically}, then~$\E \models \varphi$
-  implies~$\E \models \psi$.}
-\end{frame}
-
-\note{\justifying
-  More generally, for an object~$U$ of a topos~$\E$, we define the meaning of
-  \[ U \models \varphi \qquad\text{($\varphi$ holds on~$U$)}. \]
-  Writing~``$\E \models \varphi$'' is then an abbreviation for~``$1 \models
-  \varphi$'', where~``$1$'' denotes the terminal object of~$\E$.
-
-  In addition to soundness with respect to intuitionistic logic, the internal
-  language has the following two important properties:
-  \begin{itemize}
-    \item \hil{Monotonicity:} If~$p : V \to U$ is an arbitrary morphism and~$U
-    \models \varphi$, then also~$V \models \varphi$.
-    \item \hil{Locality:} If~$p : V \to U$ is an epimorphism and~$V \models
-    \varphi$, then also~$U \models \varphi$.
-  \end{itemize}
-}
-
-\note{\justifying
-  In the special case that~$\E = \Sh(X)$ is the topos of sheaves on a
-  topological space (or locale)~$X$, the rules of the Kripke--Joyal semantics
-  look as follows. We tersely write~``$U \models \varphi$'' instead
-  of~``$\Hom(\ull{\ \ }, U) \models \varphi$ for open subsets~$U \subseteq X$.
-  \scriptsize
-  \newcommand{\Ll}{:\Longleftrightarrow}
-  \[ \renewcommand{\arraystretch}{1.3}\begin{array}{@{}lcl@{}}
-    U \models f = g \? \F &\Ll& f|_U = g|_U \in \F(U) \\
-    U \models \varphi \wedge \psi &\Ll&
-      \text{$U \models \varphi$ and $U \models \psi$} \\
-    U \models \varphi \vee \psi &\Ll&
-      \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt} \\
-    && \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
-    && \quad\quad \text{$U_i \models \varphi$ or $U_i \models \psi$} \\
-    U \models \varphi \Rightarrow \psi &\Ll&
-      \text{for all open~$V \subseteq U$: }
-    \text{$V \models \varphi$ implies $V \models \psi$} \\
-    U \models \forall f \? \F\_ \varphi(f) &\Ll&
-      \text{for all sections~$f \in \F(V), V \subseteq U$: $V \models
-      \varphi(f)$} \\
-    U \models \exists f \? \F\_ \varphi(f) &\Ll&
-      \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
-    && \quad\quad \text{there exists~$f_i \in \F(U_i)$ s.\,th.
-    $U_i \models \varphi(f_i)$}
-  \end{array} \]
-}
-
-\begin{frame}\frametitle{Translating internal statements I}
-  Let~$X$ be a topological space (or locale) and let~$\alpha : \F \to \G$ be a
-  morphism of sheaves on~$X$. Then:
-  \allowdisplaybreaks
-  \begin{align*}
-    & \Sh(X) \models \speak{$\alpha$ is injective} \\[0.5em]
-    \Longleftrightarrow\
-    & \Sh(X) \models \forall s\?\F\_ \forall t\?\F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
-    \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$, sections $s \in \F(U)$:} \\
-    &\qquad
-      \text{for all open~$V \subseteq U$, sections $t \in \F(V)$:} \\
-    &\qquad\qquad
-        \text{for all open~$W \subseteq V$:} \\
-    &\qquad\qquad\qquad
-          \text{$\alpha_W(s|_W) = \alpha_W(t|_W)$ implies $s|_W = t|_W$} \\[0.5em]
-    \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
-    &\qquad
-          \text{$\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.5em]
-    \Longleftrightarrow\ &
-      \text{$\alpha$ is a monomorphism of sheaves}
-  \end{align*}
-\end{frame}
-
-\begin{frame}\frametitle{Translating internal statements II}
-  Let~$X$ be a topological space (or locale) and let~$\alpha : \F \to \G$ be a
-  morphism of sheaves on~$X$. Then:
-  \allowdisplaybreaks
-  \begin{align*}
-    & \Sh(X) \models \speak{$\alpha$ is surjective} \\[0.5em]
-    \Longleftrightarrow\
-    & \Sh(X) \models \forall t\?\G\_ \exists s\?\F\_ \alpha(s) = t \\[0.5em]
-    \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$, sections $t \in \G(U)$:} \\
-    &\qquad
-      \text{there exists an open covering~$U = \textstyle\bigcup_i U_i$ and} \\
-    &\qquad
-      \text{sections~$s_i \in \F(U_i)$ such that:} \\
-    &\qquad\qquad
-        \alpha|_{U_i}(s_i) = t|_{U_i} \\[0.5em]
-    \Longleftrightarrow\ &
-      \text{$\alpha$ is an epimorphism of sheaves}
-  \end{align*}
-\end{frame}
-
-\begin{frame}\frametitle{Translating internal statements III}
-  Let~$X$ be a topological space (or locale) and let~$s, t \in \F(X)$ be global
-  sections of a sheaf~$\F$ on~$X$. Then:
-  \allowdisplaybreaks
-  \begin{align*}
-    & \Sh(X) \models \neg\neg(s = t) \\[0.5em]
-    \Longleftrightarrow\
-    & \Sh(X) \models ((s = t) \Rightarrow \bot) \Rightarrow \bot \\[0.5em]
-    \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$ such that} \\
-    &\qquad
-      \text{for all open~$V \subseteq U$ such that} \\
-    &\qquad\qquad
-      s|_V = t|_V, \\
-    &\qquad
-      \text{it holds that~$V = \emptyset$,} \\
-    &
-      \text{it holds that~$U = \emptyset$} \\[0.5em]
-    \Longleftrightarrow\ &
-      \text{there exists a dense open set~$W \subseteq X$ such that $s|_W = t|_W$} \\
-  \end{align*}
-\end{frame}
-
 \backupend
@@
 * Relative spectrum
+
+Mention in passing that the Knaster--Tarski theorem is useful in this context.

### 2016-03-13 slides-pssl99.tex
Final version of the slides

@@
 
+\subsection{What is a scheme?}
+
 \begin{frame}\frametitle{What is a scheme?}
@@
 
+\subsection{What is a topos?}
 \begin{frame}\frametitle{What is a topos?}
@@
 
+\subsection{What is the internal language?}
 \begin{frame}\frametitle{What is the internal language?}
@@
 
+\section{Building and using a dictionary}
 \begin{frame}\frametitle{Building a dictionary}
@@
 
-\begin{frame}\frametitle{More on the internal language}
+\begin{frame}\frametitle{More on the internal language}\justifying
   More generally, for an object~$U$ of a topos~$\E$, we define the meaning of
@@
 
-\begin{frame}\frametitle{The rules of the Kripke--Joyal semantics}
+\begin{frame}\frametitle{The rules of the Kripke--Joyal semantics}\justifying
   In the special case that~$\E = \Sh(X)$ is the topos of sheaves on a
@@
 
-\begin{frame}\frametitle{The smallest dense sublocale}
+\begin{frame}\frametitle{The smallest dense sublocale}\justifying
   Let~$X$ be a reduced scheme satisfying a technical condition.
@@
 
-  \begin{itemize}
+  \begin{itemize}\justifying
     \item This is a highbrow way of saying ``rational functions are regular
@@
 
-\begin{frame}\frametitle{Transfer principles}
+\begin{frame}\frametitle{Transfer principles}\justifying
   Let~$M$ be an~$A$-module. How do~$M$ and the sheaf~$M^{\sim}$ on~$\Spec A$

### 2016-03-13 slides-pssl99.tex
Version with notes

@@
 \usetikzlibrary{calc,shapes.callouts,shapes.arrows}
+\hypersetup{colorlinks=true}
 
@@
 \newcommand{\ZZ}{\mathbb{Z}}
+\renewcommand{\P}{\mathcal{P}}
 \newcommand{\ppp}{\mathfrak{p}}
@@
 \setbeamertemplate{note page}[plain]
-\setbeameroption{hide notes}
 
@@
   \end{itemize}
-\end{frame}
+
+  \centering
+  \begin{columns}
+    \begin{column}{0.45\textwidth}
+      \centering
+      \includegraphics[height=0.25\textheight]{images/manifold} \\
+      \tiny a manifold
+      \par
+    \end{column}
+    \begin{column}{0.45\textwidth}
+      \centering
+      \includegraphics[height=0.25\textheight]{images/mumfords-treasure-map} \\
+      \tiny Mumford's treasure map of~$\Spec \ZZ[X]$
+      \par
+    \end{column}
+  \end{columns}
+\end{frame}
+
+\note{\justifying
+  A \emph{sheaf of rings} on a topological space~$X$ is a ring object
+  in~$\Sh(X)$, the category of set-valued sheaves on~$X$.
+
+  A sheaf~$\O_X$ of rings is \emph{local} if and only if all the
+  stalks~$\O_{X,x}$ are local rings. Why not demand that the sets of
+  sections~$\O_X(U)$ are local rings? This has a geometric meaning, but can
+  also be motivated from a logical point of view: A sheaf of rings is local if
+  and only if, from the point of view of the internal language of~$\Sh(X)$, it
+  is a local ring.
+
+  Think of~$\O_X$ as the sheaf of ``number-valued functions'' on~$X$. In
+  algebraic geometry, this structure sheaf is a crucial part of the data:
+  Wildly different schemes can have the same underlying topological space.
+  \par
+}
 
 \subsection{What is a topos?}
+
 \begin{frame}\frametitle{What is a topos?}
@@
 
+\note{\begin{itemize}\justifying
+  \item While technically correct, the formal definition is actually
+  misleading in a sense: A topos has lots of other vital structure, which
+  is crucial for a rounded understanding, but is not listed in the
+  definition (which is trimmed for minimality).
+
+  A more comprehensive definition is: A \emph{topos} is a locally cartesian
+  closed, finitely complete and cocomplete Heyting category which is exact,
+  extensive and has a subobject classifier.
+
+  \item Check out an
+  \href{https://ncatlab.org/publications/published/Leinster2011}{article by
+  Tom Leinster} for a leisurely introduction to topos theory.
+\end{itemize}}
+
 \subsection{What is the internal language?}
+
 \begin{frame}\frametitle{What is the internal language?}
@@
 
+\note{\begin{itemize}\justifying
+  \item Actually, the objects of~$\E$ feel more like \emph{types} instead of \emph{sets}:
+  For instance, there is no global membership relation~$\in$. Rather,
+  for each object~$A$ of~$\E$, there is a relation~${\in_A} : A \times \P(A) \to
+  \Omega$, where~$\P(A)$ is the power object of~$A$ and~$\Omega$ is the
+  object of truth values of~$\E$ (can be understood as the power object of
+  a terminal object).
+
+  \item Compare with the embedding theorem for abelian categories:
+  There, an explicit embedding into a category of modules is constructed.
+  Here, we only change perspective and talk about the same objects and
+  morphisms.
+
+  \item There exists a weaker variant of the internal language which works
+  in abelian categories. By using it, one can even pretend that the objects
+  are abelian groups (instead of modules), and when constructing morphisms
+  by appealing to the axiom of unique choice (which is a theorem), one
+  doesn't even have to check linearity. The proof that this approach
+  works uses only categorical logic.
+\end{itemize}}
+
+\note{\begin{itemize}\justifying
+  \item The translation of internal statements into external ones is
+  facilitated by an easy mechanical procedure which one quickly grows
+  accustomed to.
+  \item The internal language of a sheaf topos of a $\mathrm{T}_1$-space is
+  \emph{classical} (that is, verifies the principle of excluded middle) if and
+  only if the space is discrete. That's a not particularly interesting special
+  case.
+  \item See Section~2.4 of
+  \href{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
+  notes} for remarks on how to appreciate intuitionistic logic.
+\end{itemize}}
+
 \section{Building and using a dictionary}
+
 \begin{frame}\frametitle{Building a dictionary}
@@
 \note{\justifying
-  See the \href{https://github.com/iblech/internal-methods/raw/master/notes.pdf}{notes} for more dictionary entries.
+  See the \href{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{notes} for more dictionary entries.
 
@@
 
+\newcommand{\Ll}{:\Longleftrightarrow}
 \begin{frame}\frametitle{The rules of the Kripke--Joyal semantics}\justifying
@@
   \scriptsize
-  \newcommand{\Ll}{:\Longleftrightarrow}
   \[ \renewcommand{\arraystretch}{1.3}\begin{array}{@{}lcl@{}}
@@
 \end{frame}
+
+\note{\begin{itemize}\justifying
+  \item The rules are not all arbitrary: Rather, they are very finely concerted
+  to make the crucial properties about the internal language (monotonicity,
+  locality, soundness with respect to intuitinistic logic) true.
+
+  \item If~$\F$ is an object of~$\Sh(X)$, we write ``$f:\F$'' instead
+  of~``$f\in\F$'' to remind us that~$\F$ is not really (externally) a set
+  consisting of elements, but that we only pretend this by using the internal
+  language.
+
+  \item There are two further rules concerning the constants~$\top$
+  and~$\bot$ (truth resp. falsehood):
+    \[ \renewcommand{\arraystretch}{1.3}\begin{array}{@{}lcl@{}}
+      U \models \top &\Ll& U = U \text{ (always fulfilled)}\\
+      U \models \bot &\Ll& U = \emptyset
+    \end{array} \]
+
+  \item Negation is defined as
+  \[ \neg\varphi :\equiv (\varphi \Rightarrow \bot). \]
+\end{itemize}}
+
+\note{\justifying
+  The alternate definition ``$U \models \varphi \vee \psi :\Leftrightarrow
+  \text{$U \models \varphi$ or $U \models \psi$}$'' would not be local.
+  \par
+}
 

### 2016-03-14 slides-pssl99.tex
typo

@@
 
-  \begin{block}{Motto}A topos is a category sufficiently rich to support
+  \begin{block}{Motto}A topos is a category which is sufficiently rich to support
   an \hil{internal language}.
@@
 
-\note{\begin{itemize}\justifying
-  \item While technically correct, the formal definition is actually
+\note{\justifying
+  While technically correct, the formal definition is actually
   misleading in a sense: A topos has lots of other vital structure, which
@@
 
-  \item Check out an
+  Check out an
   \href{https://ncatlab.org/publications/published/Leinster2011}{article by
   Tom Leinster} for a leisurely introduction to topos theory.
-\end{itemize}}
+  \par
+}
 
@@
   facilitated by an easy mechanical procedure which one quickly grows
-  accustomed to.
+  accustomed to. \hyperlink{kripke-joyal}{\beamergotobutton{See more details.}}
   \item The internal language of a sheaf topos of a $\mathrm{T}_1$-space is
@@
   This gives an internal description of
-  Monique Hakim's spectrum functor~$\mathrm{RT} \to \mathrm{LRT}$.
+  Monique Hakim's spectrum functor from ringed toposes to locally ringed
+  toposes.
 \end{frame}
@@
 
-\begin{frame}\frametitle{More on the internal language}\justifying
+\begin{frame}[label=kripke-joyal]\frametitle{More on the internal language}\justifying
   More generally, for an object~$U$ of a topos~$\E$, we define the meaning of

### 2016-05-18 slides-augsburg.tex
Beginning of the slides for the talk in the Oberseminar

--- 2016-03-14 slides-pssl99.tex
@@
 \renewcommand{\G}{\mathcal{G}}
+\newcommand{\GG}{\mathbb{G}}
 \renewcommand{\O}{\mathcal{O}}
@@
 \newcommand{\RR}{\mathbb{R}}
+\newcommand{\PP}{\mathbb{P}}
 \newcommand{\ZZ}{\mathbb{Z}}
@@
 \newcommand{\Sh}{\mathrm{Sh}}
+\newcommand{\GL}{\mathrm{GL}}
+\newcommand{\Zar}{\mathrm{Zar}}
+\newcommand{\op}{\mathrm{op}}
 \newcommand{\Set}{\mathrm{Set}}
@@
 \newcommand{\ull}[1]{\underline{#1}}
+\newcommand{\affl}{\ensuremath{{\ull{\AA}^1_S}}}
+\newcommand{\Ll}{\vcentcolon\!\Longleftrightarrow}
 
@@
 \author{Ingo Blechschmidt}
-\date{March 13th, 2016}
+\date{May 19th, 2016}
 
@@
     (#1) {#1};\node[overlay,rectangle callout,%
-    callout relative pointer={(-0.2cm,0.8cm)},fill=blue!20] at ($(#1.north)+(1.8cm,-1.4cm)$) {#2};}%
+    callout relative pointer={(0.5cm,1.2cm)},fill=blue!20] at ($(#1.north)+(-1.2cm,-1.8cm)$) {#2};}%
 }%
@@
     \setlength{\fboxsep}{8pt}%
-    {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal
-    text}\parbox{0.87\textwidth}{#1}}}%
+    {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal text}\parbox{0.87\textwidth}{#1}}}%
   \end{center}%
@@
   \begin{beamercolorbox}[wd=\paperwidth,ht=2.5ex,dp=1.25ex,right,rightskip=1mm,leftskip=1mm]{frametitle right}
-    {\quad} \inserttitle \hfill \insertauthor \quad
+    {\quad} \inserttitle \hfill
     \insertframenumber\,/\,\inserttotalframenumber {\quad}
@@
 \setbeamertemplate{note page}[plain]
+\setbeameroption{hide notes}  % XXX
 
@@
   \scriptsize
-  Ingo Blechschmidt \\
-  University of Augsburg
-  \medskip
-
-  Special PSSL 99 \\
-  in honor of Jirka Adámek's retirement \\
-  \ \\
-  March 13th, 2016
+  am Tag, an dem Marc im Curry Club vorträgt
   \par
@@
   \begin{changemargin}{2.5em}{2.5em}
+    % XXX: Update
     We describe how the internal language of certain
@@
 
-\subsection{What is a scheme?}
-
-\begin{frame}\frametitle{What is a scheme?}
-  \begin{itemize}
-  \item A \hil{manifold} is a space which is \hil{locally isomorphic} to some open
-  subset of some~$\RR^n$.
-  \medskip
-  \item A \hil{scheme} is a space which is \hil{locally isomorphic} to the
-  \hil{spectrum of some (commutative) ring}:
-  \[ \Spec A \defeq \{ \ppp \subseteq A \,|\, \text{$\ppp$ is a prime ideal} \} \]
-  \item By \hil{space} we mean: topological space~$X$ equipped with a local
-  sheaf~$\O_X$ of rings.
-  \end{itemize}
-
-  \centering
-  \begin{columns}
-    \begin{column}{0.45\textwidth}
-      \centering
-      \includegraphics[height=0.25\textheight]{images/manifold} \\
-      \tiny a manifold
-      \par
-    \end{column}
-    \begin{column}{0.45\textwidth}
-      \centering
-      \includegraphics[height=0.25\textheight]{images/mumfords-treasure-map} \\
-      \tiny Mumford's treasure map of~$\Spec \ZZ[X]$
-      \par
-    \end{column}
-  \end{columns}
-\end{frame}
-
-\note{\justifying
-  A \emph{sheaf of rings} on a topological space~$X$ is a ring object
-  in~$\Sh(X)$, the category of set-valued sheaves on~$X$.
-
-  A sheaf~$\O_X$ of rings is \emph{local} if and only if all the
-  stalks~$\O_{X,x}$ are local rings. Why not demand that the sets of
-  sections~$\O_X(U)$ are local rings? This has a geometric meaning, but can
-  also be motivated from a logical point of view: A sheaf of rings is local if
-  and only if, from the point of view of the internal language of~$\Sh(X)$, it
-  is a local ring.
-
-  Think of~$\O_X$ as the sheaf of ``number-valued functions'' on~$X$. In
-  algebraic geometry, this structure sheaf is a crucial part of the data:
-  Wildly different schemes can have the same underlying topological space.
-  \par
-}
-
 \subsection{What is a topos?}
@@
 \subsection{What is the internal language?}
-
-\begin{frame}\frametitle{What is the internal language?}
-  Let~$\E$ be a topos. Then we can define the meaning of
-  \[ \hil{$\E \models \varphi \qquad\textnormal{(``$\varphi$ holds in $\E$'')}$} \]
-  for formulas~$\varphi$ over~$\E$ using the \hil{Kripke--Joyal semantics}.
-
-%  The internal language of a topos~$\E$ allows to
-%  \begin{enumerate}
-%    \item construct objects and morphisms of the topos,
-%    \item formulate statements about them and
-%    \item prove such statements
-%  \end{enumerate}
-%  in a \hil{naive element-based} language:
+\frame[t]{\frametitle{What is the internal language?}
+  The internal language of a topos~$\E$ allows to
+  \begin{enumerate}
+    \item construct objects and morphisms of the topos,
+    \item formulate statements about them and
+    \item prove such statements
+  \end{enumerate}
+  in a \hil{naive element-based} language:
 
@@
 
+  \note{
+    \begin{itemize}
+      \item Special case: The language of~$\Set$ is the usual mathematical language.
+
+      \item \begin{justify}Actually, the objects of~$\E$ feel more like \emph{types} instead of \emph{sets}:
+      For instance, there is no global membership relation~$\in$. Rather,
+      for each object~$A$ of~$\E$, there is a relation~${\in_A} : A \times \P(A) \to
+      \Omega$, where~$\P(A)$ is the power object of~$A$ and~$\Omega$ is the
+      object of truth values of~$\E$ (can be understood as the power object of
+      a terminal object).\end{justify}
+
+      \item \begin{justify}Compare with the embedding theorem for abelian categories:
+      There, an explicit embedding into a category of modules is constructed.
+      Here, we only change perspective and talk about the same objects and
+      morphisms.\end{justify}
+
+      \item \begin{justify}There exists a weaker variant of the internal language which works
+      in abelian categories. By using it, one can even pretend that the objects
+      are abelian groups (instead of modules), and when constructing morphisms
+      by appealing to the axiom of unique choice (which is a theorem), one
+      doesn't even have to check linearity. The proof that this approach
+      works uses only categorical logic.\end{justify}
+
+      \item \begin{justify}For expositions of the internal language,
+      see the books~\cite{johnstone:elephant}~(chapters~D1 and~D4)
+      and~\cite{moerdijk:maclane:sheaves} (chapter~VI) or the lecture
+      notes~\cite{streicher:ctcl} (chapter~13).\end{justify}
+    \end{itemize}
+  }
+}
+
+\frame[t]{\frametitle{The internal language of~$\Sh(X)$}
+  Let~$X$ be a topological space. Then we recursively define
+  \[ U \models \varphi \quad \text{(``$\varphi$ holds on~$U$'')} \]
+  for open subsets~$U \subseteq X$ and formulas~$\varphi$.
+  \pause
   \small
-  \mbox{If~$\varphi$ implies~$\psi$ \hil{intuitionistically}, then~$\E \models \varphi$
-  implies~$\E \models \psi$.}
-\end{frame}
+  \[ \renewcommand{\arraystretch}{1.3}\begin{array}{@{}l@{\ }c@{\ }l@{}}
+    U \models f = g \? \F &\Ll& f|_U = g|_U \in \Gamma(U, \F) \\
+    U \models \varphi \wedge \psi &\Ll&
+      \text{$U \models \varphi$ and $U \models \psi$} \\
+    U \models \varphi \vee \psi &\Ll&
+      \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt} \\
+    && \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
+    && \quad\quad \text{$U_i \models \varphi$ or $U_i \models \psi$} \\
+    U \models \varphi \Rightarrow \psi &\Ll&
+      \text{for all open~$V \subseteq U$: } 
+    \text{$V \models \varphi$ implies $V \models \psi$} \\
+    U \models \forall f \? \F\_ \varphi(f) &\Ll&
+      \text{for all sections~$f \in \Gamma(V, \F), V \subseteq U$: $V \models
+      \varphi(f)$} \\
+    U \models \exists f \? \F\_ \varphi(f) &\Ll&
+      \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
+    && \quad\quad \text{there exists~$f_i \in \Gamma(U_i, \F)$ s.\,th.
+    $U_i \models \varphi(f_i)$}
+  \end{array} \]
+
+  \note{
+    \begin{itemize}
+      \item \begin{justify}The rules are called \emph{Kripke-Joyal semantics} and can be
+      formulated over any topos (not just sheaf topoi). They are not all
+      arbitrary: Rather, they are very finely concerted to make the crucial
+      properties about the internal language (see next slide) true.\end{justify}
+
+      \item \begin{justify}If~$\F$ is an object of~$\Sh(X)$, we write
+      ``$f:\F$'' instead of~``$f\in\F$'' to remind us that~$\F$ is not really
+      (externally) a set consisting of elements, but that we only pretend this
+      by using the internal language.\end{justify}
+
+      \item There are two further rules concerning the constants~$\top$
+      and~$\bot$ (truth resp. falsehood):
+        \[ \renewcommand{\arraystretch}{1.3}\begin{array}{@{}lcl@{}}
+          U \models \top &\Ll& U = U \text{ (always fulfilled)}\\
+          U \models \bot &\Ll& U = \emptyset
+        \end{array} \]
+
+      \item Negation is defined as
+      \[ \neg\varphi :\equiv (\varphi \Rightarrow \bot). \]
+
+      \item \begin{justify}The alternate definition ``$U \models \varphi \vee \psi :\Leftrightarrow
+      \text{$U \models \varphi$ or $U \models \psi$}$'' would not be local (cf.
+      next slide).\end{justify}
+    \end{itemize}
+  }
+}
+
+\note{
+  \begin{itemize}
+    \item Let~$\alpha : \F \to \G$ be a morphism of sheaves on~$X$. Then:
+    \begin{align*}
+      & X \models \speak{$\alpha$ is injective} \\[0.5em]
+      \Longleftrightarrow\
+      & X \models \forall s,t\?\F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
+      \Longleftrightarrow\ &
+        \text{for all open~$U \subseteq X$, sections $s, t \in \Gamma(U, \F)$:} \\
+      &\qquad\qquad
+          U \models \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
+      \Longleftrightarrow\ &
+        \text{for all open~$U \subseteq X$, sections $s, t \in \Gamma(U, \F)$:} \\
+      &\qquad\qquad
+          \text{for all open~$V \subseteq U$:} \\
+      &\qquad\qquad\qquad\qquad
+            \text{$\alpha_V(s|_V) = \alpha_V(t|_V)$ implies $s|_V = t|_V$} \\[0.5em]
+      \Longleftrightarrow\ &
+        \text{for all open~$U \subseteq X$, sections $s, t \in \Gamma(U, \F)$:} \\
+      &\qquad\qquad
+            \text{$\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.5em]
+      \Longleftrightarrow\ &
+        \text{$\alpha$ is a monomorphism of sheaves}
+    \end{align*}
+
+    \item The corner quotes ``$\speak{\ldots}$'' indicate that
+    translation into formal language is left to the reader.
+  \end{itemize}
+}
+
+\note{
+  \begin{itemize}
+    \item Similarly, we have (exercise, use the rules!):
+    \begin{align*}
+      &
+        X \models \speak{$\alpha$ is surjective} \\[0.5em]
+      \Longleftrightarrow\ &
+        X \models \forall s\?\G\_ \exists t\?\F\_ \alpha(t) = s \\[0.5em]
+      \Longleftrightarrow\ &
+        \text{$\alpha$ is an epimorphism of sheaves}
+    \end{align*}
+  \end{itemize}
+}
+
+\note{
+  \begin{itemize}
+    \item One can simplify the rules for often-occuring special cases:
+    \[ \renewcommand{\arraystretch}{1.3}\begin{array}{@{}lcl@{}}
+      U \models \forall s\?\F\_ \forall t\?\G\_ \varphi(s,t)
+      &\Longleftrightarrow&
+      \text{for all open~$V \subseteq U$,} \\
+      && \text{sections~$s \in \Gamma(V,\F)$, $t \in \Gamma(V,\G)$:} \\
+      &&\qquad\qquad V \models \varphi(s,t) \\\\
+      U \models \forall s\?\F\_ \varphi(s) \Rightarrow \psi(s)
+      &\Longleftrightarrow&
+      \text{for all open~$V \subseteq U$, sections~$s \in \Gamma(V,\F)$:} \\
+      &&\qquad\qquad \text{$V \models \varphi(s)$ implies $V \models \psi(s)$} \\\\
+      U \models \exists!s\?\F\_ \varphi(s)
+      &\Longleftrightarrow&
+      \text{for all open~$V \subseteq U$,} \\
+      &&
+      \text{there is exactly one section~$s \in \Gamma(V,\F)$ with:} \\
+      &&\qquad\qquad V \models \varphi(s)
+    \end{array} \]
+  \end{itemize}
+}
+
+\note{
+  \begin{itemize}
+    \item \begin{justify}One can extend the language to allow for \emph{unbounded}
+    quantification ($\forall A$ vs.~$\forall a \in A$), see Shulman's
+    stack semantics~\cite{shulman:stack}. This is needed to formulate
+    universal properties internal to~$\Sh(X)$, for instance.\end{justify}
+
+    \item \begin{justify}One can further extend the language to be able to talk
+    about locally internal categories over~$\Sh(X)$ (in the sense of
+    Penon~\cite{penon:locint}, see for instance the appendix
+    of~\cite{johnstone:topos}): Then one can do category theory internal
+    to~$\Sh(X)$ using the internal language, see~\cite{blechschmidt:msc}.
+
+    This specific approach is, as far as I am aware, original work. But of
+    course, internal category theory has been done for a long time, see for
+    instance the textbook~\cite{johnstone:elephant}; also
+    cf.~\cite{chapman:rowbottom:relcat}.\end{justify}
+  \end{itemize}
+}
+
+\frame[t]{\frametitle{The internal language of~$\Sh(X)$}
+  \begin{block}{Crucial properties}
+  \begin{itemize}
+    \item \hil{Locality:} If~$U = \bigcup_i U_i$, then
+      $U \models \varphi$ iff $U_i \models
+      \varphi$ for all~$i$.
+    \item \hil{Soundness:} If~$U \models \varphi$ and $\varphi$
+    implies $\psi$ \only<1>{constructively}\only<2->{\pointthis{constructively}{
+      no $\varphi \vee \neg\varphi$,\ \
+      no $\neg\neg\varphi \Rightarrow \varphi$,\ \
+      no AxC
+    }}, \\[0.1cm]
+    then~$U \models \psi$.
+  \end{itemize}
+  \end{block}
+  \vspace{0.5cm}
+
+  \vfill
+
+  \pause\pause
+  \begin{block}{A simple look at the constructive nature}
+  \begin{itemize}
+    \item $U \models f = 0$
+      \tabto{3.05cm} iff $f|_U = 0 \in \Gamma(U, \F)$.
+    \item $U \models \neg\neg(f = 0)$
+      \tabto{3.05cm} iff~$f = 0$ on a dense open subset of~$U$.
+  \end{itemize}
+  \end{block}
+}
+
+\note{%
+  \vspace{-0.5em}%
+  \begin{center}\large\bf Why is constructive mathematics
+  interesting?\end{center}
+  \vspace{-1em}
+  \begin{itemize}
+    \item The internal logic of most topoi is constructive.
+
+    \item \begin{justify}From a constructive proof of a statement, it's always possible to
+    mechanically extract an \emph{algorithm} witnessing its truth. For example:
+    A proof of the infinitude of primes gives rise to an algorithm
+    which actually computes infinitely many primes (outputting one at a
+    time, never stopping).\end{justify}
+
+    \item \begin{justify}By the celebrated \emph{Curry--Howard correspondence},
+    constructive truth of a formula is equivalent to the existence of a
+    program of a certain type associated to the formula.\end{justify}
+
+    \item \begin{justify}In constructive mathematics, one can experiment with (and draw
+    useful conclusions also holding in a usual sense) \emph{anti-classical
+    dream axioms}, for instance the one of synthetic differential geometry:\end{justify}
+    \begin{quote}All functions~$\RR \to \RR$ are smooth.\end{quote}
+
+    \item \begin{justify}Constructive accounts of classical theories are sometimes more
+    elegant or point out some minor but interesting points which are not appreciated by a
+    classical perspective.\end{justify}
+
+    \item \begin{justify}The philosophical question on the \emph{meaning of truth} is
+    easier to tackle in constructive mathematics.\end{justify}
+  \end{itemize}
+}
+
+\note{%
+  \vspace{-0.5em}%
+  \begin{center}\large\bf Three rumours about constructive mathematics\end{center}
+  \vspace{-1em}
+
+  \begin{enumerate}
+    \item \begin{justify}There is a false rumour about constructive mathematics, namely that
+    the term \emph{contradiction} is generally forbidden. This is not the
+    case, one has to distinguish between
+    \begin{itemize}
+      \item a true proof by contradiction: ``Assume~$\varphi$ were false.
+      Then \ldots, contradiction. So~$\varphi$ is in fact true.''
+    \end{itemize}
+    which constructively is only a proof of the weaker
+    statement~$\neg\neg\varphi$, and
+    \begin{itemize}
+      \item a proof of a negated formula: ``Assume~$\psi$ were true. Then
+      \ldots, contradiction. So~$\neg\psi$ holds.''
+    \end{itemize}
+    which is a perfectly fine proof of~$\neg\psi$ in constructive mathematics.\end{justify}
+
+    \item \begin{justify}There is a similar rumour that constructive mathematicians \emph{deny}
+    the law of excluded middle. In fact, one can constructively prove that there is no
+    counterexample to the law: For any formula~$\varphi$, it holds
+    that~$\neg\neg(\varphi \vee \neg\varphi)$.
+
+    In constructive mathematics, one merely doesn't \emph{use} the law of
+    excluded middle. (Only in concrete models, for example as provided by the
+    internal universe of the sheaf topos on a non-discrete topological space,
+    the law of excluded middle will actually be refutable.)\end{justify}
+  \end{enumerate}
+}
+
+\note{
+  \begin{enumerate}
+    \item[3.] \begin{justify}There is one last false rumour about constructive mathematics: Namely
+    that most of mathematics breaks down in a constructive setting. This is
+    only true if interpreted naively: Often, already very small changes to the
+    definitions and statements (which are classically simply equivalent
+    reformulations) suffice to make them constructively acceptable.
+
+    In other cases, adding an additional hypotheses, which is classically
+    always satisfied, is necessary (and interesting). Here is an example: In
+    constructive mathematics, one can not show that any inhabited subset of the
+    natural numbers possesses a minimal element. [One can also not show the
+    negation -- recall the previous false rumour.] But one can show (quite
+    easily, by induction) that any inhabited and \emph{detachable} subset
+    of the natural numbers possesses a minimal element. A subset~$U \subseteq
+    \NN$ is detachable iff for any number~$n \in \NN$, it holds that~$n
+    \in U$ or~$n \not\in U$.
+
+    This has a computational interpretation: Given an arbitrary inhabited subset~$U
+    \subseteq \NN$, one cannot algorithmically find its minimal element. But it
+    \emph{is} possible if one has an algorithmic \emph{test of membership}
+    for~$U$.\end{justify}
+  \end{enumerate}
+
+  \begin{justify}See~\cite{stanford:constr,dalen:int,troelstra:dalen:constr}
+  for references about constructive mathematics. The blog~\cite{bauer:blog} is
+  also very informative.\end{justify}
+}
+
+\section{The little Zariski topos of a scheme}
+\subsection{Basic look and feel}
+
+\frame[t]{\frametitle{The little Zariski topos}
+  \begin{block}{Definition}The \hil{little Zariski topos} of a scheme~$X$ is the
+  category~$\Sh(X)$ of set-valued sheaves on~$X$.\end{block}
+
+  \begin{block}{Basic look and feel}
+  \begin{itemize}
+    \item Internally, the structure sheaf~$\O_X$ looks like \[ \text{
+    an ordinary ring.} \]
+    \item Internally, a sheaf of~$\O_X$-modules looks like \[
+    \text{an ordinary module on that ring.} \]
+  \end{itemize}
+  \end{block}
+}
 
@@
 \note{\begin{itemize}\justifying
-  \item The translation of internal statements into external ones is
-  facilitated by an easy mechanical procedure which one quickly grows
-  accustomed to. \hyperlink{kripke-joyal}{\beamergotobutton{See more details.}}
   \item The internal language of a sheaf topos of a $\mathrm{T}_1$-space is
@@
 
-\section{Building and using a dictionary}
+\subsection{Building and using a dictionary}
 
@@
   \end{center}
-
-  \visible<2>{\begin{tikzpicture}[overlay]
-    \draw[fill=white, draw=white, opacity=0.9] (0,0) rectangle (\paperwidth,7.1);
-    \node[anchor=south west,inner sep=0] (image) at (1.8,1.3) {
-      \includegraphics[width=0.7\textwidth]{images/steven-kleiman-misconceptions-about-kx}
-    };
-  \end{tikzpicture}}
 \end{frame}
@@
 
-  Further examples:
-  \begin{itemize}
-    \item Characterizing quasicoherence internally
-    \item Understanding spreading of properties in a logical way
-    \item Constructing the relative spectrum internally
+  Further topics in the little Zariski topos:
+  \begin{itemize}
+    \item The sheaf $\K_X$ of rational functions
+    \item Transfer principles~$M \leftrightarrow M^\sim$
+    \item The curious role of affine open subsets
+    \item Quasicoherence
+    \item Spreading from points to neighbourhoods
+    \item The relative spectrum
   \end{itemize}
@@
 }
+
+\subsection{The sheaf of rational functions}
+
+\begin{frame}\frametitle{The sheaf of rational functions}
+  \begin{block}{Classical definition}
+  The sheaf~$K_X$ of \hil{rational functions} on a scheme~$X$ is the
+  sheafification of the presheaf
+  \[ U \subseteq X \longmapsto \Gamma(U, \O_X)[S(U)^{-1}], \]
+  where
+  \[ S(U) = \{ s \in \Gamma(U, \O_X) \,|\, \text{$s \in \O_{X,x}$ is
+  regular for all~$x \in U$} \}. \]
+  \end{block}
+
+  \begin{block}{Internal definition}
+  $K_X$ is the total quotient ring of~$\O_X$.
+  \end{block}
+
+  \visible<2>{\begin{tikzpicture}[overlay]
+    \draw[fill=white, draw=white, opacity=0.9] (-1,0) rectangle (\paperwidth,7.1);
+    \node[anchor=south west,inner sep=0] (image) at (1.8,1.3) {
+      \includegraphics[width=0.7\textwidth]{images/steven-kleiman-misconceptions-about-kx}
+    };
+  \end{tikzpicture}}
+\end{frame}
+
+\section{Transfer principles}
+
+\begin{frame}\frametitle{Transfer principles}
+  \hil{Question:} How do the properties of
+  \begin{itemize}
+  \item an~$A$-module~$M$ in~$\Set$ and
+  \item the~$\O_X$-module~$M^\sim$ in~$\Sh(X)$, where~$X = \Spec A$,
+  \end{itemize}
+  relate?
+  \pause
+  \vfill
+
+  \hil{Observation:} $M^\sim = \underline{M}[\F^{-1}]$, where
+  \begin{itemize}
+  \item $\underline{M}$ is the constant sheaf with stalks~$M$ on~$X$ and
+  \item $\F \hookrightarrow \underline{A}$ is the \hil{generic filter}.
+  \end{itemize}
+  \pause
+  \vfill
+
+  \hil{Answer:} $M^\sim$ inherits those properties of $M$ which are stable
+  under localization.
+\end{frame}
+
+\begin{frame}\frametitle{The curious role of affine open subsets}
+  \hil{Question:} Why do the following identities hold, for quasicoherent
+  sheaves and affine open subsets~$U$?
+  \begin{align*}
+    \Gamma(U, \E/\F) &= \Gamma(U, \E) / \Gamma(U, \F) \\
+    \Gamma(U, \E \otimes_{\O_X} \F) &= \Gamma(U, \E) \otimes_{\Gamma(U,\O_X)} \Gamma(U, \F) \\
+    \Gamma(U, \E_\mathrm{tors}) &= \Gamma(U, \E)_\mathrm{tors} \\
+    \Gamma(U, \K_X) &= \operatorname{Quot} \O_X(U) \\
+    &\ \ \vdots
+  \end{align*}
+  \pause
+  \hil{Answer:} Because localization commutes with quotients, tensor products,
+  torsion submodules (sometimes), \ldots
+\end{frame}
 
@@
 \begin{frame}\frametitle{The $\Diamond$-translation}
-  Let~$\E_\Diamond \hookrightarrow \E$ be a subtopos given by a \pointthis{local
-  operator}{$\Diamond : \Omega_\E \to \Omega_\E$}~$\Diamond$. Then
+  Let~$\E_\Diamond \hookrightarrow \E$ be a subtopos given by a local
+  operator. Then
   \[ \E_\Diamond \models \varphi \qquad\text{iff}\qquad
@@
 % XXX: Give details.
+
+\section{The big Zariski topos of a scheme}
+
+\subsection{Basic look and feel}
+
+\begin{frame}\frametitle{The big Zariski topos}
+  \begin{block}{Definition}The \hil{big Zariski topos} $\Zar(S)$ of a scheme~$S$ is the
+  category $\Sh(\Sch/S)$. It consists of certain functors
+  $(\Sch/S)^\op \to \Set$.
+  \end{block}
+
+  \begin{block}{Basic look and feel}
+  \begin{itemize}
+    \item For each~$S$-scheme~$X$, its functor of points $\ull{X} =
+    \Hom_S(\cdot,X)$ is an object of~$\Zar(S)$. It feels like \hil{the set of
+    points} of~$X$.
+    \item Internally, $\affl$ (given by $\affl(X) = \Gamma(X,\O_X)$)
+    looks like a field:
+    \[ \Zar(S) \models \forall x\?\affl\_ x \neq 0 \Longrightarrow \speak{$x$ inv.} \]
+  \end{itemize}
+  \end{block}
+
+  \note{
+    \begin{itemize}
+      \item \begin{justify}The overcategory~$\Sch/S$ becomes a Grothendieck site by declaring
+      families of jointly surjective open immersions to be covers. See for
+      instance the excellent Stacks project~\cite{stacks} for details.\end{justify}
+
+      \item \begin{justify}Working in~$\Zar(S)$ amounts to incorporating the
+      philosophy of describing schemes by their functors of points into one's
+      mathematical language.\end{justify}
+
+      \item \begin{justify}Explicitly, the functor~$\ull{X}$ is given
+      by~$\ull{X}(T) = \Hom_S(T,X)$ for~$S$-schemes~$T$. Because the Zariski
+      site is \emph{subcanonical}, this functor is always a sheaf.\end{justify}
+
+      \item \begin{justify}The object~$\ull{S}$ looks like an one-element set
+      from the internal universe. This is to be expected.\end{justify}
+    \end{itemize}
+  }
+\end{frame}
+
+\note{
+  \begin{itemize}
+    \item \begin{justify}Hakim worked out a theory of schemes internal to topoi (but
+    without using the internal language), see~\cite{hakim:rel}.\end{justify}
+
+    \item \begin{justify}The internal language of~$\Zar(\Spec A)$ is related to Coquand's
+    program about dynamical methods in algebra,
+    see~\cite{coquand:completeness,coquand:logical,coquand:effective}.\end{justify}
+
+    \item \begin{justify}The observation that~$\affl$ is internally a field is due to
+    Kock~\cite{kock:univproj} (in the case~$S = \Spec\ZZ$) and implies a
+    curious meta-theorem:
+
+    Because $\Zar(\Spec\ZZ)$ is the \emph{classifying topos} for the theory
+    of local rings, any statement about local rings which is of a certain
+    logical form holds for the \emph{universal model}~$\ull{\AA}^1_{\Spec\ZZ}$
+    iff it holds for any local ring (in any universe, particularly~$\Set$).
+
+    Therefore, in proving a statement of such a form about arbitrary local
+    rings, one may assume that they even fulfil the field condition.
+
+    There is a similar story for local~$A$-algebras, see~\cite{anel:fact}.
+    See~\cite{wraith:int} for a short exposition on the usefulness of
+    classifying topoi and universal models.\end{justify}
+  \end{itemize}
+}
+
+\note{
+  \begin{itemize}
+    \item The affine line fulfils the axiom
+      \[ \Zar(S) \models \speak{every function~$\affl \to \affl$ is a
+      polynomial}. \]
+
+      Compare with the axiom of synthetic differential geometry:
+      \[ \text{every function~$\RR \to \RR$ is smooth.} \]
+
+      See~\cite{kock:sdg}.
+  \end{itemize}
+}
+
+\subsection{Group schemes}
+
+\begin{frame}\frametitle{Group schemes}
+  \hil{Motto:} Internal to~$\Zar(S)$, group schemes look like ordinary groups.
+  \bigskip
+
+  \small\hspace*{-0.6cm}%
+  \begin{tabular}{lll}
+    \toprule
+    group scheme & internal definition & functor of points: $X \mapsto \ldots$ \\\midrule
+    $\GG_\text{a}$ & $\affl$ (as additive group) & $\Gamma(X, \O_X)$ \\[0.5em]
+    $\GG_\text{m}$ & $\{ x \? \affl \,|\, \speak{$x$ inv.} \}$ & $\Gamma(X, \O_X)^\times$ \\[0.5em]
+    $\mu_n$ & $\{ x \? \affl \,|\, x^n = 1 \}$ & $\{ f \in \Gamma(X, \O_X) \,|\, f^n = 1 \}$ \\[0.5em]
+    $\GL_n$ & $\{ M \? \affl^{n \times n} \,|\, \speak{$M$ inv.} \}$ & $\GL_n(\Gamma(X, \O_X))$ \\
+    \bottomrule
+  \end{tabular}
+  \vspace{2em}
+\end{frame}
+
+\subsection{Some internal constructions}
+
+\begin{frame}\frametitle{Some internal constructions}
+  \begin{itemize}
+  \item The functor of points of~$\PP^n_S$ has the internal description
+  \[ \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\,
+    x_0 \neq 0 \vee \cdots \vee x_n \neq 0 \}/\text{scaling}. \]
+  \pause
+
+  \item
+  Let~$\A$ be an~$\O_S$-algebra. This induces an $\affl$-algebra $\A^\sim$
+  internal to~$\Zar(S)$. The functor of points of~$\RelSpec_S \A$ has the
+  internal description
+  \[ \Hom_{\mathrm{Alg}(\affl)}(\A^\sim, \affl). \]
+  \visible<2>{\begin{tikzpicture}[overlay]
+    \draw[fill=pink, draw=pink, opacity=0.99] (2.5,0.4) rectangle (7.4,1.3);
+  \end{tikzpicture}}
+  \pause\pause
+
+  \item Let~$X$ be an~$S$-scheme. The functor of points of $\RelSpec_X
+  \Omega^1_{X/S} \to X \to S$ has the internal description
+  \[ \Hom(\Delta, \ull{X}), \]
+  where~$\Delta = \{ \varepsilon : \affl \,|\, \varepsilon^2 = 0 \}$.
+  \end{itemize}
+\end{frame}
 
@@
 
-\newcommand{\Ll}{:\Longleftrightarrow}
 \begin{frame}\frametitle{The rules of the Kripke--Joyal semantics}\justifying

### 2016-05-19 slides-augsburg.tex
Slides mostly finished

@@
 \usepackage[protrusion=true,expansion=true]{microtype}
+
+\setlength\parskip{\medskipamount}
+\setlength\parindent{0pt}
+
+\title{Using the internal language of toposes in algebraic geometry}
+\author{Ingo Blechschmidt}
+\date{May 19th, 2016}
+
+\usetheme{Warsaw}
+\usecolortheme{seahorse}
+%\usefonttheme{default}?
+%\usepackage{kurier}?
+\usefonttheme{serif}
+\usepackage[T1]{fontenc}
+\usepackage{libertine}
+%\usepackage{mathpazo}
+\useinnertheme{rectangles}
 
@@
 \newcommand{\Sch}{\mathrm{Sch}}
+\newcommand{\LRS}{\mathrm{LRS}}
 \newcommand{\Hom}{\mathrm{Hom}}
 \DeclareMathOperator{\Spec}{Spec}
+\newcommand{\lra}{\longrightarrow}
 \newcommand{\RelSpec}{\operatorname{\text{\ul{$\mathrm{Spec}$}}}}
@@
 \newcommand{\Ll}{\vcentcolon\!\Longleftrightarrow}
-
-\setlength\parskip{\medskipamount}
-\setlength\parindent{0pt}
-
-\title{Using the internal language of toposes in algebraic geometry}
-\author{Ingo Blechschmidt}
-\date{May 19th, 2016}
-
-\usetheme{Warsaw}
-\usecolortheme{seahorse}
-%\usefonttheme{default}?
-%\usepackage{kurier}?
-\usefonttheme{serif}
-%\usepackage{libertine}?
-\usepackage{mathpazo}
-\useinnertheme{rectangles}
 
@@
     (#1) {#1};\node[overlay,rectangle callout,%
-    callout relative pointer={(0.5cm,1.2cm)},fill=blue!20] at ($(#1.north)+(-1.2cm,-1.8cm)$) {#2};}%
+    callout relative pointer={(-0.5cm,0.8cm)},fill=blue!20] at ($(#1.north)+(1.1cm,-1.4cm)$) {#2};}%
 }%
@@
 \frame[t]{\frametitle{The internal language of~$\Sh(X)$}
-  \begin{block}{Crucial properties}
-  \begin{itemize}
-    \item \hil{Locality:} If~$U = \bigcup_i U_i$, then
+  \begin{block}{Crucial property: Locality}
+    If~$U = \bigcup_i U_i$, then
       $U \models \varphi$ iff $U_i \models
       \varphi$ for all~$i$.
-    \item \hil{Soundness:} If~$U \models \varphi$ and $\varphi$
+  \end{block}
+
+  \begin{block}{Crucial property: Soundness}
+    If~$U \models \varphi$ and $\varphi$
     implies $\psi$ \only<1>{constructively}\only<2->{\pointthis{constructively}{
@@
       no AxC
-    }}, \\[0.1cm]
+    }},
     then~$U \models \psi$.
-  \end{itemize}
   \end{block}
-  \vspace{0.5cm}
 
@@
       finite locally free sheaf & finite free module \\
-      coherent sheaf & coherent module \\
+      % coherent sheaf & coherent module \\
       tensor product of sheaves & tensor product of modules \\
       rank function & minimal number of generators \\
-      sheaf of rational functions & total quotient ring of~$\O_X$ \\
+      % sheaf of rational functions & total quotient ring of~$\O_X$ \\
+      sheaf of Kähler differentials & module of Kähler differentials \\
+      dimension of $X$ & Krull dimension of $\O_X$ \\
       \bottomrule
@@
 
-\section{Transfer principles}
+\subsection{Transfer principles}
 
@@
   \item an~$A$-module~$M$ in~$\Set$ and
-  \item the~$\O_X$-module~$M^\sim$ in~$\Sh(X)$, where~$X = \Spec A$,
-  \end{itemize}
-  relate?
+  \item the~$\O_X$-module~$M^\sim$ in~$\Sh(X)$, where~$X = \Spec A$, relate?
+  \end{itemize}
   \pause
@@
   \end{itemize}
+  Note: $M$ and $\underline{M}$ share all first-order properties.
   \pause
@@
 
-  \hil{Answer:} $M^\sim$ inherits those properties of $M$ which are stable
-  under localization.
-\end{frame}
+  \centering
+  \hil{Answer:} $M^\sim$ inherits those properties of $M$ which are \\
+  \hil{stable under localization}.
+  \par
+\end{frame}
+
+\subsection{The curious role of affine open subsets}
 
@@
     \Gamma(U, \E \otimes_{\O_X} \F) &= \Gamma(U, \E) \otimes_{\Gamma(U,\O_X)} \Gamma(U, \F) \\
-    \Gamma(U, \E_\mathrm{tors}) &= \Gamma(U, \E)_\mathrm{tors} \\
-    \Gamma(U, \K_X) &= \operatorname{Quot} \O_X(U) \\
+    \Gamma(U, \E_\mathrm{tors}) &= \Gamma(U, \E)_\mathrm{tors} \quad\text{(sometimes)} \\
+    \Gamma(U, \K_X) &= \operatorname{Quot} \O_X(U) \quad\text{(sometimes)} \\
     &\ \ \vdots
@@
 
-\section{Quasicoherence of sheaves of modules}
+\subsection{Quasicoherence of sheaves of modules}
 
@@
 
-\section{Spreading of properties}
+\subsection{Spreading of properties}
 
@@
     (\exists x\?X\_ \varphi(x))^\Diamond &\defeqv \idiamond(\exists x\?X\_ \varphi^\Diamond(x))
-  \end{align*}}
-
-  \only<2->{
+  \end{align*}}%
+  \only<2->{%
     Let~$X$ be a scheme. Depending on~$\Diamond$,~$\Sh(X)
@@
 
-\section{The relative and internal spectrum}
+\subsection{The relative spectrum}
 
@@
   Define the frame of opens of~$\Spec A$ to be the frame of radical ideals
-  in~$A$.
-  \pause
-
-  This gives an internal description of
-  Monique Hakim's spectrum functor from ringed toposes to locally ringed
-  toposes.
+  in~$A$. Universal property:
+  \[ \Hom_{\mathrm{LRT}/\E}(T, \Spec A) \cong \Hom_{\mathrm{Ring}(\E)}(A, \mu_*\O_T) \]
+  for all locally ringed toposes~$T$ equipped with a geometric morphism~$T \xrightarrow{\mu} \E$.
 \end{frame}
@@
 \begin{frame}\frametitle{The relative spectrum}
-  Let~$X$ be a scheme and~$\O_X \xrightarrow{\varphi} \A$ be a quasicoherent algebra.
-  Can we describe~\hil{$\boldsymbol{\RelSpec_X \A}$}, a scheme over~$X$, internally?
+  Let~$X$ be a scheme and~$\O_X \xrightarrow{\varphi} \A$ be an algebra.
+  Can we describe~\hil{$\boldsymbol{\RelSpec_X \A}$} internally?
 
   Desired universal property:
-  \[ \Hom_{\Sch/X}(T, \RelSpec_X \A) \cong \Hom_{\mathrm{Alg}(\O_X)}(\A,
+  \[ \Hom_{\LRS/X}(T, \RelSpec_X \A) \cong \Hom_{\mathrm{Alg}(\O_X)}(\A,
   \mu_*\O_T) \]
-  for all~$X$-schemes~$T \xrightarrow{\mu} X$.
+  for all locally ringed spaces over~$X$.
   \pause
@@
 \begin{frame}\frametitle{The relative spectrum, reformulated}
-  Let~$B \to A$ be an algebra in topos.
+  Let~$B \to A$ be an algebra in a topos.
 
@@
   \begin{itemize}
-    \item For each~$S$-scheme~$X$, its functor of points $\ull{X} =
-    \Hom_S(\cdot,X)$ is an object of~$\Zar(S)$. It feels like \hil{the set of
+    \item For an~$S$-scheme~$X$, its functor of points \[ \ull{X} =
+    \Hom_S(\cdot,X) \] is an object of~$\Zar(S)$. It feels like \hil{the set of
     points} of~$X$.
@@
 
-\begin{frame}
-  \sloganwithoutborder{\hil{Understand notions and statements of algebraic geometry
-  as notions and statements of algebra internal to appropriate toposes.}}
+\subsection{The étale subtopos}
+
+\begin{frame}\frametitle{The étale subtopos}
+  Recall that the \hil{Kummer sequence} is not exact in~$\Zar(S)$ at the third
+  term:
+  \[ 1 \lra \mu_n \lra \GG_\text{m} \stackrel{(\cdot)^n}{\lra} \GG_\text{m} \lra 1 \]
+  But we have:
+  \[ \Zar(S) \models \forall f\?(\affl)^\times\_ \Diamond_\text{ét}(\exists
+  g\?(\affl)^\times\_ f = g^n), \]
+  where~$\Diamond_\text{ét}$ is such that $\Zar(S)_{\Diamond_\text{ét}}
+  \hookrightarrow \Zar(S)$ is the \hil{big étale topos}. It's the largest
+  subtopos of~$\Zar(S)$ where
+  \[ \speak{$\affl$ is separably closed} \]
+  holds (Wraith, Felix).
+\end{frame}
+
+\begin{frame}\frametitle{Comparing the little and the big toposes}
+  \begin{itemize}
+  \item From the point of view of~$\Sh(S)$, the big Zariski topos is $\Zar(\O_S | \O_S)$,
+  the classifying topos of local~$\O_S$-algebras which are local
+  over~$\O_S$.
+
+  \item From the point of view of~$\Zar(S)$, the little Zariski topos is the
+  largest subtopos where $\flat \affl \to \affl$ is bijective.
+  \begin{align*}
+    (\flat \affl)(X \xrightarrow{\mu} S) &= \Gamma(X, \mu^{-1} \O_S) \\
+    \affl(X \xrightarrow{\mu} S) &= \Gamma(X, \O_X)
+  \end{align*}
+  \end{itemize}
+\end{frame}
+
+\section{Open tasks}
+
+\begin{frame}\frametitle{Semi-open and open tasks}
+  \begin{itemize}
+  \item Characterize quasicoherence in the big Zariski topos.
+  \item Understand how to work with~${\flat} \dashv \sharp$.
+  \item Do cohomology in the little Zariski topos; exploit that higher direct
+  images look like ordinary sheaf cohomology from the internal point of view.
+  \item Do cohomology in the big Zariski topos.
+  \item Understand more subtoposes of the big Zariski topos.
+  \item Derive suitable axioms for synthetic algebraic geometry.
+  \end{itemize}
 
@@
   \par\medskip\vspace{-0.1em}}
-
-  \begin{itemize}
-    \item Simplify proofs and gain conceptual understanding.
-    \item Understand relative geometry as absolute geometry.
-    \item Develop a synthetic account of scheme theory.
-    \item Contribute to constructive algebra.
-  \end{itemize}
-
-  \centering
-  \hil{\href{http://tiny.cc/topos-notes}{http://tiny.cc/topos-notes}} \\
-  \begin{minipage}{1.0\textwidth}\justifying
-    \scriptsize
-    spreading of properties,
-    general transfer principles,
-    applications to constructive algebra,
-    quasicoherence,
-    internal Cartier divisors,
-    pullback along immersions $=$ internal sheafification,
-    scheme dimension $=$ internal Krull dimension of~$\O_X$,
-    dense $=$ not not,
-    modal operators,
-    relative spectrum,
-    other toposes,
-    étale topology,
-    group schemes $=$ groups,
-    \ldots
-  \end{minipage}
-  \par
-\end{frame}
-% Fun fact:
-% Very naive definition of P^n works internal to the big Zariski topos.
+\end{frame}
 
 \backupstart
-
-\begin{frame}[plain,c]
-  \centering
-  \includegraphics[scale=0.5]{images/sheafification-man}
-  \sffamily
-
-  You should totally look up:
-
-  \hil{The Adventures of Sheafification Man}
-  \par
-\end{frame}
-
 \appendix
@@
 \section{Details on the internal language}
-
-\begin{frame}[label=kripke-joyal]\frametitle{More on the internal language}\justifying
-  More generally, for an object~$U$ of a topos~$\E$, we define the meaning of
-  \[ U \models \varphi \qquad\text{($\varphi$ holds on~$U$)}. \]
-  Writing~``$\E \models \varphi$'' is then an abbreviation for~``$1 \models
-  \varphi$'', where~``$1$'' denotes the terminal object of~$\E$.
-
-  In addition to soundness with respect to intuitionistic logic, the internal
-  language has the following two important properties:
-  \begin{itemize}
-    \item \hil{Monotonicity:} If~$p : V \to U$ is an arbitrary morphism and~$U
-    \models \varphi$, then also~$V \models \varphi$.
-    \item \hil{Locality:} If~$p : V \to U$ is an epimorphism and~$V \models
-    \varphi$, then also~$U \models \varphi$.
-  \end{itemize}
-\end{frame}
-
-\begin{frame}\frametitle{The rules of the Kripke--Joyal semantics}\justifying
-  In the special case that~$\E = \Sh(X)$ is the topos of sheaves on a
-  topological space (or locale)~$X$, the rules of the Kripke--Joyal semantics
-  look as follows. We tersely write~``$U \models \varphi$'' instead
-  of~``$\Hom(\ull{\ \ }, U) \models \varphi$ for open subsets~$U \subseteq X$.
-  \scriptsize
-  \[ \renewcommand{\arraystretch}{1.3}\begin{array}{@{}lcl@{}}
-    U \models f = g \? \F &\Ll& f|_U = g|_U \in \F(U) \\
-    U \models \varphi \wedge \psi &\Ll&
-      \text{$U \models \varphi$ and $U \models \psi$} \\
-    U \models \varphi \vee \psi &\Ll&
-      \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt} \\
-    && \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
-    && \quad\quad \text{$U_i \models \varphi$ or $U_i \models \psi$} \\
-    U \models \varphi \Rightarrow \psi &\Ll&
-      \text{for all open~$V \subseteq U$: }
-    \text{$V \models \varphi$ implies $V \models \psi$} \\
-    U \models \forall f \? \F\_ \varphi(f) &\Ll&
-      \text{for all sections~$f \in \F(V), V \subseteq U$: $V \models
-      \varphi(f)$} \\
-    U \models \exists f \? \F\_ \varphi(f) &\Ll&
-      \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
-    && \quad\quad \text{there exists~$f_i \in \F(U_i)$ s.\,th.
-    $U_i \models \varphi(f_i)$}
-  \end{array} \]
-\end{frame}
-
-\note{\begin{itemize}\justifying
-  \item The rules are not all arbitrary: Rather, they are very finely concerted
-  to make the crucial properties about the internal language (monotonicity,
-  locality, soundness with respect to intuitinistic logic) true.
-
-  \item If~$\F$ is an object of~$\Sh(X)$, we write ``$f:\F$'' instead
-  of~``$f\in\F$'' to remind us that~$\F$ is not really (externally) a set
-  consisting of elements, but that we only pretend this by using the internal
-  language.
-
-  \item There are two further rules concerning the constants~$\top$
-  and~$\bot$ (truth resp. falsehood):
-    \[ \renewcommand{\arraystretch}{1.3}\begin{array}{@{}lcl@{}}
-      U \models \top &\Ll& U = U \text{ (always fulfilled)}\\
-      U \models \bot &\Ll& U = \emptyset
-    \end{array} \]
-
-  \item Negation is defined as
-  \[ \neg\varphi :\equiv (\varphi \Rightarrow \bot). \]
-\end{itemize}}
-
-\note{\justifying
-  The alternate definition ``$U \models \varphi \vee \psi :\Leftrightarrow
-  \text{$U \models \varphi$ or $U \models \psi$}$'' would not be local.
-  \par
-}
 
@@
   \end{itemize}
-\end{frame}
-
-\section{Transfer principles}
-
-\begin{frame}\frametitle{Transfer principles}\justifying
-  Let~$M$ be an~$A$-module. How do~$M$ and the sheaf~$M^{\sim}$ on~$\Spec A$
-  relate?
-
-  Observe that $M^{\sim} \cong \underline{M}[\F^{-1}]$ is the localization of~$M$ at
-  the \hil{generic prime filter} and that~$M$ shares all first-order properties
-  with the constant sheaf of modules~$\underline{M}$. Therefore:
-  \begin{center}
-    $M^{\sim}$ inherits all those properties of~$M$ which are \\
-    \hil{stable under localization}.
-  \end{center}
-  Examples: finitely generated, free, flat, \ldots
-
-  A converse holds as well, suitably formulated.
 \end{frame}
@@
 
-\section{The big Zariski topos}
-
-\begin{frame}\frametitle{The big Zariski topos}
-  Let~$X$ be a scheme. The \hil{big Zariski topos} is the topos of sheaves
-  on~$\Sch/X$ with respect to the Zariski topology. From its point of view,
-  \ldots
-
-  \begin{itemize}
-    \item \ldots{} $X$-schemes look just like sets,
-    \item \ldots{} $\mathbb{P}^n_X$ is given by the naive expression
-    \[ \{ (x_0,\ldots,x_n) \,|\, x_1 \neq 0 \vee \cdots \vee x_n \neq 0
-    \}/\text{(rescaling)}, \]
-    \item \ldots{} the cotangent ``bundle'' of an~$X$-scheme~$T$ is
-    \[ \text{the set of maps $\Delta \to \ull{T}$,} \]
-    where~$\Delta = \{ \varepsilon \in \ull{\AA}^1_X \,|\, \varepsilon^2 = 0
-    \}$.
-    \item \ldots{} affinity is a ``double dual condition'', and
-    \item \ldots{} the étale topology is the coarsest topology~$\Diamond$ s.\,th.
-    \[ \forall f : \ull{\AA}^1_X[T]\_\
-      \text{$f$ is monic separable} \Rightarrow
-      \Diamond(\exists t : \ull{\AA}^1\_ f(t) = 0). \]
-  \end{itemize}
-\end{frame}
-
-\note{
-  \begin{itemize}\justifying
-    \item The functor of points of~$\AA^1_X$, that is
-    \[ \ull{\AA}^1_X : (T/X) \longmapsto \O_T(T), \]
-    looks like a local ring and indeed like a field from the internal point of
-    view, in the sense that
-    \[ \forall f\?\ull{\AA}^1_X\_ \neg(f = 0) \Rightarrow \text{$f$
-    invertible}. \]
-    \item Let~$\A$ be a quasicoherent~$\O_X$-algebra. Let~$\E$ be the
-    induced~$\ull{\AA}^1_X$-algebra given by~$\E(T \xrightarrow{\mu} X) \defeq
-    (\mu^* \A)(T)$. Then the internal Hom set~$[\E,
-    \ull{\AA}^1_X]_{\ull{\AA}^1_X}$ of~$\ull{\AA}^1_X$-algebra morphisms is the
-    functor of points of~$\RelSpec_X(\A)$.
-    \item Let~$\mu : T \to X$ be quasicompact and quasiseparated.
-    Then~$\mu$ is affine iff, from the internal point of view, the map
-    \[ \ull{T} \longrightarrow [ [\ull{T},\ull{\AA}^1_X]^\flat, \ull{\AA}^1_X ]_{\ull{\AA}^1_X},\ x \longmapsto
-    \ull{\ \ }(x) \]
-    into the ``double dual'' is bijective.
-  \end{itemize}
-}
-
-\note{
-  \begin{itemize}\justifying
-    \item Describing the functor of points of the projective space was
-    suggested by Zhen Lin Low.
-    \item The statement on the étale topology follows from Gavin Wraith's
-    article \emph{Generic Galois theory of local rings}.
-  \end{itemize}
-}
-
 \backupend
@@
 \end{document}
-
-1. Begin with toposes and their internal logic.
-   Gauge the audience's familiarity with sheaves!
-
-2. State the MO question by Steven Gubkin.
-
-3. Discuss basics (include praise for Mike Shulman).
-
-4. Explain the Diamond-translation.
-
-5. Finish with outlook: more complex internal statements,
-   synthetic algebraic geometry, relative algebraic geometry.
-
-Do not include:
-* Quasicoherence condition
-* Relative spectrum
-
-Mention in passing that the Knaster--Tarski theorem is useful in this context.

### 2016-05-23 slides-ihes2015.tex
typo

@@
 
-  Topos à l'IHES \\
+  Topos à l'IHÉS \\
   November 27th, 2015
@@
   algebra} explored by Coquand, Coste, Lombardi, Roy, and others. Their
-  approach applies more generally.
+  approach applies more versatile.
   \par
@@
   look as follows. We tersely write~``$U \models \varphi$'' instead
-  of~``$\Hom(\ull{\ \ }, U) \models \varphi$ for open subsets~$U \subseteq X$.
+  of~``$\Hom(\ull{\ \ }, U) \models \varphi$'' for open subsets~$U \subseteq X$.
   \scriptsize
@@
     &\qquad\qquad
-        \alpha|_{U_i}(s_i) = t|_{U_i} \\[0.5em]
+        \alpha_{U_i}(s_i) = t|_{U_i} \\[0.5em]
     \Longleftrightarrow\ &

### 2016-05-23 slides-cambridge2016.tex
Slides almost finished

--- 2016-05-19 slides-augsburg.tex
@@
-\documentclass[12pt,utf8,notheorems,compress,t]{beamer}
+\documentclass[12pt,utf8,notheorems,compress,t,draft]{beamer}
 \usepackage{etex}
@@
 \author{Ingo Blechschmidt}
-\date{May 19th, 2016}
+\date{May 24th, 2016}
 
@@
 
+\newcommand{\fmini}[2]{\framebox{\begin{minipage}{#1}#2\end{minipage}}}
+\makeatletter
+\def\underunbrace#1{\mathop{\vtop{\m@th\ialign{##\crcr
+      $\hfil\displaystyle{#1}\hfil$\crcr\noalign{\kern3\p@\nointerlineskip}
+      \crcr\noalign{\kern3\p@}}}}\limits}
+\def\overunbrace#1{\mathop{\vbox{\m@th\ialign{##\crcr\noalign{\kern3\p@}
+      \crcr\noalign{\kern3\p@\nointerlineskip}
+      $\hfil\displaystyle{#1}\hfil$\crcr}}}\limits}
+\makeatother
+
 \newenvironment{changemargin}[2]{%
@@
     \setlength{\fboxsep}{8pt}%
-    {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal text}\parbox{0.87\textwidth}{#1}}}%
+    {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal text}\parbox{0.84\textwidth}{#1}}}%
   \end{center}%
@@
   \scriptsize
-  am Tag, an dem Marc im Curry Club vorträgt
-  \par
-\end{frame}
+  Ingo Blechschmidt \\
+  University of Augsburg
+  \medskip
+
+  Category Theory Seminar \\
+  Centre for Mathematical Sciences \\
+  University of Cambridge \\
+  \ \\
+  May 24th, 2016
+  \par
+\end{frame}
+
+{\usebackgroundtemplate{\includegraphics[width=\paperwidth]{images/cambridge-university-botanic-garden}}
+\begin{frame}[plain]
+  \vspace{\textheight}\vspace{-1em}
+  \centering
+  \emph{Cambridge University Botanic Garden} \\
+  \par
+\end{frame}}
+
+\note{
+  Photo source:
+  \href{http://www.stpaulscambridge.org.uk/classes-and-room-hire/activities-and-events/newtown-news/}{Newtown
+  News}
+}
 
@@
 
+\subsection{What is a scheme?}
+
+\begin{frame}\frametitle{What is a scheme?}
+  \begin{itemize}
+  \item A \hil{manifold} is a space which is \hil{locally isomorphic} to some open
+  subset of some~$\RR^n$.
+  \medskip
+  \item A \hil{scheme} is a space which is \hil{locally isomorphic} to the
+  \hil{spectrum of some (commutative) ring}:
+  \[ \Spec A \defeq \{ \ppp \subseteq A \,|\, \text{$\ppp$ is a prime ideal} \} \]
+  \item By \hil{space} we mean: topological space~$X$ equipped with a local
+  sheaf~$\O_X$ of rings.
+  \end{itemize}
+
+  \centering
+  \begin{columns}
+    \begin{column}{0.45\textwidth}
+      \centering
+      \includegraphics[height=0.25\textheight]{images/manifold} \\
+      \tiny a manifold
+      \par
+    \end{column}
+    \begin{column}{0.45\textwidth}
+      \centering
+      \includegraphics[height=0.25\textheight]{images/mumfords-treasure-map} \\
+      \tiny Mumford's treasure map of~$\Spec \ZZ[X]$
+      \par
+    \end{column}
+  \end{columns}
+\end{frame}
+
+\note{\justifying
+  A \emph{sheaf of rings} on a topological space~$X$ is a ring object
+  in~$\Sh(X)$, the category of set-valued sheaves on~$X$.
+
+  A sheaf~$\O_X$ of rings is \emph{local} if and only if all the
+  stalks~$\O_{X,x}$ are local rings. Why not demand that the sets of
+  sections~$\O_X(U)$ are local rings? This has a geometric meaning, but can
+  also be motivated from a logical point of view: A sheaf of rings is local if
+  and only if, from the point of view of the internal language of~$\Sh(X)$, it
+  is a local ring.
+
+  Think of~$\O_X$ as the sheaf of ``number-valued functions'' on~$X$. In
+  algebraic geometry, this structure sheaf is a crucial part of the data:
+  Wildly different schemes can have the same underlying topological space.
+  \par
+}
+
+\begin{frame}\frametitle{Motivating the spectrum}
+  Let~$A$ be a commutative ring (in~$\Set$).
+
+  Is there a \hil{free local ring}~$A \to A'$ over~$A$?
+  \[ \xymatrix{
+    A \ar[rd] \ar[rrr] &&& {\substack{\text{local}\\\text{\normalsize$R$}\\\phantom{\text{local}}}} \\
+    & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!34]{\text{local}}[rru]
+  } \]
+
+  \hil{No,} if we restrict to~$\Set$.
+
+  \hil{Yes,} if we allow a change of topos:
+  Then $A \to \O_{\Spec A}$ is the universal localization.
+\end{frame}
+
+\note{\justifying
+  Details on this point of view can be found in one of Peter Arndt's very nice
+  answers on MathOverflow:
+  \begin{center}\url{http://mathoverflow.net/a/14334/31233}\end{center}
+  \par
+}
+
 \subsection{What is a topos?}
@@
 \subsection{What is the internal language?}
+
 \frame[t]{\frametitle{What is the internal language?}
@@
       epimorphism & surjective map \\
-      ring object & ring \\
-      module object & module \\
+      group object & group \\
       \bottomrule
@@
   \[ \renewcommand{\arraystretch}{1.3}\begin{array}{@{}l@{\ }c@{\ }l@{}}
-    U \models f = g \? \F &\Ll& f|_U = g|_U \in \Gamma(U, \F) \\
+    U \models f = g \? \F &\Ll& f|_U = g|_U \in \F(U) \\
     U \models \varphi \wedge \psi &\Ll&
@@
     U \models \forall f \? \F\_ \varphi(f) &\Ll&
-      \text{for all sections~$f \in \Gamma(V, \F), V \subseteq U$: $V \models
+      \text{for all sections~$f \in \F(V), V \subseteq U$: $V \models
       \varphi(f)$} \\
@@
       \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
-    && \quad\quad \text{there exists~$f_i \in \Gamma(U_i, \F)$ s.\,th.
+    && \quad\quad \text{there exists~$f_i \in \F(U_i)$ s.\,th.
     $U_i \models \varphi(f_i)$}
@@
       \Longleftrightarrow\ &
-        \text{for all open~$U \subseteq X$, sections $s, t \in \Gamma(U, \F)$:} \\
+        \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
       &\qquad\qquad
@@
       \text{for all open~$V \subseteq U$,} \\
-      && \text{sections~$s \in \Gamma(V,\F)$, $t \in \Gamma(V,\G)$:} \\
+      && \text{sections~$s \in \F(V)$, $t \in \G(V)$:} \\
       &&\qquad\qquad V \models \varphi(s,t) \\\\
@@
       &\Longleftrightarrow&
-      \text{for all open~$V \subseteq U$, sections~$s \in \Gamma(V,\F)$:} \\
+      \text{for all open~$V \subseteq U$, sections~$s \in \F(V)$:} \\
       &&\qquad\qquad \text{$V \models \varphi(s)$ implies $V \models \psi(s)$} \\\\
@@
       &&
-      \text{there is exactly one section~$s \in \Gamma(V,\F)$ with:} \\
+      \text{there is exactly one section~$s \in \F(V)$ with:} \\
       &&\qquad\qquad V \models \varphi(s)
@@
   \pause\pause
-  \begin{block}{A simple look at the constructive nature}
+  \begin{block}{A first glance at the constructive nature}
   \begin{itemize}
     \item $U \models f = 0$
-      \tabto{3.05cm} iff $f|_U = 0 \in \Gamma(U, \F)$.
+      \tabto{2.75cm} iff $f|_U = 0 \in \F(U)$.
     \item $U \models \neg\neg(f = 0)$
-      \tabto{3.05cm} iff~$f = 0$ on a dense open subset of~$U$.
+      \tabto{2.75cm} iff~$f = 0$ on a dense open subset of~$U$.
   \end{itemize}
@@
       tensor product of sheaves & tensor product of modules \\
-      rank function & minimal number of generators \\
       % sheaf of rational functions & total quotient ring of~$\O_X$ \\
       sheaf of Kähler differentials & module of Kähler differentials \\
-      dimension of $X$ & Krull dimension of $\O_X$ \\
+      sheaf of rational functions & total quotient ring of~$\O_X$ \\
       \bottomrule
@@
   \end{center}
+
+  \visible<2>{\begin{tikzpicture}[overlay]
+    \draw[fill=white, draw=white, opacity=0.9] (0,0) rectangle (\paperwidth,6.6);
+    \node[anchor=south west,inner sep=0] (image) at (1.8,1.3) {
+      \includegraphics[width=0.7\textwidth]{images/steven-kleiman-misconceptions-about-kx}
+    };
+  \end{tikzpicture}}
 \end{frame}
@@
 \begin{frame}\frametitle{Using the dictionary}
-  Let~$X$ be a scheme. Employ its \hil{small Zariski topos}:~$\Sh(X)$.
-
   \begin{center}
@@
   \begin{itemize}
-    \item The sheaf $\K_X$ of rational functions
+    \item Upper semicontinuous rank function
     \item Transfer principles~$M \leftrightarrow M^\sim$
@@
 
-\subsection{The sheaf of rational functions}
-
-\begin{frame}\frametitle{The sheaf of rational functions}
-  \begin{block}{Classical definition}
-  The sheaf~$K_X$ of \hil{rational functions} on a scheme~$X$ is the
-  sheafification of the presheaf
-  \[ U \subseteq X \longmapsto \Gamma(U, \O_X)[S(U)^{-1}], \]
-  where
-  \[ S(U) = \{ s \in \Gamma(U, \O_X) \,|\, \text{$s \in \O_{X,x}$ is
-  regular for all~$x \in U$} \}. \]
-  \end{block}
-
-  \begin{block}{Internal definition}
-  $K_X$ is the total quotient ring of~$\O_X$.
-  \end{block}
-
-  \visible<2>{\begin{tikzpicture}[overlay]
-    \draw[fill=white, draw=white, opacity=0.9] (-1,0) rectangle (\paperwidth,7.1);
-    \node[anchor=south west,inner sep=0] (image) at (1.8,1.3) {
-      \includegraphics[width=0.7\textwidth]{images/steven-kleiman-misconceptions-about-kx}
-    };
-  \end{tikzpicture}}
+\subsection{The rank function of sheaves of modules}
+
+\begin{frame}\frametitle{The rank function of sheaves of modules}
+  There is the following one-to-one correspondence:
+  \[\small
+    \underunbrace{\fmini{0.35\textwidth}{upper semi-continuous \\ functions on~$X$}}_{\text{external}}
+    \xleftrightarrow{\quad\text{1:1}\quad}
+    \underunbrace{\fmini{0.4\textwidth}{completed natural numbers}}_{\text{internal}}
+  \]
+
+  \vfill
+
+  \begin{center}
+    \small
+    \begin{minipage}{0.42\textwidth}
+      \begin{exampleblock}{}
+        \justifying
+        Let~$M$ be a f.\,g.\@ $A$-module. Assume that $A$ is a field.
+        Then~$M$ is free iff the minimal number of generators is an
+        actual natural number.
+      \end{exampleblock}
+    \end{minipage}
+    \quad\scalebox{3}{$\Rightarrow$}\quad
+    \begin{minipage}{0.37\textwidth}
+      \begin{exampleblock}{}
+        \justifying
+        Let~$\F$ be an~$\O_X$-module of finite type. Assume that~$X$ is reduced.
+        Then~$\F$ is locally free iff its rank is locally constant.
+      \end{exampleblock}
+    \end{minipage}
+  \end{center}
 \end{frame}
@@
 \begin{frame}\frametitle{Transfer principles}
+
   \hil{Question:} How do the properties of
@@
   \item an~$A$-module~$M$ in~$\Set$ and
-  \item the~$\O_X$-module~$M^\sim$ in~$\Sh(X)$, where~$X = \Spec A$, relate?
+  \item the~$\O_X$-\only<2->{module}\only<1>{\pointthis{module}{an important
+  sheaf with~$M^\sim(X) = M$}}~$M^\sim$ in~$\Sh(X)$, where~$X = \Spec A$, relate?
   \end{itemize}
@@
   \hil{Question:} Why do the following identities hold, for quasicoherent
-  sheaves and affine open subsets~$U$?
+  sheaves~$\E$ and~$\F$ and affine open subsets~$U$?
   \begin{align*}
-    \Gamma(U, \E/\F) &= \Gamma(U, \E) / \Gamma(U, \F) \\
-    \Gamma(U, \E \otimes_{\O_X} \F) &= \Gamma(U, \E) \otimes_{\Gamma(U,\O_X)} \Gamma(U, \F) \\
-    \Gamma(U, \E_\mathrm{tors}) &= \Gamma(U, \E)_\mathrm{tors} \quad\text{(sometimes)} \\
-    \Gamma(U, \K_X) &= \operatorname{Quot} \O_X(U) \quad\text{(sometimes)} \\
-    &\ \ \vdots
+    \E/\F(U) &= \E(U) / \F(U) \\
+    (\E \otimes_{\O_X} \F)(U) &= \E(U) \otimes_{\O_X(U)} \F(U) \\
+    \E_\mathrm{tors}(U) &= \E(U)_\mathrm{tors} \quad\text{(sometimes)} \\
+    \K_X(U) &= \operatorname{Quot} \O_X(U) \quad\text{(sometimes)}
+  \end{align*}
+  \pause
+  \hil{A calculation:}
+  \begin{align*}
+    M^\sim \otimes_{\O_X} N^\sim &=
+    \ull{M}[\F^{-1}] \otimes_{\ull{A}[\F^{-1}]} \ull{N}[\F^{-1}] =
+    (\ull{M} \otimes_\ull{A} \ull{N})[\F^{-1}] \\
+    &= (\ull{M \otimes_A N})[\F^{-1}] =
+    (M \otimes_A N)^\sim.
   \end{align*}
@@
 
-\begin{frame}\frametitle{A curious property}
+\begin{frame}\frametitle{A curious property of the structure sheaf}
   Let~$X$ be a scheme. Internally to~$\Sh(X)$,
@@
 
-\begin{frame}\frametitle{The absolute spectrum}
+\begin{frame}\frametitle{The absolute spectrum, revisited}
   Let~$A$ be a commutative ring (in~$\Set$).
@@
   in~$A$. Universal property:
-  \[ \Hom_{\mathrm{LRT}/\E}(T, \Spec A) \cong \Hom_{\mathrm{Ring}(\E)}(A, \mu_*\O_T) \]
+  \[ \Hom_{\mathrm{LRT}/|\E|}(T, \Spec A) \cong \Hom_{\mathrm{Ring}(\E)}(A, \mu_*\O_T) \]
   for all locally ringed toposes~$T$ equipped with a geometric morphism~$T \xrightarrow{\mu} \E$.
@@
 \begin{frame}\frametitle{The relative spectrum}
-  Let~$X$ be a scheme and~$\O_X \xrightarrow{\varphi} \A$ be an algebra.
-  Can we describe~\hil{$\boldsymbol{\RelSpec_X \A}$} internally?
+  Let~$X$ be a scheme and~$\A$ be a quasicoherent~$\O_X$-algebra.
+  Can we describe its \hil{relative spectrum}~$\RelSpec_X \A \to X$ internally?
 
@@
   \mu_*\O_T) \]
-  for all locally ringed spaces over~$X$.
+  for all locally ringed toposes~$T$ over~$X$.
+  \pause
+
+  \only<2>{\begin{exampleblock}{Beware of believing false statements}
+  \begin{itemize}
+  \item $\RelSpec_X \O_X = X$.
+  \item Every element of~$\O_X$ which is not invertible is nilpotent.
+  \item $\Spec \A$ is the one-point locale iff every element
+  of~$\A$ is invertible or nilpotent.
+  \item Thus cannot prove~$\Spec \O_X = \mathrm{pt}$ internally.
+  \end{itemize}\end{exampleblock}}
   \pause
@@
     points} of~$X$.
-    \item Internally, $\affl$ (given by $\affl(X) = \Gamma(X,\O_X)$)
+    \item Internally, $\affl$ (given by $\affl(X) = \O_X(X)$)
     looks like a field:
@@
 }
-
-\note{
-  \begin{itemize}
-    \item The affine line fulfils the axiom
-      \[ \Zar(S) \models \speak{every function~$\affl \to \affl$ is a
-      polynomial}. \]
-
-      Compare with the axiom of synthetic differential geometry:
-      \[ \text{every function~$\RR \to \RR$ is smooth.} \]
-
-      See~\cite{kock:sdg}.
-  \end{itemize}
-}
-
-\subsection{Group schemes}
-
-\begin{frame}\frametitle{Group schemes}
-  \hil{Motto:} Internal to~$\Zar(S)$, group schemes look like ordinary groups.
-  \bigskip
-
-  \small\hspace*{-0.6cm}%
-  \begin{tabular}{lll}
-    \toprule
-    group scheme & internal definition & functor of points: $X \mapsto \ldots$ \\\midrule
-    $\GG_\text{a}$ & $\affl$ (as additive group) & $\Gamma(X, \O_X)$ \\[0.5em]
-    $\GG_\text{m}$ & $\{ x \? \affl \,|\, \speak{$x$ inv.} \}$ & $\Gamma(X, \O_X)^\times$ \\[0.5em]
-    $\mu_n$ & $\{ x \? \affl \,|\, x^n = 1 \}$ & $\{ f \in \Gamma(X, \O_X) \,|\, f^n = 1 \}$ \\[0.5em]
-    $\GL_n$ & $\{ M \? \affl^{n \times n} \,|\, \speak{$M$ inv.} \}$ & $\GL_n(\Gamma(X, \O_X))$ \\
-    \bottomrule
-  \end{tabular}
-  \vspace{2em}
-\end{frame}
 
@@
 
-  \item Let~$X$ be an~$S$-scheme. The functor of points of $\RelSpec_X
-  \Omega^1_{X/S} \to X \to S$ has the internal description
+  \item Let~$X$ be an~$S$-scheme. The functor of points of its tangent bundle
+  has the internal description
   \[ \Hom(\Delta, \ull{X}), \]
   where~$\Delta = \{ \varepsilon : \affl \,|\, \varepsilon^2 = 0 \}$.
+  \end{itemize}
+\end{frame}
+
+\subsection{A strong Kock--Lawvere axiom}
+
+\begin{frame}\frametitle{A strong Kock--Lawvere axiom}
+  \begin{itemize}
+    \item The affine line fulfils the axiom
+      \[ \Zar(S) \models \speak{every function~$\affl \to \affl$ is a
+      polynomial}. \]
+    \item More generally, for any~$\affl$-algebra~$\A$ induced by a
+    quasicoherent~$\O_S$-algebra, the canonical morphism
+    \[ \A \longrightarrow \Hom(\Hom_{\mathrm{Alg}(\affl)}(\A,\affl), \affl) \]
+    is an isomorphism.
   \end{itemize}
@@
   \begin{align*}
-    (\flat \affl)(X \xrightarrow{\mu} S) &= \Gamma(X, \mu^{-1} \O_S) \\
-    \affl(X \xrightarrow{\mu} S) &= \Gamma(X, \O_X)
+    (\flat \affl)(X \xrightarrow{\mu} S) &= (\mu^{-1} \O_S)(X) \\
+    \affl(X \xrightarrow{\mu} S) &= \O_X(X)
   \end{align*}
@@
 \backupstart
+
+{\usebackgroundtemplate{\includegraphics[width=\paperwidth]{images/mathcamp-augsburg}}
+\begin{frame}[plain]
+  \vspace{\textheight}\vspace{-1em}\centering
+  \emph{Participants of Augsburg's maths camp}
+  \par
+\end{frame}}
+
+{\usebackgroundtemplate{\includegraphics[width=\paperwidth]{images/mathcamp-space}}
+\begin{frame}[plain]
+  \vspace{\textheight}\vspace{-1em}\centering
+  \emph{The sun as seen from our stratosphere balloon}
+  \par
+\end{frame}}
+
 \appendix
@@
   \end{itemize}
+\end{frame}
+
+\subsection{Group schemes}
+
+\begin{frame}\frametitle{Group schemes}
+  \hil{Motto:} Internal to~$\Zar(S)$, group schemes look like ordinary groups.
+  \bigskip
+
+  \small\hspace*{-0.6cm}%
+  \begin{tabular}{lll}
+    \toprule
+    group scheme & internal definition & functor of points: $X \mapsto \ldots$ \\\midrule
+    $\GG_\text{a}$ & $\affl$ (as additive group) & $\O_X(X)$ \\[0.5em]
+    $\GG_\text{m}$ & $\{ x \? \affl \,|\, \speak{$x$ inv.} \}$ & $\O_X(X)^\times$ \\[0.5em]
+    $\mu_n$ & $\{ x \? \affl \,|\, x^n = 1 \}$ & $\{ f \in \O_X(X) \,|\, f^n = 1 \}$ \\[0.5em]
+    $\GL_n$ & $\{ M \? \affl^{n \times n} \,|\, \speak{$M$ inv.} \}$ & $\GL_n(\O_X(X))$ \\
+    \bottomrule
+  \end{tabular}
+  \vspace{2em}
 \end{frame}
@@
 \end{document}
+
+* Krull dimension
+* Appl. to alg.
+* rank function

### 2016-05-24 slides-cambridge2016.tex
Updated note slides

@@
-\documentclass[12pt,utf8,notheorems,compress,t,draft]{beamer}
+\documentclass[12pt,utf8,notheorems,compress,t]{beamer}
 \usepackage{etex}
@@
 \setbeamertemplate{note page}[plain]
-\setbeameroption{hide notes}  % XXX
+%\setbeameroption{hide notes}  % XXX
 
 \begin{document}
+
+\addtocounter{framenumber}{-1}
 
@@
 
+\backupstart
 {\usebackgroundtemplate{\includegraphics[width=\paperwidth]{images/cambridge-university-botanic-garden}}
@@
 \frame[t]{\frametitle{Outline}\scriptsize\begin{itemize}\item[]\tableofcontents\end{itemize}}
+\backupend
 
@@
   \begin{changemargin}{2.5em}{2.5em}
-    % XXX: Update
     We describe how the internal language of certain
@@
     necessary.
+
+    % XXX: update
   \end{changemargin}
@@
   stalks~$\O_{X,x}$ are local rings. Why not demand that the sets of
-  sections~$\O_X(U)$ are local rings? This has a geometric meaning, but can
+  sections~$\O_X(U)$ are local rings? This choice has a geometric meaning, but can
   also be motivated from a logical point of view: A sheaf of rings is local if
@@
   \end{center}
-
-  \note{
-    \begin{itemize}
-      \item Special case: The language of~$\Set$ is the usual mathematical language.
-
-      \item \begin{justify}Actually, the objects of~$\E$ feel more like \emph{types} instead of \emph{sets}:
-      For instance, there is no global membership relation~$\in$. Rather,
-      for each object~$A$ of~$\E$, there is a relation~${\in_A} : A \times \P(A) \to
-      \Omega$, where~$\P(A)$ is the power object of~$A$ and~$\Omega$ is the
-      object of truth values of~$\E$ (can be understood as the power object of
-      a terminal object).\end{justify}
-
-      \item \begin{justify}Compare with the embedding theorem for abelian categories:
-      There, an explicit embedding into a category of modules is constructed.
-      Here, we only change perspective and talk about the same objects and
-      morphisms.\end{justify}
-
-      \item \begin{justify}There exists a weaker variant of the internal language which works
-      in abelian categories. By using it, one can even pretend that the objects
-      are abelian groups (instead of modules), and when constructing morphisms
-      by appealing to the axiom of unique choice (which is a theorem), one
-      doesn't even have to check linearity. The proof that this approach
-      works uses only categorical logic.\end{justify}
-
-      \item \begin{justify}For expositions of the internal language,
-      see the books~\cite{johnstone:elephant}~(chapters~D1 and~D4)
-      and~\cite{moerdijk:maclane:sheaves} (chapter~VI) or the lecture
-      notes~\cite{streicher:ctcl} (chapter~13).\end{justify}
-    \end{itemize}
-  }
 }
@@
 \frame[t]{\frametitle{The internal language of~$\Sh(X)$}
+  \small
   Let~$X$ be a topological space. Then we recursively define
   \[ U \models \varphi \quad \text{(``$\varphi$ holds on~$U$'')} \]
-  for open subsets~$U \subseteq X$ and formulas~$\varphi$.
+  for open subsets~$U \subseteq X$ and formulas~$\varphi$. Write~``$\Sh(X)
+  \models \varphi$'' to mean~$X \models \varphi$.
   \pause
-  \small
-  \[ \renewcommand{\arraystretch}{1.3}\begin{array}{@{}l@{\ }c@{\ }l@{}}
+  \[ \renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\ }c@{\ }l@{}}
     U \models f = g \? \F &\Ll& f|_U = g|_U \in \F(U) \\
@@
   \end{array} \]
+}
 
@@
   \begin{itemize}
-      \item \begin{justify}The rules are called \emph{Kripke-Joyal semantics} and can be
+    \item Special case: The language of~$\Set$ is the usual mathematical language.
+
+    \item \begin{justify}Actually, the objects of~$\E$ feel more like \emph{types} instead of \emph{sets}:
+    For instance, there is no global membership relation~$\in$. Rather,
+    for each object~$A$ of~$\E$, there is a relation~${\in_A} : A \times \P(A) \to
+    \Omega$, where~$\P(A)$ is the power object of~$A$ and~$\Omega$ is the
+    object of truth values of~$\E$ (can be understood as the power object of
+    a terminal object).\end{justify}
+
+    \item \begin{justify}Compare with the embedding theorem for abelian categories:
+    There, an explicit embedding into a category of modules is constructed.
+    Here, we only change perspective and talk about the same objects and
+    morphisms.\end{justify}
+
+    \item \begin{justify}There exists a weaker variant of the internal language which works
+    in abelian categories. By using it, one can even pretend that the objects
+    are abelian groups (instead of modules), and when constructing morphisms
+    by appealing to the axiom of unique choice (which is a theorem), one
+    doesn't even have to check linearity. The proof that this approach
+    works uses only categorical logic.\end{justify}
+
+    \item \begin{justify}For expositions of the internal language,
+    see Chapters~D1 to~D4 of the Elephant, Chapter~VI of Moerdijk and Mac
+    Lane's book, or Chapter~13 of
+    \href{http://www.mathematik.tu-darmstadt.de/~streicher/CTCL.pdf}{these
+    lecture notes by Thomas Streicher}.\end{justify}
+  \end{itemize}
+}
+
+\note{\begin{itemize}\justifying
+  \item The internal language of a sheaf topos of a $\mathrm{T}_1$-space is
+  \emph{classical} (that is, verifies the principle of excluded middle) if and
+  only if the space is discrete. That's a not particularly interesting special
+  case.
+  \item See Section~2.4 of
+  \href{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
+  notes} for remarks on how to appreciate intuitionistic logic.
+\end{itemize}}
+
+\note{
+  \begin{itemize}
+    \item \begin{justify}The rules are called \emph{Kripke--Joyal semantics} and can be
     formulated over any topos (not just sheaf topoi). They are not all
@@
     \item Negation is defined as
-      \[ \neg\varphi :\equiv (\varphi \Rightarrow \bot). \]
+    $\neg\varphi :\equiv (\varphi \Rightarrow \bot)$.
 
@@
   \end{itemize}
-  }
 }
@@
 
-\note{
+\note{\fontsize{8pt}{9.6}\selectfont
   \begin{itemize}
@@
     \item \begin{justify}One can extend the language to allow for \emph{unbounded}
-    quantification ($\forall A$ vs.~$\forall a \in A$), see Shulman's
-    stack semantics~\cite{shulman:stack}. This is needed to formulate
+    quantification ($\forall A$ vs.~$\forall a \in A$), by Shulman's
+    \emph{stack semantics}. This is needed to formulate
     universal properties internal to~$\Sh(X)$, for instance.\end{justify}
@@
     about locally internal categories over~$\Sh(X)$ (in the sense of
-    Penon~\cite{penon:locint}, see for instance the appendix
-    of~\cite{johnstone:topos}): Then one can do category theory internal
-    to~$\Sh(X)$ using the internal language, see~\cite{blechschmidt:msc}.
+    Penon, see for instance the appendix
+    of Johnstone's first topos theory book): Then one can do category theory internal
+    to~$\Sh(X)$ using the internal language.
 
@@
     course, internal category theory has been done for a long time, see for
-    instance the textbook~\cite{johnstone:elephant}; also
-    cf.~\cite{chapman:rowbottom:relcat}.\end{justify}
+    instance the Elephant and also Chapman and Rowbottom's \emph{Relative
+    Category Theory and Geometric Morphisms: A Logical Approach}.\end{justify}
   \end{itemize}
@@
 
-\note{%
+\note{\scriptsize%
   \vspace{-0.5em}%
@@
   \begin{itemize}
-    \item The internal logic of most topoi is constructive.
+    \item The internal logic of most toposes is constructive.
 
@@
     \begin{itemize}
-      \item a true proof by contradiction: ``Assume~$\varphi$ were false.
+      \item\scriptsize a true proof by contradiction: ``Assume~$\varphi$ were false.
       Then \ldots, contradiction. So~$\varphi$ is in fact true.''
@@
     \begin{itemize}
-      \item a proof of a negated formula: ``Assume~$\psi$ were true. Then
+      \item\scriptsize a proof of a negated formula: ``Assume~$\psi$ were true. Then
       \ldots, contradiction. So~$\neg\psi$ holds.''
@@
   \begin{enumerate}
-    \item[3.] \begin{justify}There is one last false rumour about constructive mathematics: Namely
+    \item[3.]\scriptsize \begin{justify}There is one last false rumour about constructive mathematics: Namely
     that most of mathematics breaks down in a constructive setting. This is
@@
 
-    In other cases, adding an additional hypotheses, which is classically
+    In other cases, adding an additional hypothesis, which is classically
     always satisfied, is necessary (and interesting). Here is an example: In
@@
 
-  \begin{justify}See~\cite{stanford:constr,dalen:int,troelstra:dalen:constr}
-  for references about constructive mathematics. The blog~\cite{bauer:blog} is
-  also very informative.\end{justify}
+  \begin{justify}\scriptsize References about constructive mathematics include:
+  \begin{itemize}
+  \item Bridges. Constructive Mathematics.
+  \item van Dalen. Intuitionistic logic.
+  \item Troelstra and van Dalen. Constructivism in Mathematics: An Introduction.
+  \end{itemize}
+  \href{http://math.andrej.com/category/constructive-math/}{Andrej Bauer's
+  blog} is also very informative.\end{justify}
 }
@@
 }
-
-\note{\begin{itemize}\justifying
-  \item Actually, the objects of~$\E$ feel more like \emph{types} instead of \emph{sets}:
-  For instance, there is no global membership relation~$\in$. Rather,
-  for each object~$A$ of~$\E$, there is a relation~${\in_A} : A \times \P(A) \to
-  \Omega$, where~$\P(A)$ is the power object of~$A$ and~$\Omega$ is the
-  object of truth values of~$\E$ (can be understood as the power object of
-  a terminal object).
-
-  \item Compare with the embedding theorem for abelian categories:
-  There, an explicit embedding into a category of modules is constructed.
-  Here, we only change perspective and talk about the same objects and
-  morphisms.
-
-  \item There exists a weaker variant of the internal language which works
-  in abelian categories. By using it, one can even pretend that the objects
-  are abelian groups (instead of modules), and when constructing morphisms
-  by appealing to the axiom of unique choice (which is a theorem), one
-  doesn't even have to check linearity. The proof that this approach
-  works uses only categorical logic.
-\end{itemize}}
-
-\note{\begin{itemize}\justifying
-  \item The internal language of a sheaf topos of a $\mathrm{T}_1$-space is
-  \emph{classical} (that is, verifies the principle of excluded middle) if and
-  only if the space is discrete. That's a not particularly interesting special
-  case.
-  \item See Section~2.4 of
-  \href{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
-  notes} for remarks on how to appreciate intuitionistic logic.
-\end{itemize}}
 
@@
       sheaf of rational functions & total quotient ring of~$\O_X$ \\
+      dimension of $X$ & Krull dimension of~$\O_X$ \\
       \bottomrule
@@
 
+\note{
+  \begin{block}{Proposition}
+    If every inhabited subset of the natural numbers has a minimum, then the
+    law of excluded middle holds. (So in constructive mathematics, one cannot
+    prove the natural numbers to be complete in this sense.)
+  \end{block}
+
+  \begin{block}{Proof}
+    Let~$\varphi$ be an arbitrary formula. Define the subset
+    \[ U \defeq \{ n \in \NN \,|\, n = 1 \vee \varphi \} \subseteq \NN, \]
+    which surely is inhabited by~$1 \in U$. So by assumption, there exists a
+    number~$z \in \NN$ which is the minimum of~$U$. We have
+    \[ z = 0 \quad\vee\quad z > 0 \]
+    (this is constructively not trivial, but can be proven by induction).
+
+    If~$z = 0$, we have~$0 \in U$, so~$0 = 1 \vee \varphi$, so~$\varphi$
+    holds.
+
+    If~$z > 0$, then~$\neg\varphi$ holds: Because if~$\varphi$ were
+    true, zero would be an element of~$U$, contradicting the minimality of~$z$.
+  \end{block}
+}
+
+\note{
+  \begin{block}{Proposition}
+    The partially ordered set
+    \[ \widehat\NN \defeq \{ A \subseteq \NN \,|\, \text{$A$ inhabited and upward
+    closed}
+    \} \]
+    is the least partially ordered set containing~$\NN$ and possessing minima
+    of arbitrary inhabited subsets.
+  \end{block}
+
+  The embedding $\NN \hookrightarrow \widehat\NN$ is given by
+  \[ n \in \NN \longmapsto {\uparrow}(n) \defeq \{ m \in \NN \,|\, m \geq n \}. \]
+
+  \begin{block}{Proof}
+    If~$M \subseteq \widehat\NN$ is an inhabited subset, its minimum is
+    \[ \min M = \bigcup M \in \widehat\NN. \]
+    The proof of the universal property is straightforward.
+  \end{block}
+}
+
+\note{
+  \begin{block}{External translation (see Mulvey's \emph{Intuitionistic algebra
+  and representations of rings})}
+    Let~$X$ be a topological space and consider the constant sheaf~$N$ with
+    $\Gamma(U, N) = \{ f : U \to \NN \,|\ \text{$f$ continuous} \}$.
+    Internally, the sheaf~$N$ plays the role of the ordinary natural numbers.
+    Then there is an one-to-one correspondence:
+    \begin{enumerate}
+      \item Let~$A \hookrightarrow N$ be a subobject which is inhabited
+      and upward closed from the internal point of view. Then
+      \[
+        x \longmapsto \inf\{ n \in \NN \,|\, n \in A_x \}
+      \]
+      is an upper semi-continuous function on~$X$.
+
+      \item Let~$\alpha : X \to \NN$ be a upper semi-continuous function. Then
+      \[ U \subseteq X \longmapsto \{ f : U \to \NN \,|\, \text{$f$
+      continuous,\ \ $f \geq \alpha$ on~$U$} \} \]
+      is a subobject of~$N$ which internally is inhabited and upward closed.
+    \end{enumerate}
+  \end{block}
+
+  Under the correspondence, locally \emph{constant}
+  functions map exactly to the [image of the] \emph{ordinary} internal natural numbers
+  [in the completed natural numbers].
+}
+
+\note{
+  \begin{itemize}
+    \item Here is an explicit example of a completed natural number which is
+    not an ordinary natural number: Let~$X =
+    \Spec k[X]$ and~$\F = k[X]/(X-a)^\sim$. The rank of~$\F$ is~$1$ at~$a$ and
+    zero elsewhere. It corresponds to the internal completed natural number
+    \begin{multline*}
+      z \defeq \min \{ n \in \NN \,|\, \speak{$\F$ can be generated by~$n$ elements} \} = \\
+        \min\{ n \in \NN \,|\, n \geq 1 \vee \speak{the element $(X-a)$ of~$\O_X$
+        is invertible} \}.
+    \end{multline*}
+    We have the internal implications
+    \begin{align*}
+      \Sh(X) \models \phantom{\neg}\,\speak{$(X-a)$ inv.} \Rightarrow z = 0\phantom{,} \\
+      \Sh(X) \models \neg\,\speak{$(X-a)$ inv.} \Rightarrow z = 1,
+    \end{align*}
+    but we do \emph{not} have
+    \begin{align*}
+      \Sh(X) \models \speak{$(X-a)$ inv.} \ \vee\ \neg\,\speak{$(X-a)$ inv.},
+    \end{align*}
+    which would imply
+    \begin{align*}
+      \Sh(X) \models z = 0 \vee z = 1,
+    \end{align*}
+    i.\,e. the false statement that~$\F$ is locally free (of ranks~$0$ resp.~$1$).
+  \end{itemize}
+}
+
+\note{
+  \begin{itemize}
+    \item \begin{justify}Here is a constructive proof of the statement that
+    finitely generated vector spaces, for which the minimal number of
+    generators is an actual natural numbers, are free:
+
+    By assumption, the minimal number~$n \in \NN$ of generators for~$M$ exists.
+    Let~$x_1,\ldots,x_n$ be a generating family of minimal length~$n$. We want
+    to verify that it's linearly independent, so that it constitutes a basis.
+
+    Let~$\sum_i \lambda_i x_i = 0$. If any~$\lambda_i$
+    were invertible, the shortened family $x_1,\ldots,x_{i-1},$ $x_{i+1},\ldots,x_n$
+    would also generate~$M$. By minimality of~$n$, this is not possible. So
+    each~$\lambda_i$ is not invertible. By the field assumption on~$A$, it
+    follows that each~$\lambda_i$ is zero.\end{justify}
+
+    \item \begin{justify}In constructive mathematics, one can not show that
+    every finitely generated vector space over a field admits a finite basis.
+    (Exercise: Prove this by showing that this would imply the law of excluded
+    middle.)
+    This is not because the space might strangely turn out to be
+    infinite-dimensional, but merely because one may not be able to explicitly
+    exhibit a finite basis.\end{justify}
+  \end{itemize}
+}
+
 \subsection{Transfer principles}
@@
   \item $\underline{M}$ is the constant sheaf with stalks~$M$ on~$X$ and
-  \item $\F \hookrightarrow \underline{A}$ is the \hil{generic filter}.
+  \item $\F \hookrightarrow \underline{A}$ is the \hil{generic prime filter}.
   \end{itemize}
@@
 \end{frame}
+
+\note{\justifying
+  The concept of a \emph{prime filter} is a direct axiomatization of what you
+  expect the complement of a prime ideal to fulfil. In classical logic,
+  complementation gives a bijection between the prime filters and the prime
+  ideals of a ring.
+
+  Prime filters are important in constructive mathematics because localizing
+  them gives rise to local rings. In contrast, localizing a ring at the
+  complement of a prime ideal doesn't usually result in a local ring.
+
+  To construct the universal localization of~$A$, one doesn't pick a particular
+  prime filter~$F$ to construct~$A[F^{-1}]$. Instead, one picks the
+  \emph{generic prime filter}~$\F$. This filter doesn't live in~$\Set$, but
+  in~$\Sh(\Spec A)$.
+  \par
+}
 
@@
   \begin{align*}
-    M^\sim \otimes_{\O_X} N^\sim &=
+    M^\sim \otimes_{\O_U} N^\sim &=
     \ull{M}[\F^{-1}] \otimes_{\ull{A}[\F^{-1}]} \ull{N}[\F^{-1}] =
@@
   if and only if, internally to~$\Sh(X)$,
-  \begin{quote}\textnormal{$\E[f^{-1}]$ is a $\Diamond_f$-sheaf for any~$f : \O_X$, \\[0.3em]
-  \qquad\qquad where~$\Diamond_f\varphi \defeqv (\text{$f$ invertible} \Rightarrow \varphi)$.}
+  \begin{quote}\textnormal{$\E[f^{-1}]$ is a $\Box_f$-sheaf for any~$f : \O_X$, \\[0.3em]
+  \qquad\qquad where~$\Box_f\varphi \defeqv (\text{$f$ invertible} \Rightarrow \varphi)$.}
   \end{quote}
@@
   The sheaf condition and the sheafification functor can be described purely
-  internally. An object~$M$ is \emph{separated} with respect to~$\Diamond$ if
+  internally. An object~$M$ is \emph{separated} with respect to~$\Box$ if
   and only if, from the internal point of view,
-  \[ \forall x,y : M\_\ \Diamond(x = y) \Rightarrow x = y. \]
-  It is a \emph{sheaf} with respect to~$\Diamond$, if furthermore
+  \[ \forall x,y : M\_\ \Box(x = y) \Rightarrow x = y. \]
+  It is a \emph{sheaf} with respect to~$\Box$, if furthermore
   \[ \forall K \subseteq M\_\
-    \Diamond(\exists x : M\_\ K = \{ x \}) \Longrightarrow
-    \exists x : M\_\ \Diamond(x \in K). \]
+    \Box(\exists x : M\_\ K = \{ x \}) \Longrightarrow
+    \exists x : M\_\ \Box(x \in K). \]
 
@@
 
-\newcommand{\idiamond}{{\usebeamercolor[fg]{item}{\boldsymbol{\Diamond}}}}
-\newcommand{\gdiamond}[1]{\textcolor{gray}{\boldsymbol{\Diamond}(}#1\textcolor{gray}{)}}
-
-\begin{frame}\frametitle{The $\Diamond$-translation}
-  Let~$\E_\Diamond \hookrightarrow \E$ be a subtopos given by a local
+\newcommand{\idiamond}{{\usebeamercolor[fg]{item}{\boldsymbol{\Box}}}}
+\newcommand{\gdiamond}[1]{\textcolor{gray}{\boldsymbol{\Box}(}#1\textcolor{gray}{)}}
+
+\begin{frame}\frametitle{The $\Box$-translation}
+  Let~$\E_\Box \hookrightarrow \E$ be a subtopos given by a local
   operator. Then
-  \[ \E_\Diamond \models \varphi \qquad\text{iff}\qquad
-    \E \models \varphi^\Diamond\only<2->{.}\only<1>{,}\]
-  \only<1>{where the translation~$\varphi \mapsto \varphi^\Diamond$ is given by:
+  \[ \E_\Box \models \varphi \qquad\text{iff}\qquad
+    \E \models \varphi^\Box\only<2->{.}\only<1>{,}\]
+  \only<1>{where the translation~$\varphi \mapsto \varphi^\Box$ is given by:
   \begin{align*}
-    (s = t)^\Diamond &\defeqv \idiamond(s=t) \\
-    (\varphi \wedge \psi)^\Diamond &\defeqv \gdiamond{\varphi^\Diamond \wedge \psi^\Diamond} \\
-    (\varphi \vee \psi)^\Diamond &\defeqv \idiamond(\varphi^\Diamond \vee \psi^\Diamond) \\
-    (\varphi \Rightarrow \psi)^\Diamond &\defeqv \gdiamond{\varphi^\Diamond \Rightarrow \psi^\Diamond} \\
-    (\forall x\?X\_ \varphi(x))^\Diamond &\defeqv \gdiamond{\forall x\?X\_ \varphi^\Diamond(x)} \\
-    (\exists x\?X\_ \varphi(x))^\Diamond &\defeqv \idiamond(\exists x\?X\_ \varphi^\Diamond(x))
+    (s = t)^\Box &\defeqv \idiamond(s=t) \\
+    (\varphi \wedge \psi)^\Box &\defeqv \gdiamond{\varphi^\Box \wedge \psi^\Box} \\
+    (\varphi \vee \psi)^\Box &\defeqv \idiamond(\varphi^\Box \vee \psi^\Box) \\
+    (\varphi \Rightarrow \psi)^\Box &\defeqv \gdiamond{\varphi^\Box \Rightarrow \psi^\Box} \\
+    (\forall x\?X\_ \varphi(x))^\Box &\defeqv \gdiamond{\forall x\?X\_ \varphi^\Box(x)} \\
+    (\exists x\?X\_ \varphi(x))^\Box &\defeqv \idiamond(\exists x\?X\_ \varphi^\Box(x))
   \end{align*}}%
   \only<2->{%
-    Let~$X$ be a scheme. Depending on~$\Diamond$,~$\Sh(X)
-    \models \Diamond\varphi$ means that~$\varphi$ holds on \ldots
+    Let~$X$ be a scheme. Depending on~$\Box$,~$\Sh(X)
+    \models \Box\varphi$ means that~$\varphi$ holds on \ldots
     \begin{itemize}
@@
     \pause
-    Can tackle the question~``$\varphi^\Diamond \stackrel{?}{\Rightarrow} \Diamond\varphi$'' logically.
+    Can tackle the question~``$\varphi^\Box \stackrel{?}{\Rightarrow} \Box\varphi$'' logically.
   }
@@
 \note{\justifying
-  The~$\Diamond$-translation is a generalization of the \emph{double negation
+  The~$\Box$-translation is a generalization of the \emph{double negation
   translation}, which is well-known in logic. The double negation translation
@@
 
-  The~$\Diamond$-translation has been studied before (see for instance Aczel:
+  The~$\Box$-translation has been studied before (see for instance Aczel:
   \emph{The Russell--Prawitz modality}, and Escardó, Oliva: \emph{The Peirce
@@
   For ease of exposition, assume that~$X$ is irreducible with generic
-  point~$\xi$. Let~$\Diamond \defeqv \neg\neg$.
-
-  Then~$\Sh(X) \models \Diamond \varphi$ means that~$\varphi$ holds on a dense
-  open subset of~$X$, while~$\Sh(X) \models \varphi^\Diamond$ means
+  point~$\xi$. Let~$\Box \defeqv \neg\neg$.
+
+  Then~$\Sh(X) \models \Box \varphi$ means that~$\varphi$ holds on a dense
+  open subset of~$X$, while~$\Sh(X) \models \varphi^\Box$ means
   that~$\varphi$ holds at the generic point (taking stalks of all involved
@@
 
-  The question~``does~$\varphi^\Diamond$ imply~$\Diamond\varphi$?'' therefore
+  The question~``does~$\varphi^\Box$ imply~$\Box\varphi$?'' therefore
   means: Does~$\varphi$ spread from the generic point to a dense open subset?
@@
 \subsection{The relative spectrum}
-
-\begin{frame}\frametitle{The absolute spectrum, revisited}
-  Let~$A$ be a commutative ring (in~$\Set$).
-
-  Is there a \hil{free local ring}~$A \to A'$ over~$A$?
-  \[ \xymatrix{
-    A \ar[rd] \ar[rrr] &&& {\substack{\text{local}\\\text{\normalsize$R$}\\\phantom{\text{local}}}} \\
-    & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!34]{\text{local}}[rru]
-  } \]
-
-  \hil{No,} if we restrict to~$\Set$.
-
-  \hil{Yes,} if we allow a change of topos:
-  Then $A \to \O_{\Spec A}$ is the universal localization.
-\end{frame}
-
-\note{\justifying
-  Details on this point of view can be found in one of Peter Arndt's very nice
-  answers on MathOverflow:
-  \begin{center}\url{http://mathoverflow.net/a/14334/31233}\end{center}
-  \par
-}
 
@@
   \pause
-  Define the frame of opens of~$\Spec A$ to be the frame of radical ideals
+  The frame of opens of~$\Spec A$ is the frame of radical ideals
   in~$A$. Universal property:
@@
 \note{\justifying
+  The axioms of a prime filter constitute a propositional geometric theory.
+  Therefore there exists the \emph{classifying locale} over prime filters. This
+  is the ring's spectrum. See Vicker's
+  \href{http://www.cs.bham.ac.uk/~sjv/LocTopSpaces.pdf}{Locales and Toposes as Spaces} and
+  \href{http://www.cs.bham.ac.uk/~sjv/GeoAspects.pdf}{Continuity and geometric
+  logic} for very accessible introductions to this topic.
+
   Monique Hakim constructed in her thesis a very general spectrum functor,
@@
   Desired universal property:
-  \[ \Hom_{\LRS/X}(T, \RelSpec_X \A) \cong \Hom_{\mathrm{Alg}(\O_X)}(\A,
+  \[ \Hom_{\mathrm{LRL}/X}(T, \RelSpec_X \A) \cong \Hom_{\mathrm{Alg}(\O_X)}(\A,
   \mu_*\O_T) \]
-  for all locally ringed toposes~$T$ over~$X$.
+  for all locally ringed locales~$T$ over~$X$.
   \pause
@@
   \item $\RelSpec_X \O_X = X$.
-  \item Every element of~$\O_X$ which is not invertible is nilpotent.
   \item $\Spec \A$ is the one-point locale iff every element
   of~$\A$ is invertible or nilpotent.
+  \item Every element of~$\O_X$ which is not invertible is nilpotent.
   \item Thus cannot prove~$\Spec \O_X = \mathrm{pt}$ internally.
@@
     looks like a field:
-    \[ \Zar(S) \models \forall x\?\affl\_ x \neq 0 \Longrightarrow \speak{$x$ inv.} \]
+    \[ \Zar(S) \models \forall x\?\affl\_ x \neq 0 \Longrightarrow \speak{$x$ invertible} \]
   \end{itemize}
@@
       families of jointly surjective open immersions to be covers. See for
-      instance the excellent Stacks project~\cite{stacks} for details.\end{justify}
+      instance the excellent Stacks project for details.\end{justify}
 
@@
 
-\note{
+\note{\scriptsize
   \begin{itemize}
     \item \begin{justify}Hakim worked out a theory of schemes internal to topoi (but
-    without using the internal language), see~\cite{hakim:rel}.\end{justify}
-
-    \item \begin{justify}The internal language of~$\Zar(\Spec A)$ is related to Coquand's
-    program about dynamical methods in algebra,
-    see~\cite{coquand:completeness,coquand:logical,coquand:effective}.\end{justify}
+    without using the internal language) in her PhD thesis.\end{justify}
+
+    \item \begin{justify}The internal language of~$\Zar(\Spec A)$ is related to
+    the programme about dynamical methods in algebra by Coquand, Coste,
+    Lombardi, Roy, and others. See Coquand's \emph{A completeness proof for
+    geometrical logic}, Coquand and Lombardi's \emph{A logical approach to
+    abstract algebra}, and Coste, Lombardi, and Roy's \emph{Dynamical methods
+    in algebra: effective Nullstellensätze}.\end{justify}
 
     \item \begin{justify}The observation that~$\affl$ is internally a field is due to
-    Kock~\cite{kock:univproj} (in the case~$S = \Spec\ZZ$) and implies a
-    curious meta-theorem:
+    Kock (in the case~$S = \Spec\ZZ$, see his \emph{Universal projective
+    geometry via topos theory}) and implies a curious meta-theorem:
 
@@
 
-    There is a similar story for local~$A$-algebras, see~\cite{anel:fact}.
-    See~\cite{wraith:int} for a short exposition on the usefulness of
+    There is a similar story for local~$A$-algebras.
+    See Wraith's \emph{Intuitionistic algebra: some recent developments in
+    topos theory} for a short exposition on the usefulness of
     classifying topoi and universal models.\end{justify}
@@
 
+\note{\justifying
+  I'm grateful to Zhen Lin Low for suggesting the example about the projective
+  space.
+
+  Explicitly, the~$\affl$-algebra~$\A^\sim$ is given by
+  \[ \A^\sim(X \xrightarrow{\mu} S) = (\mu^* \A)(X). \]
+}
+
 \subsection{A strong Kock--Lawvere axiom}
@@
       polynomial}. \]
+    More precisely, the canonical morphism
+    \[ \affl[T] \longrightarrow \Hom(\Hom_{\mathrm{Alg}(\affl)}(\affl[T],\affl), \affl) \]
+    is an isomorphism.
     \item More generally, for any~$\affl$-algebra~$\A$ induced by a
@@
   term:
-  \[ 1 \lra \mu_n \lra \GG_\text{m} \stackrel{(\cdot)^n}{\lra} \GG_\text{m} \lra 1 \]
+  \[ 1 \lra \mu_n \lra (\affl)^\times \stackrel{(\cdot)^n}{\lra} (\affl)^\times \lra 1 \]
   But we have:
-  \[ \Zar(S) \models \forall f\?(\affl)^\times\_ \Diamond_\text{ét}(\exists
+  \[ \Zar(S) \models \forall f\?(\affl)^\times\_ \Box_\text{ét}(\exists
   g\?(\affl)^\times\_ f = g^n), \]
-  where~$\Diamond_\text{ét}$ is such that $\Zar(S)_{\Diamond_\text{ét}}
-  \hookrightarrow \Zar(S)$ is the \hil{big étale topos}. It's the largest
+  where~$\Box_\text{ét}$ is such that $\Zar(S)_{\Box_\text{ét}}
+  \hookrightarrow \Zar(S)$ is the \hil{big étale topos} of~$S$. It is the largest
   subtopos of~$\Zar(S)$ where
   \[ \speak{$\affl$ is separably closed} \]
-  holds (Wraith, Felix).
-\end{frame}
+  holds [reinterpretation of Wraith, PSSL~1].
+\end{frame}
+
+\subsection{Comparing the little and the big toposes}
 
@@
   \begin{itemize}
+  \item There is a local geometric morphism~$\Zar(S) \to \Sh(S)$.
+
   \item From the point of view of~$\Sh(S)$, the big Zariski topos is $\Zar(\O_S | \O_S)$,
@@
   \vspace{\textheight}\vspace{-1em}\centering
-  \emph{The sun as seen from our stratosphere balloon}
+  \emph{The sun as seen from our high-altitude balloon}
   \par
@@
 \begin{frame}\frametitle{Translating internal statements I}
-  Let~$X$ be a topological space (or locale) and let~$\alpha : \F \to \G$ be a
-  morphism of sheaves on~$X$. Then:
-  \allowdisplaybreaks
-  \begin{align*}
-    & \Sh(X) \models \speak{$\alpha$ is injective} \\[0.5em]
-    \Longleftrightarrow\
-    & \Sh(X) \models \forall s\?\F\_ \forall t\?\F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
-    \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$, sections $s \in \F(U)$:} \\
-    &\qquad
-      \text{for all open~$V \subseteq U$, sections $t \in \F(V)$:} \\
-    &\qquad\qquad
-        \text{for all open~$W \subseteq V$:} \\
-    &\qquad\qquad\qquad
-          \text{$\alpha_W(s|_W) = \alpha_W(t|_W)$ implies $s|_W = t|_W$} \\[0.5em]
-    \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
-    &\qquad
-          \text{$\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.5em]
-    \Longleftrightarrow\ &
-      \text{$\alpha$ is a monomorphism of sheaves}
-  \end{align*}
-\end{frame}
-
-\begin{frame}\frametitle{Translating internal statements II}
   Let~$X$ be a topological space (or locale) and let~$\alpha : \F \to \G$ be a
@@
 
-\begin{frame}\frametitle{Translating internal statements III}
+\begin{frame}\frametitle{Translating internal statements II}
   Let~$X$ be a topological space (or locale) and let~$s, t \in \F(X)$ be global
@@
 
-  \small\hspace*{-0.6cm}%
+  \small%
   \begin{tabular}{lll}
@@
 
-* Krull dimension
-* Appl. to alg.
-* rank function
+* typo fulfil

### 2016-05-24 slides-cambridge2016.tex
British English spelling

@@
 \setbeamertemplate{note page}[plain]
-%\setbeameroption{hide notes}  % XXX
 
@@
 
-    A particularly interesting task is to internalize the
+    A particularly interesting task is to internalise the
     construction of the relative spectrum, which, given a quasicoherent sheaf of algebras
@@
       \[ \renewcommand{\arraystretch}{1.3}\begin{array}{@{}lcl@{}}
-        U \models \top &\Ll& U = U \text{ (always fulfilled)}\\
+        U \models \top &\Ll& U = U \text{ (always fulfiled)}\\
         U \models \bot &\Ll& U = \emptyset
@@
 
-\begin{frame}\frametitle{The absolute spectrum, internalized}
+\begin{frame}\frametitle{The absolute spectrum, internalised}
   Let~$A$ be a commutative ring in a topos~$\E$.
@@
   \begin{itemize}
-  \item Characterize quasicoherence in the big Zariski topos.
+  \item Characterise quasicoherence in the big Zariski topos.
   \item Understand how to work with~${\flat} \dashv \sharp$.
@@
 \backupstart
+
+\begin{frame}
+  \slogan{\hil{Understand notions and statements of algebraic geometry
+  as notions and statements of algebra internal to appropriate toposes.}}
+
+  {\vspace{-0.1em}\centering
+  \rotatebox{90}{\tiny\scalebox{0.5}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
+  \includegraphics[scale=0.20]{images/external-internal-small}
+  \par\medskip\vspace{-0.1em}}
+
+  \begin{itemize}
+    \item Simplify proofs and gain conceptual understanding.
+    \item Understand relative geometry as absolute geometry.
+    \item Develop a synthetic account of scheme theory.
+    \item Contribute to constructive algebra.
+  \end{itemize}
+
+  \centering
+  \hil{\href{http://tiny.cc/topos-notes}{http://tiny.cc/topos-notes}}
+  \par
+\end{frame}
 
@@
 \end{document}
-
-* typo fulfil

### 2017-02-05 slides-warwick2017.tex
Beginning of the slides for the algebraic geometry seminar in Warwick

--- 2016-05-24 slides-cambridge2016.tex
@@
 
-\title{Using the internal language of toposes in algebraic geometry}
+\title{First steps in synthetic algebraic geometry}
 \author{Ingo Blechschmidt}
-\date{May 24th, 2016}
+\date{February 7th, 2017}
 
@@
     (#1) {#1};\node[overlay,rectangle callout,%
-    callout relative pointer={(-0.5cm,0.8cm)},fill=blue!20] at ($(#1.north)+(1.1cm,-1.4cm)$) {#2};}%
+    callout relative pointer={(0.5cm,0.5cm)},fill=blue!20] at ($(#1.north)+(-1.1cm,-1.1cm)$) {#2};}%
 }%
@@
 
-  \hil{Using the internal language of toposes in \\ algebraic geometry}
+  \hil{First steps in \\ synthetic algebraic geometry}
   \medskip
@@
 
-  Category Theory Seminar \\
-  Centre for Mathematical Sciences \\
-  University of Cambridge \\
+  Algebraic Geometry Seminar \\
+  University of Warwick \\
   \ \\
-  May 24th, 2016
-  \par
-\end{frame}
-
-\backupstart
-{\usebackgroundtemplate{\includegraphics[width=\paperwidth]{images/cambridge-university-botanic-garden}}
-\begin{frame}[plain]
-  \vspace{\textheight}\vspace{-1em}
-  \centering
-  \emph{Cambridge University Botanic Garden} \\
-  \par
-\end{frame}}
-
-\note{
-  Photo source:
-  \href{http://www.stpaulscambridge.org.uk/classes-and-room-hire/activities-and-events/newtown-news/}{Newtown
-  News}
-}
+  Feburary 7th, 2017
+  \par
+\end{frame}
 
 \frame[t]{\frametitle{Outline}\scriptsize\begin{itemize}\item[]\tableofcontents\end{itemize}}
-\backupend
 
@@
   \begin{changemargin}{2.5em}{2.5em}
-    We describe how the internal language of certain
-    toposes, the associated small and big Zariski toposes of a scheme, can be
-    used to give simpler definitions and more conceptual proofs of the basic
+    We describe how the internal language of certain toposes, the
+    associated little and big Zariski toposes of a scheme, can be used
+    to give simpler definitions and more conceptual proofs of the basic
     notions and observations in algebraic geometry.
-    % This is useful for studying schemes from a local or relative point of view.
-
-    The starting point is that, from the internal point of view, sheaves of rings
-    and sheaves of modules look just like plain rings and plain modules.
-    In this way, some concepts and statements of scheme theory can be reduced to
-    concepts and statements of intuitionistic linear algebra.
-
-    Furthermore, modal operators can be used to model phrases such as ``on a
-    dense open subset it holds that'' or ``on an open neighbourhood of a given
-    point it holds that''. These operators define certain subtoposes; a
-    generalization of the double-negation translation is useful in order to
-    understand the internal universe of those subtoposes from the internal point
-    of view of the ambient topos.
-
-    A particularly interesting task is to internalise the
-    construction of the relative spectrum, which, given a quasicoherent sheaf of algebras
-    on a scheme~$X$, yields a scheme over~$X$. From the internal point of
-    view, this construction should simply reduce to an intuitionistically sensible
-    variant of the ordinary construction of the spectrum of a ring, but it turns
-    out that this expectation is too naive and that a refined approach is
-    necessary.
-
-    % XXX: update
+
+    The starting point is that, from the "internal point of view" of the
+    little Zariski topos, sheaves of rings and sheaves of modules look
+    just like plain rings and plain modules. In this way, some concepts
+    and statements of scheme theory can be reduced to concepts and
+    statements of intuitionistic linear algebra. This simplifies working
+    with sheaves and brings conceptual clarity.
+
+    The internal language of the big Zariski topos goes even further. It
+    incorporates Grothendieck's functor-of-points philosophy in order to
+    cast modern algebraic geometry, relative to an arbitrary base
+    scheme, in a naive language reminiscient of the classical Italian
+    school. The base scheme looks like the one-element set from this
+    point of view.
+
+    The talk gives an introduction to this topos-theoretic point of view
+    of algebraic geometry in the hope that it's useful to working
+    algebraic geometers. No prior knowledge about toposes is supposed.
   \end{changemargin}
@@
 
-\section{Basics}
-
-\subsection{What is a scheme?}
-
-\begin{frame}\frametitle{What is a scheme?}
-  \begin{itemize}
-  \item A \hil{manifold} is a space which is \hil{locally isomorphic} to some open
-  subset of some~$\RR^n$.
-  \medskip
-  \item A \hil{scheme} is a space which is \hil{locally isomorphic} to the
-  \hil{spectrum of some (commutative) ring}:
-  \[ \Spec A \defeq \{ \ppp \subseteq A \,|\, \text{$\ppp$ is a prime ideal} \} \]
-  \item By \hil{space} we mean: topological space~$X$ equipped with a local
-  sheaf~$\O_X$ of rings.
-  \end{itemize}
-
-  \centering
-  \begin{columns}
-    \begin{column}{0.45\textwidth}
-      \centering
-      \includegraphics[height=0.25\textheight]{images/manifold} \\
-      \tiny a manifold
-      \par
-    \end{column}
-    \begin{column}{0.45\textwidth}
-      \centering
-      \includegraphics[height=0.25\textheight]{images/mumfords-treasure-map} \\
-      \tiny Mumford's treasure map of~$\Spec \ZZ[X]$
-      \par
-    \end{column}
-  \end{columns}
-\end{frame}
-
-\note{\justifying
-  A \emph{sheaf of rings} on a topological space~$X$ is a ring object
-  in~$\Sh(X)$, the category of set-valued sheaves on~$X$.
-
-  A sheaf~$\O_X$ of rings is \emph{local} if and only if all the
-  stalks~$\O_{X,x}$ are local rings. Why not demand that the sets of
-  sections~$\O_X(U)$ are local rings? This choice has a geometric meaning, but can
-  also be motivated from a logical point of view: A sheaf of rings is local if
-  and only if, from the point of view of the internal language of~$\Sh(X)$, it
-  is a local ring.
-
-  Think of~$\O_X$ as the sheaf of ``number-valued functions'' on~$X$. In
-  algebraic geometry, this structure sheaf is a crucial part of the data:
-  Wildly different schemes can have the same underlying topological space.
-  \par
-}
-
-\begin{frame}\frametitle{Motivating the spectrum}
-  Let~$A$ be a commutative ring (in~$\Set$).
-
-  Is there a \hil{free local ring}~$A \to A'$ over~$A$?
-  \[ \xymatrix{
-    A \ar[rd] \ar[rrr] &&& {\substack{\text{local}\\\text{\normalsize$R$}\\\phantom{\text{local}}}} \\
-    & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!34]{\text{local}}[rru]
-  } \]
-
-  \hil{No,} if we restrict to~$\Set$.
-
-  \hil{Yes,} if we allow a change of topos:
-  Then $A \to \O_{\Spec A}$ is the universal localization.
-\end{frame}
-
-\note{\justifying
-  Details on this point of view can be found in one of Peter Arndt's very nice
-  answers on MathOverflow:
-  \begin{center}\url{http://mathoverflow.net/a/14334/31233}\end{center}
-  \par
-}
+\section{The internal language of toposes}
 
 \subsection{What is a topos?}
-
-\begin{frame}\frametitle{What is a topos?}
+\frame[t]{\frametitle{What is a topos?}
   \begin{block}{Formal definition}A \hil{topos} is a category which has finite limits,
@@
 
-  \begin{block}{Motto}A topos is a category which is sufficiently rich to support
+  \begin{block}{Motto}A topos is a category sufficiently rich to support
   an \hil{internal language}.
@@
     \item \hil{$\Sh(X)$:} \tabto{1.4cm} category of set-valued sheaves on a space~$X$
+    \item \hil{$\Zar(S)$:} \tabto{1.4cm} big Zariski topos of a base scheme~$S$
   \end{itemize}\end{block}
-\end{frame}
-
-\note{\justifying
-  While technically correct, the formal definition is actually
+
+  \note{
+    \begin{itemize}
+      \item \begin{justify}While technically correct, the formal definition is actually
       misleading in a sense: A topos has lots of other vital structure, which
@@
       closed, finitely complete and cocomplete Heyting category which is exact,
-  extensive and has a subobject classifier.
-
-  Check out an
-  \href{https://ncatlab.org/publications/published/Leinster2011}{article by
-  Tom Leinster} for a leisurely introduction to topos theory.
-  \par
+      extensive and has a subobject classifier.\end{justify}
+
+      \item See~\cite{leinster:topos} for a leisurely introduction to topos
+      theory.
+    \end{itemize}
+  }
 }
@@
 \subsection{What is the internal language?}
-
 \frame[t]{\frametitle{What is the internal language?}
@@
       \midrule
-      object of~$\E$ & set/type \\
+      object of~$\E$ & set \\
       morphism in~$\E$ & map of sets \\
@@
     \item \begin{justify}The rules are called \emph{Kripke--Joyal semantics} and can be
-    formulated over any topos (not just sheaf topoi). They are not all
+    formulated over any topos (not just sheaf toposes). They are not all
     arbitrary: Rather, they are very finely concerted to make the crucial
@@
   \begin{block}{Crucial property: Soundness}
-    If~$U \models \varphi$ and $\varphi$
+    If~$U \models \varphi$ and if $\varphi$
     implies $\psi$ \only<1>{constructively}\only<2->{\pointthis{constructively}{
@@
       no $\neg\neg\varphi \Rightarrow \varphi$,\ \
-      no AxC
+      no axiom of choice
     }},
@@
   \vfill
+  \vspace*{2em}
 
@@
     \item $U \models f = 0$
-      \tabto{2.75cm} iff $f|_U = 0 \in \F(U)$.
+      \tabto{3.5cm} iff $f|_U = 0 \in \Gamma(U, \F)$.
+    \item $U \models f = 0 \vee g = 0$
+      \tabto{3.5cm} iff on a cover~$U = \bigcup_i U_i$,\\
+      \tabto{3.5cm} \qquad\qquad $f|_{U_i} = 0$ or~$g|_{U_i} = 0$.
     \item $U \models \neg\neg(f = 0)$
-      \tabto{2.75cm} iff~$f = 0$ on a dense open subset of~$U$.
+      \tabto{3.5cm} iff~$f = 0$ on a dense open subset of~$U$.
   \end{itemize}
@@
 
-\note{\scriptsize%
+\note{%
   \vspace{-0.5em}%
@@
     \begin{itemize}
-      \item\scriptsize a true proof by contradiction: ``Assume~$\varphi$ were false.
+      \item a true proof by contradiction: ``Assume~$\varphi$ were false.
       Then \ldots, contradiction. So~$\varphi$ is in fact true.''
@@
     \begin{itemize}
-      \item\scriptsize a proof of a negated formula: ``Assume~$\psi$ were true. Then
+      \item a proof of a negated formula: ``Assume~$\psi$ were true. Then
       \ldots, contradiction. So~$\neg\psi$ holds.''
@@
   \begin{enumerate}
-    \item[3.]\scriptsize \begin{justify}There is one last false rumour about constructive mathematics: Namely
+    \item[3.] \begin{justify}There is one last false rumour about constructive mathematics: Namely
     that most of mathematics breaks down in a constructive setting. This is
@@
 
-    In other cases, adding an additional hypothesis, which is classically
+    In other cases, adding an additional hypotheses, which is classically
     always satisfied, is necessary (and interesting). Here is an example: In
@@
 
-  \begin{justify}\scriptsize References about constructive mathematics include:
-  \begin{itemize}
-  \item Bridges. Constructive Mathematics.
-  \item van Dalen. Intuitionistic logic.
-  \item Troelstra and van Dalen. Constructivism in Mathematics: An Introduction.
-  \end{itemize}
-  \href{http://math.andrej.com/category/constructive-math/}{Andrej Bauer's
-  blog} is also very informative.\end{justify}
+  \begin{justify}See~\cite{stanford:constr,dalen:int,troelstra:dalen:constr}
+  for references about constructive mathematics. The blog~\cite{bauer:blog} is
+  also very informative.\end{justify}
 }
@@
 \section{The little Zariski topos of a scheme}
+
 \subsection{Basic look and feel}
@@
       \midrule
-      sheaf of sets & set/type \\
+      sheaf of sets & set \\
       morphism of sheaves & map of sets \\
@@
       tensor product of sheaves & tensor product of modules \\
-      % sheaf of rational functions & total quotient ring of~$\O_X$ \\
       sheaf of Kähler differentials & module of Kähler differentials \\
@@
 
-\begin{frame}[c]\frametitle{Praise for Mike Shulman}
-  \centering
-  \includegraphics[scale=0.4]{images/mike-shulman-stack-semantics}
-  \par
-\end{frame}
-
-\note{\justifying
-  The internal language of a topos supports
-  \begin{itemize}
-  \item first-order logic,
-  \item higher-order logic (for instance quantification over subsets),
-  \item dependent types, and
-  \item unbounded quantification.
-  \end{itemize}
-
-  The first three items are standard. The fourth is due to Mike Shulman.
-  Combined, it's possible to interpret ``essentially all of constructive
-  mathematics'' internal to a topos.
-
-  Restrictions persist for operations with a ``set-theoretical flavor'' like
-  building an infinite union of iterated powersets, for example~$\bigcup_{n \in
-  \NN} P^n(\NN)$.
-  \par
-}
-
-\subsection{The rank function of sheaves of modules}
-
-\begin{frame}\frametitle{The rank function of sheaves of modules}
-  There is the following one-to-one correspondence:
-  \[\small
-    \underunbrace{\fmini{0.35\textwidth}{upper semi-continuous \\ functions on~$X$}}_{\text{external}}
-    \xleftrightarrow{\quad\text{1:1}\quad}
-    \underunbrace{\fmini{0.4\textwidth}{completed natural numbers}}_{\text{internal}}
-  \]
-
-  \vfill
-
-  \begin{center}
-    \small
-    \begin{minipage}{0.42\textwidth}
-      \begin{exampleblock}{}
-        \justifying
-        Let~$M$ be a f.\,g.\@ $A$-module. Assume that $A$ is a field.
-        Then~$M$ is free iff the minimal number of generators is an
-        actual natural number.
-      \end{exampleblock}
-    \end{minipage}
-    \quad\scalebox{3}{$\Rightarrow$}\quad
-    \begin{minipage}{0.37\textwidth}
-      \begin{exampleblock}{}
-        \justifying
-        Let~$\F$ be an~$\O_X$-module of finite type. Assume that~$X$ is reduced.
-        Then~$\F$ is locally free iff its rank is locally constant.
-      \end{exampleblock}
-    \end{minipage}
-  \end{center}
-\end{frame}
-
-\note{
-  \begin{block}{Proposition}
-    If every inhabited subset of the natural numbers has a minimum, then the
-    law of excluded middle holds. (So in constructive mathematics, one cannot
-    prove the natural numbers to be complete in this sense.)
-  \end{block}
-
-  \begin{block}{Proof}
-    Let~$\varphi$ be an arbitrary formula. Define the subset
-    \[ U \defeq \{ n \in \NN \,|\, n = 1 \vee \varphi \} \subseteq \NN, \]
-    which surely is inhabited by~$1 \in U$. So by assumption, there exists a
-    number~$z \in \NN$ which is the minimum of~$U$. We have
-    \[ z = 0 \quad\vee\quad z > 0 \]
-    (this is constructively not trivial, but can be proven by induction).
-
-    If~$z = 0$, we have~$0 \in U$, so~$0 = 1 \vee \varphi$, so~$\varphi$
-    holds.
-
-    If~$z > 0$, then~$\neg\varphi$ holds: Because if~$\varphi$ were
-    true, zero would be an element of~$U$, contradicting the minimality of~$z$.
-  \end{block}
-}
-
-\note{
-  \begin{block}{Proposition}
-    The partially ordered set
-    \[ \widehat\NN \defeq \{ A \subseteq \NN \,|\, \text{$A$ inhabited and upward
-    closed}
-    \} \]
-    is the least partially ordered set containing~$\NN$ and possessing minima
-    of arbitrary inhabited subsets.
-  \end{block}
-
-  The embedding $\NN \hookrightarrow \widehat\NN$ is given by
-  \[ n \in \NN \longmapsto {\uparrow}(n) \defeq \{ m \in \NN \,|\, m \geq n \}. \]
-
-  \begin{block}{Proof}
-    If~$M \subseteq \widehat\NN$ is an inhabited subset, its minimum is
-    \[ \min M = \bigcup M \in \widehat\NN. \]
-    The proof of the universal property is straightforward.
-  \end{block}
-}
-
-\note{
-  \begin{block}{External translation (see Mulvey's \emph{Intuitionistic algebra
-  and representations of rings})}
-    Let~$X$ be a topological space and consider the constant sheaf~$N$ with
-    $\Gamma(U, N) = \{ f : U \to \NN \,|\ \text{$f$ continuous} \}$.
-    Internally, the sheaf~$N$ plays the role of the ordinary natural numbers.
-    Then there is an one-to-one correspondence:
-    \begin{enumerate}
-      \item Let~$A \hookrightarrow N$ be a subobject which is inhabited
-      and upward closed from the internal point of view. Then
-      \[
-        x \longmapsto \inf\{ n \in \NN \,|\, n \in A_x \}
-      \]
-      is an upper semi-continuous function on~$X$.
-
-      \item Let~$\alpha : X \to \NN$ be a upper semi-continuous function. Then
-      \[ U \subseteq X \longmapsto \{ f : U \to \NN \,|\, \text{$f$
-      continuous,\ \ $f \geq \alpha$ on~$U$} \} \]
-      is a subobject of~$N$ which internally is inhabited and upward closed.
-    \end{enumerate}
-  \end{block}
-
-  Under the correspondence, locally \emph{constant}
-  functions map exactly to the [image of the] \emph{ordinary} internal natural numbers
-  [in the completed natural numbers].
-}
-
-\note{
-  \begin{itemize}
-    \item Here is an explicit example of a completed natural number which is
-    not an ordinary natural number: Let~$X =
-    \Spec k[X]$ and~$\F = k[X]/(X-a)^\sim$. The rank of~$\F$ is~$1$ at~$a$ and
-    zero elsewhere. It corresponds to the internal completed natural number
-    \begin{multline*}
-      z \defeq \min \{ n \in \NN \,|\, \speak{$\F$ can be generated by~$n$ elements} \} = \\
-        \min\{ n \in \NN \,|\, n \geq 1 \vee \speak{the element $(X-a)$ of~$\O_X$
-        is invertible} \}.
-    \end{multline*}
-    We have the internal implications
-    \begin{align*}
-      \Sh(X) \models \phantom{\neg}\,\speak{$(X-a)$ inv.} \Rightarrow z = 0\phantom{,} \\
-      \Sh(X) \models \neg\,\speak{$(X-a)$ inv.} \Rightarrow z = 1,
-    \end{align*}
-    but we do \emph{not} have
-    \begin{align*}
-      \Sh(X) \models \speak{$(X-a)$ inv.} \ \vee\ \neg\,\speak{$(X-a)$ inv.},
-    \end{align*}
-    which would imply
-    \begin{align*}
-      \Sh(X) \models z = 0 \vee z = 1,
-    \end{align*}
-    i.\,e. the false statement that~$\F$ is locally free (of ranks~$0$ resp.~$1$).
-  \end{itemize}
-}
-
-\note{
-  \begin{itemize}
-    \item \begin{justify}Here is a constructive proof of the statement that
-    finitely generated vector spaces, for which the minimal number of
-    generators is an actual natural numbers, are free:
-
-    By assumption, the minimal number~$n \in \NN$ of generators for~$M$ exists.
-    Let~$x_1,\ldots,x_n$ be a generating family of minimal length~$n$. We want
-    to verify that it's linearly independent, so that it constitutes a basis.
-
-    Let~$\sum_i \lambda_i x_i = 0$. If any~$\lambda_i$
-    were invertible, the shortened family $x_1,\ldots,x_{i-1},$ $x_{i+1},\ldots,x_n$
-    would also generate~$M$. By minimality of~$n$, this is not possible. So
-    each~$\lambda_i$ is not invertible. By the field assumption on~$A$, it
-    follows that each~$\lambda_i$ is zero.\end{justify}
-
-    \item \begin{justify}In constructive mathematics, one can not show that
-    every finitely generated vector space over a field admits a finite basis.
-    (Exercise: Prove this by showing that this would imply the law of excluded
-    middle.)
-    This is not because the space might strangely turn out to be
-    infinite-dimensional, but merely because one may not be able to explicitly
-    exhibit a finite basis.\end{justify}
-  \end{itemize}
-}
-
 \subsection{Transfer principles}
@@
   \item an~$A$-module~$M$ in~$\Set$ and
-  \item the~$\O_X$-\only<2->{module}\only<1>{\pointthis{module}{an important
-  sheaf with~$M^\sim(X) = M$}}~$M^\sim$ in~$\Sh(X)$, where~$X = \Spec A$, relate?
+  \item the~$\O_X$-module $M^\sim$ in~$\Sh(X)$, where~$X = \Spec A$, relate?
   \end{itemize}
@@
   \item $\underline{M}$ is the constant sheaf with stalks~$M$ on~$X$ and
-  \item $\F \hookrightarrow \underline{A}$ is the \hil{generic prime filter}.
+  \item $\F \hookrightarrow \underline{A}$ is the \hil{generic prime filter}
+  with stalk~$A \setminus \mathfrak{p}$ at $\mathfrak{p} \in \Spec A$.
   \end{itemize}
@@
   \begin{align*}
-    \E/\F(U) &= \E(U) / \F(U) \\
+    (\E/\F)(U) &= \E(U) / \F(U) \\
     (\E \otimes_{\O_X} \F)(U) &= \E(U) \otimes_{\O_X(U)} \F(U) \\
@@
   \begin{itemize}
-    \item \begin{justify}Hakim worked out a theory of schemes internal to topoi (but
+    \item \begin{justify}Hakim worked out a theory of schemes internal to toposes (but
     without using the internal language) in her PhD thesis.\end{justify}
@@
     topos theory} for a short exposition on the usefulness of
-    classifying topoi and universal models.\end{justify}
+    classifying toposes and universal models.\end{justify}
   \end{itemize}

### 2017-02-06 slides-warwick2017.tex
More work on the slides

@@
 
-\usetheme{Warsaw}
+\useinnertheme[shadow=true]{rounded}
+\useoutertheme{split}
+\usecolortheme{orchid}
+\usecolortheme{whale}
+\setbeamerfont{block title}{size={}}
+
+\useinnertheme{rectangles}
+
 \usecolortheme{seahorse}
-%\usefonttheme{default}?
-%\usepackage{kurier}?
+\definecolor{mypurple}{RGB}{150,0,255}
+\setbeamercolor{structure}{fg=mypurple}
+\definecolor{myred}{RGB}{150,0,0}
+\setbeamercolor*{title}{bg=myred,fg=white}
+\setbeamercolor*{titlelike}{bg=myred,fg=white}
+
 \usefonttheme{serif}
@@
 %\usepackage{mathpazo}
-\useinnertheme{rectangles}
 
@@
 \newcommand{\Sch}{\mathrm{Sch}}
+\newcommand{\Aff}{\mathrm{Aff}}
 \newcommand{\LRS}{\mathrm{LRS}}
@@
 \newcommand{\lra}{\longrightarrow}
-\newcommand{\RelSpec}{\operatorname{\text{\ul{$\mathrm{Spec}$}}}}
+\newcommand{\RelSpec}{\operatorname{Spec}}
 \renewcommand{\_}{\mathpunct{.}}
@@
     \setlength{\fboxsep}{8pt}%
-    {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal text}\parbox{0.84\textwidth}{#1}}}%
+    {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal text}\parbox{0.93\textwidth}{#1}}}%
   \end{center}%
@@
 
-\setbeamertemplate{frametitle}[default][colsep=-2bp,rounded=false,shadow=false,center]
-
-\setbeamertemplate{headline}{}
+\setbeamertemplate{frametitle}{%
+  \vskip0.7em%
+  \leavevmode%
+  \begin{beamercolorbox}[dp=1ex,center]{}%
+      \usebeamercolor[fg]{item}{\textbf{{\Large \insertframetitle}}}
+  \end{beamercolorbox}%
+}
+
 \setbeamertemplate{navigation symbols}{}
@@
 \setbeamertemplate{footline}{%
-  \begin{beamercolorbox}[wd=\paperwidth,ht=2.5ex,dp=1.25ex,right,rightskip=1mm,leftskip=1mm]{frametitle right}
-    {\quad} \inserttitle \hfill
-    \insertframenumber\,/\,\inserttotalframenumber {\quad}
-  \end{beamercolorbox}}
+  \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex,dp=1ex,right,rightskip=1mm,leftskip=1mm]{}%
+    \inserttitle \hfill
+    \insertframenumber\,/\,\inserttotalframenumber
+  \end{beamercolorbox}%
+  \vskip0pt%
+}
 
@@
 
-\section{The internal language of toposes}
+\section[Internal language]{The internal language of toposes}
 
@@
   \[ U \models \varphi \quad \text{(``$\varphi$ holds on~$U$'')} \]
-  for open subsets~$U \subseteq X$ and formulas~$\varphi$. Write~``$\Sh(X)
-  \models \varphi$'' to mean~$X \models \varphi$.
+  for open subsets~$U \subseteq X$ and formulas~$\varphi$.
+  % Write~``$\Sh(X) \models \varphi$'' to mean~$X \models \varphi$.
   \pause
@@
 
-\section{The little Zariski topos of a scheme}
-
-\subsection{Basic look and feel}
+\section[Little Zariski Topos]{The little Zariski topos of a scheme}
 
@@
 
-\subsection{Building and using a dictionary}
+\subsection[Dictionary]{Building and using a dictionary}
 
@@
       sheaf of Kähler differentials & module of Kähler differentials \\
-      sheaf of rational functions & total quotient ring of~$\O_X$ \\
+      % sheaf of rational functions & total quotient ring of~$\O_X$ \\
       dimension of $X$ & Krull dimension of~$\O_X$ \\
@@
   notions and statements of (intuitionistic) \hil{commutative
-  algebra} internal to suitable \hil{toposes}.}
+  algebra} internal to the \hil{little Zariski topos}.}
 
@@
 
-  Further topics in the little Zariski topos:
-  \begin{itemize}
-    \item Upper semicontinuous rank function
+  Further topics regarding the little Zariski topos:
+  \begin{itemize}
     \item Transfer principles~$M \leftrightarrow M^\sim$
@@
 
-\subsection{The curious role of affine open subsets}
+\subsection[Affine opens]{The curious role of affine open subsets}
 
@@
 
-\subsection{Quasicoherence of sheaves of modules}
+\subsection[Quasicoherence]{Quasicoherence of sheaves of modules}
 
@@
 
-\subsection{Spreading of properties}
+\subsection[Spreading]{Spreading of properties}
 
@@
 
-\subsection{The relative spectrum}
+\subsection[Relative spectrum]{The relative spectrum}
 
@@
 
-\section{The big Zariski topos of a scheme}
-
-\subsection{Basic look and feel}
+\section[Big Zariski topos]{The big Zariski topos of a scheme}
+
+\begin{frame}\frametitle{Synthetic algebraic geometry}
+  Usual approach to algebraic geometry: \hil{layer schemes above ordinary set theory}
+  using either
+  \begin{itemize}
+    \item locally ringed spaces
+    \begin{multline*}
+      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^2+Y^2-Z^2)$} + {} \\
+      \text{Zariski topology} + \text{structure sheaf}
+    \end{multline*}
+    \item or Grothendieck's functor-of-points approach, where a scheme is a functor~$\mathrm{Ring} \to \mathrm{Set}$.
+    \[ A \longmapsto \{ (x,y,z) \in A^ 3 \,|\, x^2+y^2-z^2=0 \} \]
+  \end{itemize}
+
+  \hil{Synthetic approach}: model schemes \hil{directly as sets} in a certain
+  nonclassical set theory.
+  \[ \{ (x,y,z) \in (\affl)^3 \,|\, x^2+y^2-z^2=0 \} \]
+\end{frame}
 
@@
   \begin{block}{Definition}The \hil{big Zariski topos} $\Zar(S)$ of a scheme~$S$ is the
-  category $\Sh(\Sch/S)$. It consists of certain functors
-  $(\Sch/S)^\op \to \Set$.
+  category $\Sh(\Aff/S)$. It consists of functors
+  $(\Aff/S)^\op \to \Set$ satisfying the sheaf condition that
+  \[ F(U) \to
+    \prod_i F(U_i) \rightrightarrows
+    \prod_{j,k} F(U_j \cap U_k) \]
+  \\[-1.5em]
+  is a limit diagram for any affine scheme~$U = \bigcup_i U_i$ over~$S$.
   \end{block}
@@
   \begin{itemize}
-    \item For an~$S$-scheme~$X$, its functor of points \[ \ull{X} =
-    \Hom_S(\cdot,X) \] is an object of~$\Zar(S)$. It feels like \hil{the set of
+    \item For an~$S$-scheme~$X$, its functor of points $\ull{X} =
+    \Hom_S(\cdot,X)$ is an object of~$\Zar(S)$. It feels like \hil{the set of
     points} of~$X$.
@@
   \begin{itemize}
-  \item The functor of points of~$\PP^n_S$ has the internal description
-  \[ \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\,
-    x_0 \neq 0 \vee \cdots \vee x_n \neq 0 \}/\text{scaling}. \]
-  \pause
-
-  \item
-  Let~$\A$ be an~$\O_S$-algebra. This induces an $\affl$-algebra $\A^\sim$
-  internal to~$\Zar(S)$. The functor of points of~$\RelSpec_S \A$ has the
-  internal description
-  \[ \Hom_{\mathrm{Alg}(\affl)}(\A^\sim, \affl). \]
-  \visible<2>{\begin{tikzpicture}[overlay]
-    \draw[fill=pink, draw=pink, opacity=0.99] (2.5,0.4) rectangle (7.4,1.3);
-  \end{tikzpicture}}
-  \pause\pause
-
-  \item Let~$X$ be an~$S$-scheme. The functor of points of its tangent bundle
-  has the internal description
-  \[ \Hom(\Delta, \ull{X}), \]
+  \item $\mathbb{P}^n_S \defeq \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\,
+    x_0 \neq 0 \vee \cdots \vee x_n \neq 0 \}/(\affl)^\times.$
+
+  \item $\RelSpec_S A \defeq \Hom_{\mathrm{Alg}(\affl)}(A, \affl).$
+
+  \item $TX = \Hom(\Delta, \ull{X})$,
   where~$\Delta = \{ \varepsilon : \affl \,|\, \varepsilon^2 = 0 \}$.
@@
 \end{document}
+
+Mention:
+* generic freeness

### 2017-02-06 slides-warwick2017.tex
Slides almost finished

@@
 
-\usepackage{mathtools}
+%\usepackage{mathtools}
+\newcommand{\vcentcolon}{:}
 \usepackage{booktabs}
@@
 \usepackage{xstring}
-\usepackage{mathtools}
 \usepackage{soul}\setul{0.3ex}{}
@@
 \usefonttheme{serif}
-\usepackage[T1]{fontenc}
-\usepackage{libertine}
-%\usepackage{mathpazo}
+%\usepackage[T1]{fontenc}
+%\usepackage{libertine}
 
@@
 \newcommand{\ull}[1]{\underline{#1}}
-\newcommand{\affl}{\ensuremath{{\ull{\AA}^1_S}}}
+\newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
 \newcommand{\Ll}{\vcentcolon\!\Longleftrightarrow}
@@
 
-\begin{frame}\frametitle{A curious property of the structure sheaf}
+\begin{frame}\frametitle{\mbox{A curious property of the structure sheaf}}
   Let~$X$ be a scheme. Internally to~$\Sh(X)$,
@@
   \par
+\end{frame}
+
+\begin{frame}\frametitle{Generic freeness}
+  % XXX: enlarge diagram
+  \vspace*{-1em}
+  \[ \xymatrix{
+    & \ar@{-}[d]^{\substack{\text{finitely}\\\text{generated}}} M \\
+    R \ar[r]_{\text{of finite type}} & S
+  } \]
+  If $R$ is reduced ($x^n = 0 \Rightarrow x = 0$), there is a \hil{dense open}~$U \subseteq \Spec R$ such that for any $f \in R$ with~$D(f)
+  \subseteq U$,
+  \begin{itemize}
+    \item $S[f^{-1}]$ and $M[f^{-1}]$ are free modules over $R[f^{-1}]$,
+    \item $R[f^{-1}] \to S[f^{-1}]$ is of finite presentation, and
+    \item $M[f^{-1}]$ is finitely presented as a module over~$S[f^{-1}]$.
+  \end{itemize}
+
+  For a trivial proof, employ~$\Sh(\Spec R)$ and exploit that
+  \begin{itemize}
+    \item $\O_{\Spec R}$ is a \hil{field}: $\neg(\text{$x$ invertible}) \Rightarrow x = 0$,
+    \item $\O_{\Spec R}$ and $\O_{\Spec R}[X_1,\ldots,X_n]$ are \hil{Noetherian}
+    in the sense that any ideal is \hil{not not} finitely generated.
+  \end{itemize}
 \end{frame}
@@
     \begin{multline*}
-      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^2+Y^2-Z^2)$} + {} \\
+      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^3+Y^3-Z^3)$} + {} \\
       \text{Zariski topology} + \text{structure sheaf}
@@
     \item or Grothendieck's functor-of-points approach, where a scheme is a functor~$\mathrm{Ring} \to \mathrm{Set}$.
-    \[ A \longmapsto \{ (x,y,z) \in A^ 3 \,|\, x^2+y^2-z^2=0 \} \]
+    \[ A \longmapsto \{ (x,y,z) \in A^ 3 \,|\, x^3+y^3-z^3=0 \} \]
   \end{itemize}
@@
   nonclassical set theory.
-  \[ \{ (x,y,z) \in (\affl)^3 \,|\, x^2+y^2-z^2=0 \} \]
+  \[ \{ (x,y,z) \in (\affl)^3 \,|\, x^3+y^3-z^3=0 \} \]
 \end{frame}
@@
   \begin{itemize}
-  \item $\mathbb{P}^n_S \defeq \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\,
+  \item $\mathbb{P}^n_S = \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\,
     x_0 \neq 0 \vee \cdots \vee x_n \neq 0 \}/(\affl)^\times.$
 
-  \item $\RelSpec_S A \defeq \Hom_{\mathrm{Alg}(\affl)}(A, \affl).$
-
-  \item $TX = \Hom(\Delta, \ull{X})$,
+  \item $\RelSpec_S A = \Hom_{\mathrm{Alg}(\affl)}(A, \affl) =
+  \text{set of $\affl$-rational points of $A$}.$
+
+  \item $TX = \Hom(\Delta, X)$,
   where~$\Delta = \{ \varepsilon : \affl \,|\, \varepsilon^2 = 0 \}$.
+
+  \item An $\affl$-module~$E$ is quasicoherent if and only if
+  \[ E \otimes_\affl A \longrightarrow \Hom(\Spec A, E) \]
+  is an isomorphism for all finitely presented $\affl$-algebras $A$.
+
+  \item In particular, any map $\affl \to \affl$ is given by a polynomial.
+
+  \item A subset $U \subseteq X$ is open if and only if for any $x : X$
+  there exists an ideal $J \subseteq \affl$ such that $\affl/J$ is quasicoherent and
+  \[ x \in U \quad\Longleftrightarrow\quad J = (1). \]
+
+  \item A subset $U \subseteq X$ is qc-open if and only if for any $x : X$
+  there exist $f_1,\ldots,f_n \subseteq \affl$ such that
+  \[ x \in U \quad\longleftrightarrow\quad \exists i\_ \text{$f_i$ is invertible}. \]
   \end{itemize}
@@
 \end{document}
-
-Mention:
-* generic freeness

### 2017-02-07 slides-warwick2017.tex
Final todo items for tomorrow

@@
 
-%\usepackage{mathtools}
-\newcommand{\vcentcolon}{:}
+\usepackage{mathtools}
+%\newcommand{\vcentcolon}{:}
 \usepackage{booktabs}
@@
 \usefonttheme{serif}
-%\usepackage[T1]{fontenc}
-%\usepackage{libertine}
+\usepackage[T1]{fontenc}
+\usepackage{libertine}
 
@@
 \begin{frame}\frametitle{Generic freeness}
-  % XXX: enlarge diagram
-  \vspace*{-1em}
-  \[ \xymatrix{
+  \vspace*{-1.6em}
+  \[ \xymatrixcolsep{3pc}\xymatrixrowsep{3pc}\xymatrix{
     & \ar@{-}[d]^{\substack{\text{finitely}\\\text{generated}}} M \\
@@
   \end{itemize}
+  \pause
 
@@
 
-  \begin{block}{Basic look and feel}
   \begin{itemize}
@@
   \end{itemize}
-  \end{block}
 
@@
 
-  \item $\RelSpec_S A = \Hom_{\mathrm{Alg}(\affl)}(A, \affl) =
-  \text{set of $\affl$-rational points of $A$}.$
+  \item \ \\[-1.2em]\mbox{$\RelSpec_S R = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
+  \text{set of $\affl$-rational points of $R$}.$}
 
@@
   where~$\Delta = \{ \varepsilon : \affl \,|\, \varepsilon^2 = 0 \}$.
+  \pause
 
   \item An $\affl$-module~$E$ is quasicoherent if and only if
-  \[ E \otimes_\affl A \longrightarrow \Hom(\Spec A, E) \]
-  is an isomorphism for all finitely presented $\affl$-algebras $A$.
+  \[ E \otimes_\affl R \longrightarrow \Hom(\Spec R, E) \]
+  is an isomorphism for all finitely presented $\affl$-algebras $R$.
 
   \item In particular, any map $\affl \to \affl$ is given by a polynomial.
+  \pause
 
@@
   there exist $f_1,\ldots,f_n \subseteq \affl$ such that
-  \[ x \in U \quad\longleftrightarrow\quad \exists i\_ \text{$f_i$ is invertible}. \]
-  \end{itemize}
-\end{frame}
+  \[ x \in U \quad\Longleftrightarrow\quad \exists i\_ \text{$f_i$ is invertible}. \]
+  \end{itemize}
+\end{frame}
+
+% Add two bigger case studies:
+% * Construction of the Grassmannian
+% * f : Delta --> X such that f(0) in U implies that f factors over U
+%
+% And add the following smaller things:
+% * qc, sep, qs
+% * universally closed
+% * surjective
 
@@
 }
-
-\subsection{A strong Kock--Lawvere axiom}
-
-\begin{frame}\frametitle{A strong Kock--Lawvere axiom}
-  \begin{itemize}
-    \item The affine line fulfils the axiom
-      \[ \Zar(S) \models \speak{every function~$\affl \to \affl$ is a
-      polynomial}. \]
-    More precisely, the canonical morphism
-    \[ \affl[T] \longrightarrow \Hom(\Hom_{\mathrm{Alg}(\affl)}(\affl[T],\affl), \affl) \]
-    is an isomorphism.
-    \item More generally, for any~$\affl$-algebra~$\A$ induced by a
-    quasicoherent~$\O_S$-algebra, the canonical morphism
-    \[ \A \longrightarrow \Hom(\Hom_{\mathrm{Alg}(\affl)}(\A,\affl), \affl) \]
-    is an isomorphism.
-  \end{itemize}
-\end{frame}
 
@@
 
-\subsection{Comparing the little and the big toposes}
-
-\begin{frame}\frametitle{Comparing the little and the big toposes}
-  \begin{itemize}
-  \item There is a local geometric morphism~$\Zar(S) \to \Sh(S)$.
-
-  \item From the point of view of~$\Sh(S)$, the big Zariski topos is $\Zar(\O_S | \O_S)$,
-  the classifying topos of local~$\O_S$-algebras which are local
-  over~$\O_S$.
-
-  \item From the point of view of~$\Zar(S)$, the little Zariski topos is the
-  largest subtopos where $\flat \affl \to \affl$ is bijective.
-  \begin{align*}
-    (\flat \affl)(X \xrightarrow{\mu} S) &= (\mu^{-1} \O_S)(X) \\
-    \affl(X \xrightarrow{\mu} S) &= \O_X(X)
-  \end{align*}
-  \end{itemize}
-\end{frame}
-
 \section{Open tasks}
@@
   \begin{itemize}
-  \item Characterise quasicoherence in the big Zariski topos.
-  \item Understand how to work with~${\flat} \dashv \sharp$.
   \item Do cohomology in the little Zariski topos; exploit that higher direct

### 2017-02-07 slides-warwick2017.tex
Slides finished

@@
 
-\frame[t]{\frametitle{Outline}\scriptsize\begin{itemize}\item[]\tableofcontents\end{itemize}}
+\frame[t]{\frametitle{Outline}\small
+  \begin{columns}[t]
+    \begin{column}{0.6\textwidth}
+      \vspace*{-1.5em}
+      \begin{itemize}\item[]\tableofcontents\end{itemize}
+      \par
+    \end{column}
+    \begin{column}{0.4\textwidth}
+      \centering
+      \begin{minipage}{0.70\textwidth}
+        \begin{exampleblock}{}
+          \justifying
+          Any finitely generated vector space does \emph{not not} possess a basis.
+        \end{exampleblock}
+      \end{minipage}
+      \medskip
+
+      \scalebox{3}{$\Downarrow$}
+
+      \begin{minipage}{0.70\textwidth}
+        \begin{exampleblock}{}
+          \justifying
+          Any sheaf of modules of finite type on a reduced scheme is locally free
+          \emph{on a dense open subset}.
+        \end{exampleblock}
+        \centering
+        \tiny Ravi Vakil: ``Important hard exercise'' (13.7.K).
+        \par
+      \end{minipage}
+    \end{column}
+  \end{columns}
+}
 
@@
   \begin{changemargin}{2.5em}{2.5em}
-    We describe how the internal language of certain toposes, the
-    associated little and big Zariski toposes of a scheme, can be used
-    to give simpler definitions and more conceptual proofs of the basic
-    notions and observations in algebraic geometry.
-
-    The starting point is that, from the "internal point of view" of the
-    little Zariski topos, sheaves of rings and sheaves of modules look
-    just like plain rings and plain modules. In this way, some concepts
-    and statements of scheme theory can be reduced to concepts and
-    statements of intuitionistic linear algebra. This simplifies working
-    with sheaves and brings conceptual clarity.
-
-    The internal language of the big Zariski topos goes even further. It
-    incorporates Grothendieck's functor-of-points philosophy in order to
-    cast modern algebraic geometry, relative to an arbitrary base
-    scheme, in a naive language reminiscient of the classical Italian
-    school. The base scheme looks like the one-element set from this
-    point of view.
-
-    The talk gives an introduction to this topos-theoretic point of view
-    of algebraic geometry in the hope that it's useful to working
-    algebraic geometers. No prior knowledge about toposes is supposed.
+    (new abstract)
   \end{changemargin}
@@
 
-\frame[t]{\frametitle{The internal language of~$\Sh(X)$}
+\frame[t]{\frametitle{The internal language of~$\boldsymbol{\Sh(X)}$}
   \small
@@
 
-\frame[t]{\frametitle{The internal language of~$\Sh(X)$}
+\frame[t]{\frametitle{The internal language of~$\boldsymbol{\Sh(X)}$}
   \begin{block}{Crucial property: Locality}
@@
     \item $U \models f = 0$
-      \tabto{3.5cm} iff $f|_U = 0 \in \Gamma(U, \F)$.
-    \item $U \models f = 0 \vee g = 0$
-      \tabto{3.5cm} iff on a cover~$U = \bigcup_i U_i$,\\
-      \tabto{3.5cm} \qquad\qquad $f|_{U_i} = 0$ or~$g|_{U_i} = 0$.
+      \tabto{2.9cm} iff $f|_U = 0 \in \Gamma(U, \F)$.
     \item $U \models \neg\neg(f = 0)$
-      \tabto{3.5cm} iff~$f = 0$ on a dense open subset of~$U$.
+      \tabto{2.9cm} iff~$f = 0$ on a dense open subset of~$U$.
   \end{itemize}
@@
 
-\section[Little Zariski Topos]{The little Zariski topos of a scheme}
+\section[Little Zariski]{The little Zariski topos of a scheme}
 
@@
   \end{itemize}
-  \end{block}
 }
@@
   \slogan{\justifying Understand notions and statements of \hil{algebraic geometry} as
-  notions and statements of (intuitionistic) \hil{commutative
+  notions and statements of (constructive) \hil{commutative
   algebra} internal to the \hil{little Zariski topos}.}
@@
     \item Transfer principles~$M \leftrightarrow M^\sim$
+    \item Understanding generic freeness
     \item The curious role of affine open subsets
@@
   \hil{Answer:} $M^\sim$ inherits those properties of $M$ which are \\
-  \hil{stable under localization}.
+  \hil{stable under localisation}.
   \par
@@
 
-  Prime filters are important in constructive mathematics because localizing
-  them gives rise to local rings. In contrast, localizing a ring at the
+  Prime filters are important in constructive mathematics because localising
+  them gives rise to local rings. In contrast, localising a ring at the
   complement of a prime ideal doesn't usually result in a local ring.
 
-  To construct the universal localization of~$A$, one doesn't pick a particular
+  To construct the universal localisation of~$A$, one doesn't pick a particular
   prime filter~$F$ to construct~$A[F^{-1}]$. Instead, one picks the
@@
 
-\subsection[Affine opens]{The curious role of affine open subsets}
-
-\begin{frame}\frametitle{The curious role of affine open subsets}
-  \hil{Question:} Why do the following identities hold, for quasicoherent
-  sheaves~$\E$ and~$\F$ and affine open subsets~$U$?
-  \begin{align*}
-    (\E/\F)(U) &= \E(U) / \F(U) \\
-    (\E \otimes_{\O_X} \F)(U) &= \E(U) \otimes_{\O_X(U)} \F(U) \\
-    \E_\mathrm{tors}(U) &= \E(U)_\mathrm{tors} \quad\text{(sometimes)} \\
-    \K_X(U) &= \operatorname{Quot} \O_X(U) \quad\text{(sometimes)}
-  \end{align*}
-  \pause
-  \hil{A calculation:}
-  \begin{align*}
-    M^\sim \otimes_{\O_U} N^\sim &=
-    \ull{M}[\F^{-1}] \otimes_{\ull{A}[\F^{-1}]} \ull{N}[\F^{-1}] =
-    (\ull{M} \otimes_\ull{A} \ull{N})[\F^{-1}] \\
-    &= (\ull{M \otimes_A N})[\F^{-1}] =
-    (M \otimes_A N)^\sim.
-  \end{align*}
-  \pause
-  \hil{Answer:} Because localization commutes with quotients, tensor products,
-  torsion submodules (sometimes), \ldots
-\end{frame}
-
-\subsection[Quasicoherence]{Quasicoherence of sheaves of modules}
+\subsection[Generic freeness]{Generic freeness}
 
@@
   } \]
-  If $R$ is reduced ($x^n = 0 \Rightarrow x = 0$), there is a \hil{dense open}~$U \subseteq \Spec R$ such that for any $f \in R$ with~$D(f)
+  If $R$ is reduced ($x^n = 0 \Rightarrow x = 0$), there is a \hil{dense open subset} $U \subseteq \Spec R$ such that for any $f \in R$ with~$D(f)
   \subseteq U$,
-  \begin{itemize}
+  \begin{enumerate}
     \item $S[f^{-1}]$ and $M[f^{-1}]$ are free modules over $R[f^{-1}]$,
@@
     \item $M[f^{-1}]$ is finitely presented as a module over~$S[f^{-1}]$.
-  \end{itemize}
+  \end{enumerate}
   \pause
@@
 
-\begin{frame}\frametitle{Quasicoherence}
-  Let~$X$ be a scheme. Let~$\E$ be an~$\O_X$-module.
-
-  Then~$\E$ is quasicoherent
-  if and only if, internally to~$\Sh(X)$,
-  \begin{quote}\textnormal{$\E[f^{-1}]$ is a $\Box_f$-sheaf for any~$f : \O_X$, \\[0.3em]
-  \qquad\qquad where~$\Box_f\varphi \defeqv (\text{$f$ invertible} \Rightarrow \varphi)$.}
-  \end{quote}
-  \pause
-
-  In particular: If~$\E$ is quasicoherent, then internally
-  \[ (\text{$f$ invertible} \Rightarrow s = 0) \Longrightarrow
-    \bigvee_{n \geq 0} f^n s = 0 \]
-  \vspace*{-1.5em}\par%
-  for any~$f : \O_X$ and~$s : \E$.
-\end{frame}
-
-\note{\justifying
-  The sheaf condition and the sheafification functor can be described purely
-  internally. An object~$M$ is \emph{separated} with respect to~$\Box$ if
-  and only if, from the internal point of view,
-  \[ \forall x,y : M\_\ \Box(x = y) \Rightarrow x = y. \]
-  It is a \emph{sheaf} with respect to~$\Box$, if furthermore
-  \[ \forall K \subseteq M\_\
-    \Box(\exists x : M\_\ K = \{ x \}) \Longrightarrow
-    \exists x : M\_\ \Box(x \in K). \]
-
-  The second condition displayed on the previous slide is equivalent to the
-  separatedness condition. In the special case~$\E = \O_X$, $s = 1$ it reduces
-  to Mulvey's ``somewhat obscure formula''. We now understand this condition in
-  its proper context.
-  \par
-}
-
-\subsection[Spreading]{Spreading of properties}
-
-\newcommand{\idiamond}{{\usebeamercolor[fg]{item}{\boldsymbol{\Box}}}}
-\newcommand{\gdiamond}[1]{\textcolor{gray}{\boldsymbol{\Box}(}#1\textcolor{gray}{)}}
-
-\begin{frame}\frametitle{The $\Box$-translation}
-  Let~$\E_\Box \hookrightarrow \E$ be a subtopos given by a local
-  operator. Then
-  \[ \E_\Box \models \varphi \qquad\text{iff}\qquad
-    \E \models \varphi^\Box\only<2->{.}\only<1>{,}\]
-  \only<1>{where the translation~$\varphi \mapsto \varphi^\Box$ is given by:
-  \begin{align*}
-    (s = t)^\Box &\defeqv \idiamond(s=t) \\
-    (\varphi \wedge \psi)^\Box &\defeqv \gdiamond{\varphi^\Box \wedge \psi^\Box} \\
-    (\varphi \vee \psi)^\Box &\defeqv \idiamond(\varphi^\Box \vee \psi^\Box) \\
-    (\varphi \Rightarrow \psi)^\Box &\defeqv \gdiamond{\varphi^\Box \Rightarrow \psi^\Box} \\
-    (\forall x\?X\_ \varphi(x))^\Box &\defeqv \gdiamond{\forall x\?X\_ \varphi^\Box(x)} \\
-    (\exists x\?X\_ \varphi(x))^\Box &\defeqv \idiamond(\exists x\?X\_ \varphi^\Box(x))
-  \end{align*}}%
-  \only<2->{%
-    Let~$X$ be a scheme. Depending on~$\Box$,~$\Sh(X)
-    \models \Box\varphi$ means that~$\varphi$ holds on \ldots
-    \begin{itemize}
-      \item \ldots{} a dense open subset.
-      \item \ldots{} a schematically dense open subset.
-      \item \ldots{} a given open subset~$U$.
-      \item \ldots{} an open subset containing a given closed subset~$A$.
-      \item \ldots{} an open neighbourhood of a given point~$x \in X$.
-    \end{itemize}
-
-    \pause
-    Can tackle the question~``$\varphi^\Box \stackrel{?}{\Rightarrow} \Box\varphi$'' logically.
-  }
-\end{frame}
-
-\note{\justifying
-  The~$\Box$-translation is a generalization of the \emph{double negation
-  translation}, which is well-known in logic. The double negation translation
-  has the following curious property: A formula~$\varphi$ admits a classical
-  proof if and only if the translated formula~$\varphi^{\neg\neg}$ admits an
-  intuitionistic proof.
-
-  The~$\Box$-translation has been studied before (see for instance Aczel:
-  \emph{The Russell--Prawitz modality}, and Escardó, Oliva: \emph{The Peirce
-  translation and the double negation shift}), but to the best of my know\-ledge,
-  this application -- expressing the internal language of
-  subtoposes in the internal language of the ambient topos -- is new.
-  \par
-}
-
-\note{\justifying
-  For ease of exposition, assume that~$X$ is irreducible with generic
-  point~$\xi$. Let~$\Box \defeqv \neg\neg$.
-
-  Then~$\Sh(X) \models \Box \varphi$ means that~$\varphi$ holds on a dense
-  open subset of~$X$, while~$\Sh(X) \models \varphi^\Box$ means
-  that~$\varphi$ holds at the generic point (taking stalks of all involved
-  sheaves).
-
-  The question~``does~$\varphi^\Box$ imply~$\Box\varphi$?'' therefore
-  means: Does~$\varphi$ spread from the generic point to a dense open subset?
-
-  For the special case of the double negation translation, a general answer to
-  this purely logical question has long been known: This holds if~$\varphi$ is
-  a \emph{geometric formula} (doesn't contain~$\Rightarrow$ and~$\forall$).
-  \par
-}
-
-\note{\justifying
-  Let~$\F$ be a sheaf of modules on a locally ringed space~$X$.
-  Assume that the stalk~$\F_x$ at some point~$x \in X$ vanishes.
-  Then in general it does \emph{not} follow that~$\F$ vanishes on some open
-  neighbourhood of~$x$.
-
-  This can be understood in logical terms: The statement that~$\F$ vanishes,
-  \[ \forall s : \F\_\ s = 0, \]
-  is not a geometric formula.
-
-  However, if~$\F$ is additionally supposed to be of finite type, then it
-  \emph{does} follow that~$\F$ vanishes on an open neighbourhood. This too can
-  be understood in logical terms: If~$\F$ is of finite type, then internally
-  there are generators~$s_1,\ldots,s_n$ of~$\F$. Thus the vanishing of~$\F$ can
-  be reformulated as
-  \[ s_1 = 0 \wedge \cdots \wedge s_n = 0, \]
-  and this condition is manifestly geometric.
-  \par
-}
-
-\subsection[Relative spectrum]{The relative spectrum}
-
-\newcommand{\defspeca}{\text{topological space of the prime ideals of $A$}}
-\newcommand{\defspecb}{\text{topological space of the prime filters of $A$}}
-\newcommand{\defspecc}{\text{locale of the prime filters of $A$}}
-
-\begin{frame}\frametitle{The absolute spectrum, internalised}
-  Let~$A$ be a commutative ring in a topos~$\E$.
-
-  To construct the \hil{free local ring} over~$A$, give a constructive account
-  of the spectrum:
-  \begin{align*}
-    \only<1>{\Spec A &\defeq \defspeca}
-    \only<2->{\Spec A &\defeq \hcancel{$\defspeca$}{0pt}{3pt}{0pt}{-2pt}}
-    \\
-    \only<3>{&\defeq \defspecb}
-    \only<4->{&\defeq \hcancel{$\defspecb$}{0pt}{3pt}{0pt}{-2pt}}
-    \\
-    \only<5->{&\defeq \defspecc}
-  \end{align*}
-  \pause
-  \pause
-  \pause
-  \pause
-  The frame of opens of~$\Spec A$ is the frame of radical ideals
-  in~$A$. Universal property:
-  \[ \Hom_{\mathrm{LRT}/|\E|}(T, \Spec A) \cong \Hom_{\mathrm{Ring}(\E)}(A, \mu_*\O_T) \]
-  for all locally ringed toposes~$T$ equipped with a geometric morphism~$T \xrightarrow{\mu} \E$.
-\end{frame}
-
-\note{\justifying
-  The axioms of a prime filter constitute a propositional geometric theory.
-  Therefore there exists the \emph{classifying locale} over prime filters. This
-  is the ring's spectrum. See Vicker's
-  \href{http://www.cs.bham.ac.uk/~sjv/LocTopSpaces.pdf}{Locales and Toposes as Spaces} and
-  \href{http://www.cs.bham.ac.uk/~sjv/GeoAspects.pdf}{Continuity and geometric
-  logic} for very accessible introductions to this topic.
-
-  Monique Hakim constructed in her thesis a very general spectrum functor,
-  taking a ringed topos to a locally ringed one, using explicit calculations
-  with sites.
-
-  Using the internal language allows to reduce these calculations to a minimum.
-  One constructs the spectrum as the sheaf topos over an internal
-  locale and then uses the general theorem that toposes over the base~$\E$ are
-  the same as toposes internal to~$\E$.
-
-  As a byproduct one obtains that Hakim's spectrum is \emph{localic} over the base.
-  \par
-}
-
-\begin{frame}\frametitle{The relative spectrum}
-  Let~$X$ be a scheme and~$\A$ be a quasicoherent~$\O_X$-algebra.
-  Can we describe its \hil{relative spectrum}~$\RelSpec_X \A \to X$ internally?
-
-  Desired universal property:
-  \[ \Hom_{\mathrm{LRL}/X}(T, \RelSpec_X \A) \cong \Hom_{\mathrm{Alg}(\O_X)}(\A,
-  \mu_*\O_T) \]
-  for all locally ringed locales~$T$ over~$X$.
-  \pause
-
-  \only<2>{\begin{exampleblock}{Beware of believing false statements}
-  \begin{itemize}
-  \item $\RelSpec_X \O_X = X$.
-  \item $\Spec \A$ is the one-point locale iff every element
-  of~$\A$ is invertible or nilpotent.
-  \item Every element of~$\O_X$ which is not invertible is nilpotent.
-  \item Thus cannot prove~$\Spec \O_X = \mathrm{pt}$ internally.
-  \end{itemize}\end{exampleblock}}
-  \pause
-
-  \hil{Solution:} Define internally the frame of~$\RelSpec_X \A$ to be the frame of
-  those radical ideals~$I \subseteq \A$ such that
-  \[ \forall f\?\O_X\_ \forall s\?\A\_
-    (\text{$f$ invertible in~$\O_X$} \Rightarrow s \in I) \Longrightarrow
-      fs \in I. \]
-  \pause
-  Its \hil{points} are those prime filters~$G$ of~$\A$ such that
-  \[ \forall f\?\O_X\_ \varphi(f) \in G \Longrightarrow \text{$f$ invertible in~$\O_X$}. \]
-\end{frame}
-
-\note{\justifying
-  The stated condition on~$I$ is, under the assumption that~$\A$ is
-  quasicoherent, equivalent to the condition that~$I$ is quasicoherent (as
-  an~$\O_X$-module).
-
-  The relative spectrum is thus constructed as a certain sublocale of the
-  absolute one. The two constructions coincide if and only if the dimension of
-  the base scheme is~$\leq 0$.
-
-  If~$X$ is not a scheme or~$\A$ is not quasicoherent, the construction still
-  gives rise to a locally ringed locale over~$X$ which satisfies the universal
-  property
-  \[ \Hom_{\mathrm{LRL}/X}(T, \RelSpec_X \A) \cong \Hom_{\mathrm{Alg}(\O_X)}(\A,
-  \mu_*\O_T) \]
-  for all locally ringed locales~$T \xrightarrow{\mu} X$ over~$X$.
-  \par
-}
-
-\begin{frame}\frametitle{The relative spectrum, reformulated}
-  Let~$B \to A$ be an algebra in a topos.
-
-  Is there a \hil{free local and local-over-$\boldsymbol{B}$ ring}~$A \to A'$ over~$A$?
-
-  \[ \xymatrix{
-    B \ar[r]\ar@/^2pc/[rrrr]^{\text{local}}\ar@/_/[rrd]_[@!-33]{\text{local}} &
-      A \ar[rd] \ar[rrr] &&&
-      {\substack{\text{local}\\\text{\normalsize$R$}\\\phantom{\text{local}}}} \\
-    && {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}[rru]_[@!35]{\text{local}}
-  } \]
-
-  \medskip
-  Form limits in the category of \hil{locally ringed locales}
-  by \hil{relocalizing} the corresponding limit in ringed locales.
-\end{frame}
-
-\note{\justifying
-  One might wonder whether the absolute spectrum or the relative one is ``more
-  fundamental''. The absolute spectrum can be expressed using the relative one,
-  since
-  \[ \Spec A = \RelSpec_{\Spec \ZZ} A^{\sim}, \]
-  but the other way is not in general possible: The absolute spectrum is always
-  (quasi-)compact, while the relative one is not in general.
-  \par
-}
-% XXX: Give details.
-
-\section[Big Zariski topos]{The big Zariski topos of a scheme}
+\section[Big Zariski]{The big Zariski topos of a scheme}
 
@@
   category $\Sh(\Aff/S)$. It consists of functors
-  $(\Aff/S)^\op \to \Set$ satisfying the sheaf condition that
+  $(\Aff/S)^\op \to \Set$ satisfying the gluing condition that
   \[ F(U) \to
@@
     \prod_{j,k} F(U_j \cap U_k) \]
-  \\[-1.5em]
+  \\[-0.6em]
   is a limit diagram for any affine scheme~$U = \bigcup_i U_i$ over~$S$.
@@
     looks like a field:
-    \[ \Zar(S) \models \forall x\?\affl\_ x \neq 0 \Longrightarrow \speak{$x$ invertible} \]
+    \[ \Zar(S) \models \forall x\?\affl\_ x \neq 0 \Longrightarrow \text{$x$ invertible} \]
   \end{itemize}
@@
 
-\subsection{Some internal constructions}
-
-\begin{frame}\frametitle{Some internal constructions}
+\subsection{Synthetic constructions}
+
+\begin{frame}\frametitle{Synthetic constructions}
   \begin{itemize}
   \item $\mathbb{P}^n_S = \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\,
-    x_0 \neq 0 \vee \cdots \vee x_n \neq 0 \}/(\affl)^\times.$
-
-  \item \ \\[-1.2em]\mbox{$\RelSpec_S R = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
-  \text{set of $\affl$-rational points of $R$}.$}
+    x_0 \neq 0 \vee \cdots \vee x_n \neq 0 \}/(\affl)^\times.$ \\[1em]
+
+  \item \ \\[-1.2em]\mbox{$\RelSpec R = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
+  \text{set of $\affl$-rational points of $R$}.$} \\[1em]
 
@@
   where~$\Delta = \{ \varepsilon : \affl \,|\, \varepsilon^2 = 0 \}$.
-  \pause
+  \end{itemize}
+\end{frame}
+
+\subsection{Synthetic formulation of properties}
+
+\begin{frame}{Synthetic formulation of properties}
+  \begin{itemize}
+  \item \ \\[-1.2em]\mbox{$\RelSpec R = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
+  \text{set of $\affl$-rational points of $R$}.$}
 
@@
 
-  \item In particular, any map $\affl \to \affl$ is given by a polynomial.
+  \item In particular, any map $\affl \to \affl$ is given by a polynomial. \bigskip
   \pause
 
-  \item A subset $U \subseteq X$ is open if and only if for any $x : X$
-  there exists an ideal $J \subseteq \affl$ such that $\affl/J$ is quasicoherent and
-  \[ x \in U \quad\Longleftrightarrow\quad J = (1). \]
-
   \item A subset $U \subseteq X$ is qc-open if and only if for any $x : X$
-  there exist $f_1,\ldots,f_n \subseteq \affl$ such that
-  \[ x \in U \quad\Longleftrightarrow\quad \exists i\_ \text{$f_i$ is invertible}. \]
+  there exist $f_1,\ldots,f_n \in \affl$ such that
+  $x \in U \Longleftrightarrow \exists i\_ f_i \neq 0$.
+
+  \item Open subsets are $\neg\neg$-stable:
+  $\neg\neg(x \in U) \Longrightarrow x \in U$.
+
+  \item If $\gamma : \Delta \to X$ is tangent vector with $\gamma(0) \in U$,
+  then~$\gamma(\varepsilon) \in U$ for all~$\varepsilon \in \Delta$.
+  \bigskip\pause
+
+  \item $X$ is separated if and only if for any $x,y : X$,
+  there exists a quasicoherent ideal~$J \subseteq \affl$ such that
+  $x = y \Longleftrightarrow J = (0)$.
   \end{itemize}
@@
 % * Construction of the Grassmannian
-% * f : Delta --> X such that f(0) in U implies that f factors over U
 %
@@
 
-\subsection{The étale subtopos}
-
-\begin{frame}\frametitle{The étale subtopos}
-  Recall that the \hil{Kummer sequence} is not exact in~$\Zar(S)$ at the third
-  term:
-  \[ 1 \lra \mu_n \lra (\affl)^\times \stackrel{(\cdot)^n}{\lra} (\affl)^\times \lra 1 \]
-  But we have:
-  \[ \Zar(S) \models \forall f\?(\affl)^\times\_ \Box_\text{ét}(\exists
-  g\?(\affl)^\times\_ f = g^n), \]
-  where~$\Box_\text{ét}$ is such that $\Zar(S)_{\Box_\text{ét}}
-  \hookrightarrow \Zar(S)$ is the \hil{big étale topos} of~$S$. It is the largest
-  subtopos of~$\Zar(S)$ where
-  \[ \speak{$\affl$ is separably closed} \]
-  holds [reinterpretation of Wraith, PSSL~1].
-\end{frame}
-
 \section{Open tasks}
@@
   \item Do cohomology in the big Zariski topos.
-  \item Understand more subtoposes of the big Zariski topos.
-  \item Derive suitable axioms for synthetic algebraic geometry.
+  \item Understand subtoposes of the big Zariski topos.
   \end{itemize}
@@
   \rotatebox{90}{\tiny\scalebox{0.5}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
-  \includegraphics[scale=0.20]{images/external-internal-small}
+  \includegraphics[scale=0.28]{images/external-internal-small}
   \par\medskip\vspace{-0.1em}}
@@
 
-\begin{frame}
-  \slogan{\hil{Understand notions and statements of algebraic geometry
-  as notions and statements of algebra internal to appropriate toposes.}}
+\begin{frame}{}
 
@@
   \rotatebox{90}{\tiny\scalebox{0.5}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
-  \includegraphics[scale=0.20]{images/external-internal-small}
+  \includegraphics[scale=0.17]{images/external-internal-small}
   \par\medskip\vspace{-0.1em}}
+
+  \slogan{\hil{Understand notions and statements of algebraic geometry
+  as notions and statements of algebra internal to the little Zariski topos.}}
+
+  \slogan{\hil{Develop a synthetic account of algebraic geometry.}}
 
@@
     \item Understand relative geometry as absolute geometry.
-    \item Develop a synthetic account of scheme theory.
     \item Contribute to constructive algebra.
@@
 
+\subsection[Affine opens]{The curious role of affine open subsets}
+
+\begin{frame}\frametitle{The curious role of affine open subsets}
+  \hil{Question:} Why do the following identities hold, for quasicoherent
+  sheaves~$\E$ and~$\F$ and affine open subsets~$U$?
+  \begin{align*}
+    (\E/\F)(U) &= \E(U) / \F(U) \\
+    (\E \otimes_{\O_X} \F)(U) &= \E(U) \otimes_{\O_X(U)} \F(U) \\
+    \E_\mathrm{tors}(U) &= \E(U)_\mathrm{tors} \quad\text{(sometimes)} \\
+    \K_X(U) &= \operatorname{Quot} \O_X(U) \quad\text{(sometimes)}
+  \end{align*}
+  \pause
+  \hil{A calculation:}
+  \begin{align*}
+    M^\sim \otimes_{\O_U} N^\sim &=
+    \ull{M}[\F^{-1}] \otimes_{\ull{A}[\F^{-1}]} \ull{N}[\F^{-1}] =
+    (\ull{M} \otimes_\ull{A} \ull{N})[\F^{-1}] \\
+    &= (\ull{M \otimes_A N})[\F^{-1}] =
+    (M \otimes_A N)^\sim.
+  \end{align*}
+  \pause
+  \hil{Answer:} Because localisation commutes with quotients, tensor products,
+  torsion submodules (sometimes), \ldots
+\end{frame}
+
+\subsection{Quasicoherence}
+
+\begin{frame}\frametitle{Quasicoherence}
+  Let~$X$ be a scheme. Let~$\E$ be an~$\O_X$-module.
+
+  Then~$\E$ is quasicoherent
+  if and only if, internally to~$\Sh(X)$,
+  \begin{quote}\textnormal{$\E[f^{-1}]$ is a $\Box_f$-sheaf for any~$f : \O_X$, \\[0.3em]
+  \qquad\qquad where~$\Box_f\varphi \defeqv (\text{$f$ invertible} \Rightarrow \varphi)$.}
+  \end{quote}
+  \pause
+
+  In particular: If~$\E$ is quasicoherent, then internally
+  \[ (\text{$f$ invertible} \Rightarrow s = 0) \Longrightarrow
+    \bigvee_{n \geq 0} f^n s = 0 \]
+  \vspace*{-1.5em}\par%
+  for any~$f : \O_X$ and~$s : \E$.
+\end{frame}
+
+\note{\justifying
+  The sheaf condition and the sheafification functor can be described purely
+  internally. An object~$M$ is \emph{separated} with respect to~$\Box$ if
+  and only if, from the internal point of view,
+  \[ \forall x,y : M\_\ \Box(x = y) \Rightarrow x = y. \]
+  It is a \emph{sheaf} with respect to~$\Box$, if furthermore
+  \[ \forall K \subseteq M\_\
+    \Box(\exists x : M\_\ K = \{ x \}) \Longrightarrow
+    \exists x : M\_\ \Box(x \in K). \]
+
+  The second condition displayed on the previous slide is equivalent to the
+  separatedness condition. In the special case~$\E = \O_X$, $s = 1$ it reduces
+  to Mulvey's ``somewhat obscure formula''. We now understand this condition in
+  its proper context.
+  \par
+}
+
+\subsection[Spreading]{Spreading of properties}
+
+\newcommand{\idiamond}{{\usebeamercolor[fg]{item}{\boldsymbol{\Box}}}}
+\newcommand{\gdiamond}[1]{\textcolor{gray}{\boldsymbol{\Box}(}#1\textcolor{gray}{)}}
+
+\begin{frame}\frametitle{The $\Box$-translation}
+  Let~$\E_\Box \hookrightarrow \E$ be a subtopos given by a local
+  operator. Then
+  \[ \E_\Box \models \varphi \qquad\text{iff}\qquad
+    \E \models \varphi^\Box\only<2->{.}\only<1>{,}\]
+  \only<1>{where the translation~$\varphi \mapsto \varphi^\Box$ is given by:
+  \begin{align*}
+    (s = t)^\Box &\defeqv \idiamond(s=t) \\
+    (\varphi \wedge \psi)^\Box &\defeqv \gdiamond{\varphi^\Box \wedge \psi^\Box} \\
+    (\varphi \vee \psi)^\Box &\defeqv \idiamond(\varphi^\Box \vee \psi^\Box) \\
+    (\varphi \Rightarrow \psi)^\Box &\defeqv \gdiamond{\varphi^\Box \Rightarrow \psi^\Box} \\
+    (\forall x\?X\_ \varphi(x))^\Box &\defeqv \gdiamond{\forall x\?X\_ \varphi^\Box(x)} \\
+    (\exists x\?X\_ \varphi(x))^\Box &\defeqv \idiamond(\exists x\?X\_ \varphi^\Box(x))
+  \end{align*}}%
+  \only<2->{%
+    Let~$X$ be a scheme. Depending on~$\Box$,~$\Sh(X)
+    \models \Box\varphi$ means that~$\varphi$ holds on \ldots
+    \begin{itemize}
+      \item \ldots{} a dense open subset.
+      \item \ldots{} a schematically dense open subset.
+      \item \ldots{} a given open subset~$U$.
+      \item \ldots{} an open subset containing a given closed subset~$A$.
+      \item \ldots{} an open neighbourhood of a given point~$x \in X$.
+    \end{itemize}
+
+    \pause
+    Can tackle the question~``$\varphi^\Box \stackrel{?}{\Rightarrow} \Box\varphi$'' logically.
+  }
+\end{frame}
+
+\note{\justifying
+  The~$\Box$-translation is a generalization of the \emph{double negation
+  translation}, which is well-known in logic. The double negation translation
+  has the following curious property: A formula~$\varphi$ admits a classical
+  proof if and only if the translated formula~$\varphi^{\neg\neg}$ admits an
+  intuitionistic proof.
+
+  The~$\Box$-translation has been studied before (see for instance Aczel:
+  \emph{The Russell--Prawitz modality}, and Escardó, Oliva: \emph{The Peirce
+  translation and the double negation shift}), but to the best of my know\-ledge,
+  this application -- expressing the internal language of
+  subtoposes in the internal language of the ambient topos -- is new.
+  \par
+}
+
+\note{\justifying
+  For ease of exposition, assume that~$X$ is irreducible with generic
+  point~$\xi$. Let~$\Box \defeqv \neg\neg$.
+
+  Then~$\Sh(X) \models \Box \varphi$ means that~$\varphi$ holds on a dense
+  open subset of~$X$, while~$\Sh(X) \models \varphi^\Box$ means
+  that~$\varphi$ holds at the generic point (taking stalks of all involved
+  sheaves).
+
+  The question~``does~$\varphi^\Box$ imply~$\Box\varphi$?'' therefore
+  means: Does~$\varphi$ spread from the generic point to a dense open subset?
+
+  For the special case of the double negation translation, a general answer to
+  this purely logical question has long been known: This holds if~$\varphi$ is
+  a \emph{geometric formula} (doesn't contain~$\Rightarrow$ and~$\forall$).
+  \par
+}
+
+\note{\justifying
+  Let~$\F$ be a sheaf of modules on a locally ringed space~$X$.
+  Assume that the stalk~$\F_x$ at some point~$x \in X$ vanishes.
+  Then in general it does \emph{not} follow that~$\F$ vanishes on some open
+  neighbourhood of~$x$.
+
+  This can be understood in logical terms: The statement that~$\F$ vanishes,
+  \[ \forall s : \F\_\ s = 0, \]
+  is not a geometric formula.
+
+  However, if~$\F$ is additionally supposed to be of finite type, then it
+  \emph{does} follow that~$\F$ vanishes on an open neighbourhood. This too can
+  be understood in logical terms: If~$\F$ is of finite type, then internally
+  there are generators~$s_1,\ldots,s_n$ of~$\F$. Thus the vanishing of~$\F$ can
+  be reformulated as
+  \[ s_1 = 0 \wedge \cdots \wedge s_n = 0, \]
+  and this condition is manifestly geometric.
+  \par
+}
+
+\subsection[Relative spectrum]{The relative spectrum}
+
+\newcommand{\defspeca}{\text{topological space of the prime ideals of $A$}}
+\newcommand{\defspecb}{\text{topological space of the prime filters of $A$}}
+\newcommand{\defspecc}{\text{locale of the prime filters of $A$}}
+
+\begin{frame}\frametitle{The absolute spectrum, internalised}
+  Let~$A$ be a commutative ring in a topos~$\E$.
+
+  To construct the \hil{free local ring} over~$A$, give a constructive account
+  of the spectrum:
+  \begin{align*}
+    \only<1>{\Spec A &\defeq \defspeca}
+    \only<2->{\Spec A &\defeq \hcancel{$\defspeca$}{0pt}{3pt}{0pt}{-2pt}}
+    \\
+    \only<3>{&\defeq \defspecb}
+    \only<4->{&\defeq \hcancel{$\defspecb$}{0pt}{3pt}{0pt}{-2pt}}
+    \\
+    \only<5->{&\defeq \defspecc}
+  \end{align*}
+  \pause
+  \pause
+  \pause
+  \pause
+  The frame of opens of~$\Spec A$ is the frame of radical ideals
+  in~$A$. Universal property:
+  \[ \Hom_{\mathrm{LRT}/|\E|}(T, \Spec A) \cong \Hom_{\mathrm{Ring}(\E)}(A, \mu_*\O_T) \]
+  for all locally ringed toposes~$T$ equipped with a geometric morphism~$T \xrightarrow{\mu} \E$.
+\end{frame}
+
+\note{\justifying
+  The axioms of a prime filter constitute a propositional geometric theory.
+  Therefore there exists the \emph{classifying locale} over prime filters. This
+  is the ring's spectrum. See Vicker's
+  \href{http://www.cs.bham.ac.uk/~sjv/LocTopSpaces.pdf}{Locales and Toposes as Spaces} and
+  \href{http://www.cs.bham.ac.uk/~sjv/GeoAspects.pdf}{Continuity and geometric
+  logic} for very accessible introductions to this topic.
+
+  Monique Hakim constructed in her thesis a very general spectrum functor,
+  taking a ringed topos to a locally ringed one, using explicit calculations
+  with sites.
+
+  Using the internal language allows to reduce these calculations to a minimum.
+  One constructs the spectrum as the sheaf topos over an internal
+  locale and then uses the general theorem that toposes over the base~$\E$ are
+  the same as toposes internal to~$\E$.
+
+  As a byproduct one obtains that Hakim's spectrum is \emph{localic} over the base.
+  \par
+}
+
+\begin{frame}\frametitle{The relative spectrum}
+  Let~$X$ be a scheme and~$\A$ be a quasicoherent~$\O_X$-algebra.
+  Can we describe its \hil{relative spectrum}~$\RelSpec_X \A \to X$ internally?
+
+  Desired universal property:
+  \[ \Hom_{\mathrm{LRL}/X}(T, \RelSpec_X \A) \cong \Hom_{\mathrm{Alg}(\O_X)}(\A,
+  \mu_*\O_T) \]
+  for all locally ringed locales~$T$ over~$X$.
+  \pause
+
+  \only<2>{\begin{exampleblock}{Beware of believing false statements}
+  \begin{itemize}
+  \item $\RelSpec_X \O_X = X$.
+  \item $\Spec \A$ is the one-point locale iff every element
+  of~$\A$ is invertible or nilpotent.
+  \item Every element of~$\O_X$ which is not invertible is nilpotent.
+  \item Thus cannot prove~$\Spec \O_X = \mathrm{pt}$ internally.
+  \end{itemize}\end{exampleblock}}
+  \pause
+
+  \hil{Solution:} Define internally the frame of~$\RelSpec_X \A$ to be the frame of
+  those radical ideals~$I \subseteq \A$ such that
+  \[ \forall f\?\O_X\_ \forall s\?\A\_
+    (\text{$f$ invertible in~$\O_X$} \Rightarrow s \in I) \Longrightarrow
+      fs \in I. \]
+  \pause
+  Its \hil{points} are those prime filters~$G$ of~$\A$ such that
+  \[ \forall f\?\O_X\_ \varphi(f) \in G \Longrightarrow \text{$f$ invertible in~$\O_X$}. \]
+\end{frame}
+
+\note{\justifying
+  The stated condition on~$I$ is, under the assumption that~$\A$ is
+  quasicoherent, equivalent to the condition that~$I$ is quasicoherent (as
+  an~$\O_X$-module).
+
+  The relative spectrum is thus constructed as a certain sublocale of the
+  absolute one. The two constructions coincide if and only if the dimension of
+  the base scheme is~$\leq 0$.
+
+  If~$X$ is not a scheme or~$\A$ is not quasicoherent, the construction still
+  gives rise to a locally ringed locale over~$X$ which satisfies the universal
+  property
+  \[ \Hom_{\mathrm{LRL}/X}(T, \RelSpec_X \A) \cong \Hom_{\mathrm{Alg}(\O_X)}(\A,
+  \mu_*\O_T) \]
+  for all locally ringed locales~$T \xrightarrow{\mu} X$ over~$X$.
+  \par
+}
+
+\begin{frame}\frametitle{The relative spectrum, reformulated}
+  Let~$B \to A$ be an algebra in a topos.
+
+  Is there a \hil{free local and local-over-$\boldsymbol{B}$ ring}~$A \to A'$ over~$A$?
+
+  \[ \xymatrix{
+    B \ar[r]\ar@/^2pc/[rrrr]^{\text{local}}\ar@/_/[rrd]_[@!-33]{\text{local}} &
+      A \ar[rd] \ar[rrr] &&&
+      {\substack{\text{local}\\\text{\normalsize$R$}\\\phantom{\text{local}}}} \\
+    && {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}[rru]_[@!35]{\text{local}}
+  } \]
+
+  \medskip
+  Form limits in the category of \hil{locally ringed locales}
+  by \hil{relocalising} the corresponding limit in ringed locales.
+\end{frame}
+
+\note{\justifying
+  One might wonder whether the absolute spectrum or the relative one is ``more
+  fundamental''. The absolute spectrum can be expressed using the relative one,
+  since
+  \[ \Spec A = \RelSpec_{\Spec \ZZ} A^{\sim}, \]
+  but the other way is not in general possible: The absolute spectrum is always
+  (quasi-)compact, while the relative one is not in general.
+  \par
+}
+% XXX: Give details.
+
+\subsection{The étale subtopos}
+
+\begin{frame}\frametitle{The étale subtopos}
+  Recall that the \hil{Kummer sequence} is not exact in~$\Zar(S)$ at the third
+  term:
+  \[ 1 \lra \mu_n \lra (\affl)^\times \stackrel{(\cdot)^n}{\lra} (\affl)^\times \lra 1 \]
+  But we have:
+  \[ \Zar(S) \models \forall f\?(\affl)^\times\_ \Box_\text{ét}(\exists
+  g\?(\affl)^\times\_ f = g^n), \]
+  where~$\Box_\text{ét}$ is such that $\Zar(S)_{\Box_\text{ét}}
+  \hookrightarrow \Zar(S)$ is the \hil{big étale topos} of~$S$. It is the largest
+  subtopos of~$\Zar(S)$ where
+  \[ \speak{$\affl$ is separably closed} \]
+  holds [reinterpretation of Wraith, PSSL~1].
+\end{frame}
+
 \backupend

### 2017-02-07 slides-warwick2017.tex
Final improvements

@@
   } \]
-  If $R$ is reduced ($x^n = 0 \Rightarrow x = 0$), there is a \hil{dense open subset} $U \subseteq \Spec R$ such that for any $f \in R$ with~$D(f)
-  \subseteq U$,
+  If $R$ is reduced ($x^n = 0 \Rightarrow x = 0$), there
+  %is a \hil{dense open subset} $U \subseteq \Spec R$ such that for any $f \in R$ with~$D(f)
+  is $f \neq 0$ in $R$ such that
   \begin{enumerate}
@@
     \item locally ringed spaces
+    \small
     \begin{multline*}
-      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^3+Y^3-Z^3)$} + {} \\
+      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$} + {} \\
       \text{Zariski topology} + \text{structure sheaf}
     \end{multline*}
-    \item or Grothendieck's functor-of-points approach, where a scheme is a functor~$\mathrm{Ring} \to \mathrm{Set}$.
-    \[ A \longmapsto \{ (x,y,z) \in A^ 3 \,|\, x^3+y^3-z^3=0 \} \]
-  \end{itemize}
+    \normalsize
+    \item or Grothendieck's functor-of-points account, where a scheme is a functor~$\mathrm{Ring} \to \mathrm{Set}$.
+    \small\[ A \longmapsto \{ (x,y,z) \in A^3 \,|\, x^n+y^n-z^n=0 \} \]
+  \end{itemize}
+  \bigskip
+  \pause
 
@@
   nonclassical set theory.
-  \[ \{ (x,y,z) \in (\affl)^3 \,|\, x^3+y^3-z^3=0 \} \]
+  \small
+  \[ \{ (x,y,z) \in (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]
 \end{frame}
@@
   \begin{itemize}
-  \item $\mathbb{P}^n_S = \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\,
-    x_0 \neq 0 \vee \cdots \vee x_n \neq 0 \}/(\affl)^\times.$ \\[1em]
+  \item $\mathbb{P}^n = \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\,
+    x_0 \neq 0 \vee \cdots \vee x_n \neq 0 \}/(\affl)^\times$ \\
+  $\phantom{\mathbb{P}^n} \cong \text{set of one-dimensional subspaces of~$(\affl)^{n+1}$}$.
+
+  \begin{itemize}
+    \item $\O(1) = (\ell^\vee)_{\ell \in \mathbb{P}^n}$
+    \item $\O(-1) = (\ell)_{\ell \in \mathbb{P}^n}$
+    \item Euler sequence: $0 \to \ell^\perp \to ((\affl)^{n+1})^\vee \to \ell^\vee \to 0$
+  \end{itemize}\bigskip
 
   \item \ \\[-1.2em]\mbox{$\RelSpec R = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
-  \text{set of $\affl$-rational points of $R$}.$} \\[1em]
-
-  \item $TX = \Hom(\Delta, X)$,
-  where~$\Delta = \{ \varepsilon : \affl \,|\, \varepsilon^2 = 0 \}$.
+  \text{set of $\affl$-valued points of $R$}.$}
+
+  \begin{itemize}
+    \item $\RelSpec \affl[X,Y,Z]/(X^n+Y^n-Z^n) \cong {}$ \\ \qquad\qquad $\{ (x,y,z) \in (\affl)^3 \,|\, x^n+y^n-z^n = 0 \}$ \medskip
+    \item $\Delta \defeq \RelSpec \affl[\varepsilon]/(\varepsilon^2) \cong \{ \varepsilon : \affl \,|\, \varepsilon^2 = 0 \}$
+  \end{itemize}\bigskip
+
+  \item $TX = \Hom(\Delta, X)$.
   \end{itemize}
@@
   \item \ \\[-1.2em]\mbox{$\RelSpec R = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
-  \text{set of $\affl$-rational points of $R$}.$}
+  \text{set of $\affl$-valued points of $R$}.$}
 

### 2017-03-13 snippets/open-subfunctors.tex
WIP regarding open subfunctors

--- 2015-11-02 abstract-ihes2015.tex
@@
 
-\title{Using the internal language of toposes in algebraic geometry}
-\author{Ingo Blechschmidt}
-\email{ingo.blechschmidt@math.uni-augsburg.de}
-
 \begin{document}
 
-\begin{abstract}
-  We describe how the internal language of certain
-  toposes, the associated petit and gros Zariski toposes of a scheme, can be
-  used to give simpler definitions and more conceptual proofs of the basic
-  notions and observations in algebraic geometry.
-  % This is useful for studying schemes from a local or relative point of view.
+\begin{prop}\label{prop:characterization-open-subfunctor}
+Let~$X \in \Zar(S)$ be a Zariski sheaf. A subfunctor~$U
+\hookrightarrow X$ is open if and only if
+\begin{multline*}
+  \Zar(S) \models
+  \forall x\?X\_
+  \exists J \subseteq \affl\_ \\
+  \speak{$J$ is an ideal} \wedge
+  \speak{$\affl/J$ is quasicoherent} \wedge
+  (x \in U \Leftrightarrow 1 \in J).
+\end{multline*}
+\end{prop}
 
-  The starting point is that, from the internal point of view, sheaves of rings
-  and sheaves of modules look just like plain rings and plain modules.
-  In this way, some concepts and statements of scheme theory can be reduced to
-  concepts and statements of intuitionistic linear algebra.
+The displayed condition is only meaningful in an intuitionistic
+context as provided by the big Zariski topos. In classical logic, the condition
+would be trivially satisfied, for any~$U$: Classically, we have~$(x \in U) \vee
+(x \not\in U)$. If~$x \in U$, we can pick~$J = (1)$,
+and if~$x \not\in U$, we can pick~$J = (0)$.
 
-  Furthermore, modal operators can be used to model phrases such as ``on a
-  dense open subset it holds that'' or ``on an open neighbourhood of a given
-  point it holds that''. These operators define certain subtoposes; a
-  generalization of the double-negation translation is useful in order to
-  understand the internal universe of those subtoposes from the internal point
-  of view of the ambient topos.
+\begin{cor}\label{cor:sufficient-criterion-open-subfunctor}
+Let~$X \in \Zar(S)$ be a Zariski sheaf. Let~$U \hookrightarrow X$ be
+a subfunctor. If
+\[
+  \Zar(S) \models
+  \forall x\?X\_
+  \bigvee_{n \geq 0} \exists f_1,\ldots,f_n\?\affl\_
+  (x \in U \Leftrightarrow \bigvee_i \speak{$f_i$ \inv}),
+\]
+then the subfunctor is open.
+\end{cor}
 
-  A particularly interesting task is to internalize the
-  construction of the relative spectrum, which, given a quasicoherent sheaf of algebras
-  on a scheme~$X$, yields a scheme over~$X$. From the internal point of
-  view, this construction should simply reduce to an intuitionistically sensible
-  variant of the ordinary construction of the spectrum of a ring, but it turns
-  out that this expectation is too naive and that a refined approach is
-  necessary.
-\end{abstract}
+\begin{proof}[Proof of Corollary~\ref{cor:sufficient-criterion-open-subfunctor}]
+We show that the assumption implies the displayed condition of
+Proposition~\ref{prop:characterization-open-subfunctor} in the internal
+language. Given elements~$f_1,\ldots,f_n$ as in the assumption, we
+construct the ideal~$J \defeq (f_1,\ldots,f_n) \subseteq \affl$. The
+quotient~$\affl/J$ is indeed quasicoherent, being finitely presented, and
+the statement that~$1 \in J$ is equivalent to one of
+the~$f_i$ being invertible by locality of~$\affl$.
+\end{proof}
 
-\maketitle
-\thispagestyle{empty}
+\begin{proof}[Proof of Proposition~\ref{prop:characterization-open-subfunctor}]
+\end{proof}
 
 \end{document}
-
-* Krull dimension
-* Cartier divisors?
-* Box translation
-* internal spectrum
-* big Zariski topos: criterion for affineness; étale topology

### 2017-03-13 snippets/open-subfunctors.tex
Further characterization of openness

@@
 Let~$X \in \Zar(S)$ be a Zariski sheaf. A subfunctor~$U
-\hookrightarrow X$ is open if and only if
+\hookrightarrow X$ is open if and only if any of the following conditions,
+formulated in the internal language of~$\Zar(S)$, is
+satisfied.
+\begin{enumerate}
+\item For all~$x \? X$ there is an ideal~$J \subseteq \affl$ such
+that~$\affl/J$ is quasicoherent and such that~$x \in U$ if and only if~$1 \in
+J$. In formal language:
 \begin{multline*}
-  \Zar(S) \models
+  \qquad\qquad\Zar(S) \models
   \forall x\?X\_
@@
   \speak{$\affl/J$ is quasicoherent} \wedge
-  (x \in U \Leftrightarrow 1 \in J).
-\end{multline*}
+  (x \in U \Leftrightarrow 1 \in J). \end{multline*}
+\item For all~$x \? X$ the quotient~$\affl/K$ modulo the radical ideal
+\[ K \defeq \{ f\?\affl \,|\, \speak{$f$ \inv} \Rightarrow (x \in U) \}
+\subseteq \affl \]
+is quasicoherent.
+\end{enumerate}
 \end{prop}
 
-The displayed condition is only meaningful in an intuitionistic
-context as provided by the big Zariski topos. In classical logic, the condition
-would be trivially satisfied, for any~$U$: Classically, we have~$(x \in U) \vee
-(x \not\in U)$. If~$x \in U$, we can pick~$J = (1)$,
-and if~$x \not\in U$, we can pick~$J = (0)$.
+The displayed conditions are only meaningful in an intuitionistic context as
+provided by the big Zariski topos. For instance, the first condition would be
+trivially satisfied in classical logic, for any subfunctor~$U \hookrightarrow
+X$: Classically, we have~$(x \in U) \vee (x \not\in U)$. If~$x \in U$, we can
+pick~$J = (1)$, and if~$x \not\in U$, we can pick~$J = (0)$ (such that the
+quotient~$\affl/J$ is isomorphic to~$\affl$, thus finitely presented and
+therefore in particular quasicoherent).\footnote{Strictly speaking,
+incompability with classical logical surfaces even earlier: in our internal
+quasicoherence condition. The map~$E \otimes_\affl A \to [\Spec(A), E]$ which
+the condition demands to be bijective has hardly any chance to be surjective if
+the law of excluded middle is available to define maps~$\Spec(A) \to E$ by case
+distinction.}
 
@@
 \begin{proof}[Proof of Proposition~\ref{prop:characterization-open-subfunctor}]
+
 \end{proof}

### 2017-03-14 snippets/properties-of-a1.tex
Nonreducedness of the affine line

--- 2017-03-13 snippets/open-subfunctors.tex
@@
 
-\begin{prop}\label{prop:characterization-open-subfunctor}
-Let~$X \in \Zar(S)$ be a Zariski sheaf. A subfunctor~$U
-\hookrightarrow X$ is open if and only if
-\begin{multline*}
-  \Zar(S) \models
-  \forall x\?X\_
-  \exists J \subseteq \affl\_ \\
-  \speak{$J$ is an ideal} \wedge
-  \speak{$\affl/J$ is quasicoherent} \wedge
-  (x \in U \Leftrightarrow 1 \in J).
-\end{multline*}
-\end{prop}
+\begin{prop}The following properties of~$\affl$, as seen from the internal
+point of view of~$\Zar(S)$, are consequences of~$\affl$ satisfying the
+quasicoherence condition (and~$\affl$ not being the trivial ring):
+\begin{enumerate}
+\item $\affl$ is a field in the sense that
+$\forall x\?\affl\_ \neg(x = 0) \Rightarrow \speak{$x$ \inv}$.
+\item $\affl$ is not a reduced ring: $\neg \forall x\?\affl\_ (\bigvee_{n \geq
+0} x^n = 0) \Rightarrow x = 0.$
+\item Any function~$\affl \to \affl$ is given by a unique polynomial
+in~$\affl[T]$.
+\item Any monic polynomial~$p \in \affl[T]$ of degree at least one does \notnot have a
+zero.
+\end{enumerate}\end{prop}
 
-The displayed condition is only meaningful in an intuitionistic
-context as provided by the big Zariski topos. In classical logic, the condition
-would be trivially satisfied, for any~$U$: Classically, we have~$(x \in U) \vee
-(x \not\in U)$. If~$x \in U$, we can pick~$J = (1)$,
-and if~$x \not\in U$, we can pick~$J = (0)$.
-
-\begin{cor}\label{cor:sufficient-criterion-open-subfunctor}
-Let~$X \in \Zar(S)$ be a Zariski sheaf. Let~$U \hookrightarrow X$ be
-a subfunctor. If
-\[
-  \Zar(S) \models
-  \forall x\?X\_
-  \bigvee_{n \geq 0} \exists f_1,\ldots,f_n\?\affl\_
-  (x \in U \Leftrightarrow \bigvee_i \speak{$f_i$ \inv}),
-\]
-then the subfunctor is open.
-\end{cor}
-
-\begin{proof}[Proof of Corollary~\ref{cor:sufficient-criterion-open-subfunctor}]
-We show that the assumption implies the displayed condition of
-Proposition~\ref{prop:characterization-open-subfunctor} in the internal
-language. Given elements~$f_1,\ldots,f_n$ as in the assumption, we
-construct the ideal~$J \defeq (f_1,\ldots,f_n) \subseteq \affl$. The
-quotient~$\affl/J$ is indeed quasicoherent, being finitely presented, and
-the statement that~$1 \in J$ is equivalent to one of
-the~$f_i$ being invertible by locality of~$\affl$.
-\end{proof}
-
-\begin{proof}[Proof of Proposition~\ref{prop:characterization-open-subfunctor}]
-\end{proof}
+\begin{proof}\begin{enumerate}
+\item Let $x\?\affl$ such that~$\neg(x=0)$. We consider the quasicoherence
+condition for the finitely presented~$\affl$-algebra~$A \defeq \affl/(x)$.
+Since~$\Spec(A) \cong \brak{x=0} = \brak{\bot} = \emptyset$, the condition says
+that the canonical homomorphism
+\[ \affl/(x) \longrightarrow [\emptyset, \affl] \]
+is an isomorphism. Since its codomain is the zero algebra, so is~$\affl/(x)$.
+Therefore~$1 \in (x)$, that is,~$x$ is invertible.
+\item Assume that~$\affl$ is reduced. Then the set~$\Delta \defeq \{
+\varepsilon \in \affl \,|\, \varepsilon^2 = 0 \}$ is equal to~$\{ 0 \}$.
+By the quasicoherence criterion applied to the finitely
+presented~$\affl$-algebra~$A \defeq \affl[T]/(T^2)$, the canonical map
+\[ \affl[T]/(T^2) \longrightarrow [\Spec(\affl[T]/(T^2)), \affl] \cong
+  [\Delta, \affl] \cong \affl \]
+is an isomorphism. It maps~$[T]$ to zero (the value of~$T$ at~$0 \in \Delta$).
+Thus~$T \in (T^2)$ and therefore~$1 = 0$ in~$\affl$. This is a contradiction.
+\item Immediate by considering the quasicoherence condition for the finitely
+presented~$\affl$-algebra~$A \defeq \affl[T]$ and noticing that~$\Spec(A) \cong
+\affl$.
+\item Let~$p \in \affl[T]$ be a monic polynomial of degree at least one. Assume
+that~$p$ doesn't have a zero in~$\affl$. Then the spectrum of~$A \defeq
+\affl[T]/(p)$ is empty. The quasicoherence condition for~$A$ therefore implies
+that~$\affl[T]/(p)$ is zero. This means that~$p$ is invertible in~$\affl[T]$.
+A basic lemma in commutative algebra (whose standard proof is constructive)
+then implies that with the exception of the constant term in~$p$, all
+coefficients are nilpotent. This contradicts the assumption that~$p$ is monic
+of degree at least one.\qedhere
+\end{enumerate}\end{proof}
 

### 2017-03-14 snippets/properties-of-a1.tex
Remarks on appreciating the properties

@@
 quasicoherence condition (and~$\affl$ not being the trivial ring):
+% Adjust paragraph below if numbering changes.
 \begin{enumerate}
@@
 
+In classical logic, the first two properties would directly contradict each
+other; only an intuitionistic context allows for fields which are not reduced.
+Also the third statement cannot be satisfied in classical logic: for infinite
+fields the existence part fails and for finite fields the uniqueness part
+fails.
+
+That~$\affl$ is not reduced, irrespective of the reducedness of the base
+scheme~$S$, should not come as a surprise: Reducedness is not stable under base
+change, but all statements of the internal language of~$\Zar(S)$ are.
+If~$\affl$ was reduced, then all~$S$-schemes (at least those contained in the
+site used to construct~$\Zar(S)$) would be reduced as well. In contrast, the
+structure sheaf~$\O_S$ is reduced from the point of view of the little Zariski
+topos if and only if~$S$ is reduced (Proposition~\ref{prop:reduced-ring}).
+
 \end{document}

### 2017-03-14 snippets/open-subfunctors.tex
Comment on wrong characterization

@@
 
+A basic concept in the functor-of-points approach to algebraic geometry is the
+concept of an \emph{open subfunctor}. It is used to delimit schemes from more
+general kinds of spaces: A functor is deemed to be a scheme if and only if it
+admits a covering by open subfunctors which are representable.
+
+The following definition is phrased in such a way as to apply to any of the
+several ways to define the big Zariski topos~$\Zar(S)$. In particular, it
+applies to the definition using the site consisting of affine schemes which are
+locally of finite presentation over~$S$. If~$S$ is affine, the definition only
+refers to affine schemes and open subschemes of affine schemes and is therefore
+suitable if one wants to found the theory of schemes using the functorial
+approach.
+
+\begin{defn}[{\cite[Définition~I.1.3.6 on page~10]{demazure:gabriel},
+\cite[Tag~01JI]{stacks-project}}]
+A subfunctor~$U \hookrightarrow X$ in~$\Zar(S)$ is an \emph{open subfunctor} if
+and only if for any object~$(T \to S)$ of the site used to define~$\Zar(S)$
+and any~$x \in X(T)$ there exists an open subscheme~$T_0 \subseteq T$
+such that for any object~$(T' \xra{f}{\to} T \to S)$ of the site used to
+define~$\Zar(S)$ the map~$T' \to T$ factors over~$T_0$ if and only if~$X(f)(x)
+\in U(T')$.
+\end{defn}
+
 \begin{prop}\label{prop:characterization-open-subfunctor}
 Let~$X \in \Zar(S)$ be a Zariski sheaf. A subfunctor~$U
-\hookrightarrow X$ is open if and only if any of the following conditions,
-formulated in the internal language of~$\Zar(S)$, is
-satisfied.
-\begin{enumerate}
-\item For all~$x \? X$ there is an ideal~$J \subseteq \affl$ such
-that~$\affl/J$ is quasicoherent and such that~$x \in U$ if and only if~$1 \in
-J$. In formal language:
+\hookrightarrow X$ is open if and only if, from the internal point of view
+of~$\Zar(S)$, it holds that for all~$x \? X$ there is an ideal~$J \subseteq
+\affl$ such that~$\affl/J$ is quasicoherent and such that~$x \in U$ if and only
+if~$1 \in J$. In formal language:
 \begin{multline*}
@@
   (x \in U \Leftrightarrow 1 \in J). \end{multline*}
-\item For all~$x \? X$ the quotient~$\affl/K$ modulo the radical ideal
-\[ K \defeq \{ f\?\affl \,|\, \speak{$f$ \inv} \Rightarrow (x \in U) \}
-\subseteq \affl \]
-is quasicoherent.
-\end{enumerate}
 \end{prop}
 
-The displayed conditions are only meaningful in an intuitionistic context as
-provided by the big Zariski topos. For instance, the first condition would be
-trivially satisfied in classical logic, for any subfunctor~$U \hookrightarrow
-X$: Classically, we have~$(x \in U) \vee (x \not\in U)$. If~$x \in U$, we can
-pick~$J = (1)$, and if~$x \not\in U$, we can pick~$J = (0)$ (such that the
-quotient~$\affl/J$ is isomorphic to~$\affl$, thus finitely presented and
-therefore in particular quasicoherent).\footnote{Strictly speaking,
-incompability with classical logical surfaces even earlier: in our internal
-quasicoherence condition. The map~$E \otimes_\affl A \to [\Spec(A), E]$ which
-the condition demands to be bijective has hardly any chance to be surjective if
-the law of excluded middle is available to define maps~$\Spec(A) \to E$ by case
-distinction.}
+The displayed condition is only meaningful in an intuitionistic context as
+provided by the big Zariski topos. In classical logic, the condition would be
+trivially satisfied for any subfunctor~$U \hookrightarrow X$: Classically, we
+have~$(x \in U) \vee (x \not\in U)$. If~$x \in U$, we can pick~$J = (1)$, and
+if~$x \not\in U$, we can pick~$J = (0)$ (such that the quotient~$\affl/J$ is
+isomorphic to~$\affl$, thus finitely presented and therefore in particular
+quasicoherent).\footnote{Strictly speaking, incompability with classical
+logical surfaces even earlier: in our internal quasicoherence condition. The
+map~$E \otimes_\affl A \to [\Spec(A), E]$ which the condition demands to be
+bijective has hardly any chance to be surjective if the law of excluded middle
+is available to define maps~$\Spec(A) \to E$ by case distinction.}
 
@@
 \begin{proof}[Proof of Proposition~\ref{prop:characterization-open-subfunctor}]
-
+...
 \end{proof}
 
+\begin{rem}The radical~$\sqrt{J}$ of the ideal~$J$ appearing in
+Proposition~\ref{prop:characterization-open-subfunctor} is unique:
+It is equal to the radical ideal
+\[ K \defeq \{ f\?\affl \,|\, \speak{$f$ \inv} \Rightarrow (x \in U) \}
+\subseteq \affl. \]
+It's obvious that~$J \subseteq K$ and therefore~$\sqrt{J} \subseteq K$.
+For the converse direction, let~$f \in K$ be given. Since~$\affl/J$ is
+quasicoherent, the canonical map
+\[ (\affl/J)[f^{-1}] \longrightarrow [ \Spec(\affl[f^{-1}]), \affl/J ] \]
+is bijective. Since~$\Spec(\affl[f^{-1}]) \cong \brak{\speak{$f$ \inv}}$, the image
+of~$1$ is zero: If~$\Spec(\affl[f^{-1}])$ is inhabited, $f$ is invertible and
+therefore~$x$ is an element of~$U$. This implies that~$1 \in J$.
+Thereby~$\affl/J = 0$. By injectivity of the canonical map, the
+algebra~$(\affl/J)[f^{-1}]$ is zero. Therefore~$f^n \in J$ for some natural
+number~$n$.\end{rem}
+
+\begin{rem}In view of the previous remark, one cannot hope to simplify the
+condition in Proposition~\ref{prop:characterization-open-subfunctor} to read as
+follows: ``For any~$x\?X$, the quotient~$\affl/K$ modulo the ideal~$K = \{
+f\?\affl \,|\, \speak{$f$ \inv} \Rightarrow (x \in U) \}$ is quasicoherent.
+This statement implies the condition in the proposition, but the converse
+direction does not hold, since~$\affl/K \cong \affl/\sqrt{J}$ might fail to be
+quasicoherent. For instance that's the case if~$U = \emptyset$; then~$K =
+\sqrt{(0)}$ is not quasicoherent by Remark~\ref{rem:radical-not-qcoh}.\end{rem}
+
 \end{document}

### 2017-03-14 snippets/open-subfunctors.tex
Comparision with Penon's notion of openness

@@
 and any~$x \in X(T)$ there exists an open subscheme~$T_0 \subseteq T$
-such that for any object~$(T' \xra{f}{\to} T \to S)$ of the site used to
+such that for any object~$(T' \xra{f} T \to S)$ of the site used to
 define~$\Zar(S)$ the map~$T' \to T$ factors over~$T_0$ if and only if~$X(f)(x)
@@
 \end{defn}
+
+From the point of view of the internal language of~$\Zar(S)$, a subfunctor~$U
+\hookrightarrow X$ looks like the inclusion of a subset. The following
+proposition answers the natural question how one can characterize those
+inclusions which externally correspond to open subfunctors.
 
@@
 have~$(x \in U) \vee (x \not\in U)$. If~$x \in U$, we can pick~$J = (1)$, and
-if~$x \not\in U$, we can pick~$J = (0)$ (such that the quotient~$\affl/J$ is
+if~$x \not\in U$, we can pick~$J = (0)$ (whereby the quotient~$\affl/J$ is
 isomorphic to~$\affl$, thus finitely presented and therefore in particular
@@
 is bijective. Since~$\Spec(\affl[f^{-1}]) \cong \brak{\speak{$f$ \inv}}$, the image
-of~$1$ is zero: If~$\Spec(\affl[f^{-1}])$ is inhabited, $f$ is invertible and
+of~$1$ is zero: If~$\Spec(\affl[f^{-1}])$ is inhabited, the element~$f$ is invertible and
 therefore~$x$ is an element of~$U$. This implies that~$1 \in J$.
@@
 
-\begin{rem}In view of the previous remark, one cannot hope to simplify the
-condition in Proposition~\ref{prop:characterization-open-subfunctor} to read as
+\begin{rem}In view of the previous remark, one might hope to be able to simplify the
+condition in Proposition~\ref{prop:characterization-open-subfunctor} as
 follows: ``For any~$x\?X$, the quotient~$\affl/K$ modulo the ideal~$K = \{
-f\?\affl \,|\, \speak{$f$ \inv} \Rightarrow (x \in U) \}$ is quasicoherent.
+f\?\affl \,|\, \speak{$f$ \inv} \Rightarrow (x \in U) \}$ is quasicoherent.''
+However, this doesn't work out.
 This statement implies the condition in the proposition, but the converse
@@
 
+\begin{rem}In the internal language of toposes used to carry out synthetic
+differential geometry, there is the concept of an \emph{Penon-open}
+subset~\cite[Chapitre~III]{penon}: A subset~$U \subseteq X$ is Penon-open if
+and only if
+\[ \forall x \in U\_ \forall y\?X\_
+  (x \neq y) \vee (y \in U). \]
+This notion is not useful in synthetic algebraic geometry, since it is much too
+weak: Any subset of the one-element set~$1$ is Penon-open. However, not every
+subfunctor of the terminal functor in~$\Zar(S)$ is an open subfunctor.
+\end{rem}
+
+% negneg-stability
+% definition of open truth values
+
 \end{document}

### 2017-03-15 snippets/open-subfunctors.tex
Notion of open truth values

@@
 
-% negneg-stability
-% definition of open truth values
+It's worthwhile to introduce the notion of an \emph{open truth value}. This
+files our approach to synthetic algebraic geometry in the established and more
+general framework of synthetic topology~\cite{XXX}:
+
+\begin{defn}In the context of a specified local ring, as for instance~$\affl$
+of the big Zariski topos of a scheme, a truth value~$U \subseteq 1$ is
+\emph{open} if and only if there exists an ideal~$J \subseteq \affl$ such
+that~$\affl/J$ is quasicoherent and such that~$U$ holds if and only if~$1 \in
+J$. (Section~\ref{sect:basics-on-truth-values} contains generalities on truth
+values.)\end{defn}
+
+\begin{prop}Let~$X$ be a Zariski sheaf. A subfunctor~$U \hookrightarrow X$ is
+open if and only if, from the internal point of view of~$\Zar(S)$, for
+any~$x\?X$ the truth value of~``$x \in U$'' is an open truth value.\end{prop}
+\begin{proof}Immediate from the definition and the characterization given in
+Proposition~\ref{prop:characterization-open-subfunctor}.\end{proof}
+
+\begin{prop}In~$\Zar(S)$, open truth values are~$\neg\neg$-stable: If~$U
+\subseteq 1$ is an open truth value, and if~$U$ does \notnot hold, then~$U$
+actually holds.\end{prop}
+
+\begin{proof}Since~$U$ is an open truth value, there exists an ideal~$J
+\subseteq \affl$ such that~$\affl/J$ is quasicoherent and such that~$U$ holds
+if and only if~$1 \in J$. By assumption, the element~1 is \notnot an element
+of~$J$; we want to verify that it's actually an element of~$J$.
+
+Since~$\affl/J$ is quasicoherent, the canonical homomorphism
+\[ \affl/J \longrightarrow XXX \]
+XXX
+\end{proof}
 

### 2017-03-15 snippets/choice-site.tex
On the proper choice of a suitable site for constructing the big Zariski topos

--- 2017-03-14 snippets/properties-of-a1.tex
@@
 \renewcommand{\S}{\mathcal{S}}
+\renewcommand{\U}{\mathcal{U}}
 \newcommand{\NN}{\mathbb{N}}
@@
 \newcommand{\Sch}{\mathrm{Sch}}
+\newcommand{\Aff}{\mathrm{Aff}}
 \newcommand{\Mod}{\mathrm{Mod}}
@@
 \newcommand{\tors}{\mathrm{tors}}
+\newcommand{\lfp}{\mathrm{lfp}}
+\newcommand{\fp}{\mathrm{fp}}
 \DeclareMathOperator{\Spec}{Spec}
@@
 
-\begin{prop}The following properties of~$\affl$, as seen from the internal
-point of view of~$\Zar(S)$, are consequences of~$\affl$ satisfying the
-quasicoherence condition (and~$\affl$ not being the trivial ring):
-\begin{enumerate}
-\item $\affl$ is a field in the sense that
-$\forall x\?\affl\_ \neg(x = 0) \Rightarrow \speak{$x$ \inv}$.
-\item $\affl$ is not a reduced ring: $\neg \forall x\?\affl\_ (\bigvee_{n \geq
-0} x^n = 0) \Rightarrow x = 0.$
-\item Any function~$\affl \to \affl$ is given by a unique polynomial
-in~$\affl[T]$.
-\item Any monic polynomial~$p \in \affl[T]$ of degree at least one does \notnot have a
-zero.
-\end{enumerate}\end{prop}
+Unlike with the construction of the little Zariski topos, set-theoretical
+issues of size arise when constructing the big Zariski topos. These can be
+solved in several different manners, yielding toposes which are not equivalent,
+and actually differ in some important aspects, but otherwise enjoy very similar
+properties.
 
-\begin{proof}\begin{enumerate}
-\item Let $x\?\affl$ such that~$\neg(x=0)$. We consider the quasicoherence
-condition for the finitely presented~$\affl$-algebra~$A \defeq \affl/(x)$.
-Since~$\Spec(A) \cong \brak{x=0} = \brak{\bot} = \emptyset$, the condition says
-that the canonical homomorphism
-\[ \affl/(x) \longrightarrow [\emptyset, \affl] \]
-is an isomorphism. Since its codomain is the zero algebra, so is~$\affl/(x)$.
-Therefore~$1 \in (x)$, that is,~$x$ is invertible.
-\item Assume that~$\affl$ is reduced. Then the set~$\Delta \defeq \{
-\varepsilon \in \affl \,|\, \varepsilon^2 = 0 \}$ is equal to~$\{ 0 \}$.
-By the quasicoherence criterion applied to the finitely
-presented~$\affl$-algebra~$A \defeq \affl[T]/(T^2)$, the canonical map
-\[ \affl[T]/(T^2) \longrightarrow [\Spec(\affl[T]/(T^2)), \affl] \cong
-  [\Delta, \affl] \cong \affl \]
-is an isomorphism. It maps~$[T]$ to zero (the value of~$T$ at~$0 \in \Delta$).
-Thus~$T \in (T^2)$ and therefore~$1 = 0$ in~$\affl$. This is a contradiction.
-\item Immediate by considering the quasicoherence condition for the finitely
-presented~$\affl$-algebra~$A \defeq \affl[T]$ and noticing that~$\Spec(A) \cong
-\affl$.
-\item Let~$p \in \affl[T]$ be a monic polynomial of degree at least one. Assume
-that~$p$ doesn't have a zero in~$\affl$. Then the spectrum of~$A \defeq
-\affl[T]/(p)$ is empty. The quasicoherence condition for~$A$ therefore implies
-that~$\affl[T]/(p)$ is zero. This means that~$p$ is invertible in~$\affl[T]$.
-A basic lemma in commutative algebra (whose standard proof is constructive)
-then implies that with the exception of the constant term in~$p$, all
-coefficients are nilpotent. This contradicts the assumption that~$p$ is monic
-of degree at least one.\qedhere
-\end{enumerate}\end{proof}
+Some authors construct the big Zariski topos of~$S$ as the topos of
+sheaves over the site~$\Sch/S$ of all schemes over~$S$. This option is quite
+attractive since the Yoneda functor~$\Sch/S \to \Sh(\Sch/S)$, which sends
+an~$S$-scheme to its functor of points, is fully faithful. However,
+since~$\Sch/S$ is not essentially small, forming the sheaf topos is not
+possible in plain Zermelo--Fraenkel set theory.
+Since it's still possible to meaningfully speak of individual
+functors~$(\Sch/S)^\op \to S$, we can attach a Kripke--Joyal semantics
+to~$\Sh(\Sch/S)$, as long as one keeps in mind that~$\Sh(\Sch/S)$ might
+not contain a subobject classifier.
+
+We could also assume the existence of a Grothendieck universe~$\U$
+containing~$S$ and construct~$\Zar(S)$ as the topos of sheaves over the
+small site~$\Sch_\U/S$, the category of~$S$-schemes contained in~$U$.
+By the \emph{comparison lemma}~\ref{XXX}, we could also construct~$\Zar(S)$ as
+the topos of sheaves over~$\Aff_\U/S$, the category of~$S$-schemes in~$U$ which
+are affine (as absolute schemes), and obtain an equivalent topos.
+In this case, the Yoneda functor~$\Sch/S \to \Zar(S)$ might not be faithful,
+but the restricted Yoneda functor~$\Sch_\U/S \to \Zar(S)$ is.
+
+The Stacks Project proposes a more nuanced approach, namely expanding a given
+set~$M_0$ of schemes containing~$S$ to a superset~$M$ which is closed (up to
+isomorphism) under several constructions~\cite[Tag~000H]{stacks-project}: fiber
+products, countable coproducts, domains of open and closed immersions and of
+morphisms of finite type, spectra of local rings~$\O_{X,x}$, spectra of residue
+fields, and others. The Stacks Project then defines~$\Zar(S)$
+as~$\Sh(\Sch_M/S)$, where~$\Sch_M/S$ is the small category of~$S$-schemes
+in~$M$, or equivalently as~$\Sh(\Aff_M/S)$. This approach has the advantage
+that one doesn't have to assume the existence of a Grothendieck universe.
+
+From a topos-theoretical point of view, it's natural to settle for an even
+more parsimonious site: the site~$(\Sch/S)_\lfp$ consisting of the~$S$-schemes
+which are locally of finite presentation over~$S$, or equivalently the
+essentially small site~$(\Aff/S)_\lfp$ of the~$S$-schemes which are locally of
+finite presentation over~$S$ and affine (as absolute schemes).\footnote{It's
+not reasonable to restrict to the even smaller site consisting of the finitely
+presented~$S$-schemes, since open immersions can fail to be finitely presented.
+We want the site used to construct~$\Zar(S)$ to be closed under open immersions
+for various reasons, for instance to facilitate a comparison with the little
+Zariski topos~$\Sh(S)$, whose site does contain all open subsets of~$S$.
+Furthermore, since a finitely presented~$S$-scheme might not admit an open
+covering by finitely presented~$S$-schemes which are affine (as absolute
+schemes), the toposes~$\Sh((\Sch/S)_\fp)$ and~$\Sh((\Aff/S)_\fp)$ can differ.}
+
+In the special case that~$S = \Spec(A)$ is affine, this site is the
+dual of the category of finitely presented~$A$-algebras; in this case the
+topos-theoretic points of the resulting topos are precisely the local~$A$-algebras,
+and moreover, the resulting topos is the classifying topos of the theory of
+local~$A$-algebras, such that for any Grothendieck topos~$\E$, geometric
+morphisms~$\E \to \Sh((\Aff/S)_\lfp)$ correspond to local~$A$-algebras internal
+to~$\E$.
+
+In contrast, the toposes arising when using the larger sites have categories of
+points which contain further objects in addition to all local~$A$-algebras; and
+no simple description of the theory they classify is known.
+
+A disadvantage of employing the sites with the presentation condition is that
+the Yoneda functor is only fully faithful when restricted to~$(\Sch/S)_\lfp$.
+For instance, in the case~$S = \Spec(\ZZ)$, the schemes~$\Spec(\QQ)$ and the
+empty scheme have isomorphic functors of points, whereby~$\Spec(\QQ)$ looks
+like the empty set from the internal point of view.\footnote{XXX explicit proof}
+
+In the following, we try to encompass XXX point out XXX
 

### 2017-03-15 snippets/choice-site.tex
Conclusion of the site discussion

@@
 
+\subsubsection*{Naive approach}
 Some authors construct the big Zariski topos of~$S$ as the topos of
@@
 attractive since the Yoneda functor~$\Sch/S \to \Sh(\Sch/S)$, which sends
-an~$S$-scheme to its functor of points, is fully faithful. However,
-since~$\Sch/S$ is not essentially small, forming the sheaf topos is not
-possible in plain Zermelo--Fraenkel set theory.
+an~$S$-scheme to its functor of points, is fully faithful, therefore the
+internal language of~$\Sh(\Sch/S)$ can distinguish arbitrary schemes.
+
+However, since~$\Sch/S$ is not essentially small, forming the sheaf topos is
+not possible in plain Zermelo--Fraenkel set theory.
+
 Since it's still possible to meaningfully speak of individual
 functors~$(\Sch/S)^\op \to S$, we can attach a Kripke--Joyal semantics
-to~$\Sh(\Sch/S)$, as long as one keeps in mind that~$\Sh(\Sch/S)$ might
-not contain a subobject classifier.
+to~$\Sh(\Sch/S)$, as long as we keep in mind that~$\Sh(\Sch/S)$ might
+not contain a subobject classifier and might not be cartesian closed. From the
+internal point of view, powersets and function sets might therefore not exist.
 
+\subsubsection*{Using Grothendieck universes}
 We could also assume the existence of a Grothendieck universe~$\U$
@@
 small site~$\Sch_\U/S$, the category of~$S$-schemes contained in~$U$.
+
 By the \emph{comparison lemma}~\ref{XXX}, we could also construct~$\Zar(S)$ as
@@
 are affine (as absolute schemes), and obtain an equivalent topos.
+
 In this case, the Yoneda functor~$\Sch/S \to \Zar(S)$ might not be faithful,
-but the restricted Yoneda functor~$\Sch_\U/S \to \Zar(S)$ is.
+but the restricted Yoneda functor~$\Sch_\U/S \to \Zar(S)$ will.
 
+\subsubsection*{Approach of the Stacks Project}
 The Stacks Project proposes a more nuanced approach, namely expanding a given
@@
 morphisms of finite type, spectra of local rings~$\O_{X,x}$, spectra of residue
-fields, and others. The Stacks Project then defines~$\Zar(S)$
-as~$\Sh(\Sch_M/S)$, where~$\Sch_M/S$ is the small category of~$S$-schemes
-in~$M$, or equivalently as~$\Sh(\Aff_M/S)$. This approach has the advantage
-that one doesn't have to assume the existence of a Grothendieck universe.
+fields, and others.
 
+The Stacks Project then defines~$\Zar(S)$ as~$\Sh(\Sch_M/S)$, where~$\Sch_M/S$
+is the small category of~$S$-schemes in~$M$, or equivalently
+as~$\Sh(\Aff_M/S)$. This approach has the advantage that one doesn't have to
+assume the existence of a Grothendieck universe.
+
+\subsubsection*{Emplyoing parsimonious sites}
 From a topos-theoretical point of view, it's natural to settle for an even
@@
 
-A disadvantage of employing the sites with the presentation condition is that
-the Yoneda functor is only fully faithful when restricted to~$(\Sch/S)_\lfp$.
-For instance, in the case~$S = \Spec(\ZZ)$, the schemes~$\Spec(\QQ)$ and the
-empty scheme have isomorphic functors of points, whereby~$\Spec(\QQ)$ looks
-like the empty set from the internal point of view.\footnote{XXX explicit proof}
+A further advantage of these parsimonious sites is that they don't require arbitrary
+choices of a starting set~$M_0$ or a way of expanding~$M_0$ to a sufficiently
+ample set~$M$ of schemes.
 
-In the following, we try to encompass XXX point out XXX
+However, the parsimonious sites also have a serious disadvantage, namely that
+with them, the Yoneda functor is only fully faithful when restricted
+to~$(\Sch/S)_\lfp$. For instance, in the case~$S = \Spec(\ZZ)$, the
+schemes~$\Spec(\QQ)$ and the empty scheme have isomorphic functors of points,
+whereby~$\Spec(\QQ)$ looks like the empty set from the internal point of
+view.\footnote{XXX explicit proof}
+
+In the following, we do not commit to a single one of these options for
+resolving the set-theoretical size issues, but rather keep any of these options
+in mind. This approach will sometimes necessitate phrases such as ``for
+any~$S$-scheme~$T$ contained in the site used to define~$\Zar(S)$'', which might
+seem awkward for a seasoned topos-theorist when taken out of context, since
+the site used to construct a Grothendieck topos is not at all uniquely
+determined by the resulting topos.
+
+We will indicate the few places where the choice of site makes a difference.
 

### 2017-03-15 snippets/choice-site.tex
negneg-stability of open truth values

@@
 as~$\Sh(\Aff_M/S)$. This approach has the advantage that one doesn't have to
-assume the existence of a Grothendieck universe.
+assume the existence of a Grothendieck universe; the \emph{partial
+universe}~$M$ can be constructed entirely within ZFC set theory using
+transfinite recursion.
 

### 2017-03-15 snippets/open-subfunctors.tex
negneg-stability of open truth values

@@
 
-\begin{prop}In~$\Zar(S)$, open truth values are~$\neg\neg$-stable: If~$U
-\subseteq 1$ is an open truth value, and if~$U$ does \notnot hold, then~$U$
-actually holds.\end{prop}
+In many flavours of synthetic topology, open truth values~$\varphi$
+are~$\neg\neg$-stable in that~$\neg\neg\varphi$ implies~$\varphi$. With a
+small caveat, this is true for open truth values in the big Zariski topos as
+well.
 
-\begin{proof}Since~$U$ is an open truth value, there exists an ideal~$J
+\begin{prop}\label{prop:open-truth-values-stable}
+Let~$U \hookrightarrow 1$ be a subfunctor in~$\Zar(S)$ such
+that~$\Zar(S) \models \neg\neg U$. Then in any of the following
+situations it follows that~$\Zar(S) \models U$:
+\begin{enumerate}
+\item $U$ is a quasicompact open truth value.
+\item $U$ is an arbitrary open truth value and the site used to
+define~$\Zar(S)$ is closed under domains of closed immersions.
+\end{enumerate}
+\end{prop}
+
+\begin{proof}We give two proofs, an internal one and an external one, since
+they pursue different ideas.
+
+\emph{Internal proof.} Since~$U$ is an open truth value, there exists an ideal~$J
 \subseteq \affl$ such that~$\affl/J$ is quasicoherent and such that~$U$ holds
@@
 
-Since~$\affl/J$ is quasicoherent, the canonical homomorphism
-\[ \affl/J \longrightarrow XXX \]
-XXX
+By Proposition~\ref{prop:XXX}, the canonical homomorphism
+\[ \affl/J \longrightarrow [\Spec(\affl/J), \affl] \]
+is bijective; the assumptions of that proposition are satisfied in either of
+the two situations. The set~$\Spec(\affl/J)$ is isomorphic to~$\brak{J = (0)}$.
+Since~$\neg\neg(1 \in J)$, we also have~$\neg(J = (0))$.
+Therefore~$\Spec(\affl/J)$ is empty and the codomain of the displayed
+isomorphism is the zero algebra. Thus~$\affl/J$ is trivial as well, showing~$1
+\in J$.
+
+\emph{External proof.} Since~$U \hookrightarrow 1$ is an open subfunctor, there
+is an open subscheme~$S_0 \subseteq S$ such that a morphism~$f : T \to S$
+factors over~$S_0$ if and only if~$U(T)$ is inhabited. In both situations it's
+possible to endow~$X \defeq S \subseteq S_0$ with the structure of a closed
+subscheme such that~$X$ is contained in the site used to define~$\Zar(S)$.
+By the universal property of~$S_0$, we have~$X \models \neg U$. Since~$\Zar(S)
+\models \neg\neg U$, it follows that~$X$ is empty. Therefore~$S_0 = S$ and~$U$
+is globally inhabited.
+\end{proof}
+
+\begin{cor}Let~$\gamma : \Delta \to X$ be a morphism in~$\Zar(S)$.
+Let~$U \hookrightarrow X$ be an open subfunctor such that~$\Zar(S) \models
+\gamma(0) \in U$. Then, in any of the situations in
+Proposition~\ref{prop:open-truth-values-stable}, the morphism~$\gamma$ factors
+over~$U$.
+\end{cor}
+
+\begin{proof}We give an internal proof. Let~$\varepsilon \in \Delta$.
+Then~$\neg\neg(\varepsilon = 0)$. Therefore~$\neg\neg(\gamma(\varepsilon) \in
+U)$. Since being an element of~$U$ is~$\neg\neg$-stable, it follows
+that~$\gamma(\varepsilon) \in U$.
 \end{proof}

### 2017-03-15 snippets/properties-of-a1.tex
Nullstellensatz and infinitude

@@
 quasicoherence condition (and~$\affl$ not being the trivial ring):
-% Adjust paragraph below if numbering changes.
+% Adjust paragraph below the proof if numbering changes.
+% Proof of sixth property references first property by number.
 \begin{enumerate}
@@
 in~$\affl[T]$.
-\item Any monic polynomial~$p \in \affl[T]$ of degree at least one does \notnot have a
-zero.
+\item $\affl$ is \emph{weakly algebraically closed}, in the following sense:
+Any monic polynomial~$p \in \affl[T]$ of degree at least one does \notnot have
+a zero.
+\item $\affl$ is infinite in the following sense: For any number~$n \geq 0$
+and any given elements~$x_1,\ldots,x_n \? \affl$, there is \notnot an element~$y$
+which is distinct from all of the~$x_i$.
+\item $\affl$ fulfills the conclusion of the Nullstellensatz:
+Let~$f_1,\ldots,f_m \in \affl[X_1,\ldots,X_n]$ be polynomials without a common
+zero in~$(\affl)^n$. Then there are polynomials~$g_1,\ldots,g_m \in
+\affl[X_1,\ldots,X_n]$ such that~$\sum_i g_i f_i = 1$.
 \end{enumerate}\end{prop}
@@
 coefficients are nilpotent. This contradicts the assumption that~$p$ is monic
-of degree at least one.\qedhere
+of degree at least one.
+\item The polynomial~$f(T) \defeq (T - x_1) \cdots (T - x_n) + 1$
+does \notnot have a zero~$y\?\affl$, since~$\affl$ is weakly algebraically
+closed. This element cannot equal any~$x_i$, since~$f(x_i) = 1$ is not zero.
+\item We consider the quasicoherence condition for the finitely
+presented~$\affl$-algebra~$A \defeq \affl[X_1,\ldots,X_n]/(f_1,\ldots,f_m)$.
+Since~$\Spec(A) \cong \{ x \in (\affl)^n \,|\, f_1(x) = \ldots = f_m(x) = 0 \}
+= \emptyset$, the condition implies that~$A$ is the zero algebra just as in the
+verification of the first property.\qedhere
 \end{enumerate}\end{proof}

### 2017-03-15 snippets/choice-site.tex
Affine schemes from the internal point of view of the big Zariski topos

@@
 
-\subsubsection*{Emplyoing parsimonious sites}
+\subsubsection*{Employing parsimonious sites}
 From a topos-theoretical point of view, it's natural to settle for an even

### 2017-03-15 snippets/schemes.tex
Affine schemes from the internal point of view of the big Zariski topos

--- 2017-03-14 snippets/properties-of-a1.tex
@@
 
-\begin{prop}The following properties of~$\affl$, as seen from the internal
-point of view of~$\Zar(S)$, are consequences of~$\affl$ satisfying the
-quasicoherence condition (and~$\affl$ not being the trivial ring):
+\begin{prop}A sheaf~$X \in \Zar(S)$ is isomorphic to the functor of points of
+an affine~$S$-scheme (an~$S$-scheme whose structure morphism to~$S$ is affine)
+if, in the internal language of~$\Zar(S)$,
+the~$\affl$-algebra~$[X,\affl]$ is quasicoherent and the canonical map~$X \to
+\Spec([X,\affl])$ is bijective. The converse holds in any of the following
+situations:
 \begin{enumerate}
-\item $\affl$ is a field in the sense that
-$\forall x\?\affl\_ \neg(x = 0) \Rightarrow \speak{$x$ \inv}$.
-\item $\affl$ is not a reduced ring: $\neg \forall x\?\affl\_ (\bigvee_{n \geq
-0} x^n = 0) \Rightarrow x = 0.$
-\item Any function~$\affl \to \affl$ is given by a unique polynomial
-in~$\affl[T]$.
-\item Any monic polynomial~$p \in \affl[T]$ of degree at least one does \notnot have a
-zero.
+\item The affine~$S$-scheme which~$X$ represents is of finite presentation
+over~$S$.
+\item The site used to define~$\Zar(S)$ is defined using a partial universe
+and the affine~$S$-scheme which~$X$ represents if of finite type over~$S$.
+\item The affine~$S$-scheme which~$X$ represents is contained in the site used
+to define~$\Zar(S)$. (This situation subsumes the previous ones.)
 \end{enumerate}\end{prop}
 
-\begin{proof}\begin{enumerate}
-\item Let $x\?\affl$ such that~$\neg(x=0)$. We consider the quasicoherence
-condition for the finitely presented~$\affl$-algebra~$A \defeq \affl/(x)$.
-Since~$\Spec(A) \cong \brak{x=0} = \brak{\bot} = \emptyset$, the condition says
-that the canonical homomorphism
-\[ \affl/(x) \longrightarrow [\emptyset, \affl] \]
-is an isomorphism. Since its codomain is the zero algebra, so is~$\affl/(x)$.
-Therefore~$1 \in (x)$, that is,~$x$ is invertible.
-\item Assume that~$\affl$ is reduced. Then the set~$\Delta \defeq \{
-\varepsilon \in \affl \,|\, \varepsilon^2 = 0 \}$ is equal to~$\{ 0 \}$.
-By the quasicoherence criterion applied to the finitely
-presented~$\affl$-algebra~$A \defeq \affl[T]/(T^2)$, the canonical map
-\[ \affl[T]/(T^2) \longrightarrow [\Spec(\affl[T]/(T^2)), \affl] \cong
-  [\Delta, \affl] \cong \affl \]
-is an isomorphism. It maps~$[T]$ to zero (the value of~$T$ at~$0 \in \Delta$).
-Thus~$T \in (T^2)$ and therefore~$1 = 0$ in~$\affl$. This is a contradiction.
-\item Immediate by considering the quasicoherence condition for the finitely
-presented~$\affl$-algebra~$A \defeq \affl[T]$ and noticing that~$\Spec(A) \cong
-\affl$.
-\item Let~$p \in \affl[T]$ be a monic polynomial of degree at least one. Assume
-that~$p$ doesn't have a zero in~$\affl$. Then the spectrum of~$A \defeq
-\affl[T]/(p)$ is empty. The quasicoherence condition for~$A$ therefore implies
-that~$\affl[T]/(p)$ is zero. This means that~$p$ is invertible in~$\affl[T]$.
-A basic lemma in commutative algebra (whose standard proof is constructive)
-then implies that with the exception of the constant term in~$p$, all
-coefficients are nilpotent. This contradicts the assumption that~$p$ is monic
-of degree at least one.\qedhere
-\end{enumerate}\end{proof}
+\begin{proof}The ``if'' direction is straightforward, since the assumption
+expresses~$X$ as the functor of points of the relative spectrum of a
+quasicoherent~$\O_S$-algebra.
+
+For the ``only if'' direction, we abuse notation and denote the given
+affine~$S$-scheme whose functor of points is~$X$ by~``$f : X \to S$''.
+Then~$f_*\O_X$ is quasicoherent and the canonical morphism~$X \to
+\RelSpec{S}{f_*\O_X}$ is an isomorphism. In any of the listed situations, the
+internal Hom~$[X,\affl]$ is canonically isomorphic to~$(f_*\O_X)^\Zar$, since
+for any object~$T \xra{g} S$ of the site used to define~$\Zar(S)$ we have that
+\begin{align*}
+  [X,\affl](T) &\cong
+  \Hom_{\Zar(S)}(\underline{T}, [X,\affl]) \cong
+  \Hom_{\Zar(S)}(\underline{T} \times X,\affl) \\
+  &\cong \Hom_{\Zar(S)}(\underline{T} \times \underline{X},\affl) \cong
+  \Hom_{\Zar(S)}(\underline{T \times_S X},\affl) \\
+  &\cong \affl(T \times_S X) \cong
+  (g^* f_* \O_X)(T) =
+  (f_* \O_X)^\Zar(T).
+\end{align*}
+Therefore~$[X,\affl]$ is quasicoherent. The map induced by the isomorphism~$X \to
+\RelSpec{S}{f_*\O_X}$ on the level of functors of points is precisely the
+canonical map~$X \to \Spec([X,\affl])$ as defined in the internal language;
+therefore this map is bijective from the internal point of view.
+\end{proof}
 

### 2017-03-16 snippets/qcoh.tex
Proof of the quasicoherence criterion in the big Zariski topos

\documentclass[10pt,reqno]{amsart}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{etex}
\usepackage{amsmath,amsthm,amssymb,stmaryrd,color,graphicx,multirow}
\usepackage{mathtools}
\usepackage{setspace}
\usepackage{bussproofs}
\usepackage{xspace}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{array}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage[bookmarksdepth=2,pdfencoding=auto]{hyperref}
\usepackage[all]{xy}
\usepackage{soul}\setul{0.3ex}{}

\usepackage{tikz}
\usetikzlibrary{calc,shapes.callouts,shapes.arrows}
\newcommand{\hcancel}[5]{%
    \tikz[baseline=(tocancel.base)]{
        \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
        \draw[red, line width=0.3mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
    }%
}

\usepackage[natbib=true,style=numeric]{biblatex}
\usepackage[babel]{csquotes}
\bibliography{bibliography}

\theoremstyle{definition}
\newtheorem{defn}{Definition}[section]
\newtheorem{ex}[defn]{Example}

\theoremstyle{plain}

\newtheorem{prop}[defn]{Proposition}
\newtheorem{cor}[defn]{Corollary}
\newtheorem{lemma}[defn]{Lemma}
\newtheorem{thm}[defn]{Theorem}

\theoremstyle{remark}
\newtheorem{rem}[defn]{Remark}
\newtheorem{question}[defn]{Question}

\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\FF}{\mathbb{F}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\A}{\mathcal{A}}
\renewcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\renewcommand{\H}{\mathcal{H}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\L}{\mathcal{L}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\GG}{\mathbb{G}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\qqq}{\mathfrak{q}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\nnn}{\mathfrak{n}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\HOM}{\mathcal{H}\mathrm{om}}
\newcommand{\id}{\mathrm{id}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\placeholder}{\underline{\quad}}
\let\oldul\ul
\renewcommand{\ul}[1]{\text{\oldul{$#1$}}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Grp}{\mathrm{Grp}}
\newcommand{\Vect}{\mathrm{Vect}}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\PSh}{\mathrm{PSh}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\Et}{\mathrm{\acute{E}t}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Mod}{\mathrm{Mod}}
\newcommand{\Alg}{\mathrm{Alg}}
\newcommand{\Ring}{\mathrm{Ring}}
\newcommand{\LRL}{\mathrm{LRL}}
\newcommand{\pt}{\mathrm{pt}}
\newcommand{\tors}{\mathrm{tors}}
\DeclareMathOperator{\Spec}{Spec}
\newcommand{\RelSpek}{\operatorname{\ul{\mathrm{Spec}}}}
\newcommand{\QcohSpec}[2]{\mathrm{Spec}^{\mathrm{qcoh}}_{#1}{#2}}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\colim}{colim}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\Ann}{Ann}
\DeclareMathOperator{\Int}{int}
\DeclareMathOperator{\Clos}{cl}
\DeclareMathOperator{\Kernel}{ker}
\DeclareMathOperator{\supp}{supp}
\newcommand{\Ass}{\mathrm{Ass}}
\newcommand{\Open}{\T}
\newcommand{\?}{\,{:}\,}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\Ll}{\vcentcolon\Longleftrightarrow}
\newcommand{\notat}[1]{{!#1}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\lhra}{\ensuremath{\lhook\joinrel\relbar\joinrel\rightarrow}}
\newcommand{\hra}{\hookrightarrow}
\newcommand{\brak}[1]{{\llbracket{#1}\rrbracket}}
\newcommand{\sdense}{{\widehat\Box}}
\newcommand{\sdenseother}{\Box}
\newcommand{\ie}{i.\,e.\@\xspace}
\newcommand{\eg}{e.\,g.\@\xspace}
\newcommand{\vs}{vs.\@\xspace}
\newcommand{\resp}{resp.\@\xspace}
\newcommand{\inv}{inv.\@}
\newcommand{\notnot}{\emph{not not}\xspace}
\newcommand{\affl}{\ensuremath{{\ul{\AA}^1_S}}\xspace}
\newcommand{\afflx}{\ensuremath{{\ul{\AA}^1_X}}\xspace}
\newcommand{\affla}{\ensuremath{{\ul{\AA}^1_{\Spec A}}}\xspace}
\newcommand{\xra}{\xrightarrow}

\newcommand{\XXX}[1]{\textbf{XXX: #1}}

\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defequiv}{\vcentcolon\equiv}
\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}

\newcommand{\stacksproject}[1]{\cite[{\href{http://stacks.math.columbia.edu/tag/#1}{Tag~#1}}]{stacks-project}}

\definecolor{gray}{rgb}{0.7,0.7,0.7}

\begin{document}

\begin{defn}An~$R$-module~$E$ is \emph{synthetically quasicoherent} if and only if,
for any finitely presented~$R$-algebra~$A$, the canonical~$R$-algebra
homomorphism
\[ E \otimes_R A \longrightarrow [\Spec(A), E] = [[A, R]_R, E] \]
which maps a pure tensor~$x \otimes f$ to~$(\varphi \mapsto \varphi(f) x)$ is
bijective.\end{defn}

\begin{thm}Let~$E \in \Zar(S)$ be an~$\affl$-algebra.
If~$E$ is quasicoherent, that is of
the form~$\E_0^\Zar$ for some quasicoherent~$\O_S$-module~$\E_0$,
then~$E$ is synthetically quasicoherent from the internal point of view of~$\Zar(S)$.
The converse holds in any of the following situations:
\begin{enumerate}
\item The site used to construct~$\Zar(S)$ is one of the parsimonious sites.
\item The base scheme~$S$ is concentrated (quasicompact and quasiseparated) and
the functor~$E$ maps directed limits of inverse systems of~$S$-schemes with
affine transition morphisms to colimits in~$\Set$.
\end{enumerate}
\end{thm}

\begin{proof}To verify that~$E$ is synthetically quasicoherent, we have to
verify a condition for~$\affl$-algebras~$A$ in any
slice~$\Zar(S)/\ul{T}$. If such an algebra is finitely presented from
the internal point of view, then there is a covering~$T = \bigcup_i T_i$ such
that the each of the restrictions of the algebra to the~$T_i$ is of the
form~$(\A_0)^\Zar$ for some finitely presented~$\O_{T_i}$-algebra~$\A_0$.
Without loss of generality, we will just assume that~$A$ itself is of the
form~$(\A_0)^\Zar$ for a finitely presented~$\O_S$-algebra~$\A_0$.

By Proposition~\ref{XXX}, the internal expression~$\Spec(A)$ is the functor of
points of~$\RelSpek_S \A_0$. For any~$S$-scheme~$f : T \to S$ contained in the site
used to define~$\Zar(S)$, we consider the fiber product
\[ \xymatrix{
  \RelSpek_T(f^*\A_0) \ar[r]^{f'} \ar[d]_{p'} & \RelSpek_S\A_0 \ar[d]^p \\
  T \ar[r]_f & S.
} \]
Since~$\RelSpek_T(f^*\A_0) \to S$ is contained in the site (for any of our
admissible sites), we may conclude using the following chain of isomorphisms:
\begin{align*}
  [\Spec(A), E](T) &\cong
  \Hom_{\Zar(S)}(\ul{T}, [\Spec(A), E])
  \cong \Hom_{\Zar(S)}(\ul{T} \times \Spec(A), E) \\
  &\cong \Hom_{\Zar(S)}(\underline{T \times_S \RelSpek_S\A_0}, E)
  \cong E(\RelSpek_T(f^*\A_0)) \\
  &\cong \Gamma(\RelSpek_T(f^*\A_0), (p')^* f^* \E_0)
  \cong \Gamma(T, (p')_* (p')^* f_* \E_0) \\
  &\cong \Gamma(T, f^*\E_0 \otimes_{\O_T} f^*\A_0)
  \cong \Gamma(T, (\E_0 \otimes_{\O_S} \A_0)^\Zar) \\
  &\cong \Gamma(T, \E_0^\Zar \otimes_\affl \A_0^\Zar)
  \cong \Gamma(T, E \otimes_\affl A).
\end{align*}
The antepenultimate isomorphism is because pullback of modules in~$\Sh(S)$ to
modules in~$\Sh(T)$ commutes with tensor product. The penultimate isomorphism
is because pullback of a sheaf in~$\Sh(S)$ to a sheaf in~$\Zar(S)$ commutes
with tensor product (Lemma~\ref{lemma:zar-tensor-product-commutes}).

For the converse direction, we first verify that the restrictions~$E|_{\Sh(T)}$
to the little Zariski topos of each~$S$-scheme~$T$ contained in the site used
to define~$\Zar(S)$ are quasicoherent~$\O_T$-modules. For this, we employ the
quasicoherence criterion of Theorem~\ref{thm:qcoh-sheafchar}: For any open
affine subset~$T' \subseteq T$ and any function~$h \in \Gamma(T', \O_T)$ we
verify that the canonical morphism
\begin{equation}\label{eqn:want-iso}\tag{$\dagger$}
  E|_{\Sh(T)}[h^{-1}] \longrightarrow j_*(E|_{\Sh(D(h))})
\end{equation}
is an isomorphism, where~$j : D(h) \hookrightarrow T'$ denotes the inclusion.
This follows from the assumption of synthetic quasicoherence by considering
the~$\affl$-module~$A \defeq \affl[h^{-1}]$ (in the slice~$\Zar(S)/\ul{T'}$):
This expresses that the canonical morphism
\begin{equation}\label{eqn:have-iso}\tag{$\ddagger$}
  E \otimes_\affl \affl[h^{-1}] \longrightarrow [\Spec(A), E]
\end{equation}
is an isomorphism (of~$\affl$-modules in~$\Zar(S)/\ul{T'}$). Restricting the
domain to~$\Sh(T')$ yields the sheaf~$E|_{\Sh(T')} \otimes_{\O_{T'}}
\O_{T'}[h^{-1}]$, since restricting commutes with the geometric constructions
``forming the tensor product'' and ``localizing away from~$h$''.
Since~$\Spec(A)$ is the functor of points of~$D(h)$, restricting
the codomain to~$\Sh(T')$ yields the sheaf~$j_*(E|_{\Sh(D(h))})$.
The canonical morphism~\eqref{eqn:want-iso} which we want to recognize as an
isomorphism is therefore the restriction of the canonical
morphism~\eqref{eqn:have-iso} which we know to be an isomorphism.

A natural candidate for an quasicoherent~$\O_S$-module~$\E_0$ with~$E \cong
\E_0^\Zar$ is~$\E_0 \defeq E|_{\Sh(S)}$. We'll show that this is indeed true.
Let~$f : T \to S$ be any~$S$-scheme contained in the site used to
define~$\Zar(S)$. We assume for the time being that~$f$ is of finite
presentation and affine, so~$T \cong \RelSpek_S \A_0$ for some finitely
presented~$\O_S$-algebra~$\A_0$. We want to verify that the canonical morphism
\begin{equation}\label{eqn:want-iso2}\tag{§}
  f^*(E|_{\Sh(S)}) \longrightarrow E|_{\Sh(T)}
\end{equation}
is an isomorphism. Since the functor~$f_*$ from quasicoherent~$\O_T$-modules to
quasicoherent~$\O_S$-modules is fully faithful (the morphism~$f$ being affine)
and domain and codomain of that morphism are quasicoherent, it suffices to
verify that its image under~$f_*$ is an isomorphism. This image is the
canonical morphism
\[ E|_{\Sh(S)} \otimes_{\O_S} \A_0 \longrightarrow f_*(E|_{\Sh(T)}). \]
The assumption of synthetic quasicoherence, applied to the~$\affl$-algebra~$A
\defeq \A_0^\Zar$, shows that this morphism is an isomorphism.

In situation~(1), the only step left to do is to generalize the argument is the
previous paragraph to morphisms~$f : T \to S$ which are locally of finite
presentation. This works out because there are open covers of~$S$ and~$T$ such
that the appropriate restrictions of~$f$ are of finite presentation and affine.
The assumption of synthetic quasicoherence then needs to be applied to
to~$\affl$-algebras in suitable slices of~$\Zar(S)$, showing that the canonical
morphism~\eqref{eqn:want-iso2} is locally an isomorphism and therefore globally
as well.

In situation~(2), we employ the technique of approximating general~$S$-schemes
by~$S$-schemes of finite presentation. Specifically, let~$f : T \to S$ be an
arbitrary~$S$-scheme contained in the site used to define~$\Zar(S)$. Without
loss of generality, we may assume that~$T$ is an affine scheme. Thus~$T$ is
quasicompact and quasiseparated, and~$S$ is quasiseparated. We may therefore
apply the lemma of relative approximation~\stacksproject{09MV} to deduce
that~$T$ is a directed limit of an inverse system of~$S$-schemes~$f_i : T_i \to
S$ of finite presentation with affine transition maps. These~$S$-schemes
inherit quasicompactness and quasiseparatedness from~$S$. Therefore we can
apply a comparison result on the categories of quasicoherent
modules~\stacksproject{01Z0}:
\[
  E(T) = E(\lim_i T_i) \cong \colim_i E(T_i) \cong
  \colim_i \Gamma(T_i, f_i^* \E_0)
  \cong \Gamma(T, f^* \E_0). \qedhere
\]
\end{proof}

\end{document}

### 2017-03-16 snippets/qcoh.tex
Interpretation of the synthetic definition

@@
 
-\begin{thm}Let~$E \in \Zar(S)$ be an~$\affl$-algebra.
+This definition has the following interpretation. The codomain of the displayed
+canonical map is the set of all~$E$-valued functions on~$\Spec(A)$. Elements
+of~$E \otimes_R A$ induce such functions; these induced functions can
+reasonably be called ``algebraic''. In a synthetic context, there should be no
+other~$E$-valued functions as these algebraic ones, and different algebraic
+expressions should yield different functions. This is precisely what the
+postulated bijectivity expresses.
+
+\begin{thm}\label{thm:qcoh-big-char}
+Let~$E \in \Zar(S)$ be an~$\affl$-module.
 If~$E$ is quasicoherent, that is of

### 2017-03-16 snippets/qcoh.tex
Easy scholium of the proof

@@
 \newtheorem{thm}[defn]{Theorem}
+\newtheorem{scholium}[defn]{Scholium}
 
@@
 loss of generality, we may assume that~$T$ is an affine scheme. Thus~$T$ is
-quasicompact and quasiseparated, and~$S$ is quasiseparated. We may therefore
+quasicompact and quasiseparated, and~$S$ is quasiseparated by assumption. We may therefore
 apply the lemma of relative approximation~\stacksproject{09MV} to deduce
@@
 S$ of finite presentation with affine transition maps. These~$S$-schemes
-inherit quasicompactness and quasiseparatedness from~$S$. Therefore we can
-apply a comparison result on the categories of quasicoherent
+are contained in the site used to define~$\Zar(S)$. Furthermore, they inherit
+quasicompactness and quasiseparatedness from~$S$. Therefore we can apply a
+comparison result on the categories of quasicoherent
 modules~\stacksproject{01Z0}:
@@
 
+\begin{scholium}Let~$E \in \Zar(S)$ be a quasicoherent~$\affl$-module.
+Let~$A \in \Zar(S)$ be a quasicoherent~$\affl$-algebra such that~$\Spec(A) \in
+\Zar(S)$ is representable by an object of the site used to define~$\Zar(S)$.
+Then the canonical morphism
+\[ E \otimes_\affl A \longrightarrow [\Spec(A), E] \]
+is an isomorphism.
+\end{scholium}
+
+\begin{proof}The second paragraph of the proof of
+Theorem~\ref{thm:qcoh-big-char} applies.\end{proof}
+
+% PSh
+
 \end{document}

### 2017-03-17 snippets/qcoh.tex
Necessity of the representation property

@@
 
-\begin{scholium}Let~$E \in \Zar(S)$ be a quasicoherent~$\affl$-module.
+\begin{scholium}\label{scholium}
+Let~$E \in \Zar(S)$ be a quasicoherent~$\affl$-module.
 Let~$A \in \Zar(S)$ be a quasicoherent~$\affl$-algebra such that~$\Spec(A) \in
 \Zar(S)$ is representable by an object of the site used to define~$\Zar(S)$.
+% XXX: "locally representable"?
 Then the canonical morphism
@@
 
-% PSh
+\begin{rem}The condition in Scholium~\ref{scholium} that~$\Spec(A)$ is
+representable by an object of the site used to define~$\Zar(S)$ cannot be
+dropped. For instance, if we employ the parsimonious sites and consider~$S =
+\Spec \ZZ$,~$A = \K_S^\Zar$, and~$E = \affl$, then~$\Spec(A) = \emptyset$.
+Therefore the codomain of the canonical morphism is the zero algebra, but the
+domain is not.
+\end{rem}
 

### 2017-03-17 snippets/qcoh.tex
More details on the counterexample

@@
 If~$E$ is quasicoherent, that is of
-the form~$\E_0^\Zar$ for some quasicoherent~$\O_S$-module~$\E_0$,
+the form~$(\E_0)^\Zar$ for some quasicoherent~$\O_S$-module~$\E_0$,
 then~$E$ is synthetically quasicoherent from the internal point of view of~$\Zar(S)$.
@@
   \cong \Gamma(T, (\E_0 \otimes_{\O_S} \A_0)^\Zar) \\
-  &\cong \Gamma(T, \E_0^\Zar \otimes_\affl \A_0^\Zar)
+  &\cong \Gamma(T, (\E_0)^\Zar \otimes_\affl (\A_0)^\Zar)
   \cong \Gamma(T, E \otimes_\affl A).
@@
 A natural candidate for an quasicoherent~$\O_S$-module~$\E_0$ with~$E \cong
-\E_0^\Zar$ is~$\E_0 \defeq E|_{\Sh(S)}$. We'll show that this is indeed true.
+(\E_0)^\Zar$ is~$\E_0 \defeq E|_{\Sh(S)}$. We'll show that this is indeed true.
 Let~$f : T \to S$ be any~$S$-scheme contained in the site used to
@@
 The assumption of synthetic quasicoherence, applied to the~$\affl$-algebra~$A
-\defeq \A_0^\Zar$, shows that this morphism is an isomorphism.
+\defeq (\A_0)^\Zar$, shows that this morphism is an isomorphism.
 
@@
 dropped. For instance, if we employ the parsimonious sites and consider~$S =
-\Spec \ZZ$,~$A = \K_S^\Zar$, and~$E = \affl$, then~$\Spec(A) = \emptyset$.
-Therefore the codomain of the canonical morphism is the zero algebra, but the
+\Spec \ZZ$,~$E = \affl$, and $A = \K_S^\Zar$ (where~$\K_S$ is the sheaf of
+rational functions on~$S$, which in this case is the constant sheaf~$\ul{\QQ}$),
+then~$\Spec(A)$ is the functor of points of the~$\ZZ$-scheme~$\Spec(\QQ)$.
+This functor coincides with the functor of points of the empty~$\ZZ$-scheme on
+the parsimonious sites; therefore~$\Spec(A) = \emptyset$ from the internal point
+of view. Thus the codomain of the canonical morphism is the zero algebra, but the
 domain is not.

### 2017-03-18 snippets/qcoh.tex
Weaker and more natural condition

@@
 \Zar(S)$ is representable by an object of the site used to define~$\Zar(S)$.
-% XXX: "locally representable"?
 Then the canonical morphism
@@
 \begin{rem}The condition in Scholium~\ref{scholium} that~$\Spec(A)$ is
-representable by an object of the site used to define~$\Zar(S)$ cannot be
-dropped. For instance, if we employ the parsimonious sites and consider~$S =
-\Spec \ZZ$,~$E = \affl$, and $A = \K_S^\Zar$ (where~$\K_S$ is the sheaf of
-rational functions on~$S$, which in this case is the constant sheaf~$\ul{\QQ}$),
-then~$\Spec(A)$ is the functor of points of the~$\ZZ$-scheme~$\Spec(\QQ)$.
-This functor coincides with the functor of points of the empty~$\ZZ$-scheme on
-the parsimonious sites; therefore~$\Spec(A) = \emptyset$ from the internal point
-of view. Thus the codomain of the canonical morphism is the zero algebra, but the
-domain is not.
+representable by an object of the site used to define~$\Zar(S)$ is slightly
+unnatural from a topos-theoretic point of view, since the conclusion of the
+Scholium depends only on the topos over the site and not the site itself.
+In fact, the condition can be weakened and made more natural at the
+same time: It suffices to require that~$\Spec(A)$ is \emph{locally}
+representable by an object of the site.
+
+However, the condition can't be dropped completely. For instance, if we employ the
+parsimonious sites and consider~$S = \Spec \ZZ$,~$E = \affl$, and $A =
+\K_S^\Zar$ (where~$\K_S$ is the sheaf of rational functions on~$S$, which in
+this case is the constant sheaf~$\ul{\QQ}$), then~$\Spec(A)$ is the functor of
+points of the~$\ZZ$-scheme~$\Spec(\QQ)$. This functor coincides with the
+functor of points of the empty~$\ZZ$-scheme on the parsimonious sites;
+therefore~$\Spec(A) = \emptyset$ from the internal point of view. Thus the
+codomain of the canonical morphism is the zero algebra, but the domain is not.
 \end{rem}

### 2017-03-18 snippets/properties-of-a1.tex
Two further properties of A^1

@@
 \begin{prop}The following properties of~$\affl$, as seen from the internal
-point of view of~$\Zar(S)$, are consequences of~$\affl$ satisfying the
-quasicoherence condition (and~$\affl$ not being the trivial ring):
+point of view of~$\Zar(S)$, are consequences of~$\affl$ being (a local ring and) synthetically
+quasicoherent:
 % Adjust paragraph below the proof if numbering changes.
-% Proof of sixth property references first property by number.
+% Proof of third and seventh property references first property by number.
 \begin{enumerate}
-\item $\affl$ is a field in the sense that
-$\forall x\?\affl\_ \neg(x = 0) \Rightarrow \speak{$x$ \inv}$.
+\item $\affl$ is a field in the sense that any element which is not zero is
+invertible: $\forall x\?\affl\_ \neg(x = 0) \Rightarrow \speak{$x$ \inv}$. More generally,
+for any number~$n \geq 0$,
+\begin{multline*}
+  \qquad\qquad \forall x_1,\ldots,x_n\?\affl\_
+  \neg(x_1 = 0 \wedge \cdots \wedge x_n = 0) \Longrightarrow \\
+  (\speak{$x_1$ \inv} \vee \cdots \vee \speak{$x_n$ \inv}). \qquad\qquad
+\end{multline*}
 \item $\affl$ is not a reduced ring: $\neg \forall x\?\affl\_ (\bigvee_{n \geq
 0} x^n = 0) \Rightarrow x = 0.$
+\item Any element of $\affl$ which is not invertible is nilpotent.
 \item Any function~$\affl \to \affl$ is given by a unique polynomial
@@
 \begin{proof}\begin{enumerate}
-\item Let $x\?\affl$ such that~$\neg(x=0)$. We consider the quasicoherence
+\item Let $x\?\affl$ be such that~$\neg(x=0)$. We consider the quasicoherence
 condition for the finitely presented~$\affl$-algebra~$A \defeq \affl/(x)$.
@@
 Therefore~$1 \in (x)$, that is,~$x$ is invertible.
+
+The more general statement follows in the same way, by using the quasicoherence
+condition for~$A \defeq \affl/(x_1,\ldots,x_n)$. This yields~$1 \in
+(x_1,\ldots,x_n)$. Since~$\affl$ is a local ring, one of the~$x_i$ is
+invertible.
+
 \item Assume that~$\affl$ is reduced. Then the set~$\Delta \defeq \{
@@
 Thus~$T \in (T^2)$ and therefore~$1 = 0$ in~$\affl$. This is a contradiction.
+
+\item Let~$x \? \affl$ be not invertible. We consider the quasicoherence
+condition for the finitely presented~$\affl$-algebra~$A \defeq \affl[x^{-1}]$.
+Since~$\Spec(A) \cong \brak{\speak{$x$ \inv}} = \emptyset$, it follows
+that~$\affl[x^{-1}] = 0$, similarly to the proof of the first claim. Thus~$x$
+is nilpotent.
+
 \item Immediate by considering the quasicoherence condition for the finitely
@@
 \affl$.
+
 \item Let~$p \in \affl[T]$ be a monic polynomial of degree at least one. Assume
@@
 of degree at least one.
+
 \item The polynomial~$f(T) \defeq (T - x_1) \cdots (T - x_n) + 1$
@@
 closed. This element cannot equal any~$x_i$, since~$f(x_i) = 1$ is not zero.
+
 \item We consider the quasicoherence condition for the finitely

### 2017-03-19 snippets/grassmannian.tex
Grassmannian in the synthetic approach

--- 2017-03-14 snippets/open-subfunctors.tex
@@
 \renewcommand{\S}{\mathcal{S}}
+\newcommand{\V}{\mathcal{V}}
 \newcommand{\NN}{\mathbb{N}}
@@
 \newcommand{\Ass}{\mathrm{Ass}}
+\newcommand{\Sym}{\mathrm{Sym}}
 \newcommand{\Open}{\T}
@@
 
+\newcommand{\Gr}{\mathrm{Gr}}
+
 \definecolor{gray}{rgb}{0.7,0.7,0.7}
@@
 
-A basic concept in the functor-of-points approach to algebraic geometry is the
-concept of an \emph{open subfunctor}. It is used to delimit schemes from more
-general kinds of spaces: A functor is deemed to be a scheme if and only if it
-admits a covering by open subfunctors which are representable.
+Let~$S$ be a base scheme and~$\V$ a finite locally free~$\O_S$-module.
+We want to illustrate the synthetic approach by verifying using the internal
+language of~$\Zar(S)$ the basic fact that the Grassmannian~$\Gr(\V,r)$ of
+rank-$r$ locally free quotients of~$\V$, defined as a certain functor of points,
+is representable by an~$S$-scheme of finite presentation.
 
-The following definition is phrased in such a way as to apply to any of the
-several ways to define the big Zariski topos~$\Zar(S)$. In particular, it
-applies to the definition using the site consisting of affine schemes which are
-locally of finite presentation over~$S$. If~$S$ is affine, the definition only
-refers to affine schemes and open subschemes of affine schemes and is therefore
-suitable if one wants to found the theory of schemes using the functorial
-approach.
-
-\begin{defn}[{\cite[Définition~I.1.3.6 on page~10]{demazure:gabriel},
-\cite[Tag~01JI]{stacks-project}}]
-A subfunctor~$U \hookrightarrow X$ in~$\Zar(S)$ is an \emph{open subfunctor} if
-and only if for any object~$(T \to S)$ of the site used to define~$\Zar(S)$
-and any~$x \in X(T)$ there exists an open subscheme~$T_0 \subseteq T$
-such that for any object~$(T' \xra{f} T \to S)$ of the site used to
-define~$\Zar(S)$ the map~$T' \to T$ factors over~$T_0$ if and only if~$X(f)(x)
-\in U(T')$.
+\begin{defn}The \emph{Grassmannian}~$\Gr(\V,r)$ is the functor which associates
+to an~$S$-scheme~$f : T \to S$ the set
+\[ \Gr(\V,r)(T) \defeq \{
+  \text{$U \subseteq f^*\V$ sub-$\O_T$-module} \,|\,
+  \text{$(f^*\V)/U$ is locally free of rank~$r$} \}. \]
 \end{defn}
 
-From the point of view of the internal language of~$\Zar(S)$, a subfunctor~$U
-\hookrightarrow X$ looks like the inclusion of a subset. The following
-proposition answers the natural question how one can characterize those
-inclusions which externally correspond to open subfunctors.
+\begin{defn}The \emph{synthetic Grassmannian} of rank-$r$ quotients of a
+module~$V$ is the set
+\[ \Gr(V,r) \defeq \{ \text{$U \subseteq V$ submodule} \,|\,
+  \text{$V/U$ is free of rank~$r$} \}. \]
+\end{defn}
 
-\begin{prop}\label{prop:characterization-open-subfunctor}
-Let~$X \in \Zar(S)$ be a Zariski sheaf. A subfunctor~$U
-\hookrightarrow X$ is open if and only if, from the internal point of view
-of~$\Zar(S)$, it holds that for all~$x \? X$ there is an ideal~$J \subseteq
-\affl$ such that~$\affl/J$ is quasicoherent and such that~$x \in U$ if and only
-if~$1 \in J$. In formal language:
-\begin{multline*}
-  \qquad\qquad\Zar(S) \models
-  \forall x\?X\_
-  \exists J \subseteq \affl\_ \\
-  \speak{$J$ is an ideal} \wedge
-  \speak{$\affl/J$ is quasicoherent} \wedge
-  (x \in U \Leftrightarrow 1 \in J). \end{multline*}
+\begin{prop}The synthetic Grassmannian of~$\V$, as constructed by the internal
+language of~$\Zar(S)$ where~$\V$ looks like an ordinary free module, coincides
+with the functorially defined Grassmannian.\end{prop}
+
+\begin{proof}Immediate from
+Definition~\ref{defn:interpretation-internal-constructions} and
+the proof of Proposition~\ref{prop:locally-free-big-zariski}.
+\end{proof}
+
+Having established that the internally constructed synthetic Grassmannian
+actually describes the external Grassmannian which we're interested in, we can
+switch to a fully internal perspective. We'll reflect this switch notationally
+by referring to the~$\affl$-module~$V \defeq \V^\Zar$ instead of~$\V$.
+
+We define for any free submodule~$W \subseteq V$ of rank~$r$ the subset
+\[ G_W \defeq \{ U \in \Gr(V,r) \,|\, \text{$W \to V \to V/U$ is bijective} \}. \]
+This sets admits a more concrete description, since it is in canonical bijection
+to the set
+\[ G_W' \defeq \{ \pi : V \to W \,|\, \pi \circ \iota = \id \} \]
+of all splittings of the inclusion~$\iota : W \hookrightarrow V$: An element~$U
+\in G_W$ corresponds to the splitting~$V \twoheadrightarrow V/U
+\xrightarrow{({\cong})^{-1}} W$. Conversely, a splitting~$\pi$ corresponds to~$U
+\defeq \ker(\pi) \in G_W$.
+
+\begin{prop}The union of the subsets~$G_W$ is~$\Gr(V,r)$.\end{prop}
+
+\begin{proof}Let~$U \in \Gr(V,r)$. Then there exists a
+basis~$([v_1],\ldots,[v_r])$ of~$V/U$. The family~$(v_1,\ldots,v_r)$ is
+linearly independent in~$V$, therefore the submodule~$W \defeq
+\operatorname{span}(v_1,\ldots,v_r) \subseteq V$ is free of rank~$r$. The
+canonical linear map~$W \hookrightarrow V \twoheadrightarrow V/U$ maps
+the basis~$(v_i)_i$ to the basis~$([v_i])_i$ and is therefore bijective. Thus~$U
+\in G_W$.\end{proof}
+
+\begin{prop}The sets~$G_W$ are (quasicompact-)open subsets of~$\Gr(V,r)$.
 \end{prop}
 
-The displayed condition is only meaningful in an intuitionistic context as
-provided by the big Zariski topos. In classical logic, the condition would be
-trivially satisfied for any subfunctor~$U \hookrightarrow X$: Classically, we
-have~$(x \in U) \vee (x \not\in U)$. If~$x \in U$, we can pick~$J = (1)$, and
-if~$x \not\in U$, we can pick~$J = (0)$ (whereby the quotient~$\affl/J$ is
-isomorphic to~$\affl$, thus finitely presented and therefore in particular
-quasicoherent).\footnote{Strictly speaking, incompability with classical
-logical surfaces even earlier: in our internal quasicoherence condition. The
-map~$E \otimes_\affl A \to [\Spec(A), E]$ which the condition demands to be
-bijective has hardly any chance to be surjective if the law of excluded middle
-is available to define maps~$\Spec(A) \to E$ by case distinction.}
+\begin{proof}Let~$U \in \Gr(V,r)$. Then~$U \in G_W$ if and only if
+the canonical linear map~$W \hookrightarrow V \twoheadrightarrow V/U$
+is bijective. Since~$W$ and~$V/U$ are both free modules of rank~$r$, this map is
+given by an~$(r \times r)$-matrix~$M$ over~$\affl$; therefore it's bijective if
+and only if the determinant of~$M$ is invertible.
 
-\begin{cor}\label{cor:sufficient-criterion-open-subfunctor}
-Let~$X \in \Zar(S)$ be a Zariski sheaf. Let~$U \hookrightarrow X$ be
-a subfunctor. If
-\[
-  \Zar(S) \models
-  \forall x\?X\_
-  \bigvee_{n \geq 0} \exists f_1,\ldots,f_n\?\affl\_
-  (x \in U \Leftrightarrow \bigvee_i \speak{$f_i$ \inv}),
-\]
-then the subfunctor is open.
-\end{cor}
+Thus we've found a number which is invertible if and only if~$U \in G_W$. By
+Corollary~\ref{cor:sufficient-criterion-open-subfunctor}, the truth value of~``$U
+\in G_W$'' is open.\end{proof}
 
-\begin{proof}[Proof of Corollary~\ref{cor:sufficient-criterion-open-subfunctor}]
-We show that the assumption implies the displayed condition of
-Proposition~\ref{prop:characterization-open-subfunctor} in the internal
-language. Given elements~$f_1,\ldots,f_n$ as in the assumption, we
-construct the ideal~$J \defeq (f_1,\ldots,f_n) \subseteq \affl$. The
-quotient~$\affl/J$ is indeed quasicoherent, being finitely presented, and
-the statement that~$1 \in J$ is equivalent to one of
-the~$f_i$ being invertible by locality of~$\affl$.
+\begin{prop}The sets $G_W$ are affine. Moreover, the algebras which the~$G_W$ are
+spectra of are finitely presented.\end{prop}
+
+\begin{proof}The set of all linear maps~$V \to W$ is the spectrum of
+the~$\affl$-algebra~$A \defeq \Sym(\Hom_\affl(V,W)^\vee)$, since
+\begin{align*}
+  \Spec(A) &=
+  \Hom_{\mathrm{Alg}(\affl)}(\Sym(\Hom_{\mathrm{Mod}(\affl)}(V,W)^\vee), \affl) \\
+  &\cong \Hom_{\mathrm{Mod}(\affl)}(\Hom_{\mathrm{Mod}(\affl)}(V,W)^\vee, \affl) \\
+  &=\Hom_{\mathrm{Mod}(\affl)}(V,W)^{\vee\vee} \\
+  &\cong \Hom_{\mathrm{Mod}(\affl)}(V,W).
+\end{align*}
+In the last step the assumption that not only~$W$, but also~$V$ is a free module
+of finite rank enters. (This is the first time in this development that we need
+this assumption.)
+
+The set~$G_W'$ is a closed subset of this spectrum, namely the locus where the
+generic linear map~$V \to W$ is a splitting of the inclusion~$\iota : W
+\hookrightarrow V$. If we choose bases of~$V$ and~$W$,
+whereby~$\Sym(\Hom_\affl(V,W)^\vee)$ is isomorphic
+to~$\affl[M_{11},\ldots,M_{rn}]$, we can be more explicit: The set~$G_W'$ is
+isomorphic to
+\[ \Spec(k[M_{11},\ldots,M_{rn}]/(MN-I)), \]
+where~$I$ is the~$(r \times r)$ identity matrix, $M$ is the generic matrix~$M =
+(M_{ij})_{ij}$, and~$N$ is the matrix of~$\iota$ with respect to the chosen
+bases. The notation~``$(MN-I)$'' denotes the ideal generated by the entries
+of~$MN-I$.
 \end{proof}
 
-\begin{proof}[Proof of Proposition~\ref{prop:characterization-open-subfunctor}]
-...
+\end{document}
+
+\begin{cor}Die Grassmannsche~$\Gr(V,r)$ ist ein Schema von endlichem
+Typ.\end{cor}
+
+\begin{proof}Die Grassmannsche~$\Gr(V,r)$ besitzt eine offene
+Überdeckung durch die affinen Schemata~$G_W$ und ist daher ein Schema.
+
+Wenn wir einen Isomorphismus~$V \cong k^n$ wählen, sehen wir, dass
+schon~$\binom{n}{r}$ viele dieser affinen Schemata genügen: nämlich diese,
+wo~$W$ einer der Standarduntermoduln von~$k^n$ ist (erzeugt durch
+Einheitsvektoren).
+
+Denn ist~$U \in \Gr(k^n,r)$, so bildet die Surjektion~$V \to V/U$ die Basis von
+mindestens einem dieser Standarduntermoduln auf eine Basis ab und ist daher
+bijektiv. (Aus einer surjektiven~$(r \times n)$-Matrix über einem lokalen Ring
+kann man stets~$r$ Spalten auswählen, die eine linear unabhängige Familie
+bilden.)
 \end{proof}
 
-\begin{rem}The radical~$\sqrt{J}$ of the ideal~$J$ appearing in
-Proposition~\ref{prop:characterization-open-subfunctor} is unique:
-It is equal to the radical ideal
-\[ K \defeq \{ f\?\affl \,|\, \speak{$f$ \inv} \Rightarrow (x \in U) \}
-\subseteq \affl. \]
-It's obvious that~$J \subseteq K$ and therefore~$\sqrt{J} \subseteq K$.
-For the converse direction, let~$f \in K$ be given. Since~$\affl/J$ is
-quasicoherent, the canonical map
-\[ (\affl/J)[f^{-1}] \longrightarrow [ \Spec(\affl[f^{-1}]), \affl/J ] \]
-is bijective. Since~$\Spec(\affl[f^{-1}]) \cong \brak{\speak{$f$ \inv}}$, the image
-of~$1$ is zero: If~$\Spec(\affl[f^{-1}])$ is inhabited, the element~$f$ is invertible and
-therefore~$x$ is an element of~$U$. This implies that~$1 \in J$.
-Thereby~$\affl/J = 0$. By injectivity of the canonical map, the
-algebra~$(\affl/J)[f^{-1}]$ is zero. Therefore~$f^n \in J$ for some natural
-number~$n$.\end{rem}
+\begin{prop}Für den Tangentialraum an~$U \in \Gr(V,r)$ gilt:
+$T_U \Gr(V,r) \cong \Hom(U, V/U)$.\end{prop}
 
-\begin{rem}In view of the previous remark, one might hope to be able to simplify the
-condition in Proposition~\ref{prop:characterization-open-subfunctor} as
-follows: ``For any~$x\?X$, the quotient~$\affl/K$ modulo the ideal~$K = \{
-f\?\affl \,|\, \speak{$f$ \inv} \Rightarrow (x \in U) \}$ is quasicoherent.''
-However, this doesn't work out.
-This statement implies the condition in the proposition, but the converse
-direction does not hold, since~$\affl/K \cong \affl/\sqrt{J}$ might fail to be
-quasicoherent. For instance that's the case if~$U = \emptyset$; then~$K =
-\sqrt{(0)}$ is not quasicoherent by Remark~\ref{rem:radical-not-qcoh}.\end{rem}
+\begin{proof}Die Menge der Tangentialvektoren an~$U$ kann kanonisch mit den
+Abbildungen~$\gamma : \Delta \to \Gr(V,r)$ mit~$\gamma(0) = U$ identifiziert
+werden. Dabei ist~$\Delta \defeq \{ \varepsilon \in k \,|\, \varepsilon^2 = 0 \}$.
+Eine solche Abbildung liftet stets zu einer Abbildung von~$\Delta$ in die Menge
+der linear unabhängigen Familien der Länge~$r$ in~$V$. Der Rest sei als
+Übungsaufgabe überlassen. Willkommen in der wunderbaren Welt synthetischer
+Geometrie.
+\end{proof}
 
-\begin{rem}In the internal language of toposes used to carry out synthetic
-differential geometry, there is the concept of an \emph{Penon-open}
-subset~\cite[Chapitre~III]{penon}: A subset~$U \subseteq X$ is Penon-open if
-and only if
-\[ \forall x \in U\_ \forall y\?X\_
-  (x \neq y) \vee (y \in U). \]
-This notion is not useful in synthetic algebraic geometry, since it is much too
-weak: Any subset of the one-element set~$1$ is Penon-open. However, not every
-subfunctor of the terminal functor in~$\Zar(S)$ is an open subfunctor.
+\enlargethispage{2em}
+
+\begin{rem}Wiederholt man genau dieselben Argumente in einem anderen Topos --
+einem, der für Differentialgeometrie angepasst ist -- erhält man mehr oder
+weniger die Darstellbarkeit der Grassmannschen als Mannigfaltigkeit. Das
+einzige, was fehlt, ist ein Nachweis der Glattheit.
 \end{rem}
 
-% negneg-stability
-% definition of open truth values
-
 \end{document}

### 2017-03-19 snippets/properties-of-a1.tex
A further property of A^1

@@
 
+The ring object~$\affl$ in the big Zariski topos enjoys several special
+properties, some of which are unique in that they're only possible in an
+intuitionistic context. The following proposition compiles a short list of such
+properties. At least one of these, the field property, was already noticed in
+the 1970s by Anders Kock~\cite{kock:univ-proj-geometry}.
+
 \begin{prop}The following properties of~$\affl$, as seen from the internal
@@
 % Adjust paragraph below the proof if numbering changes.
-% Proof of third and seventh property references first property by number.
+% Proof of third and eigth property references first property by number.
 \begin{enumerate}
@@
 \item Any element of $\affl$ which is not invertible is nilpotent.
+\item An element of~$\affl$ is nilpotent if and only if it is \notnot zero.
 \item Any function~$\affl \to \affl$ is given by a unique polynomial
@@
 
+\item Follows from the first and third property and the observation that
+nilpotent elements are not invertible (else~$1 = 0$ in contradiction to~$\affl$
+being a local ring).
+
 \item Immediate by considering the quasicoherence condition for the finitely

### 2017-03-19 snippets/grassmannian.tex
More on the Grassmannian

@@
 
+\begin{cor}The Grassmannian~$\Gr(V,r)$ is a finitely presented scheme.\end{cor}
+
+\begin{proof}We need to verify that~$\Gr(V,r)$ admits a finite covering by
+spectra of finitely presented~$\affl$-algebras. We already know that~$\Gr(V,r)$
+can be covered by the open subsets~$G_W$ and that these sets are spectra of
+finitely presented algebras. Therefore it remains to prove that finitely many of
+these subsets suffice to cover~$\Gr(V,r)$.
+
+In fact, if we choose an isomorphism~$V \cong (\affl)^n$, we see
+that~$\binom{n}{r}$ of these subsets suffice: namely those where~$W$ is one of
+the standard submodules of~$(\affl)^n$ (generated by the standard basis
+vectors). For if~$U \in \Gr((\affl)^n,r)$, the surjection~$V \to V/U$ maps
+the basis of at least one of these standard submodules to a basis and is
+therefore bijective. This is because from any surjective~$(r \times n)$-matrix
+over a local ring one can select~$r$ columns which form an linearly independent
+family.
+\end{proof}
+
 \end{document}
-
-\begin{cor}Die Grassmannsche~$\Gr(V,r)$ ist ein Schema von endlichem
-Typ.\end{cor}
-
-\begin{proof}Die Grassmannsche~$\Gr(V,r)$ besitzt eine offene
-Überdeckung durch die affinen Schemata~$G_W$ und ist daher ein Schema.
-
-Wenn wir einen Isomorphismus~$V \cong k^n$ wählen, sehen wir, dass
-schon~$\binom{n}{r}$ viele dieser affinen Schemata genügen: nämlich diese,
-wo~$W$ einer der Standarduntermoduln von~$k^n$ ist (erzeugt durch
-Einheitsvektoren).
-
-Denn ist~$U \in \Gr(k^n,r)$, so bildet die Surjektion~$V \to V/U$ die Basis von
-mindestens einem dieser Standarduntermoduln auf eine Basis ab und ist daher
-bijektiv. (Aus einer surjektiven~$(r \times n)$-Matrix über einem lokalen Ring
-kann man stets~$r$ Spalten auswählen, die eine linear unabhängige Familie
-bilden.)
-\end{proof}
 
@@
 \end{rem}
-
-\end{document}

### 2017-03-19 snippets/injective.tex
Injective and flabby sheaves

\documentclass[10pt,reqno]{amsart}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{etex}
\usepackage{amsmath,amsthm,amssymb,stmaryrd,color,graphicx,multirow}
\usepackage{mathtools}
\usepackage{setspace}
\usepackage{bussproofs}
\usepackage{xspace}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{array}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage[bookmarksdepth=2,pdfencoding=auto]{hyperref}
\usepackage[all]{xy}
\usepackage{soul}\setul{0.3ex}{}

\usepackage{tikz}
\usetikzlibrary{calc,shapes.callouts,shapes.arrows}
\newcommand{\hcancel}[5]{%
    \tikz[baseline=(tocancel.base)]{
        \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
        \draw[red, line width=0.3mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
    }%
}

\usepackage[natbib=true,style=numeric]{biblatex}
\usepackage[babel]{csquotes}
\bibliography{bibliography}

\theoremstyle{definition}
\newtheorem{defn}{Definition}[section]
\newtheorem{ex}[defn]{Example}

\theoremstyle{plain}

\newtheorem{prop}[defn]{Proposition}
\newtheorem{cor}[defn]{Corollary}
\newtheorem{lemma}[defn]{Lemma}
\newtheorem{thm}[defn]{Theorem}
\newtheorem{scholium}[defn]{Scholium}

\theoremstyle{remark}
\newtheorem{rem}[defn]{Remark}
\newtheorem{question}[defn]{Question}

\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\FF}{\mathbb{F}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\A}{\mathcal{A}}
\renewcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\renewcommand{\H}{\mathcal{H}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\L}{\mathcal{L}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\GG}{\mathbb{G}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\qqq}{\mathfrak{q}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\nnn}{\mathfrak{n}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\HOM}{\mathcal{H}\mathrm{om}}
\newcommand{\id}{\mathrm{id}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\placeholder}{\underline{\quad}}
\let\oldul\ul
\renewcommand{\ul}[1]{\text{\oldul{$#1$}}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Grp}{\mathrm{Grp}}
\newcommand{\Vect}{\mathrm{Vect}}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\PSh}{\mathrm{PSh}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\Et}{\mathrm{\acute{E}t}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Mod}{\mathrm{Mod}}
\newcommand{\Alg}{\mathrm{Alg}}
\newcommand{\Ring}{\mathrm{Ring}}
\newcommand{\LRL}{\mathrm{LRL}}
\newcommand{\pt}{\mathrm{pt}}
\newcommand{\tors}{\mathrm{tors}}
\DeclareMathOperator{\Spec}{Spec}
\newcommand{\RelSpek}{\operatorname{\ul{\mathrm{Spec}}}}
\newcommand{\QcohSpec}[2]{\mathrm{Spec}^{\mathrm{qcoh}}_{#1}{#2}}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\colim}{colim}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\Ann}{Ann}
\DeclareMathOperator{\Int}{int}
\DeclareMathOperator{\Clos}{cl}
\DeclareMathOperator{\Kernel}{ker}
\DeclareMathOperator{\supp}{supp}
\newcommand{\Ass}{\mathrm{Ass}}
\newcommand{\Open}{\T}
\newcommand{\?}{\,{:}\,}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\Ll}{\vcentcolon\Longleftrightarrow}
\newcommand{\notat}[1]{{!#1}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\lhra}{\ensuremath{\lhook\joinrel\relbar\joinrel\rightarrow}}
\newcommand{\hra}{\hookrightarrow}
\newcommand{\brak}[1]{{\llbracket{#1}\rrbracket}}
\newcommand{\sdense}{{\widehat\Box}}
\newcommand{\sdenseother}{\Box}
\newcommand{\ie}{i.\,e.\@\xspace}
\newcommand{\eg}{e.\,g.\@\xspace}
\newcommand{\vs}{vs.\@\xspace}
\newcommand{\resp}{resp.\@\xspace}
\newcommand{\inv}{inv.\@}
\newcommand{\notnot}{\emph{not not}\xspace}
\newcommand{\affl}{\ensuremath{{\ul{\AA}^1_S}}\xspace}
\newcommand{\afflx}{\ensuremath{{\ul{\AA}^1_X}}\xspace}
\newcommand{\affla}{\ensuremath{{\ul{\AA}^1_{\Spec A}}}\xspace}
\newcommand{\xra}{\xrightarrow}

\newcommand{\XXX}[1]{\textbf{XXX: #1}}

\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defequiv}{\vcentcolon\equiv}
\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}

\newcommand{\stacksproject}[1]{\cite[{\href{http://stacks.math.columbia.edu/tag/#1}{Tag~#1}}]{stacks-project}}

\definecolor{gray}{rgb}{0.7,0.7,0.7}

\begin{document}

\subsection{Flabby sheaves}

Recall that a sheaf~$\F$ of sets on a topological space (or locale)~$X$ is
\emph{flabby} if and only if, for any open subset~$U \subseteq X$ the
restriction map~$\F(X) \to \F(U)$ is surjective.

Flabbiness of a sheaf is a local property, even though it doesn't seem like that
at first sight: If the restrictions~$\F|_{U_i}$ of~$\F$ to the members of an
open covering~$X = \bigcup_i U_i$ are flabby, then the verification that~$\F$ is
flabby can't proceed as follows. ``Let~$s \in \F(U)$ be an arbitrary section.
Since each~$\F|_{U_i}$ is flabby, the section~$s|_{U \cap U_i}$ extends to a
section on~$U_i$.'' The reason is that the individual extensions obtained in
this way might not glue.

A correct proof employs Zorn's lemma in a typical way, considering a maximal
extension and then verifying that the subset this maximal extension is defined
on is all of~$X$.

Since flabbiness is a local property, it's not unreasonable to expect that
flabbiness could be characterized in the internal language. The following
proposition shows that this is indeed the case.

\begin{prop}\label{prop:internal-char-flabbiness}
Let~$\F$ be a sheaf of sets on a topological space (or locale~$X$).
Then the following statements are equivalent:
\begin{enumerate}
\item $\F$ is flabby.

\item For any open subset $U \subseteq X$ and any section $s \in \F(U)$ there is
an open covering $X = \bigcup_i V_i$ such that, for each $i$, there is an
extension of $s$ to $U \cup V_i$ (that is, a section $s' \in \F(U \cup V_i)$
such that $s'|_U = s$).

(If $X$ is a space instead of a locale, this can be equivalently formulated as
follows: For any open subset $U \subseteq X$, any section $s \in \F(U)$, and any
point $x \in X$, there is an open neighbourhood $V$ of $x$ and an extension of
$s$ to $U \cup V$.)

% Proof of "internally injective ==> externally injective" references the
% following condition by number.
\item From the point of view of the internal language of~$\Sh(X)$, for any
subsingleton $K \subseteq \F$ there exists an element $s \? \F$ such that~$s \in
K$ if~$K$ is inhabited. More precisely,
\begin{multline*}
  \qquad\qquad\Sh(X) \models
  \forall K \subseteq \F\_
  (\forall s,s'\?K\_ s = s') \Longrightarrow \\
  \exists s\?\F\_ (\text{$K$ is inhabited} \Rightarrow s \in K).\qquad\qquad
\end{multline*}

\item The canonical map $\F \to \P_{\leq 1}(\F), s \mapsto \{s\}$ is
final from the internal point of view, that is
\[ \Sh(X) \models
  \forall K \? \P_{\leq 1}(\F)\_
  \exists s \? \F\_
  K \subseteq \{s\}, \]
where $\P_{\leq 1}(\F)$ is the object of subsingletons of $\F$.
\end{enumerate}
\end{prop}

\begin{proof}
The implication~``1~$\Rightarrow$~2'' is trivial. The converse direction uses a
typical argument with Zorn's lemma, considering a maximal extension. The
equivalence~``$2~\Leftrightarrow~3$'' is routine, using the Kripke--Joyal
semantics to interpret the internal statement. Condition~4 is a straightforward
reformulation of condition~3.
\end{proof}

Condition~2 of the proposition is, unlike the standard definition of flabbiness,
manifestly local. Also its equivalence with Condition~3 and Condition~4 is
intuitionistically valid; therefore one might consider to adopt Condition~2 as
the definition of flabbiness.

\begin{rem}The object~$\P_{\leq 1}(\F)$ of subsingletons of $\F$ can be
interpreted as the object of \emph{partially-defined
elements} of $\F$. In this view, the empty subset is the maximally undefined
element and a singleton is a maximally defined element. The proposition shows
that~$\F$ is flabby if and only if any such partially-defined element can be
refined to an honest element of~$\F$.\end{rem}

\subsection{Injective sheaves}

Recall that an object~$I$ of a category~$\C$ is \emph{injective} if and only if,
for any monomorphism~$X \to Y$ in~$\C$ and any morphism~$I \to X$, there
is a lifting such that the diagram
\[ \xymatrix{
  X \ar@{^{(}->}[r]\ar[d] & Y \ar@{-->}[dl] \\
  I
} \]
commutes. Equivalently, an object~$I$ is injective if and only if the Hom
functor~$\Hom_\C(\placeholder, I) : \C^\op \to \Set$ maps monomorphisms in~$\C$ to
surjective maps. This general definition is often specialized to one these cases:
to the category of modules over a ring, to the category of set-valued sheaves
on a topological space, and to the category of sheaves of~$\O_X$-modules on a
ringed space~$(X,\O_X)$.

The definition is seldomly applied in the category of sets, since in a classical
context it's easy to show that a set is injective if and only if it's
inhabited, thereby completely settling the question of which objects are
injective in a trivial manner.

The question is more interesting in an intuitionistic setting, since
intuitionistically one cannot prove that inhabited sets are
injective~\cite{aczel-et-al:injective}; but one can still verify that any set embeds
into an injective set. (The powerset~$\P(X)$ and even the smaller
set~$\P_{\leq1}(X)$ of subsingletons of a given set~$X$ are injective. We verify
this in the proof of Lemma~\ref{lemma:enough-flabby}.)

For a cartesian closed category~$\C$, there is also the notion of an
\emph{internally injective} object. This is an object~$I$ such that the
internal Hom functor~$[\placeholder, I] : \C^\op \to \C$ maps monomorphisms
in~$\C$ to epimorphisms. In the specical case that~$\C$ is a elementary topos
with a natural numbers object, such as the topos of set-valued sheaves on a
space, this condition can be rephrased in several ways. The following
proposition lists five of these conditions. The equivalence of the first four is
due to Harting~\cite{harting}.

\begin{prop}\label{prop:notions-of-internal-injectivity}
Let~$\E$ be an elementary topos with a natural numbers object. Then
the following statements about an object~$I \in \E$ are equivalent.
\begin{enumerate}
\item $I$ is internally injective.
\item The functor~$[\placeholder, I] : \E^\op \to \E$ maps monomorphisms in $\E$
to morphisms for which any global element of the target locally (after change of
base along an epimorphism) possesses a preimage.
\item For any morphism $p : A \to 1$ in $\E$, the object $p^*I$ has property~(1)
as an object of $\E/A$.
\item For any morphism $p : A \to 1$ in $\E$, the object $p^*I$ has property~(2)
as an object of $\E/A$.
\item From the point of view of the internal language of~$\E$, the object~$I$
is injective.\footnote{We have only introduced the internal language for sheaf
toposes. The general definition is in~\cite[Section~XXX]{shulman:stack}.}
\end{enumerate}
\end{prop}

\begin{proof}
The implications ``(1)~$\Rightarrow$~(2)'', ``3 $\Rightarrow$~(4)'', ``(3)
$\Rightarrow$~(1)'', and ``(4)~$\Rightarrow$~(2)'' are trivial.

The equivalence ``(3)~$\Leftrightarrow$~(5)'' follows directly from the
interpretation rules of the stack semantics.

The implication ``(2)~$\Rightarrow$~(4)'' employs the
extra left adjoint $p_! : \E/A \to \E$ of $p^* : \E
\to \E/A$~(which maps an object~$(X \to A)$ to~$X$), as in the usual proof that
injective sheaves remain injective when
restricted to smaller open subsets: We have that $p_* \circ [\placeholder, p^*I]_{\E/A}
\cong [\placeholder, I]_\E \circ p_!$, the functor $p_!$ preserves monomorphisms, and one
can check that $p_*$ reflects the property that global elements locally possess
preimages. Details are in~\cite[Thm.~(1)1]{harting}.

The implication ``(4)~$\Rightarrow$~(3)'' follows by performing an extra change of
base, since any non-global element becomes a global element after a suitable
change of base.
\end{proof}

Somewhat surprisingly, and in stark contrast with the situation for internally
projective objects (which are defined dually), internal injectivity coincides
with external injecticity for sheaf toposes over spaces.

\begin{thm}\label{thm:char-injectivity}
Let~$X$ be a topological space (or a locale). An object~$\I \in \Sh(X)$ is
injective if and only if it is internally injective.
\end{thm}

\begin{proof}For the ``only if'' direction, let~$\I$ be an injective sheaf of
sets. Then~$\I$ satisfies Condition~(2) in
Proposition~\ref{prop:notions-of-internal-injectivity}, even without having to
pass to covers.

For the ``if'' direction, let~$\I$ be an internally injective object. Let~$m :
\E \to \F$ be a monomorphism in~$\Sh(X)$ and let~$k : \E \to \I$ be an arbitrary
morphism. We want to show that there exists an extension $\F \to \I$ of~$k$
along~$m$. To this end, we consider the sheaf defined by the internal expression
\[ \G \defeq \brak{\{ k' \? [\F,\I] | k' \circ m = k \}}. \]
Global sections of~$\G$ are extensions of the kind we're looking for.
Therefore it suffices to show that~$\G$ is flabby. We do this by verifying
Condition~(3) of Proposition~\ref{prop:internal-char-flabbiness} in the internal
language of~$\Sh(X)$.

Let~$K \subseteq \G$ be a subsingleton. We consider the injectivity diagram
\[ \xymatrix{
  m[\E] \cup \F' \ar[r]\ar[d] & \F \ar@{-->}[ld] \\
  \I,
} \]
where~$\F' \defeq \{ s \? \F \,|\, \text{$K$ is inhabited} \}$ and the solid
vertical arrow is defined in the following way: It should map an element~$s \in
\F'$ to~$k'(s)$, where~$k'$ is any element of~$K$; and it should map an
element~$m[u] \in m[\E]$ to~$k(u)$. These prescriptions determine a well-defined
map.

Since~$\I$ is injective from the internal point of view we're taking up here,
there exists a dotted map rendering the diagram commutative. This map is an
element of~$\G$. Furthermore, this map is an element of~$K$, if~$K$ is
inhabited.
\end{proof}

\begin{thm}\label{thm:char-injectivity-modules}
Let~$(X,\O_X)$ be a ringed topological space (or a ringed locale).
An~$\O_X$-module~$\I$ is injective if and only if it is internally injective.
\end{thm}

\begin{proof}Proposition~\ref{prop:notions-of-internal-injectivity} can be
adapted from sheaves to sets to sheaves of modules, with the same proof.
% XXX: check that.

The proof of Theorem~\ref{thm:char-injectivity} can be adopted as well.
It suffices to change~``$m[\E] \cup \F'$'' to~``$m[\E] + \F''$'', where~$\F''
\defeq \{ s \? \F \,|\, \text{$s = 0$ or $K$ is inhabited} \}$.
\end{proof}

\subsection{Internal proofs of common lemmas}

\begin{lemma}A sheaf of set or a sheaf of modules is injective if and only if it
is locally injective.\end{lemma}

\begin{proof}By Theorem~\ref{thm:char-injectivity} respectively
Theorem~\ref{thm:char-injectivity-modules}, Injectivity can be characterized in the
internal language. Any such property is local.\end{proof}

\begin{lemma}Let~$X$ be a topological space (or a locale).
\begin{enumerate}
\item Let~$\I$ be an injective sheaf of sets over~$X$. Let~$\F$ be an arbitrary
sheaf of sets. Then~$\HOM(\F,\I)$ is flabby.
\item Let~$\I$ be an injective sheaf of modules over some sheaf~$\O_X$ of rings
over~$X$. Let~$\F$ be an arbitrary sheaf of modules.
Then~$\HOM_{\O_X}(\F,\I)$ is flabby.
\end{enumerate}
\end{lemma}

\begin{proof}
We cover the case of sheaf of sets first. By Theorem~\ref{thm:char-injectivity}
and Proposition~\ref{prop:internal-char-flabbiness}, it suffices to give an
intuitionistic proof of the following statement: If~$I$ is an injective set
and~$F$ is an arbitrary set, then partially defined elements of the set~$[F,I]$
of all maps~$F \to I$ can be refined to honest elements.

Thus let a subsingleton~$K \subseteq [F,I]$ be given. We consider the
injectivity diagram
\[ \xymatrix{
  F' \ar[r]\ar[d] & F \ar@{-->}[ld] \\
  I,
} \]
where~$F' \defeq \{ s \? F \,|\, \text{$K$ is inhabited} \} \subseteq F$ and the
solid vertical map sends~$s \in F'$ to~$f(s)$, where~$f$ is an arbitrary element
of~$K$. This association is well-defined. Since~$I$ is injective, a dotted lift
as indicated exists. If~$K$ is inhabited, this lift is an element of~$K$.

The same kind of proof works for the case of sheaves of modules, relying on
Theorem~\ref{thm:char-injectivity-modules} and substituting the definition
of~$F'$ by~``$F' \defeq \{ s \? F \,|\, \text{$s = 0$ or $K$ is inhabited}
\}$''.
\end{proof}

\begin{cor}Injective sheaves of sets and injective sheaves of modules are
flabby.\end{cor}

\begin{proof}Follows from the previous lemma by considering the special cases~$\F
\defeq 1$ respectively~$\F \defeq \O_X$.\end{proof}

\begin{lemma}\label{lemma:enough-flabby}
Let~$X$ be a topological space (or a locale). Any sheaf of sets
over~$X$ can be embedded into a injective (therefore flabby) sheaf of sets.
\end{lemma}

\begin{proof}By Proposition~\ref{prop:internal-char-flabbiness}, it suffices to
give an intuitionistic proof of the following statement: Any set~$F$ can be
embedded into an injective set.

As already indicated there at least two simple ways that~$F$ can be embedded
into an injective set: by embedding~$F$ in its powerset~$\P(F)$ or by
embedding~$F$ in~$\P_{\leq1}(F)$, the set of subsingletons of~$F$. For
conciseness, we only verify that~$\P_{\leq1}(F)$ is injective.

So let~$m : A \hookrightarrow B$ be an injective map and let~$k : A \to
\P_{\leq1}(F)$ be an arbitrary map. Then we can extend~$k$ to a map~$k' : B \to
\P_{\leq1}(F)$ by defining for~$y \? B$
\begin{align*}
  k'(y) &\defeq \bigcup k[m^{-1}[\{y\}]] \\
  &\phantom{\vcentcolon}= \{ s\?F \,|\, \text{$s \in k(x)$ for some~$x \in A$ such that~$m(x) = y$} \}.
  \qedhere
\end{align*}
\end{proof}

\begin{rem}The \emph{Godement construction} is a well-known way of embedding an
inhabited sheaf of sets~$\F$ into an injective sheaf, namely embedding it into
the sheaf of not necessarily continuous sections of the étale space of~$\F$:
\[ U \subseteq X \quad\longmapsto\quad
  \prod_{x \in U} \F_x. \]
The sheaf~$\P_{\leq1}(\F)$ does not coincide with this construction.
Instead, it is the sheaf with
\[ U \subseteq X \quad\longmapsto\quad
  \{ \langle V, s \rangle \,|\,
    \text{$V \subseteq U$ open, $s \in \F(V)$} \}. \]
It's not possible to describe the Godement construction in the internal language
of~$\Sh(X)$, since in the Godement construction the underlying set of~$X$
enters. But the sheaf topos of~$X$ doesn't remember this set. For instance, let~$X$
be an inhabited indiscrete topological space. Then~$\Sh(X)$ is equivalent
to~$\Set$.
\end{rem}

\begin{rem}It's not known to me whether it's possible to intuitionistically
prove that any module can be embedded into a module which satisfies the internal
flabbiness criterion of Proposition~\ref{prop:internal-char-flabbiness}. This
would imply the corollary that any sheaf of modules can be embedded into a
flabby sheaf of modules. The naive candidates don't work: The
set~$\P_{\leq1}(F)$ doesn't admit a canonical module structure, and the free
module over that set is not flabby in general.

It's certainly not possible to intuitionistically prove that any module can be
embedded into an injective module, since it's consistent with Zermelo--Fraenkel
set theory that no nontrivial injective abelian groups
exist~\cite{blass:inj-proj-axc}.
\end{rem}

\end{document}

### 2017-03-20 snippets/injective.tex
typo

@@
 semantics to interpret the internal statement. Condition~4 is a straightforward
-reformulation of condition~3.
+reformulation of Condition~3.
 \end{proof}
@@
 
-\begin{rem}The object~$\P_{\leq 1}(\F)$ of subsingletons of $\F$ can be
+The object~$\P_{\leq 1}(\F)$ of subsingletons of $\F$ can be
 interpreted as the object of \emph{partially-defined
@@
 that~$\F$ is flabby if and only if any such partially-defined element can be
-refined to an honest element of~$\F$.\end{rem}
+refined to an honest element of~$\F$.
 
@@
 \[ \xymatrix{
-  m[\E] \cup \F' \ar[r]\ar[d] & \F \ar@{-->}[ld] \\
+  m[\E] \cup \F' \ar@{^{(}->}[r]\ar[d] & \F \ar@{-->}[ld] \\
   \I,
@@
 
+% XXX: Remark that proof requires AxC and that it's a unique feature of sheaf toposes.
+% Counterexample?
+
 \subsection{Internal proofs of common lemmas}
 
-\begin{lemma}A sheaf of set or a sheaf of modules is injective if and only if it
+\begin{lemma}A sheaf of sets or a sheaf of modules is injective if and only if it
 is locally injective.\end{lemma}
@@
 of~$\Sh(X)$, since in the Godement construction the underlying set of~$X$
-enters. But the sheaf topos of~$X$ doesn't remember this set. For instance, let~$X$
-be an inhabited indiscrete topological space. Then~$\Sh(X)$ is equivalent
+enters. But the sheaf topos of~$X$ doesn't remember this set. For instance, if~$X$
+is an inhabited indiscrete topological spac, then~$\Sh(X)$ is equivalent
 to~$\Set$.

### 2017-03-20 snippets/injective.tex
typo

@@
 Since flabbiness is a local property, it's not unreasonable to expect that
-flabbiness could be characterized in the internal language. The following
+flabbiness can be characterized in the internal language. The following
 proposition shows that this is indeed the case.
@@
 \begin{prop}\label{prop:internal-char-flabbiness}
-Let~$\F$ be a sheaf of sets on a topological space (or locale~$X$).
+Let~$\F$ be a sheaf of sets on a topological space~$X$ (or a locale).
 Then the following statements are equivalent:
@@
 
-\item For any open subset $U \subseteq X$ and any section $s \in \F(U)$ there is
+\item ``Any section of~$\F$ can be locally extended'':
+For any open subset $U \subseteq X$ and any section $s \in \F(U)$ there is
 an open covering $X = \bigcup_i V_i$ such that, for each $i$, there is an
@@
 \begin{proof}
-The implication~``1~$\Rightarrow$~2'' is trivial. The converse direction uses a
+The implication~``(1)~$\Rightarrow$~(2)'' is trivial. The converse direction uses a
 typical argument with Zorn's lemma, considering a maximal extension. The
-equivalence~``$2~\Leftrightarrow~3$'' is routine, using the Kripke--Joyal
+equivalence~``(2)~$\Leftrightarrow$~(3)'' is routine, using the Kripke--Joyal
 semantics to interpret the internal statement. Condition~4 is a straightforward
@@
 Recall that an object~$I$ of a category~$\C$ is \emph{injective} if and only if,
-for any monomorphism~$X \to Y$ in~$\C$ and any morphism~$I \to X$, there
+for any monomorphism~$X \to Y$ in~$\C$ and any morphism~$X \to I$, there
 is a lifting such that the diagram
@@
 functor~$\Hom_\C(\placeholder, I) : \C^\op \to \Set$ maps monomorphisms in~$\C$ to
-surjective maps. This general definition is often specialized to one these cases:
+surjective maps. This general definition is often specialized to one of these cases:
 to the category of modules over a ring, to the category of set-valued sheaves
@@
 context it's easy to show that a set is injective if and only if it's
-inhabited, thereby completely settling the question of which objects are
+inhabited, thereby completely settling the question which objects are
 injective in a trivial manner.
@@
 injective~\cite{aczel-et-al:injective}; but one can still verify that any set embeds
-into an injective set. (The powerset~$\P(X)$ and even the smaller
+into an injective set: The powerset~$\P(X)$ and even the smaller
 set~$\P_{\leq1}(X)$ of subsingletons of a given set~$X$ are injective. We verify
-this in the proof of Lemma~\ref{lemma:enough-flabby}.)
+this in the proof of Lemma~\ref{lemma:enough-flabby}.
 
@@
 \item From the point of view of the internal language of~$\E$, the object~$I$
-is injective.\footnote{We have only introduced the internal language for sheaf
-toposes. The general definition is in~\cite[Section~XXX]{shulman:stack}.}
+is injective.\footnote{In Section~\ref{sect:internal-language}, we have only
+introduced the internal language for sheaf toposes. The general definition is
+in~\cite[Section~7]{shulman:stack}.}
 \end{enumerate}
@@
 \begin{proof}By Theorem~\ref{thm:char-injectivity} respectively
-Theorem~\ref{thm:char-injectivity-modules}, Injectivity can be characterized in the
+Theorem~\ref{thm:char-injectivity-modules}, injectivity can be characterized in the
 internal language. Any such property is local.\end{proof}
@@
   F' \ar[r]\ar[d] & F \ar@{-->}[ld] \\
-  I,
+  I
 } \]
-where~$F' \defeq \{ s \? F \,|\, \text{$K$ is inhabited} \} \subseteq F$ and the
+where~$F'$ is the subset~$\{ s \? F \,|\, \text{$K$ is inhabited} \} \subseteq F$ and the
 solid vertical map sends~$s \in F'$ to~$f(s)$, where~$f$ is an arbitrary element
@@
 
-The same kind of proof works for the case of sheaves of modules, relying on
-Theorem~\ref{thm:char-injectivity-modules} and substituting the definition
-of~$F'$ by~``$F' \defeq \{ s \? F \,|\, \text{$s = 0$ or $K$ is inhabited}
-\}$''.
+The same kind of argument applies to the case of sheaves of modules, relying on
+Theorem~\ref{thm:char-injectivity-modules} and defining~$F'$ as the
+submodule~``$\{ s \? F \,|\, \text{$s = 0$ or $K$ is inhabited} \}$''.
 \end{proof}
@@
 Let~$X$ be a topological space (or a locale). Any sheaf of sets
-over~$X$ can be embedded into a injective (therefore flabby) sheaf of sets.
+over~$X$ can be embedded into an injective (therefore flabby) sheaf of sets.
 \end{lemma}
@@
 The sheaf~$\P_{\leq1}(\F)$ does not coincide with this construction.
-Instead, it is the sheaf with
+Instead by Definition~\ref{defn:interpretation-internal-constructions}, it is the sheaf with
 \[ U \subseteq X \quad\longmapsto\quad

### 2017-03-20 snippets/injective.tex
Remark on expectations of provability

@@
 flabbiness criterion of Proposition~\ref{prop:internal-char-flabbiness}. This
-would imply the corollary that any sheaf of modules can be embedded into a
+would give an internal proof that any sheaf of modules can be embedded into a
 flabby sheaf of modules. The naive candidates don't work: The
@@
 
-It's certainly not possible to intuitionistically prove that any module can be
-embedded into an injective module, since it's consistent with Zermelo--Fraenkel
-set theory that no nontrivial injective abelian groups
+Since by the Godement construction the statement is true in many models of
+intuitionistic logic, the sheaf toposes over topological spaces, and
+furthermore the proof that the Godement construction yields a flabby sheaf is
+intuitionistically valid,\footnote{In order for the Godement construction to
+work in a intuitionistic metatheory, one has to tweak its definition a little
+bit.  Instead of mapping an open subset~$U$ to~$\prod_{x \in U} \F_x$, one has
+to map~$U$ to~$\prod_{x \in U} \P_{\leq1}(\F_x)$. This has the added advantage
+that it works even if~$\F$ is not inhabited.} it's not unreasonable to believe
+that such an intuitionistic proof is possible.
+
+On the other hand, it's certainly not possible to intuitionistically prove that
+any module can be embedded into an injective module, since it's consistent with
+Zermelo--Fraenkel set theory that no nontrivial injective abelian groups
 exist~\cite{blass:inj-proj-axc}.

### 2017-03-20 snippets/injective.tex
Common lemmas about flabby sheaves

@@
 
+\begin{lemma}Let~$X$ be a ringed space (or ringed locale). Let~$0 \to \E'
+\xra{\alpha} \E \xra{\beta} \E'' \to 0$ be a short exact sequence
+of~$\O_X$-modules. If~$\E'$ is flabby, then the induced sequence
+\[ 0 \longrightarrow \Gamma(X,\E')
+  \longrightarrow \Gamma(X,\E)
+  \longrightarrow \Gamma(X,\E'') \longrightarrow 0 \]
+is exact.
+\end{lemma}
+
+\begin{proof}Since taking global sections is left exact (being a right adjoint
+functor), it suffices to verify that the map~$\Gamma(X,\E) \to \Gamma(X,\E'')$
+is surjective. We'll do this by showing, in the internal language of~$\Sh(X)$,
+that the sheaf of preimages of a given global section~$s \in \Gamma(X,\E'')$ is
+flabby and therefore has a global section.
+
+In the internal language, this sheaf has the description~$F \defeq \{ u \? \E
+\,|\, \beta(u) = s \}$. To verify the internal condition of
+Proposition~\ref{prop:internal-char-flabbiness}, let a subsingleton~$K \subseteq
+F$ be given. Since~$\beta$ is surjective, there is a preimage~$u_0 \in F$.
+The translated set~$K - u_0 \subseteq \E$ is still a subsingleton, and its
+preimage under~$\alpha$ is as well. By the assumption on~$\E'$, there is an
+element~$v \? \E$ such that~$v \in \alpha^{-1}[K - u_0]$ if~$K - u_0$ is
+inhabited. We'll now verify that~$u_0 + \alpha(v) \in K$ if~$K$ is inhabited.
+
+So assume that~$K$ is inhabited. Then~$K - u_0$ is as well. Since the image of
+its unique element under~$\beta$ is zero and the given sequence is exact, the
+set~$\alpha^{-1}[K - u_0]$ is inhabited as well. Therefore~$v \in \alpha^{-1}[K
+- u_0]$. Thus~$u_0 + \alpha(v) \in K$.
+\end{proof}
+
+\begin{lemma}Let~$X$ be a ringed space (or ringed locale). Let~$0 \to \E'
+\xra{\alpha} \E \xra{\beta} \E'' \to 0$ be a short exact sequence
+of~$\O_X$-modules. If~$\E'$ and~$\E''$ are flabby, then~$\E$ is flabby as well.
+\end{lemma}
+
+\begin{proof}We verify the condition of
+Proposition~\ref{prop:internal-char-flabbiness} in the internal language
+of~$\Sh(X)$.
+
+Let~$K \subseteq \E$ be a subsingleton. Then its image~$\beta[K] \subseteq \E''$
+is a subsingleton as well. Since partial elements of~$\E''$ can be refined to
+honest elements, there is an element~$s \? \E''$ such that~$\beta[K] \subseteq
+\{ s \}$.
+
+Since~$\beta$ is surjective, there is an element~$t_0 \? \E$ such
+that~$\beta(t_0) = s$.
+
+The preimage~$\alpha^{-1}[K - t_0] \subseteq \E'$ is a subsingleton. This
+partial element can be refined to an honest element, that is there exists an
+element~$u \? \E'$ such that~$\alpha^{-1}[K - t_0] \subseteq \{u\}$.
+
+The partial element~$K$ can thereby refined to the honest element~$t \defeq t_0 + \alpha(u)$.
+\end{proof}
+
 \end{document}

### 2017-03-20 snippets/qcoh.tex
Internal construction of the relative spectrum
Integrated the section on the relative spectrum construction in the big Zariski topos

@@
 
-By Proposition~\ref{XXX}, the internal expression~$\Spec(A)$ is the functor of
+By Proposition~\ref{prop:relative-spectrum-big-zariski}, the internal expression~$\Spec(A)$ is the functor of
 points of~$\RelSpek_S \A_0$. For any~$S$-scheme~$f : T \to S$ contained in the site
@@
 
-\begin{rem}The condition in Scholium~\ref{scholium} that~$\Spec(A)$ is
+\begin{rem}\label{rem:local-representability}
+The condition in Scholium~\ref{scholium} that~$\Spec(A)$ is
 representable by an object of the site used to define~$\Zar(S)$ is slightly

### 2017-07-21 slides-duesseldorf2017.tex
Slides of today's talk

--- 2017-02-07 slides-warwick2017.tex
@@
 \author{Ingo Blechschmidt}
-\date{February 7th, 2017}
+\date{July 21st, 2017}
 
@@
 
-  Algebraic Geometry Seminar \\
-  University of Warwick \\
+  Oberseminar Algebra/Geometrie/Topologie \\
+  Heinrich-Heine-Universität Düsseldorf \\
   \ \\
-  Feburary 7th, 2017
-  \par
-\end{frame}
-
-\frame[t]{\frametitle{Outline}\small
-  \begin{columns}[t]
-    \begin{column}{0.6\textwidth}
-      \vspace*{-1.5em}
-      \begin{itemize}\item[]\tableofcontents\end{itemize}
-      \par
-    \end{column}
-    \begin{column}{0.4\textwidth}
+  July 21st, 2017
+  \par
+\end{frame}
+
+\backupstart
+{\usebackgroundtemplate{\includegraphics[width=\paperwidth]{images/unterbacher-see}}
+\begin{frame}[plain]
+  \vspace{\textheight}\vspace{-0em}
   \centering
-      \begin{minipage}{0.70\textwidth}
-        \begin{exampleblock}{}
-          \justifying
-          Any finitely generated vector space does \emph{not not} possess a basis.
-        \end{exampleblock}
-      \end{minipage}
-      \medskip
-
-      \scalebox{3}{$\Downarrow$}
-
-      \begin{minipage}{0.70\textwidth}
-        \begin{exampleblock}{}
-          \justifying
-          Any sheaf of modules of finite type on a reduced scheme is locally free
-          \emph{on a dense open subset}.
-        \end{exampleblock}
-        \centering
-        \tiny Ravi Vakil: ``Important hard exercise'' (13.7.K).
-        \par
-      \end{minipage}
-    \end{column}
-  \end{columns}
-}
-
-\note{\justifying\fontsize{8pt}{9.6}\selectfont
-  \begin{center}\large\textbf{Abstract}\end{center}
-
-  \begin{changemargin}{2.5em}{2.5em}
-    (new abstract)
-  \end{changemargin}
-}
+  \textcolor{white}{\emph{Camping at Unterbacher See}} \\
+  \par
+\end{frame}}
+
+% Orally: Laws of reasoning do not only apply to ordinary mathematical objects.
+% They also apply to mathematical objects of different toposes. By cleverly
+% choosing those topos, we can pretend that ...
+\frame[t]{\frametitle{Quick summary}
+  By employing the internal language of toposes in various ways, you can
+  pretend that:
+  \begin{enumerate}
+    \item Reduced rings are Noetherian and in fact fields.
+    \item Sheaves of modules are plain modules.
+    \item Schemes are sets: \[ \PP^2_S = \{
+    [x_0\mathbin{:}x_1\mathbin{:}x_2] \,|\, x_0 \neq 0 \vee x_1 \neq 0 \vee x_2 \neq
+    0 \}. \]
+  \end{enumerate}
+}
+% Any theorem automatically yields infinitely new theorems; sometimes, these
+% are useful, and sometimes, they are indeed very nice.
+
+%\frame[t]{\frametitle{Outline}\small
+%  \begin{columns}[t]
+%    \begin{column}{0.6\textwidth}
+%      \vspace*{-1.5em}
+%      \begin{itemize}\item[]\tableofcontents\end{itemize}
+%      \par
+%    \end{column}
+%    \begin{column}{0.4\textwidth}
+%      \centering
+%      \begin{minipage}{0.70\textwidth}
+%        \begin{exampleblock}{}
+%          \justifying
+%          Any finitely generated vector space does \emph{not not} possess a basis.
+%        \end{exampleblock}
+%      \end{minipage}
+%      \medskip
+%
+%      \scalebox{3}{$\Downarrow$}
+%
+%      \begin{minipage}{0.70\textwidth}
+%        \begin{exampleblock}{}
+%          \justifying
+%          Any sheaf of modules of finite type on a reduced scheme is locally free
+%          \emph{on a dense open subset}.
+%        \end{exampleblock}
+%        \centering
+%        \tiny Ravi Vakil: ``Important hard exercise'' (13.7.K).
+%        \par
+%      \end{minipage}
+%    \end{column}
+%  \end{columns}
+%}
+%
+%\note{\justifying\fontsize{8pt}{9.6}\selectfont
+%  \begin{center}\large\textbf{Abstract}\end{center}
+%
+%  \begin{changemargin}{2.5em}{2.5em}
+%    (new abstract)
+%  \end{changemargin}
+%}
 
@@
   \begin{itemize}
+    \item Unique features of the internal world
     \item Transfer principles~$M \leftrightarrow M^\sim$
-    \item Understanding generic freeness
     \item The curious role of affine open subsets
@@
 
-\subsection{Transfer principles}
-
-\begin{frame}\frametitle{Transfer principles}
-
-  \hil{Question:} How do the properties of
-  \begin{itemize}
-  \item an~$A$-module~$M$ in~$\Set$ and
-  \item the~$\O_X$-module $M^\sim$ in~$\Sh(X)$, where~$X = \Spec A$, relate?
-  \end{itemize}
-  \pause
-  \vfill
-
-  \hil{Observation:} $M^\sim = \underline{M}[\F^{-1}]$, where
-  \begin{itemize}
-  \item $\underline{M}$ is the constant sheaf with stalks~$M$ on~$X$ and
-  \item $\F \hookrightarrow \underline{A}$ is the \hil{generic prime filter}
-  with stalk~$A \setminus \mathfrak{p}$ at $\mathfrak{p} \in \Spec A$.
-  \end{itemize}
-  Note: $M$ and $\underline{M}$ share all first-order properties.
-  \pause
-  \vfill
-
-  \centering
-  \hil{Answer:} $M^\sim$ inherits those properties of $M$ which are \\
-  \hil{stable under localisation}.
-  \par
-\end{frame}
-
-\note{\justifying
-  The concept of a \emph{prime filter} is a direct axiomatization of what you
-  expect the complement of a prime ideal to fulfil. In classical logic,
-  complementation gives a bijection between the prime filters and the prime
-  ideals of a ring.
-
-  Prime filters are important in constructive mathematics because localising
-  them gives rise to local rings. In contrast, localising a ring at the
-  complement of a prime ideal doesn't usually result in a local ring.
-
-  To construct the universal localisation of~$A$, one doesn't pick a particular
-  prime filter~$F$ to construct~$A[F^{-1}]$. Instead, one picks the
-  \emph{generic prime filter}~$\F$. This filter doesn't live in~$\Set$, but
-  in~$\Sh(\Spec A)$.
-  \par
-}
-
-\subsection[Generic freeness]{Generic freeness}
-
-\begin{frame}\frametitle{\mbox{A curious property of the structure sheaf}}
+\subsection[Unique features]{Unique features of the internal world}
+
+\begin{frame}\frametitle{Unique features of the internal world}
   Let~$X$ be a scheme. Internally to~$\Sh(X)$,
@@
 
+  \only<1>{
     \centering
@@
     \par
-\end{frame}
+  }
+  \pause
+
+  If~$X$ is reduced, this implies:
+  \begin{itemize}
+    \item $\O_X$ is a \hil{field} in that $\neg(\text{$x$ invertible})
+    \Rightarrow x = 0$.
+    \item $\O_X$ has \hil{$\neg\neg$-stable equality}: $\neg\neg(x = 0)
+    \Rightarrow x = 0$.
+    \item $\O_X$ is \hil{anonymously Noetherian}: Any ideal of~$\O_X$ is \mbox{\hil{not
+    not}} finitely generated.
+  \end{itemize}
+\end{frame}
+
+\subsection[Generic freeness]{Grothendieck's generic freeness lemma}
 
@@
   \vspace*{-1.6em}
-  \[ \xymatrixcolsep{3pc}\xymatrixrowsep{3pc}\xymatrix{
+  \[ \xymatrixcolsep{3.5pc}\xymatrixrowsep{2pc}\xymatrix{
     & \ar@{-}[d]^{\substack{\text{finitely}\\\text{generated}}} M \\
@@
   } \]
-  If $R$ is reduced ($x^n = 0 \Rightarrow x = 0$), there
+  \hil{Lemma.} If $R$ is reduced, there
   %is a \hil{dense open subset} $U \subseteq \Spec R$ such that for any $f \in R$ with~$D(f)
@@
   For a trivial proof, employ~$\Sh(\Spec R)$ and exploit that
-  \begin{itemize}
-    \item $\O_{\Spec R}$ is a \hil{field}: $\neg(\text{$x$ invertible}) \Rightarrow x = 0$,
-    \item $\O_{\Spec R}$ and $\O_{\Spec R}[X_1,\ldots,X_n]$ are \hil{Noetherian}
-    in the sense that any ideal is \hil{not not} finitely generated.
-  \end{itemize}
-\end{frame}
+  $\O_{\Spec R}$ is a \hil{field} and \hil{anonymously Noetherian}.
+\end{frame}
+
+\subsection{Transfer principles}
+
+\begin{frame}\frametitle{Transfer principles}
+
+  \hil{Question:} How do the properties of
+  \begin{itemize}
+  \item an~$A$-module~$M$ in~$\Set$ and
+  \item the~$\O_X$-module $M^\sim$ in~$\Sh(X)$, where~$X = \Spec A$, relate?
+  \end{itemize}
+  \pause
+  \vfill
+
+  \hil{Observation:} $M^\sim = \underline{M}[\F^{-1}]$, where
+  \begin{itemize}
+  \item $\underline{M}$ is the constant sheaf with stalks~$M$ on~$X$ and
+  \item $\F \hookrightarrow \underline{A}$ is the \hil{generic prime filter}
+  with stalk~$A \setminus \mathfrak{p}$ at $\mathfrak{p} \in \Spec A$.
+  \end{itemize}
+  Note: $M$ and $\underline{M}$ share all first-order properties.
+  \pause
+  \vfill
+
+  \centering
+  \hil{Answer:} $M^\sim$ inherits those properties of $M$ which are \\
+  \hil{stable under localisation}.
+  \par
+\end{frame}
+
+\note{\justifying
+  The concept of a \emph{prime filter} is a direct axiomatization of what you
+  expect the complement of a prime ideal to fulfil. In classical logic,
+  complementation gives a bijection between the prime filters and the prime
+  ideals of a ring.
+
+  Prime filters are important in constructive mathematics because localising
+  them gives rise to local rings. In contrast, localising a ring at the
+  complement of a prime ideal doesn't usually result in a local ring.
+
+  To construct the universal localisation of~$A$, one doesn't pick a particular
+  prime filter~$F$ to construct~$A[F^{-1}]$. Instead, one picks the
+  \emph{generic prime filter}~$\F$. This filter doesn't live in~$\Set$, but
+  in~$\Sh(\Spec A)$.
+  \par
+}
 
@@
     \item or Grothendieck's functor-of-points account, where a scheme is a functor~$\mathrm{Ring} \to \mathrm{Set}$.
-    \small\[ A \longmapsto \{ (x,y,z) \in A^3 \,|\, x^n+y^n-z^n=0 \} \]
+    \small\[ A \longmapsto \{ (x,y,z) \? A^3 \,|\, x^n+y^n-z^n=0 \} \]
   \end{itemize}
@@
 
-  \hil{Synthetic approach}: model schemes \hil{directly as sets} in a certain
+  \hil{Synthetic approach:} model schemes \hil{directly as sets} in a certain
   nonclassical set theory.
   \small
-  \[ \{ (x,y,z) \in (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]
+  \[ \{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]
 \end{frame}
@@
   \begin{itemize}
-    \item $\O(1) = (\ell^\vee)_{\ell \in \mathbb{P}^n}$
-    \item $\O(-1) = (\ell)_{\ell \in \mathbb{P}^n}$
+    \item $\O(1) = (\ell^\vee)_{\ell : \mathbb{P}^n}$
+    \item $\O(-1) = (\ell)_{\ell : \mathbb{P}^n}$
     \item Euler sequence: $0 \to \ell^\perp \to ((\affl)^{n+1})^\vee \to \ell^\vee \to 0$
@@
   \begin{itemize}
-    \item $\RelSpec \affl[X,Y,Z]/(X^n+Y^n-Z^n) \cong {}$ \\ \qquad\qquad $\{ (x,y,z) \in (\affl)^3 \,|\, x^n+y^n-z^n = 0 \}$ \medskip
+    \item $\RelSpec \affl[X,Y,Z]/(X^n+Y^n-Z^n) \cong {}$ \\ \qquad\qquad $\{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n = 0 \}$ \medskip
     \item $\Delta \defeq \RelSpec \affl[\varepsilon]/(\varepsilon^2) \cong \{ \varepsilon : \affl \,|\, \varepsilon^2 = 0 \}$
@@
 
-  \item An $\affl$-module~$E$ is quasicoherent if and only if
+  \item An $\affl$-module~$E$ is \hil{quasicoherent} if and only if
   \[ E \otimes_\affl R \longrightarrow \Hom(\Spec R, E) \]
@@
   \item A subset $U \subseteq X$ is qc-open if and only if for any $x : X$
-  there exist $f_1,\ldots,f_n \in \affl$ such that
+  there exist $f_1,\ldots,f_n \? \affl$ such that
   $x \in U \Longleftrightarrow \exists i\_ f_i \neq 0$.
@@
 
-  \item If $\gamma : \Delta \to X$ is tangent vector with $\gamma(0) \in U$,
+  \item If $\gamma : \Delta \to X$ is a tangent vector with $\gamma(0) \in U$,
   then~$\gamma(\varepsilon) \in U$ for all~$\varepsilon \in \Delta$.

### 2017-09-15 slides-pssl101.tex
Slides for PSSL 101 (WIP)

--- 2017-07-21 slides-duesseldorf2017.tex
@@
 
-\title{First steps in synthetic algebraic geometry}
+\title{Using the internal language of toposes in commutative algebra}
 \author{Ingo Blechschmidt}
-\date{July 21st, 2017}
+\date{September 17st, 2017}
 
@@
 
-  \hil{First steps in \\ synthetic algebraic geometry}
+  \hil{Using the internal language of toposes \\ in commutative algebra}
   \medskip
@@
 
-  Oberseminar Algebra/Geometrie/Topologie \\
-  Heinrich-Heine-Universität Düsseldorf \\
+  PSSL 101 in Leeds \\
   \ \\
-  July 21st, 2017
-  \par
-\end{frame}
-
-\backupstart
-{\usebackgroundtemplate{\includegraphics[width=\paperwidth]{images/unterbacher-see}}
-\begin{frame}[plain]
-  \vspace{\textheight}\vspace{-0em}
-  \centering
-  \textcolor{white}{\emph{Camping at Unterbacher See}} \\
-  \par
-\end{frame}}
+  September 17st, 2017
+  \par
+\end{frame}
 
@@
   \begin{enumerate}
-    \item Reduced rings are Noetherian and in fact fields.
     \item Sheaves of modules are plain modules.
@@
     0 \}. \]
+    \item Reduced rings are Noetherian and in fact fields.
   \end{enumerate}
@@
 
+\frame[t]{\frametitle{The little Zariski topos}
+  Let~$A$ be a ring.
+  Is there a \hil{free local ring}~$A \to A'$ over~$A$?
+  \[ \xymatrix{
+    A \ar[rd] \ar[rrr] &&& {\substack{\text{local}\\\text{\normalsize$R$}\\\phantom{\text{local}}}} \\
+    & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!34]{\text{local}}[rru]
+  } \]
+
+  \hil{No,} if we restrict to~$\Set$.
+  \hil{Yes,} if we allow a change of topos:
+  Then $A \to \O_{\Spec(A)}$ is the universal localisation.
+
+  \begin{block}{Definition}The \hil{little Zariski topos} of~$A$ is the
+  category~$\Sh(\Spec(A))$ of set-valued sheaves on the \hil{spectrum} of~$A$.\end{block}
+
+  The topology on~$\Spec(A)$ is generated by open subsets~$D(f)$ for~$f \in A$
+  such that~$\Spec(A) = \bigcup_i D(f_i)$ if and only if~$1 = \sum_i g_i f_i$
+  for some~$g_i$.
+}
+
+% Wieso hat LocRing --> Ring keinen Linksadjungierten?
+% Sollte also nicht Limiten bewahren.
+
+% Recall the definition of a local ring and a local homomorphism!
+
 \subsection{What is the internal language?}
@@
 \frame[t]{\frametitle{The internal language of~$\boldsymbol{\Sh(X)}$}
-  \begin{block}{Crucial property: Locality}
+  \begin{block}{Locality}
     If~$U = \bigcup_i U_i$, then
@@
 
-  \begin{block}{Crucial property: Soundness}
+  \begin{block}{Soundness}
     If~$U \models \varphi$ and if $\varphi$
@@
     \item $U \models f = 0$
-      \tabto{2.9cm} iff $f|_U = 0 \in \Gamma(U, \F)$.
+      \tabto{2.9cm} iff $f|_U = 0 \in \F(U)$.
     \item $U \models \neg\neg(f = 0)$
@@
 
-\section[Little Zariski]{The little Zariski topos of a scheme}
+\section[Little Zariski]{The little Zariski topos of a ring}
+
+\begin{frame}{XXX}
+\end{frame}
+
+% O_{Spec(A)} as the reification of all stalks
+% O_{Spec(A)} integral domain iff all stalks integral domains
+% Transfer theorem
+
+% *Constructed* to yield a universal localisation (mention classifying topos),
+% *observed* to fulfil further properties.
+% Give Mulvey/Tierney citation, state background (quasicoherence)
+% also neg neg (f = 0) ==> f = 0
+
+% Anonymous Noetherian condition
+% Prove it
+
+% Grothendieck's generic freeness lemma
+
+% Outlook
 
 \frame[t]{\frametitle{The little Zariski topos}
-  \begin{block}{Definition}The \hil{little Zariski topos} of a scheme~$X$ is the
-  category~$\Sh(X)$ of set-valued sheaves on~$X$.\end{block}
+  Let~$A$ be a commutative ring (in~$\Set$).
+
+  Is there a \hil{free local ring}~$A \to A'$ over~$A$?
+  \[ \xymatrix{
+    A \ar[rd] \ar[rrr] &&& {\substack{\text{local}\\\text{\normalsize$R$}\\\phantom{\text{local}}}} \\
+    & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!34]{\text{local}}[rru]
+  } \]
+
+  \hil{No,} if we restrict to~$\Set$.
+
+  \hil{Yes,} if we allow a change of topos:
+  Then $A \to \O_{\Spec(A)}$ is the universal localisation.
+
+  \begin{block}{Definition}The \hil{little Zariski topos} of~$A$ is the
+  category~$\Sh(\Spec(A))$ of set-valued sheaves on the \hil{spectrum} of~$A$.\end{block}
+
+  The topology on~$\Spec(A)$ is generated by open subsets~$D(f)$ for~$f \in A$
+  such that~$\Spec(A) = \bigcup_i D(f_i)$ if and only if~$1 = \sum_i g_i f_i$
+  for some~$g_i$.
 

### 2017-09-15 slides-pssl101.tex
Slides almost finished

@@
 
+  \small
+  ``Without loss of generality, any reduced ring is Noetherian and a field.''
+
   \hil{Using the internal language of toposes \\ in commutative algebra}
@@
 
-  PSSL 101 in Leeds \\
-  \ \\
+  101st Peripatetic Seminar on Sheaves and Logic (Leeds) \\
   September 17st, 2017
@@
 
-\section[Internal language]{The internal language of toposes}
+\section[Internal language of toposes]{The internal language of toposes}
 
@@
   \begin{block}{Examples}\begin{itemize}
-    \item \hil{$\Set$:} \tabto{1.4cm} category of sets
-    \item \hil{$\Sh(X)$:} \tabto{1.4cm} category of set-valued sheaves on a space~$X$
-    \item \hil{$\Zar(S)$:} \tabto{1.4cm} big Zariski topos of a base scheme~$S$
+    \item \hil{$\Set$:} \tabto{1.4cm} the category of sets
+    \item \hil{$\Sh(X)$:} \tabto{1.4cm} the category of set-valued sheaves on a space~$X$
+    \item \hil{$\Zar(S)$:} \tabto{1.4cm} the big Zariski topos of a base scheme~$S$
   \end{itemize}\end{block}
@@
 
-\frame[t]{\frametitle{The little Zariski topos}
+\subsection{Universal localisation}
+\frame[t]{\frametitle{Universal localisation}
+  \setbeamercolor{block title}{bg=black!30,fg=black}
+  \begin{block}{Recall}
+    A ring is local iff it has precisely one maximal ideal.
+
+    A homomorphism is local iff it reflects invertibility.
+  \end{block}
+
   Let~$A$ be a ring.
@@
   \hil{Yes,} if we allow a change of topos:
-  Then $A \to \O_{\Spec(A)}$ is the universal localisation.
-
-  \begin{block}{Definition}The \hil{little Zariski topos} of~$A$ is the
-  category~$\Sh(\Spec(A))$ of set-valued sheaves on the \hil{spectrum} of~$A$.\end{block}
-
-  The topology on~$\Spec(A)$ is generated by open subsets~$D(f)$ for~$f \in A$
-  such that~$\Spec(A) = \bigcup_i D(f_i)$ if and only if~$1 = \sum_i g_i f_i$
-  for some~$g_i$.
-}
-
-% Wieso hat LocRing --> Ring keinen Linksadjungierten?
-% Sollte also nicht Limiten bewahren.
-
-% Recall the definition of a local ring and a local homomorphism!
+  Then $A \to A^\sim$ is the universal localisation.
+}
 
 \subsection{What is the internal language?}
-
 \frame[t]{\frametitle{What is the internal language?}
@@
 
-\section[Little Zariski]{The little Zariski topos of a ring}
-
-\begin{frame}{XXX}
-\end{frame}
-
-% O_{Spec(A)} as the reification of all stalks
-% O_{Spec(A)} integral domain iff all stalks integral domains
-% Transfer theorem
+\begin{frame}[c]\frametitle{Praise for Mike Shulman}
+  \centering
+  \includegraphics[scale=0.4]{images/mike-shulman-stack-semantics}
+  \par
+\end{frame}
+
+\section[The little Zariski topos]{The little Zariski topos of a ring}
+
+\begin{frame}{The little Zariski topos}
+  Let~$A$ be a ring. Its \hil{spectrum~$\boldsymbol{\Spec(A)}$} is
+  \begin{itemize}
+    \item generated by opens~$D(f)$ for~$f \in A$
+    \item subject to~$\Spec(A) = \bigcup_i D(f_i)$ iff~$1 = \sum_i g_i f_i$
+    for some~$g_i$.
+  \end{itemize}
+
+  The \hil{little Zariski topos} of~$A$ is the category~$\Sh(\Spec(A))$ of
+  set-valued sheaves on its spectrum. It contains a ring object~$A^\sim$
+  with~$A^\sim(D(f)) = A[f^{-1}]$.
+
+  \begin{block}{Motto}The structure sheaf~$A^\sim$ is a reification of
+  all of the stalks~$A_\ppp$.\end{block}
+
+  For instance, all stalks~$A_\ppp$ are integral domains if and only if \[ \Spec(A) \models
+  \speak{$A^\sim$ is an integral domain}. \]
+\end{frame}
+
+\subsection{Transfer principles}
+\begin{frame}{Transfer principles}
+  \begin{block}{Theorem}
+    The structure sheaf~$A^\sim$ inherits all first-order properties
+    of~$A$ which are stable under localisation.
+  \end{block}
+
+  \hil{Proof.} The structure sheaf~$A^\sim$ is the localisation
+  \[ \underline{A}[\F^{-1}] \]
+  of the constant sheaf~$\underline{A}$ at the \hil{generic filter}~$\F$, a
+  sheaf with
+  \[ \F_\ppp = A \setminus \ppp. \]
+  The rings~$A$ and~$\underline{A}$ share all first-order
+  properties.
+\end{frame}
+
+\note{\justifying
+  The concept of a \emph{prime filter} is a direct axiomatization of what you
+  expect the complement of a prime ideal to fulfil. In classical logic,
+  complementation gives a bijection between the prime filters and the prime
+  ideals of a ring.
+
+  Prime filters are important in constructive mathematics because localising
+  them gives rise to local rings. In contrast, localising a ring at the
+  complement of a prime ideal doesn't usually result in a local ring.
+
+  To construct the universal localisation of~$A$, one doesn't pick a particular
+  prime filter~$F$ to construct~$A[F^{-1}]$. Instead, one picks the
+  \emph{generic prime filter}~$\F$. This filter doesn't live in~$\Set$, but
+  in~$\Sh(\Spec A)$.
+  \par
+}
+
+\subsection[Unique features]{Unique features of the internal world}
+\begin{frame}{Unique features of the internal world}
+  Internally to~$\Sh(\Spec(A))$,
+  \begin{center}
+    \vspace*{-1em}
+    \hil{any non-invertible element of~$\boldsymbol{A^\sim}$ is nilpotent.}
+  \end{center}
+
+  \only<1>{
+    {
+      \centering
+      \includegraphics[scale=0.25]{images/tierney-on-the-spectrum-of-a-ringed-topos} \\
+      \tiny
+      %Miles Tierney. ``On the spectrum of a ringed topos''. \\ In: \emph{Algebra,
+      %Topology, and Category Theory. A Collection of Papers in Honor of Samuel
+      %Eilenberg}. Ed.\@ by A.~Heller and M.~Tierney. Academic Press, 1976,
+      %pp.~189--210.
+      Miles Tierney. On the spectrum of a ringed topos. 1976.
+      \par
+    }
+    \bigskip
+
+    \scriptsize\justifying
+    A sheaf~$\E$ of~$A^\sim$-modules is quasicoherent if and only if,
+    internally to~$\Sh(\Spec(A))$, the set $\E[f^{-1}]$ is a $\nabla_f$-sheaf for
+    any~$f : A^\sim$, where~$\nabla_f\varphi \defeqv (\text{$f$
+    invertible} \Rightarrow \varphi)$.
+    \par
+  }
+
+  \only<2>{
+    If~$A$ is reduced, then furthermore~$A^\sim$ is
+    \begin{itemize}
+      \item reduced,
+      \item a \hil{field} in that non-invertible elements
+      are zero,
+      \item \hil{anonymously Noetherian} in that
+      any ideal is \hil{not not} finitely generated,
+      \item and has~\hil{$\boldsymbol{\neg\neg}$-stable equality}:
+      $\neg\neg(f = 0) \Longrightarrow f = 0$
+    \end{itemize}
+  }
+\end{frame}
+
+\subsection[Generic freeness]{Grothendieck's generic freeness lemma}
+\begin{frame}\frametitle{Generic freeness}
+  Let~$A$ be a reduced ring and~$B, M$ as follows:
+
+  \vspace*{-4.8em}
+  ${\qquad\qquad\quad\qquad\qquad\qquad\qquad\qquad\qquad\qquad}\xymatrixcolsep{3.5pc}\xymatrixrowsep{2pc}\xymatrix{
+    & \ar@{-}[d]^{\substack{\text{finitely}\\\text{generated}}} M \\
+    A \ar[r]_{\text{of finite type}} & B
+  }$
+
+  \hil{Theorem.} There exists $f \neq 0$ in $A$ such that
+  \begin{enumerate}
+    \item $B[f^{-1}]$ and $M[f^{-1}]$ are free modules over $A[f^{-1}]$,
+    \item $A[f^{-1}] \to B[f^{-1}]$ is of finite presentation, and
+    \item $M[f^{-1}]$ is finitely presented as a module over~$B[f^{-1}]$.
+  \end{enumerate}
+  \pause
+
+  \hil{Constructive version.} If zero is the only
+  element~$f \in A$ such that
+  {\renewcommand{\insertenumlabel}{1}\usebeamertemplate{enumerate item}},
+  {\renewcommand{\insertenumlabel}{2}\usebeamertemplate{enumerate item}}, and
+  {\renewcommand{\insertenumlabel}{3}\usebeamertemplate{enumerate item}},
+  then~$1 = 0 \in A$.
+  \pause
+
+  \hil{Proof.} Internally in~$\Sh(\Spec(A))$,
+  \begin{enumerate}
+    \item $B^\sim$ and~$M^\sim$ are \hil{not not} free
+    over~$A^\sim$
+    \item $A^\sim \to B^\sim$ is \hil{not not} of finite presentation,
+    and
+    \item $M^\sim$ is \hil{not not} finitely presented as a module
+    over~$B^\sim$.
+  \end{enumerate}
+\end{frame}
+
+\end{document}
 
@@
 % *observed* to fulfil further properties.
-% Give Mulvey/Tierney citation, state background (quasicoherence)
-% also neg neg (f = 0) ==> f = 0
-
-% Anonymous Noetherian condition
-% Prove it
-
-% Grothendieck's generic freeness lemma
-
-% Outlook
-
-\frame[t]{\frametitle{The little Zariski topos}
-  Let~$A$ be a commutative ring (in~$\Set$).
-
-  Is there a \hil{free local ring}~$A \to A'$ over~$A$?
-  \[ \xymatrix{
-    A \ar[rd] \ar[rrr] &&& {\substack{\text{local}\\\text{\normalsize$R$}\\\phantom{\text{local}}}} \\
-    & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!34]{\text{local}}[rru]
-  } \]
-
-  \hil{No,} if we restrict to~$\Set$.
-
-  \hil{Yes,} if we allow a change of topos:
-  Then $A \to \O_{\Spec(A)}$ is the universal localisation.
-
-  \begin{block}{Definition}The \hil{little Zariski topos} of~$A$ is the
-  category~$\Sh(\Spec(A))$ of set-valued sheaves on the \hil{spectrum} of~$A$.\end{block}
-
-  The topology on~$\Spec(A)$ is generated by open subsets~$D(f)$ for~$f \in A$
-  such that~$\Spec(A) = \bigcup_i D(f_i)$ if and only if~$1 = \sum_i g_i f_i$
-  for some~$g_i$.
-
-  \begin{itemize}
-    \item Internally, the structure sheaf~$\O_X$ looks like \[ \text{
-    an ordinary ring.} \]
-    \item Internally, a sheaf of~$\O_X$-modules looks like \[
-    \text{an ordinary module on that ring.} \]
-  \end{itemize}
-}
 
@@
 \end{frame}
-
-\subsection[Unique features]{Unique features of the internal world}
-
-\begin{frame}\frametitle{Unique features of the internal world}
-  Let~$X$ be a scheme. Internally to~$\Sh(X)$,
-  \begin{center}
-    \hil{any non-invertible element of~$\boldsymbol{\O_X}$ is nilpotent.}
-  \end{center}
-
-  \only<1>{
-    \centering
-    \includegraphics[scale=0.3]{images/tierney-on-the-spectrum-of-a-ringed-topos} \\
-    \tiny
-    %Miles Tierney. ``On the spectrum of a ringed topos''. \\ In: \emph{Algebra,
-    %Topology, and Category Theory. A Collection of Papers in Honor of Samuel
-    %Eilenberg}. Ed.\@ by A.~Heller and M.~Tierney. Academic Press, 1976,
-    %pp.~189--210.
-    Miles Tierney. On the spectrum of a ringed topos. 1976.
-    \par
-  }
-  \pause
-
-  If~$X$ is reduced, this implies:
-  \begin{itemize}
-    \item $\O_X$ is a \hil{field} in that $\neg(\text{$x$ invertible})
-    \Rightarrow x = 0$.
-    \item $\O_X$ has \hil{$\neg\neg$-stable equality}: $\neg\neg(x = 0)
-    \Rightarrow x = 0$.
-    \item $\O_X$ is \hil{anonymously Noetherian}: Any ideal of~$\O_X$ is \mbox{\hil{not
-    not}} finitely generated.
-  \end{itemize}
-\end{frame}
-
-\subsection[Generic freeness]{Grothendieck's generic freeness lemma}
-
-\begin{frame}\frametitle{Generic freeness}
-  \vspace*{-1.6em}
-  \[ \xymatrixcolsep{3.5pc}\xymatrixrowsep{2pc}\xymatrix{
-    & \ar@{-}[d]^{\substack{\text{finitely}\\\text{generated}}} M \\
-    R \ar[r]_{\text{of finite type}} & S
-  } \]
-  \hil{Lemma.} If $R$ is reduced, there
-  %is a \hil{dense open subset} $U \subseteq \Spec R$ such that for any $f \in R$ with~$D(f)
-  is $f \neq 0$ in $R$ such that
-  \begin{enumerate}
-    \item $S[f^{-1}]$ and $M[f^{-1}]$ are free modules over $R[f^{-1}]$,
-    \item $R[f^{-1}] \to S[f^{-1}]$ is of finite presentation, and
-    \item $M[f^{-1}]$ is finitely presented as a module over~$S[f^{-1}]$.
-  \end{enumerate}
-  \pause
-
-  For a trivial proof, employ~$\Sh(\Spec R)$ and exploit that
-  $\O_{\Spec R}$ is a \hil{field} and \hil{anonymously Noetherian}.
-\end{frame}
-
-\subsection{Transfer principles}
-
-\begin{frame}\frametitle{Transfer principles}
-
-  \hil{Question:} How do the properties of
-  \begin{itemize}
-  \item an~$A$-module~$M$ in~$\Set$ and
-  \item the~$\O_X$-module $M^\sim$ in~$\Sh(X)$, where~$X = \Spec A$, relate?
-  \end{itemize}
-  \pause
-  \vfill
-
-  \hil{Observation:} $M^\sim = \underline{M}[\F^{-1}]$, where
-  \begin{itemize}
-  \item $\underline{M}$ is the constant sheaf with stalks~$M$ on~$X$ and
-  \item $\F \hookrightarrow \underline{A}$ is the \hil{generic prime filter}
-  with stalk~$A \setminus \mathfrak{p}$ at $\mathfrak{p} \in \Spec A$.
-  \end{itemize}
-  Note: $M$ and $\underline{M}$ share all first-order properties.
-  \pause
-  \vfill
-
-  \centering
-  \hil{Answer:} $M^\sim$ inherits those properties of $M$ which are \\
-  \hil{stable under localisation}.
-  \par
-\end{frame}
-
-\note{\justifying
-  The concept of a \emph{prime filter} is a direct axiomatization of what you
-  expect the complement of a prime ideal to fulfil. In classical logic,
-  complementation gives a bijection between the prime filters and the prime
-  ideals of a ring.
-
-  Prime filters are important in constructive mathematics because localising
-  them gives rise to local rings. In contrast, localising a ring at the
-  complement of a prime ideal doesn't usually result in a local ring.
-
-  To construct the universal localisation of~$A$, one doesn't pick a particular
-  prime filter~$F$ to construct~$A[F^{-1}]$. Instead, one picks the
-  \emph{generic prime filter}~$\F$. This filter doesn't live in~$\Set$, but
-  in~$\Sh(\Spec A)$.
-  \par
-}
 

### 2017-09-16 slides-pssl101.tex
to do
Slides almost finished

@@
 \usepackage{tikz}
-\usetikzlibrary{calc,shapes.callouts,shapes.arrows}
+\usetikzlibrary{calc,shapes.callouts,shapes.arrows,patterns}
 \hypersetup{colorlinks=true}
@@
   \end{enumerate}
+
+% \scriptsize\begin{itemize}\item[]\tableofcontents\end{itemize}
 }
@@
 
-\subsection{What is the internal language?}
-\frame[t]{\frametitle{What is the internal language?}
+\subsection[Kripke--Joyal semantics]{Kripke--Joyal semantics}
+\frame[t]{\frametitle{The internal language of a topos}
   The internal language of a topos~$\E$ allows to
@@
   set-valued sheaves on its spectrum. It contains a ring object~$A^\sim$
-  with~$A^\sim(D(f)) = A[f^{-1}]$.
-
-  \begin{block}{Motto}The structure sheaf~$A^\sim$ is a reification of
+  with~$A^\sim(D(f)) = A[f^{-1}]$, called the \hil{structure sheaf}.
+
+  \begin{block}{Motto (to be amended)}The structure sheaf~$A^\sim$ is a reification of
   all of the stalks~$A_\ppp$.\end{block}
@@
   properties.
+
+  \hil{Remark.} Same statement for~$A$-modules~$M$ and their mirror
+  image~$M^\sim = \underline{M}[\F^{-1}]$.
 \end{frame}
@@
 
-  \hil{Theorem.} There exists $f \neq 0$ in $A$ such that
+  \hil{Theorem.} If~$1 \neq 0$ in~$A$, there exists $f \neq 0$ in $A$ such that
   \begin{enumerate}
@@
 
-  \hil{Proof.} Internally in~$\Sh(\Spec(A))$,
+  \hil{Proof.} The claim is the translation of the fact that, internally
+  in~$\Sh(\Spec(A))$, it is \hil{not not} the case that
   \begin{enumerate}
-    \item $B^\sim$ and~$M^\sim$ are \hil{not not} free
-    over~$A^\sim$
-    \item $A^\sim \to B^\sim$ is \hil{not not} of finite presentation,
-    and
-    \item $M^\sim$ is \hil{not not} finitely presented as a module
+    \item $B^\sim$ and~$M^\sim$ are free over~$A^\sim$,
+    \item $A^\sim \to B^\sim$ is of finite presentation, and
+    \item $M^\sim$ is finitely presented as a module
     over~$B^\sim$.
@@
 
+\begin{frame}
+  Assume that~$B^\sim$ is generated by~$(x^i y^j)_{i,j\geq0}$ as an~$A^\sim$-module.
+  It's \hil{not not} the case that either some generator can be expressed as a
+  linear combination of others with smaller index, or not.
+
+  \centering
+  \begin{tikzpicture}[scale=0.4]
+    \begin{scope}
+      \draw[step=1cm,gray,very thin] (0,0) grid (8,8);
+
+      \draw
+        (0,8) -- (0,0) -- (8,0);
+
+      \fill[fill=blue!40!white,opacity=0.5]
+        (0,8) -- (0,0) -- (6,0) -- (6,5) -- (5,5) -- (5,8);
+
+      \fill[fill=red!40!white,opacity=0.5]
+        (5,5) -- (6,5) -- (6,6) -- (5,6);
+
+      \node[shape=circle,draw,inner sep=2pt] at (1,1) {1};
+    \end{scope}
+
+    \begin{scope}[xshift=9cm]
+      \draw[step=1cm,gray,very thin] (0,0) grid (8,8);
+
+      \draw
+        (0,8) -- (0,0) -- (8,0);
+
+      \fill[pattern=north east lines,pattern color=gray,opacity=0.5]
+        (5,8) -- (5,5) -- (8,5) -- (8,8);
+      \draw
+        (5,8) -- (5,5) -- (8,5);
+
+      \fill[fill=blue!40!white,opacity=0.5]
+        (0,8) -- (0,0) -- (3,0) -- (3,5) -- (2,5) -- (2,8);
+
+      \fill[fill=red!40!white,opacity=0.5]
+        (2,5) -- (3,5) -- (3,6) -- (2,6);
+
+      \node[shape=circle,draw,inner sep=2pt] at (1,1) {2};
+    \end{scope}
+
+    \begin{scope}[xshift=18cm]
+      \draw[step=1cm,gray,very thin] (0,0) grid (8,8);
+
+      \draw
+        (0,8) -- (0,0) -- (8,0);
+
+      \fill[pattern=north east lines,pattern color=gray,opacity=0.5]
+        (2,8) -- (2,5) -- (8,5) -- (8,8);
+      \draw
+        (2,8) -- (2,5) -- (8,5);
+
+      \fill[fill=blue!40!white,opacity=0.5]
+        (0,8) -- (0,0) -- (7,0) -- (7,2) -- (6,2) -- (6,5) -- (2,5) -- (2,8);
+
+      \fill[fill=red!40!white,opacity=0.5]
+        (6,2) -- (7,2) -- (7,3) -- (6,3);
+
+      \node[shape=circle,draw,inner sep=2pt] at (1,1) {3};
+    \end{scope}
+
+    \begin{scope}[yshift=-9cm]
+      \draw[step=1cm,gray,very thin] (0,0) grid (8,8);
+
+      \draw
+        (0,8) -- (0,0) -- (8,0);
+
+      \fill[pattern=north east lines,pattern color=gray,opacity=0.5]
+        (2,8) -- (2,5) -- (6,5) -- (6,2) -- (8,2) -- (8,8);
+      \draw
+        (2,8) -- (2,5) -- (6,5) -- (6,2) -- (8,2);
+
+      \fill[fill=blue!40!white,opacity=0.5]
+        (0,8) -- (0,0) -- (6,0) -- (6,3) -- (5,3) -- (5,5) -- (2,5) -- (2,8);
+
+      \fill[fill=red!40!white,opacity=0.5]
+        (5,3) -- (6,3) -- (6,4) -- (5,4);
+
+      \node[shape=circle,draw,inner sep=2pt] at (1,1) {4};
+    \end{scope}
+
+    \begin{scope}[yshift=-9cm,xshift=9cm]
+      \draw[step=1cm,gray,very thin] (0,0) grid (8,8);
+
+      \draw
+        (2,8) -- (2,5) -- (5,5) -- (5,3) -- (6,3) -- (6,2) -- (8,2);
+
+      \draw
+        (0,8) -- (0,0) -- (8,0);
+
+      \fill[pattern=north east lines,pattern color=gray,opacity=0.5]
+        (2,8) -- (2,5) -- (5,5) -- (5,3) -- (6,3) -- (6,2) -- (8,2) -- (8,8);
+
+      \fill[fill=blue!40!white,opacity=0.5]
+        (0,8) -- (0,0) -- (4,0) -- (4,3) -- (3,3) -- (3,5) -- (2,5) -- (2,8);
+
+      \fill[fill=red!40!white,opacity=0.5]
+        (3,3) -- (4,3) -- (4,4) -- (3,4);
+
+      \node[shape=circle,draw,inner sep=2pt] at (1,1) {5};
+    \end{scope}
+
+    \begin{scope}[yshift=-9cm,xshift=18cm]
+      \draw[step=1cm,gray,very thin] (0,0) grid (8,8);
+
+      \draw
+        (0,8) -- (0,0) -- (8,0);
+
+      \fill[pattern=north east lines,pattern color=gray,opacity=0.5]
+        (2,8) -- (2,5) -- (3,5) -- (3,3) -- (6,3) -- (6,2) -- (8,2) -- (8,8);
+      \draw
+        (2,8) -- (2,5) -- (3,5) -- (3,3) -- (6,3) -- (6,2) -- (8,2);
+
+      \node[shape=circle,draw,inner sep=2pt] at (1,1) {6};
+    \end{scope}
+  \end{tikzpicture}
+  \par
+\end{frame}
+
+\section{Outlook}
+
+\begin{frame}{Outlook}
+  \begin{itemize}
+    \item Are there general guidelines for when using the internal perspective
+    pays off?
+    \item What about finer topologies (constructible, étale, fppf, \ldots)?
+    \item How should synthetic algebraic geometry be extended?
+  \end{itemize}
+
+  {\vspace{-0.1em}\centering
+  \rotatebox{90}{\tiny\scalebox{0.5}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
+  \includegraphics[scale=0.27]{images/external-internal-small}
+  \par\medskip\vspace{-0.5em}}
+
+  \centering
+  \hil{\href{https://tiny.cc/topos-notes}{https://tiny.cc/topos-notes}}
+  \par
+\end{frame}
+
 \end{document}
@@
 % *observed* to fulfil further properties.
+
+% Outlook (see research statement and mention algebraic geometry)
+% Also mention synthetic algebraic geometry
+% Include dictionary
 

### 2017-09-16 slides-pssl101.tex
typo

@@
 
-  \hil{Remark.} Same statement for~$A$-modules~$M$ and their mirror
+  \hil{Remark.} The analogous statement holds for~$A$-modules~$M$ and their mirror
   image~$M^\sim = \underline{M}[\F^{-1}]$.
@@
 \note{\justifying
-  The concept of a \emph{prime filter} is a direct axiomatization of what you
+  The concept of a \emph{prime filter} is a direct axiomatisation of what you
   expect the complement of a prime ideal to fulfil. In classical logic,

### 2017-10-15 slides-cambridge2016.tex
typo

@@
 
-    % XXX: update
+    We also discuss how the little Zariski topos can be described using the
+    internal language of the big Zariski topos, and vice versa; here too there
+    is a small surprise.
   \end{changemargin}
@@
         \justifying
-        The kernel of any matrix over a principial ideal domain is finitely
+        The kernel of any matrix over a principal ideal domain is finitely
         generated.

### 2017-10-15 slides-phd-defense.tex
Slides for the PhD defense (WIP)

--- 2017-09-16 slides-pssl101.tex
@@
 
-\usepackage[english]{babel}
+\usepackage[ngerman]{babel}
 
 \usepackage{mathtools}
-%\newcommand{\vcentcolon}{:}
 \usepackage{booktabs}
@@
 
-\title{Using the internal language of toposes in commutative algebra}
+\title{Anwendungen der internen Sprache von Topoi in der algebraischen
+Geometrie}
 \author{Ingo Blechschmidt}
-\date{September 17st, 2017}
+\date{October 16th, 2017}
 
@@
     (#1) {#1};\node[overlay,rectangle callout,%
-    callout relative pointer={(0.5cm,0.5cm)},fill=blue!20] at ($(#1.north)+(-1.1cm,-1.1cm)$) {#2};}%
+    callout relative pointer={(-0.2cm,0.5cm)},fill=blue!20] at ($(#1.north)+(0.4cm,-1.1cm)$) {#2};}%
 }%
@@
     \setlength{\fboxsep}{8pt}%
-    {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal text}\parbox{0.93\textwidth}{#1}}}%
+    {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal text}\parbox{0.95\textwidth}{#1}}}%
   \end{center}%
@@
     {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal
-    text}\parbox{0.9\textwidth}{\begin{center}#1\end{center}}}}%
+    text}\parbox{1.02\textwidth}{\begin{center}#1\end{center}}}}%
   \end{center}%
@@
 
-  \small
-  ``Without loss of generality, any reduced ring is Noetherian and a field.''
-
-  \hil{Using the internal language of toposes \\ in commutative algebra}
+  \hil{Anwendungen der internen Sprache von Topoi \\ in der algebraischen
+  Geometrie}
   \medskip
@@
   \scriptsize
-  Ingo Blechschmidt \\
-  University of Augsburg
+  Ingo Blechschmidt
   \medskip
 
-  101st Peripatetic Seminar on Sheaves and Logic (Leeds) \\
-  September 17st, 2017
+  Dissertationsprüfung am \\
+  16. Oktober 2017
+  \par
+\end{frame}
+
+\begin{frame}[plain,c]
+  \centering
+  \includegraphics[scale=0.37]{images/math-overflow-steven-gubkin}
   \par
@@
 % choosing those topos, we can pretend that ...
-\frame[t]{\frametitle{Quick summary}
-  By employing the internal language of toposes in various ways, you can
-  pretend that:
-  \begin{enumerate}
-    \item Sheaves of modules are plain modules.
-    \item Schemes are sets: \[ \PP^2_S = \{
-    [x_0\mathbin{:}x_1\mathbin{:}x_2] \,|\, x_0 \neq 0 \vee x_1 \neq 0 \vee x_2 \neq
-    0 \}. \]
-    \item Reduced rings are Noetherian and in fact fields.
-  \end{enumerate}
-
-% \scriptsize\begin{itemize}\item[]\tableofcontents\end{itemize}
-}
 % Any theorem automatically yields infinitely new theorems; sometimes, these
@@
 
-%\frame[t]{\frametitle{Outline}\small
-%  \begin{columns}[t]
-%    \begin{column}{0.6\textwidth}
-%      \vspace*{-1.5em}
-%      \begin{itemize}\item[]\tableofcontents\end{itemize}
-%      \par
-%    \end{column}
-%    \begin{column}{0.4\textwidth}
-%      \centering
-%      \begin{minipage}{0.70\textwidth}
-%        \begin{exampleblock}{}
-%          \justifying
-%          Any finitely generated vector space does \emph{not not} possess a basis.
-%        \end{exampleblock}
-%      \end{minipage}
-%      \medskip
-%
-%      \scalebox{3}{$\Downarrow$}
-%
-%      \begin{minipage}{0.70\textwidth}
-%        \begin{exampleblock}{}
-%          \justifying
-%          Any sheaf of modules of finite type on a reduced scheme is locally free
-%          \emph{on a dense open subset}.
-%        \end{exampleblock}
-%        \centering
-%        \tiny Ravi Vakil: ``Important hard exercise'' (13.7.K).
-%        \par
-%      \end{minipage}
-%    \end{column}
-%  \end{columns}
-%}
-%
-%\note{\justifying\fontsize{8pt}{9.6}\selectfont
-%  \begin{center}\large\textbf{Abstract}\end{center}
-%
-%  \begin{changemargin}{2.5em}{2.5em}
-%    (new abstract)
-%  \end{changemargin}
-%}
-
-\section[Internal language of toposes]{The internal language of toposes}
-
-\subsection{What is a topos?}
-\frame[t]{\frametitle{What is a topos?}
-  \begin{block}{Formal definition}A \hil{topos} is a category which has finite limits,
-  is cartesian closed and has a subobject classifier.
+\section[Interne Sprache]{Die interne Sprache von Topoi}
+
+\subsection{Was ist ein Topos?}
+\frame[t]{\frametitle{Was ist ein Topos?}
+  \begin{block}{Formale Definition}Ein \hil{Topos} ist eine kartesisch
+  abgeschlossene Kategorie, die alle endliche Limiten und einen
+  Unterobjektklassifizierer besitzt.
   \end{block}
 
-  \begin{block}{Motto}A topos is a category which is sufficiently rich to support
-  an \hil{internal language}.
+  \begin{block}{Motto}Ein Topos ist eine Kategorie, die hinreichend reich ist,
+  um eine \hil{interne Sprache} zu erlauben.
   \end{block}
 
-  \begin{block}{Examples}\begin{itemize}
-    \item \hil{$\Set$:} \tabto{1.4cm} the category of sets
-    \item \hil{$\Sh(X)$:} \tabto{1.4cm} the category of set-valued sheaves on a space~$X$
-    \item \hil{$\Zar(S)$:} \tabto{1.4cm} the big Zariski topos of a base scheme~$S$
+  \begin{block}{Beispiele}\begin{itemize}
+    \item \hil{$\boldsymbol{\Set}$:} \tabto{1.5cm} die Kategorie der Mengen
+    \item \hil{$\boldsymbol{\Sh(X)}$:} \tabto{1.5cm} die Kategorie der Garben auf einem Raum~$X$
+    \item \hil{$\boldsymbol{\Zar(S)}$:} \tabto{1.5cm} der große Zariskitopos eines Basisschemas~$S$
   \end{itemize}\end{block}
-
-  \note{
-    \begin{itemize}
-      \item \begin{justify}While technically correct, the formal definition is actually
-      misleading in a sense: A topos has lots of other vital structure, which
-      is crucial for a rounded understanding, but is not listed in the
-      definition (which is trimmed for minimality).
-
-      A more comprehensive definition is: A \emph{topos} is a locally cartesian
-      closed, finitely complete and cocomplete Heyting category which is exact,
-      extensive and has a subobject classifier.\end{justify}
-
-      \item See~\cite{leinster:topos} for a leisurely introduction to topos
-      theory.
-    \end{itemize}
-  }
-}
-
-\subsection{Universal localisation}
-\frame[t]{\frametitle{Universal localisation}
-  \setbeamercolor{block title}{bg=black!30,fg=black}
-  \begin{block}{Recall}
-    A ring is local iff it has precisely one maximal ideal.
-
-    A homomorphism is local iff it reflects invertibility.
-  \end{block}
-
-  Let~$A$ be a ring.
-  Is there a \hil{free local ring}~$A \to A'$ over~$A$?
-  \[ \xymatrix{
-    A \ar[rd] \ar[rrr] &&& {\substack{\text{local}\\\text{\normalsize$R$}\\\phantom{\text{local}}}} \\
-    & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!34]{\text{local}}[rru]
-  } \]
-
-  \hil{No,} if we restrict to~$\Set$.
-  \hil{Yes,} if we allow a change of topos:
-  Then $A \to A^\sim$ is the universal localisation.
-}
-
-\subsection[Kripke--Joyal semantics]{Kripke--Joyal semantics}
-\frame[t]{\frametitle{The internal language of a topos}
-  The internal language of a topos~$\E$ allows to
+}
+
+\subsection{Was ist die interne Sprache?}
+
+\frame[t]{\frametitle{Was ist die interne Sprache?}
+  Die interne Sprache eines Topos~$\E$ erlaubt es, in einer \hil{naiven
+  Element-basierten Sprache}
   \begin{enumerate}
-    \item construct objects and morphisms of the topos,
-    \item formulate statements about them and
-    \item prove such statements
+    \item Objekte und Morphismen des Topos zu konstruieren,
+    \item Aussagen über diese zu formulieren und
+    \item solche Aussagen zu beweisen.
   \end{enumerate}
-  in a \hil{naive element-based} language:
 
@@
       \toprule
-      externally & internally to $\E$ \\
+      extern & intern in $\E$ \\
       \midrule
-      object of~$\E$ & set \\
-      morphism in~$\E$ & map of sets \\
-      monomorphism & injective map \\
-      epimorphism & surjective map \\
-      group object & group \\
+      Objekt von~$\E$ & Menge \\
+      Morphismus in~$\E$ & Abbildung zwischen Mengen \\
+      Monomorphismus & injektive Abbildung \\
+      Epimorphismus & surjektive Abbildung \\
+      Gruppenobjekt & Gruppe \\
       \bottomrule
@@
 }
-
-\frame[t]{\frametitle{The internal language of~$\boldsymbol{\Sh(X)}$}
+% Interne Sprache von Set nicht vergessen zu erwähnen!
+
+\begin{frame}{Die interne Sprache von~$\boldsymbol{\Sh(X)}$}
   \small
-  Let~$X$ be a topological space. Then we recursively define
-  \[ U \models \varphi \quad \text{(``$\varphi$ holds on~$U$'')} \]
-  for open subsets~$U \subseteq X$ and formulas~$\varphi$.
-  % Write~``$\Sh(X) \models \varphi$'' to mean~$X \models \varphi$.
+  Sei~$X$ ein topologischer Raum. Dann definieren wir rekursiv
+  \[ U \models \varphi \quad \text{("`$\varphi$ gilt auf~$U$"')} \]
+  für offene Teilmengen~$U \subseteq X$ und Formeln~$\varphi$.
   \pause
@@
     U \models \varphi \wedge \psi &\Ll&
-      \text{$U \models \varphi$ and $U \models \psi$} \\
+      \text{$U \models \varphi$ und $U \models \psi$} \\
     U \models \varphi \vee \psi &\Ll&
-      \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt} \\
-    && \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
-    && \quad\quad \text{$U_i \models \varphi$ or $U_i \models \psi$} \\
+      \hcancel{\text{$U \models \varphi$ oder $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt} \\
+    && \text{es gibt eine Überdeckung $U = \bigcup_i U_i$ sd. für alle~$i$:} \\
+    && \quad\quad \text{$U_i \models \varphi$ oder $U_i \models \psi$} \\
     U \models \varphi \Rightarrow \psi &\Ll&
-      \text{for all open~$V \subseteq U$: } 
-    \text{$V \models \varphi$ implies $V \models \psi$} \\
+      \text{für alle Offenen~$V \subseteq U$: } 
+    \text{$V \models \varphi$ impliziert $V \models \psi$} \\
     U \models \forall f \? \F\_ \varphi(f) &\Ll&
-      \text{for all sections~$f \in \F(V), V \subseteq U$: $V \models
+      \text{für alle Schnitte~$f \in \F(V)$ auf~$V \subseteq U$: $V \models
       \varphi(f)$} \\
     U \models \exists f \? \F\_ \varphi(f) &\Ll&
-      \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
-    && \quad\quad \text{there exists~$f_i \in \F(U_i)$ s.\,th.
-    $U_i \models \varphi(f_i)$}
+      \text{es gibt eine Überdeckung~$U = \bigcup_i U_i$ sd. für alle~$i$} \\
+    && \quad\quad \text{$f_i \in \F(U_i)$ mit
+    $U_i \models \varphi(f_i)$ existiert}
   \end{array} \]
-}
-
-\note{\fontsize{8pt}{9.6}\selectfont
-  \begin{itemize}
-    \item Special case: The language of~$\Set$ is the usual mathematical language.
-
-    \item \begin{justify}Actually, the objects of~$\E$ feel more like \emph{types} instead of \emph{sets}:
-    For instance, there is no global membership relation~$\in$. Rather,
-    for each object~$A$ of~$\E$, there is a relation~${\in_A} : A \times \P(A) \to
-    \Omega$, where~$\P(A)$ is the power object of~$A$ and~$\Omega$ is the
-    object of truth values of~$\E$ (can be understood as the power object of
-    a terminal object).\end{justify}
-
-    \item \begin{justify}Compare with the embedding theorem for abelian categories:
-    There, an explicit embedding into a category of modules is constructed.
-    Here, we only change perspective and talk about the same objects and
-    morphisms.\end{justify}
-
-    \item \begin{justify}There exists a weaker variant of the internal language which works
-    in abelian categories. By using it, one can even pretend that the objects
-    are abelian groups (instead of modules), and when constructing morphisms
-    by appealing to the axiom of unique choice (which is a theorem), one
-    doesn't even have to check linearity. The proof that this approach
-    works uses only categorical logic.\end{justify}
-
-    \item \begin{justify}For expositions of the internal language,
-    see Chapters~D1 to~D4 of the Elephant, Chapter~VI of Moerdijk and Mac
-    Lane's book, or Chapter~13 of
-    \href{http://www.mathematik.tu-darmstadt.de/~streicher/CTCL.pdf}{these
-    lecture notes by Thomas Streicher}.\end{justify}
-  \end{itemize}
-}
-
-\note{\begin{itemize}\justifying
-  \item The internal language of a sheaf topos of a $\mathrm{T}_1$-space is
-  \emph{classical} (that is, verifies the principle of excluded middle) if and
-  only if the space is discrete. That's a not particularly interesting special
-  case.
-  \item See Section~2.4 of
-  \href{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
-  notes} for remarks on how to appreciate intuitionistic logic.
-\end{itemize}}
-
-\note{
-  \begin{itemize}
-    \item \begin{justify}The rules are called \emph{Kripke--Joyal semantics} and can be
-    formulated over any topos (not just sheaf toposes). They are not all
-    arbitrary: Rather, they are very finely concerted to make the crucial
-    properties about the internal language (see next slide) true.\end{justify}
-
-    \item \begin{justify}If~$\F$ is an object of~$\Sh(X)$, we write
-    ``$f:\F$'' instead of~``$f\in\F$'' to remind us that~$\F$ is not really
-    (externally) a set consisting of elements, but that we only pretend this
-    by using the internal language.\end{justify}
-
-    \item There are two further rules concerning the constants~$\top$
-    and~$\bot$ (truth resp. falsehood):
-      \[ \renewcommand{\arraystretch}{1.3}\begin{array}{@{}lcl@{}}
-        U \models \top &\Ll& U = U \text{ (always fulfiled)}\\
-        U \models \bot &\Ll& U = \emptyset
-      \end{array} \]
-
-    \item Negation is defined as
-    $\neg\varphi :\equiv (\varphi \Rightarrow \bot)$.
-
-    \item \begin{justify}The alternate definition ``$U \models \varphi \vee \psi :\Leftrightarrow
-    \text{$U \models \varphi$ or $U \models \psi$}$'' would not be local (cf.
-    next slide).\end{justify}
-  \end{itemize}
-}
-
-\note{
-  \begin{itemize}
-    \item Let~$\alpha : \F \to \G$ be a morphism of sheaves on~$X$. Then:
-    \begin{align*}
-      & X \models \speak{$\alpha$ is injective} \\[0.5em]
-      \Longleftrightarrow\
-      & X \models \forall s,t\?\F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
-      \Longleftrightarrow\ &
-        \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
-      &\qquad\qquad
-          U \models \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
-      \Longleftrightarrow\ &
-        \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
-      &\qquad\qquad
-          \text{for all open~$V \subseteq U$:} \\
-      &\qquad\qquad\qquad\qquad
-            \text{$\alpha_V(s|_V) = \alpha_V(t|_V)$ implies $s|_V = t|_V$} \\[0.5em]
-      \Longleftrightarrow\ &
-        \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
-      &\qquad\qquad
-            \text{$\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.5em]
-      \Longleftrightarrow\ &
-        \text{$\alpha$ is a monomorphism of sheaves}
-    \end{align*}
-
-    \item The corner quotes ``$\speak{\ldots}$'' indicate that
-    translation into formal language is left to the reader.
-  \end{itemize}
-}
-
-\note{
-  \begin{itemize}
-    \item Similarly, we have (exercise, use the rules!):
-    \begin{align*}
-      &
-        X \models \speak{$\alpha$ is surjective} \\[0.5em]
-      \Longleftrightarrow\ &
-        X \models \forall s\?\G\_ \exists t\?\F\_ \alpha(t) = s \\[0.5em]
-      \Longleftrightarrow\ &
-        \text{$\alpha$ is an epimorphism of sheaves}
-    \end{align*}
-  \end{itemize}
-}
-
-\note{\fontsize{8pt}{9.6}\selectfont
-  \begin{itemize}
-    \item One can simplify the rules for often-occuring special cases:
-    \[ \renewcommand{\arraystretch}{1.3}\begin{array}{@{}lcl@{}}
-      U \models \forall s\?\F\_ \forall t\?\G\_ \varphi(s,t)
-      &\Longleftrightarrow&
-      \text{for all open~$V \subseteq U$,} \\
-      && \text{sections~$s \in \F(V)$, $t \in \G(V)$:} \\
-      &&\qquad\qquad V \models \varphi(s,t) \\\\
-      U \models \forall s\?\F\_ \varphi(s) \Rightarrow \psi(s)
-      &\Longleftrightarrow&
-      \text{for all open~$V \subseteq U$, sections~$s \in \F(V)$:} \\
-      &&\qquad\qquad \text{$V \models \varphi(s)$ implies $V \models \psi(s)$} \\\\
-      U \models \exists!s\?\F\_ \varphi(s)
-      &\Longleftrightarrow&
-      \text{for all open~$V \subseteq U$,} \\
-      &&
-      \text{there is exactly one section~$s \in \F(V)$ with:} \\
-      &&\qquad\qquad V \models \varphi(s)
-    \end{array} \]
-  \end{itemize}
-}
-
-\note{
-  \begin{itemize}
-    \item \begin{justify}One can extend the language to allow for \emph{unbounded}
-    quantification ($\forall A$ vs.~$\forall a \in A$), by Shulman's
-    \emph{stack semantics}. This is needed to formulate
-    universal properties internal to~$\Sh(X)$, for instance.\end{justify}
-
-    \item \begin{justify}One can further extend the language to be able to talk
-    about locally internal categories over~$\Sh(X)$ (in the sense of
-    Penon, see for instance the appendix
-    of Johnstone's first topos theory book): Then one can do category theory internal
-    to~$\Sh(X)$ using the internal language.
-
-    This specific approach is, as far as I am aware, original work. But of
-    course, internal category theory has been done for a long time, see for
-    instance the Elephant and also Chapman and Rowbottom's \emph{Relative
-    Category Theory and Geometric Morphisms: A Logical Approach}.\end{justify}
-  \end{itemize}
-}
-
-\frame[t]{\frametitle{The internal language of~$\boldsymbol{\Sh(X)}$}
-  \begin{block}{Locality}
-    If~$U = \bigcup_i U_i$, then
-      $U \models \varphi$ iff $U_i \models
-      \varphi$ for all~$i$.
+\end{frame}
+
+\begin{frame}{Die interne Sprache von~$\boldsymbol{\Sh(X)}$}
+  \begin{block}{Lokalität}
+    Falls~$U = \bigcup_i U_i$, dann $U \models \varphi$ genau dann, falls $U_i
+    \models \varphi$ für alle~$i$.
   \end{block}
 
-  \begin{block}{Soundness}
-    If~$U \models \varphi$ and if $\varphi$
-    implies $\psi$ \only<1>{constructively}\only<2->{\pointthis{constructively}{
-      no $\varphi \vee \neg\varphi$,\ \
-      no $\neg\neg\varphi \Rightarrow \varphi$,\ \
-      no axiom of choice
-    }},
-    then~$U \models \psi$.
+  \begin{block}{Korrektheit}
+    Falls~$U \models \varphi$ und falls $\varphi$
+    \only<1>{konstruktiv}\only<2->{\pointthis{konstruktiv}{
+      kein $\varphi \vee \neg\varphi$,\ \
+      kein $\neg\neg\varphi \Rightarrow \varphi$,\ \
+      kein Auswahlaxiom
+    }} $\psi$ impliziert,
+    dann~$U \models \psi$.
   \end{block}
@@
   \pause\pause
-  \begin{block}{A first glance at the constructive nature}
+  \begin{block}{Ein erster Eindruck der konstruktiven Natur}
   \begin{itemize}
     \item $U \models f = 0$
-      \tabto{2.9cm} iff $f|_U = 0 \in \F(U)$.
+      \tabto{2.9cm} genau dann, falls $f|_U = 0 \in \F(U)$.
     \item $U \models \neg\neg(f = 0)$
-      \tabto{2.9cm} iff~$f = 0$ on a dense open subset of~$U$.
+      \tabto{2.9cm} genau dann, falls~$f = 0$ auf einer dichten offenen
+      Teilmenge von~$U$.
   \end{itemize}
   \end{block}
-}
-
-\note{%
-  \vspace{-0.5em}%
-  \begin{center}\large\bf Why is constructive mathematics
-  interesting?\end{center}
-  \vspace{-1em}
-  \begin{itemize}
-    \item The internal logic of most toposes is constructive.
-
-    \item \begin{justify}From a constructive proof of a statement, it's always possible to
-    mechanically extract an \emph{algorithm} witnessing its truth. For example:
-    A proof of the infinitude of primes gives rise to an algorithm
-    which actually computes infinitely many primes (outputting one at a
-    time, never stopping).\end{justify}
-
-    \item \begin{justify}By the celebrated \emph{Curry--Howard correspondence},
-    constructive truth of a formula is equivalent to the existence of a
-    program of a certain type associated to the formula.\end{justify}
-
-    \item \begin{justify}In constructive mathematics, one can experiment with (and draw
-    useful conclusions also holding in a usual sense) \emph{anti-classical
-    dream axioms}, for instance the one of synthetic differential geometry:\end{justify}
-    \begin{quote}All functions~$\RR \to \RR$ are smooth.\end{quote}
-
-    \item \begin{justify}Constructive accounts of classical theories are sometimes more
-    elegant or point out some minor but interesting points which are not appreciated by a
-    classical perspective.\end{justify}
-
-    \item \begin{justify}The philosophical question on the \emph{meaning of truth} is
-    easier to tackle in constructive mathematics.\end{justify}
+\end{frame}
+
+\section[Kleiner Zariski]{Der kleine Zariskitopos eines Schemas}
+
+\begin{frame}{Der kleine Zariskitopos}
+  \begin{block}{Definition}Der \hil{kleine Zariskitopos} eines Schemas~$X$ ist
+  die Kategorie~$\Sh(X)$ der mengenwertigen Garben auf~$X$.\end{block}
+  \bigskip
+
+  Aus interner Sicht von~$\Sh(X)$ \ldots
+  \begin{itemize}
+    \item \ \\[-1.2em]\mbox{sieht die Strukturgarbe~$\O_X$ wie ein \hil{gewöhnlicher Ring} und}
+    \item \ \\[-1.2em]\mbox{eine Garbe von~$\O_X$-Moduln wie ein \hil{gewöhnlicher Modul}} über
+    diesem Ring aus.
   \end{itemize}
-}
-
-\note{%
-  \vspace{-0.5em}%
-  \begin{center}\large\bf Three rumours about constructive mathematics\end{center}
-  \vspace{-1em}
-
-  \begin{enumerate}
-    \item \begin{justify}There is a false rumour about constructive mathematics, namely that
-    the term \emph{contradiction} is generally forbidden. This is not the
-    case, one has to distinguish between
-    \begin{itemize}
-      \item a true proof by contradiction: ``Assume~$\varphi$ were false.
-      Then \ldots, contradiction. So~$\varphi$ is in fact true.''
+\end{frame}
+
+\subsection[Wörterbuch]{Entwicklung eines Wörterbuchs}
+
+\begin{frame}{Entwicklung eines Wörterbuchs}
+  \sloganwithoutborder{\hil{Verstehe Konzepte aus der algebraischen Geometrie
+  als Konzepte aus der Algebra in der internen Welt von~$\boldsymbol{\Sh(X)}$.}}
+  \begin{center}
+    \small
+    \scalebox{0.93}{\begin{tabular}{ll}
+      \toprule
+      extern & intern in $\Sh(X)$ \\
+      \midrule
+      Garbe von Mengen & Mengen \\
+      Morphismus von Garben & Abbildung zwischen Mengen \\
+      Monomorphismus & injektive Abbildung \\
+      Epimorphismus & surjektive Abbildung \\
+      \midrule
+      Garbe von Ringen & Ring \\
+      Garbe von Moduln & Modul \\
+      Garbe von endlichem Typ & endlich erzeugter Modul \\
+      lokal endlich freie Garbe & endlich freier Modul \\
+      % coherent sheaf & coherent module \\
+      Tensorprodukt von Garben & Tensorprodukt von Moduln \\
+      Garbe der Differentialformen & Modul der Kählerdifferentiale \\
+      % sheaf of rational functions & total quotient ring of~$\O_X$ \\
+      Dimension von $X$ & Krulldimension von~$\O_X$ \\
+      \bottomrule
+    \end{tabular}}
+  \end{center}
+
+  \visible<2>{\begin{tikzpicture}[overlay]
+    \draw[fill=white, draw=white, opacity=0.9] (0,0) rectangle (\paperwidth,6.6);
+    \node[anchor=south west,inner sep=0] (image) at (1.8,1.3) {
+      \includegraphics[width=0.7\textwidth]{images/steven-kleiman-misconceptions-about-kx}
+    };
+  \end{tikzpicture}}
+\end{frame}
+
+\begin{frame}{Anwendungen des Wörterbuchs}
+  \vspace*{-2em}
+  \begin{columns}[t]
+    \begin{column}{0.51\textwidth}
+      \begin{center}
+        \begin{exampleblock}{}
+          \justifying
+          Sei~$0 \to M' \to M \to M'' \to 0$ eine kurze exakte Sequenz von
+          Moduln. Sind~$M'$ und~$M''$ endlich erzeugt, so auch~$M$.
+        \end{exampleblock}
+        \medskip
+
+        \scalebox{3}{$\Downarrow$}
+
+        \begin{exampleblock}{}
+          \justifying
+          Sei $0 \to \F' \to \F \to \F'' \to 0$ eine kurze exakte Sequenz
+          von~$\O_X$-Modulgarben. Sind~$\F'$ und~$\F''$ von endlichem Typ,
+          so auch~$\F$.
+        \end{exampleblock}
+      \end{center}
+    \end{column}
+    \pause
+
+    \begin{column}{0.39\textwidth}
+      \begin{center}
+        \begin{exampleblock}{}
+          \justifying
+          Jeder endlich erzeugte Vektorraum besitzt \emph{nicht nicht} eine
+          Basis.
+        \end{exampleblock}
+        \medskip
+
+        \scalebox{3}{$\Downarrow$}
+
+        \begin{exampleblock}{}
+          \justifying
+          Jede Modulgarbe von endlichem Typ auf einem reduzierten Schema ist
+          \emph{auf einer dichten offenen Teilmenge} lokal frei.
+        \end{exampleblock}
+        \centering
+        \tiny Ravi Vakil: "`Important hard exercise"' (13.7.K).
+        \par
+      \end{center}
+    \end{column}
+  \end{columns}
+\end{frame}
+
+\begin{frame}{Mehr zum kleinen Zariskitopos}
+  \vspace*{-2em}
+  \slogan{\justifying Verstehe Konzepte und Sätze der \hil{algebraischen
+  Geometrie} als Konzepte und Sätze der (konstruktiven) \hil{kommutativen
+  Algebra} in der internen Welt des \hil{kleinen Zariskitopos}.}
+
+%  In order to:
+%  \begin{itemize}
+%    \item Gain conceptual understanding.
+%    \item Simplify proofs.
+%    \item Develop a synthetic account of scheme theory.
+%    \item Contribute to constructive algebra.
+%  \end{itemize}
+
+  \begin{itemize}
+    \item Transferprinzipien~$M \leftrightarrow M^\sim$
+    \item Besondere Eigenschaften der internen Welt
+    \item Interne Charakterisierung von Quasikohärenz
+    \item Logische Deutung von Ausbreitungsphänomenen
+    \item Interne Konstruktion des relativen Spektrums
+    \item Interner Beweis von Grothendiecks Lemma über generische Freiheit
   \end{itemize}
-    which constructively is only a proof of the weaker
-    statement~$\neg\neg\varphi$, and
-    \begin{itemize}
-      \item a proof of a negated formula: ``Assume~$\psi$ were true. Then
-      \ldots, contradiction. So~$\neg\psi$ holds.''
-    \end{itemize}
-    which is a perfectly fine proof of~$\neg\psi$ in constructive mathematics.\end{justify}
-
-    \item \begin{justify}There is a similar rumour that constructive mathematicians \emph{deny}
-    the law of excluded middle. In fact, one can constructively prove that there is no
-    counterexample to the law: For any formula~$\varphi$, it holds
-    that~$\neg\neg(\varphi \vee \neg\varphi)$.
-
-    In constructive mathematics, one merely doesn't \emph{use} the law of
-    excluded middle. (Only in concrete models, for example as provided by the
-    internal universe of the sheaf topos on a non-discrete topological space,
-    the law of excluded middle will actually be refutable.)\end{justify}
-  \end{enumerate}
-}
-
-\note{
-  \begin{enumerate}
-    \item[3.] \begin{justify}There is one last false rumour about constructive mathematics: Namely
-    that most of mathematics breaks down in a constructive setting. This is
-    only true if interpreted naively: Often, already very small changes to the
-    definitions and statements (which are classically simply equivalent
-    reformulations) suffice to make them constructively acceptable.
-
-    In other cases, adding an additional hypotheses, which is classically
-    always satisfied, is necessary (and interesting). Here is an example: In
-    constructive mathematics, one can not show that any inhabited subset of the
-    natural numbers possesses a minimal element. [One can also not show the
-    negation -- recall the previous false rumour.] But one can show (quite
-    easily, by induction) that any inhabited and \emph{detachable} subset
-    of the natural numbers possesses a minimal element. A subset~$U \subseteq
-    \NN$ is detachable iff for any number~$n \in \NN$, it holds that~$n
-    \in U$ or~$n \not\in U$.
-
-    This has a computational interpretation: Given an arbitrary inhabited subset~$U
-    \subseteq \NN$, one cannot algorithmically find its minimal element. But it
-    \emph{is} possible if one has an algorithmic \emph{test of membership}
-    for~$U$.\end{justify}
-  \end{enumerate}
-
-  \begin{justify}See~\cite{stanford:constr,dalen:int,troelstra:dalen:constr}
-  for references about constructive mathematics. The blog~\cite{bauer:blog} is
-  also very informative.\end{justify}
-}
-
-\begin{frame}[c]\frametitle{Praise for Mike Shulman}
-  \centering
-  \includegraphics[scale=0.4]{images/mike-shulman-stack-semantics}
-  \par
-\end{frame}
-
-\section[The little Zariski topos]{The little Zariski topos of a ring}
-
-\begin{frame}{The little Zariski topos}
-  Let~$A$ be a ring. Its \hil{spectrum~$\boldsymbol{\Spec(A)}$} is
-  \begin{itemize}
-    \item generated by opens~$D(f)$ for~$f \in A$
-    \item subject to~$\Spec(A) = \bigcup_i D(f_i)$ iff~$1 = \sum_i g_i f_i$
-    for some~$g_i$.
-  \end{itemize}
-
-  The \hil{little Zariski topos} of~$A$ is the category~$\Sh(\Spec(A))$ of
-  set-valued sheaves on its spectrum. It contains a ring object~$A^\sim$
-  with~$A^\sim(D(f)) = A[f^{-1}]$, called the \hil{structure sheaf}.
-
-  \begin{block}{Motto (to be amended)}The structure sheaf~$A^\sim$ is a reification of
-  all of the stalks~$A_\ppp$.\end{block}
-
-  For instance, all stalks~$A_\ppp$ are integral domains if and only if \[ \Spec(A) \models
-  \speak{$A^\sim$ is an integral domain}. \]
-\end{frame}
-
-\subsection{Transfer principles}
-\begin{frame}{Transfer principles}
+\end{frame}
+
+\subsection{Transferprinzipien}
+
+\begin{frame}\frametitle{Transferprinzipien}
   \begin{block}{Theorem}
-    The structure sheaf~$A^\sim$ inherits all first-order properties
-    of~$A$ which are stable under localisation.
+    Die Strukturgarbe~$\O_{\Spec(A)}$ erbt all solche Eigenschaften erster
+    Ordnung von~$A$, die stabil unter Lokalisierung sind.
   \end{block}
 
-  \hil{Proof.} The structure sheaf~$A^\sim$ is the localisation
+  \hil{Beweis.} Die Strukturgarbe~$\O_{\Spec(A)}$ ist die Lokalisierung
   \[ \underline{A}[\F^{-1}] \]
-  of the constant sheaf~$\underline{A}$ at the \hil{generic filter}~$\F$, a
-  sheaf with
-  \[ \F_\ppp = A \setminus \ppp. \]
-  The rings~$A$ and~$\underline{A}$ share all first-order
-  properties.
-
-  \hil{Remark.} Same statement for~$A$-modules~$M$ and their mirror
-  image~$M^\sim = \underline{M}[\F^{-1}]$.
-\end{frame}
-
-\note{\justifying
-  The concept of a \emph{prime filter} is a direct axiomatization of what you
-  expect the complement of a prime ideal to fulfil. In classical logic,
-  complementation gives a bijection between the prime filters and the prime
-  ideals of a ring.
-
-  Prime filters are important in constructive mathematics because localising
-  them gives rise to local rings. In contrast, localising a ring at the
-  complement of a prime ideal doesn't usually result in a local ring.
-
-  To construct the universal localisation of~$A$, one doesn't pick a particular
-  prime filter~$F$ to construct~$A[F^{-1}]$. Instead, one picks the
-  \emph{generic prime filter}~$\F$. This filter doesn't live in~$\Set$, but
-  in~$\Sh(\Spec A)$.
-  \par
-}
-
-\subsection[Unique features]{Unique features of the internal world}
-\begin{frame}{Unique features of the internal world}
-  Internally to~$\Sh(\Spec(A))$,
+  \justifying
+  der konstanten Garbe~$\underline{A}$ am \hil{generischen Filter}~$\F$,
+  einer Garbe mit
+  $\F_\ppp = A \setminus \ppp$.
+  Die Ringe~$A$ und~$\underline{A}$ haben die gleichen Eigenschaften erster
+  Ordnung.
+
+  \hil{Bemerkung.} Die analoge Aussage gilt für~$A$-Moduln~$M$ und ihre Spiegelbilder~$M^\sim =
+  \underline{M}[\F^{-1}]$.
+\end{frame}
+
+\subsection[Besonderheiten]{Besondere Eigenschaften der internen Welt}
+
+\begin{frame}{Besonderheiten der internen Welt}
+  Sei~$X$ ein Schema. Intern in~$\Sh(X)$ gilt:
   \begin{center}
-    \vspace*{-1em}
-    \hil{any non-invertible element of~$\boldsymbol{A^\sim}$ is nilpotent.}
+    \hil{Jedes nicht-invertierbare Element \\ von~$\boldsymbol{\O_X}$ ist nilpotent.}
   \end{center}
@@
   \only<1>{
-    {
     \centering
@@
   }
-    \bigskip
-
-    \scriptsize\justifying
-    A sheaf~$\E$ of~$A^\sim$-modules is quasicoherent if and only if,
-    internally to~$\Sh(\Spec(A))$, the set $\E[f^{-1}]$ is a $\nabla_f$-sheaf for
-    any~$f : A^\sim$, where~$\nabla_f\varphi \defeqv (\text{$f$
-    invertible} \Rightarrow \varphi)$.
-    \par
-  }
 
   \only<2>{
-    If~$A$ is reduced, then furthermore~$A^\sim$ is
-    \begin{itemize}
-      \item reduced,
-      \item a \hil{field} in that non-invertible elements
-      are zero,
-      \item \hil{anonymously Noetherian} in that
-      any ideal is \hil{not not} finitely generated,
-      \item and has~\hil{$\boldsymbol{\neg\neg}$-stable equality}:
-      $\neg\neg(f = 0) \Longrightarrow f = 0$
+    Ist~$X$ reduziert, so folgt daraus für die interne Welt:
+    \begin{itemize}
+      \item $\O_X$ ist ein \hil{Körper}: Nichteinheiten sind Null.
+      \item $\O_X$ hat \hil{$\boldsymbol{\neg\neg}$-stabile Gleichheit}: $\neg\neg(x = 0)
+      \Rightarrow x = 0$.
+      \item $\O_X$ is \hil{anonym noethersch}: Jedes Ideal von~$\O_X$ ist
+      \mbox{\hil{nicht nicht}} endlich erzeugt.
     \end{itemize}
@@
 
-\subsection[Generic freeness]{Grothendieck's generic freeness lemma}
-\begin{frame}\frametitle{Generic freeness}
-  Let~$A$ be a reduced ring and~$B, M$ as follows:
-
-  \vspace*{-4.8em}
-  ${\qquad\qquad\quad\qquad\qquad\qquad\qquad\qquad\qquad\qquad}\xymatrixcolsep{3.5pc}\xymatrixrowsep{2pc}\xymatrix{
-    & \ar@{-}[d]^{\substack{\text{finitely}\\\text{generated}}} M \\
-    A \ar[r]_{\text{of finite type}} & B
+\subsection[Gen. Freiheit]{Grothendiecks Lemma über generische Freiheit}
+
+\begin{frame}{Generische Freiheit}
+  Sei~$A$ ein reduzierter Ring und seien~$B, M$ wie folgt:
+
+  \vspace*{-4.75em}
+  ${\qquad\qquad\qquad\quad\qquad\qquad\qquad\qquad\qquad\qquad\qquad}\xymatrixcolsep{3.5pc}\xymatrixrowsep{2pc}\xymatrix{
+    & \ar@{-}[d]_{\substack{\text{endlich}\\\text{erzeugt}}} M \\
+    A \ar[r]_{\text{veT}} & B
   }$
 
-  \hil{Theorem.} If~$1 \neq 0$ in~$A$, there exists $f \neq 0$ in $A$ such that
+  \hil{Theorem.} Ist~$1 \neq 0$ in~$A$, so gibt es $f \neq 0$ in $A$ sodass
   \begin{enumerate}
-    \item $B[f^{-1}]$ and $M[f^{-1}]$ are free modules over $A[f^{-1}]$,
-    \item $A[f^{-1}] \to B[f^{-1}]$ is of finite presentation, and
-    \item $M[f^{-1}]$ is finitely presented as a module over~$B[f^{-1}]$.
+    \item $B[f^{-1}]$ und $M[f^{-1}]$ als Moduln über $A[f^{-1}]$ frei sind,
+    \item $A[f^{-1}] \to B[f^{-1}]$ von endlicher Präsentation ist und
+    \item $M[f^{-1}]$ als Modul über~$B[f^{-1}]$ endlich präsentiert ist.
   \end{enumerate}
@@
 
-  \hil{Constructive version.} If zero is the only
-  element~$f \in A$ such that
+  \hil{Konstruktive Version.} Ist Null das einzige Element~$f \in A$ mit
   {\renewcommand{\insertenumlabel}{1}\usebeamertemplate{enumerate item}},
-  {\renewcommand{\insertenumlabel}{2}\usebeamertemplate{enumerate item}}, and
+  {\renewcommand{\insertenumlabel}{2}\usebeamertemplate{enumerate item}} und
   {\renewcommand{\insertenumlabel}{3}\usebeamertemplate{enumerate item}},
-  then~$1 = 0 \in A$.
-  \pause
-
-  \hil{Proof.} The claim is the translation of the fact that, internally
-  in~$\Sh(\Spec(A))$, it is \hil{not not} the case that
+  so~$1 = 0 \in A$.
+  \pause
+
+  \hil{Beweis.} Die Behauptung ist die Übersetzung des trivialen Fakts, dass
+  intern in~$\Sh(\Spec(A))$ es \hil{nicht nicht} der Fall ist, dass
   \begin{enumerate}
-    \item $B^\sim$ and~$M^\sim$ are free over~$A^\sim$,
-    \item $A^\sim \to B^\sim$ is of finite presentation, and
-    \item $M^\sim$ is finitely presented as a module
-    over~$B^\sim$.
+    \item $B^\sim$ und~$M^\sim$ als Moduln über~$A^\sim$ frei sind,
+    \item $A^\sim \to B^\sim$ von endlicher Präsentation ist und
+    \item $M^\sim$ als Modul über~$B^\sim$ endlich präsentiert ist.
   \end{enumerate}
@@
 \begin{frame}
-  Assume that~$B^\sim$ is generated by~$(x^i y^j)_{i,j\geq0}$ as an~$A^\sim$-module.
-  It's \hil{not not} the case that either some generator can be expressed as a
-  linear combination of others with smaller index, or not.
+  \small\justifying
+  Sei~$B^\sim$ als~$A^\sim$-Modul von~$(x^i y^j)_{i,j\geq0}$ erzeugt.
+  Es ist \hil{nicht nicht} der Fall, dass entweder ein Erzeuger als
+  Linearkombination von Erzeugern mit kleinerem Index ausgedrückt werden kann,
+  oder nicht.\par
 
@@
 
-\section{Outlook}
-
-\begin{frame}{Outlook}
-  \begin{itemize}
-    \item Are there general guidelines for when using the internal perspective
-    pays off?
-    \item What about finer topologies (constructible, étale, fppf, \ldots)?
-    \item How should synthetic algebraic geometry be extended?
-  \end{itemize}
-
-  {\vspace{-0.1em}\centering
-  \rotatebox{90}{\tiny\scalebox{0.5}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
-  \includegraphics[scale=0.27]{images/external-internal-small}
-  \par\medskip\vspace{-0.5em}}
-
-  \centering
-  \hil{\href{https://tiny.cc/topos-notes}{https://tiny.cc/topos-notes}}
-  \par
-\end{frame}
-
-\end{document}
-
-% *Constructed* to yield a universal localisation (mention classifying topos),
-% *observed* to fulfil further properties.
-
-% Outlook (see research statement and mention algebraic geometry)
-% Also mention synthetic algebraic geometry
-% Include dictionary
-
-\subsection[Dictionary]{Building and using a dictionary}
-
-\begin{frame}\frametitle{Building a dictionary}
-  \sloganwithoutborder{\hil{Understand notions of algebraic geometry
-  as notions of algebra internal to~$\boldsymbol{\Sh(X)}$.}}
-  \begin{center}
-    \small
-    \scalebox{0.93}{\begin{tabular}{ll}
-      \toprule
-      externally & internally to $\Sh(X)$ \\
-      \midrule
-      sheaf of sets & set \\
-      morphism of sheaves & map of sets \\
-      monomorphism & injective map \\
-      epimorphism & surjective map \\
-      \midrule
-      sheaf of rings & ring \\
-      sheaf of modules & module \\
-      sheaf of finite type & finitely generated module \\
-      finite locally free sheaf & finite free module \\
-      % coherent sheaf & coherent module \\
-      tensor product of sheaves & tensor product of modules \\
-      sheaf of Kähler differentials & module of Kähler differentials \\
-      % sheaf of rational functions & total quotient ring of~$\O_X$ \\
-      dimension of $X$ & Krull dimension of~$\O_X$ \\
-      \bottomrule
-    \end{tabular}}
-  \end{center}
-
-  \visible<2>{\begin{tikzpicture}[overlay]
-    \draw[fill=white, draw=white, opacity=0.9] (0,0) rectangle (\paperwidth,6.6);
-    \node[anchor=south west,inner sep=0] (image) at (1.8,1.3) {
-      \includegraphics[width=0.7\textwidth]{images/steven-kleiman-misconceptions-about-kx}
-    };
-  \end{tikzpicture}}
-\end{frame}
-
-\note{\justifying
-  See the \href{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{notes} for more dictionary entries.
-
-  The simple definition of~$\K_X$ allows to give an internal account of the
-  basics of the theory of Cartier divisors, for instance giving an easy
-  description of the line bundle associated to a Cartier divisor.
-  \par
-}
-
-\begin{frame}\frametitle{Using the dictionary}
-  \begin{center}
-    \begin{minipage}{0.70\textwidth}
-      \begin{exampleblock}{}
-        \justifying
-        Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence of
-        modules. If~$M'$ and~$M''$ are finitely generated, so is~$M$.
-      \end{exampleblock}
-    \end{minipage}
-    \medskip
-
-    \scalebox{3}{$\Downarrow$}
-
-    \begin{minipage}{0.70\textwidth}
-      \begin{exampleblock}{}
-        \justifying
-        Let $0 \to \F' \to \F \to \F'' \to 0$ be a short exact sequence
-        of sheaves of~$\O_X$-modules. If~$\F'$ and~$\F''$ are of finite type,
-        so is~$\F$.
-      \end{exampleblock}
-    \end{minipage}
-  \end{center}
-\end{frame}
-
-\begin{frame}[c]\frametitle{Using the dictionary}
-  \begin{center}
-    \begin{minipage}{0.70\textwidth}
-      \begin{exampleblock}{}
-        \justifying
-        Any finitely generated vector space does \emph{not not} possess a basis.
-      \end{exampleblock}
-    \end{minipage}
-    \medskip
-
-    \scalebox{3}{$\Downarrow$}
-
-    \begin{minipage}{0.70\textwidth}
-      \begin{exampleblock}{}
-        \justifying
-        Any sheaf of modules of finite type on a reduced scheme is locally free
-        \emph{on a dense open subset}.
-      \end{exampleblock}
-      \centering
-      \tiny Ravi Vakil: ``Important hard exercise'' (13.7.K).
-      \par
-    \end{minipage}
-  \end{center}
-\end{frame}
-
-\begin{frame}\frametitle{The objective}
-  \slogan{\justifying Understand notions and statements of \hil{algebraic geometry} as
-  notions and statements of (constructive) \hil{commutative
-  algebra} internal to the \hil{little Zariski topos}.}
-
-%  In order to:
-%  \begin{itemize}
-%    \item Gain conceptual understanding.
-%    \item Simplify proofs.
-%    \item Develop a synthetic account of scheme theory.
-%    \item Contribute to constructive algebra.
-%  \end{itemize}
-
-  Further topics regarding the little Zariski topos:
-  \begin{itemize}
-    \item Unique features of the internal world
-    \item Transfer principles~$M \leftrightarrow M^\sim$
-    \item The curious role of affine open subsets
-    \item Quasicoherence
-    \item Spreading from points to neighbourhoods
-    \item The relative spectrum
-  \end{itemize}
-\end{frame}
-
-\section[Big Zariski]{The big Zariski topos of a scheme}
-
-\begin{frame}\frametitle{Synthetic algebraic geometry}
-  Usual approach to algebraic geometry: \hil{layer schemes above ordinary set theory}
-  using either
-  \begin{itemize}
-    \item locally ringed spaces
+\section[Großer Zariski]{Der große Zariskitopos eines Schemas}
+
+\begin{frame}{Synthetische algebraische Geometrie}
+  Üblicher Zugang zur Schematheorie: \hil{Entwickle Schematheorie über
+  gewöhnlicher Mengenlehre} mit
+  \begin{itemize}
+    \item lokal geringten Räumen oder
     \small
     \begin{multline*}
-      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$} + {} \\
-      \text{Zariski topology} + \text{structure sheaf}
+      \text{Menge der Primideale von~$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$} + {} \\
+      \text{Zariskitopologie} + \text{Strukturgarbe}
     \end{multline*}
     \normalsize
-    \item or Grothendieck's functor-of-points account, where a scheme is a functor~$\mathrm{Ring} \to \mathrm{Set}$.
+    \item Grothendiecks Punktefunktoransatz, bei dem ein Schema ein
+    Funktor~$\mathrm{Ring} \to \mathrm{Set}$ ist.
     \small\[ A \longmapsto \{ (x,y,z) \? A^3 \,|\, x^n+y^n-z^n=0 \} \]
@@
 
-  \hil{Synthetic approach:} model schemes \hil{directly as sets} in a certain
-  nonclassical set theory.
+  \hil{Synthetischer Zugang:} Modelliere Schemata \hil{direkt als Mengen} in
+  einer nichtklassischen Mengentheorie.
   \small
@@
 
-\begin{frame}\frametitle{The big Zariski topos}
-  \begin{block}{Definition}The \hil{big Zariski topos} $\Zar(S)$ of a scheme~$S$ is the
-  category $\Sh(\Aff/S)$. It consists of functors
-  $(\Aff/S)^\op \to \Set$ satisfying the gluing condition that
+\begin{frame}{Der große Zariskitopos}
+  \begin{block}{Definition}\justifying
+  Der \hil{große Zariskitopos} $\Zar(S)$ eines
+  Schemas~$S$ ist die Kategorie $\Sh(\Aff/S)$ derjenigen Funktoren
+  $(\Aff/S)^\op \to \Set$, die die Verklebebedingung erfüllen: Für jedes
+  affine Schema~$U = \bigcup_i U_i$ über~$S$ soll
   \[ F(U) \to
@@
   \\[-0.6em]
-  is a limit diagram for any affine scheme~$U = \bigcup_i U_i$ over~$S$.
+  ein Limesdiagramm sein.
   \end{block}
@@
   \begin{itemize}
-    \item For an~$S$-scheme~$X$, its functor of points $\ull{X} =
-    \Hom_S(\cdot,X)$ is an object of~$\Zar(S)$. It feels like \hil{the set of
-    points} of~$X$.
-    \item Internally, $\affl$ (given by $\affl(X) = \O_X(X)$)
-    looks like a field:
-    \[ \Zar(S) \models \forall x\?\affl\_ x \neq 0 \Longrightarrow \text{$x$ invertible} \]
+    \item Der Punktefunktor~$\ull{X} =
+    \Hom_S(\cdot,X)$ eines~$S$-Schemas~$X$ ist ein Objekt von~$\Zar(S)$.
+    "`Die \hil{Menge der Punkte} von~$X$."'
+    \item In der internen Welt ist $\affl$
+    ein Körper:
+    \[ \Zar(S) \models \forall x\?\affl\_ x \neq 0 \Longrightarrow \text{$x$ invertierbar} \]
   \end{itemize}
-
-  \note{
-    \begin{itemize}
-      \item \begin{justify}The overcategory~$\Sch/S$ becomes a Grothendieck site by declaring
-      families of jointly surjective open immersions to be covers. See for
-      instance the excellent Stacks project for details.\end{justify}
-
-      \item \begin{justify}Working in~$\Zar(S)$ amounts to incorporating the
-      philosophy of describing schemes by their functors of points into one's
-      mathematical language.\end{justify}
-
-      \item \begin{justify}Explicitly, the functor~$\ull{X}$ is given
-      by~$\ull{X}(T) = \Hom_S(T,X)$ for~$S$-schemes~$T$. Because the Zariski
-      site is \emph{subcanonical}, this functor is always a sheaf.\end{justify}
-
-      \item \begin{justify}The object~$\ull{S}$ looks like an one-element set
-      from the internal universe. This is to be expected.\end{justify}
-    \end{itemize}
-  }
-\end{frame}
-
-\note{\scriptsize
-  \begin{itemize}
-    \item \begin{justify}Hakim worked out a theory of schemes internal to toposes (but
-    without using the internal language) in her PhD thesis.\end{justify}
-
-    \item \begin{justify}The internal language of~$\Zar(\Spec A)$ is related to
-    the programme about dynamical methods in algebra by Coquand, Coste,
-    Lombardi, Roy, and others. See Coquand's \emph{A completeness proof for
-    geometrical logic}, Coquand and Lombardi's \emph{A logical approach to
-    abstract algebra}, and Coste, Lombardi, and Roy's \emph{Dynamical methods
-    in algebra: effective Nullstellensätze}.\end{justify}
-
-    \item \begin{justify}The observation that~$\affl$ is internally a field is due to
-    Kock (in the case~$S = \Spec\ZZ$, see his \emph{Universal projective
-    geometry via topos theory}) and implies a curious meta-theorem:
-
-    Because $\Zar(\Spec\ZZ)$ is the \emph{classifying topos} for the theory
-    of local rings, any statement about local rings which is of a certain
-    logical form holds for the \emph{universal model}~$\ull{\AA}^1_{\Spec\ZZ}$
-    iff it holds for any local ring (in any universe, particularly~$\Set$).
-
-    Therefore, in proving a statement of such a form about arbitrary local
-    rings, one may assume that they even fulfil the field condition.
-
-    There is a similar story for local~$A$-algebras.
-    See Wraith's \emph{Intuitionistic algebra: some recent developments in
-    topos theory} for a short exposition on the usefulness of
-    classifying toposes and universal models.\end{justify}
-  \end{itemize}
-}
-
-\subsection{Synthetic constructions}
-
-\begin{frame}\frametitle{Synthetic constructions}
+\end{frame}
+
+\subsection{Synthetische Konstruktionen}
+
+\begin{frame}{Synthetische Konstruktionen}
   \begin{itemize}
@@
     x_0 \neq 0 \vee \cdots \vee x_n \neq 0 \}/(\affl)^\times$ \\
-  $\phantom{\mathbb{P}^n} \cong \text{set of one-dimensional subspaces of~$(\affl)^{n+1}$}$.
+  $\phantom{\mathbb{P}^n} \cong \text{Menge der eindimensionalen UVR von~$(\affl)^{n+1}$}$.
 
@@
     \item $\O(-1) = (\ell)_{\ell : \mathbb{P}^n}$
-    \item Euler sequence: $0 \to \ell^\perp \to ((\affl)^{n+1})^\vee \to \ell^\vee \to 0$
+    \item Eulersequenz: $0 \to \ell^\perp \to ((\affl)^{n+1})^\vee \to \ell^\vee \to 0$
   \end{itemize}\bigskip
 
-  \item \ \\[-1.2em]\mbox{$\RelSpec R = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
-  \text{set of $\affl$-valued points of $R$}.$}
+  \item $\RelSpec R = \Hom_{\mathrm{Alg}(\affl)}(R, \affl)$ \\
+  $\phantom{\RelSpec R} = \text{Menge der $\affl$-wertigen Punkte von $R$}.$
 
@@
 
-\subsection{Synthetic formulation of properties}
-
-\begin{frame}{Synthetic formulation of properties}
-  \begin{itemize}
-  \item \ \\[-1.2em]\mbox{$\RelSpec R = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
-  \text{set of $\affl$-valued points of $R$}.$}
-
-  \item An $\affl$-module~$E$ is \hil{quasicoherent} if and only if
-  \[ E \otimes_\affl R \longrightarrow \Hom(\Spec R, E) \]
-  is an isomorphism for all finitely presented $\affl$-algebras $R$.
-
-  \item In particular, any map $\affl \to \affl$ is given by a polynomial. \bigskip
-  \pause
-
-  \item A subset $U \subseteq X$ is qc-open if and only if for any $x : X$
-  there exist $f_1,\ldots,f_n \? \affl$ such that
-  $x \in U \Longleftrightarrow \exists i\_ f_i \neq 0$.
-
-  \item Open subsets are $\neg\neg$-stable:
-  $\neg\neg(x \in U) \Longrightarrow x \in U$.
-
-  \item If $\gamma : \Delta \to X$ is a tangent vector with $\gamma(0) \in U$,
-  then~$\gamma(\varepsilon) \in U$ for all~$\varepsilon \in \Delta$.
-  \bigskip\pause
-
-  \item $X$ is separated if and only if for any $x,y : X$,
-  there exists a quasicoherent ideal~$J \subseteq \affl$ such that
-  $x = y \Longleftrightarrow J = (0)$.
+\subsection{Synthetische Eigenschaften}
+
+\begin{frame}{Formulierung von Eigenschaften}
+  \begin{block}{Definition}
+    \small
+    Ein~$\affl$-Modul~$E$ ist genau dann \hil{synthetisch quasikohärent}, wenn
+    \vspace*{-0.5em}
+    \[ E \otimes_\affl R \longrightarrow E^{\Spec R} \]
+    \vspace*{-1.7em}
+
+    für alle endlich präsentierten~$\affl$-Algebren~$R$ ein Isomorphismus ist.
+    (Dabei $\Spec R = \Hom_{\mathrm{Alg}(\affl)}(R, \affl)$.)
+  \end{block}
+
+  \begin{itemize}
+  \item Aus der synthetischen Quasikohärenz von~$\affl$ folgt:
+
+  Jede Abbildung $\affl \to \affl$ ist polynomiell.
+
+  \item\justifying Offene und abgeschlossene Immersionen, quasikompakte und
+  quasiseparierte Morphismen, synthetische Schemata, eigentliche Morphismen,
+  \ldots
+
+  \item Charakterisierung von wichtigen Untertopoi wie dem \hil{étalen Topos}
+  oder dem \hil{fppf-Topos}
   \end{itemize}
@@
 
-% Add two bigger case studies:
-% * Construction of the Grassmannian
-%
-% And add the following smaller things:
-% * qc, sep, qs
-% * universally closed
-% * surjective
-
-\note{\justifying
-  I'm grateful to Zhen Lin Low for suggesting the example about the projective
-  space.
-
-  Explicitly, the~$\affl$-algebra~$\A^\sim$ is given by
-  \[ \A^\sim(X \xrightarrow{\mu} S) = (\mu^* \A)(X). \]
-}
-
-\section{Open tasks}
-
-\begin{frame}\frametitle{Semi-open and open tasks}
-  \begin{itemize}
-  \item Do cohomology in the little Zariski topos; exploit that higher direct
-  images look like ordinary sheaf cohomology from the internal point of view.
-  \item Do cohomology in the big Zariski topos.
-  \item Understand subtoposes of the big Zariski topos.
+\section{Offene Fragen}
+
+\begin{frame}{Offene Fragen}
+  \begin{itemize}
+    \small
+    \item Wie geht Kohomologie in der internen Welt wirklich?
+    \item Welche Theorien klassifizieren relevante Untertopoi?
+    \item Welche Anwendungen liefern andere Topologien?
+    \item Was ist der Bezug zu den dynamischen Methoden der Algebra?
+    \item Wie sollte synthetische algebraische Geometrie weiter entwickelt
+    werden?
   \end{itemize}
@@
 
-\begin{frame}{}
-
-  {\vspace{-0.1em}\centering
-  \rotatebox{90}{\tiny\scalebox{0.5}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
-  \includegraphics[scale=0.17]{images/external-internal-small}
-  \par\medskip\vspace{-0.1em}}
-
-  \slogan{\hil{Understand notions and statements of algebraic geometry
-  as notions and statements of algebra internal to the little Zariski topos.}}
-
-  \slogan{\hil{Develop a synthetic account of algebraic geometry.}}
-
-  \begin{itemize}
-    \item Simplify proofs and gain conceptual understanding.
-    \item Understand relative geometry as absolute geometry.
-    \item Contribute to constructive algebra.
-  \end{itemize}
-
-  \centering
-  \hil{\href{http://tiny.cc/topos-notes}{http://tiny.cc/topos-notes}}
-  \par
-\end{frame}
-
-{\usebackgroundtemplate{\includegraphics[width=\paperwidth]{images/mathcamp-augsburg}}
-\begin{frame}[plain]
-  \vspace{\textheight}\vspace{-1em}\centering
-  \emph{Participants of Augsburg's maths camp}
-  \par
-\end{frame}}
-
-{\usebackgroundtemplate{\includegraphics[width=\paperwidth]{images/mathcamp-space}}
-\begin{frame}[plain]
-  \vspace{\textheight}\vspace{-1em}\centering
-  \emph{The sun as seen from our high-altitude balloon}
-  \par
-\end{frame}}
-
 \appendix
-
-\section{Details on the internal language}
-
-\begin{frame}\frametitle{Translating internal statements I}
-  Let~$X$ be a topological space (or locale) and let~$\alpha : \F \to \G$ be a
-  morphism of sheaves on~$X$. Then:
-  \allowdisplaybreaks
-  \begin{align*}
-    & \Sh(X) \models \speak{$\alpha$ is surjective} \\[0.5em]
-    \Longleftrightarrow\
-    & \Sh(X) \models \forall t\?\G\_ \exists s\?\F\_ \alpha(s) = t \\[0.5em]
-    \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$, sections $t \in \G(U)$:} \\
-    &\qquad
-      \text{there exists an open covering~$U = \textstyle\bigcup_i U_i$ and} \\
-    &\qquad
-      \text{sections~$s_i \in \F(U_i)$ such that:} \\
-    &\qquad\qquad
-        \alpha_{U_i}(s_i) = t|_{U_i} \\[0.5em]
-    \Longleftrightarrow\ &
-      \text{$\alpha$ is an epimorphism of sheaves}
-  \end{align*}
-\end{frame}
-
-\begin{frame}\frametitle{Translating internal statements II}
-  Let~$X$ be a topological space (or locale) and let~$s, t \in \F(X)$ be global
-  sections of a sheaf~$\F$ on~$X$. Then:
-  \allowdisplaybreaks
-  \begin{align*}
-    & \Sh(X) \models \neg\neg(s = t) \\[0.5em]
-    \Longleftrightarrow\
-    & \Sh(X) \models ((s = t) \Rightarrow \bot) \Rightarrow \bot \\[0.5em]
-    \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$ such that} \\
-    &\qquad
-      \text{for all open~$V \subseteq U$ such that} \\
-    &\qquad\qquad
-      s|_V = t|_V, \\
-    &\qquad
-      \text{it holds that~$V = \emptyset$,} \\
-    &
-      \text{it holds that~$U = \emptyset$} \\[0.5em]
-    \Longleftrightarrow\ &
-      \text{there exists a dense open set~$W \subseteq X$ such that $s|_W = t|_W$} \\
-  \end{align*}
-\end{frame}
 
@@
 
-\note{\justifying
-  Statements 1 and 2 follow from \emph{one} proof in the internal language,
-  applied to two different modal operators.
-
-  Similarly with statements~5 and~6.
-  \par
-}
-
-\section{The meromorphic functions revisited}
-
-\begin{frame}\frametitle{The smallest dense sublocale}\justifying
-  Let~$X$ be a reduced scheme satisfying a technical condition.
-  Let~$i : X_{\neg\neg} \to X$ be the inclusion of the smallest dense sublocale
-  of~$X$.
-
-  Then~$i_* i^{-1} \O_X \cong \K_X$.
-
-  \begin{itemize}\justifying
-    \item This is a highbrow way of saying ``rational functions are regular
-    functions which are defined on a dense open subset''.
-    \item Another reformulation is that~$\K_X$ is the sheafification of~$\O_X$
-    with respect to the~$\neg\neg$-modality.
-    \item There is a generalization to nonreduced schemes.
-  \end{itemize}
-\end{frame}
-
-\subsection{Group schemes}
-
-\begin{frame}\frametitle{Group schemes}
-  \hil{Motto:} Internal to~$\Zar(S)$, group schemes look like ordinary groups.
-  \bigskip
-
-  \small%
-  \begin{tabular}{lll}
-    \toprule
-    group scheme & internal definition & functor of points: $X \mapsto \ldots$ \\\midrule
-    $\GG_\text{a}$ & $\affl$ (as additive group) & $\O_X(X)$ \\[0.5em]
-    $\GG_\text{m}$ & $\{ x \? \affl \,|\, \speak{$x$ inv.} \}$ & $\O_X(X)^\times$ \\[0.5em]
-    $\mu_n$ & $\{ x \? \affl \,|\, x^n = 1 \}$ & $\{ f \in \O_X(X) \,|\, f^n = 1 \}$ \\[0.5em]
-    $\GL_n$ & $\{ M \? \affl^{n \times n} \,|\, \speak{$M$ inv.} \}$ & $\GL_n(\O_X(X))$ \\
-    \bottomrule
-  \end{tabular}
-  \vspace{2em}
-\end{frame}
-
-\section{Applications in algebra}
-
-\begin{frame}\frametitle{Applications in algebra}
-  Let~$A$ be a commutative ring.
-  The internal language of~$\Sh(\Spec A)$ allows you to say ``without loss of
-  generality, we may assume that~$A$ is local'', even constructively.
-
-  \begin{center}
-    \begin{minipage}{0.75\textwidth}
-      \begin{exampleblock}{}
-        \justifying
-        The kernel of any matrix over a principial ideal domain is finitely
-        generated.
-      \end{exampleblock}
-    \end{minipage}
-    \medskip
-
-    \scalebox{3}{$\Downarrow$}
-
-    \begin{minipage}{0.75\textwidth}
-      \begin{exampleblock}{}
-        \justifying
-        The kernel of any matrix over a Prüfer domain is finitely generated.
-      \end{exampleblock}
-    \end{minipage}
-  \end{center}
-\end{frame}
-
-\begin{frame}\frametitle{Hilbert's program in algebra}
-  \scriptsize\justifying
-  There is a way to combine some of the powerful tools of classical ring theory
-  with the advantages that constructive reasoning provides, for instance
-  exhibiting explicit witnesses. Namely we can devise
-  a language in which we can usefully talk about prime ideals, but which
-  substitutes non-constructive arguments by constructive arguments ``behind
-  the scenes''. The key idea is to substitute the phrase ``for all prime ideals''
-  (or equivalently ``for all prime filters'') by ``for the generic prime filter''.
-
-  More specifically, simply interpret a given proof using prime filters
-  in~$\Sh(\Spec A)$ and let it refer to~$\F \hookrightarrow \underline{A}$.
-
-  \hspace*{-0.75cm}%
-  \begin{tabular}{lll}
-    \toprule
-    Statement & constructive substitution & meaning \\\midrule
-    $x \in \ppp$ for all~$\ppp$. &
-    $x \not\in \F$. &
-    $x$ is nilpotent. \\
-    $x \in \ppp$ for all~$\ppp$ such that~$y \in \ppp$. &
-    $x \in \F \Rightarrow y \in \F$. &
-    $x \in \sqrt{(y)}$. \\
-    $x$ is regular in all stalks~$A_\ppp$. &
-    $x$ is regular in~$\ull{A}[\F^{-1}]$. &
-    $x$ is regular in~$A$. \\
-    The stalks~$A_\ppp$ are reduced. &
-    $\ull{A}[\F^{-1}]$ is reduced. &
-    $A$ is reduced. \\
-    The stalks~$M_\ppp$ vanish. &
-    $\ull{M}[\F^{-1}] = 0$. &
-    $M = 0$. \\
-%   The stalks~$M_\ppp$ are fin.\@ gen.\@ over~$A_\ppp$. &
-%   $\ull{M}[\F^{-1}]$ is fin.\@ gen.\@ over
-%   $\ull{A}[\F^{-1}]$. &
-%   $M$ is fin.\@ gen.\@ over~$A$. \\
-    The stalks~$M_\ppp$ are flat over~$A_\ppp$. &
-    $\ull{M}[\F^{-1}]$ is flat over~$\ull{A}[\F^{-1}]$. &
-    $M$ is flat over~$A$. \\
-    The maps~$M_\ppp \to N_\ppp$ are injective. &
-    $\ull{M}[\F^{-1}] \to \ull{N}[\F^{-1}]$ is injective. &
-    $M \to N$ is injective. \\
-    The maps~$M_\ppp \to N_\ppp$ are surjective. &
-    $\ull{M}[\F^{-1}] \to \ull{N}[\F^{-1}]$ is surjective. &
-    $M \to N$ is surjective. \\
-    \bottomrule
-  \end{tabular}
-
-  This is related (in a few cases equivalent) to the \emph{dynamical methods in
-  algebra} explored by Coquand, Coste, Lombardi, Roy, and others. Their
-  approach is more versatile.
-  \par
-\end{frame}
-
-\subsection[Affine opens]{The curious role of affine open subsets}
-
-\begin{frame}\frametitle{The curious role of affine open subsets}
-  \hil{Question:} Why do the following identities hold, for quasicoherent
-  sheaves~$\E$ and~$\F$ and affine open subsets~$U$?
-  \begin{align*}
-    (\E/\F)(U) &= \E(U) / \F(U) \\
-    (\E \otimes_{\O_X} \F)(U) &= \E(U) \otimes_{\O_X(U)} \F(U) \\
-    \E_\mathrm{tors}(U) &= \E(U)_\mathrm{tors} \quad\text{(sometimes)} \\
-    \K_X(U) &= \operatorname{Quot} \O_X(U) \quad\text{(sometimes)}
-  \end{align*}
-  \pause
-  \hil{A calculation:}
-  \begin{align*}
-    M^\sim \otimes_{\O_U} N^\sim &=
-    \ull{M}[\F^{-1}] \otimes_{\ull{A}[\F^{-1}]} \ull{N}[\F^{-1}] =
-    (\ull{M} \otimes_\ull{A} \ull{N})[\F^{-1}] \\
-    &= (\ull{M \otimes_A N})[\F^{-1}] =
-    (M \otimes_A N)^\sim.
-  \end{align*}
-  \pause
-  \hil{Answer:} Because localisation commutes with quotients, tensor products,
-  torsion submodules (sometimes), \ldots
-\end{frame}
-
 \subsection{Quasicoherence}

### 2017-11-01 slides-augsburg2016.tex
Reorganization

--- 2016-05-19 slides-augsburg.tex
@@
         \justifying
-        Any finitely generated vector space does \emph{not not} possess a basis.
+        Any finitely generated vector space does \mbox{\emph{not not}} possess a basis.
       \end{exampleblock}

### 2017-11-01 slides-warwick2017.tex
typo

@@
 
-  \hil{Synthetic approach}: model schemes \hil{directly as sets} in a certain
+  \hil{Synthetic approach:} model schemes \hil{directly as sets} in a certain
   nonclassical set theory.

### 2018-02-11 slides-pssl103.tex
Abstract for PSSL 103

"Any map from the reals to the reals is smooth." This statement holds in the
context of synthetic differential geometry, a well-developed account of smooth
manifolds and related notions which allows us to bring formal reasoning much
closer to geometric intuition.

We employ the internal language of the big Zariski topos of a base scheme to
give a similar account of algebraic geometry, reminiscent of the language of
the classical Italian school and incorporating Grothendieck's functor-of-points
philosophy. From the point of view of the Zariski topos, a scheme over the base
will look like a plain old set, and the affine line will look like a certain
field. Central to the synthetic account is the notion of "synthetic
quasicoherence", which doesn't have an analog in synthetic differential
geometry and which gives the account a distinct algebraic flavor.

The higher-order axiom of synthetic quasicoherence implies all known internal
properties of the affine line, for instance that it is a field and that it is
algebraically closed in a weak sense. We surmise that this is for a deeper
reason, related to the age-old question "which nongeometric sequents hold in
the classifying topos of a geometric theory?". The second part of the talk
reports on work in progress about this topic.

### 2018-03-20 abstract-unilog2018.tex
Abstract for UNILOG 2018

\documentclass{article}
\begin{document}

\subsection*{Exploring the internal language of toposes}
\addcontentsline{toc}{subsection}{Exploring the internal language of toposes, by Ingo
Blechschmidt}

\noindent
{\scshape Ingo Blechschmidt}\index{Blechschmidt, Ingo}\\
{\scshape Institute for Mathematics,
University of Augsburg, Augsburg, Germany}\\
{\scshape ingo.blechschmidt@math.uni-augsburg.de}\\

Since the work of the early pioneers in the 1970s, it's known that any topos
supports an \emph{internal language}, which allows to speak and reason about
its objects and morphisms in a naive element-based language: From the internal
perspective, objects of the topos look like sets, morphisms look like maps
between sets, epimorphisms look like surjective maps, group objects look like
plain groups and so on; and any theorem which has an intuitionistic proof also
holds in the internal universe of a topos.

With recent discoveries of new applications of the internal language in
algebra, geometry, homotopy theory, mathematical physics and measure theory,
the study of the internal language of toposes is currently experiencing a
resurgence. Our goal is give an introduction to this topic and illustrate the
usefulness of the internal language with two concrete examples.

Firstly, the internal language of the ``little Zariski topos'' allows us to
assume without loss of generality that any reduced ring is Noetherian and in
fact a field, as long as we restrict to intuitionistic reasoning. This
technique yields for instance a simple one-paragraph proof of
Grothendieck's generic freeness lemma, because it is trivial for fields. We
thereby improve on the substantially longer and somewhat convoluted previously known proofs.

Secondly, the internal language of the ``big Zariski topos'' can be used to
develop a synthetic account of algebraic geometry, in which schemes appear as
plain sets and morphisms of schemes appear as maps between these sets.
Fundamental to this account is the notion of ``synthetic quasicoherence'',
which doesn't have a counterpart in synthetic differential geometry and which
endows the internal universe with a distinctive algebraic flavor.

Somewhat surprisingly, the work on synthetic algebraic geometry is related to
an age-old question in the study of classifying toposes. The talk closes with
an invitation to the many open problems of the field.

\end{document}

### 2018-04-05 slides-pssl103.tex
Slides for PSSL 103 (WIP)

@@
+\documentclass[12pt,utf8,notheorems,compress,t]{beamer}
+\usepackage{etex}
+
+\usepackage[english]{babel}
+
+\usepackage{mathtools}
+\usepackage{booktabs}
+\usepackage{array}
+\usepackage{ragged2e}
+\usepackage{multicol}
+\usepackage{tabto}
+\usepackage{xstring}
+\usepackage{soul}\setul{0.3ex}{}
+\usepackage[all]{xy}
+\xyoption{rotate}
+\usepackage{tikz}
+\usetikzlibrary{calc,shapes.callouts,shapes.arrows,patterns}
+\hypersetup{colorlinks=true}
+
+\usepackage{pifont}
+\newcommand{\cmark}{\ding{51}}
+\newcommand{\xmark}{\ding{55}}
+
+\graphicspath{{images/}}
+
+\usepackage[protrusion=true,expansion=true]{microtype}
+
+\setlength\parskip{\medskipamount}
+\setlength\parindent{0pt}
+
+\title{Synthetic algebraic geometry and the phenomenon of nongeometric sequents}
+\author{Ingo Blechschmidt}
+\date{April 7th, 2018}
+
+\useinnertheme[shadow=true]{rounded}
+\useoutertheme{split}
+\usecolortheme{orchid}
+\usecolortheme{whale}
+\setbeamerfont{block title}{size={}}
+
+\useinnertheme{rectangles}
+
+\usecolortheme{seahorse}
+\definecolor{mypurple}{RGB}{150,0,255}
+\setbeamercolor{structure}{fg=mypurple}
+\definecolor{myred}{RGB}{150,0,0}
+\setbeamercolor*{title}{bg=myred,fg=white}
+\setbeamercolor*{titlelike}{bg=myred,fg=white}
+
+\usefonttheme{serif}
+\usepackage[T1]{fontenc}
+\usepackage{libertine}
+
+\newcommand{\A}{\mathcal{A}}
+\renewcommand{\AA}{\mathbb{A}}
+\newcommand{\E}{\mathcal{E}}
+\newcommand{\F}{\mathcal{F}}
+\renewcommand{\G}{\mathcal{G}}
+\newcommand{\GG}{\mathbb{G}}
+\renewcommand{\O}{\mathcal{O}}
+\newcommand{\K}{\mathcal{K}}
+\newcommand{\NN}{\mathbb{N}}
+\newcommand{\RR}{\mathbb{R}}
+\newcommand{\PP}{\mathbb{P}}
+\newcommand{\ZZ}{\mathbb{Z}}
+\renewcommand{\P}{\mathcal{P}}
+\newcommand{\ppp}{\mathfrak{p}}
+\newcommand{\defeq}{\vcentcolon=}
+\newcommand{\defeqv}{\vcentcolon\equiv}
+\newcommand{\Sh}{\mathrm{Sh}}
+\newcommand{\GL}{\mathrm{GL}}
+\newcommand{\Zar}{\mathrm{Zar}}
+\newcommand{\op}{\mathrm{op}}
+\newcommand{\Set}{\mathrm{Set}}
+\newcommand{\Eff}{\mathrm{Ef{}f}}
+\newcommand{\Sch}{\mathrm{Sch}}
+\newcommand{\Aff}{\mathrm{Aff}}
+\newcommand{\LRS}{\mathrm{LRS}}
+\newcommand{\Hom}{\mathrm{Hom}}
+\DeclareMathOperator{\Spec}{Spec}
+\newcommand{\lra}{\longrightarrow}
+\newcommand{\RelSpec}{\operatorname{Spec}}
+\renewcommand{\_}{\mathpunct{.}}
+\newcommand{\?}{\,{:}\,}
+\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
+\newcommand{\ull}[1]{\underline{#1}}
+\newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
+\newcommand{\Ll}{\vcentcolon\!\Longleftrightarrow}
+
+\setbeamertemplate{blocks}[rounded][shadow=false]
+
+\newcommand{\fmini}[2]{\framebox{\begin{minipage}{#1}#2\end{minipage}}}
+\makeatletter
+\def\underunbrace#1{\mathop{\vtop{\m@th\ialign{##\crcr
+      $\hfil\displaystyle{#1}\hfil$\crcr\noalign{\kern3\p@\nointerlineskip}
+      \crcr\noalign{\kern3\p@}}}}\limits}
+\def\overunbrace#1{\mathop{\vbox{\m@th\ialign{##\crcr\noalign{\kern3\p@}
+      \crcr\noalign{\kern3\p@\nointerlineskip}
+      $\hfil\displaystyle{#1}\hfil$\crcr}}}\limits}
+\makeatother
+
+\newenvironment{changemargin}[2]{%
+  \begin{list}{}{%
+    \setlength{\topsep}{0pt}%
+    \setlength{\leftmargin}{#1}%
+    \setlength{\rightmargin}{#2}%
+    \setlength{\listparindent}{\parindent}%
+    \setlength{\itemindent}{\parindent}%
+    \setlength{\parsep}{\parskip}%
+  }%
+  \item[]}{\end{list}}
+
+\newcommand{\pointthis}[2]{%
+  \tikz[remember picture,baseline]{\node[anchor=base,inner sep=0,outer sep=0]%
+    (#1) {#1};\node[overlay,rectangle callout,%
+    callout relative pointer={(-0.5cm,1.0cm)},fill=blue!20] at ($(#1.north)+(0.9cm,-1.6cm)$) {#2};}%
+}%
+
+\newcommand{\hcancel}[5]{%
+  \tikz[baseline=(tocancel.base)]{
+    \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
+    \draw[red, line width=0.4mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
+  }%
+}
+
+\newcommand{\slogan}[1]{%
+  \begin{center}%
+    \setlength{\fboxrule}{2pt}%
+    \setlength{\fboxsep}{8pt}%
+    {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal text}\parbox{0.93\textwidth}{#1}}}%
+  \end{center}%
+}
+
+\newcommand{\sloganwithoutborder}[1]{%
+  \begin{center}%
+    \setlength{\fboxrule}{0pt}%
+    \setlength{\fboxsep}{-14pt}%
+    {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal
+    text}\parbox{0.9\textwidth}{\begin{center}#1\end{center}}}}%
+  \end{center}%
+}
+
+% Adapted from https://latex.org/forum/viewtopic.php?t=2251 (Stefan Kottwitz)
+\newenvironment<>{varblock}[2]{
+  \begin{center}
+    \begin{minipage}{#1}
+      \setlength{\textwidth}{#1}
+      \begin{actionenv}#3
+	\def\insertblocktitle{\centering #2}
+	\par
+	\usebeamertemplate{block begin}}{
+        \par
+        \usebeamertemplate{block end}
+      \end{actionenv}
+    \end{minipage}
+  \end{center}}
+
+\setbeamertemplate{frametitle}{%
+  \vskip0.7em%
+  \leavevmode%
+  \begin{beamercolorbox}[dp=1ex,center]{}%
+      \usebeamercolor[fg]{item}{\textbf{{\Large \insertframetitle}}}
+  \end{beamercolorbox}%
+}
+
+\setbeamertemplate{navigation symbols}{}
+
+\newcounter{framenumberpreappendix}
+\newcommand{\backupstart}{
+  \setcounter{framenumberpreappendix}{\value{framenumber}}
+}
+\newcommand{\backupend}{
+  \addtocounter{framenumberpreappendix}{-\value{framenumber}}
+  \addtocounter{framenumber}{\value{framenumberpreappendix}} 
+}
+
+\setbeamertemplate{footline}{%
+  \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex,dp=1ex,right,rightskip=1mm,leftskip=1mm]{}%
+    \inserttitle \hfill
+    \insertframenumber\,/\,\inserttotalframenumber
+  \end{beamercolorbox}%
+  \vskip0pt%
+}
+
+\newcommand{\hil}[1]{{\usebeamercolor[fg]{item}{\textbf{#1}}}}
+
+\IfSubStr{\jobname}{\detokenize{nonotes}}{
+  \setbeameroption{hide notes}
+}{
+  \setbeameroption{show notes}
+}
+\setbeamertemplate{note page}[plain]
+\setbeameroption{hide notes}
+
+\begin{document}
+
+\addtocounter{framenumber}{-1}
+
+\begin{frame}[c]
+  \centering
+  \includegraphics[scale=0.3]{images/external-internal-small}
+  \medskip
+
+  \hil{Synthetic algebraic geometry} \\
+  \small\emph{a case study in applied topos theory}
+
+  \large\&
+
+  \normalsize
+
+  \hil{the phenomenon of nongeometric sequents}
+
+  \medskip
+
+  \scriptsize
+  Ingo Blechschmidt \\
+  University of Augsburg
+  \medskip
+
+  103rd Peripatetic Seminar on Sheaves and Logic (Brno) \\
+  April 7th, 2018
+  \par
+\end{frame}
+
+\section{Preliminaries}
+
+\subsection{Synthetic differential geometry}
+
+\begin{frame}{Synthetic differential geometry}
+  \vspace*{-1em}
+  \begin{varblock}{9.5cm}{The axiom of microaffinity}
+    \justifying
+    Let~$\Delta = \{ \varepsilon \in \RR \,|\, \varepsilon^2 = 0 \}$.
+    For any function $f : \Delta \to \RR$,
+    there are unique numbers~$a, b \in \RR$ such that
+    $f(\varepsilon) = a + b \varepsilon$
+    for all~$\varepsilon \in \Delta$.
+  \end{varblock}
+
+  \begin{center}
+    \includegraphics[width=0.5\textwidth]{microaffinity}
+  \end{center}
+
+  \vspace*{-1em}
+  \begin{itemize}
+    \item The \hil{derivative} of~$f$ as above at zero is~$b$.
+    \item Manifolds are \hil{just sets}.
+    \item A \hil{tangent vector} to~$M$ is a map~$\Delta \to M$.
+  \end{itemize}
+
+  \centering
+  \hil{Toposes provide models for this theory.}
+\end{frame}
+
+\subsection{Toposes}
+
+\begin{frame}{The internal universe of a topos}
+  A \hil{topos} is a category which has finite limits,
+  is cartesian closed and has a subobject classifier, for instance
+  \begin{itemize}
+    \item \hil{$\boldsymbol{\Set}$}, the category of sets;
+    \item \hil{$\boldsymbol{\Sh(X)}$}, the category of set-valued sheaves over
+    a space~$X$;
+    \item \hil{$\boldsymbol{\Eff}$}, the ef{}fective topos (roughly: a
+    category of data types).
+  \end{itemize}
+
+  \slogan{Any topos supports an \hil{internal language},
+  which is sound with respect to
+  \only<1>{\hil{intuitionistic reasoning}}%
+  \only<2->{\hil{\pointthis{intuitionistic reasoning}{
+    \textnormal{\textcolor{black}{
+    no $\varphi \vee \neg\varphi$,\ \
+    no $\neg\neg\varphi \Rightarrow \varphi$,\ \
+    no axiom of choice}}}}}.}
+\end{frame}
+
+\newcommand{\intex}[3]{#1 \qquad #2\par #3\medskip}
+
+\begin{frame}{Curious universes}
+  \small
+  \intex{
+    $\Eff \models \text{``There are infinitely many prime numbers.''}$
+  }{\textcolor{green!90}{\cmark}}{
+    External meaning: There is a (Turing) \hil{machine} producing arbitrarily many prime
+    numbers.
+  }
+
+  \intex{
+    $\Eff \not\models \text{``Any Turing machine halts or doesn't halt.''}$
+  }{\textcolor{red!80}{\xmark}}{
+    External meaning: There is a \hil{halting oracle} which determines whether
+    any given machine halts or doesn't halt.
+  }
+
+  \intex{
+    $\Sh(X) \not\models \text{``Any cont. function with opposite signs has
+    a zero.''}$
+  }{\textcolor{red!80}{\xmark}}{
+    External meaning: Zeros can be picked \hil{locally continuously} in
+    continuous families of continuous functions.
+  }
+\end{frame}
+
+\end{document}
+
 "Any map from the reals to the reals is smooth." This statement holds in the

### 2018-04-05 slides-pssl103.tex
Video illustrating the failure of the IVT
Slides mostly finished

@@
 \hypersetup{colorlinks=true}
+\usepackage{multimedia}
+\newcommand{\video}[2]{\movie[width=#2,height=#2,autostart,loop,poster]{}{#1}}
 
@@
 \newcommand{\RR}{\mathbb{R}}
+\newcommand{\TT}{\mathbb{T}}
 \newcommand{\PP}{\mathbb{P}}
@@
 \newcommand{\Hom}{\mathrm{Hom}}
-\DeclareMathOperator{\Spec}{Spec}
+\newcommand{\Spec}{\mathrm{Spec}}
 \newcommand{\lra}{\longrightarrow}
@@
 \newcommand{\Ll}{\vcentcolon\!\Longleftrightarrow}
+\newcommand{\inv}{inv.\@}
+\newcommand{\seq}{\vdash_{\!\!\!\vec x}}
 
@@
     (#1) {#1};\node[overlay,rectangle callout,%
-    callout relative pointer={(-0.5cm,1.0cm)},fill=blue!20] at ($(#1.north)+(0.9cm,-1.6cm)$) {#2};}%
+    callout relative pointer={(0.3cm,1.0cm)},fill=blue!20] at ($(#1.north)+(-0.1cm,-1.6cm)$) {#2};}%
 }%
@@
 
-  \slogan{Any topos supports an \hil{internal language},
+  \begin{varblock}{9cm}{}Any topos supports an \hil{internal language},
   which is sound with respect to
@@
     no $\neg\neg\varphi \Rightarrow \varphi$,\ \
-    no axiom of choice}}}}}.}
+    no axiom of choice}}}}}.\end{varblock}
 \end{frame}
 
-\newcommand{\intex}[3]{#1 \qquad #2\par #3\medskip}
+\newcommand{\intex}[3]{#1 \quad #2\par #3\medskip}
 
@@
   }{\textcolor{green!90}{\cmark}}{
-    External meaning: There is a (Turing) \hil{machine} producing arbitrarily many prime
+    External meaning: There is a Turing \hil{machine} producing arbitrarily many prime
     numbers.
@@
   }
+
+  \vspace*{-1em}
+  \hfill\video{zeros-in-families.mp4}{2cm}
+\end{frame}
+
+\section{Synthetic algebraic geometry}
+
+\begin{frame}{Approaches to algebraic geometry}
+  Usual approach to algebraic geometry: \hil{layer schemes above ordinary set theory}
+  using either
+  \begin{itemize}
+    \item locally ringed spaces
+    \small
+    \begin{multline*}
+      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$} + {} \\
+      \text{Zariski topology} + \text{structure sheaf}
+    \end{multline*}
+    \normalsize
+    \item or Grothendieck's functor-of-points account, where a scheme is a functor~$\mathrm{Ring} \to \mathrm{Set}$.
+    \small\[ A \longmapsto \{ (x,y,z) \in A^3 \,|\, x^n+y^n-z^n=0 \} \]
+  \end{itemize}
+  \bigskip
+  \pause
+
+  \hil{Synthetic approach:} model schemes \hil{directly as sets} in a certain
+  nonclassical set theory.
+  \small
+  \[ \{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]
+\end{frame}
+
+\begin{frame}{The big Zariski topos}
+  Let~$S$ be a fixed base scheme.
+
+  \begin{block}{\centering Definition}\small
+  The \hil{big Zariski topos} $\Zar(S)$ is the
+  category $\Sh(\Sch/S)$. It consists of functors
+  $(\Sch/S)^\op \to \Set$ satisfying the gluing condition that
+  \[ F(U) \to
+    \prod_i F(U_i) \rightrightarrows
+    \prod_{j,k} F(U_j \cap U_k) \]
+  \\[-0.6em]
+  is a limit diagram for any scheme~$U = \bigcup_i U_i$ over~$S$.
+  \end{block}
+
+  \begin{itemize}
+    \item For an~$S$-scheme~$X$, its functor of points $\ull{X} =
+    \Hom_S(\cdot,X)$ is an object of~$\Zar(S)$. It feels like \hil{the set of
+    points} of~$X$.
+    \item In particular, there is the ring object $\affl$.
+    \item $\Zar(S)$ classifies local~$\O_S$-algebras which are local
+    over~$\O_S$.
+  \end{itemize}
+\end{frame}
+
+\begin{frame}{Properties of the affine line}
+  \begin{itemize}
+    \item $\affl$ is a local ring:
+    \[ 1 \neq 0 \qquad\qquad \text{$x + y$ \inv} \Longrightarrow \text{$x$ \inv} \vee
+    \text{$y$ \inv} \]
+    \item $\affl$ is a field:
+    \begin{align*}
+      \neg(x = 0) &\Longleftrightarrow \text{$x$ \inv} \\
+      \neg(\text{$x$ \inv}) &\Longleftrightarrow \text{$x$ nilpotent}
+    \end{align*}
+    \item $\affl$ satisfies the axiom of microaffinity.
+    \item $\affl$ is anonymously algebraically closed: Any monic polynomial
+    does \emph{not not} have a zero.
+    \item $\affl$ is of unbounded Krull dimension.
+    \item Any function $\affl \to \affl$ is a polynomial.
+  \end{itemize}
+\end{frame}
+
+\begin{frame}{Synthetic constructions}
+  \small
+  $\hil{$\boldsymbol{\mathbb{A}^n}$} = (\affl)^n = \affl \times \cdots \times \affl$
+
+  $\begin{array}{@{}r@{}c@{}l@{}}
+    \hil{$\boldsymbol{\mathbb{P}^n}$} &\phantom{}=\phantom{}& \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee
+    \cdots \vee x_n \neq 0 \}/(\affl)^\times \\
+    &\cong& \text{set of one-dimensional subspaces of~$(\affl)^{n+1}$} \\
+    &&\qquad \text{(with~$\O(-1) = (\ell)_{\ell \? \mathbb{P}^n}$, $\O(1) =
+    (\ell^\vee)_{\ell \? \mathbb{P}^n}$)}
+  \end{array}$
+
+  $\hil{$\boldsymbol{\Spec(R)}$} = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
+    \text{set of $\affl$-valued points of $R$}$
+
+  $\hil{$\boldsymbol{TX}$} = \Hom(\Delta, X)$, where $\Delta = \{ \varepsilon \? \affl \,|\, \varepsilon^2 = 0 \}$
+
+  A subset $U \subseteq X$ is \hil{qc-open} if and only if for any $x : X$
+  there exist $f_1,\ldots,f_n \? \affl$ such that $x \in U \Longleftrightarrow
+  \exists i\_ f_i \neq 0$.
+
+  A \hil{synthetic affine scheme} is a set which is in bijection
+  with~$\Spec(R)$ for some finitely presented algebra~$R$.
+
+  A \hil{synthetic scheme} is a set which can be covered by finitely many
+  qc-open synthetic affine schemes~$U_i$ such that the intersections~$U_i \cap
+  U_j$ can be covered by finitely many qc-open synthetic affine schemes.
+\end{frame}
+
+\begin{frame}{Synthetic quasicoherence}
+  Recall~$\Spec(R) = \Hom_{\mathrm{Alg}(\affl)}(R, \affl)$ and consider the statement
+  \[ \text{``the canonical map~$
+    \begin{array}[t]{rcl}
+      R &\longrightarrow& \Hom(\Spec(R), \affl) \\
+      f &\longmapsto& (\alpha \mapsto \alpha(f))
+    \end{array}
+  $ is bijective''}.
+  \]
+  \vspace*{-1em}
+
+  \begin{itemize}
+    \item True for~$R = \affl[X]/(X^2)$ (microaffinity).
+    \item True for~$R = \affl[X]$ (every function is a polynomial).
+    \item True for \hil{any} finitely presented~$\affl$-algebra~$R$.
+  \end{itemize}
+
+  Any known property of~$\affl$ follows from this \hil{synthetic
+  quasicoherence}.
+
+  \emph{Example.} Let~$f \? \affl$ such that~$f \neq 0$. Set~$R = \affl/(f)$.
+  Then~$\Spec(R) = \emptyset$. Thus~$\Hom(\Spec(R),\affl)$ is a singleton.
+  Hence~$R = 0$. Therefore~$f$ is invertible.
+\end{frame}
+
+\section{Nongeometric sequents}
+
+\begin{frame}{Nongeometric sequents}
+  \small
+  Let~$\TT$ be a \hil{geometric theory} (rings, intervals, \ldots).
+
+  \vspace*{-2em}
+  \begin{varblock}{11cm}{}  %Fundamental theorem about classifying toposes}
+    For a \hil{geometric sequent}~$\varphi \seq \psi$, the following
+    are equivalent:
+    \begin{enumerate}
+      \item The theory~$\TT$ \hil{proves}~$\varphi \seq \psi$. \\[-1em]
+      \item $\forall \vec x\_ (\varphi \Rightarrow \psi)$ holds
+      \hil{for all models} of~$\TT$ in all toposes. \\[-1em]
+      \item $\forall \vec x\_ (\varphi \Rightarrow \psi)$ holds
+      for the \hil{generic model} of~$\TT$ in its \hil{classifying topos}.
+    \end{enumerate}
+  \end{varblock}
+
+  \begin{itemize}
+    \item Additional \hil{nongeometric sequents} may hold in a classifying
+    topos, for instance~``$\affl$ is synthetically quasicoherent''
+    in~$\Zar(S)$. \\[-1em]
+    \item These are \hil{$\boldsymbol{\TT}$-redundant}, but the converse is
+    false. \\[-1em]  % (Bezem--Buchholtz--Coquand 2017). \\[-1em]
+    \item \ \\[-1.2em]\mbox{Are they precisely the consequences of synthetic
+    quasicoherence?} \\[-1em]
+    \item Applications: synthetic algebraic geometry,
+    generic freeness, \ldots
+  \end{itemize}
 \end{frame}

### 2018-04-05 slides-pssl103.tex
Small layout changes

@@
 \newcommand{\video}[2]{\movie[width=#2,height=#2,autostart,loop,poster]{}{#1}}
+\hypersetup{colorlinks=false}
 
@@
   \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex,dp=1ex,right,rightskip=1mm,leftskip=1mm]{}%
-    \inserttitle \hfill
+    \hfill
     \insertframenumber\,/\,\inserttotalframenumber
@@
   \centering
+  \medskip
   \includegraphics[scale=0.3]{images/external-internal-small}
@@
 
-  103rd Peripatetic Seminar on Sheaves and Logic (Brno) \\
+  103rd Peripatetic Seminar on Sheaves and Logic in Brno \\
   April 7th, 2018
@@
 
-\newcommand{\intex}[3]{#1 \quad #2\par #3\medskip}
+\newcommand{\intex}[3]{#1 \quad #2\par #3\bigskip\medskip}
 
 \begin{frame}{Curious universes}
+  \begin{changemargin}{-1.1em}{0em}
     \small
-  \intex{
+    \begin{itemize}
+      \item \intex{
         $\Eff \models \text{``There are infinitely many prime numbers.''}$
@@
 
-  \intex{
+      \item \intex{
         $\Eff \not\models \text{``Any Turing machine halts or doesn't halt.''}$
@@
 
-  \intex{
+      \item \intex{
         $\Sh(X) \not\models \text{``Any cont. function with opposite signs has
@@
       }
+    \end{itemize}
+  \end{changemargin}
 
-  \vspace*{-1em}
+  \vspace*{-2em}
   \hfill\video{zeros-in-families.mp4}{2cm}
@@
   \begin{varblock}{11cm}{}  %Fundamental theorem about classifying toposes}
-    For a \hil{geometric sequent}~$\varphi \seq \psi$, the following
+    For a \hil{geometric sequent}~$\forall \vec x\_ (\varphi \Rightarrow \psi)$, the following
     are equivalent:
     \begin{enumerate}
-      \item The theory~$\TT$ \hil{proves}~$\varphi \seq \psi$. \\[-1em]
-      \item $\forall \vec x\_ (\varphi \Rightarrow \psi)$ holds
+      \item It is \hil{provable} by~$\TT$. \\[-1em]
+      \item It holds
       \hil{for all models} of~$\TT$ in all toposes. \\[-1em]
-      \item $\forall \vec x\_ (\varphi \Rightarrow \psi)$ holds
+      \item It holds
       for the \hil{generic model} of~$\TT$ in its \hil{classifying topos}.

### 2018-04-06 slides-pssl103.tex
Minor improvements

@@
       }{\textcolor{green!90}{\cmark}}{
-        External meaning: There is a Turing \hil{machine} producing arbitrarily many prime
+        External meaning: There is a \hil{Turing machine} producing arbitrarily many prime
         numbers.
@@
       }{\textcolor{red!80}{\xmark}}{
-        External meaning: Zeros can be picked \hil{locally continuously} in
+        External meaning: Zeros can locally be picked \hil{continuously} in
         continuous families of continuous functions.
@@
 
+  \begin{changemargin}{-1.0em}{-0.5em}
   \begin{itemize}
@@
     \Hom_S(\cdot,X)$ is an object of~$\Zar(S)$. It feels like \hil{the set of
-    points} of~$X$.
-    \item In particular, there is the ring object $\affl$.
+    points} of~$X$. \\[0.2em]
+    \item \ \\[-1.2em]\mbox{In particular, there is the ring object $\affl$ with~$\affl(U) =
+    \O_U(U)$.} \\[0.2em]
     \item $\Zar(S)$ classifies local~$\O_S$-algebras which are local
@@
   \end{itemize}
+  \end{changemargin}
+\end{frame}
+
+\begin{frame}{Synthetic constructions}
+  \small
+  $\hil{$\boldsymbol{\mathbb{A}^n}$} = (\affl)^n = \affl \times \cdots \times \affl$
+
+  $\begin{array}{@{}r@{}c@{}l@{}}
+    \hil{$\boldsymbol{\mathbb{P}^n}$} &\phantom{}=\phantom{}& \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee
+    \cdots \vee x_n \neq 0 \}/(\affl)^\times \\
+    &\cong& \text{set of one-dimensional subspaces of~$(\affl)^{n+1}$} \\
+    &&\qquad \text{(with~$\O(-1) = (\ell)_{\ell \? \mathbb{P}^n}$, $\O(1) =
+    (\ell^\vee)_{\ell \? \mathbb{P}^n}$)}
+  \end{array}$
+
+  $\hil{$\boldsymbol{\Spec(R)}$} = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
+    \text{set of $\affl$-valued points of $R$}$
+
+  $\hil{$\boldsymbol{TX}$} = X^\Delta$, where $\Delta = \{ \varepsilon \? \affl \,|\, \varepsilon^2 = 0 \}$
+
+  A subset $U \subseteq X$ is \hil{qc-open} if and only if for any $x : X$
+  there exist $f_1,\ldots,f_n \? \affl$ such that $x \in U \Longleftrightarrow
+  \exists i\_ f_i \neq 0$.
+
+  A \hil{synthetic affine scheme} is a set which is in bijection
+  with~$\Spec(R)$ for some finitely presented algebra~$R$.
+
+  A \hil{synthetic scheme} is a set which can be covered by finitely many
+  qc-open synthetic affine schemes~$U_i$ such that the intersections~$U_i \cap
+  U_j$ can be covered by finitely many qc-open synthetic affine schemes.
 \end{frame}
@@
 
-\begin{frame}{Synthetic constructions}
-  \small
-  $\hil{$\boldsymbol{\mathbb{A}^n}$} = (\affl)^n = \affl \times \cdots \times \affl$
-
-  $\begin{array}{@{}r@{}c@{}l@{}}
-    \hil{$\boldsymbol{\mathbb{P}^n}$} &\phantom{}=\phantom{}& \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee
-    \cdots \vee x_n \neq 0 \}/(\affl)^\times \\
-    &\cong& \text{set of one-dimensional subspaces of~$(\affl)^{n+1}$} \\
-    &&\qquad \text{(with~$\O(-1) = (\ell)_{\ell \? \mathbb{P}^n}$, $\O(1) =
-    (\ell^\vee)_{\ell \? \mathbb{P}^n}$)}
-  \end{array}$
-
-  $\hil{$\boldsymbol{\Spec(R)}$} = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
-    \text{set of $\affl$-valued points of $R$}$
-
-  $\hil{$\boldsymbol{TX}$} = \Hom(\Delta, X)$, where $\Delta = \{ \varepsilon \? \affl \,|\, \varepsilon^2 = 0 \}$
-
-  A subset $U \subseteq X$ is \hil{qc-open} if and only if for any $x : X$
-  there exist $f_1,\ldots,f_n \? \affl$ such that $x \in U \Longleftrightarrow
-  \exists i\_ f_i \neq 0$.
-
-  A \hil{synthetic affine scheme} is a set which is in bijection
-  with~$\Spec(R)$ for some finitely presented algebra~$R$.
-
-  A \hil{synthetic scheme} is a set which can be covered by finitely many
-  qc-open synthetic affine schemes~$U_i$ such that the intersections~$U_i \cap
-  U_j$ can be covered by finitely many qc-open synthetic affine schemes.
-\end{frame}
-
 \begin{frame}{Synthetic quasicoherence}
@@
     \begin{array}[t]{rcl}
-      R &\longrightarrow& \Hom(\Spec(R), \affl) \\
+      R &\longrightarrow& (\affl)^{\Spec(R)} \\
       f &\longmapsto& (\alpha \mapsto \alpha(f))
@@
 
-  \emph{Example.} Let~$f \? \affl$ such that~$f \neq 0$. Set~$R = \affl/(f)$.
-  Then~$\Spec(R) = \emptyset$. Thus~$\Hom(\Spec(R),\affl)$ is a singleton.
-  Hence~$R = 0$. Therefore~$f$ is invertible.
+  \emph{Example.} Let~$x \? \affl$ such that~$x \neq 0$. Set~$R = \affl/(x)$.
+  Then~$\Spec(R) = \emptyset$. Thus~$(\affl)^{\Spec(R)}$ is a singleton.
+  Hence~$R = 0$. Therefore~$x$ is invertible.
 \end{frame}

### 2018-04-06 slides-pssl103.tex
Some reordering

@@
 
-\subsection{Synthetic differential geometry}
+\subsection{Approaches to algebraic geometry}
 
-\begin{frame}{Synthetic differential geometry}
-  \vspace*{-1em}
-  \begin{varblock}{9.5cm}{The axiom of microaffinity}
-    \justifying
-    Let~$\Delta = \{ \varepsilon \in \RR \,|\, \varepsilon^2 = 0 \}$.
-    For any function $f : \Delta \to \RR$,
-    there are unique numbers~$a, b \in \RR$ such that
-    $f(\varepsilon) = a + b \varepsilon$
-    for all~$\varepsilon \in \Delta$.
-  \end{varblock}
+\begin{frame}{Approaches to algebraic geometry}
+  Usual approach to algebraic geometry: \hil{layer schemes above ordinary set theory}
+  using either
+  \begin{itemize}
+    \item locally ringed spaces
+    \small
+    \begin{multline*}
+      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$} + {} \\
+      \text{Zariski topology} + \text{structure sheaf}
+    \end{multline*}
+    \normalsize
+    \item or Grothendieck's functor-of-points account, where a scheme is a functor~$\mathrm{Ring} \to \mathrm{Set}$.
+    \small\[ A \longmapsto \{ (x,y,z) \in A^3 \,|\, x^n+y^n-z^n=0 \} \]
+  \end{itemize}
+  \bigskip
 
-  \begin{center}
-    \includegraphics[width=0.5\textwidth]{microaffinity}
-  \end{center}
-
-  \vspace*{-1em}
-  \begin{itemize}
-    \item The \hil{derivative} of~$f$ as above at zero is~$b$.
-    \item Manifolds are \hil{just sets}.
-    \item A \hil{tangent vector} to~$M$ is a map~$\Delta \to M$.
-  \end{itemize}
-
-  \centering
-  \hil{Toposes provide models for this theory.}
+  \hil{Synthetic approach:} model schemes \hil{directly as sets} in a certain
+  nonclassical set theory.
+  \small
+  \[ \{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]
 \end{frame}
 
-\subsection{Toposes}
+\subsection{Toposes as universes}
 
-\begin{frame}{The internal universe of a topos}
+\begin{frame}{Toposes as mathematical universes}
   A \hil{topos} is a category which has finite limits,
@@
       \item \intex{
-        $\Eff \not\models \text{``Any Turing machine halts or doesn't halt.''}$
+        $\Eff \models \text{``Any Turing machine halts or doesn't halt.''}$
       }{\textcolor{red!80}{\xmark}}{
@@
       \item \intex{
-        $\Sh(X) \not\models \text{``Any cont. function with opposite signs has
+        $\Sh(X) \models \text{``Any cont. function with opposite signs has
         a zero.''}$
@@
 
+\begin{frame}{Synthetic differential geometry}
+  \vspace*{-1em}
+  \begin{varblock}{9.5cm}{The axiom of microaffinity}
+    \justifying
+    Let~$\Delta = \{ \varepsilon \in \RR \,|\, \varepsilon^2 = 0 \}$.
+    For any function $f : \Delta \to \RR$,
+    there are unique numbers~$a, b \in \RR$ such that
+    $f(\varepsilon) = a + b \varepsilon$
+    for all~$\varepsilon \in \Delta$.
+  \end{varblock}
+
+  \begin{center}
+    \includegraphics[width=0.5\textwidth]{microaffinity}
+  \end{center}
+
+  \vspace*{-1em}
+  \begin{itemize}
+    \item The \hil{derivative} of~$f$ as above at zero is~$b$.
+    \item Manifolds are \hil{just sets}.
+    \item A \hil{tangent vector} to~$M$ is a map~$\Delta \to M$.
+  \end{itemize}
+
+  \centering
+  \hil{Toposes provide models for this theory.}
+\end{frame}
+
 \section{Synthetic algebraic geometry}
 
-\begin{frame}{Approaches to algebraic geometry}
-  Usual approach to algebraic geometry: \hil{layer schemes above ordinary set theory}
-  using either
-  \begin{itemize}
-    \item locally ringed spaces
-    \small
-    \begin{multline*}
-      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$} + {} \\
-      \text{Zariski topology} + \text{structure sheaf}
-    \end{multline*}
-    \normalsize
-    \item or Grothendieck's functor-of-points account, where a scheme is a functor~$\mathrm{Ring} \to \mathrm{Set}$.
-    \small\[ A \longmapsto \{ (x,y,z) \in A^3 \,|\, x^n+y^n-z^n=0 \} \]
-  \end{itemize}
-  \bigskip
-  \pause
-
-  \hil{Synthetic approach:} model schemes \hil{directly as sets} in a certain
-  nonclassical set theory.
-  \small
-  \[ \{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]
-\end{frame}
+\subsection{The big Zariski topos}
 
@@
 \end{frame}
+
+\subsection{Synthetic constructions}
 
@@
 
+\subsection[Quasicoherence]{Synthetic quasicoherence}
+
 \begin{frame}{Properties of the affine line}
@@
     \begin{align*}
-      \neg(x = 0) &\Longleftrightarrow \text{$x$ \inv} \\
-      \neg(\text{$x$ \inv}) &\Longleftrightarrow \text{$x$ nilpotent}
+      \neg(x = 0) &\Longleftrightarrow \text{$x$ invertible} \\
+      \neg(\text{$x$ invertible}) &\Longleftrightarrow \text{$x$ nilpotent}
     \end{align*}
-    \item $\affl$ satisfies the axiom of microaffinity.
+    \item $\affl$ satisfies the axiom of microaffinity: Any map $f : \Delta \to
+    \affl$ is of the form~$f(\varepsilon) = a + b \varepsilon$ for unique
+    values~$a,b \? \affl$, where~$\Delta = \{ \varepsilon \? \affl \,|\,
+    \varepsilon^2 = 0 \}$.
+    \item Any function $\affl \to \affl$ is a polynomial.
     \item $\affl$ is anonymously algebraically closed: Any monic polynomial
@@
     \item $\affl$ is of unbounded Krull dimension.
-    \item Any function $\affl \to \affl$ is a polynomial.
   \end{itemize}

### 2018-04-07 slides-pssl103.tex
Final changes

@@
   $(\Sch/S)^\op \to \Set$ satisfying the gluing condition that
-  \[ F(U) \to
+  \[ F(T) \to
     \prod_i F(U_i) \rightrightarrows
@@
   \\[-0.6em]
-  is a limit diagram for any scheme~$U = \bigcup_i U_i$ over~$S$.
+  is a limit diagram for any scheme~$T = \bigcup_i U_i$ over~$S$.
   \end{block}
@@
     points} of~$X$. \\[0.2em]
-    \item \ \\[-1.2em]\mbox{In particular, there is the ring object $\affl$ with~$\affl(U) =
-    \O_U(U)$.} \\[0.2em]
+    \item \ \\[-1.2em]\mbox{In particular, there is the ring object $\affl$ with~$\affl(T) =
+    \O_T(T)$.} \\[0.2em]
     \item $\Zar(S)$ classifies local~$\O_S$-algebras which are local
@@
 
+\section{Further research}
+
+\begin{frame}{Further research}
+  \vspace*{-1em}
+  \small
+  \begin{itemize}
+    \item Push synthetic algebraic geometry further:
+          true cohomology, intersection theory, derived categories, \ldots
+          \\[-1.9em]
+
+    \item What do the various subtoposes of~$\Zar(S)$ classify
+          (étale, fppf, ph, $\neg\neg$, \ldots)? What about the crystalline
+          topos?
+          \\[-1.9em]
+
+    \item Understand quasicoherence.
+          \\[-1.9em]
+
+    \item Find further applications of nongeometric sequents, for instance in
+    constructive algebra.
+  \end{itemize}
+
+  {\vspace{-0.1em}\centering
+  \rotatebox{90}{\tiny\scalebox{0.5}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
+  \includegraphics[scale=0.18]{images/external-internal-small}
+  \par\medskip\vspace{-0.5em}}
+
+  \centering
+  Expository notes: \\
+  \hil{\href{https://www.ingo-blechschmidt.eu/}{https://www.ingo-blechschmidt.eu/}}
+  \par
+\end{frame}
+
+\addtocounter{framenumber}{-1}
+
 \end{document}

### 2018-04-09 slides-pssl103.tex
Minor changes after the talk

@@
   \vspace*{-2em}
-  \hfill\video{zeros-in-families.mp4}{2cm}
+  \hfill\video{zeros-in-families.mp4}{2cm} \\
+  \hfill \scriptsize
+  (see
+  \textcolor{red}{\href{https://rawgit.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{video}}
+  for counterexample)
 \end{frame}
@@
   A \hil{synthetic affine scheme} is a set which is in bijection
-  with~$\Spec(R)$ for some finitely presented algebra~$R$.
+  with~$\Spec(R)$ for some synthetically quasicoherent~$\affl$-algebra~$R$.
 
@@
     \begin{align*}
-      \neg(x = 0) &\Longleftrightarrow \text{$x$ invertible} \\
+      \neg(x = 0) &\Longleftrightarrow \text{$x$ invertible \quad [Kock 1976]} \\
       \neg(\text{$x$ invertible}) &\Longleftrightarrow \text{$x$ nilpotent}
@@
 
-  Any known property of~$\affl$ follows from this \hil{synthetic
-  quasicoherence}.
+  Any known property of~$\affl$ follows from this \\
+  \hil{synthetic quasicoherence}.
 
@@
     \item These are \hil{$\boldsymbol{\TT}$-redundant}, but the converse is
-    false. \\[-1em]  % (Bezem--Buchholtz--Coquand 2017). \\[-1em]
+    false. [Bezem--Buchholtz--Coquand 2017; answering a question by Wraith
+    possibly raised at PSSL 1.] \\[-1em]
     \item \ \\[-1.2em]\mbox{Are they precisely the consequences of synthetic

### 2018-04-10 slides-munich2018.tex
Slides for tomorrow's talk in Munich

\documentclass[12pt,utf8,notheorems,compress,t]{beamer}
\usepackage{etex}

\usepackage[english]{babel}

\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{soul}\setul{0.3ex}{}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes.callouts,shapes.arrows,patterns}
\hypersetup{colorlinks=true}
\usepackage{multimedia}
\newcommand{\video}[2]{\movie[width=#2,height=#2,autostart,loop,poster]{}{#1}}
\hypersetup{colorlinks=false}

\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

\graphicspath{{images/}}

\usepackage[protrusion=true,expansion=true]{microtype}

\setlength\parskip{\medskipamount}
\setlength\parindent{0pt}

\title{Without loss of generality, any reduced ring is a field}
\author{Ingo Blechschmidt}
\date{April 11th, 2018}

\useinnertheme[shadow=true]{rounded}
\useoutertheme{split}
\usecolortheme{orchid}
\usecolortheme{whale}
\setbeamerfont{block title}{size={}}

\useinnertheme{rectangles}

\usecolortheme{seahorse}
\definecolor{mypurple}{RGB}{150,0,255}
\setbeamercolor{structure}{fg=mypurple}
\definecolor{myred}{RGB}{150,0,0}
\setbeamercolor*{title}{bg=myred,fg=white}
\setbeamercolor*{titlelike}{bg=myred,fg=white}

\usefonttheme{serif}
\usepackage[T1]{fontenc}
\usepackage{libertine}

\newcommand{\A}{\mathcal{A}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ull}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
\newcommand{\Ll}{\vcentcolon\!\Longleftrightarrow}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}{\vdash_{\!\!\!\vec x}}

\setbeamertemplate{blocks}[rounded][shadow=false]

\newcommand{\fmini}[2]{\framebox{\begin{minipage}{#1}#2\end{minipage}}}
\makeatletter
\def\underunbrace#1{\mathop{\vtop{\m@th\ialign{##\crcr
      $\hfil\displaystyle{#1}\hfil$\crcr\noalign{\kern3\p@\nointerlineskip}
      \crcr\noalign{\kern3\p@}}}}\limits}
\def\overunbrace#1{\mathop{\vbox{\m@th\ialign{##\crcr\noalign{\kern3\p@}
      \crcr\noalign{\kern3\p@\nointerlineskip}
      $\hfil\displaystyle{#1}\hfil$\crcr}}}\limits}
\makeatother

\newenvironment{changemargin}[2]{%
  \begin{list}{}{%
    \setlength{\topsep}{0pt}%
    \setlength{\leftmargin}{#1}%
    \setlength{\rightmargin}{#2}%
    \setlength{\listparindent}{\parindent}%
    \setlength{\itemindent}{\parindent}%
    \setlength{\parsep}{\parskip}%
  }%
  \item[]}{\end{list}}

\newcommand{\pointthis}[2]{%
  \tikz[remember picture,baseline]{\node[anchor=base,inner sep=0,outer sep=0]%
    (#1) {#1};\node[overlay,rectangle callout,%
    callout relative pointer={(0.3cm,1.0cm)},fill=blue!20] at ($(#1.north)+(-0.1cm,-1.6cm)$) {#2};}%
}%

\newcommand{\hcancel}[5]{%
  \tikz[baseline=(tocancel.base)]{
    \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
    \draw[red, line width=0.4mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
  }%
}
% Adapted from https://latex.org/forum/viewtopic.php?t=2251 (Stefan Kottwitz)
\newenvironment<>{varblock}[2]{\begin{varblockextra}{#1}{#2}{}}{\end{varblockextra}}
\newenvironment<>{varblockextra}[3]{
  \begin{center}
    \begin{minipage}{#1}
      \begin{actionenv}#4
        {\centering \hil{#2}\par}
	\def\insertblocktitle{}%\centering #2}
        \def\varblockextraend{#3}
	\usebeamertemplate{block begin}}{
        \par
        \usebeamertemplate{block end}
        \varblockextraend
      \end{actionenv}
    \end{minipage}
  \end{center}}

\setbeamertemplate{frametitle}{%
  \vskip0.7em%
  \leavevmode%
  \begin{beamercolorbox}[dp=1ex,center]{}%
      \usebeamercolor[fg]{item}{\textbf{{\Large \insertframetitle}}}
  \end{beamercolorbox}%
}

\setbeamertemplate{navigation symbols}{}

\newcounter{framenumberpreappendix}
\newcommand{\backupstart}{
  \setcounter{framenumberpreappendix}{\value{framenumber}}
}
\newcommand{\backupend}{
  \addtocounter{framenumberpreappendix}{-\value{framenumber}}
  \addtocounter{framenumber}{\value{framenumberpreappendix}} 
}

\setbeamertemplate{footline}{%
  \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex,dp=1ex,right,rightskip=1mm,leftskip=1mm]{}%
    \hfill
    \insertframenumber\,/\,\inserttotalframenumber
  \end{beamercolorbox}%
  \vskip0pt%
}

\newcommand{\hil}[1]{{\usebeamercolor[fg]{item}{\textbf{#1}}}}

\IfSubStr{\jobname}{\detokenize{nonotes}}{
  \setbeameroption{hide notes}
}{
  \setbeameroption{show notes}
}
\setbeamertemplate{note page}[plain]
\setbeameroption{hide notes}

\begin{document}

\addtocounter{framenumber}{-1}

\begin{frame}[c]
  \centering
  \bigskip
  \bigskip
  \bigskip
  \bigskip

  \large
  \hil{Without loss of generality, \\ any reduced ring is a field.}

  \bigskip

  \emph{\small Interruptions welcome at any point.}

  \bigskip

  \scriptsize
  Ingo Blechschmidt \\
  University of Augsburg
  \bigskip

  Oberseminar Mathematische Logik \\
  Ludwig-Maximilians-Universität München \\
  April 11th, 2018
  \par
\end{frame}

\section{Introduction}

\begin{frame}{Summary}
  \begin{itemize}
    \item For any reduced ring~$A$, there is a semantics with
    \[ A \models \bigl(\forall x\_ \neg(\exists y\_ xy = 1) \Rightarrow x = 0\bigr). \]

    \item This semantics is sound with respect to intuitionistic logic.

    \item \ \\[-1.2em]\mbox{It has uses in classical and constructive commutative
    algebra.}
  \end{itemize}

  \pause
  \vspace*{-0.7em}

  \begin{columns}[t]
    \begin{column}[t]{0.47\textwidth}
      \centering

      \begin{varblock}{\textwidth}{A baby application}
        \justifying
        Let~$M$ be a surjective matrix with more rows than columns over a ring~$A$.
        Then~$A = 0$.
      \end{varblock}

      \scalebox{0.8}{$\begin{pmatrix}
        \cdot & \cdot \\
        \cdot & \cdot \\
        \cdot & \cdot
      \end{pmatrix}$}
    \end{column}

    \begin{column}[t]{0.47\textwidth}
      \centering

      \begin{varblock}{\textwidth}{Generic freeness\phantom{p}}
        \justifying
        Generically, any finitely generated module over a reduced ring is free.
      \end{varblock}
      \vspace*{-0.5em}

      \includegraphics[width=0.6\textwidth]{generic-freeness}
    \end{column}
  \end{columns}
\end{frame}

\section{The semantics}

\subsection[Motivation]{Motivating the semantics}

\begin{frame}{Motivating the semantics}
  \centering
  \begin{varblockextra}{0.8\textwidth}{}{
    \hil{Examples:}\phantom{Non-}\,\! $k,\ k[[X]],\ \CC\{z\},\ \ZZ_{(p)}$ \\[0.2em]
    \hil{Non-examples:} $\ZZ,\ k[X],\ \ZZ/(pq)$
  }
    \justifying
    A ring is \hil{local} iff~$1 \neq 0$ and~$x + y = 1$ implies that~$x$ is
    invertible or~$y$ is invertible.
  \end{varblockextra}

  \begin{varblockextra}{0.8\textwidth}{}{
    Let~$x + y = 1$ in a ring~$A$.
    Then:
    \begin{itemize}
      \item The element $x$ is invertible in~$A[x^{-1}]$.
      \item The element $y$ is invertible in~$A[y^{-1}]$.
    \end{itemize}
  }
    \hil{Locally,} any ring is local.
  \end{varblockextra}
\end{frame}

\subsection{Definition}

\begin{frame}{The semantics}
  \small
  Let~$A$ be a fixed ring. Let~``$A \models \varphi$'' be a shorthand for~``$1
  \models \varphi$''.
  \only<1>{\[ \renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\quad}c@{\quad}l@{}}
    f \models \top &\text{iff}& \top \\
    f \models \bot &\text{iff}& \text{$f$ is nilpotent} \\
    f \models x = y &\text{iff}& x = y \in A[f^{-1}] \\
    f \models \varphi \wedge \psi &\text{iff}&
      \text{$f \models \varphi$ and $f \models \psi$} \\
    f \models \varphi \vee \psi &\text{iff}&
      \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
    &&\quad\text{for each~$i$, $fg_i \models \varphi$ or $fg_i \models \psi$} \\
    f \models \varphi \Rightarrow \psi &\text{iff}&
      \text{for all~$g \in A$, $fg \models \varphi$ implies $fg \models \psi$} \\
    f \models \forall x\?A^\sim\_ \varphi &\text{iff}&
      \text{for all~$g \in A$ and $x_0 \in A[(fg)^{-1}]$, $fg \models \varphi[x_0/x]$} \\
    f \models \exists x\?A^\sim\_ \varphi &\text{iff}&
      \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
    &&\quad\text{for each~$i$, $g_i \models \varphi[x_0/x]$ for some~$x_0 \in A[(fg_i)^{-1}]$}
  \end{array} \]}
  \only<2->{\[ \renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\quad}c@{\quad}l@{}}
    f \models x = y &\text{iff}& x = y \in A[f^{-1}] \\
    f \models \varphi \wedge \psi &\text{iff}&
      \text{$f \models \varphi$ and $f \models \psi$} \\
    f \models \varphi \vee \psi &\text{iff}&
      \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
    &&\quad\text{for each~$i$, $fg_i \models \varphi$ or $fg_i \models \psi$}
  \end{array} \]}

  \begin{columns}
    \begin{column}{0.50\textwidth}
      \begin{varblock}{\textwidth}{Monotonicity}{}
        If~$f \models \varphi$, then also~$fg \models \varphi$.
      \end{varblock}
    \end{column}

    \begin{column}{0.50\textwidth}
      \begin{varblock}{\textwidth}{Locality}{}
        \justifying
        If~$f^n = fg_1 + \cdots + fg_m$ and~$fg_i \models \varphi$ for all~$i$,
        then also~$f \models \varphi$.
      \end{varblock}
    \end{column}
  \end{columns}

  \begin{columns}
    \begin{column}{0.50\textwidth}
      \begin{varblock}{\textwidth}{Soundness\phantom{p}}{}
        If~$\varphi \vdash \psi$ and~$f \models \varphi$,
        then~$f \models \psi$.
      \end{varblock}
    \end{column}

    \begin{column}{0.50\textwidth}
      \begin{varblock}{\textwidth}{Forced properties}{}
        $A \models \speak{$A^\sim$ is a local ring}$.
      \end{varblock}
    \end{column}
  \end{columns}
\end{frame}

\subsection{A baby application}

\begin{frame}{A baby application}
  \vspace*{-1em}
  \begin{varblock}{\textwidth}{}
    Let~$M \in A^{n \times m}$ be a surjective matrix over a ring~$A$.
    If~$n > m$, then~$1 = 0 \in A$.
  \end{varblock}

  \justifying
  \emph{Classical proof.} Assume to the contrary that~$1 \neq 0 \in A$. Pick a maximal
  ideal~$\mmm$ of~$A$. Then~$M$ is surjective as a matrix over the
  field~$A/\mmm$. This is in contradiction to basic linear algebra. \qed

  \emph{Constructive proof.} We verify that
  $A \models \speak{$M$ is surjective}$.
  Since the claim admits an intuitionistic proof in the case that the ring
  is local, soundness implies that $A \models 1 = 0$.
  Thus~$1 = 0 \in A$. \qed

  \centering
  \includegraphics[width=0.8\textwidth]{nontrivial-uses-of-trivial-rings}
\end{frame}

\subsection[Properties]{Investigating the forcing model}

\begin{frame}{Investigating the forcing model}
  \vspace*{-1em}
  \begin{varblockextra}{\textwidth}{}{
    \hil{Examples:} being local, reduced, an integral domain.
  }
    \justifying
    Assuming the Boolean prime ideal theorem, any first-order
    formula ``$\forall \ldots \forall\_ (\cdots \Longrightarrow \cdots\!\,)$'',
    where the two subformulas may not contain~``$\Rightarrow$'' and~``$\forall$'',
    holds for~$A^\sim$ iff it holds for all stalks~$A_\ppp$.
  \end{varblockextra}
  \pause
  \begin{align*}
    \intertext{The forcing model has additional \hil{unique properties}, e.\,g.\@}
    A &\models \forall x\?A^\sim\_ \neg(\speak{$x$ inv.}) \Rightarrow \speak{$x$ nilpotent} \\
    \intertext{which if~$A$ is reduced implies the \hil{field condition}}
    A &\models \forall x\?A^\sim\_ \neg(\speak{$x$ inv.}) \Rightarrow x = 0.
  \end{align*}
  \emph{Translation.} For any element~$x \in A$, if~$f = 0$ is the only element such
  that~$x$ is invertible in~$A[f^{-1}]$, then~$x = 0$.
\end{frame}

\section{Grothendieck's generic freeness}

\subsection{Statement}

\begin{frame}{Grothendieck's generic freeness}
  \small
  Let~$A$ be a reduced ring. \\
  Let~$B$ be an~$A$-algebra of finite type ($\cong A[X_1,\ldots,X_n]/\aaa$). \\
  Let~$M$ be a finitely generated~$B$-module ($\cong B^m/U$).

  \vspace{-1em}
  \begin{varblock}{\textwidth}{}
    \hil{Theorem.} If~$1 \neq 0$ in~$A$, there exists $f \neq 0$ in $A$ such that
    \vspace*{-0.5em}
    \begin{enumerate}
      \item $B[f^{-1}]$ and $M[f^{-1}]$ are free modules over $A[f^{-1}]$,
      \item \vspace*{-0.2cm}$A[f^{-1}] \to B[f^{-1}]$ is of finite presentation, and
      \item \vspace*{-0.2cm}$M[f^{-1}]$ is finitely presented as a module over~$B[f^{-1}]$.
    \end{enumerate}
  \end{varblock}

  \begin{columns}[t]
    \begin{column}{0.34\textwidth}
      \centering
      \ \\[0.5em]
      \includegraphics[width=0.7\textwidth]{generic-freeness}

      \small
      $A = k[X]$, \\ $B = M = k[X,Y]/(XY)$
    \end{column}

    \pause

    \begin{column}{0.68\textwidth}
      \vspace*{-0.5em}
      \begin{itemize}
        \item No generalization to unreduced rings.
        \item \vspace*{-0.2cm}Implies the law of excluded middle.
        \pause
        \item \vspace*{-0.2cm}\hil{Constructive restatement.} \\ If zero is the only
        element~$f \in A$ such that
        {\!\renewcommand{\insertenumlabel}{1}\usebeamertemplate{enumerate item}},
        {\!\renewcommand{\insertenumlabel}{2}\usebeamertemplate{enumerate item}}, and
        {\!\renewcommand{\insertenumlabel}{3}\usebeamertemplate{enumerate item}},
        then~$1 = 0 \in A$.
      \end{itemize}
    \end{column}
  \end{columns}
\end{frame}

\subsection{Proof}

\begin{frame}{A constructive proof}
  \small
  Let~$A$ be a reduced ring. \\
  Let~$B$ be an~$A$-algebra of finite type ($\cong A[X_1,\ldots,X_n]/\aaa$). \\
  Let~$M$ be a finitely generated~$B$-module ($\cong B^m/U$).

  \only<1>{\vspace{-1em}
  \begin{varblock}{\textwidth}{}
    \hil{Theorem.} If zero is the only element~$f \in A$ such that
    \vspace*{-0.5em}
    \begin{enumerate}
      \item $B[f^{-1}]$ and $M[f^{-1}]$ are free modules over $A[f^{-1}]$,
      \item \vspace*{-0.2cm}$A[f^{-1}] \to B[f^{-1}]$ is of finite presentation, and
      \item \vspace*{-0.2cm}$M[f^{-1}]$ is finitely presented as a module over~$B[f^{-1}]$,
    \end{enumerate}
    \vspace*{-0.5em}
    then~$1 = 0 \in A$.
  \end{varblock}}

  \emph{Constructive proof.} Observe that the theorem amounts to
  \vspace*{-0.5em}
  \begin{multline*}A \models \ulcorner\text{It's \hil{not not} the case that} \\
    \begin{minipage}{9cm}\vspace*{-0.2cm}\begin{enumerate}
      \item $B^\sim$ and $M^\sim$ are free modules over $A^\sim$,
      \item \vspace*{-0.3cm}$A^\sim \to B^\sim$ is of finite presentation, and
      \item \vspace*{-0.3cm}$M^\sim$ is finitely presented as a module
      over~$B^\sim$$\urcorner$.
    \end{enumerate}\end{minipage}\end{multline*}
  \pause
  \justifying
  Claims~{\!\renewcommand{\insertenumlabel}{2}\usebeamertemplate{enumerate item}}
  and~{\!\renewcommand{\insertenumlabel}{3}\usebeamertemplate{enumerate item}}
  follow from the fact that~$A^\sim$ is \hil{anonymously Noetherian} (any ideal
  is \hil{not not} finitely generated) which entails
  that~$A^\sim[X_1,\ldots,X_n]$ is anonymously Noetherian.

  Claim~{\!\renewcommand{\insertenumlabel}{1}\usebeamertemplate{enumerate item}}
  follows from a careful rendition of the standard linear algebra proof,
  employing Dickson's lemma to ensure termination. \qed
\end{frame}

\backupstart

\begin{frame}
  Assume that~$B^\sim$ is generated by~$(x^i y^j)_{i,j\geq0}$ as an~$A^\sim$-module.
  It's \hil{not not} the case that either some generator can be expressed as a
  linear combination of others with smaller index, or not.

  \centering
  \begin{tikzpicture}[scale=0.4]
    \begin{scope}
      \draw[step=1cm,gray,very thin] (0,0) grid (8,8);

      \draw
        (0,8) -- (0,0) -- (8,0);

      \fill[fill=blue!40!white,opacity=0.5]
        (0,8) -- (0,0) -- (6,0) -- (6,5) -- (5,5) -- (5,8);

      \fill[fill=red!40!white,opacity=0.5]
        (5,5) -- (6,5) -- (6,6) -- (5,6);

      \node[shape=circle,draw,inner sep=2pt] at (1,1) {1};
    \end{scope}

    \begin{scope}[xshift=9cm]
      \draw[step=1cm,gray,very thin] (0,0) grid (8,8);

      \draw
        (0,8) -- (0,0) -- (8,0);

      \fill[pattern=north east lines,pattern color=gray,opacity=0.5]
        (5,8) -- (5,5) -- (8,5) -- (8,8);
      \draw
        (5,8) -- (5,5) -- (8,5);

      \fill[fill=blue!40!white,opacity=0.5]
        (0,8) -- (0,0) -- (3,0) -- (3,5) -- (2,5) -- (2,8);

      \fill[fill=red!40!white,opacity=0.5]
        (2,5) -- (3,5) -- (3,6) -- (2,6);

      \node[shape=circle,draw,inner sep=2pt] at (1,1) {2};
    \end{scope}

    \begin{scope}[xshift=18cm]
      \draw[step=1cm,gray,very thin] (0,0) grid (8,8);

      \draw
        (0,8) -- (0,0) -- (8,0);

      \fill[pattern=north east lines,pattern color=gray,opacity=0.5]
        (2,8) -- (2,5) -- (8,5) -- (8,8);
      \draw
        (2,8) -- (2,5) -- (8,5);

      \fill[fill=blue!40!white,opacity=0.5]
        (0,8) -- (0,0) -- (7,0) -- (7,2) -- (6,2) -- (6,5) -- (2,5) -- (2,8);

      \fill[fill=red!40!white,opacity=0.5]
        (6,2) -- (7,2) -- (7,3) -- (6,3);

      \node[shape=circle,draw,inner sep=2pt] at (1,1) {3};
    \end{scope}

    \begin{scope}[yshift=-9cm]
      \draw[step=1cm,gray,very thin] (0,0) grid (8,8);

      \draw
        (0,8) -- (0,0) -- (8,0);

      \fill[pattern=north east lines,pattern color=gray,opacity=0.5]
        (2,8) -- (2,5) -- (6,5) -- (6,2) -- (8,2) -- (8,8);
      \draw
        (2,8) -- (2,5) -- (6,5) -- (6,2) -- (8,2);

      \fill[fill=blue!40!white,opacity=0.5]
        (0,8) -- (0,0) -- (6,0) -- (6,3) -- (5,3) -- (5,5) -- (2,5) -- (2,8);

      \fill[fill=red!40!white,opacity=0.5]
        (5,3) -- (6,3) -- (6,4) -- (5,4);

      \node[shape=circle,draw,inner sep=2pt] at (1,1) {4};
    \end{scope}

    \begin{scope}[yshift=-9cm,xshift=9cm]
      \draw[step=1cm,gray,very thin] (0,0) grid (8,8);

      \draw
        (2,8) -- (2,5) -- (5,5) -- (5,3) -- (6,3) -- (6,2) -- (8,2);

      \draw
        (0,8) -- (0,0) -- (8,0);

      \fill[pattern=north east lines,pattern color=gray,opacity=0.5]
        (2,8) -- (2,5) -- (5,5) -- (5,3) -- (6,3) -- (6,2) -- (8,2) -- (8,8);

      \fill[fill=blue!40!white,opacity=0.5]
        (0,8) -- (0,0) -- (4,0) -- (4,3) -- (3,3) -- (3,5) -- (2,5) -- (2,8);

      \fill[fill=red!40!white,opacity=0.5]
        (3,3) -- (4,3) -- (4,4) -- (3,4);

      \node[shape=circle,draw,inner sep=2pt] at (1,1) {5};
    \end{scope}

    \begin{scope}[yshift=-9cm,xshift=18cm]
      \draw[step=1cm,gray,very thin] (0,0) grid (8,8);

      \draw
        (0,8) -- (0,0) -- (8,0);

      \fill[pattern=north east lines,pattern color=gray,opacity=0.5]
        (2,8) -- (2,5) -- (3,5) -- (3,3) -- (6,3) -- (6,2) -- (8,2) -- (8,8);
      \draw
        (2,8) -- (2,5) -- (3,5) -- (3,3) -- (6,3) -- (6,2) -- (8,2);

      \node[shape=circle,draw,inner sep=2pt] at (1,1) {6};
    \end{scope}
  \end{tikzpicture}
  \par
\end{frame}

\begin{frame}{An explicit constructive proof}
  \justifying\small\fontsize{10pt}{12.3}\selectfont
  \vspace*{-0.5em}

  \hil{Lemma.} Let~$A$ be a ring. Let~$M$ be an~$A$-module with generating
  family~$(x_1,\ldots,x_n)$. Assume that the only element~$g \in A$ such that one of
  the~$x_i$ is an~$A[g^{-1}]$-linear combination in~$A[g^{-1}]$ of the other generators
  is~$g = 0$. Then~$M$ is free with~$(x_1,\ldots,x_n)$ as a basis.

  \emph{Proof.} Let~$\textstyle\sum_i a_i x_i = 0$. Let $i$ be arbitrary. In $M[a_i^{-1}]$,
  the generator~$x_i$ is a linear combination of the other generators.
  Thus~$a_i = 0$. \qed

  \hil{Theorem.} Let~$A$ be a reduced ring. Let~$M$ be a finitely
  generated~$A$-module. If zero is the only element~$f \in A$ such
  that~$M[f^{-1}]$ is finite free as an~$A[f^{-1}]$-module, then~$1 = 0$
  in~$A$.

  \emph{Proof.}
  By induction on the length~$n$ of a generating family
  $(x_1,\ldots,x_n)$ of~$M$.

  % Note that we'll apply the induction hypothesis not to the ring~$A$, but to
  % some localizations of~$A$.
  We verify the assumption of the lemma.
  Thus let~$g \in A$ be given such that one of the~$x_i$ is an
  $A[g^{-1}]$-linear combination of the others in~$M[g^{-1}]$. Therefore the
  $A[g^{-1}]$-module~$M[g^{-1}]$ can be generated by $n-1$ elements. By the
  induction hypothesis (applied to~$A[g^{-1}]$ and its
  module~$M[g^{-1}]$) it follows that $A[g^{-1}] = 0$. Therefore $g = 0$.

  Thus~$M$ is free. We finish by using the assumption for $f = 1$.
  \qed
\end{frame}

\begin{frame}{An explicit constructive proof}
  \justifying\small
  \hil{Theorem.} Let~$A$ be a reduced ring. Let~$B$ by a finitely
  generated~$A$-algebra. If zero is the only element~$f \in A$ such
  that~$B[f^{-1}]$ is finitely presented as an~$A[f^{-1}]$-algebra, then~$1 =
  0$ in~$A$.

  \emph{Proof.} Write~$B = A[X_1,\ldots,X_n]/\aaa$. We describe only the
  case~$n = 0$.

  As a first step, we verify~$\aaa = (0)$. Let~$f \in \aaa$. Then~$B[f^{-1}] =
  0$. Thus~$f = 0$ by assumption.

  We now use the assumption again, this time for~$f = 1$.
  \qed
\end{frame}

\backupend

\end{document}

### 2018-04-10 slides-munich2018.tex
typo

@@
     \intertext{The forcing model has additional \hil{unique properties}, e.\,g.\@}
-    A &\models \forall x\?A^\sim\_ \neg(\speak{$x$ inv.}) \Rightarrow \speak{$x$ nilpotent} \\
+    A &\models \forall x\?A^\sim\_ \neg(\speak{$x$ inv.}) \Longrightarrow \speak{$x$ nilpotent} \\
     \intertext{which if~$A$ is reduced implies the \hil{field condition}}
-    A &\models \forall x\?A^\sim\_ \neg(\speak{$x$ inv.}) \Rightarrow x = 0.
+    A &\models \forall x\?A^\sim\_ \neg(\speak{$x$ inv.}) \Longrightarrow x = 0
+    \quad\text{and also} \\
+    A &\models \forall x\?A^\sim\_ \neg\neg(x = 0) \Longrightarrow x = 0.
   \end{align*}

### 2018-04-14 paper-generic-freeness.tex
Begin of paper on generic freeness

\documentclass{amsart}

\usepackage[utf8]{inputenc}
\usepackage{amsthm,mathtools}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{hyperref}

\title{An elementary and constructive proof of Grothendieck's generic freeness
lemma}
\author{Ingo Blechschmidt}

\theoremstyle{definition}
\newtheorem{defn}{Definition}[]
\newtheorem{ex}[defn]{Example}

\theoremstyle{plain}
\newtheorem{prop}[defn]{Proposition}
\newtheorem{cor}[defn]{Corollary}
\newtheorem{lemma}[defn]{Lemma}
\newtheorem{thm}[defn]{Theorem}
\newtheorem{scholium}[defn]{Scholium}

\theoremstyle{remark}
\newtheorem{rem}[defn]{Remark}
\newtheorem{question}[defn]{Question}
\newtheorem{speculation}[defn]{Speculation}
\newtheorem{caveat}[defn]{Caveat}
\newtheorem{conjecture}[defn]{Conjecture}

\newcommand{\XXX}[1]{\textbf{XXX: #1}}
\newcommand{\defeq}{\vcentcolon=}
\DeclareMathOperator{\Spec}{Spec}

\begin{document}

\begin{abstract}
  We present a new and direct proof of Grothendieck's generic freeness
  lemma. Unlike the previously published proofs, it doesn't require a series of
  reduction steps and is fully constructive, not involving the axiom of choice
  or even the law of excluded middle. It was found using a general
  topos-theoretic technique. \XXX{self-contained...}
\end{abstract}

\maketitle

We prove Grothendieck's generic freeness lemma for rings and modules in the
following form.

\begin{thm}\label{thm:algebraic}Let~$A$ be a reduced ring. Let~$B$ be
an~$A$-algebra of finite type. Let~$M$ be a finitely generated~$B$-module.
If~$f = 0$ is the only element of~$A$ such that
\begin{enumerate}
\item the~$A[f^{-1}]$-modules $B[f^{-1}]$ and $M[f^{-1}]$ are free,
\item the~$A[f^{-1}]$-algebra~$B[f^{-1}]$ is of finite presentation, and
\item the~$B[f^{-1}]$-module~$M[f^{-1}]$ is finitely presented,
\end{enumerate}
then~$1 = 0$ in~$A$.
\end{thm}

Previously known proofs proceed in a series of intermediate steps, reducing
to the case that~$A$ is a Noetherian integral domain where one can argue by
\emph{dévissage}; but in fact, a direct proof is possible and shorter.

Grothendieck's generic freeness lemma is often presented in contrapositive form
or in the following geometric variant:

\begin{thm}\label{thm:geometric}Let~$A$ be a reduced ring. Let~$B$ be
an~$A$-algebra of finite type. Let~$M$ be a finitely generated~$B$-module. Then
the space~$\Spec(A)$ contains a dense open~$U$ such that over~$U$,
\begin{enumerate}
\item[(a)] $B^\sim$ and~$M^\sim$ are free as sheaves of~$A^\sim$-modules,
\item[(b)] $B^\sim$ is of finite presentation as a sheaf of~$A^\sim$-algebras, and
\item[(c)] $M^\sim$ is finitely presented as a sheaf of~$B^\sim$-modules.
\end{enumerate}
\end{thm}

Theorem~\ref{thm:geometric} immediately follows from
Theorem~\ref{thm:algebraic} by defining~$U$ as the union of all the basic
opens~$D(f)$ such that~(1),~(2), and~(3) hold. It's clear that~(a),~(b),
and~(c) hold over~$U$, and~$U$ is dense for if~$V$ is an arbitrary open
such that~$U \cap V = \emptyset$, the open~$V$ is itself empty: Let~$h \in A$
such that~$D(h) \subseteq V$. The hypothesis implies the assumptions of
Theorem~\ref{thm:algebraic} for the datum~$(A[h^{-1}], B[h^{-1}], M[h^{-1}])$.
Thus~$1 = 0 \in A[h^{-1}]$, so~$h$ is nilpotent and~$D(h) = \emptyset$.

The new proof was found using a general topos-theoretical technique which we
believe to be useful in other situations as well. This technique allows to view
reduced rings and their modules from a different point of view, one from which
reduced rings look like fields. Using this technique, Grothendieck's generic freeness
lemma follows trivially, since the lemma is trivial for fields. The proof
presented here was obtained by unwinding the topos-theoretic proof, yielding a
self-contained argument without any references to topos theory.

\XXX{Acknowledgments, especially Martin Brandenburg}

\XXX{References to prior proofs: EGA, Stacks Project, Charles Staats. Check
\url{https://math.stackexchange.com/questions/1048496/question-on-generic-freeness-ref-matsumura-page-185}}

\section{The proof}

\begin{lemma}\label{lemma:basis}
Let~$A$ be a ring. Let~$M$ be an~$A$-module with generating
family~$(x_i)_{i \in I}$ where~$I$ is a totally ordered set. Assume that the
only element~$g \in A$ such that one of the~$x_i$ is an~$A[g^{-1}]$-linear
combination in~$M[g^{-1}]$ of other generators with smaller index is~$g = 0$.
Then~$M$ is free with~$(x_i)_{i \in I}$ as a basis.
\end{lemma}

\begin{proof}Let~$\sum_i a_i x_i = 0$. Starting with the greatest
index~$i$ which appears in that sum, we see that in~$M[a_i^{-1}]$, the
element~$x_i$ is an~$A[g^{-1}]$-linear combination of other generators with
smaller index. Thus~$a_i = 0$ by assumption.\end{proof}

\begin{thm}Let~$A$ be a reduced ring. Let~$B$ be an~$A$-algebra of finite type.
Let~$M$ be a finitely generated~$B$-module. If~$f = 0$ is the only element
of~$A$ such that~$M[f^{-1}]$ is free as an~$A[f^{-1}]$-module, then~$1 = 0$
in~$A$.\end{thm}

\begin{proof}Let~$B$ be generated by~$(x_1,\ldots,x_n)$ as an~$A$-algebra and
let~$M$ be generated by~$(v_1,\ldots,v_m)$ as an~$B$-module. We endow the set
\[ I \defeq \{ (\ell, j_1,\ldots,j_n) \,|\,
  \ell \in \{ 1,\ldots,m \},
  j_1,\ldots,j_n \in \{ 0,1,\ldots \} \} \]
with the lexicographic order. The family~$(w_i)_{i \in I} \defeq (x_1^{j_1}
\ldots x_n^{j_n} v_\ell)_{(\ell,j_1,\ldots,j_n) \in I}$ thus generates~$M$ as
an~$A$-module, and we'll call a subfamily~$(w_i)_{i \in I' \subseteq I}$
\emph{good} if and only if for all~$i \in I$, the vector~$w_i$ is a linear
combination of the vectors~$(w_{i'})_{i' \in I, i' \preceq i}$,
and if~$(\ell,j_1,\ldots,j_n) \not\in I'$ implies~$(\ell,k_1,\ldots,k_n) \not\in
I'$ for all~$k_1 \geq j_1, \ldots, k_n \geq j_n$. \XXX{Looks like in Figure XXX}

We induct on the shape of a good generating system, starting with~$(w_i)_{i \in I}$.

We show that~$(w_i)_{i \in I}$ is a basis of~$M$ by verifying the assumptions
of Lemma~\ref{lemma:basis}. Thus let~$g \in A$ be given such that one of
the~$w_i$ is an $A[g^{-1}]$-linear combination of generators with smaller index
in~$M[g^{-1}]$. Removing~$w_i = x_1^{j_1} \ldots x_n^{j_n} v_\ell$ and also all
vectors~$x_1^{k_1} \ldots x_n^{k_n} v_\ell$ where~$k_1 \geq j_1, \ldots, k_n
\geq j_n$, we obtain a subfamily which is still good. By induction, applied
to~$A[g^{-1}]$ and its module~$M[g^{-1}]$, it therefore follows that $A[g^{-1}] = 0$.
This implies that $g = 0$ since~$A$ is reduced.

Thus~$M$ is free. We finish by using the assumption for~$f = 1$.
\end{proof}

\end{document}

### 2018-04-14 slides-munich2018.tex
typo (thanks to Matthias!)
typo
Begin of paper on generic freeness

@@
 \author{Ingo Blechschmidt}
-\date{April 11th, 2018}
+\date{April 11st, 2018}
 
@@
   Ludwig-Maximilians-Universität München \\
-  April 11th, 2018
+  April 11st, 2018
   \par
@@
   family~$(x_1,\ldots,x_n)$. Assume that the only element~$g \in A$ such that one of
-  the~$x_i$ is an~$A[g^{-1}]$-linear combination in~$A[g^{-1}]$ of the other generators
+  the~$x_i$ is an~$A[g^{-1}]$-linear combination in~$M[g^{-1}]$ of the other generators
   is~$g = 0$. Then~$M$ is free with~$(x_1,\ldots,x_n)$ as a basis.
@@
   \justifying\small
-  \hil{Theorem.} Let~$A$ be a reduced ring. Let~$B$ by a finitely
+  \hil{Theorem.} Let~$A$ be a reduced ring. Let~$B$ be a finitely
   generated~$A$-algebra. If zero is the only element~$f \in A$ such

### 2018-04-15 paper-generic-freeness.tex
Proof of the finitely-generated case

@@
 
-\section{The proof}
+\section{The proof of the general case}
 
@@
 
+\section{The proof of the finitely-generated case}
+
+The following proposition is just an instance of Grothendieck's generic
+freeness lemma. We prove it here because it admits an easier proof.
+
+\begin{prop}Let~$A$ be a reduced ring. Let~$M$ be a finitely
+generated~$A$-module. If~$f = 0$ is the only element of~$A$ such
+that~$M[f^{-1}]$ is a finite free~$A[f^{-1}]$-module, then~$1 = 0$ in~$A$.
+\end{prop}
+
+\begin{proof}We proceed by induction on the length of a given generating family
+of~$M$. Let~$M$ be generated by~$(v_1,\ldots,v_m)$.
+
+We show that the family~$(v_1,\ldots,v_m)$ is linear independent. Let~$\sum_i
+a_i v_i = 0$. Over~$A[a_i^{-1}]$, the vector~$v_i \in M[a_i^{-1}]$ is a linear
+combination of the other generators. Thus~$M[a_i^{-1}]$ can be generated as
+an~$A[a_i^{-1}]$-module by fewer than~$m$ generators. The induction hypothesis,
+applied to this module, yields that~$1 = 0$ in~$A[a_i^{-1}]$. Since~$A$ is
+reduced, this amounts to~$a_i = 0$.
+
+We finish by using the assumption for~$f = 1$.
+\end{proof}
+
 \end{document}

### 2018-04-15 paper-generic-freeness.tex
On weakly Noetherian modules

@@
 \newcommand{\XXX}[1]{\textbf{XXX: #1}}
+\newcommand{\aaa}{\mathfrak{a}}
+\newcommand{\bbb}{\mathfrak{b}}
 \newcommand{\defeq}{\vcentcolon=}
@@
   We present a new and direct proof of Grothendieck's generic freeness
-  lemma. Unlike the previously published proofs, it doesn't require a series of
-  reduction steps and is fully constructive, not involving the axiom of choice
-  or even the law of excluded middle. It was found using a general
-  topos-theoretic technique. \XXX{self-contained...}
+  lemma in its general form. Unlike the previously published proofs, it doesn't
+  proceed in a series of reduction steps and is fully constructive, not
+  involving the axiom of choice or even the law of excluded middle. It was
+  found by unwinding the result of a general topos-theoretic technique.
 \end{abstract}
@@
 reduced rings and their modules from a different point of view, one from which
-reduced rings look like fields. Using this technique, Grothendieck's generic freeness
-lemma follows trivially, since the lemma is trivial for fields. The proof
-presented here was obtained by unwinding the topos-theoretic proof, yielding a
-self-contained argument without any references to topos theory.
+reduced rings look like fields. Since Grothendieck's generic freeness
+is trivial for fields, this technique yields a trivial proof for reduced rings.
+The proof presented here was obtained by unwinding the topos-theoretic proof,
+yielding a self-contained argument without any references to topos theory.
+We refer readers who want to learn about this technique to an accompanying
+note~\cite{blechschmidt:wlog}.
 
@@
 \XXX{References to prior proofs: EGA, Stacks Project, Charles Staats. Check
-\url{https://math.stackexchange.com/questions/1048496/question-on-generic-freeness-ref-matsumura-page-185}}
+\href{https://math.stackexchange.com/questions/1048496/question-on-generic-freeness-ref-matsumura-page-185}{this
+MO question}}
 
@@
 
+Recall that a module is Noetherian if and only if all of its submodules are
+finitely generated. We call a module~$M$ over an~$A$-algebra~$B$ \emph{weakly
+Noetherian} (over~$A$) if and only if, for all~$B$-submodules~$U \subseteq M$
+and all~$f \in A$ the condition
+\begin{multline*}
+  \text{for all~$g \in A$, if $U[(fg)^{-1}]$ is finitely generated as
+an~$B[(fg)^{-1}]$-module,} \\
+  \text{then~$fg$ is nilpotent in~$A$} \end{multline*}
+implies that~$f$ is nilpotent. Geometrically, this expresses that all
+quasicoherent sheaves of submodules of~$M^\sim$ are, over a dense open
+of~$\Spec(A)$, of finite type over~$B^\sim$. An~$A$-algebra~$B$
+is \emph{weakly Noetherian} if and only if it is weakly Noetherian as a module
+over itself.
+
+This notion is interesting for the following reason.
+
+\begin{lemma}\label{lemma:self-noetherian}
+Let~$A$ be a reduced ring. Then~$A$, regarded as an algebra over itself, is
+weakly Noetherian.\end{lemma}
+
+\begin{proof}Let~$\aaa \subseteq A$ be an ideal. Let~$f \in A$ such that, for
+any~$g \in A$, if~$\aaa[(fg)^{-1}]$ is finitely generated, then~$fg = 0$.
+We verify that~$f = 0$.
+
+We first show that~$\aaa[f^{-1}] = (0)$. Let~$g \in \aaa$. Then~$\aaa[(fg)^{-1}] =
+(1)$ is finitely generated. Thus~$fg = 0$, hence~$g = 0$ in~$A[f^{-1}]$.
+
+Since~$\aaa[f^{-1}]$ is finitely generated, we finish by using the
+assumption for~$g = 1$.
+\end{proof}
+
+The weak Noetherian condition enjoys the same basic stability properties as the
+usual Noetherian condition:
+
+\begin{lemma}\label{lemma:noetherian-stability}
+Let~$A$ be a ring. Let~$B$ be an~$A$-algebra.
+\begin{enumerate}
+\item If~$B$ is weakly Noetherian, then so is~$B[X]$.
+\item Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence
+of~$B$-modules. Then~$M$ is weakly Noetherian if and only if~$M'$ and~$M''$
+are.
+\end{enumerate}
+\end{lemma}
+
+We can use these observations to finish the proof of Grothendieck's generic
+freeness lemma.
+
+\begin{prop}Let~$A$ be a reduced ring. Let~$B$ be an~$A$-algebra of finite type.
+If~$f = 0$ is the only element of~$A$ such that~$B[f^{-1}]$ is of finite
+presentation as an~$A[f^{-1}]$-algebra, then~$1 = 0$ in~$A$.
+\end{prop}
+
+\begin{proof}We write~$B = A[X_1,\ldots,X_n]/\aaa$. By Lemma~\ref{lemma:self-noetherian} and
+Lemma~\ref{lemma:noetherian-stability}, the~$A$-algebra~$A[X_1,\ldots,X_n]$ is
+weakly Noetherian. We can therefore conclude if we can show, for any~$g \in A$,
+that if the ideal~$\aaa[g^{-1}]$ is finitely generated, then~$g = 0$. This
+follows immediately from the assumption (for~$f = g$).
+\end{proof}
+
+\begin{prop}Let~$A$ be a reduced ring. Let~$B$ be an~$A$-algebra of finite type.
+Let~$M$ be a finitely generated~$B$-module. If~$f = 0$ is the only element
+of~$A$ such that~$M[f^{-1}]$ is finitely presented as an~$B[f^{-1}]$-module,
+then~$1 = 0$ in~$A$.
+\end{prop}
+
+\begin{proof}We write~$M = B^n/U$ for an~$B$-submodule~$U$. By Lemma~\ref{lemma:self-noetherian} and
+Lemma~\ref{lemma:noetherian-stability}, the~$B$-module~$B^n$ is weakly
+Noetherian. We can therefore conclude if we can show, for any~$g \in A$, that if
+the submodule~$U[g^{-1}]$ is finitely generated, then~$g = 0$. This follows
+immediately from the assumption (for~$f = g$).
+\end{proof}
+
 \section{The proof of the finitely-generated case}

### 2018-04-15 paper-generic-freeness.tex
Illustration of a good generating family

@@
 \usepackage[utf8]{inputenc}
-\usepackage{amsthm,mathtools}
+\usepackage{amsthm,mathtools,tikz}
+\usetikzlibrary{patterns,matrix}
 \usepackage[protrusion=true,expansion=true]{microtype}
@@
 and if~$(\ell,j_1,\ldots,j_n) \not\in I'$ implies~$(\ell,k_1,\ldots,k_n) \not\in
-I'$ for all~$k_1 \geq j_1, \ldots, k_n \geq j_n$. \XXX{Looks like in Figure XXX}
+I'$ for all~$k_1 \geq j_1, \ldots, k_n \geq j_n$. Figure~\ref{fig:good} shows
+how a good generating family can look like.
 
-We induct on the shape of a good generating system, starting with~$(w_i)_{i \in I}$.
+\begin{figure}
+  \begin{tikzpicture}[scale=1.0]
+    \draw[step=1cm,gray,very thin] (0,0) grid (8,8);
+
+    \draw
+      (0,8) -- (0,0) -- (8,0);
+
+    \fill[pattern=north east lines,pattern color=gray,opacity=0.5]
+      (2,8) -- (2,5) -- (5,5) -- (5,3) -- (6,3) -- (6,2) -- (8,2) -- (8,8);
+    \draw
+      (2,8) -- (2,5) -- (5,5) -- (5,3) -- (6,3) -- (6,2) -- (8,2);
+
+    \fill[fill=blue!40!white,opacity=0.5]
+      (0,8) -- (0,0) -- (4,0) -- (4,3) -- (3,3) -- (3,5) -- (2,5) -- (2,8);
+
+    \fill[fill=red!40!white,opacity=0.5]
+      (3,3) -- (4,3) -- (4,4) -- (3,4);
+
+    % !perl -we 'for my $y (reverse 0..7) { for my $x (0..7) { print % "\$x^$x\Ey^$y\$ & " } print "\\\\ \n" }'
+    \begin{scope}[yshift=4cm, xshift=4cm]
+      \matrix[matrix of nodes,nodes={inner sep=0pt,text width=1.0cm,align=center,minimum height=1.0cm}]{
+        \tiny $x^0y^7v_1$ & \tiny $x^1y^7v_1$ & \tiny $x^2y^7v_1$ & \tiny $x^3y^7v_1$ & \tiny $x^4y^7v_1$ & \tiny $x^5y^7v_1$ & \tiny $x^6y^7v_1$ & \tiny $x^7y^7v_1$ & \\
+        \tiny $x^0y^6v_1$ & \tiny $x^1y^6v_1$ & \tiny $x^2y^6v_1$ & \tiny $x^3y^6v_1$ & \tiny $x^4y^6v_1$ & \tiny $x^5y^6v_1$ & \tiny $x^6y^6v_1$ & \tiny $x^7y^6v_1$ & \\
+        \tiny $x^0y^5v_1$ & \tiny $x^1y^5v_1$ & \tiny $x^2y^5v_1$ & \tiny $x^3y^5v_1$ & \tiny $x^4y^5v_1$ & \tiny $x^5y^5v_1$ & \tiny $x^6y^5v_1$ & \tiny $x^7y^5v_1$ & \\
+        \tiny $x^0y^4v_1$ & \tiny $x^1y^4v_1$ & \tiny $x^2y^4v_1$ & \tiny $x^3y^4v_1$ & \tiny $x^4y^4v_1$ & \tiny $x^5y^4v_1$ & \tiny $x^6y^4v_1$ & \tiny $x^7y^4v_1$ & \\
+        \tiny $x^0y^3v_1$ & \tiny $x^1y^3v_1$ & \tiny $x^2y^3v_1$ & \tiny $x^3y^3v_1$ & \tiny $x^4y^3v_1$ & \tiny $x^5y^3v_1$ & \tiny $x^6y^3v_1$ & \tiny $x^7y^3v_1$ & \\
+        \tiny $x^0y^2v_1$ & \tiny $x^1y^2v_1$ & \tiny $x^2y^2v_1$ & \tiny $x^3y^2v_1$ & \tiny $x^4y^2v_1$ & \tiny $x^5y^2v_1$ & \tiny $x^6y^2v_1$ & \tiny $x^7y^2v_1$ & \\
+        \tiny $x^0y^1v_1$ & \tiny $x^1y^1v_1$ & \tiny $x^2y^1v_1$ & \tiny $x^3y^1v_1$ & \tiny $x^4y^1v_1$ & \tiny $x^5y^1v_1$ & \tiny $x^6y^1v_1$ & \tiny $x^7y^1v_1$ & \\
+        \tiny $x^0y^0v_1$ & \tiny $x^1y^0v_1$ & \tiny $x^2y^0v_1$ & \tiny $x^3y^0v_1$ & \tiny $x^4y^0v_1$ & \tiny $x^5y^0v_1$ & \tiny $x^6y^0v_1$ & \tiny $x^7y^0v_1$ & \\
+      };
+    \end{scope}
+  \end{tikzpicture}
+  \caption{\label{fig:good}A graphical depiction of a good generating family in
+  the special case~$n = 2, m = 1$. The hatched cells indicate vectors which
+  have already been removed from the family. If the vector in the red cell will
+  be found to be expressible as a linear combination of vectors with smaller
+  index (blue cells), it will be removed, along with the vectors in all cells
+  to the top and to the right of the red cell.}
+\end{figure}
+
+We induct on the shape of a good generating family, starting with~$(w_i)_{i \in I}$.
 

### 2018-04-16 paper-generic-freeness.tex
typo
The full proof, now even without stuff on weakly Noetherian modules

@@
 
-\begin{thm}Let~$A$ be a reduced ring. Let~$B$ be an~$A$-algebra of finite type.
-Let~$M$ be a finitely generated~$B$-module. If~$f = 0$ is the only element
-of~$A$ such that~$M[f^{-1}]$ is free as an~$A[f^{-1}]$-module, then~$1 = 0$
-in~$A$.\end{thm}
-
-\begin{proof}Let~$B$ be generated by~$(x_1,\ldots,x_n)$ as an~$A$-algebra and
-let~$M$ be generated by~$(v_1,\ldots,v_m)$ as an~$B$-module. We endow the set
-\[ I \defeq \{ (\ell, j_1,\ldots,j_n) \,|\,
-  \ell \in \{ 1,\ldots,m \},
-  j_1,\ldots,j_n \in \{ 0,1,\ldots \} \} \]
-with the lexicographic order. The family~$(w_i)_{i \in I} \defeq (x_1^{j_1}
-\ldots x_n^{j_n} v_\ell)_{(\ell,j_1,\ldots,j_n) \in I}$ thus generates~$M$ as
-an~$A$-module, and we'll call a subfamily~$(w_i)_{i \in I' \subseteq I}$
-\emph{good} if and only if for all~$i \in I$, the vector~$w_i$ is a linear
-combination of the vectors~$(w_{i'})_{i' \in I, i' \preceq i}$,
-and if~$(\ell,j_1,\ldots,j_n) \not\in I'$ implies~$(\ell,k_1,\ldots,k_n) \not\in
-I'$ for all~$k_1 \geq j_1, \ldots, k_n \geq j_n$. Figure~\ref{fig:good} shows
+\begin{proof}[Proof of Theorem~\ref{thm:algebraic}]
+Let~$B$ be generated by~$(x_1,\ldots,x_n)$ as an~$A$-algebra and
+let~$M$ be generated by~$(v_1,\ldots,v_m)$ as a~$B$-module. We endow the sets
+\begin{align*}
+  I &\defeq \{ (i_1,\ldots,i_n) \,|\, i_1,\ldots,i_n \in \{ 0,1,\ldots \} \}
+  \quad\text{and} \\
+  J &\defeq \{ (\ell, i_1,\ldots,i_n) \,|\, \ell \in \{ 1,\ldots,m \}, i_1,\ldots,i_n \in \{ 0,1,\ldots \} \}
+\end{align*}
+with the lexicographic order. The family~$(w_j)_{j \in J} \defeq (x_1^{i_1}
+\ldots x_n^{i_n} v_\ell)_{(\ell,i_1,\ldots,i_n) \in J}$ thus generates~$M$ as
+an~$A$-module, and we'll call a subfamily~$(w_j)_{j \in J' \subseteq J}$
+\emph{good} if and only if for all~$j \in J$, the vector~$w_j$ is a linear
+combination of the vectors~$(w_{j'})_{j' \in J, j' \preceq j}$,
+and if~$(\ell,i_1,\ldots,i_n) \not\in J'$ implies~$(\ell,k_1,\ldots,k_n) \not\in
+I'$ for all~$k_1 \geq i_1, \ldots, k_n \geq i_n$. Figure~\ref{fig:good} shows
 how a good generating family can look like.
@@
       (2,8) -- (2,5) -- (5,5) -- (5,3) -- (6,3) -- (6,2) -- (8,2);
+
+    \fill[fill=black] (2,5.25) -- (2,5) -- (2.25,5) -- (2.25,5.25);
+    \fill[fill=black] (5,3.25) -- (5,3) -- (5.25,3) -- (5.25,3.25);
+    \fill[fill=black] (6,2.25) -- (6,2) -- (6.25,2) -- (6.25,2.25);
 
@@
   index (blue cells), it will be removed, along with the vectors in all cells
-  to the top and to the right of the red cell.}
+  to the top and to the right of the red cell. Corners are indicated by the
+  small black squares.}
 \end{figure}
 
-We induct on the shape of a good generating family, starting with~$(w_i)_{i \in I}$.
+Similarly, we define when a subfamily of the canonical generating
+family~$(x_1^{i_1} \cdots x_n^{i_n})_{(i_1,\ldots,i_n) \in I}$ of~$B$ is good
+(which is just the special case~$m = 1$). We then proceed by induction on the
+shapes of a given good generating family~$(w_j)_{j \in J'}$ for~$M$ and a given
+good generating family~$(s_i)_{i \in I'}$ for~$B$, starting with the canonical
+ones.
 
-We show that~$(w_i)_{i \in I}$ is a basis of~$M$ by verifying the assumptions
+We show that~$(w_j)_{j \in J'}$ is a basis of~$M$ by verifying the assumptions
 of Lemma~\ref{lemma:basis}. Thus let~$g \in A$ be given such that one of
-the~$w_i$ is an $A[g^{-1}]$-linear combination of generators with smaller index
-in~$M[g^{-1}]$. Removing~$w_i = x_1^{j_1} \ldots x_n^{j_n} v_\ell$ and also all
-vectors~$x_1^{k_1} \ldots x_n^{k_n} v_\ell$ where~$k_1 \geq j_1, \ldots, k_n
-\geq j_n$, we obtain a subfamily which is still good. By induction, applied
+the~$w_j$ is an $A[g^{-1}]$-linear combination of generators with smaller index
+in~$M[g^{-1}]$. Removing~$w_j = x_1^{i_1} \cdots x_n^{i_n} v_\ell$ and also all
+vectors~$x_1^{k_1} \cdots x_n^{k_n} v_\ell$ where~$k_1 \geq i_1, \ldots,
+k_n \geq i_n$, we obtain a subfamily which is still good for
+the~$A[g^{-1}]$-module~$M[g^{-1}]$. By induction, applied
 to~$A[g^{-1}]$ and its module~$M[g^{-1}]$, it therefore follows that $A[g^{-1}] = 0$.
@@
 
-Thus~$M$ is free. We finish by using the assumption for~$f = 1$.
+Similarly, we show that the given good generating family~$(s_i)_{i \in I'}$ is
+a basis. Thus~$M$ and~$B$ are free over~$A$. We fix for any corner~$j =
+(\ell,i_1,\ldots,i_n) \in J$, as indicated in Figure~\ref{fig:good},
+a way of expressing~$x^{i_1} \cdots x^{i_n} v_\ell = \sum_{(p,k_1,\ldots,k_n)}
+a_{j; p,k_1,\ldots,k_n} x^{k_1} \cdots x^{k_n} v_p$ as a linear combination of
+generators with strictly smaller index. Then~$M$ is isomorphic to
+the~$B$-module
+\[ B\langle V_1,\ldots, V_m\rangle/
+  (x^{i_1} \cdots x^{i_n} V_\ell - \textstyle\sum_{(p,k_1,\ldots,k_n)}
+  a_{j; p,k_1,\ldots,k_n} x_{k_1} \cdots x_{k_n} V_p)_j,
+\]
+where the index~$j$ ranges over all corners of~$J$, and therefore finitely
+presented. In a similar vain, a quotient algebra of~$A[X_1,\ldots,X_n]$, where
+we mod out a suitable ideal with as many generators as corners of~$I$, is
+isomorphic to~$B$.
+
+We finish by using the assumption for~$f = 1$.
 \end{proof}
+
+\section{The proof of the finitely-generated case}
+
+The following proposition is just an instance of Grothendieck's generic
+freeness lemma. We prove it here because it admits an easier proof.
+
+\begin{prop}Let~$A$ be a reduced ring. Let~$M$ be a finitely
+generated~$A$-module. If~$f = 0$ is the only element of~$A$ such
+that~$M[f^{-1}]$ is a finite free~$A[f^{-1}]$-module, then~$1 = 0$ in~$A$.
+\end{prop}
+
+\begin{proof}We proceed by induction on the length of a given generating family
+of~$M$. Let~$M$ be generated by~$(v_1,\ldots,v_m)$.
+
+We show that the family~$(v_1,\ldots,v_m)$ is linearly independent. Let~$\sum_i
+a_i v_i = 0$. Over~$A[a_i^{-1}]$, the vector~$v_i \in M[a_i^{-1}]$ is a linear
+combination of the other generators. Thus~$M[a_i^{-1}]$ can be generated as
+an~$A[a_i^{-1}]$-module by fewer than~$m$ generators. The induction hypothesis,
+applied to this module, yields that~$1 = 0$ in~$A[a_i^{-1}]$. Since~$A$ is
+reduced, this amounts to~$a_i = 0$.
+
+We finish by using the assumption for~$f = 1$.
+\end{proof}
+
+\end{document}
 
@@
 
-\section{The proof of the finitely-generated case}
-
-The following proposition is just an instance of Grothendieck's generic
-freeness lemma. We prove it here because it admits an easier proof.
-
-\begin{prop}Let~$A$ be a reduced ring. Let~$M$ be a finitely
-generated~$A$-module. If~$f = 0$ is the only element of~$A$ such
-that~$M[f^{-1}]$ is a finite free~$A[f^{-1}]$-module, then~$1 = 0$ in~$A$.
-\end{prop}
-
-\begin{proof}We proceed by induction on the length of a given generating family
-of~$M$. Let~$M$ be generated by~$(v_1,\ldots,v_m)$.
-
-We show that the family~$(v_1,\ldots,v_m)$ is linear independent. Let~$\sum_i
-a_i v_i = 0$. Over~$A[a_i^{-1}]$, the vector~$v_i \in M[a_i^{-1}]$ is a linear
-combination of the other generators. Thus~$M[a_i^{-1}]$ can be generated as
-an~$A[a_i^{-1}]$-module by fewer than~$m$ generators. The induction hypothesis,
-applied to this module, yields that~$1 = 0$ in~$A[a_i^{-1}]$. Since~$A$ is
-reduced, this amounts to~$a_i = 0$.
-
-We finish by using the assumption for~$f = 1$.
-\end{proof}
-
 \end{document}

### 2018-04-17 paper-generic-freeness.tex
References

@@
 
-\title{An elementary and constructive proof of Grothendieck's generic freeness
-lemma}
+\usepackage[natbib=true,style=numeric,maxnames=10]{biblatex}
+\usepackage[babel]{csquotes}
+\bibliography{paper-generic-freeness.bib}
+
+\title[Grothendieck's generic freeness lemma]{An elementary and constructive
+proof of Grothendieck's generic freeness lemma}
 \author{Ingo Blechschmidt}
@@
 \DeclareMathOperator{\Spec}{Spec}
+
+\newcommand{\stacksproject}[1]{\cite[{\href{https://stacks.math.columbia.edu/tag/#1}{Tag~#1}}]{stacks-project}}
 
@@
 
-Previously known proofs proceed in a series of intermediate steps, reducing
-to the case that~$A$ is a Noetherian integral domain where one can argue by
-\emph{dévissage}; but in fact, a direct proof is possible and shorter.
+Previously known proofs either only cover the case where~$A$ is a Noetherian
+integral domain, where one can argue by \emph{dévissage} (see for
+instance~\cite[Lemme~6.9.2]{ega-4-2},
+\cite[Thm.~24.1]{matsumura:commutative-ring-theory}, or
+\cite[Thm.~14.4]{eisenbud:commutative-algebra}), or proceed in a series of
+intermediate steps, reducing to that case (see for
+instance~\cite{staats:generic-freeness} or~\stacksproject{051Q}); but in fact,
+a direct proof is possible and shorter.
 
@@
 yielding a self-contained argument without any references to topos theory.
-We refer readers who want to learn about this technique to an accompanying
-note~\cite{blechschmidt:wlog}.
+We refer readers who want to learn about this technique to a companion 
+paper~\cite{blechschmidt:wlog}.
 
-\XXX{Acknowledgments, especially Martin Brandenburg}
-
-\XXX{References to prior proofs: EGA, Stacks Project, Charles Staats. Check
-\href{https://math.stackexchange.com/questions/1048496/question-on-generic-freeness-ref-matsumura-page-185}{this
-MO question}}
+\textbf{Acknowledgments.} The proof presented here was prompted by
+user~HeinrichD on MathOverflow~\cite{mo:kernel} and greatly
+benefited from discussions with Martin Brandenburg, who needed the constructive
+version in paper of him~\cite{brandenburg:schur}. I'm grateful to Marc
+Nieper-Wißkirchen for carefully guiding my PhD thesis, where \XXX{...}
 
@@
 with the lexicographic order. The family~$(w_j)_{j \in J} \defeq (x_1^{i_1}
-\ldots x_n^{i_n} v_\ell)_{(\ell,i_1,\ldots,i_n) \in J}$ thus generates~$M$ as
+\cdots x_n^{i_n} v_\ell)_{(\ell,i_1,\ldots,i_n) \in J}$ generates~$M$ as
 an~$A$-module, and we'll call a subfamily~$(w_j)_{j \in J' \subseteq J}$
@@
 
-\begin{figure}
+\begin{figure}[b]
   \begin{tikzpicture}[scale=1.0]
@@
   the special case~$n = 2, m = 1$. The hatched cells indicate vectors which
-  have already been removed from the family. If the vector in the red cell will
-  be found to be expressible as a linear combination of vectors with smaller
-  index (blue cells), it will be removed, along with the vectors in all cells
-  to the top and to the right of the red cell. Corners are indicated by the
-  small black squares.}
+  have already been removed from the family. The small black squares indicate
+  \emph{corners}. If the vector in the red cell will be found to be expressible
+  as a linear combination of vectors with smaller index (blue cells), it will
+  be removed, along with the vectors in all cells to the top and to the right
+  of the red cell.}
 \end{figure}
@@
 \]
-where the index~$j$ ranges over all corners of~$J$, and therefore finitely
+where the index~$j$ ranges over all corners of~$J$, and is therefore finitely
 presented. In a similar vain, a quotient algebra of~$A[X_1,\ldots,X_n]$, where
@@
 The following proposition is just an instance of Grothendieck's generic
-freeness lemma. We prove it here because it admits an easier proof.
+freeness lemma. We present it here because it admits an easier proof.
 
@@
 \end{proof}
+
+\printbibliography
 

### 2018-04-17 paper-generic-freeness.tex
typo

@@
 
+\noindent
 We prove Grothendieck's generic freeness lemma for rings and modules in the
@@
 benefited from discussions with Martin Brandenburg, who needed the constructive
-version in paper of him~\cite{brandenburg:schur}. I'm grateful to Marc
+version in a paper of his~\cite{brandenburg:schur}. I'm grateful to Marc
 Nieper-Wißkirchen for carefully guiding my PhD thesis, where \XXX{...}

### 2018-05-15 paper-generic-freeness.tex
Corrected minor typos

@@
 \item the~$A[f^{-1}]$-modules $B[f^{-1}]$ and $M[f^{-1}]$ are free,
-\item the~$A[f^{-1}]$-algebra~$B[f^{-1}]$ is of finite presentation, and
+\item the~$A[f^{-1}]$-algebra~$B[f^{-1}]$ is of finite presentation and
 \item the~$B[f^{-1}]$-module~$M[f^{-1}]$ is finitely presented,
@@
 instance~\cite[Lemme~6.9.2]{ega-4-2},
-\cite[Thm.~24.1]{matsumura:commutative-ring-theory}, or
+\cite[Thm.~24.1]{matsumura:commutative-ring-theory} or
 \cite[Thm.~14.4]{eisenbud:commutative-algebra}), or proceed in a series of
@@
 \item[(a)] $B^\sim$ and~$M^\sim$ are free as sheaves of~$A^\sim$-modules,
-\item[(b)] $B^\sim$ is of finite presentation as a sheaf of~$A^\sim$-algebras, and
+\item[(b)] $B^\sim$ is of finite presentation as a sheaf of~$A^\sim$-algebras and
 \item[(c)] $M^\sim$ is finitely presented as a sheaf of~$B^\sim$-modules.
@@
 Theorem~\ref{thm:geometric} immediately follows from
-Theorem~\ref{thm:algebraic} by defining~$U$ as the union of all the basic
-opens~$D(f)$ such that~(1),~(2), and~(3) hold. It's clear that~(a),~(b),
+Theorem~\ref{thm:algebraic} by defining~$U$ as the union of all those basic
+opens~$D(f)$ such that~(1),~(2) and~(3) hold. It's clear that~(a),~(b)
 and~(c) hold over~$U$, and~$U$ is dense for if~$V$ is an arbitrary open
-such that~$U \cap V = \emptyset$, the open~$V$ is itself empty: Let~$h \in A$
+such that~$U \cap V = \emptyset$, the open~$V$ is itself empty: Let~$h \in A$ be
 such that~$D(h) \subseteq V$. The hypothesis implies the assumptions of
@@
 yielding a self-contained argument without any references to topos theory.
-We refer readers who want to learn about this technique to a companion 
+We refer readers who want to learn about this technique to a forthcoming companion 
 paper~\cite{blechschmidt:wlog}.
@@
 user~HeinrichD on MathOverflow~\cite{mo:kernel} and greatly
-benefited from discussions with Martin Brandenburg, who needed the constructive
+benefited from discussions with Martin Brandenburg, who employed the constructive
 version in a paper of his~\cite{brandenburg:schur}. I'm grateful to Marc
-Nieper-Wißkirchen for carefully guiding my PhD thesis, where \XXX{...}
+Nieper-Wißkirchen for carefully guiding my PhD studies at the University of
+Augsburg, where most of the work for this paper was carried out.
 
@@
   \end{tikzpicture}
-  \caption{\label{fig:good}A graphical depiction of a good generating family in
-  the special case~$n = 2, m = 1$. The hatched cells indicate vectors which
+  \caption{\label{fig:good}A graphical depiction of a good generating family (the non-hatched cells) in
+  the special case~$n = 2, m = 1$, writing~``$x$'' and~``$y$'' for~$x_1$ and~$x_2$. The hatched cells indicate vectors which
   have already been removed from the family. The small black squares indicate
@@
 k_n \geq i_n$, we obtain a subfamily which is still good for
-the~$A[g^{-1}]$-module~$M[g^{-1}]$. By induction, applied
+the~$A[g^{-1}]$-module~$M[g^{-1}]$. By the induction hypothesis, applied
 to~$A[g^{-1}]$ and its module~$M[g^{-1}]$, it therefore follows that $A[g^{-1}] = 0$.
@@
 presented. In a similar vain, a quotient algebra of~$A[X_1,\ldots,X_n]$, where
-we mod out a suitable ideal with as many generators as corners of~$I$, is
+we mod out a suitable ideal with as many generators as corners of~$I'$, is
 isomorphic to~$B$.

### 2018-05-15 paper-generic-freeness.tex
Moved the special case to the front

@@
 
+\section{The proof of the finitely-generated case}
+
+The following proposition is just a special instance of Grothendieck's generic
+freeness lemma. We include it here because we want to present its proof, which
+is easier and shorter than the proof of the general case. The general proof
+will be self-contained and not refer to this one.
+
+\begin{prop}Let~$A$ be a reduced ring. Let~$M$ be a finitely
+generated~$A$-module. If~$f = 0$ is the only element of~$A$ such
+that~$M[f^{-1}]$ is a finite free~$A[f^{-1}]$-module, then~$1 = 0$ in~$A$.
+\end{prop}
+
+\begin{proof}We proceed by induction on the length of a given generating family
+of~$M$. Let~$M$ be generated by~$(v_1,\ldots,v_m)$.
+
+We show that the family~$(v_1,\ldots,v_m)$ is linearly independent. Let~$\sum_i
+a_i v_i = 0$. Over~$A[a_i^{-1}]$, the vector~$v_i \in M[a_i^{-1}]$ is a linear
+combination of the other generators. Thus~$M[a_i^{-1}]$ can be generated as
+an~$A[a_i^{-1}]$-module by fewer than~$m$ generators. The induction hypothesis,
+applied to this module, yields that~$1 = 0$ in~$A[a_i^{-1}]$. Since~$A$ is
+reduced, this amounts to~$a_i = 0$.
+
+We finish by using the assumption for~$f = 1$.
+\end{proof}
+
 \section{The proof of the general case}
@@
 
-\section{The proof of the finitely-generated case}
-
-The following proposition is just an instance of Grothendieck's generic
-freeness lemma. We present it here because it admits an easier proof.
-
-\begin{prop}Let~$A$ be a reduced ring. Let~$M$ be a finitely
-generated~$A$-module. If~$f = 0$ is the only element of~$A$ such
-that~$M[f^{-1}]$ is a finite free~$A[f^{-1}]$-module, then~$1 = 0$ in~$A$.
-\end{prop}
-
-\begin{proof}We proceed by induction on the length of a given generating family
-of~$M$. Let~$M$ be generated by~$(v_1,\ldots,v_m)$.
-
-We show that the family~$(v_1,\ldots,v_m)$ is linearly independent. Let~$\sum_i
-a_i v_i = 0$. Over~$A[a_i^{-1}]$, the vector~$v_i \in M[a_i^{-1}]$ is a linear
-combination of the other generators. Thus~$M[a_i^{-1}]$ can be generated as
-an~$A[a_i^{-1}]$-module by fewer than~$m$ generators. The induction hypothesis,
-applied to this module, yields that~$1 = 0$ in~$A[a_i^{-1}]$. Since~$A$ is
-reduced, this amounts to~$a_i = 0$.
-
-We finish by using the assumption for~$f = 1$.
-\end{proof}
-
 \printbibliography

### 2018-05-15 paper-generic-freeness.tex
Improved the presentation; fought the index war.

@@
 \usepackage[utf8]{inputenc}
-\usepackage{amsthm,mathtools,tikz}
+\usepackage{amsthm,mathtools,tikz,float,caption}
 \usetikzlibrary{patterns,matrix}
@@
 \newcommand{\bbb}{\mathfrak{b}}
+\newcommand{\I}{\mathcal{I}}
+\newcommand{\J}{\mathcal{J}}
 \newcommand{\defeq}{\vcentcolon=}
@@
 The following proposition is just a special instance of Grothendieck's generic
-freeness lemma. We include it here because we want to present its proof, which
-is easier and shorter than the proof of the general case. The general proof
-will be self-contained and not refer to this one.
+freeness lemma. Its proof is easier and shorter than the proof of the general
+case, which is why we present it here. The general proof
+will not refer to this one.
 
@@
 
+We remark that proof takes a somewhat curious course: Our goal is to
+verify~$1 = 0$, but as an intermediate step we verify that~$M$ is free, which
+after the fact will be a trivial statement.
+
 \section{The proof of the general case}
-
-\begin{lemma}\label{lemma:basis}
-Let~$A$ be a ring. Let~$M$ be an~$A$-module with generating
-family~$(x_i)_{i \in I}$ where~$I$ is a totally ordered set. Assume that the
-only element~$g \in A$ such that one of the~$x_i$ is an~$A[g^{-1}]$-linear
-combination in~$M[g^{-1}]$ of other generators with smaller index is~$g = 0$.
-Then~$M$ is free with~$(x_i)_{i \in I}$ as a basis.
-\end{lemma}
-
-\begin{proof}Let~$\sum_i a_i x_i = 0$. Starting with the greatest
-index~$i$ which appears in that sum, we see that in~$M[a_i^{-1}]$, the
-element~$x_i$ is an~$A[g^{-1}]$-linear combination of other generators with
-smaller index. Thus~$a_i = 0$ by assumption.\end{proof}
 
@@
 \begin{align*}
-  I &\defeq \{ (i_1,\ldots,i_n) \,|\, i_1,\ldots,i_n \in \{ 0,1,\ldots \} \}
+  \I &\defeq \{ (i_1,\ldots,i_n) \,|\, i_1,\ldots,i_n \geq 0 \}
   \quad\text{and} \\
-  J &\defeq \{ (\ell, i_1,\ldots,i_n) \,|\, \ell \in \{ 1,\ldots,m \}, i_1,\ldots,i_n \in \{ 0,1,\ldots \} \}
+  \J &\defeq \{ (\ell, i_1,\ldots,i_n) \,|\, 1 \leq \ell \leq m, i_1,\ldots,i_n \geq 0 \}
 \end{align*}
-with the lexicographic order. The family~$(w_j)_{j \in J} \defeq (x_1^{i_1}
-\cdots x_n^{i_n} v_\ell)_{(\ell,i_1,\ldots,i_n) \in J}$ generates~$M$ as
-an~$A$-module, and we'll call a subfamily~$(w_j)_{j \in J' \subseteq J}$
-\emph{good} if and only if for all~$j \in J$, the vector~$w_j$ is a linear
-combination of the vectors~$(w_{j'})_{j' \in J, j' \preceq j}$,
-and if~$(\ell,i_1,\ldots,i_n) \not\in J'$ implies~$(\ell,k_1,\ldots,k_n) \not\in
-I'$ for all~$k_1 \geq i_1, \ldots, k_n \geq i_n$. Figure~\ref{fig:good} shows
+with the lexicographic order. The family~$(w_J)_{J \in \J} \defeq (x_1^{i_1}
+\cdots x_n^{i_n} v_\ell)_{(\ell,i_1,\ldots,i_n) \in \J}$ generates~$M$ as
+an~$A$-module, and we'll call a subfamily~$(w_J)_{J \in \J' \subseteq \J}$
+\emph{good} if and only if for all~$J \in \J$, the vector~$w_J$ is a linear
+combination of the vectors~$(w_{J'})_{J' \in \J', J' \preceq J}$,
+and if~$(\ell,i_1,\ldots,i_n) \not\in \J'$ implies~$(\ell,k_1,\ldots,k_n) \not\in
+\J'$ for all~$k_1 \geq i_1, \ldots, k_n \geq i_n$. Figure~\ref{fig:good} shows
 how a good generating family can look like.
+Similarly, we define when a subfamily of the canonical generating
+family~$(x_1^{i_1} \cdots x_n^{i_n})_{(i_1,\ldots,i_n) \in \I}$ of~$B$ is good
+(which is just the special case~$m = 1$).
 
-\begin{figure}[b]
-  \begin{tikzpicture}[scale=1.0]
+\begin{figure}[t]
+  \captionsetup{width=.9\linewidth}
+  \begin{tikzpicture}[scale=0.9]
     \draw[step=1cm,gray,very thin] (0,0) grid (8,8);
@@
     \begin{scope}[yshift=4cm, xshift=4cm]
-      \matrix[matrix of nodes,nodes={inner sep=0pt,text width=1.0cm,align=center,minimum height=1.0cm}]{
+      \matrix[matrix of nodes,nodes={inner sep=0pt,text width=0.9cm,align=center,minimum height=0.9cm}]{
         \tiny $x^0y^7v_1$ & \tiny $x^1y^7v_1$ & \tiny $x^2y^7v_1$ & \tiny $x^3y^7v_1$ & \tiny $x^4y^7v_1$ & \tiny $x^5y^7v_1$ & \tiny $x^6y^7v_1$ & \tiny $x^7y^7v_1$ & \\
@@
 
-Similarly, we define when a subfamily of the canonical generating
-family~$(x_1^{i_1} \cdots x_n^{i_n})_{(i_1,\ldots,i_n) \in I}$ of~$B$ is good
-(which is just the special case~$m = 1$). We then proceed by induction on the
-shapes of a given good generating family~$(w_j)_{j \in J'}$ for~$M$ and a given
-good generating family~$(s_i)_{i \in I'}$ for~$B$, starting with the canonical
-ones.
+We then proceed by induction on the
+shapes of a given good generating family~$(w_J)_{J \in \J'}$ for~$M$ and a given
+good generating family~$(s_I)_{I \in \I'}$ for~$B$, starting with the canonical ones. It's
+reasonably obvious that this induction is well-founded; the formal statement
+that it is so is known as \emph{Dickson's Lemma} in mathematical logic.
 
-We show that~$(w_j)_{j \in J'}$ is a basis of~$M$ by verifying the assumptions
-of Lemma~\ref{lemma:basis}. Thus let~$g \in A$ be given such that one of
-the~$w_j$ is an $A[g^{-1}]$-linear combination of generators with smaller index
-in~$M[g^{-1}]$. Removing~$w_j = x_1^{i_1} \cdots x_n^{i_n} v_\ell$ and also all
+We show that~$(w_J)_{J \in \J'}$ is a basis of~$M$ by verifying linear
+independence. Thus let~$\sum_J a_J w_J = 0$ in~$M$. We show that all
+coefficients in this sum are zero, starting with the largest
+appearing index~$J$: In the module~$M[a_J^{-1}]$ over the localized ring~$A[a_J^{-1}]$,
+the vector~$w_J$ is a linear combination of generators with
+smaller index. Removing~$w_J = x_1^{i_1} \cdots x_n^{i_n} v_\ell$ and also all
 vectors~$x_1^{k_1} \cdots x_n^{k_n} v_\ell$ where~$k_1 \geq i_1, \ldots,
 k_n \geq i_n$, we obtain a subfamily which is still good for
-the~$A[g^{-1}]$-module~$M[g^{-1}]$. By the induction hypothesis, applied
-to~$A[g^{-1}]$ and its module~$M[g^{-1}]$, it therefore follows that $A[g^{-1}] = 0$.
-This implies that $g = 0$ since~$A$ is reduced.
+the localized module. The induction hypothesis, applied
+to~$A[a_J^{-1}]$ and its module~$M[a_J^{-1}]$, therefore implies that $A[a_J^{-1}] = 0$.
+Thus $a_J = 0$ since~$A$ is reduced.
+\enlargethispage{2em}
 
-Similarly, we show that the given good generating family~$(s_i)_{i \in I'}$ is
-a basis. Thus~$M$ and~$B$ are free over~$A$. We fix for any corner~$j =
-(\ell,i_1,\ldots,i_n) \in J$, as indicated in Figure~\ref{fig:good},
-a way of expressing~$x^{i_1} \cdots x^{i_n} v_\ell = \sum_{(p,k_1,\ldots,k_n)}
-a_{j; p,k_1,\ldots,k_n} x^{k_1} \cdots x^{k_n} v_p$ as a linear combination of
-generators with strictly smaller index. Then~$M$ is isomorphic to
-the~$B$-module
-\[ B\langle V_1,\ldots, V_m\rangle/
-  (x^{i_1} \cdots x^{i_n} V_\ell - \textstyle\sum_{(p,k_1,\ldots,k_n)}
-  a_{j; p,k_1,\ldots,k_n} x_{k_1} \cdots x_{k_n} V_p)_j,
+Similarly, we show that the given good generating family~$(s_I)_{I \in \I'}$ is
+a basis. Thus~$M$ and~$B$ are free over~$A$. We fix for any corner~$J$
+of~$\J'$, as indicated in Figure~\ref{fig:good},
+a way of expressing~$w_J = \sum_K a_{JK} w_K$ as a linear combination of
+generators with strictly smaller index. Let~$\widehat w_{(\ell,i_1,\ldots,i_n)}
+\defeq x_1^{i_1} \cdots x_n^{i_n} V_\ell$ in the free~$B$-module~$B\langle
+V_1,\ldots, V_m\rangle$. The canonical map
+\[ \widehat M \defeq B\langle V_1,\ldots, V_m\rangle/
+  (\widehat w_J - \textstyle\sum_K a_{JK} \widehat w_K)_{\text{$J$ corner of~$\J'$}} \longrightarrow M
 \]
-where the index~$j$ ranges over all corners of~$J$, and is therefore finitely
-presented. In a similar vain, a quotient algebra of~$A[X_1,\ldots,X_n]$, where
-we mod out a suitable ideal with as many generators as corners of~$I'$, is
-isomorphic to~$B$.
+is trivially well-defined and surjective. It is also injective, since any
+element of~$\widehat M$ can be written as an~$A$-linear combination of the
+vectors~$(\widehat w_J)_{J \in \J'}$ by employing the corner relations a finite
+number of times.  Therefore~$M$ is finitely presented as a~$B$-module.
+
+In a similar vein, a quotient algebra of~$A[X_1,\ldots,X_n]$, where
+we mod out by a suitable ideal with as many generators as corners of~$\I'$, is
+isomorphic to~$B$. Thus~$B$ is finitely presented as an~$A$-algebra.
 

### 2018-05-16 paper-generic-freeness.tex
Added pointer to Richman's jewel

@@
 
-We remark that proof takes a somewhat curious course: Our goal is to
+We remark that the proof takes a somewhat curious course: Our goal is to
 verify~$1 = 0$, but as an intermediate step we verify that~$M$ is free, which
-after the fact will be a trivial statement.
+after the fact will be a trivial statement. This approach is reminiscient of
+Richman's uses of trivial rings~\cite{richman:trivial-rings}.
+\enlargethispage{1em}
 
@@
 
+\nocite{blechschmidt:phd}
 \printbibliography

### 2018-05-16 paper-generic-freeness.tex
Final adjustments

@@
   proceed in a series of reduction steps and is fully constructive, not
-  involving the axiom of choice or even the law of excluded middle. It was
+  using the axiom of choice or even the law of excluded middle. It was
   found by unwinding the result of a general topos-theoretic technique.
@@
 \noindent
-We prove Grothendieck's generic freeness lemma for rings and modules in the
+We prove Grothendieck's generic freeness lemma in the
 following form.
@@
 instance~\cite{staats:generic-freeness} or~\stacksproject{051Q}); but in fact,
-a direct proof is possible and shorter.
+a direct proof is possible and shorter. The new proof unveils a certain
+combinatorial aspect to Grothendieck's generic freeness lemma, doesn't require
+any advanced prerequisites in commutative algebra and doesn't use the axiom of
+choice or the law of excluded middle. It is purely element-based, not referring
+to ideals of~$A$.
+\enlargethispage{3em}
 
@@
 
-The new proof was found using a general topos-theoretical technique which we
+The new proof was found using a general topos-theoretic technique which we
 believe to be useful in other situations as well. This technique allows to view
@@
 benefited from discussions with Martin Brandenburg, who employed the constructive
-version in a paper of his~\cite{brandenburg:schur}. I'm grateful to Marc
-Nieper-Wißkirchen for carefully guiding my PhD studies at the University of
+version in a paper of his~\cite{brandenburg:schur}. I'm grateful to Thierry
+Coquand and Peter Schuster for valuable advice and to Marc
+Nieper-Wißkirchen for carefully guiding my PhD studies~\cite{blechschmidt:phd} at the University of
 Augsburg, where most of the work for this paper was carried out.
@@
 
-\nocite{blechschmidt:phd}
 \printbibliography

### 2018-05-17 paper-wlog.tex
Beginnings of the "without loss of generality" paper

\documentclass{amsart}

\usepackage[utf8]{inputenc}
\usepackage{amsthm,mathtools,tikz,float,caption,xspace}
\usetikzlibrary{patterns,matrix}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{hyperref}
\usepackage{cleveref}

\usepackage[natbib=true,style=numeric,maxnames=10]{biblatex}
\usepackage[babel]{csquotes}
\bibliography{paper-wlog.bib}

\title{Without loss of generality, \\ any reduced ring is a field}
\author{Ingo Blechschmidt}

\theoremstyle{definition}
\newtheorem{defn}{Definition}[]
\newtheorem{ex}[defn]{Example}

\theoremstyle{plain}
\newtheorem{prop}[defn]{Proposition}
\newtheorem{cor}[defn]{Corollary}
\newtheorem{lemma}[defn]{Lemma}
\newtheorem{thm}[defn]{Theorem}
\newtheorem{scholium}[defn]{Scholium}

\theoremstyle{remark}
\newtheorem{rem}[defn]{Remark}
\newtheorem{question}[defn]{Question}
\newtheorem{speculation}[defn]{Speculation}
\newtheorem{caveat}[defn]{Caveat}
\newtheorem{conjecture}[defn]{Conjecture}

\crefname{defn}{Definition}{Definitions}
\crefname{thm}{Theorem}{Theorems}
\crefname{lemma}{Lemma}{Lemmas}
\crefname{prop}{Proposition}{Propositions}
\crefname{ex}{Example}{Examples}
\crefname{section}{Section}{Sections}

% Workaround from David Carlisle described at https://tex.stackexchange.com/a/316037
\makeatletter
\let\xx@thm\@thm
\AtBeginDocument{\let\@thm\xx@thm}
\makeatother

\newcommand{\XXX}[1]{\textbf{XXX: #1}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\defeq}{\vcentcolon=}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Quot}{Quot}
\newcommand{\?}{\,{:}\,}
\newcommand{\notnot}{\emph{not not}\xspace}
\newcommand{\negg}{\neg\!\!\!\neg}

\newcommand{\stacksproject}[1]{\cite[{\href{https://stacks.math.columbia.edu/tag/#1}{Tag~#1}}]{stacks-project}}

\begin{document}

\begin{abstract}
  XXX
\end{abstract}

\maketitle
\thispagestyle{empty}

\noindent
XXX

\section{Case studies}

\subsection{Linear surjections between finite free modules}

\begin{prop}\label{surjection}
Let~$M \in A^{n \times m}$ be a matrix over a ring~$A$ such that~$n
> m$. Assume that the induced linear map~$A^m \to A^n$ is surjective. Then~$1 =
0 \in A$.
\end{prop}

\begin{proof}[Proof~A (classical)]Assume to the contrary that~$1 \neq 0 \in A$.
Then there exists a maximal ideal~$\mmm \subseteq A$. Tensoring the given linear map
over~$A$ with~$A/\mmm$ yields a linear map~$(A/\mmm)^m \to (A/\mmm)^n$ which is
still surjective, since tensoring is right exact. Since~$A/\mmm$ is a field,
this is a contradiction by basic linear algebra.
\end{proof}

This proof uses the axiom of choice in the guise of the lemma that nontrivial
rings contain maximal ideals in order to reduce from general rings to
fields.\footnote{It's possible to avoid the full axiom of choice here and only
assume the Boolean Prime Ideal Theorem, at the expense of slightly complicating
the proof. This theorem is strong enough to conclude that a nontrivial ring has
a prime ideal~$\ppp$. One then first passes to the quotient~$A/\ppp$, where
surjectivity is preserved thanks to right-exactness of tensoring, and then to
the field of fractions~$\Quot(A/\ppp)$, where surjectivity is preserved thanks
to exactness of localization.} In fact, the statement even admits a fully
constructive proof. It's not hard to give such a proof explicitly, for instance
using determinants~\cite{richman:trivial-rings}, but in this section we want to
see how the forcing technique can be used to translate the classical proof into
a constructive one.

Starting point is the observation that \Cref{surjection} admits an easy proof
in the case that the ring~$A$ is local:

\begin{proof}[Proof~B (constructive, but only of the special case
that~$A$ is a local ring)]
Since~$M$ is surjective, some linear combination of its columns sums to the
first canonical basis vector~$e_1$. Since~$A$ is local, some entry of the first row
is invertible. By employing elementary row and column transformations, we can
change this entry to the unit of the ring, move it to the top left, and kill
any remaining entries in the first row and in the first column. Continuing in
this fashion, we can assume~$A$ to be a rectangular diagonal matrix. Since~$A$
has more rows than columns, all entries in the last row are zero. On the other
hand, one of its entries is the unit. Thus~$1 = 0$.
\end{proof}

\begin{proof}[Proof~C (constructive, general case)]
By~XXX, the given matrix~$M$ is also surjective as a matrix over~$A^\sim$.
Since~$A^\sim$ is a local ring, the previous proof applies and shows
that~$A^\sim \models 1 = 0 \? A^\sim$. This is equivalent to~$1 = 0 \in A$ by
definition.
\end{proof}

This attempt succeeds in eliminating the axiom of choice and the law of
excluded middle; but it isn't yet a faithful translation of Proof~A, since the
standard linear algebra proof referenced in Proof~A proceeds slightly
differently than Proof~B. To obtain such a translation, we use the special
field property of~$A^\sim$.

\begin{proof}[Proof~D (constructive, general case)]
By passing from~$A$ to its quotient ring~$A/\sqrt{(0)}$, which preserves
surjectivity since tensoring is right-exact, we may assume that~$A$ is a
reduced ring. Thus~$A^\sim$ is a field in the sense that non-invertible
elements are zero. It therefore suffices to give a constructive proof of
\Cref{surjection} in the special that~$A$ is such a kind of field.

Every entry of the given matrix is \notnot invertible or not invertible. Since
the statement we want to verify,~$1 = 0$, is~$\neg\neg$-stable, we may
assume that every entry is actually invertible or not invertible. If
one of the entries is invertible, we can change it to the unit, move it to the
top left, kill the remaining entries in the first row and in the first
column, and then continue in this fashion. We may thus assume that~$M$ is a
rectangular diagonal matrix having only zeros and ones on the diagonal.

XXX
\end{proof}

\subsection{Linear injections between finite free modules: McCoy's theorem}

\begin{proof}Let~$a \? A^\sim$ such that~$a \cdot (\Lambda^m M) = (0)$. We want
to verify that~$a = 0$. Instead of employing~$\neg\neg$ like on previous
occasions, we employ~$\negg\negg$, where~$\negg\varphi$ doesn't mean~$(\varphi
\Rightarrow \bot)$, but~$(\varphi \Rightarrow a = 0)$.

For any entry~$M_{ij}$ of the matrix, we~$\negg\negg$-have that it's invertible
or~$\negg($invertible$)$. The latter is equivalent to some power XXX
\end{proof}

\subsection{Nilpotent polynomials}

XXX

\subsection{Grothendieck's generic freeness lemma}

XXX

\end{document}

### 2018-05-22 slides-filmat2018.tex
Beginning of the FilMat 2018 slides

--- 2018-04-06 slides-pssl103.tex
@@
-\documentclass[12pt,utf8,notheorems,compress,t]{beamer}
+\documentclass[13pt,utf8,notheorems,compress,t]{beamer}
 \usepackage{etex}
@@
 \usepackage{xstring}
+\usepackage{ifthen}
 \usepackage{soul}\setul{0.3ex}{}
@@
 \usepackage{tikz}
-\usetikzlibrary{calc,shapes.callouts,shapes.arrows,patterns}
+\usetikzlibrary{calc,shapes.callouts,shapes.arrows,patterns,fit,backgrounds}
 \hypersetup{colorlinks=true}
@@
 
-\title{Synthetic algebraic geometry and the phenomenon of nongeometric sequents}
+\title{Exploring mathematical objects from custom-tailored mathematical universes}
 \author{Ingo Blechschmidt}
-\date{April 7th, 2018}
+\date{May 26th, 2018}
 
@@
 
-  \hil{Synthetic algebraic geometry} \\
-  \small\emph{a case study in applied topos theory}
-
-  \large\&
-
-  \normalsize
-
-  \hil{the phenomenon of nongeometric sequents}
+  \hil{Exploring mathematical objects from custom-tailored mathematical universes} \\
 
@@
   \scriptsize
+  Max Planck Institute for Mathematics in the Sciences \\
   Ingo Blechschmidt \\
-  University of Augsburg
+
   \medskip
 
-  103rd Peripatetic Seminar on Sheaves and Logic in Brno \\
-  April 7th, 2018
+  Third international conference of the Italian Network for the Philosophy of
+  Mathematics \\
+  May 26th, 2018
   \par
@@
 
-\section{Preliminaries}
+\begin{frame}[fragile]{A glimpse of the toposophic landscape}
+  \tikzstyle{topos} = [draw=mypurple, very thick,
+      rectangle, rounded corners, inner sep=10pt, inner ysep=20pt]
+  \tikzstyle{title} = [fill=mypurple, text=white]
 
-\subsection{Approaches to algebraic geometry}
+  \input{images/primes.tex}
 
-\begin{frame}{Approaches to algebraic geometry}
-  Usual approach to algebraic geometry: \hil{layer schemes above ordinary set theory}
-  using either
-  \begin{itemize}
-    \item locally ringed spaces
-    \small
-    \begin{multline*}
-      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$} + {} \\
-      \text{Zariski topology} + \text{structure sheaf}
-    \end{multline*}
-    \normalsize
-    \item or Grothendieck's functor-of-points account, where a scheme is a functor~$\mathrm{Ring} \to \mathrm{Set}$.
-    \small\[ A \longmapsto \{ (x,y,z) \in A^3 \,|\, x^n+y^n-z^n=0 \} \]
-  \end{itemize}
-  \bigskip
-
-  \hil{Synthetic approach:} model schemes \hil{directly as sets} in a certain
-  nonclassical set theory.
-  \small
-  \[ \{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]
-\end{frame}
-
-\subsection{Toposes as universes}
-
-\begin{frame}{Toposes as mathematical universes}
-  A \hil{topos} is a category which has finite limits,
-  is cartesian closed and has a subobject classifier, for instance
-  \begin{itemize}
-    \item \hil{$\boldsymbol{\Set}$}, the category of sets;
-    \item \hil{$\boldsymbol{\Sh(X)}$}, the category of set-valued sheaves over
-    a space~$X$;
-    \item \hil{$\boldsymbol{\Eff}$}, the ef{}fective topos (roughly: a
-    category of data types).
-  \end{itemize}
-
-  \begin{varblock}{9cm}{}Any topos supports an \hil{internal language},
-  which is sound with respect to
-  \only<1>{\hil{intuitionistic reasoning}}%
-  \only<2->{\hil{\pointthis{intuitionistic reasoning}{
-    \textnormal{\textcolor{black}{
-    no $\varphi \vee \neg\varphi$,\ \
-    no $\neg\neg\varphi \Rightarrow \varphi$,\ \
-    no axiom of choice}}}}}.\end{varblock}
-\end{frame}
-
-\newcommand{\intex}[3]{#1 \quad #2\par #3\bigskip\medskip}
-
-\begin{frame}{Curious universes}
-  \begin{changemargin}{-1.1em}{0em}
-    \small
-    \begin{itemize}
-      \item \intex{
-        $\Eff \models \text{``There are infinitely many prime numbers.''}$
-      }{\textcolor{green!90}{\cmark}}{
-        External meaning: There is a \hil{Turing machine} producing arbitrarily many prime
-        numbers.
+  \newcommand{\drawbox}[3]{
+    \node[topos] [fit = #3] (#1) {};
+    \node[title] at (#1.north) {#2};
   }
 
-      \item \intex{
-        $\Eff \models \text{``Any Turing machine halts or doesn't halt.''}$
-      }{\textcolor{red!80}{\xmark}}{
-        External meaning: There is a \hil{halting oracle} which determines whether
-        any given machine halts or doesn't halt.
+  \newcommand{\muchstuff}{
+    \includegraphics[height=3em]{filmat}
+    \scalebox{0.5}{\sieve{14}{2}}
   }
 
-      \item \intex{
-        $\Sh(X) \models \text{``Any cont. function with opposite signs has
-        a zero.''}$
-      }{\textcolor{red!80}{\xmark}}{
-        External meaning: Zeros can locally be picked \hil{continuously} in
-        continuous families of continuous functions.
+  \newcommand{\fewstuff}{
+    \includegraphics[height=3em]{filmat}
+    \scalebox{0.5}{\sieve{7}{2}}
   }
-    \end{itemize}
-  \end{changemargin}
 
-  \vspace*{-2em}
-  \hfill\video{zeros-in-families.mp4}{2cm}
-\end{frame}
+  \newcommand{\threeblobs}{
+    \colorbox{mypurple}{\ \ }\quad
+    \colorbox{mypurple}{\ \ }\quad
+    \colorbox{mypurple}{\ \ }
+  }
 
-\begin{frame}{Synthetic differential geometry}
-  \vspace*{-1em}
-  \begin{varblock}{9.5cm}{The axiom of microaffinity}
-    \justifying
-    Let~$\Delta = \{ \varepsilon \in \RR \,|\, \varepsilon^2 = 0 \}$.
-    For any function $f : \Delta \to \RR$,
-    there are unique numbers~$a, b \in \RR$ such that
-    $f(\varepsilon) = a + b \varepsilon$
-    for all~$\varepsilon \in \Delta$.
-  \end{varblock}
+  \begin{tikzpicture}
+    \node (objs-set0) at (0,0) {\muchstuff};
+    \node[scale=0.4] (objs-set1) at (-3.5,-2.5) {\fewstuff};
+    \node[scale=0.4] (objs-eff1) at (3.5,-2.5) {\fewstuff};
+    \node[scale=0.4] (objs-sh1)  at (0,-2.5) {\fewstuff};
 
-  \begin{center}
-    \includegraphics[width=0.5\textwidth]{microaffinity}
-  \end{center}
+    \node (prop-eff1) [below of=objs-eff1, align=left] {
+      Every function \\
+      is computable.
+    };
 
-  \vspace*{-1em}
-  \begin{itemize}
-    \item The \hil{derivative} of~$f$ as above at zero is~$b$.
-    \item Manifolds are \hil{just sets}.
-    \item A \hil{tangent vector} to~$M$ is a map~$\Delta \to M$.
-  \end{itemize}
+    \node (prop-sh1) [below of=objs-sh1, align=left] {
+      The axiom of \\
+      choice fails.
+    };
 
-  \centering
-  \hil{Toposes provide models for this theory.}
-\end{frame}
+    \node (prop-set1) [below of=objs-set1, align=left] {
+      The usual laws \\
+      of logic hold.
+    };
 
-\section{Synthetic algebraic geometry}
+    \node (more-eff1) [below of=prop-eff1] {\threeblobs};
+    \node (more-sh1)  [below of=prop-sh1]  {\threeblobs};
+    \node (more-set1) [below of=prop-set1] {\threeblobs};
 
-\subsection{The big Zariski topos}
-
-\begin{frame}{The big Zariski topos}
-  Let~$S$ be a fixed base scheme.
-
-  \begin{block}{\centering Definition}\small
-  The \hil{big Zariski topos} $\Zar(S)$ is the
-  category $\Sh(\Sch/S)$. It consists of functors
-  $(\Sch/S)^\op \to \Set$ satisfying the gluing condition that
-  \[ F(U) \to
-    \prod_i F(U_i) \rightrightarrows
-    \prod_{j,k} F(U_j \cap U_k) \]
-  \\[-0.6em]
-  is a limit diagram for any scheme~$U = \bigcup_i U_i$ over~$S$.
-  \end{block}
-
-  \begin{changemargin}{-1.0em}{-0.5em}
-  \begin{itemize}
-    \item For an~$S$-scheme~$X$, its functor of points $\ull{X} =
-    \Hom_S(\cdot,X)$ is an object of~$\Zar(S)$. It feels like \hil{the set of
-    points} of~$X$. \\[0.2em]
-    \item \ \\[-1.2em]\mbox{In particular, there is the ring object $\affl$ with~$\affl(U) =
-    \O_U(U)$.} \\[0.2em]
-    \item $\Zar(S)$ classifies local~$\O_S$-algebras which are local
-    over~$\O_S$.
-  \end{itemize}
-  \end{changemargin}
-\end{frame}
-
-\subsection{Synthetic constructions}
-
-\begin{frame}{Synthetic constructions}
-  \small
-  $\hil{$\boldsymbol{\mathbb{A}^n}$} = (\affl)^n = \affl \times \cdots \times \affl$
-
-  $\begin{array}{@{}r@{}c@{}l@{}}
-    \hil{$\boldsymbol{\mathbb{P}^n}$} &\phantom{}=\phantom{}& \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee
-    \cdots \vee x_n \neq 0 \}/(\affl)^\times \\
-    &\cong& \text{set of one-dimensional subspaces of~$(\affl)^{n+1}$} \\
-    &&\qquad \text{(with~$\O(-1) = (\ell)_{\ell \? \mathbb{P}^n}$, $\O(1) =
-    (\ell^\vee)_{\ell \? \mathbb{P}^n}$)}
-  \end{array}$
-
-  $\hil{$\boldsymbol{\Spec(R)}$} = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
-    \text{set of $\affl$-valued points of $R$}$
-
-  $\hil{$\boldsymbol{TX}$} = X^\Delta$, where $\Delta = \{ \varepsilon \? \affl \,|\, \varepsilon^2 = 0 \}$
-
-  A subset $U \subseteq X$ is \hil{qc-open} if and only if for any $x : X$
-  there exist $f_1,\ldots,f_n \? \affl$ such that $x \in U \Longleftrightarrow
-  \exists i\_ f_i \neq 0$.
-
-  A \hil{synthetic affine scheme} is a set which is in bijection
-  with~$\Spec(R)$ for some finitely presented algebra~$R$.
-
-  A \hil{synthetic scheme} is a set which can be covered by finitely many
-  qc-open synthetic affine schemes~$U_i$ such that the intersections~$U_i \cap
-  U_j$ can be covered by finitely many qc-open synthetic affine schemes.
-\end{frame}
-
-\subsection[Quasicoherence]{Synthetic quasicoherence}
-
-\begin{frame}{Properties of the affine line}
-  \begin{itemize}
-    \item $\affl$ is a local ring:
-    \[ 1 \neq 0 \qquad\qquad \text{$x + y$ \inv} \Longrightarrow \text{$x$ \inv} \vee
-    \text{$y$ \inv} \]
-    \item $\affl$ is a field:
-    \begin{align*}
-      \neg(x = 0) &\Longleftrightarrow \text{$x$ invertible} \\
-      \neg(\text{$x$ invertible}) &\Longleftrightarrow \text{$x$ nilpotent}
-    \end{align*}
-    \item $\affl$ satisfies the axiom of microaffinity: Any map $f : \Delta \to
-    \affl$ is of the form~$f(\varepsilon) = a + b \varepsilon$ for unique
-    values~$a,b \? \affl$, where~$\Delta = \{ \varepsilon \? \affl \,|\,
-    \varepsilon^2 = 0 \}$.
-    \item Any function $\affl \to \affl$ is a polynomial.
-    \item $\affl$ is anonymously algebraically closed: Any monic polynomial
-    does \emph{not not} have a zero.
-    \item $\affl$ is of unbounded Krull dimension.
-  \end{itemize}
-\end{frame}
-
-\begin{frame}{Synthetic quasicoherence}
-  Recall~$\Spec(R) = \Hom_{\mathrm{Alg}(\affl)}(R, \affl)$ and consider the statement
-  \[ \text{``the canonical map~$
-    \begin{array}[t]{rcl}
-      R &\longrightarrow& (\affl)^{\Spec(R)} \\
-      f &\longmapsto& (\alpha \mapsto \alpha(f))
-    \end{array}
-  $ is bijective''}.
-  \]
-  \vspace*{-1em}
-
-  \begin{itemize}
-    \item True for~$R = \affl[X]/(X^2)$ (microaffinity).
-    \item True for~$R = \affl[X]$ (every function is a polynomial).
-    \item True for \hil{any} finitely presented~$\affl$-algebra~$R$.
-  \end{itemize}
-
-  Any known property of~$\affl$ follows from this \hil{synthetic
-  quasicoherence}.
-
-  \emph{Example.} Let~$x \? \affl$ such that~$x \neq 0$. Set~$R = \affl/(x)$.
-  Then~$\Spec(R) = \emptyset$. Thus~$(\affl)^{\Spec(R)}$ is a singleton.
-  Hence~$R = 0$. Therefore~$x$ is invertible.
-\end{frame}
-
-\section{Nongeometric sequents}
-
-\begin{frame}{Nongeometric sequents}
-  \small
-  Let~$\TT$ be a \hil{geometric theory} (rings, intervals, \ldots).
-
-  \vspace*{-2em}
-  \begin{varblock}{11cm}{}  %Fundamental theorem about classifying toposes}
-    For a \hil{geometric sequent}~$\forall \vec x\_ (\varphi \Rightarrow \psi)$, the following
-    are equivalent:
-    \begin{enumerate}
-      \item It is \hil{provable} by~$\TT$. \\[-1em]
-      \item It holds
-      \hil{for all models} of~$\TT$ in all toposes. \\[-1em]
-      \item It holds
-      for the \hil{generic model} of~$\TT$ in its \hil{classifying topos}.
-    \end{enumerate}
-  \end{varblock}
-
-  \begin{itemize}
-    \item Additional \hil{nongeometric sequents} may hold in a classifying
-    topos, for instance~``$\affl$ is synthetically quasicoherent''
-    in~$\Zar(S)$. \\[-1em]
-    \item These are \hil{$\boldsymbol{\TT}$-redundant}, but the converse is
-    false. \\[-1em]  % (Bezem--Buchholtz--Coquand 2017). \\[-1em]
-    \item \ \\[-1.2em]\mbox{Are they precisely the consequences of synthetic
-    quasicoherence?} \\[-1em]
-    \item Applications: synthetic algebraic geometry,
-    generic freeness, \ldots
-  \end{itemize}
+    \drawbox{set1}{$\mathrm{Set}$}{(objs-set1) (prop-set1)}
+    \drawbox{eff1}{Ef{}f}{(objs-eff1) (prop-eff1)}
+    \drawbox{sh1}{$\operatorname{Sh} X$}{(objs-sh1) (prop-sh1)}
+    \drawbox{set0}{$\mathrm{Set}$}{(objs-set0) (set1) (eff1) (sh1)}
+  \end{tikzpicture}
 \end{frame}
@@
 \end{document}
-
-"Any map from the reals to the reals is smooth." This statement holds in the
-context of synthetic differential geometry, a well-developed account of smooth
-manifolds and related notions which allows us to bring formal reasoning much
-closer to geometric intuition.
-
-We employ the internal language of the big Zariski topos of a base scheme to
-give a similar account of algebraic geometry, reminiscent of the language of
-the classical Italian school and incorporating Grothendieck's functor-of-points
-philosophy. From the point of view of the Zariski topos, a scheme over the base
-will look like a plain old set, and the affine line will look like a certain
-field. Central to the synthetic account is the notion of "synthetic
-quasicoherence", which doesn't have an analog in synthetic differential
-geometry and which gives the account a distinct algebraic flavor.
-
-The higher-order axiom of synthetic quasicoherence implies all known internal
-properties of the affine line, for instance that it is a field and that it is
-algebraically closed in a weak sense. We surmise that this is for a deeper
-reason, related to the age-old question "which nongeometric sequents hold in
-the classifying topos of a geometric theory?". The second part of the talk
-reports on work in progress about this topic.

### 2018-05-23 paper-wlog.tex
Injectivity statement
Injective matrices

@@
 
-\title{Without loss of generality, \\ any reduced ring is a field}
+\title[Without loss of generality, any reduced ring is a field]{Without loss of generality, \\ any reduced ring is a field}
 \author{Ingo Blechschmidt}
@@
 \newcommand{\J}{\mathcal{J}}
+\newcommand{\ZZ}{\mathbb{Z}}
 \newcommand{\defeq}{\vcentcolon=}
@@
 
+\begin{prop}\label{injection}
+Let~$M \in A^{n \times m}$ be a matrix over a ring~$A$ such that~$n
+< m$. Assume that the induced linear map~$A^m \to A^n$ is injective. Then~$1 =
+0 \in A$.
+\end{prop}
+
+\begin{proof}[Proof~A (classical)]
+Assume to the contrary that~$1 \neq 0 \in A$. Then there exists a minimal prime
+ideal~$\ppp$. Since localization is exact, the induced linear map~$A_\ppp^m \to
+A_\ppp^n$ is still injective. Since the stalk~$A_\ppp$ has precisely one prime
+ideal, any element of~$A_\ppp$ is either invertible or nilpotent.
+
+If~$A$ happens to be reduced, a shortcut is available: In this case~$A_\ppp$ is
+also reduced, therefore a field. Thus injectivity of the linear map~$A_\ppp^m
+\to A_\ppp^n$ is a contradiction to basic linear algebra.
+
+In the general case, we argue as follows. Any entry of the matrix is invertible
+or nilpotent as an element of~$A_\ppp$. If the former case occurs for at least
+one entry, we can employ elementary row and column transformations to change it
+into the unit, move it to the top left, kill the remaining entries in the first
+row and in the first column, and continue in this fashion with the
+resulting~$((n-1) \times (m-1))$-submatrix.
+
+We eventually obtain a matrix of the
+form~$\left(\begin{smallmatrix}I&0\\0&M'\end{smallmatrix}\right)$, where~$I$ is
+an identity matrix of appropriate size, the submatrix~$M'$ is still injective,
+and all entries of~$M'$ are nilpotent. Let~$\aaa$ be the ideal of~$A_\ppp$
+generated by the entries of~$M'$. Since all generators of~$\aaa$ are nilpotent,
+there is a natural number~$n \geq 0$ such that~$\aaa^n = (0)$.
+
+If~$n \geq 1$, then in fact~$\aaa^{n-1} = (0)$, since if~$x \in \aaa^{n-1}$,
+then~$M' \cdot (x,0,\ldots,0)^t = 0$ (the coefficients of the resulting vector are
+elements of~$\aaa^n$) and thus~$x = 0$ by injectivity. Repeating this argument
+yields~$(1) = \aaa^0 = (0)$, thus~$1 = 0 \in A_\ppp$; a contradiction.
+\end{proof}
+
+\begin{proof}[Proof B (constructive)]By XXX, the given matrix~$M$ is also
+injective as a matrix over~$A^\sim$. We may therefore pass from~$A$ to the
+forcing model~$A^\sim$, allowing us to assume that that~$A$ is almost a field
+in the sense that any element is \notnot invertible or nilpotent.
+
+Since the statement we want to verify,~$1 = 0 \? A^\sim$, is~$\neg\neg$-stable,
+we may assume that any entry of the matrix is invertible or nilpotent. We can
+therefore proceed exactly as in Proof~A, only referring to~$A^\sim$ where that
+proof refers to the stalk~$A_\ppp$.
+\end{proof}
+
+\begin{prop}Let~$M \in A^{n \times m}$ be a matrix over a ring~$A$. Assume that
+the induced linear map~$A^m \to A^n$ is injective. Then~$(\Lambda^m A)$, the
+ideal of~$m$-minors of~$A$, is regular.
+\end{prop}
+
 \begin{proof}Let~$a \? A^\sim$ such that~$a \cdot (\Lambda^m M) = (0)$. We want

### 2018-05-23 paper-wlog.tex
Nilpotent polynomials

@@
 
-XXX
+\begin{prop}Let~$f \in A[X]$ be a nilpotent polynomial over a ring~$A$. Then
+the coefficients of~$f$ are nilpotent as elements of~$A$.
+\end{prop}
+
+\begin{proof}[Proof A (classical)]It suffices to show that the coefficients
+of~$f$ are contained in every prime ideal of~$A$. Thus let~$\ppp$ be an
+arbitrary prime ideal. Since some power of~$f$ is zero as an element of~$A[X]$,
+a~forteriori it is zero as an element of~$(A/\ppp)[X]$. Since~$A/\ppp$ is an
+integral domain, so is~$(A/\ppp)[X]$ (this requires a quick verification using
+induction). Thus~$f$ is zero as an element of~$(A/\ppp)[X]$, so all of its
+coefficients are elements of~$\ppp$.
+\end{proof}
+
+Proof~A is ineffective: It doesn't give any bounds on the indices of nilpotency
+of the coefficients in terms of the index of nilpotency of~$f$.\footnote{The
+proof does tell us that there is \emph{some} uniform bound, since we can apply
+it to the generic situation, that is the polynomial~$f = \sum_i A_i X^i$ over the
+ring~$\ZZ[A_0,\ldots,A_d]/(\ldots)$, where we put the individual coefficients of~$(\sum_i
+A_i X^i)^n$ at~``$\ldots$''.} The standard way to mine Proof~A for effective
+content is to employ a technique suitable for capturing the \emph{generic prime
+ideal}, for instance \cite{XXXcoquand} or \cite{XXXschuster}. These techniques
+come down to the forcing language of the space~$\Spec(A)$, but equipped with
+the \emph{flat topology} (also called \emph{co-Zariski topology}), and allow
+for a constructive interpretation of a verbatim copy of Proof~A.
+
+\begin{proof}[Proof B (constructive)]
+\end{proof}
 

### 2018-05-24 slides-filmat2018.tex
A glimpse of the toposophic landscape

@@
 \usepackage{tikz}
-\usetikzlibrary{calc,shapes.callouts,shapes.arrows,patterns,fit,backgrounds}
+\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing}
 \hypersetup{colorlinks=true}
@@
 \hypersetup{colorlinks=false}
+
+% http://www.texample.net/tikz/examples/hand-drawn-lines/
+\pgfdeclaredecoration{penciline}{initial}{
+    \state{initial}[width=+\pgfdecoratedinputsegmentremainingdistance,
+    auto corner on length=1mm,]{
+        \pgfpathcurveto%
+        {% From
+            \pgfqpoint{\pgfdecoratedinputsegmentremainingdistance}
+                      {\pgfdecorationsegmentamplitude}
+        }
+        {%  Control 1
+        \pgfmathrand
+        \pgfpointadd{\pgfqpoint{\pgfdecoratedinputsegmentremainingdistance}{0pt}}
+                    {\pgfqpoint{-\pgfdecorationsegmentaspect
+                     \pgfdecoratedinputsegmentremainingdistance}%
+                               {\pgfmathresult\pgfdecorationsegmentamplitude}
+                    }
+        }
+        {%TO
+        \pgfpointadd{\pgfpointdecoratedinputsegmentlast}{\pgfpoint{1pt}{1pt}}
+        }
+    }
+    \state{final}{}
+}
 
@@
 
-  \hil{Exploring mathematical objects from custom-tailored mathematical universes} \\
+  \hil{Exploring mathematical objects from \\ custom-tailored mathematical universes} \\
+
+  \bigskip
+
+  \scriptsize
+  Ingo Blechschmidt \\
+  Max Planck Institute for Mathematics in the Sciences, Leipzig \\
 
@@
 
-  \scriptsize
-  Max Planck Institute for Mathematics in the Sciences \\
-  Ingo Blechschmidt \\
-
+  Third international conference of the \\
+  \emph{Italian Network for the Philosophy of Mathematics} \\
   \medskip
-
-  Third international conference of the Italian Network for the Philosophy of
-  Mathematics \\
   May 26th, 2018
@@
 \begin{frame}[fragile]{A glimpse of the toposophic landscape}
-  \tikzstyle{topos} = [draw=mypurple, very thick,
-      rectangle, rounded corners, inner sep=10pt, inner ysep=20pt]
+  \tikzstyle{topos} = [draw=mypurple, very thick, rectangle, rounded corners, inner sep=5pt, inner ysep=10pt]
   \tikzstyle{title} = [fill=mypurple, text=white]
+  \tikzset{
+    invisible/.style={opacity=0,text opacity=0},
+    visible on/.style={alt={#1{}{invisible}}},
+    alt/.code args={<#1>#2#3}{%
+      \alt<#1>{\pgfkeysalso{#2}}{\pgfkeysalso{#3}}}
+  }
 
@@
 
-  \newcommand{\drawbox}[3]{
-    \node[topos] [fit = #3] (#1) {};
+  \newcommand{\drawbox}[4]{
+    \node[topos, #4] [fit = #3] (#1) {};
     \node[title] at (#1.north) {#2};
@@
     \scalebox{0.5}{\sieve{14}{2}}
+  }
+
+  \newcommand{\muchstuffplaceholder}{
+    \includegraphics[height=3em]{filmat-placeholder}
+    \scalebox{0.5}{\fakesieve{14}{2}}
   }
@@
   \begin{tikzpicture}
-    \node (objs-set0) at (0,0) {\muchstuff};
-    \node[scale=0.4] (objs-set1) at (-3.5,-2.5) {\fewstuff};
-    \node[scale=0.4] (objs-eff1) at (3.5,-2.5) {\fewstuff};
-    \node[scale=0.4] (objs-sh1)  at (0,-2.5) {\fewstuff};
+    \node (objs-set0) at (0,0) {
+      \only<1-2>{\muchstuffplaceholder}
+      \only<3>{\muchstuff}
+    };
+    \node[scale=0.4] (objs-set1) at (-3.5,-2.5) {
+      \only<1->{\fewstuff}
+    };
+    \node[scale=0.4] (objs-eff1) at (3.5,-2.5) {
+      \only<2->{\fewstuff}
+    };
+    \node[scale=0.4] (objs-sh1)  at (0,-2.5) {
+      \only<2->{\fewstuff}
+    };
+
+    \node (prop-set1) [below of=objs-set1, align=left] {
+      \only<1->{%
+        The usual laws \\
+        of logic hold.
+      }
+    };
 
     \node (prop-eff1) [below of=objs-eff1, align=left] {
+      \only<2->{%
         Every function \\
         is computable.
+      }
     };
@@
     \node (prop-sh1) [below of=objs-sh1, align=left] {
+      \only<2->{%
         The axiom of \\
         choice fails.
+      }
     };
 
-    \node (prop-set1) [below of=objs-set1, align=left] {
-      The usual laws \\
-      of logic hold.
+    \node (more-eff1) [below of=prop-eff1, visible on=<3->] {
+      \threeblobs
+    };
+    \node (more-sh1)  [below of=prop-sh1, visible on=<3->] {
+      \threeblobs
+    };
+    \node (more-set1) [below of=prop-set1, visible on=<3->] {
+      \threeblobs
     };
 
-    \node (more-eff1) [below of=prop-eff1] {\threeblobs};
-    \node (more-sh1)  [below of=prop-sh1]  {\threeblobs};
-    \node (more-set1) [below of=prop-set1] {\threeblobs};
-
-    \drawbox{set1}{$\mathrm{Set}$}{(objs-set1) (prop-set1)}
-    \drawbox{eff1}{Ef{}f}{(objs-eff1) (prop-eff1)}
-    \drawbox{sh1}{$\operatorname{Sh} X$}{(objs-sh1) (prop-sh1)}
-    \drawbox{set0}{$\mathrm{Set}$}{(objs-set0) (set1) (eff1) (sh1)}
+    \begin{scope}[visible on=<1->]
+      \drawbox{set1}{$\mathrm{Set}$}{(objs-set1) (prop-set1) (more-set1)}{}
+    \end{scope}
+    \begin{scope}[visible on=<2->]
+      \drawbox{eff1}{Ef{}f}{(objs-eff1) (prop-eff1) (more-eff1)}{tape}
+    \end{scope}
+    \begin{scope}[visible on=<2->]
+      \drawbox{sh1}{$\operatorname{Sh} X$}{(objs-sh1) (prop-sh1) (more-sh1)}{draw=none}
+      \def\R{8pt}
+      \begin{pgfonlayer}{background}
+      \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple, visible on=<2->]
+        ($(sh1.south west) + (\R, 0)$) arc(270:180:\R) --
+        ($(sh1.north west) + (0, -\R)$) arc(180:90:\R) --
+        ($(sh1.north east) + (-\R, 0)$) arc(90:0:\R) --
+        ($(sh1.south east) + (0, \R)$) arc(0:-90:\R) --
+        cycle;
+      \end{pgfonlayer}
+    \end{scope}
+    \begin{scope}[visible on=<3->]
+      \drawbox{set0}{$\mathrm{Set}$}{(objs-set0) (set1) (eff1) (sh1)}{}
+    \end{scope}
   \end{tikzpicture}

### 2018-05-24 slides-filmat2018.tex
Slides mostly finished

@@
 
-\newcommand{\fmini}[2]{\framebox{\begin{minipage}{#1}#2\end{minipage}}}
-\makeatletter
-\def\underunbrace#1{\mathop{\vtop{\m@th\ialign{##\crcr
-      $\hfil\displaystyle{#1}\hfil$\crcr\noalign{\kern3\p@\nointerlineskip}
-      \crcr\noalign{\kern3\p@}}}}\limits}
-\def\overunbrace#1{\mathop{\vbox{\m@th\ialign{##\crcr\noalign{\kern3\p@}
-      \crcr\noalign{\kern3\p@\nointerlineskip}
-      $\hfil\displaystyle{#1}\hfil$\crcr}}}\limits}
-\makeatother
+% Adapted from https://latex.org/forum/viewtopic.php?t=2251 (Stefan Kottwitz)
+\newenvironment<>{hilblock}{
+  \begin{center}
+    \begin{minipage}{8.25cm}
+      \setlength{\textwidth}{8.25cm}
+      \begin{actionenv}#1
+	\def\insertblocktitle{}
+	\par
+	\usebeamertemplate{block begin}}{
+        \par
+        \usebeamertemplate{block end}
+      \end{actionenv}
+    \end{minipage}
+  \end{center}}
 
@@
     (#1) {#1};\node[overlay,rectangle callout,%
-    callout relative pointer={(0.3cm,1.0cm)},fill=blue!20] at ($(#1.north)+(-0.1cm,-1.6cm)$) {#2};}%
+    callout relative pointer={(0.3cm,0.5cm)},fill=blue!20] at ($(#1.north)+(-0.1cm,-1.1cm)$) {#2};}%
 }%
-
-\newcommand{\hcancel}[5]{%
-  \tikz[baseline=(tocancel.base)]{
-    \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
-    \draw[red, line width=0.4mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
-  }%
-}
-
-\newcommand{\slogan}[1]{%
-  \begin{center}%
-    \setlength{\fboxrule}{2pt}%
-    \setlength{\fboxsep}{8pt}%
-    {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal text}\parbox{0.93\textwidth}{#1}}}%
-  \end{center}%
-}
-
-\newcommand{\sloganwithoutborder}[1]{%
-  \begin{center}%
-    \setlength{\fboxrule}{0pt}%
-    \setlength{\fboxsep}{-14pt}%
-    {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal
-    text}\parbox{0.9\textwidth}{\begin{center}#1\end{center}}}}%
-  \end{center}%
-}
 
@@
   \hil{Exploring mathematical objects from \\ custom-tailored mathematical universes} \\
+
+  \emph{an invitation}
 
@@
 
+\begin{frame}{The internal universe of a topos}
+  For any topos~$\E$ and any statement~$\varphi$, we define
+  \begin{gather*}
+    \E \models \varphi \\
+    \text{``$\varphi$ holds in the internal universe of~$\E$''}
+  \end{gather*}
+  using a simple recursive definition, the \hil{Kripke--Joyal semantics}.
+  \medskip
+
+  \begin{columns}
+    \begin{column}{0.3\textwidth}
+      \centering
+      $\Set \models \varphi$ \\
+      ``$\varphi$ holds \\ in the usual sense.''
+    \end{column}
+
+    \begin{column}{0.3\textwidth}
+      \centering
+      $\Sh(X) \models \varphi$ \\
+      ``$\varphi$ holds continuously.''
+    \end{column}
+
+    \begin{column}{0.3\textwidth}
+      \centering
+      $\Eff \models \varphi$ \\
+      ``$\varphi$ holds computably.''
+    \end{column}
+  \end{columns}
+  \medskip
+
+  \pause
+  Any topos supports \hil{mathematical reasoning}:
+
+  \vspace*{-1em}
+  \begin{hilblock}
+    If~$\E \models \varphi$ and if~$\varphi \vdash \psi$
+    \only<1-2>{intuitionistically}%
+    \only<3>{\pointthis{intuitionistically}{%
+      no $\varphi \vee \neg\varphi$,\ \
+      no $\neg\neg\varphi \Rightarrow \varphi$,\ \
+      no axiom of choice}},
+    then~$\E \models \psi$.
+  \end{hilblock}
+\end{frame}
+
+\newcommand{\intex}[3]{#1 \quad #2\par #3\bigskip\medskip}
+
+\begin{frame}{First steps in alternate universes}
+  \begin{changemargin}{-1.1em}{0em}
+    \small
+    \begin{itemize}
+      \item \intex{
+        $\Eff \models \text{``Any number is prime or is not prime.''}$
+      }{\textcolor{green!90}{\cmark}}{
+        External meaning: There is a \hil{Turing machine} which determines of
+        any given number whether it is prime or not.
+      }
+
+      \item \intex{
+        $\Eff \models \text{``There are infinitely many prime numbers.''}$
+      }{\textcolor{green!90}{\cmark}}{
+        External meaning: There is a \hil{Turing machine} producing arbitrarily many prime
+        numbers.
+      }
+
+      \item \intex{
+        $\Eff \models \text{``Any function~$\NN \to \NN$ is the zero function or not.''}$
+      }{\textcolor{red!80}{\xmark}}{
+        External meaning: There is a \hil{Turing machine} which, given a Turing
+        machine computing a function~$f : \NN \to \NN$, determines whether~$f$
+        is the zero function or not.
+      }
+
+      \item \intex{
+        $\Eff \models \text{``Any function~$\NN \to \NN$ is computable.''}$
+      }{\textcolor{green!90}{\cmark}}{}
+
+      \item \intex{
+        $\Sh(X) \models \text{``Any cont. function with opposite signs has
+        a zero.''}$
+      }{\textcolor{red!80}{\xmark}}{
+        External meaning: Zeros can locally be picked \hil{continuously} in
+        continuous families of continuous functions.
+        \textcolor{red}{(\href{https://rawgit.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{video} for counterexample)}
+      }
+    \end{itemize}
+  \end{changemargin}
+\end{frame}
+
 \end{document}

### 2018-05-25 slides-filmat2018.tex
Nicer title slide and slide on applications

@@
-\documentclass[13pt,utf8,notheorems,compress,t]{beamer}
+\documentclass[12pt,utf8,notheorems,compress,t]{beamer}
 \usepackage{etex}
@@
 
-% http://www.texample.net/tikz/examples/hand-drawn-lines/
-\pgfdeclaredecoration{penciline}{initial}{
-    \state{initial}[width=+\pgfdecoratedinputsegmentremainingdistance,
-    auto corner on length=1mm,]{
-        \pgfpathcurveto%
-        {% From
-            \pgfqpoint{\pgfdecoratedinputsegmentremainingdistance}
-                      {\pgfdecorationsegmentamplitude}
-        }
-        {%  Control 1
-        \pgfmathrand
-        \pgfpointadd{\pgfqpoint{\pgfdecoratedinputsegmentremainingdistance}{0pt}}
-                    {\pgfqpoint{-\pgfdecorationsegmentaspect
-                     \pgfdecoratedinputsegmentremainingdistance}%
-                               {\pgfmathresult\pgfdecorationsegmentamplitude}
-                    }
-        }
-        {%TO
-        \pgfpointadd{\pgfpointdecoratedinputsegmentlast}{\pgfpoint{1pt}{1pt}}
-        }
-    }
-    \state{final}{}
-}
-
 \usepackage{pifont}
@@
 \useinnertheme[shadow=true]{rounded}
-\useoutertheme{split}
-\usecolortheme{orchid}
-\usecolortheme{whale}
+\useoutertheme[subsection=false]{miniframes}
 \setbeamerfont{block title}{size={}}
+\addtobeamertemplate{headline}{\vskip-0.5ex}{}
 
@@
 
+\usecolortheme{orchid}
 \usecolortheme{seahorse}
@@
 \newcommand{\NN}{\mathbb{N}}
+\newcommand{\QQ}{\mathbb{Q}}
 \newcommand{\RR}{\mathbb{R}}
@@
   \begin{center}
-    \begin{minipage}{8.25cm}
-      \setlength{\textwidth}{8.25cm}
+    \begin{minipage}{9.05cm}
+      \setlength{\textwidth}{9.05cm}
       \begin{actionenv}#1
@@
 
-\newcommand{\pointthis}[2]{%
-  \tikz[remember picture,baseline]{\node[anchor=base,inner sep=0,outer sep=0]%
-    (#1) {#1};\node[overlay,rectangle callout,%
-    callout relative pointer={(0.3cm,0.5cm)},fill=blue!20] at ($(#1.north)+(-0.1cm,-1.1cm)$) {#2};}%
+\tikzset{
+  invisible/.style={opacity=0,text opacity=0},
+  visible on/.style={alt={#1{}{invisible}}},
+  alt/.code args={<#1>#2#3}{%
+    \alt<#1>{\pgfkeysalso{#2}}{\pgfkeysalso{#3}}}
+}
+
+\newcommand{\pointthis}[3]{%
+  \tikz[remember picture,baseline]{
+    \node[anchor=base,inner sep=0,outer sep=0] (#2) {#2};
+    \node[visible on=#1,overlay,rectangle callout,rounded corners,callout relative pointer={(0.3cm,0.5cm)},fill=blue!20] at ($(#2.north)+(-0.1cm,-1.1cm)$) {#3};
   }%
+}
 
@@
   \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex,dp=1ex,right,rightskip=1mm,leftskip=1mm]{}%
+    \inserttitle
     \hfill
@@
 
+{\title{}
 \begin{frame}[c]
@@
   \medskip
-  \includegraphics[scale=0.3]{images/external-internal-small}
+  \includegraphics[height=0.4\textheight]{acquaviva-platani-street}%
+  \includegraphics[height=0.4\textheight]{mussomeli-sheep}%
+  \includegraphics[height=0.4\textheight]{mussomeli-performance}%
+  \includegraphics[height=0.4\textheight]{mussomeli-castle}%
   \medskip
@@
 
-  \emph{an invitation}
+  \emph{-- an invitation --}
 
@@
   Ingo Blechschmidt \\
+  University of Augsburg / \\
   Max Planck Institute for Mathematics in the Sciences, Leipzig \\
@@
   \emph{Italian Network for the Philosophy of Mathematics} \\
+  in Mussomeli
   \medskip
+
   May 26th, 2018
   \par
-\end{frame}
+\end{frame}}
+
+\section[Toposophic landscape]{A glimpse of the toposophic landscape}
 
@@
   \tikzstyle{title} = [fill=mypurple, text=white]
-  \tikzset{
-    invisible/.style={opacity=0,text opacity=0},
-    visible on/.style={alt={#1{}{invisible}}},
-    alt/.code args={<#1>#2#3}{%
-      \alt<#1>{\pgfkeysalso{#2}}{\pgfkeysalso{#3}}}
-  }
 
   \input{images/primes.tex}
+  %\renewcommand{\sieve}[2]{SIEVE}
+  %\renewcommand{\fakesieve}[2]{SIEVE}
 
@@
     \begin{scope}[visible on=<2->]
-      \drawbox{sh1}{$\operatorname{Sh} X$}{(objs-sh1) (prop-sh1) (more-sh1)}{draw=none}
+      \drawbox{sh1}{$\mathrm{Sh}\, X$}{(objs-sh1) (prop-sh1) (more-sh1)}{draw=none}
       \def\R{8pt}
@@
 
+\section[Internal universe]{The internal universe of a topos}
+
 \begin{frame}{The internal universe of a topos}
-  For any topos~$\E$ and any statement~$\varphi$, we define
-  \begin{gather*}
-    \E \models \varphi \\
-    \text{``$\varphi$ holds in the internal universe of~$\E$''}
-  \end{gather*}
-  using a simple recursive definition, the \hil{Kripke--Joyal semantics}.
-  \medskip
+  For any topos~$\E$ and any statement~$\varphi$, we define the meaning of
+  \vspace*{-0.5em}
+  \[
+    \text{``$\E \models \varphi$''} \quad
+    \text{(``$\varphi$ holds in the internal universe of~$\E$'')}
+  \]
 
+  \vspace*{-1.0em}
+  using the \hil{Kripke--Joyal semantics}.
+
+  \vspace*{-1em}
   \begin{columns}
-    \begin{column}{0.3\textwidth}
+    \def\insertblocktitle{}
+    \begin{column}{0.3\textwidth}\usebeamertemplate{block begin}
       \centering
       $\Set \models \varphi$ \\
-      ``$\varphi$ holds \\ in the usual sense.''
-    \end{column}
+      ``$\varphi$ holds in the usual sense.''
+    \usebeamertemplate{block end}\end{column}
 
-    \begin{column}{0.3\textwidth}
+    \begin{column}{0.3\textwidth}\usebeamertemplate{block begin}
       \centering
@@
       ``$\varphi$ holds continuously.''
-    \end{column}
+    \usebeamertemplate{block end}\end{column}
 
-    \begin{column}{0.3\textwidth}
+    \begin{column}{0.3\textwidth}\usebeamertemplate{block begin}
       \centering
@@
       ``$\varphi$ holds computably.''
-    \end{column}
+    \usebeamertemplate{block end}\end{column}
   \end{columns}
@@
     If~$\E \models \varphi$ and if~$\varphi \vdash \psi$
-    \only<1-2>{intuitionistically}%
-    \only<3>{\pointthis{intuitionistically}{%
+    \pointthis{<3>}{intuitionistically}{%
       no $\varphi \vee \neg\varphi$,\ \
       no $\neg\neg\varphi \Rightarrow \varphi$,\ \
-      no axiom of choice}},
+      no axiom of choice},
     then~$\E \models \psi$.
@@
 \end{frame}
+
+\section[First steps]{First steps in alternate universes}
 
@@
   \begin{changemargin}{-1.1em}{0em}
-    \small
+    \fontsize{10pt}{12pt}\selectfont
     \begin{itemize}
@@
       }{\textcolor{green!90}{\cmark}}{
-        External meaning: There is a \hil{Turing machine} which determines of
+        There is a \hil{Turing machine} which determines of
         any given number whether it is prime or not.
@@
       }{\textcolor{green!90}{\cmark}}{
-        External meaning: There is a \hil{Turing machine} producing arbitrarily many prime
+        There is a \hil{Turing machine} producing arbitrarily many prime
         numbers.
@@
       }{\textcolor{red!80}{\xmark}}{
-        External meaning: There is a \hil{Turing machine} which, given a Turing
+        There is a \hil{Turing machine} which, given a Turing
         machine computing a function~$f : \NN \to \NN$, determines whether~$f$
@@
       }{\textcolor{red!80}{\xmark}}{
-        External meaning: Zeros can locally be picked \hil{continuously} in
+        Zeros can locally be picked \hil{continuously} in
         continuous families of continuous functions.
@@
 
+\section[Applications]{Applications in commutative algebra}
+
+\begin{frame}{Applications in commutative algebra}
+  Let~$A$ be a reduced commutative ring.
+
+  \hil{For example:}\ \ $\ZZ$,\ \ $\ZZ[X]$,\ \ $\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$,\ \ $\QQ$,\ \ $\RR$
+
+  \mbox{The \hil{little Zariski topos} of~$A$ contains a \hil{mirror image} of~$A$:
+  $A^\sim$.}
+
+  \setlength\columnsep{0pt}
+  \begin{multicols*}{2}
+    \begin{enumerate}
+      \item $A^\sim$ is always a \hil{field}. \columnbreak
+      \item $A^\sim$ is still \hil{very close} to~$A$.
+    \end{enumerate}
+  \end{multicols*}
+  % \item Studying~$A^\sim$ is the same as studying~$A$ \hil{from a different
+  % point of view}.
+
+  \vspace*{-2em}
+  \begin{columns}[t]
+    \begin{column}[t]{0.47\textwidth}
+      \centering
+
+      \begin{varblock}{\textwidth}{A baby application}
+        \justifying
+        Let~$M$ be a surjective matrix with more rows than columns over a ring~$A$.
+        Then~$A = 0$.
+      \end{varblock}
+
+      \scalebox{0.8}{$\begin{pmatrix}
+        \cdot & \cdot \\
+        \cdot & \cdot \\
+        \cdot & \cdot
+      \end{pmatrix}$}
+    \end{column}
+
+    \begin{column}[t]{0.47\textwidth}
+      \centering
+
+      \begin{varblock}{\textwidth}{Generic freeness\phantom{p}}
+        \justifying
+        Generically, any finitely generated module over a reduced ring is free.
+      \end{varblock}
+      \vspace*{-0.5em}
+
+      \includegraphics[width=0.6\textwidth]{generic-freeness}
+    \end{column}
+  \end{columns}
+\end{frame}
+
+\section{Conclusion}
+
+\begin{frame}{Topos theory \ldots}
+  \begin{itemize}
+    \item enriches the platonism debate,
+    %\item yields a new view on ``continuous dependence on parameters'',
+    \item uncovers further relations between objects,
+    \item allows to view objects from a different point of view,
+    \item has applications in mathematical practice.
+  \end{itemize}
+
+  \centering
+  \href{https://www.oliviacaramello.com/Papers/Papers.htm}{\includegraphics[height=0.45\textheight]{olivia-tst}}
+  \href{http://math.andrej.com/2014/03/04/intuitionistic-mathematics-and-realizability-in-the-physical-world/}{\includegraphics[height=0.45\textheight]{zenil-computable-universe}}
+  \href{https://pizzaseminar.speicherleck.de/skript2/zariski-topos-klein.pdf}{\includegraphics[height=0.45\textheight]{fun-with-the-zariski-topos}}
+  \href{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{\includegraphics[height=0.45\textheight]{phd-cover}}
+\end{frame}
+
 \end{document}

### 2018-05-25 slides-filmat2018.tex
Bonus slide on the little Zariski topos

@@
       }{\textcolor{green!90}{\cmark}}{
-        There is a \hil{Turing machine} which determines of
+        Meaning: There is a \hil{Turing machine} which determines of
         any given number whether it is prime or not.
@@
       }{\textcolor{green!90}{\cmark}}{
-        There is a \hil{Turing machine} producing arbitrarily many prime
-        numbers.
+        Meaning: There is a \hil{Turing machine} producing arbitrarily many
+        primes.\\[0em]
       }
@@
       }{\textcolor{red!80}{\xmark}}{
-        There is a \hil{Turing machine} which, given a Turing
+        Meaning: There is a \hil{Turing machine} which, given a Turing
         machine computing a function~$f : \NN \to \NN$, determines whether~$f$
-        is the zero function or not.
+        is zero or not.
       }
@@
       }{\textcolor{red!80}{\xmark}}{
-        Zeros can locally be picked \hil{continuously} in
+        Meaning: Zeros can locally be picked \hil{continuously} in
         continuous families of continuous functions.
@@
 
-  \hil{For example:}\ \ $\ZZ$,\ \ $\ZZ[X]$,\ \ $\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$,\ \ $\QQ$,\ \ $\RR$
+  \hil{For instance:}\ \ $\ZZ$,\ \ $\ZZ[X]$,\ \ $\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$,\ \ $\QQ$,\ \ $\RR$
 
@@
 
+{\renewcommand{\insertframenumber}{4a}
+\begin{frame}{The little Zariski topos in more detail}
+  Recall~$A[f^{-1}] = \bigl\{ \frac{u}{f^n} \,|\, u \in A, n \in \NN \bigr\}$.
+
+  \begin{itemize}
+    \item $\Sh(\Spec(A)) \models \text{``For all~$x \in A^\sim$, \ldots''}$
+
+          Meaning: For all~$f \in A$ and all~$x \in A[f^{-1}]$, \ldots
+          \medskip
+
+    \item $\Sh(\Spec(A)) \models \text{``There is~$x \in A^\sim$ such that \ldots''}$
+
+          \mbox{Meaning: There is a partition of unity,~$1 = f_1 + \cdots + f_n \in A$,}
+          such that for each~$i$, there exists~$x_i \in A[f_i^{-1}]$
+          with~\ldots
+          \medskip
+
+    \item $\Sh(\Spec(A)) \models \text{``$\varphi$ implies $\psi$''}$
+
+          Meaning: For all~$f \in A$, if~$\varphi$ on stage~$f$, then~$\psi$ on
+          stage~$f$.
+  \end{itemize}
+\end{frame}
+\addtocounter{framenumber}{-1}}
+
 \section{Conclusion}
@@
 \begin{frame}{Topos theory \ldots}
+  \vspace*{-0.5em}
   \begin{itemize}

### 2018-05-30 abstract-como2018.tex
Abstract for the topos theory conference in Como!

\documentclass{article}
\begin{document}

\subsection*{How topos theory can help commutative algebra}

\noindent
{\scshape Ingo Blechschmidt}\index{Blechschmidt, Ingo}\\
{\scshape University of Augsburg / Max Planck Institute for
Mathematics in the Sciences, Leipzig}
\medskip

Topos theory and commutative algebra are closely linked: In view of the mantra
``toposes are rings'', commutative algebra has been informing topos theory, and
by employing the internal language of toposes, topos theory has been used to
transfer results from commutative algebra to subjects such as algebraic
geometry and differential geometry. In this talk, we explore a third link: How
topos theory can help commutative algebra and neighboring disciplines.

One such relation is given by new reduction techniques. For instance, there is
a way how we can assume, without loss of generality, that any reduced ring is a
field. This technique allows to give a short and simple proof of Grothendieck's
generic freeness lemma, a basic theorem used in the setup of the theory of
moduli spaces, which substantially improves on the previous somewhat
convoluted proofs.

These topos-theoretic reduction techniques cannot generally be mimicked by traditional
commutative algebra, and in the special cases where they can,
they improve on the traditional methods by yielding fully constructive proofs.
The precise sense in which all this is true will be carefully explained in the
talk.

A further relation is given by synthetic approaches to algebraic geometry,
allowing to treat schemes with all their complex algebro-geometrical structure
as plain sets and morphisms between schemes as maps between these sets.
Fundamental to this account is the notion of ``synthetic quasicoherence'',
which doesn't have a counterpart in synthetic differential geometry and which
endows the relevant internal universes with a distinctive algebraic flavor.

Somewhat surprisingly, the work on synthetic algebraic geometry is related to
an age-old question in the study of classifying toposes. We hope to report on
recent results by Matthias Hutzler in this regard, and close with an invitation
to the many open problems of the field. The talk will begin with an
introduction to the internal language of toposes, so as to be accessible to
audience members who are not familiar with it and to provide value
outside of commutative algebra.

\end{document}

### 2018-05-30 slides-filmat2018.tex
typo
Slides in the version they went live

@@
         continuous families of continuous functions.
-        \textcolor{red}{(\href{https://rawgit.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{video} for counterexample)}
+        \textcolor{red!80}{(\href{https://rawgit.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{video} for counterexample)}
       }
@@
 
+\appendix
 \section{Conclusion}
@@
     \item uncovers further relations between objects,
-    \item allows to view objects from a different point of view,
+    \item allows to study objects from a different point of view,
     \item has applications in mathematical practice.
@@
 
+\addtocounter{framenumber}{-1}
+
 \end{document}

### 2018-06-02 paper-generic-freeness.tex
Version to be submitted
typo
Incorporated the easy suggestions by Rosolini

@@
 \author{Ingo Blechschmidt}
+\address{Max Planck Institute for Mathematics in the Sciences \\
+Inselstraße 22 \\
+04103 Leipzig, Germany}
+\email{ingo.blechschmidt@mis.mpg.de}
 
@@
   We present a new and direct proof of Grothendieck's generic freeness
-  lemma in its general form. Unlike the previously published proofs, it doesn't
+  lemma in its general form. Unlike the previously published proofs, it does not
   proceed in a series of reduction steps and is fully constructive, not
@@
 a direct proof is possible and shorter. The new proof unveils a certain
-combinatorial aspect to Grothendieck's generic freeness lemma, doesn't require
-any advanced prerequisites in commutative algebra and doesn't use the axiom of
+combinatorial aspect to Grothendieck's generic freeness lemma, does not require
+any advanced prerequisites in commutative algebra and does not use the axiom of
 choice or the law of excluded middle. It is purely element-based, not referring
@@
 \begin{enumerate}
-\item[(a)] $B^\sim$ and~$M^\sim$ are free as sheaves of~$A^\sim$-modules,
+\item[(a)] $B^\sim$ and~$M^\sim$ are locally free as sheaves of~$A^\sim$-modules,
 \item[(b)] $B^\sim$ is of finite presentation as a sheaf of~$A^\sim$-algebras and
@@
 Theorem~\ref{thm:algebraic} by defining~$U$ as the union of all those basic
-opens~$D(f)$ such that~(1),~(2) and~(3) hold. It's clear that~(a),~(b)
+opens~$D(f)$ such that~(1),~(2) and~(3) hold. It is clear that~(a),~(b)
 and~(c) hold over~$U$, and~$U$ is dense for if~$V$ is an arbitrary open
@@
 
-\textbf{Acknowledgments.} The proof presented here was prompted by
+\textbf{Acknowledgments.} The proof presented here was prompted by a question by
 user~HeinrichD on MathOverflow~\cite{mo:kernel} and greatly
 benefited from discussions with Martin Brandenburg, who employed the constructive
-version in a paper of his~\cite{brandenburg:schur}. I'm grateful to Thierry
+version in a paper of his~\cite{brandenburg:schur}. We are grateful to Thierry
 Coquand and Peter Schuster for valuable advice and to Marc
@@
 verify~$1 = 0$, but as an intermediate step we verify that~$M$ is free, which
-after the fact will be a trivial statement. This approach is reminiscient of
+after the fact will be a trivial statement. The general proof in the next
+section will have a similar style. This approach is reminiscient of
 Richman's uses of trivial rings~\cite{richman:trivial-rings}.
@@
 \cdots x_n^{i_n} v_\ell)_{(\ell,i_1,\ldots,i_n) \in \J}$ generates~$M$ as
-an~$A$-module, and we'll call a subfamily~$(w_J)_{J \in \J' \subseteq \J}$
+an~$A$-module, and we will call a subfamily~$(w_J)_{J \in \J' \subseteq \J}$
 \emph{good} if and only if for all~$J \in \J$, the vector~$w_J$ is a linear
@@
 shapes of a given good generating family~$(w_J)_{J \in \J'}$ for~$M$ and a given
-good generating family~$(s_I)_{I \in \I'}$ for~$B$, starting with the canonical ones. It's
+good generating family~$(s_I)_{I \in \I'}$ for~$B$, starting with the canonical ones. It is
 reasonably obvious that this induction is well-founded; the formal statement
-that it is so is known as \emph{Dickson's Lemma} in mathematical logic.
+that it is so is known as \emph{Dickson's Lemma} (see, for
+instance,~\cite[Thm.~1.1]{veldman:kruskal}).
 
@@
 \end{proof}
+
+\section{Conclusion}
+
+% History
+
+% Application
 

### 2018-06-02 paper-generic-freeness.tex
Incorporated the nontrivial suggestion by Rosolini

@@
-\documentclass{amsart}
+\documentclass[oneside]{amsart}
 
@@
 \usepackage{hyperref}
+
+\usepackage{geometry}
 
@@
 \newcommand{\bbb}{\mathfrak{b}}
+\newcommand{\mmm}{\mathfrak{m}}
 \newcommand{\I}{\mathcal{I}}
 \newcommand{\J}{\mathcal{J}}
+\newcommand{\E}{\mathcal{E}}
+\newcommand{\B}{\mathcal{B}}
+\renewcommand{\O}{\mathcal{O}}
 \newcommand{\defeq}{\vcentcolon=}
@@
 
-\begin{thm}\label{thm:algebraic}Let~$A$ be a reduced ring. Let~$B$ be
+\begin{thm}\label{thm:algebraic}Let~$A$ be a reduced ring (commutative, with unit). Let~$B$ be
 an~$A$-algebra of finite type. Let~$M$ be a finitely generated~$B$-module.
@@
 choice or the law of excluded middle. It is purely element-based, not referring
-to ideals of~$A$.
-\enlargethispage{3em}
+to ideals of~$A$, and doesn't use Noether normalization.
 
@@
 Richman's uses of trivial rings~\cite{richman:trivial-rings}.
-\enlargethispage{1em}
+\enlargethispage{8em}
 
-\section{The proof of the general case}
-
-\begin{proof}[Proof of Theorem~\ref{thm:algebraic}]
-Let~$B$ be generated by~$(x_1,\ldots,x_n)$ as an~$A$-algebra and
-let~$M$ be generated by~$(v_1,\ldots,v_m)$ as a~$B$-module. We endow the sets
-\begin{align*}
-  \I &\defeq \{ (i_1,\ldots,i_n) \,|\, i_1,\ldots,i_n \geq 0 \}
-  \quad\text{and} \\
-  \J &\defeq \{ (\ell, i_1,\ldots,i_n) \,|\, 1 \leq \ell \leq m, i_1,\ldots,i_n \geq 0 \}
-\end{align*}
-with the lexicographic order. The family~$(w_J)_{J \in \J} \defeq (x_1^{i_1}
-\cdots x_n^{i_n} v_\ell)_{(\ell,i_1,\ldots,i_n) \in \J}$ generates~$M$ as
-an~$A$-module, and we will call a subfamily~$(w_J)_{J \in \J' \subseteq \J}$
-\emph{good} if and only if for all~$J \in \J$, the vector~$w_J$ is a linear
-combination of the vectors~$(w_{J'})_{J' \in \J', J' \preceq J}$,
-and if~$(\ell,i_1,\ldots,i_n) \not\in \J'$ implies~$(\ell,k_1,\ldots,k_n) \not\in
-\J'$ for all~$k_1 \geq i_1, \ldots, k_n \geq i_n$. Figure~\ref{fig:good} shows
-how a good generating family can look like.
-Similarly, we define when a subfamily of the canonical generating
-family~$(x_1^{i_1} \cdots x_n^{i_n})_{(i_1,\ldots,i_n) \in \I}$ of~$B$ is good
-(which is just the special case~$m = 1$).
-
-\begin{figure}[t]
+\begin{figure}[H]
   \captionsetup{width=.9\linewidth}
@@
 
+\section{The proof of the general case}
+
+\begin{proof}[Proof of Theorem~\ref{thm:algebraic}]
+Let~$B$ be generated by~$(x_1,\ldots,x_n)$ as an~$A$-algebra and
+let~$M$ be generated by~$(v_1,\ldots,v_m)$ as a~$B$-module. We endow the sets
+\begin{align*}
+  \I &\defeq \{ (i_1,\ldots,i_n) \,|\, i_1,\ldots,i_n \geq 0 \}
+  \quad\text{and} \\
+  \J &\defeq \{ (\ell, i_1,\ldots,i_n) \,|\, 1 \leq \ell \leq m, i_1,\ldots,i_n \geq 0 \}
+\end{align*}
+with the lexicographic order. The family~$(w_J)_{J \in \J} \defeq (x_1^{i_1}
+\cdots x_n^{i_n} v_\ell)_{(\ell,i_1,\ldots,i_n) \in \J}$ generates~$M$ as
+an~$A$-module, and we will call a subfamily~$(w_J)_{J \in \J' \subseteq \J}$
+\emph{good} if and only if for all~$J \in \J$, the vector~$w_J$ is a linear
+combination of the vectors~$(w_{J'})_{J' \in \J', J' \preceq J}$,
+and if~$(\ell,i_1,\ldots,i_n) \not\in \J'$ implies~$(\ell,k_1,\ldots,k_n) \not\in
+\J'$ for all~$k_1 \geq i_1, \ldots, k_n \geq i_n$. Figure~\ref{fig:good} shows
+how a good generating family can look like.
+Similarly, we define when a subfamily of the canonical generating
+family~$(x_1^{i_1} \cdots x_n^{i_n})_{(i_1,\ldots,i_n) \in \I}$ of~$B$ is good
+(which is just the special case~$m = 1$).
+
 We then proceed by induction on the
@@
 Thus $a_J = 0$ since~$A$ is reduced.
-\enlargethispage{2em}
 
@@
 
-% History
+Commutative algebra abounds with techniques which allow us to reduce quite
+general situations to easier ones. These techniques often yield short and slick
+proofs; however, they come at an expense: They are typically nonconstructive in
+nature, employing for instance the axiom of choice, and do not argue using only
+the data at hand, but using additional auxiliary objects such as maximal
+ideals. We feel that once a subject is better understood, it is desirable to
+have more informative, direct proofs available which illuminate the
+statement more clearly; similar as to how bijective proofs are preferred over
+calculational inductive ones in combinatorics.
 
-% Application
+Let us consider as a specific example the statement that the existence of a linear
+surjection~$A^n \to A^m$ with~$n < m$ between finite free modules over an
+arbitrary ring~$A$ implies~$1 = 0 \in A$. The standard proof of this fact proceeds
+by contradiction and passes to the quotient~$A/\mmm$, where~$\mmm$ is a maximal
+ideal of~$A$, thereby reducing to the situation that the ring is a field. In
+contrast, a direct proof such as Richman's~\cite{richman:trivial-rings}
+refers only to objects mentioned in the statement itself and
+explicitly tells us how to deduce the equation~$1 = 0$ from the~$m$ equations
+which express that each basis vector of~$A^m$ has a preimage.
+
+In a similar fashion, the new proof of Grothendieck's generic freeness lemma
+explicitly tells us how to deduce~$1 = 0$ from the given conditional
+equations expressing that~$f = 0$ is the only element with properties~(1),~(2)
+and~(3). The history of Grothendieck's generic freeness lemma goes back more
+than fifty years; we are slightly surprised that a direct proof was
+discovered only now.
+
+Direct proofs sometimes generalize to new situations where the reduction techniques
+employed in more abstract proofs cannot be applied. This is the case for
+Grothendieck's generic freeness lemma, which allows for the following
+generalization:
+
+\begin{thm}\label{thm:general-generic-freeness}
+Let~$(X,\O_X)$ be a ringed space (or ringed locale, or ringed topos), such
+that for every local section~$s$ of~$\O_X$, if the only open on which~$s$ is
+invertible is the empty one, then~$s = 0$. Let~$\B$ be a sheaf
+of~$\O_X$-algebras of finite type. Let~$\E$ be a sheaf of~$\B$-modules of finite type. Then there is a dense
+open~$U$ such that over~$U$,
+\begin{enumerate}
+\item[(a)] $\B$ and~$\E$ are locally free as sheaves of~$\O_X$-modules,
+\item[(b)] $\B$ is of finite presentation as a sheaf of~$\O_X$-algebras and
+\item[(c)] $\E$ is finitely presented as a sheaf of~$\B$-modules.
+\end{enumerate}
+\end{thm}
+
+\begin{proof}The proof of Theorem~\ref{thm:algebraic} can be easily adapted to
+this more general setting. Where that proof concludes~$a_J = 0$ for ring
+elements~$a_J$ by considering the localized situation~$A[a_J^{-1}]$, we now
+conclude~$a_J = 0$ for local sections~$a_J$ of~$\O_X$ by considering the
+situation over the restriction to~$D(a_J)$. It is then the assumption on~$X$
+instead of the reducedness assumption on~$A$ which allows us to conclude~$a_J = 0$.
+We omit further details.
+\end{proof}
+
+The spectrum of a ring~$A$ is a space of the kind required by
+Theorem~\ref{thm:general-generic-freeness} if and only if~$A$ is reduced,
+thus Theorem~\ref{thm:general-generic-freeness} indeed
+generalizes Theorem~\ref{thm:geometric}. Further examples for admissible spaces
+are given by any topological, smooth or complex manifold; a corollary
+of Theorem~\ref{thm:general-generic-freeness} for these examples is that
+quotients of vector bundles, computed in the category of sheaves of modules,
+are again vector bundles after restricting to suitable dense opens.
+
+It is hard to say with certainty that a given proof does not generalize to a
+new situation; but we do not see how this could be the case with the proofs
+cited in the introduction.
 

### 2018-06-03 paper-generic-freeness.tex
Layout with shorter rows
typo

@@
 \usepackage{hyperref}
-
-\usepackage{geometry}
 
@@
 Richman's uses of trivial rings~\cite{richman:trivial-rings}.
-\enlargethispage{8em}
-
-\begin{figure}[H]
-  \captionsetup{width=.9\linewidth}
-  \begin{tikzpicture}[scale=0.9]
-    \draw[step=1cm,gray,very thin] (0,0) grid (8,8);
-
-    \draw
-      (0,8) -- (0,0) -- (8,0);
-
-    \fill[pattern=north east lines,pattern color=gray,opacity=0.5]
-      (2,8) -- (2,5) -- (5,5) -- (5,3) -- (6,3) -- (6,2) -- (8,2) -- (8,8);
-    \draw
-      (2,8) -- (2,5) -- (5,5) -- (5,3) -- (6,3) -- (6,2) -- (8,2);
-
-    \fill[fill=black] (2,5.25) -- (2,5) -- (2.25,5) -- (2.25,5.25);
-    \fill[fill=black] (5,3.25) -- (5,3) -- (5.25,3) -- (5.25,3.25);
-    \fill[fill=black] (6,2.25) -- (6,2) -- (6.25,2) -- (6.25,2.25);
-
-    \fill[fill=blue!40!white,opacity=0.5]
-      (0,8) -- (0,0) -- (4,0) -- (4,3) -- (3,3) -- (3,5) -- (2,5) -- (2,8);
-
-    \fill[fill=red!40!white,opacity=0.5]
-      (3,3) -- (4,3) -- (4,4) -- (3,4);
-
-    % !perl -we 'for my $y (reverse 0..7) { for my $x (0..7) { print % "\$x^$x\Ey^$y\$ & " } print "\\\\ \n" }'
-    \begin{scope}[yshift=4cm, xshift=4cm]
-      \matrix[matrix of nodes,nodes={inner sep=0pt,text width=0.9cm,align=center,minimum height=0.9cm}]{
-        \tiny $x^0y^7v_1$ & \tiny $x^1y^7v_1$ & \tiny $x^2y^7v_1$ & \tiny $x^3y^7v_1$ & \tiny $x^4y^7v_1$ & \tiny $x^5y^7v_1$ & \tiny $x^6y^7v_1$ & \tiny $x^7y^7v_1$ & \\
-        \tiny $x^0y^6v_1$ & \tiny $x^1y^6v_1$ & \tiny $x^2y^6v_1$ & \tiny $x^3y^6v_1$ & \tiny $x^4y^6v_1$ & \tiny $x^5y^6v_1$ & \tiny $x^6y^6v_1$ & \tiny $x^7y^6v_1$ & \\
-        \tiny $x^0y^5v_1$ & \tiny $x^1y^5v_1$ & \tiny $x^2y^5v_1$ & \tiny $x^3y^5v_1$ & \tiny $x^4y^5v_1$ & \tiny $x^5y^5v_1$ & \tiny $x^6y^5v_1$ & \tiny $x^7y^5v_1$ & \\
-        \tiny $x^0y^4v_1$ & \tiny $x^1y^4v_1$ & \tiny $x^2y^4v_1$ & \tiny $x^3y^4v_1$ & \tiny $x^4y^4v_1$ & \tiny $x^5y^4v_1$ & \tiny $x^6y^4v_1$ & \tiny $x^7y^4v_1$ & \\
-        \tiny $x^0y^3v_1$ & \tiny $x^1y^3v_1$ & \tiny $x^2y^3v_1$ & \tiny $x^3y^3v_1$ & \tiny $x^4y^3v_1$ & \tiny $x^5y^3v_1$ & \tiny $x^6y^3v_1$ & \tiny $x^7y^3v_1$ & \\
-        \tiny $x^0y^2v_1$ & \tiny $x^1y^2v_1$ & \tiny $x^2y^2v_1$ & \tiny $x^3y^2v_1$ & \tiny $x^4y^2v_1$ & \tiny $x^5y^2v_1$ & \tiny $x^6y^2v_1$ & \tiny $x^7y^2v_1$ & \\
-        \tiny $x^0y^1v_1$ & \tiny $x^1y^1v_1$ & \tiny $x^2y^1v_1$ & \tiny $x^3y^1v_1$ & \tiny $x^4y^1v_1$ & \tiny $x^5y^1v_1$ & \tiny $x^6y^1v_1$ & \tiny $x^7y^1v_1$ & \\
-        \tiny $x^0y^0v_1$ & \tiny $x^1y^0v_1$ & \tiny $x^2y^0v_1$ & \tiny $x^3y^0v_1$ & \tiny $x^4y^0v_1$ & \tiny $x^5y^0v_1$ & \tiny $x^6y^0v_1$ & \tiny $x^7y^0v_1$ & \\
-      };
-    \end{scope}
-  \end{tikzpicture}
-  \caption{\label{fig:good}A graphical depiction of a good generating family (the non-hatched cells) in
-  the special case~$n = 2, m = 1$, writing~``$x$'' and~``$y$'' for~$x_1$ and~$x_2$. The hatched cells indicate vectors which
-  have already been removed from the family. The small black squares indicate
-  \emph{corners}. If the vector in the red cell will be found to be expressible
-  as a linear combination of vectors with smaller index (blue cells), it will
-  be removed, along with the vectors in all cells to the top and to the right
-  of the red cell.}
-\end{figure}
 
@@
 
+\begin{figure}[t]
+  \captionsetup{width=.9\linewidth}
+  \begin{tikzpicture}[scale=0.9]
+    \draw[step=1cm,gray,very thin] (0,0) grid (8,8);
+
+    \draw
+      (0,8) -- (0,0) -- (8,0);
+
+    \fill[pattern=north east lines,pattern color=gray,opacity=0.5]
+      (2,8) -- (2,5) -- (5,5) -- (5,3) -- (6,3) -- (6,2) -- (8,2) -- (8,8);
+    \draw
+      (2,8) -- (2,5) -- (5,5) -- (5,3) -- (6,3) -- (6,2) -- (8,2);
+
+    \fill[fill=black] (2,5.25) -- (2,5) -- (2.25,5) -- (2.25,5.25);
+    \fill[fill=black] (5,3.25) -- (5,3) -- (5.25,3) -- (5.25,3.25);
+    \fill[fill=black] (6,2.25) -- (6,2) -- (6.25,2) -- (6.25,2.25);
+
+    \fill[fill=blue!40!white,opacity=0.5]
+      (0,8) -- (0,0) -- (4,0) -- (4,3) -- (3,3) -- (3,5) -- (2,5) -- (2,8);
+
+    \fill[fill=red!40!white,opacity=0.5]
+      (3,3) -- (4,3) -- (4,4) -- (3,4);
+
+    % !perl -we 'for my $y (reverse 0..7) { for my $x (0..7) { print % "\$x^$x\Ey^$y\$ & " } print "\\\\ \n" }'
+    \begin{scope}[yshift=4cm, xshift=4cm]
+      \matrix[matrix of nodes,nodes={inner sep=0pt,text width=0.9cm,align=center,minimum height=0.9cm}]{
+        \tiny $x^0y^7v_1$ & \tiny $x^1y^7v_1$ & \tiny $x^2y^7v_1$ & \tiny $x^3y^7v_1$ & \tiny $x^4y^7v_1$ & \tiny $x^5y^7v_1$ & \tiny $x^6y^7v_1$ & \tiny $x^7y^7v_1$ & \\
+        \tiny $x^0y^6v_1$ & \tiny $x^1y^6v_1$ & \tiny $x^2y^6v_1$ & \tiny $x^3y^6v_1$ & \tiny $x^4y^6v_1$ & \tiny $x^5y^6v_1$ & \tiny $x^6y^6v_1$ & \tiny $x^7y^6v_1$ & \\
+        \tiny $x^0y^5v_1$ & \tiny $x^1y^5v_1$ & \tiny $x^2y^5v_1$ & \tiny $x^3y^5v_1$ & \tiny $x^4y^5v_1$ & \tiny $x^5y^5v_1$ & \tiny $x^6y^5v_1$ & \tiny $x^7y^5v_1$ & \\
+        \tiny $x^0y^4v_1$ & \tiny $x^1y^4v_1$ & \tiny $x^2y^4v_1$ & \tiny $x^3y^4v_1$ & \tiny $x^4y^4v_1$ & \tiny $x^5y^4v_1$ & \tiny $x^6y^4v_1$ & \tiny $x^7y^4v_1$ & \\
+        \tiny $x^0y^3v_1$ & \tiny $x^1y^3v_1$ & \tiny $x^2y^3v_1$ & \tiny $x^3y^3v_1$ & \tiny $x^4y^3v_1$ & \tiny $x^5y^3v_1$ & \tiny $x^6y^3v_1$ & \tiny $x^7y^3v_1$ & \\
+        \tiny $x^0y^2v_1$ & \tiny $x^1y^2v_1$ & \tiny $x^2y^2v_1$ & \tiny $x^3y^2v_1$ & \tiny $x^4y^2v_1$ & \tiny $x^5y^2v_1$ & \tiny $x^6y^2v_1$ & \tiny $x^7y^2v_1$ & \\
+        \tiny $x^0y^1v_1$ & \tiny $x^1y^1v_1$ & \tiny $x^2y^1v_1$ & \tiny $x^3y^1v_1$ & \tiny $x^4y^1v_1$ & \tiny $x^5y^1v_1$ & \tiny $x^6y^1v_1$ & \tiny $x^7y^1v_1$ & \\
+        \tiny $x^0y^0v_1$ & \tiny $x^1y^0v_1$ & \tiny $x^2y^0v_1$ & \tiny $x^3y^0v_1$ & \tiny $x^4y^0v_1$ & \tiny $x^5y^0v_1$ & \tiny $x^6y^0v_1$ & \tiny $x^7y^0v_1$ & \\
+      };
+    \end{scope}
+  \end{tikzpicture}
+  \caption{\label{fig:good}A graphical depiction of a good generating family (the non-hatched cells) in
+  the special case~$n = 2, m = 1$, writing~``$x$'' and~``$y$'' for~$x_1$ and~$x_2$. The hatched cells indicate vectors which
+  have already been removed from the family. The small black squares indicate
+  \emph{corners}. If the vector in the red cell will be found to be expressible
+  as a linear combination of vectors with smaller index (blue cells), it will
+  be removed, along with the vectors in all cells to the top and to the right
+  of the red cell.}
+\end{figure}
+
 \section{Conclusion}
@@
 Direct proofs sometimes generalize to new situations where the reduction techniques
-employed in more abstract proofs cannot be applied. This is the case for
+employed by more abstract proofs cannot be applied. This is the case for
 Grothendieck's generic freeness lemma, which allows for the following
@@
 It is hard to say with certainty that a given proof does not generalize to a
-new situation; but we do not see how this could be the case with the proofs
+new situation; but we do not see how this could be the case for the proofs
 cited in the introduction.

### 2018-06-08 slides-filmat2018.tex
Beginning on notes slides for the FilMat talk

@@
 
-\IfSubStr{\jobname}{\detokenize{nonotes}}{
-  \setbeameroption{hide notes}
-}{
+\setbeamertemplate{note page}[plain]
 \setbeameroption{show notes}
-}
-\setbeamertemplate{note page}[plain]
-\setbeameroption{hide notes}
 
@@
 \end{frame}
+
+\note{\justifying
+  The concept of a platonic heaven, where the
+  objects live which mathematicians study, can be mathematically realized by
+  the topos~$\Set$. As a category, it consists of all sets and all maps between
+  these sets.
+
+  But besides~$\Set$, there are many more toposes. Each such topos can be
+  thought of as a mathematical universe, containing their own versions of the
+  natural numbers,
+}
 

### 2018-06-09 paper-flabby-objects.tex
Beginning of paper on flabbiness

\documentclass[oneside]{amsart}

\usepackage[utf8]{inputenc}
\usepackage{amsthm,mathtools,tikz,float,caption}
\usetikzlibrary{patterns,matrix,decorations.pathreplacing}
\usepackage[all]{xy}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{hyperref}

\usepackage[natbib=true,style=numeric,maxnames=10]{biblatex}
\usepackage[babel]{csquotes}
\bibliography{paper-generic-freeness.bib}

\title{Flabby and injective objects in toposes}
\author{Ingo Blechschmidt}
\address{Max Planck Institute for Mathematics in the Sciences \\
Inselstraße 22 \\
04103 Leipzig, Germany}
\email{ingo.blechschmidt@mis.mpg.de}

\theoremstyle{definition}
\newtheorem{defn}{Definition}[section]
\newtheorem{ex}[defn]{Example}

\theoremstyle{plain}
\newtheorem{prop}[defn]{Proposition}
\newtheorem{cor}[defn]{Corollary}
\newtheorem{lemma}[defn]{Lemma}
\newtheorem{thm}[defn]{Theorem}
\newtheorem{scholium}[defn]{Scholium}

\theoremstyle{remark}
\newtheorem{rem}[defn]{Remark}
\newtheorem{question}[defn]{Question}
\newtheorem{speculation}[defn]{Speculation}
\newtheorem{caveat}[defn]{Caveat}
\newtheorem{conjecture}[defn]{Conjecture}

\newcommand{\xra}[1]{\xrightarrow{#1}}
\newcommand{\XXX}[1]{\textbf{XXX: #1}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\renewcommand{\_}{\mathpunct{.}\,}

\newcommand{\stacksproject}[1]{\cite[{\href{https://stacks.math.columbia.edu/tag/#1}{Tag~#1}}]{stacks-project}}

\begin{document}

\begin{abstract}
  We introduce a general notion of \emph{flabby objects} in elementary toposes
  and study its basic properties. In the special case of localic toposes, this
  notion reduces to the common notion of flabby sheaves, yielding a
  site-independent characterization of flabby sheaves. We use this notion to
  show that an internal notion of injective objects coincides with the
  corresponding external notion, in stark contrast with the situation for
  projective objects. As an application, we give internal characterizations of
  sheaf cohomology and show that higher direct images can be understood as
  internal cohomology.
\end{abstract}

\maketitle
\thispagestyle{empty}

\noindent

\section{Flabby sheaves}

A sheaf~$F$ on a topological space (or locale)~$X$ is \emph{flabby} if and only
if all restriction maps~$F(X) \to F(U)$ are surjective. The following
properties of flabby sheaves render them fundamental to the theory of sheaf
cohomology:
\begin{enumerate}
\item Let~$(U_i)_i$ be an open covering of~$X$.
A sheaf~$F$ is flabby if and only if all of its restrictions~$F|_{U_i}$
are flabby as sheaves on~$U_i$.
\item Let~$f : X \to Y$ be a continuous map. If~$F$ is a flabby sheaf on~$X$,
then~$f_*(F)$ is a flabby sheaf on~$Y$.
\item[(3a)] Let~$0 \to F \to G \to H \to 0$ be a short exact sequence of sheaves of
modules. If~$F$ is flabby, then this sequence is also exact as a sequence of
presheaves.
\item[(3b)] Let~$0 \to F \to G \to H \to 0$ be a short exact sequence of sheaves of
modules. If~$F$ and~$H$ are flabby, then so is~$G$.
% XXX: more?
\item[(4a)] Any sheaf can be embedded into a flabby sheaf.
\item[(4b)] Any sheaf of modules can be embedded into a flabby sheaf of modules.
\end{enumerate}

Since we want to develop an analogous theory for flabby objects in elementary
toposes, it is worthwhile to analyze the logical and set-theoretic commitments
which are required to establish these properties. The standard proofs of
properties~(1),~(3a),~(3b) and~(4b) require Zorn's lemma to construct maximal
extensions of given sections. Properties~(2) and~(4a) can be verified purely
intuitionistically.

There is an alternative definition of flabbiness, to be introduced below, which
is equivalent to the usual one in presence of Zorn's lemma and which
requires different commitments: For the alternative definition,
properties~(1),~(3b) and~(4a) can be verified purely intuitionistically, and
property~(2) can be verified purely intuitionistically for open maps~$f$.

Both definitions can be generalized to yield notions of flabby objects in
elementary toposes; but for toposes which are not localic, the two resulting
notions will differ, and only the one induced by the alternative definition
can be characterized in the internal language. We therefore adapt in this paper the
alternative one as the official definition.

\begin{defn}\label{defn:flabby-sheaf}
A sheaf~$F$ on a topological space (or locale)~$X$ is \emph{flabby}
if and only if for all opens~$U$ and all sections~$s \in F(U)$, there is an
open covering~$X = \bigcup_i U_i$ such that, for all~$i$, the section~$s$ can
be extended to a section on~$U \cup U_i$.\end{defn}

If~$F$ is a flabby sheaf in the traditional sense, then~$F$ is obviously also
flabby in the sense of Definition~\ref{defn:flabby-sheaf} -- singleton
coverings will do. Conversely, let~$F$ be a flabby sheaf in the sense of
Definition~\ref{defn:flabby-sheaf}. Let~$s \in F(U)$ be a local section. Zorn's
lemma implies that there is a maximal extension~$s' \in F(U')$. By assumption,
there is an open covering~$X = \bigcup_i U_i$ such that, for all~$i$, the
section~$s'$ can be extended to~$U' \cup U_i$. Since~$s'$ is maximal, $U' \cup
U_i = U'$ for all~$i$. Therefore~$X = \bigcup_i U_i \subseteq U'$; hence~$s'$ is a
global section, as desired.

We remark that unlike the traditional definition of flabbiness,
Definition~\ref{defn:flabby-sheaf} exhibits flabbiness as a manifestly local
notion.

\section{Flabby sets}

We intend this section to be applied in the internal language of an elementary
topos. We will therefore reason purely intuitionistically. We
adopt the terminology of~\cite{XXX:benno} regarding subterminals and
subsingletons: A subset~$K \subseteq X$ is \emph{subterminal} if and only if any given
elements are equal ($\forall x,y \in K\_ x = y$), and it is a
\emph{subsingleton} if and only if there is an element~$x \in X$ such that~$K
\subseteq \{ x \}$. Any subsingleton is trivially subterminal, but the converse
might fail.

\begin{defn}A set~$X$ is \emph{flabby} if and only if any subterminal subset
of~$X$ is a subsingleton, that is, if and only if for any subset~$K \subseteq
X$ such that~$\forall x,y \in K\_ x = y$, there exists an element~$x \in X$
such that~$K \subseteq \{ x \}$.
\end{defn}

In the presence of the law of excluded middle, a set is flabby if and only if
it is inhabited. Let~$\P_{\leq 1}(X)$ be the set of subterminals of~$X$.

\begin{prop}A set~$X$ is flabby if and only if the canonical map~$X \to
\P_{\leq 1}(X)$ which sends an element~$x$ to the singleton set~$\{x\}$ is
final.
\end{prop}

\begin{proof}By definition.\end{proof}

The set~$\P_{\leq 1}(X)$ of subterminals of $X$ can be interpreted as the set
of \emph{partially-defined elements} of $X$. In this view, the empty subset is
the maximally undefined element and a singleton is a maximally defined element.
A set is flabby if and only if any of its partially-defined elements can be
refined to an honest element.

\begin{defn}A set~$X$ is \emph{injective} if and only if, for any injection~$i
: A \to B$, any map~$f : A \to X$ can be extended to a map~$B \to X$.
An~$R$-module~$M$ is \emph{injective} if and only if, for any linear
injection~$i : A \to B$ between~$R$-modules, any linear map~$f : A \to M$ can
be extended to a linear map~$B \to X$.\end{defn}

In the presence of the law of excluded middle, a set is injective if and only
if it is inhabited. In the presence of the axiom of choice, an abelian group is
injective (as a~$\ZZ$-module) if and only if it is divisible. Injective sets
and modules have been intensively studied before~\cite{XXX}; the following
properties are well-known:

\begin{prop}\label{prop:basics-injective}
\begin{enumerate}
% Take care if numbering changes (referenced below)
\item Any set can be embedded into an injective set.
\item Any injective module is also injective as a set.
\item Assuming the axiom of choice, any module can be embedded into an
injective module.
\end{enumerate}\end{prop}

\begin{proof}\begin{enumerate}
\item One can check that, for instance, the full powerset~$\P(X)$ and the set of
subterminals~$\P_{\leq 1}(X)$ are each injective.~\cite{XXX}
\item The forgetful functor from modules to sets possesses a left exact left
adjoint. More explicitly, if~$i : A \to B$ is an injective map between sets and
if~$f : A \to M$ is an arbitrary map, then the induced map~$R\langle A \rangle
\to R\langle B \rangle$ between free modules is also injective, the given
map~$f$ lifts to a linear map~$R\langle A \rangle \to M$, and an
extension~$R\langle B \rangle \to M$ induces an extension~$B \to M$ of~$f$.
\item One verifies that any abelian group can be embedded into a divisible
abelian group. By the axiom of choice, divisible abelian groups are injective.
The result for modules over arbitrary rings then follows purely
formally~\cite{XXX}. \qedhere
\end{enumerate}\end{proof}

\begin{prop}\label{prop:injective-flabby}
Any injective set is flabby.\end{prop}

\begin{proof}Let~$X$ be an injective set. Let~$K \subseteq X$ be a subterminal.
The inclusion~$f : K \to X$ extends along the injection~$K \to 1 = \{\star\}$
to a map~$1 \to X$. The unique image~$x$ of that map has the property that~$K
\subseteq \{x\}$.\end{proof}

\begin{cor}\label{cor:enough-flabby-sets}
Any set can be embedded into a flabby set.\end{cor}

\begin{proof}Immediate by Proposition~\ref{prop:basics-injective}(1) and
Proposition~\ref{prop:injective-flabby}.\end{proof}

\begin{prop}\label{prop:hom-flabby}
\begin{enumerate}
\item Let~$I$ be an injective set. Let~$T$ be an arbitrary set. Then the
set~$I^T$ of maps from~$T$ to~$I$ is flabby.
\item Let~$I$ be an injective module. Let~$T$ be an arbitrary module. Then the
set~$\Hom_R(T,I)$ of linear maps from~$T$ to~$I$ is flabby.
\end{enumerate}\end{prop}

\begin{proof}We first cover the case of sets. Let~$K \subseteq I^T$ be a
subterminal. We consider the injectivity diagram
\[ \xymatrix{
  T' \ar[r]\ar[d] & T \ar@{-->}[ld] \\
  I
} \]
where~$T'$ is the subset~$\{ s \in T \,|\, \text{$K$ is inhabited} \} \subseteq T$ and the
solid vertical map sends~$s \in T'$ to~$g(s)$, where~$g$ is an arbitrary element
of~$K$. This association is well-defined. Since~$I$ is injective, a dotted lift
as indicated exists. If~$K$ is inhabited, this lift is an element of~$K$.

The same kind of argument applies to the case of modules. If~$K$ is a
subterminal of~$\Hom_R(T,I)$, we define~$T'$ to be the submodule
$\{ s \in I \,|\, \text{$s = 0$ or $K$ is inhabited} \}$ and consider the
analogous injectivity diagram, where the solid vertical map~$f : T' \to I$ is now
defined by cases: Let~$s \in T'$. If~$s = 0$, then we set~$f(s) = 0$; if~$K$ is
inhabited, then we set~$f(s) \defeq g(s)$, where~$g$ is an arbitrary element
of~$K$. This association is again well-defined, and a dotted lift yields the
desired element of~$\Hom_R(T,I)$.
\end{proof}

Proposition~\ref{prop:hom-flabby} can be used to give an alternative proof of
Proposition~\ref{prop:injective-flabby} and to generalize
Proposition~\ref{prop:injective-flabby} to modules: If~$I$ is an injective set,
then the set~$I^1 \cong I$ is flabby. If~$I$ is an injective module, then the
set~$\Hom_R(R,I) \cong I$ is flabby.

\begin{prop}\label{prop:set-of-extensions-flabby}
Let~$I$ be an injective set. Let~$i : A \to B$ be an injection.
Let~$f : A \to I$ be an arbitrary map. Then the set of extensions of~$f$ to~$B$
is flabby.\end{prop}

\begin{proof}Set~$X \defeq \{ \bar{f} \in I^B \,|\, \bar{f} \circ i =
f \}$. Let~$K \subseteq X$ be a subsingleton. We consider the injectivity diagram
\[ \xymatrix{
  i[A] \cup B' \ar@{^{(}->}[r]\ar[d]_g & B \ar@{-->}[ld] \\
  I
} \]
where~$B'$ is the set~$\{ s \in B \,|\, \text{$K$ is inhabited} \}$ and the solid
vertical arrow~$g$ is defined in the following way: Let~$s \in i[A] \cup B'$.
If~$s \in i[A]$, then we set~$g(s) \defeq f(a)$, where~$a \in A$ is an element such
that~$s = i(a)$. If~$s \in B'$, then we set~$g(s) \defeq \bar{f}(s)$,
where~$\bar{f}$ is any element of~$K$. These prescriptions determine a well-defined
map.

Since~$I$ is injective, there exists a dotted map rendering the diagram
commutative. This map is an element of~$X$. Furthermore, if~$K$ is inhabited,
then this map is an element of~$K$.
\end{proof}

\begin{prop}Let~$0 \to M' \xra{i} M \xra{p} M'' \to 0$ be a short exact
sequence of modules. Let~$s \in M''$. If~$M'$ is flabby, then the set of
preimages of~$s$ under~$p$ is flabby.
\end{prop}

\begin{proof}For the first statement, let~$X \defeq \{ u \in M \,|\, p(u) = s
\}$. Let~$K \subseteq X$ be a subterminal. Since~$p$ is surjective, there is a
preimage~$u_0 \in X$. The translated set~$K - u_0 \subseteq M$ is still a
subterminal, and its preimage under~$i$ is as well. Since~$M'$ is flabby, there
is an element~$v \in M'$ such that~$i^{-1}[K - u_0] \subseteq \{v\}$. We verify
that~$K \subseteq \{u_0 + i(v)\}$.

Thus let~$u \in K$ be given. Then~$p(u - u_0) = 0$, so by exactness the
set~$i^{-1}[K - u_0]$ is inhabited. It therefore contains~$v$. Thus~$i(v) \in K
- u_0$. Since~$K = \{u\}$, it follows that~$i(v) = u - u_0$, so~$u \in \{u_0 +
i(v)\}$ as claimed.
\end{proof}

\begin{prop}Let~$0 \to M' \xra{i} M \xra{p} M'' \to 0$ be a short exact
sequence of modules. If~$M'$ and~$M''$ are flabby, so is~$M$.
\end{prop}

\begin{proof}Let~$K \subseteq M$ be a subterminal. Then its image~$p[K] \subseteq M''$
is a subterminal as well. Since~$M''$ is flabby, there is an element~$s \in
M''$ such that~$p[K] \subseteq \{ s \}$.

Since~$p$ is surjective, there is an element~$u_0 \in M$ such that~$p(u_0) =
s$.

The preimage~$i^{-1}[K - u_0] \subseteq M'$ is a subterminal. Since~$M'$ is
flabby, there exists an element~$v \in M'$ such that~$i^{-1}[K - u_0] \subseteq
\{v\}$.

Thus~$K \subseteq \{ u_0 + i(v) \}$.
\end{proof}

\section{Flabby objects}

\begin{defn}An object~$X$ of an elementary topos~$\E$ is \emph{flabby} if and
only if it is a flabby set from the point of view of the internal language
of~$\E$.\end{defn}

This definition amounts to the following: An object~$X$ of an elementary
topos~$\E$ is flabby if and only if, for any object~$A$ of~$\E$, the object~$X'
\defeq X \times A \in \E/A$ has the following property:
\begin{quote}
For any monomorphism~$K \to 1$ in~$\E/A$ and any morphism~$K \to X'$%
\tikz[remember picture] \node[coordinate,yshift=0.7em,xshift=1.0em] (n1) {};
in~$\E/A$, there exists an epimorphism~$B \to 1$ in~$\E/A$ and a morphism~$B
\to X'$ in~$\E/A$ such that the following diagram commutes.
\end{quote}
\[ \xymatrix{
  K \times_A B \ar[r]\ar[rd] & X' \times_A B \ar[r] & X' \\
  & B \ar[ru]
} \]
\tikz[remember picture] \node[coordinate,yshift=0.3em] (n2) {};
\begin{tikzpicture}[overlay,remember picture]
  \path (n2) -| node[coordinate] (n3) {} (n1);
  \draw[thick,decorate,decoration={brace,amplitude=3pt}]
  (n1) -- (n3) node[midway, right=4pt] {$({\star})$};
\end{tikzpicture}

Instead of referencing arbitrary stages~$A \in \E$, one can also just reference
the generic stage: Let~$\P_{\leq1}(X)$ denote the \emph{object of subterminals}
of~$X$; this object is a certain suboject of~$\P(X) = [X,\Omega_\E]_\E$, the
powerobject of~$X$. The subobject~$K$ of~$\P_{\leq1}(X) \times X$ classified by the
evaluation morphism~$\P_{\leq1}(X) \times X \to \P(X) \times X \to \Omega_\E$
is the \emph{generic subterminal} of~$X$. The object~$X$ is flabby if and only
if there exists an epimorphism~$B \to \P_{\leq1}(X)$ and a morphism~$B \to X$
such that the following diagram commutes.
\[ \xymatrix{
  K \times_{\P_{\leq1}(X)} B \ar[r]\ar[rd] & X \times B \ar[r] & X \\
  & B \ar[ru]
} \]

\begin{prop}\label{prop:flabby-sheaves-objects}
Let~$F$ be a sheaf on a topological space~$X$ (or a locale).
Then~$F$ is flabby as a sheaf if and only if~$F$ is a flabby object of the
sheaf topos~$\Sh(X)$.
\end{prop}

\begin{proof}It suffices to verify condition~$({\star})$ for stages of the
form~$A = \Hom(\cdot,U)$, where~$U$ is an open of~$X$. XXX
\end{proof}

\begin{prop}Let~$f : \E \to \F$ be an open geometric morphism between
elementary toposes. Let~$X \in \E$ be a flabby object. Then~$f_*(X)$ is flabby
as well.\end{prop}

\begin{proof}Without loss of generality, it suffices to verify
condition~$({\star})$ for the terminal stage~$A = 1$. Let~$k : K \to 1$ be a
monomorphism in~$\F$ and let~$x : K \to f_*(X)$ be an arbitrary morphism.
Then~$f^*(k) : f^*(K) \to 1$ is a monomorphism in~$\E$ and~$x^t : f^*(K) \to X$
is a morphism in~$\E$. Since~$X$ is flabby, there is an epimorphism~$B \to 1$
in~$\E$ and a morphism~$y : B \to X$ such that the morphism~$f^*(K) \times B \to X
\times B \to B$ factors over~$y$. XXX
\end{proof}

\begin{defn}An object~$I$ of an elementary topos~$\E$ is \emph{externally
injective} if and only if for any monomorphism~$A \to B$ in~$\E$, the canonical
map~$\Hom_\E(B,I) \to \Hom_\E(A,I)$ is surjective. It is \emph{internally
injective} if and only if for any monomorphism~$A \to B$ in~$\E$, the canonical
morphism~$[B,I] \to [A,I]$ between Hom objects is an epimorphism in~$\E$.
\end{defn}

If~$R$ is a ring in an elementary topos~$\E$, a similar definition can be given
for~$R$-modules in~$\E$, referring only to the set respectively the object of
linear maps. The condition for an object to be internally injective can be
rephrased in various ways. The following proposition lists five of these
conditions. The equivalence of the first four is due to
Harting~\cite{harting:XXX}.

\begin{prop}\label{prop:notions-of-internal-injectivity}
Let~$\E$ be an elementary topos. Then the following statements about an
object~$I \in \E$ are equivalent.
\begin{enumerate}
\item $I$ is internally injective.
\item The functor~$[\cdot, I] : \E^\op \to \E$ maps monomorphisms in $\E$
to morphisms for which any global element of the target locally (after change of
base along an epimorphism) possesses a preimage.
\item For any morphism $p : A \to 1$ in $\E$, the object $p^*(I)$ has property~(1)
as an object of $\E/A$.
\item For any morphism $p : A \to 1$ in $\E$, the object $p^*(I)$ has property~(2)
as an object of $\E/A$.
\item From the point of view of the internal language of~$\E$, the object~$I$
is an injective set.
\end{enumerate}
\end{prop}

\begin{proof}
The implications ``(1)~$\Rightarrow$~(2)'', ``(3)~$\Rightarrow$~(4)'',
``(3)~$\Rightarrow$~(1)'', and ``(4)~$\Rightarrow$~(2)'' are trivial.

The equivalence ``(3)~$\Leftrightarrow$~(5)'' follows directly from the
interpretation rules of the stack semantics.

The implication ``(2)~$\Rightarrow$~(4)'' employs the
extra left adjoint $p_! : \E/A \to \E$ of $p^* : \E
\to \E/A$~(which maps an object~$(X \to A)$ to~$X$), as in the usual proof that
injective sheaves remain injective when
restricted to smaller open subsets: We have that $p_* \circ [\cdot, p^*(I)]_{\E/A}
\cong [\cdot, I]_\E \circ p_!$, the functor $p_!$ preserves monomorphisms, and one
can check that $p_*$ reflects the property that global elements locally possess
preimages. Details are in~\cite[Thm.~1.1]{harting}.\footnote{Harting formulates
the statement for abelian group objects, and has to assume that~$\E$ contains a
natural numbers object to ensure the existence of an abelian version of~$p_!$.}

The implication ``(4)~$\Rightarrow$~(3)'' follows by performing an extra change of
base, since any non-global element becomes a global element after a suitable
change of base.
\end{proof}

Let~$R$ be a ring in~$\E$. Then the analogue of
Proposition~\ref{prop:notions-of-internal-injectivity} holds for~$R$-modules
in~$\E$, if~$\E$ is assumed to have a natural numbers object. The extra
assumption is needed in order to construct the left adjoint~$p_! :
\Mod_{\E/A}(R \times A) \to \Mod_\E(R)$. Phrased in the internal language, this
adjoint maps a family~$(M_a)_{a \in A}$ of~$R$-modules to the direct
sum~$\bigoplus_{a \in A} M_a$.

% XXX: Explain construction. Show that p_! preserves monos.

Somewhat surprisingly, and in stark contrast with the situation for internally
projective objects (which are defined dually), internal injectivity coincides
with external injectivity for localic toposes.

\begin{thm}\label{thm:injectivity-external-internal}
Let~$I$ be an object of an elementary topos~$\E$. If~$I$ is externally
injective, then~$I$ is also internally injective. The converse holds if~$\E$ is
localic and Zorn's lemma is available.
\end{thm}

\begin{proof}For the ``only if'' direction, let~$I$ be an object
which is externally injective. Then~$I$ satisfies Condition~(2) in
Proposition~\ref{prop:notions-of-internal-injectivity}, even without having to
pass to covers.

For the ``if'' direction, write~$\E = \Sh(X)$ and let~$I$ be an internally
injective object. Let~$i : A \to B$ be a monomorphism in~$\Sh(X)$ and let~$f :
A \to I$ be an arbitrary morphism. We want to show that there exists an
extension $B \to I$ of~$f$ along~$i$. To this end, we consider the sheaf of
such extensions, defined by the internal expression
\[ F \defeq \{ \bar{f} \in [B,I] \,|\, \bar{f} \circ i = f \}. \]
Global elements of~$F$ are extensions of the kind we are looking for.
By Proposition~\ref{prop:set-of-extensions-flabby}, this sheaf is a flabby
object and therefore by Proposition~\ref{prop:flabby-sheaves-objects} a flabby
sheaf. Zorn's lemma implies that~$F$ has a global element.
\end{proof}

\section{Characterizing cohomology}

% EXT, TOR
% Higher direct images

\section{Conclusion}

We originally set out to develop an intuitionistic account of Grothendieck's
sheaf cohomology. Čech methods can be carried out constructively, and
XXX(Barakat), but it appears that there is not a general framework for sheaf
cohomology which works in an intuitionistic metatheory.

The main obstacle preventing Grothendieck's theory to be interpreted
constructively is its reliance on injective resolutions. It is known that in
the absence of the axiom of choice, much less in a purely intuitionistic
context, there might not be any nontrivial injective abelian
group~\cite{blass:XXX}.

In principle, this problem could be remedied by employing flabby resolutions
instead of injective ones. There are, however, two problems with this
suggestion. Firstly, we needed Zorn's lemma to show that flabby sheaves are
acyclic for the global sections functor (Proposition~XXX); however, we noticed
the substitute property in XXX which might be good enough for further
developments. The more serious problem is that it is an open question whether
one can show, purely intuitionistically, that any sheaf of modules embeds into
a flabby sheaf of modules. The following is known about this problem:

\begin{enumerate}
\item There is a purely intuitionistic proof that any sheaf of sets embeds into
a flabby sheaf of sets (Corollary~\ref{cor:enough-flabby-sets}).
\item The existence of enough flabby modules, and even the existence of enough
injective modules, is \emph{not} a constructive taboo, that is, they do not
entail a classical principle like the law of excluded middle or the principle
of omniscience. We know this since assuming the axiom of choice, any
Grothendieck topos has enough injective (and therefore flabby) modules. 
\item There is a way of embedding any module into a flabby module if one is
prepared to ignore set-theoretical difficulties. Namely, let~$M$ be an~$R$-module.
Inductively construct a collection~$T$ of terms by the following clauses: $0 \in T$;
if~$x,y \in T$, then~$x + y \in T$; if~$x \in T$ and~$r \in R$, then~$rx \in
T$; if~$x \in M$, then~$\underline{x} \in T$; if~$K \subseteq T$ is a
subterminal, then~$\varepsilon_K \in T$. Then~$T/{\sim}$, where~$({\sim})$ is
the finest equivalence relation such that the module axioms hold, such
that~$\underline{x+y} \sim \underline{x} + \underline{y}$, $\underline{rx} \sim
r \underline{x}$, $\underline{0} \sim 0$ and such that~$\varepsilon_{\{t\}} \sim
t$ for all terms~$t$, is a flabby~$R$-module into which~$M$ embeds.
\end{enumerate}

We currently believe that it is not possible to give a constructive account of
a global cohomology functor. However, XXX.

\end{document}

### 2018-06-12 paper-flabby-objects.tex
Flabbiness of the set of linear extensions

@@
 
+\begin{prop}\label{prop:set-of-linear-extensions-flabby}
+Let~$I$ be an injective~$R$-module. Let~$i : A \to B$ be a linear injection.
+Let~$f : A \to I$ be an arbitrary linear map. Then the set of linear extensions of~$f$ to~$B$
+is flabby.\end{prop}
+
+\begin{proof}The proof is similar to the proof of
+Proposition~\ref{prop:set-of-extensions-flabby}. We
+set~$X \defeq \{ \bar{f} \in \Hom_R(B,I) \,|\, \bar{f} \circ i =
+f \}$. Let~$K \subseteq X$ be a subsingleton. We consider the injectivity diagram
+\[ \xymatrix{
+  i[A] + B' \ar@{^{(}->}[r]\ar[d]_g & B \ar@{-->}[ld] \\
+  I
+} \]
+where~$B'$ is the submodule~$\{ s \in B \,|\, \text{$s = 0$ or $K$ is
+inhabited} \} \subseteq B$ and the solid vertical arrow~$g$ is defined in the following
+way: Let~$s \in i[A] + B'$. Then~$s = i(a) + t$ for an element~$a \in A$ and an
+element~$t \in B'$. Since~$t \in B'$, $t = 0$ or~$K$ is inhabited. If~$t = 0$,
+we set~$g(s) \defeq f(a)$. If~$K$ is inhabited, we set~$g(s) \defeq f(a) +
+\bar{f}(s)$, where~$\bar{f}$ is any element of~$K$. These prescriptions
+determine a well-defined map.
+
+Since~$I$ is injective, there exists a dotted map rendering the diagram
+commutative. This map is an element of~$X$. Furthermore, if~$K$ is inhabited,
+then this map is an element of~$K$.
+\end{proof}
+
 \begin{prop}Let~$0 \to M' \xra{i} M \xra{p} M'' \to 0$ be a short exact

### 2018-06-12 paper-flabby-objects.tex
A constructive taboo
Various additions

@@
 \usepackage[babel]{csquotes}
-\bibliography{paper-generic-freeness.bib}
+\bibliography{paper-flabby-objects.bib}
 
@@
 \DeclareMathOperator{\Sh}{Sh}
+\DeclareMathOperator{\PSh}{PSh}
+\newcommand{\Set}{\mathrm{Set}}
 \renewcommand{\_}{\mathpunct{.}\,}
@@
 
-A sheaf~$F$ on a topological space (or locale)~$X$ is \emph{flabby} if and only
+A sheaf~$F$ on a topological space (or locale)~$X$ is \emph{flabby} (flasque) if and only
 if all restriction maps~$F(X) \to F(U)$ are surjective. The following
@@
 \item Let~$(U_i)_i$ be an open covering of~$X$.
-A sheaf~$F$ is flabby if and only if all of its restrictions~$F|_{U_i}$
+A sheaf~$F$ on~$X$ is flabby if and only if all of its restrictions~$F|_{U_i}$
 are flabby as sheaves on~$U_i$.
@@
 then~$f_*(F)$ is a flabby sheaf on~$Y$.
-\item[(3a)] Let~$0 \to F \to G \to H \to 0$ be a short exact sequence of sheaves of
-modules. If~$F$ is flabby, then this sequence is also exact as a sequence of
+\item[(3)] Let~$0 \to F \to G \to H \to 0$ be a short exact sequence of sheaves of
+modules.
+\begin{enumerate}
+\item If~$F$ is flabby, then this sequence is also exact as a sequence of
 presheaves.
-\item[(3b)] Let~$0 \to F \to G \to H \to 0$ be a short exact sequence of sheaves of
-modules. If~$F$ and~$H$ are flabby, then so is~$G$.
-% XXX: more?
+\item If~$F$ and~$H$ are flabby, then so is~$G$.
+\item If~$F$ and~$G$ are flabby, then so is~$H$.
+\end{enumerate}
 \item[(4a)] Any sheaf can be embedded into a flabby sheaf.
@@
 which are required to establish these properties. The standard proofs of
-properties~(1),~(3a),~(3b) and~(4b) require Zorn's lemma to construct maximal
-extensions of given sections. Properties~(2) and~(4a) can be verified purely
+properties~(1),~(3a),~(3b) and~(3c) require Zorn's lemma to construct maximal
+extensions of given sections. The standard proof of property~(4b) requires the
+law of excluded middle, to ensure that the Godement construction actually
+yields a flabby sheaf. Properties~(2) and~(4a) can be verified purely
 intuitionistically.
@@
 properties~(1),~(3b) and~(4a) can be verified purely intuitionistically, and
-property~(2) can be verified purely intuitionistically for open maps~$f$.
+property~(2) can be verified purely intuitionistically for open injections~$f$.
+There is a substitute for property~(3a) which can be verified purely
+intuitionistically. We comment on the status of property~(4b) in
+Section~\ref{sect:conclusion}.
 
@@
 topos. We will therefore reason purely intuitionistically. We
-adopt the terminology of~\cite{XXX:benno} regarding subterminals and
+adopt the terminology of~\cite{kock:partial-maps} regarding subterminals and
 subsingletons: A subset~$K \subseteq X$ is \emph{subterminal} if and only if any given
@@
 In the presence of the law of excluded middle, a set is flabby if and only if
-it is inhabited. Let~$\P_{\leq 1}(X)$ be the set of subterminals of~$X$.
+it is inhabited. In fact, this characterization is a constructive taboo:
+
+\begin{prop}If any inhabited set is flabby, then the law of excluded middle
+holds.
+\end{prop}
+
+\begin{proof}Let~$\varphi$ be a truth value. The set~$X \defeq \{ 0 \}
+\cup \{ 1 \,|\, \varphi \} \subseteq \{ 0,1 \}$ is inhabited by~$0$ and
+contains~$1$ if and only if~$\varphi$ holds. Let~$K$ be the subterminal~$\{ 1 \,|\, \varphi
+\} \subseteq X$. Flabbiness of~$X$ implies that there exists an element~$x \in
+X$ such that~$K \subseteq \{x\}$. We have~$x = 0$ or~$x = 1$. The first case
+entails~$\neg\varphi$. The second case entails~$1 \in X$, so~$\varphi$.
+\end{proof}
+
+Let~$\P_{\leq 1}(X)$ be the set of subterminals of~$X$.
 
@@
 
+\begin{prop}Any singleton set is flabby. The cartesian product of flabby sets
+is flabby.\end{prop}
+
+\begin{proof}Immediate.\end{proof}
+
+Subsets of flabby sets are in general not flabby, as else any set would be
+flabby in view of Corollary~\ref{cor:enough-flabby-sets}.
+
 \begin{prop}\label{prop:hom-flabby}
@@
 set~$I^T$ of maps from~$T$ to~$I$ is flabby.
-\item Let~$I$ be an injective module. Let~$T$ be an arbitrary module. Then the
+\item Let~$I$ be an injective~$R$-module. Let~$T$ be an arbitrary~$R$-module. Then the
 set~$\Hom_R(T,I)$ of linear maps from~$T$ to~$I$ is flabby.
@@
 adjoint maps a family~$(M_a)_{a \in A}$ of~$R$-modules to the direct
-sum~$\bigoplus_{a \in A} M_a$.
-
-% XXX: Explain construction. Show that p_! preserves monos.
+sum~$\bigoplus_{a \in A} M_a$. Details on this construction, phrased in the
+language of sets but interpretable in the internal language, can for instance
+be found in~\cite[page~54]{mines-richman-ruitenburg:constructive-algebra}.
 
@@
 
+\begin{rem}A generalization of the traditional notion of a flabby sheaf to
+elementary toposes is the following. An object~$X$ of an elementary topos~$\E$
+is \emph{strongly flabby} if and only if, for any monomorphism~$K \to 1$
+in~$\E$, any morphism~$K \to X$ lifts to a morphism~$1 \to X$.
+
+One can verify, purely intuitionistically, that a sheaf~$F$ on a space~$T$ is
+flabby in the traditional sense if and only if~$F$ is a strongly flabby object
+of~$\Sh(T)$.
+
+The notion of strongly flabby objects is, however, not stable under base change
+and therefore cannot be characterized in the internal language. A specific
+example is the~$G$-set~$G$ (with the translation action), considered as an
+object of the topos~$\PSh(BG)$ of~$G$-sets, where~$G$ is a nontrivial group.
+This object is not strongly flabby, since the morphism~$\emptyset \to G$ does
+not lift, but its pullback to the slice~$\PSh(BG)/G \simeq \Set$ is (assuming
+the law of excluded middle).
+\end{rem}
+
 \section{Characterizing cohomology}
@@
 \section{Conclusion}
+\label{sect:conclusion}
 
@@
 context, there might not be any nontrivial injective abelian
-group~\cite{blass:XXX}.
+group~\cite{blass:inj-proj-axc}.
 
@@
 \item The existence of enough flabby modules, and even the existence of enough
-injective modules, is \emph{not} a constructive taboo, that is, they do not
+injective modules, is \emph{not} a constructive taboo, that is, these statements do not
 entail a classical principle like the law of excluded middle or the principle
-of omniscience. We know this since assuming the axiom of choice, any
+of omniscience. This is because assuming the axiom of choice, any
 Grothendieck topos has enough injective (and therefore flabby) modules. 
@@
 prepared to ignore set-theoretical difficulties. Namely, let~$M$ be an~$R$-module.
-Inductively construct a collection~$T$ of terms by the following clauses: $0 \in T$;
-if~$x,y \in T$, then~$x + y \in T$; if~$x \in T$ and~$r \in R$, then~$rx \in
+Inductively construct a collection~$T$ of terms by the following clauses: $0
+\in T$ (where~$0$ is a formal symbol);
+if~$t,s \in T$, then~$t + s \in T$; if~$t \in T$ and~$r \in R$, then~$rt \in
 T$; if~$x \in M$, then~$\underline{x} \in T$; if~$K \subseteq T$ is a
-subterminal, then~$\varepsilon_K \in T$. Then~$T/{\sim}$, where~$({\sim})$ is
-the finest equivalence relation such that the module axioms hold, such
+subterminal, then~$\varepsilon_K \in T$. Let~$({\sim})$ be
+the finest equivalence relation on~$T$ such that~$t + (s + u) \sim (t + s) +
+u$, $t + s \sim s + t$, $t + 0 \sim t \sim 0 + t$, $1t \sim t$, $r(t+s) \sim rt
++ rs$, $(r+r')t \sim rt + r't$ and such
 that~$\underline{x+y} \sim \underline{x} + \underline{y}$, $\underline{rx} \sim
 r \underline{x}$, $\underline{0} \sim 0$ and such that~$\varepsilon_{\{t\}} \sim
-t$ for all terms~$t$, is a flabby~$R$-module into which~$M$ embeds.
+t$ for all terms~$t$. Then the quotient~$T/({\sim})$ is a flabby~$R$-module
+into which~$M$ embeds.
 \end{enumerate}

### 2018-06-14 paper-flabby-objects.tex
Nicer wording of the definition of flabby objects

@@
 \begin{defn}An object~$X$ of an elementary topos~$\E$ is \emph{flabby} if and
-only if it is a flabby set from the point of view of the internal language
+only if the statement~``$X$ is a flabby set'' holds in the stack semantics
 of~$\E$.\end{defn}
@@
 This definition amounts to the following: An object~$X$ of an elementary
-topos~$\E$ is flabby if and only if, for any object~$A$ of~$\E$, the object~$X'
-\defeq X \times A \in \E/A$ has the following property:
-\begin{quote}
-For any monomorphism~$K \to 1$ in~$\E/A$ and any morphism~$K \to X'$%
-\tikz[remember picture] \node[coordinate,yshift=0.7em,xshift=1.0em] (n1) {};
-in~$\E/A$, there exists an epimorphism~$B \to 1$ in~$\E/A$ and a morphism~$B
-\to X'$ in~$\E/A$ such that the following diagram commutes.
-\end{quote}
+topos~$\E$ is flabby if and only if, for any monomorphism~$K \to A$ and any
+morphism~$K \to X$, there exists an epimorphism~$B \to A$ and a morphism~$B
+\to X$ such that the following diagram commutes.
 \[ \xymatrix{
-  K \times_A B \ar[r]\ar[rd] & X' \times_A B \ar[r] & X' \\
-  & B \ar[ru]
+  K \times_A B \ar@{^{(}->}[r]\ar[d] & B \ar@{-->}@/^/[ld] \\
+  X
 } \]
-\tikz[remember picture] \node[coordinate,yshift=0.3em] (n2) {};
-\begin{tikzpicture}[overlay,remember picture]
-  \path (n2) -| node[coordinate] (n3) {} (n1);
-  \draw[thick,decorate,decoration={brace,amplitude=3pt}]
-  (n1) -- (n3) node[midway, right=4pt] {$({\star})$};
-\end{tikzpicture}
 
@@
 of~$X$; this object is a certain suboject of~$\P(X) = [X,\Omega_\E]_\E$, the
-powerobject of~$X$. The subobject~$K$ of~$\P_{\leq1}(X) \times X$ classified by the
+powerobject of~$X$. The subobject~$K_0$ of~$\P_{\leq1}(X) \times X$ classified by the
 evaluation morphism~$\P_{\leq1}(X) \times X \to \P(X) \times X \to \Omega_\E$
@@
 \[ \xymatrix{
-  K \times_{\P_{\leq1}(X)} B \ar[r]\ar[rd] & X \times B \ar[r] & X \\
-  & B \ar[ru]
+  K_0 \times_{\P_{\leq1}(X)} B \ar@{^{(}->}[r]\ar[d] & \P_{\leq1}(X) \ar@{-->}@/^/[ld] \\
+  X
 } \]
+
+\begin{prop}\label{prop:basic-properties-of-flabby-objects}
+Let~$X$ and~$T$ be objects of an elementary topos~$\E$.
+\begin{enumerate}
+\item If~$X$ is flabby, so is~$X \times T$ as an object in~$\E/T$.
+\item The converse holds if the unique morphism~$T \to 1$ is an epimorphism.
+\end{enumerate}
+\end{prop}
+
+\begin{proof}These properties are shared by any property which can be defined
+in the stack semantics~\cite[Lemma~7.3]{shulman:stack-semantics}.
+\end{proof}
 

### 2018-06-14 paper-flabby-objects.tex
Proof that flabby sheaves are flabby objects

@@
 
-\begin{proof}It suffices to verify condition~$({\star})$ for stages of the
-form~$A = \Hom(\cdot,U)$, where~$U$ is an open of~$X$. XXX
+\begin{proof}The proof is routine; we only verify ``only if'' direction.
+Let~$F$ be flabby as a sheaf. It suffices to verify the definition for stages
+of the form~$A = \Hom(\cdot,U)$, where~$U$ is an open of~$X$. A monomorphism~$K
+\to A$ then amounts to an open~$V \subseteq U$ (the union of all opens on
+which~$K$ is inhabited). A morphism~$K \to F$ amounts to a section~$s \in
+F(V)$. Since~$F$ is flabby as a sheaf, there is an open covering~$X =
+\bigcup_{i \in I} V_i$ such that, for all~$i$, the section~$s$ can be extended
+to a section~$s_i$ of~$V \cup V_i$. The desired epimorphism is~$B \defeq
+\coprod_i \Hom(\cdot,(V \cup V_i) \cap U) \to A$, and the desired morphism~$B
+\to X$ is given by the sections~$s_i|_{(V \cup V_i) \cap U}$.
+
+As stated, the argument in the previous paragraph requires the axiom of choice
+to pick the extensions~$s_i$; this can be avoided by a standard trick of
+expanding the index set of the coproduct to include the choices: We redefine $B \defeq
+\coprod_{(i,t) \in I'} \Hom(\cdot, (V \cup V_i) \cap U)$, where~$I' = \{ (i \in
+I, t \in F(V \cup V_i)) \,|\, t|_V = s \}$ and define the morphism~$B \to X$ on
+the~$(i,t)$-summand by~$t|_{(V \cup V_i) \cap U}$.
 \end{proof}

### 2018-06-14 paper-flabby-objects.tex
Pushforward of flabby objects

@@
 
-\begin{prop}Let~$f : \E \to \F$ be an open geometric morphism between
-elementary toposes. Let~$X \in \E$ be a flabby object. Then~$f_*(X)$ is flabby
-as well.\end{prop}
+\begin{prop}\label{prop:pushforward-of-flabby-objects}
+Let~$f : \F \to \E$ be a geometric morphism. If~$f_*$ preserves epimorphisms,
+then~$f_*$ preserves flabby objects.\end{prop}
 
-\begin{proof}Without loss of generality, it suffices to verify
-condition~$({\star})$ for the terminal stage~$A = 1$. Let~$k : K \to 1$ be a
-monomorphism in~$\F$ and let~$x : K \to f_*(X)$ be an arbitrary morphism.
-Then~$f^*(k) : f^*(K) \to 1$ is a monomorphism in~$\E$ and~$x^t : f^*(K) \to X$
-is a morphism in~$\E$. Since~$X$ is flabby, there is an epimorphism~$B \to 1$
-in~$\E$ and a morphism~$y : B \to X$ such that the morphism~$f^*(K) \times B \to X
-\times B \to B$ factors over~$y$. XXX
+\begin{proof}Let~$X \in \F$ be a flabby object. Without loss of generality, it
+suffices to verify condition~$({\star})$ for the terminal stage~$A = 1 \in \E$.
+Let~$k : K \to 1$ be a monomorphism in~$\E$ and let~$x : K \to f_*(X)$ be an arbitrary morphism.
+Then~$f^*(k) : f^*(K) \to 1$ is a monomorphism in~$\F$ and~$x^t : f^*(K) \to X$
+is a morphism in~$\F$. Since~$X$ is flabby, there is an epimorphism~$B \to 1$
+in~$\F$ and a morphism~$y : B \to X$ such that the morphism~$f^*(K) \times B
+\to X$ factors over~$y$. Hence~$x$ factors over~$f_*(y) : f_*(B) \to f_*(X)$.
+By assumption, the morphism~$f_*(B) \to f_*(1)$ is an epimorphism.
 \end{proof}
+
+The assumption on~$f_*$ of Proposition~\ref{prop:pushforward-of-flabby-objects}
+is for instance satisfied if~$f$ is a local geometric morphism or if~$f$ is
+induced by an open continuous injection between topological spaces. % XXX
 

### 2018-06-14 paper-flabby-objects.tex
More on flabby objects

@@
   notion reduces to the common notion of flabby sheaves, yielding a
-  site-independent characterization of flabby sheaves. We use this notion to
+  site-independent characterization of flabby sheaves. Continuing a line of
+  research started by Roswitha Harting, we use this notion to
   show that an internal notion of injective objects coincides with the
@@
 
-\begin{prop}Let~$0 \to M' \xra{i} M \xra{p} M'' \to 0$ be a short exact
+\begin{prop}\label{prop:set-of-preimages-flabby}
+Let~$0 \to M' \xra{i} M \xra{p} M'' \to 0$ be a short exact
 sequence of modules. Let~$s \in M''$. If~$M'$ is flabby, then the set of
@@
 conditions. The equivalence of the first four is due to
-Harting~\cite{harting:XXX}.
+Harting~\cite{harting:locally-injective}.
 
@@
 be found in~\cite[page~54]{mines-richman-ruitenburg:constructive-algebra}.
+It seems that Harting was not aware that this construction is possible, as she
+(mistakingly) states in~\cite[page~233]{harting:remark} that ``the notions of
+injectivity and internal injectivity do not coincide'' for modules.
 
@@
 \end{proof}
+
+Since we were careful in Section~\ref{sect:flabby-sets} to use the law of
+excluded middle or the axiom of choice only where needed, most results of that
+section carry over to flabby and internally injective objects. Specifically, we
+have:
+
+\begin{scholium}For any elementary topos~$\E$:
+\begin{enumerate}
+\item Any object can be embedded into an internally injective object.
+\item The underlying unstructured object of an internally injective module is
+internally injective.
+\item Any internally injective object is flabby.
+\item Any object can be embedded into a flabby object.
+\item The terminal object is flabby. The product of flabby objects if flabby.
+\item Let~$I$ be an internally injective object. Let~$T$ be an arbitrary
+object. Then~$[T,I]$ is a flabby object.
+\item (If~$\E$ has a natural numbers object.) Let~$I$ be an internally
+injective~$R$-module. Let~$T$ be an arbitrary~$R$-module. Then~$[T,I]_R$, the
+subobject of the internal Hom consisting only of the linear maps, is a flabby
+object.
+\item Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence
+of~$R$-modules in~$\E$. If~$M'$ and~$M''$ are flabby objects, so is~$M$.
+\end{enumerate}
+\end{scholium}
+
+\begin{proof}The analogous statements were established purely
+intuitionistically in Section~\ref{sect:flabby-sets}, and the stack semantics
+is sound with respect to intuitionistic logic.
+\end{proof}
+
+\begin{scholium}\label{scholium:exact-as-presheaves}
+Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence
+of~$R$-modules in an elementary topos~$\E$. Let~$M'$ be a flabby object.
+If~$\E$ is localic and Zorn's lemma is available, the induced sequence~$0 \to
+\Gamma(M') \to \Gamma(M) \to \Gamma(M') \to 0$ of~$\Gamma(R)$-modules is exact,
+where~$\Gamma(X) = \Hom_\E(1,X)$.\end{scholium}
+
+\begin{proof}We only have to verify exactness at~$\Gamma(M'')$, so let~$s \in
+\Gamma(M')$. Interpreting Proposition~\ref{prop:set-of-preimages-flabby}
+in~$\E$, we see that the object of preimages of~$s$ is object. Since~$\E$ is
+localic, this object is a flabby sheaf; since Zorn's lemma is available, it
+possesses a global element. Such an element is the desired preimage of~$s$
+in~$\Gamma(M)$.\end{proof}
+
+If~$\E$ is not necessarily localic or Zorn's lemma is not available, only a
+weaker substitute for Scholium~\ref{scholium:exact-as-presheaves} is available:
+Given~$s \in \Gamma(M'')$, the object of preimages of~$s$ is flabby. XXX
 

### 2018-06-19 paper-flabby-objects.tex
typo
Lots of minor improvements

@@
 \newcommand{\Set}{\mathrm{Set}}
+\newcommand{\Eff}{\mathrm{Ef{}f}}
 \renewcommand{\_}{\mathpunct{.}\,}
@@
 
+% Take care: This paragraph is referenced in prop:global-elements.
 If~$F$ is a flabby sheaf in the traditional sense, then~$F$ is obviously also
@@
 
-\section{Flabby sets}
+\section{Flabby sets}\label{sect:flabby-sets}
 
 We intend this section to be applied in the internal language of an elementary
-topos. We will therefore reason purely intuitionistically. We
+topos; we will speak about sets and maps between sets, but intend our arguments
+to be applied to objects and morphisms in toposes. We will therefore be careful
+to reason purely intuitionistically. We
 adopt the terminology of~\cite{kock:partial-maps} regarding subterminals and
@@
 In the presence of the law of excluded middle, a set is flabby if and only if
-it is inhabited. In fact, this characterization is a constructive taboo:
+it is inhabited. This characterization is a constructive taboo:
 
-\begin{prop}If any inhabited set is flabby, then the law of excluded middle
+\begin{prop}\label{prop:taboo}
+If any inhabited set is flabby, then the law of excluded middle
 holds.
@@
 
-\begin{defn}A set~$X$ is \emph{injective} if and only if, for any injection~$i
+\begin{defn}\begin{enumerate}
+\item A set~$X$ is \emph{injective} if and only if, for any injection~$i
 : A \to B$, any map~$f : A \to X$ can be extended to a map~$B \to X$.
-An~$R$-module~$M$ is \emph{injective} if and only if, for any linear
+\item An~$R$-module~$M$ is \emph{injective} if and only if, for any linear
 injection~$i : A \to B$ between~$R$-modules, any linear map~$f : A \to M$ can
-be extended to a linear map~$B \to X$.\end{defn}
+be extended to a linear map~$B \to M$.
+\end{enumerate}
+\[ \xymatrix{
+  A \ar@{^{(}->}[r]\ar[d] & B \ar@{-->}@/^/[ld] &&
+  A \ar@{^{(}->}[r]\ar[d] & B \ar@{-->}@/^/[ld] \\
+  X &&& M
+} \]
+\end{defn}
 
@@
 \to R\langle B \rangle$ between free modules is also injective, the given
-map~$f$ lifts to a linear map~$R\langle A \rangle \to M$, and an
+map~$f$ lifts to a linear map~$R\langle A \rangle \to M$, and an~$R$-linear
 extension~$R\langle B \rangle \to M$ induces an extension~$B \to M$ of~$f$.
@@
 
+A further corollary of Proposition~\ref{prop:injective-flabby} is that the
+statement ``any inhabited set is injective'' is a constructive taboo: If any
+inhabited set is injective, then any inhabited set is flabby, thus the law of
+excluded middle follows by Proposition~\ref{prop:taboo}.
+
 \begin{prop}Any singleton set is flabby. The cartesian product of flabby sets
@@
 \[ \xymatrix{
-  T' \ar[r]\ar[d] & T \ar@{-->}[ld] \\
+  T' \ar@{^{(}->}[r]\ar[d] & T \ar@{-->}@/^/[ld] \\
   I
@@
 subterminal of~$\Hom_R(T,I)$, we define~$T'$ to be the submodule
-$\{ s \in I \,|\, \text{$s = 0$ or $K$ is inhabited} \}$ and consider the
+$\{ s \in T \,|\, \text{$s = 0$ or $K$ is inhabited} \}$ and consider the
 analogous injectivity diagram, where the solid vertical map~$f : T' \to I$ is now
@@
 \[ \xymatrix{
-  i[A] \cup B' \ar@{^{(}->}[r]\ar[d]_g & B \ar@{-->}[ld] \\
+  i[A] \cup B' \ar@{^{(}->}[r]\ar[d]_g & B \ar@{-->}@/^/[ld] \\
   I
@@
 \[ \xymatrix{
-  i[A] + B' \ar@{^{(}->}[r]\ar[d]_g & B \ar@{-->}[ld] \\
+  i[A] + B' \ar@{^{(}->}[r]\ar[d]_g & B \ar@{-->}@/^/[ld] \\
   I
@@
 
-\begin{proof}For the first statement, let~$X \defeq \{ u \in M \,|\, p(u) = s
-\}$. Let~$K \subseteq X$ be a subterminal. Since~$p$ is surjective, there is a
-preimage~$u_0 \in X$. The translated set~$K - u_0 \subseteq M$ is still a
+\begin{proof}Let~$X \defeq \{ u \in M \,|\, p(u) = s
+\}$. Let~$K \subseteq X$ be a subterminal. Since~$p$ is surjective, there is an
+element~$u_0 \in X$. The translated set~$K - u_0 \subseteq M$ is still a
 subterminal, and its preimage under~$i$ is as well. Since~$M'$ is flabby, there
@@
 
+Noticeably missing here is XXX ...
+
 \section{Flabby objects}
@@
 of~$X$; this object is a certain suboject of~$\P(X) = [X,\Omega_\E]_\E$, the
-powerobject of~$X$. The subobject~$K_0$ of~$\P_{\leq1}(X) \times X$ classified by the
-evaluation morphism~$\P_{\leq1}(X) \times X \to \P(X) \times X \to \Omega_\E$
+powerobject of~$X$. The subobject~$K_0$ of~$X \times \P_{\leq1}(X)$ classified by the
+evaluation morphism~$X \times \P_{\leq1}(X) \to X \times \P(X) \to \Omega_\E$
 is the \emph{generic subterminal} of~$X$. The object~$X$ is flabby if and only
@@
 
-\begin{proof}These properties are shared by any property which can be defined
+\begin{proof}This holds for any property which can be defined
 in the stack semantics~\cite[Lemma~7.3]{shulman:stack-semantics}.
@@
 
+\begin{prop}\label{prop:global-elements}
+Let~$X$ be a flabby object in an elementary topos~$\E$. If~$\E$ is localic and
+Zorn's lemma is available, then~$X$ possesses a global element (a morphism~$1 \to X$).
+\end{prop}
+
+\begin{proof}This is a restatement of the discussion following
+Definition~\ref{defn:flabby-sheaf}.
+\end{proof}
+
 \begin{prop}\label{prop:pushforward-of-flabby-objects}
@@
 
-\begin{proof}Let~$X \in \F$ be a flabby object. Without loss of generality, it
-suffices to verify condition~$({\star})$ for the terminal stage~$A = 1 \in \E$.
-Let~$k : K \to 1$ be a monomorphism in~$\E$ and let~$x : K \to f_*(X)$ be an arbitrary morphism.
-Then~$f^*(k) : f^*(K) \to 1$ is a monomorphism in~$\F$ and~$x^t : f^*(K) \to X$
+\begin{proof}Let~$X \in \F$ be a flabby object.
+Let~$k : K \to A$ be a monomorphism in~$\E$ and let~$x : K \to f_*(X)$ be an
+arbitary morphism. Without loss of generality, we may assume that~$A$ is the
+terminal object~$1$ of~$\E$. Then~$f^*(k) : f^*(K) \to 1$ is a monomorphism
+in~$\F$ and the adjoint transpose~$x^t : f^*(K) \to X$
 is a morphism in~$\F$. Since~$X$ is flabby, there is an epimorphism~$B \to 1$
@@
 \to X$ factors over~$y$. Hence~$x$ factors over~$f_*(y) : f_*(B) \to f_*(X)$.
-By assumption, the morphism~$f_*(B) \to f_*(1)$ is an epimorphism.
+We conclude because the morphism~$f_*(B) \to f_*(1)$ is an epimorphism by
+assumption.
 \end{proof}
@@
 is for instance satisfied if~$f$ is a local geometric morphism or if~$f$ is
-induced by an open continuous injection between topological spaces. % XXX
+induced by an open continuous injection between topological spaces. XXX
 
@@
 object~$I \in \E$ are equivalent.
+% Take care if numbering changes (is referenced below).
 \begin{enumerate}
-\item $I$ is internally injective.
-\item The functor~$[\cdot, I] : \E^\op \to \E$ maps monomorphisms in $\E$
+\item[(1)] $I$ is internally injective.
+\item[(1')] For any morphism $p : A \to 1$ in $\E$, the object $p^*(I)$ has property~(1)
+as an object of $\E/A$.
+\item[(2)] The functor~$[\cdot, I] : \E^\op \to \E$ maps monomorphisms in $\E$
 to morphisms for which any global element of the target locally (after change of
 base along an epimorphism) possesses a preimage.
-\item For any morphism $p : A \to 1$ in $\E$, the object $p^*(I)$ has property~(1)
+\item[(2')] For any morphism $p : A \to 1$ in $\E$, the object $p^*(I)$ has property~(2)
 as an object of $\E/A$.
-\item For any morphism $p : A \to 1$ in $\E$, the object $p^*(I)$ has property~(2)
-as an object of $\E/A$.
-\item From the point of view of the internal language of~$\E$, the object~$I$
-is an injective set.
+\item[(3)] The statement~``$I$ is an injective set'' holds in the stack
+semantics of~$\E$.
 \end{enumerate}
@@
 \begin{proof}
-The implications ``(1)~$\Rightarrow$~(2)'', ``(3)~$\Rightarrow$~(4)'',
-``(3)~$\Rightarrow$~(1)'', and ``(4)~$\Rightarrow$~(2)'' are trivial.
+The implications (1)~$\Rightarrow$~(2), (1')~$\Rightarrow$~(2'),
+(1')~$\Rightarrow$~(1) and (2')~$\Rightarrow$~(2) are trivial.
 
-The equivalence ``(3)~$\Leftrightarrow$~(5)'' follows directly from the
+The equivalence (1')~$\Leftrightarrow$~(3) follows directly from the
 interpretation rules of the stack semantics.
 
-The implication ``(2)~$\Rightarrow$~(4)'' employs the
+The implication (2)~$\Rightarrow$~(2') employs the
 extra left adjoint $p_! : \E/A \to \E$ of $p^* : \E
@@
 
-The implication ``(4)~$\Rightarrow$~(3)'' follows by performing an extra change of
+The implication (2')~$\Rightarrow$~(1') follows by performing an extra change of
 base, since any non-global element becomes a global element after a suitable
@@
 be found in~\cite[page~54]{mines-richman-ruitenburg:constructive-algebra}.
-It seems that Harting was not aware that this construction is possible, as she
-(mistakingly) states in~\cite[page~233]{harting:remark} that ``the notions of
-injectivity and internal injectivity do not coincide'' for modules.
 
@@
 
-For the ``if'' direction, write~$\E = \Sh(X)$ and let~$I$ be an internally
-injective object. Let~$i : A \to B$ be a monomorphism in~$\Sh(X)$ and let~$f :
+For the ``if'' direction, let~$I$ be an internally
+injective object. Let~$i : A \to B$ be a monomorphism in~$\E$ and let~$f :
 A \to I$ be an arbitrary morphism. We want to show that there exists an
-extension $B \to I$ of~$f$ along~$i$. To this end, we consider the sheaf of
+extension $B \to I$ of~$f$ along~$i$. To this end, we consider the object of
 such extensions, defined by the internal expression
@@
 Global elements of~$F$ are extensions of the kind we are looking for.
-By Proposition~\ref{prop:set-of-extensions-flabby}, this sheaf is a flabby
-object and therefore by Proposition~\ref{prop:flabby-sheaves-objects} a flabby
-sheaf. Zorn's lemma implies that~$F$ has a global element.
+By Proposition~\ref{prop:set-of-extensions-flabby}, this object is flabby.
+By Proposition~\ref{prop:global-elements}, it has a global element.
 \end{proof}
+
+The analogue of Theorem~\ref{thm:injectivity-external-internal} for modules
+holds as well, if~$\E$ is assumed to have a natural numbers object. The proof
+carries over word for word, only referencing
+Proposition~\ref{prop:set-of-extensions-flabby} instead of
+Proposition~\ref{prop:set-of-linear-extensions-flabby}.
+It seems that Harting was not aware of this, even though she did show that
+injectivity of sheaves of modules over topological spaces is a local
+notion~\cite{XXX}, as she
+(mistakingly) states in~\cite[page~233]{harting:remark} that ``the notions of
+injectivity and internal injectivity do not coincide'' for modules.
+% XXX "mistakenly"? comparison with Harting's proof (Barr's lemma...)
 
@@
 
-\begin{scholium}For any elementary topos~$\E$:
+\begin{scholium}\label{scholium:properties-of-flabby-objects}
+For any elementary topos~$\E$:
 \begin{enumerate}
 \item Any object can be embedded into an internally injective object.
-\item The underlying unstructured object of an internally injective module is
-internally injective.
+\item (If~$\E$ has a natural numbers object.) The underlying unstructured
+object of an internally injective module is internally injective.
 \item Any internally injective object is flabby.
+% Take care if numbering changes (referenced below).
 \item Any object can be embedded into a flabby object.
@@
 \Gamma(M')$. Interpreting Proposition~\ref{prop:set-of-preimages-flabby}
-in~$\E$, we see that the object of preimages of~$s$ is object. Since~$\E$ is
+in~$\E$, we see that the object of preimages of~$s$ is flabby. Since~$\E$ is
 localic, this object is a flabby sheaf; since Zorn's lemma is available, it
@@
 
-\begin{rem}A generalization of the traditional notion of a flabby sheaf to
+\begin{rem}A direct generalization of the traditional notion of a flabby sheaf, as
+opposed to our reimagining in Definition~\ref{defn:flabby-sheaf}, to
 elementary toposes is the following. An object~$X$ of an elementary topos~$\E$
@@
 
+XXX
 % EXT, TOR
@@
 
-The main obstacle preventing Grothendieck's theory to be interpreted
+The main obstacle preventing Grothendieck's theory of derived functors to be interpreted
 constructively is its reliance on injective resolutions. It is known that in
@@
 suggestion. Firstly, we needed Zorn's lemma to show that flabby sheaves are
-acyclic for the global sections functor (Proposition~XXX); however, we noticed
-the substitute property in XXX which might be good enough for further
-developments. The more serious problem is that it is an open question whether
+acyclic for the global sections functor (Proposition~XXX). This problem might
+be mitigated by relying on the substitute property XXX.
+The more serious problem is that it is an open question whether
 one can show, purely intuitionistically, that any sheaf of modules embeds into
@@
 \item There is a purely intuitionistic proof that any sheaf of sets embeds into
-a flabby sheaf of sets (Corollary~\ref{cor:enough-flabby-sets}).
+a flabby sheaf of sets
+(Scholium~\ref{scholium:properties-of-flabby-objects}(4)).
+
 \item The existence of enough flabby modules, and even the existence of enough
@@
 Grothendieck topos has enough injective (and therefore flabby) modules.
+
 \item There is a way of embedding any module into a flabby module if one is
@@
 r \underline{x}$, $\underline{0} \sim 0$ and such that~$\varepsilon_{\{t\}} \sim
-t$ for all terms~$t$. Then the quotient~$T/({\sim})$ is a flabby~$R$-module
-into which~$M$ embeds.
+t$ for all terms~$t$. Then the quotient~$T/{\sim}$ is a flabby~$R$-module
+into which~$M$ embeds. However, it is not clear that~$T/{\sim}$ is a set.
+
+\item In the ef{}fective topos, the statement ``if~$X$ is flabby, then any map~$X
+\to X$ has a fixed point'' holds for ef{}fective sets~$X$. In particular, the
+only module in the ef{}fective topos which is flabby and an ef{}fective set is the
+trivial module (XXX proof). On the other hand, assuming the law of excluded middle, the
+statement ``any module with~$\neg\neg$-stable equality can be embedded into a
+flabby module'' holds in the ef{}fective topos. This is because for such a
+module~$M$, the canonical morphism~$M \to \Delta(\Gamma(M))$ is a monomorphism,
+the set~$\Gamma(M)$ is flabby by virtue of being inhabited, and~$\Delta$
+preserves flabby objects by Proposition~\ref{prop:pushforward-of-flabby-objects}.
+Here~$(\Delta \dashv \Gamma) : \Set \to \Eff$ is the inclusion of the
+double-negation sheaves.
 \end{enumerate}

### 2018-06-19 paper-generic-freeness.tex
Fixed minor typos and added Giuseppe Rosolini to the acknowledgments

@@
 version in a paper of his~\cite{brandenburg:schur}. We are grateful to Thierry
-Coquand and Peter Schuster for valuable advice and to Marc
+Coquand and Peter Schuster for valuable advice, to Giuseppe Rosolini for
+comments regarding the presentation of the paper, and to Marc
 Nieper-Wißkirchen for carefully guiding my PhD studies~\cite{blechschmidt:phd} at the University of
@@
 have more informative, direct proofs available which illuminate the
-statement more clearly; similar as to how bijective proofs are preferred over
+proven claims more clearly; similar as to how bijective proofs are preferred over
 calculational inductive ones in combinatorics.
@@
 
-\begin{proof}The proof of Theorem~\ref{thm:algebraic} can be easily adapted to
+\begin{proof}Our proof of Theorem~\ref{thm:algebraic} can be easily adapted to
 this more general setting. Where that proof concludes~$a_J = 0$ for ring
@@
 conclude~$a_J = 0$ for local sections~$a_J$ of~$\O_X$ by considering the
-situation over the restriction to~$D(a_J)$. It is then the assumption on~$X$
-instead of the reducedness assumption on~$A$ which allows us to conclude~$a_J = 0$.
+situation over the restriction to~$D(a_J)$. Whereas before this type of argument was
+powered by the reducedness assumption on~$A$, it is not supported by the
+assumption on~$X$.
 We omit further details.

### 2018-06-20 paper-flabby-objects.tex
Flabby objects in the effective topos

@@
 \usepackage[utf8]{inputenc}
-\usepackage{amsthm,mathtools,tikz,float,caption}
+\usepackage{amsthm,mathtools,tikz,float,caption,stmaryrd}
 \usetikzlibrary{patterns,matrix,decorations.pathreplacing}
@@
 \newcommand{\B}{\mathcal{B}}
+\newcommand{\NN}{\mathbb{N}}
 \newcommand{\ZZ}{\mathbb{Z}}
@@
 
+\section{Flabby objects in the effective topos}
+
+The notion of flabby objects originates from the notion of flabby shaves and is
+therefore closely connected to Grothendieck toposes. It is therefore instructive
+to study flabby objects in elementary toposes which are not Grothendieck
+toposes, away from their original conceptual home. We begin this study with
+establishing the following observations on flabby objects in the effective
+topos.
+
+\begin{prop}\label{prop:flabby-effective-sets}
+Let~$X$ be a flabby object in the effective topos. Let~$f : X \to X$
+be a morphism. If~$X$ is effective, the statement ``$f$ has a fixed point''
+holds in the effective topos.
+\end{prop}
+
+\begin{prop}\label{prop:semienough-flabby-modules}
+Assuming the law of excluded middle, any~$\neg\neg$-separated
+module in the effective topos can be embedded into a flabby module.
+\end{prop}
+
+We follow the terminology of~\cite{hyland:effective-topos}(XXX). The intuitive
+reason for why Proposition~\ref{prop:flabby-effective-sets} holds is the
+following. Let~$X$ be a flabby object of the effective topos. Then there is a
+procedure which computes for any subterminal~$K \subseteq X$ an element~$x_K$
+such that~$K \subseteq \{ x_K \}$. This element might not depend extensionally
+on~$K$, but this fine point is not important for the further discussion. Let
+now~$f : X \to X$ be a morphism. We construct the self-referential subset~$K
+\defeq \{ f(x_K) \}$; the formal proof below will indicate how this can be
+done. Then~$K \subseteq \{ x_K \}$, so~$f(x_K) = x_K$.
+
+A corollary of Proposition~\ref{prop:flabby-effective-sets} is that the trivial
+module is the only flabby module in the effective topos whose underlying
+unstructured object is an effective set: Given such a flabby module~$M$, let~$v
+\in M$ be an arbitrary element. Then the morphism~$x \mapsto v + x$ has a fixed
+point; thus~$v + x = x$ for some element~$x$, and hence~$v = 0$.
+
+It is the self-referentiality which makes the proof of
+Proposition~\ref{prop:flabby-effective-sets} work, but the blame for paucity
+of flabby objects in the effective topos is to put on the realizers for
+statements of the form~``$K = K$'', where~$({=})$ is the nonstandard equality
+predicate of the powerobject~$\P(X)$. A procedure witnessing flabbiness has to
+compute a reflexivity realizer for a suitable element~$x_K$ from a reflexivity
+realizer for a given element~$K$. However, such realizers are not very
+informative. Metaphorically speaking, a procedure witnessing flabbiness has to
+conjure elements out of thin air.
+
+This problem does not manifest with objects~$X$ which are not effective sets.
+Reflexivity realizers for these objects may not be very informative themselves;
+a procedure witnessing flabbiness therefore only has to turn one kind of
+non-informative realizers into another kind. The flabby modules featuring in
+the proof of Proposition~\ref{prop:semienough-flabby-modules} will accordingly
+be not effective sets.
+
+\begin{proof}[Proof of Proposition~\ref{prop:flabby-effective-sets}]
+For any Turing machine~$e$, let~$v_e$ be the set
+\begin{multline*}
+  v_e = \{ m \in \NN \,|\,
+  \text{there is an element~$x_0 \in |X|$ such that} \\
+  \text{$e$ terminates with an element of~$\llbracket x_0 = x_0 \rrbracket$ and
+  $m \in \llbracket x = x_0 \rrbracket$} \}
+\end{multline*}
+and let~$K_e \in \Sigma^{|X| \times \Sigma}$ be the nonstandard predicate given by
+\[ K_e(x,u) = \llbracket (x = x) \wedge (u \leftrightarrow v_e) \rrbracket. \]
+One can explicitly construct a realizer~$a_e$ of the statement~``$K_e = K_e$'',
+where~$({=})$ is the nonstandard equality predicate of the object~$\P_{\leq1}(X)$
+of subterminals of~$X$. This is where the assumption that~$X$ is effective is
+important; without it, we could only verify~``$K_e = K_e$'' where~$({=})$ is
+the nonstandard equality predicate of the full powerobject~$\P(X)$.
+
+Since~$X$ is flabby, there is a realizer~$r$ for the statement~``$\forall K \in
+\P_{\leq1}(X)\_ \exists x \in X\_ \forall y \in X\_ (y \in K \Rightarrow y =
+x)$''. Let~$s$ be a realizer for the statement~``$\forall x \in X\_ \exists y
+\in Y\_ y = f(x)$''. Let~$e$ be the particular Turing machine which proceeds as
+follows:
+\begin{enumerate}
+\item[1.] Simulate~$r$ on input~$a_e$ in order to obtain a realizer~$b \in
+\llbracket x = x \rrbracket$ for some~$x \in |X|$.
+\item[2.] Simulate~$s$ on input~$b$ in order to obtain a realizer~$c \in \llbracket
+f(x) = f(x) \rrbracket$.
+\item[3.] Output~$c$.
+\end{enumerate}
+The description of the machine~$e$ makes use of the number~$e$ coding it;
+the recursion theorem yields a general reason why this self-referentiality is
+possible. Here we can even do without this theorem, since a close inspection of
+the construction of~$a_e$ shows that~$a_e$ is actually independent of~$e$. This
+should not come as a surprise, as reflexivity realizers of~$\P(X)$
+and~$\P_{\leq1}(X)$ are known to be not very informative.
+
+Passing~$a_e$ to~$r$ yields a reflexivity realizer of some element~$x_{K_e} \in
+|X|$. Therefore the Turing machine~$e$ does terminate, with a reflexivity
+realizer for~$f(x_{K_e})$. Thus the statement~``$f(x_{K_e}) \in K$'' is
+realized; hence~``$f(x_{K_e}) = x_{K_e}$'' is as well.
+\end{proof}
+
+\begin{proof}[Proof of Proposition~\ref{prop:semienough-flabby-modules}]
+Let~$(\Delta \dashv \Gamma) : \Set \to \Eff$ be the inclusion of the
+double-negation sheaves. For a~$\neg\neg$-separated module~$M$ in the effective
+topos, the canonical morphism~$M \to \Delta(\Gamma(M))$ is a monomorphism; the
+set~$\Gamma(M)$ is flabby by virtue of being inhabited; and~$\Delta$ preserves
+flabby objects by Proposition~\ref{prop:pushforward-of-flabby-objects}.
+\end{proof}
+
 \section{Conclusion}
@@
 
-\item In the ef{}fective topos, the statement ``if~$X$ is flabby, then any map~$X
-\to X$ has a fixed point'' holds for ef{}fective sets~$X$. In particular, the
-only module in the ef{}fective topos which is flabby and an ef{}fective set is the
-trivial module (XXX proof). On the other hand, assuming the law of excluded middle, the
-statement ``any module with~$\neg\neg$-stable equality can be embedded into a
-flabby module'' holds in the ef{}fective topos. This is because for such a
-module~$M$, the canonical morphism~$M \to \Delta(\Gamma(M))$ is a monomorphism,
-the set~$\Gamma(M)$ is flabby by virtue of being inhabited, and~$\Delta$
-preserves flabby objects by Proposition~\ref{prop:pushforward-of-flabby-objects}.
-Here~$(\Delta \dashv \Gamma) : \Set \to \Eff$ is the inclusion of the
-double-negation sheaves.
+\item There appears to be some tension regarding the effective topos:
+Proposition~\ref{prop:semienough-flabby-modules} shows that at
+least~$\neg\neg$-separated modules in the effective topos always embed into
+flabby modules, assuming the law of excluded middle, while
+Proposition~\ref{prop:flabby-effective-sets} shows that no nontrivial effective
+module is flabby.
 \end{enumerate}

### 2018-06-28 slides-como2018.tex
Slides for Como (almost finished)

\documentclass[11pt,utf8,notheorems,compress,t]{beamer}
\usepackage{etex}

\usepackage[english]{babel}

\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{ifthen}
\usepackage{soul}\setul{0.3ex}{}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing}
\hypersetup{colorlinks=true}
\usepackage{multimedia}
\newcommand{\video}[2]{\movie[width=#2,height=#2,autostart,loop,poster]{}{#1}}
\hypersetup{colorlinks=false}

\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}
\DeclareSymbolFont{extraup}{U}{zavm}{m}{n}
\DeclareMathSymbol{\varheart}{\mathalpha}{extraup}{86}

\graphicspath{{images/}}

\usepackage[protrusion=true,expansion=true]{microtype}

\setlength\parskip{\medskipamount}
\setlength\parindent{0pt}

\title{Exploring the internal language of toposes}
\author{Ingo Blechschmidt}
\date{June 22nd, 2018 XXX}

\useinnertheme[shadow=true]{rounded}
\useoutertheme[subsection=false]{miniframes}
\setbeamerfont{block title}{size={}}

\useinnertheme{rectangles}

\usecolortheme{orchid}
\usecolortheme{seahorse}
\definecolor{mypurple}{RGB}{150,0,255}
\setbeamercolor{structure}{fg=mypurple}
\definecolor{myred}{RGB}{150,0,0}
\setbeamercolor*{title}{bg=myred,fg=white}
\setbeamercolor*{titlelike}{bg=myred,fg=white}
\setbeamercolor{frame}{bg=black}

\usefonttheme{serif}
\usepackage[T1]{fontenc}
\usepackage{libertine}

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\renewcommand{\AA}{\mathbb{A}}
\renewcommand{\C}{\mathcal{C}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ull}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
\newcommand{\Ll}{\text{iff}}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}{\vdash_{\!\!\!\vec x}}

\setbeamertemplate{blocks}[rounded][shadow=false]

% Adapted from https://latex.org/forum/viewtopic.php?t=2251 (Stefan Kottwitz)
\newenvironment<>{hilblock}{
  \begin{center}
    \begin{minipage}{9.05cm}
      \setlength{\textwidth}{9.05cm}
      \begin{actionenv}#1
        \def\insertblocktitle{}
        \par
        \usebeamertemplate{block begin}}{
        \par
        \usebeamertemplate{block end}
      \end{actionenv}
    \end{minipage}
  \end{center}}

\newcommand{\bignumber}[1]{
  \renewcommand{\insertenumlabel}{#1}\scalebox{1.5}{\usebeamertemplate{enumerate item}}
}
\newcommand{\bigheart}[1]{\scalebox{1.5}{\hil{$\varheart$}}}

\newenvironment{changemargin}[2]{%
  \begin{list}{}{%
    \setlength{\topsep}{0pt}%
    \setlength{\leftmargin}{#1}%
    \setlength{\rightmargin}{#2}%
    \setlength{\listparindent}{\parindent}%
    \setlength{\itemindent}{\parindent}%
    \setlength{\parsep}{\parskip}%
  }%
  \item[]}{\end{list}}

\newcommand{\hcancel}[5]{%
  \tikz[baseline=(tocancel.base)]{
    \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
    \draw[red, line width=0.4mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
  }%
}

\tikzset{
  invisible/.style={opacity=0,text opacity=0},
  visible on/.style={alt={#1{}{invisible}}},
  alt/.code args={<#1>#2#3}{%
    \alt<#1>{\pgfkeysalso{#2}}{\pgfkeysalso{#3}}}
}

\newcommand{\pointthis}[3]{%
  \tikz[remember picture,baseline]{
    \node[anchor=base,inner sep=0,outer sep=0] (#2) {#2};
    \node[visible on=#1,overlay,rectangle callout,rounded corners,callout relative pointer={(0.3cm,0.5cm)},fill=blue!20] at ($(#2.north)+(-0.1cm,-1.1cm)$) {#3};
  }%
}

% Adapted from https://latex.org/forum/viewtopic.php?t=2251 (Stefan Kottwitz)
\newenvironment<>{varblock}[2]{
  \begin{center}
    \begin{minipage}{#1}
      %\setlength{\textwidth}{#1}
      \begin{actionenv}#3
  \def\insertblocktitle{\centering #2}
  \par
  \usebeamertemplate{block begin}}{
        \par
        \usebeamertemplate{block end}
      \end{actionenv}
    \end{minipage}
  \end{center}}

\setbeamertemplate{frametitle}{%
  \vskip0.7em%
  \leavevmode%
  \begin{beamercolorbox}[dp=1ex,center]{}%
      \usebeamercolor[fg]{item}{\textbf{{\Large \insertframetitle}}}
  \end{beamercolorbox}%
}

\setbeamertemplate{navigation symbols}{}

\newcounter{framenumberpreappendix}
\newcommand{\backupstart}{
  \setcounter{framenumberpreappendix}{\value{framenumber}}
}
\newcommand{\backupend}{
  \addtocounter{framenumberpreappendix}{-\value{framenumber}}
  \addtocounter{framenumber}{\value{framenumberpreappendix}} 
}

\setbeamertemplate{headline}{%
  \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex]{}%
    \insertnavigation{\paperwidth}%
  \end{beamercolorbox}%
  \vskip0pt%
}

\setbeamertemplate{footline}{%
  \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex,dp=1ex,right,rightskip=1mm,leftskip=1mm]{}%
    % \inserttitle
    \hfill
    \insertframenumber\,/\,\inserttotalframenumber
  \end{beamercolorbox}%
  \vskip0pt%
}

\newcommand{\hil}[1]{{\usebeamercolor[fg]{item}{\textbf{#1}}}}

\newcommand{\bad}[1]{\textcolor{red!90}{\textnormal{#1}}}

\begin{document}

\addtocounter{framenumber}{-1}

\begin{frame}[c]
  \centering
  \medskip
  \hspace*{-3.9em}%
%  \includegraphics[height=0.3\textheight]{vichy-2}%
%  \includegraphics[height=0.3\textheight]{vichy-1}%
%  \includegraphics[height=0.3\textheight]{vichy-4}%
%  \includegraphics[height=0.3\textheight]{vichy-5}%

  \hil{Exploring the internal language of toposes} \\

  \emph{-- an invitation --}

  \bigskip

  \begin{columns}
    \small
    \begin{column}{0.35\textwidth}
      \centering
      \bignumber{1}

      \hil{Alternate universes}
      \medskip

    \end{column}
    \hspace*{0.5em}
    \begin{column}{0.25\textwidth}
      \centering
      \bignumber{2}

      \hil{Applications}
      \medskip

%      \includegraphics[height=5.5em]{calabi-yau}
    \end{column}
    \begin{column}{0.4\textwidth}
      \centering
      \bignumber{3}

      \hil{Vision for the future}
      \medskip

%      \includegraphics[height=5.5em]{question-mark}
    \end{column}
  \end{columns}

  \scriptsize
  Ingo Blechschmidt (MPI Leipzig) \\
  Toposes in Como \\
  June 29th, 20s18
  \par
\end{frame}

\begin{frame}{Motivating testcases}
  Let $A$ be a ring (commutative, with unit, $1 = 0$ allowed). \\
  Assume that $A$ is reduced: If $x^n = 0$, then $x = 0$.

  \only<1-3>{\begin{columns}[t]
  \begin{column}[t]{0.50\textwidth}
    \centering

    \scalebox{0.8}{$\begin{pmatrix}
      \cdot & \cdot \\
      \cdot & \cdot \\
      \cdot & \cdot
      \end{pmatrix}$}
    \vspace*{-1em}

    \begin{varblock}{\textwidth}{A baby application}
      \justifying
      Let~$M$ be a surjective matrix over~$A$ with more rows than columns.
      Then~$1 = 0$ in~$A$.
    \end{varblock}

    \justifying
    \only<2>{\textbf{Proof.} \bad{Assume not.} Then there is~a \bad{maximal
    ideal} $\mmm$. The matrix is surjective over the field~$A/\mmm$. This is a
    contradiction to basic linear algebra.}

    \only<3>{\textbf{Proof.} \bad{Assume not.} Then there is~a \bad{prime
    ideal} $\ppp$. The matrix is surjective over the
    field~$\mathrm{Quot}(A/\ppp)$. This is a contradiction to basic linear
    algebra.}
  \end{column}

  \begin{column}[t]{0.5\textwidth}
    \centering

    \scalebox{0.8}{$\begin{pmatrix}
      \cdot & \cdot & \cdot & \cdot \\
      \cdot & \cdot & \cdot & \cdot \\
      \cdot & \cdot & \cdot & \cdot
    \end{pmatrix}$}
    \vspace*{-1em}

    \begin{varblock}{\textwidth}{A child application}
      \justifying
      Let~$M$ be an injective matrix over~$A$ with more columns than rows.
      Then~$1 = 0$ in~$A$.
    \end{varblock}

    \justifying
    \visible<2-3>{\textbf{Proof.} \bad{Assume not.} Then there is a \bad{minimal
    prime ideal} $\ppp$. The matrix is injective over the field~$A_\ppp = A[(A
    \setminus \ppp)^{-1}]$. This is a contradiction to basic linear algebra.}
    \end{column}
  \end{columns}}

  \pause
  \pause
  \pause

  \centering

  % XXX: \includegraphics[height=3em]{generic-freeness}
  \vspace*{-1em}

  \begin{varblock}{0.9\textwidth}{Generic freeness$\phantom{p}$}
    \justifying
    Let~$B$ be an~$A$-algebra of finite type ($\cong A[X_1,\ldots,X_n]/\aaa$). \\
    Let~$M$ be a finitely generated~$B$-module ($\cong B^m/U$).

    If~$f = 0$ is the only element of~$A$ such that
    \begin{enumerate}
      \item $B[f^{-1}]$ and $M[f^{-1}]$ are free modules over $A[f^{-1}]$,
      \item $A[f^{-1}] \to B[f^{-1}]$ is of finite presentation and
      \item $M[f^{-1}]$ is finitely presented as a module over~$B[f^{-1}]$,
    \end{enumerate}
    then $1 = 0$ in~$A$.
  \end{varblock}
  \vspace*{-0.5em}

  \pause\raggedright
  \textbf{Proof.} See [Stacks Project, Tag 051Q].
\end{frame}

\begin{frame}{The internal language of a topos}
  For any topos~$\E$ and any formula~$\varphi$, we define the meaning of
  \vspace*{-0.5em}
  \[
    \text{``$\E \models \varphi$''} \quad
    \text{(``$\varphi$ holds in the internal universe of~$\E$'')}
  \]

  \vspace*{-1.0em}
  using (Shulman's extension of) the \hil{Kripke--Joyal semantics}.

  \vspace*{-1em}
  \begin{columns}
    \def\insertblocktitle{}
    \begin{column}{0.3\textwidth}\usebeamertemplate{block begin}
      \centering
      $\Set \models \varphi$ \\
      ``$\varphi$ holds in the \\ usual sense.''
      \usebeamertemplate{block end}\end{column}

    \begin{column}{0.3\textwidth}\usebeamertemplate{block begin}
      \centering
      $\Sh(X) \models \varphi$ \\
      ``$\varphi$ holds continuously.''
      \usebeamertemplate{block end}\end{column}

    \begin{column}{0.3\textwidth}\usebeamertemplate{block begin}
      \centering
      $\Eff \models \varphi$ \\
      ``$\varphi$ holds \\ computably.''
      \usebeamertemplate{block end}\end{column}
  \end{columns}
  \medskip

  Any topos supports \hil{mathematical reasoning}:

  \vspace*{-1.5em}
  \begin{hilblock}
    If~$\E \models \varphi$ and if~$\varphi$ entails~$\psi$
    \pointthis{<2>}{intuitionistically}{%
      no $\varphi \vee \neg\varphi$,\ \
      no $\neg\neg\varphi \Rightarrow \varphi$,\ \
      no axiom of choice},
    then~$\E \models \psi$.
  \end{hilblock}
\end{frame}

\begin{frame}{The internal language of~$\boldsymbol{\Sh(X)}$}
  \small
  Let~$X$ be a topological space. We recursively define
  \[ U \models \varphi \quad \text{(``$\varphi$ holds on~$U$'')} \]
  \mbox{for open subsets~$U \subseteq X$ and formulas~$\varphi$.
  Write~``$\Sh(X) \models \varphi$'' to mean~$X \models \varphi$.}
  \[ \renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\ }c@{\ }l@{}}
  U \models f = g \? F &\Ll& f|_U = g|_U \in F(U) \\
  U \models \varphi \wedge \psi &\Ll&
  \text{$U \models \varphi$ and $U \models \psi$} \\
  U \models \varphi \vee \psi &\Ll&
  \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $U = \bigcup_i U_i$ s.\,th.} \\
  && \quad\quad\text{for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
  U \models \varphi \Rightarrow \psi &\Ll&
  \text{for all open~$V \subseteq U$: } 
  \text{$V \models \varphi$ implies $V \models \psi$} \\
  U \models \forall f \? F\_ \varphi(f) &\Ll&
  \text{for all open $V \subseteq U$ and sections~$f_0 \in F(V)$: $V \models \varphi(f_0)$} \\
  U \models \forall F\_ \varphi(F) &\Ll&
  \text{for all open $V \subseteq U$ and sheaves~$F_0$ over~$V$: $V \models \varphi(F_0)$} \\
  U \models \exists f \? F\_ \varphi(f) &\Ll&
  \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
  && \quad\quad \text{there exists~$f_0 \in F(U_i)$ s.\,th. $U_i \models \varphi(f_0)$} \\
  U \models \exists F\_ \varphi(F) &\Ll&
  \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
  && \quad\quad \text{there exists a sheaf~$F_0$ on $U_i$ s.\,th. $U_i \models \varphi(F_0)$}
  \end{array} \]
\end{frame}

\begin{frame}{Internalizing parameter-dependence}
  \justifying
  Let~$X$ be a space. A continuous family~$(f_x)_{x \in X}$ of continuous
  functions (that is, a continuous function~$f : X \times \RR \to \RR$;
  $f_x(a) = f(x,a)$) induces an endomorphism of the sheaf~$\C$ of continuous
  functions:
  \[
    \bar f : \C \longrightarrow \C,\
    \text{on $U$:}\ s \longmapsto (x \mapsto f_x(s(x))).
  \]
  \begin{itemize}
    \item $\Sh(X) \models \speak{The set $\C$ is the set of (Dedekind) reals}$.
    \item $\Sh(X) \models \speak{The function $\bar f : \RR \to \RR$ is continuous}$.
    \item Iff $f_x(-1) < 0$ for all $x$, then $\Sh(X) \models \bar f(-1) < 0$.
    \item Iff $f_x(+1) > 0$ for all $x$, then $\Sh(X) \models \bar f(+1) > 0$.
    \item Iff all $f_x$ are increasing, then $\Sh(X) \models \speak{$\bar f$ is increasing}$.
    \item \ \\[-1.2em]\mbox{Iff there is an open cover~$X = \bigcup_i U_i$ such
    that for each~$i$ there is a} \mbox{continuous function~$s : U_i \to \RR$
    with~$f_x(s(x)) = 0$ for all~$x \in U_i$,} then $\Sh(X) \models \exists s
    \? \RR\_ \bar f(s) = 0$.
  \end{itemize}
\end{frame}

\begin{frame}{The little Zariski topos}
  Let~$A$ be a ring. Its \hil{little Zariski topos} is equivalently
  \begin{enumerate}
    \item the classifying topos of \hil{local localizations} of $A$,
    \item the classifying topos of \hil{prime filters} of $A$,
    \item the locale given by the frame of \hil{radical ideals} of $A$,
    \item the topos of sheaves over the poset $A$ with $f \preceq g$ iff $f \in \sqrt{(g)}$ and with $(f_i \to f)_i$ deemed covering iff $f \in \sqrt{(f_i)_i}$ or
    \item the topos of sheaves over $\Spec(A)$.
  \end{enumerate}
  Its associated topological space of points is the \hil{classical spectrum}
  \[ \{ \fff \subseteq A \,|\, \text{$\fff$ prime filter} \} + \text{Zariski topology}. \]
  It has \hil{enough points} if the Boolean Prime Ideal Theorem holds.

  \footnotesize
  Prime ideal:\, $0 \in \ppp$;\, $x \in \ppp \wedge y \in \ppp \Rightarrow x+y \in \ppp$;\, $1 \not\in \ppp$;\, $xy \in \ppp \Leftrightarrow x \in \ppp \vee y \in \ppp$

  Prime filter:\, $0 \not\in \fff$;\,\,
  $x+y \in \fff \Rightarrow x \in \fff \vee y \in \fff$;
  \hspace*{0.5pt}\,\,
  $1 \in \fff$;\,
  $xy \in \fff \Leftrightarrow x \in \fff \wedge y \in \fff$
\end{frame}

\begin{frame}{First steps in the little Zariski topos}
  Let~$A$ be a ring. Let~$\fff_0$ be the \hil{generic prime filter} of~$A$; it
  is a subobject of the constant sheaf~$\ull{A}$ of the little Zariski topos.

  \begin{itemize}
    \item The ring $A^\sim \defeq \ull{A}[\fff_0^{-1}]$ is the generic local
    localization of $A$.
    \item Given an~$A$-module~$M$, we have the~$A^\sim$-module $M^\sim \defeq
    \ull{M}[\fff_0^{-1}]$.
  \end{itemize}

  \only<1>{\begin{center}
    \includegraphics[width=0.9\textwidth]{hartshorne-tilde-construction} \\
    \footnotesize
    Robin Hartshorne. Algebraic Geometry. 1977.
  \end{center}}

  \pause

  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{varblock}{\textwidth}{}
        \justifying
        Assuming the Boolean prime ideal theorem, a geometric sequent
        ``$\forall \ldots \forall\_ (\cdots \Longrightarrow \cdots\!\,)$'',
        where the two subformulas may not contain~``$\Rightarrow$'' and~``$\forall$'',
        holds for~$M^\sim$ iff it holds for all stalks~$M_\ppp$.
      \end{varblock}
    \end{column}

    \begin{column}{0.5\textwidth}
      \begin{varblock}{\textwidth}{}
        $M^\sim$ inherits any property of~$M$ which is \hil{localization-stable}.
      \end{varblock}

      \bigskip
      If $A$ is reduced ($x^n = 0 \Rightarrow x = 0$):

      \vspace*{-1em}
      \setbeamercolor{block body}{bg=red!30}
      \setbeamercolor{structure}{fg=purple}
      \begin{varblock}{\textwidth}{}
        $A^\sim$ is a \hil{field}
        (nonunits are zero).

        $A^\sim$ has \hil{$\boldsymbol{\neg\neg}$-stable equality}.

        \mbox{$A^\sim$ is \hil{anonymously Noetherian}.}\\[-1.2em]
      \end{varblock}
    \end{column}
  \end{columns}

  \visible<3>{\begin{tikzpicture}[overlay]
    \draw[fill=white, draw=white, opacity=0.95] (-1,0) rectangle (\paperwidth,7.4);
    \node[anchor=south west,inner sep=0] (image) at (0,0.8) {\vbox{
      \centering
      \includegraphics[width=0.9\textwidth]{tierney-on-the-spectrum-of-a-ringed-topos} \\
      \footnotesize
      Miles Tierney. On the spectrum of a ringed topos. 1976.
    }};
  \end{tikzpicture}}
\end{frame}

\begin{frame}{Revisiting the testcases}
  Let $A$ be a reduced ring.

  \only<1>{\begin{columns}[t]
    \begin{column}[t]{0.50\textwidth}
      \centering

      \scalebox{0.8}{$\begin{pmatrix}
        \cdot & \cdot \\
        \cdot & \cdot \\
        \cdot & \cdot
        \end{pmatrix}$}
      \vspace*{-1em}

      \begin{varblock}{\textwidth}{A baby application}
        \justifying
        Let~$M$ be a surjective matrix over~$A$ with more rows than columns.
        Then~$1 = 0$ in~$A$.
      \end{varblock}

      \justifying
      \textbf{Proof.} The matrix is surjective over the field~$A^\sim$. This is
      a contradiction to basic linear algebra. Hence $\Spec(A) \models \bot$,
      thus $1 = 0$ in $A$.
    \end{column}
    \begin{column}[t]{0.5\textwidth}
      \centering

      \scalebox{0.8}{$\begin{pmatrix}
        \cdot & \cdot & \cdot & \cdot \\
        \cdot & \cdot & \cdot & \cdot \\
        \cdot & \cdot & \cdot & \cdot
        \end{pmatrix}$}
      \vspace*{-1em}

      \begin{varblock}{\textwidth}{A child application}
        \justifying
        Let~$M$ be an injective matrix over~$A$ with more columns than rows.
        Then~$1 = 0$ in~$A$.
      \end{varblock}

      \justifying
      \textbf{Proof.} The matrix is injective over the field~$A^\sim$. This is
      a contradiction to basic linear algebra. Hence $\Spec(A) \models \bot$,
      thus $1 = 0$ in $A$.
    \end{column}
  \end{columns}}

  \pause

  \centering

  % XXX: \includegraphics[height=3em]{generic-freeness}
  \vspace*{-1em}

  \begin{varblock}{\textwidth}{Generic freeness$\phantom{p}$}
    \justifying
    Let~$B$ be an~$A$-algebra of finite type ($\cong A[X_1,\ldots,X_n]/\aaa$). \\
    Let~$M$ be a finitely generated~$B$-module ($\cong B^m/U$).

    If~$f = 0$ is the only element of~$A$ such that
    \begin{enumerate}
      \item $B[f^{-1}]$ and $M[f^{-1}]$ are free modules over $A[f^{-1}]$, \\[-1.6em]
      \item $A[f^{-1}] \to B[f^{-1}]$ is of finite presentation and \\[-1.6em]
      \item $M[f^{-1}]$ is finitely presented as a module over~$B[f^{-1}]$,
    \end{enumerate}
    then $1 = 0$ in~$A$.
  \end{varblock}
  \vspace*{-0.5em}

  \pause\raggedright\bigskip
  \textbf{Proof.} In the little Zariski topos it's \hil{not not} the case that % \\[-1.4em]
  \begin{enumerate}
    \item $B^\sim$ and $M^\sim$ are free modules over $A^\sim$,
    \item $A^\sim \to B^\sim$ is of finite presentation and
    \item $M^\sim$ is finitely presented as a module over $B^\sim$
  \end{enumerate}
  by basic linear algebra over the field $A^\sim$. The claim is precisely the
  external translation of this fact.
\end{frame}

% Großer Zariskitopos
% Zusammenhang zum kleinen Topos
% Synthetische Quasikohärenz
% Klassifizierende Topoi

\begin{frame}{Understanding algebraic geometry}
  \vspace*{-1.5em}
  \begin{varblock}{0.9\textwidth}{}
    \justifying
    Understand \hil{notions of algebraic geometry} over a scheme~$X$ as
    \hil{notions of algebra} internal to~$\Sh(X)$.
  \end{varblock}

  \only<1>{
    \centering
    \rotatebox{90}{\tiny\scalebox{0.8}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
    \includegraphics[width=0.7\textwidth]{external-internal-small}
    \par
  }

  \pause

  \small\centering
  \scalebox{0.83}{\begin{tabular}{ll}
    \toprule
    externally & internally to $\Sh(X)$ \\
    \midrule
    sheaf of sets & set \\
    %sheaf of rings & ring \\
    sheaf of modules & module \\
    sheaf of finite type & finitely generated module \\
    % finite locally free sheaf & finite free module \\
    % coherent sheaf & coherent module \\
    tensor product of sheaves & tensor product of modules \\
    % sheaf of Kähler differentials & module of Kähler differentials \\
    sheaf of rational functions & total quotient ring of~$\O_X$ \\
    dimension of $X$ & Krull dimension of~$\O_X$ \\
    spectrum of a sheaf of~$\O_X$-algebras & ordinary spectrum [with a twist] \\
    big Zariski topos of $X$ & big Zariski topos of the ring $\O_X$ [with a twist] \\
    higher direct images & sheaf cohomology \\
    \bottomrule
  \end{tabular}}

  \begin{columns}[c]
    \begin{column}{0.43\textwidth}
      \begin{exampleblock}{}
        \justifying
        Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence of
        modules. If~$M'$ and~$M''$ are finitely generated, so is~$M$.
      \end{exampleblock}
    \end{column}

    \scalebox{3}{$\Rightarrow$}
    \hspace*{0.5em}

    \begin{column}{0.47\textwidth}
      \begin{exampleblock}{}
        \justifying
        Let $0 \to \F' \to \F \to \F'' \to 0$ be a short exact sequence
        of sheaves of~$\O_X$-modules. If~$\F'$ and~$\F''$ are of finite type,
        so is~$\F$.
      \end{exampleblock}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Synthetic algebraic geometry}
  Usual approach to algebraic geometry: \hil{layer schemes above ordinary set theory}
  using either
  \begin{itemize}
    \item locally ringed spaces
    \small
    \begin{multline*}
      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$} + {} \\
      \text{Zariski topology} + \text{structure sheaf}
    \end{multline*}
    \normalsize
    \item or Grothendieck's functor-of-points account, where a scheme is a
    functor~$\mathrm{Ring} \to \mathrm{Set}$.
    \small\[ A \longmapsto \{ (x,y,z) \in A^3 \,|\, x^n+y^n-z^n=0 \} \]
  \end{itemize}
  \bigskip

  \hil{Synthetic approach:} model schemes \hil{directly as sets} in a certain
  nonclassical set theory, the internal universe of the \mbox{\hil{big Zariski
  topos}} of a base scheme.
  \small
  \[ \{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]
\end{frame}

\begin{frame}{The big Zariski topos}
  The \hil{big Zariski topos} $\Zar(S)$ of a scheme~$S$ is equivalently
  \begin{enumerate}
    \item the topos of sheaves over~$(\Aff/S)_{\mathrm{lofp}}$,
    \item the classifying topos of local rings over~$S$ or
    \item the classifying~$\Sh(S)$-topos of local~$\O_S$-algebras which are local
    over~$\O_S$.
  \end{enumerate}

  \begin{itemize}
    \item For an~$S$-scheme~$X$, its functor of points $\ull{X} =
    \Hom_S(\cdot,X)$ is an object of~$\Zar(S)$. It feels like \hil{the set of
    points} of~$X$.
    \item The ring $\affl$, given by $\affl(T) = \O_T(T)$,
    is a \hil{field} (nonzero implies invertible).
    \item \ \\[-1.2em]\mbox{$\ull{\mathbb{P}}^n = \llbracket \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\,
      x_0 \neq 0 \vee \cdots \vee x_n \neq 0 \}/(\affl)^\times \rrbracket$}
    \item $\ull{\Spec_S(\B)} = \llbracket \Hom_\affl(\B^\Zar,\affl) \rrbracket$
    \item $\ull{U} \hookrightarrow \ull{X}$ qc-open immersion iff
    \[ \Zar(S) \models \forall x\?\ull{X}\_ \bigvee_{n\geq0} \exists f_1,\ldots,f_n\?\affl\_ x \in \ull{U} \Leftrightarrow f_1 \neq 0 \vee \cdots \vee f_n \neq 0. \]
  \end{itemize}
\end{frame}

% XXX: Mehr Leute zitieren!
% XXX: Einordnung in Gesamtkontext (etwa SDG, Kock--Lawvere, Coquand & Co., ...)
% XXX: Folie zu klassifizierenden Topoi

\end{document}

\begin{frame}{The Kripke--Joyal semantics for the little Zariski topos}
  Recall~$A[f^{-1}] = \bigl\{ \frac{u}{f^n} \,|\, u \in A, n \in \NN \bigr\}$.

  \begin{itemize}
    \item $\Sh(\Spec(A)) \models \text{``For all~$x \in A^\sim$, \ldots''}$

          Meaning: For all~$f \in A$ and all~$x \in A[f^{-1}]$, \ldots
          \medskip

    \item $\Sh(\Spec(A)) \models \text{``There is~$x \in A^\sim$ such that \ldots''}$

          \mbox{Meaning: There is a partition of unity,~$1 = f_1 + \cdots + f_n \in A$,}
          such that for each~$i$, there exists~$x_i \in A[f_i^{-1}]$
          with~\ldots
          \medskip

    \item $\Sh(\Spec(A)) \models \text{``$\varphi$ implies $\psi$''}$

          Meaning: For all~$f \in A$, if~$\varphi$ on stage~$f$, then~$\psi$ on
          stage~$f$.
  \end{itemize}
\end{frame}

\section{Vision}

\begin{frame}{Vision for the future}
  \begin{columns}
    \small
    \begin{column}{0.35\textwidth}
      \centering
      \bigheart
      \par

      constructive algebra \\
      conceptual proofs \\
      proof assistants
    \end{column}

    \begin{column}{0.47\textwidth}
      \centering
      \bigheart
      \par

      constructive algebraic geometry \\
      synthetic algebraic geometry \\
      intersection theory \\
      derived categories
    \end{column}

    \begin{column}{0.25\textwidth}
      \centering
      \bigheart
      \par

      quasicoherence \\
      nongeometric mysteries \\
      further fields
    \end{column}
  \end{columns}

  \vfill

  \centering
  \href{https://www.oliviacaramello.com/Papers/Papers.htm}{\includegraphics[height=0.45\textheight]{olivia-tst}}
  \href{http://math.andrej.com/2014/03/04/intuitionistic-mathematics-and-realizability-in-the-physical-world/}{\includegraphics[height=0.45\textheight]{zenil-computable-universe}}
  \href{https://pizzaseminar.speicherleck.de/skript2/zariski-topos-klein.pdf}{\includegraphics[height=0.45\textheight]{fun-with-the-zariski-topos}}
  \href{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{\includegraphics[height=0.45\textheight]{phd-cover}}
\end{frame}

\addtocounter{framenumber}{-1}

\end{document}

### 2018-06-28 slides-como2018.tex
More slides

@@
 \renewcommand{\G}{\mathcal{G}}
+\newcommand{\J}{\mathcal{J}}
 \newcommand{\GG}{\mathbb{G}}
@@
   \item[]}{\end{list}}
+
+\newenvironment{indentblock}{%
+  \list{}{\leftmargin\leftmargin}%
+  \item\relax
+}{%
+  \endlist
+}
 
@@
 
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses}\end{minipage}}
 \begin{frame}[c]
@@
 
-  \hil{Exploring the internal language of toposes} \\
+  \hil{How topos theory can help algebra} \\
 
@@
 
+  \bigskip
   \bigskip
@@
 
-      \hil{Alternate universes}
+      \hil{Background on the internal language}
       \medskip
@@
     \hspace*{0.5em}
-    \begin{column}{0.25\textwidth}
+    \begin{column}{0.35\textwidth}
       \centering
@@
 
-      \hil{Applications}
+      \hil{Applications in commutative algebra}
       \medskip
@@
 
-      \hil{Vision for the future}
+      \hil{The mystery of nongeometric sequents}
       \medskip
@@
   Toposes in Como \\
-  June 29th, 20s18
+  June 29th, 2018
   \par
-\end{frame}
+\end{frame}}
+
+\section{Motivation}
 
@@
 
-  % XXX: \includegraphics[height=3em]{generic-freeness}
+  \includegraphics[height=3em]{generic-freeness}
   \vspace*{-1em}
@@
 
-  \pause\raggedright
+  \pause\center
+  \begin{minipage}{0.9\textwidth}
     \textbf{Proof.} See [Stacks Project, Tag 051Q].
+  \end{minipage}
 \end{frame}
+
+\section{Internal language}
 
@@
     \def\insertblocktitle{}
-    \begin{column}{0.3\textwidth}\usebeamertemplate{block begin}
+    \begin{column}{0.25\textwidth}\usebeamertemplate{block begin}
       \centering
@@
 
-    \begin{column}{0.3\textwidth}\usebeamertemplate{block begin}
+    \begin{column}{0.25\textwidth}\usebeamertemplate{block begin}
       \centering
@@
   \]
+
+  \vspace*{-1em}
+
   \begin{itemize}
     \item $\Sh(X) \models \speak{The set $\C$ is the set of (Dedekind) reals}$.
-    \item $\Sh(X) \models \speak{The function $\bar f : \RR \to \RR$ is continuous}$.
+    \item $\Sh(X) \models \speak{The function $\bar f : \RR \to \RR$ is continuous}$. \bigskip
     \item Iff $f_x(-1) < 0$ for all $x$, then $\Sh(X) \models \bar f(-1) < 0$.
     \item Iff $f_x(+1) > 0$ for all $x$, then $\Sh(X) \models \bar f(+1) > 0$.
-    \item Iff all $f_x$ are increasing, then $\Sh(X) \models \speak{$\bar f$ is increasing}$.
+    \item Iff all $f_x$ are increasing, then $\Sh(X) \models \speak{$\bar f$ is
+    increasing}$. \bigskip
     \item \ \\[-1.2em]\mbox{Iff there is an open cover~$X = \bigcup_i U_i$ such
@@
 
+\section{The little Zariski topos}
+
 \begin{frame}{The little Zariski topos}
@@
     \item the classifying topos of \hil{local localizations} of $A$,
-    \item the classifying topos of \hil{prime filters} of $A$,
+    \item the classifying locale of \hil{prime filters} of $A$,
     \item the locale given by the frame of \hil{radical ideals} of $A$,
@@
 
-  \only<1>{\begin{center}
+  \pause
+  \only<2>{\begin{center}
     \includegraphics[width=0.9\textwidth]{hartshorne-tilde-construction} \\
@@
   \end{center}}
-
   \pause
 
+  \vspace*{-1.5em}
+
   \begin{columns}
-    \begin{column}{0.5\textwidth}
+    \begin{column}{0.48\textwidth}
       \begin{varblock}{\textwidth}{}
@@
       \end{varblock}
+
+      \vspace*{-1.5em}
+
+      \begin{varblock}{\textwidth}{}
+        $M^\sim$ inherits any property of~$M$ which is \hil{localization-stable}.
+      \end{varblock}
     \end{column}
@@
     \begin{column}{0.5\textwidth}
-      \begin{varblock}{\textwidth}{}
-        $M^\sim$ inherits any property of~$M$ which is \hil{localization-stable}.
-      \end{varblock}
+      \vspace*{1.5em}
 
-      \bigskip
       If $A$ is reduced ($x^n = 0 \Rightarrow x = 0$):
 
-      \vspace*{-1em}
+      \vspace*{-1.0em}
+
       \setbeamercolor{block body}{bg=red!30}
@@
 
-  \visible<3>{\begin{tikzpicture}[overlay]
+  \visible<4-5>{\begin{tikzpicture}[overlay]
     \draw[fill=white, draw=white, opacity=0.95] (-1,0) rectangle (\paperwidth,7.4);
     \node[anchor=south west,inner sep=0] (image) at (0,0.8) {\vbox{
+      \only<4>{
 	\centering
@@
 	Miles Tierney. On the spectrum of a ringed topos. 1976.
+      }
+
+      \only<5>{
+	\begin{varblock}{\textwidth}{Complexity reduction}
+	  The external meaning of
+	  \[
+	    \Spec(A) \models
+	      \speak{$A^\sim[X_1,\ldots,X_n]$ is anonymously Noetherian}
+	  \]
+	  is:
+	  \medskip
+
+	  \begin{indentblock}
+	  For any element~$f \in A$ and any (not necessarily quasicoherent) sheaf of
+	  ideals~$\J \hookrightarrow A^\sim[X_1,\ldots,X_n]|_{D(f)}$: If
+	  \begin{indentblock}
+	  for any element~$g \in A$ the condition that
+	  \begin{indentblock}
+	  the sheaf~$\J$ is of finite type on~$D(g)$
+	  \end{indentblock}
+	  implies that~$g = 0$,
+	  \end{indentblock}
+	  then~$f = 0$.
+	  \end{indentblock}
+	\end{varblock}
+
+	\vspace*{-2em}
+      }
     }};
@@
   \only<1>{\begin{columns}[t]
-    \begin{column}[t]{0.50\textwidth}
+    \begin{column}[t]{0.47\textwidth}
       \centering
@@
       \justifying
-      \textbf{Proof.} The matrix is surjective over the field~$A^\sim$. This is
-      a contradiction to basic linear algebra. Hence $\Spec(A) \models \bot$,
-      thus $1 = 0$ in $A$.
+      \textbf{Proof.} The matrix is surjective over the field~$A^\sim$.
+      This is a contradiction to basic linear algebra. Hence $\Spec(A) \models
+      \bot$, thus $1 = 0$ in $A$.
     \end{column}
-    \begin{column}[t]{0.5\textwidth}
+    \begin{column}[t]{0.47\textwidth}
       \centering
@@
 
+  \vspace*{-5.3em}\hspace*{21em}\begin{minipage}{3cm}
+    $\xymatrixcolsep{3.5pc}\xymatrixrowsep{1.2pc}\xymatrix{
+      & \ar@{-}[d]^{\substack{\text{finitely}\\\text{generated}}} M \\
+      A \ar[r]_{\text{of finite type}} & B
+    }$
+  \end{minipage}
+  \vspace*{-1em}
+
   \centering
 
-  % XXX: \includegraphics[height=3em]{generic-freeness}
-  \vspace*{-1em}
-
-  \begin{varblock}{\textwidth}{Generic freeness$\phantom{p}$}
+  \begin{varblock}{0.9\textwidth}{Generic freeness$\phantom{p}$}
     \justifying
@@
     \begin{enumerate}
-      \item $B[f^{-1}]$ and $M[f^{-1}]$ are free modules over $A[f^{-1}]$, \\[-1.6em]
-      \item $A[f^{-1}] \to B[f^{-1}]$ is of finite presentation and \\[-1.6em]
+      \item $B[f^{-1}]$ and $M[f^{-1}]$ are free modules over $A[f^{-1}]$,
+      \item $A[f^{-1}] \to B[f^{-1}]$ is of finite presentation and
       \item $M[f^{-1}]$ is finitely presented as a module over~$B[f^{-1}]$,
@@
 
-  \pause\raggedright\bigskip
+  \pause
+
+  \scalebox{0.9}{\begin{minipage}{1\textwidth}
     \textbf{Proof.} In the little Zariski topos it's \hil{not not} the case that % \\[-1.4em]
@@
       \item $B^\sim$ and $M^\sim$ are free modules over $A^\sim$,
-    \item $A^\sim \to B^\sim$ is of finite presentation and
-    \item $M^\sim$ is finitely presented as a module over $B^\sim$
+      \item \vspace*{-0.3em}$A^\sim \to B^\sim$ is of finite presentation and
+      \item \vspace*{-0.3em}$M^\sim$ is finitely presented as a module over $B^\sim$,
     \end{enumerate}
@@
     external translation of this fact.
+  \end{minipage}}
 \end{frame}
@@
 % Klassifizierende Topoi
+
+\section{Algebraic geometry}
 
@@
   \begin{columns}[c]
-    \begin{column}{0.43\textwidth}
-      \begin{exampleblock}{}
-        \justifying
-        Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence of
-        modules. If~$M'$ and~$M''$ are finitely generated, so is~$M$.
-      \end{exampleblock}
-    \end{column}
-
-    \scalebox{3}{$\Rightarrow$}
-    \hspace*{0.5em}
-
     \begin{column}{0.47\textwidth}
@@
         so is~$\F$.
+      \end{exampleblock}
+    \end{column}
+
+    \begin{column}{0.1\textwidth}
+      \vspace*{0.7em}
+      \scalebox{3}{$\Leftarrow$}
+    \end{column}
+
+    \begin{column}{0.44\textwidth}
+      \begin{exampleblock}{}
+        \justifying
+        Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence of
+        modules. If~$M'$ and~$M''$ are finitely generated, so is~$M$.
       \end{exampleblock}
@@
 
-  \hil{Synthetic approach:} model schemes \hil{directly as sets} in a certain
-  nonclassical set theory, the internal universe of the \mbox{\hil{big Zariski
-  topos}} of a base scheme.
+  \hil{Synthetic approach:} model schemes \hil{directly as sets} in
+  the internal universe of the \mbox{\hil{big Zariski topos}} of a base scheme.
   \small
@@
 \begin{frame}{The big Zariski topos}
+  Let~$S$ be a fixed base scheme.
+  \begin{varblock}{0.9\textwidth}{Definition}
     The \hil{big Zariski topos} $\Zar(S)$ of a scheme~$S$ is equivalently
@@
       \item the classifying topos of local rings over~$S$ or
-    \item the classifying~$\Sh(S)$-topos of local~$\O_S$-algebras which are local
-    over~$\O_S$.
+      \item the classifying~$\Sh(S)$-topos of local~$\O_S$-algebras which are
+      local over~$\O_S$.
     \end{enumerate}
+  \end{varblock}
 
@@
     points} of~$X$.
-    \item The ring $\affl$, given by $\affl(T) = \O_T(T)$,
-    is a \hil{field} (nonzero implies invertible).
-    \item \ \\[-1.2em]\mbox{$\ull{\mathbb{P}}^n = \llbracket \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\,
-      x_0 \neq 0 \vee \cdots \vee x_n \neq 0 \}/(\affl)^\times \rrbracket$}
-    \item $\ull{\Spec_S(\B)} = \llbracket \Hom_\affl(\B^\Zar,\affl) \rrbracket$
-    \item $\ull{U} \hookrightarrow \ull{X}$ qc-open immersion iff
-    \[ \Zar(S) \models \forall x\?\ull{X}\_ \bigvee_{n\geq0} \exists f_1,\ldots,f_n\?\affl\_ x \in \ull{U} \Leftrightarrow f_1 \neq 0 \vee \cdots \vee f_n \neq 0. \]
+    \item In particular, there is the ring object~$\affl$ with~$\affl(T) =
+    \O_T(T)$.
+    \item This ring object is a \hil{field}: nonzero implies
+    invertible. \\{} [Kock 1976]
   \end{itemize}
+\end{frame}
+
+\begin{frame}{Synthetic constructions}
+  \small
+  $\hil{$\boldsymbol{\mathbb{A}^n}$} = (\affl)^n = \affl \times \cdots \times \affl$
+
+  $\begin{array}{@{}r@{}c@{}l@{}}
+    \hil{$\boldsymbol{\mathbb{P}^n}$} &\phantom{}=\phantom{}& \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee
+    \cdots \vee x_n \neq 0 \}/(\affl)^\times \\
+    &\cong& \text{set of one-dimensional subspaces of~$(\affl)^{n+1}$} \\
+    &&\qquad \text{(with~$\O(-1) = (\ell)_{\ell \? \mathbb{P}^n}$, $\O(1) =
+    (\ell^\vee)_{\ell \? \mathbb{P}^n}$)}
+  \end{array}$
+
+  $\hil{$\boldsymbol{\Spec(R)}$} = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
+    \text{set of $\affl$-valued points of $R$}$
+
+  $\hil{$\boldsymbol{TX}$} = X^\Delta$, where $\Delta = \{ \varepsilon \? \affl \,|\, \varepsilon^2 = 0 \}$
+
+  A subset $U \subseteq X$ is \hil{qc-open} if and only if for any $x : X$
+  there exist $f_1,\ldots,f_n \? \affl$ such that $x \in U \Longleftrightarrow
+  \exists i\_ f_i \neq 0$.
+
+  A \hil{synthetic affine scheme} is a set which is in bijection
+  with~$\Spec(R)$ for some synthetically quasicoherent~$\affl$-algebra~$R$.
+
+  A \hil{synthetic scheme} is a set which can be covered by finitely many
+  qc-open synthetic affine schemes~$U_i$ such that the intersections~$U_i \cap
+  U_j$ can be covered by finitely many qc-open synthetic affine schemes.
+\end{frame}
+
+\section{The mystery of nongeometric sequents}
+
+\begin{frame}{Properties of the affine line}
+  \begin{itemize}
+    \item $\affl$ is a local ring:
+    \[ 1 \neq 0 \qquad\qquad \text{$x + y$ \inv} \Longrightarrow \text{$x$ \inv} \vee
+    \text{$y$ \inv} \]
+    \item $\affl$ is a field:
+    \begin{align*}
+      \neg(x = 0) &\Longleftrightarrow \text{$x$ invertible \quad [Kock 1976]} \\
+      \neg(\text{$x$ invertible}) &\Longleftrightarrow \text{$x$ nilpotent}
+    \end{align*}
+    \item $\affl$ satisfies the axiom of microaffinity: Any map $f : \Delta \to
+    \affl$ is of the form~$f(\varepsilon) = a + b \varepsilon$ for unique
+    values~$a,b \? \affl$, where~$\Delta = \{ \varepsilon \? \affl \,|\,
+    \varepsilon^2 = 0 \}$.
+    \item Any map $\affl \to \affl$ is a polynomial.
+    \item $\affl$ is anonymously algebraically closed: Any monic polynomial
+    does \emph{not not} have a zero.
+  \end{itemize}
+\end{frame}
+
+\begin{frame}{Synthetic quasicoherence}
+  Recall~$\Spec(R) = \Hom_{\mathrm{Alg}(\affl)}(R, \affl)$ and consider the statement
+  \[ \text{``the canonical map~$
+    \begin{array}[t]{rcl}
+      R &\longrightarrow& (\affl)^{\Spec(R)} \\
+      f &\longmapsto& (\alpha \mapsto \alpha(f))
+    \end{array}
+  $ is bijective''}.
+  \]
+  \vspace*{-1em}
+
+  \begin{itemize}
+    \item True for~$R = \affl[X]/(X^2)$ (microaffinity).
+    \item True for~$R = \affl[X]$ (every function is a polynomial).
+    \item True for \hil{any} finitely presented~$\affl$-algebra~$R$.
+  \end{itemize}
+
+  Any known property of~$\affl$ follows from this \\
+  \hil{synthetic quasicoherence}.
+
+  \centering
+  \hil{The mystery of nongeometric sequents}
+
+% \emph{Example.} Let~$x \? \affl$ such that~$x \neq 0$. Set~$R = \affl/(x)$.
+% Then~$\Spec(R) = \emptyset$. Thus~$(\affl)^{\Spec(R)}$ is a singleton.
+% Hence~$R = 0$. Therefore~$x$ is invertible.
+\end{frame}
+
+\begin{frame}{Classifying toposes in algebraic geometry}
+  \small
+  \centering
+  \begin{tabular}{ll}
+    \toprule
+    (Big) topos & classified theory \\\midrule
+    Zariski & local rings [Hakim 1972] \\[0.6em]
+    étale & separably closed local rings [Hakim 1972, Wraith 1979] \\[0.6em]
+    \usebeamercolor[fg]{item}{fppf} & \usebeamercolor[fg]{item}{fppf-local rings} \\
+    & \textcolor{red!90}{(conjecturally: algebraically closed local rings)} \\[0.6em]
+    \textcolor{red!80}{ph} & \textcolor{red!80}{?? (conjecturally: algebraically closed valuation} \\
+    & \qquad \textcolor{red!80}{rings validating the projective Nullstellensatz)} \\[0.6em]
+    \usebeamercolor[fg]{item}{surjective} & \usebeamercolor[fg]{item}{algebraically closed geometric fields} \\[0.6em]
+    \textcolor{red!80}{$\neg\neg$} & \textcolor{red!80}{?? (conjecturally: algebraically closed geometric} \\
+    & \qquad \textcolor{red!80}{fields are integral over the base)} \\[0.6em]
+    \usebeamercolor[fg]{item}{infinitesimal} & \usebeamercolor[fg]{item}{XXX [Hutzler 2018]} \\[0.6em]
+    \textcolor{red!100}{crystalline} & \textcolor{red!100}{??} \\
+    \bottomrule
+  \end{tabular}
 \end{frame}

### 2018-06-28 slides-como2018.tex
Slide on relations between the toposes

@@
 
-% Großer Zariskitopos
-% Zusammenhang zum kleinen Topos
-% Synthetische Quasikohärenz
-% Klassifizierende Topoi
-
 \section{Algebraic geometry}
@@
 
+\begin{frame}{Relations between the Zariski toposes}
+  The big Zariski topos is a topos over the small Zariski topos:
+  \[ \begin{array}{crcl}
+    \pi : & \Zar(A) &\longrightarrow& \Spec(A) \\
+    & \text{local~$A$-algebra~$(A \xrightarrow{\alpha} B)$} &\longmapsto& (A \to A[(\alpha^{-1}[B^\times])^{-1}])
+  \end{array} \]
+  This morphism is \hil{connected} ($\pi^{-1}$ is fully faithful) and
+  \hil{local}, so there is a preinverse
+  \[ \begin{array}{rcl}
+    \Spec(A) &\longrightarrow& \Zar(A) \\
+    \text{local localization~$(A \to B)$} &\longmapsto& (A \to B)
+  \end{array} \]
+  which is a subtopos inclusion inducing an idempotent monad~$\sharp$ and an
+  idempotent comonad~$\flat$ on~$\Zar(S)$.
+
+  \begin{itemize}
+  \justifying
+    \item Internally to~$\Zar(S)$, $\Spec(S)$ can be constructed as the
+    \hil{largest subtopos} where~$\flat \affl \to \affl$ is bijective.
+    \item Internally to~$\Spec(S)$, $\Zar(S)$ can be constructed as the
+    \hil{classifying topos} of local~$\O_S$-algebras which are local over~$\O_S$.
+    \item $\Zar(A)$ is the \hil{lax pullback}~$(\Set
+    \Rightarrow_{\Set[\mathrm{Ring}]} \Set[\mathrm{LocRing}])$.
+  \end{itemize}
+\end{frame}
+
 \begin{frame}{Synthetic constructions}
@@
   \centering
-  \hil{The mystery of nongeometric sequents}
+  \begin{varblock}{0.6344\textwidth}{}
+    \centering
+    \hil{the mystery of nongeometric sequents}
+    \par
+  \end{varblock}
 

### 2018-06-28 slides-como2018.tex
Final version of the slides

@@
 
+{\usebackgroundtemplate{\includegraphics[width=\paperwidth]{slides-como2018-titlepage-blurred}}
+\begin{frame}[plain]\end{frame}}
+
+\addtocounter{framenumber}{-1}
+
 {\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses}\end{minipage}}
@@
       \bignumber{1}
+      \smallskip
 
@@
       \medskip
-
     \end{column}
@@
       \bignumber{2}
+      \smallskip
 
@@
       \medskip
-
-%      \includegraphics[height=5.5em]{calabi-yau}
     \end{column}
@@
       \bignumber{3}
+      \smallskip
 
@@
       \medskip
-
-%      \includegraphics[height=5.5em]{question-mark}
     \end{column}
@@
 
+\begin{frame}{Synthetic constructions}
+  \small
+  $\hil{$\boldsymbol{\mathbb{A}^n}$} = (\affl)^n = \affl \times \cdots \times \affl$
+
+  $\begin{array}{@{}r@{}c@{}l@{}}
+    \hil{$\boldsymbol{\mathbb{P}^n}$} &\phantom{}=\phantom{}& \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee
+    \cdots \vee x_n \neq 0 \}/(\affl)^\times \\
+    &\cong& \text{set of one-dimensional subspaces of~$(\affl)^{n+1}$} \\
+    &&\qquad \text{(with~$\O(-1) = (\ell)_{\ell \? \mathbb{P}^n}$, $\O(1) =
+    (\ell^\vee)_{\ell \? \mathbb{P}^n}$)}
+  \end{array}$
+
+  $\hil{$\boldsymbol{\Spec(R)}$} = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
+    \text{set of $\affl$-valued points of $R$}$
+
+  $\hil{$\boldsymbol{TX}$} = X^\Delta$, where $\Delta = \{ \varepsilon \? \affl \,|\, \varepsilon^2 = 0 \}$
+
+  A subset $U \subseteq X$ is \hil{qc-open} if and only if for any $x : X$
+  there exist $f_1,\ldots,f_n \? \affl$ such that $x \in U \Longleftrightarrow
+  \exists i\_ f_i \neq 0$.
+
+  A \hil{synthetic affine scheme} is a set which is in bijection
+  with~$\Spec(R)$ for some synthetically quasicoherent~$\affl$-algebra~$R$.
+
+  A \hil{finitely presented synthetic scheme} is a set which can be covered by
+  finitely many qc-open f.p. synthetic affine schemes~$U_i$ such that the
+  intersections~$U_i \cap U_j$ can be covered by finitely many qc-open f.p.
+  synthetic affine schemes.
+\end{frame}
+
 \begin{frame}{Relations between the Zariski toposes}
@@
   \end{itemize}
-\end{frame}
-
-\begin{frame}{Synthetic constructions}
-  \small
-  $\hil{$\boldsymbol{\mathbb{A}^n}$} = (\affl)^n = \affl \times \cdots \times \affl$
-
-  $\begin{array}{@{}r@{}c@{}l@{}}
-    \hil{$\boldsymbol{\mathbb{P}^n}$} &\phantom{}=\phantom{}& \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee
-    \cdots \vee x_n \neq 0 \}/(\affl)^\times \\
-    &\cong& \text{set of one-dimensional subspaces of~$(\affl)^{n+1}$} \\
-    &&\qquad \text{(with~$\O(-1) = (\ell)_{\ell \? \mathbb{P}^n}$, $\O(1) =
-    (\ell^\vee)_{\ell \? \mathbb{P}^n}$)}
-  \end{array}$
-
-  $\hil{$\boldsymbol{\Spec(R)}$} = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
-    \text{set of $\affl$-valued points of $R$}$
-
-  $\hil{$\boldsymbol{TX}$} = X^\Delta$, where $\Delta = \{ \varepsilon \? \affl \,|\, \varepsilon^2 = 0 \}$
-
-  A subset $U \subseteq X$ is \hil{qc-open} if and only if for any $x : X$
-  there exist $f_1,\ldots,f_n \? \affl$ such that $x \in U \Longleftrightarrow
-  \exists i\_ f_i \neq 0$.
-
-  A \hil{synthetic affine scheme} is a set which is in bijection
-  with~$\Spec(R)$ for some synthetically quasicoherent~$\affl$-algebra~$R$.
-
-  A \hil{synthetic scheme} is a set which can be covered by finitely many
-  qc-open synthetic affine schemes~$U_i$ such that the intersections~$U_i \cap
-  U_j$ can be covered by finitely many qc-open synthetic affine schemes.
 \end{frame}
@@
     & \textcolor{red!90}{(conjecturally: algebraically closed local rings)} \\[0.6em]
-    \textcolor{red!80}{ph} & \textcolor{red!80}{?? (conjecturally: algebraically closed valuation} \\
-    & \qquad \textcolor{red!80}{rings validating the projective Nullstellensatz)} \\[0.6em]
+    \textcolor{red!80}{ph} & \textcolor{red!80}{?? (conjecturally: algebraically closed valuation rings} \\
+    & \qquad \textcolor{red!80}{validating the projective Nullstellensatz)} \\[0.6em]
     \usebeamercolor[fg]{item}{surjective} & \usebeamercolor[fg]{item}{algebraically closed geometric fields} \\[0.6em]
@@
     & \qquad \textcolor{red!80}{fields are integral over the base)} \\[0.6em]
-    \usebeamercolor[fg]{item}{infinitesimal} & \usebeamercolor[fg]{item}{XXX [Hutzler 2018]} \\[0.6em]
+    \usebeamercolor[fg]{item}{infinitesimal} & \usebeamercolor[fg]{item}{local algebras together with a nilpotent ideal [Hutzler 2018]} \\[0.6em]
     \textcolor{red!100}{crystalline} & \textcolor{red!100}{??} \\

### 2018-07-02 slides-como2018.tex
Notes almost finished

@@
 \usepackage{etex}
+
+\usepackage{pgfpages}
+\setbeameroption{show notes on second screen}
+\setbeamertemplate{note page}[plain]
+\newcommand{\jnote}[2]{\only<#1>{\note{\justifying#2\par}}}
 
@@
 \hypersetup{colorlinks=true}
-\usepackage{multimedia}
-\newcommand{\video}[2]{\movie[width=#2,height=#2,autostart,loop,poster]{}{#1}}
-\hypersetup{colorlinks=false}
-
-\usepackage{pifont}
-\newcommand{\cmark}{\ding{51}}
-\newcommand{\xmark}{\ding{55}}
-\DeclareSymbolFont{extraup}{U}{zavm}{m}{n}
-\DeclareMathSymbol{\varheart}{\mathalpha}{extraup}{86}
 
@@
 \newcommand{\F}{\mathcal{F}}
+\newcommand{\M}{\mathcal{M}}
 \renewcommand{\G}{\mathcal{G}}
@@
 }
-\newcommand{\bigheart}[1]{\scalebox{1.5}{\hil{$\varheart$}}}
-
-\newenvironment{changemargin}[2]{%
-  \begin{list}{}{%
-    \setlength{\topsep}{0pt}%
-    \setlength{\leftmargin}{#1}%
-    \setlength{\rightmargin}{#2}%
-    \setlength{\listparindent}{\parindent}%
-    \setlength{\itemindent}{\parindent}%
-    \setlength{\parsep}{\parskip}%
-  }%
-  \item[]}{\end{list}}
 
@@
 
-\addtocounter{framenumber}{-1}
-
-{\usebackgroundtemplate{\includegraphics[width=\paperwidth]{slides-como2018-titlepage-blurred}}
-\begin{frame}[plain]\end{frame}}
+%\addtocounter{framenumber}{-1}
+%{\usebackgroundtemplate{\includegraphics[width=\paperwidth]{slides-como2018-titlepage-blurred}}
+%\begin{frame}[plain]\end{frame}}
 
@@
   \hspace*{-3.9em}%
-%  \includegraphics[height=0.3\textheight]{vichy-2}%
-%  \includegraphics[height=0.3\textheight]{vichy-1}%
-%  \includegraphics[height=0.3\textheight]{vichy-4}%
-%  \includegraphics[height=0.3\textheight]{vichy-5}%
 
@@
     \small
-    \begin{column}{0.35\textwidth}
+    \begin{column}{0.33\textwidth}
       \centering
@@
     \end{column}
-    \hspace*{0.5em}
-    \begin{column}{0.35\textwidth}
+    \begin{column}{0.33\textwidth}
       \centering
@@
     \end{column}
-    \begin{column}{0.4\textwidth}
+    \begin{column}{0.33\textwidth}
       \centering
@@
 
-      \hil{The mystery of nongeometric sequents}
+      \hil{The mystery of \mbox{\!\!nongeometric sequents}}
       \medskip
@@
   Assume that $A$ is reduced: If $x^n = 0$, then $x = 0$.
+
+  \jnote{1-3}{The two displayed statements are trivial for fields. It is
+  therefore natural to try to reduce the general situation to the field
+  situation.}
+
+  \jnote{2-3}{The displayed proofs, which could have been taken from any
+  standard textbook on commutative algebra, succeed in this reduction quite
+  easily by employing maximal ideals or minimal prime ideals. However, this way
+  of reducing comes at a cost: It requires the Boolean Prime Ideal Theorem
+  (for ensuring the existence of a prime ideal and for ensuring that stalks at
+  minimal prime ideals are fields) and even the full axiom of choice (for
+  ensuring the existence of a minimal prime ideal).
+
+  It therefore doesn't work in the internal universe of most toposes, and in
+  any case it obscures explicit computational content: Statements so simple as
+  the two displayed ones should admit explicit, computational proofs.
+
+  We'll learn how the internal language of a certain well-chosen topos provides
+  a way to perform the reduction in an entirely constructive manner. If so
+  desired, the resulting topos-theoretic proofs can be unwinded to yield fully
+  explicit, topos-free, direct proofs.
+
+  Beautiful constructive proofs can also be found in Richman's note on
+  \href{https://www.ams.org/journals/proc/1988-103-04/S0002-9939-1988-0954974-5/S0002-9939-1988-0954974-5.pdf}{nontrivial
+  uses of nontrivial rings} and in the
+  \href{https://arxiv.org/abs/1605.04832}{recent textbook by Lombardi and
+  Quitté} on constructive commutative algebra.}
+
+  \jnote{4-}{Grothendieck's generic freeness lemma is an important theorem in
+  algebraic geometry, where it is usually stated in the following geometric
+  form:
+  \begin{indentblock}Let~$X$ be a reduced scheme. Let~$\B$ be
+  an~$\O_X$-algebra of finite type. Let~$\M$ be a~$\B$-module of finite type.
+  Then over a dense open,
+  \begin{enumerate}
+  \item[(a)] $\B$ and~$\M$ are locally free as sheaves of~$\O_X$-modules,
+  \item[(b)] $\B$ is of finite presentation as a sheaf of~$\O_X$-algebras and
+  \item[(c)] $\M$ is of finite presentation as a sheaf of~$\B$-modules.
+  \end{enumerate}\end{indentblock}}
+
+  \jnote{5}{All previously known proofs proceed in a series of reduction steps,
+  finally culminating in the case where~$A$ is a Noetherian integral domain.
+  They are somewhat convoluted (for instance, the proof in the Stacks Project
+  is three pages long) and employ several results in commutative algebra which
+  have not yet been constructivized.
+
+  Using the internal language of toposes, we will give a short, conceptual and
+  constructive proof of Grothendieck's generic freeness lemma. Again, if so
+  desired, one can unwind the internal proof to obtain a
+  constructive proof which doesn't reference topos theory. The proof obtained
+  in this way is still an improvement on the previously known proofs, requiring
+  no advanced prerequisites in commutative algebra, and takes
+  \href{https://rawgit.com/iblech/internal-methods/master/paper-generic-freeness.pdf}{about
+  a page}.}
 
@@
   using (Shulman's extension of) the \hil{Kripke--Joyal semantics}.
+
+  \jnote{1}{The internal language of a topos allows to construct objects and
+  morphisms, formulate statements about them and prove such statements in a
+  naive element-based language. From the internal point of view, objects look
+  like sets [more precisely, types]; morphisms look like maps; epimorphisms
+  look like surjections; group objects look like groups; and so on.
+
+  To determine whether a statement~$\varphi$ holds in the internal universe of a
+  given topos, one can use the Kripke--Joyal semantics to translate it into an
+  ordinary external statement and then check the validity of the external
+  translation.
+
+  For instance, in the ef{}fective topos the curious statement~``any function
+  $\NN \to \NN$ is computable'' holds, for its external meaning is the
+  triviality ``there is a Turing machine which given a Turing machine computing
+  some function~$f : \NN \to \NN$ outputs a Turing machine computing~$f$''. In
+  contrast, the statement~``any function~$\NN \to \NN$ is either the zero
+  function or not'' does not hold in the ef{}fective topos, since its external
+  meaning is ``there exists a Turing machine which given a Turing machine
+  computing some function~$f : \NN \to \NN$ decides whether~$f$ is the zero
+  function or not''.}
+
+  \jnote{2}{Any theorem which has an intuitionistic proof holds in
+  the internal universe of any topos. The restriction to intuitionistic logic
+  is not due to philosophical concerns; it is a fact of life that only very few
+  toposes validate the law of excluded middle (for instance, sheaf toposes over
+  discrete topological space do if the law of excluded middle is available in
+  the metatheory). Luckily, vast amounts of mathematics can be developed in a
+  purely intuitionistic setting.
+
+  The internal language machinery itself can be developed in an intuitionistic
+  setting.
+
+  The standard internal language of toposes in not enough for our purposes, as
+  it misses unbounded quantification (``for all groups'', ``for all rings'')
+  and dependent types. Shulman's \href{https://arxiv.org/abs/1004.3802}{stack
+  semantics} offers what we need. No knowledge of stacks is necessary to enjoy
+  his paper. Prior work includes
+  \href{https://www.cl.cam.ac.uk/~amp12/papers/polist/polist.pdf}{Polymorphism
+  is Set Theoretic, Constructively} by Pitts and
+  \href{http://www.phil.cmu.edu/projects/ast/Papers/Awodey-Butz-Simpson-Streicher-APAL-2013.pdf}{Relating
+  first-order set theories, toposes and categories of classes} by Awodey, Butz,
+  Simpson and Streicher (obtained independently and published after Shulman).}
 
@@
   \[ \renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\ }c@{\ }l@{}}
-  U \models f = g \? F &\Ll& f|_U = g|_U \in F(U) \\
+  U \models s = t \? F &\Ll& s|_U = t|_U \in F(U) \\
   U \models \varphi \wedge \psi &\Ll&
@@
   \text{$V \models \varphi$ implies $V \models \psi$} \\
-  U \models \forall f \? F\_ \varphi(f) &\Ll&
-  \text{for all open $V \subseteq U$ and sections~$f_0 \in F(V)$: $V \models \varphi(f_0)$} \\
+  U \models \forall s \? F\_ \varphi(s) &\Ll&
+  \text{for all open $V \subseteq U$ and sections~$s_0 \in F(V)$: $V \models \varphi(s_0)$} \\
   U \models \forall F\_ \varphi(F) &\Ll&
   \text{for all open $V \subseteq U$ and sheaves~$F_0$ over~$V$: $V \models \varphi(F_0)$} \\
-  U \models \exists f \? F\_ \varphi(f) &\Ll&
+  U \models \exists s \? F\_ \varphi(s) &\Ll&
   \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
-  && \quad\quad \text{there exists~$f_0 \in F(U_i)$ s.\,th. $U_i \models \varphi(f_0)$} \\
+  && \quad\quad \text{there exists~$s_0 \in F(U_i)$ s.\,th. $U_i \models \varphi(s_0)$} \\
   U \models \exists F\_ \varphi(F) &\Ll&
@@
   \end{array} \]
+
+  \jnote{1}{Many interesting sheaves have few global sections, which is why a
+  definition should as ``$U \models \forall s \? F\_ \varphi(s)$ iff $U \models
+  \varphi(s_0)$ for all~$s_0 \in F(U)$'' would miss the point. More precisely,
+  changing the definitions like this would yield the internal language of the
+  topos of presheaves on~$X$.
+
+  Here is an explicit example of the translation procedure. Let~$\alpha : \F
+  \to \G$ be a morphism of sheaves on~$X$. Then:
+  \begin{align*}
+    & X \models \speak{$\alpha$ is injective} \\[0.5em]
+    \Longleftrightarrow\
+    & X \models \forall s,t\?\F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
+    &\qquad\qquad
+        U \models \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
+    &\qquad\qquad
+        \text{for all open~$V \subseteq U$:} \\
+    &\qquad\qquad\qquad\qquad
+          \text{$\alpha_V(s|_V) = \alpha_V(t|_V)$ implies $s|_V = t|_V$} \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
+    &\qquad\qquad
+          \text{$\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.5em]
+    \Longleftrightarrow\ &
+      \text{$\alpha$ is a monomorphism of sheaves}
+  \end{align*}}
 \end{frame}
@@
   \end{itemize}
+
+  \jnote{1}{This slide, unrelated to commutative algebra or algebraic geometry,
+  aims to illustrate one of the basic uses of the internal language of toposes:
+  Upgrading any theorem admitting an intuitionistic proof to a
+  parameter-dependent version.
+
+  Constructively, there are several non-equivalent forms of the intermediate
+  value theorem. The following version doesn't admit an intuitionistic proof:
+  \begin{indentblock}Let~$g : \RR \to \RR$ be a function between the (Dedekind)
+  reals which is continuous in the usual epsilon/delta sense. Assume~$g(-1) < 0
+  < g(1)$. Then there exists a number~$x \in \RR$ such that~$g(x) = 0$.
+  \end{indentblock}
+  If there was an intuitionistic proof, the statement would hold in any topos,
+  so in particular in sheaf toposes over topological spaces. By the
+  translations shown on the slide (the corner quotes ``$\speak{\ldots}$''
+  indicate that translation into formal language is left to the reader), this
+  would amount to the following strengthening of the intermediate value
+  theorem: In continuous families of continuous functions, zeros can locally be
+  picked continuously. However, this strengthening is invalid, as
+  \href{https://raw.githubusercontent.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{this
+  video shows}.
+
+  \centering\includegraphics[width=0.1\textwidth]{zeros-in-families-still}}
+
+  \jnote{2}{In contrast, the following version does admit an intuitionistic
+  proof. We can therefore interpret it in sheaf toposes over topological
+  spaces and thereby obtain the strengthening that in continuous families of
+  strictly increasing continuous functions, zeros can locally be picked
+  continuously. You are invited to prove this strengthening directly, without
+  reference to the internal language.
+  \begin{indentblock}Let~$g : \RR \to \RR$ be a function between the (Dedekind)
+  reals which is continuous in the usual epsilon/delta sense and which is
+  strictly increasing ($a < b$ implies~$g(a) < g(b)$). Assume~$g(-1) < 0 <
+  g(1)$. Then there exists a number~$x \in \RR$ such that~$g(x) =
+  0$.\end{indentblock}}
 \end{frame}
@@
   $xy \in \fff \Leftrightarrow x \in \fff \wedge y \in \fff$
+
+  \jnote{1}{Any geometric theory has a classifying topos; if the theory under
+  consideration is propositional (doesn't have any sorts), then its classifying
+  topos can be chosen to be the topos of sheaves over a locale. One can also
+  give a direct account of classifying locales, as a pedagogical stepping stone
+  to the full theory of classifying toposes.
+
+  The slide contains a small lie: The classical definition of the spectrum of a
+  ring is via the set of prime ideals of~$A$, not prime filters. If the law of
+  excluded middle is available, there is no difference between these
+  definitions since the complement of a prime ideal is a prime filter and vice
+  versa.
+
+  One can also consider the classifying local of prime \emph{ideals} of~$A$.
+  Its associated topological space of points is the the set of prime ideal
+  of~$A$ equipped with the \emph{constructible topology}.
+
+  In an intuitionistic context, any of the (generalized) spaces of items~1--4
+  can be adopted as sensible definitions of the spectrum of~$A$. Item~5 is then
+  a tautology. The classical definition of the spectrum as a topological space
+  doesn't work very well, because verifying the universal property one expects
+  of it requires the Boolean Prime Ideal Theorem. Most dramatically, there are
+  rings which are not trivial yet don't have any prime ideals or prime filters.
+  The classical definition yields in this case the empty space.}
 \end{frame}
@@
   \end{itemize}
+
+  \jnote{1-3}{The generic prime filter~$\fff_0$ can be described in explicit
+  terms. For ring elements~$f$ and~$s$, $D(f) \models (s \in \fff_0)$ iff~$f
+  \in \sqrt{(s)}$.}
+
+  \jnote{2-3}{Our description of~$M^\sim$ reveals a precise sense in
+  which~$M^\sim$ and~$M$ are related: $M^\sim$ is simply a localization of~$M$
+  (first lifted to another universe by the constant sheaf construction). The
+  classical descriptions don't make the relation evident.}
+
+  \jnote{3}{One can show, assuming that the little Zariski topos is
+  \emph{overt}, that the module~$M$ in~$\Set$ and the module~$\ull{M}$ of the
+  little Zariski topos share all first-order properties. This observation
+  explains the metatheorem displayed at the bottom left. The assumption is
+  satisfied if any element of~$A$ is nilpotent or not nilpotent, so it's always
+  satisfied if the law of excluded middle is available. In an intuitionistic
+  context, it's still ``morally satisfied''. Details are in Section~12.9 of
+  \href{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
+  notes}.}
+
+  \jnote{3}{As a first approximation, the module~$M^\sim$ can be thought of as
+  a reification of all the stalks of~$M$ as a single object. The metatheorem
+  makes this precise and also shows the limits of this view: It is only correct
+  for geometric sequents. When considering nongeometric sequents, phenomena
+  appear which are unique to~$M^\sim$ in the sense that they are in general not
+  shared by~$M$, its stalks or its quotients.}
+
+  \jnote{4}{Assuming that~$A$ is reduced, the following nongeometric sequents
+  hold in the little Zariski topos (among others):
+
+  $A^\sim$ is a field in the sense that zero is the only noninvertible
+  element. This field property was already observed in the 1970s by Mulvey, who
+  didn't know a deeper reason for this property. We now know that it's a
+  shadow of an internal property whose external translation expresses
+  that~$A^\sim$ is quasicoherent.
+
+  $A^\sim$ has~$\neg\neg$-stable equality in the sense that
+  \[ \Spec(A) \models \forall s\?A^\sim\_ \neg\neg(s = 0) \Rightarrow s = 0. \]
+  Classically, every set has~$\neg\neg$-stable equality; intuitionistically,
+  this is a special property of some sets. It's quite useful, as some theorems
+  of classical commutative algebra can only be proven intuitionistically when
+  weakened by double negation. The stability then allows, in some cases, to
+  obtain the original conclusion.
+
+  $A^\sim$ is anonymously Noetherian in the sense that any of its ideals
+  is \mbox{\emph{not not}} finitely generated. A philosophically-motivated
+  constructivist might be offended by this notion, since it runs counter to the
+  maxim that constructive mathematics should be informative. However, in the
+  internal context it is a useful notion: Hilbert's basis theorem holds for it,
+  and we'll put it to good use in our proof of Grothendieck's generic freeness
+  lemma.}
+
+  \jnote{5}{Are there theorems which can only be proven using the internal
+  language and not be proven without?
+
+  No. Just as the translation from internal statements to external statements
+  is entirely mechanical, so is the translation from internal proofs to
+  external proofs. Any proof employing the internal language can be unwinded to
+  yield an external proof not referencing the internal language.
+
+  However, depending on the logical complexity of the statements occuring in a
+  given proof, the resulting external proof might be (much) more complex than
+  the internal proof. This is particularly the case if the proof involves
+  double negation, for much the same reason as that in computer science,
+  continuations can twist the control flow in nontrivial ways which are
+  sometimes hard to understand. It is in these cases where we can extract the
+  most value of the internal language, unlocking notions and proofs which might
+  otherwise be hard to obtain.
+
+  The slide shows a specific example. The internal statement
+  that~$A^\sim[X_1,\ldots,X_n]$ is anonymously Noetherian is quite simple; its
+  external translation is quite convoluted.}
 
@@
 
-\section{Vision}
-
-\begin{frame}{Vision for the future}
-  \begin{columns}
-    \small
-    \begin{column}{0.35\textwidth}
-      \centering
-      \bigheart
-      \par
-
-      constructive algebra \\
-      conceptual proofs \\
-      proof assistants
-    \end{column}
-
-    \begin{column}{0.47\textwidth}
-      \centering
-      \bigheart
-      \par
-
-      constructive algebraic geometry \\
-      synthetic algebraic geometry \\
-      intersection theory \\
-      derived categories
-    \end{column}
-
-    \begin{column}{0.25\textwidth}
-      \centering
-      \bigheart
-      \par
-
-      quasicoherence \\
-      nongeometric mysteries \\
-      further fields
-    \end{column}
-  \end{columns}
-
-  \vfill
-
-  \centering
-  \href{https://www.oliviacaramello.com/Papers/Papers.htm}{\includegraphics[height=0.45\textheight]{olivia-tst}}
-  \href{http://math.andrej.com/2014/03/04/intuitionistic-mathematics-and-realizability-in-the-physical-world/}{\includegraphics[height=0.45\textheight]{zenil-computable-universe}}
-  \href{https://pizzaseminar.speicherleck.de/skript2/zariski-topos-klein.pdf}{\includegraphics[height=0.45\textheight]{fun-with-the-zariski-topos}}
-  \href{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{\includegraphics[height=0.45\textheight]{phd-cover}}
-\end{frame}
-
-\addtocounter{framenumber}{-1}
-
 \end{document}

### 2018-07-03 slides-como2018.tex
More notes

@@
 
-  \jnote{1-3}{The two displayed statements are trivial for fields. It is
+  \jnote{1-2}{The two displayed statements are trivial for fields. It is
   therefore natural to try to reduce the general situation to the field
@@
 
-  \jnote{2-3}{The displayed proofs, which could have been taken from any
+  \jnote{2}{The displayed proofs, which could have been taken from any
   standard textbook on commutative algebra, succeed in this reduction quite
@@
   Quitté} on constructive commutative algebra.}
+
+  \jnote{3}{We can slightly reduce the requirements of the proof of the first
+  statement by employing not a maximal ideal, but a prime ideal. The existence
+  of maximal ideals is nontrivial rings is equivalent to the axiom of choice,
+  while the existence of prime ideals is equivalent to the weaker Boolean Prime
+  Ideal Theorem. However, this improvement doesn't change the main point; the
+  proof is still wildly unconstructive.}
 
@@
   of it requires the Boolean Prime Ideal Theorem. Most dramatically, there are
-  rings which are not trivial yet don't have any prime ideals or prime filters.
+  rings which are not trivial yet have neither prime ideals nor prime filters.
   The classical definition yields in this case the empty space.}
@@
   Let $A$ be a reduced ring.
+
+  \jnote{1}{XXX ... succeeded ...}
 

### 2018-07-03 slides-como2018.tex
Work around bug in the PDF specification

@@
 \newcommand{\jnote}[2]{\only<#1>{\note{\justifying#2\par}}}
+
+% Workaround for the issue described at
+% https://tex.stackexchange.com/questions/164406/beamer-using-href-in-notes.
+\newcommand{\fixedhref}[2]{\makebox[0pt][l]{\hspace*{\paperwidth}\href{#1}{#2}}\href{#1}{#2}}
 
@@
   Beautiful constructive proofs can also be found in Richman's note on
-  \href{https://www.ams.org/journals/proc/1988-103-04/S0002-9939-1988-0954974-5/S0002-9939-1988-0954974-5.pdf}{nontrivial
+  \fixedhref{https://www.ams.org/journals/proc/1988-103-04/S0002-9939-1988-0954974-5/S0002-9939-1988-0954974-5.pdf}{nontrivial
   uses of nontrivial rings} and in the
-  \href{https://arxiv.org/abs/1605.04832}{recent textbook by Lombardi and
+  \fixedhref{https://arxiv.org/abs/1605.04832}{recent textbook by Lombardi and
   Quitté} on constructive commutative algebra.}
@@
   no advanced prerequisites in commutative algebra, and takes
-  \href{https://rawgit.com/iblech/internal-methods/master/paper-generic-freeness.pdf}{about
+  \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-generic-freeness.pdf}{about
   a page}.}
@@
   it misses unbounded quantification (``for all groups'', ``for all rings'')
-  and dependent types. Shulman's \href{https://arxiv.org/abs/1004.3802}{stack
+  and dependent types. Shulman's \fixedhref{https://arxiv.org/abs/1004.3802}{stack
   semantics} offers what we need. No knowledge of stacks is necessary to enjoy
   his paper. Prior work includes
-  \href{https://www.cl.cam.ac.uk/~amp12/papers/polist/polist.pdf}{Polymorphism
+  \fixedhref{https://www.cl.cam.ac.uk/~amp12/papers/polist/polist.pdf}{Polymorphism
   is Set Theoretic, Constructively} by Pitts and
-  \href{http://www.phil.cmu.edu/projects/ast/Papers/Awodey-Butz-Simpson-Streicher-APAL-2013.pdf}{Relating
+  \fixedhref{http://www.phil.cmu.edu/projects/ast/Papers/Awodey-Butz-Simpson-Streicher-APAL-2013.pdf}{Relating
   first-order set theories, toposes and categories of classes} by Awodey, Butz,
@@
   picked continuously. However, this strengthening is invalid, as
-  \href{https://raw.githubusercontent.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{this
+  \fixedhref{https://raw.githubusercontent.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{this
   video shows}.
@@
   context, it's still ``morally satisfied''. Details are in Section~12.9 of
-  \href{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
+  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
   notes}.}

### 2018-07-03 paper-generic-freeness.tex
Version for the arXiv

@@
 \usetikzlibrary{patterns,matrix}
-\usepackage[protrusion=true,expansion=true]{microtype}
+\usepackage[protrusion=true]{microtype}
+\usepackage{breakurl}
 \usepackage{hyperref}
@@
 comments regarding the presentation of the paper, and to Marc
-Nieper-Wißkirchen for carefully guiding my PhD studies~\cite{blechschmidt:phd} at the University of
+Nieper-Wißkirchen for carefully guiding our PhD studies~\cite{blechschmidt:phd} at the University of
 Augsburg, where most of the work for this paper was carried out.

### 2018-07-04 slides-munich2018.tex
typo

@@
       \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
-    &&\quad\text{for each~$i$, $g_i \models \varphi[x_0/x]$ for some~$x_0 \in A[(fg_i)^{-1}]$}
+    &&\quad\text{for each~$i$, $fg_i \models \varphi[x_0/x]$ for some~$x_0 \in A[(fg_i)^{-1}]$}
   \end{array} \]}

### 2018-07-04 slides-como2018.tex
More notes

@@
 
-\title{Exploring the internal language of toposes}
+\title{How topos theory can help commutative algebra}
 \author{Ingo Blechschmidt}
-\date{June 22nd, 2018 XXX}
+\date{June 29th, 2018}
 
@@
 
-  \hil{How topos theory can help algebra} \\
+  \hil{How topos theory can help commutative algebra} \\
 
@@
 
-\begin{frame}{The internal language of~$\boldsymbol{\Sh(X)}$}
-  \small
+\begin{frame}{The Kripke--Joyal semantics of~$\boldsymbol{\Sh(X)}$}
+  \small\vspace*{-0.5em}
   Let~$X$ be a topological space. We recursively define
@@
   Write~``$\Sh(X) \models \varphi$'' to mean~$X \models \varphi$.}
-  \[ \renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\ }c@{\ }l@{}}
+  \footnotesize
+  \[ \renewcommand{\arraystretch}{1.08}\begin{array}{@{}l@{\ }c@{\ }l@{}}
+  U \models \top &\Ll& \text{true} \\
+  U \models \bot &\Ll& \hcancel{\text{false}}{0pt}{3pt}{0pt}{-2pt}\ U = \emptyset \\
   U \models s = t \? F &\Ll& s|_U = t|_U \in F(U) \\
@@
   U \models \varphi \vee \psi &\Ll&
-  \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $U = \bigcup_i U_i$ s.\,th.} \\
-  && \quad\quad\text{for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
+  \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $U = \bigcup_i U_i$} \\
+  && \quad\quad\text{such that for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
   U \models \varphi \Rightarrow \psi &\Ll&
@@
   U \models \exists s \? F\_ \varphi(s) &\Ll&
-  \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
-  && \quad\quad \text{there exists~$s_0 \in F(U_i)$ s.\,th. $U_i \models \varphi(s_0)$} \\
+  \hcancel{\text{there exists $s_0 \in F(U)$ such that $U \models \varphi(s_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
+  && \quad\quad
+  \text{there exists a covering $U = \bigcup_i U_i$ such that for all~$i$:} \\
+  && \quad\quad\quad\quad \text{there exists~$s_0 \in F(U_i)$ such that $U_i \models \varphi(s_0)$} \\
   U \models \exists F\_ \varphi(F) &\Ll&
-  \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
-  && \quad\quad \text{there exists a sheaf~$F_0$ on $U_i$ s.\,th. $U_i \models \varphi(F_0)$}
+  \hcancel{\text{there exists a sheaf $F_0$ on $U$ such that $U \models
+  \varphi(F_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
+  && \quad\quad
+  \text{there exists a covering $U = \bigcup_i U_i$ such that for all~$i$:} \\
+  && \quad\quad\quad\quad \text{there exists a sheaf~$F_0$ on $U_i$ such that $U_i \models \varphi(F_0)$}
   \end{array} \]
@@
 
-  Here is an explicit example of the translation procedure. Let~$\alpha : \F
-  \to \G$ be a morphism of sheaves on~$X$. Then:
+  Here is an explicit example of the translation procedure. Let~$\alpha : F
+  \to G$ be a morphism of sheaves on~$X$. Then (the corner quotes ``$\speak{\ldots}$''
+  indicate that translation into formal language is left to the reader):
+  \footnotesize
   \begin{align*}
-    & X \models \speak{$\alpha$ is injective} \\[0.5em]
+    & X \models \speak{$\alpha$ is injective} \\[0.2em]
     \Longleftrightarrow\
-    & X \models \forall s,t\?\F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
+    & X \models \forall s\?F\_ \forall t\?F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.2em]
     \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
+      \text{for all open~$U \subseteq X$, sections $s_0 \in F(U)$:} \\
     &\qquad\qquad
-        U \models \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
+        U \models \forall t\?F\_ \alpha(s_0) = \alpha(t) \Rightarrow s_0 = t \\[0.2em]
     \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
+      \text{for all open~$U \subseteq X$, sections $s_0 \in F(U)$:} \\
     &\qquad\qquad
-        \text{for all open~$V \subseteq U$:} \\
+        \text{for all open~$V \subseteq U$, sections $t_0 \in F(V)$:} \\
     &\qquad\qquad\qquad\qquad
-          \text{$\alpha_V(s|_V) = \alpha_V(t|_V)$ implies $s|_V = t|_V$} \\[0.5em]
+          V \models \alpha(s_0) = \alpha(t_0) \Rightarrow s_0 = t_0 \\[0.2em]
     \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
+      \text{for all open~$U \subseteq X$, sections $s_0 \in F(U)$:} \\
     &\qquad\qquad
-          \text{$\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.5em]
+        \text{for all open~$V \subseteq U$, sections $t_0 \in F(V)$:} \\
+    &\qquad\qquad\qquad\qquad
+          \text{for all open $W \subseteq V$: $\alpha_V(s_0|_W) = \alpha_V(t_0|_W)$ implies $s_0|_W = t_0|_W$} \\[0.2em]
+    \Longleftrightarrow\ &
+      \text{for all open~$U \subseteq X$, sections $s, t \in F(U)$:
+        $\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.2em]
     \Longleftrightarrow\ &
@@
   so in particular in sheaf toposes over topological spaces. By the
-  translations shown on the slide (the corner quotes ``$\speak{\ldots}$''
-  indicate that translation into formal language is left to the reader), this
+  translations shown on the slide, this
   would amount to the following strengthening of the intermediate value
@@
 
-  \jnote{1-3}{The generic prime filter~$\fff_0$ can be described in explicit
+  \jnote{1}{The Kripke--Joyal semantics for the little Zariski topos amounts to
+  the following: $\Spec(A) \models \varphi$ iff~$D(1) \models \varphi$, and the
+  clauses for~$D(f) \models \varphi$, where~$f$ ranges over the elements
+  of~$A$, are given by the following table.
+  \[\footnotesize\renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\quad}c@{\quad}l@{}}
+    D(f) \models \top &\text{iff}& \text{true} \\
+    D(f) \models \bot &\text{iff}& \text{$f$ is nilpotent} \\
+    D(f) \models x = y &\text{iff}& x = y \in A[f^{-1}] \\
+    D(f) \models \varphi \wedge \psi &\text{iff}&
+      \text{$D(f) \models \varphi$ and $D(f) \models \psi$} \\
+    D(f) \models \varphi \vee \psi &\text{iff}&
+      \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
+    &&\quad\text{for each~$i$, $D(fg_i) \models \varphi$ or $D(fg_i) \models \psi$} \\
+    D(f) \models \varphi \Rightarrow \psi &\text{iff}&
+      \text{for all~$g \in A$, $D(fg) \models \varphi$ implies $D(fg) \models \psi$} \\
+    D(f) \models \forall x\?M^\sim\_ \varphi(x) &\text{iff}&
+      \text{for all~$g \in A$ and $x_0 \in M[(fg)^{-1}]$, $D(fg) \models \varphi[x_0/x]$} \\
+    D(f) \models \exists x\?M^\sim\_ \varphi(x) &\text{iff}&
+      \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
+    &&\quad\text{for each~$i$, $D(fg_i) \models \varphi[x_0/x]$ for some~$x_0 \in M[(fg_i)^{-1}]$}
+  \end{array} \]
+  The generic prime filter~$\fff_0$ can also be described in explicit
   terms. For ring elements~$f$ and~$s$, $D(f) \models (s \in \fff_0)$ iff~$f
@@
 
-  \jnote{2-3}{Our description of~$M^\sim$ reveals a precise sense in
+  \jnote{2}{Our description of~$M^\sim$ reveals a precise sense in
   which~$M^\sim$ and~$M$ are related: $M^\sim$ is simply a localization of~$M$
   (first lifted to another universe by the constant sheaf construction). The
-  classical descriptions don't make the relation evident.}
+  classical descriptions don't make the relation evident.
+
+  As a first approximation, the module~$M^\sim$ can be thought of as a
+  reification of all the stalks of~$M$ as a single object. The metatheorem
+  displayed at the top left on the next slides makes this precise and also
+  exposes the limits of this view: It is only correct for geometric sequents.
+  When considering nongeometric sequents, phenomena appear which are unique
+  to~$M^\sim$ in the sense that they are in general not shared by~$M$, its
+  stalks or its quotients.}
 
@@
   satisfied if any element of~$A$ is nilpotent or not nilpotent, so it's always
-  satisfied if the law of excluded middle is available. In an intuitionistic
-  context, it's still ``morally satisfied''. Details are in Section~12.9 of
+  satisfied if the law of excluded middle is available in the metatheory. In an
+  intuitionistic context, it's still ``morally satisfied''. Details are in
+  Section~12.9 of
   \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
-  notes}.}
-
-  \jnote{3}{As a first approximation, the module~$M^\sim$ can be thought of as
-  a reification of all the stalks of~$M$ as a single object. The metatheorem
-  makes this precise and also shows the limits of this view: It is only correct
-  for geometric sequents. When considering nongeometric sequents, phenomena
-  appear which are unique to~$M^\sim$ in the sense that they are in general not
-  shared by~$M$, its stalks or its quotients.}
+  notes}.
+
+  The metatheorem soups up a number of lemmas of algebraic geometry, there
+  stated in geometric language. For instance, if~$M$ is finitely generated,
+  then~$M^\sim$ is of finite type. If~$M$ is finitely presented, then~$M^\sim$
+  is of finite presentation. If~$M$ is coherent, then~$M^\sim$ is coherent.}
 
@@
 
-  \jnote{1}{XXX ... succeeded ...}
+  \jnote{1}{This slide delivers on the promise made earlier: Using the internal
+  language of the little Zariski topos, we can reduce to the case of fields
+  without having to employ maximal ideals, prime ideals or minimal prime ideals.
+  Since the internal language machinery is itself constructive, the displayed
+  proofs can be unwinded to yield external constructive proofs which don't
+  reference topos theory. Details on these external proofs can soon be found in
+  a forthcoming paper titled
+  \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-wlog.pdf}{Without
+  loss of generality, any reduced ring is a field}.}
+
+  \jnote{2}{The testcase of Grothendieck's generic freeness lemma illustrates
+  that the internal language of toposes can help commutative algebra even if
+  one is not interested in constructivity issues. The previously known proofs
+  are somewhat long and somewhat convoluted; the new proof is arguably short
+  and simple.
+
+  Details on the internal proof are in Section~11.5 of
+  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
+  notes}. The
+  \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-generic-freeness.pdf}{external
+  proof} obtained by unwinding the internal one is still quite direct, compared
+  to the previously published proofs, and interestingly follows a curious
+  course: It starts with verifying, in an inductive manner, that~$B$ and~$M$
+  are free; that~$A \to B$ is of finite presentation; and that~$M$ is finitely
+  presented as a~$B$-module. Then the assumption for~$f = 1$ is used, rendering
+  the prior steps moot, since over the zero ring any module is free and any
+  algebra is of finite presentation. External translations of internal proofs
+  which use double negation will always take such a course.}
+
+  \jnote{3}{XXX}
 

### 2018-07-04 paper-generic-freeness.tex
typo
Improvement suggested by Yakov Varshavsky

@@
 version in a paper of his~\cite{brandenburg:schur}. We are grateful to Thierry
-Coquand and Peter Schuster for valuable advice, to Giuseppe Rosolini for
-comments regarding the presentation of the paper, and to Marc
+Coquand and Peter Schuster for valuable advice, to Giuseppe Rosolini and to
+Yakov Varshavsky for
+comments improving the presentation of the paper, and to Marc
 Nieper-Wißkirchen for carefully guiding our PhD studies~\cite{blechschmidt:phd} at the University of
@@
 an~$A[a_i^{-1}]$-module by fewer than~$m$ generators. The induction hypothesis,
-applied to this module, yields that~$1 = 0$ in~$A[a_i^{-1}]$. Since~$A$ is
+applied to this module over~$A[a_i^{-1}]$, yields that~$1 = 0$ in~$A[a_i^{-1}]$. Since~$A$ is
 reduced, this amounts to~$a_i = 0$.
@@
 situation over the restriction to~$D(a_J)$. Whereas before this type of argument was
-powered by the reducedness assumption on~$A$, it is not supported by the
+powered by the reducedness assumption on~$A$, it is now supported by the
 assumption on~$X$.

### 2018-07-04 slides-como2018.tex
More notes

@@
 \newcommand{\seq}{\vdash_{\!\!\!\vec x}}
+\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates
 
@@
   then~$M^\sim$ is of finite type. If~$M$ is finitely presented, then~$M^\sim$
-  is of finite presentation. If~$M$ is coherent, then~$M^\sim$ is coherent.}
+  is of finite presentation. If~$M$ is coherent, then~$M^\sim$ is coherent.
+
+  As an aside, the little Zariski topos is seldomly Boolean (validates the law of
+  excluded middle). A necessary condition is that~$A$ is of dimension~$\leq 0$.}
 
@@
 
-  \jnote{3}{XXX}
+  \jnote{3}{Here's a rough sketch why the double negations occur. In
+  undergraduate linear algebra, we learn that any finitely generated vector
+  space is free (has a basis), by the following argument:
+  Let~$(v_1,\ldots,v_n)$ be a generating family. Either one of the generators can
+  be expressed as a linear combination of the others or not. In the latter
+  case, the family is linearly independent and we're done. In the former, we
+  remove the redundant generator and continue by induction.
+
+  This argument relies on the law of excluded middle and can therefore not put
+  to work in the internal universe as is. However, intuitionistically we do
+  have the weaker statement~$\neg\neg(\varphi \vee \neg\varphi)$, and we can
+  use this law in place of the law of excluded middle to intuitionistially
+  deduce that any finitely generated vector space does \emph{not not} have a
+  basis.}
 
@@
 
+  \jnote{1}{At the
+  \fixedhref{https://sbseminar.wordpress.com/2009/08/06/algebraic-geometry-without-prime-ideals/}{Secret
+  Blogging Seminar}, there was an insightful long-running discussion on the
+  merits of the functorial approach. Two disadvantages of the approach using
+  locally ringed spaces is that the underlying topological spaces don't
+  actually parametrize ``honest'', ``geometric'' points, but irreducible closed
+  subsets; and that they don't work well in a constructive setting. (For this,
+  they would have to be replaced by locally ringed locales.)
+
+  Given a functor~$F : \mathrm{Ring} \to \Set$, we imagine~$F(A)$ to be the set
+  of~``$A$-valued points'' of the hypothetical scheme described by~$F$, the set
+  of ``points with coordinates in~$A$''. These sets are geometrically important.
+
+  The functorial approach is more economical, philosophically rewarding, and
+  works constructively. However, typically only field-valued points are easy
+  to describe. For instance, the functor representing projective~$n$-space is
+  given on fields by
+  \[ \begin{array}{r@{}c@{}l}
+  K &{}\longmapsto{}& \text{the set of lines through the origin in~$K^{n+1}$} \\
+  && \qquad \cong \{ [x_0 \hg \cdots \hg x_n] \,|\, \text{$x_i \neq 0$ for some~$i$} \},
+  \end{array} \]
+  whereas on general rings it is given by
+  \[ \begin{array}{r@{}c@{}l}
+    A &{}\longmapsto{}& \text{the set of quotients~$A^{n+1} \twoheadrightarrow P$,
+    where~$P$ is projective,} \\
+    && \qquad \text{modulo isomorphism}.
+  \end{array} \]
+  It are these more general kinds of points which impart a meaningful sense of
+  cohesion on the field-valued points, so they can't simply be dropped from
+  consideration.}
+
+  \jnote{2}{This tension is resolved by observing that the category of
+  functors~$\Ring \to \Set$ is a topos (the big Zariski topos of~$\Spec(\ZZ)$)
+  and that we can therefore employ its internal language. This language takes
+  care of managing stages behind the scenes. For instance, projective~$n$-space
+  can be described by the naive expression
+  \[ \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee \cdots \vee x_n
+  \neq 0 \}/(\affl)^\times. \]
+  This example illustrates our goal: to develop a synthetic account of
+  algebraic geometry, in which schemes are plain sets and morphisms between
+  schemes are maps between those sets. It turns out that there are many
+  similarities with the well-developed synthetic account of differential
+  geometry, but also important differences.}
+
+  \pause
+
   \hil{Synthetic approach:} model schemes \hil{directly as sets} in

### 2018-07-04 slides-como2018.tex
More notes

@@
 \newcommand{\Set}{\mathrm{Set}}
+\newcommand{\Ring}{\mathrm{Ring}}
 \newcommand{\Eff}{\mathrm{Ef{}f}}
@@
     \item or Grothendieck's functor-of-points account, where a scheme is a
-    functor~$\mathrm{Ring} \to \mathrm{Set}$.
+    functor~$\Ring \to \Set$.
     \small\[ A \longmapsto \{ (x,y,z) \in A^3 \,|\, x^n+y^n-z^n=0 \} \]
@@
 
-  \jnote{1}{At the
+  \jnote{1}{\footnotesize At the
   \fixedhref{https://sbseminar.wordpress.com/2009/08/06/algebraic-geometry-without-prime-ideals/}{Secret
   Blogging Seminar}, there was an insightful long-running discussion on the
-  merits of the functorial approach. Two disadvantages of the approach using
+  merits of the two approaches. Two disadvantages of the approach using
   locally ringed spaces is that the underlying topological spaces don't
@@
 
-  Given a functor~$F : \mathrm{Ring} \to \Set$, we imagine~$F(A)$ to be the set
-  of~``$A$-valued points'' of the hypothetical scheme described by~$F$, the set
-  of ``points with coordinates in~$A$''. These sets are geometrically important.
-
   The functorial approach is more economical, philosophically rewarding, and
-  works constructively. However, typically only field-valued points are easy
+  works constructively. Given a functor~$F : \Ring \to \Set$, we imagine~$F(A)$
+  to be the set of~``$A$-valued points'' of the hypothetical scheme described
+  by~$F$, the set of ``points with coordinates in~$A$''. These sets have direct
+  geometric meaning. However, typically only field-valued points are easy
   to describe. For instance, the functor representing projective~$n$-space is
@@
   \jnote{2}{This tension is resolved by observing that the category of
-  functors~$\Ring \to \Set$ is a topos (the big Zariski topos of~$\Spec(\ZZ)$)
+  functors~$\Ring \to \Set$ is a topos (the \emph{big Zariski topos} of~$\Spec(\ZZ)$)
   and that we can therefore employ its internal language. This language takes
-  care of managing stages behind the scenes. For instance, projective~$n$-space
+  care of juggling stages behind the scenes. For instance, projective~$n$-space
   can be described by the naive expression
@@
   \end{itemize}
+
+  \jnote{1}{The objects of the category~$(\Aff/S)_{\mathrm{lofp}}$ are
+  morphisms of the form~$\Spec(R) \to S$ which are locally of finite
+  presentation. (Other choices of resolving set-theoretical issues of size are
+  also possible.)
+
+  A functor~$F : (\Aff/S)_{\mathrm{lofp}}^\op \to \Set$ is a sheaf
+  for the Zariski topology if and only if the diagram
+  \[ F(T) \to
+    \prod_i F(U_i) \rightrightarrows
+    \prod_{j,k} F(U_j \cap U_k) \]
+  is a limit diagram for any open covering~$T = \bigcup_i U_i$ of any
+  scheme~$T \in (\Aff/S)_{\mathrm{lofp}}$.
+
+  In the case that~$S = \Spec(A)$ is affine, the big Zariski topos of~$S$ is also
+  simply called ``big Zariski topos of~$A$''. It is a subtopos of the topos of
+  functors $\mathrm{Alg}(A) \to \Set$ and classifies local~$A$-algebras.
+
+  From the internal point of view of~$\Sh(S)$, the sheaf~$\O_S$ of rings is
+  just an ordinary ring, and one can construct internally to~$\Sh(S)$ the big
+  Zariski topos of~$\O_S$. Externally, this construction will yield a certain
+  bounded topos over~$\Sh(S)$. However, as indicated on the slide, this topos
+  will \emph{not} coincide with the true big Zariski topos of~$S$. To
+  construct the true big Zariski topos, one has to build, internally
+  to~$\Sh(S)$, the classifying topos of local \emph{and local-over-$\O_S$}
+  $\O_S$-algebras.}
 \end{frame}

### 2018-07-06 slides-como2018.tex
Notes finished

@@
 \newcommand{\Ring}{\mathrm{Ring}}
+\newcommand{\LocRing}{\mathrm{LocRing}}
 \newcommand{\Eff}{\mathrm{Ef{}f}}
@@
     \end{column}
+
     \begin{column}{0.33\textwidth}
@@
   no advanced prerequisites in commutative algebra, and takes
-  \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-generic-freeness.pdf}{about
+  \fixedhref{https://arxiv.org/abs/1807.01231}{about
   a page}.}
@@
   notes}. The
-  \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-generic-freeness.pdf}{external
+  \fixedhref{https://arxiv.org/abs/1807.01231}{external
   proof} obtained by unwinding the internal one is still quite direct, compared
@@
   \end{varblock}
+
+  \jnote{1}{One doesn't need to be an expert in topos theory in order to know
+  that many notions in algebraic geometry are inspired by notions in algebra
+  and that proofs in algebraic geometry often proceed by reducing to algebra.
+  The internal language is a way of making this connection precise: In many
+  cases, the former are simply interpretations of the latter internal
+  to~$\Sh(X)$. Because this connection is precise instead of informal,
+  additional value is gained: We can skip many basic proofs in algebraic
+  geometry because they're just externalizations of proofs in algebra carried
+  out internally to~$\Sh(X)$.}
+
+  \jnote{2}{A basic example is as follows. A short exact sequence of sheaves of
+  modules looks like a short exact sequence of plain modules from the internal
+  point of view of~$\Sh(X)$. If the two outer sheaves are of finite type, then
+  from the internal point of view, the two outer modules will look like
+  finitely generated modules. Because the standard proof of the proposition
+  quoted on the lower right is intuitionistically valid, it follows that, from
+  the internal point of view, the middle module is too finitely generated.
+  Consulting the dictionary a second time, this amounts to saying that the
+  middle sheaf is of finite type.
+
+  More details on this research program can be found in
+  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
+  notes}, partly reported on at the
+  \fixedhref{https://www.youtube.com/watch?v=7S8--bIKaWQ}{2015 IHÉS
+  conference}. Even though many important dictionary entries
+  are still missing (for instance pertaining derived categories and
+  intersection theory), I believe that it is already at its current stage
+  useful to working algebraic geometers.
+
+  The next few slides illustrate an entirely different way of approaching
+  algebraic geometry using topos theory.}
 
@@
   locally ringed spaces is that the underlying topological spaces don't
-  actually parametrize ``honest'', ``geometric'' points, but irreducible closed
-  subsets; and that they don't work well in a constructive setting. (For this,
-  they would have to be replaced by locally ringed locales.)
+  actually parametrize ``honest'', ``geometric'' points, but the more complex
+  notion of irreducible closed subsets; and that they don't work well in a
+  constructive setting. (For this, they would have to be replaced by locally
+  ringed locales.)
 
@@
   \end{array} \]
-  It are these more general kinds of points which impart a meaningful sense of
+  It is these more general kinds of points which impart a meaningful sense of
   cohesion on the field-valued points, so they can't simply be dropped from
@@
   synthetic affine schemes.
+
+  \jnote{1}{In the internal universe of the big Zariski topos of a base
+  scheme~$S$, $S$-schemes can simply be modeled by sets (enjoying the special
+  property that, in a certain precise sense, they are locally affine). This
+  slide expresses some of the basic constructions of~$S$-schemes in that
+  language.
+
+  Particularly nice are the following items.
+  \begin{itemize}\justifying
+    \item Projective~$n$-space can be given by the any of the two quite naive
+    expressions displayed on the slide.
+    \item Let~$X$ be an~$S$-scheme. We often think about a sheaves
+    of~$\O_X$-modules over~$X$ by their fibers; but for a rigorous
+    treatment in the standard foundations, we have to take the full sheaf
+    structure into account; the fibers do not determine a sheaf
+    uniquely.
+
+    From the internal point of view of~$\Zar(S)$, a sheaf of~$\O_X$-modules is
+    indeed simply a family of~$\affl$-modules, one~$\affl$-module for each
+    element of~$\ull{X}$. The slide illustrates how we can define the Serre
+    twisting sheaves in this language.
+  \end{itemize}}
+
+  \jnote{2}{\begin{itemize}\justifying
+    \item The spectrum of an~$\affl$-algebra can be given by the naive
+    expression displayed on the slide. It looks like this expression can't be
+    right, ignoring any non-maximal ideals; however, it is.
+    \item The big Zariski topos of an~$S$-scheme~$X$ is, from the internal
+    point of view of~$\Zar(S)$, simply the slice topos~$\Set/\ull{X}$. Hence to
+    give an~$X$-scheme simply amounts to giving an~$\ull{X}$-indexed family of
+    sets.
+  \end{itemize}
+
+  Synthetic algebraic geometry has been developed up to the point of étale
+  geometric morphisms. Much remains to be done: For instance, as of yet there
+  is only an account of Čech methods for computing cohomology, there is not yet
+  a synthetic treatment of true cohomology. Derived categories and intersection
+  theory are also missing.}
 \end{frame}
@@
     \item $\Zar(A)$ is the \hil{lax pullback}~$(\Set
-    \Rightarrow_{\Set[\mathrm{Ring}]} \Set[\mathrm{LocRing}])$.
+    \Rightarrow_{\Set[\Ring]} \Set[\LocRing])$.
   \end{itemize}
+
+  \jnote{1}{Let~$A$ be a ring. By definition, we obtain a geometric
+  morphism~$\Set \to \Set[\Ring]$ into the classifying topos of rings.
+  There is also a geometric morphism~$\Set[\LocRing] \to \Set[\Ring]$, obtained
+  by realizing that any local ring is in particular a ring. These morphisms fit
+  together in a lax pullback square as follows:
+  \[ \xymatrix{
+    \Zar(A) \ar[r] \ar[d] & \Set[\LocRing] \ar[d] \\
+    \Set \ar[r] \ar@{}[ur]^(.3){}="a"^(.7){}="b" \ar@{=>} "a";"b" & \Set[\Ring]
+  } \]
+  This observation is joint with Peter Arndt and Matthias Hutzler.
+
+  Incidentally, the pseudo pullback of the morphism~$\Set[\LocRing] \to
+  \Set[\Ring]$ along~$\Set \to \Set[\Ring]$ is not very interesting: It's the
+  largest subtopos of~$\Set$ where~$A$ is a local ring. Assuming the law of
+  excluded middle, this subtopos is either the trivial topos (if~$A$ is not
+  local) or~$\Set$ (if~$A$ is local).
+
+  There is also a way of realizing the little Zariski topos of~$A$ as a pseudo
+  pullback, exploiting that the (localic) spectrum construction is geometric.
+  See Section~12.6 of
+  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
+  notes} for details.
+  }
 \end{frame}
@@
   \end{itemize}
+
+  \jnote{1}{The axiom of microaffinity is a special instance of the
+  \emph{Kock--Lawvere axiom} known from synthetic differential geometry. We'll
+  see on the next slide that~$\affl$ validates an unusually strong form of the
+  Kock--Lawvere axiom, not at all satisfied in the usual well-adapted models of
+  synthetic differential geometry.
+
+  The fact that, internally to~$\Zar(S)$, any map~$\affl \to \affl$ is a
+  polynomial can be seen as a formal version of the general motto that in
+  algebraic geometry, ``morphisms are polynomials''.}
 \end{frame}
@@
 
-% \emph{Example.} Let~$x \? \affl$ such that~$x \neq 0$. Set~$R = \affl/(x)$.
-% Then~$\Spec(R) = \emptyset$. Thus~$(\affl)^{\Spec(R)}$ is a singleton.
-% Hence~$R = 0$. Therefore~$x$ is invertible.
+  \jnote{1}{Let~$R$ be an~$\affl$-algebra. An element~$f \in R$ induces
+  an~$\affl$-valued function on~$\Spec(R)$; functions of this form can
+  reasonably be called ``algebraic''. In a synthetic context, there should be
+  no other~$\affl$-valued functions on~$\Spec(R)$ as these algebraic ones, and
+  different algebraic expressions should yield different functions. This is
+  precisely what the bijectivity of the displayed map expresses (in a positive
+  way).
+
+  In synthetic differential geometry, the closest cousin of synthetic algebraic
+  geometry, the analogue of the displayed map is only bijective for Weil
+  algebras such as~$\affl[X]/(X^2)$ or $\affl[X,Y]/(X^2,XY)$, not for arbitrary
+  finitely presented~$\affl$-algebras. This is a major difference to synthetic
+  differential geometry.}
+
+  \jnote{2}{The notion of synthetic quasicoherence is interesting for a number
+  of reasons:\vspace*{-1.2em}
+  \begin{itemize}\justifying
+    \item All currently known properties of~$\affl$, such as all the properties
+    listed on the previous slide, follow from the statement that~$\affl$ is
+    synthetically quasicoherent.
+
+    For instance, here is how we can verify the field property. Let~$x \?
+    \affl$ such that~$x \neq 0$. Set~$R = \affl/(x)$. Then~$\Spec(R) =
+    \emptyset$. Thus~$(\affl)^{\Spec(R)}$ is a singleton. Hence~$R = 0$.
+    Therefore~$x$ is invertible.
+    \item Given an~$\affl$-module~$E$, we can formulate the following variant
+    of the axiom of synthetic quasicoherence: ``For any finitely
+    presented~$\affl$-algebra~$R$, the canonical map~$R \otimes_\affl E \to
+    E^{\Spec(R)}$ is bijective.'' This axiom is satisfied if and only if~$E$ is
+    induced by a quasicoherent sheaf of~$\O_S$-modules.
+    \item The notion of synthetic quasicoherence is central to synthetic
+    algebraic geometry. The notions of synthetic open immersions, closed
+    immersion, schemes and several others all refer to synthetic
+    quasicoherence.
+  \end{itemize}}
+
+  \jnote{3}{An anlogue of synthetic quasicoherence holds in the classifying
+  topos of rings, demonstrating that even presheaf toposes can validate
+  interesting nontrivial nongeometric sequents.
+
+  We believe that an analogue of synthetic quasicoherence holds for the generic
+  model of any geometric theory. This is work in progress. If true, this would
+  yield a major source of nongeometric sequents in classifying toposes. Because
+  of the many applications on nongeometric sequents, it's very desirable to
+  possess such a source.
+
+  The mystery of nongeometric sequents is this: On the one hand, they are very
+  useful to have because of surprising applications; on the other hand, they
+  are as of yet quite elusive.}
 \end{frame}
 
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
 \begin{frame}{Classifying toposes in algebraic geometry}
@@
     \toprule
-    (Big) topos & classified theory \\\midrule
+    (big) topos & classified theory \\\midrule
     Zariski & local rings [Hakim 1972] \\[0.6em]
@@
     \textcolor{red!80}{$\neg\neg$} & \textcolor{red!80}{?? (conjecturally: algebraically closed geometric} \\
-    & \qquad \textcolor{red!80}{fields are integral over the base)} \\[0.6em]
+    & \qquad \textcolor{red!80}{fields which are integral over the base)} \\[0.6em]
     \usebeamercolor[fg]{item}{infinitesimal} & \usebeamercolor[fg]{item}{local algebras together with a nilpotent ideal [Hutzler 2018]} \\[0.6em]
@@
   \end{tabular}
-\end{frame}
+
+  \jnote{1}{Toposes, and also more specifically classifying toposes, originated
+  in algebraic geometry. It is therefore deeply embarassing that as of now,
+  still very little is known about the theories classified by the major toposes
+  in active use by algebraic geometers.
+  \begin{center}
+    \includegraphics[width=0.99\textwidth]{wraith-generic-galois-theory} \\
+    \footnotesize
+    Gavin Wraith. Generic Galois theory of local rings. 1979.
+  \end{center}}
+
+  \jnote{2}{For almost forty years, only the big Zariski topos and its étale
+  subtopos were understood in that way.
+  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{These~2017
+  notes} answer the question for the fppf topology and the surjective topology
+  (in Section~21) and state conjectures for the ph topology and the
+  double negation topology. However, while good to have, the answer for the
+  fppf topology remains unsatisfactory, since Wraith's conjecture that the fppf
+  topos classifies the simpler theory of algebraically closed local rings has
+  neither been confirmed nor refuted.
+
+  A couple of weeks ago, Matthias Hutzler managed to determine the theory
+  classified by the big infinitesimal topos of a ring~$A$: It classifies
+  pairs~$(B,\aaa)$ consisting of a local~$A$-algebra~$B$ and a nilpotent
+  ideal~$\aaa \subseteq B$. Details will be in his forthcoming Master's thesis.
+  He is currently working on answering the question for the closely related big
+  crystalline topos.
+
+  It will be exciting to learn what the crystalline topos and the many other
+  toposes in algebraic geometry classify; how algebraic geometry can profit
+  from these discoveries; and which new flavors of synthetic algebraic geometry
+  they unlock.}
+\end{frame}}
 
@@
 
-\begin{frame}{The Kripke--Joyal semantics for the little Zariski topos}
-  Recall~$A[f^{-1}] = \bigl\{ \frac{u}{f^n} \,|\, u \in A, n \in \NN \bigr\}$.
-
-  \begin{itemize}
-    \item $\Sh(\Spec(A)) \models \text{``For all~$x \in A^\sim$, \ldots''}$
-
-          Meaning: For all~$f \in A$ and all~$x \in A[f^{-1}]$, \ldots
-          \medskip
-
-    \item $\Sh(\Spec(A)) \models \text{``There is~$x \in A^\sim$ such that \ldots''}$
-
-          \mbox{Meaning: There is a partition of unity,~$1 = f_1 + \cdots + f_n \in A$,}
-          such that for each~$i$, there exists~$x_i \in A[f_i^{-1}]$
-          with~\ldots
-          \medskip
-
-    \item $\Sh(\Spec(A)) \models \text{``$\varphi$ implies $\psi$''}$
-
-          Meaning: For all~$f \in A$, if~$\varphi$ on stage~$f$, then~$\psi$ on
-          stage~$f$.
-  \end{itemize}
-\end{frame}
-
 \end{document}

### 2018-07-06 paper-flabby-objects.tex
typo

@@
   We introduce a general notion of \emph{flabby objects} in elementary toposes
-  and study its basic properties. In the special case of localic toposes, this
+  and study their basic properties. In the special case of localic toposes, this
   notion reduces to the common notion of flabby sheaves, yielding a
   site-independent characterization of flabby sheaves. Continuing a line of
-  research started by Roswitha Harting, we use this notion to
+  research started by Roswitha Harting, we use flabby objects to
   show that an internal notion of injective objects coincides with the
@@
 \noindent
+
+% XXX: In Einleitung interne Sprache erinnern.
 
@@
 elementary toposes; but for toposes which are not localic, the two resulting
-notions will differ, and only the one induced by the alternative definition
+notions will differ, and only the one obtained from the alternative definition
+is stable under pullback and
 can be characterized in the internal language. We therefore adapt in this paper the
@@
 Let~$F$ be a sheaf on a topological space~$X$ (or a locale).
-Then~$F$ is flabby as a sheaf if and only if~$F$ is a flabby object of the
+Then~$F$ is flabby as a sheaf if and only if~$F$ is flabby as an object of the
 sheaf topos~$\Sh(X)$.
@@
 Let~$X$ be a flabby object in an elementary topos~$\E$. If~$\E$ is localic and
-Zorn's lemma is available, then~$X$ possesses a global element (a morphism~$1 \to X$).
+Zorn's lemma is available in the metatheory, then~$X$ possesses a global element (a morphism~$1 \to X$).
 \end{prop}
@@
 notion~\cite{XXX}, as she
-(mistakingly) states in~\cite[page~233]{harting:remark} that ``the notions of
+(mistakenly) states in~\cite[page~233]{harting:remark} that ``the notions of
 injectivity and internal injectivity do not coincide'' for modules.
-% XXX "mistakenly"? comparison with Harting's proof (Barr's lemma...)
+% XXX comparison with Harting's proof (Barr's lemma...)
 
@@
 \item Any object can be embedded into a flabby object.
-\item The terminal object is flabby. The product of flabby objects if flabby.
+\item The terminal object is flabby. The product of flabby objects is flabby.
 \item Let~$I$ be an internally injective object. Let~$T$ be an arbitrary
@@
 example is the~$G$-set~$G$ (with the translation action), considered as an
-object of the topos~$\PSh(BG)$ of~$G$-sets, where~$G$ is a nontrivial group.
+object of the topos~$BG$ of~$G$-sets, where~$G$ is a nontrivial group.
 This object is not strongly flabby, since the morphism~$\emptyset \to G$ does
-not lift, but its pullback to the slice~$\PSh(BG)/G \simeq \Set$ is (assuming
+not lift, but its pullback to the slice~$BG/G \simeq \Set$ is (assuming
 the law of excluded middle).
@@
 establishing the following observations on flabby objects in the effective
-topos.
+topos. We follow the terminology of~\cite{hyland:effective-topos}.
 
@@
 
-We follow the terminology of~\cite{hyland:effective-topos}(XXX). The intuitive
+The intuitive
 reason for why Proposition~\ref{prop:flabby-effective-sets} holds is the
-following. Let~$X$ be a flabby object of the effective topos. Then there is a
+following. Let~$X$ be a flabby object in the effective topos. Then there is a
 procedure which computes for any subterminal~$K \subseteq X$ an element~$x_K$
 such that~$K \subseteq \{ x_K \}$. This element might not depend extensionally
-on~$K$, but this fine point is not important for the further discussion. Let
+on~$K$, but this fine point is not important for this discussion. Let
 now~$f : X \to X$ be a morphism. We construct the self-referential subset~$K
@@
 This problem does not manifest with objects~$X$ which are not effective sets.
-Reflexivity realizers for these objects may not be very informative themselves;
+Reflexivity realizers for these objects are themselves not very informative;
 a procedure witnessing flabbiness therefore only has to turn one kind of
@@
 the proof of Proposition~\ref{prop:semienough-flabby-modules} will accordingly
-be not effective sets.
+not be effective sets.
 
@@
 + rs$, $(r+r')t \sim rt + r't$ and such
-that~$\underline{x+y} \sim \underline{x} + \underline{y}$, $\underline{rx} \sim
-r \underline{x}$, $\underline{0} \sim 0$ and such that~$\varepsilon_{\{t\}} \sim
+that~$\underline{0} \sim 0$, $\underline{x+y} \sim \underline{x} + \underline{y}$, $\underline{rx} \sim
+r \underline{x}$, and such that~$\varepsilon_{\{t\}} \sim
 t$ for all terms~$t$. Then the quotient~$T/{\sim}$ is a flabby~$R$-module
@@
 least~$\neg\neg$-separated modules in the effective topos always embed into
-flabby modules, assuming the law of excluded middle, while
+flabby modules, assuming the law of excluded middle in the metatheory, while
 Proposition~\ref{prop:flabby-effective-sets} shows that no nontrivial effective

### 2018-07-07 paper-macneille-reals.tex
Begin of the proof of the uncountability of the MacNeille reals

--- 2018-07-03 paper-generic-freeness.tex
@@
 
-\usepackage[natbib=true,style=numeric,maxnames=10]{biblatex}
-\usepackage[babel]{csquotes}
-\bibliography{paper-generic-freeness.bib}
+\newcommand{\NN}{\mathbb{N}}
+\newcommand{\RR}{\mathbb{R}}
 
-\title[Grothendieck's generic freeness lemma]{An elementary and constructive
-proof of Grothendieck's generic freeness lemma}
-\author{Ingo Blechschmidt}
-\address{Max Planck Institute for Mathematics in the Sciences \\
-Inselstraße 22 \\
-04103 Leipzig, Germany}
-\email{ingo.blechschmidt@mis.mpg.de}
+\title{A constructive Knaster--Tarski proof of the uncountability of the reals}
+\author{Ingo Blechschmidt and Matthias Hutzler}
+%\address{Max Planck Institute for Mathematics in the Sciences \\
+%Inselstraße 22 \\
+%04103 Leipzig, Germany}
+\email{ingo.blechschmidt@mis.mpg.de, matthias.matthias.ralph.hutzler@student.uni-augsburg.de}
 
@@
 \begin{abstract}
-  We present a new and direct proof of Grothendieck's generic freeness
-  lemma in its general form. Unlike the previously published proofs, it does not
-  proceed in a series of reduction steps and is fully constructive, not
-  using the axiom of choice or even the law of excluded middle. It was
-  found by unwinding the result of a general topos-theoretic technique.
+  Leaning heavily on Levy's unusual proof of the uncountability of the reals,
+  we give an uncountability proof of the (MacNeille) reals which ... XXX
 \end{abstract}
@@
 
-\noindent
-We prove Grothendieck's generic freeness lemma in the
-following form.
+\begin{thm}Let~$f : \NN \to \RR$ be a map. Then there is a number~$x_0 \in \RR$
+such that for no~$n \in \NN$, $f(n) = x_0$.\end{thm}
 
-\begin{thm}\label{thm:algebraic}Let~$A$ be a reduced ring (commutative, with unit). Let~$B$ be
-an~$A$-algebra of finite type. Let~$M$ be a finitely generated~$B$-module.
-If~$f = 0$ is the only element of~$A$ such that
-\begin{enumerate}
-\item the~$A[f^{-1}]$-modules $B[f^{-1}]$ and $M[f^{-1}]$ are free,
-\item the~$A[f^{-1}]$-algebra~$B[f^{-1}]$ is of finite presentation and
-\item the~$B[f^{-1}]$-module~$M[f^{-1}]$ is finitely presented,
-\end{enumerate}
-then~$1 = 0$ in~$A$.
-\end{thm}
+\begin{proof}The map
+\[ g : \RR \longrightarrow \RR,\
+  x \longmapsto \sup_M \sum_{n \in M} 2^{-n}, \]
+where~$M$ ranges over all those finite detachable subsets of~$\NN$ such
+that~$f[M] < x$, is well-defined, monotone, has a postfixpoint ($0 \leq g(0)$),
+and has an upper bound for all of its postfixpoints (if~$x \leq g(x)$, then~$x
+\leq 2$). By the Knaster--Tarski fixed point theorem, it therefore has a
+greatest postfixpoint~$x_0$.
 
-Previously known proofs either only cover the case where~$A$ is a Noetherian
-integral domain, where one can argue by \emph{dévissage} (see for
-instance~\cite[Lemme~6.9.2]{ega-4-2},
-\cite[Thm.~24.1]{matsumura:commutative-ring-theory} or
-\cite[Thm.~14.4]{eisenbud:commutative-algebra}), or proceed in a series of
-intermediate steps, reducing to that case (see for
-instance~\cite{staats:generic-freeness} or~\stacksproject{051Q}); but in fact,
-a direct proof is possible and shorter. The new proof unveils a certain
-combinatorial aspect to Grothendieck's generic freeness lemma, does not require
-any advanced prerequisites in commutative algebra and does not use the axiom of
-choice or the law of excluded middle. It is purely element-based, not referring
-to ideals of~$A$, and doesn't use Noether normalization.
-
-Grothendieck's generic freeness lemma is often presented in contrapositive form
-or in the following geometric variant:
-
-\begin{thm}\label{thm:geometric}Let~$A$ be a reduced ring. Let~$B$ be
-an~$A$-algebra of finite type. Let~$M$ be a finitely generated~$B$-module. Then
-the space~$\Spec(A)$ contains a dense open~$U$ such that over~$U$,
-\begin{enumerate}
-\item[(a)] $B^\sim$ and~$M^\sim$ are locally free as sheaves of~$A^\sim$-modules,
-\item[(b)] $B^\sim$ is of finite presentation as a sheaf of~$A^\sim$-algebras and
-\item[(c)] $M^\sim$ is finitely presented as a sheaf of~$B^\sim$-modules.
-\end{enumerate}
-\end{thm}
-
-Theorem~\ref{thm:geometric} immediately follows from
-Theorem~\ref{thm:algebraic} by defining~$U$ as the union of all those basic
-opens~$D(f)$ such that~(1),~(2) and~(3) hold. It is clear that~(a),~(b)
-and~(c) hold over~$U$, and~$U$ is dense for if~$V$ is an arbitrary open
-such that~$U \cap V = \emptyset$, the open~$V$ is itself empty: Let~$h \in A$ be
-such that~$D(h) \subseteq V$. The hypothesis implies the assumptions of
-Theorem~\ref{thm:algebraic} for the datum~$(A[h^{-1}], B[h^{-1}], M[h^{-1}])$.
-Thus~$1 = 0 \in A[h^{-1}]$, so~$h$ is nilpotent and~$D(h) = \emptyset$.
-
-The new proof was found using a general topos-theoretic technique which we
-believe to be useful in other situations as well. This technique allows to view
-reduced rings and their modules from a different point of view, one from which
-reduced rings look like fields. Since Grothendieck's generic freeness
-is trivial for fields, this technique yields a trivial proof for reduced rings.
-The proof presented here was obtained by unwinding the topos-theoretic proof,
-yielding a self-contained argument without any references to topos theory.
-We refer readers who want to learn about this technique to a forthcoming companion 
-paper~\cite{blechschmidt:wlog}.
-
-\textbf{Acknowledgments.} The proof presented here was prompted by a question by
-user~HeinrichD on MathOverflow~\cite{mo:kernel} and greatly
-benefited from discussions with Martin Brandenburg, who employed the constructive
-version in a paper of his~\cite{brandenburg:schur}. We are grateful to Thierry
-Coquand and Peter Schuster for valuable advice, to Giuseppe Rosolini for
-comments regarding the presentation of the paper, and to Marc
-Nieper-Wißkirchen for carefully guiding our PhD studies~\cite{blechschmidt:phd} at the University of
-Augsburg, where most of the work for this paper was carried out.
-
-\section{The proof of the finitely-generated case}
-
-The following proposition is just a special instance of Grothendieck's generic
-freeness lemma. Its proof is easier and shorter than the proof of the general
-case, which is why we present it here. The general proof
-will not refer to this one.
-
-\begin{prop}Let~$A$ be a reduced ring. Let~$M$ be a finitely
-generated~$A$-module. If~$f = 0$ is the only element of~$A$ such
-that~$M[f^{-1}]$ is a finite free~$A[f^{-1}]$-module, then~$1 = 0$ in~$A$.
-\end{prop}
-
-\begin{proof}We proceed by induction on the length of a given generating family
-of~$M$. Let~$M$ be generated by~$(v_1,\ldots,v_m)$.
-
-We show that the family~$(v_1,\ldots,v_m)$ is linearly independent. Let~$\sum_i
-a_i v_i = 0$. Over~$A[a_i^{-1}]$, the vector~$v_i \in M[a_i^{-1}]$ is a linear
-combination of the other generators. Thus~$M[a_i^{-1}]$ can be generated as
-an~$A[a_i^{-1}]$-module by fewer than~$m$ generators. The induction hypothesis,
-applied to this module, yields that~$1 = 0$ in~$A[a_i^{-1}]$. Since~$A$ is
-reduced, this amounts to~$a_i = 0$.
-
-We finish by using the assumption for~$f = 1$.
-\end{proof}
-
-We remark that the proof takes a somewhat curious course: Our goal is to
-verify~$1 = 0$, but as an intermediate step we verify that~$M$ is free, which
-after the fact will be a trivial statement. The general proof in the next
-section will have a similar style. This approach is reminiscient of
-Richman's uses of trivial rings~\cite{richman:trivial-rings}.
-
-\section{The proof of the general case}
-
-\begin{proof}[Proof of Theorem~\ref{thm:algebraic}]
-Let~$B$ be generated by~$(x_1,\ldots,x_n)$ as an~$A$-algebra and
-let~$M$ be generated by~$(v_1,\ldots,v_m)$ as a~$B$-module. We endow the sets
-\begin{align*}
-  \I &\defeq \{ (i_1,\ldots,i_n) \,|\, i_1,\ldots,i_n \geq 0 \}
-  \quad\text{and} \\
-  \J &\defeq \{ (\ell, i_1,\ldots,i_n) \,|\, 1 \leq \ell \leq m, i_1,\ldots,i_n \geq 0 \}
-\end{align*}
-with the lexicographic order. The family~$(w_J)_{J \in \J} \defeq (x_1^{i_1}
-\cdots x_n^{i_n} v_\ell)_{(\ell,i_1,\ldots,i_n) \in \J}$ generates~$M$ as
-an~$A$-module, and we will call a subfamily~$(w_J)_{J \in \J' \subseteq \J}$
-\emph{good} if and only if for all~$J \in \J$, the vector~$w_J$ is a linear
-combination of the vectors~$(w_{J'})_{J' \in \J', J' \preceq J}$,
-and if~$(\ell,i_1,\ldots,i_n) \not\in \J'$ implies~$(\ell,k_1,\ldots,k_n) \not\in
-\J'$ for all~$k_1 \geq i_1, \ldots, k_n \geq i_n$. Figure~\ref{fig:good} shows
-how a good generating family can look like.
-Similarly, we define when a subfamily of the canonical generating
-family~$(x_1^{i_1} \cdots x_n^{i_n})_{(i_1,\ldots,i_n) \in \I}$ of~$B$ is good
-(which is just the special case~$m = 1$).
-
-We then proceed by induction on the
-shapes of a given good generating family~$(w_J)_{J \in \J'}$ for~$M$ and a given
-good generating family~$(s_I)_{I \in \I'}$ for~$B$, starting with the canonical ones. It is
-reasonably obvious that this induction is well-founded; the formal statement
-that it is so is known as \emph{Dickson's Lemma} (see, for
-instance,~\cite[Thm.~1.1]{veldman:kruskal}).
-
-We show that~$(w_J)_{J \in \J'}$ is a basis of~$M$ by verifying linear
-independence. Thus let~$\sum_J a_J w_J = 0$ in~$M$. We show that all
-coefficients in this sum are zero, starting with the largest
-appearing index~$J$: In the module~$M[a_J^{-1}]$ over the localized ring~$A[a_J^{-1}]$,
-the vector~$w_J$ is a linear combination of generators with
-smaller index. Removing~$w_J = x_1^{i_1} \cdots x_n^{i_n} v_\ell$ and also all
-vectors~$x_1^{k_1} \cdots x_n^{k_n} v_\ell$ where~$k_1 \geq i_1, \ldots,
-k_n \geq i_n$, we obtain a subfamily which is still good for
-the localized module. The induction hypothesis, applied
-to~$A[a_J^{-1}]$ and its module~$M[a_J^{-1}]$, therefore implies that $A[a_J^{-1}] = 0$.
-Thus $a_J = 0$ since~$A$ is reduced.
-
-Similarly, we show that the given good generating family~$(s_I)_{I \in \I'}$ is
-a basis. Thus~$M$ and~$B$ are free over~$A$. We fix for any corner~$J$
-of~$\J'$, as indicated in Figure~\ref{fig:good},
-a way of expressing~$w_J = \sum_K a_{JK} w_K$ as a linear combination of
-generators with strictly smaller index. Let~$\widehat w_{(\ell,i_1,\ldots,i_n)}
-\defeq x_1^{i_1} \cdots x_n^{i_n} V_\ell$ in the free~$B$-module~$B\langle
-V_1,\ldots, V_m\rangle$. The canonical map
-\[ \widehat M \defeq B\langle V_1,\ldots, V_m\rangle/
-  (\widehat w_J - \textstyle\sum_K a_{JK} \widehat w_K)_{\text{$J$ corner of~$\J'$}} \longrightarrow M
-\]
-is trivially well-defined and surjective. It is also injective, since any
-element of~$\widehat M$ can be written as an~$A$-linear combination of the
-vectors~$(\widehat w_J)_{J \in \J'}$ by employing the corner relations a finite
-number of times.  Therefore~$M$ is finitely presented as a~$B$-module.
-
-In a similar vein, a quotient algebra of~$A[X_1,\ldots,X_n]$, where
-we mod out by a suitable ideal with as many generators as corners of~$\I'$, is
-isomorphic to~$B$. Thus~$B$ is finitely presented as an~$A$-algebra.
-
-We finish by using the assumption for~$f = 1$.
-\end{proof}
-
-\begin{figure}[t]
-  \captionsetup{width=.9\linewidth}
-  \begin{tikzpicture}[scale=0.9]
-    \draw[step=1cm,gray,very thin] (0,0) grid (8,8);
-
-    \draw
-      (0,8) -- (0,0) -- (8,0);
-
-    \fill[pattern=north east lines,pattern color=gray,opacity=0.5]
-      (2,8) -- (2,5) -- (5,5) -- (5,3) -- (6,3) -- (6,2) -- (8,2) -- (8,8);
-    \draw
-      (2,8) -- (2,5) -- (5,5) -- (5,3) -- (6,3) -- (6,2) -- (8,2);
-
-    \fill[fill=black] (2,5.25) -- (2,5) -- (2.25,5) -- (2.25,5.25);
-    \fill[fill=black] (5,3.25) -- (5,3) -- (5.25,3) -- (5.25,3.25);
-    \fill[fill=black] (6,2.25) -- (6,2) -- (6.25,2) -- (6.25,2.25);
-
-    \fill[fill=blue!40!white,opacity=0.5]
-      (0,8) -- (0,0) -- (4,0) -- (4,3) -- (3,3) -- (3,5) -- (2,5) -- (2,8);
-
-    \fill[fill=red!40!white,opacity=0.5]
-      (3,3) -- (4,3) -- (4,4) -- (3,4);
-
-    % !perl -we 'for my $y (reverse 0..7) { for my $x (0..7) { print % "\$x^$x\Ey^$y\$ & " } print "\\\\ \n" }'
-    \begin{scope}[yshift=4cm, xshift=4cm]
-      \matrix[matrix of nodes,nodes={inner sep=0pt,text width=0.9cm,align=center,minimum height=0.9cm}]{
-        \tiny $x^0y^7v_1$ & \tiny $x^1y^7v_1$ & \tiny $x^2y^7v_1$ & \tiny $x^3y^7v_1$ & \tiny $x^4y^7v_1$ & \tiny $x^5y^7v_1$ & \tiny $x^6y^7v_1$ & \tiny $x^7y^7v_1$ & \\
-        \tiny $x^0y^6v_1$ & \tiny $x^1y^6v_1$ & \tiny $x^2y^6v_1$ & \tiny $x^3y^6v_1$ & \tiny $x^4y^6v_1$ & \tiny $x^5y^6v_1$ & \tiny $x^6y^6v_1$ & \tiny $x^7y^6v_1$ & \\
-        \tiny $x^0y^5v_1$ & \tiny $x^1y^5v_1$ & \tiny $x^2y^5v_1$ & \tiny $x^3y^5v_1$ & \tiny $x^4y^5v_1$ & \tiny $x^5y^5v_1$ & \tiny $x^6y^5v_1$ & \tiny $x^7y^5v_1$ & \\
-        \tiny $x^0y^4v_1$ & \tiny $x^1y^4v_1$ & \tiny $x^2y^4v_1$ & \tiny $x^3y^4v_1$ & \tiny $x^4y^4v_1$ & \tiny $x^5y^4v_1$ & \tiny $x^6y^4v_1$ & \tiny $x^7y^4v_1$ & \\
-        \tiny $x^0y^3v_1$ & \tiny $x^1y^3v_1$ & \tiny $x^2y^3v_1$ & \tiny $x^3y^3v_1$ & \tiny $x^4y^3v_1$ & \tiny $x^5y^3v_1$ & \tiny $x^6y^3v_1$ & \tiny $x^7y^3v_1$ & \\
-        \tiny $x^0y^2v_1$ & \tiny $x^1y^2v_1$ & \tiny $x^2y^2v_1$ & \tiny $x^3y^2v_1$ & \tiny $x^4y^2v_1$ & \tiny $x^5y^2v_1$ & \tiny $x^6y^2v_1$ & \tiny $x^7y^2v_1$ & \\
-        \tiny $x^0y^1v_1$ & \tiny $x^1y^1v_1$ & \tiny $x^2y^1v_1$ & \tiny $x^3y^1v_1$ & \tiny $x^4y^1v_1$ & \tiny $x^5y^1v_1$ & \tiny $x^6y^1v_1$ & \tiny $x^7y^1v_1$ & \\
-        \tiny $x^0y^0v_1$ & \tiny $x^1y^0v_1$ & \tiny $x^2y^0v_1$ & \tiny $x^3y^0v_1$ & \tiny $x^4y^0v_1$ & \tiny $x^5y^0v_1$ & \tiny $x^6y^0v_1$ & \tiny $x^7y^0v_1$ & \\
-      };
-    \end{scope}
-  \end{tikzpicture}
-  \caption{\label{fig:good}A graphical depiction of a good generating family (the non-hatched cells) in
-  the special case~$n = 2, m = 1$, writing~``$x$'' and~``$y$'' for~$x_1$ and~$x_2$. The hatched cells indicate vectors which
-  have already been removed from the family. The small black squares indicate
-  \emph{corners}. If the vector in the red cell will be found to be expressible
-  as a linear combination of vectors with smaller index (blue cells), it will
-  be removed, along with the vectors in all cells to the top and to the right
-  of the red cell.}
-\end{figure}
-
-\section{Conclusion}
-
-Commutative algebra abounds with techniques which allow us to reduce quite
-general situations to easier ones. These techniques often yield short and slick
-proofs; however, they come at an expense: They are typically nonconstructive in
-nature, employing for instance the axiom of choice, and do not argue using only
-the data at hand, but using additional auxiliary objects such as maximal
-ideals. We feel that once a subject is better understood, it is desirable to
-have more informative, direct proofs available which illuminate the
-proven claims more clearly; similar as to how bijective proofs are preferred over
-calculational inductive ones in combinatorics.
-
-Let us consider as a specific example the statement that the existence of a linear
-surjection~$A^n \to A^m$ with~$n < m$ between finite free modules over an
-arbitrary ring~$A$ implies~$1 = 0 \in A$. The standard proof of this fact proceeds
-by contradiction and passes to the quotient~$A/\mmm$, where~$\mmm$ is a maximal
-ideal of~$A$, thereby reducing to the situation that the ring is a field. In
-contrast, a direct proof such as Richman's~\cite{richman:trivial-rings}
-refers only to objects mentioned in the statement itself and
-explicitly tells us how to deduce the equation~$1 = 0$ from the~$m$ equations
-which express that each basis vector of~$A^m$ has a preimage.
-
-In a similar fashion, the new proof of Grothendieck's generic freeness lemma
-explicitly tells us how to deduce~$1 = 0$ from the given conditional
-equations expressing that~$f = 0$ is the only element with properties~(1),~(2)
-and~(3). The history of Grothendieck's generic freeness lemma goes back more
-than fifty years; we are slightly surprised that a direct proof was
-discovered only now.
-
-Direct proofs sometimes generalize to new situations where the reduction techniques
-employed by more abstract proofs cannot be applied. This is the case for
-Grothendieck's generic freeness lemma, which allows for the following
-generalization:
-
-\begin{thm}\label{thm:general-generic-freeness}
-Let~$(X,\O_X)$ be a ringed space (or ringed locale, or ringed topos), such
-that for every local section~$s$ of~$\O_X$, if the only open on which~$s$ is
-invertible is the empty one, then~$s = 0$. Let~$\B$ be a sheaf
-of~$\O_X$-algebras of finite type. Let~$\E$ be a sheaf of~$\B$-modules of finite type. Then there is a dense
-open~$U$ such that over~$U$,
-\begin{enumerate}
-\item[(a)] $\B$ and~$\E$ are locally free as sheaves of~$\O_X$-modules,
-\item[(b)] $\B$ is of finite presentation as a sheaf of~$\O_X$-algebras and
-\item[(c)] $\E$ is finitely presented as a sheaf of~$\B$-modules.
-\end{enumerate}
-\end{thm}
-
-\begin{proof}Our proof of Theorem~\ref{thm:algebraic} can be easily adapted to
-this more general setting. Where that proof concludes~$a_J = 0$ for ring
-elements~$a_J$ by considering the localized situation~$A[a_J^{-1}]$, we now
-conclude~$a_J = 0$ for local sections~$a_J$ of~$\O_X$ by considering the
-situation over the restriction to~$D(a_J)$. Whereas before this type of argument was
-powered by the reducedness assumption on~$A$, it is not supported by the
-assumption on~$X$.
-We omit further details.
-\end{proof}
-
-The spectrum of a ring~$A$ is a space of the kind required by
-Theorem~\ref{thm:general-generic-freeness} if and only if~$A$ is reduced,
-thus Theorem~\ref{thm:general-generic-freeness} indeed
-generalizes Theorem~\ref{thm:geometric}. Further examples for admissible spaces
-are given by any topological, smooth or complex manifold; a corollary
-of Theorem~\ref{thm:general-generic-freeness} for these examples is that
-quotients of vector bundles, computed in the category of sheaves of modules,
-are again vector bundles after restricting to suitable dense opens.
-
-It is hard to say with certainty that a given proof does not generalize to a
-new situation; but we do not see how this could be the case for the proofs
-cited in the introduction.
-
-\printbibliography
-
-\end{document}
-
-Recall that a module is Noetherian if and only if all of its submodules are
-finitely generated. We call a module~$M$ over an~$A$-algebra~$B$ \emph{weakly
-Noetherian} (over~$A$) if and only if, for all~$B$-submodules~$U \subseteq M$
-and all~$f \in A$ the condition
-\begin{multline*}
-  \text{for all~$g \in A$, if $U[(fg)^{-1}]$ is finitely generated as
-an~$B[(fg)^{-1}]$-module,} \\
-  \text{then~$fg$ is nilpotent in~$A$} \end{multline*}
-implies that~$f$ is nilpotent. Geometrically, this expresses that all
-quasicoherent sheaves of submodules of~$M^\sim$ are, over a dense open
-of~$\Spec(A)$, of finite type over~$B^\sim$. An~$A$-algebra~$B$
-is \emph{weakly Noetherian} if and only if it is weakly Noetherian as a module
-over itself.
-
-This notion is interesting for the following reason.
-
-\begin{lemma}\label{lemma:self-noetherian}
-Let~$A$ be a reduced ring. Then~$A$, regarded as an algebra over itself, is
-weakly Noetherian.\end{lemma}
-
-\begin{proof}Let~$\aaa \subseteq A$ be an ideal. Let~$f \in A$ such that, for
-any~$g \in A$, if~$\aaa[(fg)^{-1}]$ is finitely generated, then~$fg = 0$.
-We verify that~$f = 0$.
-
-We first show that~$\aaa[f^{-1}] = (0)$. Let~$g \in \aaa$. Then~$\aaa[(fg)^{-1}] =
-(1)$ is finitely generated. Thus~$fg = 0$, hence~$g = 0$ in~$A[f^{-1}]$.
-
-Since~$\aaa[f^{-1}]$ is finitely generated, we finish by using the
-assumption for~$g = 1$.
-\end{proof}
-
-The weak Noetherian condition enjoys the same basic stability properties as the
-usual Noetherian condition:
-
-\begin{lemma}\label{lemma:noetherian-stability}
-Let~$A$ be a ring. Let~$B$ be an~$A$-algebra.
-\begin{enumerate}
-\item If~$B$ is weakly Noetherian, then so is~$B[X]$.
-\item Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence
-of~$B$-modules. Then~$M$ is weakly Noetherian if and only if~$M'$ and~$M''$
-are.
-\end{enumerate}
-\end{lemma}
-
-We can use these observations to finish the proof of Grothendieck's generic
-freeness lemma.
-
-\begin{prop}Let~$A$ be a reduced ring. Let~$B$ be an~$A$-algebra of finite type.
-If~$f = 0$ is the only element of~$A$ such that~$B[f^{-1}]$ is of finite
-presentation as an~$A[f^{-1}]$-algebra, then~$1 = 0$ in~$A$.
-\end{prop}
-
-\begin{proof}We write~$B = A[X_1,\ldots,X_n]/\aaa$. By Lemma~\ref{lemma:self-noetherian} and
-Lemma~\ref{lemma:noetherian-stability}, the~$A$-algebra~$A[X_1,\ldots,X_n]$ is
-weakly Noetherian. We can therefore conclude if we can show, for any~$g \in A$,
-that if the ideal~$\aaa[g^{-1}]$ is finitely generated, then~$g = 0$. This
-follows immediately from the assumption (for~$f = g$).
-\end{proof}
-
-\begin{prop}Let~$A$ be a reduced ring. Let~$B$ be an~$A$-algebra of finite type.
-Let~$M$ be a finitely generated~$B$-module. If~$f = 0$ is the only element
-of~$A$ such that~$M[f^{-1}]$ is finitely presented as an~$B[f^{-1}]$-module,
-then~$1 = 0$ in~$A$.
-\end{prop}
-
-\begin{proof}We write~$M = B^n/U$ for an~$B$-submodule~$U$. By Lemma~\ref{lemma:self-noetherian} and
-Lemma~\ref{lemma:noetherian-stability}, the~$B$-module~$B^n$ is weakly
-Noetherian. We can therefore conclude if we can show, for any~$g \in A$, that if
-the submodule~$U[g^{-1}]$ is finitely generated, then~$g = 0$. This follows
-immediately from the assumption (for~$f = g$).
+If~$x_0 = f(n_0)$ for a number~$n_0 \in \NN$, then for any finite detachable
+subset~$M$ of~$\NN$ such that~$f[M] < x_0$,
+\[ \sum_{n \in M} 2^{-n} + 2^{-n_0} = \sum_{n \in M \cup \{n_0\}} 2^{-n}
+  \leq g(x_0 + 2^{-n_0}), \]
+hence~$g(x_0 + 2^{-n_0}) \geq g(x_0) + 2^{-n_0} \geq x_0 + 2^{-n_0}$.
+Thus~$x_0 + 2^{-n_0}$ is a greater postfixpoint that~$x_0$, a contradiction.
 \end{proof}

### 2018-07-07 paper-macneille-reals.tex
Introduction and elementary unwinding

@@
 \usetikzlibrary{patterns,matrix}
-\usepackage[protrusion=true]{microtype}
+\usepackage[expansion=true,protrusion=true]{microtype}
 \usepackage{breakurl}
@@
 \newtheorem{thm}[defn]{Theorem}
+\newtheorem*{thm*}{Theorem}
 \newtheorem{scholium}[defn]{Scholium}
@@
 \begin{abstract}
-  Leaning heavily on Levy's unusual proof of the uncountability of the reals,
-  we give an uncountability proof of the (MacNeille) reals which ... XXX
+  We give an uncountability proof of the reals which does not rely on their
+  sequential completeness, but on their order completeness. We use neither a
+  form of the axiom of choice nor the law of excluded middle, therefore
+  the proof applies to the MacNeille reals in any flavor of constructive
+  mathematics. The proof leans heavily on Levy's unusual proof of the
+  uncountability of the reals.
 \end{abstract}
@@
 
-\begin{thm}Let~$f : \NN \to \RR$ be a map. Then there is a number~$x_0 \in \RR$
-such that for no~$n \in \NN$, $f(n) = x_0$.\end{thm}
+One way to verify the uncountability of the reals is as follows. We first
+observe that the usual diagonalization technique shows that the powerset of the
+naturals is uncountable. We then show that this powerset is in bijection with
+the reals. In constructive mathematics, more precisely the kind of mathematics
+which can be carried out in any topos, the first step is still valid, while the
+second might fail -- for any of the several possible flavors of the reals
+which in constructive mathematics might not coincide. Therefore a different
+approach is needed.
+
+This note gives a constructive proof that a specific flavor of the reals,
+namely the MacNeille reals, is uncountable. To the best of our knowledge, this
+is the first result in that direction. However, it is just a baby step towards
+an understanding whether any of the more interesting flavors of the reals --
+for instance the Cauchy reals or the Dedekind reals -- can constructively be
+shown to be uncountable. The proof presented here is made possible because the
+MacNeille reals -- unlike the Cauchy or Dedekind reals -- can constructively be
+shown to be order-complete.
+
+Sensibilities of constructive mathematics aside, the proof presented here is
+interesting because it uses only the order completeness of the reals, not their
+sequential completeness, and because it puts the Knaster--Tarski fixed point
+theorem to good use. This fixed point theorem is fundamental to theoretical
+computer science, but appears to be seldomly used in classical analysis.
+The proof is an adaption of Levy's unusual proof~\cite{levy}.
+
+\begin{thm*}Let~$f : \NN \to \RR$ be a map. Then there is a number~$x_0 \in \RR$
+such that for no~$n \in \NN$, $f(n) = x_0$.\end{thm*}
 
@@
 where~$M$ ranges over all those finite detachable subsets of~$\NN$ such
-that~$f[M] < x$, is well-defined, monotone, has a postfixpoint ($0 \leq g(0)$),
-and has an upper bound for all of its postfixpoints (if~$x \leq g(x)$, then~$x
-\leq 2$). By the Knaster--Tarski fixed point theorem, it therefore has a
-greatest postfixpoint~$x_0$.
+that~$f[M] < x$, is well-defined (because the sets the suprema are taken of are
+inhabited by zero and bounded from above by~$2$), monotone, has a postfixpoint
+($0 \leq g(0)$), and has an upper bound for all of its postfixpoints (if~$x
+\leq g(x)$, then~$x \leq 2$). By the Knaster--Tarski fixed point theorem, it
+therefore has a greatest postfixpoint~$x_0$.
 
@@
 
+It is possible to unwind the application of the Knaster--Tarski fixed point
+theorem to obtain an entirely elementary proof of uncountability. This
+unwinding makes the impredicative nature of the proof manifest.
+
+\begin{proof}[Second proof]We consider the same map~$g : \RR \to \RR$ as in the
+first proof. Let~$x_0$ be the supremum of the set~$A \defeq \{ x \in \RR \,|\,
+x \leq g(x) \}$; this supremum exists because that set is inhabited (by zero)
+and bounded from above (by~$2$).
+
+If~$x_0 = f(n_0)$ for a number~$n_0 \in \NN$, then~$x_0 - 2^{-n_0}$ is an
+upper bound for~$A$, contradicting the fact that~$x_0$ is the least upper bound
+of~$A$: Let~$x \in A$. If~$x > x_0 - 2^{-n_0}$, then~$g(x +
+2^{-n_0}) \geq g(x) + 2^{-n_0} \geq x + 2^{-n_0}$, hence~$x + 2^{-n_0} \in A$,
+thus~$x + 2^{-n_0} \leq x_0$, a contradiction. Hence~$x \leq x_0 - 2^{-n_0}$.
+\end{proof}
+
 \end{document}

### 2018-07-08 paper-macneille-reals.tex
typo

@@
 %04103 Leipzig, Germany}
-\email{ingo.blechschmidt@mis.mpg.de, matthias.matthias.ralph.hutzler@student.uni-augsburg.de}
+\email{ingo.blechschmidt@mis.mpg.de, matthias.ralph.hutzler@student.uni-augsburg.de}
 
@@
 computer science, but appears to be seldomly used in classical analysis.
-The proof is an adaption of Levy's unusual proof~\cite{levy}.
+The proof is an adaptation of Levy's unusual proof~\cite{levy}.
 
@@
 first proof. Let~$x_0$ be the supremum of the set~$A \defeq \{ x \in \RR \,|\,
-x \leq g(x) \}$; this supremum exists because that set is inhabited (by zero)
+x \leq g(x) \}$; this supremum exists because~$A$ is inhabited (by zero)
 and bounded from above (by~$2$).

### 2018-07-09 slides-leipzig2018.tex
Begin on Leipzig slides

--- 2018-05-30 slides-filmat2018.tex
@@
 \usepackage{booktabs}
+\usepackage{stmaryrd}
 \usepackage{array}
@@
 \usepackage{tikz}
-\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing}
-\hypersetup{colorlinks=true}
-\usepackage{multimedia}
-\newcommand{\video}[2]{\movie[width=#2,height=#2,autostart,loop,poster]{}{#1}}
+\usetikzlibrary{mindmap,calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing}
 \hypersetup{colorlinks=false}
@@
 
-\title{Exploring mathematical objects from custom-tailored mathematical universes}
+\title{How topos theory can help algebra and geometry}
 \author{Ingo Blechschmidt}
-\date{May 26th, 2018}
+\date{July 10th, 2018}
 
@@
 \setbeamerfont{block title}{size={}}
-\addtobeamertemplate{headline}{\vskip-0.5ex}{}
 
@@
 \setbeamercolor*{titlelike}{bg=myred,fg=white}
+\setbeamercolor{frame}{bg=black}
 
@@
 \newcommand{\A}{\mathcal{A}}
+\newcommand{\B}{\mathcal{B}}
 \renewcommand{\AA}{\mathbb{A}}
+\renewcommand{\C}{\mathcal{C}}
 \newcommand{\E}{\mathcal{E}}
 \newcommand{\F}{\mathcal{F}}
+\newcommand{\M}{\mathcal{M}}
 \renewcommand{\G}{\mathcal{G}}
+\newcommand{\J}{\mathcal{J}}
 \newcommand{\GG}{\mathbb{G}}
@@
 \renewcommand{\P}{\mathcal{P}}
+\newcommand{\aaa}{\mathfrak{a}}
+\newcommand{\fff}{\mathfrak{f}}
 \newcommand{\ppp}{\mathfrak{p}}
+\newcommand{\mmm}{\mathfrak{m}}
 \newcommand{\defeq}{\vcentcolon=}
@@
 \newcommand{\Set}{\mathrm{Set}}
+\newcommand{\Ring}{\mathrm{Ring}}
+\newcommand{\LocRing}{\mathrm{LocRing}}
 \newcommand{\Eff}{\mathrm{Ef{}f}}
@@
 \newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
-\newcommand{\Ll}{\vcentcolon\!\Longleftrightarrow}
+\newcommand{\Ll}{\text{iff}}
 \newcommand{\inv}{inv.\@}
 \newcommand{\seq}{\vdash_{\!\!\!\vec x}}
+\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates
 
@@
   \end{center}}
+
+\newcommand{\bignumber}[1]{
+  \renewcommand{\insertenumlabel}{#1}\scalebox{1.5}{\usebeamertemplate{enumerate item}}
+}
+
+\newenvironment{indentblock}{%
+  \list{}{\leftmargin\leftmargin}%
+  \item\relax
+}{%
+  \endlist
+}
+
+\newcommand{\hcancel}[5]{%
+  \tikz[baseline=(tocancel.base)]{
+    \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
+    \draw[red, line width=0.4mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
+  }%
+}
 
@@
     \begin{minipage}{#1}
-      \setlength{\textwidth}{#1}
+      %\setlength{\textwidth}{#1}
       \begin{actionenv}#3
@@
   \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex,dp=1ex,right,rightskip=1mm,leftskip=1mm]{}%
-    \inserttitle
+    % \inserttitle
     \hfill
@@
 
-\IfSubStr{\jobname}{\detokenize{nonotes}}{
-  \setbeameroption{hide notes}
-}{
-  \setbeameroption{show notes}
-}
-\setbeamertemplate{note page}[plain]
-\setbeameroption{hide notes}
+\newcommand{\bad}[1]{\textcolor{red!90}{\textnormal{#1}}}
 
@@
 
-\addtocounter{framenumber}{-1}
-
-{\title{}
-\begin{frame}[c]
-  \centering
-  \medskip
-  \includegraphics[height=0.4\textheight]{acquaviva-platani-street}%
-  \includegraphics[height=0.4\textheight]{mussomeli-sheep}%
-  \includegraphics[height=0.4\textheight]{mussomeli-performance}%
-  \includegraphics[height=0.4\textheight]{mussomeli-castle}%
-  \medskip
-
-  \hil{Exploring mathematical objects from \\ custom-tailored mathematical universes} \\
-
-  \emph{-- an invitation --}
-
-  \bigskip
-
-  \scriptsize
-  Ingo Blechschmidt \\
-  University of Augsburg / \\
-  Max Planck Institute for Mathematics in the Sciences, Leipzig \\
-
-  \medskip
-
-  Third international conference of the \\
-  \emph{Italian Network for the Philosophy of Mathematics} \\
-  in Mussomeli
-  \medskip
-
-  May 26th, 2018
-  \par
-\end{frame}}
-
-\section[Toposophic landscape]{A glimpse of the toposophic landscape}
-
-\begin{frame}[fragile]{A glimpse of the toposophic landscape}
 \tikzstyle{topos} = [draw=mypurple, very thick, rectangle, rounded corners, inner sep=5pt, inner ysep=10pt]
@@
 
+\addtocounter{framenumber}{-1}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses}\end{minipage}}
+\begin{frame}[c]
+  \centering
+  \medskip
+  \hspace*{-3.9em}%
+
+  \hil{How topos theory can help algebra and geometry} \\
+
+  \emph{-- an invitation --}
+
+  \vfill
+
+  \scriptsize
+  Ingo Blechschmidt \\
+  July 10th, 2018
+  \par
+\end{frame}}
+
+\begin{frame}{Toposes are \ldots}
+  \begin{columns}
+    \begin{column}{0.5\textwidth}
+      \centering
+      \hil{generalized spaces}
+      \medskip
+
+      \includegraphics[height=0.35\textwidth]{grothendieck}
+      \quad
+      \includegraphics[height=0.35\textwidth]{elliptic-curve}
+
+      étale topos of a scheme
+    \end{column}
+
+    \begin{column}{0.5\textwidth}
+      \centering
+      \hil{mathematical universes}
+      \medskip
+
+      \begin{tikzpicture}
+        \node (overview) {
+          \scalebox{0.32}{\sieve{6}{4}}
+        };
+        \def\R{8pt}
+        \begin{pgfonlayer}{background}
+        \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple]
+          ($(overview.south west) + (\R, 0)$) arc(270:180:\R) --
+          ($(overview.north west) + (0, -\R)$) arc(180:90:\R) --
+          ($(overview.north east) + (-\R, 0)$) arc(90:0:\R) --
+          ($(overview.south east) + (0, \R)$) arc(0:-90:\R) --
+          cycle;
+        \end{pgfonlayer}
+      \end{tikzpicture}
+    \end{column}
+  \end{columns}
+
+  \bigskip
+  \bigskip
+
+  \begin{columns}
+    \begin{column}{0.5\textwidth}
+      \centering
+      \hil{categories of sheaves}
+      \medskip
+
+      \includegraphics[height=0.35\textwidth]{sheaf}
+
+      \footnotesize
+      A topos is a finitely complete cartesian closed category with a
+      subobject classifier.
+    \end{column}
+
+    \begin{column}{0.5\textwidth}
+      \centering
+      \hil{embodiments of theories}
+      \medskip
+
+      \includegraphics[height=0.35\textwidth]{olivia-lattices}
+
+      ``Let~$G$ be a group.''
+    \end{column}
+  \end{columns}
+\end{frame}
+
+\section{A glimpse of the toposophic landscape}
+
+\begin{frame}[fragile]{A glimpse of the toposophic landscape}
+  \hspace*{-1em}%
   \begin{tikzpicture}
@@
     };
-    \node[scale=0.4] (objs-set1) at (-3.5,-2.5) {
+    \node[scale=0.4] (objs-set1) at (-4.0,-2.5) {
       \only<1->{\fewstuff}
     };
-    \node[scale=0.4] (objs-eff1) at (3.5,-2.5) {
+    \node[scale=0.4] (objs-eff1) at (4.0,-2.5) {
       \only<2->{\fewstuff}
@@
 
-    \node (prop-sh1) [below of=objs-sh1, align=left] {
+    \node (prop-sh1) [below of=objs-sh1, align=left, inner sep=0pt] {
       \only<2->{%
-        The axiom of \\
-        choice fails.
+        The intermediate \\
+        value theorem fails.
       }
@@
 \end{frame}
-
-\section[Internal universe]{The internal universe of a topos}
 
@@
 
-  \vspace*{-1em}
+  \vspace*{-1.5em}
   \begin{hilblock}
@@
   \end{hilblock}
+  \bigskip
 \end{frame}
-
-\section[First steps]{First steps in alternate universes}
 
@@
 
-\section[Applications]{Applications in commutative algebra}
-
-\begin{frame}{Applications in commutative algebra}
-  Let~$A$ be a reduced commutative ring.
-
-  \hil{For instance:}\ \ $\ZZ$,\ \ $\ZZ[X]$,\ \ $\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$,\ \ $\QQ$,\ \ $\RR$
-
-  \mbox{The \hil{little Zariski topos} of~$A$ contains a \hil{mirror image} of~$A$:
-  $A^\sim$.}
-
-  \setlength\columnsep{0pt}
-  \begin{multicols*}{2}
-    \begin{enumerate}
-      \item $A^\sim$ is always a \hil{field}. \columnbreak
-      \item $A^\sim$ is still \hil{very close} to~$A$.
-    \end{enumerate}
-  \end{multicols*}
-  % \item Studying~$A^\sim$ is the same as studying~$A$ \hil{from a different
-  % point of view}.
-
-  \vspace*{-2em}
-  \begin{columns}[t]
-    \begin{column}[t]{0.47\textwidth}
-      \centering
-
-      \begin{varblock}{\textwidth}{A baby application}
-        \justifying
-        Let~$M$ be a surjective matrix with more rows than columns over a ring~$A$.
-        Then~$A = 0$.
-      \end{varblock}
-
-      \scalebox{0.8}{$\begin{pmatrix}
-        \cdot & \cdot \\
-        \cdot & \cdot \\
-        \cdot & \cdot
-      \end{pmatrix}$}
-    \end{column}
-
-    \begin{column}[t]{0.47\textwidth}
-      \centering
-
-      \begin{varblock}{\textwidth}{Generic freeness\phantom{p}}
-        \justifying
-        Generically, any finitely generated module over a reduced ring is free.
-      \end{varblock}
-      \vspace*{-0.5em}
-
-      \includegraphics[width=0.6\textwidth]{generic-freeness}
-    \end{column}
-  \end{columns}
-\end{frame}
-
-{\renewcommand{\insertframenumber}{4a}
-\begin{frame}{The little Zariski topos in more detail}
-  Recall~$A[f^{-1}] = \bigl\{ \frac{u}{f^n} \,|\, u \in A, n \in \NN \bigr\}$.
-
-  \begin{itemize}
-    \item $\Sh(\Spec(A)) \models \text{``For all~$x \in A^\sim$, \ldots''}$
-
-          Meaning: For all~$f \in A$ and all~$x \in A[f^{-1}]$, \ldots
-          \medskip
-
-    \item $\Sh(\Spec(A)) \models \text{``There is~$x \in A^\sim$ such that \ldots''}$
-
-          \mbox{Meaning: There is a partition of unity,~$1 = f_1 + \cdots + f_n \in A$,}
-          such that for each~$i$, there exists~$x_i \in A[f_i^{-1}]$
-          with~\ldots
-          \medskip
-
-    \item $\Sh(\Spec(A)) \models \text{``$\varphi$ implies $\psi$''}$
-
-          Meaning: For all~$f \in A$, if~$\varphi$ on stage~$f$, then~$\psi$ on
-          stage~$f$.
-  \end{itemize}
-\end{frame}
-\addtocounter{framenumber}{-1}}
-
-\appendix
-\section{Conclusion}
-
-\begin{frame}{Topos theory \ldots}
-  \vspace*{-0.5em}
-  \begin{itemize}
-    \item enriches the platonism debate,
-    %\item yields a new view on ``continuous dependence on parameters'',
-    \item uncovers further relations between objects,
-    \item allows to study objects from a different point of view,
-    \item has applications in mathematical practice.
-  \end{itemize}
-
-  \centering
-  \href{https://www.oliviacaramello.com/Papers/Papers.htm}{\includegraphics[height=0.45\textheight]{olivia-tst}}
-  \href{http://math.andrej.com/2014/03/04/intuitionistic-mathematics-and-realizability-in-the-physical-world/}{\includegraphics[height=0.45\textheight]{zenil-computable-universe}}
-  \href{https://pizzaseminar.speicherleck.de/skript2/zariski-topos-klein.pdf}{\includegraphics[height=0.45\textheight]{fun-with-the-zariski-topos}}
-  \href{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{\includegraphics[height=0.45\textheight]{phd-cover}}
-\end{frame}
-
-\addtocounter{framenumber}{-1}
-
 \end{document}

### 2018-07-09 slides-como2018.tex
typo

@@
   \fixedhref{https://www.ams.org/journals/proc/1988-103-04/S0002-9939-1988-0954974-5/S0002-9939-1988-0954974-5.pdf}{nontrivial
-  uses of nontrivial rings} and in the
+  uses of trivial rings} and in the
   \fixedhref{https://arxiv.org/abs/1605.04832}{recent textbook by Lombardi and
@@
     sheaf of sets & set \\
-    %sheaf of rings & ring \\
+    % sheaf of rings & ring \\
     sheaf of modules & module \\
@@
     big Zariski topos of $X$ & big Zariski topos of the ring $\O_X$ [with a twist] \\
-    higher direct images & sheaf cohomology \\
+    higher direct image & sheaf cohomology \\
     \bottomrule
@@
 \end{document}
-
-\end{document}

### 2018-07-09 slides-leipzig2018.tex
Slides for the Leipzig talk finished

@@
 
-      \includegraphics[height=0.35\textwidth]{grothendieck}
+      \includegraphics[height=0.32\textwidth]{grothendieck}
       \quad
-      \includegraphics[height=0.35\textwidth]{elliptic-curve}
+      \includegraphics[height=0.32\textwidth]{elliptic-curve}
 
-      étale topos of a scheme
+      \footnotesize
+      étale topos of a scheme \\
+      field with one element
     \end{column}
@@
         \node (overview) {
-          \scalebox{0.32}{\sieve{6}{4}}
+          \scalebox{0.44}{\sieve{6}{4}}
         };
@@
 
-      \includegraphics[height=0.35\textwidth]{sheaf}
+      \includegraphics[height=0.32\textwidth]{sheaf}
 
@@
 
-      \includegraphics[height=0.35\textwidth]{olivia-lattices}
+      \includegraphics[height=0.32\textwidth]{olivia-lattices}
 
@@
 
+\newcommand{\appl}[3]{\centering\includegraphics[height=3.5em]{#1}\par\hil{#2}\\\footnotesize#3}
+
+\begin{frame}{Applications}
+  \bigskip
+
+  \small
+  \begin{columns}
+    \begin{column}{0.35\textwidth}
+      \appl{analysis}{analysis}{
+        internalizing parameter-dependence
+      }
+    \end{column}
+
+    \begin{column}{0.39\textwidth}
+      \appl{calabi-yau}{algebraic geometry}{
+        \mbox{reducing geometry to algebra} \\
+        \mbox{reducing relative to absolute} \\
+        synthetic account
+      }
+    \end{column}
+
+    \begin{column}{0.35\textwidth}
+      \appl{klein-bottle}{differential geometry}{
+        reflection principles \\
+        synthetic account
+      }
+    \end{column}
+  \end{columns}
+
+  \bigskip
+  \bigskip
+
+  \begin{columns}
+    \begin{column}{0.35\textwidth}
+      \appl{torus-rainbow}{homotopy theory}{
+        synthetic account
+        computer-assisted proofs
+        generalizations
+      }
+    \end{column}
+
+    \begin{column}{0.39\textwidth}
+      \appl{generic-freeness}{commutative algebra}{
+        local-to-global principles \\
+        reduction techniques \\
+        constructive proofs
+      }
+    \end{column}
+
+    \begin{column}{0.35\textwidth}
+      \appl{bohr-topos}{further subjects}{
+        synth.~computability th. \\
+        synth.~measure theory \\
+        Bohr topos for QM
+      }
+    \end{column}
+  \end{columns}
+\end{frame}
+
+\section[Analysis]{Applications in analysis}
+
+\begin{frame}{The topos of sheaves over a space}
+  \small\vspace*{-0.8em}
+  Let~$X$ be a topological space. We recursively define
+  \[ U \models \varphi \quad \text{(``$\varphi$ holds on~$U$'')} \]
+  for open subsets~$U \subseteq X$ and formulas~$\varphi$.
+  Write~``$\Sh(X) \models \varphi$'' to mean~$X \models \varphi$.
+  Let~$\C(U)$ be the set of continuous functions~$U \to \RR$.
+
+  \vspace*{-1.3em}
+  \footnotesize
+  \[ \renewcommand{\arraystretch}{1.15}\begin{array}{@{}l@{\ \ }c@{\ \ }l@{}}
+  U \models \top &\Ll& \text{true} \\
+  U \models \bot &\Ll& \hcancel{\text{false}}{0pt}{3pt}{0pt}{-2pt}\ U = \emptyset \\
+  U \models s = t &\Ll& \text{$s(x) = t(x)$ for all $x \in U$} \\
+  U \models \varphi \wedge \psi &\Ll&
+  \text{$U \models \varphi$ and $U \models \psi$} \\
+  U \models \varphi \vee \psi &\Ll&
+  \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $U = \bigcup_i U_i$} \\
+  && \quad\quad\text{such that for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
+  U \models \varphi \Rightarrow \psi &\Ll&
+  \text{for all open~$V \subseteq U$: }
+  \text{$V \models \varphi$ implies $V \models \psi$} \\
+  U \models \forall s \? \RR\_ \varphi(s) &\Ll&
+  \text{for all open $V \subseteq U$ and functions~$s_0 \in \C(V)$: $V \models \varphi(s_0)$} \\
+  U \models \exists s \? \RR\_ \varphi(s) &\Ll&
+  \hcancel{\text{there exists $s_0 \in \C(U)$ such that $U \models \varphi(s_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
+  &&
+  \text{there exists a covering $U = \bigcup_i U_i$ such that for all~$i$:} \\
+  && \quad\quad \text{there exists~$s_0 \in \C(U_i)$ such that $U_i \models \varphi(s_0)$}
+  \end{array} \]
+\end{frame}
+
+\section[Differential geometry]{Applications in differential geometry}
+
+\begin{frame}{Synthetic differential geometry}
+  \vspace*{-1em}
+  \begin{center}
+    \includegraphics[width=0.5\textwidth]{microaffinity}
+  \end{center}
+
+  \vspace*{-2em}
+  \begin{varblock}{10.8cm}{The axiom of microaffinity}
+    \justifying
+    Let~$\Delta = \{ \varepsilon \in \RR \,|\, \varepsilon^2 = 0 \}$.
+    For any function $f : \Delta \to \RR$,
+    there is a unique number~$a \in \RR$ such that
+    $f(\varepsilon) = f(0) + a \varepsilon$
+    for all~$\varepsilon \in \Delta$.
+  \end{varblock}
+
+  \vspace*{-1em}
+  \small
+  \begin{itemize}
+    \item The \hil{derivative} of a function~$f : \RR \to \RR$ at~$x_0 \in \RR$
+    is the unique number~$a \in \RR$ such that~$f(x_0 + \varepsilon) = f(x_0) +
+    a \varepsilon$ for all~$\varepsilon \in \Delta$. \\[-1.2em]
+    \item Manifolds are \hil{just sets}. \\[-1.2em]
+    \item A \hil{tangent vector} to~$M$ is a map~$\Delta \to M$.
+  \end{itemize}
+\end{frame}
+
+\section[Commutative algebra]{Applications to commutative algebra}
+
+\begin{frame}{The little Zariski topos of a ring}
+  \small\vspace*{-0.8em}
+  Let~$A$ be a ring (commutative, with unit). We recursively define
+  \[ D(f) \models \varphi \quad \text{(``$\varphi$ holds away from the zeros of~$f$'')} \]
+  for~$f \in A$ and formulas~$\varphi$. Write
+  ``$\Spec(A) \models \varphi$'' to mean~$D(1) \models \varphi$.
+
+  \vspace*{-1em}
+  \[\footnotesize\renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\ }c@{\ }l@{}}
+    D(f) \models \top &\text{iff}& \text{true} \\
+    D(f) \models \bot &\text{iff}& \text{$f$ is nilpotent} \\
+    D(f) \models x = y &\text{iff}& x = y \in A[f^{-1}] \\
+    D(f) \models \varphi \wedge \psi &\text{iff}&
+      \text{$D(f) \models \varphi$ and $D(f) \models \psi$} \\
+    D(f) \models \varphi \vee \psi &\text{iff}&
+      \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
+    &&\quad\text{for each~$i$, $D(fg_i) \models \varphi$ or $D(fg_i) \models \psi$} \\
+    D(f) \models \varphi \Rightarrow \psi &\text{iff}&
+      \text{for all~$g \in A$, $D(fg) \models \varphi$ implies $D(fg) \models \psi$} \\
+    D(f) \models \forall x\?M^\sim\_ \varphi(x) &\text{iff}&
+      \text{for all~$g \in A$ and $x_0 \in M[(fg)^{-1}]$, $D(fg) \models \varphi[x_0/x]$} \\
+    D(f) \models \exists x\?M^\sim\_ \varphi(x) &\text{iff}&
+      \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
+    &&\quad\text{for each~$i$, $D(fg_i) \models \varphi[x_0/x]$ for some~$x_0 \in M[(fg_i)^{-1}]$}
+  \end{array} \]
+\end{frame}
+
+\section[Algebraic geometry]{Applications in algebraic geometry}
+
+\begin{frame}{Understanding algebraic geometry}
+  \vspace*{-1.5em}\small
+  \begin{varblock}{0.9\textwidth}{}
+    \justifying
+    Understand \hil{notions of algebraic geometry} over a scheme~$X$ as
+    \hil{notions of algebra} internal to~$\Sh(X)$.
+  \end{varblock}
+
+  \only<1>{
+    \centering
+    \rotatebox{90}{\tiny\scalebox{0.8}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
+    \includegraphics[width=0.7\textwidth]{external-internal-small}
+    \par
+  }
+
+  \pause
+
+  \small\centering
+  \scalebox{0.83}{\begin{tabular}{ll}
+    \toprule
+    externally & internally to $\Sh(X)$ \\
+    \midrule
+    sheaf of sets & set \\
+    % sheaf of rings & ring \\
+    sheaf of modules & module \\
+    % sheaf of finite type & finitely generated module \\
+    finite locally free sheaf & finite free module \\
+    % coherent sheaf & coherent module \\
+    tensor product of sheaves & tensor product of modules \\
+    % sheaf of Kähler differentials & module of Kähler differentials \\
+    sheaf of rational functions & total quotient ring of~$\O_X$ \\
+    dimension of $X$ & Krull dimension of~$\O_X$ \\
+    spectrum of a sheaf of~$\O_X$-algebras & ordinary spectrum [with a twist] \\
+    % big Zariski topos of $X$ & big Zariski topos of the ring $\O_X$ [with a twist] \\
+    higher direct image & sheaf cohomology \\
+    \bottomrule
+  \end{tabular}}
+
+  \footnotesize
+
+  \begin{columns}[c]
+    \begin{column}{0.47\textwidth}
+      \begin{exampleblock}{}
+        \justifying
+        Let $0 \to \F' \to \F \to \F'' \to 0$ be a short exact sequence
+        of sheaves of~$\O_X$-modules. If~$\F'$ and~$\F''$ are of finite type,
+        so is~$\F$.
+      \end{exampleblock}
+    \end{column}
+
+    \begin{column}{0.1\textwidth}
+      \vspace*{0.7em}
+      \scalebox{3}{$\Leftarrow$}
+    \end{column}
+
+    \begin{column}{0.44\textwidth}
+      \begin{exampleblock}{}
+        \justifying
+        Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence of
+        modules. If~$M'$ and~$M''$ are finitely generated, so is~$M$.
+      \end{exampleblock}
+    \end{column}
+  \end{columns}
+\end{frame}
+
+\begin{frame}{Synthetic algebraic geometry}
+  Usual approach to algebraic geometry: \hil{layer schemes above ordinary set theory}
+  using either
+  \begin{itemize}
+    \item locally ringed spaces
+    \small
+    \begin{multline*}
+      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$} + {} \\
+      \text{Zariski topology} + \text{structure sheaf}
+    \end{multline*}
+    \normalsize
+    \item or Grothendieck's functor-of-points account, where a scheme is a
+    functor~$\Ring \to \Set$.
+    \small\[ A \longmapsto \{ (x,y,z) \in A^3 \,|\, x^n+y^n-z^n=0 \} \]
+  \end{itemize}
+  \bigskip
+
+  \hil{Synthetic approach:} model schemes \hil{directly as sets} in
+  the internal universe of the \mbox{\hil{big Zariski topos}} of a base scheme.
+  \small
+  \[ \{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]
+\end{frame}
+
 \end{document}

### 2018-07-09 slides-como2018.tex
typo (many found by Matthias Hutzler, thanks!)

@@
 
-\begin{frame}{Motivating testcases}
+\begin{frame}{Motivating test cases}
   Let $A$ be a ring (commutative, with unit, $1 = 0$ allowed). \\
@@
   a way to perform the reduction in an entirely constructive manner. If so
-  desired, the resulting topos-theoretic proofs can be unwinded to yield fully
+  desired, the resulting topos-theoretic proofs can be unwound to yield fully
   explicit, topos-free, direct proofs.
@@
   statement by employing not a maximal ideal, but a prime ideal. The existence
-  of maximal ideals is nontrivial rings is equivalent to the axiom of choice,
+  of maximal ideals in nontrivial rings is equivalent to the axiom of choice,
   while the existence of prime ideals is equivalent to the weaker Boolean Prime
   Ideal Theorem. However, this improvement doesn't change the main point; the
-  proof is still wildly unconstructive.}
+  proof is still wildly nonconstructive.}
 
@@
   \jnote{1}{The internal language of a topos allows to construct objects and
-  morphisms, formulate statements about them and prove such statements in a
-  naive element-based language. From the internal point of view, objects look
-  like sets [more precisely, types]; morphisms look like maps; epimorphisms
+  morphisms of the topos, formulate statements about them and prove such
+  statements \emph{in a naive element-based language}. From the internal point
+  of view, objects look like sets [more precisely, types]; morphisms look like
+  maps; monomorphisms look like injections; epimorphisms
   look like surjections; group objects look like groups; and so on.
@@
   toposes validate the law of excluded middle (for instance, sheaf toposes over
-  discrete topological space do if the law of excluded middle is available in
+  discrete topological spaces do if the law of excluded middle is available in
   the metatheory). Luckily, vast amounts of mathematics can be developed in a
@@
   \jnote{1}{Many interesting sheaves have few global sections, which is why a
-  definition should as ``$U \models \forall s \? F\_ \varphi(s)$ iff $U \models
-  \varphi(s_0)$ for all~$s_0 \in F(U)$'' would miss the point. More precisely,
-  changing the definitions like this would yield the internal language of the
-  topos of presheaves on~$X$.
+  definition such as ``$U \models \forall s \? F\_ \varphi(s)$ iff $U \models
+  \varphi(s_0)$ for all~$s_0 \in F(U)$'' would miss the point.
 
@@
       \text{for all open~$U \subseteq X$, sections $s, t \in F(U)$:
-        $\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.2em]
+        $\alpha_U(s) = \alpha_U(t)$ implies $s = t$} \\[0.2em]
     \Longleftrightarrow\ &
@@
 
-  One can also consider the classifying local of prime \emph{ideals} of~$A$.
-  Its associated topological space of points is the the set of prime ideal
+  One can also consider the classifying locale of prime \emph{ideals} of~$A$.
+  Its associated topological space of points is the set of prime ideals
   of~$A$ equipped with the \emph{constructible topology}.
@@
 
-  As an aside, the little Zariski topos is seldomly Boolean (validates the law of
+  As an aside, the little Zariski topos is rarely Boolean (validates the law of
   excluded middle). A necessary condition is that~$A$ is of dimension~$\leq 0$.}
@@
   is entirely mechanical, so is the translation from internal proofs to
-  external proofs. Any proof employing the internal language can be unwinded to
+  external proofs. Any proof employing the internal language can be unwound to
   yield an external proof not referencing the internal language.
 
-  However, depending on the logical complexity of the statements occuring in a
+  However, depending on the logical complexity of the statements occurring in a
   given proof, the resulting external proof might be (much) more complex than
@@
 
-\begin{frame}{Revisiting the testcases}
+\begin{frame}{Revisiting the test cases}
   Let $A$ be a reduced ring.
@@
   Since the internal language machinery is itself constructive, the displayed
-  proofs can be unwinded to yield external constructive proofs which don't
+  proofs can be unwound to yield external constructive proofs which don't
   reference topos theory. Details on these external proofs can soon be found in
@@
 
-  \jnote{2}{The testcase of Grothendieck's generic freeness lemma illustrates
+  \jnote{2}{The test case of Grothendieck's generic freeness lemma illustrates
   that the internal language of toposes can help commutative algebra even if
@@
   have the weaker statement~$\neg\neg(\varphi \vee \neg\varphi)$, and we can
-  use this law in place of the law of excluded middle to intuitionistially
+  use this law in place of the law of excluded middle to intuitionistically
   deduce that any finitely generated vector space does \emph{not not} have a
@@
   \begin{itemize}
-    \item $\affl$ is a local ring:
-    \[ 1 \neq 0 \qquad\qquad \text{$x + y$ \inv} \Longrightarrow \text{$x$ \inv} \vee
-    \text{$y$ \inv} \]
+%   \item $\affl$ is a local ring:
+%   \[ 1 \neq 0 \qquad\qquad \text{$x + y$ \inv} \Longrightarrow \text{$x$ \inv} \vee
+%   \text{$y$ \inv} \]
     \item $\affl$ is a field:
@@
     \varepsilon^2 = 0 \}$.
-    \item Any map $\affl \to \affl$ is a polynomial.
+    \bigskip
+    \item Any map $\affl \to \affl$ is a polynomial function.
+    \bigskip
     \item $\affl$ is anonymously algebraically closed: Any monic polynomial
@@
 
-  \jnote{3}{An anlogue of synthetic quasicoherence holds in the classifying
+  \jnote{3}{An analogue of synthetic quasicoherence holds in the classifying
   topos of rings, demonstrating that even presheaf toposes can validate
@@
   \jnote{1}{Toposes, and also more specifically classifying toposes, originated
-  in algebraic geometry. It is therefore deeply embarassing that as of now,
+  in algebraic geometry. It is therefore deeply embarrassing that as of now,
   still very little is known about the theories classified by the major toposes

### 2018-07-09 slides-como2018.tex
typo

@@
   \footnotesize
-  \[ \renewcommand{\arraystretch}{1.08}\begin{array}{@{}l@{\ }c@{\ }l@{}}
+  \[ \renewcommand{\arraystretch}{1.08}\begin{array}{@{}l@{\ \ }c@{\ \ }l@{}}
   U \models \top &\Ll& \text{true} \\
@@
   \hcancel{\text{there exists $s_0 \in F(U)$ such that $U \models \varphi(s_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
-  && \quad\quad
+  &&
   \text{there exists a covering $U = \bigcup_i U_i$ such that for all~$i$:} \\
-  && \quad\quad\quad\quad \text{there exists~$s_0 \in F(U_i)$ such that $U_i \models \varphi(s_0)$} \\
+  && \quad\quad \text{there exists~$s_0 \in F(U_i)$ such that $U_i \models \varphi(s_0)$} \\
   U \models \exists F\_ \varphi(F) &\Ll&
@@
   \varphi(F_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
-  && \quad\quad
+  &&
   \text{there exists a covering $U = \bigcup_i U_i$ such that for all~$i$:} \\
-  && \quad\quad\quad\quad \text{there exists a sheaf~$F_0$ on $U_i$ such that $U_i \models \varphi(F_0)$}
+  && \quad\quad \text{there exists a sheaf~$F_0$ on $U_i$ such that $U_i \models \varphi(F_0)$}
   \end{array} \]
@@
     \footnotesize
-    Gavin Wraith. Generic Galois theory of local rings. 1979.
+    Gavin Wraith. Generic Galois theory of local rings. 1979. \\
+    (A cult classic and must-read for anyone interested in \\
+    the intersection of topos theory and algebraic geometry.)
   \end{center}}

### 2018-07-09 slides-leipzig2018.tex
typo

@@
   \begin{center}
-    \begin{minipage}{9.05cm}
-      \setlength{\textwidth}{9.05cm}
+    \begin{minipage}{9.85cm}
+      \setlength{\textwidth}{9.85cm}
       \begin{actionenv}#1
@@
 
-      \includegraphics[height=0.32\textwidth]{sheaf}
+      \includegraphics[height=0.304\textwidth]{sheaf}
 
@@
 
+      \footnotesize
       ``Let~$G$ be a group.''
@@
   \begin{hilblock}
-    If~$\E \models \varphi$ and if~$\varphi \vdash \psi$
+    If~$\E \models \varphi$ and if~$\varphi$
     \pointthis{<3>}{intuitionistically}{%
@@
       no $\neg\neg\varphi \Rightarrow \varphi$,\ \
-      no axiom of choice},
+      no axiom of choice} entails~$\psi$,
     then~$\E \models \psi$.
@@
   \[ U \models \varphi \quad \text{(``$\varphi$ holds on~$U$'')} \]
-  for open subsets~$U \subseteq X$ and formulas~$\varphi$.
+  for open subsets~$U \subseteq X$ and statements~$\varphi$.
   Write~``$\Sh(X) \models \varphi$'' to mean~$X \models \varphi$.
@@
   \[ D(f) \models \varphi \quad \text{(``$\varphi$ holds away from the zeros of~$f$'')} \]
-  for~$f \in A$ and formulas~$\varphi$. Write
+  for~$f \in A$ and statements~$\varphi$. Write
   ``$\Spec(A) \models \varphi$'' to mean~$D(1) \models \varphi$.

### 2018-07-10 slides-como2018.tex
typo

@@
     D(f) \models \forall x\?M^\sim\_ \varphi(x) &\text{iff}&
-      \text{for all~$g \in A$ and $x_0 \in M[(fg)^{-1}]$, $D(fg) \models \varphi[x_0/x]$} \\
+      \text{for all~$g \in A$ and $x_0 \in M[(fg)^{-1}]$, $D(fg) \models \varphi(x_0)$} \\
     D(f) \models \exists x\?M^\sim\_ \varphi(x) &\text{iff}&
       \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
-    &&\quad\text{for each~$i$, $D(fg_i) \models \varphi[x_0/x]$ for some~$x_0 \in M[(fg_i)^{-1}]$}
+    &&\quad\text{for each~$i$, $D(fg_i) \models \varphi(x_0)$ for some~$x_0 \in M[(fg_i)^{-1}]$}
   \end{array} \]

### 2018-07-10 slides-leipzig2018.tex
typo

@@
     D(f) \models \forall x\?M^\sim\_ \varphi(x) &\text{iff}&
-      \text{for all~$g \in A$ and $x_0 \in M[(fg)^{-1}]$, $D(fg) \models \varphi[x_0/x]$} \\
+      \text{for all~$g \in A$ and $x_0 \in M[(fg)^{-1}]$, $D(fg) \models \varphi(x_0)$} \\
     D(f) \models \exists x\?M^\sim\_ \varphi(x) &\text{iff}&
       \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
-    &&\quad\text{for each~$i$, $D(fg_i) \models \varphi[x_0/x]$ for some~$x_0 \in M[(fg_i)^{-1}]$}
+    &&\quad\text{for each~$i$, $D(fg_i) \models \varphi(x_0)$ for some~$x_0 \in M[(fg_i)^{-1}]$}
   \end{array} \]
 \end{frame}
+
+\backupstart
 
@@
 
+\backupend
+
 \end{document}

### 2018-07-10 slides-como2018.tex
typo

@@
     D(f) \models \bot &\text{iff}& \text{$f$ is nilpotent} \\
-    D(f) \models x = y &\text{iff}& x = y \in A[f^{-1}] \\
+    D(f) \models x = y &\text{iff}& x = y \in M[f^{-1}] \\
     D(f) \models \varphi \wedge \psi &\text{iff}&

### 2018-07-10 slides-leipzig2018.tex
typo

@@
     \node[anchor=base,inner sep=0,outer sep=0] (#2) {#2};
-    \node[visible on=#1,overlay,rectangle callout,rounded corners,callout relative pointer={(0.3cm,0.5cm)},fill=blue!20] at ($(#2.north)+(-0.1cm,-1.1cm)$) {#3};
+    \node[visible on=#1,overlay,rectangle callout,rounded corners,callout relative pointer={(-0.1cm,0.5cm)},fill=blue!20] at ($(#2.north)+(-0.1cm,-1.1cm)$) {#3};
   }%
@@
   \begin{hilblock}
-    If~$\E \models \varphi$ and if~$\varphi$
+    If~$\E \models \varphi$ and if~$\varphi$ entails~$\psi$
     \pointthis{<3>}{intuitionistically}{%
@@
       no $\neg\neg\varphi \Rightarrow \varphi$,\ \
-      no axiom of choice} entails~$\psi$,
+      no axiom of choice},
     then~$\E \models \psi$.
@@
   U \models \varphi \vee \psi &\Ll&
-  \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $U = \bigcup_i U_i$} \\
+  \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists an open covering $U = \bigcup_i U_i$} \\
   && \quad\quad\text{such that for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
@@
   &&
-  \text{there exists a covering $U = \bigcup_i U_i$ such that for all~$i$:} \\
+  \text{there exists an open covering $U = \bigcup_i U_i$ such that for all~$i$:} \\
   && \quad\quad \text{there exists~$s_0 \in \C(U_i)$ such that $U_i \models \varphi(s_0)$}
@@
     \begin{column}{0.1\textwidth}
-      \vspace*{0.7em}
+      \vspace*{1.1em}
       \scalebox{3}{$\Leftarrow$}

### 2018-07-10 slides-leipzig2018.tex
Beginning of notes for the Leipzig talk

@@
 \usepackage{etex}
+
+\usepackage{pgfpages}
+\setbeameroption{show notes on second screen}
+\setbeamertemplate{note page}[plain]
+\newcommand{\jnote}[2]{\only<#1>{\note{\footnotesize\justifying#2\par}}}
 
@@
 \newcommand{\RR}{\mathbb{R}}
+\newcommand{\CC}{\mathbb{C}}
 \newcommand{\TT}{\mathbb{T}}
@@
   \end{columns}
+
+  \jnote{1}{Toposes were invented in the 1960s by Grothendieck in order to
+  solve concrete problems in algebraic geometry. Their raison d'être is the
+  following. In algebraic geometry, we want (and sometimes have to) work over
+  fields other than~$\RR$ and~$\CC$ and even over arbitrary commutative rings.
+  For the geometric objects in such settings, the \emph{schemes}, the Euclidean
+  topology is not available; we have to make do with the \emph{Zariski
+  topology}. However, important tools as singular cohomology don't work well
+  with this topology.
+
+  The problem was solved by inventing \emph{étale topology} as an enhancement
+  of the Zariski topology. However, contrary to its name, the étale topology
+  isn't a true topology in the usual sense. Putting the étale topology on a
+  scheme doesn't yield a refined topological space, but a \emph{topos}.
+
+  Toposes generalize topological spaces in two ways: Firstly, the ``open sets''
+  of toposes don't actually have to be sets of points; they can be more general
+  kinds of objects such as coverings. Toposes can even have no points at all
+  and still be nontrivial (this is for instance the case for the \emph{topos of
+  random sequences}). Secondly, while classically a given open
+  subset is either contained in a further open subset or not, the opens of
+  toposes can be contained in further opens in many different ways.
+
+  Recently, toposes are being used to help the mysterious field with
+  one element come into being.}
+
+  \jnote{2}{Since the 1960s, many more aspects of toposes were discovered.
+  A popular reference on toposes starts with a list of 13 ways of viewing
+  toposes.
+
+  In this talk, we'll focus on the view of toposes as \emph{alternate
+  mathematical universes}. We can do mathematics inside these alternate
+  universes just as well as we can do mathematics inside the \emph{standard
+  universe} (which is represented by a particular topos called~``$\Set$'').
+  Each topos contains own versions of all the familiar mathematical objects --
+  numbers, functions, manifolds -- but the properties they enjoy differ
+  slightly from topos to topos.
+
+  The definition of what a topos is, displayed in the lower left of the slide,
+  has two problems. Firstly, it's only useful if one knows the relevant
+  category-theoretic jargon. Secondly, a topos has lots of further vital
+  structure, which is crucial for a rounded understanding. However, the
+  displayed definition is trimmed for minimality and therefore doesn't mention
+  it. A more comprehensive definition is: A \emph{topos} is a locally cartesian
+  closed, finitely complete and cocomplete Heyting category which is exact,
+  extensive and has a subobject classifier. We won't need either definition in
+  this talk.}
+
+  \jnote{3}{Toposes can also be thought as reifying the ``semantic essence'' of
+  mathematical theories (the theory of groups, the theory of rings, the theory
+  of intervals, \ldots). Given any such theory~$T$, there is a so-called
+  \emph{classifying topos}~$\Set[T]$. Its points are precisely the set-based
+  models of~$T$ (the groups, the rings, the intervals, \ldots), and it contains
+  a \emph{generic model} which has exactly those properties which all models
+  have. (This generic model is what mathematicians implicitly refer to when
+  they say ``Let~$G$ be a group.''.)
+
+  Crucially, two theories~$T$ and~$T'$ can have equivalent classifying
+  toposes even when they are not syntactially related in any way. This
+  observation is the starting point of Olivia Caramello's \emph{bridge
+  technique}, a grand research program with applications in many different
+  fields.
+
+  \[ \xymatrix{
+    & \Set[T] \simeq \Set[T'] \ar@{--}@/^1pc/[rd] \ar@{--}@/_1pc/[ld] \\
+    T && T'
+  } \]}
 \end{frame}
@@
   \end{tikzpicture}
+
+  \jnote{1}{The topos~$\Set$ is the standard topos. This topos is where
+  ordinary mathematics takes place.}
+
+  \jnote{2}{Besides~$\Set$, there is a proper class worth of further toposes.
+  We'll get to know some of these better during the remaining course of the
+  talk.
+
+  For any topological space~$X$, there is a topos~$\Sh(X)$ of \emph{sheaves
+  over~$X$}. They are useful in analysis for internalizing
+  parameter-dependence. Apart from pathological cases like~$X$
+  being discrete, the intermediate value theorem in the form
+  \begin{indentblock}Let~$f : \RR \to \RR$ be a continuous function.
+  Assume~$f(-1) < 0 < f(1)$. Then there is a number~$x \in \RR$ such
+  that~$f(x) = 0$.\end{indentblock}
+  fails in these toposes -- for very meaningful reasons discussed below.
+  The following, classically equivalent, version does hold:
+  \begin{indentblock}Let~$f : \RR \to \RR$ be a continuous function.
+  Assume~$f(-1) < 0 < f(1)$. Then, for every~$\varepsilon > 0$, there is
+  a number~$x \in \RR$ such that~$|f(x)| < \varepsilon$.\end{indentblock}
+
+  The \emph{effective topos}~$\Eff$ is a computer scientist's dream come true:
+  In it, any function~$\NN \to \NN$ is computable by a Turing machine. The
+  effective topos and its close cousins can be used to study the differences
+  between the many models of computation, particularly those differences which
+  are only visible at higher types.}
 \end{frame}

### 2018-07-10 slides-leipzig2018.tex
typo

@@
     If~$\E \models \varphi$ and if~$\varphi$ entails~$\psi$
-    \pointthis{<3>}{intuitionistically}{%
+    \pointthis{<3>}{constructively}{%
       no $\varphi \vee \neg\varphi$,\ \

### 2018-07-10 slides-leipzig2018.tex
More notes

@@
 \newcommand{\jnote}[2]{\only<#1>{\note{\footnotesize\justifying#2\par}}}
+
+% Workaround for the issue described at
+% https://tex.stackexchange.com/questions/164406/beamer-using-href-in-notes.
+\newcommand{\fixedhref}[2]{\makebox[0pt][l]{\hspace*{\paperwidth}\href{#1}{#2}}\href{#1}{#2}}
 
@@
   topology}. However, important tools as singular cohomology don't work well
-  with this topology.
+  with this topology (too few opens).
 
@@
   of the Zariski topology. However, contrary to its name, the étale topology
-  isn't a true topology in the usual sense. Putting the étale topology on a
+  isn't actually a topology in the usual sense. Putting the étale topology on a
   scheme doesn't yield a refined topological space, but a \emph{topos}.
@@
   toposes can be contained in further opens in many different ways.
+  % XXX: Link to Alex Simpson
 
@@
   one element come into being.}
+  % XXX: Link to John Baez
 
@@
   Each topos contains own versions of all the familiar mathematical objects --
-  numbers, functions, manifolds -- but the properties they enjoy differ
-  slightly from topos to topos.
+  numbers, functions, manifolds -- but the properties they enjoy can differ
+  slightly from the properties they enjoy in the standard topos.
 
@@
   category-theoretic jargon. Secondly, a topos has lots of further vital
-  structure, which is crucial for a rounded understanding. However, the
-  displayed definition is trimmed for minimality and therefore doesn't mention
-  it. A more comprehensive definition is: A \emph{topos} is a locally cartesian
+  structure, which is crucial for a rounded understanding, but not listed in the
+  displayed definition (which is trimmed for minimality).
+  A more comprehensive definition is: A \emph{topos} is a locally cartesian
   closed, finitely complete and cocomplete Heyting category which is exact,
@@
   fields.
+  % XXX: Link Olivia
 
@@
   \jnote{2}{Besides~$\Set$, there is a proper class worth of further toposes.
-  We'll get to know some of these better during the remaining course of the
+  We'll get to know some of these better during the course of the
   talk.
 
-  For any topological space~$X$, there is a topos~$\Sh(X)$ of \emph{sheaves
+  For any topological space~$X$, there is the topos~$\Sh(X)$ of \emph{sheaves
   over~$X$}. They are useful in analysis for internalizing
@@
   are only visible at higher types.}
+
+  \jnote{3}{This talk is still part of ordinary mathematics, that is, of the
+  topos~$\Set$. Therefore a more accurate picture depicts the three examples as
+  part of~$\Set$. (Technically, while they might not be ``toposes
+  over~$\Set$'', they are still locally-internal to~$\Set$ in the sense of
+  Penon.)
+
+  Most of the toposes in active use are either toposes of sheaves (over a
+  topological space or a XXXnLab \emph{site}), realizability toposes (such
+  as~$\Eff$ or variants constructed using different models of computation), or
+  arise from those using topos-theoretic constructions; but this is not a
+  complete classification.}
 \end{frame}
@@
   using the \hil{Kripke--Joyal semantics}.
+
+  \jnote{1-2}{We can picture the Kripke--Joyal semantics as a kind of translation
+  engine. When we're talking with Anna, a mathematician who lives in the
+  ef{}fective topos, at first we might be weirded out when she states ``it's a basic
+  fact of life that any function~$\NN \to \NN$ is computable''. But if we
+  remember to switch on the Kripke--Joyal translation, we instead hear~``it's a
+  basic fact of life that there is a Turing machine which, given a Turing
+  machine computing a function~$f : \NN \to \NN$, outputs a Turing machine
+  computing~$f$'' which we can easily agree with.
+
+  The precise translation rules will be explained by osmosis for the
+  ef{}fective topos, on the next slide, and by a formal definition for sheaf
+  toposes and for the little Zariski topos, further below.}
+
+  \jnote{2}{When exploring a new topos for the first time, the only way to
+  find out which statements hold in it is to translate them using the
+  Kripke--Joyal semantics and check whether the translation holds in the usual
+  mathematical sense. As soon as we have established a certain Grundstock XXX,
+  we can switch to a more efficient procedure: We just use mathematical
+  reasoning to deduce from known statements new ones.}
+
+  \jnote{3}{Irrespective of philosophical preferences, it's a fact of life that
+  most toposes only support constructive reasoning; in most toposes, proof by
+  contradiction is not valid. (Examples for toposes in which this is possible
+  are sheaf toposes~$\Sh(X)$ over discrete topological spaces, but not many
+  more than that.)
+
+  One might fear that most of mathematics breaks down in a constructive
+  setting. This is only true if interpreted naively: Often, already very small
+  changes to the definitions and statements suffice to make them constructively
+  valid (and are classically simply equivalent reformulations).
+  In other cases, we need to add interesting additional hypotheses --
+  hypotheses which are classically always satisfied. Here are a couple of
+  examples.
+
+  \begin{enumerate}\justifying
+    \item The usual proof that~$\sqrt{2}$ is not rational is perfectly fine
+    from a constructive point of view. It shows that the assumption
+    that~$\sqrt{2}$ is rational entails a contradiction. This is just the
+    definition of what it means to be not rational. (There's a difference
+    XXX Link Andrej
+    between a true proof by contradiction and a proof of a negated statement.
+    Only the former can't generally be carried out in constructive
+    mathematics.)
+    \item Constructively it's still true that there is no such thing as a
+    statement being neither true nor false: That is, we still
+    have~$\neg\neg(\varphi \vee \neg\varphi)$.
+  \end{enumerate}}
+
+  \jnote{4}{\begin{enumerate}\justifying\addtocounter{enumi}{2}
+    \item Constructively, we can't show that any inhabited subset of the
+    natural numbers has a minimal element. [We can also not show the
+    negation of that statement -- any valid constructive proof is a fortiori
+    a valid classical proof.] But we can show (quite
+    easily, by induction) that any inhabited and \emph{detachable} subset
+    of the natural numbers has a minimal element: A subset~$U \subseteq
+    \NN$ is detachable iff for any number~$n \in \NN$,~$n
+    \in U$ or~$n \not\in U$. Weakening the conclusion, we can also show that
+    any inhabited subset of the natural numbers does \emph{not not} have a
+    minimal element.
+
+    Both the failure and the two fixes can be interpreted computationally:
+    Given just the promise of an inhabited subset, we can't algorithmically
+    determine its minimum. But we can do so when given a \emph{membership
+    oracle}, or if it's okay to return the result in the
+    \emph{continuation monad}. XXX Link Ingo
+
+    \item We can't constructively prove that any finitely generated vector
+    space admits a basis. We can, however, constructively verify that any
+    finitely generated vector does \emph{not not} admit a basis. (By exploiting
+    that given a generating family~$(x_1,\ldots,x_n)$ it's \emph{not not} the
+    case that either one of the generators can be expressed as a linear
+    combination of the others, or not.) We'll see below that this particular
+    example implies that any sheaf of finite type over a reduced scheme is
+    locally free on a dense open.
+  \end{enumerate}}
+
+  \jnote{5}{\begin{enumerate}\justifying\addtocounter{enumi}{4}
+    \item We can't constructively verify the fundamental theorem of Galois
+    theory for arbitrary (not necessarily finite) Galois extensions in its
+    usual formulation. But if we pass from the topological Galois group
+    to the \emph{localic Galois group}, we can. Some aspects of the proof even
+    get simpler that way.
+
+    \item Similarly, we can't constructively verify the Gelfand--Naimark
+    correspondence between commutative~$C^*$-algebras with unit and compact Hausdorff
+    spaces. But we can do so if we pass from compact Hausdorff spaces to
+    compact Hausdorff locales.
+    XXX Naimark
+  \end{enumerate}
+
+  A recommendation for more on constructive mathematics is the informative and
+  entertaining talk \emph{Five Stages of Accepting Constructive Mathematics}
+  by Andrej Bauer (video, notes). XXX Link}
 
@@
     If~$\E \models \varphi$ and if~$\varphi$ entails~$\psi$
-    \pointthis{<3>}{constructively}{%
+    \pointthis{<3->}{constructively}{%
       no $\varphi \vee \neg\varphi$,\ \
@@
   \end{changemargin}
+
+  \jnote{1}{There is a variant of the effective topos which is not built using
+  Turing machines, but using \emph{infinite-time Turing machines}, a popular
+  model for hypercomputation. In that variant, the statement ``any
+  function~$\NN \to \NN$ is the zero function or not'' is true; the statement
+  ``any function~$\NN \to \NN$ is computable by a Turing machine'' is false;
+  and the statement~``any function~$\NN \to \NN$ is computable by an
+  infinite-time Turing machine'' is true again. Details can be found in
+  \fixedhref{...}{this set of slides}. XXX
+
+  \begin{center}\includegraphics[width=0.6\textwidth]{exploring-hypercomputation-slides}\end{center}}
+
+  \jnote{2}{We can also build a variant of the effective topos which uses
+  \emph{machines of the real world} instead of idealized Turing machines. In
+  doing so, we leave the realm of rigorous mathematics, but obtain interesting
+  connections with philosophy and physics. For instance, in that variant the
+  statement~``any function~$\RR \to \RR$ is continuous'' is true if machines in
+  the real world can only perform finitely many computational steps in finite
+  time and if it's possible to build hidden communication channels. Details can
+  be found in the book chapter \emph{\fixedhref{...}{Realizability for the Real
+  World}} by Andrej Bauer.
+  \begin{center}\includegraphics[width=0.3\textwidth]{zenil-computable-universe}\end{center}}
 \end{frame}
@@
   \end{array} \]
+
+  \jnote{1}{It's an instructive exercise to verify that~$\Sh(X) \models
+  \neg\neg\varphi$ if and only if there is a dense open subset~$U \subseteq X$
+  such that~$U \models \varphi$. This equivalence gives geometric meaning to
+  the failure of classical logic in~$\Sh(X)$.}
+\end{frame}
+
+\begin{frame}{Internalizing parameter-dependence}
+  \justifying
+  Let~$(f_x)_{x \in X}$ be a continuous family of continuous
+  functions (that is, let a continuous function~$X \times \RR \to \RR,\,(x,a) \mapsto
+  f_x(a)$ be given). From the internal point of view of~$\Sh(X)$, this family
+  looks like a single function~$f : \RR \to \RR$.
+  \begin{itemize}
+    \item $\Sh(X) \models (\text{The function $f : \RR \to \RR$ is
+    continuous})$.\medskip
+    \item Iff $f_x(-1) < 0$ for all $x \in X$, then $\Sh(X) \models f(-1) < 0$.
+    \item Iff $f_x(+1) > 0$ for all $x \in X$, then $\Sh(X) \models f(+1) > 0$.
+    \item Iff all $f_x$ are increasing, then $\Sh(X) \models (\text{$f$ is
+    increasing})$.\medskip
+    \item Iff there is an open cover~$X = \bigcup_i U_i$ such
+    that for each~$i$, there is a continuous function~$s : U_i \to \RR$
+    with~$f_x(s(x)) = 0$ for all~$x \in U_i$, then $\Sh(X) \models \exists s
+    \? \RR\_ f(s) = 0$.
+  \end{itemize}
+
+  \jnote{1}{Hence:
+  \begin{enumerate}\justifying
+    \item The standard formulation of the intermediate value theorem fails
+    in~$\Sh(X)$, because its external interpretation is that in continuous
+    families of continuous functions, zeros can locally be picked continuously.
+    That claim is false, as
+    \fixedhref{https://rawgit.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{this
+    counterexample} demonstrates.
+
+    As a corollary, we deduce that the standard formulation of the
+    intermediate value theorem is not constructively provable.
+
+    \item The approximative version of the intermediate value theorem (stating
+    that for any~$\varepsilon > 0$, there is a number~$x$ such that~$|f(x)| <
+    \varepsilon$) has a constructive proof and therefore holds in~$\Sh(X)$. The
+    external interpretation is that in continuous families of continuous
+    functions, approximate zeros can locally be picked continuously.
+    % XXX Link Matt F.
+
+    \item The monotone intermediate value theorem, stating that a strictly
+    increasing continuous function with opposite signs has a unique zero,
+    admits a constructive proof and therefore holds in~$\Sh(X)$. The external
+    interpretation is that in continuous families of strictly increasing
+    continuous functions, zeros can globally be picked continuously. You are
+    invited to prove this fact directly, without reference to the internal
+    language. This exercise isn't particularly hard, but it's not trivial either.
+  \end{enumerate}}
 \end{frame}
@@
   \end{itemize}
+
+  \jnote{1}{Differential geometry abounds with intuition which can't be
+  formalized in the standard account of differential geometry. For instance, we
+  can't literally define a tangent vector to be an infinitesimal piece of a
+  curve -- we have to resort to germs of smooth functions or derivations --
+  or speak about infinitesimal deformations of the identity transformation.
+
+  Synthetic differential geometry was born to provide an account of
+  differential geometry which is both rigorous and closer to intuition than the
+  standard account. Its starting point is the axiom of microaffinity, which is
+  very useful but wildly false in classical mathematics. The fundamental
+  theorem about synthetic differential geometry, connecting it with the
+  ordinary world of smooth manifolds, is that there exist \emph{well-adapted
+  models} for synthetic differential geometry -- toposes~$\E$ such that:
+  \begin{enumerate}\justifying
+    \item There is a functorial way of associating to any smooth manifold~$M$
+    an object~$y(M)$ of~$\E$ (something which~$\E$ believes to be a set) and
+    to any smooth map~$f : M \to N$ a morphism~$y(f) : y(M) \to y(N)$ of~$\E$
+    (something which~$\E$ believes to be map).
+    % (item is referenced below by number)
+    \item Any morphism in~$\E$ of type~$y(M) \to y(N)$ is
+    of the form~$y(f)$ for a smooth map~$f$, and if~$\E \models y(f) = y(g)$,
+    then actually~$f = g$.
+    \item In~$\E$, the axiom of microaffinity and related axioms hold for the
+    ring~$y(\RR^1)$.
+  \end{enumerate}}
+
+  \jnote{2}{Here are two examples for synthetic reasoning and synthetic
+  constructions.
+
+  \begin{itemize}\justifying
+    \item We can compute the derivative of~$f$ with~$f(x) = x^2$ as follows:
+    For all~$\varepsilon \in \Delta$, $f(x + \varepsilon) = (x + \varepsilon)^2
+    = x^2 + 2x\varepsilon$, hence~$f'(x) = 2x$ by the definition on the slide.
+
+    \item Write~``$M^\Delta$'' for the set of all maps~$\Delta \to M$. A
+    \emph{vector field} on a manifold (set)~$M$ is just a map~$X : M \to
+    M^\Delta$ such that~$X(p)(0) = p$ for all~$p \in M$. A tangent field~$X$
+    induces an infinitesimal path~$\gamma : \Delta \to M^M$ in the space (set)
+    of transformations of~$M$, by setting~$\gamma(\varepsilon) = (p \mapsto
+    X(p)(\varepsilon))$.
+
+    Conversely, a path~$\gamma : \Delta \to M^M$ such that~$\gamma(0) =
+    \mathrm{id}_M$ yields a vector field~$X : M \to M^\Delta$ by setting~$X(p) =
+    (\varepsilon \mapsto \gamma(\varepsilon)(p)$. No continuity or smoothness
+    checks have to be carried out.
+  \end{itemize}
+
+  There are also notions in synthetic differential geometry which
+  don't have a classical counterpart. For instance, in synthetic differential
+  geometry it's possible to view differential forms on~$M$ -- which are usually
+  thought of as \emph{functionals} on (exterior powers of) the tangent bundle
+  -- as \emph{quantities} (maps from~$M$ into a certain nonclassical object).}
+
+  \jnote{3}{Recall the following standard reflection principle taught in
+  undergraduate calculus: Even though continuity is officially a property which
+  depends only on the input/output behaviour of a function, reflecting on the
+  \emph{term} used to define a given function often allows us to quickly
+  conclude that it's continuous. For instance, any function defined by a
+  polynomial expression is continuous.
+
+  Synthetic differential geometry yields a new reflection principle, for
+  ordinary manifolds, which improves both on the assumption and on the
+  conclusion: Any map between manifolds which is definable in constructive
+  mathematics (so in particular, any map which is given by a polynomial
+  expression) is smooth (so in particular continuous). This is due to item~2
+  above.
+
+  Synthetic differential geometry is very well-developed and at your disposal.
+  References include:
+  \begin{itemize}
+    \item \fixedhref{...}{Notes for highschool students} (in German)
+    \item \fixedhref{...}{...}, a blog post by Andrej Bauer
+    \item \fixedhref{...}{...}, a short introduction by ... (Primer)
+    \item \fixedhref{...}{...}, the definitive book by Anders Kock
+  \end{itemize}
+
+  [Technical comment: The object~$y(\RR^1)$ will usually not be the object of
+  Dedekind real numbers in~$\E$.]}
 \end{frame}
@@
   \end{array} \]
+
+  \jnote{1}{Irrespective of whether~$A$ is a local ring, its mirror
+  image~$A^\sim$ is always a local ring (that is, the axioms of what it means
+  to be a local ring hold in~$\Spec(A)$).
+
+  A basic application of the internal language of~$\Spec(A)$ are
+  local-to-global principles. For instance:
+  \begin{itemize}\justifying
+    \item The statement ``the kernel of a surjective matrix over a local ring
+    is finite free'' admits a constructive proof. It therefore holds
+    in~$\Spec(A)$. Its external meaning is that the kernel of a surjective
+    matrix~$M$ over~$A$ is finite locally free (there exists a partition~$1 =
+    f_1 + \cdots + f_n$ such that for each~$i$, the localized
+    module~$(\operatorname{ker} M)[f_i^{-1}]$ is finite free).
+
+    \item The ring~$A$ is a Prüfer domain if and only if~$A^\sim$ is a Bézout
+    domain. Therefore any constructive theorem about Bézout domains entails a
+    corresponding theorem about Prüfer domains. Bézout domains are quite rare,
+    while Prüfer domains abound (for instance the ring of integers of any
+    number field is a Prüfer domain).
+  \end{itemize}}
+
+  \jnote{2}{More advanced applications are rendered possible by the observation
+  that, if~$A$ is a reduced ring ($x^n = 0 \Rightarrow x = 0$), its mirror
+  image~$A^\sim$ is \emph{anonymously Noetherian} (every ideal is \emph{not
+  not} finitely generated) and \emph{a field}. This fact doesn't have a
+  classical counterpart -- in general, neither~$A$ nor its stalks nor
+  its quotients nor its subrings are Noetherian or fields.
+
+  This reduction technique allows to give a new proof of
+  Grothendieck's generic freeness lemma which substantially improves on the
+  previously known proofs in length and clarity: from approximately three
+  pages (requiring several advanced prerequisites in commutative algebra) to a
+  single paragraph (requiring only the tiny bit of topos theory needed in order to
+  setup the internal language).
+
+  Details are \fixedhref{...}{here} (XXX link Munich slides).
+
+  As mentioned before, any proof involving the internal language can be
+  \emph{unwound} to yield a direct proof not referencing toposes. Depending on
+  the logical complexity of the statements appearing in the internal proof,
+  this process can substantially lengthen the proof. In the case of the new
+  proof of Grothendieck's generic freeness lemma, we were lucky to
+  \fixedhref{obtain a one-page proof}{...} using this process. (This crucially
+  rested on the fact that we were able to eliminate the use of the Noetherian
+  property from the internal proof; else the unwound proof would have been much
+  longer.)}
 \end{frame}

### 2018-07-10 slides-leipzig2018.tex
Finished the notes

@@
 \usetikzlibrary{mindmap,calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing}
-\hypersetup{colorlinks=false}
+\hypersetup{colorlinks=true}
 
@@
   \par
+
+  \jnote{1}{This talk has two parts. In the first part, we'll learn that
+  besides the standard mathematical universe, in which ordinary mathematics
+  takes place, there is a host of alternate mathematical universes. In these
+  alternate universes, the usual objects of mathematics enjoy slightly
+  different properties. For instance, we'll encounter universes in which the
+  intermediate value theorem fails or in which any map~$\RR \to \RR$ is
+  continuous.
+
+  In the second part, we'll see that these alternate universes, while seeming
+  strange on first contact, yield concrete applications in algebra and
+  geometry.}
 \end{frame}}
@@
   kinds of objects such as coverings. Toposes can even have no points at all
-  and still be nontrivial (this is for instance the case for the \emph{topos of
-  random sequences}). Secondly, while classically a given open
+  and still be nontrivial (this is for instance the case for the
+  \fixedhref{https://homepages.inf.ed.ac.uk/als/Talks/ccc09.pdf}{\emph{topos of
+  random sequences}}). Secondly, while classically a given open
   subset is either contained in a further open subset or not, the opens of
   toposes can be contained in further opens in many different ways.
-  % XXX: Link to Alex Simpson
-
-  Recently, toposes are being used to help the mysterious field with
-  one element come into being.}
-  % XXX: Link to John Baez
+
+  Recently, toposes are being used to help the mythical
+  \fixedhref{http://math.ucr.edu/home/baez/week259.html}{field with
+  one element} come into being.}
 
@@
   toposes even when they are not syntactially related in any way. This
-  observation is the starting point of Olivia Caramello's \emph{bridge
-  technique}, a grand research program with applications in many different
+  observation is the starting point of Olivia Caramello's
+  \fixedhref{https://www.oliviacaramello.com/}{\emph{bridge
+  technique}}, a grand research program with applications in many different
   fields.
-  % XXX: Link Olivia
 
@@
 
-  \jnote{2}{Besides~$\Set$, there is a proper class worth of further toposes.
+  \jnote{2}{Besides~$\Set$, there is a proper class' worth of further toposes.
   We'll get to know some of these better during the course of the
@@
   Most of the toposes in active use are either toposes of sheaves (over a
-  topological space or a XXXnLab \emph{site}), realizability toposes (such
-  as~$\Eff$ or variants constructed using different models of computation), or
-  arise from those using topos-theoretic constructions; but this is not a
-  complete classification.}
+  topological space or a
+  \fixedhref{https://ncatlab.org/nlab/show/site}{\emph{site}}), realizability
+  toposes (such as~$\Eff$ or variants constructed using different models of
+  computation), or arise from those using topos-theoretic constructions; but
+  this is not a complete classification. (The topos~$\Set$ is equivalent to the
+  topos of sheaves over the one-point space.)}
 \end{frame}
@@
   Kripke--Joyal semantics and check whether the translation holds in the usual
-  mathematical sense. As soon as we have established a certain Grundstock XXX,
-  we can switch to a more efficient procedure: We just use mathematical
-  reasoning to deduce from known statements new ones.}
+  mathematical sense. As soon as we have established a certain stock of
+  statements in this way, we can switch to a more efficient procedure: We can
+  just use mathematical reasoning to deduce new statements from known ones.}
 
@@
     that~$\sqrt{2}$ is rational entails a contradiction. This is just the
-    definition of what it means to be not rational. (There's a difference
-    XXX Link Andrej
-    between a true proof by contradiction and a proof of a negated statement.
+    definition of what it means to be not rational.
+    (\fixedhref{http://math.andrej.com/2010/03/29/proof-of-negation-and-proof-by-contradiction/}{There's a difference}
+    between a proof by contradiction and a proof of a negated statement.
     Only the former can't generally be carried out in constructive
@@
     \item Constructively it's still true that there is no such thing as a
-    statement being neither true nor false: That is, we still
+    statement which is neither true nor false: That is, we still
     have~$\neg\neg(\varphi \vee \neg\varphi)$.
@@
     determine its minimum. But we can do so when given a \emph{membership
-    oracle}, or if it's okay to return the result in the
-    \emph{continuation monad}. XXX Link Ingo
+    oracle}, or if it's
+    \fixedhref{https://rawgit.com/iblech/talk-constructive-mathematics/master/hal2015-notes.pdf}{okay to return the result in the
+    \emph{continuation monad}}.
 
@@
     space admits a basis. We can, however, constructively verify that any
-    finitely generated vector does \emph{not not} admit a basis. (By exploiting
-    that given a generating family~$(x_1,\ldots,x_n)$ it's \emph{not not} the
-    case that either one of the generators can be expressed as a linear
-    combination of the others, or not.) We'll see below that this particular
-    example implies that any sheaf of finite type over a reduced scheme is
-    locally free on a dense open.
+    finitely generated vector does \emph{not not} admit a basis, (By exploiting
+    that given a generating family~$(x_1,\ldots,x_n)$, it's \emph{not not} the
+    case that either one of the generators is a linear combination of the
+    others, or not.) We'll see below that this particular example entails that
+    any sheaf of finite type over a reduced scheme is locally free on a dense
+    open.
   \end{enumerate}}
@@
 
-    \item Similarly, we can't constructively verify the Gelfand--Naimark
+    \item Similarly, we can't constructively verify the Gelfand--Neumark
     correspondence between commutative~$C^*$-algebras with unit and compact Hausdorff
@@
     compact Hausdorff locales.
-    XXX Naimark
   \end{enumerate}
@@
   entertaining talk \emph{Five Stages of Accepting Constructive Mathematics}
-  by Andrej Bauer (video, notes). XXX Link}
+  by Andrej Bauer
+  (\fixedhref{https://www.youtube.com/watch?v=21qPOReu4FI}{video},
+  \fixedhref{http://www.ams.org/journals/bull/2017-54-03/S0273-0979-2016-01556-4/}{notes}).}
 
@@
       \item \intex{
+        $\Eff \models \text{``Any function~$\RR \to \RR$ is continuous.''}$
+      }{\textcolor{green!90}{\cmark}}{}
+
+      \item \intex{
         $\Sh(X) \models \text{``Any cont. function with opposite signs has
@@
   infinite-time Turing machine'' is true again. Details can be found in
-  \fixedhref{...}{this set of slides}. XXX
-
-  \begin{center}\includegraphics[width=0.6\textwidth]{exploring-hypercomputation-slides}\end{center}}
+  \fixedhref{https://rawgit.com/iblech/mathezirkel-kurs/master/superturingmaschinen/slides-warwick2017.pdf}{this set of slides}.
+
+  \begin{center}\fixedhref{https://rawgit.com/iblech/mathezirkel-kurs/master/superturingmaschinen/slides-warwick2017.pdf}{\includegraphics[width=0.6\textwidth]{exploring-hypercomputation-slides}}\end{center}}
 
@@
   time and if it's possible to build hidden communication channels. Details can
-  be found in the book chapter \emph{\fixedhref{...}{Realizability for the Real
-  World}} by Andrej Bauer.
-  \begin{center}\includegraphics[width=0.3\textwidth]{zenil-computable-universe}\end{center}}
+  be found in the book chapter
+  \emph{\fixedhref{http://math.andrej.com/2014/03/04/intuitionistic-mathematics-and-realizability-in-the-physical-world/}{Intuitionistic Mathematics and Realizability in the Physical World}} by Andrej Bauer.
+  \begin{center}\fixedhref{http://math.andrej.com/2014/03/04/intuitionistic-mathematics-and-realizability-in-the-physical-world/}{\includegraphics[width=0.3\textwidth]{zenil-computable-universe}}\end{center}}
 \end{frame}
@@
   \end{columns}
+
+  \jnote{1}{The applications of the internal language of toposes in several
+  branches of mathematics are driven by the following fundamental empirical fact:
+  \begin{center}The external translations of \\
+  some easy-to-prove internal statements of suitable toposes \\
+  are \\
+  nontrivial statements about the subject matter at hand.\end{center}
+  For instance, Grothendieck's generic freeness lemma, an important theorem in
+  commutative algebra and algebraic geometry, is the external translation of
+  the basic observation ``finitely generated modules over finitely generated
+  algebras over fields are \emph{not not} free over the field and \emph{not
+  not} finitely presented over the algebra'' of constructive linear algebra.
+
+  Is there a theorem which can only be proven by topos-theoretic methods?
+  \emph{No,} for any proof employing the internal language can be unwound to
+  yield an external proof not referencing topos theory. Just as the translation
+  from internal statements to external ones is entirely mechanical, so is the
+  translation from internal proofs to external ones. However, this translation
+  will always make the proof longer, and, depending on the
+  syntactical complexity of the involved statements, the resulting external
+  proof might be quite complex.
+
+  The next slide shows two examples.}
+
+  \jnote{2}{\begin{itemize}\justifying
+    \item The statement ``if the two outer sheaves in a short exact sequence of
+    sheaves of modules are of finite type, then so is the middle one'' is the
+    external translation of the basic fact ``if the two outer modules
+    in a short exact sequence of modules are finitely generated, then so is the
+    middle one'' of constructive linear algebra.
+
+    The syntactical complexity of the internal statement is quite low (just one
+    implication sign, on toplevel, no long chains of quantifiers of mixed
+    types). Therefore the resulting external proof obtained by unwinding will
+    still be quite short and straightforward. In fact, this external proof will
+    coincide with what anyone well-versed in scheme theory will produce,
+    judiciously juggling with open subsets.
+
+    Using the internal language of toposes here is therefore mostly for mental
+    hygiene, making rigorous the intuitive idea that the two statements are
+    related to each other.
+
+    \item A version of Hilbert's basis theorem, ``polynomial rings over
+    anonymously Noetherian rings are again anonymously Noetherian'', can be put
+    to use in the proof of Grothendieck's generic freeness lemma alluded to
+    above. This statement has nontrivial syntactical complexity (the
+    Noetherian condition involves a double negation, therefore nested
+    implications). For this reason, already the translation of the
+    \emph{statement} is quite convoluted (a part of it is
+    \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-como2018.pdf}{displayed
+    in this set of slides} for your viewing pleasure). The translation of its
+    \emph{proof} contains a deformed copy of the standard proof of Hilbert's
+    basis theorem (in a way which doesn't allow to simply \emph{apply}
+    Hilbert's basis theorem, one has to actually redo its proof).
+
+    In situations like this, the internal language is of real value.
+  \end{itemize}}
+
+  \jnote{3}{We're used to the fact that the usual laws of mathematical
+  reasoning apply to mathematical objects. The internal language of toposes
+  encapsulates and makes useful the following observation: The usual laws of
+  reasoning apply to our favorite mathematical objects \emph{also in
+  nontrivial, different ways}. I find this quite astonishing.
+
+  The next slide shows a basic application of this train of thought in
+  analysis. The presented example is of a rather simple nature and serves only
+  to explain the translation process in an explicit and rigorous fashion,
+  contrasting the previous slides which were somewhat short on details.
+  Details on the application to homotopy theory can be found in the
+  \fixedhref{https://homotopytypetheory.org/book/}{HoTT book}.}
 \end{frame}
@@
     that for any~$\varepsilon > 0$, there is a number~$x$ such that~$|f(x)| <
-    \varepsilon$) has a constructive proof and therefore holds in~$\Sh(X)$. The
+    \varepsilon$)
+    \fixedhref{https://mathoverflow.net/questions/253059/approximate-intermediate-value-theorem-in-pure-constructive-mathematics}{has
+    a constructive proof} and therefore holds in~$\Sh(X)$. The
     external interpretation is that in continuous families of continuous
     functions, approximate zeros can locally be picked continuously.
-    % XXX Link Matt F.
 
@@
     \emph{vector field} on a manifold (set)~$M$ is just a map~$X : M \to
-    M^\Delta$ such that~$X(p)(0) = p$ for all~$p \in M$. A tangent field~$X$
+    M^\Delta$ such that~$X(p)(0) = p$ for all~$p \in M$. A vector field~$X$
     induces an infinitesimal path~$\gamma : \Delta \to M^M$ in the space (set)
@@
   \begin{itemize}
-    \item \fixedhref{...}{Notes for highschool students} (in German)
-    \item \fixedhref{...}{...}, a blog post by Andrej Bauer
-    \item \fixedhref{...}{...}, a short introduction by ... (Primer)
-    \item \fixedhref{...}{...}, the definitive book by Anders Kock
+    \item \emph{\fixedhref{https://rawgit.com/iblech/mathezirkel-kurs/master/thema05-sdg/blatt05.pdf}{Synthetische
+    Differentialgeometrie}}, notes for high school students (in German)
+    \item \emph{\fixedhref{http://math.andrej.com/2008/08/13/intuitionistic-mathematics-for-physics/}{Intuitionistic
+    mathematics for physics}}, a blog post by Andrej Bauer
+    \item \emph{A Primer of Infinitesimal Analysis}, a short introduction by John Bell
+    \item \emph{\fixedhref{http://home.imf.au.dk/kock/sdg99.pdf}{Synthetic
+    Differential Geometry}}, the definitive book by Anders Kock
   \end{itemize}
@@
 
-\section[Commutative algebra]{Applications to commutative algebra}
+\section[Commutative algebra]{Applications in commutative algebra}
 
@@
     while Prüfer domains abound (for instance the ring of integers of any
-    number field is a Prüfer domain).
-  \end{itemize}}
+    number field is a Prüfer domain, even constructively so).
+  \end{itemize}
+
+  There are
+  \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-munich2018.pdf}{general
+  metatheorems} which state a precise connection between~$A$ and~$A^\sim$,
+  allowing in some cases to quickly pass from one to the other.}
 
@@
 
-  Details are \fixedhref{...}{here} (XXX link Munich slides).
-
-  As mentioned before, any proof involving the internal language can be
+  Sometimes, the effects of this topos-theoretic reduction technique can be
+  mimicked by classical techniques in commutative algebra such as passing
+  to quotient rings or stalks. To the best of my knowledge, that's not the case
+  for Grothendieck's generic freeness lemma. (In other cases, where they can,
+  the classical techniques require the axiom of choice, while the
+  topos-theoretic technique doesn't and therefore yields a more calculational,
+  informative proof.)
+
+  Details can be found
+  \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-munich2018.pdf}{in
+  this set of slides}.}
+
+  \jnote{3}{As mentioned before, any proof involving the internal language can be
   \emph{unwound} to yield a direct proof not referencing toposes. Depending on
@@
   proof of Grothendieck's generic freeness lemma, we were lucky to
-  \fixedhref{obtain a one-page proof}{...} using this process. (This crucially
-  rested on the fact that we were able to eliminate the use of the Noetherian
-  property from the internal proof; else the unwound proof would have been much
-  longer.)}
+  \fixedhref{https://arxiv.org/abs/1807.01231}{obtain a one-page proof} using
+  this process. (That the resulting external proof was so short is because it
+  was possible to eliminate the use of the Noetherian property from the
+  internal proof; without that elimination, the unwound proof was much longer.)}
 \end{frame}
-
-\backupstart
 
@@
   }
+
+  \jnote{1}{One doesn't need to be an expert in topos theory in order to know
+  that many notions in algebraic geometry are inspired by notions in algebra
+  and that proofs in algebraic geometry often proceed by reducing to algebra.
+  The internal language is a way of making this connection precise: In many
+  cases, the former are simply interpretations of the latter internal
+  to~$\Sh(X)$. Because this connection is precise instead of informal,
+  additional value is gained: We can skip many basic proofs in algebraic
+  geometry because they're just externalizations of proofs in algebra carried
+  out internally to~$\Sh(X)$.}
+
+  \jnote{2}{A basic example is as follows. A short exact sequence of sheaves of
+  modules looks like a short exact sequence of plain modules from the internal
+  point of view of~$\Sh(X)$. If the two outer sheaves are of finite type, then
+  from the internal point of view, the two outer modules will look like
+  finitely generated modules. Because the standard proof of the proposition
+  quoted on the lower right is constructively valid, it follows that, from
+  the internal point of view, the middle module is too finitely generated.
+  Consulting the dictionary a second time, this amounts to saying that the
+  middle sheaf is of finite type.
+
+  A more advanced example is: The theorem ``any finitely generated vector space
+  does \emph{not not} have a basis'' of constructive linear algebra entails, by
+  interpretation in~$\Sh(X)$, that any sheaf of finite type over a reduced
+  scheme is finite locally free on a \emph{dense open subset}.
+
+  More details on this research program can be found in
+  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
+  notes}, partly reported on at the
+  \fixedhref{https://www.youtube.com/watch?v=7S8--bIKaWQ}{2015 IHÉS topos
+  theory conference}. Even though many important dictionary entries are still
+  missing (for instance pertaining to derived categories and intersection
+  theory), I believe that it is already in its current form useful to working
+  algebraic geometers.
+
+  The next slide illustrates a further, different way of approaching algebraic
+  geometry using topos theory.}
 
@@
 
-  \hil{Synthetic approach:} model schemes \hil{directly as sets} in
+  \visible<2>{\hil{Synthetic approach:} model schemes \hil{directly as sets} in
   the internal universe of the \mbox{\hil{big Zariski topos}} of a base scheme.
   \small
-  \[ \{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]
+  \[ \{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]}
+
+  \jnote{1}{At the
+  \fixedhref{https://sbseminar.wordpress.com/2009/08/06/algebraic-geometry-without-prime-ideals/}{Secret
+  Blogging Seminar}, there was an insightful long-running discussion on the
+  merits of the two approaches. Two disadvantages of the approach using
+  locally ringed spaces is that the underlying topological spaces don't
+  actually parametrize ``honest'', ``geometric'' points, but the more complex
+  notion of irreducible closed subsets; and that they don't work well in a
+  constructive setting. (For this, they would have to be replaced by locally
+  ringed locales.)
+
+  The functorial approach is more economical, philosophically rewarding, and
+  works constructively. Given a functor~$F : \Ring \to \Set$, we imagine~$F(A)$
+  to be the set of~``$A$-valued points'' of the hypothetical scheme described
+  by~$F$, the set of ``points with coordinates in~$A$''. These sets have direct
+  geometric meaning. However, typically only field-valued points are easy
+  to describe. For instance, the functor representing projective~$n$-space is
+  given on fields by
+  \[ \begin{array}{r@{}c@{}l}
+  K &{}\longmapsto{}& \text{the set of lines through the origin in~$K^{n+1}$} \\
+  && \qquad \cong \{ [x_0 \hg \cdots \hg x_n] \,|\, \text{$x_i \neq 0$ for some~$i$} \},
+  \end{array} \]
+  whereas on general rings it is given by
+  \[ \begin{array}{r@{}c@{}l}
+    A &{}\longmapsto{}& \text{the set of quotients~$A^{n+1} \twoheadrightarrow P$,
+    where~$P$ is projective,} \\
+    && \qquad \text{modulo isomorphism}.
+  \end{array} \]
+  It is these more general kinds of points which impart a sense of
+  cohesion on the field-valued points, so they can't simply be dropped from
+  consideration.\bigskip}
+
+  \jnote{2}{This tension is resolved by observing that the category of
+  functors~$\Ring \to \Set$ is a topos (the \emph{big Zariski topos} of~$\Spec(\ZZ)$)
+  and that we can therefore employ its internal language. This language takes
+  care of juggling stages behind the scenes. For instance, projective~$n$-space
+  can be described by the naive expression
+  \[ \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee \cdots \vee x_n
+  \neq 0 \}/(\affl)^\times. \]
+  This example illustrates the goal: to develop a synthetic account of
+  algebraic geometry, in which schemes are plain sets and morphisms between
+  schemes are maps between those sets. It turns out that there are many
+  similarities with the well-developed synthetic account of differential
+  geometry, but also important differences, and it also turns out that
+  synthetic algebraic geometry has close connections to a certain age-old
+  unresolved problem in topos theory, the \emph{mystery of nongeometric
+  sequents}.
+
+  Details are in
+  \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-como2018.pdf}{this
+  set of slides}.}
 \end{frame}
 
-\backupend
-
 \end{document}

### 2018-07-10 slides-como2018.tex
typo

@@
   conference}. Even though many important dictionary entries
-  are still missing (for instance pertaining derived categories and
+  are still missing (for instance pertaining to derived categories and
   intersection theory), I believe that it is already at its current stage

### 2018-07-10 slides-leipzig2018.tex
typo

@@
 
-\newcommand{\bignumber}[1]{
-  \renewcommand{\insertenumlabel}{#1}\scalebox{1.5}{\usebeamertemplate{enumerate item}}
-}
-
 \newenvironment{indentblock}{%
@@
 \newcommand{\hil}[1]{{\usebeamercolor[fg]{item}{\textbf{#1}}}}
-
-\newcommand{\bad}[1]{\textcolor{red!90}{\textnormal{#1}}}
 
@@
   Crucially, two theories~$T$ and~$T'$ can have equivalent classifying
-  toposes even when they are not syntactially related in any way. This
+  toposes even when they are not syntactically related in any way. This
   observation is the starting point of Olivia Caramello's
@@
   being discrete, the intermediate value theorem in the form
-  \begin{indentblock}Let~$f : \RR \to \RR$ be a continuous function.
+  \begin{indentblock}\emph{Let~$f : \RR \to \RR$ be a continuous function.
   Assume~$f(-1) < 0 < f(1)$. Then there is a number~$x \in \RR$ such
-  that~$f(x) = 0$.\end{indentblock}
+  that~$f(x) = 0$.}\end{indentblock}
   fails in these toposes -- for very meaningful reasons discussed below.
   The following, classically equivalent, version does hold:
-  \begin{indentblock}Let~$f : \RR \to \RR$ be a continuous function.
+  \begin{indentblock}\emph{Let~$f : \RR \to \RR$ be a continuous function.
   Assume~$f(-1) < 0 < f(1)$. Then, for every~$\varepsilon > 0$, there is
-  a number~$x \in \RR$ such that~$|f(x)| < \varepsilon$.\end{indentblock}
+  a number~$x \in \RR$ such that~$|f(x)| < \varepsilon$.}\end{indentblock}
 
@@
   \jnote{1-2}{We can picture the Kripke--Joyal semantics as a kind of translation
-  engine. When we're talking with Anna, a mathematician who lives in the
-  ef{}fective topos, at first we might be weirded out when she states ``it's a basic
+  engine. When we're talking over the phone with Anna, a mathematician who lives in the
+  ef{}fective topos, at first we might feel uncomfortable when she states ``it's a basic
   fact of life that any function~$\NN \to \NN$ is computable''. But if we
@@
   (\fixedhref{https://www.youtube.com/watch?v=21qPOReu4FI}{video},
-  \fixedhref{http://www.ams.org/journals/bull/2017-54-03/S0273-0979-2016-01556-4/}{notes}).}
+  \fixedhref{https://www.ams.org/journals/bull/2017-54-03/S0273-0979-2016-01556-4/}{notes}).}
 
@@
   branches of mathematics are driven by the following fundamental empirical fact:
+  \vspace*{-1em}
   \begin{center}The external translations of \\
@@
   nontrivial statements about the subject matter at hand.\end{center}
+  \vspace*{-1em}
   For instance, Grothendieck's generic freeness lemma, an important theorem in
@@
 
-  The next slide shows two examples.}
-
-  \jnote{2}{\begin{itemize}\justifying
+  The next two slides show examples regarding this increase in complexity.}
+
+  \jnote{2}{\begin{enumerate}\justifying
     \item The statement ``if the two outer sheaves in a short exact sequence of
@@
     The syntactical complexity of the internal statement is quite low (just one
-    implication sign, on toplevel, no long chains of quantifiers of mixed
+    implication sign, at toplevel, no long chains of quantifiers of mixed
     types). Therefore the resulting external proof obtained by unwinding will
@@
     coincide with what anyone well-versed in scheme theory will produce,
-    judiciously juggling with open subsets.
-
-    Using the internal language of toposes here is therefore mostly for mental
-    hygiene, making rigorous the intuitive idea that the two statements are
-    related to each other.
-
+    judiciously juggling open subsets.
+
+    Using the internal language of toposes in situations like this is therefore
+    mostly for mental hygiene, making rigorous the intuitive idea that the two
+    statements are related to each other.
+  \end{enumerate}}
+
+  \jnote{3}{\begin{enumerate}\justifying\addtocounter{enumi}{1}
     \item A version of Hilbert's basis theorem, ``polynomial rings over
@@
     to use in the proof of Grothendieck's generic freeness lemma alluded to
-    above. This statement has nontrivial syntactical complexity (the
+    above.
+
+    This statement has nontrivial syntactical complexity (the
     Noetherian condition involves a double negation, therefore nested
@@
     In situations like this, the internal language is of real value.
-  \end{itemize}}
-
-  \jnote{3}{We're used to the fact that the usual laws of mathematical
+  \end{enumerate}}
+
+  \jnote{4}{We're used to the fact that the usual laws of mathematical
   reasoning apply to mathematical objects. The internal language of toposes
@@
   reasoning apply to our favorite mathematical objects \emph{also in
-  nontrivial, different ways}. I find this quite astonishing.
+  nontrivial ways different from the accustomed one}. I find this quite astonishing.
 
@@
   Details on the application to homotopy theory can be found in the
-  \fixedhref{https://homotopytypetheory.org/book/}{HoTT book}.}
+  \fixedhref{https://homotopytypetheory.org/book/}{HoTT book}.
+
+  A tour of applications in algebraic geometry can be found in
+  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
+  notes}.}
 \end{frame}
@@
   differential geometry which is both rigorous and closer to intuition than the
-  standard account. Its starting point is the axiom of microaffinity, which is
-  very useful but wildly false in classical mathematics. The fundamental
+  standard account, allowing for instance to read the classic works of Sophus
+  Lie in a literal manner.
+
+  Its starting point is the axiom of microaffinity, which is
+  quite useful but wildly false in classical mathematics. The fundamental
   theorem about synthetic differential geometry, connecting it with the
-  ordinary world of smooth manifolds, is that there exist \emph{well-adapted
-  models} for synthetic differential geometry -- toposes~$\E$ such that:
+  ordinary world of smooth manifolds, states that there exist \emph{well-adapted
+  models} for synthetic differential geometry -- toposes~$\E$ such that:\par
+  \scriptsize
   \begin{enumerate}\justifying
@@
     % (item is referenced below by number)
-    \item Any morphism in~$\E$ of type~$y(M) \to y(N)$ is
+    \item Any morphism of~$\E$ of type~$y(M) \to y(N)$ is
     of the form~$y(f)$ for a smooth map~$f$, and if~$\E \models y(f) = y(g)$,
@@
     \item In~$\E$, the axiom of microaffinity and related axioms hold for the
-    ring~$y(\RR^1)$.
+    ring~$y(\RR^1)$. [Technical comment: This ring will usually not be the ring of
+    Dedekind real numbers in~$\E$.]
   \end{enumerate}}
@@
 
-  \jnote{3}{Recall the following standard reflection principle taught in
+  \jnote{3}{The following reflection principle is routinely used in
   undergraduate calculus: Even though continuity is officially a property which
@@
     Differential Geometry}}, the definitive book by Anders Kock
-  \end{itemize}
-
-  [Technical comment: The object~$y(\RR^1)$ will usually not be the object of
-  Dedekind real numbers in~$\E$.]}
+  \end{itemize}}
 \end{frame}
@@
   the logical complexity of the statements appearing in the internal proof,
-  this process can substantially lengthen the proof. In the case of the new
+  this process can substantially lengthen the proof.
+
+  In the case of the new
   proof of Grothendieck's generic freeness lemma, we were lucky to
@@
   synthetic algebraic geometry has close connections to a certain age-old
-  unresolved problem in topos theory, the \emph{mystery of nongeometric
+  unsolved problem in topos theory, the \emph{mystery of nongeometric
   sequents}.

### 2018-07-10 slides-leipzig2018.tex
typo

@@
 
-    \item The approximative version of the intermediate value theorem (stating
+    \item The \emph{approximate} intermediate value theorem (stating
     that for any~$\varepsilon > 0$, there is a number~$x$ such that~$|f(x)| <
@@
 
-    \item The monotone intermediate value theorem, stating that a strictly
+    \item The \emph{monotone} intermediate value theorem, stating that a strictly
     increasing continuous function with opposite signs has a unique zero,

### 2018-08-31 paper-macneille-reals.tex
typo

@@
 second might fail -- for any of the several possible flavors of the reals
-which in constructive mathematics might not coincide. Therefore a different
-approach is needed.
+such as the Cauchy reals, the Dedekind reals or the MacNeille reals (which all
+coincide in classical mathematics, but might differ in constructive
+mathematics). Therefore a different approach is needed.
 
-This note gives a constructive proof that a specific flavor of the reals,
-namely the MacNeille reals, is uncountable. To the best of our knowledge, this
+This note gives a constructive proof that one of these flavors,
+the MacNeille reals, is uncountable. To the best of our knowledge, this
 is the first result in that direction. However, it is just a baby step towards
-an understanding whether any of the more interesting flavors of the reals --
-for instance the Cauchy reals or the Dedekind reals -- can constructively be
+an understanding whether any of the more interesting flavors of the reals can constructively be
 shown to be uncountable. The proof presented here is made possible because the

### 2018-10-03 paper-flabby-objects.tex
typo

@@
 cohomology:
+% XXX: Je nachdem, wie die Einleitung aussieht, hier die Betonung anders
+% setzen. Fundamental ist nur, dass man welke Auflösungen zur Berechnung von
+% Garbenkohomologie verwenden kann und dass sie von f_* bewahrt werden. Die
+% restlichen Eigenschaften sind wichtig, aber nicht fundamental.
 \begin{enumerate}
@@
 There is a substitute for property~(3a) which can be verified purely
-intuitionistically. We comment on the status of property~(4b) in
+intuitionistically. We do not know whether property~(4b) can be established
+purely intuitionistically, but we give a rudimentary analysis in
 Section~\ref{sect:conclusion}.
@@
 is stable under pullback and
-can be characterized in the internal language. We therefore adapt in this paper the
+can be characterized in the internal language. We therefore adopt in this paper the
 alternative one as the official definition.
@@
 
-\begin{prop}\label{prop:set-of-extensions-flabby}
-Let~$I$ be an injective set. Let~$i : A \to B$ be an injection.
+\begin{lemma}\label{lemma:set-of-extensions-flabby}
+\begin{enumerate}
+\item Let~$I$ be an injective set. Let~$i : A \to B$ be an injection.
 Let~$f : A \to I$ be an arbitrary map. Then the set of extensions of~$f$ to~$B$
-is flabby.\end{prop}
+is flabby.
+\item Let~$I$ be an injective~$R$-module. Let~$i : A \to B$ be a linear injection.
+Let~$f : A \to I$ be an arbitrary linear map. Then the set of linear extensions of~$f$ to~$B$
+is flabby.
+\end{enumerate}
+\end{lemma}
 
-\begin{proof}Set~$X \defeq \{ \bar{f} \in I^B \,|\, \bar{f} \circ i =
+\begin{proof}For the first claim, we set~$X \defeq \{ \bar{f} \in I^B \,|\, \bar{f} \circ i =
 f \}$. Let~$K \subseteq X$ be a subsingleton. We consider the injectivity diagram
@@
 vertical arrow~$g$ is defined in the following way: Let~$s \in i[A] \cup B'$.
-If~$s \in i[A]$, then we set~$g(s) \defeq f(a)$, where~$a \in A$ is an element such
-that~$s = i(a)$. If~$s \in B'$, then we set~$g(s) \defeq \bar{f}(s)$,
+If~$s \in i[A]$, we set~$g(s) \defeq f(a)$, where~$a \in A$ is an element such
+that~$s = i(a)$. If~$s \in B'$, we set~$g(s) \defeq \bar{f}(s)$,
 where~$\bar{f}$ is any element of~$K$. These prescriptions determine a well-defined
@@
 then this map is an element of~$K$.
-\end{proof}
 
-\begin{prop}\label{prop:set-of-linear-extensions-flabby}
-Let~$I$ be an injective~$R$-module. Let~$i : A \to B$ be a linear injection.
-Let~$f : A \to I$ be an arbitrary linear map. Then the set of linear extensions of~$f$ to~$B$
-is flabby.\end{prop}
-
-\begin{proof}The proof is similar to the proof of
-Proposition~\ref{prop:set-of-extensions-flabby}. We
+The proof of the second claim is similar. We
 set~$X \defeq \{ \bar{f} \in \Hom_R(B,I) \,|\, \bar{f} \circ i =
@@
 \begin{enumerate}
-\item If~$X$ is flabby, so is~$X \times T$ as an object in~$\E/T$.
+\item If~$X$ is flabby, so is~$X \times T$ as an object of~$\E/T$.
 \item The converse holds if the unique morphism~$T \to 1$ is an epimorphism.
@@
 
-\begin{proof}The proof is routine; we only verify ``only if'' direction.
-Let~$F$ be flabby as a sheaf. It suffices to verify the definition for stages
+\begin{proof}The proof is routine; we only verify the ``only if'' direction.
+Let~$F$ be flabby as a sheaf. It suffices to verify the definining condition for stages
 of the form~$A = \Hom(\cdot,U)$, where~$U$ is an open of~$X$. A monomorphism~$K
@@
 \begin{prop}\label{prop:global-elements}
-Let~$X$ be a flabby object in an elementary topos~$\E$. If~$\E$ is localic and
+Let~$X$ be a flabby object of a localic topos~$\E$. If
 Zorn's lemma is available in the metatheory, then~$X$ possesses a global element (a morphism~$1 \to X$).
@@
 Global elements of~$F$ are extensions of the kind we are looking for.
-By Proposition~\ref{prop:set-of-extensions-flabby}, this object is flabby.
+By Lemma~\ref{lemma:set-of-extensions-flabby}, this object is flabby.
 By Proposition~\ref{prop:global-elements}, it has a global element.
@@
 carries over word for word, only referencing
-Proposition~\ref{prop:set-of-extensions-flabby} instead of
-Proposition~\ref{prop:set-of-linear-extensions-flabby}.
+Lemma~\ref{lemma:set-of-extensions-flabby}(2) instead of
+Lemma~\ref{lemma:set-of-extensions-flabby}(1).
 It seems that Harting was not aware of this, even though she did show that
@@
 \begin{prop}\label{prop:semienough-flabby-modules}
-Assuming the law of excluded middle, any~$\neg\neg$-separated
+Assuming the law of excluded middle in the metatheory, any~$\neg\neg$-separated
 module in the effective topos can be embedded into a flabby module.
 \end{prop}
+
+% Δ(X) ist welk, falls X bewohnt.
+% Δ(X) ist separiert.
+% Ist A separierter Modul, so lässt sich A in einen welken Modul einbetten
+% (der seinerseits separiert ist, er ist nämlich Δ(Γ(A))).
+% Sind welke Mengen separiert? Nein: Omega.
+% Sind welke Moduln separiert? Nein: Ich denke mittlerweile, dass sich jeder
+% Modul in einen welken einbetten lässt, über etwas, was wie eine induktive
+% Konstruktion aussieht.
+
+% Rem.: diskret =/=> welk, sep =/=> welk (Bsp.: konstante Garbe Z auf Mnf.)
+% Umkehrung auch nicht (Bsp.: Godemontkonstruktion, Omega)
 
@@
 \begin{proof}[Proof of Proposition~\ref{prop:flabby-effective-sets}]
-For any Turing machine~$e$, let~$v_e$ be the set
+For any Turing machine~$e$, let~$v_e : |X| \to \Sigma$ be the nonstandard
+predicate given by
 \begin{multline*}
-  v_e = \{ m \in \NN \,|\,
+  v_e(x) = \{ m \in \NN \,|\,
   \text{there is an element~$x_0 \in |X|$ such that} \\
@@
 and let~$K_e \in \Sigma^{|X| \times \Sigma}$ be the nonstandard predicate given by
-\[ K_e(x,u) = \llbracket (x = x) \wedge (u \leftrightarrow v_e) \rrbracket. \]
+\[ K_e(x,u) = \llbracket (x = x) \wedge (u \leftrightarrow v_e(x)) \rrbracket. \]
 One can explicitly construct a realizer~$a_e$ of the statement~``$K_e = K_e$'',
@@
 \begin{proof}[Proof of Proposition~\ref{prop:semienough-flabby-modules}]
-Let~$(\Delta \dashv \Gamma) : \Set \to \Eff$ be the inclusion of the
+Let~$(\Gamma \dashv \Delta) : \Set \to \Eff$ be the inclusion of the
 double-negation sheaves. For a~$\neg\neg$-separated module~$M$ in the effective

### 2018-10-03 paper-flabby-objects.tex
typo

@@
 the finest equivalence relation on~$T$ such that~$t + (s + u) \sim (t + s) +
-u$, $t + s \sim s + t$, $t + 0 \sim t \sim 0 + t$, $1t \sim t$, $r(t+s) \sim rt
+u$, $t + s \sim s + t$, $t + 0 \sim t \sim 0 + t$, $0t \sim 0$, $1t \sim t$, $r(t+s) \sim rt
 + rs$, $(r+r')t \sim rt + r't$ and such

### 2018-10-03 slides-pssl104.tex
Beginning of PSSL104 slides

--- 2018-07-10 slides-como2018.tex
@@
 \usepackage{etex}
-
-\usepackage{pgfpages}
-\setbeameroption{show notes on second screen}
-\setbeamertemplate{note page}[plain]
-\newcommand{\jnote}[2]{\only<#1>{\note{\justifying#2\par}}}
-
-% Workaround for the issue described at
-% https://tex.stackexchange.com/questions/164406/beamer-using-href-in-notes.
-\newcommand{\fixedhref}[2]{\makebox[0pt][l]{\hspace*{\paperwidth}\href{#1}{#2}}\href{#1}{#2}}
 
@@
 \hypersetup{colorlinks=true}
+\usepackage{multimedia}
+\newcommand{\video}[2]{\movie[width=#2,height=#2,autostart,loop,poster]{}{#1}}
+\hypersetup{colorlinks=false}
+
+\usepackage{pifont}
+\newcommand{\cmark}{\ding{51}}
+\newcommand{\xmark}{\ding{55}}
+\DeclareSymbolFont{extraup}{U}{zavm}{m}{n}
+\DeclareMathSymbol{\varheart}{\mathalpha}{extraup}{86}
 
@@
 
-\title{How topos theory can help commutative algebra}
+\title{How not to constructivize cohomology}
 \author{Ingo Blechschmidt}
-\date{June 29th, 2018}
+\date{October 6th, 2018}
 
 \useinnertheme[shadow=true]{rounded}
-\useoutertheme[subsection=false]{miniframes}
 \setbeamerfont{block title}{size={}}
@@
 \newcommand{\A}{\mathcal{A}}
-\newcommand{\B}{\mathcal{B}}
 \renewcommand{\AA}{\mathbb{A}}
-\renewcommand{\C}{\mathcal{C}}
 \newcommand{\E}{\mathcal{E}}
 \newcommand{\F}{\mathcal{F}}
-\newcommand{\M}{\mathcal{M}}
 \renewcommand{\G}{\mathcal{G}}
-\newcommand{\J}{\mathcal{J}}
 \newcommand{\GG}{\mathbb{G}}
@@
 \renewcommand{\P}{\mathcal{P}}
-\newcommand{\aaa}{\mathfrak{a}}
-\newcommand{\fff}{\mathfrak{f}}
 \newcommand{\ppp}{\mathfrak{p}}
-\newcommand{\mmm}{\mathfrak{m}}
 \newcommand{\defeq}{\vcentcolon=}
@@
 \newcommand{\Set}{\mathrm{Set}}
-\newcommand{\Ring}{\mathrm{Ring}}
-\newcommand{\LocRing}{\mathrm{LocRing}}
 \newcommand{\Eff}{\mathrm{Ef{}f}}
@@
 \newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
-\newcommand{\Ll}{\text{iff}}
+\newcommand{\Ll}{\vcentcolon\!\Longleftrightarrow}
 \newcommand{\inv}{inv.\@}
 \newcommand{\seq}{\vdash_{\!\!\!\vec x}}
-\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates
 
@@
 }
+\newcommand{\bigheart}{\includegraphics{heart}}
 
-\newenvironment{indentblock}{%
-  \list{}{\leftmargin\leftmargin}%
-  \item\relax
-}{%
-  \endlist
-}
-
-\newcommand{\hcancel}[5]{%
-  \tikz[baseline=(tocancel.base)]{
-    \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
-    \draw[red, line width=0.4mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
+\newenvironment{changemargin}[2]{%
+  \begin{list}{}{%
+    \setlength{\topsep}{0pt}%
+    \setlength{\leftmargin}{#1}%
+    \setlength{\rightmargin}{#2}%
+    \setlength{\listparindent}{\parindent}%
+    \setlength{\itemindent}{\parindent}%
+    \setlength{\parsep}{\parskip}%
   }%
-}
+  \item[]}{\end{list}}
 
@@
 
+\setbeamertemplate{headline}{%
+  \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex]{}%
+    \insertsectionnavigationhorizontal{\paperwidth}{}{}%
+  \end{beamercolorbox}%
+  \vskip0pt%
+}
+
 \setbeamertemplate{frametitle}{%
-  \vskip0.7em%
+  \vskip1.2em%
   \leavevmode%
@@
 
-\setbeamertemplate{headline}{%
-  \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex]{}%
-    \insertnavigation{\paperwidth}%
-  \end{beamercolorbox}%
-  \vskip0pt%
-}
-
+\newcommand{\insertframeextra}{}
 \setbeamertemplate{footline}{%
@@
     \hfill
-    \insertframenumber\,/\,\inserttotalframenumber
+    \insertframenumber\insertframeextra\,/\,\inserttotalframenumber
   \end{beamercolorbox}%
@@
 \newcommand{\hil}[1]{{\usebeamercolor[fg]{item}{\textbf{#1}}}}
-
-\newcommand{\bad}[1]{\textcolor{red!90}{\textnormal{#1}}}
+\newcommand{\bad}[1]{\textcolor{red!90}{\textbf{#1}}}
 
 \begin{document}
-
-%\addtocounter{framenumber}{-1}
-%{\usebackgroundtemplate{\includegraphics[width=\paperwidth]{slides-como2018-titlepage-blurred}}
-%\begin{frame}[plain]\end{frame}}
 
@@
   \centering
-  \medskip
-  \hspace*{-3.9em}%
-
-  \hil{How topos theory can help commutative algebra} \\
-
-  \emph{-- an invitation --}
 
   \bigskip
+  \includegraphics[width=0.2\textwidth]{sheaf}
   \bigskip
 
-  \begin{columns}
-    \small
-    \begin{column}{0.33\textwidth}
-      \centering
-      \bignumber{1}
-      \smallskip
-
-      \hil{Background on the internal language}
-      \medskip
-    \end{column}
-
-    \begin{column}{0.33\textwidth}
-      \centering
-      \bignumber{2}
-      \smallskip
-
-      \hil{Applications in commutative algebra}
-      \medskip
-    \end{column}
-
-    \begin{column}{0.33\textwidth}
-      \centering
-      \bignumber{3}
-      \smallskip
-
-      \hil{The mystery of \mbox{\!\!nongeometric sequents}}
-      \medskip
-    \end{column}
-  \end{columns}
+  \hil{How not to constructivize cohomology}
 
   \scriptsize
-  Ingo Blechschmidt (MPI Leipzig) \\
-  Toposes in Como \\
-  June 29th, 2018
+  \textit{-- interruptions welcome at any point --}
+  \bigskip
+
+  Ingo Blechschmidt \\
+  University of Verona
+  \bigskip
+
+  104th Peripatetic Seminar on Sheaves and Logic in Amsterdam \\
+  October 6th, 2018
   \par
@@
 
-\section{Motivation}
+\section{Sheaf cohomology}
 
-\begin{frame}{Motivating test cases}
-  Let $A$ be a ring (commutative, with unit, $1 = 0$ allowed). \\
-  Assume that $A$ is reduced: If $x^n = 0$, then $x = 0$.
-
-  \jnote{1-2}{The two displayed statements are trivial for fields. It is
-  therefore natural to try to reduce the general situation to the field
-  situation.}
-
-  \jnote{2}{The displayed proofs, which could have been taken from any
-  standard textbook on commutative algebra, succeed in this reduction quite
-  easily by employing maximal ideals or minimal prime ideals. However, this way
-  of reducing comes at a cost: It requires the Boolean Prime Ideal Theorem
-  (for ensuring the existence of a prime ideal and for ensuring that stalks at
-  minimal prime ideals are fields) and even the full axiom of choice (for
-  ensuring the existence of a minimal prime ideal).
-
-  It therefore doesn't work in the internal universe of most toposes, and in
-  any case it obscures explicit computational content: Statements so simple as
-  the two displayed ones should admit explicit, computational proofs.
-
-  We'll learn how the internal language of a certain well-chosen topos provides
-  a way to perform the reduction in an entirely constructive manner. If so
-  desired, the resulting topos-theoretic proofs can be unwound to yield fully
-  explicit, topos-free, direct proofs.
-
-  Beautiful constructive proofs can also be found in Richman's note on
-  \fixedhref{https://www.ams.org/journals/proc/1988-103-04/S0002-9939-1988-0954974-5/S0002-9939-1988-0954974-5.pdf}{nontrivial
-  uses of trivial rings} and in the
-  \fixedhref{https://arxiv.org/abs/1605.04832}{recent textbook by Lombardi and
-  Quitté} on constructive commutative algebra.}
-
-  \jnote{3}{We can slightly reduce the requirements of the proof of the first
-  statement by employing not a maximal ideal, but a prime ideal. The existence
-  of maximal ideals in nontrivial rings is equivalent to the axiom of choice,
-  while the existence of prime ideals is equivalent to the weaker Boolean Prime
-  Ideal Theorem. However, this improvement doesn't change the main point; the
-  proof is still wildly nonconstructive.}
-
-  \jnote{4-}{Grothendieck's generic freeness lemma is an important theorem in
-  algebraic geometry, where it is usually stated in the following geometric
-  form:
-  \begin{indentblock}Let~$X$ be a reduced scheme. Let~$\B$ be
-  an~$\O_X$-algebra of finite type. Let~$\M$ be a~$\B$-module of finite type.
-  Then over a dense open,
-  \begin{enumerate}
-  \item[(a)] $\B$ and~$\M$ are locally free as sheaves of~$\O_X$-modules,
-  \item[(b)] $\B$ is of finite presentation as a sheaf of~$\O_X$-algebras and
-  \item[(c)] $\M$ is of finite presentation as a sheaf of~$\B$-modules.
-  \end{enumerate}\end{indentblock}}
-
-  \jnote{5}{All previously known proofs proceed in a series of reduction steps,
-  finally culminating in the case where~$A$ is a Noetherian integral domain.
-  They are somewhat convoluted (for instance, the proof in the Stacks Project
-  is three pages long) and employ several results in commutative algebra which
-  have not yet been constructivized.
-
-  Using the internal language of toposes, we will give a short, conceptual and
-  constructive proof of Grothendieck's generic freeness lemma. Again, if so
-  desired, one can unwind the internal proof to obtain a
-  constructive proof which doesn't reference topos theory. The proof obtained
-  in this way is still an improvement on the previously known proofs, requiring
-  no advanced prerequisites in commutative algebra, and takes
-  \fixedhref{https://arxiv.org/abs/1807.01231}{about
-  a page}.}
-
-  \only<1-3>{\begin{columns}[t]
-  \begin{column}[t]{0.50\textwidth}
-    \centering
-
-    \scalebox{0.8}{$\begin{pmatrix}
-      \cdot & \cdot \\
-      \cdot & \cdot \\
-      \cdot & \cdot
-      \end{pmatrix}$}
-    \vspace*{-1em}
-
-    \begin{varblock}{\textwidth}{A baby application}
-      \justifying
-      Let~$M$ be a surjective matrix over~$A$ with more rows than columns.
-      Then~$1 = 0$ in~$A$.
-    \end{varblock}
+\begin{frame}
+  \begin{columns}
+    \begin{column}{0.5\textwidth}
+      \centering{\usebeamercolor[fg]{item}{\textbf{\Large Sheaf cohomology}}}\bigskip
 
       \justifying
-    \only<2>{\textbf{Proof.} \bad{Assume not.} Then there is~a \bad{maximal
-    ideal} $\mmm$. The matrix is surjective over the field~$A/\mmm$. This is a
-    contradiction to basic linear algebra.}
+      Let~$E$ be a sheaf of modules over a space~$X$.
+      Choose an \bad{injective resolution} $0 \to E \to I^0 \to I^1 \to \cdots$.
+      Then $H^n(X; E)$ is the $n$-th cohomology of
+      \[ (0 \xrightarrow{\phantom{d^0}} I^0(X) \xrightarrow{d^0} I^1(X) \xrightarrow{d^1} \cdots), \]
+      that is~$\operatorname{ker}(d^n)\mathop{/}\operatorname{im}(d^{n-1})$.
+    %  \begin{align*}
+    %    H^n(X; E) &\defeq \text{$n$-th cohomology of $(0 \to I^0(X) \to I^1(X) \to \cdots)$} \\
+    %    &\phantom{\vcentcolon}= \operatorname{ker}(I^n(X) \to I^{n+1}(X))\mathop{/}\operatorname{im}(I^{n-1}(X) \to I^n(X))
+    %  \end{align*}
+    %  is well-defined up to isomorphism.
 
-    \only<3>{\textbf{Proof.} \bad{Assume not.} Then there is~a \bad{prime
-    ideal} $\ppp$. The matrix is surjective over the
-    field~$\mathrm{Quot}(A/\ppp)$. This is a contradiction to basic linear
-    algebra.}
+      \begin{itemize}
+        \item The modules~$H^n(X; E)$ are important invariants.
+        \item Let~$A$ be a ring. Then $H^n(X; \underline{A}) = H_{\text{sing}}^n(X;
+        A)$, if~$X$ is locally semiconnected [XXX].
+      \end{itemize}
     \end{column}
 
-  \begin{column}[t]{0.5\textwidth}
-    \centering
-
-    \scalebox{0.8}{$\begin{pmatrix}
-      \cdot & \cdot & \cdot & \cdot \\
-      \cdot & \cdot & \cdot & \cdot \\
-      \cdot & \cdot & \cdot & \cdot
-    \end{pmatrix}$}
-    \vspace*{-1em}
-
-    \begin{varblock}{\textwidth}{A child application}
-      \justifying
-      Let~$M$ be an injective matrix over~$A$ with more columns than rows.
-      Then~$1 = 0$ in~$A$.
-    \end{varblock}
+    \begin{column}{0.6\textwidth}
+      \centering{\usebeamercolor[fg]{item}{\textbf{\Large Constructive mathematics}}}\bigskip
 
       \justifying
-    \visible<2-3>{\textbf{Proof.} \bad{Assume not.} Then there is a \bad{minimal
-    prime ideal} $\ppp$. The matrix is injective over the field~$A_\ppp = A[(A
-    \setminus \ppp)^{-1}]$. This is a contradiction to basic linear algebra.}
-    \end{column}
-  \end{columns}}
-
-  \pause
-  \pause
-  \pause
-
-  \centering
-
-  \includegraphics[height=3em]{generic-freeness}
-  \vspace*{-1em}
-
-  \begin{varblock}{0.9\textwidth}{Generic freeness$\phantom{p}$}
-    \justifying
-    Let~$B$ be an~$A$-algebra of finite type ($\cong A[X_1,\ldots,X_n]/\aaa$). \\
-    Let~$M$ be a finitely generated~$B$-module ($\cong B^m/U$).
-
-    If~$f = 0$ is the only element of~$A$ such that
-    \begin{enumerate}
-      \item $B[f^{-1}]$ and $M[f^{-1}]$ are free modules over $A[f^{-1}]$,
-      \item $A[f^{-1}] \to B[f^{-1}]$ is of finite presentation and
-      \item $M[f^{-1}]$ is finitely presented as a module over~$B[f^{-1}]$,
-    \end{enumerate}
-    then $1 = 0$ in~$A$.
-  \end{varblock}
-  \vspace*{-0.5em}
-
-  \pause\center
-  \begin{minipage}{0.9\textwidth}
-    \textbf{Proof.} See [Stacks Project, Tag 051Q].
-  \end{minipage}
-\end{frame}
-
-\section{Internal language}
-
-\begin{frame}{The internal language of a topos}
-  For any topos~$\E$ and any formula~$\varphi$, we define the meaning of
-  \vspace*{-0.5em}
-  \[
-    \text{``$\E \models \varphi$''} \quad
-    \text{(``$\varphi$ holds in the internal universe of~$\E$'')}
-  \]
-
-  \vspace*{-1.0em}
-  using (Shulman's extension of) the \hil{Kripke--Joyal semantics}.
-
-  \jnote{1}{The internal language of a topos allows to construct objects and
-  morphisms of the topos, formulate statements about them and prove such
-  statements \emph{in a naive element-based language}. From the internal point
-  of view, objects look like sets [more precisely, types]; morphisms look like
-  maps; monomorphisms look like injections; epimorphisms
-  look like surjections; group objects look like groups; and so on.
-
-  To determine whether a statement~$\varphi$ holds in the internal universe of a
-  given topos, one can use the Kripke--Joyal semantics to translate it into an
-  ordinary external statement and then check the validity of the external
-  translation.
-
-  For instance, in the ef{}fective topos the curious statement~``any function
-  $\NN \to \NN$ is computable'' holds, for its external meaning is the
-  triviality ``there is a Turing machine which given a Turing machine computing
-  some function~$f : \NN \to \NN$ outputs a Turing machine computing~$f$''. In
-  contrast, the statement~``any function~$\NN \to \NN$ is either the zero
-  function or not'' does not hold in the ef{}fective topos, since its external
-  meaning is ``there exists a Turing machine which given a Turing machine
-  computing some function~$f : \NN \to \NN$ decides whether~$f$ is the zero
-  function or not''.}
-
-  \jnote{2}{Any theorem which has an intuitionistic proof holds in
-  the internal universe of any topos. The restriction to intuitionistic logic
-  is not due to philosophical concerns; it is a fact of life that only very few
-  toposes validate the law of excluded middle (for instance, sheaf toposes over
-  discrete topological spaces do if the law of excluded middle is available in
-  the metatheory). Luckily, vast amounts of mathematics can be developed in a
-  purely intuitionistic setting.
-
-  The internal language machinery itself can be developed in an intuitionistic
-  setting.
-
-  The standard internal language of toposes in not enough for our purposes, as
-  it misses unbounded quantification (``for all groups'', ``for all rings'')
-  and dependent types. Shulman's \fixedhref{https://arxiv.org/abs/1004.3802}{stack
-  semantics} offers what we need. No knowledge of stacks is necessary to enjoy
-  his paper. Prior work includes
-  \fixedhref{https://www.cl.cam.ac.uk/~amp12/papers/polist/polist.pdf}{Polymorphism
-  is Set Theoretic, Constructively} by Pitts and
-  \fixedhref{http://www.phil.cmu.edu/projects/ast/Papers/Awodey-Butz-Simpson-Streicher-APAL-2013.pdf}{Relating
-  first-order set theories, toposes and categories of classes} by Awodey, Butz,
-  Simpson and Streicher (obtained independently and published after Shulman).}
-
-  \vspace*{-1em}
-  \begin{columns}
-    \def\insertblocktitle{}
-    \begin{column}{0.25\textwidth}\usebeamertemplate{block begin}
-      \centering
-      $\Set \models \varphi$ \\
-      ``$\varphi$ holds in the \\ usual sense.''
-      \usebeamertemplate{block end}\end{column}
-
-    \begin{column}{0.25\textwidth}\usebeamertemplate{block begin}
-      \centering
-      $\Sh(X) \models \varphi$ \\
-      ``$\varphi$ holds continuously.''
-      \usebeamertemplate{block end}\end{column}
-
-    \begin{column}{0.25\textwidth}\usebeamertemplate{block begin}
-      \centering
-      $\Eff \models \varphi$ \\
-      ``$\varphi$ holds \\ computably.''
-      \usebeamertemplate{block end}\end{column}
-  \end{columns}
-  \medskip
-
-  Any topos supports \hil{mathematical reasoning}:
-
-  \vspace*{-1.5em}
-  \begin{hilblock}
-    If~$\E \models \varphi$ and if~$\varphi$ entails~$\psi$
-    \pointthis{<2>}{intuitionistically}{%
-      no $\varphi \vee \neg\varphi$,\ \
-      no $\neg\neg\varphi \Rightarrow \varphi$,\ \
-      no axiom of choice},
-    then~$\E \models \psi$.
-  \end{hilblock}
-\end{frame}
-
-\begin{frame}{The Kripke--Joyal semantics of~$\boldsymbol{\Sh(X)}$}
-  \small\vspace*{-0.5em}
-  Let~$X$ be a topological space. We recursively define
-  \[ U \models \varphi \quad \text{(``$\varphi$ holds on~$U$'')} \]
-  \mbox{for open subsets~$U \subseteq X$ and formulas~$\varphi$.
-  Write~``$\Sh(X) \models \varphi$'' to mean~$X \models \varphi$.}
-  \footnotesize
-  \[ \renewcommand{\arraystretch}{1.08}\begin{array}{@{}l@{\ \ }c@{\ \ }l@{}}
-  U \models \top &\Ll& \text{true} \\
-  U \models \bot &\Ll& \hcancel{\text{false}}{0pt}{3pt}{0pt}{-2pt}\ U = \emptyset \\
-  U \models s = t \? F &\Ll& s|_U = t|_U \in F(U) \\
-  U \models \varphi \wedge \psi &\Ll&
-  \text{$U \models \varphi$ and $U \models \psi$} \\
-  U \models \varphi \vee \psi &\Ll&
-  \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $U = \bigcup_i U_i$} \\
-  && \quad\quad\text{such that for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
-  U \models \varphi \Rightarrow \psi &\Ll&
-  \text{for all open~$V \subseteq U$: }
-  \text{$V \models \varphi$ implies $V \models \psi$} \\
-  U \models \forall s \? F\_ \varphi(s) &\Ll&
-  \text{for all open $V \subseteq U$ and sections~$s_0 \in F(V)$: $V \models \varphi(s_0)$} \\
-  U \models \forall F\_ \varphi(F) &\Ll&
-  \text{for all open $V \subseteq U$ and sheaves~$F_0$ over~$V$: $V \models \varphi(F_0)$} \\
-  U \models \exists s \? F\_ \varphi(s) &\Ll&
-  \hcancel{\text{there exists $s_0 \in F(U)$ such that $U \models \varphi(s_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
-  &&
-  \text{there exists a covering $U = \bigcup_i U_i$ such that for all~$i$:} \\
-  && \quad\quad \text{there exists~$s_0 \in F(U_i)$ such that $U_i \models \varphi(s_0)$} \\
-  U \models \exists F\_ \varphi(F) &\Ll&
-  \hcancel{\text{there exists a sheaf $F_0$ on $U$ such that $U \models
-  \varphi(F_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
-  &&
-  \text{there exists a covering $U = \bigcup_i U_i$ such that for all~$i$:} \\
-  && \quad\quad \text{there exists a sheaf~$F_0$ on $U_i$ such that $U_i \models \varphi(F_0)$}
-  \end{array} \]
-
-  \jnote{1}{Many interesting sheaves have few global sections, which is why a
-  definition such as ``$U \models \forall s \? F\_ \varphi(s)$ iff $U \models
-  \varphi(s_0)$ for all~$s_0 \in F(U)$'' would miss the point.
-
-  Here is an explicit example of the translation procedure. Let~$\alpha : F
-  \to G$ be a morphism of sheaves on~$X$. Then (the corner quotes ``$\speak{\ldots}$''
-  indicate that translation into formal language is left to the reader):
-  \footnotesize
-  \begin{align*}
-    & X \models \speak{$\alpha$ is injective} \\[0.2em]
-    \Longleftrightarrow\
-    & X \models \forall s\?F\_ \forall t\?F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.2em]
-    \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$, sections $s_0 \in F(U)$:} \\
-    &\qquad\qquad
-        U \models \forall t\?F\_ \alpha(s_0) = \alpha(t) \Rightarrow s_0 = t \\[0.2em]
-    \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$, sections $s_0 \in F(U)$:} \\
-    &\qquad\qquad
-        \text{for all open~$V \subseteq U$, sections $t_0 \in F(V)$:} \\
-    &\qquad\qquad\qquad\qquad
-          V \models \alpha(s_0) = \alpha(t_0) \Rightarrow s_0 = t_0 \\[0.2em]
-    \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$, sections $s_0 \in F(U)$:} \\
-    &\qquad\qquad
-        \text{for all open~$V \subseteq U$, sections $t_0 \in F(V)$:} \\
-    &\qquad\qquad\qquad\qquad
-          \text{for all open $W \subseteq V$: $\alpha_V(s_0|_W) = \alpha_V(t_0|_W)$ implies $s_0|_W = t_0|_W$} \\[0.2em]
-    \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$, sections $s, t \in F(U)$:
-        $\alpha_U(s) = \alpha_U(t)$ implies $s = t$} \\[0.2em]
-    \Longleftrightarrow\ &
-      \text{$\alpha$ is a monomorphism of sheaves}
-  \end{align*}}
-\end{frame}
-
-\begin{frame}{Internalizing parameter-dependence}
-  \justifying
-  Let~$X$ be a space. A continuous family~$(f_x)_{x \in X}$ of continuous
-  functions (that is, a continuous function~$f : X \times \RR \to \RR$;
-  $f_x(a) = f(x,a)$) induces an endomorphism of the sheaf~$\C$ of continuous
-  functions:
-  \[
-    \bar f : \C \longrightarrow \C,\
-    \text{on $U$:}\ s \longmapsto (x \mapsto f_x(s(x))).
-  \]
-
-  \vspace*{-1em}
-
-  \begin{itemize}
-    \item $\Sh(X) \models \speak{The set $\C$ is the set of (Dedekind) reals}$.
-    \item $\Sh(X) \models \speak{The function $\bar f : \RR \to \RR$ is continuous}$. \bigskip
-    \item Iff $f_x(-1) < 0$ for all $x$, then $\Sh(X) \models \bar f(-1) < 0$.
-    \item Iff $f_x(+1) > 0$ for all $x$, then $\Sh(X) \models \bar f(+1) > 0$.
-    \item Iff all $f_x$ are increasing, then $\Sh(X) \models \speak{$\bar f$ is
-    increasing}$. \bigskip
-    \item \ \\[-1.2em]\mbox{Iff there is an open cover~$X = \bigcup_i U_i$ such
-    that for each~$i$ there is a} \mbox{continuous function~$s : U_i \to \RR$
-    with~$f_x(s(x)) = 0$ for all~$x \in U_i$,} then $\Sh(X) \models \exists s
-    \? \RR\_ \bar f(s) = 0$.
-  \end{itemize}
-
-  \jnote{1}{This slide, unrelated to commutative algebra or algebraic geometry,
-  aims to illustrate one of the basic uses of the internal language of toposes:
-  Upgrading any theorem admitting an intuitionistic proof to a
-  parameter-dependent version.
-
-  Constructively, there are several non-equivalent forms of the intermediate
-  value theorem. The following version doesn't admit an intuitionistic proof:
-  \begin{indentblock}Let~$g : \RR \to \RR$ be a function between the (Dedekind)
-  reals which is continuous in the usual epsilon/delta sense. Assume~$g(-1) < 0
-  < g(1)$. Then there exists a number~$x \in \RR$ such that~$g(x) = 0$.
-  \end{indentblock}
-  If there was an intuitionistic proof, the statement would hold in any topos,
-  so in particular in sheaf toposes over topological spaces. By the
-  translations shown on the slide, this
-  would amount to the following strengthening of the intermediate value
-  theorem: In continuous families of continuous functions, zeros can locally be
-  picked continuously. However, this strengthening is invalid, as
-  \fixedhref{https://raw.githubusercontent.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{this
-  video shows}.
-
-  \centering\includegraphics[width=0.1\textwidth]{zeros-in-families-still}}
-
-  \jnote{2}{In contrast, the following version does admit an intuitionistic
-  proof. We can therefore interpret it in sheaf toposes over topological
-  spaces and thereby obtain the strengthening that in continuous families of
-  strictly increasing continuous functions, zeros can locally be picked
-  continuously. You are invited to prove this strengthening directly, without
-  reference to the internal language.
-  \begin{indentblock}Let~$g : \RR \to \RR$ be a function between the (Dedekind)
-  reals which is continuous in the usual epsilon/delta sense and which is
-  strictly increasing ($a < b$ implies~$g(a) < g(b)$). Assume~$g(-1) < 0 <
-  g(1)$. Then there exists a number~$x \in \RR$ such that~$g(x) =
-  0$.\end{indentblock}}
-\end{frame}
-
-\section{The little Zariski topos}
-
-\begin{frame}{The little Zariski topos}
-  Let~$A$ be a ring. Its \hil{little Zariski topos} is equivalently
-  \begin{enumerate}
-    \item the classifying topos of \hil{local localizations} of $A$,
-    \item the classifying locale of \hil{prime filters} of $A$,
-    \item the locale given by the frame of \hil{radical ideals} of $A$,
-    \item the topos of sheaves over the poset $A$ with $f \preceq g$ iff $f \in \sqrt{(g)}$ and with $(f_i \to f)_i$ deemed covering iff $f \in \sqrt{(f_i)_i}$ or
-    \item the topos of sheaves over $\Spec(A)$.
-  \end{enumerate}
-  Its associated topological space of points is the \hil{classical spectrum}
-  \[ \{ \fff \subseteq A \,|\, \text{$\fff$ prime filter} \} + \text{Zariski topology}. \]
-  It has \hil{enough points} if the Boolean Prime Ideal Theorem holds.
-
-  \footnotesize
-  Prime ideal:\, $0 \in \ppp$;\, $x \in \ppp \wedge y \in \ppp \Rightarrow x+y \in \ppp$;\, $1 \not\in \ppp$;\, $xy \in \ppp \Leftrightarrow x \in \ppp \vee y \in \ppp$
-
-  Prime filter:\, $0 \not\in \fff$;\,\,
-  $x+y \in \fff \Rightarrow x \in \fff \vee y \in \fff$;
-  \hspace*{0.5pt}\,\,
-  $1 \in \fff$;\,
-  $xy \in \fff \Leftrightarrow x \in \fff \wedge y \in \fff$
-
-  \jnote{1}{Any geometric theory has a classifying topos; if the theory under
-  consideration is propositional (doesn't have any sorts), then its classifying
-  topos can be chosen to be the topos of sheaves over a locale. One can also
-  give a direct account of classifying locales, as a pedagogical stepping stone
-  to the full theory of classifying toposes.
-
-  The slide contains a small lie: The classical definition of the spectrum of a
-  ring is via the set of prime ideals of~$A$, not prime filters. If the law of
-  excluded middle is available, there is no difference between these
-  definitions since the complement of a prime ideal is a prime filter and vice
-  versa.
-
-  One can also consider the classifying locale of prime \emph{ideals} of~$A$.
-  Its associated topological space of points is the set of prime ideals
-  of~$A$ equipped with the \emph{constructible topology}.
-
-  In an intuitionistic context, any of the (generalized) spaces of items~1--4
-  can be adopted as sensible definitions of the spectrum of~$A$. Item~5 is then
-  a tautology. The classical definition of the spectrum as a topological space
-  doesn't work very well, because verifying the universal property one expects
-  of it requires the Boolean Prime Ideal Theorem. Most dramatically, there are
-  rings which are not trivial yet have neither prime ideals nor prime filters.
-  The classical definition yields in this case the empty space.}
-\end{frame}
-
-\begin{frame}{First steps in the little Zariski topos}
-  Let~$A$ be a ring. Let~$\fff_0$ be the \hil{generic prime filter} of~$A$; it
-  is a subobject of the constant sheaf~$\ull{A}$ of the little Zariski topos.
-
-  \begin{itemize}
-    \item The ring $A^\sim \defeq \ull{A}[\fff_0^{-1}]$ is the generic local
-    localization of $A$.
-    \item Given an~$A$-module~$M$, we have the~$A^\sim$-module $M^\sim \defeq
-    \ull{M}[\fff_0^{-1}]$.
-  \end{itemize}
-
-  \jnote{1}{The Kripke--Joyal semantics for the little Zariski topos amounts to
-  the following: $\Spec(A) \models \varphi$ iff~$D(1) \models \varphi$, and the
-  clauses for~$D(f) \models \varphi$, where~$f$ ranges over the elements
-  of~$A$, are given by the following table.
-  \[\footnotesize\renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\quad}c@{\quad}l@{}}
-    D(f) \models \top &\text{iff}& \text{true} \\
-    D(f) \models \bot &\text{iff}& \text{$f$ is nilpotent} \\
-    D(f) \models x = y &\text{iff}& x = y \in M[f^{-1}] \\
-    D(f) \models \varphi \wedge \psi &\text{iff}&
-      \text{$D(f) \models \varphi$ and $D(f) \models \psi$} \\
-    D(f) \models \varphi \vee \psi &\text{iff}&
-      \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
-    &&\quad\text{for each~$i$, $D(fg_i) \models \varphi$ or $D(fg_i) \models \psi$} \\
-    D(f) \models \varphi \Rightarrow \psi &\text{iff}&
-      \text{for all~$g \in A$, $D(fg) \models \varphi$ implies $D(fg) \models \psi$} \\
-    D(f) \models \forall x\?M^\sim\_ \varphi(x) &\text{iff}&
-      \text{for all~$g \in A$ and $x_0 \in M[(fg)^{-1}]$, $D(fg) \models \varphi(x_0)$} \\
-    D(f) \models \exists x\?M^\sim\_ \varphi(x) &\text{iff}&
-      \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
-    &&\quad\text{for each~$i$, $D(fg_i) \models \varphi(x_0)$ for some~$x_0 \in M[(fg_i)^{-1}]$}
-  \end{array} \]
-  The generic prime filter~$\fff_0$ can also be described in explicit
-  terms. For ring elements~$f$ and~$s$, $D(f) \models (s \in \fff_0)$ iff~$f
-  \in \sqrt{(s)}$.}
-
-  \jnote{2}{Our description of~$M^\sim$ reveals a precise sense in
-  which~$M^\sim$ and~$M$ are related: $M^\sim$ is simply a localization of~$M$
-  (first lifted to another universe by the constant sheaf construction). The
-  classical descriptions don't make the relation evident.
-
-  As a first approximation, the module~$M^\sim$ can be thought of as a
-  reification of all the stalks of~$M$ as a single object. The metatheorem
-  displayed at the top left on the next slides makes this precise and also
-  exposes the limits of this view: It is only correct for geometric sequents.
-  When considering nongeometric sequents, phenomena appear which are unique
-  to~$M^\sim$ in the sense that they are in general not shared by~$M$, its
-  stalks or its quotients.}
-
-  \jnote{3}{One can show, assuming that the little Zariski topos is
-  \emph{overt}, that the module~$M$ in~$\Set$ and the module~$\ull{M}$ of the
-  little Zariski topos share all first-order properties. This observation
-  explains the metatheorem displayed at the bottom left. The assumption is
-  satisfied if any element of~$A$ is nilpotent or not nilpotent, so it's always
-  satisfied if the law of excluded middle is available in the metatheory. In an
-  intuitionistic context, it's still ``morally satisfied''. Details are in
-  Section~12.9 of
-  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
-  notes}.
-
-  The metatheorem soups up a number of lemmas of algebraic geometry, there
-  stated in geometric language. For instance, if~$M$ is finitely generated,
-  then~$M^\sim$ is of finite type. If~$M$ is finitely presented, then~$M^\sim$
-  is of finite presentation. If~$M$ is coherent, then~$M^\sim$ is coherent.
-
-  As an aside, the little Zariski topos is rarely Boolean (validates the law of
-  excluded middle). A necessary condition is that~$A$ is of dimension~$\leq 0$.}
-
-  \jnote{4}{Assuming that~$A$ is reduced, the following nongeometric sequents
-  hold in the little Zariski topos (among others):
-
-  $A^\sim$ is a field in the sense that zero is the only noninvertible
-  element. This field property was already observed in the 1970s by Mulvey, who
-  didn't know a deeper reason for this property. We now know that it's a
-  shadow of an internal property whose external translation expresses
-  that~$A^\sim$ is quasicoherent.
-
-  $A^\sim$ has~$\neg\neg$-stable equality in the sense that
-  \[ \Spec(A) \models \forall s\?A^\sim\_ \neg\neg(s = 0) \Rightarrow s = 0. \]
-  Classically, every set has~$\neg\neg$-stable equality; intuitionistically,
-  this is a special property of some sets. It's quite useful, as some theorems
-  of classical commutative algebra can only be proven intuitionistically when
-  weakened by double negation. The stability then allows, in some cases, to
-  obtain the original conclusion.
-
-  $A^\sim$ is anonymously Noetherian in the sense that any of its ideals
-  is \mbox{\emph{not not}} finitely generated. A philosophically-motivated
-  constructivist might be offended by this notion, since it runs counter to the
-  maxim that constructive mathematics should be informative. However, in the
-  internal context it is a useful notion: Hilbert's basis theorem holds for it,
-  and we'll put it to good use in our proof of Grothendieck's generic freeness
-  lemma.}
-
-  \jnote{5}{Are there theorems which can only be proven using the internal
-  language and not be proven without?
-
-  No. Just as the translation from internal statements to external statements
-  is entirely mechanical, so is the translation from internal proofs to
-  external proofs. Any proof employing the internal language can be unwound to
-  yield an external proof not referencing the internal language.
-
-  However, depending on the logical complexity of the statements occurring in a
-  given proof, the resulting external proof might be (much) more complex than
-  the internal proof. This is particularly the case if the proof involves
-  double negation, for much the same reason as that in computer science,
-  continuations can twist the control flow in nontrivial ways which are
-  sometimes hard to understand. It is in these cases where we can extract the
-  most value of the internal language, unlocking notions and proofs which might
-  otherwise be hard to obtain.
-
-  The slide shows a specific example. The internal statement
-  that~$A^\sim[X_1,\ldots,X_n]$ is anonymously Noetherian is quite simple; its
-  external translation is quite convoluted.}
-
-  \pause
-  \only<2>{\begin{center}
-    \includegraphics[width=0.9\textwidth]{hartshorne-tilde-construction} \\
-    \footnotesize
-    Robin Hartshorne. Algebraic Geometry. 1977.
-  \end{center}}
-  \pause
-
-  \vspace*{-1.5em}
-
-  \begin{columns}
-    \begin{column}{0.48\textwidth}
-      \begin{varblock}{\textwidth}{}
-        \justifying
-        Assuming the Boolean prime ideal theorem, a geometric sequent
-        ``$\forall \ldots \forall\_ (\cdots \Longrightarrow \cdots\!\,)$'',
-        where the two subformulas may not contain~``$\Rightarrow$'' and~``$\forall$'',
-        holds for~$M^\sim$ iff it holds for all stalks~$M_\ppp$.
-      \end{varblock}
-
-      \vspace*{-1.5em}
-
-      \begin{varblock}{\textwidth}{}
-        $M^\sim$ inherits any property of~$M$ which is \hil{localization-stable}.
-      \end{varblock}
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
-      \vspace*{1.5em}
-
-      If $A$ is reduced ($x^n = 0 \Rightarrow x = 0$):
-
-      \vspace*{-1.0em}
-
-      \setbeamercolor{block body}{bg=red!30}
-      \setbeamercolor{structure}{fg=purple}
-      \begin{varblock}{\textwidth}{}
-        $A^\sim$ is a \hil{field}
-        (nonunits are zero).
-
-        $A^\sim$ has \hil{$\boldsymbol{\neg\neg}$-stable equality}.
-
-        \mbox{$A^\sim$ is \hil{anonymously Noetherian}.}\\[-1.2em]
-      \end{varblock}
-    \end{column}
-  \end{columns}
-
-  \visible<4-5>{\begin{tikzpicture}[overlay]
-    \draw[fill=white, draw=white, opacity=0.95] (-1,0) rectangle (\paperwidth,7.4);
-    \node[anchor=south west,inner sep=0] (image) at (0,0.8) {\vbox{
-      \only<4>{
-	\centering
-	\includegraphics[width=0.9\textwidth]{tierney-on-the-spectrum-of-a-ringed-topos} \\
-	\footnotesize
-	Miles Tierney. On the spectrum of a ringed topos. 1976.
-      }
-
-      \only<5>{
-	\begin{varblock}{\textwidth}{Complexity reduction}
-	  The external meaning of
-	  \[
-	    \Spec(A) \models
-	      \speak{$A^\sim[X_1,\ldots,X_n]$ is anonymously Noetherian}
-	  \]
-	  is:
-	  \medskip
-
-	  \begin{indentblock}
-	  For any element~$f \in A$ and any (not necessarily quasicoherent) sheaf of
-	  ideals~$\J \hookrightarrow A^\sim[X_1,\ldots,X_n]|_{D(f)}$: If
-	  \begin{indentblock}
-	  for any element~$g \in A$ the condition that
-	  \begin{indentblock}
-	  the sheaf~$\J$ is of finite type on~$D(g)$
-	  \end{indentblock}
-	  implies that~$g = 0$,
-	  \end{indentblock}
-	  then~$f = 0$.
-	  \end{indentblock}
-	\end{varblock}
-
-	\vspace*{-2em}
-      }
-    }};
-  \end{tikzpicture}}
-\end{frame}
-
-\begin{frame}{Revisiting the test cases}
-  Let $A$ be a reduced ring.
-
-  \jnote{1}{This slide delivers on the promise made earlier: Using the internal
-  language of the little Zariski topos, we can reduce to the case of fields
-  without having to employ maximal ideals, prime ideals or minimal prime ideals.
-  Since the internal language machinery is itself constructive, the displayed
-  proofs can be unwound to yield external constructive proofs which don't
-  reference topos theory. Details on these external proofs can soon be found in
-  a forthcoming paper titled
-  \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-wlog.pdf}{Without
-  loss of generality, any reduced ring is a field}.}
-
-  \jnote{2}{The test case of Grothendieck's generic freeness lemma illustrates
-  that the internal language of toposes can help commutative algebra even if
-  one is not interested in constructivity issues. The previously known proofs
-  are somewhat long and somewhat convoluted; the new proof is arguably short
-  and simple.
-
-  Details on the internal proof are in Section~11.5 of
-  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
-  notes}. The
-  \fixedhref{https://arxiv.org/abs/1807.01231}{external
-  proof} obtained by unwinding the internal one is still quite direct, compared
-  to the previously published proofs, and interestingly follows a curious
-  course: It starts with verifying, in an inductive manner, that~$B$ and~$M$
-  are free; that~$A \to B$ is of finite presentation; and that~$M$ is finitely
-  presented as a~$B$-module. Then the assumption for~$f = 1$ is used, rendering
-  the prior steps moot, since over the zero ring any module is free and any
-  algebra is of finite presentation. External translations of internal proofs
-  which use double negation will always take such a course.}
-
-  \jnote{3}{Here's a rough sketch why the double negations occur. In
-  undergraduate linear algebra, we learn that any finitely generated vector
-  space is free (has a basis), by the following argument:
-  Let~$(v_1,\ldots,v_n)$ be a generating family. Either one of the generators can
-  be expressed as a linear combination of the others or not. In the latter
-  case, the family is linearly independent and we're done. In the former, we
-  remove the redundant generator and continue by induction.
-
-  This argument relies on the law of excluded middle and can therefore not put
-  to work in the internal universe as is. However, intuitionistically we do
-  have the weaker statement~$\neg\neg(\varphi \vee \neg\varphi)$, and we can
-  use this law in place of the law of excluded middle to intuitionistically
-  deduce that any finitely generated vector space does \emph{not not} have a
-  basis.}
-
-  \only<1>{\begin{columns}[t]
-    \begin{column}[t]{0.47\textwidth}
-      \centering
-
-      \scalebox{0.8}{$\begin{pmatrix}
-        \cdot & \cdot \\
-        \cdot & \cdot \\
-        \cdot & \cdot
-        \end{pmatrix}$}
-      \vspace*{-1em}
-
-      \begin{varblock}{\textwidth}{A baby application}
-        \justifying
-        Let~$M$ be a surjective matrix over~$A$ with more rows than columns.
-        Then~$1 = 0$ in~$A$.
-      \end{varblock}
-
-      \justifying
-      \textbf{Proof.} The matrix is surjective over the field~$A^\sim$.
-      This is a contradiction to basic linear algebra. Hence $\Spec(A) \models
-      \bot$, thus $1 = 0$ in $A$.
-    \end{column}
-    \begin{column}[t]{0.47\textwidth}
-      \centering
-
-      \scalebox{0.8}{$\begin{pmatrix}
-        \cdot & \cdot & \cdot & \cdot \\
-        \cdot & \cdot & \cdot & \cdot \\
-        \cdot & \cdot & \cdot & \cdot
-        \end{pmatrix}$}
-      \vspace*{-1em}
-
-      \begin{varblock}{\textwidth}{A child application}
-        \justifying
-        Let~$M$ be an injective matrix over~$A$ with more columns than rows.
-        Then~$1 = 0$ in~$A$.
-      \end{varblock}
-
-      \justifying
-      \textbf{Proof.} The matrix is injective over the field~$A^\sim$. This is
-      a contradiction to basic linear algebra. Hence $\Spec(A) \models \bot$,
-      thus $1 = 0$ in $A$.
-    \end{column}
-  \end{columns}}
-
-  \pause
-
-  \vspace*{-5.3em}\hspace*{21em}\begin{minipage}{3cm}
-    $\xymatrixcolsep{3.5pc}\xymatrixrowsep{1.2pc}\xymatrix{
-      & \ar@{-}[d]^{\substack{\text{finitely}\\\text{generated}}} M \\
-      A \ar[r]_{\text{of finite type}} & B
-    }$
-  \end{minipage}
-  \vspace*{-1em}
-
-  \centering
-
-  \begin{varblock}{0.9\textwidth}{Generic freeness$\phantom{p}$}
-    \justifying
-    Let~$B$ be an~$A$-algebra of finite type ($\cong A[X_1,\ldots,X_n]/\aaa$). \\
-    Let~$M$ be a finitely generated~$B$-module ($\cong B^m/U$).
-
-    If~$f = 0$ is the only element of~$A$ such that
-    \begin{enumerate}
-      \item $B[f^{-1}]$ and $M[f^{-1}]$ are free modules over $A[f^{-1}]$,
-      \item $A[f^{-1}] \to B[f^{-1}]$ is of finite presentation and
-      \item $M[f^{-1}]$ is finitely presented as a module over~$B[f^{-1}]$,
-    \end{enumerate}
-    then $1 = 0$ in~$A$.
-  \end{varblock}
-  \vspace*{-0.5em}
-
-  \pause
-
-  \scalebox{0.9}{\begin{minipage}{1\textwidth}
-    \textbf{Proof.} In the little Zariski topos it's \hil{not not} the case that % \\[-1.4em]
-    \begin{enumerate}
-      \item $B^\sim$ and $M^\sim$ are free modules over $A^\sim$,
-      \item \vspace*{-0.3em}$A^\sim \to B^\sim$ is of finite presentation and
-      \item \vspace*{-0.3em}$M^\sim$ is finitely presented as a module over $B^\sim$,
-    \end{enumerate}
-    by basic linear algebra over the field $A^\sim$. The claim is precisely the
-    external translation of this fact.
-  \end{minipage}}
-\end{frame}
-
-\section{Algebraic geometry}
-
-\begin{frame}{Understanding algebraic geometry}
-  \vspace*{-1.5em}
-  \begin{varblock}{0.9\textwidth}{}
-    \justifying
-    Understand \hil{notions of algebraic geometry} over a scheme~$X$ as
-    \hil{notions of algebra} internal to~$\Sh(X)$.
-  \end{varblock}
-
-  \jnote{1}{One doesn't need to be an expert in topos theory in order to know
-  that many notions in algebraic geometry are inspired by notions in algebra
-  and that proofs in algebraic geometry often proceed by reducing to algebra.
-  The internal language is a way of making this connection precise: In many
-  cases, the former are simply interpretations of the latter internal
-  to~$\Sh(X)$. Because this connection is precise instead of informal,
-  additional value is gained: We can skip many basic proofs in algebraic
-  geometry because they're just externalizations of proofs in algebra carried
-  out internally to~$\Sh(X)$.}
-
-  \jnote{2}{A basic example is as follows. A short exact sequence of sheaves of
-  modules looks like a short exact sequence of plain modules from the internal
-  point of view of~$\Sh(X)$. If the two outer sheaves are of finite type, then
-  from the internal point of view, the two outer modules will look like
-  finitely generated modules. Because the standard proof of the proposition
-  quoted on the lower right is intuitionistically valid, it follows that, from
-  the internal point of view, the middle module is too finitely generated.
-  Consulting the dictionary a second time, this amounts to saying that the
-  middle sheaf is of finite type.
-
-  More details on this research program can be found in
-  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
-  notes}, partly reported on at the
-  \fixedhref{https://www.youtube.com/watch?v=7S8--bIKaWQ}{2015 IHÉS
-  conference}. Even though many important dictionary entries
-  are still missing (for instance pertaining derived categories and
-  intersection theory), I believe that it is already at its current stage
-  useful to working algebraic geometers.
-
-  The next few slides illustrate an entirely different way of approaching
-  algebraic geometry using topos theory.}
-
-  \only<1>{
-    \centering
-    \rotatebox{90}{\tiny\scalebox{0.8}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
-    \includegraphics[width=0.7\textwidth]{external-internal-small}
-    \par
-  }
-
-  \pause
-
-  \small\centering
-  \scalebox{0.83}{\begin{tabular}{ll}
-    \toprule
-    externally & internally to $\Sh(X)$ \\
-    \midrule
-    sheaf of sets & set \\
-    % sheaf of rings & ring \\
-    sheaf of modules & module \\
-    sheaf of finite type & finitely generated module \\
-    % finite locally free sheaf & finite free module \\
-    % coherent sheaf & coherent module \\
-    tensor product of sheaves & tensor product of modules \\
-    % sheaf of Kähler differentials & module of Kähler differentials \\
-    sheaf of rational functions & total quotient ring of~$\O_X$ \\
-    dimension of $X$ & Krull dimension of~$\O_X$ \\
-    spectrum of a sheaf of~$\O_X$-algebras & ordinary spectrum [with a twist] \\
-    big Zariski topos of $X$ & big Zariski topos of the ring $\O_X$ [with a twist] \\
-    higher direct image & sheaf cohomology \\
-    \bottomrule
-  \end{tabular}}
-
-  \begin{columns}[c]
-    \begin{column}{0.47\textwidth}
-      \begin{exampleblock}{}
-        \justifying
-        Let $0 \to \F' \to \F \to \F'' \to 0$ be a short exact sequence
-        of sheaves of~$\O_X$-modules. If~$\F'$ and~$\F''$ are of finite type,
-        so is~$\F$.
-      \end{exampleblock}
-    \end{column}
-
-    \begin{column}{0.1\textwidth}
-      \vspace*{0.7em}
-      \scalebox{3}{$\Leftarrow$}
-    \end{column}
-
-    \begin{column}{0.44\textwidth}
-      \begin{exampleblock}{}
-        \justifying
-        Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence of
-        modules. If~$M'$ and~$M''$ are finitely generated, so is~$M$.
-      \end{exampleblock}
     \end{column}
@@
 
-\begin{frame}{Synthetic algebraic geometry}
-  Usual approach to algebraic geometry: \hil{layer schemes above ordinary set theory}
-  using either
-  \begin{itemize}
-    \item locally ringed spaces
-    \small
-    \begin{multline*}
-      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$} + {} \\
-      \text{Zariski topology} + \text{structure sheaf}
-    \end{multline*}
-    \normalsize
-    \item or Grothendieck's functor-of-points account, where a scheme is a
-    functor~$\Ring \to \Set$.
-    \small\[ A \longmapsto \{ (x,y,z) \in A^3 \,|\, x^n+y^n-z^n=0 \} \]
-  \end{itemize}
-  \bigskip
-
-  \jnote{1}{\footnotesize At the
-  \fixedhref{https://sbseminar.wordpress.com/2009/08/06/algebraic-geometry-without-prime-ideals/}{Secret
-  Blogging Seminar}, there was an insightful long-running discussion on the
-  merits of the two approaches. Two disadvantages of the approach using
-  locally ringed spaces is that the underlying topological spaces don't
-  actually parametrize ``honest'', ``geometric'' points, but the more complex
-  notion of irreducible closed subsets; and that they don't work well in a
-  constructive setting. (For this, they would have to be replaced by locally
-  ringed locales.)
-
-  The functorial approach is more economical, philosophically rewarding, and
-  works constructively. Given a functor~$F : \Ring \to \Set$, we imagine~$F(A)$
-  to be the set of~``$A$-valued points'' of the hypothetical scheme described
-  by~$F$, the set of ``points with coordinates in~$A$''. These sets have direct
-  geometric meaning. However, typically only field-valued points are easy
-  to describe. For instance, the functor representing projective~$n$-space is
-  given on fields by
-  \[ \begin{array}{r@{}c@{}l}
-  K &{}\longmapsto{}& \text{the set of lines through the origin in~$K^{n+1}$} \\
-  && \qquad \cong \{ [x_0 \hg \cdots \hg x_n] \,|\, \text{$x_i \neq 0$ for some~$i$} \},
-  \end{array} \]
-  whereas on general rings it is given by
-  \[ \begin{array}{r@{}c@{}l}
-    A &{}\longmapsto{}& \text{the set of quotients~$A^{n+1} \twoheadrightarrow P$,
-    where~$P$ is projective,} \\
-    && \qquad \text{modulo isomorphism}.
-  \end{array} \]
-  It is these more general kinds of points which impart a meaningful sense of
-  cohesion on the field-valued points, so they can't simply be dropped from
-  consideration.}
-
-  \jnote{2}{This tension is resolved by observing that the category of
-  functors~$\Ring \to \Set$ is a topos (the \emph{big Zariski topos} of~$\Spec(\ZZ)$)
-  and that we can therefore employ its internal language. This language takes
-  care of juggling stages behind the scenes. For instance, projective~$n$-space
-  can be described by the naive expression
-  \[ \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee \cdots \vee x_n
-  \neq 0 \}/(\affl)^\times. \]
-  This example illustrates our goal: to develop a synthetic account of
-  algebraic geometry, in which schemes are plain sets and morphisms between
-  schemes are maps between those sets. It turns out that there are many
-  similarities with the well-developed synthetic account of differential
-  geometry, but also important differences.}
-
-  \pause
-
-  \hil{Synthetic approach:} model schemes \hil{directly as sets} in
-  the internal universe of the \mbox{\hil{big Zariski topos}} of a base scheme.
-  \small
-  \[ \{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]
-\end{frame}
-
-\begin{frame}{The big Zariski topos}
-  Let~$S$ be a fixed base scheme.
-  \begin{varblock}{0.9\textwidth}{Definition}
-    The \hil{big Zariski topos} $\Zar(S)$ of a scheme~$S$ is equivalently
-    \begin{enumerate}
-      \item the topos of sheaves over~$(\Aff/S)_{\mathrm{lofp}}$,
-      \item the classifying topos of local rings over~$S$ or
-      \item the classifying~$\Sh(S)$-topos of local~$\O_S$-algebras which are
-      local over~$\O_S$.
-    \end{enumerate}
-  \end{varblock}
-
-  \begin{itemize}
-    \item For an~$S$-scheme~$X$, its functor of points $\ull{X} =
-    \Hom_S(\cdot,X)$ is an object of~$\Zar(S)$. It feels like \hil{the set of
-    points} of~$X$.
-    \item In particular, there is the ring object~$\affl$ with~$\affl(T) =
-    \O_T(T)$.
-    \item This ring object is a \hil{field}: nonzero implies
-    invertible. \\{} [Kock 1976]
-  \end{itemize}
-
-  \jnote{1}{The objects of the category~$(\Aff/S)_{\mathrm{lofp}}$ are
-  morphisms of the form~$\Spec(R) \to S$ which are locally of finite
-  presentation. (Other choices of resolving set-theoretical issues of size are
-  also possible.)
-
-  A functor~$F : (\Aff/S)_{\mathrm{lofp}}^\op \to \Set$ is a sheaf
-  for the Zariski topology if and only if the diagram
-  \[ F(T) \to
-    \prod_i F(U_i) \rightrightarrows
-    \prod_{j,k} F(U_j \cap U_k) \]
-  is a limit diagram for any open covering~$T = \bigcup_i U_i$ of any
-  scheme~$T \in (\Aff/S)_{\mathrm{lofp}}$.
-
-  In the case that~$S = \Spec(A)$ is affine, the big Zariski topos of~$S$ is also
-  simply called ``big Zariski topos of~$A$''. It is a subtopos of the topos of
-  functors $\mathrm{Alg}(A) \to \Set$ and classifies local~$A$-algebras.
-
-  From the internal point of view of~$\Sh(S)$, the sheaf~$\O_S$ of rings is
-  just an ordinary ring, and one can construct internally to~$\Sh(S)$ the big
-  Zariski topos of~$\O_S$. Externally, this construction will yield a certain
-  bounded topos over~$\Sh(S)$. However, as indicated on the slide, this topos
-  will \emph{not} coincide with the true big Zariski topos of~$S$. To
-  construct the true big Zariski topos, one has to build, internally
-  to~$\Sh(S)$, the classifying topos of local \emph{and local-over-$\O_S$}
-  $\O_S$-algebras.}
-\end{frame}
-
-\begin{frame}{Synthetic constructions}
-  \small
-  $\hil{$\boldsymbol{\mathbb{A}^n}$} = (\affl)^n = \affl \times \cdots \times \affl$
-
-  $\begin{array}{@{}r@{}c@{}l@{}}
-    \hil{$\boldsymbol{\mathbb{P}^n}$} &\phantom{}=\phantom{}& \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee
-    \cdots \vee x_n \neq 0 \}/(\affl)^\times \\
-    &\cong& \text{set of one-dimensional subspaces of~$(\affl)^{n+1}$} \\
-    &&\qquad \text{(with~$\O(-1) = (\ell)_{\ell \? \mathbb{P}^n}$, $\O(1) =
-    (\ell^\vee)_{\ell \? \mathbb{P}^n}$)}
-  \end{array}$
-
-  $\hil{$\boldsymbol{\Spec(R)}$} = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
-    \text{set of $\affl$-valued points of $R$}$
-
-  $\hil{$\boldsymbol{TX}$} = X^\Delta$, where $\Delta = \{ \varepsilon \? \affl \,|\, \varepsilon^2 = 0 \}$
-
-  A subset $U \subseteq X$ is \hil{qc-open} if and only if for any $x : X$
-  there exist $f_1,\ldots,f_n \? \affl$ such that $x \in U \Longleftrightarrow
-  \exists i\_ f_i \neq 0$.
-
-  A \hil{synthetic affine scheme} is a set which is in bijection
-  with~$\Spec(R)$ for some synthetically quasicoherent~$\affl$-algebra~$R$.
-
-  A \hil{finitely presented synthetic scheme} is a set which can be covered by
-  finitely many qc-open f.p. synthetic affine schemes~$U_i$ such that the
-  intersections~$U_i \cap U_j$ can be covered by finitely many qc-open f.p.
-  synthetic affine schemes.
-
-  \jnote{1}{In the internal universe of the big Zariski topos of a base
-  scheme~$S$, $S$-schemes can simply be modeled by sets (enjoying the special
-  property that, in a certain precise sense, they are locally affine). This
-  slide expresses some of the basic constructions of~$S$-schemes in that
-  language.
-
-  Particularly nice are the following items.
-  \begin{itemize}\justifying
-    \item Projective~$n$-space can be given by the any of the two quite naive
-    expressions displayed on the slide.
-    \item Let~$X$ be an~$S$-scheme. We often think about a sheaves
-    of~$\O_X$-modules over~$X$ by their fibers; but for a rigorous
-    treatment in the standard foundations, we have to take the full sheaf
-    structure into account; the fibers do not determine a sheaf
-    uniquely.
-
-    From the internal point of view of~$\Zar(S)$, a sheaf of~$\O_X$-modules is
-    indeed simply a family of~$\affl$-modules, one~$\affl$-module for each
-    element of~$\ull{X}$. The slide illustrates how we can define the Serre
-    twisting sheaves in this language.
-  \end{itemize}}
-
-  \jnote{2}{\begin{itemize}\justifying
-    \item The spectrum of an~$\affl$-algebra can be given by the naive
-    expression displayed on the slide. It looks like this expression can't be
-    right, ignoring any non-maximal ideals; however, it is.
-    \item The big Zariski topos of an~$S$-scheme~$X$ is, from the internal
-    point of view of~$\Zar(S)$, simply the slice topos~$\Set/\ull{X}$. Hence to
-    give an~$X$-scheme simply amounts to giving an~$\ull{X}$-indexed family of
-    sets.
-  \end{itemize}
-
-  Synthetic algebraic geometry has been developed up to the point of étale
-  geometric morphisms. Much remains to be done: For instance, as of yet there
-  is only an account of Čech methods for computing cohomology, there is not yet
-  a synthetic treatment of true cohomology. Derived categories and intersection
-  theory are also missing.}
-\end{frame}
-
-\begin{frame}{Relations between the Zariski toposes}
-  The big Zariski topos is a topos over the small Zariski topos:
-  \[ \begin{array}{crcl}
-    \pi : & \Zar(A) &\longrightarrow& \Spec(A) \\
-    & \text{local~$A$-algebra~$(A \xrightarrow{\alpha} B)$} &\longmapsto& (A \to A[(\alpha^{-1}[B^\times])^{-1}])
-  \end{array} \]
-  This morphism is \hil{connected} ($\pi^{-1}$ is fully faithful) and
-  \hil{local}, so there is a preinverse
-  \[ \begin{array}{rcl}
-    \Spec(A) &\longrightarrow& \Zar(A) \\
-    \text{local localization~$(A \to B)$} &\longmapsto& (A \to B)
-  \end{array} \]
-  which is a subtopos inclusion inducing an idempotent monad~$\sharp$ and an
-  idempotent comonad~$\flat$ on~$\Zar(S)$.
-
-  \begin{itemize}
-  \justifying
-    \item Internally to~$\Zar(S)$, $\Spec(S)$ can be constructed as the
-    \hil{largest subtopos} where~$\flat \affl \to \affl$ is bijective.
-    \item Internally to~$\Spec(S)$, $\Zar(S)$ can be constructed as the
-    \hil{classifying topos} of local~$\O_S$-algebras which are local over~$\O_S$.
-    \item $\Zar(A)$ is the \hil{lax pullback}~$(\Set
-    \Rightarrow_{\Set[\Ring]} \Set[\LocRing])$.
-  \end{itemize}
-
-  \jnote{1}{Let~$A$ be a ring. By definition, we obtain a geometric
-  morphism~$\Set \to \Set[\Ring]$ into the classifying topos of rings.
-  There is also a geometric morphism~$\Set[\LocRing] \to \Set[\Ring]$, obtained
-  by realizing that any local ring is in particular a ring. These morphisms fit
-  together in a lax pullback square as follows:
-  \[ \xymatrix{
-    \Zar(A) \ar[r] \ar[d] & \Set[\LocRing] \ar[d] \\
-    \Set \ar[r] \ar@{}[ur]^(.3){}="a"^(.7){}="b" \ar@{=>} "a";"b" & \Set[\Ring]
-  } \]
-  This observation is joint with Peter Arndt and Matthias Hutzler.
-
-  Incidentally, the pseudo pullback of the morphism~$\Set[\LocRing] \to
-  \Set[\Ring]$ along~$\Set \to \Set[\Ring]$ is not very interesting: It's the
-  largest subtopos of~$\Set$ where~$A$ is a local ring. Assuming the law of
-  excluded middle, this subtopos is either the trivial topos (if~$A$ is not
-  local) or~$\Set$ (if~$A$ is local).
-
-  There is also a way of realizing the little Zariski topos of~$A$ as a pseudo
-  pullback, exploiting that the (localic) spectrum construction is geometric.
-  See Section~12.6 of
-  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
-  notes} for details.
-  }
-\end{frame}
-
-\section{The mystery of nongeometric sequents}
-
-\begin{frame}{Properties of the affine line}
-  \begin{itemize}
-%   \item $\affl$ is a local ring:
-%   \[ 1 \neq 0 \qquad\qquad \text{$x + y$ \inv} \Longrightarrow \text{$x$ \inv} \vee
-%   \text{$y$ \inv} \]
-    \item $\affl$ is a field:
-    \begin{align*}
-      \neg(x = 0) &\Longleftrightarrow \text{$x$ invertible \quad [Kock 1976]} \\
-      \neg(\text{$x$ invertible}) &\Longleftrightarrow \text{$x$ nilpotent}
-    \end{align*}
-    \item $\affl$ satisfies the axiom of microaffinity: Any map $f : \Delta \to
-    \affl$ is of the form~$f(\varepsilon) = a + b \varepsilon$ for unique
-    values~$a,b \? \affl$, where~$\Delta = \{ \varepsilon \? \affl \,|\,
-    \varepsilon^2 = 0 \}$.
-    \bigskip
-    \item Any map $\affl \to \affl$ is a polynomial function.
-    \bigskip
-    \item $\affl$ is anonymously algebraically closed: Any monic polynomial
-    does \emph{not not} have a zero.
-  \end{itemize}
-
-  \jnote{1}{The axiom of microaffinity is a special instance of the
-  \emph{Kock--Lawvere axiom} known from synthetic differential geometry. We'll
-  see on the next slide that~$\affl$ validates an unusually strong form of the
-  Kock--Lawvere axiom, not at all satisfied in the usual well-adapted models of
-  synthetic differential geometry.
-
-  The fact that, internally to~$\Zar(S)$, any map~$\affl \to \affl$ is a
-  polynomial can be seen as a formal version of the general motto that in
-  algebraic geometry, ``morphisms are polynomials''.}
-\end{frame}
-
-\begin{frame}{Synthetic quasicoherence}
-  Recall~$\Spec(R) = \Hom_{\mathrm{Alg}(\affl)}(R, \affl)$ and consider the statement
-  \[ \text{``the canonical map~$
-    \begin{array}[t]{rcl}
-      R &\longrightarrow& (\affl)^{\Spec(R)} \\
-      f &\longmapsto& (\alpha \mapsto \alpha(f))
-    \end{array}
-  $ is bijective''}.
-  \]
-  \vspace*{-1em}
-
-  \begin{itemize}
-    \item True for~$R = \affl[X]/(X^2)$ (microaffinity).
-    \item True for~$R = \affl[X]$ (every function is a polynomial).
-    \item True for \hil{any} finitely presented~$\affl$-algebra~$R$.
-  \end{itemize}
-
-  Any known property of~$\affl$ follows from this \\
-  \hil{synthetic quasicoherence}.
-
-  \centering
-  \begin{varblock}{0.6344\textwidth}{}
-    \centering
-    \hil{the mystery of nongeometric sequents}
-    \par
-  \end{varblock}
-
-  \jnote{1}{Let~$R$ be an~$\affl$-algebra. An element~$f \in R$ induces
-  an~$\affl$-valued function on~$\Spec(R)$; functions of this form can
-  reasonably be called ``algebraic''. In a synthetic context, there should be
-  no other~$\affl$-valued functions on~$\Spec(R)$ as these algebraic ones, and
-  different algebraic expressions should yield different functions. This is
-  precisely what the bijectivity of the displayed map expresses (in a positive
-  way).
-
-  In synthetic differential geometry, the closest cousin of synthetic algebraic
-  geometry, the analogue of the displayed map is only bijective for Weil
-  algebras such as~$\affl[X]/(X^2)$ or $\affl[X,Y]/(X^2,XY)$, not for arbitrary
-  finitely presented~$\affl$-algebras. This is a major difference to synthetic
-  differential geometry.}
-
-  \jnote{2}{The notion of synthetic quasicoherence is interesting for a number
-  of reasons:\vspace*{-1.2em}
-  \begin{itemize}\justifying
-    \item All currently known properties of~$\affl$, such as all the properties
-    listed on the previous slide, follow from the statement that~$\affl$ is
-    synthetically quasicoherent.
-
-    For instance, here is how we can verify the field property. Let~$x \?
-    \affl$ such that~$x \neq 0$. Set~$R = \affl/(x)$. Then~$\Spec(R) =
-    \emptyset$. Thus~$(\affl)^{\Spec(R)}$ is a singleton. Hence~$R = 0$.
-    Therefore~$x$ is invertible.
-    \item Given an~$\affl$-module~$E$, we can formulate the following variant
-    of the axiom of synthetic quasicoherence: ``For any finitely
-    presented~$\affl$-algebra~$R$, the canonical map~$R \otimes_\affl E \to
-    E^{\Spec(R)}$ is bijective.'' This axiom is satisfied if and only if~$E$ is
-    induced by a quasicoherent sheaf of~$\O_S$-modules.
-    \item The notion of synthetic quasicoherence is central to synthetic
-    algebraic geometry. The notions of synthetic open immersions, closed
-    immersion, schemes and several others all refer to synthetic
-    quasicoherence.
-  \end{itemize}}
-
-  \jnote{3}{An analogue of synthetic quasicoherence holds in the classifying
-  topos of rings, demonstrating that even presheaf toposes can validate
-  interesting nontrivial nongeometric sequents.
-
-  We believe that an analogue of synthetic quasicoherence holds for the generic
-  model of any geometric theory. This is work in progress. If true, this would
-  yield a major source of nongeometric sequents in classifying toposes. Because
-  of the many applications on nongeometric sequents, it's very desirable to
-  possess such a source.
-
-  The mystery of nongeometric sequents is this: On the one hand, they are very
-  useful to have because of surprising applications; on the other hand, they
-  are as of yet quite elusive.}
-\end{frame}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
-\begin{frame}{Classifying toposes in algebraic geometry}
-  \small
-  \centering
-  \begin{tabular}{ll}
-    \toprule
-    (big) topos & classified theory \\\midrule
-    Zariski & local rings [Hakim 1972] \\[0.6em]
-    étale & separably closed local rings [Hakim 1972, Wraith 1979] \\[0.6em]
-    \usebeamercolor[fg]{item}{fppf} & \usebeamercolor[fg]{item}{fppf-local rings} \\
-    & \textcolor{red!90}{(conjecturally: algebraically closed local rings)} \\[0.6em]
-    \textcolor{red!80}{ph} & \textcolor{red!80}{?? (conjecturally: algebraically closed valuation rings} \\
-    & \qquad \textcolor{red!80}{validating the projective Nullstellensatz)} \\[0.6em]
-    \usebeamercolor[fg]{item}{surjective} & \usebeamercolor[fg]{item}{algebraically closed geometric fields} \\[0.6em]
-    \textcolor{red!80}{$\neg\neg$} & \textcolor{red!80}{?? (conjecturally: algebraically closed geometric} \\
-    & \qquad \textcolor{red!80}{fields which are integral over the base)} \\[0.6em]
-    \usebeamercolor[fg]{item}{infinitesimal} & \usebeamercolor[fg]{item}{local algebras together with a nilpotent ideal [Hutzler 2018]} \\[0.6em]
-    \textcolor{red!100}{crystalline} & \textcolor{red!100}{??} \\
-    \bottomrule
-  \end{tabular}
-
-  \jnote{1}{Toposes, and also more specifically classifying toposes, originated
-  in algebraic geometry. It is therefore deeply embarrassing that as of now,
-  still very little is known about the theories classified by the major toposes
-  in active use by algebraic geometers.
-  \begin{center}
-    \includegraphics[width=0.99\textwidth]{wraith-generic-galois-theory} \\
-    \footnotesize
-    Gavin Wraith. Generic Galois theory of local rings. 1979. \\
-    (A cult classic and must-read for anyone interested in \\
-    the intersection of topos theory and algebraic geometry.)
-  \end{center}}
-
-  \jnote{2}{For almost forty years, only the big Zariski topos and its étale
-  subtopos were understood in that way.
-  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{These~2017
-  notes} answer the question for the fppf topology and the surjective topology
-  (in Section~21) and state conjectures for the ph topology and the
-  double negation topology. However, while good to have, the answer for the
-  fppf topology remains unsatisfactory, since Wraith's conjecture that the fppf
-  topos classifies the simpler theory of algebraically closed local rings has
-  neither been confirmed nor refuted.
-
-  A couple of weeks ago, Matthias Hutzler managed to determine the theory
-  classified by the big infinitesimal topos of a ring~$A$: It classifies
-  pairs~$(B,\aaa)$ consisting of a local~$A$-algebra~$B$ and a nilpotent
-  ideal~$\aaa \subseteq B$. Details will be in his forthcoming Master's thesis.
-  He is currently working on answering the question for the closely related big
-  crystalline topos.
-
-  It will be exciting to learn what the crystalline topos and the many other
-  toposes in algebraic geometry classify; how algebraic geometry can profit
-  from these discoveries; and which new flavors of synthetic algebraic geometry
-  they unlock.}
-\end{frame}}
-
-% XXX: Mehr Leute zitieren!
-% XXX: Einordnung in Gesamtkontext (etwa SDG, Kock--Lawvere, Coquand & Co., ...)
-% XXX: Folie zu klassifizierenden Topoi
-
 \end{document}

### 2018-10-04 slides-pssl104.tex
Slides almost finished

@@
     \end{minipage}
-  \end{center}}
+  \end{center}
+}
 
@@
 \setbeamertemplate{frametitle}{%
-  \vskip1.2em%
+  \vskip1.0em%
   \leavevmode%
@@
 \newcommand{\hil}[1]{{\usebeamercolor[fg]{item}{\textbf{#1}}}}
-\newcommand{\bad}[1]{\textcolor{red!90}{\textbf{#1}}}
+\newcommand{\bad}[1]{\textcolor{red!90}{#1}}
 
@@
 
-\begin{frame}
-  \begin{columns}
-    \begin{column}{0.5\textwidth}
-      \centering{\usebeamercolor[fg]{item}{\textbf{\Large Sheaf cohomology}}}\bigskip
-
+%{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[width=\paperwidth]{leray-shaded}\end{minipage}}
+\begin{frame}{Sheaf cohomology}
+  \vspace*{-1.2em}
+  \begin{varblock}{\textwidth}{}
     \justifying
     Let~$E$ be a sheaf of modules over a space~$X$.
+    Let~$\Gamma$ be the global sections functor.
     Choose an \bad{injective resolution} $0 \to E \to I^0 \to I^1 \to \cdots$.
-      Then $H^n(X; E)$ is the $n$-th cohomology of
-      \[ (0 \xrightarrow{\phantom{d^0}} I^0(X) \xrightarrow{d^0} I^1(X) \xrightarrow{d^1} \cdots), \]
-      that is~$\operatorname{ker}(d^n)\mathop{/}\operatorname{im}(d^{n-1})$.
-    %  \begin{align*}
-    %    H^n(X; E) &\defeq \text{$n$-th cohomology of $(0 \to I^0(X) \to I^1(X) \to \cdots)$} \\
-    %    &\phantom{\vcentcolon}= \operatorname{ker}(I^n(X) \to I^{n+1}(X))\mathop{/}\operatorname{im}(I^{n-1}(X) \to I^n(X))
-    %  \end{align*}
-    %  is well-defined up to isomorphism.
+    Then the \hil{$\boldsymbol{n}$-th cohomology of $\boldsymbol{E}$ is}
+    \begin{align*}
+      H^n(X; E) &\defeq \text{$n$-th cohomology of $(0 \to \Gamma I^0 \to \Gamma I^1 \to \cdots)$} \\
+      &\phantom{\vcentcolon}= \operatorname{ker}(\Gamma I^n \to \Gamma I^{n+1})\mathop{/}\operatorname{im}(\Gamma I^{n-1} \to \Gamma I^n).
+    \end{align*}
+  \end{varblock}
 
@@
     \item The modules~$H^n(X; E)$ are important invariants.
-        \item Let~$A$ be a ring. Then $H^n(X; \underline{A}) = H_{\text{sing}}^n(X;
-        A)$, if~$X$ is locally semiconnected [XXX].
+
+          [ $\chi(X; \O_X) = 1 - \operatorname{genus}_X$, $(C \cdot C') = \chi(\O_C \otimes_{\O_X}^{\mathbb{L}} \O_C')$ ]
+          \\[0.6em]
+    \item Let~$A$ be an abelian group. Let~$X$ be semi-locally contractible.
+          Then $H^n(X; \underline{A}) = H_{\text{sing}}^n(X; A)$.
+          \\[0.6em]
+    \item Let~$f : X \to Y$ be continuous. Then there is a spectral sequence
+          $H^i(Y; R^jf_*(E)) \Rightarrow H^{i+j}(X; E)$.
   \end{itemize}
+\end{frame}
+%}
+
+\section{Constructive mathematics}
+
+\begin{frame}{Constructive mathematics}
+  \centering
+
+  \vspace*{-1em}
+  mathematics without $\varphi \vee \neg\varphi$,\ \ $\neg\neg\varphi \Rightarrow \varphi$,\ \ axiom of choice
+  \bigskip
+
+  \centering\includegraphics[width=0.5\textwidth]{andrej-bauer-five-stages} \\
+  \scriptsize\href{https://video.ias.edu/members/1213/0318-AndrejBauer}{Andrej Bauer, talk at the IAS}\par
+  \small
+  \bigskip
+
+  \begin{columns}
+    \begin{column}{0.50\textwidth}
+      \centering
+      \hil{Axiomatic freedom}
+
+      \footnotesize
+      \mbox{``Every map~$\NN \to \NN$ is computable.''}
+      \mbox{``Every map~$\RR \to \RR$ is continuous.''}
+      \mbox{``Every map~$\affl \to \affl$ is polynomial.''}
+      \mbox{``Heyting Arithmetic has exactly one model.''}
+      \mbox{``The subsets of~$\{\heartsuit\}$ form a proper class.''}
+      \mbox{``There is an injection~$\RR \to \NN$.''}
     \end{column}
 
-    \begin{column}{0.6\textwidth}
-      \centering{\usebeamercolor[fg]{item}{\textbf{\Large Constructive mathematics}}}\bigskip
+    \begin{column}{0.50\textwidth}
+      \centering
+      \hil{Applications}
 
-      \justifying
+      \footnotesize
+      \mbox{program extraction} \\
+      \mbox{synthetic differential geometry} \\
+      \mbox{synthetic algebraic geometry} \\
+      \mbox{synthetic domain theory} \\
+      \mbox{new reduction techniques in algebra} \\
+      \mbox{Bohr topos for quantum mechanics}
     \end{column}
@@
 
+\section{Relativization by internalization}
+
+\begin{frame}{Relativization by internalization}
+  \vspace*{-0.5em}
+  Let~$X$ be a space. The \hil{internal language} of the topos~$\Sh(X)$ allows us to reason
+  about sheaves on~$X$ in \hil{naive element-based terms}.
+
+  \only<1>{
+    \centering
+    \rotatebox{90}{\tiny\scalebox{0.8}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
+    \includegraphics[width=0.7\textwidth]{external-internal-small}
+    \par
+  }
+  \pause
+
+  \begin{center}
+    \small
+    \begin{tabular}{ll}
+      \toprule
+      externally & internally to $\Sh(X)$ \\
+      \midrule
+      sheaf & set/type \\
+      morphism of sheaves & map between sets \\
+      %monomorphism of sheaves & injective map between sets \\
+      %epimorphism of sheaves & surjective map between sets \\
+      sheaf of cont.\@ real-valued functions & set of Dedekind reals \\
+      %sheaf of modules & module \\
+      %coherent sheaf of modules & coherent module \\
+      over-locale~$f : Y \to X$ & locale~$I(Y)$ \\
+      sheaf over~$Y$ & sheaf over~$I(Y)$ \\
+      higher direct image $R^n f_* E$ & \bad{??} \\
+      \midrule
+      \pause
+      \begin{minipage}{5.1cm}
+        In continuous families of continuous functions with opposite signs,
+        zeros can locally be picked continuously.
+      \end{minipage} &
+      \begin{minipage}{4.4cm}
+        \raggedright
+        The intermediate value theorem holds.
+      \end{minipage} \\
+      \midrule
+      \begin{minipage}{5.1cm}
+        Every finite type sheaf of modules is finite locally free \emph{on a dense open}.
+      \end{minipage} &
+      \begin{minipage}{4.4cm}
+        Every finitely generated vector space is \emph{not not} finite free.
+      \end{minipage} \\
+      \midrule
+      \begin{minipage}{5.1cm}
+        \raggedright
+        Grothendieck's generic freeness lemma holds.
+      \end{minipage} &
+      \begin{minipage}{4.2cm}
+        \raggedright
+        (Some trivial observation about modules over fields.)
+      \end{minipage} \\
+      \bottomrule
+    \end{tabular}
+  \end{center}
+\end{frame}
+
+\section{Internalizing higher direct images}
+
+\begin{frame}{On injective objects}
+  \begin{block}{}
+    A set~$M$ is \hil{injective} iff for any injection~$A \to B$,
+    any map~$A \to M$ extends to a map on~$B$.
+  \end{block}
+  \vspace*{-1em}
+
+  \begin{itemize}
+    \small
+    \item ``A set is injective iff it's inhabited'' is
+          a \hil{constructive taboo}.
+    \item Constructively, there are still \hil{enough injective sets}.
+          % Every set~$M$ embeds into an injective set, for instance~$P(M)$ or~$P_{\leq1}(M)$.
+  \end{itemize}
+  \pause
+
+  \begin{block}{}
+    A module~$M$ is \hil{injective} iff for any linear injection~$A \to B$,
+    any linear map~$A \to M$ extends to a linear map on~$B$.
+  \end{block}
+  \vspace*{-1em}
+
+  \begin{itemize}
+    \small
+    \item It's consistent with \textbf{ZF} that there are no injective modules [Blass 1979].
+    \item The existence of enough injective modules is \hil{constructively neutral}.
+  \end{itemize}
+  \pause
+
+  \begin{block}{}
+    \justifying
+    A sheaf of modules~$M$ is \hil{injective} iff for any linear monomorphism~$A \to B$,
+    any linear morphism~$A \to M$ extends to a linear morphism on~$B$.
+  \end{block}
+  \vspace*{-1em}
+
+  \begin{itemize}
+    \small
+    \item Assuming choice, there are enough injectives over any site.
+    \item Assuming Zorn's lemma, a sheaf of modules over a locale~$X$ is injective iff, from the internal
+    point of view of~$\Sh(X)$, it is an injective module.
+  \end{itemize}
+\end{frame}
+
+\section{Flabby resolutions}
+
+\begin{frame}{Flabby resolutions}
+  \begin{block}{}
+    A sheaf~$E$ on a space~$X$ is \hil{flabby} iff any local section~$s \in
+    E(U)$ on an open~$U$ extends to a global section~$\bar s \in E(X)$: $\bar s|_U = s$.
+    % XXX: change
+  \end{block}
+  \vspace*{-1em}
+
+  \begin{itemize}
+    \small
+    \item Injective sheaves of sets and injectives sheaves of modules are flabby.
+    \item Every sheaf embeds into a flabby sheaf.
+    \item Assuming the law of excluded middle, any sheaf of modules over a
+    topological space embeds into a flabby sheaf of modules.
+    \item Assuming Zorn's lemma, flabby sheaves of modules are \hil{acyclic for the global sections functor}.
+    \item Assuming \bad{??}, sheaf cohomology and higher direct images can be computed using \hil{flabby resolutions}.
+  \end{itemize}
+
+  \begin{block}{}
+    A set~$M$ is \hil{flabby} iff for any subterminal~$K \subseteq M$, there is
+    an element~$x \in M$ such that~$K \subseteq \{x\}$.
+  \end{block}
+  \vspace*{-1em}
+
+  \begin{itemize}
+    \small
+    \item A sheaf is flabby iff, from the internal point of view, it's a flabby set.
+    \item Every set embeds into a flabby set.
+    \item \bad{Open question:} Does every module embed into a flabby module?
+  \end{itemize}
+\end{frame}
+
+% XXX: Results about Eff
+% XXX: Summary slide
+
 \end{document}

### 2018-10-05 slides-pssl104.tex
More work on slides

@@
   \tikz[remember picture,baseline]{
-    \node[anchor=base,inner sep=0,outer sep=0] (#2) {#2};
-    \node[visible on=#1,overlay,rectangle callout,rounded corners,callout relative pointer={(0.3cm,0.5cm)},fill=blue!20] at ($(#2.north)+(-0.1cm,-1.1cm)$) {#3};
+    \node[anchor=base,inner sep=0,outer sep=0,color=blue!90] (#2) {#2};
+    \node[visible on=#1,overlay,rectangle callout,rounded corners,callout
+    relative pointer={(0.0cm,-0.5cm)},fill=blue!20] at ($(#2.north)+(-0.1cm,0.7cm)$) {#3};
+  }%
+}
+
+\newcommand{\pointthisbelow}[3]{%
+  \tikz[remember picture,baseline]{
+    \node[anchor=base,inner sep=0,outer sep=0,color=blue!90] (#2) {#2};
+    \node[visible on=#1,overlay,rectangle callout,rounded corners,callout
+    relative pointer={(0.0cm,0.5cm)},fill=blue!20] at ($(#2.north)+(-0.1cm,-1.0cm)$) {#3};
   }%
@@
 
-\section{Sheaf cohomology}
+\section[Cohomology]{Sheaf cohomology}
 
@@
 
-          [ $\chi(X; \O_X) = 1 - \operatorname{genus}_X$, $(C \cdot C') = \chi(\O_C \otimes_{\O_X}^{\mathbb{L}} \O_C')$ ]
+          [ $\chi(X; \O_X) = 1 - \operatorname{genus}_X$,
+          $(C \cdot C') = \chi(\O_C \otimes_{\O_X}^{\mathbb{L}} \O_C')$, \ldots ]
           \\[0.6em]
@@
           \\[0.6em]
-    \item Let~$f : X \to Y$ be continuous. Then there is a spectral sequence
-          $H^i(Y; R^jf_*(E)) \Rightarrow H^{i+j}(X; E)$.
+    \item Let~$f : X \to B$ be continuous. Then there is a spectral sequence
+          $H^i(B; R^jf_*(E)) \Longrightarrow H^{i+j}(X; E)$.
   \end{itemize}
@@
 
-\section{Constructive mathematics}
+\section[Constructivism]{Constructive mathematics}
 
@@
   \centering\includegraphics[width=0.5\textwidth]{andrej-bauer-five-stages} \\
-  \scriptsize\href{https://video.ias.edu/members/1213/0318-AndrejBauer}{Andrej Bauer, talk at the IAS}\par
+  \scriptsize\href{https://video.ias.edu/members/1213/0318-AndrejBauer}{Andrej Bauer at an IAS talk}\par
   \small
@@
       \mbox{``The subsets of~$\{\heartsuit\}$ form a proper class.''}
-      \mbox{``There is an injection~$\RR \to \NN$.''}
+      \mbox{``There is an injection~$\RR \to \NN$.''} \\
+      \rotatebox{90}{\dots}
     \end{column}
@@
       \mbox{new reduction techniques in algebra} \\
-      \mbox{Bohr topos for quantum mechanics}
+      \mbox{Bohr topos for quantum mechanics} \\
+      \rotatebox{90}{\dots}
     \end{column}
@@
   \only<1>{
+    \vspace*{0.5em}
     \centering
@@
 
-\begin{frame}{On injective objects}
+\begin{frame}{Internalizing higher direct images}
   \begin{block}{}
@@
     \justifying
-    A sheaf of modules~$M$ is \hil{injective} iff for any linear monomorphism~$A \to B$,
-    any linear morphism~$A \to M$ extends to a linear morphism on~$B$.
+    A sheaf of modules~$M$ is \hil{injective} iff for any linear monomorphism
+    \mbox{$A \!\!\to\!\! B$,
+    any linear morphism~$A \!\!\to\!\! M$ extends to a linear morphism on~$B$.}
   \end{block}
@@
   \begin{itemize}
-    \small
-    \item Injective sheaves of sets and injectives sheaves of modules are flabby.
+    \small\justifying
+    %\item Injective sheaves of sets and injectives sheaves of modules are flabby.
     \item Every sheaf embeds into a flabby sheaf.
@@
     \item Assuming Zorn's lemma, flabby sheaves of modules are \hil{acyclic for the global sections functor}.
-    \item Assuming \bad{??}, sheaf cohomology and higher direct images can be computed using \hil{flabby resolutions}.
+    Hence, assuming \bad{??}, sheaf cohomology and higher direct images can be computed using \hil{flabby resolutions}.
   \end{itemize}
 
+  \bigskip
+  \bigskip
+
   \begin{block}{}
-    A set~$M$ is \hil{flabby} iff for any subterminal~$K \subseteq M$, there is
-    an element~$x \in M$ such that~$K \subseteq \{x\}$.
+    A set~$M$ is \hil{flabby} iff any \pointthis{<1>}{subterminal}{$\forall x,y
+    \in K\_ x = y$}~$K \subseteq M$ is a \pointthis{<1>}{subsingleton}{$\exists
+    a \in M\_ \forall x \in K\_ x = a$}.
   \end{block}
@@
 
-% XXX: Results about Eff
-% XXX: Summary slide
+\begin{frame}{Flabbiness as an organizing principle}
+  \begin{block}{}
+    A set~$M$ is \hil{flabby} iff any \pointthisbelow{<1>}{subterminal}{$\forall x,y
+    \in K\_ x = y$}~$K \subseteq M$ is a \pointthisbelow{<1>}{subsingleton}{$\exists
+    a \in M\_ \forall x \in K\_ x = a$}.
+  \end{block}
+  \bigskip
+  \bigskip
+  \justifying
+
+  \textbf{Proposition.} Let~$E$ be a sheaf over a locale~$X$. Then~$E$ is flabby
+  iff it is a flabby set from the point of view of~$\Sh(X)$.
+
+  \textbf{Proof.} Routine/Zorn's lemma.
+
+  \textbf{Proposition.} Let~$M$ be a sheaf of modules over a locale~$X$.
+  Then~$M$ is injective iff it is injective from the point of view of~$\Sh(X)$.
+
+  \textbf{Proof.} (Only ``$\Leftarrow$''.) Let~$i : A \to B$ be a linear
+  monomorphism. Let~$f : A \to M$ be a linear morphism. Then verify, internally, that
+  the set~$E \defeq \{ \bar f : B \to M \,|\, \bar f \circ i = f \}$ is flabby.
+%  Let~$K \subseteq E$ be a subterminal.
+%  We consider the injectivity diagram
+%  \[ \xymatrix{
+%    i[A] + B' \ar@{^{(}->}[r]\ar[d]_g & B \ar@{-->}@/^/[ld] \\
+%    I
+%  } \]
+%  where~$B'$ is the submodule~$\{ s \in B \,|\, \text{$s = 0$ or $K$ is
+%  inhabited} \} \subseteq B$ and the solid vertical arrow~$g$ is defined in the following
+%  way: Let~$s \in i[A] + B'$. Then~$s = i(a) + t$ for an element~$a \in A$ and an
+%  element~$t \in B'$. Since~$t \in B'$, $t = 0$ or~$K$ is inhabited. If~$t = 0$,
+%  we set~$g(s) \defeq f(a)$. If~$K$ is inhabited, we set~$g(s) \defeq f(a) +
+%  \bar{f}(s)$, where~$\bar{f}$ is any element of~$K$. These prescriptions
+%  determine a well-defined map.
+\end{frame}
+
+\section[In the effective topos]{Flabbiness in the effective topos}
+
+\begin{frame}{Flabbiness in the effective topos}
+  \begin{block}{}
+    A set~$M$ is \hil{flabby} iff any \pointthisbelow{<1>}{subterminal}{$\forall x,y
+    \in K\_ x = y$}~$K \subseteq M$ is a \pointthisbelow{<1>}{subsingleton}{$\exists
+    a \in M\_ \forall x \in K\_ x = a$}.
+  \end{block}
+  \bigskip
+  \bigskip
+  \bigskip
+
+  \justifying
+  \textbf{Proposition.}
+  Let~$X$ be an effective object in the effective topos. Then
+  \[ \text{``If~$X$ is flabby, any endomap on~$X$ has a fixed point.''} \]
+  from the point of view of the effective topos.
+  \medskip
+
+  \textbf{Proposition.}
+  Assuming the law of excluded middle, any~$\neg\neg$-separated
+  module in the effective topos can be embedded into a flabby module.
+  \medskip
+
+  \textbf{Question.} Are there enough flabby modules in the effective
+  topos?
+\end{frame}
+
+% XXX: ctcl.pdf, realizability book
+% XXX: Harting
+
+\begin{frame}{State of affairs}
+  \centering
+
+  \textcolor{blue!90}{\Huge :-)}
+
+  The existence of enough flabby modules is constructively neutral. \\
+  Higher direct images can be understood as internal sheaf cohomology.
+
+  \bigskip
+
+  \bad{\Huge \scalebox{2}{:-(}}
+
+  Flabby sheaves can fail to be acyclic, constructively. \\
+  There is still no general constructive framework for sheaf cohomology.
+  Even though:
+
+  \begin{itemize}
+    \item Basic homological algebra is entirely constructive.
+    \item There are algorithms for computing cohomology [Barakat XXX].
+    \item Čech methods work constructively, even in a synthetic context.
+  \end{itemize}
+\end{frame}
+\addtocounter{framenumber}{-1}
 

### 2018-10-05 slides-pssl104.tex
Slide on singular cohomology

@@
 \usepackage{booktabs}
-\usepackage{stmaryrd}
+\usepackage{stmaryrd,wasysym}
 \usepackage{array}
@@
 
+\newcommand{\smallsphere}{\ensuremath{\vcenter{\hbox{\includegraphics[height=1em]{sphere}}}}}
+\newcommand{\smalltorus}{\ensuremath{\vcenter{\hbox{\includegraphics[height=1em]{torus}}}}}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace{5.6cm}\includegraphics[width=\paperwidth]{leray}\end{minipage}}
+\begin{frame}{Singular cohomology}
+  \centering
+
+  Is $\vcenter{\hbox{\includegraphics[height=0.1\textwidth]{sphere}}}$ homeomorphic to
+  $\vcenter{\hbox{\includegraphics[height=0.1\textwidth]{torus}}}$? No:
+  \begin{align*}
+    H^0(\smallsphere,\ZZ) &\cong \ZZ &
+    H^1(\smallsphere,\ZZ) &\cong 0   &
+    H^2(\smallsphere,\ZZ) &\cong \ZZ \\
+    H^0(\smalltorus,\ZZ) & \cong \ZZ &
+    H^2(\smalltorus,\ZZ) & \cong \ZZ \oplus \ZZ &
+    H^2(\smalltorus,\ZZ) & \cong \ZZ
+  \end{align*}
+
+  \begin{block}{}
+  \justifying
+  Given~$f : X \to B$, can we compute the cohomology of~$X$ if we
+  understand the cohomology of~$B$ and the cohomology of the fibers of~$f$?
+  \end{block}{}
+\end{frame}}
+
 %{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[width=\paperwidth]{leray-shaded}\end{minipage}}
@@
     \begin{align*}
-      H^n(X; E) &\defeq \text{$n$-th cohomology of $(0 \to \Gamma I^0 \to \Gamma I^1 \to \cdots)$} \\
+      H^n(X, E) &\defeq \text{$n$-th cohomology of $(0 \to \Gamma I^0 \to \Gamma I^1 \to \cdots)$} \\
       &\phantom{\vcentcolon}= \operatorname{ker}(\Gamma I^n \to \Gamma I^{n+1})\mathop{/}\operatorname{im}(\Gamma I^{n-1} \to \Gamma I^n).
@@
   \begin{itemize}
-    \item The modules~$H^n(X; E)$ are important invariants.
+    \item The modules~$H^n(X, E)$ are important invariants.
 
-          [ $\chi(X; \O_X) = 1 - \operatorname{genus}_X$,
+          [ $\chi(X, \O_X) = 1 - \operatorname{genus}_X$,
           $(C \cdot C') = \chi(\O_C \otimes_{\O_X}^{\mathbb{L}} \O_C')$, \ldots ]
@@
     \item Let~$A$ be an abelian group. Let~$X$ be semi-locally contractible.
-          Then $H^n(X; \underline{A}) = H_{\text{sing}}^n(X; A)$.
+          Then $H^n(X, \underline{A}) = H_{\text{sing}}^n(X, A)$.
           \\[0.6em]
     \item Let~$f : X \to B$ be continuous. Then there is a spectral sequence
-          $H^i(B; R^jf_*(E)) \Longrightarrow H^{i+j}(X; E)$.
+          $H^i(B, R^jf_*(E)) \Longrightarrow H^{i+j}(X, E)$.
   \end{itemize}
@@
 
-\section{Flabby resolutions}
+\section{Flabby objects}
 
@@
 
-  \textcolor{blue!90}{\Huge :-)}
+  \textcolor{blue!90}{\Huge \smiley}
 
-  The existence of enough flabby modules is constructively neutral. \\
+  The existence of enough injective modules is constructively neutral. \\
   Higher direct images can be understood as internal sheaf cohomology.
@@
 
-  \bad{\Huge \scalebox{2}{:-(}}
+  \bad{\Huge \scalebox{2}{\frownie}}
 
   Flabby sheaves can fail to be acyclic, constructively. \\
-  There is still no general constructive framework for sheaf cohomology.
+  There is still no general constructive framework for sheaf cohomology. \\
   Even though:
@@
     \item Basic homological algebra is entirely constructive.
-    \item There are algorithms for computing cohomology [Barakat XXX].
+    \item There are algorithms for computing cohomology [Barakat, \ldots].
     \item Čech methods work constructively, even in a synthetic context.

### 2018-10-05 paper-flabby-objects.tex
Slides almost finished
More work on slides
Much nicer beginning (I hope)

@@
 \begin{proof}For the first claim, we set~$X \defeq \{ \bar{f} \in I^B \,|\, \bar{f} \circ i =
-f \}$. Let~$K \subseteq X$ be a subsingleton. We consider the injectivity diagram
+f \}$. Let~$K \subseteq X$ be a subterminal. We consider the injectivity diagram
 \[ \xymatrix{
@@
 set~$X \defeq \{ \bar{f} \in \Hom_R(B,I) \,|\, \bar{f} \circ i =
-f \}$. Let~$K \subseteq X$ be a subsingleton. We consider the injectivity diagram
+f \}$. Let~$K \subseteq X$ be a subterminal. We consider the injectivity diagram
 \[ \xymatrix{
@@
 } \]
-where~$B'$ is the submodule~$\{ s \in B \,|\, \text{$s = 0$ or $K$ is
+where~$B'$ is the submodule~$\{ t \in B \,|\, \text{$t = 0$ or $K$ is
 inhabited} \} \subseteq B$ and the solid vertical arrow~$g$ is defined in the following
@@
 
+% In Eff gilt: Jede negneg-separierte abelsche Gruppe bettet in eine injektive
+% ein. Nämlich A --> ΔΓA --> ΔI für ΓA --> I Einbettung in eine injektive
+% Gruppe.
+
 The intuitive

### 2018-10-05 slides-pssl104.tex
Much nicer beginning (I hope)

@@
 
+\begin{frame}{Flabby sets}
+  Let~$M$ be a set. A subset~$K \subseteq M$ is \ldots
+  \begin{itemize}
+    \item a \hil{subterminal} iff $\forall x,y \in K\_ x = y$.
+    \item \mbox{a \hil{subsingleton} iff $\exists a \in M\_ \forall x \in K\_ x = a$, that is,} \\
+    \mbox{\phantom{a \hil{subsingleton}} iff $K \subseteq \{ a \}$ for some~$a \in M$.}
+  \end{itemize}
+
+  Trivially, any subsingleton is a subterminal.
+
+  \textbf{Definition.}\visible<2->{\hil{$^\star$}}
+  The set~$M$ is \hil{flabby} iff the converse holds.
+
+  Any flabby set is inhabited.
+  %Conversely, any inhabited set is flabby iff the law of excluded middle holds.
+
+  \visible<4->{
+    \textbf{Proposition.} Any set embeds into a flabby set.
+
+    \justifying
+    \textbf{Proof.} We have~$M \hookrightarrow P(M)$, and~$P(M)$ is flabby:
+    Let~$K \subseteq P(M)$ be a subterminal. Then~$K \subseteq \{ \bigcup K \}$.
+    % for if~$A \in K$, then~$K = \{ A \}$ and hence~$\bigcup K = K$.
+  }
+
+  \visible<5->{
+    \bad{\textbf{Open question.}} Does any module embed into a flabby module?
+  }
+
+  \vfill
+  \visible<3->{
+    \small
+    \hil{$^\star$ This talk is set in the context of constructive mathematics:} \\
+    \hil{\phantom{$^\star$}} mathematics without $\varphi \vee \neg\varphi$,\ \ $\neg\neg\varphi \Rightarrow \varphi$,\ \ axiom of choice
+  }
+\end{frame}
+
 \section[Cohomology]{Sheaf cohomology}
@@
       \mbox{synthetic algebraic geometry} \\
+      \mbox{synthetic probability theory} \\
       \mbox{synthetic domain theory} \\
@@
           % Every set~$M$ embeds into an injective set, for instance~$P(M)$ or~$P_{\leq1}(M)$.
+    \item Any injective set is flabby.
   \end{itemize}
@@
     E(U)$ on an open~$U$ extends to a global section~$\bar s \in E(X)$: $\bar s|_U = s$.
-    % XXX: change
   \end{block}
@@
   \begin{itemize}
-    \small\justifying
+    \justifying
     %\item Injective sheaves of sets and injectives sheaves of modules are flabby.
-    \item Every sheaf embeds into a flabby sheaf.
-    \item Assuming the law of excluded middle, any sheaf of modules over a
-    topological space embeds into a flabby sheaf of modules.
+    \item Assuming Zorn's lemma:
+
+          A sheaf is flabby iff, from the internal point of view, it's a flabby
+          set. \\[0.6em]
+    \item Assuming the law of excluded middle:
+
+          Any sheaf of modules over a topological space embeds into a flabby
+          sheaf of modules. \\[0.6em]
     \item Assuming Zorn's lemma, flabby sheaves of modules are \hil{acyclic for the global sections functor}.
     Hence, assuming \bad{??}, sheaf cohomology and higher direct images can be computed using \hil{flabby resolutions}.
-  \end{itemize}
-
-  \bigskip
-  \bigskip
-
-  \begin{block}{}
-    A set~$M$ is \hil{flabby} iff any \pointthis{<1>}{subterminal}{$\forall x,y
-    \in K\_ x = y$}~$K \subseteq M$ is a \pointthis{<1>}{subsingleton}{$\exists
-    a \in M\_ \forall x \in K\_ x = a$}.
-  \end{block}
-  \vspace*{-1em}
-
-  \begin{itemize}
-    \small
-    \item A sheaf is flabby iff, from the internal point of view, it's a flabby set.
-    \item Every set embeds into a flabby set.
-    \item \bad{Open question:} Does every module embed into a flabby module?
   \end{itemize}
@@
 \begin{frame}{Flabbiness as an organizing principle}
-  \begin{block}{}
-    A set~$M$ is \hil{flabby} iff any \pointthisbelow{<1>}{subterminal}{$\forall x,y
-    \in K\_ x = y$}~$K \subseteq M$ is a \pointthisbelow{<1>}{subsingleton}{$\exists
-    a \in M\_ \forall x \in K\_ x = a$}.
-  \end{block}
-  \bigskip
-  \bigskip
-  \justifying
+%  \begin{block}{}
+%    A set~$M$ is \hil{flabby} iff any \pointthisbelow{<1>}{subterminal}{$\forall x,y
+%    \in K\_ x = y$}~$K \subseteq M$ is a \pointthisbelow{<1>}{subsingleton}{$\exists
+%    a \in M\_ \forall x \in K\_ x = a$}.
+%  \end{block}
+%  \bigskip
+%  \bigskip
+%  \justifying
 
-  \textbf{Proposition.} Let~$E$ be a sheaf over a locale~$X$. Then~$E$ is flabby
-  iff it is a flabby set from the point of view of~$\Sh(X)$.
+%  \textbf{Proposition.} Let~$E$ be a sheaf over a locale~$X$. Then~$E$ is flabby
+%  iff it is a flabby set from the point of view of~$\Sh(X)$.
+%
+%  \textbf{Proof.} Routine/Zorn's lemma.
 
-  \textbf{Proof.} Routine/Zorn's lemma.
-
+  \justifying\small
   \textbf{Proposition.} Let~$M$ be a sheaf of modules over a locale~$X$.
@@
   the set~$E \defeq \{ \bar f : B \to M \,|\, \bar f \circ i = f \}$ is flabby.
-%  Let~$K \subseteq E$ be a subterminal.
-%  We consider the injectivity diagram
-%  \[ \xymatrix{
-%    i[A] + B' \ar@{^{(}->}[r]\ar[d]_g & B \ar@{-->}@/^/[ld] \\
-%    I
-%  } \]
-%  where~$B'$ is the submodule~$\{ s \in B \,|\, \text{$s = 0$ or $K$ is
-%  inhabited} \} \subseteq B$ and the solid vertical arrow~$g$ is defined in the following
-%  way: Let~$s \in i[A] + B'$. Then~$s = i(a) + t$ for an element~$a \in A$ and an
-%  element~$t \in B'$. Since~$t \in B'$, $t = 0$ or~$K$ is inhabited. If~$t = 0$,
-%  we set~$g(s) \defeq f(a)$. If~$K$ is inhabited, we set~$g(s) \defeq f(a) +
-%  \bar{f}(s)$, where~$\bar{f}$ is any element of~$K$. These prescriptions
-%  determine a well-defined map.
+  \pause
+
+  Let~$K \subseteq E$ be a subterminal.
+  We consider the injectivity diagram
+  \[ \xymatrix{
+    i[A] + B' \ar@{^{(}->}[r]\ar[d]_g & B \ar@{-->}@/^/[ld]^{\bar g} \\
+    I
+  } \]
+  where~$B' \defeq \{ t \in B \,|\, \text{$t = 0$ or $K$ is
+  inhabited} \} \subseteq B$ and~$g$ is defined as follows:
+  Let~$s \in i[A] + B'$. Then~$s = i(a) + t$ for some~$a \in A$ and~$t \in B'$.
+  Since~$t \in B'$, $t = 0$ or~$K$ is inhabited. If~$t = 0$,
+  we set~$g(s) \defeq f(a)$. If~$K$ is inhabited, we set~$g(s) \defeq f(a) +
+  \bar{f}(s)$, where~$\bar{f}$ is any element of~$K$.
+
+  Since~$M$ is injective, there exists a dotted map~$\bar g \in E$.
+  We have~$K \subseteq \{ \bar g \}$.
 \end{frame}
@@
   from the point of view of the effective topos.
+
+  \textbf{Proof (sketch).}
+  We have a procedure which computes for any subterminal~$K \subseteq X$ an
+  element~$x_K$ such that~$K \subseteq \{ x_K \}$. Let~$f : X \to X$ be a map.
+  Construct~$K \defeq \{ f(x_K) \}$. Then~$K \subseteq \{ x_K \}$, so~$f(x_K) =
+  x_K$.
   \medskip
@@
   module in the effective topos can be embedded into a flabby module.
+
+  \textbf{Proof.} We have~$M \hookrightarrow \Delta\Gamma M$.
   \medskip

### 2018-10-05 slides-pssl104.tex
Final changes

@@
 
-  \textbf{Definition.}\visible<2->{\hil{$^\star$}}
-  The set~$M$ is \hil{flabby} iff the converse holds.
+  \mbox{\textbf{Definition.}\visible<2->{\hil{$^\star$}}
+  The set~$M$ is \hil{flabby} iff any subterminal is a subsingleton.}
 
@@
     \textbf{Proof.} We have~$M \hookrightarrow P(M)$, and~$P(M)$ is flabby:
-    Let~$K \subseteq P(M)$ be a subterminal. Then~$K \subseteq \{ \bigcup K \}$.
-    % for if~$A \in K$, then~$K = \{ A \}$ and hence~$\bigcup K = K$.
+    Let~$K \subseteq P(M)$ be a subterminal. Then~$K \subseteq \{ \bigcup K \}$,
+    for if~$A \in K$, then~$K = \{ A \}$ and hence~$A \in \{ \bigcup K \} = \{ A \}$.
   }
@@
   \begin{align*}
-    H^0(\smallsphere,\ZZ) &\cong \ZZ &
-    H^1(\smallsphere,\ZZ) &\cong 0   &
-    H^2(\smallsphere,\ZZ) &\cong \ZZ \\
-    H^0(\smalltorus,\ZZ) & \cong \ZZ &
-    H^2(\smalltorus,\ZZ) & \cong \ZZ \oplus \ZZ &
-    H^2(\smalltorus,\ZZ) & \cong \ZZ
+    H^0_{\text{sing}}(\smallsphere,\ZZ) &\cong \ZZ &
+    H^1_{\text{sing}}(\smallsphere,\ZZ) &\cong 0   &
+    H^2_{\text{sing}}(\smallsphere,\ZZ) &\cong \ZZ \\
+    H^0_{\text{sing}}(\smalltorus,\ZZ) & \cong \ZZ &
+    H^1_{\text{sing}}(\smalltorus,\ZZ) & \cong \ZZ \oplus \ZZ &
+    H^2_{\text{sing}}(\smalltorus,\ZZ) & \cong \ZZ
   \end{align*}
@@
     \item Let~$A$ be an abelian group. Let~$X$ be semi-locally contractible.
-          Then $H^n(X, \underline{A}) = H_{\text{sing}}^n(X, A)$.
+          Then $H^n(X, \underline{A}) = H_{\text{sing}}^n(X, A)$ [XXX].
           \\[0.6em]
@@
       \mbox{synthetic algebraic geometry} \\
-      \mbox{synthetic probability theory} \\
+      % \mbox{synthetic probability theory} \\
       \mbox{synthetic domain theory} \\
@@
 \end{frame}
+
+% DURING THE TALK: Zitat vorlesen
 
@@
       sheaf over~$Y$ & sheaf over~$I(Y)$ \\
-      higher direct image $R^n f_* E$ & \bad{??} \\
+      higher direct image $R^n f_* E$ & \bad{??} sheaf cohomology $H^n(I(Y), E)$ \\
       \midrule
       \pause
+      \begin{minipage}{5.1cm}
+        Every finite type sheaf of modules is finite locally free \emph{on a dense open}.
+      \end{minipage} &
+      \begin{minipage}{4.4cm}
+        Every finitely generated vector space is \emph{not not} finite free.
+      \end{minipage} \\
+      \midrule
       \begin{minipage}{5.1cm}
@@
         The intermediate value theorem holds.
-      \end{minipage} \\
-      \midrule
-      \begin{minipage}{5.1cm}
-        Every finite type sheaf of modules is finite locally free \emph{on a dense open}.
-      \end{minipage} &
-      \begin{minipage}{4.4cm}
-        Every finitely generated vector space is \emph{not not} finite free.
       \end{minipage} \\
@@
 \end{frame}
+
+% DURING THE TALK: Mention Set as a special case
+% DURING THE TALK: Mention Mike and Awodey, Butz, Simpson, Streicher
 
@@
   Since~$M$ is injective, there exists a dotted map~$\bar g \in E$.
-  We have~$K \subseteq \{ \bar g \}$.
+  We have~$K \subseteq \{ \bar g \}$. \qed
 \end{frame}
 
-\section[In the effective topos]{Flabbiness in the effective topos}
+\section[In the ef{}fective topos]{Flabbiness in the ef{}fective topos}
 
-\begin{frame}{Flabbiness in the effective topos}
+\begin{frame}{Flabbiness in the ef{}fective topos}
+  \small
   \begin{block}{}
@@
     \in K\_ x = y$}~$K \subseteq M$ is a \pointthisbelow{<1>}{subsingleton}{$\exists
-    a \in M\_ \forall x \in K\_ x = a$}.
+    a \in M\_ K \subseteq \{a\}$}.
   \end{block}
@@
   \textbf{Proposition.}
-  Let~$X$ be an effective object in the effective topos. Then
+  Let~$X$ be an ef{}fective object in the ef{}fective topos. Then
   \[ \text{``If~$X$ is flabby, any endomap on~$X$ has a fixed point.''} \]
-  from the point of view of the effective topos.
+  from the point of view of the ef{}fective topos.
 
@@
   We have a procedure which computes for any subterminal~$K \subseteq X$ an
-  element~$x_K$ such that~$K \subseteq \{ x_K \}$. Let~$f : X \to X$ be a map.
-  Construct~$K \defeq \{ f(x_K) \}$. Then~$K \subseteq \{ x_K \}$, so~$f(x_K) =
-  x_K$.
+  element~$a_K$ such that~$K \subseteq \{ a_K \}$. Let~$f : X \to X$ be a map.
+  Construct~$K \defeq \{ f(a_K) \}$. Then~$K \subseteq \{ a_K \}$, so~$f(a_K) =
+  a_K$. \qed
   \medskip
+
+  \textbf{Corollary.}
+  The only ef{}fective flabby module~$M$ is the zero module.
+
+  \textbf{Proof.} Let~$x \in M$.
+  Then~$x + a = a$ for some~$a \in M$; hence~$x = 0$. \qed
 
@@
   Assuming the law of excluded middle, any~$\neg\neg$-separated
-  module in the effective topos can be embedded into a flabby module.
+  module in the ef{}fective topos can be embedded into a flabby module.
 
-  \textbf{Proof.} We have~$M \hookrightarrow \Delta\Gamma M$.
+  \textbf{Proof.} We have~$M \hookrightarrow \Delta\Gamma M$. \qed
   \medskip
 
-  \textbf{Question.} Are there enough flabby modules in the effective
-  topos?
+  \textbf{Question.} Are there enough flabby modules in the ef{}fective topos?
 \end{frame}
@@
 % XXX: ctcl.pdf, realizability book
-% XXX: Harting
+% DURING THE TALK: Harting
 

### 2018-10-11 slides-como2018.tex
fight against bitrot
Notes slides for PSSL104 almost finished

@@
 \setbeamertemplate{note page}[plain]
-\newcommand{\jnote}[2]{\only<#1>{\note{\justifying#2\par}}}
+\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying#2\par}}}
 
@@
   \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex]{}%
-    \insertnavigation{\paperwidth}%
+    \insertsectionnavigationhorizontal{\paperwidth}{}{}%
   \end{beamercolorbox}%
@@
   To determine whether a statement~$\varphi$ holds in the internal universe of a
-  given topos, one can use the Kripke--Joyal semantics to translate it into an
+  given topos, we can use the Kripke--Joyal semantics to translate it into an
   ordinary external statement and then check the validity of the external
@@
   From the internal point of view of~$\Sh(S)$, the sheaf~$\O_S$ of rings is
-  just an ordinary ring, and one can construct internally to~$\Sh(S)$ the big
+  just an ordinary ring, and we can construct internally to~$\Sh(S)$ the big
   Zariski topos of~$\O_S$. Externally, this construction will yield a certain
@@
   will \emph{not} coincide with the true big Zariski topos of~$S$. To
-  construct the true big Zariski topos, one has to build, internally
+  construct the true big Zariski topos, we have to build, internally
   to~$\Sh(S)$, the classifying topos of local \emph{and local-over-$\O_S$}

### 2018-10-11 slides-pssl104.tex
typo
Notes slides for PSSL104 almost finished

@@
 \usepackage{etex}
+
+\usepackage{pgfpages}
+\setbeameroption{show notes on second screen}
+\setbeamertemplate{note page}[plain]
+\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\footnotesize\justifying#2\par}}}
+
+% Workaround for the issue described at
+% https://tex.stackexchange.com/questions/164406/beamer-using-href-in-notes.
+\newcommand{\fixedhref}[2]{\makebox[0pt][l]{\hspace*{\paperwidth}\href{#1}{#2}}\href{#1}{#2}}
 
@@
 \hypersetup{colorlinks=true}
-\usepackage{multimedia}
-\newcommand{\video}[2]{\movie[width=#2,height=#2,autostart,loop,poster]{}{#1}}
-\hypersetup{colorlinks=false}
 
@@
   \mbox{\textbf{Definition.}\visible<2->{\hil{$^\star$}}
-  The set~$M$ is \hil{flabby} iff any subterminal is a subsingleton.}
+  A set~$M$ is \hil{flabby} iff any subterminal is a subsingleton.}
 
@@
   }
+
+  \jnote{1-}{Any flabby set is inhabited, for there is always the empty
+  subterminal.
+
+  Conversely, given a set~$M$ inhabited by some element~$x_0 \in M$, it
+  might appear that we have an easy proof that~$M$ is flabby: Any
+  subterminal~$K \subseteq M$ is empty or of the form~$K = \{ x \}$ for some~$x
+  \in M$. In the first case, $K$ is a subsingleton for~$K \subseteq \{x_0\}$,
+  and in the second case,~$K$ is a subsingleton for~$K \subseteq \{x\}$.
+
+  Hence it might appear that the definition doesn't make much sense, since it
+  might appear that flabbiness is equivalent to being inhabited.}
+
+  \jnote{2-}{However, one always has to look out for the fine print.}
+
+  \jnote{3-}{In constructive mathematics, the condition for a set to be flabby
+  is nontrivial. We'll later see that the condition is also interesting for
+  classical mathematics, if interpreted internally to suitable toposes.}
+
+  \jnote{4-}{Even though constructively we can't show that any inhabited set is
+  flabby, we can still verify that there are \emph{enough flabby sets}.}
+
+  \jnote{5-}{A module is \emph{flabby} if and only if its underlying set is.
+  We'll see what the significance of this open question is later on.}
 \end{frame}
@@
   \end{block}{}
+
+  \jnote{1-}{Associated to any topological space~$X$ and any abelian group~$A$
+  are the groups~$H^n_{\text{sing}}(X, A)$, the singular cohomology groups
+  of~$X$ with coefficients in~$A$. They depend functorially on~$X$; hence one
+  of many of their applications is to verify that given spaces are not
+  homeomorphic.
+
+  Given a space~$X$, we can hope that we can write~$X$ as the
+  codomain of a continuous map~$f : X \to B$, in such a way that the base space~$B$
+  and the fibers of~$f$ are in some sense easy to understand. In particular, we
+  could ask whether the cohomology of~$X$ can be computed from the cohomology
+  of~$B$ and the cohomology of the fibers.
+
+  The answer, given by Jean Leray in the 1940s, is: Yes, we can, but the
+  framework of singular cohomology is too rigid. For a positive answer we have
+  to generalize to sheaf cohomology. And thus, the notion of sheaves was born.}
 \end{frame}}
 
-%{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[width=\paperwidth]{leray-shaded}\end{minipage}}
 \begin{frame}{Sheaf cohomology}
@@
     \item Let~$A$ be an abelian group. Let~$X$ be semi-locally contractible.
-          Then $H^n(X, \underline{A}) = H_{\text{sing}}^n(X, A)$ [XXX].
+          Then $H^n(X, \underline{A}) = H_{\text{sing}}^n(X, A)$
+          \href{https://arxiv.org/abs/1602.06674}{[Sella 2016]}.
           \\[0.6em]
@@
   \end{itemize}
+
+  \jnote{1-}{An injective resolution is a sequence of sheaves and modules and
+  linear morphisms as indicated such that the sequence is \emph{exact}
+  (the kernel of any outgoing morphism equals the image of the respective incoming
+  morphism) and such that the sheaves~$I^n$ are injective (a notion recalled
+  below).
+
+  The fundamental fact of homological algebra is: Even though the sequence~$0
+  \to I^0 \to I^1 \to \cdots$ can only fail to be exact at the front, the
+  sequence~$0 \to \Gamma I^0 \to \Gamma I^1 \to \cdots$ of global sections can
+  fail to be exact at any place. Sheaf cohomology measures the extent of this
+  failure.
+
+  The standard proof of the existence of injective resolutions requires Zorn's
+  lemma and the law of excluded middle. Injective resolutions are not unique,
+  but the resulting sheaf cohomology modules are unique up to isomorphism.
+  A primer on these matters is located
+  \fixedhref{https://rawgit.com/iblech/talk-homological-algebra/master/notes-derived-functors.pdf}{here}.
+
+  The positive answer to the question posed on the previous slide is given by
+  the spectral sequence displayed at the bottom of this slide. The sheaves~$R^n
+  f_*(E)$ are called the \emph{higher direct images of~$E$} (along~$f$). Even
+  if~$E$ is a constant sheaf, its higher direct images might not be. This is
+  the reason why singular cohomology is too restrictive.
+
+  The higher direct images~$R^n f_*(E)$ are defined exactly as the sheaf
+  cohomology~$H^n(X, E)$, only with the global sections functor~$\Gamma$
+  replaced by the pushforward functor~$f_*$. They are dubbed ``relative
+  cohomology'', for instance because under some conditions, there are
+  isomorphisms~$(R^n f_*(E))_b \cong H^n(X_b, E|_{X_b})$ where~$X_b$ is the
+  fiber of~$b$ under~$f$. This talk presents a rigorous and general way to
+  regard higher direct images as sheaf cohomology.}
 \end{frame}
-%}
 
@@
   \end{columns}
+
+  \jnote{1-}{Constructive mathematics can be studied for philosophical reasons
+  or out of general mathematical curiosity. But restricting to constructive
+  reasoning in our proofs also yields concrete gains for
+  classical mathematics.
+
+  One of these is \emph{program extraction}: From any constructive proof, we can
+  mechanically extract a program witnessing the proven statement. A basic
+  example is that any constructive proof of the infinitude of primes yields an
+  algorithm for computing primes (together with a termination and correctness
+  proof).
+
+  Another is that, since constructive mathematics is consistent with a number
+  of anti-classical dream axioms, constructive mathematics allows to develop
+  \emph{synthetic accounts} of several subjects. For instance, in synthetic
+  algebraic geometry, a scheme is just a set, a morphism of schemes is just
+  a map of sets, and any map of the ground field into itself is polynomial.
+
+  There are also \emph{reduction techniques} which propose interesting deals. For
+  instance, there is a technique which allows us to pretend that reduced rings
+  are Noetherian and in fact fields -- if in return we switch from classical
+  reasoning to constructive. This particular technique has been used to turn
+  the slightly convoluted multi-page proof of Grothendieck's generic
+  freeness lemma into a simple one-paragraph argument.
+
+  An informative and entertaining primer on constructive mathematics can be
+  found in the linked talk recording by Andrej Bauer or his
+  \fixedhref{http://www.ams.org/journals/bull/2017-54-03/S0273-0979-2016-01556-4/S0273-0979-2016-01556-4.pdf}{written
+  notes} on the subject. Don't worry, the standard proof that~$\sqrt{2}$ is not rational is
+  perfectly fine in constructive mathematics.
+  }
 \end{frame}
@@
 % DURING THE TALK: Zitat vorlesen
+% Thus, contrary to how it may appear on the surface, doing mathematics
+% ``constructively'' does not usually involve giving up important theorems, but
+% rather finding the best way to state the definitions so as to make the
+% important theorems constructively provable. That is, we may freely use the [law of
+% excluded middle] when first investigating a subject, but once that subject is
+% better understood, we can hope to refine its definitions and proofs so as to
+% avoid that axiom.
 
@@
   \end{center}
+
+  \jnote{2-}{The internal language of a topos~$\E$ is a device which defines
+  for any formula~$\varphi$ of a certain language (a form of extensional type
+  theory) what it means for~$\varphi$ to hold \emph{internally to~$\E$},
+  written~``$\E \models \varphi$''. This translation process is sound with
+  respect to intuitionistic logic; hence any theorem of constructive
+  mathematics is valid in any topos. Only few toposes validate classical logic
+  (for instance~$\Sh(X)$ does if~$X$ is a discrete space and the law of
+  excluded middle is available in the metatheory).
+
+  As a special case, the internal language of the topos~$\Set$ is just the
+  usual mathematical language; more formally, $\Set \models \varphi$ if and
+  only if~$\varphi$.}
+
+  % XXX: Cite Shulman and ABSS. Also Streicher's lecture notes.
+
+  \jnote{3-}{The intermediate value theorem (``any continuous function with
+  opposite signs has a zero'') doesn't admit a constructive proof, because for
+  most spaces~$X$ the external translation~$\Sh(X) \models \text{IVT}$ is not
+  true -- it's not true that in continuous families of continuous functions
+  with opposite signs, zeros can locally be picked continuously, as
+  \fixedhref{https://raw.githubusercontent.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{this
+  video shows}.
+
+  Over reduced schemes, every finite type sheaf of modules is finite locally free on a
+  dense open. This statement (``important hard exercise'' 13.7.K in
+  \fixedhref{http://math.stanford.edu/~vakil/216blog/FOAGnov1817public.pdf}{[Vakil
+  2017]}) is just the external translation of the easy-to-prove internal
+  statement that every finitely generated vector space does \emph{not not}
+  admit a finite basis. (A scheme is reduced if and only if its structure sheaf
+  looks like a field from the internal point of view (in the sense that $1 \neq
+  0$ and~$\neg(\text{$x$ invertible}) \Rightarrow x = 0$). This is why the
+  reducedness condition is important.)}
 \end{frame}
@@
     \small
-    \item It's consistent with \textbf{ZF} that there are no injective modules [Blass 1979].
+    \item It's consistent with \textbf{ZF} that there are no injective modules
+    \href{http://www.ams.org/journals/tran/1979-255-00/S0002-9947-1979-0542870-6/S0002-9947-1979-0542870-6.pdf}{[Blass 1979]}.
     \item The existence of enough injective modules is \hil{constructively neutral}.
@@
   \end{itemize}
+
+  \jnote{2-3}{Somewhat surprisingly, even though the standard proof that there
+  are enough injective modules requires the axiom of choice and even though
+  it's consistent with Zermelo--Fraenkel set theory that the zero module is the
+  only injective~$\ZZ$-module, the existence of enough injective modules is
+  \emph{constructively neutral}, that is, does not imply a fundamentally
+  nonconstructive principle like the law of excluded middle.
+
+  Indeed, assuming the axiom of choice in the metatheory, the statement~``any
+  module can be embedded into an injective module'' holds in the internal
+  language of any Grothendieck topos. This is because, assuming the axiom of
+  choice in the metatheory, any sheaf of modules over a site can be embedded
+  into an injective sheaf of modules and, somewhat
+  surprisingly, \ldots}
+
+  \jnote{3}{\ldots{} a sheaf of modules is injective if and only if it is an
+  injective module from the internal point of view. The~``$\Rightarrow$''
+  direction is straightforward; the~``$\Leftarrow$'' direction is nontrivial:
+  The external meaning of the internal existential quantifier is \emph{local}
+  existence. Hence linear morphisms into a sheaf of modules which is injective from the internal
+  point of view can \emph{locally} be extended. But
+  these extensions need not be compatible, hence might not glue to a global
+  extension. For the case of sheaves of abelian groups, this result is due to
+  Roswitha Harting in an
+  \fixedhref{https://www.tandfonline.com/doi/pdf/10.1080/00927878308822853}{1983
+  paper of her}. The case of sheaves of modules is arguably also due to her,
+  even though she states that the result doesn't hold for sheaves of modules.
+  (Technology has improved since then, and using flabbiness as an organizing
+  principle one can give a reasonably straightforward proof of the general
+  statement.)
+
+  In contrast, internally and externally projective modules do not coincide at
+  all.}
+
+  \jnote{4}{A consequence of the fact that internal and external injectivity
+  coincides for sheaves of modules over locales is that we can interpret the
+  higher direct images~$R^n f_*(Y, E)$ of a sheaf~$E$ of modules over an
+  over-locale~$f : Y \to X$ as sheaf cohomology~$H^n(I(Y), E)$, where~$I(Y)$ is
+  the internal locale of~$\Sh(X)$ corresponding to~$Y$.
+
+  A basic application is the following. Any student in algebraic geometry
+  needs, at some point in her life, compute the cohomology of projective space.
+  At a later point she needs to compute higher direct images
+  along~$\mathbb{P}^n_S \to S$, where~$\mathbb{P}^n_S$ is a relative version of
+  projective space. Since higher direct images are just internal
+  sheaf cohomology, she can in fact skip the second computation.}
 \end{frame}
+
+% DURING THE TALK: Harting
 
@@
   \end{itemize}
+
+  \jnote{1-}{Since we cannot show the existence of enough injective sheaves of
+  modules (or even just plain modules) constructively, the definition of sheaf
+  cohomology using injective resolutions doesn't work in a constructive
+  setting.}
 \end{frame}
@@
   \textbf{Proof.} (Only ``$\Leftarrow$''.) Let~$i : A \to B$ be a linear
-  monomorphism. Let~$f : A \to M$ be a linear morphism. Then verify, internally, that
+  monomorphism. Let~$f : A \to M$ be a linear morphism. We verify, internally, that
   the set~$E \defeq \{ \bar f : B \to M \,|\, \bar f \circ i = f \}$ is flabby.
@@
 % XXX: ctcl.pdf, realizability book
-% DURING THE TALK: Harting
 

### 2018-10-13 paper-flabby-objects.tex
gen pdf
Note slides almost finished

@@
 \author{Ingo Blechschmidt}
-\address{Max Planck Institute for Mathematics in the Sciences \\
-Inselstraße 22 \\
-04103 Leipzig, Germany}
-\email{ingo.blechschmidt@mis.mpg.de}
+\address{Università di Verona \\
+Department of Computer Science \\
+Strada le Grazie 15 \\
+37134 Verona, Italy}
+\email{iblech@speicherleck.de}
 
@@
 
+{\centering \textbf{Rough draft, comments welcome! iblech@speicherleck.de}\par}
+\bigskip
+
 \noindent
 
+XXX: introduction
 % XXX: In Einleitung interne Sprache erinnern.
@@
 The notion of flabby objects originates from the notion of flabby shaves and is
-therefore closely connected to Grothendieck toposes. It is therefore instructive
+therefore closely connected to Grothendieck toposes. Hence it is instructive
 to study flabby objects in elementary toposes which are not Grothendieck
@@
 
+\printbibliography
+
 \end{document}

### 2018-10-13 slides-pssl104.tex
Note slides almost finished

@@
   \jnote{3-}{In constructive mathematics, the condition for a set to be flabby
-  is nontrivial. We'll later see that the condition is also interesting for
+  is nontrivial. We'll see later that the condition is also interesting for
   classical mathematics, if interpreted internally to suitable toposes.}
@@
   codomain of a continuous map~$f : X \to B$, in such a way that the base space~$B$
-  and the fibers of~$f$ are in some sense easy to understand. In particular, we
+  and the fibers of~$f$ are in some sense easy to understand. In such a situation we
   could ask whether the cohomology of~$X$ can be computed from the cohomology
@@
   The answer, given by Jean Leray in the 1940s, is: Yes, we can, but the
-  framework of singular cohomology is too rigid. For a positive answer we have
+  framework of singular cohomology is too rigid for this task. For a positive answer we have
   to generalize to sheaf cohomology. And thus, the notion of sheaves was born.}
@@
 
-  \jnote{1-}{An injective resolution is a sequence of sheaves and modules and
+  \jnote{1-}{An injective resolution is a sequence of sheaves of modules and
   linear morphisms as indicated such that the sequence is \emph{exact}
@@
   There are also \emph{reduction techniques} which propose interesting deals. For
-  instance, there is a technique which allows us to pretend that reduced rings
-  are Noetherian and in fact fields -- if in return we switch from classical
+  instance, there is a technique which allows us to pretend that a reduced ring
+  is Noetherian and in fact a field -- if in return we switch from classical
   reasoning to constructive. This particular technique has been used to turn
@@
   A basic application is the following. Any student in algebraic geometry
-  needs, at some point in her life, compute the cohomology of projective space.
+  needs, at some point in her life, to compute the cohomology of projective space.
   At a later point she needs to compute higher direct images
@@
   cohomology using injective resolutions doesn't work in a constructive
-  setting.}
+  setting. Classically it's known that \emph{flabby resolutions} can also be
+  used to compute sheaf cohomology. There are more flabby sheaves than injective
+  ones, they have better stability properties (flabby sheaves are preserved
+  under pushforward) and the axiom of choice is not needed to construct flabby
+  resolutions (the standard proof uses only the law of excluded middle, and not
+  Zorn's lemma). Hence it seems reasonable to base a constructively sensible
+  definition of sheaf cohomology on flabby resolutions. We tried to do so, and
+  failed.
+
+  Assuming Zorn's lemma, the notion of a flabby sheaf is a local notion,
+  meaning that a sheaf is flabby if and only if its restrictions to every
+  member of an open covering are, but this fact is not obvious from the
+  definition. In contrast, the notion that a sheaf~$E$ is flabby from the
+  internal point of view is local without any assumptions (as is any
+  internal notion), hence maybe we should consider adopting internal flabbiness as the
+  official definition of flabbiness. Its external translation is:
+
+  A sheaf~$E$ is flabby from the internal point of view if and only if for any
+  local section~$s \in E(U)$, there is an open covering~$X = \bigcup_i U_i$
+  such that for all~$i$, the section~$s$ extends to a section on~$U \cup U_i$.}
 \end{frame}
@@
 \begin{frame}{Flabbiness as an organizing principle}
-%  \begin{block}{}
-%    A set~$M$ is \hil{flabby} iff any \pointthisbelow{<1>}{subterminal}{$\forall x,y
-%    \in K\_ x = y$}~$K \subseteq M$ is a \pointthisbelow{<1>}{subsingleton}{$\exists
-%    a \in M\_ \forall x \in K\_ x = a$}.
-%  \end{block}
-%  \bigskip
-%  \bigskip
-%  \justifying
-
-%  \textbf{Proposition.} Let~$E$ be a sheaf over a locale~$X$. Then~$E$ is flabby
-%  iff it is a flabby set from the point of view of~$\Sh(X)$.
-%
-%  \textbf{Proof.} Routine/Zorn's lemma.
-
   \justifying\small
@@
   We have~$K \subseteq \{ \bar g \}$. \qed
+
+  \jnote{1-}{The notion of being flabby from the internal point of view turns
+  out to have valuable organizing power. For instance, both of the
+  following statements can be proven by first verifying that a certain sheaf is
+  internally flabby (which can be done entirely constructively) and then
+  appealing to Zorn's lemma in order to obtain a global section of that sheaf.
+
+  \begin{itemize}\justifying
+    \item Flabby sheaves are acyclic for the global sections functor: Let~$0
+    \to E \to F \to G \to 0$ be a short exact sequence of sheaves of modules
+    and let~$E$ be flabby. Then the sequence remains exact after taking global
+    sections.
+
+    (Verify that the sheaf of local preimages of a given section~$s \in G(X)$ is
+    flabby.)
+
+    \item Internally injective modules are externally injective.
+
+    (See proof on the slide.)
+  \end{itemize}}
 \end{frame}
@@
   \textbf{Question.} Are there enough flabby modules in the ef{}fective topos?
+
+  \jnote{1-}{The notion of flabby sets was conceived to model the notion of
+  flabby sheaves and is therefore closely connected to Grothendieck toposes.
+  Hence it is instructive to study flabby objects in elementary toposes which
+  are not Grothendieck toposes, away from their original conceptual home.
+
+  In particular, we hope to prove the conjecture that the statement ``any
+  module embeds into a flabby module'' is not constructively provable by
+  verifying that it doesn't hold in the effective topos.
+
+  To this end, the slide displays two results.
+
+  Details on the self-referential construction~``$K \defeq \{ f(a_K) \}$'' are in
+  \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-flabby-objects.pdf}{this
+  draft paper}.
+
+  A module is~$\neg\neg$-separated if and only if~$\neg\neg(x = 0)$ implies~$x = 0$.}
 \end{frame}
@@
   \end{itemize}
+
+  \jnote{1-}{Even if we could constructively prove that there are enough flabby
+  modules, there is still the problem that the proof that flabby
+  sheaves of modules are acyclic for the global sections functor (appears to)
+  require Zorn's lemma.}
 \end{frame}

### 2018-10-15 slides-pssl104.tex
Notes for every slide finished, now fine-tuning

@@
   sheaves of modules are acyclic for the global sections functor (appears to)
-  require Zorn's lemma.}
+  require Zorn's lemma.
+
+  Hence it appears that the simple idea of basing sheaf cohomology on flabby
+  resolutions doesn't work, constructively. More work is needed to give a
+  constructive account of sheaf cohomology and, more generally, derived functors.
+
+  I hope that some day, we can study the cohomology of the the smallest dense
+  sublocale of the one-point space.$^\star$
+
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+
+  $^\star$ Assuming the law of excluded middle, this locale is just
+  the one-point space again. Hence cohomology of this locale should measure the
+  extent to which we're nonclassical, being zero if and only if the law of
+  excluded middle holds.
+
+  An alternative way to phrase this question is as follows.
+  Let~$\Set_{\neg\neg}$ be the smallest dense subtopos of~$\Set$, the topos of
+  double-negation sheaves. The forgetful functor~$\mathrm{Ab}(\Set_{\neg\neg})
+  \to \mathrm{Ab}$ is left-exact, but might not be right-exact, since
+  a map~$f : A \to B$ is an epimorphism
+  in~$\mathrm{Ab}(\Set_{\neg\neg})$ if and only if~$\forall y \in B\_
+  \neg\neg(\exists x \in A\_ f(x) = y)$, which is weaker than being surjective.
+  What do its right derived functors look like?\par}
 \end{frame}

### 2018-10-15 slides-pssl104.tex
typo

@@
   flabby, we can still verify that there are \emph{enough flabby sets}.}
-
-  \jnote{5-}{A module is \emph{flabby} if and only if its underlying set is.
+  \jnote{5-}{However it's unknown whether there are \emph{enough flabby
+  modules}. (A module is \emph{flabby} if and only if its underlying set is.)
   We'll see what the significance of this open question is later on.}

### 2018-10-15 slides-cambridge2016.tex
use https where possible

@@
   Photo source:
-  \href{http://www.stpaulscambridge.org.uk/classes-and-room-hire/activities-and-events/newtown-news/}{Newtown
+  \href{https://www.stpaulscambridge.org.uk/classes-and-room-hire/activities-and-events/newtown-news/}{Newtown
   News}
@@
   answers on MathOverflow:
-  \begin{center}\url{http://mathoverflow.net/a/14334/31233}\end{center}
+  \begin{center}\url{https://mathoverflow.net/a/14334/31233}\end{center}
   \par
@@
     Lane's book, or Chapter~13 of
-    \href{http://www.mathematik.tu-darmstadt.de/~streicher/CTCL.pdf}{these
+    \href{https://www.mathematik.tu-darmstadt.de/~streicher/CTCL.pdf}{these
     lecture notes by Thomas Streicher}.\end{justify}
@@
   is the ring's spectrum. See Vicker's
-  \href{http://www.cs.bham.ac.uk/~sjv/LocTopSpaces.pdf}{Locales and Toposes as Spaces} and
-  \href{http://www.cs.bham.ac.uk/~sjv/GeoAspects.pdf}{Continuity and geometric
+  \href{https://www.cs.bham.ac.uk/~sjv/LocTopSpaces.pdf}{Locales and Toposes as Spaces} and
+  \href{https://www.cs.bham.ac.uk/~sjv/GeoAspects.pdf}{Continuity and geometric
   logic} for very accessible introductions to this topic.
@@
   \centering
-  \hil{\href{http://tiny.cc/topos-notes}{http://tiny.cc/topos-notes}}
+  \hil{\href{https://tiny.cc/topos-notes}{http://tiny.cc/topos-notes}}
   \par

### 2018-10-15 slides-duesseldorf2017.tex
typo
Slides finished
use https where possible

@@
     \item or Grothendieck's functor-of-points account, where a scheme is a functor~$\mathrm{Ring} \to \mathrm{Set}$.
-    \small\[ A \longmapsto \{ (x,y,z) \? A^3 \,|\, x^n+y^n-z^n=0 \} \]
+    \small\[ A \longmapsto \{ (x,y,z) \in A^3 \,|\, x^n+y^n-z^n=0 \} \]
   \end{itemize}
@@
   is the ring's spectrum. See Vicker's
-  \href{http://www.cs.bham.ac.uk/~sjv/LocTopSpaces.pdf}{Locales and Toposes as Spaces} and
+  \href{https://www.cs.bham.ac.uk/~sjv/LocTopSpaces.pdf}{Locales and Toposes as Spaces} and
   \href{http://www.cs.bham.ac.uk/~sjv/GeoAspects.pdf}{Continuity and geometric

### 2018-10-15 slides-ihes2015.tex
use https where possible

@@
   \centering
-  \hil{\href{http://tiny.cc/topos-notes}{http://tiny.cc/topos-notes}} \\
+  \hil{\href{https://tiny.cc/topos-notes}{https://tiny.cc/topos-notes}} \\
   \begin{minipage}{1.0\textwidth}\justifying

### 2018-10-15 slides-pssl101.tex
to do
typo
use https where possible

@@
 \end{frame}
+% XXX: Mention Martin Brandenburg!
 

### 2018-10-15 slides-pssl104.tex
use https where possible

@@
   found in the linked talk recording by Andrej Bauer or his
-  \fixedhref{http://www.ams.org/journals/bull/2017-54-03/S0273-0979-2016-01556-4/S0273-0979-2016-01556-4.pdf}{written
+  \fixedhref{https://www.ams.org/journals/bull/2017-54-03/S0273-0979-2016-01556-4/S0273-0979-2016-01556-4.pdf}{written
   notes} on the subject. Don't worry, the standard proof that~$\sqrt{2}$ is not rational is
@@
     \item It's consistent with \textbf{ZF} that there are no injective modules
-    \href{http://www.ams.org/journals/tran/1979-255-00/S0002-9947-1979-0542870-6/S0002-9947-1979-0542870-6.pdf}{[Blass 1979]}.
+    \href{https://www.ams.org/journals/tran/1979-255-00/S0002-9947-1979-0542870-6/S0002-9947-1979-0542870-6.pdf}{[Blass 1979]}.
     \item The existence of enough injective modules is \hil{constructively neutral}.

### 2018-10-16 slides-pssl104.tex
Final version of the PSSL slides with notes

@@
       \mbox{``Every map~$\affl \to \affl$ is polynomial.''}
-      \mbox{``Heyting Arithmetic has exactly one model.''}
-      \mbox{``The subsets of~$\{\heartsuit\}$ form a proper class.''}
-      \mbox{``There is an injection~$\RR \to \NN$.''} \\
+      \href{https://www.staff.science.uu.nl/~ooste110/realizability/arithcatsubmit.pdf}{\mbox{``Heyting Arithmetic has exactly one model.''}}
+      \href{http://www1.maths.leeds.ac.uk/~pmtng/Slides/gambino-lisbon-jan2018.pdf}{\mbox{``The subsets of~$\{\heartsuit\}$ form a proper class.''}}
+      \href{https://rawgit.com/iblech/mathezirkel-kurs/master/superturingmaschinen/slides-warwick2017.pdf}{\mbox{``There is an injection~$\RR \to \NN$.''}} \\
       \rotatebox{90}{\dots}
@@
       \mbox{program extraction} \\
-      \mbox{synthetic differential geometry} \\
-      \mbox{synthetic algebraic geometry} \\
-      % \mbox{synthetic probability theory} \\
-      \mbox{synthetic domain theory} \\
-      \mbox{new reduction techniques in algebra} \\
-      \mbox{Bohr topos for quantum mechanics} \\
+      \href{http://math.andrej.com/2008/08/13/intuitionistic-mathematics-for-physics/}{\mbox{synthetic differential geometry}} \\
+      \href{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{\mbox{synthetic algebraic geometry}} \\
+      \href{https://www.dpmms.cam.ac.uk/~martin/Research/Oldpapers/synthetic91.pdf}{\mbox{synthetic domain theory}} \\
+      \href{https://rawgit.com/iblech/internal-methods/master/slides-bayreuth2018.pdf}{\mbox{new reduction techniques in algebra}} \\
+      \href{https://rawgit.com/iblech/internal-methods/master/}{\mbox{Bohr topos for quantum mechanics}} \\
       \rotatebox{90}{\dots}
@@
 
-  \jnote{2-}{The internal language of a topos~$\E$ is a device which defines
+  \jnote{2-3}{The internal language of a topos~$\E$ is a device which defines
   for any formula~$\varphi$ of a certain language (a form of extensional type
@@
 
-  % XXX: Cite Shulman and ABSS. Also Streicher's lecture notes.
-
-  \jnote{3-}{The intermediate value theorem (``any continuous function with
+  \jnote{3}{The intermediate value theorem (``any continuous function with
   opposite signs has a zero'') doesn't admit a constructive proof, because for
@@
   reducedness condition is important.)}
+
+  \jnote{4-}{Excellent references on the internal language include:
+  \begin{center}
+    \fixedhref{https://www.oliviacaramello.com/Papers/Papers.htm}{\includegraphics[height=3.5cm]{olivia-tst}}
+    \fixedhref{https://store.doverpublications.com/0486450260.html}{\includegraphics[height=3.5cm]{goldblatt-logic}}
+    \fixedhref{https://www.worldcat.org/title/sheaves-in-geometry-and-logic-a-first-introduction-to-topos-theory/oclc/24428855}{\includegraphics[height=3.5cm]{moerdijk-mac-lane-sheaves}}
+    \fixedhref{https://www2.mathematik.tu-darmstadt.de/~streicher/CTCL.pdf}{\includegraphics[height=3.5cm]{streicher-ctcl}}
+  \end{center}
+
+  We use an extension of the original form of the internal language which
+  allows for unbounded quantification, Mike Shulman's
+  \fixedhref{https://arxiv.org/abs/1004.3802}{\emph{stack semantics}}.
+  (Independently, Steve Awodey, Carsten Butz, Alex Simpson and Thomas Streicher
+  developed a
+  \fixedhref{http://homepages.inf.ed.ac.uk/als/Research/Sources/set-models.pdf}{similar
+  semantics}.)}
 \end{frame}
@@
   projective space. Since higher direct images are just internal
-  sheaf cohomology, she can in fact skip the second computation.}
+  sheaf cohomology, she can in fact skip the second computation.
+
+  Further progress along these lines is hindered by the fact that we don't yet
+  have a constructive account of sheaf cohomology.}
 \end{frame}
@@
   module embeds into a flabby module'' is not constructively provable by
-  verifying that it doesn't hold in the effective topos.
+  verifying that it doesn't hold in the ef{}fective topos.
 
@@
   \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-flabby-objects.pdf}{this
-  draft paper}.
+  draft paper}. A module is~$\neg\neg$-separated if and only if~$\neg\neg(x = 0)$ implies~$x = 0$.
+  References on the ef{}fective topos include Martin Hyland's
+  \fixedhref{https://webdpmms.maths.cam.ac.uk/~martin/Research/Oldpapers/hyland-effectivetopos.pdf}{survey
+  paper} and the canonical book by one of our honoraries:
 
-  A module is~$\neg\neg$-separated if and only if~$\neg\neg(x = 0)$ implies~$x = 0$.}
+  {\centering\fixedhref{https://www.staff.science.uu.nl/~ooste110/boekbegin.pdf}{\includegraphics[height=3.5cm]{oosten-realizability}}\par}}
 \end{frame}
-
-% XXX: ctcl.pdf, realizability book
 
@@
 
-  Hence it appears that the simple idea of basing sheaf cohomology on flabby
-  resolutions doesn't work, constructively. More work is needed to give a
-  constructive account of sheaf cohomology and, more generally, derived functors.
+  Hence it appears that the simple idea of basing a constructive account of sheaf cohomology on flabby
+  resolutions doesn't work.
 
-  I hope that some day, we can study the cohomology of the the smallest dense
+  More work is needed.
+  I hope that some day, we can study the cohomology of the smallest dense
   sublocale of the one-point space.$^\star$
 
-  \bigskip
-  \bigskip
-  \bigskip
-  \bigskip
+  {\centering\includegraphics[height=2cm]{phantoms}\par}
 
@@
 
-  An alternative way to phrase this question is as follows.
+  An alternative way of putting this question is as follows.
   Let~$\Set_{\neg\neg}$ be the smallest dense subtopos of~$\Set$, the topos of
-  double-negation sheaves. The forgetful functor~$\mathrm{Ab}(\Set_{\neg\neg})
+  double negation sheaves. The forgetful functor~$\mathrm{Ab}(\Set_{\neg\neg})
   \to \mathrm{Ab}$ is left-exact, but might not be right-exact, since

### 2018-10-16 paper-flabby-objects.tex
typo

@@
 
-Noticeably missing here is XXX ...
+Noticeably missing here is a statement as follows: ``Let~$0 \to M' \to M \to
+M'' \to 0$ be a short exact sequence of modules. If~$M'$ and~$M$ are flabby, so
+is~$M''$.'' Assuming Zorn's lemma in the metatheory, this statement is true in
+every topos of sheaves over a locale, but we don't know whether it has an
+intuitionistic proof and in fact we surmise that it has not.
 
@@
 
-The notion of strongly flabby objects is, however, not stable under base change
+The notion of strongly flabby objects is, however, not local (in the same sense
+that the notion of flabby objects is, as stated in
+Proposition~\ref{prop:basic-properties-of-flabby-objects})
 and therefore cannot be characterized in the internal language. A specific
@@
 not lift, but its pullback to the slice~$BG/G \simeq \Set$ is (assuming
-the law of excluded middle).
+the law of excluded middle in the metatheory), and the unique morphism~$G \to
+1$ is indeed an epimorphism.
 \end{rem}
@@
 We currently believe that it is not possible to give a constructive account of
-a global cohomology functor. However, XXX.
+a global cohomology functor which would associate to any sheaf of modules its
+cohomology. However, it should be possible to do so for a restricted class of
+sheaves, while still preserving the good formal properties expected from
+derived functors.
 

### 2018-10-16 paper-flabby-objects.tex
A connection to QITs (quotient inductive types)
With valuable input from Thorsten Altenkirch.

@@
 
-\item There is a way of embedding any module into a flabby module if one is
-prepared to ignore set-theoretical difficulties. Namely, let~$M$ be an~$R$-module.
-Inductively construct a collection~$T$ of terms by the following clauses: $0
-\in T$ (where~$0$ is a formal symbol);
-if~$t,s \in T$, then~$t + s \in T$; if~$t \in T$ and~$r \in R$, then~$rt \in
-T$; if~$x \in M$, then~$\underline{x} \in T$; if~$K \subseteq T$ is a
-subterminal, then~$\varepsilon_K \in T$. Let~$({\sim})$ be
-the finest equivalence relation on~$T$ such that~$t + (s + u) \sim (t + s) +
-u$, $t + s \sim s + t$, $t + 0 \sim t \sim 0 + t$, $0t \sim 0$, $1t \sim t$, $r(t+s) \sim rt
-+ rs$, $(r+r')t \sim rt + r't$ and such
-that~$\underline{0} \sim 0$, $\underline{x+y} \sim \underline{x} + \underline{y}$, $\underline{rx} \sim
-r \underline{x}$, and such that~$\varepsilon_{\{t\}} \sim
-t$ for all terms~$t$. Then the quotient~$T/{\sim}$ is a flabby~$R$-module
-into which~$M$ embeds. However, it is not clear that~$T/{\sim}$ is a set.
+%\item There is a way of embedding any module into a flabby module if one is
+%prepared to ignore set-theoretical difficulties. Namely, let~$M$ be an~$R$-module.
+%Inductively construct a collection~$T$ of terms by the following clauses: $0
+%\in T$ (where~$0$ is a formal symbol);
+%if~$t,s \in T$, then~$t + s \in T$; if~$t \in T$ and~$r \in R$, then~$rt \in
+%T$; if~$x \in M$, then~$\underline{x} \in T$; if~$K \subseteq T$ is a
+%subterminal, then~$\varepsilon_K \in T$. Let~$({\sim})$ be
+%the finest equivalence relation on~$T$ such that~$t + (s + u) \sim (t + s) +
+%u$, $t + s \sim s + t$, $t + 0 \sim t \sim 0 + t$, $0t \sim 0$, $1t \sim t$, $r(t+s) \sim rt
+%+ rs$, $(r+r')t \sim rt + r't$ and such
+%that~$\underline{0} \sim 0$, $\underline{x+y} \sim \underline{x} + \underline{y}$, $\underline{rx} \sim
+%r \underline{x}$, and such that~$\varepsilon_{\{t\}} \sim
+%t$ for all terms~$t$. Then the quotient~$T/{\sim}$ is a flabby~$R$-module
+%into which~$M$ embeds. However, it is not clear that~$T/{\sim}$ is a set.
+
+\item There is a way of embedding any module into a flabby module if
+\emph{quotient inductive types}, as suggested by Altenkirch and
+Kaposi~\cite{altenkirch-kaposi:qits}, are available. These generalize ordinary
+inductive~$W$-types, which exist in any
+topos~\cite{moerdijk-palmgren:wellfounded-trees,berg-moerdijk:w-types-in-sheaves,berg-kouwenhoven-gentil:w-types-in-eff}
+and whose existence
+can indeed be verified in an intuitionistic set theory like IZF, by allowing to
+give constructors and state identifications at the same time. More
+specifically, given an~$R$-module~$M$, we can construct a flabby envelope~$T$
+of~$M$ as the quotient inductive type generated by the following clauses:~$0
+\in T$ (where~$0$ is a formal symbol); if~$t,s \in T$, then~$t + s \in T$;
+if~$t \in T$ and~$r \in R$, then~$rt \in T$; if~$x \in M$, then~$\underline{x}
+\in T$; if~$K \subseteq T$ is a subterminal, then~$\varepsilon_K \in T$;
+if~$t,s,u \in T$ and~$r,r' \in R$, then~$t + (s + u) = (t + s) + u$, $t + s = s + t$, $t + 0 = t
+= t + 0$, $0t = 0$, $1t = t$, $r(t+u) = rt + ru$, $(r+r')t = rt + r't$; if~$x,y
+\in M$ and~$r \in R$, then~$\underline{0} = 0$, $\underline{x + y} =
+\underline{x} + \underline{y}$, $\underline{rx} = r \underline{x}$; and if~$t
+\in M$, then~$\varepsilon_{\{t\}} = t$.
+
+However, it is an open question under which circumstances quotient inductive
+types can be shown to exist. Zermelo--Fraenkel with choice certainly suffices,
+while Zermelo--Fraenkel without choice doesn't~\cite[Section~9]{shulman-lumsdaine:hits},\footnote{With
+quotient inductive types, any infinitary algebraic theory admits free algebras.
+However, it is consistent with Zermelo--Fraenkel set theory that some such
+theories don't admit free algebras~\cite{blass:free-algebras}.} hence IZF also
+doesn't. Quotient inductive types seem to be, as the existence of enough
+injective modules, \emph{constructively neutral}.
+
+\item There is a number of simple constructions which come close to providing
+flabby envelopes for arbitrary modules~$M$. For instance, we could equip~$T
+\defeq \P_{\leq1}(X)/{\sim}$, where~$K \sim L$ if and only if~$K = L$ or~$K
+\cup L \subseteq \{0\}$, with a module structure given by~$0 \defeq [\{0\}]$,
+$[K]+[L] \defeq [K+L]$ and~$r [K] \defeq rK$. The resulting module admits a
+linear injection from~$M$, sending an element~$x$ to~$[\{x\}]$. However, it
+fails to be flabby. Given a subterminal~$E \subseteq \P_{\leq1}(X)/{\sim}$,
+there is the well-defined element~$v \defeq [\{x \in M\,|\,\text{$x \in K$ for
+some~$[K] \in E$}\}]$, but we cannot verify~$E \subseteq \{v\}$.
 

### 2018-10-18 paper-flabby-objects.tex
Beginning of section on internalizing higher direct images

@@
 
-\section{Characterizing cohomology}
+\section{Higher direct images as internal sheaf cohomology}
 
-XXX
-% EXT, TOR
-% Higher direct images
+Let~$X$ be a locale and let~$f : Y \to X$ be an over-locale. By the fundamental
+relation between locales and topological spaces, this situation arises
+for instance, when given a sober topological space and a topological
+space over it, as is often the case in algebraic topology or algebraic
+geometry. Let a sheaf~$\O_Y$ of rings on~$Y$ be given. Then the
+traditional way to define the \emph{higher direct images} of a sheaf~$E$
+of~$\O_Y$-modules is to pick an injective resolution~$0 \to E \to I^\bullet$
+and set~$R^n f_*(E) \defeq H^n(\Gamma(I^\bullet))$.
+
+We adopt in this section the axiom of choice, so that enough injectives are
+available to ensure that this recipe can be carried out. The resulting sheaf
+of~$f_*\O_Y$-modules is well-defined in the following sense: Given a further
+injective resolution~$0 \to E \to J^\bullet$, there is up to homotopy precisely
+one morphism~$I^\bullet \to J^\bullet$ compatible with the identity on~$E$, and
+this morphism induces an isomorphism on cohomology.
+
+Higher direct images are pictured as a ``relative'' version of sheaf
+cohomology. Due to the result that injectivity of sheaves of modules can be
+characterized in the internal language, we can give the following precise
+rendering of this slogan.
+
+The over-locale~$Y$ corresponds to a locale~$I(Y)$ internal to~$\Sh(X)$, in
+such a way that the category of internal sheaves over this internal locale
+coincides with~$\Sh(Y)$~\cite{XXX}; in particular, a given sheaf~$E$ of~$\O_Y$-modules can
+be regarded as a sheaf over~$I(Y)$. Under this equivalence, the morphism~$f : Y
+\to X$ corresponds to the unique morphism~$I(Y) \to \mathrm{pt}$ to the
+internal one-point locale. Hence it makes sense to construct, from the internal
+point of view of~$\Sh(X)$, the sheaf cohomology~$H^n(I(Y), E)$ of~$E$.
+
+\begin{thm}The sheaf on~$X$ obtained by evaluating the expression~``$H^n(I(Y),
+E)$'' of the internal language of~$\Sh(X)$ coincides with~$R^n
+f_*(E)$.\end{thm}
+
+\begin{proof}We recall that the stack semantics is idempotent in the sense
+that~$\Sh(X) \models (\Sh(I(Y)) \models \varphi)$ if and only if~$\Sh(Y)
+\models \varphi$. Hence a sheaf of modules on~$Y$ is injective if and only if
+it is an injective module from the point of view of~$\Sh(Y)$, if and only if it
+is ... XXX Since we have assumed the axiom of choice in the metatheory and
+since internal and external injectivity coincide by
+Theorem~\ref{thm:injectivity-external-internal}, the statement~``any module has
+an injective resolution'' holds internally to~$\Sh(Y)$. Therefore the
+statement XXX
+\end{proof}
 

### 2018-10-20 paper-flabby-objects.tex
More references

@@
 injective (as a~$\ZZ$-module) if and only if it is divisible. Injective sets
-and modules have been intensively studied before~\cite{XXX}; the following
-properties are well-known:
+and modules have been intensively studied in the context of foundations
+before~\cite{blass:inj-proj-axc,harting:locally-injective,kenney:injective-choice,aczel-berg-granstroem-schuster:injective};
+the following properties are well-known:
 
@@
 \end{proof}
+
+Kenney stressed that the notion of an injective set should be regarded as an
+interesting strengthening of the constructively rather ill-behaved notion of a
+nonempty set~\cite{kenney:injective-choice}. For instance, while the statements
+``there is a choice function for every set of nonempty sets'' and even ``there
+is a choice function for every set of inhabited sets'' are constructive taboos,
+the statement ``there is a choice function for every set of injective sets'' is
+constructively neutral. Proposition~\ref{prop:set-of-preimages-flabby} demonstrates
+that the notion of a flabby set can be regarded as an interesting intermediate
+notion: In the situation of Proposition~\ref{prop:set-of-preimages-flabby}, the
+set of preimages is not only not empty or inhabited, but even flabby.
 

### 2018-10-20 paper-flabby-objects.tex
Remark on infinite time Turing machines
typo

@@
 Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence
-of~$R$-modules in an elementary topos~$\E$. Let~$M'$ be a flabby object.
-If~$\E$ is localic and Zorn's lemma is available, the induced sequence~$0 \to
-\Gamma(M') \to \Gamma(M) \to \Gamma(M') \to 0$ of~$\Gamma(R)$-modules is exact,
+of~$R$-modules in a localic topos~$\E$. Let~$M'$ be a flabby object.
+Assuming Zorn's lemma in the metatheory, the induced sequence~$0 \to
+\Gamma(M') \to \Gamma(M) \to \Gamma(M'') \to 0$ of~$\Gamma(R)$-modules is exact,
 where~$\Gamma(X) = \Hom_\E(1,X)$.\end{scholium}
@@
 \begin{proof}We only have to verify exactness at~$\Gamma(M'')$, so let~$s \in
-\Gamma(M')$. Interpreting Proposition~\ref{prop:set-of-preimages-flabby}
+\Gamma(M'')$. Interpreting Proposition~\ref{prop:set-of-preimages-flabby}
 in~$\E$, we see that the object of preimages of~$s$ is flabby. Since~$\E$ is
@@
 weaker substitute for Scholium~\ref{scholium:exact-as-presheaves} is available:
-Given~$s \in \Gamma(M'')$, the object of preimages of~$s$ is flabby. XXX
+Given~$s \in \Gamma(M'')$, the object of preimages of~$s$ is flabby. In
+particular, given any point of~$\E$, we can extend any local preimage of~$s$ to
+a preimage which is defined on an open neighborhood of that point. We believe
+that there are situations in which this weaker substitute is good enough,
+similar to how in constructive algebra often the existence of a sufficiently
+large field extension is good enough where one would classically blithely pass
+to an algebraic closure.
 
@@
 object of the topos~$BG$ of~$G$-sets, where~$G$ is a nontrivial group.
-This object is not strongly flabby, since the morphism~$\emptyset \to G$ does
+This object is not strongly flabby, since the morphism~$\emptyset \to 1$ does
 not lift, but its pullback to the slice~$BG/G \simeq \Set$ is (assuming
@@
 |X|$. Therefore the Turing machine~$e$ does terminate, with a reflexivity
-realizer for~$f(x_{K_e})$. Thus the statement~``$f(x_{K_e}) \in K$'' is
+realizer for~$f(x_{K_e})$. Thus the statement~``$f(x_{K_e}) \in K_e$'' is
 realized; hence~``$f(x_{K_e}) = x_{K_e}$'' is as well.
@@
 \end{proof}
+
+\begin{rem}The analogues of Proposition~\ref{prop:flabby-effective-sets} and
+Proposition~\ref{prop:semienough-flabby-modules} for the realizability topos
+constructed using infinite time Turing
+machines~\cite{bauer:injection,hamkins-lewis:ittm} are true as well, with the
+same proofs.\end{rem}
 
@@
 types can be shown to exist. Zermelo--Fraenkel with choice certainly suffices,
-while Zermelo--Fraenkel without choice doesn't~\cite[Section~9]{shulman-lumsdaine:hits},\footnote{With
-quotient inductive types, any infinitary algebraic theory admits free algebras.
-However, it is consistent with Zermelo--Fraenkel set theory that some such
-theories don't admit free algebras~\cite{blass:free-algebras}.} hence IZF also
-doesn't. Quotient inductive types seem to be, as the existence of enough
-injective modules, \emph{constructively neutral}.
+while Zermelo--Fraenkel without choice doesn't~\cite[Section~9]{shulman-lumsdaine:hits},
+hence IZF also doesn't.\footnote{With quotient inductive types, any infinitary
+algebraic theory admits free algebras. However, it is consistent with
+Zermelo--Fraenkel set theory that some such theories don't admit free
+algebras~\cite{blass:free-algebras}.} The existence of quotient inductive types
+seem to be, as the existence of enough injective modules, \emph{constructively
+neutral}.
 
 \item There is a number of simple constructions which come close to providing
-flabby envelopes for arbitrary modules~$M$. For instance, we could equip~$T
+flabby envelopes for arbitrary modules~$M$. For instance, we could equip the set~$T
 \defeq \P_{\leq1}(X)/{\sim}$, where~$K \sim L$ if and only if~$K = L$ or~$K

### 2018-10-20 paper-flabby-objects.tex
typo

@@
 \} \subseteq X$. Flabbiness of~$X$ implies that there exists an element~$x \in
-X$ such that~$K \subseteq \{x\}$. We have~$x = 0$ or~$x = 1$. The first case
+X$ such that~$K \subseteq \{x\}$. We have~$x \neq 1$ or~$x = 1$. The first case
 entails~$\neg\varphi$. The second case entails~$1 \in X$, so~$\varphi$.
@@
 
+\begin{rem}Although we will see in Section~\ref{sect:in-eff} that there is some
+relation between flabby sets and~$\neg\neg$-separated sets, neither notion
+encompasses the other. The set~$\Omega$ is flabby, but might fail to
+be~$\neg\neg$-separated; the set~$\NN$ is~$\neg\neg$-separated, even discrete,
+but might fail to be flabby.
+\end{rem}
+
 \begin{defn}\begin{enumerate}
-\item A set~$X$ is \emph{injective} if and only if, for any injection~$i
-: A \to B$, any map~$f : A \to X$ can be extended to a map~$B \to X$.
-\item An~$R$-module~$M$ is \emph{injective} if and only if, for any linear
-injection~$i : A \to B$ between~$R$-modules, any linear map~$f : A \to M$ can
-be extended to a linear map~$B \to M$.
+\item A set~$I$ is \emph{injective} if and only if, for any injection~$i
+: A \to B$, any map~$f : A \to I$ can be extended to a map~$B \to I$.
+\item An~$R$-module~$I$ is \emph{injective} if and only if, for any linear
+injection~$i : A \to B$ between~$R$-modules, any linear map~$f : A \to I$ can
+be extended to a linear map~$B \to I$.
 \end{enumerate}
 \[ \xymatrix{
-  A \ar@{^{(}->}[r]\ar[d] & B \ar@{-->}@/^/[ld] &&
   A \ar@{^{(}->}[r]\ar[d] & B \ar@{-->}@/^/[ld] \\
-  X &&& M
+  I
 } \]
@@
 adjoint. More explicitly, if~$i : A \to B$ is an injective map between sets and
-if~$f : A \to M$ is an arbitrary map, then the induced map~$R\langle A \rangle
+if~$f : A \to I$ is an arbitrary map, then the induced map~$R\langle A \rangle
 \to R\langle B \rangle$ between free modules is also injective, the given
-map~$f$ lifts to a linear map~$R\langle A \rangle \to M$, and an~$R$-linear
-extension~$R\langle B \rangle \to M$ induces an extension~$B \to M$ of~$f$.
+map~$f$ lifts to a linear map~$R\langle A \rangle \to I$, and an~$R$-linear
+extension~$R\langle B \rangle \to I$ induces an extension~$B \to M$ of~$f$.
 \item One verifies that any abelian group can be embedded into a divisible
-abelian group. By the axiom of choice, divisible abelian groups are injective.
+abelian group. By Baer's criterion (which requires the axiom of choice), divisible abelian groups are injective.
 The result for modules over arbitrary rings then follows purely
@@
 
-\begin{proof}Let~$X$ be an injective set. Let~$K \subseteq X$ be a subterminal.
-The inclusion~$f : K \to X$ extends along the injection~$K \to 1 = \{\star\}$
-to a map~$1 \to X$. The unique image~$x$ of that map has the property that~$K
+\begin{proof}Let~$I$ be an injective set. Let~$K \subseteq I$ be a subterminal.
+The inclusion~$f : K \to I$ extends along the injection~$K \to 1 = \{\star\}$
+to a map~$1 \to I$. The unique image~$x$ of that map has the property that~$K
 \subseteq \{x\}$.\end{proof}
@@
 \begin{cor}\label{cor:enough-flabby-sets}
-Any set can be embedded into a flabby set.\end{cor}
+Any set embeds into a flabby set.\end{cor}
 
@@
 \section{Flabby objects in the effective topos}
+\label{sect:in-eff}
 
@@
 % Konstruktion aussieht.
-
-% Rem.: diskret =/=> welk, sep =/=> welk (Bsp.: konstante Garbe Z auf Mnf.)
-% Umkehrung auch nicht (Bsp.: Godemontkonstruktion, Omega)
 
@@
 instead of injective ones. There are, however, two problems with this
-suggestion. Firstly, we needed Zorn's lemma to show that flabby sheaves are
-acyclic for the global sections functor (Proposition~XXX). This problem might
-be mitigated by relying on the substitute property XXX.
-The more serious problem is that it is an open question whether
+suggestion. Firstly, all proofs known to us that flabby sheaves are
+acyclic for the global sections functor require Zorn's lemma. This problem might
+be mitigated by relying on the substitute property discussed
+following Scholium~\ref{scholium:exact-as-presheaves}.
+But secondly, it is an open question whether
 one can show, purely intuitionistically, that any sheaf of modules embeds into

### 2018-10-20 paper-flabby-objects.tex
typo

@@
 \end{enumerate}
-\item[(4a)] Any sheaf can be embedded into a flabby sheaf.
-\item[(4b)] Any sheaf of modules can be embedded into a flabby sheaf of modules.
+\item[(4a)] Any sheaf embeds into a flabby sheaf.
+\item[(4b)] Any sheaf of modules embeds into a flabby sheaf of modules.
 \end{enumerate}
@@
 % Take care if numbering changes (referenced below)
-\item Any set can be embedded into an injective set.
+\item Any set embeds into an injective set.
 \item Any injective module is also injective as a set.
-\item Assuming the axiom of choice, any module can be embedded into an
+\item Assuming the axiom of choice, any module embeds into an
 injective module.
@@
 extension~$R\langle B \rangle \to I$ induces an extension~$B \to M$ of~$f$.
-\item One verifies that any abelian group can be embedded into a divisible
+\item One verifies that any abelian group embeds into a divisible
 abelian group. By Baer's criterion (which requires the axiom of choice), divisible abelian groups are injective.
@@
 \begin{enumerate}
-\item Any object can be embedded into an internally injective object.
+\item Any object embeds into an internally injective object.
 \item (If~$\E$ has a natural numbers object.) The underlying unstructured
@@
 % Take care if numbering changes (referenced below).
-\item Any object can be embedded into a flabby object.
+\item Any object embeds into a flabby object.
 \item The terminal object is flabby. The product of flabby objects is flabby.
@@
 Assuming the law of excluded middle in the metatheory, any~$\neg\neg$-separated
-module in the effective topos can be embedded into a flabby module.
+module in the effective topos embeds into a flabby module.
 \end{prop}

### 2018-10-20 paper-flabby-objects.tex
Finished proof that higher direct images are internal sheaf cohomology

@@
 
-We adopt in this section the axiom of choice, so that enough injectives are
-available to ensure that this recipe can be carried out. The resulting sheaf
+Assuming the axiom of choice, there are enough injective sheaves of modules so
+that this recipe can be carried out. The resulting sheaf
 of~$f_*\O_Y$-modules is well-defined in the following sense: Given a further
@@
 cohomology. Due to the result that injectivity of sheaves of modules can be
-characterized in the internal language, we can give the following precise
-rendering of this slogan.
-
-The over-locale~$Y$ corresponds to a locale~$I(Y)$ internal to~$\Sh(X)$, in
+characterized in the internal language, we can give a precise
+rendering of this slogan: We can understand higher direct images as
+internal sheaf cohomology.
+
+The details are as follows. The over-locale~$Y$ corresponds to a locale~$I(Y)$ internal to~$\Sh(X)$, in
 such a way that the category of internal sheaves over this internal locale
-coincides with~$\Sh(Y)$~\cite{XXX}; in particular, a given sheaf~$E$ of~$\O_Y$-modules can
+coincides with~$\Sh(Y)$; in particular, a given sheaf~$E$ of~$\O_Y$-modules can
 be regarded as a sheaf over~$I(Y)$. Under this equivalence, the morphism~$f : Y
@@
 
-\begin{thm}The sheaf on~$X$ obtained by evaluating the expression~``$H^n(I(Y),
-E)$'' of the internal language of~$\Sh(X)$ coincides with~$R^n
-f_*(E)$.\end{thm}
-
-\begin{proof}We recall that the stack semantics is idempotent in the sense
-that~$\Sh(X) \models (\Sh(I(Y)) \models \varphi)$ if and only if~$\Sh(Y)
-\models \varphi$. Hence a sheaf of modules on~$Y$ is injective if and only if
-it is an injective module from the point of view of~$\Sh(Y)$, if and only if it
-is ... XXX Since we have assumed the axiom of choice in the metatheory and
-since internal and external injectivity coincide by
-Theorem~\ref{thm:injectivity-external-internal}, the statement~``any module has
-an injective resolution'' holds internally to~$\Sh(Y)$. Therefore the
-statement XXX
+Usually one would not expect an internal construction which depends on
+arbitary choices to yield a globally-defined sheaf on~$X$ -- following the
+definition of the stack semantics we only obtain a family of sheaves defined on
+members of some open covering of~$X$; but we verify in
+Theorem~\ref{thm:higher-direct-images-as-internal-sheaf-cohomology} that in our
+case, it does, and that the resulting sheaf coincides with the~$n$-th higher
+direct image of~$E$.
+
+\begin{lemma}\label{lemma:notions-of-internal-injectivity}
+Let~$Y$ be a ringed locale over a locale~$X$. Let~$I$ be a sheaf of modules on~$Y$.
+Assuming Zorn's lemma in the metatheory, the following statements are equivalent:
+\begin{enumerate}
+\item $I$ is an injective sheaf of modules.
+\item From the point of view of~$\Sh(Y)$, $I$ is an injective module.
+\item From the point of view of~$\Sh(X)$, $I$ is an injective module from the
+point of view of~$\Sh(I(Y))$.
+\item From the point of view of~$\Sh(X)$, $I$ is an injective sheaf of modules
+on~$I(Y)$.
+\end{enumerate}
+\end{lemma}
+
+\begin{proof}The equivalence of the first two statements is by
+Theorem~\ref{thm:injectivity-external-internal}. The equivalence~$\text{(2)}
+\Leftrightarrow \text{(3)}$ is by the idempotency of the stack semantics:
+$\Sh(Y) \models \varphi$ if and only if~$\Sh(X) \models (\Sh(I(Y)) \models
+\varphi)$. (Shulman stated and proved a restricted version of this idempotency property
+in his original paper on the stack semantics~\cite[Lemma~7.20]{shulman:stack-semantics}.
+A proof of the general case is slightly less
+accessible~\cite[Lemma~1.20]{blechschmidt:master}.) The equivalence~$\text{(3)}
+\Leftrightarrow \text{(4)}$ is by interpreting
+Theorem~\ref{thm:injectivity-external-internal} internally to~$\Sh(X)$. This
+requires Zorn's lemma to hold internally to~$\Sh(X)$; this is indeed the case
+since we assume Zorn's lemma in the metatheory and since the validity of Zorn's
+lemma passes from the metatheory to localic
+toposes~\cite[Proposition~D4.5.14]{johnstone:elephant}.
+\end{proof}
+
+\begin{thm}\label{thm:higher-direct-images-as-internal-sheaf-cohomology}
+Let~$f : Y \to X$ be a ringed locale over a locale~$X$. Let~$E$ be a sheaf of modules over~$Y$.
+Assuming the axiom of choice in the metatheory, the expression~``$H^n(I(Y), E)$''
+of the internal language of~$\Sh(X)$ denotes a globally-defined sheaf
+on~$X$, and this sheaf coincides with~$R^n f_*(E)$.\end{thm}
+
+\begin{proof}By Lemma~\ref{lemma:notions-of-internal-injectivity} and by the
+fact that every sheaf of modules over~$Y$ admits an injective resolution, every sheaf
+of modules over~$I(Y)$ admits an injective resolution from the point of view
+of~$\Sh(X)$. Hence we can, internally to~$\Sh(X)$, carry out the construction of~$H^n(I(Y), E)$.
+Externally, this yields an open covering of~$X$ such that we have, for each
+member~$U$ of that covering
+\begin{itemize}
+\item a sheaf~$M$ on~$U$,
+\item a module structure on~$M$,
+\item a resolution~$0 \to E|_{f^{-1}U} \to I^\bullet$ by sheaves of modules which are
+internally and hence externally injective and
+\item data exhibiting~$M$ as the~$n$-th cohomology
+of~$(f|_{f^{-1}U})_*(I^\bullet)$.
+\end{itemize}
+On intersections of such opens~$U$ and~$U'$, there is exactly one
+isomorphism~$M|_{U \cap U'} \to M'|_{U \cap U'}$ of sheaves of modules induced by
+a morphism of resolutions which is compatible with the identity on~$E$.
+Hence the cocycle condition for these isomorphisms is satisfied, ensuring that
+the individual sheaves~$M$ glue to a globally-defined sheaf of modules on~$X$.
+(The individual injective resolutions need not glue to a global injective
+resolution.)
+
+The claim that this sheaf coincides with~$R^n f_*(E)$ follows from the fact
+that we can pick as internal resolution of~$E$ (considered as a sheaf
+over~$I(Y)$), the particular injective resolution of~$E$ (considered as a sheaf
+over~$Y$) used to define~$R^n f_*(E)$.
 \end{proof}

### 2018-10-21 paper-flabby-objects.tex
Applications of the internal characterization

@@
 
+The internal characterization provided by
+Theorem~\ref{thm:higher-direct-images-as-internal-sheaf-cohomology} gives, as a
+simple application, a geometric explanation that higher direct images along
+the identity~$\mathrm{id} : X \to X$ vanish: From the internal point of view
+of~$\Sh(X)$, the over-locale~$X$ corresponds to the one-point locale, and the
+higher cohomology of the one-point locale vanishes.
+
+In algebraic geometry, the internal characterization can be used to immediately
+deduce the explicit description of the higher direct images of Serre's twisting
+sheaves along the projection~$\mathbb{P}^n_S \to S$, where~$S$ is an arbitrary
+base scheme (or even base locally ringed locale), from a computation of the
+cohomology of projective~$n$-space. Background on carrying out scheme
+theory internally to a topos, is given in~\cite[Section~12]{blechschmidt:phd}.
+
 \section{Flabby objects in the effective topos}

### 2018-10-21 paper-flabby-objects.tex
typo

@@
 \usepackage{hyperref}
+\usepackage{xspace}
 
@@
 \renewcommand{\_}{\mathpunct{.}\,}
+\newcommand{\effective}{ef{}fective\xspace}
 
@@
   corresponding external notion, in stark contrast with the situation for
-  projective objects. As an application, we give internal characterizations of
-  sheaf cohomology and show that higher direct images can be understood as
-  internal cohomology.
+  projective objects. We show as an application that higher direct images can
+  be understood as internal cohomology, and study flabby objects in the
+  \effective topos.
 \end{abstract}
@@
 
-Kenney stressed that the notion of an injective set should be regarded as an
+Toby Kenney stressed that the notion of an injective set should be regarded as an
 interesting strengthening of the constructively rather ill-behaved notion of a
@@
 \section{Flabby objects}
+\label{sect:flabby-objects}
 
@@
 conditions. The equivalence of the first four is due to
-Harting~\cite{harting:locally-injective}.
+Roswitha Harting~\cite{harting:locally-injective}.
 
@@
 can check that $p_*$ reflects the property that global elements locally possess
-preimages. Details are in~\cite[Thm.~1.1]{harting}.\footnote{Harting formulates
+preimages. Details are in~\cite[Thm.~1.1]{harting:locally-injective}.\footnote{Harting formulates
 the statement for abelian group objects, and has to assume that~$\E$ contains a
@@
 projective objects (which are defined dually), internal injectivity coincides
-with external injectivity for localic toposes.
+with external injectivity for localic toposes. In the special case of sheaves
+of abelian groups, this result is due to Roswitha
+Harting~\cite[Proposition~2.1]{harting:locally-injective}.
 
@@
 injective, then~$I$ is also internally injective. The converse holds if~$\E$ is
-localic and Zorn's lemma is available.
+localic and Zorn's lemma is available in the metatheory.
 \end{thm}
@@
 Lemma~\ref{lemma:set-of-extensions-flabby}(1).
-It seems that Harting was not aware of this, even though she did show that
+It seems that Roswitha Harting was not aware of this generalization, even though she did show that
 injectivity of sheaves of modules over topological spaces is a local
-notion~\cite{XXX}, as she
+notion~\cite[Remark~5]{harting:remark}, as she
 (mistakenly) states in~\cite[page~233]{harting:remark} that ``the notions of
@@
 
-\begin{proof}The analogous statements were established purely
+\begin{proof}We established the analogous statements for sets and modules purely
 intuitionistically in Section~\ref{sect:flabby-sets}, and the stack semantics
@@
 \section{Higher direct images as internal sheaf cohomology}
+\label{sect:higher-direct-images}
 
@@
 of~$\O_Y$-modules is to pick an injective resolution~$0 \to E \to I^\bullet$
-and set~$R^n f_*(E) \defeq H^n(\Gamma(I^\bullet))$.
+and set~$R^n f_*(E) \defeq H^n(f_*(I^\bullet))$.
 
@@
 members of some open covering of~$X$; but we verify in
-Theorem~\ref{thm:higher-direct-images-as-internal-sheaf-cohomology} that in our
-case, it does, and that the resulting sheaf coincides with the~$n$-th higher
-direct image of~$E$.
+Theorem~\ref{thm:higher-direct-images-as-internal-sheaf-cohomology} below that in our
+case, it does, and that the resulting sheaf coincides with~$R^n f_*(E)$.
 
@@
 Theorem~\ref{thm:higher-direct-images-as-internal-sheaf-cohomology} gives, as a
-simple application, a geometric explanation that higher direct images along
+simple application, a logical explanation that higher direct images along
 the identity~$\mathrm{id} : X \to X$ vanish: From the internal point of view
@@
 
-\section{Flabby objects in the effective topos}
+\section{Flabby objects in the \effective topos}
 \label{sect:in-eff}
@@
 toposes, away from their original conceptual home. We begin this study with
-establishing the following observations on flabby objects in the effective
-topos. We follow the terminology of~\cite{hyland:effective-topos}.
+establishing the following observations on flabby objects in the \effective
+topos. We follow the terminology of Martin Hyland's survey on the \effective
+topos~\cite{hyland:effective-topos}.
 
 \begin{prop}\label{prop:flabby-effective-sets}
-Let~$X$ be a flabby object in the effective topos. Let~$f : X \to X$
-be a morphism. If~$X$ is effective, the statement ``$f$ has a fixed point''
-holds in the effective topos.
+Let~$X$ be a flabby object in the \effective topos. Let~$f : X \to X$
+be a morphism. If~$X$ is \effective, the statement ``$f$ has a fixed point''
+holds in the \effective topos.
 \end{prop}
@@
 Assuming the law of excluded middle in the metatheory, any~$\neg\neg$-separated
-module in the effective topos embeds into a flabby module.
+module in the \effective topos embeds into a flabby module.
 \end{prop}
@@
 reason for why Proposition~\ref{prop:flabby-effective-sets} holds is the
-following. Let~$X$ be a flabby object in the effective topos. Then there is a
+following. Let~$X$ be a flabby object in the \effective topos. Then there is a
 procedure which computes for any subterminal~$K \subseteq X$ an element~$x_K$
@@
 A corollary of Proposition~\ref{prop:flabby-effective-sets} is that the trivial
-module is the only flabby module in the effective topos whose underlying
-unstructured object is an effective set: Given such a flabby module~$M$, let~$v
+module is the only flabby module in the \effective topos whose underlying
+unstructured object is an \effective set: Given such a flabby module~$M$, let~$v
 \in M$ be an arbitrary element. Then the morphism~$x \mapsto v + x$ has a fixed
@@
 Proposition~\ref{prop:flabby-effective-sets} work, but the blame for paucity
-of flabby objects in the effective topos is to put on the realizers for
+of flabby objects in the \effective topos is to put on the realizers for
 statements of the form~``$K = K$'', where~$({=})$ is the nonstandard equality
@@
 
-This problem does not manifest with objects~$X$ which are not effective sets.
+This problem does not manifest with objects~$X$ which are not \effective sets.
 Reflexivity realizers for these objects are themselves not very informative;
@@
 the proof of Proposition~\ref{prop:semienough-flabby-modules} will accordingly
-not be effective sets.
+not be \effective sets.
 
@@
 where~$({=})$ is the nonstandard equality predicate of the object~$\P_{\leq1}(X)$
-of subterminals of~$X$. This is where the assumption that~$X$ is effective is
+of subterminals of~$X$. This is where the assumption that~$X$ is \effective is
 important; without it, we could only verify~``$K_e = K_e$'' where~$({=})$ is
@@
 Let~$(\Gamma \dashv \Delta) : \Set \to \Eff$ be the inclusion of the
-double-negation sheaves. For a~$\neg\neg$-separated module~$M$ in the effective
+double-negation sheaves. For a~$\neg\neg$-separated module~$M$ in the \effective
 topos, the canonical morphism~$M \to \Delta(\Gamma(M))$ is a monomorphism; the
@@
 
-\item There appears to be some tension regarding the effective topos:
+\item There appears to be some tension regarding the \effective topos:
 Proposition~\ref{prop:semienough-flabby-modules} shows that at
-least~$\neg\neg$-separated modules in the effective topos always embed into
+least~$\neg\neg$-separated modules in the \effective topos always embed into
 flabby modules, assuming the law of excluded middle in the metatheory, while
-Proposition~\ref{prop:flabby-effective-sets} shows that no nontrivial effective
+Proposition~\ref{prop:flabby-effective-sets} shows that no nontrivial \effective
 module is flabby.

### 2018-10-21 paper-flabby-objects.tex
Detail's on Harting's proof

@@
 injectivity and internal injectivity do not coincide'' for modules.
-% XXX comparison with Harting's proof (Barr's lemma...)
+
+It is worth noting that, because the internal language machinery was at that
+point not as well-developed as it is today, she had to go to considerable
+length to construct internal direct sums of abelian group
+objects~\cite{harting:coproduct}, and in order to verify that taking internal
+direct sums is faithful she felt the need to employ Barr's
+metatheorem~\cite[Theorem~1.7]{harting:effacements}. Nowadays we can verify
+both statements by simply carrying out an intuitionistic proof in the case of
+the topos of sets and then trusting the internal language to obtain the
+generalization to arbitrary elementary toposes with a natural numbers object.
 

### 2018-10-21 paper-flabby-objects.tex
First approximation to an introduction

@@
 
-XXX: introduction
-% XXX: In Einleitung interne Sprache erinnern.
+As is nowadays well-established, any topos supports an \emph{internal language}
+which can be used to reason about the objects and morphisms of the topos in a
+naive element-based language, allowing us to pretend that the objects are plain
+sets (or types) and the morphisms are plain maps between those sets. The
+internal language is sound with respect to intuitionistic reasoning, whereby
+any intuitionistic theorem holds in any topos.
+
+The internal language of a sheaf topos enables \emph{relativization by
+internalization}. For instance, by interpreting the
+proposition \begin{quote}``in any short exact sequence of modules, if the two
+outer ones are finitely generated then so is the middle one''\end{quote} of
+intuitionistic commutative algebra internally to the topos of sheaves over a
+space~$X$, we obtain the geometric analogue \begin{quote}``in any short exact
+sequence of sheaves of modules over~$X$, if the two outer ones are of finite
+type then so is the middle one''.\end{quote}
+This way of deducing geometric theorems provides conceptual clarity, reduces
+technical overhead and justifies certain kinds of ``fast and loose reasoning''
+typical of informal algebraic geometry. As soon as we go beyond the fragment of
+geometric sequents and consider more involved first-order or even higher-order
+statements, also significant improvements in proof length and proof complexity
+can be gained. For instance, Grothendieck's generic freeness lemma admits a
+short and simple proof in this framework, while previously-published proofs
+proceed in a somewhat involved series of reduction steps and require a fair
+amount of prerequisites in commutative
+algebra~\cite{blechschmidt:phd,blechschmidt:generic-freeness}.
+
+The practicality of this approach hinges on the extent to which the dictionary
+between internal and external notions has been worked out. For instance, the
+simple example displayed above hinges on the dictionary entry stating that a
+sheaf of modules is of finite type if and only if it looks like a
+finitely-generated module from the internal point of view. The motivation for
+this note was to find internal characterizations of flabbiness and of higher
+direct images, and the resulting entries are laid out in
+Section~\ref{sect:flabby-objects} and in
+Section~\ref{sect:higher-direct-images}: A sheaf is flabby if and only if, from
+the internal point of view, it is a flabby set -- a notion introduced in
+Section~\ref{sect:flabby-sets} below; and higher direct images look like sheaf
+cohomology from the internal point view.
+
+As a byproduct, we demonstrate how the notion of flabby sets is a useful
+organizing principle in the study of injective objects. We employ flabby sets
+to give a new proof of Roswitha Harting's result that injectivity of sheaves is
+a local notion~\cite{harting:remark} and that a sheaf is injective if and only
+if it is injective from the internal point of
+view~\cite{harting:locally-injective}, which she stated (in slightly different
+language) for sheaves of abelian groups. We use the opportunity to correct a
+small mistake of her, namely claiming that no generalization to sheaves of
+modules is possible.
+
+A further motivation for this note was our desire to seek a constructive
+account of sheaf cohomology. Sheaf cohomology is commonly defined using
+injective resolutions, which can fail to exist in the absence of the axiom of
+choice~\cite{blass:inj-proj-axc}, but it is also known that flabby resolutions
+can be used in their stead, making them the obvious candidate for a
+constructively sensible replacement of the usual definition. However, we show
+in Section~\ref{sect:in-eff} and in Section~\ref{sect:conclusion} that flabby
+resolutions present their own challenges, and in summary we failed to reach our
+goal. To the best of our knowledge, the problem of giving a constructive accout
+of sheaf cohomology is still open.
+
+In view of almost 80 years of sheaf cohomology, this state of affairs is
+slightly embarassing, challenging the call that ``once [a] subject is better
+understood, we can hope to refine its definitions and proofs so as to
+avoid that axiom''~\cite[Section~3.4]{hott}.
+
+A constructive account of sheaf cohomology would be highly desirable, not only
+out of a philosophical desire to obtain a deeper understanding of the
+foundations of sheaf cohomology, but also to: use the tools of sheaf cohomology
+in the internal setting of toposes, thereby extending the applicability of
+relativization by internalization; and to carry out \emph{integrated
+developments} of algorithms for computing sheaf cohomology, where we would
+extract algorithms together with termination and correctness proofs from a
+hypothetical constructive account.
 

### 2018-10-21 paper-flabby-objects.tex
Background on the internal language

@@
 \noindent
-
 As is nowadays well-established, any topos supports an \emph{internal language}
@@
 naive element-based language, allowing us to pretend that the objects are plain
-sets (or types) and the morphisms are plain maps between those sets. The
-internal language is sound with respect to intuitionistic reasoning, whereby
-any intuitionistic theorem holds in any topos.
+sets (or types) and that the morphisms are plain maps between those sets
+(\cite[Chapter~6]{borceux:handbook3}, \cite[Section~1.3]{caramello:ttt},
+\cite[Chapter~14]{goldblatt:topoi},
+\cite[Chapter~VI]{moerdijk-maclane:sheaves-logic}). The internal language is
+sound with respect to intuitionistic reasoning, whereby any intuitionistic
+theorem holds in any topos.
 
@@
 finitely-generated module from the internal point of view. The motivation for
-this note was to find internal characterizations of flabbiness and of higher
+this note was to find internal characterizations of flabby sheaves and of higher
 direct images, and the resulting entries are laid out in
@@
 Section~\ref{sect:higher-direct-images}: A sheaf is flabby if and only if, from
-the internal point of view, it is a flabby set -- a notion introduced in
+the internal point of view, it is a flabby set, a notion introduced in
 Section~\ref{sect:flabby-sets} below; and higher direct images look like sheaf
@@
 small mistake of her, namely claiming that no generalization to sheaves of
-modules is possible.
+modules would be possible.
+
+When employing the internal language, we are always referring to Mike Shulman's
+extension of the usual language, his \emph{stack
+semantics}~\cite{shulman:stack-semantics}. This extension allows to internalize
+unbounded quantification, which among other things is required to express the
+internal injectivity condition and the internal construction of sheaf
+cohomology.
 
@@
 
+\textbf{Acknowledgments.} We are grateful to Thorsten Altenkirch, Simon Henry
+and Maria Emilia Maietti for insightful discussions and pointers to prior work,
+and to Jürgen Jost, Marc Nieper-Wißkirchen and Peter Schuster for valuable
+guidance. Most of the work for this note was carried out at the University of
+Augsburg and the Max Planck Institute for Mathematics in the Sciences in
+Leipzig.
+
 \section{Flabby sheaves}
 
-A sheaf~$F$ on a topological space (or locale)~$X$ is \emph{flabby} (flasque) if and only
+A sheaf~$F$ on a topological space or a locale~$X$ is \emph{flabby} (flasque) if and only
 if all restriction maps~$F(X) \to F(U)$ are surjective. The following
@@
 cohomology:
-% XXX: Je nachdem, wie die Einleitung aussieht, hier die Betonung anders
-% setzen. Fundamental ist nur, dass man welke Auflösungen zur Berechnung von
-% Garbenkohomologie verwenden kann und dass sie von f_* bewahrt werden. Die
-% restlichen Eigenschaften sind wichtig, aber nicht fundamental.
 \begin{enumerate}
@@
 \item One can check that, for instance, the full powerset~$\P(X)$ and the set of
-subterminals~$\P_{\leq 1}(X)$ are each injective.~\cite{XXX}
+subterminals~$\P_{\leq 1}(X)$ are each injective.
 \item The forgetful functor from modules to sets possesses a left exact left

### 2018-10-21 paper-flabby-objects.tex
to do
Final draft

@@
   projective objects. We show as an application that higher direct images can
-  be understood as internal cohomology, and study flabby objects in the
+  be understood as internal cohomology, and we study flabby objects in the
   \effective topos.
@@
 \thispagestyle{empty}
-
-{\centering \textbf{Rough draft, comments welcome! iblech@speicherleck.de}\par}
-\bigskip
 
@@
 Section~\ref{sect:flabby-sets} below; and higher direct images look like sheaf
-cohomology from the internal point view.
+cohomology from the internal point of view.
 
@@
 language) for sheaves of abelian groups. We use the opportunity to correct a
-small mistake of her, namely claiming that no generalization to sheaves of
+small mistake of hers, namely claiming that no generalization to sheaves of
 modules would be possible.
@@
 understood, we can hope to refine its definitions and proofs so as to
-avoid that axiom''~\cite[Section~3.4]{hott}.
+avoid [the law of excluded middle]''~\cite[Section~3.4]{hott}.
 
@@
 foundations of sheaf cohomology, but also to: use the tools of sheaf cohomology
-in the internal setting of toposes, thereby extending the applicability of
+in the internal setting of toposes, thereby extending their applicability by
 relativization by internalization; and to carry out \emph{integrated
@@
 
+% XXX eff topos
+
 \textbf{Acknowledgments.} We are grateful to Thorsten Altenkirch, Simon Henry
 and Maria Emilia Maietti for insightful discussions and pointers to prior work,
-and to Jürgen Jost, Marc Nieper-Wißkirchen and Peter Schuster for valuable
-guidance. Most of the work for this note was carried out at the University of
+to Jürgen Jost, Marc Nieper-Wißkirchen and Peter Schuster for valuable
+guidance, and to Daniel Albert and Caterina Cozzi for their careful readings of
+earlier drafts. Most of the work for this note was carried out at the University of
 Augsburg and the Max Planck Institute for Mathematics in the Sciences in
@@
 requires different commitments: For the alternative definition,
-properties~(1),~(3b) and~(4a) can be verified purely intuitionistically, and
-property~(2) can be verified purely intuitionistically for open injections~$f$.
+properties~(1),~(3b) and~(4a) can be verified purely intuitionistically.
 There is a substitute for property~(3a) which can be verified purely
@@
 encompasses the other. The set~$\Omega$ is flabby, but might fail to
-be~$\neg\neg$-separated; the set~$\NN$ is~$\neg\neg$-separated, even discrete,
-but might fail to be flabby.
+be~$\neg\neg$-separated; the set~$\ZZ$ is~$\neg\neg$-separated, even discrete,
+but might fail to be flabby. This can abstractly be seen by adapting the proof of
+Proposition~\ref{prop:taboo}. An explicit model in which~$\ZZ$ is not flabby
+can be obtained by picking any topological space~$T$ such that~$H^1(T,
+\underline{\ZZ}) \neq 0$. Then the constant sheaf~$\underline{\ZZ}$ is not
+flabby and hence, by Proposition~\ref{prop:flabby-sheaves-objects} below, not a
+flabby set from the internal point of view of~$\Sh(T)$.
 \end{rem}
@@
 injection~$i : A \to B$ between~$R$-modules, any linear map~$f : A \to I$ can
-be extended to a linear map~$B \to I$.
+be extended to a linear map~$B \to I$, as in the diagram below.
 \end{enumerate}
@@
 map~$f$ lifts to a linear map~$R\langle A \rangle \to I$, and an~$R$-linear
-extension~$R\langle B \rangle \to I$ induces an extension~$B \to M$ of~$f$.
+extension~$R\langle B \rangle \to I$ induces an extension~$B \to I$ of~$f$.
 \item One verifies that any abelian group embeds into a divisible
@@
 The result for modules over arbitrary rings then follows purely
-formally~\cite{XXX}. \qedhere
+formally, since the functor~$A \mapsto \Hom(R,A)$ from abelian groups
+to~$R$-modules has a left exact left adjoint with monic unit. \qedhere
 \end{enumerate}\end{proof}
@@
 is~$M''$.'' Assuming Zorn's lemma in the metatheory, this statement is true in
-every topos of sheaves over a locale, but we don't know whether it has an
+every topos of sheaves over a locale, but we do not know whether it has an
 intuitionistic proof and in fact we surmise that it has not.
@@
 The assumption on~$f_*$ of Proposition~\ref{prop:pushforward-of-flabby-objects}
-is for instance satisfied if~$f$ is a local geometric morphism or if~$f$ is
-induced by an open continuous injection between topological spaces. XXX
+is for instance satisfied if~$f$ is a local geometric morphism.
 
@@
 sheaf cohomology. Čech methods can be carried out constructively, and
-XXX(Barakat), but it appears that there is not a general framework for sheaf
+there are constructive accounts of special cases, resulting even in
+efficient-in-practice algorithms~\cite{barakat-lh:homalg,barakat-lh:ext},
+but it appears that there is not a general framework for sheaf
 cohomology which works in an intuitionistic metatheory.
@@
 types can be shown to exist. Zermelo--Fraenkel with choice certainly suffices,
-while Zermelo--Fraenkel without choice doesn't~\cite[Section~9]{shulman-lumsdaine:hits},
-hence IZF also doesn't.\footnote{With quotient inductive types, any infinitary
+while Zermelo--Fraenkel without choice does not~\cite[Section~9]{shulman-lumsdaine:hits},
+hence IZF also does not.\footnote{With quotient inductive types, any infinitary
 algebraic theory admits free algebras. However, it is consistent with
-Zermelo--Fraenkel set theory that some such theories don't admit free
+Zermelo--Fraenkel set theory that some such theories do not admit free
 algebras~\cite{blass:free-algebras}.} The existence of quotient inductive types
@@
 \printbibliography
+\enlargethispage{1em}
 

### 2018-10-22 paper-flabby-objects.tex
typo

@@
 sheaf of modules is of finite type if and only if it looks like a
-finitely-generated module from the internal point of view. The motivation for
+finitely generated module from the internal point of view. The motivation for
 this note was to find internal characterizations of flabby sheaves and of higher
@@
 organizing principle in the study of injective objects. We employ flabby sets
-to give a new proof of Roswitha Harting's result that injectivity of sheaves is
+to give a new proof of Roswitha Harting's results that injectivity of sheaves is
 a local notion~\cite{harting:remark} and that a sheaf is injective if and only
@@
 language) for sheaves of abelian groups. We use the opportunity to correct a
-small mistake of hers, namely claiming that no generalization to sheaves of
-modules would be possible.
-
-When employing the internal language, we are always referring to Mike Shulman's
-extension of the usual language, his \emph{stack
+small mistake of hers, namely claiming that the analogous results for sheaves of
+modules would be false.
+
+When employing the internal language of a topos, we are always referring to Mike Shulman's
+extension of the usual internal language, his \emph{stack
 semantics}~\cite{shulman:stack-semantics}. This extension allows to internalize
@@
 injective resolutions, which can fail to exist in the absence of the axiom of
-choice~\cite{blass:inj-proj-axc}, but it is also known that flabby resolutions
-can be used in their stead, making them the obvious candidate for a
+choice~\cite{blass:inj-proj-axc}, but flabby resolutions
+can also be used in their stead, making them the obvious candidate for a
 constructively sensible replacement of the usual definition. However, we show
@@
 resolutions present their own challenges, and in summary we failed to reach our
-goal. To the best of our knowledge, the problem of giving a constructive accout
+goal. To the best of our knowledge, the problem of giving a constructive account
 of sheaf cohomology is still open.
@@
 Since~$I$ is injective, there exists a dotted map rendering the diagram
-commutative. This map is an element of~$X$. Furthermore, if~$K$ is inhabited,
-then this map is an element of~$K$.
+commutative. This map is an element of~$X$. If~$K$ is inhabited,
+this map is an element of~$K$.
 
@@
 If~$R$ is a ring in an elementary topos~$\E$, a similar definition can be given
-for~$R$-modules in~$\E$, referring only to the set respectively the object of
+for~$R$-modules in~$\E$, referring to the set respectively the object of
 linear maps. The condition for an object to be internally injective can be
@@
 preimages. Details are in~\cite[Thm.~1.1]{harting:locally-injective}.\footnote{Harting formulates
-the statement for abelian group objects, and has to assume that~$\E$ contains a
+her theorem for abelian group objects, and has to assume that~$\E$ contains a
 natural numbers object to ensure the existence of an abelian version of~$p_!$.}
@@
 The implication (2')~$\Rightarrow$~(1') follows by performing an extra change of
-base, since any non-global element becomes a global element after a suitable
+base, exploting that any non-global element becomes a global element after a suitable
 change of base.
@@
 Global elements of~$F$ are extensions of the kind we are looking for.
-By Lemma~\ref{lemma:set-of-extensions-flabby}, this object is flabby.
+By Lemma~\ref{lemma:set-of-extensions-flabby}(1), interpreted in~$\E$, this object is flabby.
 By Proposition~\ref{prop:global-elements}, it has a global element.
@@
 It is worth noting that, because the internal language machinery was at that
-point not as well-developed as it is today, she had to go to considerable
+point not as well-developed as it is today, Harting had to go to considerable
 length to construct internal direct sums of abelian group
@@
 Since we were careful in Section~\ref{sect:flabby-sets} to use the law of
-excluded middle or the axiom of choice only where needed, most results of that
+excluded middle and the axiom of choice only where needed, most results of that
 section carry over to flabby and internally injective objects. Specifically, we
@@
 Usually one would not expect an internal construction which depends on
-arbitary choices to yield a globally-defined sheaf on~$X$ -- following the
+arbitary choices to yield a globally-defined sheaf over~$X$ -- following the
 definition of the stack semantics we only obtain a family of sheaves defined on
@@
 \begin{lemma}\label{lemma:notions-of-internal-injectivity}
-Let~$Y$ be a ringed locale over a locale~$X$. Let~$I$ be a sheaf of modules on~$Y$.
+Let~$Y$ be a ringed locale over a locale~$X$. Let~$I$ be a sheaf of modules over~$Y$.
 Assuming Zorn's lemma in the metatheory, the following statements are equivalent:
@@
 of the internal language of~$\Sh(X)$ denotes a globally-defined sheaf
-on~$X$, and this sheaf coincides with~$R^n f_*(E)$.\end{thm}
+over~$X$, and this sheaf coincides with~$R^n f_*(E)$.\end{thm}
 
@@
 \begin{itemize}
-\item a sheaf~$M$ on~$U$,
+\item a sheaf~$M$ over~$U$,
 \item a module structure on~$M$,
@@
 that we can pick as internal resolution of~$E$ (considered as a sheaf
-over~$I(Y)$), the particular injective resolution of~$E$ (considered as a sheaf
+over~$I(Y)$) the particular injective resolution of~$E$ (considered as a sheaf
 over~$Y$) used to define~$R^n f_*(E)$.
@@
 cohomology of projective~$n$-space. Background on carrying out scheme
-theory internally to a topos, is given in~\cite[Section~12]{blechschmidt:phd}.
+theory internally to a topos is given in~\cite[Section~12]{blechschmidt:phd}.
 
@@
 
-The notion of flabby objects originates from the notion of flabby shaves and is
+The notion of flabby objects originates from the notion of flabby sheaves and is
 therefore closely connected to Grothendieck toposes. Hence it is instructive
@@
 such that~$K \subseteq \{ x_K \}$. This element might not depend extensionally
-on~$K$, but this fine point is not important for this discussion. Let
-now~$f : X \to X$ be a morphism. We construct the self-referential subset~$K
+on~$K$, but this fine point is not important for this discussion. Let~$f : X \to X$
+be a morphism. We construct the self-referential subset~$K
 \defeq \{ f(x_K) \}$; the formal proof below will indicate how this can be
@@
 \end{enumerate}
-The description of the machine~$e$ makes use of the number~$e$ coding it;
+This description of the machine~$e$ makes use of the number~$e$ coding it;
 the recursion theorem yields a general reason why this self-referentiality is
@@
 but it appears that there is not a general framework for sheaf
-cohomology which works in an intuitionistic metatheory.
+cohomology which would work in an intuitionistic metatheory.
 
@@
 \underline{x} + \underline{y}$, $\underline{rx} = r \underline{x}$; and if~$t
-\in M$, then~$\varepsilon_{\{t\}} = t$.
+\in T$, then~$\varepsilon_{\{t\}} = t$.
 
@@
 
-\item There is a number of simple constructions which come close to providing
+\item There are a number of simple constructions which come close to providing
 flabby envelopes for arbitrary modules~$M$. For instance, we could equip the set~$T

### 2018-10-22 paper-flabby-objects.tex
more typos (thanks to aspell)

@@
 In view of almost 80 years of sheaf cohomology, this state of affairs is
-slightly embarassing, challenging the call that ``once [a] subject is better
+slightly embarrassing, challenging the call that ``once [a] subject is better
 understood, we can hope to refine its definitions and proofs so as to
@@
 the generic stage: Let~$\P_{\leq1}(X)$ denote the \emph{object of subterminals}
-of~$X$; this object is a certain suboject of~$\P(X) = [X,\Omega_\E]_\E$, the
+of~$X$; this object is a certain subobject of~$\P(X) = [X,\Omega_\E]_\E$, the
 powerobject of~$X$. The subobject~$K_0$ of~$X \times \P_{\leq1}(X)$ classified by the
@@
 \begin{proof}The proof is routine; we only verify the ``only if'' direction.
-Let~$F$ be flabby as a sheaf. It suffices to verify the definining condition for stages
+Let~$F$ be flabby as a sheaf. It suffices to verify the defining condition for stages
 of the form~$A = \Hom(\cdot,U)$, where~$U$ is an open of~$X$. A monomorphism~$K
@@
 Let~$k : K \to A$ be a monomorphism in~$\E$ and let~$x : K \to f_*(X)$ be an
-arbitary morphism. Without loss of generality, we may assume that~$A$ is the
+arbitrary morphism. Without loss of generality, we may assume that~$A$ is the
 terminal object~$1$ of~$\E$. Then~$f^*(k) : f^*(K) \to 1$ is a monomorphism
@@
 The implication (2')~$\Rightarrow$~(1') follows by performing an extra change of
-base, exploting that any non-global element becomes a global element after a suitable
+base, exploiting that any non-global element becomes a global element after a suitable
 change of base.
@@
 Usually one would not expect an internal construction which depends on
-arbitary choices to yield a globally-defined sheaf over~$X$ -- following the
+arbitrary choices to yield a globally-defined sheaf over~$X$ -- following the
 definition of the stack semantics we only obtain a family of sheaves defined on

### 2018-10-27 abstract-filmat2018.tex
Beginning of the paper for the FILMAT proceedings

\documentclass[a4paper,english,11pt]{scrartcl}

\usepackage[utf8]{inputenc}

\usepackage[english]{babel}

\usepackage[protrusion=true,expansion=true]{microtype}

\usepackage[T1]{fontenc}
\usepackage{libertine}

\begin{document}

% Exploring mathematical objects from custom-tailored mathematical universes

Mathematicians study mathematical objects from the point of view of an ambient
mathematical universe. Feelings vary on the ontological status of the ambient
universe and on foundational details; but in any case it remains fixed for all
objects of study, providing a common home for relations with other objects and
setting the allowed laws of reasoning.

It is, however, also possible to study an object from the point of view of
alternate mathematical universes -- toposes -- which can be custom-tailored to
the object in question. These universes typically paint a different picture of
the studied object, which may from the alternate point of view lack properties
which it had when embedded in the usual universe, and which conversely may enjoy
unique new properties.

Studying objects from the point of view of nonstandard toposes is not done out
of abstract curiosity. It's rather done because there are certain precise
relationships between the properties from the point of view of nonstandard
toposes and properties from the point of view of the usual ambient universe.
In this way, we can gain insights we otherwise couldn't.

For instance, this approach has yielded a memorable and constructive proof of
Grothendieck's generic freeness lemma, a certain statement in commutative
algebra about modules and their localizations, for which previously only
somewhat convoluted proofs were known. The technique originated in algebraic
geometry and has yielded applications there, in commutative algebra, and in
constructive mathematics.

A basic fact of this approach is that relevant alternate universes typically
don't validate full classical logic, but only intuitionistic logic. The law of
excluded middle and the law of double negation elimination are therefore not
generally applicable in these universes; this phenomenon is irrespective of
philosophical convictions about the metalevel.

The proposed talk will explain this approach in an expository manner, without
requiring prerequisites in topos theory. The talk will center around an
easy-to-grasp example, namely that provided by the ``Zariski topos of a ring''
This topos is an alternate universe, custom-tailored to a given ring, such that
from its point of view the ring is always a field -- even if, from the usual
point of view, it is not. This observation allows to apply any intuitionistic
theorem about fields in the alternate universe. The results obtained in this
way translate to results about the ring as judged by the usual ambient
universe. The result about Grothendieck's generic freeness lemma alluded to
above is obtained in this fashion: The freeness lemma is trivial to prove
for fields and therefore holds in the Zariski topos; translated, it yields the
usual freeness lemma for rings.

The talk doesn't explicitly feature content of philosophy of mathematics. It's
rather an invitation to philosophers of mathematics to explore this technique
from a philosophical point of view and to take nonstandard toposes seriously. I
believe that this will be very worthwile, since this area has as of yet only
been explored mathematically, not philosophically.

\end{document}

### 2018-10-27 paper-filmat.tex
Beginning of the paper for the FILMAT proceedings

--- 2018-10-21 paper-flabby-objects.tex
@@
 \usepackage[utf8]{inputenc}
-\usepackage{amsthm,mathtools,tikz,float,caption,stmaryrd}
-\usetikzlibrary{patterns,matrix,decorations.pathreplacing}
+\usepackage{amsthm,mathtools,stmaryrd}
 \usepackage[all]{xy}
-\usepackage[protrusion=true,expansion=true]{microtype}
-\usepackage{hyperref}
+%\usepackage[protrusion=true,expansion=true]{microtype}
+\usepackage[hyphens]{url}
+\usepackage[breaklinks=true]{hyperref}
 \usepackage{xspace}
@@
 
-\title{Flabby and injective objects in toposes}
+\title{Exploring mathematical objects from custom-tailored mathematical universes}
 \author{Ingo Blechschmidt}
@@
 \begin{abstract}
-  We introduce a general notion of \emph{flabby objects} in elementary toposes
-  and study their basic properties. In the special case of localic toposes, this
-  notion reduces to the common notion of flabby sheaves, yielding a
-  site-independent characterization of flabby sheaves. Continuing a line of
-  research started by Roswitha Harting, we use flabby objects to
-  show that an internal notion of injective objects coincides with the
-  corresponding external notion, in stark contrast with the situation for
-  projective objects. We show as an application that higher direct images can
-  be understood as internal cohomology, and we study flabby objects in the
-  \effective topos.
+  foo
 \end{abstract}
@@
 \noindent
-As is nowadays well-established, any topos supports an \emph{internal language}
-which can be used to reason about the objects and morphisms of the topos in a
-naive element-based language, allowing us to pretend that the objects are plain
-sets (or types) and that the morphisms are plain maps between those sets
-(\cite[Chapter~6]{borceux:handbook3}, \cite[Section~1.3]{caramello:ttt},
-\cite[Chapter~14]{goldblatt:topoi},
-\cite[Chapter~VI]{moerdijk-maclane:sheaves-logic}). The internal language is
-sound with respect to intuitionistic reasoning, whereby any intuitionistic
-theorem holds in any topos.
-
-The internal language of a sheaf topos enables \emph{relativization by
-internalization}. For instance, by interpreting the
-proposition \begin{quote}``in any short exact sequence of modules, if the two
-outer ones are finitely generated then so is the middle one''\end{quote} of
-intuitionistic commutative algebra internally to the topos of sheaves over a
-space~$X$, we obtain the geometric analogue \begin{quote}``in any short exact
-sequence of sheaves of modules over~$X$, if the two outer ones are of finite
-type then so is the middle one''.\end{quote}
-This way of deducing geometric theorems provides conceptual clarity, reduces
-technical overhead and justifies certain kinds of ``fast and loose reasoning''
-typical of informal algebraic geometry. As soon as we go beyond the fragment of
-geometric sequents and consider more involved first-order or even higher-order
-statements, also significant improvements in proof length and proof complexity
-can be gained. For instance, Grothendieck's generic freeness lemma admits a
-short and simple proof in this framework, while previously-published proofs
-proceed in a somewhat involved series of reduction steps and require a fair
-amount of prerequisites in commutative
-algebra~\cite{blechschmidt:phd,blechschmidt:generic-freeness}.
-
-The practicality of this approach hinges on the extent to which the dictionary
-between internal and external notions has been worked out. For instance, the
-simple example displayed above hinges on the dictionary entry stating that a
-sheaf of modules is of finite type if and only if it looks like a
-finitely-generated module from the internal point of view. The motivation for
-this note was to find internal characterizations of flabby sheaves and of higher
-direct images, and the resulting entries are laid out in
-Section~\ref{sect:flabby-objects} and in
-Section~\ref{sect:higher-direct-images}: A sheaf is flabby if and only if, from
-the internal point of view, it is a flabby set, a notion introduced in
-Section~\ref{sect:flabby-sets} below; and higher direct images look like sheaf
-cohomology from the internal point of view.
-
-As a byproduct, we demonstrate how the notion of flabby sets is a useful
-organizing principle in the study of injective objects. We employ flabby sets
-to give a new proof of Roswitha Harting's result that injectivity of sheaves is
-a local notion~\cite{harting:remark} and that a sheaf is injective if and only
-if it is injective from the internal point of
-view~\cite{harting:locally-injective}, which she stated (in slightly different
-language) for sheaves of abelian groups. We use the opportunity to correct a
-small mistake of hers, namely claiming that no generalization to sheaves of
-modules would be possible.
-
-When employing the internal language, we are always referring to Mike Shulman's
-extension of the usual language, his \emph{stack
-semantics}~\cite{shulman:stack-semantics}. This extension allows to internalize
-unbounded quantification, which among other things is required to express the
-internal injectivity condition and the internal construction of sheaf
-cohomology.
-
-A further motivation for this note was our desire to seek a constructive
-account of sheaf cohomology. Sheaf cohomology is commonly defined using
-injective resolutions, which can fail to exist in the absence of the axiom of
-choice~\cite{blass:inj-proj-axc}, but it is also known that flabby resolutions
-can be used in their stead, making them the obvious candidate for a
-constructively sensible replacement of the usual definition. However, we show
-in Section~\ref{sect:in-eff} and in Section~\ref{sect:conclusion} that flabby
-resolutions present their own challenges, and in summary we failed to reach our
-goal. To the best of our knowledge, the problem of giving a constructive accout
-of sheaf cohomology is still open.
-
-In view of almost 80 years of sheaf cohomology, this state of affairs is
-slightly embarassing, challenging the call that ``once [a] subject is better
-understood, we can hope to refine its definitions and proofs so as to
-avoid [the law of excluded middle]''~\cite[Section~3.4]{hott}.
-
-A constructive account of sheaf cohomology would be highly desirable, not only
-out of a philosophical desire to obtain a deeper understanding of the
-foundations of sheaf cohomology, but also to: use the tools of sheaf cohomology
-in the internal setting of toposes, thereby extending their applicability by
-relativization by internalization; and to carry out \emph{integrated
-developments} of algorithms for computing sheaf cohomology, where we would
-extract algorithms together with termination and correctness proofs from a
-hypothetical constructive account.
-
-% XXX eff topos
-
-\textbf{Acknowledgments.} We are grateful to Thorsten Altenkirch, Simon Henry
-and Maria Emilia Maietti for insightful discussions and pointers to prior work,
-to Jürgen Jost, Marc Nieper-Wißkirchen and Peter Schuster for valuable
-guidance, and to Daniel Albert and Caterina Cozzi for their careful readings of
-earlier drafts. Most of the work for this note was carried out at the University of
-Augsburg and the Max Planck Institute for Mathematics in the Sciences in
-Leipzig.
-
-\section{Flabby sheaves}
-
-A sheaf~$F$ on a topological space or a locale~$X$ is \emph{flabby} (flasque) if and only
-if all restriction maps~$F(X) \to F(U)$ are surjective. The following
-properties of flabby sheaves render them fundamental to the theory of sheaf
-cohomology:
-\begin{enumerate}
-\item Let~$(U_i)_i$ be an open covering of~$X$.
-A sheaf~$F$ on~$X$ is flabby if and only if all of its restrictions~$F|_{U_i}$
-are flabby as sheaves on~$U_i$.
-\item Let~$f : X \to Y$ be a continuous map. If~$F$ is a flabby sheaf on~$X$,
-then~$f_*(F)$ is a flabby sheaf on~$Y$.
-\item[(3)] Let~$0 \to F \to G \to H \to 0$ be a short exact sequence of sheaves of
-modules.
-\begin{enumerate}
-\item If~$F$ is flabby, then this sequence is also exact as a sequence of
-presheaves.
-\item If~$F$ and~$H$ are flabby, then so is~$G$.
-\item If~$F$ and~$G$ are flabby, then so is~$H$.
-\end{enumerate}
-\item[(4a)] Any sheaf embeds into a flabby sheaf.
-\item[(4b)] Any sheaf of modules embeds into a flabby sheaf of modules.
-\end{enumerate}
-
-Since we want to develop an analogous theory for flabby objects in elementary
-toposes, it is worthwhile to analyze the logical and set-theoretic commitments
-which are required to establish these properties. The standard proofs of
-properties~(1),~(3a),~(3b) and~(3c) require Zorn's lemma to construct maximal
-extensions of given sections. The standard proof of property~(4b) requires the
-law of excluded middle, to ensure that the Godement construction actually
-yields a flabby sheaf. Properties~(2) and~(4a) can be verified purely
-intuitionistically.
-
-There is an alternative definition of flabbiness, to be introduced below, which
-is equivalent to the usual one in presence of Zorn's lemma and which
-requires different commitments: For the alternative definition,
-properties~(1),~(3b) and~(4a) can be verified purely intuitionistically.
-There is a substitute for property~(3a) which can be verified purely
-intuitionistically. We do not know whether property~(4b) can be established
-purely intuitionistically, but we give a rudimentary analysis in
-Section~\ref{sect:conclusion}.
-
-Both definitions can be generalized to yield notions of flabby objects in
-elementary toposes; but for toposes which are not localic, the two resulting
-notions will differ, and only the one obtained from the alternative definition
-is stable under pullback and
-can be characterized in the internal language. We therefore adopt in this paper the
-alternative one as the official definition.
-
-\begin{defn}\label{defn:flabby-sheaf}
-A sheaf~$F$ on a topological space (or locale)~$X$ is \emph{flabby}
-if and only if for all opens~$U$ and all sections~$s \in F(U)$, there is an
-open covering~$X = \bigcup_i U_i$ such that, for all~$i$, the section~$s$ can
-be extended to a section on~$U \cup U_i$.\end{defn}
-
-% Take care: This paragraph is referenced in prop:global-elements.
-If~$F$ is a flabby sheaf in the traditional sense, then~$F$ is obviously also
-flabby in the sense of Definition~\ref{defn:flabby-sheaf} -- singleton
-coverings will do. Conversely, let~$F$ be a flabby sheaf in the sense of
-Definition~\ref{defn:flabby-sheaf}. Let~$s \in F(U)$ be a local section. Zorn's
-lemma implies that there is a maximal extension~$s' \in F(U')$. By assumption,
-there is an open covering~$X = \bigcup_i U_i$ such that, for all~$i$, the
-section~$s'$ can be extended to~$U' \cup U_i$. Since~$s'$ is maximal, $U' \cup
-U_i = U'$ for all~$i$. Therefore~$X = \bigcup_i U_i \subseteq U'$; hence~$s'$ is a
-global section, as desired.
-
-We remark that unlike the traditional definition of flabbiness,
-Definition~\ref{defn:flabby-sheaf} exhibits flabbiness as a manifestly local
-notion.
-
-\section{Flabby sets}\label{sect:flabby-sets}
-
-We intend this section to be applied in the internal language of an elementary
-topos; we will speak about sets and maps between sets, but intend our arguments
-to be applied to objects and morphisms in toposes. We will therefore be careful
-to reason purely intuitionistically. We
-adopt the terminology of~\cite{kock:partial-maps} regarding subterminals and
-subsingletons: A subset~$K \subseteq X$ is \emph{subterminal} if and only if any given
-elements are equal ($\forall x,y \in K\_ x = y$), and it is a
-\emph{subsingleton} if and only if there is an element~$x \in X$ such that~$K
-\subseteq \{ x \}$. Any subsingleton is trivially subterminal, but the converse
-might fail.
-
-\begin{defn}A set~$X$ is \emph{flabby} if and only if any subterminal subset
-of~$X$ is a subsingleton, that is, if and only if for any subset~$K \subseteq
-X$ such that~$\forall x,y \in K\_ x = y$, there exists an element~$x \in X$
-such that~$K \subseteq \{ x \}$.
-\end{defn}
-
-In the presence of the law of excluded middle, a set is flabby if and only if
-it is inhabited. This characterization is a constructive taboo:
-
-\begin{prop}\label{prop:taboo}
-If any inhabited set is flabby, then the law of excluded middle
-holds.
-\end{prop}
-
-\begin{proof}Let~$\varphi$ be a truth value. The set~$X \defeq \{ 0 \}
-\cup \{ 1 \,|\, \varphi \} \subseteq \{ 0,1 \}$ is inhabited by~$0$ and
-contains~$1$ if and only if~$\varphi$ holds. Let~$K$ be the subterminal~$\{ 1 \,|\, \varphi
-\} \subseteq X$. Flabbiness of~$X$ implies that there exists an element~$x \in
-X$ such that~$K \subseteq \{x\}$. We have~$x \neq 1$ or~$x = 1$. The first case
-entails~$\neg\varphi$. The second case entails~$1 \in X$, so~$\varphi$.
-\end{proof}
-
-Let~$\P_{\leq 1}(X)$ be the set of subterminals of~$X$.
-
-\begin{prop}A set~$X$ is flabby if and only if the canonical map~$X \to
-\P_{\leq 1}(X)$ which sends an element~$x$ to the singleton set~$\{x\}$ is
-final.
-\end{prop}
-
-\begin{proof}By definition.\end{proof}
-
-The set~$\P_{\leq 1}(X)$ of subterminals of $X$ can be interpreted as the set
-of \emph{partially-defined elements} of $X$. In this view, the empty subset is
-the maximally undefined element and a singleton is a maximally defined element.
-A set is flabby if and only if any of its partially-defined elements can be
-refined to an honest element.
-
-\begin{rem}Although we will see in Section~\ref{sect:in-eff} that there is some
-relation between flabby sets and~$\neg\neg$-separated sets, neither notion
-encompasses the other. The set~$\Omega$ is flabby, but might fail to
-be~$\neg\neg$-separated; the set~$\ZZ$ is~$\neg\neg$-separated, even discrete,
-but might fail to be flabby. This can abstractly be seen by adapting the proof of
-Proposition~\ref{prop:taboo}. An explicit model in which~$\ZZ$ is not flabby
-can be obtained by picking any topological space~$T$ such that~$H^1(T,
-\underline{\ZZ}) \neq 0$. Then the constant sheaf~$\underline{\ZZ}$ is not
-flabby and hence, by Proposition~\ref{prop:flabby-sheaves-objects} below, not a
-flabby set from the internal point of view of~$\Sh(T)$.
-\end{rem}
-
-\begin{defn}\begin{enumerate}
-\item A set~$I$ is \emph{injective} if and only if, for any injection~$i
-: A \to B$, any map~$f : A \to I$ can be extended to a map~$B \to I$.
-\item An~$R$-module~$I$ is \emph{injective} if and only if, for any linear
-injection~$i : A \to B$ between~$R$-modules, any linear map~$f : A \to I$ can
-be extended to a linear map~$B \to I$, as in the diagram below.
-\end{enumerate}
-\[ \xymatrix{
-  A \ar@{^{(}->}[r]\ar[d] & B \ar@{-->}@/^/[ld] \\
-  I
-} \]
-\end{defn}
-
-In the presence of the law of excluded middle, a set is injective if and only
-if it is inhabited. In the presence of the axiom of choice, an abelian group is
-injective (as a~$\ZZ$-module) if and only if it is divisible. Injective sets
-and modules have been intensively studied in the context of foundations
-before~\cite{blass:inj-proj-axc,harting:locally-injective,kenney:injective-choice,aczel-berg-granstroem-schuster:injective};
-the following properties are well-known:
-
-\begin{prop}\label{prop:basics-injective}
-\begin{enumerate}
-% Take care if numbering changes (referenced below)
-\item Any set embeds into an injective set.
-\item Any injective module is also injective as a set.
-\item Assuming the axiom of choice, any module embeds into an
-injective module.
-\end{enumerate}\end{prop}
-
-\begin{proof}\begin{enumerate}
-\item One can check that, for instance, the full powerset~$\P(X)$ and the set of
-subterminals~$\P_{\leq 1}(X)$ are each injective.
-\item The forgetful functor from modules to sets possesses a left exact left
-adjoint. More explicitly, if~$i : A \to B$ is an injective map between sets and
-if~$f : A \to I$ is an arbitrary map, then the induced map~$R\langle A \rangle
-\to R\langle B \rangle$ between free modules is also injective, the given
-map~$f$ lifts to a linear map~$R\langle A \rangle \to I$, and an~$R$-linear
-extension~$R\langle B \rangle \to I$ induces an extension~$B \to I$ of~$f$.
-\item One verifies that any abelian group embeds into a divisible
-abelian group. By Baer's criterion (which requires the axiom of choice), divisible abelian groups are injective.
-The result for modules over arbitrary rings then follows purely
-formally, since the functor~$A \mapsto \Hom(R,A)$ from abelian groups
-to~$R$-modules has a left exact left adjoint with monic unit. \qedhere
-\end{enumerate}\end{proof}
-
-\begin{prop}\label{prop:injective-flabby}
-Any injective set is flabby.\end{prop}
-
-\begin{proof}Let~$I$ be an injective set. Let~$K \subseteq I$ be a subterminal.
-The inclusion~$f : K \to I$ extends along the injection~$K \to 1 = \{\star\}$
-to a map~$1 \to I$. The unique image~$x$ of that map has the property that~$K
-\subseteq \{x\}$.\end{proof}
-
-\begin{cor}\label{cor:enough-flabby-sets}
-Any set embeds into a flabby set.\end{cor}
-
-\begin{proof}Immediate by Proposition~\ref{prop:basics-injective}(1) and
-Proposition~\ref{prop:injective-flabby}.\end{proof}
-
-A further corollary of Proposition~\ref{prop:injective-flabby} is that the
-statement ``any inhabited set is injective'' is a constructive taboo: If any
-inhabited set is injective, then any inhabited set is flabby, thus the law of
-excluded middle follows by Proposition~\ref{prop:taboo}.
-
-\begin{prop}Any singleton set is flabby. The cartesian product of flabby sets
-is flabby.\end{prop}
-
-\begin{proof}Immediate.\end{proof}
-
-Subsets of flabby sets are in general not flabby, as else any set would be
-flabby in view of Corollary~\ref{cor:enough-flabby-sets}.
-
-\begin{prop}\label{prop:hom-flabby}
-\begin{enumerate}
-\item Let~$I$ be an injective set. Let~$T$ be an arbitrary set. Then the
-set~$I^T$ of maps from~$T$ to~$I$ is flabby.
-\item Let~$I$ be an injective~$R$-module. Let~$T$ be an arbitrary~$R$-module. Then the
-set~$\Hom_R(T,I)$ of linear maps from~$T$ to~$I$ is flabby.
-\end{enumerate}\end{prop}
-
-\begin{proof}We first cover the case of sets. Let~$K \subseteq I^T$ be a
-subterminal. We consider the injectivity diagram
-\[ \xymatrix{
-  T' \ar@{^{(}->}[r]\ar[d] & T \ar@{-->}@/^/[ld] \\
-  I
-} \]
-where~$T'$ is the subset~$\{ s \in T \,|\, \text{$K$ is inhabited} \} \subseteq T$ and the
-solid vertical map sends~$s \in T'$ to~$g(s)$, where~$g$ is an arbitrary element
-of~$K$. This association is well-defined. Since~$I$ is injective, a dotted lift
-as indicated exists. If~$K$ is inhabited, this lift is an element of~$K$.
-
-The same kind of argument applies to the case of modules. If~$K$ is a
-subterminal of~$\Hom_R(T,I)$, we define~$T'$ to be the submodule
-$\{ s \in T \,|\, \text{$s = 0$ or $K$ is inhabited} \}$ and consider the
-analogous injectivity diagram, where the solid vertical map~$f : T' \to I$ is now
-defined by cases: Let~$s \in T'$. If~$s = 0$, then we set~$f(s) = 0$; if~$K$ is
-inhabited, then we set~$f(s) \defeq g(s)$, where~$g$ is an arbitrary element
-of~$K$. This association is again well-defined, and a dotted lift yields the
-desired element of~$\Hom_R(T,I)$.
-\end{proof}
-
-Proposition~\ref{prop:hom-flabby} can be used to give an alternative proof of
-Proposition~\ref{prop:injective-flabby} and to generalize
-Proposition~\ref{prop:injective-flabby} to modules: If~$I$ is an injective set,
-then the set~$I^1 \cong I$ is flabby. If~$I$ is an injective module, then the
-set~$\Hom_R(R,I) \cong I$ is flabby.
-
-\begin{lemma}\label{lemma:set-of-extensions-flabby}
-\begin{enumerate}
-\item Let~$I$ be an injective set. Let~$i : A \to B$ be an injection.
-Let~$f : A \to I$ be an arbitrary map. Then the set of extensions of~$f$ to~$B$
-is flabby.
-\item Let~$I$ be an injective~$R$-module. Let~$i : A \to B$ be a linear injection.
-Let~$f : A \to I$ be an arbitrary linear map. Then the set of linear extensions of~$f$ to~$B$
-is flabby.
-\end{enumerate}
-\end{lemma}
-
-\begin{proof}For the first claim, we set~$X \defeq \{ \bar{f} \in I^B \,|\, \bar{f} \circ i =
-f \}$. Let~$K \subseteq X$ be a subterminal. We consider the injectivity diagram
-\[ \xymatrix{
-  i[A] \cup B' \ar@{^{(}->}[r]\ar[d]_g & B \ar@{-->}@/^/[ld] \\
-  I
-} \]
-where~$B'$ is the set~$\{ s \in B \,|\, \text{$K$ is inhabited} \}$ and the solid
-vertical arrow~$g$ is defined in the following way: Let~$s \in i[A] \cup B'$.
-If~$s \in i[A]$, we set~$g(s) \defeq f(a)$, where~$a \in A$ is an element such
-that~$s = i(a)$. If~$s \in B'$, we set~$g(s) \defeq \bar{f}(s)$,
-where~$\bar{f}$ is any element of~$K$. These prescriptions determine a well-defined
-map.
-
-Since~$I$ is injective, there exists a dotted map rendering the diagram
-commutative. This map is an element of~$X$. Furthermore, if~$K$ is inhabited,
-then this map is an element of~$K$.
-
-The proof of the second claim is similar. We
-set~$X \defeq \{ \bar{f} \in \Hom_R(B,I) \,|\, \bar{f} \circ i =
-f \}$. Let~$K \subseteq X$ be a subterminal. We consider the injectivity diagram
-\[ \xymatrix{
-  i[A] + B' \ar@{^{(}->}[r]\ar[d]_g & B \ar@{-->}@/^/[ld] \\
-  I
-} \]
-where~$B'$ is the submodule~$\{ t \in B \,|\, \text{$t = 0$ or $K$ is
-inhabited} \} \subseteq B$ and the solid vertical arrow~$g$ is defined in the following
-way: Let~$s \in i[A] + B'$. Then~$s = i(a) + t$ for an element~$a \in A$ and an
-element~$t \in B'$. Since~$t \in B'$, $t = 0$ or~$K$ is inhabited. If~$t = 0$,
-we set~$g(s) \defeq f(a)$. If~$K$ is inhabited, we set~$g(s) \defeq f(a) +
-\bar{f}(s)$, where~$\bar{f}$ is any element of~$K$. These prescriptions
-determine a well-defined map.
-
-Since~$I$ is injective, there exists a dotted map rendering the diagram
-commutative. This map is an element of~$X$. Furthermore, if~$K$ is inhabited,
-then this map is an element of~$K$.
-\end{proof}
-
-\begin{prop}\label{prop:set-of-preimages-flabby}
-Let~$0 \to M' \xra{i} M \xra{p} M'' \to 0$ be a short exact
-sequence of modules. Let~$s \in M''$. If~$M'$ is flabby, then the set of
-preimages of~$s$ under~$p$ is flabby.
-\end{prop}
-
-\begin{proof}Let~$X \defeq \{ u \in M \,|\, p(u) = s
-\}$. Let~$K \subseteq X$ be a subterminal. Since~$p$ is surjective, there is an
-element~$u_0 \in X$. The translated set~$K - u_0 \subseteq M$ is still a
-subterminal, and its preimage under~$i$ is as well. Since~$M'$ is flabby, there
-is an element~$v \in M'$ such that~$i^{-1}[K - u_0] \subseteq \{v\}$. We verify
-that~$K \subseteq \{u_0 + i(v)\}$.
-
-Thus let~$u \in K$ be given. Then~$p(u - u_0) = 0$, so by exactness the
-set~$i^{-1}[K - u_0]$ is inhabited. It therefore contains~$v$. Thus~$i(v) \in K
-- u_0$. Since~$K = \{u\}$, it follows that~$i(v) = u - u_0$, so~$u \in \{u_0 +
-i(v)\}$ as claimed.
-\end{proof}
-
-Toby Kenney stressed that the notion of an injective set should be regarded as an
-interesting strengthening of the constructively rather ill-behaved notion of a
-nonempty set~\cite{kenney:injective-choice}. For instance, while the statements
-``there is a choice function for every set of nonempty sets'' and even ``there
-is a choice function for every set of inhabited sets'' are constructive taboos,
-the statement ``there is a choice function for every set of injective sets'' is
-constructively neutral. Proposition~\ref{prop:set-of-preimages-flabby} demonstrates
-that the notion of a flabby set can be regarded as an interesting intermediate
-notion: In the situation of Proposition~\ref{prop:set-of-preimages-flabby}, the
-set of preimages is not only not empty or inhabited, but even flabby.
-
-\begin{prop}Let~$0 \to M' \xra{i} M \xra{p} M'' \to 0$ be a short exact
-sequence of modules. If~$M'$ and~$M''$ are flabby, so is~$M$.
-\end{prop}
-
-\begin{proof}Let~$K \subseteq M$ be a subterminal. Then its image~$p[K] \subseteq M''$
-is a subterminal as well. Since~$M''$ is flabby, there is an element~$s \in
-M''$ such that~$p[K] \subseteq \{ s \}$.
-
-Since~$p$ is surjective, there is an element~$u_0 \in M$ such that~$p(u_0) =
-s$.
-
-The preimage~$i^{-1}[K - u_0] \subseteq M'$ is a subterminal. Since~$M'$ is
-flabby, there exists an element~$v \in M'$ such that~$i^{-1}[K - u_0] \subseteq
-\{v\}$.
-
-Thus~$K \subseteq \{ u_0 + i(v) \}$.
-\end{proof}
-
-Noticeably missing here is a statement as follows: ``Let~$0 \to M' \to M \to
-M'' \to 0$ be a short exact sequence of modules. If~$M'$ and~$M$ are flabby, so
-is~$M''$.'' Assuming Zorn's lemma in the metatheory, this statement is true in
-every topos of sheaves over a locale, but we do not know whether it has an
-intuitionistic proof and in fact we surmise that it has not.
-
-\section{Flabby objects}
-\label{sect:flabby-objects}
-
-\begin{defn}An object~$X$ of an elementary topos~$\E$ is \emph{flabby} if and
-only if the statement~``$X$ is a flabby set'' holds in the stack semantics
-of~$\E$.\end{defn}
-
-This definition amounts to the following: An object~$X$ of an elementary
-topos~$\E$ is flabby if and only if, for any monomorphism~$K \to A$ and any
-morphism~$K \to X$, there exists an epimorphism~$B \to A$ and a morphism~$B
-\to X$ such that the following diagram commutes.
-\[ \xymatrix{
-  K \times_A B \ar@{^{(}->}[r]\ar[d] & B \ar@{-->}@/^/[ld] \\
-  X
-} \]
-
-Instead of referencing arbitrary stages~$A \in \E$, one can also just reference
-the generic stage: Let~$\P_{\leq1}(X)$ denote the \emph{object of subterminals}
-of~$X$; this object is a certain suboject of~$\P(X) = [X,\Omega_\E]_\E$, the
-powerobject of~$X$. The subobject~$K_0$ of~$X \times \P_{\leq1}(X)$ classified by the
-evaluation morphism~$X \times \P_{\leq1}(X) \to X \times \P(X) \to \Omega_\E$
-is the \emph{generic subterminal} of~$X$. The object~$X$ is flabby if and only
-if there exists an epimorphism~$B \to \P_{\leq1}(X)$ and a morphism~$B \to X$
-such that the following diagram commutes.
-\[ \xymatrix{
-  K_0 \times_{\P_{\leq1}(X)} B \ar@{^{(}->}[r]\ar[d] & \P_{\leq1}(X) \ar@{-->}@/^/[ld] \\
-  X
-} \]
-
-\begin{prop}\label{prop:basic-properties-of-flabby-objects}
-Let~$X$ and~$T$ be objects of an elementary topos~$\E$.
-\begin{enumerate}
-\item If~$X$ is flabby, so is~$X \times T$ as an object of~$\E/T$.
-\item The converse holds if the unique morphism~$T \to 1$ is an epimorphism.
-\end{enumerate}
-\end{prop}
-
-\begin{proof}This holds for any property which can be defined
-in the stack semantics~\cite[Lemma~7.3]{shulman:stack-semantics}.
-\end{proof}
-
-\begin{prop}\label{prop:flabby-sheaves-objects}
-Let~$F$ be a sheaf on a topological space~$X$ (or a locale).
-Then~$F$ is flabby as a sheaf if and only if~$F$ is flabby as an object of the
-sheaf topos~$\Sh(X)$.
-\end{prop}
-
-\begin{proof}The proof is routine; we only verify the ``only if'' direction.
-Let~$F$ be flabby as a sheaf. It suffices to verify the definining condition for stages
-of the form~$A = \Hom(\cdot,U)$, where~$U$ is an open of~$X$. A monomorphism~$K
-\to A$ then amounts to an open~$V \subseteq U$ (the union of all opens on
-which~$K$ is inhabited). A morphism~$K \to F$ amounts to a section~$s \in
-F(V)$. Since~$F$ is flabby as a sheaf, there is an open covering~$X =
-\bigcup_{i \in I} V_i$ such that, for all~$i$, the section~$s$ can be extended
-to a section~$s_i$ of~$V \cup V_i$. The desired epimorphism is~$B \defeq
-\coprod_i \Hom(\cdot,(V \cup V_i) \cap U) \to A$, and the desired morphism~$B
-\to X$ is given by the sections~$s_i|_{(V \cup V_i) \cap U}$.
-
-As stated, the argument in the previous paragraph requires the axiom of choice
-to pick the extensions~$s_i$; this can be avoided by a standard trick of
-expanding the index set of the coproduct to include the choices: We redefine $B \defeq
-\coprod_{(i,t) \in I'} \Hom(\cdot, (V \cup V_i) \cap U)$, where~$I' = \{ (i \in
-I, t \in F(V \cup V_i)) \,|\, t|_V = s \}$ and define the morphism~$B \to X$ on
-the~$(i,t)$-summand by~$t|_{(V \cup V_i) \cap U}$.
-\end{proof}
-
-\begin{prop}\label{prop:global-elements}
-Let~$X$ be a flabby object of a localic topos~$\E$. If
-Zorn's lemma is available in the metatheory, then~$X$ possesses a global element (a morphism~$1 \to X$).
-\end{prop}
-
-\begin{proof}This is a restatement of the discussion following
-Definition~\ref{defn:flabby-sheaf}.
-\end{proof}
-
-\begin{prop}\label{prop:pushforward-of-flabby-objects}
-Let~$f : \F \to \E$ be a geometric morphism. If~$f_*$ preserves epimorphisms,
-then~$f_*$ preserves flabby objects.\end{prop}
-
-\begin{proof}Let~$X \in \F$ be a flabby object.
-Let~$k : K \to A$ be a monomorphism in~$\E$ and let~$x : K \to f_*(X)$ be an
-arbitary morphism. Without loss of generality, we may assume that~$A$ is the
-terminal object~$1$ of~$\E$. Then~$f^*(k) : f^*(K) \to 1$ is a monomorphism
-in~$\F$ and the adjoint transpose~$x^t : f^*(K) \to X$
-is a morphism in~$\F$. Since~$X$ is flabby, there is an epimorphism~$B \to 1$
-in~$\F$ and a morphism~$y : B \to X$ such that the morphism~$f^*(K) \times B
-\to X$ factors over~$y$. Hence~$x$ factors over~$f_*(y) : f_*(B) \to f_*(X)$.
-We conclude because the morphism~$f_*(B) \to f_*(1)$ is an epimorphism by
-assumption.
-\end{proof}
-
-The assumption on~$f_*$ of Proposition~\ref{prop:pushforward-of-flabby-objects}
-is for instance satisfied if~$f$ is a local geometric morphism.
-
-\begin{defn}An object~$I$ of an elementary topos~$\E$ is \emph{externally
-injective} if and only if for any monomorphism~$A \to B$ in~$\E$, the canonical
-map~$\Hom_\E(B,I) \to \Hom_\E(A,I)$ is surjective. It is \emph{internally
-injective} if and only if for any monomorphism~$A \to B$ in~$\E$, the canonical
-morphism~$[B,I] \to [A,I]$ between Hom objects is an epimorphism in~$\E$.
-\end{defn}
-
-If~$R$ is a ring in an elementary topos~$\E$, a similar definition can be given
-for~$R$-modules in~$\E$, referring only to the set respectively the object of
-linear maps. The condition for an object to be internally injective can be
-rephrased in various ways. The following proposition lists five of these
-conditions. The equivalence of the first four is due to
-Roswitha Harting~\cite{harting:locally-injective}.
-
-\begin{prop}\label{prop:notions-of-internal-injectivity}
-Let~$\E$ be an elementary topos. Then the following statements about an
-object~$I \in \E$ are equivalent.
-% Take care if numbering changes (is referenced below).
-\begin{enumerate}
-\item[(1)] $I$ is internally injective.
-\item[(1')] For any morphism $p : A \to 1$ in $\E$, the object $p^*(I)$ has property~(1)
-as an object of $\E/A$.
-\item[(2)] The functor~$[\cdot, I] : \E^\op \to \E$ maps monomorphisms in $\E$
-to morphisms for which any global element of the target locally (after change of
-base along an epimorphism) possesses a preimage.
-\item[(2')] For any morphism $p : A \to 1$ in $\E$, the object $p^*(I)$ has property~(2)
-as an object of $\E/A$.
-\item[(3)] The statement~``$I$ is an injective set'' holds in the stack
-semantics of~$\E$.
-\end{enumerate}
-\end{prop}
-
-\begin{proof}
-The implications (1)~$\Rightarrow$~(2), (1')~$\Rightarrow$~(2'),
-(1')~$\Rightarrow$~(1) and (2')~$\Rightarrow$~(2) are trivial.
-
-The equivalence (1')~$\Leftrightarrow$~(3) follows directly from the
-interpretation rules of the stack semantics.
-
-The implication (2)~$\Rightarrow$~(2') employs the
-extra left adjoint $p_! : \E/A \to \E$ of $p^* : \E
-\to \E/A$~(which maps an object~$(X \to A)$ to~$X$), as in the usual proof that
-injective sheaves remain injective when
-restricted to smaller open subsets: We have that $p_* \circ [\cdot, p^*(I)]_{\E/A}
-\cong [\cdot, I]_\E \circ p_!$, the functor $p_!$ preserves monomorphisms, and one
-can check that $p_*$ reflects the property that global elements locally possess
-preimages. Details are in~\cite[Thm.~1.1]{harting:locally-injective}.\footnote{Harting formulates
-the statement for abelian group objects, and has to assume that~$\E$ contains a
-natural numbers object to ensure the existence of an abelian version of~$p_!$.}
-
-The implication (2')~$\Rightarrow$~(1') follows by performing an extra change of
-base, since any non-global element becomes a global element after a suitable
-change of base.
-\end{proof}
-
-Let~$R$ be a ring in~$\E$. Then the analogue of
-Proposition~\ref{prop:notions-of-internal-injectivity} holds for~$R$-modules
-in~$\E$, if~$\E$ is assumed to have a natural numbers object. The extra
-assumption is needed in order to construct the left adjoint~$p_! :
-\Mod_{\E/A}(R \times A) \to \Mod_\E(R)$. Phrased in the internal language, this
-adjoint maps a family~$(M_a)_{a \in A}$ of~$R$-modules to the direct
-sum~$\bigoplus_{a \in A} M_a$. Details on this construction, phrased in the
-language of sets but interpretable in the internal language, can for instance
-be found in~\cite[page~54]{mines-richman-ruitenburg:constructive-algebra}.
-
-Somewhat surprisingly, and in stark contrast with the situation for internally
-projective objects (which are defined dually), internal injectivity coincides
-with external injectivity for localic toposes. In the special case of sheaves
-of abelian groups, this result is due to Roswitha
-Harting~\cite[Proposition~2.1]{harting:locally-injective}.
-
-\begin{thm}\label{thm:injectivity-external-internal}
-Let~$I$ be an object of an elementary topos~$\E$. If~$I$ is externally
-injective, then~$I$ is also internally injective. The converse holds if~$\E$ is
-localic and Zorn's lemma is available in the metatheory.
-\end{thm}
-
-\begin{proof}For the ``only if'' direction, let~$I$ be an object
-which is externally injective. Then~$I$ satisfies Condition~(2) in
-Proposition~\ref{prop:notions-of-internal-injectivity}, even without having to
-pass to covers.
-
-For the ``if'' direction, let~$I$ be an internally
-injective object. Let~$i : A \to B$ be a monomorphism in~$\E$ and let~$f :
-A \to I$ be an arbitrary morphism. We want to show that there exists an
-extension $B \to I$ of~$f$ along~$i$. To this end, we consider the object of
-such extensions, defined by the internal expression
-\[ F \defeq \{ \bar{f} \in [B,I] \,|\, \bar{f} \circ i = f \}. \]
-Global elements of~$F$ are extensions of the kind we are looking for.
-By Lemma~\ref{lemma:set-of-extensions-flabby}, this object is flabby.
-By Proposition~\ref{prop:global-elements}, it has a global element.
-\end{proof}
-
-The analogue of Theorem~\ref{thm:injectivity-external-internal} for modules
-holds as well, if~$\E$ is assumed to have a natural numbers object. The proof
-carries over word for word, only referencing
-Lemma~\ref{lemma:set-of-extensions-flabby}(2) instead of
-Lemma~\ref{lemma:set-of-extensions-flabby}(1).
-It seems that Roswitha Harting was not aware of this generalization, even though she did show that
-injectivity of sheaves of modules over topological spaces is a local
-notion~\cite[Remark~5]{harting:remark}, as she
-(mistakenly) states in~\cite[page~233]{harting:remark} that ``the notions of
-injectivity and internal injectivity do not coincide'' for modules.
-
-It is worth noting that, because the internal language machinery was at that
-point not as well-developed as it is today, she had to go to considerable
-length to construct internal direct sums of abelian group
-objects~\cite{harting:coproduct}, and in order to verify that taking internal
-direct sums is faithful she felt the need to employ Barr's
-metatheorem~\cite[Theorem~1.7]{harting:effacements}. Nowadays we can verify
-both statements by simply carrying out an intuitionistic proof in the case of
-the topos of sets and then trusting the internal language to obtain the
-generalization to arbitrary elementary toposes with a natural numbers object.
-
-Since we were careful in Section~\ref{sect:flabby-sets} to use the law of
-excluded middle or the axiom of choice only where needed, most results of that
-section carry over to flabby and internally injective objects. Specifically, we
-have:
-
-\begin{scholium}\label{scholium:properties-of-flabby-objects}
-For any elementary topos~$\E$:
-\begin{enumerate}
-\item Any object embeds into an internally injective object.
-\item (If~$\E$ has a natural numbers object.) The underlying unstructured
-object of an internally injective module is internally injective.
-\item Any internally injective object is flabby.
-% Take care if numbering changes (referenced below).
-\item Any object embeds into a flabby object.
-\item The terminal object is flabby. The product of flabby objects is flabby.
-\item Let~$I$ be an internally injective object. Let~$T$ be an arbitrary
-object. Then~$[T,I]$ is a flabby object.
-\item (If~$\E$ has a natural numbers object.) Let~$I$ be an internally
-injective~$R$-module. Let~$T$ be an arbitrary~$R$-module. Then~$[T,I]_R$, the
-subobject of the internal Hom consisting only of the linear maps, is a flabby
-object.
-\item Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence
-of~$R$-modules in~$\E$. If~$M'$ and~$M''$ are flabby objects, so is~$M$.
-\end{enumerate}
-\end{scholium}
-
-\begin{proof}We established the analogous statements for sets and modules purely
-intuitionistically in Section~\ref{sect:flabby-sets}, and the stack semantics
-is sound with respect to intuitionistic logic.
-\end{proof}
-
-\begin{scholium}\label{scholium:exact-as-presheaves}
-Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence
-of~$R$-modules in a localic topos~$\E$. Let~$M'$ be a flabby object.
-Assuming Zorn's lemma in the metatheory, the induced sequence~$0 \to
-\Gamma(M') \to \Gamma(M) \to \Gamma(M'') \to 0$ of~$\Gamma(R)$-modules is exact,
-where~$\Gamma(X) = \Hom_\E(1,X)$.\end{scholium}
-
-\begin{proof}We only have to verify exactness at~$\Gamma(M'')$, so let~$s \in
-\Gamma(M'')$. Interpreting Proposition~\ref{prop:set-of-preimages-flabby}
-in~$\E$, we see that the object of preimages of~$s$ is flabby. Since~$\E$ is
-localic, this object is a flabby sheaf; since Zorn's lemma is available, it
-possesses a global element. Such an element is the desired preimage of~$s$
-in~$\Gamma(M)$.\end{proof}
-
-If~$\E$ is not necessarily localic or Zorn's lemma is not available, only a
-weaker substitute for Scholium~\ref{scholium:exact-as-presheaves} is available:
-Given~$s \in \Gamma(M'')$, the object of preimages of~$s$ is flabby. In
-particular, given any point of~$\E$, we can extend any local preimage of~$s$ to
-a preimage which is defined on an open neighborhood of that point. We believe
-that there are situations in which this weaker substitute is good enough,
-similar to how in constructive algebra often the existence of a sufficiently
-large field extension is good enough where one would classically blithely pass
-to an algebraic closure.
-
-\begin{rem}A direct generalization of the traditional notion of a flabby sheaf, as
-opposed to our reimagining in Definition~\ref{defn:flabby-sheaf}, to
-elementary toposes is the following. An object~$X$ of an elementary topos~$\E$
-is \emph{strongly flabby} if and only if, for any monomorphism~$K \to 1$
-in~$\E$, any morphism~$K \to X$ lifts to a morphism~$1 \to X$.
-
-One can verify, purely intuitionistically, that a sheaf~$F$ on a space~$T$ is
-flabby in the traditional sense if and only if~$F$ is a strongly flabby object
-of~$\Sh(T)$.
-
-The notion of strongly flabby objects is, however, not local (in the same sense
-that the notion of flabby objects is, as stated in
-Proposition~\ref{prop:basic-properties-of-flabby-objects})
-and therefore cannot be characterized in the internal language. A specific
-example is the~$G$-set~$G$ (with the translation action), considered as an
-object of the topos~$BG$ of~$G$-sets, where~$G$ is a nontrivial group.
-This object is not strongly flabby, since the morphism~$\emptyset \to 1$ does
-not lift, but its pullback to the slice~$BG/G \simeq \Set$ is (assuming
-the law of excluded middle in the metatheory), and the unique morphism~$G \to
-1$ is indeed an epimorphism.
-\end{rem}
-
-\section{Higher direct images as internal sheaf cohomology}
-\label{sect:higher-direct-images}
-
-Let~$X$ be a locale and let~$f : Y \to X$ be an over-locale. By the fundamental
-relation between locales and topological spaces, this situation arises
-for instance, when given a sober topological space and a topological
-space over it, as is often the case in algebraic topology or algebraic
-geometry. Let a sheaf~$\O_Y$ of rings on~$Y$ be given. Then the
-traditional way to define the \emph{higher direct images} of a sheaf~$E$
-of~$\O_Y$-modules is to pick an injective resolution~$0 \to E \to I^\bullet$
-and set~$R^n f_*(E) \defeq H^n(f_*(I^\bullet))$.
-
-Assuming the axiom of choice, there are enough injective sheaves of modules so
-that this recipe can be carried out. The resulting sheaf
-of~$f_*\O_Y$-modules is well-defined in the following sense: Given a further
-injective resolution~$0 \to E \to J^\bullet$, there is up to homotopy precisely
-one morphism~$I^\bullet \to J^\bullet$ compatible with the identity on~$E$, and
-this morphism induces an isomorphism on cohomology.
-
-Higher direct images are pictured as a ``relative'' version of sheaf
-cohomology. Due to the result that injectivity of sheaves of modules can be
-characterized in the internal language, we can give a precise
-rendering of this slogan: We can understand higher direct images as
-internal sheaf cohomology.
-
-The details are as follows. The over-locale~$Y$ corresponds to a locale~$I(Y)$ internal to~$\Sh(X)$, in
-such a way that the category of internal sheaves over this internal locale
-coincides with~$\Sh(Y)$; in particular, a given sheaf~$E$ of~$\O_Y$-modules can
-be regarded as a sheaf over~$I(Y)$. Under this equivalence, the morphism~$f : Y
-\to X$ corresponds to the unique morphism~$I(Y) \to \mathrm{pt}$ to the
-internal one-point locale. Hence it makes sense to construct, from the internal
-point of view of~$\Sh(X)$, the sheaf cohomology~$H^n(I(Y), E)$ of~$E$.
-
-Usually one would not expect an internal construction which depends on
-arbitary choices to yield a globally-defined sheaf on~$X$ -- following the
-definition of the stack semantics we only obtain a family of sheaves defined on
-members of some open covering of~$X$; but we verify in
-Theorem~\ref{thm:higher-direct-images-as-internal-sheaf-cohomology} below that in our
-case, it does, and that the resulting sheaf coincides with~$R^n f_*(E)$.
-
-\begin{lemma}\label{lemma:notions-of-internal-injectivity}
-Let~$Y$ be a ringed locale over a locale~$X$. Let~$I$ be a sheaf of modules on~$Y$.
-Assuming Zorn's lemma in the metatheory, the following statements are equivalent:
-\begin{enumerate}
-\item $I$ is an injective sheaf of modules.
-\item From the point of view of~$\Sh(Y)$, $I$ is an injective module.
-\item From the point of view of~$\Sh(X)$, $I$ is an injective module from the
-point of view of~$\Sh(I(Y))$.
-\item From the point of view of~$\Sh(X)$, $I$ is an injective sheaf of modules
-on~$I(Y)$.
-\end{enumerate}
-\end{lemma}
-
-\begin{proof}The equivalence of the first two statements is by
-Theorem~\ref{thm:injectivity-external-internal}. The equivalence~$\text{(2)}
-\Leftrightarrow \text{(3)}$ is by the idempotency of the stack semantics:
-$\Sh(Y) \models \varphi$ if and only if~$\Sh(X) \models (\Sh(I(Y)) \models
-\varphi)$. (Shulman stated and proved a restricted version of this idempotency property
-in his original paper on the stack semantics~\cite[Lemma~7.20]{shulman:stack-semantics}.
-A proof of the general case is slightly less
-accessible~\cite[Lemma~1.20]{blechschmidt:master}.) The equivalence~$\text{(3)}
-\Leftrightarrow \text{(4)}$ is by interpreting
-Theorem~\ref{thm:injectivity-external-internal} internally to~$\Sh(X)$. This
-requires Zorn's lemma to hold internally to~$\Sh(X)$; this is indeed the case
-since we assume Zorn's lemma in the metatheory and since the validity of Zorn's
-lemma passes from the metatheory to localic
-toposes~\cite[Proposition~D4.5.14]{johnstone:elephant}.
-\end{proof}
-
-\begin{thm}\label{thm:higher-direct-images-as-internal-sheaf-cohomology}
-Let~$f : Y \to X$ be a ringed locale over a locale~$X$. Let~$E$ be a sheaf of modules over~$Y$.
-Assuming the axiom of choice in the metatheory, the expression~``$H^n(I(Y), E)$''
-of the internal language of~$\Sh(X)$ denotes a globally-defined sheaf
-on~$X$, and this sheaf coincides with~$R^n f_*(E)$.\end{thm}
-
-\begin{proof}By Lemma~\ref{lemma:notions-of-internal-injectivity} and by the
-fact that every sheaf of modules over~$Y$ admits an injective resolution, every sheaf
-of modules over~$I(Y)$ admits an injective resolution from the point of view
-of~$\Sh(X)$. Hence we can, internally to~$\Sh(X)$, carry out the construction of~$H^n(I(Y), E)$.
-Externally, this yields an open covering of~$X$ such that we have, for each
-member~$U$ of that covering
-\begin{itemize}
-\item a sheaf~$M$ on~$U$,
-\item a module structure on~$M$,
-\item a resolution~$0 \to E|_{f^{-1}U} \to I^\bullet$ by sheaves of modules which are
-internally and hence externally injective and
-\item data exhibiting~$M$ as the~$n$-th cohomology
-of~$(f|_{f^{-1}U})_*(I^\bullet)$.
-\end{itemize}
-On intersections of such opens~$U$ and~$U'$, there is exactly one
-isomorphism~$M|_{U \cap U'} \to M'|_{U \cap U'}$ of sheaves of modules induced by
-a morphism of resolutions which is compatible with the identity on~$E$.
-Hence the cocycle condition for these isomorphisms is satisfied, ensuring that
-the individual sheaves~$M$ glue to a globally-defined sheaf of modules on~$X$.
-(The individual injective resolutions need not glue to a global injective
-resolution.)
-
-The claim that this sheaf coincides with~$R^n f_*(E)$ follows from the fact
-that we can pick as internal resolution of~$E$ (considered as a sheaf
-over~$I(Y)$), the particular injective resolution of~$E$ (considered as a sheaf
-over~$Y$) used to define~$R^n f_*(E)$.
-\end{proof}
-
-The internal characterization provided by
-Theorem~\ref{thm:higher-direct-images-as-internal-sheaf-cohomology} gives, as a
-simple application, a logical explanation that higher direct images along
-the identity~$\mathrm{id} : X \to X$ vanish: From the internal point of view
-of~$\Sh(X)$, the over-locale~$X$ corresponds to the one-point locale, and the
-higher cohomology of the one-point locale vanishes.
-
-In algebraic geometry, the internal characterization can be used to immediately
-deduce the explicit description of the higher direct images of Serre's twisting
-sheaves along the projection~$\mathbb{P}^n_S \to S$, where~$S$ is an arbitrary
-base scheme (or even base locally ringed locale), from a computation of the
-cohomology of projective~$n$-space. Background on carrying out scheme
-theory internally to a topos, is given in~\cite[Section~12]{blechschmidt:phd}.
-
-\section{Flabby objects in the \effective topos}
-\label{sect:in-eff}
-
-The notion of flabby objects originates from the notion of flabby shaves and is
-therefore closely connected to Grothendieck toposes. Hence it is instructive
-to study flabby objects in elementary toposes which are not Grothendieck
-toposes, away from their original conceptual home. We begin this study with
-establishing the following observations on flabby objects in the \effective
-topos. We follow the terminology of Martin Hyland's survey on the \effective
-topos~\cite{hyland:effective-topos}.
-
-\begin{prop}\label{prop:flabby-effective-sets}
-Let~$X$ be a flabby object in the \effective topos. Let~$f : X \to X$
-be a morphism. If~$X$ is \effective, the statement ``$f$ has a fixed point''
-holds in the \effective topos.
-\end{prop}
-
-\begin{prop}\label{prop:semienough-flabby-modules}
-Assuming the law of excluded middle in the metatheory, any~$\neg\neg$-separated
-module in the \effective topos embeds into a flabby module.
-\end{prop}
-
-% Δ(X) ist welk, falls X bewohnt.
-% Δ(X) ist separiert.
-% Ist A separierter Modul, so lässt sich A in einen welken Modul einbetten
-% (der seinerseits separiert ist, er ist nämlich Δ(Γ(A))).
-% Sind welke Mengen separiert? Nein: Omega.
-% Sind welke Moduln separiert? Nein: Ich denke mittlerweile, dass sich jeder
-% Modul in einen welken einbetten lässt, über etwas, was wie eine induktive
-% Konstruktion aussieht.
-
-% In Eff gilt: Jede negneg-separierte abelsche Gruppe bettet in eine injektive
-% ein. Nämlich A --> ΔΓA --> ΔI für ΓA --> I Einbettung in eine injektive
-% Gruppe.
-
-The intuitive
-reason for why Proposition~\ref{prop:flabby-effective-sets} holds is the
-following. Let~$X$ be a flabby object in the \effective topos. Then there is a
-procedure which computes for any subterminal~$K \subseteq X$ an element~$x_K$
-such that~$K \subseteq \{ x_K \}$. This element might not depend extensionally
-on~$K$, but this fine point is not important for this discussion. Let
-now~$f : X \to X$ be a morphism. We construct the self-referential subset~$K
-\defeq \{ f(x_K) \}$; the formal proof below will indicate how this can be
-done. Then~$K \subseteq \{ x_K \}$, so~$f(x_K) = x_K$.
-
-A corollary of Proposition~\ref{prop:flabby-effective-sets} is that the trivial
-module is the only flabby module in the \effective topos whose underlying
-unstructured object is an \effective set: Given such a flabby module~$M$, let~$v
-\in M$ be an arbitrary element. Then the morphism~$x \mapsto v + x$ has a fixed
-point; thus~$v + x = x$ for some element~$x$, and hence~$v = 0$.
-
-It is the self-referentiality which makes the proof of
-Proposition~\ref{prop:flabby-effective-sets} work, but the blame for paucity
-of flabby objects in the \effective topos is to put on the realizers for
-statements of the form~``$K = K$'', where~$({=})$ is the nonstandard equality
-predicate of the powerobject~$\P(X)$. A procedure witnessing flabbiness has to
-compute a reflexivity realizer for a suitable element~$x_K$ from a reflexivity
-realizer for a given element~$K$. However, such realizers are not very
-informative. Metaphorically speaking, a procedure witnessing flabbiness has to
-conjure elements out of thin air.
-
-This problem does not manifest with objects~$X$ which are not \effective sets.
-Reflexivity realizers for these objects are themselves not very informative;
-a procedure witnessing flabbiness therefore only has to turn one kind of
-non-informative realizers into another kind. The flabby modules featuring in
-the proof of Proposition~\ref{prop:semienough-flabby-modules} will accordingly
-not be \effective sets.
-
-\begin{proof}[Proof of Proposition~\ref{prop:flabby-effective-sets}]
-For any Turing machine~$e$, let~$v_e : |X| \to \Sigma$ be the nonstandard
-predicate given by
-\begin{multline*}
-  v_e(x) = \{ m \in \NN \,|\,
-  \text{there is an element~$x_0 \in |X|$ such that} \\
-  \text{$e$ terminates with an element of~$\llbracket x_0 = x_0 \rrbracket$ and
-  $m \in \llbracket x = x_0 \rrbracket$} \}
-\end{multline*}
-and let~$K_e \in \Sigma^{|X| \times \Sigma}$ be the nonstandard predicate given by
-\[ K_e(x,u) = \llbracket (x = x) \wedge (u \leftrightarrow v_e(x)) \rrbracket. \]
-One can explicitly construct a realizer~$a_e$ of the statement~``$K_e = K_e$'',
-where~$({=})$ is the nonstandard equality predicate of the object~$\P_{\leq1}(X)$
-of subterminals of~$X$. This is where the assumption that~$X$ is \effective is
-important; without it, we could only verify~``$K_e = K_e$'' where~$({=})$ is
-the nonstandard equality predicate of the full powerobject~$\P(X)$.
-
-Since~$X$ is flabby, there is a realizer~$r$ for the statement~``$\forall K \in
-\P_{\leq1}(X)\_ \exists x \in X\_ \forall y \in X\_ (y \in K \Rightarrow y =
-x)$''. Let~$s$ be a realizer for the statement~``$\forall x \in X\_ \exists y
-\in Y\_ y = f(x)$''. Let~$e$ be the particular Turing machine which proceeds as
-follows:
-\begin{enumerate}
-\item[1.] Simulate~$r$ on input~$a_e$ in order to obtain a realizer~$b \in
-\llbracket x = x \rrbracket$ for some~$x \in |X|$.
-\item[2.] Simulate~$s$ on input~$b$ in order to obtain a realizer~$c \in \llbracket
-f(x) = f(x) \rrbracket$.
-\item[3.] Output~$c$.
-\end{enumerate}
-The description of the machine~$e$ makes use of the number~$e$ coding it;
-the recursion theorem yields a general reason why this self-referentiality is
-possible. Here we can even do without this theorem, since a close inspection of
-the construction of~$a_e$ shows that~$a_e$ is actually independent of~$e$. This
-should not come as a surprise, as reflexivity realizers of~$\P(X)$
-and~$\P_{\leq1}(X)$ are known to be not very informative.
-
-Passing~$a_e$ to~$r$ yields a reflexivity realizer of some element~$x_{K_e} \in
-|X|$. Therefore the Turing machine~$e$ does terminate, with a reflexivity
-realizer for~$f(x_{K_e})$. Thus the statement~``$f(x_{K_e}) \in K_e$'' is
-realized; hence~``$f(x_{K_e}) = x_{K_e}$'' is as well.
-\end{proof}
-
-\begin{proof}[Proof of Proposition~\ref{prop:semienough-flabby-modules}]
-Let~$(\Gamma \dashv \Delta) : \Set \to \Eff$ be the inclusion of the
-double-negation sheaves. For a~$\neg\neg$-separated module~$M$ in the \effective
-topos, the canonical morphism~$M \to \Delta(\Gamma(M))$ is a monomorphism; the
-set~$\Gamma(M)$ is flabby by virtue of being inhabited; and~$\Delta$ preserves
-flabby objects by Proposition~\ref{prop:pushforward-of-flabby-objects}.
-\end{proof}
-
-\begin{rem}The analogues of Proposition~\ref{prop:flabby-effective-sets} and
-Proposition~\ref{prop:semienough-flabby-modules} for the realizability topos
-constructed using infinite time Turing
-machines~\cite{bauer:injection,hamkins-lewis:ittm} are true as well, with the
-same proofs.\end{rem}
-
-\section{Conclusion}
-\label{sect:conclusion}
-
-We originally set out to develop an intuitionistic account of Grothendieck's
-sheaf cohomology. Čech methods can be carried out constructively, and
-there are constructive accounts of special cases, resulting even in
-efficient-in-practice algorithms~\cite{barakat-lh:homalg,barakat-lh:ext},
-but it appears that there is not a general framework for sheaf
-cohomology which works in an intuitionistic metatheory.
-
-The main obstacle preventing Grothendieck's theory of derived functors to be interpreted
-constructively is its reliance on injective resolutions. It is known that in
-the absence of the axiom of choice, much less in a purely intuitionistic
-context, there might not be any nontrivial injective abelian
-group~\cite{blass:inj-proj-axc}.
-
-In principle, this problem could be remedied by employing flabby resolutions
-instead of injective ones. There are, however, two problems with this
-suggestion. Firstly, all proofs known to us that flabby sheaves are
-acyclic for the global sections functor require Zorn's lemma. This problem might
-be mitigated by relying on the substitute property discussed
-following Scholium~\ref{scholium:exact-as-presheaves}.
-But secondly, it is an open question whether
-one can show, purely intuitionistically, that any sheaf of modules embeds into
-a flabby sheaf of modules. The following is known about this problem:
-
-\begin{enumerate}
-\item There is a purely intuitionistic proof that any sheaf of sets embeds into
-a flabby sheaf of sets
-(Scholium~\ref{scholium:properties-of-flabby-objects}(4)).
-
-\item The existence of enough flabby modules, and even the existence of enough
-injective modules, is \emph{not} a constructive taboo, that is, these statements do not
-entail a classical principle like the law of excluded middle or the principle
-of omniscience. This is because assuming the axiom of choice, any
-Grothendieck topos has enough injective (and therefore flabby) modules.
-
-%\item There is a way of embedding any module into a flabby module if one is
-%prepared to ignore set-theoretical difficulties. Namely, let~$M$ be an~$R$-module.
-%Inductively construct a collection~$T$ of terms by the following clauses: $0
-%\in T$ (where~$0$ is a formal symbol);
-%if~$t,s \in T$, then~$t + s \in T$; if~$t \in T$ and~$r \in R$, then~$rt \in
-%T$; if~$x \in M$, then~$\underline{x} \in T$; if~$K \subseteq T$ is a
-%subterminal, then~$\varepsilon_K \in T$. Let~$({\sim})$ be
-%the finest equivalence relation on~$T$ such that~$t + (s + u) \sim (t + s) +
-%u$, $t + s \sim s + t$, $t + 0 \sim t \sim 0 + t$, $0t \sim 0$, $1t \sim t$, $r(t+s) \sim rt
-%+ rs$, $(r+r')t \sim rt + r't$ and such
-%that~$\underline{0} \sim 0$, $\underline{x+y} \sim \underline{x} + \underline{y}$, $\underline{rx} \sim
-%r \underline{x}$, and such that~$\varepsilon_{\{t\}} \sim
-%t$ for all terms~$t$. Then the quotient~$T/{\sim}$ is a flabby~$R$-module
-%into which~$M$ embeds. However, it is not clear that~$T/{\sim}$ is a set.
-
-\item There is a way of embedding any module into a flabby module if
-\emph{quotient inductive types}, as suggested by Altenkirch and
-Kaposi~\cite{altenkirch-kaposi:qits}, are available. These generalize ordinary
-inductive~$W$-types, which exist in any
-topos~\cite{moerdijk-palmgren:wellfounded-trees,berg-moerdijk:w-types-in-sheaves,berg-kouwenhoven-gentil:w-types-in-eff}
-and whose existence
-can indeed be verified in an intuitionistic set theory like IZF, by allowing to
-give constructors and state identifications at the same time. More
-specifically, given an~$R$-module~$M$, we can construct a flabby envelope~$T$
-of~$M$ as the quotient inductive type generated by the following clauses:~$0
-\in T$ (where~$0$ is a formal symbol); if~$t,s \in T$, then~$t + s \in T$;
-if~$t \in T$ and~$r \in R$, then~$rt \in T$; if~$x \in M$, then~$\underline{x}
-\in T$; if~$K \subseteq T$ is a subterminal, then~$\varepsilon_K \in T$;
-if~$t,s,u \in T$ and~$r,r' \in R$, then~$t + (s + u) = (t + s) + u$, $t + s = s + t$, $t + 0 = t
-= t + 0$, $0t = 0$, $1t = t$, $r(t+u) = rt + ru$, $(r+r')t = rt + r't$; if~$x,y
-\in M$ and~$r \in R$, then~$\underline{0} = 0$, $\underline{x + y} =
-\underline{x} + \underline{y}$, $\underline{rx} = r \underline{x}$; and if~$t
-\in M$, then~$\varepsilon_{\{t\}} = t$.
-
-However, it is an open question under which circumstances quotient inductive
-types can be shown to exist. Zermelo--Fraenkel with choice certainly suffices,
-while Zermelo--Fraenkel without choice does not~\cite[Section~9]{shulman-lumsdaine:hits},
-hence IZF also does not.\footnote{With quotient inductive types, any infinitary
-algebraic theory admits free algebras. However, it is consistent with
-Zermelo--Fraenkel set theory that some such theories do not admit free
-algebras~\cite{blass:free-algebras}.} The existence of quotient inductive types
-seem to be, as the existence of enough injective modules, \emph{constructively
-neutral}.
-
-\item There is a number of simple constructions which come close to providing
-flabby envelopes for arbitrary modules~$M$. For instance, we could equip the set~$T
-\defeq \P_{\leq1}(X)/{\sim}$, where~$K \sim L$ if and only if~$K = L$ or~$K
-\cup L \subseteq \{0\}$, with a module structure given by~$0 \defeq [\{0\}]$,
-$[K]+[L] \defeq [K+L]$ and~$r [K] \defeq rK$. The resulting module admits a
-linear injection from~$M$, sending an element~$x$ to~$[\{x\}]$. However, it
-fails to be flabby. Given a subterminal~$E \subseteq \P_{\leq1}(X)/{\sim}$,
-there is the well-defined element~$v \defeq [\{x \in M\,|\,\text{$x \in K$ for
-some~$[K] \in E$}\}]$, but we cannot verify~$E \subseteq \{v\}$.
-
-\item There appears to be some tension regarding the \effective topos:
-Proposition~\ref{prop:semienough-flabby-modules} shows that at
-least~$\neg\neg$-separated modules in the \effective topos always embed into
-flabby modules, assuming the law of excluded middle in the metatheory, while
-Proposition~\ref{prop:flabby-effective-sets} shows that no nontrivial \effective
-module is flabby.
-\end{enumerate}
-
-We currently believe that it is not possible to give a constructive account of
-a global cohomology functor which would associate to any sheaf of modules its
-cohomology. However, it should be possible to do so for a restricted class of
-sheaves, while still preserving the good formal properties expected from
-derived functors.
 
 \printbibliography
-\enlargethispage{1em}
 
 \end{document}
+
+Outline:
+
+Stuff that should be mentioned:
+
+* Andrej's realizability in the real world
+* Cauchy vs. Dedekind numbers (physical quantities, ...)
+* Syntactical vs. semantical interpretation
+* phone call analogy
+* detailed explanation of the pretty picture; analogy with "inner models"
+  of set theory
+* different intention with the alternative universes in comparison to set theory
+* quick overview of the several aspects of toposes (maybe at the end,
+  as an outlook?)
+* enrichment of platonism debate (find better term for this!)
+* uncovering further relations between objects
+* allowing a switch of perspective
+* applications in mathematical practice
+* int. logic as common denominator
+* arbitrariness of the "standard axioms"
+* models of ZF yield toposes (models of ETCS), including a quick discussion of
+  equivalence
+* what giving up classical logic actually amounts to in practice
+* examples in Eff
+* examples in Sh(X)
+* examples in Spec(A), "reifying all the individual localizations into a
+  single coherent entity which can be reasoned about as if it were a single
+  ring"
+* importance of having an adapted language
+* SDG (infinitesimals are well-studied in philosophy of mathematics)
+* beautiful and intriguing fact: laws of logic apply to mathematical objects
+  not only on the face, but also in different ways
+* quick remark on the internal language being based on types instead of sets

### 2018-10-30 paper-filmat.tex
First steps in the effective topos

@@
 \usepackage{amsthm,mathtools,stmaryrd}
+\usepackage{booktabs}
 \usepackage[all]{xy}
-%\usepackage[protrusion=true,expansion=true]{microtype}
-\usepackage[hyphens]{url}
-\usepackage[breaklinks=true]{hyperref}
+\usepackage[protrusion=true,expansion=true]{microtype}
 \usepackage{xspace}
@@
 \usepackage[babel]{csquotes}
-\bibliography{paper-flabby-objects.bib}
+\bibliography{paper-filmat.bib}
 
-\title{Exploring mathematical objects from custom-tailored mathematical universes}
+\usepackage{pifont}
+\newcommand{\cmark}{\ding{51}}
+\newcommand{\xmark}{\ding{55}}
+
+\title[]{Exploring mathematical objects from custom-tailored mathematical universes}
 \author{Ingo Blechschmidt}
@@
 \newcommand{\effective}{ef{}fective\xspace}
+\newcommand{\?}{\,{:}\,}
 
@@
 \noindent
+
+\section{Exploring the \effective topos}
+
+% XXX introduction
+
+\begin{tabular}{lll}
+  \toprule
+  Statement & in $\Set$ & in $\Eff$ \\
+  \midrule
+  Any natural number is prime or not prime. & \cmark{} (trivially so) & \cmark \\
+  There are infinitely many primes. & \cmark & \cmark \\
+  Any function $\NN \to \NN$ is the zero function or not. & \cmark{} (trivially so) & \xmark \\
+  Any function $\NN \to \NN$ is computable by a Turing machine. & \xmark & \cmark{} (trivially so) \\
+  Any function $\RR \to \RR$ is continuous. & \xmark & \cmark \\
+  \bottomrule
+\end{tabular}
+
+\subsection*{``Any natural number is prime or not.''} Even without knowing what
+a prime number is, one can safely judge this statement to be true in
+the standard topos, since it is just an instance of the law of excluded middle.
+
+By the Kripke--Joyal semantics, saying that it's true in the \effective topos
+amounts to saying that there is a Turing machine which, given a natural
+number~$n$ as input, terminates with a correct judgment whether~$n$ is prime or
+not. Such a Turing machine indeed exists -- writing such a program is often a
+first exercise in programming courses. Hence the statement is also true in the
+\effective topos, but for the nontrivial reason that such a machine exists.
+
+\subsection*{``There are infinitely many primes.''} A first-order formalization
+of this statement is ``for any natural number~$n$, there is a natural
+number~$p$ which is greater than~$n$'', and is known to be true in the standard
+topos by any of the many proofs of this fact.
+
+Its external meaning when interpreted in the \effective topos is that there exists
+a Turing machine which, given a natural number~$n$ as input, terminates with a
+prime number~$p > n$ as output. Such a Turing machine exists, hence the
+statement is true in the \effective topos.
+
+\subsection*{``Any function~$\NN \to \NN$ is the zero function or not.''} More
+formally, the statement is
+\[ \forall f \? \NN^\NN\_
+  \bigl((\forall n \? \NN\_ f(n) = 0) \vee
+  \neg
+  (\forall n \? \NN\_ f(n) = 0)\bigr). \]
+By the law of excluded middle, this statement is trivially true in the standard
+topos.
+
+Its meaning when interpreted in the \effective topos is that there exists a
+Turing machine~$M$ which, given the description of a Turing machine~$F$ which
+computes a function~$f : \NN \to \NN$ as input, terminates with a correct
+judgment of whether~$f$ is the zero function or not. Such a machine~$M$ does
+not exist, hence the statement is false in the \effective topos.
+
+A formal proof that such a machine~$M$ does not exist will reduce its assumed
+existence to the undecidability of the halting problem. Intuitively, the issue
+is the following. Turing machines are able to simulate other Turing machines.
+Hence~$M$ could simulate~$F$ on various inputs to search the list of
+function values~$f(0), f(1), \ldots$ for a nonzero number. In case that after
+a certain number of steps a nonzero function value is found, the machine~$M$
+can correctly output the judgment that~$f$ is not the zero function. But if the
+search only turned up zero values, it cannot come to any verdict -- it cannot
+rule out that a nonzero function value will show up in the as yet unexplored
+part of the function.
+
+\subsection*{``Any function~$\NN \to \NN$ is computable by a Turing machine.''}
+The preceding examples could give the impression that what is true in the
+\effective topos is simply a subset of what is true in the standard topos. This
+statement shows that the relation between these two toposes is more nuanced.
+
+The fundamental observation of computability theory is that, in the standard
+topos, there are functions~$\NN \to \NN$ which are not computable by a Turing
+machine. Explicit examples include the \emph{halting
+function}, which maps a number~$n$ to zero or one depending on whether
+the~$n$'th Turing machine (in some fixed enumeration of all Turing machines)
+terminates or not, and the \emph{busy beaver function}. Cardinality arguments
+even show that most functions~$\NN \to \NN$ are not computable: There
+are~$\aleph_0^{\aleph_0} = 2^{\aleph_0}$ functions~$\NN \to \NN$, but
+only~$\aleph_0$ Turing machines and hence only~$\aleph_0$ functions which are
+computable by a Turing machine.
+
+In contrast, in the \effective topos, any function~$\NN \to \NN$ is computable
+by a Turing machine: The external meaning of this internal statement is that
+there exists a Turing machine~$M$ which, given a description of a Turing
+machine~$F$ computing a function~$f : \NN \to \NN$, outputs a description of a
+Turing machine computing~$f$. It is trivial to program such a machine~$M$; the
+machine~$M$ simply has to echo its input back to the caller.
+
+To avert a paradox, we should point out where the proof of the fundamental
+observation of computability theory employs nonconstructive reasoning, for if
+it would admit a constructive proof, it would also hold internally to the
+\effective topos, in contradiction to the fact that it does not. The halting
+function~$h : \NN \to \NN$, defined using the case distinction
+\[ h : n \mapsto \begin{cases}
+  1, & \text{if the $n$'th Turing machine terminates}, \\
+  0, & \text{if the $n$'th Turing machine does not terminate},
+\end{cases} \]
+cannot be given as a counterexample in the \effective topos since, in the
+\effective topos, it is not actually a total function from~$\NN$ to~$\NN$. It
+is only defined on those numbers~$n$ for which the~$n$'th Turing machine
+terminates or does not terminate. Assuming the law of excluded middle, this is
+a trivial condition; but intuitionistically, it is not. The definition of the
+busy beaver function requires a similar case distinction and therefore also
+does not give rise to a well-defined counterexample within the \effective
+topos.
+
+\subsection*{``Any function~$\RR \to \RR$ is continuous.''} In the standard
+topos, this statement is plainly false, with the signum and Heaviside functions
+being prominent counterexamples. In the \effective topos, this statement is
+true. A formal proof is not entirely straightforward~\cite{XXX}; the intuitive
+reason for why this is so is as follows.
+
+What the \effective topos believes to be a real number is, from the external
+point of view, a Turing machine~$X$ which outputs, when called with a natural
+number~$n$ as input, a rational approximation~$X(n)$. These approximations are
+required to be \emph{consistent} in the sense that~$|X(n) - X(m)|
+\leq 1/(n+1) + 1/(m+1)$. Intuitively, such a machine~$X$ denotes the real
+number~$\lim_{n \to \infty} X(n)$, and the approximations~$X(n)$ must be
+within~$1/(n+1)$ of the limit.
+
+A function~$f : \RR \to \RR$ in the \effective topos is therefore given by a
+Turing machine~$M$ which, given the description of such a Turing machine~$X$ as
+input, outputs the description of a similar such Turing machine~$Y$ as output.
+To compute a rational approximation~$Y(n)$, the machine~$Y$ may simulate~$X$
+and can therefore determine arbitrarily many rational approximations~$X(m)$.
+However, within finite time, the machine~$Y$ can only acquire finitely many
+such approximations. Hence a function such as the signum function, for which
+even rough rational approximations of~$\sgn(x)$ require infinite precision in
+the input~$x$, do not exist in the \effective topos.
+
+% XXX formal rules
 
@@
 * examples in Eff
+* examples in Eff(STM)
 * examples in Sh(X)
@@
 * quick remark on the internal language being based on types instead of sets
+* reference to Bohr topos approach
+* Source https://plato.stanford.edu/entries/category-theory/ for relevant
+  literatur
+* the internal language as a generalized modal operator
+
+Send to Georg.

### 2018-10-30 paper-filmat.tex
Formal translation rules

@@
 \usepackage[utf8]{inputenc}
-\usepackage{amsthm,mathtools,stmaryrd}
+\usepackage{amsthm,mathtools,stmaryrd,amssymb}
 \usepackage{booktabs}
@@
 \newcommand{\?}{\,{:}\,}
+\newcommand{\realizes}{\Vdash}
 
@@
 such approximations. Hence a function such as the signum function, for which
-even rough rational approximations of~$\sgn(x)$ require infinite precision in
-the input~$x$, do not exist in the \effective topos.
+even rough rational approximations of~$\operatorname{sgn}(x)$ require infinite
+precision in the input~$x$, do not exist in the \effective topos.
 
-% XXX formal rules
+\subsection*{The formal translation rules} The previous examples were picked to
+convey an intuitive understanding of what statements in the \effective topos
+externally mean, and to showcase a variety of different situations. The
+formal translation rules are given in Table~\ref{table:eff}.
+
+\begin{table}
+  \begin{tabbing}
+    $\Eff \models (\forall n\?\NN\_ \varphi(n))\!\!\!$ \= \kill
+    $\Eff \models \varphi$ \> iff there is a Turing machine~$M$ such that~$M
+    \realizes \varphi$. \\\\
+    $M \realizes \top$ \> is true for any machine~$M$. \\
+    $M \realizes \bot$ \> is false for any machine~$M$. \\
+    $M \realizes (\varphi \wedge \psi)$ \> iff~$M$ terminates on inputs~$0$
+    and~$1$ such that $M(0) \realizes \varphi$ and~$M(1) \realizes \psi$. \\
+    $M \realizes (\varphi \vee \psi)$ \> iff~$M$ terminates on inputs~$0$ and~$1$
+    such that, \\ \> \qquad if~$M(0) = 0$ then~$M(1) \realizes \varphi$, and if~$M(0) \neq
+    0$ then~$M(1) \realizes \psi$. \\
+    $M \realizes (\varphi \Rightarrow \psi)$ \> iff for any Turing machine~$E$
+    such that~$E \realizes \varphi$, \\ \> \qquad $M$ terminates on input~$E$ such
+    that~$M(E) \realizes \psi$. \\
+    $M \realizes (\forall n\?\NN\_ \varphi(n))$ \> iff for any natural number~$n_0
+    \in \NN$, \\ \> \qquad $M$ terminates on input~$n_0$ such that~$M(n_0)
+    \realizes \varphi(n_0)$. \\
+    $M \realizes (\exists n\?\NN\_ \varphi(n))$ \> iff~$M$ terminates on
+    inputs~$0$ and~$1$ such that $M(1) \realizes \varphi(M(0))$.
+  \end{tabbing}
+
+  \caption{\label{table:eff} The (first-order fragment of) the translation
+  rules defining the meaning of statements internal to the \effective topos.}
+\end{table}
 

### 2018-10-31 paper-filmat.tex
typo
Some higher-order rules

--- 2018-10-30 paper-filmat.tex
@@
 % XXX introduction
+% XXX mention connection to realizability
 
@@
 function}, which maps a number~$n$ to zero or one depending on whether
-the~$n$'th Turing machine (in some fixed enumeration of all Turing machines)
+the~$n$-th Turing machine (in some fixed enumeration of all Turing machines)
 terminates or not, and the \emph{busy beaver function}. Cardinality arguments
@@
 \[ h : n \mapsto \begin{cases}
-  1, & \text{if the $n$'th Turing machine terminates}, \\
-  0, & \text{if the $n$'th Turing machine does not terminate},
+  1, & \text{if the $n$-th Turing machine terminates}, \\
+  0, & \text{if the $n$-th Turing machine does not terminate},
 \end{cases} \]
@@
 \effective topos, it is not actually a total function from~$\NN$ to~$\NN$. It
-is only defined on those numbers~$n$ for which the~$n$'th Turing machine
+is only defined on those numbers~$n$ for which the~$n$-th Turing machine
 terminates or does not terminate. Assuming the law of excluded middle, this is
@@
 such approximations. Hence a function such as the signum function, for which
-even rough rational approximations of~$\sgn(x)$ require infinite precision in
-the input~$x$, do not exist in the \effective topos.
+even rough rational approximations of~$\operatorname{sgn}(x)$ require infinite
+precision in the input~$x$, do not exist in the \effective topos.
 
-% XXX formal rules
+\subsection*{The formal translation rules} The previous examples were picked to
+convey an intuitive understanding of what statements in the \effective topos
+externally mean, and to showcase a variety of different situations. The
+formal translation rules are given in Table~\ref{table:eff}.
+
+\begin{table}
+  \begin{tabbing}
+    $\Eff \models (\forall f\?\NN^\NN \varphi(n))\!\!$ \= \kill
+    $\Eff \models \varphi$ \> iff there is a natural number~$e$ such that~$e
+    \realizes \varphi$. \\\\
+    $e \realizes \top$ \> is true for any number~$e$. \\
+    $e \realizes \bot$ \> is false for any number~$e$. \\
+    $e \realizes (\varphi \wedge \psi)$ \> iff~$e \cdot 0 \downarrow$ and~$e
+    \cdot 1 \downarrow$ and $e\cdot0 \realizes \varphi$ and~$e\cdot1 \realizes \psi$. \\
+    $e \realizes (\varphi \vee \psi)$ \> iff~$e \cdot 0 \downarrow$ and~$e
+    \cdot 1 \downarrow$ and \\ \> \qquad if~$e\cdot0 = 0$ then~$e\cdot1 \realizes
+    \varphi$, and \\ \> \qquad if~$e\cdot0 \neq 0$ then~$e\cdot1 \realizes \psi$. \\
+    $e \realizes (\varphi \Rightarrow \psi)$ \> iff for any number~$r$
+    such that~$r \realizes \varphi$, $e \cdot r \downarrow$ and~$e \cdot r \realizes \psi$. \\
+    $e \realizes (\forall n\?\NN\_ \varphi(n))$ \> iff for any natural number~$n_0
+    \in \NN$, $e \cdot n_0 \downarrow$ and~$e \cdot n_0 \realizes \varphi(n_0)$. \\
+    $e \realizes (\exists n\?\NN\_ \varphi(n))$ \> iff~$e\cdot0 \downarrow$ and~$e\cdot1 \downarrow$
+    terminate and~$e\cdot1 \realizes \varphi(e\cdot0)$. \\
+    $e \realizes (\forall f\?\NN^\NN\_ \varphi(f))$ \> iff for any function~$f_0
+    : \NN \to \NN$ and any number~$r_0$ such that \\ \> \qquad $f_0$ is computed by the~$r_0$-th
+    Turing machine, \\ \> \qquad
+    $e \cdot r_0 \downarrow$ and~$e \cdot r_0 \realizes \varphi(f_0)$. \\
+    $e \realizes (\exists f\?\NN\_ \varphi(f))$ \> iff~$e \cdot 0 \downarrow$
+    and~$e \cdot 1 \downarrow$ such that
+    the $(e \cdot 0)$-th Turing machine \\ \> \qquad computes a function~$f_0 : \NN \to \NN$
+    and $e \cdot 1 \realizes \varphi(f_0)$.
+  \end{tabbing}
+
+  \caption{\label{table:eff} A (fragment of) the translation
+  rules defining the meaning of statements internal to the \effective topos.
+  We write~``$e \cdot n \downarrow$'' to mean that calling the~$e$-th Turing
+  machine on input~$n$ terminates, and in this case denote the result by~``$e
+  \cdot n$''.}
+\end{table}
 

### 2018-11-01 paper-filmat.tex
On the internal language of sheaf toposes

@@
 \usepackage[utf8]{inputenc}
-\usepackage{amsthm,mathtools,stmaryrd,amssymb}
+\usepackage{amsthm,mathtools,stmaryrd,amssymb,graphicx}
 \usepackage{booktabs}
@@
 \bibliography{paper-filmat.bib}
+
+\graphicspath{{images/}}
 
@@
 \newcommand{\realizes}{\Vdash}
+\newcommand{\notnot}{\emph{not~not}\xspace}
 
@@
 
+\section{Exploring toposes of sheaves}
+
+Associated to any topological space~$X$ (such as Euclidean space), there is the
+\emph{topos of sheaves over~$X$}, $\Sh(X)$. To a first approximation, a
+statement is true in~$\Sh(X)$ if and only if it ``holds locally on~$X$'';
+what~$\Sh(X)$ believes to be a set is a ``continuous family of sets, one set
+for each point of~$X$''. The precise rules of the Kripke--Joyal semantics
+of~$\Sh(X)$ are listed in Table~\ref{table:sheaf}.
+
+\subsection{A geometric interpretation of double negation}
+In intuitionistic logic, the double negation~$\neg\neg\varphi$ of a
+statement~$\varphi$ is a slight weakening of~$\varphi$; while~$(\varphi
+\Rightarrow \neg\neg\varphi)$ is an intuitionistic tautology, the converse can
+only be shown for some specific statements. The internal language of~$\Sh(X)$
+gives geometric meaning to this logical peculiarity.
+
+Namely, one can show that~$\Sh(X) \models \neg\neg\varphi$ is equivalent to the
+existence of a \emph{dense open}~$U$ of~$X$ such that~$U \models \varphi$.
+If~$\Sh(X) \models \varphi$, that is if~$X \models \varphi$, then there
+obviously exists such a dense open, namely~$X$ itself; however the converse
+usually fails.
+
+The only case that the law of excluded middle does hold internally to~$\Sh(X)$
+is when the only dense open of~$X$ is~$X$ itself; assuming classical logic in
+the metatheory, this holds if and only if every open is also closed. This is
+for instance satisfied if~$X$ is discrete.
+
+An important special case is when~$X$ is the one-point space. In this
+case~$\Sh(X)$ is equivalent (as categories and hence toposes) to the standard
+topos. If mathematics within~$\Sh(X)$ can be described as ``mathematics
+over~$X$'', then this observation justifies saying that ``ordinary mathematics
+is mathematics over the point''.
+
+\subsection{Real numbers}
+As detailled in Section~\ref{sect:eff-continuous}, what the \effective topos believes to be a real number is
+actually a Turing machine computing arbitrarily-good consistent rational
+approximations. A similarly drastic shift in meaning, though in an orthogonal
+direction, occurs with~$\Sh(X)$. What~$\Sh(X)$ believes to be a (Dedekind) real
+number~$a$ is actually a continuous function~$a : X \to \RR$, that is, a continuous
+family of real numbers on~$X$.
+
+Such a function is everywhere positive on~$X$ if and only if, from the internal point of
+view~$\Sh(X)$, the number~$a$ is positive; it is everywhere zero if and only
+if, internally, the number~$a$ is zero; and it is everywhere negative if and
+only if, internally, the number~$a$ is negative.
+
+The law of trichotomy, stating that any real number is either negative, zero or
+positive, generally fails in~$\Sh(X)$. By the Kripke--Joyal semantics, the external
+meaning of this internal statement is that for any continuous function~$a : U
+\to \RR$ defined on any open~$U$ of~$X$, there is an open covering~$U =
+\bigcup_i U_i$ such that on each member~$U_i$ of this covering, the function~$a$ is
+either everywhere negative on~$U_i$, everywhere zero on~$U_i$ or everywhere
+positive on~$U_i$. But this statement is, for most base spaces~$X$, false.
+Figure~\ref{fig:trichotomy}(c) shows a counterexample.
+
+The weaker statement that for any real number~$a$ it's \notnot the case that~$a < 0$
+or~$a = 0$ or~$a > 0$ does hold in~$\Sh(X)$, for this statement is an
+intuitionistic tautology. Its meaning is that there exists a
+dense open~$U$ such that~$U$ can be covered by opens on which~$a$ is either
+everywhere negative, everywhere zero or everywhere positive. In the example
+given in Figure~\ref{fig:trichotomy}(c), this open~$U$ could be taken as~$X$
+with the unique zero removed.
+
+\begin{table}
+  \begin{tabbing}
+    $U \models (\forall x\?\RR\_ \varphi(x))$ \= \kill
+    $\Sh(X) \models \varphi$ \> iff $X \models \varphi$. \\\\
+    $U \models a = b$ \> iff~$a = b$ on~$U$. \\
+    $U \models \top$ \> is true for any open~$U$. \\
+    $U \models \bot$ \> iff~$U$ is the empty open. \\
+    $U \models (\varphi \wedge \psi)$ \> iff~$U \models \varphi$ and~$U \models \psi$. \\
+    $U \models (\varphi \vee \psi)$ \> iff there is an open covering~$U =
+    \bigcup_i U_i$ such that, \\ \> \qquad for each~$i$, $U_i \models \varphi$
+    or $U_i \models \psi$. \\
+    $U \models (\varphi \Rightarrow \psi)$ \> iff for every open~$V \subseteq U$,
+    $V \models \varphi$ implies~$V \models \psi$. \\
+    $U \models (\forall a\?\RR\_ \varphi(x))$ \> iff for every open~$V
+    \subseteq U$ and any continuous function~$a_0 : V \to \RR$, $V \models
+    \varphi(a_0)$. \\
+    $U \models (\exists a\?\RR\_ \varphi(x))$ \> iff there is an open
+    covering~$U = \bigcup_i U_i$ such that, \\ \> \qquad for each~$i$,
+    there exists a continuous function~$a_0 : U_i \to \RR$ with~$U_i \models
+    \varphi(a_0)$.
+  \end{tabbing}
+
+  \caption{\label{table:sheaf} A (fragment of) the translation rules defining
+  the meaning of statements internal to~$\Sh(X)$, the topos of sheaves over a
+  topological space~$X$.}
+\end{table}
+
+\begin{figure}
+  \includegraphics[height=3cm]{trichotomy-1}
+  \includegraphics[height=3cm]{trichotomy-2}
+  \includegraphics[height=3cm]{trichotomy-3}
+  \caption{\label{fig:trichotomy} Three examples for what the topos~$\Sh(X)$
+  each believes to be a single real number, where the base space~$X$ is the
+  unit interval. (a) A positive real number. (b) A negative real number. (c) A
+  number which is neither negative nor zero nor positive. Externally speaking,
+  no open neighborhood~$V$ of the unique zero of the depicted function~$a$ has
+  the property that~$a$ is everywhere negative, everywhere zero or everywhere
+  positive on~$V$.}
+\end{figure}
+
 \printbibliography

### 2018-11-01 paper-filmat.tex
typo

@@
 
-\subsection*{``Any natural number is prime or not.''} Even without knowing what
+\subsection{``Any natural number is prime or not.''} Even without knowing what
 a prime number is, one can safely judge this statement to be true in
@@
 
-\subsection*{``There are infinitely many primes.''} A first-order formalization
+\subsection{``There are infinitely many primes.''} A first-order formalization
 of this statement is ``for any natural number~$n$, there is a natural
@@
 
-\subsection*{``Any function~$\NN \to \NN$ is the zero function or not.''} More
+\subsection{``Any function~$\NN \to \NN$ is the zero function or not.''} More
 formally, the statement is
@@
 
-\subsection*{``Any function~$\NN \to \NN$ is computable by a Turing machine.''}
+\subsection{``Any function~$\NN \to \NN$ is computable by a Turing machine.''}
 The preceding examples could give the impression that what is true in the
@@
 
-\subsection*{``Any function~$\RR \to \RR$ is continuous.''} In the standard
-topos, this statement is plainly false, with the signum and Heaviside functions
+\subsection{``Any function~$\RR \to \RR$ is continuous.''}\label{sect:eff-continuous}
+In the standard topos, this statement is plainly false, with the signum and Heaviside functions
 being prominent counterexamples. In the \effective topos, this statement is
@@
 
-\subsection*{The formal translation rules} The previous examples were picked to
+\subsection{The formal translation rules} The previous examples were picked to
 convey an intuitive understanding of what statements in the \effective topos
@@
   \begin{tabbing}
-    $\Eff \models (\forall f\?\NN^\NN \varphi(n))\!\!$ \= \kill
+    $e \models (\forall f\?\NN^\NN\_ \varphi(n))$ \= \kill
     $\Eff \models \varphi$ \> iff there is a natural number~$e$ such that~$e
     \realizes \varphi$. \\\\
+    $e \realizes s = t$ \> iff $s = t$. \\
     $e \realizes \top$ \> is true for any number~$e$. \\

### 2018-11-01 paper-filmat.tex
On real functions in Sh(X)

@@
 direction, occurs with~$\Sh(X)$. What~$\Sh(X)$ believes to be a (Dedekind) real
-number~$a$ is actually a continuous function~$a : X \to \RR$, that is, a continuous
-family of real numbers on~$X$.
+number~$a$ is actually a continuous family of real numbers on~$X$, that is, a
+function~$a : X \to \RR$.
 
@@
 
+\subsection{Real functions} Let~$(f_x)_{x \in X}$ be a continuous family of
+continuous real-valued functions; that is, each of the individual
+functions~$f_x : \RR \to \RR$ should be continuous and moreover the map~$\RR
+\times X \to \RR, (a,x) \mapsto f_x(a)$ should be continuous.
+From the point of view of~$\Sh(X)$, this family looks like a single continuous
+function~$f : \RR \to \RR$.
+
+The internal statement that~$f(-1) < 0$ means that~$f_x(-1) < 0$ for all~$x \in
+X$, and similarly so for being positive. More generally, if~$a$ and~$b$ are
+continuous functions~$X \to \RR$ (hence real numbers from the internal point of
+view), the internal statement~$f(a) < b$ means that~$f_x(a(x)) < b(x)$ for
+all~$x \in X$.
+
+The internal statement that~$f$ possesses a zero, that is, that there exists a
+number~$a$ such that~$f(a) = 0$, means that all the functions~$f_x$ each
+possess a zero and that moreover, these zeros can locally be picked in a
+continuous fashion. More precisely, this statement means that there is an open
+covering~$X = \bigcup_i U_i$ such that, for each~$i$, there is a continuous
+function~$a : U_i \to \RR$ such that~$f_x(a(x)) = 0$ for all~$x \in U_i$. (On
+overlaps~$U_i \cap U_j$, the zero-picking functions~$a$ need not agree.)
+
+From these observations we can deduce that the intermediate value theorem of
+basic analysis does in general not hold in~$\Sh(X)$ and hence does not allow
+for an intuitionistic proof. This theorem states: ``If~$f : \RR \to \RR$
+is a continuous function such that~$f(-1) < 0$ and~$f(1) > 0$, there exists a
+number~$a$ such that~$f(a) = 0$.'' The external meaning of this statement is
+that in any continuous family~$(f_x)_x$ of continuous functions with~$f_x(-1) <
+0$ and~$f_x(1) > 0$ for all~$x \in X$, it's locally possible to pick zeros of
+the family in a continuous fashion. Figure~\ref{fig:ivt} shows a counterexample
+to this claim.
+
 \printbibliography

### 2018-12-04 paper-filmat.tex
Topos built using infinite-time Turing machines

@@
 being prominent counterexamples. In the \effective topos, this statement is
-true. A formal proof is not entirely straightforward~\cite{XXX}; the intuitive
-reason for why this is so is as follows.
+true. A formal proof is not entirely straightforward~\cite{XXX}, but an intuitive
+explanation is as follows.
 
@@
     \realizes \varphi$. \\\\
+    \begin{minipage}{\textwidth}
+    In the following, we write~``$e \cdot n \downarrow$'' to mean that calling
+    the~$e$-th Turing machine on input~$n$ terminates, and in this case denote
+    the result by~``$e \cdot n$''.\end{minipage} \\\\
     $e \realizes s = t$ \> iff $s = t$. \\
@@
   \caption{\label{table:eff} A (fragment of) the translation
-  rules defining the meaning of statements internal to the \effective topos.
-  We write~``$e \cdot n \downarrow$'' to mean that calling the~$e$-th Turing
-  machine on input~$n$ terminates, and in this case denote the result by~``$e
-  \cdot n$''.}
+  rules defining the meaning of statements internal to the \effective topos.}
 \end{table}
+
+\subsection{Variants of the \effective topos} The \effective topos belongs to a
+wider class of \emph{realizability toposes}. These can be obtained by repeating
+the construction of the \effective topos with any other reasonable model of
+computation in place of Turing machines. The resulting toposes will in general
+not be equivalent and reflect higher-order properties of the employed models.
+Two of these further toposes are of special philosphical interest.
+
+Firstly, in place of ordinary Turing machines, one can employ the
+\emph{infinite-time Turing machines} pioneered by Hamkins and
+Lewis~\cite{hamkins-lewis:ittm}. These machines model \emph{hypercomputation}
+in that they can run for ``longer than infinity''; more precisely, the
+computational steps are indexed by the ordinal numbers instead of the natural
+numbers.
+
+In the realizability topos made using infinite-time Turing machines, the full
+law of excluded middle still fails, but some instances which are wrong in the
+\effective topos do hold in this topos. For instance, the statement ``any
+function~$\NN \to \NN$ is the zero function or not'' does: Its external meaning
+is that there is an infinite-time Turing machine~$M$ which, given the
+description of an infinite-time Turing machine~$F$ computing a function~$f :
+\NN \to \NN$ as input, terminates (at some ordinal time step) with a correct
+judgment of whether~$f$ is the zero function or not. Such a machine~$M$ indeed
+exists: It simply has to simulate~$F$ on all inputs~$0,1,\ldots$ in order and
+check whether one of the resulting function values is not zero. This
+will require a transfinite amount of time (not least because simulating~$F$ on
+just one input might require a transfinite amount of time), but as an
+infinite-time Turing machine,~$M$ is capable of carrying out this procedure.
+
+This realizability topos provides an intriguing environment challenging many
+mathematical intuitions shaped by classical logic. For instance, while from the
+point of view of this topos the reals are still uncountable in the sense that
+there is no surjection~$\NN \to \RR$, there is an injection~$\RR \to
+\NN$~\cite{bauer:injection}.
+
+A second ......
 

### 2018-12-05 slides-padova2018.tex
Slides for today's talk at Milly's group

--- 2018-10-11 slides-como2018.tex
@@
-\documentclass[11pt,utf8,notheorems,compress,t]{beamer}
+\documentclass[12pt,utf8,notheorems,compress,t]{beamer}
 \usepackage{etex}
-
-\usepackage{pgfpages}
-\setbeameroption{show notes on second screen}
-\setbeamertemplate{note page}[plain]
-\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying#2\par}}}
-
-% Workaround for the issue described at
-% https://tex.stackexchange.com/questions/164406/beamer-using-href-in-notes.
-\newcommand{\fixedhref}[2]{\makebox[0pt][l]{\hspace*{\paperwidth}\href{#1}{#2}}\href{#1}{#2}}
 
@@
 
-\title{How topos theory can help commutative algebra}
+\title{New reduction techniques in commutative algebra driven by logical methods}
 \author{Ingo Blechschmidt}
-\date{June 29th, 2018}
+\date{October 24th, 2018}
 
 \useinnertheme[shadow=true]{rounded}
-\useoutertheme[subsection=false]{miniframes}
 \setbeamerfont{block title}{size={}}
@@
 \newcommand{\A}{\mathcal{A}}
-\newcommand{\B}{\mathcal{B}}
 \renewcommand{\AA}{\mathbb{A}}
-\renewcommand{\C}{\mathcal{C}}
 \newcommand{\E}{\mathcal{E}}
 \newcommand{\F}{\mathcal{F}}
-\newcommand{\M}{\mathcal{M}}
 \renewcommand{\G}{\mathcal{G}}
@@
 \newcommand{\ZZ}{\mathbb{Z}}
+\newcommand{\CC}{\mathbb{C}}
 \renewcommand{\P}{\mathcal{P}}
-\newcommand{\aaa}{\mathfrak{a}}
+\newcommand{\ppp}{\mathfrak{p}}
 \newcommand{\fff}{\mathfrak{f}}
-\newcommand{\ppp}{\mathfrak{p}}
-\newcommand{\mmm}{\mathfrak{m}}
 \newcommand{\defeq}{\vcentcolon=}
@@
 \newcommand{\Aff}{\mathrm{Aff}}
+\newcommand{\Ring}{\mathrm{Ring}}
+\newcommand{\LocRing}{\mathrm{LocRing}}
 \newcommand{\LRS}{\mathrm{LRS}}
@@
 \newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
-\newcommand{\Ll}{\text{iff}}
+\newcommand{\Ll}{\vcentcolon\!\Longleftrightarrow}
 \newcommand{\inv}{inv.\@}
 \newcommand{\seq}{\vdash_{\!\!\!\vec x}}
-\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates
 
 \setbeamertemplate{blocks}[rounded][shadow=false]
+
+\newenvironment{indentblock}{%
+  \list{}{\leftmargin\leftmargin}%
+  \item\relax
+}{%
+  \endlist
+}
 
@@
 }
-
-\newenvironment{indentblock}{%
-  \list{}{\leftmargin\leftmargin}%
-  \item\relax
-}{%
-  \endlist
-}
-
-\newcommand{\hcancel}[5]{%
-  \tikz[baseline=(tocancel.base)]{
-    \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
-    \draw[red, line width=0.4mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
+\newcommand{\bigheart}{\includegraphics{heart}}
+
+\newenvironment{changemargin}[2]{%
+  \begin{list}{}{%
+    \setlength{\topsep}{0pt}%
+    \setlength{\leftmargin}{#1}%
+    \setlength{\rightmargin}{#2}%
+    \setlength{\listparindent}{\parindent}%
+    \setlength{\itemindent}{\parindent}%
+    \setlength{\parsep}{\parskip}%
   }%
-}
+  \item[]}{\end{list}}
 
@@
 % Adapted from https://latex.org/forum/viewtopic.php?t=2251 (Stefan Kottwitz)
-\newenvironment<>{varblock}[2]{
+\newenvironment<>{varblock}[2]{\begin{varblockextra}{#1}{#2}{}}{\end{varblockextra}}
+\newenvironment<>{varblockextra}[3]{
   \begin{center}
     \begin{minipage}{#1}
-      %\setlength{\textwidth}{#1}
-      \begin{actionenv}#3
-  \def\insertblocktitle{\centering #2}
-  \par
+      \begin{actionenv}#4
+        {\centering \hil{#2}\par}
+	\def\insertblocktitle{}%\centering #2}
+        \def\varblockextraend{#3}
 	\usebeamertemplate{block begin}}{
@@
         \usebeamertemplate{block end}
+        \varblockextraend
       \end{actionenv}
@@
 \setbeamertemplate{frametitle}{%
-  \vskip0.7em%
+  \vskip1.2em%
   \leavevmode%
@@
 
-\setbeamertemplate{headline}{%
+\newcommand{\mynav}[3]{%
   \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex]{}%
-    \insertsectionnavigationhorizontal{\paperwidth}{}{}%
+    \begin{columns}
+      \begin{column}{0.333\textwidth}
+        \raggedright
+        \textcolor{#1}{\qquad\qquad Summary}
+        \par
+      \end{column}
+      \begin{column}{0.333\textwidth}
+        \centering
+        \textcolor{#2}{The forcing model}
+        \par
+      \end{column}
+      \begin{column}{0.333\textwidth}
+        \raggedleft
+        \textcolor{#3}{Revisiting the test cases \qquad\qquad}
+        \par
+      \end{column}
+    \end{columns}
   \end{beamercolorbox}%
-  \vskip0pt%
-}
-
+}
+
+\newcommand{\insertframeextra}{}
 \setbeamertemplate{footline}{%
@@
 \newcommand{\hil}[1]{{\usebeamercolor[fg]{item}{\textbf{#1}}}}
-
 \newcommand{\bad}[1]{\textcolor{red!90}{\textnormal{#1}}}
 
+\newcommand{\jnote}[2]{}
+
 \begin{document}
 
-%\addtocounter{framenumber}{-1}
-%{\usebackgroundtemplate{\includegraphics[width=\paperwidth]{slides-como2018-titlepage-blurred}}
-%\begin{frame}[plain]\end{frame}}
-
 \addtocounter{framenumber}{-1}
+
+\setbeamertemplate{headline}{\mynav{gray}{gray}{gray}}
 
@@
   \centering
-  \medskip
-  \hspace*{-3.9em}%
-
-  \hil{How topos theory can help commutative algebra} \\
-
-  \emph{-- an invitation --}
 
   \bigskip
+  \includegraphics[width=0.4\textwidth]{external-internal-small}
   \bigskip
 
-  \begin{columns}
-    \small
-    \begin{column}{0.33\textwidth}
-      \centering
-      \bignumber{1}
-      \smallskip
-
-      \hil{Background on the internal language}
-      \medskip
-    \end{column}
-
-    \begin{column}{0.33\textwidth}
-      \centering
-      \bignumber{2}
-      \smallskip
-
-      \hil{Applications in commutative algebra}
-      \medskip
-    \end{column}
-
-    \begin{column}{0.33\textwidth}
-      \centering
-      \bignumber{3}
-      \smallskip
-
-      \hil{The mystery of \mbox{\!\!nongeometric sequents}}
-      \medskip
-    \end{column}
-  \end{columns}
+  \hil{New reduction techniques in commutative algebra driven by logical methods}
 
   \scriptsize
-  Ingo Blechschmidt (MPI Leipzig) \\
-  Toposes in Como \\
-  June 29th, 2018
+  \textit{-- an invitation --}
+  \bigskip
+
+  Ingo Blechschmidt \\
+  Università di Verona
+  \bigskip
+
+  Logic Seminar Padova \\
+  December 5st, 2018
   \par
@@
 
-\section{Motivation}
-
-\begin{frame}{Motivating test cases}
-  Let $A$ be a ring (commutative, with unit, $1 = 0$ allowed). \\
-  Assume that $A$ is reduced: If $x^n = 0$, then $x = 0$.
-
-  \jnote{1-2}{The two displayed statements are trivial for fields. It is
-  therefore natural to try to reduce the general situation to the field
-  situation.}
-
-  \jnote{2}{The displayed proofs, which could have been taken from any
-  standard textbook on commutative algebra, succeed in this reduction quite
-  easily by employing maximal ideals or minimal prime ideals. However, this way
-  of reducing comes at a cost: It requires the Boolean Prime Ideal Theorem
-  (for ensuring the existence of a prime ideal and for ensuring that stalks at
-  minimal prime ideals are fields) and even the full axiom of choice (for
-  ensuring the existence of a minimal prime ideal).
-
-  It therefore doesn't work in the internal universe of most toposes, and in
-  any case it obscures explicit computational content: Statements so simple as
-  the two displayed ones should admit explicit, computational proofs.
-
-  We'll learn how the internal language of a certain well-chosen topos provides
-  a way to perform the reduction in an entirely constructive manner. If so
-  desired, the resulting topos-theoretic proofs can be unwound to yield fully
-  explicit, topos-free, direct proofs.
-
-  Beautiful constructive proofs can also be found in Richman's note on
-  \fixedhref{https://www.ams.org/journals/proc/1988-103-04/S0002-9939-1988-0954974-5/S0002-9939-1988-0954974-5.pdf}{nontrivial
-  uses of trivial rings} and in the
-  \fixedhref{https://arxiv.org/abs/1605.04832}{recent textbook by Lombardi and
-  Quitté} on constructive commutative algebra.}
-
-  \jnote{3}{We can slightly reduce the requirements of the proof of the first
-  statement by employing not a maximal ideal, but a prime ideal. The existence
-  of maximal ideals in nontrivial rings is equivalent to the axiom of choice,
-  while the existence of prime ideals is equivalent to the weaker Boolean Prime
-  Ideal Theorem. However, this improvement doesn't change the main point; the
-  proof is still wildly nonconstructive.}
-
-  \jnote{4-}{Grothendieck's generic freeness lemma is an important theorem in
-  algebraic geometry, where it is usually stated in the following geometric
-  form:
-  \begin{indentblock}Let~$X$ be a reduced scheme. Let~$\B$ be
-  an~$\O_X$-algebra of finite type. Let~$\M$ be a~$\B$-module of finite type.
-  Then over a dense open,
+\section{Summary}
+\setbeamertemplate{headline}{\mynav{black}{gray}{gray}}
+
+\begin{frame}{Summary}
+  \vspace*{-1em}
+
+  \visible<4>{
+    \begin{changemargin}{-2.0em}{-0.5em}
+      \begin{itemize}
+        \item \ \\[-1.2em]\mbox{For any reduced ring~$A$, there is a ring~$A^\sim$ in a certain topos with}
+        \[ \models \bigl(\forall x\?A^\sim\_ \neg(\exists y\?A^\sim\_ xy = 1) \Rightarrow x = 0\bigr). \]
+
+        \item This semantics is sound with respect to intuitionistic logic.
+
+        \item \ \\[-1.2em]\mbox{It has uses in classical and constructive commutative
+        algebra.}
+      \end{itemize}
+    \end{changemargin}
+  }
+
+  \vspace*{-1.5em}
+
+  \begin{columns}[t]
+    \begin{column}[t]{0.52\textwidth}
+      \centering
+
+      \begin{varblock}{\textwidth}{A baby example}
+        \justifying
+        Let~$M$ be an injective matrix with more columns than rows over a
+        reduced ring~$A$.
+        Then~$1 = 0$ in~$A$.
+      \end{varblock}
+      \vspace*{-0.5em}
+
+      \only<1>{
+        \scalebox{0.8}{$\begin{pmatrix}
+          \cdot & \cdot & \cdot & \cdot & \cdot \\
+          \cdot & \cdot & \cdot & \cdot & \cdot \\
+          \cdot & \cdot & \cdot & \cdot & \cdot
+        \end{pmatrix}$}
+      }
+
+      \visible<2->{
+        \justifying
+        \textbf{Proof.} \bad{Assume not.} Then there is a \bad{minimal
+        prime ideal} $\ppp \subseteq A$. The matrix is injective over the \bad{field}~$A_\ppp = A[(A
+        \setminus \ppp)^{-1}]$; contradiction to basic linear algebra.
+      }
+    \end{column}
+
+    \begin{column}[t]{0.46\textwidth}
+      \centering
+
+      \begin{varblock}{\textwidth}{Generic freeness\phantom{p}}
+        \justifying
+        Generically, any finitely generated module over a reduced ring
+        is free.\phantom{g}
+      \end{varblock}
+      \vspace*{-0.5em}
+
+      \only<1-2>{{
+        \scriptsize\raggedright
+        (A ring is reduced iff $x^n=0$ implies $x=0$.)
+        \par
+      }}
+
+      \only<1-2>{\includegraphics[width=0.73\textwidth]{generic-freeness}}
+
+      \visible<3->{
+        \justifying
+        \textbf{Proof.} See [Stacks Project].
+      }
+    \end{column}
+  \end{columns}
+\end{frame}
+
+\section{The forcing model}
+\setbeamertemplate{headline}{\mynav{gray}{black}{gray}}
+
+\begin{frame}{Motivating the semantics}
+  \centering
+  \begin{varblockextra}{0.8\textwidth}{}{
+    \hil{Examples:}\phantom{Non-}\,\!\ \ $k,\ k[[X]],\ \CC\{z\},\ \ZZ_{(p)}$ \\[0.2em]
+    \hil{Non-examples:}\ \ $\ZZ,\ k[X],\ \ZZ/(pq)$
+  }
+    \justifying
+    A ring is \hil{local} iff~$1 \neq 0$ and if~$x + y = 1$ implies that~$x$ is
+    invertible or~$y$ is invertible.
+  \end{varblockextra}
+
+  \begin{varblockextra}{0.8\textwidth}{}{
+    Let~$x + y = 1$ in a ring~$A$.
+    Then:
+    \begin{itemize}
+      \item The element $x$ is invertible in~$A[x^{-1}]$.
+      \item The element $y$ is invertible in~$A[y^{-1}]$.
+    \end{itemize}
+  }
+    \hil{Locally,} any ring is local.
+  \end{varblockextra}
+\end{frame}
+
+\begin{frame}{The Kripke--Joyal semantics}
+  \small
+  \mbox{\!\!\!Recall~$A[f^{-1}] = \bigl\{ \frac{u}{f^n} \,|\, u \in A, n \in \NN \bigr\}$.
+  Let~``$\models \varphi$'' be short for~``$1 \models \varphi$''.}
+  \only<1>{\[ \renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\quad}c@{\quad}l@{}}
+    f \models \top &\text{iff}& \top \\
+    f \models \bot &\text{iff}& \text{$f$ is nilpotent} \\
+    f \models x = y &\text{iff}& x = y \in A[f^{-1}] \\
+    f \models \varphi \wedge \psi &\text{iff}&
+      \text{$f \models \varphi$ and $f \models \psi$} \\
+    f \models \varphi \vee \psi &\text{iff}&
+      \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
+    &&\quad\text{for each~$i$, $fg_i \models \varphi$ or $fg_i \models \psi$} \\
+    f \models \varphi \Rightarrow \psi &\text{iff}&
+      \text{for all~$g \in A$, $fg \models \varphi$ implies $fg \models \psi$} \\
+    f \models \forall x\?A^\sim\_ \varphi &\text{iff}&
+      \text{for all~$g \in A$ and all $x_0 \in A[(fg)^{-1}]$, $fg \models \varphi[x_0/x]$} \\
+    f \models \exists x\?A^\sim\_ \varphi &\text{iff}&
+      \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
+    &&\quad\text{for each~$i$, $fg_i \models \varphi[x_0/x]$ for some~$x_0 \in A[(fg_i)^{-1}]$}
+  \end{array} \]}
+  \only<2->{\[ \renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\quad}c@{\quad}l@{}}
+    f \models x = y &\text{iff}& x = y \in A[f^{-1}] \\
+    f \models \varphi \wedge \psi &\text{iff}&
+      \text{$f \models \varphi$ and $f \models \psi$} \\
+    f \models \varphi \vee \psi &\text{iff}&
+      \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
+    &&\quad\text{for each~$i$, $fg_i \models \varphi$ or $fg_i \models \psi$}
+  \end{array} \]}
+
+  \pause
+
+  \begin{columns}
+    \begin{column}{0.50\textwidth}
+      \begin{varblock}{\textwidth}{Monotonicity}{}
+        If~$f \models \varphi$, then also~$fg \models \varphi$.
+      \end{varblock}
+    \end{column}
+
+    \begin{column}{0.50\textwidth}
+      \begin{varblock}{\textwidth}{Locality}{}
+        \justifying
+        If~$f^n = fg_1 + \cdots + fg_m$ and~$fg_i \models \varphi$ for all~$i$,
+        then also~$f \models \varphi$.
+      \end{varblock}
+    \end{column}
+  \end{columns}
+
+  \begin{columns}
+    \begin{column}{0.50\textwidth}
+      \begin{varblock}{\textwidth}{Soundness\phantom{p}}{}
+        If~$\varphi \vdash \psi$ and~$f \models \varphi$,
+        then~$f \models \psi$.
+      \end{varblock}
+    \end{column}
+
+    \begin{column}{0.50\textwidth}
+      \begin{varblock}{\textwidth}{Forced properties}{}
+        $\models \speak{$A^\sim$ is a local ring}$.
+      \end{varblock}
+    \end{column}
+  \end{columns}
+\end{frame}
+
+\tikzstyle{topos} = [draw=mypurple, very thick, rectangle, rounded corners, inner sep=5pt, inner ysep=10pt]
+\tikzstyle{title} = [fill=mypurple, text=white]
+
+\input{images/primes.tex}
+%\renewcommand{\sieve}[2]{SIEVE}
+%\renewcommand{\fakesieve}[2]{SIEVE}
+
+\newcommand{\drawbox}[4]{
+  \node[topos, #4] [fit = #3] (#1) {};
+  \node[title] at (#1.north) {#2};
+}
+
+\newcommand{\muchstuff}{
+  \includegraphics[height=3em]{filmat}
+  \scalebox{0.5}{\sieve{14}{2}}
+}
+
+\newcommand{\muchstuffplaceholder}{
+  \includegraphics[height=3em]{filmat-placeholder}
+  \scalebox{0.5}{\fakesieve{14}{2}}
+}
+
+\newcommand{\fewstuff}{
+  \includegraphics[height=3em]{filmat}
+  \scalebox{0.5}{\sieve{7}{2}}
+}
+
+\begin{frame}[fragile]{A universal property}
+  The displayed semantics is the first-order fragment of the \hil{higher-order
+  internal language} of the \hil{little Zariski topos}.
+
+  \begin{tikzpicture}
+    \node[scale=0.4] (objs-set1) at (-4.0,-2.5) {
+      \only<1->{\fewstuff}
+    };
+    \node[scale=0.4] (objs-eff1) at (4.0,-2.5) {
+      \only<1->{\fewstuff}
+    };
+    \node[scale=0.4] (objs-sh1)  at (0,-2.5) {
+      \only<1->{\fewstuff}
+    };
+
+    \node (prop-set1) [below of=objs-set1, align=left, inner ysep=0pt] {
+      The usual laws \\
+      of logic hold.
+    };
+
+    \node (prop-eff1) [below of=objs-eff1, align=left, inner ysep=0pt] {
+      Every function \\
+      is computable.
+    };
+
+    \node (prop-sh1) [below of=objs-sh1, align=left, inner sep=0pt] {
+      The intermediate \\
+      value theorem fails.
+    };
+
+    \begin{scope}
+      \drawbox{set1}{$\mathrm{Set}$}{(objs-set1) (prop-set1)}{}
+    \end{scope}
+    \begin{scope}
+      \drawbox{eff1}{Ef{}f}{(objs-eff1) (prop-eff1)}{tape}
+    \end{scope}
+    \begin{scope}
+      \drawbox{sh1}{$\mathrm{Sh}\, X$}{(objs-sh1) (prop-sh1)}{draw=none}
+      \def\R{8pt}
+      \begin{pgfonlayer}{background}
+      \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple]
+        ($(sh1.south west) + (\R, 0)$) arc(270:180:\R) --
+        ($(sh1.north west) + (0, -\R)$) arc(180:90:\R) --
+        ($(sh1.north east) + (-\R, 0)$) arc(90:0:\R) --
+        ($(sh1.south east) + (0, \R)$) arc(0:-90:\R) --
+        cycle;
+      \end{pgfonlayer}
+    \end{scope}
+  \end{tikzpicture}
+
+  \pause
+
+  Is there a \hil{free local ring}~$A \to A'$ over~$A$?
+  \begin{columns}[t]
+    \begin{column}{0.4\textwidth}
+      $\xymatrix{
+        A \ar[rd] \ar[rrr]^\alpha &&& {\substack{\phantom{\text{local}}\\\text{\normalsize$R$}\\\text{local}}} \\
+        & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!35]{\text{local}}[rru]
+      }$
+    \end{column}
+
+    \begin{column}{0.50\textwidth}
+      \small\justifying
+      For a fixed ring~$R$, the localization $A' \defeq A[S^{-1}]$ with $S \defeq
+      \alpha^{-1}[R^\times]$ would do the job.
+      \medskip
+
+      Hence we need the \hil{generic filter}.
+    \end{column}
+  \end{columns}
+\end{frame}
+
+\begin{frame}{The little Zariski topos}
+  \small
+
+  Let~$A$ be a ring. Its \hil{little Zariski topos} is equivalently
+  \vspace*{-0.5em}
   \begin{enumerate}
-  \item[(a)] $\B$ and~$\M$ are locally free as sheaves of~$\O_X$-modules,
-  \item[(b)] $\B$ is of finite presentation as a sheaf of~$\O_X$-algebras and
-  \item[(c)] $\M$ is of finite presentation as a sheaf of~$\B$-modules.
-  \end{enumerate}\end{indentblock}}
-
-  \jnote{5}{All previously known proofs proceed in a series of reduction steps,
-  finally culminating in the case where~$A$ is a Noetherian integral domain.
-  They are somewhat convoluted (for instance, the proof in the Stacks Project
-  is three pages long) and employ several results in commutative algebra which
-  have not yet been constructivized.
-
-  Using the internal language of toposes, we will give a short, conceptual and
-  constructive proof of Grothendieck's generic freeness lemma. Again, if so
-  desired, one can unwind the internal proof to obtain a
-  constructive proof which doesn't reference topos theory. The proof obtained
-  in this way is still an improvement on the previously known proofs, requiring
-  no advanced prerequisites in commutative algebra, and takes
-  \fixedhref{https://arxiv.org/abs/1807.01231}{about
-  a page}.}
-
-  \only<1-3>{\begin{columns}[t]
-  \begin{column}[t]{0.50\textwidth}
-    \centering
-
-    \scalebox{0.8}{$\begin{pmatrix}
-      \cdot & \cdot \\
-      \cdot & \cdot \\
-      \cdot & \cdot
-      \end{pmatrix}$}
-    \vspace*{-1em}
-
-    \begin{varblock}{\textwidth}{A baby application}
-      \justifying
-      Let~$M$ be a surjective matrix over~$A$ with more rows than columns.
-      Then~$1 = 0$ in~$A$.
-    \end{varblock}
-
-    \justifying
-    \only<2>{\textbf{Proof.} \bad{Assume not.} Then there is~a \bad{maximal
-    ideal} $\mmm$. The matrix is surjective over the field~$A/\mmm$. This is a
-    contradiction to basic linear algebra.}
-
-    \only<3>{\textbf{Proof.} \bad{Assume not.} Then there is~a \bad{prime
-    ideal} $\ppp$. The matrix is surjective over the
-    field~$\mathrm{Quot}(A/\ppp)$. This is a contradiction to basic linear
-    algebra.}
-  \end{column}
-
-  \begin{column}[t]{0.5\textwidth}
-    \centering
-
-    \scalebox{0.8}{$\begin{pmatrix}
-      \cdot & \cdot & \cdot & \cdot \\
-      \cdot & \cdot & \cdot & \cdot \\
-      \cdot & \cdot & \cdot & \cdot
-    \end{pmatrix}$}
-    \vspace*{-1em}
-
-    \begin{varblock}{\textwidth}{A child application}
-      \justifying
-      Let~$M$ be an injective matrix over~$A$ with more columns than rows.
-      Then~$1 = 0$ in~$A$.
-    \end{varblock}
-
-    \justifying
-    \visible<2-3>{\textbf{Proof.} \bad{Assume not.} Then there is a \bad{minimal
-    prime ideal} $\ppp$. The matrix is injective over the field~$A_\ppp = A[(A
-    \setminus \ppp)^{-1}]$. This is a contradiction to basic linear algebra.}
-    \end{column}
-  \end{columns}}
-
-  \pause
-  \pause
-  \pause
-
-  \centering
-
-  \includegraphics[height=3em]{generic-freeness}
-  \vspace*{-1em}
-
-  \begin{varblock}{0.9\textwidth}{Generic freeness$\phantom{p}$}
-    \justifying
-    Let~$B$ be an~$A$-algebra of finite type ($\cong A[X_1,\ldots,X_n]/\aaa$). \\
-    Let~$M$ be a finitely generated~$B$-module ($\cong B^m/U$).
-
-    If~$f = 0$ is the only element of~$A$ such that
-    \begin{enumerate}
-      \item $B[f^{-1}]$ and $M[f^{-1}]$ are free modules over $A[f^{-1}]$,
-      \item $A[f^{-1}] \to B[f^{-1}]$ is of finite presentation and
-      \item $M[f^{-1}]$ is finitely presented as a module over~$B[f^{-1}]$,
-    \end{enumerate}
-    then $1 = 0$ in~$A$.
-  \end{varblock}
-  \vspace*{-0.5em}
-
-  \pause\center
-  \begin{minipage}{0.9\textwidth}
-    \textbf{Proof.} See [Stacks Project, Tag 051Q].
-  \end{minipage}
-\end{frame}
-
-\section{Internal language}
-
-\begin{frame}{The internal language of a topos}
-  For any topos~$\E$ and any formula~$\varphi$, we define the meaning of
-  \vspace*{-0.5em}
-  \[
-    \text{``$\E \models \varphi$''} \quad
-    \text{(``$\varphi$ holds in the internal universe of~$\E$'')}
-  \]
-
-  \vspace*{-1.0em}
-  using (Shulman's extension of) the \hil{Kripke--Joyal semantics}.
-
-  \jnote{1}{The internal language of a topos allows to construct objects and
-  morphisms of the topos, formulate statements about them and prove such
-  statements \emph{in a naive element-based language}. From the internal point
-  of view, objects look like sets [more precisely, types]; morphisms look like
-  maps; monomorphisms look like injections; epimorphisms
-  look like surjections; group objects look like groups; and so on.
-
-  To determine whether a statement~$\varphi$ holds in the internal universe of a
-  given topos, we can use the Kripke--Joyal semantics to translate it into an
-  ordinary external statement and then check the validity of the external
-  translation.
-
-  For instance, in the ef{}fective topos the curious statement~``any function
-  $\NN \to \NN$ is computable'' holds, for its external meaning is the
-  triviality ``there is a Turing machine which given a Turing machine computing
-  some function~$f : \NN \to \NN$ outputs a Turing machine computing~$f$''. In
-  contrast, the statement~``any function~$\NN \to \NN$ is either the zero
-  function or not'' does not hold in the ef{}fective topos, since its external
-  meaning is ``there exists a Turing machine which given a Turing machine
-  computing some function~$f : \NN \to \NN$ decides whether~$f$ is the zero
-  function or not''.}
-
-  \jnote{2}{Any theorem which has an intuitionistic proof holds in
-  the internal universe of any topos. The restriction to intuitionistic logic
-  is not due to philosophical concerns; it is a fact of life that only very few
-  toposes validate the law of excluded middle (for instance, sheaf toposes over
-  discrete topological spaces do if the law of excluded middle is available in
-  the metatheory). Luckily, vast amounts of mathematics can be developed in a
-  purely intuitionistic setting.
-
-  The internal language machinery itself can be developed in an intuitionistic
-  setting.
-
-  The standard internal language of toposes in not enough for our purposes, as
-  it misses unbounded quantification (``for all groups'', ``for all rings'')
-  and dependent types. Shulman's \fixedhref{https://arxiv.org/abs/1004.3802}{stack
-  semantics} offers what we need. No knowledge of stacks is necessary to enjoy
-  his paper. Prior work includes
-  \fixedhref{https://www.cl.cam.ac.uk/~amp12/papers/polist/polist.pdf}{Polymorphism
-  is Set Theoretic, Constructively} by Pitts and
-  \fixedhref{http://www.phil.cmu.edu/projects/ast/Papers/Awodey-Butz-Simpson-Streicher-APAL-2013.pdf}{Relating
-  first-order set theories, toposes and categories of classes} by Awodey, Butz,
-  Simpson and Streicher (obtained independently and published after Shulman).}
-
-  \vspace*{-1em}
-  \begin{columns}
-    \def\insertblocktitle{}
-    \begin{column}{0.25\textwidth}\usebeamertemplate{block begin}
-      \centering
-      $\Set \models \varphi$ \\
-      ``$\varphi$ holds in the \\ usual sense.''
-      \usebeamertemplate{block end}\end{column}
-
-    \begin{column}{0.25\textwidth}\usebeamertemplate{block begin}
-      \centering
-      $\Sh(X) \models \varphi$ \\
-      ``$\varphi$ holds continuously.''
-      \usebeamertemplate{block end}\end{column}
-
-    \begin{column}{0.25\textwidth}\usebeamertemplate{block begin}
-      \centering
-      $\Eff \models \varphi$ \\
-      ``$\varphi$ holds \\ computably.''
-      \usebeamertemplate{block end}\end{column}
-  \end{columns}
-  \medskip
-
-  Any topos supports \hil{mathematical reasoning}:
-
-  \vspace*{-1.5em}
-  \begin{hilblock}
-    If~$\E \models \varphi$ and if~$\varphi$ entails~$\psi$
-    \pointthis{<2>}{intuitionistically}{%
-      no $\varphi \vee \neg\varphi$,\ \
-      no $\neg\neg\varphi \Rightarrow \varphi$,\ \
-      no axiom of choice},
-    then~$\E \models \psi$.
-  \end{hilblock}
-\end{frame}
-
-\begin{frame}{The Kripke--Joyal semantics of~$\boldsymbol{\Sh(X)}$}
-  \small\vspace*{-0.5em}
-  Let~$X$ be a topological space. We recursively define
-  \[ U \models \varphi \quad \text{(``$\varphi$ holds on~$U$'')} \]
-  \mbox{for open subsets~$U \subseteq X$ and formulas~$\varphi$.
-  Write~``$\Sh(X) \models \varphi$'' to mean~$X \models \varphi$.}
-  \footnotesize
-  \[ \renewcommand{\arraystretch}{1.08}\begin{array}{@{}l@{\ \ }c@{\ \ }l@{}}
-  U \models \top &\Ll& \text{true} \\
-  U \models \bot &\Ll& \hcancel{\text{false}}{0pt}{3pt}{0pt}{-2pt}\ U = \emptyset \\
-  U \models s = t \? F &\Ll& s|_U = t|_U \in F(U) \\
-  U \models \varphi \wedge \psi &\Ll&
-  \text{$U \models \varphi$ and $U \models \psi$} \\
-  U \models \varphi \vee \psi &\Ll&
-  \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $U = \bigcup_i U_i$} \\
-  && \quad\quad\text{such that for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
-  U \models \varphi \Rightarrow \psi &\Ll&
-  \text{for all open~$V \subseteq U$: }
-  \text{$V \models \varphi$ implies $V \models \psi$} \\
-  U \models \forall s \? F\_ \varphi(s) &\Ll&
-  \text{for all open $V \subseteq U$ and sections~$s_0 \in F(V)$: $V \models \varphi(s_0)$} \\
-  U \models \forall F\_ \varphi(F) &\Ll&
-  \text{for all open $V \subseteq U$ and sheaves~$F_0$ over~$V$: $V \models \varphi(F_0)$} \\
-  U \models \exists s \? F\_ \varphi(s) &\Ll&
-  \hcancel{\text{there exists $s_0 \in F(U)$ such that $U \models \varphi(s_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
-  &&
-  \text{there exists a covering $U = \bigcup_i U_i$ such that for all~$i$:} \\
-  && \quad\quad \text{there exists~$s_0 \in F(U_i)$ such that $U_i \models \varphi(s_0)$} \\
-  U \models \exists F\_ \varphi(F) &\Ll&
-  \hcancel{\text{there exists a sheaf $F_0$ on $U$ such that $U \models
-  \varphi(F_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
-  &&
-  \text{there exists a covering $U = \bigcup_i U_i$ such that for all~$i$:} \\
-  && \quad\quad \text{there exists a sheaf~$F_0$ on $U_i$ such that $U_i \models \varphi(F_0)$}
-  \end{array} \]
-
-  \jnote{1}{Many interesting sheaves have few global sections, which is why a
-  definition such as ``$U \models \forall s \? F\_ \varphi(s)$ iff $U \models
-  \varphi(s_0)$ for all~$s_0 \in F(U)$'' would miss the point.
-
-  Here is an explicit example of the translation procedure. Let~$\alpha : F
-  \to G$ be a morphism of sheaves on~$X$. Then (the corner quotes ``$\speak{\ldots}$''
-  indicate that translation into formal language is left to the reader):
-  \footnotesize
-  \begin{align*}
-    & X \models \speak{$\alpha$ is injective} \\[0.2em]
-    \Longleftrightarrow\
-    & X \models \forall s\?F\_ \forall t\?F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.2em]
-    \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$, sections $s_0 \in F(U)$:} \\
-    &\qquad\qquad
-        U \models \forall t\?F\_ \alpha(s_0) = \alpha(t) \Rightarrow s_0 = t \\[0.2em]
-    \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$, sections $s_0 \in F(U)$:} \\
-    &\qquad\qquad
-        \text{for all open~$V \subseteq U$, sections $t_0 \in F(V)$:} \\
-    &\qquad\qquad\qquad\qquad
-          V \models \alpha(s_0) = \alpha(t_0) \Rightarrow s_0 = t_0 \\[0.2em]
-    \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$, sections $s_0 \in F(U)$:} \\
-    &\qquad\qquad
-        \text{for all open~$V \subseteq U$, sections $t_0 \in F(V)$:} \\
-    &\qquad\qquad\qquad\qquad
-          \text{for all open $W \subseteq V$: $\alpha_V(s_0|_W) = \alpha_V(t_0|_W)$ implies $s_0|_W = t_0|_W$} \\[0.2em]
-    \Longleftrightarrow\ &
-      \text{for all open~$U \subseteq X$, sections $s, t \in F(U)$:
-        $\alpha_U(s) = \alpha_U(t)$ implies $s = t$} \\[0.2em]
-    \Longleftrightarrow\ &
-      \text{$\alpha$ is a monomorphism of sheaves}
-  \end{align*}}
-\end{frame}
-
-\begin{frame}{Internalizing parameter-dependence}
-  \justifying
-  Let~$X$ be a space. A continuous family~$(f_x)_{x \in X}$ of continuous
-  functions (that is, a continuous function~$f : X \times \RR \to \RR$;
-  $f_x(a) = f(x,a)$) induces an endomorphism of the sheaf~$\C$ of continuous
-  functions:
-  \[
-    \bar f : \C \longrightarrow \C,\
-    \text{on $U$:}\ s \longmapsto (x \mapsto f_x(s(x))).
-  \]
-
-  \vspace*{-1em}
-
-  \begin{itemize}
-    \item $\Sh(X) \models \speak{The set $\C$ is the set of (Dedekind) reals}$.
-    \item $\Sh(X) \models \speak{The function $\bar f : \RR \to \RR$ is continuous}$. \bigskip
-    \item Iff $f_x(-1) < 0$ for all $x$, then $\Sh(X) \models \bar f(-1) < 0$.
-    \item Iff $f_x(+1) > 0$ for all $x$, then $\Sh(X) \models \bar f(+1) > 0$.
-    \item Iff all $f_x$ are increasing, then $\Sh(X) \models \speak{$\bar f$ is
-    increasing}$. \bigskip
-    \item \ \\[-1.2em]\mbox{Iff there is an open cover~$X = \bigcup_i U_i$ such
-    that for each~$i$ there is a} \mbox{continuous function~$s : U_i \to \RR$
-    with~$f_x(s(x)) = 0$ for all~$x \in U_i$,} then $\Sh(X) \models \exists s
-    \? \RR\_ \bar f(s) = 0$.
-  \end{itemize}
-
-  \jnote{1}{This slide, unrelated to commutative algebra or algebraic geometry,
-  aims to illustrate one of the basic uses of the internal language of toposes:
-  Upgrading any theorem admitting an intuitionistic proof to a
-  parameter-dependent version.
-
-  Constructively, there are several non-equivalent forms of the intermediate
-  value theorem. The following version doesn't admit an intuitionistic proof:
-  \begin{indentblock}Let~$g : \RR \to \RR$ be a function between the (Dedekind)
-  reals which is continuous in the usual epsilon/delta sense. Assume~$g(-1) < 0
-  < g(1)$. Then there exists a number~$x \in \RR$ such that~$g(x) = 0$.
-  \end{indentblock}
-  If there was an intuitionistic proof, the statement would hold in any topos,
-  so in particular in sheaf toposes over topological spaces. By the
-  translations shown on the slide, this
-  would amount to the following strengthening of the intermediate value
-  theorem: In continuous families of continuous functions, zeros can locally be
-  picked continuously. However, this strengthening is invalid, as
-  \fixedhref{https://raw.githubusercontent.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{this
-  video shows}.
-
-  \centering\includegraphics[width=0.1\textwidth]{zeros-in-families-still}}
-
-  \jnote{2}{In contrast, the following version does admit an intuitionistic
-  proof. We can therefore interpret it in sheaf toposes over topological
-  spaces and thereby obtain the strengthening that in continuous families of
-  strictly increasing continuous functions, zeros can locally be picked
-  continuously. You are invited to prove this strengthening directly, without
-  reference to the internal language.
-  \begin{indentblock}Let~$g : \RR \to \RR$ be a function between the (Dedekind)
-  reals which is continuous in the usual epsilon/delta sense and which is
-  strictly increasing ($a < b$ implies~$g(a) < g(b)$). Assume~$g(-1) < 0 <
-  g(1)$. Then there exists a number~$x \in \RR$ such that~$g(x) =
-  0$.\end{indentblock}}
-\end{frame}
-
-\section{The little Zariski topos}
-
-\begin{frame}{The little Zariski topos}
-  Let~$A$ be a ring. Its \hil{little Zariski topos} is equivalently
-  \begin{enumerate}
-    \item the classifying topos of \hil{local localizations} of $A$,
-    \item the classifying locale of \hil{prime filters} of $A$,
-    \item the locale given by the frame of \hil{radical ideals} of $A$,
-    \item the topos of sheaves over the poset $A$ with $f \preceq g$ iff $f \in \sqrt{(g)}$ and with $(f_i \to f)_i$ deemed covering iff $f \in \sqrt{(f_i)_i}$ or
+    \item the classifying locale of \hil{prime filters} of $A$, \\[-1.2em]
+    \item the classifying topos of \hil{local localizations} of $A$, \\[-1.2em]
+    \item the locale given by the frame of \hil{radical ideals} of $A$, \\[-1.2em]
+    \item the topos of sheaves over the poset $A$ with $f \preceq g$ iff $f \in \sqrt{(g)}$ and with $(f_i \to f)_i$ deemed covering iff $f \in \sqrt{(f_i)_i}$ or \\[-1.2em]
     \item the topos of sheaves over $\Spec(A)$.
@@
 
-  \footnotesize
+  \scriptsize
   Prime ideal:\, $0 \in \ppp$;\, $x \in \ppp \wedge y \in \ppp \Rightarrow x+y \in \ppp$;\, $1 \not\in \ppp$;\, $xy \in \ppp \Leftrightarrow x \in \ppp \vee y \in \ppp$
@@
   $xy \in \fff \Leftrightarrow x \in \fff \wedge y \in \fff$
-
-  \jnote{1}{Any geometric theory has a classifying topos; if the theory under
-  consideration is propositional (doesn't have any sorts), then its classifying
-  topos can be chosen to be the topos of sheaves over a locale. One can also
-  give a direct account of classifying locales, as a pedagogical stepping stone
-  to the full theory of classifying toposes.
-
-  The slide contains a small lie: The classical definition of the spectrum of a
-  ring is via the set of prime ideals of~$A$, not prime filters. If the law of
-  excluded middle is available, there is no difference between these
-  definitions since the complement of a prime ideal is a prime filter and vice
-  versa.
-
-  One can also consider the classifying locale of prime \emph{ideals} of~$A$.
-  Its associated topological space of points is the set of prime ideals
-  of~$A$ equipped with the \emph{constructible topology}.
-
-  In an intuitionistic context, any of the (generalized) spaces of items~1--4
-  can be adopted as sensible definitions of the spectrum of~$A$. Item~5 is then
-  a tautology. The classical definition of the spectrum as a topological space
-  doesn't work very well, because verifying the universal property one expects
-  of it requires the Boolean Prime Ideal Theorem. Most dramatically, there are
-  rings which are not trivial yet have neither prime ideals nor prime filters.
-  The classical definition yields in this case the empty space.}
-\end{frame}
-
-\begin{frame}{First steps in the little Zariski topos}
-  Let~$A$ be a ring. Let~$\fff_0$ be the \hil{generic prime filter} of~$A$; it
-  is a subobject of the constant sheaf~$\ull{A}$ of the little Zariski topos.
-
+\end{frame}
+
+\begin{frame}{Investigating the forcing model}
+% Let~$A$ be a reduced commutative ring ($x^n = 0 \Rightarrow x = 0$).
+  \small
+
+  The \hil{little Zariski topos} of a ring~$A$ is equivalently
+  \vspace*{-0.5em}
   \begin{itemize}
-    \item The ring $A^\sim \defeq \ull{A}[\fff_0^{-1}]$ is the generic local
-    localization of $A$.
-    \item Given an~$A$-module~$M$, we have the~$A^\sim$-module $M^\sim \defeq
-    \ull{M}[\fff_0^{-1}]$.
+    \item the topos of sheaves over~$\Spec(A)$, \\[-1.9em]
+    \item the locale given by the frame of radical ideals of~$A$, \\[-1.9em]
+%   \item the classifying topos of local localizations of~$A$ or
+    \item the classifying locale of filters of~$A$
   \end{itemize}
-
-  \jnote{1}{The Kripke--Joyal semantics for the little Zariski topos amounts to
-  the following: $\Spec(A) \models \varphi$ iff~$D(1) \models \varphi$, and the
-  clauses for~$D(f) \models \varphi$, where~$f$ ranges over the elements
-  of~$A$, are given by the following table.
-  \[\footnotesize\renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\quad}c@{\quad}l@{}}
-    D(f) \models \top &\text{iff}& \text{true} \\
-    D(f) \models \bot &\text{iff}& \text{$f$ is nilpotent} \\
-    D(f) \models x = y &\text{iff}& x = y \in M[f^{-1}] \\
-    D(f) \models \varphi \wedge \psi &\text{iff}&
-      \text{$D(f) \models \varphi$ and $D(f) \models \psi$} \\
-    D(f) \models \varphi \vee \psi &\text{iff}&
-      \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
-    &&\quad\text{for each~$i$, $D(fg_i) \models \varphi$ or $D(fg_i) \models \psi$} \\
-    D(f) \models \varphi \Rightarrow \psi &\text{iff}&
-      \text{for all~$g \in A$, $D(fg) \models \varphi$ implies $D(fg) \models \psi$} \\
-    D(f) \models \forall x\?M^\sim\_ \varphi(x) &\text{iff}&
-      \text{for all~$g \in A$ and $x_0 \in M[(fg)^{-1}]$, $D(fg) \models \varphi(x_0)$} \\
-    D(f) \models \exists x\?M^\sim\_ \varphi(x) &\text{iff}&
-      \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
-    &&\quad\text{for each~$i$, $D(fg_i) \models \varphi(x_0)$ for some~$x_0 \in M[(fg_i)^{-1}]$}
-  \end{array} \]
-  The generic prime filter~$\fff_0$ can also be described in explicit
-  terms. For ring elements~$f$ and~$s$, $D(f) \models (s \in \fff_0)$ iff~$f
-  \in \sqrt{(s)}$.}
-
-  \jnote{2}{Our description of~$M^\sim$ reveals a precise sense in
-  which~$M^\sim$ and~$M$ are related: $M^\sim$ is simply a localization of~$M$
-  (first lifted to another universe by the constant sheaf construction). The
-  classical descriptions don't make the relation evident.
-
-  As a first approximation, the module~$M^\sim$ can be thought of as a
-  reification of all the stalks of~$M$ as a single object. The metatheorem
-  displayed at the top left on the next slides makes this precise and also
-  exposes the limits of this view: It is only correct for geometric sequents.
-  When considering nongeometric sequents, phenomena appear which are unique
-  to~$M^\sim$ in the sense that they are in general not shared by~$M$, its
-  stalks or its quotients.}
-
-  \jnote{3}{One can show, assuming that the little Zariski topos is
-  \emph{overt}, that the module~$M$ in~$\Set$ and the module~$\ull{M}$ of the
-  little Zariski topos share all first-order properties. This observation
-  explains the metatheorem displayed at the bottom left. The assumption is
-  satisfied if any element of~$A$ is nilpotent or not nilpotent, so it's always
-  satisfied if the law of excluded middle is available in the metatheory. In an
-  intuitionistic context, it's still ``morally satisfied''. Details are in
-  Section~12.9 of
-  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
-  notes}.
-
-  The metatheorem soups up a number of lemmas of algebraic geometry, there
-  stated in geometric language. For instance, if~$M$ is finitely generated,
-  then~$M^\sim$ is of finite type. If~$M$ is finitely presented, then~$M^\sim$
-  is of finite presentation. If~$M$ is coherent, then~$M^\sim$ is coherent.
-
-  As an aside, the little Zariski topos is rarely Boolean (validates the law of
-  excluded middle). A necessary condition is that~$A$ is of dimension~$\leq 0$.}
-
-  \jnote{4}{Assuming that~$A$ is reduced, the following nongeometric sequents
-  hold in the little Zariski topos (among others):
-
-  $A^\sim$ is a field in the sense that zero is the only noninvertible
-  element. This field property was already observed in the 1970s by Mulvey, who
-  didn't know a deeper reason for this property. We now know that it's a
-  shadow of an internal property whose external translation expresses
-  that~$A^\sim$ is quasicoherent.
-
-  $A^\sim$ has~$\neg\neg$-stable equality in the sense that
-  \[ \Spec(A) \models \forall s\?A^\sim\_ \neg\neg(s = 0) \Rightarrow s = 0. \]
-  Classically, every set has~$\neg\neg$-stable equality; intuitionistically,
-  this is a special property of some sets. It's quite useful, as some theorems
-  of classical commutative algebra can only be proven intuitionistically when
-  weakened by double negation. The stability then allows, in some cases, to
-  obtain the original conclusion.
-
-  $A^\sim$ is anonymously Noetherian in the sense that any of its ideals
-  is \mbox{\emph{not not}} finitely generated. A philosophically-motivated
-  constructivist might be offended by this notion, since it runs counter to the
-  maxim that constructive mathematics should be informative. However, in the
-  internal context it is a useful notion: Hilbert's basis theorem holds for it,
-  and we'll put it to good use in our proof of Grothendieck's generic freeness
-  lemma.}
-
-  \jnote{5}{Are there theorems which can only be proven using the internal
-  language and not be proven without?
-
-  No. Just as the translation from internal statements to external statements
-  is entirely mechanical, so is the translation from internal proofs to
-  external proofs. Any proof employing the internal language can be unwound to
-  yield an external proof not referencing the internal language.
-
-  However, depending on the logical complexity of the statements occurring in a
-  given proof, the resulting external proof might be (much) more complex than
-  the internal proof. This is particularly the case if the proof involves
-  double negation, for much the same reason as that in computer science,
-  continuations can twist the control flow in nontrivial ways which are
-  sometimes hard to understand. It is in these cases where we can extract the
-  most value of the internal language, unlocking notions and proofs which might
-  otherwise be hard to obtain.
-
-  The slide shows a specific example. The internal statement
-  that~$A^\sim[X_1,\ldots,X_n]$ is anonymously Noetherian is quite simple; its
-  external translation is quite convoluted.}
-
-  \pause
-  \only<2>{\begin{center}
-    \includegraphics[width=0.9\textwidth]{hartshorne-tilde-construction} \\
-    \footnotesize
-    Robin Hartshorne. Algebraic Geometry. 1977.
-  \end{center}}
-  \pause
-
-  \vspace*{-1.5em}
-
-  \begin{columns}
-    \begin{column}{0.48\textwidth}
+  \vspace*{-0.5em}
+  and contains a \hil{mirror image} of~$A$, the sheaf of rings $A^\sim$.
+
+  \vspace*{-1.9em}
+
+  \begin{columns}[t]
+    \begin{column}{0.5\textwidth}
       \begin{varblock}{\textwidth}{}
         \justifying
-        Assuming the Boolean prime ideal theorem, a geometric sequent
-        ``$\forall \ldots \forall\_ (\cdots \Longrightarrow \cdots\!\,)$'',
+        Assuming the Boolean Prime Ideal Theorem, a first-order
+        formula ``$\forall \ldots \forall\_ (\cdots \Longrightarrow \cdots\!\,)$'',
         where the two subformulas may not contain~``$\Rightarrow$'' and~``$\forall$'',
-        holds for~$M^\sim$ iff it holds for all stalks~$M_\ppp$.
-      \end{varblock}
-
-      \vspace*{-1.5em}
+        holds for~$A^\sim$ iff it holds for all stalks~$A_\ppp$.
+      \end{varblock}
+
+      \vspace*{-2em}
 
       \begin{varblock}{\textwidth}{}
-        $M^\sim$ inherits any property of~$M$ which is \hil{localization-stable}.
+        $A^\sim$ inherits any property of~$A$ which is
+        \hil{localization-stable}.
       \end{varblock}
@@
     \begin{column}{0.5\textwidth}
-      \vspace*{1.5em}
-
-      If $A$ is reduced ($x^n = 0 \Rightarrow x = 0$):
-
-      \vspace*{-1.0em}
-
+      \vspace*{1.7em}
+
+      If~$A$ is reduced ($x^n = 0 \Rightarrow x = 0$):
+
+      \vspace*{-0.9em}
       \setbeamercolor{block body}{bg=red!30}
@@
       \begin{varblock}{\textwidth}{}
-        $A^\sim$ is a \hil{field}
-        (nonunits are zero).
+        $A^\sim$ is a \hil{field}.
 
@@
 
-  \visible<4-5>{\begin{tikzpicture}[overlay]
-    \draw[fill=white, draw=white, opacity=0.95] (-1,0) rectangle (\paperwidth,7.4);
-    \node[anchor=south west,inner sep=0] (image) at (0,0.8) {\vbox{
-      \only<4>{
+  \visible<2->{\begin{tikzpicture}[overlay]
+    \draw[fill=white, draw=white, opacity=0.85] (-1,0) rectangle (\paperwidth,8.0);
+    \node[anchor=south west,inner sep=0] (image) at (0,1.0) {\vbox{
+      \only<2>{
         \centering
@@
 
-      \only<5>{
-	\begin{varblock}{\textwidth}{Complexity reduction}
+      \only<3>{
+	\begin{varblock}{\textwidth}{}
 	  The external meaning of
 	  \[
-	    \Spec(A) \models
+	    \models
 	      \speak{$A^\sim[X_1,\ldots,X_n]$ is anonymously Noetherian}
@@
 \end{frame}
+\renewcommand{\insertframeextra}{}
+
+\section{Revisiting the test cases}
+\setbeamertemplate{headline}{\mynav{gray}{gray}{black}}
 
 \begin{frame}{Revisiting the test cases}
-  Let $A$ be a reduced ring.
-
-  \jnote{1}{This slide delivers on the promise made earlier: Using the internal
-  language of the little Zariski topos, we can reduce to the case of fields
-  without having to employ maximal ideals, prime ideals or minimal prime ideals.
-  Since the internal language machinery is itself constructive, the displayed
-  proofs can be unwound to yield external constructive proofs which don't
-  reference topos theory. Details on these external proofs can soon be found in
-  a forthcoming paper titled
-  \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-wlog.pdf}{Without
-  loss of generality, any reduced ring is a field}.}
-
-  \jnote{2}{The test case of Grothendieck's generic freeness lemma illustrates
-  that the internal language of toposes can help commutative algebra even if
-  one is not interested in constructivity issues. The previously known proofs
-  are somewhat long and somewhat convoluted; the new proof is arguably short
-  and simple.
-
-  Details on the internal proof are in Section~11.5 of
-  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
-  notes}. The
-  \fixedhref{https://arxiv.org/abs/1807.01231}{external
-  proof} obtained by unwinding the internal one is still quite direct, compared
-  to the previously published proofs, and interestingly follows a curious
-  course: It starts with verifying, in an inductive manner, that~$B$ and~$M$
-  are free; that~$A \to B$ is of finite presentation; and that~$M$ is finitely
-  presented as a~$B$-module. Then the assumption for~$f = 1$ is used, rendering
-  the prior steps moot, since over the zero ring any module is free and any
-  algebra is of finite presentation. External translations of internal proofs
-  which use double negation will always take such a course.}
-
-  \jnote{3}{Here's a rough sketch why the double negations occur. In
-  undergraduate linear algebra, we learn that any finitely generated vector
-  space is free (has a basis), by the following argument:
-  Let~$(v_1,\ldots,v_n)$ be a generating family. Either one of the generators can
-  be expressed as a linear combination of the others or not. In the latter
-  case, the family is linearly independent and we're done. In the former, we
-  remove the redundant generator and continue by induction.
-
-  This argument relies on the law of excluded middle and can therefore not put
-  to work in the internal universe as is. However, intuitionistically we do
-  have the weaker statement~$\neg\neg(\varphi \vee \neg\varphi)$, and we can
-  use this law in place of the law of excluded middle to intuitionistically
-  deduce that any finitely generated vector space does \emph{not not} have a
-  basis.}
-
-  \only<1>{\begin{columns}[t]
-    \begin{column}[t]{0.47\textwidth}
+  \vspace*{-1em}
+  Let~$A$ be a reduced commutative ring ($x^n = 0 \Rightarrow x = 0$). \\
+  Let~$A^\sim$ be its mirror image in the little Zariski topos.
+
+  \begin{columns}[t]
+    \begin{column}[t]{0.48\textwidth}
       \centering
 
-      \scalebox{0.8}{$\begin{pmatrix}
-        \cdot & \cdot \\
-        \cdot & \cdot \\
-        \cdot & \cdot
+      \scalebox{0.5}{$\begin{pmatrix}
+        \cdot & \cdot & \cdot & \cdot & \cdot \\
+        \cdot & \cdot & \cdot & \cdot & \cdot \\
+        \cdot & \cdot & \cdot & \cdot & \cdot
       \end{pmatrix}$}
-      \vspace*{-1em}
-
-      \begin{varblock}{\textwidth}{A baby application}
-        \justifying
-        Let~$M$ be a surjective matrix over~$A$ with more rows than columns.
-        Then~$1 = 0$ in~$A$.
-      \end{varblock}
-
-      \justifying
-      \textbf{Proof.} The matrix is surjective over the field~$A^\sim$.
-      This is a contradiction to basic linear algebra. Hence $\Spec(A) \models
-      \bot$, thus $1 = 0$ in $A$.
-    \end{column}
-    \begin{column}[t]{0.47\textwidth}
-      \centering
-
-      \scalebox{0.8}{$\begin{pmatrix}
-        \cdot & \cdot & \cdot & \cdot \\
-        \cdot & \cdot & \cdot & \cdot \\
-        \cdot & \cdot & \cdot & \cdot
-        \end{pmatrix}$}
-      \vspace*{-1em}
-
-      \begin{varblock}{\textwidth}{A child application}
+      \vspace*{-0.5em}
+
+      \begin{varblock}{\textwidth}{A baby example}
         \justifying
@@
       \justifying
-      \textbf{Proof.} The matrix is injective over the field~$A^\sim$. This is
-      a contradiction to basic linear algebra. Hence $\Spec(A) \models \bot$,
-      thus $1 = 0$ in $A$.
-    \end{column}
-  \end{columns}}
-
-  \pause
-
-  \vspace*{-5.3em}\hspace*{21em}\begin{minipage}{3cm}
-    $\xymatrixcolsep{3.5pc}\xymatrixrowsep{1.2pc}\xymatrix{
-      & \ar@{-}[d]^{\substack{\text{finitely}\\\text{generated}}} M \\
-      A \ar[r]_{\text{of finite type}} & B
-    }$
-  \end{minipage}
-  \vspace*{-1em}
-
+      \textbf{Proof.} $M$ is also injective as a matrix over~$A^\sim$.
+      Since~$A^\sim$ is a field, this is a contradiction by basic linear
+      algebra. Thus~$\models \bot$. This amounts to~$1 = 0$ in~$A$.
+    \end{column}
+
+    \begin{column}[t]{0.57\textwidth}
       \centering
 
-  \begin{varblock}{0.9\textwidth}{Generic freeness$\phantom{p}$}
-    \justifying
-    Let~$B$ be an~$A$-algebra of finite type ($\cong A[X_1,\ldots,X_n]/\aaa$). \\
-    Let~$M$ be a finitely generated~$B$-module ($\cong B^m/U$).
-
-    If~$f = 0$ is the only element of~$A$ such that
-    \begin{enumerate}
-      \item $B[f^{-1}]$ and $M[f^{-1}]$ are free modules over $A[f^{-1}]$,
-      \item $A[f^{-1}] \to B[f^{-1}]$ is of finite presentation and
-      \item $M[f^{-1}]$ is finitely presented as a module over~$B[f^{-1}]$,
-    \end{enumerate}
-    then $1 = 0$ in~$A$.
-  \end{varblock}
+      \includegraphics[height=1.9em]{generic-freeness}
       \vspace*{-0.5em}
 
-  \pause
-
-  \scalebox{0.9}{\begin{minipage}{1\textwidth}
-    \textbf{Proof.} In the little Zariski topos it's \hil{not not} the case that % \\[-1.4em]
-    \begin{enumerate}
-      \item $B^\sim$ and $M^\sim$ are free modules over $A^\sim$,
-      \item \vspace*{-0.3em}$A^\sim \to B^\sim$ is of finite presentation and
-      \item \vspace*{-0.3em}$M^\sim$ is finitely presented as a module over $B^\sim$,
-    \end{enumerate}
-    by basic linear algebra over the field $A^\sim$. The claim is precisely the
-    external translation of this fact.
-  \end{minipage}}
-\end{frame}
-
-\section{Algebraic geometry}
-
-\begin{frame}{Understanding algebraic geometry}
+      \begin{varblock}{\textwidth}{Generic freeness\phantom{p}}
+        \justifying
+        Let~$M$ be a finitely generated~$A$-module.
+        If~$f = 0$ is the only element of~$A$ such that~$M[f^{-1}]$ is a
+        free~$A[f^{-1}]$-module, then~$1 = 0$ in~$A$.
+      \end{varblock}
+      \vspace*{-0.1em}
+
+      \justifying
+      \textbf{Proof.} The claim amounts to \mbox{$\models
+      \text{``$M^\sim$}$}$\text{
+      is \hil{not not} free''}$. Since~$A^\sim$ is a field, this follows from
+      basic linear algebra.
+    \end{column}
+  \end{columns}
+\end{frame}
+
+\backupstart
+\setbeamertemplate{headline}{\mynav{gray}{gray}{gray}}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses}\end{minipage}}
+\begin{frame}
+  \bigskip
+  \centering
+  \includegraphics[height=3em]{heart}
+  \par
+  \raggedright
+
+  The Zariski topos and related toposes have applications in:
+  \begin{itemize}
+    \item classical algebra and classical algebraic geometry
+    \item constructive algebra and constructive algebraic geometry
+    \item synthetic algebraic geometry (``schemes are just sets'')
+  \end{itemize}
+
+  Connections with:
+  \begin{itemize}
+    \item understanding quasicoherence
+    \item the age-old mystery of nongeometric sequents
+  \end{itemize}
+
+\end{frame}}
+
+\begin{frame}[plain,c]
+  \centering%
+  \hil{Further reading}
+  \medskip
+
+  \framebox{\href{https://pizzaseminar.speicherleck.de/skript2/zariski-topos-klein.pdf}{\includegraphics[height=0.8\textheight]{fun-with-the-zariski-topos}}}\qquad%
+  \framebox{\href{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{\includegraphics[height=0.8\textheight]{phd-cover}}}
+  \par
+\end{frame}
+
+\begin{frame}{Applications in algebraic geometry}
   \vspace*{-1.5em}
@@
   \end{varblock}
-
-  \jnote{1}{One doesn't need to be an expert in topos theory in order to know
-  that many notions in algebraic geometry are inspired by notions in algebra
-  and that proofs in algebraic geometry often proceed by reducing to algebra.
-  The internal language is a way of making this connection precise: In many
-  cases, the former are simply interpretations of the latter internal
-  to~$\Sh(X)$. Because this connection is precise instead of informal,
-  additional value is gained: We can skip many basic proofs in algebraic
-  geometry because they're just externalizations of proofs in algebra carried
-  out internally to~$\Sh(X)$.}
-
-  \jnote{2}{A basic example is as follows. A short exact sequence of sheaves of
-  modules looks like a short exact sequence of plain modules from the internal
-  point of view of~$\Sh(X)$. If the two outer sheaves are of finite type, then
-  from the internal point of view, the two outer modules will look like
-  finitely generated modules. Because the standard proof of the proposition
-  quoted on the lower right is intuitionistically valid, it follows that, from
-  the internal point of view, the middle module is too finitely generated.
-  Consulting the dictionary a second time, this amounts to saying that the
-  middle sheaf is of finite type.
-
-  More details on this research program can be found in
-  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
-  notes}, partly reported on at the
-  \fixedhref{https://www.youtube.com/watch?v=7S8--bIKaWQ}{2015 IHÉS
-  conference}. Even though many important dictionary entries
-  are still missing (for instance pertaining to derived categories and
-  intersection theory), I believe that it is already at its current stage
-  useful to working algebraic geometers.
-
-  The next few slides illustrate an entirely different way of approaching
-  algebraic geometry using topos theory.}
-
-  \only<1>{
-    \centering
-    \rotatebox{90}{\tiny\scalebox{0.8}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
-    \includegraphics[width=0.7\textwidth]{external-internal-small}
-    \par
-  }
-
-  \pause
 
@@
     sheaf of sets & set \\
-    % sheaf of rings & ring \\
+    %sheaf of rings & ring \\
     sheaf of modules & module \\
@@
     spectrum of a sheaf of~$\O_X$-algebras & ordinary spectrum [with a twist] \\
-    big Zariski topos of $X$ & big Zariski topos of the ring $\O_X$ [with a twist] \\
-    higher direct image & sheaf cohomology \\
+    higher direct images & sheaf cohomology \\
     \bottomrule
@@
     \normalsize
-    \item or Grothendieck's functor-of-points account, where a scheme is a
-    functor~$\Ring \to \Set$.
+    \item or Grothendieck's functor-of-points account, where a scheme is a functor~$\mathrm{Ring} \to \mathrm{Set}$.
     \small\[ A \longmapsto \{ (x,y,z) \in A^3 \,|\, x^n+y^n-z^n=0 \} \]
@@
 
-  \jnote{1}{\footnotesize At the
-  \fixedhref{https://sbseminar.wordpress.com/2009/08/06/algebraic-geometry-without-prime-ideals/}{Secret
-  Blogging Seminar}, there was an insightful long-running discussion on the
-  merits of the two approaches. Two disadvantages of the approach using
-  locally ringed spaces is that the underlying topological spaces don't
-  actually parametrize ``honest'', ``geometric'' points, but the more complex
-  notion of irreducible closed subsets; and that they don't work well in a
-  constructive setting. (For this, they would have to be replaced by locally
-  ringed locales.)
-
-  The functorial approach is more economical, philosophically rewarding, and
-  works constructively. Given a functor~$F : \Ring \to \Set$, we imagine~$F(A)$
-  to be the set of~``$A$-valued points'' of the hypothetical scheme described
-  by~$F$, the set of ``points with coordinates in~$A$''. These sets have direct
-  geometric meaning. However, typically only field-valued points are easy
-  to describe. For instance, the functor representing projective~$n$-space is
-  given on fields by
-  \[ \begin{array}{r@{}c@{}l}
-  K &{}\longmapsto{}& \text{the set of lines through the origin in~$K^{n+1}$} \\
-  && \qquad \cong \{ [x_0 \hg \cdots \hg x_n] \,|\, \text{$x_i \neq 0$ for some~$i$} \},
-  \end{array} \]
-  whereas on general rings it is given by
-  \[ \begin{array}{r@{}c@{}l}
-    A &{}\longmapsto{}& \text{the set of quotients~$A^{n+1} \twoheadrightarrow P$,
-    where~$P$ is projective,} \\
-    && \qquad \text{modulo isomorphism}.
-  \end{array} \]
-  It is these more general kinds of points which impart a meaningful sense of
-  cohesion on the field-valued points, so they can't simply be dropped from
-  consideration.}
-
-  \jnote{2}{This tension is resolved by observing that the category of
-  functors~$\Ring \to \Set$ is a topos (the \emph{big Zariski topos} of~$\Spec(\ZZ)$)
-  and that we can therefore employ its internal language. This language takes
-  care of juggling stages behind the scenes. For instance, projective~$n$-space
-  can be described by the naive expression
-  \[ \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee \cdots \vee x_n
-  \neq 0 \}/(\affl)^\times. \]
-  This example illustrates our goal: to develop a synthetic account of
-  algebraic geometry, in which schemes are plain sets and morphisms between
-  schemes are maps between those sets. It turns out that there are many
-  similarities with the well-developed synthetic account of differential
-  geometry, but also important differences.}
-
-  \pause
-
-  \hil{Synthetic approach:} model schemes \hil{directly as sets} in
-  the internal universe of the \mbox{\hil{big Zariski topos}} of a base scheme.
+  \hil{Synthetic approach:} model schemes \hil{directly as sets} in a certain
+  nonclassical set theory, the internal universe of the \mbox{\hil{big Zariski
+  topos}} of a base scheme.
   \small
@@
 \end{frame}
+
+\fontsize{11pt}{13.2}
 
@@
 
+\fontsize{10pt}{12}
 {\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
 \begin{frame}{Classifying toposes in algebraic geometry}
-  \small
+  \footnotesize
   \centering
@@
 \end{frame}}
-
-% XXX: Mehr Leute zitieren!
-% XXX: Einordnung in Gesamtkontext (etwa SDG, Kock--Lawvere, Coquand & Co., ...)
-% XXX: Folie zu klassifizierenden Topoi
+\backupend
 

### 2018-12-05 slides-padova2018.tex
Some notes slides

@@
 \usepackage{etex}
+
+\usepackage{pgfpages}
+\setbeameroption{show notes on second screen}
+\setbeamertemplate{note page}[plain]
+\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}
+
+% Workaround for the issue described at
+% https://tex.stackexchange.com/questions/164406/beamer-using-href-in-notes.
+\newcommand{\fixedhref}[2]{\makebox[0pt][l]{\hspace*{\paperwidth}\href{#1}{#2}}\href{#1}{#2}}
 
@@
 \newcommand{\A}{\mathcal{A}}
+\newcommand{\B}{\mathcal{B}}
+\newcommand{\M}{\mathcal{M}}
 \renewcommand{\AA}{\mathbb{A}}
@@
 \renewcommand{\P}{\mathcal{P}}
+\newcommand{\aaa}{\mathfrak{a}}
 \newcommand{\ppp}{\mathfrak{p}}
@@
 
-\newcommand{\jnote}[2]{}
-
 \begin{document}
@@
         \justifying
-        \textbf{Proof.} See [Stacks Project].
+        \textbf{Proof.} See \href{https://stacks.math.columbia.edu/tag/051Q}{[Stacks Project]}.
       }
@@
   \end{columns}
+
+  \jnote{1}{The two displayed statements are trivial for fields. It is
+  therefore natural to try to reduce the general situation to the field
+  situation.}
+
+  \jnote{2}{The displayed proof, which could have been taken from any
+  standard textbook on commutative algebra, succeeds in this reduction
+  by employing proof by contradition and minimal prime ideals. However, this way
+  of reducing comes at a cost: It requires the Boolean Prime Ideal Theorem
+  (for ensuring the existence of a prime ideal and for ensuring that stalks at
+  minimal prime ideals are fields) and even the full axiom of choice (for
+  ensuring the existence of a minimal prime ideal).
+
+  We should hope that such a simple statement admits a more informative,
+  explicit, computational proof: There should be an explicit method for
+  transforming the given conditional equations expressing injectivity into the
+  equation~$1 = 0$. And indeed there is: Beautiful constructive proofs can
+  be found in Richman's note on
+  \fixedhref{https://www.ams.org/journals/proc/1988-103-04/S0002-9939-1988-0954974-5/S0002-9939-1988-0954974-5.pdf}{nontrivial
+  uses of trivial rings} and in the
+  \fixedhref{https://arxiv.org/abs/1605.04832}{recent textbook by Lombardi and
+  Quitté} on constructive commutative algebra.
+
+  The new reduction technique presented in this talk provides a way of
+  performing the reduction in an entirely constructive manner, avoiding the
+  axiom of choice. If so desired, resulting topos-theoretic proofs can be
+  unwound to yield fully explicit, topos-free, direct proofs.}
+
+  \jnote{3}{The baby example demonstrates that the reduction technique of this
+  talk is of interest to constructive commutative algebra. What about classical
+  commutative algebra? This is what the second example aims at.
+  Grothendieck's generic freeness lemma is an important theorem in algebraic
+  geometry, where it is usually stated in the following geometric form:
+
+  \vspace*{-1.2em}
+  \begin{indentblock}Let~$X$ be a reduced scheme. Let~$\B$ be
+  an~$\O_X$-algebra of finite type. Let~$\M$ be a~$\B$-module of finite type.
+  Then over a dense open,
+  \begin{enumerate}
+  \item[(a)] $\B$ and~$\M$ are locally free as sheaves of~$\O_X$-modules,
+  \item[(b)] $\B$ is of finite presentation as a sheaf of~$\O_X$-algebras and
+  \item[(c)] $\M$ is of finite presentation as a sheaf of~$\B$-modules.
+  \end{enumerate}\end{indentblock}
+  \vspace*{-1.2em}
+
+  All previously known proofs proceed in a series of reduction steps,
+  finally culminating in the case where~$A$ is a Noetherian integral domain.
+  They are somewhat convoluted (spanning several pages) and require nontrivial
+  prerequisites in commutative algebra.
+
+  Using the new reduction technique, there is a short (one-paragraph) and
+  conceptual proof of Grothendieck's generic freeness lemma. It is constructive
+  as a bonus; and if desired, one can unwind the resulting proof to obtain a
+  constructive proof which doesn't reference topos theory. The proof obtained
+  in this way is still an improvement on the previously known proofs, requiring
+  no advanced prerequisites in commutative algebra, and takes
+  \fixedhref{https://arxiv.org/abs/1807.01231}{about a page}.}
 \end{frame}
@@
     \end{itemize}
+    (Recall~$A[f^{-1}] = \bigl\{ \frac{u}{f^n} \,|\, u \in A, n \in \NN \bigr\}$.)
   }
@@
   \end{varblockextra}
+
+  \jnote{1}{In topos theory, we have lots of experience of ``changing
+  universes'' in order to ``force'' some statements to be true. However,
+  because the field condition we are aiming at is not a ``geometric sequent'', these
+  techniques do not work here. Hence we try to take it more slowly and devise a
+  semantics which forces the given ring only to be local.
+
+  The key insight is that \emph{locally} (in the sense of topology/geometry),
+  any ring is a local ring. That is, we may pretend that any given ring is
+  local if we are prepared to pass to numerous localizations during the course
+  of an argument. The semantics displayed on the next slide manages this
+  localization-juggling for us.}
 \end{frame}
@@
 \begin{frame}{The Kripke--Joyal semantics}
-  \small
-  \mbox{\!\!\!Recall~$A[f^{-1}] = \bigl\{ \frac{u}{f^n} \,|\, u \in A, n \in \NN \bigr\}$.
-  Let~``$\models \varphi$'' be short for~``$1 \models \varphi$''.}
+  \small\vspace*{-0.7em}
+  \only<1>{Let~$A$ be a ring (commutative, with unit). We recursively define
+  \[ f \models \varphi \quad \text{(``$\varphi$ holds away from the zeros of~$f$'')} \]
+  for elements~$f \in A$ and statements~$\varphi$. Write
+  ``$\models \varphi$'' to mean~$1 \models \varphi$.}
   \only<1>{\[ \renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\quad}c@{\quad}l@{}}
-    f \models \top &\text{iff}& \top \\
+    f \models \top &\text{is}& \text{true} \\
     f \models \bot &\text{iff}& \text{$f$ is nilpotent} \\
@@
   \end{array} \]}
+  \only<2->{Write ``$\models \varphi$'' to mean~$1 \models \varphi$.}
   \only<2->{\[ \renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\quad}c@{\quad}l@{}}
@@
   \end{array} \]}
+
+  \jnote{1}{The clause for~``$\vee$'' is made exactly in such a way as to
+  ensure, if~$x + y = 1$, that~$1 \models ((\exists z\?A^\sim\_ xz = 1) \vee
+  (\exists z\?A^\sim\_ yz = 1))$.
+
+  The definition of the semantics is reminiscient of Kripke and Beth models.
+  Indeed, it is a fragment of the Kripke--Joyal semantics of the \emph{internal
+  language of a topos}, and this general semantics encompasses Kripke and Beth
+  models as special cases.}
+
+  \jnote{2}{The soundness lemma states: If~$f \models \varphi$, and
+  if~$\varphi$ intuitionistically entails a further statement~$\psi$, then
+  also~$f \models \psi$. In this way we can \emph{reason} with the forcing
+  model, similarly as if~$A^\sim$ would actually exist as a ring instead of
+  merely being a convenient syntactic fiction.
+
+  If we want~$A^\sim$ to actually exist, not just as a figure of speech, then
+  we have to broaden our notion of existence and accept ring objects in
+  toposes. More on this on the next slide.
+
+  Irrespective of whether~$A$ is a local ring, its mirror
+  image~$A^\sim$ is always a local ring (that is, the axioms of what it means
+  to be a local ring hold under the translation rules specified by the
+  semantics). A basic application of this forcing model are local-to-global
+  principles. For instance:
+  \begin{itemize}\justifying
+    \item The statement ``the kernel of a surjective matrix over a local ring
+    is finite free'' admits a constructive proof. It therefore holds
+    for~$A^\sim$. Its external meaning is that the kernel of a surjective
+    matrix~$M$ over~$A$ is finite locally free (there exists a partition~$1 =
+    f_1 + \cdots + f_n$ such that for each~$i$, the localized
+    module~$(\operatorname{ker} M)[f_i^{-1}]$ is finite free).
+
+    \item The ring~$A$ is a Prüfer domain if and only if~$A^\sim$ is a Bézout
+    domain. Therefore any constructive theorem about Bézout domains entails a
+    corresponding theorem about Prüfer domains. Bézout domains are quite rare,
+    while Prüfer domains abound (for instance the ring of integers of any
+    number field is a Prüfer domain, even constructively so).
+  \end{itemize}}
 
@@
 {\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
-\begin{frame}{Classifying toposes in algebraic geometry}
+\begin{frame}{\!\!\!\mbox{Classifying toposes in algebraic geometry}}
   \footnotesize

### 2018-12-06 slides-padova2018.tex
Notes slides finished

@@
 
-  \visible<4>{
+  \visible<4->{
     \begin{changemargin}{-2.0em}{-0.5em}
@@
   \jnote{1}{The two displayed statements are trivial for fields. It is
-  therefore natural to try to reduce the general situation to the field
-  situation.}
+  therefore natural to try to reduce the general situation to that of fields.}
 
@@
   \fixedhref{https://arxiv.org/abs/1807.01231}{about a page}.}
+
+  \jnote{4}{Instead of passing from a given ring~$A$ to one of its
+  stalks~$A_\ppp$ or quotient rings~$A/\aaa$, the reduction technique presented
+  in this talk passes from~$A$ to a \emph{forcing model}~$A^\sim$.
+
+  Unlike stalks or quotient rings, which are honest rings, the forcing
+  model~$A^\sim$ is not a ring in the strict sense of the word: It doesn't
+  have an underlying \emph{set} of elements, but instead an underlying
+  \emph{sheaf} of elements. It is a \emph{ring object} in a certain category,
+  the \emph{little Zariski topos} of~$A$. But as long as we restrict to intuitionistic
+  reasoning, this difference is immaterial. Any intuitionistic theorem about
+  rings applies to~$A^\sim$ as if~$A$ were a proper ring.
+
+  Studying~$A^\sim$ is in fact the same as studying~$A$ from a certain
+  different, \emph{local} point of view. The precise meaning of this statement
+  will be explained on slide~3.}
+
+  \jnote{5}{Intuitionistic logic is the same as classical logic, but without:
+  \vspace*{-1em}
+  \begin{itemize}
+    \item the law of excluded middle: $\varphi \vee \neg\varphi$ \\[-1.5em]
+    \item the law of double negation elimination: $\neg\neg\varphi \Rightarrow \varphi$ \\[-1.5em]
+    \item the axiom of choice
+  \end{itemize}
+  \vspace*{-1em}
+  If one is unfamiliar with constructive mathematics, then doing without these
+  three laws seems unmotivated, rather peculiar and overtly restrictive. Here, the
+  restriction to intuitionistic logic is not by some philosophical choice.
+  Rather, it's by mathematical necessity. It's just a fact that in general, the
+  laws of classical logic don't apply to~$A^\sim$. (Assuming a classical
+  metatheory and assuming~$A$ to be Noetherian, they do iff~$A$ is of Krull
+  dimension~$\leq 0$.)
+
+  Luckily, vast amounts of commutative algebra work in the intuitionistic
+  setting, as for instance evidenced by the recent 1000$^+$-page
+  \fixedhref{https://arxiv.org/abs/1605.04832}{textbook by
+  Lombardi and Quitté}. This claim extends to statements which are usually
+  proven using maximal ideals or minimal ideal prime ideals and hence require
+  Zorn's lemma. Indeed, the technique presented in this talk allows to
+  constructivize some results of this kind.
+
+  Background on constructive mathematics can for instance be found in a
+  \fixedhref{https://video.ias.edu/members/1213/0318-AndrejBauer}{talk by
+  Andrej Bauer}
+  (\fixedhref{https://www.ams.org/journals/bull/2017-54-03/S0273-0979-2016-01556-4/S0273-0979-2016-01556-4.pdf}{written
+  notes}). The standard proof that~$\sqrt{2}$ is not rational is perfectly fine
+  in constructive mathematics.}
 \end{frame}
@@
 
-  \jnote{1}{In topos theory, we have lots of experience of ``changing
-  universes'' in order to ``force'' some statements to be true. However,
-  because the field condition we are aiming at is not a ``geometric sequent'', these
-  techniques do not work here. Hence we try to take it more slowly and devise a
-  semantics which forces the given ring only to be local.
+  \jnote{1}{In topos theory, we have lots of experience of \emph{changing
+  universes} in order to \emph{force} some statements to become true. However,
+  because the field condition we are aiming at is not a \emph{geometric
+  sequent}, these techniques do not work here. Hence we'll take it more slowly
+  and only devise a semantics which forces the given ring to be local.
+
+  The displayed definition of a local ring is, in the presence of the axiom of
+  choice, equivalent to the more common one (ring with exactly one maximal
+  ideal). In constructive mathematics, the displayed definition usually works
+  better.
 
@@
   of an argument. The semantics displayed on the next slide manages this
-  localization-juggling for us.}
+  localization-juggling for us.
+
+  By~$A[f^{-1}]$, we mean the localization of~$A$ away from~$f$. This
+  construction makes sense even if~$f$ is a zerodivisor, in which
+  case~$A[f^{-1}]$ is the zero ring.}
 \end{frame}
@@
 
-  Irrespective of whether~$A$ is a local ring, its mirror
+  The four lemmas displayed on this slide, as well as all the claims on
+  further slides, can be proven in very weak intuitionistic metatheories.}
+
+  \jnote{3}{Irrespective of whether~$A$ is a local ring, its mirror
   image~$A^\sim$ is always a local ring (that is, the axioms of what it means
@@
 
+  \jnote{1}{A topos is a special kind of category. Every topos has an
+  associated \emph{internal language} which can be used to do mathematics
+  \emph{internally to the topos}.
+
+  The prototypical example of a topos is the category~$\Set$. Doing mathematics
+  internally to~$\Set$ amounts to just doing mathematics in the usual sense.
+  A primer on the topos-theoretic landscape is contained in
+  \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-leipzig2018.pdf}{these
+  slides}. These slides also explain the geometric reason why the intermediate value
+  theorem fails in most toposes of sheaves, and why in the ef{}fective topos
+  any function~$\NN \to \NN$ is computable.}
+
+  \jnote{2-}{A free local ring~$A'$ over~$A$ is a local ring~$A'$
+  together with a ring homomorphism~$A \to A'$ such that any ring
+  homomorphim~$A \to R$ to a local ring uniquely factors over a local ring
+  homomorphism~$A' \to R$. (A ring homomorphism is \emph{local} iff it reflects
+  invertibility.)
+
+  Assuming the Boolean Prime Ideal Theorem, one can show that there is a free
+  local ring over~$A$ if and only if~$A$ has exactly one prime ideal. In this
+  case~$A$ is already local, and we can take~$A' \defeq A$. If we want every
+  ring to possess a free local ring over it, we need to accept ring objects of
+  different toposes than~$\Set$.
+
+  The little Zariski topos contains the \emph{generic filter} of~$A$.
+  Localizing~$A$ at this filter yields the desired free local ring. It is
+  precisely what was called~$A^\sim$ before. It is also known as the structure
+  sheaf of~$\Spec(A)$.}
+
   \pause
@@
       For a fixed ring~$R$, the localization $A' \defeq A[S^{-1}]$ with $S \defeq
-      \alpha^{-1}[R^\times]$ would do the job.
+      \alpha^{-1}[R^\times]$ would do the job. ($S$ is a \emph{filter}.)
       \medskip
@@
   $xy \in \fff \Leftrightarrow x \in \fff \wedge y \in \fff$
+
+  \jnote{1}{Any geometric theory has a \emph{classifying topos} which contains
+  the \emph{generic model} of that theory (any model in any topos is uniquely
+  the pullback of the generic one); if the theory under consideration is
+  propositional (doesn't have any sorts), then its classifying topos can be
+  chosen to be the topos of sheaves over a locale. One can also give a direct
+  account of classifying locales, as a pedagogical stepping stone to the full
+  theory of classifying toposes.
+
+  The slide contains a small lie: The classical definition of the spectrum of a
+  ring is via the set of prime ideals of~$A$, not filters. If the law of
+  excluded middle is available, there is no difference between these
+  definitions since the complement of a prime ideal is a filter and vice
+  versa.
+  One can also consider the classifying locale of prime ideals of~$A$.
+  Its associated topological space of points is the set of prime ideals
+  of~$A$ equipped with the \emph{constructible topology}.
+
+  In an intuitionistic (but still impredicative) context, any of the
+  (generalized) spaces of items~1--4 can be adopted as sensible definitions of
+  the spectrum. Item~5 is then a tautology. The classical definition of
+  the spectrum as a topological space doesn't work very well, because verifying
+  the universal property one expects of it requires the Boolean Prime Ideal
+  Theorem. Most dramatically, in some toposes there are rings which are not
+  trivial yet have neither prime ideals nor filters. The classical definition
+  yields in this case the empty space.}
 \end{frame}
@@
 
-  \visible<2->{\begin{tikzpicture}[overlay]
+  \jnote{1}{For working with~$A^\sim$, it's important to know how properties
+  of~$A$ relate to properties of~$A^\sim$.
+
+  The first displayed metatheorem justifies that, to a first approximation, the
+  forcing model~$A^\sim$ is a reification of all the stalks of~$A$ into a
+  single coherent entity. But crucially, this slogan is only correct for
+  properties which can be put in the displayed syntactical form (called
+  \emph{geometric sequents}). The reductive power of passing from~$A$
+  to~$A^\sim$ results from surprising non-geometric sequents which are
+  satisfied by~$A^\sim$ and not shared by~$A$, its localizations or its
+  quotients.
+
+  A slight generalization of the second metatheorem soups up a number of basic
+  lemmas of algebraic geometry, there stated in geometric language. For
+  instance, if~$M$ is finitely generated, then~$M^\sim$ is of finite type.
+  If~$M$ is finitely presented, then~$M^\sim$ is of finite presentation. If~$M$
+  is coherent, then~$M^\sim$ is coherent.}
+
+  \jnote{2}{Surprisingly and significantly, in case that~$A$ is reduced, there
+  are a number of non-geometric sequents validated by~$A^\sim$. These are
+  unique features of the forcing model.
+
+  $A^\sim$ is a field in the sense that zero is the only
+  noninvertible element.
+
+  $A^\sim$ has~$\neg\neg$-stable equality in the sense that
+  \[ \models \forall x\?A^\sim\_ \forall y\?A^\sim\_ \neg\neg(x = y) \Rightarrow x = y. \]
+  Classically, every set has~$\neg\neg$-stable equality; intuitionistically,
+  this is a special property of some sets. It's quite useful, as some theorems
+  of classical commutative algebra can only be proven intuitionistically when
+  weakened by double negation. The stability then allows, in some cases, to
+  obtain the original conclusion.
+
+  $A^\sim$ is \emph{anonymously Noetherian} in the sense that any of its ideals is
+  \mbox{\emph{not not}} finitely generated. A philosophically-motivated
+  constructivist might be offended by this notion, since it runs counter to the
+  maxim that constructive mathematics should be informative, telling us only
+  that there can't not be finite generating families. However, in the forcing
+  context it is a useful notion: Hilbert's basis theorem holds for it, and it
+  can be put to good use in the proof of (the general case of) Grothendieck's
+  generic freeness lemma.}
+
+  \jnote{3}{The field property was already observed in the 1970s
+  by Mulvey, but apparently back then its significance for applications was
+  overlooked and no deeper reason for this property was known. We now know
+  that it's a shadow of a forced higher-order property whose external translation
+  expresses that~$A^\sim$ is quasicoherent (details are in Section~3.3 of
+  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
+  notes}).}
+
+  \jnote{4}{Some properties of the forcing model, which are easy to state and
+  prove as properties about~$A^\sim$, have quite complex meanings when
+  unravelled to refer directly to~$A$. In this way the forcing model unlocks
+  observations which might otherwise be too unwieldy to manage.}
+
+  \visible<3->{\begin{tikzpicture}[overlay]
     \draw[fill=white, draw=white, opacity=0.85] (-1,0) rectangle (\paperwidth,8.0);
     \node[anchor=south west,inner sep=0] (image) at (0,1.0) {\vbox{
-      \only<2>{
+      \only<3>{
         \centering
@@
 
-      \only<3>{
+      \only<4>{
 	\begin{varblock}{\textwidth}{}
@@
   \end{columns}
+
+  \jnote{1}{With the baby example, we see that the new reduction technique
+  presented in this talk allows to reinterpret the core content of the
+  classical proof shown on slide~1 in a constructive fashion. Since the forcing
+  model was set up in a constructive way, we could mechanically extract an
+  explicit procedure from the new proof.
+
+  For the second example, a remark might be in order. A basic theorem of
+  undergraduate linear algebra is that any finitely generated vector space has
+  a basis. In this form, the theorem cannot quite be proven intuitionistically.
+  Indeed, the standard proof uses the law of excluded middle in the first step:
+  \begin{indentblock}
+    \scriptsize
+    Let~$(x_1,\ldots,x_n)$ be a given finite generating family.
+    It might be the case that one of the vectors~$x_i$ can be expressed as a
+    linear combination of the others, or not.
+    In the second case, the family is linearly independent and thus the vector
+    space is shown to have a basis.
+    In the first case, we continue with the smaller generating
+    family~$(x_1,\ldots,x_{i-1},x_{i+1},\ldots,x_n)$ in an inductive fashion.
+  \end{indentblock}
+  Intuitionistically, the law of excluded middle is not available; however its
+  double negation ($\neg\neg(\varphi \vee \neg\varphi)$) is. Threading the
+  double negation through the rest of the proof yields the following theorem of
+  intuitionistic linear algebra: Any finitely generated vector space is
+  \emph{not not} free. It is this theorem which the proof on the slide refers
+  to.}
 \end{frame}
@@
 
-\end{frame}}
-
-\begin{frame}[plain,c]
-  \centering%
+  \jnote{1}{Summarizing, we can associate to any reduced ring~$A$ the forcing
+  model~$A^\sim$.
+  \begin{itemize}
+    \item It has the pleasant property that it is a field.
+    \item Reasoning about it requires that we restrict ourselves to
+    intuitionistic logic.
+  \end{itemize}
+  Depending on the application, this trade-off can not be useful at all, or be
+  quite valuable. It certainly is so for proving Grothendieck's generic
+  freeness lemma, simplifying a multi-page argument to a single and conceptual
+  paragraph. (The previous slide only showed a fragment of the general
+  statement of Grothendieck's generic freeness lemma. The technique outlined in
+  these slides is amenable to the full version. Details are in Section~11.5 of
+  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
+  notes}.)
+
+  The ideas underlying this new reduction technique can also be used for
+  different purposes: for understanding, in a rigorous way, notions of algebraic
+  geometry as notions of algebra, and for developing a synthetic account of
+  algebraic geometry. This is very briefly touched upon on the next two
+  slides.}
+
+  \jnote{2}{
+    \centering
     \hil{Further reading}
@@
 
-  \framebox{\href{https://pizzaseminar.speicherleck.de/skript2/zariski-topos-klein.pdf}{\includegraphics[height=0.8\textheight]{fun-with-the-zariski-topos}}}\qquad%
-  \framebox{\href{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{\includegraphics[height=0.8\textheight]{phd-cover}}}
+    \framebox{\fixedhref{https://pizzaseminar.speicherleck.de/skript2/zariski-topos-klein.pdf}{\includegraphics[height=0.8\textheight]{fun-with-the-zariski-topos}}}\qquad%
+    \framebox{\fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{\includegraphics[height=0.8\textheight]{phd-cover}}}
     \par
-\end{frame}
+  }
+\end{frame}}
 
@@
   \end{columns}
+
+  \jnote{1}{One doesn't need to be an expert in topos theory in order to know
+  that many notions in algebraic geometry are inspired by notions in algebra
+  and that proofs in algebraic geometry often proceed by reducing to algebra.
+
+  If~$X$ is a scheme, the internal language of the topos~$\Sh(X)$ is a way of
+  making this connection precise: In many cases, the former are simply
+  interpretations of the latter internal to~$\Sh(X)$. Because this connection
+  is precise instead of informal, additional value is gained: We can skip many
+  basic proofs in algebraic geometry because they're just externalizations of
+  proofs in algebra carried out internally to~$\Sh(X)$.}
+
+  \jnote{2}{A basic example is as follows. A short exact sequence of sheaves of
+  modules looks like a short exact sequence of plain modules from the internal
+  point of view of~$\Sh(X)$. If the two outer sheaves are of finite type, then
+  from the internal point of view, the two outer modules will look like
+  finitely generated modules. Because the standard proof of the proposition
+  quoted on the lower right is intuitionistically valid, it follows that, from
+  the internal point of view, the middle module is too finitely generated.
+  Consulting the dictionary a second time, this amounts to saying that the
+  middle sheaf is of finite type.
+
+  A more advanced example is: The theorem ``any finitely generated vector space
+  does \emph{not not} have a basis'' of constructive linear algebra entails, by
+  interpretation in~$\Sh(X)$, that any sheaf of finite type over a reduced
+  scheme is finite locally free on a \emph{dense open subset}.
+
+  More details on this research program can be found in
+  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
+  notes}, partly reported on at the
+  \fixedhref{https://www.youtube.com/watch?v=7S8--bIKaWQ}{2015 IHÉS topos
+  theory conference}. Even though many important dictionary entries are still
+  missing (for instance pertaining to derived categories and intersection
+  theory), I believe that it is already in its current form useful to working
+  algebraic geometers.
+
+  The next slide illustrates a further, different way of approaching algebraic
+  geometry using topos theory.}
 \end{frame}
@@
 
+  \jnote{1}{At the
+  \fixedhref{https://sbseminar.wordpress.com/2009/08/06/algebraic-geometry-without-prime-ideals/}{Secret
+  Blogging Seminar}, there was an insightful long-running discussion on the
+  merits of the two approaches to algebraic geometry. Two disadvantages of the approach using
+  locally ringed spaces is that the underlying topological spaces don't
+  actually parametrize ``honest'', ``geometric'' points, but the more complex
+  notion of irreducible closed subsets; and that they don't work well in a
+  constructive setting. (For this, they would have to be replaced by locally
+  ringed locales.)
+
+  The functorial approach is more economical, philosophically rewarding, and
+  works constructively. Given a functor~$F : \Ring \to \Set$, we imagine~$F(A)$
+  to be the set of~``$A$-valued points'' of the hypothetical scheme described
+  by~$F$, the set of ``points with coordinates in~$A$''. These sets have direct
+  geometric meaning. However, typically only field-valued points are easy
+  to describe. For instance, the functor representing projective~$n$-space is
+  given on fields by
+  \[ \begin{array}{r@{}c@{}l}
+  K &{}\longmapsto{}& \text{the set of lines through the origin in~$K^{n+1}$} \\
+  && \qquad \cong \{ [x_0 \hg \cdots \hg x_n] \,|\, \text{$x_i \neq 0$ for some~$i$} \},
+  \end{array} \]
+  whereas on general rings it is given by
+  \[ \begin{array}{r@{}c@{}l}
+    A &{}\longmapsto{}& \text{the set of quotients~$A^{n+1} \twoheadrightarrow P$,
+    where~$P$ is projective,} \\
+    && \qquad \text{modulo isomorphism}.
+  \end{array} \]
+  It is these more general kinds of points which impart a sense of
+  cohesion on the field-valued points, so they can't simply be dropped from
+  consideration.\bigskip}
+
+  \pause
+
   \hil{Synthetic approach:} model schemes \hil{directly as sets} in a certain
@@
   \[ \{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]
+
+  \jnote{2}{This tension is resolved by observing that the category of
+  functors~$\Ring \to \Set$ is a topos (the \emph{big Zariski topos} of~$\Spec(\ZZ)$)
+  and that we can therefore employ its internal language. This language takes
+  care of juggling stages behind the scenes. For instance, projective~$n$-space
+  can be described by the naive expression
+  \[ \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee \cdots \vee x_n
+  \neq 0 \}/(\affl)^\times. \]
+  This example illustrates the goal: to develop a synthetic account of
+  algebraic geometry, in which schemes are plain sets and morphisms between
+  schemes are maps between those sets. It turns out that there are many
+  similarities with the well-developed synthetic account of differential
+  geometry, but also important differences, and it also turns out that
+  synthetic algebraic geometry has close connections to a certain age-old
+  unsolved problem in topos theory, the \emph{mystery of nongeometric
+  sequents}.
+
+  Details are in
+  \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-como2018.pdf}{this
+  set of slides}.}
 \end{frame}
 
-\fontsize{11pt}{13.2}
-
-\begin{frame}{The big Zariski topos}
-  Let~$S$ be a fixed base scheme.
-  \begin{varblock}{0.9\textwidth}{Definition}
-    The \hil{big Zariski topos} $\Zar(S)$ of a scheme~$S$ is equivalently
-    \begin{enumerate}
-      \item the topos of sheaves over~$(\Aff/S)_{\mathrm{lofp}}$,
-      \item the classifying topos of local rings over~$S$ or
-      \item the classifying~$\Sh(S)$-topos of local~$\O_S$-algebras which are
-      local over~$\O_S$.
-    \end{enumerate}
-  \end{varblock}
-
-  \begin{itemize}
-    \item For an~$S$-scheme~$X$, its functor of points $\ull{X} =
-    \Hom_S(\cdot,X)$ is an object of~$\Zar(S)$. It feels like \hil{the set of
-    points} of~$X$.
-    \item In particular, there is the ring object~$\affl$ with~$\affl(T) =
-    \O_T(T)$.
-    \item This ring object is a \hil{field}: nonzero implies
-    invertible. \\{} [Kock 1976]
-  \end{itemize}
-
-  \jnote{1}{The objects of the category~$(\Aff/S)_{\mathrm{lofp}}$ are
-  morphisms of the form~$\Spec(R) \to S$ which are locally of finite
-  presentation. (Other choices of resolving set-theoretical issues of size are
-  also possible.)
-
-  A functor~$F : (\Aff/S)_{\mathrm{lofp}}^\op \to \Set$ is a sheaf
-  for the Zariski topology if and only if the diagram
-  \[ F(T) \to
-    \prod_i F(U_i) \rightrightarrows
-    \prod_{j,k} F(U_j \cap U_k) \]
-  is a limit diagram for any open covering~$T = \bigcup_i U_i$ of any
-  scheme~$T \in (\Aff/S)_{\mathrm{lofp}}$.
-
-  In the case that~$S = \Spec(A)$ is affine, the big Zariski topos of~$S$ is also
-  simply called ``big Zariski topos of~$A$''. It is a subtopos of the topos of
-  functors $\mathrm{Alg}(A) \to \Set$ and classifies local~$A$-algebras.
-
-  From the internal point of view of~$\Sh(S)$, the sheaf~$\O_S$ of rings is
-  just an ordinary ring, and we can construct internally to~$\Sh(S)$ the big
-  Zariski topos of~$\O_S$. Externally, this construction will yield a certain
-  bounded topos over~$\Sh(S)$. However, as indicated on the slide, this topos
-  will \emph{not} coincide with the true big Zariski topos of~$S$. To
-  construct the true big Zariski topos, we have to build, internally
-  to~$\Sh(S)$, the classifying topos of local \emph{and local-over-$\O_S$}
-  $\O_S$-algebras.}
-\end{frame}
-
-\begin{frame}{Synthetic constructions}
-  \small
-  $\hil{$\boldsymbol{\mathbb{A}^n}$} = (\affl)^n = \affl \times \cdots \times \affl$
-
-  $\begin{array}{@{}r@{}c@{}l@{}}
-    \hil{$\boldsymbol{\mathbb{P}^n}$} &\phantom{}=\phantom{}& \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee
-    \cdots \vee x_n \neq 0 \}/(\affl)^\times \\
-    &\cong& \text{set of one-dimensional subspaces of~$(\affl)^{n+1}$} \\
-    &&\qquad \text{(with~$\O(-1) = (\ell)_{\ell \? \mathbb{P}^n}$, $\O(1) =
-    (\ell^\vee)_{\ell \? \mathbb{P}^n}$)}
-  \end{array}$
-
-  $\hil{$\boldsymbol{\Spec(R)}$} = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
-    \text{set of $\affl$-valued points of $R$}$
-
-  $\hil{$\boldsymbol{TX}$} = X^\Delta$, where $\Delta = \{ \varepsilon \? \affl \,|\, \varepsilon^2 = 0 \}$
-
-  A subset $U \subseteq X$ is \hil{qc-open} if and only if for any $x : X$
-  there exist $f_1,\ldots,f_n \? \affl$ such that $x \in U \Longleftrightarrow
-  \exists i\_ f_i \neq 0$.
-
-  A \hil{synthetic affine scheme} is a set which is in bijection
-  with~$\Spec(R)$ for some synthetically quasicoherent~$\affl$-algebra~$R$.
-
-  A \hil{finitely presented synthetic scheme} is a set which can be covered by
-  finitely many qc-open f.p. synthetic affine schemes~$U_i$ such that the
-  intersections~$U_i \cap U_j$ can be covered by finitely many qc-open f.p.
-  synthetic affine schemes.
-
-  \jnote{1}{In the internal universe of the big Zariski topos of a base
-  scheme~$S$, $S$-schemes can simply be modeled by sets (enjoying the special
-  property that, in a certain precise sense, they are locally affine). This
-  slide expresses some of the basic constructions of~$S$-schemes in that
-  language.
-
-  Particularly nice are the following items.
-  \begin{itemize}\justifying
-    \item Projective~$n$-space can be given by the any of the two quite naive
-    expressions displayed on the slide.
-    \item Let~$X$ be an~$S$-scheme. We often think about a sheaves
-    of~$\O_X$-modules over~$X$ by their fibers; but for a rigorous
-    treatment in the standard foundations, we have to take the full sheaf
-    structure into account; the fibers do not determine a sheaf
-    uniquely.
-
-    From the internal point of view of~$\Zar(S)$, a sheaf of~$\O_X$-modules is
-    indeed simply a family of~$\affl$-modules, one~$\affl$-module for each
-    element of~$\ull{X}$. The slide illustrates how we can define the Serre
-    twisting sheaves in this language.
-  \end{itemize}}
-
-  \jnote{2}{\begin{itemize}\justifying
-    \item The spectrum of an~$\affl$-algebra can be given by the naive
-    expression displayed on the slide. It looks like this expression can't be
-    right, ignoring any non-maximal ideals; however, it is.
-    \item The big Zariski topos of an~$S$-scheme~$X$ is, from the internal
-    point of view of~$\Zar(S)$, simply the slice topos~$\Set/\ull{X}$. Hence to
-    give an~$X$-scheme simply amounts to giving an~$\ull{X}$-indexed family of
-    sets.
-  \end{itemize}
-
-  Synthetic algebraic geometry has been developed up to the point of étale
-  geometric morphisms. Much remains to be done: For instance, as of yet there
-  is only an account of Čech methods for computing cohomology, there is not yet
-  a synthetic treatment of true cohomology. Derived categories and intersection
-  theory are also missing.}
-\end{frame}
-
-\begin{frame}{Relations between the Zariski toposes}
-  The big Zariski topos is a topos over the small Zariski topos:
-  \[ \begin{array}{crcl}
-    \pi : & \Zar(A) &\longrightarrow& \Spec(A) \\
-    & \text{local~$A$-algebra~$(A \xrightarrow{\alpha} B)$} &\longmapsto& (A \to A[(\alpha^{-1}[B^\times])^{-1}])
-  \end{array} \]
-  This morphism is \hil{connected} ($\pi^{-1}$ is fully faithful) and
-  \hil{local}, so there is a preinverse
-  \[ \begin{array}{rcl}
-    \Spec(A) &\longrightarrow& \Zar(A) \\
-    \text{local localization~$(A \to B)$} &\longmapsto& (A \to B)
-  \end{array} \]
-  which is a subtopos inclusion inducing an idempotent monad~$\sharp$ and an
-  idempotent comonad~$\flat$ on~$\Zar(S)$.
-
-  \begin{itemize}
-  \justifying
-    \item Internally to~$\Zar(S)$, $\Spec(S)$ can be constructed as the
-    \hil{largest subtopos} where~$\flat \affl \to \affl$ is bijective.
-    \item Internally to~$\Spec(S)$, $\Zar(S)$ can be constructed as the
-    \hil{classifying topos} of local~$\O_S$-algebras which are local over~$\O_S$.
-    \item $\Zar(A)$ is the \hil{lax pullback}~$(\Set
-    \Rightarrow_{\Set[\Ring]} \Set[\LocRing])$.
-  \end{itemize}
-
-  \jnote{1}{Let~$A$ be a ring. By definition, we obtain a geometric
-  morphism~$\Set \to \Set[\Ring]$ into the classifying topos of rings.
-  There is also a geometric morphism~$\Set[\LocRing] \to \Set[\Ring]$, obtained
-  by realizing that any local ring is in particular a ring. These morphisms fit
-  together in a lax pullback square as follows:
-  \[ \xymatrix{
-    \Zar(A) \ar[r] \ar[d] & \Set[\LocRing] \ar[d] \\
-    \Set \ar[r] \ar@{}[ur]^(.3){}="a"^(.7){}="b" \ar@{=>} "a";"b" & \Set[\Ring]
-  } \]
-  This observation is joint with Peter Arndt and Matthias Hutzler.
-
-  Incidentally, the pseudo pullback of the morphism~$\Set[\LocRing] \to
-  \Set[\Ring]$ along~$\Set \to \Set[\Ring]$ is not very interesting: It's the
-  largest subtopos of~$\Set$ where~$A$ is a local ring. Assuming the law of
-  excluded middle, this subtopos is either the trivial topos (if~$A$ is not
-  local) or~$\Set$ (if~$A$ is local).
-
-  There is also a way of realizing the little Zariski topos of~$A$ as a pseudo
-  pullback, exploiting that the (localic) spectrum construction is geometric.
-  See Section~12.6 of
-  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
-  notes} for details.
-  }
-\end{frame}
-
-\section{The mystery of nongeometric sequents}
-
-\begin{frame}{Properties of the affine line}
-  \begin{itemize}
-%   \item $\affl$ is a local ring:
-%   \[ 1 \neq 0 \qquad\qquad \text{$x + y$ \inv} \Longrightarrow \text{$x$ \inv} \vee
-%   \text{$y$ \inv} \]
-    \item $\affl$ is a field:
-    \begin{align*}
-      \neg(x = 0) &\Longleftrightarrow \text{$x$ invertible \quad [Kock 1976]} \\
-      \neg(\text{$x$ invertible}) &\Longleftrightarrow \text{$x$ nilpotent}
-    \end{align*}
-    \item $\affl$ satisfies the axiom of microaffinity: Any map $f : \Delta \to
-    \affl$ is of the form~$f(\varepsilon) = a + b \varepsilon$ for unique
-    values~$a,b \? \affl$, where~$\Delta = \{ \varepsilon \? \affl \,|\,
-    \varepsilon^2 = 0 \}$.
-    \bigskip
-    \item Any map $\affl \to \affl$ is a polynomial function.
-    \bigskip
-    \item $\affl$ is anonymously algebraically closed: Any monic polynomial
-    does \emph{not not} have a zero.
-  \end{itemize}
-
-  \jnote{1}{The axiom of microaffinity is a special instance of the
-  \emph{Kock--Lawvere axiom} known from synthetic differential geometry. We'll
-  see on the next slide that~$\affl$ validates an unusually strong form of the
-  Kock--Lawvere axiom, not at all satisfied in the usual well-adapted models of
-  synthetic differential geometry.
-
-  The fact that, internally to~$\Zar(S)$, any map~$\affl \to \affl$ is a
-  polynomial can be seen as a formal version of the general motto that in
-  algebraic geometry, ``morphisms are polynomials''.}
-\end{frame}
-
-\begin{frame}{Synthetic quasicoherence}
-  Recall~$\Spec(R) = \Hom_{\mathrm{Alg}(\affl)}(R, \affl)$ and consider the statement
-  \[ \text{``the canonical map~$
-    \begin{array}[t]{rcl}
-      R &\longrightarrow& (\affl)^{\Spec(R)} \\
-      f &\longmapsto& (\alpha \mapsto \alpha(f))
-    \end{array}
-  $ is bijective''}.
-  \]
-  \vspace*{-1em}
-
-  \begin{itemize}
-    \item True for~$R = \affl[X]/(X^2)$ (microaffinity).
-    \item True for~$R = \affl[X]$ (every function is a polynomial).
-    \item True for \hil{any} finitely presented~$\affl$-algebra~$R$.
-  \end{itemize}
-
-  Any known property of~$\affl$ follows from this \\
-  \hil{synthetic quasicoherence}.
-
-  \centering
-  \begin{varblock}{0.6344\textwidth}{}
-    \centering
-    \hil{the mystery of nongeometric sequents}
-    \par
-  \end{varblock}
-
-  \jnote{1}{Let~$R$ be an~$\affl$-algebra. An element~$f \in R$ induces
-  an~$\affl$-valued function on~$\Spec(R)$; functions of this form can
-  reasonably be called ``algebraic''. In a synthetic context, there should be
-  no other~$\affl$-valued functions on~$\Spec(R)$ as these algebraic ones, and
-  different algebraic expressions should yield different functions. This is
-  precisely what the bijectivity of the displayed map expresses (in a positive
-  way).
-
-  In synthetic differential geometry, the closest cousin of synthetic algebraic
-  geometry, the analogue of the displayed map is only bijective for Weil
-  algebras such as~$\affl[X]/(X^2)$ or $\affl[X,Y]/(X^2,XY)$, not for arbitrary
-  finitely presented~$\affl$-algebras. This is a major difference to synthetic
-  differential geometry.}
-
-  \jnote{2}{The notion of synthetic quasicoherence is interesting for a number
-  of reasons:\vspace*{-1.2em}
-  \begin{itemize}\justifying
-    \item All currently known properties of~$\affl$, such as all the properties
-    listed on the previous slide, follow from the statement that~$\affl$ is
-    synthetically quasicoherent.
-
-    For instance, here is how we can verify the field property. Let~$x \?
-    \affl$ such that~$x \neq 0$. Set~$R = \affl/(x)$. Then~$\Spec(R) =
-    \emptyset$. Thus~$(\affl)^{\Spec(R)}$ is a singleton. Hence~$R = 0$.
-    Therefore~$x$ is invertible.
-    \item Given an~$\affl$-module~$E$, we can formulate the following variant
-    of the axiom of synthetic quasicoherence: ``For any finitely
-    presented~$\affl$-algebra~$R$, the canonical map~$R \otimes_\affl E \to
-    E^{\Spec(R)}$ is bijective.'' This axiom is satisfied if and only if~$E$ is
-    induced by a quasicoherent sheaf of~$\O_S$-modules.
-    \item The notion of synthetic quasicoherence is central to synthetic
-    algebraic geometry. The notions of synthetic open immersions, closed
-    immersion, schemes and several others all refer to synthetic
-    quasicoherence.
-  \end{itemize}}
-
-  \jnote{3}{An analogue of synthetic quasicoherence holds in the classifying
-  topos of rings, demonstrating that even presheaf toposes can validate
-  interesting nontrivial nongeometric sequents.
-
-  We believe that an analogue of synthetic quasicoherence holds for the generic
-  model of any geometric theory. This is work in progress. If true, this would
-  yield a major source of nongeometric sequents in classifying toposes. Because
-  of the many applications on nongeometric sequents, it's very desirable to
-  possess such a source.
-
-  The mystery of nongeometric sequents is this: On the one hand, they are very
-  useful to have because of surprising applications; on the other hand, they
-  are as of yet quite elusive.}
-\end{frame}
-
-\fontsize{10pt}{12}
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
-\begin{frame}{\!\!\!\mbox{Classifying toposes in algebraic geometry}}
-  \footnotesize
-  \centering
-  \begin{tabular}{ll}
-    \toprule
-    (big) topos & classified theory \\\midrule
-    Zariski & local rings [Hakim 1972] \\[0.6em]
-    étale & separably closed local rings [Hakim 1972, Wraith 1979] \\[0.6em]
-    \usebeamercolor[fg]{item}{fppf} & \usebeamercolor[fg]{item}{fppf-local rings} \\
-    & \textcolor{red!90}{(conjecturally: algebraically closed local rings)} \\[0.6em]
-    \textcolor{red!80}{ph} & \textcolor{red!80}{?? (conjecturally: algebraically closed valuation rings} \\
-    & \qquad \textcolor{red!80}{validating the projective Nullstellensatz)} \\[0.6em]
-    \usebeamercolor[fg]{item}{surjective} & \usebeamercolor[fg]{item}{algebraically closed geometric fields} \\[0.6em]
-    \textcolor{red!80}{$\neg\neg$} & \textcolor{red!80}{?? (conjecturally: algebraically closed geometric} \\
-    & \qquad \textcolor{red!80}{fields which are integral over the base)} \\[0.6em]
-    \usebeamercolor[fg]{item}{infinitesimal} & \usebeamercolor[fg]{item}{local algebras together with a nilpotent ideal [Hutzler 2018]} \\[0.6em]
-    \textcolor{red!100}{crystalline} & \textcolor{red!100}{??} \\
-    \bottomrule
-  \end{tabular}
-
-  \jnote{1}{Toposes, and also more specifically classifying toposes, originated
-  in algebraic geometry. It is therefore deeply embarrassing that as of now,
-  still very little is known about the theories classified by the major toposes
-  in active use by algebraic geometers.
-  \begin{center}
-    \includegraphics[width=0.99\textwidth]{wraith-generic-galois-theory} \\
-    \footnotesize
-    Gavin Wraith. Generic Galois theory of local rings. 1979. \\
-    (A cult classic and must-read for anyone interested in \\
-    the intersection of topos theory and algebraic geometry.)
-  \end{center}}
-
-  \jnote{2}{For almost forty years, only the big Zariski topos and its étale
-  subtopos were understood in that way.
-  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{These~2017
-  notes} answer the question for the fppf topology and the surjective topology
-  (in Section~21) and state conjectures for the ph topology and the
-  double negation topology. However, while good to have, the answer for the
-  fppf topology remains unsatisfactory, since Wraith's conjecture that the fppf
-  topos classifies the simpler theory of algebraically closed local rings has
-  neither been confirmed nor refuted.
-
-  A couple of weeks ago, Matthias Hutzler managed to determine the theory
-  classified by the big infinitesimal topos of a ring~$A$: It classifies
-  pairs~$(B,\aaa)$ consisting of a local~$A$-algebra~$B$ and a nilpotent
-  ideal~$\aaa \subseteq B$. Details will be in his forthcoming Master's thesis.
-  He is currently working on answering the question for the closely related big
-  crystalline topos.
-
-  It will be exciting to learn what the crystalline topos and the many other
-  toposes in algebraic geometry classify; how algebraic geometry can profit
-  from these discoveries; and which new flavors of synthetic algebraic geometry
-  they unlock.}
-\end{frame}}
 \backupend

### 2018-12-06 slides-padova2018.tex
Final changes

@@
   \par
+
+  \jnote{1}{We can associate to any reduced ring~$A$ a \emph{forcing
+  model}~$A^\sim$.
+  \begin{itemize}
+    \item The forcing model has the pleasant property that it is a field.
+    \item Reasoning about it requires that we restrict ourselves to
+    intuitionistic logic.
+  \end{itemize}
+  Details are on the following slides.}
 \end{frame}}
@@
   standard textbook on commutative algebra, succeeds in this reduction
-  by employing proof by contradition and minimal prime ideals. However, this way
+  by employing proof by contradiction and minimal prime ideals. However, this way
   of reducing comes at a cost: It requires the Boolean Prime Ideal Theorem
@@
   We should hope that such a simple statement admits a more informative,
-  explicit, computational proof: There should be an explicit method for
+  explicit, computational proof not employing transfinite methods: There should
+  be an explicit method for
   transforming the given conditional equations expressing injectivity into the
@@
   the \emph{little Zariski topos} of~$A$. But as long as we restrict to intuitionistic
-  reasoning, this difference is immaterial. Any intuitionistic theorem about
-  rings applies to~$A^\sim$ as if~$A$ were a proper ring.
+  reasoning, this difference is immaterial. A metatheorem displayed on a later
+  slide states that any intuitionistic theorem about
+  rings applies to~$A^\sim$ just as if~$A$ were a proper, ordinary ring.
 
@@
   By~$A[f^{-1}]$, we mean the localization of~$A$ away from~$f$. This
-  construction makes sense even if~$f$ is a zerodivisor, in which
+  construction makes sense even if~$f$ is a zero divisor, in which
   case~$A[f^{-1}]$ is the zero ring.}
@@
 
-  The definition of the semantics is reminiscient of Kripke and Beth models.
+  The definition of the semantics is reminiscent of Kripke and Beth models.
   Indeed, it is a fragment of the Kripke--Joyal semantics of the \emph{internal
@@
   together with a ring homomorphism~$A \to A'$ such that any ring
-  homomorphim~$A \to R$ to a local ring uniquely factors over a local ring
+  homomorphism~$A \to R$ to a local ring uniquely factors over a local ring
   homomorphism~$A' \to R$. (A ring homomorphism is \emph{local} iff it reflects
@@
   \begin{itemize}
-    \item It has the pleasant property that it is a field.
+    \item The forcing model has the pleasant property that it is a field.
     \item Reasoning about it requires that we restrict ourselves to

### 2019-01-09 paper-filmat.tex
to do
Some details on the relation to models of set theory

@@
 \noindent
+
+\section{Unsorted}
+
+Toposes can be pictured as ``universes in which we can do mathematics in'' in
+much the same way as models of set theory can be viewed in this way. In fact,
+to any model~$M$ of a set theory such as~$\mathbf{ZF}$ or~$\mathbf{ZFC}$, there
+is a topos~$\Set_M$ such that a statement holds in the internal language
+of~$\Set_M$ if and only if it holds in~$M$.\footnote{The topos~$\Set_M$ can be
+described as follows: Its objects are the elements of~$M$, that is the things
+which~$M$ believes to be sets, and its morphisms are those things which~$M$
+believes to be maps. The topos~$\Set_M$ validates the axioms of
+$\mathbf{ETCS}$~\cite{XXX}, and for models which are not elementarily
+equivalent, their associated toposes will not be equivalent.} This
+includes~$V$, the trivial inner model consisting of all sets there are, in case
+such a thing exists in one's chosen ontology; in this case the associated
+topos~$\Set_V$ is the category of all sets, and a statement holds in~$\Set_V$
+if and only if it's true in the ordinary mathematical sense.
+
+Carrying the analogy further, just as using forcing and other techniques we can
+construct new models of set theory from given ones, thereby exploring the
+set-theoretic multiverse, we can construct new toposes from given toposes.
+However, there are two important differences between the notion of mathematical
+universe as provided by toposes and as provided by models of set theory, both
+regarding the subject matter and the reasons for why we are interested in them.
+
+Firstly, toposes are more general than models of set theory. By definition, a
+model of~$\mathbf{ZFC}$ will always satisfy the axioms of~$\mathbf{ZFC}$; in
+contrast, most toposes do not even validate the law of excluded middle, much
+less so the axiom of choice (which implies the law of excluded middle in
+presence of the remaining axioms of intuitionistic Zermelo--Fraenkel set
+theory~\cite{XXX}). This failure of classicality is often for interesting
+geometric or algebraic reasons instead of because of logical issues, as will be
+detailed in Section~\ref{several-toposes}.
+
+Secondly, there is a difference in motivation. The main philosophical reason
+for studying models of set theory is to study which notions of sets are
+coherent: Does the cardinality of the reals need to be the cardinal directly
+succeeding~$\aleph_0$, the cardinality of the naturals? No, there are models of
+set theory in which the continuum hypothesis fails. Do non-measurable sets of
+reals need to exist? No, in models of~$\mathbf{ZF}+\mathbf{AD}$,
+Zermelo--Fraenkel set theory plus the axiom of determinacy, it's a theorem that
+every subset of~$\RR^n$ is Lebesgue-measurable. Can the axiom of choice be
+added to the axioms of~$\mathbf{ZF}$ without causing inconsistency? Yes, if~$M$
+is a model of~$\mathbf{ZF}$ then~$L^M$, the set of definable sets of~$M$, is a
+model of~$\mathbf{ZFC}$.~\cite{XXX, maybe SEP?}
+
+While toposes can be used for similar such purposes, and indeed have been,
+especially to explore the various intuitionistic notions of sets, an important
+aspect of topos theory is that toposes are used to explore the standard
+mathematical universe: Truth in the \effective topos tells us what is
+computable; truth in sheaf toposes tells us what's true locally, in a fashion
+which depends continuously on parameters; the little Zariski topos of a general
+commutative ring can be used to pretend that the given ring is a field; toposes
+adapted to synthetic differential geometry can be used to rigorously work with
+infinitesimals. All of these examples will be presented in more detail in
+Section~\ref{several-toposes}.
+
+In a certain precise way, toposes allow us to study the common objects of
+mathematics from a different point of view -- one such view for every topos --
+and it is a beautiful and intriguing fact that with the sole exception of the
+law of excluded middle, the laws of logic apply to mathematical objects also
+when viewed through the lens of a specific topos.
 
@@
 * what giving up classical logic actually amounts to in practice
-* examples in Eff
-* examples in Eff(STM)
-* examples in Sh(X)
 * examples in Spec(A), "reifying all the individual localizations into a
@@
 * SDG (infinitesimals are well-studied in philosophy of mathematics)
-* beautiful and intriguing fact: laws of logic apply to mathematical objects
-  not only on the face, but also in different ways
 * quick remark on the internal language being based on types instead of sets
@@
 Send to Georg.
+Also Giuseppe Cammarata.

### 2019-01-09 paper-filmat.tex
More on variants of the effective topos

@@
 \newcommand{\stacksproject}[1]{\cite[{\href{https://stacks.math.columbia.edu/tag/#1}{Tag~#1}}]{stacks-project}}
+
+\renewcommand{\paragraph}[1]{\noindent\textbf{#1.}}
 
@@
 
+\paragraph{Hypercomputation}
 Firstly, in place of ordinary Turing machines, one can employ the
@@
 computational steps are indexed by the ordinal numbers instead of the natural
-numbers.
+numbers. For instance, an infinite-time Turing machine can trivially decide the
+twin prime conjecture, by simply walking along the natural number line and
+recording any twin primes it finds. Then, on day~$\omega$, it can observe
+whether it has found finitely many twins or not.
 
@@
 
-A second ......
+\paragraph{Machines of the real physical world} A second variant of the
+\effective topos is obtained by using machines of the real physical world
+instead of abstract Turing machines in its construction. In doing so, we of
+course leave the realm of mathematics, as real-world machines are not objects
+of mathematical studies, but still it is interesting to see which commitments
+about the natural of the physical world imply which internal statements of the
+resulting topos.
+
+For instance, Andrej Bauer proved that inside this topos any function~$\RR \to
+\RR$ is continuous if, in the physical world, only finitely many computational
+steps can be carried out in finite time and if it is possible to form
+tamper-free private communication channels~\cite{XXX}.  % XXX: look up precise formulation
 

### 2019-01-09 paper-filmat.tex
typo

@@
 given in Figure~\ref{fig:trichotomy}(c), this open~$U$ could be taken as~$X$
-with the unique zero removed.
+with the unique zero of~$a$ removed.
 
@@
   number which is neither negative nor zero nor positive. Externally speaking,
-  no open neighborhood~$V$ of the unique zero of the depicted function~$a$ has
-  the property that~$a$ is everywhere negative, everywhere zero or everywhere
-  positive on~$V$.}
+  there is no covering of the unit interval by open subsets on which the
+  depicted function~$a$ is either everywhere negative, everywhere zero or everywhere
+  positive.}
 \end{figure}

### 2019-01-09 paper-filmat.tex
to do

@@
   of set theory
-* different intention with the alternative universes in comparison to set theory
 * quick overview of the several aspects of toposes (maybe at the end,
@@
 
+* use the phrase "very rich and very deep" (or similarly) about set theory
+
 Send to Georg.

### 2019-01-09 paper-reflection.tex
Started the note on the reflection principle

\documentclass[oneside,reqno]{amsart}

\usepackage[utf8]{inputenc}
\usepackage{amsthm,mathtools,stmaryrd,amssymb,graphicx}
\usepackage{booktabs}
\usepackage[all]{xy}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{xspace}

\usepackage[natbib=true,style=numeric,maxnames=10]{biblatex}
\usepackage[babel]{csquotes}
\bibliography{paper-filmat.bib}

\graphicspath{{images/}}

\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

\title[]{Reflections on reflection for \\ intuitionistic set theories}
\author{}
\email{iblech@speicherleck.de}

\theoremstyle{definition}
\newtheorem{defn}{Definition}[section]
\newtheorem{ex}[defn]{Example}

\theoremstyle{plain}
\newtheorem{prop}[defn]{Proposition}
\newtheorem{cor}[defn]{Corollary}
\newtheorem{lemma}[defn]{Lemma}
\newtheorem{thm}[defn]{Theorem}
\newtheorem{scholium}[defn]{Scholium}

\theoremstyle{remark}
\newtheorem{rem}[defn]{Remark}
\newtheorem{question}[defn]{Question}
\newtheorem{speculation}[defn]{Speculation}
\newtheorem{caveat}[defn]{Caveat}
\newtheorem{conjecture}[defn]{Conjecture}

\newcommand{\xra}[1]{\xrightarrow{#1}}
\newcommand{\XXX}[1]{\textbf{XXX: #1}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\SS}{\mathbb{S}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\PSh}{PSh}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\effective}{ef{}fective\xspace}
\newcommand{\?}{\,{:}\,}
\newcommand{\realizes}{\Vdash}
\newcommand{\notnot}{\emph{not~not}\xspace}

\newcommand{\stacksproject}[1]{\cite[{\href{https://stacks.math.columbia.edu/tag/#1}{Tag~#1}}]{stacks-project}}

\renewcommand{\paragraph}[1]{\noindent\textbf{#1.}}

\newcommand{\ZF}{\textsc{zf}}
\newcommand{\IZF}{\textsc{izf}}
\newcommand{\CZF}{\textsc{czf}}
\newcommand{\ZFC}{\textsc{zfc}}
\newcommand{\ZFCS}{\textsc{zfc/s}}
\newcommand{\RRS}{\textsc{rrs}}
\newcommand{\ES}{(\IZF+\RRS)/\textsc{s}}

\begin{document}

\begin{abstract}
  We study the well-known reflection principle of Zermelo--Fraenkel set theory
  in the context of intuitionistic Zermelo--Fraenkel set theory~$\IZF$. We show
  that the reflection principle is equivalent to Aczel's relation reflection scheme~$\RRS$. As
  applications, we give a new proof that relativized dependent choice is
  equivalent to the conjunction of~$\RRS$ and dependent choice and we present
  an intuitionistic version of Feferman's~$\ZFC/S$, a conservative extension
  of~$\ZFC$ which is useful as a foundation for category theory.
\end{abstract}

\maketitle
\thispagestyle{empty}

\section{Introduction}

In its basic form, the reflection principle for Zermelo--Fraenkel set
theory~$\ZF$ states:
\begin{thm}Let~$\varphi(x_1,\ldots,x_n)$ be a formula in the language of set
theory with free variables as indicated. Then~$\ZF$ proves
\[ \forall M\_
  \exists S \supseteq M\_
  \forall x_1,\ldots,x_n \in S\_
  \varphi(x_1,\ldots,x_n) \Leftrightarrow \varphi^S(x_1,\ldots,x_n), \]
where~$\varphi^S$ is the~$S$-relativization of~$\varphi$, obtained by
substituting any occurence of~``$\forall x$'' and~``$\exists x$'' by~``$\forall
x \in S$'' and~``$\exists x \in S$''.
\end{thm}
Furthermore, the resulting set~$S$ may be supposed to be transitive, closed
under subsets or even to be a stage~$V_\alpha$ of the cumulative hierarchy.
% XXX: "hence" instead of "or even"?

The reflection principle is important both for philosophical and for practical
reasons: Philosophically, it tells us that truth of any formula can already be
checked in an initial segment of the universe. Practically, it allows to
transfer results obtained for a restricted class of objects to all such
objects. For instance, if we manage to verify (the $S$-relativization of) some
group-theoretic statement for all groups contained in an arbitrary set~$X$,
then we may deduce by the reflection principle that in fact the statement holds
for all groups in the universe.

This observation has been used by Feferman to construct~$\ZFCS$ (``$\ZFC$ with
smallness''), a conservative extension of~$\ZFC$ which provides a useful
foundation of category theory~\cite{feferman:zfcs}. This theory extends~$\ZFC$
by the addition of a new constant symbol~$\SS$ together with axioms stating
that~$\SS$ is transitive, closed under subsets and reflective with respect to
every formula~$\varphi(x_1,\ldots,x_n)$ of the original language:
\[ \label{ax:refl}\tag{$\star$}
\forall x_1,\ldots,x_n \in \SS\_ \varphi(x_1,\ldots,x_n) \Leftrightarrow
\varphi^{\SS}(x_1,\ldots,x_n). \]
This theory is conservative over~$\ZFC$ because any given proof in~$\ZFCS$ uses
only a finite number of instances of this axiom scheme and hence the reflection
principle yields a honest set~$S$ which validates just the same equivalences.

Elements of~$\SS$ are deemed ``small'', so that~$\SS$ is the set of all small
sets. The system~$\ZFCS$ is useful as a foundation for category theory because
it allows a native treatment of large structures -- for instance, the category
of all small sets can be formed entirely withing~$\ZFCS$, without resorting to
classes -- while not requiring new set-theoretical commitments such as the
existence of Grothendieck universes. We invite the reader who wants to learn
more about the merits of~$\ZFCS$ to a survey by
Shulman~\cite[Section~11]{shulman:sets}.

To our naive eyes, the passage from~$\ZFC$ to~$\ZFCS$ looked sufficiently
innocent so that we set out to create an intuitionistic version of~$\ZFCS$: To
our mind, size issues were entirely separate from issues of constructivity, and
we hence opined that they should be dealt with separately. Such a separation
would lead to improved mental hygiene and would allow us to use the benefits
of~$\ZFCS$ in situations where the law of excluded middle and the axiom of
choice are not available, such as in realizability semantics, sheaf semantics
or quite generally topos semantics.

We expected this modification to be entirely straightforward. However, the
situation turned out to be more subtle and we failed in our original goal of
verifying the reflection principle for intuitionistic Zermelo--Fraenkel set
theory~$\IZF$. The situation for~$\CZF$, the predicative subtheory of~$\IZF$
commonly heralded as the largest common denominator of all flavors of
constructive set theory, is even more mysterious.

More specifically, the main result of this note is the following.

\begin{thm}The reflection principle is equivalent, over~$\IZF$, to Aczel's relation reflection
scheme~$\RRS$.\end{thm}

Aczel's relation reflection scheme will be reviewed below. It is validated by
all known models of~$\IZF$, hence might be regarded as not entirely
unconstructive, even though it's conjectured to be independent of~$\IZF$. As a
result, the question whether the reflection principle holds for~$\IZF$ remains
open (though conjectured to be false), and for the stronger system~$\IZF+\RRS$
we can give a variant ``with smallness'' which can serve as a set-theoretic
foundation for category theory.

\begin{conjecture}The system $\IZF$ does not prove the reflection
principle.\end{conjecture}

\begin{defn}The system $\ES$ is obtained from~$\IZF+\RRS$ by
adding a constant symbol~$\SS$ together with axioms
stating that~$\SS$ is transitive, closed under subsets and reflective for all
formulas of the original language (that is, adding one instance
of~\eqref{ax:refl} for each formula of~$\IZF$).\end{defn}

\begin{cor}The system~$\ES$ is conservative
over~$\IZF+\RRS$.\end{cor}

This note is organized as follows. Section~1 reviews the classical proof of the
reflection principle in the context of~$\ZF$ set theory. Section~2 reviews
Aczel's relation reflection scheme and presents our main result. We conclude in
Section~3 and Section~4 with two short applications.

\textbf{Acknowledgments.} XXX

\end{document}

% XXX: Topos which validates exactly those statements which are true
% externally.

% XXX: Review axioms of IZF.

### 2019-01-09 paper-reflection.tex
Review of the classical proof

@@
 \DeclareMathOperator{\PSh}{PSh}
+\DeclareMathOperator{\rank}{rank}
 \newcommand{\Set}{\mathrm{Set}}
@@
 
+\section{Review of the classical reflection principle}
+
+A basic proof of the reflection principle for~$\ZF$ runs as follows. Our proof
+of the reflection principle for~$\IZF+\RRS$ will follow the same outline.
+
+\begin{lemma}\label{lemma:zf-onestep}
+Let~$\varphi(u_1,\ldots,u_n,x)$ be a formula in the language of set
+theory with free variables as indicated. Then~$\ZF$ proves
+\begin{multline*}
+  {\qquad}\forall M\_
+  \exists S \supseteq M\_
+  \forall u_1,\ldots,u_n \in S\_ \\
+  (\exists x\_ \varphi(u_1,\ldots,u_n,x)) \Longrightarrow
+  (\exists x \in S\_ \varphi(u_1,\ldots,u_n,x)).{\qquad}
+\end{multline*}
+Furthermore:
+\begin{enumerate}
+\item We may suppose that~$S$ is transitive.
+\item We may suppose that~$S$ is closed under subsets.
+\item Given a finite list~$\varphi_1,\ldots,\varphi_k$ of formulas instead of
+the single formula~$\varphi$, we may suppose that~$S$ bounds all of
+the~$\varphi_i$.
+% XXX: better explanation
+\end{enumerate}
+\end{lemma}
+
+\begin{proof}Given a class~$X$ (as commonly understood as the comprehension of
+a formula), we denote by~$X^\sim$ its subclass~$\{ x \in X \,|\, \forall y
+\in X\_ \rank(x) \leq \rank(y) \}$, where the rank function refers to the stage
+in the cumulative hierarchy. The two fundamental properties of this construction are:
+This subclass is equal to a set,\footnote{If~$X$ is empty, then
+this claim is trivial; if~$X$ is inhabited by some element~$x_0$, then~$X^\sim$ can
+be obtained using separation from~$V_{\rank(x_0)+1}$; and in fact, by an argument
+using the set of truth values and unbounded separation, the claim can also be
+proven in~$\IZF$.} and it is inhabited if and only if~$X$ is.
+
+Starting with~$S_0 \defeq M$, we construct~$S_{k+1}$ from~$S_k$ as the union of
+sets
+\[ S_{k+1} \defeq S_k \cup \bigcup_{u_1,\ldots,u_n \in S_k} \{ x \,|\,
+\varphi(u_1,\ldots,u_n,x) \}^\sim. \]
+It is then easy to check that~$S \defeq \bigcup_{k \in \NN} S_k$ has the
+required property.
+
+For addendum~(1), we redeclare~$S \defeq \bigcup_{k \in \NN} S_k'$,
+where~$S_k'$ is the transitive closure of~$S_k$.
+
+To ensure both addendum~(1) and addendum~(2), we redeclare~$S \defeq \bigcup_{k
+\in \NN} P^\omega(S_k')$, where~$P^\omega(X) \defeq \bigcup_{\ell \in \NN}
+P^{(\ell)}(X)$ is the union of iterated powersets.
+
+For addendum~(3), we change the definition of~$S_{k+1}$ to include one summand
+for each formula~$\varphi_i$.
+\end{proof}
+
+The proof of Lemma~\ref{lemma:zf-onestep} is mostly constructive; however,
+there is one issue with nontrivial ramifications: While~$\IZF$ does still show
+that the subclass~$X^\sim$ of a given class~$X$ is a set, it does not verify
+that~$X^\sim$ is inhabited if~$X$ is. This would amount to the constructive
+taboo that any inhabited set contains a rank-minimal element; and in fact, by
+a result of Friedman and Scedrov~\cite{XXX}, no definable substitute
+for~$X^\sim$ exists. The remedy presented in
+Section~\ref{sect:constructive-proof} will construct~$X^\sim$ in a non-unique
+fashion and then take the union using some additional care.
+
+\begin{thm}\label{thm:zf-bigstep}
+Let~$\varphi(x_1,\ldots,x_n)$ be a formula in the language of set
+theory with free variables as indicated. Then~$\ZF$ proves
+\[ \forall M\_
+  \exists S \supseteq M\_
+  \forall x_1,\ldots,x_n \in S\_
+  \varphi(x_1,\ldots,x_n) \Leftrightarrow \varphi^S(x_1,\ldots,x_n), \]
+where~$\varphi^S$ is the~$S$-relativization of~$\varphi$.
+\end{thm}
+
+\begin{proof}Let a set~$M$ be given. We obtain~$S$ by applying
+Lemma~\ref{lemma:zf-onestep} to the list of all subformulas of~$\varphi$. We
+then verify that~$S$ has the required property by an induction on the structure
+of~$\varphi$. The cases~$\text{``$=$''}, \text{``$\in$''}, \text{``$\top$''},
+\text{``$\bot$''}, \text{``$\wedge$''},
+\text{``$\vee$''}, \text{``$\Rightarrow$''}$ follow
+trivially from the induction hypothesis. The case~``$\forall$'' does
+not need to be treated, since we may assume without loss of generality that all
+universal quantifiers in~$\varphi$ have been rewritten
+as~``$\neg\exists\neg$''.
+
+The remaining case is where~$\varphi$ is of the form~$\varphi \equiv (\exists
+x\_ \psi(u_1,\ldots,u_m,x))$. In this case, the claim is that
+\[ \forall u_1,\ldots,u_m \in S\_
+  (\exists x\_ \psi(u_1,\ldots,u_m,x)) \Longrightarrow
+  (\exists x \in S\_ \psi^S(u_1,\ldots,u_m,x)), \]
+and this follows by the property of~$S$ guaranteed by
+Lemma~\ref{lemma:zf-onestep} and by the induction hypothesis concerning the
+subformula~$\psi^S(u_1,\ldots,u_m,x)$.
+\end{proof}
+
+The proof of Theorem~\ref{thm:zf-bigstep} is constructive with the sole
+exception of treating the case of universal quantifiers by appealing to the law
+of excluded middle. For the constructive proof in
+Section~\ref{sect:constructive-proof}, we will instead appeal to a strengthened
+version of Lemma~\ref{lemma:zf-onestep}.
+
 \end{document}

### 2019-01-10 paper-reflection.tex
typo

@@
   applications, we give a new proof that relativized dependent choice is
-  equivalent to the conjunction of~$\RRS$ and dependent choice and we present
-  an intuitionistic version of Feferman's~$\ZFC/S$, a conservative extension
-  of~$\ZFC$ which is useful as a foundation for category theory.
+  equivalent to the conjunction of the relation reflection scheme and dependent
+  choice and we present an intuitionistic version of Feferman's~$\ZFCS$, a
+  conservative extension of~$\ZFC$ which is useful as a foundation for category
+  theory.
 \end{abstract}
@@
 
-In its basic form, the reflection principle for Zermelo--Fraenkel set
-theory~$\ZF$ states:
+The basic form of the reflection principle for Zermelo--Fraenkel set
+theory~$\ZF$ is the following.
 \begin{thm}Let~$\varphi(x_1,\ldots,x_n)$ be a formula in the language of set
@@
 x \in S$'' and~``$\exists x \in S$''.
-\end{thm}
 Furthermore, the resulting set~$S$ may be supposed to be transitive, closed
@@
 % XXX: "hence" instead of "or even"?
+\end{thm}
 
@@
 checked in an initial segment of the universe. Practically, it allows to
+% XXX: be more precise about philosophy
 transfer results obtained for a restricted class of objects to all such
@@
 group-theoretic statement for all groups contained in an arbitrary set~$X$,
-then we may deduce by the reflection principle that in fact the statement holds
+then we may deduce by the reflection principle that the statement holds
 for all groups in the universe.
@@
 smallness''), a conservative extension of~$\ZFC$ which provides a useful
-foundation of category theory~\cite{feferman:zfcs}. This theory extends~$\ZFC$
+foundation of category theory~\cite{feferman:zfcs}. This system extends~$\ZFC$
 by the addition of a new constant symbol~$\SS$ together with axioms stating
@@
 \varphi^{\SS}(x_1,\ldots,x_n). \]
-This theory is conservative over~$\ZFC$ because any given proof in~$\ZFCS$ uses
-only a finite number of instances of this axiom scheme and hence the reflection
-principle yields a honest set~$S$ which validates just the same equivalences.
+The system~$\ZFCS$ is conservative over~$\ZFC$ because any given proof in~$\ZFCS$ uses
+only a finite number of instances of the axioms~\eqref{ax:refl}, whereby the reflection
+principle can be used to yield an honest set~$S$ which validates just the same equivalences
+and can hence be used in place of~$\SS$.
 
@@
 sets. The system~$\ZFCS$ is useful as a foundation for category theory because
-it allows a native treatment of large structures -- for instance, the category
-of all small sets can be formed entirely withing~$\ZFCS$, without resorting to
-classes -- while not requiring new set-theoretical commitments such as the
-existence of Grothendieck universes. We invite the reader who wants to learn
-more about the merits of~$\ZFCS$ to a survey by
+it supports, without requiring new set-theoretical commitments such as the
+existence of Grothendieck universes, a native treatment of large structures.
+for instance, the category of all small sets can be formed entirely
+withing~$\ZFCS$, without resorting to classes. We invite the reader wanting
+to learn more about the merits of~$\ZFCS$ to a survey by
 Shulman~\cite[Section~11]{shulman:sets}.
@@
 we hence opined that they should be dealt with separately. Such a separation
-would lead to improved mental hygiene and would allow us to use the benefits
+would not only lead to improved mental hygiene and better understanding, but
+would also allow us to use the benefits
 of~$\ZFCS$ in situations where the law of excluded middle and the axiom of
@@
 
-More specifically, the main result of this note is the following.
+However, we suceeded in verifying the reflection principle for a slight
+extension of~$\IZF$:
 
@@
 
-This note is organized as follows. Section~1 reviews the classical proof of the
-reflection principle in the context of~$\ZF$ set theory. Section~2 reviews
-Aczel's relation reflection scheme and presents our main result. We conclude in
-Section~3 and Section~4 with two short applications.
+This note is organized as follows. Section~\ref{sect:review} reviews the
+classical proof of the reflection principle in the context of~$\ZF$ set theory.
+Section~\ref{sect:classical-proof} reviews Aczel's relation reflection scheme
+and presents our main result. We conclude in Section~\ref{sect:appl-rdc} and
+Section~\ref{sect:appl-smallness} with two short applications.
 
@@
 \section{Review of the classical reflection principle}
+\label{sect:review}
 
@@
 
-\begin{lemma}\label{lemma:zf-onestep}
+\begin{lemma}\label{lemma:zf-smallstep}
 Let~$\varphi(u_1,\ldots,u_n,x)$ be a formula in the language of set
@@
 \item We may suppose that~$S$ is closed under subsets.
-\item Given a finite list~$\varphi_1,\ldots,\varphi_k$ of formulas instead of
+\item Given a finite list~$\varphi_1,\ldots,\varphi_s$ of formulas instead of
 the single formula~$\varphi$, we may suppose that~$S$ bounds all of
-the~$\varphi_i$.
+the formulas~$\varphi_i$.
 % XXX: better explanation
@@
 
-Starting with~$S_0 \defeq M$, we construct~$S_{k+1}$ from~$S_k$ as the union of
-sets
+Starting with~$S_0 \defeq M$, we construct~$S_{k+1}$ from~$S_k$ as the union
 \[ S_{k+1} \defeq S_k \cup \bigcup_{u_1,\ldots,u_n \in S_k} \{ x \,|\,
 \varphi(u_1,\ldots,u_n,x) \}^\sim. \]
-It is then easy to check that~$S \defeq \bigcup_{k \in \NN} S_k$ has the
+It is then easy to check that~$S \defeq \bigcup_{k \in \NN} S_k$ is a set with the
 required property.
 
-For addendum~(1), we redeclare~$S \defeq \bigcup_{k \in \NN} S_k'$,
-where~$S_k'$ is the transitive closure of~$S_k$.
+For addendum~(1), we redeclare~$S \defeq \bigcup_{k \in \NN} S_k^t$,
+where~$S_k^t$ is the transitive closure of~$S_k$.
 
@@
 
-The proof of Lemma~\ref{lemma:zf-onestep} is mostly constructive; however,
-there is one issue with nontrivial ramifications: While~$\IZF$ does still show
+The proof of Lemma~\ref{lemma:zf-smallstep} is mostly constructive; however,
+there is one issue with nontrivial ramifications: While~$\IZF$ does show
 that the subclass~$X^\sim$ of a given class~$X$ is a set, it does not verify
@@
 Section~\ref{sect:constructive-proof} will construct~$X^\sim$ in a non-unique
-fashion and then take the union using some additional care.
+fashion and then deal with the resulting fallout that taking the union requires
+additional care.
 
@@
 \begin{proof}Let a set~$M$ be given. We obtain~$S$ by applying
-Lemma~\ref{lemma:zf-onestep} to the list of all subformulas of~$\varphi$. We
-then verify that~$S$ has the required property by an induction on the structure
+Lemma~\ref{lemma:zf-smallstep} to the list of all subformulas of~$\varphi$. That
+the resulting set~$S$ has the required property can be checked by an induction on the structure
 of~$\varphi$. The cases~$\text{``$=$''}, \text{``$\in$''}, \text{``$\top$''},
@@
 trivially from the induction hypothesis. The case~``$\forall$'' does
-not need to be treated, since we may assume without loss of generality that all
+not need to be treated since we may assume without loss of generality that all
 universal quantifiers in~$\varphi$ have been rewritten
@@
 \[ \forall u_1,\ldots,u_m \in S\_
-  (\exists x\_ \psi(u_1,\ldots,u_m,x)) \Longrightarrow
-  (\exists x \in S\_ \psi^S(u_1,\ldots,u_m,x)), \]
-and this follows by the property of~$S$ guaranteed by
-Lemma~\ref{lemma:zf-onestep} and by the induction hypothesis concerning the
+  (\exists x\_ \psi(u_1,\ldots,u_m,x)) \Longleftrightarrow
+  (\exists x \in S\_ \psi^S(u_1,\ldots,u_m,x)). \]
+This claim follows by the property of~$S$ guaranteed by
+Lemma~\ref{lemma:zf-smallstep} and by the induction hypothesis concerning the
 subformula~$\psi^S(u_1,\ldots,u_m,x)$.
@@
 Section~\ref{sect:constructive-proof}, we will instead appeal to a strengthened
-version of Lemma~\ref{lemma:zf-onestep}.
+version of Lemma~\ref{lemma:zf-smallstep}.
 

### 2019-01-10 paper-reflection.tex
Proof of the reflection principle for IZF+RRS

@@
 \newcommand{\RRS}{\textsc{rrs}}
+\newcommand{\MDC}{\textsc{mdc}}
 \newcommand{\ES}{(\IZF+\RRS)/\textsc{s}}
@@
 
+\section{Constructivizing the reflection principle}
+\label{sect:constructive-proof}
+
+Our constructive rendition of the reflection principle will require Aczel's
+relation reflection scheme~$\RRS$ displayed in Definition~\ref{defn:rrs}. This
+result is the best possible, as conversely the reflection principle entails the
+relation reflection scheme. The relation reflection scheme first surfaced in
+the theory of coinductive definitions of classes and enjoys substantial
+stability properties, as it passes from the metatheory to XXX[all kinds of]
+models. Background on~$\RRS$ can be found in Aczel's original article
+introducing it~\cite{aczel:rrs}.
+
+\begin{defn}[$\RRS$, Aczel's relation reflection scheme]\label{defn:rrs}
+Let~$X$ and~$R \subseteq X \times X$ be classes. Let~$A \subseteq X$ be a
+subset. If~$\forall x \in X\_ \exists y \in X\_ \langle x,y \rangle \in R$,
+then there is a set~$B$ such that~$A \subseteq B \subseteq X$ and such
+that~$\forall x \in B\_ \exists y \in B\_ \langle x,y \rangle \in R$.
+\end{defn}
+
+For our purposes, Palmgren's multivalued dependent choice is slightly more
+convenient to work with than~$\RRS$. It is equivalent, over~$\CZF$ and
+a~fortiori over~$\IZF$, to~$\RRS$~\cite{palmgren:mdc}.
+
+\begin{defn}[$\MDC$, Palmgren's multivalued dependent choice]
+Let~$X$ and~$R \subseteq X \times X$ be classes. Let~$A \subseteq X$ be a
+subset. If~$\forall x \in X\_ \exists y \in X\_ \langle x,y \rangle \in R$,
+then there is a function~$f : \NN \to P(X)$ (the class of all subsets of~$X$)
+such that~$A \subseteq f(0)$ and such that~$\forall x \in f(n)\_ \exists y \in
+f(n+1)\_ \langle x,y \rangle \in R$ for every~$n$.\end{defn}
+
+Even though superficially similar, the following lemma is \emph{not} yet a
+constructivization of Lemma~\ref{lemma:zf-smallstep}; these two lemmas differ
+in the set from which~$u_1,\ldots,u_n$ are drawn.
+
+\begin{lemma}\label{lemma:izf-microstep}
+Let~$\varphi(u_1,\ldots,u_n,x)$ be a formula in the language of set
+theory with free variables as indicated. Then~$\IZF$ proves
+\begin{multline*}
+  {\qquad}\forall H\_
+  \exists H' \supseteq H\_
+  \forall u_1,\ldots,u_n \in H\_ \\
+  \begin{array}{@{}rcl@{}}
+  (\exists x\_ \varphi(u_1,\ldots,u_n,x)) &\Longrightarrow&
+  (\exists x \in H'\_ \varphi(u_1,\ldots,u_n,x)) \quad\mathop{\wedge} \\
+  (\forall x\_ \varphi(u_1,\ldots,u_n,x)) &\Longleftarrow&
+  (\forall x \in H'\_ \varphi(u_1,\ldots,u_n,x)).\end{array}
+  {\qquad}
+\end{multline*}
+Furthermore, we may assume that~$H'$ is transitive and closed under subsets;
+and given a finite list~$\varphi_1,\ldots,\varphi_s$ of formulas instead of
+the single formula~$\varphi$, we may suppose that~$H'$ has the displayed
+property for each of the formulas~$\varphi_i$.
+\end{lemma}
+
+\begin{proof}Let~$\Omega \defeq P(\{0\})$ be the set of truth values.
+For given elements~$u_1,\ldots,u_n \in H$, we have
+\[ \begin{array}{@{}l@{}l@{}}
+  \forall a \in \{ a \in \{0\} \,|\, \exists x\_ \varphi(u_1,\ldots,u_n,x) \}\_ &
+  \exists x\_ \varphi(u_1,\ldots,u_n,x) \quad\text{and} \\
+  \forall p \in \{ p \in \Omega \,|\, \exists x\_ (0 \in p \Leftrightarrow
+  \varphi(u_1,\ldots,u_n,x)) \}\_ &
+  \exists x\_ (0 \in p \Leftrightarrow \varphi(u_1,\ldots,u_n,x)).
+\end{array} \]
+Hence, by collection, there are sets~$C$ and~$D$ such that
+\[ \begin{array}{@{}rcl@{}}
+  (\exists x\_ \varphi(u_1,\ldots,u_n,x)) &\Longrightarrow&
+  (\exists x \in C\_ \varphi(u_1,\ldots,u_n,x)) \quad\text{and} \\
+  (\forall x\_ \varphi(u_1,\ldots,u_n,x)) &\Longleftarrow&
+  (\forall x \in D\_ \varphi(u_1,\ldots,u_n,x)).
+\end{array} \]
+The union~$C \cup D$ satisfies both of these conditions at once.
+
+Applying collection again, there is a set~$X$ such that for
+any~$u_1,\ldots,u_n \in H$ there exists a set~$E \in X$ such that
+\[ \begin{array}{@{}rcl@{}}
+  (\exists x\_ \varphi(u_1,\ldots,u_n,x)) &\Longrightarrow&
+  (\exists x \in E\_ \varphi(u_1,\ldots,u_n,x)) \quad\text{and} \\
+  (\forall x\_ \varphi(u_1,\ldots,u_n,x)) &\Longleftarrow&
+  (\forall x \in E\_ \varphi(u_1,\ldots,u_n,x)).
+\end{array} \]
+Hence the set~$H' \defeq M \cup \bigcup X$ has the required property.
+
+To ensure that~$H'$ is transitive and closed under subsets, we simply pass
+first to its transitive closure and then compute the union of all its
+finitely-iterated powersets.
+
+In order to accomodate more than a single formula~$\varphi$, we add one summand
+in the definition of~$H'$ for each formula~$\varphi_i$.
+\end{proof}
+
+\begin{lemma}\label{lemma:izf-smallstep}
+Let~$\varphi(u_1,\ldots,u_n,x)$ be a formula in the language of set
+theory with free variables as indicated. Then~$\IZF+\RRS$ proves
+\begin{multline*}
+  {\qquad}\forall M\_
+  \exists S \supseteq M\_
+  \forall u_1,\ldots,u_n \in S\_ \\
+  \begin{array}{@{}rcl@{}}
+  (\exists x\_ \varphi(u_1,\ldots,u_n,x)) &\Longrightarrow&
+  (\exists x \in S\_ \varphi(u_1,\ldots,u_n,x)) \quad\mathop{\wedge} \\
+  (\forall x\_ \varphi(u_1,\ldots,u_n,x)) &\Longleftarrow&
+  (\forall x \in S\_ \varphi(u_1,\ldots,u_n,x)).\end{array}
+  {\qquad}
+\end{multline*}
+Furthermore:
+\begin{enumerate}
+\item We may suppose that~$S$ is transitive.
+\item We may suppose that~$S$ is closed under subsets.
+\item Given a finite list~$\varphi_1,\ldots,\varphi_s$ of formulas instead of
+the single formula~$\varphi$, we may suppose that~$S$ bounds all of
+the formulas~$\varphi_i$.
+% XXX: better explanation
+\end{enumerate}
+\end{lemma}
+
+\begin{proof}By~$\MDC$ and Lemma~\ref{lemma:izf-microstep}, there is a
+function~$f : \NN \to P(V)$, where~$V$ is the universe, such that~$M \in
+f(0)$ and such that for any number~$k$ and any set~$H \in f(k)$, there is a
+set~$H' \in f(k+1)$ such that~$H$ and~$H'$ are related as in the conclusion of
+Lemma~\ref{lemma:izf-microstep}.
+
+We set~$S \defeq \bigcup \bigl(\bigcup_{k \in \NN} f(k)\bigr)$. This set~$S$
+has the required property; to verify this, it is useful to observe that
+given~$u_1,\ldots,u_n \in S$, there exists a common number~$k$ such
+that~$u_1,\ldots,u_n \in H$ for some~$H \in f(k)$.
+
+In order to ensure addendum~(1) and~(2), we apply~$\MDC$ in a slightly
+different way such that for any number~$k$ and any set~$H \in f(k)$, there is a
+set~$H' \in f(k+1)$ such that~$H$ and~$H'$ are related as in the conclusion of
+Lemma~\ref{lemma:izf-microstep} and such that~$H'$ is transitive and closed
+under subsets. Even though it cannot be expected that for any number~$k$,
+any set~$H \in f(k)$ will be transitive and closed under subsets, the union~$S$
+will.
+\end{proof}
+
+\begin{thm}\label{thm:izf-bigstep}
+Let~$\varphi(x_1,\ldots,x_n)$ be a formula in the language of set
+theory with free variables as indicated. Then~$\IZF+\RRS$ proves
+\[ \forall M\_
+  \exists S \supseteq M\_
+  \forall x_1,\ldots,x_n \in S\_
+  \varphi(x_1,\ldots,x_n) \Leftrightarrow \varphi^S(x_1,\ldots,x_n), \]
+where~$\varphi^S$ is the~$S$-relativization of~$\varphi$.
+\end{thm}
+
+\begin{proof}The proof of Theorem~\ref{thm:zf-bigstep} carries over. The only
+difference is that instead of Lemma~\ref{lemma:zf-smallstep},
+Lemma~\ref{lemma:izf-smallstep} has to be used, and that the case for the
+universal quantifier has to be treated just as the case for the existential
+quantifier is.\end{proof}
+
 \end{document}

### 2019-01-10 paper-reflection.tex
typo

@@
 \end{multline*}
-Furthermore:
-\begin{enumerate}
-\item We may suppose that~$S$ is transitive.
-\item We may suppose that~$S$ is closed under subsets.
-\item Given a finite list~$\varphi_1,\ldots,\varphi_s$ of formulas instead of
-the single formula~$\varphi$, we may suppose that~$S$ bounds all of
-the formulas~$\varphi_i$.
+Furthermore: (1)~We may suppose that~$S$ is transitive. (2)~We may suppose that~$S$ is
+closed under subsets. (3)~Given a finite list~$\varphi_1,\ldots,\varphi_s$
+of formulas instead of the single formula~$\varphi$, we may suppose that~$S$
+bounds all of the formulas~$\varphi_i$.
 % XXX: better explanation
-\end{enumerate}
 \end{lemma}
@@
 \end{multline*}
-Furthermore, we may assume that~$H'$ is transitive and closed under subsets;
-and given a finite list~$\varphi_1,\ldots,\varphi_s$ of formulas instead of
-the single formula~$\varphi$, we may suppose that~$H'$ has the displayed
-property for each of the formulas~$\varphi_i$.
+Furthermore: (1)~We may suppose that~$H'$ is transitive. (2)~We may suppose
+that~$H'$ is closed under subsets. (3)~Given a finite
+list~$\varphi_1,\ldots,\varphi_s$ of formulas instead of the single
+formula~$\varphi$, we may suppose that~$H'$ has the displayed property for each
+of the formulas~$\varphi_i$.
 \end{lemma}
@@
 \end{multline*}
-Furthermore:
-\begin{enumerate}
-\item We may suppose that~$S$ is transitive.
-\item We may suppose that~$S$ is closed under subsets.
-\item Given a finite list~$\varphi_1,\ldots,\varphi_s$ of formulas instead of
-the single formula~$\varphi$, we may suppose that~$S$ bounds all of
-the formulas~$\varphi_i$.
+Furthermore: (1)~We may suppose that~$S$ is transitive. (2)~We may suppose
+that~$S$ is closed under subsets. (3)~Given a finite
+list~$\varphi_1,\ldots,\varphi_s$ of formulas instead of the single
+formula~$\varphi$, we may suppose that~$S$ bounds all of the
+formulas~$\varphi_i$.
 % XXX: better explanation
-\end{enumerate}
 \end{lemma}

### 2019-01-10 paper-reflection.tex
Optimality result

@@
 x \in S$'' and~``$\exists x \in S$''.
-Furthermore, the resulting set~$S$ may be supposed to be transitive, closed
-under subsets or even to be a stage~$V_\alpha$ of the cumulative hierarchy.
+Furthermore, the resulting set~$S$ may be supposed to be transitive, to be closed
+under subsets or even to be a stage~$V_\alpha$ of the cumulative hierarchy;
 % XXX: "hence" instead of "or even"?
+and given not a single formula~$\varphi$ but a finite
+list~$\varphi_1,\ldots,\varphi_s$ of formulas, we may suppose that~$S$ reflects
+all of them.
 \end{thm}
@@
   \varphi(x_1,\ldots,x_n) \Leftrightarrow \varphi^S(x_1,\ldots,x_n), \]
-where~$\varphi^S$ is the~$S$-relativization of~$\varphi$.
+where~$\varphi^S$ is the~$S$-relativization of~$\varphi$. Furthermore, the
+resulting set~$S$ may be supposed to be transitive and to be closed under
+subsets; and given not a single formula~$\varphi$ but a finite
+list~$\varphi_1,\ldots,\varphi_s$ of formulas, we may suppose that~$S$ reflects
+all of them.
 \end{thm}
@@
 where~$\varphi^S$ is the~$S$-relativization of~$\varphi$.
+Furthermore, the resulting set~$S$ may be supposed to be transitive and to be closed
+under subsets; and given not a single formula~$\varphi$ but a finite
+list~$\varphi_1,\ldots,\varphi_s$ of formulas, we may suppose that~$S$ reflects
+all of them.
 \end{thm}
@@
 
+\begin{thm}\label{thm:refl-entails-rrs}
+Over~$\IZF$, each instance of Aczel's relation reflection scheme~$\RRS$ can be
+deduced from suitable instances of the assumption that, given a finite list
+of formulas, for every set~$M$ there is a set~$S \supseteq M$ reflecting the
+given formulas.
+\end{thm}
+
+\begin{proof}We verify the equivalent~$\MDC$ instead of~$\RRS$. Let~$X$ and~$R
+\subseteq X \times X$ be classes. Let~$A \subseteq X$ be a subset and
+suppose~$\forall x \in X\_ \exists y \in X\_ \langle x,y \rangle \in R$.
+
+By assumption, there is a set~$S \supseteq A$ which reflects the three
+formulas~``$x \in X$'', ``$\langle x,y \rangle \in R$'' and~``$\exists y \in X\_ \langle
+x,y \rangle \in R$''.
+
+By separation, the class~$S \cap X$ is a set, and hence we may set~$f(k) \defeq S \cap X$
+for all numbers~$k$. Then~$A \subseteq f(0)$, any given~$x \in f(k)$, by
+assumption and the reflecting property of~$S$ there exists~$y \in S$ such
+that~$y \in X$ (hence~$y \in f(k+1)$) and~$\langle x,y \rangle \in R$.
+\end{proof}
+
 \end{document}
@@
 % XXX: Review axioms of IZF.
+
+% XXX: Think about pairs.

### 2019-01-10 paper-reflection.tex
Bibliography
typo

@@
 \usepackage[babel]{csquotes}
-\bibliography{paper-filmat.bib}
+\bibliography{paper-reflection.bib}
 
@@
 \author{}
-\email{iblech@speicherleck.de}
+\email{}
 
@@
 the theory of coinductive definitions of classes and enjoys substantial
-stability properties, as it passes from the metatheory to XXX[all kinds of]
+stability properties, as it passes from the meta theory to XXX[all kinds of]
 models. Background on~$\RRS$ can be found in Aczel's original article
@@
 then there is a function~$f : \NN \to P(X)$ (the class of all subsets of~$X$)
-such that~$A \subseteq f(0)$ and such that~$\forall x \in f(n)\_ \exists y \in
-f(n+1)\_ \langle x,y \rangle \in R$ for every~$n$.\end{defn}
+such that~$A \subseteq f(0)$ and such that~$\forall x \in f(k)\_ \exists y \in
+f(k+1)\_ \langle x,y \rangle \in R$ for every number~$k$.\end{defn}
 

### 2019-01-10 paper-reflection.tex
Two tiny applications

@@
 
+\newcommand{\PRA}{\textsc{pra}}
 \newcommand{\ZF}{\textsc{zf}}
@@
 \newcommand{\ZFCS}{\textsc{zfc/s}}
+\newcommand{\ZFCI}{\textsc{zfc+i}}
 \newcommand{\RRS}{\textsc{rrs}}
+\newcommand{\RDC}{\textsc{rdc}}
+\newcommand{\DC}{\textsc{dc}}
 \newcommand{\MDC}{\textsc{mdc}}
@@
 By separation, the class~$S \cap X$ is a set, and hence we may set~$f(k) \defeq S \cap X$
-for all numbers~$k$. Then~$A \subseteq f(0)$, any given~$x \in f(k)$, by
+for all numbers~$k$. Then~$A \subseteq f(0)$; any given~$x \in f(k)$, by
 assumption and the reflecting property of~$S$ there exists~$y \in S$ such
@@
 
+\section{A new proof of~$\RDC = \RRS + \DC$}
+\label{sect:appl-rdc}
+
+When he introduced~$\RRS$, Aczel proved that over $\CZF$ without subset
+collection, relative dependent choice $\RDC$ is equivalent to the conjunction
+of~$\RRS$ and dependent choice~$\DC$~\cite[Theorem~2.4]{aczel:rrs}. Using
+reflection, we can provide a new proof of this fact, although over the much
+stronger base theory~$\IZF$ instead of~$\CZF^{-}$. The idea is that reflection
+allows to reduce~$\RDC$ to~$\DC$.
+
+\begin{defn}[$\DC$, dependent choice]
+Let~$X$ and~$R \subseteq X \times X$ be sets. Let~$x_0 \in X$.
+If~$\forall x \in X\_ \exists y \in X\_ \langle x,y \rangle \in R$,
+then there is a function~$f : \NN \to X$ such that~$f(0) = x_0$ and
+such that~$\langle f(k), f(k+1) \rangle \in R$ for all numbers~$k$.
+\end{defn}
+
+\begin{defn}[$\RDC$, relative dependent choice]
+Let~$X$ and~$R \subseteq X \times X$ be classes. Let~$x_0 \in X$.
+If~$\forall x \in X\_ \exists y \in X\_ \langle x,y \rangle \in R$,
+then there is a function~$f : \NN \to X$ such that~$f(0) = x_0$ and
+such that~$\langle f(k), f(k+1) \rangle \in R$ for all numbers~$k$.
+\end{defn}
+
+\begin{prop}Over~$\IZF$, $\RDC$ is equivalent to~$\RRS + \DC$.
+\end{prop}
+
+\begin{proof}It's clear that~$\RDC$ implies~$\RRS + \DC$. % XXX: check
+
+Conversely, assume~$\RRS$ and~$\DC$. In order to verify~$\RDC$, let classes~$X$
+and~$R \subseteq X \times X$ be given. Let~$x_0 \in X$ and assume~$\forall x
+\in X\_ \exists y \in X\_ \langle x,y \rangle \in R$.
+
+By Theorem~\ref{thm:izf-bigstep}, reflection is available; hence there is a
+set~$S \supseteq \{x_0\}$ which reflects the three formulas~``$x \in X$'', ``$\langle x,y
+\rangle \in R$'' and~``$\forall x\_ x \in X \Rightarrow \exists y\_ (y \in X \wedge \langle x,y
+\rangle \in R)$''.
+
+Hence~$\forall x \in X \cap S\_ \exists y \in X \cap S\_ \langle x,y \rangle
+\in R$. By~$\DC$, there is a choice function~$f : \NN \to X \cap S$ such
+that~$f(0) = x_0$ and~$\langle f(k), f(k+1) \rangle \in R$ for all numbers~$k$.
+This is a function of the kind required by~$\RDC$.
+\end{proof}
+
+\section{An intuitionistic version of Feferman's~$\ZFCS$}
+
+XXX: repeat definition and conservativity statement?
+
+The system~$\ES$ is also interesting from the point of view of topos theory.
+Any topos supports an internal language which can be used to reason about the
+objects and morphisms of the topos in a naive element-based language, allowing
+us to pretend that the objects are plain sets (or types) and that the morphisms
+are plain maps between those sets (\cite[Chapter~6]{borceux:handbook3},
+\cite[Section~1.3]{caramello:ttt}, \cite[Chapter~14]{goldblatt:topoi},
+\cite[Chapter~VI]{moerdijk-maclane:sheaves-logic}).
+
+Given a topos~$\mathcal{E}$ and a formula~$\varphi$ in its internal language,
+we write~``$\mathcal{E} \models \varphi$'' to mean that~$\varphi$ holds
+in~$\E$. As a special case, truth in the topos~$\Set$, the category of all sets,
+coincides with truth in the background theory; symbolically: $\Set \models
+\varphi$ iff~$\varphi$.
+
+However, in the context of~($\IZF$ or)~$\ZFC$, it is difficult to make this
+claim precise. Because in~$\ZFC$ the category~$\Set$ of all sets can not be
+coded as a set,~$\ZFC$ cannot define truth in~$\Set$. We must therefore resort
+to a meta theory in order to express this claim, for instance by stating
+that primitive recursive arithmetic~$\PRA$ proves that for any
+formula~$\varphi$ of~$\ZFC$, $\ZFC$ proves that~$(\Set \models \varphi)
+\Leftrightarrow \varphi$, where~``$\Set \models \varphi$'' is to be unrolled
+by~$\PRA$.
+
+An alternative is offered by~$\ZFCI$, $\ZFC$ plus the existence of a
+strongly inaccessible cardinal. In this system, there is a Grothendieck
+universe~$U$, we can form the category~$\Set_U$ of all sets of~$U$ as an honest
+set, can define truth in~$\Set_U$ and prove, within the system, that for any
+formula~$\varphi$, $(\Set_U \models \varphi) \Leftrightarrow (U \models
+\varphi)$. This even holds for formulas of the full infinitary language of
+toposes, which allows for infinite disjunctions and infinite conjunctions; this
+could not be mimicked by resorting to~$\PRA$ as indicated above.
+
+However, truth in a Grothendieck universe~$U$ need not be related to actual
+truth. A solution is provided by~$\ZFCS$ and by~$\ES$. In these systems, we can
+form the topos~$\Set_\SS$, define truth in it, prove for all formulas~$\varphi$
+that~$(\Set_\SS \models \varphi) \Leftrightarrow (\SS \models \varphi)$; and
+reflectin for~$\SS$ ensures that for each (external, standard)
+formula~$\varphi$, the system proves~``$(\Set_\SS \models \varphi)
+\Leftrightarrow \varphi$''.
+
+\printbibliography
+
 \end{document}
-
-% XXX: Topos which validates exactly those statements which are true
-% externally.
 

### 2019-01-10 paper-reflection.tex
typo

@@
 
-Aczel's relation reflection scheme will be reviewed below. It is validated by
+Aczel's relation reflection scheme will be reviewed below. It is validated not
+only by~$\ZFC$, but also by~$\ZF$, and furthermore by
 all known models of~$\IZF$, hence might be regarded as not entirely
@@
 relation reflection scheme~$\RRS$ displayed in Definition~\ref{defn:rrs}. This
-result is the best possible, as conversely the reflection principle entails the
+result is the best possible, as we verify in Theorem~\ref{thm:refl-entails-rrs}
+that conversely the reflection principle entails the
 relation reflection scheme. The relation reflection scheme first surfaced in
@@
 % XXX: Think about pairs.
+
+% XXX: Remark that the new proof of reflection also works over ZF, but that the
+% usual proof is probably preferable.

### 2019-01-10 paper-reflection.tex
to do
typo

@@
 where~$\varphi^S$ is the~$S$-relativization of~$\varphi$, obtained by
-substituting any occurence of~``$\forall x$'' and~``$\exists x$'' by~``$\forall
+substituting any occurrence of~``$\forall x$'' and~``$\exists x$'' by~``$\forall
 x \in S$'' and~``$\exists x \in S$''.
@@
 existence of Grothendieck universes, a native treatment of large structures.
-for instance, the category of all small sets can be formed entirely
-withing~$\ZFCS$, without resorting to classes. We invite the reader wanting
-to learn more about the merits of~$\ZFCS$ to a survey by
+For instance, the category of all small sets can be formed entirely
+within~$\ZFCS$, without resorting to classes. We invite the reader wanting
+to learn more about the merits of~$\ZFCS$ to study the survey by
 Shulman~\cite[Section~11]{shulman:sets}.
@@
 verifying the reflection principle for intuitionistic Zermelo--Fraenkel set
-theory~$\IZF$. The situation for~$\CZF$, the predicative subtheory of~$\IZF$
+theory~$\IZF$. The situation for~$\CZF$, the predicative subsystem of~$\IZF$
 commonly heralded as the largest common denominator of all flavors of
@@
 
-However, we suceeded in verifying the reflection principle for a slight
+However, we succeeded in verifying the reflection principle for a slight
 extension of~$\IZF$:
@@
 
-In order to accomodate more than a single formula~$\varphi$, we add one summand
+In order to accommodate more than a single formula~$\varphi$, we add one summand
 in the definition of~$H'$ for each formula~$\varphi_i$.
@@
 that~$(\Set_\SS \models \varphi) \Leftrightarrow (\SS \models \varphi)$; and
-reflectin for~$\SS$ ensures that for each (external, standard)
+reflection for~$\SS$ ensures that for each (external, standard)
 formula~$\varphi$, the system proves~``$(\Set_\SS \models \varphi)
@@
 % usual proof is probably preferable.
+
+% XXX: Über Verallgemeinerung zu CZF spekulieren.
+
+% XXX: Note that the result fails over IZF_R.
+
+% XXX: Reference more choice axioms?

### 2019-01-10 paper-reflection.tex
typo

@@
 To ensure both addendum~(1) and addendum~(2), we redeclare~$S \defeq \bigcup_{k
-\in \NN} P^\omega(S_k')$, where~$P^\omega(X) \defeq \bigcup_{\ell \in \NN}
+\in \NN} P^\omega(S_k^t)$, where~$P^\omega(X) \defeq \bigcup_{\ell \in \NN}
 P^{(\ell)}(X)$ is the union of iterated powersets.
@@
 
-To ensure that~$H'$ is transitive and closed under subsets, we simply pass
+To ensure that~$H'$ is transitive and closed under subsets, we pass
 first to its transitive closure and then compute the union of all its
@@
 that~$u_1,\ldots,u_n \in H$ for some~$H \in f(k)$.
+% XXX
 

### 2019-01-10 paper-reflection.tex
Effortlessly fix non-critical error

@@
 
-For addendum~(1), we redeclare~$S \defeq \bigcup_{k \in \NN} S_k^t$,
-where~$S_k^t$ is the transitive closure of~$S_k$.
-
-To ensure both addendum~(1) and addendum~(2), we redeclare~$S \defeq \bigcup_{k
-\in \NN} P^\omega(S_k^t)$, where~$P^\omega(X) \defeq \bigcup_{\ell \in \NN}
-P^{(\ell)}(X)$ is the union of iterated powersets.
-
-For addendum~(3), we change the definition of~$S_{k+1}$ to include one summand
-for each formula~$\varphi_i$.
+For addendum~(1), we change the definition of~$S_{k+1}$ to be the transitive
+closure of what is was before. To further accommodate addendum~(2), we change
+this definition again, to be~$P^\omega$ of what is was before,
+where~$P^\omega(X) \defeq \bigcup_{\ell \in \NN} P^{(\ell)}(X)$ is the union of
+iterated powersets. For addendum~(3), we change the definition of~$S_{k+1}$ to
+include one summand for each formula~$\varphi_i$.
 \end{proof}

### 2019-01-10 paper-reflection.tex
Mark critical error

@@
 
-\begin{proof}By~$\MDC$ and Lemma~\ref{lemma:izf-microstep}, there is a
+\begin{proof}We first consider the case~$n = 1$.
+
+By~$\MDC$ and Lemma~\ref{lemma:izf-microstep}, there is a
 function~$f : \NN \to P(V)$, where~$V$ is the universe, such that~$M \in
@@
 We set~$S \defeq \bigcup \bigl(\bigcup_{k \in \NN} f(k)\bigr)$. This set~$S$
-has the required property; to verify this, it is useful to observe that
-given~$u_1,\ldots,u_n \in S$, there exists a common number~$k$ such
-that~$u_1,\ldots,u_n \in H$ for some~$H \in f(k)$.
-% XXX
+has the required property.
 
@@
 any set~$H \in f(k)$ will be transitive and closed under subsets, the union~$S$
-will.
+will. Addendum~(3) can be ensured because of addendum~(3) of
+Lemma~\ref{lemma:izf-microstep}.
+
+For the case~$n \neq 1$, \textbf{???}
 \end{proof}

### 2019-01-11 paper-reflection.tex
to do
Fix critical error

@@
 
+\newenvironment{indentblock}{%
+  \list{}{\leftmargin\leftmargin}%
+  \item\relax
+}{%
+  \endlist
+}
+
 \newcommand{\xra}[1]{\xrightarrow{#1}}
@@
 \DeclareMathOperator{\rank}{rank}
+\DeclareMathOperator{\length}{length}
+\DeclareMathOperator{\List}{List}
 \newcommand{\Set}{\mathrm{Set}}
@@
 \newcommand{\MDC}{\textsc{mdc}}
-\newcommand{\ES}{(\IZF+\RRS)/\textsc{s}}
+\newcommand{\ES}{(\IZF+\RRS_2)/\textsc{s}}
 
@@
   in the context of intuitionistic Zermelo--Fraenkel set theory~$\IZF$. We show
-  that the reflection principle is equivalent to Aczel's relation reflection scheme~$\RRS$. As
+  that the reflection principle is equivalent to~$\RRS_2$, a strengthened
+  version of Aczel's relation reflection scheme. As
   applications, we give a new proof that relativized dependent choice is
@@
 
-\begin{thm}The reflection principle is equivalent, over~$\IZF$, to Aczel's relation reflection
-scheme~$\RRS$.\end{thm}
+\begin{thm}The reflection principle is equivalent, over~$\IZF$, to the strong
+relation reflection scheme~$\RRS_2$.\end{thm}
 
-Aczel's relation reflection scheme will be reviewed below. It is validated not
+We also give a weaker version of the reflection principle which is equivalent
+to Aczel's original reflection scheme~$\RRS$.
+Both~$\RRS$ and~$\RRS_2$ will be reviewed below. They are validated not
 only by~$\ZFC$, but also by~$\ZF$, and furthermore by
 all known models of~$\IZF$, hence might be regarded as not entirely
-unconstructive, even though it's conjectured to be independent of~$\IZF$. As a
+% XXX: be more precise in "all known"
+unconstructive, even though they are conjectured to be independent of~$\IZF$. As a
 result, the question whether the reflection principle holds for~$\IZF$ remains
-open (though conjectured to be false), and for the stronger system~$\IZF+\RRS$
+open (though conjectured to be false), and for the stronger system~$\IZF+\RRS_2$
 we can give a variant ``with smallness'' which can serve as a set-theoretic
@@
 
-\begin{defn}The system $\ES$ is obtained from~$\IZF+\RRS$ by
+\begin{defn}The system $\ES$ is obtained from~$\IZF+\RRS_2$ by
 adding a constant symbol~$\SS$ together with axioms
@@
 \begin{cor}The system~$\ES$ is conservative
-over~$\IZF+\RRS$.\end{cor}
+over~$\IZF+\RRS_2$.\end{cor}
 
@@
 classical proof of the reflection principle in the context of~$\ZF$ set theory.
-Section~\ref{sect:classical-proof} reviews Aczel's relation reflection scheme
-and presents our main result. We conclude in Section~\ref{sect:appl-rdc} and
+Section~\ref{sect:axioms} reviews Aczel's relation reflection scheme and its
+variants. Our main result is presented in Section~\ref{sect:constructive-proof}.
+We conclude in Section~\ref{sect:appl-rdc} and
 Section~\ref{sect:appl-smallness} with two short applications.
@@
 
-A basic proof of the reflection principle for~$\ZF$ runs as follows. Our proof
-of the reflection principle for~$\IZF+\RRS$ will follow the same outline.
+A basic proof of the reflection principle in~$\ZF$ runs as follows. Our proof
+of the reflection principle in~$\IZF+\RRS_2$ will follow the same outline.
 
@@
 Lemma~\ref{lemma:zf-smallstep} to the list of all subformulas of~$\varphi$. That
+% XXX: "all" subformulas
 the resulting set~$S$ has the required property can be checked by an induction on the structure
@@
 
+\section{Aczel's relation reflection axiom scheme and its variants}
+\label{sect:axioms}
+
+We write~``$R : X \rightrightarrows Y$'' to mean~``$\forall x \in X\_ \exists y
+\in Y\_ \langle x,y \rangle \in R$''.
+
+\begin{defn}Let~$X$ and~$R \subseteq X \times X$ be classes. Let~$x_0 \in X$.
+\begin{itemize}
+\setlength{\itemsep}{0.3em}
+\item[$\DC$] \emph{dependent choice} \\ If~$X$ and~$R$ are sets and
+if~$R : X \rightrightarrows X$, then there is a function~$f : \NN \to X$ such that~$f(0) =
+x_0$ and~$\langle f(k), f(k+1) \rangle \in R$ for all numbers~$k$.
+\item[$\RDC$] \emph{relativized dependent choice} \\ If~$R : X
+\rightrightarrows X$, then there is a function~$f : \NN \to X$ such that~$f(0)
+= x_0$ and~$\langle f(k), f(k+1) \rangle \in R$ for all numbers~$k$.
+\item[$\RRS$] \emph{Aczel's relation reflection scheme} \\
+If~$R : X \rightrightarrows X$, then there is
+a set~$B$ such that~$x_0 \in B \subseteq X$ and~$R : B \rightrightarrows B$.
+\item[$\MDC$] \emph{Palmgren's multivalued dependent choice} \\
+If~$R : X \rightrightarrows X$, then there is a function~$f : \NN \to P(X)$
+(the class of all subsets of~$X$) such that~$x_0 \in f(0)$ and such
+that~$\forall x \in f(k)\_ \exists y \in f(k+1)\_ \langle x,y \rangle \in R$
+for every number~$k$.
+\end{itemize}
+\end{defn}
+
+\begin{defn}\label{defn:rrs2}
+Let~$X$ and~$R \subseteq X \times X \times X$ be classes. Let~$x_0 \in X$.
+\begin{itemize}
+\setlength{\itemsep}{0.3em}
+\item[$\RRS_2$]
+If~$R : X \times X \rightrightarrows X$, then there is a set~$B$ such that~$x_0
+\in B \subseteq X$ and~$R : B \times B \rightrightarrows B$.
+\item[$\MDC_2$]
+If~$R : X \times X \rightrightarrows X$, then there is a function~$f : \NN \to
+P(X)$ such that~$x_0 \in f(0)$ and such that~$\forall x \in f(k)\_ \forall x'
+\in f(k')\_ \exists y \in f(\max\{k,k'\}+1)\_ \langle x,x',y \rangle \in R$ for
+any numbers~$k, k'$.
+\end{itemize}
+\end{defn}
+
+Aczel's relation reflection scheme~$\RRS$ first surfaced in
+the theory of coinductive definitions of classes and enjoys substantial
+stability properties, as it passes from the meta theory to XXX[all kinds of]
+models. Background on~$\RRS$ can be found in Aczel's original article
+introducing it~\cite{aczel:rrs}. $\RRS$ is equivalent, over~$\CZF$ and
+a~fortiori over~$\IZF$, to Palmgren's multivalued dependent
+choice~\cite{palmgren:mdc}; this observation makes the relationship to
+dependent choice more visible.
+
+\begin{rem}Most published renderings of~$\RRS$ and~$\MDC$ do not refer to an initial
+element~$x_0 \in X$, but to a set~$A \subseteq X$ and then require instead
+of that~$x_0 \in B$, that~$A \subseteq B$. This difference is immaterial, thanks in
+one direction to the existence of singleton sets and in the other to strong
+collection. The same is true for~$\RRS_2$ and~$\MDC_2$, though the proof is not
+as easy.
+\end{rem}
+
+% First prove that in RRS_2^∈, we may suppose that not only a single given
+% element x0 is contained in B, but also a further given element x1.
+% Do this by applying RRS_2^∈ to X^ := { ∅ } U { {x} | x ∈ X } and
+% R^(p,p',q) given by:
+% * p = ∅   and p' = ∅    and q = {x0}  or
+% * p = ∅   and p' = {x0} and q = {x1}  or  the other way round  or
+% * p = {x} and p' = {x'} and q = {y} and R(x,x',y).
+%
+% Now let R : X × X ⇉ X and a set A ⊆ X be given. We want to verify RRS_2^⊆
+% for this data.
+%
+% Given H ∈ P(X), H' ∈ P(X), consider:
+%
+%    Let x ∈ H, x' ∈ H'.
+%    Then by the strengthened RRS_2^∈, there is a set B ⊆ X
+%    such that x, x' ∈ B and R : B × B ⇉ B.
+%
+% We can use collection to obtain a set J of such B's, and consider U J.
+%
+% We can then apply RRS_2^∈ to this relation to obtain a set T such that
+% A ∈ T ⊆ P(X) and ... Set C := U T. Then A ⊆ C ⊆ X and R : C × C ⇉ C.
+
+Aczel's $\RRS$ should not in itself be regarded as a choice principle; however in
+its presence, ordinary~$\DC$ entails~$\RDC$. This result is due to
+Aczel~\cite[Theorem~2.4]{aczel:rrs}, who proved the equivalence~$\RDC = \RRS +
+\DC$ over~$\CZF^-$, and we give a new proof of this equivalence over the
+stronger theory~$\IZF$, in Section~\ref{sect:appl-rdc}.
+
+In the presence of~$\DC$, $\RRS$ is equivalent to~$\RRS_2$:
+
+\begin{prop}\label{prop:rdc-rrs2-dc}
+Over~$\CZF^-$, $\RDC$ is equivalent to~$\RRS_2 + \DC$.
+\end{prop}
+
+\begin{proof}Trivially,~$\RDC$ entails~$\DC$, and~$\RRS_2 + \DC$
+entails~$\RDC$ by Aczel's result since~$\RRS_2$ entails~$\RRS$.
+
+To verify that~$\RDC$ entails~$\RRS_2$, let classes~$X$ and~$R \subseteq X
+\times X \times X$ be given, let~$x_0 \in X$ be an element and assume~$R : X
+\times X \rightrightarrows X$. Let~$\List(X)$ be the class of finite lists with
+entries in~$X$. We declare a class~$\widehat X$ by
+\[ \widehat X \defeq \{ \langle i,j,v \rangle \,|\, i,j \in
+\NN, v \in \List(X), i,j < \length(v) \} \]
+and a relation~$\widehat R \subseteq \widehat X \times \widehat X$ by
+defining~$\langle \langle i,j,v \rangle, \langle i',j',v' \rangle \rangle \in
+\widehat R$ to be equivalent to
+\begin{indentblock}
+there exists an element~$y \in X$ such that
+\begin{enumerate}
+\item $\langle v!i, v!j, y \rangle \in R$ (where~$v!k$ is the element of the
+list~$v$ at position~$k$),
+\item $v'$ is obtained from~$v$ by adding the single element~$y$ at the end and
+\item $\langle i',j' \rangle$ is the next point after~$\langle i,j
+\rangle$ in some fixed enumeration of~$\NN^2$ which, for each number~$n$, first visits
+all points~$\langle k,l \rangle$ with~$k,l < n$ before it visits any of
+the other points.
+\end{enumerate}
+\end{indentblock}
+Then~$\widehat R : \widehat X \rightrightarrows \widehat X$. Applying~$\RDC$
+with initial value~$\langle 0,0, [x_0] \rangle \in \widehat X$ yields a
+function~$f : \NN \to \widehat X$. Let~$B$ be the set of all
+entries of the lists contained in the tuples~$f(k)$. Then~$x_0 \in B
+\subseteq X$ and~$R : B \times B \rightrightarrows B$.
+\end{proof}
+
+A consequence of Proposition~\ref{prop:rdc-rrs2-dc} is that~$\RRS_2$ holds in
+Aczel's type-theoretic ``sets as trees'' model
+of~$\CZF$~\cite{aczel:sets-as-trees}, since that model validates~$\RDC$.
+
+\begin{prop}Over~$\CZF^-$, $\RRS_2$ is equivalent to~$\MDC_2$.\end{prop}
+
+\begin{proof}The proof in~\cite{palmgren:mdc} carries over.\end{proof}
+
+% XXX: RRS_2 holds in ZF.
+
 \section{Constructivizing the reflection principle}
@@
 
-Our constructive rendition of the reflection principle will require Aczel's
-relation reflection scheme~$\RRS$ displayed in Definition~\ref{defn:rrs}. This
-result is the best possible, as we verify in Theorem~\ref{thm:refl-entails-rrs}
-that conversely the reflection principle entails the
-relation reflection scheme. The relation reflection scheme first surfaced in
-the theory of coinductive definitions of classes and enjoys substantial
-stability properties, as it passes from the meta theory to XXX[all kinds of]
-models. Background on~$\RRS$ can be found in Aczel's original article
-introducing it~\cite{aczel:rrs}.
-
-\begin{defn}[$\RRS$, Aczel's relation reflection scheme]\label{defn:rrs}
-Let~$X$ and~$R \subseteq X \times X$ be classes. Let~$A \subseteq X$ be a
-subset. If~$\forall x \in X\_ \exists y \in X\_ \langle x,y \rangle \in R$,
-then there is a set~$B$ such that~$A \subseteq B \subseteq X$ and such
-that~$\forall x \in B\_ \exists y \in B\_ \langle x,y \rangle \in R$.
-\end{defn}
-
-For our purposes, Palmgren's multivalued dependent choice is slightly more
-convenient to work with than~$\RRS$. It is equivalent, over~$\CZF$ and
-a~fortiori over~$\IZF$, to~$\RRS$~\cite{palmgren:mdc}.
-
-\begin{defn}[$\MDC$, Palmgren's multivalued dependent choice]
-Let~$X$ and~$R \subseteq X \times X$ be classes. Let~$A \subseteq X$ be a
-subset. If~$\forall x \in X\_ \exists y \in X\_ \langle x,y \rangle \in R$,
-then there is a function~$f : \NN \to P(X)$ (the class of all subsets of~$X$)
-such that~$A \subseteq f(0)$ and such that~$\forall x \in f(k)\_ \exists y \in
-f(k+1)\_ \langle x,y \rangle \in R$ for every number~$k$.\end{defn}
+Our constructive rendition of the reflection principle will require the axiom
+scheme~$\RRS_2$ displayed in Definition~\ref{defn:rrs2}. This result is the
+best possible, as we verify in Theorem~\ref{thm:refl-entails-rrs2} that
+conversely the reflection principle entails~$\RRS_2$.
 
@@
 Let~$\varphi(u_1,\ldots,u_n,x)$ be a formula in the language of set
-theory with free variables as indicated. Then~$\IZF+\RRS$ proves
+theory with free variables as indicated. Then~$\IZF+\RRS_2$ proves
 \begin{multline*}
@@
 
-\begin{proof}We first consider the case~$n = 1$.
+\begin{proof}By~$\RRS_2$, there is a set~$B$ such that~$M \in B$ and such that
+for any~$H_1, H_2 \in B$, there is a set~$H' \in B$ such that~$H \defeq H_1
+\cup H_2$ and~$H'$ are related as in the conclusion of Lemma~\ref{lemma:izf-microstep}.
 
-By~$\MDC$ and Lemma~\ref{lemma:izf-microstep}, there is a
-function~$f : \NN \to P(V)$, where~$V$ is the universe, such that~$M \in
-f(0)$ and such that for any number~$k$ and any set~$H \in f(k)$, there is a
-set~$H' \in f(k+1)$ such that~$H$ and~$H'$ are related as in the conclusion of
+We set~$S \defeq \bigcup B$. This set has the required property. To verify
+this, it is useful to observe that given~$u_1,\ldots,u_n \in S$, there is a
+common set~$H \in B$ such that~$u_1,\ldots,u_n \in H$.
+
+In order to ensure addendum~(1) and~(2), we apply~$\RRS_2$ in a slightly
+different way to guarantee that for any sets~$H_1, H_2 \in B$, there is a
+set~$H' \in B$ such that~$H \defeq H_1 \cup H_2$ and~$H'$ are related as in the
+conclusion of Lemma~\ref{lemma:izf-microstep} and such that furthermore~$H'$ is
+transitive and closed under subsets. Even though it cannot be expected that any
+particular set~$H \in B$ will be transitive and closed under subsets, the
+union~$S$ will. Addendum~(3) can be ensured because of addendum~(3) of
 Lemma~\ref{lemma:izf-microstep}.
+\end{proof}
 
-We set~$S \defeq \bigcup \bigl(\bigcup_{k \in \NN} f(k)\bigr)$. This set~$S$
-has the required property.
-
-In order to ensure addendum~(1) and~(2), we apply~$\MDC$ in a slightly
-different way such that for any number~$k$ and any set~$H \in f(k)$, there is a
-set~$H' \in f(k+1)$ such that~$H$ and~$H'$ are related as in the conclusion of
-Lemma~\ref{lemma:izf-microstep} and such that~$H'$ is transitive and closed
-under subsets. Even though it cannot be expected that for any number~$k$,
-any set~$H \in f(k)$ will be transitive and closed under subsets, the union~$S$
-will. Addendum~(3) can be ensured because of addendum~(3) of
-Lemma~\ref{lemma:izf-microstep}.
-
-For the case~$n \neq 1$, \textbf{???}
-\end{proof}
+\begin{rem}In the special case~$n = 1$, the proof of Lemma~\ref{lemma:izf-smallstep} can
+be simplified to only use~$\RRS$ instead of~$\RRS_2$.\end{rem}
 
@@
 Let~$\varphi(x_1,\ldots,x_n)$ be a formula in the language of set
-theory with free variables as indicated. Then~$\IZF+\RRS$ proves
+theory with free variables as indicated. Then~$\IZF+\RRS_2$ proves
 \[ \forall M\_
@@
 universal quantifier has to be treated just as the case for the existential
-quantifier is.\end{proof}
+quantifier has to.\end{proof}
 
-\begin{thm}\label{thm:refl-entails-rrs}
-Over~$\IZF$, each instance of Aczel's relation reflection scheme~$\RRS$ can be
+% XXX: State mini-reflection theorem.
+
+\begin{thm}\label{thm:refl-entails-rrs2}
+Over~$\IZF$, each instance of Aczel's relation reflection scheme~$\RRS_2$ can be
 deduced from suitable instances of the assumption that, given a finite list
@@
 
-\begin{proof}We verify the equivalent~$\MDC$ instead of~$\RRS$. Let~$X$ and~$R
-\subseteq X \times X$ be classes. Let~$A \subseteq X$ be a subset and
-suppose~$\forall x \in X\_ \exists y \in X\_ \langle x,y \rangle \in R$.
+\begin{proof}Let~$X$ and~$R \subseteq X \times X \times X$ be classes. Let~$x_0 \in X$
+be an element and suppose~$R : X \times X \rightrightarrows X$.
 
-By assumption, there is a set~$S \supseteq A$ which reflects the three
-formulas~``$x \in X$'', ``$\langle x,y \rangle \in R$'' and~``$\exists y \in X\_ \langle
-x,y \rangle \in R$''.
+By assumption, there is a set~$S \ni x_0$ which reflects the three
+formulas~``$x \in X$'', ``$\langle x,x',y \rangle \in R$'' and~``$\exists y \in X\_ \langle
+x,x',y \rangle \in R$''.
 
-By separation, the class~$S \cap X$ is a set, and hence we may set~$f(k) \defeq S \cap X$
-for all numbers~$k$. Then~$A \subseteq f(0)$; any given~$x \in f(k)$, by
-assumption and the reflecting property of~$S$ there exists~$y \in S$ such
-that~$y \in X$ (hence~$y \in f(k+1)$) and~$\langle x,y \rangle \in R$.
+The class~$B \defeq X \cap S$ is a set by separation and contains~$x_0$.
+Given~$x,x' \in B$, there is a set~$y$ such that~$y \in X$ and~$\langle x,y,y'
+\rangle \in R$. By the reflecting property of~$S$, we can assume that such an
+element~$y$ exists in~$S$.
+
+Hence~$R : B \times B \rightrightarrows B$.
 \end{proof}
@@
 
-\begin{defn}[$\DC$, dependent choice]
-Let~$X$ and~$R \subseteq X \times X$ be sets. Let~$x_0 \in X$.
-If~$\forall x \in X\_ \exists y \in X\_ \langle x,y \rangle \in R$,
-then there is a function~$f : \NN \to X$ such that~$f(0) = x_0$ and
-such that~$\langle f(k), f(k+1) \rangle \in R$ for all numbers~$k$.
-\end{defn}
-
-\begin{defn}[$\RDC$, relative dependent choice]
-Let~$X$ and~$R \subseteq X \times X$ be classes. Let~$x_0 \in X$.
-If~$\forall x \in X\_ \exists y \in X\_ \langle x,y \rangle \in R$,
-then there is a function~$f : \NN \to X$ such that~$f(0) = x_0$ and
-such that~$\langle f(k), f(k+1) \rangle \in R$ for all numbers~$k$.
-\end{defn}
-
 \begin{prop}Over~$\IZF$, $\RDC$ is equivalent to~$\RRS + \DC$.
@@
 
-\begin{proof}It's clear that~$\RDC$ implies~$\RRS + \DC$. % XXX: check
+\begin{proof}Trivially,~$\RDC$ implies~$\DC$, and~$\RDC$ implies~$\RRS$ by a
+similar, though much simpler, argument as in the proof of
+Proposition~\ref{prop:rdc-rrs2-dc}.
 
@@
 
+\textbf{XXX: Use mini-reflection theorem instead.}
 By Theorem~\ref{thm:izf-bigstep}, reflection is available; hence there is a
@@
 \section{An intuitionistic version of Feferman's~$\ZFCS$}
+\label{sect:appl-smallness}
 
@@
 that primitive recursive arithmetic~$\PRA$ proves that for any
-formula~$\varphi$ of~$\ZFC$, $\ZFC$ proves that~$(\Set \models \varphi)
-\Leftrightarrow \varphi$, where~``$\Set \models \varphi$'' is to be unrolled
+formula~$\varphi$ of~$\ZFC$, $\ZFC$ proves~``$(\Set \models \varphi)
+\Leftrightarrow \varphi$'', where~``$\Set \models \varphi$'' is to be unrolled
 by~$\PRA$.
@@
 strongly inaccessible cardinal. In this system, there is a Grothendieck
-universe~$U$, we can form the category~$\Set_U$ of all sets of~$U$ as an honest
-set, can define truth in~$\Set_U$ and prove, within the system, that for any
+universe~$U$; we can form the category~$\Set_U$ of all sets of~$U$ as an honest
+set; can define truth in~$\Set_U$; and prove, within the system, that for any
 formula~$\varphi$, $(\Set_U \models \varphi) \Leftrightarrow (U \models
 \varphi)$. This even holds for formulas of the full infinitary language of
-toposes, which allows for infinite disjunctions and infinite conjunctions; this
-could not be mimicked by resorting to~$\PRA$ as indicated above.
+toposes, which allows infinite disjunctions and infinite conjunctions; this
+extended language could not be treated by resorting to~$\PRA$ as indicated above.
 
@@
 
-% XXX: Reference more choice axioms?
+% XXX: Reference more choice axioms? e.g. Aczel and Rathjen; and that one PhD
+% thesis

### 2019-01-11 paper-reflection.tex
Baby form of the reflection theorem

@@
 
-\begin{rem}In the special case~$n = 1$, the proof of Lemma~\ref{lemma:izf-smallstep} can
-be simplified to only use~$\RRS$ instead of~$\RRS_2$.\end{rem}
+\begin{rem}\label{rem:izf-smallstep-1}
+In the special case~$n = 1$, the proof of Lemma~\ref{lemma:izf-smallstep} can
+be simplified to only use~$\RRS$ instead of~$\RRS_2$, because in this case it
+suffices for the set~$B$ to be such that for any~$H \in B$, there is a set~$H'
+\in B$ such that~$H$ and~$H'$ are related as in the conclusion of
+Lemma~\ref{lemma:izf-microstep}.
+\end{rem}
 
@@
 
-% XXX: State mini-reflection theorem.
+\begin{scholium}\label{scholium:izf-bigstep-1}
+Let~$\varphi(x_1,\ldots,x_n)$ be a formula in the language of set
+theory with free variables as indicated. Assume that the surrounding scope of
+any unbounded quantifier in~$\varphi$ contains at most one free variable.
+Then~$\IZF+\RRS$ proves the same conclusion as stated in
+Theorem~\ref{thm:izf-bigstep}.
+\end{scholium}
+
+\begin{proof}The condition on the number of free variables allows the proof of
+Theorem~\ref{thm:izf-bigstep} to be adapted to employ the version of
+Lemma~\ref{lemma:izf-smallstep} outlined in Remark~\ref{rem:izf-smallstep-1},
+which requires only~$\IZF+\RRS$ instead of~$\IZF+\RRS_2$.
+\end{proof}
 
@@
 
-\textbf{XXX: Use mini-reflection theorem instead.}
-By Theorem~\ref{thm:izf-bigstep}, reflection is available; hence there is a
-set~$S \supseteq \{x_0\}$ which reflects the three formulas~``$x \in X$'', ``$\langle x,y
-\rangle \in R$'' and~``$\forall x\_ x \in X \Rightarrow \exists y\_ (y \in X \wedge \langle x,y
-\rangle \in R)$''.
+We cannot apply Scholium~\ref{scholium:izf-bigstep-1} to the formula~``$\forall
+x \in X\_ \exists y \in X\_ \langle x,y \rangle \in R$'' since unbounded
+quantifiers implicitly appearing in the formulas~``$x \in X$'' and~``$\langle
+x,y \rangle \in R$'' (recalling that~$X$ and~$R$ are classes) may violate the
+condition on the number of free variables. However, we can opt to leave these
+subformulas to be untranslated when carrying out the~$S$-relativization; with
+this understanding, Scholium~\ref{scholium:izf-bigstep-1} can be applied to
+yield a set~$S \ni x_0$ such that
+\[ \forall x \in S\_ (x \in X \Rightarrow \exists y \in S\_
+  (y \in X \wedge \langle x,y \rangle \in R)). \]
 

### 2019-01-12 paper-reflection.tex
Tied some loose ends

@@
 \newcommand{\xra}[1]{\xrightarrow{#1}}
-\newcommand{\XXX}[1]{\textbf{XXX: #1}}
 \newcommand{\aaa}{\mathfrak{a}}
@@
 \newcommand{\B}{\mathcal{B}}
+\newcommand{\EE}{\mathbb{E}}
 \newcommand{\NN}{\mathbb{N}}
@@
 \newcommand{\IZF}{\textsc{izf}}
+\newcommand{\IZFS}{\textsc{izf/s}}
 \newcommand{\CZF}{\textsc{czf}}
@@
 \begin{thm}Let~$\varphi(x_1,\ldots,x_n)$ be a formula in the language of set
-theory with free variables as indicated. Then~$\ZF$ proves
+theory with some of its free variables as indicated (and further free variables
+allowed). Then~$\ZF$ proves
 \[ \forall M\_
@@
 
-The reflection principle is important both for philosophical and for practical
-reasons: Philosophically, it tells us that truth of any formula can already be
-checked in an initial segment of the universe. Practically, it allows to
-% XXX: be more precise about philosophy
+The reflection principle expresses that truth of any formula can already be
+checked in an initial segment of the universe. This observation
+is important both for philosophical and for practical
+reasons: Philosophically, it tells us that the set-theoretic universe cannot be
+distinguished from its initial segments by any set-theoretical property.
+Practically, it allows to
 transfer results obtained for a restricted class of objects to all such
@@
 then we may deduce by the reflection principle that the statement holds
-for all groups in the universe.
+for all groups in the universe. Examples from sheaf cohomology and more
+generally category theory abound; an example from set theory is presented in
+Proposition~\ref{prop:zf-rrs2}.
 
-This observation has been used by Feferman to construct~$\ZFCS$ (``$\ZFC$ with
+The reflection principle has been used by Feferman to construct~$\ZFCS$ (``$\ZFC$ with
 smallness''), a conservative extension of~$\ZFC$ which provides a useful
 foundation of category theory~\cite{feferman:zfcs}. This system extends~$\ZFC$
-by the addition of a new constant symbol~$\SS$ together with axioms stating
+by a new constant symbol~$\SS$ together with axioms stating
 that~$\SS$ is transitive, closed under subsets and reflective with respect to
@@
 To our naive eyes, the passage from~$\ZFC$ to~$\ZFCS$ looked sufficiently
-innocent so that we set out to create an intuitionistic version of~$\ZFCS$: To
-our mind, size issues were entirely separate from issues of constructivity, and
+innocent so that we set out to develop an intuitionistic version of~$\ZFCS$: To
+our mind, size issues were entirely different concerns than issues of constructivity, and
 we hence opined that they should be dealt with separately. Such a separation
@@
 situation turned out to be more subtle and we failed in our original goal of
-verifying the reflection principle for intuitionistic Zermelo--Fraenkel set
+verifying the reflection principle in intuitionistic Zermelo--Fraenkel set
 theory~$\IZF$. The situation for~$\CZF$, the predicative subsystem of~$\IZF$
 commonly heralded as the largest common denominator of all flavors of
-constructive set theory, is even more mysterious.
+constructive set theory, remains even more elusive.
 
-However, we succeeded in verifying the reflection principle for a slight
+However, we succeeded in verifying the reflection principle in only a slight
 extension of~$\IZF$:
@@
 
-\begin{conjecture}The system $\IZF$ does not prove the reflection
-principle.\end{conjecture}
-
-\begin{defn}The system $\ES$ is obtained from~$\IZF+\RRS_2$ by
-adding a constant symbol~$\SS$ together with axioms
-stating that~$\SS$ is transitive, closed under subsets and reflective for all
-formulas of the original language (that is, adding one instance
-of~\eqref{ax:refl} for each formula of~$\IZF$).\end{defn}
-
-\begin{cor}The system~$\ES$ is conservative
-over~$\IZF+\RRS_2$.\end{cor}
+% Since~$\ZF$ proves~$\RRS_2$, our proof of the reflection principle also serves
+% as a new classical proof. However, the standard proof is shorter, XXX
+% Hint that we don't need regularity.
 
@@
 Let~$\varphi(u_1,\ldots,u_n,x)$ be a formula in the language of set
-theory with free variables as indicated. Then~$\ZF$ proves
+theory. Then~$\ZF$ proves
 \begin{multline*}
@@
 bounds all of the formulas~$\varphi_i$.
-% XXX: better explanation
 \end{lemma}
@@
 Let~$\varphi(x_1,\ldots,x_n)$ be a formula in the language of set
-theory with free variables as indicated. Then~$\ZF$ proves
+theory. Then~$\ZF$ proves
 \[ \forall M\_
@@
   \forall x_1,\ldots,x_n \in S\_
-  \varphi(x_1,\ldots,x_n) \Leftrightarrow \varphi^S(x_1,\ldots,x_n), \]
-where~$\varphi^S$ is the~$S$-relativization of~$\varphi$. Furthermore, the
+  \varphi(x_1,\ldots,x_n) \Leftrightarrow \varphi^S(x_1,\ldots,x_n). \]
+Furthermore, the
 resulting set~$S$ may be supposed to be transitive and to be closed under
@@
 \begin{proof}Let a set~$M$ be given. We obtain~$S$ by applying
-Lemma~\ref{lemma:zf-smallstep} to the list of all subformulas of~$\varphi$. That
-% XXX: "all" subformulas
-the resulting set~$S$ has the required property can be checked by an induction on the structure
+Lemma~\ref{lemma:zf-smallstep} to the list of all subformulas of~$\varphi$
+which start with an existential quantifier. That
+the resulting set~$S$ has the required property can then be checked by an induction on the structure
 of~$\varphi$. The cases~$\text{``$=$''}, \text{``$\in$''}, \text{``$\top$''},
@@
 The proof of Theorem~\ref{thm:zf-bigstep} is constructive with the sole
-exception of treating the case of universal quantifiers by appealing to the law
-of excluded middle. For the constructive proof in
-Section~\ref{sect:constructive-proof}, we will instead appeal to a strengthened
+exceptions of treating the case of universal quantifiers by appealing to the law
+of excluded middle and referring to the unconstructive Lemma~\ref{lemma:zf-smallstep}. For the constructive proof in
+Section~\ref{sect:constructive-proof}, we will solve both issues by instead appealing to a strengthened
 version of Lemma~\ref{lemma:zf-smallstep}.
@@
 \label{sect:axioms}
+
+We follow the usual convention that~$\IZF$ is the set theory with the following
+axioms: extensionality, pair, union, empty set, infinity, unbounded separation,
+collection, powerset and~$\in$-induction. We direct the reader wishing for a
+survey of~$\IZF$ and other nonclassical set theories to Crosilla's entry in the
+SEP~\cite{crosilla:cst-izf}.
 
@@
 models. Background on~$\RRS$ can be found in Aczel's original article
-introducing it~\cite{aczel:rrs}. $\RRS$ is equivalent, over~$\CZF$ and
+introducing it~\cite{aczel:rrs}, and more information on choice axioms in
+general is contained in the book draft by Aczel and
+Rathjen~\cite[Section~10]{aczel-rathjen:cst}.
+$\RRS$ is equivalent, over~$\CZF$ and
 a~fortiori over~$\IZF$, to Palmgren's multivalued dependent
@@
 stronger theory~$\IZF$, in Section~\ref{sect:appl-rdc}.
+
+The axiom scheme~$\RRS_2$ does not appear to have been studied much. Apart from
+the PhD thesis by Ziegler~\cite{ziegler:phd}, where it is called the
+\emph{strong relation reflection principle}, we have not been able to track
+down further mentions of it in the literature; hence it seems prudent to verify
+some of its basic properties here.
 
@@
 
-% XXX: RRS_2 holds in ZF.
+\begin{prop}\label{prop:zf-rrs2}
+$\ZF$ proves~$\RRS_2$.\end{prop}
+
+\begin{proof}Let~$X$ and~$R \subseteq X \times X$ be classes. Let~$x_0 \in X$
+and assume~$R : X \rightrightarrows X$. By the reflection principle, there is a
+set~$S \ni x_0$ such that~$R : X \cap S \rightrightarrows X \cap S$. This
+concludes the proof as the class~$B \defeq X \cap S$ is a set by separation.
+\end{proof}
 
@@
 Let~$\varphi(u_1,\ldots,u_n,x)$ be a formula in the language of set
-theory with free variables as indicated. Then~$\IZF$ proves
+theory. Then~$\IZF$ proves
 \begin{multline*}
@@
 
+The proof of Lemma~\ref{lemma:izf-microstep} makes crucial use of unbounded
+separation and the powerset axiom. Hence we do not believe that it can
+be improved to work over~$\CZF$. Since we do not have any uniqueness guarantee
+on the~``$\exists x$'' quantifiers in the proof, it also does not work
+over~$\IZF_\text{Rep}$, the variant of~$\IZF$ with replacement instead of
+collection.
+
 \begin{lemma}\label{lemma:izf-smallstep}
 Let~$\varphi(u_1,\ldots,u_n,x)$ be a formula in the language of set
-theory with free variables as indicated. Then~$\IZF+\RRS_2$ proves
+theory. Then~$\IZF+\RRS_2$ proves
 \begin{multline*}
@@
 formulas~$\varphi_i$.
-% XXX: better explanation
 \end{lemma}
@@
 Let~$\varphi(x_1,\ldots,x_n)$ be a formula in the language of set
-theory with free variables as indicated. Then~$\IZF+\RRS_2$ proves
+theory. Then~$\IZF+\RRS_2$ proves
 \[ \forall M\_
@@
 Let~$\varphi(x_1,\ldots,x_n)$ be a formula in the language of set
-theory with free variables as indicated. Assume that the surrounding scope of
+theory. Assume that the surrounding scope of
 any unbounded quantifier in~$\varphi$ contains at most one free variable.
@@
 
-XXX: repeat definition and conservativity statement?
+\begin{defn}The system $\ES$ is obtained from~$\IZF+\RRS_2$ by
+adding a constant symbol~$\SS$ together with axioms
+stating that~$\SS$ is transitive, closed under subsets and reflective for all
+formulas of the original language.\end{defn}
+
+\begin{prop}The system~$\ES$ is conservative
+over~$\IZF+\RRS_2$.\end{prop}
+
+\begin{proof}Because the reflection principle is available in~$\IZF+\RRS_2$,
+the same argument as for~$\ZFCS$ applies.\end{proof}
+
+Just as~$\ZFCS$ can serve as a set-theoretic foundation for category theory in
+a classical context, we argue that~$\ES$ can serve as such a foundation in an
+intuitionistic context (provided, of course, one is willing to
+accept~$\RRS_2$).
 
 The system~$\ES$ is also interesting from the point of view of topos theory.
-Any topos supports an internal language which can be used to reason about the
+We recall that any topos supports an internal language which can be used to reason about the
 objects and morphisms of the topos in a naive element-based language, allowing
@@
 \cite[Section~1.3]{caramello:ttt}, \cite[Chapter~14]{goldblatt:topoi},
-\cite[Chapter~VI]{moerdijk-maclane:sheaves-logic}).
+\cite[Chapter~VI]{moerdijk-maclane:sheaves-logic}). We refer
+to~\cite[Sections~1 and~2]{blechschmidt:phd} for a short introduction and a
+review of some of the applications of the internal language.
 
@@
 However, truth in a Grothendieck universe~$U$ need not be related to actual
-truth. A solution is provided by~$\ZFCS$ and by~$\ES$. In these systems, we can
+truth. A solution to this problem is provided by~$\ZFCS$ and by~$\ES$. In these systems, we can
 form the topos~$\Set_\SS$, define truth in it, prove for all formulas~$\varphi$
@@
 
+\section{Outlook}
+
+We proved that, over~$\IZF$, the reflection principle is equivalent
+to~$\RRS_2$. This gives credence to the claim that the reflection principle is
+not provable over~$\IZF$ alone. However, the following question remains open:
+
+\begin{question}Does~$\IZF$ prove~$\RRS_2$?\end{question}
+
+If the answer is in the negative, as is most likely, then no conservative
+extension could include a constant symbol~$\SS$ should that~$\SS$-relativized
+truth is the same as absolute truth.
+
+However, truth in toposes is a more flexible notion than~$S$-relativized
+truth for any set~$S$. Hence one might hope that even in this case, the
+following question does have a positive answer:
+
+\begin{question}Is there a conservative extension~$\IZF'$ of~$\IZF$, containing
+a constant symbol~$\EE$, such that~$\IZF'$ proves that~$\EE$ is an elementary
+topos and such that~$\IZF'$ proves~``$(\EE \models \varphi) \Leftrightarrow
+\varphi$'' for any formula~$\varphi$?\end{question}
+
 \printbibliography
@@
 \end{document}
-
-% XXX: Review axioms of IZF.
-
-% XXX: Think about pairs.
-
-% XXX: Remark that the new proof of reflection also works over ZF, but that the
-% usual proof is probably preferable.
-
-% XXX: Über Verallgemeinerung zu CZF spekulieren.
-
-% XXX: Note that the result fails over IZF_R.
-
-% XXX: Reference more choice axioms? e.g. Aczel and Rathjen; and that one PhD
-% thesis

### 2019-01-12 paper-reflection.tex
typo

@@
 \usepackage[utf8]{inputenc}
+\usepackage[english]{babel}
 \usepackage{amsthm,mathtools,stmaryrd,amssymb,graphicx}
@@
 
-\title[]{Reflections on reflection for \\ intuitionistic set theories}
+\title[Reflections on reflection for intuitionistic set theories]{Reflections on reflection for \\ intuitionistic set theories}
 \author{}
@@
 objects. For instance, if we manage to verify (the $S$-relativization of) some
-group-theoretic statement for all groups contained in an arbitrary set~$X$,
-then we may deduce by the reflection principle that the statement holds
+group-theoretic statement for all groups contained in an arbitrary set~$S$,
+then we may deduce that the statement holds
 for all groups in the universe. Examples from sheaf cohomology and more
@@
 \DC$ over~$\CZF^-$, and we give a new proof of this equivalence over the
-stronger theory~$\IZF$, in Section~\ref{sect:appl-rdc}.
+stronger theory~$\IZF$ in Section~\ref{sect:appl-rdc}.
 
@@
   \forall x_1,\ldots,x_n \in S\_
-  \varphi(x_1,\ldots,x_n) \Leftrightarrow \varphi^S(x_1,\ldots,x_n), \]
-where~$\varphi^S$ is the~$S$-relativization of~$\varphi$.
+  \varphi(x_1,\ldots,x_n) \Leftrightarrow \varphi^S(x_1,\ldots,x_n). \]
 Furthermore, the resulting set~$S$ may be supposed to be transitive and to be closed
@@
 strongly inaccessible cardinal. In this system, there is a Grothendieck
-universe~$U$; we can form the category~$\Set_U$ of all sets of~$U$ as an honest
+universe~$U$; we can form the category~$\Set_U$ of all sets in~$U$ as an honest
 set; can define truth in~$\Set_U$; and prove, within the system, that for any
@@
 truth. A solution to this problem is provided by~$\ZFCS$ and by~$\ES$. In these systems, we can
-form the topos~$\Set_\SS$, define truth in it, prove for all formulas~$\varphi$
+form the topos~$\Set_\SS$ of all sets in~$\SS$, define truth in it, prove for all formulas~$\varphi$
 that~$(\Set_\SS \models \varphi) \Leftrightarrow (\SS \models \varphi)$; and

### 2019-01-30 paper-reflection.tex
typo (thanks to Daniel Albert)

@@
 We expected this modification to be entirely straightforward. However, the
-situation turned out to be more subtle and we failed in our original goal of
+situation turned out to be more subtle, and we failed in our original goal of
 verifying the reflection principle in intuitionistic Zermelo--Fraenkel set
@@
 
-\textbf{Acknowledgments.} XXX
+\textbf{Acknowledgments.} We are grateful to Daniel Albert for his careful
+reading of earlier drafts. XXX
 
@@
 of formulas instead of the single formula~$\varphi$, we may suppose that~$S$
-bounds all of the formulas~$\varphi_i$.
+bounds all the formulas~$\varphi_i$.
 \end{lemma}
@@
 The proof of Theorem~\ref{thm:zf-bigstep} is constructive with the sole
-exceptions of treating the case of universal quantifiers by appealing to the law
+exception of treating the case of universal quantifiers by appealing to the law
 of excluded middle and referring to the unconstructive Lemma~\ref{lemma:zf-smallstep}. For the constructive proof in
@@
 list~$\varphi_1,\ldots,\varphi_s$ of formulas instead of the single
-formula~$\varphi$, we may suppose that~$S$ bounds all of the
+formula~$\varphi$, we may suppose that~$S$ bounds all the
 formulas~$\varphi_i$.

### 2019-02-01 paper-reflection.tex
typo

@@
 If the answer is in the negative, as is most likely, then no conservative
-extension could include a constant symbol~$\SS$ should that~$\SS$-relativized
+extension could include a constant symbol~$\SS$ such that~$\SS$-relativized
 truth is the same as absolute truth.

### 2019-02-20 paper-macneille-reals.tex
typo
Version as uploaded to the arXiv

@@
 \usetikzlibrary{patterns,matrix}
-\usepackage[expansion=true,protrusion=true]{microtype}
-\usepackage{breakurl}
+%\usepackage[expansion=true,protrusion=true]{microtype}
+%\usepackage{breakurl}
 \usepackage{hyperref}
+
+\usepackage[natbib=true,style=numeric,maxnames=10]{biblatex}
+\usepackage[babel]{csquotes}
+\bibliography{paper-macneille-reals.bib}
 
@@
 
-\title{A constructive Knaster--Tarski proof of the uncountability of the reals}
-\author{Ingo Blechschmidt and Matthias Hutzler}
-%\address{Max Planck Institute for Mathematics in the Sciences \\
-%Inselstraße 22 \\
-%04103 Leipzig, Germany}
-\email{ingo.blechschmidt@mis.mpg.de, matthias.ralph.hutzler@student.uni-augsburg.de}
+\title[]{A constructive Knaster--Tarski proof of the uncountability of the reals}
+
+\author{Ingo Blechschmidt}
+\address{Università di Verona \\
+Department of Computer Science \\
+Strada le Grazie 15 \\
+37134 Verona, Italy}
+\email{ingo.blechschmidt@univr.it}
+
+\author{Matthias Hutzler}
+\address{Universität Augsburg \\
+Institut für Mathematik \\
+Universitätsstr. 14 \\
+86159 Augsburg, Germany}
+\email{matthias.ralph.hutzler@student.uni-augsburg.de}
 
@@
 \begin{abstract}
-  We give an uncountability proof of the reals which does not rely on their
-  sequential completeness, but on their order completeness. We use neither a
+  We give an uncountability proof of the reals which relies on their
+  order completeness instead of their sequential completeness. We use neither a
   form of the axiom of choice nor the law of excluded middle, therefore
@@
 the reals. In constructive mathematics, more precisely the kind of mathematics
-which can be carried out in any topos, the first step is still valid, while the
-second might fail -- for any of the several possible flavors of the reals
+which can be carried out in any topos or the kind of mathematics formalizable
+in Intuitionistic Zermelo--Fraenkel set theory, the first step is still valid, while the
+second might fail. This failure may occur for any of the several possible flavors of the reals
 such as the Cauchy reals, the Dedekind reals or the MacNeille reals (which all
@@
 an understanding whether any of the more interesting flavors of the reals can constructively be
-shown to be uncountable. The proof presented here is made possible because the
+shown to be uncountable, a problem posed to us by Andrej Bauer who we
+gratefully acknowledge. The proof presented here is made possible because the
 MacNeille reals -- unlike the Cauchy or Dedekind reals -- can constructively be
-shown to be order-complete.
+shown to be (conditionally) order complete~\cite[Lemma~D4.7.7]{johnstone:elephant}.
 
@@
 computer science, but appears to be seldomly used in classical analysis.
-The proof is an adaptation of Levy's unusual proof~\cite{levy}.
+The proof is an adaptation of Levy's unusual proof~\cite{levy:uncountability}.
 
 \begin{thm*}Let~$f : \NN \to \RR$ be a map. Then there is a number~$x_0 \in \RR$
-such that for no~$n \in \NN$, $f(n) = x_0$.\end{thm*}
+such that for no~$n_0 \in \NN$, $f(n_0) = x_0$.\end{thm*}
 
@@
   x \longmapsto \sup_M \sum_{n \in M} 2^{-n}, \]
-where~$M$ ranges over all those finite detachable subsets of~$\NN$ such
+where~$M$ ranges over all those (Bishop-)finite subsets of~$\NN$ such
 that~$f[M] < x$, is well-defined (because the sets the suprema are taken of are
@@
 
-If~$x_0 = f(n_0)$ for a number~$n_0 \in \NN$, then for any finite detachable
+If~$x_0 = f(n_0)$ for a number~$n_0 \in \NN$, then for any finite
 subset~$M$ of~$\NN$ such that~$f[M] < x_0$,
@@
 hence~$g(x_0 + 2^{-n_0}) \geq g(x_0) + 2^{-n_0} \geq x_0 + 2^{-n_0}$.
-Thus~$x_0 + 2^{-n_0}$ is a greater postfixpoint that~$x_0$, a contradiction.
+Thus~$x_0 + 2^{-n_0}$ is a greater postfixpoint than~$x_0$, a contradiction.
 \end{proof}
@@
 of~$A$: Let~$x \in A$. If~$x > x_0 - 2^{-n_0}$, then~$g(x +
-2^{-n_0}) \geq g(x) + 2^{-n_0} \geq x + 2^{-n_0}$, hence~$x + 2^{-n_0} \in A$,
+2^{-n_0}) \geq g(x) + 2^{-n_0} \geq x + 2^{-n_0}$ (where the first inequality
+is as in the first proof, exploiting that~$x \leq x_0$ by definition of~$x_0$), hence~$x + 2^{-n_0} \in A$,
 thus~$x + 2^{-n_0} \leq x_0$, a contradiction. Hence~$x \leq x_0 - 2^{-n_0}$.

### 2019-03-03 paper-flabby-objects.tex
Be more economical in the use of packages
typo

@@
 \usepackage[utf8]{inputenc}
-\usepackage{amsthm,mathtools,tikz,float,caption,stmaryrd}
-\usetikzlibrary{patterns,matrix,decorations.pathreplacing}
+\usepackage{amsthm,mathtools,stmaryrd}
 \usepackage[all]{xy}
@@
 algebras~\cite{blass:free-algebras}.} The existence of quotient inductive types
-seem to be, as the existence of enough injective modules, \emph{constructively
+seems to be, as the existence of enough injective modules, \emph{constructively
 neutral}.

### 2019-03-05 paper-qcoh.tex
Begin of the paper on synthetic quasicoherence

\documentclass[oneside,reqno]{amsart}

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsthm,mathtools,stmaryrd,amssymb,graphicx}
\usepackage{booktabs}
\usepackage[all]{xy}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{xspace}

\usepackage[natbib=true,style=numeric,maxnames=10]{biblatex}
\usepackage[babel]{csquotes}
\bibliography{paper-qcoh.bib}

\title{A general Nullstellensatz for generalized spaces}
\author{Ingo Blechschmidt}
\email{ingo.blechschmidt@univr.it}

\theoremstyle{definition}
\newtheorem{defn}{Definition}[section]
\newtheorem{ex}[defn]{Example}

\theoremstyle{plain}
\newtheorem{prop}[defn]{Proposition}
\newtheorem{cor}[defn]{Corollary}
\newtheorem{lemma}[defn]{Lemma}
\newtheorem{thm}[defn]{Theorem}
\newtheorem{scholium}[defn]{Scholium}

\theoremstyle{remark}
\newtheorem{rem}[defn]{Remark}
\newtheorem{question}[defn]{Question}
\newtheorem{speculation}[defn]{Speculation}
\newtheorem{caveat}[defn]{Caveat}
\newtheorem{conjecture}[defn]{Conjecture}

\newenvironment{indentblock}{%
  \list{}{\leftmargin\leftmargin}%
  \item\relax
}{%
  \endlist
}

\newcommand{\xra}[1]{\xrightarrow{#1}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\B}{\mathcal{B}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\EE}{\mathbb{E}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\SS}{\mathbb{S}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\PSh}{PSh}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\length}{length}
\DeclareMathOperator{\List}{List}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\effective}{ef{}fective\xspace}
\newcommand{\?}{\,{:}\,}
\newcommand{\realizes}{\Vdash}
\newcommand{\notnot}{\emph{not~not}\xspace}
\usepackage{soul}\setul{0.3ex}{}
\let\oldul\ul
\renewcommand{\ul}[1]{\text{\oldul{$#1$}}}
\newcommand{\affl}{\ensuremath{{\ul{\AA}^1}}\xspace}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}

\newcommand{\stacksproject}[1]{\cite[{\href{https://stacks.math.columbia.edu/tag/#1}{Tag~#1}}]{stacks-project}}

\renewcommand{\paragraph}[1]{\noindent\textbf{#1.}}

\newcommand{\ZFC}{\textsc{zfc}}

\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}

\begin{document}

\begin{abstract}
  We give a general Nullstellensatz for the generic model of a
  geometric theory, useful as a source of nongeometric sequents validated by
  the generic model, and characterize the first-order formulas validated by the
  genetric model. XXX
\end{abstract}

\maketitle
\thispagestyle{empty}

\section{Introduction}

\paragraph{Generic models} Let~$\TT$ be a geometric theory, such as the theory
of rings, local rings or intervals. We follow Caramello's
terminology~\cite{caramello:tst} to mean by \emph{geometric theory} a system
given by a set of sorts, a set of finitary function symbols, a set of finitary
relation symbols and a set of axioms, consisting of geometric sequents
(sequents of the form~$\varphi \seq{\vec x} \psi$ where~$\varphi$ and~$\psi$
are geometric formulas, that is formulas built from equality and the relation
symbols by the logical connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$
and by arbitrary set-indexed disjunctions~$\bigvee$).

A fundamental result is that there is a \emph{generic model}~$U_\TT$ of~$\TT$,
a model such that for any geometric sequent~$\sigma$, the following notions
coincide: \begin{enumerate} \item The sequent~$\sigma$ is provable
modulo~$\TT$. \item The sequent~$\sigma$ holds for any~$\TT$-model in any
Grothendieck topos. \item The sequent~$\sigma$ holds for~$U_\TT$.
\end{enumerate} One could argue that it is this model which mathematicians
implicitly refer to when they utter ``Let~$M$ be a~$\TT$-model.''. It can
typically not be realized as a set-theoretic model, consisting of a set for
each sort, a function for each function symbol and so on; instead it is a model
in a custom-tailored syntactically constructed Grothendieck topos, the
\emph{classifying topos}~$\Set[\TT]$ of~$\TT$, hence consists of an object
of~$\Set[\TT]$ for each sort, a morphism for each function symbol and so on.

To state what it means for a~$\TT$-structure in a topos~$\E$ to verify the
axioms of~$\TT$, rendering it a model, the \emph{internal language} of~$\E$ is
used, roughly reviewed in Section~\ref{sect:review} below. We write~``$\E
\models \alpha$'' to mean that the statement~$\alpha$ holds from the internal
point of view of~$\E$. Since this language is a form of a higher-order
dependently-typed intuitionistic logic, the classifying topos~$\Set[\TT]$ can
be regarded as a higher-order completion of the geometric theory~$\TT$. The
generic model enjoys the universal property that any~$\TT$-model in any
(Grothendieck) topos~$\E$ is the pullback of~$U_\TT$ along an essentially
unique geometric morphism~$\E \to \Set[\TT]$.
\medskip

\paragraph{Nongeometric sequents} Crucially, the
equivalence~$(1)\Leftrightarrow(2)\Leftrightarrow(3)$ relating
provability and truth in~$\Set[\TT]$ only pertains to geometric sequents. The
generic model may validate additional nongeometric sequents which are not
provable from the axioms of~$\TT$ in first-order or even higher-order logic,
and these nongeometric sequents may be quite surprising and have useful
consequences.

One of the most celebrated such sequents arises in the case that~$\TT$ is the
theory of local rings. In this case, the classifying topos~$\Set[\TT]$ is also
known as the \emph{big Zariski topos} of~$\Spec(\ZZ)$ from algebraic geometry,
the topos of sheaves over the site of schemes locally of finite presentation,
and the generic model is the functor~$\affl$ of points of the affine line, the
functor which maps any (l.o.f.p.) scheme~$X$ to~$\Hom(X,\mathbb{A}^1) =
\O_X(X)$.

From the point of view of the topos, the ring object~$\affl$ is not only a
local ring, but even a field in the sense that
\[ \text{any nonzero element is invertible}. \]
As this condition is of nongeometric form, it is not inherited by arbitrary
local rings, which are indeed typically not fields. However, any intuitionistic
consequence of this condition which is of geometric form is inherited
by any local ring in any topos. Hence we may, when verifying a general fact
about local rings, suppose without loss of generality that the given ring is a
field. This observation is due to Kock~\cite{kock:univ-proj-geometry}, who exploited it to
develop projective geometry over local rings, and was further used by Reyes to
prove a Jacobian criterion for étale morphisms~\cite{reyes:cramer}.

A related nongeometric sequent is valid in the little Zariski topos of the
spectrum of a ring~$A$, the classifying topos of local localizations of~$A$.
If~$A$ is reduced, the generic model validates the dual condition that any
noninvertible element is zero. This property has been used to give a short and
even constructive proof of Grothendieck's generic freeness lemma,
substantially improving on previously published
proofs~\cite{blechschmidt:generic-freeness}.

In time, further nongeometric sequents holding in the big Zariski topos of an
arbitrary base scheme have been found~\cite[Section~18.4]{blechschmidt:phd}. These include:
\begin{itemize}
\item $\affl$ is \emph{anonymously algebraically closed} in the sense that
any monic polynomial~$p \? \affl[T]$ of degree at least one does \notnot have a
zero.
\item $\affl$ fulfills the following version of the Nullstellensatz:
Let $f_1,\ldots,f_m \in \affl[X_1,\ldots,X_n]$ be polynomials without a common
zero in~$(\affl)^n$. Then there are polynomials~$g_1,\ldots,g_m \in
\affl[X_1,\ldots,X_n]$ such that~$\sum_i g_i f_i = 1$.
\item Any function~$\affl \to \affl$ is given by a unique polynomial.
\item $\affl$ fulfills the axiom of microaffinity: Let~$\Delta = \{ \varepsilon \?
\affl \,|\, \varepsilon^2 = 0 \}$. Let~$f : \Delta \to \affl$ be an arbitrary
function. Then there are unique elements~$a, b \? \affl$ such
that~$f(\varepsilon) = a + b \varepsilon$ for all~$\varepsilon \? \Delta$.
\item $\affl$ is \emph{synthetically quasicoherent} in the sense that for any
finitely presented~$\affl$-algebra~$A$, the canonical homomorphism~$A \to
(\affl)^{\Spec(A)}$, where~$\Spec(A)$ is defined to be the set
of~$\affl$-algebra-homomorphisms~$A \to \affl$, is bijective.
\end{itemize}
All of these nongeometric sequents are useful for the purposes of synthetic
algebraic geometry, the desire to carry out algebraic geometry in a language
close to the simple language on the 19th and the beginning of the 20th century
while still being fully rigorous and fully general, working over arbitrary base
schemes instead of restricting to the field of complex numbers.
\medskip

\paragraph{Characterizing nongeometric sequents} Referring to one of the previous
examples, Tierney remarked around the time that those sequents were first
studied that ``[it] is surely important, though its precise significance is
still somewhat obscure---as is the case with many such nongeometric
formulas''~\cite[p.~209]{tierney:spectrum}. In view of their importance, is
there a way to discover nongeometric sequents in a systematic fashion? To
characterize the nongeometric sequents holding in classifying toposes? To this
end, Wraith put forward a specific conjecture~\cite[p.~336]{wraith:intuitionistic-algebra}:
\begin{quote}
\emph{The problem of characterising all the non-geometric properties of a generic
model appears to be difficult. If the generic model of a geometric theory~$\TT$
satisfies a sentence~$\alpha$ then any geometric consequence of $\TT + \alpha$ has to be a
consequence of~$\TT$. We might call~$\alpha$ $\TT$-redundant. Does the
generic~$\TT$-model satisfy all~$\TT$-redundant sentences?}
\end{quote}
This question was recently answered in the negative by Bezem, Buchholtz and
Coquand~\cite{bezem-buchholtz-coquand:syntactic-forcing-models}; hence the
characterization we propose is necessarily more nuanced.

Our starting point was the empirical
observation~\cite[p.~164]{blechschmidt:phd} that in the case of the big
Zariski topos, every true known nongeometric sequent followed from just a
single such, namely the synthetic quasicoherence of the generic
model, and in earlier work we surmised that one could formulate an appropriate
metatheorem explaining this observation and generalizing it to arbitrary
classifying toposes~\cite[Speculation~22.1]{blechschmidt:phd}. This hope turned
out to be true, in the sense we will now make precise.
\medskip

\paragraph{A general Nullstellensatz} To explain the relevant background, the
somewhat vague question ``to which extent does the classifying
topos~$\Set[\TT]$ realize that it is the classifying topos for~$\TT$?'' is a
useful guiding principle.

From the point of view of~$\Set[\TT]$, the model~$U_\TT$ is an ordinary
set-theoretic model. While in general properties of a specific set-theoretic
model do not follow solely from the axioms of the theory, it is on first sight
conceivable that this could be true for~$U_\TT$. That it, is is reasonable to
wonder whether
\[ \Set[\TT] \models \speak{a geometric sequent~$\sigma$ holds for~$U_\TT$ iff~$\ul{\TT}$ proves~$\sigma$} \]
holds.\footnote{Here~$\ul{\TT}$ is the internal geometric theory induced
by~$\TT$, obtained by pulling back the set of sorts, the set of function
symbols and so on along the geometric morphism~$\Set[\TT] \to \Set$. More
details will be given in Section~\ref{sect:review}. The corner quotes indicate
that for sake of readability, the translation into formal language is to be
carried out by the reader.} While the ``if'' direction
certainly holds (observing that the generic model~$U_\TT$ is, from the point of
view of~$\Set[\TT]$, a model of~$\ul{\TT}$), the ``only if'' direction
typically fails.

For instance, let~$\TT$ be the theory of rings and let~$\sigma$ be
the sequent~$(\top \vdash 1 + 1 = 0)$. Since there is an intuitionistic proof
that~$\TT$ does not prove~$\sigma$ and toposes are sound with respect to
intuitionistic logic, the statement~$\speak{\ul{\TT} proves~$\sigma$}$ is false
from the internal point of view of~$\Set[\TT]$. However, it is not the case
that the statement~$\speak{$1 + 1 = 0$ in~$U_\TT$}$ is false from the internal
point of view. In fact, this statement holds in a nontrivial slice
of~$\Set[\TT]$, the open subtopos coinciding with the classifying topos of
the theory of rings of characteristic two.

Intuitively, the problem is that while the meaning of~$\speak{$\ul{\TT} \vdash
\sigma$}$ is fixed, the meaning of~$\speak{$\sigma$ holds for~$U_\TT$}$ can vary
with the slice. This problem can be solved by passing from~$\ul{\TT}$ to a
varying theory, the internal theory~$\ul{\TT}/U_\TT$ defined below.
Unlike~$\ul{\TT}$, this theory is not the pullback of an external geometric
theory.

\begin{defn}Let~$\TT$ be a geometric theory. The theory~$\ul{\TT}/U_\TT$ is the
geometric theory internal to~$\Set[\TT]$ which arises from~$\ul{\TT}$
by adding additional constant symbols~$e_x$, one for each
element~$x : U_\TT$, axioms~$(\top \vdash f(e_{x_1},\ldots,e_{x_n}) =
e_{f(x_1,\ldots,x_n)})$ for each function symbol~$f$ and~$n$-tuple of elements
of~$U_\TT$, and axioms~$(\top \vdash R(e_{x_1},\ldots,e_{x_n}))$ for each
relation symbol~$R$ and~$n$-tuple~$(e_{x_1},\ldots,e_{x_n})$ such
that~$R(e_{x_1},\ldots,e_{x_n})$ holds for~$U_\TT$.
\end{defn}

From the point of view of~$\Set[\TT]$, a model of~$\ul{\TT}/U_\TT$ is a model
of~$\ul{\TT}$ equipped with a~$\ul{\TT}$-homomorphism from~$U_\TT$. In
particular, the identity~$(U_\TT \to U_\TT)$ is a model of~$\ul{\TT}/U_\TT$.

\begin{thm}Let~$\TT$ be a geometric theory. Then, internally to~$\Set[\TT]$:
\begin{quote}
A geometric sequent~$\sigma$ over the signature of~$\ul{\TT}/U_\TT$ holds
for~$U_\TT$ if and only if~$\ul{\TT}/U_\TT$ proves~$\sigma$.
\end{quote}
\end{thm}

\begin{thm}Let~$\TT$ be a geometric theory. Let~$\alpha$ be a first-order
formula over the signature of~$\TT$. Then the following statements are
equivalent.
\begin{enumerate}
\item The formula~$\alpha$ holds for~$U_\TT$.
\item The formula~$\alpha$ is provable in first-order intuitionistic logic
modulo the axioms of~$\TT$ and the additional axioms
\[ \sigma \quad\Longleftrightarrow\quad
  \ul{\TT}/U \vdash' \sigma \]
where~$\sigma$ ranges over all geometric sequents.
\end{enumerate}
\end{thm}

\section{Background on classifying toposes}

\section{Proofs of the main theorems}

\section{The special case of Horn theories}

\section{Applications}

%We do have, for any geometric sequent~$\sigma$, that
%\[ \Set[\TT] \models \speak{$\sigma$ holds for~$U_\TT$} \quad\text{iff}\quad
%  \Set[\TT] \models \speak{$\ul{\TT}$ proves~$\sigma$}, \]
%since~$\Set[\TT] \models \speak{$\sigma$ holds for~$U_\TT$}$ implies that~$\TT$
%proves~$\sigma$ and since~$\TT \vdash \sigma$ implies~$\E \models
%(\ul{\TT} \vdash \sigma)$ for any topos~$\E$ over~$\Set$. However, this
%equivalence is not stable under slicing ...

\printbibliography

\end{document}

* Discuss case of SDG: Why does the Kock--Lawvere there hold only for Weil
  algebras? Answer: It doesn't really! We do have that [U,U] is isomorphic to
  the free C^∞-ring on one generator. But this observation doesn't appear to be
  useful for further developments.

* Cite https://docplayer.net/55796150-Intuitionistic-algebra-some-recent-developments-in-topos-theory.html
  (in general and for internal theories)

### 2019-03-05 paper-qcoh.tex
More work on the introduction

@@
 \end{enumerate} One could argue that it is this model which mathematicians
-implicitly refer to when they utter ``Let~$M$ be a~$\TT$-model.''. It can
+implicitly refer to when they utter the phrase ``Let~$M$ be a~$\TT$-model.''. It can
 typically not be realized as a set-theoretic model, consisting of a set for
@@
 local ring, but even a field in the sense that
-\[ \text{any nonzero element is invertible}. \]
+any nonzero element is invertible.
 As this condition is of nongeometric form, it is not inherited by arbitrary
@@
 any monic polynomial~$p \? \affl[T]$ of degree at least one does \notnot have a
-zero.
-\item $\affl$ fulfills the following version of the Nullstellensatz:
+zero. \smallskip
+\item The Nullstellensatz holds:
 Let $f_1,\ldots,f_m \in \affl[X_1,\ldots,X_n]$ be polynomials without a common
 zero in~$(\affl)^n$. Then there are polynomials~$g_1,\ldots,g_m \in
-\affl[X_1,\ldots,X_n]$ such that~$\sum_i g_i f_i = 1$.
-\item Any function~$\affl \to \affl$ is given by a unique polynomial.
-\item $\affl$ fulfills the axiom of microaffinity: Let~$\Delta = \{ \varepsilon \?
+\affl[X_1,\ldots,X_n]$ such that~$\sum_i g_i f_i = 1$. \smallskip
+\item Any function~$\affl \to \affl$ is given by a unique polynomial. \smallskip
+\item $\affl$ is microaffine: Let~$\Delta = \{ \varepsilon \?
 \affl \,|\, \varepsilon^2 = 0 \}$. Let~$f : \Delta \to \affl$ be an arbitrary
 function. Then there are unique elements~$a, b \? \affl$ such
-that~$f(\varepsilon) = a + b \varepsilon$ for all~$\varepsilon \? \Delta$.
-\item $\affl$ is \emph{synthetically quasicoherent} in the sense that for any
+that~$f(\varepsilon) = a + b \varepsilon$ for all~$\varepsilon \? \Delta$. \smallskip
+\item $\affl$ is \emph{synthetically quasicoherent}: For any
 finitely presented~$\affl$-algebra~$A$, the canonical homomorphism~$A \to
-(\affl)^{\Spec(A)}$, where~$\Spec(A)$ is defined to be the set
+(\affl)^{\Spec(A)}$, where~$\Spec(A)$ is defined as the set
 of~$\affl$-algebra-homomorphisms~$A \to \affl$, is bijective.
@@
 topos~$\Set[\TT]$ realize that it is the classifying topos for~$\TT$?'' is a
-useful guiding principle.
+useful guiding principle. This is easiest to visualize with a concrete example
+for~$\TT$, such as the theory of rings.
 
-From the point of view of~$\Set[\TT]$, the model~$U_\TT$ is an ordinary
-set-theoretic model. While in general properties of a specific set-theoretic
-model do not follow solely from the axioms of the theory, it is on first sight
-conceivable that this could be true for~$U_\TT$. That it, is is reasonable to
-wonder whether
-\[ \Set[\TT] \models \speak{a geometric sequent~$\sigma$ holds for~$U_\TT$ iff~$\ul{\TT}$ proves~$\sigma$} \]
-holds.\footnote{Here~$\ul{\TT}$ is the internal geometric theory induced
+Let~$A$ be a ring. A simple version of the classical Nullstellensatz states:
+\emph{For any polynomials~$f$ and~$g$ over~$A$, if any zero of~$f$ is also a
+zero of~$g$, then there is a polynomial~$h$ such that~$g = hf$.} The
+polynomial~$h$ can be regarded as an ``algebraic certificate'' of the
+hypothesis. This principle holds for instance in the case that~$A$ is an
+algebraically closed field and that~$g$ is the unit polynomial. It is also
+true, without any restriction on~$g$, for the generic ring.
+
+We could try to generalize the Nullstellensatz to arbitrary geometric theories~$\TT$ as
+follows. \emph{For any geometric sequent~$\sigma$, if~$\sigma$ holds for a
+given~$\TT$-model~$M$ then~$\sigma$ is provable modulo~$\TT$.} In place of the
+algebraic certificate we now have a logical certificate, a \emph{proof}
+of~$\sigma$. However, this generalized statement is typically false, even for
+the generic model~$U_\TT$: The statement
+\begin{multline*}
+  \qquad\qquad\qquad\Set[\TT] \models \ulcorner\text{for any geometric sequent~$\sigma$,} \\
+    \text{if~$\sigma$ holds for~$U_\TT$ then~$\ul{\TT}$
+    proves~$\sigma$}\urcorner\qquad\qquad\qquad
+\end{multline*}
+does not hold.\footnote{Here~$\ul{\TT}$ is the internal geometric theory induced
 by~$\TT$, obtained by pulling back the set of sorts, the set of function
-symbols and so on along the geometric morphism~$\Set[\TT] \to \Set$. More
-details will be given in Section~\ref{sect:review}. The corner quotes indicate
+symbols and so on along the geometric morphism~$\Set[\TT] \to \Set$. For
+instance, if~$\TT$ is the theory of rings, then from the internal point of view
+of~$\Set[\TT]$ the theory~$\ul{\TT}$ will again be the theory of rings.
+More details will be given in Section~\ref{sect:review}. The corner quotes indicate
 that for sake of readability, the translation into formal language is to be
-carried out by the reader.} While the ``if'' direction
-certainly holds (observing that the generic model~$U_\TT$ is, from the point of
-view of~$\Set[\TT]$, a model of~$\ul{\TT}$), the ``only if'' direction
-typically fails.
+carried out by the reader.}
 
@@
 
-Intuitively, the problem is that while the meaning of~$\speak{$\ul{\TT} \vdash
-\sigma$}$ is fixed, the meaning of~$\speak{$\sigma$ holds for~$U_\TT$}$ can vary
+Intuitively, the problem is that while the meaning of~$\speak{$\ul{\TT}$
+proves~$\sigma$}$ is fixed, the meaning of~$\speak{$\sigma$ holds for~$U_\TT$}$ can vary
 with the slice. This problem can be solved by passing from~$\ul{\TT}$ to a
-varying theory, the internal theory~$\ul{\TT}/U_\TT$ defined below.
+varying theory, the internal theory~$\ul{\TT}/U_\TT$ defined in
+Section~\ref{sect:main}. If~$\TT$ is the theory of rings, then~$\ul{\TT}/U_\TT$
+is the~$\Set[\TT]$-theory of~$U_\TT$-algebras.
 Unlike~$\ul{\TT}$, this theory is not the pullback of an external geometric
-theory.
+theory. We then have, subject to some qualifications made precise in
+Section~\ref{sect:main}:
 
-\begin{defn}Let~$\TT$ be a geometric theory. The theory~$\ul{\TT}/U_\TT$ is the
-geometric theory internal to~$\Set[\TT]$ which arises from~$\ul{\TT}$
-by adding additional constant symbols~$e_x$, one for each
-element~$x : U_\TT$, axioms~$(\top \vdash f(e_{x_1},\ldots,e_{x_n}) =
-e_{f(x_1,\ldots,x_n)})$ for each function symbol~$f$ and~$n$-tuple of elements
-of~$U_\TT$, and axioms~$(\top \vdash R(e_{x_1},\ldots,e_{x_n}))$ for each
-relation symbol~$R$ and~$n$-tuple~$(e_{x_1},\ldots,e_{x_n})$ such
-that~$R(e_{x_1},\ldots,e_{x_n})$ holds for~$U_\TT$.
-\end{defn}
-
-From the point of view of~$\Set[\TT]$, a model of~$\ul{\TT}/U_\TT$ is a model
-of~$\ul{\TT}$ equipped with a~$\ul{\TT}$-homomorphism from~$U_\TT$. In
-particular, the identity~$(U_\TT \to U_\TT)$ is a model of~$\ul{\TT}/U_\TT$.
-
-\begin{thm}Let~$\TT$ be a geometric theory. Then, internally to~$\Set[\TT]$:
-\begin{quote}
-A geometric sequent~$\sigma$ over the signature of~$\ul{\TT}/U_\TT$ holds
-for~$U_\TT$ if and only if~$\ul{\TT}/U_\TT$ proves~$\sigma$.
-\end{quote}
+\begin{thm}\label{thm:nullstellensatz0}
+Let~$\TT$ be a geometric theory. Then, internally to~$\Set[\TT]$:
+\begin{equation}\tag{$\ddagger$}\label{eq:nullstellensatz}
+\text{A geometric$^*$ sequent~$\sigma$ holds for~$U_\TT$ if and only
+if~$\ul{\TT}/U_\TT$ proves$^*$~$\sigma$.}
+\end{equation}
 \end{thm}
 
-\begin{thm}Let~$\TT$ be a geometric theory. Let~$\alpha$ be a first-order
+To illustrate Theorem~\ref{thm:nullstellensatz0}, let~$\TT$ be the theory of
+rings and let~$\sigma$ be of the form~$(f(x) = 0 \seq{x} g(x) = 0)$ for some
+polynomials~$f$ and~$g$. To say that~$\sigma$ holds for~$U_\TT$ amounts to
+saying that any zero~$x \? U_\TT$ of~$f$ is also a zero of~$g$, and to say
+that~$\ul{\TT}/U_\TT$ proves~$\sigma$ amounts to saying that
+in~$U_\TT[X]/(f(X))$, the free~$U_\TT$-algebra on one generator~$X$ subject to
+the relation~$f(X) = 0$, the relation~$g([X]) = 0$ holds. Hence we obtain
+\[ \Set[\TT] \models
+  \forall f,g \? U_\TT[X]\_ \bigl(
+    (\forall x \? U_\TT\_ f(x) = 0 \Rightarrow g(x) = 0) \Longleftrightarrow
+      \exists h \in U_\TT[X]\_ g = hf\bigr). \]
+
+Theorem~\ref{thm:nullstellensatz0} is a source of nongeometric sequents. Indeed,
+it is the universal such source in the sense that any first-order formula
+holding for~$U_\TT$ can be deduced from~\eqref{eq:nullstellensatz}:
+
+\begin{thm}\label{thm:characterization0}
+Let~$\TT$ be a geometric theory. Let~$\alpha$ be a first-order
 formula over the signature of~$\TT$. Then the following statements are
@@
 \item The formula~$\alpha$ is provable in first-order intuitionistic logic
-modulo the axioms of~$\TT$ and the additional axioms
-\[ \sigma \quad\Longleftrightarrow\quad
-  \ul{\TT}/U \vdash' \sigma \]
-where~$\sigma$ ranges over all geometric sequents.
+modulo the axioms of~$\TT$ and the additional axiom
+scheme~\eqref{eq:nullstellensatz}.
 \end{enumerate}
 \end{thm}
+
+\paragraph{Acknowledgments} XXX
 
@@
   (in general and for internal theories)
+
+\begin{defn}Let~$\TT$ be a geometric theory. The theory~$\ul{\TT}/U_\TT$ is the
+geometric theory internal to~$\Set[\TT]$ which arises from~$\ul{\TT}$
+by adding additional constant symbols~$e_x$, one for each
+element~$x : U_\TT$, axioms~$(\top \vdash f(e_{x_1},\ldots,e_{x_n}) =
+e_{f(x_1,\ldots,x_n)})$ for each function symbol~$f$ and~$n$-tuple of elements
+of~$U_\TT$, and axioms~$(\top \vdash R(e_{x_1},\ldots,e_{x_n}))$ for each
+relation symbol~$R$ and~$n$-tuple~$(e_{x_1},\ldots,e_{x_n})$ such
+that~$R(e_{x_1},\ldots,e_{x_n})$ holds for~$U_\TT$.
+\end{defn}
+
+From the point of view of~$\Set[\TT]$, a model of~$\ul{\TT}/U_\TT$ is a model
+of~$\ul{\TT}$ equipped with a~$\ul{\TT}$-homomorphism from~$U_\TT$. In
+particular, the identity~$(U_\TT \to U_\TT)$ is a model of~$\ul{\TT}/U_\TT$.
+We then have:
+
+% We would not expect the statement
+% \[ \text{``for any geometric sequent~$\sigma$, if~$\sigma$ holds for~$A$
+% then the theory of rings proves~$\sigma$''} \]
+% to hold for an ordinary ring~$A$ in the topos of sets.
+%
+% Let~$M$ be a~$\TT$-model. We then trivially have
+% signature of~$\TT$. While we always have the implication
+% \[ \text{$\TT$ proves~$\sigma$} \quad\Longrightarrow\quad
+%   \text{$\sigma$ holds for~$M$}, \]
+% the converse is of course typically false,
+%
+% Furthermore, in a first introduction, it is useful to
+% restrict to an instructive special case, such as the theory of (commutative) rings.
+%
+% A simple version of the classical Nullstellensatz, valid for algebraically
+% closed fields, states: Let~$f \in A[X]$ be a polynomial. If
+% \[ \forall x \in A\_ (f(x) = 0 \in A \Rightarrow 1 = 0 \in A), \]
+% then there is a polynomial~$g \in A[X]$ such that~$gf = 1$. This polynomial~$g$
+% can be regarded as an ``algebraic certificate'' that the equation~$f(x) = 0$
+% only has a solution in the trivial case~$1 = 0$. The conclusion can also be phrased as
+% follows: In the free~$A$-algebra on one generator modulo the relation~$f(X) =
+% 0$, the relation~$1 = 0$ holds.
+%
+% Let~$A$ be a specific ring. Let~$f,g \in \ZZ[X]$ be
+% polynomials. While we always have the implication
+% \[ g([X]) = 0 \in \ZZ[X]/(f) \quad\Longrightarrow\quad
+%   \forall x \in A\_ (f(x) = 0 \Rightarrow g(x) = 0), \]
+% the converse is typically false: Just that~$g(x) = 0$ holds for every zero
+% of~$f$ in~$A$ does not imply that~$g([X]) = 0$ holds for the generic zero
+% of~$f$ in the free ring on one generator~$X$ modulo the relation~$f(X) = 0$.
+%
+%
+% the case that~$\TT$ is a Horn theory (where the axioms are of the
+% form~$(\phi \seq{\vec x} \psi)$ with the formulas~$\phi$ and~$\psi$ being
+% finite conjunctions of atomic propositions (equality and the given relations)),
+% such that we can switch from a logical terminology to a more algebraic one.
+% In the Horn case, for any set~$X$ and any set~$R$ of atomic propositions the
+% term algebra construction yields~$\TT\langle X | R\rangle$, the
+% free~$\TT$-model generated by~$X$ modulo the relations~$R$. We will also abuse
+% notation to pretend that~$\TT$ is single-sorted.
+%
+% Let~$M$ be a specific~$\TT$-model and assume that for
+% all~$n$-tuples~$x = (x_1,\ldots,x_n)$ of elements of~$M$, if some set~$R_1$ of
+% atomic propositions is satisfied for~$x$ then so is some set~$R_2$. In this
+% case it is unreasonable to expect that~$R_2$ holds for the~$n$-tuple of
+% generators of~$\TT\langle e_1,\ldots,e_n | R_1\rangle$.
+%
+% let~$R_1$ and~$R_2$ be sets of atomic
+% propositions with~$n$ free variables. Then
+% and assume, for some sets of relations~$R_1$
+% and~$R_2$
+%
+% The term algebra of a Horn theory~$\TT$ is always a model of~$\TT$, such that

### 2019-03-05 paper-qcoh.tex
Some statements and proofs

@@
 \DeclareMathOperator{\Hom}{Hom}
-\DeclareMathOperator{\Mod}{Mod}
 \DeclareMathOperator{\Sh}{Sh}
@@
 \newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
+\newcommand{\brak}[1]{\llbracket #1 \rrbracket}
+\newcommand{\Mod}[1]{{#1}\mathrm{\text{-}mod}}
 
@@
 
+In case that~$\TT$ is the quotient of a Horn theory XXX
+
 Theorem~\ref{thm:nullstellensatz0} is a source of nongeometric sequents. Indeed,
@@
 
+The following lemma shows that for coherent theories, there is no difference
+between provability and provability$^*$.
+
+\begin{lemma}Let~$\E$ be a Grothendieck topos. Let~$\TT$ be a coherent theory
+internal to~$\E$. Let~$\sigma$ be a coherent sequent over the signature
+of~$\TT$. Then the following statements are equivalent.
+\begin{enumerate}
+\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of externally
+finite shape}.$
+\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of arbitrary
+external shape}.$
+\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of arbitrary
+internal shape}.$
+\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of internally
+finite shape}.$
+\end{enumerate}
+\end{lemma}
+
+\begin{proof}It is obvious that~(1) implies~(2) implies~(3).
+
+To verify that~(3) implies~(4), we can mimick the usual proof of this fact in
+the internal language of~$\E$: One can construct a variant of the syntactic
+site of~$\TT$ which is built using only coherent sequents and finitary
+derivability~\cite[Section~1.4]{caramello:tst}. The sheaf topos over this site
+is another model for the classifying topos of~$\TT$. If~$\sigma$
+is~$\TT$-derivable by a proof tree of arbitrary shape, then~$\sigma$ holds in
+this model of the classifying topos. By the analogue of
+Proposition~\ref{prop:classifying-topos} for this model, we obtain
+that~$\sigma$ is~$\TT$-derivable by a proof tree of finite shape.
+
+That~(4) implies~(1) is a routine exercise exploiting that
+\[ \E \models \forall X\_ \speak{$X$ is Kuratowski-finite} \Leftrightarrow
+  \bigvee_{n \geq 0} \exists x_1,\ldots,x_n \? X\_
+  \forall x \? X\_ \bigvee_{i=1}^n x = x_i. \]
+\end{proof}
+
 \section{The special case of Horn theories}
+
+Throughout this section, let~$\TT$ be a Horn theory.
+
+\begin{lemma}\label{lemma:free-models}
+Let~$X$ be a set equipped with a morphism~$X \to S$ to the set of sorts
+of the signature~$\Sigma$ of~$\TT$. Let~$R$ be a set of atomic propositions in which the
+elements of~$X$ may appear as new constants of the respective sorts. Then there
+is~$\TT\langle X | R \rangle$, the free~$\TT$-model on the generators~$X$ modulo
+the relations~$R$.\end{lemma}
+
+\begin{proof}The desired model can be constructed as a term algebra. As a set,
+it consists of the terms (in the empty context) of the signature~$\Sigma + X$
+modulo the equivalence relation identifying two terms if and only if~$\TT + R$
+proves them to be equal. The function symbols~$f$ of~$\Sigma$ are interpreted
+by declaring~$\brak{f}([t_1],\ldots,[t_n]) = [f(t_1,\ldots,t_n)]$ and the
+relation symbols~$R$ are interpreted by declaring~$([t_1],\ldots,[t_n]) \in
+\brak{S} \Leftrightarrow (\TT + R \vdash S(t_1,\ldots,t_n))$.
+
+We omit the required verifications and only remark that while the same
+construction can be carried out if~$\TT$ is a general geometric theory, the
+resulting object would in general not be a model of~$\TT$.
+\end{proof}
+
+\begin{lemma}The category of~$\TT$-models is complete and
+cocomplete.\end{lemma}
+
+\begin{proof}Limits are computed as the limits of the underlying sets, colimits
+are computed by using the construction of Lemma~\ref{lemma:free-models}. For
+instance, the coproduct of~$\TT\langle X | R \rangle$ and~$\TT\langle X' | R'
+\rangle$ is~$\TT\langle X \amalg X' \,|\, R, R' \rangle$.\end{proof}
+
+Having the special case of the theory of rings in mind, we write the coproduct
+in the category of~$\TT$-models as~``$\otimes$''.
+
+\begin{lemma}Let~$\sigma = (\varphi_1 \wedge \cdots \wedge \varphi_n
+\seq{x_1,\ldots,x_k} \psi_1 \wedge \cdots \wedge \psi_m)$ be a Horn sequent
+over the signature of~$\TT$. Then the following statements are equivalent.
+\begin{enumerate}
+\item The theory~$\TT$ proves~$\sigma$.
+\item In~$\TT\langle x_1,\ldots,x_k | \varphi_1,\ldots,\varphi_n \rangle$, the
+propositions~$\psi_1,\ldots,\psi_m$ hold for
+the~$k$-tuple~$([x_1],\ldots,[x_k])$.
+\end{enumerate}
+\end{lemma}
+
+\begin{proof}By construction of the term algebra.\end{proof}
+
+\begin{lemma}A~$\TT$-model is a finitely presented object of the category
+of~$\TT$-models if and only if it is isomorphic to a model of the
+form~$\TT\langle X | R \rangle$ where~$X$ is Bishop-finite and~$R$ is
+Kuratowski-finite.
+\end{lemma}
+
+\begin{proof}It is an instructive exercise to verify that models of the stated
+form are compact. Conversely, let a~$\TT$-model~$M$ be given. Then~$\TT$ is the
+filtered colimit of all models over~$M$ which are of the stated form. If~$M$ is
+compact, the identity on~$M$ factors over such a model. Hence~$M$ is a
+retract of such a model and hence itself isomorphic to a model of this form.
+\end{proof}
+
+% Set[T] einführen
+
+Any~$\TT$-model~$A$ has a mirror image in the topos~$\Set[\TT]$, namely the
+functor~$A^\sim : \Mod{\TT}_\mathrm{fp} \to \Set$ given by~$T \mapsto A \otimes T$.
+This object is in a canonical way a~$\TT$-model over~$U_\TT$, hence from the
+point of view of~$\Set[\TT]$ a~$\ul{\TT}/U_\TT$-model.
+
+\begin{lemma}The functor~$(\cdot)^\sim$ from~$\TT$-models to~$\ul{\TT}/U_\TT$-models
+in~$\Set[\TT]$ is left adjoint to functor~$\Gamma = \Hom(1, \cdot)$ computing
+global elements.
+\end{lemma}
+
+\begin{proof}An~$U_\TT$-algebra-homomorphism~$\alpha : A^\sim \to M$ yields
+the~$\TT$-model-homo\-mor\-phism~$\alpha_0 : A \to M(0) = \Gamma(M)$, where~$0$ is the
+initial~$\TT$-model. Conversely, a~$\TT$-model-homomorphism~$\beta : A \to
+\Gamma(M)$ yields an~$U_\TT$-algebra-homomorphism by summing~$A \to M(0) \to
+M(T)$ with the structure morphism~$T = U_\TT(T) \to
+M(T)$.\end{proof}
+
+\begin{defn}The \emph{spectrum}~$\Spec(M)$ of a~$U_\TT$-algebra~$M$ in~$\Set[\TT]$
+is the result of constructing, internally to~$\Set[\TT]$, the set
+of~$U_\TT$-algebra-homomorphisms~$M \to U_\TT$.
+\end{defn}
+
+Externally, the spectrum of~$M$ is the functor
+mapping a finitely presented~$\TT$-model~$T$ to~$\Hom_T(M(T), T)$, the set
+of~$\TT$-homomorphisms~$M(T) \to T$ compatible with the structure morphisms~$T
+\to M(T)$ and~$T \to T$.
 

### 2019-03-06 paper-qcoh.tex
More statements and proofs

@@
 
-In case that~$\TT$ is the quotient of a Horn theory XXX
+The statement~\eqref{eq:nullstellensatz} is not a geometric sequent. Therefore
+it is not to be expected that it passes from~$\Set[\TT]$ to a
+subtopos~$\Set[\TT']$ corresponding to a quotient theory~$\TT'$ of~$\TT$, and
+indeed in general it does not. We formulate a substitute as Theorem~XXX,
+substantially broading the scope of the Nullstellensatz.
 
@@
 it is the universal such source in the sense that any first-order formula
-holding for~$U_\TT$ can be deduced from~\eqref{eq:nullstellensatz}:
+which holds for~$U_\TT$ can be deduced from~\eqref{eq:nullstellensatz}:
 
@@
   \bigvee_{n \geq 0} \exists x_1,\ldots,x_n \? X\_
-  \forall x \? X\_ \bigvee_{i=1}^n x = x_i. \]
+  \forall x \? X\_ \bigvee_{i=1}^n x = x_i. \qedhere \]
 \end{proof}
+
+\begin{thm}\label{thm:nullstellensatz-horn}
+Let~$\TT$ be a geometric theory. Let~$\TT'$ be a quotient theory
+of~$\TT$. Assume that the generic model~$U_\TT$ is a sheaf for the topology
+on~$\Set[\TT]$ cutting out the subtopos~$\Set[\TT']$. Then the following
+statement holds internally to~$\Set[\TT']$:
+\[ \text{A geometric$^\star$ sequent~$\sigma$ with Horn consequent holds for~$U_{\TT'}$
+  iff~$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$.} \]
+\end{thm}
+
+\begin{proof}In general, the generic model of~$\TT'$ is the pullback of the
+generic model of~$\TT$ to the
+subtopos~$\Set[\TT']$~\cite[Lemma~2.3]{caramello:definability}. By the sheaf
+assumption, the objects~$U_{\TT'}$ and~$U_\TT$ actually agree, that is~$U_\TT$
+is contained in the subtopos and has the universal property of~$U_{\TT'}$.
+
+The ``if'' direction is trivial, as~$U_{\TT'}$ is a~$\ul{\TT}/U_\TT$-model.
+
+For the ``only if'' direction, we use that a statement holds in~$\Set[\TT']$ if
+and only if its~$\nabla$-translation holds in~$\Set[\TT]$, where~$\nabla$ is
+the modal operator associated to the topology cutting
+out~$\Set[\TT']$~\cite[Theorem~6.31]{blechschmidt:phd}. Exploiting some of the
+simplification rules of the~$\nabla$-translation~\cite[Section~6.6]{blechschmidt:phd},
+it hence suffices to verify, internally to~$\Set[\TT]$, that:
+\begin{multline*}
+  \text{For any geometric$^\star$ sequent~$\sigma = (\varphi \seq{\vec x} \psi)$ where~$\psi$ is a Horn formula,} \\
+    \text{if~$\forall x_1,\ldots,x_n\?U_\TT'\_ (\varphi \Rightarrow \nabla
+    \psi)$, then~$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$.}
+\end{multline*}
+Since~$\nabla$ commutes with finite conjunctions and since the sheaf assumption
+implies that $\nabla(s = t)$ is equivalent
+to~$s = t$ and that, for relation symbols~$R$, the
+statement~$\nabla(R(s_1,\ldots,s_m))$ is equivalent to~$R(s_1,\ldots,s_m)$,
+the statement~$\nabla\psi$ is equivalent to~$\psi$. Hence the claim follows
+from Theorem~\ref{thm:nullstellensatz}.
+\end{proof}
+
+A situation in which the sheaf assumption of Theorem~\ref{thm:nullstellensatz-horn}
+is satisfied is when~$\TT$ is a Horn theory and the topology cutting
+out~$\Set[\TT']$ is subcanonical. For instance, this is the case if~$\TT$ is
+the theory of rings and~$\Set[\TT']$ is any of the well-known toposes in
+algebraic geometry such as the big Zariski topos, the big étale topos or the
+big fppf topos.
 
@@
 
+\begin{lemma}Let~$A$ be a~$\TT$-model. Then~$\Spec(A^\sim)$ coincides
+with~$yA$, where~$y : \Mod{\TT} \to \Set[\TT]$ is the Yoneda embedding.
+\end{lemma}
+
+\begin{proof}XXX\end{proof}
+
+\begin{lemma}Let~$A$ and~$B$ be finitely presented~$\TT$-models. Then the
+canonical morphism
+\[ \Hom_{U_\TT}(A^\sim, B^\sim) \longrightarrow \Spec(A^\sim)^{\Spec(B^\sim)}
+\]
+is an isomorphism.
+\end{lemma}
+
 \section{Applications}
@@
 % The term algebra of a Horn theory~$\TT$ is always a model of~$\TT$, such that
+
+* Sei T' subkanonisch für T.
+
+* Sei φ eine Hornformel. Gelte ∇φ. Folgt dann φ?
+
+* Für φ = (x = y) stimmt das, da U separiert ist.
+
+* Für φ = (R(x,y)) stimmt, das ...?
+
+  Was heißt das denn? Habe R \--> U x U. Gehe davon aus, dass ∇φ bedeutet, dass
+  ∇((x,y) ∈ R). Daraus folgt aber (x,y) ∈ R:
+
+  K = { (x,y) | (x,y) ∈ R } ist Subsingleton von U x U und ∇-bewohnt.
+  Also existiert (a,b) ∈ U x U mit ∇... Hilft nicht.
+
+  Aber: K = { (a,b) ∈ R | a = x & b = y } ist Subsingleton von R.
+  Und ∇-bewohnt. Also existiert (a,b) ∈ R mit ∇(a=x & b = y).
+  Wegen Separiertheit folgt a = x & b = y. Fertig. :-)

### 2019-03-06 paper-qcoh.tex
typo
Some clarification
to do

@@
 
+Summarizing, the situation is as follows.
+\begin{itemize}
+\item The generic model~$U_\TT$ is a
+conservative~$\TT$-model. \smallskip
+\item The topos~$\Set[\TT]$ does not believe that~$U_\TT$ is a
+conservative~$\ul{\TT}$-model. \smallskip
+\item The topos~$\Set[\TT]$ does believe that~$U_\TT$
+is a conservative$^\star$~$\ul{\TT}/U_\TT$-model.
+\end{itemize}
+
 Theorem~\ref{thm:nullstellensatz0} is a source of nongeometric sequents. Indeed,
@@
   \text{For any geometric$^\star$ sequent~$\sigma = (\varphi \seq{\vec x} \psi)$ where~$\psi$ is a Horn formula,} \\
-    \text{if~$\forall x_1,\ldots,x_n\?U_\TT'\_ (\varphi \Rightarrow \nabla
+    \text{if~$\forall x_1,\ldots,x_n\?U_\TT\_ (\varphi \Rightarrow \nabla
     \psi)$, then~$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$.}
@@
   Wegen Separiertheit folgt a = x & b = y. Fertig. :-)
+
+TODO:
+
+* Nichttrivialität des Nullstellensatzes diskutieren (Spezialfall top |- φ
+  betrachten, dieser ist nämlich schon mehr oder weniger trivial)
+
+* Anmerken, dass der Nullstellensatz für first-order sequents NICHT stimmt
+  (Beispiel: ¬(x=0) ⇒ (x inv.))

### 2019-03-07 paper-composition.tex
Stub for joint paper on composition of internal theories

--- 2019-03-05 paper-qcoh.tex
@@
 \usepackage[babel]{csquotes}
-\bibliography{paper-qcoh.bib}
+\bibliography{paper-composition.bib}
 
-\title{A general Nullstellensatz for generalized spaces}
+\title{Composition of internal theories}
+
 \author{Ingo Blechschmidt}
 \email{ingo.blechschmidt@univr.it}
+
+\author{Matthias Hutzler}
+\email{matthias.ralph.hutzler@student.uni-augsburg.de}
+
+\author{Alexander Oldenziel}
+\email{a.f.d.a.gietelinkoldenziel@gmail.com}
 
@@
 \newcommand{\B}{\mathcal{B}}
+\newcommand{\C}{\mathcal{C}}
 \renewcommand{\AA}{\mathbb{A}}
@@
 \begin{abstract}
-  We give a general Nullstellensatz for the generic model of a
-  geometric theory, useful as a source of nongeometric sequents validated by
-  the generic model, and characterize the first-order formulas validated by the
-  genetric model. XXX
+  XXX
 \end{abstract}
@@
 
-\paragraph{Generic models} Let~$\TT$ be a geometric theory, such as the theory
-of rings, local rings or intervals. We follow Caramello's
-terminology~\cite{caramello:tst} to mean by \emph{geometric theory} a system
-given by a set of sorts, a set of finitary function symbols, a set of finitary
-relation symbols and a set of axioms, consisting of geometric sequents
-(sequents of the form~$\varphi \seq{\vec x} \psi$ where~$\varphi$ and~$\psi$
-are geometric formulas, that is formulas built from equality and the relation
-symbols by the logical connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$
-and by arbitrary set-indexed disjunctions~$\bigvee$).
+XXX state main lemma (to be proved in Section 3)
 
-A fundamental result is that there is a \emph{generic model}~$U_\TT$ of~$\TT$,
-a model such that for any geometric sequent~$\sigma$, the following notions
-coincide: \begin{enumerate} \item The sequent~$\sigma$ is provable
-modulo~$\TT$. \item The sequent~$\sigma$ holds for any~$\TT$-model in any
-Grothendieck topos. \item The sequent~$\sigma$ holds for~$U_\TT$.
-\end{enumerate} One could argue that it is this model which mathematicians
-implicitly refer to when they utter the phrase ``Let~$M$ be a~$\TT$-model.''. It can
-typically not be realized as a set-theoretic model, consisting of a set for
-each sort, a function for each function symbol and so on; instead it is a model
-in a custom-tailored syntactically constructed Grothendieck topos, the
-\emph{classifying topos}~$\Set[\TT]$ of~$\TT$, hence consists of an object
-of~$\Set[\TT]$ for each sort, a morphism for each function symbol and so on.
+\begin{defn}A geometric morphism~$\F \to \E$ is \emph{of presheaf type} if and
+only if it is equivalent (XXX be more precise) to the canonical
+functor~$\E^{\C^\op} \to \E$ for an internal category~$\C$ in~$\E$.\end{defn}
 
-To state what it means for a~$\TT$-structure in a topos~$\E$ to verify the
-axioms of~$\TT$, rendering it a model, the \emph{internal language} of~$\E$ is
-used, roughly reviewed in Section~\ref{sect:review} below. We write~``$\E
-\models \alpha$'' to mean that the statement~$\alpha$ holds from the internal
-point of view of~$\E$. Since this language is a form of a higher-order
-dependently-typed intuitionistic logic, the classifying topos~$\Set[\TT]$ can
-be regarded as a higher-order completion of the geometric theory~$\TT$. The
-generic model enjoys the universal property that any~$\TT$-model in any
-(Grothendieck) topos~$\E$ is the pullback of~$U_\TT$ along an essentially
-unique geometric morphism~$\E \to \Set[\TT]$.
-\medskip
+XXX This notion is the relativization of the notion of a presheaf category; a
+geometric theory~$\TT$ is of presheaf type if and only if the canonical
+geometric morphism~$\Set[\TT] \to \Set$ is of presheaf type.
 
-\paragraph{Nongeometric sequents} Crucially, the
-equivalence~$(1)\Leftrightarrow(2)\Leftrightarrow(3)$ relating
-provability and truth in~$\Set[\TT]$ only pertains to geometric sequents. The
-generic model may validate additional nongeometric sequents which are not
-provable from the axioms of~$\TT$ in first-order or even higher-order logic,
-and these nongeometric sequents may be quite surprising and have useful
-consequences.
+\begin{prop}The composition of geometric morphisms of presheaf type is of
+presheaf type.\end{prop}
 
-One of the most celebrated such sequents arises in the case that~$\TT$ is the
-theory of local rings. In this case, the classifying topos~$\Set[\TT]$ is also
-known as the \emph{big Zariski topos} of~$\Spec(\ZZ)$ from algebraic geometry,
-the topos of sheaves over the site of schemes locally of finite presentation,
-and the generic model is the functor~$\affl$ of points of the affine line, the
-functor which maps any (l.o.f.p.) scheme~$X$ to~$\Hom(X,\mathbb{A}^1) =
-\O_X(X)$.
-
-From the point of view of the topos, the ring object~$\affl$ is not only a
-local ring, but even a field in the sense that
-any nonzero element is invertible.
-As this condition is of nongeometric form, it is not inherited by arbitrary
-local rings, which are indeed typically not fields. However, any intuitionistic
-consequence of this condition which is of geometric form is inherited
-by any local ring in any topos. Hence we may, when verifying a general fact
-about local rings, suppose without loss of generality that the given ring is a
-field. This observation is due to Kock~\cite{kock:univ-proj-geometry}, who exploited it to
-develop projective geometry over local rings, and was further used by Reyes to
-prove a Jacobian criterion for étale morphisms~\cite{reyes:cramer}.
-
-A related nongeometric sequent is valid in the little Zariski topos of the
-spectrum of a ring~$A$, the classifying topos of local localizations of~$A$.
-If~$A$ is reduced, the generic model validates the dual condition that any
-noninvertible element is zero. This property has been used to give a short and
-even constructive proof of Grothendieck's generic freeness lemma,
-substantially improving on previously published
-proofs~\cite{blechschmidt:generic-freeness}.
-
-In time, further nongeometric sequents holding in the big Zariski topos of an
-arbitrary base scheme have been found~\cite[Section~18.4]{blechschmidt:phd}. These include:
-\begin{itemize}
-\item $\affl$ is \emph{anonymously algebraically closed} in the sense that
-any monic polynomial~$p \? \affl[T]$ of degree at least one does \notnot have a
-zero. \smallskip
-\item The Nullstellensatz holds:
-Let $f_1,\ldots,f_m \in \affl[X_1,\ldots,X_n]$ be polynomials without a common
-zero in~$(\affl)^n$. Then there are polynomials~$g_1,\ldots,g_m \in
-\affl[X_1,\ldots,X_n]$ such that~$\sum_i g_i f_i = 1$. \smallskip
-\item Any function~$\affl \to \affl$ is given by a unique polynomial. \smallskip
-\item $\affl$ is microaffine: Let~$\Delta = \{ \varepsilon \?
-\affl \,|\, \varepsilon^2 = 0 \}$. Let~$f : \Delta \to \affl$ be an arbitrary
-function. Then there are unique elements~$a, b \? \affl$ such
-that~$f(\varepsilon) = a + b \varepsilon$ for all~$\varepsilon \? \Delta$. \smallskip
-\item $\affl$ is \emph{synthetically quasicoherent}: For any
-finitely presented~$\affl$-algebra~$A$, the canonical homomorphism~$A \to
-(\affl)^{\Spec(A)}$, where~$\Spec(A)$ is defined as the set
-of~$\affl$-algebra-homomorphisms~$A \to \affl$, is bijective.
-\end{itemize}
-All of these nongeometric sequents are useful for the purposes of synthetic
-algebraic geometry, the desire to carry out algebraic geometry in a language
-close to the simple language on the 19th and the beginning of the 20th century
-while still being fully rigorous and fully general, working over arbitrary base
-schemes instead of restricting to the field of complex numbers.
-\medskip
-
-\paragraph{Characterizing nongeometric sequents} Referring to one of the previous
-examples, Tierney remarked around the time that those sequents were first
-studied that ``[it] is surely important, though its precise significance is
-still somewhat obscure---as is the case with many such nongeometric
-formulas''~\cite[p.~209]{tierney:spectrum}. In view of their importance, is
-there a way to discover nongeometric sequents in a systematic fashion? To
-characterize the nongeometric sequents holding in classifying toposes? To this
-end, Wraith put forward a specific conjecture~\cite[p.~336]{wraith:intuitionistic-algebra}:
-\begin{quote}
-\emph{The problem of characterising all the non-geometric properties of a generic
-model appears to be difficult. If the generic model of a geometric theory~$\TT$
-satisfies a sentence~$\alpha$ then any geometric consequence of $\TT + \alpha$ has to be a
-consequence of~$\TT$. We might call~$\alpha$ $\TT$-redundant. Does the
-generic~$\TT$-model satisfy all~$\TT$-redundant sentences?}
-\end{quote}
-This question was recently answered in the negative by Bezem, Buchholtz and
-Coquand~\cite{bezem-buchholtz-coquand:syntactic-forcing-models}; hence the
-characterization we propose is necessarily more nuanced.
-
-Our starting point was the empirical
-observation~\cite[p.~164]{blechschmidt:phd} that in the case of the big
-Zariski topos, every true known nongeometric sequent followed from just a
-single such, namely the synthetic quasicoherence of the generic
-model, and in earlier work we surmised that one could formulate an appropriate
-metatheorem explaining this observation and generalizing it to arbitrary
-classifying toposes~\cite[Speculation~22.1]{blechschmidt:phd}. This hope turned
-out to be true, in the sense we will now make precise.
-\medskip
-
-\paragraph{A general Nullstellensatz} To explain the relevant background, the
-somewhat vague question ``to which extent does the classifying
-topos~$\Set[\TT]$ realize that it is the classifying topos for~$\TT$?'' is a
-useful guiding principle. This is easiest to visualize with a concrete example
-for~$\TT$, such as the theory of rings.
-
-Let~$A$ be a ring. A simple version of the classical Nullstellensatz states:
-\emph{For any polynomials~$f$ and~$g$ over~$A$, if any zero of~$f$ is also a
-zero of~$g$, then there is a polynomial~$h$ such that~$g = hf$.} The
-polynomial~$h$ can be regarded as an ``algebraic certificate'' of the
-hypothesis. This principle holds for instance in the case that~$A$ is an
-algebraically closed field and that~$g$ is the unit polynomial. It is also
-true, without any restriction on~$g$, for the generic ring.
-
-We could try to generalize the Nullstellensatz to arbitrary geometric theories~$\TT$ as
-follows. \emph{For any geometric sequent~$\sigma$, if~$\sigma$ holds for a
-given~$\TT$-model~$M$ then~$\sigma$ is provable modulo~$\TT$.} In place of the
-algebraic certificate we now have a logical certificate, a \emph{proof}
-of~$\sigma$. However, this generalized statement is typically false, even for
-the generic model~$U_\TT$: The statement
-\begin{multline*}
-  \qquad\qquad\qquad\Set[\TT] \models \ulcorner\text{for any geometric sequent~$\sigma$,} \\
-    \text{if~$\sigma$ holds for~$U_\TT$ then~$\ul{\TT}$
-    proves~$\sigma$}\urcorner\qquad\qquad\qquad
-\end{multline*}
-does not hold.\footnote{Here~$\ul{\TT}$ is the internal geometric theory induced
-by~$\TT$, obtained by pulling back the set of sorts, the set of function
-symbols and so on along the geometric morphism~$\Set[\TT] \to \Set$. For
-instance, if~$\TT$ is the theory of rings, then from the internal point of view
-of~$\Set[\TT]$ the theory~$\ul{\TT}$ will again be the theory of rings.
-More details will be given in Section~\ref{sect:review}. The corner quotes indicate
-that for sake of readability, the translation into formal language is to be
-carried out by the reader.}
-
-For instance, let~$\TT$ be the theory of rings and let~$\sigma$ be
-the sequent~$(\top \vdash 1 + 1 = 0)$. Since there is an intuitionistic proof
-that~$\TT$ does not prove~$\sigma$ and toposes are sound with respect to
-intuitionistic logic, the statement~$\speak{\ul{\TT} proves~$\sigma$}$ is false
-from the internal point of view of~$\Set[\TT]$. However, it is not the case
-that the statement~$\speak{$1 + 1 = 0$ in~$U_\TT$}$ is false from the internal
-point of view. In fact, this statement holds in a nontrivial slice
-of~$\Set[\TT]$, the open subtopos coinciding with the classifying topos of
-the theory of rings of characteristic two.
-
-Intuitively, the problem is that while the meaning of~$\speak{$\ul{\TT}$
-proves~$\sigma$}$ is fixed, the meaning of~$\speak{$\sigma$ holds for~$U_\TT$}$ can vary
-with the slice. This problem can be solved by passing from~$\ul{\TT}$ to a
-varying theory, the internal theory~$\ul{\TT}/U_\TT$ defined in
-Section~\ref{sect:main}. If~$\TT$ is the theory of rings, then~$\ul{\TT}/U_\TT$
-is the~$\Set[\TT]$-theory of~$U_\TT$-algebras.
-Unlike~$\ul{\TT}$, this theory is not the pullback of an external geometric
-theory. We then have, subject to some qualifications made precise in
-Section~\ref{sect:main}:
-
-\begin{thm}\label{thm:nullstellensatz0}
-Let~$\TT$ be a geometric theory. Then, internally to~$\Set[\TT]$:
-\begin{equation}\tag{$\ddagger$}\label{eq:nullstellensatz}
-\text{A geometric$^*$ sequent~$\sigma$ holds for~$U_\TT$ if and only
-if~$\ul{\TT}/U_\TT$ proves$^*$~$\sigma$.}
-\end{equation}
-\end{thm}
-
-To illustrate Theorem~\ref{thm:nullstellensatz0}, let~$\TT$ be the theory of
-rings and let~$\sigma$ be of the form~$(f(x) = 0 \seq{x} g(x) = 0)$ for some
-polynomials~$f$ and~$g$. To say that~$\sigma$ holds for~$U_\TT$ amounts to
-saying that any zero~$x \? U_\TT$ of~$f$ is also a zero of~$g$, and to say
-that~$\ul{\TT}/U_\TT$ proves~$\sigma$ amounts to saying that
-in~$U_\TT[X]/(f(X))$, the free~$U_\TT$-algebra on one generator~$X$ subject to
-the relation~$f(X) = 0$, the relation~$g([X]) = 0$ holds. Hence we obtain
-\[ \Set[\TT] \models
-  \forall f,g \? U_\TT[X]\_ \bigl(
-    (\forall x \? U_\TT\_ f(x) = 0 \Rightarrow g(x) = 0) \Longleftrightarrow
-      \exists h \in U_\TT[X]\_ g = hf\bigr). \]
-
-In case that~$\TT$ is the quotient of a Horn theory XXX
-
-Theorem~\ref{thm:nullstellensatz0} is a source of nongeometric sequents. Indeed,
-it is the universal such source in the sense that any first-order formula
-holding for~$U_\TT$ can be deduced from~\eqref{eq:nullstellensatz}:
-
-\begin{thm}\label{thm:characterization0}
-Let~$\TT$ be a geometric theory. Let~$\alpha$ be a first-order
-formula over the signature of~$\TT$. Then the following statements are
-equivalent.
-\begin{enumerate}
-\item The formula~$\alpha$ holds for~$U_\TT$.
-\item The formula~$\alpha$ is provable in first-order intuitionistic logic
-modulo the axioms of~$\TT$ and the additional axiom
-scheme~\eqref{eq:nullstellensatz}.
-\end{enumerate}
-\end{thm}
-
-\paragraph{Acknowledgments} XXX
-
-\section{Background on classifying toposes}
-
-\section{Proofs of the main theorems}
-
-The following lemma shows that for coherent theories, there is no difference
-between provability and provability$^*$.
-
-\begin{lemma}Let~$\E$ be a Grothendieck topos. Let~$\TT$ be a coherent theory
-internal to~$\E$. Let~$\sigma$ be a coherent sequent over the signature
-of~$\TT$. Then the following statements are equivalent.
-\begin{enumerate}
-\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of externally
-finite shape}.$
-\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of arbitrary
-external shape}.$
-\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of arbitrary
-internal shape}.$
-\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of internally
-finite shape}.$
-\end{enumerate}
-\end{lemma}
-
-\begin{proof}It is obvious that~(1) implies~(2) implies~(3).
-
-To verify that~(3) implies~(4), we can mimick the usual proof of this fact in
-the internal language of~$\E$: One can construct a variant of the syntactic
-site of~$\TT$ which is built using only coherent sequents and finitary
-derivability~\cite[Section~1.4]{caramello:tst}. The sheaf topos over this site
-is another model for the classifying topos of~$\TT$. If~$\sigma$
-is~$\TT$-derivable by a proof tree of arbitrary shape, then~$\sigma$ holds in
-this model of the classifying topos. By the analogue of
-Proposition~\ref{prop:classifying-topos} for this model, we obtain
-that~$\sigma$ is~$\TT$-derivable by a proof tree of finite shape.
-
-That~(4) implies~(1) is a routine exercise exploiting that
-\[ \E \models \forall X\_ \speak{$X$ is Kuratowski-finite} \Leftrightarrow
-  \bigvee_{n \geq 0} \exists x_1,\ldots,x_n \? X\_
-  \forall x \? X\_ \bigvee_{i=1}^n x = x_i. \]
+\begin{proof}(This is what Sina confirmed for us.) Spell out construction.
+(Maybe in Section 4.)
 \end{proof}
 
-\section{The special case of Horn theories}
+\section{Background on internal geometric theories}
 
-Throughout this section, let~$\TT$ be a Horn theory.
+XXX: State definition of geometric theory, being careful to only use geometric
+constructions
 
-\begin{lemma}\label{lemma:free-models}
-Let~$X$ be a set equipped with a morphism~$X \to S$ to the set of sorts
-of the signature~$\Sigma$ of~$\TT$. Let~$R$ be a set of atomic propositions in which the
-elements of~$X$ may appear as new constants of the respective sorts. Then there
-is~$\TT\langle X | R \rangle$, the free~$\TT$-model on the generators~$X$ modulo
-the relations~$R$.\end{lemma}
+XXX: State definition of model of a geometric theory, being careful to only use
+geometric constructions
 
-\begin{proof}The desired model can be constructed as a term algebra. As a set,
-it consists of the terms (in the empty context) of the signature~$\Sigma + X$
-modulo the equivalence relation identifying two terms if and only if~$\TT + R$
-proves them to be equal. The function symbols~$f$ of~$\Sigma$ are interpreted
-by declaring~$\brak{f}([t_1],\ldots,[t_n]) = [f(t_1,\ldots,t_n)]$ and the
-relation symbols~$R$ are interpreted by declaring~$([t_1],\ldots,[t_n]) \in
-\brak{S} \Leftrightarrow (\TT + R \vdash S(t_1,\ldots,t_n))$.
+XXX: Explain internal geometric theories
 
-We omit the required verifications and only remark that while the same
-construction can be carried out if~$\TT$ is a general geometric theory, the
-resulting object would in general not be a model of~$\TT$.
-\end{proof}
+XXX: Remark subtletly with respect to infinitary disjunctions and infinitary
+proofs
 
-\begin{lemma}The category of~$\TT$-models is complete and
-cocomplete.\end{lemma}
+\section{Proof of the main result}
 
-\begin{proof}Limits are computed as the limits of the underlying sets, colimits
-are computed by using the construction of Lemma~\ref{lemma:free-models}. For
-instance, the coproduct of~$\TT\langle X | R \rangle$ and~$\TT\langle X' | R'
-\rangle$ is~$\TT\langle X \amalg X' \,|\, R, R' \rangle$.\end{proof}
+XXX: State and prove main lemma about~$\Set[\TT][\TT']$
 
-Having the special case of the theory of rings in mind, we write the coproduct
-in the category of~$\TT$-models as~``$\otimes$''.
+\section{Criteria for theories to be of presheaf type}
 
-\begin{lemma}Let~$\sigma = (\varphi_1 \wedge \cdots \wedge \varphi_n
-\seq{x_1,\ldots,x_k} \psi_1 \wedge \cdots \wedge \psi_m)$ be a Horn sequent
-over the signature of~$\TT$. Then the following statements are equivalent.
-\begin{enumerate}
-\item The theory~$\TT$ proves~$\sigma$.
-\item In~$\TT\langle x_1,\ldots,x_k | \varphi_1,\ldots,\varphi_n \rangle$, the
-propositions~$\psi_1,\ldots,\psi_m$ hold for
-the~$k$-tuple~$([x_1],\ldots,[x_k])$.
-\end{enumerate}
-\end{lemma}
+XXX: Give semantic criterion for being of presheaf type: Let~$\TT$ be of
+presheaf type and let~$\TT''$ be an extension of~$\TT$ such that there is a
+theory~$\TT'$ internal to~$\Set[\TT]$ such that~$\Set[\TT][\TT'] =
+\Set[\TT'']$. If~$\TT'$ is of presheaf type from the point of view
+of~$\Set[\TT]$, then~$\TT''$ is of presheaf type.
 
-\begin{proof}By construction of the term algebra.\end{proof}
+XXX: Explain that in the previous paragraph, the existence of~$\TT'$ is not
+actually an assumption, but is in fact always satisfied. (?)
 
-\begin{lemma}A~$\TT$-model is a finitely presented object of the category
-of~$\TT$-models if and only if it is isomorphic to a model of the
-form~$\TT\langle X | R \rangle$ where~$X$ is Bishop-finite and~$R$ is
-Kuratowski-finite.
-\end{lemma}
-
-\begin{proof}It is an instructive exercise to verify that models of the stated
-form are compact. Conversely, let a~$\TT$-model~$M$ be given. Then~$\TT$ is the
-filtered colimit of all models over~$M$ which are of the stated form. If~$M$ is
-compact, the identity on~$M$ factors over such a model. Hence~$M$ is a
-retract of such a model and hence itself isomorphic to a model of this form.
-\end{proof}
-
-% Set[T] einführen
-
-Any~$\TT$-model~$A$ has a mirror image in the topos~$\Set[\TT]$, namely the
-functor~$A^\sim : \Mod{\TT}_\mathrm{fp} \to \Set$ given by~$T \mapsto A \otimes T$.
-This object is in a canonical way a~$\TT$-model over~$U_\TT$, hence from the
-point of view of~$\Set[\TT]$ a~$\ul{\TT}/U_\TT$-model.
-
-\begin{lemma}The functor~$(\cdot)^\sim$ from~$\TT$-models to~$\ul{\TT}/U_\TT$-models
-in~$\Set[\TT]$ is left adjoint to functor~$\Gamma = \Hom(1, \cdot)$ computing
-global elements.
-\end{lemma}
-
-\begin{proof}An~$U_\TT$-algebra-homomorphism~$\alpha : A^\sim \to M$ yields
-the~$\TT$-model-homo\-mor\-phism~$\alpha_0 : A \to M(0) = \Gamma(M)$, where~$0$ is the
-initial~$\TT$-model. Conversely, a~$\TT$-model-homomorphism~$\beta : A \to
-\Gamma(M)$ yields an~$U_\TT$-algebra-homomorphism by summing~$A \to M(0) \to
-M(T)$ with the structure morphism~$T = U_\TT(T) \to
-M(T)$.\end{proof}
-
-\begin{defn}The \emph{spectrum}~$\Spec(M)$ of a~$U_\TT$-algebra~$M$ in~$\Set[\TT]$
-is the result of constructing, internally to~$\Set[\TT]$, the set
-of~$U_\TT$-algebra-homomorphisms~$M \to U_\TT$.
-\end{defn}
-
-Externally, the spectrum of~$M$ is the functor
-mapping a finitely presented~$\TT$-model~$T$ to~$\Hom_T(M(T), T)$, the set
-of~$\TT$-homomorphisms~$M(T) \to T$ compatible with the structure morphisms~$T
-\to M(T)$ and~$T \to T$.
+XXX: Draw conclusions, for instance that adding a further constant preserves
+being of presheaf type (find lots of further such syntactical examples).
 
@@
 
-%We do have, for any geometric sequent~$\sigma$, that
-%\[ \Set[\TT] \models \speak{$\sigma$ holds for~$U_\TT$} \quad\text{iff}\quad
-%  \Set[\TT] \models \speak{$\ul{\TT}$ proves~$\sigma$}, \]
-%since~$\Set[\TT] \models \speak{$\sigma$ holds for~$U_\TT$}$ implies that~$\TT$
-%proves~$\sigma$ and since~$\TT \vdash \sigma$ implies~$\E \models
-%(\ul{\TT} \vdash \sigma)$ for any topos~$\E$ over~$\Set$. However, this
-%equivalence is not stable under slicing ...
+XXX: Results from the Master thesis of Matthias
+
+XXX: Show that the naive definition of the big Zariski topos does not work.
+(From the point of view of Sh(S), the big Zariski topos of S is \emph{not} the
+topos of local~$\O_S$-algebras, but the topos of local~$\O_S$-algebras which
+are local over~$\O_S$.)
+
+\section{XXX general todo}
+
+XXX cite Martha Bunge's characterization of functors of presheaf type
+
+XXX cite Shawn Henry's PhD thesis
+
+XXX cite that other guy who mentioned internal geometric theories (forgot whom
+right now, might be Freyd or Andreas Blass, either in the result that the
+object of Kuratowski-finite subsets is preserved under pullback or in the paper
+about existential fixed-point logic)
+
+XXX cite Azriel Lévy
+\url{https://www.ams.org/journals/proc/1961-012-02/S0002-9939-1961-0122702-2/S0002-9939-1961-0122702-2.pdf}
 
@@
 \end{document}
-
-* Discuss case of SDG: Why does the Kock--Lawvere there hold only for Weil
-  algebras? Answer: It doesn't really! We do have that [U,U] is isomorphic to
-  the free C^∞-ring on one generator. But this observation doesn't appear to be
-  useful for further developments.
-
-* Cite https://docplayer.net/55796150-Intuitionistic-algebra-some-recent-developments-in-topos-theory.html
-  (in general and for internal theories)
-
-\begin{defn}Let~$\TT$ be a geometric theory. The theory~$\ul{\TT}/U_\TT$ is the
-geometric theory internal to~$\Set[\TT]$ which arises from~$\ul{\TT}$
-by adding additional constant symbols~$e_x$, one for each
-element~$x : U_\TT$, axioms~$(\top \vdash f(e_{x_1},\ldots,e_{x_n}) =
-e_{f(x_1,\ldots,x_n)})$ for each function symbol~$f$ and~$n$-tuple of elements
-of~$U_\TT$, and axioms~$(\top \vdash R(e_{x_1},\ldots,e_{x_n}))$ for each
-relation symbol~$R$ and~$n$-tuple~$(e_{x_1},\ldots,e_{x_n})$ such
-that~$R(e_{x_1},\ldots,e_{x_n})$ holds for~$U_\TT$.
-\end{defn}
-
-From the point of view of~$\Set[\TT]$, a model of~$\ul{\TT}/U_\TT$ is a model
-of~$\ul{\TT}$ equipped with a~$\ul{\TT}$-homomorphism from~$U_\TT$. In
-particular, the identity~$(U_\TT \to U_\TT)$ is a model of~$\ul{\TT}/U_\TT$.
-We then have:
-
-% We would not expect the statement
-% \[ \text{``for any geometric sequent~$\sigma$, if~$\sigma$ holds for~$A$
-% then the theory of rings proves~$\sigma$''} \]
-% to hold for an ordinary ring~$A$ in the topos of sets.
-%
-% Let~$M$ be a~$\TT$-model. We then trivially have
-% signature of~$\TT$. While we always have the implication
-% \[ \text{$\TT$ proves~$\sigma$} \quad\Longrightarrow\quad
-%   \text{$\sigma$ holds for~$M$}, \]
-% the converse is of course typically false,
-%
-% Furthermore, in a first introduction, it is useful to
-% restrict to an instructive special case, such as the theory of (commutative) rings.
-%
-% A simple version of the classical Nullstellensatz, valid for algebraically
-% closed fields, states: Let~$f \in A[X]$ be a polynomial. If
-% \[ \forall x \in A\_ (f(x) = 0 \in A \Rightarrow 1 = 0 \in A), \]
-% then there is a polynomial~$g \in A[X]$ such that~$gf = 1$. This polynomial~$g$
-% can be regarded as an ``algebraic certificate'' that the equation~$f(x) = 0$
-% only has a solution in the trivial case~$1 = 0$. The conclusion can also be phrased as
-% follows: In the free~$A$-algebra on one generator modulo the relation~$f(X) =
-% 0$, the relation~$1 = 0$ holds.
-%
-% Let~$A$ be a specific ring. Let~$f,g \in \ZZ[X]$ be
-% polynomials. While we always have the implication
-% \[ g([X]) = 0 \in \ZZ[X]/(f) \quad\Longrightarrow\quad
-%   \forall x \in A\_ (f(x) = 0 \Rightarrow g(x) = 0), \]
-% the converse is typically false: Just that~$g(x) = 0$ holds for every zero
-% of~$f$ in~$A$ does not imply that~$g([X]) = 0$ holds for the generic zero
-% of~$f$ in the free ring on one generator~$X$ modulo the relation~$f(X) = 0$.
-%
-%
-% the case that~$\TT$ is a Horn theory (where the axioms are of the
-% form~$(\phi \seq{\vec x} \psi)$ with the formulas~$\phi$ and~$\psi$ being
-% finite conjunctions of atomic propositions (equality and the given relations)),
-% such that we can switch from a logical terminology to a more algebraic one.
-% In the Horn case, for any set~$X$ and any set~$R$ of atomic propositions the
-% term algebra construction yields~$\TT\langle X | R\rangle$, the
-% free~$\TT$-model generated by~$X$ modulo the relations~$R$. We will also abuse
-% notation to pretend that~$\TT$ is single-sorted.
-%
-% Let~$M$ be a specific~$\TT$-model and assume that for
-% all~$n$-tuples~$x = (x_1,\ldots,x_n)$ of elements of~$M$, if some set~$R_1$ of
-% atomic propositions is satisfied for~$x$ then so is some set~$R_2$. In this
-% case it is unreasonable to expect that~$R_2$ holds for the~$n$-tuple of
-% generators of~$\TT\langle e_1,\ldots,e_n | R_1\rangle$.
-%
-% let~$R_1$ and~$R_2$ be sets of atomic
-% propositions with~$n$ free variables. Then
-% and assume, for some sets of relations~$R_1$
-% and~$R_2$
-%
-% The term algebra of a Horn theory~$\TT$ is always a model of~$\TT$, such that

### 2019-03-09 paper-qcoh.tex
to do
typo

@@
 by any local ring in any topos. Hence we may, when verifying a general fact
-about local rings, suppose without loss of generality that the given ring is a
+about local rings which is expressible as a geometric sequent,
+suppose without loss of generality that the given ring is a
 field. This observation is due to Kock~\cite{kock:univ-proj-geometry}, who exploited it to
@@
 hypothesis. This principle holds for instance in the case that~$A$ is an
-algebraically closed field and that~$g$ is the unit polynomial. It is also
+algebraically closed field and that~$g$ is the unit polynomial. We will see
+below that it is also
 true, without any restriction on~$g$, for the generic ring.
@@
 We could try to generalize the Nullstellensatz to arbitrary geometric theories~$\TT$ as
-follows. \emph{For any geometric sequent~$\sigma$, if~$\sigma$ holds for a
+follows: \emph{For any geometric sequent~$\sigma$, if~$\sigma$ holds for a
 given~$\TT$-model~$M$ then~$\sigma$ is provable modulo~$\TT$.} In place of the
@@
 that for sake of readability, the translation into formal language is to be
-carried out by the reader.}
+carried out by the reader. \par The displayed statement is much stronger than
+the statement that for any geometric sequent~$\sigma$, if $\Set[\TT] \models
+\speak{$\sigma$ holds for~$U_\TT$}$ then~$\TT$ proves~$\sigma$. This latter statement,
+where the universal quantifier and the ``if \ldots then'' have been pulled out,
+is true.}
 
@@
 theory. We then have, subject to some qualifications made precise in
-Section~\ref{sect:main}:
+Section~\ref{sect:main}, the following general Nullstellensatz:
 
@@
 To illustrate Theorem~\ref{thm:nullstellensatz0}, let~$\TT$ be the theory of
-rings and let~$\sigma$ be of the form~$(f(x) = 0 \seq{x} g(x) = 0)$ for some
+rings and let~$\sigma$ be the sequent~$(f(x) = 0 \seq{x} g(x) = 0)$ for some
 polynomials~$f$ and~$g$. To say that~$\sigma$ holds for~$U_\TT$ amounts to
@@
 indeed in general it does not. We formulate a substitute as Theorem~XXX,
-substantially broading the scope of the Nullstellensatz.
+substantially broadening the scope of the Nullstellensatz.
 
@@
 out~$\Set[\TT']$ is subcanonical. For instance, this is the case if~$\TT$ is
-the theory of rings and~$\Set[\TT']$ is any of the well-known toposes in
+the theory of rings and~$\Set[\TT']$ is one of several well-known toposes in
 algebraic geometry such as the big Zariski topos, the big étale topos or the
@@
 by declaring~$\brak{f}([t_1],\ldots,[t_n]) = [f(t_1,\ldots,t_n)]$ and the
-relation symbols~$R$ are interpreted by declaring~$([t_1],\ldots,[t_n]) \in
+relation symbols~$S$ are interpreted by declaring~$([t_1],\ldots,[t_n]) \in
 \brak{S} \Leftrightarrow (\TT + R \vdash S(t_1,\ldots,t_n))$.
@@
 We omit the required verifications and only remark that while the same
-construction can be carried out if~$\TT$ is a general geometric theory, the
+construction could be carried out if~$\TT$ is a general geometric theory, the
 resulting object would in general not be a model of~$\TT$.
@@
 
-\begin{lemma}A~$\TT$-model is a finitely presented object of the category
+\begin{lemma}A~$\TT$-model is finitely presentable as an object of the category
 of~$\TT$-models if and only if it is isomorphic to a model of the
@@
 \begin{lemma}Let~$A$ be a~$\TT$-model. Then~$\Spec(A^\sim)$ coincides
-with~$yA$, where~$y : \Mod{\TT} \to \Set[\TT]$ is the Yoneda embedding.
+with~$yA$, the functor~$\Hom_{\Mod{\TT}}(A, \cdot)$.
 \end{lemma}
@@
   (in general and for internal theories)
+
+* Discuss related work, especially: Olivia 6.1.3, ...
 
@@
   (Beispiel: ¬(x=0) ⇒ (x inv.))
+
+* Anmerken, dass wir in konstruktiver Metalogik arbeiten.

### 2019-03-11 paper-qcoh.tex
Proofs of the main theorems

@@
 a model such that for any geometric sequent~$\sigma$, the following notions
-coincide: \begin{enumerate} \item The sequent~$\sigma$ is provable
-modulo~$\TT$. \item The sequent~$\sigma$ holds for any~$\TT$-model in any
-Grothendieck topos. \item The sequent~$\sigma$ holds for~$U_\TT$.
-\end{enumerate} One could argue that it is this model which mathematicians
+coincide:
+\begin{enumerate}
+\item The sequent~$\sigma$ is provable modulo~$\TT$.
+\item The sequent~$\sigma$ holds for any~$\TT$-model in any Grothendieck topos.
+\item The sequent~$\sigma$ holds for~$U_\TT$.
+\end{enumerate}
+One could argue that it is this model which mathematicians
 implicitly refer to when they utter the phrase ``Let~$M$ be a~$\TT$-model.''. It can
@@
 used, roughly reviewed in Section~\ref{sect:review} below. We write~``$\E
-\models \alpha$'' to mean that the statement~$\alpha$ holds from the internal
+\models \alpha$'' to mean that the formula~$\alpha$ holds from the internal
 point of view of~$\E$. Since this language is a form of a higher-order
@@
 where the universal quantifier and the ``if \ldots then'' have been pulled out,
-is true.}
+is true.} In this sense~$\Set[\TT]$ does not believe that~$U_\TT$ is the
+generic~$\ul{\TT}$-model.
 
-For instance, let~$\TT$ be the theory of rings and let~$\sigma$ be
+A concrete counterexample is as follows. Let~$\TT$ be the theory of rings and let~$\sigma$ be
 the sequent~$(\top \vdash 1 + 1 = 0)$. Since there is an intuitionistic proof
@@
 
-\begin{proof}It is obvious that~(1) implies~(2) implies~(3).
+\begin{proof}It is trivial that~(1) implies~(2) implies~(3).
 
@@
 \end{proof}
+
+\begin{lemma}\label{lemma:truth-to-provability}
+Let~$\TT$ be a geometric theory. Let~$\alpha$ be a geometric formula over the
+signature of~$\TT$ in a context~$\vec x$. Then
+\[ \{ \vec x\_ \alpha \} \models \speak{$\ul{\TT}/U_\TT$ proves $(\top
+\vdash_{[]} \alpha)$}, \]
+where the free variables~$\vec x$ occurring in~$\alpha$ are interpreted by
+their generic values over~$\{ \vec x\_ \alpha \}$.
+\end{lemma}
+
+% XXX fill in proof
+
+\begin{lemma}\label{lemma:locally-constant}
+Let~$\TT$ be a geometric theory.
+Let~$\varphi$ be a section of the sheaf of geometric$^\star$ formulas over the signature
+of~$\ul{\TT}/U_\TT$ over a stage~$A$. Then there is a covering~$(A_i \to A)_i$
+of~$A$ such that for each index~$i$, there is a formula~$\varphi_i$ over the signature
+of~$\TT/U_\TT(A_i)$ such that~$A_i \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$
+$(\varphi \dashv\vdash \varphi_i)$}$.
+\end{lemma}
+
+\begin{proof}By passing to a covering, we may suppose that~$\varphi$ is given
+by an (external) geometric formula over the signature
+of~$\ul{\TT}(A)/U_\TT(A)$. Any function symbol and relation symbol
+of~$\ul{\TT}(A)$ occuring in~$\varphi$ is locally given by a symbol in~$\TT$.
+Hence the claim would be trivial if~$\varphi$ were a coherent formula, for in
+this case we would just have to pass to further coverings, one for each
+occuring symbol, a finite number of times in total.
+
+However, in general, we cannot conclude as easily. Write~$A = \{ \vec x\_
+\alpha \}$. Let~$R$ be a relation symbol of~$\ul{\TT}(A)$ occuring
+in~$\varphi$. Then there is a covering~$(\{ \vec y_j\_ \alpha_j \}
+\xrightarrow{[\theta_j]} \{ \vec x\_ \alpha \})_j$ such that, restricted to~$\{
+\vec y_j\_ \alpha_j \}$, $R$ is given by a symbol~$R_j$ of~$\TT$. To construct
+the desired formula~$\varphi'$, we replace any such occurrence~$R(\ldots)$
+in~$\varphi$ by
+\[ \bigvee_j \bigl((\exists \vec y_j\_ \theta_j) \wedge R_j(\ldots)\bigr). \]
+In a similar vein we treat any occurrence of function symbols.
+
+The verification of~$A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$
+$(\varphi \dashv\vdash \varphi')$}$ rests on the observation
+\[ A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$ $\bigl((\exists \vec y_j\_
+\theta_j) \seq{\vec x} \bigvee_j \{ \top \,|\, \text{$(\exists \vec y_j\_ \theta_j)$
+holds for~$U_\TT$} \}\bigr)$} \]
+which in turn follows from Lemma~\ref{lemma:truth-to-provability} exploiting
+that~$(\exists \vec y_j\_ \theta_j) \wedge (\exists \vec y_k\_ \theta_k)
+\seq{\vec x} \bigvee \{ \top \,|\, j = k \}$.
+\end{proof}
+
+\begin{thm}\label{thm:nullstellensatz}
+Let~$\TT$ be a geometric theory. Then, internally to~$\Set[\TT]$, for any
+geometric$^\star$ sequent~$\sigma$ over the signature of~$\ul{\TT}/U_\TT$, the
+following statements are equivalent:
+\begin{enumerate}
+\item The sequent~$\sigma$ holds for~$U_\TT$.
+\item The sequent~$\sigma$ is provable$^\star$ modulo~$\ul{\TT}/U_\TT$.
+\end{enumerate}
+\end{thm}
+
+\begin{proof}The direction~$(2) \Rightarrow (1)$ is immediate because~$U_\TT$ is, from the
+internal point of view of~$\Set[\TT]$, a~$\ul{\TT}/U_\TT$-model. Hence even the
+following stronger statement holds internally: \emph{For any geometric
+sequent~$\sigma$, if~$\ul{\TT}/U_\TT$ proves~$\sigma$, then~$\sigma$ holds
+for~$U_\TT$.}
+
+For the direction~$(1) \Rightarrow (2)$ we have to verify that, given any
+stage~$A$ and any section~$\sigma$ of the sheaf of
+geometric$^\star$ sequents over~$A$, if~$A \models
+\speak{$\sigma$ holds for~$U_\TT$}$ then~$A \models
+\speak{$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$}$. By
+Lemma~\ref{lemma:locally-constant} we may suppose that~$\sigma$ is an
+(external) geometric sequent over the signature of~$\TT/U_\TT(A)$.
+
+Writing~$A = \{ \vec x\_ \alpha \}$ and~$\sigma = (\varphi \seq{\vec y} \psi)$,
+we have~$\{ \vec x\_ \alpha \} \models \forall y\_ (\varphi \Rightarrow \psi)$,
+hence~$\{ \vec x, \vec y\_ \alpha \wedge \varphi \} \models \psi$. Thus~$\TT$
+proves~$(\alpha \wedge \varphi \seq{\vec x, \vec y} \psi)$. This proof can be
+pulled back from~$\Set$ to~$\Set[\TT]/A$ to show~$A \models
+\speak{$\ul{\TT}/U_\TT$ proves$^\star$~$(\alpha
+\wedge \varphi \seq{\vec x, \vec y} \psi)$}$. By
+Lemma~\ref{lemma:truth-to-provability}, we also have~$A \models
+\speak{$\ul{\TT}/U_\TT$ proves$^\star$~$(\top \seq{[]} \alpha)$}$ (where the free
+variables occuring in~$\alpha$ are interpreted as the generic values available
+over~$A$), hence~$A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$~$(\varphi
+\seq{\vec y} \psi)$}$.
+\end{proof}
+
+% XXX: check contexts.
 
@@
 big fppf topos.
+
+\begin{thm}\label{thm:characterization}
+Let~$\TT$ be a geometric theory. Let~$\alpha$ be a first-order
+formula over the signature of~$\TT$. Then the following statements are
+equivalent.
+\begin{enumerate}
+\item The formula~$\alpha$ holds for~$U_\TT$.
+\item The formula~$\alpha$ is provable in first-order intuitionistic logic
+modulo the axioms of~$\TT$ and the additional axiom
+scheme~\eqref{eq:nullstellensatz}.
+\end{enumerate}
+\end{thm}
+
+\begin{proof}The direction~$(1) \Rightarrow (2)$ is immediate, as the internal
+language of~$\Set[\TT]$ validates first-order intuitionistic logic and the
+axiom scheme~\eqref{eq:nullstellensatz} by Theorem~\ref{thm:nullstellensatz}.
+
+For the converse direction, we note that, given a geometric sequent~$\sigma$
+over the signature of~$\TT$, the statement~$\speak{$\ul{\TT}/U_\TT$
+proves$^\star$~$\sigma$}$ can be expressed as a geometric formula over the
+signature of~$\TT$. Applying this observation successively to subformulas
+of the given formula~$\alpha$, Theorem~\ref{thm:nullstellensatz} on the
+semantic side and the axiom scheme~\eqref{eq:nullstellensatz} imply that we may
+assume that~$\alpha$ is in fact a geometric formula. Hence we are reduced to
+the basic fact that, for geometric formulas~$\varphi$, $\Set[\TT] \models
+\varphi$ implies that~$\TT$ proves~$\varphi$.
+\end{proof}
+
+% XXX coherent case
 
@@
 
+\section{The generalization to the higher-order case}
+
+\begin{thm}\label{thm:higher-order-nullstellensatz}
+Let~$\TT$ be a geometric theory. Let~$x_1\?X_1,\ldots,x_n\?X_n$ be a context over the
+signature of~$\TT$. Then the canonical morphism
+\[ \mathrm{Form}^\star_{\vec x}(\ul{\TT}/U_\TT)/(\dashv\vdash_{\vec x}) \longrightarrow P(X_1 \times
+\cdots \times X_n) \]
+sending the equivalence class of a geometric$^\star$ formula~$\varphi$ over the
+signature of~$\ul{\TT}/U_\TT$ to the subset~$\{ (x_1,\ldots,x_n) \,|\, \varphi
+\}$ is an isomorphism.
+% XXX explain equivalence relation
+\end{thm}
+
+An immediate corollary of Theorem~\ref{thm:higher-order-nullstellensatz} is the
+Nullstellensatz of Theorem~\ref{thm:nullstellensatz}. Indeed, arguing
+internally to~$\Set[\TT]$, let~$\sigma = (\varphi \seq{\vec x} \psi)$ be a
+geometric$^\star$ sequent over the signature of~$\ul{\TT}/U_\TT$. Assume
+that~$\sigma$ holds for~$U_\TT$. Then the subsets~$\{ (\vec x) \,|\, \varphi
+\}$ and~$\{ (\vec x) \,|\, \varphi \wedge \psi \}$ are equal. Hence, by
+Theorem~\ref{thm:higher-order-nullstellensatz}, the formulas~$\varphi$
+and~$\varphi \wedge \psi$ are provably~$^\star$ equivalent.
+Thus~$\ul{\TT}/U_\TT$ proves$^\star$~$(\varphi \seq{\vec x} \psi)$.
+
+% XXX corollary: function spaces
+
+% XXX prove higher-order characterization?
+
 \section{Applications}
+
+% Zariski
+
+% Kock--Lawvere in SDG
 
@@
 * Discuss related work, especially: Olivia 6.1.3, ...
+
+* Discuss higher-order case
 

### 2019-03-11 paper-qcoh.tex
Proof of basic stage-to-provability lemma

@@
 \newcommand{\mmm}{\mathfrak{m}}
+\newcommand{\C}{\mathcal{C}}
 \newcommand{\I}{\mathcal{I}}
@@
 Let~$\TT$ be a geometric theory. Let~$\alpha$ be a geometric formula over the
-signature of~$\TT$ in a context~$\vec x$. Then
+signature of~$\TT$ in a context~$x_1\?X_1,\ldots,x_n\?X_n$. Then
 \[ \{ \vec x\_ \alpha \} \models \speak{$\ul{\TT}/U_\TT$ proves $(\top
@@
 where the free variables~$\vec x$ occurring in~$\alpha$ are interpreted by
-their generic values over~$\{ \vec x\_ \alpha \}$.
+their generic values over~$\{ \vec x\_ \alpha \}$, that is the projection
+maps~$\{ \vec x\_ \alpha \} \to \{ x_i\?X_i\_ \top \}$.
 \end{lemma}
 
-% XXX fill in proof
+\begin{proof}By induction on the structure of~$\alpha$. The cases
+of~``$\top$'' and ``$\wedge$'' are trivial; the cases of~``$\bigvee$''
+and~``$\exists$'' follow from passing to suitable coverings; and the case of
+atomic propositions is by definition of~$\ul{\TT}/U_\TT$.
+\end{proof}
 

### 2019-03-11 paper-qcoh.tex
More careful rendition of contexts

@@
 
-To verify that~(3) implies~(4), we can mimick the usual proof of this fact in
-the internal language of~$\E$: One can construct a variant of the syntactic
+To verify that~(3) implies~(4), we can mimic the usual proof of this fact in
+the internal language of~$\E$: There is a variant of the syntactic
 site of~$\TT$ which is built using only coherent sequents and finitary
 derivability~\cite[Section~1.4]{caramello:tst}. The sheaf topos over this site
-is another model for the classifying topos of~$\TT$. If~$\sigma$
+is another model for the classifying topos of~$\TT$, and still validates, like
+any Grothendieck topos, full infinitary logic. Hence, if~$\sigma$
 is~$\TT$-derivable by a proof tree of arbitrary shape, then~$\sigma$ holds in
@@
 Let~$\varphi$ be a section of the sheaf of geometric$^\star$ formulas over the signature
-of~$\ul{\TT}/U_\TT$ over a stage~$A$. Then there is a covering~$(A_i \to A)_i$
+of~$\ul{\TT}/U_\TT$ over a stage~$A \in \C_\TT$. Then there is a covering~$(A_i \to A)_i$
 of~$A$ such that for each index~$i$, there is a formula~$\varphi_i$ over the signature
@@
 by an (external) geometric formula over the signature
-of~$\ul{\TT}(A)/U_\TT(A)$. Any function symbol and relation symbol
-of~$\ul{\TT}(A)$ occuring in~$\varphi$ is locally given by a symbol in~$\TT$.
+of~$\ul{\TT}(A)/U_\TT(A)$.
+
+Any function symbol and relation symbol
+of~$\ul{\TT}(A)$ occurring in~$\varphi$ is locally given by a symbol of~$\TT$.
 Hence the claim would be trivial if~$\varphi$ were a coherent formula, for in
@@
 \alpha \}$. Let~$R$ be a relation symbol of~$\ul{\TT}(A)$ occuring
-in~$\varphi$. Then there is a covering~$(\{ \vec y_j\_ \alpha_j \}
+in~$\varphi$. By the explicit description of constant sheaves as sheaves of
+locally constant maps, there is a covering~$(\{ \vec y_j\_ \alpha_j \}
 \xrightarrow{[\theta_j]} \{ \vec x\_ \alpha \})_j$ such that, restricted to~$\{
-\vec y_j\_ \alpha_j \}$, $R$ is given by a symbol~$R_j$ of~$\TT$. To construct
+\vec y_j\_ \alpha_j \}$, $R$ is given by a relation symbol~$R_j$ of~$\TT$. To construct
 the desired formula~$\varphi'$, we replace any such occurrence~$R(\ldots)$
@@
 $(\varphi \dashv\vdash \varphi')$}$ rests on the observation
-\[ A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$ $\bigl((\exists \vec y_j\_
-\theta_j) \seq{\vec x} \bigvee_j \{ \top \,|\, \text{$(\exists \vec y_j\_ \theta_j)$
+\[ A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$ $\bigl((\exists \vec y_k\_
+\theta_k) \seq{[]} \bigvee_k \{ \top \,|\, \text{$(\exists \vec y_k\_ \theta_k)$
 holds for~$U_\TT$} \}\bigr)$} \]
-which in turn follows from Lemma~\ref{lemma:truth-to-provability} exploiting
-that~$(\exists \vec y_j\_ \theta_j) \wedge (\exists \vec y_k\_ \theta_k)
-\seq{\vec x} \bigvee \{ \top \,|\, j = k \}$.
+which in turn can be checked on the covering~$(\{ \vec y_j\_ \alpha_j \}
+\xrightarrow{[\theta_j]} \{ \vec x\_ \alpha \})_j$, applying
+Lemma~\ref{lemma:truth-to-provability} and using that~$\TT$ (and
+hence~$\ul{\TT}$) proves~$(\exists \vec y_j\_ \theta_j) \wedge (\exists \vec
+y_k\_ \theta_k) \seq{\vec x} \bigvee \{ \top \,|\, j = k \}$.
 \end{proof}
@@
 For the direction~$(1) \Rightarrow (2)$ we have to verify that, given any
-stage~$A$ and any section~$\sigma$ of the sheaf of
+stage~$A \in \C_\TT$ and any section~$\sigma$ of the sheaf of
 geometric$^\star$ sequents over~$A$, if~$A \models
@@
 \end{proof}
-
-% XXX: check contexts.
 
@@
 \begin{multline*}
-  \text{For any geometric$^\star$ sequent~$\sigma = (\varphi \seq{\vec x} \psi)$ where~$\psi$ is a Horn formula,} \\
-    \text{if~$\forall x_1,\ldots,x_n\?U_\TT\_ (\varphi \Rightarrow \nabla
-    \psi)$, then~$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$.}
+  \text{\emph{For any geometric$^\star$ sequent~$\sigma = (\varphi \seq{\vec x} \psi)$ where~$\psi$ is a Horn formula,}} \\
+    \text{\emph{if~$\forall x_1,\ldots,x_n\?U_\TT\_ (\varphi \Rightarrow \nabla
+    \psi)$, then~$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$.}}
 \end{multline*}
@@
 of the given formula~$\alpha$, Theorem~\ref{thm:nullstellensatz} on the
-semantic side and the axiom scheme~\eqref{eq:nullstellensatz} imply that we may
+semantic side and the axiom scheme~\eqref{eq:nullstellensatz} on the syntactic side imply that we may
 assume that~$\alpha$ is in fact a geometric formula. Hence we are reduced to

### 2019-03-11 paper-qcoh.tex
Further statements

@@
 
-% Set[T] einführen
+% XXX Set[T] einführen
 
@@
 
-% XXX corollary: function spaces
+\begin{cor}Let~$\TT$ be a geometric theory. Let~$\{\vec x\_ \varphi\}$
+and~$\{\vec y\_ \psi\}$ be geometric formulas in given contexts. Then,
+internally to~$\Set[\TT]$, the canonical map from the set of equivalence classes
+of~$\ul{\TT}/U$-provably$^\star$ functional geometric$^\star$ formulas
+from~$\{\vec x\_ \varphi\}$ to~$\{\vec y\_ \psi\}$ to the set of maps~$\{(\vec y)
+\,|\, \psi \}^{\{(\vec x) \,|\, \varphi\}}$ is a bijection.
+\end{cor}
 
-% XXX prove higher-order characterization?
+% XXX fill in proof
+
+\begin{thm}\label{thm:higher-order-nullstellensatz}
+Let~$\TT$ be a geometric theory. Then, internally to~$\Set[\TT]$, for any
+higher-order geometric$^\star$ sequent~$\sigma$ over the signature
+of~$\ul{\TT}/U$, the following statements are equivalent:
+\begin{enumerate}
+\item The sequent~$\sigma$ holds for~$U_\TT$.
+\item The sequent~$\sigma$ is provable$^\star$ modulo~$\TT/U_\TT$ in geometric
+higher-order logic.
+\end{enumerate}
+\end{thm}
+
+% XXX fill in proof
+
+\begin{thm}\label{thm:higher-order-characterization}
+Let~$\TT$ be a geometric theory. Let~$\alpha$ be a higher-order formula over
+the signature of~$\TT$. Then the following statements are equivalent:
+\begin{enumerate}
+\item The formula~$\alpha$ holds for~$U_\TT$.
+\item The formula~$\alpha$ is provable in higher-order intuitionistic logic
+modulo the axioms of~$\TT$ and the additional axiom scheme~XXX.
+\end{enumerate}
+\end{thm}
+
+% XXX fill in proof
 

### 2019-03-13 paper-qcoh.tex
typo

@@
 \DeclareMathOperator{\List}{List}
+\DeclareMathOperator{\id}{id}
 \newcommand{\Set}{\mathrm{Set}}
@@
 symbols by the logical connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$
-and by arbitrary set-indexed disjunctions~$\bigvee$).
+and by arbitrary set-indexed disjunctions~$\bigvee$). By \emph{first-order
+formula} we will mean a formula which may contain, additionally to the
+connectives allowed for geometric formulas, the connectives~${\Rightarrow}$
+and~${\forall}$.
 
@@
 used, roughly reviewed in Section~\ref{sect:review} below. We write~``$\E
-\models \alpha$'' to mean that the formula~$\alpha$ holds from the internal
+\models \alpha$'' to mean that a formula~$\alpha$ holds from the internal
 point of view of~$\E$. Since this language is a form of a higher-order
-dependently-typed intuitionistic logic, the classifying topos~$\Set[\TT]$ can
+intuitionistic extensional dependent type theory, the classifying topos~$\Set[\TT]$ can
 be regarded as a higher-order completion of the geometric theory~$\TT$. The
@@
 (\affl)^{\Spec(A)}$, where~$\Spec(A)$ is defined as the set
-of~$\affl$-algebra-homomorphisms~$A \to \affl$, is bijective.
+of~$\affl$-algebra homomorphisms~$A \to \affl$, is bijective.
 \end{itemize}
@@
 classifying toposes~\cite[Speculation~22.1]{blechschmidt:phd}. This hope turned
-out to be true, in the sense we will now make precise.
+out to be true, in the sense we will now indicate.
 \medskip
@@
 
-An immediate corollary of Theorem~\ref{thm:higher-order-nullstellensatz} is the
+An immediate corollary of Theorem~\ref{thm:definability} is the
 Nullstellensatz of Theorem~\ref{thm:nullstellensatz}. Indeed, arguing
@@
 \}$ and~$\{ (\vec x) \,|\, \varphi \wedge \psi \}$ are equal. Hence, by
-Theorem~\ref{thm:higher-order-nullstellensatz}, the formulas~$\varphi$
-and~$\varphi \wedge \psi$ are provably~$^\star$ equivalent.
+Theorem~\ref{thm:definability}, the formulas~$\varphi$
+and~$\varphi \wedge \psi$ are provably$^\star$ equivalent.
 Thus~$\ul{\TT}/U_\TT$ proves$^\star$~$(\varphi \seq{\vec x} \psi)$.

### 2019-03-13 paper-qcoh.tex
Remark on explicitness of characterization

@@
 \end{thm}
+
+Theorem~\ref{thm:characterization0} characterizes which first-order formulas
+hold for the generic model. We could of course wish for a more explicit
+characterization; but since even the characterization of geometric sequents
+holding for the generic model (they are precisely those which are provable in
+geometric logic modulo~$\TT$) is of a rather implicit nature, this wish appears
+unfounded.
+
+We stress that our characterization is more explicit than the tautologous
+characterization (``a first-order formula holds for~$U_\TT$ iff it is provable
+in~$\TT'$, where~$\TT'$ is the first-order theory whose set of axioms is the
+set of first-order formulas satisfied by~$U_\TT$'') and the (incorrect)
+characterization ``a first-order formula holds for~$U_\TT$ iff it
+is~$\TT$-redundant''. Indeed, if~$\TT$ happens to be coherent and recursively
+axiomatizable, then in stating Theorem~\ref{thm:characterization0} we may
+restrict to coherent existential fixed-point logic, and the resulting theory
+will again be recursively axiomatizable.
+\medskip
 

### 2019-03-13 paper-qcoh.tex
Started section on background

@@
 
-\section{Background on classifying toposes}
+\section{Background on \ldots}
+
+\subsection{the internal language of Grothendieck toposes}
+
+% XXX
+
+\subsection{internal geometric theories}
+
+% XXX, especially the notion of provability and
+% provability*
+
+Given an internal theory~$\Sigma$ internal to a Grothendieck topos~$\E$ (or
+elementary topos with a natural numbers object), we can successively build the
+object of contexts (the object of lists of sorts); the object of terms
+(equipped with a morphism to the object of contexts); the object of atomic
+propositions (again equipped with such a morphism); the object of geometric
+formulas (again so); the object of geometric sequents (again so); and, given an
+internal theory~$\TT$ over~$\Sigma$, the object of proof trees of~$\TT$.
+
+\subsection{classifying toposes}
+
+\begin{defn}The \emph{syntactic site}~$\C_\TT$ of a geometric theory~$\TT$ has:
+\begin{enumerate}
+\item as objects ``geometric formulas in contexts''~$\{x_1\?X_1,\ldots,x_n\?X_n\_
+\varphi\}$ where~$\varphi$ is a geometric formula over the signature of~$\TT$
+in the displayed context;
+\item as set of morphisms~$\Hom_{\C_\TT}(\{\vec x\_
+\varphi\}, \{\vec y\_ \psi\})$ the set of formulas~$\theta$ in the
+context~$\vec x, \vec y$ which are~$\TT$-provably functional,
+modulo~$\TT$-provable equivalence of such formulas;
+\item as covering families those families~$(\{\vec x_i\_ \varphi_i\} \xrightarrow{\theta_i}
+\{\vec y\_ \psi\})_i$ for which~$\TT$ proves~$(\psi \seq{\vec y} \bigvee_i \exists \vec x_i\_
+\theta_i)$.
+\end{enumerate}
+\end{defn}
+
+\begin{defn}The \emph{classifying topos}~$\Set[\TT]$ of a geometric
+theory~$\TT$ is the topos of set-valued sheaves on~$\C_\TT$.\end{defn}
+
+Writing~$y : \C_\TT \to \Set[\TT]$ for the Yoneda embedding, the \emph{generic
+model}~$U_\TT$ of~$\TT$ interprets a sort~$X$ of~$\TT$ as the sheaf~$y\{x\?X\_ \top\}$,
+a function symbol~$f : X_1 \cdots X_n \to Y$ as the morphism given by
+the~$\TT$-provably functional formula~$f(x_1,\ldots,x_n) = y$ and a relation
+symbol~$R \rightarrowtail X_1 \cdots X_n$ by the subobject~$y\{\vec x\_ R(\vec
+x)\} \rightarrowtail y\{\vec x\_ \top\}$.
+
+\begin{thm}The generic model is universal in the sense that for any
+Grothendieck topos~$\E$, the functor
+\[ \text{(category of geometric morphisms~$\E \to \Set[\TT]$)} \longrightarrow
+\text{(category of~$\TT$-models in~$\E$)} \]
+given by~$f \mapsto f^*U_\TT$ is an equivalence of categories.
+\end{thm}
+
+\begin{prop}Let~$\alpha$ and~$\varphi$ be geometric formulas in a context~$\vec
+x$ over the signature of a geometric theory~$\TT$. Then the following
+statements are equivalent:
+\begin{enumerate}
+\item $\Set[\TT] \models \forall \vec x\_ (\alpha \Rightarrow \varphi)$.
+\item $\{\vec x\_ \alpha\} \models \varphi$.
+\item $\TT$ proves~$(\alpha \seq{\vec x} \varphi)$.
+\end{enumerate}
+\end{prop}
+
+\begin{proof}The equivalence~$(1) \Leftrightarrow (2)$ follows immediately by
+unrolling the Kripke--Joyal semantics. The equivalence~$(2) \Leftrightarrow
+(3)$ is by induction on the structure of~$\varphi$.\end{proof}
 

### 2019-03-13 paper-qcoh.tex
More work on the Horn case

@@
 
-\begin{proof}An~$U_\TT$-algebra-homomorphism~$\alpha : A^\sim \to M$ yields
-the~$\TT$-model-homo\-mor\-phism~$\alpha_0 : A \to M(0) = \Gamma(M)$, where~$0$ is the
-initial~$\TT$-model. Conversely, a~$\TT$-model-homomorphism~$\beta : A \to
-\Gamma(M)$ yields an~$U_\TT$-algebra-homomorphism by summing~$A \to M(0) \to
+\begin{proof}An~$U_\TT$-algebra homomorphism~$\alpha : A^\sim \to M$ yields
+the~$\TT$-model homo\-mor\-phism~$\alpha_0 : A \to M(0) = \Gamma(M)$, where~$0$ is the
+initial~$\TT$-model. Conversely, a~$\TT$-model homomorphism~$\beta : A \to
+\Gamma(M)$ yields an~$U_\TT$-algebra homomorphism by summing~$A \to M(0) \to
 M(T)$ with the structure morphism~$T = U_\TT(T) \to
@@
 is the result of constructing, internally to~$\Set[\TT]$, the set
-of~$U_\TT$-algebra-homomorphisms~$M \to U_\TT$.
+of~$U_\TT$-algebra homomorphisms~$M \to U_\TT$.
 \end{defn}
 
-Externally, the spectrum of~$M$ is the functor
-mapping a finitely presented~$\TT$-model~$T$ to~$\Hom_T(M(T), T)$, the set
-of~$\TT$-homomorphisms~$M(T) \to T$ compatible with the structure morphisms~$T
-\to M(T)$ and~$T \to T$.
+% This is wrong:
+% Externally, the spectrum of~$M$ is the functor
+% mapping a finitely presented~$\TT$-model~$T$ to~$\Hom_T(M(T), T)$, the set
+% of~$\TT$-homomorphisms~$M(T) \to T$ compatible with the structure morphisms~$T
+% \to M(T)$ and~$T \to T$.
 
-\begin{lemma}Let~$A$ be a~$\TT$-model. Then~$\Spec(A^\sim)$ coincides
+\begin{lemma}\label{lemma:spec-sim-representable}
+Let~$A$ be a~$\TT$-model. Then~$\Spec(A^\sim)$ coincides
 with~$yA$, the functor~$\Hom_{\Mod{\TT}}(A, \cdot)$.
@@
 
-\begin{proof}XXX\end{proof}
+\begin{proof}By the Yoneda lemma, the sections of the sheaf~$\Spec(A^\sim) :
+\Mod{\TT}_\mathrm{fp} \to \Set$ on an object~$T$ are given by the set
+\begin{align*}
+  \Spec(A^\sim)(T) &\cong \Hom(yT, \Spec(A^\sim)) =
+  \Hom(yT, [A^\sim,U_\TT]_{U_\TT}) \\
+  &\cong \Hom(yT \times A^\sim, U_\TT)_{\text{$U_\TT$-algebra homomorphism in second
+  argument}} \\
+  &\cong \Hom_{U_\TT}(A^\sim, (U_\TT)^{yT})
+  \cong \Hom_{U_\TT}(A^\sim, U_\TT|T),
+\end{align*}
+where~$[A^\sim,U_\TT]_{U_\TT}$ is the object of~$U_\TT$-algebra homomorphisms from~$A^\sim$
+to~$U_\TT$ (a subobject of the internal Hom~$U_\TT^{A^\sim}$); $\Hom_{U_\TT}$
+denotes the set of~$U_\TT$-algebra homomorphisms; $(U_\TT)^{yT}$ is the object of morphisms
+from~$yT$ to~$U_\TT$; and~$U_\TT|T$ is the functor~$U_\TT(T \times \cdot)$, that is~$S
+\mapsto T \otimes S$.
 
-\begin{lemma}Let~$A$ and~$B$ be finitely presented~$\TT$-models. Then the
-canonical morphism
+An arbitrary element~$f \in (yA)(T)$, that is an arbitrary~$\TT$-model
+homomorphism~$f : A \to T$, induces a~$U_\TT$-algebra homomorphism~$g : A^\sim \to
+U_\TT|T$ by setting~$g_S \defeq f \otimes \id_S : A \otimes S \to T \otimes S$. The
+given homomorphism~$f$ can be recovered by~$f = g_0$, the component of~$g$ at
+the initial model.
+
+Conversely, a~$U_\TT$-algebra homomorphism~$g : A^\sim \to U_\TT|T$ induces
+a~$\TT$-model homomorphism~$f : A \to T$ by setting~$f \defeq g_0$. Because~$g$
+is a natural transformation and because~$g$ is compatible with the structure
+morphisms~$U_\TT \to A^\sim$ and~$U_\TT \to U_\TT|T$, the morphism~$g$ is determined
+by~$f$.
+\end{proof}
+
+\begin{lemma}\label{lemma:fp-double-dual}
+Let~$A$ be a finitely presented~$\TT$-model. Then the canonical morphism
+\[ A^\sim \longrightarrow (U_\TT)^{\Spec(A^\sim)} \]
+is an isomorphism of~$U_\TT$-algebras.
+\end{lemma}
+
+\begin{proof}By Lemma~\ref{lemma:spec-sim-representable}, the
+functor~$\Spec(A^\sim)$ coincides with~$yA$. Since by assumption~$A$ is
+contained in the site defining~$\Set[\TT]$, the exponential~$(U_\TT)^{yA}$
+coincides with~$U_\TT|A$ (notation as in the proof in
+Lemma~\ref{lemma:spec-sim-representable}), that is, the~$U_\TT$-algebra~$A^\sim$.
+\end{proof}
+
+\begin{cor}Let~$A$ and~$B$ be~$\TT$-models. Assume that~$B$ is finitely
+presented. Then the canonical morphism
 \[ \Hom_{U_\TT}(A^\sim, B^\sim) \longrightarrow \Spec(A^\sim)^{\Spec(B^\sim)}
@@
 is an isomorphism.
-\end{lemma}
+\end{cor}
+
+\begin{proof}We have the chain of isomorphisms
+\begin{align*}
+  \Spec(A^\sim)^{\Spec(B^\sim)} &=
+  {[A^\sim,U_\TT]_{U_\TT}}^{\Spec(B^\sim)} \cong
+  [\Spec(B^\sim) \times A^\sim, U_\TT]_{U_\TT} \\
+  &\cong
+  [A^\sim, U_\TT^{\Spec(B^\sim)}]_{U_\TT} \cong
+  [A^\sim, B^\sim],
+\end{align*}
+where the final isomorphism is by Lemma~\ref{lemma:fp-double-dual}.
+\end{proof}
 

### 2019-03-13 paper-qcoh.tex
More work on the higher-order case

@@
 
-\begin{thm}\label{thm:higher-order-nullstellensatz}
+% XXX introduction
+
+By \emph{geometric higher-order logic} we mean the extension of geometric
+logic where we are allowed to form, in addition to the basic sorts supplied by
+a given signature: finite limits of sorts; set-indexed colimits of sorts; and
+powers of sorts. These derived sorts come with respective term constructors
+(tuple formers, coprojections, set comprehension) and the usual rules governing
+these constructors.
+
+A \emph{geometric higher-order formula} is a formula built from equality and
+relation symbols by the logical
+connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$
+and by arbitrary set-indexed disjunctions~$\bigvee$. The existential
+quantification can be over any sort, including the derived sorts. A
+\emph{geometric higher-order sequent} is a sequent of the form~$(\varphi
+\seq{\vec x} \psi)$ where~$\varphi$ and~$\psi$ are geometric higher-order
+formulas and the sorts of the variables~$\vec x$ may be derived sorts.
+
+The truth of geometric higher-order sequents is in general not preserved under
+pullback along geometric morphisms; however for the fragment not containing
+powersorts, this is true. In fact, there is a canonical way of associating to
+any geometric higher-order sequent~$\sigma$ in this fragment a
+family~$(\sigma_i)_i$ of ordinary geometric sequents such that~$\sigma$ is
+provable in geometric higher-order logic if and only if all the
+sequents~$\sigma_i$ are provable in ordinary geometric logic.
+
+\begin{thm}\label{thm:definability}
 Let~$\TT$ be a geometric theory. Let~$x_1\?X_1,\ldots,x_n\?X_n$ be a context over the
@@
 
-% XXX fill in proof
+\begin{proof}We argue internally to~$\Set[\TT]$.
+The canonical map sends an equivalence class~$[\theta]$ to the unique map~$f :
+\{(\vec x) \,|\, \varphi\} \to \{(\vec y) \,|\, \psi \}$ whose graph is given
+by the set~$\{ (\vec x, \vec y) \,|\, \theta \}$.
+
+For verifying surjectivity, let a map~$f : \{(\vec x) \,|\, \varphi\} \to
+\{(\vec y) \,|\, \psi \}$ be given. Then its graph is a subset of~$\vec X
+\times \vec Y$, hence by Theorem~\ref{thm:definability} given by a
+geometric$^\star$ formula~$\theta$. Because~$f$ is a map, this formula is
+functional; and by the Nullstellensatz, it is~$\ul{\TT}/U_\TT$-provably$^\star$ so.
+
+For verifying injectivity, let~$\theta$ and~$\theta'$
+be~$\ul{\TT}/U_\TT$-provably$^\star$ functional formulas which give rise to
+identical maps. Then they also give rise to identical graphs, hence
+are~$\ul{\TT}/U_\TT$-provably$^\star$ equivalent by
+Theorem~\ref{thm:definability}.
+\end{proof}
 

### 2019-03-13 paper-qcoh.tex
Description of internal provability

@@
 
-\section{Background on \ldots}
-
-\subsection{the internal language of Grothendieck toposes}
+\section{Background}
+
+\subsection{Background on the internal language of Grothendieck toposes}
 
@@
 
-\subsection{internal geometric theories}
-
-% XXX, especially the notion of provability and
-% provability*
+\subsection{Background on internal geometric theories}
+
+% XXX
 
@@
 formulas (again so); the object of geometric sequents (again so); and, given an
-internal theory~$\TT$ over~$\Sigma$, the object of proof trees of~$\TT$.
-
-\subsection{classifying toposes}
+internal geometric theory~$\TT$ over~$\Sigma$, the object of proof trees of~$\TT$.
+
+From the internal point of view of~$\E$, these objects can be obtained by
+simply carrying out the usual constructions of the set of contexts, the set of
+terms and so on. Having the object of proof trees at hand, we can define
+provability:
+
+\begin{defn}\label{defn:provability}
+Let~$\TT$ be a geometric theory internal to a Grothendieck topos~$\E$.
+Internally to~$\E$, an element~$\sigma$ of the object of geometric sequents is
+\emph{provable modulo~$\TT$} if and only if there is an element of the object
+of proof trees of~$\TT$ which has~$\sigma$ as its conclusion.\end{defn}
+
+\begin{ex}An ordinary geometric theory is the same as a geometric theory
+internal to the topos~$\Set$. A geometric sequent is provable in the ordinary
+sense if and only if, from the internal point of view of~$\Set$, it is provable
+in the sense of Definition~\ref{defn:provability}.\end{ex}
+
+Disjunctions appearing in internal geometric formulas may be indexed by
+arbitrary objects of the topos, just like disjunctions appearing in ordinary
+external geometric formulas over an ordinary signature may be indexed by
+arbitrary sets. If~$\E$ is a Grothendieck topos, the object of geometric
+formulas over an internal signature~$\Sigma$ in~$\E$ has an important
+subobject, the subobject of those formulas such that locally, any appearing
+disjunction is indexed by a constant sheaf. Such internal geometric formulas
+will be called \emph{geometric$^\star$ formulas}.
+
+Correspondingly, there is a subobject of the object of proof trees, the object
+of~\emph{proof$^\star$ trees} where any occurring geometric formulas is a
+geometric$^\star$ formula. The shape of such a proof$^\star$ tree is locally
+given by the shape of an ordinary external proof tree.
+
+\begin{defn}\label{defn:provability*}
+Let~$\TT$ be a geometric theory internal to a Grothendieck topos~$\E$.
+Internally to~$\E$, an element~$\sigma$ of the object of geometric$^\star$
+sequents is \emph{provable$^\star$ modulo~$\TT$} if and only if there is an
+element of the object of proof$^\star$ trees of~$\TT$ which has~$\sigma$ as its
+conclusion.\end{defn}
+
+The following lemma shows that for coherent theories, there is no difference
+between provability and provability$^*$.
+
+\begin{lemma}Let~$\E$ be a Grothendieck topos. Let~$\TT$ be a coherent theory
+internal to~$\E$. Let~$\sigma$ be a coherent sequent over the signature
+of~$\TT$. Then the following statements are equivalent.
+\begin{enumerate}
+\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of externally
+finite shape}.$
+\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of arbitrary
+external shape}$, that is~$\E \models \speak{$\sigma$ is provable$^\star$
+modulo~$\TT$}$.
+\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of arbitrary
+internal shape}$, that is~$\E \models \speak{$\sigma$ is provable
+modulo~$\TT$}$.
+\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of internally
+finite shape}.$
+\end{enumerate}
+\end{lemma}
+
+\begin{proof}It is trivial that~(1) implies~(2) implies~(3).
+
+To verify that~(3) implies~(4), we can mimic the usual proof of this fact in
+the internal language of~$\E$: There is a variant of the syntactic
+site of~$\TT$ which is built using only coherent sequents and finitary
+derivability~\cite[Section~1.4]{caramello:tst}. The sheaf topos over this site
+is another model for the classifying topos of~$\TT$, and still validates, like
+any Grothendieck topos, full infinitary logic. Hence, if~$\sigma$
+is~$\TT$-derivable by a proof tree of arbitrary shape, then~$\sigma$ holds in
+this model of the classifying topos. By the analogue of
+Proposition~\ref{prop:classifying-topos} for this model, we obtain
+that~$\sigma$ is~$\TT$-derivable by a proof tree of finite shape.
+
+That~(4) implies~(1) is a routine exercise exploiting that
+\[ \E \models \forall X\_ \speak{$X$ is Kuratowski-finite} \Leftrightarrow
+  \bigvee_{n \geq 0} \exists x_1,\ldots,x_n \? X\_
+  \forall x \? X\_ \bigvee_{i=1}^n x = x_i. \qedhere \]
+\end{proof}
+
+\subsection{Background on classifying toposes}
 
@@
 between provability and provability$^*$.
-
-\begin{lemma}Let~$\E$ be a Grothendieck topos. Let~$\TT$ be a coherent theory
-internal to~$\E$. Let~$\sigma$ be a coherent sequent over the signature
-of~$\TT$. Then the following statements are equivalent.
-\begin{enumerate}
-\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of externally
-finite shape}.$
-\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of arbitrary
-external shape}.$
-\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of arbitrary
-internal shape}.$
-\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of internally
-finite shape}.$
-\end{enumerate}
-\end{lemma}
-
-\begin{proof}It is trivial that~(1) implies~(2) implies~(3).
-
-To verify that~(3) implies~(4), we can mimic the usual proof of this fact in
-the internal language of~$\E$: There is a variant of the syntactic
-site of~$\TT$ which is built using only coherent sequents and finitary
-derivability~\cite[Section~1.4]{caramello:tst}. The sheaf topos over this site
-is another model for the classifying topos of~$\TT$, and still validates, like
-any Grothendieck topos, full infinitary logic. Hence, if~$\sigma$
-is~$\TT$-derivable by a proof tree of arbitrary shape, then~$\sigma$ holds in
-this model of the classifying topos. By the analogue of
-Proposition~\ref{prop:classifying-topos} for this model, we obtain
-that~$\sigma$ is~$\TT$-derivable by a proof tree of finite shape.
-
-That~(4) implies~(1) is a routine exercise exploiting that
-\[ \E \models \forall X\_ \speak{$X$ is Kuratowski-finite} \Leftrightarrow
-  \bigvee_{n \geq 0} \exists x_1,\ldots,x_n \? X\_
-  \forall x \? X\_ \bigvee_{i=1}^n x = x_i. \qedhere \]
-\end{proof}
 

### 2019-03-13 paper-qcoh.tex
Description of T/U_T

@@
 
-The following lemma shows that for coherent theories, there is no difference
-between provability and provability$^*$.
+Given a geometric theory~$\TT$, the main theorems reference a certain internal
+theory~$\ul{\TT}/U_\TT$ internal to~$\Set[\TT]$. This theory is defined as
+follows.
+
+\begin{defn}Let~$\TT$ be a geometric theory. The theory~$\ul{\TT}/U_\TT$ is the
+geometric theory internal to~$\Set[\TT]$ which arises from the pulled-back
+theory~$\ul{\TT}$ by adding additional constant symbols~$e_x$ of the
+appropriate sorts, one for each element~$x : U_\TT$, axioms~$(\top \vdash
+f(e_{x_1},\ldots,e_{x_n}) = e_{f(x_1,\ldots,x_n)})$ for each function
+symbol~$f$ and~$n$-tuple of elements of~$U_\TT$ (of the appropriate sorts), and
+axioms~$(\top \vdash R(e_{x_1},\ldots,e_{x_n}))$ for each relation symbol~$R$
+and~$n$-tuple~$(x_1,\ldots,x_n)$ (of the appropriate sorts) such
+that~$R(x_1,\ldots,x_n)$ holds for~$U_\TT$.\end{defn}
+
+From the point of view of~$\Set[\TT]$, a model of~$\ul{\TT}/U_\TT$ is a model
+of~$\ul{\TT}$ equipped with a~$\ul{\TT}$-homomorphism from~$U_\TT$. In
+particular, the identity~$(U_\TT \to U_\TT)$ is a model of~$\ul{\TT}/U_\TT$.
+This is what we mean when we say that~$U_\TT$ is in a canonical way
+a~$\ul{\TT}/U_\TT$-model.
+
+\begin{ex}Let~$\TT$ be the theory of rings. Then~$\ul{\TT}/U_\TT$ is, from the
+internal point of view of~$\Set[\TT]$, the theory of~$U_\TT$-algebras.\end{ex}
+
+\begin{ex}Let~$\TT$ be a geometric theory. Let~$M$ be a model of~$\TT$ in the
+category of sets. Let~$f : \Set \to \Set[\TT]$ be the corresponding geometric
+morphism. Then~$f^*(\ul{\TT}/U_\TT)$ is the theory of~$M$-algebras
+($\TT$-models equipped with a~$\TT$-homomorphism from~$M$). This is
+because~$f^*\ul{\TT} = \TT$, $f^*U_\TT = M$ and because the construction of the
+theory~$\ul{\TT}/U_\TT$ is geometric.\end{ex}
 
@@
 * Discuss related work, especially: Olivia 6.1.3, ...
-
-* Discuss higher-order case
-
-\begin{defn}Let~$\TT$ be a geometric theory. The theory~$\ul{\TT}/U_\TT$ is the
-geometric theory internal to~$\Set[\TT]$ which arises from~$\ul{\TT}$
-by adding additional constant symbols~$e_x$, one for each
-element~$x : U_\TT$, axioms~$(\top \vdash f(e_{x_1},\ldots,e_{x_n}) =
-e_{f(x_1,\ldots,x_n)})$ for each function symbol~$f$ and~$n$-tuple of elements
-of~$U_\TT$, and axioms~$(\top \vdash R(e_{x_1},\ldots,e_{x_n}))$ for each
-relation symbol~$R$ and~$n$-tuple~$(e_{x_1},\ldots,e_{x_n})$ such
-that~$R(e_{x_1},\ldots,e_{x_n})$ holds for~$U_\TT$.
-\end{defn}
-
-From the point of view of~$\Set[\TT]$, a model of~$\ul{\TT}/U_\TT$ is a model
-of~$\ul{\TT}$ equipped with a~$\ul{\TT}$-homomorphism from~$U_\TT$. In
-particular, the identity~$(U_\TT \to U_\TT)$ is a model of~$\ul{\TT}/U_\TT$.
-We then have:
 

### 2019-03-14 paper-qcoh.tex
Externalization of Set[T][T/U_T]

@@
 theory~$\ul{\TT}/U_\TT$ is geometric.\end{ex}
+
+\begin{rem}From the internal point of view of~$\Set[\TT]$, we can construct the
+classifying topos of~$\ul{\TT}/U_\TT$. Externally, this construction gives rise
+to a bounded topos over~$\Set[\TT]$, hence to a Grothendieck topos. Using for
+instance the technique described
+in~\cite{blechschmidt-hutzler-oldenziel:composition}, one can show that this topos classifies the
+theory of homomorphisms between~$\TT$-models. It can also be obtained as the
+lax pullback~$(\Set[\TT] \Rightarrow_{\Set[\TT]} \Set[\TT])$.\end{rem}
 

### 2019-03-15 paper-qcoh.tex
typo
Follow Emily Riehl (and others) in using the Hiragana "よ" for the Yoneda embedding

@@
 \usepackage{xspace}
+
+% lifted from https://arxiv.org/abs/1506.08870
+\DeclareFontFamily{U}{min}{}
+\DeclareFontShape{U}{min}{m}{n}{<-> udmj30}{}
+\newcommand\yon{\!\text{\usefont{U}{min}{m}{n}\symbol{'210}}\!}
 
@@
 
-% XXX
+% XXX, esp. pullback
 
@@
 
-Writing~$y : \C_\TT \to \Set[\TT]$ for the Yoneda embedding, the \emph{generic
-model}~$U_\TT$ of~$\TT$ interprets a sort~$X$ of~$\TT$ as the sheaf~$y\{x\?X\_ \top\}$,
+Writing~$\yon : \C_\TT \to \Set[\TT]$ for the Yoneda embedding, the \emph{generic
+model}~$U_\TT$ of~$\TT$ interprets a sort~$X$ of~$\TT$ as the sheaf~$\yon\{x\?X\_ \top\}$,
 a function symbol~$f : X_1 \cdots X_n \to Y$ as the morphism given by
 the~$\TT$-provably functional formula~$f(x_1,\ldots,x_n) = y$ and a relation
-symbol~$R \rightarrowtail X_1 \cdots X_n$ by the subobject~$y\{\vec x\_ R(\vec
-x)\} \rightarrowtail y\{\vec x\_ \top\}$.
+symbol~$R \rightarrowtail X_1 \cdots X_n$ by the subobject~$\yon\{\vec x\_ R(\vec
+x)\} \rightarrowtail \yon\{\vec x\_ \top\}$.
 
@@
 
+The resulting formula~$\varphi'$ is a geometric formula over the signature
+of~$\TT/U_\TT(A)$.
 The verification of~$A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$
@@
 \[ A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$ $\bigl((\exists \vec y_k\_
-\theta_k) \seq{[]} \bigvee_k \{ \top \,|\, \text{$(\exists \vec y_k\_ \theta_k)$
+\theta_k) \seq{[]} \bigvee_\ell \{ \top \,|\, \text{$(\exists \vec y_\ell\_ \theta_\ell)$
 holds for~$U_\TT$} \}\bigr)$} \]
@@
 Writing~$A = \{ \vec x\_ \alpha \}$ and~$\sigma = (\varphi \seq{\vec y} \psi)$,
-we have~$\{ \vec x\_ \alpha \} \models \forall y\_ (\varphi \Rightarrow \psi)$,
+we have~$\{ \vec x\_ \alpha \} \models \forall \vec y\_ (\varphi \Rightarrow \psi)$,
 hence~$\{ \vec x, \vec y\_ \alpha \wedge \varphi \} \models \psi$. Thus~$\TT$
 proves~$(\alpha \wedge \varphi \seq{\vec x, \vec y} \psi)$. This proof can be
-pulled back from~$\Set$ to~$\Set[\TT]/A$ to show~$A \models
+pulled back from~$\Set$ to~$\Set[\TT]/\yon A$ to show~$A \models
 \speak{$\ul{\TT}/U_\TT$ proves$^\star$~$(\alpha
@@
 Let~$A$ be a~$\TT$-model. Then~$\Spec(A^\sim)$ coincides
-with~$yA$, the functor~$\Hom_{\Mod{\TT}}(A, \cdot)$.
+with~$\yon A$, the functor~$\Hom_{\Mod{\TT}}(A, \cdot)$.
 \end{lemma}
@@
 \begin{align*}
-  \Spec(A^\sim)(T) &\cong \Hom(yT, \Spec(A^\sim)) =
-  \Hom(yT, [A^\sim,U_\TT]_{U_\TT}) \\
-  &\cong \Hom(yT \times A^\sim, U_\TT)_{\text{$U_\TT$-algebra homomorphism in second
+  \Spec(A^\sim)(T) &\cong \Hom(\yon T, \Spec(A^\sim)) =
+  \Hom(\yon T, [A^\sim,U_\TT]_{U_\TT}) \\
+  &\cong \Hom(\yon T \times A^\sim, U_\TT)_{\text{$U_\TT$-algebra homomorphism in second
   argument}} \\
-  &\cong \Hom_{U_\TT}(A^\sim, (U_\TT)^{yT})
+  &\cong \Hom_{U_\TT}(A^\sim, (U_\TT)^{\yon T})
   \cong \Hom_{U_\TT}(A^\sim, U_\TT|T),
@@
 to~$U_\TT$ (a subobject of the internal Hom~$U_\TT^{A^\sim}$); $\Hom_{U_\TT}$
-denotes the set of~$U_\TT$-algebra homomorphisms; $(U_\TT)^{yT}$ is the object of morphisms
-from~$yT$ to~$U_\TT$; and~$U_\TT|T$ is the functor~$U_\TT(T \times \cdot)$, that is~$S
+denotes the set of~$U_\TT$-algebra homomorphisms; $(U_\TT)^{\yon T}$ is the object of morphisms
+from~$\yon T$ to~$U_\TT$; and~$U_\TT|T$ is the functor~$U_\TT(T \times \cdot)$, that is~$S
 \mapsto T \otimes S$.
 
-An arbitrary element~$f \in (yA)(T)$, that is an arbitrary~$\TT$-model
+An arbitrary element~$f \in (\yon A)(T)$, that is an arbitrary~$\TT$-model
 homomorphism~$f : A \to T$, induces a~$U_\TT$-algebra homomorphism~$g : A^\sim \to
@@
 \begin{proof}By Lemma~\ref{lemma:spec-sim-representable}, the
-functor~$\Spec(A^\sim)$ coincides with~$yA$. Since by assumption~$A$ is
-contained in the site defining~$\Set[\TT]$, the exponential~$(U_\TT)^{yA}$
+functor~$\Spec(A^\sim)$ coincides with~$\yon A$. Since by assumption~$A$ is
+contained in the site defining~$\Set[\TT]$, the exponential~$(U_\TT)^{\yon A}$
 coincides with~$U_\TT|A$ (notation as in the proof in

### 2019-03-15 paper-qcoh.tex
typo

@@
 
-\begin{prop}Let~$\alpha$ and~$\varphi$ be geometric formulas in a context~$\vec
+\begin{prop}\label{prop:basic-truth}
+Let~$\alpha$ and~$\varphi$ be geometric formulas in a context~$\vec
 x$ over the signature of a geometric theory~$\TT$. Then the following
@@
 \item $\Set[\TT] \models \forall \vec x\_ (\alpha \Rightarrow \varphi)$.
-\item $\{\vec x\_ \alpha\} \models \varphi$.
+\item $\{\vec x\_ \alpha\} \models \varphi$, where the free variables
+in~$\varphi$ are interpreted as their generic values over~$\{\vec
+x\_\alpha\}$, that is the projection maps~$\{ \vec x\_ \alpha \} \to \{
+x_i\?X_i\_ \top \})$.
 \item $\TT$ proves~$(\alpha \seq{\vec x} \varphi)$.
@@
 \vdash_{[]} \alpha)$}, \]
-where the free variables~$\vec x$ occurring in~$\alpha$ are interpreted by
-their generic values over~$\{ \vec x\_ \alpha \}$, that is the projection
-maps~$\{ \vec x\_ \alpha \} \to \{ x_i\?X_i\_ \top \}$.
+where the free variables~$\vec x$ occurring in~$\alpha$ are interpreted as in
+Proposition~\ref{prop:basic-truth}.
 \end{lemma}
@@
 this case we would just have to pass to further coverings, one for each
-occuring symbol, a finite number of times in total.
+occurring symbol, a finite number of times in total.
 
 However, in general, we cannot conclude as easily. Write~$A = \{ \vec x\_
-\alpha \}$. Let~$R$ be a relation symbol of~$\ul{\TT}(A)$ occuring
+\alpha \}$. Let~$R$ be a relation symbol of~$\ul{\TT}(A)$ occurring
 in~$\varphi$. By the explicit description of constant sheaves as sheaves of
@@
 \speak{$\ul{\TT}/U_\TT$ proves$^\star$~$(\top \seq{[]} \alpha)$}$ (where the free
-variables occuring in~$\alpha$ are interpreted as the generic values available
+variables occurring in~$\alpha$ are interpreted as the generic values available
 over~$A$), hence~$A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$~$(\varphi

### 2019-03-15 paper-qcoh.tex
Main theorem in the Horn case

@@
 \item $\affl$ is \emph{synthetically quasicoherent}: For any
-finitely presented~$\affl$-algebra~$A$, the canonical homomorphism~$A \to
+finitely presentable~$\affl$-algebra~$A$, the canonical homomorphism~$A \to
 (\affl)^{\Spec(A)}$, where~$\Spec(A)$ is defined as the set
@@
 
-\begin{lemma}A~$\TT$-model is finitely presentable as an object of the category
+\begin{lemma}\label{lemma:char-fp-models}
+A~$\TT$-model is finitely presentable as an object of the category
 of~$\TT$-models if and only if it is isomorphic to a model of the
@@
 \begin{lemma}\label{lemma:fp-double-dual}
-Let~$A$ be a finitely presented~$\TT$-model. Then the canonical morphism
+Let~$A$ be a finitely presentable~$\TT$-model. Then the canonical morphism
 \[ A^\sim \longrightarrow (U_\TT)^{\Spec(A^\sim)} \]
@@
 \begin{cor}Let~$A$ and~$B$ be~$\TT$-models. Assume that~$B$ is finitely
-presented. Then the canonical morphism
+presentable. Then the canonical morphism
 \[ \Hom_{U_\TT}(A^\sim, B^\sim) \longrightarrow \Spec(A^\sim)^{\Spec(B^\sim)}
@@
 where the final isomorphism is by Lemma~\ref{lemma:fp-double-dual}.
+\end{proof}
+
+% XXX terminology "U_T-algebra". location of the tensor explanation
+
+\begin{thm}The generic~$\TT$-model is \emph{quasicoherent} in the following sense:
+From the point of view of~$\Set[\TT]$, for any finitely
+presentable~$U_\TT$-algebra~$A$ (finitely presented object in the category
+of~$U_\TT$-algebras), the canonical~$U_\TT$-algebra homomorphism
+\[ A \longrightarrow (U_\TT)^{\Spec(A)} \]
+is an isomorphism.
+\end{thm}
+
+\begin{proof}The proof of Lemma~\ref{lemma:char-fp-models} is constructive and
+thus valid in the internal language of~$\Set[\TT]$. Hence we can apply it,
+internally, to the theory~$\ul{\TT}/U_\TT$. XXX
 \end{proof}

### 2019-03-15 paper-qcoh.tex
typo

@@
 of~$\Set[\TT]$ the theory~$\ul{\TT}$ will again be the theory of rings.
-More details will be given in Section~\ref{sect:review}. The corner quotes indicate
+More details will be given in Section~\ref{sect:review-theories}. The corner quotes indicate
 that for sake of readability, the translation into formal language is to be
@@
 subtopos~$\Set[\TT']$ corresponding to a quotient theory~$\TT'$ of~$\TT$, and
-indeed in general it does not. We formulate a substitute as Theorem~XXX,
-substantially broadening the scope of the Nullstellensatz.
+indeed in general it does not. However, there is still a useful substitute,
+which formulate as Theorem~\ref{thm:nullstellensatz-horn}. This substitute
+substantially broadens the scope of the Nullstellensatz.
 
@@
 scheme~\eqref{eq:nullstellensatz}.
+% XXX be more precise about "scheme"
 \end{enumerate}
@@
 
-Theorem~\ref{thm:characterization0} characterizes which first-order formulas
-hold for the generic model. We could of course wish for a more explicit
+Theorem~\ref{thm:characterization0} characterizes the first-order formulas
+which hold for the generic model. We could of course wish for a more explicit
 characterization; but since even the characterization of geometric sequents
@@
 characterization (``a first-order formula holds for~$U_\TT$ iff it is provable
-in~$\TT'$, where~$\TT'$ is the first-order theory whose set of axioms is the
+modulo~$\TT'$, where~$\TT'$ is the first-order theory whose set of axioms is the
 set of first-order formulas satisfied by~$U_\TT$'') and the (incorrect)
@@
 \subsection{Background on internal geometric theories}
+\label{sect:review-theories}
 
@@
 
-Given an internal theory~$\Sigma$ internal to a Grothendieck topos~$\E$ (or
+Given an internal signature~$\Sigma$ internal to a Grothendieck topos~$\E$ (or
 elementary topos with a natural numbers object), we can successively build the
@@
 terms and so on. Having the object of proof trees at hand, we can define
-provability:
+the notion of provability:
 
@@
 Correspondingly, there is a subobject of the object of proof trees, the object
-of~\emph{proof$^\star$ trees} where any occurring geometric formulas is a
+of~\emph{proof$^\star$ trees} where any occurring geometric formula is a
 geometric$^\star$ formula. The shape of such a proof$^\star$ tree is locally
@@
 this model of the classifying topos. By the analogue of
-Proposition~\ref{prop:classifying-topos} for this model, we obtain
+Proposition~\ref{prop:basic-truth} for this model, we obtain
 that~$\sigma$ is~$\TT$-derivable by a proof tree of finite shape.
@@
 \subsection{Background on classifying toposes}
+% XXX move up to avoid future reference
 
@@
 \section{Proofs of the main theorems}
+\label{sect:main}
 
@@
 
-\begin{defn}Let~$\TT$ be a geometric theory. The theory~$\ul{\TT}/U_\TT$ is the
+\begin{defn}\label{defn:tu}
+Let~$\TT$ be a geometric theory. The theory~$\ul{\TT}/U_\TT$ is the
 geometric theory internal to~$\Set[\TT]$ which arises from the pulled-back
@@
 \begin{proof}The direction~$(1) \Rightarrow (2)$ is immediate, as the internal
-language of~$\Set[\TT]$ validates first-order intuitionistic logic and the
+language of~$\Set[\TT]$ validates first-order intuitionistic logic and as it
+validates the
 axiom scheme~\eqref{eq:nullstellensatz} by Theorem~\ref{thm:nullstellensatz}.
@@
 We omit the required verifications and only remark that while the same
-construction could be carried out if~$\TT$ is a general geometric theory, the
+construction could be carried out if~$\TT$ was a general geometric theory, the
 resulting object would in general not be a model of~$\TT$.
@@
 \item The theory~$\TT$ proves~$\sigma$.
-\item In~$\TT\langle x_1,\ldots,x_k | \varphi_1,\ldots,\varphi_n \rangle$, the
+\item In~$\TT\langle x_1,\ldots,x_k \,|\, \varphi_1,\ldots,\varphi_n \rangle$, the
 propositions~$\psi_1,\ldots,\psi_m$ hold for
@@
 
-\begin{proof}An~$U_\TT$-algebra homomorphism~$\alpha : A^\sim \to M$ yields
+\begin{proof}A~$U_\TT$-algebra homomorphism~$\alpha : A^\sim \to M$ yields
 the~$\TT$-model homo\-mor\-phism~$\alpha_0 : A \to M(0) = \Gamma(M)$, where~$0$ is the
 initial~$\TT$-model. Conversely, a~$\TT$-model homomorphism~$\beta : A \to
-\Gamma(M)$ yields an~$U_\TT$-algebra homomorphism by summing~$A \to M(0) \to
+\Gamma(M)$ yields a~$U_\TT$-algebra homomorphism by summing~$A \to M(0) \to
 M(T)$ with the structure morphism~$T = U_\TT(T) \to

### 2019-03-15 paper-qcoh.tex
Proof of main theorem in the Horn case

@@
 thus valid in the internal language of~$\Set[\TT]$. Hence we can apply it,
-internally, to the theory~$\ul{\TT}/U_\TT$. XXX
+internally, to the theory~$\ul{\TT}/U_\TT$ to deduce that a~$U_\TT$-algebra~$A$
+is finitely presentable if and only if it is isomorphic to a~$U_\TT$-algebra of
+the form~$(\ul{\TT}/U_\TT)\langle X | R \rangle$ with~$X$ Bishop-finite and~$R$
+Kuratowski-finite.
+
+We therefore have to verify the following internal
+statement: \emph{For any number~$n$, for any sorts~$X_1,\ldots,X_n$
+of~$\ul{\TT}/U_\TT$, for any number~$m$, for any atomic
+propositions~$R_1,\ldots,R_m$ over the signature of~$\ul{\TT}/U_\TT$ extended
+by constants~$e_1 \? X_1, \ldots, e_n \? X_n$, the canonical map~$A \to
+(U_\TT)^{\Spec(A)}$ where~$A \defeq (\ul{\TT}/U_\TT)\langle
+e_1\?X_1,\ldots,e_n\?X_n \,|\, R_1,\ldots,R_m \rangle$ is an isomorphism.}
+
+Following the Kripke--Joyal translation of this statement, let a stage~$T \in
+\Mod{\TT}_\mathrm{fp}$,~$T$-elements~$X_1,\ldots,X_n$ of the object of sorts of
+the signature of~$\ul{\TT}/U_\TT$ (that is the constant sheaf on the set of
+sorts of~$\TT$), and~$T$-elements~$R_1,\ldots,R_m$ of the object of atomic
+propositions over the signature of~$\ul{\TT}/U_\TT$ be given. By passing to a
+covering of~$T$, we may assume that the~$X_i$ are given by sorts of~$\TT$ and
+that the~$R_j$ are given by atomic propositions over the signature
+of~$\TT/U_\TT(T)$.
+
+Since the slice~$\Set[\TT]/\yon T$ is equivalent to~$\Set[\TT/T]$,
+hence again the classifying topos of a Horn theory, we may without loss of
+generality assume that~$T$ is the initial~$\TT$-model.
+
+In this case the claim follows from Lemma~\ref{lemma:fp-double-dual}, since
+the result of constructing, internally to~$\Set[\TT]$, the
+model~$(\ul{\TT}/U_\TT)\langle
+e_1\?X_1,\ldots,e_n\?X_n \,|\, R_1,\ldots,R_m \rangle$ coincides with
+the~$U_\TT$-algebra~$(\TT\langle
+e_1\?X_1,\ldots,e_n\?X_n \,|\, R_1,\ldots,R_m \rangle)^\sim$.
 \end{proof}

### 2019-03-15 paper-qcoh.tex
Proof of internal definability

@@
 \cdots \times X_n) \]
-sending the equivalence class of a geometric$^\star$ formula~$\varphi$ over the
-signature of~$\ul{\TT}/U_\TT$ to the subset~$\{ (x_1,\ldots,x_n) \,|\, \varphi
+sending, internally speaking, the equivalence class of a geometric$^\star$ formula~$\varphi$ over the
+signature of~$\ul{\TT}/U_\TT$ in the context~$\vec x$ to the subset~$\{ (x_1,\ldots,x_n) \,|\, \varphi
 \}$ is an isomorphism.
@@
 
-An immediate corollary of Theorem~\ref{thm:definability} is the
-Nullstellensatz of Theorem~\ref{thm:nullstellensatz}. Indeed, arguing
-internally to~$\Set[\TT]$, let~$\sigma = (\varphi \seq{\vec x} \psi)$ be a
-geometric$^\star$ sequent over the signature of~$\ul{\TT}/U_\TT$. Assume
-that~$\sigma$ holds for~$U_\TT$. Then the subsets~$\{ (\vec x) \,|\, \varphi
-\}$ and~$\{ (\vec x) \,|\, \varphi \wedge \psi \}$ are equal. Hence, by
-Theorem~\ref{thm:definability}, the formulas~$\varphi$
-and~$\varphi \wedge \psi$ are provably$^\star$ equivalent.
-Thus~$\ul{\TT}/U_\TT$ proves$^\star$~$(\varphi \seq{\vec x} \psi)$.
+\begin{proof}Injectivity is by the Nullstellensatz of
+Theorem~\ref{thm:nullstellensatz}. Surjectivity is by the definability
+result~\cite[Theorem~2.2]{caramello:definability}, exploiting that the internal
+statement localizes well by Lemma~\ref{lemma:truth-to-provability}.
+% XXX check reference number
+\end{proof}
+
+% An immediate corollary of Theorem~\ref{thm:definability} is the
+% Nullstellensatz of Theorem~\ref{thm:nullstellensatz}. Indeed, arguing
+% internally to~$\Set[\TT]$, let~$\sigma = (\varphi \seq{\vec x} \psi)$ be a
+% geometric$^\star$ sequent over the signature of~$\ul{\TT}/U_\TT$. Assume
+% that~$\sigma$ holds for~$U_\TT$. Then the subsets~$\{ (\vec x) \,|\, \varphi
+% \}$ and~$\{ (\vec x) \,|\, \varphi \wedge \psi \}$ are equal. Hence, by
+% Theorem~\ref{thm:definability}, the formulas~$\varphi$
+% and~$\varphi \wedge \psi$ are provably$^\star$ equivalent.
+% Thus~$\ul{\TT}/U_\TT$ proves$^\star$~$(\varphi \seq{\vec x} \psi)$.
 
@@
 
-% XXX fill in proof
+\begin{proof}The implication~$(2) \Rightarrow (1)$ is immediate since~$U_\TT$
+is a model of~$\TT/U_\TT$. XXX
+\end{proof}
 

### 2019-03-16 paper-qcoh.tex
typo

@@
   geometric theory, useful as a source of nongeometric sequents validated by
-  the generic model, and characterize the first-order formulas validated by the
-  genetric model. XXX
+  the generic model, and characterize the first-order and higher-order formulas validated by the
+  genetric model.
 \end{abstract}
@@
 \paragraph{Generic models} Let~$\TT$ be a geometric theory, such as the theory
-of rings, local rings or intervals. We follow Caramello's
+of rings, of local rings or of intervals. We follow Caramello's
 terminology~\cite{caramello:tst} to mean by \emph{geometric theory} a system
@@
 symbols by the logical connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$
-and by arbitrary set-indexed disjunctions~$\bigvee$). By \emph{first-order
+and by arbitrary set-indexed disjunctions~$\bigvee$). By (infinitary) \emph{first-order
 formula} we will mean a formula which may contain, additionally to the
@@
 \end{enumerate}
-One could argue that it is this model which mathematicians
-implicitly refer to when they utter the phrase ``Let~$M$ be a~$\TT$-model.''. It can
+One could argue that it is this model which we
+implicitly refer to when we utter the phrase ``Let~$M$ be a~$\TT$-model.''. It can
 typically not be realized as a set-theoretic model, consisting of a set for
@@
 axioms of~$\TT$, rendering it a model, the \emph{internal language} of~$\E$ is
-used, roughly reviewed in Section~\ref{sect:review} below. We write~``$\E
+used, roughly reviewed in Section~\ref{sect:review-language} below. We write~``$\E
 \models \alpha$'' to mean that a formula~$\alpha$ holds from the internal
@@
 hypothesis. This principle holds for instance in the case that~$A$ is an
-algebraically closed field and that~$g$ is the unit polynomial. We will see
+algebraically closed field and~$g$ is the unit polynomial. We will see
 below that it is also
@@
     (\forall x \? U_\TT\_ f(x) = 0 \Rightarrow g(x) = 0) \Longleftrightarrow
-      \exists h \in U_\TT[X]\_ g = hf\bigr). \]
+      \exists h \? U_\TT[X]\_ g = hf\bigr). \]
 
@@
 indeed in general it does not. However, there is still a useful substitute,
-which formulate as Theorem~\ref{thm:nullstellensatz-horn}. This substitute
+which we formulate as Theorem~\ref{thm:nullstellensatz-horn}. This substitute
 substantially broadens the scope of the Nullstellensatz.
@@
 \item The formula~$\alpha$ is provable in first-order intuitionistic logic
-modulo the axioms of~$\TT$ and the additional axiom
-scheme~\eqref{eq:nullstellensatz}.
-% XXX be more precise about "scheme"
+modulo the axioms of~$\TT$ and the additional axiom~\eqref{eq:nullstellensatz}.
 \end{enumerate}
@@
 
+% XXX outline and higher-order
+
 \paragraph{Acknowledgments} XXX
@@
 \subsection{Background on the internal language of Grothendieck toposes}
+\label{sect:review-language}
 
@@
 
-\begin{proof}The direction~$(1) \Rightarrow (2)$ is immediate, as the internal
+\begin{proof}The direction~$(2) \Rightarrow (1)$ is immediate, as the internal
 language of~$\Set[\TT]$ validates first-order intuitionistic logic and as it
@@
 \begin{lemma}The functor~$(\cdot)^\sim$ from~$\TT$-models to~$\ul{\TT}/U_\TT$-models
-in~$\Set[\TT]$ is left adjoint to functor~$\Gamma = \Hom(1, \cdot)$ computing
+in~$\Set[\TT]$ is left adjoint to the functor~$\Gamma = \Hom(1, \cdot)$ computing
 global elements.
@@
 
-% XXX fill in proof
+\begin{proof}Theorem~\ref{thm:definability} on the semantic side and the axiom
+scheme on the syntactic side allow us to replace any mention of a
+powersort~$P(X)$ in~$\alpha$
+by~$\mathrm{Form}^\star_{x\!\?\!X}(\ul{\TT}/U_\TT)/({\dashv\vdash_x})$. Then we
+can argue as in the proof of Theorem~\ref{thm:characterization}, noting that
+the axiom scheme indeed entails the axiom scheme posited by
+Theorem~\ref{thm:characterization}.\end{proof}
+
+% XXX check
+
+%In einem ersten Schritt schreiben wir alle Existenzquantifikationen und freie Variablen
+%über Potenzsorten um. Aus "exists A : P(X)" wird "exists φ :
+%Form*(T/U_T)_{x:X}/-||-". (Dabei kann X immer noch eine abgeleitete Sorte sein!) Aus
+%der freien Variable "A : P(X)" wird analog eine freie Variable.
+%
+%Entsprechend verändern wir den Rumpf. Aus "x0 ∈ A" wird φ[x0/x].
+%
+%Dann sind wir fertig mit dem Nullstellensatz für extended geometric logic.
+%
+%Noch klären:
+%* P(X) x Y
+%* Form* Teil von extended geometric logic?
+%* Beweisbarkeit modulo extended geometric logic ist geometrische Formel?
 
 \section{Applications}
+
+% XXX
 
@@
 
-* Cite https://docplayer.net/55796150-Intuitionistic-algebra-some-recent-developments-in-topos-theory.html
+* XXX Cite https://docplayer.net/55796150-Intuitionistic-algebra-some-recent-developments-in-topos-theory.html
   (in general and for internal theories)
 
-* Discuss related work, especially: Olivia 6.1.3, ...
-
-% We would not expect the statement
-% \[ \text{``for any geometric sequent~$\sigma$, if~$\sigma$ holds for~$A$
-% then the theory of rings proves~$\sigma$''} \]
-% to hold for an ordinary ring~$A$ in the topos of sets.
-%
-% Let~$M$ be a~$\TT$-model. We then trivially have
-% signature of~$\TT$. While we always have the implication
-% \[ \text{$\TT$ proves~$\sigma$} \quad\Longrightarrow\quad
-%   \text{$\sigma$ holds for~$M$}, \]
-% the converse is of course typically false,
-%
-% Furthermore, in a first introduction, it is useful to
-% restrict to an instructive special case, such as the theory of (commutative) rings.
-%
-% A simple version of the classical Nullstellensatz, valid for algebraically
-% closed fields, states: Let~$f \in A[X]$ be a polynomial. If
-% \[ \forall x \in A\_ (f(x) = 0 \in A \Rightarrow 1 = 0 \in A), \]
-% then there is a polynomial~$g \in A[X]$ such that~$gf = 1$. This polynomial~$g$
-% can be regarded as an ``algebraic certificate'' that the equation~$f(x) = 0$
-% only has a solution in the trivial case~$1 = 0$. The conclusion can also be phrased as
-% follows: In the free~$A$-algebra on one generator modulo the relation~$f(X) =
-% 0$, the relation~$1 = 0$ holds.
-%
-% Let~$A$ be a specific ring. Let~$f,g \in \ZZ[X]$ be
-% polynomials. While we always have the implication
-% \[ g([X]) = 0 \in \ZZ[X]/(f) \quad\Longrightarrow\quad
-%   \forall x \in A\_ (f(x) = 0 \Rightarrow g(x) = 0), \]
-% the converse is typically false: Just that~$g(x) = 0$ holds for every zero
-% of~$f$ in~$A$ does not imply that~$g([X]) = 0$ holds for the generic zero
-% of~$f$ in the free ring on one generator~$X$ modulo the relation~$f(X) = 0$.
-%
-%
-% the case that~$\TT$ is a Horn theory (where the axioms are of the
-% form~$(\phi \seq{\vec x} \psi)$ with the formulas~$\phi$ and~$\psi$ being
-% finite conjunctions of atomic propositions (equality and the given relations)),
-% such that we can switch from a logical terminology to a more algebraic one.
-% In the Horn case, for any set~$X$ and any set~$R$ of atomic propositions the
-% term algebra construction yields~$\TT\langle X | R\rangle$, the
-% free~$\TT$-model generated by~$X$ modulo the relations~$R$. We will also abuse
-% notation to pretend that~$\TT$ is single-sorted.
-%
-% Let~$M$ be a specific~$\TT$-model and assume that for
-% all~$n$-tuples~$x = (x_1,\ldots,x_n)$ of elements of~$M$, if some set~$R_1$ of
-% atomic propositions is satisfied for~$x$ then so is some set~$R_2$. In this
-% case it is unreasonable to expect that~$R_2$ holds for the~$n$-tuple of
-% generators of~$\TT\langle e_1,\ldots,e_n | R_1\rangle$.
-%
-% let~$R_1$ and~$R_2$ be sets of atomic
-% propositions with~$n$ free variables. Then
-% and assume, for some sets of relations~$R_1$
-% and~$R_2$
-%
-% The term algebra of a Horn theory~$\TT$ is always a model of~$\TT$, such that
-
-* Sei T' subkanonisch für T.
-
-* Sei φ eine Hornformel. Gelte ∇φ. Folgt dann φ?
-
-* Für φ = (x = y) stimmt das, da U separiert ist.
-
-* Für φ = (R(x,y)) stimmt, das ...?
-
-  Was heißt das denn? Habe R \--> U x U. Gehe davon aus, dass ∇φ bedeutet, dass
-  ∇((x,y) ∈ R). Daraus folgt aber (x,y) ∈ R:
-
-  K = { (x,y) | (x,y) ∈ R } ist Subsingleton von U x U und ∇-bewohnt.
-  Also existiert (a,b) ∈ U x U mit ∇... Hilft nicht.
-
-  Aber: K = { (a,b) ∈ R | a = x & b = y } ist Subsingleton von R.
-  Und ∇-bewohnt. Also existiert (a,b) ∈ R mit ∇(a=x & b = y).
-  Wegen Separiertheit folgt a = x & b = y. Fertig. :-)
+* XXX Discuss related work, especially: Olivia 6.1.3, ...
 
@@
 
-* Nichttrivialität des Nullstellensatzes diskutieren (Spezialfall top |- φ
+* XXX Nichttrivialität des Nullstellensatzes diskutieren (Spezialfall top |- φ
   betrachten, dieser ist nämlich schon mehr oder weniger trivial)
 
-* Anmerken, dass der Nullstellensatz für first-order sequents NICHT stimmt
+* XXX Anmerken, dass der Nullstellensatz für first-order sequents NICHT stimmt
   (Beispiel: ¬(x=0) ⇒ (x inv.))
 
-* Anmerken, dass wir in konstruktiver Metalogik arbeiten.
+* XXX Anmerken, dass wir in konstruktiver Metalogik arbeiten.

### 2019-03-16 paper-qcoh.tex
Proof of the higher-order case

@@
 
-By \emph{geometric higher-order logic} we mean the extension of geometric
-logic where we are allowed to form, in addition to the basic sorts supplied by
-a given signature: finite limits of sorts; set-indexed colimits of sorts; and
-powers of sorts. These derived sorts come with respective term constructors
-(tuple formers, coprojections, set comprehension) and the usual rules governing
-these constructors.
-
-A \emph{geometric higher-order formula} is a formula built from equality and
+By \emph{extended geometric logic} we mean the extension of geometric logic
+where we are allowed to form, in addition to the basic sorts supplied by
+a given signature, finite limits of sorts and set-indexed colimits of sorts. By
+(intuitionistic) \emph{higher-order logic}, we mean the further extension where
+we may also form powersorts. These derived sorts come with respective term
+constructors (tuple formers, coprojections, set comprehension) and the usual
+rules governing these constructors.
+
+An \emph{extended geometric formula} is a formula of extended geometric logic built from equality and
 relation symbols by the logical
@@
 and by arbitrary set-indexed disjunctions~$\bigvee$. The existential
-quantification can be over any sort, including the derived sorts. A
-\emph{geometric higher-order sequent} is a sequent of the form~$(\varphi
-\seq{\vec x} \psi)$ where~$\varphi$ and~$\psi$ are geometric higher-order
+quantification can be over any sorts of extended geometric logic, including the derived sorts. An
+\emph{extended geometric sequent} is a sequent of the form~$(\varphi
+\seq{\vec x} \psi)$ where~$\varphi$ and~$\psi$ are extended geometric
 formulas and the sorts of the variables~$\vec x$ may be derived sorts.
 
-The truth of geometric higher-order sequents is in general not preserved under
-pullback along geometric morphisms; however for the fragment not containing
-powersorts, this is true. In fact, there is a canonical way of associating to
-any geometric higher-order sequent~$\sigma$ in this fragment a
-family~$(\sigma_i)_i$ of ordinary geometric sequents such that~$\sigma$ is
-provable in geometric higher-order logic if and only if all the
-sequents~$\sigma_i$ are provable in ordinary geometric logic.
+It is possible to extend the Kripke--Joyal semantics so that higher-order logic
+can be interpreted in any Grothendieck topos. The truth of a higher-order
+sequent~$(\varphi \seq{\vec x} \psi)$ is in general not preserved under
+pullback along geometric morphisms, even if~$\varphi$ and~$\psi$ do not
+contain~${\forall}$ and~$\Rightarrow$, since powerobjects are in general not
+preserved under pullback. However, as can be deduced from the following lemma,
+the truth of extended geometric sequents is preserved; as is folklore, extended
+geometric logic is just a thin layer over ordinary geometric logic.
+
+\begin{lemma}\label{lemma:extended-conservative}
+Let~$\sigma$ be an extended geometric sequent over the signature
+of a geometric theory~$\TT$. Then there is a set-indexed family~$(\sigma_i)_i$ of ordinary
+geometric sequents over the same signature, so that~$\sigma$ is provable in
+extended geometric logic if and only if all the sequents~$\sigma_i$ are
+provable in ordinary geometric logic.\end{lemma}
+
+\begin{proof}Any existential quantification of the form~``$\exists p \? X
+\times Y$'' can be replaced by the string~``$\exists x \? X\_ \exists y \?
+Y$'', and similarly for free variables of product sorts appearing in the
+context of~$\sigma$. In a similar vein more general finite limits are treated.
+
+An existential quantification of the form~``$\exists x \? \coprod_i X_i$'' can
+be replaced by the string~``$\bigvee_i \exists x \? X_i$''.
+
+Finally, for any occurrence of a free variable~$x \?
+\coprod_i X_i$ in the context of~$\sigma$, we can replace~$\sigma$ by the
+family of sequents~$(\sigma_i)_i$, where the sequent~$\sigma_i$ is the same
+as~$\sigma$ only that the free variable~$x$ is changed to be of sort~$X_i$ (and
+the corresponding change in the consequent and the antecedent is applied,
+applying the appropriate coprojection).
+
+After carrying out these steps, the free variables are only of the basic sorts
+supplied by the signature of~$\TT$ and existential quantifications only
+range over the basic sorts. However, in the consequents and antecedents, still
+tuple formers and coprojections may appear. These can be replaced as suggested
+by the rules governing these. For instance, an equation~``$\langle x,y \rangle
+= \langle x',y' \rangle$'' can be replaced by the conjunction~``$x = x' \wedge
+y = y'$'', and an equation~``$\iota_i(x) = \iota_j(y)$'' (where~$\iota_i$
+and~$\iota_j$ are coprojections associated with coproduct sorts) can be
+replaced by the subsingleton-indexed disjunction~``$\bigvee\{ x = y \,|\, i = j \}$''.
+\end{proof}
 
@@
 Let~$\TT$ be a geometric theory. Then, internally to~$\Set[\TT]$, for any
-higher-order geometric$^\star$ sequent~$\sigma$ over the signature
+extended geometric$^\star$ sequent~$\sigma$ over the signature
 of~$\ul{\TT}/U$, the following statements are equivalent:
@@
 \item The sequent~$\sigma$ holds for~$U_\TT$.
-\item The sequent~$\sigma$ is provable$^\star$ modulo~$\TT/U_\TT$ in geometric
-higher-order logic.
+\item The sequent~$\sigma$ is provable$^\star$ modulo~$\TT/U_\TT$ in extended
+geometric logic.
 \end{enumerate}
@@
 
-\begin{proof}The implication~$(2) \Rightarrow (1)$ is immediate since~$U_\TT$
-is a model of~$\TT/U_\TT$. XXX
+\begin{proof}The implication~$(2) \Rightarrow (1)$ is
+immediate since~$U_\TT$ is a model of~$\TT/U_\TT$. The converse direction is by
+Lemma~\ref{lemma:extended-conservative}, which holds internally in~$\Set[\TT]$
+as the proof we supplied is constructive, and the Nullstellensatz for ordinary
+geometric logic of Theorem~\ref{thm:nullstellensatz}.
 \end{proof}

### 2019-03-16 paper-qcoh.tex
typo
Two non-strengthening remarks

@@
 
+\begin{rem}Theorem~\ref{thm:nullstellensatz} cannot be strengthened to
+arbitrary geometric sequents. For instance, in the case that~$\TT$ is the
+theory of rings, the internal geometric sequent~$(\top \seq{x\?U_\TT}
+\bigvee_{a\?U_\TT} (x = e_a))$ trivially holds for~$U_\TT$. However, it is not
+provable modulo~$\TT/U_\TT$, as for instance the polynomial algebra~$U_\TT[X]$
+does not validate it.\end{rem}
+
 \begin{thm}\label{thm:nullstellensatz-horn}
@@
 
+Theorem~\ref{thm:nullstellensatz} cannot be strengthened to arbitrary
+first-order (or first-order$^\star$ or even just coherent) formulas in place of
+geometric$^\star$ sequents. For instance, in the case that~$\TT$ is the theory
+of local rings, the generic model~$U_\TT$ validates the coherent
+formula~$\speak{any element which is not zero is invertible}$, but~$\TT/U_\TT$
+does not prove this fact, as it is for instance not validated by the polynomial
+algebra~$U_\TT[X]$. However, Theorem~\ref{thm:nullstellensatz} still plays an
+important role in understanding first-order formulas:
+
 \begin{thm}\label{thm:characterization}
@@
 scheme~\eqref{eq:nullstellensatz}.
+% XXX be more explicit about that being a scheme
 \end{enumerate}
@@
   \Spec(A^\sim)^{\Spec(B^\sim)} &=
-  {[A^\sim,U_\TT]_{U_\TT}}^{\Spec(B^\sim)} \cong
+  ([A^\sim,U_\TT]_{U_\TT})^{\Spec(B^\sim)} \cong
   [\Spec(B^\sim) \times A^\sim, U_\TT]_{U_\TT} \\
@@
 
-* XXX Anmerken, dass der Nullstellensatz für first-order sequents NICHT stimmt
-  (Beispiel: ¬(x=0) ⇒ (x inv.))
-
 * XXX Anmerken, dass wir in konstruktiver Metalogik arbeiten.

### 2019-03-16 paper-qcoh.tex
Outline

@@
 
-% XXX outline and higher-order
+\paragraph{Outline} In Section~\ref{sect:review}, we review background on the
+internal language of toposes, internal geometric theories and classifying
+toposes. Section~\ref{sect:main} contains proofs of the main theorems in the
+full generality of geometric theories. Restricting to Horn theories allows for
+a treatment which is more algebraic and less logical in flavor. For the benefit
+of readers with a more algebraic background, we include a mostly self-contained
+account of the Horn case as Section~\ref{sect:horn}. We generalize our main
+theorems to the higher-order case in Section~\ref{sect:higher-order} and
+conclude with applications in Section~\ref{sect:applications}.
+
+Throughout we work in a constructive metatheory, to allow our results to be
+interpreted internally to toposes.
+\medskip
 
@@
 \section{Background}
+\label{sect:review}
 
@@
 \section{The special case of Horn theories}
+\label{sect:horn}
 
@@
 \section{The generalization to the higher-order case}
+\label{sect:higher-order}
 
@@
 \section{Applications}
+\label{sect:applications}
 
@@
   betrachten, dieser ist nämlich schon mehr oder weniger trivial)
-
-* XXX Anmerken, dass wir in konstruktiver Metalogik arbeiten.

### 2019-03-17 paper-qcoh.tex
Internal signatures

@@
 
-% XXX, esp. pullback
+The notions of signatures, geometric theories and classifying toposes can be
+relativized to the internal world of arbitrary toposes with natural numbers
+objects. Basics on internal signatures, internal geometric
+theories and internal classifying toposes are
+folklore~\cite[p.~334]{wraith:intuitionistic-algebra}; a careful treatment is
+due to Shawn Henry~\cite{henry:phd}.
+
+Briefly, an internal signature~$\Sigma$ internal to a topos~$\E$ consists of an
+object of sorts, an object of function symbols, an object of relation symbols,
+and various morphisms indicating the sorts involved with the function and
+relation symbols.
 
@@
 propositions (again equipped with such a morphism); the object of geometric
-formulas (again so); the object of geometric sequents (again so); and, given an
-internal geometric theory~$\TT$ over~$\Sigma$, the object of proof trees of~$\TT$.
+formulas (again so); the object of geometric sequents (again so). A internal
+geometric theory~$\TT$ over~$\Sigma$ is then given by a subobject of the object
+of geometric sequents, interpreted as the object of axioms of~$\TT$. Given such
+an internal theory~$\TT$, we can build the object of proof trees of~$\TT$.
 
@@
 
+\begin{ex}An ordinary geometric theory~$\TT$ over an ordinary
+signature~$\Sigma$ can be pulled back along a geometric morphism~$\E \to \Set$
+to yield an internal geometric theory~$\ul{\TT}$ over the internal
+signature~$\ul{\Sigma}$ in~$\E$. The object of sorts of~$\ul{\Sigma}$ is the pullback of
+the set of sorts of~$\Sigma$, and so on.\end{ex}
+
 Disjunctions appearing in internal geometric formulas may be indexed by
@@
 external geometric formulas over an ordinary signature may be indexed by
-arbitrary sets. If~$\E$ is a Grothendieck topos, the object of geometric
-formulas over an internal signature~$\Sigma$ in~$\E$ has an important
+arbitrary sets. If~$\Sigma$ is an internal signature in a Grothendieck topos~$\E$, the object of geometric
+formulas over~$\Sigma$ has an important
 subobject, the subobject of those formulas such that locally, any appearing

### 2019-03-17 paper-qcoh.tex
Move to avoid forward reference

@@
 \paragraph{Outline} In Section~\ref{sect:review}, we review background on the
-internal language of toposes, internal geometric theories and classifying
-toposes. Section~\ref{sect:main} contains proofs of the main theorems in the
+internal language of toposes, classifying toposes and internal geometric theories.
+Section~\ref{sect:main} contains proofs of the main theorems in the
 full generality of geometric theories. Restricting to Horn theories allows for
@@
 % XXX
+
+\subsection{Background on classifying toposes}
+\label{sect:review-classifying-toposes}
+
+\begin{defn}The \emph{syntactic site}~$\C_\TT$ of a geometric theory~$\TT$ has:
+\begin{enumerate}
+\item as objects ``geometric formulas in contexts''~$\{x_1\?X_1,\ldots,x_n\?X_n\_
+\varphi\}$ where~$\varphi$ is a geometric formula over the signature of~$\TT$
+in the displayed context;
+\item as set of morphisms~$\Hom_{\C_\TT}(\{\vec x\_
+\varphi\}, \{\vec y\_ \psi\})$ the set of formulas~$\theta$ in the
+context~$\vec x, \vec y$ which are~$\TT$-provably functional,
+modulo~$\TT$-provable equivalence of such formulas;
+\item as covering families those families~$(\{\vec x_i\_ \varphi_i\} \xrightarrow{\theta_i}
+\{\vec y\_ \psi\})_i$ for which~$\TT$ proves~$(\psi \seq{\vec y} \bigvee_i \exists \vec x_i\_
+\theta_i)$.
+\end{enumerate}
+\end{defn}
+
+\begin{defn}The \emph{classifying topos}~$\Set[\TT]$ of a geometric
+theory~$\TT$ is the topos of set-valued sheaves on~$\C_\TT$.\end{defn}
+
+Writing~$\yon : \C_\TT \to \Set[\TT]$ for the Yoneda embedding, the \emph{generic
+model}~$U_\TT$ of~$\TT$ interprets a sort~$X$ of~$\TT$ as the sheaf~$\yon\{x\?X\_ \top\}$,
+a function symbol~$f : X_1 \cdots X_n \to Y$ as the morphism given by
+the~$\TT$-provably functional formula~$f(x_1,\ldots,x_n) = y$ and a relation
+symbol~$R \rightarrowtail X_1 \cdots X_n$ by the subobject~$\yon\{\vec x\_ R(\vec
+x)\} \rightarrowtail \yon\{\vec x\_ \top\}$.
+
+\begin{thm}The generic model is universal in the sense that for any
+Grothendieck topos~$\E$, the functor
+\[ \text{(category of geometric morphisms~$\E \to \Set[\TT]$)} \longrightarrow
+\text{(category of~$\TT$-models in~$\E$)} \]
+given by~$f \mapsto f^*U_\TT$ is an equivalence of categories.
+\end{thm}
+
+\begin{prop}\label{prop:basic-truth}
+Let~$\alpha$ and~$\varphi$ be geometric formulas in a context~$\vec
+x$ over the signature of a geometric theory~$\TT$. Then the following
+statements are equivalent:
+\begin{enumerate}
+\item $\Set[\TT] \models \forall \vec x\_ (\alpha \Rightarrow \varphi)$.
+\item $\{\vec x\_ \alpha\} \models \varphi$, where the free variables
+in~$\varphi$ are interpreted as their generic values over~$\{\vec
+x\_\alpha\}$, that is the projection maps~$\{ \vec x\_ \alpha \} \to \{
+x_i\?X_i\_ \top \})$.
+\item $\TT$ proves~$(\alpha \seq{\vec x} \varphi)$.
+\end{enumerate}
+\end{prop}
+
+\begin{proof}The equivalence~$(1) \Leftrightarrow (2)$ follows immediately by
+unrolling the Kripke--Joyal semantics. The equivalence~$(2) \Leftrightarrow
+(3)$ is by induction on the structure of~$\varphi$.\end{proof}
 
@@
 \end{proof}
-
-\subsection{Background on classifying toposes}
-% XXX move up to avoid future reference
-
-\begin{defn}The \emph{syntactic site}~$\C_\TT$ of a geometric theory~$\TT$ has:
-\begin{enumerate}
-\item as objects ``geometric formulas in contexts''~$\{x_1\?X_1,\ldots,x_n\?X_n\_
-\varphi\}$ where~$\varphi$ is a geometric formula over the signature of~$\TT$
-in the displayed context;
-\item as set of morphisms~$\Hom_{\C_\TT}(\{\vec x\_
-\varphi\}, \{\vec y\_ \psi\})$ the set of formulas~$\theta$ in the
-context~$\vec x, \vec y$ which are~$\TT$-provably functional,
-modulo~$\TT$-provable equivalence of such formulas;
-\item as covering families those families~$(\{\vec x_i\_ \varphi_i\} \xrightarrow{\theta_i}
-\{\vec y\_ \psi\})_i$ for which~$\TT$ proves~$(\psi \seq{\vec y} \bigvee_i \exists \vec x_i\_
-\theta_i)$.
-\end{enumerate}
-\end{defn}
-
-\begin{defn}The \emph{classifying topos}~$\Set[\TT]$ of a geometric
-theory~$\TT$ is the topos of set-valued sheaves on~$\C_\TT$.\end{defn}
-
-Writing~$\yon : \C_\TT \to \Set[\TT]$ for the Yoneda embedding, the \emph{generic
-model}~$U_\TT$ of~$\TT$ interprets a sort~$X$ of~$\TT$ as the sheaf~$\yon\{x\?X\_ \top\}$,
-a function symbol~$f : X_1 \cdots X_n \to Y$ as the morphism given by
-the~$\TT$-provably functional formula~$f(x_1,\ldots,x_n) = y$ and a relation
-symbol~$R \rightarrowtail X_1 \cdots X_n$ by the subobject~$\yon\{\vec x\_ R(\vec
-x)\} \rightarrowtail \yon\{\vec x\_ \top\}$.
-
-\begin{thm}The generic model is universal in the sense that for any
-Grothendieck topos~$\E$, the functor
-\[ \text{(category of geometric morphisms~$\E \to \Set[\TT]$)} \longrightarrow
-\text{(category of~$\TT$-models in~$\E$)} \]
-given by~$f \mapsto f^*U_\TT$ is an equivalence of categories.
-\end{thm}
-
-\begin{prop}\label{prop:basic-truth}
-Let~$\alpha$ and~$\varphi$ be geometric formulas in a context~$\vec
-x$ over the signature of a geometric theory~$\TT$. Then the following
-statements are equivalent:
-\begin{enumerate}
-\item $\Set[\TT] \models \forall \vec x\_ (\alpha \Rightarrow \varphi)$.
-\item $\{\vec x\_ \alpha\} \models \varphi$, where the free variables
-in~$\varphi$ are interpreted as their generic values over~$\{\vec
-x\_\alpha\}$, that is the projection maps~$\{ \vec x\_ \alpha \} \to \{
-x_i\?X_i\_ \top \})$.
-\item $\TT$ proves~$(\alpha \seq{\vec x} \varphi)$.
-\end{enumerate}
-\end{prop}
-
-\begin{proof}The equivalence~$(1) \Leftrightarrow (2)$ follows immediately by
-unrolling the Kripke--Joyal semantics. The equivalence~$(2) \Leftrightarrow
-(3)$ is by induction on the structure of~$\varphi$.\end{proof}
 

### 2019-03-17 paper-qcoh.tex
Axiom schemes vs. axioms

@@
 \renewcommand{\SS}{\mathbb{S}}
-\renewcommand{\P}{\mathcal{P}}
 \renewcommand{\O}{\mathcal{O}}
@@
 Let~$\TT$ be a geometric theory. Then, internally to~$\Set[\TT]$:
-\begin{equation}\tag{$\ddagger$}\label{eq:nullstellensatz}
+\begin{equation}\tag{$\dagger$}\label{eq:nullstellensatz0}
 \text{A geometric$^*$ sequent~$\sigma$ holds for~$U_\TT$ if and only
@@
 
-The statement~\eqref{eq:nullstellensatz} is not a geometric sequent. Therefore
+The statement~\eqref{eq:nullstellensatz0} is not a geometric sequent. Therefore
 it is not to be expected that it passes from~$\Set[\TT]$ to a
@@
 it is the universal such source in the sense that any first-order formula
-which holds for~$U_\TT$ can be deduced from~\eqref{eq:nullstellensatz}:
+which holds for~$U_\TT$ can be deduced from~\eqref{eq:nullstellensatz0}:
 
@@
 \item The formula~$\alpha$ is provable in first-order intuitionistic logic
-modulo the axioms of~$\TT$ and the additional axiom~\eqref{eq:nullstellensatz}.
+modulo the axioms of~$\TT$ and the additional axiom~\eqref{eq:nullstellensatz0}.
 \end{enumerate}
@@
 \item The formula~$\alpha$ is provable in first-order intuitionistic logic
-modulo the axioms of~$\TT$ and the additional axiom
-scheme~\eqref{eq:nullstellensatz}.
-% XXX be more explicit about that being a scheme
+modulo the axioms of~$\TT$ and the additional axioms
+\begin{equation}\label{eq:nullstellensatz}\tag{$\ddagger$}
+  \speak{$\sigma$ holds for~$U_\TT$} \Longrightarrow \speak{$\ul{\TT}/U_\TT$
+  proves$^\star$~$\sigma$}
+\end{equation}
+where~$\sigma$ ranges over the geometric formulas over the signature of~$\TT$.
 \end{enumerate}
 \end{thm}
+
+% XXX handholding
 
@@
 \item The formula~$\alpha$ is provable in higher-order intuitionistic logic
-modulo the axioms of~$\TT$ and the additional axiom scheme~XXX.
+modulo the axioms of~$\TT$ and the additional axiom scheme
+\begin{equation}\label{eq:higher-order-nullstellensatz}\tag{$\P$}
+  \speak{the map~$\mathrm{Form}^\star_{\vec
+  x}(\ul{\TT}/U_\TT)/(\dashv\vdash_{\vec x}) \longrightarrow P(X_1 \times
+  \cdots \times X_n)$ is bijective},
+\end{equation}
+where~$X_1,\ldots,X_n$ is any list of sorts.
 \end{enumerate}
@@
 \begin{proof}Theorem~\ref{thm:definability} on the semantic side and the axiom
-scheme on the syntactic side allow us to replace any mention of a
+scheme~\eqref{eq:higher-order-nullstellensatz} on the syntactic side allow us to replace any mention of a
 powersort~$P(X)$ in~$\alpha$
-by~$\mathrm{Form}^\star_{x\!\?\!X}(\ul{\TT}/U_\TT)/({\dashv\vdash_x})$. Then we
+by~$\mathrm{Form}^\star_{x\!\?\!X}(\ul{\TT}/U_\TT)/({\dashv\vdash_x})$,
+similarly to how the proof of Lemma~\ref{lemma:extended-conservative} compiles
+extended geometric logic to ordinary geometric. Then we
 can argue as in the proof of Theorem~\ref{thm:characterization}, noting that
-the axiom scheme indeed entails the axiom scheme posited by
+the axiom scheme indeed entails the axiom scheme~\eqref{eq:nullstellensatz} posited by
 Theorem~\ref{thm:characterization}.\end{proof}

### 2019-03-17 paper-qcoh.tex
typo

@@
 \begin{enumerate}
-\item The formula~$\alpha$ holds for~$U_\TT$.
+\item The formula~$\alpha$ holds for~$U_\TT$. \smallskip
 \item The formula~$\alpha$ is provable in first-order intuitionistic logic
@@
 
+We use the usual convention of abbreviating~``$x_1\?X_1,\ldots,x_n\?X_n$''
+as~``$\vec x \? \vec X$'' or even just~``$\vec x$''.
+
 \begin{defn}The \emph{syntactic site}~$\C_\TT$ of a geometric theory~$\TT$ has:
@@
 \varphi\}$ where~$\varphi$ is a geometric formula over the signature of~$\TT$
-in the displayed context;
+in the displayed context; \smallskip
 \item as set of morphisms~$\Hom_{\C_\TT}(\{\vec x\_
@@
 context~$\vec x, \vec y$ which are~$\TT$-provably functional,
-modulo~$\TT$-provable equivalence of such formulas;
+modulo~$\TT$-provable equivalence of such formulas; \smallskip
 \item as covering families those families~$(\{\vec x_i\_ \varphi_i\} \xrightarrow{\theta_i}
-\{\vec y\_ \psi\})_i$ for which~$\TT$ proves~$(\psi \seq{\vec y} \bigvee_i \exists \vec x_i\_
+\{\vec y\_ \psi\})_i$ for which~$\TT$ proves the sequent~$(\psi \seq{\vec y} \bigvee_i \exists \vec x_i\_
 \theta_i)$.
@@
 \item $\Set[\TT] \models \forall \vec x\_ (\alpha \Rightarrow \varphi)$.
+\smallskip
 \item $\{\vec x\_ \alpha\} \models \varphi$, where the free variables
@@
 x\_\alpha\}$, that is the projection maps~$\{ \vec x\_ \alpha \} \to \{
-x_i\?X_i\_ \top \})$.
+x_i\?X_i\_ \top \})$. \smallskip
 \item $\TT$ proves~$(\alpha \seq{\vec x} \varphi)$.
@@
 \item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of externally
-finite shape}.$
+finite shape}.$ \smallskip
 \item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of arbitrary
 external shape}$, that is~$\E \models \speak{$\sigma$ is provable$^\star$
-modulo~$\TT$}$.
+modulo~$\TT$}$. \smallskip
 \item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of arbitrary
 internal shape}$, that is~$\E \models \speak{$\sigma$ is provable
-modulo~$\TT$}$.
+modulo~$\TT$}$. \smallskip
 \item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of internally
@@
 
-Given a geometric theory~$\TT$, the main theorems reference a certain internal
+Given a geometric theory~$\TT$ with its generic model~$U_\TT$ in~$\Set[\TT]$,
+the main theorems reference a certain internal
 theory~$\ul{\TT}/U_\TT$ internal to~$\Set[\TT]$. This theory is defined as
@@
 \begin{enumerate}
-\item The sequent~$\sigma$ holds for~$U_\TT$.
+\item The sequent~$\sigma$ holds for~$U_\TT$. \smallskip
 \item The sequent~$\sigma$ is provable$^\star$ modulo~$\ul{\TT}/U_\TT$.
@@
 \begin{enumerate}
-\item The theory~$\TT$ proves~$\sigma$.
+\item The theory~$\TT$ proves~$\sigma$. \smallskip
 \item In~$\TT\langle x_1,\ldots,x_k \,|\, \varphi_1,\ldots,\varphi_n \rangle$, the
@@
 where~$[A^\sim,U_\TT]_{U_\TT}$ is the object of~$U_\TT$-algebra homomorphisms from~$A^\sim$
-to~$U_\TT$ (a subobject of the internal Hom~$U_\TT^{A^\sim}$); $\Hom_{U_\TT}$
+to~$U_\TT$ (a subobject of the internal Hom~$(U_\TT)^{A^\sim}$); $\Hom_{U_\TT}$
 denotes the set of~$U_\TT$-algebra homomorphisms; $(U_\TT)^{\yon T}$ is the object of morphisms
@@
 \begin{enumerate}
-\item The sequent~$\sigma$ holds for~$U_\TT$.
+\item The sequent~$\sigma$ holds for~$U_\TT$. \smallskip
 \item The sequent~$\sigma$ is provable$^\star$ modulo~$\TT/U_\TT$ in extended
@@
 \begin{enumerate}
-\item The formula~$\alpha$ holds for~$U_\TT$.
+\item The formula~$\alpha$ holds for~$U_\TT$. \smallskip
 \item The formula~$\alpha$ is provable in higher-order intuitionistic logic

### 2019-03-17 paper-qcoh.tex
Explanation of abuse of notation

@@
 
-% XXX handholding
-
-\begin{proof}The direction~$(2) \Rightarrow (1)$ is immediate, as the internal
+In stating the axiom scheme~\eqref{eq:nullstellensatz}, we are slightly abusing
+notation. At face value, this axiom scheme would only make sense in the internal
+language of~$\Set[\TT]$, where the generic model~$U_\TT$ actually lives.
+However, here we are interpreting it in a purely syntactic context; when
+writing~``$U_\TT$'', we actually mean the tautologous ``model'' in which any
+sort of~$\TT$ is interpreted by the sort itself. For instance, if~$\sigma =
+(\varphi \seq{x_1\!\?\!X_1,\ldots,x_n\!\?\!X_n} \psi)$ is a sequent,
+then~$\speak{$\sigma$ holds for~$U_\TT$}$ is to be interpreted as the
+formula~``$\forall x_1\?X_1\_ \ldots \forall x_n\?X_n\_ (\varphi \Rightarrow
+\psi)$''.
+
+\begin{proof}[Proof of Theorem~\ref{thm:characterization}]
+The direction~$(2) \Rightarrow (1)$ is immediate, as the internal
 language of~$\Set[\TT]$ validates first-order intuitionistic logic and as it

### 2019-03-17 paper-qcoh.tex
The coherent case

@@
 
-\begin{lemma}Let~$\E$ be a Grothendieck topos. Let~$\TT$ be a coherent theory
+\begin{lemma}\label{lemma:provability-in-coherent-case}
+Let~$\E$ be a Grothendieck topos. Let~$\TT$ be a coherent theory
 internal to~$\E$. Let~$\sigma$ be a coherent sequent over the signature
@@
 
-% XXX coherent case
+The Nullstellensatz of Theorem~\ref{thm:nullstellensatz} and the
+characterization of Theorem~\ref{thm:characterization} pertain to geometric
+logic and hence require some care in dealing with the flexibility of
+disjunctions. In particular, the internal statement of
+Theorem~\ref{thm:nullstellensatz} requires external ingredients in referring to
+geometric$^\star$ sequents and provability$^\star$. A natural question is
+therefore whether, in the case that~$\TT$ is a coherent theory, the statements
+of these two theorems can be simplified.
+
+As stated in Lemma~\ref{lemma:provability-in-coherent-case}, for coherent
+sequents modulo a coherent theory there is no difference between provability in
+geometric logic, provability$^\star$ in geometric logic and provability in
+coherent logic. However, while geometric logic is powerful enough to express
+provability$^\star$ in geometric logic, coherent logic is not powerful enough
+to express provability in coherent logic. ``Coherent logic cannot eat itself.''
+Hence the the analogue of Theorem~\ref{thm:characterization} cannot be
+formulated for coherent logic.
+
+A fragment of logic which is still finitary, but is powerful enough to express
+its own provability predicate, is coherent existential fixed-point
+logic~\cite{blass:topoi-and-computation,blass:existential-fixed-point-logic}.
+For this fragment, we have the following characterization.
+
+\begin{scholium}Let~$\TT$ be a coherent theory (or more generally theory in coherent
+existential fixed-point logic). Let~$\alpha$ be a finitary first-order
+formula over the signature of~$\TT$. Then the following statements are
+equivalent.
+\begin{enumerate}
+\item The formula~$\alpha$ holds for~$U_\TT$. \smallskip
+\item The formula~$\alpha$ is provable in coherent existential fixed-point logic
+modulo the axioms of~$\TT$ and the additional axioms
+\begin{equation}
+  \speak{$\sigma$ holds for~$U_\TT$} \Longrightarrow \speak{$\ul{\TT}/U_\TT$
+  proves~$\sigma$}
+\end{equation}
+where~$\sigma$ ranges over the formulas of coherent existential fixed-point
+logic over the signature of~$\TT$.
+\end{enumerate}
+\end{scholium}
+
+\begin{proof}The proofs of Theorem~\ref{thm:nullstellensatz} and
+Theorem~\ref{thm:characterization} can be adapted to the setting of coherent
+existential fixed-point logic.
+\end{proof}
 

### 2019-03-17 paper-qcoh.tex
Review of basic facts about the classifying topos of Horn theories

@@
 
-Throughout this section, let~$\TT$ be a Horn theory.
+% XXX introduction
+
+Throughout this section, let~$\TT$ be a Horn theory. It is a basic result that
+Horn theories are of presheaf type, such that the classifying topos~$\Set[\TT]$
+can be taken as the topos of functors~$\Mod{\TT}_\mathrm{fp} \to \Set$,
+where~$\Mod{\TT}_\mathrm{fp}$ is the full subcategory of the category
+of~$\TT$-models (in~$\Set$) on the finitely presentable objects. The
+(underlying object of the) generic model is the tautologous functor~$U_\TT : T
+\mapsto T$. By~\emph{$U_\TT$-algebra} we mean, in analogy with the terminology in
+the case that~$\TT$ is the theory of rings, a~$\TT$-model~$M$ equipped with
+a~$\TT$-homomorphism~$U_\TT \to M$.
 
@@
 \end{proof}
-
-\begin{lemma}The category of~$\TT$-models is complete and
-cocomplete.\end{lemma}
-
-\begin{proof}Limits are computed as the limits of the underlying sets, colimits
-are computed by using the construction of Lemma~\ref{lemma:free-models}. For
-instance, the coproduct of~$\TT\langle X | R \rangle$ and~$\TT\langle X' | R'
-\rangle$ is~$\TT\langle X \amalg X' \,|\, R, R' \rangle$.\end{proof}
-
-Having the special case of the theory of rings in mind, we write the coproduct
-in the category of~$\TT$-models as~``$\otimes$''.
 
@@
 
-% XXX Set[T] einführen
+\begin{lemma}The category of~$\TT$-models (in~$\Set$) is complete and
+cocomplete.\end{lemma}
+
+\begin{proof}Limits are computed as the limits of the underlying sets, colimits
+are computed by using the construction of Lemma~\ref{lemma:free-models}. For
+instance, the coproduct of~$\TT\langle X | R \rangle$ and~$\TT\langle X' | R'
+\rangle$ is~$\TT\langle X \amalg X' \,|\, R, R' \rangle$.\end{proof}
+
+Having the special case of the theory of rings in mind, we write the coproduct
+in the category of~$\TT$-models as~``$\otimes$''.
 
@@
 
-% XXX terminology "U_T-algebra". location of the tensor explanation
-
 \begin{thm}The generic~$\TT$-model is \emph{quasicoherent} in the following sense:

### 2019-03-17 paper-qcoh.tex
Better explain the equivalence relation

@@
 Let~$\TT$ be a geometric theory. Let~$x_1\?X_1,\ldots,x_n\?X_n$ be a context over the
-signature of~$\TT$. Then the canonical morphism
+signature of~$\TT$. Let~$\mathrm{Form}^\star_{\vec
+x}(\ul{\TT}/U_\TT)/(\dashv\vdash_{\vec x})$ be the~$\Set[\TT]$-object of
+geometric$^\star$ formulas over the signature of~$\ul{\TT}/U_\TT$ in the
+context~$\vec x$, where any two such formulas are identified if and only
+if~$\ul{\TT}/U_\TT$ proves them equivalent.
+Then the canonical morphism
 \[ \mathrm{Form}^\star_{\vec x}(\ul{\TT}/U_\TT)/(\dashv\vdash_{\vec x}) \longrightarrow P(X_1 \times
 \cdots \times X_n) \]
-sending, internally speaking, the equivalence class of a geometric$^\star$ formula~$\varphi$ over the
-signature of~$\ul{\TT}/U_\TT$ in the context~$\vec x$ to the subset~$\{ (x_1,\ldots,x_n) \,|\, \varphi
-\}$ is an isomorphism.
-% XXX explain equivalence relation
+sending, internally speaking, the equivalence class of a geometric$^\star$
+formula~$\varphi$ to the subset~$\{ (x_1,\ldots,x_n) \,|\, \varphi \}$ is an isomorphism.
 \end{thm}

### 2019-03-17 paper-qcoh.tex
typo

@@
 Theorem~\ref{thm:nullstellensatz}. Surjectivity is by the definability
-result~\cite[Theorem~2.2]{caramello:definability}, exploiting that the internal
+result~\cite[Theorem~2.4]{caramello:definability}, exploiting that the internal
 statement localizes well by Lemma~\ref{lemma:truth-to-provability}.
-% XXX check reference number
 \end{proof}

### 2019-03-17 paper-qcoh.tex
Give details how definability entails the Nullstellensatz

@@
 
-% An immediate corollary of Theorem~\ref{thm:definability} is the
-% Nullstellensatz of Theorem~\ref{thm:nullstellensatz}. Indeed, arguing
-% internally to~$\Set[\TT]$, let~$\sigma = (\varphi \seq{\vec x} \psi)$ be a
-% geometric$^\star$ sequent over the signature of~$\ul{\TT}/U_\TT$. Assume
-% that~$\sigma$ holds for~$U_\TT$. Then the subsets~$\{ (\vec x) \,|\, \varphi
-% \}$ and~$\{ (\vec x) \,|\, \varphi \wedge \psi \}$ are equal. Hence, by
-% Theorem~\ref{thm:definability}, the formulas~$\varphi$
-% and~$\varphi \wedge \psi$ are provably$^\star$ equivalent.
-% Thus~$\ul{\TT}/U_\TT$ proves$^\star$~$(\varphi \seq{\vec x} \psi)$.
+\begin{rem}\label{rem:definability-entails-nullstellensatz}
+The proof of Theorem~\ref{thm:definability} used the Nullstellensatz
+of Theorem~\ref{thm:nullstellensatz}. Conversely, the internal statement of
+Theorem~\ref{thm:definability} entails the internal
+statement~\eqref{eq:nullstellensatz} of Theorem~\ref{thm:nullstellensatz} in
+extended geometric logic. Indeed, leaving the translation to formal logic to
+the reader, let~$\sigma = (\varphi \seq{\vec x} \psi)$ be a
+geometric$^\star$ sequent over the signature of~$\ul{\TT}/U_\TT$ such
+that~$\sigma$ holds for~$U_\TT$. Then the subsets~$\{ (\vec x) \,|\, \varphi
+\}$ and~$\{ (\vec x) \,|\, \varphi \wedge \psi \}$ are equal. Hence the
+assumption implies that the formulas~$\varphi$
+and~$\varphi \wedge \psi$ are provably$^\star$ equivalent.
+Thus~$\ul{\TT}/U_\TT$ proves$^\star$~$(\varphi \seq{\vec x} \psi)$.
+\end{rem}
 
@@
 can argue as in the proof of Theorem~\ref{thm:characterization}, noting that
-the axiom scheme indeed entails the axiom scheme~\eqref{eq:nullstellensatz} posited by
-Theorem~\ref{thm:characterization}.\end{proof}
+the axiom scheme indeed entails the axiom scheme~\eqref{eq:nullstellensatz} by
+Remark~\ref{rem:definability-entails-nullstellensatz}.\end{proof}
 

### 2019-03-17 paper-qcoh.tex
Cite Blass's /Seven trees in one/
to do
Discuss nontriviality of the Nullstellensatz

@@
 hence~$\ul{\TT}$) proves~$(\exists \vec y_j\_ \theta_j) \wedge (\exists \vec
-y_k\_ \theta_k) \seq{\vec x} \bigvee \{ \top \,|\, j = k \}$.
+y_k\_ \theta_k) \seq{\vec x} \bigvee \{ \top \,|\, j = k \}$. (This kind of
+reasoning also appears, for instance, in~\cite[p.~20]{blass:seven-trees}.)
 \end{proof}
@@
 
+The force of the Nullstellensatz of Theorem~\ref{thm:nullstellensatz} is for
+sequents~$\sigma$ which are not of the form~$(\top \seq{[]} \varphi)$. Indeed,
+for such sequents, even~$\TT$ proves~$(\speak{$\sigma$ holds for~$U_\TT$}
+\Rightarrow \speak{$\ul{\TT}/U_\TT$ proves~$(\top \seq{[]} \varphi)$})$, as can
+be seen by a simple induction on the structure of~$\varphi$. For more general
+sequents, Theorem~\ref{thm:nullstellensatz} expresses a nontrivial fact: that
+truth of a conditional, universally-quantified statement implies a single
+unconditional statement.
+
 \begin{rem}Theorem~\ref{thm:nullstellensatz} cannot be strengthened to
@@
 theory of rings, the internal geometric sequent~$(\top \seq{x\?U_\TT}
-\bigvee_{a\?U_\TT} (x = e_a))$ trivially holds for~$U_\TT$. However, it is not
-provable modulo~$\TT/U_\TT$, as for instance the polynomial algebra~$U_\TT[X]$
+\bigvee_{a\?U_\TT} (x = e_a))$ trivially holds for~$U_\TT$. However, this sequent is not
+provable modulo~$\ul{\TT}/U_\TT$, as for instance the polynomial algebra~$U_\TT[X]$
 does not validate it.\end{rem}
@@
 
-TODO:
-
-* XXX Nichttrivialität des Nullstellensatzes diskutieren (Spezialfall top |- φ
-  betrachten, dieser ist nämlich schon mehr oder weniger trivial)
+* XXX Cite Filip Bár?

### 2019-03-17 paper-qcoh.tex
Terminology clarification

@@
 A fragment of logic which is still finitary, but is powerful enough to express
-its own provability predicate, is coherent existential fixed-point
-logic~\cite{blass:topoi-and-computation,blass:existential-fixed-point-logic}.
-For this fragment, we have the following characterization.
+its own provability predicate, is \emph{coherent existential fixed-point
+logic}, coherent logic enriched by list sorts and the fixed-point operator of
+existential fixed-point logic~\cite{blass:topoi-and-computation,blass:existential-fixed-point-logic}.
+The list sorts are necessary to express raw terms, raw formulas and raw
+sequents; the fixed-point operator is then used to express well-formedness of
+raw terms, raw formulas and raw sequents; and to express provability. For this
+fragment, we have the following characterization.
 
@@
 modulo the axioms of~$\TT$ and the additional axioms
-\begin{equation}
+\[
   \speak{$\sigma$ holds for~$U_\TT$} \Longrightarrow \speak{$\ul{\TT}/U_\TT$
   proves~$\sigma$}
-\end{equation}
+\]
 where~$\sigma$ ranges over the formulas of coherent existential fixed-point

### 2019-03-17 paper-qcoh.tex
Simpler example

@@
 arbitrary geometric sequents. For instance, in the case that~$\TT$ is the
-theory of rings, the internal geometric sequent~$(\top \seq{x\?U_\TT}
+theory of objects, the internal geometric sequent~$(\top \seq{x\?U_\TT}
 \bigvee_{a\?U_\TT} (x = e_a))$ trivially holds for~$U_\TT$. However, this sequent is not
-provable modulo~$\ul{\TT}/U_\TT$, as for instance the polynomial algebra~$U_\TT[X]$
-does not validate it.\end{rem}
+provable modulo~$\ul{\TT}/U_\TT$, as for instance the model~$U_\TT \amalg
+\{\star\}$ does not validate it.\end{rem}
 

### 2019-03-17 paper-qcoh.tex
Slight strengthening
typo

@@
 
+\begin{rem}In the case that the~$\Set[\TT']$ is a dense subtopos
+of~$\Set[\TT]$, Theorem~\ref{thm:nullstellensatz-horn} can be strengthened to
+allow~$\bot$ as consequent, since in this case~$(\nabla\bot) \Rightarrow \bot$.\end{rem}
+
 Theorem~\ref{thm:nullstellensatz} cannot be strengthened to arbitrary
@@
 geometric$^\star$ sequents. For instance, in the case that~$\TT$ is the theory
-of local rings, the generic model~$U_\TT$ validates the coherent
-formula~$\speak{any element which is not zero is invertible}$, but~$\TT/U_\TT$
+of rings, the generic model~$U_\TT$ validates the coherent
+formula~$\speak{any element~$x$ for which~$(x = 0 \Rightarrow 1 = 0)$ is invertible}$, but~$\ul{\TT}/U_\TT$
 does not prove this fact, as it is for instance not validated by the polynomial
@@
 where~$\sigma$ ranges over the geometric formulas over the signature of~$\TT$.
+% XXX include free variables
 \end{enumerate}

### 2019-03-18 paper-qcoh.tex
Started section on applications

@@
 \section{Introduction}
+\label{sect:introduction}
 
@@
 \end{itemize}
+% Take care: In sect:applications, we state that each of these nongeometric
+% sequents also holds in the ring classifier.
 All of these nongeometric sequents are useful for the purposes of synthetic
 algebraic geometry, the desire to carry out algebraic geometry in a language
-close to the simple language on the 19th and the beginning of the 20th century
+close to the simple language of the 19th and the beginning of the 20th century
 while still being fully rigorous and fully general, working over arbitrary base
@@
 
-% XXX introduction
+The purpose of this section is to redo the development in the special case of
+Horn theories. XXX
 
@@
 
-\begin{thm}The generic~$\TT$-model is \emph{quasicoherent} in the following sense:
+\begin{thm}\label{thm:quasicoherence}
+The generic~$\TT$-model is \emph{quasicoherent} in the following sense:
 From the point of view of~$\Set[\TT]$, for any finitely
@@
 \end{thm}
+
+% XXX justification name quasicoherent
 
@@
 \end{proof}
+
+% XXX subtopos case
 
@@
 
+\section{Nongeometric sequents in the ring classifier}
+
+Each of the nongeometric sequents mentioned in Section~\ref{sect:introduction}
+can be deduced from the first-order or higher-order Nullstellensatz. Here we
+briefly indicate how this works using two examples.
+
+Let~$\TT$ be the theory of rings. Then, internally to~$\Set[\TT]$, we have for
+any element~$x \? U_\TT$:
+\begin{align*}
+  (x = 0 \Rightarrow 1 = 0) &\Longrightarrow \speak{$x$ is invertible} \quad\text{and} \\
+  (\speak{$x$ is invertible} \Rightarrow 1 = 0) &\Longrightarrow \speak{$x$ is nilpotent}.
+\end{align*}
+To verify the first claim, assume~$(x = 0 \Rightarrow 1 = 0)$. By the
+Nullstellensatz of Theorem~\ref{thm:nullstellensatz}, the
+theory~$\ul{\TT}/U_\TT$ proves this fact. Hence in particular it holds in
+the~$U_\TT$-algebra~$U_\TT/(x)$. Because~$x = 0$ in~$U_\TT/(x)$, we have~$1 =
+0$ in~$U_\TT/(x)$, that is~$1 \in (x)$, hence~$x$ is invertible.
+
+To verify the second claim, assume~$(\speak{$x$ is invertible} \Rightarrow 1 =
+0)$. By the Nullstellensatz, the theory~$\ul{\TT}/U_\TT$ proves this fact,
+hence in particular it holds in~$U_\TT[x^{-1}]$. Because~$x$ is invertible in
+this~$U_\TT$-algebra, we have~$1 = 0$ in~$U_\TT[x^{-1}]$. Hence~$x$ is
+nilpotent.
+
+Thanks to the version of the Nullstellensatz pertaining Horn consequents given
+in Theorem~\ref{thm:nullstellensatz-horn}, these two arguments can also be
+carried out in the classifying topos of local rings. It is instructive to
+consider how we would have to argue if we wanted to use only the unmodified
+Nullstellensatz:
+
+Assume~$(x = 0 \Rightarrow 1 = 0)$. By the Nullstellensatz, the theory of local
+rings proves this fact. From this we may not conclude that it holds
+for~$U_\TT/(x)$, as this~$U_\TT$-algebra might not be local. But we may
+conclude that it holds for~$\O_{\Spec(U_\TT/(x))}$, the structure sheaf of the affine scheme
+given by~$U_\TT/(x)$. This sheaf is not a model of~$\ul{\TT}/U_\TT$, but it is
+so from the point of view of the topos of sheaves over~$\Spec(U_\TT/(x))$. From
+the point of view of that topos,~$x = 0$ in~$\O_{\Spec(U_\TT/(x))}$, hence~$1 =
+0$ in~$\O_{\Spec(U_\TT/(x))}$. This amounts to~$U_\TT/(x)$ being the zero ring,
+hence~$x$ is invertible in~$U_\TT$.
+
+\section{On the Kock--Lawvere axiom in synthetic differential geometry}
+
+The generic local ring~$\affl$ validates the following strong form of the Kock--Lawvere
+axiom (Theorem~\ref{thm:quasicoherence}): \emph{For any finitely
+presented~$\affl$-algebra~$A$, the canonical map~$A \to (\affl)^{\Spec(A)}$ is
+an isomorphism of~$\affl$-algebras.} The special case~$A \defeq \affl[X]$
+implies that any map~$\affl \to \affl$ is given by a polynomial,
+since~$\Spec(\affl[X]) = \Hom_\affl(\affl[X],\affl) \cong \affl$.
+
+In contrast, in most models of synthetic differential geometry, we only have
+the following weaker statement: \emph{For any Weil algebra~$A$, the canonical
+map~$A \to \RR^{\Spec(A)}$ is an isomorphism of~$\RR$-algebras.} Here~$\RR$
+refers to the canonical line object (the image of the manifold~$\RR^1$ in the
+model) and~$\Spec(A)$ is the set of~$\RR$-algebra homomorphisms~$A \to \RR$.
+Any Weil algebra is finitely presented, but the converse fails; in particular,
+the polynomial algebra~$\RR[X]$ is not a Weil algebra, hence there is no reason
+to believe that any map~$\RR \to \RR$ is given by a polynomial.
+
+The quasicoherence statement of
+Theorem~\ref{thm:quasicoherence} can shed light on this discrepancy. % XXX
+
 % XXX

### 2019-03-19 paper-qcoh.tex
to do
typo
Some more nongeometric sequents

@@
 \newcommand{\TT}{\mathbb{T}}
+\newcommand{\OO}{\mathbb{O}}
 \newcommand{\ZZ}{\mathbb{Z}}
@@
 
-\section{Nongeometric sequents in the ring classifier}
+\subsection{Nongeometric sequents in the object classifier}
+
+Johnstone~\cite{XXX} lists several nongeometric sequents(XXX) holding in the
+classifying topos of the theory~$\OO$ of objects, the theory with a single sort and
+no function symbols, relations symbols or axioms, including:
+\begin{align*}
+  \forall x,y\?U_\OO\_ \neg\neg(x = y) &&
+  \forall x_1,\ldots,x_n\?U_\OO\_ \neg \forall y\?U_\OO\_ \bigvee_{i=1}^n (y = x_i)
+\end{align*}
+Intuitively, the first expresses that~$U_\OO$ is close to a subsingleton, while
+the second expresses that~$U_\OO$ is close to being infinite. Using the
+Nullstellensatz, these formulas can be verified as follows.
+
+We argue internally to~$\Set[\OO]$. Let~$x,y\?U_\OO$ and assume~$\neg(x = y)$.
+Then~$\ul{\OO}/U_\OO$ proves~$(x = y \seq{[]} \bot)$. Hence this sequent holds
+of any model of~$\ul{\OO}/U_\OO$, in particular of the
+quotient~$U_\OO/(x \sim y)$. Hence~$\bot$ holds for this model, that is, we
+have a contradiction.
+
+For the second formula, let~$x_1,\ldots,x_n\?U_\OO$ and assume~$\forall
+y\?U_\OO\_ \bigvee_{i=1}^n (y = x_i)$. By the Nullstellensatz, this universal
+statement is provable and hence holds for every~$\ul{\OO}/U_\OO$-model. But it
+does not hold for~$U_\OO \amalg \{\star\}$.
+
+The argument using the Nullstellensatz makes it transparent that both formulas
+are also satisfied by the generic model of any Horn theory~$\TT$: In that case,
+instead of constructing the quotient set~$U_\OO/(x \sim y)$, we construct the
+quotient model~$(\ul{\TT}/U_\TT)\langle \,|\, x = y \rangle$; and instead of~$U_\OO
+\amalg \{\star\}$, we construct the free model~$(\ul{\TT}/U_\TT)\langle \star
+\,|\, \rangle$ (XXX need to exclude trivial singleton theory).
+
+\subsection{Nongeometric sequents in the ring classifier}
 
@@
 
-\section{On the Kock--Lawvere axiom in synthetic differential geometry}
+\subsection{On the Kock--Lawvere axiom in synthetic differential geometry}
 
@@
 
+% XXX object classifier and related
+
 %We do have, for any geometric sequent~$\sigma$, that
@@
 
-* XXX Cite https://docplayer.net/55796150-Intuitionistic-algebra-some-recent-developments-in-topos-theory.html
-  (in general and for internal theories)
-
 * XXX Discuss related work, especially: Olivia 6.1.3, ...

### 2019-03-27 paper-qcoh.tex
to do
typo

@@
 From the point of view of~$\Set[\TT]$, for any finitely
-presentable~$U_\TT$-algebra~$A$ (finitely presented object in the category
+presentable~$U_\TT$-algebra~$A$ (finitely presentable object in the category
 of~$U_\TT$-algebras), the canonical~$U_\TT$-algebra homomorphism
@@
 * XXX Cite Filip Bár?
+
+* XXX restrict to finitary first-order formulas in the scholium

### 2019-03-28 paper-qcoh.tex
typo

@@
 coincides with~$U_\TT|A$ (notation as in the proof in
-Lemma~\ref{lemma:spec-sim-representable}), that is, the~$U_\TT$-algebra~$A^\sim$.
+Lemma~\ref{lemma:spec-sim-representable}), that is, with the~$U_\TT$-algebra~$A^\sim$.
 \end{proof}

### 2019-04-04 slides-birmingham2019.tex
Beginning of the slides for 6WFTop

--- 2018-12-05 slides-padova2018.tex
@@
 
-\title{New reduction techniques in commutative algebra driven by logical methods}
+\title{A general Nullstellensatz for generalized spaces}
 \author{Ingo Blechschmidt}
-\date{October 24th, 2018}
+\date{April 11st, 2019}
 
@@
   \begin{beamercolorbox}[dp=1ex,center]{}%
-      \usebeamercolor[fg]{item}{\textbf{{\Large \insertframetitle}}}
+  %   \usebeamercolor[fg]{item}{\textbf{{\Large \insertframetitle}}}
+    \begin{tikzpicture}
+      \def\R{8pt}
+      \node (title) {\hil{\insertframetitle}};
+      \begin{pgfonlayer}{background}
+      \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple]
+        ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
+        ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
+        ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
+        ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
+        cycle;
+      \end{pgfonlayer}
+    \end{tikzpicture}
   \end{beamercolorbox}%
@@
   \bigskip
-  \includegraphics[width=0.4\textwidth]{external-internal-small}
+  \includegraphics[width=0.4\textwidth]{phantoms}
   \bigskip
 
-  \hil{New reduction techniques in commutative algebra driven by logical methods}
+  \begin{tikzpicture}
+    \def\R{8pt}
+    \node (title) {\hil{\ A general Nullstellensatz for generalized spaces\ }};
+    \begin{pgfonlayer}{background}
+    \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple]
+      ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
+      ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
+      ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
+      ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
+      cycle;
+    \end{pgfonlayer}
+  \end{tikzpicture}
 
@@
 
-  Logic Seminar Padova \\
-  December 5st, 2018
+  6th Workshop on Formal Topology in Birmingham \\
+  April 11st, 2019
   \par
+
+  \jnote{1}{We can associate to any reduced ring~$A$ a \emph{forcing
+  model}~$A^\sim$.
+  \begin{itemize}
+    \item The forcing model has the pleasant property that it is a field.
+    \item Reasoning about it requires that we restrict ourselves to
+    intuitionistic logic.
+  \end{itemize}
+  Details are on the following slides.}
 \end{frame}}
 
-\section{Summary}
-\setbeamertemplate{headline}{\mynav{black}{gray}{gray}}
+\begin{frame}{The mystery of nongeometric sequents}
+  Let~$\TT$ be a geometric theory, for instance the theory of rings.
 
-\begin{frame}{Summary}
-  \vspace*{-1em}
+  % Illustration
 
-  \visible<4>{
-    \begin{changemargin}{-2.0em}{-0.5em}
-      \begin{itemize}
-        \item \ \\[-1.2em]\mbox{For any reduced ring~$A$, there is a ring~$A^\sim$ in a certain topos with}
-        \[ \models \bigl(\forall x\?A^\sim\_ \neg(\exists y\?A^\sim\_ xy = 1) \Rightarrow x = 0\bigr). \]
+  \textbf{Theorem.} There is a \hil{generic model}~$U_\TT$, a model such that
+  for any geometric sequent~$\sigma$ the following notions coincide:
+  \begin{enumerate}
+    \item The sequent~$\sigma$ holds for~$U_\TT$.
+    \item The sequent~$\sigma$ holds for any~$\TT$-model in any topos.
+    \item The sequent~$\sigma$ is provable modulo~$\TT$.
+  \end{enumerate}
 
-        \item This semantics is sound with respect to intuitionistic logic.
-
-        \item \ \\[-1.2em]\mbox{It has uses in classical and constructive commutative
-        algebra.}
-      \end{itemize}
-    \end{changemargin}
-  }
-
-  \vspace*{-1.5em}
-
-  \begin{columns}[t]
-    \begin{column}[t]{0.52\textwidth}
-      \centering
-
-      \begin{varblock}{\textwidth}{A baby example}
   \justifying
-        Let~$M$ be an injective matrix with more columns than rows over a
-        reduced ring~$A$.
-        Then~$1 = 0$ in~$A$.
-      \end{varblock}
-      \vspace*{-0.5em}
-
-      \only<1>{
-        \scalebox{0.8}{$\begin{pmatrix}
-          \cdot & \cdot & \cdot & \cdot & \cdot \\
-          \cdot & \cdot & \cdot & \cdot & \cdot \\
-          \cdot & \cdot & \cdot & \cdot & \cdot
-        \end{pmatrix}$}
-      }
-
-      \visible<2->{
-        \justifying
-        \textbf{Proof.} \bad{Assume not.} Then there is a \bad{minimal
-        prime ideal} $\ppp \subseteq A$. The matrix is injective over the \bad{field}~$A_\ppp = A[(A
-        \setminus \ppp)^{-1}]$; contradiction to basic linear algebra.
-      }
-    \end{column}
-
-    \begin{column}[t]{0.46\textwidth}
-      \centering
-
-      \begin{varblock}{\textwidth}{Generic freeness\phantom{p}}
-        \justifying
-        Generically, any finitely generated module over a reduced ring
-        is free.\phantom{g}
-      \end{varblock}
-      \vspace*{-0.5em}
-
-      \only<1-2>{{
-        \scriptsize\raggedright
-        (A ring is reduced iff $x^n=0$ implies $x=0$.)
-        \par
-      }}
-
-      \only<1-2>{\includegraphics[width=0.73\textwidth]{generic-freeness}}
-
-      \visible<3->{
-        \justifying
-        \textbf{Proof.} See \href{https://stacks.math.columbia.edu/tag/051Q}{[Stacks Project]}.
-      }
-    \end{column}
-  \end{columns}
-
-  \jnote{1}{The two displayed statements are trivial for fields. It is
-  therefore natural to try to reduce the general situation to the field
-  situation.}
-
-  \jnote{2}{The displayed proof, which could have been taken from any
-  standard textbook on commutative algebra, succeeds in this reduction
-  by employing proof by contradition and minimal prime ideals. However, this way
-  of reducing comes at a cost: It requires the Boolean Prime Ideal Theorem
-  (for ensuring the existence of a prime ideal and for ensuring that stalks at
-  minimal prime ideals are fields) and even the full axiom of choice (for
-  ensuring the existence of a minimal prime ideal).
-
-  We should hope that such a simple statement admits a more informative,
-  explicit, computational proof: There should be an explicit method for
-  transforming the given conditional equations expressing injectivity into the
-  equation~$1 = 0$. And indeed there is: Beautiful constructive proofs can
-  be found in Richman's note on
-  \fixedhref{https://www.ams.org/journals/proc/1988-103-04/S0002-9939-1988-0954974-5/S0002-9939-1988-0954974-5.pdf}{nontrivial
-  uses of trivial rings} and in the
-  \fixedhref{https://arxiv.org/abs/1605.04832}{recent textbook by Lombardi and
-  Quitté} on constructive commutative algebra.
-
-  The new reduction technique presented in this talk provides a way of
-  performing the reduction in an entirely constructive manner, avoiding the
-  axiom of choice. If so desired, resulting topos-theoretic proofs can be
-  unwound to yield fully explicit, topos-free, direct proofs.}
-
-  \jnote{3}{The baby example demonstrates that the reduction technique of this
-  talk is of interest to constructive commutative algebra. What about classical
-  commutative algebra? This is what the second example aims at.
-  Grothendieck's generic freeness lemma is an important theorem in algebraic
-  geometry, where it is usually stated in the following geometric form:
-
-  \vspace*{-1.2em}
-  \begin{indentblock}Let~$X$ be a reduced scheme. Let~$\B$ be
-  an~$\O_X$-algebra of finite type. Let~$\M$ be a~$\B$-module of finite type.
-  Then over a dense open,
-  \begin{enumerate}
-  \item[(a)] $\B$ and~$\M$ are locally free as sheaves of~$\O_X$-modules,
-  \item[(b)] $\B$ is of finite presentation as a sheaf of~$\O_X$-algebras and
-  \item[(c)] $\M$ is of finite presentation as a sheaf of~$\B$-modules.
-  \end{enumerate}\end{indentblock}
-  \vspace*{-1.2em}
-
-  All previously known proofs proceed in a series of reduction steps,
-  finally culminating in the case where~$A$ is a Noetherian integral domain.
-  They are somewhat convoluted (spanning several pages) and require nontrivial
-  prerequisites in commutative algebra.
-
-  Using the new reduction technique, there is a short (one-paragraph) and
-  conceptual proof of Grothendieck's generic freeness lemma. It is constructive
-  as a bonus; and if desired, one can unwind the resulting proof to obtain a
-  constructive proof which doesn't reference topos theory. The proof obtained
-  in this way is still an improvement on the previously known proofs, requiring
-  no advanced prerequisites in commutative algebra, and takes
-  \fixedhref{https://arxiv.org/abs/1807.01231}{about a page}.}
+  \textbf{Main observation.} The generic model can validate nongeometric
+  sequents which are not provable modulo~$\TT$ and which do not hold for
+  arbitrary models in arbitrary toposes.
 \end{frame}
 
-\section{The forcing model}
-\setbeamertemplate{headline}{\mynav{gray}{black}{gray}}
-
-\begin{frame}{Motivating the semantics}
-  \centering
-  \begin{varblockextra}{0.8\textwidth}{}{
-    \hil{Examples:}\phantom{Non-}\,\!\ \ $k,\ k[[X]],\ \CC\{z\},\ \ZZ_{(p)}$ \\[0.2em]
-    \hil{Non-examples:}\ \ $\ZZ,\ k[X],\ \ZZ/(pq)$
-  }
-    \justifying
-    A ring is \hil{local} iff~$1 \neq 0$ and if~$x + y = 1$ implies that~$x$ is
-    invertible or~$y$ is invertible.
-  \end{varblockextra}
-
-  \begin{varblockextra}{0.8\textwidth}{}{
-    Let~$x + y = 1$ in a ring~$A$.
-    Then:
-    \begin{itemize}
-      \item The element $x$ is invertible in~$A[x^{-1}]$.
-      \item The element $y$ is invertible in~$A[y^{-1}]$.
-    \end{itemize}
-    (Recall~$A[f^{-1}] = \bigl\{ \frac{u}{f^n} \,|\, u \in A, n \in \NN \bigr\}$.)
-  }
-    \hil{Locally,} any ring is local.
-  \end{varblockextra}
-
-  \jnote{1}{In topos theory, we have lots of experience of ``changing
-  universes'' in order to ``force'' some statements to be true. However,
-  because the field condition we are aiming at is not a ``geometric sequent'', these
-  techniques do not work here. Hence we try to take it more slowly and devise a
-  semantics which forces the given ring only to be local.
-
-  The key insight is that \emph{locally} (in the sense of topology/geometry),
-  any ring is a local ring. That is, we may pretend that any given ring is
-  local if we are prepared to pass to numerous localizations during the course
-  of an argument. The semantics displayed on the next slide manages this
-  localization-juggling for us.}
-\end{frame}
-
-\begin{frame}{The Kripke--Joyal semantics}
-  \small\vspace*{-0.7em}
-  \only<1>{Let~$A$ be a ring (commutative, with unit). We recursively define
-  \[ f \models \varphi \quad \text{(``$\varphi$ holds away from the zeros of~$f$'')} \]
-  for elements~$f \in A$ and statements~$\varphi$. Write
-  ``$\models \varphi$'' to mean~$1 \models \varphi$.}
-  \only<1>{\[ \renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\quad}c@{\quad}l@{}}
-    f \models \top &\text{is}& \text{true} \\
-    f \models \bot &\text{iff}& \text{$f$ is nilpotent} \\
-    f \models x = y &\text{iff}& x = y \in A[f^{-1}] \\
-    f \models \varphi \wedge \psi &\text{iff}&
-      \text{$f \models \varphi$ and $f \models \psi$} \\
-    f \models \varphi \vee \psi &\text{iff}&
-      \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
-    &&\quad\text{for each~$i$, $fg_i \models \varphi$ or $fg_i \models \psi$} \\
-    f \models \varphi \Rightarrow \psi &\text{iff}&
-      \text{for all~$g \in A$, $fg \models \varphi$ implies $fg \models \psi$} \\
-    f \models \forall x\?A^\sim\_ \varphi &\text{iff}&
-      \text{for all~$g \in A$ and all $x_0 \in A[(fg)^{-1}]$, $fg \models \varphi[x_0/x]$} \\
-    f \models \exists x\?A^\sim\_ \varphi &\text{iff}&
-      \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
-    &&\quad\text{for each~$i$, $fg_i \models \varphi[x_0/x]$ for some~$x_0 \in A[(fg_i)^{-1}]$}
-  \end{array} \]}
-  \only<2->{Write ``$\models \varphi$'' to mean~$1 \models \varphi$.}
-  \only<2->{\[ \renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\quad}c@{\quad}l@{}}
-    f \models x = y &\text{iff}& x = y \in A[f^{-1}] \\
-    f \models \varphi \wedge \psi &\text{iff}&
-      \text{$f \models \varphi$ and $f \models \psi$} \\
-    f \models \varphi \vee \psi &\text{iff}&
-      \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
-    &&\quad\text{for each~$i$, $fg_i \models \varphi$ or $fg_i \models \psi$}
-  \end{array} \]}
-
-  \jnote{1}{The clause for~``$\vee$'' is made exactly in such a way as to
-  ensure, if~$x + y = 1$, that~$1 \models ((\exists z\?A^\sim\_ xz = 1) \vee
-  (\exists z\?A^\sim\_ yz = 1))$.
-
-  The definition of the semantics is reminiscient of Kripke and Beth models.
-  Indeed, it is a fragment of the Kripke--Joyal semantics of the \emph{internal
-  language of a topos}, and this general semantics encompasses Kripke and Beth
-  models as special cases.}
-
-  \jnote{2}{The soundness lemma states: If~$f \models \varphi$, and
-  if~$\varphi$ intuitionistically entails a further statement~$\psi$, then
-  also~$f \models \psi$. In this way we can \emph{reason} with the forcing
-  model, similarly as if~$A^\sim$ would actually exist as a ring instead of
-  merely being a convenient syntactic fiction.
-
-  If we want~$A^\sim$ to actually exist, not just as a figure of speech, then
-  we have to broaden our notion of existence and accept ring objects in
-  toposes. More on this on the next slide.
-
-  Irrespective of whether~$A$ is a local ring, its mirror
-  image~$A^\sim$ is always a local ring (that is, the axioms of what it means
-  to be a local ring hold under the translation rules specified by the
-  semantics). A basic application of this forcing model are local-to-global
-  principles. For instance:
-  \begin{itemize}\justifying
-    \item The statement ``the kernel of a surjective matrix over a local ring
-    is finite free'' admits a constructive proof. It therefore holds
-    for~$A^\sim$. Its external meaning is that the kernel of a surjective
-    matrix~$M$ over~$A$ is finite locally free (there exists a partition~$1 =
-    f_1 + \cdots + f_n$ such that for each~$i$, the localized
-    module~$(\operatorname{ker} M)[f_i^{-1}]$ is finite free).
-
-    \item The ring~$A$ is a Prüfer domain if and only if~$A^\sim$ is a Bézout
-    domain. Therefore any constructive theorem about Bézout domains entails a
-    corresponding theorem about Prüfer domains. Bézout domains are quite rare,
-    while Prüfer domains abound (for instance the ring of integers of any
-    number field is a Prüfer domain, even constructively so).
-  \end{itemize}}
-
-  \pause
-
-  \begin{columns}
-    \begin{column}{0.50\textwidth}
-      \begin{varblock}{\textwidth}{Monotonicity}{}
-        If~$f \models \varphi$, then also~$fg \models \varphi$.
-      \end{varblock}
-    \end{column}
-
-    \begin{column}{0.50\textwidth}
-      \begin{varblock}{\textwidth}{Locality}{}
-        \justifying
-        If~$f^n = fg_1 + \cdots + fg_m$ and~$fg_i \models \varphi$ for all~$i$,
-        then also~$f \models \varphi$.
-      \end{varblock}
-    \end{column}
-  \end{columns}
-
-  \begin{columns}
-    \begin{column}{0.50\textwidth}
-      \begin{varblock}{\textwidth}{Soundness\phantom{p}}{}
-        If~$\varphi \vdash \psi$ and~$f \models \varphi$,
-        then~$f \models \psi$.
-      \end{varblock}
-    \end{column}
-
-    \begin{column}{0.50\textwidth}
-      \begin{varblock}{\textwidth}{Forced properties}{}
-        $\models \speak{$A^\sim$ is a local ring}$.
-      \end{varblock}
-    \end{column}
-  \end{columns}
-\end{frame}
-
-\tikzstyle{topos} = [draw=mypurple, very thick, rectangle, rounded corners, inner sep=5pt, inner ysep=10pt]
-\tikzstyle{title} = [fill=mypurple, text=white]
-
-\input{images/primes.tex}
-%\renewcommand{\sieve}[2]{SIEVE}
-%\renewcommand{\fakesieve}[2]{SIEVE}
-
-\newcommand{\drawbox}[4]{
-  \node[topos, #4] [fit = #3] (#1) {};
-  \node[title] at (#1.north) {#2};
-}
-
-\newcommand{\muchstuff}{
-  \includegraphics[height=3em]{filmat}
-  \scalebox{0.5}{\sieve{14}{2}}
-}
-
-\newcommand{\muchstuffplaceholder}{
-  \includegraphics[height=3em]{filmat-placeholder}
-  \scalebox{0.5}{\fakesieve{14}{2}}
-}
-
-\newcommand{\fewstuff}{
-  \includegraphics[height=3em]{filmat}
-  \scalebox{0.5}{\sieve{7}{2}}
-}
-
-\begin{frame}[fragile]{A universal property}
-  The displayed semantics is the first-order fragment of the \hil{higher-order
-  internal language} of the \hil{little Zariski topos}.
-
-  \begin{tikzpicture}
-    \node[scale=0.4] (objs-set1) at (-4.0,-2.5) {
-      \only<1->{\fewstuff}
-    };
-    \node[scale=0.4] (objs-eff1) at (4.0,-2.5) {
-      \only<1->{\fewstuff}
-    };
-    \node[scale=0.4] (objs-sh1)  at (0,-2.5) {
-      \only<1->{\fewstuff}
-    };
-
-    \node (prop-set1) [below of=objs-set1, align=left, inner ysep=0pt] {
-      The usual laws \\
-      of logic hold.
-    };
-
-    \node (prop-eff1) [below of=objs-eff1, align=left, inner ysep=0pt] {
-      Every function \\
-      is computable.
-    };
-
-    \node (prop-sh1) [below of=objs-sh1, align=left, inner sep=0pt] {
-      The intermediate \\
-      value theorem fails.
-    };
-
-    \begin{scope}
-      \drawbox{set1}{$\mathrm{Set}$}{(objs-set1) (prop-set1)}{}
-    \end{scope}
-    \begin{scope}
-      \drawbox{eff1}{Ef{}f}{(objs-eff1) (prop-eff1)}{tape}
-    \end{scope}
-    \begin{scope}
-      \drawbox{sh1}{$\mathrm{Sh}\, X$}{(objs-sh1) (prop-sh1)}{draw=none}
-      \def\R{8pt}
-      \begin{pgfonlayer}{background}
-      \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple]
-        ($(sh1.south west) + (\R, 0)$) arc(270:180:\R) --
-        ($(sh1.north west) + (0, -\R)$) arc(180:90:\R) --
-        ($(sh1.north east) + (-\R, 0)$) arc(90:0:\R) --
-        ($(sh1.south east) + (0, \R)$) arc(0:-90:\R) --
-        cycle;
-      \end{pgfonlayer}
-    \end{scope}
-  \end{tikzpicture}
-
-  \pause
-
-  Is there a \hil{free local ring}~$A \to A'$ over~$A$?
-  \begin{columns}[t]
-    \begin{column}{0.4\textwidth}
-      $\xymatrix{
-        A \ar[rd] \ar[rrr]^\alpha &&& {\substack{\phantom{\text{local}}\\\text{\normalsize$R$}\\\text{local}}} \\
-        & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!35]{\text{local}}[rru]
-      }$
-    \end{column}
-
-    \begin{column}{0.50\textwidth}
-      \small\justifying
-      For a fixed ring~$R$, the localization $A' \defeq A[S^{-1}]$ with $S \defeq
-      \alpha^{-1}[R^\times]$ would do the job.
-      \medskip
-
-      Hence we need the \hil{generic filter}.
-    \end{column}
-  \end{columns}
-\end{frame}
-
-\begin{frame}{The little Zariski topos}
-  \small
-
-  Let~$A$ be a ring. Its \hil{little Zariski topos} is equivalently
-  \vspace*{-0.5em}
-  \begin{enumerate}
-    \item the classifying locale of \hil{prime filters} of $A$, \\[-1.2em]
-    \item the classifying topos of \hil{local localizations} of $A$, \\[-1.2em]
-    \item the locale given by the frame of \hil{radical ideals} of $A$, \\[-1.2em]
-    \item the topos of sheaves over the poset $A$ with $f \preceq g$ iff $f \in \sqrt{(g)}$ and with $(f_i \to f)_i$ deemed covering iff $f \in \sqrt{(f_i)_i}$ or \\[-1.2em]
-    \item the topos of sheaves over $\Spec(A)$.
-  \end{enumerate}
-  Its associated topological space of points is the \hil{classical spectrum}
-  \[ \{ \fff \subseteq A \,|\, \text{$\fff$ prime filter} \} + \text{Zariski topology}. \]
-  It has \hil{enough points} if the Boolean Prime Ideal Theorem holds.
-
-  \scriptsize
-  Prime ideal:\, $0 \in \ppp$;\, $x \in \ppp \wedge y \in \ppp \Rightarrow x+y \in \ppp$;\, $1 \not\in \ppp$;\, $xy \in \ppp \Leftrightarrow x \in \ppp \vee y \in \ppp$
-
-  Prime filter:\, $0 \not\in \fff$;\,\,
-  $x+y \in \fff \Rightarrow x \in \fff \vee y \in \fff$;
-  \hspace*{0.5pt}\,\,
-  $1 \in \fff$;\,
-  $xy \in \fff \Leftrightarrow x \in \fff \wedge y \in \fff$
-\end{frame}
-
-\begin{frame}{Investigating the forcing model}
-% Let~$A$ be a reduced commutative ring ($x^n = 0 \Rightarrow x = 0$).
-  \small
-
-  The \hil{little Zariski topos} of a ring~$A$ is equivalently
-  \vspace*{-0.5em}
-  \begin{itemize}
-    \item the topos of sheaves over~$\Spec(A)$, \\[-1.9em]
-    \item the locale given by the frame of radical ideals of~$A$, \\[-1.9em]
-%   \item the classifying topos of local localizations of~$A$ or
-    \item the classifying locale of filters of~$A$
-  \end{itemize}
-  \vspace*{-0.5em}
-  and contains a \hil{mirror image} of~$A$, the sheaf of rings $A^\sim$.
-
-  \vspace*{-1.9em}
-
-  \begin{columns}[t]
-    \begin{column}{0.5\textwidth}
-      \begin{varblock}{\textwidth}{}
-        \justifying
-        Assuming the Boolean Prime Ideal Theorem, a first-order
-        formula ``$\forall \ldots \forall\_ (\cdots \Longrightarrow \cdots\!\,)$'',
-        where the two subformulas may not contain~``$\Rightarrow$'' and~``$\forall$'',
-        holds for~$A^\sim$ iff it holds for all stalks~$A_\ppp$.
-      \end{varblock}
-
-      \vspace*{-2em}
-
-      \begin{varblock}{\textwidth}{}
-        $A^\sim$ inherits any property of~$A$ which is
-        \hil{localization-stable}.
-      \end{varblock}
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
-      \vspace*{1.7em}
-
-      If~$A$ is reduced ($x^n = 0 \Rightarrow x = 0$):
-
-      \vspace*{-0.9em}
-      \setbeamercolor{block body}{bg=red!30}
-      \setbeamercolor{structure}{fg=purple}
-      \begin{varblock}{\textwidth}{}
-        $A^\sim$ is a \hil{field}.
-
-        $A^\sim$ has \hil{$\boldsymbol{\neg\neg}$-stable equality}.
-
-        \mbox{$A^\sim$ is \hil{anonymously Noetherian}.}\\[-1.2em]
-      \end{varblock}
-    \end{column}
-  \end{columns}
-
-  \visible<2->{\begin{tikzpicture}[overlay]
-    \draw[fill=white, draw=white, opacity=0.85] (-1,0) rectangle (\paperwidth,8.0);
-    \node[anchor=south west,inner sep=0] (image) at (0,1.0) {\vbox{
-      \only<2>{
-        \centering
-        \includegraphics[width=0.9\textwidth]{tierney-on-the-spectrum-of-a-ringed-topos} \\
-        \footnotesize
-        Miles Tierney. On the spectrum of a ringed topos. 1976.
-      }
-
-      \only<3>{
-	\begin{varblock}{\textwidth}{}
-	  The external meaning of
-	  \[
-	    \models
-	      \speak{$A^\sim[X_1,\ldots,X_n]$ is anonymously Noetherian}
-	  \]
-	  is:
-	  \medskip
-
-	  \begin{indentblock}
-	  For any element~$f \in A$ and any (not necessarily quasicoherent) sheaf of
-	  ideals~$\J \hookrightarrow A^\sim[X_1,\ldots,X_n]|_{D(f)}$: If
-	  \begin{indentblock}
-	  for any element~$g \in A$ the condition that
-	  \begin{indentblock}
-	  the sheaf~$\J$ is of finite type on~$D(g)$
-	  \end{indentblock}
-	  implies that~$g = 0$,
-	  \end{indentblock}
-	  then~$f = 0$.
-	  \end{indentblock}
-	\end{varblock}
-
-	\vspace*{-2em}
-      }
-    }};
-  \end{tikzpicture}}
-\end{frame}
-\renewcommand{\insertframeextra}{}
-
-\section{Revisiting the test cases}
-\setbeamertemplate{headline}{\mynav{gray}{gray}{black}}
-
-\begin{frame}{Revisiting the test cases}
-  \vspace*{-1em}
-  Let~$A$ be a reduced commutative ring ($x^n = 0 \Rightarrow x = 0$). \\
-  Let~$A^\sim$ be its mirror image in the little Zariski topos.
-
-  \begin{columns}[t]
-    \begin{column}[t]{0.48\textwidth}
-      \centering
-
-      \scalebox{0.5}{$\begin{pmatrix}
-        \cdot & \cdot & \cdot & \cdot & \cdot \\
-        \cdot & \cdot & \cdot & \cdot & \cdot \\
-        \cdot & \cdot & \cdot & \cdot & \cdot
-      \end{pmatrix}$}
-      \vspace*{-0.5em}
-
-      \begin{varblock}{\textwidth}{A baby example}
-        \justifying
-        Let~$M$ be an injective matrix over~$A$ with more columns than rows.
-        Then~$1 = 0$ in~$A$.
-      \end{varblock}
-
-      \justifying
-      \textbf{Proof.} $M$ is also injective as a matrix over~$A^\sim$.
-      Since~$A^\sim$ is a field, this is a contradiction by basic linear
-      algebra. Thus~$\models \bot$. This amounts to~$1 = 0$ in~$A$.
-    \end{column}
-
-    \begin{column}[t]{0.57\textwidth}
-      \centering
-
-      \includegraphics[height=1.9em]{generic-freeness}
-      \vspace*{-0.5em}
-
-      \begin{varblock}{\textwidth}{Generic freeness\phantom{p}}
-        \justifying
-        Let~$M$ be a finitely generated~$A$-module.
-        If~$f = 0$ is the only element of~$A$ such that~$M[f^{-1}]$ is a
-        free~$A[f^{-1}]$-module, then~$1 = 0$ in~$A$.
-      \end{varblock}
-      \vspace*{-0.1em}
-
-      \justifying
-      \textbf{Proof.} The claim amounts to \mbox{$\models
-      \text{``$M^\sim$}$}$\text{
-      is \hil{not not} free''}$. Since~$A^\sim$ is a field, this follows from
-      basic linear algebra.
-    \end{column}
-  \end{columns}
-\end{frame}
-
-\backupstart
-\setbeamertemplate{headline}{\mynav{gray}{gray}{gray}}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses}\end{minipage}}
-\begin{frame}
-  \bigskip
-  \centering
-  \includegraphics[height=3em]{heart}
-  \par
-  \raggedright
-
-  The Zariski topos and related toposes have applications in:
-  \begin{itemize}
-    \item classical algebra and classical algebraic geometry
-    \item constructive algebra and constructive algebraic geometry
-    \item synthetic algebraic geometry (``schemes are just sets'')
-  \end{itemize}
-
-  Connections with:
-  \begin{itemize}
-    \item understanding quasicoherence
-    \item the age-old mystery of nongeometric sequents
-  \end{itemize}
-
-\end{frame}}
-
-\begin{frame}[plain,c]
-  \centering%
-  \hil{Further reading}
-  \medskip
-
-  \framebox{\href{https://pizzaseminar.speicherleck.de/skript2/zariski-topos-klein.pdf}{\includegraphics[height=0.8\textheight]{fun-with-the-zariski-topos}}}\qquad%
-  \framebox{\href{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{\includegraphics[height=0.8\textheight]{phd-cover}}}
-  \par
-\end{frame}
-
-\begin{frame}{Applications in algebraic geometry}
-  \vspace*{-1.5em}
-  \begin{varblock}{0.9\textwidth}{}
-    \justifying
-    Understand \hil{notions of algebraic geometry} over a scheme~$X$ as
-    \hil{notions of algebra} internal to~$\Sh(X)$.
-  \end{varblock}
-
-  \small\centering
-  \scalebox{0.83}{\begin{tabular}{ll}
-    \toprule
-    externally & internally to $\Sh(X)$ \\
-    \midrule
-    sheaf of sets & set \\
-    %sheaf of rings & ring \\
-    sheaf of modules & module \\
-    sheaf of finite type & finitely generated module \\
-    % finite locally free sheaf & finite free module \\
-    % coherent sheaf & coherent module \\
-    tensor product of sheaves & tensor product of modules \\
-    % sheaf of Kähler differentials & module of Kähler differentials \\
-    sheaf of rational functions & total quotient ring of~$\O_X$ \\
-    dimension of $X$ & Krull dimension of~$\O_X$ \\
-    spectrum of a sheaf of~$\O_X$-algebras & ordinary spectrum [with a twist] \\
-    higher direct images & sheaf cohomology \\
-    \bottomrule
-  \end{tabular}}
-
-  \begin{columns}[c]
-    \begin{column}{0.47\textwidth}
-      \begin{exampleblock}{}
-        \justifying
-        Let $0 \to \F' \to \F \to \F'' \to 0$ be a short exact sequence
-        of sheaves of~$\O_X$-modules. If~$\F'$ and~$\F''$ are of finite type,
-        so is~$\F$.
-      \end{exampleblock}
-    \end{column}
-
-    \begin{column}{0.1\textwidth}
-      \vspace*{0.7em}
-      \scalebox{3}{$\Leftarrow$}
-    \end{column}
-
-    \begin{column}{0.44\textwidth}
-      \begin{exampleblock}{}
-        \justifying
-        Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence of
-        modules. If~$M'$ and~$M''$ are finitely generated, so is~$M$.
-      \end{exampleblock}
-    \end{column}
-  \end{columns}
-\end{frame}
-
-\begin{frame}{Synthetic algebraic geometry}
-  Usual approach to algebraic geometry: \hil{layer schemes above ordinary set theory}
-  using either
-  \begin{itemize}
-    \item locally ringed spaces
-    \small
-    \begin{multline*}
-      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$} + {} \\
-      \text{Zariski topology} + \text{structure sheaf}
-    \end{multline*}
-    \normalsize
-    \item or Grothendieck's functor-of-points account, where a scheme is a functor~$\mathrm{Ring} \to \mathrm{Set}$.
-    \small\[ A \longmapsto \{ (x,y,z) \in A^3 \,|\, x^n+y^n-z^n=0 \} \]
-  \end{itemize}
-  \bigskip
-
-  \hil{Synthetic approach:} model schemes \hil{directly as sets} in a certain
-  nonclassical set theory, the internal universe of the \mbox{\hil{big Zariski
-  topos}} of a base scheme.
-  \small
-  \[ \{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]
-\end{frame}
-
-\fontsize{11pt}{13.2}
-
-\begin{frame}{The big Zariski topos}
-  Let~$S$ be a fixed base scheme.
-  \begin{varblock}{0.9\textwidth}{Definition}
-    The \hil{big Zariski topos} $\Zar(S)$ of a scheme~$S$ is equivalently
-    \begin{enumerate}
-      \item the topos of sheaves over~$(\Aff/S)_{\mathrm{lofp}}$,
-      \item the classifying topos of local rings over~$S$ or
-      \item the classifying~$\Sh(S)$-topos of local~$\O_S$-algebras which are
-      local over~$\O_S$.
-    \end{enumerate}
-  \end{varblock}
-
-  \begin{itemize}
-    \item For an~$S$-scheme~$X$, its functor of points $\ull{X} =
-    \Hom_S(\cdot,X)$ is an object of~$\Zar(S)$. It feels like \hil{the set of
-    points} of~$X$.
-    \item In particular, there is the ring object~$\affl$ with~$\affl(T) =
-    \O_T(T)$.
-    \item This ring object is a \hil{field}: nonzero implies
-    invertible. \\{} [Kock 1976]
-  \end{itemize}
-
-  \jnote{1}{The objects of the category~$(\Aff/S)_{\mathrm{lofp}}$ are
-  morphisms of the form~$\Spec(R) \to S$ which are locally of finite
-  presentation. (Other choices of resolving set-theoretical issues of size are
-  also possible.)
-
-  A functor~$F : (\Aff/S)_{\mathrm{lofp}}^\op \to \Set$ is a sheaf
-  for the Zariski topology if and only if the diagram
-  \[ F(T) \to
-    \prod_i F(U_i) \rightrightarrows
-    \prod_{j,k} F(U_j \cap U_k) \]
-  is a limit diagram for any open covering~$T = \bigcup_i U_i$ of any
-  scheme~$T \in (\Aff/S)_{\mathrm{lofp}}$.
-
-  In the case that~$S = \Spec(A)$ is affine, the big Zariski topos of~$S$ is also
-  simply called ``big Zariski topos of~$A$''. It is a subtopos of the topos of
-  functors $\mathrm{Alg}(A) \to \Set$ and classifies local~$A$-algebras.
-
-  From the internal point of view of~$\Sh(S)$, the sheaf~$\O_S$ of rings is
-  just an ordinary ring, and we can construct internally to~$\Sh(S)$ the big
-  Zariski topos of~$\O_S$. Externally, this construction will yield a certain
-  bounded topos over~$\Sh(S)$. However, as indicated on the slide, this topos
-  will \emph{not} coincide with the true big Zariski topos of~$S$. To
-  construct the true big Zariski topos, we have to build, internally
-  to~$\Sh(S)$, the classifying topos of local \emph{and local-over-$\O_S$}
-  $\O_S$-algebras.}
-\end{frame}
-
-\begin{frame}{Synthetic constructions}
-  \small
-  $\hil{$\boldsymbol{\mathbb{A}^n}$} = (\affl)^n = \affl \times \cdots \times \affl$
-
-  $\begin{array}{@{}r@{}c@{}l@{}}
-    \hil{$\boldsymbol{\mathbb{P}^n}$} &\phantom{}=\phantom{}& \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee
-    \cdots \vee x_n \neq 0 \}/(\affl)^\times \\
-    &\cong& \text{set of one-dimensional subspaces of~$(\affl)^{n+1}$} \\
-    &&\qquad \text{(with~$\O(-1) = (\ell)_{\ell \? \mathbb{P}^n}$, $\O(1) =
-    (\ell^\vee)_{\ell \? \mathbb{P}^n}$)}
-  \end{array}$
-
-  $\hil{$\boldsymbol{\Spec(R)}$} = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
-    \text{set of $\affl$-valued points of $R$}$
-
-  $\hil{$\boldsymbol{TX}$} = X^\Delta$, where $\Delta = \{ \varepsilon \? \affl \,|\, \varepsilon^2 = 0 \}$
-
-  A subset $U \subseteq X$ is \hil{qc-open} if and only if for any $x : X$
-  there exist $f_1,\ldots,f_n \? \affl$ such that $x \in U \Longleftrightarrow
-  \exists i\_ f_i \neq 0$.
-
-  A \hil{synthetic affine scheme} is a set which is in bijection
-  with~$\Spec(R)$ for some synthetically quasicoherent~$\affl$-algebra~$R$.
-
-  A \hil{finitely presented synthetic scheme} is a set which can be covered by
-  finitely many qc-open f.p. synthetic affine schemes~$U_i$ such that the
-  intersections~$U_i \cap U_j$ can be covered by finitely many qc-open f.p.
-  synthetic affine schemes.
-
-  \jnote{1}{In the internal universe of the big Zariski topos of a base
-  scheme~$S$, $S$-schemes can simply be modeled by sets (enjoying the special
-  property that, in a certain precise sense, they are locally affine). This
-  slide expresses some of the basic constructions of~$S$-schemes in that
-  language.
-
-  Particularly nice are the following items.
-  \begin{itemize}\justifying
-    \item Projective~$n$-space can be given by the any of the two quite naive
-    expressions displayed on the slide.
-    \item Let~$X$ be an~$S$-scheme. We often think about a sheaves
-    of~$\O_X$-modules over~$X$ by their fibers; but for a rigorous
-    treatment in the standard foundations, we have to take the full sheaf
-    structure into account; the fibers do not determine a sheaf
-    uniquely.
-
-    From the internal point of view of~$\Zar(S)$, a sheaf of~$\O_X$-modules is
-    indeed simply a family of~$\affl$-modules, one~$\affl$-module for each
-    element of~$\ull{X}$. The slide illustrates how we can define the Serre
-    twisting sheaves in this language.
-  \end{itemize}}
-
-  \jnote{2}{\begin{itemize}\justifying
-    \item The spectrum of an~$\affl$-algebra can be given by the naive
-    expression displayed on the slide. It looks like this expression can't be
-    right, ignoring any non-maximal ideals; however, it is.
-    \item The big Zariski topos of an~$S$-scheme~$X$ is, from the internal
-    point of view of~$\Zar(S)$, simply the slice topos~$\Set/\ull{X}$. Hence to
-    give an~$X$-scheme simply amounts to giving an~$\ull{X}$-indexed family of
-    sets.
-  \end{itemize}
-
-  Synthetic algebraic geometry has been developed up to the point of étale
-  geometric morphisms. Much remains to be done: For instance, as of yet there
-  is only an account of Čech methods for computing cohomology, there is not yet
-  a synthetic treatment of true cohomology. Derived categories and intersection
-  theory are also missing.}
-\end{frame}
-
-\begin{frame}{Relations between the Zariski toposes}
-  The big Zariski topos is a topos over the small Zariski topos:
-  \[ \begin{array}{crcl}
-    \pi : & \Zar(A) &\longrightarrow& \Spec(A) \\
-    & \text{local~$A$-algebra~$(A \xrightarrow{\alpha} B)$} &\longmapsto& (A \to A[(\alpha^{-1}[B^\times])^{-1}])
-  \end{array} \]
-  This morphism is \hil{connected} ($\pi^{-1}$ is fully faithful) and
-  \hil{local}, so there is a preinverse
-  \[ \begin{array}{rcl}
-    \Spec(A) &\longrightarrow& \Zar(A) \\
-    \text{local localization~$(A \to B)$} &\longmapsto& (A \to B)
-  \end{array} \]
-  which is a subtopos inclusion inducing an idempotent monad~$\sharp$ and an
-  idempotent comonad~$\flat$ on~$\Zar(S)$.
-
-  \begin{itemize}
-  \justifying
-    \item Internally to~$\Zar(S)$, $\Spec(S)$ can be constructed as the
-    \hil{largest subtopos} where~$\flat \affl \to \affl$ is bijective.
-    \item Internally to~$\Spec(S)$, $\Zar(S)$ can be constructed as the
-    \hil{classifying topos} of local~$\O_S$-algebras which are local over~$\O_S$.
-    \item $\Zar(A)$ is the \hil{lax pullback}~$(\Set
-    \Rightarrow_{\Set[\Ring]} \Set[\LocRing])$.
-  \end{itemize}
-
-  \jnote{1}{Let~$A$ be a ring. By definition, we obtain a geometric
-  morphism~$\Set \to \Set[\Ring]$ into the classifying topos of rings.
-  There is also a geometric morphism~$\Set[\LocRing] \to \Set[\Ring]$, obtained
-  by realizing that any local ring is in particular a ring. These morphisms fit
-  together in a lax pullback square as follows:
-  \[ \xymatrix{
-    \Zar(A) \ar[r] \ar[d] & \Set[\LocRing] \ar[d] \\
-    \Set \ar[r] \ar@{}[ur]^(.3){}="a"^(.7){}="b" \ar@{=>} "a";"b" & \Set[\Ring]
-  } \]
-  This observation is joint with Peter Arndt and Matthias Hutzler.
-
-  Incidentally, the pseudo pullback of the morphism~$\Set[\LocRing] \to
-  \Set[\Ring]$ along~$\Set \to \Set[\Ring]$ is not very interesting: It's the
-  largest subtopos of~$\Set$ where~$A$ is a local ring. Assuming the law of
-  excluded middle, this subtopos is either the trivial topos (if~$A$ is not
-  local) or~$\Set$ (if~$A$ is local).
-
-  There is also a way of realizing the little Zariski topos of~$A$ as a pseudo
-  pullback, exploiting that the (localic) spectrum construction is geometric.
-  See Section~12.6 of
-  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
-  notes} for details.
-  }
-\end{frame}
-
-\section{The mystery of nongeometric sequents}
-
-\begin{frame}{Properties of the affine line}
-  \begin{itemize}
-%   \item $\affl$ is a local ring:
-%   \[ 1 \neq 0 \qquad\qquad \text{$x + y$ \inv} \Longrightarrow \text{$x$ \inv} \vee
-%   \text{$y$ \inv} \]
-    \item $\affl$ is a field:
-    \begin{align*}
-      \neg(x = 0) &\Longleftrightarrow \text{$x$ invertible \quad [Kock 1976]} \\
-      \neg(\text{$x$ invertible}) &\Longleftrightarrow \text{$x$ nilpotent}
-    \end{align*}
-    \item $\affl$ satisfies the axiom of microaffinity: Any map $f : \Delta \to
-    \affl$ is of the form~$f(\varepsilon) = a + b \varepsilon$ for unique
-    values~$a,b \? \affl$, where~$\Delta = \{ \varepsilon \? \affl \,|\,
-    \varepsilon^2 = 0 \}$.
-    \bigskip
-    \item Any map $\affl \to \affl$ is a polynomial function.
-    \bigskip
-    \item $\affl$ is anonymously algebraically closed: Any monic polynomial
-    does \emph{not not} have a zero.
-  \end{itemize}
-
-  \jnote{1}{The axiom of microaffinity is a special instance of the
-  \emph{Kock--Lawvere axiom} known from synthetic differential geometry. We'll
-  see on the next slide that~$\affl$ validates an unusually strong form of the
-  Kock--Lawvere axiom, not at all satisfied in the usual well-adapted models of
-  synthetic differential geometry.
-
-  The fact that, internally to~$\Zar(S)$, any map~$\affl \to \affl$ is a
-  polynomial can be seen as a formal version of the general motto that in
-  algebraic geometry, ``morphisms are polynomials''.}
-\end{frame}
-
-\begin{frame}{Synthetic quasicoherence}
-  Recall~$\Spec(R) = \Hom_{\mathrm{Alg}(\affl)}(R, \affl)$ and consider the statement
-  \[ \text{``the canonical map~$
-    \begin{array}[t]{rcl}
-      R &\longrightarrow& (\affl)^{\Spec(R)} \\
-      f &\longmapsto& (\alpha \mapsto \alpha(f))
-    \end{array}
-  $ is bijective''}.
-  \]
-  \vspace*{-1em}
-
-  \begin{itemize}
-    \item True for~$R = \affl[X]/(X^2)$ (microaffinity).
-    \item True for~$R = \affl[X]$ (every function is a polynomial).
-    \item True for \hil{any} finitely presented~$\affl$-algebra~$R$.
-  \end{itemize}
-
-  Any known property of~$\affl$ follows from this \\
-  \hil{synthetic quasicoherence}.
-
-  \centering
-  \begin{varblock}{0.6344\textwidth}{}
-    \centering
-    \hil{the mystery of nongeometric sequents}
-    \par
-  \end{varblock}
-
-  \jnote{1}{Let~$R$ be an~$\affl$-algebra. An element~$f \in R$ induces
-  an~$\affl$-valued function on~$\Spec(R)$; functions of this form can
-  reasonably be called ``algebraic''. In a synthetic context, there should be
-  no other~$\affl$-valued functions on~$\Spec(R)$ as these algebraic ones, and
-  different algebraic expressions should yield different functions. This is
-  precisely what the bijectivity of the displayed map expresses (in a positive
-  way).
-
-  In synthetic differential geometry, the closest cousin of synthetic algebraic
-  geometry, the analogue of the displayed map is only bijective for Weil
-  algebras such as~$\affl[X]/(X^2)$ or $\affl[X,Y]/(X^2,XY)$, not for arbitrary
-  finitely presented~$\affl$-algebras. This is a major difference to synthetic
-  differential geometry.}
-
-  \jnote{2}{The notion of synthetic quasicoherence is interesting for a number
-  of reasons:\vspace*{-1.2em}
-  \begin{itemize}\justifying
-    \item All currently known properties of~$\affl$, such as all the properties
-    listed on the previous slide, follow from the statement that~$\affl$ is
-    synthetically quasicoherent.
-
-    For instance, here is how we can verify the field property. Let~$x \?
-    \affl$ such that~$x \neq 0$. Set~$R = \affl/(x)$. Then~$\Spec(R) =
-    \emptyset$. Thus~$(\affl)^{\Spec(R)}$ is a singleton. Hence~$R = 0$.
-    Therefore~$x$ is invertible.
-    \item Given an~$\affl$-module~$E$, we can formulate the following variant
-    of the axiom of synthetic quasicoherence: ``For any finitely
-    presented~$\affl$-algebra~$R$, the canonical map~$R \otimes_\affl E \to
-    E^{\Spec(R)}$ is bijective.'' This axiom is satisfied if and only if~$E$ is
-    induced by a quasicoherent sheaf of~$\O_S$-modules.
-    \item The notion of synthetic quasicoherence is central to synthetic
-    algebraic geometry. The notions of synthetic open immersions, closed
-    immersion, schemes and several others all refer to synthetic
-    quasicoherence.
-  \end{itemize}}
-
-  \jnote{3}{An analogue of synthetic quasicoherence holds in the classifying
-  topos of rings, demonstrating that even presheaf toposes can validate
-  interesting nontrivial nongeometric sequents.
-
-  We believe that an analogue of synthetic quasicoherence holds for the generic
-  model of any geometric theory. This is work in progress. If true, this would
-  yield a major source of nongeometric sequents in classifying toposes. Because
-  of the many applications on nongeometric sequents, it's very desirable to
-  possess such a source.
-
-  The mystery of nongeometric sequents is this: On the one hand, they are very
-  useful to have because of surprising applications; on the other hand, they
-  are as of yet quite elusive.}
-\end{frame}
-
-\fontsize{10pt}{12}
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
-\begin{frame}{\!\!\!\mbox{Classifying toposes in algebraic geometry}}
-  \footnotesize
-  \centering
-  \begin{tabular}{ll}
-    \toprule
-    (big) topos & classified theory \\\midrule
-    Zariski & local rings [Hakim 1972] \\[0.6em]
-    étale & separably closed local rings [Hakim 1972, Wraith 1979] \\[0.6em]
-    \usebeamercolor[fg]{item}{fppf} & \usebeamercolor[fg]{item}{fppf-local rings} \\
-    & \textcolor{red!90}{(conjecturally: algebraically closed local rings)} \\[0.6em]
-    \textcolor{red!80}{ph} & \textcolor{red!80}{?? (conjecturally: algebraically closed valuation rings} \\
-    & \qquad \textcolor{red!80}{validating the projective Nullstellensatz)} \\[0.6em]
-    \usebeamercolor[fg]{item}{surjective} & \usebeamercolor[fg]{item}{algebraically closed geometric fields} \\[0.6em]
-    \textcolor{red!80}{$\neg\neg$} & \textcolor{red!80}{?? (conjecturally: algebraically closed geometric} \\
-    & \qquad \textcolor{red!80}{fields which are integral over the base)} \\[0.6em]
-    \usebeamercolor[fg]{item}{infinitesimal} & \usebeamercolor[fg]{item}{local algebras together with a nilpotent ideal [Hutzler 2018]} \\[0.6em]
-    \textcolor{red!100}{crystalline} & \textcolor{red!100}{??} \\
-    \bottomrule
-  \end{tabular}
-
-  \jnote{1}{Toposes, and also more specifically classifying toposes, originated
-  in algebraic geometry. It is therefore deeply embarrassing that as of now,
-  still very little is known about the theories classified by the major toposes
-  in active use by algebraic geometers.
-  \begin{center}
-    \includegraphics[width=0.99\textwidth]{wraith-generic-galois-theory} \\
-    \footnotesize
-    Gavin Wraith. Generic Galois theory of local rings. 1979. \\
-    (A cult classic and must-read for anyone interested in \\
-    the intersection of topos theory and algebraic geometry.)
-  \end{center}}
-
-  \jnote{2}{For almost forty years, only the big Zariski topos and its étale
-  subtopos were understood in that way.
-  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{These~2017
-  notes} answer the question for the fppf topology and the surjective topology
-  (in Section~21) and state conjectures for the ph topology and the
-  double negation topology. However, while good to have, the answer for the
-  fppf topology remains unsatisfactory, since Wraith's conjecture that the fppf
-  topos classifies the simpler theory of algebraically closed local rings has
-  neither been confirmed nor refuted.
-
-  A couple of weeks ago, Matthias Hutzler managed to determine the theory
-  classified by the big infinitesimal topos of a ring~$A$: It classifies
-  pairs~$(B,\aaa)$ consisting of a local~$A$-algebra~$B$ and a nilpotent
-  ideal~$\aaa \subseteq B$. Details will be in his forthcoming Master's thesis.
-  He is currently working on answering the question for the closely related big
-  crystalline topos.
-
-  It will be exciting to learn what the crystalline topos and the many other
-  toposes in algebraic geometry classify; how algebraic geometry can profit
-  from these discoveries; and which new flavors of synthetic algebraic geometry
-  they unlock.}
-\end{frame}}
-\backupend
-
 \end{document}

### 2019-04-05 slides-birmingham2019.tex
Explanations for some terms

@@
 
+\tikzset{
+  invisible/.style={opacity=0,text opacity=0},
+  visible on/.style={alt={#1{}{invisible}}},
+  alt/.code args={<#1>#2#3}{%
+    \alt<#1>{\pgfkeysalso{#2}}{\pgfkeysalso{#3}}}
+}
+
+\newcommand{\explain}[7]{%
+  \tikz[remember picture,baseline]{
+    \node[anchor=base,inner sep=2pt,outer sep=0,fill=#3,rounded corners] (label) {#1};
+    \node[anchor=north,visible on=<#2>,overlay,rectangle callout,rounded corners,callout
+    relative pointer={(0.0cm,0.5cm)+(0.0cm,#6)},fill=#3] at ($(label.south)+(0,-0.3cm)+(#4,#5)$) {#7};
+  }%
+}
+
 % Adapted from https://latex.org/forum/viewtopic.php?t=2251 (Stefan Kottwitz)
@@
 \setbeamertemplate{frametitle}{%
-  \vskip1.2em%
+  \vskip0.6em%
   \leavevmode%
@@
   \end{beamercolorbox}%
+  \vskip-0.6em%
 }
@@
 \newcommand{\mynav}[3]{%
-  \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex]{}%
-    \begin{columns}
-      \begin{column}{0.333\textwidth}
-        \raggedright
-        \textcolor{#1}{\qquad\qquad Summary}
-        \par
-      \end{column}
-      \begin{column}{0.333\textwidth}
-        \centering
-        \textcolor{#2}{The forcing model}
-        \par
-      \end{column}
-      \begin{column}{0.333\textwidth}
-        \raggedleft
-        \textcolor{#3}{Revisiting the test cases \qquad\qquad}
-        \par
-      \end{column}
-    \end{columns}
-  \end{beamercolorbox}%
+  foo
 }
@@
   \par
-
-  \jnote{1}{We can associate to any reduced ring~$A$ a \emph{forcing
-  model}~$A^\sim$.
-  \begin{itemize}
-    \item The forcing model has the pleasant property that it is a field.
-    \item Reasoning about it requires that we restrict ourselves to
-    intuitionistic logic.
-  \end{itemize}
-  Details are on the following slides.}
 \end{frame}}
@@
 \begin{frame}{The mystery of nongeometric sequents}
-  Let~$\TT$ be a geometric theory, for instance the theory of rings.
+  Let~$\TT$ be a \explain{geometric theory}{4-}{red!20}{-1cm}{0cm}{0cm}{\begin{minipage}{4.3cm}
+  sorts,
+  function symbols,
+  relation symbols,
+  geometric sequents as axioms
+  \end{minipage}}, for instance the
+  \explain{theory of rings}{4-}{blue!20}{-1cm}{0cm}{0cm}{\begin{minipage}{4.3cm}
+    \begin{tabbing}
+      fun. symb.: \= \kill
+      sorts: \> $R$ \\
+      fun. symb.: \> $0$, $1$, $+$, $-$, $\cdot$ \\
+      % rel. symb.: \> \emph{none} \\
+      axioms: \> $(\top \vdash_{x,y:R} x y = y x)$, \ldots
+    \end{tabbing}
+  \end{minipage}}.
 
   % Illustration
+  \vspace{2.5cm}
+
+  \pause
 
   \textbf{Theorem.} There is a \hil{generic model}~$U_\TT$, a model such that
-  for any geometric sequent~$\sigma$ the following notions coincide:
+  for any \explain{geometric sequent}{4-}{yellow!70}{0cm}{0cm}{0cm}{
+    a sequent of the form~$(\varphi \vdash_{x_1,\ldots,x_n} \psi)$
+    where~$\varphi$ and~$\psi$ are geometric formulas
+  }~$\sigma$ the following notions coincide:
   \begin{enumerate}
     \item The sequent~$\sigma$ holds for~$U_\TT$.
-    \item The sequent~$\sigma$ holds for any~$\TT$-model in any topos.
+    \item The sequent~$\sigma$ holds for any~$\TT$-model in any
+    \explain{topos}{4-}{purple!20}{0cm}{0cm}{0cm}{foo}.
     \item The sequent~$\sigma$ is provable modulo~$\TT$.
@@
 
+  \pause
+
   \justifying
-  \textbf{Main observation.} The generic model can validate nongeometric
-  sequents which are not provable modulo~$\TT$ and which do not hold for
-  arbitrary models in arbitrary toposes.
+  %\textbf{Main observation.} The generic model can validate nongeometric
+  %sequents which are not provable modulo~$\TT$ and which do not hold for
+  %arbitrary models in arbitrary toposes.
+
+  \textbf{Observation (Kock).} The generic local ring is a field:
+  \[ (x = 0 \Rightarrow \bot) \vdash_{x:R} (\exists y\?R\_ xy = 1) \]
 \end{frame}

### 2019-04-09 slides-birmingham2019.tex
Outline
More work on the slides

@@
 \usepackage{ifthen}
+\usepackage[normalem]{ulem}
 \usepackage{soul}\setul{0.3ex}{}
@@
 
+% lifted from https://arxiv.org/abs/1506.08870
+\DeclareFontFamily{U}{min}{}
+\DeclareFontShape{U}{min}{m}{n}{<-> udmj30}{}
+\newcommand\yon{\!\text{\usefont{U}{min}{m}{n}\symbol{'210}}\!}
+
 \newcommand{\A}{\mathcal{A}}
 \newcommand{\B}{\mathcal{B}}
+\renewcommand{\C}{\mathcal{C}}
 \newcommand{\M}{\mathcal{M}}
@@
 \newcommand{\inv}{inv.\@}
-\newcommand{\seq}{\vdash_{\!\!\!\vec x}}
+\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}
 \newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates
@@
 
+\newcommand{\hcancel}[5]{%
+  \tikz[baseline=(tocancel.base)]{
+    \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
+    \draw[red, line width=0.4mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
+  }%
+}
+
 \newcommand{\explain}[7]{%
@@
 }
+
+\newcommand{\squiggly}[1]{\textcolor{red}{#1}}  % XXX
 
@@
   \end{center}}
+
+\setbeamertemplate{headline}{%
+  \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex]{}%
+    \insertsectionnavigationhorizontal{\paperwidth}{}{}%
+  \end{beamercolorbox}%
+  \vskip0pt%
+}
 
@@
 
-\setbeamertemplate{headline}{\mynav{gray}{gray}{gray}}
+%\setbeamertemplate{headline}{\mynav{gray}{gray}{gray}}
 
@@
 
+\section{The generic model}
+
+\subsection{The mystery of nongeometric sequents}
+
 \begin{frame}{The mystery of nongeometric sequents}
@@
       sorts: \> $R$ \\
-      fun. symb.: \> $0$, $1$, $+$, $-$, $\cdot$ \\
+      fun. symb.: \> $0$, $1$, $-$, $+$, $\cdot$ \\
       % rel. symb.: \> \emph{none} \\
@@
 
-  \textbf{Theorem.} There is a \hil{generic model}~$U_\TT$, a model such that
+  \justifying
+  \textbf{Theorem.} There is a \hil{generic model}~$U_\TT$. It is
+  \hil{conservative} in that
   for any \explain{geometric sequent}{4-}{yellow!70}{0cm}{0cm}{0cm}{
@@
   }~$\sigma$ the following notions coincide:
+  \vspace*{-1.2em}
   \begin{enumerate}
@@
 
+\subsection{Construction of the generic model}
+
+\begin{frame}{Construction of the generic model}
+  The generic model is \hil{not} the same as \ldots
+  \begin{itemize}
+    \item the \hil{initial model}, think~$\ZZ$, or
+    \item the \hil{free model on one generator}, think~$\ZZ[X]$.
+  \end{itemize}
+  Set-theoretic models are \hil{too inflexible}.
+
+  \textbf{Definition.} The \hil{syntactic site}~$\C_\TT$ has \ldots
+  \begin{enumerate}
+    \item objects: $\{ x_1\?X_1, \ldots, x_n\?X_n\_ \varphi \}$ (shorter:
+    $\{ \vec x\_ \varphi \}$)
+    \item morphisms: eqv. classes of \explain{provably functional
+    formulas}{2-}{red!20}{-2.5cm}{-1.5cm}{0cm}{$\Hom(\{\vec x\_ \varphi\},\{\vec
+    y\_ \psi\}) = \{ \theta \,|\, (\theta \seq{\vec x, \vec y} \varphi
+    \wedge \psi), (\varphi \seq{\vec x} \exists! \vec
+    y\_ \theta) \text{ mod~$\TT$} \}/({\dashv\vdash})$}
+    \item coverings: provably jointly surjective families
+  \end{enumerate}
+  The topos of sheaves over~$\C_\TT$ is the \hil{classifying
+  topos}~$\Set[\TT]$. The generic model interprets a sort~$X$ by~$\yon\{x\?X\_\top\}$.
+\end{frame}
+
+\subsection{Going internal}
+
+\begin{frame}{Going internal}
+  \small
+  Let~$\C$ be a site. We recursively define
+  \[
+    U \models \varphi \quad
+    \text{(``$\varphi$ holds on~$U$'')}
+  \]
+  for objects~$U \in \C$ and \squiggly{formulas}~$\varphi$.
+  Write~``$\Sh(\C) \models \varphi$'' to mean~$1_\CC \models \varphi$.
+  \footnotesize
+  \[ \renewcommand{\arraystretch}{1.08}\begin{array}{@{}l@{\ \ }c@{\ \ }l@{}}
+  U \models \top &\Ll& \text{true} \\
+  U \models \bot &\Ll& \hcancel{\text{false}}{0pt}{3pt}{0pt}{-2pt}\ \text{the
+  empty family is a covering of~$U$} \\
+  U \models s = t \? F &\Ll& s|_U = t|_U \in F(U) \\
+  U \models \varphi \wedge \psi &\Ll&
+  \text{$U \models \varphi$ and $U \models \psi$} \\
+  U \models \varphi \vee \psi &\Ll&
+  \hcancel{\text{$U \models \varphi$ or $U \models
+  \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $(U_i \to U)_i$} \\
+  && \quad\quad\text{such that for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
+  U \models \varphi \Rightarrow \psi &\Ll&
+  \text{for all~$V \to U$: }
+  \text{$V \models \varphi$ implies $V \models \psi$} \\
+  U \models \forall s \? F\_ \varphi(s) &\Ll&
+  \text{for all $V \to U$ and sections~$s_0 \in F(V)$: $V \models \varphi(s_0)$} \\
+  % U \models \forall F\_ \varphi(F) &\Ll&
+  % \text{for all $V \to U$ and sheaves~$F_0$ over~$V$: $V \models \varphi(F_0)$} \\
+  U \models \exists s \? F\_ \varphi(s) &\Ll&
+  \hcancel{\text{there exists $s_0 \in F(U)$ such that $U \models \varphi(s_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
+  &&
+  \text{there exists a covering $(U_i \to U)_i$ such that for all~$i$:} \\
+  && \quad\quad \text{there exists~$s_0 \in F(U_i)$ such that $U_i \models \varphi(s_0)$}
+  %U \models \exists F\_ \varphi(F) &\Ll&
+  %\hcancel{\text{there exists a sheaf $F_0$ on $U$ such that $U \models
+  %\varphi(F_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
+  %&&
+  %\text{there exists a covering $(U_i \to U)_i$ such that for all~$i$:} \\
+  %&& \quad\quad \text{there exists a sheaf~$F_0$ on $U_i$ such that $U_i \models \varphi(F_0)$}
+  \end{array} \]
+\end{frame}
+
+% \begin{document}
+
+\section{Examples for nongeometric properties}
+
+\begin{frame}{Examples for nongeometric properties}
+  The \hil{generic object} validates [XXX]:
+  \begin{enumerate}
+    \item $\forall x,y\?U_\TT\_ \neg\neg(x = y)$.
+    \item $\forall x_1,\ldots,x_n\?U_\TT\_ \neg \forall y\?U_\TT\_
+    \bigvee_{i=1}^n y = x_i.$
+  \end{enumerate}
+
+  The \hil{generic ring} validates:
+  \begin{enumerate}
+    \item $\forall x\?U_\TT\_ \neg\neg(x = 0)$.
+    \item $\forall x\?U_\TT\_ (x = 0 \Rightarrow 1 = 0) \Rightarrow (\exists
+    y\?U_\TT\_ xy = 1)$.
+  \end{enumerate}
+
+  The \hil{generic local ring} validates:
+  \begin{enumerate}
+    \item $\neg \forall x\?U_\TT\_ \neg\neg(x = 0)$ (as~$1 \neq 0$).
+    \item $\forall a_0,\ldots,a_{n-1}\?U_\TT\_ \neg\neg \exists x\?U_\TT\_
+      x^n + a_{n-1}x^{n-1} + \cdots + a_0 x^0 = 0$.
+    \item \justifying Let~$\Delta = \{ \varepsilon \? U_\TT \,|\, \varepsilon^2 = 0 \}$.
+    For any map~$f : \Delta \to U_\TT$, there are unique elements~$a,b \?
+    U_\TT$ such that~$f(\varepsilon) = a + b\varepsilon$ for all~$\varepsilon
+    \? \Delta$.
+%   \item For any finitely presented~$U_\TT$-algebra~$A$,
+%   the canonical map~$A \to (U_\TT)^{\Hom_{U_\TT}(A,U_\TT)}$
+%   is an isomorphism.
+  \end{enumerate}
+\end{frame}
+
+\section{Motivation for generic models}
+
+\begin{frame}{Motivation for generic models}
+  Let~$A$ be a ring. Is there a \hil{free local ring}~$A \to A'$ over~$A$?
+  \begin{columns}[t]
+    \begin{column}{0.4\textwidth}
+      $\xymatrix{
+        A \ar[rd] \ar[rrr]^\alpha &&& {\substack{\phantom{\text{local}}\\\text{\normalsize$R$}\\\text{local}}} \\
+        & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!35]{\text{local}}[rru]
+      }$
+    \end{column}
+
+    \begin{column}{0.50\textwidth}
+      \small\justifying
+      For a fixed ring~$R$, the localization $A' \defeq A[S^{-1}]$ with $S \defeq
+      \alpha^{-1}[R^\times]$ would do the job. ($S$ is a \emph{filter}.)
+      \medskip
+
+      Hence we need the \hil{generic filter}.
+    \end{column}
+  \end{columns}
+
+  \pause
+
+  \justifying
+  The free local ring over~$A$ is~$A^\sim \defeq \underline{A}[F^{-1}]$,
+  where~$F$ is the generic filter, living in~$\Spec(A)$, the classifying topos
+  of filters of~$A$.
+
+  \pause
+
+  If~$A$ is reduced ($x^n = 0 \Rightarrow x = 0$):
+
+  \vspace*{-0.8em}
+  \setbeamercolor{block body}{bg=red!30}
+  \setbeamercolor{structure}{fg=purple}
+  \begin{varblock}{\textwidth}{}
+    $A^\sim$ is a \hil{field}: $\forall x\?A^\sim\_ (\neg(\exists y\?A^\sim\_
+    xy = 1) \Rightarrow x = 0)$.
+
+    $A^\sim$ has \hil{$\boldsymbol{\neg\neg}$-stable equality}:
+    $\forall x,y\?A^\sim\_ \neg\neg(x = y) \Rightarrow x = y$.
+
+    \mbox{$A^\sim$ is \hil{anonymously Noetherian}.}\\[-1.2em]
+  \end{varblock}
+\end{frame}
+
+\section{Internal conservativity}
+
+\begin{frame}{Internal conservativity}
+  The generic model~$U_\TT$ is conservative in the sense that:
+  \begin{multline*}
+    \qquad\textit{For any geometric sequent~$\sigma$,} \\
+    \textit{if $\Set[\TT] \models \speak{$U_\TT \models \sigma$}$,
+    then~$\TT$ proves~$\sigma$}.\qquad
+  \end{multline*}
+  But~$U_\TT$ \hil{does not believe} that it is conservative in the sense that
+  \hil{neither}
+  \begin{multline*}
+    \qquad\textit{For any geometric sequent~$\sigma$,} \\
+    \Set[\TT] \models (\speak{$U_\TT \models \sigma$} \Rightarrow
+    \speak{$\TT$ proves~$\sigma$}) \qquad
+  \end{multline*}
+  \hil{nor}
+  \begin{multline*}
+    \qquad\Set[\TT] \models \ulcorner\text{For any geometric sequent~$\sigma$,} \\
+      \text{if~$U_\TT \models \sigma$ then~$\TT$
+      proves~$\sigma$}\urcorner\qquad
+  \end{multline*}
+  hold.
+\end{frame}
+
+\begin{frame}{The Nullstellensatz}
+  \justifying
+  \textbf{The algebraic Nullstellensatz.}
+  Let~$A$ be a ring.
+  Let~$f,g \in A[X]$ be polynomials.
+  Then, subject to some conditions:
+  \[
+    \underbrace{\bigl(\forall x \in A\_ (f(x) = 0 \Rightarrow g(x) =
+    0)\bigr)}_{\text{algebraic truth}} \Longrightarrow
+    \underbrace{\bigl(\exists h \in A[X]\_ g = hf\bigr)}_{\text{algebraic certificate}}
+  \]
+
+  \justifying
+  \textbf{The topos-theoretic Nullstellensatz.}
+  Internally to~$\Set[\TT]$, for any geometric$^\star$ sequent~$\sigma$ over
+  the signature of~$\underline{\TT}/U_\TT$,
+  \[ \text{if~$U_\TT \models \sigma$, then~$\underline{\TT}/U_\TT$
+    proves$^\star$~$\sigma$.}
+  \]
+\end{frame}
+
+% Analogy with the harmonic oscillator in the study of physics
+
+% * The generic model
+%   * The mystery of nongeometric sequents
+%   * Construction of the generic model
+% * Kripke--Joyal semantics; focus on existence of extensions
+% * Examples for theories and nongeometric sequents with a focus on the wlog
+%   aspect
+% * Why we are interested in the generic models themselves
+%   * Universal localization
+%   * Grothendieck's generic freeness lemma
+% * Synthetic quasicoherence
+% * Conservativity from the internal point as guiding question
+%   * Naive formulation and why it doesn't work
+%   * Recall of the algebraic Nullstellensatz
+% * The new Nullstellensatz
+%   * What Set[T][T/U_T] classifies
+%   * Kock again
+%   * Subtleties regarding infinites
+%   * Improvement in the Horn case
+% * Universality of the source provided by the Nullstellensatz
+
 \end{document}
+
+% Are there size issues when KJ'ing internally?
+% Comparison to the approach by Vickers?
+% When does a theory prove the Nullstellensatz?
+
+% Sei T eine Theorie, für die ein konservatives Modell in Set existiert.
+% Dann haben wir eine Surjektion Set --> Set[T]. Also existiert eine
+% geometric expansion von T, welche Morita-äquivalent zur leeren Theorie ist.

### 2019-04-09 paper-qcoh.tex
typo
More work on the slides

@@
 and~$n$-tuple~$(x_1,\ldots,x_n)$ (of the appropriate sorts) such
-that~$R(x_1,\ldots,x_n)$ holds for~$U_\TT$.\end{defn}
+that~$R(x_1,\ldots,x_n)$.\end{defn}
 
@@
 to express provability in coherent logic. ``Coherent logic cannot eat itself.''
-Hence the the analogue of Theorem~\ref{thm:characterization} cannot be
+Hence the analogue of Theorem~\ref{thm:characterization} cannot be
 formulated for coherent logic.

### 2019-04-09 slides-birmingham2019.tex
More work on the slides

@@
 \usepackage[normalem]{ulem}
-\usepackage{soul}\setul{0.3ex}{}
 \usepackage[all]{xy}
@@
 \newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
-\newcommand{\ull}[1]{\underline{#1}}
+\newcommand{\ul}[1]{\underline{#1}}
 \newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
@@
   \begin{enumerate}
-    \item $\neg \forall x\?U_\TT\_ \neg\neg(x = 0)$ (as~$1 \neq 0$).
+    \item $\neg \forall x\?U_\TT\_ \neg\neg(x = 0)$.
     \item $\forall a_0,\ldots,a_{n-1}\?U_\TT\_ \neg\neg \exists x\?U_\TT\_
@@
 \begin{frame}{Internal conservativity}
-  The generic model~$U_\TT$ is conservative in the sense that:
+  The generic model~$U_\TT$ is conservative in the sense that
   \begin{multline*}
-    \qquad\textit{For any geometric sequent~$\sigma$,} \\
-    \textit{if $\Set[\TT] \models \speak{$U_\TT \models \sigma$}$,
-    then~$\TT$ proves~$\sigma$}.\qquad
+    \quad\textit{for any geometric sequent~$\sigma$,} \\
+    \textit{if $\Set[\TT] \models \speak{$\sigma$ holds for~$U_\TT$}$,
+    then~$\TT$ proves~$\sigma$}.\quad
   \end{multline*}
-  But~$U_\TT$ \hil{does not believe} that it is conservative in the sense that
-  \hil{neither}
+  But~$U_\TT$ \hil{does not believe} that it's conservative in the sense that
   \begin{multline*}
-    \qquad\textit{For any geometric sequent~$\sigma$,} \\
-    \Set[\TT] \models (\speak{$U_\TT \models \sigma$} \Rightarrow
-    \speak{$\TT$ proves~$\sigma$}) \qquad
+    \quad\Set[\TT] \models \ulcorner\text{for any geometric sequent~$\sigma$,} \\
+      \text{if~$\sigma$ holds for~$U_\TT$, then~$\ul{\TT}$
+      proves~$\sigma$}\urcorner\quad
   \end{multline*}
-  \hil{nor}
-  \begin{multline*}
-    \qquad\Set[\TT] \models \ulcorner\text{For any geometric sequent~$\sigma$,} \\
-      \text{if~$U_\TT \models \sigma$ then~$\TT$
-      proves~$\sigma$}\urcorner\qquad
-  \end{multline*}
-  hold.
+  does \hil{not} hold.
+
+  \textbf{Example.} Let~$\TT$ be the theory of rings. Then
+  \begin{align*}
+    \Set[\TT] &\models \neg(\speak{$\ul{\TT}$ proves $(\top \vdash 1 + 1 =
+    0)$}) \\
+    \text{but}\quad \Set[\TT] & \not\models \neg(1 + 1 = 0).
+  \end{align*}
 \end{frame}
@@
 
-  \justifying
+%  \textbf{Definition.} The theory~$\ul{\TT}/U_\TT$ is the
+%  internal geometric theory which arises from~$\ul{\TT}$ by adding:
+%  \begin{enumerate}
+%    \justifying
+%    \item constant symbols~$e_x$, one for each element~$x : U_\TT$,
+%    \item axioms~$(\top \vdash f(e_{x_1},\ldots,e_{x_n}) = e_{f(x_1,\ldots,x_n)})$ for each function
+%    symbol~$f$ and~$n$-tuple~$(x_1,\ldots,x_n) \in (U_\TT)^n$,
+%    \item axioms~$(\top \vdash R(e_{x_1},\ldots,e_{x_n}))$ for each relation symbol~$R$
+%    and~$n$-tuple~$(x_1,\ldots,x_n) \in (U_\TT)^n$ such
+%    that~$R(x_1,\ldots,x_n)$.
+%  \end{enumerate}
+
+  \textbf{The false naive topos-theoretic Nullstellensatz.}
+  Internally to~$\Set[\TT]$, for any geometric sequent~$\sigma$ over
+  the signature of~$\underline{\TT}$,
+  \[ \text{if~$\sigma$ holds for~$U_\TT$, then~$\underline{\TT}$
+    proves~$\sigma$.}
+  \]
+
   \textbf{The topos-theoretic Nullstellensatz.}
@@
   the signature of~$\underline{\TT}/U_\TT$,
-  \[ \text{if~$U_\TT \models \sigma$, then~$\underline{\TT}/U_\TT$
+  \[ \text{if~$\sigma$ holds for~$U_\TT$, then~$\underline{\TT}/U_\TT$
     proves$^\star$~$\sigma$.}

### 2019-04-10 slides-birmingham2019.tex
Slides almost finished

--- 2019-04-09 slides-birmingham2019.tex
@@
 
-\usepackage{pgfpages}
-\setbeameroption{show notes on second screen}
-\setbeamertemplate{note page}[plain]
-\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}
+%\usepackage{pgfpages}
+%\setbeameroption{show notes on second screen}
+%\setbeamertemplate{note page}[plain]
+\newcommand{\jnote}[2]{}%\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}
 
@@
 \newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
-\newcommand{\ull}[1]{\underline{#1}}
-\newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
+\newcommand{\ul}[1]{\underline{#1}}
+\newcommand{\affl}{\ensuremath{{\ul{\ensuremath{\AA}}^1}}}
 \newcommand{\Ll}{\vcentcolon\!\Longleftrightarrow}
@@
 
+\newcommand{\explainstub}[2]{%
+  \tikz[remember picture,baseline]{
+    \node[anchor=base,inner sep=2pt,outer sep=0,fill=#2,rounded corners] (label) {#1};
+  }%
+}
+
 \newcommand{\squiggly}[1]{\textcolor{red}{#1}}  % XXX
@@
       \def\R{8pt}
-      \node (title) {\hil{\insertframetitle}};
+      \node (title) {\hil{\large\insertframetitle}};
       \begin{pgfonlayer}{background}
@@
   \hil{conservative} in that
-  for any \explain{geometric sequent}{4-}{yellow!70}{0cm}{0cm}{0cm}{
-    a sequent of the form~$(\varphi \vdash_{x_1,\ldots,x_n} \psi)$
-    where~$\varphi$ and~$\psi$ are geometric formulas
-  }~$\sigma$ the following notions coincide:
+  for any \explainstub{geometric sequent}{yellow!70}~$\sigma$ the following notions coincide:
   \vspace*{-1.2em}
@@
     \item The sequent~$\sigma$ holds for any~$\TT$-model in any
-    \explain{topos}{4-}{purple!20}{0cm}{0cm}{0cm}{foo}.
+    \explainstub{topos}{purple!20}.
     \item The sequent~$\sigma$ is provable modulo~$\TT$.
@@
   \begin{itemize}
-    \item the \hil{initial model}, think~$\ZZ$, or
-    \item the \hil{free model on one generator}, think~$\ZZ[X]$.
+    \item the \hil{initial model} (think~$\ZZ$) or
+    \item the \hil{free model on one generator} (think~$\ZZ[X]$).
   \end{itemize}
@@
     $\{ \vec x\_ \varphi \}$)
-    \item morphisms: eqv. classes of \explain{provably functional
-    formulas}{2-}{red!20}{-2.5cm}{-1.5cm}{0cm}{$\Hom(\{\vec x\_ \varphi\},\{\vec
-    y\_ \psi\}) = \{ \theta \,|\, (\theta \seq{\vec x, \vec y} \varphi
-    \wedge \psi), (\varphi \seq{\vec x} \exists! \vec
-    y\_ \theta) \text{ mod~$\TT$} \}/({\dashv\vdash})$}
-    \item coverings: provably jointly surjective families
+    \item morphisms: eqv. classes of \explainstub{provably functional
+    formulas}{red!20}
+    \item coverings: \explainstub{provably jointly surjective
+    families}{purple!30}
   \end{enumerate}
@@
 
-\begin{frame}{Going internal}
+\begin{frame}{Working internally to toposes}
   \small
@@
 
-\section{Examples for nongeometric properties}
+\section[Nongeometric properties]{A selection of nongeometric properties}
 
-\begin{frame}{Examples for nongeometric properties}
-  The \hil{generic object} validates [XXX]:
+\begin{frame}{A selection of nongeometric properties}
+  The \hil{generic object} validates:
   \begin{enumerate}
@@
     \bigvee_{i=1}^n y = x_i.$
+    \item $(U_\TT)^{U_\TT} \cong 1 \amalg U_\TT$.
   \end{enumerate}
@@
     For any map~$f : \Delta \to U_\TT$, there are unique elements~$a,b \?
-    U_\TT$ such that~$f(\varepsilon) = a + b\varepsilon$ for all~$\varepsilon
+    U_\TT$ s.\,th.~$f(\varepsilon) = a + b\varepsilon$ for all~$\varepsilon
     \? \Delta$.
@@
 
-\section{Motivation for generic models}
+\section{Affine schemes}
 
-\begin{frame}{Motivation for generic models}
+\begin{frame}{Affine schemes}
   Let~$A$ be a ring. Is there a \hil{free local ring}~$A \to A'$ over~$A$?
@@
   \justifying
-  The free local ring over~$A$ is~$A^\sim \defeq \underline{A}[F^{-1}]$,
+  The free local ring over~$A$ is~$A^\sim \defeq \ul{A}[F^{-1}]$,
   where~$F$ is the generic filter, living in~$\Spec(A)$, the classifying topos
@@
 
-\section{Internal conservativity}
+% \begin{document}
 
-\begin{frame}{Internal conservativity}
-  The generic model~$U_\TT$ is conservative in the sense that:
-  \begin{multline*}
-    \qquad\textit{For any geometric sequent~$\sigma$,} \\
-    \textit{if $\Set[\TT] \models \speak{$U_\TT \models \sigma$}$,
-    then~$\TT$ proves~$\sigma$}.\qquad
-  \end{multline*}
-  But~$U_\TT$ \hil{does not believe} that it is conservative in the sense that
-  \hil{neither}
-  \begin{multline*}
-    \qquad\textit{For any geometric sequent~$\sigma$,} \\
-    \Set[\TT] \models (\speak{$U_\TT \models \sigma$} \Rightarrow
-    \speak{$\TT$ proves~$\sigma$}) \qquad
-  \end{multline*}
-  \hil{nor}
-  \begin{multline*}
-    \qquad\Set[\TT] \models \ulcorner\text{For any geometric sequent~$\sigma$,} \\
-      \text{if~$U_\TT \models \sigma$ then~$\TT$
-      proves~$\sigma$}\urcorner\qquad
-  \end{multline*}
-  hold.
+\section[A systematic source]{A systematic source of nongeometricity?}
+
+\begin{frame}{A systematic source of nongeometricity?}
+  \justifying
+  \textbf{Empirical fact.}
+  In \hil{synthetic algebraic geometry}, every known property of~$\affl$ followed from
+  its \hil{synthetic quasicoherence}:
+
+  \begin{quote}
+    \emph{\mbox{For any finitely presented~$\affl$-algebra~$A$, the canonical map}
+    \[ A \longrightarrow (\affl)^{\Hom_\affl(A,\affl)},\ s \longmapsto (x
+    \mapsto x(s)) \]
+    is an isomorphism of~$\affl$-algebras.}
+  \end{quote}
+
+  \vspace*{1.0em}
+  \begin{enumerate}
+    \item Does a general metatheorem explain this observation?
+    \item Is there a systematic source in any classifying topos?
+    \item Is there even an exhaustive source?
+  \end{enumerate}
+  \vspace*{-3.3em}
+
+  \visible<2>{\begin{tikzpicture}[overlay]
+    \draw[fill=white, draw=white, opacity=0.95] (-1,1.4) rectangle (\paperwidth,5.4);
+    \node[anchor=south west,inner sep=0] (image) at (0,1.0) {\vbox{
+      \only<2>{
+	\centering
+	\includegraphics[width=0.9\textwidth]{wraith-some-recent-developments-in-topos-theory} \\
+	\scriptsize
+        Gavin Wraith. Some recent developments in topos theory. \\ In:
+        Proc.\@ of the ICM (Helsinki, 1978).
+      }
+    }};
+  \end{tikzpicture}}%
+
+  \[ \xymatrix{
+    \text{$\TT$ proves~$\varphi$\;} \ar@/_1pc/@{=>}[r] &
+    \text{\;$\varphi$ holds for~$U_\TT$\;} \ar@/_1pc/@{=>}[r]
+    \ar@/_1pc/@{=>}@[red][l]
+    |{\SelectTips{cm}{}\textbf{\scalebox{3}{\textcolor{red}{\texttimes}}}}|{} &
+    \text{\;$\varphi$ is~$\TT$-redundant}
+    \ar@/_1pc/@{=>}@[red][l]
+    |{\SelectTips{cm}{}\textbf{\scalebox{3}{\textcolor{red}{\texttimes}}}}|{}
+  } \]
 \end{frame}
 
-\begin{frame}{The Nullstellensatz}
+\section{A topos-theoretic Nullstellensatz}
+
+\newcommand{\nullstellensatz}{%
   \justifying
+  \textbf{Theorem.} Internally to~$\Set[\TT]$:
+
+  \begin{quote}\emph{\mbox{For any \explainstub{geometric$^\star$ sequent}{red!20}~$\sigma$ over
+  the \explainstub{signature of~$\ul{\TT}/U_\TT$}{blue!20},}
+  if~\explainstub{$\sigma$ holds for~$U_\TT$}{yellow!70}, then~\explainstub{$\ul{\TT}/U_\TT$
+  proves$^\star$~$\sigma$}{yellow!70}.}
+  \end{quote}
+  \vspace*{-0.6em}
+}
+
+\begin{frame}{A topos-theoretic Nullstellensatz}
+  \nullstellensatz
+
+  \pause
+
   \textbf{The algebraic Nullstellensatz.}
@@
   \]
+  \vspace*{-0.9em}
 
+  \pause
+
+  \textbf{A naive version.}
+  ``Internally to~$\Set[\TT]$, for any geometric sequent~$\sigma$ over
+  the signature of~$\ul{\TT}$,
+  if~$\sigma$ holds for~$U_\TT$, then~$\ul{\TT}$
+  proves~$\sigma$.'' \textbf{False,} for instance with the theory of rings we
+  have
+  \begin{align*}
+    \Set[\TT] &\models \neg(\speak{$\ul{\TT}$ proves $(\top \vdash 1 + 1 =
+    0)$}) \\
+    \text{but}\quad \Set[\TT] & \not\models \neg(1 + 1 = 0).
+  \end{align*}
+\end{frame}
+
+\begin{frame}{A varying internal theory}
+  \nullstellensatz
+
+  \explainstub{\textbf{Definition.}}{blue!20} The theory~$\ul{\TT}/U_\TT$ is the
+  internal geometric theory of~\hil{$U_\TT$-algebras}, the theory which arises from~$\ul{\TT}$ by adding:
+  \begin{enumerate}
+    \justifying\small
+    \item for each element~$x : U_\TT$, a constant symbols~$e_x$,
+    \item for each function symbol~$f$ and~$n$-tuple~$(x_1,\ldots,x_n) \in (U_\TT)^n$, the axiom
+    $(\top \vdash f(e_{x_1},\ldots,e_{x_n}) = e_{f(x_1,\ldots,x_n)})$,
+    \item for each relation symbol~$R$ and~$n$-tuple~$(x_1,\ldots,x_n) \in (U_\TT)^n$ such
+    that~$R(x_1,\ldots,x_n)$, the axiom~$(\top \vdash R(e_{x_1},\ldots,e_{x_n}))$.
+  \end{enumerate}
+
+  \textbf{Remark.} Externalizing the internal classifying
+  topos~$\Set[\TT][\ul{\TT}/U_\TT]$ yields the classifying topos
+  of~$\TT$-homomorphisms.
+\end{frame}
+
+\begin{frame}{Example applications}
+  \nullstellensatz
+
+  \textbf{In the object classifier.}
+  Let~$x \? U_\TT$. Assume~$\neg(x = 0)$. By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves~$(x = 0
+  \Rightarrow \bot)$. But this is false in the~$\ul{\TT}/U_\TT$-model~$(U_\TT
+  \amalg \{ \heartsuit \})$.
+
+  \textbf{In the ring classifier.}
+  Let~$f,g \? U_\TT[X]$ such that any zero of~$f$ is a zero of~$g$.
+  By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves this fact.
+  Hence it holds in the~$\ul{\TT}/U_\TT$-model~$U_\TT[X]/(f)$. Hence~$g = 0$
+  in~$U_\TT[X]/(f)$, that is~$g = hf$ for some~$h \? U_\TT[X]$.
+\end{frame}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\hfill\mbox{\includegraphics[width=0.13\paperwidth]{hen}\;\;}\end{minipage}}
+\begin{frame}{Exhaustion and extensions}
   \justifying
-  \textbf{The topos-theoretic Nullstellensatz.}
-  Internally to~$\Set[\TT]$, for any geometric$^\star$ sequent~$\sigma$ over
-  the signature of~$\underline{\TT}/U_\TT$,
-  \[ \text{if~$U_\TT \models \sigma$, then~$\underline{\TT}/U_\TT$
-    proves$^\star$~$\sigma$.}
-  \]
-\end{frame}
+  \textbf{Theorem.} A first-order formula holds for~$U_\TT$ iff it is
+  intuitionistically provable from the axioms of~$\TT$ and
+  the scheme
+  \begin{equation}\tag{Nullstellensatz} \speak{$\sigma$ holds} \implies \speak{$\ul{\TT}/U_\TT$
+  proves$^{\star\star}$~$\sigma$}. \end{equation}
+
+  \textbf{Theorem.}
+  Let~$\TT'$ be a quotient theory
+  of~$\TT$. Assume that~$U_\TT$ is contained in the subtopos~$\Set[\TT']$. Then internally
+  to~$\Set[\TT']$:
+  \begin{quote}
+    \emph{A geometric$^\star$ sequent~$\sigma$ with Horn consequent holds for~$U_{\TT'}$
+    iff~$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$.}
+  \end{quote}
+  \vspace*{-0.3em}
+
+  \textbf{Theorem.}
+  % Let~$\mathrm{Form}^\star_{x\?X}(\ul{\TT}/U_\TT)/(\dashv\vdash_{x\?X})$ be the~$\Set[\TT]$-object of
+  % geometric$^\star$ formulas over the signature of~$\ul{\TT}/U_\TT$ in the
+  % context~$x\?X$, where any two such formulas are identified if and only
+  % if~$\ul{\TT}/U_\TT$ proves them equivalent.
+  The morphism~$\mathrm{ev}$ is an isomorphism.
+  \[ \mathrm{ev} : \mathrm{FunctFormulas}^\star(\ul{\TT}/U_\TT)/(\dashv\vdash)
+  \longrightarrow P(U_\TT) \]
+
+  \textbf{Theorem.} A higher-order formula holds for~$U_\TT$ iff it is
+  provable in intuitionistic higher-order logic from the axioms of~$\TT$ and
+  the higher-order Nullstellensatz scheme.
+\end{frame}}
 
@@
 % When does a theory prove the Nullstellensatz?
+% Look up what Kock and Reyes used the wlog for.
 

### 2019-04-11 slides-birmingham2019.tex
Slides finished

@@
 \usepackage{tikz}
-\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing}
+\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing,positioning}
 \hypersetup{colorlinks=true}
@@
 
-\title{A general Nullstellensatz for generalized spaces}
+\title{A general Nullstellensatz for generalised spaces}
 \author{Ingo Blechschmidt}
-\date{April 11st, 2019}
+\date{April 11th, 2019}
 
@@
 \newcommand{\affl}{\ensuremath{{\ul{\ensuremath{\AA}}^1}}}
-\newcommand{\Ll}{\vcentcolon\!\Longleftrightarrow}
+\newcommand{\Ll}{\text{iff}}
 \newcommand{\inv}{inv.\@}
@@
     \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
-    \draw[red, line width=0.4mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
+    \draw[red!80, line width=0.4mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
   }%
@@
 
-\newcommand{\squiggly}[1]{\textcolor{red}{#1}}  % XXX
+\newcommand{\squiggly}[1]{%
+  \tikz[remember picture,baseline]{
+    \node[anchor=base,inner sep=0,outer sep=0] (label) {#1};
+    \draw[thick,color=red!80,decoration={snake,amplitude=0.5pt,segment
+    length=3pt},decorate] ($(label.south west) + (0,-2pt)$) -- ($(label.south east) + (0,-2pt)$);
+  }%
+}
 
@@
 \setbeamertemplate{frametitle}{%
-  \vskip0.6em%
+  \vskip0.4em%
   \leavevmode%
@@
     \def\R{8pt}
-    \node (title) {\hil{\ A general Nullstellensatz for generalized spaces\ }};
+    \node (title) {\hil{A general Nullstellensatz for generalised spaces}};
     \begin{pgfonlayer}{background}
@@
 
+  6th Workshop on Formal Topology in Birmingham \\
+  April 11th, 2019
+  \bigskip
+
   Ingo Blechschmidt \\
   Università di Verona
-  \bigskip
-
-  6th Workshop on Formal Topology in Birmingham \\
-  April 11st, 2019
   \par
@@
 \begin{frame}{The mystery of nongeometric sequents}
-  Let~$\TT$ be a \explain{geometric theory}{4-}{red!20}{-1cm}{0cm}{0cm}{\begin{minipage}{4.3cm}
+  Let~$\TT$ be a \explain{geometric theory}{1-}{red!20}{-1cm}{0cm}{0cm}{\begin{minipage}{4.3cm}
+    \small
   sorts,
@@
   \end{minipage}}, for instance the
-  \explain{theory of rings}{4-}{blue!20}{-1cm}{0cm}{0cm}{\begin{minipage}{4.3cm}
+  \explain{theory of rings}{1-}{blue!20}{-1cm}{0cm}{0cm}{\begin{minipage}{4.3cm}
+    \small
     \begin{tabbing}
@@
 
-  % Illustration
-  \vspace{2.5cm}
+  \vspace*{1.8cm}
 
+  \begin{center}
+    \begin{tikzpicture}[ultra thick, node distance=7mm]
+      \node[rectangle, draw=blue!80, fill=blue!20] (a) {$\ZZ$};
+      \node[rectangle, draw=blue!80, fill=blue!20, right=of a] (b) {$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$};
+      \node[regular polygon, regular polygon sides=5, draw=orange!80,
+      fill=orange!20, right=of b, inner sep=0cm] (c) {$\O_X$};
+      \node[star, star points=10, inner sep=0cm, draw=purple!80, fill=purple!20, right=of c] {$U_\TT$};
+    \end{tikzpicture}
+  \end{center}
+
+  \vspace*{-0.5cm}
   \pause
@@
   \textbf{Observation (Kock).} The generic local ring is a field:
+  \vspace*{-0.2cm}
   \[ (x = 0 \Rightarrow \bot) \vdash_{x:R} (\exists y\?R\_ xy = 1) \]
@@
   \end{itemize}
-  Set-theoretic models are \hil{too inflexible}.
+  Set-based models are \hil{too inflexible}.
 
@@
 \begin{frame}{Working internally to toposes}
-  \small
+  \justifying
   Let~$\C$ be a site. We recursively define
@@
   for objects~$U \in \C$ and \squiggly{formulas}~$\varphi$.
-  Write~``$\Sh(\C) \models \varphi$'' to mean~$1_\CC \models \varphi$.
+  Write~``$\Sh(\C) \models \varphi$'' for~$1 \models \varphi$.
   \footnotesize
@@
 \end{frame}
-
-% \begin{document}
 
@@
       $\xymatrix{
-        A \ar[rd] \ar[rrr]^\alpha &&& {\substack{\phantom{\text{local}}\\\text{\normalsize$R$}\\\text{local}}} \\
+        A \ar[rd] \ar[rrr]^f &&& {\substack{\phantom{\text{local}}\\\text{\normalsize$R$}\\\text{local}}} \\
         & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!35]{\text{local}}[rru]
@@
       For a fixed ring~$R$, the localization $A' \defeq A[S^{-1}]$ with $S \defeq
-      \alpha^{-1}[R^\times]$ would do the job. ($S$ is a \emph{filter}.)
+      f^{-1}[R^\times]$ would do the job. ($S$ is a \emph{filter}.)
       \medskip
@@
 
-% \begin{document}
-
 \section[A systematic source]{A systematic source of nongeometricity?}
@@
     \draw[fill=white, draw=white, opacity=0.95] (-1,1.4) rectangle (\paperwidth,5.4);
-    \node[anchor=south west,inner sep=0] (image) at (0,1.0) {\vbox{
+    \node[anchor=south west,inner sep=0] (image) at (0,0.9) {\vbox{
       \only<2>{
@@
   \[ \xymatrix{
-    \text{$\TT$ proves~$\varphi$\;} \ar@/_1pc/@{=>}[r] &
-    \text{\;$\varphi$ holds for~$U_\TT$\;} \ar@/_1pc/@{=>}[r]
+    \text{$\TT$ proves~$\alpha$\;} \ar@/_1pc/@{=>}[r] &
+    \text{\;$\alpha$ holds for~$U_\TT$\;} \ar@/_1pc/@{=>}[r]
     \ar@/_1pc/@{=>}@[red][l]
-    |{\SelectTips{cm}{}\textbf{\scalebox{3}{\textcolor{red}{\texttimes}}}}|{} &
-    \text{\;$\varphi$ is~$\TT$-redundant}
+    |{\SelectTips{cm}{}\textbf{\scalebox{3}{\textcolor{red}{\!\texttimes\!}}}}|{} &
+    \text{\;$\alpha$ is~$\TT$-redundant}
     \ar@/_1pc/@{=>}@[red][l]
-    |{\SelectTips{cm}{}\textbf{\scalebox{3}{\textcolor{red}{\texttimes}}}}|{}
+    |{\SelectTips{cm}{}\textbf{\scalebox{3}{\textcolor{red}{\!\texttimes\!}}}}|{}
   } \]
@@
 
-  \begin{quote}\emph{\mbox{For any \explainstub{geometric$^\star$ sequent}{red!20}~$\sigma$ over
+  \begin{quote}\emph{\mbox{For any \explainstub{geometric$^\star$ sequent~$\sigma$}{red!20} over
   the \explainstub{signature of~$\ul{\TT}/U_\TT$}{blue!20},}
@@
     \justifying\small
-    \item for each element~$x : U_\TT$, a constant symbols~$e_x$,
-    \item for each function symbol~$f$ and~$n$-tuple~$(x_1,\ldots,x_n) \in (U_\TT)^n$, the axiom
+    \item for each element~$x : U_\TT$ a constant symbol~$e_x$,
+    \item for each function symbol~$f$ and~$n$-tuple~$(x_1,\ldots,x_n) \in (U_\TT)^n$ the axiom
     $(\top \vdash f(e_{x_1},\ldots,e_{x_n}) = e_{f(x_1,\ldots,x_n)})$,
     \item for each relation symbol~$R$ and~$n$-tuple~$(x_1,\ldots,x_n) \in (U_\TT)^n$ such
-    that~$R(x_1,\ldots,x_n)$, the axiom~$(\top \vdash R(e_{x_1},\ldots,e_{x_n}))$.
+    that~$R(x_1,\ldots,x_n)$ the axiom~$(\top \vdash R(e_{x_1},\ldots,e_{x_n}))$.
   \end{enumerate}
@@
 
-\begin{frame}{Example applications}
+\begin{frame}{Revisiting the test cases}
   \nullstellensatz
+  \medskip
 
   \textbf{In the object classifier.}
-  Let~$x \? U_\TT$. Assume~$\neg(x = 0)$. By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves~$(x = 0
-  \Rightarrow \bot)$. But this is false in the~$\ul{\TT}/U_\TT$-model~$(U_\TT
-  \amalg \{ \heartsuit \})$.
+%  Assume that~$U_\TT$ is empty, that is $\forall x \? U_\TT\_ (\top \Rightarrow
+%  \bot)$. By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves~$(\top \seq{x} \bot)$.
+%  But this is false in the~$\ul{\TT}/U_\TT$-model~$U_\TT \amalg
+%  \{\heartsuit\}$.
+  Let~$x,y \? U_\TT$. Assume that~$\neg(x = y)$.
+  By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves~$(e_x = e_y \vdash \bot)$.
+  But this is false in the~$\ul{\TT}/U_\TT$-model~$U_\TT/(x \sim y)$.
+  \medskip
 
@@
   By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves this fact.
-  Hence it holds in the~$\ul{\TT}/U_\TT$-model~$U_\TT[X]/(f)$. Hence~$g = 0$
+  Hence it holds in the~$\ul{\TT}/U_\TT$-model~$U_\TT[X]/(f)$. In this
+  model~$f$ has the zero~$[X]$. Hence also~$g([X]) = 0$
   in~$U_\TT[X]/(f)$, that is~$g = hf$ for some~$h \? U_\TT[X]$.
@@
 
+% An Tafel: geometric formula, local ring
+% Erwähnen: Marc Bezem, Ulrik Buchholtz, Thierry Coquand
+
 % Sei T eine Theorie, für die ein konservatives Modell in Set existiert.

### 2019-04-11 slides-birmingham2019.tex
Rounded corners

@@
     \begin{tikzpicture}[ultra thick, node distance=7mm]
-      \node[rectangle, draw=blue!80, fill=blue!20] (a) {$\ZZ$};
-      \node[rectangle, draw=blue!80, fill=blue!20, right=of a] (b) {$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$};
+      \node[rectangle, rounded corners=1pt, draw=blue!80, fill=blue!20] (a) {$\ZZ$};
+      \node[rectangle, rounded corners=1pt, draw=blue!80, fill=blue!20, right=of a] (b) {$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$};
       \node[regular polygon, regular polygon sides=5, draw=orange!80,
-      fill=orange!20, right=of b, inner sep=0cm] (c) {$\O_X$};
-      \node[star, star points=10, inner sep=0cm, draw=purple!80, fill=purple!20, right=of c] {$U_\TT$};
+      fill=orange!20, right=of b, rounded corners=1pt, inner sep=0cm] (c) {$\O_X$};
+      \node[star, rounded corners=1pt, star points=10, inner sep=0cm, draw=purple!80, fill=purple!20, right=of c] {$U_\TT$};
     \end{tikzpicture}

### 2019-04-11 slides-birmingham2019.tex
Picture of the black hole in M87

@@
 
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\hspace*{-1.7cm}\includegraphics[height=\paperheight]{black-hole-m87}\end{minipage}}
+\begin{frame}[plain]
+  \vspace*{\paperheight}\vspace*{-1cm}\scriptsize
+  \centering
+  \textcolor{white}{Fig.: The Event Horizon Telescope picture of the central black
+  hole in the galaxy M87}
+  \par
+\end{frame}}
+\addtocounter{framenumber}{-1}
+
 \section{The generic model}
@@
 
-  \textbf{Remark.} Externalizing the internal classifying
+  \textbf{Remark.} Externalising the internal classifying
   topos~$\Set[\TT][\ul{\TT}/U_\TT]$ yields the classifying topos

### 2019-04-14 slides-birmingham2019.tex
Grothendieck quote
Slides as they went live
Beginning of notes slides

@@
 
-%\usepackage{pgfpages}
-%\setbeameroption{show notes on second screen}
-%\setbeamertemplate{note page}[plain]
-\newcommand{\jnote}[2]{}%\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}
+\usepackage{pgfpages}
+\setbeameroption{show notes on second screen}
+\setbeamertemplate{note page}[plain]
+\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}
 
@@
   \par
+
+  \jnote{1}{\textcolor{red}{This set of slides will be updated soon with more
+  annotations.} Right now important references are not properly mentioned, including:
+  \begin{itemize}
+    \item Marc Bezem, Ulrik Buchholtz and Thierry Coquand. \emph{Syntactic forcing
+    models for coherent logic.}
+    \item Olivia Caramello. \emph{Universal models and definability.}
+    \item Carsten Butz and Peter Johnstone. \emph{Classifying toposes for
+    first-order theories.}
+  \end{itemize}}
 \end{frame}}
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\hspace*{-1.7cm}\includegraphics[height=\paperheight]{black-hole-m87}\end{minipage}}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\hspace*{-1.7cm}\includegraphics[height=\paperheight,trim=0 0 16.9cm 0,clip]{black-hole-m87}\end{minipage}}
 \begin{frame}[plain]
-  \vspace*{\paperheight}\vspace*{-1cm}\scriptsize
+  \vspace*{\paperheight}\vspace*{-0.8cm}\scriptsize
   \centering
@@
   \par
+
+  \jnote{1}{At this meeting we had lots of discussions on the notion of space.
+  Here is a picture of actual space, or rather the lack thereof, released only
+  yesterday.}
 \end{frame}}
@@
 \begin{frame}{The mystery of nongeometric sequents}
+  \jnote{1}{A \emph{geometric sequent} is a syntactical expression of the
+  form~$(\varphi \seq{x_1\?X_1,\ldots,x_n\?X_n} \psi)$,
+  where~$x_1\?X_1,\ldots,x_n\?X_n$ is a list of variable declarations,
+  the~$X_i$ ranging over the available sorts, and~$\varphi$ and~$\psi$ are
+  \emph{geometric formulas}. Often the variable context is abbreviated to~$\vec
+  x \? \vec X$ or even just~$\vec x$. Such a sequent is read as ``in the
+  context of variables~$\vec x$, $\varphi$ entails~$\psi$''.
+
+  Geometric formulas are built from atomic propositions (using equality or the
+  relation symbols) using the connectives~$\top$, $\bot$, $\wedge$, $\bigvee$
+  (set-indexed disjunction) and~$\exists$. Geometric formulas may not
+  contain~$\neg$, $\Rightarrow$, $\forall$.
+
+  There is a notion of a \emph{model} of a geometric theory in
+  a given topos. For instance, a ring in the usual sense is a model of the
+  theory of rings in the topos~$\Set$. The structure sheaf of a scheme~$X$ is a
+  model in the topos~$\Sh(X)$ of set-valued sheaves on~$X$.}
+
+  \jnote{2-}{Among all models in any topos, the \emph{universal} or
+  \emph{generic} one is special. It enjoys the universal property that any
+  model in any topos can be obtained from it by pullback along an essentially
+  unique geometric morphism. It is intriguing from a logical point of view
+  because it has exactly those properties which are shared by any model in any
+  topos.
+
+  One could argue, with a certain amount of success, that the generic model of
+  the theory of rings is what a mathematician implicitly refers to when she
+  utters the phrase ``Let~$R$ be a ring''. This point of view is fundamental to
+  the slogan \emph{continuity is geometricity}, as expounded for instance in
+  \fixedhref{http://www.cs.bham.ac.uk/~sjv/GeoAspects.pdf}{Continuity and
+  geometric logic} by Steve Vickers.}
+
+  \jnote{3-}{Crucially, the conservativity statement only pertains to
+  properties which can be put as geometric sequents. Generic models may have
+  additional nongeometric properties. Because conservativity does not apply to
+  them, they are not shared by all models in all toposes -- but any consequences
+  which can be put as geometric sequents are.
+
+  Hence one reason why these nongeometric sequents are interesting is because
+  they provide us with new reduction strategies (``without loss of
+  generality'').}
+
   Let~$\TT$ be a \explain{geometric theory}{1-}{red!20}{-1cm}{0cm}{0cm}{\begin{minipage}{4.3cm}
@@
   topos}~$\Set[\TT]$. The generic model interprets a sort~$X$ by~$\yon\{x\?X\_\top\}$.
+
+  \jnote{1-}{In case the theory~$\TT$ is a Horn theory, term algebras ... XXX}
 \end{frame}
@@
 
+% Grothendieck quote for the title picture:
+%
+% Ces “nuages probabilistes”, remplaçant les rassurantes particules matérielles
+% d’antan, me rappellent étrangement les élusifs “voisinages ouverts” qui
+% peuplent les topos, tels des fantômes évanescents, pour entourer des “points”
+% imaginaires, ...
+%
+% Diese "Wahrscheinlichkeitswolken", welche die beruhigenden materiellen
+% Partikel von früher ersetzen, erinnern mich irgendwie an die flüchtigen
+% "offenen Umgebungen" der Topoi -- wie dahinschwindende Phantome, um
+% die fiktiven "Punkte" zu umgeben, ...
+%
+% These “probability clouds”, replacing the reassuring material particles of
+% before, remind me strangely of the elusive “open neighborhoods” that populate
+% the topoi, like evanescent phantoms, to surround the imaginary “points”.
+
 % Sei T eine Theorie, für die ein konservatives Modell in Set existiert.

### 2019-04-15 slides-birmingham2019.tex
More notes

@@
 
+  For instance, if we want to verify a geometric sequent for all local rings,
+  we may freely use the displayed field axiom.
   Hence one reason why these nongeometric sequents are interesting is because
@@
     \begin{tikzpicture}[ultra thick, node distance=7mm]
-      \node[rectangle, rounded corners=1pt, draw=blue!80, fill=blue!20] (a) {$\ZZ$};
-      \node[rectangle, rounded corners=1pt, draw=blue!80, fill=blue!20, right=of a] (b) {$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$};
+      \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40] (a) {$\ZZ$};
+      \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40, right=of a] (b) {$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$};
       \node[regular polygon, regular polygon sides=5, draw=orange!80,
@@
 
-  \jnote{1-}{In case the theory~$\TT$ is a Horn theory, term algebras ... XXX}
+  \jnote{1-}{In case the theory~$\TT$ is a Horn theory (for instance if it is
+  an equational theory), the \emph{term algebra} (the set of terms in the
+  empty context modulo provable equality) is a model of~$\TT$. While such
+  models do enjoy some nice categorical properties, they are in general
+  \emph{not} the generic model.}
 \end{frame}

### 2019-04-15 slides-birmingham2019.tex
More notes

@@
 
+% \begin{document}
+
 \begin{frame}{The mystery of nongeometric sequents}
@@
   theory of rings in the topos~$\Set$. The structure sheaf of a scheme~$X$ is a
-  model in the topos~$\Sh(X)$ of set-valued sheaves on~$X$.}
+  model in the topos~$\Sh(X)$ of set-valued sheaves on~$X$.
+
+  With \emph{topos} we mean Grothendieck topos, and as metatheory we use a
+  constructive but impredicative flavour of English (which could be formalised
+  by what is supported by the internal language of elementary toposes with an NNO).
+  However the Nullstellensatz presented later makes no use of the
+  subobject classifier, hence the results can likely be generalised to hold in
+  a predicative metatheory or to hold for arithmetic universes.}
 
@@
 
-  \jnote{1-}{In case the theory~$\TT$ is a Horn theory (for instance if it is
+  \jnote{1}{In case the theory~$\TT$ is a Horn theory (for instance if it is
   an equational theory), the \emph{term algebra} (the set of terms in the
@@
   models do enjoy some nice categorical properties, they are in general
-  \emph{not} the generic model.}
+  \emph{not} the generic model.
+
+  In general, the generic model cannot be realised as a set-based model (with
+  a set for each sort, a map for each function symbol and so on). Sets are too
+  constant for this purpose; the flexibility of sheaves (``variable sets'') is
+  required: The generic model lives in the topos of set-valued
+  sheaves over~$\C_\TT$.
+
+  A morphism~$A = \{\vec x\_ \varphi\} \to \{\vec y\_ \psi\} = B$ in~$\C_\TT$ is the
+  equivalence class (modulo provable equivalence) of a geometric
+  formula~$\theta$ such that~$\TT$ proves
+  \begin{enumerate}
+    \item ``$\theta$ is a relation on~$A \times B$'': $(\theta \seq{\vec x,
+    \vec y} \varphi \wedge \psi)$
+    \item ``$\theta$ is total'': $(\varphi \seq{\vec x} \exists \vec y\_
+    \theta)$
+    \item ``$\theta$ is single-valued'': $(\theta \wedge \theta[\vec y'/\vec y]
+    \seq{\vec x,\vec y,\vec y'} \vec y = \vec y')$
+  \end{enumerate}
+
+  \vspace*{-0.6em}
+  A family~$(\{ \vec x_i\_ \varphi_i \} \xrightarrow{[\theta_i]} \{ \vec y\_
+  \psi \})_i$ is a covering iff~$\TT$ proves~$(\psi \seq{\vec y} \bigvee_i
+  \exists \vec x_i\_ \theta_i)$.
+
+  The slides experiment with using the symbol~``$\yon$'' for the Yoneda
+  embedding~$\C_\TT \to \Set[\TT]$, as in \fixedhref{XXX}{Elements
+  of~$(\infty,1)$-category theory} by Emily Riehl and Dominic Verity.}
+
+  \jnote{2}{The special case that the generic model of a theory~$\TT$ can be realised as
+  a model in~$\Set$ occurs iff~$\TT$ is Morita-equivalent to the
+  empty theory, that is, iff~$\TT$ has exactly one model in any topos.
+
+  The special case that there exists at least some conservative~$\TT$-model
+  in~$\Set$ occurs iff~$\TT$ has a conservative geometric expansion to a theory
+  which is Morita-equivalent to the empty theory.}
 \end{frame}
@@
   \end{array} \]
+
+  \jnote{1-}{
+    Using the internal language of a topos, we can import all$^\star$ of
+    constructive impredicative mathematics into the internal universe of a
+    topos. XXX
+  }
 \end{frame}

### 2019-04-22 slides-birmingham2019.tex
More notes

@@
 
-  \jnote{1-}{
-    Using the internal language of a topos, we can import all$^\star$ of
-    constructive impredicative mathematics into the internal universe of a
-    topos. XXX
+  \jnote{1}{
+    The internal language of a (Grothendieck or elementary) topos~$\E$ is a
+    device which allows us to speak and reason about the objects and morphisms
+    of~$\E$ in a naive element-based language close to the usual formal
+    mathematical language. Using this language, objects of~$\E$ look like plain
+    old sets [or types]; morphisms look like plain old maps between those sets;
+    epimorphisms look like surjections; group objects look like groups; and so on.
+
+    In particular, we can use the internal language to define when a
+    given~$\TT$-structure in~$\E$ is a model -- namely iff it looks like a
+    model from the internal point of view.
+
+    The internal language is implemented by the \emph{Kripke--Joyal semantics},
+    a translation procedure which converts formulas of the internal language
+    into external statements about the objects and morphisms of~$\E$. The slide
+    displays some of the translation rules in the case that~$\E$ is a
+    Grothendieck topos.
+
+    We can actually do mathematics internally because the Kripke--Joyal
+    semantics is sound with respect to intuitionistic logic: If~$\E \models
+    \varphi$ and if~$\varphi$ intuitionistically entails a further
+    formula~$\psi$, then~$\E \models \psi$.
+
+    An instructive special case is provided by the topos~$\Set$, because~$\Set
+    \models \varphi$ iff~$\varphi$ holds in the usual mathematical sense.
+    topos.
+  }
+
+  \jnote{2}{
+    The Kripke--Joyal semantics can be extended to interpret unbounded
+    quantification (``for all sets'' as opposed to ``for all elements of the
+    particular set~$X$'') and dependent types. With these extensions, we can
+    import all of everyday constructive impredicative mathematics into the
+    internal world of a topos.
   }
@@
   \end{enumerate}
+
+  \jnote{1}{
+    The generic object, the generic model of the theory which has exactly one
+    sort and no function symbols, relations symbols or axioms, appears to be
+    slightly indecisive: On the one hand, up to a double negation, it is a
+    subsingleton; on the other hand, it is infinite. This observation is due to
+    Carsten Butz and Peter Johnstone (\fixedhref{XXX}{Classifying toposes for
+    first-order-theories}).
+
+    The generic ring, the generic model of the theory of rings, is similarly
+    indecisive. It is infinite in the following sense:
+    \[ \forall x_1,\ldots,x_n \? U_\TT\_
+      \bigl(\forall y \? U_\TT\_ \bigvee_{i=1}^n (y = x_i)\bigr) \Rightarrow
+      1 = 0. \]
+
+    The theory of local rings is the quotient theory of the theory of rings
+    obtained by adding the axioms~$(1 = 0 \vdash \bot)$ and~$((\exists z\_
+    (x+y)z = 1) \seq{x,y} (\exists z\_ xz = 1) \vee (\exists z\_ yz = 1))$.
+    (Assuming the Boolean prime ideal theorem, a ring is local in this sense
+    iff it local in the usual sense (has exactly one maximal ideal).)
+
+    The first displayed property of the generic local ring illustrates
+    that nongeometric sequents need not be inherited by quotient theories.
+  }
+
+  \jnote{2}{
+    All of the displayed properties give rise to reduction techniques:
+    If we want to verify a geometric sequent for all rings, it suffices to
+    verify it for the generic ring; but the generic ring has additional
+    nongeometric properties not shared by every ring, such as the two displayed
+    ones. (This was for instance used by XXX.)
+
+    However, we face some challenges when pursuing these techniques, including
+    the following:
+    \begin{enumerate}\justifying
+      \item It is not easy to determine interesting and useful properties of
+      the generic model.
+      \item The set of validated nongeometric sequents changes slightly
+      unpredictably when passing to quotient theories. For instance, when
+      proving that a geometric sequent holds for all rings, we may assume that
+      any element is \emph{not not} zero. But we may not assume this
+      simplification if we want to verify a geometric sequent for all local
+      rings (and exploit this locality in the proof).
+      \item There is only so much we want to state and prove in full generality
+      for all rings, all local rings, all modules, and so on. We are often much
+      more interested in properties of particular mathematical objects.
+    \end{enumerate}
+  }
+
+  \jnote{3}{
+    The firstly-mentioned problem on the previous page is alleviated by the
+    Nullstellensatz presented in this talk, which gives a systematic and
+    universal source of nongeometric sequents validated by the generic model.
+    However, manual work is still required to reduce this set of sequents to a
+    smaller, manageable one consisting of memorable properties while hopefully
+    still preserving universality.
+  }
 \end{frame}

### 2019-04-24 slides-birmingham2019.tex
More notes

@@
       simplification if we want to verify a geometric sequent for all local
-      rings (and exploit this locality in the proof).
+      rings (and if we want to exploit the given locality in the proof).
       \item There is only so much we want to state and prove in full generality
@@
     still preserving universality.
+
+    To counter the third problem, it's prudent to consider geometric theories
+    which depend on a given mathematical object of interest. For instance,
+    given a ring~$A$, we can consider the theory of prime ideals of~$A$, of
+    complemented prime ideals, of filters, and so on. The classifying toposes
+    of these theories are of independent interest -- in fact they are sheaf
+    toposes over certain important locales in algebraic geometry -- and
+    nongeometric sequents validated by their generic models bundle nontrivial
+    information about~$A$. More details are on the following slide.
   }
 \end{frame}
+
+% XXX prudent
 
@@
 \begin{frame}{Affine schemes}
+  \jnote{1}{
+    A ring is \emph{local} iff every invertible sum contains an invertible
+    summand, that is if~$1 \neq 0$ and if~$x + y$ invertible implies~$x$
+    invertible or~$y$ invertible. Assuming the axiom of choice, this elementary
+    definition is equivalent to the textbook definition of a local ring (a ring
+    with exactly one maximal ideal). A ring homomorphism is \emph{local} iff it
+    reflects invertibility.
+
+    The notion of a \emph{filter} in a ring~$A$ is a direct axiomatisation of
+    what classically would be the complement of a prime ideal. The filter
+    axioms are: $1 \in F$, $(xy \in F) \Leftrightarrow ((x \in F) \wedge (y \in
+    F))$, $\neg(0 \in F)$, $(x+y \in F) \Rightarrow ((x \in F) \vee (y \in F))$.
+
+    A \emph{free local ring} over~$A$ is a homomorphism into a local
+    ring~$A'$ such that any homomorphism into a local ring~$R$ factors uniquely
+    over~$A'$ via a local homomorphism.
+
+    For any particular local ring~$A \xrightarrow{f} R$, the
+    localisation~$A[S^{-1}]$ with~$S = f^{-1}[R^\times]$ is a local ring which
+    fits into the displayed diagram. However, in general there is no single
+    choice of~$S$ which would work for any local ring~$R$. Indeed, classically
+    one can show that a free local ring over~$A$ exists if and only if~$A$
+    contains exactly one prime ideal, in which case~$A$ itself is the free
+    local ring.
+  }
+
+  \jnote{2-}{
+    If we want a free local ring to exist for any ring~$A$, we have to broaden
+    our notion of existence and embrace rings which live in toposes other
+    than~$\Set$. There is a notion of a homomorphism between rings living in
+    arbitrary toposes, and using this notion one can verify:
+
+    The free local ring over~$A$ can be built in the classifying topos of
+    filters of~$A$, as the localisation (in that topos) of~$A$ at the generic
+    filter (more precisely, the localisation of the mirror image of~$A$ in that
+    topos, that is the constant sheaf~$\underline{A}$).
+  }
+
   Let~$A$ be a ring. Is there a \hil{free local ring}~$A \to A'$ over~$A$?
@@
       \small\justifying
-      For a fixed ring~$R$, the localization $A' \defeq A[S^{-1}]$ with $S \defeq
+      For a fixed ring~$R$, the localisation $A' \defeq A[S^{-1}]$ with $S \defeq
       f^{-1}[R^\times]$ would do the job. ($S$ is a \emph{filter}.)
@@
 % * Why we are interested in the generic models themselves
-%   * Universal localization
+%   * Universal localisation
 %   * Grothendieck's generic freeness lemma

### 2019-04-24 slides-birmingham2019.tex
More notes

@@
 
-  \jnote{2-}{
+  \jnote{2}{
     If we want a free local ring to exist for any ring~$A$, we have to broaden
@@
     filters of~$A$, as the localisation (in that topos) of~$A$ at the generic
-    filter (more precisely, the localisation of the mirror image of~$A$ in that
-    topos, that is the constant sheaf~$\underline{A}$).
+    filter. (More precisely, as the localisation of the mirror image of~$A$ in that
+    topos, that is the constant sheaf~$\underline{A}$.)
+
+    The classifying topos of filters of~$A$ coincides with the topos of sheaves
+    over what's called the \emph{spectrum of~$A$} in algebraic geometry, and
+    under this equivalence the free local ring coincides with the structure
+    sheaf of spectrum. In fact the classifying topos serves as a good
+    constructive substitute for the classical spectrum construction,
+    enjoying the expected universal property even if the axiom of choice is not
+    available, which is why it is simply denoted~``$\Spec(A)$'' on the slide.
+
+    (The classifying topos of prime ideals of~$A$ is also interesting; it
+    coincides with the topos of sheaves over the spectrum of~$A$ equipped with
+    the constructible topology.)
+  }
+
+  \jnote{3}{
+    Assuming the Boolean prime ideal theorem, the geometric sequents
+    validated by~$A^\sim$ are easy to describe: They are precisely those which
+    are validated by all the stalks~$A_\ppp$ of~$A$.
+
+    But~$A^\sim$ enjoys further unique properties which are not shared by the
+    stalks of~$A$, other localisations of~$A$, quotients of~$A$ or indeed any
+    reasonable construction. Three of these are displayed on the slide.
+    (A ring is \emph{anonymously Noetherian} iff each of its ideals is
+    \emph{not not} finitely generated. Textbook proofs of Hilbert's basis theorem
+    are constructively acceptable for this Noetherian condition.)
+
+    The object~$A^\sim$ strikes a fine balance: On the one hand, it is still
+    close to~$A$, so that information learned about~$A^\sim$ teaches us
+    about~$A$; on the other hand, it enjoys a unique properties rendering it
+    simpler than~$A$.
+
+    This balance allows for a simple and conceptually satisfying proof of
+    \emph{Gro\-then\-dieck's generic freeness lemma}, an important theorem in
+    algebraic geometry. Details can be found in
+    \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-padova2018.pdf}{this
+    set of slides}.
   }
@@
 \begin{frame}{A systematic source of nongeometricity?}
+  \jnote{1-2}{
+    Mimicking the synthetic approach to differential geometry, synthetic
+    algebraic geometry is a framework for algebraic geometry in which schemes
+    can be modelled by plain old sets, morphisms of schemes by plain old maps
+    between those sets, group schemes by plain old groups, and so on. Unlike
+    its close cousin, it is far less developed; some first steps are outlined
+    in Sections~19 and~20 of
+    \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these notes}.
+
+    Synthetic algebraic geometry is carried out internally to the \emph{big
+    Zariski topos} of a given base scheme; in the special case that the base
+    scheme is the terminal scheme~$\Spec(\ZZ)$, this topos is just the
+    classifying topos of local rings. The relevant generic model living in the
+    big Zariski topos is denoted~``$\affl$'' because it coincides with the
+    functor of points of the affine line.
+  }
+
+  \jnote{2}{
+    Marc Bezem, Ulrik Buchholtz and Thierry Coquand answered in their 2017 paper
+    \fixedhref{https://arxiv.org/abs/1712.07743}{Syntactic forcing
+    models for coherent logic} Gavin Wraith's question in the negative. (As
+    Thierry remarked during the talk, this is even if one takes care to phrase
+    the question in a way to exclude the trivial counterexamples given by any
+    instance of the law of excluded middle in the language of~$\TT$.) If the
+    answer had been positive, this would have given a neat, if somewhat hard to
+    check, characterisation of the formulas validated by the generic model.
+  }
+
+  \jnote{3}{
+    The topos-theoretic \emph{Nullstellensatz}, to be presented on the next
+    slides, answers the displayed three questions in the affirmative.
+
+    Briefly, the Nullstellensatz is a certain statement in the language of a
+    given geometric theory~$\TT$ which is
+    \begin{itemize}\justifying
+      \item validated by the generic~$\TT$-model,
+      \item typically not validated by other~$\TT$-models, and
+      \item such that any statement validated by the generic~$\TT$-model can be
+      deduced, in intuitionistic logic, from the axioms of~$\TT$ and the
+      Nullstellensatz.
+    \end{itemize}
+
+    We believe that this characterisation is as explicit as it can get, but
+    would be delighted to be surprised by a future improvement.
+  }
+
   \justifying
@@
 	\scriptsize
-        Gavin Wraith. Some recent developments in topos theory. \\ In:
+        Gavin Wraith. \emph{Some recent developments in topos theory.} \\ In:
         Proc.\@ of the ICM (Helsinki, 1978).
@@
 \begin{frame}{A topos-theoretic Nullstellensatz}
+  \jnote{1-2}{
+    By~$\ul{\TT}$ we mean the geometric theory internal to~$\Set[\TT]$ obtained
+    by pulling back the set of sorts of~$\TT$, the set of function symbols and so on
+    along the unique geometric morphism~$\Set[\TT] \to \Set$. For instance,
+    if~$\TT$ is the theory of rings, then from the internal point of view
+    of~$\Set[\TT]$, the theory~$\ul{\TT}$ will again be the theory of rings.
+
+    The theory~$\ul{\TT}/U_\TT$ will be defined on the next slide. It is a certain
+    geometric theory internal to~$\Set[\TT]$.
+
+    The asterisks in \emph{geometric$^\star$ sequent} and
+    \emph{provability$^\star$} indicate that any infinities used to index
+    disjunctions have to be come from the base topos. This restriction is an important
+    subtlety, though not vital to this talk. If~$\TT$ is a coherent
+    theory, then for coherent sequents there is no difference between
+    provability in coherent logic, provability in geometric logic and
+    provability$^\star$.
+  }
+
+  \jnote{2}{
+    The algebraic Nullstellensatz states that, in some cases, algebraic truths
+    are witnessed by explicit \emph{algebraic certificates} -- syntactical
+    objects giving a priori reasons for why a given truth is to be expected.
+
+    In the topos-theoretic Nullstellensatz, algebraic truths are replaced by
+    arbitrary truths of the generic model, subject only to the condition that
+    they can be expressed as a geometric sequent, and algebraic certificates
+    are replaced by \emph{logical certificates}: proofs.
+  }
+
+  \jnote{3}{
+    While, as stated on slide 1/20, the generic model~$U_\TT$ is a
+    conservative~$\TT$-model, the classifying topos~$\Set[\TT]$ does not
+    believe this fact. That is, the statement~``$U_\TT$ is a
+    conservative~$\ul{\TT}$-model'' is not true internally to~$\Set[\TT]$.
+    What is true is the modified statement~``$U_\TT$ is a
+    conservative$^\star$~$\ul{\TT}/U_\TT$-model''.
+  }
+
   \nullstellensatz
@@
   of~$\TT$-homomorphisms.
+
+  \jnote{1}{
+    Just as locales internal to a topos~$\E$ can be externalised to yield
+    localic geometric morphisms into~$\E$, internal Grothendieck toposes can be
+    externalised to yield bounded geometric morphisms. Since the composition of
+    bounded geometric morphisms is bounded, the externalisation of a
+    Grothendieck topos internally to a Grothendieck topos is itself a
+    Grothendieck topos, hence the classifying topos of some geometric theory.
+
+    Constructing internally to~$\Set[\TT]$, where~$\ul{\TT}/U_\TT$ is just an
+    ordinary geometric theory, the classifying topos of that theory, and then
+    externalising the resulting Grothendieck topos results in the classifying
+    topos of~$\TT$-homomorphisms. There are two canonical geometric morphisms
+    from this topos to~$\Set[\TT]$, the morphism computing the domain and the
+    morphism computing the codomain, and the morphism obtained by the
+    externalisation procedure is the former.
+  }
 \end{frame}
@@
   in~$U_\TT[X]/(f)$, that is~$g = hf$ for some~$h \? U_\TT[X]$.
+
+  \jnote{1}{
+    This slide gives two examples how to use the Nullstellensatz to deduce
+    properties of the generic model. A couple of remarks are in order.
+
+    The Nullstellensatz is trivial for sequents~$\sigma$ of the form~$(\top
+    \vdash \psi)$. The Nullstellensatz is only interesting in case
+    that~$\sigma$ has a nontrivial antecedent or is set in a nonempty context.
+
+    Since the converse direction in the Nullstellensatz also holds
+    (because~$U_\TT$ is a~$\ul{\TT}/U_\TT$-model), the
+    statements~$\speak{$\sigma$ holds for~$U_\TT$}$
+    and~$\speak{$\ul{\TT}/U_\TT$ proves$^\star$ $\sigma$}$ are equivalent. This
+    equivalence is intriguing from a logical point of view, since the former
+    statement is a geometric implication while the latter can be put as a
+    geometric formula. (Up to a subtle issue indicated on the next slide.)
+
+    To apply the Nullstellensatz, no description of a site defining~$\Set[\TT]$
+    is required.
+
+    Often when using the Nullstellensatz, we go from an (assumed) truth
+    of~$U_\TT$ via provability$^\star$ to another model~$M$
+    of~$\ul{\TT}/U_\TT$. That is, we use provability$^\star$ as a (one-way)
+    \emph{bridge}:
+    \[ \xymatrix{
+      & \text{$\ul{\TT}/U_\TT$ proves$^\star$ $\sigma$} \\
+      \text{$\sigma$ holds for $U_\TT$} \ar@/^1.1pc/@{<=>}[ur] &&
+      \text{$\sigma$ holds for $M$} \ar@/_1.1pc/@{<=}[ul]
+    } \]
+  }
 \end{frame}
@@
   \end{quote}
-  \vspace*{-0.3em}
+  \vspace*{-1.6em}
 
@@
   the higher-order Nullstellensatz scheme.
+
+  \jnote{1}{
+    The first theorem states that the source provided by the Nullstellensatz is
+    exhaustive. The notion of provability$^{\star\star}$ is a strengthening of
+    the notion of provability$^\star$, which in turn is a strengthening of the
+    ordinary notion of provability in geometric logic. We are using it here
+    because while the notion of provability$^\star$ can be expressed in the
+    internal language of a topos, it cannot be expressed in intuitionistic
+    logic.
+
+    The second theorem provides a useful variant of the Nullstellensatz. Its
+    assumptions are for instance satisfied if~$\TT$ is the theory of rings
+    and~$\TT'$ is the theory of local rings. When applicable, it can be used to
+    avoid doubly-internal toposes. It also explains, for instance, why in the
+    formulation of synthetic quasicoherence no local rings appear even though
+    the relevant topos is the classifying topos of local rings.
+
+    The third and fourth theorems generalise the Nullstellensatz to the
+    higher-order setting. The map~$\mathrm{ev}$ takes (the equivalence class
+    of) a~$\ul{\TT}/U_\TT$-provably$^\star$ geometric$^\star$ formula~$\theta$
+    in one free variable to the subset~$\{ x \? U_\TT \,|\, \theta(x) \}$.
+  }
+
+  \jnote{2}{
+    XXX future work
+  }
 \end{frame}}

### 2019-05-01 slides-birmingham2019.tex
Notes almost finished

@@
 
+  For instance, if~$\TT$ is the theory of rings, then the initial model
+  is~$\ZZ$. This model validates some geometric sequents which are not
+  validated by all rings, for instance~$(x^2 = 0 \seq{x:R} x = 0)$ or $(1 = 0
+  \vdash \bot)$.
+
   In general, the generic model cannot be realised as a set-based model (with
@@
   required: The generic model lives in the topos of set-valued
-  sheaves over~$\C_\TT$.
-
-  A morphism~$A = \{\vec x\_ \varphi\} \to \{\vec y\_ \psi\} = B$ in~$\C_\TT$ is the
+  sheaves over~$\C_\TT$.}
+
+  \jnote{2}{A morphism~$A = \{\vec x\_ \varphi\} \to \{\vec y\_ \psi\} = B$ in~$\C_\TT$ is the
   equivalence class (modulo provable equivalence) of a geometric
@@
   The slides experiment with using the symbol~``$\yon$'' for the Yoneda
-  embedding~$\C_\TT \to \Set[\TT]$, as in \fixedhref{XXX}{Elements
+  embedding~$\C_\TT \to \Set[\TT]$, as in \fixedhref{http://www.math.jhu.edu/~eriehl/elements.pdf}{Elements
   of~$(\infty,1)$-category theory} by Emily Riehl and Dominic Verity.}
 
-  \jnote{2}{The special case that the generic model of a theory~$\TT$ can be realised as
+  \jnote{3}{The special case that the generic model of a theory~$\TT$ can be realised as
   a model in~$\Set$ occurs iff~$\TT$ is Morita-equivalent to the
@@
 
-    In particular, we can use the internal language to define when a
-    given~$\TT$-structure in~$\E$ is a model -- namely iff it looks like a
+    In particular, we can use the internal language to define what it means for a
+    given~$\TT$-structure in~$\E$ to be a model -- namely iff it looks like a
     model from the internal point of view.
 
-    The internal language is implemented by the \emph{Kripke--Joyal semantics},
+    The internal language can be implemented by the \emph{Kripke--Joyal semantics},
     a translation procedure which converts formulas of the internal language
@@
     \models \varphi$ iff~$\varphi$ holds in the usual mathematical sense.
-    topos.
   }
@@
     quantification (``for all sets'' as opposed to ``for all elements of the
-    particular set~$X$'') and dependent types. With these extensions, we can
+    particular set~$X$'') and dependent types. The former are for instance
+    required to express universal properties (``for all groups'', ``for all
+    rings''), and the latter are all over the place, even if their use might
+    not be particularly highlighted.
+
+    With these extensions, we can
     import all of everyday constructive impredicative mathematics into the
     internal world of a topos.
+
+    Some illustrations of working with the internal language can be found in these
+    sets of slides:
+    \begin{itemize}
+      \item
+      \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-leipzig2018.pdf}{Slides
+      for Jürgen Jost's group seminar at the MPI Leipzig}
+      \item
+      \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-como2018.pdf}{Slides
+      for \emph{Toposes in Como}}
+      (recording
+      \fixedhref{https://www.youtube.com/watch?v=SuELfC5P_R8}{available})
+    \end{itemize}
+    A longer exposition, with pointers to the literature, can be found in Section~2 of
+    \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these notes}.
   }
@@
     subsingleton; on the other hand, it is infinite. This observation is due to
-    Carsten Butz and Peter Johnstone (\fixedhref{XXX}{Classifying toposes for
-    first-order-theories}).
+    Carsten Butz and Peter Johnstone
+    (\fixedhref{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.16.4102&rep=rep1&type=pdf}{Classifying toposes for
+    first-order theories}).
 
@@
     nongeometric properties not shared by every ring, such as the two displayed
-    ones. (This was for instance used by XXX.)
-
-    However, we face some challenges when pursuing these techniques, including
+    ones. (This was for instance used
+    \fixedhref{https://www.sciencedirect.com/science/article/pii/0022404976900025}{by
+    Anders Kock} and
+    \fixedhref{https://link.springer.com/chapter/10.1007\%2FBFb0061835}{by
+    Gonzalo Reyes}.)
+
+    However, we face some challenges when pursuing these reduction techniques, including
     the following:
@@
     of these theories are of independent interest -- in fact they are sheaf
-    toposes over certain important locales in algebraic geometry -- and
+    toposes over certain important spaces in algebraic geometry -- and
     nongeometric sequents validated by their generic models bundle nontrivial
@@
 \end{frame}
-
-% XXX prudent
 
@@
     close to~$A$, so that information learned about~$A^\sim$ teaches us
-    about~$A$; on the other hand, it enjoys a unique properties rendering it
+    about~$A$; on the other hand, it enjoys unique properties rendering it
     simpler than~$A$.
@@
     Thierry remarked during the talk, this is even if one takes care to phrase
-    the question in a way to exclude the trivial counterexamples given by any
-    instance of the law of excluded middle in the language of~$\TT$.) If the
+    the question in a way to exclude the trivial counterexamples given by
+    instances of the law of excluded middle in the language of~$\TT$.) If the
     answer had been positive, this would have given a neat, if somewhat hard to
-    check, characterisation of the formulas validated by the generic model.
+    use in practice, characterisation of the formulas validated by the generic model.
   }
@@
     We believe that this characterisation is as explicit as it can get, but
-    would be delighted to be surprised by a future improvement.
+    would be delighted to be surprised by a future improvement. We stumbled on
+    it by playing with the synthetic quasicoherence statement, not least thanks
+    to encouragement by Alexander Oldenziel. However the route from that
+    statement to the Nullstellensatz is not quite direct; it turns out that
+    synthetic quasicoherence is a corollary of a specialisation of a
+    higher-order version of the Nullstellensatz to Horn theories.
   }
@@
     in one free variable to the subset~$\{ x \? U_\TT \,|\, \theta(x) \}$.
+
+    Details on all of this are slowly emerging at
+    \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-qcoh.pdf}{this
+    preprint}.
   }

### 2019-05-02 slides-birmingham2019.tex
Notes finished

@@
   \par
-
-  \jnote{1}{\textcolor{red}{This set of slides will be updated soon with more
-  annotations.} Right now important references are not properly mentioned, including:
-  \begin{itemize}
-    \item Marc Bezem, Ulrik Buchholtz and Thierry Coquand. \emph{Syntactic forcing
-    models for coherent logic.}
-    \item Olivia Caramello. \emph{Universal models and definability.}
-    \item Carsten Butz and Peter Johnstone. \emph{Classifying toposes for
-    first-order theories.}
-  \end{itemize}}
 \end{frame}}
@@
 \subsection{The mystery of nongeometric sequents}
-
-% \begin{document}
 
@@
     The theory of local rings is the quotient theory of the theory of rings
-    obtained by adding the axioms~$(1 = 0 \vdash \bot)$ and~$((\exists z\_
-    (x+y)z = 1) \seq{x,y} (\exists z\_ xz = 1) \vee (\exists z\_ yz = 1))$.
-    (Assuming the Boolean prime ideal theorem, a ring is local in this sense
+    obtained by adding the axioms
+    \[ (1 = 0 \vdash \bot) \ \text{and}\ ((\exists z\_
+    (x+y)z = 1) \seq{x,y} (\exists z\_ xz = 1) \vee (\exists z\_ yz = 1)). \]
+    (Assuming the axiom of choice, a ring is local in this sense
     iff it local in the usual sense (has exactly one maximal ideal).)
@@
   \justifying
-  \textbf{Theorem.} A first-order formula holds for~$U_\TT$ iff it is
+  \textbf{Theorem 1.} A first-order formula holds for~$U_\TT$ iff it is
   intuitionistically provable from the axioms of~$\TT$ and
@@
 
-  \textbf{Theorem.}
+  \textbf{Theorem 2.}
   Let~$\TT'$ be a quotient theory
@@
   to~$\Set[\TT']$:
+
+  \vspace*{-0.4em}
   \begin{quote}
@@
   \end{quote}
-  \vspace*{-1.6em}
-
-  \textbf{Theorem.}
+
+  \textbf{Theorem 3.}
   % Let~$\mathrm{Form}^\star_{x\?X}(\ul{\TT}/U_\TT)/(\dashv\vdash_{x\?X})$ be the~$\Set[\TT]$-object of
@@
   % if~$\ul{\TT}/U_\TT$ proves them equivalent.
-  The morphism~$\mathrm{ev}$ is an isomorphism.
-  \[ \mathrm{ev} : \mathrm{FunctFormulas}^\star(\ul{\TT}/U_\TT)/(\dashv\vdash)
-  \longrightarrow P(U_\TT) \]
-
-  \textbf{Theorem.} A higher-order formula holds for~$U_\TT$ iff it is
+  The morphism~$\mathrm{ev}$ below is an isomorphism.
+  \begin{equation}\label{eq:hon}\tag{$\star$}
+    \mathrm{ev} : \mathrm{FunctFormulas}^\star(\ul{\TT}/U_\TT)/(\dashv\vdash) \longrightarrow P(U_\TT)
+  \end{equation}
+
+  \textbf{Theorem 4.} A higher-order formula holds for~$U_\TT$ iff it is
   provable in intuitionistic higher-order logic from the axioms of~$\TT$ and
-  the higher-order Nullstellensatz scheme.
+  the higher-order Nullstellensatz scheme~($\star$).
 
   \jnote{1}{
-    The first theorem states that the source provided by the Nullstellensatz is
+    Theorem~1 states that the source provided by the Nullstellensatz is
     exhaustive. The notion of provability$^{\star\star}$ is a strengthening of
@@
 
-    The second theorem provides a useful variant of the Nullstellensatz. Its
+    Theorem~2 provides a useful variant of the Nullstellensatz. Its
     assumptions are for instance satisfied if~$\TT$ is the theory of rings
@@
 
-    The third and fourth theorems generalise the Nullstellensatz to the
-    higher-order setting. The map~$\mathrm{ev}$ takes (the equivalence class
+    Theorems~3 and~4 generalise the Nullstellensatz to the
+    higher-order setting. The map~$\mathrm{ev}$ maps (the equivalence class
     of) a~$\ul{\TT}/U_\TT$-provably$^\star$ geometric$^\star$ formula~$\theta$
@@
 
-    Details on all of this are slowly emerging at
-    \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-qcoh.pdf}{this
-    preprint}.
+    Written details on all of this are
+    \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-qcoh.pdf}{slowly
+    emerging}.
   }
@@
   \jnote{2}{
-    XXX future work
+    The Nullstellensatz is related to several precursors. A corollary of the
+    Nullstellensatz is that, over the first-order theory validated by~$U_\TT$, any
+    first-order formula is in fact logically equivalent to a geometric formula.
+    This corollary has already been observed by Carsten Butz and Peter
+    Johnstone in their paper
+    \fixedhref{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.16.4102&rep=rep1&type=pdf}{Classifying
+    toposes for first-order theories} (Lemma~4.2 there). At that point, a
+    characterisation of the first-order formulas in the general case, of the
+    form as in Theorem~1, was still missing.
+
+    Theorem~3 is a relativisation of Olivia Caramello's completeness theorem,
+    Theorem~2.4(ii) in her paper
+    \fixedhref{https://arxiv.org/abs/0906.3061}{Universal models and
+    definability}. The passage from the external to the internal phrasing
+    requires going from~$\TT$ to~$\ul{\TT}/U_\TT$.
+  }
+
+  \jnote{2}{
+    Plans for the future include:
+    \begin{itemize}\justifying
+      \item Developing an Agda library for dealing with the internal language
+      of toposes and related kinds of categories, employing Agda's
+      meta-programming facilities; with such a library at hand,
+      formalising the Nullstellensatz in Agda.
+      \item Exploring Nullstellensatz-style results for arithmetic universes.
+      \item Applying the Nullstellensatz in constructive algebra and algebraic
+      geometry, along the lines of generic freeness and synthetic algebraic
+      geometry. Most of the toposes in geometric use are actually uncharted
+      territory from a logical point of view.
+    \end{itemize}
   }

### 2019-05-03 paper-qcoh.tex
Slides almost finished
to do
More work on the introduction

@@
 \usepackage[english]{babel}
-\usepackage{amsthm,mathtools,stmaryrd,amssymb,graphicx}
+\usepackage{amsthm,mathtools,stmaryrd,amssymb,graphicx,color}
 \usepackage{booktabs}
@@
   the generic model, and characterize the first-order and higher-order formulas validated by the
-  genetric model.
+  generic model.
 \end{abstract}
@@
 
+\begin{center}
+  \textcolor{red}{\textbf{-- rough draft --}}
+
+  \emph{comments welcome at ingo.blechschmidt@univr.it}
+\end{center}
+
 \section{Introduction}
@@
 \paragraph{Generic models} Let~$\TT$ be a geometric theory, such as the theory
-of rings, of local rings or of intervals. We follow Caramello's
+of rings, the theory of local rings or the theory of intervals. We follow Caramello's
 terminology~\cite{caramello:tst} to mean by \emph{geometric theory} a system
 given by a set of sorts, a set of finitary function symbols, a set of finitary
-relation symbols and a set of axioms, consisting of geometric sequents
+relation symbols and a set of axioms consisting of geometric sequents
 (sequents of the form~$\varphi \seq{\vec x} \psi$ where~$\varphi$ and~$\psi$
@@
 
-A fundamental result is that there is a \emph{generic model}~$U_\TT$ of~$\TT$,
-a model such that for any geometric sequent~$\sigma$, the following notions
+A fundamental result is that there is a \emph{generic model}~$U_\TT$ of~$\TT$.
+This model is \emph{conservative} in that
+for any geometric sequent~$\sigma$, the following notions
 coincide:
@@
 One could argue that it is this model which we
-implicitly refer to when we utter the phrase ``Let~$M$ be a~$\TT$-model.''. It can
+implicitly refer to when we utter the phrase ``Let~$M$ be
+a~$\TT$-model.''.\footnote{For instance, this point of view is fundamental to
+the slogan \emph{continuity is geometricity}, as stressed by
+Vickers~\cite{vickers:continuity}.}
+It can
 typically not be realized as a set-theoretic model, consisting of a set for
@@
 and the generic model is the functor~$\affl$ of points of the affine line, the
-functor which maps any (l.o.f.p.) scheme~$X$ to~$\Hom(X,\mathbb{A}^1) =
+functor which maps any (l.o.f.p.) scheme~$X$ to~$\Hom_\mathrm{Sch}(X,\mathbb{A}^1) =
 \O_X(X)$.
 
-From the point of view of the topos, the ring object~$\affl$ is not only a
+From the point of view of the Zariski topos, the ring object~$\affl$ is not only a
 local ring, but even a field in the sense that
@@
 about local rings which is expressible as a geometric sequent,
-suppose without loss of generality that the given ring is a
-field. This observation is due to Kock~\cite{kock:univ-proj-geometry}, who exploited it to
+suppose without loss of generality that the field axioms holds.
+This observation is due to Kock~\cite{kock:univ-proj-geometry}, who exploited it to
 develop projective geometry over local rings, and was further used by Reyes to
@@
 
-A related nongeometric sequent is valid in the little Zariski topos of the
-spectrum of a ring~$A$, the classifying topos of local localizations of~$A$.
-If~$A$ is reduced, the generic model validates the dual condition that any
+We surmise that many more reduction techniques along these lines exist for
+other kinds of algebraic objects. However, when actually using such techniques
+in practice, we face the challenge that while we can use them to prove results
+about \emph{all} local rings, \emph{all} modules and so on, it is difficult to
+incorporate specific information about a particular local ring or a particular
+module at hand. This difficulty is compounded by the fact that interesting
+nongeometric properties are typically not inherited by the generic model of
+quotient theories -- for instance the generic ring validates the formula~$\forall
+x\?U_\TT\_ \neg\neg(x = 0)$ while the generic local ring does not.
+
+Hence it is useful to turn to geometric theories which refer to a given
+mathematical object. For instance, given a ring~$A$, there is the theory of
+local localizations of~$A$, and its classifying topos is known in algebraic
+geometry as the little Zariski topos of~$A$, the topos of sheaves over the
+spectrum of~$A$. If~$A$ is reduced, this topos validates the dual condition that any
 noninvertible element is zero. This property has been used to give a short and
@@
 schemes instead of restricting to the field of complex numbers.
+
+Nontrivial nongeometric sequents are not an exclusive feature of sheaf toposes.
+The generic model of a theory of presheaf type typically also validates such
+sequents. For instance, all of the aforementioned properties of the generic
+local ring are shared by the generic ring, which lives in the presheaf
+topos~$[\mathrm{Ring}_\mathrm{fp}, \Set]$.
 \medskip
 
-\paragraph{Characterizing nongeometric sequents} Referring to one of the previous
-examples, Tierney remarked around the time that those sequents were first
+\paragraph{Characterizing nongeometric sequents} Referring to the field condition in the little Zariski topos,
+Tierney remarked around the time that those sequents were first
 studied that ``[it] is surely important, though its precise significance is
@@
 \end{quote}
-This question was recently answered in the negative by Bezem, Buchholtz and
-Coquand~\cite{bezem-buchholtz-coquand:syntactic-forcing-models}; hence the
-characterization we propose is necessarily more nuanced.
+Because classical logic is conservative over intuitionistic logic for geometric
+sequents, this question has a trivial negative answer: No, any instance of the
+law of excluded middle over the signature of~$\TT$ is~$\TT$-redundant but
+typically not validated by the generic~$\TT$-model. Moreover,
+Bezem, Buchholtz and
+Coquand recently showed that the answer is still negative even if appropriate
+care is taken to exclude these
+counterexamples~\cite{bezem-buchholtz-coquand:syntactic-forcing-models}. Hence our
+characterization of the first-order theory validated by the generic~$\TT$-model
+is necessarily more nuanced.
 
@@
 Intuitively, the problem is that while the meaning of~$\speak{$\ul{\TT}$
-proves~$\sigma$}$ is fixed, the meaning of~$\speak{$\sigma$ holds for~$U_\TT$}$ can vary
-with the slice. This problem can be solved by passing from~$\ul{\TT}$ to a
+proves~$\sigma$}$ is fixed, the meaning of~$\speak{$\sigma$ holds for~$U_\TT$}$ varies
+with the slice, as~$U_\TT$ changes form on different stages. This mismatch is solved by passing from~$\ul{\TT}$ to a
 varying theory, the internal theory~$\ul{\TT}/U_\TT$ defined in
@@
 \text{A geometric$^*$ sequent~$\sigma$ holds for~$U_\TT$ if and only
-if~$\ul{\TT}/U_\TT$ proves$^*$~$\sigma$.}
+if~$\ul{\TT}/U_\TT$ proves~$\sigma$.}
 \end{equation}
@@
 rings and let~$\sigma$ be the sequent~$(f(x) = 0 \seq{x} g(x) = 0)$ for some
-polynomials~$f$ and~$g$. To say that~$\sigma$ holds for~$U_\TT$ amounts to
+polynomials~$f$ and~$g$ with coefficients in~$U_\TT$. To say that~$\sigma$ holds for~$U_\TT$ amounts to
 saying that any zero~$x \? U_\TT$ of~$f$ is also a zero of~$g$, and to say
@@
 which we formulate as Theorem~\ref{thm:nullstellensatz-horn}. This substitute
-substantially broadens the scope of the Nullstellensatz.
+broadens the scope of the Nullstellensatz and is, whenever applicable, useful
+for simplifying computations.
 
@@
 
+It is not entirely obvious how to correctly phrase
+axiom~\eqref{eq:nullstellensatz0} in first-order logic, an issue
+which we will discuss when we restate Theorem~\ref{thm:characterization0} in
+the body of this note (Theorem~\ref{thm:characterization}).
+
 Theorem~\ref{thm:characterization0} characterizes the first-order formulas
@@
 
+% XXX: mention internal language in general, and SDG?
+\paragraph{Related work} The topos-theoretic Nullstellensatz is related to
+several precursors. A corollary of the Nullstellensatz is that, over the
+first-order theory validated by~$U_\TT$, any first-order formula is in fact
+logically equivalent to a geometric formula. This corollary has already been
+observed by Butz and Johnstone~\cite[Lemma~4.2]{butz-johnstone:first-order}. At
+that point, a characterisation of the first-order formulas in the general case,
+of the form as in Theorem~\ref{thm:characterization0}, was still missing.
+
+The higher-order variant of our Nullstellensatz
+(Theorem~\ref{thm:definability}) relativizes Cara\-mel\-lo's completeness
+theorem~\cite[Theorem~2.4(ii)]{caramello:definability}. Her theorem is the
+external statement that any subobject of~$U_\TT$ is given by the interpretation
+of a geometric formula over the signature of~$\TT$; our relativization states
+that, internally to~$\Set[\TT]$, any subset of~$U_\TT$ is given by a
+geometric$^\star$ formula over the signature of~$\ul{\TT}/U_\TT$. As in the
+first-order case, the passage from the external to the internal phrasing
+necessitates the switch from~$\TT$ to~$\ul{\TT}/U_\TT$.
+
+The Nullstellensatz yields for a given (perhaps conditional) truth about the
+generic model a proof modulo the axioms of~$\ul{\TT}/U_\TT$. As illustrated in
+Section~\ref{sect:applications}, the procedure we typically use to extract
+fruitful information from such a proof is to apply it to a
+further~$\ul{\TT}/U_\TT$-model which we specifically construct for the purposes
+at hand. Hence we go from truth in~$U_\TT$ via provability to truth in a
+further model, that is, we use provability as a (one-way) \emph{bridge}:
+\[ \xymatrix{
+  & \text{$\ul{\TT}/U_\TT$ proves $\sigma$} \\
+  \text{$\sigma$ holds for $U_\TT$} \ar@/^1.1pc/@{<=>}[ur] &&
+  \text{$\sigma$ holds for $M$} \ar@/_1.1pc/@{<=}[ul]
+} \]
+This perspective is inspired by Caramello's research
+program~\cite{caramello:tst}, though it is not an instance of her main technical
+device for establishing bridges, namely exploiting the fact that a single topos
+may admit descriptions using quite different sites.
+\medskip
+
+\paragraph{Outlook} The Nullstellensatz yields a description of the first-order
+and higher-order theory validated by the generic model of a geometric theory.
+As already indicated, with more examples and details to be given in
+Section~\ref{sect:applications}, this description explains and puts into
+perspective several established results and observations.
+
+From the point of view of applications, the main task for the future is to
+explore the description in the case of particular geometric theories of
+interest. Just as the reduction technique ``any reduced ring is a field'' --
+valid because the generic localization of a reduced ring is a field -- enabled
+a short and simple proof of Grothendieck's generic freeness
+lemma~\cite[Section~11.5]{blechschmidt:phd}, more reduction techniques along
+these lines should exist in a wide range of subjects. We hope that with the
+Nullstellensatz at hand, discovering such techniques can progress in a more
+systematic fashion.
+
+From the point of view of theory, the appropriate context for the
+Nullstellensatz should be determined. We formulate the Nullstellensatz in the
+context of geometric theories and their classifying Grothendieck toposes, but
+there should be analogues for the Nullstellensatz in other contexts where it
+makes sense to speak of ``classifying gadgets'', and these analogues should
+shed light on the topos-theoretic version and ideally even explain it from deeper
+principles. In particular, preliminary computations indicate that there is a
+version of the Nullstellensatz for arithmetic universes, the predicative cousin
+of toposes introduced by Joyal~\cite{XXX} and recently an important object of
+consideration by Maietti and Vickers~\cite{XXX}.
+\medskip
+
 \paragraph{Outline} In Section~\ref{sect:review}, we review background on the
@@
 \paragraph{Acknowledgments} XXX
+% don't forget: Thierry, and Steve and Martín for 6WFTop; also of course
+% Alexander. Also include grant statement.
 
@@
 
-Johnstone~\cite{XXX} lists several nongeometric sequents(XXX) holding in the
+Butz and Johnstone~\cite{butz-johnstone:first-order} list several nongeometric sequents holding in the
 classifying topos of the theory~$\OO$ of objects, the theory with a single sort and
@@
 
-* XXX Discuss related work, especially: Olivia 6.1.3, ...
-
 * XXX Cite Filip Bár?
@@
 * XXX restrict to finitary first-order formulas in the scholium
+
+* XXX Upgrade Horn section to cartesian theories?

### 2019-05-04 paper-qcoh.tex
to do

@@
 \begin{enumerate}
-\item as objects ``geometric formulas in contexts''~$\{x_1\?X_1,\ldots,x_n\?X_n\_
+\item as objects \emph{geometric formulas in contexts}~$\{x_1\?X_1,\ldots,x_n\?X_n\_
 \varphi\}$ where~$\varphi$ is a geometric formula over the signature of~$\TT$
 in the displayed context; \smallskip
-\item as set of morphisms~$\Hom_{\C_\TT}(\{\vec x\_
-\varphi\}, \{\vec y\_ \psi\})$ the set of formulas~$\theta$ in the
+\item as set~$\Hom_{\C_\TT}(\{\vec x\_ \varphi\}, \{\vec y\_ \psi\})$ of morphisms
+the set of formulas~$\theta$ in the
 context~$\vec x, \vec y$ which are~$\TT$-provably functional,
@@
 
-Writing~$\yon : \C_\TT \to \Set[\TT]$ for the Yoneda embedding, the \emph{generic
+Writing~$\yon : \C_\TT \to \Set[\TT]$ for the Yoneda embedding,\footnote{With
+this notational choice we are following Riehl and
+Verity~\cite{riehl-verity:elements}.}
+the \emph{generic
 model}~$U_\TT$ of~$\TT$ interprets a sort~$X$ of~$\TT$ as the sheaf~$\yon\{x\?X\_ \top\}$,
@@
 \end{thm}
+
+\begin{proof}See, for instance,~\cite[Theorem~2.1.8]{caramello:tst}
+or~\cite[discussion before Proposition~D3.1.12]{johnstone:elephant}.\end{proof}
 
@@
 propositions (again equipped with such a morphism); the object of geometric
-formulas (again so); the object of geometric sequents (again so). A internal
+formulas (again so); and the object of geometric sequents (again so). A internal
 geometric theory~$\TT$ over~$\Sigma$ is then given by a subobject of the object
@@
 
-\section{Proofs of the main theorems}
+\section{The first-order Nullstellensatz}
 \label{sect:main}
@@
 Given a geometric theory~$\TT$ with its generic model~$U_\TT$ in~$\Set[\TT]$,
-the main theorems reference a certain internal
+our main theorems will reference a certain internal
 theory~$\ul{\TT}/U_\TT$ internal to~$\Set[\TT]$. This theory is defined as
@@
 theory of homomorphisms between~$\TT$-models. It can also be obtained as the
-lax pullback~$(\Set[\TT] \Rightarrow_{\Set[\TT]} \Set[\TT])$.\end{rem}
+lax pullback~$(\Set[\TT] \Rightarrow_{\Set[\TT]} \Set[\TT])$.
+There are two canonical geometric morphisms from this topos to~$\Set[\TT]$, the
+morphism computing the domain and the morphism computing the codomain; the
+morphism obtained by externalizing the internal construction is the former.
+\end{rem}
 
@@
 \vdash_{[]} \alpha)$}, \]
-where the free variables~$\vec x$ occurring in~$\alpha$ are interpreted as in
-Proposition~\ref{prop:basic-truth}.
+where the free variables~$\vec x$ occurring in~$\alpha$ are first interpreted as in
+Proposition~\ref{prop:basic-truth} and then regarded as the induced constant
+symbols.
 \end{lemma}
@@
 Lemma~\ref{lemma:truth-to-provability} and using that~$\TT$ (and
-hence~$\ul{\TT}$) proves~$(\exists \vec y_j\_ \theta_j) \wedge (\exists \vec
-y_k\_ \theta_k) \seq{\vec x} \bigvee \{ \top \,|\, j = k \}$. (This kind of
+hence~$\ul{\TT}$) proves~$((\exists \vec y_j\_ \theta_j) \wedge (\exists \vec
+y_k\_ \theta_k) \seq{\vec x} \bigvee \{ \top \,|\, j = k \})$. (This kind of
 reasoning also appears, for instance, in~\cite[p.~20]{blass:seven-trees}.)
 \end{proof}
+% XXX check last paragraph; change proves* to proves.
 
@@
 sequents~$\sigma$ which are not of the form~$(\top \seq{[]} \varphi)$. Indeed,
-for such sequents, even~$\TT$ proves~$(\speak{$\sigma$ holds for~$U_\TT$}
+for those special sequents, even~$\TT$ proves~$(\speak{$\sigma$ holds for~$U_\TT$}
 \Rightarrow \speak{$\ul{\TT}/U_\TT$ proves~$(\top \seq{[]} \varphi)$})$, as can
@@
 sequents, Theorem~\ref{thm:nullstellensatz} expresses a nontrivial fact: that
-truth of a conditional, universally-quantified statement implies a single
-unconditional statement.
-
-\begin{rem}Theorem~\ref{thm:nullstellensatz} cannot be strengthened to
-arbitrary geometric sequents. For instance, in the case that~$\TT$ is the
+truth of an universally-quantified conditional statement implies a single
+unquantified unconditional statement.
+
+\begin{rem}The restriction in Theorem~\ref{thm:nullstellensatz} to
+geometric$^\star$ sequents cannot be lifted, that is the generalization
+of~Theorem~\ref{thm:nullstellensatz} to
+arbitrary internal geometric sequents is false. For instance, in the case that~$\TT$ is the
 theory of objects, the internal geometric sequent~$(\top \seq{x\?U_\TT}
@@
 of~$\Set[\TT]$, Theorem~\ref{thm:nullstellensatz-horn} can be strengthened to
-allow~$\bot$ as consequent, since in this case~$(\nabla\bot) \Rightarrow \bot$.\end{rem}
+allow~$\bot$ as consequent, since in this case we have~$(\nabla\psi)
+\Rightarrow \psi$ also for~$\psi \equiv \bot$.\end{rem}
 
@@
 where~$\Mod{\TT}_\mathrm{fp}$ is the full subcategory of the category
-of~$\TT$-models (in~$\Set$) on the finitely presentable objects. The
+of~$\TT$-models (in~$\Set$) on the finitely presentable objects~\cite[Theorem~2.1.21]{caramello:tst}. The
 (underlying object of the) generic model is the tautologous functor~$U_\TT : T
@@
 a~$\TT$-homomorphism~$U_\TT \to M$.
+
+% XXX: upgrade to cartesian case? maybe as a remark?
 
@@
 relation symbols~$S$ are interpreted by declaring~$([t_1],\ldots,[t_n]) \in
-\brak{S} \Leftrightarrow (\TT + R \vdash S(t_1,\ldots,t_n))$.
+\brak{S} \Leftrightarrow (\TT + R \,\vdash S(t_1,\ldots,t_n))$.
 
@@
 construction could be carried out if~$\TT$ was a general geometric theory, the
-resulting object would in general not be a model of~$\TT$.
+term algebra would in general not be a model of~$\TT$.
 \end{proof}
@@
 form are compact. Conversely, let a~$\TT$-model~$M$ be given. Then~$\TT$ is the
+% XXX cite Adamek/Rosicky
 filtered colimit of all models over~$M$ which are of the stated form. If~$M$ is
@@
 denotes the set of~$U_\TT$-algebra homomorphisms; $(U_\TT)^{\yon T}$ is the object of morphisms
-from~$\yon T$ to~$U_\TT$; and~$U_\TT|T$ is the functor~$U_\TT(T \times \cdot)$, that is~$S
-\mapsto T \otimes S$.
+from~$\yon T$ to~$U_\TT$; and~$U_\TT|T$ is the functor~$U_\TT(T \times \cdot)$, that is
+the functor~$S \mapsto T \otimes S$.
 
@@
 rules governing these constructors.
+% XXX cite some reference for these "usual rules", perhaps Milly?
 
@@
 connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$
-and by arbitrary set-indexed disjunctions~$\bigvee$. The existential
+and by arbitrary set-indexed disjunctions~$\bigvee$. Existential
 quantification can be over any sorts of extended geometric logic, including the derived sorts. An
@@
 statement~\eqref{eq:nullstellensatz} of Theorem~\ref{thm:nullstellensatz} in
+% XXX "the internal statement"? be more clear...
 extended geometric logic. Indeed, leaving the translation to formal logic to
@@
 \item The sequent~$\sigma$ holds for~$U_\TT$. \smallskip
-\item The sequent~$\sigma$ is provable$^\star$ modulo~$\TT/U_\TT$ in extended
+\item The sequent~$\sigma$ is provable$^\star$ modulo~$\ul{\TT}/U_\TT$ in extended
 geometric logic.
@@
 \begin{proof}The implication~$(2) \Rightarrow (1)$ is
-immediate since~$U_\TT$ is a model of~$\TT/U_\TT$. The converse direction is by
+immediate since~$U_\TT$ is a model of~$\ul{\TT}/U_\TT$. The converse direction is by
 Lemma~\ref{lemma:extended-conservative}, which holds internally in~$\Set[\TT]$
-as the proof we supplied is constructive, and the Nullstellensatz for ordinary
+as the proof we supplied is constructive, and by the Nullstellensatz for ordinary
 geometric logic of Theorem~\ref{thm:nullstellensatz}.
@@
 where~$X_1,\ldots,X_n$ is any list of sorts.
+% XXX think about variables...
 \end{enumerate}

### 2019-05-04 paper-qcoh.tex
typo

@@
 observed by Butz and Johnstone~\cite[Lemma~4.2]{butz-johnstone:first-order}. At
-that point, a characterisation of the first-order formulas in the general case,
+that point, a characterization of the first-order formulas in the general case,
 of the form as in Theorem~\ref{thm:characterization0}, was still missing.
@@
 context of geometric theories and their classifying Grothendieck toposes, but
-there should be analogues for the Nullstellensatz in other contexts where it
+there should be analogues of the Nullstellensatz in other contexts where it
 makes sense to speak of ``classifying gadgets'', and these analogues should
@@
 
-To verify that~(3) implies~(4), we can mimic the usual proof of this fact in
+To verify that~(3) implies~(4), we can mimic the usual proof of that for
+coherent sequents, being derivable in geometric logic implies being derivable
+in coherent logic, in
 the internal language of~$\E$: There is a variant of the syntactic

### 2019-05-05 paper-qcoh.tex
Refactored to remove the notion of provability* (and to properly deal with free variables)

@@
 rings and let~$\sigma$ be the sequent~$(f(x) = 0 \seq{x} g(x) = 0)$ for some
-polynomials~$f$ and~$g$ with coefficients in~$U_\TT$. To say that~$\sigma$ holds for~$U_\TT$ amounts to
+polynomials~$f$ and~$g$. To say that~$\sigma$ holds for~$U_\TT$ amounts to
 saying that any zero~$x \? U_\TT$ of~$f$ is also a zero of~$g$, and to say
@@
 \begin{thm}\label{thm:characterization0}
-Let~$\TT$ be a geometric theory. Let~$\alpha$ be a first-order
+Let~$\TT$ be a geometric theory. Let~$\varphi$ be a first-order
 formula over the signature of~$\TT$. Then the following statements are
@@
 \begin{enumerate}
-\item The formula~$\alpha$ holds for~$U_\TT$. \smallskip
-\item The formula~$\alpha$ is provable in first-order intuitionistic logic
+\item The formula~$\varphi$ holds for~$U_\TT$. \smallskip
+\item The formula~$\varphi$ is provable in first-order intuitionistic logic
 modulo the axioms of~$\TT$ and the additional axiom~\eqref{eq:nullstellensatz0}.
@@
 
-It is not entirely obvious how to correctly phrase
-axiom~\eqref{eq:nullstellensatz0} in first-order logic, an issue
-which we will discuss when we restate Theorem~\ref{thm:characterization0} in
-the body of this note (Theorem~\ref{thm:characterization}).
+It is not entirely obvious how to correctly formalize
+axiom~\eqref{eq:nullstellensatz0} in unadorned (XXX) first-order logic, and in
+fact, we do not believe that an entirely faithful formalization is possible.
+We will resolve this issue in the body of this note, where we will restate
+Theorem~\ref{thm:characterization0} in a more precise form (Theorem~\ref{thm:characterization}).
 
@@
 
+\begin{rem}\label{rem:unique-descriptions}
+Sections of the sheaf~$\yon\{ y \? Y\_ \top \}$ over a stage~$A = \{ \vec x\_
+\varphi \} \in C_\TT$ are in one-to-one correspondence with ``unique
+descriptions'' over~$A$, that is formulas~$\theta$ in the context~$\vec x, y$
+such that~$\TT$ proves~$(\varphi \seq{\vec x} \exists!y\?Y\_ \theta)$, up to
+provable equivalence.
+\end{rem}
+
 \begin{thm}The generic model is universal in the sense that for any
@@
 in~$\varphi$ are interpreted as their generic values over~$\{\vec
-x\_\alpha\}$, that is the projection maps~$\{ \vec x\_ \alpha \} \to \{
-x_i\?X_i\_ \top \})$. \smallskip
+x\_\alpha\}$, that is as the projection maps~$\{ \vec x\_ \alpha \} \to \{
+x_i\?X_i\_ \top \}$. \smallskip
 \item $\TT$ proves~$(\alpha \seq{\vec x} \varphi)$.
@@
 geometric theory~$\TT$ over~$\Sigma$ is then given by a subobject of the object
-of geometric sequents, interpreted as the object of axioms of~$\TT$. Given such
-an internal theory~$\TT$, we can build the object of proof trees of~$\TT$.
-
-From the internal point of view of~$\E$, these objects can be obtained by
-simply carrying out the usual constructions of the set of contexts, the set of
-terms and so on. Having the object of proof trees at hand, we can define
-the notion of provability:
-
-\begin{defn}\label{defn:provability}
-Let~$\TT$ be a geometric theory internal to a Grothendieck topos~$\E$.
-Internally to~$\E$, an element~$\sigma$ of the object of geometric sequents is
-\emph{provable modulo~$\TT$} if and only if there is an element of the object
-of proof trees of~$\TT$ which has~$\sigma$ as its conclusion.\end{defn}
+of geometric sequents, interpreted as the object of axioms of~$\TT$.
+Given such an internal theory~$\TT$, we can build the subobject of the provable
+sequents.
+
+For the most part, these objects can be obtained by simply carrying out the
+familiar constructions of the set of contexts, of the set of terms and so on in
+the internal language of~$\E$. Some care is required in constructing the
+object of geometric formulas: Inductively closing the object of basic
+propositions under the connectives of geometric logic will fail because of size
+issues -- even if we carry this out in~$\Set$, we will end up with a proper
+class. Instead the object of geometric formulas should be constructed so as to
+only contain geometric formulas in canonical form~(``$\bigvee \exists \cdots
+\exists\_ \varphi_1 \wedge \cdots \wedge \varphi_n$'' for atomic
+propositions~$\varphi_i$). There is no real loss in this restriction since in
+foundations where it makes sense to say this, any geometric formula is provably
+equivalent to a geometric formula in canonical form.
+
+% XXX similarily
+Similarly, we cannot construct the subobject of provable sequents by first
+constructing an object of raw (arbitrarily-branching) trees and then cutting
+down to an object of correctly formed proof trees. Instead, the subobject of
+provable sequents can be obtained by intersecting, internally to~$\E$, all
+subsets of the set of sequents which are closed under the rules of geometric
+logic.\footnote{Equivalently one can appeal to the Knaster--Tarski fixed point
+theorem, which is constructively valid in the form we require
+here~\cite{bauer-lumsdaine:bourbaki-witt}, to construct the least fixed point
+of the operator which, given a set~$M$ of geometric sequents, computes the set of
+geometric sequents which can be derived from those in~$M$ in at most one step.
+Details on how to carry out such kinds of inductive constructions can be found
+in~\cite{blass:induction} and more specifically in
+\cite[Chapter~III]{henry:phd}.}
+
+None of these complications arise when setting up the theory of internal
+coherent theories.
 
@@
 internal to the topos~$\Set$. A geometric sequent is provable in the ordinary
-sense if and only if, from the internal point of view of~$\Set$, it is provable
-in the sense of Definition~\ref{defn:provability}.\end{ex}
+sense if and only if, from the internal point of view of~$\Set$, it is
+contained in the object of provable sequents.\end{ex}
 
@@
 signature~$\ul{\Sigma}$ in~$\E$. The object of sorts of~$\ul{\Sigma}$ is the pullback of
-the set of sorts of~$\Sigma$, and so on.\end{ex}
+the set of sorts of~$\Sigma$, the object of function symbols of~$\ul{\Sigma}$
+is the pullback of the set of function symbols of~$\Sigma$, and so on.\end{ex}
 
 Disjunctions appearing in internal geometric formulas may be indexed by
-arbitrary objects of the topos, just like disjunctions appearing in ordinary
+arbitrary objects of the topos (that is, disjunctions may be taken over
+arbitrary subobjects of disjuncts), just like disjunctions appearing in ordinary
 external geometric formulas over an ordinary signature may be indexed by
@@
 
-Correspondingly, there is a subobject of the object of proof trees, the object
-of~\emph{proof$^\star$ trees} where any occurring geometric formula is a
-geometric$^\star$ formula. The shape of such a proof$^\star$ tree is locally
-given by the shape of an ordinary external proof tree.
-
-\begin{defn}\label{defn:provability*}
-Let~$\TT$ be a geometric theory internal to a Grothendieck topos~$\E$.
-Internally to~$\E$, an element~$\sigma$ of the object of geometric$^\star$
-sequents is \emph{provable$^\star$ modulo~$\TT$} if and only if there is an
-element of the object of proof$^\star$ trees of~$\TT$ which has~$\sigma$ as its
-conclusion.\end{defn}
-
-The following lemma shows that for coherent theories, there is no difference
-between provability and provability$^*$.
-
-\begin{lemma}\label{lemma:provability-in-coherent-case}
-Let~$\E$ be a Grothendieck topos. Let~$\TT$ be a coherent theory
-internal to~$\E$. Let~$\sigma$ be a coherent sequent over the signature
-of~$\TT$. Then the following statements are equivalent.
-\begin{enumerate}
-\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of externally
-finite shape}.$ \smallskip
-\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of arbitrary
-external shape}$, that is~$\E \models \speak{$\sigma$ is provable$^\star$
-modulo~$\TT$}$. \smallskip
-\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of arbitrary
-internal shape}$, that is~$\E \models \speak{$\sigma$ is provable
-modulo~$\TT$}$. \smallskip
-\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of internally
-finite shape}.$
-\end{enumerate}
-\end{lemma}
-
-\begin{proof}It is trivial that~(1) implies~(2) implies~(3).
-
-To verify that~(3) implies~(4), we can mimic the usual proof of that for
-coherent sequents, being derivable in geometric logic implies being derivable
-in coherent logic, in
-the internal language of~$\E$: There is a variant of the syntactic
-site of~$\TT$ which is built using only coherent sequents and finitary
-derivability~\cite[Section~1.4]{caramello:tst}. The sheaf topos over this site
-is another model for the classifying topos of~$\TT$, and still validates, like
-any Grothendieck topos, full infinitary logic. Hence, if~$\sigma$
-is~$\TT$-derivable by a proof tree of arbitrary shape, then~$\sigma$ holds in
-this model of the classifying topos. By the analogue of
-Proposition~\ref{prop:basic-truth} for this model, we obtain
-that~$\sigma$ is~$\TT$-derivable by a proof tree of finite shape.
-
-That~(4) implies~(1) is a routine exercise exploiting that
-\[ \E \models \forall X\_ \speak{$X$ is Kuratowski-finite} \Leftrightarrow
-  \bigvee_{n \geq 0} \exists x_1,\ldots,x_n \? X\_
-  \forall x \? X\_ \bigvee_{i=1}^n x = x_i. \qedhere \]
-\end{proof}
+\begin{rem}A priori, there are two different notions of what it could mean that
+a geometric$^\star$ formula is provable: It could be provable when regarded as
+a geometric formula, or we could allow only geometric$^\star$ formulas as
+intermediate formulas in proofs. One can show that these two notions coincide,
+similarly to how a coherent sequent is provable in geometric logic if and only
+if it is provable in coherent logic. Since we do not require this fact in the
+course of this note, we omit a detailed verification.
+\end{rem}
 
@@
 Proposition~\ref{prop:basic-truth} and then regarded as the induced constant
-symbols.
+symbols provided by the enlarged signature of~$\ul{\TT}/U_\TT$.
 \end{lemma}
@@
 of~$A$ such that for each index~$i$, there is a formula~$\varphi_i$ over the signature
-of~$\TT/U_\TT(A_i)$ such that~$A_i \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$
+of~$\TT/U_\TT(A_i)$ such that~$A_i \models \speak{$\ul{\TT}/U_\TT$ proves
 $(\varphi \dashv\vdash \varphi_i)$}$.
@@
 \[ \bigvee_j \bigl((\exists \vec y_j\_ \theta_j) \wedge R_j(\ldots)\bigr). \]
+(The formulas~$\theta_j$ will typically contain instances of the
+variables~$\vec x$. When writing down this replacement, we treat these as in
+Lemma~\ref{lemma:truth-to-provability}. Hence this replacement is set in the
+same context as~$\varphi$, only that new constant symbols might occur.)
 In a similar vein we treat any occurrence of function symbols.
@@
 of~$\TT/U_\TT(A)$.
-The verification of~$A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$
+The verification of~$A \models \speak{$\ul{\TT}/U_\TT$ proves
 $(\varphi \dashv\vdash \varphi')$}$ rests on the observation
-\[ A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$ $\bigl((\exists \vec y_k\_
-\theta_k) \seq{[]} \bigvee_\ell \{ \top \,|\, \text{$(\exists \vec y_\ell\_ \theta_\ell)$
+\[ A \models \speak{$\ul{\TT}/U_\TT$ proves $\bigl((\exists \vec y_k\_
+\theta_k) \seq{[]} \bigvee \{ \top \,|\, \text{$(\exists \vec y_k\_ \theta_k)$
 holds for~$U_\TT$} \}\bigr)$} \]
@@
 \end{proof}
-% XXX check last paragraph; change proves* to proves.
 
@@
 \item The sequent~$\sigma$ holds for~$U_\TT$. \smallskip
-\item The sequent~$\sigma$ is provable$^\star$ modulo~$\ul{\TT}/U_\TT$.
+\item The sequent~$\sigma$ is provable modulo~$\ul{\TT}/U_\TT$.
 \end{enumerate}
@@
 \begin{proof}The direction~$(2) \Rightarrow (1)$ is immediate because~$U_\TT$ is, from the
-internal point of view of~$\Set[\TT]$, a~$\ul{\TT}/U_\TT$-model. Hence even the
-following stronger statement holds internally: \emph{For any geometric
-sequent~$\sigma$, if~$\ul{\TT}/U_\TT$ proves~$\sigma$, then~$\sigma$ holds
-for~$U_\TT$.}
+internal point of view of~$\Set[\TT]$, a~$\ul{\TT}/U_\TT$-model.
 
@@
 \speak{$\sigma$ holds for~$U_\TT$}$ then~$A \models
-\speak{$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$}$. By
+\speak{$\ul{\TT}/U_\TT$ proves~$\sigma$}$. By
 Lemma~\ref{lemma:locally-constant} we may suppose that~$\sigma$ is an
@@
 Writing~$A = \{ \vec x\_ \alpha \}$ and~$\sigma = (\varphi \seq{\vec y} \psi)$,
-we have~$\{ \vec x\_ \alpha \} \models \forall \vec y\_ (\varphi \Rightarrow \psi)$,
-hence~$\{ \vec x, \vec y\_ \alpha \wedge \varphi \} \models \psi$. Thus~$\TT$
-proves~$(\alpha \wedge \varphi \seq{\vec x, \vec y} \psi)$. This proof can be
-pulled back from~$\Set$ to~$\Set[\TT]/\yon A$ to show~$A \models
-\speak{$\ul{\TT}/U_\TT$ proves$^\star$~$(\alpha
+we have~$\{ \vec x\_ \alpha \} \models \forall \vec y\_ (\varphi \Rightarrow \psi)$ by assumption,
+hence~$\{ \vec x, \vec y\_ \alpha \wedge \varphi \} \models \psi$ (where we
+inline any occurence of an element of~$U_\TT(A)$ as a constant symbol
+in~$\varphi$, recalling Remark~\ref{rem:unique-descriptions}).
+Thus~$\TT$
+proves~$(\alpha \wedge \varphi \seq{\vec x, \vec y} \psi)$ (where we now have
+to do the same inlining for~$\psi$ as well). This proof can be
+pulled back from~$\Set$ to~$\Set[\TT]/\yon A$ to obtain~$A \models
+\speak{$\ul{\TT}/U_\TT$ proves~$(\alpha
 \wedge \varphi \seq{\vec x, \vec y} \psi)$}$. By
 Lemma~\ref{lemma:truth-to-provability}, we also have~$A \models
-\speak{$\ul{\TT}/U_\TT$ proves$^\star$~$(\top \seq{[]} \alpha)$}$ (where the free
+\speak{$\ul{\TT}/U_\TT$ proves~$(\top \seq{[]} \alpha)$}$ (where the free
 variables occurring in~$\alpha$ are interpreted as the generic values available
-over~$A$), hence~$A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$~$(\varphi
+over~$A$), hence~$A \models \speak{$\ul{\TT}/U_\TT$ proves~$(\varphi
 \seq{\vec y} \psi)$}$.
@@
 sequents~$\sigma$ which are not of the form~$(\top \seq{[]} \varphi)$. Indeed,
-for those special sequents, even~$\TT$ proves~$(\speak{$\sigma$ holds for~$U_\TT$}
-\Rightarrow \speak{$\ul{\TT}/U_\TT$ proves~$(\top \seq{[]} \varphi)$})$, as can
-be seen by a simple induction on the structure of~$\varphi$. For more general
-sequents, Theorem~\ref{thm:nullstellensatz} expresses a nontrivial fact: that
+for those special sequents, the implication~$(\text{$\sigma$ holds for~$U_\TT$}
+\Rightarrow \text{$\ul{\TT}/U_\TT$ proves~$(\top \seq{[]} \varphi)$})$
+can be proven by a simple induction on the structure of~$\varphi$. Only for more general
+sequents does Theorem~\ref{thm:nullstellensatz} express a nontrivial fact: that
 truth of an universally-quantified conditional statement implies a single
@@
 \[ \text{A geometric$^\star$ sequent~$\sigma$ with Horn consequent holds for~$U_{\TT'}$
-  iff~$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$.} \]
+  iff~$\ul{\TT}/U_\TT$ proves~$\sigma$.} \]
 \end{thm}
@@
     \text{\emph{if~$\forall x_1,\ldots,x_n\?U_\TT\_ (\varphi \Rightarrow \nabla
-    \psi)$, then~$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$.}}
+    \psi)$, then~$\ul{\TT}/U_\TT$ proves~$\sigma$.}}
 \end{multline*}
@@
 does not prove this fact, as it is for instance not validated by the polynomial
-algebra~$U_\TT[X]$. However, Theorem~\ref{thm:nullstellensatz} still plays an
-important role in understanding first-order formulas:
+algebra~$U_\TT[X]$.
+
+However, Theorem~\ref{thm:nullstellensatz} still plays an
+important role in understanding first-order formulas, as it is at the core of
+our proposed characterization of the first-order formulas validated by the
+generic model.
+
+\begin{scholium}\label{scholium:nullstellensatz-more-specific}
+Let~$\TT$ be a geometric theory. Let~$\vec x$ be a context over the signature
+of~$\TT$. Let~$(\sigma_i)_i$ be a set of geometric sequents, where each context
+is of the form~$\vec x, \vec y_i$ for some additional list~$\vec y_i$.
+Write~$\sigma_i = (\varphi_i \seq{\vec x, \vec y_i} \psi_i)$. Then, internally
+to~$\Set[\TT]$, it holds that
+\[ \forall \vec x\_
+  \bigl(\bigwedge_i (\forall \vec y_i\_ \varphi_i \Rightarrow \psi_i)\bigr) \Longrightarrow
+  \bigvee_\alpha \alpha,
+\]
+where the disjunction ranges over all those geometric formulas~$\alpha$ in the
+context~$\vec x$ such that for each index~$i$, the theory~$\TT$ proves~$(\alpha
+\wedge \varphi_i \seq{\vec x, \vec y_i} \psi_i)$.
+\end{scholium}
+
+\begin{proof}Immediate from the proof of Theorem~\ref{thm:nullstellensatz}.
+\end{proof}
 
 \begin{thm}\label{thm:characterization}
-Let~$\TT$ be a geometric theory. Let~$\alpha$ be a first-order
+Let~$\TT$ be a geometric theory. Let~$\chi$ be a (finitary or infinitary) first-order
 formula over the signature of~$\TT$. Then the following statements are
@@
 \begin{enumerate}
-\item The formula~$\alpha$ holds for~$U_\TT$. \smallskip
-\item The formula~$\alpha$ is provable in first-order intuitionistic logic
-modulo the axioms of~$\TT$ and the additional axioms
+\item The formula~$\chi$ holds for~$U_\TT$. \smallskip
+\item The formula~$\chi$ is provable in infinitary first-order intuitionistic
+logic modulo the axioms of~$\TT$ adjoined by, for each geometric
+sequent~$\sigma = (\varphi \seq{\vec x, \vec y} \psi)$ and for each splitting
+of its context into two parts~$\vec x, \vec y$, the additional axiom
 \begin{equation}\label{eq:nullstellensatz}\tag{$\ddagger$}
-  \speak{$\sigma$ holds for~$U_\TT$} \Longrightarrow \speak{$\ul{\TT}/U_\TT$
-  proves$^\star$~$\sigma$}
+  \forall \vec x\_
+  \bigl((\forall \vec y\_ \varphi \Rightarrow \psi) \Longrightarrow
+  \bigvee_\alpha \alpha\bigr),
 \end{equation}
-where~$\sigma$ ranges over the geometric formulas over the signature of~$\TT$.
-% XXX include free variables
+where the disjunction ranges over all those geometric formulas~$\alpha$ in the
+context~$\vec x$ such that~$\TT$ proves~$(\alpha \wedge \varphi \seq{\vec x,\vec
+y} \psi)$.
 \end{enumerate}
@@
 
-In stating the axiom scheme~\eqref{eq:nullstellensatz}, we are slightly abusing
-notation. At face value, this axiom scheme would only make sense in the internal
-language of~$\Set[\TT]$, where the generic model~$U_\TT$ actually lives.
-However, here we are interpreting it in a purely syntactic context; when
-writing~``$U_\TT$'', we actually mean the tautologous ``model'' in which any
-sort of~$\TT$ is interpreted by the sort itself. For instance, if~$\sigma =
-(\varphi \seq{x_1\!\?\!X_1,\ldots,x_n\!\?\!X_n} \psi)$ is a sequent,
-then~$\speak{$\sigma$ holds for~$U_\TT$}$ is to be interpreted as the
-formula~``$\forall x_1\?X_1\_ \ldots \forall x_n\?X_n\_ (\varphi \Rightarrow
-\psi)$''.
-
+To a first approximation, axiom scheme~\eqref{eq:nullstellensatz} is just a
+rendition of the (nontrivial direction of the) statement of the Nullstellensatz
+in infinitary first-order logic. However, the conclusion ``$\sigma$ is provable
+modulo~$\ul{\TT}/U_\TT$'' does not seem to be expressible in this logic.
+The conclusion~``$\bigvee_\alpha \alpha$'' of axiom
+scheme~\eqref{eq:nullstellensatz} is a strengthening of ``$\sigma$ is provable
+modulo~$\ul{\TT}/U_\TT$'' which can. We recall that in infinitary first-order
+formulas, we allow set-indexed disjunctions (but not set-indexed conjunctions).
+
+% XXX hyphenation scholium
 \begin{proof}[Proof of Theorem~\ref{thm:characterization}]
-The direction~$(2) \Rightarrow (1)$ is immediate, as the internal
-language of~$\Set[\TT]$ validates first-order intuitionistic logic and as it
-validates the
-axiom scheme~\eqref{eq:nullstellensatz} by Theorem~\ref{thm:nullstellensatz}.
-
-For the converse direction, we note that, given a geometric sequent~$\sigma$
-over the signature of~$\TT$, the statement~$\speak{$\ul{\TT}/U_\TT$
-proves$^\star$~$\sigma$}$ can be expressed as a geometric formula over the
-signature of~$\TT$. Applying this observation successively to subformulas
-of the given formula~$\alpha$, Theorem~\ref{thm:nullstellensatz} on the
-semantic side and the axiom scheme~\eqref{eq:nullstellensatz} on the syntactic side imply that we may
-assume that~$\alpha$ is in fact a geometric formula. Hence we are reduced to
-the basic fact that, for geometric formulas~$\varphi$, $\Set[\TT] \models
-\varphi$ implies that~$\TT$ proves~$\varphi$.
+Any infinitary first-order formula can be simplified to an equivalent geometric
+formula -- on the semantic side by repeatedly applying
+Scho\-lium~\ref{scholium:nullstellensatz-more-specific}, on the semantic side by
+repeatedly applying axiom scheme~\eqref{eq:nullstellensatz}. Hence we are
+reduced to the basic fact (Proposition~\ref{prop:basic-truth}) that, for
+geometric formulas~$\varphi$, $\Set[\TT] \models
+\varphi$ if and only if~$\TT$ proves~$\varphi$.
 \end{proof}
@@
 Theorem~\ref{thm:nullstellensatz} requires external ingredients in referring to
-geometric$^\star$ sequents and provability$^\star$. A natural question is
+geometric$^\star$ sequents. A natural question is
 therefore whether, in the case that~$\TT$ is a coherent theory, the statements
@@
 
-As stated in Lemma~\ref{lemma:provability-in-coherent-case}, for coherent
-sequents modulo a coherent theory there is no difference between provability in
-geometric logic, provability$^\star$ in geometric logic and provability in
-coherent logic. However, while geometric logic is powerful enough to express
-provability$^\star$ in geometric logic, coherent logic is not powerful enough
+However, while geometric logic is powerful enough to express
+provability in geometric logic (or rather the strenghtening we employed in
+axiom scheme~\eqref{eq:nullstellensatz}),
+coherent logic is not powerful enough
 to express provability in coherent logic. ``Coherent logic cannot eat itself.''
@@
 existential fixed-point logic~\cite{blass:topoi-and-computation,blass:existential-fixed-point-logic}.
-The list sorts are necessary to express raw terms, raw formulas and raw
-sequents; the fixed-point operator is then used to express well-formedness of
+The list sorts can be used to express raw terms, raw formulas and raw
+sequents; the fixed-point operator can then be used to express well-formedness of
 raw terms, raw formulas and raw sequents; and to express provability. For this
@@
 
-\begin{scholium}Let~$\TT$ be a coherent theory (or more generally theory in coherent
+\begin{scholium}Let~$\TT$ be a coherent theory (or more generally a theory in coherent
 existential fixed-point logic). Let~$\alpha$ be a finitary first-order
@@
 where~$\sigma$ ranges over the formulas of coherent existential fixed-point
-logic over the signature of~$\TT$.
+logic over the signature of~$\TT$. (When referring to~``$U_\TT$'', we here mean
+the tautologous ``model'' in which any
+sort of~$\TT$ is interpreted by the sort itself. For instance, if~$\sigma =
+(\varphi \seq{x_1\!\?\!X_1,\ldots,x_n\!\?\!X_n} \psi)$ is a sequent,
+then~$\speak{$\sigma$ holds for~$U_\TT$}$ is to be interpreted as the
+formula~``$\forall x_1\?X_1\_ \ldots \forall x_n\?X_n\_ (\varphi \Rightarrow
+\psi)$''.)
 \end{enumerate}
@@
 existential fixed-point logic.
+\end{proof}
+
+On the other end, we can ask for a generalization of
+Theorem~\ref{thm:characterization} to the extension of infinitary first-order
+logic where we additionally allow set-indexed conjunctions. Such a
+generalization is given by the following scholium.
+
+\begin{scholium}
+Let~$\TT$ be a geometric theory. Let~$\chi$ be a formula in infinitary
+first-order formula enriched by set-indexed conjunctions over the signature
+of~$\TT$. Then the following statements are equivalent.
+\begin{enumerate}
+\item The formula~$\chi$ holds for~$U_\TT$. \smallskip
+\item The formula~$\chi$ is provable in infinitary first-order intuitionistic
+logic enriched by set-indexed conjunctions modulo the axioms of~$\TT$ adjoined
+by the displayed formulas of
+Scholium~\ref{scholium:nullstellensatz-more-specific}.
+\end{enumerate}
+\end{scholium}
+
+\begin{proof}The proof of Theorem~\ref{thm:characterization} carries over word
+for word.
 \end{proof}
@@
 and by arbitrary set-indexed disjunctions~$\bigvee$. Existential
-quantification can be over any sorts of extended geometric logic, including the derived sorts. An
+quantification can be over any of the sorts of extended geometric logic, including the derived sorts. An
 \emph{extended geometric sequent} is a sequent of the form~$(\varphi
@@
 of a geometric theory~$\TT$. Then there is a set-indexed family~$(\sigma_i)_i$ of ordinary
-geometric sequents over the same signature, so that~$\sigma$ is provable in
+geometric sequents over the same signature such that~$\sigma$ is provable in
 extended geometric logic if and only if all the sequents~$\sigma_i$ are
@@
 Then the canonical morphism
-\[ \mathrm{Form}^\star_{\vec x}(\ul{\TT}/U_\TT)/(\dashv\vdash_{\vec x}) \longrightarrow P(X_1 \times
-\cdots \times X_n) \]
+\[ \mathrm{Form}^\star_{\vec x}(\ul{\TT}/U_\TT)/(\dashv\vdash_{\vec x})
+  \longrightarrow P(\yon X_1 \times \cdots \times \yon X_n) \]
 sending, internally speaking, the equivalence class of a geometric$^\star$
-formula~$\varphi$ to the subset~$\{ (x_1,\ldots,x_n) \,|\, \varphi \}$ is an isomorphism.
+formula~$\varphi$ to the subset~$\{ (x_1,\ldots,x_n) \,|\, \text{$\varphi$
+holds for~$(x_1,\ldots,x_n)$} \}$ is an isomorphism.
 \end{thm}
@@
 The proof of Theorem~\ref{thm:definability} used the Nullstellensatz
-of Theorem~\ref{thm:nullstellensatz}. Conversely, the internal statement of
-Theorem~\ref{thm:definability} entails the internal
-statement~\eqref{eq:nullstellensatz} of Theorem~\ref{thm:nullstellensatz} in
-% XXX "the internal statement"? be more clear...
-extended geometric logic. Indeed, leaving the translation to formal logic to
-the reader, let~$\sigma = (\varphi \seq{\vec x} \psi)$ be a
+of Theorem~\ref{thm:nullstellensatz}. Conversely,
+Theorem~\ref{thm:definability} implies Theorem~\ref{thm:nullstellensatz}.
+Indeed, arguing internally, let~$\sigma = (\varphi \seq{\vec x} \psi)$ be a
 geometric$^\star$ sequent over the signature of~$\ul{\TT}/U_\TT$ such
@@
 assumption implies that the formulas~$\varphi$
-and~$\varphi \wedge \psi$ are provably$^\star$ equivalent.
-Thus~$\ul{\TT}/U_\TT$ proves$^\star$~$(\varphi \seq{\vec x} \psi)$.
+and~$\varphi \wedge \psi$ are provably equivalent.
+Thus~$\ul{\TT}/U_\TT$ proves~$(\varphi \seq{\vec x} \psi)$.
 \end{rem}
@@
 internally to~$\Set[\TT]$, the canonical map from the set of equivalence classes
-of~$\ul{\TT}/U$-provably$^\star$ functional geometric$^\star$ formulas
+of~$\ul{\TT}/U$-provably functional geometric$^\star$ formulas
 from~$\{\vec x\_ \varphi\}$ to~$\{\vec y\_ \psi\}$ to the set of maps~$\{(\vec y)
@@
 geometric$^\star$ formula~$\theta$. Because~$f$ is a map, this formula is
-functional; and by the Nullstellensatz, it is~$\ul{\TT}/U_\TT$-provably$^\star$ so.
+functional; and by the Nullstellensatz, it is~$\ul{\TT}/U_\TT$-provably so.
 
 For verifying injectivity, let~$\theta$ and~$\theta'$
-be~$\ul{\TT}/U_\TT$-provably$^\star$ functional formulas which give rise to
+be~$\ul{\TT}/U_\TT$-provably functional formulas which give rise to
 identical maps. Then they also give rise to identical graphs, hence
-are~$\ul{\TT}/U_\TT$-provably$^\star$ equivalent by
+are~$\ul{\TT}/U_\TT$-provably equivalent by
 Theorem~\ref{thm:definability}.
@@
 \item The sequent~$\sigma$ holds for~$U_\TT$. \smallskip
-\item The sequent~$\sigma$ is provable$^\star$ modulo~$\ul{\TT}/U_\TT$ in extended
+\item The sequent~$\sigma$ is provable modulo~$\ul{\TT}/U_\TT$ in extended
 geometric logic.
@@
 \begin{thm}\label{thm:higher-order-characterization}
-Let~$\TT$ be a geometric theory. Let~$\alpha$ be a higher-order formula over
+Let~$\TT$ be a geometric theory. Let~$\chi$ be a higher-order formula over
 the signature of~$\TT$. Then the following statements are equivalent:
 \begin{enumerate}
-\item The formula~$\alpha$ holds for~$U_\TT$. \smallskip
-\item The formula~$\alpha$ is provable in higher-order intuitionistic logic
+\item The formula~$\chi$ holds for~$U_\TT$. \smallskip
+\item The formula~$\chi$ is provable in higher-order intuitionistic logic
 modulo the axioms of~$\TT$ and the additional axiom scheme
@@
 scheme~\eqref{eq:higher-order-nullstellensatz} on the syntactic side allow us to replace any mention of a
-powersort~$P(X)$ in~$\alpha$
+powersort~$P(X)$ in~$\chi$
 by~$\mathrm{Form}^\star_{x\!\?\!X}(\ul{\TT}/U_\TT)/({\dashv\vdash_x})$,
@@
 * XXX Upgrade Horn section to cartesian theories?
+
+* XXX fix "first-order" usage everywhere

### 2019-05-05 paper-qcoh.tex
A couple more references

@@
 It is not entirely obvious how to correctly formalize
-axiom~\eqref{eq:nullstellensatz0} in unadorned (XXX) first-order logic, and in
+axiom~\eqref{eq:nullstellensatz0} in unadorned first-order logic, and in
 fact, we do not believe that an entirely faithful formalization is possible.
@@
 version of the Nullstellensatz for arithmetic universes, the predicative cousin
-of toposes introduced by Joyal~\cite{XXX} and recently an important object of
-consideration by Maietti and Vickers~\cite{XXX}.
+of toposes introduced by Joyal and recently an important object of
+consideration by Maietti and Vickers~\cite{maietti:au,maietti-vickers:induction,vickers:sketches}.
 \medskip
@@
 
-% XXX similarily
 Similarly, we cannot construct the subobject of provable sequents by first
@@
 
-% XXX hyphenation scholium
 \begin{proof}[Proof of Theorem~\ref{thm:characterization}]
@@
 formula -- on the semantic side by repeatedly applying
-Scho\-lium~\ref{scholium:nullstellensatz-more-specific}, on the semantic side by
+Scho\-li\-um~\ref{scholium:nullstellensatz-more-specific}, on the semantic side by
 repeatedly applying axiom scheme~\eqref{eq:nullstellensatz}. Hence we are
@@
 \begin{proof}It is an instructive exercise to verify that models of the stated
-form are compact. Conversely, let a~$\TT$-model~$M$ be given. Then~$\TT$ is the
-% XXX cite Adamek/Rosicky
+form are compact (in a slightly different context, this is done
+in~\cite[Theorem~3.12]{adamek-rosicky:presentable}). Conversely, let
+a~$\TT$-model~$M$ be given. Then~$\TT$ is the
 filtered colimit of all models over~$M$ which are of the stated form. If~$M$ is
@@
 
-* XXX restrict to finitary first-order formulas in the scholium
-
 * XXX Upgrade Horn section to cartesian theories?

### 2019-05-05 paper-qcoh.tex
Explanation of term "quasicoherence"

@@
 of~$U_\TT$-algebras), the canonical~$U_\TT$-algebra homomorphism
-\[ A \longrightarrow (U_\TT)^{\Spec(A)} \]
+\[ A \longrightarrow (U_\TT)^{\Spec(A)},\ x \longmapsto -(x) \]
 is an isomorphism.
@@
 
-% XXX justification name quasicoherent
-
-\begin{proof}The proof of Lemma~\ref{lemma:char-fp-models} is constructive and
+We use the term \emph{quasicoherent} in reference to algebraic geometry: For
+an~$\affl$-module~$M$ in the big Zariski topos of a scheme~$S$, where~$\affl$
+is the functor of points of the affine line over~$S$, there is a
+well-established notion of what it means that~$M$ is quasicoherent. This
+property can be characterized in the internal language: Such a module~$M$ is
+quasicoherent if and only if, from the internal point of view of the big
+Zariski topos, the canonical map~$A \otimes_\affl M \to M^{\Spec(A)}$ is an
+isomorphism for any finitely
+presented~$\affl$-algebra~$A$~\cite[Theorem~18.19]{blechschmidt:phd}.
+Specializing to the case~$M = \affl$, we obtain the quasicoherence condition of
+Theorem~\ref{thm:quasicoherence}.
+
+\begin{proof}[Proof of Theorem~\ref{thm:quasicoherence}]The proof of
+Lemma~\ref{lemma:char-fp-models} is constructive and
 thus valid in the internal language of~$\Set[\TT]$. Hence we can apply it,

### 2019-05-05 paper-qcoh.tex
Beginning of the Horn subtopos case

@@
 indeed in general it does not. However, there is still a useful substitute,
-which we formulate as Theorem~\ref{thm:nullstellensatz-horn}. This substitute
+which we formulate as Corollary~\ref{corollary:nullstellensatz-horn}. This substitute
 broadens the scope of the Nullstellensatz and is, whenever applicable, useful
@@
 
-\begin{thm}\label{thm:nullstellensatz-horn}
+\begin{cor}\label{corollary:nullstellensatz-horn}
 Let~$\TT$ be a geometric theory. Let~$\TT'$ be a quotient theory
@@
   iff~$\ul{\TT}/U_\TT$ proves~$\sigma$.} \]
-\end{thm}
+\end{cor}
 
@@
 
-A situation in which the sheaf assumption of Theorem~\ref{thm:nullstellensatz-horn}
+A situation in which the sheaf assumption of Corollary~\ref{corollary:nullstellensatz-horn}
 is satisfied is when~$\TT$ is a Horn theory and the topology cutting
@@
 \begin{rem}In the case that the~$\Set[\TT']$ is a dense subtopos
-of~$\Set[\TT]$, Theorem~\ref{thm:nullstellensatz-horn} can be strengthened to
+of~$\Set[\TT]$, Corollary~\ref{corollary:nullstellensatz-horn} can be strengthened to
 allow~$\bot$ as consequent, since in this case we have~$(\nabla\psi)
@@
 
-% XXX subtopos case
+% XXX The internal statement occurring in Theorem~\ref{thm:quasicoherent} is
+% higher-order, therefore
+
+\begin{cor}Let~$\TT'$ be a quotient theory of~$\TT$, not necessarily Horn.
+Assume that the generic model~$U_\TT$ is a sheaf for the topology
+on~$\Set[\TT]$ cutting out the subtopos~$\Set[\TT']$. Then, from the internal
+point of view of~$\Set[\TT']$, the canonical map~$A \to (U_{\TT'})^{\Spec(A)}$
+is an isomorphism of~$U_\TT$-algebras for every finitely
+presented~$U_\TT$-algebra~$A$. (By~``$\Spec(A)$'', we mean the set
+of~$U_\TT$-algebra homomorphisms from~$A$ to~$U_{\TT'}$.)
+\end{cor}
 
@@
 Thanks to the version of the Nullstellensatz pertaining Horn consequents given
-in Theorem~\ref{thm:nullstellensatz-horn}, these two arguments can also be
+in Corollary~\ref{corollary:nullstellensatz-horn}, these two arguments can also be
 carried out in the classifying topos of local rings. It is instructive to

### 2019-05-08 paper-qcoh.tex
More on the subtopos case

@@
 
+\begin{proof}In general, the generic model~$U_{\TT'}$ is the sheafification
+of~$U_\TT$~\cite[Lemma~2.3]{caramello:definability}. By the sheaf
+assumption, the objects~$U_{\TT'}$ and~$U_\TT$ actually agree as functors
+on~$\Mod{\TT}_\mathrm{fp}$. Employing the lemma that computing the internal Hom
+to a sheaf commutes with sheafification twice, we see that
+the canonical isomorphism of Lemma~\ref{lemma:fp-double-dual} induces an
+isomorphism~$a(A^\sim) \to (U_{\TT'})^{\Spec(a(A^\sim))}$ ... XXX
+\end{proof}
+
 \section{The generalization to the higher-order case}

### 2019-05-08 paper-qcoh.tex
Subtopos case finished

--- 2019-05-05 paper-qcoh.tex
@@
 
-% XXX The internal statement occurring in Theorem~\ref{thm:quasicoherent} is
-% higher-order, therefore
-
-\begin{cor}Let~$\TT'$ be a quotient theory of~$\TT$, not necessarily Horn.
-Assume that the generic model~$U_\TT$ is a sheaf for the topology
+\begin{thm}Let~$\TT'$ be a quotient theory of~$\TT$, not necessarily Horn.
+Assume that~$U_\TT$ is a sheaf for the topology
 on~$\Set[\TT]$ cutting out the subtopos~$\Set[\TT']$. Then, from the internal
@@
 is an isomorphism of~$U_\TT$-algebras for every finitely
-presented~$U_\TT$-algebra~$A$. (By~``$\Spec(A)$'', we mean the set
+presented~$U_\TT$-algebra~$A$. (By~``$\Spec(A)$'', we here mean the set
 of~$U_\TT$-algebra homomorphisms from~$A$ to~$U_{\TT'}$.)
-\end{cor}
+\end{thm}
+
+\begin{proof}In general, the generic model~$U_{\TT'}$ is the sheafification
+of~$U_\TT$~\cite[Lemma~2.3]{caramello:definability}. By the sheaf assumption,
+the objects~$U_{\TT'}$ and~$U_\TT$ actually agree as functors
+on~$\Mod{\TT}_\mathrm{fp}$.
+
+Following the Kripke--Joyal translation of the claim, similar as in the proof
+of Theorem~\ref{thm:quasicoherence}, it suffices to show that
+applying the sheafification functor to the isomorphisms provided by
+Lemma~\ref{lemma:fp-double-dual} (over arbitrary slices) results in
+isomorphisms of the same form. This fact follows from the lemma (applied twice)
+that computing the internal Hom to a sheaf commutes with sheafification and
+from the observation that the construction~``free~$U_\TT$-module modulo
+relations'' is geometric, hence in particular commutes with sheafification.
+\end{proof}
 

### 2019-05-08 paper-qcoh.tex
Remark regarding a more minimal formulation

@@
 
+\begin{rem}Both in Scholium~\ref{scholium:nullstellensatz-more-specific} and in
+the axiom scheme~\eqref{eq:nullstellensatz}, it suffices to restrict the
+disjunction~``$\bigvee_\alpha \alpha$'' to those formulas~$\alpha$ which are of
+the form~``$\exists\cdots\exists\_ \varphi$'' with~$\varphi$ of Horn type.
+\end{rem}
+
 The Nullstellensatz of Theorem~\ref{thm:nullstellensatz} and the

### 2019-05-10 slides-munich2019.tex
Slides for tomorrow's talk in Munich

--- 2019-05-02 slides-birmingham2019.tex
@@
 
-\title{A general Nullstellensatz for generalised spaces}
+\title{A general Nullstellensatz for generalized spaces}
 \author{Ingo Blechschmidt}
-\date{April 11th, 2019}
+\date{May 10th, 2019}
 
@@
       \node (title) {\hil{\large\insertframetitle}};
-      \begin{pgfonlayer}{background}
-        \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple]
-          ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
-          ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
-          ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
-          ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
-          cycle;
-      \end{pgfonlayer}
     \end{tikzpicture}
@@
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses}\end{minipage}}
 \begin{frame}[c]
@@
   \bigskip
-  \includegraphics[width=0.4\textwidth]{phantoms}
   \bigskip
@@
     \def\R{8pt}
-    \node (title) {\hil{A general Nullstellensatz for generalised spaces}};
-    \begin{pgfonlayer}{background}
-      \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple]
-        ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
-        ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
-        ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
-        ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
-        cycle;
-    \end{pgfonlayer}
+    \node (title) {\hil{A general Nullstellensatz for generalized spaces}};
   \end{tikzpicture}
@@
   \scriptsize
-  \textit{-- an invitation --}
   \bigskip
 
-  6th Workshop on Formal Topology in Birmingham \\
-  April 11th, 2019
+  Logic Workshop in Munich \\
+  May 10th, 2019
   \bigskip
@@
   \par
-\end{frame}}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\hspace*{-1.7cm}\includegraphics[height=\paperheight,trim=0 0 16.9cm 0,clip]{black-hole-m87}\end{minipage}}
-\begin{frame}[plain]
-  \vspace*{\paperheight}\vspace*{-0.8cm}\scriptsize
-  \centering
-  \textcolor{white}{Fig.: The Event Horizon Telescope picture of the central black
-  hole in the galaxy M87}
-  \par
-
-  \jnote{1}{At this meeting we had lots of discussions on the notion of space.
-  Here is a picture of actual space, or rather the lack thereof, released only
-  yesterday.}
-\end{frame}}
-\addtocounter{framenumber}{-1}
+\end{frame}
 
@@
   With \emph{topos} we mean Grothendieck topos, and as metatheory we use a
-  constructive but impredicative flavour of English (which could be formalised
+  constructive but impredicative flavour of English (which could be formalized
   by what is supported by the internal language of elementary toposes with an NNO).
   However the Nullstellensatz presented later makes no use of the
-  subobject classifier, hence the results can likely be generalised to hold in
+  subobject classifier, hence the results can likely be generalized to hold in
   a predicative metatheory or to hold for arithmetic universes.}
@@
     \item The sequent~$\sigma$ holds for~$U_\TT$.
-    \item The sequent~$\sigma$ holds for any~$\TT$-model in any
-    \explainstub{topos}{purple!20}.
+    \item The sequent~$\sigma$ holds for any~(sheaf) model of~$\TT$.
     \item The sequent~$\sigma$ is provable modulo~$\TT$.
@@
 
-\begin{frame}{Construction of the generic model}
+\begin{frame}{On the generic model}
   The generic model is \hil{not} the same as \ldots
@@
 
-  \textbf{Definition.} The \hil{syntactic site}~$\C_\TT$ has \ldots
-  \begin{enumerate}
-    \item objects: $\{ x_1\?X_1, \ldots, x_n\?X_n\_ \varphi \}$ (shorter:
-    $\{ \vec x\_ \varphi \}$)
-    \item morphisms: eqv. classes of \explainstub{provably functional
-    formulas}{red!20}
-    \item coverings: \explainstub{provably jointly surjective
-    families}{purple!30}
-  \end{enumerate}
-  The topos of sheaves over~$\C_\TT$ is the \hil{classifying
-  topos}~$\Set[\TT]$. The generic model interprets a sort~$X$ by~$\yon\{x\?X\_\top\}$.
+  The generic model is a \hil{sheaf model}.
 
@@
 
-  In general, the generic model cannot be realised as a set-based model (with
+  In general, the generic model cannot be realized as a set-based model (with
   a set for each sort, a map for each function symbol and so on). Sets are too
@@
 
-  \jnote{2}{A morphism~$A = \{\vec x\_ \varphi\} \to \{\vec y\_ \psi\} = B$ in~$\C_\TT$ is the
-  equivalence class (modulo provable equivalence) of a geometric
-  formula~$\theta$ such that~$\TT$ proves
-  \begin{enumerate}
-    \item ``$\theta$ is a relation on~$A \times B$'': $(\theta \seq{\vec x,
-    \vec y} \varphi \wedge \psi)$
-    \item ``$\theta$ is total'': $(\varphi \seq{\vec x} \exists \vec y\_
-    \theta)$
-    \item ``$\theta$ is single-valued'': $(\theta \wedge \theta[\vec y'/\vec y]
-    \seq{\vec x,\vec y,\vec y'} \vec y = \vec y')$
-  \end{enumerate}
-
-  \vspace*{-0.6em}
-  A family~$(\{ \vec x_i\_ \varphi_i \} \xrightarrow{[\theta_i]} \{ \vec y\_
-  \psi \})_i$ is a covering iff~$\TT$ proves~$(\psi \seq{\vec y} \bigvee_i
-  \exists \vec x_i\_ \theta_i)$.
-
-  The slides experiment with using the symbol~``$\yon$'' for the Yoneda
-  embedding~$\C_\TT \to \Set[\TT]$, as in \fixedhref{http://www.math.jhu.edu/~eriehl/elements.pdf}{Elements
-  of~$(\infty,1)$-category theory} by Emily Riehl and Dominic Verity.}
-
-  \jnote{3}{The special case that the generic model of a theory~$\TT$ can be realised as
-  a model in~$\Set$ occurs iff~$\TT$ is Morita-equivalent to the
+  \jnote{2}{The special case that the generic model of a theory~$\TT$ can be realized as
+  a set-based model occurs iff~$\TT$ is Morita-equivalent to the
   empty theory, that is, iff~$\TT$ has exactly one model in any topos.
 
-  The special case that there exists at least some conservative~$\TT$-model
-  in~$\Set$ occurs iff~$\TT$ has a conservative geometric expansion to a theory
+  The special case that there exists at least some conservative set-based~$\TT$-model
+  occurs iff~$\TT$ has a conservative geometric expansion to a theory
   which is Morita-equivalent to the empty theory.}
@@
 
-\begin{frame}{Working internally to toposes}
+\begin{frame}{A primer on sheaf semantics}
   \justifying
-  Let~$\C$ be a site. We recursively define
+  \vspace*{-0.6em}
+  Write~``$\Set[\TT] \models \varphi$'' for~$\{\_\top\} \models \varphi$. We
+  define
   \[
-    U \models \varphi \quad
-    \text{(``$\varphi$ holds on~$U$'')}
+    \{\vec x\_ \alpha\} \models \varphi \quad
+    \text{(``$\varphi$ holds on stage~$\{\vec x\_\alpha\}$'')}
   \]
-  for objects~$U \in \C$ and \squiggly{formulas}~$\varphi$.
-  Write~``$\Sh(\C) \models \varphi$'' for~$1 \models \varphi$.
+  for Horn formulas~$\alpha$ (in contexts~$\vec x$) and first-order
+  formulas~$\varphi$.\par
+  \vspace*{-1.2em}
   \footnotesize
-  \[ \renewcommand{\arraystretch}{1.08}\begin{array}{@{}l@{\ \ }c@{\ \ }l@{}}
-  U \models \top &\Ll& \text{true} \\
-  U \models \bot &\Ll& \hcancel{\text{false}}{0pt}{3pt}{0pt}{-2pt}\ \text{the
-  empty family is a covering of~$U$} \\
-  U \models s = t \? F &\Ll& s|_U = t|_U \in F(U) \\
-  U \models \varphi \wedge \psi &\Ll&
-  \text{$U \models \varphi$ and $U \models \psi$} \\
-  U \models \varphi \vee \psi &\Ll&
-  \hcancel{\text{$U \models \varphi$ or $U \models
-  \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $(U_i \to U)_i$} \\
-  && \quad\quad\text{such that for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
-  U \models \varphi \Rightarrow \psi &\Ll&
-  \text{for all~$V \to U$: }
-  \text{$V \models \varphi$ implies $V \models \psi$} \\
-  U \models \forall s \? F\_ \varphi(s) &\Ll&
-  \text{for all $V \to U$ and sections~$s_0 \in F(V)$: $V \models \varphi(s_0)$} \\
+  \[
+  \renewcommand{\arraystretch}{1.08}\begin{array}{@{\!\!\!\!\!\!\!\!\!\!\!}l@{\ \ }c@{\ \ }l@{}}
+  \{\vec x\_ \alpha\} \models \top &\Ll& \text{true} \\
+  \{\vec x\_ \alpha\} \models \bot &\Ll& \hcancel{\text{false}}{0pt}{3pt}{0pt}{-2pt}\
+  \alpha \seq{\vec x} \bot \\
+  \{\vec x\_ \alpha\} \models s = t \? F &\Ll& s = t \in F(\{\vec x\_\alpha\}) \\
+  \{\vec x\_ \alpha\} \models \varphi \wedge \psi &\Ll&
+  \text{$\{\vec x\_ \alpha\} \models \varphi$ and $\{\vec x\_ \alpha\} \models \psi$} \\
+  \{\vec x\_ \alpha\} \models \varphi \vee \psi &\Ll&
+  \hcancel{\text{$\{\vec x\_ \alpha\} \models \varphi$ or $\{\vec x\_ \alpha\} \models
+  \psi$}}{0pt}{3pt}{0pt}{-2pt} \\
+  && \text{there exists a covering~$(\{ \vec y_i\_ \beta_i \} \xrightarrow{p_i} \{ \vec x\_ \alpha \})_i$} \\
+  && \quad\quad\text{such that for all~$i$: $\{\vec y_i\_ \beta_i\} \models
+  p_i^*\varphi$ or $\{\vec y_i\_ \beta_i\} \models p_i^*\psi$} \\
+  \{\vec x\_ \alpha\} \models \varphi \Rightarrow \psi &\Ll&
+  \text{for all~$\{\vec y\_\beta\} \xrightarrow{p} \{\vec x\_\alpha\}$: }
+  \text{$\{\vec y\_\beta\} \models p^*\varphi$ implies $\{\vec y\_\beta\} \models
+  p^*\psi$} \\
+  \{\vec x\_ \alpha\} \models \forall s \? F\_ \varphi &\Ll&
+  \text{for all $\{\vec y\_\beta\} \xrightarrow{p} \{\vec x\_\alpha\}$ and~$s_0
+  \in F(\{\vec y\_\beta\})$: $\{\vec y\_\beta\} \models (p^*\varphi)[s_0/s]$} \\
   % U \models \forall F\_ \varphi(F) &\Ll&
   % \text{for all $V \to U$ and sheaves~$F_0$ over~$V$: $V \models \varphi(F_0)$} \\
-  U \models \exists s \? F\_ \varphi(s) &\Ll&
-  \hcancel{\text{there exists $s_0 \in F(U)$ such that $U \models \varphi(s_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
+  \{\vec x\_ \alpha\} \models \exists s \? F\_ \varphi &\Ll&
+  \hcancel{\text{there exists $s_0 \in F(\{\vec x\_\alpha\})$ such that $\{\vec
+  x\_\alpha\} \models \varphi[s_0/s]$}}{0pt}{3pt}{0pt}{-2pt} \\
   &&
-  \text{there exists a covering $(U_i \to U)_i$ such that for all~$i$:} \\
-  && \quad\quad \text{there exists~$s_0 \in F(U_i)$ such that $U_i \models \varphi(s_0)$}
+  \text{there exists a covering $(\{ \vec y_i\_ \beta_i \} \xrightarrow{p_i} \{
+  \vec x\_ \alpha \})_i$ such that for all~$i$:} \\
+  && \quad\quad \text{there exists~$s_0 \in F(\{\vec y_i\_\beta_i\})$ such that
+  $\{\vec y_i\_\beta_i\} \models (p_i^*\varphi)[s_0/s]$}
   %U \models \exists F\_ \varphi(F) &\Ll&
@@
   if~\explainstub{$\sigma$ holds for~$U_\TT$}{yellow!70}, then~\explainstub{$\ul{\TT}/U_\TT$
-  proves$^\star$~$\sigma$}{yellow!70}.}
+  proves~$\sigma$}{yellow!70}.}
   \end{quote}
@@
 
-  \textbf{Remark.} Externalising the internal classifying
-  topos~$\Set[\TT][\ul{\TT}/U_\TT]$ yields the classifying topos
-  of~$\TT$-homomorphisms.
-
   \jnote{1}{
-    Just as locales internal to a topos~$\E$ can be externalised to yield
+    Just as locales internal to a topos~$\E$ can be externalized to yield
     localic geometric morphisms into~$\E$, internal Grothendieck toposes can be
-    externalised to yield bounded geometric morphisms. Since the composition of
+    externalized to yield bounded geometric morphisms. Since the composition of
     bounded geometric morphisms is bounded, the externalisation of a
@@
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\hfill\mbox{\includegraphics[width=0.13\paperwidth]{hen}\;\;}\end{minipage}}
 \begin{frame}{Exhaustion and extensions}
@@
   \begin{equation}\tag{Nullstellensatz} \speak{$\sigma$ holds} \implies \speak{$\ul{\TT}/U_\TT$
-  proves$^{\star\star}$~$\sigma$}. \end{equation}
+  proves~$\sigma$}. \end{equation}
 
@@
   Let~$\TT'$ be a quotient theory
-  of~$\TT$. Assume that~$U_\TT$ is contained in the subtopos~$\Set[\TT']$. Then internally
+  of~$\TT$. Assume that~$U_\TT$ is a sheaf for the topology cutting out~$\Set[\TT']$. Then internally
   to~$\Set[\TT']$:
@@
     \emph{A geometric$^\star$ sequent~$\sigma$ with Horn consequent holds for~$U_{\TT'}$
-    iff~$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$.}
+    iff~$\ul{\TT}/U_\TT$ proves~$\sigma$.}
   \end{quote}
 
-  \textbf{Theorem 3.}
-  % Let~$\mathrm{Form}^\star_{x\?X}(\ul{\TT}/U_\TT)/(\dashv\vdash_{x\?X})$ be the~$\Set[\TT]$-object of
-  % geometric$^\star$ formulas over the signature of~$\ul{\TT}/U_\TT$ in the
-  % context~$x\?X$, where any two such formulas are identified if and only
-  % if~$\ul{\TT}/U_\TT$ proves them equivalent.
-  The morphism~$\mathrm{ev}$ below is an isomorphism.
-  \begin{equation}\label{eq:hon}\tag{$\star$}
-    \mathrm{ev} : \mathrm{FunctFormulas}^\star(\ul{\TT}/U_\TT)/(\dashv\vdash) \longrightarrow P(U_\TT)
-  \end{equation}
-
-  \textbf{Theorem 4.} A higher-order formula holds for~$U_\TT$ iff it is
+  \textbf{Theorem 3.} A higher-order formula holds for~$U_\TT$ iff it is
   provable in intuitionistic higher-order logic from the axioms of~$\TT$ and
-  the higher-order Nullstellensatz scheme~($\star$).
+  the higher-order Nullstellensatz scheme.
 
@@
 
-    Theorems~3 and~4 generalise the Nullstellensatz to the
+    Theorems~3 and~4 generalize the Nullstellensatz to the
     higher-order setting. The map~$\mathrm{ev}$ maps (the equivalence class
@@
   }
-\end{frame}}
+\end{frame}
 

### 2019-05-11 paper-reflection.tex
typo, to do

@@
   equivalent to the conjunction of the relation reflection scheme and dependent
-  choice and we present an intuitionistic version of Feferman's~$\ZFCS$, a
+  choice, and we present an intuitionistic version of Feferman's~$\ZFCS$, a
   conservative extension of~$\ZFC$ which is useful as a foundation for category
@@
 For instance, the category of all small sets can be formed entirely
-within~$\ZFCS$, without resorting to classes. We invite the reader wanting
+within~$\ZFCS$, without resorting to classes. We invite the reader desiring
 to learn more about the merits of~$\ZFCS$ to study the survey by
@@
 commonly heralded as the largest common denominator of all flavors of
-constructive set theory, remains even more elusive.
+constructive set theory, remains even more elusive. (XXX, update: $\CZF$ does
+not, and cannot be expected to, verify the reflection principle.)
 
@@
 We conclude in Section~\ref{sect:appl-rdc} and
-Section~\ref{sect:appl-smallness} with two short applications.
+Section~\ref{sect:appl-smallness} with two applications.
 
@@
 reading of earlier drafts. XXX
+
+% XXX briefly recall the language of IZF/ZF.
+% XXX state free variables more precisely
 
@@
 
-The proof of Lemma~\ref{lemma:zf-smallstep} is mostly constructive; however,
+The proof constructed by Lemma~\ref{lemma:zf-smallstep} is mostly constructive; however,
 there is one issue with nontrivial ramifications: While~$\IZF$ does show
@@
 that~$X^\sim$ is inhabited if~$X$ is. This would amount to the constructive
-taboo that any inhabited set contains a rank-minimal element; and in fact, by
+taboo that any inhabited set contains a rank-minimal element. Moreover, by
 a result of Friedman and Scedrov~\cite{XXX}, no definable substitute
@@
 The remaining case is where~$\varphi$ is of the form~$\varphi \equiv (\exists
-x\_ \psi(u_1,\ldots,u_m,x))$. In this case, the claim is that
+x\_ \psi(u_1,\ldots,u_m,x))$. In this case the claim is that
 \[ \forall u_1,\ldots,u_m \in S\_
@@
 as easy.
+% XXX give the proof
 \end{rem}
+
+% XXX introduce CZF^-
 
@@
 \end{array} \]
-Hence the set~$H' \defeq M \cup \bigcup X$ has the required property.
+Hence the set~$H' \defeq H \cup \bigcup X$ has the required property.
 
@@
 finitely-iterated powersets.
+% XXX what do we need for this?
 
@@
 We set~$S \defeq \bigcup B$. This set has the required property. To verify
-this, it is useful to observe that given~$u_1,\ldots,u_n \in S$, there is a
+this claim, it is useful to observe that given~$u_1,\ldots,u_n \in S$, there is a
 common set~$H \in B$ such that~$u_1,\ldots,u_n \in H$.
 
-In order to ensure addendum~(1) and~(2), we apply~$\RRS_2$ in a slightly
+In order to ensure addenda~(1) and~(2), we apply~$\RRS_2$ in a slightly
 different way to guarantee that for any sets~$H_1, H_2 \in B$, there is a
@@
 universe~$U$; we can form the category~$\Set_U$ of all sets in~$U$ as an honest
-set; can define truth in~$\Set_U$; and prove, within the system, that for any
+set; define truth in~$\Set_U$; and prove, within the system, that for any
 formula~$\varphi$, $(\Set_U \models \varphi) \Leftrightarrow (U \models

### 2019-05-15 paper-reflection.tex
RRS_2 passes from the metatheory to sheaf models (WIP)

@@
 \newcommand{\ZZ}{\mathbb{Z}}
+\newcommand{\CC}{\mathbb{C}}
 \renewcommand{\SS}{\mathbb{S}}
@@
 \newcommand{\notnot}{\emph{not~not}\xspace}
+\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
 
@@
 
+Fourman and independently Hayashi have shown how to interpret~$\IZF$ in any
+Grothendieck topos~\cite{fourman:izf,hayashi:izf}. The following proposition
+shows that the axiom scheme~$\RRS_2$ passes from the metatheory to any such
+model:
+
+\begin{prop}Over~$\IZF + \RRS_2$, $\RRS_2$ holds in any Grothendieck topos.
+\end{prop}
+
+\begin{proof}We employ Streicher's version of the interpretation
+of~$\IZF$~\cite{streicher:forcizf}, and follow him also in the notation.
+
+Let~$\CC$ be a Grothendieck site, assumed without loss of generality to possess
+a terminal object~$1$ and pullbacks. Let~$V$ be the universe sheaf over~$\CC$. Let
+formulas~$\chi$ and~$\varrho$ be given, with parameters from~$V(1)$, defining
+classes~$X$ and~$R$ in~$\Sh(\CC)$. Let an element~$x_0 \in V(1)$ be given and assume
+\[ 1 \models \speak{$R \subseteq X \times X \times X$} \qquad\text{and}\qquad
+  1 \models \speak{$R : X \times X \rightrightarrows X$}, \]
+where the corner quotes indicate that translation to formal language is left to
+the reader.
+
+For an object~$I \in \CC$, let~$V_I$ be the (external) class of pairs~$(J
+\xrightarrow{u} I, x \in V(J))$. Define the (external) classes
+\begin{align*}
+  X' &\defeq \{ (I,F) \,|\, I \in \CC, F \in P(V_I), \mathrm{fst}[F] \in
+  \mathrm{Cov}(I), \forall (J \xrightarrow{u} I, x \in V(J)) \in F\_ J \models x \in X \}
+\end{align*}
+XXX
+\end{proof}
+
 \section{Constructivizing the reflection principle}

### 2019-07-07 slides-edinburgh2019.tex
Slides for CT 2019 (almost finished)

--- 2019-05-02 slides-birmingham2019.tex
@@
 
-\usepackage{pgfpages}
-\setbeameroption{show notes on second screen}
-\setbeamertemplate{note page}[plain]
-\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}
+%\usepackage{pgfpages}
+%\setbeameroption{show notes on second screen}
+%\setbeamertemplate{note page}[plain]
+\newcommand{\jnote}[2]{}
 
@@
       \begin{pgfonlayer}{background}
-        \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple]
+        \draw[decorate, very thick, draw=mypurple]
           ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
@@
   \bigskip
-  \includegraphics[width=0.4\textwidth]{phantoms}
+  \[ \xymatrix{
+    & \text{$\ul{\TT}/U_\TT$ proves $\sigma$} \\
+    \text{$\sigma$ holds for $U_\TT$} \ar@/^1.1pc/@{<=>}[ur] &&
+    \text{$\sigma$ holds for $M$} \ar@/_1.1pc/@{<=}[ul]
+  } \]
   \bigskip
@@
     \begin{pgfonlayer}{background}
-      \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple]
+      \draw[decorate, very thick, draw=mypurple]
         ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
@@
 
-  6th Workshop on Formal Topology in Birmingham \\
-  April 11th, 2019
+  Category Theory 2019 in Edinburgh \\
+  July 9th, 2019
   \bigskip
@@
   \begin{quote}
-    \emph{\mbox{For any finitely presented~$\affl$-algebra~$A$, the canonical map}
+    \emph{\mbox{For any finitely presented~$\affl$-algebra~$A$, the ``double dual'' map}
     \[ A \longrightarrow (\affl)^{\Hom_\affl(A,\affl)},\ s \longmapsto (x
@@
 
-    The asterisks in \emph{geometric$^\star$ sequent} and
-    \emph{provability$^\star$} indicate that any infinities used to index
+    The asterisks in \emph{geometric$^\star$ sequent}
+    indicate that any infinities used to index
     disjunctions have to be come from the base topos. This restriction is an important
@@
   \nullstellensatz
-  \medskip
-
+  \[ \xymatrix{
+    & \text{$\ul{\TT}/U_\TT$ proves $\sigma$} \\
+    \text{$\sigma$ holds for $U_\TT$} \ar@/^1.1pc/@{<=>}[ur] &&
+    \text{$\sigma$ holds for $M$} \ar@/_1.1pc/@{<=}[ul]
+  } \]
+
+  \only<1>{
     \textbf{In the object classifier.}
@@
     Let~$x,y \? U_\TT$. Assume that~$\neg(x = y)$.
-  By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves~$(e_x = e_y \vdash \bot)$.
-  But this is false in the~$\ul{\TT}/U_\TT$-model~$U_\TT/(x \sim y)$.
-  \medskip
-
+    By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves~$\sigma \defeq (e_x = e_y \vdash \bot)$.
+    But this is false in the~$\ul{\TT}/U_\TT$-model~$M \defeq U_\TT/(x \sim y)$.
+  }
+
+  \only<2->{
     \textbf{In the ring classifier.}
@@
     in~$U_\TT[X]/(f)$, that is~$g = hf$ for some~$h \? U_\TT[X]$.
+  }
 
@@
     statements~$\speak{$\sigma$ holds for~$U_\TT$}$
-    and~$\speak{$\ul{\TT}/U_\TT$ proves$^\star$ $\sigma$}$ are equivalent. This
+    and~$\speak{$\ul{\TT}/U_\TT$ proves $\sigma$}$ are equivalent. This
     equivalence is intriguing from a logical point of view, since the former
@@
     Often when using the Nullstellensatz, we go from an (assumed) truth
-    of~$U_\TT$ via provability$^\star$ to another model~$M$
+    of~$U_\TT$ via provability to another model~$M$
     of~$\ul{\TT}/U_\TT$. That is, we use provability$^\star$ as a (one-way)
@@
     \[ \xymatrix{
-      & \text{$\ul{\TT}/U_\TT$ proves$^\star$ $\sigma$} \\
+      & \text{$\ul{\TT}/U_\TT$ proves $\sigma$} \\
       \text{$\sigma$ holds for $U_\TT$} \ar@/^1.1pc/@{<=>}[ur] &&
@@
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\hfill\mbox{\includegraphics[width=0.13\paperwidth]{hen}\;\;}\end{minipage}}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{7.95cm}\hfill\mbox{\includegraphics[width=0.13\paperwidth]{hen}\;\;}\end{minipage}}
 \begin{frame}{Exhaustion and extensions}
@@
   \begin{equation}\tag{Nullstellensatz} \speak{$\sigma$ holds} \implies \speak{$\ul{\TT}/U_\TT$
-  proves$^{\star\star}$~$\sigma$}. \end{equation}
+  proves~$\sigma$}. \end{equation}
+  \vspace*{-1.5em}
 
@@
   to~$\Set[\TT']$:
-
-  \vspace*{-0.4em}
   \begin{quote}
     \emph{A geometric$^\star$ sequent~$\sigma$ with Horn consequent holds for~$U_{\TT'}$
-    iff~$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$.}
+    iff~$\ul{\TT}/U_\TT$ proves~$\sigma$.}\par
   \end{quote}
 
-  \textbf{Theorem 3.}
-  % Let~$\mathrm{Form}^\star_{x\?X}(\ul{\TT}/U_\TT)/(\dashv\vdash_{x\?X})$ be the~$\Set[\TT]$-object of
-  % geometric$^\star$ formulas over the signature of~$\ul{\TT}/U_\TT$ in the
-  % context~$x\?X$, where any two such formulas are identified if and only
-  % if~$\ul{\TT}/U_\TT$ proves them equivalent.
-  The morphism~$\mathrm{ev}$ below is an isomorphism.
-  \begin{equation}\label{eq:hon}\tag{$\star$}
-    \mathrm{ev} : \mathrm{FunctFormulas}^\star(\ul{\TT}/U_\TT)/(\dashv\vdash) \longrightarrow P(U_\TT)
-  \end{equation}
-
-  \textbf{Theorem 4.} A higher-order formula holds for~$U_\TT$ iff it is
+  \textbf{Theorem 3.} A higher-order formula holds for~$U_\TT$ iff it is
   provable in intuitionistic higher-order logic from the axioms of~$\TT$ and
-  the higher-order Nullstellensatz scheme~($\star$).
+  the higher-order Nullstellensatz scheme.
 

### 2019-07-08 slides-edinburgh2019.tex
Final version of the slides

@@
 
+\usepackage{graphbox}
 \usepackage{mathtools}
@@
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\hspace*{-1.7cm}\includegraphics[height=\paperheight,trim=0 0 16.9cm 0,clip]{black-hole-m87}\end{minipage}}
-\begin{frame}[plain]
-  \vspace*{\paperheight}\vspace*{-0.8cm}\scriptsize
-  \centering
-  \textcolor{white}{Fig.: The Event Horizon Telescope picture of the central black
-  hole in the galaxy M87}
-  \par
-
-  \jnote{1}{At this meeting we had lots of discussions on the notion of space.
-  Here is a picture of actual space, or rather the lack thereof, released only
-  yesterday.}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\hspace*{-1.7cm}\includegraphics[height=\paperheight]{arthurs-seat-4}\end{minipage}}
+\begin{frame}[plain]%
+  \vspace*{\paperheight}\vspace*{-2.35cm}\hspace*{-1cm}%
+  \includegraphics[width=\paperwidth/4]{arthurs-seat-7}%
+  \includegraphics[width=\paperwidth/4]{arthurs-seat-0}%
+  \includegraphics[width=\paperwidth/4]{arthurs-seat-5}%
+  \includegraphics[width=\paperwidth/4]{arthurs-seat-9}%
 \end{frame}}
@@
   \begin{enumerate}
-    \item objects: $\{ x_1\?X_1, \ldots, x_n\?X_n\_ \varphi \}$ (shorter:
-    $\{ \vec x\_ \varphi \}$)
+    \item objects: \explainstub{formulas-in-contexts}{red!20} $\{ x_1\?X_1, \ldots, x_n\?X_n\_ \varphi \}$
     \item morphisms: eqv. classes of \explainstub{provably functional
-    formulas}{red!20}
+    formulas}{red!30}
     \item coverings: \explainstub{provably jointly surjective
@@
   The topos of sheaves over~$\C_\TT$ is the \hil{classifying
-  topos}~$\Set[\TT]$. The generic model interprets a sort~$X$ by~$\yon\{x\?X\_\top\}$.
+  topos}~$\Set[\TT]$. The generic model interprets a sort~$X$ by the sheaf~$\yon\{x\?X\_\top\}$.
 
@@
   for objects~$U \in \C$ and \squiggly{formulas}~$\varphi$.
-  Write~``$\Sh(\C) \models \varphi$'' for~$1 \models \varphi$.
-  \footnotesize
+  Write~``$\Sh(\C) \models \varphi$'' for~$U \models \varphi$ for all~$U$.
+  \only<1>{\footnotesize
   \[ \renewcommand{\arraystretch}{1.08}\begin{array}{@{}l@{\ \ }c@{\ \ }l@{}}
@@
   %&& \quad\quad \text{there exists a sheaf~$F_0$ on $U_i$ such that $U_i \models \varphi(F_0)$}
-  \end{array} \]
+  \end{array} \]}
+
+  \only<2>{
+    \bigskip
+    \begin{columns}[t]
+      \begin{column}{0.45\textwidth}\justifying
+        \textbf{Example.}
+        $\Set \models \varphi$ if and only if $\varphi$ is true.
+        \bigskip
+
+        \textbf{Theorem.}
+        If~$\Sh(\C) \models \varphi$ and if $\varphi$ intuitionistically
+        entails~$\psi$, then also~$\Sh(\C) \models \psi$.
+      \end{column}
+
+      \begin{column}{0.45\textwidth}
+        \includegraphics[width=\textwidth,align=t]{mike-shulman-stack-semantics}
+        \vspace*{1.5em}
+      \end{column}
+    \end{columns}
+  }
 
@@
 
-\section{Affine schemes}
-
-\begin{frame}{Affine schemes}
-  \jnote{1}{
-    A ring is \emph{local} iff every invertible sum contains an invertible
-    summand, that is if~$1 \neq 0$ and if~$x + y$ invertible implies~$x$
-    invertible or~$y$ invertible. Assuming the axiom of choice, this elementary
-    definition is equivalent to the textbook definition of a local ring (a ring
-    with exactly one maximal ideal). A ring homomorphism is \emph{local} iff it
-    reflects invertibility.
-
-    The notion of a \emph{filter} in a ring~$A$ is a direct axiomatisation of
-    what classically would be the complement of a prime ideal. The filter
-    axioms are: $1 \in F$, $(xy \in F) \Leftrightarrow ((x \in F) \wedge (y \in
-    F))$, $\neg(0 \in F)$, $(x+y \in F) \Rightarrow ((x \in F) \vee (y \in F))$.
-
-    A \emph{free local ring} over~$A$ is a homomorphism into a local
-    ring~$A'$ such that any homomorphism into a local ring~$R$ factors uniquely
-    over~$A'$ via a local homomorphism.
-
-    For any particular local ring~$A \xrightarrow{f} R$, the
-    localisation~$A[S^{-1}]$ with~$S = f^{-1}[R^\times]$ is a local ring which
-    fits into the displayed diagram. However, in general there is no single
-    choice of~$S$ which would work for any local ring~$R$. Indeed, classically
-    one can show that a free local ring over~$A$ exists if and only if~$A$
-    contains exactly one prime ideal, in which case~$A$ itself is the free
-    local ring.
-  }
-
-  \jnote{2}{
-    If we want a free local ring to exist for any ring~$A$, we have to broaden
-    our notion of existence and embrace rings which live in toposes other
-    than~$\Set$. There is a notion of a homomorphism between rings living in
-    arbitrary toposes, and using this notion one can verify:
-
-    The free local ring over~$A$ can be built in the classifying topos of
-    filters of~$A$, as the localisation (in that topos) of~$A$ at the generic
-    filter. (More precisely, as the localisation of the mirror image of~$A$ in that
-    topos, that is the constant sheaf~$\underline{A}$.)
-
-    The classifying topos of filters of~$A$ coincides with the topos of sheaves
-    over what's called the \emph{spectrum of~$A$} in algebraic geometry, and
-    under this equivalence the free local ring coincides with the structure
-    sheaf of spectrum. In fact the classifying topos serves as a good
-    constructive substitute for the classical spectrum construction,
-    enjoying the expected universal property even if the axiom of choice is not
-    available, which is why it is simply denoted~``$\Spec(A)$'' on the slide.
-
-    (The classifying topos of prime ideals of~$A$ is also interesting; it
-    coincides with the topos of sheaves over the spectrum of~$A$ equipped with
-    the constructible topology.)
-  }
-
-  \jnote{3}{
-    Assuming the Boolean prime ideal theorem, the geometric sequents
-    validated by~$A^\sim$ are easy to describe: They are precisely those which
-    are validated by all the stalks~$A_\ppp$ of~$A$.
-
-    But~$A^\sim$ enjoys further unique properties which are not shared by the
-    stalks of~$A$, other localisations of~$A$, quotients of~$A$ or indeed any
-    reasonable construction. Three of these are displayed on the slide.
-    (A ring is \emph{anonymously Noetherian} iff each of its ideals is
-    \emph{not not} finitely generated. Textbook proofs of Hilbert's basis theorem
-    are constructively acceptable for this Noetherian condition.)
-
-    The object~$A^\sim$ strikes a fine balance: On the one hand, it is still
-    close to~$A$, so that information learned about~$A^\sim$ teaches us
-    about~$A$; on the other hand, it enjoys unique properties rendering it
-    simpler than~$A$.
-
-    This balance allows for a simple and conceptually satisfying proof of
-    \emph{Gro\-then\-dieck's generic freeness lemma}, an important theorem in
-    algebraic geometry. Details can be found in
-    \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-padova2018.pdf}{this
-    set of slides}.
-  }
-
-  Let~$A$ be a ring. Is there a \hil{free local ring}~$A \to A'$ over~$A$?
-  \begin{columns}[t]
-    \begin{column}{0.4\textwidth}
-      $\xymatrix{
-        A \ar[rd] \ar[rrr]^f &&& {\substack{\phantom{\text{local}}\\\text{\normalsize$R$}\\\text{local}}} \\
-        & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!35]{\text{local}}[rru]
-      }$
-    \end{column}
-
-    \begin{column}{0.50\textwidth}
-      \small\justifying
-      For a fixed ring~$R$, the localisation $A' \defeq A[S^{-1}]$ with $S \defeq
-      f^{-1}[R^\times]$ would do the job. ($S$ is a \emph{filter}.)
-      \medskip
-
-      Hence we need the \hil{generic filter}.
-    \end{column}
-  \end{columns}
-
-  \pause
-
-  \justifying
-  The free local ring over~$A$ is~$A^\sim \defeq \ul{A}[F^{-1}]$,
-  where~$F$ is the generic filter, living in~$\Spec(A)$, the classifying topos
-  of filters of~$A$.
-
-  \pause
-
-  If~$A$ is reduced ($x^n = 0 \Rightarrow x = 0$):
-
-  \vspace*{-0.8em}
+\begin{frame}{A different kind of example}
+  Let~$A$ be a reduced ring ($x^n = 0 \Rightarrow x = 0$).
+
+  The \hil{generic local localisation}~$A^\sim$ of~$A$ validates:
+
+  \vspace*{-1.8em}
   \setbeamercolor{block body}{bg=red!30}
@@
   \end{varblock}
+
+  Consequence: short and conceptual proof of Grothendieck's generic freeness
+  lemma
 \end{frame}
+
+%\section{Affine schemes}
+%
+%\begin{frame}{Affine schemes}
+%  \jnote{1}{
+%    A ring is \emph{local} iff every invertible sum contains an invertible
+%    summand, that is if~$1 \neq 0$ and if~$x + y$ invertible implies~$x$
+%    invertible or~$y$ invertible. Assuming the axiom of choice, this elementary
+%    definition is equivalent to the textbook definition of a local ring (a ring
+%    with exactly one maximal ideal). A ring homomorphism is \emph{local} iff it
+%    reflects invertibility.
+%
+%    The notion of a \emph{filter} in a ring~$A$ is a direct axiomatisation of
+%    what classically would be the complement of a prime ideal. The filter
+%    axioms are: $1 \in F$, $(xy \in F) \Leftrightarrow ((x \in F) \wedge (y \in
+%    F))$, $\neg(0 \in F)$, $(x+y \in F) \Rightarrow ((x \in F) \vee (y \in F))$.
+%
+%    A \emph{free local ring} over~$A$ is a homomorphism into a local
+%    ring~$A'$ such that any homomorphism into a local ring~$R$ factors uniquely
+%    over~$A'$ via a local homomorphism.
+%
+%    For any particular local ring~$A \xrightarrow{f} R$, the
+%    localisation~$A[S^{-1}]$ with~$S = f^{-1}[R^\times]$ is a local ring which
+%    fits into the displayed diagram. However, in general there is no single
+%    choice of~$S$ which would work for any local ring~$R$. Indeed, classically
+%    one can show that a free local ring over~$A$ exists if and only if~$A$
+%    contains exactly one prime ideal, in which case~$A$ itself is the free
+%    local ring.
+%  }
+%
+%  \jnote{2}{
+%    If we want a free local ring to exist for any ring~$A$, we have to broaden
+%    our notion of existence and embrace rings which live in toposes other
+%    than~$\Set$. There is a notion of a homomorphism between rings living in
+%    arbitrary toposes, and using this notion one can verify:
+%
+%    The free local ring over~$A$ can be built in the classifying topos of
+%    filters of~$A$, as the localisation (in that topos) of~$A$ at the generic
+%    filter. (More precisely, as the localisation of the mirror image of~$A$ in that
+%    topos, that is the constant sheaf~$\underline{A}$.)
+%
+%    The classifying topos of filters of~$A$ coincides with the topos of sheaves
+%    over what's called the \emph{spectrum of~$A$} in algebraic geometry, and
+%    under this equivalence the free local ring coincides with the structure
+%    sheaf of spectrum. In fact the classifying topos serves as a good
+%    constructive substitute for the classical spectrum construction,
+%    enjoying the expected universal property even if the axiom of choice is not
+%    available, which is why it is simply denoted~``$\Spec(A)$'' on the slide.
+%
+%    (The classifying topos of prime ideals of~$A$ is also interesting; it
+%    coincides with the topos of sheaves over the spectrum of~$A$ equipped with
+%    the constructible topology.)
+%  }
+%
+%  \jnote{3}{
+%    Assuming the Boolean prime ideal theorem, the geometric sequents
+%    validated by~$A^\sim$ are easy to describe: They are precisely those which
+%    are validated by all the stalks~$A_\ppp$ of~$A$.
+%
+%    But~$A^\sim$ enjoys further unique properties which are not shared by the
+%    stalks of~$A$, other localisations of~$A$, quotients of~$A$ or indeed any
+%    reasonable construction. Three of these are displayed on the slide.
+%    (A ring is \emph{anonymously Noetherian} iff each of its ideals is
+%    \emph{not not} finitely generated. Textbook proofs of Hilbert's basis theorem
+%    are constructively acceptable for this Noetherian condition.)
+%
+%    The object~$A^\sim$ strikes a fine balance: On the one hand, it is still
+%    close to~$A$, so that information learned about~$A^\sim$ teaches us
+%    about~$A$; on the other hand, it enjoys unique properties rendering it
+%    simpler than~$A$.
+%
+%    This balance allows for a simple and conceptually satisfying proof of
+%    \emph{Gro\-then\-dieck's generic freeness lemma}, an important theorem in
+%    algebraic geometry. Details can be found in
+%    \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-padova2018.pdf}{this
+%    set of slides}.
+%  }
+%
+%  Let~$A$ be a ring. Is there a \hil{free local ring}~$A \to A'$ over~$A$?
+%  \begin{columns}[t]
+%    \begin{column}{0.4\textwidth}
+%      $\xymatrix{
+%        A \ar[rd] \ar[rrr]^f &&& {\substack{\phantom{\text{local}}\\\text{\normalsize$R$}\\\text{local}}} \\
+%        & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!35]{\text{local}}[rru]
+%      }$
+%    \end{column}
+%
+%    \begin{column}{0.50\textwidth}
+%      \small\justifying
+%      For a fixed ring~$R$, the localisation $A' \defeq A[S^{-1}]$ with $S \defeq
+%      f^{-1}[R^\times]$ would do the job. ($S$ is a \emph{filter}.)
+%      \medskip
+%
+%      Hence we need the \hil{generic filter}.
+%    \end{column}
+%  \end{columns}
+%
+%  \pause
+%
+%  \justifying
+%  The free local ring over~$A$ is~$A^\sim \defeq \ul{A}[F^{-1}]$,
+%  where~$F$ is the generic filter, living in~$\Spec(A)$, the classifying topos
+%  of filters of~$A$.
+%
+%  \pause
+%
+%  If~$A$ is reduced ($x^n = 0 \Rightarrow x = 0$):
+%
+%  \vspace*{-0.8em}
+%  \setbeamercolor{block body}{bg=red!30}
+%  \setbeamercolor{structure}{fg=purple}
+%  \begin{varblock}{\textwidth}{}
+%    $A^\sim$ is a \hil{field}: $\forall x\?A^\sim\_ (\neg(\exists y\?A^\sim\_
+%    xy = 1) \Rightarrow x = 0)$.
+%
+%    $A^\sim$ has \hil{$\boldsymbol{\neg\neg}$-stable equality}:
+%    $\forall x,y\?A^\sim\_ \neg\neg(x = y) \Rightarrow x = y$.
+%
+%    \mbox{$A^\sim$ is \hil{anonymously Noetherian}.}\\[-1.2em]
+%  \end{varblock}
+%\end{frame}
 
@@
 
+  \vspace*{-0.3em}
   \begin{quote}\emph{\mbox{For any \explainstub{geometric$^\star$ sequent~$\sigma$}{red!20} over
@@
     \item for each element~$x : U_\TT$ a constant symbol~$e_x$,
-    \item for each function symbol~$f$ and~$n$-tuple~$(x_1,\ldots,x_n) \in (U_\TT)^n$ the axiom
+    \item for each function symbol~$f$ and~$n$-tuple~$(x_1,\ldots,x_n) : (U_\TT)^n$ the axiom
     $(\top \vdash f(e_{x_1},\ldots,e_{x_n}) = e_{f(x_1,\ldots,x_n)})$,
-    \item for each relation symbol~$R$ and~$n$-tuple~$(x_1,\ldots,x_n) \in (U_\TT)^n$ such
+    \item for each relation symbol~$R$ and~$n$-tuple~$(x_1,\ldots,x_n) : (U_\TT)^n$ such
     that~$R(x_1,\ldots,x_n)$ the axiom~$(\top \vdash R(e_{x_1},\ldots,e_{x_n}))$.
@@
   \nullstellensatz
+  \vspace*{-2em}
   \[ \xymatrix{

### 2019-07-10 slides-birmingham2019.tex
Note slides finished

@@
     \emph{provability$^\star$} indicate that any infinities used to index
-    disjunctions have to be come from the base topos. This restriction is an important
+    disjunctions have to come from the base topos. This restriction is an important
     subtlety, though not vital to this talk. If~$\TT$ is a coherent
@@
   \jnote{3}{
-    While, as stated on slide 1/20, the generic model~$U_\TT$ is a
+    While, as stated on slide 1/10, the generic model~$U_\TT$ is a
     conservative~$\TT$-model, the classifying topos~$\Set[\TT]$ does not

### 2019-07-10 slides-edinburgh2019.tex
Note slides finished

@@
 
-%\usepackage{pgfpages}
-%\setbeameroption{show notes on second screen}
-%\setbeamertemplate{note page}[plain]
-\newcommand{\jnote}[2]{}
+\usepackage{pgfpages}
+\setbeameroption{show notes on second screen}
+\setbeamertemplate{note page}[plain]
+\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}
 
@@
   \par
+
+  \jnote{1}{\textbf{This talk in a nutshell:}
+
+  There is a certain ring, the \emph{generic ring}, which is special in that
+  it is \emph{conservative}: It has exactly those properties which are shared
+  by any ring whatsoever. One could argue that it is this ring which we're
+  implicitly referring to when we utter the phrase ``Let~$R$ be a ring.''.
+
+  A surprising observation is that the generic ring is a field. This is not a
+  contradiction because conservativity only pertains to properties which can be
+  put in a certain syntactic form, dubbed \emph{geometric sequents}, and the
+  field condition is not of that form.
+
+  As a consequence, when verifying a geometric sequent for all rings, we may
+  freely assume the field condition: Our proof will apply to the generic ring,
+  because this ring happens to be a field, and its conclusion will spread to
+  any ring whatsoever thanks to its syntactic form.
+
+  An analogous story can be told for the generic group, the generic vector
+  space and quite generally for the generic model of any \emph{geometric
+  theory}. It is an interesting task to discover properties of such
+  generic models. This talk presents a source of such properties, the
+  \emph{topos-theoretic Nullstellensatz}, which is
+  universal in the sense that any such property is a logical consequence of the
+  Nullstellensatz.}
 \end{frame}}
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\hspace*{-1.7cm}\includegraphics[height=\paperheight]{arthurs-seat-4}\end{minipage}}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\hspace*{-1.7cm}\includegraphics[height=\paperheight,trim=0 0 5.1cm 0,clip]{arthurs-seat-4}\end{minipage}}
 \begin{frame}[plain]%
@@
   \includegraphics[width=\paperwidth/4]{arthurs-seat-9}%
+  \jnote{1}{This talk owes a special debt to Alexander Oldenziel, who
+  repeatedly suggested that the topic of this talk should be studied.
+
+  The photos on the slide are of \emph{Arthur's Seat}, a beautiful extinct
+  volcano in Edinburgh. It gives an excellent view of the city and is very much
+  recommended for a hiking tour.
+  }%
 \end{frame}}
@@
 \begin{frame}{The mystery of nongeometric sequents}
-  \jnote{1}{A \emph{geometric sequent} is a syntactical expression of the
+  \jnote{1}{A \emph{geometric sequent} is a syntactic expression of the
   form~$(\varphi \seq{x_1\?X_1,\ldots,x_n\?X_n} \psi)$,
@@
   the~$X_i$ ranging over the available sorts, and~$\varphi$ and~$\psi$ are
-  \emph{geometric formulas}. Often the variable context is abbreviated to~$\vec
-  x \? \vec X$ or even just~$\vec x$. Such a sequent is read as ``in the
-  context of variables~$\vec x$, $\varphi$ entails~$\psi$''.
-
-  Geometric formulas are built from atomic propositions (using equality or the
-  relation symbols) using the connectives~$\top$, $\bot$, $\wedge$, $\bigvee$
+  \emph{geometric formulas}. Such a sequent is read as ``in the
+  context~$x_1\?X_1,\ldots,x_n\?X_n$, $\varphi$ entails $\psi$''.
+  Geometric formulas are built from atomic propositions (equality and the
+  available relation symbols) using the connectives~$\top$, $\bot$, $\wedge$, $\bigvee$
   (set-indexed disjunction) and~$\exists$. Geometric formulas may not
@@
   There is a notion of a \emph{model} of a geometric theory in
-  a given topos. For instance, a ring in the usual sense is a model of the
+  any given topos. For instance, a ring in the usual sense is a model of the
   theory of rings in the topos~$\Set$. The structure sheaf of a scheme~$X$ is a
@@
   Hence one reason why these nongeometric sequents are interesting is because
-  they provide us with new reduction strategies (``without loss of
+  they provide us with new reduction strategies (proofs by ``without loss of
   generality'').}
@@
   an equational theory), the \emph{term algebra} (the set of terms in the
-  empty context modulo provable equality) is a model of~$\TT$. While such
-  models do enjoy some nice categorical properties, they are in general
+  empty context modulo provable equality) is a model of~$\TT$. While the
+  term algebra does enjoy some nice categorical properties, it is in general
   \emph{not} the generic model.
@@
   required: The generic model lives in the topos of set-valued
-  sheaves over~$\C_\TT$.}
-
-  \jnote{2}{A morphism~$A = \{\vec x\_ \varphi\} \to \{\vec y\_ \psi\} = B$ in~$\C_\TT$ is the
+  sheaves over~$\C_\TT$, called the \emph{classifying topos} of~$\TT$.}
+
+  \jnote{2}{The notation in the definition of the objects of~$\C_\TT$ is meant
+  to be suggestive, but should not be taken literally. The objects are
+  syntactic expressions consisting of a context and a geometric formula in that
+  context.
+
+  A morphism~$A = \{\vec x\_ \varphi\} \to \{\vec y\_ \psi\} = B$ in~$\C_\TT$ is the
   equivalence class (modulo provable equivalence) of a geometric
@@
     Grothendieck topos.
-
-    We can actually do mathematics internally because the Kripke--Joyal
-    semantics is sound with respect to intuitionistic logic: If~$\E \models
-    \varphi$ and if~$\varphi$ intuitionistically entails a further
-    formula~$\psi$, then~$\E \models \psi$.
-
-    An instructive special case is provided by the topos~$\Set$, because~$\Set
-    \models \varphi$ iff~$\varphi$ holds in the usual mathematical sense.
   }
@@
   \jnote{2}{
+    We can actually \emph{do mathematics} internally to a topos because the
+    Kripke--Joyal semantics is sound with respect to intuitionistic logic:
+    If~$\E \models \varphi$ and if~$\varphi$ intuitionistically entails a
+    further formula~$\psi$, then~$\E \models \psi$.
+
     The Kripke--Joyal semantics can be extended to interpret unbounded
@@
     The generic object, the generic model of the theory which has exactly one
-    sort and no function symbols, relations symbols or axioms, appears to be
+    sort and no function symbols, relations symbols or axioms, is
     slightly indecisive: On the one hand, up to a double negation, it is a
@@
 
-\begin{frame}{A different kind of example}
+\begin{frame}{An application in commutative algebra}
   Let~$A$ be a reduced ring ($x^n = 0 \Rightarrow x = 0$).
@@
 
-  Consequence: short and conceptual proof of Grothendieck's generic freeness
-  lemma
+  This observation unlocks a short and conceptual proof of Grothendieck's
+  generic freeness lemma.
+
+  \jnote{1}{A \emph{local localisation} of a ring~$A$ is a
+  localisation~$A[S^{-1}]$ which is a local ring. The classifying topos of
+  local localisations of~$A$ is none other than the topos of sheaves over the
+  spectrum of~$A$, as studied in algebraic geometry; and the generic model is
+  also known as the structure sheaf~$\O_{\operatorname{Spec}(A)}$.
+
+  Assuming the Boolean prime ideal theorem, the geometric sequents
+  validated by~$A^\sim$ are easy to describe: They are precisely those which
+  are validated by all the stalks~$A_\ppp$ of~$A$.
+
+  But~$A^\sim$ enjoys further unique properties which are not shared by the
+  stalks of~$A$, other localisations of~$A$, quotients of~$A$ or indeed any
+  reasonable construction. Three of these are displayed on the slide.
+  (A ring is \emph{anonymously Noetherian} iff each of its ideals is
+  \emph{not not} finitely generated. Textbook proofs of Hilbert's basis theorem
+  are constructively valid for this Noetherian condition.)
+
+  The object~$A^\sim$ strikes a fine balance: On the one hand, it is still
+  close to~$A$, so that information learned about~$A^\sim$ teaches us
+  about~$A$; on the other hand, it enjoys unique properties rendering it
+  simpler than~$A$.
+
+  This balance allows for a simple and conceptually satisfying proof of
+  \emph{Gro\-then\-dieck's generic freeness lemma}, an important theorem in
+  algebraic geometry. Details can be found in
+  \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-padova2018.pdf}{this
+  set of slides}.}
 \end{frame}
-
-%\section{Affine schemes}
-%
-%\begin{frame}{Affine schemes}
-%  \jnote{1}{
-%    A ring is \emph{local} iff every invertible sum contains an invertible
-%    summand, that is if~$1 \neq 0$ and if~$x + y$ invertible implies~$x$
-%    invertible or~$y$ invertible. Assuming the axiom of choice, this elementary
-%    definition is equivalent to the textbook definition of a local ring (a ring
-%    with exactly one maximal ideal). A ring homomorphism is \emph{local} iff it
-%    reflects invertibility.
-%
-%    The notion of a \emph{filter} in a ring~$A$ is a direct axiomatisation of
-%    what classically would be the complement of a prime ideal. The filter
-%    axioms are: $1 \in F$, $(xy \in F) \Leftrightarrow ((x \in F) \wedge (y \in
-%    F))$, $\neg(0 \in F)$, $(x+y \in F) \Rightarrow ((x \in F) \vee (y \in F))$.
-%
-%    A \emph{free local ring} over~$A$ is a homomorphism into a local
-%    ring~$A'$ such that any homomorphism into a local ring~$R$ factors uniquely
-%    over~$A'$ via a local homomorphism.
-%
-%    For any particular local ring~$A \xrightarrow{f} R$, the
-%    localisation~$A[S^{-1}]$ with~$S = f^{-1}[R^\times]$ is a local ring which
-%    fits into the displayed diagram. However, in general there is no single
-%    choice of~$S$ which would work for any local ring~$R$. Indeed, classically
-%    one can show that a free local ring over~$A$ exists if and only if~$A$
-%    contains exactly one prime ideal, in which case~$A$ itself is the free
-%    local ring.
-%  }
-%
-%  \jnote{2}{
-%    If we want a free local ring to exist for any ring~$A$, we have to broaden
-%    our notion of existence and embrace rings which live in toposes other
-%    than~$\Set$. There is a notion of a homomorphism between rings living in
-%    arbitrary toposes, and using this notion one can verify:
-%
-%    The free local ring over~$A$ can be built in the classifying topos of
-%    filters of~$A$, as the localisation (in that topos) of~$A$ at the generic
-%    filter. (More precisely, as the localisation of the mirror image of~$A$ in that
-%    topos, that is the constant sheaf~$\underline{A}$.)
-%
-%    The classifying topos of filters of~$A$ coincides with the topos of sheaves
-%    over what's called the \emph{spectrum of~$A$} in algebraic geometry, and
-%    under this equivalence the free local ring coincides with the structure
-%    sheaf of spectrum. In fact the classifying topos serves as a good
-%    constructive substitute for the classical spectrum construction,
-%    enjoying the expected universal property even if the axiom of choice is not
-%    available, which is why it is simply denoted~``$\Spec(A)$'' on the slide.
-%
-%    (The classifying topos of prime ideals of~$A$ is also interesting; it
-%    coincides with the topos of sheaves over the spectrum of~$A$ equipped with
-%    the constructible topology.)
-%  }
-%
-%  \jnote{3}{
-%    Assuming the Boolean prime ideal theorem, the geometric sequents
-%    validated by~$A^\sim$ are easy to describe: They are precisely those which
-%    are validated by all the stalks~$A_\ppp$ of~$A$.
-%
-%    But~$A^\sim$ enjoys further unique properties which are not shared by the
-%    stalks of~$A$, other localisations of~$A$, quotients of~$A$ or indeed any
-%    reasonable construction. Three of these are displayed on the slide.
-%    (A ring is \emph{anonymously Noetherian} iff each of its ideals is
-%    \emph{not not} finitely generated. Textbook proofs of Hilbert's basis theorem
-%    are constructively acceptable for this Noetherian condition.)
-%
-%    The object~$A^\sim$ strikes a fine balance: On the one hand, it is still
-%    close to~$A$, so that information learned about~$A^\sim$ teaches us
-%    about~$A$; on the other hand, it enjoys unique properties rendering it
-%    simpler than~$A$.
-%
-%    This balance allows for a simple and conceptually satisfying proof of
-%    \emph{Gro\-then\-dieck's generic freeness lemma}, an important theorem in
-%    algebraic geometry. Details can be found in
-%    \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-padova2018.pdf}{this
-%    set of slides}.
-%  }
-%
-%  Let~$A$ be a ring. Is there a \hil{free local ring}~$A \to A'$ over~$A$?
-%  \begin{columns}[t]
-%    \begin{column}{0.4\textwidth}
-%      $\xymatrix{
-%        A \ar[rd] \ar[rrr]^f &&& {\substack{\phantom{\text{local}}\\\text{\normalsize$R$}\\\text{local}}} \\
-%        & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!35]{\text{local}}[rru]
-%      }$
-%    \end{column}
-%
-%    \begin{column}{0.50\textwidth}
-%      \small\justifying
-%      For a fixed ring~$R$, the localisation $A' \defeq A[S^{-1}]$ with $S \defeq
-%      f^{-1}[R^\times]$ would do the job. ($S$ is a \emph{filter}.)
-%      \medskip
-%
-%      Hence we need the \hil{generic filter}.
-%    \end{column}
-%  \end{columns}
-%
-%  \pause
-%
-%  \justifying
-%  The free local ring over~$A$ is~$A^\sim \defeq \ul{A}[F^{-1}]$,
-%  where~$F$ is the generic filter, living in~$\Spec(A)$, the classifying topos
-%  of filters of~$A$.
-%
-%  \pause
-%
-%  If~$A$ is reduced ($x^n = 0 \Rightarrow x = 0$):
-%
-%  \vspace*{-0.8em}
-%  \setbeamercolor{block body}{bg=red!30}
-%  \setbeamercolor{structure}{fg=purple}
-%  \begin{varblock}{\textwidth}{}
-%    $A^\sim$ is a \hil{field}: $\forall x\?A^\sim\_ (\neg(\exists y\?A^\sim\_
-%    xy = 1) \Rightarrow x = 0)$.
-%
-%    $A^\sim$ has \hil{$\boldsymbol{\neg\neg}$-stable equality}:
-%    $\forall x,y\?A^\sim\_ \neg\neg(x = y) \Rightarrow x = y$.
-%
-%    \mbox{$A^\sim$ is \hil{anonymously Noetherian}.}\\[-1.2em]
-%  \end{varblock}
-%\end{frame}
 
@@
     algebraic geometry is a framework for algebraic geometry in which schemes
-    can be modelled by plain old sets, morphisms of schemes by plain old maps
+    are modelled by plain old sets, morphisms of schemes by plain old maps
     between those sets, group schemes by plain old groups, and so on. Unlike
@@
     models for coherent logic} Gavin Wraith's question in the negative. (As
-    Thierry remarked during the talk, this is even if one takes care to phrase
+    Thierry remarked at 6WFTop, this is even if one takes care to phrase
     the question in a way to exclude the trivial counterexamples given by
@@
 
-    The asterisks in \emph{geometric$^\star$ sequent}
+    The asterisk in \emph{geometric$^\star$ sequent}
     indicate that any infinities used to index
-    disjunctions have to be come from the base topos. This restriction is an important
+    disjunctions have to come from the base topos. This restriction is an important
     subtlety, though not vital to this talk. If~$\TT$ is a coherent
-    theory, then for coherent sequents there is no difference between
-    provability in coherent logic, provability in geometric logic and
-    provability$^\star$.
+    theory, then we can restrict to coherent sequents and thereby vanquish the
+    subtlety.
   }
@@
     The algebraic Nullstellensatz states that, in some cases, algebraic truths
-    are witnessed by explicit \emph{algebraic certificates} -- syntactical
+    are witnessed by explicit \emph{algebraic certificates} -- syntactic
     objects giving a priori reasons for why a given truth is to be expected.
@@
     arbitrary truths of the generic model, subject only to the condition that
-    they can be expressed as a geometric sequent, and algebraic certificates
+    they can be expressed as geometric sequents, and algebraic certificates
     are replaced by \emph{logical certificates}: proofs.
@@
     conservative$^\star$~$\ul{\TT}/U_\TT$-model''.
+
+    To be more precise, the following is true: \\
+    \mbox{For any geometric sequent~$\sigma$, if~$\Set[\TT] \models
+    \speak{$\sigma$ holds for~$U_\TT$}$, then $\TT$ proves~$\sigma$.}
+
+    The following is false in general: \\
+    \mbox{$\Set[\TT] \models \speak{For any geometric$^\star$ sequent~$\sigma$,
+    if~$\sigma$ holds for~$U_\TT$, then $\ul{\TT}$ proves~$\sigma$}$.}
+
+    The following is true: \\
+    \mbox{$\Set[\TT] \models \speak{For any geometric$^\star$ sequent~$\sigma$,
+    if~$\sigma$ holds for~$U_\TT$, then $\ul{\TT}/U_\TT$ proves~$\sigma$}$.}
   }
@@
 
-    Constructing internally to~$\Set[\TT]$, where~$\ul{\TT}/U_\TT$ is just an
+    Constructing internally to~$\Set[\TT]$, where~$\ul{\TT}/U_\TT$ looks just like an
     ordinary geometric theory, the classifying topos of that theory, and then
@@
   %  \{\heartsuit\}$.
-    Let~$x,y \? U_\TT$. Assume that~$\neg(x = y)$.
+    Let~$x,y \? U_\TT$. We want to verify~$\neg\neg(x = y)$. Hence assume that~$\neg(x = y)$.
     By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves~$\sigma \defeq (e_x = e_y \vdash \bot)$.
-    But this is false in the~$\ul{\TT}/U_\TT$-model~$M \defeq U_\TT/(x \sim y)$.
+    Thus~$\sigma$ holds in any~$\ul{\TT}/U_\TT$-model.
+    But~$\sigma$ is false in the~$\ul{\TT}/U_\TT$-model~$M \defeq U_\TT/(x \sim
+    y)$. Hence~$\bot$.
   }
@@
   \jnote{1}{
-    This slide gives two examples how to use the Nullstellensatz to deduce
+    This slide gives an example how to use the Nullstellensatz to deduce
     properties of the generic model. A couple of remarks are in order.
@@
 
-    Often when using the Nullstellensatz, we go from an (assumed) truth
+    Often when using the Nullstellensatz, we go from a (perhaps assumed) truth
     of~$U_\TT$ via provability to another model~$M$
-    of~$\ul{\TT}/U_\TT$. That is, we use provability$^\star$ as a (one-way)
-    \emph{bridge}:
-    \[ \xymatrix{
-      & \text{$\ul{\TT}/U_\TT$ proves $\sigma$} \\
-      \text{$\sigma$ holds for $U_\TT$} \ar@/^1.1pc/@{<=>}[ur] &&
-      \text{$\sigma$ holds for $M$} \ar@/_1.1pc/@{<=}[ul]
-    } \]
+    of~$\ul{\TT}/U_\TT$. That is, we use provability as a (one-way)
+    \emph{bridge}.
   }
@@
     Theorem~1 states that the source provided by the Nullstellensatz is
-    exhaustive. The notion of provability$^{\star\star}$ is a strengthening of
-    the notion of provability$^\star$, which in turn is a strengthening of the
-    ordinary notion of provability in geometric logic. We are using it here
-    because while the notion of provability$^\star$ can be expressed in the
-    internal language of a topos, it cannot be expressed in intuitionistic
-    logic.
+    exhaustive. It is possible, but not entirely obvious, how to formulate the
+    Nullstellensatz as an (infinitary) first-order formula.
 
@@
 
-    Theorems~3 and~4 generalise the Nullstellensatz to the
-    higher-order setting. The map~$\mathrm{ev}$ maps (the equivalence class
-    of) a~$\ul{\TT}/U_\TT$-provably$^\star$ geometric$^\star$ formula~$\theta$
-    in one free variable to the subset~$\{ x \? U_\TT \,|\, \theta(x) \}$.
+    Theorems~3 indicates that the Nullstellensatz can be generalised to the
+    higher-order setting.
 
@@
 
-    Theorem~3 is a relativisation of Olivia Caramello's completeness theorem,
+    Theorem~3 rests on a relativisation of Olivia Caramello's completeness theorem,
     Theorem~2.4(ii) in her paper
@@
 
-  \jnote{2}{
+  \jnote{3}{
     Plans for the future include:
@@
       geometry. Most of the toposes in geometric use are actually uncharted
-      territory from a logical point of view.
+      territory from a logical point of view. The talk by Matthias Hutzler will
+      present new results in that direction.
     \end{itemize}

### 2019-07-10 slides-edinburgh2019.tex
Final version of the Edinburgh slides

@@
   The photos on the slide are of \emph{Arthur's Seat}, a beautiful extinct
-  volcano in Edinburgh. It gives an excellent view of the city and is very much
-  recommended for a hiking tour.
+  volcano in Edinburgh where parts of these slides have been created. It gives
+  an excellent view of the city and is very much recommended for a hiking tour.
   }%
@@
   any given topos. For instance, a ring in the usual sense is a model of the
-  theory of rings in the topos~$\Set$. The structure sheaf of a scheme~$X$ is a
-  model in the topos~$\Sh(X)$ of set-valued sheaves on~$X$.
+  theory of rings in the topos~$\Set$. The structure sheaf~$\O_X$ of a scheme~$X$ is a
+  model in the topos~$\Sh(X)$ of set-valued sheaves over~$X$.
 
@@
       \begin{column}{0.45\textwidth}
-        \includegraphics[width=\textwidth,align=t]{mike-shulman-stack-semantics}
+        \href{https://arxiv.org/abs/1004.3802}{\includegraphics[width=\textwidth,align=t]{mike-shulman-stack-semantics}}
         \vspace*{1.5em}
@@
   This observation unlocks a short and conceptual proof of Grothendieck's
-  generic freeness lemma.
+  generic freeness lemma in algebraic geometry.
 

### 2019-07-31 paper-flabby-objects.tex
Fixed mathematical typo (thanks to the anonymous referee)
typo

@@
 subterminals~$\P_{\leq 1}(X)$ are each injective.
-\item The forgetful functor from modules to sets possesses a left exact left
+\item The forgetful functor from modules to sets possesses a monomorphism-preserving left
 adjoint. More explicitly, if~$i : A \to B$ is an injective map between sets and
@@
 \begin{lemma}\label{lemma:notions-of-internal-injectivity}
-Let~$Y$ be a ringed locale over a locale~$X$. Let~$I$ be a sheaf of modules over~$Y$.
+Let~$Y$ be a ringed locale over a locale~$X$. Let~$J$ be a sheaf of modules over~$Y$.
 Assuming Zorn's lemma in the metatheory, the following statements are equivalent:
 \begin{enumerate}
-\item $I$ is an injective sheaf of modules.
-\item From the point of view of~$\Sh(Y)$, $I$ is an injective module.
-\item From the point of view of~$\Sh(X)$, $I$ is an injective module from the
+\item $J$ is an injective sheaf of modules.
+\item From the point of view of~$\Sh(Y)$, $J$ is an injective module.
+\item From the point of view of~$\Sh(X)$, $J$ is an injective module from the
 point of view of~$\Sh(I(Y))$.
-\item From the point of view of~$\Sh(X)$, $I$ is an injective sheaf of modules
+\item From the point of view of~$\Sh(X)$, $J$ is an injective sheaf of modules
 on~$I(Y)$.
@@
 Theorem~\ref{thm:higher-direct-images-as-internal-sheaf-cohomology} gives, as a
-simple application, a logical explanation that higher direct images along
+simple application, a logical explanation of the basic fact that higher direct images along
 the identity~$\mathrm{id} : X \to X$ vanish: From the internal point of view

### 2019-07-31 paper-flabby-objects.tex
A couple of small fixes and additions (thanks to the anonymous referee)

@@
 properties~(1),~(3a),~(3b) and~(3c) require Zorn's lemma to construct maximal
-extensions of given sections. The standard proof of property~(4b) requires the
+extensions of given sections.\footnote{We are careful to distinguish between
+the axiom of choice and Zorn's lemma. The former implies the latter, but the
+converse requires the law of excluded middle.}
+The standard proof of property~(4b) requires the
 law of excluded middle, to ensure that the Godement construction actually
@@
 
-Instead of referencing arbitrary stages~$A \in \E$, one can also just reference
+Instead of referring to arbitrary stages~$A \in \E$, one can also just refer to
 the generic stage: Let~$\P_{\leq1}(X)$ denote the \emph{object of subterminals}
@@
 
-\begin{proof}For the ``only if'' direction, let~$I$ be an object
+\begin{proof}Let~$I$ be an object
 which is externally injective. Then~$I$ satisfies Condition~(2) in
@@
 
-For the ``if'' direction, let~$I$ be an internally
+For the converse direction, let~$I$ be an internally
 injective object. Let~$i : A \to B$ be a monomorphism in~$\E$ and let~$f :

### 2019-09-19 slides-herrsching2019.tex
Beginning of the slides for Proof and Computation 2019

--- 2019-07-10 slides-edinburgh2019.tex
@@
 
-\usepackage{pgfpages}
-\setbeameroption{show notes on second screen}
-\setbeamertemplate{note page}[plain]
-\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}
+%\usepackage{pgfpages}
+%\setbeameroption{show notes on second screen}
+%\setbeamertemplate{note page}[plain]
+%\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}
+\newcommand{\jnote}[2]{}
 
@@
 
-\title{A general Nullstellensatz for generalised spaces}
+\title{Generalized spaces for constructive algebra}
 \author{Ingo Blechschmidt}
-\date{April 11th, 2019}
+\date{September 2019}
 
@@
 
-\newcommand{\mynav}[3]{%
-  foo
-}
-
 \newcommand{\insertframeextra}{}
@@
 
+\tikzstyle{topos} = [draw=mypurple, very thick, rectangle, rounded corners, inner sep=5pt, inner ysep=10pt]
+\tikzstyle{title} = [fill=mypurple, text=white]
+
+\input{images/primes.tex}
+%\renewcommand{\sieve}[2]{SIEVE}
+%\renewcommand{\fakesieve}[2]{SIEVE}
+
+\newcommand{\drawbox}[4]{
+  \node[topos, #4] [fit = #3] (#1) {};
+  \node[title] at (#1.north) {#2};
+}
+
+\newcommand{\muchstuff}{
+  \includegraphics[height=3em]{filmat}
+  \scalebox{0.5}{\sieve{14}{2}}
+}
+
+\newcommand{\muchstuffplaceholder}{
+  \includegraphics[height=3em]{filmat-placeholder}
+  \scalebox{0.5}{\fakesieve{14}{2}}
+}
+
+\newcommand{\fewstuff}{
+  \includegraphics[height=3em]{filmat}
+  \scalebox{0.5}{\sieve{7}{2}}
+}
+
+\newcommand{\threeblobs}{
+  \colorbox{mypurple}{\ \ }\quad
+  \colorbox{mypurple}{\ \ }\quad
+  \colorbox{mypurple}{\ \ }
+}
+
 %\setbeamertemplate{headline}{\mynav{gray}{gray}{gray}}
@@
 
-  \bigskip
-  \[ \xymatrix{
-    & \text{$\ul{\TT}/U_\TT$ proves $\sigma$} \\
-    \text{$\sigma$ holds for $U_\TT$} \ar@/^1.1pc/@{<=>}[ur] &&
-    \text{$\sigma$ holds for $M$} \ar@/_1.1pc/@{<=}[ul]
-  } \]
-  \bigskip
-
   \begin{tikzpicture}
     \def\R{8pt}
-    \node (title) {\hil{A general Nullstellensatz for generalised spaces}};
+    \node (title) {\hil{\,Generalized spaces for constructive algebra\,}};
     \begin{pgfonlayer}{background}
@@
   \end{tikzpicture}
+  \bigskip
+
+  \begin{columns}
+    \small
+    \begin{column}{0.35\textwidth}
+      \centering
+      \bignumber{1}
+
+      \phantom{x}\hil{Locales}
+      \medskip
+
+      \includegraphics[height=4.5em]{phantoms.png}
+    \end{column}
+    \hspace*{0.5em}
+    \begin{column}{0.28\textwidth}
+      \centering
+      \bignumber{2}
+
+      \hil{Sheaf models}
+      \medskip
+
+      \begin{tikzpicture}
+        \node (overview) {
+          \scalebox{0.35}{\sieve{4}{3}}
+        };
+        \def\R{8pt}
+        \begin{pgfonlayer}{background}
+        \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple]
+          ($(overview.south west) + (\R, 0)$) arc(270:180:\R) --
+          ($(overview.north west) + (0, -\R)$) arc(180:90:\R) --
+          ($(overview.north east) + (-\R, 0)$) arc(90:0:\R) --
+          ($(overview.south east) + (0, \R)$) arc(0:-90:\R) --
+          cycle;
+        \end{pgfonlayer}
+      \end{tikzpicture}
+    \end{column}
+    \begin{column}{0.4\textwidth}
+      \centering
+      \bignumber{3}
+
+      \hil{Constructive algebra}
+      \medskip
+
+      \includegraphics[height=4.5em]{calabi-yau}
+    \end{column}
+  \end{columns}
+  \bigskip
 
@@
 
-  Category Theory 2019 in Edinburgh \\
-  July 9th, 2019
+  Proof and Computation \\
+  September 20th to 26th, 2019
   \bigskip
@@
 
-  \jnote{1}{\textbf{This talk in a nutshell:}
-
-  There is a certain ring, the \emph{generic ring}, which is special in that
-  it is \emph{conservative}: It has exactly those properties which are shared
-  by any ring whatsoever. One could argue that it is this ring which we're
-  implicitly referring to when we utter the phrase ``Let~$R$ be a ring.''.
-
-  A surprising observation is that the generic ring is a field. This is not a
-  contradiction because conservativity only pertains to properties which can be
-  put in a certain syntactic form, dubbed \emph{geometric sequents}, and the
-  field condition is not of that form.
-
-  As a consequence, when verifying a geometric sequent for all rings, we may
-  freely assume the field condition: Our proof will apply to the generic ring,
-  because this ring happens to be a field, and its conclusion will spread to
-  any ring whatsoever thanks to its syntactic form.
-
-  An analogous story can be told for the generic group, the generic vector
-  space and quite generally for the generic model of any \emph{geometric
-  theory}. It is an interesting task to discover properties of such
-  generic models. This talk presents a source of such properties, the
-  \emph{topos-theoretic Nullstellensatz}, which is
-  universal in the sense that any such property is a logical consequence of the
-  Nullstellensatz.}
 \end{frame}}
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\hspace*{-1.7cm}\includegraphics[height=\paperheight,trim=0 0 5.1cm 0,clip]{arthurs-seat-4}\end{minipage}}
-\begin{frame}[plain]%
-  \vspace*{\paperheight}\vspace*{-2.35cm}\hspace*{-1cm}%
-  \includegraphics[width=\paperwidth/4]{arthurs-seat-7}%
-  \includegraphics[width=\paperwidth/4]{arthurs-seat-0}%
-  \includegraphics[width=\paperwidth/4]{arthurs-seat-5}%
-  \includegraphics[width=\paperwidth/4]{arthurs-seat-9}%
-  \jnote{1}{This talk owes a special debt to Alexander Oldenziel, who
-  repeatedly suggested that the topic of this talk should be studied.
-
-  The photos on the slide are of \emph{Arthur's Seat}, a beautiful extinct
-  volcano in Edinburgh. It gives an excellent view of the city and is very much
-  recommended for a hiking tour.
-  }%
-\end{frame}}
-\addtocounter{framenumber}{-1}
-
-\section{The generic model}
-
-\subsection{The mystery of nongeometric sequents}
-
-\begin{frame}{The mystery of nongeometric sequents}
-  \jnote{1}{A \emph{geometric sequent} is a syntactic expression of the
-  form~$(\varphi \seq{x_1\?X_1,\ldots,x_n\?X_n} \psi)$,
-  where~$x_1\?X_1,\ldots,x_n\?X_n$ is a list of variable declarations,
-  the~$X_i$ ranging over the available sorts, and~$\varphi$ and~$\psi$ are
-  \emph{geometric formulas}. Such a sequent is read as ``in the
-  context~$x_1\?X_1,\ldots,x_n\?X_n$, $\varphi$ entails $\psi$''.
-  Geometric formulas are built from atomic propositions (equality and the
-  available relation symbols) using the connectives~$\top$, $\bot$, $\wedge$, $\bigvee$
-  (set-indexed disjunction) and~$\exists$. Geometric formulas may not
-  contain~$\neg$, $\Rightarrow$, $\forall$.
-
-  There is a notion of a \emph{model} of a geometric theory in
-  any given topos. For instance, a ring in the usual sense is a model of the
-  theory of rings in the topos~$\Set$. The structure sheaf of a scheme~$X$ is a
-  model in the topos~$\Sh(X)$ of set-valued sheaves on~$X$.
-
-  With \emph{topos} we mean Grothendieck topos, and as metatheory we use a
-  constructive but impredicative flavour of English (which could be formalised
-  by what is supported by the internal language of elementary toposes with an NNO).
-  However the Nullstellensatz presented later makes no use of the
-  subobject classifier, hence the results can likely be generalised to hold in
-  a predicative metatheory or to hold for arithmetic universes.}
-
-  \jnote{2-}{Among all models in any topos, the \emph{universal} or
-  \emph{generic} one is special. It enjoys the universal property that any
-  model in any topos can be obtained from it by pullback along an essentially
-  unique geometric morphism. It is intriguing from a logical point of view
-  because it has exactly those properties which are shared by any model in any
-  topos.
-
-  One could argue, with a certain amount of success, that the generic model of
-  the theory of rings is what a mathematician implicitly refers to when she
-  utters the phrase ``Let~$R$ be a ring''. This point of view is fundamental to
-  the slogan \emph{continuity is geometricity}, as expounded for instance in
-  \fixedhref{http://www.cs.bham.ac.uk/~sjv/GeoAspects.pdf}{Continuity and
-  geometric logic} by Steve Vickers.}
-
-  \jnote{3-}{Crucially, the conservativity statement only pertains to
-  properties which can be put as geometric sequents. Generic models may have
-  additional nongeometric properties. Because conservativity does not apply to
-  them, they are not shared by all models in all toposes -- but any consequences
-  which can be put as geometric sequents are.
-
-  For instance, if we want to verify a geometric sequent for all local rings,
-  we may freely use the displayed field axiom.
-  Hence one reason why these nongeometric sequents are interesting is because
-  they provide us with new reduction strategies (proofs by ``without loss of
-  generality'').}
-
-  Let~$\TT$ be a \explain{geometric theory}{1-}{red!20}{-1cm}{0cm}{0cm}{\begin{minipage}{4.3cm}
-    \small
-    sorts,
-    function symbols,
-    relation symbols,
-    geometric sequents as axioms
-  \end{minipage}}, for instance the
-  \explain{theory of rings}{1-}{blue!20}{-1cm}{0cm}{0cm}{\begin{minipage}{4.3cm}
-    \small
-    \begin{tabbing}
-      fun. symb.: \= \kill
-      sorts: \> $R$ \\
-      fun. symb.: \> $0$, $1$, $-$, $+$, $\cdot$ \\
-      % rel. symb.: \> \emph{none} \\
-      axioms: \> $(\top \vdash_{x,y:R} x y = y x)$, \ldots
-    \end{tabbing}
-  \end{minipage}}.
-
-  \vspace*{1.8cm}
-
-  \begin{center}
-    \begin{tikzpicture}[ultra thick, node distance=7mm]
-      \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40] (a) {$\ZZ$};
-      \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40, right=of a] (b) {$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$};
-      \node[regular polygon, regular polygon sides=5, draw=orange!80,
-      fill=orange!20, right=of b, rounded corners=1pt, inner sep=0cm] (c) {$\O_X$};
-      \node[star, rounded corners=1pt, star points=10, inner sep=0cm, draw=purple!80, fill=purple!20, right=of c] {$U_\TT$};
-    \end{tikzpicture}
-  \end{center}
-
-  \vspace*{-0.5cm}
-  \pause
-
-  \justifying
-  \textbf{Theorem.} There is a \hil{generic model}~$U_\TT$. It is
-  \hil{conservative} in that
-  for any \explainstub{geometric sequent}{yellow!70}~$\sigma$ the following notions coincide:
-  \vspace*{-1.2em}
-  \begin{enumerate}
-    \item The sequent~$\sigma$ holds for~$U_\TT$.
-    \item The sequent~$\sigma$ holds for any~$\TT$-model in any
-    \explainstub{topos}{purple!20}.
-    \item The sequent~$\sigma$ is provable modulo~$\TT$.
-  \end{enumerate}
-
-  \pause
-
-  \justifying
-  %\textbf{Main observation.} The generic model can validate nongeometric
-  %sequents which are not provable modulo~$\TT$ and which do not hold for
-  %arbitrary models in arbitrary toposes.
-
-  \textbf{Observation (Kock).} The generic local ring is a field:
-  \vspace*{-0.2cm}
-  \[ (x = 0 \Rightarrow \bot) \vdash_{x:R} (\exists y\?R\_ xy = 1) \]
-\end{frame}
-
-\subsection{Construction of the generic model}
-
-\begin{frame}{Construction of the generic model}
-  The generic model is \hil{not} the same as \ldots
-  \begin{itemize}
-    \item the \hil{initial model} (think~$\ZZ$) or
-    \item the \hil{free model on one generator} (think~$\ZZ[X]$).
-  \end{itemize}
-  Set-based models are \hil{too inflexible}.
-
-  \textbf{Definition.} The \hil{syntactic site}~$\C_\TT$ has \ldots
-  \begin{enumerate}
-    \item objects: \explainstub{formulas-in-contexts}{red!20} $\{ x_1\?X_1, \ldots, x_n\?X_n\_ \varphi \}$
-    \item morphisms: eqv. classes of \explainstub{provably functional
-    formulas}{red!30}
-    \item coverings: \explainstub{provably jointly surjective
-    families}{purple!30}
-  \end{enumerate}
-  The topos of sheaves over~$\C_\TT$ is the \hil{classifying
-  topos}~$\Set[\TT]$. The generic model interprets a sort~$X$ by the sheaf~$\yon\{x\?X\_\top\}$.
-
-  \jnote{1}{In case the theory~$\TT$ is a Horn theory (for instance if it is
-  an equational theory), the \emph{term algebra} (the set of terms in the
-  empty context modulo provable equality) is a model of~$\TT$. While the
-  term algebra does enjoy some nice categorical properties, it is in general
-  \emph{not} the generic model.
-
-  For instance, if~$\TT$ is the theory of rings, then the initial model
-  is~$\ZZ$. This model validates some geometric sequents which are not
-  validated by all rings, for instance~$(x^2 = 0 \seq{x:R} x = 0)$ or $(1 = 0
-  \vdash \bot)$.
-
-  In general, the generic model cannot be realised as a set-based model (with
-  a set for each sort, a map for each function symbol and so on). Sets are too
-  constant for this purpose; the flexibility of sheaves (``variable sets'') is
-  required: The generic model lives in the topos of set-valued
-  sheaves over~$\C_\TT$, called the \emph{classifying topos} of~$\TT$.}
-
-  \jnote{2}{The notation in the definition of the objects of~$\C_\TT$ is meant
-  to be suggestive, but should not be taken literally. The objects are
-  syntactic expressions consisting of a context and a geometric formula in that
-  context.
-
-  A morphism~$A = \{\vec x\_ \varphi\} \to \{\vec y\_ \psi\} = B$ in~$\C_\TT$ is the
-  equivalence class (modulo provable equivalence) of a geometric
-  formula~$\theta$ such that~$\TT$ proves
-  \begin{enumerate}
-    \item ``$\theta$ is a relation on~$A \times B$'': $(\theta \seq{\vec x,
-    \vec y} \varphi \wedge \psi)$
-    \item ``$\theta$ is total'': $(\varphi \seq{\vec x} \exists \vec y\_
-    \theta)$
-    \item ``$\theta$ is single-valued'': $(\theta \wedge \theta[\vec y'/\vec y]
-    \seq{\vec x,\vec y,\vec y'} \vec y = \vec y')$
-  \end{enumerate}
-
-  \vspace*{-0.6em}
-  A family~$(\{ \vec x_i\_ \varphi_i \} \xrightarrow{[\theta_i]} \{ \vec y\_
-  \psi \})_i$ is a covering iff~$\TT$ proves~$(\psi \seq{\vec y} \bigvee_i
-  \exists \vec x_i\_ \theta_i)$.
-
-  The slides experiment with using the symbol~``$\yon$'' for the Yoneda
-  embedding~$\C_\TT \to \Set[\TT]$, as in \fixedhref{http://www.math.jhu.edu/~eriehl/elements.pdf}{Elements
-  of~$(\infty,1)$-category theory} by Emily Riehl and Dominic Verity.}
-
-  \jnote{3}{The special case that the generic model of a theory~$\TT$ can be realised as
-  a model in~$\Set$ occurs iff~$\TT$ is Morita-equivalent to the
-  empty theory, that is, iff~$\TT$ has exactly one model in any topos.
-
-  The special case that there exists at least some conservative~$\TT$-model
-  in~$\Set$ occurs iff~$\TT$ has a conservative geometric expansion to a theory
-  which is Morita-equivalent to the empty theory.}
-\end{frame}
-
-\subsection{Going internal}
-
-\begin{frame}{Working internally to toposes}
-  \justifying
-  Let~$\C$ be a site. We recursively define
-  \[
-    U \models \varphi \quad
-    \text{(``$\varphi$ holds on~$U$'')}
-  \]
-  for objects~$U \in \C$ and \squiggly{formulas}~$\varphi$.
-  Write~``$\Sh(\C) \models \varphi$'' for~$U \models \varphi$ for all~$U$.
-  \only<1>{\footnotesize
-  \[ \renewcommand{\arraystretch}{1.08}\begin{array}{@{}l@{\ \ }c@{\ \ }l@{}}
-  U \models \top &\Ll& \text{true} \\
-  U \models \bot &\Ll& \hcancel{\text{false}}{0pt}{3pt}{0pt}{-2pt}\ \text{the
-  empty family is a covering of~$U$} \\
-  U \models s = t \? F &\Ll& s|_U = t|_U \in F(U) \\
-  U \models \varphi \wedge \psi &\Ll&
-  \text{$U \models \varphi$ and $U \models \psi$} \\
-  U \models \varphi \vee \psi &\Ll&
-  \hcancel{\text{$U \models \varphi$ or $U \models
-  \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $(U_i \to U)_i$} \\
-  && \quad\quad\text{such that for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
-  U \models \varphi \Rightarrow \psi &\Ll&
-  \text{for all~$V \to U$: }
-  \text{$V \models \varphi$ implies $V \models \psi$} \\
-  U \models \forall s \? F\_ \varphi(s) &\Ll&
-  \text{for all $V \to U$ and sections~$s_0 \in F(V)$: $V \models \varphi(s_0)$} \\
-  % U \models \forall F\_ \varphi(F) &\Ll&
-  % \text{for all $V \to U$ and sheaves~$F_0$ over~$V$: $V \models \varphi(F_0)$} \\
-  U \models \exists s \? F\_ \varphi(s) &\Ll&
-  \hcancel{\text{there exists $s_0 \in F(U)$ such that $U \models \varphi(s_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
-  &&
-  \text{there exists a covering $(U_i \to U)_i$ such that for all~$i$:} \\
-  && \quad\quad \text{there exists~$s_0 \in F(U_i)$ such that $U_i \models \varphi(s_0)$}
-  %U \models \exists F\_ \varphi(F) &\Ll&
-  %\hcancel{\text{there exists a sheaf $F_0$ on $U$ such that $U \models
-  %\varphi(F_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
-  %&&
-  %\text{there exists a covering $(U_i \to U)_i$ such that for all~$i$:} \\
-  %&& \quad\quad \text{there exists a sheaf~$F_0$ on $U_i$ such that $U_i \models \varphi(F_0)$}
-  \end{array} \]}
-
-  \only<2>{
-    \bigskip
-    \begin{columns}[t]
-      \begin{column}{0.45\textwidth}\justifying
-        \textbf{Example.}
-        $\Set \models \varphi$ if and only if $\varphi$ is true.
-        \bigskip
-
-        \textbf{Theorem.}
-        If~$\Sh(\C) \models \varphi$ and if $\varphi$ intuitionistically
-        entails~$\psi$, then also~$\Sh(\C) \models \psi$.
-      \end{column}
-
-      \begin{column}{0.45\textwidth}
-        \includegraphics[width=\textwidth,align=t]{mike-shulman-stack-semantics}
-        \vspace*{1.5em}
-      \end{column}
-    \end{columns}
-  }
-
-  \jnote{1}{
-    The internal language of a (Grothendieck or elementary) topos~$\E$ is a
-    device which allows us to speak and reason about the objects and morphisms
-    of~$\E$ in a naive element-based language close to the usual formal
-    mathematical language. Using this language, objects of~$\E$ look like plain
-    old sets [or types]; morphisms look like plain old maps between those sets;
-    epimorphisms look like surjections; group objects look like groups; and so on.
-
-    In particular, we can use the internal language to define what it means for a
-    given~$\TT$-structure in~$\E$ to be a model -- namely iff it looks like a
-    model from the internal point of view.
-
-    The internal language can be implemented by the \emph{Kripke--Joyal semantics},
-    a translation procedure which converts formulas of the internal language
-    into external statements about the objects and morphisms of~$\E$. The slide
-    displays some of the translation rules in the case that~$\E$ is a
-    Grothendieck topos.
-  }
-
-  \jnote{2}{
-    We can actually \emph{do mathematics} internally to a topos because the
-    Kripke--Joyal semantics is sound with respect to intuitionistic logic:
-    If~$\E \models \varphi$ and if~$\varphi$ intuitionistically entails a
-    further formula~$\psi$, then~$\E \models \psi$.
-
-    The Kripke--Joyal semantics can be extended to interpret unbounded
-    quantification (``for all sets'' as opposed to ``for all elements of the
-    particular set~$X$'') and dependent types. The former are for instance
-    required to express universal properties (``for all groups'', ``for all
-    rings''), and the latter are all over the place, even if their use might
-    not be particularly highlighted.
-
-    With these extensions, we can
-    import all of everyday constructive impredicative mathematics into the
-    internal world of a topos.
-
-    Some illustrations of working with the internal language can be found in these
-    sets of slides:
-    \begin{itemize}
-      \item
-      \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-leipzig2018.pdf}{Slides
-      for Jürgen Jost's group seminar at the MPI Leipzig}
-      \item
-      \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-como2018.pdf}{Slides
-      for \emph{Toposes in Como}}
-      (recording
-      \fixedhref{https://www.youtube.com/watch?v=SuELfC5P_R8}{available})
-    \end{itemize}
-    A longer exposition, with pointers to the literature, can be found in Section~2 of
-    \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these notes}.
-  }
-\end{frame}
-
-\section[Nongeometric properties]{A selection of nongeometric properties}
-
-\begin{frame}{A selection of nongeometric properties}
-  The \hil{generic object} validates:
-  \begin{enumerate}
-    \item $\forall x,y\?U_\TT\_ \neg\neg(x = y)$.
-    \item $\forall x_1,\ldots,x_n\?U_\TT\_ \neg \forall y\?U_\TT\_
-    \bigvee_{i=1}^n y = x_i.$
-    \item $(U_\TT)^{U_\TT} \cong 1 \amalg U_\TT$.
-  \end{enumerate}
-
-  The \hil{generic ring} validates:
-  \begin{enumerate}
-    \item $\forall x\?U_\TT\_ \neg\neg(x = 0)$.
-    \item $\forall x\?U_\TT\_ (x = 0 \Rightarrow 1 = 0) \Rightarrow (\exists
-    y\?U_\TT\_ xy = 1)$.
-  \end{enumerate}
-
-  The \hil{generic local ring} validates:
-  \begin{enumerate}
-    \item $\neg \forall x\?U_\TT\_ \neg\neg(x = 0)$.
-    \item $\forall a_0,\ldots,a_{n-1}\?U_\TT\_ \neg\neg \exists x\?U_\TT\_
-      x^n + a_{n-1}x^{n-1} + \cdots + a_0 x^0 = 0$.
-    \item \justifying Let~$\Delta = \{ \varepsilon \? U_\TT \,|\, \varepsilon^2 = 0 \}$.
-    For any map~$f : \Delta \to U_\TT$, there are unique elements~$a,b \?
-    U_\TT$ s.\,th.~$f(\varepsilon) = a + b\varepsilon$ for all~$\varepsilon
-    \? \Delta$.
-%   \item For any finitely presented~$U_\TT$-algebra~$A$,
-%   the canonical map~$A \to (U_\TT)^{\Hom_{U_\TT}(A,U_\TT)}$
-%   is an isomorphism.
-  \end{enumerate}
-
-  \jnote{1}{
-    The generic object, the generic model of the theory which has exactly one
-    sort and no function symbols, relations symbols or axioms, is
-    slightly indecisive: On the one hand, up to a double negation, it is a
-    subsingleton; on the other hand, it is infinite. This observation is due to
-    Carsten Butz and Peter Johnstone
-    (\fixedhref{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.16.4102&rep=rep1&type=pdf}{Classifying toposes for
-    first-order theories}).
-
-    The generic ring, the generic model of the theory of rings, is similarly
-    indecisive. It is infinite in the following sense:
-    \[ \forall x_1,\ldots,x_n \? U_\TT\_
-      \bigl(\forall y \? U_\TT\_ \bigvee_{i=1}^n (y = x_i)\bigr) \Rightarrow
-      1 = 0. \]
-
-    The theory of local rings is the quotient theory of the theory of rings
-    obtained by adding the axioms
-    \[ (1 = 0 \vdash \bot) \ \text{and}\ ((\exists z\_
-    (x+y)z = 1) \seq{x,y} (\exists z\_ xz = 1) \vee (\exists z\_ yz = 1)). \]
-    (Assuming the axiom of choice, a ring is local in this sense
-    iff it local in the usual sense (has exactly one maximal ideal).)
-
-    The first displayed property of the generic local ring illustrates
-    that nongeometric sequents need not be inherited by quotient theories.
-  }
-
-  \jnote{2}{
-    All of the displayed properties give rise to reduction techniques:
-    If we want to verify a geometric sequent for all rings, it suffices to
-    verify it for the generic ring; but the generic ring has additional
-    nongeometric properties not shared by every ring, such as the two displayed
-    ones. (This was for instance used
-    \fixedhref{https://www.sciencedirect.com/science/article/pii/0022404976900025}{by
-    Anders Kock} and
-    \fixedhref{https://link.springer.com/chapter/10.1007\%2FBFb0061835}{by
-    Gonzalo Reyes}.)
-
-    However, we face some challenges when pursuing these reduction techniques, including
-    the following:
-    \begin{enumerate}\justifying
-      \item It is not easy to determine interesting and useful properties of
-      the generic model.
-      \item The set of validated nongeometric sequents changes slightly
-      unpredictably when passing to quotient theories. For instance, when
-      proving that a geometric sequent holds for all rings, we may assume that
-      any element is \emph{not not} zero. But we may not assume this
-      simplification if we want to verify a geometric sequent for all local
-      rings (and if we want to exploit the given locality in the proof).
-      \item There is only so much we want to state and prove in full generality
-      for all rings, all local rings, all modules, and so on. We are often much
-      more interested in properties of particular mathematical objects.
-    \end{enumerate}
-  }
-
-  \jnote{3}{
-    The firstly-mentioned problem on the previous page is alleviated by the
-    Nullstellensatz presented in this talk, which gives a systematic and
-    universal source of nongeometric sequents validated by the generic model.
-    However, manual work is still required to reduce this set of sequents to a
-    smaller, manageable one consisting of memorable properties while hopefully
-    still preserving universality.
-
-    To counter the third problem, it's prudent to consider geometric theories
-    which depend on a given mathematical object of interest. For instance,
-    given a ring~$A$, we can consider the theory of prime ideals of~$A$, of
-    complemented prime ideals, of filters, and so on. The classifying toposes
-    of these theories are of independent interest -- in fact they are sheaf
-    toposes over certain important spaces in algebraic geometry -- and
-    nongeometric sequents validated by their generic models bundle nontrivial
-    information about~$A$. More details are on the following slide.
-  }
-\end{frame}
-
-\begin{frame}{An application in commutative algebra}
-  Let~$A$ be a reduced ring ($x^n = 0 \Rightarrow x = 0$).
-
-  The \hil{generic local localisation}~$A^\sim$ of~$A$ validates:
-
-  \vspace*{-1.8em}
-  \setbeamercolor{block body}{bg=red!30}
-  \setbeamercolor{structure}{fg=purple}
-  \begin{varblock}{\textwidth}{}
-    $A^\sim$ is a \hil{field}: $\forall x\?A^\sim\_ (\neg(\exists y\?A^\sim\_
-    xy = 1) \Rightarrow x = 0)$.
-
-    $A^\sim$ has \hil{$\boldsymbol{\neg\neg}$-stable equality}:
-    $\forall x,y\?A^\sim\_ \neg\neg(x = y) \Rightarrow x = y$.
-
-    \mbox{$A^\sim$ is \hil{anonymously Noetherian}.}\\[-1.2em]
-  \end{varblock}
-
-  This observation unlocks a short and conceptual proof of Grothendieck's
-  generic freeness lemma.
-
-  \jnote{1}{A \emph{local localisation} of a ring~$A$ is a
-  localisation~$A[S^{-1}]$ which is a local ring. The classifying topos of
-  local localisations of~$A$ is none other than the topos of sheaves over the
-  spectrum of~$A$, as studied in algebraic geometry; and the generic model is
-  also known as the structure sheaf~$\O_{\operatorname{Spec}(A)}$.
-
-  Assuming the Boolean prime ideal theorem, the geometric sequents
-  validated by~$A^\sim$ are easy to describe: They are precisely those which
-  are validated by all the stalks~$A_\ppp$ of~$A$.
-
-  But~$A^\sim$ enjoys further unique properties which are not shared by the
-  stalks of~$A$, other localisations of~$A$, quotients of~$A$ or indeed any
-  reasonable construction. Three of these are displayed on the slide.
-  (A ring is \emph{anonymously Noetherian} iff each of its ideals is
-  \emph{not not} finitely generated. Textbook proofs of Hilbert's basis theorem
-  are constructively valid for this Noetherian condition.)
-
-  The object~$A^\sim$ strikes a fine balance: On the one hand, it is still
-  close to~$A$, so that information learned about~$A^\sim$ teaches us
-  about~$A$; on the other hand, it enjoys unique properties rendering it
-  simpler than~$A$.
-
-  This balance allows for a simple and conceptually satisfying proof of
-  \emph{Gro\-then\-dieck's generic freeness lemma}, an important theorem in
-  algebraic geometry. Details can be found in
-  \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-padova2018.pdf}{this
-  set of slides}.}
-\end{frame}
-
-\section[A systematic source]{A systematic source of nongeometricity?}
-
-\begin{frame}{A systematic source of nongeometricity?}
-  \jnote{1-2}{
-    Mimicking the synthetic approach to differential geometry, synthetic
-    algebraic geometry is a framework for algebraic geometry in which schemes
-    are modelled by plain old sets, morphisms of schemes by plain old maps
-    between those sets, group schemes by plain old groups, and so on. Unlike
-    its close cousin, it is far less developed; some first steps are outlined
-    in Sections~19 and~20 of
-    \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these notes}.
-
-    Synthetic algebraic geometry is carried out internally to the \emph{big
-    Zariski topos} of a given base scheme; in the special case that the base
-    scheme is the terminal scheme~$\Spec(\ZZ)$, this topos is just the
-    classifying topos of local rings. The relevant generic model living in the
-    big Zariski topos is denoted~``$\affl$'' because it coincides with the
-    functor of points of the affine line.
-  }
-
-  \jnote{2}{
-    Marc Bezem, Ulrik Buchholtz and Thierry Coquand answered in their 2017 paper
-    \fixedhref{https://arxiv.org/abs/1712.07743}{Syntactic forcing
-    models for coherent logic} Gavin Wraith's question in the negative. (As
-    Thierry remarked at 6WFTop, this is even if one takes care to phrase
-    the question in a way to exclude the trivial counterexamples given by
-    instances of the law of excluded middle in the language of~$\TT$.) If the
-    answer had been positive, this would have given a neat, if somewhat hard to
-    use in practice, characterisation of the formulas validated by the generic model.
-  }
-
-  \jnote{3}{
-    The topos-theoretic \emph{Nullstellensatz}, to be presented on the next
-    slides, answers the displayed three questions in the affirmative.
-
-    Briefly, the Nullstellensatz is a certain statement in the language of a
-    given geometric theory~$\TT$ which is
-    \begin{itemize}\justifying
-      \item validated by the generic~$\TT$-model,
-      \item typically not validated by other~$\TT$-models, and
-      \item such that any statement validated by the generic~$\TT$-model can be
-      deduced, in intuitionistic logic, from the axioms of~$\TT$ and the
-      Nullstellensatz.
-    \end{itemize}
-
-    We believe that this characterisation is as explicit as it can get, but
-    would be delighted to be surprised by a future improvement. We stumbled on
-    it by playing with the synthetic quasicoherence statement, not least thanks
-    to encouragement by Alexander Oldenziel. However the route from that
-    statement to the Nullstellensatz is not quite direct; it turns out that
-    synthetic quasicoherence is a corollary of a specialisation of a
-    higher-order version of the Nullstellensatz to Horn theories.
-  }
-
-  \justifying
-  \textbf{Empirical fact.}
-  In \hil{synthetic algebraic geometry}, every known property of~$\affl$ followed from
-  its \hil{synthetic quasicoherence}:
-
-  \begin{quote}
-    \emph{\mbox{For any finitely presented~$\affl$-algebra~$A$, the ``double dual'' map}
-    \[ A \longrightarrow (\affl)^{\Hom_\affl(A,\affl)},\ s \longmapsto (x
-    \mapsto x(s)) \]
-    is an isomorphism of~$\affl$-algebras.}
-  \end{quote}
-
-  \vspace*{1.0em}
-  \begin{enumerate}
-    \item Does a general metatheorem explain this observation?
-    \item Is there a systematic source in any classifying topos?
-    \item Is there even an exhaustive source?
-  \end{enumerate}
-  \vspace*{-3.3em}
-
-  \visible<2>{\begin{tikzpicture}[overlay]
-    \draw[fill=white, draw=white, opacity=0.95] (-1,1.4) rectangle (\paperwidth,5.4);
-    \node[anchor=south west,inner sep=0] (image) at (0,0.9) {\vbox{
-      \only<2>{
-	\centering
-	\includegraphics[width=0.9\textwidth]{wraith-some-recent-developments-in-topos-theory} \\
-	\scriptsize
-        Gavin Wraith. \emph{Some recent developments in topos theory.} \\ In:
-        Proc.\@ of the ICM (Helsinki, 1978).
-      }
-    }};
-  \end{tikzpicture}}%
-
-  \[ \xymatrix{
-    \text{$\TT$ proves~$\alpha$\;} \ar@/_1pc/@{=>}[r] &
-    \text{\;$\alpha$ holds for~$U_\TT$\;} \ar@/_1pc/@{=>}[r]
-    \ar@/_1pc/@{=>}@[red][l]
-    |{\SelectTips{cm}{}\textbf{\scalebox{3}{\textcolor{red}{\!\texttimes\!}}}}|{} &
-    \text{\;$\alpha$ is~$\TT$-redundant}
-    \ar@/_1pc/@{=>}@[red][l]
-    |{\SelectTips{cm}{}\textbf{\scalebox{3}{\textcolor{red}{\!\texttimes\!}}}}|{}
-  } \]
-\end{frame}
-
-\section{A topos-theoretic Nullstellensatz}
-
-\newcommand{\nullstellensatz}{%
-  \justifying
-  \textbf{Theorem.} Internally to~$\Set[\TT]$:
-
-  \vspace*{-0.3em}
-  \begin{quote}\emph{\mbox{For any \explainstub{geometric$^\star$ sequent~$\sigma$}{red!20} over
-  the \explainstub{signature of~$\ul{\TT}/U_\TT$}{blue!20},}
-  if~\explainstub{$\sigma$ holds for~$U_\TT$}{yellow!70}, then~\explainstub{$\ul{\TT}/U_\TT$
-  proves~$\sigma$}{yellow!70}.}
-  \end{quote}
-  \vspace*{-0.6em}
-}
-
-\begin{frame}{A topos-theoretic Nullstellensatz}
-  \jnote{1-2}{
-    By~$\ul{\TT}$ we mean the geometric theory internal to~$\Set[\TT]$ obtained
-    by pulling back the set of sorts of~$\TT$, the set of function symbols and so on
-    along the unique geometric morphism~$\Set[\TT] \to \Set$. For instance,
-    if~$\TT$ is the theory of rings, then from the internal point of view
-    of~$\Set[\TT]$, the theory~$\ul{\TT}$ will again be the theory of rings.
-
-    The theory~$\ul{\TT}/U_\TT$ will be defined on the next slide. It is a certain
-    geometric theory internal to~$\Set[\TT]$.
-
-    The asterisk in \emph{geometric$^\star$ sequent}
-    indicate that any infinities used to index
-    disjunctions have to come from the base topos. This restriction is an important
-    subtlety, though not vital to this talk. If~$\TT$ is a coherent
-    theory, then we can restrict to coherent sequents and thereby vanquish the
-    subtlety.
-  }
-
-  \jnote{2}{
-    The algebraic Nullstellensatz states that, in some cases, algebraic truths
-    are witnessed by explicit \emph{algebraic certificates} -- syntactic
-    objects giving a priori reasons for why a given truth is to be expected.
-
-    In the topos-theoretic Nullstellensatz, algebraic truths are replaced by
-    arbitrary truths of the generic model, subject only to the condition that
-    they can be expressed as geometric sequents, and algebraic certificates
-    are replaced by \emph{logical certificates}: proofs.
-  }
-
-  \jnote{3}{
-    While, as stated on slide 1/10, the generic model~$U_\TT$ is a
-    conservative~$\TT$-model, the classifying topos~$\Set[\TT]$ does not
-    believe this fact. That is, the statement~``$U_\TT$ is a
-    conservative~$\ul{\TT}$-model'' is not true internally to~$\Set[\TT]$.
-    What is true is the modified statement~``$U_\TT$ is a
-    conservative$^\star$~$\ul{\TT}/U_\TT$-model''.
-
-    To be more precise, the following is true: \\
-    \mbox{For any geometric sequent~$\sigma$, if~$\Set[\TT] \models
-    \speak{$\sigma$ holds for~$U_\TT$}$, then $\TT$ proves~$\sigma$.}
-
-    The following is false in general: \\
-    \mbox{$\Set[\TT] \models \speak{For any geometric$^\star$ sequent~$\sigma$,
-    if~$\sigma$ holds for~$U_\TT$, then $\ul{\TT}$ proves~$\sigma$}$.}
-
-    The following is true: \\
-    \mbox{$\Set[\TT] \models \speak{For any geometric$^\star$ sequent~$\sigma$,
-    if~$\sigma$ holds for~$U_\TT$, then $\ul{\TT}/U_\TT$ proves~$\sigma$}$.}
-  }
-
-  \nullstellensatz
-
-  \pause
-
-  \textbf{The algebraic Nullstellensatz.}
-  Let~$A$ be a ring.
-  Let~$f,g \in A[X]$ be polynomials.
-  Then, subject to some conditions:
-  \[
-    \underbrace{\bigl(\forall x \in A\_ (f(x) = 0 \Rightarrow g(x) =
-    0)\bigr)}_{\text{algebraic truth}} \Longrightarrow
-    \underbrace{\bigl(\exists h \in A[X]\_ g = hf\bigr)}_{\text{algebraic certificate}}
-  \]
-  \vspace*{-0.9em}
-
-  \pause
-
-  \textbf{A naive version.}
-  ``Internally to~$\Set[\TT]$, for any geometric sequent~$\sigma$ over
-  the signature of~$\ul{\TT}$,
-  if~$\sigma$ holds for~$U_\TT$, then~$\ul{\TT}$
-  proves~$\sigma$.'' \textbf{False,} for instance with the theory of rings we
-  have
-  \begin{align*}
-    \Set[\TT] &\models \neg(\speak{$\ul{\TT}$ proves $(\top \vdash 1 + 1 =
-    0)$}) \\
-    \text{but}\quad \Set[\TT] & \not\models \neg(1 + 1 = 0).
-  \end{align*}
-\end{frame}
-
-\begin{frame}{A varying internal theory}
-  \nullstellensatz
-
-  \explainstub{\textbf{Definition.}}{blue!20} The theory~$\ul{\TT}/U_\TT$ is the
-  internal geometric theory of~\hil{$U_\TT$-algebras}, the theory which arises from~$\ul{\TT}$ by adding:
-  \begin{enumerate}
-    \justifying\small
-    \item for each element~$x : U_\TT$ a constant symbol~$e_x$,
-    \item for each function symbol~$f$ and~$n$-tuple~$(x_1,\ldots,x_n) : (U_\TT)^n$ the axiom
-    $(\top \vdash f(e_{x_1},\ldots,e_{x_n}) = e_{f(x_1,\ldots,x_n)})$,
-    \item for each relation symbol~$R$ and~$n$-tuple~$(x_1,\ldots,x_n) : (U_\TT)^n$ such
-    that~$R(x_1,\ldots,x_n)$ the axiom~$(\top \vdash R(e_{x_1},\ldots,e_{x_n}))$.
-  \end{enumerate}
-
-  \textbf{Remark.} Externalising the internal classifying
-  topos~$\Set[\TT][\ul{\TT}/U_\TT]$ yields the classifying topos
-  of~$\TT$-homomorphisms.
-
-  \jnote{1}{
-    Just as locales internal to a topos~$\E$ can be externalised to yield
-    localic geometric morphisms into~$\E$, internal Grothendieck toposes can be
-    externalised to yield bounded geometric morphisms. Since the composition of
-    bounded geometric morphisms is bounded, the externalisation of a
-    Grothendieck topos internally to a Grothendieck topos is itself a
-    Grothendieck topos, hence the classifying topos of some geometric theory.
-
-    Constructing internally to~$\Set[\TT]$, where~$\ul{\TT}/U_\TT$ looks just like an
-    ordinary geometric theory, the classifying topos of that theory, and then
-    externalising the resulting Grothendieck topos results in the classifying
-    topos of~$\TT$-homomorphisms. There are two canonical geometric morphisms
-    from this topos to~$\Set[\TT]$, the morphism computing the domain and the
-    morphism computing the codomain, and the morphism obtained by the
-    externalisation procedure is the former.
-  }
-\end{frame}
-
-\begin{frame}{Revisiting the test cases}
-  \nullstellensatz
-  \vspace*{-2em}
-  \[ \xymatrix{
-    & \text{$\ul{\TT}/U_\TT$ proves $\sigma$} \\
-    \text{$\sigma$ holds for $U_\TT$} \ar@/^1.1pc/@{<=>}[ur] &&
-    \text{$\sigma$ holds for $M$} \ar@/_1.1pc/@{<=}[ul]
-  } \]
-
-  \only<1>{
-    \textbf{In the object classifier.}
-  %  Assume that~$U_\TT$ is empty, that is $\forall x \? U_\TT\_ (\top \Rightarrow
-  %  \bot)$. By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves~$(\top \seq{x} \bot)$.
-  %  But this is false in the~$\ul{\TT}/U_\TT$-model~$U_\TT \amalg
-  %  \{\heartsuit\}$.
-    Let~$x,y \? U_\TT$. We want to verify~$\neg\neg(x = y)$. Hence assume that~$\neg(x = y)$.
-    By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves~$\sigma \defeq (e_x = e_y \vdash \bot)$.
-    Thus~$\sigma$ holds in any~$\ul{\TT}/U_\TT$-model.
-    But~$\sigma$ is false in the~$\ul{\TT}/U_\TT$-model~$M \defeq U_\TT/(x \sim
-    y)$. Hence~$\bot$.
-  }
-
-  \only<2->{
-    \textbf{In the ring classifier.}
-    Let~$f,g \? U_\TT[X]$ such that any zero of~$f$ is a zero of~$g$.
-    By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves this fact.
-    Hence it holds in the~$\ul{\TT}/U_\TT$-model~$U_\TT[X]/(f)$. In this
-    model~$f$ has the zero~$[X]$. Hence also~$g([X]) = 0$
-    in~$U_\TT[X]/(f)$, that is~$g = hf$ for some~$h \? U_\TT[X]$.
-  }
-
-  \jnote{1}{
-    This slide gives an example how to use the Nullstellensatz to deduce
-    properties of the generic model. A couple of remarks are in order.
-
-    The Nullstellensatz is trivial for sequents~$\sigma$ of the form~$(\top
-    \vdash \psi)$. The Nullstellensatz is only interesting in case
-    that~$\sigma$ has a nontrivial antecedent or is set in a nonempty context.
-
-    Since the converse direction in the Nullstellensatz also holds
-    (because~$U_\TT$ is a~$\ul{\TT}/U_\TT$-model), the
-    statements~$\speak{$\sigma$ holds for~$U_\TT$}$
-    and~$\speak{$\ul{\TT}/U_\TT$ proves $\sigma$}$ are equivalent. This
-    equivalence is intriguing from a logical point of view, since the former
-    statement is a geometric implication while the latter can be put as a
-    geometric formula. (Up to a subtle issue indicated on the next slide.)
-
-    To apply the Nullstellensatz, no description of a site defining~$\Set[\TT]$
-    is required.
-
-    Often when using the Nullstellensatz, we go from a (perhaps assumed) truth
-    of~$U_\TT$ via provability to another model~$M$
-    of~$\ul{\TT}/U_\TT$. That is, we use provability as a (one-way)
-    \emph{bridge}.
-  }
-\end{frame}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{7.95cm}\hfill\mbox{\includegraphics[width=0.13\paperwidth]{hen}\;\;}\end{minipage}}
-\begin{frame}{Exhaustion and extensions}
-  \justifying
-  \textbf{Theorem 1.} A first-order formula holds for~$U_\TT$ iff it is
-  intuitionistically provable from the axioms of~$\TT$ and
-  the scheme
-  \begin{equation}\tag{Nullstellensatz} \speak{$\sigma$ holds} \implies \speak{$\ul{\TT}/U_\TT$
-  proves~$\sigma$}. \end{equation}
-  \vspace*{-1.5em}
-
-  \textbf{Theorem 2.}
-  Let~$\TT'$ be a quotient theory
-  of~$\TT$. Assume that~$U_\TT$ is contained in the subtopos~$\Set[\TT']$. Then internally
-  to~$\Set[\TT']$:
-  \begin{quote}
-    \emph{A geometric$^\star$ sequent~$\sigma$ with Horn consequent holds for~$U_{\TT'}$
-    iff~$\ul{\TT}/U_\TT$ proves~$\sigma$.}\par
-  \end{quote}
-
-  \textbf{Theorem 3.} A higher-order formula holds for~$U_\TT$ iff it is
-  provable in intuitionistic higher-order logic from the axioms of~$\TT$ and
-  the higher-order Nullstellensatz scheme.
-
-  \jnote{1}{
-    Theorem~1 states that the source provided by the Nullstellensatz is
-    exhaustive. It is possible, but not entirely obvious, how to formulate the
-    Nullstellensatz as an (infinitary) first-order formula.
-
-    Theorem~2 provides a useful variant of the Nullstellensatz. Its
-    assumptions are for instance satisfied if~$\TT$ is the theory of rings
-    and~$\TT'$ is the theory of local rings. When applicable, it can be used to
-    avoid doubly-internal toposes. It also explains, for instance, why in the
-    formulation of synthetic quasicoherence no local rings appear even though
-    the relevant topos is the classifying topos of local rings.
-
-    Theorems~3 indicates that the Nullstellensatz can be generalised to the
-    higher-order setting.
-
-    Written details on all of this are
-    \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-qcoh.pdf}{slowly
-    emerging}.
-  }
-
-  \jnote{2}{
-    The Nullstellensatz is related to several precursors. A corollary of the
-    Nullstellensatz is that, over the first-order theory validated by~$U_\TT$, any
-    first-order formula is in fact logically equivalent to a geometric formula.
-    This corollary has already been observed by Carsten Butz and Peter
-    Johnstone in their paper
-    \fixedhref{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.16.4102&rep=rep1&type=pdf}{Classifying
-    toposes for first-order theories} (Lemma~4.2 there). At that point, a
-    characterisation of the first-order formulas in the general case, of the
-    form as in Theorem~1, was still missing.
-
-    Theorem~3 rests on a relativisation of Olivia Caramello's completeness theorem,
-    Theorem~2.4(ii) in her paper
-    \fixedhref{https://arxiv.org/abs/0906.3061}{Universal models and
-    definability}. The passage from the external to the internal phrasing
-    requires going from~$\TT$ to~$\ul{\TT}/U_\TT$.
-  }
-
-  \jnote{3}{
-    Plans for the future include:
-    \begin{itemize}\justifying
-      \item Developing an Agda library for dealing with the internal language
-      of toposes and related kinds of categories, employing Agda's
-      meta-programming facilities; with such a library at hand,
-      formalising the Nullstellensatz in Agda.
-      \item Exploring Nullstellensatz-style results for arithmetic universes.
-      \item Applying the Nullstellensatz in constructive algebra and algebraic
-      geometry, along the lines of generic freeness and synthetic algebraic
-      geometry. Most of the toposes in geometric use are actually uncharted
-      territory from a logical point of view. The talk by Matthias Hutzler will
-      present new results in that direction.
-    \end{itemize}
-  }
-\end{frame}}
-
-% Analogy with the harmonic oscillator in the study of physics
-
-% * The generic model
-%   * The mystery of nongeometric sequents
-%   * Construction of the generic model
-% * Kripke--Joyal semantics; focus on existence of extensions
-% * Examples for theories and nongeometric sequents with a focus on the wlog
-%   aspect
-% * Why we are interested in the generic models themselves
-%   * Universal localisation
-%   * Grothendieck's generic freeness lemma
-% * Synthetic quasicoherence
-% * Conservativity from the internal point as guiding question
-%   * Naive formulation and why it doesn't work
-%   * Recall of the algebraic Nullstellensatz
-% * The new Nullstellensatz
-%   * What Set[T][T/U_T] classifies
-%   * Kock again
-%   * Subtleties regarding infinites
-%   * Improvement in the Horn case
-% * Universality of the source provided by the Nullstellensatz
-
 \end{document}
-
-% Are there size issues when KJ'ing internally?
-% Comparison to the approach by Vickers?
-% When does a theory prove the Nullstellensatz?
-% Look up what Kock and Reyes used the wlog for.
-
-% An Tafel: geometric formula, local ring
-% Erwähnen: Marc Bezem, Ulrik Buchholtz, Thierry Coquand
 
@@
 % the topoi, like evanescent phantoms, to surround the imaginary “points”.
-
-% Sei T eine Theorie, für die ein konservatives Modell in Set existiert.
-% Dann haben wir eine Surjektion Set --> Set[T]. Also existiert eine
-% geometric expansion von T, welche Morita-äquivalent zur leeren Theorie ist.

### 2019-09-19 slides-herrsching2019.tex
More work on the slides

@@
 
+% \begin{document}
+
+\section{Locales}
+
+\begin{frame}{Locales}
+  \begin{hilblock}Locales are a kind of space in which \hil{opens} instead of
+  \hil{points} are fundamental.\end{hilblock}
+
+  \begin{columns}[t]
+    \begin{column}[t]{0.4\textwidth}
+      \textbf{Def.:} A \hil{topological space} consists of a \hil{set of
+      points}~$X$ together with a set~$\O(X)$ of point sets which are deemed
+      \hil{open} such that unions of XXX
+    \end{column}
+
+    \begin{column}[t]{0.6\textwidth}
+      \begin{tikzpicture}[sibling distance=8em, level distance=2.5em,
+        every node/.style = {shape=rectangle, rounded corners,
+          draw, align=center}]
+        \node {euclidean spaces} [grow=up]
+          child { node {metric spaces}
+            child { node {sober topological spaces}
+              child { node {locales}
+                child { node {toposes}
+                  child { node {$\infty$-toposes} } } }
+              child { node {topological spaces} } } };
+      \end{tikzpicture}
+    \end{column}
+  \end{columns}
+\end{frame}
+
 \end{document}

### 2019-09-20 slides-herrsching2019.tex
Slides for Lecture I finished

--- 2019-09-19 slides-herrsching2019.tex
@@
 \newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates
+\newcommand{\brak}[1]{{\llbracket{#1}\rrbracket}}
 
@@
       \def\R{8pt}
-      \node (title) {\hil{\large\insertframetitle}};
+      \node (title) {\hil{\large\,\!\insertframetitle}};
       \begin{pgfonlayer}{background}
@@
 
-      \hil{Sheaf models}
+      \phantom{x}\hil{Sheaf models}
       \medskip
 
-      \begin{tikzpicture}
+      \phantom{x}\!\!\begin{tikzpicture}
         \node (overview) {
@@
 
-  Proof and Computation \\
-  September 20th to 26th, 2019
+  Autumn schoon on Proof and Computation \\
+  September 20th to 26th, 2019 \\
+  Herrsching
   \bigskip
@@
 
+% \begin{document}
+
+\section{Locales}
+
+\begin{frame}
+  \centering\bigskip
+  \includegraphics[height=9.5em]{phantoms}
+
+  \large Lecture I:
+  \hil{Locales}
+
+  \normalsize
+
+  \begin{hilblock}Locales are a kind of space in which \hil{opens} instead of
+  \hil{points} are fundamental.\end{hilblock}
+
+\end{frame}
+
+\begin{frame}{Locales in a nutshell}
+  \justifying
+  \textbf{Definition:} A \hil{topological space}~$X$ consists of a \hil{set of
+  points} together with a set~$\O(X)$ of point sets which are deemed
+  \hil{open} such that unions and finite intersections of open sets are open.
+  \bigskip
+
+  \centering\small
+  \begin{tikzpicture}[sibling distance=8em, level distance=2.5em,
+    every node/.style = {shape=rectangle, rounded corners,
+      draw, align=center}, semithick, draw=mypurple]
+    \node {euclidean spaces} [grow=up]
+      child { node {metric spaces}
+        child[dashed] { node [solid] {sober topological spaces}
+          child[solid] { node {locales}
+            child { node {toposes}
+              child { node {$\infty$-toposes} } } }
+          child[solid] { node {topological spaces} } } };
+  \end{tikzpicture}
+\end{frame}
+
+\begin{frame}{Nontrivial spaces without points}
+  The following locales don't have any points and are nontrivial:
+  \begin{enumerate}
+    \item the locale of surjections $\NN \twoheadrightarrow \RR$
+    \item $\QQ \cap (\RR \setminus \QQ)$
+    \item the pairwise intersections in the Banach--Tarski paradox
+    \item Alex Simpson's locale of random binary sequences
+  \end{enumerate}
+
+  \begin{hilblock}
+    Relinquishing points increases flexibility.
+  \end{hilblock}
+
+  \centering
+  \includegraphics[height=5em]{banach-tarski}
+\end{frame}
+
+\begin{frame}{Issues of constructivity}
+  \begin{enumerate}\justifying
+    \item The unit interval~$[0,1]$ is \hil{compact}.
+    \bigskip
+
+    \item \hil{The fundamental theorem of Galois theory:}
+    Let~$L | k$ be a Galois extension. Then there is a bijection between
+    \[ \small\begin{array}{@{}rcl@{}}
+      \begin{minipage}{0.25\textwidth}\centering the intermediate extensions~$L|E|k$\end{minipage}
+      &\text{and}&
+      \begin{minipage}{0.33\textwidth}\centering the closed subgroups
+      of~$\mathrm{Aut}_k(L)$.\end{minipage} \\
+      \\
+      E &\longmapsto& \{ \sigma \in \mathrm{Aut}_k(L) \,|\, \sigma|_E = \operatorname{id} \} \\
+      L^H &\longmapsfrom& H
+    \end{array} \]
+    \medskip
+
+    \item \hil{Gelfand duality:} The categories of
+    \[ \small\begin{array}{@{}rcl@{}}
+      \begin{minipage}{0.38\textwidth}\centering compact Hausdorff spaces\end{minipage}
+      &\text{and}&
+      \left(\begin{minipage}{0.33\textwidth}\centering commutative $C^\star$-algebras with unit\end{minipage}\right)^{\mathrm{op}} \\[1.1em]
+      X &\longmapsto& \{ f : X \to \CC \,|\, \text{$f$ continuous} \}
+    \end{array} \]
+    are equivalent.
+  \end{enumerate}
+\end{frame}
+
+\begin{frame}{Localic basics}
+  \justifying
+  \textbf{Definition:} A \hil{topological space}~$X$ consists of a \hil{set of
+  points} together with a set~$\O(X)$ of point sets which are deemed
+  \hil{open} such that unions and finite intersections of open sets are open.
+
+  \textbf{Observation:} The partial order~$\O(X)$ of open sets has
+  \[ \small\begin{array}{@{}ccc@{}}
+    \text{arbitrary joins}
+    &\text{and}&
+    \text{finite meets}, \\
+    \bigvee && \wedge
+  \end{array} \]
+  and finite meets distribute over arbitrary joins:
+  \[ U \wedge \bigvee_i V_i = \bigvee_i (U \wedge V_i). \]
+  \vspace*{-1.2em}
+  \pause
+
+  \textbf{Definition:} A \hil{frame} is a partial order with arbitrary joins
+  and finite meets such that the distributive law holds.
+
+  \textbf{Definition:} A \hil{locale}~$X$ is given by a \hil{frame~$\O(X)$ of opens}.
+\end{frame}
+
+\begin{frame}{Locales presented by theories}
+  \justifying
+  \textbf{Definition:} A \hil{geometric signature} consists of
+  \begin{enumerate}
+    \item a set of sorts: $X$, $Y$, $Z$, \ldots
+    \item a set of function symbols: $f : X \times Y \to Z$, \ldots
+    \item a set of relation symbols: $R \hookrightarrow X \times Y \times Z$, \ldots
+  \end{enumerate}
+  % A geometric signature is \hil{propositional} iff its set
+  % of sorts is empty.
+  A \hil{geometric theory}~$\TT$ over a geometric
+  signature~$\Sigma$ consists of a set of axioms (geometric sequents
+  over~$\Sigma$).
+
+  \only<1>{
+    \textbf{Examples:}
+    \begin{enumerate}
+      \item the theory of rings
+      \item the theory of surjections~$\NN \twoheadrightarrow \RR$
+      \item the theory of Dedekind cuts
+    \end{enumerate}
+  }
+  \pause
+
+  \textbf{Definition:} A \hil{set-based model}~$M$ of a theory~$\TT$ consists of
+  \begin{enumerate}
+    \item a set~$\brak{X}$ for each sort~$X$,
+    \item a function~$\brak{f} : \brak{X_1} \times \cdots \times \brak{X_n} \to
+    \brak{Y}$ for \\
+    each function symbol~$f : X_1 \times \cdots \times X_n \to
+    Y$, and
+    \item a relation~$\brak{R} \subseteq \brak{X_1} \times \cdots \times
+    \brak{X_n}$ for \\
+    each relation symbol~$R \hookrightarrow X_1 \times \cdots \times X_n$
+  \end{enumerate}
+  such that~$M$ validates the axioms of~$\TT$.
+\end{frame}
+
 \end{document}
+
+    \begin{column}[t]{0.4\textwidth}
+      \justifying
+      \textbf{Contrast with:} A \hil{topological space} consists of a \hil{set of
+      points}~$X$ together with a set~$\O(X)$ of point sets which are deemed
+      \hil{open} such that \ldots
+    \end{column}
 

### 2019-09-21 slides-herrsching2019.tex
Some notes slides

@@
 
-%\usepackage{pgfpages}
-%\setbeameroption{show notes on second screen}
-%\setbeamertemplate{note page}[plain]
-%\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}
-\newcommand{\jnote}[2]{}
+\usepackage{pgfpages}
+\setbeameroption{show notes on second screen}
+\setbeamertemplate{note page}[plain]
+\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}
 
@@
 \newcommand{\brak}[1]{{\llbracket{#1}\rrbracket}}
+\newcommand{\pt}{\mathrm{pt}}
+\newcommand{\Loc}{\mathrm{Loc}}
+\newcommand{\Top}{\mathrm{Top}}
 
@@
 
-  Autumn schoon on Proof and Computation \\
+  Autumn school on Proof and Computation \\
   September 20th to 26th, 2019 \\
@@
 
+  \jnote{1}{
+    These ``probability clouds'', replacing the reassuring material particles of
+    before, remind me strangely of the elusive ``open neighborhoods'' that populate
+    the topoi, like evanescent phantoms, to surround the imaginary ``points''.
+
+    \hfill -- Alexander Grothendieck
+
+    \bigskip
+    \bigskip
+
+    References include:
+    \begin{enumerate}
+      \item Peter Johnstone.
+      \fixedhref{http://pointlesstopology.com/the-point-of-pointless-topology.pdf}{The point of pointless topology}, 1983.
+      \item Peter Johnstone.
+      \fixedhref{http://www.heldermann.de/R&E/RAE18/ctw06.pdf}{The art of
+      pointless thinking: a student's guide to the category of locales}, 1990.
+    \end{enumerate}
+
+    Our metatheory is a constructive (no law of excluded middle, no choice
+    principles of any kind) but impredicative (free use of the powerset axiom)
+    flavor of English. The contents of this course could be formalized in the
+    internal language of toposes or in~\textsc{izf}. The main ideas of this
+    course are far more robust than this particular presentation, in
+    particular, they can also be made sense of in a predicative setting (such
+    as~\textsc{czf} or the internal language of arithmetic universes).
+  }
 \end{frame}
 
-\begin{frame}{Locales in a nutshell}
+\begin{frame}{Locales in context}
   \justifying
-  \textbf{Definition:} A \hil{topological space}~$X$ consists of a \hil{set of
+  \textbf{Definition.} A \hil{topological space}~$X$ consists of a \hil{set of
   points} together with a set~$\O(X)$ of point sets which are deemed
@@
   \includegraphics[height=5em]{banach-tarski}
+
+  \jnote{1}{
+    The locale of surjections $\NN \twoheadrightarrow \RR$ doesn't have any
+    points, but it has lots of interesting opens, such as~$U_{n,x}$, the ``open of those
+    surjections~$f$ such that~$f(n) = x$. If~$x$ is a fixed real number and~$n$
+    ranges over the naturals, these opens cover the full locale.
+
+    The Banach--Tarski paradox is the unintuitive statement that a
+    three-dimensional solid ball of radius~$r$ can be partitioned into six
+    disjoint subsets in such a way that rearranging those subsets yields two
+    solid balls of radius~$r$ each. These subsets are not measurable and
+    require the axiom of choice for their construction.
+
+    The Banach--Tarski paradox can be resolved by adopting the axiom of
+    determinacy instead of the axiom of choice, which entails that all subsets
+    of~$\RR^3$ are measurable, or by passing from the topological space~$\RR^3$
+    to its localic cousin. The localic counterparts of the six pieces will
+    still not have any points in common, but their locale-theoretic pairwise
+    intersection will not be trivial.
+
+    Tom Leinster
+    \fixedhref{https://golem.ph.utexas.edu/category/2011/12/on_the_law_of_large_numbers_su.html}{published
+    an accessible exposition} of the locale of random sequences at the n-category café.
+  }
 \end{frame}
@@
   \end{enumerate}
+
+  \jnote{1}{
+    By \emph{compact}, we mean any open covering has a (Kuratowski-)finite
+    subcovering. The topological space~$[0,1]$ fails to be compact in some
+    flavors of constructive mathematics, but the localic unit interval is
+    always compact.
+
+    In verifying the fundamental theorem of Galois theory, at some point one
+    has to extend a given~$k$-homomorphism~$E \to L$ which is defined on some
+    finite intermediate extension~$E$ to~$L$. One-step extensions to larger
+    intermediate extensions are no problem, but extending to all of~$L$
+    requires some form of choice. By passing from the topological Galois group
+    to its localic group, this issue vanishes. References include the papers
+    \fixedhref{https://core.ac.uk/download/pdf/82660137.pdf}{Galois theory in a topos} and
+    \fixedhref{http://www.numdam.org/article/CTGDC_1981__22_1_61_0.pdf}{Localic groups} by Gavin Wraith.
+    Olivia Caramello's paper
+    \fixedhref{https://arxiv.org/abs/1301.0300}{Topological galois theory} is
+    also relevant.
+
+    A similar issue arises with Gelfand duality. A fully constructive treatment
+    is possible by passing from compact Hausdorff (topological) spaces to
+    completely regular locales. This treatment unlocks the Bohr topos approach
+    to quantum mechanics.
+  }
 \end{frame}
@@
   \justifying
-  \textbf{Definition:} A \hil{topological space}~$X$ consists of a \hil{set of
+  \textbf{Definition.} A \hil{topological space}~$X$ consists of a \hil{set of
   points} together with a set~$\O(X)$ of point sets which are deemed
@@
 
-  \textbf{Observation:} The partial order~$\O(X)$ of open sets has
+  \textbf{Observation.} The partial order~$\O(X)$ of open sets has
   \[ \small\begin{array}{@{}ccc@{}}
@@
 
-  \textbf{Definition:} A \hil{frame} is a partial order with arbitrary joins
+  \textbf{Definition.} A \hil{frame} is a partial order with arbitrary joins
   and finite meets such that the distributive law holds.
 
-  \textbf{Definition:} A \hil{locale}~$X$ is given by a \hil{frame~$\O(X)$ of opens}.
+  \textbf{Definition.} A \hil{locale}~$X$ is given by a \hil{frame~$\O(X)$ of opens}.
+
+  \jnote{2}{
+    \textbf{Example.} Any topological space~$Y$ induces a locale~$L(Y)$ by
+    setting~$\O(L(Y)) \defeq \O(Y)$.
+
+    \textbf{Example.} The \emph{one-point locale}~$\pt$ is the locale induced
+    by the one-point topological space~$\{\heartsuit\}$. Its frame of opens is
+    the powerset of~$\{\heartsuit\}$, also known as the set~$\Omega$ of truth
+    values. Its least element is~$\bot = \emptyset$ and its largest element
+    is~$\top = \{\heartsuit\}$.
+
+    \textbf{Definition.} A \emph{frame homomorphism}~$L \to L'$ is a monotone
+    map~$L \to L'$ which preserves arbitrary joins and finite meets.
+
+    \textbf{Example.} A continuous map~$f : Y \to Y'$ induces a frame
+    homomorphism~$\O(Y') \to \O(Y)$ by mapping~$U \mapsto f^{-1}[U]$.
+
+    \textbf{Definition.} A \emph{locale morphism}~$X \to X'$ is given by a
+    frame homomorphism~$\O(X') \to \O(X)$.
+
+    \textbf{Example.} A continuous map~$Y \to Y'$ induces a locale
+    morphism~$L(Y) \to L(Y')$ in the same direction.
+
+    \textbf{Definition.} A \emph{point} of a locale~$X$ is a locale
+    morphism~$\pt \to X$, or equivalently a completely prime filter of~$\O(X)$.
+    (A frame morphism~$\alpha : \O(X) \to \Omega$ gives rise to the completely prime
+    filter~$F \defeq \{ u \in \O(X) \,|\, \alpha(u) = \top \}$.)
+  }
+
+  \jnote{3}{
+    The set of points of a locale can be made into a topological space, giving
+    rise to a functor~$\pt : \Loc \to \Top$. This functor is right adjoint
+    to~$L$. A locale~$X$ is \emph{spatial} iff the canonical
+    morphism~$L(\pt(X)) \to X$ is an isomorphism, and a topological space~$Y$
+    is \emph{sober} iff the canonical morphism~$Y \to \pt(L(Y))$ is a
+    homeomorphism.
+  }
+\end{frame}
+
+\section{Sheaf models}
+
+\begin{frame}
+  \centering\bigskip
+  \begin{tikzpicture}
+    \node (overview) {
+      \scalebox{1.00}{\sieve{4}{3}}
+    };
+    \def\R{8pt}
+    \begin{pgfonlayer}{background}
+    \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple]
+      ($(overview.south west) + (\R, 0)$) arc(270:180:\R) --
+      ($(overview.north west) + (0, -\R)$) arc(180:90:\R) --
+      ($(overview.north east) + (-\R, 0)$) arc(90:0:\R) --
+      ($(overview.south east) + (0, \R)$) arc(0:-90:\R) --
+      cycle;
+    \end{pgfonlayer}
+  \end{tikzpicture}
+
+  \large Lecture II:
+  \hil{Sheaf models}
+
+  \normalsize
 \end{frame}
@@
   \justifying
-  \textbf{Definition:} A \hil{geometric signature} consists of
+  \textbf{Definition.} A \hil{geometric signature} consists of
   \begin{enumerate}
@@
   \only<1>{
-    \textbf{Examples:}
+    \textbf{Examples.}
     \begin{enumerate}
@@
 
-  \textbf{Definition:} A \hil{set-based model}~$M$ of a theory~$\TT$ consists of
+  \textbf{Definition.} A \hil{set-based model}~$M$ of a theory~$\TT$ consists of
   \begin{enumerate}
@@
 \end{document}
-
-    \begin{column}[t]{0.4\textwidth}
-      \justifying
-      \textbf{Contrast with:} A \hil{topological space} consists of a \hil{set of
-      points}~$X$ together with a set~$\O(X)$ of point sets which are deemed
-      \hil{open} such that \ldots
-    \end{column}
 

### 2019-09-22 slides-herrsching2019.tex
Slides for Lecture II finished

@@
 
+% \begin{document}
+
 \section{Sheaf models}
@@
   \normalsize
+
+  \begin{hilblock}Sheaves allow us to explore mathematical objects from
+  custom-tailored mathematical universes.\phantom{p}\end{hilblock}
 \end{frame}
 
-\begin{frame}{Locales presented by theories}
+\begin{frame}{Frames presented by theories}
   \justifying
-  \textbf{Definition.} A \hil{geometric signature} consists of
+  \textbf{Definition.} A \hil{geometric theory} consists of
   \begin{enumerate}
+    \small
     \item a set of sorts: $X$, $Y$, $Z$, \ldots
@@
     \item a set of relation symbols: $R \hookrightarrow X \times Y \times Z$, \ldots
+    \item a set of axioms: $\varphi \vdash_{x:X, y:Y} \psi$, \ldots
   \end{enumerate}
@@
   % of sorts is empty.
-  A \hil{geometric theory}~$\TT$ over a geometric
-  signature~$\Sigma$ consists of a set of axioms (geometric sequents
-  over~$\Sigma$).
 
-  \only<1>{
+  \only<1->{
     \textbf{Examples.}
-    \begin{enumerate}
-      \item the theory of rings
-      \item the theory of surjections~$\NN \twoheadrightarrow \RR$
-      \item the theory of Dedekind cuts
-    \end{enumerate}
+    The theory of rings,
+    of surjections~\mbox{$\NN \twoheadrightarrow \RR$},
+    of Dedekind cuts,
+    of prime ideals of a given ring, \ldots
   }
-  \pause
 
@@
   \begin{enumerate}
+    \small
     \item a set~$\brak{X}$ for each sort~$X$,
     \item a function~$\brak{f} : \brak{X_1} \times \cdots \times \brak{X_n} \to
-    \brak{Y}$ for \\
-    each function symbol~$f : X_1 \times \cdots \times X_n \to
-    Y$, and
+    \brak{Y}$ \\
+    for each function symbol~$f : X_1 \times \cdots \times X_n \to Y$, and
     \item a relation~$\brak{R} \subseteq \brak{X_1} \times \cdots \times
-    \brak{X_n}$ for \\
-    each relation symbol~$R \hookrightarrow X_1 \times \cdots \times X_n$
+    \brak{X_n}$ \\
+    for each relation symbol~$R \hookrightarrow X_1 \times \cdots \times X_n$
   \end{enumerate}
   such that~$M$ validates the axioms of~$\TT$.
+
+  \jnote{1}{
+    A geometric theory is called \emph{propositional} iff its set of
+    sorts is empty. Hence the only ingredients of a propositional geometric
+    theory are a set of nullary relation symbols (atomic propositions) and a
+    set of axioms.
+
+    Associated to any propositional geometric theory~$\TT$ is its
+    \emph{Lindenbaum algebra}. This is the set of formulas over the signature
+    of the theory modulo provable equivalence. Endowed with the partial order
+    given by
+    \[ [\varphi] \leq [\psi] \quad\vcentcolon\Longleftrightarrow\quad
+      \text{$\TT$ proves $\varphi \vdash \psi$}, \]
+    the Lindenbaum algebra becomes a frame.
+
+    The Lindenbaum algebra of~$\TT$ is the free frame generated by the
+    atomic propositions of~$\TT$ modulo the axioms of~$\TT$. The associated
+    locale~$L(\TT)$ is the \emph{classifying locale of~$\TT$}. Its points are
+    in canonical bijection with the set-based models of~$\TT$.
+  }
+\end{frame}
+
+\begin{frame}{Sheaves}
+  \justifying
+  \textbf{Definition.} A \hil{presheaf}~$F$ on a locale~$X$ is given by
+  \begin{enumerate}
+    \item a set~$F(U)$ for each open~$U$ of~$X$ and
+    \item a map~$(\cdot)|^U_V : F(U) \to F(V)$ for each opens~$V \leq U$
+  \end{enumerate}
+  such that $(\cdot)|^U_U = \operatorname{id}_{F(U)}$ for all~$U$ and
+  $(\cdot)|^V_W \circ (\cdot)|^U_V = (\cdot)|^U_W$ for all~$W \leq V
+  \leq U$.
+  A \hil{compatible family} with respect to a covering~$U = \bigvee_i U_i$
+  is a family~$(s_i)_i$ of sections~$s_i \in F(U_i)$ such that~$s_i|^{U_i}_{U_i
+  \wedge U_j} = s_j|^{U_j}_{U_i \wedge U_j}$. $F$ is a \hil{sheaf} iff for any
+  such family there is a unique section~$s \in F(U)$ such that~$s|^U_{U_i} =
+  s_i$ for all~$i$.
+
+  \textbf{Example.} The \hil{sheaf of continuous functions} on a locale~$X$ is
+  given by~$\C(U) = \Hom(U,\RR)$.
+
+  \textbf{Non-example.} The \hil{presheaf of constant
+  functions},~$\C_{\text{c}}(U) = \text{``the set of constant functions~$U \to
+  \RR$''}$, it usually not a sheaf.
+\end{frame}
+
+\begin{frame}{Sheaf semantics}
+  \justifying\small
+  Let~$X$ be a locale. We recursively define what it means for a
+  \hil{first-order formula over an open~$U$ of~$X$} to be forced:
+  \[ \footnotesize\renewcommand{\arraystretch}{1.08}\begin{array}{@{}l@{\ \ }c@{\ \ }l@{}}
+  U \models \top &\Ll& \text{true} \\
+  U \models \bot &\Ll& \hcancel{\text{false}}{0pt}{3pt}{0pt}{-2pt}\ U = \bot \\
+  U \models s = t &\Ll& \text{$s = t$ when evaluated as elements of~$F(U)$} \\
+  U \models \varphi \wedge \psi &\Ll&
+  \text{$U \models \varphi$ and $U \models \psi$} \\
+  U \models \varphi \vee \psi &\Ll&
+  \hcancel{\text{$U \models \varphi$ or $U \models
+  \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $U = \bigvee_i U_i$} \\
+  && \quad\quad\text{such that for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
+  U \models \varphi \Rightarrow \psi &\Ll&
+  \text{for all~$V \leq U$: }
+  \text{$V \models \varphi$ implies $V \models \psi$} \\
+  U \models \forall s \? F\_ \varphi(s) &\Ll&
+  \text{for all $V \leq U$ and sections~$s_0 \in F(V)$: $V \models \varphi(s_0)$} \\
+  U \models \exists s \? F\_ \varphi(s) &\Ll&
+  \hcancel{\text{there exists $s_0 \in F(U)$ such that $U \models \varphi(s_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
+  &&
+  \text{there exists a covering $U = \bigvee_i U_i$ such that for all~$i$:} \\
+  && \quad\quad \text{there exists~$s_0 \in F(U_i)$ such that $U_i \models \varphi(s_0)$}
+  \end{array} \]
+  \textbf{Theorem.} If~$U \models \varphi$ and if~$\varphi$ entails~$\psi$
+  intuitionistically, then~$U \models \psi$.
+
+  \jnote{1}{
+    A \emph{formula over an open~$U$} of~$X$ is a first-order formula (made up
+    using~${=}\ {\top}\ {\wedge}\ {\bot}\ {\vee}\ {\Rightarrow}\ {\forall}\
+    {\exists}$) over the signature which has one sort for each sheaf~$F$, one
+    constant symbol of sort~$F$ for each section~$s \in F(U)$, one function
+    symbol~$f : F \to G$ for each morphism of sheaves, and so on.
+  }
+\end{frame}
+
+\begin{frame}{Internalizing parameter-dependence}
+  \justifying
+  \small
+  Let~$X$ be a topological space. Let~$(f_x)_{x \in X}$ be a continuous family of continuous
+  functions~$\RR \to \RR$ (that is, let a continuous function $X \times \RR \to \RR,\,(x,a) \mapsto
+  f_x(a)$ be given). This family defines a sheaf morphism~$f : \C \to
+  \C$ by~$f_U : \C(U) \to \C(U),\,s \mapsto (x \mapsto f_x(s(x)))$.
+  \normalsize
+
+  \begin{itemize}\justifying
+    \item $X \models \text{``$\C$ is the set of Dedekind real numbers''}.$
+    \item $X \models \text{``the function $f : \C \to \C$ is
+    continuous''}$.\medskip
+    \item Iff $f_x(-1) < 0$ for all $x \in X$, then $X \models f(-1) < 0$.
+    \item Iff $f_x(+1) > 0$ for all $x \in X$, then $X \models f(+1) > 0$.
+    \item Iff all $f_x$ are increasing, then $X \models \text{``$f$ is
+    increasing''}$.\medskip
+    \item Iff there is an open cover~$X = \bigcup_i U_i$ such
+    that for each~$i$, there is a continuous function~$s : U_i \to \RR$
+    with~$f_x(s(x)) = 0$ for all~$x \in U_i$, then $X \models \exists s
+    \? \C\_ f(s) = 0$.
+  \end{itemize}
+
+  \jnote{1}{Hence:
+  \begin{enumerate}\justifying
+    \item The standard formulation of the intermediate value theorem fails
+    in~$\Sh(X)$, because its external interpretation is that in continuous
+    families of continuous functions, zeros can locally be picked continuously.
+    That claim is false, as
+    \fixedhref{https://rawgit.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{this
+    counterexample} demonstrates.
+
+    As a corollary, we deduce that the standard formulation of the
+    intermediate value theorem is not constructively provable.
+
+    \item The \emph{approximate} intermediate value theorem (stating
+    that for any~$\varepsilon > 0$, there is a number~$x$ such that~$|f(x)| <
+    \varepsilon$)
+    \fixedhref{https://mathoverflow.net/questions/253059/approximate-intermediate-value-theorem-in-pure-constructive-mathematics}{has
+    a constructive proof} and therefore holds in~$\Sh(X)$. The
+    external interpretation is that in continuous families of continuous
+    functions, approximate zeros can locally be picked continuously.
+
+    \item The \emph{monotone} intermediate value theorem, stating that a strictly
+    increasing continuous function with opposite signs has a unique zero,
+    admits a constructive proof and therefore holds in~$\Sh(X)$. The external
+    interpretation is that in continuous families of strictly increasing
+    continuous functions, zeros can globally be picked continuously. You are
+    invited to prove this fact directly, without reference to the internal
+    language. This exercise isn't particularly hard, but it's not trivial either.
+  \end{enumerate}}
 \end{frame}

### 2019-09-23 slides-herrsching2019.tex
Some notes slides

@@
 \newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}
+%\newcommand{\jnote}[2]{}
 
@@
   \end{enumerate}
-  % A geometric signature is \hil{propositional} iff its set
-  % of sorts is empty.
 
@@
   \jnote{1}{
+    The formulas appearing to the left and the right of the turnstile in axioms
+    of a geometric theory have to be \emph{geometric formula}. A formula is
+    \emph{geometric} iff it is built using only~${=}\ {\top}\ {\wedge}\
+    {\bot}\ {\vee}\ {\bigvee}\ {\exists}$ (but no ${\Rightarrow}\
+    {\forall}$). Universal quantification and implication can to some extent be
+    emulated by using free variables or the turnstile; but this allows
+    universal quantification and implication only to be used once, on top
+    level, not in nested subformulas.
+
     A geometric theory is called \emph{propositional} iff its set of
@@
     set of axioms.
+  }
 
+  \jnote{2}{
+    \textbf{Example.} The theory of rings has:
+    \begin{enumerate}
+      \item one sort: $R$
+      \item five constant symbols: $0$ and $1$ (nullary), $-$ (unary), $+$ and
+      $\cdot$ (binary)
+      \item no relation symbols
+      \item the usual axioms, such as $\top \vdash_{x:R,y:R} x + y = y + x$
+    \end{enumerate}
+
+    \textbf{Example.} The theory of surjections~$\NN \twoheadrightarrow \RR$ has:
+    \begin{enumerate}\justifying
+      \item no sorts
+      \item no function symbols
+      \item several nullary relation symbols, one for each pair~$\langle n,x
+      \rangle \in \NN \times \RR$, written~``$\varphi_{nx}$''
+      \item axioms: the axiom~$\top \vdash \bigvee_{x \in \RR} \varphi_{nx}$
+      for each~$n \in \NN$; the axiom~$\varphi_{nx} \wedge \varphi_{ny} \vdash
+      \bigvee\{ \top \,|\, x = y \}$ for each~$n \in \NN$, $x,y \in \RR$ (the
+      disjunction appearing in that axiom is over a subsingleton of formulas);
+      the axiom~$\top \vdash \bigvee_{n \in \NN} \varphi_{nx}$ for each~$x \in
+      \RR$
+    \end{enumerate}
+  }
+
+  \jnote{3}{
     Associated to any propositional geometric theory~$\TT$ is its
@@
     in canonical bijection with the set-based models of~$\TT$.
+
+    \textbf{Definition.} The \emph{localic real line} is the classifying locale
+    of the theory of Dedekind cuts.
+    The \emph{locale of surjections~$\NN \twoheadrightarrow \RR$} is the
+    classifying locale of the theory of surjections~$\NN \twoheadrightarrow
+    \RR$.
+    The \emph{empty locale} is the classifying locale of the \emph{inconsistent
+    theory} (which has no sorts, no function symbols, no relations symbols, and
+    one axiom, namely~$\top \vdash \bot$).
+
+    The points of the localic real line are in canonical bijection with
+    the Dedekind cuts, hence with the elements of the set of (Dedekind) real
+    numbers. The points of the locale of surjections~$\NN \twoheadrightarrow
+    \RR$ are in canonical bijection with the surjections~$\NN
+    \twoheadrightarrow \RR$. Assuming a classical metatheory, there are no such
+    points, but the locale is still nontrivial.
+  }
+
+  \jnote{4}{
+    Any locale is (isomorphic to) the classifying locale of some propositional
+    geometric theory, namely the ``theory of its points''. The same locale can
+    often be presented by widely different-looking theories, and the
+    topos-theoretic version of this observation is the basis for Olivia
+    Caramello's research program.
+
+    \textbf{Exercise.} Which explicit theory is the two-point locale (the
+    locale induced from the discrete topological space~$\{0,1\}$) the
+    classifying locale of?
   }
@@
   functions},~$\C_{\text{c}}(U) = \text{``the set of constant functions~$U \to
-  \RR$''}$, it usually not a sheaf.
+  \RR$''}$, is usually not a sheaf.
+
+  \jnote{1}{
+    All definitions on this slide literally also make sense for topological
+    spaces instead of locales, since only the notion of opens and their
+    inclusion relation is used.
+
+    The maps~$(\cdot)|^U_V$ are called \emph{restriction maps}. For the
+    sheaf~$\C$ of continuous real-valued functions, they are given by actual
+    restriction of functions to smaller domains, that is by
+    \[ \C(U) \longrightarrow \C(V),\ s \longmapsto s|_V. \]
+
+    \textbf{Exercise.} Find, in the case~$X = [0,1]$, a compatible family of
+    sections of the presheaf~$\C_\text{c}$ which shows that this
+    presheaf is not a sheaf.
+  }
 \end{frame}
@@
     symbol~$f : F \to G$ for each morphism of sheaves, and so on.
+
+    \textbf{Proposition.} The sheaf semantics is monotone and local in the
+    following sense: If~$V \leq U$, then~$U \models \varphi$ implies~$V \models
+    \varphi$. If~$U = \bigvee_i U_i$ and if~$U_i \models \varphi$ for all~$i$,
+    then~$U \models \varphi$.
+
+    \textbf{Exercise.} For any formula~$\varphi$ over any open~$U$ of any
+    locale~$X$, there is a largest open~$V \leq U$ such that~$V \models
+    \varphi$. (Hint: Set~$V \defeq \bigvee\{ W \leq U \,|\, W \models \varphi \}$
+    and exploit the monotonicity and locality of the sheaf semantics.)
+
+    \textbf{Exercise.} For any formula~$\varphi$ over any open~$U$ of any
+    locale~$X$, $U \models \neg\neg\varphi$ iff there exists an open~$V \leq U$
+    which is dense in~$U$ such that~$V \models \varphi$.
   }

### 2019-09-23 slides-herrsching2019.tex
Slides for Lecture III finished

@@
 
+% \begin{document}
+
+\section{Applications in constructive algebra}
+
+\begin{frame}
+  \centering\bigskip
+  \includegraphics[height=8em]{calabi-yau}
+
+  \large Lecture III:
+  \hil{Applications in constructive algebra}
+
+  \normalsize
+
+  \begin{hilblock}Without loss of generality, any reduced ring is a field.\end{hilblock}
+\end{frame}
+
+\begin{frame}{A remarkable sheaf}
+  \justifying
+  Let~$A$ be a reduced ring ($x^n = 0 \Rightarrow x = 0$).
+  Then there is a certain sheaf~$A^\sim$ of rings on a certain locale~$X$ such
+  that \ldots
+
+  {\centering $A^\sim$ is close to~$A$:\par}
+  \vspace*{-1.4em}
+  \begin{varblock}{\textwidth}{}
+    \justifying
+    There is a \hil{canonical bijection}~$A \to A^\sim(X)$.
+
+    $A^\sim$ inherits any property of~$A$ which is
+    \hil{localization-stable}.
+
+    A geometric sequent holds for~$A^\sim$ iff$^\star$ it holds for \hil{all
+    stalks}~$A_{\mathfrak{f}}$.
+  \end{varblock}
+
+  {\centering $A^\sim$ has better properties than~$A$:\par}
+  \vspace*{-1.6em}
+  \setbeamercolor{block body}{bg=red!30}
+  \setbeamercolor{structure}{fg=purple}
+  \begin{varblock}{\textwidth}{}
+    $A^\sim$ is a \hil{field}: $\forall x\?A^\sim\_ (\neg(\exists y\?A^\sim\_
+    xy = 1) \Rightarrow x = 0)$.
+
+    $A^\sim$ has \hil{$\boldsymbol{\neg\neg}$-stable equality}:
+    $\forall x,y\?A^\sim\_ \neg\neg(x = y) \Rightarrow x = y$.
+
+    \mbox{$A^\sim$ is \hil{anonymously Noetherian}.}\\[-1.2em]
+  \end{varblock}
+
+  {\centering This observation can be exploited to give \\ short, conceptual and
+  constructive proofs.\par}
+\end{frame}
+
+\begin{frame}{Examples}
+  \vspace*{-1.2em}
+  \begin{columns}[t]
+    \begin{column}[t]{0.58\textwidth}
+      \centering
+
+      \begin{varblock}{\textwidth}{Injective matrices}
+        \justifying
+        Let~$M$ be an injective matrix with more columns than rows over a ring~$A$.
+        Then~$1 = 0$ in~$A$.
+      \end{varblock}
+
+      \justifying
+      \textbf{Proof.} \bad{Assume not.} Then there is a \bad{maximal
+      prime filter} $\fff \subseteq A$. The matrix is injective over the \bad{field}~$A_\fff$;
+      contradiction to basic linear algebra.\medskip
+
+      \textbf{Proof.} $M$ is also injective as a matrix over~$A^\sim$.
+      This is a contradiction by basic linear
+      algebra. Thus~$X \models \bot$. This amounts to~$1 = 0$ in~$A$.
+    \end{column}
+
+    \begin{column}[t]{0.46\textwidth}
+      \centering
+
+      \begin{varblock}{\textwidth}{Generic freeness\phantom{p}}
+        \justifying
+        Let~$M$ be a finitely generated~$A$-module.
+        If~$f = 0$ is the only element of~$A$ such that~$M[f^{-1}]$ is a
+        free~$A[f^{-1}]$-module, then~$1 = 0$ in~$A$.
+      \end{varblock}
+
+      \justifying
+      \textbf{Proof.} See [Stacks Project].
+      \medskip
+
+      \textbf{Proof.} The claim amounts to \mbox{$X \models
+      \text{``$M^\sim$}$}$\text{ is \hil{not not} free''}$. This statement
+      follows from basic linear algebra over the field~$A^\sim$.
+    \end{column}
+  \end{columns}
+\end{frame}
+
+\begin{frame}[plain]
+  \centering
+  \includegraphics[height=\textheight]{lombardi-quitte}
+
+  \jnote{1}{
+    The book is \fixedhref{https://arxiv.org/abs/1605.04832}{available on the
+    arXiv} and warmly recommended, as are all slides by Thierry Coquand on
+    constructive algebra.«
+  }
+\end{frame}
+\addtocounter{framenumber}{-1}
+
 \end{document}

### 2020-01-08 paper-filmat.tex
to do
typo

@@
 theory~\cite{XXX}). This failure of classicality is often for interesting
-geometric or algebraic reasons instead of because of logical issues, as will be
+geometric or algebraic reasons instead of logical issues, as will be
 detailed in Section~\ref{several-toposes}.
@@
 \subsection{``There are infinitely many primes.''} A first-order formalization
-of this statement is ``for any natural number~$n$, there is a natural
+of this statement is ``for any natural number~$n$, there is a prime
 number~$p$ which is greater than~$n$'', and is known to be true in the standard
@@
 Also Giuseppe Cammarata.
+
+Mention forcing

### 2020-01-08 slides-herrsching2019.tex
more precision

@@
   U \models \bot &\Ll& \hcancel{\text{false}}{0pt}{3pt}{0pt}{-2pt}\ U = \bot \\
-  U \models s = t &\Ll& \text{$s = t$ when evaluated as elements of~$F(U)$} \\
+  U \models s =_F t &\Ll& \text{$s = t$ when evaluated as elements of~$F(U)$} \\
   U \models \varphi \wedge \psi &\Ll&
@@
       \justifying
-      \textbf{Proof.} See [Stacks Project].
+      \textbf{Proof.} See \href{https://stacks.math.columbia.edu/tag/051Q}{[Stacks Project]}.
       \medskip

### 2020-01-11 paper-filmat.tex
Beginnings of a beginning

@@
 \noindent
+Toposes can be pictured as mathematical universes in which we can do
+mathematics in. Most mathematicians spend all their professional life in just a
+single topos, the so-called \emph{standard topos}. However, besides the
+standard topos, there is a colorful host of alternate toposes which are just as
+worthy of mathematical study and in which mathematics plays out slightly
+differently.
 
-\section{Unsorted}
+For instance, there are toposes in which the axiom of choice and the
+intermediate value theorem from undergraduate calculus fail, toposes in which
+any function~$\RR \to \RR$ is continuous, toposes in which infinitesimal
+numbers exist and toposes whose properties depend on certain circumstances of
+our physical world.
+
+The purpose of this contribution is twofold.
+\begin{enumerate}
+\item We give a glimpse of the toposophic landscape, presenting several
+specific toposes and exploring their peculiar properties.
+
+\item We explicate how each topos provides a distinct lens through which the
+usual mathematical objects of the standard topos can be viewed.
+\end{enumerate}
+
+Viewed through such a lens, a given mathematical object can have different
+properties than it has when considered normally. In particular, it can have
+better properties for the purposes of specific applications, especially if
+the topos is custom-tailored to the object in question. This change of
+perspective has been used in mathematical practice and demonstrates that
+toposes go much beyond being logicians' testbeds. To give just a taste of what
+is possible, through the lens provided by an appropriate topos, any given ring
+can look like a field.
+
+We argue that toposes and specifically the change in perspective provided by
+toposes are ripe for philosophical analysis. In particular, there are the
+following connections with topics in the philosophy of mathematics:
+\begin{enumerate}
+\item Toposes enrich the realism/anti-realism debate in that they point the
+picture that the platonic heaven of mathematical objects is not unique: besides
+the standard heaven of the standard topos, we could fathom the alternate
+heavens of all other toposes, all embedded in a secord-order heaven.
+\item Mathematics is not only about studying mathematical objects, but also
+about studying the relations between mathematical objects. The distinct view
+on mathematical objects provided by any topos uncovers relations which
+otherwise remain hidden.
+\item In some cases, a mathematical relation can be expressed quite succinctly
+using the language of a specific topos and not so succinctly using the language
+of the standard topos. This phenomenon showcases the importance of
+\emph{appropriate language}.
+\item Toposes provide new impetus to study constructive mathematics and
+intuitionistic logic, in particular also to restrict to intuitionistic
+logic on the meta level and to consider the idea that the platonic heaven might
+be governed by intuitionistic logic.
+\end{enumerate}
+This note only touches on these topics and invites further research.
+
+\bigskip
+\paragraph{Acknowledgments} We are grateful to XXX for many invaluable
+discussions shaping this work and to XXX for their careful criticism of earlier
+drafts. XXX conference organizer and participants
+
+\section{Toposes as alternate mathematical universes}
+
+Formally, a topos is a certain kind of \emph{category}, containing objects and
+morphisms between those objects. For instance, the standard topos contains all
+sets as its objects and all maps between sets as morphisms. 
+% The term \emph{standard topos} reflects that all objects of
+% mathematical study can be encoded as sets -- a group is a \emph{set} of
+% elements together ...
+% \begin{ex}The \emph{standard topos}~$\Set$ contains all
+% sets as its objects and all maps between sets as morphisms.\end{ex}
+A careful definition of the notion of a topos requires some amount
+of category theory, but, as will be outlined in the following sections,
+exploring the mathematical universe of a given topos does not.
+
+Put briefly, the axioms state that a topos should share several categorical
+properties with the category of sets (namely having finite limits, being
+cartesian closed and having a subobject classifier); they ensure that each topos contains its
+own versions of familiar mathematical objects such as natural numbers, real
+numbers, groups and manifolds, and is closed under the usual constructions
+such as cartesian products or quotients.
+
+Given a topos~$\E$, we write~``$\E \models \varphi$'' to denote that a
+mathematical statement~$\varphi$ \emph{holds in~$\E$}. The meaning of~``$\E \models
+\varphi$'' is defined by recursion on the structure of~$\varphi$ following the
+so-called \emph{Kripke--Joyal translation rules}. For instance, the rule for
+translating conjunction reads
+\[ \E \models (\alpha \wedge \beta) \qquad\text{iff}\qquad
+  \E \models \alpha \quad\text{and}\quad \E \models \beta. \]
+The remaining translation rules are more involved; we do not list them here for
+the case of a general topos~$\E$, but we will state them in the next sections
+for several specific toposes.
+
+In the definition of~$\E \models \varphi$, the statement~$\varphi$ can be any
+statement in the language of a general version of higher-order predicate
+calculus with dependent types. In practice almost any mathematical statement
+can be interpreted in a given topos.\footnote{The main exceptions are
+statements from set theory, which typically make substantial use of a global
+membership predicate~``$\in$''. Toposes only support a typed \emph{local}
+membership predicate, where we may write~``$x \in A$'' only in the context of
+some fixed sort type~$B$ such that~$x$ is of type~$B$ and~$A$ is of
+type~$P(B)$, the power type of~$B$.}
+
+It is by the Kripke--Joyal translation rules that we can access the alternate
+universe of a topos. In the special case of the standard topos~$\Set$, the
+definition of~``$\Set \models \varphi$'' unfolds to~$\varphi$ for any
+statement~$\varphi$. Hence a statement holds in the standard topos if and only
+if it holds in the usual mathematical sense.
+
+\subsection{The logic of toposes} By their definition as special kinds of
+categories, toposes are merely algebraic structures not unlike groups and
+vector spaces are. Hence we need to argue why we picture toposes as
+mathematical universes while we do not elevate other kinds of algebraic structures
+in the same way. For us, the reason is given by the following metatheorem:
+
+\begin{thm}Let~$\E$ be a topos and~$\varphi$ be a statement such that~$\E \models
+\varphi$. If~$\varphi$ intuitionistically entails a further statement~$\psi$,
+then~$\E \models \psi$.\end{thm}
+
+XXX explain intuitionistically
+
+XXX explain that this metatheorem allows us to \emph{reason} in a topos
+
+XXX include something like this: A natural question is this: \emph{Which of
+the familiar mathematical facts of the standard topos carry over to arbitrary
+toposes~$\E$?} For any given mathematical statement~$\varphi$ and topos~$\E$,
+we can unroll the definition of~$\E \models \varphi$ to try to check
+whether~$\varphi$ holds in~$\E$ on an individual case-by-case basis; but it
+turns out that there is a general metatheorem ...
+
+XXX Mathematicians are familiar with the fact that the usual objects of
+mathematical study are governed by the laws of ordinary \emph{classical
+reasoning}. ...
+
+\subsection{Models of set theory}
 
@@
 much the same way as models of set theory can be viewed in this way. In fact,
-to any model~$M$ of a set theory such as~$\mathbf{ZF}$ or~$\mathbf{ZFC}$, there
+to any model~$M$ of a set theory such as~\textsc{zf} or~\textsc{zfc}, there
 is a topos~$\Set_M$ such that a statement holds in the internal language

### 2020-01-12 paper-filmat.tex
Sorted the unsorted

@@
 Toposes can be pictured as mathematical universes in which we can do
-mathematics in. Most mathematicians spend all their professional life in just a
+mathematics. Most mathematicians spend all their professional life in just a
 single topos, the so-called \emph{standard topos}. However, besides the
@@
 Viewed through such a lens, a given mathematical object can have different
-properties than it has when considered normally. In particular, it can have
+properties than when considered normally. In particular, it can have
 better properties for the purposes of specific applications, especially if
@@
 is possible, through the lens provided by an appropriate topos, any given ring
-can look like a field.
+can look like a field and hence mathematical techniques for fields also apply,
+through the lens, to rings.
 
@@
 \begin{enumerate}
-\item Toposes enrich the realism/anti-realism debate in that they point the
+\item Toposes enrich the realism/anti-realism debate in that they paint the
 picture that the platonic heaven of mathematical objects is not unique: besides
@@
 otherwise remain hidden.
+% XXX: mention Olivia?
 \item In some cases, a mathematical relation can be expressed quite succinctly
@@
 \end{enumerate}
-This note only touches on these topics and invites further research.
+This note touches on these topics and invites further research.
 
@@
 Formally, a topos is a certain kind of \emph{category}, containing objects and
-morphisms between those objects. For instance, the standard topos contains all
-sets as its objects and all maps between sets as morphisms. 
-% The term \emph{standard topos} reflects that all objects of
-% mathematical study can be encoded as sets -- a group is a \emph{set} of
-% elements together ...
-% \begin{ex}The \emph{standard topos}~$\Set$ contains all
-% sets as its objects and all maps between sets as morphisms.\end{ex}
-A careful definition of the notion of a topos requires some amount
-of category theory, but, as will be outlined in the following sections,
-exploring the mathematical universe of a given topos does not.
+morphisms between those objects. The formal definition, recorded here only for
+reference, requires some amount of category theory, but, as will be outlined in
+the following sections, exploring the mathematical universe of a given topos
+does not.
+
+\begin{defn}A \emph{topos} (more precisely \emph{elementary topos} or
+\emph{logos} with a natural numbers object) is a category which has all finite
+limits, is cartesian closed and has a subobject classifier.\end{defn}
 
 Put briefly, the axioms state that a topos should share several categorical
-properties with the category of sets (namely having finite limits, being
-cartesian closed and having a subobject classifier); they ensure that each topos contains its
+properties with the category of sets; they ensure that each topos contains its
 own versions of familiar mathematical objects such as natural numbers, real
@@
 such as cartesian products or quotients.
+The prototypical topos is the standard topos:
+
+\begin{ex}The \emph{standard topos}~$\Set$ contains all
+sets as its objects and all maps between sets as morphisms.\end{ex}
 
@@
 membership predicate, where we may write~``$x \in A$'' only in the context of
-some fixed sort type~$B$ such that~$x$ is of type~$B$ and~$A$ is of
+some fixed type~$B$ such that~$x$ is of type~$B$ and~$A$ is of
 type~$P(B)$, the power type of~$B$.}
@@
 \subsection{The logic of toposes} By their definition as special kinds of
-categories, toposes are merely algebraic structures not unlike groups and
-vector spaces are. Hence we need to argue why we picture toposes as
-mathematical universes while we do not elevate other kinds of algebraic structures
-in the same way. For us, the reason is given by the following metatheorem:
+categories, toposes are merely algebraic structures not unlike groups or vector
+spaces. Hence we need to argue why we picture toposes as mathematical universes
+while we do not elevate other kinds of algebraic structures in the same way.
+For us, this usage is justified by the following metatheorem:
 
-\begin{thm}Let~$\E$ be a topos and~$\varphi$ be a statement such that~$\E \models
-\varphi$. If~$\varphi$ intuitionistically entails a further statement~$\psi$,
-then~$\E \models \psi$.\end{thm}
+\begin{thm}\label{thm:reasoning}Let~$\E$ be a topos and let~$\varphi$ be a
+statement such that~$\E \models \varphi$. If~$\varphi$ intuitionistically
+entails a further statement~$\psi$ (that is, if it is provable in
+intuitionistic logic that~$\varphi$ entails~$\psi$), then~$\E \models
+\psi$.\end{thm}
 
-XXX explain intuitionistically
+This metatheorem allows us to \emph{reason} in toposes. When first exploring a
+new topos~$\E$, we need to employ the Kripke--Joyal translation rules each time
+we want to check whether a statement holds in~$\E$. But as soon as we
+have amassed a stock of statements known to be true in~$\E$, we can find more
+by deducing their logical consequences.
 
-XXX explain that this metatheorem allows us to \emph{reason} in a topos
+For instance, in any topos where the statement ``any map~$\RR \to \RR$ is
+continuous'' is true, also the statement ``any map~$\RR \to \RR^2$ is
+continuous'' is, since there is an intuitionistic proof that a map into a
+higher-dimensional Euclidean space is continuous if its individual components
+are.
 
-XXX include something like this: A natural question is this: \emph{Which of
-the familiar mathematical facts of the standard topos carry over to arbitrary
-toposes~$\E$?} For any given mathematical statement~$\varphi$ and topos~$\E$,
-we can unroll the definition of~$\E \models \varphi$ to try to check
-whether~$\varphi$ holds in~$\E$ on an individual case-by-case basis; but it
-turns out that there is a general metatheorem ...
+The only caveat of Theorem~\ref{thm:reasoning} is that toposes generally only
+support intuitionistic reasoning, not the full power of the ordinary
+\emph{classical reasoning}. That is, within most toposes, the law of excluded
+middle ($\varphi \vee \neg\varphi$) and the law of double negation elimination
+($\neg\neg\varphi \Rightarrow \varphi$) are not available.
 
-XXX Mathematicians are familiar with the fact that the usual objects of
-mathematical study are governed by the laws of ordinary \emph{classical
-reasoning}. ...
+While it may appear that these two laws pervade any mathematical theory, in
+fact a substantial amount of mathematics can be developed intuitionistically
+(see for
+instance~\cite{mines-richman-ruitenburg:constructive-algebra,lombardi-quitte:constructive-algebra}
+for constructive algebra,~\cite{bishop-bridges:constructive-analysis} for
+constructive analysis
+and~\cite{bauer:int-mathematics,bauer:video,melikhov:intuitionistic-logic} for
+further references) and hence the alternate universes provided by toposes
+cannot be too strange: In any topos, there are infinitely many prime numbers,
+the square root of two is not rational and the powerset of the naturals is
+uncountable.
 
-\subsection{Models of set theory}
+\begin{defn}A topos~$\E$ is \emph{boolean} if and only if the laws of classical
+logic are true in~$\E$.\end{defn}
 
-Toposes can be pictured as ``universes in which we can do mathematics in'' in
-much the same way as models of set theory can be viewed in this way. In fact,
+The standard topos is boolean if and only if, as is commonly supposed, the laws
+of classical logic hold on the meta level. Most toposes of interest are not
+boolean, irrespective of one's philosophical commitments about the meta level,
+and conversely some toposes are boolean even if classical logic is not
+available on the meta level.
+
+\begin{rem}The axiom of choice (which is strictly speaking not part of
+classical logic, but of classical set theory) is also not available in most
+toposes. By \emph{Diaconescu's theorem}, the axiom of choice implies the law of
+excluded middle in presence of other axioms which are available in any topos.
+\end{rem}
+
+%A natural question is this: \emph{Which of
+%the familiar mathematical facts of the standard topos carry over to arbitrary
+%toposes?} For any given mathematical statement~$\varphi$ and topos~$\E$,
+%we can unroll the definition of~$\E \models \varphi$ to try to check
+%whether~$\varphi$ holds in~$\E$ on an individual case-by-case basis; but since
+%any topos supports intuitionistic reasoning, we may at once conclude: \emph{Any
+%theorem of constructive mathematics, that is any theorem deduced using only
+%intuitionistic logic, is valid in any topos.}
+
+%Mathematicians are familiar with the fact that the usual objects of
+%mathematical study are governed by the laws of ordinary \emph{classical
+%reasoning}. ...
+
+\subsection{Relation to models of set theory} In set theory, philosophy and
+logic, models of set theories ... XXX
+
+Toposes can be pictured as
+``universes in which we can do mathematics'' in much the same way as models of
+set theory. Just as using forcing and other techniques we can construct new
+models of set theory from given ones, thereby exploring the set-theoretic
+multiverse, we can construct new toposes from given toposes.
+
+However, there are important differences between the notion of mathematical
+universe as provided by toposes and as provided by models of set theory, both
+regarding the subject matter and the reasons for why we are interested in them.
+
+Firstly, toposes are more general than models of set theory. By definition, a
+model of \textsc{zfc} will always satisfy the axioms of \textsc{zfc}; in
+contrast, most toposes do not even validate the law of excluded middle, much
+less so the axiom of choice.
+
+Secondly, there is a shift in emphasis. An important philosophical objective
+for studying models of set theory is to explore which notions of sets are
+coherent: Does the cardinality of the reals need to be the cardinal directly
+succeeding~$\aleph_0$, the cardinality of the naturals? No, there are models of
+set theory in which the continuum hypothesis fails. Do non-measurable sets of
+reals need to exist? No, in models of~$\textsc{zf}+\textsc{ad}$,
+Zermelo--Fraenkel set theory plus the axiom of determinacy, it is a theorem that
+every subset of~$\RR^n$ is Lebesgue-measurable. Can the axiom of choice be
+added to the axioms of~\textsc{zf} without causing inconsistency? Yes, if~$M$
+is a model of~\textsc{zf} then~$L^M$, the set of definable sets of~$M$, is a
+model of~\textsc{zfc}.  %~\cite{XXX, maybe SEP?}
+
+While toposes can be used for similar such purposes, and indeed have been,
+especially to explore the various intuitionistic notions of sets, an important
+aspect of topos theory is that toposes are used to explore the standard
+mathematical universe: truth in the \effective topos tells us what is
+computable; truth in sheaf toposes tells us what's true locally; toposes
+adapted to synthetic differential geometry can be used to rigorously work with
+infinitesimals. All of these examples will be presented in more detail in the
+next sections.
+
+In a sense which can be made precise, toposes allow us to study the usual
+objects of mathematics from a different point of view -- one such view for
+every topos -- and it is a beautiful and intriguing fact that with the sole
+exception of the law of excluded middle, the laws of logic apply to
+mathematical objects also when viewed through the lens of a specific topos.
+
+In fact,
 to any model~$M$ of a set theory such as~\textsc{zf} or~\textsc{zfc}, there
@@
 if and only if it's true in the ordinary mathematical sense.
-
-Carrying the analogy further, just as using forcing and other techniques we can
-construct new models of set theory from given ones, thereby exploring the
-set-theoretic multiverse, we can construct new toposes from given toposes.
-However, there are two important differences between the notion of mathematical
-universe as provided by toposes and as provided by models of set theory, both
-regarding the subject matter and the reasons for why we are interested in them.
-
-Firstly, toposes are more general than models of set theory. By definition, a
-model of~$\mathbf{ZFC}$ will always satisfy the axioms of~$\mathbf{ZFC}$; in
-contrast, most toposes do not even validate the law of excluded middle, much
-less so the axiom of choice (which implies the law of excluded middle in
-presence of the remaining axioms of intuitionistic Zermelo--Fraenkel set
-theory~\cite{XXX}). This failure of classicality is often for interesting
-geometric or algebraic reasons instead of logical issues, as will be
-detailed in Section~\ref{several-toposes}.
-
-Secondly, there is a difference in motivation. The main philosophical reason
-for studying models of set theory is to study which notions of sets are
-coherent: Does the cardinality of the reals need to be the cardinal directly
-succeeding~$\aleph_0$, the cardinality of the naturals? No, there are models of
-set theory in which the continuum hypothesis fails. Do non-measurable sets of
-reals need to exist? No, in models of~$\mathbf{ZF}+\mathbf{AD}$,
-Zermelo--Fraenkel set theory plus the axiom of determinacy, it's a theorem that
-every subset of~$\RR^n$ is Lebesgue-measurable. Can the axiom of choice be
-added to the axioms of~$\mathbf{ZF}$ without causing inconsistency? Yes, if~$M$
-is a model of~$\mathbf{ZF}$ then~$L^M$, the set of definable sets of~$M$, is a
-model of~$\mathbf{ZFC}$.~\cite{XXX, maybe SEP?}
-
-While toposes can be used for similar such purposes, and indeed have been,
-especially to explore the various intuitionistic notions of sets, an important
-aspect of topos theory is that toposes are used to explore the standard
-mathematical universe: Truth in the \effective topos tells us what is
-computable; truth in sheaf toposes tells us what's true locally, in a fashion
-which depends continuously on parameters; the little Zariski topos of a general
-commutative ring can be used to pretend that the given ring is a field; toposes
-adapted to synthetic differential geometry can be used to rigorously work with
-infinitesimals. All of these examples will be presented in more detail in
-Section~\ref{several-toposes}.
-
-In a certain precise way, toposes allow us to study the common objects of
-mathematics from a different point of view -- one such view for every topos --
-and it is a beautiful and intriguing fact that with the sole exception of the
-law of excluded middle, the laws of logic apply to mathematical objects also
-when viewed through the lens of a specific topos.
 

### 2020-01-12 paper-filmat.tex
More on models of set theory

@@
 
-\begin{ex}The \emph{standard topos}~$\Set$ contains all
-sets as its objects and all maps between sets as morphisms.\end{ex}
+\begin{defn}The \emph{standard topos}~$\Set$ is the category which contains all
+sets as its objects and all maps between sets as morphisms.\end{defn}
 
@@
 \subsection{Relation to models of set theory} In set theory, philosophy and
-logic, models of set theories ... XXX
+logic, models of set theories are studied. These are structures~$(M,\in)$
+validating the axioms of some set theory such as Zermelo--Fraenkel set theory
+with choice~\textsc{zfc}, and they can be pictured as ``universes in which we
+can do mathematics'' in much the same way as toposes.
 
-Toposes can be pictured as
-``universes in which we can do mathematics'' in much the same way as models of
-set theory. Just as using forcing and other techniques we can construct new
+In fact, to any model~$(M,\in)$ of a set theory such as~\textsc{zf}
+or~\textsc{zfc}, there is a topos~$\Set_M$ such that a statement holds
+in~$\Set_M$ if and only if it holds in~$M$.\footnote{The topos~$\Set_M$ can be
+described as follows: Its objects are the elements of~$M$, that is the entities
+which~$M$ believes to be sets, and its morphisms are those entities which~$M$
+believes to be maps. The topos~$\Set_M$ validates the axioms of
+\textsc{etcs}~\cite{XXX}, and models are elementarily equivalent if their
+associated toposes are equivalent as categories.}
+
+\begin{ex}The topos~$\Set_V$ associated to the universe~$V$ of all sets (if
+this structure is available in one's chosen ontology) coincides with the
+standard topos~$\Set$.\end{ex}
+
+In set theory, we use forcing and other techniques to construct new
 models of set theory from given ones, thereby exploring the set-theoretic
-multiverse, we can construct new toposes from given toposes.
+multiverse. There are similar techniques available for constructing new toposes
+from given ones, and some of these correspond to the techniques from set
+theory.
 
-However, there are important differences between the notion of mathematical
-universe as provided by toposes and as provided by models of set theory, both
+However, there are also important differences between the notion of mathematical
+universes as provided by toposes and as provided by models of set theory, both
 regarding the subject matter and the reasons for why we are interested in them.
@@
 mathematical objects also when viewed through the lens of a specific topos.
-
-In fact,
-to any model~$M$ of a set theory such as~\textsc{zf} or~\textsc{zfc}, there
-is a topos~$\Set_M$ such that a statement holds in the internal language
-of~$\Set_M$ if and only if it holds in~$M$.\footnote{The topos~$\Set_M$ can be
-described as follows: Its objects are the elements of~$M$, that is the things
-which~$M$ believes to be sets, and its morphisms are those things which~$M$
-believes to be maps. The topos~$\Set_M$ validates the axioms of
-$\mathbf{ETCS}$~\cite{XXX}, and for models which are not elementarily
-equivalent, their associated toposes will not be equivalent.} This
-includes~$V$, the trivial inner model consisting of all sets there are, in case
-such a thing exists in one's chosen ontology; in this case the associated
-topos~$\Set_V$ is the category of all sets, and a statement holds in~$\Set_V$
-if and only if it's true in the ordinary mathematical sense.
 
@@
 not be equivalent and reflect higher-order properties of the employed models.
-Two of these further toposes are of special philosphical interest.
+Two of these further toposes are of special philosophical interest.
 
+\bigskip
 \paragraph{Hypercomputation}
@@
 
+\bigskip
 \paragraph{Machines of the real physical world} A second variant of the
@@
 steps can be carried out in finite time and if it is possible to form
-tamper-free private communication channels~\cite{XXX}.  % XXX: look up precise formulation
+tamper-free private communication channels~\cite{bauer:int-mathematics}.
 
@@
 number~$a$ is actually a continuous family of real numbers on~$X$, that is, a
-function~$a : X \to \RR$.
+continuous function~$a : X \to \RR$.
 

### 2020-01-13 paper-filmat.tex
Infinitesimal numbers

@@
 
+% \begin{document}
+
+\section{XXX ... infinitesimal numbers ...}
+
+The idea of \emph{infinitesimal numbers} -- numbers which lie between~$-1/n$
+and~$1/n$ for any natural number~$n$ -- has a long and rich history. They are
+not part of today's standard setup of the reals, but they are still intriguing
+as calculational tools and as a device to bring mathematical intuition and
+mathematical formalism closer together.
+
+For instance, employing numbers~$\varepsilon$ such that~$\varepsilon^2 = 0$, we can
+compute derivatives blithely as follows, without requiring the notion of
+limits:
+\begin{align*}
+  (x + \varepsilon)^2 - x^2 &= x^2 + 2x\varepsilon + \varepsilon^2 - x^2 = 2x\varepsilon \\
+  (x + \varepsilon)^3 - x^3 &= x^3 + 3x^2\varepsilon + 3x\varepsilon^2 + \varepsilon^3 - x^3 = 3x^2\varepsilon
+\end{align*}
+In each case, the derivative is visible as the coefficient of~$\varepsilon$ in
+the result. A further example is from geometry: Having a nontrivial
+set~$\Delta$ of infinitesimal numbers available allows us to define a
+\emph{tangent vector} to a manifold~$M$ to be a map~$\gamma : \Delta \to M$. This
+definition precisely captures the intuition that a tangent vector is an
+infinitesimal curve.
+
+\subsection{Hyperreal numbers} There are several ways for introducing
+infinitesimals into rigorous mathematics. One is Robinson's \emph{nonstandard
+analysis}, where we enlarge the field~$\RR$ of real numbers to a
+field~$^\star\RR$ of \emph{hyperreal numbers} by means of a non-principal
+ultrafilter.
+
+The hyperreals contain an isomorphic copy of the ordinary reals as the
+so-called \emph{standard elements}, and they also contain infinitesimal numbers
+and their inverses, transfinite numbers, and support a powerful \emph{transfer
+principle}: Any statement which doesn't refer to standardness is true for the
+hyperreals if and only if it is true for the ordinary reals.
+
+In the ``if'' direction, the transfer principle is useful for importing
+knowledge about the ordinary reals in the hyperreal realm. For instance,
+addition of hyperreals is commutative because addition of reals is.
+By the ``only if'' direction, a theorem established for the hyperreals also
+holds for the ordinary reals. In this way, the infinitesimal numbers of
+nonstandard analysis can be viewed as a convenient fiction, generating a
+conservative extension of the usual setup of mathematics.
+
+However, the realization of this fiction crucically rests on a non-principal
+ultrafilter, whose existence requires principles which go beyond the means of
+Zermelo--Fraenkel set theory~\textsc{zf}.\footnote{A hyperreal number is
+represented by an infinite sequence~$(x_0,x_1,x_2,\ldots)$ of ordinary real
+numbers. For instance, the sequence~$(1,1,1,\ldots)$ represents the hyperreal
+version of the number~$1$, the sequence~$(1,\frac{1}{2},\frac{1}{3},\ldots)$
+represents an infinitesimal number and its inverse~$(1,2,3,\ldots)$ represents
+a transfinite number.
+%
+The sequence~$(1,1,1,\ldots)$ is deemed positive, and so is~$(-1,1,1,1,\ldots)$,
+which differs from the former only in finitely many places. But
+should~$(1,-1,1,-1,\ldots)$ be deemed positive or negative? Whatever the
+answer, our decision has consequences for other sequences. For
+instance~$(-1,1,-1,1,\ldots)$ should be assigned the opposite sign
+and~$(\tan(1),\tan(-1),\tan(1),\tan(-1),\ldots)$ the same.
+%
+A non-principal ultrafilter is a set-theoretic gadget which fixes all such
+decisions once and for all in a coherent manner. Having such an ultrafilter
+available, a sequence~$(x_0,x_1,x_2,\ldots)$ is deemed positive if and only if
+the set~$\{i \in \NN \,|\, x_i > 0\}$ is part of the ultrafilter.}
+Non-principal ultrafilters cannot be described in explicit terms, and
+they are also not at all canonical structures: \textsc{zfc} proves that there
+are~$2^{2^{\aleph_0}}$ many~\cite{pospisil:ultrafilters}.
+
+A practical consequence of this non-constructivity is that it can be hard to
+unwind proofs which employ hyperreal numbers to direct proofs, and even where
+possible there is no general procedure for doing so.
+
+\subsection{Topos-theoretic alternatives to the hyperreal numbers} Topos theory
+provides several constructive alternatives for realizing infinitesimals.
+One such is ``cheap nonstandard analysis'' by Terence Tao~\cite{tao:cheap-nsa}
+and is related to Robinson's nonstandard analysis like potential infinity is
+related to actual infinity. Instead of appealing to the axiom of choice to
+obtain a completed ultrafilter, cheap nonstandard analysis constructs larger
+and larger approximations to an ideal ultrafilter on the go.
+
+The following section presents the basics of \emph{synthetic differential
+geometry}~\cite{kock:sdg,kock:new-methods}. This is a further topos-theoretic
+approach to infinitesimals which is suited to illustrate the philosophy of toposes
+as lenses. Its stated goal is to devise a rigorous context in which the
+writings of Sophus Lie, who freely employed infinitesimals in his seminal work
+on differential geometry, can be effortlessly interpreted, staying close to the
+original.
+
+\subsection{XXX ...}
+
 \printbibliography
@@
 * use the phrase "very rich and very deep" (or similarly) about set theory
+* PA doesn't have models
+* Sven(?)
 

### 2020-01-14 paper-filmat.tex
Details on the Zariski topos

@@
-\documentclass[oneside]{amsart}
+\documentclass[oneside,reqno]{amsart}
 
@@
 \author{Ingo Blechschmidt}
-\address{Università di Verona \\
-Department of Computer Science \\
-Strada le Grazie 15 \\
-37134 Verona, Italy}
-\email{iblech@speicherleck.de}
+\address{Universität Augsburg \\
+Institut für Mathematik \\
+Universitätsstr. 14 \\
+86159 Augsburg, Germany}
+\email{ingo.blechschmidt@math.uni-augsburg.de}
 
@@
 \newtheorem{conjecture}[defn]{Conjecture}
+
+\newenvironment{indentblock}{%
+  \list{}{\leftmargin\leftmargin}%
+  \item\relax
+}{%
+  \endlist
+}
 
@@
 \DeclareMathOperator{\Sh}{Sh}
+\DeclareMathOperator{\Zar}{Zar}
 \DeclareMathOperator{\PSh}{PSh}
@@
 added to the axioms of~\textsc{zf} without causing inconsistency? Yes, if~$M$
-is a model of~\textsc{zf} then~$L^M$, the set of definable sets of~$M$, is a
+is a model of~\textsc{zf} then~$L^M$, the constructible sets of~$M$ form a
 model of~\textsc{zfc}.  %~\cite{XXX, maybe SEP?}
@@
     $U \models (\varphi \vee \psi)$ \> iff there is an open covering~$U =
-    \bigcup_i U_i$ such that, \\ \> \qquad for each~$i$, $U_i \models \varphi$
+    \bigcup_i U_i$ such that, \\ \> \qquad for each index~$i$, $U_i \models \varphi$
     or $U_i \models \psi$. \\
@@
     $V \models \varphi$ implies~$V \models \psi$. \\
-    $U \models (\forall a\?\RR\_ \varphi(x))$ \> iff for every open~$V
-    \subseteq U$ and any continuous function~$a_0 : V \to \RR$, $V \models
+    $U \models (\forall a\?\RR\_ \varphi(a))$ \> iff for every open~$V
+    \subseteq U$ and \\\>\qquad any continuous function~$a_0 : V \to \RR$, $V \models
     \varphi(a_0)$. \\
-    $U \models (\exists a\?\RR\_ \varphi(x))$ \> iff there is an open
-    covering~$U = \bigcup_i U_i$ such that, \\ \> \qquad for each~$i$,
-    there exists a continuous function~$a_0 : U_i \to \RR$ with~$U_i \models
+    $U \models (\exists a\?\RR\_ \varphi(a))$ \> iff there is an open
+    covering~$U = \bigcup_i U_i$ such that, \\ \> \qquad for each index~$i$,
+    there exists a \\\>\qquad continuous function~$a_0 : U_i \to \RR$ with~$U_i \models
     \varphi(a_0)$.
@@
 limits:
-\begin{align*}
+\begin{equation}\label{eq:derivative}\tag{$\star$}
+  \begin{aligned}
     (x + \varepsilon)^2 - x^2 &= x^2 + 2x\varepsilon + \varepsilon^2 - x^2 = 2x\varepsilon \\
     (x + \varepsilon)^3 - x^3 &= x^3 + 3x^2\varepsilon + 3x\varepsilon^2 + \varepsilon^3 - x^3 = 3x^2\varepsilon
-\end{align*}
+  \end{aligned}
+\end{equation}
 In each case, the derivative is visible as the coefficient of~$\varepsilon$ in
@@
 
-The following section presents the basics of \emph{synthetic differential
+The following section presents a topos used in \emph{synthetic differential
 geometry}~\cite{kock:sdg,kock:new-methods}. This is a further topos-theoretic
@@
 as lenses. Its stated goal is to devise a rigorous context in which the
-writings of Sophus Lie, who freely employed infinitesimals in his seminal work
-on differential geometry, can be effortlessly interpreted, staying close to the
-original.
+writings of Sophus Lie, who freely employed infinitesimals in his seminal
+works, can be effortlessly interpreted, staying close to the
+original and requiring no coding.
 
-\subsection{XXX ...}
+\subsection{The Zariski topos} The starting point is the observation that
+while the field~$\RR$ of ordinary real numbers doesn't contain infinitesimals
+(except for zero), the ring~$\RR[\varepsilon]/(\varepsilon^2)$ of \emph{dual
+numbers} does. This ring has the cartesian product~$\RR \times \RR$ as its
+underlying set and the ring operations are defined such
+that~$\varepsilon^2 = 0$, where~$\varepsilon \defeq \langle 0,1 \rangle$:
+\[
+  \langle a,b \rangle + \langle a',b' \rangle \defeq
+  \langle a+a', b+b' \rangle \qquad
+  \langle a,b \rangle \cdot \langle a',b' \rangle \defeq
+  \langle aa', ab'+a'b \rangle
+\]
+We write~$\langle a,b \rangle$ more clearly
+as~$a+b\varepsilon$.
+
+The dual numbers contain infinitesimal numbers, more precisely \emph{nilsquare
+numbers}, namely all numbers of the form~$b \varepsilon$ for~$b \in \RR$, and
+they are sufficient to rigorously reproduce derivative computations of
+polynomials such as~\eqref{eq:derivative}.
+
+However, the dual numbers are severely lacking in other aspects. Firstly, they
+don't contain any nilcubic numbers which are not already nilsquare. These are
+required in order to extend calculations like~\eqref{eq:derivative} to second
+derivatives, as in
+\[ (x+\varepsilon)^3 - x^3 = x^3 + 3x^2\varepsilon + \frac{1}{2!} 6x
+\varepsilon^2. \]
+Secondly, the dual numbers contain, up to scaling, only a single infinitesimal
+number. Further independent infinitesimals are required in order to deal with
+functions of several variables, as in
+\[ f(x+\varepsilon,y+\varepsilon') - f(x,y) = D_xf(x,y)\varepsilon +
+D_yf(x,y)\varepsilon'. \]
+Thirdly, and perhaps most importantly, the ring of dual numbers fails to be a
+field. The only invertible dual numbers are the numbers of the form~$a +
+b\varepsilon$ with~$a$ invertible in the reals; it is not true that any nonzero
+dual number is invertible.
+
+The first deficiency could be fixed by passing
+from~$\RR[\varepsilon]/(\varepsilon^2)$ to~$\RR[\varepsilon]/(\varepsilon^3)$
+(a ring whose elements are triples and whose ring operations are defined such
+that~$\langle0,0,1\rangle^3 = 0$) and the second by passing
+from~$\RR[\varepsilon]/(\varepsilon^2)$
+to~$\RR[\varepsilon,\varepsilon']/(\varepsilon^2,\varepsilon'^2,\varepsilon\varepsilon')$.
+In a sense, both of these proposed replacements are better \emph{stages} than
+the basic~$\RR[\varepsilon]/(\varepsilon)^2$ or even~$\RR$ itself. However,
+similar criticisms can be mounted against any of these better stages, and the
+problem that all these substitutes are not fields persists.
+
+\bigskip
+\paragraph{Introducing the topos}
+The \emph{Zariski topos of~$\RR$},~$\Zar(\RR)$, meets all of these challenges. It
+contains a ring~$R$, the so-called \emph{ring of smooth numbers}, which reifies
+the real numbers, the dual numbers, the two proposed better stages and indeed
+any finitely presented~$\RR$-algebra into a single coherent entity. The
+Kripke--Joyal translations rules of~$\Zar(\RR)$ are listed in
+Table~\ref{table:zariski}. Any evaluation of an internal statement starts out
+with the most basic stage of all, the ordinary reals~$\RR$; then, during the
+course of evaluation, the current stage is successively refined to better
+stages (further finitely presented~$\RR$-algebras).
+
+For instance, universal quantification~``$\forall x \? R$'' not only refers to
+all elements of the current stage, but also to any elements of any refinement
+of the current stage. Similarly, negation~``$\neg\varphi$'' does not only mean
+that~$\varphi$ would imply~$1 = 0$ in the current stage, but also that it does
+so at any later stage.
+
+For reference purposes only, we include the precise definition of the Zariski topos.
+\begin{defn}The \emph{Zariski topos of~$\RR$},~$\Zar(\RR)$, is the category of
+functors from finitely presented~$\RR$-algebras to sets. The object~$R$
+of~$\Zar(\RR)$ is the tautologous functor~$A \mapsto A$.
+\end{defn}
+
+\paragraph{Properties of the smooth numbers}
+As a concrete example, the Kripke--Joyal translation of the statement that~$R$ is a field,
+\[ \Zar(\RR) \models \forall x\?R\_ (\neg(x = 0) \Rightarrow \exists y\?R\_ xy = 1), \]
+is this:
+\begin{indentblock}
+For any stage~$A$ and any element~$x \in A$, \\
+${\qquad}$ if for any later stage~$B$ of~$A$ \\
+${\qquad\qquad}$ in which~$x = 0$ holds \\
+${\qquad\qquad}$ also~$1 = 0$ holds, \\
+${\qquad}$ then~$A$ can be covered by later stages~$B_i$ such that, for each
+index~$i$, \\
+${\qquad\qquad}$ there is an element~$y \in B_i$ such that~$xy = 1$ in~$B_i$.
+\end{indentblock}
+And indeed, this statement is true. Let a stage~$A$ (a finitely
+presented~$\RR$-algebra) and an element~$x \in A$ which fulfills the stated
+condition be given. Then~$x = 0$ holds in the refinement~$B \defeq A/(x)$.
+Hence~$1 = 0$ holds in~$B$. By elementary algebra, this means that~$x$ is
+invertible in~$A$. Hence the conclusion holds for the singleton covering
+of~$A$ by~$B_1 \defeq A$.
+
+Within~$\Zar(\RR)$, we set~$\Delta \defeq \{ \varepsilon \? R \,|\,
+\varepsilon^2 = 0 \}$. Then~$R$ has the following properties:
+\begin{enumerate}
+\item Law of cancellation: $\forall x \? R\_ \forall y \? R\_ \bigl((\forall
+\varepsilon \? \Delta\_ x\varepsilon = y\varepsilon)
+\Rightarrow x = y\bigr)$
+\item Axiom of micro-affinity: $\forall f \? R^\Delta\_ \exists! a \? R\_
+\forall \varepsilon \? \Delta\_ f(\varepsilon) = f(0) + a\varepsilon$
+\end{enumerate}
+The unique number~$a$ in the axiom of micro-affinity deserves to be
+called~``$f'(0)$''; this is how we synthetically define the derivative in
+synthetic differential geometry.
+
+Having motivated the Zariski topos by the desire to devise a universe with
+infinitesimals, the actual ontological status of the infinitesimal numbers in
+the Zariski topos is more nuanced. The law of cancellation implies that,
+within~$\Zar(\RR)$, it is not the case that zero is the only nilsquare number.
+However, this doesn't mean that there actually \emph{is} a nilsquare number
+in~$R$. In fact, any nilsquare number cannot be nonzero, as nonzero numbers are
+invertible while nilsquare numbers are not. Hence any nilsquare number in~$R$
+is \emph{not not} zero. This state of affairs is only possible in an
+intuitionistic context.
+
+\begin{rem}The ring~$R$ of smooth numbers does not coincide with the Cauchy reals,
+the Dedekind reals or indeed any well-known construction of the reals
+within~$\Zar(\RR)$. This observation explains why~$R$ can satisfy the law of
+cancellation even though it is an intuitionistic theorem that the only
+nilquadratic number in any flavor of the reals is zero.
+\end{rem}
+
+\subsection{Well-adapted models} The Zariski topos of~$\RR$ contains
+infinitesimals in a satisfying manner. However it is not suited as a home for
+synthetic differential geometry, a first indication being that in~$\Zar(\RR)$,
+any function~$R \to R$ is polynomial. Hence important functions such as the
+exponential function do not exist in~$\Zar(\RR)$. More comprehensively, the
+Zariski topos is not a well-adapted model in the sense of the following
+definition.
+
+\begin{defn}\label{defn:well-adapted}A \emph{well-adapted model} of synthetic
+differential geometry is a topos~$\E$ together with a ring~$R$ in~$\E$ such
+that:
+\begin{enumerate}
+\item The ring~$R$ is a field.
+\item The ring~$R$ validates the axiom of micro-affinity and several related
+axioms.
+\item There is a fully faithful functor~$i : \mathrm{Mnf} \to \E$ embedding the
+category of smooth manifolds into~$\E$.
+\item The ring~$R$ is coincides with~$i(\RR^1)$, the image of the real line
+in~$\E$.
+\end{enumerate}
+\end{defn}
+
+Is is then the culmination of a long line of research by several authors that
+several well-adapted models of synthetic differential geometry exist. By the
+conditions imposed by Definition~\ref{defn:well-adapted}, for any such
+topos~$\E$ the following transfer principle holds: If~$f,g : \RR \to \RR$ are
+smooth functions, then~$f' = g$ (in the ordinary sense of the derivative) if
+and only if~$i(f)' = i(g)$ in~$\E$ (in the synthetic sense of the derivative).
+
+Hence the nilquadratic infinitesimal numbers of synthetic differential geometry
+may freely be employed as a convenient fiction when computing derivatives.
+Because the theorem on the existence of well-adapted models has a constructive
+proof, any proof making using of these infinitesimals may mechanically be
+unwound to a (longer and more complex) proof which only refers to the ordinary
+reals.
+
+\subsection{On the importance of language}
+
+XXX translation unwieldy, lens, view, ...
 
@@
 * Sven(?)
+* Set[T]
+* free topos
+* semantic view vs. syntactic view
+* Kripke models
 

### 2020-01-14 paper-filmat.tex
Picture of the toposophic landscape

@@
 \usepackage{amsthm,mathtools,stmaryrd,amssymb,graphicx}
+\usepackage{tikz}
+\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing}
 \usepackage{booktabs}
@@
 \newcommand{\xmark}{\ding{55}}
+\definecolor{mypurple}{RGB}{150,0,255}
 
@@
 \begin{abstract}
-  foo
+  XXX
 \end{abstract}
@@
 worthy of mathematical study and in which mathematics plays out slightly
-differently.
+differently (Figure~\ref{fig:landscape}).
+
+\begin{figure}
+  \tikzstyle{topos} = [draw=mypurple, very thick, rectangle, rounded corners, inner sep=5pt, inner ysep=10pt]
+  \tikzstyle{title} = [fill=mypurple, text=white]
+
+  \input{images/primes.tex}
+
+  \newcommand{\drawbox}[4]{
+    \node[topos, #4] [fit = #3] (#1) {};
+    \node[title] at (#1.north) {#2};
+  }
+
+  \newcommand{\muchstuff}{
+    \includegraphics[height=3em]{filmat}
+    \scalebox{0.5}{\sieve{14}{2}}
+  }
+
+  \newcommand{\muchstuffplaceholder}{
+    \includegraphics[height=3em]{filmat-placeholder}
+    \scalebox{0.5}{\fakesieve{14}{2}}
+  }
+
+  \newcommand{\fewstuff}{
+    \includegraphics[height=3em]{filmat}
+    \scalebox{0.5}{\sieve{7}{2}}
+  }
+
+  \newcommand{\threeblobs}{
+    \colorbox{mypurple}{\ \ }\quad
+    \colorbox{mypurple}{\ \ }\quad
+    \colorbox{mypurple}{\ \ }
+  }
+
+  \begin{tikzpicture}
+    \node (objs-set0) at (0,0) {
+      \muchstuff
+    };
+    \node[scale=0.4] (objs-set1) at (-3.5,-2.5) {
+      \fewstuff
+    };
+    \node[scale=0.4] (objs-eff1) at (3.5,-2.5) {
+      \fewstuff
+    };
+    \node[scale=0.4] (objs-sh1)  at (0,-2.5) {
+      \fewstuff
+    };
+
+    \node (prop-set1) [below of=objs-set1, align=left] {
+      The usual laws \\
+      of logic hold.
+    };
+
+    \node (prop-eff1) [below of=objs-eff1, align=left] {
+      Every function \\
+      is computable.
+    };
+
+    \node (prop-sh1) [below of=objs-sh1, align=left] {
+      The axiom of \\
+      choice fails.
+    };
+
+    \node (more-eff1) [below of=prop-eff1] {
+      \threeblobs
+    };
+    \node (more-sh1)  [below of=prop-sh1] {
+      \threeblobs
+    };
+    \node (more-set1) [below of=prop-set1] {
+      \threeblobs
+    };
+
+    \drawbox{set1}{$\mathrm{Set}$}{(objs-set1) (prop-set1) (more-set1)}{}
+    \drawbox{eff1}{Ef{}f}{(objs-eff1) (prop-eff1) (more-eff1)}{tape}
+    \drawbox{sh1}{$\mathrm{Sh}\, X$}{(objs-sh1) (prop-sh1) (more-sh1)}{draw=none}
+    \def\R{8pt}
+    \begin{pgfonlayer}{background}
+    \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple]
+      ($(sh1.south west) + (\R, 0)$) arc(270:180:\R) --
+      ($(sh1.north west) + (0, -\R)$) arc(180:90:\R) --
+      ($(sh1.north east) + (-\R, 0)$) arc(90:0:\R) --
+      ($(sh1.south east) + (0, \R)$) arc(0:-90:\R) --
+      cycle;
+    \end{pgfonlayer}
+    \drawbox{set0}{$\mathrm{Set}$}{(objs-set0) (set1) (eff1) (sh1)}{}
+  \end{tikzpicture}
+
+  \caption{\label{fig:landscape}A glimpse of the toposophic landscape,
+  displaying alongside the standard topos~$\Set$ two further toposes.}
+  % Each topos comes with its own copy of the natural numbers and all the other
+  % familiar mathematical objects.
+  % Toposes are still mathematical structures and hence part of the standard
+  % topos, which is why the standard topos is also drawn to encompass ...
+\end{figure}
+
+% \begin{document}
 
@@
 \begin{enumerate}
-\item Toposes enrich the realism/anti-realism debate in that they paint the
+\item Toposes enrich the realism/anti-realism debate in that they paint the larger
 picture that the platonic heaven of mathematical objects is not unique: besides
@@
 mathematical objects also when viewed through the lens of a specific topos.
+
+\subsection{A glimpse of the toposophic landscape}
 
@@
 
-% \begin{document}
-
 \section{XXX ... infinitesimal numbers ...}
 
-The idea of \emph{infinitesimal numbers} -- numbers which lie between~$-1/n$
-and~$1/n$ for any natural number~$n$ -- has a long and rich history. They are
+The idea of \emph{infinitesimal numbers} -- numbers which lie between~$-\frac{1}{n}$
+and~$\frac{1}{n}$ for any natural number~$n$ -- has a long and rich history. They are
 not part of today's standard setup of the reals, but they are still intriguing
@@
 
-XXX translation unwieldy, lens, view, ...
+XXX translation unwieldy, lens, view, illustrates custom-tailored, ...
 

### 2020-01-14 paper-filmat.tex
Glimpse of the toposophic landscape

@@
 \newcommand{\NN}{\mathbb{N}}
+\newcommand{\TT}{\mathbb{T}}
 \newcommand{\RR}{\mathbb{R}}
@@
 \newcommand{\notnot}{\emph{not~not}\xspace}
+\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}
 
@@
 \subsection{A glimpse of the toposophic landscape}
+XXX explanation of Figure 1
+
+\begin{enumerate}
+\item $\Set$, the \emph{standard topos}. A statement is true in~$\Set$ iff it is true
+in the ordinary mathematical sense.
+\item $\Set_M$ associated to any model~$(M,\in)$ of~\textsc{zf}.
+\item $\Eff$, the \emph{\effective topos}. A statement is true in~$\Eff$ iff
+if it has a \emph{computable witness} as detailed in
+Section~\ref{sect:effective-topos}. In~$\Eff$, any function~$\NN \to \NN$ is
+computable.
+\item $\Sh(X)$, the \emph{topos of sheaves} over any space~$X$. A statement is true
+in~$\Sh(X)$ iff it holds \emph{locally on~$X$}, as detailed in
+Section~\ref{sect:sheaf-toposes}. For most interesting choices of~$X$, the
+axiom of choice and the intermediate value theorem will fail for geometric
+reasons in~$\Sh(X)$.
+\item $\Zar(A)$, the \emph{Zariski topos} of a ring~$A$ presented in
+Section~\ref{sect:smooth}. This topos contains a mirror
+image of~$A$ which is a field, even if~$A$ is not.
+\item $\mathrm{Bohr}(A)$, the \emph{Bohr topos} associated to a noncommutative
+C\textsuperscript{*}\kern-.1ex-algebra~$A$. This topos contains a mirror image
+of~$A$ which is commutative. In this sense, quantum mechanical systems (which are
+described by noncommutative C\textsuperscript{*}\kern-.1ex-algebras) can be
+regarded as classical mechanical systems (which are described by commutative
+algebras)~\cite{butterfield-hamilton-isham:bohr,heunen-landsman-spitters:aqt}.
+\item $\Set[\TT]$, the \emph{classifying topos} of a geometric theory~$\TT$.\footnote{A
+geometric theory is a theory whose axioms can be put as \emph{geometric
+sequents}, sequents of the form~$\varphi \seq{\vec x} \psi$ where~$\varphi$
+and~$\psi$ are geometric formulas (formulas built from equality and specified
+relation symbols by the logical
+connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$ and by arbitrary
+set-indexed disjunctions~$\bigvee$).} This topos contains the
+\emph{generic~$\TT$-model}. For instance, the classifying topos of the theory
+of groups contains the \emph{generic group}. This group has exactly those
+properties which are shared by any group whatsoever. It is arguably this group
+which we implicitly refer to when we utter the phrase ``Let~$G$ be a group.''. XXX
+\item $T(\mathcal{L}_0)$, the \emph{free topos}. A statement is true in the free topos iff it is
+intuitionistically provable. Lambek and Scott proposed that the free topos can
+reconcile moderate platonism (because this topos has a certain universal
+property which can be used to single it out among the plenitude of toposes),
+moderate formalism (because it is constructed in a purely syntactic way) and
+moderate logicism (because, as a topos, it supports an intuitionistic type
+theory)~\cite{lambek:incompatible,couture-lambek:reflections}.
+\end{enumerate}
+
+Constructions, given a topos~$\E$: XXX
+\begin{enumerate}
+\item Given an object~$X$ of~$\E$, the \emph{slice topos}~$\E/X$
+contains a \emph{generic element} of~$X$.
+\item The topos~$\E[\mathbb{O}]$ contains a \emph{generic object}.
+\item There is a ``smallest dense'' subtopos $\Sh_{\neg\neg}(\E)$. This topos
+is always boolean, even if~$\E$ is not. For a mathematician who employs
+intuitionistic logic on their meta level, XXX
+\item XXX ultrapower
+\end{enumerate}
 
 \section{Exploring the \effective topos}
-
-% XXX introduction
-% XXX mention connection to realizability
+\label{sect:effective-topos}
+
+XXX introduction: the objects of the effective topos are data types and the
+morphisms are computable maps.
+
+XXX mention connection to realizability
 
@@
 \section{Exploring toposes of sheaves}
+\label{sect:sheaf-toposes}
 
@@
 
-\section{XXX ... infinitesimal numbers ...}
+\section{Exploring toposes adapted to synthetic differential geometry}
+\label{sect:smooth}
 
@@
 
-* Andrej's realizability in the real world
 * Cauchy vs. Dedekind numbers (physical quantities, ...)
@@
   as an outlook?)
-* enrichment of platonism debate (find better term for this!)
-* uncovering further relations between objects
 * allowing a switch of perspective
@@
 * arbitrariness of the "standard axioms"
-* models of ZF yield toposes (models of ETCS), including a quick discussion of
-  equivalence
 * what giving up classical logic actually amounts to in practice
-* examples in Spec(A), "reifying all the individual localizations into a
-  single coherent entity which can be reasoned about as if it were a single
-  ring"
 * importance of having an adapted language
@@
 * quick remark on the internal language being based on types instead of sets
-* reference to Bohr topos approach
 * Source https://plato.stanford.edu/entries/category-theory/ for relevant
@@
 * Sven(?)
-* Set[T]
-* free topos
 * semantic view vs. syntactic view
 * Kripke models
+* more pointers to topos literature
 

### 2020-01-16 paper-filmat.tex
Suggestions by Matthias

@@
 \newcommand{\NN}{\mathbb{N}}
+\newcommand{\QQ}{\mathbb{Q}}
 \newcommand{\TT}{\mathbb{T}}
@@
 intermediate value theorem from undergraduate calculus fail, toposes in which
-any function~$\RR \to \RR$ is continuous, toposes in which infinitesimal
-numbers exist and toposes whose properties depend on certain circumstances of
-our physical world.
+any function~$\RR \to \RR$ is continuous and toposes in which infinitesimal
+numbers exist.
 
@@
 
-\item We explicate how each topos provides a distinct lens through which the
+\item We explicate how toposes provides distinct lenses through which the
 usual mathematical objects of the standard topos can be viewed.
 \end{enumerate}
+% XXX check everywhere
 
@@
 Turing machine~$M$ which, given the description of such a Turing machine~$X$ as
-input, outputs the description of a similar such Turing machine~$Y$ as output.
+input, outputs the description of a similar such Turing machine~$Y$.
 To compute a rational approximation~$Y(n)$, the machine~$Y$ may simulate~$X$
@@
 recording any twin primes it finds. Then, on day~$\omega$, it can observe
-whether it has found finitely many twins or not.
+whether it has found infinitely many twins or not.
 
@@
 \bigskip
-\paragraph{Machines of the real physical world} A second variant of the
+\paragraph{Machines of the physical world} A second variant of the
 \effective topos is obtained by using machines of the real physical world
@@
 of mathematical studies, but still it is interesting to see which commitments
-about the natural of the physical world imply which internal statements of the
+about the nature of the physical world imply which internal statements of the
 resulting topos.
@@
 the metatheory, this holds if and only if every open is also closed. This is
-for instance satisfied if~$X$ is discrete.
+essentially only satisfied if~$X$ is discrete.
 
@@
 provides several constructive alternatives for realizing infinitesimals.
-One such is ``cheap nonstandard analysis'' by Terence Tao~\cite{tao:cheap-nsa}
-and is related to Robinson's nonstandard analysis like potential infinity is
-related to actual infinity. Instead of appealing to the axiom of choice to
+One such is ``cheap nonstandard analysis'' by Terence Tao~\cite{tao:cheap-nsa}.
+It is to Robinson's nonstandard analysis what potential infinity is
+to actual infinity. Instead of appealing to the axiom of choice to
 obtain a completed ultrafilter, cheap nonstandard analysis constructs larger
@@
 \end{defn}
+% XXX sheaf condition?
 
@@
 synthetic differential geometry, a first indication being that in~$\Zar(\RR)$,
-any function~$R \to R$ is polynomial. Hence important functions such as the
+any function~$R \to R$ is a polynomial function. Hence important functions such as the
 exponential function do not exist in~$\Zar(\RR)$. More comprehensively, the
@@
 category of smooth manifolds into~$\E$.
-\item The ring~$R$ is coincides with~$i(\RR^1)$, the image of the real line
+\item The ring~$R$ coincides with~$i(\RR^1)$, the image of the real line
 in~$\E$.
@@
 Because the theorem on the existence of well-adapted models has a constructive
-proof, any proof making using of these infinitesimals may mechanically be
+proof, any proof making use of these infinitesimals may mechanically be
 unwound to a (longer and more complex) proof which only refers to the ordinary
@@
 * more pointers to topos literature
+* A topos is more than its semantics
+* Check references everywhere (give good pointers)
+* Format rules in a better way
 

### 2020-01-16 paper-filmat.tex
Tackled lots of XXX's

@@
 picture that the platonic heaven of mathematical objects is not unique: besides
-the standard heaven of the standard topos, we could fathom the alternate
+the standard heaven of the standard topos, we can fathom the alternate
 heavens of all other toposes, all embedded in a secord-order heaven.
@@
 be governed by intuitionistic logic.
+% XXX contingent: To some extent, what the commonly agreed-upon rules of
+% mathematics are is dependent on contingent historical factors. It is
+% conceivable, for instance, that the intuitionists ...
 \end{enumerate}
@@
 
-Put briefly, the axioms state that a topos should share several categorical
+Put briefly, these axioms state that a topos should share several categorical
 properties with the category of sets; they ensure that each topos contains its
@@
 membership predicate, where we may write~``$x \in A$'' only in the context of
-some fixed type~$B$ such that~$x$ is of type~$B$ and~$A$ is of
-type~$P(B)$, the power type of~$B$.}
+some fixed type~$M$ such that~$x$ is of type~$M$ and~$A$ is of
+type~$P(M)$, the power type of~$M$.}
 
@@
 
+That said, intuitionistic logic still allows for a considerable amount of
+freedom, and in many toposes statements are true which are baffling if one has
+only received training in classical logic. For instance, on first sight it
+looks like the signum function
+\[ \operatorname{sgn} : \RR \longrightarrow \RR,\ x \longmapsto \begin{cases}
+  -1, & \text{if $x < 0$,} \\
+  0, & \text{if $x = 0$,} \\
+  1, & \text{if $x > 0$,}
+\end{cases} \]
+is an obvious counterexample to the statement ``any map~$\RR \to \RR$ is
+continuous''. However, closer inspection reveals that the signum function
+cannot be proven to be a total function~$\RR \to \RR$ if only intuitionistic
+logic is available. The domain of the signum
+function is the subset~$\{ x \in \RR \,|\, x < 0 \vee x = 0 \vee x > 0 \}
+\subseteq \RR$, and in intuitionistic logic this subset cannot be shown to
+coincide with~$\RR$.
+
+Sections~\ref{sect:effective-topos} to~\ref{sect:smooth} present several
+examples for such anti-classical statements and explain how to make sense of
+them.
+
 \begin{defn}A topos~$\E$ is \emph{boolean} if and only if the laws of classical
@@
 
-The standard topos is boolean if and only if, as is commonly supposed, the laws
+Since exactly those statements hold in the standard topos which hold on the
+meta level, the standard topos is boolean if and only if, as is commonly supposed, the laws
 of classical logic hold on the meta level. Most toposes of interest are not
@@
 which~$M$ believes to be sets, and its morphisms are those entities which~$M$
-believes to be maps. The topos~$\Set_M$ validates the axioms of
-\textsc{etcs}~\cite{XXX}, and models are elementarily equivalent if their
+believes to be maps. The topos~$\Set_M$ validates the axioms of the structural
+set theory \textsc{etcs}~\cite{XXX}, and models are isomorphic if and only if their
 associated toposes are equivalent as categories.}
+% XXX reference
 
@@
 added to the axioms of~\textsc{zf} without causing inconsistency? Yes, if~$M$
-is a model of~\textsc{zf} then~$L^M$, the constructible sets of~$M$ form a
-model of~\textsc{zfc}.  %~\cite{XXX, maybe SEP?}
-
-While toposes can be used for similar such purposes, and indeed have been,
-especially to explore the various intuitionistic notions of sets, an important
-aspect of topos theory is that toposes are used to explore the standard
+is a model of~\textsc{zf} then~$L^M$, the structure of the constructible sets of~$M$, forms a
+model of~\textsc{zfc}.
+
+Toposes can be used for similar such purposes, and indeed have been,
+especially to explore the various intuitionistic notions of sets. However, an important
+% XXX reference: Fourman for IZF? Gambini?
+aspect of topos theory is that toposes are used to explore the \emph{standard}
 mathematical universe: truth in the \effective topos tells us what is
-computable; truth in sheaf toposes tells us what's true locally; toposes
+computable; truth in sheaf toposes tells us what is true locally; toposes
 adapted to synthetic differential geometry can be used to rigorously work with
@@
 \subsection{A glimpse of the toposophic landscape}
-XXX explanation of Figure 1
+There is a proper class of toposes. Figure~\ref{fig:landscape} depicts three
+toposes side by side: the standard topos, a sheaf topos and the \effective
+topos. Each of these toposes tells a different story of mathematics, and any
+topos which is not the standard topos invites us to ponder alternative ways how
+mathematics could unfold.
 
 \begin{enumerate}
+\item The \emph{trivial topos}. In the trivial topos, any statement whatsoever
+is true. The trivial topos is not interesting on its own, but its existence
+streamlines the theory and it can be an interesting question whether a given
+topos coincides with the trivial topos.
 \item $\Set$, the \emph{standard topos}. A statement is true in~$\Set$ iff it is true
 in the ordinary mathematical sense.
-\item $\Set_M$ associated to any model~$(M,\in)$ of~\textsc{zf}.
+\item $\Set_M$, the topos associated to any model~$(M,\in)$ of~\textsc{zf}.
 \item $\Eff$, the \emph{\effective topos}. A statement is true in~$\Eff$ iff
@@
 in~$\Sh(X)$ iff it holds \emph{locally on~$X$}, as detailed in
-Section~\ref{sect:sheaf-toposes}. For most interesting choices of~$X$, the
-axiom of choice and the intermediate value theorem will fail for geometric
-reasons in~$\Sh(X)$.
+Section~\ref{sect:sheaf-toposes}. For most choices of~$X$, the
+axiom of choice and the intermediate value theorem fail in~$\Sh(X)$, and this
+failure is for geometric reasons.
 \item $\Zar(A)$, the \emph{Zariski topos} of a ring~$A$ presented in
@@
 \item $\Set[\TT]$, the \emph{classifying topos} of a geometric theory~$\TT$.\footnote{A
-geometric theory is a theory whose axioms can be put as \emph{geometric
+geometric theory is a theory in many-sorted first-order logic whose axioms can be put as \emph{geometric
 sequents}, sequents of the form~$\varphi \seq{\vec x} \psi$ where~$\varphi$
@@
 \emph{generic~$\TT$-model}. For instance, the classifying topos of the theory
-of groups contains the \emph{generic group}. This group has exactly those
-properties which are shared by any group whatsoever. It is arguably this group
-which we implicitly refer to when we utter the phrase ``Let~$G$ be a group.''. XXX
+of groups contains the \emph{generic group}. Arguably it is this group
+which we implicitly refer to when we utter the phrase ``Let~$G$ be a group.''.
+The generic group has has exactly those
+properties which are shared by any group whatsoever.\footnote{More precisely,
+this is only true for those properties which can be formulated as geometric
+sequents. For arbitrary properties~$\varphi$, the statements~``the generic
+group has property~$\varphi$'' and~``all groups have property~$\varphi$'' need
+not be equivalent. This situation is explored
+in~\cite{blechschmidt:nullstellensatz}.}
+% FUTURE: Give Set[T] its own section.
 \item $T(\mathcal{L}_0)$, the \emph{free topos}. A statement is true in the free topos iff it is
@@
 
-Constructions, given a topos~$\E$: XXX
+There are several constructions which produce a new toposes from a given
+topos~$\E$. A non-exhaustive list is the following.
 \begin{enumerate}
 \item Given an object~$X$ of~$\E$, the \emph{slice topos}~$\E/X$
-contains a \emph{generic element} of~$X$.
-\item The topos~$\E[\mathbb{O}]$ contains a \emph{generic object}.
+contains a \emph{generic element}~$x_0$ of~$X$. This generic element can be
+pictured as the element we implicitly refer to when we utter the
+phrase~``Let~$x$ be an element of~$X$.''. A statement~$\varphi(x_0)$
+about~$x_0$ is true in~$\E/X$ if and only if in~$\E$ the statement~$\forall x
+\? X\_ \varphi(x)$ is true.
+
+For instance, the topos~$\Set/\QQ$ contains the generic rational number~$x_0$.
+Neither the statement~``$x_0$ is zero'' nor the statement~``$x_0$ is not zero''
+hold in~$\Set/\QQ$, as it is neither the case that any rational number
+in~$\Set$ is zero nor that any rational number in~$\Set$ is not zero. Like any
+rational number, the number~$x_0$ can be written as a fraction~$\frac{a}{b}$,
+where just as~$x_0$ itself, the numbers~$a$ and~$b$ are quite indetermined.
+\item Given a statement~$\varphi$ (which may contain objects of~$\E$ as
+parameters but which must be formalizable as a geometric sequent), there is a
+largest subtopos of~$\E$ in which~$\varphi$ holds. This construction is useful
+if neither~$\varphi$ nor~$\neg\varphi$ hold in~$\E$ and we want to
+force~$\varphi$ to be true. If~$\E \models \neg\varphi$, then the resulting
+topos is the trivial topos.
+% FUTURE: The topos~$\E[\mathbb{O}]$ contains a \emph{generic object}.
 \item There is a ``smallest dense'' subtopos $\Sh_{\neg\neg}(\E)$. This topos
-is always boolean, even if~$\E$ is not. For a mathematician who employs
-intuitionistic logic on their meta level, XXX
-\item XXX ultrapower
+is always boolean, even if~$\E$ and the meta level are not. For a mathematician who employs
+intuitionistic logic on their meta level, non-constructive results by their
+classical colleagues do not appear to make sense in~$\Set$, but they hold
+in~$\Sh_{\neg\neg}(\Set)$. If classical logic holds on the meta level,
+then~$\Set$ and~$\Sh_{\neg\neg}(\Set)$ coincide.
+
+The topos~ $\Sh_{\neg\neg}(\E)$ is related to the \emph{double negation
+translation}~$\varphi \mapsto \varphi^{\neg\neg}$ from classical logic into
+intuitionistic logic: A statement~$\varphi$ holds in~$\Sh_{\neg\neg}(\E)$ if
+and only if~$\varphi^{\neg\neg}$ holds in~$\E$. XXX reference
+% FUTURE: ultrapowers, gluing
 \end{enumerate}
+
+Toposes are still mathematical structures, and as long as we study
+toposes within the usual setup of mathematics, our toposes are all part of the
+standard topos. This is why Figure~\ref{fig:landscape} pictures the standard
+topos twice. Hence the toposes which we can study in mathematics do not tell us
+all possible stories how mathematics could unfold, only those which appear
+coherent from the point of view of the standard topos, and the topos-theoretic
+multiverse which we have access to is just a small part of an even larger
+landscape.\footnote{This paragraph employs an overly narrow conception of
+``mathematics'', for instance excluding any predicative flavors of mathematics.
+Toposes are impredicative in the sense that any object of a topos has a
+powerobject. A predicative cousin of toposes are the \emph{arithmetic
+universes} introduced by Joyal which have recently been an important object of
+consideration by Maietti and
+Vickers~\cite{maietti:au,maietti-vickers:induction,vickers:sketches}.}
+
+To obtain just a hint of how the true landscape looks like, we can study topos
+theory from the inside of toposes; the resulting picture can look quite
+different than the picture which emerges from within the standard topos.
+
+For instance, from within the standard topos, we can write down the
+construction which yields the standard topos and the construction which yields
+the \effective topos~$\Eff$ and observe that the resulting toposes are not at
+all equivalent: If~$\Eff$, any function~$\RR \to \RR$ is continuous
+while~$\Set$ abounds with discontinuous functions (at least assuming a
+classical meta level). In contrast, if we carry out these two constructions
+from within the \effective topos, we obtain toposes which are elementarily
+equivalent. More precisely, for any statement~$\varphi$ of higher order
+arithmetic,
+\[ \Eff \models (\Set \models \varphi) \qquad\text{iff}\qquad\Eff \models
+  (\Eff \models \varphi). \]
+In this sense the construction which yields the \effective topos is
+\emph{idempotent}.
 
@@
 
-By the Kripke--Joyal semantics, saying that it's true in the \effective topos
-amounts to saying that there is a Turing machine which, given a natural
+By the Kripke--Joyal semantics, stating that this statement is true in the \effective topos
+amounts to stating that there is a Turing machine which, given a natural
 number~$n$ as input, terminates with a correct judgment whether~$n$ is prime or
@@
 
+% XXX external ...
 Its external meaning when interpreted in the \effective topos is that there exists
@@
 Its meaning when interpreted in the \effective topos is that there exists a
-Turing machine~$M$ which, given the description of a Turing machine~$F$ which
+Turing machine~$M$ which, when given the description of a Turing machine~$F$ which
 computes a function~$f : \NN \to \NN$ as input, terminates with a correct
@@
 
-A formal proof that such a machine~$M$ does not exist will reduce its assumed
-existence to the undecidability of the halting problem. Intuitively, the issue
-is the following. Turing machines are able to simulate other Turing machines.
-Hence~$M$ could simulate~$F$ on various inputs to search the list of
+Intuitively, the issue
+is the following. Turing machines are able to simulate other Turing machines,
+hence~$M$ could simulate~$F$ on various inputs to search the list of
 function values~$f(0), f(1), \ldots$ for a nonzero number. In case that after
@@
 
+A rigorous proof that such a machine~$M$ does not exist reduce its assumed
+existence to the undecidability of the halting problem.
+
 \subsection{``Any function~$\NN \to \NN$ is computable by a Turing machine.''}
-The preceding examples could give the impression that what is true in the
-\effective topos is simply a subset of what is true in the standard topos. This
-statement shows that the relation between these two toposes is more nuanced.
+The preceding examples give the impression that what is true in the
+\effective topos is simply a subset of what is true in the standard topos. The
+example of this subsection shows that the relation between the two toposes is more nuanced.
+
+% XXX explanation computable function
 
@@
 by a Turing machine: The external meaning of this internal statement is that
-there exists a Turing machine~$M$ which, given a description of a Turing
+there exists a Turing machine~$M$ which, when given a description of a Turing
 machine~$F$ computing a function~$f : \NN \to \NN$, outputs a description of a
-Turing machine computing~$f$. It is trivial to program such a machine~$M$; the
+Turing machine computing~$f$. It is trivial to program such a machine~$M$: the
 machine~$M$ simply has to echo its input back to the caller.
 
-To avert a paradox, we should point out where the proof of the fundamental
-observation of computability theory employs nonconstructive reasoning, for if
+To avert a paradox, we should point out where the classical proof of the
+existence of noncomputable functions theory employs nonconstructive reasoning, for if
+% XXX the introduction of this paragraph doesn't match the rest
 it would admit a constructive proof, it would also hold internally to the
@@
 being prominent counterexamples. In the \effective topos, this statement is
-true. A formal proof is not entirely straightforward~\cite{XXX}, but an intuitive
+true. A rigorous proof is not entirely straightforward~\cite{XXX}, but an intuitive
 explanation is as follows.
@@
 required to be \emph{consistent} in the sense that~$|X(n) - X(m)|
-\leq 1/(n+1) + 1/(m+1)$. Intuitively, such a machine~$X$ denotes the real
+\leq 2^{-n} + 2^{-m}$. Intuitively, such a machine~$X$ denotes the real
 number~$\lim_{n \to \infty} X(n)$, and the approximations~$X(n)$ must be
-within~$1/(n+1)$ of the limit.
+within~$2^{-n}$ of the limit.
 
 A function~$f : \RR \to \RR$ in the \effective topos is therefore given by a
-Turing machine~$M$ which, given the description of such a Turing machine~$X$ as
+Turing machine~$M$ which, when given the description of such a Turing machine~$X$ as
 input, outputs the description of a similar such Turing machine~$Y$.
@@
 and can therefore determine arbitrarily many rational approximations~$X(m)$.
-However, within finite time, the machine~$Y$ can only acquire finitely many
+However, within in a finite amount of time, the machine~$Y$ can only learn finitely many
 such approximations. Hence a function such as the signum function, for which
@@
     \begin{minipage}{\textwidth}
-    In the following, we write~``$e \cdot n \downarrow$'' to mean that calling
+    In the following, we write~``$e \cdot n \downarrow$'' to mean that running
     the~$e$-th Turing machine on input~$n$ terminates, and in this case denote
@@
     $e \realizes s = t$ \> iff $s = t$. \\
-    $e \realizes \top$ \> is true for any number~$e$. \\
-    $e \realizes \bot$ \> is false for any number~$e$. \\
+    $e \realizes \top$ \> iff $1 = 1$. \\
+    $e \realizes \bot$ \> iff $1 = 0$. \\
     $e \realizes (\varphi \wedge \psi)$ \> iff~$e \cdot 0 \downarrow$ and~$e
@@
     \varphi$, and \\ \> \qquad if~$e\cdot0 \neq 0$ then~$e\cdot1 \realizes \psi$. \\
-    $e \realizes (\varphi \Rightarrow \psi)$ \> iff for any number~$r$
+    $e \realizes (\varphi \Rightarrow \psi)$ \> iff for any number~$r \in \NN$
     such that~$r \realizes \varphi$, $e \cdot r \downarrow$ and~$e \cdot r \realizes \psi$. \\
-    $e \realizes (\forall n\?\NN\_ \varphi(n))$ \> iff for any natural number~$n_0
+    $e \realizes (\forall n\?\NN\_ \varphi(n))$ \> iff for any number~$n_0
     \in \NN$, $e \cdot n_0 \downarrow$ and~$e \cdot n_0 \realizes \varphi(n_0)$. \\
     $e \realizes (\exists n\?\NN\_ \varphi(n))$ \> iff~$e\cdot0 \downarrow$ and~$e\cdot1 \downarrow$
-    terminate and~$e\cdot1 \realizes \varphi(e\cdot0)$. \\
+    and~$e\cdot1 \realizes \varphi(e\cdot0)$. \\
     $e \realizes (\forall f\?\NN^\NN\_ \varphi(f))$ \> iff for any function~$f_0
@@
     $e \cdot r_0 \downarrow$ and~$e \cdot r_0 \realizes \varphi(f_0)$. \\
-    $e \realizes (\exists f\?\NN\_ \varphi(f))$ \> iff~$e \cdot 0 \downarrow$
-    and~$e \cdot 1 \downarrow$ such that
+    $e \realizes (\exists f\?\NN^\NN\_ \varphi(f))$ \> iff~$e \cdot 0 \downarrow$
+    and~$e \cdot 1 \downarrow$ and
     the $(e \cdot 0)$-th Turing machine \\ \> \qquad computes a function~$f_0 : \NN \to \NN$
@@
 
-In the realizability topos made using infinite-time Turing machines, the full
+In the realizability topos constructed using infinite-time Turing machines, the full
 law of excluded middle still fails, but some instances which are wrong in the
-\effective topos do hold in this topos. For instance, the statement ``any
+\effective topos do hold in this topos. For instance, the instance ``any
 function~$\NN \to \NN$ is the zero function or not'' does: Its external meaning
-is that there is an infinite-time Turing machine~$M$ which, given the
+is that there is an infinite-time Turing machine~$M$ which, when given the
 description of an infinite-time Turing machine~$F$ computing a function~$f :
@@
 
-This realizability topos provides an intriguing environment challenging many
+The realizability topos given by infinite-time Turing machines provides an intriguing environment challenging many
 mathematical intuitions shaped by classical logic. For instance, while from the
@@
 course leave the realm of mathematics, as real-world machines are not objects
-of mathematical studies, but still it is interesting to see which commitments
+of mathematical study, but still it is interesting to see which commitments
 about the nature of the physical world imply which internal statements of the
@@
 
-Namely, one can show that~$\Sh(X) \models \neg\neg\varphi$ is equivalent to the
+Namely, it is an instructive exercise that~$\Sh(X) \models \neg\neg\varphi$ is equivalent to the
 existence of a \emph{dense open}~$U$ of~$X$ such that~$U \models \varphi$.
@@
 topos. If mathematics within~$\Sh(X)$ can be described as ``mathematics
-over~$X$'', then this observation justifies saying that ``ordinary mathematics
+over~$X$'', then this observation justifies the slogan that ``ordinary mathematics
 is mathematics over the point''.
@@
 \subsection{Real numbers}
-As detailled in Section~\ref{sect:eff-continuous}, what the \effective topos believes to be a real number is
+As detailed in Section~\ref{sect:eff-continuous}, what the \effective topos believes to be a real number is
 actually a Turing machine computing arbitrarily-good consistent rational
@@
 positive, generally fails in~$\Sh(X)$. By the Kripke--Joyal semantics, the external
+% XXX external
 meaning of this internal statement is that for any continuous function~$a : U
@@
 
-The weaker statement that for any real number~$a$ it's \notnot the case that~$a < 0$
+The weaker statement that for any real number~$a$ it is \notnot the case that~$a < 0$
 or~$a = 0$ or~$a > 0$ does hold in~$\Sh(X)$, for this statement is an
@@
 
-The internal statement that~$f(-1) < 0$ means that~$f_x(-1) < 0$ for all~$x \in
+The internal statement~``$f(-1) < 0$'' means that~$f_x(-1) < 0$ for all~$x \in
 X$, and similarly so for being positive. More generally, if~$a$ and~$b$ are
 continuous functions~$X \to \RR$ (hence real numbers from the internal point of
-view), the internal statement~$f(a) < b$ means that~$f_x(a(x)) < b(x)$ for
+view), the internal statement~``$f(a) < b$'' means that~$f_x(a(x)) < b(x)$ for
 all~$x \in X$.
 
-The internal statement that~$f$ possesses a zero, that is, that there exists a
-number~$a$ such that~$f(a) = 0$, means that all the functions~$f_x$ each
+The internal statement~``$f$ possesses a zero'', that is ``there exists a
+number~$a$ such that~$f(a) = 0$'', means that all the functions~$f_x$ each
 possess a zero and that moreover, these zeros can locally be picked in a
 continuous fashion. More precisely, this statement means that there is an open
-covering~$X = \bigcup_i U_i$ such that, for each~$i$, there is a continuous
+covering~$X = \bigcup_i U_i$ such that, for each index~$i$, there is a continuous
 function~$a : U_i \to \RR$ such that~$f_x(a(x)) = 0$ for all~$x \in U_i$. (On
@@
 From these observations we can deduce that the intermediate value theorem of
-basic analysis does in general not hold in~$\Sh(X)$ and hence does not allow
+undergraduate calculus does in general not hold in~$\Sh(X)$ and hence does not allow
 for an intuitionistic proof. This theorem states: ``If~$f : \RR \to \RR$
@@
 that in any continuous family~$(f_x)_x$ of continuous functions with~$f_x(-1) <
-0$ and~$f_x(1) > 0$ for all~$x \in X$, it's locally possible to pick zeros of
+0$ and~$f_x(1) > 0$ for all~$x \in X$, it is locally possible to pick zeros of
 the family in a continuous fashion. Figure~\ref{fig:ivt} shows a counterexample
 to this claim.
+% XXX create figure
 
@@
 and their inverses, transfinite numbers, and support a powerful \emph{transfer
-principle}: Any statement which doesn't refer to standardness is true for the
+principle}: Any statement which does not refer to standardness is true for the
 hyperreals if and only if it is true for the ordinary reals.
@@
 
+% XXX remark for arbitrary rings
+% XXX table Zariski rules
 \subsection{The Zariski topos} The starting point is the observation that
-while the field~$\RR$ of ordinary real numbers doesn't contain infinitesimals
+while the field~$\RR$ of ordinary real numbers does not contain infinitesimals
 (except for zero), the ring~$\RR[\varepsilon]/(\varepsilon^2)$ of \emph{dual
@@
 However, the dual numbers are severely lacking in other aspects. Firstly, they
-don't contain any nilcubic numbers which are not already nilsquare. These are
+do not contain any nilcubic numbers which are not already nilsquare. These are
 required in order to extend calculations like~\eqref{eq:derivative} to second
 derivatives, as in
-\[ (x+\varepsilon)^3 - x^3 = x^3 + 3x^2\varepsilon + \frac{1}{2!} 6x
+\[ (x+\varepsilon)^3 - x^3 = 3x^2\varepsilon + \frac{1}{2!} 6x
 \varepsilon^2. \]
@@
 within~$\Zar(\RR)$, it is not the case that zero is the only nilsquare number.
-However, this doesn't mean that there actually \emph{is} a nilsquare number
+However, this does not mean that there actually \emph{is} a nilsquare number
 in~$R$. In fact, any nilsquare number cannot be nonzero, as nonzero numbers are
@@
 * more pointers to topos literature
+* Reference for the effective topos
+* Reference idempotency
 * A topos is more than its semantics
+* Reference: Moerdijk
 * Check references everywhere (give good pointers)
 * Format rules in a better way
+* Format examples in Eff in a better way
 

### 2020-01-16 paper-filmat.tex
typo (thanks to Moritz L)

@@
 the standard heaven of the standard topos, we can fathom the alternate
-heavens of all other toposes, all embedded in a secord-order heaven.
+heavens of all other toposes, all embedded in a second-order heaven.
 \item Mathematics is not only about studying mathematical objects, but also

### 2020-01-16 paper-filmat.tex
to do

@@
 * Format examples in Eff in a better way
+* toposes are not small (Figure 1)
 

### 2020-01-18 paper-flabby-objects.tex
typo

@@
 such that~$K \subseteq \{ x_K \}$. This element might not depend extensionally
-on~$K$, but this fine point is not important for this discussion. Let~$f : X \to X$
+on~$K$, but this fine point is not important for the present discussion. Let~$f : X \to X$
 be a morphism. We construct the self-referential subset~$K
@@
 It is the self-referentiality which makes the proof of
-Proposition~\ref{prop:flabby-effective-sets} work, but the blame for paucity
+Proposition~\ref{prop:flabby-effective-sets} work, but the blame for the paucity
 of flabby objects in the \effective topos is to put on the realizers for

### 2020-01-18 paper-filmat.tex
Introduction to the effective topos

@@
 \newcommand{\effective}{ef{}fective\xspace}
+\newcommand{\effectively}{ef{}fectively\xspace}
 \newcommand{\?}{\,{:}\,}
@@
 
-\section{Exploring the \effective topos}
+\section{The \effective topos, a convenient home for computability theory}
 \label{sect:effective-topos}
 
-XXX introduction: the objects of the effective topos are data types and the
-morphisms are computable maps.
+A basic question in computability theory is: Which computational tasks are
+solvable in principle by computer programs? For instance, there is an algorithm
+for computing the greatest common divisor of any pair of natural numbers, and
+hence we say ``any pair of natural numbers \emph{\effectively} has a greatest
+common divisor''  or ``the function~$\NN \times \NN \to \NN,\,(n,m) \mapsto
+\operatorname{gcd}(n,m)$ is \emph{computable}''.
+
+In such questions of computability, practical issues such as resource
+constraints and hardware malfunctions are ignored; we work with the theoretical
+notion of \emph{Turing machines}, a mathematical abstraction of the computers
+of the real world.
+
+A basic observation in computability theory is that there are computational tasks
+which are not solvable even for these idealized Turing machines. The paramount
+example is the \emph{halting problem}: Given a turing machine~$M$, determine
+whether~$M$ terminates (comes to a stop after having carried out a finite
+number of computational steps) or not.
+
+A Turing machine~$H$ which would solve this problem, that is read the
+description of a Turing machine~$M$ as input and output one or zero depending
+on whether~$M$ terminates or not, would be called a \emph{halting oracle}, and
+a basic result is that there are no halting oracles. If we fix some \effective
+enumeration of all Turing machines, then we can express the undecidability of
+the halting problem also by saying that the \emph{halting function}
+\[ h : \NN \longrightarrow \NN,\ n \longmapsto \begin{cases}
+  1, & \text{if the~$n$-th Turing machine terminates}, \\
+  0, & \text{otherwise,}
+\end{cases} \]
+is not computable.
+
+The \emph{\effective topos}~$\Eff$ is a convenient home for computability
+theory. A statement is true in~$\Eff$ if and only if it has a \emph{computable
+witness}. For instance, a computable witness of a statement of the
+form~``$\forall x\_ \exists y\_ \varphi(x,y)$'' is a Turing machine which, when
+given an input~$x$, computes an output~$y$ together with a computable witness
+for~$\varphi(x,y)$.
+
+Section~\ref{sect:eff-examples} presents several examples to convey an
+intuitive understanding of truth in the \effective topos; the precise
+translation rules are displayed in Table~\ref{table:eff}. A precise definition
+of the \effective topos requires notions of category theory which we do not
+want to suppose here; it is included only for reference.
+
+\begin{defn}\begin{enumerate}
+\item An \emph{assembly} is a set~$X$ together with a
+relation~$({\realizes_X}) \subseteq \NN \times X$ such that for every element~$x
+\in X$, there is a number~$n$ such that~$n \realizes_X x$.
+\item A \emph{morphism of
+assemblies}~$(X,{\realizes_X}) \to (Y,{\realizes_Y})$ is a map~$f : X \to Y$
+which is \emph{tracked} by a Turing machine, that is for which there exists a
+Turing machine~$M$ such that for any element~$x \in X$ and any number~$n$ such
+that~$n \realizes x$, the computation~$M(n)$ terminates and~$M(n) \realizes
+f(x)$.\end{enumerate}\end{defn}
+
+A number~$n$ such that~$n \realizes_X x$ is called a \emph{realizer} for~$x$
+and can be pictured as a concrete representation of the abstract element~$x$.
+The \emph{assembly of natural numbers} is the assembly~$(\NN,{=_\NN})$ and the
+\emph{assembly of functions~$\NN \to \NN$} is the
+assembly~$(X,{\realizes})$ where~$X$ is the set of computable functions~$\NN
+\to \NN$ and~$n \realizes f$ if and only if the~$n$-th Turing machine
+computes~$f$.
+
+\begin{defn}The \emph{\effective topos}~$\Eff$ is the ex/reg completion of the
+category of assemblies.\end{defn}
 
 XXX mention connection to realizability
+
+\subsection{Exploring the \effective topos}
+\label{sect:eff-examples}
+Due to its computational nature, truth in the \effective topos is quite
+different from truth in the standard topos. The following examples are XXX
 
@@
 precision in the input~$x$, do not exist in the \effective topos.
-
-\subsection{The formal translation rules} The previous examples were picked to
-convey an intuitive understanding of what statements in the \effective topos
-externally mean, and to showcase a variety of different situations. The
-formal translation rules are given in Table~\ref{table:eff}.
 
@@
 
-\section{Exploring toposes of sheaves}
+\section{Toposes of sheaves, a convenient home for local truth}
 \label{sect:sheaf-toposes}
@@
 
-\section{Exploring toposes adapted to synthetic differential geometry}
+\section{Topos adapted to synthetic differential geometry}
 \label{sect:smooth}

### 2020-01-18 paper-filmat.tex
to do

@@
   Any function $\RR \to \RR$ is continuous. & \xmark & \cmark \\
+  Heyting arithmetic is categorical. & \xmark & \cmark \\
   \bottomrule
@@
 
-* Cauchy vs. Dedekind numbers (physical quantities, ...)
-* Syntactical vs. semantical interpretation
-* phone call analogy
-* detailed explanation of the pretty picture; analogy with "inner models"
+* XXX Cauchy vs. Dedekind numbers (physical quantities, ...)
+* XXX Syntactical vs. semantical interpretation
+* XXX phone call analogy
+* XXX detailed explanation of the pretty picture; analogy with "inner models"
   of set theory
-* quick overview of the several aspects of toposes (maybe at the end,
+* XXX quick overview of the several aspects of toposes (maybe at the end,
   as an outlook?)
-* allowing a switch of perspective
-* applications in mathematical practice
-* int. logic as common denominator
-* arbitrariness of the "standard axioms"
-* what giving up classical logic actually amounts to in practice
-* importance of having an adapted language
-* SDG (infinitesimals are well-studied in philosophy of mathematics)
-* quick remark on the internal language being based on types instead of sets
-* Source https://plato.stanford.edu/entries/category-theory/ for relevant
+* XXX allowing a switch of perspective
+* XXX applications in mathematical practice
+* XXX int. logic as common denominator
+* XXX arbitrariness of the "standard axioms"
+* XXX what giving up classical logic actually amounts to in practice
+* XXX importance of having an adapted language
+* XXX SDG (infinitesimals are well-studied in philosophy of mathematics)
+* XXX quick remark on the internal language being based on types instead of sets
+* XXX Source https://plato.stanford.edu/entries/category-theory/ for relevant
   literatur
-* the internal language as a generalized modal operator
-
-* use the phrase "very rich and very deep" (or similarly) about set theory
-* PA doesn't have models
-* Sven(?)
-* semantic view vs. syntactic view
-* Kripke models
-* more pointers to topos literature
-* Reference for the effective topos
-* Reference idempotency
-* A topos is more than its semantics
-* Reference: Moerdijk
-* Check references everywhere (give good pointers)
-* Format rules in a better way
-* Format examples in Eff in a better way
-* toposes are not small (Figure 1)
+* XXX the internal language as a generalized modal operator
+
+* XXX use the phrase "very rich and very deep" (or similarly) about set theory
+* XXX PA doesn't have models
+* XXX Sven(?)
+* XXX semantic view vs. syntactic view
+* XXX Kripke models
+* XXX more pointers to topos literature
+* XXX Reference for the effective topos
+* XXX Reference idempotency
+* XXX A topos is more than its semantics
+* XXX Reference: Moerdijk
+* XXX Check references everywhere (give good pointers)
+* XXX Format rules in a better way
+* XXX Format examples in Eff in a better way
+* XXX toposes are not small (Figure 1)
 
@@
 Also Giuseppe Cammarata.
-
-Mention forcing

### 2020-01-18 paper-filmat.tex
typo

@@
 
+\tableofcontents
+
 \noindent
@@
 \emph{logos} with a natural numbers object) is a category which has all finite
-limits, is cartesian closed and has a subobject classifier.\end{defn}
+limits, is cartesian closed, has a subobject classifier and contains a natural
+numbers object.\end{defn}
 
@@
 some fixed type~$M$ such that~$x$ is of type~$M$ and~$A$ is of
-type~$P(M)$, the power type of~$M$.}
+type~$P(M)$, the power type of~$M$. We refer to~\cite{XXX} for ways around this
+restriction.}
 
@@
 freedom, and in many toposes statements are true which are baffling if one has
-only received training in classical logic. For instance, on first sight it
+only received training in mathematics based on classical logic. For instance, on first sight it
 looks like the signum function
@@
 examples for such anti-classical statements and explain how to make sense of
-them.
+them. Some toposes are closer to the standard topos and do not validate such
+anti-classical statements:
 
@@
 
+Some of the most prominent toposes are the following.
+
 \begin{enumerate}
@@
 Section~\ref{sect:effective-topos}. In~$\Eff$, any function~$\NN \to \NN$ is
-computable.
+computable, any function~$\RR \to \RR$ is continuous and the countable axiom of
+choice holds (even if it does not on the meta level).
 \item $\Sh(X)$, the \emph{topos of sheaves} over any space~$X$. A statement is true
@@
 in~$\Set$ is zero nor that any rational number in~$\Set$ is not zero. Like any
-rational number, the number~$x_0$ can be written as a fraction~$\frac{a}{b}$,
-where just as~$x_0$ itself, the numbers~$a$ and~$b$ are quite indetermined.
+rational number, the number~$x_0$ can be written as a fraction~$\frac{a}{b}$.
+Just as~$x_0$ itself, the numbers~$a$ and~$b$ are quite indetermined.
 \item Given a statement~$\varphi$ (which may contain objects of~$\E$ as
@@
 is always boolean, even if~$\E$ and the meta level are not. For a mathematician who employs
-intuitionistic logic on their meta level, non-constructive results by their
+intuitionistic logic on their meta level, the nonconstructive results of their
 classical colleagues do not appear to make sense in~$\Set$, but they hold
@@
 ``mathematics'', for instance excluding any predicative flavors of mathematics.
-Toposes are impredicative in the sense that any object of a topos has a
+Toposes are impredicative in the sense that any object of a topos is required to have a
 powerobject. A predicative cousin of toposes are the \emph{arithmetic
@@
 the \effective topos~$\Eff$ and observe that the resulting toposes are not at
-all equivalent: If~$\Eff$, any function~$\RR \to \RR$ is continuous
-while~$\Set$ abounds with discontinuous functions (at least assuming a
+all equivalent: In~$\Eff$, any function~$\RR \to \RR$ is continuous
+while~$\Set$ abounds with discontinuous functions (at least if we assume a
 classical meta level). In contrast, if we carry out these two constructions
@@
 By the Kripke--Joyal semantics, stating that this statement is true in the \effective topos
-amounts to stating that there is a Turing machine which, given a natural
+amounts to stating that there is a Turing machine which, when given a natural
 number~$n$ as input, terminates with a correct judgment whether~$n$ is prime or
@@
 Its external meaning when interpreted in the \effective topos is that there exists
-a Turing machine which, given a natural number~$n$ as input, terminates with a
+a Turing machine which, when given a natural number~$n$ as input, terminates with a
 prime number~$p > n$ as output. Such a Turing machine exists, hence the
@@
 
-A rigorous proof that such a machine~$M$ does not exist reduce its assumed
+A rigorous proof that such a machine~$M$ does not exist reduces its assumed
 existence to the undecidability of the halting problem.
@@
 
-To avert a paradox, we should point out where the classical proof of the
+To avert a paradox, we should point out where the usual proof of the
 existence of noncomputable functions theory employs nonconstructive reasoning, for if
 % XXX the introduction of this paragraph doesn't match the rest
-it would admit a constructive proof, it would also hold internally to the
+it would not go beyond intuitionistic reasoning, it would also hold internally to the
 \effective topos, in contradiction to the fact that it does not. The halting
@@
 \paragraph{Machines of the physical world} A second variant of the
-\effective topos is obtained by using machines of the real physical world
-instead of abstract Turing machines in its construction. In doing so, we of
+\effective topos is obtained by using machines of the physical world
+instead of abstract Turing machines. In doing so, we of
 course leave the realm of mathematics, as real-world machines are not objects
@@
 
-For instance, Andrej Bauer proved that inside this topos any function~$\RR \to
+For instance, Andrej Bauer showed that inside this topos any function~$\RR \to
 \RR$ is continuous if, in the physical world, only finitely many computational
@@
 case~$\Sh(X)$ is equivalent (as categories and hence toposes) to the standard
-topos. If mathematics within~$\Sh(X)$ can be described as ``mathematics
-over~$X$'', then this observation justifies the slogan that ``ordinary mathematics
+topos. To the extent that mathematics within~$\Sh(X)$ can be described as ``mathematics
+over~$X$'', this observation justifies the slogan that ``ordinary mathematics
 is mathematics over the point''.
@@
 % XXX external
-meaning of this internal statement is that for any continuous function~$a : U
+meaning of the internal statement~``$\forall a:\RR\_ a<0 \vee a=0 \vee a>0$'' is that for any continuous function~$a : U
 \to \RR$ defined on any open~$U$ of~$X$, there is an open covering~$U =
@@
 undergraduate calculus does in general not hold in~$\Sh(X)$ and hence does not allow
-for an intuitionistic proof. This theorem states: ``If~$f : \RR \to \RR$
+for an intuitionistic proof. The intermediate value theorem states: ``If~$f : \RR \to \RR$
 is a continuous function such that~$f(-1) < 0$ and~$f(1) > 0$, there exists a
@@
 so-called \emph{standard elements}, and they also contain infinitesimal numbers
-and their inverses, transfinite numbers, and support a powerful \emph{transfer
+and their inverses, transfinite numbers. Additionally, they support a powerful \emph{transfer
 principle}: Any statement which does not refer to standardness is true for the
@@
 
-A practical consequence of this non-constructivity is that it can be hard to
+A practical consequence of this nonconstructivity is that it can be hard to
 unwind proofs which employ hyperreal numbers to direct proofs, and even where
@@
 It is to Robinson's nonstandard analysis what potential infinity is
-to actual infinity. Instead of appealing to the axiom of choice to
+to actual infinity: Instead of appealing to the axiom of choice to
 obtain a completed ultrafilter, cheap nonstandard analysis constructs larger
@@
 
-The following section presents a topos used in \emph{synthetic differential
-geometry}~\cite{kock:sdg,kock:new-methods}. This is a further topos-theoretic
+The following section presents a (variant of a) topos used in \emph{synthetic differential
+geometry}~\cite{kock:sdg,kock:new-methods}. This subject is a further topos-theoretic
 approach to infinitesimals which is suited to illustrate the philosophy of toposes
-as lenses. Its stated goal is to devise a rigorous context in which the
+as lenses. A major motivation for the development of synthetic differential
+geometry was to devise a rigorous context in which the
 writings of Sophus Lie, who freely employed infinitesimals in his seminal
@@
 (a ring whose elements are triples and whose ring operations are defined such
-that~$\langle0,0,1\rangle^3 = 0$) and the second by passing
+that~$\langle0,1,0\rangle^3 = 0$) and the second by passing
 from~$\RR[\varepsilon]/(\varepsilon^2)$
@@
 In a sense, both of these proposed replacements are better \emph{stages} than
-the basic~$\RR[\varepsilon]/(\varepsilon)^2$ or even~$\RR$ itself. However,
+the basic ring~$\RR[\varepsilon]/(\varepsilon)^2$ or even~$\RR$ itself. However,
 similar criticisms can be mounted against any of these better stages, and the
@@
 
-\subsection{Well-adapted models} The Zariski topos of~$\RR$ contains
+\subsection{Well-adapted models} The Zariski topos of~$\RR$ allows to compute with
 infinitesimals in a satisfying manner. However it is not suited as a home for
@@
 
-Is is then the culmination of a long line of research by several authors that
+It is the culmination of a long line of research by several authors that
 several well-adapted models of synthetic differential geometry exist. By the
-conditions imposed by Definition~\ref{defn:well-adapted}, for any such
+conditions imposed in Definition~\ref{defn:well-adapted}, for any such
 topos~$\E$ the following transfer principle holds: If~$f,g : \RR \to \RR$ are
@@
 * XXX toposes are not small (Figure 1)
+* XXX hypercomputation or supercomputation?
 

### 2020-01-19 paper-filmat.tex
Categoricity of arithmetic

@@
 
+\bigskip
+\dotparagraph{``Heyting arithmetic is categorical.''} In addition to the standard
+model~$\NN$, the standard topos contains uncountably many nonstandard models of
+Peano arithmetic (at least if we assume a classical meta level). By a theorem
+of Benno van den Berg and Jaap van Oosten, the situation
+is quite different in the \effective topos~\cite{XXX}:
+\begin{enumerate}
+\item Heyting arithmetic, the intuitionistic cousin of Peano arithmetic, has
+exactly one model up to isomorphism, namely~$\NN$.
+\item In fact, even the finitely axiomatizable subsystem of Heyting arithmetic
+where the induction scheme is restricted to~$\Sigma_1$-formulas has exactly one
+model up to isomorphism, again~$\NN$. As a consequence, Heyting arithmetic is
+finitely axiomatizable.
+\item Peano arithmetic is ``quasi-inconsistent'' in that it does not have any
+models, for any model of Peano arithmetic would also be a model of Heyting
+arithmetic, but the only model of Heyting arithmetic is~$\NN$ and~$\NN$ does
+not validate the theorem ``any Turing machine terminates or does not
+terminate'' of Peano arithmetic.
+\end{enumerate}
+As a consequence, Gödel's completeness theorem fails in the \effective topos:
+In the \effective topos, Peano arithmetic is consistent (because it is
+equiconsistent to Heyting arithmetic, which has a model) but does not have a
+model.
+
+The reason for~(1) is related to the fact that all the nonstandard models of
+Peano arithmetic in the standard topos are noncomputable~\cite{XXX}.
+
 \begin{table}
@@
 * XXX use the phrase "very rich and very deep" (or similarly) about set theory
-* XXX PA doesn't have models
 * XXX Sven(?)

### 2020-01-19 paper-filmat.tex
typo

@@
 \end{tabular}
-
-\subsection{``Any natural number is prime or not.''} Even without knowing what
+% XXX formatting
+
+\newcommand{\dotparagraph}[1]{\noindent\textbf{#1}}
+
+\bigskip
+\dotparagraph{``Any natural number is prime or not.''} Even without knowing what
 a prime number is, one can safely judge this statement to be true in
@@
 
-\subsection{``There are infinitely many primes.''} A first-order formalization
+\bigskip
+\dotparagraph{``There are infinitely many primes.''} A first-order formalization
 of this statement is ``for any natural number~$n$, there is a prime
@@
 
-\subsection{``Any function~$\NN \to \NN$ is the zero function or not.''} More
+\bigskip
+\dotparagraph{``Any function~$\NN \to \NN$ is the zero function or not.''} More
 formally, the statement is
@@
 
-\subsection{``Any function~$\NN \to \NN$ is computable by a Turing machine.''}
+\bigskip
+\dotparagraph{``Any function~$\NN \to \NN$ is computable by a Turing machine.''}
 The preceding examples give the impression that what is true in the
@@
 
-\subsection{``Any function~$\RR \to \RR$ is continuous.''}\label{sect:eff-continuous}
+\bigskip
+\dotparagraph{``Any function~$\RR \to \RR$ is continuous.''}\label{sect:eff-continuous}
 In the standard topos, this statement is plainly false, with the signum and Heaviside functions
@@
 
-* XXX Cauchy vs. Dedekind numbers (physical quantities, ...)
 * XXX Syntactical vs. semantical interpretation
 * XXX phone call analogy
-* XXX detailed explanation of the pretty picture; analogy with "inner models"
-  of set theory
-* XXX quick overview of the several aspects of toposes (maybe at the end,
-  as an outlook?)
+* XXX quick overview of the several aspects of toposes (maybe at the end, as an outlook?)
 * XXX allowing a switch of perspective
@@
 * XXX arbitrariness of the "standard axioms"
-* XXX what giving up classical logic actually amounts to in practice
 * XXX importance of having an adapted language
-* XXX SDG (infinitesimals are well-studied in philosophy of mathematics)
 * XXX quick remark on the internal language being based on types instead of sets
-* XXX Source https://plato.stanford.edu/entries/category-theory/ for relevant
-  literatur
+* XXX Source https://plato.stanford.edu/entries/category-theory/ for relevant literature
 * XXX the internal language as a generalized modal operator
-
 * XXX use the phrase "very rich and very deep" (or similarly) about set theory
@@
 * XXX Format rules in a better way
-* XXX Format examples in Eff in a better way
 * XXX toposes are not small (Figure 1)

### 2020-01-19 paper-filmat.tex
typo

@@
 
-% XXX explanation computable function
-
-The fundamental observation of computability theory is that, in the standard
-topos, there are functions~$\NN \to \NN$ which are not computable by a Turing
-machine. Explicit examples include the \emph{halting
-function}, which maps a number~$n$ to zero or one depending on whether
-the~$n$-th Turing machine (in some fixed enumeration of all Turing machines)
-terminates or not, and the \emph{busy beaver function}. Cardinality arguments
+As recalled above, in the standard topos there are functions~$\NN \to \NN$ which are not computable by a Turing
+machine. Cardinality arguments
 even show that most functions~$\NN \to \NN$ are not computable: There
@@
 In contrast, in the \effective topos, any function~$\NN \to \NN$ is computable
-by a Turing machine: The external meaning of this internal statement is that
+by a Turing machine. The external meaning of this internal statement is that
 there exists a Turing machine~$M$ which, when given a description of a Turing
@@
 existence of noncomputable functions theory employs nonconstructive reasoning, for if
-% XXX the introduction of this paragraph doesn't match the rest
-it would not go beyond intuitionistic reasoning, it would also hold internally to the
-\effective topos, in contradiction to the fact that it does not. The halting
-function~$h : \NN \to \NN$, defined using the case distinction
+the proof would not go beyond intuitionistic reasoning, it would also hold internally to the
+\effective topos, in contradiction to the fact that in the \effective topos
+all functions~$\NN \to \NN$ are computable.
+
+The usual proof sets up the halting function~$h : \NN \to \NN$, defined using
+the case distinction
 \[ h : n \mapsto \begin{cases}
@@
 \end{cases} \]
-cannot be given as a counterexample in the \effective topos since, in the
-\effective topos, it is not actually a total function from~$\NN$ to~$\NN$. It
-is only defined on those numbers~$n$ for which the~$n$-th Turing machine
+and proceeds to show that~$h$ is not computable. However, in the \effective
+topos, this definition does not give rise to a total function from~$\NN$
+to~$\NN$. The actual domain is the subset of those natural numbers~$n$
+for which the~$n$-th Turing machine
 terminates or does not terminate. Assuming the law of excluded middle, this is
-a trivial condition; but intuitionistically, it is not. The definition of the
-busy beaver function requires a similar case distinction and therefore also
-does not give rise to a well-defined counterexample within the \effective
-topos.
+a trivial condition; but intuitionistically, it is not.
 
@@
 * XXX importance of having an adapted language
-* XXX quick remark on the internal language being based on types instead of sets
 * XXX Source https://plato.stanford.edu/entries/category-theory/ for relevant literature

### 2020-01-19 paper-filmat.tex
References to the automatic continuity

@@
 being prominent counterexamples. In the \effective topos, this statement is
-true. A rigorous proof is not entirely straightforward~\cite{XXX}, but an intuitive
+true and independently due to
+Kreisel--Lacombe--Shoenfield~\cite{kreisel-lacombe-shoenfield:cont} and
+Ceĭtin~\cite{ceitin:cont}. A rigorous proof is not entirely
+straightforward (a textbook reference
+is~\cite[Thm.~9.2.1]{longley-normann:higher-order-computability}), but an intuitive
 explanation is as follows.

### 2020-01-19 paper-filmat.tex
References for categoricity of arithmetic

@@
 of Benno van den Berg and Jaap van Oosten, the situation
-is quite different in the \effective topos~\cite{XXX}:
+is quite different in the \effective topos~\cite{berg-oosten:arithmetic}:
 \begin{enumerate}
@@
 
-The reason for~(1) is related to the fact that all the nonstandard models of
-Peano arithmetic in the standard topos are noncomputable~\cite{XXX}.
+The reason for~(1) is related to the fact that no nonstandard model of
+Peano arithmetic in the standard topos is computable~\cite{tennenbaum:models}.
 
@@
 Also Giuseppe Cammarata.
+
+* XXX Markov's principle
+* XXX formal Church--Turing thesis

### 2020-01-19 paper-filmat.tex
Markov's principle in the effective topos

@@
 \bigskip
+\dotparagraph{``Markov's principle holds.''} Markov's principle is the
+following statement:
+\begin{equation}\label{eq:markov}\tag{MP}
+  \forall f \? \NN^\NN\_ \bigl((\neg\neg\exists n\?\NN\_ f(n) = 0)
+  \Longrightarrow \exists n\?\NN\_ f(n) = 0\bigr).
+\end{equation}
+It is an instance of the law of double negation elimination and hence trivially
+true in the standard topos, at least if we subscribe to classical logic on the
+meta level.
+
+The \effective topos inherits Markov's principle from the meta level.
+The statement~``$\Eff \models \text{\eqref{eq:markov}}$'' means that there is a
+Turing machine~$M$ which, when given the description of a Turing machine~$F$
+computing a function~$f : \NN \to \NN$, outputs the description of a Turing
+machine~$S_F$ which, when given a witness of~``$\neg\neg \exists n\?\NN\_ f(n)
+= 0$'', outputs a witness of~``$\exists n\?\NN\_ f(n) = 0$'' (up to trivial
+conversions, a number~$n$ such that~$f(n) = 0$).
+
+By the translation rules listed in Table~\ref{table:eff}, a number~$e$
+realizes~``$\neg\neg \exists n\?\NN\_ f(n) = 0$'' if and only if it is
+\emph{not not} the case that there is some number~$e'$ such that~$e'$
+realizes~``$\exists n\?\NN\_ f(n) = 0$''. Hence, if ``$\exists n\?\NN\_ f(n) =
+0$'' is realized at all, then any number is a witness of~``$\neg\neg
+\exists n\?\NN\_ f(n) = 0$''.
+
+As a consequence, the input given to machine~$S_F$ is entirely uninformative
+and~$S_F$ cannot make direct computational use of it, but it ensures that an
+\emph{unbounded search} will not fail: The machine~$S_F$ can simulate~$F$ in
+order to compute, in turn, the values~$f(0), f(1), f(2), \ldots$, and stop
+with output~$n$ as soon as it determines that some function value~$f(n)$ is zero.
+
+\bigskip
 \dotparagraph{``Heyting arithmetic is categorical.''} In addition to the standard
@@
 
-* XXX Markov's principle
 * XXX formal Church--Turing thesis

### 2020-01-19 paper-filmat.tex
Reference for idempotency of the effective topos

@@
 In this sense the construction which yields the \effective topos is
-\emph{idempotent}.
+\emph{idempotent}~\cite[Section~3.8.3]{oosten:realizability}.
 
@@
 * XXX Reference for the effective topos
-* XXX Reference idempotency
 * XXX A topos is more than its semantics

### 2020-01-19 paper-filmat.tex
Tackled a couple of XXX's

@@
 
-XXX mention connection to realizability
+% FUTURE (in separate notes about the effective topos): Relationship to
+% realizability
 
@@
 Due to its computational nature, truth in the \effective topos is quite
-different from truth in the standard topos. The following examples are XXX
-
+different from truth in the standard topos. This section explores the following
+examples:
+
+\bigskip
+\begin{center}
 \begin{tabular}{lll}
@@
   \midrule
-  Any natural number is prime or not prime. & \cmark{} (trivially so) & \cmark \\
+  Any natural number is prime or not prime. & \cmark{} (trivially) & \cmark \\
   There are infinitely many primes. & \cmark & \cmark \\
-  Any function $\NN \to \NN$ is the zero function or not. & \cmark{} (trivially so) & \xmark \\
-  Any function $\NN \to \NN$ is computable by a Turing machine. & \xmark & \cmark{} (trivially so) \\
+  Any function $\NN \to \NN$ is constantly zero or not. & \cmark{} (trivially) & \xmark \\
+  Any function $\NN \to \NN$ is computable. & \xmark & \cmark{} (trivially) \\
   Any function $\RR \to \RR$ is continuous. & \xmark & \cmark \\
+  Markov's principle holds. & \cmark{} (trivially) & \cmark \\
   Heyting arithmetic is categorical. & \xmark & \cmark \\
@@
 \end{tabular}
-% XXX formatting
+\end{center}
 
@@
 
-% XXX external ...
 Its external meaning when interpreted in the \effective topos is that there exists
-a Turing machine which, when given a natural number~$n$ as input, terminates with a
+a Turing machine~$M$ which, when given a natural number~$n$ as input, terminates with a
 prime number~$p > n$ as output. Such a Turing machine exists, hence the
-statement is true in the \effective topos.
+statement is true in the \effective topos.\footnote{More precisely, the
+machine~$M$ should also output the description of a Turing machine which
+witnesses that~$p$ is prime. However, the statement~``$p$ is prime''
+is~$\neg\neg$-stable (even decidable), and for those statements witnesses are
+redundant.}
 
 \bigskip
-\dotparagraph{``Any function~$\NN \to \NN$ is the zero function or not.''} More
+\dotparagraph{``Any function~$\NN \to \NN$ is constantly zero or not.''} More
 formally, the statement is
@@
 \bigskip
-\dotparagraph{``Any function~$\NN \to \NN$ is computable by a Turing machine.''}
+\dotparagraph{``Any function~$\NN \to \NN$ is computable.''}
 The preceding examples give the impression that what is true in the
@@
 positive, generally fails in~$\Sh(X)$. By the Kripke--Joyal semantics, the external
-% XXX external
 meaning of the internal statement~``$\forall a:\RR\_ a<0 \vee a=0 \vee a>0$'' is that for any continuous function~$a : U
@@
 * XXX hypercomputation or supercomputation?
+* XXX internal/external
 

### 2020-01-19 paper-filmat.tex
Tackled some XXX's

@@
 \usepackage[utf8]{inputenc}
-\usepackage{amsthm,mathtools,stmaryrd,amssymb,graphicx}
+\usepackage{amsthm,mathtools,stmaryrd,amssymb,graphicx,ragged2e}
 \usepackage{tikz}
 \usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing}
-\usepackage{booktabs}
+\usepackage{booktabs,framed}
 \usepackage[all]{xy}
@@
 
-\newcommand{\xra}[1]{\xrightarrow{#1}}
-\newcommand{\XXX}[1]{\textbf{XXX: #1}}
-\newcommand{\aaa}{\mathfrak{a}}
-\newcommand{\bbb}{\mathfrak{b}}
-\newcommand{\mmm}{\mathfrak{m}}
-\newcommand{\I}{\mathcal{I}}
-\newcommand{\J}{\mathcal{J}}
 \newcommand{\E}{\mathcal{E}}
-\newcommand{\F}{\mathcal{F}}
 \newcommand{\B}{\mathcal{B}}
+\newcommand{\BB}{\mathbb{B}}
 \newcommand{\NN}{\mathbb{N}}
@@
 
-\item We explicate how toposes provides distinct lenses through which the
+\item We explicate how toposes provide distinct lenses through which the
 usual mathematical objects of the standard topos can be viewed.
@@
 drafts. XXX conference organizer and participants
+% Todd Lehman for creating parts of Figure~1
 
@@
 some fixed type~$M$ such that~$x$ is of type~$M$ and~$A$ is of
-type~$P(M)$, the power type of~$M$. We refer to~\cite{XXX} for ways around this
-restriction.}
+type~$P(M)$, the power type of~$M$. We refer
+to~\cite{fourman:sheaf-models,streicher:forcizf,awodey-butz-simpson-streicher:bist}
+for ways around this restriction.}
 
@@
 believes to be maps. The topos~$\Set_M$ validates the axioms of the structural
-set theory \textsc{etcs}~\cite{XXX}, and models are isomorphic if and only if their
+set theory \textsc{etcs}~\cite{mclarty:structuralism,marquis:foundations,barton-friedman:structures}, and models are isomorphic if and only if their
 associated toposes are equivalent as categories.}
-% XXX reference
 
@@
 especially to explore the various intuitionistic notions of sets. However, an important
-% XXX reference: Fourman for IZF? Gambini?
 aspect of topos theory is that toposes are used to explore the \emph{standard}
@@
 
-The topos~ $\Sh_{\neg\neg}(\E)$ is related to the \emph{double negation
+The topos~$\Sh_{\neg\neg}(\E)$ is related to the \emph{double negation
 translation}~$\varphi \mapsto \varphi^{\neg\neg}$ from classical logic into
 intuitionistic logic: A statement~$\varphi$ holds in~$\Sh_{\neg\neg}(\E)$ if
-and only if~$\varphi^{\neg\neg}$ holds in~$\E$. XXX reference
+and only if~$\varphi^{\neg\neg}$ holds
+in~$\E$~\cite[Theorem~6.31]{blechschmidt:phd}.
 % FUTURE: ultrapowers, gluing
@@
 
+\begin{rem}Quite surprisingly, there are infinite sets~$X$ for which any flavor of constructive
+mathematics, in particular the kind which is valid in any topos, verifies the
+\emph{omniscience principle}
+\[ \forall f\?\BB^X\_ \bigl((\exists x\?X\_ f(x) = 0) \vee (\forall x\?X\_ f(x)
+= 1)\bigr), \]
+where~$\BB = \{ 0, 1 \}$ is the set of booleans. This is not the case for~$X =
+\NN$, but it is for instance the case for the one-point compactification~$X =
+\NN_\infty$ of the naturals. This phenomenon has been thoroughly explored by
+Martin Escardó~\cite{escardo:omniscience}.\end{rem}
+
 \bigskip
@@
 \effective topos is simply a subset of what is true in the standard topos. The
-example of this subsection shows that the relation between the two toposes is more nuanced.
+example of this subsection, the so-called \emph{formal Church--Turing thesis},
+shows that the relation between the two toposes is more nuanced.
 
@@
 true in the standard topos, at least if we subscribe to classical logic on the
-meta level.
-
-The \effective topos inherits Markov's principle from the meta level.
+meta level. A useful consequence of Markov's principle is that Turing machines
+which do not run forever (that is, which do \emph{not not} terminate) actually
+terminate; this follows by applying Markov's principle to the function~$f : \NN
+\to \NN$ where~$f(n)$ is zero or one depending on whether a given Turing
+machine has terminated within~$n$ computational steps or not.
+
+The \effective topos inherits Markov's principle from the meta level:
 The statement~``$\Eff \models \text{\eqref{eq:markov}}$'' means that there is a
 Turing machine~$M$ which, when given the description of a Turing machine~$F$
-computing a function~$f : \NN \to \NN$, outputs the description of a Turing
+computing a function~$f : \NN \to \NN$ as input, outputs the description of a Turing
 machine~$S_F$ which, when given a witness of~``$\neg\neg \exists n\?\NN\_ f(n)
@@
 As a consequence, the input given to machine~$S_F$ is entirely uninformative
-and~$S_F$ cannot make direct computational use of it, but it ensures that an
-\emph{unbounded search} will not fail: The machine~$S_F$ can simulate~$F$ in
-order to compute, in turn, the values~$f(0), f(1), f(2), \ldots$, and stop
+and~$S_F$ cannot make direct computational use of it. But its existence ensures that an
+\emph{unbounded search} will not fail (and hence succeed, by an appeal to Markov's
+principle on the meta level): The machine~$S_F$ can simulate~$F$ to
+compute the values~$f(0), f(1), f(2), \ldots$ in turn, and stop
 with output~$n$ as soon as it determines that some function value~$f(n)$ is zero.
@@
 \begin{table}
-  \begin{tabbing}
+  \begin{framed}\begin{tabbing}
     $e \models (\forall f\?\NN^\NN\_ \varphi(n))$ \= \kill
@@
     \realizes \varphi$. \\\\
-    \begin{minipage}{\textwidth}
-    In the following, we write~``$e \cdot n \downarrow$'' to mean that running
-    the~$e$-th Turing machine on input~$n$ terminates, and in this case denote
+    \begin{minipage}{1.04\textwidth}
+    A number~$e$ such that~$e \realizes \varphi$ is called a
+    \emph{realizer} for~$\varphi$. It is the precise version of what is called
+    \emph{computational witness} in the main text.
+    In the following, we write~``$e \cdot n \downarrow$'' to mean that
+    the~$e$-th Turing machine terminates on input~$n$, and in this case denote
     the result by~``$e \cdot n$''.\end{minipage} \\\\
@@
     and $e \cdot 1 \realizes \varphi(f_0)$.
-  \end{tabbing}
+  \end{tabbing}\end{framed}
+  \bigskip
 
@@
 \begin{table}
-  \begin{tabbing}
+  \begin{framed}\begin{tabbing}
     $U \models (\forall x\?\RR\_ \varphi(x))$ \= \kill
@@
     or $U_i \models \psi$. \\
-    $U \models (\varphi \Rightarrow \psi)$ \> iff for every open~$V \subseteq U$,
+    $U \models (\varphi \Rightarrow \psi)$ \> iff for any open~$V \subseteq U$,
     $V \models \varphi$ implies~$V \models \psi$. \\
-    $U \models (\forall a\?\RR\_ \varphi(a))$ \> iff for every open~$V
+    $U \models (\forall a\?\RR\_ \varphi(a))$ \> iff for any open~$V
     \subseteq U$ and \\\>\qquad any continuous function~$a_0 : V \to \RR$, $V \models
@@
     \varphi(a_0)$.
-  \end{tabbing}
+  \end{tabbing}\end{framed}
+  \bigskip
 
@@
   \includegraphics[height=3cm]{trichotomy-3}
-  \caption{\label{fig:trichotomy} Three examples for what the topos~$\Sh(X)$
+  \caption{\label{fig:trichotomy} Three examples of what the topos~$\Sh(X)$
   each believes to be a single real number, where the base space~$X$ is the
@@
   positive.}
+\end{figure}
+
+\begin{figure}
+  \includegraphics[height=3cm]{zeros-in-families-still}
+  \caption{\label{fig:ivt} A single member~$f_{x_0}$ of a continuous
+  family~$(f_x)_{x \in X}$ of continuous functions~$f_x : \RR \to \RR$. The
+  parameter space is~$X = [0,1]$ (not shown) and the other functions~$f_x$ are
+  obtained from~$f_{x_0}$ by moving the horizontal plateau further up or further
+  down. The depicted member~$f_{x_0}$ has a unique
+  zero and there is an open neighborhood~$U$ of~$x_0$ on which zeros of the
+  functions~$f_x$, $x \in U$ can be picked continuously. However, there is no
+  such neighborhood of that particular parameter value~$x_1$ for which the
+  horizontal plateau lies on the $x$-axis.}
 \end{figure}
@@
 to this claim.
-% XXX create figure
 
@@
 
-% XXX remark for arbitrary rings
-% XXX table Zariski rules
 \subsection{The Zariski topos} The starting point is the observation that
@@
 The \emph{Zariski topos of~$\RR$},~$\Zar(\RR)$, meets all of these challenges. It
-contains a ring~$R$, the so-called \emph{ring of smooth numbers}, which reifies
+contains a ring~$\RR^\sim$, the so-called \emph{ring of smooth numbers}, which reifies
 the real numbers, the dual numbers, the two proposed better stages and indeed
@@
 
-For instance, universal quantification~``$\forall x \? R$'' not only refers to
+For instance, universal quantification~``$\forall x \? \RR^\sim$'' not only refers to
 all elements of the current stage, but also to any elements of any refinement
 of the current stage. Similarly, negation~``$\neg\varphi$'' does not only mean
-that~$\varphi$ would imply~$1 = 0$ in the current stage, but also that it does
+that~$\varphi$ would imply~$\bot$ in the current stage, but also that it does
 so at any later stage.
@@
 For reference purposes only, we include the precise definition of the Zariski topos.
-\begin{defn}The \emph{Zariski topos of~$\RR$},~$\Zar(\RR)$, is the category of
-functors from finitely presented~$\RR$-algebras to sets. The object~$R$
-of~$\Zar(\RR)$ is the tautologous functor~$A \mapsto A$.
+\begin{defn}The \emph{Zariski topos of~$\RR$},~$\Zar(\RR)$, is a certain full subcategory of
+the category functors from finitely presented~$\RR$-algebras to sets, namely of
+the \emph{Zariski sheaves}. The object~$\RR^\sim$ of~$\Zar(\RR)$ is the tautologous functor~$A \mapsto A$.
 \end{defn}
-% XXX sheaf condition?
+
+\begin{table}
+  \begin{framed}
+    \begin{tabbing}
+      $\Zar(\RR) \models (\forall f\?\NN^\NN\_ \varphi(n))$ \= \kill
+      $\Zar(\RR) \models \varphi$ \> iff $\RR \models \varphi$. \\\\
+      $A \models s = t$ \> iff $s = t$ as elements of~$A$. \\
+      $A \models \top$ \> iff $1 = 1$ in~$A$. \\
+      $A \models \bot$ \> iff $1 = 0$ in~$A$. \\
+      $A \models (\varphi \wedge \psi)$ \> iff~$A \models \varphi$ and~$A \models \psi$. \\
+      $A \models (\varphi \vee \psi)$ \> iff there exists a partition~$1 = f_1 + \cdots + f_n \in A$ such that, \\
+      \> \qquad for each index~$i$, $A[f_i^{-1}] \models \varphi$ or~$A[f_i^{-1}] \models \psi$. \\
+      $A \models (\varphi \Rightarrow \psi)$ \> iff for any finitely
+      presented~$A$-algebra~$B$, \\
+      \> \qquad $B \models \varphi$ implies~$B \models \psi$. \\
+      $A \models (\forall x\?\RR^\sim\_ \varphi(x))$ \> iff for any finitely
+      presented~$A$-algebra~$B$ and \\
+      \> \qquad any element~$x_0 \in B$, $B \models \varphi(x_0)$. \\
+      $A \models (\exists x\?\RR^\sim\_ \varphi(x))$ \> iff there exists a partition~$1 = f_1 + \cdots + f_n \in A$ such that, \\
+      \> \qquad for each index~$i$, there is an element~$x_0 \in A[f_i^{-1}]$ \\
+      \> \qquad such that $A[f_i^{-1}] \models \varphi(x_0)$.
+    \end{tabbing}
+
+    \bigskip
+    \justify
+    A \emph{covering} of an~$\RR$-algebra~$A$ is a family of~$A$-algebras of the
+    form~$(A[f_i^{-1}])_i$ such that~$1 = f_1 + \cdots + f_n \in A$.
+  \end{framed}
+  \bigskip
+
+  \caption{\label{table:zariski} A (fragment of) the Kripke--Joyal translation
+  rules of the Zariski topos~$\Zar(\RR)$.}
+\end{table}
 
 \paragraph{Properties of the smooth numbers}
-As a concrete example, the Kripke--Joyal translation of the statement that~$R$ is a field,
-\[ \Zar(\RR) \models \forall x\?R\_ (\neg(x = 0) \Rightarrow \exists y\?R\_ xy = 1), \]
+As a concrete example, the Kripke--Joyal translation of the statement that~$\RR^\sim$ is a field,
+\[ \Zar(\RR) \models \forall x\?\RR^\sim\_ (\neg(x = 0) \Rightarrow \exists y\?\RR^\sim\_ xy = 1), \]
 is this:
@@
 For any stage~$A$ and any element~$x \in A$, \\
-${\qquad}$ if for any later stage~$B$ of~$A$ \\
-${\qquad\qquad}$ in which~$x = 0$ holds \\
-${\qquad\qquad}$ also~$1 = 0$ holds, \\
-${\qquad}$ then~$A$ can be covered by later stages~$B_i$ such that, for each
-index~$i$, \\
-${\qquad\qquad}$ there is an element~$y \in B_i$ such that~$xy = 1$ in~$B_i$.
+${\qquad}$ for any later stage~$B$ of~$A$, \\
+${\qquad\qquad}$ if for any later stage~$C$ of~$B$ \\
+${\qquad\qquad\qquad}$ in which~$x = 0$ holds \\
+${\qquad\qquad\qquad}$ also~$1 = 0$ holds, \\
+${\qquad\qquad}$ then~$B$ can be covered by later stages~$C_i$ such that, \\
+${\qquad\qquad\qquad}$ for each
+index~$i$, there is an element~$y \in C_i$ with~$xy = 1$ in~$C_i$.
 \end{indentblock}
@@
 presented~$\RR$-algebra) and an element~$x \in A$ which fulfills the stated
-condition be given. Then~$x = 0$ holds in the refinement~$B \defeq A/(x)$.
-Hence~$1 = 0$ holds in~$B$. By elementary algebra, this means that~$x$ is
-invertible in~$A$. Hence the conclusion holds for the singleton covering
-of~$A$ by~$B_1 \defeq A$.
-
-Within~$\Zar(\RR)$, we set~$\Delta \defeq \{ \varepsilon \? R \,|\,
-\varepsilon^2 = 0 \}$. Then~$R$ has the following properties:
+condition be given. Let~$B$ be any later stage of~$A$ (any finitely
+presented~$A$-algebra -- such an algebra is also finitely presented as
+an~$\RR$-algebra). Then~$x = 0$ holds in the refinement~$C \defeq B/(x)$.
+Hence~$1 = 0$ holds in~$C$. By elementary algebra, this means that~$x$ is
+invertible in~$B$. Hence the conclusion holds for the singleton covering
+of~$B$ by~$C_1 \defeq B$.
+
+\begin{rem}The Zariski topos can also be set up with an arbitrary commutative
+ring~$S$ in place of~$\RR$. The resulting topos~$\Zar(S)$ contains a mirror
+image~$S^\sim$ of~$S$, a reification of all finitely presented~$S$-algebras into a
+single entity. The computation we just carried out also applies in this
+more general context and shows that~$S^\sim$ is a field. It is in this sense that the
+topos~$\Zar(S)$ provides a lens through which~$S$ looks like a field.
+
+A small variant of this lens has been used to give a new proof of
+\emph{Grothendieck's generic freeness lemma}, a fundamental theorem in
+algebraic geometry. The new proof uses the lens to reduce to the case
+of fields, where the claim is trivial~\cite[Section~11.5]{blechschmidt:phd}.
+\end{rem}
+
+Within~$\Zar(\RR)$, we set~$\Delta \defeq \{ \varepsilon \? \RR^\sim \,|\,
+\varepsilon^2 = 0 \}$. Then~$\RR^\sim$ has the following properties:
 \begin{enumerate}
-\item Law of cancellation: $\forall x \? R\_ \forall y \? R\_ \bigl((\forall
+\item Law of cancellation: $\forall x \? \RR^\sim\_ \forall y \? \RR^\sim\_ \bigl((\forall
 \varepsilon \? \Delta\_ x\varepsilon = y\varepsilon)
 \Rightarrow x = y\bigr)$
-\item Axiom of micro-affinity: $\forall f \? R^\Delta\_ \exists! a \? R\_
+\item Axiom of micro-affinity: $\forall f \? {\RR^\sim}^\Delta\_ \exists! a \? \RR^\sim\_
 \forall \varepsilon \? \Delta\_ f(\varepsilon) = f(0) + a\varepsilon$
@@
 However, this does not mean that there actually \emph{is} a nilsquare number
-in~$R$. In fact, any nilsquare number cannot be nonzero, as nonzero numbers are
-invertible while nilsquare numbers are not. Hence any nilsquare number in~$R$
+in~$\RR^\sim$. In fact, any nilsquare number cannot be nonzero, as nonzero numbers are
+invertible while nilsquare numbers are not. Hence any nilsquare number in~$\RR^\sim$
 is \emph{not not} zero. This state of affairs is only possible in an
@@
 
-\begin{rem}The ring~$R$ of smooth numbers does not coincide with the Cauchy reals,
+\begin{rem}The ring~$\RR^\sim$ of smooth numbers does not coincide with the Cauchy reals,
 the Dedekind reals or indeed any well-known construction of the reals
-within~$\Zar(\RR)$. This observation explains why~$R$ can satisfy the law of
+within~$\Zar(\RR)$. This observation explains why~$\RR^\sim$ can satisfy the law of
 cancellation even though it is an intuitionistic theorem that the only
@@
 synthetic differential geometry, a first indication being that in~$\Zar(\RR)$,
-any function~$R \to R$ is a polynomial function. Hence important functions such as the
+any function~$\RR^\sim \to \RR^\sim$ is a polynomial function. Hence important functions such as the
 exponential function do not exist in~$\Zar(\RR)$. More comprehensively, the
@@
 \begin{defn}\label{defn:well-adapted}A \emph{well-adapted model} of synthetic
-differential geometry is a topos~$\E$ together with a ring~$R$ in~$\E$ such
+differential geometry is a topos~$\E$ together with a ring~$\RR^\sim$ in~$\E$ such
 that:
 \begin{enumerate}
-\item The ring~$R$ is a field.
-\item The ring~$R$ validates the axiom of micro-affinity and several related
+\item The ring~$\RR^\sim$ is a field.
+\item The ring~$\RR^\sim$ validates the axiom of micro-affinity and several related
 axioms.
@@
 category of smooth manifolds into~$\E$.
-\item The ring~$R$ coincides with~$i(\RR^1)$, the image of the real line
+\item The ring~$\RR^\sim$ coincides with~$i(\RR^1)$, the image of the real line
 in~$\E$.
@@
 It is the culmination of a long line of research by several authors that
-several well-adapted models of synthetic differential geometry exist. By the
+several well-adapted models of synthetic differential geometry exist~\cite{moerdijk-reyes:models}. By the
 conditions imposed in Definition~\ref{defn:well-adapted}, for any such
@@
 * XXX Sven(?)
-* XXX semantic view vs. syntactic view
 * XXX Kripke models
@@
 * XXX Check references everywhere (give good pointers)
-* XXX Format rules in a better way
 * XXX toposes are not small (Figure 1)
-* XXX hypercomputation or supercomputation?
 * XXX internal/external
@@
 
-* XXX formal Church--Turing thesis
+* XXX Mike
+* XXX Hamkins multiverse

### 2020-01-20 paper-filmat.tex
Pointers to introductory literature

@@
 This note touches on these topics and invites further research.
+
+We intend this contribution to be self-contained and do not assume familiarity with
+topos theory or category theory. Readers who would like to learn more details
+are directed to a gentle introduction to topos theory by Tom
+Leinster~\cite{leinster:introduction}.
+Standard references for the internal
+language of toposes include~\cite[Chapter~VI]{moerdijk-maclane:sheaves-logic},
+\cite[Chapter~14]{goldblatt:topoi},
+\cite{caramello:preliminaries}, \cite{streicher:ctcl}, \cite{shulman:categorical-logic},
+\cite[Chapter~6]{borceux:handbook3} and~\cite[Part~D]{johnstone:elephant}.
 
@@
 
+Introductory literature on the \effective topos
+includes~\cite{hyland:effective-topos,oosten:realizability,phoa:effective,bauer:c2c}.
+
 \begin{defn}\begin{enumerate}
@@
 
-Outline:
-
 Stuff that should be mentioned:
@@
 * XXX more pointers to topos literature
-* XXX Reference for the effective topos
 * XXX A topos is more than its semantics
-* XXX Reference: Moerdijk
 * XXX Check references everywhere (give good pointers)
@@
 
-* XXX Mike
 * XXX Hamkins multiverse

### 2020-01-20 paper-filmat.tex
Toposes are usually not small

@@
 standard topos. This is why Figure~\ref{fig:landscape} pictures the standard
-topos twice. Hence the toposes which we can study in mathematics do not tell us
+topos twice.\footnote{There is a fine print to consider. Technically, if we
+work within~\textsc{zf} or its intuitionistic cousin~\textsc{izf}, most toposes
+of interest are proper classes, not sets. In particular~$\Set$ itself is a
+proper class. Hence Figure~\ref{fig:landscape} should not be interpreted as
+indicating that toposes are contained in~$\Set$ as objects, which most are not.
+In this regard toposes are similar to class-sized inner models in set theory.
+We believe that ``our toposes are all part of the standard topos'' is still an
+apt description of the situation.}
+The toposes which we can study in mathematics do not tell us
 all possible stories how mathematics could unfold, only those which appear
@@
 * XXX Check references everywhere (give good pointers)
-* XXX toposes are not small (Figure 1)
 * XXX internal/external

### 2020-01-20 paper-filmat.tex
Two more references

@@
 topos theory or category theory. Readers who would like to learn more details
-are directed to a gentle introduction to topos theory by Tom
-Leinster~\cite{leinster:introduction}.
-Standard references for the internal
+are directed to the survey of category theory by Jean-Pierre
+Marquis~\cite{sep:category-theory} and to a gentle introduction to topos theory
+by Tom Leinster~\cite{leinster:introduction}. Standard references for the internal
 language of toposes include~\cite[Chapter~VI]{moerdijk-maclane:sheaves-logic},
@@
 landscape.\footnote{This paragraph employs an overly narrow conception of
-``mathematics'', for instance excluding any predicative flavors of mathematics.
+``mathematics'', for instance excluding any predicative flavors of mathematics
+(Laura Crosilla's survey~\cite{crosilla:predicative} is an excellent introduction).
 Toposes are impredicative in the sense that any object of a topos is required to have a
@@
 * XXX importance of having an adapted language
-* XXX Source https://plato.stanford.edu/entries/category-theory/ for relevant literature
 * XXX the internal language as a generalized modal operator

### 2020-01-20 paper-filmat.tex
Definition of Zariski sheaves

@@
 \usepackage[utf8]{inputenc}
-\usepackage{amsthm,mathtools,stmaryrd,amssymb,graphicx,ragged2e}
+\usepackage{amsthm,mathtools,amssymb,graphicx,ragged2e,extpfeil}
 \usepackage{tikz}
@@
 
+\newextarrow{\xbigtoto}{{20}{20}{20}{20}}{\bigRelbar\bigRelbar{\bigtwoarrowsleft\rightarrow\rightarrow}}
 \newcommand{\E}{\mathcal{E}}
@@
 \begin{defn}The \emph{Zariski topos of~$\RR$},~$\Zar(\RR)$, is a certain full subcategory of
-the category functors from finitely presented~$\RR$-algebras to sets, namely of
-the \emph{Zariski sheaves}. The object~$\RR^\sim$ of~$\Zar(\RR)$ is the tautologous functor~$A \mapsto A$.
+the category of functors from finitely presented~$\RR$-algebras to sets, namely of
+the Zariski sheaves. Such a functor is a \emph{Zariski sheaf} if and only if,
+for any covering~$(A[f_i^{-1}])_i$ of any finitely presented~$\RR$-algebra~$A$,
+the diagram
+\[ F(A) \longrightarrow \prod_i F(A[f_i^{-1}]) \xbigtoto{} \prod_{j,k}
+F(A[(f_jf_k)^{-1}]) \]
+is a coequalizer diagram. The object~$\RR^\sim$ of~$\Zar(\RR)$ is the tautologous functor~$A \mapsto A$.
 \end{defn}
@@
     \begin{tabbing}
-      $\Zar(\RR) \models (\forall f\?\NN^\NN\_ \varphi(n))$ \= \kill
+      $A \models (\exists x\?\RR^\sim\_ \varphi(x))$ \= \kill
       $\Zar(\RR) \models \varphi$ \> iff $\RR \models \varphi$. \\\\

### 2020-01-20 paper-filmat.tex
Relation to Kripke models

@@
 \item $\Set_M$, the topos associated to any model~$(M,\in)$ of~\textsc{zf}.
+\item $\Set^W$, the category of functors~$(W,\leq) \to \Set$ associated to
+any Kripke model~$(W,\leq)$. A statement is true in this topos iff it is valid
+with respect to the ordinary Kripke semantics of~$(W,\leq)$. This example shows
+that the Kripke--Joyal semantics of toposes generalizes the more familiar
+Kripke semantics.
 \item $\Eff$, the \emph{\effective topos}. A statement is true in~$\Eff$ iff
@@
 * XXX Sven(?)
-* XXX Kripke models
 * XXX more pointers to topos literature

### 2020-01-20 paper-filmat.tex
Comparison with Hamkin's multiverse philosophy

@@
 
+\begin{rem}The picture of a topos-theoretic multiverse is related to Hamkin's
+multiverse view in set theory~\ref{hamkins:multiverse}. In fact, the
+topos-theoretic multiverse can be regarded as an extension of the set-theoretic
+multiverse: While Hamkins proposes to embrace all models of set theory (not
+necessarily all of them not equally), we propose to embrace all toposes (again
+not necessarily all of them not equally). As every model~$M$ of set theory
+gives rise to a topos~$\Set_M$, the set-theoretic multiverse is contained in the
+topos-theoretic one.
+
+However, a central and intriguing feature of the multiverse view in set theory
+has, as of yet, no counterpart in topos theory: namely a systematic study of
+its modal logic with respect to various notions of relations between toposes.\end{rem}
+
 \section{The \effective topos, a convenient home for computability theory}
@@
 Also Giuseppe Cammarata.
-
-* XXX Hamkins multiverse

### 2020-01-20 paper-filmat.tex
Definition of negation

@@
     the~$e$-th Turing machine terminates on input~$n$, and in this case denote
-    the result by~``$e \cdot n$''.\end{minipage} \\\\
+    the result by~``$e \cdot n$''. No separate clause for negation is listed,
+    as~``$\neg\varphi$'' is an abbreviation for~``$(\varphi \Rightarrow
+    \bot)$''.\end{minipage} \\\\
     $e \realizes s = t$ \> iff $s = t$. \\

### 2020-01-20 paper-filmat.tex
Application of NSA in symplectic geometry

@@
 
-However, the realization of this fiction crucically rests on a non-principal
+There is a growing body of research in mathematics which employs hyperreal
+numbers in this sense. To exemplarily cite just one example, a recent
+application of nonstandard analysis in symplectic geometry is due to Oliver
+Fabert~\cite{fabert:non-squeezing,fabert:floer}, who verified an
+infinite-dimensional analogue of the Arnold conjecture.
+
+However, the realization of the fiction of infinitesimal numbers in nonstandard analysis crucically rests on a non-principal
 ultrafilter, whose existence requires principles which go beyond the means of
@@
 unwind proofs which employ hyperreal numbers to direct proofs, and even where
-possible there is no general procedure for doing so.
+possible there is no general procedure for doing so. For instance, Fabert has
+not obtained a direct proof of his result, and not for the lack of trying
+(personal communication).
 
@@
 * XXX use the phrase "very rich and very deep" (or similarly) about set theory
-* XXX Sven(?)
 * XXX more pointers to topos literature

### 2020-01-20 paper-filmat.tex
typo

@@
 middle ($\varphi \vee \neg\varphi$) and the law of double negation elimination
-($\neg\neg\varphi \Rightarrow \varphi$) are not available.
+($\neg\neg\varphi \Rightarrow \varphi$) are not available. It is intuitionistic
+logic and not classical logic which is the common denominator of all toposes.
 
@@
 examples for such anti-classical statements and explain how to make sense of
-them. Some toposes are closer to the standard topos and do not validate such
+them. There are also toposes which are closer to the standard topos and do not validate such
 anti-classical statements:
@@
 
-There are several constructions which produce a new toposes from a given
+There are several constructions which produce new toposes from a given
 topos~$\E$. A non-exhaustive list is the following.
@@
 \item Given an object~$X$ of~$\E$, the \emph{slice topos}~$\E/X$
-contains a \emph{generic element}~$x_0$ of~$X$. This generic element can be
+contains the \emph{generic element}~$x_0$ of~$X$. This generic element can be
 pictured as the element we implicitly refer to when we utter the
@@
 
-Within~$\Zar(\RR)$, we set~$\Delta \defeq \{ \varepsilon \? \RR^\sim \,|\,
-\varepsilon^2 = 0 \}$. Then~$\RR^\sim$ has the following properties:
+Within~$\Zar(\RR)$, we can construct the set~$\Delta \defeq \{ \varepsilon \? \RR^\sim \,|\,
+\varepsilon^2 = 0 \}$ of nilsquare numbers. Then~$\RR^\sim$ has the following properties:
 \begin{enumerate}
@@
 * XXX applications in mathematical practice
-* XXX int. logic as common denominator
 * XXX arbitrariness of the "standard axioms"
@@
 * XXX use the phrase "very rich and very deep" (or similarly) about set theory
-* XXX more pointers to topos literature
 * XXX A topos is more than its semantics

### 2020-01-20 paper-filmat.tex
Beginning of the final section

@@
 machine~$M$ should also output the description of a Turing machine which
-witnesses that~$p$ is prime. However, the statement~``$p$ is prime''
+witnesses that~$p$ is prime and that~$p > n$. However, the statement~``$p$ is prime and~$p > n$''
 is~$\neg\neg$-stable (even decidable), and for those statements witnesses are
@@
 
-\subsection{On the importance of language}
-
-XXX translation unwieldy, lens, view, illustrates custom-tailored, ...
+\section{On the importance of language}
+
+Having explored the alternate universes of several toposes in
+Sections~\ref{sect:effective-topos} to~\ref{sect:smooth}, we are now in a
+position to explain our view of toposes as lenses in more details.
+
+Toposes can be seen in two quite different ways:
+\begin{enumerate}
+\item (semantic view) A topos is an alternate mathematical universe. Any topos
+contains its own stock of mathematical objects. A ``transfer theorem'' relates
+properties of those objects with properties of objects of the standard topos: A
+statement~$\varphi$ about the objects of a topos~$\E$ holds in~$\E$ iff the
+statement~``$\E \models \varphi$'' holds in the standard topos.
+\item (syntactic view) A topos is merely an index to a syntactical translation
+process. Any topos~$\E$ gives rise to a ``generalized modal operator'' which
+turns a statement~$\varphi$ (about ordinary mathematical objects) into the
+statement~``$\E \models \varphi$'' of the same kind (again about ordinary
+mathematical objects).
+\end{enumerate}
+
+For instance, in the semantic view, the \effective topos is an alternative
+universe which contains its own version of the natural numbers. These naturals
+cannot be directly compared with the naturals of the standard topos, for they
+live in distinct universes, but by the transfer theorem they are still linked in a
+nontrivial way: For instance, the statement ``there are infinitely many primes
+in~$\Eff$'' (a statement about natural numbers in~$\Eff$) is equivalent to the statement
+``for any number~$n$, there \emph{\effectively} exists a prime number~$p > n$''
+(a statement about natural numbers and computability in the standard topos).
+
+In the syntactic view, the \effective topos merely provides a coherent way of
+adding the qualification ``\effective'' to mathematical statements, for
+instance turning the statement ``for any number~$n$, there exists a prime number~$p
+> n$'' into ``for any number~$n$, there \emph{\effectively} exists a prime
+number~$p > n$''. Similarly, a sheaf topos~$\Sh(X)$ provides a coherent way for
+turning statements about real numbers and real functions into statements about
+continuous~$X$-indexed families of real numbers and real functions.
+
+The crucial point is that the translation scheme provided by any topos is sound
+with respect to intuitionistic logic. XXX
 
@@
 
-* XXX Syntactical vs. semantical interpretation
 * XXX phone call analogy
@@
 * XXX arbitrariness of the "standard axioms"
-* XXX importance of having an adapted language
-* XXX the internal language as a generalized modal operator
 * XXX use the phrase "very rich and very deep" (or similarly) about set theory

### 2020-01-21 paper-filmat.tex
typo

@@
 only received training in mathematics based on classical logic. For instance, on first sight it
-looks like the signum function
+looks like the sign function
 \[ \operatorname{sgn} : \RR \longrightarrow \RR,\ x \longmapsto \begin{cases}
@@
 is an obvious counterexample to the statement ``any map~$\RR \to \RR$ is
-continuous''. However, closer inspection reveals that the signum function
+continuous''. However, closer inspection reveals that the sign function
 cannot be proven to be a total function~$\RR \to \RR$ if only intuitionistic
-logic is available. The domain of the signum
+logic is available. The domain of the sign
 function is the subset~$\{ x \in \RR \,|\, x < 0 \vee x = 0 \vee x > 0 \}
@@
 which are not solvable even for these idealized Turing machines. The paramount
-example is the \emph{halting problem}: Given a turing machine~$M$, determine
+example is the \emph{halting problem}: Given a Turing machine~$M$, determine
 whether~$M$ terminates (comes to a stop after having carried out a finite
@@
 \dotparagraph{``Any function~$\RR \to \RR$ is continuous.''}\label{sect:eff-continuous}
-In the standard topos, this statement is plainly false, with the signum and Heaviside functions
+In the standard topos, this statement is plainly false, with the sign and Heaviside functions
 being prominent counterexamples. In the \effective topos, this statement is
@@
 However, within in a finite amount of time, the machine~$Y$ can only learn finitely many
-such approximations. Hence a function such as the signum function, for which
+such approximations. Hence a function such as the sign function, for which
 even rough rational approximations of~$\operatorname{sgn}(x)$ require infinite
@@
 
+% XXX introduction...
+
 Associated to any topological space~$X$ (such as Euclidean space), there is the
@@
 
-However, the realization of the fiction of infinitesimal numbers in nonstandard analysis crucically rests on a non-principal
+However, the realization of the fiction of infinitesimal numbers in nonstandard analysis crucially rests on a non-principal
 ultrafilter, whose existence requires principles which go beyond the means of
@@
 However, the dual numbers are severely lacking in other aspects. Firstly, they
-do not contain any nilcubic numbers which are not already nilsquare. These are
+do not contain any nilcube numbers which are not already nilsquare. These are
 required in order to extend calculations like~\eqref{eq:derivative} to second
@@
 cancellation even though it is an intuitionistic theorem that the only
-nilquadratic number in any flavor of the reals is zero.
+nilsquare number in any flavor of the reals is zero.
 \end{rem}
@@
 
-Hence the nilquadratic infinitesimal numbers of synthetic differential geometry
+Hence the nilsquare infinitesimal numbers of synthetic differential geometry
 may freely be employed as a convenient fiction when computing derivatives.
@@
 The crucial point is that the translation scheme provided by any topos is sound
-with respect to intuitionistic logic. XXX
+with respect to intuitionistic logic. Hence, independent of our position
+towards toposes as alternative universe, we XXX
 

### 2020-01-21 paper-qcoh.tex
to do
typo

@@
 terminology~\cite{caramello:tst} to mean by \emph{geometric theory} a system
+% XXX: The terminology is not due to Olivia, right? Find better expression
 given by a set of sorts, a set of finitary function symbols, a set of finitary
@@
 and the generic model is the functor~$\affl$ of points of the affine line, the
-functor which maps any (l.o.f.p.) scheme~$X$ to~$\Hom_\mathrm{Sch}(X,\mathbb{A}^1) =
+functor which maps any scheme~$X$ (l.o.f.p.) to~$\Hom_\mathrm{Sch}(X,\mathbb{A}^1) =
 \O_X(X)$.
@@
 somewhat vague question ``to which extent does the classifying
-topos~$\Set[\TT]$ realize that it is the classifying topos for~$\TT$?'' is a
-useful guiding principle. This is easiest to visualize with a concrete example
+topos~$\Set[\TT]$ realize that it is the classifying topos for~$\TT$?'' is useful as
+a guiding principle. This is easiest to visualize with a concrete example
 for~$\TT$, such as the theory of rings.
@@
 propositions (again equipped with such a morphism); the object of geometric
-formulas (again so); and the object of geometric sequents (again so). A internal
+formulas (again so); and the object of geometric sequents (again so). An internal
 geometric theory~$\TT$ over~$\Sigma$ is then given by a subobject of the object
@@
 Disjunctions appearing in internal geometric formulas may be indexed by
-arbitrary objects of the topos (that is, disjunctions may be taken over
-arbitrary subobjects of disjuncts), just like disjunctions appearing in ordinary
+arbitrary objects of the topos, just like disjunctions appearing in ordinary
 external geometric formulas over an ordinary signature may be indexed by
@@
 where the disjunction ranges over all those geometric formulas~$\alpha$ in the
-context~$\vec x$ such that for each index~$i$, the theory~$\TT$ proves~$(\alpha
+context~$\vec x$ such that for every index~$i$, the theory~$\TT$ proves~$(\alpha
 \wedge \varphi_i \seq{\vec x, \vec y_i} \psi_i)$.
@@
 formula -- on the semantic side by repeatedly applying
-Scho\-li\-um~\ref{scholium:nullstellensatz-more-specific}, on the semantic side by
+Scho\-li\-um~\ref{scholium:nullstellensatz-more-specific}, on the syntactic side by
 repeatedly applying axiom scheme~\eqref{eq:nullstellensatz}. Hence we are
@@
 of~$\TT$-models (in~$\Set$) on the finitely presentable objects~\cite[Theorem~2.1.21]{caramello:tst}. The
+% XXX: besser formulieren family of objects
 (underlying object of the) generic model is the tautologous functor~$U_\TT : T
@@
 by declaring~$\brak{f}([t_1],\ldots,[t_n]) = [f(t_1,\ldots,t_n)]$ and the
-relation symbols~$S$ are interpreted by declaring~$([t_1],\ldots,[t_n]) \in
-\brak{S} \Leftrightarrow (\TT + R \,\vdash S(t_1,\ldots,t_n))$.
+relation symbols~$H$ are interpreted by declaring~$([t_1],\ldots,[t_n]) \in
+\brak{H} \Leftrightarrow (\TT + R \,\vdash H(t_1,\ldots,t_n))$.
 
@@
 \rangle$ is~$\TT\langle X \amalg X' \,|\, R, R' \rangle$.\end{proof}
+% XXX sagen, dass jedes Modell (beliebig) präsentiert wird
 
@@
 of~$\TT/U_\TT(T)$.
+% XXX: No passing to covers is necessary, as we're in a *pre*sheaf topos!
 
@@
 classifying topos of the theory~$\OO$ of objects, the theory with a single sort and
-no function symbols, relations symbols or axioms, including:
+no function symbols, relation symbols or axioms, including:
 \begin{align*}

### 2020-01-22 paper-filmat.tex
typo

@@
 \end{enumerate}
-This note touches on these topics and invites further research.
+We invite further research on these connections.
 
@@
 
-Formally, a topos is a certain kind of \emph{category}, containing objects and
-morphisms between those objects. The formal definition, recorded here only for
-reference, requires some amount of category theory, but, as will be outlined in
+A topos is a certain kind of \emph{category}, containing objects and
+morphisms between those objects. The precise definition is recorded here only for
+reference. Appreciating it requires some amount of category theory, but, as will be demonstrated in
 the following sections, exploring the mathematical universe of a given topos
@@
 
-\begin{defn}The \emph{standard topos}~$\Set$ is the category which contains all
+\begin{defn}The \emph{standard topos}~$\Set$ is the category which has all
 sets as its objects and all maps between sets as morphisms.\end{defn}
@@
   \E \models \alpha \quad\text{and}\quad \E \models \beta. \]
-The remaining translation rules are more involved; we do not list them here for
+The remaining translation rules are more
+involved~\cite[Section~VI.7]{moerdijk-maclane:sheaves-logic}; we do not list them here for
 the case of a general topos~$\E$, but we will state them in the next sections
@@
 ($\neg\neg\varphi \Rightarrow \varphi$) are not available. It is intuitionistic
-logic and not classical logic which is the common denominator of all toposes.
+logic and not classical logic which is the common denominator of all toposes;
+we cannot generally argue by contradiction in a topos.
 
@@
 and~\cite{bauer:int-mathematics,bauer:video,melikhov:intuitionistic-logic} for
-further references) and hence the alternate universes provided by toposes
+accessible surveys on appreciating intuitionistic logic) and hence the alternate universes provided by toposes
 cannot be too strange: In any topos, there are infinitely many prime numbers,
-the square root of two is not rational and the powerset of the naturals is
-uncountable.
+the square root of two is not rational, the fundamental theorem of Galois
+theory holds and the powerset of the naturals is uncountable.
 
@@
 in~\cite{blechschmidt:nullstellensatz}.}
+% FUTURE: Write just a little more on Set[T], that geom.impl. consequences spread.
 % FUTURE: Give Set[T] its own section.
@@
 landscape.\footnote{This paragraph employs an overly narrow conception of
-``mathematics'', for instance excluding any predicative flavors of mathematics
-(Laura Crosilla's survey~\cite{crosilla:predicative} is an excellent introduction).
+``mathematics'', focusing only on those mathematical worlds which form toposes
+and for instance excluding any predicative flavors of mathematics
+(Laura Crosilla's survey~\cite{crosilla:predicativity} is an excellent introduction).
 Toposes are impredicative in the sense that any object of a topos is required to have a
@@
 In such questions of computability, practical issues such as resource
-constraints and hardware malfunctions are ignored; we work with the theoretical
+constraints or hardware malfunctions are ignored; we employ the theoretical
 notion of \emph{Turing machines}, a mathematical abstraction of the computers
@@
   Heyting arithmetic is categorical. & \xmark & \cmark \\
+  % FUTURE: Countable axiom of choice
   \bottomrule
@@
 first exercise in programming courses. Hence the statement is also true in the
-\effective topos, but for the nontrivial reason that such a machine exists.
+\effective topos, but for the nontrivial reason that primality can be
+algorithmically tested.
 
@@
 \bigskip
-\dotparagraph{``Any function~$\NN \to \NN$ is constantly zero or not.''} More
-formally, the statement is
+\dotparagraph{``Any function~$\NN \to \NN$ is constantly zero or not.''} Precisely,
+the statement is
 \[ \forall f \? \NN^\NN\_
@@
 The preceding examples give the impression that what is true in the
-\effective topos is simply a subset of what is true in the standard topos. The
+\effective topos is solely a subset of what is true in the standard topos. The
 example of this subsection, the so-called \emph{formal Church--Turing thesis},
@@
 Turing machine computing~$f$. It is trivial to program such a machine~$M$: the
-machine~$M$ simply has to echo its input back to the caller.
+machine~$M$ simply has to echo its input back to the user.
 
@@
 Lewis~\cite{hamkins-lewis:ittm}. These machines model \emph{hypercomputation}
-in that they can run for ``longer than infinity''; more precisely, the
+in that they can run for ``longer than infinity''; more precisely, their
 computational steps are indexed by the ordinal numbers instead of the natural
@@
 
-The weaker statement that for any real number~$a$ it is \notnot the case that~$a < 0$
-or~$a = 0$ or~$a > 0$ does hold in~$\Sh(X)$, for this statement is an
+The weaker statement ``for any real number~$a$ it is \notnot the case that~$a < 0$
+or~$a = 0$ or~$a > 0$'' does hold in~$\Sh(X)$, for this statement is an
 intuitionistic tautology. Its meaning is that there exists a
@@
 mathematical formalism closer together.
+% FUTURE: Give references to philosophical accounts of infinitesimal numbers.
 
@@
 In the ``if'' direction, the transfer principle is useful for importing
-knowledge about the ordinary reals in the hyperreal realm. For instance,
+knowledge about the ordinary reals into the hyperreal realm. For instance,
 addition of hyperreals is commutative because addition of reals is.
@@
 In a sense, both of these proposed replacements are better \emph{stages} than
-the basic ring~$\RR[\varepsilon]/(\varepsilon)^2$ or even~$\RR$ itself. However,
+the basic ring~$\RR[\varepsilon]/(\varepsilon^2)$ or even~$\RR$ itself. However,
 similar criticisms can be mounted against any of these better stages, and the
@@
 Within~$\Zar(\RR)$, we can construct the set~$\Delta \defeq \{ \varepsilon \? \RR^\sim \,|\,
-\varepsilon^2 = 0 \}$ of nilsquare numbers. Then~$\RR^\sim$ has the following properties:
+\varepsilon^2 = 0 \}$ of nilsquare numbers. Then~$\RR^\sim$ validates the following laws:
 \begin{enumerate}
@@
 Sections~\ref{sect:effective-topos} to~\ref{sect:smooth}, we are now in a
-position to explain our view of toposes as lenses in more details.
+position to explain our view of toposes as lenses in more detail.
 
@@
 \item (syntactic view) A topos is merely an index to a syntactical translation
-process. Any topos~$\E$ gives rise to a ``generalized modal operator'' which
+procedure. Any topos~$\E$ gives rise to a ``generalized modal operator'' which
 turns a statement~$\varphi$ (about ordinary mathematical objects) into the

### 2020-01-22 paper-filmat.tex
A syntactic account and on the importance of language

@@
 
+\subsection{A syntactic account of toposes} We introduced toposes from a semantic
+point of view. There is also a second, purely syntactic point of view on
+toposes:
+\begin{enumerate}
+\item (semantic view) A topos is an alternate mathematical universe. Any topos
+contains its own stock of mathematical objects. A ``transfer theorem'' relates
+properties of those objects with properties of objects of the standard topos: A
+statement~$\varphi$ about the objects of a topos~$\E$ holds in~$\E$ iff the
+statement~``$\E \models \varphi$'' holds in the standard topos.
+\item (syntactic view) A topos is merely an index to a syntactical translation
+procedure. Any topos~$\E$ gives rise to a ``generalized modal operator'' which
+turns a statement~$\varphi$ (about ordinary mathematical objects) into the
+statement~``$\E \models \varphi$'' of the same kind (again about ordinary
+mathematical objects).
+\end{enumerate}
+
+For instance, in the semantic view, the \effective topos is an alternative
+universe which contains its own version of the natural numbers. These naturals
+cannot be directly compared with the naturals of the standard topos, for they
+live in distinct universes, but by the transfer theorem they are still linked in a
+nontrivial way: For instance, the statement ``there are infinitely many primes
+in~$\Eff$'' (a statement about natural numbers in~$\Eff$) is equivalent to the statement
+``for any number~$n$, there \emph{\effectively} exists a prime number~$p > n$''
+(a statement about natural numbers and computability in the standard topos).
+(The meaning of~\emph{effectivity} will be recalled in
+Section~\ref{sect:effective-topos}.)
+
+In the syntactic view, the \effective topos merely provides a coherent way of
+adding the qualification ``\effective'' to mathematical statements, for
+instance turning the statement ``for any number~$n$, there exists a prime number~$p
+> n$'' into ``for any number~$n$, there \emph{\effectively} exists a prime
+number~$p > n$''. Similarly, a sheaf topos~$\Sh(X)$ provides a coherent way for
+turning statements about real numbers and real functions into statements about
+continuous~$X$-indexed families of real numbers and real functions.
+
+The crucial point is that the translation scheme provided by any topos is sound
+with respect to intuitionistic logic. Hence, regardless of our actual position
+on toposes as alternate universes, working under the lens of a given topos
+\emph{feels like} working in an alternate universe.
+
 \section{The \effective topos, a convenient home for computability theory}
@@
 
-\subsection{The Zariski topos} The starting point is the observation that
+\subsection{The Zariski topos}
+\label{sect:the-zariski-topos}
+The starting point is the observation that
 while the field~$\RR$ of ordinary real numbers does not contain infinitesimals
@@
 
-\paragraph{Properties of the smooth numbers}
+\paragraph{Properties of the smooth numbers}\label{page:field-property}
 As a concrete example, the Kripke--Joyal translation of the statement that~$\RR^\sim$ is a field,
@@
 \Rightarrow x = y\bigr)$
-\item Axiom of micro-affinity: $\forall f \? {\RR^\sim}^\Delta\_ \exists! a \? \RR^\sim\_
+\item Axiom of micro-affinity: $\forall f \? (\RR^\sim)^\Delta\_ \exists! a \? \RR^\sim\_
 \forall \varepsilon \? \Delta\_ f(\varepsilon) = f(0) + a\varepsilon$
@@
 
-\section{On the importance of language}
-
-Having explored the alternate universes of several toposes in
-Sections~\ref{sect:effective-topos} to~\ref{sect:smooth}, we are now in a
-position to explain our view of toposes as lenses in more detail.
-
-Toposes can be seen in two quite different ways:
+\subsection{On the importance of language} The verification of the field
+property of~$\RR^\sim$ in Section~\ref{sect:the-zariski-topos} on
+page~\pageref{page:field-property} demonstrates a basic feature of the
+Kripke--Joyal translation rules: The translation~``$\E \models \varphi$'' of a
+statement~$\varphi$ is usually quite complex, even if~$\varphi$ is reasonably
+perspicuous.
+
+This property is where the language of toposes derives its usefulness for
+mathematical practice from: In some cases, the easiest way to prove a result
+(about objects of the standard topos) is
 \begin{enumerate}
-\item (semantic view) A topos is an alternate mathematical universe. Any topos
-contains its own stock of mathematical objects. A ``transfer theorem'' relates
-properties of those objects with properties of objects of the standard topos: A
-statement~$\varphi$ about the objects of a topos~$\E$ holds in~$\E$ iff the
-statement~``$\E \models \varphi$'' holds in the standard topos.
-\item (syntactic view) A topos is merely an index to a syntactical translation
-procedure. Any topos~$\E$ gives rise to a ``generalized modal operator'' which
-turns a statement~$\varphi$ (about ordinary mathematical objects) into the
-statement~``$\E \models \varphi$'' of the same kind (again about ordinary
-mathematical objects).
+\item to observe that the claim is
+equivalent to the Kripke--Joyal translation of a different (typically more
+perspicuous) claim about objects of some problem-specific relevant topos and
+then
+\item to verify this different claim, reasoning internally to the topos.
 \end{enumerate}
 
-For instance, in the semantic view, the \effective topos is an alternative
-universe which contains its own version of the natural numbers. These naturals
-cannot be directly compared with the naturals of the standard topos, for they
-live in distinct universes, but by the transfer theorem they are still linked in a
-nontrivial way: For instance, the statement ``there are infinitely many primes
-in~$\Eff$'' (a statement about natural numbers in~$\Eff$) is equivalent to the statement
-``for any number~$n$, there \emph{\effectively} exists a prime number~$p > n$''
-(a statement about natural numbers and computability in the standard topos).
-
-In the syntactic view, the \effective topos merely provides a coherent way of
-adding the qualification ``\effective'' to mathematical statements, for
-instance turning the statement ``for any number~$n$, there exists a prime number~$p
-> n$'' into ``for any number~$n$, there \emph{\effectively} exists a prime
-number~$p > n$''. Similarly, a sheaf topos~$\Sh(X)$ provides a coherent way for
-turning statements about real numbers and real functions into statements about
-continuous~$X$-indexed families of real numbers and real functions.
-
-The crucial point is that the translation scheme provided by any topos is sound
-with respect to intuitionistic logic. Hence, independent of our position
-towards toposes as alternative universe, we XXX
+One can always mechanically eliminate the topos machinery from such a proof, by
+translating all intermediate statements following the Kripke--Joyal translation
+rules and unwinding the constructive soundness proof of
+Theorem~\ref{thm:reasoning}. This unwinding typically turns transparent
+internal proofs into complex external proofs -- proofs which one may not have
+found without the problem-adapted internal language provided by a
+custom-tailored topos.
+
+% FUTURE: Cite blechschmidt:generic-freeness?
 

### 2020-01-22 paper-filmat.tex
Prepared for double-blind review
Down to two XXX's

@@
 
-\title[]{Exploring mathematical objects from custom-tailored mathematical universes}
-\author{Ingo Blechschmidt}
-\address{Universität Augsburg \\
-Institut für Mathematik \\
-Universitätsstr. 14 \\
-86159 Augsburg, Germany}
-\email{ingo.blechschmidt@math.uni-augsburg.de}
+%\title[]{Exploring mathematical objects from custom-tailored mathematical universes}
+%\author{Ingo Blechschmidt}
+%\address{Universität Augsburg \\
+%Institut für Mathematik \\
+%Universitätsstr. 14 \\
+%86159 Augsburg, Germany}
+%\email{ingo.blechschmidt@math.uni-augsburg.de}
 
@@
 \newcommand{\RR}{\mathbb{R}}
+\newcommand{\CC}{\mathbb{C}}
 \newcommand{\ZZ}{\mathbb{Z}}
@@
 
-\begin{abstract}
-  XXX
-\end{abstract}
-
-\maketitle
+%\begin{abstract}
+%  FUTURE: abstract
+%\end{abstract}
+
+%\maketitle
 \thispagestyle{empty}
 
-\tableofcontents
+\begin{center}
+  \begin{minipage}{0.6\textwidth}
+  \emph{To the anonymous referees:}
+  \bigskip
+
+  Thank you for your volunteer work. It is very much appreciated. I am looking
+  forward to your comments.
+
+  \hfill -- the anonymous author
+  \end{minipage}
+\end{center}
+\bigskip
 
@@
 
-\begin{figure}
+\begin{figure}[b]
   \tikzstyle{topos} = [draw=mypurple, very thick, rectangle, rounded corners, inner sep=5pt, inner ysep=10pt]
@@
 \end{enumerate}
-% XXX check everywhere
 
@@
 heavens of all other toposes, all embedded in a second-order heaven.
+\item To some extent, the mathematical landscape depends on the commonly agreed-upon rules of
+mathematics. These are not entirely absolute; for instance, it is conceivable
+that from the foundational crisis Brouwer's intuitionism would have emerged as
+the main school of thought and we would now all reject the law of excluded
+middle. Toposes allow us to explore alternatives to how history
+has played out.
 \item Mathematics is not only about studying mathematical objects, but also
@@
 on mathematical objects provided by any topos uncovers relations which
-otherwise remain hidden.
-% XXX: mention Olivia?
+otherwise remain hidden.\footnote{Olivia Caramello's research
+program~\cite{caramello:tst} provides a further topos-theoretic way for
+uncovering hidden relations, though not between objects but between mathematical
+theories.}
 \item In some cases, a mathematical relation can be expressed quite succinctly
@@
 be governed by intuitionistic logic.
-% XXX contingent: To some extent, what the commonly agreed-upon rules of
-% mathematics are is dependent on contingent historical factors. It is
-% conceivable, for instance, that the intuitionists ...
 \end{enumerate}
@@
 \bigskip
-\paragraph{Acknowledgments} We are grateful to XXX for many invaluable
-discussions shaping this work and to XXX for their careful criticism of earlier
-drafts. XXX conference organizer and participants
-% Todd Lehman for creating parts of Figure~1
+\paragraph{Acknowledgments} We are grateful to Andrej Bauer, Martin
+Brandenburg, Sina Hazratpour, Matthias Hutzler, Marc Nieper-Wißkirchen and
+Alexander Oldenziel for invaluable discussions shaping this work, to Moritz
+Laudahn, Matthias Hutzler and Milan Zerbin for their careful criticism of
+earlier drafts and to Todd Lehman for preparing parts of
+Figure~\ref{fig:landscape}. We thank the organizers and all participants of the
+Mussomeli conference of the Italian Network for the Philosophy in Mathematics,
+where this work was presented, for creating an exceptionally beautiful meeting. In
+particular, we are grateful to Neil Barton, Danielle Macbeth, Gianluigi Oliveri
+and Lorenzo Rossi for valuable comments.
+
+\tableofcontents
 
@@
 the case of a general topos~$\E$, but we will state them in the next sections
-for several specific toposes.
+for several specific toposes. We refer to~``$\E \models \varphi$'' also as the
+``external meaning of the internal statement~$\varphi$''.
 
@@
 
-\subsection{The logic of toposes} By their definition as special kinds of
+\subsection{The logic of toposes}\label{sect:logic-of-toposes}
+By their definition as special kinds of
 categories, toposes are merely algebraic structures not unlike groups or vector
@@
 %reasoning}. ...
+
+At this point in the text, all prerequisites for exploring toposes have been
+introduced. The reader who wishes to develop, by explicit examples, intuition
+for working internally to toposes is invited to skip ahead to
+Section~\ref{sect:effective-topos}.
 
@@
 \end{center}
+\bigskip
 
@@
 
-\bigskip
-\dotparagraph{``Any natural number is prime or not.''} Even without knowing what
+\begin{ex}\textbf{``Any natural number is prime or not.''} Even without knowing what
 a prime number is, one can safely judge this statement to be true in
@@
 algorithmically tested.
-
-\bigskip
-\dotparagraph{``There are infinitely many primes.''} A first-order formalization
+\end{ex}
+
+\begin{ex}\textbf{``There are infinitely many primes.''} A first-order formalization
 of this statement is ``for any natural number~$n$, there is a prime
@@
 redundant.}
-
-\bigskip
-\dotparagraph{``Any function~$\NN \to \NN$ is constantly zero or not.''} Precisely,
+\end{ex}
+
+\begin{ex}\textbf{``Any function~$\NN \to \NN$ is constantly zero or not.''} Precisely,
 the statement is
@@
 existence to the undecidability of the halting problem.
+\end{ex}
 
@@
 
-\bigskip
-\dotparagraph{``Any function~$\NN \to \NN$ is computable.''}
+\begin{ex}\textbf{``Any function~$\NN \to \NN$ is computable.''}
 The preceding examples give the impression that what is true in the
@@
 a trivial condition; but intuitionistically, it is not.
-
-\bigskip
-\dotparagraph{``Any function~$\RR \to \RR$ is continuous.''}\label{sect:eff-continuous}
+\end{ex}
+
+\begin{ex}\textbf{``Any function~$\RR \to \RR$ is continuous.''}\label{sect:eff-continuous}
 In the standard topos, this statement is plainly false, with the sign and Heaviside functions
@@
 straightforward (a textbook reference
-is~\cite[Thm.~9.2.1]{longley-normann:higher-order-computability}), but an intuitive
+is~\cite[Theorem~9.2.1]{longley-normann:higher-order-computability}), but an intuitive
 explanation is as follows.
@@
 precision in the input~$x$, do not exist in the \effective topos.
-
-\bigskip
-\dotparagraph{``Markov's principle holds.''} Markov's principle is the
+\end{ex}
+
+\begin{ex}\textbf{``Markov's principle holds.''} Markov's principle is the
 following statement:
@@
 with output~$n$ as soon as it determines that some function value~$f(n)$ is zero.
-
-\bigskip
-\dotparagraph{``Heyting arithmetic is categorical.''} In addition to the standard
+\end{ex}
+
+\begin{ex}\textbf{``Heyting arithmetic is categorical.''} In addition to the standard
 model~$\NN$, the standard topos contains uncountably many nonstandard models of
@@
 Peano arithmetic in the standard topos is computable~\cite{tennenbaum:models}.
+\end{ex}
 
@@
 law of excluded middle still fails, but some instances which are wrong in the
-\effective topos do hold in this topos. For instance, the instance ``any
+\effective topos do hold in this topos. For example, the instance ``any
 function~$\NN \to \NN$ is the zero function or not'' does: Its external meaning
@@
 
-% XXX introduction...
-
 Associated to any topological space~$X$ (such as Euclidean space), there is the
@@
 
+Just as the \effective topos provides a coherent setting for studying
+computability using a naive language, the sheaf topos~$\Sh(X)$ provides a
+coherent setting for studying continuous~$X$-indexed families of objects (sets,
+numbers, functions) as if they were single objects.
+% FUTURE: Explain better.
+% FUTURE: History, sites, flexibility
+
 \subsection{A geometric interpretation of double negation}
@@
 
-\subsection{Real numbers}
+\subsection{Reifying continuous families of real numbers as single real numbers}
 As detailed in Section~\ref{sect:eff-continuous}, what the \effective topos believes to be a real number is
@@
 number~$a$ is actually a continuous family of real numbers on~$X$, that is, a
-continuous function~$a : X \to \RR$.
+continuous function~$a : X \to \RR$~\cite[Corollary~D4.7.5]{johnstone:elephant}.
 
@@
 
-\subsection{Real functions} Let~$(f_x)_{x \in X}$ be a continuous family of
+\subsection{Reifying continuous families of real functions as single real functions}
+Let~$(f_x)_{x \in X}$ be a continuous family of
 continuous real-valued functions; that is, each of the individual
@@
 
+\begin{ex}
 From these observations we can deduce that the intermediate value theorem of
@@
 number~$a$ such that~$f(a) = 0$.'' The external meaning of this statement is
-that in any continuous family~$(f_x)_x$ of continuous functions with~$f_x(-1) <
+that for any continuous family~$(f_x)_x$ of continuous functions with~$f_x(-1) <
 0$ and~$f_x(1) > 0$ for all~$x \in X$, it is locally possible to pick zeros of
@@
 
-\section{Topos adapted to synthetic differential geometry}
+In contrast, the intermediate value theorem for monotone functions does have an
+intuitionistic proof and hence applies in the internal universe of~$\Sh(X)$.
+Thus for any continuous family~$(f_x)_x$ of continuous monotone functions
+with~$f_x(-1) < 0$ and~$f_x(1) > 0$ for all~$x \in X$, it is locally possible
+to pick zeros of the family in a continuous fashion.\footnote{While the
+Kripke--Joyal translation of~``$\exists$'' is by definition \emph{local
+existence}, one can show that the Kripke--Joyal translation of~``$\exists!$''
+is \emph{unique existence on all opens}, in particular \emph{unique global
+existence}. Because the conclusion of the intermediate value theorem for
+monotone functions can be strengthened from ``has a zero'' to ``has a unique
+zero'', this observation shows that the zeros can even globally be picked in a
+continuous fashion.}
+\end{ex}
+
+\begin{ex}The fundamental theorem of algebra generally fails in~$\Sh(X)$, even
+for quadratic polynomials. What~$\Sh(X)$ believes to be a (Dedekind)
+complex number is externally a continuous function~$X \to \CC$. Let~$X$ be the
+complex plane. Then the identity function~$\mathrm{id}_X$ is a single complex number from the
+internal point of view of~$\Sh(X)$. The fundamental theorem of algebra would
+predict~``$\Sh(X) \models \exists a\?\CC\_ a^2 - \mathrm{id}_X = 0$'',
+hence that there is an open covering~$X = \bigcup_i U_i$ such that on each
+open~$U_i$, there is a continuous function~$a : U_i \to \CC$ such that~$a(z)^2
+- z = 0$ for all~$z \in U_i$. However, it is a basic fact of complex analysis
+that such a function does not exist if~$0 \in U_i$.
+\end{ex}
+
+\begin{ex}The standard proof of Banach's fixed point theorem employs only
+intuitionistic reasoning, hence applies internally to~$\Sh(X)$. Interpreting the
+internal Banach fixed point theorem by the Kripke--Joyal translation rules
+yields the statement that fixed points of continuous families of contractions
+depend continuously on parameters.
+\end{ex}
+
+\section{Toposes adapted to synthetic differential geometry}
 \label{sect:smooth}
@@
 * XXX phone call analogy
-* XXX quick overview of the several aspects of toposes (maybe at the end, as an outlook?)
-* XXX allowing a switch of perspective
-* XXX applications in mathematical practice
-* XXX arbitrariness of the "standard axioms"
-* XXX use the phrase "very rich and very deep" (or similarly) about set theory
+FUTURE: quick overview of the several aspects of toposes (maybe at the end, as an outlook?)
+FUTURE: use the phrase "very rich and very deep" (or similarly) about set theory
 * XXX A topos is more than its semantics
-* XXX Check references everywhere (give good pointers)
-* XXX internal/external
+FUTURE: Check references everywhere (give good pointers)
 

### 2020-01-22 paper-filmat.tex
Version as just handed in

@@
 \begin{center}
+  \begin{tikzpicture}
+    \node (comment) at (0,0) {
       \begin{minipage}{0.6\textwidth}
@@
       forward to your comments.
+      \bigskip
 
@@
       \end{minipage}
+    };
+
+    \node[draw=mypurple, very thick, rectangle, rounded corners, inner sep=5pt] [fit = (comment)] (foo) {};
+  \end{tikzpicture}
 \end{center}
@@
 that from the foundational crisis Brouwer's intuitionism would have emerged as
-the main school of thought and we would now all reject the law of excluded
+the main school of thought and that we would now all reject the law of excluded
 middle. Toposes allow us to explore alternatives to how history
@@
 uncovering hidden relations, though not between objects but between mathematical
-theories.}
+theories. \textbf{\textcolor{red}{NOTE TO EDITOR: Insert cross reference to Olivia's chapter of
+this book.}}}
 \item In some cases, a mathematical relation can be expressed quite succinctly
@@
 if it holds in the usual mathematical sense.
+
+% FUTURE:
+% \begin{rem}It might be helpful to picture mathematicians in a topos~$\E$. For
+% exchanging ideas, they use the same mathematical language like we do, but their
+% mathematical results seem unfamiliar or even wrong to us. Only when we remember
+% to switch on the Kripke--Joyal translation engine their claims start making
+% sense.\end{rem}
 
@@
 \begin{rem}The picture of a topos-theoretic multiverse is related to Hamkin's
-multiverse view in set theory~\ref{hamkins:multiverse}. In fact, the
+multiverse view in set theory~\cite{hamkins:multiverse}. In fact, the
 topos-theoretic multiverse can be regarded as an extension of the set-theoretic
@@
 Theorem~\ref{thm:reasoning}. This unwinding typically turns transparent
-internal proofs into complex external proofs -- proofs which one may not have
+internal proofs into complex external proofs -- proofs which one might not have
 found without the problem-adapted internal language provided by a
@@
 
-* XXX phone call analogy
 FUTURE: quick overview of the several aspects of toposes (maybe at the end, as an outlook?)
 FUTURE: use the phrase "very rich and very deep" (or similarly) about set theory
-* XXX A topos is more than its semantics
+FUTURE: A topos is more than its semantics
 FUTURE: Check references everywhere (give good pointers)

### 2020-02-13 announcement-agda-workshop.tex
Announcement of the Agda workshop in Padova

--- 2019-05-01 slides-birmingham2019.tex
@@
 \usepackage{pgfpages}
-\setbeameroption{show notes on second screen}
-\setbeamertemplate{note page}[plain]
-\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}
-
-% Workaround for the issue described at
-% https://tex.stackexchange.com/questions/164406/beamer-using-href-in-notes.
-\newcommand{\fixedhref}[2]{\makebox[0pt][l]{\hspace*{\paperwidth}\href{#1}{#2}}\href{#1}{#2}}
 
@@
 
-\usepackage{mathtools}
-\usepackage{booktabs}
-\usepackage{stmaryrd}
-\usepackage{array}
 \usepackage{ragged2e}
-\usepackage{multicol}
-\usepackage{tabto}
-\usepackage{xstring}
-\usepackage{ifthen}
-\usepackage[normalem]{ulem}
-\usepackage[all]{xy}
-\xyoption{rotate}
-\usepackage{tikz}
-\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing,positioning}
-\hypersetup{colorlinks=true}
-
 \usepackage{pifont}
+\usepackage{accents}
+\usepackage{graphbox}
 \newcommand{\cmark}{\ding{51}}
@@
 \setlength\parindent{0pt}
-
-\title{A general Nullstellensatz for generalised spaces}
-\author{Ingo Blechschmidt}
-\date{April 11th, 2019}
 
@@
 \newcommand\yon{\!\text{\usefont{U}{min}{m}{n}\symbol{'210}}\!}
-
-\newcommand{\A}{\mathcal{A}}
-\newcommand{\B}{\mathcal{B}}
-\renewcommand{\C}{\mathcal{C}}
-\newcommand{\M}{\mathcal{M}}
-\renewcommand{\AA}{\mathbb{A}}
-\newcommand{\E}{\mathcal{E}}
-\newcommand{\F}{\mathcal{F}}
-\renewcommand{\G}{\mathcal{G}}
-\newcommand{\J}{\mathcal{J}}
-\newcommand{\GG}{\mathbb{G}}
-\renewcommand{\O}{\mathcal{O}}
-\newcommand{\K}{\mathcal{K}}
-\newcommand{\NN}{\mathbb{N}}
-\newcommand{\QQ}{\mathbb{Q}}
-\newcommand{\RR}{\mathbb{R}}
-\newcommand{\TT}{\mathbb{T}}
-\newcommand{\PP}{\mathbb{P}}
-\newcommand{\ZZ}{\mathbb{Z}}
-\newcommand{\CC}{\mathbb{C}}
-\renewcommand{\P}{\mathcal{P}}
-\newcommand{\aaa}{\mathfrak{a}}
-\newcommand{\ppp}{\mathfrak{p}}
-\newcommand{\fff}{\mathfrak{f}}
-\newcommand{\defeq}{\vcentcolon=}
-\newcommand{\defeqv}{\vcentcolon\equiv}
-\newcommand{\Sh}{\mathrm{Sh}}
-\newcommand{\GL}{\mathrm{GL}}
-\newcommand{\Zar}{\mathrm{Zar}}
-\newcommand{\op}{\mathrm{op}}
-\newcommand{\Set}{\mathrm{Set}}
-\newcommand{\Eff}{\mathrm{Ef{}f}}
-\newcommand{\Sch}{\mathrm{Sch}}
-\newcommand{\Aff}{\mathrm{Aff}}
-\newcommand{\Ring}{\mathrm{Ring}}
-\newcommand{\LocRing}{\mathrm{LocRing}}
-\newcommand{\LRS}{\mathrm{LRS}}
-\newcommand{\Hom}{\mathrm{Hom}}
-\newcommand{\Spec}{\mathrm{Spec}}
-\newcommand{\lra}{\longrightarrow}
-\newcommand{\RelSpec}{\operatorname{Spec}}
-\renewcommand{\_}{\mathpunct{.}}
-\newcommand{\?}{\,{:}\,}
-\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
-\newcommand{\ul}[1]{\underline{#1}}
-\newcommand{\affl}{\ensuremath{{\ul{\ensuremath{\AA}}^1}}}
-\newcommand{\Ll}{\text{iff}}
-\newcommand{\inv}{inv.\@}
-\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}
-\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates
-
-\setbeamertemplate{blocks}[rounded][shadow=false]
-
-\newenvironment{indentblock}{%
-  \list{}{\leftmargin\leftmargin}%
-  \item\relax
-}{%
-  \endlist
-}
-
-% Adapted from https://latex.org/forum/viewtopic.php?t=2251 (Stefan Kottwitz)
-\newenvironment<>{hilblock}{
-  \begin{center}
-    \begin{minipage}{9.05cm}
-      \setlength{\textwidth}{9.05cm}
-      \begin{actionenv}#1
-        \def\insertblocktitle{}
-        \par
-        \usebeamertemplate{block begin}}{
-        \par
-        \usebeamertemplate{block end}
-      \end{actionenv}
-    \end{minipage}
-  \end{center}}
-
-\newcommand{\bignumber}[1]{
-  \renewcommand{\insertenumlabel}{#1}\scalebox{1.5}{\usebeamertemplate{enumerate item}}
-}
-\newcommand{\bigheart}{\includegraphics{heart}}
-
-\newenvironment{changemargin}[2]{%
-  \begin{list}{}{%
-    \setlength{\topsep}{0pt}%
-    \setlength{\leftmargin}{#1}%
-    \setlength{\rightmargin}{#2}%
-    \setlength{\listparindent}{\parindent}%
-    \setlength{\itemindent}{\parindent}%
-    \setlength{\parsep}{\parskip}%
-  }%
-  \item[]}{\end{list}}
-
-\tikzset{
-  invisible/.style={opacity=0,text opacity=0},
-  visible on/.style={alt={#1{}{invisible}}},
-  alt/.code args={<#1>#2#3}{%
-    \alt<#1>{\pgfkeysalso{#2}}{\pgfkeysalso{#3}}}
-}
-
-\newcommand{\pointthis}[3]{%
-  \tikz[remember picture,baseline]{
-    \node[anchor=base,inner sep=0,outer sep=0] (#2) {#2};
-    \node[visible on=#1,overlay,rectangle callout,rounded corners,callout relative pointer={(0.3cm,0.5cm)},fill=blue!20] at ($(#2.north)+(-0.1cm,-1.1cm)$) {#3};
-  }%
-}
-
-\tikzset{
-  invisible/.style={opacity=0,text opacity=0},
-  visible on/.style={alt={#1{}{invisible}}},
-  alt/.code args={<#1>#2#3}{%
-    \alt<#1>{\pgfkeysalso{#2}}{\pgfkeysalso{#3}}}
-}
-
-\newcommand{\hcancel}[5]{%
-  \tikz[baseline=(tocancel.base)]{
-    \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
-    \draw[red!80, line width=0.4mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
-  }%
-}
-
-\newcommand{\explain}[7]{%
-  \tikz[remember picture,baseline]{
-    \node[anchor=base,inner sep=2pt,outer sep=0,fill=#3,rounded corners] (label) {#1};
-    \node[anchor=north,visible on=<#2>,overlay,rectangle callout,rounded corners,callout
-    relative pointer={(0.0cm,0.5cm)+(0.0cm,#6)},fill=#3] at ($(label.south)+(0,-0.3cm)+(#4,#5)$) {#7};
-  }%
-}
-
-\newcommand{\explainstub}[2]{%
-  \tikz[remember picture,baseline]{
-    \node[anchor=base,inner sep=2pt,outer sep=0,fill=#2,rounded corners] (label) {#1};
-  }%
-}
-
-\newcommand{\squiggly}[1]{%
-  \tikz[remember picture,baseline]{
-    \node[anchor=base,inner sep=0,outer sep=0] (label) {#1};
-    \draw[thick,color=red!80,decoration={snake,amplitude=0.5pt,segment
-    length=3pt},decorate] ($(label.south west) + (0,-2pt)$) -- ($(label.south east) + (0,-2pt)$);
-  }%
-}
-
-% Adapted from https://latex.org/forum/viewtopic.php?t=2251 (Stefan Kottwitz)
-\newenvironment<>{varblock}[2]{\begin{varblockextra}{#1}{#2}{}}{\end{varblockextra}}
-\newenvironment<>{varblockextra}[3]{
-  \begin{center}
-    \begin{minipage}{#1}
-      \begin{actionenv}#4
-        {\centering \hil{#2}\par}
-	\def\insertblocktitle{}%\centering #2}
-        \def\varblockextraend{#3}
-	\usebeamertemplate{block begin}}{
-        \par
-        \usebeamertemplate{block end}
-        \varblockextraend
-      \end{actionenv}
-    \end{minipage}
-  \end{center}}
 
@@
 
-\newcounter{framenumberpreappendix}
-\newcommand{\backupstart}{
-  \setcounter{framenumberpreappendix}{\value{framenumber}}
-}
-\newcommand{\backupend}{
-  \addtocounter{framenumberpreappendix}{-\value{framenumber}}
-  \addtocounter{framenumber}{\value{framenumberpreappendix}}
-}
-
-\newcommand{\mynav}[3]{%
-  foo
-}
-
 \newcommand{\insertframeextra}{}
-\setbeamertemplate{footline}{%
-  \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex,dp=1ex,right,rightskip=1mm,leftskip=1mm]{}%
-    % \inserttitle
-    \hfill
-    \insertframenumber\insertframeextra\,/\,\inserttotalframenumber
-  \end{beamercolorbox}%
-  \vskip0pt%
-}
+\setbeamertemplate{footline}{}
 
@@
 
-\addtocounter{framenumber}{-1}
+\begin{frame}
+  \centering
+  \vspace*{-0.3em}
+  \includegraphics[width=0.25\textwidth]{hen}
+  \\[-0.3em]
+  \huge
+  \hil{Agda}, \mbox{a
+  \textit{beaut$\accentset{\,\,\includegraphics[width=0.3em]{butterfly}}{\text{\textit{\i}}}$ful} proof assistant}
 
-%\setbeamertemplate{headline}{\mynav{gray}{gray}{gray}}
+  \tiny
+  second part of the course \textsc{teoria dei tipi} \textbullet{} begin April
+  29th, 2020 \textbullet{} by Ingo Blechschmidt
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses}\end{minipage}}
-\begin{frame}[c]
-  \centering
+  \normalsize
+  \raggedright
+  \bigskip
+
+  \begin{columns}
+    \begin{column}{0.47\textwidth}
+      Agda is \ldots
+      \begin{enumerate}
+        \item a programming language
+        \item a \hil{proof language}
+        \item<3-> somewhat hard to learn on one's own
+        \item<3-> \hil{fun and easy} to learn as part of a course
+      \end{enumerate}
+    \end{column}
+
+    \begin{column}[t]{0.61\textwidth}
+      \only<1>{\includegraphics[width=1.0\textwidth,align=t]{agda-example}}
+      \only<2->{
+        With Agda you can \ldots
+        \begin{enumerate}
+          \item ensure \hil{correctness} of proofs
+          \item \mbox{\hil{practically explore} type theory}
+          \item appreciate mathematics from a new point of view
+        \end{enumerate}
 
         \bigskip
-  \includegraphics[width=0.4\textwidth]{phantoms}
-  \bigskip
-
-  \begin{tikzpicture}
-    \def\R{8pt}
-    \node (title) {\hil{A general Nullstellensatz for generalised spaces}};
-    \begin{pgfonlayer}{background}
-      \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple]
-        ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
-        ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
-        ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
-        ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
-        cycle;
-    \end{pgfonlayer}
-  \end{tikzpicture}
-
-  \scriptsize
-  \textit{-- an invitation --}
-  \bigskip
-
-  6th Workshop on Formal Topology in Birmingham \\
-  April 11th, 2019
-  \bigskip
-
-  Ingo Blechschmidt \\
-  Università di Verona
-  \par
-
-  \jnote{1}{\textcolor{red}{This set of slides will be updated soon with more
-  annotations.} Right now important references are not properly mentioned, including:
-  \begin{itemize}
-    \item Marc Bezem, Ulrik Buchholtz and Thierry Coquand. \emph{Syntactic forcing
-    models for coherent logic.}
-    \item Olivia Caramello. \emph{Universal models and definability.}
-    \item Carsten Butz and Peter Johnstone. \emph{Classifying toposes for
-    first-order theories.}
-  \end{itemize}}
-\end{frame}}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\hspace*{-1.7cm}\includegraphics[height=\paperheight,trim=0 0 16.9cm 0,clip]{black-hole-m87}\end{minipage}}
-\begin{frame}[plain]
-  \vspace*{\paperheight}\vspace*{-0.8cm}\scriptsize
-  \centering
-  \textcolor{white}{Fig.: The Event Horizon Telescope picture of the central black
-  hole in the galaxy M87}
-  \par
-
-  \jnote{1}{At this meeting we had lots of discussions on the notion of space.
-  Here is a picture of actual space, or rather the lack thereof, released only
-  yesterday.}
-\end{frame}}
-\addtocounter{framenumber}{-1}
-
-\section{The generic model}
-
-\subsection{The mystery of nongeometric sequents}
-
-% \begin{document}
-
-\begin{frame}{The mystery of nongeometric sequents}
-  \jnote{1}{A \emph{geometric sequent} is a syntactical expression of the
-  form~$(\varphi \seq{x_1\?X_1,\ldots,x_n\?X_n} \psi)$,
-  where~$x_1\?X_1,\ldots,x_n\?X_n$ is a list of variable declarations,
-  the~$X_i$ ranging over the available sorts, and~$\varphi$ and~$\psi$ are
-  \emph{geometric formulas}. Often the variable context is abbreviated to~$\vec
-  x \? \vec X$ or even just~$\vec x$. Such a sequent is read as ``in the
-  context of variables~$\vec x$, $\varphi$ entails~$\psi$''.
-
-  Geometric formulas are built from atomic propositions (using equality or the
-  relation symbols) using the connectives~$\top$, $\bot$, $\wedge$, $\bigvee$
-  (set-indexed disjunction) and~$\exists$. Geometric formulas may not
-  contain~$\neg$, $\Rightarrow$, $\forall$.
-
-  There is a notion of a \emph{model} of a geometric theory in
-  a given topos. For instance, a ring in the usual sense is a model of the
-  theory of rings in the topos~$\Set$. The structure sheaf of a scheme~$X$ is a
-  model in the topos~$\Sh(X)$ of set-valued sheaves on~$X$.
-
-  With \emph{topos} we mean Grothendieck topos, and as metatheory we use a
-  constructive but impredicative flavour of English (which could be formalised
-  by what is supported by the internal language of elementary toposes with an NNO).
-  However the Nullstellensatz presented later makes no use of the
-  subobject classifier, hence the results can likely be generalised to hold in
-  a predicative metatheory or to hold for arithmetic universes.}
-
-  \jnote{2-}{Among all models in any topos, the \emph{universal} or
-  \emph{generic} one is special. It enjoys the universal property that any
-  model in any topos can be obtained from it by pullback along an essentially
-  unique geometric morphism. It is intriguing from a logical point of view
-  because it has exactly those properties which are shared by any model in any
-  topos.
-
-  One could argue, with a certain amount of success, that the generic model of
-  the theory of rings is what a mathematician implicitly refers to when she
-  utters the phrase ``Let~$R$ be a ring''. This point of view is fundamental to
-  the slogan \emph{continuity is geometricity}, as expounded for instance in
-  \fixedhref{http://www.cs.bham.ac.uk/~sjv/GeoAspects.pdf}{Continuity and
-  geometric logic} by Steve Vickers.}
-
-  \jnote{3-}{Crucially, the conservativity statement only pertains to
-  properties which can be put as geometric sequents. Generic models may have
-  additional nongeometric properties. Because conservativity does not apply to
-  them, they are not shared by all models in all toposes -- but any consequences
-  which can be put as geometric sequents are.
-
-  For instance, if we want to verify a geometric sequent for all local rings,
-  we may freely use the displayed field axiom.
-  Hence one reason why these nongeometric sequents are interesting is because
-  they provide us with new reduction strategies (``without loss of
-  generality'').}
-
-  Let~$\TT$ be a \explain{geometric theory}{1-}{red!20}{-1cm}{0cm}{0cm}{\begin{minipage}{4.3cm}
-    \small
-    sorts,
-    function symbols,
-    relation symbols,
-    geometric sequents as axioms
-  \end{minipage}}, for instance the
-  \explain{theory of rings}{1-}{blue!20}{-1cm}{0cm}{0cm}{\begin{minipage}{4.3cm}
-    \small
-    \begin{tabbing}
-      fun. symb.: \= \kill
-      sorts: \> $R$ \\
-      fun. symb.: \> $0$, $1$, $-$, $+$, $\cdot$ \\
-      % rel. symb.: \> \emph{none} \\
-      axioms: \> $(\top \vdash_{x,y:R} x y = y x)$, \ldots
-    \end{tabbing}
-  \end{minipage}}.
-
-  \vspace*{1.8cm}
-
-  \begin{center}
-    \begin{tikzpicture}[ultra thick, node distance=7mm]
-      \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40] (a) {$\ZZ$};
-      \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40, right=of a] (b) {$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$};
-      \node[regular polygon, regular polygon sides=5, draw=orange!80,
-      fill=orange!20, right=of b, rounded corners=1pt, inner sep=0cm] (c) {$\O_X$};
-      \node[star, rounded corners=1pt, star points=10, inner sep=0cm, draw=purple!80, fill=purple!20, right=of c] {$U_\TT$};
-    \end{tikzpicture}
-  \end{center}
-
-  \vspace*{-0.5cm}
-  \pause
-
-  \justifying
-  \textbf{Theorem.} There is a \hil{generic model}~$U_\TT$. It is
-  \hil{conservative} in that
-  for any \explainstub{geometric sequent}{yellow!70}~$\sigma$ the following notions coincide:
-  \vspace*{-1.2em}
-  \begin{enumerate}
-    \item The sequent~$\sigma$ holds for~$U_\TT$.
-    \item The sequent~$\sigma$ holds for any~$\TT$-model in any
-    \explainstub{topos}{purple!20}.
-    \item The sequent~$\sigma$ is provable modulo~$\TT$.
-  \end{enumerate}
-
-  \pause
-
-  \justifying
-  %\textbf{Main observation.} The generic model can validate nongeometric
-  %sequents which are not provable modulo~$\TT$ and which do not hold for
-  %arbitrary models in arbitrary toposes.
-
-  \textbf{Observation (Kock).} The generic local ring is a field:
-  \vspace*{-0.2cm}
-  \[ (x = 0 \Rightarrow \bot) \vdash_{x:R} (\exists y\?R\_ xy = 1) \]
-\end{frame}
-
-\subsection{Construction of the generic model}
-
-\begin{frame}{Construction of the generic model}
-  The generic model is \hil{not} the same as \ldots
-  \begin{itemize}
-    \item the \hil{initial model} (think~$\ZZ$) or
-    \item the \hil{free model on one generator} (think~$\ZZ[X]$).
-  \end{itemize}
-  Set-based models are \hil{too inflexible}.
-
-  \textbf{Definition.} The \hil{syntactic site}~$\C_\TT$ has \ldots
-  \begin{enumerate}
-    \item objects: $\{ x_1\?X_1, \ldots, x_n\?X_n\_ \varphi \}$ (shorter:
-    $\{ \vec x\_ \varphi \}$)
-    \item morphisms: eqv. classes of \explainstub{provably functional
-    formulas}{red!20}
-    \item coverings: \explainstub{provably jointly surjective
-    families}{purple!30}
-  \end{enumerate}
-  The topos of sheaves over~$\C_\TT$ is the \hil{classifying
-  topos}~$\Set[\TT]$. The generic model interprets a sort~$X$ by~$\yon\{x\?X\_\top\}$.
-
-  \jnote{1}{In case the theory~$\TT$ is a Horn theory (for instance if it is
-  an equational theory), the \emph{term algebra} (the set of terms in the
-  empty context modulo provable equality) is a model of~$\TT$. While such
-  models do enjoy some nice categorical properties, they are in general
-  \emph{not} the generic model.
-
-  For instance, if~$\TT$ is the theory of rings, then the initial model
-  is~$\ZZ$. This model validates some geometric sequents which are not
-  validated by all rings, for instance~$(x^2 = 0 \seq{x:R} x = 0)$ or $(1 = 0
-  \vdash \bot)$.
-
-  In general, the generic model cannot be realised as a set-based model (with
-  a set for each sort, a map for each function symbol and so on). Sets are too
-  constant for this purpose; the flexibility of sheaves (``variable sets'') is
-  required: The generic model lives in the topos of set-valued
-  sheaves over~$\C_\TT$.}
-
-  \jnote{2}{A morphism~$A = \{\vec x\_ \varphi\} \to \{\vec y\_ \psi\} = B$ in~$\C_\TT$ is the
-  equivalence class (modulo provable equivalence) of a geometric
-  formula~$\theta$ such that~$\TT$ proves
-  \begin{enumerate}
-    \item ``$\theta$ is a relation on~$A \times B$'': $(\theta \seq{\vec x,
-    \vec y} \varphi \wedge \psi)$
-    \item ``$\theta$ is total'': $(\varphi \seq{\vec x} \exists \vec y\_
-    \theta)$
-    \item ``$\theta$ is single-valued'': $(\theta \wedge \theta[\vec y'/\vec y]
-    \seq{\vec x,\vec y,\vec y'} \vec y = \vec y')$
-  \end{enumerate}
-
-  \vspace*{-0.6em}
-  A family~$(\{ \vec x_i\_ \varphi_i \} \xrightarrow{[\theta_i]} \{ \vec y\_
-  \psi \})_i$ is a covering iff~$\TT$ proves~$(\psi \seq{\vec y} \bigvee_i
-  \exists \vec x_i\_ \theta_i)$.
-
-  The slides experiment with using the symbol~``$\yon$'' for the Yoneda
-  embedding~$\C_\TT \to \Set[\TT]$, as in \fixedhref{http://www.math.jhu.edu/~eriehl/elements.pdf}{Elements
-  of~$(\infty,1)$-category theory} by Emily Riehl and Dominic Verity.}
-
-  \jnote{3}{The special case that the generic model of a theory~$\TT$ can be realised as
-  a model in~$\Set$ occurs iff~$\TT$ is Morita-equivalent to the
-  empty theory, that is, iff~$\TT$ has exactly one model in any topos.
-
-  The special case that there exists at least some conservative~$\TT$-model
-  in~$\Set$ occurs iff~$\TT$ has a conservative geometric expansion to a theory
-  which is Morita-equivalent to the empty theory.}
-\end{frame}
-
-\subsection{Going internal}
-
-\begin{frame}{Working internally to toposes}
-  \justifying
-  Let~$\C$ be a site. We recursively define
-  \[
-    U \models \varphi \quad
-    \text{(``$\varphi$ holds on~$U$'')}
-  \]
-  for objects~$U \in \C$ and \squiggly{formulas}~$\varphi$.
-  Write~``$\Sh(\C) \models \varphi$'' for~$1 \models \varphi$.
-  \footnotesize
-  \[ \renewcommand{\arraystretch}{1.08}\begin{array}{@{}l@{\ \ }c@{\ \ }l@{}}
-  U \models \top &\Ll& \text{true} \\
-  U \models \bot &\Ll& \hcancel{\text{false}}{0pt}{3pt}{0pt}{-2pt}\ \text{the
-  empty family is a covering of~$U$} \\
-  U \models s = t \? F &\Ll& s|_U = t|_U \in F(U) \\
-  U \models \varphi \wedge \psi &\Ll&
-  \text{$U \models \varphi$ and $U \models \psi$} \\
-  U \models \varphi \vee \psi &\Ll&
-  \hcancel{\text{$U \models \varphi$ or $U \models
-  \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $(U_i \to U)_i$} \\
-  && \quad\quad\text{such that for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
-  U \models \varphi \Rightarrow \psi &\Ll&
-  \text{for all~$V \to U$: }
-  \text{$V \models \varphi$ implies $V \models \psi$} \\
-  U \models \forall s \? F\_ \varphi(s) &\Ll&
-  \text{for all $V \to U$ and sections~$s_0 \in F(V)$: $V \models \varphi(s_0)$} \\
-  % U \models \forall F\_ \varphi(F) &\Ll&
-  % \text{for all $V \to U$ and sheaves~$F_0$ over~$V$: $V \models \varphi(F_0)$} \\
-  U \models \exists s \? F\_ \varphi(s) &\Ll&
-  \hcancel{\text{there exists $s_0 \in F(U)$ such that $U \models \varphi(s_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
-  &&
-  \text{there exists a covering $(U_i \to U)_i$ such that for all~$i$:} \\
-  && \quad\quad \text{there exists~$s_0 \in F(U_i)$ such that $U_i \models \varphi(s_0)$}
-  %U \models \exists F\_ \varphi(F) &\Ll&
-  %\hcancel{\text{there exists a sheaf $F_0$ on $U$ such that $U \models
-  %\varphi(F_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
-  %&&
-  %\text{there exists a covering $(U_i \to U)_i$ such that for all~$i$:} \\
-  %&& \quad\quad \text{there exists a sheaf~$F_0$ on $U_i$ such that $U_i \models \varphi(F_0)$}
-  \end{array} \]
-
-  \jnote{1}{
-    The internal language of a (Grothendieck or elementary) topos~$\E$ is a
-    device which allows us to speak and reason about the objects and morphisms
-    of~$\E$ in a naive element-based language close to the usual formal
-    mathematical language. Using this language, objects of~$\E$ look like plain
-    old sets [or types]; morphisms look like plain old maps between those sets;
-    epimorphisms look like surjections; group objects look like groups; and so on.
-
-    In particular, we can use the internal language to define what it means for a
-    given~$\TT$-structure in~$\E$ to be a model -- namely iff it looks like a
-    model from the internal point of view.
-
-    The internal language can be implemented by the \emph{Kripke--Joyal semantics},
-    a translation procedure which converts formulas of the internal language
-    into external statements about the objects and morphisms of~$\E$. The slide
-    displays some of the translation rules in the case that~$\E$ is a
-    Grothendieck topos.
-
-    We can actually do mathematics internally because the Kripke--Joyal
-    semantics is sound with respect to intuitionistic logic: If~$\E \models
-    \varphi$ and if~$\varphi$ intuitionistically entails a further
-    formula~$\psi$, then~$\E \models \psi$.
-
-    An instructive special case is provided by the topos~$\Set$, because~$\Set
-    \models \varphi$ iff~$\varphi$ holds in the usual mathematical sense.
+        \centering\hil{``proving $=$ programming''}
       }
-
-  \jnote{2}{
-    The Kripke--Joyal semantics can be extended to interpret unbounded
-    quantification (``for all sets'' as opposed to ``for all elements of the
-    particular set~$X$'') and dependent types. The former are for instance
-    required to express universal properties (``for all groups'', ``for all
-    rings''), and the latter are all over the place, even if their use might
-    not be particularly highlighted.
-
-    With these extensions, we can
-    import all of everyday constructive impredicative mathematics into the
-    internal world of a topos.
-
-    Some illustrations of working with the internal language can be found in these
-    sets of slides:
-    \begin{itemize}
-      \item
-      \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-leipzig2018.pdf}{Slides
-      for Jürgen Jost's group seminar at the MPI Leipzig}
-      \item
-      \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-como2018.pdf}{Slides
-      for \emph{Toposes in Como}}
-      (recording
-      \fixedhref{https://www.youtube.com/watch?v=SuELfC5P_R8}{available})
-    \end{itemize}
-    A longer exposition, with pointers to the literature, can be found in Section~2 of
-    \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these notes}.
-  }
-\end{frame}
-
-\section[Nongeometric properties]{A selection of nongeometric properties}
-
-\begin{frame}{A selection of nongeometric properties}
-  The \hil{generic object} validates:
-  \begin{enumerate}
-    \item $\forall x,y\?U_\TT\_ \neg\neg(x = y)$.
-    \item $\forall x_1,\ldots,x_n\?U_\TT\_ \neg \forall y\?U_\TT\_
-    \bigvee_{i=1}^n y = x_i.$
-    \item $(U_\TT)^{U_\TT} \cong 1 \amalg U_\TT$.
-  \end{enumerate}
-
-  The \hil{generic ring} validates:
-  \begin{enumerate}
-    \item $\forall x\?U_\TT\_ \neg\neg(x = 0)$.
-    \item $\forall x\?U_\TT\_ (x = 0 \Rightarrow 1 = 0) \Rightarrow (\exists
-    y\?U_\TT\_ xy = 1)$.
-  \end{enumerate}
-
-  The \hil{generic local ring} validates:
-  \begin{enumerate}
-    \item $\neg \forall x\?U_\TT\_ \neg\neg(x = 0)$.
-    \item $\forall a_0,\ldots,a_{n-1}\?U_\TT\_ \neg\neg \exists x\?U_\TT\_
-      x^n + a_{n-1}x^{n-1} + \cdots + a_0 x^0 = 0$.
-    \item \justifying Let~$\Delta = \{ \varepsilon \? U_\TT \,|\, \varepsilon^2 = 0 \}$.
-    For any map~$f : \Delta \to U_\TT$, there are unique elements~$a,b \?
-    U_\TT$ s.\,th.~$f(\varepsilon) = a + b\varepsilon$ for all~$\varepsilon
-    \? \Delta$.
-%   \item For any finitely presented~$U_\TT$-algebra~$A$,
-%   the canonical map~$A \to (U_\TT)^{\Hom_{U_\TT}(A,U_\TT)}$
-%   is an isomorphism.
-  \end{enumerate}
-
-  \jnote{1}{
-    The generic object, the generic model of the theory which has exactly one
-    sort and no function symbols, relations symbols or axioms, appears to be
-    slightly indecisive: On the one hand, up to a double negation, it is a
-    subsingleton; on the other hand, it is infinite. This observation is due to
-    Carsten Butz and Peter Johnstone
-    (\fixedhref{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.16.4102&rep=rep1&type=pdf}{Classifying toposes for
-    first-order theories}).
-
-    The generic ring, the generic model of the theory of rings, is similarly
-    indecisive. It is infinite in the following sense:
-    \[ \forall x_1,\ldots,x_n \? U_\TT\_
-      \bigl(\forall y \? U_\TT\_ \bigvee_{i=1}^n (y = x_i)\bigr) \Rightarrow
-      1 = 0. \]
-
-    The theory of local rings is the quotient theory of the theory of rings
-    obtained by adding the axioms~$(1 = 0 \vdash \bot)$ and~$((\exists z\_
-    (x+y)z = 1) \seq{x,y} (\exists z\_ xz = 1) \vee (\exists z\_ yz = 1))$.
-    (Assuming the Boolean prime ideal theorem, a ring is local in this sense
-    iff it local in the usual sense (has exactly one maximal ideal).)
-
-    The first displayed property of the generic local ring illustrates
-    that nongeometric sequents need not be inherited by quotient theories.
-  }
-
-  \jnote{2}{
-    All of the displayed properties give rise to reduction techniques:
-    If we want to verify a geometric sequent for all rings, it suffices to
-    verify it for the generic ring; but the generic ring has additional
-    nongeometric properties not shared by every ring, such as the two displayed
-    ones. (This was for instance used
-    \fixedhref{https://www.sciencedirect.com/science/article/pii/0022404976900025}{by
-    Anders Kock} and
-    \fixedhref{https://link.springer.com/chapter/10.1007\%2FBFb0061835}{by
-    Gonzalo Reyes}.)
-
-    However, we face some challenges when pursuing these reduction techniques, including
-    the following:
-    \begin{enumerate}\justifying
-      \item It is not easy to determine interesting and useful properties of
-      the generic model.
-      \item The set of validated nongeometric sequents changes slightly
-      unpredictably when passing to quotient theories. For instance, when
-      proving that a geometric sequent holds for all rings, we may assume that
-      any element is \emph{not not} zero. But we may not assume this
-      simplification if we want to verify a geometric sequent for all local
-      rings (and if we want to exploit the given locality in the proof).
-      \item There is only so much we want to state and prove in full generality
-      for all rings, all local rings, all modules, and so on. We are often much
-      more interested in properties of particular mathematical objects.
-    \end{enumerate}
-  }
-
-  \jnote{3}{
-    The firstly-mentioned problem on the previous page is alleviated by the
-    Nullstellensatz presented in this talk, which gives a systematic and
-    universal source of nongeometric sequents validated by the generic model.
-    However, manual work is still required to reduce this set of sequents to a
-    smaller, manageable one consisting of memorable properties while hopefully
-    still preserving universality.
-
-    To counter the third problem, it's prudent to consider geometric theories
-    which depend on a given mathematical object of interest. For instance,
-    given a ring~$A$, we can consider the theory of prime ideals of~$A$, of
-    complemented prime ideals, of filters, and so on. The classifying toposes
-    of these theories are of independent interest -- in fact they are sheaf
-    toposes over certain important spaces in algebraic geometry -- and
-    nongeometric sequents validated by their generic models bundle nontrivial
-    information about~$A$. More details are on the following slide.
-  }
-\end{frame}
-
-\section{Affine schemes}
-
-\begin{frame}{Affine schemes}
-  \jnote{1}{
-    A ring is \emph{local} iff every invertible sum contains an invertible
-    summand, that is if~$1 \neq 0$ and if~$x + y$ invertible implies~$x$
-    invertible or~$y$ invertible. Assuming the axiom of choice, this elementary
-    definition is equivalent to the textbook definition of a local ring (a ring
-    with exactly one maximal ideal). A ring homomorphism is \emph{local} iff it
-    reflects invertibility.
-
-    The notion of a \emph{filter} in a ring~$A$ is a direct axiomatisation of
-    what classically would be the complement of a prime ideal. The filter
-    axioms are: $1 \in F$, $(xy \in F) \Leftrightarrow ((x \in F) \wedge (y \in
-    F))$, $\neg(0 \in F)$, $(x+y \in F) \Rightarrow ((x \in F) \vee (y \in F))$.
-
-    A \emph{free local ring} over~$A$ is a homomorphism into a local
-    ring~$A'$ such that any homomorphism into a local ring~$R$ factors uniquely
-    over~$A'$ via a local homomorphism.
-
-    For any particular local ring~$A \xrightarrow{f} R$, the
-    localisation~$A[S^{-1}]$ with~$S = f^{-1}[R^\times]$ is a local ring which
-    fits into the displayed diagram. However, in general there is no single
-    choice of~$S$ which would work for any local ring~$R$. Indeed, classically
-    one can show that a free local ring over~$A$ exists if and only if~$A$
-    contains exactly one prime ideal, in which case~$A$ itself is the free
-    local ring.
-  }
-
-  \jnote{2}{
-    If we want a free local ring to exist for any ring~$A$, we have to broaden
-    our notion of existence and embrace rings which live in toposes other
-    than~$\Set$. There is a notion of a homomorphism between rings living in
-    arbitrary toposes, and using this notion one can verify:
-
-    The free local ring over~$A$ can be built in the classifying topos of
-    filters of~$A$, as the localisation (in that topos) of~$A$ at the generic
-    filter. (More precisely, as the localisation of the mirror image of~$A$ in that
-    topos, that is the constant sheaf~$\underline{A}$.)
-
-    The classifying topos of filters of~$A$ coincides with the topos of sheaves
-    over what's called the \emph{spectrum of~$A$} in algebraic geometry, and
-    under this equivalence the free local ring coincides with the structure
-    sheaf of spectrum. In fact the classifying topos serves as a good
-    constructive substitute for the classical spectrum construction,
-    enjoying the expected universal property even if the axiom of choice is not
-    available, which is why it is simply denoted~``$\Spec(A)$'' on the slide.
-
-    (The classifying topos of prime ideals of~$A$ is also interesting; it
-    coincides with the topos of sheaves over the spectrum of~$A$ equipped with
-    the constructible topology.)
-  }
-
-  \jnote{3}{
-    Assuming the Boolean prime ideal theorem, the geometric sequents
-    validated by~$A^\sim$ are easy to describe: They are precisely those which
-    are validated by all the stalks~$A_\ppp$ of~$A$.
-
-    But~$A^\sim$ enjoys further unique properties which are not shared by the
-    stalks of~$A$, other localisations of~$A$, quotients of~$A$ or indeed any
-    reasonable construction. Three of these are displayed on the slide.
-    (A ring is \emph{anonymously Noetherian} iff each of its ideals is
-    \emph{not not} finitely generated. Textbook proofs of Hilbert's basis theorem
-    are constructively acceptable for this Noetherian condition.)
-
-    The object~$A^\sim$ strikes a fine balance: On the one hand, it is still
-    close to~$A$, so that information learned about~$A^\sim$ teaches us
-    about~$A$; on the other hand, it enjoys unique properties rendering it
-    simpler than~$A$.
-
-    This balance allows for a simple and conceptually satisfying proof of
-    \emph{Gro\-then\-dieck's generic freeness lemma}, an important theorem in
-    algebraic geometry. Details can be found in
-    \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-padova2018.pdf}{this
-    set of slides}.
-  }
-
-  Let~$A$ be a ring. Is there a \hil{free local ring}~$A \to A'$ over~$A$?
-  \begin{columns}[t]
-    \begin{column}{0.4\textwidth}
-      $\xymatrix{
-        A \ar[rd] \ar[rrr]^f &&& {\substack{\phantom{\text{local}}\\\text{\normalsize$R$}\\\text{local}}} \\
-        & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!35]{\text{local}}[rru]
-      }$
-    \end{column}
-
-    \begin{column}{0.50\textwidth}
-      \small\justifying
-      For a fixed ring~$R$, the localisation $A' \defeq A[S^{-1}]$ with $S \defeq
-      f^{-1}[R^\times]$ would do the job. ($S$ is a \emph{filter}.)
-      \medskip
-
-      Hence we need the \hil{generic filter}.
     \end{column}
   \end{columns}
-
-  \pause
-
-  \justifying
-  The free local ring over~$A$ is~$A^\sim \defeq \ul{A}[F^{-1}]$,
-  where~$F$ is the generic filter, living in~$\Spec(A)$, the classifying topos
-  of filters of~$A$.
-
-  \pause
-
-  If~$A$ is reduced ($x^n = 0 \Rightarrow x = 0$):
-
-  \vspace*{-0.8em}
-  \setbeamercolor{block body}{bg=red!30}
-  \setbeamercolor{structure}{fg=purple}
-  \begin{varblock}{\textwidth}{}
-    $A^\sim$ is a \hil{field}: $\forall x\?A^\sim\_ (\neg(\exists y\?A^\sim\_
-    xy = 1) \Rightarrow x = 0)$.
-
-    $A^\sim$ has \hil{$\boldsymbol{\neg\neg}$-stable equality}:
-    $\forall x,y\?A^\sim\_ \neg\neg(x = y) \Rightarrow x = y$.
-
-    \mbox{$A^\sim$ is \hil{anonymously Noetherian}.}\\[-1.2em]
-  \end{varblock}
 \end{frame}
-
-\section[A systematic source]{A systematic source of nongeometricity?}
-
-\begin{frame}{A systematic source of nongeometricity?}
-  \jnote{1-2}{
-    Mimicking the synthetic approach to differential geometry, synthetic
-    algebraic geometry is a framework for algebraic geometry in which schemes
-    can be modelled by plain old sets, morphisms of schemes by plain old maps
-    between those sets, group schemes by plain old groups, and so on. Unlike
-    its close cousin, it is far less developed; some first steps are outlined
-    in Sections~19 and~20 of
-    \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these notes}.
-
-    Synthetic algebraic geometry is carried out internally to the \emph{big
-    Zariski topos} of a given base scheme; in the special case that the base
-    scheme is the terminal scheme~$\Spec(\ZZ)$, this topos is just the
-    classifying topos of local rings. The relevant generic model living in the
-    big Zariski topos is denoted~``$\affl$'' because it coincides with the
-    functor of points of the affine line.
-  }
-
-  \jnote{2}{
-    Marc Bezem, Ulrik Buchholtz and Thierry Coquand answered in their 2017 paper
-    \fixedhref{https://arxiv.org/abs/1712.07743}{Syntactic forcing
-    models for coherent logic} Gavin Wraith's question in the negative. (As
-    Thierry remarked during the talk, this is even if one takes care to phrase
-    the question in a way to exclude the trivial counterexamples given by
-    instances of the law of excluded middle in the language of~$\TT$.) If the
-    answer had been positive, this would have given a neat, if somewhat hard to
-    use in practice, characterisation of the formulas validated by the generic model.
-  }
-
-  \jnote{3}{
-    The topos-theoretic \emph{Nullstellensatz}, to be presented on the next
-    slides, answers the displayed three questions in the affirmative.
-
-    Briefly, the Nullstellensatz is a certain statement in the language of a
-    given geometric theory~$\TT$ which is
-    \begin{itemize}\justifying
-      \item validated by the generic~$\TT$-model,
-      \item typically not validated by other~$\TT$-models, and
-      \item such that any statement validated by the generic~$\TT$-model can be
-      deduced, in intuitionistic logic, from the axioms of~$\TT$ and the
-      Nullstellensatz.
-    \end{itemize}
-
-    We believe that this characterisation is as explicit as it can get, but
-    would be delighted to be surprised by a future improvement. We stumbled on
-    it by playing with the synthetic quasicoherence statement, not least thanks
-    to encouragement by Alexander Oldenziel. However the route from that
-    statement to the Nullstellensatz is not quite direct; it turns out that
-    synthetic quasicoherence is a corollary of a specialisation of a
-    higher-order version of the Nullstellensatz to Horn theories.
-  }
-
-  \justifying
-  \textbf{Empirical fact.}
-  In \hil{synthetic algebraic geometry}, every known property of~$\affl$ followed from
-  its \hil{synthetic quasicoherence}:
-
-  \begin{quote}
-    \emph{\mbox{For any finitely presented~$\affl$-algebra~$A$, the canonical map}
-    \[ A \longrightarrow (\affl)^{\Hom_\affl(A,\affl)},\ s \longmapsto (x
-    \mapsto x(s)) \]
-    is an isomorphism of~$\affl$-algebras.}
-  \end{quote}
-
-  \vspace*{1.0em}
-  \begin{enumerate}
-    \item Does a general metatheorem explain this observation?
-    \item Is there a systematic source in any classifying topos?
-    \item Is there even an exhaustive source?
-  \end{enumerate}
-  \vspace*{-3.3em}
-
-  \visible<2>{\begin{tikzpicture}[overlay]
-    \draw[fill=white, draw=white, opacity=0.95] (-1,1.4) rectangle (\paperwidth,5.4);
-    \node[anchor=south west,inner sep=0] (image) at (0,0.9) {\vbox{
-      \only<2>{
-	\centering
-	\includegraphics[width=0.9\textwidth]{wraith-some-recent-developments-in-topos-theory} \\
-	\scriptsize
-        Gavin Wraith. \emph{Some recent developments in topos theory.} \\ In:
-        Proc.\@ of the ICM (Helsinki, 1978).
-      }
-    }};
-  \end{tikzpicture}}%
-
-  \[ \xymatrix{
-    \text{$\TT$ proves~$\alpha$\;} \ar@/_1pc/@{=>}[r] &
-    \text{\;$\alpha$ holds for~$U_\TT$\;} \ar@/_1pc/@{=>}[r]
-    \ar@/_1pc/@{=>}@[red][l]
-    |{\SelectTips{cm}{}\textbf{\scalebox{3}{\textcolor{red}{\!\texttimes\!}}}}|{} &
-    \text{\;$\alpha$ is~$\TT$-redundant}
-    \ar@/_1pc/@{=>}@[red][l]
-    |{\SelectTips{cm}{}\textbf{\scalebox{3}{\textcolor{red}{\!\texttimes\!}}}}|{}
-  } \]
-\end{frame}
-
-\section{A topos-theoretic Nullstellensatz}
-
-\newcommand{\nullstellensatz}{%
-  \justifying
-  \textbf{Theorem.} Internally to~$\Set[\TT]$:
-
-  \begin{quote}\emph{\mbox{For any \explainstub{geometric$^\star$ sequent~$\sigma$}{red!20} over
-  the \explainstub{signature of~$\ul{\TT}/U_\TT$}{blue!20},}
-  if~\explainstub{$\sigma$ holds for~$U_\TT$}{yellow!70}, then~\explainstub{$\ul{\TT}/U_\TT$
-  proves$^\star$~$\sigma$}{yellow!70}.}
-  \end{quote}
-  \vspace*{-0.6em}
-}
-
-\begin{frame}{A topos-theoretic Nullstellensatz}
-  \jnote{1-2}{
-    By~$\ul{\TT}$ we mean the geometric theory internal to~$\Set[\TT]$ obtained
-    by pulling back the set of sorts of~$\TT$, the set of function symbols and so on
-    along the unique geometric morphism~$\Set[\TT] \to \Set$. For instance,
-    if~$\TT$ is the theory of rings, then from the internal point of view
-    of~$\Set[\TT]$, the theory~$\ul{\TT}$ will again be the theory of rings.
-
-    The theory~$\ul{\TT}/U_\TT$ will be defined on the next slide. It is a certain
-    geometric theory internal to~$\Set[\TT]$.
-
-    The asterisks in \emph{geometric$^\star$ sequent} and
-    \emph{provability$^\star$} indicate that any infinities used to index
-    disjunctions have to be come from the base topos. This restriction is an important
-    subtlety, though not vital to this talk. If~$\TT$ is a coherent
-    theory, then for coherent sequents there is no difference between
-    provability in coherent logic, provability in geometric logic and
-    provability$^\star$.
-  }
-
-  \jnote{2}{
-    The algebraic Nullstellensatz states that, in some cases, algebraic truths
-    are witnessed by explicit \emph{algebraic certificates} -- syntactical
-    objects giving a priori reasons for why a given truth is to be expected.
-
-    In the topos-theoretic Nullstellensatz, algebraic truths are replaced by
-    arbitrary truths of the generic model, subject only to the condition that
-    they can be expressed as a geometric sequent, and algebraic certificates
-    are replaced by \emph{logical certificates}: proofs.
-  }
-
-  \jnote{3}{
-    While, as stated on slide 1/20, the generic model~$U_\TT$ is a
-    conservative~$\TT$-model, the classifying topos~$\Set[\TT]$ does not
-    believe this fact. That is, the statement~``$U_\TT$ is a
-    conservative~$\ul{\TT}$-model'' is not true internally to~$\Set[\TT]$.
-    What is true is the modified statement~``$U_\TT$ is a
-    conservative$^\star$~$\ul{\TT}/U_\TT$-model''.
-  }
-
-  \nullstellensatz
-
-  \pause
-
-  \textbf{The algebraic Nullstellensatz.}
-  Let~$A$ be a ring.
-  Let~$f,g \in A[X]$ be polynomials.
-  Then, subject to some conditions:
-  \[
-    \underbrace{\bigl(\forall x \in A\_ (f(x) = 0 \Rightarrow g(x) =
-    0)\bigr)}_{\text{algebraic truth}} \Longrightarrow
-    \underbrace{\bigl(\exists h \in A[X]\_ g = hf\bigr)}_{\text{algebraic certificate}}
-  \]
-  \vspace*{-0.9em}
-
-  \pause
-
-  \textbf{A naive version.}
-  ``Internally to~$\Set[\TT]$, for any geometric sequent~$\sigma$ over
-  the signature of~$\ul{\TT}$,
-  if~$\sigma$ holds for~$U_\TT$, then~$\ul{\TT}$
-  proves~$\sigma$.'' \textbf{False,} for instance with the theory of rings we
-  have
-  \begin{align*}
-    \Set[\TT] &\models \neg(\speak{$\ul{\TT}$ proves $(\top \vdash 1 + 1 =
-    0)$}) \\
-    \text{but}\quad \Set[\TT] & \not\models \neg(1 + 1 = 0).
-  \end{align*}
-\end{frame}
-
-\begin{frame}{A varying internal theory}
-  \nullstellensatz
-
-  \explainstub{\textbf{Definition.}}{blue!20} The theory~$\ul{\TT}/U_\TT$ is the
-  internal geometric theory of~\hil{$U_\TT$-algebras}, the theory which arises from~$\ul{\TT}$ by adding:
-  \begin{enumerate}
-    \justifying\small
-    \item for each element~$x : U_\TT$ a constant symbol~$e_x$,
-    \item for each function symbol~$f$ and~$n$-tuple~$(x_1,\ldots,x_n) \in (U_\TT)^n$ the axiom
-    $(\top \vdash f(e_{x_1},\ldots,e_{x_n}) = e_{f(x_1,\ldots,x_n)})$,
-    \item for each relation symbol~$R$ and~$n$-tuple~$(x_1,\ldots,x_n) \in (U_\TT)^n$ such
-    that~$R(x_1,\ldots,x_n)$ the axiom~$(\top \vdash R(e_{x_1},\ldots,e_{x_n}))$.
-  \end{enumerate}
-
-  \textbf{Remark.} Externalising the internal classifying
-  topos~$\Set[\TT][\ul{\TT}/U_\TT]$ yields the classifying topos
-  of~$\TT$-homomorphisms.
-
-  \jnote{1}{
-    Just as locales internal to a topos~$\E$ can be externalised to yield
-    localic geometric morphisms into~$\E$, internal Grothendieck toposes can be
-    externalised to yield bounded geometric morphisms. Since the composition of
-    bounded geometric morphisms is bounded, the externalisation of a
-    Grothendieck topos internally to a Grothendieck topos is itself a
-    Grothendieck topos, hence the classifying topos of some geometric theory.
-
-    Constructing internally to~$\Set[\TT]$, where~$\ul{\TT}/U_\TT$ is just an
-    ordinary geometric theory, the classifying topos of that theory, and then
-    externalising the resulting Grothendieck topos results in the classifying
-    topos of~$\TT$-homomorphisms. There are two canonical geometric morphisms
-    from this topos to~$\Set[\TT]$, the morphism computing the domain and the
-    morphism computing the codomain, and the morphism obtained by the
-    externalisation procedure is the former.
-  }
-\end{frame}
-
-\begin{frame}{Revisiting the test cases}
-  \nullstellensatz
-  \medskip
-
-  \textbf{In the object classifier.}
-%  Assume that~$U_\TT$ is empty, that is $\forall x \? U_\TT\_ (\top \Rightarrow
-%  \bot)$. By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves~$(\top \seq{x} \bot)$.
-%  But this is false in the~$\ul{\TT}/U_\TT$-model~$U_\TT \amalg
-%  \{\heartsuit\}$.
-  Let~$x,y \? U_\TT$. Assume that~$\neg(x = y)$.
-  By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves~$(e_x = e_y \vdash \bot)$.
-  But this is false in the~$\ul{\TT}/U_\TT$-model~$U_\TT/(x \sim y)$.
-  \medskip
-
-  \textbf{In the ring classifier.}
-  Let~$f,g \? U_\TT[X]$ such that any zero of~$f$ is a zero of~$g$.
-  By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves this fact.
-  Hence it holds in the~$\ul{\TT}/U_\TT$-model~$U_\TT[X]/(f)$. In this
-  model~$f$ has the zero~$[X]$. Hence also~$g([X]) = 0$
-  in~$U_\TT[X]/(f)$, that is~$g = hf$ for some~$h \? U_\TT[X]$.
-
-  \jnote{1}{
-    This slide gives two examples how to use the Nullstellensatz to deduce
-    properties of the generic model. A couple of remarks are in order.
-
-    The Nullstellensatz is trivial for sequents~$\sigma$ of the form~$(\top
-    \vdash \psi)$. The Nullstellensatz is only interesting in case
-    that~$\sigma$ has a nontrivial antecedent or is set in a nonempty context.
-
-    Since the converse direction in the Nullstellensatz also holds
-    (because~$U_\TT$ is a~$\ul{\TT}/U_\TT$-model), the
-    statements~$\speak{$\sigma$ holds for~$U_\TT$}$
-    and~$\speak{$\ul{\TT}/U_\TT$ proves$^\star$ $\sigma$}$ are equivalent. This
-    equivalence is intriguing from a logical point of view, since the former
-    statement is a geometric implication while the latter can be put as a
-    geometric formula. (Up to a subtle issue indicated on the next slide.)
-
-    To apply the Nullstellensatz, no description of a site defining~$\Set[\TT]$
-    is required.
-
-    Often when using the Nullstellensatz, we go from an (assumed) truth
-    of~$U_\TT$ via provability$^\star$ to another model~$M$
-    of~$\ul{\TT}/U_\TT$. That is, we use provability$^\star$ as a (one-way)
-    \emph{bridge}:
-    \[ \xymatrix{
-      & \text{$\ul{\TT}/U_\TT$ proves$^\star$ $\sigma$} \\
-      \text{$\sigma$ holds for $U_\TT$} \ar@/^1.1pc/@{<=>}[ur] &&
-      \text{$\sigma$ holds for $M$} \ar@/_1.1pc/@{<=}[ul]
-    } \]
-  }
-\end{frame}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\hfill\mbox{\includegraphics[width=0.13\paperwidth]{hen}\;\;}\end{minipage}}
-\begin{frame}{Exhaustion and extensions}
-  \justifying
-  \textbf{Theorem.} A first-order formula holds for~$U_\TT$ iff it is
-  intuitionistically provable from the axioms of~$\TT$ and
-  the scheme
-  \begin{equation}\tag{Nullstellensatz} \speak{$\sigma$ holds} \implies \speak{$\ul{\TT}/U_\TT$
-  proves$^{\star\star}$~$\sigma$}. \end{equation}
-
-  \textbf{Theorem.}
-  Let~$\TT'$ be a quotient theory
-  of~$\TT$. Assume that~$U_\TT$ is contained in the subtopos~$\Set[\TT']$. Then internally
-  to~$\Set[\TT']$:
-  \begin{quote}
-    \emph{A geometric$^\star$ sequent~$\sigma$ with Horn consequent holds for~$U_{\TT'}$
-    iff~$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$.}
-  \end{quote}
-  \vspace*{-1.6em}
-
-  \textbf{Theorem.}
-  % Let~$\mathrm{Form}^\star_{x\?X}(\ul{\TT}/U_\TT)/(\dashv\vdash_{x\?X})$ be the~$\Set[\TT]$-object of
-  % geometric$^\star$ formulas over the signature of~$\ul{\TT}/U_\TT$ in the
-  % context~$x\?X$, where any two such formulas are identified if and only
-  % if~$\ul{\TT}/U_\TT$ proves them equivalent.
-  The morphism~$\mathrm{ev}$ is an isomorphism.
-  \[ \mathrm{ev} : \mathrm{FunctFormulas}^\star(\ul{\TT}/U_\TT)/(\dashv\vdash)
-  \longrightarrow P(U_\TT) \]
-
-  \textbf{Theorem.} A higher-order formula holds for~$U_\TT$ iff it is
-  provable in intuitionistic higher-order logic from the axioms of~$\TT$ and
-  the higher-order Nullstellensatz scheme.
-
-  \jnote{1}{
-    The first theorem states that the source provided by the Nullstellensatz is
-    exhaustive. The notion of provability$^{\star\star}$ is a strengthening of
-    the notion of provability$^\star$, which in turn is a strengthening of the
-    ordinary notion of provability in geometric logic. We are using it here
-    because while the notion of provability$^\star$ can be expressed in the
-    internal language of a topos, it cannot be expressed in intuitionistic
-    logic.
-
-    The second theorem provides a useful variant of the Nullstellensatz. Its
-    assumptions are for instance satisfied if~$\TT$ is the theory of rings
-    and~$\TT'$ is the theory of local rings. When applicable, it can be used to
-    avoid doubly-internal toposes. It also explains, for instance, why in the
-    formulation of synthetic quasicoherence no local rings appear even though
-    the relevant topos is the classifying topos of local rings.
-
-    The third and fourth theorems generalise the Nullstellensatz to the
-    higher-order setting. The map~$\mathrm{ev}$ takes (the equivalence class
-    of) a~$\ul{\TT}/U_\TT$-provably$^\star$ geometric$^\star$ formula~$\theta$
-    in one free variable to the subset~$\{ x \? U_\TT \,|\, \theta(x) \}$.
-
-    Details on all of this are slowly emerging at
-    \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-qcoh.pdf}{this
-    preprint}.
-  }
-
-  \jnote{2}{
-    XXX future work
-  }
-\end{frame}}
-
-% Analogy with the harmonic oscillator in the study of physics
-
-% * The generic model
-%   * The mystery of nongeometric sequents
-%   * Construction of the generic model
-% * Kripke--Joyal semantics; focus on existence of extensions
-% * Examples for theories and nongeometric sequents with a focus on the wlog
-%   aspect
-% * Why we are interested in the generic models themselves
-%   * Universal localisation
-%   * Grothendieck's generic freeness lemma
-% * Synthetic quasicoherence
-% * Conservativity from the internal point as guiding question
-%   * Naive formulation and why it doesn't work
-%   * Recall of the algebraic Nullstellensatz
-% * The new Nullstellensatz
-%   * What Set[T][T/U_T] classifies
-%   * Kock again
-%   * Subtleties regarding infinites
-%   * Improvement in the Horn case
-% * Universality of the source provided by the Nullstellensatz
 
@@
 
-% Are there size issues when KJ'ing internally?
-% Comparison to the approach by Vickers?
-% When does a theory prove the Nullstellensatz?
-% Look up what Kock and Reyes used the wlog for.
+addictive
 
-% An Tafel: geometric formula, local ring
-% Erwähnen: Marc Bezem, Ulrik Buchholtz, Thierry Coquand
+do not learn Agda!
 
-% Grothendieck quote for the title picture:
-%
-% Ces “nuages probabilistes”, remplaçant les rassurantes particules matérielles
-% d’antan, me rappellent étrangement les élusifs “voisinages ouverts” qui
-% peuplent les topos, tels des fantômes évanescents, pour entourer des “points”
-% imaginaires, ...
-%
-% Diese "Wahrscheinlichkeitswolken", welche die beruhigenden materiellen
-% Partikel von früher ersetzen, erinnern mich irgendwie an die flüchtigen
-% "offenen Umgebungen" der Topoi -- wie dahinschwindende Phantome, um
-% die fiktiven "Punkte" zu umgeben, ...
-%
-% These “probability clouds”, replacing the reassuring material particles of
-% before, remind me strangely of the elusive “open neighborhoods” that populate
-% the topoi, like evanescent phantoms, to surround the imaginary “points”.
+Hello! I'm Ingo Blechschmidt, and I'm here to tell you about what's probably
+the single coolest thing I've learned in the last couple years. I'm here to
+tell you about Agda, a system put forward by Catarina Coquand in 1999.
 
-% Sei T eine Theorie, für die ein konservatives Modell in Set existiert.
-% Dann haben wir eine Surjektion Set --> Set[T]. Also existiert eine
-% geometric expansion von T, welche Morita-äquivalent zur leeren Theorie ist.
+First and foremost, Agda is an ordinary programming language. You can write
+computer programs in Agda.
+
+But, much more to the point, Agda is also a proof language. You can write
+mathematical proofs in Agda, and Agda will check their correctness.
+Over the years, using Agda or similar systems, substantial errors and gaps in
+published mathematical proofs have been found.
+
+Agda is based on dependent type theory which is introduced in the first part of
+the course by professor Maietti. Using Agda, you can actually play with
+type theory in the computer, and unlike your pen or your paper, Agda will
+always give you immediate feedback.
+
+% There is only one catch. Namely, Agda is somewhat hard to learn on one's own.
+% Sometimes, error messages are cryptic and documentation is sparse. Normally,
+% this would be a reason to worry. But in this case, you don't need to worry!
+
+We will learn Agda in a series of practical sessions. It will be easy and fun!
+
+If you are a computer scientist, learning Agda will help you understand type
+theory and architect correct programs, programs which will not go wrong.
+
+And if you are a mathematician, by learning Agda you will appreciate mathematics
+from a new point of view, enjoying proofs from an angle you haven't yet looked at.
+
+Please feel invited. Questions will always be welcome. We will do this together.
+Agda is pure elegance, and I'm sure you will enjoy it.

### 2020-02-20 paper-filmat.tex
With proper title

@@
 
-%\title[]{Exploring mathematical objects from custom-tailored mathematical universes}
-%\author{Ingo Blechschmidt}
-%\address{Universität Augsburg \\
-%Institut für Mathematik \\
-%Universitätsstr. 14 \\
-%86159 Augsburg, Germany}
-%\email{ingo.blechschmidt@math.uni-augsburg.de}
+\title[]{Exploring mathematical objects from custom-tailored mathematical universes}
+\author{Ingo Blechschmidt}
+\address{Universität Augsburg \\
+Institut für Mathematik \\
+Universitätsstr. 14 \\
+86159 Augsburg, Germany}
+\email{ingo.blechschmidt@math.uni-augsburg.de}
 
@@
 
-%\maketitle
+\maketitle
 \thispagestyle{empty}
-
-\begin{center}
-  \begin{tikzpicture}
-    \node (comment) at (0,0) {
-      \begin{minipage}{0.6\textwidth}
-      \emph{To the anonymous referees:}
-      \bigskip
-
-      Thank you for your volunteer work. It is very much appreciated. I am looking
-      forward to your comments.
-      \bigskip
-
-      \hfill -- the anonymous author
-      \end{minipage}
-    };
-
-    \node[draw=mypurple, very thick, rectangle, rounded corners, inner sep=5pt] [fit = (comment)] (foo) {};
-  \end{tikzpicture}
-\end{center}
-\bigskip
 

### 2020-03-25 slides-herrsching2019.tex
typo

@@
 
+% XXX In note slides, explain "iff*".
+
 \begin{frame}{Examples}
@@
 
+% XXX in note slides, explain which space X to take
+
 \begin{frame}[plain]
@@
     arXiv} and warmly recommended, as are all slides by Thierry Coquand on
-    constructive algebra.«
+    constructive algebra.
   }

### 2020-03-27 paper-herrsching2019/main.tex
Stub for Herrsching 2019 notes

\documentclass{ws-rv9x6}
\usepackage{ws-rv-thm}
\usepackage{ws-rv-van}
\makeindex
%\newindex{aindx}{adx}{and}{Author Index}       % author index
%\renewindex{default}{idx}{ind}{Subject Index}  % subject index

\begin{document}

\chapter{Generalized spaces for constructive algebra}

\author[I. Blechschmidt]{Ingo Blechschmidt}
% \index[aindx]{Blechschmidt, I.}

\address{Universität Augsburg \\
Institut für Mathematik \\
Universitätsstr. 14 \\
86159 Augsburg, Germany}

\begin{abstract}
foo
\end{abstract}
%\markright{Customized Running Head for Odd Page} % default is Chapter Title.
\body

%\tableofcontents

\section{Locales}

\section{Sheaf models}

\section{Applications in constructive algebra}

\end{document}

### 2020-03-28 paper-herrsching2019/main.tex
Section on locales

@@
 \documentclass{ws-rv9x6}
-\usepackage{ws-rv-thm}
+\usepackage[onethmnum]{ws-rv-thm}
 \usepackage{ws-rv-van}
+\usepackage{mathtools,stmaryrd}
 \makeindex
@@
 %\renewindex{default}{idx}{ind}{Subject Index}  % subject index
+
+\newcommand{\A}{\mathcal{A}}
+\newcommand{\B}{\mathcal{B}}
+\newcommand{\C}{\mathcal{C}}
+\newcommand{\M}{\mathcal{M}}
+\renewcommand{\AA}{\mathbb{A}}
+\newcommand{\E}{\mathcal{E}}
+\newcommand{\F}{\mathcal{F}}
+\newcommand{\G}{\mathcal{G}}
+\newcommand{\J}{\mathcal{J}}
+\newcommand{\GG}{\mathbb{G}}
+\renewcommand{\O}{\mathcal{O}}
+\newcommand{\K}{\mathcal{K}}
+\newcommand{\NN}{\mathbb{N}}
+\newcommand{\QQ}{\mathbb{Q}}
+\newcommand{\RR}{\mathbb{R}}
+\newcommand{\TT}{\mathbb{T}}
+\newcommand{\PP}{\mathbb{P}}
+\newcommand{\ZZ}{\mathbb{Z}}
+\newcommand{\CC}{\mathbb{C}}
+\renewcommand{\P}{\mathcal{P}}
+\newcommand{\aaa}{\mathfrak{a}}
+\newcommand{\ppp}{\mathfrak{p}}
+\newcommand{\fff}{\mathfrak{f}}
+\newcommand{\defeq}{\vcentcolon=}
+\newcommand{\defeqv}{\vcentcolon\equiv}
+\newcommand{\Sh}{\mathrm{Sh}}
+\newcommand{\GL}{\mathrm{GL}}
+\newcommand{\Zar}{\mathrm{Zar}}
+\newcommand{\op}{\mathrm{op}}
+\newcommand{\Set}{\mathrm{Set}}
+\newcommand{\Eff}{\mathrm{Ef{}f}}
+\newcommand{\Sch}{\mathrm{Sch}}
+\newcommand{\Aff}{\mathrm{Aff}}
+\newcommand{\Ring}{\mathrm{Ring}}
+\newcommand{\LocRing}{\mathrm{LocRing}}
+\newcommand{\LRS}{\mathrm{LRS}}
+\newcommand{\Hom}{\mathrm{Hom}}
+\newcommand{\Spec}{\mathrm{Spec}}
+\newcommand{\Gal}{\mathrm{Gal}}
+\newcommand{\lra}{\longrightarrow}
+\newcommand{\RelSpec}{\operatorname{Spec}}
+\renewcommand{\_}{\mathpunct{.}}
+\newcommand{\?}{\,{:}\,}
+\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
+\newcommand{\ul}[1]{\underline{#1}}
+\newcommand{\affl}{\ensuremath{{\ul{\ensuremath{\AA}}^1}}}
+\newcommand{\Ll}{\text{iff}}
+\newcommand{\inv}{inv.\@}
+\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}
+\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates
+\newcommand{\brak}[1]{{\llbracket{#1}\rrbracket}}
+\newcommand{\pt}{\mathrm{pt}}
+\newcommand{\Loc}{\mathrm{Loc}}
+\newcommand{\Top}{\mathrm{Top}}
 
@@
 \begin{abstract}
-foo
+XXX
 \end{abstract}
-%\markright{Customized Running Head for Odd Page} % default is Chapter Title.
 \body
@@
 
+The notion of \emph{space} is fundamental to large parts of mathematics. It
+exists in various flavors, ranging from the basic cartesian spaces~$\RR^n$ to
+the more general metric and topological spaces and also including several more
+and slightly exotic flavors such as the diffeological spaces of differential topology.
+
+Common to all of the mentioned flavors of space is that \emph{points are their
+building blocks}: Cartesian, metric, topological and diffeological spaces are,
+first and foremost, sets of points. Their spatial structure -- a metric, a
+topology or a diffeology -- is additional data to the underlying set:
+
+\begin{definition}A \emph{metric space} consists of a set~$X$ of points
+together with a metric~$d : X \times X \to \RR_{\geq 0}$ satisfying the metric
+axioms. \end{definition}
+
+\begin{definition}A \emph{topological space} consists of a set~$X$ of points
+together with a set~$\O(X) \subseteq \P(X)$ of point sets which are deemed
+\emph{open} such that (arbitrary, set-indexed) unions and finite intersections
+of open sets are open.
+\end{definition}
+
+Locales are a further and particularly unique flavor of the notion of space
+which turn this classical picture upside down: Locales embrace \emph{opens
+instead of points as primitive building blocks}. With locales, points are a
+derived concept. In particular, the opens of a locale are not sets of points;
+in fact, they needn't be sets of anything in peculiar.
+
+The formal definition of a locale will be given below as
+Definition~\ref{defn:locale}, but before we will review examples and comment on
+the relevance of locales to constructive mathematics.
+
+\subsection{Spaces without points}
+\label{sect:examples-no-points}
+
+A metric or topological space without any points is not very interesting: It is
+empty, and up to isomorphism there is only one such space. In contrast, a
+locale can be nontrivial even if it does not contain any points. This
+phenomenon is one instance of the general guiding principle that relinquishing
+points increases flexibility.
+
+\paragraph{The locale of surjections~$\NN \twoheadrightarrow \RR$.} As is
+well-known, there are no surjections from~$\NN$ to~$\RR$.\footnote{More
+precisely, it is a theorem of classical mathematics that the reals are
+uncountable in the sense of admitting no surjection from the naturals. The
+situation is more subtle in constructive mathematics. Firstly, in the absence
+of countable choice, the reals bifurcate into several distinct flavors, hence
+one needs to state which flavor of the reals one is referring to. Secondly, all
+known proofs of the uncountability of the Cauchy and the Dedekind reals assume
+either the law of excluded middle or the axiom of countable choice. In the
+absence of either of these axioms, only the MacNeille reals are known to be
+uncountable~\cite{blechschmidt-hutzler:macneille}.} Hence there is no
+interesting topological space of those surjections. However, there is a
+well-defined and nontrivial locale~$X$ of those surjections.
+
+The points of~$X$ are in canonical one-to-one correspondence with the
+surjections~$\NN \twoheadrightarrow \RR$, hence~$X$ does not have any points.
+But this locale does have uncountably many basic opens~$U_{n,x}$, where~$n$
+ranges over the naturals and~$x$ ranges over the reals. We picture~$U_{n,x}$ as
+the open of those surjections~$f$ for which~$f(n) = x$, we can compute with
+these opens and consider functions on~$X$. For instance:
+\begin{enumerate}
+\item If~$x \neq y$, then the intersection of~$U_{n,x}$ with~$U_{n,y}$
+is truly empty.
+\item The union of the~$U_{n,x}$, where~$x$ is a fixed real
+number and~$n$ ranges over the naturals, is all of~$X$. No finite number
+of these opens covers~$X$, hence~$X$ is not compact.
+\item For any real~$x$, there is a well-defined continuous function from~$X$ to
+(the localic version of) the naturals which, on the level of points, would map
+a surjection~$f : \NN \to \RR$ to the smallest preimage of~$x$.
+\item There is a well-defined continuous function from~$X$ to (the localic
+version of) the reals which, on the level of points, would map a surjection~$f$
+to the number~$\sum_{n=0}^\infty 2^{-n} \arctan(f(n))$. Unlike the previous
+example, this function is not locally constant.
+\item\label{item:intersection-sublocales} The locale~$X$ is a sublocale of the
+locale~$Y$ of arbitrary functions~$\NN \to \RR$ (which can also be realized as
+a topological space). It can be obtained as the intersection of the uncountably
+many sublocales~$Y_x$, where~$Y_x$ is the sublocale of~$Y$ consisting of those
+functions which hit the real~$x$.
+\item A certain sublocale of~$X$, the locale~$X'$ of those functions~$\NN \to
+\RR$ for which any real has infinitely many preimages, has a fractal nature: It
+is covered by the~$U_{0,x}$, where~$x$ ranges over the reals; the pairwise
+intersection of these opens is (truly) empty; and they are each isomorphic
+to~$X'$ -- on the level of points, by mapping a surjection~$f$ in~$U_{0,x}$ to
+the surjection~$(n \mapsto f(n+1))$.
+\end{enumerate}
+
+There is nothing special about the real numbers which makes this example work;
+in fact, the example works just as well with any set~$M$ in place of the reals.
+For any set~$M$, there is a locale of surjections~$\NN \twoheadrightarrow M$,
+and this locale is trivial (isomorphic to the empty locale) if and only if~$M$
+is empty. Locales of these kind are used as an important reduction step in the
+extension of Grothendieck's Galois theory by Joyal and
+Tierney~\cite[Section~V.3]{joyal-tierney:galois-theory}.
+
+\paragraph{Intersection of dense sublocales.} In ordinary topology, the
+intersection of dense subspaces need not be dense. A simple example is the
+intersection of~$\QQ$ with its complement in~$\RR$. In contrast, the
+intersection of (even an arbitrary set of) dense sublocales is always again
+dense -- even if the intersection might well have no points.
+
+For instance, the locale-theoretic intersection of (the localic version
+of)~$\QQ$ with its complement does not have any points and is dense in (the
+localic version of) the reals. Intuitively, while these two sublocales do not
+have any points in common, there still is nontrivial ``localic glue''.
+
+% This basic fact about locale-theoretic denseness illustrates a more general
+% phenomenon of locales: Locales tend to enjoy better formal properties.
+
+Another example is given by item~(\ref{item:intersection-sublocales}) above:
+Each of the sublocales~$Y_x$ is dense in~$Y$, hence their intersection~$X$ is
+so as well.
+
+\paragraph{The Banach--Tarski paradox.} The Banach--Tarski paradox is the
+unintuitive statement that a three-dimensional solid ball in~$\RR^3$ of
+radius~$r$ can be partitioned into six disjoint subsets in such a way that
+rearranging those subsets using only euclidean motions yields two disjoint
+solid balls of radius~$r$ each. The axiom of choice is required to construct
+these subsets, and the Banach--Tarski paradox is not in
+contradiction with the basic properties of the (Lebesgue) measure in~$\RR^3$
+because these intermediate subsets are not measurable.
+
+The traditional way to avoid the Banach--Tarski paradox is to adopt the
+\emph{axiom of determinacy} instead of the axiom of choice. Just as the axiom
+of choice posits that a certain property of the finite domain also holds for the
+infinite, the axiom of determinacy is a certain statement whose finitary
+analogue is provable in unadorned Zermelo--Fraenkel set theory. It entails that
+all subsets of~$\RR^n$ are measurable.
+
+The Banach--Tarski paradox can also be avoided by adopting a localic point of
+view: While the localic counterparts of the six pieces do not have any points
+in common, the locale-theoretic pairwise intersections are still nontrivial.
+Hence one would not expect the rearrangement of these sublocales to have the
+same volume as the original solid ball. [XXX: give reference]
+
+\paragraph{Random sequences.} XXX
+
+\subsection{Constructive concerns}
+
+By relinquishing points, locales provide a more flexible notion of space.
+Section~\ref{sect:examples-no-points} illustrates[XXX:word] this observation
+with several examples of nontrivial locales without any points. Constructive
+mathematics gives a further, orthogonal motivation to study locales: There are
+situations in which the relevant spaces do have enough points, but only if one
+subscribes to the axiom of choice or similar non-constructive principles. In
+these situations, the pointfree approach allowed by locales can help to give
+constructive versions of classical results.
+
+XXX ideal objects
+
+\paragraph{Compactness of the unit interval.} The unit interval, when realized
+as a topological space, can fail to be be compact in constructive
+mathematics. For instance, it fails in the Russian school because the Kleene
+tree provides a computable open covering of~$[0,1]$ with no computable finite
+subcovering [XXX: reference].
+
+In contrast, the localic version of the unit interval is always compact. The
+proof is by an explicit computation with its basic opens. [XXX: reference]
+
+\paragraph{Galois theory.} Let~$L|k$ be a Galois extension. The
+fundamental theorem of Galois theory states that there is a bijection between
+the intermediate extensions~$L|E|k$ and the closed subgroups of the topological
+Galois group~$\Gal(L|k)$. The bijection maps an intermediate extension~$L|E|k$
+to the subgroup~$\Gal(L|E)$ and its inverse maps a closed subgroup~$H \subseteq
+\Gal(L|k)$ to the intermediate extension~$L^H$.
+
+Much of the proof of the fundamental theorem of Galois theory is constructive,
+but some parts use the law of excluded middle and Zorn's lemma in order to
+construct certain extensions of given field homomorphisms.\footnote{For
+instace, the statement~$E \subseteq L^{\Gal(L|E)}$ is trivial. For the converse
+inclusion, let~$x \in L^{\Gal(L|E)}$. Assume for the sake of contradiction
+that~$x \not\in E$. Using Zorn's lemma and the law of excluded middle, we find
+a homomorphism~$\sigma : L \to L$ with~$\sigma|_E = \operatorname{id}$
+and~$\sigma(x) \neq x$. This is a contradiction to~$x \in L^{\Gal(L|E)}$.}
+As a consequence, the fundamental theorem of Galois theory is not provable in
+constructive mathematics.
+
+However, this failure is not for fundamental Galois-theoretic reasons, but because of an
+unfortunate choice in the definitions. There is a notion of a localic group (a
+locale~$G$ equipped with continuous maps~$G \times G \xrightarrow{\circ} G$, $G
+\xrightarrow{(\cdot)^{-1}} G$, $1 \xrightarrow{e} G$ satisfying the group
+axioms), and the topological Galois group has a localic counterpart. The
+fundamental theorem can be reformulated to refer to this localic Galois group,
+and the proof of this reformulation is entirely constructive.
+[XXX: references]
+
+If one is so inclined, then one can obtain the topological version of the
+fundamental theorem as a corollary of the localic version; the required
+non-constructive principles for this step are neatly packaged up in the study
+of the relation of the localic Galois group with the topological one.
+
+\begin{remark}Incidentally, the fundamental theorem of Galois theory also
+showcases a related general phenomenon, namely that classical mathematics
+allows to push back topological concerns for longer than in constructive
+mathematics, whereas in constructive mathematics we have to embrace topology
+(in a sufficiently pointfree form such as locales) from the beginning.
+
+To be more specific, classically, the basic version of the fundamental theorem
+(intermediate extensions correspond to subgroups) only holds for finite Galois
+extensions. For infinite Galois extensions, we have to restrict to
+\emph{closed} subgroups. Constructively, the basic version cannot even be shown
+for finite extensions; we have to employ spatial language even for those, or
+else settle for a discrete version of the fundamental theorem: For finite field
+extensions, finite intermediate extensions correspond to finite subgroups.
+Classically, intermediate extensions of finite extensions and subgroups of
+finite groups are automatically finite, but constructively this can fail.
+\end{remark}
+
+\paragraph{Gelfand duality} XXX
+
+\subsection{The basics of the theory of locales}
+
+The starting point to arrive at the definition of a locale is the following
+fundamental observation: The partial order~$\O(X)$ of open subsets of a topological
+space~$X$ has
+\[ \begin{array}{@{}ccc@{}}
+\text{arbitrary joins}
+&\text{and}&
+\text{finite meets}, \\
+\bigvee && \wedge
+\end{array} \]
+and finite meets distribute over arbitrary joins:
+\[ U \wedge \bigvee_i V_i = \bigvee_i (U \wedge V_i). \]
+
+This observation motivates the following definition. [XXX: which abstracts...]
+\begin{definition}A \emph{frame} is a partial order with (arbitrary,
+set-indexed) joins and finite meets such that the distributive law
+holds. A \emph{frame homomorphism~$\alpha : A \to A'$} is a monotone map~$A
+\to A'$ which preserves arbitrary joins and finite meets.\end{definition}
+
+The notion of a frame is (infinitarily) algebraic. To obtain a geometric
+notion, we ``reverse the direction of the arrows'':
+
+\begin{definition}\label{defn:locale}
+A \emph{locale}~$X$ is given by a frame~$\O(X)$, the ``frame
+of opens of~$X$''. A \emph{morphism~$f : X \to X'$ of locales} (or ``continuous
+map of locales'') is a frame homomorphism~$\O(X') \to \O(X)$.\end{definition}
+
+In place of the open sets of points, locales have arbitrary \emph{opens}, the
+elements of their underlying frame. The opens of locales behave similar to the
+open sets in topology in that arbitrary unions and finite intersections make
+sense; but unlike before, they need not be sets of points.
+
+Examples for locales include the following.
+\begin{enumerate}
+\item Any topological space~$Y$ induces a locale~$L(Y)$ by
+setting~$\O(L(Y)) \defeq \O(Y)$. A continuous map~$f : Y \to Y'$ of topological
+spaces induces the frame homomorphism~$\O(Y') \to \O(Y),\,U \mapsto f^{-1}[U]$
+in the other direction and hence a morphism~$L(Y) \to L(Y')$ of locales in the
+same direction.
+\item The \emph{one-point locale}~$\pt$ is the locale induced
+by the one-point topological space~$\{\star\}$. Its frame of opens is
+the powerset of~$\{\star\}$, also known as the set~$\Omega$ of truth
+values. Its least element is~$\bot = \emptyset$ and its largest element
+is~$\top = \{\star\}$, and potentially not all elements of~$\Omega$ are equal
+to of these two.
+\item The locale of surjections~$\NN \twoheadrightarrow \RR$ and the localic
+version of the reals of Section~\ref{sect:examples-no-points} are best
+constructed as \emph{classifying locales}, a notion to be introduced below.
+\end{enumerate}
+
+Points do not appear in the definition of a locale, but they can be defined as
+a derived concept:
+\begin{definition}A \emph{point} of a locale~$X$ is a locale morphism~$\pt \to
+X$.\end{definition}
+
+The underlying frame homomorphism~$\alpha : \O(X) \to \Omega$ of a point~$x$
+of~$X$ can be pictured as mapping each open of~$X$ to the truth value to which
+extent~$x$ belongs to~$U$. Symbolically, we write~$x \inplus u$ iff~$\alpha(u)
+= \top$.
+
+The point~$x$ is completely determined by the
+set~$\{ u \in \O(X) \,|\, \alpha(u) = \top \}$. This set is a \emph{completely
+prime filter}, that is a subset~$\fff \subseteq \O(X)$ which is upward-closed,
+closed under finite meets and for which~$\bigvee_i u_i \in \fff$ implies~$u_i
+\in \fff$ for some index~$i$; and conversely, any such completely prime filter
+gives rise to a point of~$X$.
+
+\begin{definition}A locale~$X$ is \emph{spatial} if and only if its points
+suffice to distinguish its opens, that is if for any opens~$u, v \in \O(X)$,
+if~$x \inplus u \Leftrightarrow x \inplus v$ for all points~$x$ of~$X$, then~$u =
+v$.\end{definition}
+
+For instance, any locale induced by a topological space is spatial. The locale
+of surjections~$\NN \to \RR$ is a drastic example of a locale which fails to be
+spatial, and the localic real line is spatial in classical mathematics and can
+fail to be spatial in constructive mathematics.
+
+\begin{remark}The set of points of a locale~$X$ can be made into a topological space, giving
+rise to a functor~$\pt : \Loc \to \Top$. This functor is right adjoint
+to the functor~$L : \Top \to \Loc$. A locale~$X$ is spatial iff the canonical
+morphism~$L(\pt(X)) \to X$ is an isomorphism, and a topological space~$Y$
+is \emph{sober} iff the canonical morphism~$Y \to \pt(L(Y))$ is a
+homeomorphism. The space~$\pt(L(Y))$ is the \emph{sobrification} of~$Y$; for
+instance, the sobrification of any inhabited indiscrete space is the one-point
+space.\end{remark}
+
 \section{Sheaf models}
@@
 \end{document}
+
+XXX cite https://apcz.umk.pl/czasopisma/index.php/LLP/article/viewFile/LLP.2013.009/732

### 2020-03-28 paper-filmat.tex
typo

@@
 is an obvious counterexample to the statement ``any map~$\RR \to \RR$ is
-continuous''. However, closer inspection reveals that the sign function
+continuous''. However, a closer inspection reveals that the sign function
 cannot be proven to be a total function~$\RR \to \RR$ if only intuitionistic
@@
 \leq 2^{-n} + 2^{-m}$. Intuitively, such a machine~$X$ denotes the real
-number~$\lim_{n \to \infty} X(n)$, and the approximations~$X(n)$ must be
+number~$\lim_{n \to \infty} X(n)$, and each approximation~$X(n)$ must be
 within~$2^{-n}$ of the limit.
@@
 = 0$'', outputs a witness of~``$\exists n\?\NN\_ f(n) = 0$'' (up to trivial
-conversions, a number~$n$ such that~$f(n) = 0$).
+conversions, this is a number~$n$ such that~$f(n) = 0$).
 
@@
   number which is neither negative nor zero nor positive. Externally speaking,
-  there is no covering of the unit interval by open subsets on which the
+  there is no covering of the unit interval by opens on which the
   depicted function~$a$ is either everywhere negative, everywhere zero or everywhere
@@
   parameter space is~$X = [0,1]$ (not shown) and the other functions~$f_x$ are
-  obtained from~$f_{x_0}$ by moving the horizontal plateau further up or further
+  obtained from~$f_{x_0}$ by moving the horizontal plateau further up or
   down. The depicted member~$f_{x_0}$ has a unique
@@
 
-In contrast, the intermediate value theorem for monotone functions does have an
+In contrast, the intermediate value theorem for (strictly) monotone functions does have an
 intuitionistic proof and hence applies in the internal universe of~$\Sh(X)$.
@@
 
-\subsection{Hyperreal numbers} There are several ways for introducing
+\subsection{Hyperreal numbers} There are several ways of introducing
 infinitesimals into rigorous mathematics. One is Robinson's \emph{nonstandard
@@
 the Zariski sheaves. Such a functor is a \emph{Zariski sheaf} if and only if,
-for any covering~$(A[f_i^{-1}])_i$ of any finitely presented~$\RR$-algebra~$A$,
-the diagram
+for any covering~$(A[f_i^{-1}])_i$ of any finitely presented~$\RR$-algebra~$A$
+(this notion is defined in Table~\ref{table:zariski}), the diagram
 \[ F(A) \longrightarrow \prod_i F(A[f_i^{-1}]) \xbigtoto{} \prod_{j,k}
@@
 And indeed, this statement is true. Let a stage~$A$ (a finitely
-presented~$\RR$-algebra) and an element~$x \in A$ which fulfills the stated
-condition be given. Let~$B$ be any later stage of~$A$ (any finitely
+presented~$\RR$-algebra) and an element~$x \in A$
+be given. Let~$B$ be any later stage of~$A$ (any finitely
 presented~$A$-algebra -- such an algebra is also finitely presented as
-an~$\RR$-algebra). Then~$x = 0$ holds in the refinement~$C \defeq B/(x)$.
+an~$\RR$-algebra). Assume that for any later stage~$C$ of~$B$ in which~$x = 0$
+holds also~$1 = 0$ holds. Trivially,~$x = 0$ holds in the particular refinement~$C \defeq B/(x)$.
 Hence~$1 = 0$ holds in~$C$. By elementary algebra, this means that~$x$ is
 invertible in~$B$. Hence the conclusion holds for the singleton covering
-of~$B$ by~$C_1 \defeq B$.
+of~$B$ given by~$C_1 \defeq B$.
 

### 2020-03-28 paper-filmat.tex
Details on the injection ℝ → ℕ, prompted by Franka Waaldijk
MO thread: https://mathoverflow.net/questions/355936/contrasting-theorems-in-classical-logic-and-constructivism/

@@
 there is no surjection~$\NN \to \RR$, there is an injection~$\RR \to
-\NN$~\cite{bauer:injection}.
+\NN$~\cite{bauer:injection}.\footnote{What the realizability topos
+given by infinite-time Turing machines believes to be a real number is, from
+the external point of view, an infinite-time Turing machine~$X$ which outputs,
+when called with a natural number~$n$ as input, a rational
+approximation~$X(n)$. As with the original \effective topos, these
+approximations have to be consistent in the sense that~$|X(n) - X(m)| \leq
+2^{-n} + 2^{-m}$, and two such machines~$X, X'$ represent the same real
+iff~$|X(n) - X'(m)| \leq 2^{-n} + 2^{-m}$ for all natural numbers~$n,m$.
+
+A map~$\RR \to \NN$ in this topos is hence given by an infinite-time
+Turing machine~$M$ which, when given the description of such an infinite-time
+Turing machine~$X$ as input, outputs a certain natural number~$M(X)$. If~$X$
+and~$X'$ represent the same real, then~$M(X)$ has to coincide with~$M(X')$.
+This map~$\RR \to \NN$ is injective iff conversely,~$M(X) = M(X')$ implies
+that~$X$ and~$X'$ represent the same real.
+
+We can program such a machine~$M$ as follows: Read the description of an
+infinite-time Turing machine~$X$ representing a real number as input. Then
+simulate, in a dovetailing fashion, all infinite-time Turing machines and
+compare their outputs with the outputs of~$X$. As soon as a machine~$X'$ is
+found which happens to terminate on all inputs in such a way that~$|X(n) -
+X'(m)| \leq 2^{-n} + 2^{-m}$ for all natural numbers~$n,m$, output the number
+of this machine (in the chosen enumeration of all infinite-time Turing
+machines) and halt.
+
+The number~$M(X)$ computed by~$M$ depends on the input/output behaviour of~$X$,
+the chosen order of infinite-time Turing machines, and on details of the
+interleaving simulation and comparison procedure -- but it does not depend on
+the implementation of~$X$ or on its specific choice of rational
+approximations~$X(n)$. The search terminates since there is at least one
+infinite-time Turing machine which represents the same real number as~$X$ does,
+namely~$X$ itself.}
 

### 2020-03-29 paper-herrsching2019/main.tex
Section on sheaf models

@@
 \usepackage{mathtools,stmaryrd}
+
+\usepackage{tikz}
+\usetikzlibrary{calc}
+\newcommand{\hcancel}[5]{%
+  \tikz[baseline=(tocancel.base)]{
+    \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
+    \draw[red!80, line width=0.4mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
+  }%
+}
+
 \makeindex
@@
 \newcommand{\aaa}{\mathfrak{a}}
+\newcommand{\bbb}{\mathfrak{b}}
 \newcommand{\ppp}{\mathfrak{p}}
@@
 radius~$r$ can be partitioned into six disjoint subsets in such a way that
-rearranging those subsets using only euclidean motions yields two disjoint
+rearranging those subsets using only Euclidean motions yields two disjoint
 solid balls of radius~$r$ each. The axiom of choice is required to construct
@@
 proof is by an explicit computation with its basic opens. [XXX: reference]
+
+More generally, Tychonoff's theorem that the product of any number of compact
+topological spaces is again compact is equivalent to the axiom of choice. Its
+localic counterpart, the statement that the product of any number of compact
+locales is again compact, can be verified without any nonconstructive
+principles. [XXX: reference]
 
@@
 
-The notion of a frame is (infinitarily) algebraic. To obtain a geometric
-notion, we ``reverse the direction of the arrows'':
+The least element of a frame (the empty join) is denoted~``$\bot$'' and the
+largest element (the empty meet)~``$\top$''. The notion of a frame is
+(infinitarily) algebraic. To obtain a geometric notion, we ``reverse the
+direction of the arrows'':
 
@@
 
+Several notions in topology only refer to open sets and not to points. Such
+notions have an immediate counterpart in locale theory. For instance, a
+locale~$X$ is \emph{compact} iff for any family~$(U_i)_{i \in I}$ of opens
+of~$X$ such that~$\top = \bigvee_{i \in I} U_i$, there is a (Kuratowski-)finite
+subset~$I' \subseteq I$ such that~$\top = \bigvee_{i \in I'} U_i$. A
+topological space~$Y$ is compact iff its induced locale~$L(Y)$ is.
+
 Points do not appear in the definition of a locale, but they can be defined as
@@
 
+This definition is inspired from the situation with topological spaces, where
+continuous maps from the one-point space to a topological space~$Y$ are in
+canonical bijection with the points of~$Y$.
+
 The underlying frame homomorphism~$\alpha : \O(X) \to \Omega$ of a point~$x$
-of~$X$ can be pictured as mapping each open of~$X$ to the truth value to which
-extent~$x$ belongs to~$U$. Symbolically, we write~$x \inplus u$ iff~$\alpha(u)
+of~$X$ can be pictured as mapping each open~$U$ of~$X$ to the truth value to which
+extent~$x$ belongs to~$U$. Symbolically, we write~$x \inplus U$ iff~$\alpha(U)
 = \top$.
@@
 The point~$x$ is completely determined by the
-set~$\{ u \in \O(X) \,|\, \alpha(u) = \top \}$. This set is a \emph{completely
+set~$\{ U \in \O(X) \,|\, \alpha(U) = \top \}$. This set is a \emph{completely
 prime filter}, that is a subset~$\fff \subseteq \O(X)$ which is upward-closed,
-closed under finite meets and for which~$\bigvee_i u_i \in \fff$ implies~$u_i
+closed under finite meets and for which~$\bigvee_i U_i \in \fff$ implies~$U_i
 \in \fff$ for some index~$i$; and conversely, any such completely prime filter
@@
 \begin{definition}A locale~$X$ is \emph{spatial} if and only if its points
-suffice to distinguish its opens, that is if for any opens~$u, v \in \O(X)$,
-if~$x \inplus u \Leftrightarrow x \inplus v$ for all points~$x$ of~$X$, then~$u =
-v$.\end{definition}
+suffice to distinguish its opens, that is if for any opens~$U, V \in \O(X)$,
+if~$x \inplus U \Leftrightarrow x \inplus V$ for all points~$x$ of~$X$, then~$U =
+V$.\end{definition}
 
@@
 of surjections~$\NN \to \RR$ is a drastic example of a locale which fails to be
-spatial, and the localic real line is spatial in classical mathematics and can
+spatial. The localic real line is spatial in classical mathematics and can
 fail to be spatial in constructive mathematics.
@@
 
+\subsection{Geometric theories}
+
+\begin{definition}A \emph{geometric theory} consists of
+\begin{enumerate}
+  \item a set of sorts: $X$, $Y$, $Z$, \ldots
+  \item a set of function symbols: $f : X \times Y \to Z$, \ldots
+  \item a set of relation symbols: $R \hookrightarrow X \times Y \times Z$, \ldots
+  \item a set of geometric sequents as axioms: $\varphi \vdash_{x:X, y:Y} \psi$, \ldots
+\end{enumerate}
+A \emph{geometric sequent} (in some context~$x_1\?X_1,\ldots,x_n\?X_n$) is a
+formula built using only the ingredients ${=}\ {\top}\ {\wedge}\ {\bot}\
+{\vee}\ {\bigvee}\ {\exists}$ and the relation symbols (but no ${\Rightarrow}\
+{\forall}$). The symbol~``$\bigvee$'' refers to disjunctions of arbitrary
+set-indexed families of formulas.
+\end{definition}
+
+Even though superficially similar, geometric theories play a
+substantially different role than formal systems such as Peano arithmetic or
+Zermelo--Fraenkel set theory. There are a number of notable differences:
+\begin{enumerate}
+\item Formal systems are typically of foundational interest and can be
+fruitfully employed as metatheories. Geometric theories are more interesting
+for applications, for instance for presenting spaces. This aspect is discussed
+in Section~\ref{sect:presenting-frames}.
+\item Formal systems typically come with an intended model; geometric theories
+do not. For instance, the geometric theory of groups has all groups as models.
+\item Formal systems are typically required to be recursively
+axiomatizable. Many interesting geometric theories are not, and indeed contain
+an uncountable number of axioms. Geometric theories also allow for arbitrary
+set-indexed disjunctions; for this to make sense, the study of geometric
+theories can only be carried out in a sufficiently rich background theory.
+\item For formal systems, typically a version of Gödel's completeness theorem
+holds: A formula is derivable iff it holds in all models. In contrast,
+geometric theories can be consistent yet do not admit set-based models. An
+example is the geometric theory of surjections~$\NN \twoheadrightarrow \RR$
+reviewed below.
+\item Perhaps the most important difference to formal systems is that geometric
+theories often depend on given mathematical objects. For instance, for each
+ring~$A$, there is the geometric theory of prime ideals of~$A$.
+\end{enumerate}
+
+\begin{definition}A \emph{set-based model}~$M$ (or ``Tarski model'') of a
+geometric theory~$\TT$ consists of
+\begin{enumerate}
+  \item a set~$\brak{X}$ for each sort~$X$,
+  \item a function~$\brak{f} : \brak{X_1} \times \cdots \times \brak{X_n} \to
+  \brak{Y}$
+  for each function symbol~$f : X_1 \times \cdots \times X_n \to Y$ and
+  \item a relation~$\brak{R} \subseteq \brak{X_1} \times \cdots \times \brak{X_n}$
+  for each relation symbol~$R \hookrightarrow X_1 \times \cdots \times X_n$
+\end{enumerate}
+such that~$M$ validates the axioms of~$\TT$.
+\end{definition}
+
+\paragraph{The geometric theory of rings.} This theory has one sort, $R$; five function symbols:
+$0$ and $1$ (nullary), $-$ (unary), $+$ and $\cdot$ (binary); no relation
+symbols; the usual axioms, such as $\top \vdash_{x:R,y:R} x + y = y + x$.
+
+A set-based model of the theory of rings is an ordinary ring.
+
+In a similar vein, there is the geometric theory of monoids, of groups, and so
+on.
+
+\paragraph{The geometric theory of objects.} This theory has one sort, $X$, and no function
+symbols, relation symbols or axioms. A set-based model of this theory is just a
+set.
+
+A variant of this theory is the theory of inhabited objects, which has the
+axiom~$\top \vdash \exists x\?X\_ \top$. Its set-based models are precisely the
+inhabited sets.
+
+\paragraph{The geometric theory of surjections~$f : \NN \twoheadrightarrow
+\RR$.} This theory has no
+sorts; no function symbols; a nullary relation symbol~$\varphi_{nx}$ for each
+pair~$\langle n,x \rangle \in \NN \times \RR$ (to be read as~``$f$ maps~$n$
+to~$x$''); and the following axioms:
+\begin{enumerate}
+\item For each~$n \in \NN$, the axiom~$\top \vdash \bigvee_{x \in \RR}
+\varphi_{nx}$, expressing that~$f$ is total.
+\item For each~$n \in \NN$ and each~$x,y \in \RR$, the axiom~$\varphi_{nx}
+\wedge \varphi_{ny} \vdash \bigvee\{ \top \,|\, x = y \}$,
+expressing that~$f$ is single-valued.\footnote{The disjunction is taken over
+the set~$\{ \top \,|\, x = y \}$. This is a certain subsingleton set of
+formulas; it is inhabited (by the formula~$\top$) iff~$x = y$. Hence, if~$x =
+y$, this axiom reads~$\varphi_{nx}
+\wedge \varphi_{ny} \vdash \top$ and could also be omitted; if~$x \neq y$,
+this axiom reads~$\varphi_{nx} \wedge \varphi_{ny} \vdash \bot$.}
+\item For each~$x \in \RR$, the axiom~$\top \vdash \bigvee_{n \in \NN}
+\varphi_{nx}$, expressing that~$f$ is surjective.
+\end{enumerate}
+
+A model of this geometric theory consists of an~$(\NN \times \RR)$-indexed
+family of truth values, that is a subset~$G$ of~$\NN \times \RR$, satisfying
+precisely those axioms which render~$G$ the graph of a surjection~$\NN
+\twoheadrightarrow \RR$.
+
+\paragraph{The geometric theory of Dedekind cuts.} This theory has no sorts; no
+function symbols; nullary relation symbols~$\alpha_x$ and~$\beta_x$ for each
+rational number~$x$ (to be read as~``$x$ is contained in the lower respectively
+the upper cut''); and the following axioms, spelling out that the cut is bounded,
+rounded, open and located:
+\begin{align*}
+  \top &\vdash \textstyle\bigvee_{x \in \QQ} \alpha_x \\
+  \top &\vdash \textstyle\bigvee_{x \in \QQ} \beta_x \\
+  \alpha_y &\vdash \alpha_x & \text{(for each~$x,y \in \QQ$ such that~$x < y$)} \\
+  \beta_x &\vdash \beta_y & \text{(for each~$x,y \in \QQ$ such that~$x < y$)} \\
+  \alpha_x &\vdash \textstyle\bigvee_{y > x} \alpha_y & \text{(for each~$x \in \QQ$)} \\
+  \beta_y &\vdash \textstyle\bigvee_{x < y} \beta_x & \text{(for each~$y \in \QQ$)} \\
+  \alpha_x \wedge \beta_y &\vdash \bot & \text{(for each~$x,y \in \QQ$ such that~$x \geq y$)} \\
+  \top &\vdash \alpha_x \vee \beta_y & \text{(for each~$x,y \in \QQ$ such that~$x < y$)}
+\end{align*}
+
+A model of this geometric theory consists of two families of truth values, both
+indexed by the rational numbers, hence two subsets~$L,U \subseteq \QQ$, in such
+a way that~$\langle L,U \rangle$ is a Dedekind cut.
+
+\paragraph{The geometric theory of prime ideals of a given ring~$A$.} In classical
+commutative algebra, the notion of a \emph{prime ideal} of a ring~$A$ is
+fundamental. A prime ideal is a subset~$\ppp \subseteq A$ such that~$0 \in
+\ppp$, $(x \in \ppp) \wedge (y \in \ppp) \Rightarrow (x + y \in \ppp)$, $(x \in
+\ppp) \Rightarrow rx \in \ppp$ (for any~$r \in A$),~$1 \not\in \ppp$ and~$(xy
+\in \ppp) \Rightarrow (x \in \ppp) \vee (y \in \ppp)$.
+
+Correspondingly, the geometric theory of prime ideals has no sorts; no function
+symbols; one relation symbol~$\alpha_x$ for each element~$x \in A$; and the
+following axioms:
+\begin{align*}
+  \top &\vdash \alpha_0 \\
+  \alpha_x \wedge \alpha_y &\vdash \alpha_{x+y} & \text{(for each~$x,y \in A$)} \\
+  \alpha_x &\vdash \alpha_{rx} & \text{(for each~$r,x \in A$)} \\
+  \alpha_1 &\vdash \bot \\
+  \alpha_{xy} &\vdash \alpha_x \vee \alpha_y & \text{(for each~$x,y \in A$)}
+\end{align*}
+
+For many purposes, the theory of prime \emph{filters} is actually more
+relevant. Classically, a prime filter of a ring is simply the complement of a
+prime ideal; constructively, it is prudent to axiomatize this notion directly.
+Hence the theory of prime filters of~$A$ has no sorts, no function symbols, one
+relation symbol~$\alpha_x$ for each element~$x \in A$, and the following
+axioms:
+\begin{align*}
+  \alpha_0 &\vdash \bot \\
+  \alpha_{x+y} &\vdash \alpha_x \vee \alpha_y & \text{(for each~$x,y \in A$)} \\
+  \alpha_{rx} &\vdash \alpha_x & \text{(for each~$r,x \in A$)} \\
+  \top &\vdash \alpha_1 \\
+  \alpha_x \wedge \alpha_y &\vdash \alpha_{xy} & \text{(for each~$x,y \in A$)}
+\end{align*}
+
+\paragraph{The inconsistent geometric theory.} This theory has no sorts,
+function symbols or relations, but the single axiom $\top \vdash \bot$.
+This theory does not admit any set-based models.
+
+\paragraph{The empty geometric theory.} This theory does not have any sorts,
+function symbols, relations or axioms. There is exactly one set-based model of
+this theory, the empty structure.
+
+\subsection{Presenting frames by theories}
+\label{sect:presenting-frames}
+
+\begin{definition}A geometric theory~$\TT$ is \emph{propositional} if and only
+if its set of sorts is empty.\end{definition}
+
+As a consequence, a propositional geometric theory consists just of a set of
+nullary relation symbols and a set of axioms. Associated to any such
+theory~$\TT$ is its \emph{Lindenbaum algebra}: This is the partially ordered
+set of the formulas over the signature of~$\TT$ modulo~$\TT$-provable
+equivalence, equipped with the ordering~$[\varphi] \preceq [\psi]$ iff~$\TT$
+proves~$\varphi \vdash \psi$.
+
+The Lindenbaum algebra is a frame, with finite meets given by~$[\varphi] \wedge
+[\psi] = [\varphi \wedge \psi]$ and set-indexed joins given by~$\bigvee_i
+[\varphi_i] = [\bigvee_i \varphi_i]$;\footnote{This short description of the
+set-indexed joins picks representatives from each equivalence class and is
+hence only sensible in the presence of the axiom of choice. In a
+constructive metatheory, we should rather write~$\bigvee_{i \in I} M_i =
+[\bigvee_{\varphi \in K} \varphi]$, where~$K = \{ \varphi \,|\, \exists i \in
+I\_ \varphi \in M_i \}$.} we can regard the Lindenbaum algebra as the free
+frame generated by the nullary relation symbols of~$\TT$ modulo the axioms
+of~$\TT$. Our interest in the Lindenbaum algebra is because it gives rise to a
+locale:
+
+\begin{definition}The \emph{classifying locale}~$L(\TT)$ of a propositional
+geometric theory is the locale which has the Lindenbaum algebra of~$\TT$ as its
+underlying frame.\end{definition}
+
+It is an instructive exercise to verify that the points of~$L(\TT)$ are in
+canonical one-to-one correspondence with the set-based models of~$\TT$. This
+observation also explains why we cannot hope to construct classifying
+locales of arbitrary, nonpropositional, geometric theories: Any locale only has
+a set of points, but arbitrary geometric theories can have a proper class of
+set-based models.
+
+Many locales are fruitfully described as the classifying locale of a certain
+propositional geometric theory. For instance:
+\begin{enumerate}
+\item The \emph{localic real line} is the classifying locale of the theory of
+Dedekind cuts. Its points are the models of that theory, that is, the Dedekind
+reals, and the induced topology on the set of points coincides with the usual
+Euclidean topology on the reals.
+
+The localic open unit interval is the classifying locale of the
+same theory, but with the added axiom~$\top \vdash \alpha_0 \wedge \beta_1$.
+\item The \emph{locale of surjections~$\NN \twoheadrightarrow \RR$} is the
+classifying locale of the theory of surjections~$\NN \twoheadrightarrow \RR$.
+\item The \emph{spectrum} of a ring~$A$ is the classifying locale of the
+theory of prime filters of~$A$. The topological space of points of this locale
+is the Zariski spectrum as familiar from algebraic
+geometry.\footnote{Classicaly, the spectrum is usually defined as the
+topological space of prime ideals instead of prime filters. However,
+classically there is a canonical one-to-one correspondence between prime ideals
+and prime filters, hence one could just as well use filters instead of ideals
+in the definition. The classifying locale of prime ideals of~$A$ also exists,
+and has the prime ideals of~$A$ as its point, but the induced topology on its
+topological space of points is the \emph{flat topology} or \emph{co-Zariski
+topology}~\cite{tarizadeh:flat} instead of the ordinary
+Zariski topology~\cite[Proposition~4.5]{johnstone:rings-fields-and-spectra}.}
+\item The \emph{localic Galois group} of a field extension~$L|k$ is the
+classifying locale of the theory of ring automorphisms~$L \to L$ which fix~$k$.
+\item The \emph{empty locale} is the classifying locale of the inconsistent theory.
+\item The \emph{one-point locale} is the classifying locale of the empty theory.
+%\item The \emph{discrete locale on a set~$M$} is the classifying locale of the
+%theory which has one nullary relation symbol~$\varphi_x$ for each element~$x
+%\in M$, the axiom~$\top \vdash \bigvee_{x \in M} \varphi_x$ and for each~$x,y
+%\in M$ the axiom~$\varphi_x \wedge \varphi_y \vdash \bigvee\{ \top \,|\, x = y
+%\}$.
+\end{enumerate}
+
+Just as we often refer to a topological space only by its points (``the
+topological space of prime filters of~$A$''), not mentioning its topology, it is
+customary to abbreviate ``the classifying locale of the propositional geometric
+theory of prime filters of~$A$'' as ``the locale of prime filters of~$A$''.
+
+\begin{remark}The traditional way of constructing the real line or the Zariski
+spectrum as a topological space proceeds in three steps: (1) Write down the
+axioms (of Dedekind cuts respectively prime filter). (2) Using the powerset
+axiom and separation, construct the set of Dedekind reals respectively the set
+of prime filters. (3) Devise a useful topology on the resulting set.
+
+Locales provide us with a more economic and also more conceptual way of
+arriving at these topological spaces: They can be obtained as the topological
+space of points of the corresponding classifying locale. In particular,
+step~(3) of manually devising a topology is not necessary when pursuing the
+localic route.
+\end{remark}
+
+In general, different theories can give rise to isomorphic classifying locales;
+such theories are called \emph{Morita-equivalent}. A vast research program is
+built on this observation~\cite{caramello:tst}.
+
+Conversely, any locale~$X$ is the classifying locale of a certain theory, namely
+of the \emph{theory of points of~$X$}. This theory has one nullary relation
+symbol~$\varphi_u$ for each open~$U \in \O(X)$ (read as ``the point belongs
+to~$U$'') and the following axioms:
+\begin{align*}
+  \varphi_U &\vdash \varphi_V & \text{(for all~$U,V \in \O(X)$ such that~$U \preceq V$)} \\
+  \top &\vdash \varphi_\top \\
+  \varphi_U \wedge \varphi_V &\vdash \varphi_{U \wedge V} & \text{(for all~$U,V \in \O(X)$)} \\
+  \varphi_{\bigvee_i U_i} &\vdash \bigvee_i \varphi_{U_i} & \text{(for each set-indexed family~$(U_i)_i$ of opens)}
+\end{align*}
+These are exactly the axioms for a completely prime filter of~$\O(X)$. The
+isomorphism from~$\O(X)$ to the Lindenbaum algebra of this theory maps an
+open~$U$ to~$[\varphi_U]$.
+
+\subsection{Sheaves on locales}
+
+\begin{definition}\label{defn:presheaf}
+A \emph{presheaf}~$F$ on a locale~$X$ is a functor~$\O(X)^\op
+\to \Set$, that is
+\begin{enumerate}
+  \item a set~$F(U)$ for each open~$U \in \O(X)$ and
+  \item a map~$(\cdot)|^U_V : F(U) \to F(V)$ for each opens~$V \preceq U$
+\end{enumerate}
+such that $(\cdot)|^U_U = \operatorname{id}_{F(U)}$ for all~$U \in \O(X)$ and
+$(\cdot)|^V_W \circ (\cdot)|^U_V = (\cdot)|^U_W$ for all~$W \preceq V
+\preceq U$.\end{definition}
+
+Definition~\ref{defn:presheaf}, and also any other definition in this section,
+also makes sense for topological spaces instead of locales, since only the
+notion of opens and their inclusion relation is used. The elements of~$F(U)$ are
+called \emph{sections of~$F$ over~$U$} and the maps~$(\cdot)|^U_V$
+are called \emph{restriction maps}.
+
+When first learning about presheaves, one can get boggled down[XXX:wording] by the vast
+amount of data encoded in a single presheaf (one set for each open). The
+situation should be compared to Kripke models for a first-order language~$L$. The
+preorder~$W$ of nodes of such a model is like the frame of opens of a locale, and the
+family~$(M_w)_{w \in W}$ of~$L$-structures, one for each node, is like a
+presheaf. In fact, the notion of a locale can be generalized to that of a
+\emph{site} so that both locales and Kripke frames induce sites and that
+presheaves on sites can be defined; however, this shall not be pursued here.
+[XXX: reference]
+
+The prototypical example of a presheaf (and also a sheaf) is the presheaf~$\C$ of
+continuous real-valued functions on a space~$X$. For this presheaf, the
+set~$\C(U)$ is the set of continuous real-valued maps~$U \to \RR$, and the
+maps~$(\cdot)|^U_V$ are given by actual restriction of functions to smaller
+domains, that is by
+\[ \C(U) \longrightarrow \C(V),\ s \longmapsto s|_V. \]
+
+\begin{definition}
+\begin{enumerate}
+\item A \emph{compatible family} of a presheaf~$F$ with respect to an open
+covering~$U = \bigvee_i U_i$ is a family~$(s_i)_i$ of sections~$s_i \in F(U_i)$
+such that~$s_i|^{U_i}_{U_i \wedge U_j} = s_j|^{U_j}_{U_i \wedge U_j}$ for all
+indices~$i,j$.
+\item A presheaf~$F$ is a \emph{sheaf}
+iff for any compatible family~$(s_i)_i$ with respect to an open covering~$U =
+\bigvee_i U_i$ there is a unique section~$s \in F(U)$ such
+that~$s|^U_{U_i} = s_i$ for all~$i$.
+\end{enumerate}
+\end{definition}
+
+For instance, the presheaf~$\C$ of continuous real-valued functions is a sheaf,
+while its subpresheaf~$\C_\text{c}$ of constant real-valued functions is
+usually not a sheaf. For instance, on~$X = \RR$, the two constant functions~$f
+: (-1,0) \to \RR,\,x \mapsto -1$ and~$g : (0,1) \to \RR,\,x \mapsto 1$ agree on
+the intersection of their domains (which is empty), but there is no constant
+function~$h : (-1,0) \cup (0,1) \to \RR$ which restricts to~$f$ on~$(-1,0)$ and
+to~$g$ on~$(0,1)$.
+
+\begin{definition}A \emph{morphism of presheaves}~$\eta : F \to G$ on a
+locale~$X$ is a family~$(\eta_U)_{U \in \O(X)}$ of maps~$\eta_U : F(U) \to
+G(U)$ such that for all sections~$s \in F(U)$ and all opens~$V \preceq U$
+of~$X$, $\eta_U(s)|^U_V = \eta_V(s|^U_V)$. A \emph{morphism of sheaves} is a
+morphism of the underlying presheaves.
+\end{definition}
+
+\subsection{Sheaf semantics}
+
+Let~$X$ be a locale. A \emph{formula over an open~$U \in \O(X)$} is a
+first-order formula (made up using~${=}\ {\top}\ {\wedge}\ {\bot}\ {\vee}\
+{\Rightarrow}\ {\forall}\ {\exists}$) over the signature which has one sort for
+each sheaf~$F$, one constant symbol of sort~$F$ for each section~$s \in F(U)$,
+one function symbol~$f : F \to G$ for each morphism of sheaves, and so on.
+
+Table~\ref{table:sheaf-semantics} defines by recursion on the structure of
+formulas what it means for a formula~$\varphi$ over an open~$U$ to be
+\emph{forced} or \emph{true over~$U$}, written~``$U \models \varphi$''. The
+main differences with the perhaps more familiar Kripke semantics are marked.
+
+\begin{table}[ht]
+  \tbl{The (first-order fragment of) the sheaf semantics of a locale~$X$}{\begin{tabular}{@{}l@{\ \ }c@{\ \ }l@{}}
+    $U \models s =_F t$ &\Ll& $s = t$ when evaluated as elements of~$F(U)$ \\
+    $U \models \top$ &\Ll& true \\
+    $U \models \varphi \wedge \psi$ &\Ll& $U \models \varphi$ and $U \models \psi$ \\
+    $U \models \bot$ &\Ll& \hcancel{false}{0pt}{3pt}{0pt}{-2pt}\ $U = \bot$ \\
+    $U \models \varphi \vee \psi$ &\Ll& \hcancel{$U \models \varphi$ or $U \models \psi$}{0pt}{3pt}{0pt}{-2pt}\ there exists an open covering $U = \bigvee_i U_i$ \\
+    && \quad\quad such that for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$ \\
+    $U \models \varphi \Rightarrow \psi$ &\Ll& for all~$V \preceq U$: $V \models \varphi$ implies $V \models \psi$ \\
+    $U \models \forall s \? F\_ \varphi(s)$ &\Ll& for all $V \preceq U$ and sections~$s_0 \in F(V)$: $V \models \varphi(s_0)$ \\
+    $U \models \exists s \? F\_ \varphi(s)$ &\Ll& \hcancel{there exists $s_0 \in F(U)$ such that $U \models \varphi(s_0)$}{0pt}{3pt}{0pt}{-2pt} \\
+    && there exists an open covering $U = \bigvee_i U_i$ such that for all~$i$: \\
+    && \quad\quad there exists~$s_0 \in F(U_i)$ such that $U_i \models \varphi(s_0)$
+  \end{tabular}}
+  \label{table:sheaf-semantics}
+\end{table}
+
+Some clauses in Table~\ref{table:sheaf-semantics} contain slight abuses of
+notation. Namely, if~$\varphi$ is a formula over an open~$U$ and if~$V$ is an
+open such that~$V \preceq U$, then~``$V \models \varphi$'' is officially not
+well-defined, as~$\varphi$ is not a formula over~$V$. When we still write~``$V
+\models \varphi$'', we mean that any constant symbol~$s \in F(U)$ appearing
+in~$\varphi$ should be interpreted as its restriction~$s|^U_V \in F(V)$, so
+that the resulting formula is a formula over~$V$.
+
+The principal properties of the sheaf semantics is that it is monotone, local
+and sound with respect to intuitionistic first-order logic:
+
+\begin{theorem}Let~$X$ be a locale. Let~$U$ be an open of~$X$. Let~$\varphi$ be
+a formula over~$U$.
+\begin{enumerate}
+\item If~$V$ is an open such that~$V \preceq U$, then~$U \models
+\varphi$ implies~$V \models \varphi$.
+\item Let~$U = \bigvee_i U_i$ be an open covering. If~$U_i \models
+\varphi$ for all~$i$, then~$U \models \varphi$.
+\item Let~$\psi$ be a further formula over~$U$. If~$U \models
+\varphi$ and if~$\varphi$ entails~$\psi$ intuitionistically, then also~$U
+\models \psi$.
+\end{enumerate}
+\end{theorem}
+
+\begin{proof}The proof of~(1) is by a routine induction on the structure
+of~$\varphi$. For the clauses pertaining disjunction and existential
+quantification, the distributive law has to be used.
+
+The proof of~(2) is also a routine induction proof. For the clause pertaining
+equality, the sheaf condition has to be used.
+
+To verify statement~(3), it is prudent to generalize the claim slightly: An
+induction on the structure of derivations shows that if a sequent~$\varphi
+\seq{x_1{:}X_1,\ldots,x_n{:}X_n} \psi$ is derivable in intuitionistic first-order
+logic, then~$U \models \forall x_1\?X_1\_ \ldots \forall x_n\?X_n\_ (\varphi
+\Rightarrow \psi)$.
+\end{proof}
+
+\begin{definition}A \emph{sheaf model}~$M$ of a geometric theory~$\TT$ over a
+locale~$X$ consists of
+\begin{enumerate}
+  \item a sheaf~$\brak{S}$ on~$X$ for each sort~$S$,
+  \item a morphism~$\brak{f} : \brak{S_1} \times \cdots \times \brak{S_n} \to
+  \brak{T}$ of sheaves for each function symbol~$f : S_1 \times \cdots \times S_n \to T$ and
+  \item a subsheaf~$\brak{R} \subseteq \brak{S_1} \times \cdots \times \brak{S_n}$
+  for each relation symbol~$R \hookrightarrow S_1 \times \cdots \times S_n$
+\end{enumerate}
+such that~$M$ validates the axioms of~$\TT$, that is, such that the axioms
+of~$\TT$ are forced on~$\top \in \O(X)$.
+\end{definition}
+
+XXX ``internalizing parameter-dependence''
+
 \section{Applications in constructive algebra}
+
+XXX: just a stub
+
+Let~$A$ be a reduced ring ($x^n = 0 \Rightarrow x = 0$).
+Then there is a certain sheaf~$A^\sim$ of rings on a certain locale~$X$ such
+that \ldots
+
+{\centering $A^\sim$ is close to~$A$:\par}
+\begin{enumerate}
+\item There is a canonical bijection~$A \to A^\sim(X)$.
+\item $A^\sim$ inherits any property of~$A$ which is
+localization-stable.
+\item A geometric sequent holds for~$A^\sim$ iff$^\star$ it holds for all
+stalks~$A_{\mathfrak{f}}$.
+\end{enumerate}
+
+{\centering $A^\sim$ has better properties than~$A$:\par}
+\begin{enumerate}
+\item $A^\sim$ is a field: $\forall x\?A^\sim\_ (\neg(\exists y\?A^\sim\_
+  xy = 1) \Rightarrow x = 0)$.
+\item $A^\sim$ has $\neg\neg$-stable equality:
+$\forall x,y\?A^\sim\_ \neg\neg(x = y) \Rightarrow x = y$.
+\item $A^\sim$ is \emph{anonymously Noetherian}.
+\end{enumerate}
+
+{\centering This observation can be exploited to give \\ short, conceptual and
+constructive proofs.\par}
+
+\begin{theorem}Let~$A$ be a ring. The Lindenbaum algebra of the theory of prime
+filters of~$A$ is canonically isomorphic to the frame of radical ideals in~$A$.
+\end{theorem}
+
+\begin{proof}The set of radical ideals is ordered by inclusion. The meet~$\aaa
+\wedge \bbb$ of two radical ideals is their intersection and the
+join~$\bigvee_i \aaa_i$ of radical ideals is~$\sqrt{\sum_i \aaa_i}$.
+
+The desired isomorphism maps~$[\alpha_x]$ to~$\sqrt{(x)}$ and maps finite meets
+to finite meets and arbitrary joins to arbitrary joins, as it should be. A
+computation shows that it is well-defined and monotone. For instance, in the
+Lindenbaum algebra, we have~$[\alpha_x] \wedge [\alpha_y] \preceq
+[\alpha_{xy}]$, and indeed~$\sqrt{(x)} \cap \sqrt{(y)} \subseteq \sqrt{(xy)}$.
+XXX: surjective and order-reflecting
+\end{proof}
+
+XXX injective matrices
+
+XXX generic freeness
+
+XXX spell out details of field property
 
@@
 XXX cite https://apcz.umk.pl/czasopisma/index.php/LLP/article/viewFile/LLP.2013.009/732
+
+XXX include rules for geometric logic?

### 2020-03-30 paper-herrsching2019/main.tex
to do
Found the story for the third section, now need to tell it

@@
 \usepackage{ws-rv-van}
-\usepackage{mathtools,stmaryrd}
+\usepackage{mathtools,stmaryrd,xspace}
 
+\usepackage[normalem]{ulem}
 \usepackage{tikz}
@@
 \newcommand{\Top}{\mathrm{Top}}
+\newcommand{\BPIT}{\textsc{bpit}\xspace}
 
@@
 of~$X$ such that~$\top = \bigvee_{i \in I} U_i$, there is a (Kuratowski-)finite
-subset~$I' \subseteq I$ such that~$\top = \bigvee_{i \in I'} U_i$. A
-topological space~$Y$ is compact iff its induced locale~$L(Y)$ is.
+subset~$I' \subseteq I$ such that~$\top = \bigvee_{i \in I'} U_i$. For
+instance, a topological space~$Y$ is compact iff its induced locale~$L(Y)$ is.
 
@@
 For instance, any locale induced by a topological space is spatial. The locale
-of surjections~$\NN \to \RR$ is a drastic example of a locale which fails to be
+of surjections~$\NN \twoheadrightarrow \RR$ is a drastic example of a locale which fails to be
 spatial. The localic real line is spatial in classical mathematics and can
@@
 \item Formal systems typically come with an intended model; geometric theories
-do not. For instance, the geometric theory of groups has all groups as models.
+do not. For instance, the geometric theory of groups has all groups as models;
+when we are writing down the axioms of a group, we are not setting out to
+capture the properties of any one specific group.
 \item Formal systems are typically required to be recursively
@@
 commutative algebra, the notion of a \emph{prime ideal} of a ring~$A$ is
-fundamental. A prime ideal is a subset~$\ppp \subseteq A$ such that~$0 \in
-\ppp$, $(x \in \ppp) \wedge (y \in \ppp) \Rightarrow (x + y \in \ppp)$, $(x \in
-\ppp) \Rightarrow rx \in \ppp$ (for any~$r \in A$),~$1 \not\in \ppp$ and~$(xy
-\in \ppp) \Rightarrow (x \in \ppp) \vee (y \in \ppp)$.
-
-Correspondingly, the geometric theory of prime ideals has no sorts; no function
+fundamental. Corresponding to the definition (recalled in
+Section~\ref{sect:algebraic-preliminaries}), the geometric theory of prime
+ideals has no sorts; no function
 symbols; one relation symbol~$\alpha_x$ for each element~$x \in A$; and the
@@
 Hence the theory of prime filters of~$A$ has no sorts, no function symbols, one
-relation symbol~$\alpha_x$ for each element~$x \in A$, and the following
+relation symbol~$\alpha_x$ for each element~$x \in A$ and the following
 axioms:
@@
 in the definition. The classifying locale of prime ideals of~$A$ also exists,
-and has the prime ideals of~$A$ as its point, but the induced topology on its
+and has the prime ideals of~$A$ as its points, but the induced topology on its
 topological space of points is the \emph{flat topology} or \emph{co-Zariski
@@
 of the \emph{theory of points of~$X$}. This theory has one nullary relation
-symbol~$\varphi_u$ for each open~$U \in \O(X)$ (read as ``the point belongs
+symbol~$\varphi_U$ for each open~$U \in \O(X)$ (read as ``the point belongs
 to~$U$'') and the following axioms:
@@
   \item a set~$F(U)$ for each open~$U \in \O(X)$ and
-  \item a map~$(\cdot)|^U_V : F(U) \to F(V)$ for each opens~$V \preceq U$
+  \item a map~$(\cdot)|^U_V : F(U) \to F(V)$ for each pair of opens~$V \preceq U$
 \end{enumerate}
@@
 
-Definition~\ref{defn:presheaf}, and also any other definition in this section,
+Definition~\ref{defn:presheaf}, as well as any other definition in this section,
 also makes sense for topological spaces instead of locales, since only the
@@
 
-XXX: just a stub
+Commutative algebra abounds with techniques to reduce given situations to
+simpler ones, for instance passing to a quotient or passing to a
+localization. These techniques facilitate short and elegant proofs, such as the
+following. (Section~\ref{sect:algebraic-preliminaries} contains some algebraic
+preliminaries.)
 
-Let~$A$ be a reduced ring ($x^n = 0 \Rightarrow x = 0$).
-Then there is a certain sheaf~$A^\sim$ of rings on a certain locale~$X$ such
-that \ldots
+\begin{quote}
+\textbf{Theorem.} Let~$M$ be an injective matrix with more columns than rows
+over a reduced ring~$A$. Then~$1 = 0$ in~$A$.
 
-{\centering $A^\sim$ is close to~$A$:\par}
+\textbf{Proof.} \uwave{Assume not.} Then there is a \uwave{minimal
+prime ideal} $\ppp \subseteq A$. The matrix~$M$ is also injective when
+considered as a matrix over the stalk~$A_\ppp$; since~$A_\ppp$ is a
+\uwave{field}, this is a contradiction to basic linear algebra. \qed
+\end{quote}
+
+However, these reduction techniques typically require \emph{transfinite
+methods}. The displayed proof contains three appeals to the transfinite:
+\begin{enumerate}
+\item The proof proceeds by contradiction, hence requires the law of excluded
+middle, the strongest of the omniscience principles.
+\item The proof then requires a minimal prime ideal. Such an ideal can be
+obtained in two steps: First, the Boolean Prime Ideal
+Theorem~\BPIT is invoked to obtain some prime ideal. Then
+Zorn's lemma fabricates a minimal one.\footnote{Standard textbooks prove the statement ``any nontrivial
+ring has a prime ideal'' by using Zorn's lemma, which
+in the presence of the law of excluded middle is equivalent to the full axiom
+of choice; however, the statement is actually equivalent to the weaker Boolean
+Prime Ideal Theorem [XXX: reference].
+
+The combined statement ``any ring has a minimal prime ideal'' is equivalent to
+the full axiom of choice. This observation might be folklore; we learned the
+following argument from Will Savin~[XXX: reference]: Let~$(M_i)_{i \in I}$ be a
+disjoint family of inhabited sets. We consider the free ring~$\ZZ[\bigcup_i M_i]$ and
+its ideal~$I$ generated by~$\{ ab \,|\, i \in I, a,b \in M_i, a \neq b \}$. Assume
+that the factor ring~$\ZZ[\bigcup_i M_i]/I$ contains a minimal prime ideal. Its
+preimage~$\ppp \subseteq \ZZ[\bigcup_i M_i]$ is then a prime ideal which is
+minimal among all prime ideals containing~$I$. By primality, for any index~$i$,
+there is at most one element of~$M_i$ not in~$\ppp$. By minimality, for each
+index~$i$, there is exactly one element of~$M_i$ not in~$\ppp$: If there were
+none, then pick an element~$a \in M_i$ XXX}
+% https://mathoverflow.net/questions/98731/minimal-prime-ideals-and-axiom-of-choicerevised-version
+\item Finally, the proof exploits that the stalks of reduced rings at minimal
+prime ideals are fields. This requires two further invokations
+of~\BPIT.\footnote{By~\BPIT, the intersection of all prime ideals of~$A_\ppp$
+is its nilradical. Since~$A_\ppp$ is reduced, the nilradical is the zero ideal.
+Since the prime ideals of~$A_\ppp$ are in bijection with those prime ideals
+of~$A$ which are contained in~$\ppp$, the ring~$A_\ppp$ has exactly one prime
+ideal. By~\BPIT, any noninvertible element of~$A_\ppp$ is contained in some
+prime ideal, hence in the zero ideal. Thus~$A_\ppp$ is a field.}
+\end{enumerate}
+
+This state of affairs is not satisfactory. A statement as simple as the
+displayed example should admit an explicit calculational proof, presenting a
+concrete method for transforming the given conditional equations expressing
+injectivity into the equation~$1 = 0$ without any appeals to the transfinite.
+
+The completeness theorem for coherent logic~[XXX: reference (Elephant)] even
+gives an a priori reason why (for a given matrix size) a finitary proof has to
+exist. However, since the proof of the completeness theorem itself uses~\BPIT
+(and indeed is equivalent to it), it does not give any indication how such a
+proof could be found. (Still, beautiful constructive proofs are known and
+presented, for instance, in a celebrated short note by Richman on nontrivial
+uses of trivial rings~\cite{richman:trivial-rings} and in the recent textbook by
+Lombardi and Quitté on constructive commutative
+algebra~\cite{lombardi-quitte:constructive-commutative-algebra}.)
+
+The key issue with the usual reduction techniques in commutative algebra is
+their dependence on \emph{ideal objects} such as prime or maximal ideals. In
+general, those objects can only be obtained by transfinite methods. However, in
+practice many arguments do not actually require completed ideal objects:
+Their computational core applies just as well to finite approximations of
+these ideal objects.
+
+It is then a challenging task in mathematical logic to devise efficient means to
+extract this hidden constructive content in a mostly mechanical manner. We
+envision to reinterpret any given classical proof employing ideal objects in a
+constructive fashion, realizing ideal objects as convenient fictions.
+
+Locales and sheaf models contribute to this program by providing new reduction
+techniques for commutative algebra. These techniques have a similar effect as the
+classical reduction techniques, but are powered by purely constructive
+underpinnings. In a nutshell, the idea is as follows.
+
+\begin{enumerate}
+\item Instead of replacing a given ring with another (for instance a ring~$A$ with
+one of its stalks~$A_\ppp$ at a minimal prime ideal), we replace a given ring
+with a \emph{sheaf of rings}.
+\item We then maintain the illusion that we are working with a plain old ring
+by employing the sheaf semantics.
+\end{enumerate}
+
+The goal of this section is to give a leisurely introduction to this procedure.
+We recall in Section~\ref{sect:algebraic-preliminaries} algebraic
+preliminaries. Section~\ref{sect:a-remarkable-sheaf} explores the sheaf to use.
+There is an interesting origin story to this sheaf, which we tell in
+Section~\ref{sect:origins}. Finally, we illustrate how to use the technique in
+a series of concrete examples in Section~\ref{sect:example-applications}.
+
+\subsection{Algebraic preliminaries}
+\label{sect:algebraic-preliminaries}
+
+By \emph{ring}, we mean commutative ring with unit. We do not require the unit
+to be distinct from the zero; indeed, trivial rings have intriguing
+applications~\cite{richman:trivial-rings}.
+
+A ring is \emph{reduced} iff zero is its only nilpotent; that is, if~$x^n = 0$
+implies~$x = 0$.
+
+An \emph{ideal} is a subset~$\aaa \subseteq A$ such that~$0 \in \aaa$, $(x \in
+\ppp) \wedge (y \in \ppp) \Rightarrow (x + y \in \ppp)$ and $(x \in
+\ppp) \Rightarrow (rx \in \ppp)$ (for any~$r \in A$).
+An ideal~$\ppp$ is \emph{prime} iff~$1 \not\in \ppp$ and~$(xy
+\in \ppp) \Rightarrow (x \in \ppp) \vee (y \in \ppp)$.
+
+A subset~$\fff \subseteq A$ is a \emph{prime filter} iff
+$0 \not\in \fff$, $(x + y \in \fff) \Rightarrow (x \in \fff) \vee (y \in
+\fff)$, $1 \in \fff$ and $(xy \in \fff) \Leftrightarrow (x \in \fff) \wedge (y
+\in \fff)$.
+
+An ideal~$\aaa \subseteq A$ is a \emph{radical ideal} iff~$x^n \in \aaa$
+implies~$x \in \aaa$ for any~$x \in A$ and natural number~$n$. The
+\emph{radical}~$\sqrt{\aaa}$ of an arbitrary ideal~$\aaa$ is the radical
+ideal~$\{ x \in A \,|\, \exists n \in \NN\_ x^n \in \aaa \}$.
+
+A ring is \emph{local} iff a finite sum is invertible only if one of its
+summands also is. More precisely, this means that~$1 \neq 0$ and that
+whenever~$a + b$ is invertible, $a$ is invertible or~$b$ is invertible.
+This definition is an elementary rephrasing of the textbook definition,
+according to which a ring is local iff it has exactly one maximal ideal. The
+elementary definition works better in constructive contexts.
+
+If~$S$ is a multiplicatively closed subset of a ring~$A$, the
+\emph{localization}~$A[S^{-1}]$ is the ring of formal fractions~$\frac{a}{s}$ with~$a
+\in A$ and~$s \in S$, where two such fractions~$\frac{a}{s}$ and~$\frac{b}{t}$ are deemed
+equivalent iff there is an element~$u \in S$ such that~$uta = usb$. We do not
+require that~$0 \not\in S$; instead we plainly observe that ~$A[S^{-1}]$ is the
+zero ring iff~$0 \in S$.
+
+An important special case is \emph{localizing away from an element}: If~$f$ is
+an element of~$A$, then~$A[f^{-1}]$ is~$A[\{1,f,f^2,f^3,\ldots\}^{-1}]$. This
+ring is the zero ring iff~$f$ is nilpotent.
+
+The \emph{stalk at a prime ideal}~$\ppp \subseteq A$ is the
+localization~$A_\ppp \defeq A[(A \setminus \ppp)^{-1}]$. Assuming de Morgan's
+laws, such stalks are local rings. The \emph{stalk at a filter}~$\fff \subseteq A$ is
+defined as~$A_\fff \defeq A[\fff^{-1}]$ and is a local ring even
+constructively. Conversely, if a localization~$A[S^{-1}]$ is local then~$S$ can
+be saturated to a filter.
+
+A property of rings is \emph{localization-stable} iff it is inherited by any
+localization. For instance, if~$A$ is a reduced ring, then so is any
+localization~$A[S^{-1}]$.
+
+\subsection{A remarkable sheaf}
+\label{sect:a-remarkable-sheaf}
+
+Let~$A$ be a reduced ring. Then there is a certain locale~$X$ which contains a
+certain ``mirror image'' of~$A$, a sheaf~$A^\sim$ of rings. On the one hand,
+this mirror image is ``close to~$A$'', such that studying~$A^\sim$ pays off for
+learning about~$A$. On the other hand, it has much better properties than~$A$
+has.
+
+\begin{center}
+\vbox{$A^\sim$ is close to~$A$: \\[0.6em]
+\framebox{\parbox{0.9\textwidth}{\vspace*{-0.7em}
 \begin{enumerate}
@@
 stalks~$A_{\mathfrak{f}}$.
-\end{enumerate}
+\end{enumerate}\vspace*{-0.7em}}}}
 
-{\centering $A^\sim$ has better properties than~$A$:\par}
+\vbox{$A^\sim$ has better properties than~$A$: \\[0.6em]
+\framebox{\parbox{0.9\textwidth}{\vspace*{-0.7em}
 \begin{enumerate}
-\item $A^\sim$ is a field: $\forall x\?A^\sim\_ (\neg(\exists y\?A^\sim\_
-  xy = 1) \Rightarrow x = 0)$.
+\item $A^\sim$ is a local ring:
+% $\neg(1=0)$, $\forall x,y\?A^\sim\_ x + y = 1 \Rightarrow (\exists z\?A^\sim\_ xz = 1) \vee (\exists z\?A^\sim\_ yz = 1)$.
+\item $A^\sim$ is a field: $\forall x\?A^\sim\_ (\neg(\exists y\?A^\sim\_ xy = 1) \Rightarrow x = 0)$.
 \item $A^\sim$ has $\neg\neg$-stable equality:
@@
 \item $A^\sim$ is \emph{anonymously Noetherian}.
-\end{enumerate}
+\end{enumerate}\vspace*{-0.7em}}}}
+\end{center}
 
-{\centering This observation can be exploited to give \\ short, conceptual and
-constructive proofs.\par}
+%{\centering This observation can be exploited to give \\ short, conceptual and
+%constructive proofs.\par}
+
+XXX from here on just a stub
+
+\begin{proposition}The sheaf of rings~$A^\sim$ has $\neg\neg$-stable equality:
+\[ \Spec(A) \models \forall x,y\?A^\sim\_ \neg\neg(x = y) \Rightarrow x = y. \]
+\end{proposition}
+
+\begin{proof}We argue internally. Let~$x,y \? A^\sim$ be given such
+that~$\neg\neg(x = y)$. We will verify that~$x - y$ is not invertible; by the
+field condition, this will imply~$x - y = 0$, hence~$x = y$.
+
+So assume that~$x - y$ is invertible. Then~$\neg(x - y = 0)$, since if~$x - y =
+0$, then~$1 = (x-y) (x-y)^{-1} = 0$ in contradiction to~$1 \neq 0$. This is a
+contradiction to the assumption~$\neg\neg(x-y=0)$.\end{proof}
+
+\begin{proposition}The sheaf of rings~$A^\sim$ is anonymously
+Noetherian.\end{proposition}
+
+\begin{proof}We argue internally. Let~$\aaa \subseteq A^\sim$ be an ideal.
+Assume that~$\aaa$ is not finitely generated. We will verify that then~$\aaa =
+(0)$, hence that~$\aaa$ is finitely generated; contradiction.
+
+Let~$x \in \aaa$. If~$x$ is invertible, then~$\aaa = (1)$ is finitely
+generated. Hence~$x$ is not invertible. Thus~$x = 0$ because~$A^\sim$ is a
+field.
+\end{proof}
 

### 2020-03-31 paper-herrsching2019/main.tex
More on A^\sim

@@
 \newcommand{\ppp}{\mathfrak{p}}
+\newcommand{\qqq}{\mathfrak{q}}
+\newcommand{\mmm}{\mathfrak{m}}
 \newcommand{\fff}{\mathfrak{f}}
@@
 suffice to distinguish its opens, that is if for any opens~$U, V \in \O(X)$,
-if~$x \inplus U \Leftrightarrow x \inplus V$ for all points~$x$ of~$X$, then~$U =
-V$.\end{definition}
+if~$x \inplus U \Rightarrow x \inplus V$ for all points~$x$ of~$X$, then~$U
+\preceq V$.\end{definition}
+
+An equivalent conditon is: if for any opens~$U, V \in \O(X)$, if~$x \inplus U
+\Leftrightarrow x \inplus V$ for all points~$x$ of~$X$, then~$U = V$.
 
@@
 spatial. The localic real line is spatial in classical mathematics and can
-fail to be spatial in constructive mathematics.
+fail to be spatial in constructive mathematics (Theorem~\ref{thm:spatiality-reals}).
 
@@
 spectrum as a topological space proceeds in three steps: (1) Write down the
-axioms (of Dedekind cuts respectively prime filter). (2) Using the powerset
+axioms (of Dedekind cuts respectively prime filters). (2) Using the powerset
 axiom and separation, construct the set of Dedekind reals respectively the set
@@
 open~$U$ to~$[\varphi_U]$.
+
+The question whether a given locale is spatial is intimately related with the
+existence of appropriate \emph{ideal objects}. Their existence typically hinges
+on the axiom of choice or some of its variants. For instance, there are the
+following results (where we allow ourselves a modicum in classical logic in
+order to state the results in a familiar form).
+\begin{enumerate}
+\item The localic spectrum of a ring~$A$ is spatial iff every nontrivial ideal
+(of a certain class of rings related with~$A$) is contained in a prime ideal.
+\item The localic Galois group of a Galois extension~$L|k$ is spatial iff
+any~$k$-homomorphism~$\sigma : E \to L$ defined on a finite intermediate
+extension~$L|E|k$ can be extended to a~$k$-homomorphism~$L \to L$.
+\item The localic real line is spatial iff the every open covering of the
+set~$[0,1]$ of Dedekind reals has a Lebesgue number (that is, is Heine--Borel
+compact).
+\item The statement that any classifying locale of a propositional coherent
+theory (that is, a propositional geometric theory in which only finite
+disjunctions occur) is spatial is equivalent to the Boolean Prime Ideal
+Theorem~\BPIT.
+\end{enumerate}
+These results explain why in classical mathematics, where all of these
+conditions are satisfied, we can blithely and fruitfully use the topological
+spectrum of a ring, the topological Galois group or the topological space of
+reals. In classical mathematics, employing the localic counterparts is merely
+for convenience or for aesthetic reasons; in constructive mathematics, the
+localic replacements (or similar replacements in other pointfree approaches to
+topology) are vital.
+
+%\begin{theorem}The localic real line is spatial if and only if the set~$[0,1]$ of
+%Dedekind real numbers is compact in the sense of Heine--Borel.
+%\end{theorem}
+%
+%\begin{proof}XXX\end{proof}
+
+%\begin{theorem}Let~$A$ be a ring and~$\Spec(A)$ its spectrum as defined above as the locale of
+%prime filters of~$A$. Then~$\Spec(A)$ is spatial if (the law of excluded middle and) the
+%Boolean Prime Ideal Theorem~\BPIT hold. More specifically, $\Spec(A)$ is
+%spatial if and only if for any ideal~$\aaa \subseteq A$ and any element~$x \in A$,
+%\[ x \in \sqrt{\aaa} \quad\textnormal{if}\quad
+%\textnormal{for all filters~$\fff \subseteq A$ with~$x \in \fff$, $\fff \between \aaa$}. \]
+%\end{theorem}
+%
+%The symbol~``$\between$'' denotes that the two sets have an element in common.
+%
+%\begin{proof}Assume that~$\Spec(A)$ is spatial. Let an ideal~$\aaa \subseteq A$
+%and an element~$x \in A$ be given such that for all filters~$\fff \subseteq A$
+%with~$x \in \fff$, $\fff \between \aaa$. Then for all filters~$\fff$, $\fff
+%\inplus \alpha_x \Rightarrow \fff \inplus \bigvee_{y\in \aaa} \alpha_y$.
+%Hence ...
+%Using the filter axioms, any open of~$\Spec(A)$ can be written
+%as~$[\bigvee_i \alpha_{x_i}]$.
+%\end{proof}
 
@@
 
+A \emph{minimal prime ideal} is a prime ideal~$\ppp$ such that for any prime
+ideal~$\qqq$ with~$\qqq \subseteq \ppp$, $\qqq = \ppp$.
+
 A subset~$\fff \subseteq A$ is a \emph{prime filter} iff
@@
 
-Let~$A$ be a reduced ring. Then there is a certain locale~$X$ which contains a
+Let~$A$ be a ring. Then there is a certain locale~$X$ which contains a
 certain ``mirror image'' of~$A$, a sheaf~$A^\sim$ of rings. On the one hand,
@@
 \framebox{\parbox{0.9\textwidth}{\vspace*{-0.7em}
-\begin{enumerate}
-\item There is a canonical bijection~$A \to A^\sim(X)$.
-\item $A^\sim$ inherits any property of~$A$ which is
+\begin{Alphlist}[(C)]
+\item\label{item:iso}
+There is a canonical isomorphism~$A \to A^\sim(X)$ of rings.
+\item\label{item:loc-stable}
+$A^\sim$ inherits any property of~$A$ which is
 localization-stable.
-\item A geometric sequent holds for~$A^\sim$ iff$^\star$ it holds for all
+\item\label{item:stalks}
+A geometric sequent holds for~$A^\sim$ iff$^\star$ it holds for all
 stalks~$A_{\mathfrak{f}}$.
-\end{enumerate}\vspace*{-0.7em}}}}
+\end{Alphlist}\vspace*{-0.7em}}}}
+\hfill $^\star$assuming \BPIT for the ``if'' direction{\quad}
+\bigskip
 
@@
 \framebox{\parbox{0.9\textwidth}{\vspace*{-0.7em}
-\begin{enumerate}
-\item $A^\sim$ is a local ring:
-% $\neg(1=0)$, $\forall x,y\?A^\sim\_ x + y = 1 \Rightarrow (\exists z\?A^\sim\_ xz = 1) \vee (\exists z\?A^\sim\_ yz = 1)$.
-\item $A^\sim$ is a field: $\forall x\?A^\sim\_ (\neg(\exists y\?A^\sim\_ xy = 1) \Rightarrow x = 0)$.
-\item $A^\sim$ has $\neg\neg$-stable equality:
+\begin{Alphlist}[(D)]
+\addtocounter{enumi}{3}
+\item\label{item:local} $A^\sim$ is a local ring.
+\end{Alphlist}
+
+Assuming that~$A$ is reduced: \\[-1.8em]
+\begin{Alphlist}[(G)]
+\addtocounter{enumi}{4}
+\item\label{item:field} $A^\sim$ is a field: $\forall x\?A^\sim\_ (\neg(\exists y\?A^\sim\_ xy = 1) \Rightarrow x = 0)$.
+\item\label{item:negneg-stable} $A^\sim$ has $\neg\neg$-stable equality:
 $\forall x,y\?A^\sim\_ \neg\neg(x = y) \Rightarrow x = y$.
-\item $A^\sim$ is \emph{anonymously Noetherian}.
-\end{enumerate}\vspace*{-0.7em}}}}
+\item\label{item:noetherian} $A^\sim$ is \emph{anonymously Noetherian}.
+\end{Alphlist}\vspace*{-0.7em}}}}
 \end{center}
 
-%{\centering This observation can be exploited to give \\ short, conceptual and
-%constructive proofs.\par}
-
-XXX from here on just a stub
+Recall that when we ascribe some ring-theoretic property to~$A^\sim$, we
+are implicitly using the sheaf semantics. For instance, when we say
+that~$A^\sim$ is a local ring, we actually mean
+\begin{multline*}
+  \Sh(X) \models \neg(1 =_{A^\sim} 0) \quad\text{and} \\
+  \Sh(X) \models \forall x,y\?A^\sim\_ \bigl(x + y =_{A^\sim} 1 \Longrightarrow \\
+  (\exists
+  z\?A^\sim\_ xz =_{A^\sim} 1) \vee (\exists z\?A^\sim\_ yz =_{A^\sim} 1)\bigr).
+\end{multline*}
+
+The locale~$X$ on which this theater plays out is none other than the
+\emph{spectrum of~$A$} from algebraic geometry, and the sheaf~$A^\sim$ is the
+sheaf of regular functions. We will give details on its construction in
+Section~\ref{sect:origins}, but first want to reflect on the displayed
+statements.
+
+\paragraph{A reification of all stalks.}
+Statement~(\ref{item:stalks}) explains that, to a first approximation, the
+sheaf~$A^\sim$ is a reification of all the stalks of~$A$ into a single coherent
+entity: The sheaf~$A^\sim$ has exactly those properties which are shared by all
+the stalks~$A_\fff$. For instance, the sheaf~$A^\sim$ is an integral domain iff
+all the stalks are.
+
+But crucially, this metatheorem only pertains to properties which can be
+formulated as geometric sequents. Being an integral domain is such a property
+($1 = 0 \vdash \bot$, $xy = 0 \seq{x:R,y:R} x = 0 \vee y = 0$), but many others
+are not. The reductive power of harnessing~$A^\sim$ stems from the fact
+that~$A^\sim$ enjoys properties which are not shared by~$A$, its stalks or
+indeed any localizations or quotients.
+
+\paragraph{Back and forth.}
+Along with statement~(\ref{item:stalks}), statement~(\ref{item:iso}) is the main
+reason why results about~$A^\sim$ imply results about~$A$. For instance, if we
+want to show that a certain ring element~$x \in A$ is zero, we can just as well
+show that its mirror image in~$A^\sim$ is zero.
+
+For this step to be useful, we need to know what~$A^\sim$ looks like.
+Statements~(\ref{item:stalks}) and~(\ref{item:loc-stable}) both allow us to
+transfer our knowledge about~$A$ to~$A^\sim$; while
+statements~(\ref{item:stalks}) requires a syntactic restriction on the property
+under consideration, statement~(\ref{item:loc-stable}) requires a semantic one.
+
+\paragraph{The field property.}
+The fact that~$A^\sim$ is a field is where~$A^\sim$ derives its main usefulness
+from. While statements~(\ref{item:loc-stable}), (\ref{item:stalks}) and
+(\ref{item:local}) are immediate consequences of the way~$A^\sim$ is
+constructed (Section~\ref{sect:origins}), the field property~(\ref{item:field})
+requires an actual computation for its verification and was a surprising
+discovery by Mulvey in the 1970s.
+
+Tierney commented around that time that the field property ``is surely important,
+though its precise significance is still somewhat obscure---as is the case with
+many such nongeometric formulas''~\cite[p.~209]{tierney:spectrum}. Even though
+the property has been known for a long time, only recently was it put in a
+wider context and its its usefulness for constructive algebra
+appreciated~\cite[Sections~3.3 and~11.4]{blechschmidt:phd}.
+
+The field property is a unique feature of~$A^\sim$; it cannot be mimicked by
+classical techniques in commutative algebra. Several constructions come close,
+but fail in other ways:
+\begin{enumerate}
+\item The stalks~$A_\mmm$ at maximal ideals: These are not fields.
+\item The stalks~$A_\ppp$ at minimal prime ideals: These are fields, but a ring
+element~$x \in A$ which is zero in all these stalks~$A_\ppp$ is not necessarily
+zero in~$A$. Also, an~$A$-module~$M$ which is finitely generated and whose
+stalks~$M_\ppp$ at minimal prime ideals are free is ``free on a dense open'',
+but this is false without the finiteness hypothesis. The analogous statement
+for~$M^\sim$ and~$A^\sim$ holds in full generality (XXX reviewed below).
+\item The quotients~$A/\mmm$: These are fields, but an element~$x \in A$ which is
+zero in all these quotients~$A/\mmm$ is not necessarily zero in~$A$, and
+passing to~$A/\mmm$ is not exact (does not preserve injectivity of linear
+maps).
+\item The quotients~$A/\ppp$ modulo the prime ideals: These are merely integral
+domains.
+\end{enumerate}
+%\item If a ring element~$x \in A$ is zero in all stalk~$A_\mmm$ at maximal
+%ideals, then~$x$ is zero in~$A$. However, the rings~$A_\mmm$ are not fields.
+%\item The rings~$A/\mmm$ are fields, but a ring element~$x \in A$ which is zero
+%in all these quotients~$A/\mmm$ is not necessarily zero in~$A$.
+%\item If a ring element~$x \in A$ is zero in all quotients~$A/\ppp$,
+%where~$\ppp$ ranges over the prime ideals, then~$x$ is zero in~$A$. However,
+%the rings~$A/\ppp$ are merely integral domains.
+%\item The stalks~$A_\ppp$, where~$\ppp$ ranges over the minimal prime ideals,
+%are fields. However, a ring element~$x \in A$ which is zero
+%in all these stalks~$A_\ppp$ is not necessarily zero in~$A$.
+
+The field condition displayed in statement~(\ref{item:field}) is not the only
+such condition used in constructive algebra. For instance, a stronger condition
+is ``any element is zero or invertible'' and a further, incomparable, condition
+is ``any nonzero element is invertible''.
+
+XXX when those other conditions satisfied
+
+\paragraph{Double negation stability.}
+XXX explain that it is surprising and explain usefulness, then give proof
 
@@
 contradiction to the assumption~$\neg\neg(x-y=0)$.\end{proof}
+
+\paragraph{The Noetherian property.}
+XXX not shared by stalks; ``true constructivists would be offended''
 
@@
 \end{proof}
+
+\subsection{The origins of the sheaf}
+\label{sect:origins}
+XXX better title
+
+XXX universal localization...
+
+XXX proofs of (B), (C), (D)
 
@@
 
+\subsection{Applications}
+\label{sect:example-applications}
+XXX better title
+
 XXX injective matrices
@@
 XXX include rules for geometric logic?
+
+X \models phi

### 2020-03-31 paper-herrsching2019/main.tex
Origin story and main workhorse

@@
 \usepackage[normalem]{ulem}
+\usepackage[all]{xy}
 \usepackage{tikz}
@@
 \newcommand{\GL}{\mathrm{GL}}
+\newcommand{\ab}{\mathrm{ab}}
 \newcommand{\Zar}{\mathrm{Zar}}
@@
 \newcommand{\Gal}{\mathrm{Gal}}
+\newcommand{\Rad}{\mathrm{Rad}}
 \newcommand{\lra}{\longrightarrow}
@@
 
-%\tableofcontents
+\tableofcontents
 
@@
 The question whether a given locale is spatial is intimately related with the
-existence of appropriate \emph{ideal objects}. Their existence typically hinges
-on the axiom of choice or some of its variants. For instance, there are the
-following results (where we allow ourselves a modicum in classical logic in
+existence of appropriate \emph{ideal objects} whose existence typically hinges
+on the axiom of choice or one of its variants. For instance, there are the
+following results (where we allow ourselves a modicum of classical logic in
 order to state the results in a familiar form).
@@
 any~$k$-homomorphism~$\sigma : E \to L$ defined on a finite intermediate
-extension~$L|E|k$ can be extended to a~$k$-homomorphism~$L \to L$.
-\item The localic real line is spatial iff the every open covering of the
-set~$[0,1]$ of Dedekind reals has a Lebesgue number (that is, is Heine--Borel
-compact).
-\item The statement that any classifying locale of a propositional coherent
+extension~$L|E|k$ admits an extension to a~$k$-homomorphism~$L \to L$.
+\item The localic real line is spatial iff every open covering of the
+metric space~$[0,1]$ of Dedekind reals has a Lebesgue number (that is, if the
+metric space~$[0,1]$ is Heine--Borel compact).
+\item The statement that the classifying locale of any propositional coherent
 theory (that is, a propositional geometric theory in which only finite
@@
 \begin{definition}\label{defn:presheaf}
-A \emph{presheaf}~$F$ on a locale~$X$ is a functor~$\O(X)^\op
-\to \Set$, that is
+A \emph{presheaf}~$F$ on a locale~$X$ is a functor~$\O(X)^\op \to \Set$, that is
 \begin{enumerate}
@@
 \ppp) \wedge (y \in \ppp) \Rightarrow (x + y \in \ppp)$ and $(x \in
-\ppp) \Rightarrow (rx \in \ppp)$ (for any~$r \in A$).
+\ppp) \Rightarrow (rx \in \ppp)$ (for any~$r \in A$). A family~$(x_i)_i$ of
+ring elements generates an ideal, also denoted~``$(x_i)_i$'', consisting of
+(finite) sums of multiples of the~$x_i$.
+
 An ideal~$\ppp$ is \emph{prime} iff~$1 \not\in \ppp$ and~$(xy
 \in \ppp) \Rightarrow (x \in \ppp) \vee (y \in \ppp)$.
-
 A \emph{minimal prime ideal} is a prime ideal~$\ppp$ such that for any prime
@@
 localization~$A_\ppp \defeq A[(A \setminus \ppp)^{-1}]$. Assuming de Morgan's
-laws, such stalks are local rings. The \emph{stalk at a filter}~$\fff \subseteq A$ is
+laws, such stalks are local rings. The \emph{stalk at a prime filter}~$\fff \subseteq A$ is
 defined as~$A_\fff \defeq A[\fff^{-1}]$ and is a local ring even
 constructively. Conversely, if a localization~$A[S^{-1}]$ is local then~$S$ can
-be saturated to a filter.
+be saturated to a prime filter.
 
@@
 
-Let~$A$ be a ring. Then there is a certain locale~$X$ which contains a
-certain ``mirror image'' of~$A$, a sheaf~$A^\sim$ of rings. On the one hand,
+Let~$A$ be a ring. Then there is a certain locale~$X$ on which a
+certain ``mirror image'' of~$A$ exists, a sheaf~$A^\sim$ of rings on~$X$. On the one hand,
 this mirror image is ``close to~$A$'', such that studying~$A^\sim$ pays off for
@@
 $A^\sim$ inherits any property of~$A$ which is
-localization-stable.
+localization-stable.$^\star$
 \item\label{item:stalks}
-A geometric sequent holds for~$A^\sim$ iff$^\star$ it holds for all
-stalks~$A_{\mathfrak{f}}$.
+A geometric sequent holds for~$A^\sim$ iff$^{\star\!\star}$ it holds for all
+stalks~$A_\fff$.
 \end{Alphlist}\vspace*{-0.7em}}}}
-\hfill $^\star$assuming \BPIT for the ``if'' direction{\quad}
+\hfill $^\star$precise formulation in
+Proposition~{\ref{prop:immediate-consequences}}{\quad}\phantom{\,\,}\\
+\hfill $^{\star\!\star}$assuming \BPIT for the ``if'' direction{\quad}
 \bigskip
@@
   \Sh(X) \models \neg(1 =_{A^\sim} 0) \quad\text{and} \\
-  \Sh(X) \models \forall x,y\?A^\sim\_ \bigl(x + y =_{A^\sim} 1 \Longrightarrow \\
+  \Sh(X) \models \forall x,y\?A^\sim\_ \bigl((\exists z\?A^\sim\_ (x + y)z
+  =_{A^\sim} 1) \Longrightarrow \\
   (\exists
@@
 (\ref{item:local}) are immediate consequences of the way~$A^\sim$ is
-constructed (Section~\ref{sect:origins}), the field property~(\ref{item:field})
+constructed (Proposition~\ref{prop:immediate-consequences}), the field property~(\ref{item:field})
 requires an actual computation for its verification and was a surprising
@@
 the property has been known for a long time, only recently was it put in a
-wider context and its its usefulness for constructive algebra
+wider context (recognizing it as a small shadow of algebraic geometry's
+\emph{quasicoherence}) and its its usefulness for constructive algebra
 appreciated~\cite[Sections~3.3 and~11.4]{blechschmidt:phd}.
@@
 
-\subsection{The origins of the sheaf}
+\subsection{An algebraic origin story}
 \label{sect:origins}
-XXX better title
-
-XXX universal localization...
-
-XXX proofs of (B), (C), (D)
-
-\begin{theorem}Let~$A$ be a ring. The Lindenbaum algebra of the theory of prime
-filters of~$A$ is canonically isomorphic to the frame of radical ideals in~$A$.
+
+The purpose of this section is to properly motivate the construction
+leading to~$A^\sim$. Briefly, the story is as follows, with details given
+below.
+\begin{enumerate}
+\item Given a ring~$A$, we set out to construct the \emph{free local ring}
+over~$A$.
+\item In the strict sense of the word, this endeavor will fail, but in a wider
+sense the quest will succeed and yield the sheaf~$A^\sim$.
+\item It is then a surprising observation that~$A^\sim$ is a field, even though
+we only set out to construct a local ring.
+\item There is general machinery for free constructions of this kind, and this
+machinery is \emph{not} applicable to construct free fields.
+\end{enumerate}
+
+\paragraph{Free constructions.}
+Let~$G$ be a group. Then there is a universal way to turn~$G$ into an abelian
+group~$G^\ab$ equipped with a group homomorphism~$G \to G^\ab$, the \emph{free
+abelian group over~$G$}. This abelianization has the universal property that
+for any homomorphism~$G \to M$ into an abelian group, there is exactly one
+group homomorphism~$G^\ab \to M$ rendering the diagram
+\[ \xymatrix{
+  G \ar[rd] \ar[rrr] &&& {\substack{\text{abelian}\\\text{\normalsize$\!M$}\\\phantom{\text{abelian}}}} \\
+  & {\substack{\text{\normalsize$G^\ab$}\\\text{abelian}}} \ar@{-->}_[@!29]{}[rru]
+} \]
+commutative. In textbooks, the abelianization is usually constructed as~$G^\ab
+= G/[N,N]$, that is as the quotient by the subgroup generated by the
+commutators. This explicit construction makes it appear that the abelianization
+of a group is specific result in algebra.
+
+We can also cast the contruction in logical language, which will make it
+evident that abelianization is just a special case of a more general procedure.
+The free abelian group over~$G$ can be obtained as the \emph{term model} of the
+theory of abelian groups extended by one constant~$x_g$ for each element~$g \in
+G$ and the axioms
+\begin{align*}
+  \top &\vdash x_{g \circ h} = x_g \circ x_h & \text{(for each~$g,h \in G$)} \\
+  \top &\vdash x_g^{-1} = x_{g^{-1}} & \text{(for each~$g \in G$)} \\
+  \top &\vdash x_e = e
+\end{align*}
+That is, $G^\ab$ contains exactly those elements and exactly those
+identifications between elements such that~$G^\ab$ is an abelian group and such
+that~$G$ can be interpreted in it (that is, that there is a homomorphism~$G \to
+G^\ab$).
+
+In exactly the same manner, we can construct the free group over a monoid, the
+free~$\ZZ/(2)$-algebra over a ring, the free ring over a set or the free
+Heyting algebra over a partially ordered set. This is all well-understood; a
+particularly perspicious account in the general setting of partial Horn
+theories is due to Palmgren and Vickers~\cite{palmgren-vickers:partial-horn}.
+
+However, this procedure cannot be carried out for any pair of theories. For
+instance, there are no \emph{free fields}. The free field on a set~$X$ would be
+a field~$K$ together with a map~$X \to K$ such that any map~$X \to L$ factors
+over the given map~$X \to K$ by a unique field homomorphism~$K \to L$.
+% Exercise: proof this!
+
+For instance, in the case~$X = \{ t \}$, the field~$\QQ(t)$ of rational
+functions comes close, but fails because there are no field homomorphisms~$\QQ(t)
+\to L$ which could map~$t$ to an algebraic element. We could also try to
+concoct a term model, but the set of terms in the language of a field extended
+by a constant~$t$ modulo provable equivalence is only the ring~$\ZZ[t]$, not a
+field.
+
+The non-existence of free fields is explained by the observation that the
+theory of fields is, for any of its variants,\footnote{By ``theory of fields'',
+we could mean the theory of rings extended by either the coherent axiom~$\top
+\seq{x:R} x = 0 \vee (\exists y\?R\_ xy = 1)$ or by the axiom~$\neg(\exists
+y\?R\_ xy = 1) \seq{x:R} x = 0$ or by the axiom~$\neg(x = 0) \seq{x:R} \exists
+(y\?R\_ xy = 1)$.} not a partial Horn theory.
+XXX: Are general conditions known for existence of free models?
+
+\paragraph{Free local rings.} The theory of local rings, that is the theory of
+rings extended by the two coherent axioms
+\begin{align*}
+  1 = 0 &\vdash \bot \\
+  (\exists z\?R\_ (x+y)z = 1) &\seq{x:R,y:R} (\exists z\?R\_ xz = 1) \vee
+  (\exists z\?R\_ yz = 1),
+\end{align*}
+is also not a partial Horn theory. Hence there is no reason to expect that free
+local rings exist, and indeed, in general they do not. However, despite this
+negative outlook, we want to analyze the situation in more detail.
+
+The correct notion of a morphism between local rings is that of a \emph{local
+ring homomorphism}, a ring homomorphism~$f : R \to S$ which \emph{reflects
+invertibility}, that is for which~$f(x)$ being invertible implies that~$x$ is
+invertible.
+
+Hence a \emph{free local ring} over a ring~$A$ is to be a local ring~$A'$
+together with a ring homomorphism~$A \to A'$ such that the following universal
+property holds: Any ring homomorphism~$A \to B$ into a local ring uniquely
+factors via a local homomorphism over~$A \to A'$ as indicated in the diagram.
+\[ \xymatrix{
+  A \ar[rd] \ar[rrr] &&& {\substack{\text{local}\\\text{\normalsize$\!\!\!B$}\\\phantom{\text{local}}}} \\
+  & {\substack{\text{\normalsize$\!\!A'$}\\\text{local}}} \ar@{-->}_[@!31]{\text{local}}[rru]
+} \]
+
+To develop intuition for what this universal property requests, let a
+homomorphism~$f : A \to B$ into a local ring be given. The subset~$B^\times
+\subseteq B$ is then a \emph{prime filter}, due to the locality of~$B$. Its
+preimage~$\fff \defeq f^{-1}[B^\times]$ is a prime filter of~$A$. Hence the
+localization~$A_\fff = A[\fff^{-1}]$ is a local ring and~$f$ factors via the
+well-defined map~$f' : A_\fff \to B,\,\frac{x}{s} \mapsto f(s)^{-1}f(x)$ over
+the canonical morphism~$A \to A_\fff,\,x \mapsto \frac{x}{1}$. Furthermore,
+this map is local.
+
+Hence the stalks~$A_\fff$, where~$\fff$ ranges over the prime filters of~$A$,
+can be regarded as approximations to the hypothetical true local ring over~$A$.
+Each of these stalks is local, but might not validate the universal property
+for all homomorphisms~$A \to B$ into local rings. We are thus led to the
+following conclusions.
+
+\begin{enumerate}
+\item A free local ring~$A \to A'$ over~$A$ exists if~$A$ has exactly one prime
+filter. (The converse also holds.) In classical mathematics, this condition is
+satisfied if and only if~$A$ is local ring of Krull dimension zero. In this
+case, the unique prime filter is~$A^\times$ and the free local ring over~$A$ is given
+by~$A$ itself. Hence in general, the problem of constructing the free local
+ring over a given ring is not solvable.
+\item However, we could solve the problem in full generality if we would have a
+special prime filter~$\fff_0$ which could somehow shift shape, that is, turn into any
+specific prime filter~$\fff$ on demand.
+\end{enumerate}
+
+The existence of such a shape-shifting prime filter~$\fff_0$ is not a pipe dream.
+While it does not exist as a set, it can be incarnated as a \emph{sheaf}.
+
+\begin{definition}\label{defn:generic-prime-filter}
+The \emph{generic prime filter}~$\fff_0$ of a ring~$A$ is the
+generic model of the theory of prime filters of~$A$.\end{definition}
+
+According to this definition, the generic prime filter is a particular
+\emph{sheaf model} over the classifying locale of prime filters of~$A$, the
+\emph{spectrum} of~$A$. By localizing at the generic prime filter, we solve
+the problem of constructing the free local ring in a generalized sense: \emph{The
+free local ring over~$A$ exists as a certain sheaf of rings.} It has a
+universal property not only with respect to all ordinary rings, but also all
+sheaves of rings on arbitrary locales.
+
+\paragraph{Constructing the sheaf~$A^\sim$.} We recall that the given ring~$A$
+has a counterpart in the universe of sheaves over~$\Spec(A)$, namely the
+constant sheaf~$\underline{A}$. We construct the sheaf~$A^\sim$ as
+\[ A^\sim \defeq \underline{A}[\fff_0^{-1}], \]
+that is as the localization of~$\underline{A}$ at its prime
+filter~$\fff_0$.\footnote{This description contains a slight abuse of notation.
+According to Definition~\ref{defn:generic-prime-filter}, the generic prime
+filter $\fff_0$ is a model of the theory of prime filters, hence a
+family~$(U_x)_{x \in A}$ of opens of~$\Spec(A)$, one for each nullary relation
+symbol of the theory. In fact, by the construction of the generic model, the
+open~$U_x$ is simply the element~$[D(x)]$ of the Lindenbaum algebra. We
+turn this family into the subsheaf of~$\underline{A}$ given by $U \mapsto \{ g
+\in \Hom(U,A) \,|\, \forall x \in A\_ f^{-1}[\{a\}] \preceq [D(a)] \}$. Abusing
+notation, we denote this subsheaf also as~``$\fff_0$''. The
+statement~``$\fff_0$ is a prime filter'' is then validated by the sheaf semantics.}
+
+With this definition, a number of statements from
+Section~\ref{sect:a-remarkable-sheaf} are immediate.
+
+\begin{proposition}\label{prop:immediate-consequences}
+Let~$A$ be a ring. Then the statements~(\ref{item:loc-stable}), (\ref{item:stalks}) and
+(\ref{item:local}) displayed above hold:
+\begin{Alphlist}[(D)]
+\item[(\ref{item:loc-stable})]
+The ring~$A^\sim$ has every property~$\varphi$ that~$A$ has, assuming
+that~$\varphi$ is a property of rings which can be put in the formal
+higher-order language of toposes; that there is an intuitionistic proof
+that~$\varphi$ is localization-stable; and that every element of~$A$ is
+nilpotent or not.
+\item[(\ref{item:stalks})]
+If a geometric sequent holds for~$A^\sim$, then it also holds for all
+stalks~$A_\fff$ at prime filters. The converse holds if~$\Spec(A)$ is spatial,
+or instance if~\BPIT is available.
+\item[(\ref{item:local})] The ring~$A^\sim$ is local.
+\end{Alphlist}
+\end{proposition}
+
+\begin{proof}
+\begin{Alphlist}[(D)]
+\item[(\ref{item:local})] The ring~$A^\sim$ is local as it is the
+localization at a prime filter.
+\item[(\ref{item:loc-stable})] Any property of~$A$ passes to~$\underline{A}$
+(XXX: ref) and the ring~$A^\sim$ is a localization of~$\underline{A}$.
+\item[(\ref{item:stalks})] We recall that the points of~$\Spec(A)$ are
+precisely the prime filters of~$A$. The stalk of the constant
+sheaf~$\underline{A}$ at any point is precisely~$A$, and by construction, the
+stalk of the generic prime filter~$\fff_0$ (considered as a subsheaf
+of~$\underline{A}$) at a prime filter~$\fff$ is~$\fff$. Since localization is a
+geometric construction, it is preserved by the operation of computing
+(sheaf-theoretic) stalks. Hence the stalk of~$A^\sim$ at a point~$\fff$ is
+$A[\fff^{-1}]$, that is, the ring-theoretic stalk of~$A$ at~$\fff$. The claim
+therefore follows from Proposition~XXX.\qedhere
+\end{Alphlist}
+\end{proof}
+
+Furthermore, the sheaf~$A^\sim$ is the free local ring over~$A$ in the
+following sense. If~$R$ is a sheaf of rings on a locale~$X$ and~$S$ is a sheaf
+of rings on a locale~$Y$, then by a morphism~$R \to S$ we mean a pair
+consisting of a morphism~$f : Y \to X$ of locales together with a
+morphism~$f^\sharp : f^{-1}R \to S$ of sheaves of rings on~$Y$.
+In this way a category of sheaves of rings on arbitrary locales is set up, into
+which the category of ordinary rings embeds.
+
+For~$A^\sim$ to be the local ring over~$A$, there has to be a morphism~$A \to
+A^\sim$. For this we take the pair~$\langle g, g^\sharp \rangle$ where~$g :
+\Spec(A) \to \pt$ is the unique morphism into the one-point locale
+and~$g^\sharp : g^{-1}A = \underline{A} \to A^\sim$ is the localization
+morphism.
+
+If~$\langle f,f^\sharp \rangle : A \to B$ is a morphism into a local sheaf of
+rings on an arbitrary locale~$X$, we can construct, in the universe of sheaves
+over~$X$, the factorization~$\underline{A} \to
+\underline{A}[\fff^{-1}] \to B$ where~$\fff \defeq
+(f^\sharp)^{-1}[B^\times]$. Since~$\fff$ is a model of the theory of prime
+filters of~$A$, there is a unique morphism~$g : X \to \Spec(A)$ of locales such
+that~$g^{-1}\fff_0 = \fff$. Since~$g^{-1}A^\sim = \underline{A}[\fff^{-1}]$,
+the localization just constructed gives rise to a morphism~$A^\sim \to B$ of
+local sheaves of rings.
+
+\subsection{Understanding the sheaf model}
+
+In Section~\ref{sect:origins}, we constructed the sheaf~$A^\sim$ as the free
+local ring over~$A$, more specifically as the localization of the constant
+sheaf~$\underline{A}$ on~$\Spec(A)$ at the generic prime filter~$\fff_0$
+of~$A$. Several of the properties mentioned in
+Section~\ref{sect:a-remarkable-sheaf} are immediate consequences of this
+abstract description.
+
+However, for a more detailed identification of~$A^\sim$ -- especially for
+verifying its field property -- we require a more concrete description. In
+particular, we require a concrete description of the underlying frame
+of~$\Spec(A)$. Since this frame is defined as the Lindenbaum algebra of the
+theory of prime filters of~$A$, we need a thorough grasp of the consequences of
+this theory.
+
+Such an understanding is imparted by the following theorem. This theorem is the
+central workhorse of our approach.
+
+\begin{theorem}\label{thm:workhorse}
+Let~$A$ be a ring. The Lindenbaum algebra of the theory~$\TT$ of prime filters
+of~$A$ is canonically isomorphic to the frame of radical ideals in~$A$, via an
+isomorphism which maps
+\[ [D(f)] \longmapsto \sqrt{(f)}. \]
+In particular:
+\[
+  \textnormal{$\TT$ proves~$D(f) \vdash \bigvee_{i \in I} D(g_i)$} \quad\text{iff}\quad
+    f \in \sqrt{(g_i)_i}. \]
 \end{theorem}
 
-\begin{proof}The set of radical ideals is ordered by inclusion. The meet~$\aaa
-\wedge \bbb$ of two radical ideals is their intersection and the
-join~$\bigvee_i \aaa_i$ of radical ideals is~$\sqrt{\sum_i \aaa_i}$.
-
-The desired isomorphism maps~$[\alpha_x]$ to~$\sqrt{(x)}$ and maps finite meets
-to finite meets and arbitrary joins to arbitrary joins, as it should be. A
-computation shows that it is well-defined and monotone. For instance, in the
-Lindenbaum algebra, we have~$[\alpha_x] \wedge [\alpha_y] \preceq
-[\alpha_{xy}]$, and indeed~$\sqrt{(x)} \cap \sqrt{(y)} \subseteq \sqrt{(xy)}$.
-XXX: surjective and order-reflecting
+\begin{proof}The set of radical ideals~$\Rad(A)$ is ordered by inclusion. Its
+least element is~$\sqrt{(0)}$, the meet~$\aaa \wedge \bbb$ of two radical
+ideals is their intersection and the join~$\bigvee_i \aaa_i$ of radical ideals
+is~$\sqrt{\sum_i \aaa_i}$. In this way the radical ideals form a frame.
+
+The Lindenbaum algebra is the free frame generated by the symbols~$D(x)$,
+where~$x$ ranges over the elements of~$A$, modulo the axioms for a prime
+filter. Hence to give a frame homomorphism~$\O(\Spec(A)) \to \Rad(A)$, we need
+to specify images in~$\Rad(A)$ for each generator in such a way that the
+axioms for a prime filter are satisfied. This is achieved by declaring~$D(f)
+\mapsto \sqrt{(f)}$:
+\begin{align*}
+  D(0) &\vdash \bot & \sqrt{(0)} &\subseteq \sqrt{(0)} \\
+  D(f+g) &\vdash D(f) \vee D(g) & \sqrt{(f+g)} &\subseteq\textstyle \sqrt{\sqrt{(f)} + \sqrt{(g)}} \\
+  \top &\vdash D(1) & \sqrt{(1)} &\subseteq \sqrt{(1)} \\
+  D(f) \wedge D(g) \dashv\!&\vdash D(fg) & \sqrt{(f)} \cap \sqrt{(g)} &= \sqrt{(fg)}
+\end{align*}
+The most interesting of these radical inclusions is probably~$\sqrt{(f)} \cap
+\sqrt{(g)} \subseteq \sqrt{(fg)}$, as this illustrates why passing to the
+radical is vital; in general, we do not have~$(f) \cap (g) \subseteq (fg)$.
+
+The frame homomorphism obtained in this manner is surjective, since a preimage
+for a radical ideal~$\aaa$ is given by~$\bigvee_{f \in \aaa} [D(f)]$. It
+remains to show that the homomorphism reflects the ordering.
+
+To this end, let an element~$f \in A$ and a family~$(g_i)_i$ of elements be
+given and assume~$\sqrt{(f)} \subseteq \sqrt{\sum_i \sqrt{(g_i)}}$. Then there
+are values~$n,m \in \NN$, $i_1,\ldots,i_m \in \NN$ and~$u_1,\ldots,u_m \in
+A$ such that~$f^n = u_1 g_{i_1} + \cdots + u_m g_{i_m}$. We then have the
+following chain of entailments.
+\[
+  D(f) \vdash
+  D(f^n) \vdash
+  \bigvee_{k=1}^m D(u_k g_{i_k}) \vdash
+  \bigvee_{k=1}^m D(g_{i_k}) \vdash
+  \bigvee_i D(g_i). \qedhere
+\]
 \end{proof}
+
+Theorem~\ref{thm:workhorse} shows that any derivation in the theory of prime
+filters of~$A$ can be put into a \emph{normal form}. An expression of the
+form~``$f^n = u_1 g_{i_1} + \cdots + u_m g_{i_m}$'' can be regarded as an
+\emph{algebraic certificate} of the entailment~$D(f) \vdash \bigvee_i D(g_i)$.
 
@@
 X \models phi
+
+  &\text{iff}\quad \bigvee_{n \in \NN} \exists m \in \NN\_ \exists i_1,\ldots,i_n \in I\_ \exists
+  u_1,\ldots,u_n \in A\_ f^m = \sum_{j=1}^n u_j g_{i_j}

### 2020-04-01 paper-herrsching2019/main.tex
Second workhorse and algebraic recasting

@@
 
-The notion of \emph{space} is fundamental to large parts of mathematics. It
+The notion of \emph{space} is fundamental to large parts of mathematics. This notion
 exists in various flavors, ranging from the basic cartesian spaces~$\RR^n$ to
@@
 
-\begin{definition}A \emph{metric space} consists of a set~$X$ of points
+\begin{quote}
+\textbf{Definition.} A \emph{metric space} consists of a set~$X$ of points
 together with a metric~$d : X \times X \to \RR_{\geq 0}$ satisfying the metric
-axioms. \end{definition}
-
-\begin{definition}A \emph{topological space} consists of a set~$X$ of points
+axioms.
+
+\textbf{Definition.} A \emph{topological space} consists of a set~$X$ of points
 together with a set~$\O(X) \subseteq \P(X)$ of point sets which are deemed
@@
 of open sets are open.
-\end{definition}
+\end{quote}
 
@@
 The formal definition of a locale will be given below as
-Definition~\ref{defn:locale}, but before we will review examples and comment on
+Definition~\ref{defn:locale}, but we first want to review examples and comment on
 the relevance of locales to constructive mathematics.
@@
 locale can be nontrivial even if it does not contain any points. This
-phenomenon is one instance of the general guiding principle that relinquishing
+phenomenon is one instance of a general guiding principle, namely that relinquishing
 points increases flexibility.
@@
 surjections~$\NN \twoheadrightarrow \RR$, hence~$X$ does not have any points.
-But this locale does have uncountably many basic opens~$U_{n,x}$, where~$n$
-ranges over the naturals and~$x$ ranges over the reals. We picture~$U_{n,x}$ as
-the open of those surjections~$f$ for which~$f(n) = x$, we can compute with
-these opens and consider functions on~$X$. For instance:
-\begin{enumerate}
-\item If~$x \neq y$, then the intersection of~$U_{n,x}$ with~$U_{n,y}$
+But this locale does have uncountably many basic opens~$U_{nx}$, where~$n$
+ranges over the naturals and~$x$ ranges over the reals. We picture~$U_{nx}$ as
+the ``open of those surjections~$f$ for which~$f(n) = x$'', we can compute with
+these opens, consider functions on~$X$ and ponder sublocales of~$X$. For instance:
+\begin{enumerate}
+\item If~$x \neq y$, then the intersection of~$U_{nx}$ with~$U_{ny}$
 is truly empty.
-\item The union of the~$U_{n,x}$, where~$x$ is a fixed real
+\item The union of the~$U_{nx}$, where~$x$ is a fixed real
 number and~$n$ ranges over the naturals, is all of~$X$. No finite number
@@
 (the localic version of) the naturals which, on the level of points, would map
-a surjection~$f : \NN \to \RR$ to the smallest preimage of~$x$.
+a surjection~$f : \NN \to \RR$ to the smallest preimage of~$x$ under~$f$.
 \item There is a well-defined continuous function from~$X$ to (the localic
@@
 \RR$ for which any real has infinitely many preimages, has a fractal nature: It
-is covered by the~$U_{0,x}$, where~$x$ ranges over the reals; the pairwise
+is covered by the (intersections with~$X'$ of the)~$U_{0x}$, where~$x$ ranges over the reals; the pairwise
 intersection of these opens is (truly) empty; and they are each isomorphic
-to~$X'$ -- on the level of points, by mapping a surjection~$f$ in~$U_{0,x}$ to
+to~$X'$ -- on the level of points, by mapping a surjection~$f$ in~$U_{0x}$ to
 the surjection~$(n \mapsto f(n+1))$.
@@
 
-There is nothing special about the real numbers which makes this example work;
+There is nothing special about the real numbers in this example;
 in fact, the example works just as well with any set~$M$ in place of the reals.
@@
 these subsets, and the Banach--Tarski paradox is not in
-contradiction with the basic properties of the (Lebesgue) measure in~$\RR^3$
+contradiction with the basic properties of the Lebesgue measure in~$\RR^3$
 because these intermediate subsets are not measurable.
@@
 
-More generally, Tychonoff's theorem that the product of any number of compact
+Similarly, Tychonoff's theorem that the product of any number of compact
 topological spaces is again compact is equivalent to the axiom of choice. Its
@@
 construct certain extensions of given field homomorphisms.\footnote{For
-instace, the statement~$E \subseteq L^{\Gal(L|E)}$ is trivial. For the converse
+instance, the statement~$E \subseteq L^{\Gal(L|E)}$ is trivial. For the converse
 inclusion, let~$x \in L^{\Gal(L|E)}$. Assume for the sake of contradiction
@@
 and~$\sigma(x) \neq x$. This is a contradiction to~$x \in L^{\Gal(L|E)}$.}
-As a consequence, the fundamental theorem of Galois theory is not provable in
-constructive mathematics.
+As a consequence, the fundamental theorem of Galois theory as stated is not
+provable in constructive mathematics.
 
@@
 showcases a related general phenomenon, namely that classical mathematics
-allows to push back topological concerns for longer than in constructive
-mathematics, whereas in constructive mathematics we have to embrace topology
+allows to push back topological concerns for a bit longer,
+whereas in constructive mathematics we have to embrace topology
 (in a sufficiently pointfree form such as locales) from the beginning.
@@
 extensions, finite intermediate extensions correspond to finite subgroups.
+[XXX: reference]
 Classically, intermediate extensions of finite extensions and subgroups of
-finite groups are automatically finite, but constructively this can fail.
+finite groups are automatically finite, but constructively this can
+fail.\footnote{For instance, the field extension~$\QQ(\sqrt{2})|\QQ$ is finite.
+However, if the intermediate extension~$E \defeq \{ x \in \QQ(\sqrt{2}) \,|\, x
+\in \QQ \vee \varphi \}$ is finite, then~$\varphi \vee \neg\varphi$.}
 \end{remark}
@@
 
-The starting point to arrive at the definition of a locale is the following
-fundamental observation: The partial order~$\O(X)$ of open subsets of a topological
-space~$X$ has
+The starting point of the theory of locales is the following
+observation. The set~$\O(X)$ of open subsets of a topological
+space~$X$ forms a partially ordered set which has
 \[ \begin{array}{@{}ccc@{}}
@@
 
-This observation motivates the following definition. [XXX: which abstracts...]
-\begin{definition}A \emph{frame} is a partial order with (arbitrary,
+The key idea of locales is to elevate this observation to a definition, in the
+process dropping the requirement for the elements of the ordered set to be sets
+of points:
+\begin{definition}A \emph{frame} is a partially ordered set with (arbitrary,
 set-indexed) joins and finite meets such that the distributive law
@@
 is~$\top = \{\star\}$, and potentially not all elements of~$\Omega$ are equal
-to of these two.
+to one of these two.
 \item The locale of surjections~$\NN \twoheadrightarrow \RR$ and the localic
@@
 of~$X$ such that~$\top = \bigvee_{i \in I} U_i$, there is a (Kuratowski-)finite
-subset~$I' \subseteq I$ such that~$\top = \bigvee_{i \in I'} U_i$. For
-instance, a topological space~$Y$ is compact iff its induced locale~$L(Y)$ is.
+subset~$I' \subseteq I$ such that~$\top = \bigvee_{i \in I'} U_i$.
+With this definition, a topological space~$Y$ is compact iff its induced locale~$L(Y)$ is.
 
@@
 gives rise to a point of~$X$.
+% XXX exercise
 
@@
 For instance, any locale induced by a topological space is spatial. The locale
-of surjections~$\NN \twoheadrightarrow \RR$ is a drastic example of a locale which fails to be
+of surjections~$\NN \twoheadrightarrow \RR$ is a striking example of a locale which fails to be
 spatial. The localic real line is spatial in classical mathematics and can
@@
 axiomatizable. Many interesting geometric theories are not, and indeed contain
-an uncountable number of axioms. Geometric theories also allow for arbitrary
+an uncountable number of sorts, function symbols, relation symbols or axioms. Geometric theories also allow for arbitrary
 set-indexed disjunctions; for this to make sense, the study of geometric
@@
 theories often depend on given mathematical objects. For instance, for each
-ring~$A$, there is the geometric theory of prime ideals of~$A$.
+ring~$A$, there is the geometric theory of prime ideals of~$A$. This theory
+will play a crucial role in Section~\ref{sect:appl}.
 \end{enumerate}
@@
 $0$ and $1$ (nullary), $-$ (unary), $+$ and $\cdot$ (binary); no relation
-symbols; the usual axioms, such as $\top \vdash_{x:R,y:R} x + y = y + x$.
+symbols; and the usual axioms, such as $\top \vdash_{x:R,y:R} x + y = y + x$.
 
@@
 ideals has no sorts; no function
-symbols; one relation symbol~$\alpha_x$ for each element~$x \in A$; and the
+symbols; one relation symbol~``$V(x)$'' for each element~$x \in A$; and the
 following axioms:
 \begin{align*}
-  \top &\vdash \alpha_0 \\
-  \alpha_x \wedge \alpha_y &\vdash \alpha_{x+y} & \text{(for each~$x,y \in A$)} \\
-  \alpha_x &\vdash \alpha_{rx} & \text{(for each~$r,x \in A$)} \\
-  \alpha_1 &\vdash \bot \\
-  \alpha_{xy} &\vdash \alpha_x \vee \alpha_y & \text{(for each~$x,y \in A$)}
+  \top &\vdash V(0) \\
+  V(x) \wedge V(y) &\vdash V(x+y) & \text{(for each~$x,y \in A$)} \\
+  V(x) &\vdash V(xy) & \text{(for each~$x,y \in A$)} \\
+  V(1) &\vdash \bot \\
+  V(xy) &\vdash V(x) \vee V(y) & \text{(for each~$x,y \in A$)}
 \end{align*}
@@
 Hence the theory of prime filters of~$A$ has no sorts, no function symbols, one
-relation symbol~$\alpha_x$ for each element~$x \in A$ and the following
+relation symbol~``$D(x)$'' for each element~$x \in A$ and the following
 axioms:
 \begin{align*}
-  \alpha_0 &\vdash \bot \\
-  \alpha_{x+y} &\vdash \alpha_x \vee \alpha_y & \text{(for each~$x,y \in A$)} \\
-  \alpha_{rx} &\vdash \alpha_x & \text{(for each~$r,x \in A$)} \\
-  \top &\vdash \alpha_1 \\
-  \alpha_x \wedge \alpha_y &\vdash \alpha_{xy} & \text{(for each~$x,y \in A$)}
+  D(0) &\vdash \bot \\
+  D(x+y) &\vdash D(x) \vee D(y) & \text{(for each~$x,y \in A$)} \\
+  D(xy) &\vdash D(x) & \text{(for each~$x,y \in A$)} \\
+  \top &\vdash D(1) \\
+  D(x) \wedge D(y) &\vdash D(xy) & \text{(for each~$x,y \in A$)}
 \end{align*}
@@
 The localic open unit interval is the classifying locale of the
-same theory, but with the added axiom~$\top \vdash \alpha_0 \wedge \beta_1$.
+same theory, but with the additional axiom~$\top \vdash \alpha_0 \wedge \beta_1$.
 \item The \emph{locale of surjections~$\NN \twoheadrightarrow \RR$} is the
@@
 In general, different theories can give rise to isomorphic classifying locales;
-such theories are called \emph{Morita-equivalent}. A vast research program is
-built on this observation~\cite{caramello:tst}.
+such theories are called \emph{Morita-equivalent}. Olivia Caramello has built a vast research program
+on this observation~\cite{caramello:tst}.
 
@@
 amount of data encoded in a single presheaf (one set for each open). The
-situation should be compared to Kripke models for a first-order language~$L$. The
+situation should be compared to the perhaps more familiar Kripke models for a first-order language~$L$. The
 preorder~$W$ of nodes of such a model is like the frame of opens of a locale, and the
@@
 \item A presheaf~$F$ is a \emph{sheaf}
-iff for any compatible family~$(s_i)_i$ with respect to an open covering~$U =
+iff for any compatible family~$(s_i)_i$ with respect to any open covering~$U =
 \bigvee_i U_i$ there is a unique section~$s \in F(U)$ such
-that~$s|^U_{U_i} = s_i$ for all~$i$.
+that~$s|^U_{U_i} = s_i$ for all indices~$i$ (``compatible sections glue'').
 \end{enumerate}
@@
 notation. Namely, if~$\varphi$ is a formula over an open~$U$ and if~$V$ is an
-open such that~$V \preceq U$, then~``$V \models \varphi$'' is officially not
+open such that~$V \preceq U$, then~``$V \models \varphi$'' is formally not
 well-defined, as~$\varphi$ is not a formula over~$V$. When we still write~``$V
@@
 \section{Applications in constructive algebra}
+\label{sect:appl}
 
@@
 There is an interesting origin story to this sheaf, which we tell in
-Section~\ref{sect:origins}. Finally, we illustrate how to use the technique in
+Section~\ref{sect:origins}. For computations, it is vital to have an explicit
+description of this sheaf available; such a description is presented in
+Section~\ref{sect:understanding}. Finally, we illustrate how to use the technique in
 a series of concrete examples in Section~\ref{sect:example-applications}.
@@
 Let~$A$ be a ring. Then there is a certain locale~$X$ on which a
-certain ``mirror image'' of~$A$ exists, a sheaf~$A^\sim$ of rings on~$X$. On the one hand,
+certain ``mirror image'' of~$A$ exists, a sheaf~$A^\sim$ of rings. On the one hand,
 this mirror image is ``close to~$A$'', such that studying~$A^\sim$ pays off for
@@
 \paragraph{Back and forth.}
-Along with statement~(\ref{item:stalks}), statement~(\ref{item:iso}) is the main
-reason why results about~$A^\sim$ imply results about~$A$. For instance, if we
+Statement~(\ref{item:iso}) is a kind of conservativity result.
+For instance, if we
 want to show that a certain ring element~$x \in A$ is zero, we can just as well
@@
 wider context (recognizing it as a small shadow of algebraic geometry's
-\emph{quasicoherence}) and its its usefulness for constructive algebra
+\emph{quasicoherence}) and its usefulness for constructive algebra
 appreciated~\cite[Sections~3.3 and~11.4]{blechschmidt:phd}.
@@
 we only set out to construct a local ring.
-\item There is general machinery for free constructions of this kind, and this
-machinery is \emph{not} applicable to construct free fields.
+\item The field property of~$A^\sim$ should be appropriately appreciated: While
+there is a general machinery for free constructions of this kind, this
+machinery is \emph{not} applicable for constructing free fields.
 \end{enumerate}
@@
 for any homomorphism~$G \to M$ into an abelian group, there is exactly one
-group homomorphism~$G^\ab \to M$ rendering the diagram
+homomorphism~$G^\ab \to M$ rendering the diagram
 \[ \xymatrix{
@@
 The free abelian group over~$G$ can be obtained as the \emph{term model} of the
-theory of abelian groups extended by one constant~$x_g$ for each element~$g \in
+theory of abelian groups extended by a constant~$x_g$ for each element~$g \in
 G$ and the axioms
@@
 that~$G$ can be interpreted in it (that is, that there is a homomorphism~$G \to
-G^\ab$).
+G^\ab$), and nothing more.
 
@@
 over the given map~$X \to K$ by a unique field homomorphism~$K \to L$.
-% Exercise: proof this!
+% XXX Exercise: proof this!
 
@@
 property holds: Any ring homomorphism~$A \to B$ into a local ring uniquely
-factors via a local homomorphism over~$A \to A'$ as indicated in the diagram.
+factors via a local ring homomorphism over~$A \to A'$ as indicated in the diagram.
 \[ \xymatrix{
@@
 well-defined map~$f' : A_\fff \to B,\,\frac{x}{s} \mapsto f(s)^{-1}f(x)$ over
-the canonical morphism~$A \to A_\fff,\,x \mapsto \frac{x}{1}$. Furthermore,
+the localization morphism~$A \to A_\fff,\,x \mapsto \frac{x}{1}$. Furthermore,
 this map is local.
@@
 \subsection{Understanding the sheaf model}
+\label{sect:understanding}
 
@@
 
+\begin{remark}An immediate corollary of Theorem~\ref{thm:workhorse} is
+that~$\Spec(A)$ is compact. However, the stronger result that~$\Spec(A)$ is
+even a \emph{coherent locale} can also be deduced from a general theorem (XXX: ref to
+exercise) since the theory of prime filters is a coherent theory.\end{remark}
+
+Having identified the underlying frame of~$\Spec(A)$, the second workhorse is
+an explicit description of the sheaf~$A^\sim$.
+
+\begin{proposition}\label{prop:identification}
+Let~$A$ be a ring. For any~$f \in A$, there is a canonical isomorphism
+\[ A[f^{-1}] \longrightarrow A^\sim([D(f)]). \]
+In particular, for~$f = 1$, we obtain statement~(\ref{item:iso}) from
+Section~\ref{sect:a-remarkable-sheaf}.
+\end{proposition}
+
+\begin{proof}For~$f \in A$, we set~$S_f \defeq \{ g \in A \,|\, \exists n \in
+\NN\_ \exists u \in A\_ f^n = ug \} \subseteq A$. This multiplicative system is
+the \emph{saturation} of~$\{ f^0, f^1, \ldots \}$. The localization~$A[S_f^{-1}]$
+is canonically isomorphic to~$A[f^{-1}]$, but has the advantage that for
+elements~$f,g \in A$ such that~$\sqrt{(f)} = \sqrt{(g)}$, the
+rings~$A[S_f^{-1}]$ and~$A[S_g^{-1}]$ are actually \emph{the same}
+while~$A[f^{-1}]$ and~$A[g^{-1}]$ are merely canonically isomorphic. This
+property eases formalization in set theory; it is otherwise not important.
+
+On the basis of~$\Spec(A)$, the opens of the form~$[D(f)]$, we define a partial
+presheaf by
+\[ [D(f)] \longmapsto A[S_f^{-1}]. \]
+This partial presheaf is, on the basis where it is defined, a sheaf. Unraveling
+the definitions, this claim boils down to the following basic result in
+commutative algebra:
+\begin{quote}
+Let~$B$ be a ring. Let~$1 = f_1 + \cdots + f_m \in B$ be a partition of unity. Let
+elements~$s_i \in B[f_i^{-1}]$ be given. Assume that~$s_i = s_j$
+in~$B[(f_if_j)^{-1}]$ for all pairs of indices. Then there is exactly one
+element~$s \in B$ such that, for all indices~$i$,~$s = s_i$ in~$B[f_i^{-1}]$.
+\end{quote}
+The remainder of the argument consists of sheaf-theoretic generalities.
+XXX
+\end{proof}
+
+More generally, for any~$A$-module~$M$, we can construct the
+localization~$M^\sim \defeq \underline{M}[\fff_0^{-1}]$. In the special case
+that~$M$ is~$A$, this definition coincides with our original definition
+of~$A^\sim$. The proof of Proposition~\ref{prop:identification} carries over to
+show that~$M^\sim([D(f)]) \cong M[f^{-1}]$.
+
+As a corollary of Theorem~\ref{thm:workhorse} and
+Proposition~\ref{prop:identification}, we can unwind the definitions to recast
+the sheaf semantics in entirely explicit algebraic terms, with no locales or
+sheaves in sight. By a \emph{formula over a ring element}~$f \in A$, we mean a
+first-order formula over the signature which has
+\begin{enumerate}
+\item a sort~``$A^\sim$'' and function symbols for the structure of a ring,
+\item for each~$A$-module~$M$, a sort~``$M^\sim$'' and function symbols for the
+structure of an~$A^\sim$-module,
+\item for each linear map~$M \to N$, a function symbol~$M^\sim \to N^\sim$,
+\item a constant of sort~$A^\sim$ for each element of~$A[f^{-1}]$ and
+\item for any~$A$-module~$M$, a constant of sort~$M^\sim$ for each element of~$M[f^{-1}]$.
+\end{enumerate}
+
+\begin{corollary}Let~$A$ be a ring. Let~$\varphi$ be a formula over~$f \in A$.
+Then~$[D(f)] \models \varphi$ in the sense of Definition~XXX if and only
+if~$f \models \varphi$ in the sense of Table~\ref{table:algebraic-kripke-joyal}.
+\end{corollary}
+
+\begin{proof}Induction on the structure of~$\varphi$, harnessing the explicit
+descriptions provided by Theorem~\ref{thm:workhorse} and
+Proposition~\ref{prop:identification}.\end{proof}
+
+\begin{table}[ht]
+  \tbl{A purely algebraic presentation of the sheaf semantics of~$\Spec(A)$}{\begin{tabular}{@{}l@{\ \ }c@{\ \ }l@{}}
+    $f \models \top$ &iff& true \\
+    $f \models \bot$ &iff& $f$ is nilpotent \\
+    $f \models x =_{M^\sim} y$ &iff& $x = y \in M[f^{-1}]$ \\
+    $f \models \varphi \wedge \psi$ &iff&
+      $f \models \varphi$ and $f \models \psi$ \\
+    $f \models \varphi \vee \psi$ &iff&
+      there exists a partition~$f^n = fg_1 + \cdots + fg_m$ such that, \\
+    &&\quad for each~$i$, $fg_i \models \varphi$ or $fg_i \models \psi$ \\
+    $f \models \varphi \Rightarrow \psi$ &iff&
+      for all~$g \in A$, $fg \models \varphi$ implies $fg \models \psi$ \\
+    $f \models \forall x\?A^\sim\_ \varphi(x)$ &iff&
+      for all~$g \in A$ and all $x_0 \in A[(fg)^{-1}]$, $fg \models \varphi(x_0)$ \\
+    $f \models \exists x\?A^\sim\_ \varphi(x)$ &iff&
+      there exists a partition~$f^n = fg_1 + \cdots + fg_m$ such that, \\
+    &&\quad for each~$i$, $fg_i \models \varphi(x_0)$ for some~$x_0 \in A[(fg_i)^{-1}]$
+  \end{tabular}}
+  \label{table:algebraic-kripke-joyal}
+\end{table}
+
 \subsection{Applications}

### 2020-04-01 paper-herrsching2019/main.tex
Field property and examples (stubs)

@@
 
-\begin{theorem}Let~$X$ be a locale. Let~$U$ be an open of~$X$. Let~$\varphi$ be
+\begin{theorem}\label{thm:basic-properties-sheaf-semantics}
+Let~$X$ be a locale. Let~$U$ be an open of~$X$. Let~$\varphi$ be
 a formula over~$U$.
@@
 \textbf{Proof.} \uwave{Assume not.} Then there is a \uwave{minimal
-prime ideal} $\ppp \subseteq A$. The matrix~$M$ is also injective when
-considered as a matrix over the stalk~$A_\ppp$; since~$A_\ppp$ is a
+prime ideal} $\ppp \subseteq A$. Since localization is exact, the matrix~$M$ is also injective when
+considered as a matrix over the stalk~$A_\ppp$. Since~$A_\ppp$ is a
 \uwave{field}, this is a contradiction to basic linear algebra. \qed
@@
 equivalent iff there is an element~$u \in S$ such that~$uta = usb$. We do not
-require that~$0 \not\in S$; instead we plainly observe that ~$A[S^{-1}]$ is the
-zero ring iff~$0 \in S$.
+require that~$0 \not\in S$; instead we plainly observe that~$A[S^{-1}]$ is the
+zero ring iff~$0 \in S$. We also remark that~$\frac{a}{s} = 0$ in~$A[S^{-1}]$
+iff~$ua = 0$ for some element~$u \in S$.
 
@@
 \addtocounter{enumi}{4}
-\item\label{item:field} $A^\sim$ is a field: $\forall x\?A^\sim\_ (\neg(\exists y\?A^\sim\_ xy = 1) \Rightarrow x = 0)$.
+\item\label{item:field} $A^\sim$ is a field: $\forall x\?A^\sim\_ \Bigl((\neg(\exists y\?A^\sim\_ xy = 1)) \Rightarrow x = 0\bigr)$.
 \item\label{item:negneg-stable} $A^\sim$ has $\neg\neg$-stable equality:
@@
 over the given map~$X \to K$ by a unique field homomorphism~$K \to L$.
-% XXX Exercise: proof this!
+% XXX Exercise: prove this!
 
@@
 
+\begin{remark}\label{rem:consistency}
+A further corollary of Theorem~\ref{thm:workhorse} is that the theory of prime
+filters of~$A$ is inconsistent (that it, that it proves the sequent~$\top
+\vdash \bot$, which can also be written as~$D(1) \vdash D(0)$) if and only
+if~$A$ is the trivial ring. Contrapositively, if~$A$ is not the trivial ring,
+then the theory is consistent. This observation can be regarded as a finitary
+substitute for the statement (equivalent to~\BPIT in general) ``nontrivial rings contain prime filters''.
+\end{remark}
+
 Having identified the underlying frame of~$\Spec(A)$, the second workhorse is
@@
 
+\paragraph{An elementary reformulation.}
 As a corollary of Theorem~\ref{thm:workhorse} and
@@
 the sheaf semantics in entirely explicit algebraic terms, with no locales or
-sheaves in sight. By a \emph{formula over a ring element}~$f \in A$, we mean a
+sheaves in sight. Notwithstanding the impredicative nature of locale theory,
+the resulting formulation will even make sense in a metatheories without a
+powerset operation such as~\textsc{czf} or arithmetic universes~\cite{XXX}. In
+fact, it is now a purely syntactic translation procedure which can be carried
+out within~\textsc{pra}.
+
+By a \emph{formula over a ring element}~$f \in A$, we mean a
 first-order formula over the signature which has
@@
 
-\begin{corollary}Let~$A$ be a ring. Let~$\varphi$ be a formula over~$f \in A$.
+\begin{corollary}\label{cor:algebraic-reformulation}
+Let~$A$ be a ring. Let~$\varphi$ be a formula over~$f \in A$.
 Then~$[D(f)] \models \varphi$ in the sense of Definition~XXX if and only
@@
 
+The analogue of Theorem~\ref{thm:basic-properties-sheaf-semantics} for the new
+formulation of the sheaf semantics reads as follows.
+
+\begin{theorem}Let~$A$ be a ring. Let~$f \in A$. Let~$\varphi$ be a formula
+over~$f$.
+\begin{enumerate}
+\item Let~$g \in A$. If~$f \models \varphi$, then~$fg \models \varphi$.
+\item Let~$f^n = fg_1 + \cdots + fg_m$. If~$fg_i \models \varphi$ for all
+indices~$i$, then~$f \models \varphi$.
+\item Let~$\psi$ be a further formula over~$f$. If~$f \models
+\varphi$ and if~$\varphi$ entails~$\psi$ intuitionistically, then also~$f
+\models \psi$.
+\end{enumerate}
+\end{theorem}
+
+\begin{proof}Combination of Theorem~\ref{thm:basic-properties-sheaf-semantics}
+and Corollary~\ref{cor:algebraic-reformulation}.\end{proof}
+
+\paragraph{The field property.} We are now in a position to verify
+statement~(\ref{item:field}) of Section~\ref{sect:a-remarkable-sheaf}.
+
+\begin{proposition}\label{prop:field-property}
+Let~$A$ be a reduced ring. Then~$A^\sim$ is a field in that
+\[ 1 \models \forall x\?A^\sim\_ \bigl((\neg(\exists y\?A^\sim\_ xy = 1)) \Rightarrow
+x = 0\bigr). \]
+\end{proposition}
+
+\begin{proof}By Table~\ref{table:algebraic-kripke-joyal}, we have to verify the
+following claim.
+\begin{quote}
+For any element~$f \in A$ and any element~$x \in A[f^{-1}]$, \\
+${\qquad}$ for any element~$g \in A$, \\
+${\qquad\qquad}$ if for any~$h \in A$, \\
+${\qquad\qquad\qquad}$ if there is~$(fgh)^n = \sum_{i=1}^m fghp_i$
+such that for each~$i$, \\
+${\qquad\qquad\qquad\qquad}$ there exists~$y \in
+A[(fghp_i)^{-1}]$ such that \\
+${\qquad\qquad\qquad\qquad\qquad}$ $xy = 1$ in~$A[(fghp_i)^{-1}]$, \\
+${\qquad\qquad\qquad}$ then~$fgh$ is nilpotent, \\
+${\qquad\qquad}$ then~$x = 0$ in~$A[(fg)^{-1}]$.
+\end{quote}
+Hence let elements~$f \in A$, $x \in A[f^{-1}]$ and~$g \in A$ be given and
+assume that for any~$h \in A$, the displayed condition holds. We are to show
+that~$x = 0$ in~$A[(fg)^{-1}]$.
+
+Write~$x = \frac{x'}{f}$. For~$h \defeq x'$, we have the
+partition~$(fgh)^1 = fgh \cdot 1$ with~$m = 1$ and~$p_1 = 1$. For~$y \defeq
+\frac{f}{x'} = \frac{f^2g}{fgh}$, we have~$xy = 1$ in~$A[(fghp_1)^{-1}]$.
+By assumption, the element~$fgh$ is nilpotent, hence zero. Thus~$x' = 0$
+in~$A[(fg)^{-1}]$. In particular, we have~$x = 0$ in~$A[(fg)^{-1}]$.
+\end{proof}
+
+The proof of Proposition~\ref{prop:field-property} visibly demonstrates that
+unwinding the definition of the sheaf semantics quickly results in formulas
+which are quite convoluted. This behavior is a basic consequence of the fact
+that, with the exception of the clauses for~``$\top$'' and~``$\wedge$'', all
+clauses in Table~\ref{table:algebraic-kripke-joyal} introduce additional
+quantifiers.
+
+This increase in complexity should be properly appreciated: The sheaf semantics
+allows us to harness convoluted properties in simple language. The unrolled
+monstrosity in the proof of Proposition~\ref{prop:field-property} is not
+memorable at all and unlikely to find its way into human-written proofs.
+However, in the form~``$A^\sim$ is a field'', it becomes easily accessible.
+
+This state of affairs is even more pronounced for higher-order formulas such as
+the condition to be anonymously Noetherian. For instance, XXX
+
+\begin{quote}
+For any element~$f \in A$ and any (not necessarily quasicoherent) sheaf of
+ideals~$\J \hookrightarrow A^\sim[X_1,\ldots,X_n]|_{D(f)}$: If
+\begin{quote}
+for any element~$g \in A$ the condition that
+\begin{quote}
+the sheaf~$\J$ is of finite type on~$D(g)$
+\end{quote}
+implies that~$g = 0$,
+\end{quote}
+then~$f = 0$.
+\end{quote}
+
 \subsection{Applications}
@@
 
-XXX injective matrices
-
-XXX generic freeness
-
-XXX spell out details of field property
+\begin{proposition}Let~$M$ be an injective matrix with more columns than rows
+over a reduced ring~$A$. Then~$1 = 0$ in~$A$.
+\end{proposition}
+
+\begin{proof}Since localization is exact, the matrix~$M$ is also injective when
+considered as a matrix over~$A^\sim$ (statement (\ref{item:loc-stable})).
+Since~$A^\sim$ is a field, this is a contradiction to basic (intuitionistic)
+linear algebra. In other words, $\Spec(A) \models \bot$. By
+Remark~\ref{rem:consistency}, this amounts to~$1 = 0$ in~$A$.
+\end{proof}
+
+\begin{theorem}(Grothendieck's generic freeness lemma, simple version)
+Let~$A$ be a reduced ring. Let~$M$ be a finitely generated~$A$-module.
+If~$f = 0$ is the only element of~$A$ such that~$M[f^{-1}]$ is a
+free~$A[f^{-1}]$-module, then~$1 = 0$ in~$A$.
+\end{theorem}
+
+\begin{proof}The claim amounts to the statement that~$M^\sim$ is \emph{not not}
+free. Since~$A^\sim$ is a field, this follows from
+basic (intuitionistic) linear algebra.\footnote{XXX case distinction}\end{proof}
+
+XXX can be unwound, see ...; give pointers to existing proofs
 
@@
 X \models phi
-
-  &\text{iff}\quad \bigvee_{n \in \NN} \exists m \in \NN\_ \exists i_1,\ldots,i_n \in I\_ \exists
-  u_1,\ldots,u_n \in A\_ f^m = \sum_{j=1}^n u_j g_{i_j}

### 2020-04-01 paper-herrsching2019/main.tex
Many exercises

@@
 \usepackage{mathtools,stmaryrd,xspace}
+
+\newlength{\exerciseskip}
+\setlength{\exerciseskip}{1.5em}
+\newcounter{exercisenr}
+\newenvironment{exercise}[1]{
+  \refstepcounter{exercisenr}
+  \noindent\textbf{Exercise \theexercisenr{}.} \emph{#1} \smallskip\par\noindent%
+}{\par\vspace{\exerciseskip}}
 
@@
 \newcommand{\pt}{\mathrm{pt}}
+\newcommand{\Pt}{\mathrm{Pt}}
 \newcommand{\Loc}{\mathrm{Loc}}
@@
 \newcommand{\BPIT}{\textsc{bpit}\xspace}
+\newcommand{\notnot}{\emph{not not}\xspace}
+\newcommand{\negg}{\neg\!\!\!\neg}
+\newcommand{\bott}{\bot\!\!\!\!\bot}
 
@@
 \begin{remark}The set of points of a locale~$X$ can be made into a topological space, giving
-rise to a functor~$\pt : \Loc \to \Top$. This functor is right adjoint
+rise to a functor~$\Pt : \Loc \to \Top$. This functor is right adjoint
 to the functor~$L : \Top \to \Loc$. A locale~$X$ is spatial iff the canonical
-morphism~$L(\pt(X)) \to X$ is an isomorphism, and a topological space~$Y$
-is \emph{sober} iff the canonical morphism~$Y \to \pt(L(Y))$ is a
-homeomorphism. The space~$\pt(L(Y))$ is the \emph{sobrification} of~$Y$; for
+morphism~$L(\Pt(X)) \to X$ is an isomorphism, and a topological space~$Y$
+is \emph{sober} iff the canonical morphism~$Y \to \Pt(L(Y))$ is a
+homeomorphism. The space~$\Pt(L(Y))$ is the \emph{sobrification} of~$Y$; for
 instance, the sobrification of any inhabited indiscrete space is the one-point
 space.\end{remark}
+
+\begin{remark}\label{rem:sober-embed}
+Since continuous maps between spaces induce locale morphisms, we
+have a canonical map~$\Hom_\Top(Y,Y') \to \Hom_\Loc(L(Y),L(Y'))$. If~$Y'$ is
+sober, then this map is a bijection. In particular, the set-theoretic points
+of~$Y'$ and the locale-theoretic points of~$L(Y')$ are in canonical
+one-to-one correspondence.
+\end{remark}
+
+\subsection{Exercises}
+
+\begin{exercise}{The topological space of points of a locale}%
+\label{ex:canonical-topology}%
+Devise a canonical topology on the set of points of a locale.
+\end{exercise}
+
+\begin{exercise}{On the embedding of sober topological spaces into locales}%
+Verify the statements in Remark~\ref{rem:sober-embed}.
+\end{exercise}
+
+\begin{exercise}{Isomorphisms of frames}%
+Show that a frame homomorphism~$\alpha : A \to A'$ is an isomorphism (that is,
+admits an inverse frame homomorphism) if and only if~$\alpha$ is surjective and
+reflects the ordering (that is,~$\alpha(U) \preceq \alpha(V) \Rightarrow U
+\preceq V$).
+\end{exercise}
+
+\begin{exercise}{Points as complete prime filters}%
+Verify that the points of a locale~$X$ are in canonical one-to-one correspondence
+with the completely prime filters of~$\O(X)$.
+\end{exercise}
+
+\begin{exercise}{Characterizing spatial locales}%
+Let~$X$ be a locale. Show that the following conditions are equivalent.
+\begin{enumerate}
+\item For any opens~$U, V \in \O(X)$: If~$x \inplus U \Rightarrow y \inplus V$
+for all points~$x$ of~$X$, then~$U \preceq V$.
+\item For any opens~$U, V \in \O(X)$: If~$x \inplus U \Leftrightarrow y \inplus V$
+for all points~$x$ of~$X$, then~$U = V$.
+\item There is a topological space~$Y$ and a surjective morphism~$L(Y) \to X$
+of locales. (A morphism of locales is \emph{surjective} iff its underlying
+frame homomorphism is injective.)
+\item The canonical morphism~$L(\Pt(X)) \to X$ is an isomorphism.
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}{Coproducts of locales}%
+\begin{enumerate}
+\item Show that the category of frames has small products.
+\item Deduce that the category of locales has small coproducts.
+\item Let~$M$ be a set and endow it with the discrete topology. Show
+that~$L(M)$ is the coproduct~$\coprod_{x \in M} \pt$.
+\end{enumerate}
+The category of locales also has small products;
+Exercise~\ref{ex:products-of-locales} is devoted to this fact.
+\end{exercise}
+
+\begin{exercise}{Local locales}%
+A locale~$X$ is \emph{local} iff for any open covering~$\top = \bigvee_i U_i$,
+there is an index~$i$ such that~$\top = U_i$. (For instance, the spectrum of a
+ring~$A$ is local iff~$A$ is local.)
+Show that any local locale has a canonical point, the \emph{focal point}.
+\end{exercise}
+
+\begin{exercise}{Localic maps into discrete spaces}%
+Let~$X$ be a locale. Let~$A$ be a set and endow it with the discrete topology. Show
+that morphisms~$X \to L(A)$ are in canonical correspondence with
+families~$(U_a)_{a \in A}$ of opens of~$X$ such that~$\top = \bigvee_{a \in A}
+U_a$ and~$U_a \wedge U_b \preceq \bigvee\{ \top \,|\, a = b \}$.
+\end{exercise}
 
@@
 
-\paragraph{The geometric theory of Dedekind cuts.} This theory has no sorts; no
+\paragraph{The geometric theory of Dedekind cuts.}\label{par:dedekind-cuts}
+This theory has no sorts; no
 function symbols; nullary relation symbols~$\alpha_x$ and~$\beta_x$ for each
@@
 The prototypical example of a presheaf (and also a sheaf) is the presheaf~$\C$ of
-continuous real-valued functions on a space~$X$. For this presheaf, the
+continuous real-valued functions on a topological space~$X$. For this presheaf, the
 set~$\C(U)$ is the set of continuous real-valued maps~$U \to \RR$, and the
@@
 XXX ``internalizing parameter-dependence''
+
+\subsection{Exercises}
+
+\begin{exercise}{Discrete locales as classifying locales}%
+\begin{enumerate}
+\item Find a theory which is classified by the two-point locale (the locale
+associated to the discrete two-point topological space).
+\item Let~$M$ be a set and endow it with the discrete topology. Find a theory
+which is classified by~$L(M)$.
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}{The localic unit interval}%
+Extend the theory of Dedekind cuts in such a way that its models are those
+Dedekind reals which are elements of~$[0,1]$.
+\end{exercise}
+
+\begin{exercise}{More on exotic locales}
+\begin{enumerate}
+\item Define the locale of injections~$\RR \hookrightarrow \NN$.
+\item Define the locale of partial surjections~$\NN \twoheadrightarrow \NN^\NN$.
+
+{\scriptsize\emph{Note.} In Russian constructivism, as for instance reigning in
+the ef{}fective topos~\cite{XXX}, the Baire space~$\NN^\NN$ is subcountable.
+However, in this setting any function~$\NN \to \NN$ is computable.
+The universe of sheaves over this locale was the first example of a setting in
+which the Baire space is subcountable and still any external number-theoretic
+function can be imported~\cite[Section~4.3]{scedrov:forcing}.\par}
+\item Define the locale of ``open covers of~$[0,1]$ by intervals with rational
+endpoints such that the union of any finite number of those intervals has total
+Lebesgue measure less than~$\frac{1}{2}$''.
+
+{\scriptsize\emph{Note.} In the universe of sheaves over this locale, the set
+of unit interval Dedekind real numbers fails to be (Heine--Borel) compact and
+the localic real line fails to be
+spatial~\cite[Example~D4.7.13]{johnstohne:elephant}.\par}
+\item An open~$U$ of a locale~$X$ is \emph{complemented} iff there is an
+open~$V$ such that~$U \vee V = \top$ and~$U \wedge V = \bot$. A locale is
+\emph{zero-dimensional} iff any open is a join of complemented opens. Assuming
+that any two reals are equal or not, show that the locale of surjections~$\NN
+\to \RR$ is zero-dimensional.
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}{Models of theories as points of the classifying locales}%
+Let~$\TT$ be a propositional geometric theory. Verify that the points
+of~$L(\TT)$ are in canonical bijection with the set-based models of~$\TT$.
+\end{exercise}
+
+\begin{exercise}{Products of locales}\label{ex:products-of-locales}%
+Let~$X$ and~$X'$ be locales, classifying some propositional geometric
+theory~$\TT$ respectively~$\TT'$. Let~$\TT + \TT'$ be the combined theory
+(whose set of sorts is the disjoint union of the sets of sorts of~$\TT$
+and~$\TT'$, and so on). Show that~$L(\TT + \TT')$ is the product of~$X$
+and~$X'$ in the category of locales.
+\end{exercise}
+
+\begin{exercise}{Coproducts of locales, revisited}%
+Let~$X$ and~$X'$ be locales, classifying some propositional geometric
+theory~$\TT$ respectively~$\TT'$. Find a theory in terms of~$\TT$
+and~$\TT'$ which is classified by~$X \amalg X'$.
+\end{exercise}
+
+\begin{exercise}{Spatiality of the localic reals}%
+\newcommand{\fa}{\ensuremath{(\!\cdot}}%
+\newcommand{\fb}{\ensuremath{\cdot\!)}}%
+Let~$\RR$ be the localic real line, defined as the classifying locale of the
+theory~$\TT$ of Dedekind cuts as as introduced on
+page~\pageref{par:dedekind-cuts}. For rational numbers~$x,y$, let~``$\fa x,y
+\fb$'' denote the formula~$\alpha_x \wedge \beta_y$. Let~$(x,y)$ denote the
+open interval~$\{ \xi \in \Pt(\RR) \,|\, x < \xi < y \}$ in the set of Dedekind reals.
+\begin{enumerate}
+\item Extend the notation to allow for~$x,y=\pm\infty$.
+\item Show that the canonical topology on~$\Pt(\RR)$, defined in
+Exercise~\ref{ex:canonical-topology}, coincides with the usual Euclidean
+topology.
+\item Show that any open of~$\RR$ can be written in the form~$\bigvee_i
+[\fa x_i,y_i \fb]$ with~$x_i,y_i \in \QQ \cup \{\pm\infty\}$.
+\item Let rational numbers be given such that~$(x,y) \subseteq \bigcup_i
+(x_i,y_i)$ in~$\Pt(\RR)$. Assume that this open covering has a Lebesgue number.
+Show that~$\TT$ proves~$\fa x,y \fb \vdash \bigvee_i \fa x_i,y_i \fb$.
+\item Assume that closed intervals in the set of Dedekind reals are
+(Heine--Borel) compact. Then redo part~(d), but without the assumption on the
+existence of a Lebesgue number.
+\item Under the same assumption as in part~(e), show that the localic real line
+is spatial.
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}{The localic Galois group of a field extension}%
+Let~$L|k$ be a field extension. The \emph{localic Galois group}~$\Gal(L|k)$
+of~$L|k$ is the classifying locale of the theory which has one nullary relation
+symbol~``$\sigma_{xy}$'' for each pair~$\langle x,y \rangle \in L$ and the
+following axioms:
+\begin{align*}
+  \top &\vdash \bigvee_y \sigma_{xy} & \text{(for each~$x \in L$)} \\
+  \top &\vdash \bigvee_x \sigma_{xy} & \text{(for each~$y \in L$)} \\
+  \sigma_{xy} \wedge \sigma_{xy'} &\vdash \bigvee\{ \top \,|\, y = y' \} & \text{(for each~$x,y,y' \in L$)} \\
+  \sigma_{xy} \wedge \sigma_{x'y} &\vdash \bigvee\{ \top \,|\, x = x' \} & \text{(for each~$x,x',y' \in L$)} \\
+  \top &\vdash \sigma_{00} \\
+  \top &\vdash \sigma_{11} \\
+  \sigma_{xy} \wedge \sigma_{x'y'} &\vdash \sigma_{x+x',y+y'} & \text{(for each~$x,x',y,y' \in L$)} \\
+  \sigma_{xy} \wedge \sigma_{x'y'} &\vdash \sigma_{xx',yy'} & \text{(for each~$x,x',y,y' \in L$)} \\
+  \top &\vdash \sigma_{xx} & \text{(for each~$x \in k$)}
+\end{align*}
+\begin{enumerate}
+\item Explain how to obtain a morphism~$\Gal(L|k) \to \Gal(L|k)$ which on the
+level of points should send an automorphism~$\sigma$ to its inverse~$\sigma^{-1}$.
+\item Explain how to obtain a morphism~$\Gal(L|k) \times \Gal(L|k) \to \Gal(L|k)$ which on the
+level of points should send a pair~$\langle \sigma, \tau \rangle$ to~$\sigma \circ \tau$.
+\item Assume from now on that~$L$ is a \emph{geometric field} in the sense that any
+element is zero or invertible. Show that the axiom expressing injectivity can
+be dropped without changing the set of provable sequents.
+\item Let~$f \in k[X]$ be a monic polynomial such that~$f(X) =
+(X-x_1)\cdots(X-x_n)$ over~$L$. Let~$x$ be one of the zeros of~$f$. Show that
+the theory shows~$\top \vdash \bigvee_{i=1}^n \sigma_{x,x_i}$.
+\item Now furthermore assume that~$L$ is \emph{Galois over}~$k$ in the sense that
+any element of~$L$ is the zero of a monic polynomial~$f \in k[X]$ which is
+separable (that is,~$(f,f') = (1) \subseteq k[X]$) and splits over~$L$ into
+linear factors. Show that the axioms expressing surjectivity can be dropped.
+\item Show that the topological space of points of~$\Gal(L|k)$ is canonically
+homeomorphic to the topological Galois group of~$L|k$.
+\item Show that~$\Gal(L|k)$ is spatial in classical mathematics.
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}{Coherent locales and coherent theories}%
+An open~$U$ of a locale is \emph{compact} iff, whenever~$U
+\preceq \bigvee_{i \in I} V_i$, there is a (Kuratowski-)finite subset~$I'
+\subseteq I$ such that~$U \preceq \bigvee_{i \in I'} V_i$. A locale is
+\emph{coherent} iff its compact opens are closed under finite meets and if
+every open is a join of compact opens. A geometric formula is \emph{coherent} iff only
+finite disjunctions occur in it. A geometric theory is \emph{coherent}
+iff all of its axioms consist of coherent formulas.
+\begin{enumerate}
+\item Show that any coherent locale classifies some coherent theory.
+\item Let~$\TT$ be a propositional coherent theory. Show that~$L(\TT)$ is
+coherent.
+
+{\scriptsize\emph{Hint.} For coherent formulas~$\alpha$ and geometric
+formulas~$\varphi$, define a semantics by the clauses below. Show by an induction on the
+structure of geometric derivations that, if~$\TT$ proves a geometric
+sequent~$\varphi \vdash \psi$, then for all coherent formulas~$\alpha$ such
+that~$\alpha \models \varphi$, $\alpha \models \psi$. Show for coherent
+formulas~$\alpha,\beta$ that~$\alpha \models \beta$ iff~$\TT$ proves~$\alpha
+\vdash \beta$. Use these ingredients to show that, if~$\alpha$ is a coherent
+formula and~$(\beta_i)_{i \in I}$ is a family of coherent formulas, if~$\TT$
+proves~$\alpha \vdash \bigvee_{i \in I} \beta_i$ then there is a
+(Kuratowski-)finite subset~$I' \subseteq I$ such that~$\TT$ proves~$\alpha
+\vdash \bigvee_{i \in I'} \beta_i$.
+
+\begin{center}\tablefont
+\begin{tabular}{@{}l@{\ \ }c@{\ \ }l@{}}
+  $\alpha \models \beta$ &iff& $\TT$ proves~$\alpha \vdash \beta$ \qquad(for~$\beta$ a nullary relation symbol) \\
+  $\alpha \models \top$ &iff& true \\
+  $\alpha \models \varphi \wedge \psi$ &iff& $\alpha \models \varphi$ and $\alpha \models \psi$ \\
+  $\alpha \models \bigvee_i \varphi_i$ &iff& there are coherent
+  formulas~$\alpha_1,\ldots,\alpha_n$ such that \\ &&\qquad $\TT$ proves~$\alpha \vdash
+  \alpha_1 \vee \cdots \vee \alpha_n$ and such for each~$k \in \{1,\ldots,n\}$,
+  \\ &&\qquad\qquad there is an index~$i \in I$ such that~$\alpha_k \models \varphi_i$
+\end{tabular}
+\end{center}}
+
+\item Show that an arbitrary (set-indexed) product of coherent locales is
+coherent.
+
+\item \emph{Deligne's completeness theorem} states that coherent locales (in
+fact, even coherent \emph{toposes}) are spatial. Show that Deligne's
+completeness theorem implies that propositional coherent theories
+without any set-based models are inconsistent.
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}{A geometric interpretation of double negation}%
+An open~$U$ of a locale is \emph{dense} iff for any open~$V$, $U \wedge V =
+\bot$ implies~$U = \bot$. Let~$\varphi$ be a formula over a locale~$X$.
+\begin{enumerate}
+\item Show that there is a largest open,
+denoted~``$\llbracket\varphi\rrbracket$'', such that~$U \models \varphi$.
+
+{\scriptsize\emph{Hint.} Consider the join of all opens~$V$ such that~$V
+\models \varphi$.\par}
+
+\item Show that~$X \models \neg\neg\varphi$ iff~$\llbracket\varphi\rrbracket$
+is dense.
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}{Unique local existence is global existence}%
+\label{ex:global-existence}%
+Let~$F$ be a sheaf on a locale~$X$. Let~$\varphi$ be a formula over~$X$. Assume
+that~$\Sh(X) \models \exists!s\?F\_ \varphi(s)$, that is
+\[ \Sh(X) \models \exists s\?F\_ \varphi(s) \quad\text{and}\quad
+  \Sh(X) \models \forall s\?F\_ \forall t\?F\_ (\varphi(s) \wedge \varphi(t)
+  \Rightarrow s = t). \]
+Show that for any open~$U \in \O(X)$, there is a unique section~$s \in F(U)$
+such that~$U \models \varphi(s)$.
+\end{exercise}
+
+\begin{exercise}{The sheaf of real functions as a field}%
+Let~$X$ be a topological space. Let~$\C$ be the sheaf of continuous real-valued
+functions on~$X$.
+\begin{enumerate}
+\item Let~$U$ be an open of~$X$. Let~$f \in \C(U)$. Show using the result
+of Exercise~\ref{ex:global-existence} that~$U \models (\exists g\?\C\_ fg =_\C
+1)$ iff there is a function~$g \in \C(U)$ such that~$fg = 1$.
+\item Show that~$\C$ is a field in that
+$X \models \forall f\?\C\_
+  \bigl(\neg(\exists g\?\C\_ fg = 1)\bigr) \Rightarrow f = 0$.
+\item Give an example of space~$X$ such that it is not the case that~$\C$ is a
+field in the stronger sense that
+$\Sh(X) \models \forall f\?\C\_ (\exists g\?\C\_ fg = 1) \vee f = 0$.
+\item Let~$X = \CC$ and let~$\O$ be the sheaf of holomorphic functions on~$X$.
+Show that, in classical mathematics, the sheaf~$\O$ is \emph{discrete} in that
+$\Sh(\CC) \models \forall f\?\O\_ \forall g\?\O\_ f = g \vee \neg(f = g)$.
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}{On the failure of the fundamental theorem of algebra}%
+Let~$F$ be the sheaf of continuous complex-valued functions on~$\CC$. Give an
+example of a function~$f : \CC \to \CC$ such that it is not the case that
+\[ \Sh(X) \models \exists z\?F\_ z^2 = f. \]
+{\scriptsize\emph{Note.} The sheaf~$F$ can be obtained by constructing, in the
+universe of sheaves on~$\CC$, the set of pairs of Dedekind reals. Hence the
+example demonstrates that the usual formulation of the fundamental theorem of
+algebra does not admit a constructive proof. However, there are replacements
+available~\cite{XXX}.\par}
+\end{exercise}
 
@@
 
-\begin{theorem}Let~$A$ be a ring. Let~$f \in A$. Let~$\varphi$ be a formula
-over~$f$.
+\begin{theorem}\label{thm:basic-properties-sheaf-semantics-algebraic}
+Let~$A$ be a ring. Let~$f \in A$. Let~$\varphi$ be a formula over~$f$.
 \begin{enumerate}
@@
 Since~$A^\sim$ is a field, this is a contradiction to basic (intuitionistic)
-linear algebra. In other words, $\Spec(A) \models \bot$. By
+linear algebra.\footnote{We are referencing the following theorem: ``Let~$M$ be
+an injective matrix with more columns than rows over a field~$k$ in the sense
+that any element of~$k$ is either zero or invertible. Then~$\bot$.'' This
+theorem is proven by repeatedly applying elementary row and column
+transformations until the matrix is of the
+form~$\left(\begin{smallmatrix}I&0\\0&0\end{smallmatrix}\right)$. By assumption
+on the size of~$M$, there is at least one zero column. This contradicts
+injectivity.
+
+However, since the ring~$A^\sim$ is only a field in the weaker sense that
+nonunits are zero, this theorem does not directly apply. We only have that any
+element of~$A^\sim$ is \emph{not not} zero or invertible. We hence employ a
+standard trick: In intuitionistic proofs of~$\bot$, we may use the law of
+excluded middle a finite number of times. This is because in intuitionistic
+(and even minimal) logic, we have~$\neg\neg(\varphi \vee \neg\varphi)$;
+$\neg\neg\varphi \wedge (\varphi \Rightarrow \neg\neg\psi) \Longrightarrow
+\neg\neg\psi$; and~$\neg\neg\bot \Leftrightarrow \bot$.}
+In other words, $\Spec(A) \models \bot$. By
 Remark~\ref{rem:consistency}, this amounts to~$1 = 0$ in~$A$.
@@
 
-\begin{theorem}(Grothendieck's generic freeness lemma, simple version)
+%\begin{proposition}[McCoy's theorem]
+%Let~$M \in A^{n \times m}$ be an injective matrix over a reduced ring~$A$.
+%Then~$(\Lambda^m M)$, the ideal of~$m$-minors of~$M$, is regular.
+%\end{proposition}
+%
+%\begin{proof}Let~$x \in A$ such that~$x \cdot (\Lambda^M) = (0)$ be given. We
+%will show~$x = 0$~in~$A$ by verifying~$x = 0$ in~$A^\sim$ under the sheaf
+%semantics.
+%
+%To this end, it is prudent to introduce a variant of the usual
+%negation~$\neg\varphi \equiv (\varphi \Rightarrow \bot)$, namely~$\negg\varphi
+%\equiv (\varphi \Rightarrow a = 0)$. Our claim is then~$\bott \defeq
+%\negg\top$. As usual, the double negation of the law of excluded middle is
+%available, that is we have
+%\[ \forall x\?A^\sim\_ \negg\negg\bigl((\exists y\?A\_ xy = 1) \vee \negg(\exists
+%y\?A\_ xy = 1)\bigr), \]
+%and as our goal is~$\bott$, we may use with law of excluded middle a finite
+%number of times.
+%
+%Repeatedly exploiting that any matrix element is invertible or
+%$\negg\text{invertible}$ and applying elementary row and column
+%transformations, we may assume that~$M$ is of the
+%form~$\left(\begin{smallmatrix}I&0\\0&N\end{smallmatrix}\right)$ with all
+%elements of~$N$ $\negg\text{invertible}$.
+%
+%By Remark~\ref{XXX},
+%\end{proof}
+% XXX
+
+\begin{proposition}[Grothendieck's generic freeness lemma, simple version]
+\label{thm:baby-grothendieck}
 Let~$A$ be a reduced ring. Let~$M$ be a finitely generated~$A$-module.
@@
 free~$A[f^{-1}]$-module, then~$1 = 0$ in~$A$.
-\end{theorem}
+\end{proposition}
 
@@
 free. Since~$A^\sim$ is a field, this follows from
-basic (intuitionistic) linear algebra.\footnote{XXX case distinction}\end{proof}
+basic (intuitionistic) linear algebra.\footnote{A standard result of linear
+algebra is: ``Any finitely generated vector space over a field is finite free
+(that is, has a finite basis).'' The proof proceeds by considering a given
+generating family and using the law of excluded middle to determine whether one
+of the generators can be expressed using the others or not. In the latter case,
+the family is linearly independent and hence a basis; in the former case, the
+proof removes the redundant generator and continues by induction.
+
+Intuitionistically, we only have the double negation of the law of excluded
+middle available; this is why intuitionistically we only have that finitely
+generated vector spaces are \notnot finite free. The field condition required
+by the proof is exactly the one satisfied by~$A^\sim$.}\end{proof}
+
+\begin{theorem}[generic freeness lemma, general version]
+Let~$A$ be a reduced ring. Let~$B$ be an~$A$-algebra of finite type. Let~$M$ be
+a finitely generated~$B$-module.
+If~$f = 0$ is the only element of~$A$ such that
+\begin{enumerate}
+\item $B[f^{-1}]$ and $M[f^{-1}]$ are free modules over $A[f^{-1}]$,
+\item $A[f^{-1}] \to B[f^{-1}]$ is of finite presentation and
+\item $M[f^{-1}]$ is finitely presented as a module over~$B[f^{-1}]$,
+\end{enumerate}
+then~$1 = 0$ in~$A$.
+\end{theorem}
+
+\begin{proof}One can check that the claim is the interpretation by the sheaf
+semantics of the following statement about the sheaves: It is \notnot the case that
+\begin{enumerate}
+  \item $B^\sim$ and $M^\sim$ are free modules over $A^\sim$ (by basic linear algebra),
+  \item $A^\sim \to B^\sim$ is of finite presentation, and
+  \item $M^\sim$ is finitely presented as a module over $B^\sim$,
+\end{enumerate}
+
+Item~(1) is by basic linear algebra as in the proof of
+Theorem~\ref{thm:baby-grothendieck}, but a bit more involved as we now need to
+deal with countable generating
+families~\cite[Theorem~11.16]{blechschmidt:phd}. Item~(2) is because finitely
+generated algebras over anonymously Noetherian rings are \notnot finitely
+presented (in a description of the form~$R[X_1,\ldots,X_n]/\aaa$, the
+ideal~$\aaa$ is \notnot finitely generated).
+Item~(3) is because finitely generated modules over anonymously Noetherian are
+\notnot finitely presented.
+\end{proof}
 
@@
 
+\subsection{Exercises}
+
+\begin{exercise}{Nontriviality of~$A^\sim$}%
+Let~$A$ be a ring. In Proposition~\ref{prop:immediate-consequences}, we gave an
+abstract proof that~$A^\sim$ is local and hence nontrivial in the sense that~$1
+\neq 0$ (even if~$A$ itself happens to be trivial). Give a new proof of this
+fact, by verifying directly~$1 \models \neg(1 =_{A^\sim} 0)$ using the
+semantics given in Table~\ref{table:algebraic-kripke-joyal}.
+\end{exercise}
+
+\begin{exercise}{A generalization of the field property}%
+Let~$A$ be a ring. We extend the sheaf semantics to infinitary first-order
+formulas by adding the following clauses to
+Table~\ref{table:algebraic-kripke-joyal}:
+\begin{center}\tablefont
+\begin{tabular}{@{}l@{\ \ }c@{\ \ }l@{}}
+  $f \models \bigwedge_{k \in K} \varphi_k$ &iff&
+    for all indices~$k \in K$, $f \models \varphi_k$ \\
+  $f \models \bigvee_{k \in K} \varphi_k$ &iff&
+    there exists a partition~$f^n = fg_1 + \cdots + fg_m$ such that, \\
+  &&\quad for each~$i$, there is an index~$k \in K$ such that~$fg_i \models \varphi_k$
+\end{tabular}
+\end{center}
+Theorem~\ref{thm:basic-properties-sheaf-semantics-algebraic}
+also apllies to this extension, and moreover, this extension in sound with
+respect to infinitary intuitionistic logic.
+\begin{enumerate}
+\item Prove that:
+\[ 1 \models \forall f\?A^\sim\_ \forall x\?A^\sim\_
+  \bigl((\exists z\?A^\sim\_ fz = 1) \Rightarrow x = 0\bigr) \Longrightarrow \bigvee_{n \in \NN} (f^n x = 0). \]
+\item Deduce that~$A^\sim$ is ``almost a field'' in the sense that any nonunit
+is nilpotent.
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}{Bridging modules and their induced sheaves}%
+Let~$M$ be a module over a ring~$A$.
+\begin{enumerate}
+\item Show that~$M$ is finitely generated iff~$M^\sim$ is.
+\item Show that~$M$ is finite locally free (that is, that there is a
+partition~$1 = f_1 + \cdots + f_n$ such that each module~$M[f_i^{-1}]$ is
+finite free over~$A[f_i^{-1}]$) iff~$M^\sim$ is finite free.
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}{Local injectivity and surjectivity}%
+Let~$M$ an~$N$ be modules over a ring~$A$. Let~$f : M \to N$ be a linear map.
+\begin{enumerate}
+\item Show that~$f$ is injective iff~$f^\sim : M^\sim \to N^\sim$ is injective.
+\item Show that~$f$ is surjective iff~$f^\sim : M^\sim \to N^\sim$ is surjective.
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}{The units of~$A^\sim$}%
+Let~$A$ be a ring. Let~$f \in A$. Let~$x \in A[f^{-1}]$. Show using the result
+of Exercise~\ref{ex:global-existence} that~$f \models \exists y\?A^\sim\_ xy
+= 1$ if and only if~$x$ is invertible in~$A[f^{-1}]$.
+\end{exercise}
+
+\begin{exercise}{Identifying the triviality of the spectrum}%
+Let~$A$ be a ring. Verify, without assuming~\BPIT, that~$\Spec(A)$ is
+isomorphic to the one-point locale iff any element of~$A$ is either
+invertible or nilpotent.
+\end{exercise}
+
+\begin{exercise}{Prüfer domains as valuation domains}%
+An \emph{integral domain} is a ring such that~$1 \neq 0$ and such that~$xy = 0$
+implies~$x = 0$ or~$y = 0$. A \emph{valuation domain} is an integral domain
+such that for any two elements, one divides the other. A \emph{Prüfer domain}
+is an integral domain such that any finitely generated ideal is locally a
+principal ideal.
+\begin{enumerate}
+\item Let~$A$ be a ring. Show that, if~$A$ is an integral domain then so
+is~$A^\sim$. Does the converse hold?
+\item Let~$A$ be a valuation domain. Show that any matrix over~$A$ can be put
+into diagonal form by elementary row and column operations.
+\item Let~$A$ be an integral domain. Show that~$A$ is a Prüfer domain if and
+only if~$A^\sim$ is a valuation domain.
+\item Let~$A$ be a Prüfer domain. Show that any matrix over~$A$ can locally be
+put into diagonal form by elementary row and column operations, by applying the
+result of part~(b) to~$A^\sim$.
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}{A basic version of Kaplansky's theorem}
+\begin{enumerate}
+\item Let~$A$ be a local ring. Let~$\aaa \subseteq A$ be a finitely generated
+ideal such that~$\aaa^2 = \aaa$. Show that~$\aaa = (0)$ or~$\aaa = (1)$.
+\item Let~$A$ be a local ring. Let~$M \in A^{n \times n}$ be an idempotent
+matrix. Verify that~$M$ is similar to a diagonal matrix with entries zero and
+one by applying part~(a) to the ideals of~$k$-minors of~$M$. Deduce that the
+cokernel of~$M$ is finite free.
+\item Let~$A$ be an arbitrary ring. Let~$M \in A^{n \times n}$ be an idempotent
+matrix. Show that the cokernel of~$M$ is finite locally free, by applying the
+result of part~(b) to~$A^\sim$.
+\end{enumerate}
+\end{exercise}
+
 \end{document}
@@
 
+XXX Cite http://www.mat.uc.pt/~picado/covariant/contents.pdf
+http://www.mat.uc.pt/~picado/publicat/Chapter2.pdf
+
 X \models phi

### 2020-04-02 paper-herrsching2019/main.tex
Generic models and more

@@
 \usepackage{ws-rv-van}
-\usepackage{mathtools,stmaryrd,xspace}
+\usepackage{mathtools,stmaryrd,xspace,bussproofs}
 
@@
 is empty. Locales of these kind are used as an important reduction step in the
-extension of Grothendieck's Galois theory by Joyal and
-Tierney~\cite[Section~V.3]{joyal-tierney:galois-theory}.
+extension of Grothendieck's Galois theory by André Joyal and
+Myles Tierney~\cite[Section~V.3]{joyal-tierney:galois-theory}.
 
@@
 ring~$A$ is local iff~$A$ is local.)
-Show that any local locale has a canonical point, the \emph{focal point}.
+Show that any local locale has a canonical point, its \emph{focal point}.
 \end{exercise}
@@
 {\vee}\ {\bigvee}\ {\exists}$ and the relation symbols (but no ${\Rightarrow}\
-{\forall}$). The symbol~``$\bigvee$'' refers to disjunctions of arbitrary
-set-indexed families of formulas.
+{\forall}$). The symbol~``$\bigvee$'' refers to disjunction of arbitrary
+set-indexed families of formulas. The theory \emph{proves} a sequent~$\sigma$ if and
+only if there exists a \emph{derivation} of~$\sigma$, where the set of
+derivations is inductively generated by the axioms and the rules displayed in
+Table~\ref{table:geometric-logic}.
 \end{definition}
+
+{\makeatletter
+\newcommand*{\shifttext}[1]{%
+  \settowidth{\@tempdima}{#1}%
+  \makebox[\@tempdima]{\hspace*{\@tempdima}#1}%
+}
+\makeatother
+\begin{table}[t]
+  \tbl{The rules of geometric logic}{\begin{minipage}{0.8\textwidth}
+    \centering
+    \medskip
+
+    \textnormal{structural rules} \\\smallskip
+    \phantom{a}\hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}\UnaryInfC{$\varphi \seq{\vec x} \varphi$}\DisplayProof\hfill
+    \AxiomC{$\varphi \seq{\vec x} \psi$}\AxiomC{$\psi \seq{\vec x}
+    \chi$}\BinaryInfC{$\varphi \seq{\vec x} \chi$}\DisplayProof\hfill
+    \AxiomC{$\varphi \seq{\vec x} \psi$}\UnaryInfC{$\varphi[\vec s/\vec x]
+    \seq{\vec y} \psi[\vec s/\vec x]$}\DisplayProof
+    \makebox[0pt]{\shifttext{\tiny ($\vec y$ not bound)}}  % in~$\varphi$ and~$\psi$)}}
+    \phantom{a}\hfill
+    \bigskip
+
+    \textnormal{rules for conjunction} \\\smallskip
+    \phantom{a}\hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}\UnaryInfC{$\varphi \seq{\vec x} \top$}\DisplayProof\hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}\UnaryInfC{$\varphi \wedge \psi \seq{\vec x} \varphi$}\DisplayProof\hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}\UnaryInfC{$\varphi \wedge \psi \seq{\vec x} \psi$}\DisplayProof\hfill
+    \AxiomC{$\varphi \seq{\vec x} \psi$}\AxiomC{$\varphi \seq{\vec x} \chi$}\BinaryInfC{$\varphi \seq{\vec x} \psi \wedge \chi$}\DisplayProof
+    \phantom{a}\hfill
+    \bigskip
+
+    \textnormal{rules for finitary disjunction} \\\smallskip
+    \phantom{a}\hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}\UnaryInfC{$\bot \seq{\vec x} \varphi$}\DisplayProof\hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}\UnaryInfC{$\varphi \seq{\vec x} \varphi \vee \psi$}\DisplayProof\hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}\UnaryInfC{$\psi \seq{\vec x} \varphi \vee \psi$}\DisplayProof\hfill
+    \AxiomC{$\varphi \seq{\vec x} \chi$}\AxiomC{$\psi \seq{\vec x} \chi$}\BinaryInfC{$\varphi \vee \psi \seq{\vec x} \chi$}\DisplayProof
+    \phantom{a}\hfill
+    \bigskip
+
+    \textnormal{rules for infinitary disjunction} \\\smallskip
+    \phantom{a}\hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}\UnaryInfC{$\varphi_{i_0} \seq{\vec x} \bigvee_{i \in I} \varphi_i$}\DisplayProof\hfill
+    \AxiomC{$\varphi_i \seq{\vec x} \chi \text{ for each~$i \in I$}$}\UnaryInfC{$\bigvee_{i \in I} \varphi_i \seq{\vec x} \chi$}\DisplayProof
+    \phantom{a}\hfill
+    \bigskip
+
+    \textnormal{double rule for existential quantification} \\\smallskip
+    \phantom{a}\hfill
+    \Axiom$\varphi \ \fCenter\seq{\vec x, y} \psi$
+    \doubleLine
+    \UnaryInf$\exists y\?Y\_\! \varphi\ \fCenter\seq{\vec x} \psi$
+    \DisplayProof
+    \makebox[0pt]{\shifttext{\tiny ($y$ not free in~$\psi$)}}
+    \hfill\phantom{a}
+    \bigskip
+
+    \textnormal{rules for equality} \\\smallskip
+    \vspace{-0.5em}
+    \phantom{a}\hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}
+    \UnaryInfC{$\top \seq{x} x = x$}
+    \DisplayProof
+    \hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}
+    \UnaryInfC{$(\vec x = \vec y) \wedge \varphi \seq{\vec z} \varphi[\vec y/\vec x]$}
+    \DisplayProof
+    \hfill\phantom{a} \\[0.5em]
+    (``$\vec x = \vec y\,$'' is an abbreviation for~``$x_1 = y_1 \wedge \cdots \wedge x_n =
+    y_n$''.)%
+  \end{minipage}}
+  \label{table:geometric-logic}
+\end{table}}
 
@@
 \end{align*}
+
+\paragraph{The geometric theory of automorphisms of a field.}\label{par:theory-automorphisms}
+Let~$L|k$ be a field
+extension. The classifying locale of automorphisms~$L \to L$ which fix~$k$ has
+one nullary relation symbol~``$\sigma_{xy}$'' for each pair~$\langle x,y
+\rangle \in L$ and the following axioms:
+\begin{align*}
+  \top &\vdash \bigvee_y \sigma_{xy} & \text{(for each~$x \in L$)} \\
+  \top &\vdash \bigvee_x \sigma_{xy} & \text{(for each~$y \in L$)} \\
+  \sigma_{xy} \wedge \sigma_{xy'} &\vdash \bigvee\{ \top \,|\, y = y' \} & \text{(for each~$x,y,y' \in L$)} \\
+  \sigma_{xy} \wedge \sigma_{x'y} &\vdash \bigvee\{ \top \,|\, x = x' \} & \text{(for each~$x,x',y \in L$)} \\
+  \top &\vdash \sigma_{00} \\
+  \top &\vdash \sigma_{11} \\
+  \sigma_{xy} \wedge \sigma_{x'y'} &\vdash \sigma_{x+x',y+y'} & \text{(for each~$x,x',y,y' \in L$)} \\
+  \sigma_{xy} \wedge \sigma_{x'y'} &\vdash \sigma_{xx',yy'} & \text{(for each~$x,x',y,y' \in L$)} \\
+  \top &\vdash \sigma_{xx} & \text{(for each~$x \in k$)}
+\end{align*}
+A model of this theory (can be reorganized to be) an element of the classical
+Galois group of~$L|k$.
 
@@
 \end{enumerate}
-These results explain why in classical mathematics, where all of these
-conditions are satisfied, we can blithely and fruitfully use the topological
+
+In this picture, the opens of a locale can be regarded as finite approximations
+to the ideal points which potentially belong to them. For instance, the
+open~$[\sigma_{\sqrt{2},-\sqrt{2}} \wedge \sigma_{\sqrt{3},\sqrt{3}}]$
+of the localic Galois group of~$\overline{\QQ}|\QQ$ can be regarded as an
+approximation to a certain automorphism of~$\overline{\QQ}$ about
+which~$\sqrt{2} \mapsto -\sqrt{2}$ and~$\sqrt{3} \mapsto \sqrt{3}$ is known. Over time, this
+approximation might be refined to a smaller open, one which contains more
+information about the purported automorphism, and eventually we might actually
+obtain an automorphism which is defined on all of~$\overline{\QQ}$. \emph{The opens
+make good constructive sense even if the points remain elusive.}
+
+The results mentioned above explain why we can, in classical mathematics, where all of the stated
+conditions are satisfied, blithely and fruitfully use the topological
 spectrum of a ring, the topological Galois group or the topological space of
@@
 
-%\begin{theorem}The localic real line is spatial if and only if the set~$[0,1]$ of
-%Dedekind real numbers is compact in the sense of Heine--Borel.
-%\end{theorem}
-%
-%\begin{proof}XXX\end{proof}
-
-%\begin{theorem}Let~$A$ be a ring and~$\Spec(A)$ its spectrum as defined above as the locale of
-%prime filters of~$A$. Then~$\Spec(A)$ is spatial if (the law of excluded middle and) the
-%Boolean Prime Ideal Theorem~\BPIT hold. More specifically, $\Spec(A)$ is
-%spatial if and only if for any ideal~$\aaa \subseteq A$ and any element~$x \in A$,
-%\[ x \in \sqrt{\aaa} \quad\textnormal{if}\quad
-%\textnormal{for all filters~$\fff \subseteq A$ with~$x \in \fff$, $\fff \between \aaa$}. \]
-%\end{theorem}
-%
-%The symbol~``$\between$'' denotes that the two sets have an element in common.
-%
-%\begin{proof}Assume that~$\Spec(A)$ is spatial. Let an ideal~$\aaa \subseteq A$
-%and an element~$x \in A$ be given such that for all filters~$\fff \subseteq A$
-%with~$x \in \fff$, $\fff \between \aaa$. Then for all filters~$\fff$, $\fff
-%\inplus \alpha_x \Rightarrow \fff \inplus \bigvee_{y\in \aaa} \alpha_y$.
-%Hence ...
-%Using the filter axioms, any open of~$\Spec(A)$ can be written
-%as~$[\bigvee_i \alpha_{x_i}]$.
-%\end{proof}
+\begin{remark}The case of the locale of surjections~$\NN \twoheadrightarrow
+\RR$ is slightly different in that there, even the strong existence axioms of
+classical mathematics do not suffice to materialize a point.
+\end{remark}
 
@@
 called \emph{sections of~$F$ over~$U$} and the maps~$(\cdot)|^U_V$
-are called \emph{restriction maps}.
+are called \emph{restriction maps}. Elements of~$F(\top)$ are called
+\emph{global sections}.
 
@@
 
+% XXX Introduction...
+
 Let~$X$ be a locale. A \emph{formula over an open~$U \in \O(X)$} is a
@@
 each sheaf~$F$, one constant symbol of sort~$F$ for each section~$s \in F(U)$,
-one function symbol~$f : F \to G$ for each morphism of sheaves, and so on.
+one function symbol~$f : F \to G$ for each morphism of sheaves, and so on. As
+usual, we define~$\neg\varphi$ to be an abbreviation of~$(\varphi \Rightarrow
+\bot)$.
 
@@
 main differences with the perhaps more familiar Kripke semantics are marked.
+The sheaf semantics is also called \emph{Kripke--Joyal semantics}.
 
@@
 
-\begin{definition}A \emph{sheaf model}~$M$ of a geometric theory~$\TT$ over a
-locale~$X$ consists of
+\begin{definition}\label{defn:sheaf-model}
+A \emph{sheaf model} of a geometric theory~$\TT$ over a locale~$X$ consists of
 \begin{enumerate}
@@
 \end{enumerate}
-such that~$M$ validates the axioms of~$\TT$, that is, such that the axioms
-of~$\TT$ are forced on~$\top \in \O(X)$.
+such that the axioms of~$\TT$ are validated under the sheaf semantics.
 \end{definition}
+
+\begin{example}The sheaf~$\C$ of continuous real-valued functions together with
+appropriate morphisms is a sheaf model of the theory of rings. Any sheaf can be
+regarded as a model of the theory of objects.\end{example}
+
+\begin{example}\label{ex:sets-as-sheaf-models}
+Any set-based model of a geometric theory~$\TT$ gives rise to a
+sheaf model over the one-point locale, by employing the constant sheaf
+construction of XXX. Hence sheaf models are strictly more general than
+set-based models.
+\end{example}
+
+In the special case of nullary function symbols or relations symbols, by
+``$\brak{S_1} \times \cdots \times \brak{S_n}$'' in
+Definition~\ref{defn:sheaf-model} we mean the empty product of sheaves, the
+\emph{terminal sheaf}~$1$ with~$1(U) = \{\star\}$ for every open~$U$ of~$X$.
+Morphisms~$1 \to F$ are in canonical bijection with global sections of~$F$.
+Subsheaves~$F \subseteq 1$ are in canonical bijection with the opens of~$X$
+(Exercise~\ref{ex:subsheaves-1}).
+
+With this in mind, we can rephrase Definition~\ref{defn:sheaf-model} in the
+special case of propositional theories in the following simpler form. A
+\emph{sheaf model} of a propositional geometric theory~$\TT$ over a locale~$X$
+consists of a family~$(U_\varphi)_\varphi$ of opens of~$X$, one for each
+nullary relation symbol~$\varphi$ of~$\TT$, in such a way that the axioms are
+satisfied. For instance, that an axiom~``$\alpha \wedge \beta \vdash \bigvee_i
+\gamma_i$'' is satisfied means that~$U_\alpha \wedge U_\beta \preceq \bigvee_i
+U_{\gamma_i}$.
+
+In other words, a model of a propositional geometric theory~$\TT$ over a
+locale~$X$ is the same as a frame homomorphism~$\O(L(\TT)) \to \O(X)$, hence a
+locale morphism~$X \to L(\TT)$.
+
+A geometric theory can well have no set-based models, but might still be
+consistent and allow for a diverse range of sheaf models on a host of different
+locales. In fact, any propositional geometric theory always has at least one,
+and very peculiar, sheaf model, namely its \emph{generic model}.
+
+\begin{definition}The \emph{generic model}~$U_\TT$ of a propositional geometric
+theory~$\TT$ is the sheaf model on~$L(\TT)$ given by~$U_\varphi \defeq
+[\varphi]$ for all nullary relation symbols~$\varphi$.\end{definition}
+
+Under the view of models as locale morphisms, the generic model~$U_\TT$
+corresponds to the identity morphism~$L(\TT) \to \L(\TT)$. If a~$\TT$-model~$M$ on
+a locale~$X$ is given by a locale morphism~$f : X \to L(\TT)$, we also say
+that~``$M$ arises from~$U_\TT$ by pullback along~$f$''.
+
+For instance, there is the \emph{generic Dedekind cut}, the generic model of
+the theory~$\TT$ of Dedekind cuts on the classifying locale of~$\TT$, the
+localic real line. By Example~\ref{ex:sets-as-sheaves}, any specific Dedekind
+real~$x$ (that is, any point~$x$ of~$L(\TT)$) gives rise to a model of~$\TT$ on
+the one-point locale~$\pt$, corresponding to the locale morphism~$x : \pt \to
+L(\TT)$.
+
+The story of generic models is at the same time tautologous and deep, and these
+short paragraphs do not do justice to this intriguing aspect of the interplay
+between topology (in the form of locales) and logic (in the form of
+propositional geometric theories). We strongly
+recommend the leisurely account by Steve Vickers~\ref{XXX} for a fuller
+picture.
 
@@
 Let~$L|k$ be a field extension. The \emph{localic Galois group}~$\Gal(L|k)$
-of~$L|k$ is the classifying locale of the theory which has one nullary relation
-symbol~``$\sigma_{xy}$'' for each pair~$\langle x,y \rangle \in L$ and the
-following axioms:
-\begin{align*}
-  \top &\vdash \bigvee_y \sigma_{xy} & \text{(for each~$x \in L$)} \\
-  \top &\vdash \bigvee_x \sigma_{xy} & \text{(for each~$y \in L$)} \\
-  \sigma_{xy} \wedge \sigma_{xy'} &\vdash \bigvee\{ \top \,|\, y = y' \} & \text{(for each~$x,y,y' \in L$)} \\
-  \sigma_{xy} \wedge \sigma_{x'y} &\vdash \bigvee\{ \top \,|\, x = x' \} & \text{(for each~$x,x',y' \in L$)} \\
-  \top &\vdash \sigma_{00} \\
-  \top &\vdash \sigma_{11} \\
-  \sigma_{xy} \wedge \sigma_{x'y'} &\vdash \sigma_{x+x',y+y'} & \text{(for each~$x,x',y,y' \in L$)} \\
-  \sigma_{xy} \wedge \sigma_{x'y'} &\vdash \sigma_{xx',yy'} & \text{(for each~$x,x',y,y' \in L$)} \\
-  \top &\vdash \sigma_{xx} & \text{(for each~$x \in k$)}
-\end{align*}
+is the classifying locale of the theory of automorphisms~$L \to L$ which
+fix~$k$, as described on page~\pageref{par:theory-automorphisms}.
 \begin{enumerate}
@@
 with a \emph{sheaf of rings}.
-\item We then maintain the illusion that we are working with a plain old ring
-by employing the sheaf semantics.
+\item We then maintain the convenient illusion that we are working with a plain old ring
+instead of a sheaf by employing the sheaf semantics.
 \end{enumerate}
@@
 \emph{spectrum of~$A$} from algebraic geometry, and the sheaf~$A^\sim$ is the
-sheaf of regular functions. We will give details on its construction in
+sheaf of regular functions, also denoted~``$\O_{\Spec(A)}$''. We will give details on its construction in
 Section~\ref{sect:origins}, but first want to reflect on the displayed
@@
 
+Not only does the ring~$A$ have a mirror image as a sheaf, also
+any~$A$-module~$M$ does. Its mirror image is the \emph{quasicoherent
+sheaf}~$M^\sim$ from algebraic geometry. As with~$A$ and~$A^\sim$, there is a
+close connection between~$M$ and~$M^\sim$; for instance, the module~$M$ is
+finitely generated over~$A$ iff~$M^\sim$ is finitely generated over~$A^\sim$
+(Exercise~\ref{ex:bridging-modules}).
+
 \paragraph{The field property.}
@@
 requires an actual computation for its verification and was a surprising
-discovery by Mulvey in the 1970s.
-
-Tierney commented around that time that the field property ``is surely important,
+discovery by Christopher Mulvey in the 1970s.
+
+Myles Tierney commented around that time that the field property ``is surely important,
 though its precise significance is still somewhat obscure---as is the case with
@@
 particularly perspicious account in the general setting of partial Horn
-theories is due to Palmgren and Vickers~\cite{palmgren-vickers:partial-horn}.
+theories is due to Erik Palmgren and Steve Vickers~\cite{palmgren-vickers:partial-horn}.
 
@@
 \begin{exercise}{Bridging modules and their induced sheaves}%
+\label{ex:bridging-modules}%
 Let~$M$ be a module over a ring~$A$.
@@
 finite free over~$A[f_i^{-1}]$) iff~$M^\sim$ is finite free.
+\item Show that~$M$ is finitely presented iff~$M^\sim$ is.
 \end{enumerate}
@@
 invertible or nilpotent.
+\end{exercise}
+
+\begin{exercise}{Spatiality of the spectrum}%
+Let~$A$ be a ring. Let~$\Spec(A)$ be its spectrum.
+\begin{enumerate}
+\item Verify that~$\Spec(A)$ is spatial if and only if for any element~$x \in
+A$ and any ideal~$\aaa \subseteq A$,
+\[ x \in \sqrt{\aaa} \quad\textnormal{if}\quad
+\text{for all prime filters~$\fff \subseteq A$ with~$x \in \fff$, $\fff \between \aaa$}, \]
+where the symbol~``$\between$'' denotes that the two sets have an element in common.
+\item Assuming the law of excluded middle, show that~$\Spec(A)$ is spatial if
+and only if for any ideal~$\aaa \subseteq A$ the familiar identity
+\[ \sqrt{\aaa} = \bigcap_{\ppp \supseteq \aaa} \ppp \]
+holds.
+\item Conclude that the spectrum of any ring is spatial iff~\BPIT holds.
+\end{enumerate}
 \end{exercise}
@@
 X \models phi
+
+FUTURE:
+* Why the name "geometric"?

### 2020-04-02 paper-herrsching2019/main.tex
Rules for first-order logic

@@
     \doubleLine
-    \UnaryInf$\exists y\?Y\_\! \varphi\ \fCenter\seq{\vec x} \psi$
+    \UnaryInf$\exists y\?Y\_ \varphi\ \fCenter\seq{\vec x} \psi$
     \DisplayProof
@@
 \begin{remark}The case of the locale of surjections~$\NN \twoheadrightarrow
-\RR$ is slightly different in that there, even the strong existence axioms of
-classical mathematics do not suffice to materialize a point.
+\RR$ is particular in that there, even the strong existence axioms of
+classical mathematics do not suffice to materialize a point. In fact, already
+the law of excluded middle or the countable axiom of choice preclude the
+existence of points.
 \end{remark}
@@
 The principal properties of the sheaf semantics is that it is monotone, local
-and sound with respect to intuitionistic first-order logic:
+and sound with respect to intuitionistic first-order logic (which we fix for
+concreteness in Table~\ref{table:first-order-logic}):
+
+{\makeatletter
+\newcommand*{\shifttext}[1]{%
+  \settowidth{\@tempdima}{#1}%
+  \makebox[\@tempdima]{\hspace*{\@tempdima}#1}%
+}
+\makeatother
+\begin{table}[t]
+  \tbl{The rules of intuitionistic first-order logic}{\begin{minipage}{0.8\textwidth}
+    \centering
+    \medskip
+
+    (structural rules and rules for~${=}\ {\top}\ {\wedge}\ {\bot}\ {\vee}\
+    {\exists}$ as in Table~\ref{table:geometric-logic})
+    \medskip
+
+    \textnormal{double rule for implication} \\\smallskip
+    \phantom{a}\hfill
+    \Axiom$\varphi \wedge \psi\ \fCenter\seq{\vec x} \chi$
+    \doubleLine
+    \UnaryInf$\varphi\ \fCenter\seq{\vec x} \psi \Rightarrow \chi$
+    \DisplayProof
+    \phantom{a}\hfill
+    \bigskip
+
+    \textnormal{double rule for universal quantification} \\\smallskip
+    \phantom{a}\hfill
+    \Axiom$\varphi \ \fCenter\seq{\vec x, y} \psi$
+    \doubleLine
+    \UnaryInf$\varphi\ \fCenter\seq{\vec x} \forall y\?Y\_ \psi$
+    \DisplayProof
+    \makebox[0pt]{\shifttext{\tiny ($y$ not free in~$\varphi$)}}
+    \hfill\phantom{a}
+    \bigskip
+
+  \end{minipage}}
+  \label{table:first-order-logic}
+\end{table}}
 
@@
 
-XXX include rules for geometric logic?
-
 XXX Cite http://www.mat.uc.pt/~picado/covariant/contents.pdf

### 2020-04-02 paper-herrsching2019/main.tex
Morphisms by mapping points

@@
 relevant. Classically, a prime filter of a ring is simply the complement of a
-prime ideal; constructively, it is prudent to axiomatize this notion directly.
+prime ideal; constructively, it is prudent to turn prime ideals upside down and axiomatize this notion directly.
 Hence the theory of prime filters of~$A$ has no sorts, no function symbols, one
@@
 
-\begin{remark}The traditional way of constructing the real line or the Zariski
+\begin{remark}\label{rem:traditional-spaces}
+The traditional way of constructing the real line or the Zariski
 spectrum as a topological space proceeds in three steps: (1) Write down the
@@
 propositional geometric theories). We strongly
-recommend the leisurely account by Steve Vickers~\ref{XXX} for a fuller
+recommend the accessible account by Steve Vickers~\ref{XXX} for a fuller
 picture.
+
+\begin{remark}The traditional way of constructing a map of topological spaces~$f : Y \to
+Y'$ proceeds in three steps: (1) Write down a map of the underlying sets. (2)
+Identify the preimages~$f^{-1}[U]$. (3) Verify that these preimages are open.
+
+Continuing Remark~\ref{rem:traditional-spaces}, locales provide more
+efficient means for constructing morphisms: To write down a morphism of
+locales~$X \to X'$, where~$X'$ is given as the classifying locale of some
+theory~$\TT$, we only need to specify a~$\TT$-model on~$X$; no separate
+verification of continuity is required.
+
+For instance, if~$f : A \to B$ is a ring homomorphism, we obtain a locale
+morphism~$\Spec(B) \to \Spec(A)$ by specifying the model~$([D(f(x))])_{x \in
+A}$ of the theory of prime filters of~$A$ on~$\Spec(B)$. On the level of
+points, this morphism induces the expected map~$\fff \mapsto f^{-1}[\fff]$.
+\end{remark}
 

### 2020-04-02 paper-herrsching2019/main.tex
Semantics of relation symbols

@@
     $U \models s =_F t$ &\Ll& $s = t$ when evaluated as elements of~$F(U)$ \\
+    $U \models R(s_1,\ldots,s_n)$ &\Ll& $\langle s_1,\ldots,s_n \rangle \in R(U)$ \\
     $U \models \top$ &\Ll& true \\

### 2020-04-02 paper-herrsching2019/main.tex
Lots of references (not enough yet)

@@
 
+% XXX \cite{belanger-marquis:pointless} and Johnstone, vickers:locales-toposes
+
 The notion of \emph{space} is fundamental to large parts of mathematics. This notion
@@
 view: While the localic counterparts of the six pieces do not have any points
-in common, the locale-theoretic pairwise intersections are still nontrivial.
-Hence one would not expect the rearrangement of these sublocales to have the
-same volume as the original solid ball. [XXX: give reference]
+in common, the locale-theoretic pairwise intersections are still nontrivial~\cite{simpson:measure}.
+Hence there is no paradox, as one would not expect an rearrangement of
+overlapping sublocales to preserve volume.
 
@@
 By relinquishing points, locales provide a more flexible notion of space.
-Section~\ref{sect:examples-no-points} illustrates[XXX:word] this observation
+Section~\ref{sect:examples-no-points} substantiates this observation
 with several examples of nontrivial locales without any points. Constructive
@@
 subscribes to the axiom of choice or similar non-constructive principles. In
-these situations, the pointfree approach allowed by locales can help to give
-constructive versions of classical results.
+these situations, the pointfree approach allowed by locales help to give
+constructive versions of classical results. We will further discuss this theme
+in Section~\ref{sect:presenting-frames}.
 
@@
 tree provides a computable open covering of~$[0,1]$ with no computable finite
-subcovering [XXX: reference].
+subcovering (a self-contained exposition of this phenomenon is
+Ref.~\refcite{bauer:kleene-tree}).
 
 In contrast, the localic version of the unit interval is always compact. The
-proof is by an explicit computation with its basic opens. [XXX: reference]
+proof is by an explicit computation with its basic opens, exploiting the inductive
+nature of derivations~\cite{cederquist-negri:heine-borel}.
 
@@
 locales is again compact, can be verified without any nonconstructive
-principles. [XXX: reference]
+principles~\cite{vickers:tychonoff}.
 
@@
 However, this failure is not for fundamental Galois-theoretic reasons, but because of an
-unfortunate choice in the definitions. There is a notion of a localic group (a
+unfortunate choice in the definitions. There is a notion of a localic group~\cite{wraith:localic-groups} (a
 locale~$G$ equipped with continuous maps~$G \times G \xrightarrow{\circ} G$, $G
@@
 fundamental theorem can be reformulated to refer to this localic Galois group,
-and the proof of this reformulation is entirely constructive.
-[XXX: references]
+and the proof of this reformulation is entirely constructive~\cite{wraith:galois-topos}.
 
@@
 else settle for a discrete version of the fundamental theorem: For finite field
-extensions, finite intermediate extensions correspond to finite subgroups.
-[XXX: reference]
+extensions, finite intermediate extensions correspond to finite
+subgroups~\cite[Theorem~8.8]{mines-richman-ruitenburg:constructive-algebra}.
 Classically, intermediate extensions of finite extensions and subgroups of
@@
 However, if the intermediate extension~$E \defeq \{ x \in \QQ(\sqrt{2}) \,|\, x
-\in \QQ \vee \varphi \}$ is finite, then~$\varphi \vee \neg\varphi$.}
+\in \QQ \vee \varphi \}$ is finite, then~$\varphi \vee \neg\varphi$. Using the
+sheaf semantics reviewed in Section~\ref{sect:sheaf-semantics}, this
+intermediate extension can be turned into a topological
+counterexample~\cite[page~65]{wraith:localic-groups}.}
 \end{remark}
@@
 \in \fff$ for some index~$i$; and conversely, any such completely prime filter
-gives rise to a point of~$X$.
-% XXX exercise
+gives rise to a point of~$X$ (Exercise~\ref{ex:points-as-filters}).
 
@@
 spatial. The localic real line is spatial in classical mathematics and can
-fail to be spatial in constructive mathematics (Theorem~\ref{thm:spatiality-reals}).
+fail to be spatial in constructive mathematics (Exercise~\ref{ex:spatiality-reals}).
 
@@
 \begin{exercise}{Points as complete prime filters}%
+\label{ex:points-as-filters}%
 Verify that the points of a locale~$X$ are in canonical one-to-one correspondence
@@
 
-When first learning about presheaves, one can get boggled down[XXX:wording] by the vast
+When first learning about presheaves, one may be intimidated by the vast
 amount of data encoded in a single presheaf (one set for each open). The
@@
 \emph{site} so that both locales and Kripke frames induce sites and that
-presheaves on sites can be defined; however, this shall not be pursued here.
-[XXX: reference]
+presheaves on sites can be defined; however, this shall not be pursued here,
+a textbook reference is Ref.~\refcite{moerdijk-maclane:sheaves-logic}.
 
@@
 \[ \C(U) \longrightarrow \C(V),\ s \longmapsto s|_V. \]
+
+Any set~$M$ gives rise to a presheaf~$\underline{M}^\mathrm{pre}$, by
+setting~$\underline{M}^\mathrm{pre}(U) = M$ for every open~$U$ and employing
+the identity map as restriction maps. In this way, the universe of sets embeds
+into the universe of presheaves.
 
@@
 to~$g$ on~$(0,1)$.
+
+There is a general construction called \emph{sheafification} which turns
+presheaves into sheaves. Applied to~$\C_\text{c}$, this process yields the
+sheaf of \emph{locally} constant real-valued functions. Similarly, the
+sheafification of the presheaf of bounded functions is the sheaf of locally
+bounded functions, and similarly for other properties of functions.
+
+The presheaf~$\underline{M}^\mathrm{pre}$ associated to a set~$M$ is almost
+never a presheaf. Its sheafification~$\underline{M}$ is the \emph{constant sheaf associated
+to~$M$}\label{page:constant-sheaf}. On an open~$U$, its sections are the locale
+morphisms~$U \to M$, where~$M$ is regarded as the induced locale associated to
+the discrete topological space~$M$ and~$U$ is regarded as a sublocale of~$X$.
+By the constant sheaf construction, the universe of sets maps to the universe
+of sheaves on~$X$. We will not require much more details on this construction.
+
+\begin{definition}A \emph{subsheaf} of a sheaf~$F$ on a locale~$X$ is a
+sheaf~$F'$ with~$F'(U) \subseteq F(U)$ for all opens~$U$ and the inherited
+restriction maps.\end{definition}
 
@@
 sheaf model over the one-point locale, by employing the constant sheaf
-construction of XXX. Hence sheaf models are strictly more general than
+construction reviewed on page~\ref{page:constant-sheaf}. Hence sheaf models are strictly more general than
 set-based models.
@@
 Under the view of models as locale morphisms, the generic model~$U_\TT$
-corresponds to the identity morphism~$L(\TT) \to \L(\TT)$. If a~$\TT$-model~$M$ on
+corresponds to the identity morphism~$L(\TT) \to L(\TT)$. If a~$\TT$-model~$M$ on
 a locale~$X$ is given by a locale morphism~$f : X \to L(\TT)$, we also say
@@
 the theory~$\TT$ of Dedekind cuts on the classifying locale of~$\TT$, the
-localic real line. By Example~\ref{ex:sets-as-sheaves}, any specific Dedekind
+localic real line. By Example~\ref{ex:sets-as-sheaf-models}, any specific Dedekind
 real~$x$ (that is, any point~$x$ of~$L(\TT)$) gives rise to a model of~$\TT$ on
@@
 propositional geometric theories). We strongly
-recommend the accessible account by Steve Vickers~\ref{XXX} for a fuller
+recommend the accessible account by Steve Vickers~\ref{vickers:continuity} for a fuller
 picture.
@@
 
-\begin{exercise}{More on exotic locales}
+\begin{exercise}{More on exotic locales}%
+\label{ex:exotic-locales}%
 \begin{enumerate}
@@
 {\scriptsize\emph{Note.} In Russian constructivism, as for instance reigning in
-the ef{}fective topos~\cite{XXX}, the Baire space~$\NN^\NN$ is subcountable.
+the ef{}fective topos~\cite{hyland:effective-topos,phoa:effective,bauer:c2c}, the Baire space~$\NN^\NN$ is subcountable.
 However, in this setting any function~$\NN \to \NN$ is computable.
@@
 the localic real line fails to be
-spatial~\cite[Example~D4.7.13]{johnstohne:elephant}.\par}
+spatial~\cite[Example~D4.7.13]{johnstone:elephant}.\par}
 \item An open~$U$ of a locale~$X$ is \emph{complemented} iff there is an
@@
 
-\begin{exercise}{Products of locales}\label{ex:products-of-locales}%
+\begin{exercise}{Product of locales}\label{ex:products-of-locales}%
 Let~$X$ and~$X'$ be locales, classifying some propositional geometric
@@
 
-\begin{exercise}{Coproducts of locales, revisited}%
+\begin{exercise}{Coproduct of locales, revisited}%
 Let~$X$ and~$X'$ be locales, classifying some propositional geometric
@@
 \begin{exercise}{Spatiality of the localic reals}%
+\label{ex:spatiality-reals}%
 \newcommand{\fa}{\ensuremath{(\!\cdot}}%
@@
 \end{enumerate}
+{\scriptsize\emph{Note.} Exercise~\ref{ex:exotic-locales} lists settings in
+which the assumption in part~(e) is not satisfied.\par}
 \end{exercise}
@@
 \begin{exercise}{Coherent locales and coherent theories}%
+\label{ex:coherent}%
 An open~$U$ of a locale is \emph{compact} iff, whenever~$U
@@
 without any set-based models are inconsistent.
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}{Subsheaves of the terminal sheaf}%
+\label{ex:subsheaves-1}%
+Let~$X$ be a locale.
+\begin{enumerate}
+\item Show that any subsheaf~$F \subseteq 1$ of the terminal sheaf on~$X$ gives rise
+to an open of~$X$ by gluing all its sections.
+\item Show that the construction of part~(a) yields a canonical bijection of
+the subsheaves of~$1$ and the opens of~$X$.
 \end{enumerate}
@@
 example demonstrates that the usual formulation of the fundamental theorem of
-algebra does not admit a constructive proof. However, there are replacements
-available~\cite{XXX}.\par}
+algebra does not admit a constructive proof. (However, there are replacements
+available~\cite{richman:fta}, and the fundamental theorem does apply to the
+Cauchy complex numbers~\cite{ruitenburg:roots} and under weak choice
+principles~\cite{bridges-richman-schuster:wcc}).\par}
 \end{exercise}
@@
 of choice; however, the statement is actually equivalent to the weaker Boolean
-Prime Ideal Theorem [XXX: reference].
+Prime Ideal Theorem~\cite{scott:bpit,banaschewski-harting:lattice-aspects}.
 
@@
 
-The completeness theorem for coherent logic~[XXX: reference (Elephant)] even
+The completeness theorem for coherent logic~\cite[Corollary~D1.5.10]{johnstone:elephant} even
 gives an a priori reason why (for a given matrix size) a finitary proof has to
@@
 Lombardi and Quitté on constructive commutative
-algebra~\cite{lombardi-quitte:constructive-commutative-algebra}.)
+algebra~\cite{lombardi-quitte:constructive-algebra}.)
 
@@
 element~$x \in A$ which is zero in all these stalks~$A_\ppp$ is not necessarily
-zero in~$A$. Also, an~$A$-module~$M$ which is finitely generated and whose
-stalks~$M_\ppp$ at minimal prime ideals are free is ``free on a dense open'',
-but this is false without the finiteness hypothesis. The analogous statement
-for~$M^\sim$ and~$A^\sim$ holds in full generality (XXX reviewed below).
+zero in~$A$. Also, properties of the stalks tend to only spread to
+dense opens of the spectrum, if they do so at all, while properties of~$A^\sim$
+are in close connection with properties of~$A$. For instance, that all
+stalks~$M_\ppp$ of an~$A$-module~$M$ at minimal (or, for that matter, all) prime ideals are finitely
+generated does not imply that~$M$ is finitely generated, while finite
+generation of~$M^\sim$ does imply finite generation of~$M$.
 \item The quotients~$A/\mmm$: These are fields, but an element~$x \in A$ which is
@@
 XXX not shared by stalks; ``true constructivists would be offended''
+
+Noetherian conditions are notoriously tricky in constructive mathematics; XXX
+There are several proposals for constructively sensible definition of
+Noetherian rings in the literature on constructive algebra, each with unique
+advantages and
+disadvantages~\cite{richman:noetherian,mines-richman-ruitenburg:constructive-algebra,perdry:noetherian,perdry:lazy,perdry-schuster:noetherian,tennenbaum:hilbert}.
+Insightful comments on why this is so can be found in the introduction and more
+specifically on page~27 of the textbook by Henri Lombardi and Claude
+Quitté~\cite{lombardi-quitte:constructive-algebra}.
+
+A ring is \emph{anonymously Noetherian} if and only if any of its ideals is
+\notnot finitely generated.
+
+%\begin{ex}There is an intuitionistic proof that the ring~$\ZZ$ is anonymously
+%Noetherian: Let~$\aaa \subseteq \ZZ$ be any ideal. Under the assumption that
+%either there exists a nonzero element in~$\aaa$ or not, the ideal~$\aaa$ is
+%\notnot finitely generated, even \notnot principal: For in the first case, a
+%minimal element~$d$ of~$\aaa \cap \NN^+$ (which \notnot exists) witnesses~$\aaa
+%= (d)$. In the second case the ideal~$\aaa$ is the zero ideal. Since the
+%assumption is \notnot satisfied, the ideal~$\aaa$ is \notnot \notnot finitely
+%generated, so \notnot finitely generated. (We remark on this proof scheme on
+%page~\pageref{proof-scheme-boxed-statements}.) \end{ex}
+%
+%It appears that this Noetherian condition has not been studied in the
+%literature on constructive algebra. Indeed, from the philosophical and the
+%computational point of view on constructive mathematics, the notion of an anonymously
+%Noetherian ring is not very useful: From an intuitionistic proof that a given
+%ideal is finitely generated one can mechanically extract explicit generators,
+%thereby satisfying computational or philosophical demands (``the generators are
+%really there''). In contrast, an intuitionistic proof that a given ideal is
+%\notnot finitely generated doesn't contain computational content in general.
+%
+%Along the same lines, one could dismiss the fact (proved below) that Hilbert's
+%basis theorem, stating that~$A[X]$ is Noetherian if~$A$ is, holds for the
+%notion of anonymously Noetherian rings. In fact, one could feel mocked by this
+%version of Hilbert's basis theorem: It promises that any ideal of~$A[X]$
+%``has'' finitely many generators in some Platonic sense (without providing any
+%clue on how one might go on to find the generators -- they remain \emph{anonymous}), provided
+%that any ideal of~$A$ ``has'' finitely many generators in the same
+%sense.\footnote{We borrowed the term ``anonymous'' from type theory, where it
+%is used with a similar meaning (see for
+%instance~\cite{kraus-escardo-coquand-altenkirch:anonymous}). However, there is
+%a subtle difference: Unique and anonymous-in-the-sense-of-type-theory existence
+%implies existence, but \notnot existence does not.}
 
@@
 However, this procedure cannot be carried out for any pair of theories. For
-instance, there are no \emph{free fields}. The free field on a set~$X$ would be
-a field~$K$ together with a map~$X \to K$ such that any map~$X \to L$ factors
-over the given map~$X \to K$ by a unique field homomorphism~$K \to L$.
-% XXX Exercise: prove this!
+instance, there are no \emph{free fields}. The free field on a set~$M$ would be
+a field~$K$ together with a map~$M \to K$ such that any map~$M \to L$ into a field~$L$ factors
+over the given map~$M \to K$ by a unique field homomorphism~$K \to L$.
 
@@
 that~$\Spec(A)$ is compact. However, the stronger result that~$\Spec(A)$ is
-even a \emph{coherent locale} can also be deduced from a general theorem (XXX: ref to
-exercise) since the theory of prime filters is a coherent theory.\end{remark}
+even a \emph{coherent locale} can also be deduced from a general theorem
+since the theory of prime filters is a coherent theory
+(Exercise~\ref{ex:coherent}).\end{remark}
 
@@
 the resulting formulation will even make sense in a metatheories without a
-powerset operation such as~\textsc{czf} or arithmetic universes~\cite{XXX}. In
+powerset operation such as~\textsc{czf}~\cite{crosilla:predicativity,aczel-rathjen:cst} or arithmetic
+universes~\cite{maietti:au,vickers:sketches}. In
 fact, it is now a purely syntactic translation procedure which can be carried
@@
 Let~$A$ be a ring. Let~$\varphi$ be a formula over~$f \in A$.
-Then~$[D(f)] \models \varphi$ in the sense of Definition~XXX if and only
+Then~$[D(f)] \models \varphi$ in the sense of Table~\ref{table:sheaf-semantics} if and only
 if~$f \models \varphi$ in the sense of Table~\ref{table:algebraic-kripke-joyal}.
@@
 
+\begin{exercise}{Free fields}
+\begin{enumerate}
+\item A \emph{free field} over a set~$M$ would be
+a field~$K$ together with a map~$M \to K$ such that any map~$M \to L$ into a
+field~$L$ factors
+over the given map~$M \to K$ by a unique field homomorphism~$K \to L$. Prove
+that free fields exist for no set~$M$.
+\item Let~$k$ be a field. Devise a hypothetical notion of
+\emph{free~$k$-fields} (where a~$k$-field is a field together with a
+homomorphism from~$k$) and show that there are no free~$k$-fields.
+\end{enumerate}
+\end{exercise}
+
 \begin{exercise}{Nontriviality of~$A^\sim$}%
@@
 
+\bibliographystyle{ws-rv-van}
+\bibliography{generalized-spaces}
+
 \end{document}
-
-XXX cite https://apcz.umk.pl/czasopisma/index.php/LLP/article/viewFile/LLP.2013.009/732
 

### 2020-04-04 paper-herrsching2019/main.tex
Gelfand duality and several smaller items

@@
 \newcommand{\Spec}{\mathrm{Spec}}
+\newcommand{\Max}{\mathrm{Max}}
 \newcommand{\Gal}{\mathrm{Gal}}
 \newcommand{\Rad}{\mathrm{Rad}}
+\newcommand{\Idl}{\mathrm{Idl}}
 \newcommand{\lra}{\longrightarrow}
@@
 
-% XXX \cite{belanger-marquis:pointless} and Johnstone, vickers:locales-toposes
+% XXX \cite{belanger-marquis:pointless} and Johnstone, vickers:locales-toposes,
+% picado-pultr:frames-and-locales
 
@@
 
+% NOTE TO EDITORS AND LAYOUTERS: It is on purpose that these two definitions
+% are not set using the usual theorem environment. This is because we are only
+% citing these definitions in order to reflect on them. We are not actually
+% introducing these two notions here for further use.
 \begin{quote}
@@
 subscribes to the axiom of choice or similar non-constructive principles. In
-these situations, the pointfree approach allowed by locales help to give
+these situations, the pointfree approach allowed by locales helps to give
 constructive versions of classical results. We will further discuss this theme
@@
 tree provides a computable open covering of~$[0,1]$ with no computable finite
-subcovering (a self-contained exposition of this phenomenon is
+subcovering (a self-contained exposition of this phenomenon is contained in
 Ref.~\refcite{bauer:kleene-tree}).
@@
 
-\paragraph{Gelfand duality} XXX
+\paragraph{Gelfand duality.} The celebrated Gelfand correspondence states that
+the mapping
+\[ X \longmapsto \Hom(X, \CC) = \{ f : X \to \CC \,|\, \text{$f$ is continuous} \} \]
+defines a contravariant equivalence between compact Hausdorff spaces and
+commutative unital~$C^\star$-algebras. In the other direction,
+a~$C^\star$-algebra~$A$ is mapped to its \emph{maximal spectrum}~$\Max(A)$, the
+topological space of multiplicative linear functionals~$A \to \CC$.
+
+Some parts of the proof require the axiom of choice, (only) for fabricating
+appropriate elements of the maximal spectrum. But as with the fundamental
+theorem of infinite Galois theory, the Gelfand correspondence has a
+constructive core, and a pointfree reformulation unearths this constructive
+content: There is a contravariant equivalence between compact completely
+regular locales and commutative unital~$C^\star$-algebras.
+% XXX Barr
 
@@
 \[ \begin{array}{@{}ccc@{}}
-\text{arbitrary joins}
-&\text{and}&
-\text{finite meets}, \\
+\text{arbitrary joins (suprema)}
+&\quad\text{and}\quad&
+\text{finite meets (infima)}, \\
 \bigvee && \wedge
@@
 version of the reals of Section~\ref{sect:examples-no-points} are best
-constructed as \emph{classifying locales}, a notion to be introduced below.
+constructed as \emph{classifying locales}, a notion to be introduced in
+Section~\ref{sect:presenting-frames}.
+\item The radical ideals of any ring~$A$ form a frame and hence give rise to a
+locale. We will identify this locale in Theorem~\ref{thm:workhorse} to be the
+\emph{spectrum} of~$A$ as studied in algebraic geometry.
 \end{enumerate}
@@
 
-\begin{remark}The set of points of a locale~$X$ can be made into a topological space, giving
+\begin{remark}\label{rem:definition-sober}
+The set of points of a locale~$X$ can be made into a topological space, giving
 rise to a functor~$\Pt : \Loc \to \Top$. This functor is right adjoint
@@
 instance, the sobrification of any inhabited indiscrete space is the one-point
-space.\end{remark}
+space.
+
+The adjunction~$L \dashv \Pt$ restricts to an equivalence between sober
+topological spaces and spatial locales. Assuming the law of excluded middle,
+both Hausdorff spaces and the schemes from algebraic geometry are sober
+(Exercise~\ref{ex:hausdorff-sober}), hence the spaces of a wide
+range of the mathematical landscape can be studied as locales.\end{remark}
 
@@
 have a canonical map~$\Hom_\Top(Y,Y') \to \Hom_\Loc(L(Y),L(Y'))$. If~$Y'$ is
-sober, then this map is a bijection. In particular, the set-theoretic points
+sober, then this map is a bijection. In particular, in this case the set-theoretic points
 of~$Y'$ and the locale-theoretic points of~$L(Y')$ are in canonical
@@
 Verify the statements in Remark~\ref{rem:sober-embed}.
+\end{exercise}
+
+\begin{exercise}{Hausdorff spaces are sober}%
+\label{ex:hausdorff-sober}%
+A \emph{meet-irreducible open subset} of a topological space~$Y$ is an
+open subset~$W \subseteq Y$ such that~$(U \cap V \subseteq W) \Rightarrow (U
+\subseteq W) \vee (V \subseteq W)$ for all open subsets~$U,V \subseteq Y$ and
+such that~$Y \not\subseteq W$.
+\begin{enumerate}
+\item Assuming the law of excluded middle, show that the completely prime
+filters of a topological space are in canonical one-to-one correspondence with
+the meet-irreducible open subsets.
+
+{\scriptsize\emph{Hint.} Given a completely prime filter~$\fff \subseteq
+\O(Y)$, the set~$\bigcup\{ U \in \O(Y) \,|\, U \not\in \fff \}$ is a
+meet-irreducible open subset.\par}
+\item Assuming the law of excluded middle, show that a topological space is
+sober in the sense of Remark~\ref{rem:definition-sober} if and only if for
+every meet-irreducible open subset~$W$ there is a unique point~$y \in Y$ such
+that~$W = Y \setminus \overline{\{y\}}$.
+(This is a classical definition of sobriety.)
+
+\item Assuming the law of excluded middle, show that Hausdorff topological
+spaces validate the condition stated in part~(b) and hence are sober.
+\end{enumerate}
 \end{exercise}
@@
 \preceq V$).
+\end{exercise}
+
+\begin{exercise}{On the covariant approach}%
+Let~$A$ be a frame.
+\begin{enumerate}
+\item Show that~$A$ contains arbitrary (set-indexed) meets (though they are not
+required to be preserved by frame homomorphisms), by the impredicative
+construction
+\[ \bigwedge_{i \in I} U_i = \bigvee\{ \text{$V \preceq U_i$ for all~$i \in I$}
+\}. \]
+\item What are the arbitrary meets in the frame of open subsets of a
+topological space?
+\item Show that for any frame homomorphism~$\alpha : A \to A'$ there exists a
+monotone map~$\beta : A' \to A$ such that~\emph{$\beta$ is right adjoint
+to~$\alpha$}, that is such that
+\[ \alpha(U) \preceq V \quad\text{iff}\quad U \preceq \beta(V) \]
+for all~$U \in A$ and~$V \in A'$.
+\end{enumerate}
+{\scriptsize\emph{Note.} Based on the observation in part~(c), a
+\emph{covariant approach} to locales can be developed, in which locale maps~$X
+\to X'$ are defined to be certain kinds of maps~$\O(X) \to
+\O(X')$~\cite{picado-pultr:covariant}.\par}
 \end{exercise}
@@
 
+\begin{exercise}{The frame of radical ideals}%
+\label{ex:frame-of-radical-ideals}%
+Let~$A$ be a ring. We order the set~$\Idl(A)$ of ideals of~$A$ by inclusion.
+(See Section~\ref{sect:algebraic-preliminaries} for some algebraic
+preliminaries.)
+\begin{enumerate}
+\item Verify that~$\Idl(A)$ has arbitrary meets and joins, given by
+intersection and sum of ideals, but that it is in general not a frame.
+
+{\scriptsize\emph{Hint.} The distributive law fails in the polynomial ring~$\QQ[X,Y]$.\par}
+\item Show that the sub-poset~$\Rad(A)$ of radical ideals is a frame, with
+the same finite meets but with joins given by the radical of the join computed
+in~$\Idl(A)$.
+\item\label{item:join-top} Let~$\varphi$ be a truth value. Let~$\{\top\,|\,\varphi\}$ be the
+subsingleton subset of~$\Rad(A)$ which contains the top element~$\sqrt{(1)}$
+iff~$\varphi$. Verify that
+\[ \bigvee\{\top\,|\,\varphi\} = \{ f \in A \,|\, \text{$f$ is nilpotent
+or~$\varphi$} \} \]
+in~$\Rad(A)$.
+\end{enumerate}
+\end{exercise}
+
 \begin{exercise}{Coproducts of locales}%
@@
 \end{enumerate}
-The category of locales also has small products;
-Exercise~\ref{ex:products-of-locales} is devoted to this fact.
+{\scriptsize\emph{Note.} The category of locales also has small products;
+Exercise~\ref{ex:products-of-locales} is devoted to this fact.\par}
 \end{exercise}
@@
 \begin{exercise}{Local locales}%
+\label{ex:local-locales}%
 A locale~$X$ is \emph{local} iff for any open covering~$\top = \bigvee_i U_i$,
@@
 function symbols; nullary relation symbols~$\alpha_x$ and~$\beta_x$ for each
-rational number~$x$ (to be read as~``$x$ is contained in the lower respectively
-the upper cut''); and the following axioms, spelling out that the cut is bounded,
+rational number~$x$ (to be read as~``$x$ is contained in the lower (upper)
+cut''); and the following axioms, spelling out that the cut is bounded,
 rounded, open and located:
@@
 spectrum as a topological space proceeds in three steps: (1) Write down the
-axioms (of Dedekind cuts respectively prime filters). (2) Using the powerset
-axiom and separation, construct the set of Dedekind reals respectively the set
-of prime filters. (3) Devise a useful topology on the resulting set.
+axioms (of Dedekind cuts or of prime filters, respectively). (2) Using the powerset
+axiom and separation, construct the set of Dedekind reals or the set
+of prime filters, respectively. (3) Devise a useful topology on the resulting set.
 
@@
 \subsection{Sheaf semantics}
+\label{sect:sheaf-semantics}
 
 % XXX Introduction...
+
+% XXX A sheaf may well be nontrivial even though it does not have any global
+% sections. This is for instance the case for the subsheaf of the sheaf of continuous
+% complex-valued functions on~$\CC$, given by~$U \mapsto \{ f : U \to \CC \,|\,
+% \text{$f$ is continuous and~$f(z)^2 = z$ for all~$z \in U$} \}$. This sheaf
+% has lots of local sections (local square root maps), but no global one.
+% This phenomenon is related to the failure of (a simple version of) the
+% fundamental theorem of algebra (Exercise~\ref{ex:fta}).
 
@@
 propositional geometric theories). We strongly
-recommend the accessible account by Steve Vickers~\ref{vickers:continuity} for a fuller
+recommend the accessible account by Steve Vickers~\cite{vickers:continuity} for a fuller
 picture.
@@
 that any two reals are equal or not, show that the locale of surjections~$\NN
-\to \RR$ is zero-dimensional.
+\twoheadrightarrow \RR$ is zero-dimensional.
 \end{enumerate}
@@
 
-\begin{exercise}{Product of locales}\label{ex:products-of-locales}%
-Let~$X$ and~$X'$ be locales, classifying some propositional geometric
-theory~$\TT$ respectively~$\TT'$. Let~$\TT + \TT'$ be the combined theory
+\begin{exercise}{The product of locales}\label{ex:products-of-locales}%
+Let~$X$ and~$X'$ be locales, classifying propositional geometric
+theories~$\TT$ and~$\TT'$ respectively. Let~$\TT + \TT'$ be the combined theory
 (whose set of sorts is the disjoint union of the sets of sorts of~$\TT$
@@
 
-\begin{exercise}{Coproduct of locales, revisited}%
+\begin{exercise}{The coproduct of locales, revisited}%
 Let~$X$ and~$X'$ be locales, classifying some propositional geometric
-theory~$\TT$ respectively~$\TT'$. Find a theory in terms of~$\TT$
+theories~$\TT$ and~$\TT'$ respectively. Find a theory in terms of~$\TT$
 and~$\TT'$ which is classified by~$X \amalg X'$.
@@
 
-\item \emph{Deligne's completeness theorem} states that coherent locales (in
-fact, even coherent \emph{toposes}) are spatial. Show that Deligne's
-completeness theorem implies that propositional coherent theories
+\item \emph{Deligne's completeness theorem} states that coherent locales (and
+even coherent \emph{toposes}) are spatial. Show that Deligne's
+result implies that propositional coherent theories
 without any set-based models are inconsistent.
@@
 \begin{exercise}{On the failure of the fundamental theorem of algebra}%
+\label{ex:fta}%
 Let~$F$ be the sheaf of continuous complex-valued functions on~$\CC$. Give an
@@
 
-Such an understanding is imparted by the following theorem. This theorem is the
+Such an understanding is imparted by the following theorem. It is the
 central workhorse of our approach.
@@
 \emph{free~$k$-fields} (where a~$k$-field is a field together with a
-homomorphism from~$k$) and show that there are no free~$k$-fields.
+homomorphism from~$k$) and show that the only set~$M$ such that there is a
+free~$k$-field over~$M$ is the empty set.
 \end{enumerate}
@@
 Let~$A$ be a ring. In Proposition~\ref{prop:immediate-consequences}, we gave an
-abstract proof that~$A^\sim$ is local and hence nontrivial in the sense that~$1
-\neq 0$ (even if~$A$ itself happens to be trivial). Give a new proof of this
+abstract proof that~$A^\sim$ is (local and hence) nontrivial in the sense that~$1
+\neq 0$. This proof did not need to assume that~$A$ itself is nontrivial. Give a new proof of this
 fact, by verifying directly~$1 \models \neg(1 =_{A^\sim} 0)$ using the
@@
 is nilpotent.
+\item Show that~$A$ is reduced iff~$A^\sim$ is.
 \end{enumerate}
@@
 invertible or nilpotent.
+
+{\noindent\scriptsize\emph{Hint.} Use the description of the unique map~$\O(\pt) \to
+\O(\Spec(A)),\,\varphi \mapsto \bigvee\{\top\,|\,\varphi\}$ computed in
+part~(\ref{item:join-top}) of Exercise~\ref{ex:frame-of-radical-ideals}.\par}
+\end{exercise}
+
+\begin{exercise}{Locality of the spectrum}%
+Let~$A$ be a ring. Show that~$\Spec(A)$ is local in the sense of
+Exercise~\ref{ex:local-locales} if and only if~$A$ is local as a ring.
 \end{exercise}
@@
 such that for any two elements, one divides the other. A \emph{Prüfer domain}
-is an integral domain such that any finitely generated ideal is locally a
-principal ideal.
-\begin{enumerate}
-\item Let~$A$ be a ring. Show that, if~$A$ is an integral domain then so
-is~$A^\sim$. Does the converse hold?
+is an integral domain such that any finitely generated ideal~$\aaa$ is locally a
+principal ideal (in the sense that there exists a partition~$1 = f_1 + \cdots +
+f_n$ such that, for each index~$i$, the ideal~$\aaa[f_i^{-1}]$ is finitely
+generated in~$A[f_i^{-1}]$).
+\begin{enumerate}
+\item Let~$A$ be a ring. Show that~$A^\sim$ is an integral domain if~$A$ is.
+Does the converse hold?
 \item Let~$A$ be a valuation domain. Show that any matrix over~$A$ can be put
@@
 
+XXX Cite Frames and Locales: Topology Without Points
+
 X \models phi

### 2020-04-05 paper-herrsching2019/main.tex
Fixed typography of exercises

@@
 \usepackage{ws-rv-van}
-\usepackage{mathtools,stmaryrd,xspace,bussproofs}
-
-\newlength{\exerciseskip}
-\setlength{\exerciseskip}{1.5em}
-\newcounter{exercisenr}
-\newenvironment{exercise}[1]{
-  \refstepcounter{exercisenr}
-  \noindent\textbf{Exercise \theexercisenr{}.} \emph{#1} \smallskip\par\noindent%
-}{\par\vspace{\exerciseskip}}
-
+
+% NOTE TO EDITORS AND LAYOUTERS: I require the following packages.
 \usepackage[normalem]{ulem}
@@
 \usetikzlibrary{calc}
+\usepackage{mathtools,stmaryrd,xspace,bussproofs}
+
+\usepackage[english]{babel}
+
+% NOTE TO EDITORS AND LAYOUTERS: This package increases the quality of the
+% typesetting. I personally recommebd using it, but my contribution does not
+% depend on it.
+\usepackage[expansion=true,protrusion=true]{microtype}
+
+\newtheorem{ex}{Exercise}
+\let\Exercisefont\upshape
+\def\Exerciseheadfont{\bfseries}
+\newenvironment{exercise}[1]{
+  \begin{ex}[#1]
+}{\end{ex}}
+
 \newcommand{\hcancel}[5]{%
@@
 
-\begin{exercise}{The topological space of points of a locale}%
+\begin{exercise}{The space of points of a locale}%
 \label{ex:canonical-topology}%
-Devise a canonical topology on the set of points of a locale.
-\end{exercise}
-
-\begin{exercise}{On the embedding of sober topological spaces into locales}%
-Verify the statements in Remark~\ref{rem:sober-embed}.
+Devise a canonical to\-po\-logy on the set of points of a locale.
+\end{exercise}
+
+\begin{exercise}{Embedding sober topological spaces into locales}%
+Verify the statements made in Remark~\ref{rem:sober-embed}.
 \end{exercise}
@@
 such that~$Y \not\subseteq W$.
-\begin{enumerate}
+\begin{alphlist}[(c)]
 \item Assuming the law of excluded middle, show that the completely prime
 filters of a topological space are in canonical one-to-one correspondence with
-the meet-irreducible open subsets.
+the meet-irreducible open subsets.\smallskip
 
@@
 spaces validate the condition stated in part~(b) and hence are sober.
-\end{enumerate}
+\end{alphlist}
+{\scriptsize\emph{Note.} Without the law of excluded middle, Hausdorff spaces
+and even metric spaces may fail to be sober. In fact, already the topological
+space of the rational numbers can fail to be
+sober~\cite[Example~8.14(iv)]{fourman-scott:sheaves-and-logic}.\par}
+\end{exercise}
+
+\begin{exercise}{Complete metric spaces are sober}
+A \emph{metric space} is a set~$Y$ together with a relation~$d \subseteq Y
+\times Y \times \QQ_{\geq0}$, written~``$d(x,y) \leq q$'', such that the
+following axioms are satisfied.
+\begin{align*}
+  d(x,y) \leq q \wedge q \leq r &\Longrightarrow d(x,y) \leq r \\
+  (\forall r \in \QQ_{> q}\_ d(x,y) \leq r) &\Longrightarrow d(x,y) \leq q \\
+  &\mathrel{\phantom{\Longrightarrow}} \exists q \in \QQ_{\geq0}\_ d(x,y) \leq q \\
+  &\mathrel{\phantom{\Longrightarrow}} d(x,x) \leq 0 \\
+  d(x,y) \leq 0 &\Longrightarrow x = y \\
+  d(x,y) \leq q &\Longrightarrow d(y,x) \leq q \\
+  d(x,y) \leq q \wedge d(y,z) \leq r &\Longrightarrow d(x,z) \leq q+r
+\end{align*}
+A \emph{Cauchy process} (or \emph{multi-valued modulated Cauchy sequence}) in a metric
+space~$Y$ is a map~$\alpha : \QQ_{>0} \to P(Y)$ such that all sets~$\alpha(q)$
+are inhabited and such that~$d(x,x') \leq q + q'$ for all~$q,q' \in \QQ_{>0}$
+and~$x \in \alpha(q), x' \in \alpha(q')$. Such a process \emph{converges
+to}~$x_0 \in Y$ iff~$d(x,x_0) \leq q$ for all~$q \in \QQ_{>0}$
+and~$x \in \alpha(q)$. A metric space~$Y$ is \emph{complete} iff every Cauchy
+process in~$Y$ converges.
+
+Verify, without using the law of excluded middle, that any complete metric
+space is sober.
 \end{exercise}
@@
 
-\begin{exercise}{On the covariant approach}%
+\begin{exercise}{The covariant approach to locales}%
 Let~$A$ be a frame.
-\begin{enumerate}
+\begin{alphlist}[(c)]
 \item Show that~$A$ contains arbitrary (set-indexed) meets (though they are not
@@
 for all~$U \in A$ and~$V \in A'$.
-\end{enumerate}
+\end{alphlist}
 {\scriptsize\emph{Note.} Based on the observation in part~(c), a
@@
 Let~$X$ be a locale. Show that the following conditions are equivalent.
-\begin{enumerate}
+\begin{alphlist}[(c)]
 \item For any opens~$U, V \in \O(X)$: If~$x \inplus U \Rightarrow y \inplus V$
@@
 \item The canonical morphism~$L(\Pt(X)) \to X$ is an isomorphism.
-\end{enumerate}
+\end{alphlist}
 \end{exercise}
@@
 preliminaries.)
-\begin{enumerate}
+\begin{alphlist}[(b)]
 \item Verify that~$\Idl(A)$ has arbitrary meets and joins, given by
-intersection and sum of ideals, but that it is in general not a frame.
+intersection and sum of ideals, but that it is in general not a
+frame.\smallskip
 
@@
 in~$\Rad(A)$.
-\end{enumerate}
+\end{alphlist}
 \end{exercise}
@@
 \begin{exercise}{Coproducts of locales}%
-\begin{enumerate}
+\begin{alphlist}[(c)]
 \item Show that the category of frames has small products.
@@
 that~$L(M)$ is the coproduct~$\coprod_{x \in M} \pt$.
-\end{enumerate}
+\end{alphlist}
 {\scriptsize\emph{Note.} The category of locales also has small products;
@@
 
-\begin{exercise}{Localic maps into discrete spaces}%
+\begin{exercise}{Maps into discrete spaces}%
 Let~$X$ be a locale. Let~$A$ be a set and endow it with the discrete topology. Show
@@
 \begin{exercise}{Discrete locales as classifying locales}%
-\begin{enumerate}
+\begin{alphlist}[(b)]
 \item Find a theory which is classified by the two-point locale (the locale
@@
 which is classified by~$L(M)$.
-\end{enumerate}
+\end{alphlist}
 \end{exercise}
@@
 \label{ex:exotic-locales}%
-\begin{enumerate}
+\begin{alphlist}[(d)]
 \item Define the locale of injections~$\RR \hookrightarrow \NN$.
-\item Define the locale of partial surjections~$\NN \twoheadrightarrow \NN^\NN$.
+\item Define the locale of partial surjections~$\NN \twoheadrightarrow \NN^\NN$.\smallskip
 
@@
 endpoints such that the union of any finite number of those intervals has total
-Lebesgue measure less than~$\frac{1}{2}$''.
+Lebesgue measure less than~$\frac{1}{2}$''.\smallskip
 
@@
 \twoheadrightarrow \RR$ is zero-dimensional.
-\end{enumerate}
-\end{exercise}
-
-\begin{exercise}{Models of theories as points of the classifying locales}%
+\end{alphlist}
+\end{exercise}
+
+\begin{exercise}{Models as points}%
 Let~$\TT$ be a propositional geometric theory. Verify that the points
@@
 open interval~$\{ \xi \in \Pt(\RR) \,|\, x < \xi < y \}$ in the set of Dedekind reals.
-\begin{enumerate}
+\begin{alphlist}[(f)]
 \item Extend the notation to allow for~$x,y=\pm\infty$.
@@
 is spatial.
-\end{enumerate}
+\end{alphlist}
 {\scriptsize\emph{Note.} Exercise~\ref{ex:exotic-locales} lists settings in
@@
 fix~$k$, as described on page~\pageref{par:theory-automorphisms}.
-\begin{enumerate}
+\begin{alphlist}[(g)]
 \item Explain how to obtain a morphism~$\Gal(L|k) \to \Gal(L|k)$ which on the
@@
 \item Show that~$\Gal(L|k)$ is spatial in classical mathematics.
-\end{enumerate}
+\end{alphlist}
 \end{exercise}
@@
 iff all of its axioms consist of coherent formulas.
-\begin{enumerate}
+\begin{alphlist}[(d)]
 \item Show that any coherent locale classifies some coherent theory.
 \item Let~$\TT$ be a propositional coherent theory. Show that~$L(\TT)$ is
-coherent.
+coherent.\smallskip
 
@@
 without any set-based models are inconsistent.
-\end{enumerate}
+\end{alphlist}
 \end{exercise}
@@
 Let~$X$ be a locale.
-\begin{enumerate}
+\begin{alphlist}[(b)]
 \item Show that any subsheaf~$F \subseteq 1$ of the terminal sheaf on~$X$ gives rise
@@
 the subsheaves of~$1$ and the opens of~$X$.
-\end{enumerate}
+\end{alphlist}
 \end{exercise}
@@
 \bot$ implies~$U = \bot$. Let~$\varphi$ be a formula over a locale~$X$.
-\begin{enumerate}
+\begin{alphlist}[(b)]
 \item Show that there is a largest open,
-denoted~``$\llbracket\varphi\rrbracket$'', such that~$U \models \varphi$.
+denoted~``$\llbracket\varphi\rrbracket$'', such that~$U \models \varphi$.\smallskip
 
@@
 is dense.
-\end{enumerate}
+\end{alphlist}
 \end{exercise}
@@
 functions on~$X$.
-\begin{enumerate}
+\begin{alphlist}[(d)]
 \item Let~$U$ be an open of~$X$. Let~$f \in \C(U)$. Show using the result
@@
 $\Sh(\CC) \models \forall f\?\O\_ \forall g\?\O\_ f = g \vee \neg(f = g)$.
-\end{enumerate}
-\end{exercise}
-
-\begin{exercise}{On the failure of the fundamental theorem of algebra}%
+\end{alphlist}
+\end{exercise}
+
+\begin{exercise}{The failure of the fundamental theorem of algebra}%
 \label{ex:fta}%
@@
 example of a function~$f : \CC \to \CC$ such that it is not the case that
-\[ \Sh(X) \models \exists z\?F\_ z^2 = f. \]
+\[ \Sh(\CC) \models \exists z\?F\_ z^2 = f. \]
 {\scriptsize\emph{Note.} The sheaf~$F$ can be obtained by constructing, in the
@@
 \item The field property of~$A^\sim$ should be appropriately appreciated: While
-there is a general machinery for free constructions of this kind, this
+there is a general machinery for free constructions of this generalized kind, this
 machinery is \emph{not} applicable for constructing free fields.
@@
 
+\begin{remark}There is general topos-theoretic machinery for generalized free
+constructions of this kind~\cite{coste:sheaf-representation,cole:spectra}, and
+this machinery can be used to construct several kinds of spectra, including the
+ordinary spectrum of a ring. However, this machinery does have limitations, and
+in particular it cannot be used to construct free fields.
+\end{remark}
+
 \subsection{Understanding the sheaf model}
@@
 \begin{remark}An immediate corollary of Theorem~\ref{thm:workhorse} is
-that~$\Spec(A)$ is compact. However, the stronger result that~$\Spec(A)$ is
+that~$\Spec(A)$ is compact, since if~$f \in \sqrt{(g_i)_i}$ then also~$f \in
+\sqrt{(g_{i_1},\ldots,g_{i_n})}$ for suitable indices~$i_1,\ldots,i_n$.
+However, the stronger result that~$\Spec(A)$ is
 even a \emph{coherent locale} can also be deduced from a general theorem
@@
 
-\begin{proposition}[Grothendieck's generic freeness lemma, simple version]
+\begin{proposition}[generic freeness lemma, simple version]
 \label{thm:baby-grothendieck}
@@
 \begin{exercise}{Free fields}
-\begin{enumerate}
+\begin{alphlist}[(b)]
 \item A \emph{free field} over a set~$M$ would be
@@
 that free fields exist for no set~$M$.
-\item Let~$k$ be a field. Devise a hypothetical notion of
+\item Let~$k$ be a field. Devise a notion of
 \emph{free~$k$-fields} (where a~$k$-field is a field together with a
@@
 free~$k$-field over~$M$ is the empty set.
-\end{enumerate}
+\end{alphlist}
 \end{exercise}
@@
 Theorem~\ref{thm:basic-properties-sheaf-semantics-algebraic}
-also apllies to this extension, and moreover, this extension in sound with
+also applies to this extension, and moreover, this extension in sound with
 respect to infinitary intuitionistic logic.
-\begin{enumerate}
+\begin{alphlist}[(c)]
 \item Prove that:
@@
 \item Show that~$A$ is reduced iff~$A^\sim$ is.
-\end{enumerate}
+\end{alphlist}
 \end{exercise}
@@
 Let~$M$ be a module over a ring~$A$.
-\begin{enumerate}
+\begin{alphlist}[(c)]
 \item Show that~$M$ is finitely generated iff~$M^\sim$ is.
@@
 \item Show that~$M$ is finitely presented iff~$M^\sim$ is.
-\end{enumerate}
+\end{alphlist}
 \end{exercise}
@@
 Let~$M$ an~$N$ be modules over a ring~$A$. Let~$f : M \to N$ be a linear map.
-\begin{enumerate}
+\begin{alphlist}[(b)]
 \item Show that~$f$ is injective iff~$f^\sim : M^\sim \to N^\sim$ is injective.
 \item Show that~$f$ is surjective iff~$f^\sim : M^\sim \to N^\sim$ is surjective.
-\end{enumerate}
-\end{exercise}
-
-\begin{exercise}{The units of~$A^\sim$}%
+\end{alphlist}
+\end{exercise}
+
+\begin{exercise}{Units of the sheaf model}%
 Let~$A$ be a ring. Let~$f \in A$. Let~$x \in A[f^{-1}]$. Show using the result
@@
 
-\begin{exercise}{Identifying the triviality of the spectrum}%
+\begin{exercise}{Triviality of the spectrum}%
 Let~$A$ be a ring. Verify, without assuming~\BPIT, that~$\Spec(A)$ is
 isomorphic to the one-point locale iff any element of~$A$ is either
-invertible or nilpotent.
+invertible or nilpotent.\smallskip
 
@@
 \begin{exercise}{Locality of the spectrum}%
-Let~$A$ be a ring. Show that~$\Spec(A)$ is local in the sense of
+Let~$A$ be a ring. Show that the spectrum of~$A$ is local in the sense of
 Exercise~\ref{ex:local-locales} if and only if~$A$ is local as a ring.
@@
 \begin{exercise}{Spatiality of the spectrum}%
-Let~$A$ be a ring. Let~$\Spec(A)$ be its spectrum.
-\begin{enumerate}
+Let~$A$ be a ring.
+\begin{alphlist}[(c)]
 \item Verify that~$\Spec(A)$ is spatial if and only if for any element~$x \in
@@
 \item Conclude that the spectrum of any ring is spatial iff~\BPIT holds.
-\end{enumerate}
+\end{alphlist}
 \end{exercise}
@@
 generated in~$A[f_i^{-1}]$).
-\begin{enumerate}
+\begin{alphlist}[(d)]
 \item Let~$A$ be a ring. Show that~$A^\sim$ is an integral domain if~$A$ is.
@@
 result of part~(b) to~$A^\sim$.
-\end{enumerate}
+\end{alphlist}
 \end{exercise}
@@
 \begin{exercise}{A basic version of Kaplansky's theorem}
-\begin{enumerate}
+\begin{alphlist}[(c)]
 \item Let~$A$ be a local ring. Let~$\aaa \subseteq A$ be a finitely generated

### 2020-04-05 paper-herrsching2019/main.tex
Frobenius rules required in geometric logic
Baby Barr (with variables)

@@
 % depend on it.
-\usepackage[expansion=true,protrusion=true]{microtype}
+%\usepackage[expansion=true,protrusion=true]{microtype}
 
@@
 
+    \textnormal{mixed rules (unnecessary in presence of the double rule for implication)} \\\smallskip
+    \vspace{-0.5em}
+    \phantom{a}\hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}
+    \UnaryInfC{$(\bigvee_i \varphi_i) \wedge \psi \seq{\vec x} \bigvee_i (\varphi_i \wedge \psi)$}
+    \DisplayProof
+    \hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}
+    \UnaryInfC{$(\exists y\?Y\_ \varphi) \wedge \psi \seq{\vec x} \exists y\?Y\_ (\varphi \wedge \psi)$}
+    \DisplayProof
+    \makebox[0pt]{\shifttext{\tiny ($y$ none of the~$\vec x$)}}
+    \hfill\phantom{a}
+    \bigskip
+
     \textnormal{rules for equality} \\\smallskip
@@
 
-\paragraph{The geometric theory of rings.} This theory has one sort, $R$; five function symbols:
+We explicitly allow empty carriers, that is we do not demand that the
+sets~$\brak{X}$ are inhabited.
+Examples for geometric theories include the following.
+\begin{enumerate}
+\item \emph{The geometric theory of rings.} This theory has one sort, $R$; five function symbols:
 $0$ and $1$ (nullary), $-$ (unary), $+$ and $\cdot$ (binary); no relation
@@
 In a similar vein, there is the geometric theory of monoids, of groups, and so
-on.
-
-\paragraph{The geometric theory of objects.} This theory has one sort, $X$, and no function
+on. Some of these theories use the infinite disjunctions supported by geometric
+logic: For instance, the theory of fields of positive characteristic or the
+theory of torsion groups have among their axioms
+\[ \top \vdash \bigvee_{n \in \NN_{\geq 1}} (\underbrace{1 + \cdots + 1}_{\text{$n$ summands}} = 0)
+  \qquad
+  \top \seq{g:G} \bigvee_{n \in \NN} (g^n = e). \]
+
+\item \emph{The geometric theory of objects.}\label{item:theory-of-objects}
+This theory has one sort, $X$, and no function
 symbols, relation symbols or axioms. A set-based model of this theory is just a
@@
 
-\paragraph{The geometric theory of surjections~$f : \NN \twoheadrightarrow
+\item \emph{The geometric theory of surjections~$f : \NN \twoheadrightarrow
 \RR$.} This theory has no
@@
 
-\paragraph{The geometric theory of Dedekind cuts.}\label{par:dedekind-cuts}
+\item \emph{The geometric theory of Dedekind cuts.}\label{par:dedekind-cuts}
 This theory has no sorts; no
@@
 
-\paragraph{The geometric theory of prime ideals of a given ring~$A$.} In classical
+\item \emph{The geometric theory of prime ideals of a given ring~$A$.} In classical
 commutative algebra, the notion of a \emph{prime ideal} of a ring~$A$ is
@@
 
-\paragraph{The geometric theory of automorphisms of a field.}\label{par:theory-automorphisms}
+\item \emph{The geometric theory of automorphisms of a field.}\label{par:theory-automorphisms}
 Let~$L|k$ be a field
@@
 
-\paragraph{The inconsistent geometric theory.} This theory has no sorts,
+\item \emph{The inconsistent geometric theory.} This theory has no sorts,
 function symbols or relations, but the single axiom $\top \vdash \bot$.
@@
 
-\paragraph{The empty geometric theory.} This theory does not have any sorts,
+\item \emph{The empty geometric theory.} This theory does not have any sorts,
 function symbols, relations or axioms. There is exactly one set-based model of
 this theory, the empty structure.
+\end{enumerate}
+
+\paragraph{Extracting constructive content from classical proofs.} Geometric
+logic strikes a fine balance between being sufficiently expressive in order to
+have rich applications, but also sufficiently restrictive so that \emph{Barr's
+theorem} holds. This theorem comes in two versions, both concerned with
+eliminating classical reasoning from proofs of geometric sequents:
+\begin{enumerate}
+\item If infinitary classical logic proves a geometric sequent, then so does
+geometric logic.
+\item If a specific version of extensional type theory with the law of excluded
+middle and the axiom of choice proves a geometric sequent, then so does
+geometric logic.
+\end{enumerate}
+For instance, consider the theory of algebraically closed geometric fields,
+that is the theory of rings extended by the axioms
+\begin{align*}
+  \top &\seq{x:R} (x = 0) \vee (\exists y\?R\_ xy = 1) \\
+  \top &\seq{a_0:R,\ldots,a_{n-1}:R} \exists x\?R\_ x^n + a_{n-1}x^{n-1} +
+  \cdots + a_1x + a_0 = 0.
+\end{align*}
+The \emph{Nullstellensatz} is the statement that for any
+system~$f_1,\ldots,f_n \in K[X_1,\ldots,X_m]$ of polynomials over an
+algebraically closed field~$K$, either the polynomials~$f_i$ have a common zero
+or there are polynomials~$g_i$ such that~$1 = \sum_i f_i g_i$ (such an
+expression is an \emph{algebraic witness} of the nonexistence of a common
+zero).
+
+The Nullstellensatz is a theorem of classical mathematics, employing the
+axiom of choice. Since it can be formulated as a geometric sequent over the
+geometric theory of algebraically closed geometric fields, Barr's theorem
+implies that there is also a constructive proof, in fact even a purely
+geometric proof.
+
+The simpler version of Barr's theorem is constructive, and although we will not
+use it in the remainder of this text, we take the opportunity to give a
+self-contained proof.
+
+However, the more powerful version is not constructive. This puts us in the
+curious situation that we are promised, unconstructively, that there exists
+a constructive proof. The textbook argument~\cite[Theorem~7.57]{johnstone:topos-theory} establishing the more powerful
+version of Barr's theorem requires
+the axiom of choice, but recently Michael Rathjen showed that~\textsc{zf}
+suffices~\cite{rathjen:barr}.
+
+\begin{theorem}Let~$\TT$ be a geometric theory. Let~$\sigma$ be a geometric
+sequent over the signature of~$\TT$. Then the following statements are
+equivalent.
+\begin{enumerate}
+\item There is a~$\TT$-derivation of~$\sigma$ in geometric logic.
+\item There is a~$\TT$-derivation of~$\sigma$ in infinitary intuitionistic logic.
+\item There is a~$\TT$-derivation of~$\sigma$ in infinitary classical logic.
+\end{enumerate}
+\end{theorem}
+
+\begin{proof}It is immediate that~(1) implies~(3).
+
+To show that~(3) implies~(2), we combine the double negation translation
+with Friedman's trick. We give a sketch of the main thrust of the argument,
+omitting lengthy verifications. Let~$\sigma \equiv (\alpha \seq{\vec x} \beta)$.
+We introduce the operator~$\nabla$ with~$\nabla\varphi \defeqv
+((\varphi \Rightarrow \beta) \Rightarrow \beta)$, where care has to be taken to
+avoid capturing the variables~$\vec x$. This operator is a \emph{local
+operator} in the sense that for any formula~$\varphi$ in some context~$\vec y$,
+infinitary intuitionistic logic proves the sequents
+\[
+\top \seq{\vec x,\vec y} (\varphi \Rightarrow \nabla\varphi),\
+\top \seq{\vec x,\vec y} (\nabla\nabla\varphi \Rightarrow \nabla\varphi),\
+\top \seq{\vec x,\vec y} (\nabla\varphi \wedge \nabla\psi \Leftrightarrow \nabla(\varphi \wedge \psi)).
+\]
+We then set up the~\emph{$\nabla$-translation}~$\varphi \mapsto \varphi^\nabla$
+for infinitary first-order formulas over the signature of~$\TT$ as described in
+Table~\ref{table:nabla-translation}.
+Again, care has to be taken to ensure that there is no variable collision with the
+variables~$\vec x$. The chosen local operator is special in that
+the~$\nabla$-translation of the law of excluded middle,
+\[ \top \seq{\vec x,\vec y} \nabla(\varphi^\nabla \vee (\varphi^\nabla
+\Rightarrow \nabla\bot)) \]
+is intuitionistically provable. Straightforward inductions on the structure of
+formulas or derivations then establish the following properties.
+\begin{alphlist}
+\item For any infinitary first-order formula~$\varphi$ in some context~$\vec y$,
+infinitary intuitionistic logic proves~$\nabla(\varphi^\nabla) \seq{\vec x,\vec
+y} \varphi^\nabla$.
+\item For any geometric formula~$\varphi$ in some context~$\vec y$, infinitary
+intuitionistic logic proves~$\varphi^\nabla \mathrel{\dashv\vdash\!\!\!_{\vec x,\vec y}} \nabla\varphi$.
+\item For any derivation of a sequent~$\varphi \seq{\vec y} \psi$ in infinitary
+classical logic, there is a derivation of~$\varphi^\nabla \seq{\vec y}
+\psi^\nabla$ in infinitary intuitionistic logic.
+\end{alphlist}
+Combining these ingredients, the given derivation of~$\sigma \equiv (\alpha
+\seq{\vec x} \beta)$ in infinitary classical logic can be transformed to give a derivation of
+\begin{align*}
+  \alpha[\vec x'/\vec x] &\seq{\vec x,\vec x'}
+  \nabla(\alpha[\vec x'/\vec x]) \seq{\vec x,\vec x'}
+  (\alpha[\vec x'/\vec x])^\nabla \\
+  &\seq{\vec x,\vec x'}
+  (\beta[\vec x'/\vec x])^\nabla \seq{\vec x,\vec x'}
+  \nabla(\beta[\vec x'/\vec x]) \mathrel{\dashv\vdash\!\!\!_{\vec x,\vec x'}}
+  ((\beta[\vec x'/\vec x] \Rightarrow \beta) \Rightarrow \beta).
+\end{align*}
+in infinitary intuitionistic logic. Substituting~$\vec x'$ by~$\vec x$, we obtain
+\[ \alpha \seq{\vec x} ((\beta \Rightarrow \beta) \Rightarrow \beta) \seq{\vec
+x} \beta. \]
+
+To verify that~(2) implies~(1), XXX
+\end{proof}
+
+\begin{table}[ht]
+  \tbl{The definition of the~$\nabla$-translation}{\begin{minipage}{\textwidth}\begin{align*}
+    (x = y)^\nabla &\defeqv \nabla(x = y) \\
+    R(y_1,\ldots,y_m)^\nabla &\defeqv \nabla(R(y_1,\ldots,y_m)) \\
+    \top^\nabla &\defeqv \nabla\top \quad \\
+    \bot^\nabla &\defeqv \nabla\bot \\
+    (\varphi \wedge \psi)^\nabla &\defeqv \varphi^\nabla \wedge \psi^\nabla &
+    \textstyle (\bigwedge_i \varphi_i)^\nabla &\defeqv \textstyle \bigwedge_i \varphi_i^\nabla \\
+    (\varphi \vee \psi)^\nabla &\defeqv \nabla(\varphi^\nabla \vee \psi^\nabla) &
+    \textstyle (\bigvee_i \varphi_i)^\nabla &\defeqv \textstyle \nabla(\bigvee_i \varphi_i^\nabla) \\
+    (\varphi \Rightarrow \psi)^\nabla &\defeqv (\varphi^\nabla \Rightarrow \psi^\nabla) \\
+    (\forall x\?X\_ \varphi)^\nabla &\defeqv (\forall x\?X\_ \varphi^\nabla) &
+    (\exists x\?X\_ \varphi)^\nabla &\defeqv \nabla(\exists x\?X\_ \varphi^\nabla)
+  \end{align*}\end{minipage}}
+  \label{table:nabla-translation}
+\end{table}
 
@@
 
-\begin{exercise}{Discrete locales as classifying locales}%
+\begin{exercise}{The theory of finite objects}%
+\begin{alphlist}[(c)]
+\item Extend the geometric theory of objects
+(page~\ref{item:theory-of-objects}) by relation symbols and axioms, but not by
+constant symbols, to ensure that the set-based models are precisely the
+two-element sets.
+\item Write down a geometric theory whose set-based models are precisely the
+(Kuratowski-)finite sets.\smallskip
+
+{\scriptsize\emph{Hint.} Employ countably many relation symbols~$R_n
+\hookrightarrow X \times \cdots \times X$, where~$R_n(x_1,\ldots,x_n)$ should
+signify that the elements~$x_1,\ldots,x_n$ exhaust all elements of~$X$. A
+solution is given in~\cite[Example~D1.1.7(k)]{johnstone:elephant}.\par}
+\end{alphlist}
+\end{exercise}
+
+\begin{exercise}{Discrete locales}%
 \begin{alphlist}[(b)]
@@
 \begin{exercise}{The localic unit interval}%
-Extend the theory of Dedekind cuts in such a way that its models are those
+Extend the theory of Dede\-kind cuts in such a way that its models are those
 Dedekind reals which are elements of~$[0,1]$.

### 2020-04-05 paper-herrsching2019/main.tex
Simplification by suppressing variables

@@
 However, the more powerful version is not constructive. This puts us in the
-curious situation that we are promised, unconstructively, that there exists
-a constructive proof. The textbook argument~\cite[Theorem~7.57]{johnstone:topos-theory} establishing the more powerful
-version of Barr's theorem requires
-the axiom of choice, but recently Michael Rathjen showed that~\textsc{zf}
-suffices~\cite{rathjen:barr}.
+curious situation that we are promised, unconstructively, that there exist
+certain constructive proofs. The textbook
+argument~\cite[Theorem~7.57]{johnstone:topos-theory} establishing the more
+powerful version of Barr's theorem requires the axiom of choice, but recently
+Michael Rathjen lowered the required metatheory to~\textsc{zf}~\cite[Remark~4.2]{rathjen:barr}.
 
@@
 with Friedman's trick. We give a sketch of the main thrust of the argument,
-omitting lengthy verifications. Let~$\sigma \equiv (\alpha \seq{\vec x} \beta)$.
+omitting lengthy verifications. Let~$\sigma \equiv (\alpha \seq{\vec a} \beta)$.
+For notational simplicity, we assume that the context~$\vec a$ is empty, though
+all of the following can be adapted to the nonempty case if care is taken to
+avoid variable capture.
+
 We introduce the operator~$\nabla$ with~$\nabla\varphi \defeqv
-((\varphi \Rightarrow \beta) \Rightarrow \beta)$, where care has to be taken to
-avoid capturing the variables~$\vec x$. This operator is a \emph{local
-operator} in the sense that for any formula~$\varphi$ in some context~$\vec y$,
-infinitary intuitionistic logic proves the sequents
+((\varphi \Rightarrow \beta) \Rightarrow \beta)$. This operator is a
+\emph{local operator}~\cite[Section~14.5]{goldblatt:topoi} in the sense that
+for any formula~$\varphi$ in some context~$\vec x$, infinitary intuitionistic
+logic proves the sequents
 \[
-\top \seq{\vec x,\vec y} (\varphi \Rightarrow \nabla\varphi),\
-\top \seq{\vec x,\vec y} (\nabla\nabla\varphi \Rightarrow \nabla\varphi),\
-\top \seq{\vec x,\vec y} (\nabla\varphi \wedge \nabla\psi \Leftrightarrow \nabla(\varphi \wedge \psi)).
+\top \seq{\vec x} (\varphi \Rightarrow \nabla\varphi),\
+\top \seq{\vec x} (\nabla\nabla\varphi \Rightarrow \nabla\varphi),\
+\top \seq{\vec x} (\nabla\varphi \wedge \nabla\psi \Leftrightarrow \nabla(\varphi \wedge \psi)).
 \]
@@
 the~$\nabla$-translation of the law of excluded middle,
-\[ \top \seq{\vec x,\vec y} \nabla(\varphi^\nabla \vee (\varphi^\nabla
-\Rightarrow \nabla\bot)) \]
+\[ \top \seq{\vec x} \nabla(\varphi^\nabla \vee (\varphi^\nabla \Rightarrow \nabla\bot)), \]
 is intuitionistically provable. Straightforward inductions on the structure of
@@
 \item For any infinitary first-order formula~$\varphi$ in some context~$\vec y$,
-infinitary intuitionistic logic proves~$\nabla(\varphi^\nabla) \seq{\vec x,\vec
-y} \varphi^\nabla$.
-\item For any geometric formula~$\varphi$ in some context~$\vec y$, infinitary
-intuitionistic logic proves~$\varphi^\nabla \mathrel{\dashv\vdash\!\!\!_{\vec x,\vec y}} \nabla\varphi$.
-\item For any derivation of a sequent~$\varphi \seq{\vec y} \psi$ in infinitary
-classical logic, there is a derivation of~$\varphi^\nabla \seq{\vec y}
+infinitary intuitionistic logic proves~$\nabla(\varphi^\nabla) \seq{\vec x} \varphi^\nabla$.
+\item For any geometric formula~$\varphi$ in some context~$\vec x$, infinitary
+intuitionistic logic proves~$\varphi^\nabla \mathrel{\dashv\vdash\!\!\!_{\vec x}} \nabla\varphi$.
+\item For any derivation of a sequent~$\varphi \seq{\vec x} \psi$ in infinitary
+classical logic, there is a derivation of~$\varphi^\nabla \seq{\vec x}
 \psi^\nabla$ in infinitary intuitionistic logic.
@@
 Combining these ingredients, the given derivation of~$\sigma \equiv (\alpha
-\seq{\vec x} \beta)$ in infinitary classical logic can be transformed to give a derivation of
-\begin{align*}
-  \alpha[\vec x'/\vec x] &\seq{\vec x,\vec x'}
-  \nabla(\alpha[\vec x'/\vec x]) \seq{\vec x,\vec x'}
-  (\alpha[\vec x'/\vec x])^\nabla \\
-  &\seq{\vec x,\vec x'}
-  (\beta[\vec x'/\vec x])^\nabla \seq{\vec x,\vec x'}
-  \nabla(\beta[\vec x'/\vec x]) \mathrel{\dashv\vdash\!\!\!_{\vec x,\vec x'}}
-  ((\beta[\vec x'/\vec x] \Rightarrow \beta) \Rightarrow \beta).
-\end{align*}
-in infinitary intuitionistic logic. Substituting~$\vec x'$ by~$\vec x$, we obtain
-\[ \alpha \seq{\vec x} ((\beta \Rightarrow \beta) \Rightarrow \beta) \seq{\vec
-x} \beta. \]
-
-To verify that~(2) implies~(1), XXX
+\vdash \beta)$ in infinitary classical logic can be transformed to give a derivation of
+\[
+  \alpha \vdash
+  \nabla\alpha \vdash
+  \alpha^\nabla \vdash
+  \beta^\nabla \vdash
+  \nabla\beta \equiv
+  ((\beta \Rightarrow \beta) \Rightarrow \beta) \vdash
+  \beta
+\]
+in infinitary intuitionistic logic.
+
+To verify that~(2) implies~(1), we define a semantics by the clauses in
+Table~\ref{table:classifying-topos}. The following properties of the semantics
+can then be established by induction:
+\begin{alphlist}[(d)]
+\item XXX % ...; idea: do only the propositional case
+\end{alphlist}
 \end{proof}
@@
   \label{table:nabla-translation}
+\end{table}
+
+\begin{table}
+  \tbl{A syntactic model for XXX}{\begin{tabular}{@{}l@{\ \ }c@{\ \ }l@{}}
+    $\alpha \models \beta$ &iff& $\TT$ proves~$\alpha \vdash \beta$ \qquad(for~$\beta$ a relation symbol) \\
+    $\alpha \models \top$ &iff& true \\
+    $\alpha \models \varphi \wedge \psi$ &iff& $\alpha \models \varphi$ and $\alpha \models \psi$ \\
+    $\alpha \models \bigvee_i \varphi_i$ &iff& there are coherent
+    formulas~$\alpha_1,\ldots,\alpha_n$ such that \\ &&\qquad $\TT$ proves~$\alpha \vdash
+    \alpha_1 \vee \cdots \vee \alpha_n$ and such for each~$k \in \{1,\ldots,n\}$,
+    \\ &&\qquad\qquad there is an index~$i \in I$ such that~$\alpha_k \models \varphi_i$
+  \end{tabular}}
+  \label{table:classifying-topos}
 \end{table}

### 2020-04-06 paper-herrsching2019/main.tex
Barr's theorem and more

@@
 % NOTE TO EDITORS AND LAYOUTERS: This package increases the quality of the
-% typesetting. I personally recommebd using it, but my contribution does not
+% typesetting. I personally recommend using it, but my contribution does not
 % depend on it.
@@
 content: There is a contravariant equivalence between compact completely
-regular locales and commutative unital~$C^\star$-algebras.
-% XXX Barr
+regular locales and commutative
+unital~$C^\star$-algebras~\cite{banaschewski-mulvey:gelfand,coquand-spitters:gelfand,henry:gelfand}.
+
+\begin{remark}For a time, we only knew that~\textsc{zfc} proved that there is a
+constructive proof of the localic version of Gelfand duality without directly
+knowing such a constructive proof. This situation arose because the first proof
+of the localic duality~\cite{banaschewski-mulvey:gelfand} employed \emph{Barr's
+theorem}, to be reviewed on page~\pageref{par:barr}.\end{remark}
 
@@
 
-    \textnormal{mixed rules (unnecessary in presence of the double rule for implication)} \\\smallskip
+    \textnormal{mixed rules (unnecessary in presence of the rule for implication)} \\\smallskip
     \vspace{-0.5em}
@@
   \label{table:geometric-logic}
+\end{table}}
+
+{\makeatletter
+\newcommand*{\shifttext}[1]{%
+  \settowidth{\@tempdima}{#1}%
+  \makebox[\@tempdima]{\hspace*{\@tempdima}#1}%
+}
+\makeatother
+\begin{table}[t]
+  \tbl{The rules of infinitary intuitionistic first-order logic}{\begin{minipage}{0.8\textwidth}
+    \centering
+    \medskip
+
+    (structural rules and rules for~${=}\ {\top}\ {\wedge}\ {\bot}\ {\vee}\
+    {\exists}$ as in Table~\ref{table:geometric-logic})
+    \medskip
+
+    \textnormal{rules for infinitary conjunction} \\\smallskip
+    \phantom{a}\hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}\UnaryInfC{$\bigwedge_{i \in I} \varphi_i \seq{\vec x} \varphi_{i_0}$}\DisplayProof\hfill
+    \AxiomC{$\psi \seq{\vec x} \varphi_i \text{ for each~$i \in I$}$}\UnaryInfC{$\psi \seq{\vec x} \bigwedge_{i \in I} \varphi_i$}\DisplayProof
+    \phantom{a}\hfill
+    \bigskip
+
+    \textnormal{double rule for implication} \\\smallskip
+    \phantom{a}\hfill
+    \Axiom$\varphi \wedge \psi\ \fCenter\seq{\vec x} \chi$
+    \doubleLine
+    \UnaryInf$\varphi\ \fCenter\seq{\vec x} \psi \Rightarrow \chi$
+    \DisplayProof
+    \phantom{a}\hfill
+    \bigskip
+
+    \textnormal{double rule for universal quantification} \\\smallskip
+    \phantom{a}\hfill
+    \Axiom$\varphi \ \fCenter\seq{\vec x, y} \psi$
+    \doubleLine
+    \UnaryInf$\varphi\ \fCenter\seq{\vec x} \forall y\?Y\_ \psi$
+    \DisplayProof
+    \makebox[0pt]{\shifttext{\tiny ($y$ not free in~$\varphi$)}}
+    \hfill\phantom{a}
+    \bigskip
+
+  \end{minipage}}
+  \label{table:first-order-logic}
 \end{table}}
@@
 
+By adding the axiom
+\begin{align*}
+  \top &\seq{x:R} (x = 0) \vee (\exists y\?R\_ xy = 1),
+\intertext{we describe \emph{geometric fields}, and by additionally adding for
+each natural number~$n \geq 1$ the axiom}
+  \top &\seq{a_0:R,\ldots,a_{n-1}:R} \exists x\?R\_ x^n + a_{n-1}x^{n-1} +
+  \cdots + a_1x + a_0 = 0,
+\end{align*}
+we describe \emph{algebraically closed geometric fields}. (This field
+condition is called ``geometric'' because it can be put as a geometric sequent.
+Other, constructively not equivalent conditions such as ``any nonzero element
+is invertible'' or ``any nonunit is zero'' are not geometric sequents.)
+
 In a similar vein, there is the geometric theory of monoids, of groups, and so
@@
 \wedge \varphi_{ny} \vdash \top$ and could also be omitted; if~$x \neq y$,
-this axiom reads~$\varphi_{nx} \wedge \varphi_{ny} \vdash \bot$.}
+this axiom reads~$\varphi_{nx} \wedge \varphi_{ny} \vdash \bot$. If we work in
+a constructive metatheory, we cannot perform this case distinction.}
 \item For each~$x \in \RR$, the axiom~$\top \vdash \bigvee_{n \in \NN}
@@
 
-\paragraph{Extracting constructive content from classical proofs.} Geometric
-logic strikes a fine balance between being sufficiently expressive in order to
+\paragraph{Extracting constructive content from classical proofs.}\label{par:barr}
+Geometric logic strikes a fine balance between being sufficiently expressive in order to
 have rich applications, but also sufficiently restrictive so that \emph{Barr's
-theorem} holds. This theorem comes in two versions, both concerned with
-eliminating classical reasoning from proofs of geometric sequents:
+theorem} holds. This theorem is concerned with eliminating classical reasoning
+from proofs of geometric sequents, and it comes in a simple yet still useful
+and a more sophisticated version:
 \begin{enumerate}
@@
 \end{enumerate}
-For instance, consider the theory of algebraically closed geometric fields,
-that is the theory of rings extended by the axioms
-\begin{align*}
-  \top &\seq{x:R} (x = 0) \vee (\exists y\?R\_ xy = 1) \\
-  \top &\seq{a_0:R,\ldots,a_{n-1}:R} \exists x\?R\_ x^n + a_{n-1}x^{n-1} +
-  \cdots + a_1x + a_0 = 0.
-\end{align*}
-The \emph{Nullstellensatz} is the statement that for any
+For instance, the \emph{Nullstellensatz} is the statement that for any
 system~$f_1,\ldots,f_n \in K[X_1,\ldots,X_m]$ of polynomials over an
@@
 The simpler version of Barr's theorem is constructive, and although we will not
-use it in the remainder of this text, we take the opportunity to give a
-self-contained proof.
-
-However, the more powerful version is not constructive. This puts us in the
-curious situation that we are promised, unconstructively, that there exist
-certain constructive proofs. The textbook
-argument~\cite[Theorem~7.57]{johnstone:topos-theory} establishing the more
-powerful version of Barr's theorem requires the axiom of choice, but recently
-Michael Rathjen lowered the required metatheory to~\textsc{zf}~\cite[Remark~4.2]{rathjen:barr}.
-
-\begin{theorem}Let~$\TT$ be a geometric theory. Let~$\sigma$ be a geometric
-sequent over the signature of~$\TT$. Then the following statements are
-equivalent.
+use it in the remainder of this text, we use the opportunity to present a
+self-contained proof. The proof describes an explicit method for eliminating
+the law of excluded middle from proofs of geometric sequents.
+
+In contrast, the full version is not constructive.\footnote{%
+The textbook argument~\cite[Theorem~7.57]{johnstone:topos-theory} establishing
+the full version of Barr's theorem requires the axiom of choice. But recently
+Michael Rathjen was able to lower the required metatheory
+to~\textsc{zf}~\cite[Remark~4.2]{rathjen:barr}, by working internally to a
+relative version of Gödel's constructible universe~$L$ over a forcing
+extension, where the Barwise completeness theorem applies.}
+This puts us in the curious situation that we are promised, unconstructively,
+that there exist certain constructive proofs, without being given any indication how
+these proofs could be found or what an upper bound on their length is.
+
+However, knowing that these proofs exist platonically is still worthwhile, as
+we can be assured that looking for these proofs is not bound to fail. Indeed,
+for the example of the Nullstellensatz, a constructive proof is given
+as~\cite[Theorem~9.7]{lombardi-quitte:constructive-algebra}.
+
+Also, if one is interested in constructive proofs not because of their
+algorithmic content or of the mathematical insights they communicate but more
+pragmatically for their applicability in, for instance, sheaf models, not
+having them explicitly available is entirely bearable.
+
+For the precise statement of the simple version of Barr's theorem, we recall
+that the rules of geometric logic have been defined in
+Table~\ref{table:geometric-logic}; that the rules of infinitary intuitionistic
+first-order logic have been defined in Table~\ref{table:first-order-logic}; and
+that infinitary classical logic is infinitary intuitionistic logic extended by
+the law of excluded middle.
+\begin{theorem}[Barr's theorem, simple version]
+\label{thm:baby-barr}
+Let~$\TT$ be a geometric theory.
+Let~$\sigma$ be a geometric sequent over the signature of~$\TT$. Then the
+following statements are equivalent.
 \begin{enumerate}
 \item There is a~$\TT$-derivation of~$\sigma$ in geometric logic.
-\item There is a~$\TT$-derivation of~$\sigma$ in infinitary intuitionistic logic.
+\item There is a~$\TT$-derivation of~$\sigma$ in infinitary intuitionistic
+logic.
 \item There is a~$\TT$-derivation of~$\sigma$ in infinitary classical logic.
@@
 To show that~(3) implies~(2), we combine the double negation translation
-with Friedman's trick. We give a sketch of the main thrust of the argument,
+with Friedman's ``nontrivial exit continuation'' trick. We give a sketch of the
+main thrust of the argument,
 omitting lengthy verifications. Let~$\sigma \equiv (\alpha \seq{\vec a} \beta)$.
@@
 \top \seq{\vec x} (\nabla\nabla\varphi \Rightarrow \nabla\varphi),\
-\top \seq{\vec x} (\nabla\varphi \wedge \nabla\psi \Leftrightarrow \nabla(\varphi \wedge \psi)).
+\top \seq{\vec x} ((\nabla\varphi \wedge \nabla\psi) \Leftrightarrow \nabla(\varphi \wedge \psi)).
 \]
@@
 Table~\ref{table:nabla-translation}.
-Again, care has to be taken to ensure that there is no variable collision with the
-variables~$\vec x$. The chosen local operator is special in that
-the~$\nabla$-translation of the law of excluded middle,
-\[ \top \seq{\vec x} \nabla(\varphi^\nabla \vee (\varphi^\nabla \Rightarrow \nabla\bot)), \]
-is intuitionistically provable. Straightforward inductions on the structure of
+Lengthy but straightforward computations by induction on the structure of
 formulas or derivations then establish the following properties.
@@
 classical logic, there is a derivation of~$\varphi^\nabla \seq{\vec x}
-\psi^\nabla$ in infinitary intuitionistic logic.
+\psi^\nabla$ in infinitary intuitionistic logic. This property rests on the
+fact that the~$\nabla$-translation of the law of excluded middle,
+\[ \top \seq{\vec x} \nabla(\varphi^\nabla \vee (\varphi^\nabla \Rightarrow \nabla\bot)), \]
+is intuitionistically provable.
 \end{alphlist}
@@
 
-To verify that~(2) implies~(1), we define a semantics by the clauses in
-Table~\ref{table:classifying-topos}. The following properties of the semantics
-can then be established by induction:
+To verify that~(2) implies~(1), we define a semantics~``$\alpha \models
+\varphi$'' for geometric formulas~$\alpha$ and infinitary first-order
+formulas~$\varphi$ by the clauses in Table~\ref{table:classifying-topos}. For
+notational simplicity, we restrict to the propositional case where the set of
+sorts of~$\TT$ is empty.
+The following properties of the semantics can then be established by induction:
 \begin{alphlist}[(d)]
-\item XXX % ...; idea: do only the propositional case
+\item If~$\TT$ proves~$\beta \vdash \alpha$ (in geometric logic), then~$\alpha \models \varphi$ implies~$\beta
+\models \varphi$.
+\item If~$\TT$ proves~$\alpha \dashv\vdash \bigvee_i \beta_i$ (geometrically) and~$\beta_i \models \varphi$ for
+all indices~$i$, then~$\alpha \models \varphi$.
+\item\label{item:baby-geom} If~$\varphi$ is a geometric formula, then~$\alpha
+\models \varphi$ iff~$\TT$ proves~$\alpha \vdash \varphi$ (geometrically).
+\item\label{item:baby-sound} If a first-order sequent~$(\varphi \vdash \psi)$
+is derivable from the axioms of~$\TT$ in infinitary intuitionistic logic,
+then~$\top \models (\varphi \Rightarrow \psi)$.
 \end{alphlist}
+We can then conclude as follows. Assume that~$(\varphi \vdash \psi)$
+is derivable in infinitary intuitionistic logic. By~(\ref{item:baby-sound}), we
+have~$\top \models (\varphi \Rightarrow \psi)$. By the clause for implication
+in Table~\ref{table:classifying-topos}, we have~$\varphi \models \psi$.
+By~(\ref{item:baby-geom}), there is a~$\TT$-derivation of~$\varphi \models
+\psi$ in geometric logic.
 \end{proof}
+
+\begin{remark}The semantics set up in the proof of Theorem~\ref{thm:baby-barr}
+coincides with the sheaf semantics of the \emph{classifying locale} of~$\TT$
+reviewed in Section~\ref{sect:presenting-frames} and
+Section~\ref{sect:sheaf-semantics}. In the nonpropositional case, the semantics
+coincides with the slightly more general topos semantics of the
+\emph{classifying topos} of~$\TT$~\cite[Section~2]{caramello:tst}.\end{remark}
 
@@
     (x = y)^\nabla &\defeqv \nabla(x = y) \\
-    R(y_1,\ldots,y_m)^\nabla &\defeqv \nabla(R(y_1,\ldots,y_m)) \\
-    \top^\nabla &\defeqv \nabla\top \quad \\
+    R(x_1,\ldots,x_n)^\nabla &\defeqv \nabla(R(x_1,\ldots,x_n)) \\
+    \top^\nabla &\defeqv \top \quad \\
     \bot^\nabla &\defeqv \nabla\bot \\
@@
 
-\begin{table}
-  \tbl{A syntactic model for XXX}{\begin{tabular}{@{}l@{\ \ }c@{\ \ }l@{}}
-    $\alpha \models \beta$ &iff& $\TT$ proves~$\alpha \vdash \beta$ \qquad(for~$\beta$ a relation symbol) \\
+\begin{table}[ht]
+  \tbl{The semantics of the \emph{classifying topos of~$\TT$} (propositional case)}{\begin{tabular}{@{}l@{\ \ }c@{\ \ }l@{}}
+    $\alpha \models \beta$ &iff& $\TT$ proves~$\alpha \vdash \beta$ (in geometric logic) \qquad(for~$\beta$ a relation symbol) \\
     $\alpha \models \top$ &iff& true \\
     $\alpha \models \varphi \wedge \psi$ &iff& $\alpha \models \varphi$ and $\alpha \models \psi$ \\
-    $\alpha \models \bigvee_i \varphi_i$ &iff& there are coherent
-    formulas~$\alpha_1,\ldots,\alpha_n$ such that \\ &&\qquad $\TT$ proves~$\alpha \vdash
-    \alpha_1 \vee \cdots \vee \alpha_n$ and such for each~$k \in \{1,\ldots,n\}$,
-    \\ &&\qquad\qquad there is an index~$i \in I$ such that~$\alpha_k \models \varphi_i$
+    $\alpha \models \bigvee_i \varphi_i$ &iff& there is a family~$(\beta_k)_{k \in K}$ of geometric
+    formulas such that \\ &&\qquad $\TT$ proves~$\alpha \dashv\vdash \bigvee_k \beta_k$ and such for each index~$k \in K$,
+    \\ &&\qquad\qquad there is an index~$i \in I$ such that~$\beta_k \models \varphi_i$ \\
+    $\alpha \models (\varphi \Rightarrow \psi)$ &iff& for any geometric
+    formula~$\beta$ such that~$\TT$ proves~$\beta \vdash \alpha$, \\
+    &&\qquad if~$\beta \models \varphi$, then~$\beta \models \psi$
   \end{tabular}}
@@
 The principal properties of the sheaf semantics is that it is monotone, local
-and sound with respect to intuitionistic first-order logic (which we fix for
-concreteness in Table~\ref{table:first-order-logic}):
-
-{\makeatletter
-\newcommand*{\shifttext}[1]{%
-  \settowidth{\@tempdima}{#1}%
-  \makebox[\@tempdima]{\hspace*{\@tempdima}#1}%
-}
-\makeatother
-\begin{table}[t]
-  \tbl{The rules of intuitionistic first-order logic}{\begin{minipage}{0.8\textwidth}
-    \centering
-    \medskip
-
-    (structural rules and rules for~${=}\ {\top}\ {\wedge}\ {\bot}\ {\vee}\
-    {\exists}$ as in Table~\ref{table:geometric-logic})
-    \medskip
-
-    \textnormal{double rule for implication} \\\smallskip
-    \phantom{a}\hfill
-    \Axiom$\varphi \wedge \psi\ \fCenter\seq{\vec x} \chi$
-    \doubleLine
-    \UnaryInf$\varphi\ \fCenter\seq{\vec x} \psi \Rightarrow \chi$
-    \DisplayProof
-    \phantom{a}\hfill
-    \bigskip
-
-    \textnormal{double rule for universal quantification} \\\smallskip
-    \phantom{a}\hfill
-    \Axiom$\varphi \ \fCenter\seq{\vec x, y} \psi$
-    \doubleLine
-    \UnaryInf$\varphi\ \fCenter\seq{\vec x} \forall y\?Y\_ \psi$
-    \DisplayProof
-    \makebox[0pt]{\shifttext{\tiny ($y$ not free in~$\varphi$)}}
-    \hfill\phantom{a}
-    \bigskip
-
-  \end{minipage}}
-  \label{table:first-order-logic}
-\end{table}}
+and sound with respect to (infinitary) intuitionistic first-order logic as
+defined in Table~\ref{table:first-order-logic}:
 
@@
 \begin{example}The sheaf~$\C$ of continuous real-valued functions together with
-appropriate morphisms is a sheaf model of the theory of rings. Any sheaf can be
-regarded as a model of the theory of objects.\end{example}
+appropriate morphisms is a sheaf model of the theory of rings.\end{example}
+
+\begin{example}Any sheaf can be
+regarded as a sheaf model of the theory of objects.\end{example}
 
@@
 
+\begin{exercise}{Decidability of invertibility}
+Show that it is not the case that any element of~$\ZZ^\sim$ is invertible or
+not invertible.\smallskip
+
+{\noindent\scriptsize\emph{Note.} Since~$\ZZ^\sim$ is a (sheaf) model of the
+theory of local rings, this observation shows that the theory of
+local rings does not prove that invertibility is decidable. Using much the same
+technique, this result has also been proven by Thierry
+Coquand~\cite{coquand:local-rings}.
+
+More generally, one can show: A ring~$A$ is of Krull dimension~$\leq n$
+iff~$A^\sim$ is; and a local ring (such as~$A^\sim$) is of Krull
+dimension~$\leq 0$ iff any element is invertible or nilpotent. Hence
+invertibility of elements of~$A^\sim$ is decidable iff~$A$ is of Krull
+dimension~$\leq 0$. In this case, up to a technical condition, the universe of
+sheaves over~$\Spec(A)$ is classical, hence any statement whatsoever is
+decidable. \emph{The ring controls the logic:} Assuming that the metatheory is
+classical and assuming the technical condition, the universe of sheaves
+over~$\Spec(A)$ is governed by classical logic iff~$A^\sim$ is of Krull
+dimension~$\leq 0$.
+
+The for these purposes correct constructive notion of Krull dimension can be
+found in Refs.~\refcite{coquand-ducos-lombardi-quitte:krull-integral,coquand-lombardi-roy:char-krull}.
+It has been used to give a short and constructive proof that~$\dim
+k[X_1,\ldots,X_n] = n$, where~$k$ is a geometric
+field~\cite{coquand-lombardi:krull-dim-polynomial-ring}. Proofs of the stated
+relation between the dimension of~$A$ and the dimension of~$A^\sim$ are in
+Ref.~\cite[Section~3.4]{blechschmidt:phd}.\par}
+\end{exercise}
+
 \begin{exercise}{Triviality of the spectrum}%

### 2020-04-06 paper-herrsching2019/main.tex
Random sequences

@@
 
-\paragraph{Random sequences.} XXX
+\paragraph{Random sequences.} \emph{Cantor space} is the space~$2^\NN$ of
+infinite binary sequences, equipped with the product topology. A subset~$U
+\subseteq 2^\NN$ is open with respect to this topology if and only if for
+every~$\alpha \in U$, there is a number~$n \in \NN$ such that
+\[ \{ \beta \in 2^\NN \,|\, \forall i < n\_ \alpha(i) = \beta(i) \} \subseteq U. \]
+
+The open subsets of~$2^\NN$ are precisely those subsets which encode
+\emph{observable} (or \emph{semidecidable}) properties of binary sequences,
+those properties for which a finite record of digits suffices to verify that a
+sequence has it (but not necessarily to falsify it). For instance, the sets~$V
+= \{ \alpha \in 2^\NN \,|\, \alpha(0) = \alpha(1) = 1 \}$ and~$W = \{
+\alpha \in 2^\NN \,|\, \exists n\in\NN\_ \alpha(n) = 1 \}$ are open and the sets
+\begin{align*}
+  A &\defeq \{ \alpha \in 2^\NN \,|\, \forall n\in\NN\_ \alpha(n) = 0 \} \\
+  B &\defeq \Bigl\{ \alpha \in 2^\NN \,\Big|\, \lim_{n \to \infty} \frac{1}{n} \sum_{i=0}^{n-1} \alpha(i) = \frac{1}{2} \Bigr\}
+\end{align*}
+are not. With respect to the standard Lebesgue measure on~$2^\NN$, the sets~$V$
+and~$W$ have measure~$\frac{1}{4}$ and~$1$, respectively.
+
+Unlike a regular sequence such as~$(0,0,0,\ldots)$, we expect that, if the
+digits of a \emph{random sequence} are revealed step by step, we will
+eventually verify that it is an element of~$W$; and moreover, that it has any
+observable property~$U$ with Lebesgue measure~$1$.
+
+Elevating this expectation to a definition, it seems prudent to declare that a
+sequence is \emph{random} if and only if it is contained in the intersection of
+all open subsets~$U \subseteq 2^\NN$ with Lebesgue measure~$1$. One could then
+endeavor to setup a develop a theory of probability on this notion.
+
+However, the resulting theory will be trivial for the plain reason that
+according to this definition, there are no random sequences: For any
+sequence~$\alpha$, the set~$\{ \beta \in 2^\NN \,|\, \exists n \in \NN\_
+\alpha(n) \neq \beta(n) \}$ is open, has measure~$1$ and does not
+contain~$\alpha$.
+
+Despite this fundamental issue, one can still argue that this approach has
+merit~\cite{simpson:randomness}, and hence it is worthwhile to find a
+mathematical foundation which is capable of formalizing it. Topological
+spaces are too restrictive, but Alex Simpson observed that locales are
+sufficiently flexible: There is a \emph{locale of random sequences}, defined as
+the locale-theoretic intersection of all measure-$1$ opens of~$2^\NN$. This
+locale is nontrivial despite having no points.
 

### 2020-04-06 paper-herrsching2019/main.tex
Taking care with respect to infinite disjunctions

@@
 derivations is inductively generated by the axioms and the rules displayed in
-Table~\ref{table:geometric-logic}.
+Table~\ref{table:geometric-logic}.\footnote{Because of the infinitary
+disjunctions, some care is needed to not ``built in'' the axiom of choice in
+the definition of derivations. More specifically, we allow inhabited sets of
+subdeductions instead of requiring that subdeductions are given by a function.
+Details on the implementation in~\textsc{czf} can be found in
+Ref.~\cite[Definition~5.2]{rathjen:barr}.}
 \end{definition}

### 2020-04-06 paper-herrsching2019/main.tex
Noetherianity

@@
 
-Noetherian conditions are notoriously tricky in constructive mathematics; XXX
+Noetherian conditions are notoriously tricky in constructive algebra, not
+least because of their overuse in classical commutative algebra which render
+attempts to extract constructive content more challenging.
+
 There are several proposals for constructively sensible definition of
@@
 specifically on page~27 of the textbook by Henri Lombardi and Claude
-Quitté~\cite{lombardi-quitte:constructive-algebra}.
-
+Quitté~\cite{lombardi-quitte:constructive-algebra}. To this list of conditions,
+we add the following.
+
+\begin{definition}
+\label{defn:anonymously-noetherian}
 A ring is \emph{anonymously Noetherian} if and only if any of its ideals is
 \notnot finitely generated.
+\end{definition}
+
+% Ex.: Z finitely generated
+
+This definition is at odds with the idea that constructive mathematics should
+be informative, because it only expresses that it is impossible for no finite
+generating family to exist without requiring that such a family should be given
+-- it may remain \emph{anonymous}.\footnote{We borrowed the term ``anonymous''
+from type theory, where it is used with a similar meaning (see for instance
+Ref.~\refcite{kraus-escardo-coquand-altenkirch:anonymous}). However, there is a
+subtle difference: The conjunction of ``there is at most one element~$x$'' and
+``there is anonymously an element~$x$ in the sense of type theory'' implies
+that there actually exists an element~$x$. In contrast, \notnot existence does
+not.}
+
+However, it is still useful in constructive algebra when interpreted by the
+sheaf semantics: If~$A$ is reduced, then~$A^\sim$ is anonymously Noetherian.
+XXX
 
@@
 %that any ideal of~$A$ ``has'' finitely many generators in the same
-%sense.\footnote{We borrowed the term ``anonymous'' from type theory, where it
-%is used with a similar meaning (see for
-%instance~\cite{kraus-escardo-coquand-altenkirch:anonymous}). However, there is
-%a subtle difference: Unique and anonymous-in-the-sense-of-type-theory existence
-%implies existence, but \notnot existence does not.}
+%sense.\footnote{}
 
@@
 
+\begin{exercise}{Noetherian properties of the integers}%
+\begin{enumerate}
+\item Show that the law of excluded middle holds if the ring~$\ZZ$ of integers
+is Noetherian in the sense that any ideal is finitely generated.\smallskip
+
+{\scriptsize\emph{Hint.} For a truth value~$\varphi$, consider the ideal~$\{ x
+\in \ZZ \,|\, x = 0 \vee \varphi \}$.\par}
+\item Give an example of a space~$X$ and a subsheaf~$F$ of the constant
+sheaf~$\underline{\ZZ}$ such that, under the sheaf semantics, the subsheaf~$F$
+is an ideal, and such that it is not the case that, again under the sheaf
+semantics, the ideal~$F$ is finitely generated.
+\item Show that~$\ZZ$ is anonymously Noetherian in the sense of
+Definition~\ref{defn:anonymously-noetherian}.
+\end{enumerate}
+\end{exercise}
+
 \begin{exercise}{Free fields}

### 2020-04-06 paper-herrsching2019/main.tex
McCoy's theorem

@@
 \newcommand{\Top}{\mathrm{Top}}
+\newcommand{\ann}{\operatorname{ann}}
 \newcommand{\BPIT}{\textsc{bpit}\xspace}
@@
 
-%\begin{proposition}[McCoy's theorem]
-%Let~$M \in A^{n \times m}$ be an injective matrix over a reduced ring~$A$.
-%Then~$(\Lambda^m M)$, the ideal of~$m$-minors of~$M$, is regular.
-%\end{proposition}
-%
-%\begin{proof}Let~$x \in A$ such that~$x \cdot (\Lambda^M) = (0)$ be given. We
-%will show~$x = 0$~in~$A$ by verifying~$x = 0$ in~$A^\sim$ under the sheaf
-%semantics.
-%
-%To this end, it is prudent to introduce a variant of the usual
-%negation~$\neg\varphi \equiv (\varphi \Rightarrow \bot)$, namely~$\negg\varphi
-%\equiv (\varphi \Rightarrow a = 0)$. Our claim is then~$\bott \defeq
-%\negg\top$. As usual, the double negation of the law of excluded middle is
-%available, that is we have
-%\[ \forall x\?A^\sim\_ \negg\negg\bigl((\exists y\?A\_ xy = 1) \vee \negg(\exists
-%y\?A\_ xy = 1)\bigr), \]
-%and as our goal is~$\bott$, we may use with law of excluded middle a finite
-%number of times.
-%
-%Repeatedly exploiting that any matrix element is invertible or
-%$\negg\text{invertible}$ and applying elementary row and column
-%transformations, we may assume that~$M$ is of the
-%form~$\left(\begin{smallmatrix}I&0\\0&N\end{smallmatrix}\right)$ with all
-%elements of~$N$ $\negg\text{invertible}$.
-%
-%By Remark~\ref{XXX},
-%\end{proof}
-% XXX
-
 \begin{proposition}[generic freeness lemma, simple version]
@@
 XXX can be unwound, see ...; give pointers to existing proofs
+
+\paragraph{McCoy's theorem.}\label{par:mccoy}
+It is a familiar fact from linear algebra that an injective~$(n \times
+m)$-matrix~$M$ over a geometric field has at least one invertible~$m$-minor.
+Like many proofs in linear algebra, the proof proceeds by first applying
+elementary row and column transformations to put~$M$ into the form~$N \defeq
+\left(\begin{smallmatrix}I&0\\0&M'\end{smallmatrix}\right)$. This step exploits
+that any element is zero or invertible. All entries in the submatrix~$M'$ are
+zero, and since~$M$ is injective, the number of columns of~$M'$ is zero.
+Hence~$N = \left(\begin{smallmatrix}I\\0\end{smallmatrix}\right)$ has an
+invertible~$m$-minor, and thus~$M$ as well.
+
+A version of \emph{McCoy's theorem} states that this fact
+generalizes to arbitrary rings:
+
+\begin{quote}
+\textbf{Theorem.} Let~$M \in A^{n \times m}$ be an injective matrix.
+Then~$(\Lambda^m M)$, the ideal of~$m$-minors of~$M$, is regular.
+\end{quote}
+
+The sheaf-theoretic reduction technique allows us to give a constructive proof
+of McCoy's theorem which is close to the linear algebra proof. In case that the
+ring~$A$ is reduced, we can even copy the original proof almost word for word;
+for the general case, we require the following lemma.
+
+\begin{quote}
+\textbf{Lemma.} Let~$M \in A^{n \times m}$ be an injective matrix with~$m > 0$. Assume that all entries
+of~$M$ are nilpotent. Then~$1 = 0$ in~$A$.
+
+\textbf{Proof of the lemma.} Let~$(M) \subseteq A$ be the ideal generated by the entries
+of~$M$. The assumption implies that there is a number~$K \in \NN$ such that~$(M)^K = (0)$.
+If~$K \geq 1$, then also~$(M)^{K-1} = (0)$, since for~$x \in (M)^{K-1}$ the
+coefficients of~$M (x,0,\ldots,0)^t$ are elements of~$(M)^K$, hence zero,
+and~$M$ is injective. We may thus assume~$K = 0$. Hence~$1 \in (M)^0 = (0)$. \qed
+\end{quote}
+The proof now proceeds as follows.
+\begin{quote}
+\textbf{Proof of the theorem.} Let~$x \in A$ be such that~$x (\Lambda^m M) = (0)$. We are to
+show~$x = 0$. By passing to the factor ring~$A/\ann(x)$, we may
+assume~$(\Lambda^m M) = (0)$ and~$x = 1$; the new goal is to verify~$1 = 0$. We
+do this by validating~$\bot$ in the sheaf semantics.
+
+We may employ the law of excluded middle a finite number of times because we
+are to verify~$\bot$. By repeatedly exploiting that any matrix entry is
+invertible or not, we may use row and column transformations to put~$M$ into
+the form~$N \defeq \left(\begin{smallmatrix}I&0\\0&M'\end{smallmatrix}\right)$ where all
+entries of~$M'$ are not invertible. By~XXX, the entries of~$M'$ are
+nilpotent, and since~$M'$ is injective and~$1 \neq 0$, the preceding lemma
+shows that the number of columns of~$M'$ is zero.
+
+Hence~$N = \left(\begin{smallmatrix}I\\0\end{smallmatrix}\right)$ and~$(1) =
+(\Lambda^m N) = (\Lambda^m M) = (0)$, a contradiction. \qed
+\end{quote}
+
+\begin{remark}Not of pragmatic, but of logical interest is that the proof can
+also be recast as follows. Instead of reducing to the factor ring~$A/\ann(x)$,
+we continue considering~$A$. Working under the sheaf sematics, we then
+introduce a variant of the usual negation~$\neg\varphi \equiv (\varphi
+\Rightarrow \bot)$, namely~$\negg\varphi \equiv (\varphi \Rightarrow a = 0)$.
+We may still use the law of excluded middle for this new negation and hence
+exploit that any element~$x$ of~$A^\sim$ is invertible or ``not'' invertible. By
+Exercise~\ref{ex:gen-field-property}, this amounts to saying that some power
+of~$x$ annihilates~$a$. The remainder of the proof carries over with minor
+modifications.
+\end{remark}
+
+As with the other examples, the proof can mechanically be unwound to obtain a proof
+which is fully explicit:
+
+\begin{quote}
+\textbf{Proof.} Let~$x \in A$ be such that~$x (\Lambda^m M) = (0)$. We are to
+show~$x = 0$. By passing to the factor ring~$A/\ann(x)$, we may
+assume~$(\Lambda^m M) = (0)$ and~$x = 1$; the new goal is to verify~$1 = 0$.
+
+If~$m = 0$, the claim is immediate because in this case~$(1) = (\Lambda^m M) =
+(0)$. So assume~$m > 0$.
+
+Over the localized ring~$A[M_{ij}^{-1}]$, the matrix entry~$M_{ij}$ is
+invertible and hence~$M$ can be put into the
+form~$\left(\begin{smallmatrix}1&0\\0&M'\end{smallmatrix}\right)$ by elementary
+row and column transformation. The matrix~$M'$ is still injective,
+and~$(\Lambda^{m-1} M') = (\Lambda^m M)$, hence we may assume by induction
+that~$1 = 0$ in~$A[M_{ij}^{-1}]$.
+
+Hence all entries of~$M$ are nilpotent and the claim follows from the preceding
+lemma. \qed
+\end{quote}
+
+Other short and explicit proofs of McCoy's theorem are given in
+Refs.~\refcite{blocki:mccoy},
+\cite[Theorem~5.22]{lombardi-quitte:constructive-algebra} and
+\cite[Exercise~5.23A(3)]{lam:exercises}. The proof in
+Ref.~\cite[Proposition~III.8.2.3]{bourbaki:alg} employs the law of excluded
+middle. We were not able to discern that these proofs are just different
+coatings of the same idea that ours is based on; these proofs all employ
+computations with Cramer-like formulas, while our proof stresses the similarity
+to the field case.
+
+It is on purpose that the (sheaf-theoretic version of) our proof contains case
+distinctions, to closely mimick the proof in the field case. It is also
+possible to do without, as already our unrolled proof demonstrates. Thierry
+Coquand and Claude Quitté gave a beautiful proof without case distinctions an
+application of a local--global
+principle~\cite[Theorem~2.4]{coquand-quitte:constructive-finite-free-resolutions};
+if desired, their proof can also be cast in the sheaf-theoretic framework
+(Exercise~\ref{ex:local-global}).
 
@@
 \begin{exercise}{A generalization of the field property}%
+\label{ex:gen-field-property}%
 Let~$A$ be a ring. We extend the sheaf semantics to infinitary first-order
@@
 
+\begin{exercise}{A local--global principle}%
+\label{ex:local-global}%
+Let~$A$ be a ring. We introduce an operator~$\nabla$ by
+\[ \nabla\varphi \defeqv \bigl(\forall s\?A^\sim\_ ((\varphi \Rightarrow s = 0)
+\Rightarrow s = 0\bigr).\]
+\begin{alphlist}[(f)]
+\item Verify that~$\nabla$ is a \emph{local operator}, that is verify that
+\[ (\varphi \Rightarrow \nabla\varphi), \quad
+(\nabla\nabla\varphi \Rightarrow \nabla\varphi) \quad\text{and}\quad
+((\nabla\varphi \wedge \nabla\psi) \Leftrightarrow \nabla(\varphi \wedge \psi))
+\]
+hold under the sheaf semantics.
+\item Show that equality is~$\nabla$-stable in that
+\[ 1 \models \forall x\?A^\sim\_ \forall y\?A^\sim\_ \nabla(x = y) \Rightarrow x = y. \]
+
+{\scriptsize\emph{Note.} If~$A$ is reduced, then~$\nabla$ coincides
+with~$\neg\neg$ and the claim reduces to statement~(\ref{item:negneg-stable})
+on page~\pageref{item:negneg-stable}. In general, the operator~$\nabla$ is
+related to the notion of \emph{scheme-theoretic dense open
+subsets}~\cite[Lemma~9.11]{blechschmidt:phd}.\par}
+\item Verify that an element of~$A^\sim$ is regular iff it
+is~$\nabla$-invertible, that is:
+\[ 1 \models \forall f\?A^\sim\_
+  \bigl(
+    (\forall s\?A^\sim\_ fs = 0 \Rightarrow s = 0) \Longleftrightarrow
+    \nabla(\exists g\?A^\sim\_ fg = 1)
+  \bigr)
+\]
+\item Verify the analogous statement for finitely generated ideals:
+\begin{multline*}1 \models \forall f_1\?A^\sim\_ \ldots \forall f_n\?A^\sim\_ \\
+  \bigl(
+    (\forall s\?A^\sim\_ (\textstyle\bigwedge_{i=1}^n f_is = 0) \Rightarrow s = 0) \Longleftrightarrow
+    \nabla(1 \in (f_1,\ldots,f_n))
+  \bigr) \end{multline*}
+\item Explain how the local--global principle ``if~$(f_1,\ldots,f_n)$ is a
+regular ideal and if an ideal~$(g_1,\ldots,g_m)$ is regular when considered
+over each localization~$A[f_i^{-1}]$, then~$(g_1,\ldots,g_m)$ is regular
+over~$A$'' can be viewed as an instance of modus ponens for local operators
+($(\nabla\varphi \wedge (\varphi \Rightarrow \nabla\psi)) \Longrightarrow
+\nabla\psi$).
+\item Fill in the details, unwind the following proof of McCoy's theorem
+(page~\pageref{par:mccoy}) and compare the result with the proof in
+Ref.~\cite[Theorem~2.4]{coquand-quitte:constructive-finite-free-resolutions}.
+
+\begin{quote}
+\textbf{Theorem.} Let~$M \in A^{n \times m}$ be an injective matrix.
+Then~$(\Lambda^m M)$, the ideal of~$m$-minors of~$M$, is regular.
+
+\textbf{Proof.} We verify that, under~$\nabla$, there exists an
+invertible~$m$-minor, that is we verify~$\nabla(\bigvee_f (\exists g\?A^\sim\_
+fg=1))$, where the disjunction is over the~$m$-minors of~$M$.
+
+Since~$M$ is injective, the ideal generated by the first column is regular,
+hence under~$\nabla$ the unit ideal. Since our goal is~$\nabla$-stable, we may assume
+that it actually is the unit ideal. Hence one of the entries in the first
+column is invertible. Applying elementary row and column transformations, we
+can put~$M$ into the
+form~$\left(\begin{smallmatrix}1&0\\0&M'\end{smallmatrix}\right)$. Continuing
+in this fashion, we obtain the
+form~$\left(\begin{smallmatrix}I\\0\end{smallmatrix}\right)$. This matrix
+obviously has an invertible~$m$-minor, and hence the original one has one as
+well.
+\end{quote}
+\end{alphlist}
+\end{exercise}
+
 \begin{exercise}{Bridging modules and their induced sheaves}%

### 2020-04-07 paper-herrsching2019/main.tex
Lots of tiny improvements

@@
     \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
-    \draw[red!80, line width=0.4mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
+    \draw[black, line width=0.2mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
   }%
@@
 
-\address{Universität Augsburg \\
-Institut für Mathematik \\
-Universitätsstr. 14 \\
-86159 Augsburg, Germany}
+\address{Università di Verona \\
+Department of Computer Science \\
+Strada le Grazie 15 \\
+37134 Verona, Italy}
 
@@
 the more general metric and topological spaces and also including several more
-and slightly exotic flavors such as the diffeological spaces of differential topology.
+and slightly exotic flavors such as the diffeological spaces from differential topology.
 
@@
 \textbf{Definition.} A \emph{metric space} consists of a set~$X$ of points
-together with a metric~$d : X \times X \to \RR_{\geq 0}$ satisfying the metric
+together with a map~$d : X \times X \to \RR_{\geq 0}$ satisfying the metric
 axioms.
@@
 derived concept. In particular, the opens of a locale are not sets of points;
-in fact, they needn't be sets of anything in peculiar.
+in fact, they need not be sets of anything in peculiar.
 
@@
 of these opens covers~$X$, hence~$X$ is not compact.
-\item For any real~$x$, there is a well-defined continuous function from~$X$ to
+\item Assuming~$x = y \vee \neg(x = y)$ for any pair of reals, there is, for any real~$x$, a well-defined continuous function from~$X$ to
 (the localic version of) the naturals which, on the level of points, would map
@@
 unintuitive statement that a three-dimensional solid ball in~$\RR^3$ of
-radius~$r$ can be partitioned into six disjoint subsets in such a way that
+radius~$r$ can be partitioned into five disjoint subsets in such a way that
 rearranging those subsets using only Euclidean motions yields two disjoint
@@
 The Banach--Tarski paradox can also be avoided by adopting a localic point of
-view: While the localic counterparts of the six pieces do not have any points
+view: While the localic counterparts of the five pieces do not have any points
 in common, the locale-theoretic pairwise intersections are still nontrivial~\cite{simpson:measure}.
-Hence there is no paradox, as one would not expect an rearrangement of
+Hence there is no paradox, as one would not expect a rearrangement of
 overlapping sublocales to preserve volume.
@@
 eventually verify that it is an element of~$W$; and moreover, that it has any
-observable property~$U$ with Lebesgue measure~$1$.
+observable property~$U$ of Lebesgue measure~$1$.
 
@@
 subscribes to the axiom of choice or similar non-constructive principles. In
-these situations, the pointfree approach allowed by locales helps to give
+these situations, the pointfree approach facilitated by locales helps to give
 constructive versions of classical results. We will further discuss this theme
-in Section~\ref{sect:presenting-frames}.
-
-XXX ideal objects
+in Section~\ref{sect:presenting-frames}. Examples for this phenomenon include
+the following.
 
@@
 \paragraph{Galois theory.} Let~$L|k$ be a Galois extension. The
-fundamental theorem of Galois theory states that there is a bijection between
+fundamental theorem of (infinite) Galois theory states that there is a bijection between
 the intermediate extensions~$L|E|k$ and the closed subgroups of the topological
@@
 to the subgroup~$\Gal(L|E)$ and its inverse maps a closed subgroup~$H \subseteq
-\Gal(L|k)$ to the intermediate extension~$L^H$.
+\Gal(L|k)$ to the fixed field~$L^H$.
 
@@
 
-The point~$x$ is completely determined by the
-set~$\{ U \in \O(X) \,|\, \alpha(U) = \top \}$. This set is a \emph{completely
+The point~$x$ is completely determined by the information to which opens it
+belongs, that is by the the set {$\{ U \in \O(X) \,|\, \alpha(U) = \top \}$}. This set is a \emph{completely
 prime filter}, that is a subset~$\fff \subseteq \O(X)$ which is upward-closed,
@@
 \begin{definition}A locale~$X$ is \emph{spatial} if and only if its points
-suffice to distinguish its opens, that is if for any opens~$U, V \in \O(X)$,
+suffice to detect the inclusion relation on the opens of~$X$, that is if for any opens~$U, V \in \O(X)$,
 if~$x \inplus U \Rightarrow x \inplus V$ for all points~$x$ of~$X$, then~$U
@@
 
-An equivalent conditon is: if for any opens~$U, V \in \O(X)$, if~$x \inplus U
+An equivalent definition is: A locale~$X$ is \emph{spatial} if and only if its points
+suffice to distinguish its opens, that is if for any opens~$U, V \in \O(X)$, if~$x \inplus U
 \Leftrightarrow x \inplus V$ for all points~$x$ of~$X$, then~$U = V$.
@@
 have a canonical map~$\Hom_\Top(Y,Y') \to \Hom_\Loc(L(Y),L(Y'))$. If~$Y'$ is
-sober, then this map is a bijection. In particular, in this case the set-theoretic points
+sober, this map is a bijection. In particular, in this case the set-theoretic points
 of~$Y'$ and the locale-theoretic points of~$L(Y')$ are in canonical
@@
 
-\subsection{Exercises}
+\subsection*{Exercises}
+\addcontentsline{toc}{subsection}{Exercises}
 
@@
 
-\begin{exercise}{Embedding sober topological spaces into locales}%
+\begin{exercise}{Sober topological spaces as locales}%
 Verify the statements made in Remark~\ref{rem:sober-embed}.
@@
 
+% FUTURE: Fix linebreak
 \begin{exercise}{Hausdorff spaces are sober}%
@@
 and even metric spaces may fail to be sober. In fact, already the topological
-space of the rational numbers can fail to be
+space of rational numbers can fail to be
 sober~\cite[Example~8.14(iv)]{fourman-scott:sheaves-and-logic}.\par}
@@
 A \emph{Cauchy process} (or \emph{multi-valued modulated Cauchy sequence}) in a metric
-space~$Y$ is a map~$\alpha : \QQ_{>0} \to P(Y)$ such that all sets~$\alpha(q)$
+space~$Y$ is a map~$\alpha : \QQ_{>0} \to P(Y)$ such that all the sets~$\alpha(q)$
 are inhabited and such that~$d(x,x') \leq q + q'$ for all~$q,q' \in \QQ_{>0}$
@@
 Verify, without using the law of excluded middle, that any complete metric
-space is sober.
+space is sober. \smallskip
+
+\noindent{\scriptsize\emph{Note.} In the presense of the countable axiom of choice, any
+Cauchy process can be refined to a Cauchy sequence. In its absence, however, the
+familiar completion construction using equivalence classes of Cauchy sequences
+fails to yield a Cauchy-complete space. In contrast, any Cauchy sequence and
+more generally any Cauchy process converges in the space of equivalence classes
+of Cauchy processes.\par}
 \end{exercise}
@@
 \begin{exercise}{Isomorphisms of frames}%
+\label{ex:isomorphisms-of-frames}%
 Show that a frame homomorphism~$\alpha : A \to A'$ is an isomorphism (that is,
@@
 construction
-\[ \bigwedge_{i \in I} U_i = \bigvee\{ \text{$V \preceq U_i$ for all~$i \in I$}
+\[ \bigwedge_{i \in I} U_i = \bigvee\{ V \in A \,|\, \text{$V \preceq U_i$ for all~$i \in I$}
 \}. \]
@@
 
+\begin{exercise}{Dense opens}%
+\label{ex:dense-opens}%
+An open~$U$ of a locale is \emph{dense} iff for any open~$V$, $U \wedge V =
+\bot$ implies~$U = \bot$. 
+\begin{alphlist}[(b)]
+\item Let~$M$ be a set and endow it with the discrete topology. Show that an
+open~$U \in \O(L(M))$ is dense iff for any element~$x \in M$, $\neg\neg(x \in
+U)$.
+\item Let~$\varphi$ be a truth value, hence an open of the one-point
+locale~$\pt$. Show that~$\varphi$ is dense in~$\pt$ iff~$\neg\neg\varphi$.
+\end{alphlist}
+\end{exercise}
+
 \begin{exercise}{Characterizing spatial locales}%
@@
 
-\begin{exercise}{Coproducts of locales}%
+\begin{exercise}{Coproduct of locales}%
 \begin{alphlist}[(c)]
 \item Show that the category of frames has small products.
-\item Deduce that the category of locales has small coproducts.
+\item Deduce that the category of locales has small coproducts.\smallskip
+
+{\scriptsize\emph{Note.} The category of locales also has small products;
+Exercise~\ref{ex:products-of-locales} is devoted to this fact.\par}
 \item Let~$M$ be a set and endow it with the discrete topology. Show
@@
 \end{alphlist}
-{\scriptsize\emph{Note.} The category of locales also has small products;
-Exercise~\ref{ex:products-of-locales} is devoted to this fact.\par}
 \end{exercise}
@@
 
-\section{Sheaf models}
+\section{Geometric theories and sheaf models}
 
@@
 {\vee}\ {\bigvee}\ {\exists}$ and the relation symbols (but no ${\Rightarrow}\
-{\forall}$). The symbol~``$\bigvee$'' refers to disjunction of arbitrary
+{\forall}$). We follow the usual convention that negation is understood as
+syntactic sugar for implication to~$\bot$. The symbol~``$\bigvee$'' refers to disjunction of arbitrary
 set-indexed families of formulas. The theory \emph{proves} a sequent~$\sigma$ if and
-only if there exists a \emph{derivation} of~$\sigma$, where the set of
+only if there exists a \emph{derivation} of~$\sigma$, where the class of
 derivations is inductively generated by the axioms and the rules displayed in
@@
 
-    (structural rules and rules for~${=}\ {\top}\ {\wedge}\ {\bot}\ {\vee}\
+    (structural rules and rules for~${=}\ {\top}\ {\wedge}\ {\bot}\ {\vee}\ {\bigvee}\
     {\exists}$ as in Table~\ref{table:geometric-logic})
@@
 holds: A formula is derivable iff it holds in all models. In contrast,
-geometric theories can be consistent yet do not admit set-based models. An
+geometric theories can be consistent yet do not admit any set-based models. An
 example is the geometric theory of surjections~$\NN \twoheadrightarrow \RR$
@@
 \[ \top \vdash \bigvee_{n \in \NN_{\geq 1}} (\underbrace{1 + \cdots + 1}_{\text{$n$ summands}} = 0)
-  \qquad
+  \quad\text{or}\quad
   \top \seq{g:G} \bigvee_{n \in \NN} (g^n = e). \]
@@
 Let~$L|k$ be a field
-extension. The classifying locale of automorphisms~$L \to L$ which fix~$k$ has
+extension. The theory of automorphisms~$L \to L$ which fix~$k$ has
 one nullary relation symbol~``$\sigma_{xy}$'' for each pair~$\langle x,y
@@
 Geometric logic strikes a fine balance between being sufficiently expressive in order to
-have rich applications, but also sufficiently restrictive so that \emph{Barr's
+have rich applications and also being sufficiently restrictive so that \emph{Barr's
 theorem} holds. This theorem is concerned with eliminating classical reasoning
@@
 The simpler version of Barr's theorem is constructive, and although we will not
-use it in the remainder of this text, we use the opportunity to present a
+use it in the remainder of this text, we use this opportunity to present a
 self-contained proof. The proof describes an explicit method for eliminating
@@
 relative version of Gödel's constructible universe~$L$ over a forcing
-extension, where the Barwise completeness theorem applies.}
+extension, where the Barwise completeness theorem applies. To our knowledge,
+the precise strength of Barr's theorem relative to~\textsc{czf} or~\textsc{izf}
+has not yet been precisely calibrated. In particular, it is an open
+question~\cite{henry:question-barr} whether Barr's theorem implies the law of
+excluded middle, though it would be a major surprise if it did not.}
 This puts us in the curious situation that we are promised, unconstructively,
@@
 \end{alphlist}
-Combining these ingredients, the given derivation of~$\sigma \equiv (\alpha
+Combining these ingredients, a given derivation of~$\sigma \equiv (\alpha
 \vdash \beta)$ in infinitary classical logic can be transformed to give a derivation of
@@
 formulas~$\varphi$ by the clauses in Table~\ref{table:classifying-topos}. For
-notational simplicity, we restrict to the propositional case where the set of
+notational simplicity, we restrict to the propositional case, that is assume that the set of
 sorts of~$\TT$ is empty.
@@
 set of the formulas over the signature of~$\TT$ modulo~$\TT$-provable
-equivalence, equipped with the ordering~$[\varphi] \preceq [\psi]$ iff~$\TT$
+equivalence, equipped with the ordering given by~$[\varphi] \preceq [\psi]$ iff~$\TT$
 proves~$\varphi \vdash \psi$.
@@
 a set of points, but arbitrary geometric theories can have a proper class of
-set-based models.
+set-based models. (There is, however, a \emph{classifying arithmetic
+space}~\cite{vickers:sketches} of any coherent theory and a \emph{classifying
+topos}~\cite[Section~2]{caramello:tst} of any geometric theory.)
 
@@
 The localic open unit interval is the classifying locale of the
-same theory, but with the additional axiom~$\top \vdash \alpha_0 \wedge \beta_1$.
+same theory but with the additional axiom~$\top \vdash \alpha_0 \wedge \beta_1$.
 \item The \emph{locale of surjections~$\NN \twoheadrightarrow \RR$} is the
@@
 Conversely, any locale~$X$ is the classifying locale of a certain theory, namely
-of the \emph{theory of points of~$X$}. This theory has one nullary relation
+the \emph{theory of points of~$X$}. This theory has one nullary relation
 symbol~$\varphi_U$ for each open~$U \in \O(X)$ (read as ``the point belongs
@@
 spectrum of a ring, the topological Galois group or the topological space of
-reals. In classical mathematics, employing the localic counterparts is merely
+reals. In classical mathematics, employing their localic counterparts is merely
 for convenience or for aesthetic reasons; in constructive mathematics, the
@@
 \subsection{Sheaves on locales}
+\label{sect:sheaves}
 
@@
 
-When first learning about presheaves, one may be intimidated by the vast
-amount of data encoded in a single presheaf (one set for each open). The
+When first learning about presheaves, one may feel intimidated by the vast
+amount of data encoded in a single presheaf (one set for each open, of which there are often uncountably many). The
 situation should be compared to the perhaps more familiar Kripke models for a first-order language~$L$. The
@@
 There is a general construction called \emph{sheafification} which turns
-presheaves into sheaves. Applied to~$\C_\text{c}$, this process yields the
+presheaves into sheaves. Applied to the presheaf~$\C_\text{c}$, this process yields the
 sheaf of \emph{locally} constant real-valued functions. Similarly, the
 sheafification of the presheaf of bounded functions is the sheaf of locally
-bounded functions, and similarly for other properties of functions.
-
+bounded functions, and so on; this is a general principle.
+
+\label{par:constant-sheaf}%
 The presheaf~$\underline{M}^\mathrm{pre}$ associated to a set~$M$ is almost
@@
 to~$M$}\label{page:constant-sheaf}. On an open~$U$, its sections are the locale
-morphisms~$U \to M$, where~$M$ is regarded as the induced locale associated to
+morphisms~$U \to M$, where~$M$ is regarded as the locale induced by
 the discrete topological space~$M$ and~$U$ is regarded as a sublocale of~$X$.
 By the constant sheaf construction, the universe of sets maps to the universe
-of sheaves on~$X$. We will not require much more details on this construction.
+of sheaves on~$X$. We will not require much more details on this construction
+and only note two special cases:
+\begin{enumerate}
+\item If the locale~$X$ is induced by a topological space~$Y$, then
+\[ \underline{M}(U) = \{ f : U \to M \,|\, \text{$f$ is locally constant} \}. \]
+\item If~$X$ is the one-point locale, then~$\underline{M}(U) = M^U$.
+\end{enumerate}
+
+In the other direction, given a point~$x$ of a locale~$X$, any sheaf~$F$ on~$X$ gives
+rise to a set, namely the \emph{stalk of~$F$ at~$x$}. The stalk is the set of
+\emph{germs of sections} of~$F$ at~$x$, that is the set of all sections of~$F$
+on opens~$U$ such that~$x \inplus U$, where two such sections~$\langle U,s
+\rangle$ and~$\langle U',s' \rangle$ are identified iff there is an open~$V
+\preceq U \wedge U'$ such that~$x \inplus V$ and~$s|^U_V = s'|^{U'}_V$.
+
+For instance, the stalk at the origin of the sheaf of holomorphic functions
+on~$\CC$ is in canonical bijection with the set~$\CC\{z\}$ of convergent power
+series, and the stalks of a constant sheaf~$\underline{M}$ are all in canonical
+bijection with~$M$.
 
@@
 
-Let~$X$ be a locale. A \emph{formula over an open~$U \in \O(X)$} is a
-first-order formula (made up using~${=}\ {\top}\ {\wedge}\ {\bot}\ {\vee}\
-{\Rightarrow}\ {\forall}\ {\exists}$) over the signature which has one sort for
+Let~$X$ be a locale. A \emph{formula over an open~$U \in \O(X)$} is an
+infinitary first-order formula (made up using~${=}\ {\top}\ {\wedge}\ \{\bigwedge\}\ {\bot}\ {\vee}\
+{\bigvee}\ {\Rightarrow}\ {\forall}\ {\exists}$) over the signature which has one sort for
 each sheaf~$F$, one constant symbol of sort~$F$ for each section~$s \in F(U)$,
@@
 formulas what it means for a formula~$\varphi$ over an open~$U$ to be
-\emph{forced} or \emph{true over~$U$}, written~``$U \models \varphi$''. The
+\emph{forced by~$U$} or \emph{true over~$U$}, written~``$U \models \varphi$''. The
 main differences with the perhaps more familiar Kripke semantics are marked.
@@
     $U \models \varphi \wedge \psi$ &\Ll& $U \models \varphi$ and $U \models \psi$ \\
+    $U \models \bigwedge_k \varphi_k$ &\Ll& for every index~$k$, $U \models \varphi_k$ \\
     $U \models \bot$ &\Ll& \hcancel{false}{0pt}{3pt}{0pt}{-2pt}\ $U = \bot$ \\
     $U \models \varphi \vee \psi$ &\Ll& \hcancel{$U \models \varphi$ or $U \models \psi$}{0pt}{3pt}{0pt}{-2pt}\ there exists an open covering $U = \bigvee_i U_i$ \\
-    && \quad\quad such that for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$ \\
+    && \quad\quad such that for all indices~$i$: $U_i \models \varphi$ or $U_i \models \psi$ \\
+    $U \models \bigvee_k \varphi_k$ &\Ll& \hcancel{for some index~$k$, $U \models \varphi_k$}{0pt}{3pt}{0pt}{-2pt}\ there exists an open covering $U = \bigvee_i U_i$ \\
+    && \quad\quad such that for each index~$i$, $U_i \models \varphi_k$ for some index~$k$ \\
     $U \models \varphi \Rightarrow \psi$ &\Ll& for all~$V \preceq U$: $V \models \varphi$ implies $V \models \psi$ \\
@@
 induction on the structure of derivations shows that if a sequent~$\varphi
-\seq{x_1{:}X_1,\ldots,x_n{:}X_n} \psi$ is derivable in intuitionistic first-order
+\seq{x_1{:}X_1,\ldots,x_n{:}X_n} \psi$ is derivable in infinitary intuitionistic first-order
 logic, then~$U \models \forall x_1\?X_1\_ \ldots \forall x_n\?X_n\_ (\varphi
@@
 \end{proof}
+
+\begin{remark}The proof of Theorem~\ref{thm:basic-properties-sheaf-semantics}
+is constructive. Hence, by unrolling the definition of the sheaf semantics, any
+proof which uses the sheaf semantics as a simplifying device can mechanically
+be unwound to a fully explicit proof which does not reference sheaves or the
+sheaf semantics.
+
+In general, the resulting proofs will be longer and less perspicious
+than the original proofs, in particular if the original proofs employed nested
+implications (such as double negations). In these cases, there is a certain
+``exotic flavor'' to the unwound proofs, as with tricks with continuations in
+computer science. We display some such proofs which are still readable in
+Section~\ref{sect:example-applications}.
+\end{remark}
 
@@
 
-\begin{example}The sheaf~$\C$ of continuous real-valued functions together with
-appropriate morphisms is a sheaf model of the theory of rings.\end{example}
-
-\begin{example}Any sheaf can be
+\begin{example}The sheaf~$\C$ of continuous real-valued functions on any
+locale~$X$ together with appropriate morphisms is a sheaf model of the theory
+of rings. For instance, the interpretation of the axiom
+\[ \forall f\?\C\_ \forall g\?\C\_ f + g = g + f \]
+with the sheaf semantics amounts to the condition
+\begin{quote}
+For any open~$U$ and any continuous function~$f \in \C(U)$, \\
+${\qquad}$ for any open~$V$ and any continuous function~$g \in \C(V)$, \\
+${\qquad\qquad}$ $f|^U_V + g = g + f|^U_V$.
+\end{quote}
+and is readily seen to be fulfilled.
+\end{example}
+
+\begin{example}Any sheaf on any locale can be
 regarded as a sheaf model of the theory of objects.\end{example}
@@
 \begin{example}\label{ex:sets-as-sheaf-models}
-Any set-based model of a geometric theory~$\TT$ gives rise to a
+Any set-based model~$M$ of a geometric theory~$\TT$ gives rise to a
 sheaf model over the one-point locale, by employing the constant sheaf
-construction reviewed on page~\ref{page:constant-sheaf}. Hence sheaf models are strictly more general than
-set-based models.
+construction reviewed on page~\ref{page:constant-sheaf}. An infinitary
+first-order formula holds for~$M$ iff it holds for the induced sheaf model.
+Hence sheaf models are strictly more general than set-based models.
 \end{example}
@@
 locales~$X \to X'$, where~$X'$ is given as the classifying locale of some
-theory~$\TT$, we only need to specify a~$\TT$-model on~$X$; no separate
+theory~$\TT'$, we only need to specify a~$\TT'$-model~$M'$ on~$X$; no separate
 verification of continuity is required.
 
+If~$X$ classifies a theory~$\TT$, such a specification is usually carried out
+by constructing~$M'$ in terms of the generic model~$U_\TT$.
 For instance, if~$f : A \to B$ is a ring homomorphism, we obtain a locale
@@
 
-XXX ``internalizing parameter-dependence''
-
-\subsection{Exercises}
+\subsection*{Exercises}
+\addcontentsline{toc}{subsection}{Exercises}
 
@@
 \begin{alphlist}[(d)]
-\item Define the locale of injections~$\RR \hookrightarrow \NN$.
 \item Define the locale of partial surjections~$\NN \twoheadrightarrow \NN^\NN$.\smallskip
@@
 function can be imported~\cite[Section~4.3]{scedrov:forcing}.\par}
+\item Define the locale of injections~$\RR \hookrightarrow \NN$.\smallskip
+
+{\scriptsize\emph{Note.} In classical mathematics, this locale does not have
+any points. However there is a variant of the ef{}fective topos based on the
+infinite-time Turing machines by Joel David Hamkins and Andy
+Lewis~\cite{hamkins-lewis:ittm} in which an injection~$\RR \hookrightarrow \NN$
+exists~\cite{bauer:injection}. See
+Ref.~\cite[Section~2.2]{blechschmidt:custom-tailored} for an exposition of the
+proof.\par}
+% FUTURE: Give arXiv reference
+
 \item Define the locale of ``open covers of~$[0,1]$ by intervals with rational
@@
 \label{ex:coherent}%
-An open~$U$ of a locale is \emph{compact} iff, whenever~$U
+An open~$U$ of a locale is \emph{compact} if and only if, whenever~$U
 \preceq \bigvee_{i \in I} V_i$, there is a (Kuratowski-)finite subset~$I'
@@
 
-{\scriptsize\emph{Hint.} For coherent formulas~$\alpha$ and geometric
+{\scriptsize\emph{Hint.} Proceed similarly as in the proof of
+Theorem~\ref{thm:baby-barr}. For coherent formulas~$\alpha$ and geometric
 formulas~$\varphi$, define a semantics by the clauses below. Show by an induction on the
@@
   $\alpha \models \bigvee_i \varphi_i$ &iff& there are coherent
-  formulas~$\alpha_1,\ldots,\alpha_n$ such that \\ &&\qquad $\TT$ proves~$\alpha \vdash
+  formulas~$\alpha_1,\ldots,\alpha_n$ such that \\ &&\qquad $\TT$ proves~$\alpha \dashv\vdash
   \alpha_1 \vee \cdots \vee \alpha_n$ and such for each~$k \in \{1,\ldots,n\}$,
@@
 result implies that propositional coherent theories
-without any set-based models are inconsistent.
+without any set-based models are inconsistent. (This is usually phrased as
+``consistent theories have models''.)
 \end{alphlist}
@@
 \begin{exercise}{A geometric interpretation of double negation}%
-An open~$U$ of a locale is \emph{dense} iff for any open~$V$, $U \wedge V =
-\bot$ implies~$U = \bot$. Let~$\varphi$ be a formula over a locale~$X$.
-\begin{alphlist}[(b)]
+Let~$\varphi$ be a formula over an open~$U$ of a locale~$X$.
+\begin{alphlist}[(c)]
 \item Show that there is a largest open,
@@
 \item Show that~$X \models \neg\neg\varphi$ iff~$\llbracket\varphi\rrbracket$
-is dense.
+is dense in the sense of Exercise~\ref{ex:dense-opens}.
+
+\item Give a condition on a locale or a topological space such that its sheaf
+semantics validates the law of excluded middle (or equivalently the law of
+double negation elimination).
 \end{alphlist}
@@
 However, these reduction techniques typically require \emph{transfinite
-methods}. The displayed proof contains three appeals to the transfinite:
+methods}. The displayed proof appeals to the transfinite four times:
 \begin{enumerate}
@@
 proof could be found. (Still, beautiful constructive proofs are known and
-presented, for instance, in a celebrated short note by Richman on nontrivial
+presented, for instance, in a celebrated short note by Fred Richman on nontrivial
 uses of trivial rings~\cite{richman:trivial-rings} and in the recent textbook by
-Lombardi and Quitté on constructive commutative
+Henri Lombardi and Claude Quitté on constructive commutative
 algebra~\cite{lombardi-quitte:constructive-algebra}.)
+% FUTURE: Find precise reference.
 
@@
 It is then a challenging task in mathematical logic to devise efficient means to
-extract this hidden constructive content in a mostly mechanical manner. We
+extract this obscured constructive content in a mostly mechanical manner. We
 envision to reinterpret any given classical proof employing ideal objects in a
@@
 \end{enumerate}
+
+In some cases, these new techniques XXX
 
@@
 \ppp) \wedge (y \in \ppp) \Rightarrow (x + y \in \ppp)$ and $(x \in
-\ppp) \Rightarrow (rx \in \ppp)$ (for any~$r \in A$). A family~$(x_i)_i$ of
+\ppp) \Rightarrow (xy \in \ppp)$ (for any~$y \in A$). A family~$(x_i)_i$ of
 ring elements generates an ideal, also denoted~``$(x_i)_i$'', consisting of
-(finite) sums of multiples of the~$x_i$.
+(Kuratowski-finite) sums of multiples of the~$x_i$.
 
@@
 are not. The reductive power of harnessing~$A^\sim$ stems from the fact
-that~$A^\sim$ enjoys properties which are not shared by~$A$, its stalks or
-indeed any localizations or quotients.
+that~$A^\sim$ enjoys properties which are not shared by~$A$, by its stalks or
+indeed by any of its localizations or quotients.
 
@@
 transfer our knowledge about~$A$ to~$A^\sim$; while
-statements~(\ref{item:stalks}) requires a syntactic restriction on the property
+statement~(\ref{item:stalks}) requires a syntactic restriction on the property
 under consideration, statement~(\ref{item:loc-stable}) requires a semantic one.
@@
 \paragraph{The field property.}
-The fact that~$A^\sim$ is a field is where~$A^\sim$ derives its main usefulness
+The fact that~$A^\sim$ is a field is where~$A^\sim$ derives its main usefulness as a reduction technique
 from. While statements~(\ref{item:loc-stable}), (\ref{item:stalks}) and
@@
 The field condition displayed in statement~(\ref{item:field}) is not the only
-such condition used in constructive algebra. For instance, a stronger condition
+field condition used in constructive algebra. For instance, a stronger condition
 is ``any element is zero or invertible'' and a further, incomparable, condition
@@
 we add the following.
+% FUTURE: Fix reference "27"
 
@@
 = G/[N,N]$, that is as the quotient by the subgroup generated by the
-commutators. This explicit construction makes it appear that the abelianization
-of a group is specific result in algebra.
-
-We can also cast the contruction in logical language, which will make it
-evident that abelianization is just a special case of a more general procedure.
+commutators. This explicit construction makes it appear that abelianization
+is a specific result in algebra pertaining to groups.
+
+We can also cast the contruction in logical language. This change in perspective makes it evident
+that abelianization is just a special case of a more general procedure.
 The free abelian group over~$G$ can be obtained as the \emph{term model} of the
@@
 particularly perspicious account in the general setting of partial Horn
-theories is due to Erik Palmgren and Steve Vickers~\cite{palmgren-vickers:partial-horn}.
+theories is due to Erik Palmgren and Steve Vickers~\cite[Section~5]{palmgren-vickers:partial-horn}.
 
@@
 The non-existence of free fields is explained by the observation that the
-theory of fields is, for any of its variants,\footnote{By ``theory of fields'',
-we could mean the theory of rings extended by either the coherent axiom~$\top
-\seq{x:R} x = 0 \vee (\exists y\?R\_ xy = 1)$ or by the axiom~$\neg(\exists
-y\?R\_ xy = 1) \seq{x:R} x = 0$ or by the axiom~$\neg(x = 0) \seq{x:R} \exists
-(y\?R\_ xy = 1)$.} not a partial Horn theory.
+theory of fields is, for any of its variants, not a partial Horn theory.
 XXX: Are general conditions known for existence of free models?
@@
 \emph{sheaf model} over the classifying locale of prime filters of~$A$, the
-\emph{spectrum} of~$A$. By localizing at the generic prime filter, we solve
+\emph{spectrum}~$\Spec(A)$ of~$A$. By localizing at the generic prime filter, we solve
 the problem of constructing the free local ring in a generalized sense: \emph{The
@@
 
-\paragraph{Constructing the sheaf~$A^\sim$.} We recall that the given ring~$A$
-has a counterpart in the universe of sheaves over~$\Spec(A)$, namely the
+\paragraph{Constructing the sheaf~$A^\sim$.} We recall from Section~\ref{sect:sheaves} that the given ring~$A$
+has a simple counterpart in the universe of sheaves over~$\Spec(A)$, namely the
 constant sheaf~$\underline{A}$. We construct the sheaf~$A^\sim$ as
@@
 turn this family into the subsheaf of~$\underline{A}$ given by $U \mapsto \{ g
-\in \Hom(U,A) \,|\, \forall x \in A\_ f^{-1}[\{a\}] \preceq [D(a)] \}$. Abusing
+\in \Hom(U,A) \,|\, \forall x \in A\_ g^{-1}[\{a\}] \preceq [D(a)] \}$. Abusing
 notation, we denote this subsheaf also as~``$\fff_0$''. The
@@
 stalks~$A_\fff$ at prime filters. The converse holds if~$\Spec(A)$ is spatial,
-or instance if~\BPIT is available.
+for instance if~\BPIT is available.
 \item[(\ref{item:local})] The ring~$A^\sim$ is local.
@@
 localization at a prime filter.
-\item[(\ref{item:loc-stable})] Any property of~$A$ passes to~$\underline{A}$
-(XXX: ref) and the ring~$A^\sim$ is a localization of~$\underline{A}$.
+\item[(\ref{item:loc-stable})] The assumption on~$A$ implies that~$\Spec(A)$ is
+\emph{overt}. For overt locales, it is a standard fact that any (infinitary)
+first-order property of sets passes to the induced constant sheaves. Hence the
+claim follows by observing that~$A^\sim$ is a localization of~$\underline{A}$.
+% FUTURE: Exercise
 \item[(\ref{item:stalks})] We recall that the points of~$\Spec(A)$ are
-precisely the prime filters of~$A$. The stalk of the constant
+precisely the prime filters of~$A$. The sheaf-theoretic stalk of the constant
 sheaf~$\underline{A}$ at any point is precisely~$A$, and by construction, the
@@
 consisting of a morphism~$f : Y \to X$ of locales together with a
-morphism~$f^\sharp : f^{-1}R \to S$ of sheaves of rings on~$Y$.
+morphism~$f^\sharp : f^{-1}R \to S$ of sheaves of rings on~$Y$. (The
+sheaf~$f^{-1}R$ is the \emph{pullback sheaf} of~$R$ along~$f$.)
 In this way a category of sheaves of rings on arbitrary locales is set up, into
@@
 of~$\Spec(A)$. Since this frame is defined as the Lindenbaum algebra of the
-theory of prime filters of~$A$, we need a thorough grasp of the consequences of
+theory of prime filters of~$A$, we need a thorough grasp of the set of sequents proved by
 this theory.
@@
 
-The Lindenbaum algebra is the free frame generated by the symbols~$D(x)$,
+The Lindenbaum algebra of~$\TT$ is the free frame generated by the symbols~$D(x)$,
 where~$x$ ranges over the elements of~$A$, modulo the axioms for a prime
@@
 The most interesting of these radical inclusions is probably~$\sqrt{(f)} \cap
-\sqrt{(g)} \subseteq \sqrt{(fg)}$, as this illustrates why passing to the
+\sqrt{(g)} \subseteq \sqrt{(fg)}$, as this inclusion illustrates why passing to the
 radical is vital; in general, we do not have~$(f) \cap (g) \subseteq (fg)$.
@@
 The frame homomorphism obtained in this manner is surjective, since a preimage
-for a radical ideal~$\aaa$ is given by~$\bigvee_{f \in \aaa} [D(f)]$. It
+for a radical ideal~$\aaa$ is given by~$\bigvee_{f \in \aaa} [D(f)]$. By
+Exercise~\ref{ex:isomorphisms-of-frames}, it
 remains to show that the homomorphism reflects the ordering.
@@
 that~$\Spec(A)$ is compact, since if~$f \in \sqrt{(g_i)_i}$ then also~$f \in
-\sqrt{(g_{i_1},\ldots,g_{i_n})}$ for suitable indices~$i_1,\ldots,i_n$.
+\sqrt{(g_{i_1},\ldots,g_{i_m})}$ for suitable indices~$i_1,\ldots,i_m$.
 However, the stronger result that~$\Spec(A)$ is
@@
 
-On the basis of~$\Spec(A)$, the opens of the form~$[D(f)]$, we define a partial
+On the basis of~$\Spec(A)$ given by the opens of the form~$[D(f)]$, we define a partial
 presheaf by
@@
 \end{quote}
-The remainder of the argument consists of sheaf-theoretic generalities.
-XXX
+The remainder of the argument is by sheaf-theoretic generalities.
 \end{proof}
@@
 ${\qquad}$ for any element~$g \in A$, \\
-${\qquad\qquad}$ if for any~$h \in A$, \\
+${\qquad\qquad}$ if for any element~$h \in A$, \\
 ${\qquad\qquad\qquad}$ if there is~$(fgh)^n = \sum_{i=1}^m fghp_i$
@@
 partition~$(fgh)^1 = fgh \cdot 1$ with~$m = 1$ and~$p_1 = 1$. For~$y \defeq
-\frac{f}{x'} = \frac{f^2g}{fgh}$, we have~$xy = 1$ in~$A[(fghp_1)^{-1}]$.
+\frac{f}{x'} = \frac{f^2gp_1}{fghp_1}$, we have~$xy = 1$ in~$A[(fghp_1)^{-1}]$.
 By assumption, the element~$fgh$ is nilpotent, hence zero. Thus~$x' = 0$
-in~$A[(fg)^{-1}]$. In particular, we have~$x = 0$ in~$A[(fg)^{-1}]$.
+in~$A[(fg)^{-1}]$. This implies that~$x = 0$ in~$A[(fg)^{-1}]$.
 \end{proof}
@@
 \begin{quote}
-the sheaf~$\J$ is of finite type on~$D(g)$
+the sheaf~$\J$ is of finite type on~$D(fg)$
 \end{quote}
-implies that~$g = 0$,
+implies that~$fg$ is nilpotent,
 \end{quote}
-then~$f = 0$.
+then~$f$ is nilpotent.
 \end{quote}
 
-\subsection{Applications}
+\subsection{Evaluating test cases}
 \label{sect:example-applications}
-XXX better title
-
-\begin{proposition}Let~$M$ be an injective matrix with more columns than rows
+
+\paragraph{Size of injective and surjective matrices.}
+
+\begin{quote}
+\textbf{Theorem.} Let~$M$ be an injective matrix with more columns than rows
 over a reduced ring~$A$. Then~$1 = 0$ in~$A$.
-\end{proposition}
-
-\begin{proof}Since localization is exact, the matrix~$M$ is also injective when
+\end{quote}
+
+\begin{quote}
+\textbf{Proof.} Since localization is exact, the matrix~$M$ is also injective when
 considered as a matrix over~$A^\sim$ (statement (\ref{item:loc-stable})).
@@
 However, since the ring~$A^\sim$ is only a field in the weaker sense that
-nonunits are zero, this theorem does not directly apply. We only have that any
+nonunits are zero, this theorem does not apply directly. We only have that any
 element of~$A^\sim$ is \emph{not not} zero or invertible. We hence employ a
@@
 (and even minimal) logic, we have~$\neg\neg(\varphi \vee \neg\varphi)$;
-$\neg\neg\varphi \wedge (\varphi \Rightarrow \neg\neg\psi) \Longrightarrow
+$\neg\neg\chi \wedge (\chi \Rightarrow \neg\neg\psi) \Longrightarrow
 \neg\neg\psi$; and~$\neg\neg\bot \Leftrightarrow \bot$.}
-In other words, $\Spec(A) \models \bot$. By
-Remark~\ref{rem:consistency}, this amounts to~$1 = 0$ in~$A$.
-\end{proof}
-
-\begin{proposition}[generic freeness lemma, simple version]
-\label{thm:baby-grothendieck}
+In other words, $1 \models \bot$. By
+Remark~\ref{rem:consistency} or Table~\ref{table:algebraic-kripke-joyal}, this amounts to~$1 = 0$ in~$A$.\qed
+\end{quote}
+
+If desired, the proof can be mechanically unwound to a fully explicit proof
+which does not refer to the sheaf semantics. In this case, unwinding yields
+exactly the proof by Fred Richman~\cite[Theorem~2]{richman:trivial-rings} (only
+that he needs one additional paragraph to deal with the possibility that the
+ring is not reduced, which we for conciseness simply assumed):
+
+\begin{quote}
+Over the localized
+ring~$A[M_{ij}^{-1}]$, the matrix element~$M_{ij}$ is invertible and hence~$M$ can be
+put into the form~$\left(\begin{smallmatrix}1&0\\0&M'\end{smallmatrix}\right)$
+by elementary row and column transformations. The matrix~$M'$ is injective and
+has more columns than rows, hence by induction, applied to this submatrix
+over~$A[M_{ij}^{-1}]$, we have~$1 = 0$ in~$A[M_{ij}^{-1}]$. Since~$A$ is
+reduced, this amounts to~$M_{ij} = 0$ in~$A$.
+
+Hence~$M$ is the zero matrix. Hence~$M(1,0,\ldots,0)^t = 0$ and by
+injectivity~$1 = 0$ in~$A$.
+\end{quote}
+
+In exactly the same fashion, the dual statement about surjective matricen can
+be proven.
+
+\paragraph{Grothendieck's generic freeness theorem.}
+
+\begin{quote}
+\textbf{Theorem.}
 Let~$A$ be a reduced ring. Let~$M$ be a finitely generated~$A$-module.
@@
 free~$A[f^{-1}]$-module, then~$1 = 0$ in~$A$.
-\end{proposition}
-
-\begin{proof}The claim amounts to the statement that~$M^\sim$ is \emph{not not}
-free. Since~$A^\sim$ is a field, this follows from
+\end{quote}
+
+\begin{quote}
+\textbf{Proof.} The claim amounts to the statement that~$M^\sim$ is \emph{not not}
+free. Since~$A^\sim$ is a field, this statement follows from
 basic (intuitionistic) linear algebra.\footnote{A standard result of linear
@@
 generated vector spaces are \notnot finite free. The field condition required
-by the proof is exactly the one satisfied by~$A^\sim$.}\end{proof}
-
-\begin{theorem}[generic freeness lemma, general version]
+by the proof is exactly the one satisfied by~$A^\sim$.}\qed\end{quote}
+
+Again, the proof can be unwound if desired:
+
+\begin{quote}\textbf{Proof.}
+We proceed by induction on the length of a given generating family
+of~$M$. Let~$M$ be generated by~$(v_1,\ldots,v_m)$.
+
+We show that the family~$(v_1,\ldots,v_m)$ is linearly independent. Let~$\sum_i
+a_i v_i = 0$. Over~$A[a_i^{-1}]$, the vector~$v_i \in M[a_i^{-1}]$ is a linear
+combination of the other generators. Thus~$M[a_i^{-1}]$ can be generated as
+an~$A[a_i^{-1}]$-module by fewer than~$m$ generators. The induction hypothesis,
+applied to this module over~$A[a_i^{-1}]$, yields that~$1 = 0$ in~$A[a_i^{-1}]$. Since~$A$ is
+reduced, this amounts to~$a_i = 0$.
+
+We finish by using the assumption for~$f = 1$. \qed
+\end{quote}
+
+\begin{quote}
+\textbf{Theorem.}
 Let~$A$ be a reduced ring. Let~$B$ be an~$A$-algebra of finite type. Let~$M$ be
@@
 then~$1 = 0$ in~$A$.
-\end{theorem}
-
-\begin{proof}One can check that the claim is the interpretation by the sheaf
+\end{quote}
+
+\begin{quote}\textbf{Proof.} One can check that the claim is the interpretation by the sheaf
 semantics of the following statement about the sheaves: It is \notnot the case that
@@
 Item~(1) is by basic linear algebra as in the proof of
-Theorem~\ref{thm:baby-grothendieck}, but a bit more involved as we now need to
+simple version of Grothendieck's generic freeness lemma, but a bit more involved as we now need to
 deal with countable generating
@@
 Item~(3) is because finitely generated modules over anonymously Noetherian are
-\notnot finitely presented.
-\end{proof}
-
-XXX can be unwound, see ...; give pointers to existing proofs
+\notnot finitely presented.\qed
+\end{quote}
+
+A variant of this proof where the reference to the Noetherian condition has
+been replaced by an explicit computation has been unwound in
+Ref.~\cite{blechschmidt:generic-freeness}.
 
@@
 
-\begin{exercise}{Nontriviality of~$A^\sim$}%
+\begin{exercise}{Nontriviality of~$\boldsymbol{A^\sim}$}%
 Let~$A$ be a ring. In Proposition~\ref{prop:immediate-consequences}, we gave an
@@
 \begin{exercise}{Local injectivity and surjectivity}%
-Let~$M$ an~$N$ be modules over a ring~$A$. Let~$f : M \to N$ be a linear map.
+Let~$M$ and~$N$ be modules over a ring~$A$. Let~$f : M \to N$ be a linear map.
 \begin{alphlist}[(b)]
@@
 principal ideal (in the sense that there exists a partition~$1 = f_1 + \cdots +
-f_n$ such that, for each index~$i$, the ideal~$\aaa[f_i^{-1}]$ is finitely
-generated in~$A[f_i^{-1}]$).
+f_n$ such that, for each index~$i$, the ideal~$\aaa[f_i^{-1}]$ is a principal
+ideal in~$A[f_i^{-1}]$).
 \begin{alphlist}[(d)]
@@
 \item Let~$A$ be a local ring. Let~$\aaa \subseteq A$ be a finitely generated
-ideal such that~$\aaa^2 = \aaa$. Show that~$\aaa = (0)$ or~$\aaa = (1)$.
+ideal such that~$\aaa^2 = \aaa$. Show that~$\aaa = (0)$ or~$\aaa = (1)$.\smallskip
+
+{\scriptsize\emph{Hint.} Nakayama's lemma.\par}
 \item Let~$A$ be a local ring. Let~$M \in A^{n \times n}$ be an idempotent
@@
 result of part~(b) to~$A^\sim$.
+\item Verify, without using the law of excluded middle or that~$A$ is
+Noetherian, that an~$A$-module~$M$ is finitely generated and projective if and
+only if it is finite locally free.\smallskip
+
+{\scriptsize\emph{Note.} A self-contained solution is given in
+Ref.~\refcite{blechschmidt:kaplansky}.\par}
 \end{alphlist}

### 2020-04-08 paper-herrsching2019/main.tex
Proper introductions

@@
 \usepackage{ws-rv-van}
+
+\usepackage[english]{babel}
 
@@
 
-\usepackage[english]{babel}
+\usepackage{hyperref}
 
@@
 
-\section{Locales}
-
+\paragraph{Locales.}
 % XXX \cite{belanger-marquis:pointless} and Johnstone, vickers:locales-toposes,
@@
 in fact, they need not be sets of anything in peculiar.
+
+Section~\ref{sect:locales} explores the basics of the theory of locales, with a
+focus on examples and issues for constructive mathematics.
+
+\paragraph{Geometric theories and sheaf models.} The most sensible solution to
+the task
+\begin{quote}
+Determine the continuous complex-valued functions~$f$ such that~$f(z)^2 = z$.
+\end{quote}
+is not the single set
+\[ \{ f : \CC \to \CC \,|\, \text{$f$ is continuous and $f(z)^2 = z$ for all~$z \in \CC$} \}. \]
+Rather, since the solvability of this functional equation varies with the
+domain on which~$f$ should be defined (having no global solution defined on all
+of~$\CC$ but two solutions on small disks not containing the origin), we should
+consider the family
+\[ (\{ f : U \to \CC \,|\, \text{$f$ is continuous and $f(z)^2 = z$ for all~$z
+\in \CC$} \})_{U \in \O(\CC)} \]
+of sets as the solution (or perhaps, since no solution was actually given, as
+an apt description) of the task.
+
+If we are building a prime ideal~$\ppp$ step by step, adding elements to it
+when the need arises, the most sensible answer to the question
+\begin{quote}
+Is~$x$ contained in the prime ideal~$\ppp$?
+\end{quote}
+is not a single truth value, but a family of truth values, parametrized by the
+possible stages of construction of~$\ppp$.
+
+Somewhat surprisingly, these two examples can be treated by a single theory,
+the \emph{theory of sheaves}. Sheaves help to organize such families in an
+efficient manner and reify them as single coherent entities; the first family
+is (part of) a \emph{sheaf of sets} and the second family is (part of) a
+\emph{sheaf of truth values}. Sheaves will be the subject of
+Sections~\ref{sect:sheaves}--\ref{sect:sheaf-semantics}.
+
+Sheaves interact particularly well with \emph{geometric theories}, which we
+review in Section~\ref{sect:geometric-theories}. Firstly, sheaves allow us to
+generalize our notion of a \emph{model} of a geometric theory. For instance, a
+set-based model of the geometric theory of rings is just an ordinary ring; but
+now we can also consider sheaves of rings as further models.
+
+Some geometric theories only really unfold in the world of sheaves. For
+instance, the geometric theory of surjections~$\NN \twoheadrightarrow \RR$
+reviewed on page~\pageref{item:theory-of-surjections} does not have any
+set-based models, but it is consistent and it does have nontrivial sheaf
+models. For the purpose of constructive algebra, sheaf models of some geometric
+theories play a vital role, as we will discuss in Section~\ref{sect:appl}.
+
+Secondly, geometric theories allow us to efficiently construct the spaces
+(locales) on which we want to study sheaves. This is because a fundamental
+feature of geometric theories (in contrast to, say, arbitrary first-order
+theories) is that their models naturally organize to form a space, the
+space (locale) of models reviewed in Section~\ref{sect:presenting-frames}.
+
+% A peculiar feature of geometric theories is that (Barr's theorem)
+% FUTURE
+
+\paragraph{Applications in constructive algebra.}
+Commutative algebra abounds with techniques to reduce given situations to
+simpler ones, for instance passing to a quotient or passing to a
+localization. These techniques facilitate short and elegant proofs, such as the
+following. (Section~\ref{sect:algebraic-preliminaries} contains some algebraic
+preliminaries.)
+
+\begin{quote}
+\textbf{Theorem.} Let~$M$ be an injective matrix with more columns than rows
+over a reduced ring~$A$. Then~$1 = 0$ in~$A$.
+
+\textbf{Proof.} \uwave{Assume not.} Then there is a \uwave{minimal
+prime ideal} $\ppp \subseteq A$. Since localization is exact, the matrix~$M$ is also injective when
+considered as a matrix over the stalk~$A_\ppp$. Since~$A_\ppp$ is a
+\uwave{field}, this is a contradiction to basic linear algebra. \qed
+\end{quote}
+
+However, these reduction techniques typically require \emph{transfinite
+methods}. The displayed proof appeals to the transfinite four times:
+\begin{enumerate}
+\item The proof proceeds by contradiction, hence requires the law of excluded
+middle, the strongest of the omniscience principles.
+\item The proof then requires a minimal prime ideal. Such an ideal can be
+obtained in two steps: First, the Boolean Prime Ideal
+Theorem~\BPIT is invoked to obtain some prime ideal. Then
+Zorn's lemma fabricates a minimal one.\footnote{Standard textbooks prove the statement ``any nontrivial
+ring has a prime ideal'' by using Zorn's lemma, which
+in the presence of the law of excluded middle is equivalent to the full axiom
+of choice; however, the statement is actually equivalent to the weaker Boolean
+Prime Ideal Theorem~\cite{scott:bpit,banaschewski-harting:lattice-aspects}.
+
+The combined statement ``any ring has a minimal prime ideal'' is equivalent to
+the full axiom of choice. This observation might be folklore; we learned the
+following argument from Will Savin~[XXX: reference]: Let~$(M_i)_{i \in I}$ be a
+disjoint family of inhabited sets. We consider the free ring~$\ZZ[\bigcup_i M_i]$ and
+its ideal~$I$ generated by~$\{ ab \,|\, i \in I, a,b \in M_i, a \neq b \}$. Assume
+that the factor ring~$\ZZ[\bigcup_i M_i]/I$ contains a minimal prime ideal. Its
+preimage~$\ppp \subseteq \ZZ[\bigcup_i M_i]$ is then a prime ideal which is
+minimal among all prime ideals containing~$I$. By primality, for any index~$i$,
+there is at most one element of~$M_i$ not in~$\ppp$. By minimality, for each
+index~$i$, there is exactly one element of~$M_i$ not in~$\ppp$: If there were
+none, then pick an element~$a \in M_i$ XXX}
+% https://mathoverflow.net/questions/98731/minimal-prime-ideals-and-axiom-of-choicerevised-version
+\item Finally, the proof exploits that the stalks of reduced rings at minimal
+prime ideals are fields. This requires two further invokations
+of~\BPIT.\footnote{By~\BPIT, the intersection of all prime ideals of~$A_\ppp$
+is its nilradical. Since~$A_\ppp$ is reduced, the nilradical is the zero ideal.
+Since the prime ideals of~$A_\ppp$ are in bijection with those prime ideals
+of~$A$ which are contained in~$\ppp$, the ring~$A_\ppp$ has exactly one prime
+ideal. By~\BPIT, any noninvertible element of~$A_\ppp$ is contained in some
+prime ideal, hence in the zero ideal. Thus~$A_\ppp$ is a field.}
+\end{enumerate}
+
+This state of affairs is not satisfactory. A statement as simple as the
+displayed example should admit an explicit calculational proof, presenting a
+concrete method for transforming the given conditional equations expressing
+injectivity into the equation~$1 = 0$ without any appeals to the transfinite.
+
+The completeness theorem for coherent logic~\cite[Corollary~D1.5.10]{johnstone:elephant} even
+gives an a priori reason why (for a given matrix size) a finitary proof has to
+exist. However, since the proof of the completeness theorem itself uses~\BPIT
+(and indeed is equivalent to it), it does not give any indication how such a
+proof could be found. (Still, beautiful constructive proofs are known and
+presented, for instance, in a celebrated short note by Fred Richman on nontrivial
+uses of trivial rings~\cite{richman:trivial-rings} and in the recent textbook by
+Henri Lombardi and Claude Quitté on constructive commutative
+algebra~\cite{lombardi-quitte:constructive-algebra}.)
+% FUTURE: Find precise reference.
+
+The key issue with the usual reduction techniques in commutative algebra is
+their dependence on \emph{ideal objects} such as prime or maximal ideals. In
+general, those objects can only be obtained by transfinite methods. However, in
+practice many arguments do not actually require completed ideal objects:
+Their computational core applies just as well to finite approximations of
+these ideal objects.
+
+It is then a challenging task in mathematical logic to devise efficient means to
+extract this obscured constructive content in a mostly mechanical manner. We
+envision to reinterpret any given classical proof employing ideal objects in a
+constructive fashion, realizing ideal objects as convenient fictions.
+
+Locales and sheaf models contribute to this program by providing new reduction
+techniques for commutative algebra. These techniques have a similar effect as the
+classical reduction techniques, but are powered by purely constructive
+underpinnings. In a nutshell, the idea is as follows.
+
+\begin{enumerate}
+\item Instead of replacing a given ring with another (for instance a ring~$A$ with
+one of its stalks~$A_\ppp$ at a minimal prime ideal), we replace a given ring
+with a \emph{sheaf of rings}.
+\item We then maintain the convenient illusion that we are working with a plain old ring
+instead of a sheaf by employing the sheaf semantics.
+\end{enumerate}
+
+In some cases, these new techniques XXX
+
+\section{Locales}
+\label{sect:locales}
 
@@
 \subsection{Geometric theories}
+\label{sect:geometric-theories}
 
@@
 
-\item \emph{The geometric theory of surjections~$f : \NN \twoheadrightarrow
+\item\label{item:theory-of-surjections} \emph{The geometric theory of surjections~$f : \NN \twoheadrightarrow
 \RR$.} This theory has no
@@
 coincides with the slightly more general topos semantics of the
-\emph{classifying topos} of~$\TT$~\cite[Section~2]{caramello:tst}.\end{remark}
+\emph{classifying topos} of~$\TT$~\cite[Section~2]{caramello:tst}. The local
+operator defined in the proof cuts out a certain Boolean subtopos of the
+classifying topos.
+\end{remark}
 
@@
 
-% XXX Introduction...
-
-% XXX A sheaf may well be nontrivial even though it does not have any global
-% sections. This is for instance the case for the subsheaf of the sheaf of continuous
-% complex-valued functions on~$\CC$, given by~$U \mapsto \{ f : U \to \CC \,|\,
-% \text{$f$ is continuous and~$f(z)^2 = z$ for all~$z \in U$} \}$. This sheaf
-% has lots of local sections (local square root maps), but no global one.
-% This phenomenon is related to the failure of (a simple version of) the
-% fundamental theorem of algebra (Exercise~\ref{ex:fta}).
-
+\paragraph{Motivating the sheaf semantics.} Let~$F$ be a sheaf on a locale~$X$.
+What is a sensible reading of the statement~``$\forall s\?F\_ (\varphi(s)
+\Rightarrow \psi(s))$''? A first answer is ``for any global section~$s \in
+F(X)$ such that~$\varphi(s)$,~$\psi(s)$''. However, this proposal does not do
+justice to the sheaf, whose interesting behavior may well only play out with
+its sections on small opens. Indeed, many important sheaves have no or few
+global sections. Hence a more sophisticated answer is ``for any open~$U \in
+\O(X)$ and any section~$s \in F(U)$ with~$\varphi(s)$,~$\psi(s)$'',
+where~``$\varphi(s)$'' and~``$\psi(s)$'' have to be interpreted in a similar
+vein.
+
+Similarly, given a section~$s \in F(U)$, what should~``$\neg\varphi(s)$'' mean?
+Since a section may well validate additional properties when we zoom in to
+smaller opens, the sheaf-theoretically sensible answer is ``except for the
+empty open~$V = \bot$,~$s|^U_V$ does not have property~$\varphi$ for any
+open~$V \preceq U$'', where again~``$\varphi(s)$'' has to be understood in a
+sensible way.
+
+As a final motivating example, consider the statement~``$\exists s\?F\_
+\varphi(s)$''. This should mean that there is an open covering~$X = \bigvee_i
+U_i$ such that on each open~$U_i$, there is a section~$s \in F(U_i)$ such
+that~$\varphi(s)$, properly interpreted. We should not require these individual
+sections to agree on the intersections~$U_i \wedge U_j$ -- by the sheaf
+property, this would amount to saying that there is a global section~$s \in
+F(X)$ with~$\varphi(s)$.
+
+The \emph{sheaf semantics} formalizes the informal language agreement just
+outlined. A marvelous fact with wide-ranging applications is that this
+semantics is sound with respect to first-order intuitionistic logic (and even
+infinitary higher-order intuitionistic logic, and even a suitable version of
+extensional type theory, if the semantics is appropriately extended).
+
+That is, if we have established a property~$\varphi$ and if there is an
+intuitionistic proof that~$\varphi$ entails a further property~$\psi$, then we
+may conclude that also the interpretation of~$\psi$ with the sheaf semantics
+holds.
+
+Hence the sheaf semantics allows us to reason with sheaves in much the same way
+as we reason with sets, in a simple element-based language, and to import much
+of the already-existing results in constructive mathematics into the world of
+sheaves. For instance, any intuitionistic theorem about rings yields a
+corresponding theorem about sheaves of rings.
+
+There is only one caveat to keep in mind: Irrespective of its status in the
+metatheory, the interpretation of the law of excluded middle with the sheaf
+semantics is, for most locales~$X$, false (Example~\ref{ex:negneg}). People who
+are familiar with other flavors of constructive mathematics should also observe
+that the interpretation of any form of the axiom of choice, even countable
+choice or weak countable choice, is false for most locales~$X$. (For discrete
+locales, these axioms pass from the metatheory to the universe of sheaves.)
+
+Incidentally, that the sheaf semantics is not sound with respect to classical
+logic allows us to illustrate the failure of certain classical theorems in
+constructive mathematics by geometric counterexamples and thereby verify their
+unprovability in intuitionistic logic.
+
+We now begin the formal development of the sheaf semantics. More background and
+examples from theoretical computer science, analysis and differential geometry
+are contained in Ref.~\refcite{blechschmidt:custom-tailored}, which should be
+regarded as a companion paper to this contribution.
+
+\paragraph{Definition and basic properties of the sheaf semantics.}
 Let~$X$ be a locale. A \emph{formula over an open~$U \in \O(X)$} is an
@@
 
+\begin{example}\label{ex:negneg}
+For instance, let~$F$ be the sheaf
+\[ U \longmapsto \{ f : U \to \CC \,|\, \text{$f$ is continuous and~$f(z)^2 =
+z$ for all~$z \in U$} \} \]
+on~$X = \CC$. The interpretation of the statement~``$\exists s\?F\_ \top$'' is
+false, as on no open which contains the origin a continuous square root
+function exists), but the interpretation of~``$\neg\neg(\exists s\?F\_ \top)$''
+is true: Unrolling the definition, this statement expresses that
+\begin{quote}
+the only open~$U \preceq X$ which has the property that \\
+${\qquad}$ the only open~$V \preceq U$ which has the property that \\
+${\qquad\qquad}$ locally on~$V$, there exist continuous square root functions \\
+${\qquad}$ is~$V = \bot$ \\
+is~$U = \bot$.
+\end{quote}
+\end{example}
+
 The principal properties of the sheaf semantics is that it is monotone, local
@@
 
+% FUTURE: Add an introductory sentence.
+\paragraph{Sheaf models.}
 \begin{definition}\label{defn:sheaf-model}
@@
 
+For further reference we state a proposition relating truth in a sheaf model
+with truth at all points.
+% FUTURE: Give better motivation
+
+\begin{proposition}Let~$\TT$ be a geometric theory. Let~$\TT_0$ be the same
+theory as~$\TT$ but without any axioms. Let~$M$ be a sheaf model
+of~$\TT_0$ on a locale~$X$ (hence just a \emph{sheaf structure}) and consider
+the structures~$M_x$ obtained by computing the stalks at points~$x \in \Pt(X)$.
+Let~$\sigma$ be a geometric sequent.
+Then:
+\begin{enumerate}
+\item If~$M$ validates~$\sigma$, then so does~$M_x$, for any point~$x \in \Pt(X)$.
+\item The converse holds if~$X$ is spatial.
+\end{enumerate}
+\end{proposition}
+
+\begin{proof}Omitted.\end{proof}
+
+In particular, the stalks of sheaf models of a geometric theory~$\TT$ are
+set-based models of~$\TT$.
+
+\paragraph{The generic model.}
 A geometric theory can well have no set-based models, but might still be
@@
 L(\TT)$.
+
+For a geometric sequent~$\sigma$ over the signature of a geometric
+theory~$\TT$, the following statements are equivalent:
+\begin{enumerate}
+\item The generic model of~$\TT$ validates~$\sigma$.
+\item All sheaf models of~$\TT$ on arbitrary locales validate~$\sigma$.
+\item The theory~$\TT$ proves~$\sigma$:
+\end{enumerate}
+Hence we may understand the phrase~``Let~$M$ be a model of~$\TT$.'' to
+implicitly refer to the generic model. The first-order properties of the
+generic model need not be shared by all sheaf models and need not be provable
+by~$\TT$.
+% FUTURE: Explain more
 
@@
 
+\begin{exercise}{The least number principle}%
+\begin{alphlist}[(e)]
+\item A subset~$M \subseteq X$ of a set~$X$ is \emph{detachable} iff for any
+element~$x \in X$, either~$x \in M$ or~$x \not\in M$. Show that any inhabited
+detachable subset~$M \subseteq \NN$ contains a minimal element.
+\item Show that for any inhabited set~$M$ of natural numbers, it is \notnot the
+case that there is a number~$n \in M$ such that~$n \leq m$ for all~$m \in M$.
+\item Show that for every finitely generated vector space~$V$ over a field in the
+sense that nonunits are zero, it is \notnot the case that~$V$ has a basis.
+\item Verify that the assumption that any inhabited set of natural numbers
+contains a minimal element implies the law of excluded middle.\smallskip
+
+{\scriptsize\emph{Hint.} Consider a set of the form~$\{ n \in \NN \,|\, n = 1
+\vee \varphi \}$.\par}
+\item Give an explicit example of a subsheaf~$F$ of the constant
+sheaf~$\underline{\NN}$ on a topological space~$X$ such that
+\[ \Sh(X) \models \exists n\?F\_ \top
+  \quad\text{but}\quad
+  \Sh(X) \not\models \exists n\?F\_ \forall m\?F\_ n \leq m. \]
+\end{alphlist}
+\end{exercise}
+
+% FUTURE: Failure of IVT
+
 \section{Applications in constructive algebra}
@@
 
-Commutative algebra abounds with techniques to reduce given situations to
-simpler ones, for instance passing to a quotient or passing to a
-localization. These techniques facilitate short and elegant proofs, such as the
-following. (Section~\ref{sect:algebraic-preliminaries} contains some algebraic
-preliminaries.)
-
-\begin{quote}
-\textbf{Theorem.} Let~$M$ be an injective matrix with more columns than rows
-over a reduced ring~$A$. Then~$1 = 0$ in~$A$.
-
-\textbf{Proof.} \uwave{Assume not.} Then there is a \uwave{minimal
-prime ideal} $\ppp \subseteq A$. Since localization is exact, the matrix~$M$ is also injective when
-considered as a matrix over the stalk~$A_\ppp$. Since~$A_\ppp$ is a
-\uwave{field}, this is a contradiction to basic linear algebra. \qed
-\end{quote}
-
-However, these reduction techniques typically require \emph{transfinite
-methods}. The displayed proof appeals to the transfinite four times:
-\begin{enumerate}
-\item The proof proceeds by contradiction, hence requires the law of excluded
-middle, the strongest of the omniscience principles.
-\item The proof then requires a minimal prime ideal. Such an ideal can be
-obtained in two steps: First, the Boolean Prime Ideal
-Theorem~\BPIT is invoked to obtain some prime ideal. Then
-Zorn's lemma fabricates a minimal one.\footnote{Standard textbooks prove the statement ``any nontrivial
-ring has a prime ideal'' by using Zorn's lemma, which
-in the presence of the law of excluded middle is equivalent to the full axiom
-of choice; however, the statement is actually equivalent to the weaker Boolean
-Prime Ideal Theorem~\cite{scott:bpit,banaschewski-harting:lattice-aspects}.
-
-The combined statement ``any ring has a minimal prime ideal'' is equivalent to
-the full axiom of choice. This observation might be folklore; we learned the
-following argument from Will Savin~[XXX: reference]: Let~$(M_i)_{i \in I}$ be a
-disjoint family of inhabited sets. We consider the free ring~$\ZZ[\bigcup_i M_i]$ and
-its ideal~$I$ generated by~$\{ ab \,|\, i \in I, a,b \in M_i, a \neq b \}$. Assume
-that the factor ring~$\ZZ[\bigcup_i M_i]/I$ contains a minimal prime ideal. Its
-preimage~$\ppp \subseteq \ZZ[\bigcup_i M_i]$ is then a prime ideal which is
-minimal among all prime ideals containing~$I$. By primality, for any index~$i$,
-there is at most one element of~$M_i$ not in~$\ppp$. By minimality, for each
-index~$i$, there is exactly one element of~$M_i$ not in~$\ppp$: If there were
-none, then pick an element~$a \in M_i$ XXX}
-% https://mathoverflow.net/questions/98731/minimal-prime-ideals-and-axiom-of-choicerevised-version
-\item Finally, the proof exploits that the stalks of reduced rings at minimal
-prime ideals are fields. This requires two further invokations
-of~\BPIT.\footnote{By~\BPIT, the intersection of all prime ideals of~$A_\ppp$
-is its nilradical. Since~$A_\ppp$ is reduced, the nilradical is the zero ideal.
-Since the prime ideals of~$A_\ppp$ are in bijection with those prime ideals
-of~$A$ which are contained in~$\ppp$, the ring~$A_\ppp$ has exactly one prime
-ideal. By~\BPIT, any noninvertible element of~$A_\ppp$ is contained in some
-prime ideal, hence in the zero ideal. Thus~$A_\ppp$ is a field.}
-\end{enumerate}
-
-This state of affairs is not satisfactory. A statement as simple as the
-displayed example should admit an explicit calculational proof, presenting a
-concrete method for transforming the given conditional equations expressing
-injectivity into the equation~$1 = 0$ without any appeals to the transfinite.
-
-The completeness theorem for coherent logic~\cite[Corollary~D1.5.10]{johnstone:elephant} even
-gives an a priori reason why (for a given matrix size) a finitary proof has to
-exist. However, since the proof of the completeness theorem itself uses~\BPIT
-(and indeed is equivalent to it), it does not give any indication how such a
-proof could be found. (Still, beautiful constructive proofs are known and
-presented, for instance, in a celebrated short note by Fred Richman on nontrivial
-uses of trivial rings~\cite{richman:trivial-rings} and in the recent textbook by
-Henri Lombardi and Claude Quitté on constructive commutative
-algebra~\cite{lombardi-quitte:constructive-algebra}.)
-% FUTURE: Find precise reference.
-
-The key issue with the usual reduction techniques in commutative algebra is
-their dependence on \emph{ideal objects} such as prime or maximal ideals. In
-general, those objects can only be obtained by transfinite methods. However, in
-practice many arguments do not actually require completed ideal objects:
-Their computational core applies just as well to finite approximations of
-these ideal objects.
-
-It is then a challenging task in mathematical logic to devise efficient means to
-extract this obscured constructive content in a mostly mechanical manner. We
-envision to reinterpret any given classical proof employing ideal objects in a
-constructive fashion, realizing ideal objects as convenient fictions.
-
-Locales and sheaf models contribute to this program by providing new reduction
-techniques for commutative algebra. These techniques have a similar effect as the
-classical reduction techniques, but are powered by purely constructive
-underpinnings. In a nutshell, the idea is as follows.
-
+We recall from the introduction one way how locales and sheaves yield new
+reduction techniques for constructive algebra:
 \begin{enumerate}
@@
 \end{enumerate}
-
-In some cases, these new techniques XXX
 
@@
   \top &\vdash x_{g \circ h} = x_g \circ x_h & \text{(for each~$g,h \in G$)} \\
-  \top &\vdash x_g^{-1} = x_{g^{-1}} & \text{(for each~$g \in G$)} \\
+  \top &\vdash x_{g^{-1}} = x_g^{-1} & \text{(for each~$g \in G$)} \\
   \top &\vdash x_e = e
@@
 \end{remark}
+
+\begin{remark}Given a ring~$A$, there is the geometric theory of \emph{local
+localizations} of~$A$. This is the theory of local rings extended by constants~$e_a$
+for each element~$a \in A$ and the following axioms:
+\begin{align*}
+  \top &\vdash e_0 = 0 \\
+  \top &\vdash e_1 = 1 \\
+  \top &\vdash e_{a + b} = e_a + e_b & \text{(for each~$a,b \in A$)} \\
+  \top &\vdash e_{ab} = e_a e_b & \text{(for each~$a,b \in A$)} \\
+  \top &\vdash e_{-a} = -e_a & \text{(for each~$a \in A$)} \\
+  \top &\vdash_{x:R} \bigvee_{a \in A} \bigvee_{s \in A} \bigl((\exists z\?R\_
+  e_s z = 1) \wedge e_s x = e_a\bigr) \\
+  e_a = 0 &\vdash \bigvee_{\substack{s \in A\\sa = 0}} (\exists z\?R\_ e_s z = 1) & \text{(for each~$a \in A$)}
+\end{align*}
+By the first five listed entries, a model~$R$ of this theory comes equipped
+with a ring homomorphism~$f : A \to R$. The remaining two entries ensure that~$R$ is
+a localization of~$A$, more precisely that it is the localization of~$A$
+at~$f^{-1}R^\times$.
+
+Since this theory is not propositional, Section~\ref{sect:presenting-frames}
+does not apply. There is, however, still a classifying topos of this theory, and it turns
+out that it coincides with the topos of sheaves over~$\Spec(A)$ -- mainly
+because the theory of local localizations of~$A$ is equivalent to the theory of
+prime filters of~$A$. Under this identification, the generic model of the local
+localization is precisely the sheaf~$A^\sim$, and it validates first-order
+sequents not expected from a local localization; as repeatedly stressed
+(statement~(\ref{item:field})), it is a field, even though the theory of local
+localizations does not prove the field condition.
+\end{remark}
+% FUTURE: Cite Nullstellensatz paper
 
@@
 
-FUTURE:
-* Why the name "geometric"?
+FUTURE: Why the name "geometric"?
+FUTURE: Forcing

### 2020-04-08 paper-herrsching2019/main.tex
Heine­Borel for uncountable covers

@@
 {\scriptsize\emph{Note.} Exercise~\ref{ex:exotic-locales} lists settings in
-which the assumption in part~(e) is not satisfied.\par}
+which the assumption in part~(e) is not satisfied. The converse in part~(f)
+also holds. Incidentally, the spatiality of the localic real line is an example
+for a statement which requires the full version of the Heine--Borel property,
+not only Heine--Borel for countable covers as it is often studied in reverse
+mathematics~\cite{normann-sanders:uncountable}.\par}
 \end{exercise}

### 2020-04-08 paper-herrsching2019/main.tex
References on locales

@@
 \paragraph{Locales.}
-% XXX \cite{belanger-marquis:pointless} and Johnstone, vickers:locales-toposes,
-% picado-pultr:frames-and-locales
 
@@
 Section~\ref{sect:locales} explores the basics of the theory of locales, with a
-focus on examples and issues for constructive mathematics.
+focus on examples and issues for constructive mathematics. Accessible
+introductions to the theory include
+Refs.~\refcite{johnstone:art,johnstone:point}. The interplay between logic and
+geometry is particularly stressed in
+Refs.~\refcite{vickers:continuity,vickers:locales-toposes}. We recommend
+Ref.~\refcite{belanger-marquis:pointless} for a historical guide and
+Ref.~\refcite{picado-pultr:frames-and-locales} for a comprehensive modern
+textbook on the subject.
 
@@
 
-XXX Cite http://www.mat.uc.pt/~picado/covariant/contents.pdf
-http://www.mat.uc.pt/~picado/publicat/Chapter2.pdf
-
-XXX Cite Frames and Locales: Topology Without Points
-
 X \models phi
@@
 FUTURE: Forcing
+FUTURE: Grothendieck quote
+
+% schuster:formal-zariski
+% cls:spectral-schemes
+really necessary? Wouldn't it suffice to define~$\Spec(A)$ as the topological
+space of filters of~$A$? Indeed, for some time this was
+believed~\cite[Section~3]{lawvere:icm-address}; however, this hope turned out
+certain topos without any filters~\cite[pp.~200f.]{tierney:spectrum}, thus

### 2020-04-08 paper-herrsching2019/main.tex
Minimal prime ideals = AC

@@
 Prime Ideal Theorem~\cite{scott:bpit,banaschewski-harting:lattice-aspects}.
-
 The combined statement ``any ring has a minimal prime ideal'' is equivalent to
-the full axiom of choice. This observation might be folklore; we learned the
-following argument from Will Savin~[XXX: reference]: Let~$(M_i)_{i \in I}$ be a
-disjoint family of inhabited sets. We consider the free ring~$\ZZ[\bigcup_i M_i]$ and
-its ideal~$I$ generated by~$\{ ab \,|\, i \in I, a,b \in M_i, a \neq b \}$. Assume
-that the factor ring~$\ZZ[\bigcup_i M_i]/I$ contains a minimal prime ideal. Its
-preimage~$\ppp \subseteq \ZZ[\bigcup_i M_i]$ is then a prime ideal which is
-minimal among all prime ideals containing~$I$. By primality, for any index~$i$,
-there is at most one element of~$M_i$ not in~$\ppp$. By minimality, for each
-index~$i$, there is exactly one element of~$M_i$ not in~$\ppp$: If there were
-none, then pick an element~$a \in M_i$ XXX}
-% https://mathoverflow.net/questions/98731/minimal-prime-ideals-and-axiom-of-choicerevised-version
+the full axiom of choice~\cite{savin:minimal-prime-ideals}.}
+%FUTURE:
+%This observation might be folklore; we learned the
+%following argument from Will Savin~[XXX: reference]: Let~$(M_i)_{i \in I}$ be a
+%disjoint family of inhabited sets. We consider the free ring~$\ZZ[\bigcup_i M_i]$ and
+%its ideal~$I$ generated by~$\{ ab \,|\, i \in I, a,b \in M_i, a \neq b \}$. Assume
+%that the factor ring~$\ZZ[\bigcup_i M_i]/I$ contains a minimal prime ideal. Its
+%preimage~$\ppp \subseteq \ZZ[\bigcup_i M_i]$ is then a prime ideal which is
+%minimal among all prime ideals containing~$I$. By primality, for any index~$i$,
+%there is at most one element of~$M_i$ not in~$\ppp$. By minimality, for each
+%index~$i$, there is exactly one element of~$M_i$ not in~$\ppp$: If there were
+%none, then pick an element~$a \in M_i$ ...}
 \item Finally, the proof exploits that the stalks of reduced rings at minimal

### 2020-04-08 paper-herrsching2019/main.tex
Bohr topos

@@
 
+The constructive Gelfand correspondence has been used to construct phase spaces
+for quantum-mechanical systems in the Bohr topos approach to quantum
+mechanics~\cite{butterfield-hamilton-isham:bohr,heunen-landsman-spitters:bohr,henry:bohr}.
+Briefly, a quantum mechanical system is described by a~$C^\star$-algebra~$A$
+which is in all interesting cases noncommutative. As such the Gelfand correspondence cannot be
+applied to~$A$. But there is a mirror image of~$A$ as a sheaf model on a
+certain locale, the \emph{Bohr topos} of~$A$, which \emph{is} commutative.
+Since the universe of sheaves supports intuitionistic reasoning, as we will
+review in Section~\ref{sect:sheaf-semantics}, the constructive Gelfand
+correspondence can then be applied to this mirror image.
+
 \begin{remark}For a time, we only knew that~\textsc{zfc} proved that there is a

### 2020-04-08 paper-herrsching2019/main.tex
Notation for Sh(X)
References to cut elimination

@@
 open sets in topology in that arbitrary unions and finite intersections make
-sense; but unlike before, they need not be sets of points.
+sense; but unlike before, they need not be sets of points. Occasionally we
+abuse notation and denote the largest open of a locale~$X$ by~``$X$''.
 
@@
 operator defined in the proof cuts out a certain Boolean subtopos of the
-classifying topos.
+classifying topos. An alternative proof proceeds by cut
+elimination~\cite[Section~6]{rathjen:barr}; see also
+Ref.~\refcite{dyckhoff-negri:geometrisation}, and
+Ref.~\refcite{rathjen-toppel:reduction} for background on proof-theoretical
+reduction, of which Barr's theorem is an example.
 \end{remark}
@@
 The sheaf semantics is also called \emph{Kripke--Joyal semantics}.
+
+For the special case that~$U$ is the largest open of~$X$ we also write~``$\Sh(X)
+\models \varphi$'' or~``$X \models \varphi$''.
 
@@
 
-X \models phi
-
 FUTURE: Why the name "geometric"?

### 2020-04-08 paper-herrsching2019/main.tex
Tackled a couple of XXX's

@@
 %This observation might be folklore; we learned the
-%following argument from Will Savin~[XXX: reference]: Let~$(M_i)_{i \in I}$ be a
+%following argument from Will Savin~[reference]: Let~$(M_i)_{i \in I}$ be a
 %disjoint family of inhabited sets. We consider the free ring~$\ZZ[\bigcup_i M_i]$ and
@@
 
-\begin{proposition}Let~$\TT$ be a geometric theory. Let~$\TT_0$ be the same
+\begin{proposition}\label{prop:at-points}
+Let~$\TT$ be a geometric theory. Let~$\TT_0$ be the same
 theory as~$\TT$ but without any axioms. Let~$M$ be a sheaf model
@@
 
-The non-existence of free fields is explained by the observation that the
+The non-existence of free fields is put into perspective by the observation that the
 theory of fields is, for any of its variants, not a partial Horn theory.
-XXX: Are general conditions known for existence of free models?
+% FUTURE: Are general conditions known for existence of free models?
 
@@
 $A[\fff^{-1}]$, that is, the ring-theoretic stalk of~$A$ at~$\fff$. The claim
-therefore follows from Proposition~XXX.\qedhere
+therefore follows from Proposition~\ref{prop:at-points}.\qedhere
 \end{Alphlist}
@@
 \end{quote}
-The proof now proceeds as follows.
+Because we want to work in the generality of
+a ring which might not be reduced, we cannot use that~$A^\sim$ is a field.
+Instead we will employ the generalization established in
+Exercise~\ref{ex:gen-field-property}: Any nonunit of~$A^\sim$ is nilpotent. The
+proof then proceeds as follows.
 \begin{quote}
@@
 the form~$N \defeq \left(\begin{smallmatrix}I&0\\0&M'\end{smallmatrix}\right)$ where all
-entries of~$M'$ are not invertible. By~XXX, the entries of~$M'$ are
+entries of~$M'$ are not invertible. Since nonunits of~$A^\sim$ are nilpotent, the entries of~$M'$ are
 nilpotent, and since~$M'$ is injective and~$1 \neq 0$, the preceding lemma

### 2020-04-08 paper-herrsching2019/main.tex
Anonymous Noetherianity

@@
 \begin{exercise}{A geometric interpretation of double negation}%
+\label{ex:negneg-interpretation}%
 Let~$\varphi$ be a formula over an open~$U$ of a locale~$X$.
@@
 
-\paragraph{Double negation stability.}
-XXX explain that it is surprising and explain usefulness, then give proof
-
-\begin{proposition}The sheaf of rings~$A^\sim$ has $\neg\neg$-stable equality:
+\paragraph{Double negation stability.} Generally in constructive mathematics,
+two elements being \notnot equal does not imply that they are actually equal:
+\begin{enumerate}
+\item Sheaves give counterexamples: The interpretation of the statement that
+two given sections of a sheaf are \notnot equal with the sheaf semantics is
+that they agree on a dense open (Exercise~\ref{ex:negneg-interpretation}). This
+does not generally imply that they agree on all of their domain.
+\item Realizability gives counterexamples: Realizers of negated statements are
+always uninformative; in the case that a statement~$\varphi$ has a realizer at
+all, the statement~$\neg\neg\varphi$ is realized by any number whatsoever.
+Only in certain cases, where a realizer can be algorithmically reconstructed
+from the mere promise that a realizer exists, is~$\neg\neg\varphi \Rightarrow
+\varphi$ realizable.
+\end{enumerate}
+
+Hence statement~(\ref{item:negneg-stable}) comes at a surprise. It is useful to
+import some of classical logic into the constructive setting of sheaves
+on~$\Spec(A)$. For instance, the law of excluded middle is constructively valid
+in the form
+\[ \neg\neg(\varphi \vee \neg\varphi). \]
+Combined with the observation
+\[ \bigl(\neg\neg\chi \wedge (\chi \Rightarrow \neg\neg\psi)\bigr) \Longrightarrow \neg\neg\psi \]
+we can hence freely use the law of excluded middle in constructive arguments --
+but at the price that the conclusion~$\neg\neg\psi$ is doubly negated, which is
+why typically this observation is not of much use.
+
+However, double negation stability of equality does allow us to conclude~$\psi$
+from the a~priori much weaker premiss~$\neg\neg\psi$, if~$\psi$ is of the
+form~$(s =_{A^\sim} t)$.
+
+This stability is a consequence of the field property. In case that the
+ring~$A$ is not reduced and hence~$A^\sim$ not a field, a substitute is given
+in Exercise~\ref{ex:local-global}. We stress that the proof of
+statement~(\ref{item:negneg-stable}) does not require double negation stability
+of~$A$ on the metalevel; in fact, it is fully constructive.
+
+\begin{proposition}If~$A$ is reduced, then the sheaf of rings~$A^\sim$ has $\neg\neg$-stable equality:
 \[ \Spec(A) \models \forall x,y\?A^\sim\_ \neg\neg(x = y) \Rightarrow x = y. \]
@@
 
-\begin{proof}We argue internally. Let~$x,y \? A^\sim$ be given such
+\begin{proof}We argue internally in the universe of sheaves, using the sheaf
+semantics. Let~$x,y \? A^\sim$ be given such
 that~$\neg\neg(x = y)$. We will verify that~$x - y$ is not invertible; by the
@@
 So assume that~$x - y$ is invertible. Then~$\neg(x - y = 0)$, since if~$x - y =
-0$, then~$1 = (x-y) (x-y)^{-1} = 0$ in contradiction to~$1 \neq 0$. This is a
+0$, then~$1 = (x-y) (x-y)^{-1} = 0$ in contradiction to~$1 \neq 0$. Hence we have a
 contradiction to the assumption~$\neg\neg(x-y=0)$.\end{proof}
@@
 
-% Ex.: Z finitely generated
-
-This definition is at odds with the idea that constructive mathematics should
-be informative, because it only expresses that it is impossible for no finite
+\begin{example}The assumption that any ideal of~$\ZZ$ is finitely generated implies
+the law of excluded middle. But we can prove, constructively, that~$\ZZ$ is
+anonymously Noetherian: Let~$\aaa$ be an ideal of~$\ZZ$ and assume that~$\aaa$
+is not finitely generated. Then~$1 \not\in \aaa$, as else~$\aaa = (1)$ would be
+finitely generated. Also~$2 \not\in \aaa$, as else (noting
+that~$1\not\in\aaa$),~$\aaa = (2)$. Continuing in this manner, we may
+conclude~$n \not\in \aaa$ for every number~$n \geq 1$. Hence~$\aaa = (0)$. But
+this is a contradiction to the assumption that~$\aaa$ is not finitely
+generated.
+\end{example}
+
+Definition~\ref{defn:anonymously-noetherian} is at odds with the idea that constructive mathematics should
+be informative: It only expresses that it is impossible for no finite
 generating family to exist without requiring that such a family should be given
@@
 that there actually exists an element~$x$. In contrast, \notnot existence does
-not.}
-
-However, it is still useful in constructive algebra when interpreted by the
-sheaf semantics: If~$A$ is reduced, then~$A^\sim$ is anonymously Noetherian.
-XXX
-
-%\begin{ex}There is an intuitionistic proof that the ring~$\ZZ$ is anonymously
-%Noetherian: Let~$\aaa \subseteq \ZZ$ be any ideal. Under the assumption that
-%either there exists a nonzero element in~$\aaa$ or not, the ideal~$\aaa$ is
-%\notnot finitely generated, even \notnot principal: For in the first case, a
-%minimal element~$d$ of~$\aaa \cap \NN^+$ (which \notnot exists) witnesses~$\aaa
-%= (d)$. In the second case the ideal~$\aaa$ is the zero ideal. Since the
-%assumption is \notnot satisfied, the ideal~$\aaa$ is \notnot \notnot finitely
-%generated, so \notnot finitely generated. (We remark on this proof scheme on
-%page~\pageref{proof-scheme-boxed-statements}.) \end{ex}
-%
-%It appears that this Noetherian condition has not been studied in the
-%literature on constructive algebra. Indeed, from the philosophical and the
-%computational point of view on constructive mathematics, the notion of an anonymously
-%Noetherian ring is not very useful: From an intuitionistic proof that a given
-%ideal is finitely generated one can mechanically extract explicit generators,
-%thereby satisfying computational or philosophical demands (``the generators are
-%really there''). In contrast, an intuitionistic proof that a given ideal is
-%\notnot finitely generated doesn't contain computational content in general.
-%
-%Along the same lines, one could dismiss the fact (proved below) that Hilbert's
-%basis theorem, stating that~$A[X]$ is Noetherian if~$A$ is, holds for the
-%notion of anonymously Noetherian rings. In fact, one could feel mocked by this
-%version of Hilbert's basis theorem: It promises that any ideal of~$A[X]$
-%``has'' finitely many generators in some Platonic sense (without providing any
-%clue on how one might go on to find the generators -- they remain \emph{anonymous}), provided
-%that any ideal of~$A$ ``has'' finitely many generators in the same
-%sense.\footnote{}
-
-\begin{proposition}The sheaf of rings~$A^\sim$ is anonymously
+not.} And while Hilbert's basis theorem, stating that if a ring~$B$ is
+Noetherian then so is the polynomial ring~$B[X]$, does have a constructive
+proof for the anonymous version of the Noetherian condition,\footnote{For
+instance, a careful reading of the textbook proof given
+in~\cite[Theorem~7.5]{atiyah-macdonald:commutative-algebra} shows that the
+basis theorem holds intuitionistically as stated.} this result seems empty: For
+it just transforms one unconcrete promise (generators of each ideal of~$B$
+exist somewhere, platonically) into another (generators for ideals of~$B[X]$
+exist somewhere, platonically).
+
+Curiously, the anonymous version of the Noetherian condition is still useful in
+constructive algebra when interpreted by the sheaf semantics: This is firstly
+because it just so happens that~$A^\sim$ is anonymously Noetherian if~$A$ is reduced,
+and secondly because the interpretation of double negation with the sheaf
+semantics does have nontrivial content (as evidenced by
+statement~(\ref{item:negneg-stable})).
+
+\begin{proposition}\label{prop:anon-noeth}
+If~$A$ is reduced, then the sheaf of rings~$A^\sim$ is anonymously
 Noetherian.\end{proposition}
 
-\begin{proof}We argue internally. Let~$\aaa \subseteq A^\sim$ be an ideal.
+\begin{proof}We argue internally, under the sheaf semantics. Let~$\aaa \subseteq A^\sim$ be an ideal.
 Assume that~$\aaa$ is not finitely generated. We will verify that then~$\aaa =
@@
 \end{proof}
+
+\begin{remark}The condition for a ring to be (anonymously) Noetherian is a
+higher-order condition and hence out of scope of the first-order version of the
+sheaf semantics presented in Section~\ref{sect:sheaf-semantics}. However, the
+general version of the sheaf semantics is sound with respect to intuitionistic
+higher-order logic, hence the proof of Proposition~\ref{prop:anon-noeth} is
+valid.
+\end{remark}
 
@@
 However, in the form~``$A^\sim$ is a field'', it becomes easily accessible.
-
-This state of affairs is even more pronounced for higher-order formulas such as
-the condition to be anonymously Noetherian. For instance, XXX
-
+This state of affairs is the more pronounced the more negations and especially
+double negations occur in the proof.
+
+%\begin{quote}
+%For any element~$f \in A$ and any (not necessarily quasicoherent) sheaf of
+%ideals~$\J \hookrightarrow A^\sim[X_1,\ldots,X_n]|_{D(f)}$: If
+%\begin{quote}
+%for any element~$g \in A$ the condition that
+%\begin{quote}
+%the sheaf~$\J$ is of finite type on~$D(fg)$
+%\end{quote}
+%implies that~$fg$ is nilpotent,
+%\end{quote}
+%then~$f$ is nilpotent.
+%\end{quote}
+
+\begin{example}For future reference, and also for giving a different
+perspective on statement~(\ref{item:negneg-stable}), we unroll here the
+interpretation of~$f \models \neg\neg\varphi$. It is:
 \begin{quote}
-For any element~$f \in A$ and any (not necessarily quasicoherent) sheaf of
-ideals~$\J \hookrightarrow A^\sim[X_1,\ldots,X_n]|_{D(f)}$: If
-\begin{quote}
-for any element~$g \in A$ the condition that
-\begin{quote}
-the sheaf~$\J$ is of finite type on~$D(fg)$
+For any element~$g \in A$, \\
+${\qquad}$ if for any element~$h \in A$, \\
+${\qquad\qquad}$ if~$fgh \models \varphi$, \\
+${\qquad\qquad}$ then~$fgh$ is nilpotent, \\
+${\qquad}$ then~$fg$ is nilpotent.
 \end{quote}
-implies that~$fg$ is nilpotent,
-\end{quote}
-then~$f$ is nilpotent.
-\end{quote}
+In the case that~$A$ is reduced and~$f$ is the unit element of~$A$, this
+condition implies (by setting~$g \defeq 1$): ``If~$h = 0$ is the only element of~$A$
+such that~$h \models \varphi$, then~$1 = 0$ in~$A$.''
+\end{example}
 

### 2020-04-08 paper-herrsching2019/main.tex
typo
Further field condition

@@
 \begin{exercise}{The least number principle}%
+\label{ex:least-number-principle}%
 \begin{alphlist}[(e)]
@@
 \end{enumerate}
+% FUTURE: A_min
 %\item If a ring element~$x \in A$ is zero in all stalk~$A_\mmm$ at maximal
@@
 The field condition displayed in statement~(\ref{item:field}) is not the only
-field condition used in constructive algebra. For instance, a stronger condition
-is ``any element is zero or invertible'' and a further, incomparable, condition
-is ``any nonzero element is invertible''.
-
-XXX when those other conditions satisfied
+field condition used in constructive algebra. Perhaps the most important such
+condition is that any element is zero or invertible, not least because this
+condition can be expressed as a geometric formula. It is stronger than the
+condition that nonunits are zero and is satisfied by~$A^\sim$ iff~$A$ is reduced
+and of Krull dimension~$\leq 0$.
 
@@
 \paragraph{The Noetherian property.}
-XXX not shared by stalks; ``true constructivists would be offended''
 
@@
 
-\begin{example}The assumption that any ideal of~$\ZZ$ is finitely generated implies
+\begin{example}\label{ex:z-noeth}The assumption that any ideal of~$\ZZ$ is finitely generated implies
 the law of excluded middle. But we can prove, constructively, that~$\ZZ$ is
@@
 semantics does have nontrivial content (as evidenced by
-statement~(\ref{item:negneg-stable})).
+statement~(\ref{item:negneg-stable})). The Noetherian condition is also a
+unique feature of the sheaf model, not shared by the ring~$A$ or its stalks.
 
@@
 been replaced by an explicit computation has been unwound in
-Ref.~\cite{blechschmidt:generic-freeness}.
+Ref.~\refcite{blechschmidt:generic-freeness}.
 
@@
 semantics, the ideal~$F$ is finitely generated.
-\item Show that~$\ZZ$ is anonymously Noetherian in the sense of
-Definition~\ref{defn:anonymously-noetherian}.
+\item Give a proof that~$\ZZ$ is anonymously Noetherian which, unlike the proof
+given in Example~\ref{ex:z-noeth}, uses the anonymous version of the least
+numbr principle of Exercise~\ref{ex:least-number-principle}.
 \end{enumerate}

### 2020-04-08 paper-herrsching2019/main.tex
Two more XXX's

@@
 \begin{abstract}
-XXX
+XXX abstract
 \end{abstract}
@@
 presented, for instance, in a celebrated short note by Fred Richman on nontrivial
-uses of trivial rings~\cite{richman:trivial-rings} and in the recent textbook by
-Henri Lombardi and Claude Quitté on constructive commutative
-algebra~\cite{lombardi-quitte:constructive-algebra}.)
+uses of trivial rings~\cite{richman:trivial-rings} and in the recent
+textbook~\cite{lombardi-quitte:constructive-algebra} by Henri Lombardi and
+Claude Quitté on constructive commutative algebra.)
 % FUTURE: Find precise reference.
@@
 
-In some cases, these new techniques XXX
+In some cases, these new techniques go even beyond their classical role models.
+For instance, while a module need not be free ``on a dense open'' if all its
+stalks at minimal prime ideals are, a module~$M$ is free in that sense if its
+mirror image~$M^\sim$ is \notnot free. This observation is the basis for a
+new proof of Grothendieck's generic freeness lemma, reviewed in
+Section~\ref{sect:example-applications}, which is shorter and arguably more
+perspicious than all previously known classical proofs.
 
@@
 \end{enumerate}
+Barr's theorem has ramifications to constructive algebra, as pointed out by
+Gavin Wraith~\cite{wraith:intuitionistic-algebra}.
 For instance, the \emph{Nullstellensatz} is the statement that for any

### 2020-04-08 paper-herrsching2019/main.tex
Fixed a couple of typos (thanks to aspell)

@@
 \item Finally, the proof exploits that the stalks of reduced rings at minimal
-prime ideals are fields. This requires two further invokations
+prime ideals are fields. This requires two further invocations
 of~\BPIT.\footnote{By~\BPIT, the intersection of all prime ideals of~$A_\ppp$
@@
 Section~\ref{sect:example-applications}, which is shorter and arguably more
-perspicious than all previously known classical proofs.
+perspicuous than all previously known classical proofs.
 
@@
 
-\noindent{\scriptsize\emph{Note.} In the presense of the countable axiom of choice, any
+\noindent{\scriptsize\emph{Note.} In the presence of the countable axiom of choice, any
 Cauchy process can be refined to a Cauchy sequence. In its absence, however, the
@@
 is the Zariski spectrum as familiar from algebraic
-geometry.\footnote{Classicaly, the spectrum is usually defined as the
+geometry.\footnote{Classically, the spectrum is usually defined as the
 topological space of prime ideals instead of prime filters. However,
@@
 
-In general, the resulting proofs will be longer and less perspicious
+In general, the resulting proofs will be longer and less perspicuous
 than the original proofs, in particular if the original proofs employed nested
@@
 
-We can also cast the contruction in logical language. This change in perspective makes it evident
+We can also cast the construction in logical language. This change in perspective makes it evident
 that abelianization is just a special case of a more general procedure.
@@
 Heyting algebra over a partially ordered set. This is all well-understood; a
-particularly perspicious account in the general setting of partial Horn
+particularly perspicuous account in the general setting of partial Horn
 theories is due to Erik Palmgren and Steve Vickers~\cite[Section~5]{palmgren-vickers:partial-horn}.
@@
 
-In exactly the same fashion, the dual statement about surjective matricen can
+In exactly the same fashion, the dual statement about surjective matrices can
 be proven.
@@
 also be recast as follows. Instead of reducing to the factor ring~$A/\ann(x)$,
-we continue considering~$A$. Working under the sheaf sematics, we then
+we continue considering~$A$. Working under the sheaf semantics, we then
 introduce a variant of the usual negation~$\neg\varphi \equiv (\varphi
@@
 It is on purpose that the (sheaf-theoretic version of) our proof contains case
-distinctions, to closely mimick the proof in the field case. It is also
+distinctions, to closely mimic the proof in the field case. It is also
 possible to do without, as already our unrolled proof demonstrates. Thierry
@@
 given in Example~\ref{ex:z-noeth}, uses the anonymous version of the least
-numbr principle of Exercise~\ref{ex:least-number-principle}.
+number principle of Exercise~\ref{ex:least-number-principle}.
 \end{enumerate}

### 2020-04-08 paper-herrsching2019/main.tex
typo

@@
 \paragraph{Compactness of the unit interval.} The unit interval, when realized
-as a topological space, can fail to be be compact in constructive
+as a topological space, can fail to be compact in constructive
 mathematics. For instance, it fails in the Russian school because the Kleene
@@
 The point~$x$ is completely determined by the information to which opens it
-belongs, that is by the the set {$\{ U \in \O(X) \,|\, \alpha(U) = \top \}$}. This set is a \emph{completely
+belongs, that is by the set {$\{ U \in \O(X) \,|\, \alpha(U) = \top \}$}. This set is a \emph{completely
 prime filter}, that is a subset~$\fff \subseteq \O(X)$ which is upward-closed,
@@
 Let~$\RR$ be the localic real line, defined as the classifying locale of the
-theory~$\TT$ of Dedekind cuts as as introduced on
+theory~$\TT$ of Dedekind cuts as introduced on
 page~\pageref{par:dedekind-cuts}. For rational numbers~$x,y$, let~``$\fa x,y
@@
 nonunits are zero, this theorem does not apply directly. We only have that any
-element of~$A^\sim$ is \emph{not not} zero or invertible. We hence employ a
+element of~$A^\sim$ is \notnot zero or invertible. We hence employ a
 standard trick: In intuitionistic proofs of~$\bot$, we may use the law of
@@
 \begin{quote}
-\textbf{Proof.} The claim amounts to the statement that~$M^\sim$ is \emph{not not}
+\textbf{Proof.} The claim amounts to the statement that~$M^\sim$ is \notnot
 free. Since~$A^\sim$ is a field, this statement follows from

### 2020-04-08 paper-herrsching2019/main.tex
Meta-introduction

@@
 \begin{abstract}
-XXX abstract
+XXX This contribution gives an introduction to the theory of locales,
 \end{abstract}
@@
 \tableofcontents
+\bigskip
+
+\noindent
+The purpose of this contribution is to give a coherent account of a particular
+narrative which links locales, geometric theories, sheaf semantics and
+constructive commutative algebra. We are hoping to convey a firm grasp of three
+ideas:
+
+\begin{enumerate}
+\item Locales are a kind of space in which opens instead of points are
+fundamental.
+% Any propositional geometric theory gives rise to a locale of models.
+\item Sheaf semantics allows us to explore mathematical objects from custom-tailored
+mathematical universes.
+\item Without loss of generality, any reduced ring is a field.
+\end{enumerate}
+
+Unlike some other texts on locales or sheaves, this text is set in a
+constructive but impredicative metatheory. We do not use the law of excluded
+middle nor any version of the axiom of choice, and we also do not adopt any
+nonclassical principles. We freely use powersets, but more for linguistic
+convenience than by necessity.
+
+In particular, the central idea presented in
+Section~\ref{sect:appl}, the reduction technique by which we can pretend that a
+given reduced ring is a field, is very robust with respect to the metatheory.
+Even though we chose a semantic point of view for its derivation, the final
+result is a pure syntactical transformation. As it stands, this text could be
+formalized in the kind of type theory which is valid in
+toposes~\cite{maietti:modular-correspondence} or in~\textsc{izf}~\cite{crosilla:cst-izf}.
+
+For readers familiar with the topics of this contribution, we would like to
+make a couple of remarks. We present a self-contained proof of the simple
+version of Barr's theorem as Theorem~\ref{thm:baby-barr}. For reducing
+intuitionistic provability to geometric provability, we employ (what amounts
+to) the syntactic site, similar to the refined completeness result
+by Thierry Coquand~\cite{coquand:site} but closer to the standard description
+of the syntactic site. For reducing classical provability to intuitionistic
+provability, we employ a variant of the double negation translation. We were
+not able to track down a reference to this approach in the context of
+infinitary logic, but the basic idea has long been known and summarized, for
+instance, in Ref.~\refcite{aczel:russell-prawitz}. % XXX first-order completion. We probably should introduce sites but don't
+
+\subsection*{Introduction}
+\addcontentsline{toc}{section}{Introduction}
 
@@
 metatheory, the interpretation of the law of excluded middle with the sheaf
-semantics is, for most locales~$X$, false (Example~\ref{ex:negneg}). People who
-are familiar with other flavors of constructive mathematics should also observe
+semantics is, for most locales~$X$, false (Example~\ref{ex:negneg}).
+Practitioners of other flavors of constructive mathematics should also observe
 that the interpretation of any form of the axiom of choice, even countable
-choice or weak countable choice, is false for most locales~$X$. (For discrete
-locales, these axioms pass from the metatheory to the universe of sheaves.)
+choice or weak countable choice, is false for most locales~$X$. (About the only
+exception are locales induced from discrete topological spaces. For such
+locales, the law of excluded middle and countable, dependent or full choice
+pass from the metatheory to the universe of sheaves.)
 

### 2020-04-08 paper-herrsching2019/main.tex
Remarks for experts

@@
 Unlike some other texts on locales or sheaves, this text is set in a
-constructive but impredicative metatheory. We do not use the law of excluded
+constructive (but impredicative) metatheory. We do not use the law of excluded
 middle nor any version of the axiom of choice, and we also do not adopt any
@@
 In particular, the central idea presented in
-Section~\ref{sect:appl}, the reduction technique by which we can pretend that a
+Section~\ref{sect:appl}, the reduction technique by which we can pretend that any
 given reduced ring is a field, is very robust with respect to the metatheory.
 Even though we chose a semantic point of view for its derivation, the final
-result is a pure syntactical transformation. As it stands, this text could be
+result is a purely syntactical transformation. As it stands, this text could be
 formalized in the kind of type theory which is valid in
@@
 For readers familiar with the topics of this contribution, we would like to
-make a couple of remarks. We present a self-contained proof of the simple
+make a couple of remarks.
+\begin{enumerate}
+\item We present a self-contained proof of the simple
 version of Barr's theorem as Theorem~\ref{thm:baby-barr}. For reducing
@@
 not able to track down a reference to this approach in the context of
-infinitary logic, but the basic idea has long been known and summarized, for
-instance, in Ref.~\refcite{aczel:russell-prawitz}. % XXX first-order completion. We probably should introduce sites but don't
+infinitary logic, but the basic idea has long been known and is summarized, for
+instance, in Ref.~\refcite{aczel:russell-prawitz}.
+\item In Section~\ref{sect:sheaves}, where we introduce sheaves, we should
+probably also introduce sites. They would fit the narrative well and allow for
+the construction of generic models of arbitrary geometric theories and not only
+propositional ones. They would also allow us to cut down on the metatheory,
+sitting comfortably within predicative environments such as arithmetic
+universes~\cite{maietti:au,vickers:sketches} or~\textsc{czf}.
+We do not, however, mainly to not presume
+prerequisites in category theory.
+\item The technique presented in Section~\ref{sect:appl} can be
+regarded as a first-order (and, if desired, higher-order) coating of an
+underlying geometric theory. We propose to take this coating seriously; in
+particular to make good use of surprising nongeometric sequents validated by the
+generic model; and, for ease of use by algebraists, to cast it in semantic terms,
+even though it is all syntax behind the scenes.\footnote{By now it is an established idea in constructive algebra that we should
+consider the theory of prime filters instead of the prime filters themselves,
+hoping that, for instance, a given mathematical proof of the statement~``no
+prime filter contains~$x$'' can also be cast in the
+geometric language of the theory, which then yields the nilpotency of~$x$
+constructively. If the given proof uses only geometric reasoning, this is
+immediate; if it uses intuitionistic reasoning, it can be compiled down (for
+instance precisely by the technique we use); and if it uses classical reasoning, a
+version of Barr's theorem might be applicable. The elusive actual prime filters
+of the ring are replaced with the concrete and purely syntactical generic prime
+filter.
+
+This approach misses that the generic model is interesting on its own and in
+particular that it satisfies peculiar nongeometric sequents such as ``nonunits
+are nilpotent'' and ``any ideal is \notnot finitely generated''. In this sense,
+the generic model is better than the actual models it aims to replace, and we
+believe that potential applications of this insight should be properly
+explored.}
+\end{enumerate}
 

### 2020-04-08 paper-herrsching2019/main.tex
Acknowledgments

@@
 
-For readers familiar with the topics of this contribution, we would like to
+For readers familiar with the themes of this contribution, we would like to
 make a couple of remarks.
@@
 
+\paragraph{Acknowledgments.} We are grateful to Peter Arndt, Andrej Bauer,
+Martin Brandenburg, Thierry Coquand, Martín Escardó, Simon Henry, Matthias
+Hutzler, Milly Maietti, Marc Nieper-Wißkirchen, Alexander Oldenziel, Anja
+Petković, Peter Schuster, Helmut Schwichtenberg, Steve Vickers and Daniel
+Wessel for invaluable discussions shaping this work. We thank the
+organizers Klaus Mainzer, Peter Schuster and Helmut Schwich\-ten\-berg of
+\emph{Proof and Computation 2019}, where this work was presented, for their
+kind invitation and for creating a beautiful and stimulating meeting. We also
+express our thanks to the local coordinator Chuangjie Xu and to all
+participants of that meeting.
+
 \subsection*{Introduction}
@@
 \begin{quote}
-Determine the continuous complex-valued functions~$f$ such that~$f(z)^2 = z$.
+Determine the continuous complex-valued functions~$f$ with~$f(z)^2 = z$.
 \end{quote}

### 2020-04-08 paper-herrsching2019/main.tex
More references

@@
 
-\paragraph{Acknowledgments.} We are grateful to Peter Arndt, Andrej Bauer,
-Martin Brandenburg, Thierry Coquand, Martín Escardó, Simon Henry, Matthias
-Hutzler, Milly Maietti, Marc Nieper-Wißkirchen, Alexander Oldenziel, Anja
-Petković, Peter Schuster, Helmut Schwichtenberg, Steve Vickers and Daniel
-Wessel for invaluable discussions shaping this work. We thank the
-organizers Klaus Mainzer, Peter Schuster and Helmut Schwich\-ten\-berg of
-\emph{Proof and Computation 2019}, where this work was presented, for their
-kind invitation and for creating a beautiful and stimulating meeting. We also
-express our thanks to the local coordinator Chuangjie Xu and to all
-participants of that meeting.
-
 \subsection*{Introduction}
@@
 For many purposes, the theory of prime \emph{filters} is actually more
-relevant. Classically, a prime filter of a ring is simply the complement of a
+relevant.\footnote{This is not to say that all the issues with prime ideals in
+constructive mathematics evaporate when turning to prime filters. For a time it
+was believed that algebraic geometry could be developed constructively if only
+the spectrum of a ring was defined as the topological space of its prime
+filters~\cite[Section~3]{lawvere:icm-address}, but this hope turned out to be
+too naive. André Joyal gave an explicit example of a nontrivial sheaf model of
+the theory of rings without any prime filters~\cite[pp.~200f.]{tierney:spectrum}.}
+Classically, a prime filter of a ring is simply the complement of a
 prime ideal; constructively, it is prudent to turn prime ideals upside down and axiomatize this notion directly.
@@
 are contained in Ref.~\refcite{blechschmidt:custom-tailored}, which should be
-regarded as a companion paper to this contribution.
+regarded as a companion paper to this contribution. We also recommend
+Christopher Mulvey's tutorial~\cite{mulvey:repr}, which culminates in a proof
+of the Serre--Swan theorem using the sheaf semantics. Standard references for
+the sheaf semantics include Refs.
+\cite[Chapter~VI]{moerdijk-maclane:sheaves-logic},
+\cite[Chapter~14]{goldblatt:topoi},
+\cite[Section~5]{caramello:preliminaries},
+\cite[Sections~12--15]{streicher:ctcl},
+\cite[Chapter~6]{borceux:handbook3},
+\cite[Part~D]{johnstone:elephant} and~[\refcite{shulman:categorical-logic}],
+which also gives a detailed account of useful type theory.
 
@@
 
+\section*{Acknowledgments}
+
+We are grateful to Peter Arndt, Andrej Bauer,
+Martin Brandenburg, Thierry Coquand, Martín Escardó, Simon Henry, Matthias
+Hutzler, Milly Maietti, Marc Nieper-Wißkirchen, Alexander Oldenziel, Anja
+Petković, Peter Schuster, Helmut Schwichtenberg, Steve Vickers and Daniel
+Wessel for invaluable discussions shaping this work.
+
+We thank the
+organizers Klaus Mainzer, Peter Schuster and Helmut Schwich\-ten\-berg of
+\emph{Proof and Computation 2019}, where this work was presented, for their
+kind invitation and for creating a beautiful and stimulating meeting. We also
+express our thanks to the local coordinator Chuangjie Xu and to all
+participants of that meeting.
+
 \bibliographystyle{ws-rv-van}

### 2020-04-08 paper-herrsching2019/main.tex
Proper formatting of the case studies

@@
 \paragraph{Applications in constructive algebra.}
+\label{par:appl-constr-alg}
 Commutative algebra abounds with techniques to reduce given situations to
@@
 
-\begin{example}For future reference, and also for giving a different
+\begin{example}\label{ex:negneg-interpretation}
+For future reference, and also for giving a different
 perspective on statement~(\ref{item:negneg-stable}), we unroll here the
@@
 
+On page~\ref{par:appl-constr-alg}, we presented a classical proof of the
+following theorem.
 \begin{quote}
@@
 \end{quote}
+We are now in a position to show how the reduction technique developed in
+Section~\ref{sect:appl} can be used to recast the classical proof so that it
+becomes constructive:
 
@@
 \begin{quote}
-Over the localized
+\textbf{Proof.} Over the localized
 ring~$A[M_{ij}^{-1}]$, the matrix element~$M_{ij}$ is invertible and hence~$M$ can be
@@
 
-Hence~$M$ is the zero matrix. Hence~$M(1,0,\ldots,0)^t = 0$ and by
-injectivity~$1 = 0$ in~$A$.
+Hence~$M$ is the zero matrix. Thus~$M(1,0,\ldots,0)^t = 0$ and by
+injectivity~$1 = 0$ in~$A$. \qed
 \end{quote}
@@
 
-\paragraph{Grothendieck's generic freeness theorem.}
-
+\paragraph{Grothendieck's generic freeness lemma.} This theorem is a basic
+theorem in algebraic geometry, used for instance for developing the theory of
+moduli spaces. When phrased for the affine situation in a constructively
+sensible way, a simple version of this theorem reads as follows.
 \begin{quote}
@@
 \end{quote}
-
+This theorem is particularly interesting for the program of constructivizing
+classical commutative algebra, as it is an example where the constructive proof
+is much shorter and arguably more perspiciuous than classical proofs (for which
+see for instance Refs.~\cite[Lemme~6.9.2]{ega-4-2},
+\cite[Thm.~24.1]{matsumura:commutative-ring-theory} or
+\cite[Thm.~14.4]{eisenbud:commutative-algebra}, which only cover the case of
+Noetherian integral domains, or Refs.~[\refcite{staats:generic-freeness}],
+\cite[{\href{https://stacks.math.columbia.edu/tag/051Q}{Tag~051Q}}]{stacks-project}
+which proceed in a series of intermediate steps, reducing to that case).
 \begin{quote}
-\textbf{Proof.} The claim amounts to the statement that~$M^\sim$ is \notnot
+\textbf{Proof.} By Example~\ref{ex:negneg-interpretation} and
+Exercise~\ref{ex:bridging-modules}(\ref{item:free}), the claim amounts to the statement that~$M^\sim$ is \notnot
 free. Since~$A^\sim$ is a field, this statement follows from
@@
 by the proof is exactly the one satisfied by~$A^\sim$.}\qed\end{quote}
-
-Again, the proof can be unwound if desired:
+Again, the proof can be unwound if
+desired~\cite[Proposition~3]{blechschmidt:generic-freeness}:
 
@@
 \end{quote}
-
+The full version of Grothendieck's generic freeness lemma, and a constructive
+proof of it, read as follows.
 \begin{quote}
@@
 then~$1 = 0$ in~$A$.
-\end{quote}
-
-\begin{quote}\textbf{Proof.} One can check that the claim is the interpretation by the sheaf
+
+\textbf{Proof.} One can check that the claim is the interpretation by the sheaf
 semantics of the following statement about the sheaves: It is \notnot the case that
@@
 \end{quote}
-
 A variant of this proof where the reference to the Noetherian condition has
@@
 \begin{quote}
-\textbf{Proof.} Let~$x \in A$ be such that~$x (\Lambda^m M) = (0)$. We are to
+\textbf{Proof of the theorem.} Let~$x \in A$ be such that~$x (\Lambda^m M) = (0)$. We are to
 show~$x = 0$. By passing to the factor ring~$A/\ann(x)$, we may
@@
 obviously has an invertible~$m$-minor, and hence the original one has one as
-well.
+well.\qed
 \end{quote}
@@
 \item Show that~$M$ is finitely generated iff~$M^\sim$ is.
-\item Show that~$M$ is finite locally free (that is, that there is a
+\item\label{item:free} Show that~$M$ is finite locally free (that is, that there is a
 partition~$1 = f_1 + \cdots + f_n$ such that each module~$M[f_i^{-1}]$ is

### 2020-04-08 paper-herrsching2019/main.tex
typo

@@
 
-\paragraph{Double negation stability.} Generally in constructive mathematics,
+\paragraph{Double negation stability.}\label{par:double-negation-stability}
+Generally in constructive mathematics,
 two elements being \notnot equal does not imply that they are actually equal:
@@
 On page~\ref{par:appl-constr-alg}, we presented a classical proof of the
-following theorem.
+following theorem.\par
 \begin{quote}
@@
 nonunits are zero, this theorem does not apply directly. We only have that any
-element of~$A^\sim$ is \notnot zero or invertible. We hence employ a
-standard trick: In intuitionistic proofs of~$\bot$, we may use the law of
-excluded middle a finite number of times. This is because in intuitionistic
-(and even minimal) logic, we have~$\neg\neg(\varphi \vee \neg\varphi)$;
-$\neg\neg\chi \wedge (\chi \Rightarrow \neg\neg\psi) \Longrightarrow
-\neg\neg\psi$; and~$\neg\neg\bot \Leftrightarrow \bot$.}
+element of~$A^\sim$ is \notnot zero or invertible. We hence employ the
+trick described in the section on double negation stability
+(page~\pageref{par:double-negation-stability}).}
 In other words, $1 \models \bot$. By
@@
 moduli spaces. When phrased for the affine situation in a constructively
-sensible way, a simple version of this theorem reads as follows.
+sensible way, a simple version of this theorem reads as follows.\par
 \begin{quote}
@@
 \cite[{\href{https://stacks.math.columbia.edu/tag/051Q}{Tag~051Q}}]{stacks-project}
-which proceed in a series of intermediate steps, reducing to that case).
+which proceed in a series of intermediate steps, reducing to that case).\par
 \begin{quote}
@@
 The full version of Grothendieck's generic freeness lemma, and a constructive
-proof of it, read as follows.
+proof of it, read as follows.\par
 \begin{quote}
@@
 elementary row and column transformations to put~$M$ into the form~$N \defeq
-\left(\begin{smallmatrix}I&0\\0&M'\end{smallmatrix}\right)$. This step exploits
-that any element is zero or invertible. All entries in the submatrix~$M'$ are
-zero, and since~$M$ is injective, the number of columns of~$M'$ is zero.
+\left(\begin{smallmatrix}I&0\\0&M'\end{smallmatrix}\right)$ where all entries
+in the submatrix~$M'$ are zero. This step exploits
+that any element is zero or invertible. Since~$M$ is injective, the number of columns of~$M'$ is zero.
 Hence~$N = \left(\begin{smallmatrix}I\\0\end{smallmatrix}\right)$ has an
-invertible~$m$-minor, and thus~$M$ as well.
+invertible~$m$-minor, and thus~$M$ has well.
 
@@
 
-\textbf{Proof of the lemma.} Let~$(M) \subseteq A$ be the ideal generated by the entries
-of~$M$. The assumption implies that there is a number~$K \in \NN$ such that~$(M)^K = (0)$.
+\textbf{Proof of the lemma.} Let~$(M)$ be the ideal generated by the entries
+of~$M$. The assumption implies that there is a exponent~$K \in \NN$ such that~$(M)^K = (0)$.
 If~$K \geq 1$, then also~$(M)^{K-1} = (0)$, since for~$x \in (M)^{K-1}$ the
-coefficients of~$M (x,0,\ldots,0)^t$ are elements of~$(M)^K$, hence zero,
-and~$M$ is injective. We may thus assume~$K = 0$. Hence~$1 \in (M)^0 = (0)$. \qed
+entries of~$M (x,0,\ldots,0)^t$ are elements of~$(M)^K$, hence zero,
+and~$M$ is injective. We may thus assume~$K = 0$. So~$1 \in (M)^0 = (0)$. \qed
 \end{quote}
@@
 Exercise~\ref{ex:gen-field-property}: Any nonunit of~$A^\sim$ is nilpotent. The
-proof then proceeds as follows.
+proof then proceeds as follows.\par
 \begin{quote}
@@
 entries of~$M'$ are not invertible. Since nonunits of~$A^\sim$ are nilpotent, the entries of~$M'$ are
-nilpotent, and since~$M'$ is injective and~$1 \neq 0$, the preceding lemma
+nilpotent, and since~$M'$ is injective and since~$1 \neq 0$, the preceding lemma
 shows that the number of columns of~$M'$ is zero.
@@
 middle. We were not able to discern that these proofs are just different
-coatings of the same idea that ours is based on; these proofs all employ
+presentations of the same idea that ours is based on; these proofs all employ
 computations with Cramer-like formulas, while our proof stresses the similarity
@@
 possible to do without, as already our unrolled proof demonstrates. Thierry
-Coquand and Claude Quitté gave a beautiful proof without case distinctions an
-application of a local--global
+Coquand and Claude Quitté gave a beautiful proof without case distinctions by
+applying a local--global
 principle~\cite[Theorem~2.4]{coquand-quitte:constructive-finite-free-resolutions};

### 2020-04-08 paper-herrsching2019/main.tex
Reformatted introduction

@@
 \begin{abstract}
-XXX This contribution gives an introduction to the theory of locales,
-\end{abstract}
-\body
-
-\tableofcontents
-\bigskip
-
-\noindent
 The purpose of this contribution is to give a coherent account of a particular
@@
 ideas:
-
-\begin{enumerate}
-\item Locales are a kind of space in which opens instead of points are
+(1)~Locales are a kind of space in which opens instead of points are
 fundamental.
-% Any propositional geometric theory gives rise to a locale of models.
-\item Sheaf semantics allows us to explore mathematical objects from custom-tailored
+(2)~Sheaf semantics allows us to explore mathematical objects from custom-tailored
 mathematical universes.
-\item Without loss of generality, any reduced ring is a field.
-\end{enumerate}
-
-Unlike some other texts on locales or sheaves, this text is set in a
-constructive (but impredicative) metatheory. We do not use the law of excluded
-middle nor any version of the axiom of choice, and we also do not adopt any
-nonclassical principles. We freely use powersets, but more for linguistic
-convenience than by necessity.
-
-In particular, the central idea presented in
-Section~\ref{sect:appl}, the reduction technique by which we can pretend that any
-given reduced ring is a field, is very robust with respect to the metatheory.
-Even though we chose a semantic point of view for its derivation, the final
-result is a purely syntactical transformation. As it stands, this text could be
-formalized in the kind of type theory which is valid in
-toposes~\cite{maietti:modular-correspondence} or in~\textsc{izf}~\cite{crosilla:cst-izf}.
-
-For readers familiar with the themes of this contribution, we would like to
-make a couple of remarks.
-\begin{enumerate}
-\item We present a self-contained proof of the simple
-version of Barr's theorem as Theorem~\ref{thm:baby-barr}. For reducing
-intuitionistic provability to geometric provability, we employ (what amounts
-to) the syntactic site, similar to the refined completeness result
-by Thierry Coquand~\cite{coquand:site} but closer to the standard description
-of the syntactic site. For reducing classical provability to intuitionistic
-provability, we employ a variant of the double negation translation. We were
-not able to track down a reference to this approach in the context of
-infinitary logic, but the basic idea has long been known and is summarized, for
-instance, in Ref.~\refcite{aczel:russell-prawitz}.
-\item In Section~\ref{sect:sheaves}, where we introduce sheaves, we should
-probably also introduce sites. They would fit the narrative well and allow for
-the construction of generic models of arbitrary geometric theories and not only
-propositional ones. They would also allow us to cut down on the metatheory,
-sitting comfortably within predicative environments such as arithmetic
-universes~\cite{maietti:au,vickers:sketches} or~\textsc{czf}.
-We do not, however, mainly to not presume
-prerequisites in category theory.
-\item The technique presented in Section~\ref{sect:appl} can be
-regarded as a first-order (and, if desired, higher-order) coating of an
-underlying geometric theory. We propose to take this coating seriously; in
-particular to make good use of surprising nongeometric sequents validated by the
-generic model; and, for ease of use by algebraists, to cast it in semantic terms,
-even though it is all syntax behind the scenes.\footnote{By now it is an established idea in constructive algebra that we should
-consider the theory of prime filters instead of the prime filters themselves,
-hoping that, for instance, a given mathematical proof of the statement~``no
-prime filter contains~$x$'' can also be cast in the
-geometric language of the theory, which then yields the nilpotency of~$x$
-constructively. If the given proof uses only geometric reasoning, this is
-immediate; if it uses intuitionistic reasoning, it can be compiled down (for
-instance precisely by the technique we use); and if it uses classical reasoning, a
-version of Barr's theorem might be applicable. The elusive actual prime filters
-of the ring are replaced with the concrete and purely syntactical generic prime
-filter.
-
-This approach misses that the generic model is interesting on its own and in
-particular that it satisfies peculiar nongeometric sequents such as ``nonunits
-are nilpotent'' and ``any ideal is \notnot finitely generated''. In this sense,
-the generic model is better than the actual models it aims to replace, and we
-believe that potential applications of this insight should be properly
-explored.}
-\end{enumerate}
+(3)~Without loss of generality, any reduced ring is a field.
+\end{abstract}
+\body
+
+\tableofcontents
 
@@
 
+\paragraph{Remark on the metatheory.}
+Unlike some other texts on locales or sheaves, this text is set in a
+constructive (but impredicative) metatheory. We do not use the law of excluded
+middle nor any version of the axiom of choice, and we also do not adopt any
+nonclassical principles. We freely use powersets, but more for linguistic
+convenience than by necessity.
+
+In particular, the central idea presented in
+Section~\ref{sect:appl}, the reduction technique by which we can pretend that any
+given reduced ring is a field, is very robust with respect to the metatheory.
+Even though we chose a semantic point of view for its derivation, the final
+result is a purely syntactical transformation. As it stands, this text could be
+formalized in the kind of type theory which is valid in
+toposes~\cite{maietti:modular-correspondence} or in~\textsc{izf}~\cite{crosilla:cst-izf}.
+
+\paragraph{Notes for readers familiar with the themes of this contribution.}
+We present a self-contained proof of the simple
+version of Barr's theorem as Theorem~\ref{thm:baby-barr}. For reducing
+intuitionistic provability to geometric provability, we employ (what amounts
+to) the syntactic site, similar to the refined completeness result
+by Thierry Coquand~\cite{coquand:site} but closer to the standard description
+of the syntactic site. For reducing classical provability to intuitionistic
+provability, we employ a variant of the double negation translation. We were
+not able to track down a reference to this approach in the context of
+infinitary logic, but the basic idea has long been known and is summarized, for
+instance, in Ref.~\refcite{aczel:russell-prawitz}.
+
+In Section~\ref{sect:sheaves}, where we introduce sheaves, we should
+probably also introduce sites. They would fit the narrative well and allow for
+the construction of generic models of arbitrary geometric theories and not only
+propositional ones. They would also allow us to cut down on the required metatheory,
+sitting comfortably within predicative environments such as arithmetic
+universes~\cite{maietti:au,vickers:sketches} or~\textsc{czf}.
+We do not, however, mainly to not require
+prerequisites in category theory.
+
+The technique presented in Section~\ref{sect:appl} can be
+regarded as a first-order (and, if desired, higher-order) coating of an
+underlying geometric theory. We propose to take this coating seriously; in
+particular to make good use of surprising nongeometric sequents validated by the
+generic model; and, for ease of use by algebraists, to cast it in semantic terms,
+even though it is all syntax behind the scenes.\footnote{By now it is an established idea in constructive algebra that we should
+consider the theory of prime filters instead of the prime filters themselves,
+hoping that, for instance, a given mathematical proof of the statement~``no
+prime filter contains~$x$'' can also be cast in the
+geometric language of the theory, which then yields the nilpotency of~$x$
+constructively. If the given proof uses only geometric reasoning, this is
+immediate; if it uses intuitionistic reasoning, it can be compiled down (for
+instance precisely by the technique we use); and if it uses classical reasoning, a
+version of Barr's theorem might be applicable. The elusive actual prime filters
+of the ring are replaced with the concrete and purely syntactical generic prime
+filter.
+
+This approach misses that the generic model is interesting on its own and in
+particular that it satisfies peculiar nongeometric sequents such as ``nonunits
+are nilpotent'' and ``any ideal is \notnot finitely generated''. In this sense,
+the generic model is better than the actual models it aims to replace, and we
+believe that potential applications of this insight should be properly
+explored.}
+
 \section{Locales}
@@
 
-\begin{definition}A \emph{geometric theory} consists of
+\begin{definition}\label{defn:geometric-theory}
+A \emph{geometric theory} consists of
 \begin{enumerate}
@@
 \end{definition}
+
+We refer to Ref.~\cite[Section~D1.1]{johnstone:elephant} for a more detailed
+version of Definition~\ref{defn:geometric-theory}.
 
@@
 theories often depend on given mathematical objects. For instance, for each
-ring~$A$, there is the geometric theory of prime ideals of~$A$. This theory
+ring~$A$, there is the geometric theory of prime filters of~$A$. This theory
 will play a crucial role in Section~\ref{sect:appl}.
@@
 \end{enumerate}
-Barr's theorem has ramifications to constructive algebra, as pointed out by
+Barr's theorem has ramifications for constructive algebra, as pointed out by
 Gavin Wraith~\cite{wraith:intuitionistic-algebra}.
@@
 for the example of the Nullstellensatz, a constructive proof is given
-as~\cite[Theorem~9.7]{lombardi-quitte:constructive-algebra}.
+in Ref.~\cite[Theorem~9.7]{lombardi-quitte:constructive-algebra}.
 
@@
 theory~$\TT$ is its \emph{Lindenbaum algebra}: This is the partially ordered
-set of the formulas over the signature of~$\TT$ modulo~$\TT$-provable
+set of the geometric formulas over the signature of~$\TT$ modulo~$\TT$-provable
 equivalence, equipped with the ordering given by~$[\varphi] \preceq [\psi]$ iff~$\TT$
@@
 \begin{proposition}\label{prop:at-points}
-Let~$\TT$ be a geometric theory. Let~$\TT_0$ be the same
-theory as~$\TT$ but without any axioms. Let~$M$ be a sheaf model
-of~$\TT_0$ on a locale~$X$ (hence just a \emph{sheaf structure}) and consider
+Let~$\TT$ be a geometric theory without any axioms. Let~$M$ be a sheaf model
+of~$\TT$ on a locale~$X$ (hence just a \emph{sheaf structure}) and consider
 the structures~$M_x$ obtained by computing the stalks at points~$x \in \Pt(X)$.
@@
 signify that the elements~$x_1,\ldots,x_n$ exhaust all elements of~$X$. A
-solution is given in~\cite[Example~D1.1.7(k)]{johnstone:elephant}.\par}
+solution is given in Ref.~\cite[Example~D1.1.7(k)]{johnstone:elephant}.\par}
 \end{alphlist}
+\end{exercise}
+
+\begin{exercise}{Issues of size}%
+Working in~\textsc{izf}, show that the Lindenbaum algebra of a propositional
+geometric theory~$\TT$ can be realized as a set, even though the class of
+geometric formulas is typically a proper class.\smallskip
+
+\noindent{\scriptsize\emph{Hint.} Use that any geometric formula is provably equivalent
+to one of the form~$\bigvee_i \varphi_i$, where each formula~$\varphi_i$ is a
+conjunction of nullary relation symbols.\par}
 \end{exercise}

### 2020-04-09 paper-herrsching2019/main.tex
Where to go from here

@@
 
-\subsection*{Introduction}
+\subsection*{Introduction\footnote{
+Unlike some other texts on locales or sheaves, this text is set in a
+constructive (but impredicative) metatheory. We do not use the law of excluded
+middle nor any version of the axiom of choice, and we also do not adopt any
+nonclassical principles. We freely use powersets, but more for linguistic
+convenience than by necessity.
+In particular, the central idea presented in
+Section~\ref{sect:appl}, the reduction technique by which we can pretend that any
+given reduced ring is a field, is very robust with respect to the metatheory.
+Even though we chose a semantic point of view for its derivation, the final
+result is a purely syntactical transformation. As it stands, this text could be
+formalized in the kind of type theory which is valid in
+toposes~\cite{maietti:modular-correspondence} or
+in~\textsc{izf}~\cite{crosilla:cst-izf}.}}
+
 \addcontentsline{toc}{section}{Introduction}
@@
 
-\paragraph{Remark on the metatheory.}
-Unlike some other texts on locales or sheaves, this text is set in a
-constructive (but impredicative) metatheory. We do not use the law of excluded
-middle nor any version of the axiom of choice, and we also do not adopt any
-nonclassical principles. We freely use powersets, but more for linguistic
-convenience than by necessity.
-
-In particular, the central idea presented in
-Section~\ref{sect:appl}, the reduction technique by which we can pretend that any
-given reduced ring is a field, is very robust with respect to the metatheory.
-Even though we chose a semantic point of view for its derivation, the final
-result is a purely syntactical transformation. As it stands, this text could be
-formalized in the kind of type theory which is valid in
-toposes~\cite{maietti:modular-correspondence} or in~\textsc{izf}~\cite{crosilla:cst-izf}.
+\paragraph{Where to go from here.} The themes which this contribution touches
+on are embedded into a larger context, and we invite readers to follow those
+interesting tangents.
+
+Regarding spaces, the notion of locales can be modified in several ways.
+Firstly, there are \emph{sites} and closely related
+\emph{toposes}.~\cite{leinster:introduction,moerdijk-maclane:sheaves-logic}.
+Given two opens~$U$ and~$V$ of a locale, there is only a truth value as to
+whether~$U$ is contained in~$V$; with sites and toposes, there can be a
+nontrivial \emph{set} of ways that~$U$ is contained in~$V$. In an orthogonal
+direction, there is the notion of a \emph{formal
+topology}~\cite{sambin:some-points,sambin:ifs}. These implement the idea of
+pointfree topology in predicative settings.
+
+Regarding sheaf semantics, Section~\ref{sect:sheaf-semantics} only presents
+the first-order case, but the theory can be extended include unbounded
+quantification~\cite{shulman:stack-semantics}, function types and power types
+and even a fully-fledged type theory~\cite{maietti:modular-correspondence}. The
+sheaf semantics of appropriate toposes has been used to give \emph{synthethic
+accounts} of several fields, such as differential geometry~\cite{kock:sdg},
+domain theory~\cite{hyland:synthetic-domain-theory} and computability
+theory~\cite{bauer:synthetic-computability-theory}.
+
+Regarding constructive algebra, there are by now several
+textbooks~\cite{mines-richman-ruitenburg:constructive-algebra,lombardi-quitte:constructive-algebra,yengui:constructive-commutative-algebra}
+on constructive algebra available and there is a vast literature on
+the intriguing subject of constructivizing classical commutative algebra.
+Starting points include
+Refs.~\refcite{coste-lombardi-roy:dynamical-method,coquand-lombardi:logical-approach,rinaldi-schuster-wessel:edde,cederquist-coquand:entrel}.
 
@@
 certain topos without any filters~\cite[pp.~200f.]{tierney:spectrum}, thus
+
+XXX: Two views on A~
+XXX: Direct definition of A~ possible

### 2020-04-09 paper-herrsching2019/main.tex
More references

@@
 topology}~\cite{sambin:some-points,sambin:ifs}. These implement the idea of
-pointfree topology in predicative settings.
+pointfree topology in predicative settings. In this context also basics of
+algebraic geometry have been
+developed~\cite{schuster:formal-zariski,cls:spectral-schemes,cls:projective-spectrum}.
 
@@
 
+\begin{remark}Given the importance of Theorem~\ref{thm:workhorse}, it is
+natural to wonder where the proof stems from. Our proof proceeded in an ad hoc
+fashion, verifying that the frame of radical ideals has the required properties
+-- but where does the idea to use radical ideals come from? In general, determining the set
+of provable sequents of a geometric theory is a challenging problem. Our most
+efficient tool is probably the theory of \emph{entailment relations}. We refer
+to Refs.~\refcite{rinaldi-schuster-wessel:edde,cederquist-coquand:entrel} and
+the references therein and also strongly recommend the literature on dynamical
+methods in commutative
+algebra~\refcite{coste-lombardi-roy:dynamical-method,coquand-lombardi:logical-approach,coquand:site}.
+\end{remark}
+
 Having identified the underlying frame of~$\Spec(A)$, the second workhorse is
@@
 out within~\textsc{pra}.
+
+The approach using locales and sheaves provides two ingredients which go beyond
+mere syntax: Firstly, they give motivation for setting up the semantics in
+the way we do (Section~\ref{sect:origins}). Secondly, they
+conceptualize~$A^\sim$ as a single entity, just as rings like~$A/\mmm$
+or~$A_\ppp$ are single entities; nevertheless, after unrolling the definitions,
+any statement about~$A^\sim$ is merely a (logically more complex) statement
+about~$A$ and its localizations~$A[f^{-1}]$.
 
@@
 \begin{proof}Combination of Theorem~\ref{thm:basic-properties-sheaf-semantics}
-and Corollary~\ref{cor:algebraic-reformulation}.\end{proof}
+and Corollary~\ref{cor:algebraic-reformulation}, or alternatively direct proof
+by induction without using any preliminaries from earlier sections of this
+contribution.\end{proof}
 
@@
 
-% schuster:formal-zariski
-% cls:spectral-schemes
-really necessary? Wouldn't it suffice to define~$\Spec(A)$ as the topological
-space of filters of~$A$? Indeed, for some time this was
-believed~\cite[Section~3]{lawvere:icm-address}; however, this hope turned out
-certain topos without any filters~\cite[pp.~200f.]{tierney:spectrum}, thus
-
-XXX: Two views on A~
-XXX: Direct definition of A~ possible
+FUTURE: Cite Duval and Joyal.

### 2020-04-09 paper-herrsching2019/main.tex
Cleaner macros

@@
 
-\usepackage{hyperref}
-
 % NOTE TO EDITORS AND LAYOUTERS: This package increases the quality of the
@@
 
-\newcommand{\A}{\mathcal{A}}
-\newcommand{\B}{\mathcal{B}}
-\newcommand{\C}{\mathcal{C}}
-\newcommand{\M}{\mathcal{M}}
-\renewcommand{\AA}{\mathbb{A}}
-\newcommand{\E}{\mathcal{E}}
-\newcommand{\F}{\mathcal{F}}
-\newcommand{\G}{\mathcal{G}}
-\newcommand{\J}{\mathcal{J}}
-\newcommand{\GG}{\mathbb{G}}
-\renewcommand{\O}{\mathcal{O}}
-\newcommand{\K}{\mathcal{K}}
+\newcommand{\CCC}{\mathcal{C}}
+\newcommand{\OOO}{\mathcal{O}}
 \newcommand{\NN}{\mathbb{N}}
@@
 \newcommand{\TT}{\mathbb{T}}
-\newcommand{\PP}{\mathbb{P}}
 \newcommand{\ZZ}{\mathbb{Z}}
 \newcommand{\CC}{\mathbb{C}}
-\renewcommand{\P}{\mathcal{P}}
 \newcommand{\aaa}{\mathfrak{a}}
@@
 \newcommand{\Sh}{\mathrm{Sh}}
-\newcommand{\GL}{\mathrm{GL}}
 \newcommand{\ab}{\mathrm{ab}}
-\newcommand{\Zar}{\mathrm{Zar}}
 \newcommand{\op}{\mathrm{op}}
 \newcommand{\Set}{\mathrm{Set}}
-\newcommand{\Eff}{\mathrm{Ef{}f}}
-\newcommand{\Sch}{\mathrm{Sch}}
-\newcommand{\Aff}{\mathrm{Aff}}
-\newcommand{\Ring}{\mathrm{Ring}}
-\newcommand{\LocRing}{\mathrm{LocRing}}
-\newcommand{\LRS}{\mathrm{LRS}}
 \newcommand{\Hom}{\mathrm{Hom}}
@@
 \newcommand{\Idl}{\mathrm{Idl}}
-\newcommand{\lra}{\longrightarrow}
-\newcommand{\RelSpec}{\operatorname{Spec}}
-\renewcommand{\_}{\mathpunct{.}}
+\newcommand{\+}{\mathpunct{.}}
 \newcommand{\?}{\,{:}\,}
-\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
-\newcommand{\ul}[1]{\underline{#1}}
-\newcommand{\affl}{\ensuremath{{\ul{\ensuremath{\AA}}^1}}}
-\newcommand{\Ll}{\text{iff}}
-\newcommand{\inv}{inv.\@}
 \newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}
-\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates
-\newcommand{\brak}[1]{{\llbracket{#1}\rrbracket}}
 \newcommand{\pt}{\mathrm{pt}}
@@
 \newcommand{\bott}{\bot\!\!\!\!\bot}
+\newcommand{\brak}[1]{{\llbracket{#1}\rrbracket}}
 
@@
 \textbf{Definition.} A \emph{topological space} consists of a set~$X$ of points
-together with a set~$\O(X) \subseteq \P(X)$ of point sets which are deemed
+together with a set~$\OOO(X) \subseteq \P(X)$ of point sets which are deemed
 \emph{open} such that (arbitrary, set-indexed) unions and finite intersections
@@
 \[ (\{ f : U \to \CC \,|\, \text{$f$ is continuous and $f(z)^2 = z$ for all~$z
-\in \CC$} \})_{U \in \O(\CC)} \]
+\in \CC$} \})_{U \in \OOO(\CC)} \]
 of sets as the solution (or perhaps, since no solution was actually given, as
@@
 every~$\alpha \in U$, there is a number~$n \in \NN$ such that
-\[ \{ \beta \in 2^\NN \,|\, \forall i < n\_ \alpha(i) = \beta(i) \} \subseteq U. \]
+\[ \{ \beta \in 2^\NN \,|\, \forall i < n\+ \alpha(i) = \beta(i) \} \subseteq U. \]
 
@@
 = \{ \alpha \in 2^\NN \,|\, \alpha(0) = \alpha(1) = 1 \}$ and~$W = \{
-\alpha \in 2^\NN \,|\, \exists n\in\NN\_ \alpha(n) = 1 \}$ are open and the sets
+\alpha \in 2^\NN \,|\, \exists n\in\NN\+ \alpha(n) = 1 \}$ are open and the sets
 \begin{align*}
-  A &\defeq \{ \alpha \in 2^\NN \,|\, \forall n\in\NN\_ \alpha(n) = 0 \} \\
+  A &\defeq \{ \alpha \in 2^\NN \,|\, \forall n\in\NN\+ \alpha(n) = 0 \} \\
   B &\defeq \Bigl\{ \alpha \in 2^\NN \,\Big|\, \lim_{n \to \infty} \frac{1}{n} \sum_{i=0}^{n-1} \alpha(i) = \frac{1}{2} \Bigr\}
@@
 according to this definition, there are no random sequences: For any
-sequence~$\alpha$, the set~$\{ \beta \in 2^\NN \,|\, \exists n \in \NN\_
+sequence~$\alpha$, the set~$\{ \beta \in 2^\NN \,|\, \exists n \in \NN\+
 \alpha(n) \neq \beta(n) \}$ is open, has measure~$1$ and does not
@@
 The starting point of the theory of locales is the following
-observation. The set~$\O(X)$ of open subsets of a topological
+observation. The set~$\OOO(X)$ of open subsets of a topological
 space~$X$ forms a partially ordered set which has
@@
 \begin{definition}\label{defn:locale}
-A \emph{locale}~$X$ is given by a frame~$\O(X)$, the ``frame
+A \emph{locale}~$X$ is given by a frame~$\OOO(X)$, the ``frame
 of opens of~$X$''. A \emph{morphism~$f : X \to X'$ of locales} (or ``continuous
-map of locales'') is a frame homomorphism~$\O(X') \to \O(X)$.\end{definition}
+map of locales'') is a frame homomorphism~$\OOO(X') \to \OOO(X)$.\end{definition}
 
@@
 \item Any topological space~$Y$ induces a locale~$L(Y)$ by
-setting~$\O(L(Y)) \defeq \O(Y)$. A continuous map~$f : Y \to Y'$ of topological
-spaces induces the frame homomorphism~$\O(Y') \to \O(Y),\,U \mapsto f^{-1}[U]$
+setting~$\OOO(L(Y)) \defeq \OOO(Y)$. A continuous map~$f : Y \to Y'$ of topological
+spaces induces the frame homomorphism~$\OOO(Y') \to \OOO(Y),\,U \mapsto f^{-1}[U]$
 in the other direction and hence a morphism~$L(Y) \to L(Y')$ of locales in the
@@
 
-The underlying frame homomorphism~$\alpha : \O(X) \to \Omega$ of a point~$x$
+The underlying frame homomorphism~$\alpha : \OOO(X) \to \Omega$ of a point~$x$
 of~$X$ can be pictured as mapping each open~$U$ of~$X$ to the truth value to which
@@
 The point~$x$ is completely determined by the information to which opens it
-belongs, that is by the set {$\{ U \in \O(X) \,|\, \alpha(U) = \top \}$}. This set is a \emph{completely
-prime filter}, that is a subset~$\fff \subseteq \O(X)$ which is upward-closed,
+belongs, that is by the set {$\{ U \in \OOO(X) \,|\, \alpha(U) = \top \}$}. This set is a \emph{completely
+prime filter}, that is a subset~$\fff \subseteq \OOO(X)$ which is upward-closed,
 closed under finite meets and for which~$\bigvee_i U_i \in \fff$ implies~$U_i
@@
 \begin{definition}A locale~$X$ is \emph{spatial} if and only if its points
-suffice to detect the inclusion relation on the opens of~$X$, that is if for any opens~$U, V \in \O(X)$,
+suffice to detect the inclusion relation on the opens of~$X$, that is if for any opens~$U, V \in \OOO(X)$,
 if~$x \inplus U \Rightarrow x \inplus V$ for all points~$x$ of~$X$, then~$U
@@
 An equivalent definition is: A locale~$X$ is \emph{spatial} if and only if its points
-suffice to distinguish its opens, that is if for any opens~$U, V \in \O(X)$, if~$x \inplus U
+suffice to distinguish its opens, that is if for any opens~$U, V \in \OOO(X)$, if~$x \inplus U
 \Leftrightarrow x \inplus V$ for all points~$x$ of~$X$, then~$U = V$.
@@
 {\scriptsize\emph{Hint.} Given a completely prime filter~$\fff \subseteq
-\O(Y)$, the set~$\bigcup\{ U \in \O(Y) \,|\, U \not\in \fff \}$ is a
+\OOO(Y)$, the set~$\bigcup\{ U \in \OOO(Y) \,|\, U \not\in \fff \}$ is a
 meet-irreducible open subset.\par}
@@
   d(x,y) \leq q \wedge q \leq r &\Longrightarrow d(x,y) \leq r \\
-  (\forall r \in \QQ_{> q}\_ d(x,y) \leq r) &\Longrightarrow d(x,y) \leq q \\
-  &\mathrel{\phantom{\Longrightarrow}} \exists q \in \QQ_{\geq0}\_ d(x,y) \leq q \\
+  (\forall r \in \QQ_{> q}\+ d(x,y) \leq r) &\Longrightarrow d(x,y) \leq q \\
+  &\mathrel{\phantom{\Longrightarrow}} \exists q \in \QQ_{\geq0}\+ d(x,y) \leq q \\
   &\mathrel{\phantom{\Longrightarrow}} d(x,x) \leq 0 \\
@@
 \emph{covariant approach} to locales can be developed, in which locale maps~$X
-\to X'$ are defined to be certain kinds of maps~$\O(X) \to
-\O(X')$~\cite{picado-pultr:covariant}.\par}
+\to X'$ are defined to be certain kinds of maps~$\OOO(X) \to
+\OOO(X')$~\cite{picado-pultr:covariant}.\par}
 \end{exercise}
@@
 Verify that the points of a locale~$X$ are in canonical one-to-one correspondence
-with the completely prime filters of~$\O(X)$.
+with the completely prime filters of~$\OOO(X)$.
 \end{exercise}
@@
 \item Let~$M$ be a set and endow it with the discrete topology. Show that an
-open~$U \in \O(L(M))$ is dense iff for any element~$x \in M$, $\neg\neg(x \in
+open~$U \in \OOO(L(M))$ is dense iff for any element~$x \in M$, $\neg\neg(x \in
 U)$.
@@
 \begin{alphlist}[(c)]
-\item For any opens~$U, V \in \O(X)$: If~$x \inplus U \Rightarrow y \inplus V$
+\item For any opens~$U, V \in \OOO(X)$: If~$x \inplus U \Rightarrow y \inplus V$
 for all points~$x$ of~$X$, then~$U \preceq V$.
-\item For any opens~$U, V \in \O(X)$: If~$x \inplus U \Leftrightarrow y \inplus V$
+\item For any opens~$U, V \in \OOO(X)$: If~$x \inplus U \Leftrightarrow y \inplus V$
 for all points~$x$ of~$X$, then~$U = V$.
@@
     \doubleLine
-    \UnaryInf$\exists y\?Y\_ \varphi\ \fCenter\seq{\vec x} \psi$
+    \UnaryInf$\exists y\?Y\+ \varphi\ \fCenter\seq{\vec x} \psi$
     \DisplayProof
@@
     \AxiomC{$\phantom{\seq{\vec x}}$}
-    \UnaryInfC{$(\exists y\?Y\_ \varphi) \wedge \psi \seq{\vec x} \exists y\?Y\_ (\varphi \wedge \psi)$}
+    \UnaryInfC{$(\exists y\?Y\+ \varphi) \wedge \psi \seq{\vec x} \exists y\?Y\+ (\varphi \wedge \psi)$}
     \DisplayProof
@@
     \doubleLine
-    \UnaryInf$\varphi\ \fCenter\seq{\vec x} \forall y\?Y\_ \psi$
+    \UnaryInf$\varphi\ \fCenter\seq{\vec x} \forall y\?Y\+ \psi$
     \DisplayProof
@@
 \begin{align*}
-  \top &\seq{x:R} (x = 0) \vee (\exists y\?R\_ xy = 1),
+  \top &\seq{x:R} (x = 0) \vee (\exists y\?R\+ xy = 1),
 \intertext{we describe \emph{geometric fields}, and by additionally adding for
 each natural number~$n \geq 1$ the axiom}
-  \top &\seq{a_0:R,\ldots,a_{n-1}:R} \exists x\?R\_ x^n + a_{n-1}x^{n-1} +
+  \top &\seq{a_0:R,\ldots,a_{n-1}:R} \exists x\?R\+ x^n + a_{n-1}x^{n-1} +
   \cdots + a_1x + a_0 = 0,
@@
 A variant of this theory is the theory of inhabited objects, which has the
-axiom~$\top \vdash \exists x\?X\_ \top$. Its set-based models are precisely the
+axiom~$\top \vdash \exists x\?X\+ \top$. Its set-based models are precisely the
 inhabited sets.
@@
     (\varphi \Rightarrow \psi)^\nabla &\defeqv (\varphi^\nabla \Rightarrow \psi^\nabla) \\
-    (\forall x\?X\_ \varphi)^\nabla &\defeqv (\forall x\?X\_ \varphi^\nabla) &
-    (\exists x\?X\_ \varphi)^\nabla &\defeqv \nabla(\exists x\?X\_ \varphi^\nabla)
+    (\forall x\?X\+ \varphi)^\nabla &\defeqv (\forall x\?X\+ \varphi^\nabla) &
+    (\exists x\?X\+ \varphi)^\nabla &\defeqv \nabla(\exists x\?X\+ \varphi^\nabla)
   \end{align*}\end{minipage}}
@@
 [\bigvee_{\varphi \in K} \varphi]$, where~$K = \{ \varphi \,|\, \exists i \in
-I\_ \varphi \in M_i \}$.} we can regard the Lindenbaum algebra as the free
+I\+ \varphi \in M_i \}$.} we can regard the Lindenbaum algebra as the free
 frame generated by the nullary relation symbols of~$\TT$ modulo the axioms
@@
 the \emph{theory of points of~$X$}. This theory has one nullary relation
-symbol~$\varphi_U$ for each open~$U \in \O(X)$ (read as ``the point belongs
+symbol~$\varphi_U$ for each open~$U \in \OOO(X)$ (read as ``the point belongs
 to~$U$'') and the following axioms:
 \begin{align*}
-  \varphi_U &\vdash \varphi_V & \text{(for all~$U,V \in \O(X)$ such that~$U \preceq V$)} \\
+  \varphi_U &\vdash \varphi_V & \text{(for all~$U,V \in \OOO(X)$ such that~$U \preceq V$)} \\
   \top &\vdash \varphi_\top \\
-  \varphi_U \wedge \varphi_V &\vdash \varphi_{U \wedge V} & \text{(for all~$U,V \in \O(X)$)} \\
+  \varphi_U \wedge \varphi_V &\vdash \varphi_{U \wedge V} & \text{(for all~$U,V \in \OOO(X)$)} \\
   \varphi_{\bigvee_i U_i} &\vdash \bigvee_i \varphi_{U_i} & \text{(for each set-indexed family~$(U_i)_i$ of opens)}
 \end{align*}
-These are exactly the axioms for a completely prime filter of~$\O(X)$. The
-isomorphism from~$\O(X)$ to the Lindenbaum algebra of this theory maps an
+These are exactly the axioms for a completely prime filter of~$\OOO(X)$. The
+isomorphism from~$\OOO(X)$ to the Lindenbaum algebra of this theory maps an
 open~$U$ to~$[\varphi_U]$.
@@
 \begin{definition}\label{defn:presheaf}
-A \emph{presheaf}~$F$ on a locale~$X$ is a functor~$\O(X)^\op \to \Set$, that is
+A \emph{presheaf}~$F$ on a locale~$X$ is a functor~$\OOO(X)^\op \to \Set$, that is
 \begin{enumerate}
-  \item a set~$F(U)$ for each open~$U \in \O(X)$ and
+  \item a set~$F(U)$ for each open~$U \in \OOO(X)$ and
   \item a map~$(\cdot)|^U_V : F(U) \to F(V)$ for each pair of opens~$V \preceq U$
 \end{enumerate}
-such that $(\cdot)|^U_U = \operatorname{id}_{F(U)}$ for all~$U \in \O(X)$ and
+such that $(\cdot)|^U_U = \operatorname{id}_{F(U)}$ for all~$U \in \OOO(X)$ and
 $(\cdot)|^V_W \circ (\cdot)|^U_V = (\cdot)|^U_W$ for all~$W \preceq V
@@
 
-The prototypical example of a presheaf (and also a sheaf) is the presheaf~$\C$ of
+The prototypical example of a presheaf (and also a sheaf) is the presheaf~$\CCC$ of
 continuous real-valued functions on a topological space~$X$. For this presheaf, the
-set~$\C(U)$ is the set of continuous real-valued maps~$U \to \RR$, and the
+set~$\CCC(U)$ is the set of continuous real-valued maps~$U \to \RR$, and the
 maps~$(\cdot)|^U_V$ are given by actual restriction of functions to smaller
 domains, that is by
-\[ \C(U) \longrightarrow \C(V),\ s \longmapsto s|_V. \]
+\[ \CCC(U) \longrightarrow \CCC(V),\ s \longmapsto s|_V. \]
 
@@
 
-For instance, the presheaf~$\C$ of continuous real-valued functions is a sheaf,
-while its subpresheaf~$\C_\text{c}$ of constant real-valued functions is
+For instance, the presheaf~$\CCC$ of continuous real-valued functions is a sheaf,
+while its subpresheaf~$\CCC_\text{c}$ of constant real-valued functions is
 usually not a sheaf. For instance, on~$X = \RR$, the two constant functions~$f
@@
 There is a general construction called \emph{sheafification} which turns
-presheaves into sheaves. Applied to the presheaf~$\C_\text{c}$, this process yields the
+presheaves into sheaves. Applied to the presheaf~$\CCC_\text{c}$, this process yields the
 sheaf of \emph{locally} constant real-valued functions. Similarly, the
@@
 \begin{definition}A \emph{morphism of presheaves}~$\eta : F \to G$ on a
-locale~$X$ is a family~$(\eta_U)_{U \in \O(X)}$ of maps~$\eta_U : F(U) \to
+locale~$X$ is a family~$(\eta_U)_{U \in \OOO(X)}$ of maps~$\eta_U : F(U) \to
 G(U)$ such that for all sections~$s \in F(U)$ and all opens~$V \preceq U$
@@
 \paragraph{Motivating the sheaf semantics.} Let~$F$ be a sheaf on a locale~$X$.
-What is a sensible reading of the statement~``$\forall s\?F\_ (\varphi(s)
+What is a sensible reading of the statement~``$\forall s\?F\+ (\varphi(s)
 \Rightarrow \psi(s))$''? A first answer is ``for any global section~$s \in
@@
 global sections. Hence a more sophisticated answer is ``for any open~$U \in
-\O(X)$ and any section~$s \in F(U)$ with~$\varphi(s)$,~$\psi(s)$'',
+\OOO(X)$ and any section~$s \in F(U)$ with~$\varphi(s)$,~$\psi(s)$'',
 where~``$\varphi(s)$'' and~``$\psi(s)$'' have to be interpreted in a similar
@@
 
-As a final motivating example, consider the statement~``$\exists s\?F\_
+As a final motivating example, consider the statement~``$\exists s\?F\+
 \varphi(s)$''. This should mean that there is an open covering~$X = \bigvee_i
@@
 \paragraph{Definition and basic properties of the sheaf semantics.}
-Let~$X$ be a locale. A \emph{formula over an open~$U \in \O(X)$} is an
+Let~$X$ be a locale. A \emph{formula over an open~$U \in \OOO(X)$} is an
 infinitary first-order formula (made up using~${=}\ {\top}\ {\wedge}\ \{\bigwedge\}\ {\bot}\ {\vee}\
@@
   \tbl{The (first-order fragment of) the sheaf semantics of a locale~$X$}{\begin{tabular}{@{}l@{\ \ }c@{\ \ }l@{}}
-    $U \models s =_F t$ &\Ll& $s = t$ when evaluated as elements of~$F(U)$ \\
-    $U \models R(s_1,\ldots,s_n)$ &\Ll& $\langle s_1,\ldots,s_n \rangle \in R(U)$ \\
-    $U \models \top$ &\Ll& true \\
-    $U \models \varphi \wedge \psi$ &\Ll& $U \models \varphi$ and $U \models \psi$ \\
-    $U \models \bigwedge_k \varphi_k$ &\Ll& for every index~$k$, $U \models \varphi_k$ \\
-    $U \models \bot$ &\Ll& \hcancel{false}{0pt}{3pt}{0pt}{-2pt}\ $U = \bot$ \\
-    $U \models \varphi \vee \psi$ &\Ll& \hcancel{$U \models \varphi$ or $U \models \psi$}{0pt}{3pt}{0pt}{-2pt}\ there exists an open covering $U = \bigvee_i U_i$ \\
+    $U \models s =_F t$ &iff& $s = t$ when evaluated as elements of~$F(U)$ \\
+    $U \models R(s_1,\ldots,s_n)$ &iff& $\langle s_1,\ldots,s_n \rangle \in R(U)$ \\
+    $U \models \top$ &iff& true \\
+    $U \models \varphi \wedge \psi$ &iff& $U \models \varphi$ and $U \models \psi$ \\
+    $U \models \bigwedge_k \varphi_k$ &iff& for every index~$k$, $U \models \varphi_k$ \\
+    $U \models \bot$ &iff& \hcancel{false}{0pt}{3pt}{0pt}{-2pt}\ $U = \bot$ \\
+    $U \models \varphi \vee \psi$ &iff& \hcancel{$U \models \varphi$ or $U \models \psi$}{0pt}{3pt}{0pt}{-2pt}\ there exists an open covering $U = \bigvee_i U_i$ \\
     && \quad\quad such that for all indices~$i$: $U_i \models \varphi$ or $U_i \models \psi$ \\
-    $U \models \bigvee_k \varphi_k$ &\Ll& \hcancel{for some index~$k$, $U \models \varphi_k$}{0pt}{3pt}{0pt}{-2pt}\ there exists an open covering $U = \bigvee_i U_i$ \\
+    $U \models \bigvee_k \varphi_k$ &iff& \hcancel{for some index~$k$, $U \models \varphi_k$}{0pt}{3pt}{0pt}{-2pt}\ there exists an open covering $U = \bigvee_i U_i$ \\
     && \quad\quad such that for each index~$i$, $U_i \models \varphi_k$ for some index~$k$ \\
-    $U \models \varphi \Rightarrow \psi$ &\Ll& for all~$V \preceq U$: $V \models \varphi$ implies $V \models \psi$ \\
-    $U \models \forall s \? F\_ \varphi(s)$ &\Ll& for all $V \preceq U$ and sections~$s_0 \in F(V)$: $V \models \varphi(s_0)$ \\
-    $U \models \exists s \? F\_ \varphi(s)$ &\Ll& \hcancel{there exists $s_0 \in F(U)$ such that $U \models \varphi(s_0)$}{0pt}{3pt}{0pt}{-2pt} \\
+    $U \models \varphi \Rightarrow \psi$ &iff& for all~$V \preceq U$: $V \models \varphi$ implies $V \models \psi$ \\
+    $U \models \forall s \? F\+ \varphi(s)$ &iff& for all $V \preceq U$ and sections~$s_0 \in F(V)$: $V \models \varphi(s_0)$ \\
+    $U \models \exists s \? F\+ \varphi(s)$ &iff& \hcancel{there exists $s_0 \in F(U)$ such that $U \models \varphi(s_0)$}{0pt}{3pt}{0pt}{-2pt} \\
     && there exists an open covering $U = \bigvee_i U_i$ such that for all~$i$: \\
@@
 z$ for all~$z \in U$} \} \]
-on~$X = \CC$. The interpretation of the statement~``$\exists s\?F\_ \top$'' is
+on~$X = \CC$. The interpretation of the statement~``$\exists s\?F\+ \top$'' is
 false, as on no open which contains the origin a continuous square root
-function exists), but the interpretation of~``$\neg\neg(\exists s\?F\_ \top)$''
+function exists), but the interpretation of~``$\neg\neg(\exists s\?F\+ \top)$''
 is true: Unrolling the definition, this statement expresses that
@@
 \seq{x_1{:}X_1,\ldots,x_n{:}X_n} \psi$ is derivable in infinitary intuitionistic first-order
-logic, then~$U \models \forall x_1\?X_1\_ \ldots \forall x_n\?X_n\_ (\varphi
+logic, then~$U \models \forall x_1\?X_1\+ \ldots \forall x_n\?X_n\+ (\varphi
 \Rightarrow \psi)$.
@@
 
-\begin{example}The sheaf~$\C$ of continuous real-valued functions on any
+\begin{example}The sheaf~$\CCC$ of continuous real-valued functions on any
 locale~$X$ together with appropriate morphisms is a sheaf model of the theory
 of rings. For instance, the interpretation of the axiom
-\[ \forall f\?\C\_ \forall g\?\C\_ f + g = g + f \]
+\[ \forall f\?\CCC\+ \forall g\?\CCC\+ f + g = g + f \]
 with the sheaf semantics amounts to the condition
 \begin{quote}
-For any open~$U$ and any continuous function~$f \in \C(U)$, \\
-${\qquad}$ for any open~$V$ and any continuous function~$g \in \C(V)$, \\
+For any open~$U$ and any continuous function~$f \in \CCC(U)$, \\
+${\qquad}$ for any open~$V$ and any continuous function~$g \in \CCC(V)$, \\
 ${\qquad\qquad}$ $f|^U_V + g = g + f|^U_V$.
@@
 In other words, a model of a propositional geometric theory~$\TT$ over a
-locale~$X$ is the same as a frame homomorphism~$\O(L(\TT)) \to \O(X)$, hence a
+locale~$X$ is the same as a frame homomorphism~$\OOO(L(\TT)) \to \OOO(X)$, hence a
 locale morphism~$X \to L(\TT)$.
@@
 Let~$F$ be a sheaf on a locale~$X$. Let~$\varphi$ be a formula over~$X$. Assume
-that~$\Sh(X) \models \exists!s\?F\_ \varphi(s)$, that is
-\[ \Sh(X) \models \exists s\?F\_ \varphi(s) \quad\text{and}\quad
-  \Sh(X) \models \forall s\?F\_ \forall t\?F\_ (\varphi(s) \wedge \varphi(t)
+that~$\Sh(X) \models \exists!s\?F\+ \varphi(s)$, that is
+\[ \Sh(X) \models \exists s\?F\+ \varphi(s) \quad\text{and}\quad
+  \Sh(X) \models \forall s\?F\+ \forall t\?F\+ (\varphi(s) \wedge \varphi(t)
   \Rightarrow s = t). \]
-Show that for any open~$U \in \O(X)$, there is a unique section~$s \in F(U)$
+Show that for any open~$U \in \OOO(X)$, there is a unique section~$s \in F(U)$
 such that~$U \models \varphi(s)$.
@@
 \begin{exercise}{The sheaf of real functions as a field}%
-Let~$X$ be a topological space. Let~$\C$ be the sheaf of continuous real-valued
+Let~$X$ be a topological space. Let~$\CCC$ be the sheaf of continuous real-valued
 functions on~$X$.
 \begin{alphlist}[(d)]
-\item Let~$U$ be an open of~$X$. Let~$f \in \C(U)$. Show using the result
-of Exercise~\ref{ex:global-existence} that~$U \models (\exists g\?\C\_ fg =_\C
-1)$ iff there is a function~$g \in \C(U)$ such that~$fg = 1$.
-\item Show that~$\C$ is a field in that
-$X \models \forall f\?\C\_
-  \bigl(\neg(\exists g\?\C\_ fg = 1)\bigr) \Rightarrow f = 0$.
-\item Give an example of space~$X$ such that it is not the case that~$\C$ is a
+\item Let~$U$ be an open of~$X$. Let~$f \in \CCC(U)$. Show using the result
+of Exercise~\ref{ex:global-existence} that~$U \models (\exists g\?\CCC\+ fg =_\CCC
+1)$ iff there is a function~$g \in \CCC(U)$ such that~$fg = 1$.
+\item Show that~$\CCC$ is a field in that
+$X \models \forall f\?\CCC\+
+  \bigl(\neg(\exists g\?\CCC\+ fg = 1)\bigr) \Rightarrow f = 0$.
+\item Give an example of space~$X$ such that it is not the case that~$\CCC$ is a
 field in the stronger sense that
-$\Sh(X) \models \forall f\?\C\_ (\exists g\?\C\_ fg = 1) \vee f = 0$.
-\item Let~$X = \CC$ and let~$\O$ be the sheaf of holomorphic functions on~$X$.
-Show that, in classical mathematics, the sheaf~$\O$ is \emph{discrete} in that
-$\Sh(\CC) \models \forall f\?\O\_ \forall g\?\O\_ f = g \vee \neg(f = g)$.
+$\Sh(X) \models \forall f\?\CCC\+ (\exists g\?\CCC\+ fg = 1) \vee f = 0$.
+\item Let~$X = \CC$ and let~$\OOO$ be the sheaf of holomorphic functions on~$X$.
+Show that, in classical mathematics, the sheaf~$\OOO$ is \emph{discrete} in that
+$\Sh(\CC) \models \forall f\?\OOO\+ \forall g\?\OOO\+ f = g \vee \neg(f = g)$.
 \end{alphlist}
@@
 example of a function~$f : \CC \to \CC$ such that it is not the case that
-\[ \Sh(\CC) \models \exists z\?F\_ z^2 = f. \]
+\[ \Sh(\CC) \models \exists z\?F\+ z^2 = f. \]
 {\scriptsize\emph{Note.} The sheaf~$F$ can be obtained by constructing, in the
@@
 sheaf~$\underline{\NN}$ on a topological space~$X$ such that
-\[ \Sh(X) \models \exists n\?F\_ \top
+\[ \Sh(X) \models \exists n\?F\+ \top
   \quad\text{but}\quad
-  \Sh(X) \not\models \exists n\?F\_ \forall m\?F\_ n \leq m. \]
+  \Sh(X) \not\models \exists n\?F\+ \forall m\?F\+ n \leq m. \]
 \end{alphlist}
@@
 \emph{radical}~$\sqrt{\aaa}$ of an arbitrary ideal~$\aaa$ is the radical
-ideal~$\{ x \in A \,|\, \exists n \in \NN\_ x^n \in \aaa \}$.
+ideal~$\{ x \in A \,|\, \exists n \in \NN\+ x^n \in \aaa \}$.
 
@@
 \addtocounter{enumi}{4}
-\item\label{item:field} $A^\sim$ is a field: $\forall x\?A^\sim\_ \Bigl((\neg(\exists y\?A^\sim\_ xy = 1)) \Rightarrow x = 0\bigr)$.
+\item\label{item:field} $A^\sim$ is a field: $\forall x\?A^\sim\+ \Bigl((\neg(\exists y\?A^\sim\+ xy = 1)) \Rightarrow x = 0\bigr)$.
 \item\label{item:negneg-stable} $A^\sim$ has $\neg\neg$-stable equality:
-$\forall x,y\?A^\sim\_ \neg\neg(x = y) \Rightarrow x = y$.
+$\forall x,y\?A^\sim\+ \neg\neg(x = y) \Rightarrow x = y$.
 \item\label{item:noetherian} $A^\sim$ is \emph{anonymously Noetherian}.
@@
   \Sh(X) \models \neg(1 =_{A^\sim} 0) \quad\text{and} \\
-  \Sh(X) \models \forall x,y\?A^\sim\_ \bigl((\exists z\?A^\sim\_ (x + y)z
+  \Sh(X) \models \forall x,y\?A^\sim\+ \bigl((\exists z\?A^\sim\+ (x + y)z
   =_{A^\sim} 1) \Longrightarrow \\
   (\exists
-  z\?A^\sim\_ xz =_{A^\sim} 1) \vee (\exists z\?A^\sim\_ yz =_{A^\sim} 1)\bigr).
+  z\?A^\sim\+ xz =_{A^\sim} 1) \vee (\exists z\?A^\sim\+ yz =_{A^\sim} 1)\bigr).
 \end{multline*}
@@
 \emph{spectrum of~$A$} from algebraic geometry, and the sheaf~$A^\sim$ is the
-sheaf of regular functions, also denoted~``$\O_{\Spec(A)}$''. We will give details on its construction in
+sheaf of regular functions, also denoted~``$\OOO_{\Spec(A)}$''. We will give details on its construction in
 Section~\ref{sect:origins}, but first want to reflect on the displayed
@@
 \begin{proposition}If~$A$ is reduced, then the sheaf of rings~$A^\sim$ has $\neg\neg$-stable equality:
-\[ \Spec(A) \models \forall x,y\?A^\sim\_ \neg\neg(x = y) \Rightarrow x = y. \]
+\[ \Spec(A) \models \forall x,y\?A^\sim\+ \neg\neg(x = y) \Rightarrow x = y. \]
 \end{proposition}
@@
   1 = 0 &\vdash \bot \\
-  (\exists z\?R\_ (x+y)z = 1) &\seq{x:R,y:R} (\exists z\?R\_ xz = 1) \vee
-  (\exists z\?R\_ yz = 1),
+  (\exists z\?R\+ (x+y)z = 1) &\seq{x:R,y:R} (\exists z\?R\+ xz = 1) \vee
+  (\exists z\?R\+ yz = 1),
 \end{align*}
@@
 turn this family into the subsheaf of~$\underline{A}$ given by $U \mapsto \{ g
-\in \Hom(U,A) \,|\, \forall x \in A\_ g^{-1}[\{a\}] \preceq [D(a)] \}$. Abusing
+\in \Hom(U,A) \,|\, \forall x \in A\+ g^{-1}[\{a\}] \preceq [D(a)] \}$. Abusing
 notation, we denote this subsheaf also as~``$\fff_0$''. The
@@
   \top &\vdash e_{-a} = -e_a & \text{(for each~$a \in A$)} \\
-  \top &\vdash_{x:R} \bigvee_{a \in A} \bigvee_{s \in A} \bigl((\exists z\?R\_
+  \top &\vdash_{x:R} \bigvee_{a \in A} \bigvee_{s \in A} \bigl((\exists z\?R\+
   e_s z = 1) \wedge e_s x = e_a\bigr) \\
-  e_a = 0 &\vdash \bigvee_{\substack{s \in A\\sa = 0}} (\exists z\?R\_ e_s z = 1) & \text{(for each~$a \in A$)}
+  e_a = 0 &\vdash \bigvee_{\substack{s \in A\\sa = 0}} (\exists z\?R\+ e_s z = 1) & \text{(for each~$a \in A$)}
 \end{align*}
@@
 where~$x$ ranges over the elements of~$A$, modulo the axioms for a prime
-filter. Hence to give a frame homomorphism~$\O(\Spec(A)) \to \Rad(A)$, we need
+filter. Hence to give a frame homomorphism~$\OOO(\Spec(A)) \to \Rad(A)$, we need
 to specify images in~$\Rad(A)$ for each generator in such a way that the
@@
 \begin{proof}For~$f \in A$, we set~$S_f \defeq \{ g \in A \,|\, \exists n \in
-\NN\_ \exists u \in A\_ f^n = ug \} \subseteq A$. This multiplicative system is
+\NN\+ \exists u \in A\+ f^n = ug \} \subseteq A$. This multiplicative system is
 the \emph{saturation} of~$\{ f^0, f^1, \ldots \}$. The localization~$A[S_f^{-1}]$
@@
       for all~$g \in A$, $fg \models \varphi$ implies $fg \models \psi$ \\
-    $f \models \forall x\?A^\sim\_ \varphi(x)$ &iff&
+    $f \models \forall x\?A^\sim\+ \varphi(x)$ &iff&
       for all~$g \in A$ and all $x_0 \in A[(fg)^{-1}]$, $fg \models \varphi(x_0)$ \\
-    $f \models \exists x\?A^\sim\_ \varphi(x)$ &iff&
+    $f \models \exists x\?A^\sim\+ \varphi(x)$ &iff&
       there exists a partition~$f^n = fg_1 + \cdots + fg_m$ such that, \\
@@
 Let~$A$ be a reduced ring. Then~$A^\sim$ is a field in that
-\[ 1 \models \forall x\?A^\sim\_ \bigl((\neg(\exists y\?A^\sim\_ xy = 1)) \Rightarrow
+\[ 1 \models \forall x\?A^\sim\+ \bigl((\neg(\exists y\?A^\sim\+ xy = 1)) \Rightarrow
 x = 0\bigr). \]
@@
 Noetherian integral domains, or Refs.~[\refcite{staats:generic-freeness}],
-\cite[{\href{https://stacks.math.columbia.edu/tag/051Q}{Tag~051Q}}]{stacks-project}
+\cite[Tag~051Q]{stacks-project}
 which proceed in a series of intermediate steps, reducing to that case).\par
@@
 \item Prove that:
-\[ 1 \models \forall f\?A^\sim\_ \forall x\?A^\sim\_
-  \bigl((\exists z\?A^\sim\_ fz = 1) \Rightarrow x = 0\bigr) \Longrightarrow \bigvee_{n \in \NN} (f^n x = 0). \]
+\[ 1 \models \forall f\?A^\sim\+ \forall x\?A^\sim\+
+  \bigl((\exists z\?A^\sim\+ fz = 1) \Rightarrow x = 0\bigr) \Longrightarrow \bigvee_{n \in \NN} (f^n x = 0). \]
 \item Deduce that~$A^\sim$ is ``almost a field'' in the sense that any nonunit
@@
 Let~$A$ be a ring. We introduce an operator~$\nabla$ by
-\[ \nabla\varphi \defeqv \bigl(\forall s\?A^\sim\_ ((\varphi \Rightarrow s = 0)
+\[ \nabla\varphi \defeqv \bigl(\forall s\?A^\sim\+ ((\varphi \Rightarrow s = 0)
 \Rightarrow s = 0\bigr).\]
@@
 \item Show that equality is~$\nabla$-stable in that
-\[ 1 \models \forall x\?A^\sim\_ \forall y\?A^\sim\_ \nabla(x = y) \Rightarrow x = y. \]
+\[ 1 \models \forall x\?A^\sim\+ \forall y\?A^\sim\+ \nabla(x = y) \Rightarrow x = y. \]
 
@@
 is~$\nabla$-invertible, that is:
-\[ 1 \models \forall f\?A^\sim\_
+\[ 1 \models \forall f\?A^\sim\+
   \bigl(
-    (\forall s\?A^\sim\_ fs = 0 \Rightarrow s = 0) \Longleftrightarrow
-    \nabla(\exists g\?A^\sim\_ fg = 1)
+    (\forall s\?A^\sim\+ fs = 0 \Rightarrow s = 0) \Longleftrightarrow
+    \nabla(\exists g\?A^\sim\+ fg = 1)
   \bigr)
@@
 \item Verify the analogous statement for finitely generated ideals:
-\begin{multline*}1 \models \forall f_1\?A^\sim\_ \ldots \forall f_n\?A^\sim\_ \\
+\begin{multline*}1 \models \forall f_1\?A^\sim\+ \ldots \forall f_n\?A^\sim\+ \\
   \bigl(
-    (\forall s\?A^\sim\_ (\textstyle\bigwedge_{i=1}^n f_is = 0) \Rightarrow s = 0) \Longleftrightarrow
+    (\forall s\?A^\sim\+ (\textstyle\bigwedge_{i=1}^n f_is = 0) \Rightarrow s = 0) \Longleftrightarrow
     \nabla(1 \in (f_1,\ldots,f_n))
@@
 \textbf{Proof.} We verify that, under~$\nabla$, there exists an
-invertible~$m$-minor, that is we verify~$\nabla(\bigvee_f (\exists g\?A^\sim\_
+invertible~$m$-minor, that is we verify~$\nabla(\bigvee_f (\exists g\?A^\sim\+
 fg=1))$, where the disjunction is over the~$m$-minors of~$M$.
@@
 Let~$A$ be a ring. Let~$f \in A$. Let~$x \in A[f^{-1}]$. Show using the result
-of Exercise~\ref{ex:global-existence} that~$f \models \exists y\?A^\sim\_ xy
+of Exercise~\ref{ex:global-existence} that~$f \models \exists y\?A^\sim\+ xy
 = 1$ if and only if~$x$ is invertible in~$A[f^{-1}]$.
@@
 
-{\noindent\scriptsize\emph{Hint.} Use the description of the unique map~$\O(\pt) \to
-\O(\Spec(A)),\,\varphi \mapsto \bigvee\{\top\,|\,\varphi\}$ computed in
+{\noindent\scriptsize\emph{Hint.} Use the description of the unique map~$\OOO(\pt) \to
+\OOO(\Spec(A)),\,\varphi \mapsto \bigvee\{\top\,|\,\varphi\}$ computed in
 part~(\ref{item:join-top}) of Exercise~\ref{ex:frame-of-radical-ideals}.\par}

### 2020-04-09 paper-herrsching2019/main.tex
typo

@@
 \textbf{Definition.} A \emph{topological space} consists of a set~$X$ of points
-together with a set~$\OOO(X) \subseteq \P(X)$ of point sets which are deemed
+together with a set~$\OOO(X) \subseteq P(X)$ of point sets which are deemed
 \emph{open} such that (arbitrary, set-indexed) unions and finite intersections

### 2020-04-09 paper-herrsching2019/generalized-spaces.tex
More macro hygiene

--- 2020-04-09 paper-herrsching2019/main.tex
@@
 % depend on it.
-%\usepackage[expansion=true,protrusion=true]{microtype}
-
-\newtheorem{ex}{Exercise}
-\let\Exercisefont\upshape
-\def\Exerciseheadfont{\bfseries}
-\newenvironment{exercise}[1]{
-  \begin{ex}[#1]
-}{\end{ex}}
-
+\usepackage[expansion=true,protrusion=true]{microtype}
+
+\makeindex
+%\newindex{aindx}{adx}{and}{Author Index}       % author index
+%\renewindex{default}{idx}{ind}{Subject Index}  % subject index
+
+\begin{document}
+
+% NOTE TO TEX PERSONS: With the curly bracket, we are opening here a scope
+% which will close at the very end of this contribution. In this way our macros
+% do not pollute the namespace of the other contributions, that is after the
+% scope ends, our macros will no longer exist.
+{
 \newcommand{\hcancel}[5]{%
@@
 }
-
-\makeindex
-%\newindex{aindx}{adx}{and}{Author Index}       % author index
-%\renewindex{default}{idx}{ind}{Subject Index}  % subject index
+\newtheorem{exercise}{Exercise}
+\let\Exercisefont\upshape
+\def\Exerciseheadfont{\bfseries}
 
@@
 
-\begin{document}
-
 \chapter{Generalized spaces for constructive algebra}
@@
 
-\begin{exercise}{The space of points of a locale}%
+\begin{exercise}[The space of points of a locale]%
 \label{ex:canonical-topology}%
@@
 
-\begin{exercise}{Sober topological spaces as locales}%
+\begin{exercise}[Sober topological spaces as locales]%
 Verify the statements made in Remark~\ref{rem:sober-embed}.
@@
 % FUTURE: Fix linebreak
-\begin{exercise}{Hausdorff spaces are sober}%
+\begin{exercise}[Hausdorff spaces are sober]%
 \label{ex:hausdorff-sober}%
@@
 
-\begin{exercise}{Complete metric spaces are sober}
+\begin{exercise}[Complete metric spaces are sober]%
 A \emph{metric space} is a set~$Y$ together with a relation~$d \subseteq Y
@@
 
-\begin{exercise}{Isomorphisms of frames}%
+\begin{exercise}[Isomorphisms of frames]%
 \label{ex:isomorphisms-of-frames}%
@@
 
-\begin{exercise}{The covariant approach to locales}%
+\begin{exercise}[The covariant approach to locales]%
 Let~$A$ be a frame.
@@
 
-\begin{exercise}{Points as complete prime filters}%
+\begin{exercise}[Points as complete prime filters]%
 \label{ex:points-as-filters}%
@@
 
-\begin{exercise}{Dense opens}%
+\begin{exercise}[Dense opens]%
 \label{ex:dense-opens}%
@@
 
-\begin{exercise}{Characterizing spatial locales}%
+\begin{exercise}[Characterizing spatial locales]%
 Let~$X$ be a locale. Show that the following conditions are equivalent.
@@
 
-\begin{exercise}{The frame of radical ideals}%
+\begin{exercise}[The frame of radical ideals]%
 \label{ex:frame-of-radical-ideals}%
@@
 
-\begin{exercise}{Coproduct of locales}%
+\begin{exercise}[Coproduct of locales]%
 \begin{alphlist}[(c)]
@@
 
-\begin{exercise}{Local locales}%
+\begin{exercise}[Local locales]%
 \label{ex:local-locales}%
@@
 
-\begin{exercise}{Maps into discrete spaces}%
+\begin{exercise}[Maps into discrete spaces]%
 Let~$X$ be a locale. Let~$A$ be a set and endow it with the discrete topology. Show
@@
 
-\begin{exercise}{The theory of finite objects}%
+\begin{exercise}[The theory of finite objects]%
 \begin{alphlist}[(c)]
@@
 
-\begin{exercise}{Issues of size}%
+\begin{exercise}[Issues of size]%
 Working in~\textsc{izf}, show that the Lindenbaum algebra of a propositional
@@
 
-\begin{exercise}{Discrete locales}%
+\begin{exercise}[Discrete locales]%
 \begin{alphlist}[(b)]
@@
 
-\begin{exercise}{The localic unit interval}%
+\begin{exercise}[The localic unit interval]%
 Extend the theory of Dede\-kind cuts in such a way that its models are those
@@
 
-\begin{exercise}{More on exotic locales}%
+\begin{exercise}[More on exotic locales]%
 \label{ex:exotic-locales}%
@@
 
-\begin{exercise}{Models as points}%
+\begin{exercise}[Models as points]%
 Let~$\TT$ be a propositional geometric theory. Verify that the points
@@
 
-\begin{exercise}{The product of locales}\label{ex:products-of-locales}%
+\begin{exercise}[The product of locales]\label{ex:products-of-locales}%
 Let~$X$ and~$X'$ be locales, classifying propositional geometric
@@
 
-\begin{exercise}{The coproduct of locales, revisited}%
+\begin{exercise}[The coproduct of locales, revisited]%
 Let~$X$ and~$X'$ be locales, classifying some propositional geometric
@@
 
-\begin{exercise}{Spatiality of the localic reals}%
+\begin{exercise}[Spatiality of the localic reals]%
 \label{ex:spatiality-reals}%
@@
 
-\begin{exercise}{The localic Galois group of a field extension}%
+\begin{exercise}[The localic Galois group of a field extension]%
 Let~$L|k$ be a field extension. The \emph{localic Galois group}~$\Gal(L|k)$
@@
 
-\begin{exercise}{Coherent locales and coherent theories}%
+\begin{exercise}[Coherent locales and coherent theories]%
 \label{ex:coherent}%
@@
 
-\begin{exercise}{Subsheaves of the terminal sheaf}%
+\begin{exercise}[Subsheaves of the terminal sheaf]%
 \label{ex:subsheaves-1}%
@@
 
-\begin{exercise}{A geometric interpretation of double negation}%
+\begin{exercise}[A geometric interpretation of double negation]%
 \label{ex:negneg-interpretation}%
@@
 
-\begin{exercise}{Unique local existence is global existence}%
+\begin{exercise}[Unique local existence is global existence]%
 \label{ex:global-existence}%
@@
 
-\begin{exercise}{The sheaf of real functions as a field}%
+\begin{exercise}[The sheaf of real functions as a field]%
 Let~$X$ be a topological space. Let~$\CCC$ be the sheaf of continuous real-valued
@@
 
-\begin{exercise}{The failure of the fundamental theorem of algebra}%
+\begin{exercise}[The failure of the fundamental theorem of algebra]%
 \label{ex:fta}%
@@
 
-\begin{exercise}{The least number principle}%
+\begin{exercise}[The least number principle]%
 \label{ex:least-number-principle}%
@@
 
-\begin{exercise}{Noetherian properties of the integers}%
+\begin{exercise}[Noetherian properties of the integers]%
 \begin{enumerate}
@@
 
-\begin{exercise}{Free fields}
+\begin{exercise}[Free fields]%
 \begin{alphlist}[(b)]
@@
 
-\begin{exercise}{Nontriviality of~$\boldsymbol{A^\sim}$}%
+\begin{exercise}[Nontriviality of~$\boldsymbol{A^\sim}$]%
 Let~$A$ be a ring. In Proposition~\ref{prop:immediate-consequences}, we gave an
@@
 
-\begin{exercise}{A generalization of the field property}%
+\begin{exercise}[A generalization of the field property]%
 \label{ex:gen-field-property}%
@@
 
-\begin{exercise}{A local--global principle}%
+\begin{exercise}[A local--global principle]%
 \label{ex:local-global}%
@@
 
-\begin{exercise}{Bridging modules and their induced sheaves}%
+\begin{exercise}[Bridging modules and their induced sheaves]%
 \label{ex:bridging-modules}%
@@
 
-\begin{exercise}{Local injectivity and surjectivity}%
+\begin{exercise}[Local injectivity and surjectivity]%
 Let~$M$ and~$N$ be modules over a ring~$A$. Let~$f : M \to N$ be a linear map.
@@
 
-\begin{exercise}{Units of the sheaf model}%
+\begin{exercise}[Units of the sheaf model]%
 Let~$A$ be a ring. Let~$f \in A$. Let~$x \in A[f^{-1}]$. Show using the result
@@
 
-\begin{exercise}{Decidability of invertibility}
+\begin{exercise}[Decidability of invertibility]%
 Show that it is not the case that any element of~$\ZZ^\sim$ is invertible or
@@
 
-\begin{exercise}{Triviality of the spectrum}%
+\begin{exercise}[Triviality of the spectrum]%
 Let~$A$ be a ring. Verify, without assuming~\BPIT, that~$\Spec(A)$ is
@@
 
-\begin{exercise}{Locality of the spectrum}%
+\begin{exercise}[Locality of the spectrum]%
 Let~$A$ be a ring. Show that the spectrum of~$A$ is local in the sense of
@@
 
-\begin{exercise}{Spatiality of the spectrum}%
+\begin{exercise}[Spatiality of the spectrum]%
 Let~$A$ be a ring.
@@
 
-\begin{exercise}{Prüfer domains as valuation domains}%
+\begin{exercise}[Prüfer domains as valuation domains]%
 An \emph{integral domain} is a ring such that~$1 \neq 0$ and such that~$xy = 0$
@@
 
-\begin{exercise}{A basic version of Kaplansky's theorem}
+\begin{exercise}[A basic version of Kaplansky's theorem]%
 \begin{alphlist}[(c)]
@@
 
+}
+% NOTE TO TEX PERSONS: This lonely brace closes the group opened at the very
+% beginning of this contribution.
+
 \bibliographystyle{ws-rv-van}

### 2020-05-08 paper-herrsching2019/generalized-spaces.tex
typo

@@
 reviewed on page~\pageref{item:theory-of-surjections} does not have any
-set-based models, but it is consistent and it does have nontrivial sheaf
+set-based models, but this theory is consistent and it does have nontrivial sheaf
 models. For the purpose of constructive algebra, sheaf models of some geometric
@@
 \RR$ for which any real has infinitely many preimages, has a fractal nature: It
-is covered by the (intersections with~$X'$ of the)~$U_{0x}$, where~$x$ ranges over the reals; the pairwise
+is covered by the (intersections with~$X'$ of the) opens~$U_{0x}$, where~$x$ ranges over the reals; the pairwise
 intersection of these opens is (truly) empty; and they are each isomorphic
@@
 FUTURE: Cite Duval and Joyal.
+
+FUTURE: Relate algebraic quantum mechanics to topos theoryto construct new kind
+of quantum spaces.?? A spectral invariant for non-commutative algebras ??

### 2020-06-21 announcement-agda-workshop.tex
Slides for the first session of the Agda workshop in Padova

@@
         \item a \hil{proof language}
-        \item<3-> somewhat hard to learn on one's own
-        \item<3-> \hil{fun and easy} to learn as part of a course
+        \item<4-> somewhat hard to learn on one's own
+        \item<4-> \hil{fun and easy} to learn as part of a course
       \end{enumerate}
@@
           \item \mbox{\hil{practically explore} type theory}
-          \item appreciate mathematics from a new point of view
+          \item appreciate mathematics \hil{from a new point of view}
+          \item \hil{verify correctness} of programs
         \end{enumerate}
-
-        \bigskip
-        \centering\hil{``proving $=$ programming''}
       }
@@
   \end{columns}
+
+  \pause\pause
+  \vspace*{-22em}\hspace*{-1.5em}\small
+  \vbox{
+    \textit{Three mottos}: \\[-0.6em]
+    \begin{enumerate}\scriptsize
+      \item ``proving $=$ programming'' \\[-1.2em]
+      \item ``conjecturing $=$ specifying'' \\[-1.2em]
+      \item ``induction $=$ recursion''
+    \end{enumerate}
+  }
 \end{frame}
+
+\newcommand{\imgslide}[1]{{\usebackgroundtemplate{\parbox[c][\paperheight][c]{\paperwidth}{\centering\includegraphics[height=\paperheight]{#1}}}\begin{frame}[plain]\end{frame}}}
+
+\imgslide{announcement-agda-workshop/catarina-coquand}
+\imgslide{announcement-agda-workshop/martin-escardo}
+\imgslide{announcement-agda-workshop/pc2018}
+\imgslide{announcement-agda-workshop/fischbachau}
+
+\begin{frame}[plain]
+  \centering
+  \medskip
+
+  \includegraphics[width=0.8\textwidth]{agda-course/wendelstein-1}
+
+  \includegraphics[height=0.245\textwidth]{agda-course/wendelstein-2}
+  \includegraphics[height=0.245\textwidth]{agda-course/wendelstein-3}
+\end{frame}
+
+\imgslide{agda-course/agda-wiki}
 

### 2020-06-21 paper-filmat.tex
Include more stills and video of family (in response to review)

@@
 \usepackage{xspace}
+\usepackage{marginnote,animate,ocgx2}
 
@@
 \begin{figure}
-  \includegraphics[height=3cm]{zeros-in-families-still}
-  \caption{\label{fig:ivt} A single member~$f_{x_0}$ of a continuous
+\marginnote{
+  \begin{ocg}[printocg=never]{layername}{layerid1}{on}
+    \vspace*{-12em}
+    \scalebox{0.2}{
+      \animategraphics[loop,autoplay]{60}{zeros-in-families-frames/}{0000}{0188}
+    }
+    \begin{minipage}{3cm}\scriptsize\noindent
+    In the online version of this contribution, a video of the
+    continuous family can be seen here.\end{minipage}
+  \end{ocg}}
+  \includegraphics[height=3cm]{zeros-in-families-still-1}
+  \includegraphics[height=3cm]{zeros-in-families-still-2}
+  \includegraphics[height=3cm]{zeros-in-families-still-3}
+
+  \caption{\label{fig:ivt}Three members~$f_{x_0}, f_{x_1}, f_{x_2}$ of a continuous
   family~$(f_x)_{x \in X}$ of continuous functions~$f_x : \RR \to \RR$. The
-  parameter space is~$X = [0,1]$ (not shown) and the other functions~$f_x$ are
-  obtained from~$f_{x_0}$ by moving the horizontal plateau further up or
-  down. The depicted member~$f_{x_0}$ has a unique
-  zero and there is an open neighborhood~$U$ of~$x_0$ on which zeros of the
-  functions~$f_x$, $x \in U$ can be picked continuously. However, there is no
+  parameter space is~$X = [0,1]$ (not shown). The functions~$f_x$ are
+  obtained by moving the horizontal plateau up or
+  down. The leftmost depicted member~$f_{x_0}$ has a unique
+  zero, and there is an open neighborhood~$U$ of~$x_0$ on which zeros of the
+  functions~$f_x$, $x \in U$ can be picked continuously. The same is true
+  for~$x_2$ (right figure). However, there is no
   such neighborhood of that particular parameter value~$x_1$ for which the
-  horizontal plateau lies on the $x$-axis.}
+  horizontal plateau lies on the $x$-axis (middle figure).}
 \end{figure}

### 2020-06-21 paper-filmat.tex
to do
typo (in response to review)

@@
 \definecolor{mypurple}{RGB}{150,0,255}
+
+\usepackage{hyperref}
 
@@
 The only caveat of Theorem~\ref{thm:reasoning} is that toposes generally only
-support intuitionistic reasoning, not the full power of the ordinary
+support intuitionistic reasoning and not the full power of the ordinary
 \emph{classical reasoning}. That is, within most toposes, the law of excluded
@@
 which we implicitly refer to when we utter the phrase ``Let~$G$ be a group.''.
-The generic group has has exactly those
+The generic group has exactly those
 properties which are shared by any group whatsoever.\footnote{More precisely,
@@
 theory from the inside of toposes; the resulting picture can look quite
-different than the picture which emerges from within the standard topos.
+different from the picture which emerges from within the standard topos.
 
@@
 example is the \emph{halting problem}: Given a Turing machine~$M$, determine
-whether~$M$ terminates (comes to a stop after having carried out a finite
-number of computational steps) or not.
+whether~$M$ terminates (comes to a stop after having carried out finitely
+many computational steps) or not.
 
@@
 \begin{ex}\textbf{``Any function~$\RR \to \RR$ is continuous.''}\label{sect:eff-continuous}
-In the standard topos, this statement is plainly false, with the sign and Heaviside functions
+In the standard topos, this statement is plainly false, with the sign and Heaviside step functions
 being prominent counterexamples. In the \effective topos, this statement is
@@
 and can therefore determine arbitrarily many rational approximations~$X(m)$.
-However, within in a finite amount of time, the machine~$Y$ can only learn finitely many
+However, within a finite amount of time, the machine~$Y$ can only learn finitely many
 such approximations. Hence a function such as the sign function, for which
 even rough rational approximations of~$\operatorname{sgn}(x)$ require infinite
-precision in the input~$x$, do not exist in the \effective topos.
+precision in the input~$x$, does not exist in the \effective topos.
 \end{ex}
@@
 
-As a consequence, the input given to machine~$S_F$ is entirely uninformative
+As a consequence, the input given to the machine~$S_F$ is entirely uninformative
 and~$S_F$ cannot make direct computational use of it. But its existence ensures that an
@@
 
-  \caption{\label{table:eff} A (fragment of) the translation
+  \caption{\label{table:eff}A (fragment of) the translation
   rules defining the meaning of statements internal to the \effective topos.}
@@
 judgment of whether~$f$ is the zero function or not. Such a machine~$M$ indeed
-exists: It simply has to simulate~$F$ on all inputs~$0,1,\ldots$ in order and
-check whether one of the resulting function values is not zero. This
+exists: It simply has to simulate~$F$ on all inputs~$0,1,\ldots$ and
+check whether one of the resulting function values is not zero. This search
 will require a transfinite amount of time (not least because simulating~$F$ on
-just one input might require a transfinite amount of time), but as an
-infinite-time Turing machine,~$M$ is capable of carrying out this procedure.
+just one input might require a transfinite amount of time), but infinite-time
+Turing machines are capable of carrying out this procedure.
 
@@
 and~$X'$ represent the same real, then~$M(X)$ has to coincide with~$M(X')$.
-This map~$\RR \to \NN$ is injective iff conversely,~$M(X) = M(X')$ implies
+This map~$\RR \to \NN$ is injective iff conversely~$M(X) = M(X')$ implies
 that~$X$ and~$X'$ represent the same real.
@@
 the chosen order of infinite-time Turing machines, and on details of the
-interleaving simulation and comparison procedure -- but it does not depend on
+interleaving simulation and the comparison procedure -- but it does not depend on
 the implementation of~$X$ or on its specific choice of rational
@@
 Such a function is everywhere positive on~$X$ if and only if, from the internal point of
-view~$\Sh(X)$, the number~$a$ is positive; it is everywhere zero if and only
+view of~$\Sh(X)$, the number~$a$ is positive; it is everywhere zero if and only
 if, internally, the number~$a$ is zero; and it is everywhere negative if and
@@
 The weaker statement ``for any real number~$a$ it is \notnot the case that~$a < 0$
-or~$a = 0$ or~$a > 0$'' does hold in~$\Sh(X)$, for this statement is an
-intuitionistic tautology. Its meaning is that there exists a
+or~$a = 0$ or~$a > 0$'' does hold in~$\Sh(X)$, for this statement is a
+theorem of intuitionistic calculus. Its meaning is that there exists a
 dense open~$U$ such that~$U$ can be covered by opens on which~$a$ is either
@@
 
-  \caption{\label{table:sheaf} A (fragment of) the translation rules defining
+  \caption{\label{table:sheaf}A (fragment of) the translation rules defining
   the meaning of statements internal to~$\Sh(X)$, the topos of sheaves over a
@@
   \includegraphics[height=3cm]{trichotomy-3}
-  \caption{\label{fig:trichotomy} Three examples of what the topos~$\Sh(X)$
-  each believes to be a single real number, where the base space~$X$ is the
-  unit interval. (a) A positive real number. (b) A negative real number. (c) A
-  number which is neither negative nor zero nor positive. Externally speaking,
+  \caption{\label{fig:trichotomy}Three examples of what the topos~$\Sh(X)$
+  believes to be a single real number, where the base space~$X$ is the
+  unit interval. (a) A positive real number. (b) A negative real number. (c)~A~number
+  which is neither negative nor zero nor positive. Externally speaking,
   there is no covering of the unit interval by opens on which the
@@
 F(A[(f_jf_k)^{-1}]) \]
-is a coequalizer diagram. The object~$\RR^\sim$ of~$\Zar(\RR)$ is the tautologous functor~$A \mapsto A$.
+is an equalizer diagram. The object~$\RR^\sim$ of~$\Zar(\RR)$ is the tautologous functor~$A \mapsto A$.
 \end{defn}
@@
     \justify
-    A \emph{covering} of an~$\RR$-algebra~$A$ is a family of~$A$-algebras of the
-    form~$(A[f_i^{-1}])_i$ such that~$1 = f_1 + \cdots + f_n \in A$.
+    A \emph{covering} of an~$\RR$-algebra~$A$ is a finite family of~$A$-algebras of the
+    form~$(A[f_i^{-1}])_{i=1,\ldots,n}$ such that~$1 = f_1 + \cdots + f_n \in A$.
   \end{framed}
@@
 
-  \caption{\label{table:zariski} A (fragment of) the Kripke--Joyal translation
+  \caption{\label{table:zariski}A (fragment of) the Kripke--Joyal translation
   rules of the Zariski topos~$\Zar(\RR)$.}
@@
 statement~$\varphi$ is usually quite complex, even if~$\varphi$ is reasonably
-perspicuous.
+transparent.
 
@@
 Also Giuseppe Cammarata.
+
+XXX: Kleene-Vesley topos?

### 2020-06-21 paper-filmat.tex
Pointer to ex/reg completion (in response to review)

@@
 \to \NN$ and~$n \realizes f$ if and only if the~$n$-th Turing machine
-computes~$f$.
-
-\begin{defn}The \emph{\effective topos}~$\Eff$ is the ex/reg completion of the
-category of assemblies.\end{defn}
+computes~$f$. The category of assemblies is a regular category, but it is
+missing \effective quotients. The \effective topos is obtained by a suitable
+completion procedure:
+
+\begin{defn}The \emph{\effective topos}~$\Eff$ is the ex/reg completion (as
+in~\cite[Section~3.4]{menni:exact}) of the category of assemblies.\end{defn}
 

### 2020-06-21 paper-filmat.tex
typo

@@
 The topos~$\Sh_{\neg\neg}(\E)$ is related to the \emph{double negation
-translation}~$\varphi \mapsto \varphi^{\neg\neg}$ from classical logic into
-intuitionistic logic: A statement~$\varphi$ holds in~$\Sh_{\neg\neg}(\E)$ if
-and only if~$\varphi^{\neg\neg}$ holds
+translation} from classical logic into
+intuitionistic logic: A statement holds in~$\Sh_{\neg\neg}(\E)$ if
+and only if its translation holds
 in~$\E$~\cite[Theorem~6.31]{blechschmidt:phd}.

### 2020-06-21 paper-filmat.tex
Some additions (in response to review)

@@
 multiverse: While Hamkins proposes to embrace all models of set theory (not
-necessarily all of them not equally), we propose to embrace all toposes (again
-not necessarily all of them not equally). As every model~$M$ of set theory
+necessarily all of them equally -- we might prefer some models over others), we propose to embrace all toposes (again
+not necessarily all of them equally). As every model~$M$ of set theory
 gives rise to a topos~$\Set_M$, the set-theoretic multiverse is contained in the
@@
 \begin{enumerate}
-\item Heyting arithmetic, the intuitionistic cousin of Peano arithmetic, has
+\item Heyting arithmetic, the intuitionistic cousin of Peano arithmetic, is
+categorical in the sense that it has
 exactly one model up to isomorphism, namely~$\NN$.
@@
 \emph{Grothendieck's generic freeness lemma}, a fundamental theorem in
-algebraic geometry. The new proof uses the lens to reduce to the case
-of fields, where the claim is trivial~\cite[Section~11.5]{blechschmidt:phd}.
+algebraic geometry about the free locus of certain sheaves. The new proof uses the lens to reduce to the case
+of fields, where the claim is trivial~\cite[Section~11.5]{blechschmidt:phd},
+and improves in length on all previously known proofs, even if the topos
+machinery is eliminated by unrolling the appropriate
+definitions~\cite{blechschmidt:generic-freeness}.\footnote{This contribution is
+not the proper place for an exposition of Grothendieck's generic freeness
+lemma, but some aspects can already be appreciated on a syntactical level.
+Grothendieck's generic freeness lemma states that any finitely generated sheaf
+of modules on a reduced scheme is finite locally free \emph{on a dense open}.
+By employing the internal language, this statement is reduced to the following
+fact of intuitionistic linear algebra: Any finitely generated module over a
+field is \notnot finite free.}
 \end{rem}
@@
 called~``$f'(0)$''; this is how we synthetically define the derivative in
-synthetic differential geometry.
+synthetic differential geometry. (However, despite these properties the Zariski
+topos is not yet \emph{well-adapted} to synthetic differential geometry in the
+sense of Definition~\ref{defn:well-adapted} below.)
 
@@
 equivalent to the Kripke--Joyal translation of a different (typically more
-perspicuous) claim about objects of some problem-specific relevant topos and
+transparent) claim about objects of some problem-specific relevant topos and
 then
@@
 
-% FUTURE: Cite blechschmidt:generic-freeness?
-
 \printbibliography

### 2020-06-21 paper-herrsching2019/generalized-spaces.tex
to do

@@
 condition that nonunits are zero and is satisfied by~$A^\sim$ iff~$A$ is reduced
-and of Krull dimension~$\leq 0$.
+and of Krull dimension~$\leq 0$~\cite[Proposition~2.13]{blechschmidt:phd}.
+% FUTURE: dig out reason again
 
@@
 This stability is a consequence of the field property. In case that the
-ring~$A$ is not reduced and hence~$A^\sim$ not a field, a substitute is given
+ring~$A$ is not reduced and hence~$A^\sim$ is not a field, a substitute is given
 in Exercise~\ref{ex:local-global}. We stress that the proof of

### 2020-06-21 paper-filmat.tex
Some additional references/explanations (in response to review)

@@
 set theory \textsc{etcs}~\cite{mclarty:structuralism,marquis:foundations,barton-friedman:structures}, and models are isomorphic if and only if their
-associated toposes are equivalent as categories.}
+associated toposes are equivalent as
+categories~\cite[Section~VI.10]{moerdijk-maclane:sheaves-logic}.}
 
@@
 
-Firstly, toposes are more general than models of set theory. By definition, a
-model of \textsc{zfc} will always satisfy the axioms of \textsc{zfc}; in
-contrast, most toposes do not even validate the law of excluded middle, much
+Firstly, toposes are more general than models of set theory. Every model of set
+theory gives rise to a topos, but not every topos is induced in this way from a model of set
+theory. Unlike models of~\textsc{zfc}, most toposes do not validate the law of excluded middle, much
 less so the axiom of choice.
@@
 
-The reason for~(1) is related to the fact that no nonstandard model of
+Statement~(1) is reminiscent of the fact that no nonstandard model of
 Peano arithmetic in the standard topos is computable~\cite{tennenbaum:models}.
@@
 Let~$(f_x)_{x \in X}$ be a continuous family of
-continuous real-valued functions; that is, each of the individual
-functions~$f_x : \RR \to \RR$ should be continuous and moreover the map~$\RR
-\times X \to \RR, (a,x) \mapsto f_x(a)$ should be continuous.
+continuous real-valued functions; that is, not only should each of the individual
+functions~$f_x : \RR \to \RR$ be continuous, but the joint map~$\RR
+\times X \to \RR, (a,x) \mapsto f_x(a)$ should be continuous. (This stronger
+condition implies continuity of the individual functions.)
 From the point of view of~$\Sh(X)$, this family looks like a single continuous

### 2020-06-22 paper-filmat.tex
typo

@@
 existence of noncomputable functions theory employs nonconstructive reasoning, for if
-the proof would not go beyond intuitionistic reasoning, it would also hold internally to the
+the proof would only use intuitionistic reasoning, it would also hold internally to the
 \effective topos, in contradiction to the fact that in the \effective topos
@@
 The number~$M(X)$ computed by~$M$ depends on the input/output behaviour of~$X$,
-the chosen order of infinite-time Turing machines, and on details of the
+the chosen ordering of infinite-time Turing machines, and on details of the
 interleaving simulation and the comparison procedure -- but it does not depend on
@@
 course leave the realm of mathematics, as real-world machines are not objects
-of mathematical study, but still it is interesting to see which commitments
+of mathematical study. But it is still interesting to see which commitments
 about the nature of the physical world imply which internal statements of the
@@
 not part of today's standard setup of the reals, but they are still intriguing
-as calculational tools and as a device to bring mathematical intuition and
+as a calculational tool and as a device to bring mathematical intuition and
 mathematical formalism closer together.
@@
 
-There is a growing body of research in mathematics which employs hyperreal
+There is a small and growing body of research in mathematics which employs hyperreal
 numbers in this sense. To exemplarily cite just one example, a recent

### 2020-06-22 paper-filmat.tex
Clarification (in response to review)
typo

@@
 for which the~$n$-th Turing machine
-terminates or does not terminate. Assuming the law of excluded middle, this is
-a trivial condition; but intuitionistically, it is not.
+terminates or does not terminate. This condition is trivial only assuming the
+law of excluded middle; intuitionistically, this condition is nontrivial and
+cuts out a nontrivial subset of~$\NN$.
+
+XXX: subsets...
 \end{ex}
@@
 
-There is a small and growing body of research in mathematics which employs hyperreal
+There is a growing body of research in mathematics which employs hyperreal
 numbers in this sense. To exemplarily cite just one example, a recent

### 2020-06-30 paper-filmat.tex
Other aspects of toposes (in response to review)

@@
 the topos is custom-tailored to the object in question. This change of
-perspective has been used in mathematical practice and demonstrates that
-toposes go much beyond being logicians' testbeds. To give just a taste of what
+perspective has been used in mathematical practice.
+% and demonstrates that toposes go much beyond being logicians' testbeds.
+To give just a taste of what
 is possible, through the lens provided by an appropriate topos, any given ring
@@
 \bigskip
+\paragraph{Other aspects of toposes} This note focuses on just a single
+aspect of toposes, the view of toposes as alternate mathematical universes.
+This aspect is not the only one, nor did it historically come first.
+
+Toposes were originally conceived by Grothendieck in the early 1960s for the
+needs of algebraic geometry, as a general framework for constructing and
+studying invariants in classical and new geometric contexts, and it is in that
+subject that toposes saw their deepest applications. The proof of Fermat's Last
+Theorem is probably the most prominent such application, crucially resting on
+the cohomology and homotopy invariants provided by toposes.
+
+In the seminal work introducing
+toposes~\cite{artin-grothendieck-verdier:topos}, toposes are viewed as
+generalized kinds of spaces. Every topological space~$X$ gives rise to a topos,
+the \emph{topos of sheaves over~$X$}, and every continuous map gives rise to a
+\emph{geometric morphism} between the induced sheaf toposes, but not every
+topos is of this form. While the open sets of a topological space are required
+to be parts of the space, the opens of toposes are not; and while for open
+subsets~$U$ and~$V$ there is only a truth value as to whether~$U$ is contained
+in~$V$, in a general topos there can be many distinct ways how an open is
+contained in another one. This additional flexibility is required in situations
+where honest open subsets are rare, such as when studying the étale cohomology
+of a scheme~\cite{milne:etale}.
+
+That toposes could also be regarded as mathematical universes was realized only
+later, by Bill Lawvere and Myles Tierney at the end of the 1960s. They abstracted some of
+the most important categorical properties of Grothendieck's toposes into what
+is now known as the definition of an \emph{elementary topos}. Elementary
+toposes are considerably more general and less tied to geometry than the
+original toposes. The theory of elementary toposes has a substantially
+different, logical flavor, not least because a different notion of morphism
+plays an important role. To help disambiguate, there is a trend to rename
+elementary toposes to \emph{logoses}, but this text still follows the standard
+convention.
+
+A further perspective on toposes emerged in the early 1970s with the discovery
+that toposes can be regarded as embodiments of a certain kind of first-order
+theories, the \emph{geometric theories} briefly discussed on
+page~\pageref{item:classifying-topos}. The so-called \emph{classifying toposes}
+link geometrical and logical aspects and are fundamental to Olivia Caramello's
+bridge-building program~\cite{caramello:tst}. Geometrically, the classifying
+topos of a geometric theory~$\TT$ can be regarded as the generalized space of
+models of~$\TT$; this idea is due to Monique
+Hakim~\cite{hakim:relative-schemes}, though she did not cast her discovery in
+this language. Logically, the classifying topos of~$\TT$ can be regarded as a
+particular mathematical universe containing the \emph{generic~$\TT$-model}, a
+model which has exactly those properties which are shared by all models.
+
+Yet more views on toposes are fruitfully used --
+reference~\cite{johnstone:elephants} lists ten more -- but we shall not review
+them here.
+
+\bigskip
 \paragraph{Acknowledgments} We are grateful to Andrej Bauer, Martin
@@
 Laudahn, Matthias Hutzler and Milan Zerbin for their careful criticism of
-earlier drafts and to Todd Lehman for preparing parts of
-Figure~\ref{fig:landscape}. We thank the organizers and all participants of the
+earlier drafts and to Todd Lehman for the code for parts of
+Figure~\ref{fig:landscape}. This note also profited substantially from the
+thorough review by the three anonymous referees, whose efforts are much
+appreciated and gladly acknowledged. We thank the organizers and all participants of the
 Mussomeli conference of the Italian Network for the Philosophy in Mathematics,
@@
 algebras)~\cite{butterfield-hamilton-isham:bohr,heunen-landsman-spitters:aqt}.
-\item $\Set[\TT]$, the \emph{classifying topos} of a geometric theory~$\TT$.\footnote{A
+\item\label{item:classifying-topos}$\Set[\TT]$, the \emph{classifying topos} of a geometric theory~$\TT$.\footnote{A
 geometric theory is a theory in many-sorted first-order logic whose axioms can be put as \emph{geometric
@@
 group has property~$\varphi$'' and~``all groups have property~$\varphi$'' need
-not be equivalent. This situation is explored
+not be equivalent. This inbalance has mathematical applications and is explored
 in~\cite{blechschmidt:nullstellensatz}.}
@@
 
-FUTURE: quick overview of the several aspects of toposes (maybe at the end, as an outlook?)
 FUTURE: use the phrase "very rich and very deep" (or similarly) about set theory

### 2020-06-30 paper-filmat.tex
Clearer definition (in response to review)

@@
 
-\begin{defn}A \emph{topos} (more precisely \emph{elementary topos} or
-\emph{logos} with a natural numbers object) is a category which has all finite
+\begin{defn}A \emph{topos} is a category which has all finite
 limits, is cartesian closed, has a subobject classifier and contains a natural
-numbers object.\end{defn}
+numbers object.\footnote{More precisely, this is the definition of an
+\emph{elementary topos with a natural numbers object}. Since this definition is
+less tied to geometry than Grothendieck's (as categories of sheaves over
+sites), there is a trend to call these toposes \emph{logoses}. However, that
+term also has other uses.}\end{defn}
 

### 2020-06-30 paper-filmat.tex
Be more precise about subtoposes (in response to review)
Be more precise about infinitesimal numbers (in response to review)

@@
 force~$\varphi$ to be true. If~$\E \models \neg\varphi$, then the resulting
-topos is the trivial topos.
+topos is the trivial topos. (A subtopos is not simply a subcategory; rather, it
+is more like a certain kind of quotient category. We do not give, and for the
+purposes of this contribution do not need, further details.)
 % FUTURE: The topos~$\E[\mathbb{O}]$ contains a \emph{generic object}.
@@
 
-The idea of \emph{infinitesimal numbers} -- numbers which lie between~$-\frac{1}{n}$
-and~$\frac{1}{n}$ for any natural number~$n$ -- has a long and rich history. They are
+The idea of \emph{infinitesimal numbers} -- numbers which can be pictured as lying between~$-\frac{1}{n}$
+and~$\frac{1}{n}$ for any natural number~$n$ (though this intuition will not
+serve as their formal definition in this text) -- has a long and rich history. They are
 not part of today's standard setup of the reals, but they are still intriguing
@@
 
-The dual numbers contain infinitesimal numbers, more precisely \emph{nilsquare
-numbers}, namely all numbers of the form~$b \varepsilon$ for~$b \in \RR$, and
+The flavor of infinitesimal numbers supported
+by~$\RR[\varepsilon]/(\varepsilon^2)$ are the \emph{nilsquare numbers}, numbers
+which square to zero. The numbers~$b \varepsilon$ with~$b \in \RR$
+are nilsquare in~$\RR[\varepsilon]/(\varepsilon^2)$, and
 they are sufficient to rigorously reproduce derivative computations of

### 2020-06-30 paper-filmat.tex
Make intended readership more clear (in response to review)

@@
 We intend this contribution to be self-contained and do not assume familiarity with
-topos theory or category theory. Readers who would like to learn more details
+topos theory or category theory, having a diverse readership of philosophers of
+mathematics in mind. However, to make this text more substantial to
+categorically-inclined readers, some categorical definitions are included.
+These definitions can be skipped without impacting the main message of this contribution.
+
+Readers who would like to learn more details
 are directed to the survey of category theory by Jean-Pierre

### 2020-06-30 paper-filmat.tex
Be more modest (in response to review)
typo

@@
 
-\section{The \effective topos, a convenient home for computability theory}
+\section{The \effective topos, a universe shaped by computability}
 \label{sect:effective-topos}
@@
 A basic observation in computability theory is that there are computational tasks
-which are not solvable even for these idealized Turing machines. The paramount
+which are not solvable even for these idealized Turing machines. The premier
 example is the \emph{halting problem}: Given a Turing machine~$M$, determine
@@
 
-XXX: Kleene-Vesley topos?
+FUTURE: Kleene-Vesley topos

### 2020-06-30 paper-filmat.tex
Explain subsets in Eff better (in response to review)

@@
 topos, this definition does not give rise to a total function from~$\NN$
-to~$\NN$. The actual domain is the subset of those natural numbers~$n$
+to~$\NN$. The actual domain is the subset~$M$ of those natural numbers~$n$
 for which the~$n$-th Turing machine
@@
 
-XXX: subsets...
+Subobjects in the \effective topos are more than mere subsets; to give an
+element of~$M$ in the \effective topos, we need not only give a natural
+number~$n$ such that the~$n$-th Turing machine terminates or does not
+terminate, but also a computational witness of either case. For any particular
+numeral~$n_0 \in \NN$, there is such a witness (appealing to the law of
+excluded middle on the meta level), and hence the statement~``$n_0
+\in M$'' holds in the \effective topos. However, there is no program which
+could compute such witnesses for any number~$n$, hence the statement~``$\forall
+n\?\NN\_ n \in M$'' is not true in~$\Eff$ and hence the \effective topos does not believe~$M$
+and~$\NN$ to be the same.
 \end{ex}

### 2020-06-30 paper-filmat.tex
Improve sentence (in response to review)

@@
 
-This property is where the language of toposes derives its usefulness for
-mathematical practice from: In some cases, the easiest way to prove a result
+The language of toposes derives its usefulness for mathematical practice from
+this complexity reduction: In some cases, the easiest way to prove a result
 (about objects of the standard topos) is

### 2020-06-30 paper-filmat.tex
Final version of paper

@@
 Viewed through such a lens, a given mathematical object can have different
-properties than when considered normally. In particular, it can have
+properties than when viewed normally. In particular, it can have
 better properties for the purposes of specific applications, especially if
@@
 We intend this contribution to be self-contained and do not assume familiarity with
-topos theory or category theory, having a diverse readership of philosophers of
+topos theory or category theory, having a diverse readership of people interested in philosophy of
 mathematics in mind. However, to make this text more substantial to
@@
 
-Yet more views on toposes are fruitfully used --
-reference~\cite{johnstone:elephants} lists ten more -- but we shall not review
-them here.
+Yet more views on toposes are fruitfully employed --
+reference~\cite[pages~vii--viii]{johnstone:elephant} lists ten more -- but we shall not review
+them here. A historical survey can be found in
+reference~\cite{mclarty:history}.
 
@@
 \varphi$'' is defined by recursion on the structure of~$\varphi$ following the
-so-called \emph{Kripke--Joyal translation rules}. For instance, the rule for
-translating conjunction reads
-\[ \E \models (\alpha \wedge \beta) \qquad\text{iff}\qquad
-  \E \models \alpha \quad\text{and}\quad \E \models \beta. \]
+so-called \emph{Kripke--Joyal translation rules}. For instance, the rules for
+translating conjunction and falsity read
+\[ \begin{array}{lcl}
+  \E \models (\alpha \wedge \beta) &\qquad\text{iff}\qquad&
+  \E \models \alpha \quad\text{and}\quad \E \models \beta, \\[0.2em]
+  \E \models \bot &\qquad\text{iff}\qquad&
+  \text{$\E$ is the trivial topos}.
+\end{array} \]
 The remaining translation rules are more
@@
 statement in the language of a general version of higher-order predicate
-calculus with dependent types. In practice almost any mathematical statement
+calculus with dependent types, with a base type for each object of~$\E$ and
+with a constant of type~$X$ for each morphism~$1 \to X$ in~$\E$. In practice almost any mathematical statement
 can be interpreted in a given topos.\footnote{The main exceptions are
@@
 to~\cite{fourman:sheaf-models,streicher:forcizf,awodey-butz-simpson-streicher:bist}
-for ways around this restriction.}
+for ways around this restriction.} We refrain from giving a precise definition
+of the language here, but refer to the references~\cite[Section~7]{shulman:stack}
+and~\cite[Section~VI.7]{moerdijk-maclane:sheaves-logic} for details.
 
@@
 standard topos. This is why Figure~\ref{fig:landscape} pictures the standard
-topos twice.\footnote{There is a fine print to consider. Technically, if we
+topos twice, once as a particular topos next to others, and once as the universe
+covering the entirety of our mathematical discourse.\footnote{There is a fine print to consider. Technically, if we
 work within~\textsc{zf} or its intuitionistic cousin~\textsc{izf}, most toposes
@@
 In this regard toposes are similar to class-sized inner models in set theory.
-We believe that ``our toposes are all part of the standard topos'' is still an
-apt description of the situation.}
+We believe that the vague statement ``our toposes are all part of the standard topos'' is still an
+apt description of the situation. A possible formalization is (the trivial
+observation that) ``our toposes are all indexed categories over~$\Set$''.}
 The toposes which we can study in mathematics do not tell us

### 2020-06-30 paper-filmat.tex
One more comment (in response to review)

@@
 Just as the \effective topos provides a coherent setting for studying
-computability using a naive language, the sheaf topos~$\Sh(X)$ provides a
+computability using a naive element-based language, the sheaf topos~$\Sh(X)$ provides a
 coherent setting for studying continuous~$X$-indexed families of objects (sets,
 numbers, functions) as if they were single objects.
-% FUTURE: Explain better.
-% FUTURE: History, sites, flexibility
+
+Sheaf toposes take up a special place in the history of topos theory: If the
+base~$X$ is allowed to be a \emph{site} instead of a topological space, the
+resulting toposes constitute the large class of Grothendieck toposes, the
+original notion of toposes. Categorically, the passage from topological spaces
+to sites is rather small, but the resulting increase in flexibility is
+substantial and fundamental to modern algebraic geometry.
 

### 2020-07-17 paper-filmat.tex
Final version

@@
-\documentclass[oneside,reqno]{amsart}
+%\documentclass[oneside,reqno]{amsart}
+\documentclass[graybox]{svmult}
+
+\usepackage{type1cm}
+\usepackage{graphicx}
+\usepackage{multicol}
+\usepackage[bottom]{footmisc}
+\usepackage{newtxtext}
+\usepackage{newtxmath}
 
 \usepackage[utf8]{inputenc}
-\usepackage{amsthm,mathtools,amssymb,graphicx,ragged2e,extpfeil}
+\usepackage{mathtools,ragged2e}
 \usepackage{tikz}
@@
 
-\usepackage[natbib=true,style=numeric,maxnames=10]{biblatex}
-\usepackage[babel]{csquotes}
-\bibliography{paper-filmat.bib}
+\usepackage{natbib}
 
@@
 
-\title[]{Exploring mathematical objects from custom-tailored mathematical universes}
-\author{Ingo Blechschmidt}
-\address{Universität Augsburg \\
-Institut für Mathematik \\
-Universitätsstr. 14 \\
-86159 Augsburg, Germany}
-\email{ingo.blechschmidt@math.uni-augsburg.de}
-
-\theoremstyle{definition}
-\newtheorem{defn}{Definition}[section]
-\newtheorem{ex}[defn]{Example}
-
-\theoremstyle{plain}
-\newtheorem{prop}[defn]{Proposition}
-\newtheorem{cor}[defn]{Corollary}
-\newtheorem{lemma}[defn]{Lemma}
-\newtheorem{thm}[defn]{Theorem}
-\newtheorem{scholium}[defn]{Scholium}
-
-\theoremstyle{remark}
-\newtheorem{rem}[defn]{Remark}
-\newtheorem{question}[defn]{Question}
-\newtheorem{speculation}[defn]{Speculation}
-\newtheorem{caveat}[defn]{Caveat}
-\newtheorem{conjecture}[defn]{Conjecture}
+%\theoremstyle{definition}
+%\newtheorem{defn}{Definition}[section]
+%\newtheorem{ex}[defn]{Example}
+%
+%\theoremstyle{plain}
+%\newtheorem{prop}[defn]{Proposition}
+%\newtheorem{cor}[defn]{Corollary}
+%\newtheorem{lemma}[defn]{Lemma}
+%\newtheorem{thm}[defn]{Theorem}
+%\newtheorem{scholium}[defn]{Scholium}
+%
+%\theoremstyle{remark}
+%\newtheorem{rem}[defn]{Remark}
+%\newtheorem{question}[defn]{Question}
+%\newtheorem{speculation}[defn]{Speculation}
+%\newtheorem{caveat}[defn]{Caveat}
+%\newtheorem{conjecture}[defn]{Conjecture}
 
@@
 
-\newextarrow{\xbigtoto}{{20}{20}{20}{20}}{\bigRelbar\bigRelbar{\bigtwoarrowsleft\rightarrow\rightarrow}}
-\newcommand{\E}{\mathcal{E}}
+%\newextarrow{\xbigtoto}{{20}{20}{20}{20}}{\bigRelbar\bigRelbar{\bigtwoarrowsleft\rightarrow\rightarrow}}
+\renewcommand{\E}{\mathcal{E}}
 \newcommand{\B}{\mathcal{B}}
@@
 
-%\begin{abstract}
-%  FUTURE: abstract
-%\end{abstract}
+\title*{Exploring mathematical objects from custom-tailored mathematical universes}
+\author{Ingo Blechschmidt}
+\institute{Ingo Blechschmidt \at Universität Augsburg, Institut für Mathematik,
+Universitätsstr. 14,
+86159 Augsburg, Germany, \email{ingo.blechschmidt@math.uni-augsburg.de}}
 
 \maketitle
-\thispagestyle{empty}
+
+\abstract{
+  Toposes can be pictured as mathematical universes. Besides the standard topos,
+  in which most of mathematics unfolds, there is a colorful host of alternate
+  toposes in which mathematics plays out slightly
+  differently.
+  For instance, there are toposes in which the axiom of choice and the
+  intermediate value theorem from undergraduate calculus fail.
+  The purpose of this contribution is to give a glimpse of the toposophic
+  landscape, presenting several specific toposes and exploring their peculiar
+  properties, and to explicate how toposes provide distinct lenses through
+  which the usual mathematical objects of the standard topos can be viewed.
+  \emph{Key words:} topos theory, realism debate, well-adapted language,
+  constructive mathematics
+}
+\medskip
 
@@
 \begin{figure}[b]
+  \centering
   \tikzstyle{topos} = [draw=mypurple, very thick, rectangle, rounded corners, inner sep=5pt, inner ysep=10pt]
@@
 on mathematical objects provided by any topos uncovers relations which
-otherwise remain hidden.\footnote{Olivia Caramello's research
-program~\cite{caramello:tst} provides a further topos-theoretic way for
+otherwise remain hidden.\footnote{The research
+program put forward by~\cite{caramello:tst} provides a further topos-theoretic way for
 uncovering hidden relations, though not between objects but between mathematical
@@
 Readers who would like to learn more details
-are directed to the survey of category theory by Jean-Pierre
-Marquis~\cite{sep:category-theory} and to a gentle introduction to topos theory
-by Tom Leinster~\cite{leinster:introduction}. Standard references for the internal
+are directed to the survey of category theory by~\cite{sep:category-theory}
+and to a gentle introduction to topos theory
+by~\cite{leinster:introduction}. Standard references for the internal
 language of toposes include~\cite[Chapter~VI]{moerdijk-maclane:sheaves-logic},
@@
 In the seminal work introducing
-toposes~\cite{artin-grothendieck-verdier:topos}, toposes are viewed as
+toposes by~\cite{artin-grothendieck-verdier:topos}, toposes are viewed as
 generalized kinds of spaces. Every topological space~$X$ gives rise to a topos,
@@
 where honest open subsets are rare, such as when studying the étale cohomology
-of a scheme~\cite{milne:etale}.
+of a scheme as in~\cite{milne:etale}.
 
@@
 link geometrical and logical aspects and are fundamental to Olivia Caramello's
-bridge-building program~\cite{caramello:tst}. Geometrically, the classifying
+bridge-building program set out in~\cite{caramello:tst}. Geometrically, the classifying
 topos of a geometric theory~$\TT$ can be regarded as the generalized space of
-models of~$\TT$; this idea is due to Monique
-Hakim~\cite{hakim:relative-schemes}, though she did not cast her discovery in
+models of~$\TT$; this idea is due to~\cite{hakim:relative-schemes}, though she did not cast her discovery in
 this language. Logically, the classifying topos of~$\TT$ can be regarded as a
@@
 Yet more views on toposes are fruitfully employed --
-reference~\cite[pages~vii--viii]{johnstone:elephant} lists ten more -- but we shall not review
-them here. A historical survey can be found in
-reference~\cite{mclarty:history}.
+\cite[pages~vii--viii]{johnstone:elephant} lists ten more -- but we shall not review
+them here. A historical survey was compiled by~\cite{mclarty:history}.
 
@@
 Figure~\ref{fig:landscape}. This note also profited substantially from the
-thorough review by the three anonymous referees, whose efforts are much
+thorough review by three anonymous referees, whose efforts are much
 appreciated and gladly acknowledged. We thank the organizers and all participants of the
@@
 
-\tableofcontents
+%\tableofcontents
 
@@
 
-\begin{defn}A \emph{topos} is a category which has all finite
+\begin{definition}A \emph{topos} is a category which has all finite
 limits, is cartesian closed, has a subobject classifier and contains a natural
@@
 sites), there is a trend to call these toposes \emph{logoses}. However, that
-term also has other uses.}\end{defn}
+term also has other uses.}\end{definition}
 
@@
 
-\begin{defn}The \emph{standard topos}~$\Set$ is the category which has all
-sets as its objects and all maps between sets as morphisms.\end{defn}
+\begin{definition}The \emph{standard topos}~$\Set$ is the category which has all
+sets as its objects and all maps between sets as morphisms.\end{definition}
 
@@
 The remaining translation rules are more
-involved~\cite[Section~VI.7]{moerdijk-maclane:sheaves-logic}; we do not list them here for
+involved, as detailed by~\cite[Section~VI.7]{moerdijk-maclane:sheaves-logic}; we do not list them here for
 the case of a general topos~$\E$, but we will state them in the next sections
@@
 % FUTURE:
-% \begin{rem}It might be helpful to picture mathematicians in a topos~$\E$. For
+% \begin{remark}It might be helpful to picture mathematicians in a topos~$\E$. For
 % exchanging ideas, they use the same mathematical language like we do, but their
@@
 % to switch on the Kripke--Joyal translation engine their claims start making
-% sense.\end{rem}
+% sense.\end{remark}
 
@@
 
-\begin{thm}\label{thm:reasoning}Let~$\E$ be a topos and let~$\varphi$ be a
+\begin{theorem}\label{thm:reasoning}Let~$\E$ be a topos and let~$\varphi$ be a
 statement such that~$\E \models \varphi$. If~$\varphi$ intuitionistically
@@
 intuitionistic logic that~$\varphi$ entails~$\psi$), then~$\E \models
-\psi$.\end{thm}
+\psi$.\end{theorem}
 
@@
 
-\begin{defn}A topos~$\E$ is \emph{boolean} if and only if the laws of classical
-logic are true in~$\E$.\end{defn}
+\begin{definition}A topos~$\E$ is \emph{boolean} if and only if the laws of classical
+logic are true in~$\E$.\end{definition}
 
@@
 
-\begin{rem}The axiom of choice (which is strictly speaking not part of
+\begin{remark}The axiom of choice (which is strictly speaking not part of
 classical logic, but of classical set theory) is also not available in most
@@
 excluded middle in presence of other axioms which are available in any topos.
-\end{rem}
+\end{remark}
 
@@
 believes to be maps. The topos~$\Set_M$ validates the axioms of the structural
-set theory \textsc{etcs}~\cite{mclarty:structuralism,marquis:foundations,barton-friedman:structures}, and models are isomorphic if and only if their
+set theory \textsc{etcs}, see~\cite{mclarty:structuralism,marquis:foundations,barton-friedman:structures}, and models are isomorphic if and only if their
 associated toposes are equivalent as
-categories~\cite[Section~VI.10]{moerdijk-maclane:sheaves-logic}.}
-
-\begin{ex}The topos~$\Set_V$ associated to the universe~$V$ of all sets (if
+categories, see~\cite[Section~VI.10]{moerdijk-maclane:sheaves-logic}.}
+
+\begin{example}The topos~$\Set_V$ associated to the universe~$V$ of all sets (if
 this structure is available in one's chosen ontology) coincides with the
-standard topos~$\Set$.\end{ex}
+standard topos~$\Set$.\end{example}
 
@@
 regarded as classical mechanical systems (which are described by commutative
-algebras)~\cite{butterfield-hamilton-isham:bohr,heunen-landsman-spitters:aqt}.
+algebras). Details are described by~\cite{butterfield-hamilton-isham:bohr,heunen-landsman-spitters:aqt}.
 \item\label{item:classifying-topos}$\Set[\TT]$, the \emph{classifying topos} of a geometric theory~$\TT$.\footnote{A
@@
 moderate logicism (because, as a topos, it supports an intuitionistic type
-theory)~\cite{lambek:incompatible,couture-lambek:reflections}.
+theory). Details are described by~\cite{lambek:incompatible,couture-lambek:reflections}.
 \end{enumerate}
@@
 and for instance excluding any predicative flavors of mathematics
-(Laura Crosilla's survey~\cite{crosilla:predicativity} is an excellent introduction).
+(Laura Crosilla's survey in~\cite{crosilla:predicativity} is an excellent introduction).
 Toposes are impredicative in the sense that any object of a topos is required to have a
@@
 consideration by Maietti and
-Vickers~\cite{maietti:au,maietti-vickers:induction,vickers:sketches}.}
+Vickers, see~\cite{maietti:au,maietti-vickers:induction,vickers:sketches}.}
 
@@
 
-\begin{rem}The picture of a topos-theoretic multiverse is related to Hamkin's
-multiverse view in set theory~\cite{hamkins:multiverse}. In fact, the
+\begin{remark}The picture of a topos-theoretic multiverse is related to Hamkin's
+multiverse view in set theory as put forward in~\cite{hamkins:multiverse}. In fact, the
 topos-theoretic multiverse can be regarded as an extension of the set-theoretic
@@
 has, as of yet, no counterpart in topos theory: namely a systematic study of
-its modal logic with respect to various notions of relations between toposes.\end{rem}
+its modal logic with respect to various notions of relations between toposes.\end{remark}
 
@@
 Introductory literature on the \effective topos
-includes~\cite{hyland:effective-topos,oosten:realizability,phoa:effective,bauer:c2c}.
-
-\begin{defn}\begin{enumerate}
+includes the references~\cite{hyland:effective-topos,oosten:realizability,phoa:effective,bauer:c2c}.
+
+\begin{definition}\ \\[-1.2em]\begin{enumerate}
 \item An \emph{assembly} is a set~$X$ together with a
@@
 that~$n \realizes x$, the computation~$M(n)$ terminates and~$M(n) \realizes
-f(x)$.\end{enumerate}\end{defn}
+f(x)$.\end{enumerate}\end{definition}
 
@@
 
-\begin{defn}The \emph{\effective topos}~$\Eff$ is the ex/reg completion (as
-in~\cite[Section~3.4]{menni:exact}) of the category of assemblies.\end{defn}
+\begin{definition}The \emph{\effective topos}~$\Eff$ is the ex/reg completion (as
+in~\cite[Section~3.4]{menni:exact}) of the category of assemblies.\end{definition}
 
@@
 
-\begin{ex}\textbf{``Any natural number is prime or not.''} Even without knowing what
+\begin{example}\textbf{``Any natural number is prime or not.''} Even without knowing what
 a prime number is, one can safely judge this statement to be true in
@@
 algorithmically tested.
-\end{ex}
-
-\begin{ex}\textbf{``There are infinitely many primes.''} A first-order formalization
+\end{example}
+
+\begin{example}\textbf{``There are infinitely many primes.''} A first-order formalization
 of this statement is ``for any natural number~$n$, there is a prime
@@
 redundant.}
-\end{ex}
-
-\begin{ex}\textbf{``Any function~$\NN \to \NN$ is constantly zero or not.''} Precisely,
+\end{example}
+
+\begin{example}\textbf{``Any function~$\NN \to \NN$ is constantly zero or not.''} Precisely,
 the statement is
@@
 existence to the undecidability of the halting problem.
-\end{ex}
-
-\begin{rem}Quite surprisingly, there are infinite sets~$X$ for which any flavor of constructive
+\end{example}
+
+\begin{remark}Quite surprisingly, there are infinite sets~$X$ for which any flavor of constructive
 mathematics, in particular the kind which is valid in any topos, verifies the
@@
 \NN_\infty$ of the naturals. This phenomenon has been thoroughly explored by
-Martin Escardó~\cite{escardo:omniscience}.\end{rem}
-
-\begin{ex}\textbf{``Any function~$\NN \to \NN$ is computable.''}
+Escardó~\cite{escardo:omniscience}.\end{remark}
+
+\begin{example}\textbf{``Any function~$\NN \to \NN$ is computable.''}
 The preceding examples give the impression that what is true in the
@@
 and~$\NN$ to be the same.
-\end{ex}
-
-\begin{ex}\textbf{``Any function~$\RR \to \RR$ is continuous.''}\label{sect:eff-continuous}
+\end{example}
+
+\begin{example}\textbf{``Any function~$\RR \to \RR$ is continuous.''}\label{sect:eff-continuous}
 In the standard topos, this statement is plainly false, with the sign and Heaviside step functions
 being prominent counterexamples. In the \effective topos, this statement is
-true and independently due to
-Kreisel--Lacombe--Shoenfield~\cite{kreisel-lacombe-shoenfield:cont} and
-Ceĭtin~\cite{ceitin:cont}. A rigorous proof is not entirely
+true and independently due to~\cite{kreisel-lacombe-shoenfield:cont} and~\cite{ceitin:cont}. A rigorous proof is not entirely
 straightforward (a textbook reference
@@
 precision in the input~$x$, does not exist in the \effective topos.
-\end{ex}
-
-\begin{ex}\textbf{``Markov's principle holds.''} Markov's principle is the
+\end{example}
+
+\begin{example}\textbf{``Markov's principle holds.''} Markov's principle is the
 following statement:
@@
 with output~$n$ as soon as it determines that some function value~$f(n)$ is zero.
-\end{ex}
-
-\begin{ex}\textbf{``Heyting arithmetic is categorical.''} In addition to the standard
+\end{example}
+
+\begin{example}\textbf{``Heyting arithmetic is categorical.''} In addition to the standard
 model~$\NN$, the standard topos contains uncountably many nonstandard models of
 Peano arithmetic (at least if we assume a classical meta level). By a theorem
-of Benno van den Berg and Jaap van Oosten, the situation
-is quite different in the \effective topos~\cite{berg-oosten:arithmetic}:
+of~\cite{berg-oosten:arithmetic}, the situation
+is quite different in the \effective topos:
 \begin{enumerate}
@@
 
-Statement~(1) is reminiscent of the fact that no nonstandard model of
-Peano arithmetic in the standard topos is computable~\cite{tennenbaum:models}.
-\end{ex}
+Statement~(1) is reminiscent of the fact due to~\cite{tennenbaum:models} that no nonstandard model of
+Peano arithmetic in the standard topos is computable.
+\end{example}
 
 \begin{table}
+  \centering
   \begin{framed}\begin{tabbing}
@@
 Firstly, in place of ordinary Turing machines, one can employ the
-\emph{infinite-time Turing machines} pioneered by Hamkins and
-Lewis~\cite{hamkins-lewis:ittm}. These machines model \emph{hypercomputation}
+\emph{infinite-time Turing machines} pioneered by~\cite{hamkins-lewis:ittm}. These machines model \emph{hypercomputation}
 in that they can run for ``longer than infinity''; more precisely, their
@@
 there is no surjection~$\NN \to \RR$, there is an injection~$\RR \to
-\NN$~\cite{bauer:injection}.\footnote{What the realizability topos
+\NN$~\cite[Section~4]{bauer:injection}.\footnote{What the realizability topos
 given by infinite-time Turing machines believes to be a real number is, from
@@
 
-For instance, Andrej Bauer showed that inside this topos any function~$\RR \to
+For instance, \cite{bauer:int-mathematics} showed that inside this topos any function~$\RR \to
 \RR$ is continuous if, in the physical world, only finitely many computational
 steps can be carried out in finite time and if it is possible to form
-tamper-free private communication channels~\cite{bauer:int-mathematics}.
+tamper-free private communication channels.
 
@@
 \begin{table}
+  \centering
   \begin{framed}\begin{tabbing}
@@
 \begin{figure}
+  \centering
   \includegraphics[height=3cm]{trichotomy-1}
@@
 \begin{figure}
+  \centering
   \marginnote{
     \begin{ocg}[printocg=never]{layername}{layerid1}{on}
-    \vspace*{-12em}
+      \vspace*{-14em}
       \scalebox{0.2}{
@@
       \begin{minipage}{3cm}\scriptsize\noindent
-    In the online version of this contribution, a video of the
-    continuous family can be seen here.\end{minipage}
-  \end{ocg}}
+      In the online version of this book, a video of the
+      continuous family is shown here.\end{minipage}
+    \end{ocg}
+  }
   \includegraphics[height=3cm]{zeros-in-families-still-1}
@@
 
-\begin{ex}
+\begin{example}
 From these observations we can deduce that the intermediate value theorem of
@@
 continuous fashion.}
-\end{ex}
-
-\begin{ex}The fundamental theorem of algebra generally fails in~$\Sh(X)$, even
+\end{example}
+
+\begin{example}The fundamental theorem of algebra generally fails in~$\Sh(X)$, even
 for quadratic polynomials. What~$\Sh(X)$ believes to be a (Dedekind)
@@
 that such a function does not exist if~$0 \in U_i$.
-\end{ex}
-
-\begin{ex}The standard proof of Banach's fixed point theorem employs only
+\end{example}
+
+\begin{example}The standard proof of Banach's fixed point theorem employs only
 intuitionistic reasoning, hence applies internally to~$\Sh(X)$. Interpreting the
@@
 depend continuously on parameters.
-\end{ex}
+\end{example}
 
@@
 numbers in this sense. To exemplarily cite just one example, a recent
-application of nonstandard analysis in symplectic geometry is due to Oliver
-Fabert~\cite{fabert:non-squeezing,fabert:floer}, who verified an
+application of nonstandard analysis in symplectic geometry is due to~\cite{fabert:non-squeezing,fabert:floer}, who verified an
 infinite-dimensional analogue of the Arnold conjecture.
@@
 they are also not at all canonical structures: \textsc{zfc} proves that there
-are~$2^{2^{\aleph_0}}$ many~\cite{pospisil:ultrafilters}.
+are~$2^{2^{\aleph_0}}$ many, see~\cite{pospisil:ultrafilters}.
 
@@
 provides several constructive alternatives for realizing infinitesimals.
-One such is ``cheap nonstandard analysis'' by Terence Tao~\cite{tao:cheap-nsa}.
+One such is ``cheap nonstandard analysis'' by~\cite{tao:cheap-nsa}.
 It is to Robinson's nonstandard analysis what potential infinity is
@@
 The following section presents a (variant of a) topos used in \emph{synthetic differential
-geometry}~\cite{kock:sdg,kock:new-methods}. This subject is a further topos-theoretic
+geometry} as discussed by~\cite{kock:sdg,kock:new-methods}. This subject is a further topos-theoretic
 approach to infinitesimals which is suited to illustrate the philosophy of toposes
@@
 For reference purposes only, we include the precise definition of the Zariski topos.
-\begin{defn}The \emph{Zariski topos of~$\RR$},~$\Zar(\RR)$, is a certain full subcategory of
+\begin{definition}The \emph{Zariski topos of~$\RR$},~$\Zar(\RR)$, is a certain full subcategory of
 the category of functors from finitely presented~$\RR$-algebras to sets, namely of
@@
 (this notion is defined in Table~\ref{table:zariski}), the diagram
-\[ F(A) \longrightarrow \prod_i F(A[f_i^{-1}]) \xbigtoto{} \prod_{j,k}
+\[ F(A) \rightarrow \prod_i F(A[f_i^{-1}]) \rightrightarrows \prod_{j,k}
 F(A[(f_jf_k)^{-1}]) \]
 is an equalizer diagram. The object~$\RR^\sim$ of~$\Zar(\RR)$ is the tautologous functor~$A \mapsto A$.
-\end{defn}
+\end{definition}
 
 \begin{table}
+  \centering
   \begin{framed}
@@
 
-\begin{rem}The Zariski topos can also be set up with an arbitrary commutative
+\begin{remark}The Zariski topos can also be set up with an arbitrary commutative
 ring~$S$ in place of~$\RR$. The resulting topos~$\Zar(S)$ contains a mirror
@@
 machinery is eliminated by unrolling the appropriate
-definitions~\cite{blechschmidt:generic-freeness}.\footnote{This contribution is
+definitions as in~\cite{blechschmidt:generic-freeness}.\footnote{This contribution is
 not the proper place for an exposition of Grothendieck's generic freeness
@@
 field is \notnot finite free.}
-\end{rem}
+\end{remark}
 
@@
 
-\begin{rem}The ring~$\RR^\sim$ of smooth numbers does not coincide with the Cauchy reals,
+\begin{remark}The ring~$\RR^\sim$ of smooth numbers does not coincide with the Cauchy reals,
 the Dedekind reals or indeed any well-known construction of the reals
@@
 nilsquare number in any flavor of the reals is zero.
-\end{rem}
+\end{remark}
 
@@
 
-\begin{defn}\label{defn:well-adapted}A \emph{well-adapted model} of synthetic
+\begin{definition}\label{defn:well-adapted}A \emph{well-adapted model} of synthetic
 differential geometry is a topos~$\E$ together with a ring~$\RR^\sim$ in~$\E$ such
@@
 \end{enumerate}
-\end{defn}
+\end{definition}
 
 It is the culmination of a long line of research by several authors that
-several well-adapted models of synthetic differential geometry exist~\cite{moerdijk-reyes:models}. By the
+several well-adapted models of synthetic differential geometry exist, see~\cite{moerdijk-reyes:models}. By the
 conditions imposed in Definition~\ref{defn:well-adapted}, for any such
@@
 
-\printbibliography
+\bibliographystyle{chicago}
+\bibliography{paper-filmat}
 

### 2020-11-02 paper-herrsching2019/generalized-spaces.tex
typo

@@
 Common to all of the mentioned flavors of space is that \emph{points are their
-building blocks}: Cartesian, metric, topological and diffeological spaces are,
+building blocks}: cartesian, metric, topological and diffeological spaces are,
 first and foremost, sets of points. Their spatial structure -- a metric, a
@@
 set-based models, but this theory is consistent and it does have nontrivial sheaf
-models. For the purpose of constructive algebra, sheaf models of some geometric
+models. For the purposes of constructive algebra, sheaf models of some geometric
 theories play a vital role, as we will discuss in Section~\ref{sect:appl}.
@@
 Section~\ref{sect:example-applications}, which is shorter and arguably more
-perspicuous than all previously known classical proofs.
+perspicuous than the previously known classical proofs.
 
@@
 Regarding sheaf semantics, Section~\ref{sect:sheaf-semantics} only presents
-the first-order case, but the theory can be extended include unbounded
+the first-order case, but the theory can be extended to include unbounded
 quantification~\cite{shulman:stack-semantics}, function types and power types
@@
 The formal definition of a locale will be given below as
-Definition~\ref{defn:locale}, but we first want to review examples and comment on
+Definition~\ref{defn:locale}, but we first review examples and comment on
 the relevance of locales to constructive mathematics.
@@
 locale can be nontrivial even if it does not contain any points. This
-phenomenon is one instance of a general guiding principle, namely that relinquishing
+phenomenon is an instance of a general guiding principle, namely that relinquishing
 points increases flexibility.
@@
 ranges over the naturals and~$x$ ranges over the reals. We picture~$U_{nx}$ as
-the ``open of those surjections~$f$ for which~$f(n) = x$'', we can compute with
+the ``open of those surjections~$f$ for which~$f(n) = x$'' and we can compute with
 these opens, consider functions on~$X$ and ponder sublocales of~$X$. For instance:
@@
 Despite this fundamental issue, one can still argue that this approach has
-merit~\cite{simpson:randomness}, and hence it is worthwhile to find a
+merit~\cite{simpson:measure}, and hence it is worthwhile to find a
 mathematical foundation which is capable of formalizing it. Topological
@@
 \end{array} \]
-and finite meets distribute over arbitrary joins:
+and where finite meets distribute over arbitrary joins:
 \[ U \wedge \bigvee_i V_i = \bigvee_i (U \wedge V_i). \]
@@
 of~$X$ can be pictured as mapping each open~$U$ of~$X$ to the truth value to which
-extent~$x$ belongs to~$U$. Symbolically, we write~$x \inplus U$ iff~$\alpha(U)
+extent~$x$ belongs to~$U$. Mnemonically, we write~$x \inplus U$ iff~$\alpha(U)
 = \top$.
@@
 (Exercise~\ref{ex:hausdorff-sober}), hence the spaces of a wide
-range of the mathematical landscape can be studied as locales.\end{remark}
+range of the mathematical landscape can be faithfully studied as locales.\end{remark}
 
@@
 An open~$U$ of a locale is \emph{dense} iff for any open~$V$, $U \wedge V =
-\bot$ implies~$U = \bot$.
+\bot$ implies~$V = \bot$.
 \begin{alphlist}[(b)]
@@
 Table~\ref{table:geometric-logic}.\footnote{Because of the infinitary
-disjunctions, some care is needed to not ``built in'' the axiom of choice in
+disjunctions, some care is needed to not ``build in'' the axiom of choice in
 the definition of derivations. More specifically, we allow inhabited sets of
@@
 
-The Nullstellensatz is a theorem of classical mathematics, employing the
+The Nullstellensatz is a theorem of classical mathematics whose usual proof employs the
 axiom of choice. Since it can be formulated as a geometric sequent over the
@@
 \models \varphi$.
-\item If~$\TT$ proves~$\alpha \dashv\vdash \bigvee_i \beta_i$ (geometrically) and~$\beta_i \models \varphi$ for
+\item If~$\TT$ proves~$\alpha \dashv\vdash \bigvee_i \beta_i$ (geometrically) and if~$\beta_i \models \varphi$ for
 all indices~$i$, then~$\alpha \models \varphi$.
@@
 constructive metatheory, we should rather write~$\bigvee_{i \in I} M_i =
-[\bigvee_{\varphi \in K} \varphi]$, where~$K = \{ \varphi \,|\, \exists i \in
-I\+ \varphi \in M_i \}$.} we can regard the Lindenbaum algebra as the free
+[\bigvee_{\varphi \in \bigcup_i M_i} \varphi]$.}
+we can regard the Lindenbaum algebra as the free
 frame generated by the nullary relation symbols of~$\TT$ modulo the axioms
@@
 reals. In classical mathematics, employing their localic counterparts is merely
-for convenience or for aesthetic reasons; in constructive mathematics, the
-localic replacements (or similar replacements in other pointfree approaches to
-topology) are vital.
+for convenience or for aesthetic reasons; in constructive mathematics,
+localic replacement (or similar replacements in other pointfree approaches to
+topology) is vital.
 
@@
 empty open~$V = \bot$,~$s|^U_V$ does not have property~$\varphi$ for any
-open~$V \preceq U$'', where again~``$\varphi(s)$'' has to be understood in a
+open~$V \preceq U$'', where again~``$\varphi(s|^U_V)$'' has to be understood in a
 sensible way.
@@
 Let~$X$ be a locale. A \emph{formula over an open~$U \in \OOO(X)$} is an
-infinitary first-order formula (made up using~${=}\ {\top}\ {\wedge}\ \{\bigwedge\}\ {\bot}\ {\vee}\
+infinitary first-order formula (made up using~${=}\ {\top}\ {\wedge}\ {\bigwedge}\ {\bot}\ {\vee}\
 {\bigvee}\ {\Rightarrow}\ {\forall}\ {\exists}$) over the signature which has one sort for
@@
 formulas what it means for a formula~$\varphi$ over an open~$U$ to be
-\emph{forced by~$U$} or \emph{true over~$U$}, written~``$U \models \varphi$''. The
+\emph{forced by~$U$} or to be \emph{true over~$U$}, written~``$U \models \varphi$''. The
 main differences with the perhaps more familiar Kripke semantics are marked.
@@
 function exists), but the interpretation of~``$\neg\neg(\exists s\?F\+ \top)$''
-is true: Unrolling the definition, this statement expresses that
+is true: Unrolling the definition, this statement expresses that\par
 \begin{quote}
@@
 sheaf model over the one-point locale, by employing the constant sheaf
-construction reviewed on page~\ref{page:constant-sheaf}. An infinitary
+construction reviewed on page~\pageref{page:constant-sheaf}. An infinitary
 first-order formula holds for~$M$ iff it holds for the induced sheaf model.
@@
 \emph{sheaf model} of a propositional geometric theory~$\TT$ over a locale~$X$
-consists of a family~$(U_\varphi)_\varphi$ of opens of~$X$, one for each
+consists of a family~$(U_\varphi)_\varphi$ of opens of~$X$, one open for each
 nullary relation symbol~$\varphi$ of~$\TT$, in such a way that the axioms are
@@
 \begin{enumerate}
-\item If~$M$ validates~$\sigma$, then so does~$M_x$, for any point~$x \in \Pt(X)$.
+\item If~$M$ validates~$\sigma$, then so does~$M_x$ for any point~$x \in \Pt(X)$.
 \item The converse holds if~$X$ is spatial.
@@
 \item All sheaf models of~$\TT$ on arbitrary locales validate~$\sigma$.
-\item The theory~$\TT$ proves~$\sigma$:
+\item The theory~$\TT$ proves~$\sigma$.
 \end{enumerate}
@@
 \item Extend the geometric theory of objects
-(page~\ref{item:theory-of-objects}) by relation symbols and axioms, but not by
+(page~\pageref{item:theory-of-objects}) by relation symbols and axioms, but not by
 constant symbols, to ensure that the set-based models are precisely the
@@
 signify that the elements~$x_1,\ldots,x_n$ exhaust all elements of~$X$. A
-solution is given in Ref.~\cite[Example~D1.1.7(k)]{johnstone:elephant}.\par}
+solution to this exercise is given in Ref.~\cite[Example~D1.1.7(k)]{johnstone:elephant}.\par}
 \end{alphlist}
@@
 (X-x_1)\cdots(X-x_n)$ over~$L$. Let~$x$ be one of the zeros of~$f$. Show that
-the theory shows~$\top \vdash \bigvee_{i=1}^n \sigma_{x,x_i}$.
+the theory proves~$\top \vdash \bigvee_{i=1}^n \sigma_{x,x_i}$.
 \item Now furthermore assume that~$L$ is \emph{Galois over}~$k$ in the sense that
@@
 \begin{enumerate}
-\item Instead of replacing a given ring with another (for instance a ring~$A$ with
+\item Instead of replacing a given ring with another ring (for instance a ring~$A$ with
 one of its stalks~$A_\ppp$ at a minimal prime ideal), we replace a given ring
@@
 
-A ring is \emph{reduced} iff zero is its only nilpotent; that is, if~$x^n = 0$
+A ring is \emph{reduced} iff zero is its only nilpotent element; that is, if~$x^n = 0$
 implies~$x = 0$.
@@
 An ideal~$\aaa \subseteq A$ is a \emph{radical ideal} iff~$x^n \in \aaa$
-implies~$x \in \aaa$ for any~$x \in A$ and natural number~$n$. The
+implies~$x \in \aaa$ for any~$x \in A$ and any natural number~$n$. The
 \emph{radical}~$\sqrt{\aaa}$ of an arbitrary ideal~$\aaa$ is the radical
@@
 whenever~$a + b$ is invertible, $a$ is invertible or~$b$ is invertible.
-This definition is an elementary rephrasing of the textbook definition,
+This definition is an elementary rephrasing of the textbook definition
 according to which a ring is local iff it has exactly one maximal ideal. The
@@
 \paragraph{Free constructions.}
-Let~$G$ be a group. Then there is a universal way to turn~$G$ into an abelian
+Let~$G$ be a group. Then there is a universal way of turning~$G$ into an abelian
 group~$G^\ab$ equipped with a group homomorphism~$G \to G^\ab$, the \emph{free
@@
 commutative. In textbooks, the abelianization is usually constructed as~$G^\ab
-= G/[N,N]$, that is as the quotient by the subgroup generated by the
+= G/[G,G]$, that is as the quotient by the subgroup generated by the
 commutators. This explicit construction makes it appear that abelianization
@@
 
-For instance, in the case~$X = \{ t \}$, the field~$\QQ(t)$ of rational
+For instance, in the case~$M = \{ t \}$, the field~$\QQ(t)$ of rational
 functions comes close, but fails because there are no field homomorphisms~$\QQ(t)
@@
 homomorphism~$f : A \to B$ into a local ring be given. The subset~$B^\times
-\subseteq B$ is then a \emph{prime filter}, due to the locality of~$B$. Its
+\subseteq B$ is a \emph{prime filter}, due to the locality of~$B$. Its
 preimage~$\fff \defeq f^{-1}[B^\times]$ is a prime filter of~$A$. Hence the
@@
 free local ring over~$A$ exists as a certain sheaf of rings.} It has a
-universal property not only with respect to all ordinary rings, but also all
+universal property not only with respect to all ordinary rings, but also with respect to all
 sheaves of rings on arbitrary locales.
@@
 Since this theory is not propositional, Section~\ref{sect:presenting-frames}
-does not apply. There is, however, still a classifying topos of this theory, and it turns
+does not apply. There is, however, still a classifying \emph{topos} of this theory, and it turns
 out that it coincides with the topos of sheaves over~$\Spec(A)$ -- mainly
@@
 
-To this end, let an element~$f \in A$ and a family~$(g_i)_i$ of elements be
+To this end, let an element~$f \in A$ and a family~$(g_i)_{i \in I}$ of elements be
 given and assume~$\sqrt{(f)} \subseteq \sqrt{\sum_i \sqrt{(g_i)}}$. Then there
-are values~$n,m \in \NN$, $i_1,\ldots,i_m \in \NN$ and~$u_1,\ldots,u_m \in
+are values~$n,m \in \NN$, $i_1,\ldots,i_m \in I$ and~$u_1,\ldots,u_m \in
 A$ such that~$f^n = u_1 g_{i_1} + \cdots + u_m g_{i_m}$. We then have the
@@
 then the theory is consistent. This observation can be regarded as a finitary
-substitute for the statement (equivalent to~\BPIT in general) ``nontrivial rings contain prime filters''.
+substitute of the statement (equivalent to~\BPIT in general) ``nontrivial rings contain prime filters''.
 \end{remark}
@@
 the definitions, this claim boils down to the following basic result in
-commutative algebra:
+commutative algebra:\par
 \begin{quote}
 Let~$B$ be a ring. Let~$1 = f_1 + \cdots + f_m \in B$ be a partition of unity. Let
-elements~$s_i \in B[f_i^{-1}]$ be given. Assume that~$s_i = s_j$
-in~$B[(f_if_j)^{-1}]$ for all pairs of indices. Then there is exactly one
+elements~$s_i \in B[f_i^{-1}]$ be given. Assume that~$s_j = s_k$
+in~$B[(f_jf_k)^{-1}]$ for all pairs of indices. Then there is exactly one
 element~$s \in B$ such that, for all indices~$i$,~$s = s_i$ in~$B[f_i^{-1}]$.
@@
 sheaves in sight. Notwithstanding the impredicative nature of locale theory,
-the resulting formulation will even make sense in a metatheories without a
+the resulting formulation will even make sense in metatheories without a
 powerset operation such as~\textsc{czf}~\cite{crosilla:predicativity,aczel-rathjen:cst} or arithmetic
 universes~\cite{maietti:au,vickers:sketches}. In
-fact, it is now a purely syntactic translation procedure which can be carried
+fact, it is then a purely syntactic translation procedure which can be carried
 out within~\textsc{pra}.
@@
 However, in the form~``$A^\sim$ is a field'', it becomes easily accessible.
-This state of affairs is the more pronounced the more negations and especially
+This state of affairs becomes the more pronounced the more negations and especially
 double negations occur in the proof.
@@
 
-On page~\ref{par:appl-constr-alg}, we presented a classical proof of the
+On page~\pageref{par:appl-constr-alg}, we presented a classical proof of the
 following theorem.\par
@@
 \begin{enumerate}
-  \item $B^\sim$ and $M^\sim$ are free modules over $A^\sim$ (by basic linear algebra),
+  \item $B^\sim$ and $M^\sim$ are free modules over $A^\sim$,
   \item $A^\sim \to B^\sim$ is of finite presentation, and
@@
 introduce a variant of the usual negation~$\neg\varphi \equiv (\varphi
-\Rightarrow \bot)$, namely~$\negg\varphi \equiv (\varphi \Rightarrow a = 0)$.
+\Rightarrow \bot)$, namely~$\negg\varphi \equiv (\varphi \Rightarrow x = 0)$.
 We may still use the law of excluded middle for this new negation and hence
-exploit that any element~$x$ of~$A^\sim$ is invertible or ``not'' invertible. By
+exploit that any element~$f$ of~$A^\sim$ is invertible or ``not'' invertible. By
 Exercise~\ref{ex:gen-field-property}, this amounts to saying that some power
-of~$x$ annihilates~$a$. The remainder of the proof carries over with minor
+of~$f$ annihilates~$x$. The remainder of the proof carries over with minor
 modifications.
@@
 form~$\left(\begin{smallmatrix}1&0\\0&M'\end{smallmatrix}\right)$ by elementary
-row and column transformation. The matrix~$M'$ is still injective,
+row and column transformations. The matrix~$M'$ is still injective,
 and~$(\Lambda^{m-1} M') = (\Lambda^m M)$, hence we may assume by induction
@@
 \begin{alphlist}[(f)]
-\item Verify that~$\nabla$ is a \emph{local operator}, that is verify that
+\item Verify that~$\nabla$ is a \emph{local operator}, that is that
 \[ (\varphi \Rightarrow \nabla\varphi), \quad
@@
 \begin{exercise}[Units of the sheaf model]%
-Let~$A$ be a ring. Let~$f \in A$. Let~$x \in A[f^{-1}]$. Show using the result
-of Exercise~\ref{ex:global-existence} that~$f \models \exists y\?A^\sim\+ xy
+Let~$A$ be a ring. Let~$f \in A$. Let~$x \in A[f^{-1}]$. Show using
+Exercise~\ref{ex:global-existence} that~$f \models \exists y\?A^\sim\+ xy
 = 1$ if and only if~$x$ is invertible in~$A[f^{-1}]$.
@@
 field~\cite{coquand-lombardi:krull-dim-polynomial-ring}. Proofs of the stated
-relation between the dimension of~$A$ and the dimension of~$A^\sim$ are in
+relation between the dimension of~$A$ and the dimension of~$A^\sim$ are recorded in
 Ref.~\cite[Section~3.4]{blechschmidt:phd}.\par}

### 2020-11-02 paper-herrsching2019/generalized-spaces.tex
typo (thanks to the anonymous referee)

@@
 rearranging those subsets using only Euclidean motions yields two disjoint
-solid balls of radius~$r$ each. The axiom of choice is required to construct
+solid balls of radius~$r$ each. The axiom of choice is required to procure
 these subsets, and the Banach--Tarski paradox is not in
@@
 Table~\ref{table:nabla-translation}.
+
+\begin{table}[ht]
+  \tbl{The definition of the~$\nabla$-translation}{\begin{minipage}{\textwidth}\begin{align*}
+    (x = y)^\nabla &\defeqv \nabla(x = y) \\
+    R(x_1,\ldots,x_n)^\nabla &\defeqv \nabla(R(x_1,\ldots,x_n)) \\
+    \top^\nabla &\defeqv \top \quad \\
+    \bot^\nabla &\defeqv \nabla\bot \\
+    (\varphi \wedge \psi)^\nabla &\defeqv \varphi^\nabla \wedge \psi^\nabla &
+    \textstyle (\bigwedge_i \varphi_i)^\nabla &\defeqv \textstyle \bigwedge_i \varphi_i^\nabla \\
+    (\varphi \vee \psi)^\nabla &\defeqv \nabla(\varphi^\nabla \vee \psi^\nabla) &
+    \textstyle (\bigvee_i \varphi_i)^\nabla &\defeqv \textstyle \nabla(\bigvee_i \varphi_i^\nabla) \\
+    (\varphi \Rightarrow \psi)^\nabla &\defeqv (\varphi^\nabla \Rightarrow \psi^\nabla) \\
+    (\forall x\?X\+ \varphi)^\nabla &\defeqv (\forall x\?X\+ \varphi^\nabla) &
+    (\exists x\?X\+ \varphi)^\nabla &\defeqv \nabla(\exists x\?X\+ \varphi^\nabla)
+  \end{align*}\end{minipage}}
+  \label{table:nabla-translation}
+\end{table}
+
 Lengthy but straightforward computations by induction on the structure of
@@
 \begin{table}[ht]
-  \tbl{The definition of the~$\nabla$-translation}{\begin{minipage}{\textwidth}\begin{align*}
-    (x = y)^\nabla &\defeqv \nabla(x = y) \\
-    R(x_1,\ldots,x_n)^\nabla &\defeqv \nabla(R(x_1,\ldots,x_n)) \\
-    \top^\nabla &\defeqv \top \quad \\
-    \bot^\nabla &\defeqv \nabla\bot \\
-    (\varphi \wedge \psi)^\nabla &\defeqv \varphi^\nabla \wedge \psi^\nabla &
-    \textstyle (\bigwedge_i \varphi_i)^\nabla &\defeqv \textstyle \bigwedge_i \varphi_i^\nabla \\
-    (\varphi \vee \psi)^\nabla &\defeqv \nabla(\varphi^\nabla \vee \psi^\nabla) &
-    \textstyle (\bigvee_i \varphi_i)^\nabla &\defeqv \textstyle \nabla(\bigvee_i \varphi_i^\nabla) \\
-    (\varphi \Rightarrow \psi)^\nabla &\defeqv (\varphi^\nabla \Rightarrow \psi^\nabla) \\
-    (\forall x\?X\+ \varphi)^\nabla &\defeqv (\forall x\?X\+ \varphi^\nabla) &
-    (\exists x\?X\+ \varphi)^\nabla &\defeqv \nabla(\exists x\?X\+ \varphi^\nabla)
-  \end{align*}\end{minipage}}
-  \label{table:nabla-translation}
-\end{table}
-
-\begin{table}[ht]
   \tbl{The semantics of the \emph{classifying topos of~$\TT$} (propositional case)}{\begin{tabular}{@{}l@{\ \ }c@{\ \ }l@{}}
@@
 The presheaf~$\underline{M}^\mathrm{pre}$ associated to a set~$M$ is almost
-never a presheaf. Its sheafification~$\underline{M}$ is the \emph{constant sheaf associated
+never a sheaf. Its sheafification~$\underline{M}$ is the \emph{constant sheaf associated
 to~$M$}\label{page:constant-sheaf}. On an open~$U$, its sections are the locale
@@
     $U \models s =_F t$ &iff& $s = t$ when evaluated as elements of~$F(U)$ \\
-    $U \models R(s_1,\ldots,s_n)$ &iff& $\langle s_1,\ldots,s_n \rangle \in R(U)$ \\
+    $U \models R(s_1,\ldots,s_n)$ &iff& $\langle s_1,\ldots,s_n \rangle \in
+    R(U)$ \\
+    && \quad\quad \text{(where $R$ is the relation symbol corresponding to} \\
+    && \quad\quad\quad\quad \text{a subsheaf~$R \subseteq F_1 \times \cdots \times F_n$)} \\
     $U \models \top$ &iff& true \\
@@
 The existence of such a shape-shifting prime filter~$\fff_0$ is not a pipe dream.
-While it does not exist as a set, it can be incarnated as a \emph{sheaf}.
+While it does not exist as a set, it can be realized as a \emph{sheaf}.
 
@@
 moduli spaces. When phrased for the affine situation in a constructively
-sensible way, a simple version of this theorem reads as follows.\par
+sensible way, a simple version of this theorem reads as follows. We recall
+from Section~\ref{sect:algebraic-preliminaries} that
+the notation~$M[f^{-1}]$ makes sense for any element~$f \in A$; in case~$f$ is
+zero or nilpotent, this localized module will be the trivial module.\par
 \begin{quote}

### 2020-11-26 paper-herrsching2019/generalized-spaces.tex
more typos (thanks to the anonymous referee)

@@
 \item The proof proceeds by contradiction, hence requires the law of excluded
-middle, the strongest of the omniscience principles.
+middle.  % the strongest of the omniscience principles.
 \item The proof then requires a minimal prime ideal. Such an ideal can be
@@
 Prime Ideal Theorem~\cite{scott:bpit,banaschewski-harting:lattice-aspects}.
-The combined statement ``any ring has a minimal prime ideal'' is equivalent to
+The combined statement ``any nontrivial ring has a minimal prime ideal'' is equivalent to
 the full axiom of choice~\cite{savin:minimal-prime-ideals}.}
@@
 uncountable~\cite{blechschmidt-hutzler:macneille}.} Hence there is no
-interesting topological space of those surjections. However, there is a
+interesting topological space of those surjections. The set~$\{ f : \NN
+\twoheadrightarrow \RR \,|\, \text{$f$ is a surjection} \}$ is empty, and there
+is not much more to say about this state of affairs. However, there is a
 well-defined and nontrivial locale~$X$ of those surjections.
@@
 (the localic version of) the naturals which, on the level of points, would map
-a surjection~$f : \NN \to \RR$ to the smallest preimage of~$x$ under~$f$.
+a surjection~$f : \NN \to \RR$ to the smallest number~$n$ such that~$f(n) = x$.
 \item There is a well-defined continuous function from~$X$ to (the localic
@@
 
-Unlike a regular sequence such as~$(0,0,0,\ldots)$, we expect that, if the
+Unlike a regular sequence such as~$(0,0,0,\ldots)$, we intuitively expect that, if the
 digits of a \emph{random sequence} are revealed step by step, we will
@@
 of~$X$ such that~$\top = \bigvee_{i \in I} U_i$, there is a (Kuratowski-)finite
-subset~$I' \subseteq I$ such that~$\top = \bigvee_{i \in I'} U_i$.
+subset~$I' \subseteq I$ such that~$\top = \bigvee_{i \in I'} U_i$.\footnote{A
+set~$M$ is \emph{Kuratowski-finite} iff there is a surjection~$[n]
+\twoheadrightarrow M$, where~$[n] = \{1,2,\ldots,n\}$. In constructive
+mathematics, we distinguish this notion from the stronger condition for a
+set~$M$ to be \emph{Bishop-finite}, which requires that there is a
+bijection~$[n] \to M$.}
 With this definition, a topological space~$Y$ is compact iff its induced locale~$L(Y)$ is.
@@
 spatial. The localic real line is spatial in classical mathematics and can
-fail to be spatial in constructive mathematics (Exercise~\ref{ex:spatiality-reals}).
+fail to be spatial in constructive mathematics
+(Exercise~\ref{ex:spatiality-reals}). In this case, the locale induced by the
+topological space of reals should not be confused with the true localic real
+line. These locales have the same points, but the former might not be locally compact
+while the latter always is.
 
@@
 fruitfully employed as metatheories. Geometric theories are more interesting
-for applications, for instance for presenting spaces. This aspect is discussed
+for specific applications, for instance for presenting spaces. This aspect is discussed
 in Section~\ref{sect:presenting-frames}.
@@
 \item Perhaps the most important difference to formal systems is that geometric
-theories often depend on given mathematical objects. For instance, for each
+theories, unlike formal systems, often depend on given mathematical objects. For instance, for each
 ring~$A$, there is the geometric theory of prime filters of~$A$. This theory
@@
 \begin{enumerate}
-\item If infinitary classical logic proves a geometric sequent, then so does
+\item If infinitary classical logic (classical logic extended by set-indexed
+conjunctions and disjunctions) proves a geometric sequent, then so does
 geometric logic.
@@
 and sound with respect to (infinitary) intuitionistic first-order logic as
-defined in Table~\ref{table:first-order-logic}:
+defined in Table~\ref{table:first-order-logic} on
+page~\pageref{table:first-order-logic}:
 
@@
 
-\begin{proof}Omitted.\end{proof}
+\begin{proof}See~\cite[Corollary~D1.2.14(ii)]{johnstone:elephant}.\end{proof}
 
@@
 
+Concluding, one can well ask what the computational content of the equation~$1
+= 0$ is. The answer depends on the ring in question. In rings where we know a
+priori that~$1 \neq 0$, the two statements are most useful in their
+contrapositive form. But in rings of the form~$R[f^{-1}]$ or~$R/(f)$, learning
+that~$1 = 0$ can indeed be useful: In the former case, this tells us that~$f$
+is nilpotent (and a computational witness of~$1 = 0$ can be transformed into a
+number~$n$ such that~$f^n = 0$), and in the latter case, this tells us that~$f$
+is invertible.
+
 \paragraph{Grothendieck's generic freeness lemma.} This theorem is a basic
@@
 the notation~$M[f^{-1}]$ makes sense for any element~$f \in A$; in case~$f$ is
-zero or nilpotent, this localized module will be the trivial module.\par
+zero or nilpotent, the localized module~$M[f^{-1}]$ will be the trivial module.\par
 \begin{quote}
@@
 Petković, Peter Schuster, Helmut Schwichtenberg, Steve Vickers and Daniel
-Wessel for invaluable discussions shaping this work.
+Wessel for invaluable discussions shaping this work. We are also grateful to
+the anonymous referee, whose comments suggested several improvements to the
+presentation.
 

### 2020-12-27 paper-herrsching2019/generalized-spaces.tex
typo
Absolute final version

@@
 \usepackage[onethmnum]{ws-rv-thm}
-\usepackage{ws-rv-van}
+\usepackage[square]{ws-rv-van}
 
@@
 
+\newcommand{\bracketedrefcite}[1]{\citep{#1}}
+
 \chapter{Generalized spaces for constructive algebra}
@@
 introductions to the theory include
-Refs.~\refcite{johnstone:art,johnstone:point}. The interplay between logic and
+Refs.~\bracketedrefcite{johnstone:art,johnstone:point}. The interplay between logic and
 geometry is particularly stressed in
-Refs.~\refcite{vickers:continuity,vickers:locales-toposes}. We recommend
-Ref.~\refcite{belanger-marquis:pointless} for a historical guide and
-Ref.~\refcite{picado-pultr:frames-and-locales} for a comprehensive modern
+Refs.~\bracketedrefcite{vickers:continuity,vickers:locales-toposes}. We recommend
+Ref.~\bracketedrefcite{belanger-marquis:pointless} for a historical guide and
+Ref.~\bracketedrefcite{picado-pultr:frames-and-locales} for a comprehensive modern
 textbook on the subject.
@@
 Starting points include
-Refs.~\refcite{coste-lombardi-roy:dynamical-method,coquand-lombardi:logical-approach,rinaldi-schuster-wessel:edde,cederquist-coquand:entrel}.
+Refs.~\bracketedrefcite{coste-lombardi-roy:dynamical-method,coquand-lombardi:logical-approach,rinaldi-schuster-wessel:edde,cederquist-coquand:entrel}.
 
@@
 infinitary logic, but the basic idea has long been known and is summarized, for
-instance, in Ref.~\refcite{aczel:russell-prawitz}.
+instance, in Ref.~\bracketedrefcite{aczel:russell-prawitz}.
 
@@
 subcovering (a self-contained exposition of this phenomenon is contained in
-Ref.~\refcite{bauer:kleene-tree}).
+Ref.~\bracketedrefcite{bauer:kleene-tree}).
 
@@
 in Table~\ref{table:classifying-topos}, we have~$\varphi \models \psi$.
-By~(\ref{item:baby-geom}), there is a~$\TT$-derivation of~$\varphi \models
+By~(\ref{item:baby-geom}), there is a~$\TT$-derivation of~$\varphi \vdash
 \psi$ in geometric logic.
@@
 elimination~\cite[Section~6]{rathjen:barr}; see also
-Ref.~\refcite{dyckhoff-negri:geometrisation}, and
-Ref.~\refcite{rathjen-toppel:reduction} for background on proof-theoretical
+Ref.~\bracketedrefcite{dyckhoff-negri:geometrisation}, and
+Ref.~\bracketedrefcite{rathjen-toppel:reduction} for background on proof-theoretical
 reduction, of which Barr's theorem is an example.
@@
 presheaves on sites can be defined; however, this shall not be pursued here,
-a textbook reference is Ref.~\refcite{moerdijk-maclane:sheaves-logic}.
+a textbook reference is Ref.~\bracketedrefcite{moerdijk-maclane:sheaves-logic}.
 
@@
 examples from theoretical computer science, analysis and differential geometry
-are contained in Ref.~\refcite{blechschmidt:custom-tailored}, which should be
+are contained in Ref.~\bracketedrefcite{blechschmidt:custom-tailored}, which should be
 regarded as a companion paper to this contribution. We also recommend
@@
 \cite[Chapter~6]{borceux:handbook3},
-\cite[Part~D]{johnstone:elephant} and~[\refcite{shulman:categorical-logic}],
+\cite[Part~D]{johnstone:elephant} and~\bracketedrefcite{shulman:categorical-logic},
 which also gives a detailed account of useful type theory.
@@
 from type theory, where it is used with a similar meaning (see for instance
-Ref.~\refcite{kraus-escardo-coquand-altenkirch:anonymous}). However, there is a
+Ref.~\bracketedrefcite{kraus-escardo-coquand-altenkirch:anonymous}). However, there is a
 subtle difference: The conjunction of ``there is at most one element~$x$'' and
@@
 efficient tool is probably the theory of \emph{entailment relations}. We refer
-to Refs.~\refcite{rinaldi-schuster-wessel:edde,cederquist-coquand:entrel} and
+to Refs.~\bracketedrefcite{rinaldi-schuster-wessel:edde,cederquist-coquand:entrel} and
 the references therein and also strongly recommend the literature on dynamical
 methods in commutative
-algebra~\refcite{coste-lombardi-roy:dynamical-method,coquand-lombardi:logical-approach,coquand:site}.
+algebra~\bracketedrefcite{coste-lombardi-roy:dynamical-method,coquand-lombardi:logical-approach,coquand:site}.
 \end{remark}
@@
 \cite[Thm.~14.4]{eisenbud:commutative-algebra}, which only cover the case of
-Noetherian integral domains, or Refs.~[\refcite{staats:generic-freeness}],
+Noetherian integral domains, or Refs.~\bracketedrefcite{staats:generic-freeness},
 \cite[Tag~051Q]{stacks-project}
@@
 been replaced by an explicit computation has been unwound in
-Ref.~\refcite{blechschmidt:generic-freeness}.
+Ref.~\bracketedrefcite{blechschmidt:generic-freeness}.
 
@@
 Other short and explicit proofs of McCoy's theorem are given in
-Refs.~\refcite{blocki:mccoy},
+Refs.~\bracketedrefcite{blocki:mccoy},
 \cite[Theorem~5.22]{lombardi-quitte:constructive-algebra} and
@@
 The for these purposes correct constructive notion of Krull dimension can be
-found in Refs.~\refcite{coquand-ducos-lombardi-quitte:krull-integral,coquand-lombardi-roy:char-krull}.
+found in Refs.~\bracketedrefcite{coquand-ducos-lombardi-quitte:krull-integral,coquand-lombardi-roy:char-krull}.
 It has been used to give a short and constructive proof that~$\dim
@@
 {\scriptsize\emph{Note.} A self-contained solution is given in
-Ref.~\refcite{blechschmidt:kaplansky}.\par}
+Ref.~\bracketedrefcite{blechschmidt:kaplansky}.\par}
 \end{alphlist}

### 2020-12-27 paper-herrsching2019/generalized-spaces.tex
Generate hyperlinks
Fix double label

@@
 %\renewindex{default}{idx}{ind}{Subject Index}  % subject index
+
+\usepackage{hyperref}
 
@@
 
-\begin{example}\label{ex:negneg-interpretation}
+\begin{example}\label{ex:negneg-interpretation-unrolled}
 For future reference, and also for giving a different
@@
 \begin{quote}
-\textbf{Proof.} By Example~\ref{ex:negneg-interpretation} and
+\textbf{Proof.} By Example~\ref{ex:negneg-interpretation-unrolled} and
 Exercise~\ref{ex:bridging-modules}(\ref{item:free}), the claim amounts to the statement that~$M^\sim$ is \notnot

### 2021-09-01 paper-herrsching2019/generalized-spaces.tex
typo

@@
 filter. (The converse also holds.) In classical mathematics, this condition is
-satisfied if and only if~$A$ is local ring of Krull dimension zero. In this
+satisfied if and only if~$A$ is a local ring of Krull dimension zero. In this
 case, the unique prime filter is~$A^\times$ and the free local ring over~$A$ is given
@@
 
-For~$A^\sim$ to be the local ring over~$A$, there has to be a morphism~$A \to
+For~$A^\sim$ to be the free local ring over~$A$, there has to be a morphism~$A \to
 A^\sim$. For this we take the pair~$\langle g, g^\sharp \rangle$ where~$g :

### 2021-11-06 paper-qcoh.tex
typo
to do

@@
 in~\cite{blechschmidt-hutzler-oldenziel:composition}, one can show that this topos classifies the
-theory of homomorphisms between~$\TT$-models. It can also be obtained as the
+theory of homomorphisms between~$\TT$-models. This topos can also be obtained as the
 lax pullback~$(\Set[\TT] \Rightarrow_{\Set[\TT]} \Set[\TT])$.
@@
 
+% XXX explain how Nullstellensatz follows from the Horn version
+
 \section{The generalization to the higher-order case}
@@
 classifying topos of the theory~$\OO$ of objects, the theory with a single sort and
-no function symbols, relation symbols or axioms, including:
+no function symbols, relation symbols or axioms, including the following ones.
 \begin{align*}
@@
 \end{align*}
-Intuitively, the first expresses that~$U_\OO$ is close to a subsingleton, while
+Intuitively, the first sequent expresses that~$U_\OO$ is close to a subsingleton, while
 the second expresses that~$U_\OO$ is close to being infinite. Using the
@@
 We argue internally to~$\Set[\OO]$. Let~$x,y\?U_\OO$ and assume~$\neg(x = y)$.
-Then~$\ul{\OO}/U_\OO$ proves~$(x = y \seq{[]} \bot)$. Hence this sequent holds
+By the Nullstellensatz, the theory~$\ul{\OO}/U_\OO$ proves~$(x = y \seq{[]} \bot)$. Hence this sequent holds
 of any model of~$\ul{\OO}/U_\OO$, in particular of the
@@
 * XXX fix "first-order" usage everywhere
+
+* XXX Cite C2.5.4 or similar

### 2021-11-20 slides-dagstuhl2021.tex
Slides for Dagstuhl 2021 (WIP)

--- 2019-07-10 slides-edinburgh2019.tex
@@
 \usepackage{pgfpages}
-\setbeameroption{show notes on second screen}
-\setbeamertemplate{note page}[plain]
-\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}
+%\setbeameroption{show notes on second screen}
+%\setbeamertemplate{note page}[plain]
+%\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}
+\newcommand{\jnote}[2]{}
 
@@
 \usepackage{stmaryrd}
+\usepackage{amssymb}
 \usepackage{array}
@@
 \hypersetup{colorlinks=true}
+
+\newcommand*\circled[1]{\tikz[baseline=(char.base)]{%
+  \node[shape=circle,draw,inner sep=1pt] (char) {#1};}}
+
+\DeclareFontFamily{U}{bbm}{}
+\DeclareFontShape{U}{bbm}{m}{n}
+   {  <5> <6> <7> <8> <9> <10> <12> gen * bbm
+      <10.95> bbm10%
+      <14.4>  bbm12%
+      <17.28><20.74><24.88> bbm17}{}
+\DeclareFontShape{U}{bbm}{m}{sl}
+   {  <5> <6> <7> bbmsl8%
+      <8> <9> <10> <12> gen * bbmsl
+      <10.95> bbmsl10%
+      <14.4> <17.28> <20.74> <24.88> bbmsl12}{}
+\DeclareFontShape{U}{bbm}{bx}{n}
+   {  <5> <6> <7> <8> <9> <10> <12> gen * bbmbx
+      <10.95> bbmbx10%
+      <14.4> <17.28> <20.74> <24.88> bbmbx12}{}
+\DeclareFontShape{U}{bbm}{bx}{sl}
+   {  <5> <6> <7> <8> <9> <10> <10.95> <12> <14.4> <17.28>%
+      <20.74> <24.88> bbmbxsl10}{}
+\DeclareFontShape{U}{bbm}{b}{n}
+   {  <5> <6> <7> <8> <9> <10> <10.95> <12> <14.4> <17.28>%
+      <20.74> <24.88> bbmb10}{}
+\DeclareMathAlphabet{\mathbbm}{U}{bbm}{m}{n}
+\SetMathAlphabet\mathbbm{bold}{U}{bbm}{bx}{n}
 
@@
 
-\title{A general Nullstellensatz for generalised spaces}
+\title{Bridging the foundational gap: updating algebraic geometry in
+face of current challenges regarding formalizability, constructivity and
+predicativity}
 \author{Ingo Blechschmidt}
-\date{April 11th, 2019}
+\date{November 23th, 2021}
 
@@
 \newcommand{\B}{\mathcal{B}}
-\renewcommand{\C}{\mathcal{C}}
+\newcommand{\C}{\mathcal{C}}
 \newcommand{\M}{\mathcal{M}}
 \renewcommand{\AA}{\mathbb{A}}
+\newcommand{\BB}{\mathbb{B}}
+\newcommand{\pp}{\mathbbm{p}}
+\newcommand{\MM}{\mathbb{M}}
 \newcommand{\E}{\mathcal{E}}
 \newcommand{\F}{\mathcal{F}}
-\renewcommand{\G}{\mathcal{G}}
+\newcommand{\FF}{\mathbb{F}}
+\newcommand{\G}{\mathcal{G}}
 \newcommand{\J}{\mathcal{J}}
@@
   \end{center}}
-
-\newcommand{\bignumber}[1]{
-  \renewcommand{\insertenumlabel}{#1}\scalebox{1.5}{\usebeamertemplate{enumerate item}}
-}
-\newcommand{\bigheart}{\includegraphics{heart}}
 
@@
 
-\setbeamertemplate{headline}{%
-  \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex]{}%
-    \insertsectionnavigationhorizontal{\paperwidth}{}{}%
-  \end{beamercolorbox}%
-  \vskip0pt%
-}
+\setbeamertemplate{headline}{}
 
 \setbeamertemplate{frametitle}{%
-  \vskip0.4em%
+  \vskip0.5em%
   \leavevmode%
   \begin{beamercolorbox}[dp=1ex,center]{}%
-  %   \usebeamercolor[fg]{item}{\textbf{{\Large \insertframetitle}}}
-    \begin{tikzpicture}
-      \def\R{8pt}
-      \node (title) {\hil{\large\insertframetitle}};
-      \begin{pgfonlayer}{background}
-        \draw[decorate, very thick, draw=mypurple]
-          ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
-          ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
-          ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
-          ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
-          cycle;
-      \end{pgfonlayer}
-    \end{tikzpicture}
+    \usebeamercolor[fg]{item}{\textbf{{\Large\insertframetitle}}}
   \end{beamercolorbox}%
-  \vskip-0.6em%
+  \vskip-0.2em%
 }
@@
   \addtocounter{framenumber}{\value{framenumberpreappendix}}
-}
-
-\newcommand{\mynav}[3]{%
-  foo
 }
@@
 
+\newcommand{\bignumber}[1]{%
+  \renewcommand{\insertenumlabel}{#1}\scalebox{1.2}{\!\usebeamertemplate{enumerate item}\!}
+}
+\newcommand{\bigheart}{\includegraphics{heart}}
+
+\newcommand{\subhead}[1]{{\centering\textcolor{gray}{\hrulefill}\quad\textnormal{#1}\quad\textcolor{gray}{\hrulefill}\par}}
+
 \begin{document}
@@
   \bigskip
-  \[ \xymatrix{
-    & \text{$\ul{\TT}/U_\TT$ proves $\sigma$} \\
-    \text{$\sigma$ holds for $U_\TT$} \ar@/^1.1pc/@{<=>}[ur] &&
-    \text{$\sigma$ holds for $M$} \ar@/_1.1pc/@{<=}[ul]
-  } \]
+  %\includegraphics[height=0.32\textwidth]{olivia-lattices}
   \bigskip
@@
     \def\R{8pt}
-    \node (title) {\hil{A general Nullstellensatz for generalised spaces}};
+    \node (pretitle) {Bridging the foundational gap:};
+    \node (title) [below of=pretitle] {\vbox{\vspace*{-0.5em}updating \hil{algebraic
+    geometry} in face of \hil{current challenges} regarding
+    \hil{formalizability}, \hil{constructivity} and \hil{predicativity}}};
     \begin{pgfonlayer}{background}
@@
 
-  Category Theory 2019 in Edinburgh \\
-  July 9th, 2019
+  Dagstuhl Seminar 20202: \\
+  Geometric Logic, Constructivisation, and Automated Theorem Proving \\
+  \ \\
+  November 23th, 2021
   \bigskip
@@
   Ingo Blechschmidt \\
-  Università di Verona
+  University of Augsburg
   \par
-
-  \jnote{1}{\textbf{This talk in a nutshell:}
-
-  There is a certain ring, the \emph{generic ring}, which is special in that
-  it is \emph{conservative}: It has exactly those properties which are shared
-  by any ring whatsoever. One could argue that it is this ring which we're
-  implicitly referring to when we utter the phrase ``Let~$R$ be a ring.''.
-
-  A surprising observation is that the generic ring is a field. This is not a
-  contradiction because conservativity only pertains to properties which can be
-  put in a certain syntactic form, dubbed \emph{geometric sequents}, and the
-  field condition is not of that form.
-
-  As a consequence, when verifying a geometric sequent for all rings, we may
-  freely assume the field condition: Our proof will apply to the generic ring,
-  because this ring happens to be a field, and its conclusion will spread to
-  any ring whatsoever thanks to its syntactic form.
-
-  An analogous story can be told for the generic group, the generic vector
-  space and quite generally for the generic model of any \emph{geometric
-  theory}. It is an interesting task to discover properties of such
-  generic models. This talk presents a source of such properties, the
-  \emph{topos-theoretic Nullstellensatz}, which is
-  universal in the sense that any such property is a logical consequence of the
-  Nullstellensatz.}
 \end{frame}}
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\hspace*{-1.7cm}\includegraphics[height=\paperheight,trim=0 0 5.1cm 0,clip]{arthurs-seat-4}\end{minipage}}
-\begin{frame}[plain]%
-  \vspace*{\paperheight}\vspace*{-2.35cm}\hspace*{-1cm}%
-  \includegraphics[width=\paperwidth/4]{arthurs-seat-7}%
-  \includegraphics[width=\paperwidth/4]{arthurs-seat-0}%
-  \includegraphics[width=\paperwidth/4]{arthurs-seat-5}%
-  \includegraphics[width=\paperwidth/4]{arthurs-seat-9}%
-  \jnote{1}{This talk owes a special debt to Alexander Oldenziel, who
-  repeatedly suggested that the topic of this talk should be studied.
+\begin{frame}{Algebraic geometry in a nutshell}
+  Turn \hil{commutative rings} into \hil{spaces}, and \hil{glue} those
+  spaces.
+  \bigskip
 
-  The photos on the slide are of \emph{Arthur's Seat}, a beautiful extinct
-  volcano in Edinburgh. It gives an excellent view of the city and is very much
-  recommended for a hiking tour.
-  }%
-\end{frame}}
-\addtocounter{framenumber}{-1}
+  \subhead{Examples}
+  \begin{enumerate}
+    \item $\Spec(k[X_1,\ldots,X_n]) = \AA^n$.
+    \item $\Spec(k[X,Y]/(Y-X^2)) = \text{standard parabola}$.
+    \item Gluing
+    \begin{tikzpicture}[scale=0.2,baseline={([yshift=-.6ex]current bounding box.center)}]
+      \draw (0,0) circle (1);
+      \draw [fill=white] (90:1) circle (0.2);
+    \end{tikzpicture}
+    with
+    \begin{tikzpicture}[scale=0.2,baseline={([yshift=-.6ex]current bounding box.center)}]
+      \draw (0,0) circle (1);
+      \draw [fill=white] (270:1) circle (0.2);
+    \end{tikzpicture}
+    along
+    \begin{tikzpicture}[scale=0.2,baseline={([yshift=-.6ex]current bounding box.center)}]
+      \draw (0,0) circle (1);
+      \draw [fill=white] (90:1) circle (0.2);
+      \draw [fill=white] (270:1) circle (0.2);
+    \end{tikzpicture}
+    yields
+    $\PP^1$.
+  \end{enumerate}
+  \bigskip
+
+  \subhead{Concrete results}
+  Fermat's Last Theorem: For~$n \geq 3$, no positive integers satisfy
+  \[ a^n + b^n = c^n. \]
+\end{frame}
+
+\begin{frame}{Transfinite methods}
+  The current presentation of algebraic geometry employs:
+  \begin{itemize}
+    \item large structures: classes, large categories, universes, \ldots
+    \item law of excluded middle
+    \item axiom of choice
+  \end{itemize}
+  \bigskip
+
+  \subhead{despite:}
+  \begin{enumerate}
+    \item subject matter (in part) very concrete
+    \item computer algebra systems for computations exist
+    \item constructive algebra well-developed
+  \end{enumerate}
+\end{frame}
+
+\end{document}
 
@@
 
-\subsection{The mystery of nongeometric sequents}
+\subsection{The generic ring}
 
-\begin{frame}{The mystery of nongeometric sequents}
-  \jnote{1}{A \emph{geometric sequent} is a syntactic expression of the
-  form~$(\varphi \seq{x_1\?X_1,\ldots,x_n\?X_n} \psi)$,
-  where~$x_1\?X_1,\ldots,x_n\?X_n$ is a list of variable declarations,
-  the~$X_i$ ranging over the available sorts, and~$\varphi$ and~$\psi$ are
-  \emph{geometric formulas}. Such a sequent is read as ``in the
-  context~$x_1\?X_1,\ldots,x_n\?X_n$, $\varphi$ entails $\psi$''.
-  Geometric formulas are built from atomic propositions (equality and the
-  available relation symbols) using the connectives~$\top$, $\bot$, $\wedge$, $\bigvee$
-  (set-indexed disjunction) and~$\exists$. Geometric formulas may not
-  contain~$\neg$, $\Rightarrow$, $\forall$.
+\begin{frame}{The generic ring}
+  \visible<2->{``}Let~$R$ be a ring.\visible<2->{''} \visible<2->{-- Which ring does this phrase refer to?}
 
-  There is a notion of a \emph{model} of a geometric theory in
-  any given topos. For instance, a ring in the usual sense is a model of the
-  theory of rings in the topos~$\Set$. The structure sheaf of a scheme~$X$ is a
-  model in the topos~$\Sh(X)$ of set-valued sheaves on~$X$.
-
-  With \emph{topos} we mean Grothendieck topos, and as metatheory we use a
-  constructive but impredicative flavour of English (which could be formalised
-  by what is supported by the internal language of elementary toposes with an NNO).
-  However the Nullstellensatz presented later makes no use of the
-  subobject classifier, hence the results can likely be generalised to hold in
-  a predicative metatheory or to hold for arithmetic universes.}
-
-  \jnote{2-}{Among all models in any topos, the \emph{universal} or
-  \emph{generic} one is special. It enjoys the universal property that any
-  model in any topos can be obtained from it by pullback along an essentially
-  unique geometric morphism. It is intriguing from a logical point of view
-  because it has exactly those properties which are shared by any model in any
-  topos.
-
-  One could argue, with a certain amount of success, that the generic model of
-  the theory of rings is what a mathematician implicitly refers to when she
-  utters the phrase ``Let~$R$ be a ring''. This point of view is fundamental to
-  the slogan \emph{continuity is geometricity}, as expounded for instance in
-  \fixedhref{http://www.cs.bham.ac.uk/~sjv/GeoAspects.pdf}{Continuity and
-  geometric logic} by Steve Vickers.}
-
-  \jnote{3-}{Crucially, the conservativity statement only pertains to
-  properties which can be put as geometric sequents. Generic models may have
-  additional nongeometric properties. Because conservativity does not apply to
-  them, they are not shared by all models in all toposes -- but any consequences
-  which can be put as geometric sequents are.
-
-  For instance, if we want to verify a geometric sequent for all local rings,
-  we may freely use the displayed field axiom.
-  Hence one reason why these nongeometric sequents are interesting is because
-  they provide us with new reduction strategies (proofs by ``without loss of
-  generality'').}
-
-  Let~$\TT$ be a \explain{geometric theory}{1-}{red!20}{-1cm}{0cm}{0cm}{\begin{minipage}{4.3cm}
-    \small
-    sorts,
-    function symbols,
-    relation symbols,
-    geometric sequents as axioms
-  \end{minipage}}, for instance the
-  \explain{theory of rings}{1-}{blue!20}{-1cm}{0cm}{0cm}{\begin{minipage}{4.3cm}
-    \small
-    \begin{tabbing}
-      fun. symb.: \= \kill
-      sorts: \> $R$ \\
-      fun. symb.: \> $0$, $1$, $-$, $+$, $\cdot$ \\
-      % rel. symb.: \> \emph{none} \\
-      axioms: \> $(\top \vdash_{x,y:R} x y = y x)$, \ldots
-    \end{tabbing}
-  \end{minipage}}.
-
-  \vspace*{1.8cm}
-
+  \vspace*{-0.6em}
   \begin{center}
     \begin{tikzpicture}[ultra thick, node distance=7mm]
-      \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40] (a) {$\ZZ$};
-      \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40, right=of a] (b) {$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$};
-      \node[regular polygon, regular polygon sides=5, draw=orange!80,
-      fill=orange!20, right=of b, rounded corners=1pt, inner sep=0cm] (c) {$\O_X$};
-      \node[star, rounded corners=1pt, star points=10, inner sep=0cm, draw=purple!80, fill=purple!20, right=of c] {$U_\TT$};
+      \node[rectangle, rounded corners=1pt, inner sep=5pt, draw=lime!80, fill=lime!40] (a) {$\ZZ$};
+      \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40, right=of a] (b) {$\FF_2$};
+      \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40, right=of b] (c) {$\QQ[X]$};
+      \node[regular polygon, regular polygon sides=6, draw=orange!80, fill=orange!40, right=of c, rounded corners=1pt] (d) {$\RR$};
+      \node[regular polygon, regular polygon sides=5, draw=pink!80, fill=pink!40, right=of d, rounded corners=1pt, inner sep=0cm] (e) {$\O_X$};
+      \visible<3->{\node[star, rounded corners=1pt, star points=10, inner
+      sep=2pt, draw=purple!80, fill=purple!30, right=of e] {$\AA$};}
     \end{tikzpicture}
   \end{center}
+  \vspace*{-0.6em}
 
-  \vspace*{-0.5cm}
+  \pause
+  \pause
   \pause
@@
   \justifying
-  \textbf{Theorem.} There is a \hil{generic model}~$U_\TT$. It is
-  \hil{conservative} in that
-  for any \explainstub{geometric sequent}{yellow!70}~$\sigma$ the following notions coincide:
-  \vspace*{-1.2em}
+  \textbf{Thm.} For any$^\star$ property~$P$ of rings, the following are
+  equivalent: \\[0em]
   \begin{enumerate}
-    \item The sequent~$\sigma$ holds for~$U_\TT$.
-    \item The sequent~$\sigma$ holds for any~$\TT$-model in any
-    \explainstub{topos}{purple!20}.
-    \item The sequent~$\sigma$ is provable modulo~$\TT$.
+    \item The \hil{generic ring}~$\AA$ has property~$P$.
+    \item Every$^\star$ ring has property~$P$.
+    \item The ring axioms entail property~$P$.
+  \end{enumerate}
+  \pause
+
+  \textbf{Example A.} For any~$x,y,z \in \AA$, $x + (y + z) = (x + y) + z$.
+  \pause
+
+  \textbf{Example B.} \only<6>{Is~$1 + 1 = 0$ in~$\AA$?}%
+  \pause
+  \only<7->{It is \hil{not the case} that~$1 + 1 = 0$ in~$\AA$. \only<8->{But also:}} \\
+  \pause
+  \phantom{\textbf{Example B.}} It is \hil{not the case} that~$1 + 1 \neq 0$ in~$\AA$.
+  \pause
+
+  \textbf{Example C (Anders Kock).} The generic ring is a \hil{field}:
+  \[ \forall x \in \AA\_ \bigl((x = 0 \Rightarrow 1 = 0) \Rightarrow (\exists y
+  \in \AA\_ xy = 1)\bigr). \]
+
+  \pause
+  \vspace*{-0.6em}
+  \hil{Hence:} When verifying a coherent sequent for all rings, can without loss of
+  generality assume the field condition.
+\end{frame}
+
+\begin{frame}{A selection of noncoherent sequents}
+  The \hil{generic object}~$\MM$ validates:
+  \begin{enumerate}
+    \item $\forall x,y \in \MM\_ \neg\neg(x = y)$.
+    \item $\forall x_1,\ldots,x_n \in \MM\_ \neg \forall y \in \MM\_
+    y = x_1 \vee \cdots \vee y = x_n$.
   \end{enumerate}
 
+  The \hil{generic ring}~$\AA$ validates:
+  \begin{enumerate}
+    \item $\forall x \in \AA\_ (x = 0 \Rightarrow 1 = 0) \Rightarrow (\exists y \in \AA\_ xy = 1)$.
+    \item $\forall x \in \AA\_ \neg\neg(x = 0)$.
+  \end{enumerate}
+
+  The \hil{generic local ring}~$\AA'$ validates:
+  \begin{enumerate}
+    \item $\forall x \in \AA'\_ (x = 0 \Rightarrow 1 = 0) \Rightarrow (\exists y \in \AA'\_ xy = 1)$.
+    \item $\neg \forall x \in \AA'\_ \neg\neg(x = 0)$.
+    \item $\forall f \in \AA'[X]_{\text{degree} > 0}\_ \neg\neg \exists x \in \AA'\_ f(x) = 0$.
+  \end{enumerate}
+\end{frame}
+
+\begin{frame}{An application in commutative algebra}
+  \justifying
+  Let~$A$ be a reduced ring ($x^n = 0 \Rightarrow x = 0$).
+  Let~$\pp$ be the \hil{generic prime ideal$^\star$} of~$A$.
+  Then~$A_\pp \defeq A[\pp^{-1}]$ validates:
+
+  {\vspace*{-1.2em}
+  \setbeamercolor{block body}{bg=red!30}
+  \setbeamercolor{structure}{fg=purple}
+  \begin{varblock}{\textwidth}{}
+    $A_\pp$ is a \hil{field}: $\forall x \in A_\pp\_ (\neg(\exists y \in A_\pp\_
+    xy = 1) \Rightarrow x = 0)$.
+
+    $A_\pp$ has \hil{$\boldsymbol{\neg\neg}$-stable equality}:
+    $\forall x,y \in A_\pp\_ \neg\neg(x = y) \Rightarrow x = y$.
+
+    \mbox{$A_\pp$ is \hil{anonymously Noetherian}.}\\[-1.2em]
+  \end{varblock}}
+
+  This observation unlocks a short and conceptual proof of Grothen\-dieck's
+  \hil{generic freeness lemma} in algebraic geometry.
+
+  \mbox{\textbf{Thm.} (baby freeness) Let~$M$ be an~$A$-module. Then~\bignumber{1}
+  implies~\bignumber{3}.}
+
+  {\small
+  \begin{tabular}{@{}lc@{\ }l@{}}
+    \bignumber{1} $M$ is finitely generated &($\Longleftrightarrow$&
+    $M_\pp$ is finitely generated) \\
+    \bignumber{2} $M$ is locally free &($\Longleftrightarrow$&
+    $M_\pp$ is free) \\
+    \bignumber{3} $M$ is locally free \emph{on a dense open} &($\Longleftrightarrow$&
+    $M_\pp$ is \emph{not not} free) \\
+  \end{tabular}\par}
+
+  \textbf{Proof.} Elementary linear algebra over~$A_\pp$. \qed
+\end{frame}
+
+\begin{frame}{A systematic source}
+  \begin{center}
+    \includegraphics[width=0.8\textwidth]{wraith-some-recent-developments-in-topos-theory} \\
+    \scriptsize
+    Gavin Wraith. \emph{Some recent developments in topos theory.} \\ In:
+    Proc.\@ of the ICM (Helsinki, 1978).
+  \end{center}
+
+  \textbf{Thm. (Nullstellensatz):} The generic~$\TT$-model~$U_T$ validates:
+  For any coherent sequent~$\sigma$,
+  \[
+    \text{$\sigma$ holds for $U_\TT$} \quad\textcolor{purple}{\Longleftrightarrow}\quad
+    \text{$\ul{\TT}/U_\TT$ proves $\sigma$.}
+  \]
+
+  \justifying
+  \textbf{Thm. (universality):} The generic~$\TT$-model validates a first-order
+  formula~$P$ if and only if~$P$ is intuitionistically deducible from the
+  axioms of~$\TT$ and the Nullstellensatz.
+\end{frame}
+
+\begin{frame}{Arithmetic universes}
+  Places where we can do mathematics (among others):
+  \begin{multicols}{2}
+    \begin{enumerate}
+      \item $\Set$ (sets)
+      \item $\Eff$ (data types) \columnbreak
+      \item \mbox{$\mathrm{sSet}$ (simplicial sets)}
+      \item \mbox{$\Sh(X)$ (sheaves over~$X$)}
+    \end{enumerate}
+  \end{multicols}
+  \pause
+  \vspace*{-0.6em}
+  These are examples for \hil{arithmetic universes}.
   \pause
@@
   \justifying
-  %\textbf{Main observation.} The generic model can validate nongeometric
-  %sequents which are not provable modulo~$\TT$ and which do not hold for
-  %arbitrary models in arbitrary toposes.
+  \textbf{Definition.} An \emph{arithmetic universe} is a category with finite
+  limits (``$\times$''), stable finite disjoint coproducts (``$\amalg$''),
+  stable ef{}fective quotients (``$X/{\sim}$'') and parametrized list objects
+  (``$\NN$'', ``$\mathrm{List}(X)$'').
 
-  \textbf{Observation (Kock).} The generic local ring is a field:
-  \vspace*{-0.2cm}
-  \[ (x = 0 \Rightarrow \bot) \vdash_{x:R} (\exists y\?R\_ xy = 1) \]
+  \textbf{Thm.} Any statement which is provable in \hil{predicative constructive
+  mathematics} (no powersets, no $\varphi \vee \neg\varphi$, no
+  \mbox{$\neg\neg\varphi \Rightarrow \varphi$}, no axiom of choice) is true in any
+  arithmetic universe.
+  \pause
+
+  Further examples:
+
+  \begin{enumerate}
+    \addtocounter{enumi}{4}
+    \item the \hil{initial} arithmetic universe
+    \item the \hil{classifying} arithmetic universe for the theory of rings
+  \end{enumerate}
 \end{frame}
 
-\subsection{Construction of the generic model}
-
-\begin{frame}{Construction of the generic model}
-  The generic model is \hil{not} the same as \ldots
-  \begin{itemize}
-    \item the \hil{initial model} (think~$\ZZ$) or
-    \item the \hil{free model on one generator} (think~$\ZZ[X]$).
-  \end{itemize}
-  Set-based models are \hil{too inflexible}.
-
-  \textbf{Definition.} The \hil{syntactic site}~$\C_\TT$ has \ldots
-  \begin{enumerate}
-    \item objects: \explainstub{formulas-in-contexts}{red!20} $\{ x_1\?X_1, \ldots, x_n\?X_n\_ \varphi \}$
-    \item morphisms: eqv. classes of \explainstub{provably functional
-    formulas}{red!30}
-    \item coverings: \explainstub{provably jointly surjective
-    families}{purple!30}
-  \end{enumerate}
-  The topos of sheaves over~$\C_\TT$ is the \hil{classifying
-  topos}~$\Set[\TT]$. The generic model interprets a sort~$X$ by the sheaf~$\yon\{x\?X\_\top\}$.
-
-  \jnote{1}{In case the theory~$\TT$ is a Horn theory (for instance if it is
-  an equational theory), the \emph{term algebra} (the set of terms in the
-  empty context modulo provable equality) is a model of~$\TT$. While the
-  term algebra does enjoy some nice categorical properties, it is in general
-  \emph{not} the generic model.
-
-  For instance, if~$\TT$ is the theory of rings, then the initial model
-  is~$\ZZ$. This model validates some geometric sequents which are not
-  validated by all rings, for instance~$(x^2 = 0 \seq{x:R} x = 0)$ or $(1 = 0
-  \vdash \bot)$.
-
-  In general, the generic model cannot be realised as a set-based model (with
-  a set for each sort, a map for each function symbol and so on). Sets are too
-  constant for this purpose; the flexibility of sheaves (``variable sets'') is
-  required: The generic model lives in the topos of set-valued
-  sheaves over~$\C_\TT$, called the \emph{classifying topos} of~$\TT$.}
-
-  \jnote{2}{The notation in the definition of the objects of~$\C_\TT$ is meant
-  to be suggestive, but should not be taken literally. The objects are
-  syntactic expressions consisting of a context and a geometric formula in that
-  context.
-
-  A morphism~$A = \{\vec x\_ \varphi\} \to \{\vec y\_ \psi\} = B$ in~$\C_\TT$ is the
-  equivalence class (modulo provable equivalence) of a geometric
-  formula~$\theta$ such that~$\TT$ proves
-  \begin{enumerate}
-    \item ``$\theta$ is a relation on~$A \times B$'': $(\theta \seq{\vec x,
-    \vec y} \varphi \wedge \psi)$
-    \item ``$\theta$ is total'': $(\varphi \seq{\vec x} \exists \vec y\_
-    \theta)$
-    \item ``$\theta$ is single-valued'': $(\theta \wedge \theta[\vec y'/\vec y]
-    \seq{\vec x,\vec y,\vec y'} \vec y = \vec y')$
-  \end{enumerate}
-
-  \vspace*{-0.6em}
-  A family~$(\{ \vec x_i\_ \varphi_i \} \xrightarrow{[\theta_i]} \{ \vec y\_
-  \psi \})_i$ is a covering iff~$\TT$ proves~$(\psi \seq{\vec y} \bigvee_i
-  \exists \vec x_i\_ \theta_i)$.
-
-  The slides experiment with using the symbol~``$\yon$'' for the Yoneda
-  embedding~$\C_\TT \to \Set[\TT]$, as in \fixedhref{http://www.math.jhu.edu/~eriehl/elements.pdf}{Elements
-  of~$(\infty,1)$-category theory} by Emily Riehl and Dominic Verity.}
-
-  \jnote{3}{The special case that the generic model of a theory~$\TT$ can be realised as
-  a model in~$\Set$ occurs iff~$\TT$ is Morita-equivalent to the
-  empty theory, that is, iff~$\TT$ has exactly one model in any topos.
-
-  The special case that there exists at least some conservative~$\TT$-model
-  in~$\Set$ occurs iff~$\TT$ has a conservative geometric expansion to a theory
-  which is Morita-equivalent to the empty theory.}
-\end{frame}
-
-\subsection{Going internal}
-
-\begin{frame}{Working internally to toposes}
-  \justifying
-  Let~$\C$ be a site. We recursively define
-  \[
-    U \models \varphi \quad
-    \text{(``$\varphi$ holds on~$U$'')}
-  \]
-  for objects~$U \in \C$ and \squiggly{formulas}~$\varphi$.
-  Write~``$\Sh(\C) \models \varphi$'' for~$U \models \varphi$ for all~$U$.
-  \only<1>{\footnotesize
-  \[ \renewcommand{\arraystretch}{1.08}\begin{array}{@{}l@{\ \ }c@{\ \ }l@{}}
-  U \models \top &\Ll& \text{true} \\
-  U \models \bot &\Ll& \hcancel{\text{false}}{0pt}{3pt}{0pt}{-2pt}\ \text{the
-  empty family is a covering of~$U$} \\
-  U \models s = t \? F &\Ll& s|_U = t|_U \in F(U) \\
-  U \models \varphi \wedge \psi &\Ll&
-  \text{$U \models \varphi$ and $U \models \psi$} \\
-  U \models \varphi \vee \psi &\Ll&
-  \hcancel{\text{$U \models \varphi$ or $U \models
-  \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $(U_i \to U)_i$} \\
-  && \quad\quad\text{such that for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
-  U \models \varphi \Rightarrow \psi &\Ll&
-  \text{for all~$V \to U$: }
-  \text{$V \models \varphi$ implies $V \models \psi$} \\
-  U \models \forall s \? F\_ \varphi(s) &\Ll&
-  \text{for all $V \to U$ and sections~$s_0 \in F(V)$: $V \models \varphi(s_0)$} \\
-  % U \models \forall F\_ \varphi(F) &\Ll&
-  % \text{for all $V \to U$ and sheaves~$F_0$ over~$V$: $V \models \varphi(F_0)$} \\
-  U \models \exists s \? F\_ \varphi(s) &\Ll&
-  \hcancel{\text{there exists $s_0 \in F(U)$ such that $U \models \varphi(s_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
-  &&
-  \text{there exists a covering $(U_i \to U)_i$ such that for all~$i$:} \\
-  && \quad\quad \text{there exists~$s_0 \in F(U_i)$ such that $U_i \models \varphi(s_0)$}
-  %U \models \exists F\_ \varphi(F) &\Ll&
-  %\hcancel{\text{there exists a sheaf $F_0$ on $U$ such that $U \models
-  %\varphi(F_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
-  %&&
-  %\text{there exists a covering $(U_i \to U)_i$ such that for all~$i$:} \\
-  %&& \quad\quad \text{there exists a sheaf~$F_0$ on $U_i$ such that $U_i \models \varphi(F_0)$}
-  \end{array} \]}
-
-  \only<2>{
-    \bigskip
-    \begin{columns}[t]
-      \begin{column}{0.45\textwidth}\justifying
-        \textbf{Example.}
-        $\Set \models \varphi$ if and only if $\varphi$ is true.
-        \bigskip
-
-        \textbf{Theorem.}
-        If~$\Sh(\C) \models \varphi$ and if $\varphi$ intuitionistically
-        entails~$\psi$, then also~$\Sh(\C) \models \psi$.
-      \end{column}
-
-      \begin{column}{0.45\textwidth}
-        \includegraphics[width=\textwidth,align=t]{mike-shulman-stack-semantics}
-        \vspace*{1.5em}
-      \end{column}
-    \end{columns}
-  }
-
-  \jnote{1}{
-    The internal language of a (Grothendieck or elementary) topos~$\E$ is a
-    device which allows us to speak and reason about the objects and morphisms
-    of~$\E$ in a naive element-based language close to the usual formal
-    mathematical language. Using this language, objects of~$\E$ look like plain
-    old sets [or types]; morphisms look like plain old maps between those sets;
-    epimorphisms look like surjections; group objects look like groups; and so on.
-
-    In particular, we can use the internal language to define what it means for a
-    given~$\TT$-structure in~$\E$ to be a model -- namely iff it looks like a
-    model from the internal point of view.
-
-    The internal language can be implemented by the \emph{Kripke--Joyal semantics},
-    a translation procedure which converts formulas of the internal language
-    into external statements about the objects and morphisms of~$\E$. The slide
-    displays some of the translation rules in the case that~$\E$ is a
-    Grothendieck topos.
-  }
-
-  \jnote{2}{
-    We can actually \emph{do mathematics} internally to a topos because the
-    Kripke--Joyal semantics is sound with respect to intuitionistic logic:
-    If~$\E \models \varphi$ and if~$\varphi$ intuitionistically entails a
-    further formula~$\psi$, then~$\E \models \psi$.
-
-    The Kripke--Joyal semantics can be extended to interpret unbounded
-    quantification (``for all sets'' as opposed to ``for all elements of the
-    particular set~$X$'') and dependent types. The former are for instance
-    required to express universal properties (``for all groups'', ``for all
-    rings''), and the latter are all over the place, even if their use might
-    not be particularly highlighted.
-
-    With these extensions, we can
-    import all of everyday constructive impredicative mathematics into the
-    internal world of a topos.
-
-    Some illustrations of working with the internal language can be found in these
-    sets of slides:
-    \begin{itemize}
-      \item
-      \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-leipzig2018.pdf}{Slides
-      for Jürgen Jost's group seminar at the MPI Leipzig}
-      \item
-      \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-como2018.pdf}{Slides
-      for \emph{Toposes in Como}}
-      (recording
-      \fixedhref{https://www.youtube.com/watch?v=SuELfC5P_R8}{available})
-    \end{itemize}
-    A longer exposition, with pointers to the literature, can be found in Section~2 of
-    \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these notes}.
-  }
-\end{frame}
-
-\section[Nongeometric properties]{A selection of nongeometric properties}
-
-\begin{frame}{A selection of nongeometric properties}
-  The \hil{generic object} validates:
-  \begin{enumerate}
-    \item $\forall x,y\?U_\TT\_ \neg\neg(x = y)$.
-    \item $\forall x_1,\ldots,x_n\?U_\TT\_ \neg \forall y\?U_\TT\_
-    \bigvee_{i=1}^n y = x_i.$
-    \item $(U_\TT)^{U_\TT} \cong 1 \amalg U_\TT$.
-  \end{enumerate}
-
-  The \hil{generic ring} validates:
-  \begin{enumerate}
-    \item $\forall x\?U_\TT\_ \neg\neg(x = 0)$.
-    \item $\forall x\?U_\TT\_ (x = 0 \Rightarrow 1 = 0) \Rightarrow (\exists
-    y\?U_\TT\_ xy = 1)$.
-  \end{enumerate}
-
-  The \hil{generic local ring} validates:
-  \begin{enumerate}
-    \item $\neg \forall x\?U_\TT\_ \neg\neg(x = 0)$.
-    \item $\forall a_0,\ldots,a_{n-1}\?U_\TT\_ \neg\neg \exists x\?U_\TT\_
-      x^n + a_{n-1}x^{n-1} + \cdots + a_0 x^0 = 0$.
-    \item \justifying Let~$\Delta = \{ \varepsilon \? U_\TT \,|\, \varepsilon^2 = 0 \}$.
-    For any map~$f : \Delta \to U_\TT$, there are unique elements~$a,b \?
-    U_\TT$ s.\,th.~$f(\varepsilon) = a + b\varepsilon$ for all~$\varepsilon
-    \? \Delta$.
-%   \item For any finitely presented~$U_\TT$-algebra~$A$,
-%   the canonical map~$A \to (U_\TT)^{\Hom_{U_\TT}(A,U_\TT)}$
-%   is an isomorphism.
-  \end{enumerate}
-
-  \jnote{1}{
-    The generic object, the generic model of the theory which has exactly one
-    sort and no function symbols, relations symbols or axioms, is
-    slightly indecisive: On the one hand, up to a double negation, it is a
-    subsingleton; on the other hand, it is infinite. This observation is due to
-    Carsten Butz and Peter Johnstone
-    (\fixedhref{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.16.4102&rep=rep1&type=pdf}{Classifying toposes for
-    first-order theories}).
-
-    The generic ring, the generic model of the theory of rings, is similarly
-    indecisive. It is infinite in the following sense:
-    \[ \forall x_1,\ldots,x_n \? U_\TT\_
-      \bigl(\forall y \? U_\TT\_ \bigvee_{i=1}^n (y = x_i)\bigr) \Rightarrow
-      1 = 0. \]
-
-    The theory of local rings is the quotient theory of the theory of rings
-    obtained by adding the axioms
-    \[ (1 = 0 \vdash \bot) \ \text{and}\ ((\exists z\_
-    (x+y)z = 1) \seq{x,y} (\exists z\_ xz = 1) \vee (\exists z\_ yz = 1)). \]
-    (Assuming the axiom of choice, a ring is local in this sense
-    iff it local in the usual sense (has exactly one maximal ideal).)
-
-    The first displayed property of the generic local ring illustrates
-    that nongeometric sequents need not be inherited by quotient theories.
-  }
-
-  \jnote{2}{
-    All of the displayed properties give rise to reduction techniques:
-    If we want to verify a geometric sequent for all rings, it suffices to
-    verify it for the generic ring; but the generic ring has additional
-    nongeometric properties not shared by every ring, such as the two displayed
-    ones. (This was for instance used
-    \fixedhref{https://www.sciencedirect.com/science/article/pii/0022404976900025}{by
-    Anders Kock} and
-    \fixedhref{https://link.springer.com/chapter/10.1007\%2FBFb0061835}{by
-    Gonzalo Reyes}.)
-
-    However, we face some challenges when pursuing these reduction techniques, including
-    the following:
-    \begin{enumerate}\justifying
-      \item It is not easy to determine interesting and useful properties of
-      the generic model.
-      \item The set of validated nongeometric sequents changes slightly
-      unpredictably when passing to quotient theories. For instance, when
-      proving that a geometric sequent holds for all rings, we may assume that
-      any element is \emph{not not} zero. But we may not assume this
-      simplification if we want to verify a geometric sequent for all local
-      rings (and if we want to exploit the given locality in the proof).
-      \item There is only so much we want to state and prove in full generality
-      for all rings, all local rings, all modules, and so on. We are often much
-      more interested in properties of particular mathematical objects.
-    \end{enumerate}
-  }
-
-  \jnote{3}{
-    The firstly-mentioned problem on the previous page is alleviated by the
-    Nullstellensatz presented in this talk, which gives a systematic and
-    universal source of nongeometric sequents validated by the generic model.
-    However, manual work is still required to reduce this set of sequents to a
-    smaller, manageable one consisting of memorable properties while hopefully
-    still preserving universality.
-
-    To counter the third problem, it's prudent to consider geometric theories
-    which depend on a given mathematical object of interest. For instance,
-    given a ring~$A$, we can consider the theory of prime ideals of~$A$, of
-    complemented prime ideals, of filters, and so on. The classifying toposes
-    of these theories are of independent interest -- in fact they are sheaf
-    toposes over certain important spaces in algebraic geometry -- and
-    nongeometric sequents validated by their generic models bundle nontrivial
-    information about~$A$. More details are on the following slide.
-  }
-\end{frame}
-
-\begin{frame}{An application in commutative algebra}
-  Let~$A$ be a reduced ring ($x^n = 0 \Rightarrow x = 0$).
-
-  The \hil{generic local localisation}~$A^\sim$ of~$A$ validates:
-
-  \vspace*{-1.8em}
-  \setbeamercolor{block body}{bg=red!30}
-  \setbeamercolor{structure}{fg=purple}
-  \begin{varblock}{\textwidth}{}
-    $A^\sim$ is a \hil{field}: $\forall x\?A^\sim\_ (\neg(\exists y\?A^\sim\_
-    xy = 1) \Rightarrow x = 0)$.
-
-    $A^\sim$ has \hil{$\boldsymbol{\neg\neg}$-stable equality}:
-    $\forall x,y\?A^\sim\_ \neg\neg(x = y) \Rightarrow x = y$.
-
-    \mbox{$A^\sim$ is \hil{anonymously Noetherian}.}\\[-1.2em]
-  \end{varblock}
-
-  This observation unlocks a short and conceptual proof of Grothendieck's
-  generic freeness lemma.
-
-  \jnote{1}{A \emph{local localisation} of a ring~$A$ is a
-  localisation~$A[S^{-1}]$ which is a local ring. The classifying topos of
-  local localisations of~$A$ is none other than the topos of sheaves over the
-  spectrum of~$A$, as studied in algebraic geometry; and the generic model is
-  also known as the structure sheaf~$\O_{\operatorname{Spec}(A)}$.
-
-  Assuming the Boolean prime ideal theorem, the geometric sequents
-  validated by~$A^\sim$ are easy to describe: They are precisely those which
-  are validated by all the stalks~$A_\ppp$ of~$A$.
-
-  But~$A^\sim$ enjoys further unique properties which are not shared by the
-  stalks of~$A$, other localisations of~$A$, quotients of~$A$ or indeed any
-  reasonable construction. Three of these are displayed on the slide.
-  (A ring is \emph{anonymously Noetherian} iff each of its ideals is
-  \emph{not not} finitely generated. Textbook proofs of Hilbert's basis theorem
-  are constructively valid for this Noetherian condition.)
-
-  The object~$A^\sim$ strikes a fine balance: On the one hand, it is still
-  close to~$A$, so that information learned about~$A^\sim$ teaches us
-  about~$A$; on the other hand, it enjoys unique properties rendering it
-  simpler than~$A$.
-
-  This balance allows for a simple and conceptually satisfying proof of
-  \emph{Gro\-then\-dieck's generic freeness lemma}, an important theorem in
-  algebraic geometry. Details can be found in
-  \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-padova2018.pdf}{this
-  set of slides}.}
-\end{frame}
-
-\section[A systematic source]{A systematic source of nongeometricity?}
-
-\begin{frame}{A systematic source of nongeometricity?}
-  \jnote{1-2}{
-    Mimicking the synthetic approach to differential geometry, synthetic
-    algebraic geometry is a framework for algebraic geometry in which schemes
-    are modelled by plain old sets, morphisms of schemes by plain old maps
-    between those sets, group schemes by plain old groups, and so on. Unlike
-    its close cousin, it is far less developed; some first steps are outlined
-    in Sections~19 and~20 of
-    \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these notes}.
-
-    Synthetic algebraic geometry is carried out internally to the \emph{big
-    Zariski topos} of a given base scheme; in the special case that the base
-    scheme is the terminal scheme~$\Spec(\ZZ)$, this topos is just the
-    classifying topos of local rings. The relevant generic model living in the
-    big Zariski topos is denoted~``$\affl$'' because it coincides with the
-    functor of points of the affine line.
-  }
-
-  \jnote{2}{
-    Marc Bezem, Ulrik Buchholtz and Thierry Coquand answered in their 2017 paper
-    \fixedhref{https://arxiv.org/abs/1712.07743}{Syntactic forcing
-    models for coherent logic} Gavin Wraith's question in the negative. (As
-    Thierry remarked at 6WFTop, this is even if one takes care to phrase
-    the question in a way to exclude the trivial counterexamples given by
-    instances of the law of excluded middle in the language of~$\TT$.) If the
-    answer had been positive, this would have given a neat, if somewhat hard to
-    use in practice, characterisation of the formulas validated by the generic model.
-  }
-
-  \jnote{3}{
-    The topos-theoretic \emph{Nullstellensatz}, to be presented on the next
-    slides, answers the displayed three questions in the affirmative.
-
-    Briefly, the Nullstellensatz is a certain statement in the language of a
-    given geometric theory~$\TT$ which is
-    \begin{itemize}\justifying
-      \item validated by the generic~$\TT$-model,
-      \item typically not validated by other~$\TT$-models, and
-      \item such that any statement validated by the generic~$\TT$-model can be
-      deduced, in intuitionistic logic, from the axioms of~$\TT$ and the
-      Nullstellensatz.
-    \end{itemize}
-
-    We believe that this characterisation is as explicit as it can get, but
-    would be delighted to be surprised by a future improvement. We stumbled on
-    it by playing with the synthetic quasicoherence statement, not least thanks
-    to encouragement by Alexander Oldenziel. However the route from that
-    statement to the Nullstellensatz is not quite direct; it turns out that
-    synthetic quasicoherence is a corollary of a specialisation of a
-    higher-order version of the Nullstellensatz to Horn theories.
-  }
-
-  \justifying
-  \textbf{Empirical fact.}
-  In \hil{synthetic algebraic geometry}, every known property of~$\affl$ followed from
-  its \hil{synthetic quasicoherence}:
-
-  \begin{quote}
-    \emph{\mbox{For any finitely presented~$\affl$-algebra~$A$, the ``double dual'' map}
-    \[ A \longrightarrow (\affl)^{\Hom_\affl(A,\affl)},\ s \longmapsto (x
-    \mapsto x(s)) \]
-    is an isomorphism of~$\affl$-algebras.}
-  \end{quote}
-
-  \vspace*{1.0em}
-  \begin{enumerate}
-    \item Does a general metatheorem explain this observation?
-    \item Is there a systematic source in any classifying topos?
-    \item Is there even an exhaustive source?
-  \end{enumerate}
-  \vspace*{-3.3em}
-
-  \visible<2>{\begin{tikzpicture}[overlay]
-    \draw[fill=white, draw=white, opacity=0.95] (-1,1.4) rectangle (\paperwidth,5.4);
-    \node[anchor=south west,inner sep=0] (image) at (0,0.9) {\vbox{
-      \only<2>{
-	\centering
-	\includegraphics[width=0.9\textwidth]{wraith-some-recent-developments-in-topos-theory} \\
-	\scriptsize
-        Gavin Wraith. \emph{Some recent developments in topos theory.} \\ In:
-        Proc.\@ of the ICM (Helsinki, 1978).
-      }
-    }};
-  \end{tikzpicture}}%
-
-  \[ \xymatrix{
-    \text{$\TT$ proves~$\alpha$\;} \ar@/_1pc/@{=>}[r] &
-    \text{\;$\alpha$ holds for~$U_\TT$\;} \ar@/_1pc/@{=>}[r]
-    \ar@/_1pc/@{=>}@[red][l]
-    |{\SelectTips{cm}{}\textbf{\scalebox{3}{\textcolor{red}{\!\texttimes\!}}}}|{} &
-    \text{\;$\alpha$ is~$\TT$-redundant}
-    \ar@/_1pc/@{=>}@[red][l]
-    |{\SelectTips{cm}{}\textbf{\scalebox{3}{\textcolor{red}{\!\texttimes\!}}}}|{}
-  } \]
-\end{frame}
-
-\section{A topos-theoretic Nullstellensatz}
-
-\newcommand{\nullstellensatz}{%
-  \justifying
-  \textbf{Theorem.} Internally to~$\Set[\TT]$:
-
-  \vspace*{-0.3em}
-  \begin{quote}\emph{\mbox{For any \explainstub{geometric$^\star$ sequent~$\sigma$}{red!20} over
-  the \explainstub{signature of~$\ul{\TT}/U_\TT$}{blue!20},}
-  if~\explainstub{$\sigma$ holds for~$U_\TT$}{yellow!70}, then~\explainstub{$\ul{\TT}/U_\TT$
-  proves~$\sigma$}{yellow!70}.}
-  \end{quote}
-  \vspace*{-0.6em}
-}
-
-\begin{frame}{A topos-theoretic Nullstellensatz}
-  \jnote{1-2}{
-    By~$\ul{\TT}$ we mean the geometric theory internal to~$\Set[\TT]$ obtained
-    by pulling back the set of sorts of~$\TT$, the set of function symbols and so on
-    along the unique geometric morphism~$\Set[\TT] \to \Set$. For instance,
-    if~$\TT$ is the theory of rings, then from the internal point of view
-    of~$\Set[\TT]$, the theory~$\ul{\TT}$ will again be the theory of rings.
-
-    The theory~$\ul{\TT}/U_\TT$ will be defined on the next slide. It is a certain
-    geometric theory internal to~$\Set[\TT]$.
-
-    The asterisk in \emph{geometric$^\star$ sequent}
-    indicate that any infinities used to index
-    disjunctions have to come from the base topos. This restriction is an important
-    subtlety, though not vital to this talk. If~$\TT$ is a coherent
-    theory, then we can restrict to coherent sequents and thereby vanquish the
-    subtlety.
-  }
-
-  \jnote{2}{
-    The algebraic Nullstellensatz states that, in some cases, algebraic truths
-    are witnessed by explicit \emph{algebraic certificates} -- syntactic
-    objects giving a priori reasons for why a given truth is to be expected.
-
-    In the topos-theoretic Nullstellensatz, algebraic truths are replaced by
-    arbitrary truths of the generic model, subject only to the condition that
-    they can be expressed as geometric sequents, and algebraic certificates
-    are replaced by \emph{logical certificates}: proofs.
-  }
-
-  \jnote{3}{
-    While, as stated on slide 1/10, the generic model~$U_\TT$ is a
-    conservative~$\TT$-model, the classifying topos~$\Set[\TT]$ does not
-    believe this fact. That is, the statement~``$U_\TT$ is a
-    conservative~$\ul{\TT}$-model'' is not true internally to~$\Set[\TT]$.
-    What is true is the modified statement~``$U_\TT$ is a
-    conservative$^\star$~$\ul{\TT}/U_\TT$-model''.
-
-    To be more precise, the following is true: \\
-    \mbox{For any geometric sequent~$\sigma$, if~$\Set[\TT] \models
-    \speak{$\sigma$ holds for~$U_\TT$}$, then $\TT$ proves~$\sigma$.}
-
-    The following is false in general: \\
-    \mbox{$\Set[\TT] \models \speak{For any geometric$^\star$ sequent~$\sigma$,
-    if~$\sigma$ holds for~$U_\TT$, then $\ul{\TT}$ proves~$\sigma$}$.}
-
-    The following is true: \\
-    \mbox{$\Set[\TT] \models \speak{For any geometric$^\star$ sequent~$\sigma$,
-    if~$\sigma$ holds for~$U_\TT$, then $\ul{\TT}/U_\TT$ proves~$\sigma$}$.}
-  }
-
-  \nullstellensatz
-
-  \pause
-
-  \textbf{The algebraic Nullstellensatz.}
-  Let~$A$ be a ring.
-  Let~$f,g \in A[X]$ be polynomials.
-  Then, subject to some conditions:
-  \[
-    \underbrace{\bigl(\forall x \in A\_ (f(x) = 0 \Rightarrow g(x) =
-    0)\bigr)}_{\text{algebraic truth}} \Longrightarrow
-    \underbrace{\bigl(\exists h \in A[X]\_ g = hf\bigr)}_{\text{algebraic certificate}}
-  \]
-  \vspace*{-0.9em}
-
-  \pause
-
-  \textbf{A naive version.}
-  ``Internally to~$\Set[\TT]$, for any geometric sequent~$\sigma$ over
-  the signature of~$\ul{\TT}$,
-  if~$\sigma$ holds for~$U_\TT$, then~$\ul{\TT}$
-  proves~$\sigma$.'' \textbf{False,} for instance with the theory of rings we
-  have
-  \begin{align*}
-    \Set[\TT] &\models \neg(\speak{$\ul{\TT}$ proves $(\top \vdash 1 + 1 =
-    0)$}) \\
-    \text{but}\quad \Set[\TT] & \not\models \neg(1 + 1 = 0).
-  \end{align*}
-\end{frame}
-
-\begin{frame}{A varying internal theory}
-  \nullstellensatz
-
-  \explainstub{\textbf{Definition.}}{blue!20} The theory~$\ul{\TT}/U_\TT$ is the
-  internal geometric theory of~\hil{$U_\TT$-algebras}, the theory which arises from~$\ul{\TT}$ by adding:
-  \begin{enumerate}
-    \justifying\small
-    \item for each element~$x : U_\TT$ a constant symbol~$e_x$,
-    \item for each function symbol~$f$ and~$n$-tuple~$(x_1,\ldots,x_n) : (U_\TT)^n$ the axiom
-    $(\top \vdash f(e_{x_1},\ldots,e_{x_n}) = e_{f(x_1,\ldots,x_n)})$,
-    \item for each relation symbol~$R$ and~$n$-tuple~$(x_1,\ldots,x_n) : (U_\TT)^n$ such
-    that~$R(x_1,\ldots,x_n)$ the axiom~$(\top \vdash R(e_{x_1},\ldots,e_{x_n}))$.
-  \end{enumerate}
-
-  \textbf{Remark.} Externalising the internal classifying
-  topos~$\Set[\TT][\ul{\TT}/U_\TT]$ yields the classifying topos
-  of~$\TT$-homomorphisms.
-
-  \jnote{1}{
-    Just as locales internal to a topos~$\E$ can be externalised to yield
-    localic geometric morphisms into~$\E$, internal Grothendieck toposes can be
-    externalised to yield bounded geometric morphisms. Since the composition of
-    bounded geometric morphisms is bounded, the externalisation of a
-    Grothendieck topos internally to a Grothendieck topos is itself a
-    Grothendieck topos, hence the classifying topos of some geometric theory.
-
-    Constructing internally to~$\Set[\TT]$, where~$\ul{\TT}/U_\TT$ looks just like an
-    ordinary geometric theory, the classifying topos of that theory, and then
-    externalising the resulting Grothendieck topos results in the classifying
-    topos of~$\TT$-homomorphisms. There are two canonical geometric morphisms
-    from this topos to~$\Set[\TT]$, the morphism computing the domain and the
-    morphism computing the codomain, and the morphism obtained by the
-    externalisation procedure is the former.
-  }
-\end{frame}
-
-\begin{frame}{Revisiting the test cases}
-  \nullstellensatz
-  \vspace*{-2em}
-  \[ \xymatrix{
-    & \text{$\ul{\TT}/U_\TT$ proves $\sigma$} \\
-    \text{$\sigma$ holds for $U_\TT$} \ar@/^1.1pc/@{<=>}[ur] &&
-    \text{$\sigma$ holds for $M$} \ar@/_1.1pc/@{<=}[ul]
-  } \]
-
-  \only<1>{
-    \textbf{In the object classifier.}
-  %  Assume that~$U_\TT$ is empty, that is $\forall x \? U_\TT\_ (\top \Rightarrow
-  %  \bot)$. By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves~$(\top \seq{x} \bot)$.
-  %  But this is false in the~$\ul{\TT}/U_\TT$-model~$U_\TT \amalg
-  %  \{\heartsuit\}$.
-    Let~$x,y \? U_\TT$. We want to verify~$\neg\neg(x = y)$. Hence assume that~$\neg(x = y)$.
-    By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves~$\sigma \defeq (e_x = e_y \vdash \bot)$.
-    Thus~$\sigma$ holds in any~$\ul{\TT}/U_\TT$-model.
-    But~$\sigma$ is false in the~$\ul{\TT}/U_\TT$-model~$M \defeq U_\TT/(x \sim
-    y)$. Hence~$\bot$.
-  }
-
-  \only<2->{
-    \textbf{In the ring classifier.}
-    Let~$f,g \? U_\TT[X]$ such that any zero of~$f$ is a zero of~$g$.
-    By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves this fact.
-    Hence it holds in the~$\ul{\TT}/U_\TT$-model~$U_\TT[X]/(f)$. In this
-    model~$f$ has the zero~$[X]$. Hence also~$g([X]) = 0$
-    in~$U_\TT[X]/(f)$, that is~$g = hf$ for some~$h \? U_\TT[X]$.
-  }
-
-  \jnote{1}{
-    This slide gives an example how to use the Nullstellensatz to deduce
-    properties of the generic model. A couple of remarks are in order.
-
-    The Nullstellensatz is trivial for sequents~$\sigma$ of the form~$(\top
-    \vdash \psi)$. The Nullstellensatz is only interesting in case
-    that~$\sigma$ has a nontrivial antecedent or is set in a nonempty context.
-
-    Since the converse direction in the Nullstellensatz also holds
-    (because~$U_\TT$ is a~$\ul{\TT}/U_\TT$-model), the
-    statements~$\speak{$\sigma$ holds for~$U_\TT$}$
-    and~$\speak{$\ul{\TT}/U_\TT$ proves $\sigma$}$ are equivalent. This
-    equivalence is intriguing from a logical point of view, since the former
-    statement is a geometric implication while the latter can be put as a
-    geometric formula. (Up to a subtle issue indicated on the next slide.)
-
-    To apply the Nullstellensatz, no description of a site defining~$\Set[\TT]$
-    is required.
-
-    Often when using the Nullstellensatz, we go from a (perhaps assumed) truth
-    of~$U_\TT$ via provability to another model~$M$
-    of~$\ul{\TT}/U_\TT$. That is, we use provability as a (one-way)
-    \emph{bridge}.
-  }
-\end{frame}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{7.95cm}\hfill\mbox{\includegraphics[width=0.13\paperwidth]{hen}\;\;}\end{minipage}}
-\begin{frame}{Exhaustion and extensions}
-  \justifying
-  \textbf{Theorem 1.} A first-order formula holds for~$U_\TT$ iff it is
-  intuitionistically provable from the axioms of~$\TT$ and
-  the scheme
-  \begin{equation}\tag{Nullstellensatz} \speak{$\sigma$ holds} \implies \speak{$\ul{\TT}/U_\TT$
-  proves~$\sigma$}. \end{equation}
-  \vspace*{-1.5em}
-
-  \textbf{Theorem 2.}
-  Let~$\TT'$ be a quotient theory
-  of~$\TT$. Assume that~$U_\TT$ is contained in the subtopos~$\Set[\TT']$. Then internally
-  to~$\Set[\TT']$:
-  \begin{quote}
-    \emph{A geometric$^\star$ sequent~$\sigma$ with Horn consequent holds for~$U_{\TT'}$
-    iff~$\ul{\TT}/U_\TT$ proves~$\sigma$.}\par
-  \end{quote}
-
-  \textbf{Theorem 3.} A higher-order formula holds for~$U_\TT$ iff it is
-  provable in intuitionistic higher-order logic from the axioms of~$\TT$ and
-  the higher-order Nullstellensatz scheme.
-
-  \jnote{1}{
-    Theorem~1 states that the source provided by the Nullstellensatz is
-    exhaustive. It is possible, but not entirely obvious, how to formulate the
-    Nullstellensatz as an (infinitary) first-order formula.
-
-    Theorem~2 provides a useful variant of the Nullstellensatz. Its
-    assumptions are for instance satisfied if~$\TT$ is the theory of rings
-    and~$\TT'$ is the theory of local rings. When applicable, it can be used to
-    avoid doubly-internal toposes. It also explains, for instance, why in the
-    formulation of synthetic quasicoherence no local rings appear even though
-    the relevant topos is the classifying topos of local rings.
-
-    Theorems~3 indicates that the Nullstellensatz can be generalised to the
-    higher-order setting.
-
-    Written details on all of this are
-    \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-qcoh.pdf}{slowly
-    emerging}.
-  }
-
-  \jnote{2}{
-    The Nullstellensatz is related to several precursors. A corollary of the
-    Nullstellensatz is that, over the first-order theory validated by~$U_\TT$, any
-    first-order formula is in fact logically equivalent to a geometric formula.
-    This corollary has already been observed by Carsten Butz and Peter
-    Johnstone in their paper
-    \fixedhref{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.16.4102&rep=rep1&type=pdf}{Classifying
-    toposes for first-order theories} (Lemma~4.2 there). At that point, a
-    characterisation of the first-order formulas in the general case, of the
-    form as in Theorem~1, was still missing.
-
-    Theorem~3 rests on a relativisation of Olivia Caramello's completeness theorem,
-    Theorem~2.4(ii) in her paper
-    \fixedhref{https://arxiv.org/abs/0906.3061}{Universal models and
-    definability}. The passage from the external to the internal phrasing
-    requires going from~$\TT$ to~$\ul{\TT}/U_\TT$.
-  }
-
-  \jnote{3}{
-    Plans for the future include:
-    \begin{itemize}\justifying
-      \item Developing an Agda library for dealing with the internal language
-      of toposes and related kinds of categories, employing Agda's
-      meta-programming facilities; with such a library at hand,
-      formalising the Nullstellensatz in Agda.
-      \item Exploring Nullstellensatz-style results for arithmetic universes.
-      \item Applying the Nullstellensatz in constructive algebra and algebraic
-      geometry, along the lines of generic freeness and synthetic algebraic
-      geometry. Most of the toposes in geometric use are actually uncharted
-      territory from a logical point of view. The talk by Matthias Hutzler will
-      present new results in that direction.
-    \end{itemize}
-  }
-\end{frame}}
-
-% Analogy with the harmonic oscillator in the study of physics
-
-% * The generic model
-%   * The mystery of nongeometric sequents
-%   * Construction of the generic model
-% * Kripke--Joyal semantics; focus on existence of extensions
-% * Examples for theories and nongeometric sequents with a focus on the wlog
-%   aspect
-% * Why we are interested in the generic models themselves
-%   * Universal localisation
-%   * Grothendieck's generic freeness lemma
-% * Synthetic quasicoherence
-% * Conservativity from the internal point as guiding question
-%   * Naive formulation and why it doesn't work
-%   * Recall of the algebraic Nullstellensatz
-% * The new Nullstellensatz
-%   * What Set[T][T/U_T] classifies
-%   * Kock again
-%   * Subtleties regarding infinites
-%   * Improvement in the Horn case
-% * Universality of the source provided by the Nullstellensatz
-
 \end{document}
-
-% Are there size issues when KJ'ing internally?
-% Comparison to the approach by Vickers?
-% When does a theory prove the Nullstellensatz?
-% Look up what Kock and Reyes used the wlog for.
-
-% An Tafel: geometric formula, local ring
-% Erwähnen: Marc Bezem, Ulrik Buchholtz, Thierry Coquand
-
-% Grothendieck quote for the title picture:
-%
-% Ces “nuages probabilistes”, remplaçant les rassurantes particules matérielles
-% d’antan, me rappellent étrangement les élusifs “voisinages ouverts” qui
-% peuplent les topos, tels des fantômes évanescents, pour entourer des “points”
-% imaginaires, ...
-%
-% Diese "Wahrscheinlichkeitswolken", welche die beruhigenden materiellen
-% Partikel von früher ersetzen, erinnern mich irgendwie an die flüchtigen
-% "offenen Umgebungen" der Topoi -- wie dahinschwindende Phantome, um
-% die fiktiven "Punkte" zu umgeben, ...
-%
-% These “probability clouds”, replacing the reassuring material particles of
-% before, remind me strangely of the elusive “open neighborhoods” that populate
-% the topoi, like evanescent phantoms, to surround the imaginary “points”.
-
-% Sei T eine Theorie, für die ein konservatives Modell in Set existiert.
-% Dann haben wir eine Surjektion Set --> Set[T]. Also existiert eine
-% geometric expansion von T, welche Morita-äquivalent zur leeren Theorie ist.

### 2021-11-20 slides-dagstuhl2021.tex
More slides for Dagstuhl

@@
 \usepackage{pgfpages}
+\usepackage[export]{adjustbox}
 %\setbeameroption{show notes on second screen}
@@
   \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
 
@@
   \bigskip
+  \bigskip
+  \bigskip
 
@@
 
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
 \begin{frame}{Algebraic geometry in a nutshell}
@@
   \[ a^n + b^n = c^n. \]
-\end{frame}
+\end{frame}}
 
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.95cm}\includegraphics[width=\paperwidth]{staircase}\end{minipage}}
 \begin{frame}{Transfinite methods}
-  The current presentation of algebraic geometry employs:
+  The current presentation of algebraic geometry hinges on:
   \begin{itemize}
     \item large structures: classes, large categories, universes, \ldots
+    \item powersets
     \item law of excluded middle
@@
     \item subject matter (in part) very concrete
-    \item computer algebra systems for computations exist
-    \item constructive algebra well-developed
+    \item practical computer algebra systems for computations
+    \item well-developed constructive algebra
   \end{enumerate}
+\end{frame}}
+
+\begin{frame}{Formalizing algebraic geometry}
+  \bigskip
+  \bigskip
+  \bigskip
+  \includegraphics[width=\textwidth]{schemes-in-lean2}
+
+  \vspace*{-17em}
+  \includegraphics[frame=0.1em,bb=0 170 610 625,width=0.3\textwidth]{schemes-in-coq}\hfill
+  \includegraphics[frame=0.1em,bb=0 170 610 625,width=0.3\textwidth]{schemes-in-lean}
+\end{frame}
+
+\begin{frame}{A trinitarian challenge}
+  \begin{center}\tiny $1^3 + \cdots + n^3 = (1 + \cdots + n)^2$\end{center}
+  \vspace*{-0.6em}
+
+  \begin{columns}
+    \begin{column}{0.33\textwidth}
+      \centering
+      \includegraphics[height=7em]{lem}
+
+      \hil{constructivity}
+      \par
+    \end{column}
+    \begin{column}{0.33\textwidth}
+      \centering
+      \includegraphics[height=7em]{sum-of-cubics}
+
+      \hil{elegance}
+      \par
+    \end{column}
+
+    \begin{column}{0.33\textwidth}
+      \centering
+      \includegraphics[height=7em]{proof-assistants}
+
+      \par
+      \hil{formalizability}
+    \end{column}
+  \end{columns}
+  \bigskip
+  \bigskip
+
+  \subhead{Thesis}
+  \centering
+  Elegant mathematics lends itself to rewarding formalization.
 \end{frame}

### 2021-11-21 slides-dagstuhl2021.tex
Slides finished(?)

@@
 
+\begin{frame}{Foundational possibilities}
+  \small
+
+  \subhead{Schemes and scheme morphisms}
+  \only<1-2>{
+  \begin{itemize}
+    \setlength\itemsep{-0.2em}
+    %\item locally ringed spaces (or locales, distributive lattices, sites,
+    %toposes, arithmetic universes), formal geometries
+    \item locally ringed \ldots \\
+    \ldots{} spaces \tabto{4cm}-- \textit{unconstructive, hard to formalize} \\
+    \ldots{} locales \tabto{4cm}-- \textit{impredicative, superfluous opens} \\
+    \ldots{} distributive lattices \tabto{4cm}-- \textit{need extension from basis} \\
+    \ldots{} sites \tabto{4cm}-- \textit{current favorite! morphisms intricate} \\
+    \ldots{} toposes \tabto{4cm}-- \textit{large structure} \\
+    \ldots{} arithmetic universes \tabto{4cm}\mbox{-- \textit{a bit better;
+    issue with relative spectrum}}
+    \item formal geometries \tabto{4cm}-- \textit{can be regarded as sites}
+    \item functor of points \tabto{4cm}-- \textit{large structure, issue with
+    sizes or \\ \qquad
+    with schemes not of finite presentation over the base ring}
+    \item formal gluing data \tabto{4cm}-- \textit{morphisms intricate}
+  \end{itemize}
+  \textbf{A surprise of uncertain import:}
+  \pause
+  Internally to the big Zariski topos of a base scheme, the Zariski spectrum of
+  a finitely presented algebra does have enough points! [Cherubini--Coquand]}
+  \bigskip
+  \pause
+
+  \subhead{Zariski cohomology}
+  \begin{itemize}
+    \setlength\itemsep{-0.2em}
+    \small
+    \item Čech cohomology \tabto{4.5cm}-- \textit{ad hoc, but fine for}
+    \tabto{4.5cm}\phantom{--} \textit{quasicompact separated schemes}
+    \item injective resolutions \tabto{4.5cm}-- \textit{hopelessly unconstructive}
+    \item flabby resolutions \tabto{4.5cm}-- \textit{probably unconstructive}
+    \item pointwise Kan extensions \tabto{4.5cm}-- \textit{fine!}
+  \end{itemize}
+  \bigskip
+
+  \subhead{Étale cohomology}
+  \textcolor{red!90}{??}
+\end{frame}
+
 \end{document}

### 2021-11-23 slides-dagstuhl2021.tex
Finished slides

@@
 
-\title{Bridging the foundational gap: updating algebraic geometry in
+\title{Bridging the foundational gap: Updating algebraic geometry in
 face of current challenges regarding formalizability, constructivity and
@@
     \node (pretitle) {Bridging the foundational gap:};
-    \node (title) [below of=pretitle] {\vbox{\vspace*{-0.5em}updating \hil{algebraic
+    \node (title) [below of=pretitle] {\vbox{\vspace*{-0.5em}Updating \hil{algebraic
     geometry} in face of \hil{current challenges} regarding
@@
       \draw (0,0) circle (1);
-      \draw [fill=white] (90:1) circle (0.2);
+      \draw [fill=white] (90:1) circle (0.3);
     \end{tikzpicture}
@@
       \draw (0,0) circle (1);
-      \draw [fill=white] (270:1) circle (0.2);
+      \draw [fill=white] (270:1) circle (0.3);
     \end{tikzpicture}
@@
       \draw (0,0) circle (1);
-      \draw [fill=white] (90:1) circle (0.2);
-      \draw [fill=white] (270:1) circle (0.2);
+      \draw [fill=white] (90:1) circle (0.3);
+      \draw [fill=white] (270:1) circle (0.3);
     \end{tikzpicture}
@@
 \begin{frame}{Transfinite methods}
-  The current presentation of algebraic geometry hinges on:
+  The standard presentation of algebraic geometry hinges on:
   \begin{itemize}
@@
     \item subject matter (in part) very concrete
-    \item practical computer algebra systems for computations
-    \item well-developed constructive algebra
+    \item computer algebra systems for computations practical
+    \item constructive algebra well-established
+    \item high-level proofs often constructive
   \end{enumerate}
@@
     \item locally ringed \ldots \\
-    \ldots{} spaces \tabto{4cm}-- \textit{unconstructive, hard to formalize} \\
+    \ldots{} topological spaces \tabto{4cm}-- \textit{unconstructive, hard to formalize} \\
     \ldots{} locales \tabto{4cm}-- \textit{impredicative, superfluous opens} \\
@@
     \item injective resolutions \tabto{4.5cm}-- \textit{hopelessly unconstructive}
+    \item dynamical injectives \tabto{4.5cm}-- \textit{??}
     \item flabby resolutions \tabto{4.5cm}-- \textit{probably unconstructive}
-    \item pointwise Kan extensions \tabto{4.5cm}-- \textit{fine!}
+    \item pointwise Kan extensions \tabto{4.5cm}-- \textit{fine! partially
+    defined $\mathbb{R}\Gamma$;} \tabto{4.5cm}-- \textit{existence
+    verified for} \tabto{4.5cm}\phantom{--} \textit{quasicompact separated schemes;}
+    \tabto{4.5cm}\phantom{--} \textit{hyper coverings?}
   \end{itemize}
-  \bigskip
 
@@
 \end{document}
-
-\section{The generic model}
-
-\subsection{The generic ring}
-
-\begin{frame}{The generic ring}
-  \visible<2->{``}Let~$R$ be a ring.\visible<2->{''} \visible<2->{-- Which ring does this phrase refer to?}
-
-  \vspace*{-0.6em}
-  \begin{center}
-    \begin{tikzpicture}[ultra thick, node distance=7mm]
-      \node[rectangle, rounded corners=1pt, inner sep=5pt, draw=lime!80, fill=lime!40] (a) {$\ZZ$};
-      \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40, right=of a] (b) {$\FF_2$};
-      \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40, right=of b] (c) {$\QQ[X]$};
-      \node[regular polygon, regular polygon sides=6, draw=orange!80, fill=orange!40, right=of c, rounded corners=1pt] (d) {$\RR$};
-      \node[regular polygon, regular polygon sides=5, draw=pink!80, fill=pink!40, right=of d, rounded corners=1pt, inner sep=0cm] (e) {$\O_X$};
-      \visible<3->{\node[star, rounded corners=1pt, star points=10, inner
-      sep=2pt, draw=purple!80, fill=purple!30, right=of e] {$\AA$};}
-    \end{tikzpicture}
-  \end{center}
-  \vspace*{-0.6em}
-
-  \pause
-  \pause
-  \pause
-
-  \justifying
-  \textbf{Thm.} For any$^\star$ property~$P$ of rings, the following are
-  equivalent: \\[0em]
-  \begin{enumerate}
-    \item The \hil{generic ring}~$\AA$ has property~$P$.
-    \item Every$^\star$ ring has property~$P$.
-    \item The ring axioms entail property~$P$.
-  \end{enumerate}
-  \pause
-
-  \textbf{Example A.} For any~$x,y,z \in \AA$, $x + (y + z) = (x + y) + z$.
-  \pause
-
-  \textbf{Example B.} \only<6>{Is~$1 + 1 = 0$ in~$\AA$?}%
-  \pause
-  \only<7->{It is \hil{not the case} that~$1 + 1 = 0$ in~$\AA$. \only<8->{But also:}} \\
-  \pause
-  \phantom{\textbf{Example B.}} It is \hil{not the case} that~$1 + 1 \neq 0$ in~$\AA$.
-  \pause
-
-  \textbf{Example C (Anders Kock).} The generic ring is a \hil{field}:
-  \[ \forall x \in \AA\_ \bigl((x = 0 \Rightarrow 1 = 0) \Rightarrow (\exists y
-  \in \AA\_ xy = 1)\bigr). \]
-
-  \pause
-  \vspace*{-0.6em}
-  \hil{Hence:} When verifying a coherent sequent for all rings, can without loss of
-  generality assume the field condition.
-\end{frame}
-
-\begin{frame}{A selection of noncoherent sequents}
-  The \hil{generic object}~$\MM$ validates:
-  \begin{enumerate}
-    \item $\forall x,y \in \MM\_ \neg\neg(x = y)$.
-    \item $\forall x_1,\ldots,x_n \in \MM\_ \neg \forall y \in \MM\_
-    y = x_1 \vee \cdots \vee y = x_n$.
-  \end{enumerate}
-
-  The \hil{generic ring}~$\AA$ validates:
-  \begin{enumerate}
-    \item $\forall x \in \AA\_ (x = 0 \Rightarrow 1 = 0) \Rightarrow (\exists y \in \AA\_ xy = 1)$.
-    \item $\forall x \in \AA\_ \neg\neg(x = 0)$.
-  \end{enumerate}
-
-  The \hil{generic local ring}~$\AA'$ validates:
-  \begin{enumerate}
-    \item $\forall x \in \AA'\_ (x = 0 \Rightarrow 1 = 0) \Rightarrow (\exists y \in \AA'\_ xy = 1)$.
-    \item $\neg \forall x \in \AA'\_ \neg\neg(x = 0)$.
-    \item $\forall f \in \AA'[X]_{\text{degree} > 0}\_ \neg\neg \exists x \in \AA'\_ f(x) = 0$.
-  \end{enumerate}
-\end{frame}
-
-\begin{frame}{An application in commutative algebra}
-  \justifying
-  Let~$A$ be a reduced ring ($x^n = 0 \Rightarrow x = 0$).
-  Let~$\pp$ be the \hil{generic prime ideal$^\star$} of~$A$.
-  Then~$A_\pp \defeq A[\pp^{-1}]$ validates:
-
-  {\vspace*{-1.2em}
-  \setbeamercolor{block body}{bg=red!30}
-  \setbeamercolor{structure}{fg=purple}
-  \begin{varblock}{\textwidth}{}
-    $A_\pp$ is a \hil{field}: $\forall x \in A_\pp\_ (\neg(\exists y \in A_\pp\_
-    xy = 1) \Rightarrow x = 0)$.
-
-    $A_\pp$ has \hil{$\boldsymbol{\neg\neg}$-stable equality}:
-    $\forall x,y \in A_\pp\_ \neg\neg(x = y) \Rightarrow x = y$.
-
-    \mbox{$A_\pp$ is \hil{anonymously Noetherian}.}\\[-1.2em]
-  \end{varblock}}
-
-  This observation unlocks a short and conceptual proof of Grothen\-dieck's
-  \hil{generic freeness lemma} in algebraic geometry.
-
-  \mbox{\textbf{Thm.} (baby freeness) Let~$M$ be an~$A$-module. Then~\bignumber{1}
-  implies~\bignumber{3}.}
-
-  {\small
-  \begin{tabular}{@{}lc@{\ }l@{}}
-    \bignumber{1} $M$ is finitely generated &($\Longleftrightarrow$&
-    $M_\pp$ is finitely generated) \\
-    \bignumber{2} $M$ is locally free &($\Longleftrightarrow$&
-    $M_\pp$ is free) \\
-    \bignumber{3} $M$ is locally free \emph{on a dense open} &($\Longleftrightarrow$&
-    $M_\pp$ is \emph{not not} free) \\
-  \end{tabular}\par}
-
-  \textbf{Proof.} Elementary linear algebra over~$A_\pp$. \qed
-\end{frame}
-
-\begin{frame}{A systematic source}
-  \begin{center}
-    \includegraphics[width=0.8\textwidth]{wraith-some-recent-developments-in-topos-theory} \\
-    \scriptsize
-    Gavin Wraith. \emph{Some recent developments in topos theory.} \\ In:
-    Proc.\@ of the ICM (Helsinki, 1978).
-  \end{center}
-
-  \textbf{Thm. (Nullstellensatz):} The generic~$\TT$-model~$U_T$ validates:
-  For any coherent sequent~$\sigma$,
-  \[
-    \text{$\sigma$ holds for $U_\TT$} \quad\textcolor{purple}{\Longleftrightarrow}\quad
-    \text{$\ul{\TT}/U_\TT$ proves $\sigma$.}
-  \]
-
-  \justifying
-  \textbf{Thm. (universality):} The generic~$\TT$-model validates a first-order
-  formula~$P$ if and only if~$P$ is intuitionistically deducible from the
-  axioms of~$\TT$ and the Nullstellensatz.
-\end{frame}
-
-\begin{frame}{Arithmetic universes}
-  Places where we can do mathematics (among others):
-  \begin{multicols}{2}
-    \begin{enumerate}
-      \item $\Set$ (sets)
-      \item $\Eff$ (data types) \columnbreak
-      \item \mbox{$\mathrm{sSet}$ (simplicial sets)}
-      \item \mbox{$\Sh(X)$ (sheaves over~$X$)}
-    \end{enumerate}
-  \end{multicols}
-  \pause
-  \vspace*{-0.6em}
-  These are examples for \hil{arithmetic universes}.
-  \pause
-
-  \justifying
-  \textbf{Definition.} An \emph{arithmetic universe} is a category with finite
-  limits (``$\times$''), stable finite disjoint coproducts (``$\amalg$''),
-  stable ef{}fective quotients (``$X/{\sim}$'') and parametrized list objects
-  (``$\NN$'', ``$\mathrm{List}(X)$'').
-
-  \textbf{Thm.} Any statement which is provable in \hil{predicative constructive
-  mathematics} (no powersets, no $\varphi \vee \neg\varphi$, no
-  \mbox{$\neg\neg\varphi \Rightarrow \varphi$}, no axiom of choice) is true in any
-  arithmetic universe.
-  \pause
-
-  Further examples:
-
-  \begin{enumerate}
-    \addtocounter{enumi}{4}
-    \item the \hil{initial} arithmetic universe
-    \item the \hil{classifying} arithmetic universe for the theory of rings
-  \end{enumerate}
-\end{frame}
-
-\end{document}

### 2022-01-05 paper-flabby-objects.tex
First round of improvements

@@
   and study their basic properties. In the special case of localic toposes, this
-  notion reduces to the common notion of flabby sheaves, yielding a
+  notion reduces to the established notion of flabby sheaves, yielding a
   site-independent characterization of flabby sheaves. Continuing a line of
@@
 statements, also significant improvements in proof length and proof complexity
-can be gained. For instance, Grothendieck's generic freeness lemma admits a
+can be obtained. For instance, Grothendieck's generic freeness lemma admits a
 short and simple proof in this framework, while previously-published proofs
@@
 internal injectivity condition and the internal construction of sheaf
-cohomology.
+cohomology via injective resolutions.
 
@@
 in Section~\ref{sect:in-eff} and in Section~\ref{sect:conclusion} that flabby
-resolutions present their own challenges, and in summary we failed to reach our
-goal. To the best of our knowledge, the problem of giving a constructive account
+resolutions present their own challenges, and in summary we failed to reach this
+goal. The problem of giving a constructive account
 of sheaf cohomology is still open.
@@
 requires different commitments: For the alternative definition,
-properties~(1),~(3b) and~(4a) can be verified purely intuitionistically.
+properties~(1),~(3b),~(4a) and~(4b) can be verified purely intuitionistically.
 There is a substitute for property~(3a) which can be verified purely
-intuitionistically. We do not know whether property~(4b) can be established
-purely intuitionistically, but we give a rudimentary analysis in
-Section~\ref{sect:conclusion}.
+intuitionistically.
 
@@
 be extended to a section on~$U \cup U_i$.\end{defn}
+% XXX choice-free formulation
 
@@
 
-\begin{rem}Although we will see in Section~\ref{sect:in-eff} that there is some
+\begin{rem}\label{rem:constant-flabby}
+Although we will see in XXX and in Section~\ref{sect:in-eff} that there is some
 relation between flabby sets and~$\neg\neg$-separated sets, neither notion
@@
 % Take care if numbering changes (referenced below)
-\item Any set embeds into an injective set.
+\item Any set embeds canonically (in a uniform fashion) into an injective set.
 \item Any injective module is also injective as a set.
-\item Assuming the axiom of choice, any module embeds into an
+\item Assuming Zorn's lemma (but not the law of excluded middle, hence not the
+axiom of choice), any module embeds into an
 injective module.
@@
 subterminals~$\P_{\leq 1}(X)$ are each injective.
-\item The forgetful functor from modules to sets possesses a monomorphism-preserving left
-adjoint. More explicitly, if~$i : A \to B$ is an injective map between sets and
-if~$f : A \to I$ is an arbitrary map, then the induced map~$R\langle A \rangle
+\item This statement follows from general abstract nonsense, since the forgetful functor from modules to sets possesses a monomorphism-preserving left
+adjoint. More explicitly, let~$I$ be an injective~$R$-module, let~$i : A \to B$ be an injective map between sets and
+let~$f : A \to I$ be an arbitrary map. Then the induced map~$R\langle A \rangle
 \to R\langle B \rangle$ between free modules is also injective, the given
@@
 \begin{cor}\label{cor:enough-flabby-sets}
-Any set embeds into a flabby set.\end{cor}
+Any set canonically embeds into a flabby set.\end{cor}
 
@@
 
-\begin{prop}Any singleton set is flabby. The cartesian product of flabby sets
+\begin{prop}Any singleton set is flabby. The binary cartesian product of flabby sets
 is flabby.\end{prop}
@@
 flabby in view of Corollary~\ref{cor:enough-flabby-sets}.
+
+% XXX enough flabby modules
 
@@
 is an element~$v \in M'$ such that~$i^{-1}[K - u_0] \subseteq \{v\}$. We verify
-that~$K \subseteq \{u_0 + i(v)\}$.
+that~$K \subseteq \{u_0 + i(v)\}$:
 
@@
 the generic stage: Let~$\P_{\leq1}(X)$ denote the \emph{object of subterminals}
-of~$X$; this object is a certain subobject of~$\P(X) = [X,\Omega_\E]_\E$, the
-powerobject of~$X$. The subobject~$K_0$ of~$X \times \P_{\leq1}(X)$ classified by the
+of~$X$; this object is a certain subobject of the powerobject~$\P(X) = [X,\Omega_\E]_\E$.
+The subobject~$K_0$ of~$X \times \P_{\leq1}(X)$ classified by the
 evaluation morphism~$X \times \P_{\leq1}(X) \to X \times \P(X) \to \Omega_\E$
@@
 \to X$ is given by the sections~$s_i|_{(V \cup V_i) \cap U}$.
+% XXX choicefree
 
@@
 \end{proof}
+
+\begin{rem}\label{rem:flabby-global}
+Some condition on the topos is necessary for flabby objects to
+possess global elements. An example is given by the~$G$-set~$G$ (with the
+translation action), considered as an object of the topos~$BG$ of~$G$-sets,
+where~$G$ is a nontrivial group. This object is flabby (because it is inhabited
+and~$BG$ is a Boolean topos, assuming the law of excluded middle in the
+metatheory), but it does not have any global elements.
+\end{rem}
 
@@
 is for instance satisfied if~$f$ is a local geometric morphism.
+
+\begin{rem}Pullbacks of flabby objects along geometric morphisms are usually
+not flabby. For instance, constant sheaves can fail to be flabby
+(Remark~\ref{rem:constant-flabby}) but arise as pullbacks along geometric
+morphisms to the topos~$\Set$, in which most objects are flabby (assuming the
+law of excluded middle).\end{rem}
 
@@
 \begin{enumerate}
-\item Any object embeds into an internally injective object.
+\item Any object embeds canonically into an internally injective object.
 \item (If~$\E$ has a natural numbers object.) The underlying unstructured
@@
 \item Any internally injective object is flabby.
-% Take care if numbering changes (referenced below).
-\item Any object embeds into a flabby object.
+\item Any object embeds canonically into a flabby object.
+\item Any internal module embeds canonically into an internal module which is flabby.
 \item The terminal object is flabby. The product of flabby objects is flabby.
@@
 
+% XXX If Zorn's lemma is not available but E is localic, ... is at least true
+% if E is compact.
+
 \begin{rem}A direct generalization of the traditional notion of a flabby sheaf, as
@@
 flabby in the traditional sense if and only if~$F$ is a strongly flabby object
-of~$\Sh(T)$.
+in~$\Sh(T)$, if and only if~$F$ is a flabby object in the presheaf
+topos~$\PSh(T)$.
 
@@
 and therefore cannot be characterized in the internal language. A specific
-example is the~$G$-set~$G$ (with the translation action), considered as an
-object of the topos~$BG$ of~$G$-sets, where~$G$ is a nontrivial group.
+example is the~$G$-set~$G$ (with the translation action) as in
+Remark~\ref{rem:flabby-global}.
 This object is not strongly flabby, since the morphism~$\emptyset \to 1$ does
@@
 
-Assuming the axiom of choice, there are enough injective sheaves of modules so
+Assuming Zorn's lemma, there are enough injective sheaves of modules so
 that this recipe can be carried out. The resulting sheaf
@@
 such a way that the category of internal sheaves over this internal locale
-coincides with~$\Sh(Y)$; in particular, a given sheaf~$E$ of~$\O_Y$-modules can
+coincides with~$\Sh(Y)$ [XXX give reference]; in particular, a given sheaf~$E$ of~$\O_Y$-modules can
 be regarded as a sheaf over~$I(Y)$. Under this equivalence, the morphism~$f : Y
@@
 requires Zorn's lemma to hold internally to~$\Sh(X)$; this is indeed the case
-since we assume Zorn's lemma in the metatheory and since the validity of Zorn's
+since we assume Zorn's lemma in the metatheory and the validity of Zorn's
 lemma passes from the metatheory to localic
@@
 Let~$f : Y \to X$ be a ringed locale over a locale~$X$. Let~$E$ be a sheaf of modules over~$Y$.
-Assuming the axiom of choice in the metatheory, the expression~``$H^n(I(Y), E)$''
+Assuming Zorn's lemma in the metatheory, the expression~``$H^n(I(Y), E)$''
 of the internal language of~$\Sh(X)$ denotes a globally-defined sheaf
-over~$X$, and this sheaf coincides with~$R^n f_*(E)$.\end{thm}
+on~$X$, and this sheaf coincides with~$R^n f_*(E)$.\end{thm}
 
@@
 of~$\Sh(X)$. Hence we can, internally to~$\Sh(X)$, carry out the construction of~$H^n(I(Y), E)$.
-Externally, this yields an open covering of~$X$ such that we have, for each
+Externally, this construction yields an open covering of~$X$ such that we have, for each
 member~$U$ of that covering
@@
 toposes, away from their original conceptual home. We begin this study with
-establishing the following observations on flabby objects in the \effective
+establishing the following observation on flabby objects in the \effective
 topos. We follow the terminology of Martin Hyland's survey on the \effective
@@
 \begin{prop}\label{prop:flabby-effective-sets}
-Let~$X$ be a flabby object in the \effective topos. Let~$f : X \to X$
-be a morphism. If~$X$ is \effective, the statement ``$f$ has a fixed point''
-holds in the \effective topos.
+The only object in the \effective topos which is both flabby and \effective is the
+singleton object.
 \end{prop}
 
-\begin{prop}\label{prop:semienough-flabby-modules}
-Assuming the law of excluded middle in the metatheory, any~$\neg\neg$-separated
-module in the \effective topos embeds into a flabby module.
-\end{prop}
-
-% Δ(X) ist welk, falls X bewohnt.
-% Δ(X) ist separiert.
-% Ist A separierter Modul, so lässt sich A in einen welken Modul einbetten
-% (der seinerseits separiert ist, er ist nämlich Δ(Γ(A))).
-% Sind welke Mengen separiert? Nein: Omega.
-% Sind welke Moduln separiert? Nein: Ich denke mittlerweile, dass sich jeder
-% Modul in einen welken einbetten lässt, über etwas, was wie eine induktive
-% Konstruktion aussieht.
-
-% In Eff gilt: Jede negneg-separierte abelsche Gruppe bettet in eine injektive
+% XXX In Eff gilt: Jede negneg-separierte abelsche Gruppe bettet in eine injektive
 % ein. Nämlich A --> ΔΓA --> ΔI für ΓA --> I Einbettung in eine injektive
@@
 
+% XXX
 The intuitive
@@
 
-This problem does not manifest with objects~$X$ which are not \effective sets.
+This problem does not manifest with objects~$X$ which are not \effective
+objects such as double-negation sheaves.
 Reflexivity realizers for these objects are themselves not very informative;
 a procedure witnessing flabbiness therefore only has to turn one kind of
-non-informative realizers into another kind. The flabby modules featuring in
-the proof of Proposition~\ref{prop:semienough-flabby-modules} will accordingly
-not be \effective sets.
+non-informative realizers into another kind.
 
@@
 
-\begin{proof}[Proof of Proposition~\ref{prop:semienough-flabby-modules}]
-Let~$(\Gamma \dashv \Delta) : \Set \to \Eff$ be the inclusion of the
-double-negation sheaves. For a~$\neg\neg$-separated module~$M$ in the \effective
-topos, the canonical morphism~$M \to \Delta(\Gamma(M))$ is a monomorphism; the
-set~$\Gamma(M)$ is flabby by virtue of being inhabited; and~$\Delta$ preserves
-flabby objects by Proposition~\ref{prop:pushforward-of-flabby-objects}.
-\end{proof}
-
-\begin{rem}The analogues of Proposition~\ref{prop:flabby-effective-sets} and
-Proposition~\ref{prop:semienough-flabby-modules} for the realizability topos
+\begin{rem}The analogue of Proposition~\ref{prop:flabby-effective-sets}
+for the realizability topos
 constructed using infinite time Turing
-machines~\cite{bauer:injection,hamkins-lewis:ittm} are true as well, with the
-same proofs.\end{rem}
+machines~\cite{bauer:injection,hamkins-lewis:ittm} is true as well, with the
+same proof.\end{rem}
+% XXX any PCA?
 
@@
 efficient-in-practice algorithms~\cite{barakat-lh:homalg,barakat-lh:ext},
-but it appears that there is not a general framework for sheaf
+but there is no established general framework for sheaf
 cohomology which would work in an intuitionistic metatheory.
@@
 constructively is its reliance on injective resolutions. It is known that in
-the absence of the axiom of choice, much less in a purely intuitionistic
+the absence of Zorn's lemma, much less in a purely intuitionistic
 context, there might not be any nontrivial injective abelian
@@
 
-In principle, this problem could be remedied by employing flabby resolutions
-instead of injective ones. There are, however, two problems with this
+Can this issue be remedied by employing flabby resolutions
+instead of injective ones? Classically, it is known that they can be used in
+their stead. There are, however, two problems with this
 suggestion. Firstly, all proofs known to us that flabby sheaves are
@@
 be mitigated by relying on the substitute property discussed
-following Scholium~\ref{scholium:exact-as-presheaves}.
-But secondly, it is an open question whether
-one can show, purely intuitionistically, that any sheaf of modules embeds into
-a flabby sheaf of modules. The following is known about this problem:
+following Scholium~\ref{scholium:exact-as-presheaves}. XXX
 
 \begin{enumerate}
-\item There is a purely intuitionistic proof that any sheaf of sets embeds into
-a flabby sheaf of sets
-(Scholium~\ref{scholium:properties-of-flabby-objects}(4)).
-
-\item The existence of enough flabby modules, and even the existence of enough
-injective modules, is \emph{not} a constructive taboo, that is, these statements do not
-entail a classical principle like the law of excluded middle or the principle
-of omniscience. This is because assuming the axiom of choice, any
-Grothendieck topos has enough injective (and therefore flabby) modules.
+\item XXX Reformulate as remark somewhere
 
@@
 \end{document}
+
+XXX cite more references for Eff, including
+Hyland §5
+Fourman-Scott 1979
+Johnstone 1977
+toposmoeder
+
+XXX quasicompact case
+
+XXX embed into flabby module

### 2022-01-09 paper-flabby-objects.tex
"Every" instead of "any" (thanks to the anonymous referee)

@@
 \usepackage{xspace}
+\usepackage{color}
 
@@
 \newcommand{\xra}[1]{\xrightarrow{#1}}
-\newcommand{\XXX}[1]{\textbf{XXX: #1}}
+\newcommand{\XXX}[1]{\textbf{\textcolor{red}{XXX: #1}}}
 \newcommand{\aaa}{\mathfrak{a}}
@@
 \noindent
-As is nowadays well-established, any topos supports an \emph{internal language}
+As is nowadays well-established, every topos supports an \emph{internal language}
 which can be used to reason about the objects and morphisms of the topos in a
@@
 \cite[Chapter~VI]{moerdijk-maclane:sheaves-logic}). The internal language is
-sound with respect to intuitionistic reasoning, whereby any intuitionistic
-theorem holds in any topos.
+sound with respect to intuitionistic reasoning, whereby every intuitionistic
+theorem holds in every topos.
 
@@
 internalization}. For instance, by interpreting the
-proposition \begin{quote}``in any short exact sequence of modules, if the two
+proposition \begin{quote}``in every short exact sequence of modules, if the two
 outer ones are finitely generated then so is the middle one''\end{quote} of
 intuitionistic commutative algebra internally to the topos of sheaves over a
-space~$X$, we obtain the geometric analogue \begin{quote}``in any short exact
+space~$X$, we obtain the geometric analogue \begin{quote}``in every short exact
 sequence of sheaves of modules over~$X$, if the two outer ones are of finite
@@
 \end{enumerate}
-\item[(4a)] Any sheaf embeds into a flabby sheaf.
-\item[(4b)] Any sheaf of modules embeds into a flabby sheaf of modules.
+\item[(4a)] Every sheaf embeds canonically into a flabby sheaf.
+\item[(4b)] Every sheaf of modules embeds canonically into a flabby sheaf of modules.
 \end{enumerate}
@@
 be extended to a section on~$U \cup U_i$.\end{defn}
-% XXX choice-free formulation
+\XXX{choice-free formulation}
 
@@
 adopt the terminology of~\cite{kock:partial-maps} regarding subterminals and
-subsingletons: A subset~$K \subseteq X$ is \emph{subterminal} if and only if any given
+subsingletons: A subset~$K \subseteq X$ is \emph{subterminal} if and only if every given
 elements are equal ($\forall x,y \in K\_ x = y$), and it is a
 \emph{subsingleton} if and only if there is an element~$x \in X$ such that~$K
-\subseteq \{ x \}$. Any subsingleton is trivially subterminal, but the converse
+\subseteq \{ x \}$. Every subsingleton is trivially subterminal, but the converse
 might fail.
 
-\begin{defn}A set~$X$ is \emph{flabby} if and only if any subterminal subset
-of~$X$ is a subsingleton, that is, if and only if for any subset~$K \subseteq
+\begin{defn}A set~$X$ is \emph{flabby} if and only if every subterminal subset
+of~$X$ is a subsingleton, that is, if and only if for every subset~$K \subseteq
 X$ such that~$\forall x,y \in K\_ x = y$, there exists an element~$x \in X$
@@
 \begin{prop}\label{prop:taboo}
-If any inhabited set is flabby, then the law of excluded middle
+If every inhabited set is flabby, then the law of excluded middle
 holds.
@@
 the maximally undefined element and a singleton is a maximally defined element.
-A set is flabby if and only if any of its partially-defined elements can be
+A set is flabby if and only if every of its partially-defined elements can be
 refined to an honest element.
@@
 \begin{rem}\label{rem:constant-flabby}
-Although we will see in XXX and in Section~\ref{sect:in-eff} that there is some
+Although we will see in \XXX{number} and in Section~\ref{sect:in-eff} that there is some
 relation between flabby sets and~$\neg\neg$-separated sets, neither notion
@@
 \begin{defn}\begin{enumerate}
-\item A set~$I$ is \emph{injective} if and only if, for any injection~$i
-: A \to B$, any map~$f : A \to I$ can be extended to a map~$B \to I$.
-\item An~$R$-module~$I$ is \emph{injective} if and only if, for any linear
-injection~$i : A \to B$ between~$R$-modules, any linear map~$f : A \to I$ can
+\item A set~$I$ is \emph{injective} if and only if, for every injection~$i
+: A \to B$, every map~$f : A \to I$ can be extended to a map~$B \to I$.
+\item An~$R$-module~$I$ is \emph{injective} if and only if, for every linear
+injection~$i : A \to B$ between~$R$-modules, every linear map~$f : A \to I$ can
 be extended to a linear map~$B \to I$, as in the diagram below.
@@
 % Take care if numbering changes (referenced below)
-\item Any set embeds canonically (in a uniform fashion) into an injective set.
-\item Any injective module is also injective as a set.
+\item Every set embeds canonically (that is, in a uniform fashion) into an injective set.
+\item Every injective module is also injective as a set.
 \item Assuming Zorn's lemma (but not the law of excluded middle, hence not the
-axiom of choice), any module embeds into an
+axiom of choice), every module embeds into an
 injective module.
@@
 \begin{prop}\label{prop:injective-flabby}
-Any injective set is flabby.\end{prop}
+Every injective set is flabby.\end{prop}
 
@@
 \begin{cor}\label{cor:enough-flabby-sets}
-Any set canonically embeds into a flabby set.\end{cor}
+Every set canonically embeds into a flabby set.\end{cor}
 
@@
 A further corollary of Proposition~\ref{prop:injective-flabby} is that the
-statement ``any inhabited set is injective'' is a constructive taboo: If any
-inhabited set is injective, then any inhabited set is flabby, thus the law of
+statement ``every inhabited set is injective'' is a constructive taboo: If every
+inhabited set is injective, then every inhabited set is flabby, thus the law of
 excluded middle follows by Proposition~\ref{prop:taboo}.
 
-\begin{prop}Any singleton set is flabby. The binary cartesian product of flabby sets
+\begin{prop}Every singleton set is flabby. The binary cartesian product of flabby sets
 is flabby.\end{prop}
@@
 
-Subsets of flabby sets are in general not flabby, as else any set would be
+Subsets of flabby sets are in general not flabby, as else every set would be
 flabby in view of Corollary~\ref{cor:enough-flabby-sets}.
 
-% XXX enough flabby modules
+\XXX{enough flabby modules}
 
@@
 This definition amounts to the following: An object~$X$ of an elementary
-topos~$\E$ is flabby if and only if, for any monomorphism~$K \to A$ and any
+topos~$\E$ is flabby if and only if, for every monomorphism~$K \to A$ and every
 morphism~$K \to X$, there exists an epimorphism~$B \to A$ and a morphism~$B
@@
 
-\begin{proof}This holds for any property which can be defined
+\begin{proof}This holds for every property which can be defined
 in the stack semantics~\cite[Lemma~7.3]{shulman:stack-semantics}.
@@
 \to X$ is given by the sections~$s_i|_{(V \cup V_i) \cap U}$.
-% XXX choicefree
+\XXX{choicefree}
 
@@
 \begin{defn}An object~$I$ of an elementary topos~$\E$ is \emph{externally
-injective} if and only if for any monomorphism~$A \to B$ in~$\E$, the canonical
+injective} if and only if for every monomorphism~$A \to B$ in~$\E$, the canonical
 map~$\Hom_\E(B,I) \to \Hom_\E(A,I)$ is surjective. It is \emph{internally
-injective} if and only if for any monomorphism~$A \to B$ in~$\E$, the canonical
+injective} if and only if for every monomorphism~$A \to B$ in~$\E$, the canonical
 morphism~$[B,I] \to [A,I]$ between Hom objects is an epimorphism in~$\E$.
@@
 \item[(1)] $I$ is internally injective.
-\item[(1')] For any morphism $p : A \to 1$ in $\E$, the object $p^*(I)$ has property~(1)
+\item[(1')] For every morphism $p : A \to 1$ in $\E$, the object $p^*(I)$ has property~(1)
 as an object of $\E/A$.
 \item[(2)] The functor~$[\cdot, I] : \E^\op \to \E$ maps monomorphisms in $\E$
-to morphisms for which any global element of the target locally (after change of
+to morphisms for which every global element of the target locally (after change of
 base along an epimorphism) possesses a preimage.
-\item[(2')] For any morphism $p : A \to 1$ in $\E$, the object $p^*(I)$ has property~(2)
+\item[(2')] For every morphism $p : A \to 1$ in $\E$, the object $p^*(I)$ has property~(2)
 as an object of $\E/A$.
@@
 \begin{scholium}\label{scholium:properties-of-flabby-objects}
-For any elementary topos~$\E$:
+For every elementary topos~$\E$:
 \begin{enumerate}
-\item Any object embeds canonically into an internally injective object.
+\item Every object embeds canonically into an internally injective object.
 \item (If~$\E$ has a natural numbers object.) The underlying unstructured
 object of an internally injective module is internally injective.
-\item Any internally injective object is flabby.
-\item Any object embeds canonically into a flabby object.
-\item Any internal module embeds canonically into an internal module which is flabby.
-\item The terminal object is flabby. The product of flabby objects is flabby.
+\item Every internally injective object is flabby.
+\item Every object embeds canonically into a flabby object.
+\item Every internal module embeds canonically into an internal module which is flabby.
+\item The terminal object is flabby. The binary product of flabby objects is flabby.
 \item Let~$I$ be an internally injective object. Let~$T$ be an arbitrary
@@
 
-% XXX If Zorn's lemma is not available but E is localic, ... is at least true
-% if E is compact.
+\XXX{If Zorn's lemma is not available but E is localic, ... is at least true
+if E is compact.}
 
@@
 elementary toposes is the following. An object~$X$ of an elementary topos~$\E$
-is \emph{strongly flabby} if and only if, for any monomorphism~$K \to 1$
-in~$\E$, any morphism~$K \to X$ lifts to a morphism~$1 \to X$.
+is \emph{strongly flabby} if and only if, for every monomorphism~$K \to 1$
+in~$\E$, every morphism~$K \to X$ lifts to a morphism~$1 \to X$.
 
@@
 such a way that the category of internal sheaves over this internal locale
-coincides with~$\Sh(Y)$ [XXX give reference]; in particular, a given sheaf~$E$ of~$\O_Y$-modules can
+coincides with~$\Sh(Y)$ [\XXX{give reference}]; in particular, a given sheaf~$E$ of~$\O_Y$-modules can
 be regarded as a sheaf over~$I(Y)$. Under this equivalence, the morphism~$f : Y
@@
 
-% XXX
+\XXX{XXX}
 The intuitive
@@
 same proof.\end{rem}
-% XXX any PCA?
+\XXX{any PCA?}
 
@@
 be mitigated by relying on the substitute property discussed
-following Scholium~\ref{scholium:exact-as-presheaves}. XXX
+following Scholium~\ref{scholium:exact-as-presheaves}. \XXX{XXX}
 
 \begin{enumerate}
-\item XXX Reformulate as remark somewhere
+\item \XXX{Reformulate as remark somewhere}
 
@@
 while Zermelo--Fraenkel without choice does not~\cite[Section~9]{shulman-lumsdaine:hits},
-hence IZF also does not.\footnote{With quotient inductive types, any infinitary
+hence IZF also does not.\footnote{With quotient inductive types, every infinitary
 algebraic theory admits free algebras. However, it is consistent with
@@
 
+\XXX{cite more: Hyland §5,
+Fourman-Scott 1979,
+Johnstone 1977,
+toposmoeder}
+
+\XXX{quasicompact case}
+
 \end{document}
-
-XXX cite more references for Eff, including
-Hyland §5
-Fourman-Scott 1979
-Johnstone 1977
-toposmoeder
-
-XXX quasicompact case
-
-XXX embed into flabby module

### 2022-01-09 paper-flabby-objects.tex
Better anti-flabbiness explanation (thanks to the anonymous referee)

@@
 can be obtained by picking any topological space~$T$ such that~$H^1(T,
-\underline{\ZZ}) \neq 0$. Then the constant sheaf~$\underline{\ZZ}$ is not
+\underline{\ZZ}) \neq 0$, where~$\underline{\ZZ}$ is the constant sheaf with
+stalks~$\ZZ$. For instance, the space~$T$ could be connected (such that every
+global section is constant) while having an open which is the disjoint union of
+two connected components. Then the sheaf~$\underline{\ZZ}$ is not
 flabby and hence, by Proposition~\ref{prop:flabby-sheaves-objects} below, not a

### 2022-01-09 paper-flabby-objects.tex
Functionally flabby sets (WIP)

@@
 \newcommand{\defeq}{\vcentcolon=}
+\newcommand{\defeqv}{\vcentcolon\equiv}
 \newcommand{\op}{\mathrm{op}}
@@
 
+\XXX{proper exposition}
+
+\begin{defn}\label{defn:functionally-flabby}
+A set~$X$ is \emph{functionally flabby} if and only if there is a
+map~$\varepsilon : \P_{\leq1}(X) \to X$ such that for all~$K \in
+\P_{\leq1}(X)$, $K \subseteq \{ \varepsilon(K) \}$.\end{defn}
+
+A map~$\varepsilon : \P_{\leq1}(X) \to X$ satisfies the condition in
+Definition~\ref{defn:functionally-flabby} if and only if it is a retraction of
+the canonical injection~$X \to \P_{\leq1}(X)$.
+
+\begin{prop}For a set~$X$, the following conditions are equivalent:
+\begin{enumerate}
+\item $X$ is injective.
+\item $X$ is functionally flabby.
+\end{enumerate}
+\end{prop}
+
+\begin{proof}Let~$X$ be an injective set. Then the identity~$X \to X$ can be extended
+along the canonical inclusion~$X \to \P_{\leq1}(X)$ to a map~$\varepsilon :
+\P_{\leq1}(X) \to X$. This map witnesses that~$X$ is functionally flabby.
+
+Conversely, let~$X$ be a functionally flabby set with witness~$\varepsilon :
+\P_{\leq1}(X) \to X$. Let~$i : A \to B$ be an injection and let~$f : A \to X$
+be a map. Then~$x \mapsto \varepsilon(f[i^{-1}[\{x\}])$ is an extension of~$f$
+along~$i$.
+\end{proof}
+
+\begin{prop}Every module embeds canonically into a functionally flabby module.\end{prop}
+
+\begin{proof}Let~$M$ be a module. First we'll assume that~$M$
+is a sheaf for some modal operator~$\nabla$ such that for any
+formula~$\varphi$, $\nabla(\varphi \vee (\varphi \Rightarrow \nabla\bot))$. In
+this case, the module~$M$ is already functionally flabby: A witnessing choice
+function maps a subsingleton~$K \subseteq M$ to the unique element~$x \in M$
+such that~$\nabla(x \in K')$, where~$K' \defeq K \cup \{ 0 \,|\, \text{$K$
+inhabited} \Rightarrow \nabla\bot \}$. Since~$K'$ has the property
+that~$\nabla(\text{$K$ is a singleton})$, such an element~$x$ exists and is unique
+by the sheaf condition.
+
+Now let~$M$ be a module which is separated for some XXX modal operator~$\nabla$.
+Then its sheafification is a functionally flabby module into which~$M$ embeds.
+
+Finally, let~$M$ be an arbitrary module. For any element~$x \in M$,
+let~$\nabla_x$ be the modal operator with~$\nabla_x\varphi \defeqv ((\varphi \Rightarrow x
+= 0) \Rightarrow x = 0)$. The module~$M$ might not be~$\nabla_x$-separated for
+a particular element~$x$, but jointly XXX
+\end{proof}
+
 \begin{prop}\label{prop:hom-flabby}

### 2022-02-13 paper-flabby-objects.tex
Remark on a potential variant of the definition

@@
 the axiom of choice and Zorn's lemma. The former implies the latter, but the
-converse requires the law of excluded middle.}
+converse implication requires the law of excluded middle.}
 The standard proof of property~(4b) requires the
@@
 open covering~$X = \bigcup_i U_i$ such that, for all~$i$, the section~$s$ can
-be extended to a section on~$U \cup U_i$.\end{defn}
-\XXX{choice-free formulation}
+be extended to a section on~$U \cup U_i$.\footnote{Even in the absence of the
+axiom of choice it doesn't make a difference
+whether we are stipulating, as in the definition, for each index~$i \in I$ the mere
+existence of an extension~$t \in F(U \cup U_i)$, or whether we are stipulating
+the existence of a family~$(t_i)_i$ of extensions~$t_i \in F(U \cup U_i)$.
+Clearly, the existence of the family implies the existence of individual
+extensions. Conversely, if for each~$i \in I$ there exists an extension, there
+is a tautologous family of extensions over the enlarged index set~$I' \defeq \{
+(i,t) \,|\, i \in I, t \in F(U \cup U_i), t|_U = s \}$, and we still have~$X =
+\bigcup_{(i,t) \in I'} U_i$.}
+\end{defn}
 

### 2022-02-13 paper-flabby-objects.tex
More references

@@
 such a way that the category of internal sheaves over this internal locale
-coincides with~$\Sh(Y)$ [\XXX{give reference}]; in particular, a given sheaf~$E$ of~$\O_Y$-modules can
+coincides with~$\Sh(Y)$~\cite[Scholium~C1.6.4]{johnstone:elephant}; in particular, a given sheaf~$E$ of~$\O_Y$-modules can
 be regarded as a sheaf over~$I(Y)$. Under this equivalence, the morphism~$f : Y
@@
 topos. We follow the terminology of Martin Hyland's survey on the \effective
-topos~\cite{hyland:effective-topos}.
+topos~\cite{hyland:effective-topos} and refer to
+\cite{oosten:realizability,moerdijk-oosten:topos-theory,phoa:effective,bauer:c2c} for more background.
 
@@
 Fourman-Scott 1979,
-Johnstone 1977,
-toposmoeder}
+Johnstone 1977}
 

### 2022-02-13 paper-reflection.tex
typo

@@
 
-\begin{rem}Most published renderings of~$\RRS$ and~$\MDC$ do not refer to an initial
+\begin{rem}Most published renditions of~$\RRS$ and~$\MDC$ do not refer to an initial
 element~$x_0 \in X$, but to a set~$A \subseteq X$ and then require instead

### 2022-02-13 paper-flabby-objects.tex
Reference to Kempf
to do
Explanation of "over-locale"

@@
 
-Let~$X$ be a locale and let~$f : Y \to X$ be an over-locale. By the fundamental
+Let~$X$ be a locale and let~$Y$ be a locale over~$X$, that is, a morphism~$f : Y \to X$ of locales. By the fundamental
 relation between locales and topological spaces, this situation arises
@@
 of~$\O_Y$-modules is to pick an injective resolution~$0 \to E \to I^\bullet$
-and set~$R^n f_*(E) \defeq H^n(f_*(I^\bullet))$.
+and set~$R^n f_*(E) \defeq H^n(f_*(I^\bullet))$~\cite[Tag~01DZ]{stacks-project}.
 
@@
 characterized in the internal language, we can give a precise
-rendering of this slogan: We can understand higher direct images as
-internal sheaf cohomology.
+rendering of this slogan: \emph{Higher direct images are
+internal sheaf cohomology.}
 
@@
 instead of injective ones? Classically, it is known that they can be used in
-their stead. There are, however, two problems with this
+their stead. For instance, George Kempf developed the foundations of the
+cohomology of quasicoherent sheaves on this premise~\cite{kempf:cohomology}. There are, however, two problems with this
 suggestion. Firstly, all proofs known to us that flabby sheaves are
@@
 \end{document}
+
+XXX: injective as module implies functionally flabby
+
+FUTURE:
+
+* Does "flabby sheaves are acyclic on all locales" imply Zorn?

### 2022-02-14 paper-flabby-objects.tex
Rework of section on the effective topos

@@
 
-\begin{prop}Every module embeds canonically into a functionally flabby module.\end{prop}
+\begin{thm}\label{thm:enough-flabby-modules}
+Every module embeds canonically into a functionally flabby module.\end{thm}
 
@@
 
-\XXX{XXX}
 The intuitive
@@
 following. Let~$X$ be a flabby object in the \effective topos. Then there is a
-procedure which computes for any subterminal~$K \subseteq X$ an element~$x_K$
-such that~$K \subseteq \{ x_K \}$. This element might not depend extensionally
-on~$K$, but this fine point is not important for the present discussion. Let~$f : X \to X$
-be a morphism. We construct the self-referential subset~$K
-\defeq \{ f(x_K) \}$; the formal proof below will indicate how this can be
-done. Then~$K \subseteq \{ x_K \}$, so~$f(x_K) = x_K$.
-
-A corollary of Proposition~\ref{prop:flabby-effective-sets} is that the trivial
-module is the only flabby module in the \effective topos whose underlying
-unstructured object is an \effective set: Given such a flabby module~$M$, let~$v
-\in M$ be an arbitrary element. Then the morphism~$x \mapsto v + x$ has a fixed
-point; thus~$v + x = x$ for some element~$x$, and hence~$v = 0$.
-
-It is the self-referentiality which makes the proof of
-Proposition~\ref{prop:flabby-effective-sets} work, but the blame for the paucity
-of flabby objects in the \effective topos is to put on the realizers for
-statements of the form~``$K = K$'', where~$({=})$ is the nonstandard equality
-predicate of the powerobject~$\P(X)$. A procedure witnessing flabbiness has to
-compute a reflexivity realizer for a suitable element~$x_K$ from a reflexivity
-realizer for a given element~$K$. However, such realizers are not very
-informative. Metaphorically speaking, a procedure witnessing flabbiness has to
+procedure which computes for any (realizer of a) subterminal~$K \subseteq X$ a
+(realizer of an) element~$x_K$
+such that~$K \subseteq \{ x_K \}$. However, realizers for subsets are not very
+informative; the procedure cannot ask questions such as ``is a
+given element of~$X$ contained in~$K$?'' nor query its input in any way.
+Hence~$x_K$ will actually be the same element for any subterminal~$K$.
+Metaphorically speaking, a procedure witnessing flabbiness has to
 conjure elements out of thin air.
 
-This problem does not manifest with objects~$X$ which are not \effective
-objects such as double-negation sheaves.
-Reflexivity realizers for these objects are themselves not very informative;
-a procedure witnessing flabbiness therefore only has to turn one kind of
-non-informative realizers into another kind.
+This issue does not manifest with objects~$X$ which are not \effective objects
+such as double-negation sheaves. Realizers for elements of such objects are
+themselves not very informative; for those, a procedure witnessing flabbiness
+only has to turn one kind of non-informative realizers into another kind. This
+is why, in line with Theorem~\ref{thm:enough-flabby-modules}, it is still true
+in the \effective topos that any module of the \effective topos embeds into a
+flabby module.
 
 \begin{proof}[Proof of Proposition~\ref{prop:flabby-effective-sets}]
-For any Turing machine~$e$, let~$v_e : |X| \to \Sigma$ be the nonstandard
-predicate given by
-\begin{multline*}
-  v_e(x) = \{ m \in \NN \,|\,
-  \text{there is an element~$x_0 \in |X|$ such that} \\
-  \text{$e$ terminates with an element of~$\llbracket x_0 = x_0 \rrbracket$ and
-  $m \in \llbracket x = x_0 \rrbracket$} \}
-\end{multline*}
-and let~$K_e \in \Sigma^{|X| \times \Sigma}$ be the nonstandard predicate given by
-\[ K_e(x,u) = \llbracket (x = x) \wedge (u \leftrightarrow v_e(x)) \rrbracket. \]
-One can explicitly construct a realizer~$a_e$ of the statement~``$K_e = K_e$'',
-where~$({=})$ is the nonstandard equality predicate of the object~$\P_{\leq1}(X)$
-of subterminals of~$X$. This is where the assumption that~$X$ is \effective is
-important; without it, we could only verify~``$K_e = K_e$'' where~$({=})$ is
-the nonstandard equality predicate of the full powerobject~$\P(X)$.
-
-Since~$X$ is flabby, there is a realizer~$r$ for the statement~``$\forall K \in
-\P_{\leq1}(X)\_ \exists x \in X\_ \forall y \in X\_ (y \in K \Rightarrow y =
-x)$''. Let~$s$ be a realizer for the statement~``$\forall x \in X\_ \exists y
-\in Y\_ y = f(x)$''. Let~$e$ be the particular Turing machine which proceeds as
-follows:
-\begin{enumerate}
-\item[1.] Simulate~$r$ on input~$a_e$ in order to obtain a realizer~$b \in
-\llbracket x = x \rrbracket$ for some~$x \in |X|$.
-\item[2.] Simulate~$s$ on input~$b$ in order to obtain a realizer~$c \in \llbracket
-f(x) = f(x) \rrbracket$.
-\item[3.] Output~$c$.
-\end{enumerate}
-This description of the machine~$e$ makes use of the number~$e$ coding it;
-the recursion theorem yields a general reason why this self-referentiality is
-possible. Here we can even do without this theorem, since a close inspection of
-the construction of~$a_e$ shows that~$a_e$ is actually independent of~$e$. This
-should not come as a surprise, as reflexivity realizers of~$\P(X)$
-and~$\P_{\leq1}(X)$ are known to be not very informative.
-
-Passing~$a_e$ to~$r$ yields a reflexivity realizer of some element~$x_{K_e} \in
-|X|$. Therefore the Turing machine~$e$ does terminate, with a reflexivity
-realizer for~$f(x_{K_e})$. Thus the statement~``$f(x_{K_e}) \in K_e$'' is
-realized; hence~``$f(x_{K_e}) = x_{K_e}$'' is as well.
-\end{proof}
-
-\begin{rem}The analogue of Proposition~\ref{prop:flabby-effective-sets}
-for the realizability topos
+Let~$X$ be an object of the \effective topos. Then the object~$P_{\leq1}(X)$ of
+subterminals of~$X$ is a uniform object in the sense
+of~\cite[Section~3.4]{moerdijk-oosten:topos-theory}, being a retract of the
+uniform object~$P(X)$ by the surjection
+\[ P(X) \longrightarrow P_{\leq1}(X),\
+  M \longmapsto \{ x \in X \,|\, M = \{ x \} \}. \]
+Hence if~$X$ is \effective, the uniformity principle~\cite[Proposition~15.1]{hyland:effective-topos}
+\[ \forall K \in P_{\leq1}(X)\_
+  \exists x \in X\_ K \subseteq \{x\} \quad\Longrightarrow\quad
+  \exists x_0 \in X\_ \forall K \in P_{\leq1}(X)\_ K \subseteq \{x_0\} \]
+applies. Thus, if~$X$ is \effective and flabby, there is
+an element~$x_0 \in X$ such that~$K \subseteq \{x_0\}$ for any subterminal~$K$
+of~$X$. The conclusion follows by considering, for any elements~$a,b \in X$,
+the subterminals~$\{a\}$ and~$\{b\}$.
+\end{proof}
+
+\begin{rem}The analogue of Proposition~\ref{prop:flabby-effective-sets} is true
+as well for the realizability topos
 constructed using infinite time Turing
-machines~\cite{bauer:injection,hamkins-lewis:ittm} is true as well, with the
-same proof.\end{rem}
-\XXX{any PCA?}
+machines~\cite{bauer:injection,hamkins-lewis:ittm} and indeed for any realizability
+topos, with the same proof, as power objects are always
+uniform~\cite{johnstone:review-oosten}.\end{rem}
 

### 2022-02-14 paper-flabby-objects.tex
typo

@@
 this case, the module~$M$ is already functionally flabby: A witnessing choice
-function maps a subsingleton~$K \subseteq M$ to the unique element~$x \in M$
+function maps a subterminal~$K \subseteq M$ to the unique element~$x \in M$
 such that~$\nabla(x \in K')$, where~$K' \defeq K \cup \{ 0 \,|\, \text{$K$

### 2022-02-14 paper-flabby-objects.tex
to do
typo

@@
 of~\cite[Section~3.4]{moerdijk-oosten:topos-theory}, being a retract of the
-uniform object~$P(X)$ by the surjection
-\[ P(X) \longrightarrow P_{\leq1}(X),\
+uniform object~$\P(X)$ by the surjection
+\[ \P(X) \longrightarrow \P_{\leq1}(X),\
   M \longmapsto \{ x \in X \,|\, M = \{ x \} \}. \]
 Hence if~$X$ is \effective, the uniformity principle~\cite[Proposition~15.1]{hyland:effective-topos}
-\[ \forall K \in P_{\leq1}(X)\_
+\[ \forall K \in \P_{\leq1}(X)\_
   \exists x \in X\_ K \subseteq \{x\} \quad\Longrightarrow\quad
-  \exists x_0 \in X\_ \forall K \in P_{\leq1}(X)\_ K \subseteq \{x_0\} \]
+  \exists x_0 \in X\_ \forall K \in \P_{\leq1}(X)\_ K \subseteq \{x_0\} \]
 applies. Thus, if~$X$ is \effective and flabby, there is
@@
 \begin{rem}The analogue of Proposition~\ref{prop:flabby-effective-sets} is true
-as well for the realizability topos
-constructed using infinite time Turing
-machines~\cite{bauer:injection,hamkins-lewis:ittm} and indeed for any realizability
-topos, with the same proof, as power objects are always
+as well for the realizability topos constructed using infinite time Turing
+machines~\cite{bauer:injection,hamkins-lewis:ittm} and indeed for any
+realizability topos, with the same proof, as power objects are always
 uniform~\cite{johnstone:review-oosten}.\end{rem}
@@
 
+Cite! https://arxiv.org/pdf/2008.10677.pdf
+
 FUTURE:

### 2022-02-14 paper-flabby-objects.tex
Full presentation of the existence of enough flabby modules

@@
 
-\begin{defn}A set~$X$ is \emph{flabby} if and only if every subterminal subset
+\begin{defn}\label{defn:flabby-set}A set~$X$ is \emph{flabby} if and only if every subterminal subset
 of~$X$ is a subsingleton, that is, if and only if for every subset~$K \subseteq
@@
 
-\XXX{enough flabby modules}
-
-\XXX{proper exposition}
+\subsection{On the existence of enough flabby modules}
+
+For the intended application to the theory of sheaf cohomology, the existence
+of \emph{enough flabby sheaves of modules} is crucial: Any sheaf of modules
+embeds into a flabby sheaf of modules. In this section, we study the existence
+of enough flabby modules from the non-sheaf theoretic but constructive point of
+view.
+
+We have already established that every set embeds into a flabby set. However,
+the supersets suggested by the proof of Corollary~\ref{cor:enough-flabby-sets}
+do not carry a module structure even when the base set does. This deficit
+raises the following question: Given a module~$M$, does there exist a linear
+injection~$M \to M'$ into a set~$M'$ which simultaneously carries a module
+structure and is flabby?
+
+An earlier version of this text formulated this question as an open problem.
+The positive solution presented below rests on the intermediate notion of
+\emph{functionally flabby} sets and makes essential use of sheaves for modal
+operators. A short survey for the latter is contained
+in~\cite[Section~6]{blechschmidt:phd}; other references
+include~\cite[Sections~14.4f.]{goldblatt:topoi},~\cite{goldblatt:modality}
+\cite{vries:sheafification} and~\cite{fourman-scott:sheaves-and-logic}.\footnote{On page~5 of the
+preprint~\cite{vries:sheafification} there is a slight typing error:
+Fact~2.1(i) gives the characterization of~$j$-closedness, not~$j$-denseness.
+The correct characterization of~$j$-denseness in that context is~$\forall b \in
+B\_ j(b \in A)$.}
 
@@
 
+Trivially, every functionally flabby set is flabby in the sense of
+Definition~\ref{defn:flabby-set}.
 A map~$\varepsilon : \P_{\leq1}(X) \to X$ satisfies the condition in
@@
 
-\begin{proof}Let~$M$ be a module. First we'll assume that~$M$
-is a sheaf for some modal operator~$\nabla$ such that for any
+\begin{proof}We explain how to construct the required flabby envelopes
+in three steps.
+First, let~$M$ be a module which is a sheaf for some modal operator~$\nabla$
+with the property that for any
 formula~$\varphi$, $\nabla(\varphi \vee (\varphi \Rightarrow \nabla\bot))$. In
@@
 such that~$\nabla(x \in K')$, where~$K' \defeq K \cup \{ 0 \,|\, \text{$K$
-inhabited} \Rightarrow \nabla\bot \}$. Since~$K'$ has the property
-that~$\nabla(\text{$K$ is a singleton})$, such an element~$x$ exists and is unique
+inhabited} \Rightarrow \nabla\bot \}$. Since the set~$K'$ has, unlike~$K$, the property
+that~$\nabla(\text{$K'$ is a singleton})$, such an element~$x$ exists and is unique
 by the sheaf condition.
 
-Now let~$M$ be a module which is separated for some XXX modal operator~$\nabla$.
-Then its sheafification is a functionally flabby module into which~$M$ embeds.
+Second, let~$M$ be a module which is separated for some modal operator~$\nabla$
+as above. Then its sheafification is a functionally flabby module into
+which~$M$ embeds since the canonical map from~$M$ to its sheafification is
+injective by separatedness of~$M$.
 
@@
 let~$\nabla_x$ be the modal operator with~$\nabla_x\varphi \defeqv ((\varphi \Rightarrow x
-= 0) \Rightarrow x = 0)$. The module~$M$ might not be~$\nabla_x$-separated for
-a particular element~$x$, but jointly XXX
-\end{proof}
+= 0) \Rightarrow x = 0)$. This is a modal operator of the kind as above.
+The module~$M$ might not be~$\nabla_x$-separated for
+any particular element~$x$, but it is jointly so: For all elements~$a,b\in M$,
+if~$\nabla_x(a = b)$ for all~$x \in M$, then~$a = b$ (considering~$x \defeq
+a-b$). Hence the canonical map~$M \to \prod_{x \in M} M^{+_x}$ into the product
+of the plus constructions with respect to all the modal operators~$\nabla_x$ is
+injective, and the desired injection is the composition
+\[ M \longrightarrow \prod_{x \in M} M^{+_x} \longrightarrow \prod_{x \in M}
+M^{+_x+_x} \]
+into the product of the sheafifications.
+\end{proof}
+
+\begin{rem}For the purpose of verifying that any module embeds into a flabby
+module, it is essential that in the first step of the proof of
+Theorem~\ref{thm:enough-flabby-modules} the module~$M$ is not only shown to be flabby,
+but even functionally flabby, and moreover in an explicit manner, explicitly
+presenting a witnessing map~$\P_{\leq1}(M) \to M$. This is because while an
+arbitrary product of flabby sets can fail to be flabby, an arbitrary product of
+functionally flabby sets with given witnesses is again so.
+\end{rem}
+
+Given the somewhat nontrivial nature of the construction in the proof of
+Theorem~\ref{thm:enough-flabby-modules}, a natural question is whether simpler
+constructions exist as well.
+There are a number of simple constructions which come close to providing
+flabby envelopes for arbitrary modules, but all such constructions known to the
+author fail in some manner. For instance, given a module~$M$, we could equip the set~$T
+\defeq \P_{\leq1}(M)/{\sim}$, where~$K \sim L$ if and only if~$K = L$ or~$K
+\cup L \subseteq \{0\}$, with a module structure given by~$0 \defeq [\{0\}]$,
+$[K]+[L] \defeq [K+L]$ and~$r [K] \defeq rK$. The resulting module admits a
+linear injection from~$M$, sending an element~$x$ to~$[\{x\}]$. However, it
+fails to be flabby. Given a subterminal~$E \subseteq \P_{\leq1}(M)/{\sim}$,
+there is the well-defined element~$v \defeq [\{x \in M\,|\,\text{$x \in K$ for
+some~$[K] \in E$}\}]$, but we cannot verify~$E \subseteq \{v\}$.
+
+That said, there is an alternative construction if sufficiently general
+\emph{quotient inductive types}, as suggested by Altenkirch and
+Kaposi~\cite{altenkirch-kaposi:qits}, are available. These generalize ordinary
+inductive~$W$-types, which exist in any
+topos~\cite{moerdijk-palmgren:wellfounded-trees,berg-moerdijk:w-types-in-sheaves,berg-kouwenhoven-gentil:w-types-in-eff}
+and whose existence
+can indeed be verified in an intuitionistic set theory like~\textsc{izf}, by allowing to
+give constructors and state identifications at the same time. More
+specifically, given an~$R$-module~$M$, we can construct a flabby envelope~$T$
+of~$M$ as the quotient inductive type generated by the following clauses,
+starting out as the construction of the free module over the underlying set
+of~$M$:~$0
+\in T$ (where~$0$ is a formal symbol); if~$t,s \in T$, then~$t + s \in T$;
+if~$t \in T$ and~$r \in R$, then~$rt \in T$; if~$x \in M$, then~$\underline{x}
+\in T$; if~$K : I \to T$ is a family of elements of~$T$ indexed by a subterminal, then~$\varepsilon_K \in T$;
+if~$t,s,u \in T$ and~$r,r' \in R$, then~$t + (s + u) = (t + s) + u$, $t + s = s + t$, $t + 0 = t
+= t + 0$, $0t = 0$, $1t = t$, $r(t+u) = rt + ru$, $(r+r')t = rt + r't$; if~$x,y
+\in M$ and~$r \in R$, then~$\underline{0} = 0$, $\underline{x + y} =
+\underline{x} + \underline{y}$, $\underline{rx} = r \underline{x}$; and if~$K :
+I \to T$ is a family such that~$I$ is inhabited by some element~$i_0$,
+then~$\varepsilon_{K} = K(i_0)$.
+
+However, there are two issues with this approach.
+Firstly, it is an open question under which circumstances quotient inductive
+types can be shown to exist. Zermelo--Fraenkel with choice certainly suffices,
+while Zermelo--Fraenkel without choice does not~\cite[Section~9]{shulman-lumsdaine:hits},
+hence~\textsc{izf} also does not.\footnote{With quotient inductive types, every infinitary
+algebraic theory admits free algebras. However, it is consistent with
+Zermelo--Fraenkel set theory that some such theories do not admit free
+algebras~\cite{blass:free-algebras}.} The existence of quotient inductive types
+seems to be, as the existence of enough injective modules, \emph{constructively
+neutral}.
+Secondly, by referencing arbitrary families of elements, the construction
+transcends the given type-theoretic or set-theoretic universe; the resulting
+object~$T$ is not manifestly small even if~$M$ is.
+
+\subsection{Exactness properties}
 
@@
 
-%\item There is a way of embedding any module into a flabby module if one is
-%prepared to ignore set-theoretical difficulties. Namely, let~$M$ be an~$R$-module.
-%Inductively construct a collection~$T$ of terms by the following clauses: $0
-%\in T$ (where~$0$ is a formal symbol);
-%if~$t,s \in T$, then~$t + s \in T$; if~$t \in T$ and~$r \in R$, then~$rt \in
-%T$; if~$x \in M$, then~$\underline{x} \in T$; if~$K \subseteq T$ is a
-%subterminal, then~$\varepsilon_K \in T$. Let~$({\sim})$ be
-%the finest equivalence relation on~$T$ such that~$t + (s + u) \sim (t + s) +
-%u$, $t + s \sim s + t$, $t + 0 \sim t \sim 0 + t$, $0t \sim 0$, $1t \sim t$, $r(t+s) \sim rt
-%+ rs$, $(r+r')t \sim rt + r't$ and such
-%that~$\underline{0} \sim 0$, $\underline{x+y} \sim \underline{x} + \underline{y}$, $\underline{rx} \sim
-%r \underline{x}$, and such that~$\varepsilon_{\{t\}} \sim
-%t$ for all terms~$t$. Then the quotient~$T/{\sim}$ is a flabby~$R$-module
-%into which~$M$ embeds. However, it is not clear that~$T/{\sim}$ is a set.
-
-\item There is a way of embedding any module into a flabby module if
-\emph{quotient inductive types}, as suggested by Altenkirch and
-Kaposi~\cite{altenkirch-kaposi:qits}, are available. These generalize ordinary
-inductive~$W$-types, which exist in any
-topos~\cite{moerdijk-palmgren:wellfounded-trees,berg-moerdijk:w-types-in-sheaves,berg-kouwenhoven-gentil:w-types-in-eff}
-and whose existence
-can indeed be verified in an intuitionistic set theory like IZF, by allowing to
-give constructors and state identifications at the same time. More
-specifically, given an~$R$-module~$M$, we can construct a flabby envelope~$T$
-of~$M$ as the quotient inductive type generated by the following clauses:~$0
-\in T$ (where~$0$ is a formal symbol); if~$t,s \in T$, then~$t + s \in T$;
-if~$t \in T$ and~$r \in R$, then~$rt \in T$; if~$x \in M$, then~$\underline{x}
-\in T$; if~$K \subseteq T$ is a subterminal, then~$\varepsilon_K \in T$;
-if~$t,s,u \in T$ and~$r,r' \in R$, then~$t + (s + u) = (t + s) + u$, $t + s = s + t$, $t + 0 = t
-= t + 0$, $0t = 0$, $1t = t$, $r(t+u) = rt + ru$, $(r+r')t = rt + r't$; if~$x,y
-\in M$ and~$r \in R$, then~$\underline{0} = 0$, $\underline{x + y} =
-\underline{x} + \underline{y}$, $\underline{rx} = r \underline{x}$; and if~$t
-\in T$, then~$\varepsilon_{\{t\}} = t$.
-
-However, it is an open question under which circumstances quotient inductive
-types can be shown to exist. Zermelo--Fraenkel with choice certainly suffices,
-while Zermelo--Fraenkel without choice does not~\cite[Section~9]{shulman-lumsdaine:hits},
-hence IZF also does not.\footnote{With quotient inductive types, every infinitary
-algebraic theory admits free algebras. However, it is consistent with
-Zermelo--Fraenkel set theory that some such theories do not admit free
-algebras~\cite{blass:free-algebras}.} The existence of quotient inductive types
-seems to be, as the existence of enough injective modules, \emph{constructively
-neutral}.
-
-\item There are a number of simple constructions which come close to providing
-flabby envelopes for arbitrary modules~$M$. For instance, we could equip the set~$T
-\defeq \P_{\leq1}(X)/{\sim}$, where~$K \sim L$ if and only if~$K = L$ or~$K
-\cup L \subseteq \{0\}$, with a module structure given by~$0 \defeq [\{0\}]$,
-$[K]+[L] \defeq [K+L]$ and~$r [K] \defeq rK$. The resulting module admits a
-linear injection from~$M$, sending an element~$x$ to~$[\{x\}]$. However, it
-fails to be flabby. Given a subterminal~$E \subseteq \P_{\leq1}(X)/{\sim}$,
-there is the well-defined element~$v \defeq [\{x \in M\,|\,\text{$x \in K$ for
-some~$[K] \in E$}\}]$, but we cannot verify~$E \subseteq \{v\}$.
-
 \item There appears to be some tension regarding the \effective topos:
@@
 * Does "flabby sheaves are acyclic on all locales" imply Zorn?
+
+1. Zorn
+2. effective
+3. predicative, no AU's

### 2022-02-14 paper-flabby-objects.tex
Acknowledge Verona

@@
 guidance, and to Daniel Albert and Caterina Cozzi for their careful readings of
-earlier drafts. Most of the work for this note was carried out at the University of
-Augsburg and the Max Planck Institute for Mathematics in the Sciences in
-Leipzig.
+earlier drafts. The work for this note was carried out at the University of
+Augsburg, the Max Planck Institute for Mathematics in the Sciences in
+Leipzig and the University of Verona.
 

### 2022-02-14 paper-flabby-objects.tex
Reference Crosilla's survey on CST/IZF

@@
 and whose existence
-can indeed be verified in an intuitionistic set theory like~\textsc{izf}, by allowing to
+can indeed be verified in an intuitionistic set theory like~\textsc{izf}~\cite{crosilla:cst-izf}, by allowing to
 give constructors and state identifications at the same time. More

### 2022-02-14 paper-flabby-objects.tex
Rewritten conclusion

@@
 goal. The problem of giving a constructive account
-of sheaf cohomology is still open.
+of sheaf cohomology is still open~\cite{tenorio-mariano:cohomology}.
 
@@
 instead of injective ones? Classically, it is known that they can be used in
-their stead. For instance, George Kempf developed the foundations of the
-cohomology of quasicoherent sheaves on this premise~\cite{kempf:cohomology}. There are, however, two problems with this
-suggestion. Firstly, all proofs known to us that flabby sheaves are
+their stead, and moreover George Kempf developed the foundations of the
+cohomology of quasicoherent sheaves on this premise~\cite{kempf:cohomology}. There are, however, three issues with this
+suggestion.
+
+Firstly, all proofs known to us that flabby sheaves are
 acyclic for the global sections functor require Zorn's lemma. This problem might
 be mitigated by relying on the substitute property discussed
-following Scholium~\ref{scholium:exact-as-presheaves}. \XXX{XXX}
-
-\begin{enumerate}
-\item \XXX{Reformulate as remark somewhere}
-
-\item There appears to be some tension regarding the \effective topos:
-Proposition~\ref{prop:semienough-flabby-modules} shows that at
-least~$\neg\neg$-separated modules in the \effective topos always embed into
-flabby modules, assuming the law of excluded middle in the metatheory, while
-Proposition~\ref{prop:flabby-effective-sets} shows that no nontrivial \effective
-module is flabby.
-\end{enumerate}
+following Scholium~\ref{scholium:exact-as-presheaves}, but still acyclitity in
+the usual sense will most likely not be attainable.
+
+Secondly, while we have shown in Theorem~\ref{thm:enough-flabby-modules} that
+enough flabby envelopes exist constructively, the analysis in Section~\ref{sect:in-eff}
+demonstrates that the presented envelopes are in a sense deeply uncomputable.
+Intuitionistic Zermelo--Fraenkel set theory~\cite{crosilla:cst-izf} and the
+type theory of toposes~\cite{maietti:modular} do verify that they exist, but
+only by virtue of excessive reliance on powersets, precluding concrete
+computations; the construction is highly \emph{impredicative}, not meeting the
+bar of \emph{predicative mathematics}~\cite{crosilla:predicativity}.
+
+As a result, thirdly, flabby resolutions do not generalize from toposes to the
+setting of \emph{arithmetic universes}, the predicative cousins of toposes
+introduced by André Joyal which have recently been an important object of
+consideration by Milly Maietti and
+Steve Vickers~\cite{maietti:au,maietti-vickers:induction,vickers:sketches}.
+These are not only interesting on their own and as a convenient foundation
+for ensuring predicativity (and hence computability in a strict sense), but also
+because they can be used to obtain base-independent proofs for the topos
+case~\cite{vickers:classifying,hazratpour:phd}: \emph{Computing cohomology of
+concrete spaces should not require fixing a universe of sets first.}
 
@@
 sheaves, while still preserving the good formal properties expected from
-derived functors.
+derived functors. To this end, a mix of the approaches using
+pointwise Kan extensions, as cogently argued for by Emily
+Riehl~\cite[Chapter~2]{riehl:cathtpy}, and ind-objects as presented in the Stacks
+Project~\cite[Tag~05S7]{stacks-project} seems promising.
+
+This framework is sufficiently flexible to not demand derived functors to be
+defined everywhere. Rather, they will be defined just for those objects for
+which we happen to have a suitable resolution. Only classically, by using
+injective or flabby sheaves, can we pretend that every object has such a
+resolution. We hope to report on details in future work.
 
@@
 
-\XXX{cite more: Hyland §5,
-Fourman-Scott 1979,
-Johnstone 1977}
-
-\XXX{quasicompact case}
-
 \end{document}
 
-XXX: injective as module implies functionally flabby
-
-Cite! https://arxiv.org/pdf/2008.10677.pdf
-
 FUTURE:
@@
 
-1. Zorn
-2. effective
-3. predicative, no AU's
+* quasicompact/spectral case

### 2022-02-14 paper-flabby-objects.tex
More precise assumptions for enough injectives

@@
 \item Every injective module is also injective as a set.
-\item Assuming Zorn's lemma (but not the law of excluded middle, hence not the
-axiom of choice), every module embeds into an
+\item Assuming the axiom of choice (so Zorn's lemma in combination with the law of excluded middle), every module embeds into an
 injective module.
@@
 \end{enumerate}\end{proof}
+
+We note in passing that Remark~\ref{rem:enough-injective-modules-without-lem}
+below explains how to verify Proposition~\ref{prop:basics-injective}(3) also in
+the absence of the law of excluded middle (but still requiring Zorn's lemma).
 

### 2022-02-14 paper-flabby-objects.tex
to do

@@
 
-\XXX{If Zorn's lemma is not available but E is localic, ... is at least true
-if E is compact.}
+\begin{rem}We can dispose the reliance on Zorn's lemma in
+Scholium~\ref{scholium:exact-as-presheaves} if instead~$\E$ is compact. This is
+because XXX ... finitely many local preimages ... patch together ... by the
+following internal observation ...
+\end{rem}
 
@@
 
+TO DO:
+* enough injectives employing only Zorn
+* first names?
+
 FUTURE:
-
 * Does "flabby sheaves are acyclic on all locales" imply Zorn?
-
 * quasicompact/spectral case

### 2022-02-15 paper-flabby-objects.tex
Started on enough injectives without LEM

@@
 \newcommand{\ZZ}{\mathbb{Z}}
+\newcommand{\QQ}{\mathbb{Q}}
 \renewcommand{\P}{\mathcal{P}}
@@
 \item One verifies that any abelian group embeds into a divisible
-abelian group. By Baer's criterion (which requires the axiom of choice), divisible abelian groups are injective.
+abelian group. By Baer's criterion (which in this form requires the axiom of choice), divisible abelian groups are injective.
 The result for modules over arbitrary rings then follows purely
@@
 
-We note in passing that Remark~\ref{rem:enough-injective-modules-without-lem}
-below explains how to verify Proposition~\ref{prop:basics-injective}(3) also in
-the absence of the law of excluded middle (but still requiring Zorn's lemma).
+We note in passing that the multi-step technique of the proof of
+Theorem~\ref{thm:enough-flabby-modules} below can be used to verify
+Proposition~\ref{prop:basics-injective}(3) also in the absence of the law of
+excluded middle (but still requiring Zorn's lemma). Since this observation is of no
+further import for the purposes of this text, details are postponed to
+Section~\ref{sect:enough-injective-modules}.
 
@@
 pointwise Kan extensions, as cogently argued for by Emily
-Riehl~\cite[Chapter~2]{riehl:cathtpy}, and ind-objects as presented in the Stacks
+Riehl~\cite[Chapter~2]{riehl:cathtpy} (see also~\cite{hinich:what-is,maltsiniotis:adjunction}), and ind-objects as presented in the Stacks
 Project~\cite[Tag~05S7]{stacks-project} seems promising.
@@
 injective or flabby sheaves, can we pretend that every object has such a
-resolution. We hope to report on details in future work.
+resolution. We hope to report on details in future work. Can perhaps even the
+rare cases where injective resolutions exceptionally do have some link to
+computations, as on projective spaces~\cite{huang:cohomology}, be given a
+proper constructive and predicative home?
 
@@
 \enlargethispage{1em}
+
+\appendix
+
+% \begin{document}
+
+\section{On the existence of enough injective modules}
+\label{sect:enough-injective-modules}
+
+\begin{defn}\begin{enumerate}
+\item An abelian group~$I$ is \emph{divisible} if and only if for every
+element~$x \in I$ and every natural number~$n \geq 1$, there is an element~$y \in I$
+such that~$ny = x$.
+\item An abelian group~$I$ \emph{satisfies the Baer criterion} if and only if,
+for every ideal~$\aaa \subseteq \ZZ$, every ($\ZZ$-)linear map~$\aaa \to I$
+admits an extension along the inclusion to a linear map~$\ZZ \to I$.
+\end{enumerate}
+\end{defn}
+
+\begin{prop}
+\begin{enumerate}
+\item Assuming the law of excluded middle, divisible abelian groups satisfy
+the Baer criterion.
+\item Assuming Zorn's lemma, abelian groups which satisfy the Baer criterion
+are injective.
+\end{enumerate}
+\end{prop}
+
+\begin{proof}Assuming the law of excluded middle, the only ideals of~$\ZZ$ are
+the principal ideals~$(n)$ where~$n \geq 0$. For~$n = 0$ existence of
+extensions is trivial and for~$n > 0$ existence of extensions follow from
+divisibility.
+
+Assuming Zorn's lemma, let~$I$ be an abelian group satisfying the Baer
+criterion, let~$i : A \to B$ be a linear injection and let~$f : A \to I$ be a
+linear map. The poset of partial extensions contains suprema of chains and
+hence a maximal partial extension~$f_0 : A_0 \to I$ with~$A \subseteq A_0
+\subseteq B$. To verify that~$A_0 = B$, let an element~$x \in B$ be given.
+By the Baer condition, the linear map~$g : (A_0 : x) \to I$ defined on the
+ideal~$(A_0 : x) = \{ n \in \ZZ \,|\, nx \in A_0 \} \subseteq \ZZ$ given by~$g(n) =
+f_0(nx)$ can be extended to a linear map~$\overline{g} : \ZZ \to I$.
+Hence~$f_0$ can be extended to the map~$A_0 + (x) \to I$ given by~$u + nx
+\mapsto f_0(u) + \overline{g}(n)$. By maximality~$A_0 + (x) = A_0$, hence~$x \in A_0$.
+\end{proof}
+
+\begin{prop}
+\begin{enumerate}
+\item Any abelian group embeds into a divisible abelian group.
+\item Any abelian group embeds into an abelian group satisfying the Baer
+condition.
+\end{enumerate}
+\end{prop}
+
+\begin{proof}For the first part, an abelian group~$A$ embeds into the
+divisible group~$\QQ\langle A \rangle / K$, where~$\QQ\langle A \rangle$ is the
+underlying abelian group of the free~$\QQ$-vector space on~$A$ and~$K$ is the
+kernel of the canonical map~$\ZZ\langle A \rangle \to A$.
+
+The second part is more involved and requires, similar to the proof of
+Theorem~\ref{thm:enough-flabby-modules}, a strengthening of the Baer condition
+where the required extensions are explicitly given. As a first step, let~$A$ be
+an abelian group which is a sheaf for a modal operator~$\nabla$ such
+that~$\nabla(\varphi \vee (\varphi \Rightarrow \nabla\bot))$. Let~$i : \aaa
+\hookrightarrow \ZZ$ be the inclusion of an ideal of~$\ZZ$ and let~$f : \aaa
+\to A$ be a linear map. Applying the sheafification functor~$(\cdot)^{++}$, we
+have the following commutative diagram:
+\[ \xymatrix{
+  \aaa \ar@{^{(}->}[r]^i\ar[d]\ar@/_2pc/[dd]_f & \ZZ \ar[d] \\
+  \aaa^{++} \ar@{^{(}->}[r]^{i^{++}}\ar@{-->}[d] & \ZZ^{++}\ar@{-->}@/^/[ld]^{f'} \\
+  A
+} \]
+The given map~$f$ factors (uniquely) over the map~$\aaa \to \aaa^{++}$ since~$A$ is a
+sheaf. The map~$i^{++}$ is still injective as sheafification is exact. The two
+top vertical maps need not be injective, as~$\ZZ$ need not be separated
+for~$\nabla$; but this issue XXX
+\end{proof}
+
+\begin{prop}Assuming Zorn's lemma:
+\begin{enumerate}
+\item Any abelian group embeds into an injective abelian group.
+\item Any module embeds into an injective module.
+\item XXX Grothendieck topos?
+\end{enumerate}
+\end{prop}
 
@@
 * first names?
+* cite more from Barakat and friends?
 

### 2022-02-16 paper-flabby-objects.tex
Finished section on enough injectives

@@
 \begin{prop}
+\label{prop:embed1}
 \begin{enumerate}
-\item Assuming the law of excluded middle, divisible abelian groups satisfy
+\item Assuming the law of excluded middle, all divisible abelian groups satisfy
 the Baer criterion.
-\item Assuming Zorn's lemma, abelian groups which satisfy the Baer criterion
+\item Assuming Zorn's lemma, all abelian groups which satisfy the Baer criterion
 are injective.
@@
 hence a maximal partial extension~$f_0 : A_0 \to I$ with~$A \subseteq A_0
-\subseteq B$. To verify that~$A_0 = B$, let an element~$x \in B$ be given.
+\subseteq B$.
+
+To verify that~$A_0 = B$, let an element~$x \in B$ be given.
 By the Baer condition, the linear map~$g : (A_0 : x) \to I$ defined on the
@@
 
-\begin{prop}
+\begin{lemma}\label{lemma:q-mod-z}
+Let~$\nabla$ be a modal operator such
+that~$\nabla(\varphi \vee (\varphi \Rightarrow \nabla\bot))$ and
+let~$(\cdot)^+$ denote the plus construction with respect to~$\nabla$. Then:
 \begin{enumerate}
-\item Any abelian group embeds into a divisible abelian group.
-\item Any abelian group embeds into an abelian group satisfying the Baer
-condition.
+\item The sheafification~$(\QQ/\ZZ)^{++}$ satisfies the Baer condition.
+\item Assuming Zorn's lemma, given an element~$x$ of an abelian group~$A$,
+there is a linear map~$g : A \to (\QQ/\ZZ)^{++}$ such that~$g(x) = 0
+\Rightarrow \nabla(x=0)$.
+\end{enumerate}
+\end{lemma}
+
+\begin{proof}Let~$i : \aaa \hookrightarrow \ZZ$ be the inclusion of an ideal
+and let~$f : \aaa \to A$ be a linear map as in the solid part of the following
+commutative diagram:
+\[ \xymatrix{
+  \aaa \ar@{^{(}->}[r]^i\ar[d]\ar@/_2pc/[dd]_f & \ZZ \ar[d] \\
+  \aaa^{++} \ar@{^{(}->}[r]^{i^{++}}\ar@{-->}[d] & \ZZ^{++}\ar@{-->}@/^/[ld]^{f'} \\
+  (\QQ/\ZZ)^{++}
+} \]
+The given map~$f$ factors (uniquely) over the map~$\aaa \to \aaa^{++}$
+since~$(\QQ/\ZZ)^{++}$ is a sheaf. The map~$i^{++}$ is still injective as
+sheafification is exact. (The two top vertical maps need not be injective,
+as~$\ZZ$ need not be separated for~$\nabla$.) Hence we are reduced to an
+extension problem in the subtopos of~$\nabla$-sheaves, which is solvable
+because this subtopos is Boolean (Proposition~\ref{prop:embed1}(1)). More
+explicitly, the extension~$f'$ maps~$1 \in \ZZ$ to the gluing of
+\begin{multline*}
+  \{ 0 \,|\, \forall x \in \aaa. \nabla(x = 0) \} \cup {}\\
+  \{ \tfrac{1}{d} v \,|\, d \geq 1, v \in [0,1),
+    f(d) = \underline{v}, \forall x \in \aaa. \nabla(x \in \aaa)
+    \Leftrightarrow \nabla(\exists n \in \ZZ. x = nd) \}.
+\end{multline*}
+
+For the second part, let~$A$ be an abelian group and let~$x \in A$ be an
+element. Let~$i : \ZZ/(0:x) \to A$ be the linear map with~$[1] \mapsto x$,
+where~$(0:x)$ is the ideal~$\{ n \in \ZZ \,|\, nx = 0 \}$. In the Boolean
+subtopos of~$\nabla$-sheaves, this ideal is either the zero ideal or the
+principal ideal~$(d)$ of some positive generator. Let~$h : \ZZ/(0:x) \to
+(\QQ/\ZZ)^{++}$ be the linear map sending~$[1]$ to the respective gluing such
+that in the first case~$h([1]) = [\tfrac{1}{2}]$ and in the second case~$h([1])
+= [\frac{1}{d}]$. Assuming Zorn's lemma such that
+Proposition~\ref{prop:embed1}(2) applies, there is an extension~$\overline{h} :
+A \to (\QQ/\ZZ)^{++}$. This map is the desired map. If~$\overline{h}(x) =
+h([1]) = 0$, then in the subtopos~$d = 1$, hence~$\nabla(x = 0)$.
+\end{proof}
+
+\begin{prop}\label{prop:embed2}
+\begin{enumerate}
+\item Every abelian group embeds canonically into a divisible abelian group.
+\item Every abelian group maps canonically to an abelian group functionally satisfying the Baer
+condition. Assuming Zorn's lemma, this map is an embedding.
 \end{enumerate}
@@
 where the required extensions are explicitly given. As a first step, let~$A$ be
-an abelian group which is a sheaf for a modal operator~$\nabla$ such
-that~$\nabla(\varphi \vee (\varphi \Rightarrow \nabla\bot))$. Let~$i : \aaa
-\hookrightarrow \ZZ$ be the inclusion of an ideal of~$\ZZ$ and let~$f : \aaa
-\to A$ be a linear map. Applying the sheafification functor~$(\cdot)^{++}$, we
-have the following commutative diagram:
-\[ \xymatrix{
-  \aaa \ar@{^{(}->}[r]^i\ar[d]\ar@/_2pc/[dd]_f & \ZZ \ar[d] \\
-  \aaa^{++} \ar@{^{(}->}[r]^{i^{++}}\ar@{-->}[d] & \ZZ^{++}\ar@{-->}@/^/[ld]^{f'} \\
-  A
-} \]
-The given map~$f$ factors (uniquely) over the map~$\aaa \to \aaa^{++}$ since~$A$ is a
-sheaf. The map~$i^{++}$ is still injective as sheafification is exact. The two
-top vertical maps need not be injective, as~$\ZZ$ need not be separated
-for~$\nabla$; but this issue XXX
+an abelian group which is separated for a modal operator~$\nabla$ such
+that~$\nabla(\varphi \vee (\varphi \Rightarrow \nabla\bot))$. The codomain of
+the canonical map
+\[ A \longrightarrow \prod_{g : A \to (\QQ/\ZZ)^{++}} (\QQ/\ZZ)^{++},\
+  x \longmapsto (g(x))_g \]
+is a product of abelian groups which functionally satisfy the Baer criterion by
+Lemma~\ref{lemma:q-mod-z}(1) and hence satisfies the Baer condition itself.
+Assuming Zorn's lemma, this map is injective by Lemma~\ref{lemma:q-mod-z}(2)
+and~$\nabla$-separatedness of~$A$.
+
+For the general case, let an abelian group~$A$ be given. Denoting
+by~$(\cdot)^{+_x}$ the plus construction with respect to the modal
+operator~$\nabla_x$ with~$\nabla_x\varphi \defeqv ((\varphi \Rightarrow x
+= 0) \Rightarrow x = 0)$, the codomain of the canonical map
+\[ A \longrightarrow \prod_{x \in A} \prod_{g : A \to (\QQ/\ZZ)^{+_x+_x}}
+(\QQ/\ZZ)^{+_x+_x} \]
+satisfies the Baer criterion and is injective as in the proof of
+Theorem~\ref{thm:enough-flabby-modules}.
 \end{proof}
@@
 \begin{enumerate}
-\item Any abelian group embeds into an injective abelian group.
-\item Any module embeds into an injective module.
-\item XXX Grothendieck topos?
+\item Every abelian group embeds canonically into an injective abelian group.
+\item Every module embeds canonically into an injective module.
+\item Every module in any Grothendieck topos embeds canonically into an injective module.
 \end{enumerate}
 \end{prop}
+
+\begin{proof}The first claim is by Proposition~\ref{prop:embed2}.
+
+The second claim follows purely formally as in the proof of
+Proposition~\ref{prop:basics-injective}(3), using the adjunction between the
+category of modules and the category of abelian groups.
+
+For the third claim, let~$\E$ be a Grothendieck topos and let~$\pi : \F \to \E$ be its Barr
+covering \XXX{ref}. Zorn's lemma passes from the metatheory to~$\F$ since~$\F$ is a
+localic topos. Hence the first part applies in~$\F$. Purely formally, using the
+adjunction between~$\mathrm{Ab}(\F)$ and~$\mathrm{Ab}(\E)$, there are enough
+injective abelian groups in~$\E$. The existence of enough injective modules
+in~$\E$ follows from internalizing the second claim in~$\E$.
+\end{proof}
+
+\XXX{cite whom for the general argument?}
 

### 2022-02-16 paper-flabby-objects.tex
typo (thanks to Lukas Stoll)

@@
 language) for sheaves of abelian groups. We use the opportunity to correct a
-small mistake of hers, namely claiming that the analogous results for sheaves of
+small inaccuracy of hers, namely claiming that the analogous results for sheaves of
 modules would be false.
@@
 to Jürgen Jost, Marc Nieper-Wißkirchen and Peter Schuster for valuable
-guidance, and to Daniel Albert and Caterina Cozzi for their careful readings of
+guidance, and to Daniel Albert and Giacomo Cozzi for their careful readings of
 earlier drafts. The work for this note was carried out at the University of
@@
 If~$F$ is a flabby sheaf in the traditional sense, then~$F$ is obviously also
-flabby in the sense of Definition~\ref{defn:flabby-sheaf} -- singleton
-coverings will do. Conversely, let~$F$ be a flabby sheaf in the sense of
+flabby in the sense of Definition~\ref{defn:flabby-sheaf} -- the singleton
+covering~$X = X$ will do. Conversely, let~$F$ be a flabby sheaf in the sense of
 Definition~\ref{defn:flabby-sheaf}. Let~$s \in F(U)$ be a local section. Zorn's

### 2022-02-16 paper-flabby-objects.tex
typo

@@
 \begin{rem}\label{rem:constant-flabby}
-Although we will see in \XXX{number} and in Section~\ref{sect:in-eff} that there is some
+A set~$X$ is \emph{$\neg\neg$-separated} if and only
+if~$\neg\neg(x=y)\Rightarrow x=y$ for all elements~$x,y \in X$.
+Although Section~\ref{sect:enough-flabby-modules} presents some
 relation between flabby sets and~$\neg\neg$-separated sets, neither notion
@@
 \subsection{On the existence of enough flabby modules}
+\label{sect:enough-flabby-modules}
 
@@
 objects~\cite{harting:coproduct}, and in order to verify that taking internal
-direct sums is faithful she felt the need to employ Barr's
+direct sums is faithful she appealed to Barr's
 metatheorem~\cite[Theorem~1.7]{harting:effacements}. Nowadays we can verify

### 2022-02-16 paper-flabby-objects.tex
Better statement and better proof about Hom(T,I) (thanks to the anonymous reviewer)

@@
 \item Let~$I$ be an injective set. Let~$T$ be an arbitrary set. Then the
-set~$I^T$ of maps from~$T$ to~$I$ is flabby.
-\item Let~$I$ be an injective~$R$-module. Let~$T$ be an arbitrary~$R$-module. Then the
-set~$\Hom_R(T,I)$ of linear maps from~$T$ to~$I$ is flabby.
+set~$I^T$ of maps from~$T$ to~$I$ is injective.
+\item Let~$I$ be an injective~$R$-module. Let~$T$ be an arbitrary~$R$-module.
+If~$T$ is flat, then the module~$\Hom_R(T,I)$ of linear maps from~$T$ to~$I$ is
+injective. In the general case, it is at least flabby.
 \end{enumerate}\end{prop}
 
-\begin{proof}We first cover the case of sets. Let~$K \subseteq I^T$ be a
-subterminal. We consider the injectivity diagram
-\[ \xymatrix{
-  T' \ar@{^{(}->}[r]\ar[d] & T \ar@{-->}@/^/[ld] \\
+\begin{proof}The first claim follows abstractly from the fact that the Hom
+functor~$(\cdot)^T$ has a monomorphism-preserving left adjoint, namely the
+product functor~$(\cdot) \times T$. Explicitly, an extension problem as in the
+left half of the diagram
+\[
+  \xymatrix{
+    A \ar@{^{(}->}[r]^i\ar[d]_f & B \ar@{-->}@/^/[ld]^{\overline{f}} \\
+    I^T
+  }
+  \qquad
+  \xymatrix{
+    A \times T \ar@{^{(}->}[r]^{i \times T}\ar[d]_{g \defeq f^t} & B \times T \ar@{-->}@/^/[ld]^{\overline{g}} \\
     I
-} \]
-where~$T'$ is the subset~$\{ s \in T \,|\, \text{$K$ is inhabited} \} \subseteq T$ and the
-solid vertical map sends~$s \in T'$ to~$g(s)$, where~$g$ is an arbitrary element
-of~$K$. This association is well-defined. Since~$I$ is injective, a dotted lift
-as indicated exists. If~$K$ is inhabited, this lift is an element of~$K$.
-
-The same kind of argument applies to the case of modules. If~$K$ is a
-subterminal of~$\Hom_R(T,I)$, we define~$T'$ to be the submodule
-$\{ s \in T \,|\, \text{$s = 0$ or $K$ is inhabited} \}$ and consider the
-analogous injectivity diagram, where the solid vertical map~$f : T' \to I$ is now
-defined by cases: Let~$s \in T'$. If~$s = 0$, then we set~$f(s) = 0$; if~$K$ is
-inhabited, then we set~$f(s) \defeq g(s)$, where~$g$ is an arbitrary element
-of~$K$. This association is again well-defined, and a dotted lift yields the
-desired element of~$\Hom_R(T,I)$.
-\end{proof}
-
-Proposition~\ref{prop:hom-flabby} can be used to give an alternative proof of
-Proposition~\ref{prop:injective-flabby} and to generalize
-Proposition~\ref{prop:injective-flabby} to modules: If~$I$ is an injective set,
-then the set~$I^1 \cong I$ is flabby. If~$I$ is an injective module, then the
-set~$\Hom_R(R,I) \cong I$ is flabby.
+  }
+\]
+can be transposed to the extension problem as in the right half. A
+solution~$\overline{g}$ gives rise to the solution~$\overline{f}$ of the
+original problem by the setting~$\overline{f}(x) = (t \mapsto
+\overline{g}(x,t))$.
+
+The injectivity part of the second claim follows entirely analogously,
+employing the tensor product instead of the cartesian product.
+
+For the general statement, let~$K$ be a subterminal of~$\Hom_R(T,I)$. Let~$T'$
+be the submodule $\{ s \in T \,|\, \text{$s = 0$ or $K$ is inhabited} \}
+\subseteq T$ and let~$f : T' \to I$ be the linear map defined as follows:
+Let~$s \in T'$. If~$s = 0$, then we set~$f(s) = 0$; if~$K$ is inhabited, then
+we set~$f(s) \defeq g(s)$, where~$g$ is an arbitrary element of~$K$. This
+association is well-defined. Since~$I$ is injective as a module, there is a
+linear extension~$\overline{f} : T \to I$ of~$f$ along the inclusion~$T'
+\subseteq T$. If~$K$ is inhabited, this extension is an element of~$K$ as
+required.
+\end{proof}
+
+%Proposition~\ref{prop:hom-flabby} can be used to give an alternative proof of
+%Proposition~\ref{prop:injective-flabby} and to generalize
+%Proposition~\ref{prop:injective-flabby} to modules: If~$I$ is an injective set,
+%then the set~$I^1 \cong I$ is flabby. If~$I$ is an injective module, then the
+%set~$\Hom_R(R,I) \cong I$ is flabby.
 
@@
 
-% \begin{document}
-
 \section{On the existence of enough injective modules}

### 2022-03-09 paper-flabby-objects.tex
One more example

@@
 
+\begin{ex}The object of Dedekind reals of a topos is in general not flabby. For
+instance, in the case of the topos of sheaves over the real line~$\RR^1$, the
+object of Dedekind reals is the sheaf~$\C$ of continuous (Dedekind-)real valued
+functions~\XXX{cite}. The section~$1/x$ cannot be extended to opens containing
+the origin.\end{ex}
+
 \begin{prop}\label{prop:global-elements}

### 2022-04-02 paper-filmat.tex
arXiv version

@@
-%\documentclass[oneside,reqno]{amsart}
 \documentclass[graybox]{svmult}
-
+\pdfoutput=1
+
+\usepackage{amsmath,amssymb,etex,ifxetex}
 \usepackage{type1cm}
@@
 \usepackage[bottom]{footmisc}
-\usepackage{newtxtext}
-\usepackage{newtxmath}
+\usepackage{newtx}
 
@@
 \usepackage{natbib}
+
+\usepackage[{colorlinks,citecolor=[rgb]{0,0.1,0}}]{hyperref}
+
+%\usepackage{geometry}
+%\geometry{lmargin=4.6cm,rmargin=4.5cm}
 
@@
 
-\title*{Exploring mathematical objects from custom-tailored mathematical universes}
+\addtocounter{chapter}{3}
+
+\title{Exploring mathematical objects from custom-tailored mathematical universes}
 \author{Ingo Blechschmidt}
@@
   which the usual mathematical objects of the standard topos can be viewed.
-  \emph{Key words:} topos theory, realism debate, well-adapted language,
-  constructive mathematics
 }
+\keywords{topos theory, realism debate, well-adapted language,
+  constructive mathematics}
 \medskip
@@
 
-% \begin{document}
+%\newcommand{\RR}{\mathbb{R}}
 
@@
 uncovering hidden relations, though not between objects but between mathematical
-theories. \textbf{\textcolor{red}{NOTE TO EDITOR: Insert cross reference to Olivia's chapter of
+theories. \textit{\textcolor{red}{Note to editor: please insert cross reference to Olivia's chapter of
 this book.}}}
@@
 \begin{center}
-\begin{tabular}{lll}
+\begin{tabular}{ll@{\qquad}l}
   \toprule

### 2022-04-04 paper-flabby-objects.tex
typo

--- 2022-02-16 paper-flabby-objects.tex
@@
 
+\begin{ex}The object of Dedekind reals of a topos is in general not flabby. For
+instance, in the case of the topos of sheaves over the real line~$\mathbb{R}^1$, the
+object of Dedekind reals is the sheaf~$\mathcal{C}$ of continuous (Dedekind-)real valued
+functions~\XXX{cite}. The section~$1/x$ cannot be extended to opens containing
+the origin.\end{ex}
+
 \begin{prop}\label{prop:global-elements}

### 2022-04-04 paper-qcoh.tex
to do
typo
Counterexample demonstrating the necessity of the fp condition

@@
 \newcommand{\ZZ}{\mathbb{Z}}
+\newcommand{\QQ}{\mathbb{Q}}
 \renewcommand{\SS}{\mathbb{S}}
@@
 geometry as the little Zariski topos of~$A$, the topos of sheaves over the
-spectrum of~$A$. If~$A$ is reduced, this topos validates the dual condition that any
+spectrum of~$A$. If~$A$ is reduced, this topos validates the dual field condition that any
 noninvertible element is zero. This property has been used to give a short and
@@
 Lemma~\ref{lemma:fp-double-dual} (over arbitrary slices) results in
-isomorphisms of the same form. This fact follows from the lemma (applied twice)
-that computing the internal Hom to a sheaf commutes with sheafification and
+isomorphisms of the same form. This fact follows from the fact that
+the presheaf Hom into a sheaf is already a sheaf itself and
 from the observation that the construction~``free~$U_\TT$-module modulo
@@
 % XXX explain how Nullstellensatz follows from the Horn version
+
+\begin{rem}The finite presentability condition in
+Lemma~\ref{lemma:fp-double-dual} cannot be dropped. For instance, in the case
+that~$\TT$ is the theory of commutative rings with unit and~$A$ is the
+ring~$\QQ$ of integers, we have~$\Spec(A^\sim) \cong \Spec(0^\sim)$, where~$0$
+is the zero ring, as~$\QQ$ allows ring homomorphisms only in those finitely
+presented rings in which~$1 = 0$ holds. Hence~$A^\sim$
+and~$(U_\TT)^{\Spec(A^\sim)} \cong (U_\TT)^{\Spec(0^\sim)} \cong 0^\sim$ do not coincide.
+\end{rem}
 
@@
 * XXX Cite C2.5.4 or similar
+
+% XXX Check: https://arxiv.org/abs/1811.08838

### 2022-04-05 paper-qcoh.tex
typo

@@
 algebraic certificate we now have a logical certificate, a \emph{proof}
-of~$\sigma$. However, this generalized statement is typically false, even for
+of~$\sigma$.
+
+However, this generalized statement is typically false, even for
 the generic model~$U_\TT$: The statement
@@
 proves~$\sigma$}$ is fixed, the meaning of~$\speak{$\sigma$ holds for~$U_\TT$}$ varies
-with the slice, as~$U_\TT$ changes form on different stages. This mismatch is solved by passing from~$\ul{\TT}$ to a
+with the slice, as~$U_\TT$ shifts shape on different stages. This mismatch is solved by passing from~$\ul{\TT}$ to a
 varying theory, the internal theory~$\ul{\TT}/U_\TT$ defined in
@@
 the relation~$f(X) = 0$, the relation~$g([X]) = 0$ holds. Hence we obtain
+an algebraic Nullstellensatz:
 \[ \Set[\TT] \models
@@
 these lines should exist in a wide range of subjects. We hope that with the
-Nullstellensatz at hand, discovering such techniques can progress in a more
+Nullstellensatz at hand, the discovery of such techniques can progress in a more
 systematic fashion.
@@
 None of these complications arise when setting up the theory of internal
-coherent theories.
+coherent theories, where disjunctions are restricted to be finitary.
 
@@
 
-\begin{rem}In the case that the~$\Set[\TT']$ is a dense subtopos
+\begin{rem}In the case that the subtopos~$\Set[\TT']$ is a dense subtopos
 of~$\Set[\TT]$, Corollary~\ref{corollary:nullstellensatz-horn} can be strengthened to
@@
 Theorem~\ref{thm:nullstellensatz} cannot be strengthened to arbitrary
-first-order (or first-order$^\star$ or even just coherent) formulas in place of
+first-order (or first-order$^\star$ or even finitary first-order) formulas in place of
 geometric$^\star$ sequents. For instance, in the case that~$\TT$ is the theory
-of rings, the generic model~$U_\TT$ validates the coherent
+of rings, the generic model~$U_\TT$ validates the finitary first-order
 formula~$\speak{any element~$x$ for which~$(x = 0 \Rightarrow 1 = 0)$ is invertible}$, but~$\ul{\TT}/U_\TT$
@@
 Throughout this section, let~$\TT$ be a Horn theory. It is a basic result that
-Horn theories are of presheaf type, such that the classifying topos~$\Set[\TT]$
+Horn theories are of presheaf type, that is the classifying topos~$\Set[\TT]$
 can be taken as the topos of functors~$\Mod{\TT}_\mathrm{fp} \to \Set$,
@@
 instance, the coproduct of~$\TT\langle X | R \rangle$ and~$\TT\langle X' | R'
-\rangle$ is~$\TT\langle X \amalg X' \,|\, R, R' \rangle$.\end{proof}
+\rangle$ is~$\TT\langle X \amalg X' \,|\, R \cup R' \rangle$.\end{proof}
 % XXX sagen, dass jedes Modell (beliebig) präsentiert wird
@@
 \begin{lemma}\label{lemma:spec-sim-representable}
-Let~$A$ be a~$\TT$-model. Then~$\Spec(A^\sim)$ coincides
+Let~$A$ be a~$\TT$-model (not necessarily finitely presentable). Then~$\Spec(A^\sim)$ coincides
 with~$\yon A$, the functor~$\Hom_{\Mod{\TT}}(A, \cdot)$.
@@
 
-\begin{proof}By the Yoneda lemma, the sections of the sheaf~$\Spec(A^\sim) :
+\begin{proof}By the Yoneda lemma, the sections of the presheaf~$\Spec(A^\sim) :
 \Mod{\TT}_\mathrm{fp} \to \Set$ on an object~$T$ are given by the set
@@
 \Mod{\TT}_\mathrm{fp}$,~$T$-elements~$X_1,\ldots,X_n$ of the object of sorts of
-the signature of~$\ul{\TT}/U_\TT$ (that is the constant sheaf on the set of
+the signature of~$\ul{\TT}/U_\TT$ (that is the constant presheaf on the set of
 sorts of~$\TT$), and~$T$-elements~$R_1,\ldots,R_m$ of the object of atomic
-propositions over the signature of~$\ul{\TT}/U_\TT$ be given. By passing to a
-covering of~$T$, we may assume that the~$X_i$ are given by sorts of~$\TT$ and
-that the~$R_j$ are given by atomic propositions over the signature
+propositions over the signature of~$\ul{\TT}/U_\TT$ be given. The~$X_i$ are given by sorts of~$\TT$ and
+the~$R_j$ are given by atomic propositions over the signature
 of~$\TT/U_\TT(T)$.
-% XXX: No passing to covers is necessary, as we're in a *pre*sheaf topos!
 
@@
 
-\begin{thm}Let~$\TT'$ be a quotient theory of~$\TT$, not necessarily Horn.
+\begin{thm}Let~$\TT'$ be a geometric quotient theory of~$\TT$, not necessarily Horn.
 Assume that~$U_\TT$ is a sheaf for the topology
@@
 classifying topos of the theory~$\OO$ of objects, the theory with a single sort and
-no function symbols, relation symbols or axioms, including the following ones.
+no function symbols, relation symbols or axioms. In particular, they present
+the following ones.
 \begin{align*}

### 2022-04-06 paper-filmat.tex
typo (thanks to Lukas!)

@@
 By their definition as special kinds of
-categories, toposes are merely algebraic structures not unlike groups or vector
+categories, toposes are merely essentially algebraic structures not unlike groups or vector
 spaces. Hence we need to argue why we picture toposes as mathematical universes
@@
 group has property~$\varphi$'' and~``all groups have property~$\varphi$'' need
-not be equivalent. This inbalance has mathematical applications and is explored
+not be equivalent. This imbalance has mathematical applications and is explored
 in~\cite{blechschmidt:nullstellensatz}.}

### 2022-04-06 paper-filmat.tex
typo (thanks to Lukas!)
typo

@@
 interleaving simulation and the comparison procedure -- but it does not depend on
-the implementation of~$X$ or on its specific choice of rational
+the implementation of~$X$ nor on its specific choice of rational
 approximations~$X(n)$. The search terminates since there is at least one
@@
 resulting toposes constitute the large class of Grothendieck toposes, the
-original notion of toposes. Categorically, the passage from topological spaces
+original notion of toposes. Categorically, the step from topological spaces
 to sites is rather small, but the resulting increase in flexibility is
@@
 \]
-We write~$\langle a,b \rangle$ more clearly
+We write~$\langle a,b \rangle$ more suggestively
 as~$a+b\varepsilon$.

### 2022-07-05 algar-lecture-notes.tex
Beginnings of ALGAR lecture notes

\documentclass[10pt,reqno,a4paper,openany]{amsbook}
\usepackage{etex}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{etoolbox,chngcntr}
\usepackage{amsmath,amsthm,amssymb,array,stmaryrd,color,graphicx,mathtools,multirow,setspace}
\usepackage{soul}\setul{0.3ex}{}
\usepackage{bussproofs}
\usepackage{xspace}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage[bookmarksdepth=2,pdfencoding=auto]{hyperref}

\usepackage{libertine}

% Hack to load extpfeil from https://tex.stackexchange.com/a/297109/32372
\expandafter\def\csname opt@stmaryrd.sty\endcsname
{only,shortleftarrow,shortrightarrow}
\usepackage{extpfeil}
\newextarrow{\xbigtoto}{{20}{20}{20}{20}}
   {\bigRelbar\bigRelbar{\bigtwoarrowsleft\rightarrow\rightarrow}}

\usepackage[all]{xy}
\usepackage{tikz}
\usetikzlibrary{calc,shapes.callouts,shapes.arrows,matrix,patterns}
\newcommand{\hcancel}[5]{%
    \tikz[baseline=(tocancel.base)]{
        \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
        \draw[red, line width=0.3mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
    }%
}

\usepackage[natbib=true,style=numeric,maxnames=10]{biblatex}
\usepackage[babel]{csquotes}
\bibliography{bibliography}

\theoremstyle{definition}
\newtheorem{defn}{Definition}[section]
\newtheorem{ex}[defn]{Example}

\theoremstyle{plain}
\newtheorem{prop}[defn]{Proposition}
\newtheorem{cor}[defn]{Corollary}
\newtheorem{lemma}[defn]{Lemma}
\newtheorem{thm}[defn]{Theorem}
\newtheorem{scholium}[defn]{Scholium}

\theoremstyle{remark}
\newtheorem{rem}[defn]{Remark}
\newtheorem{question}[defn]{Question}
\newtheorem{speculation}[defn]{Speculation}
\newtheorem{caveat}[defn]{Caveat}
\newtheorem{conjecture}[defn]{Conjecture}

\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\FF}{\mathbb{F}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\G}{\mathcal{G}}
\let\acuteH\H
\newcommand{\konig}{K\acuteH onig}
\renewcommand{\H}{\mathcal{H}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\L}{\mathcal{L}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\U}{\mathcal{U}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\GG}{\mathbb{G}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\ccc}{\mathfrak{c}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\qqq}{\mathfrak{q}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\nnn}{\mathfrak{n}}
\newcommand{\?}{\,{:}\,}

\newenvironment{indentblock}{%
  \list{}{\leftmargin\leftmargin}%
  \item\relax
}{%
  \endlist
}

\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defequiv}{\vcentcolon\equiv}
\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}

\definecolor{gray}{rgb}{0.7,0.7,0.7}

\title{The curious world of constructive mathematics}
\author{Ingo Blechschmidt}
%\email{iblech@speicherleck.de}

\makeatletter
\counterwithout{section}{chapter}
\counterwithout{footnote}{chapter}
\counterwithout{table}{chapter}
\counterwithout{figure}{chapter}
\patchcmd{\@thm}{\let\thm@indent\indent}{\let\thm@indent\noindent}{}{}
\patchcmd{\@thm}{\thm@headfont{\scshape}}{\thm@headfont{\bfseries}}{}{}
\patchcmd{\@makechapterhead}{\chaptername}{Lecture}{}{}
\patchcmd{\@chapter}{\chaptername}{Part}{}{}
\patchcmd{\@schapter}{\chaptername}{Part}{}{}
\addto\captionsenglish{\renewcommand\chaptername{Lecture}}
\def\l@section{\@tocline{1}{0pt}{1pc}{}{}} % \bfseries}}
\def\l@chapter{\@tocline{-1}{12pt}{0pt}{}{\bfseries}}
\renewcommand\thechapter{\Roman{chapter}}
\newcommand{\nocontentsline}[3]{}
\newcommand{\tocless}[1]{\let\addcontentsline=\nocontentsline}
\normalparindent=12pt
\parindent=\normalparindent
\renewenvironment{proof}[1][\proofname]{\par
  \pushQED{\qed}%
  \normalfont \topsep6\p@\@plus6\p@\relax
  \trivlist
  \item[\hskip\labelsep
        \itshape
    #1\@addpunct{.}]\ignorespaces
}{%
  \popQED\endtrivlist\@endpefalse
}
\let\@afterindenttrue\@afterindentfalse
\def\subsection{\@startsection{subsection}{2}%
  {0pt}{.5\linespacing\@plus.7\linespacing}{-.5em}%
  {\normalfont\bfseries}}
\makeatother

\newenvironment{intro}{\begin{quote}}{\end{quote}}

\begin{document}

\begin{abstract}
Constructive mathematics is a flavor of mathematics in which we use the
axiom of choice and the technique of proof by contradiction only in
certain special cases. The square root of two is constructively still
irrational, but there might be vector spaces without a basis.

As a result, proofs are more informative (for instance regarding
bounds), finer distinctions can be made (for instance between positive
existence and mere impossibility of non-existence) and results apply
more generally: Every constructive result also has a geometric
interpretation, where it applies to continuous families, and an
algorithmic interpretation, yielding computational witnesses such as
procedures for computing the objects whose existence has been shown.

Relinquishing the axiom of choice and the law of excluded middle also
allows us to explore axioms and notions which are incompatible with
these classical laws, such as mathematical settings in which all
functions are continuous or in which the intuitive idea of a ``generic
ring'' can be put on a firm basis.
\end{abstract}

{
\renewcommand{\newpage}{\ }
\renewcommand{\vfill}{\ }
\renewcommand{\vfil}{\ }
\maketitle
}

\setcounter{tocdepth}{1}
{
\renewcommand{\newpage}{\ }
\tableofcontents
}

\chapter{A first glimpse of constructive mathematics}

\begin{intro}
\it
This lecture provides a first glimpse of constructive mathematics with a
focus on applications of constructive mathematics and on providing
intuition for quickly discerning which techniques and results hold
constructively.
\end{intro}

\vfill
{\small
\begin{verbatim}
- What constructive mathematics is about: finer distinctions, algorithmic
  interpretation, geometric interpretation (including teaser on why we
  might care)
- Proof by and proof of a contradiction: powers of irrational numbers,
  discreteness of N, Q, Qbar, failure of discreteness of R
- Positive information from negative? de Morgan, minimums of sets of
  natural numbers
- Relation to classical mathematics
- Finite sets and the axiom of choice
- Axiomatic freedom: all functions continuous, all functions
  computable, generic prime ideal
\end{verbatim}
}

\chapter{On the constructive content of classical proofs}

\begin{intro}
Decades of experience in constructive mathematics show: \emph{Most results
in classical mathematics, even those whose proof rests on
non-constructive principles like the axiom of choice or the law of
excluded middle, have a hidden constructive core.} With a mix of
experience, seasoned tools and general metatheorems, this constructive
content can be extracted from classical proofs. In this way we obtain
constructive reformulations of classical results, especially
if they are of a sufficiently concrete nature.

For instance, while the existence of maximal ideals in arbitrary rings
is equivalent to the axiom of choice, every first-order consequence of
their existence for linear algebra over rings also holds constructively.

This lecture illustrates the latent constructive nature of classical
proofs with examples and presents two general metatheorems which
elucidate proof mining for constructive content.
\end{intro}

\section{The double-negation translation}

\section{Barr's theorem}

\section{Examples from quadratic form theory}

\chapter{Generic models and their applications}

\begin{intro}
Commutative algebra progressed when the intuitive but informal notion of
``the generic element of a given field~$k$'' was reified in the form of the
specific element~$X$ of the polyomial ring~$k[X]$. The caveat is, of course,
that the expanded ring~$k[X]$ is no longer a field.

A similar story unfolds one level up. \emph{Topos theory provides us with ``the
generic ring'', the ring we are implicitly picturing when someone utters
the phrase ``Let~$R$ be a ring''.} The generic ring has exactly those
properties (of the large class of ``geometric implications'') which all rings
have. To have the generic ring in our ontology,
we need to broaden our notion of existence---the generic ring is not a
ring in the usual sense of the word, but a ring object in an enlarged
topos, and still close enough to the familiar rings in that all
constructive theorems about rings apply to it.

Bizarrely, the generic ring has the property (not formalizable as a
geometric implication) that it is even a field. Hence, if we are to
prove a geometric implication for all rings, we can just as well assume
the field property.

The lecture introduces the notion of topos-theoretic generic models in
general, focussing on the generic ring, the generic prime ideal and
concrete applications.
\end{intro}

\vfill{\small%
\begin{verbatim}
- Existence of generic models
- Generic ring is a field
- Generic prime ideal
- Kripke–Joyal semantics
\end{verbatim}
}

\end{document}

### 2022-07-06 paper-flabby-objects.tex
typo

@@
 such that~$ny = x$.
-\item An abelian group~$I$ \emph{satisfies the Baer criterion} if and only if,
+\item An abelian group~$I$ \emph{satisfies the Baer condition} if and only if,
 for every ideal~$\aaa \subseteq \ZZ$, every ($\ZZ$-)linear map~$\aaa \to I$
@@
 
+Trivially, every group satisfying the Baer condition is also divisible.
+
 \begin{prop}
@@
 \item Assuming the law of excluded middle, all divisible abelian groups satisfy
-the Baer criterion.
-\item Assuming Zorn's lemma, all abelian groups which satisfy the Baer criterion
+the Baer condition.
+\item Assuming Zorn's lemma, all abelian groups which satisfy the Baer condition
 are injective.
@@
 Assuming Zorn's lemma, let~$I$ be an abelian group satisfying the Baer
-criterion, let~$i : A \to B$ be a linear injection and let~$f : A \to I$ be a
-linear map. The poset of partial extensions contains suprema of chains and
+condition, let~$i : A \to B$ be a linear injection and let~$f : A \to I$ be a
+linear map. The poset of partial extensions of~$f$ contains suprema of chains and
 hence a maximal partial extension~$f_0 : A_0 \to I$ with~$A \subseteq A_0
@@
 \begin{proof}Let~$i : \aaa \hookrightarrow \ZZ$ be the inclusion of an ideal
-and let~$f : \aaa \to A$ be a linear map as in the solid part of the following
+and let~$f : \aaa \to (\QQ/\ZZ)^{++}$ be a linear map as in the solid part of the following
 commutative diagram:
@@
   x \longmapsto (g(x))_g \]
-is a product of abelian groups which functionally satisfy the Baer criterion by
+is a product of abelian groups which functionally satisfy the Baer condition by
 Lemma~\ref{lemma:q-mod-z}(1) and hence satisfies the Baer condition itself.
@@
 (\QQ/\ZZ)^{+_x+_x} \]
-satisfies the Baer criterion and is injective as in the proof of
+satisfies the Baer condition and is injective as in the proof of
 Theorem~\ref{thm:enough-flabby-modules}.

### 2022-08-15 algar-lecture-notes.tex
First steps into constructive mathematics

@@
 \usepackage{bussproofs}
+\usepackage{manfnt}
 \usepackage{xspace}
@@
 \theoremstyle{definition}
-\newtheorem{defn}{Definition}[section]
+\newtheorem{defn}{Definition}[chapter]
 \newtheorem{ex}[defn]{Example}
@@
 
-\newenvironment{intro}{\begin{quote}}{\end{quote}}
+\newenvironment{intro}{\begin{quote}}{\end{quote}\bigskip}
+
+\newtheorem{exercise}[defn]{Exercise}
+\renewcommand{\theenumi}{\alph{enumi}}
 
@@
 \end{intro}
+
+\begin{prop}There are irrational numbers~$x$ and~$y$ such that~$x^y$ is
+rational.
+\end{prop}
+\begin{proof}[First proof] The number~$\sqrt{2}^{\sqrt{2}}$ is rational or
+irrational. In the first case, set~$x \defeq \sqrt{2}$, $y \defeq \sqrt{2}$.
+In the second case, set~$x \defeq \sqrt{2}^{\sqrt{2}}$, $y \defeq \sqrt{2}$.
+\end{proof}
+\begin{proof}[Second proof] Set~$x \defeq \sqrt{2}$ and~$y \defeq \log_{\sqrt{2}} 3$.
+Then~$x^y = 3$ is rational. The verification that~$y$ is irrational is even
+easier than that of~$\sqrt{2}$.\footnote{Let~$y = p/q$ with~$p, q \in \ZZ$
+and~$q \neq 0$. Since~$y > 0$, we may assume~$p, q \in \NN$. Then~$3 =
+(\sqrt{2})^{p/q}$, hence~$3^{2q} = 2^p$. This is in contradiction to the
+uniqueness of the prime factor decomposition, since the factor~$3$ occurs on
+the left but not on the right.}
+\end{proof}
+
+The first proof is \emph{unconstructive}: It does not actually give us an
+example for a pair~$(x,y)$ as desired. In contrast, the second proof is
+constructive -- the existential claim is verified by an explicit construction
+of a suitable example.
+
+Of the many axioms and inference rules of classical logic, exactly one is
+responsible for enabling unconstructive arguments, namely the \emph{principle
+of excluded middle}:
+\[ \varphi \ \vee\ \neg\varphi. \]
+The first proof above used this principle in its very first step. In
+constructive mathematics, we abstain from this principle; we build
+constructive mathematics on \emph{intuitionistic logic}, which contains neither
+this principle nor the (equivalent) \emph{principle of double negation
+elimination} stating~$\neg\neg\varphi \Rightarrow \varphi$, and insofar as we
+layer a set theory on top of our logical foundation, we abstain from the axiom
+of choice (which in presence of other common set-theoretical axioms implies the
+principle of excluded middle, see Exercise~\ref{ex:diaconescu}).
+
+\marginpar{\dbend}
+In constructive mathematics, we do \emph{not} claim that the principle of
+excluded middle is false. Indeed, intuitionistic logic is downwardly compatible
+with classical logic (every intuitionistic proof is a fortiori also a classical
+proof), and some special instances of the principle of excluded middle are
+intuitionistically verifiable (an example is given in
+Proposition~\ref{prop:discreteness}). Instead, in constructive mathematics we
+merely do not use the principle of excluded middle.
+
+\begin{exercise}[Basics on negation]
+Recalling that negation is defined as implying absurdity,~$\neg\varphi
+\defequiv (\varphi \Rightarrow \bot)$, verify intuitionistically:
+\begin{enumerate}
+\item $\varphi \Rightarrow \neg\neg\varphi$
+\item $\neg\neg\neg\varphi \Leftrightarrow \neg\varphi$
+\item $\neg\neg(\varphi \vee \neg\varphi)$
+\item The principle of excluded middle, for all statements~$\varphi$, is
+equivalent to the principle of double negation elimination for all statements.
+
+{\noindent\scriptsize\emph{Hint.}
+Don't try to verify that double negation elimination for a specific
+statement~$\varphi$ implies the principle of excluded middle for that same
+statement~$\varphi$ -- this cannot be shown.\par}
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}[An epistemic riddle on transcendental numbers]
+Verify, using a proof by contradiction, that at least one of the numbers~$e +
+\pi$ and~$e - \pi$ is transcendental; and that at least one of the
+numbers~$e + \pi$ and~$e \cdot \pi$ is transcendental.
+\end{exercise}
+
+\begin{exercise}[Diaconescu's theorem]
+\label{ex:diaconescu}
+The axiom of choice can be put as: ``Every surjective map~$f$ has a
+section~$s$, that is a map in the other direction such that~$f \circ s =
+\mathrm{id}$.'' This exercise is on Diaconescu's theorem, stating that
+the axiom of choice implies the law of excluded middle. To this end, let~$\varphi$ be a
+statement and consider the subsets
+\begin{align*}
+  U &= \{ x \in X \,|\, (x = 0) \vee \varphi \} \\
+  V &= \{ x \in X \,|\, (x = 1) \vee \varphi \}
+\end{align*}
+of the discrete set~$X \defeq \{ 0,1 \}$.
+\begin{enumerate}
+\item Verify that~$U = V$ if and only if~$\varphi$.
+\item Using that~$x = y \vee x \neq y$ for all elements~$x,y \in X$, show that
+the existence of a section of the surjective map
+\[ \begin{array}{@{}rcl@{}}
+  X &\longrightarrow& \{U,V\} \\
+  0 &\longmapsto& U \\
+  1 &\longmapsto& V
+\end{array} \]
+implies~$\varphi \vee \neg\varphi$.
+\end{enumerate}
+\end{exercise}
 

### 2022-09-02 algar-lecture-notes.tex
More exercises on basics of constructive mathematics

@@
 
+\graphicspath{{images/}}
 \usepackage{libertine}
@@
 \newcommand{\?}{\,{:}\,}
+\renewcommand{\_}{\mathpunct{.}\,}
 
@@
 Recalling that negation is defined as implying absurdity,~$\neg\varphi
-\defequiv (\varphi \Rightarrow \bot)$, verify intuitionistically:
+\defequiv (\varphi \Rightarrow \bot)$, verify intuitionistically without
+recourse to truth tables:
 \begin{enumerate}
@@
 \item $\neg\neg(\varphi \vee \neg\varphi)$
-\item The principle of excluded middle, for all statements~$\varphi$, is
-equivalent to the principle of double negation elimination for all statements.
+\item $(\forall\psi\_ (\psi \vee \neg\psi))
+\Longleftrightarrow (\forall\psi\_ ((\neg\neg\psi) \Rightarrow
+\psi))$
 
@@
 Don't try to verify that double negation elimination for a specific
-statement~$\varphi$ implies the principle of excluded middle for that same
-statement~$\varphi$ -- this cannot be shown.\par}
+statement~$\psi$ implies the principle of excluded middle for that same
+statement~$\psi$ -- this cannot be shown. There are subtleties regarding the
+quantification over $\psi$ (this is not expressible in pure first-order logic),
+however the exercise is still instructive if we gloss over this issue.\par}
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}[Constructive status of classical tautologies]
+Which of the following classical tautologies can reasonably be expected to
+admit constructive proofs?
+\begin{enumerate}
+\item $\neg(\alpha \vee \beta) \ \Longrightarrow\ \neg\alpha \wedge \neg\beta$
+\item $\neg(\alpha \wedge \beta) \ \Longrightarrow\ \neg\alpha \vee \neg\beta$
+\item $(\alpha \vee \beta) \wedge \neg\alpha \ \Longrightarrow\ \beta$
+\item $\forall M \in P(X)\_ (\exists x\?X\_ x \in M) \vee M = \emptyset$
+(already interesting for~$X = \{\star\}$)
+\item $\forall n\?\NN\_ (n = 0 \vee n \neq 0)$
+\item $\forall x\?\RR\_ (x = 0 \vee x \neq 0)$
+\item $\forall x\?\RR\_ (\neg(\exists y\?\RR\_ xy=1) \Rightarrow x = 0)$
+\item $\forall z\?\overline{\QQ}\_ (z = 0 \vee z \neq 0)$
+\item $\forall f \? \NN \to \{0,1\}\_ (\neg\neg\exists n \? \NN\_ f(n) = 0)
+\Rightarrow (\exists n \? \NN\_ f(n) = 0)$
+\item $\forall f \? \NN \to \{0,1\}\_ \exists n \? \NN\_ (f(n) = 1
+\Leftrightarrow (\forall m \? \NN\_ f(m) = 1))$
+\item $\forall f \? \NN \to \{0,1\}\_ (\exists n \? \NN\_ f(n) = 0) \vee
+(\forall n \? \NN\_ f(n) = 1)$
+\item $\forall f \? \NN_\infty \to \{0,1\}\_ (\exists n \? \NN_\infty\_ f(n) = 0) \vee
+(\forall n \? \NN_\infty\_ f(n) = 1)$
+
+{\noindent\scriptsize\emph{Remark.} The set~$\NN_\infty$ is the \emph{one-point
+compactification} of~$\NN$. A sensible definition in constructive mathematics
+is as the set of decreasing binary sequences~$(x_0,x_1,x_2,\ldots)$. The
+naturals embed into~$\NN_\infty$ by mapping~$n$ to the sequence~$1^n 0^\omega =
+(1,\ldots,1,0,\ldots,0)$, and an element not in the image of this embedding
+is~$\infty \defeq 1^\omega = (1,1,\ldots)$. Assuming the law of excluded
+middle (or already weaker principles), every element of~$\NN_\infty$ is of one
+of these two forms. Martín Escardó has worked extensively on unexpected
+instances of the principle of omniscience for searchable sets
+like~$\NN_\infty$~\cite{escardo:omniscient1,escardo:omniscient2}.\par}
 \end{enumerate}
@@
 \label{ex:diaconescu}
-The axiom of choice can be put as: ``Every surjective map~$f$ has a
-section~$s$, that is a map in the other direction such that~$f \circ s =
-\mathrm{id}$.'' This exercise is on Diaconescu's theorem, stating that
+The axiom of choice can be put as: ``Every surjective map has a
+section.'' (A \emph{section}~$s$ to a surjective map~$f$ is a map in the other direction such that~$f \circ s =
+\mathrm{id}$.) It is a theorem due to Diaconescu that
 the axiom of choice implies the law of excluded middle. To this end, let~$\varphi$ be a
@@
 \end{enumerate}
+\end{exercise}
+
+\begin{exercise}[A stronger form of the irrationality of~$\sqrt{2}$]
+\label{ex:sqrt2}
+Mine the proof of Proposition~\ref{prop:sqrt2} to give an intuitionistic proof
+that for every rational number~$x$, the distance~$|\sqrt{2}-x|$ is positive.
+
+{\noindent\scriptsize\emph{Note.} Strictly speaking, this exercise presupposes
+familiarity with an intuitionistic account of the basics of undergraduate real
+analysis. Without it, one cannot really be expected to precisely think about
+these matters. One such account (though assuming the axiom of dependent choice)
+is~\cite{bridges}. However, this exercise is insightful even when carried out
+slightly informally. Keep in mind that, to show that a real number is positive,
+constructively it is not enough to merely verify that it cannot be zero or
+negative. A safe way to verify that a real number~$a$ is positive is to exhibit
+a rational number~$b$ such that~$a \geq b > 0$.\par}
 \end{exercise}
@@
 \end{document}
+
+omniscient1
+https://www.cs.bham.ac.uk/~mhe/.talks/dagstuhl2011/omniscient.pdf
+
+omniscient2
+https://www.ioc.ee/~tarmo/tsem16/escardo2605-slides.pdf

### 2022-09-02 algar-lecture-notes.tex
typo
Intepretation of statements

@@
 Then~$x^y = 3$ is rational. The verification that~$y$ is irrational is even
-easier than that of~$\sqrt{2}$.\footnote{Let~$y = p/q$ with~$p, q \in \ZZ$
-and~$q \neq 0$. Since~$y > 0$, we may assume~$p, q \in \NN$. Then~$3 =
-(\sqrt{2})^{p/q}$, hence~$3^{2q} = 2^p$. This is in contradiction to the
+easier than that of~$\sqrt{2}$.\footnote{Let~$y = a/b$ with~$a, b \in \ZZ$
+and~$b \neq 0$. Since~$y > 0$, we may assume~$a, b \in \NN$. Then~$3 =
+(\sqrt{2})^{a/b}$, hence~$3^{2b} = 2^a$. This is in contradiction to the
 uniqueness of the prime factor decomposition, since the factor~$3$ occurs on
@@
 
-\marginpar{\dbend}
 In constructive mathematics, we do \emph{not} claim that the principle of
-excluded middle is false. Indeed, intuitionistic logic is downwardly compatible
+excluded middle is false. \marginpar{\dbend}
+Indeed, intuitionistic logic is downwardly compatible
 with classical logic (every intuitionistic proof is a fortiori also a classical
@@
 merely do not use the principle of excluded middle.
+
+Also, deducing from a statement of the form~``$\exists x \in X\_ \varphi(x)$''
+that there actually is an element~$x \in X$ such that~$\varphi(x)$, and then
+using this particular element in the rest of an argument, is intuitionistically a
+valid logical inference, just as it is in classical logic. The axiom of choice
+is unrelated to this kind of proof step, even though they are sometimes
+mistaken.
+
+\section{Proof by contradiction vs. proof of a negation}
+
+A rumor about constructive mathematics states that constructively, the term
+``contradiction'' would be generally forbidden. This is false. In fact, we need
+to distinguish between two distinct figures of proof which are often conflated
+in classical informal mathematics:
+
+\begin{enumerate}
+\item[1.] ``Assume~$\neg\varphi$. Then $\ldots$, a contradiction.
+Hence~$\neg\varphi$ was false and thus~$\varphi$ holds.''
+\item[2.] ``Assume~$\psi$. Then $\ldots$, a contradiction. Hence~$\neg\psi$.''
+\end{enumerate}
+
+Arguments of the first form are proper proofs by contradiction and hence not
+generally acceptable in constructive mathematics. What is established is
+only~$\neg(\neg\varphi)$, the impossibility of~$\neg\varphi$; constructively,
+this is weaker than a positive affirmation of~$\varphi$. Though there are
+situations in which~$\neg\neg\varphi$ implies~$\varphi$, generally it does not
+and hence intuitionistically arguments of the first form fail to
+estabilish~$\varphi$.
+
+In contrast, arguments of the second kind are fine from an intuitionistic point
+of view: They are not proper proofs by contradiction, but proofs of negated
+statements. That such proofs are valid follows directly from the definition of
+negation (which texts on classical logic often also adopt) as a certain
+implication:
+\[ \neg\psi \defequiv (\psi \Rightarrow \bot), \]
+where~``$\bot$'', pronounced ``bottom'', is \emph{absurdity}, a canonical false
+statement. (Informally, absurdity is also written as~``$\lightning$'' or~``$1 = 0$''.)
+Hence, to establish~$\neg\psi$, we may give a proof of~$\bot$ under
+the assumption~$\psi$, just as we may give a proof of~$\beta$ under the
+assumption~$\alpha$ if we want to prove~$\alpha \Rightarrow \beta$.
+
+The two proofs of the following fact from number theory demonstrate the
+difference:
+\begin{prop}\label{prop:sqrt2}The number~$\sqrt{2}$ is not rational.\end{prop}
+\begin{proof}[Proof (only valid classically)]
+Assume that the claim is false, that is, that the number~$\sqrt{2}$ is
+\emph{not not} rational. Then~$\sqrt{2}$ is rational. Hence there are
+integers~$a$ and~$b$ with~$\sqrt{2} = a / b$. Thus~$2b^2 = a^2$. This identity
+contradicts the uniqueness of the prime factor decomposition, since the
+factor~$2$ occurs an odd number of times on the left, but an even number of
+times on the right.
+\end{proof}
+\begin{proof}[Proof (also valid intuitionistically)]
+Assume that the number~$\sqrt{2}$ is rational. Then there are integers~$a$
+and~$b$, \ldots, a contradiction.
+[The theorem on the uniqueness of the prime factor decomposition admits an
+intuitionistic proof.]
+\end{proof}
+
+Constructively stronger than the statement that~$\sqrt{2}$ is merely
+\emph{not rational} is the statement that for every rational number~$x$
+the distance~$|\sqrt{2} - x|$ is positive. This stronger claim admits an
+intuitionistic proof as well (Exercise~\ref{ex:sqrt2}).
+
+\section{Constructive meaning of mathematical statements}
+
+By our training in classical mathematics, abstaining from the principle of
+excluded middle can feel peculiar, perhaps even outrageous: Isn't it obvious
+that every mathematical statement is true or false?
+
+This sense of bewilderment is resolved by observing that \emph{even though
+constructive mathematicians use the same logical symbols, they have a slightly
+different meaning in mind.} When a constructive mathematician argues for some
+statement~$\varphi$, they mean that they have an \emph{explicit witness}
+for~$\varphi$. This shift in meaning from the classical interpretation is
+elaborated by the \emph{Brouwer--Heyting--Kolmogorov interpretation} sketched
+below. It is informal and philosophical in nature, but still a useful guide to
+the constructive meaning of mathematical statements.
+
+\subsection*{The Brouwer--Heyting--Kolmogorov interpretation}
+
+The notion of witnesses is compositional in nature and begins with witnesses for \emph{atomic
+statements}, those not built from substatements using the logical
+connectives~$\wedge,\vee,\Rightarrow$ or the quantifiers~$\forall,\exists$. For
+instance, in formal number theory, the atomic statements are of the form
+``$s = t$'',
+where~$s$ and~$t$ are terms for natural numbers; statements of this form are so
+simple that we can check them directly so that witnesses don't need to supply
+particular additional information. We hence decree that true atomic statements
+have trivial witnesses and that false atomic statements have no witnesses at
+all.
+
+Table~\ref{table:bhk} explains what witnesses of compound statements are. For
+instance, a witness of a statement of the form
+\[ \forall n \? \NN\_ \bigl(\varphi(n) \Rightarrow \psi(n)\bigr) \]
+is a rule explaining how, for any natural number~$n : \NN$, witnesses
+for~$\varphi(n)$ give rise to witnesses for~$\psi(n)$.
+
+\begin{table}
+  \centering
+  \small
+  \renewcommand{\arraystretch}{1.3}
+  \begin{tabular}{@{}rp{5.9cm}p{6.5cm}@{}}
+    \toprule
+    & {klassische Logik} & {intuitionistische Logik}
+    \\\midrule
+    Aussage $\varphi$ & Die Aussage $\varphi$ gilt. & Wir haben Beleg für $\varphi$. \\
+    $\bot$ & Es stimmt Falschheit. & Wir haben Beleg für Falschheit. \\
+    $\varphi \wedge \psi$ & $\varphi$ und $\psi$ stimmen. & Wir haben Beleg für~$\varphi$ und für~$\psi$. \\
+    $\varphi \vee \psi$ & $\varphi$ oder $\psi$ stimmt. & Wir haben Beleg für~$\varphi$ oder für~$\psi$. \\
+    $\varphi \Rightarrow \psi$ & Sollte~$\varphi$ stimmen, dann auch~$\psi$. &
+    Aus Belegen für~$\varphi$ können wir (gleich\-mä\-ßig) Belege für~$\psi$ konstruieren. \\
+    $\neg\varphi$ &
+      $\varphi$ stimmt nicht. &
+      Es kann keinen Beleg für~$\varphi$ geben. \\
+    $\forall x\?X\_ \varphi(x)$ & Für alle $x : X$ stimmt jeweils~$\varphi(x).$ &
+      Wir können (gleichmäßig) für alle~$x : X$ Belege für~$\varphi(x)$ konstruieren. \\
+    $\exists x\?X\_ \varphi(x)$ & \raggedright Es gibt mindestens ein~$x : X$, für das~$\varphi(x)$
+    stimmt. & {\raggedright
+      Wir haben ein~$x : X$ zusammen mit Beleg für~$\varphi(x).$} \\
+    \bottomrule
+  \end{tabular}
+  \caption{\label{bhk}Informale rekursive Definition des Belegbegriffs.}
+\end{table}
+
+\begin{ex}
+According to the Brouwer--Heyting--Kolmogorov interpretation, the principle of
+excluded middle states that there is a rule for producing, for every~$\varphi$,
+either a witness for~$\varphi$ or a witness for~$\neg\varphi$. The BHK
+interpretation does not specify which kind of rule ...
+Unter dieser Interpretation meint das Prinzip vom ausgeschlossenen Dritten, dass wir für jede
+Aussage Beleg für sie oder ihre Negation haben. Das ist aber offensichtlich
+nicht der Fall.
+\end{ex}
+
+%\begin{bsp}
+Die Interpretation von~$\neg\neg\varphi$ ist, dass es keinen Beleg
+für~$\neg\varphi$ gibt. Daraus folgt natürlich noch nicht, dass wir tatsächlich
+Beleg für~$\varphi$ haben; gewissermaßen ist eine solche Aussage~$\varphi$ nur
+"`potenziell wahr"'.
+%\end{bsp}
+
+%\begin{bsp}
+Wenn wir wissen, dass sich unser Haustürschlüssel irgendwo in der
+Wohnung befinden muss (da wir ihn letzte Nacht verwendet haben, um die Tür
+aufzusperren), wir ihn momentan aber nicht finden, so können wir konstruktiv
+nur folgende doppelt negierte Aussage vertreten:
+\[ \neg\neg (\exists x\_ \text{der Schlüssel befindet sich an Position~$x$})
+\]
+%\end{bsp}
+
+%\begin{bsp}
+Wir stehen im Supermarkt und erinnern uns, dass wir unbedingt
+gewisse Zutaten einkaufen müssen. Leider fällt uns momentan keine einzige der
+Zutaten mehr ein. Dann können wir zwar die Aussage, dass die Menge der zu
+besorgenden Zutaten nicht leer ist, vertreten, nicht jedoch die stärkere
+Aussage, dass diese Menge ein Element enthält.
+%\end{bsp}
+
+%\begin{bsp}[\cite{sigfpe:katemoss,bbc:katemoss}]
+Es war ein Video aufgetaucht, dass Kate Moss beim Konsumieren von Drogen zeigte,
+und zwar entweder solche von einem Typ~A oder solche von einem Typ~B. Welcher
+Typ aber tatsächlich vorlag, konnte nicht entschieden werden. Also gab es für
+keine der beiden Straftaten einen Beleg, Kate Moss wurde daher nicht
+strafrechtlich verfolgt.
+%\end{bsp}
+
+%\begin{bem}Die Formulierung mit dem generischen \emph{Wir} in Tafel~\ref{bhk}
+ist etwas irreführend. Wie auch in klassischer Mathematik hängen
+intuitionistische Urteile nicht von \emph{uns} oder anderen Mathematikern ab.
+Aussagen, die bisher noch nicht konstruktiv bewiesen wurden, können
+durchaus einen (noch unbekannten) Beleg besitzen. Eine genauere Diskussion
+findet sich etwa in~\cite[Seite~42f.]{kohlenbach:applprooftheory}. Die
+Brouwer--Heyting--Kolmogorov-Interpretation kann im Rahmen der
+\emph{Realisierbarkeitstheorie} formalisiert werden~\cite{bauer:realizability};
+ohne eine solche Formalisierung gibt es gewisse weitere
+Probleme~\cite{artemov:bhk,sanz:piecha:critical-bhk,dalen:bhk}.
+%\end{bem}
+
+\subsubsection*{\ldots{} über Berechenbarkeit}
+
+Es gibt noch eine zweite Interpretation, die beim Verständnis der konstruktiven
+Sichtweise sehr hilfreich ist:
+%\listedmotto{konstruktiv-gdw-computer}{Eine Aussage gilt genau dann konstruktiv, wenn es ein
+%Computerprogramm gibt, welches sie in endlicher Zeit bezeugt.}
+Etwa ist mit dieser Interpretation klar, dass die formale Aussage
+\[ \forall n \in \NN\_ \exists p \geq n\_ \text{$p$ ist eine Primzahl}, \]
+eine Formulierung der Unendlichkeit der Primzahlen, auch konstruktiv
+stimmt: Denn man kann leicht ein Computerprogramm angeben, das eine natürliche
+Zahl~$n$ als Eingabe erwartet und dann, etwa über die Sieb-Methode von
+Eratosthenes, eine Primzahl~$p \geq n$ produziert (zusammen mit einem Nachweis,
+dass~$p$ tatsächlich prim ist).
+
+Das Motto kann man tatsächlich zu einem formalen Theorem
+präzisieren, das ist Gegenstand der gefeierten
+Curry--Howard-Korrespondenz (Abschnitt~\ref{curry-howard}).
+
+\section{Exercises}
 

### 2022-09-03 algar-lecture-notes.tex
Beginnings on extracting content from classical proofs

@@
 \usepackage[babel]{csquotes}
-\bibliography{bibliography}
+\bibliography{algar-lecture-notes}
 
@@
 ohne eine solche Formalisierung gibt es gewisse weitere
-Probleme~\cite{artemov:bhk,sanz:piecha:critical-bhk,dalen:bhk}.
+Probleme~\cite{artemov:bhk,sanz-piecha:critical-bhk,dalen:bhk}.
 %\end{bem}
@@
 instances of the principle of omniscience for searchable sets
-like~$\NN_\infty$~\cite{escardo:omniscient1,escardo:omniscient2}.\par}
+like~$\NN_\infty$~\cite{escardo:omniscience1,escardo:omniscience2,escardo:omniscience}.\par}
 \end{enumerate}
@@
 these matters. One such account (though assuming the axiom of dependent choice)
-is~\cite{bridges}. However, this exercise is insightful even when carried out
+is~\cite{bishop-bridges:bible}. However, this exercise is insightful even when carried out
 slightly informally. Keep in mind that, to show that a real number is positive,
@@
 
+\begin{thm}[Dickson's lemma]\label{thm:dickson}
+Let~$k \in \NN$. Let~$f : \NN \to \NN^k$ be an arbitrary map. Then there are
+indices~$i < j$ such that~$f(i) \leq f(j)$ (componentwise).
+\end{thm}
+
+\begin{proof}[Proof (classical)]The case~$k = 0$ is trivial and we omit the cases~$k
+\geq 2$, hence let~$k = 1$. In this case, the map~$f$ attains some minimal
+value. Set~$i$ to be (one of the) positions where this minimal value is
+attained. Set~$j \defeq i+1$. Then, trivially, $f(i) \leq f(j)$.
+\end{proof}
+
+\begin{thm}\label{thm:surjective-matrix}
+Let~$M$ be a surjective matrix with more rows than columns over a commutative
+ring~$A$ with unit. Then~$1 = 0$ in~$A$.
+\end{thm}
+
+\begin{proof}[Proof (classical)]Assume not. Then there is a maximal
+ideal~$\mmm \subseteq A$. The matrix~$M$ remains surjective when considered over
+the quotient ring~$A/\mmm$, and by maximality this quotient ring is a field.
+Hence we have a contradiction to basic linear algebra, namely to the basic fact
+that matrices over fields are not surjective if they have more rows than
+columns.\end{proof}
+
+What is the meaning of these non-effective proofs? Theorem~\ref{thm:dickson}
+claims the existence of a finite object with a decidable property (a
+pair~$(i,j)$ such that~$f(i) \leq f(j)$), but the given proof employs
+transfinite methods and gives no indication how we could compute or otherwise
+find this object. Instead, the classical proof asks us to grasp the infinitude
+of all values of~$f$ so we can determine the minimum. The issue is even more
+pronounced with Theorem~\ref{thm:surjective-matrix}, since the existence of
+maximal ideals in nontrivial rings requires the axiom of
+choice~\cite{scott:prime-ideals,hodges:krull,banaschewski:krull,erne:krull,howard-rubin:ac}.
+To add one more conundrum: Assume that we have a classical proof, using the law
+of excluded middle and the axiom of choice, that some given Turing machine
+terminates. Can we then constructively accept that the machine will halt? Do we
+have an upper bound for the number of computational steps the machine carries
+out before halting?
+
+Astoundingly, it is almost always the case that from classical proofs useful
+constructive content can be extracted. In fact, due to general metatheorems,
+in many cases there are even explicit mechanical procedures for extracting this
+hidden content, while other cases require more creativity for determining
+suitable constructive reformulations. For instance:
+
+% \begin{document}
+
+\begin{enumerate}
+\item \emph{Eliminating the axiom of choice by the~$L$-translation.} Can the
+axiom of choice ever help in proving arithmetical statements, those first-order
+statements in which all quantifiers range over the natural numbers? Well, it
+might. But a result of Gödel states that~\textsc{zfc} (Zermelo--Fraenkel set
+theory with the axiom of choice) is conservative over~\textsc{zf} (ZF set
+theory without it) for such statements -- hence all appeals to this axiom can
+be mechanically eliminated from a given proof. This is true even if the proof
+transcends the arithmetical realm and includes statements which are not
+arithmetical, only the asserted claim is required to be arithmetical. Hence we
+are free to use the axiom of choice, tranquil in knowing that we could always
+reformulate our proofs without it.\footnote{Zermelo--Fraenkel set theory with
+the axiom of choice is the go-to foundation of mathematics often cited as
+supporting ``almost all'' of current mathematics, one important exception being
+some (definitions and) results in category theory dealing with large
+structures~\cite{shulman:set-theory,feferman:set-foundations}. A fundamental
+result due to Gödel is that the axiom of choice ``holds in~$L$'', the
+\emph{constructible universe}, even if it might not hold in~$V$, the true
+universe of all sets. More precisely, if~\textsc{zfc} shows some
+statement~$\varphi$, then~\textsc{zf} shows its~$L$-relativized
+version~$\varphi^L$, where all quantifiers have been restricted to range
+over~$L$ instead of~$V$. The conservation result follows because the natural
+numbers ``are absolute between~$V$
+and~$L$''~\cite{goedel:ac-gch,schoenfield:predicativity}.}
+
+\item \emph{Eliminating the law of excluded middle by the double-negation
+translation and its variants.} At the price of
+slightly modifying the asserted claim, the law of excluded middle
+can always be mechanically eliminated from a given proof. This
+elimination procedure is facilitated by the \emph{double-negation translation}
+reviewed below. In some cases, a refined translation even allows to preserve the asserted
+claim exactly; this technique is variously known as \emph{Friedman's trick},
+\emph{nontrivial exit continuation} or (the baby version of) \emph{Barr's
+theorem}. To cite a concrete instance of this phenomenon, classical~\textsc{zf}
+set theory is conservative over its intuitionistic cousin~\textsc{izf}
+for~$\Pi^0_2$-statements (statements of the form~$\forall\ldots\forall\_
+\exists\ldots\exists\_ \%$, where all quantifiers in the final~``$\%$''
+are bounded).
+
+\item \emph{Embracing generic models.} A useful companion to both of the
+aforementioned techniques is to switch from referencing all models of a certain
+kind to referencing only the \emph{generic model}. For instance, Krull's lemma
+stating that a ring element is already nilpotent if it is contained in all
+prime ideals requires the Boolean Prime Ideal Theorem, a slightly weaker
+version of the axiom of choice but still ineffectiv and unconstructive.
+However, Krull's lemma is valid in the form that a ring element is nilpotent if
+it is contained in the \emph{generic prime ideal}. This particular example has
+received lots of attention (see the references
+in~\cite{blechschmidt-schuster:constructive-maximal-ideals}) and the general
+technique will be the object of the fourth lecture.
+\end{enumerate}
+
 \section{The double-negation translation}
@@
 
+\printbibliography
+
 \end{document}
-
-omniscient1
-https://www.cs.bham.ac.uk/~mhe/.talks/dagstuhl2011/omniscient.pdf
-
-omniscient2
-https://www.ioc.ee/~tarmo/tsem16/escardo2605-slides.pdf

### 2022-09-03 algar-lecture-notes.tex
Short comment on predicativism

@@
 
+Noticeably missing in this list is any technique for eliminating uses of the
+\emph{powerset axiom} stating that the collection of all subsets of a given
+set is again a set. While this axiom is uncontested by ordinary constructive
+mathematics and doesn't receive nearly as much philosophical attention as the
+axiom of choice or the law of excluded middle, it is this axiom which actually
+and substantially increases logical strength. While~\textsc{zfc}, \textsc{zf}
+and \textsc{izf} are equiconsistent (and in fact verify the same
+arithmetical~$\Pi^0_2$-statements), systems without it such as Kripke--Platek
+set theory~(\textsc{kp}) or constructive Zermelo--Fraenkel set
+theory~(\textsc{czf}) are much weaker. We will not discuss this curious state
+of affairs; references include~\cite{crosilla:predicativity,aczel-rathjen:cst}.
+
 \section{The double-negation translation}

### 2022-09-03 algar-lecture-notes.tex
typo
to do

@@
 \item $(\alpha \vee \beta) \wedge \neg\alpha \ \Longrightarrow\ \beta$
-\item $\forall M \in P(X)\_ (\exists x\?X\_ x \in M) \vee M = \emptyset$
+\item $\forall M\?P(X)\_ (\exists x\?X\_ x \in M) \vee M = \emptyset$
 (already interesting for~$X = \{\star\}$)
@@
 \item $\forall f \? \NN \to \{0,1\}\_ (\neg\neg\exists n \? \NN\_ f(n) = 0)
-\Rightarrow (\exists n \? \NN\_ f(n) = 0)$
+\Rightarrow (\exists n \? \NN\_ f(n) = 0)$ (Markov's principle)
 \item $\forall f \? \NN \to \{0,1\}\_ \exists n \? \NN\_ (f(n) = 1
-\Leftrightarrow (\forall m \? \NN\_ f(m) = 1))$
+\Rightarrow (\forall m \? \NN\_ f(m) = 1))$ (Drinker's paradox)
 \item $\forall f \? \NN \to \{0,1\}\_ (\exists n \? \NN\_ f(n) = 0) \vee
@@
 {\noindent\scriptsize\emph{Remark.} The set~$\NN_\infty$ is the \emph{one-point
-compactification} of~$\NN$. A sensible definition in constructive mathematics
+compactification} of~$\NN$. A sensible definition of it in constructive mathematics
 is as the set of decreasing binary sequences~$(x_0,x_1,x_2,\ldots)$. The
@@
 \end{document}
+
+1
+Add Agda exercises
+
+2
+classical logic fairy tale
+Add Agda exercises
+Mention Dialectica
+something with minima (for negneg translation)
+Exercises: negneg proof, stability of the HA axioms, applications of Barr
+negneg drinker and CPS perspective
+
+3
+Definition
+Examples
+Language: definition, properties, IVT, FTA
+Exercises: dense, monotone IVT, ...
+
+4
+"A fantastical ring"
+The generic prime filter
+Explicit construction
+Exercises: (lift from Spiel & Spaß)

### 2022-09-03 algar-lecture-notes.tex
cover picture
More exercises

@@
 \renewcommand{\vfil}{\ }
+{\centering\includegraphics[width=0.3\textwidth]{phantoms}\par}
 \maketitle
@@
 
-\begin{exercise}[Basics on negation]
-Recalling that negation is defined as implying absurdity,~$\neg\varphi
-\defequiv (\varphi \Rightarrow \bot)$, verify intuitionistically without
-recourse to truth tables:
-\begin{enumerate}
-\item $\varphi \Rightarrow \neg\neg\varphi$
-\item $\neg\neg\neg\varphi \Leftrightarrow \neg\varphi$
-\item $\neg\neg(\varphi \vee \neg\varphi)$
-\item $(\forall\psi\_ (\psi \vee \neg\psi))
-\Longleftrightarrow (\forall\psi\_ ((\neg\neg\psi) \Rightarrow
-\psi))$
-
-{\noindent\scriptsize\emph{Hint.}
-Don't try to verify that double negation elimination for a specific
-statement~$\psi$ implies the principle of excluded middle for that same
-statement~$\psi$ -- this cannot be shown. There are subtleties regarding the
-quantification over $\psi$ (this is not expressible in pure first-order logic),
-however the exercise is still instructive if we gloss over this issue.\par}
-\end{enumerate}
-\end{exercise}
-
 \begin{exercise}[Constructive status of classical tautologies]
@@
 
+\begin{exercise}[Basics on negation]
+Recalling that negation is defined as implying absurdity,~$\neg\varphi
+\defequiv (\varphi \Rightarrow \bot)$, verify intuitionistically without
+recourse to truth tables:
+\begin{enumerate}
+\item $\varphi \Rightarrow \neg\neg\varphi$
+\item $\neg\neg\neg\varphi \Leftrightarrow \neg\varphi$
+\item $\neg\neg(\varphi \vee \neg\varphi)$
+\item $(\forall\psi\_ (\psi \vee \neg\psi))
+\Longleftrightarrow (\forall\psi\_ ((\neg\neg\psi) \Rightarrow
+\psi))$
+
+{\noindent\scriptsize\emph{Hint.}
+Don't try to verify that double negation elimination for a specific
+statement~$\psi$ implies the principle of excluded middle for that same
+statement~$\psi$ -- this cannot be shown. There are subtleties regarding the
+quantification over $\psi$ (this is not expressible in pure first-order logic),
+however the exercise is still instructive if we gloss over this issue.\par}
+\end{enumerate}
+\end{exercise}
+
 \begin{exercise}[An epistemic riddle on transcendental numbers]
@@
 numbers~$e + \pi$ and~$e \cdot \pi$ is transcendental.
+\end{exercise}
+
+\begin{exercise}[A stronger form of the irrationality of~$\sqrt{2}$]
+\label{ex:sqrt2}
+Mine the proof of Proposition~\ref{prop:sqrt2} to give an intuitionistic proof
+that for every rational number~$x$, the distance~$|\sqrt{2}-x|$ is positive.
+
+{\noindent\scriptsize\emph{Note.} Strictly speaking, this exercise presupposes
+familiarity with an intuitionistic account of the basics of undergraduate real
+analysis. Without it, one cannot really be expected to precisely think about
+these matters. One such account (though assuming the axiom of dependent choice)
+is~\cite{bishop-bridges:bible}. However, this exercise is insightful even when carried out
+slightly informally. Keep in mind that, to show that a real number is positive,
+constructively it is not enough to merely verify that it cannot be zero or
+negative. A safe way to verify that a real number~$a$ is positive is to exhibit
+a rational number~$b$ such that~$a \geq b > 0$.\par}
 \end{exercise}
@@
 \end{enumerate}
-\end{exercise}
-
-\begin{exercise}[A stronger form of the irrationality of~$\sqrt{2}$]
-\label{ex:sqrt2}
-Mine the proof of Proposition~\ref{prop:sqrt2} to give an intuitionistic proof
-that for every rational number~$x$, the distance~$|\sqrt{2}-x|$ is positive.
-
-{\noindent\scriptsize\emph{Note.} Strictly speaking, this exercise presupposes
-familiarity with an intuitionistic account of the basics of undergraduate real
-analysis. Without it, one cannot really be expected to precisely think about
-these matters. One such account (though assuming the axiom of dependent choice)
-is~\cite{bishop-bridges:bible}. However, this exercise is insightful even when carried out
-slightly informally. Keep in mind that, to show that a real number is positive,
-constructively it is not enough to merely verify that it cannot be zero or
-negative. A safe way to verify that a real number~$a$ is positive is to exhibit
-a rational number~$b$ such that~$a \geq b > 0$.\par}
 \end{exercise}
@@
 
+Even though Barr's theorem has wide impact, there are definitive situations in
+which, provably so, no useful constructive content can be extracted.
+Exercise~\ref{ex:no-constructive-content} gives an example for this situation.
+
 \section{Examples from quadratic form theory}
+
+\section{Exercises}
+
+\begin{exercise}[Drinker's paradox]
+The Drinker's paradox is the tautology
+\[ \forall f \? \NN \to \{0,1\}\_ \exists n \? \NN\_ (f(n) = 1
+\Rightarrow (\forall m \? \NN\_ f(m) = 1)) \]
+of classical logic. A proof proceeds as follows: By the law of
+excluded middle, either there is a number~$n$ such that~$f(n) = 0$ or not. In
+the first case, we can take such a number~$n$ as the desired~$n$. In the second
+case, we can take~$n \defeq 0$.
+\begin{enumerate}
+\item Determine the double-negation translation of the Drinker's paradox.
+\item Tell a classical logic fairy tale for Drinker's paradox similar to the
+story for Dickson's lemma. The protagonist of the story will change their mind
+regarding the correct value of~$n$; what is their first choice?
+\item Connect the Drinker's paradox to the issue of minima of sets of natural
+numbers.
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}[Stability of the axioms]
+Peano arithmetic (\textsc{pa}) is set in the language~$(0,S,+,\cdot)$ and has
+the following axioms (where leading universal quantifiers are supressed for
+brevity):
+\begin{enumerate}
+\renewcommand{\theenumi}{\arabic{enumi}}
+\item $Sx \neq 0$
+\item $Sx = Sy \Rightarrow x = y$
+\item $y = 0 \vee (\exists x\_ y = Sx)$
+\item $x + 0 = x$
+\item $x + Sy = S(x+y)$
+\item $x \cdot 0 = 0$
+\item $x \cdot Sy = (x \cdot y) + x$
+\item $P(0) \wedge (\forall n\_ P(n) \Rightarrow P(Sn)) \Longrightarrow
+(\forall n\_ P(n))$ (one axiom for each formula~$P(n)$)
+\end{enumerate}
+Heyting arithmetic~(\textsc{ha}) has exactly the same axioms, but is based on intuitionistic
+logic instead of classical logic.
+\begin{enumerate}
+\item Show that~\textsc{ha} proves the double-negation translation of each
+axiom of~\textsc{pa}.
+\item Convince yourself that~\textsc{izf} does not prove the double-negation
+translation of the axiom of choice. Hence the double-negation translation
+alone is insufficient to extract constructive content from proofs using the
+axiom of choice.
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}[Details on the double-negation translation]
+Complete the verification of the fundamental properties of the double-negation
+translation.
+\end{exercise}
+
+\begin{exercise}[No constructive content]\label{ex:no-constructive-content}
+Let~$\textsf{Prf}(p)$ be a formula of arithmetic expressing that~$p$ is a
+correct encoding of a~\textsc{pa}-proof of~$\bot$. A consequence of Gödel's
+second incompleteness theorem is that there is no~\textsc{pa}-proof of
+\[ G \defequiv (\forall p\_ \neg\textsf{Prf}(p)), \]
+even though for each number~$p_0$ it is actually the case that (and~\textsc{pa}
+can verify that)~$p_0$ does not constitute a correct encoding of
+a~\textsc{pa}-proof of~$\bot$.
+\begin{enumerate}
+\item Give a~\textsc{pa}-proof, using the law of excluded middle, of the
+statement
+\[ \exists q\_ (\textsf{Prf}(q) \vee G). \]
+\item Show that for no number~$q_0 \in \NN$, the
+statement~``$\textsf{Prf}(\underline{q_0})
+\vee G$'' admits a \textsc{pa}-proof. In this sense no witness can be
+extracted from the classical proof in~(a).
+\end{enumerate}
+\end{exercise}
 
@@
 something with minima (for negneg translation)
-Exercises: negneg proof, stability of the HA axioms, applications of Barr
-negneg drinker and CPS perspective
+Exercises: applications of Barr
 

### 2022-09-03 algar-lecture-notes.tex
More exercises

@@
 \renewcommand{\vfil}{\ }
-{\centering\includegraphics[width=0.3\textwidth]{phantoms}\par}
+{\vspace*{-6em}\centering\includegraphics[width=0.3\textwidth]{phantoms}\par}
+\thispagestyle{empty}\enlargethispage{7em}
 \maketitle
@@
 of choice (which in presence of other common set-theoretical axioms implies the
-principle of excluded middle, see Exercise~\ref{ex:diaconescu}).
+principle of excluded middle, see Exercise~\ref{ex:diaconescu}). As a
+consequence, we cannot generally reason by contradiction in constructive mathematics.
 
@@
 In contrast, arguments of the second kind are fine from an intuitionistic point
-of view: They are not proper proofs by contradiction, but proofs of negated
+of view: They do not constitute proper proofs by contradiction, but instead are proofs of negated
 statements. That such proofs are valid follows directly from the definition of
-negation (which texts on classical logic often also adopt) as a certain
-implication:
+negation as a certain
+implication (which, incidentally, texts on classical logic often also adopt):
 \[ \neg\psi \defequiv (\psi \Rightarrow \bot), \]
@@
 
-The two proofs of the following fact from number theory demonstrate the
+The two proofs below of the following fact from number theory demonstrate the
 difference:
@@
 contradicts the uniqueness of the prime factor decomposition, since the
-factor~$2$ occurs an odd number of times on the left, but an even number of
+factor~$2$ occurs an odd number of times on the left but an even number of
 times on the right.
@@
 and~$b$, \ldots, a contradiction.
-[The theorem on the uniqueness of the prime factor decomposition admits an
+[The requisite theorem on the uniqueness of the prime factor decomposition admits an
 intuitionistic proof.]
@@
 elaborated by the \emph{Brouwer--Heyting--Kolmogorov interpretation} sketched
-below. It is informal and philosophical in nature, but still a useful guide to
+below. It is informal and philosophical in nature and not without
+issues~\cite{artemov:bhk,sanz-piecha:critical-bhk,dalen:bhk}, but still a useful guide to
 the constructive meaning of mathematical statements.
@@
 
-The notion of witnesses is compositional in nature and begins with witnesses for \emph{atomic
+The notion of witnesses is compositional in nature and starts out with witnesses for \emph{atomic
 statements}, those not built from substatements using the logical
@@
 \[ \forall n \? \NN\_ \bigl(\varphi(n) \Rightarrow \psi(n)\bigr) \]
-is a rule explaining how, for any natural number~$n : \NN$, witnesses
+is a rule explaining how, for every natural number~$n : \NN$, witnesses
 for~$\varphi(n)$ give rise to witnesses for~$\psi(n)$.
@@
 
-\begin{exercise}[Constructive status of classical tautologies]
+\begin{exercise}[Constructive status of classical tautologies]\label{ex:tautologies}
 Which of the following classical tautologies can reasonably be expected to
@@
 \item $\neg(\alpha \wedge \beta) \ \Longrightarrow\ \neg\alpha \vee \neg\beta$
+\item $(\alpha \Rightarrow \beta) \ \Longrightarrow\ (\neg\alpha \vee \beta)$
 \item $(\alpha \vee \beta) \wedge \neg\alpha \ \Longrightarrow\ \beta$
@@
 \item $\neg\neg(\varphi \vee \neg\varphi)$
+\item $\neg\neg(\alpha \wedge \beta) \Leftrightarrow (\neg\neg\alpha \wedge
+\neg\neg\beta)$
+\item $(\neg\neg\alpha \wedge (\alpha \Rightarrow \neg\neg\beta)) \Longrightarrow
+\neg\neg\beta$
 \item $(\forall\psi\_ (\psi \vee \neg\psi))
@@
 
+\begin{exercise}[Brouwerian counterexamples]
+Show that each of the following statements implies the law of excluded middle,
+hence is not available in constructive mathematics.
+\begin{enumerate}
+\item Every ideal of~$\ZZ$ is finitely generated.
+
+{\noindent\scriptsize\emph{Hint.} Use that finitely generated ideals of~$\ZZ$
+are principal ideals and consider the ideal~$\aaa \defeq \{ x \in \ZZ
+\,|\, x = 0 \vee \varphi \}$.\par}
+
+{\noindent\scriptsize\emph{Remark.} The failure of every ideal of~$\ZZ$ to be
+finitely generated should not be misconstrued to exclaim that in constructive
+mathematics, there suddenly would be ideals of~$\ZZ$ of infinite rank. The
+failure is simply because, given an abstract ideal, we cannot pinpoint a finite
+system of generators.\par}
+
+\item Over every field, the polynomial~$X^2 + 1$ is either reducible or
+irreducible.
+
+{\noindent\scriptsize\emph{Hint.} Consider the field~$K \defeq \{ z \in \QQ(i)
+\,|\, z \in \QQ \vee \varphi \}$.\par}
+
+\item Subsets of Kuratowski-finite sets are Kuratowski-finite.
+
+{\noindent\scriptsize\emph{Note.} A set~$X$ is \emph{Kuratowski-finite} if and
+only if, for some number~$n \in \NN$, there is a surjective map~$[n] \to X$,
+where~$[n] = \{ 0,1,\ldots,n-1 \}$. More briefly, a set~$X$ is
+Kuratowski-finite iff its elements can be enumerated: $X = \{x_1,\ldots,x_n \}$.\par}
+\end{enumerate}
+Show that the following statement implies Markov's principle (from
+Exercise~\ref{ex:tautologies}):
+\begin{enumerate}
+\addtocounter{enumi}{3}
+\item Every real number which is not zero is invertible.
+\end{enumerate}
+\noindent
+More Brouwerian counterexamples can be
+found in~\cite{mandelkern:brouwerian-counterexamples} and
+in~\cite{mines-richman-ruitenburg:constructive-algebra}.
+\end{exercise}
+
 \begin{exercise}[Diaconescu's theorem]
@@
 section.'' (A \emph{section}~$s$ to a surjective map~$f$ is a map in the other direction such that~$f \circ s =
-\mathrm{id}$.) It is a theorem due to Diaconescu that
+\mathrm{id}$.) A theorem of Diaconescu states that
 the axiom of choice implies the law of excluded middle. To this end, let~$\varphi$ be a
@@
 suitable constructive reformulations. For instance:
-
-% \begin{document}
 
@@
 
+\chapter{Toposes and their internal language}
+
+\section{Examples}
+\newpage
+
+\section{Definition}
+\newpage
+
+\section{The Kripke--Joyal semantics of sheaf toposes}
+\newpage
+
+\section{Exercises}
+
 \chapter{Generic models and their applications}
@@
 \end{intro}
+
+\section{A fantastical ring}
+\newpage
+
+\section{The generic prime filter}
+\newpage
+
+\section{Explicit construction}
+\newpage
+
+\section{Exercises}
 
@@
 Exercises: (lift from Spiel & Spaß)
+
+Cite surveys of .bib

### 2022-09-03 algar-lecture-notes.tex
BHK rules

@@
   \renewcommand{\arraystretch}{1.3}
-  \begin{tabular}{@{}rp{5.9cm}p{6.5cm}@{}}
+  \begin{tabular}{@{}rp{4.3cm}p{5.3cm}@{}}
     \toprule
-    & {klassische Logik} & {intuitionistische Logik}
+    & {classical logic} & {intuitionistic logic}
     \\\midrule
-    Aussage $\varphi$ & Die Aussage $\varphi$ gilt. & Wir haben Beleg für $\varphi$. \\
-    $\bot$ & Es stimmt Falschheit. & Wir haben Beleg für Falschheit. \\
-    $\varphi \wedge \psi$ & $\varphi$ und $\psi$ stimmen. & Wir haben Beleg für~$\varphi$ und für~$\psi$. \\
-    $\varphi \vee \psi$ & $\varphi$ oder $\psi$ stimmt. & Wir haben Beleg für~$\varphi$ oder für~$\psi$. \\
-    $\varphi \Rightarrow \psi$ & Sollte~$\varphi$ stimmen, dann auch~$\psi$. &
-    Aus Belegen für~$\varphi$ können wir (gleich\-mä\-ßig) Belege für~$\psi$ konstruieren. \\
+    statement $\varphi$ & The statement $\varphi$ holds. & We have a witness for $\varphi$. \\
+    $\bot$ & A contradiction holds. & We have a witness for a contradiction. \\
+    $\varphi \wedge \psi$ & $\varphi$ and $\psi$ hold. & We have a witness for~$\varphi$ and for~$\psi$. \\
+    $\varphi \vee \psi$ & $\varphi$ or $\psi$ hold. & We have a witness for~$\varphi$ or for~$\psi$. \\
+    $\varphi \Rightarrow \psi$ & If~$\varphi$ holds, then so does~$\psi$. &
+    We can (uniformly) construct witnesses for~$\psi$ from witnesses
+    for~$\varphi$. \\
     $\neg\varphi$ &
-      $\varphi$ stimmt nicht. &
-      Es kann keinen Beleg für~$\varphi$ geben. \\
-    $\forall x\?X\_ \varphi(x)$ & Für alle $x : X$ stimmt jeweils~$\varphi(x).$ &
-      Wir können (gleichmäßig) für alle~$x : X$ Belege für~$\varphi(x)$ konstruieren. \\
-    $\exists x\?X\_ \varphi(x)$ & \raggedright Es gibt mindestens ein~$x : X$, für das~$\varphi(x)$
-    stimmt. & {\raggedright
-      Wir haben ein~$x : X$ zusammen mit Beleg für~$\varphi(x).$} \\
+      $\varphi$ does not hold. &
+      There is no witness for~$\varphi$. \\
+    $\forall x\?X\_ \varphi(x)$ & For all $x \? X$ it holds that~$\varphi(x).$ &
+      We can (uniformly) construct, for all~$x \? X$, witnesses for~$\varphi(x)$. \\
+    $\exists x\?X\_ \varphi(x)$ & \raggedright There is at least one~$x \? X$
+    such that~$\varphi(x)$ holds. & {\raggedright
+      We have a~$x \? X$ together with a witness for~$\varphi(x)$.} \\
     \bottomrule

### 2022-09-04 algar-lecture-notes.tex
BHK, computability and geometric interpretation

@@
 \title{The curious world of constructive mathematics}
-\author{Ingo Blechschmidt}
+\author{Ingo Blechschmidt \vspace*{-2em}}
 %\email{iblech@speicherleck.de}
@@
 
+\urlstyle{sf}
+
 \begin{document}
@@
 
-Relinquishing the axiom of choice and the law of excluded middle also
+Relinquishing the axiom of choice and the principle of excluded middle also
 allows us to explore axioms and notions which are incompatible with
@@
 \renewcommand{\vfil}{\ }
+\enlargethispage{7em}
 {\vspace*{-6em}\centering\includegraphics[width=0.3\textwidth]{phantoms}\par}
-\thispagestyle{empty}\enlargethispage{7em}
 \maketitle
@@
 }
+\thispagestyle{empty}
 
@@
 constructively.
+
+This account is an update and translation of~\cite[Section~1]{blechschmidt:pizzaseminar}
+and references
+include~\cite{bauer:five-stages-video,bauer:five-stages-article,avigad:constructive,sep:intuitionistic-logic,dummett:basis,melikhov:intuitionistic-logic,streicher:constructive}
+or more
+specifically~\cite{mines-richman-ruitenburg:constructive-algebra,lombardi-quitte:constructive-algebra}
+for constructive algebra and~\cite{bishop-bridges:bible} for constructive
+analysis.
 \end{intro}
@@
 of choice (which in presence of other common set-theoretical axioms implies the
-principle of excluded middle, see Exercise~\ref{ex:diaconescu}). As a
-consequence, we cannot generally reason by contradiction in constructive mathematics.
+principle of excluded middle, see Exercise~\ref{ex:diaconescu}) and from Zorn's lemma. As a
+consequence, we cannot generally reason by contradiction in constructive
+mathematics, and to demonstrate the existence of an object
+it is not enough that its nonexistence would entail a contradiction.
 
-In constructive mathematics, we do \emph{not} claim that the principle of
+Importantly, in constructive mathematicics we do \emph{not} claim that the principle of
 excluded middle is false. \marginpar{\dbend}
@@
 with classical logic (every intuitionistic proof is a fortiori also a classical
-proof), and some special instances of the principle of excluded middle are
+proof), and some special instances of the principle of excluded middle and with
+it some special instances of proof by contradiction are
 intuitionistically verifiable (an example is given in
@@
 A rumor about constructive mathematics states that constructively, the term
-``contradiction'' would be generally forbidden. This is false. In fact, we need
+``contradiction'' would be generally forbidden. This rumor is false. In fact, we need
 to distinguish between two distinct figures of proof which are often conflated
@@
 Arguments of the first form are proper proofs by contradiction and hence not
-generally acceptable in constructive mathematics. What is established is
+generally accepted in constructive mathematics. What they establish is
 only~$\neg(\neg\varphi)$, the impossibility of~$\neg\varphi$; constructively,
@@
 
-\subsection*{The Brouwer--Heyting--Kolmogorov interpretation}
+\subsection{The Brouwer--Heyting--Kolmogorov interpretation}
 
-The notion of witnesses is compositional in nature and starts out with witnesses for \emph{atomic
+The notion of witnesses is compositional in nature and starts out with the notion of witnesses for \emph{atomic
 statements}, those not built from substatements using the logical
@@
 
-Table~\ref{table:bhk} explains what witnesses of compound statements are. For
+Table~\ref{table:bhk} explains how compound statements should be witnessed. For
 instance, a witness of a statement of the form
@@
   \end{tabular}
-  \caption{\label{bhk}Informale rekursive Definition des Belegbegriffs.}
+  \caption{\label{table:bhk}Informal recursive definition of the notion of witnesses.}
 \end{table}
@@
 According to the Brouwer--Heyting--Kolmogorov interpretation, the principle of
-excluded middle states that there is a rule for producing, for every~$\varphi$,
-either a witness for~$\varphi$ or a witness for~$\neg\varphi$. The BHK
-interpretation does not specify which kind of rule ...
-Unter dieser Interpretation meint das Prinzip vom ausgeschlossenen Dritten, dass wir für jede
-Aussage Beleg für sie oder ihre Negation haben. Das ist aber offensichtlich
-nicht der Fall.
+excluded middle states that we can construct, for every~$\varphi$,
+either a witness for~$\varphi$ or a witness for~$\neg\varphi$. This claim is
+obviously false.
 \end{ex}
 
-%\begin{bsp}
-Die Interpretation von~$\neg\neg\varphi$ ist, dass es keinen Beleg
-für~$\neg\varphi$ gibt. Daraus folgt natürlich noch nicht, dass wir tatsächlich
-Beleg für~$\varphi$ haben; gewissermaßen ist eine solche Aussage~$\varphi$ nur
-"`potenziell wahr"'.
-%\end{bsp}
+\begin{ex}
+The BHK interpretation of a doubly negated statement~$\neg\neg\varphi$ is that
+there is no witness for~$\neg\varphi$. This state of affairs does not entail
+that we actually have a witness for~$\varphi$; in a sense, the
+statement~$\varphi$ is only ``potentially true''.
+\end{ex}
 
-%\begin{bsp}
-Wenn wir wissen, dass sich unser Haustürschlüssel irgendwo in der
-Wohnung befinden muss (da wir ihn letzte Nacht verwendet haben, um die Tür
-aufzusperren), wir ihn momentan aber nicht finden, so können wir konstruktiv
-nur folgende doppelt negierte Aussage vertreten:
-\[ \neg\neg (\exists x\_ \text{der Schlüssel befindet sich an Position~$x$})
-\]
-%\end{bsp}
+\begin{ex}
+Assuming that we cannot find our apartment key, but know that it has to be
+somewhere (because we used it last night to unlock the door), constructively we
+can only defend the doubly negated statement
+\[ \neg\neg (\exists x\_ \text{the key is at position~$x$}). \]
+\end{ex}
 
-%\begin{bsp}
-Wir stehen im Supermarkt und erinnern uns, dass wir unbedingt
-gewisse Zutaten einkaufen müssen. Leider fällt uns momentan keine einzige der
-Zutaten mehr ein. Dann können wir zwar die Aussage, dass die Menge der zu
-besorgenden Zutaten nicht leer ist, vertreten, nicht jedoch die stärkere
-Aussage, dass diese Menge ein Element enthält.
-%\end{bsp}
+\begin{ex}
+We are running errands and are remembering that we need to fetch certain
+ingredients. Unfortunately we don't recall any of them right now. Then we can
+constructively defend the statement that the set of ingredients to fetch is not
+empty, not however the stronger statement that this set is inhabited.
+\end{ex}
 
-%\begin{bsp}[\cite{sigfpe:katemoss,bbc:katemoss}]
-Es war ein Video aufgetaucht, dass Kate Moss beim Konsumieren von Drogen zeigte,
-und zwar entweder solche von einem Typ~A oder solche von einem Typ~B. Welcher
-Typ aber tatsächlich vorlag, konnte nicht entschieden werden. Also gab es für
-keine der beiden Straftaten einen Beleg, Kate Moss wurde daher nicht
-strafrechtlich verfolgt.
-%\end{bsp}
+\begin{ex}[\cite{sigfpe:katemoss,bbc:katemoss}]
+A video surfaced depicting Kate Moss taking drugs, more precisely either drugs
+of some type~A or drugs of some type~B. However from the video it wasn't clear
+which type of drug it actually was. Hence there was no evidence for either type
+of consumption; Kate Moss wasn't prosecuted.
+\end{ex}
 
-%\begin{bem}Die Formulierung mit dem generischen \emph{Wir} in Tafel~\ref{bhk}
-ist etwas irreführend. Wie auch in klassischer Mathematik hängen
-intuitionistische Urteile nicht von \emph{uns} oder anderen Mathematikern ab.
-Aussagen, die bisher noch nicht konstruktiv bewiesen wurden, können
-durchaus einen (noch unbekannten) Beleg besitzen. Eine genauere Diskussion
-findet sich etwa in~\cite[Seite~42f.]{kohlenbach:applprooftheory}. Die
-Brouwer--Heyting--Kolmogorov-Interpretation kann im Rahmen der
-\emph{Realisierbarkeitstheorie} formalisiert werden~\cite{bauer:realizability};
-ohne eine solche Formalisierung gibt es gewisse weitere
-Probleme~\cite{artemov:bhk,sanz-piecha:critical-bhk,dalen:bhk}.
-%\end{bem}
+\begin{rem}The reference of an unspecified ``we'' in table~\ref{table:bhk}
+renders our account of the BHK interpretation not only informal, but is also
+somewhat misleading. As in classical mathematics, judgments of intuitionistic
+logic do not actually depend on you, me or other mathematicians. It's perfectly
+possible that statements which haven't yet been constructively verified admit
+(as yet unknown) witnesses. More details can be found
+in~\cite[pp.~42]{kohlenbach:applprooftheory} and
+in~\cite{artemov:bhk,sanz-piecha:critical-bhk,dalen:bhk}. The
+Brouwer--Heyting--Kolmogorov interpretation is made more formal and precise by
+\emph{realizability theory}~\cite{bauer:realizability,oosten:realizability}.
+\end{rem}
 
-\subsubsection*{\ldots{} über Berechenbarkeit}
+\subsection{The computability interpretation}
 
-Es gibt noch eine zweite Interpretation, die beim Verständnis der konstruktiven
-Sichtweise sehr hilfreich ist:
-%\listedmotto{konstruktiv-gdw-computer}{Eine Aussage gilt genau dann konstruktiv, wenn es ein
-%Computerprogramm gibt, welches sie in endlicher Zeit bezeugt.}
-Etwa ist mit dieser Interpretation klar, dass die formale Aussage
-\[ \forall n \in \NN\_ \exists p \geq n\_ \text{$p$ ist eine Primzahl}, \]
-eine Formulierung der Unendlichkeit der Primzahlen, auch konstruktiv
-stimmt: Denn man kann leicht ein Computerprogramm angeben, das eine natürliche
-Zahl~$n$ als Eingabe erwartet und dann, etwa über die Sieb-Methode von
-Eratosthenes, eine Primzahl~$p \geq n$ produziert (zusammen mit einem Nachweis,
-dass~$p$ tatsächlich prim ist).
+A second interpretation of mathematical statements suited for constructive
+mathematics is given by the following motto: In a certain sense, \emph{we
+accept a mathematical statement in constructive mathematics if and only if
+there is a computer program witnessing it in finite time.} For instance, that
+the statement
+\[ \forall n \? \NN\_ \exists p \? \NN\_ p \geq n \wedge \text{$p$ is prime} \]
+is valid in constructive mathematics corresponds to the fact that there is a
+computer program which reads a number~$n$ as input and then, after some
+computation, produces a prime number~$p \geq n$ as output (together with a
+witness that~$p$ is in fact greater or equal than~$n$ and that~$p$ is prime).
 
-Das Motto kann man tatsächlich zu einem formalen Theorem
-präzisieren, das ist Gegenstand der gefeierten
-Curry--Howard-Korrespondenz (Abschnitt~\ref{curry-howard}).
+For the motto to be correct, the notion of ``witnessed by a computer program''
+has to be interpreted in sufficiently generous manner, not least because
+uncountable and uncomputable structures have their place in constructive
+mathematics just as in classical mathematics. A more formal rendition is
+provided by the celebrated \emph{Curry--Howard correspondence} and
+\emph{propositions as (some) types}.
+
+\subsection{The geometric interpretation}
+
+A third interpretation of mathematical statements suited for constructive
+mathematics, and of a substantially different kind than the BHK and the
+computability interpretation, is provided by geometry. In a certain sense,
+\emph{we accept a mathematical statement in constructive mathematics if and
+only if it holds locally in continuous families}.
+
+For instance, while it is a fact of classical mathematics that every complex
+number has a square root, a fundamental observation in complex analysis is that
+such roots cannot be picked in a locally continuous manner. Hence the standard
+formulation of the fundamental theorem of algebra fails in constructive
+mathematics.\footnote{More precisely, the fundamental theorem of algebra fails
+for the complex numbers built from pairs of Dedekind reals. Constructively, the
+reals built using Dedekind cuts are not the same as the reals built using
+Cauchy sequences; in fact, the latter are a subset of the former and serve a
+different purpose. Such a bifurcation of classically equivalent notions is
+typical of constructive mathematics. The fundamental theorem of algebra is
+correct for the flavor of complex numbers built from Cauchy
+reals~\cite{ruitenburg:roots}, and also for the algebraic
+numbers~\cite[Section~VII.4]{mines-richman-ruitenburg:constructive-algebra}
+[XXX insert better references] and for complex polynomials (of either flavor)
+which are separable. Furthermore, the fundamental theorem of algebra holds in
+an approximate sense and in a multiset sense~\cite{richman:fta}, and the
+standard version only requires mild choice
+principles~\cite{bridges-richman-schuster:wcc}.} This example will be discussed
+in more detail in Section~\ref{sect:fta}.
+
+The geometric interpretation not only gives transparent explanations to the
+logical subtleties of constructive mathematics, but is in a more precise form
+also the workhorse of many modern applications of constructive mathematics to
+mathematics in general. We will hence dedicate ample of space to the geometric
+interpretation by devoting Lecture~\ref{lect:toposes} to it, and will discuss
+some of its more advanced applications in Lecture~\ref{lect:generic}.
+
+\section{Finer distinctions supported by constructive mathematics}
 
@@
 (1,\ldots,1,0,\ldots,0)$, and an element not in the image of this embedding
-is~$\infty \defeq 1^\omega = (1,1,\ldots)$. Assuming the law of excluded
+is~$\infty \defeq 1^\omega = (1,1,\ldots)$. Assuming the principle of excluded
 middle (or already weaker principles), every element of~$\NN_\infty$ is of one
@@
 \begin{exercise}[Brouwerian counterexamples]
-Show that each of the following statements implies the law of excluded middle,
+Show that each of the following statements implies the principle of excluded middle,
 hence is not available in constructive mathematics.
@@
 \noindent
-More Brouwerian counterexamples can be
-found in~\cite{mandelkern:brouwerian-counterexamples} and
-in~\cite{mines-richman-ruitenburg:constructive-algebra}.
+Brouwerian counterexamples abound in constructive mathematics; when developing
+a constructive account of a theory, they help to clearly demarcate its limits.
+As such additional Brouwerian counterexamples can be found in most texts on
+constructive mathematics, such
+as~\cite{mines-richman-ruitenburg:constructive-algebra}; a compilation mainly
+from constructive analysis can be found
+in~\cite{mandelkern:brouwerian-counterexamples}.
 \end{exercise}
@@
 \mathrm{id}$.) A theorem of Diaconescu states that
-the axiom of choice implies the law of excluded middle. To this end, let~$\varphi$ be a
+the axiom of choice implies the principle of excluded middle. To this end, let~$\varphi$ be a
 statement and consider the subsets
@@
 
-\vfill
-{\small
-\begin{verbatim}
-- What constructive mathematics is about: finer distinctions, algorithmic
-  interpretation, geometric interpretation (including teaser on why we
-  might care)
-- Proof by and proof of a contradiction: powers of irrational numbers,
-  discreteness of N, Q, Qbar, failure of discreteness of R
-- Positive information from negative? de Morgan, minimums of sets of
-  natural numbers
-- Relation to classical mathematics
-- Finite sets and the axiom of choice
-- Axiomatic freedom: all functions continuous, all functions
-  computable, generic prime ideal
-\end{verbatim}
-}
+% - What constructive mathematics is about: finer distinctions, algorithmic
+%   interpretation, geometric interpretation (including teaser on why we
+%   might care)
+% - Proof by and proof of a contradiction: powers of irrational numbers,
+%   discreteness of N, Q, Qbar, failure of discreteness of R
+% - Positive information from negative? de Morgan, minimums of sets of
+%   natural numbers
+% - Relation to classical mathematics
+% - Finite sets and the axiom of choice
+% - Axiomatic freedom: all functions continuous, all functions
+%   computable, generic prime ideal
 
 \chapter{On the constructive content of classical proofs}
+\label{lect:constructive-content}
 
@@
 in classical mathematics, even those whose proof rests on
-non-constructive principles like the axiom of choice or the law of
+non-constructive principles like the axiom of choice or the principle of
 excluded middle, have a hidden constructive core.} With a mix of
@@
 choice~\cite{scott:prime-ideals,hodges:krull,banaschewski:krull,erne:krull,howard-rubin:ac}.
-To add one more conundrum: Assume that we have a classical proof, using the law
+To add one more conundrum: Assume that we have a classical proof, using the principle
 of excluded middle and the axiom of choice, that some given Turing machine
@@
 
-\item \emph{Eliminating the law of excluded middle by the double-negation
+\item \emph{Eliminating the principle of excluded middle by the double-negation
 translation and its variants.} At the price of
-slightly modifying the asserted claim, the law of excluded middle
+slightly modifying the asserted claim, the principle of excluded middle
 can always be mechanically eliminated from a given proof. This
@@
 mathematics and doesn't receive nearly as much philosophical attention as the
-axiom of choice or the law of excluded middle, it is this axiom which actually
+axiom of choice or the principle of excluded middle, it is this axiom which actually
 and substantially increases logical strength. While~\textsc{zfc}, \textsc{zf}
@@
 \section{The double-negation translation}
+\newpage
 
 \section{Barr's theorem}
+\newpage
 
@@
 \section{Examples from quadratic form theory}
+\newpage
 
@@
 \Rightarrow (\forall m \? \NN\_ f(m) = 1)) \]
-of classical logic. A proof proceeds as follows: By the law of
+of classical logic. A proof proceeds as follows: By the principle of
 excluded middle, either there is a number~$n$ such that~$f(n) = 0$ or not. In
@@
 \begin{enumerate}
-\item Give a~\textsc{pa}-proof, using the law of excluded middle, of the
+\item Give a~\textsc{pa}-proof, using the principle of excluded middle, of the
 statement
@@
 \chapter{Toposes and their internal language}
+\label{lect:toposes}
 
@@
 \newpage
+\ \newpage
 
@@
 \chapter{Generic models and their applications}
+\label{lect:generic}
 

### 2022-09-04 algar-lecture-notes.tex
On the double-negation embedding

@@
 
+\newcommand{\cl}{\mathrm{cl}}
+\newcommand{\notnot}{\emph{not~not}}
+
 \begin{document}
@@
 \neg\neg\beta)$
+\item $\neg(\alpha \vee \beta) \Longleftrightarrow (\neg\alpha \wedge \neg\beta)$
 \item $(\neg\neg\alpha \wedge (\alpha \Rightarrow \neg\neg\beta)) \Longrightarrow
@@
 proofs with examples and presents two general metatheorems which
-elucidate proof mining for constructive content.
+elucidate proof mining for constructive content. The author learned this
+material from diverse texts and slides of Thierry Coquand; the
+reference~\cite{coquand:classical} is a good start.
 \end{intro}
@@
 
-\begin{proof}[Proof (classical)]The case~$k = 0$ is trivial and we omit the cases~$k
+\begin{proof}[Proof (classical)]The case~$k = 0$ is trivial and we omit demonstrations of the cases~$k
 \geq 2$, hence let~$k = 1$. In this case, the map~$f$ attains some minimal
@@
 find this object. Instead, the classical proof asks us to grasp the infinitude
-of all values of~$f$ so we can determine the minimum. The issue is even more
+of all values of~$f$ and determine their minimum. The issue is even more
 pronounced with Theorem~\ref{thm:surjective-matrix}, since the existence of
@@
 transcends the arithmetical realm and includes statements which are not
-arithmetical, only the asserted claim is required to be arithmetical. Hence we
+arithmetical; only the asserted claim is required to be arithmetical. Hence we
 are free to use the axiom of choice, tranquil in knowing that we could always
@@
 \emph{nontrivial exit continuation} or (the baby version of) \emph{Barr's
-theorem}. To cite a concrete instance of this phenomenon, classical~\textsc{zf}
+theorem}. To cite a specific instance of this phenomenon, classical~\textsc{zf}
 set theory is conservative over its intuitionistic cousin~\textsc{izf}
@@
 prime ideals requires the Boolean Prime Ideal Theorem, a slightly weaker
-version of the axiom of choice but still ineffectiv and unconstructive.
+version of the axiom of choice but still ineffective and unconstructive.
 However, Krull's lemma is valid in the form that a ring element is nilpotent if
@@
 and \textsc{izf} are equiconsistent (and in fact verify the same
-arithmetical~$\Pi^0_2$-statements), systems without it such as Kripke--Platek
+arithmetical~$\Pi^0_2$-statements), systems without the powerset axiom such as Kripke--Platek
 set theory~(\textsc{kp}) or constructive Zermelo--Fraenkel set
 theory~(\textsc{czf}) are much weaker. We will not discuss this curious state
-of affairs; references include~\cite{crosilla:predicativity,aczel-rathjen:cst}.
-
-\section{The double-negation translation}
-\newpage
+of affairs and only note that rejecting the powerset axiom can be
+well-motivated and is the basis of \emph{predicative mathematics};
+references include~\cite{crosilla:predicativity,aczel-rathjen:cst}.
+
+\section{The double-negation embedding of classical into constructive mathematics}
+
+The premier difference between constructive and classical mathematics is in the
+existential quantifier~``$\exists$'' and disjunction~``$\vee$''.
+Constructively, to verify an existential statement, it is not enough to verify
+the impossibility of nonexistence; and to verify a disjunction~$\alpha \vee
+\beta$, it is not enough to verify the impossibility of~$\neg\alpha \wedge
+\neg\beta$.
+
+That is not to say, however, that the more informative versions of~``$\exists$''
+and~``$\vee$'' of constructive mathematics would be in any senser ``better''
+than their classical counterparts: There are many situations in which the
+classical semantics is exactly the appropriate one. Should we somehow combine
+classical and intuitionistic logic to form a joint logic supporting
+both the informative connectives from intuitionistic logic and the ``platonic''
+connectives from classical mathematics?
+
+Perhaps surprisingly, it turns out that no such combination is necessary, for
+the classical connectives are already definable in intuitionistic logic:
+\begin{align*}
+  \exists^\cl x \? X\_ \varphi(x)
+    &\quad\defequiv\quad \neg\neg (\exists x\?X\_ \varphi(x))
+    && \text{(this is eqv. to $\neg(\forall x\?X\_ \neg\varphi(x))$)} \\
+  \alpha \vee^\cl \beta
+    &\quad\defequiv\quad \neg\neg (\alpha \vee \beta)
+    && \text{(this is eqv. to $\neg(\neg\alpha \wedge \neg\beta)$)}
+\end{align*}
+
+This observation is the starting point of the \emph{double-negation embedding}
+of classical logic into intuitionistic logic. This embedding translates any
+statement~$\varphi$ into a related statement~$\varphi^{\neg\neg}$, substituting
+all occurences of~``$\exists$'' and~``$\vee$'' by~``$\exists^\cl$''
+and~``$\vee^\cl$'' (and prefixing all atomic statements by~``$\neg\neg$'')
+as detailed in Table~\ref{table:negneg}. Crucially, while the principle of
+excluded middle cannot be verified for~``$\vee$'', the principle of excluded
+middle for~``$\vee^\cl$'' is an intuitionistic tautology:
+
+\begin{thm}For every statement~$\varphi$,
+$\neg\neg(\varphi \vee \neg\varphi)$.
+\end{thm}
+
+\begin{proof}By definition of negation, we are to prove
+\[ \underbrace{\bigl((\varphi \vee (\varphi\Rightarrow\bot)) \Longrightarrow
+\bot\bigr)}_{\equiv\vcentcolon\, \chi}
+\Longrightarrow \bot. \]
+So assume~$\chi$; we are to show~$\bot$.
+
+Hypothetically, if~$\varphi$, then in particular~$\varphi \vee
+(\varphi\Rightarrow\bot)$ and hence, by~$\chi$, $\bot$. This argument
+establishes~$\varphi \Rightarrow \bot$.
+
+Thus in particular we have~$\varphi \vee (\varphi \Rightarrow \bot)$ and hence,
+by~$\chi$ again, $\bot$.
+\end{proof}
+
+The fundamental properties of the double-negation translation are summarized by
+the following theorem.
+
+\begin{thm}\label{thm:negneg}
+For every statement~$\varphi$, \ldots
+\begin{enumerate}
+\item classical logic proves~$\varphi \Leftrightarrow \varphi^{\neg\neg}$,
+\item intuitionistic logic proves~$\neg\neg(\varphi^{\neg\neg}) \Rightarrow
+\varphi^{\neg\neg}$,
+\item (if~$\varphi$ is a geometric formula) intuitionistic logic
+proves~$\varphi^{\neg\neg} \Leftrightarrow \neg\neg\varphi$, and
+\item classical logic proves~$\varphi$ from some set~$\Gamma$ of assumptions
+if and only if intuitionistic logic proves~$\varphi^{\neg\neg}$ from the
+set~$\Gamma^{\neg\neg}$ of translated assumptions.
+\end{enumerate}
+\end{thm}
+
+\begin{proof}Instructive exercise.\end{proof}
+
+The double-negation embedding is not only conceptually pleasing, establishing
+that intuitionistic logic can serve as a common home for both classical and
+constructive mathematics, but also supplies us with a vast source of
+constructive results.
+
+\begin{ex}Classical linear algebra teaches us that every finite system of
+generators of a vector space over a residue field\footnote{Constructively, the
+notion of a field bifurcates into several nonequivalent notions: A nontrivial
+commutative ring with unit is a \ldots
+\begin{enumerate}
+\item a \emph{geometric field} or \emph{discrete field} iff
+every element is zero or
+invertible (as in~$\QQ$ or~$\overline{\QQ}$),
+\item a \emph{residue field} iff
+every element which is not invertible is zero (as in the Cauchy or Dedekind
+reals), and
+\item a \emph{field of fractions} iff
+every element which is not zero is invertible (as in the localization of the
+Cauchy or Dedekind reals at the set of non-zeros).
+\end{enumerate}These each have their
+uses~\cite{johnstone:rings-fields-and-spectra,mulvey:repr}.}
+can be thinned to a generating family in which no vector is redundant, hence to
+a basis. By slightly massaging the result obtained by the double-negation
+embedding, we obtain the constructive theorem that every finitely generated
+vector space over a residue field does \emph{not not} possess a basis.
+
+We will learn in Section~\ref{sect:generic-freeness} that this simple theorem
+immediately entails (a basic version of) \emph{Grothendieck's generic freeness
+lemma}, a result in algebraic geometry fundamental to the theory of moduli
+spaces, if applied after localizing at the ``generic prime filter'' (this
+requires passing to a more adapted topos). It is marvelous how large the impact
+of this simple theorem from undergraduate linear algebra is, if made to apply
+in arbitrary toposes by the double-negation embedding and then specializing to
+a particularly well-adapted one.\end{ex}
 
@@
 Complete the verification of the fundamental properties of the double-negation
-translation.
+translation stated as Theorem~\ref{thm:negneg}.
+\end{exercise}
+
+\begin{exercise}[Minima of sets of natural numbers]
+In Theorem~\ref{thm:min}(b) we showed that every inhabited set of natural
+numbers does \notnot contain a minimal element. Explain how this result follows
+from the double-negation embedding applied to the classical fact that every
+inhabited set of natural numbers directly contains a minimal element.
+
+{\noindent\scriptsize\emph{Hint.} It is not enough to just cite the
+double-negation embedding. The resulting constructive theorem has to be
+massaged a bit to yield Theorem~\ref{thm:min}(b).\par}
+\end{exercise}
+
+\begin{exercise}[Grothendieck's generic freeness lemma, part one]
+XXX
 \end{exercise}
@@
 
-Cite surveys of .bib
+\[
+  \forall x_1,\ldots,x_n\?V\_
+    \bigl(\forall x\?V\_ \exists a_1,\ldots,a_n\?K\_ x = \sum_i a_i x_i\bigr)
+      \Longrightarrow
+        \]

### 2022-09-05 algar-lecture-notes.tex
Decidability of equality and minima of sets of natural numbers

@@
 
-\begin{ex}
+\begin{ex}\label{ex:key}
 Assuming that we cannot find our apartment key, but know that it has to be
@@
 \emph{propositions as (some) types}.
+
+\begin{rem}Constructive mathematics is emphatically not the same as computable
+mathematics; in particular, there are statements which do admit a witness by a
+computer program but which are rejected in (most schools of) constructive
+mathematics (see Exercise~\ref{ex:markov}). The apparent tension with the motto
+presented above is resolved in its more formal treatment. In any case, the two
+subjects are closely related and their connection provides useful
+intuition.\end{rem}
 
@@
 
+By relinquishing the blanket principle of excluded middle, constructive
+mathematics allows us to make finer distinctions -- distinctions which are
+hidden in classical mathematics, but carry algorithmic and geometric
+significance. We have already seen one instance of this increased
+expressiveness in Example~\ref{ex:key}: In classical mathematics, we are
+trained to cancel double negations as soon as they arise, but constructively
+there is a substantial difference between actually having a mathematical object
+and merely knowing that its nonexistence is impossible.
+
+\subsection{Decidability of equality}
+
+A further examples pertains to the basic notion of equality. In classical
+mathematics, any given elements~$a,b$ of a set~$X$ are either equal or not --
+trivially so, by the principle of excluded middle. Constructively, the
+situation is more nuanced.
+
+Firstly, in a positive direction, we do have the following result.
+
+\begin{prop}\label{prop:discreteness}
+For every pair of natural numbers~$a$ and~$b$, either~$a = b$ or~$a \neq b$.\end{prop}
+
+\begin{proof}[Proof (constructive)]
+We spell out only the case~$b = 0$, that is, we only spell out a proof that
+every natural number~$a$ is either zero or not. To this end, we do a induction
+on~$a$.
+
+The base step~$a = 0$ is trivial, since if~$a = 0$ then also~$a = 0 \vee a \neq
+0$ (we can always weaken statements by addition additional disjuncts).
+
+In the inductive step~$a \to a + 1$, we have~$a + 1 \neq 0$ by one of the basic
+axioms of arithmetic (for a complete list, see
+Exercise~\ref{ex:stability-axioms}) and hence in particular~$a+1 = 0 \vee
+a+1\neq0$ by weakening. (This proof is one of the rare instances of a correct
+induction proof where the inductive step does not refer to the induction
+hypothesis.)
+\end{proof}
+
+As from any constructive proof, a computer program witnessing the asserted fact
+can be extracted. It will (roughly) have type $\mathsf{Nat} \times \mathsf{Nat}
+\to \mathsf{Bool}$, computing~$\mathsf{true}$ or~$\mathsf{false}$ depending on
+whether its input agree or are distinct. Because the presented proof proceeded
+by induction, the resulting program will proceed by recursion.
+
+Sets like the set of natural numbers for which equality is decidable have a
+special name in constructive mathematics, hinting at a connection with topology:
+\begin{defn}A set~$X$ is \emph{discrete} if and only if for every
+elements~$a,b \in X$, either~$a = b$ or~$a \neq b$.\end{defn}
+In classical mathematics, all sets are discrete. Constructively, discreteness
+is a nontrivial property, and although~$\NN$ is discrete and with it also~$\ZZ$
+and~$\QQ$, there are many important sets which constructively cannot be shown
+to be discrete:
+
+\begin{prop}Already in the case~$X = \{\star\}$ it holds that if the
+powerset~$P(X)$ is discrete, then the principle of excluded middle
+holds.\end{prop}
+
+\begin{proof}Let~$\varphi$ be a statement and consider the element~$K_\varphi
+\defeq \{ x \in X \,|\, \varphi \} \in P(X)$. If~$K_\varphi = X$,
+then~$\varphi$; and if~$K_\varphi \neq X$, then~$\neg\varphi$.\end{proof}
+
+\begin{prop}If the set of (any of the usual flavors of the) reals is discrete,
+then the principle of omniscience holds (XXX this has not been
+introduced).\end{prop}
+
+The failure of the discreteness of the set of real numbers can be explained in
+computational terms as follows. In computable mathematics, a real numbers is
+represented by a program computing better and better rational approximations.
+Given two such programs, we can simulate or execute them to obtain
+approximations to a desired precision, hoping to distinguish the two
+represented real numbers. However, in case that the two numbers actually agree,
+in finite time we will never verify this fact: Computably, we cannot rule out
+the possibility that a difference between the two numbers will only be detected
+in the as yet unexplored range of rational approximations.\footnote{This
+argument can be formalized using Turing machines and realizability theory;
+see~\cite[Section~4.2.1]{blechschmidt:filmat} for a review. The analysis
+dramatically changes (see~\cite[Section~4.2.2]{blechschmidt:filmat}) if we
+instead refer to the infinite-time Turing machines of Hamkins and
+Lewis~\cite{hamkins-lewis:ittm} which can carry out an infinite amount of
+computation before halting. A philosophically intriguing (and, by necessity,
+informal) account referring to machines in the real world as opposed to
+idealized Turing machines has been put forward by Andrej
+Bauer~\cite{bauer:int-mathematics}.}
+
+Lest an impression that only topologically simple domains can be shown to be
+discrete in constructive mathematics emerges, here is a more sophisticated
+example of a discrete set.
+
+\begin{prop}The set~$\overline{\QQ}$ of algebraic numbers is
+discrete.\end{prop}
+
+\begin{proof}See~\cite[Section~XXX]{mines-richman-ruitenburg:constructive-algebra}.\end{proof}
+
+\subsection{Minima of sets of natural numbers}
+
+As an approximate rule of thumb, in the author's personal experience, every
+result of classical mathematics which has received constructive scrutiny turned
+out to either
+\begin{enumerate}
+\item be manifestly and unsurprisingly unconstructive, or
+\item admit a constructive reformulation.
+\end{enumerate}
+Of course, the jury is still out on the many results not yet studied from a
+constructive point of view.
+
+The results ``every vector space has a basis'' and ``there exist
+nonmeasurable subsets of the reals'' are of the first kind. Such results often
+provide great abstract insight to the mathematical landscape -- it is
+comforting to know that there cannot be vector spaces so bizarre that they
+don't have a basis or that the notion of measurable subsets is not trivial.
+Such results also tend to not influence concrete situations too much: How could
+an infinite basis for which there is no hope that we ever learn any actual
+description of inform our computations? (Better uncover topological structure
+and look for a Schauder basis!)
+
+Indeed, there are even metatheorems, reported on in
+Lecture~\ref{lect:constructive-content}, to the effect that in sufficiently
+concrete and logically simple situations, the axiom of choice and the principle
+of excluded middle do not allow to prove results which couldn't be proven
+without them. And as further evidence, there is Friedman's \emph{grand
+conjecture}~\cite{XXX} -- informal but not yet challenged -- that every result
+published in the Annals by a mathematician not self-identifying as a set
+theorist or logician can be (reformulated to be) provable even in~\textsc{efa},
+a foundational system much weaker than~\textsc{zfc},~\textsc{zf} or even Peano
+arithmetic.
+
+An example of a result of type~(b) us the basic observation that every
+inhabited set of natural numbers contains a minimal element. This result is not
+available in constructive mathematics as stated, but there are two constructive
+substitutes.
+
+\begin{defn}A set~$X$ is \emph{inhabited} if and only if there exists an
+element of~$X$. (Constructively, this is stronger than~$X$ not being
+empty.)\end{defn}
+
+\begin{prop}If every inhabited set of natural numbers contains a minimal
+element, then the principle of excluded middle holds.\end{prop}
+
+To salvage the classical result, we can strengthen its assumption or weaken its
+conclusion.
+
+\begin{defn}A subset~$U \subseteq X$ is \emph{detachable} if and only if for
+every element~$x \in X$, either~$x \in U$ or~$x \not\in U$.\end{defn}
+
+For instance, the subset of~$\NN$ consisting of the prime numbers is detachable,
+while the subset of those numbers~$n$ such that the~$n$-th Turing machine
+terminates cannot constructively be shown to be detachable.\footnote{A
+computable witness to detachability of this subset would be a \emph{halting
+oracle}, but a basic fact of computability theory is that there no such
+oracles.}
+
+\begin{prop}\label{prop:minima}\begin{enumerate}
+\item Every detachable inhabited subset of~$\NN$ contains a minimum.
+\item Every inhabited subst of~$\NN$ does~\notnot contain a minimum.
+\end{enumerate}
+\end{prop}
+
 \section{Exercises}
@@
 in~\cite{mandelkern:brouwerian-counterexamples}.
+\end{exercise}
+
+\begin{exercise}[Markov's principle]\label{ex:markov}
+Markov's principle is the statement that
+\[ \forall f \? \NN \to \{0,1\}\_ (\neg\neg\exists n \? \NN\_ f(n) = 0)
+\Rightarrow (\exists n \? \NN\_ f(n) = 0). \]
+It is a simple instance of the classical principle of double-negation
+elimination, but not available in (most schools of) constructive mathematics.
+\begin{enumerate}
+\item Why does Markov's principle imply that programs which do not run forever
+actually halt?
+\item Explain why Markov's principle is witnessed by a computer program (even
+though it does not admit a constructive proof). Which assumption on your
+metatheory does your argument require?
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}[Minima of sets of natural numbers, part one]\label{ex:minima1}
+\begin{enumerate}
+\item Give a constructive proof of Proposition~\ref{prop:minima}(b).
+\item What does the computational witness extracted from the proof of
+Proposition~\ref{prop:minima}(a) look like? Can you devise a different proof
+corresponding to a different algorithm?
+\end{enumerate}
 \end{exercise}
@@
 
-\begin{exercise}[Stability of the axioms]
+\begin{exercise}[Stability of the axioms]\label{ex:stability-axioms}
 Peano arithmetic (\textsc{pa}) is set in the language~$(0,S,+,\cdot)$ and has
@@
 
-\begin{exercise}[Minima of sets of natural numbers]
-In Theorem~\ref{thm:min}(b) we showed that every inhabited set of natural
-numbers does \notnot contain a minimal element. Explain how this result follows
+\begin{exercise}[Minima of sets of natural numbers, part two]
+In Proposition~\ref{prop:minima}(b) we showed that every inhabited set of natural
+numbers does \notnot contain a minimal element.
+\begin{enumerate}
+\item Redo Exercise~\ref{ex:minima1}(a) using the motto that in proofs of
+negated statements, we are allowed to use finitely many instances of the
+principle of excluded middle.
+\item Explain how Proposition~\ref{prop:minima}(b) follows
 from the double-negation embedding applied to the classical fact that every
@@
 double-negation embedding. The resulting constructive theorem has to be
-massaged a bit to yield Theorem~\ref{thm:min}(b).\par}
+massaged a bit to yield Proposition~\ref{prop:minima}(b).\par}
+\item Explain in computational terms how your proof of
+Proposition~\ref{prop:minima}(b) runs its course.
+\end{enumerate}
 \end{exercise}
@@
         \]
+
+Comment on Zorn's lemma

### 2022-09-08 algar-lecture-notes.tex
Abstract for first lecture on toposes

@@
 
+\begin{intro}
+Just as the easiest and shortest path between two truths of the real domain
+often passes through the complex domain, sometimes the easiest and most
+conceptual path to a result observes that the claim can also be formulated in
+the internal language of an alternate mathematical universe -- an alternate
+topos -- where the proof can be simple, direct and make use of the finer
+distinctions provided by intuitionistic logic.
+
+As such, toposes are the workhorse of many modern applications of constructive
+mathematics to mathematics in general. Toposes originated in algebraic
+geometry, where they have been used as generalized spaces (allowing for a
+nontrivial category of opens instead of only a partially ordered set of opens),
+before category theorists observed that there is also a strong logical aspect
+to toposes: Toposes can be pictured as mathematical universes -- universes in
+which we can do mathematics much in the same way as in the so-called standard
+topos of sets and maps, the single topos in which most mathematicians spend all
+their professional life in.
+
+Besides the standard topos, in which mathematics unfolds exactly as known and
+classical logic reigns, there is a colorful host of alternate toposes -- some
+intimately related and useful to the mathematics of the standard topos, some
+less so and important to computability theorists and logicians. In most
+alternate toposes, only intuitionistic logic is sound and it is a plain fact of
+the matter that the law of excluded middle and the axiom of choice fail.
+
+In the lecture we will focus on precise definitions, examples of toposes, and
+logical applications to mathematics in general, laying the groundwork for the
+fourth lecture on generic models and carefully elucidating why it is that most
+toposes validate only intuitionistic logic. References include
+\cite{blechschmidt:filmat,leinster:introduction,streicher:ctcl,moerdijk-maclane:sheaves-logic,goldblatt:topoi,johnstone:elephant}
+(and \cite{blechschmidt:phd} for applications in algebraic geometry). These
+themes constitute just a tiny part of topos theory. Exposition on Olivia
+Caramello's grand topos-theoretic bridge-building program~\cite{caramello:tst},
+on applications of toposes to the meta-analysis of
+logic~\cite{lambek-scott:hocl} and how topos theory can be used to reconcile
+seemingly disparate positions in the philosophy of
+mathematics~\cite{lambek:incompatible,couture-lambek:reflections} will take
+place in more informal settings than the lecture.
+\end{intro}
+
 \section{Examples}

### 2022-09-08 algar-lecture-notes.tex
Beginning of list of globalization examples
Number exercises in a more sensible way
Introduction to topos-theoretic phantoms

@@
 }
+
+\usepackage{pifont}
+\newcommand{\cmark}{\ding{51}}
+\newcommand{\xmark}{\ding{55}}
 
@@
 
-\newtheorem{exercise}[defn]{Exercise}
+\newtheorem{exercise}{Exercise}[chapter]
 \renewcommand{\theenumi}{\alph{enumi}}
@@
 
+\newpage
+
+\begin{tabular}{@{}lp{0.83\textwidth}ll@{}}
+  \toprule
+  & statement & status \\\midrule
+  1a & Every continuous function~$f : [-1,1] \to \RR$ with~$f(-1) < 0$
+  and~$f(1) > 0$ has a zero. & \cmark/\xmark \\
+  1b & For every continuous function~$f : X \times [-1,1] \to \RR$
+  with~$f(\cdot,-1) < 0$ and~$f(\cdot,1) > 0$ there is an open covering~$X =
+  \bigcup_i U_i$ such that for each index~$i$, there is a continuous
+  zero-picking function~$z : U_i \to [-1,1]$
+\end{tabular}
+
 \section{Examples}
@@
 
+The convergence radius of the Taylor series expansion of~$1/(1-x)$ around the
+origin, that is the geometric series
+$1 + x + x^2 + \cdots$,
+is~$1$. That the convergence radius isn't larger doesn't come too surprising in
+view of the fact that the function~$1/(1-x)$ has a non-removable discontinuity
+at~$x = 1$. But why is also the convergence radius of the expansion
+of~$1/(1+x^2)$,
+\[ 1/(1+x^2) = 1 - x^2 + x^4 - x^6 + \cdots, \]
+just~$1$? The function~$1/(1+x^2)$ is
+perfectly well-defined at~$\pm 1$, infinitely derivable on all of~$\RR$, neatly
+decaying for~$x \to \pm\infty$ and doesn't have any singularities.
+
+A rewarding explanation of this phenomenon is nowadays relayed in any course on
+complex analysis: The function~$1/(1+x^2)$ does have a singularity at
+distance~$1$ from the origin, namely at~$\pm i$ in the complex plane, and the disk
+of convergence is only so big that it merely touches the nearest singularity.
+
+This episode is one example of many how the ``imaginary unit''---once
+an elusive \emph{mathematical phantom}---affects the story of the real numbers even
+though it itself is not one of them. Like many mathematical phantoms before it,
+the imaginary unit obtruded its effects so convincingly that eventually we
+embraced it and broadened our notion of existence.
+
+Toposes allow us to bring further entities into being, entities which do not
+exist in the standard topos but which nevertheless affect the story of the
+standard topos. We will focus on:
+\begin{enumerate}
+\item The \emph{generic ring} -- which cryptically is also a field
+\item The \emph{generic prime filter}~$\ppp_\star$ of a given ring~$A$ -- which
+XXX
+\item The \emph{generic surjection}~$\NN \to X$ -- which exists also when~$X$
+is uncountable
+\end{enumerate}
+The complex numbers can be compiled down to pairs of real numbers and hence
+exhibit already in their very construction an intimate connection with the real
+numbers; in a similar vein, these topos-theoretic generic models can be
+compiled away in that proofs employing them can be unrolled, in a mechanical
+fashion even, so as to not mention them.
+
 \section{A fantastical ring}

### 2022-09-08 algar-lecture-notes.tex
typo

@@
 
-\newenvironment{intro}{\begin{quote}}{\end{quote}\bigskip}
+\newenvironment{intro}{\begin{quote}}{\end{quote}\bigskip\noindent}
 
@@
 every element which is not zero is invertible (as in the localization of the
-Cauchy or Dedekind reals at the set of non-zeros).
+Cauchy or Dedekind reals at the set of non-zeros or in the \emph{generic local
+ring} described in Lecture~\ref{lect:generic}).
 \end{enumerate}These each have their

### 2022-09-09 algar-lecture-notes.tex
Many more exercises

@@
 \newcommand{\B}{\mathcal{B}}
-\newcommand{\C}{\mathcal{C}}
+\newcommand{\CCC}{\mathcal{C}}
 \newcommand{\D}{\mathcal{D}}
@@
 \renewcommand{\H}{\mathcal{H}}
-\renewcommand{\O}{\mathcal{O}}
+\newcommand{\OOO}{\mathcal{O}}
 \newcommand{\K}{\mathcal{K}}
@@
 \newcommand{\notnot}{\emph{not~not}}
+\newcommand{\Sh}{\mathrm{Sh}}
 
@@
 numbers~$e + \pi$ and~$e \cdot \pi$ is transcendental.
+
+{\noindent\scriptsize\emph{Note.} At time of writing, for none of these numbers
+a (constructive or classical) proof of their transcendence is known.\par}
 \end{exercise}
@@
 Kuratowski-finite iff its elements can be enumerated: $X = \{x_1,\ldots,x_n \}$.\par}
+
+\item Every subset of the (Cauchy or Dedekind) reals which is inhabited and
+bounded from above has a supremum.
+
+{\noindent\scriptsize\emph{Note.} A \emph{supremum} of a set~$M$ of reals is a
+number~$s$ such that~$M \leq s$ (that is~$x \leq s$ for all~$x \in M$) and such
+that for every number~$s'$ with~$M \leq s'$, $s \leq s'$. In constructive
+mathematics, we can make finer distinctions between the classically equivalent
+constructions of the real numbers: The reals constructed using Cauchy
+sequences inject into the reals constructed using Dedekind cuts which in turn
+inject into the MacNeille reals, and each serve a different purpose. The Cauchy
+and Dedekind reals cannot constructively be shown to be complete in the sense
+of this exercise, while the MacNeille reals can. Conversely, the rationals can
+be shown to be dense in the first two kinds of reals but not in the MacNeille
+reals~\cite[Section~D4.7]{johnstone:elephant}.\par}
 \end{enumerate}
@@
 \begin{enumerate}
-\addtocounter{enumi}{3}
+\addtocounter{enumi}{4}
 \item Every real number which is not zero is invertible.
@@
 \begin{ex}Classical linear algebra teaches us that every finite system of
-generators of a vector space over a residue field\footnote{Constructively, the
+generators of a vector space over a residue field\footnote{\label{fn:fields}Constructively, the
 notion of a field bifurcates into several nonequivalent notions: A nontrivial
@@
 
+% \begin{document}
+
 \newpage
@@
 \section{Exercises}
+
+\begin{exercise}[{The sheaf of real functions as a
+field~\cite[Exercise~28]{blechschmidt:generalized-spaces}}]
+Let~$X$ be a topological space. Let~$\CCC$ be the sheaf of continuous real-valued
+functions on~$X$.
+\begin{enumerate}
+\item Let~$U$ be an open of~$X$. Let~$f \in \CCC(U)$. Show that~$U \models (\exists g\?\CCC\_ fg =_\CCC
+1)$ iff there is a function~$g \in \CCC(U)$ such that~$fg = 1$.
+\item Show that~$\CCC$ is a \emph{residue field} in that
+$X \models \forall f\?\CCC\_
+  \bigl(\neg(\exists g\?\CCC\_ fg = 1)\bigr) \Rightarrow f = 0$.
+\item Give an example of space~$X$ such that it is not the case that~$\CCC$ is a
+field in the stronger sense that
+$\Sh(X) \models \forall f\?\CCC\_ (\exists g\?\CCC\_ fg = 1) \vee f = 0$.
+\item Let~$X = \CC$ and let~$\OOO$ be the sheaf of holomorphic functions on~$X$,
+that is~$\OOO(U) = \{ f : U \to \CC \,|\, \text{$f$ holomorphic} \}$.
+Show that, in classical mathematics, the sheaf~$\OOO$ is \emph{discrete} in that
+\[ \Sh(\CC) \models \forall f\?\OOO\_ \forall g\?\OOO\_ f = g \vee \neg(f = g).
+\]
+\end{enumerate}
+
+\end{exercise}
+\begin{exercise}[{The geometric interpretation of double
+negation~\cite[Exercise~26]{blechschmidt:generalized-spaces}}]
+Let~$\varphi$ be a formula over an open~$U$ of a topological space~$X$.
+\begin{enumerate}
+\item Show that there is a largest open,
+denoted~``$\llbracket\varphi\rrbracket$'', such
+that~$\llbracket\varphi\rrbracket \models \varphi$.\smallskip
+
+{\noindent\scriptsize\emph{Hint.} Consider the union of all opens~$V$ such that~$V
+\models \varphi$.\par}
+
+\item Show that~$X \models \neg\neg\varphi$ iff~$\llbracket\varphi\rrbracket$
+is dense.
+
+{\noindent\scriptsize\emph{Note.} A subset~$W \subseteq X$ is \emph{dense} iff
+for every open~$T \subseteq X$, if~$W \cap T = \emptyset$ then~$T =
+\emptyset$.\par}
+
+\item Give a condition on the topological space~$X$ such that the principle of
+excluded middle (or equivalently the principle of double negation elimination)
+is valid in~$\Sh(X)$.
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}[The fundamental properties of the sheaf semantics]
+Complete the proof of Proposition~\ref{prop:sheaf-semantics}, so verify
+monotonicity and locality of the sheaf semantics.
+\end{exercise}
+
+\begin{exercise}[{Unique local existence is global existence~\cite[Exercise~27]{blechschmidt:generalized-spaces}}]
+Let~$F$ be a sheaf on a topological space~$X$. Let~$\varphi$ be a formula over~$X$. Assume
+that~$\Sh(X) \models \exists!s\?F\_ \varphi(s)$, that is
+\[ \Sh(X) \models \exists s\?F\_ \varphi(s) \quad\text{and}\quad
+  \Sh(X) \models \forall s\?F\_ \forall t\?F\_ (\varphi(s) \wedge \varphi(t)
+  \Rightarrow s = t). \]
+Show that for any open~$U \subseteq X$, there is a unique section~$s \in F(U)$
+such that~$U \models \varphi(s)$.
+
+{\noindent\scriptsize\emph{Note.} If you do not know the general definition of
+a sheaf, then consider only the specific sheaf~$\CCC$ of continuous real-valued
+functions.\par}
+\end{exercise}
 
@@
 
-\vfill{\small%
-\begin{verbatim}
-- Existence of generic models
-- Generic ring is a field
-- Generic prime ideal
-- Kripke–Joyal semantics
-\end{verbatim}
-}
+\begin{exercise}[Anonymously Noetherian rings]
+A commutative ring~$A$ is \emph{anonymously Noetherian} if for every
+ideal~$\aaa \subseteq A$ it is \notnot the case that~$\aaa$ is finitely
+generated.
+\begin{enumerate}
+\item Show that residue fields in the sense of Footnote~\ref{fn:fields} are
+anonymously Noetherian.
+\item Verify that~$\ZZ$ is anonymously Noetherian.
+
+{\noindent\scriptsize\emph{Note.} Either run a direct proof or use the
+anonymous least number principle of Proposition~\ref{prop:minima}(a).\par}
+\item Look up a proof of Hilbert's basis theorem, for instance the
+proof in~\cite[Theorem~7.5]{atiyah-macdonald:commutative-algebra}, and check
+that it can be massaged into a constructive proof of the statement that
+polynomial rings over anonymously Noetherian rings are anonymously Noetherian.
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}[{Prüfer domains as valuation domains~\cite[Exercise~43]{blechschmidt:generalized-spaces}}]
+An \emph{integral domain} is a commutative ring such that~$1 \neq 0$ and such that~$xy = 0$
+implies~$x = 0$ or~$y = 0$. A \emph{valuation domain} is an integral domain
+such that for any two elements, one divides the other. A \emph{Prüfer domain}
+is an integral domain such that every finitely generated ideal~$\aaa$ is locally a
+principal ideal (in the sense that there exists a partition~$1 = f_1 + \cdots +
+f_n$ such that, for each index~$i$, the ideal~$\aaa[f_i^{-1}]$ is a principal
+ideal in~$A[f_i^{-1}]$).
+\begin{enumerate}
+\item Let~$A$ be a ring. Show that~$A^\sim$ is an integral domain if~$A$ is.
+Does the converse hold?
+\item Let~$A$ be a valuation domain. Show that any matrix over~$A$ can be put
+into diagonal form by elementary row and column operations.
+\item Let~$A$ be an integral domain. Show that~$A$ is a Prüfer domain if and
+only if~$A^\sim$ is a valuation domain.
+\item Let~$A$ be a Prüfer domain. Show that any matrix over~$A$ can locally be
+put into diagonal form by elementary row and column operations, by applying the
+result of part~(b) to~$A^\sim$.
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}[{A basic version of Kaplansky's theorem~\cite[Exercise~44]{blechschmidt:generalized-spaces}}]
+\begin{enumerate}
+\item Let~$A$ be a local ring. Let~$\aaa \subseteq A$ be a finitely generated
+ideal such that~$\aaa^2 = \aaa$. Show that~$\aaa = (0)$ or~$\aaa = (1)$.\smallskip
+
+{\noindent\scriptsize\emph{Hint.} Nakayama's lemma.\par}
+\item Let~$A$ be a local ring. Let~$M \in A^{n \times n}$ be an idempotent
+matrix. Verify that~$M$ is similar to a diagonal matrix with entries~$0$ and~$1$
+by applying part~(a) to the ideals of~$k$-minors of~$M$. Deduce that the
+cokernel of~$M$ is finite free.
+\item Let~$A$ be an arbitrary ring. Let~$M \in A^{n \times n}$ be an idempotent
+matrix. Show that the cokernel of~$M$ is finite locally free, by applying the
+result of part~(b) to~$A^\sim$.
+\item Verify, without using the law of excluded middle or that~$A$ is
+Noetherian, that an~$A$-module~$M$ is finitely generated and projective if and
+only if it is finite locally free.\smallskip
+
+{\noindent\scriptsize\emph{Note.} A self-contained solution is given in
+Ref.~\cite{blechschmidt:kaplansky}.\par}
+\end{enumerate}
+\end{exercise}
 

### 2022-09-10 algar-lecture-notes.tex
More background on generic elements and models

@@
 \newcommand{\D}{\mathcal{D}}
-\newcommand{\E}{\mathcal{E}}
+\newcommand{\EEE}{\mathcal{E}}
 \newcommand{\F}{\mathcal{F}}
@@
 \newcommand{\Sh}{\mathrm{Sh}}
+\newcommand{\Set}{\mathrm{Set}}
 
@@
 \item The \emph{generic ring} -- which cryptically is also a field
-\item The \emph{generic prime filter}~$\ppp_\star$ of a given ring~$A$ -- which
+\item The \emph{generic prime filter} of a given ring~$A$ -- which
 XXX
@@
 
+\section{Ring-theoretic prelude: the method of indeterminates}
+
+The Cayley--Hamilton theorem of linear algebra states that every square matrix
+over every commutative ring is a root of its characteristic polynomial. Among
+its many proofs, the following one is particularly interesting from a logical
+point of view: It is enough to verify the claim for the \emph{ring-theoretic
+generic~$(n \times n)$-matrix}~$(A_{ij})_{ij}$ over the polynomial
+ring~$\ZZ[A_{ij}\,|\,i,j\in\{1,\ldots,n\}]$, since this matrix can specialize
+to every specific matrix over every commutative ring and this kind of specialization
+preserves the~$n^2$ polynomial identities expressing that the matrix is a root
+of its characteristic polynomial. Now to check this special case, it suffices to
+verify it for every choice of complex numbers~$a_{ij} \in \CC$ in place of the
+formal indeterminates~$A_{ij}$. For complex matrices, the claim follows from
+the observation that the claim is immediate for the diagonalizable complex
+matrices and that these are dense in the space of all complex matrices.
+
+This approach is known as the \emph{method of indeterminates} and is amenable
+to many situations. For instance, there is the
+\emph{ring-theoretic generic~$n$-th root} (over~$\ZZ[X]/(X^n-1)$), the
+\emph{ring-theoretic generic nilpotent element of index~$\leq n$}
+(over~$\ZZ[X]/(X^n)$), the \emph{ring-theoretic generic power series}
+(over~$\ZZ[A_0,A_1,\ldots]$), and many others. The central features of the
+method of indeterminates are:
+\begin{enumerate}
+\renewcommand{\theenumi}{\arabic{enumi}}
+\item We extend a certain base ring, often~$\ZZ$, to a larger ring~$R_0$ containing
+the \emph{generic instance} of the problem at hand.
+\item For every specific instance in every ring~$R$, there is a unique
+homomorphism~$R_0 \to R$ mapping the generic instance to that one.
+\item Since homomorphisms of rings preserve polynomial identities, every
+polynomial identity (and in fact even every property which can be expressed as a
+geometric formula in the language of rings) satisfied by the generic instance
+passes down to every specific instance. In fact, the generic instance satisfies
+exactly those polynomial identities which are provable from the ring axioms.
+\item Crucially, the generic instance is applicable to special techniques (for
+instance involving complex numbers) and has special properties which themselves
+do not pass down to every specific instance, but which are useful in
+establishing those that do.
+\end{enumerate}
+
+The generic models facilitated by topos theory generalize the ring-theoretic
+generic elements. The latter are restricted to those kinds of elements which
+can be described by polynomial identities (such as~$X^n - 1$ for~$n$-th roots).
+For example, there is no ring-theoretic generic non-nilpotent element
+or ring-theoretic generic regular element (a ring element~$x \in A$ is
+\emph{regular} if and only if for all~$y \in A$, $xy = 0$ implies~$y = 0$).
+The topos-theoretic generic models, in contrast, support arbitrary geometric
+implications as defining properties, and are not restricted to mere elements
+(or finite or infinite lists of elements): For instance, as elucidated in the
+next sections, there is the topos-theoretic \emph{generic ring} or the
+topos-theoretic \emph{generic prime filter} of any given ring.
+
+Briefly, for every geometric theory~$\TT$ (set of geometric implications as
+axioms) over any signature, there exists a topos-theoretic generic model
+of~$\TT$. The analogy with the generic elements of ring theory is as follows:
+\begin{enumerate}
+\renewcommand{\theenumi}{\arabic{enumi}}
+\item We extend the base universe, often the standard topos~$\Set$, to a larger
+topos~$\Set[\TT]$ containing the \emph{generic model}~$U_\TT$ of~$\TT$.
+\item For every specific~$\TT$-model~$M$ in any topos~$\EEE$, there is a unique
+map~$\Set[\TT] \to \EEE$ mapping~$U_\TT$ to~$M$.\footnote{To be more precise,
+the category of maps~$\mathrm{Hom}(\Set[\TT],\EEE)$ is equivalent to the
+category of~$\TT$-models in~$\EEE$, naturally in~$\EEE$. By ``map'', we mean
+what is called the ``inverse image part of geometric morphisms'' in topos
+theory.}
+\item Since such maps preserve geometric implications, every property
+which can be expressed as a geometric implication passes from~$U_\TT$ to every
+specific model. In fact, the generic model satisfies exactly those geometric
+implications which are provable from the axioms of~$\TT$.
+\item Crucially, the generic model is applicable to special techniques and has
+special properties which themselves do not pass down to every specific model,
+but which are useful in establishing those that do.
+\end{enumerate}
+
 \section{A fantastical ring}
@@
 \section{Exercises}
+
+\begin{exercise}[Limitations of ring-theoretic generics]
+Show that each of the following hypothetical ring-theoretic generics do not
+exist:
+\begin{enumerate}
+\item the generic nilpotent element (of arbitrary index)
+\item the generic regular element
+\item the generic prime element
+\end{enumerate}
+That is, show that there is no ring~$R_0$ containing an element~$x_0$ of the
+desired kind such that for every such element~$x$ of every ring~$R$, there is a
+unique homomorphism~$R_0 \to R$ mapping~$x_0$ to~$x$.
+\end{exercise}
 

### 2022-09-13 paper-flabby-objects.tex
More references (thanks to Felix)
to do
typo

@@
 extension~$R\langle B \rangle \to I$ induces an extension~$B \to I$ of~$f$.
-\item One verifies that any abelian group embeds into a divisible
+\item One verifies that every abelian group embeds into a divisible
 abelian group. By Baer's criterion (which in this form requires the axiom of choice), divisible abelian groups are injective.
@@
 * quasicompact/spectral case
+
+CITE:
+https://stacks.math.columbia.edu/tag/05NM
+Oberst relative satellites
+Amitsur derived functors
+Buchsbaum satellites
+Buchsbaum a note on homology
+Stauffer derived functors without injectives
+Felix
+Thierry
+Huber Calculation of derived functors
+Huishan WU, Revers mathematics and semisimple rings

### 2022-09-13 algar-lecture-notes.tex
More work on KJ examples and the generic prime filter

@@
 \newcommand{\cl}{\mathrm{cl}}
-\newcommand{\notnot}{\emph{not~not}}
+\newcommand{\notnot}{\emph{not~not}\xspace}
 \newcommand{\Sh}{\mathrm{Sh}}
@@
 
-\begin{tabular}{@{}lp{0.83\textwidth}ll@{}}
-  \toprule
-  & statement & status \\\midrule
-  1a & Every continuous function~$f : [-1,1] \to \RR$ with~$f(-1) < 0$
-  and~$f(1) > 0$ has a zero. & \cmark/\xmark \\
-  1b & For every continuous function~$f : X \times [-1,1] \to \RR$
-  with~$f(\cdot,-1) < 0$ and~$f(\cdot,1) > 0$ there is an open covering~$X =
-  \bigcup_i U_i$ such that for each index~$i$, there is a continuous
-  zero-picking function~$z : U_i \to [-1,1]$
-\end{tabular}
+\emph{Analysis:}
+\begin{enumerate}
+\item[\cmark/\xmark] Every continuous function~$f : [-1,1] \to \RR$ with~$f(-1) < 0$
+  and~$f(1) > 0$ has a zero.
+\item[\xmark/\xmark] Let~$X$ be a topological space and let~$f : X \times [-1,1] \to \RR$
+be a continuous function with~$f(\cdot,-1) < 0$ and~$f(\cdot,1) > 0$. Then
+there is an open covering~$X = \bigcup_i U_i$ such that for each index~$i$,
+there is a continuous zero-picking function~$z : U_i \to [-1,1]$.
+\end{enumerate}
+
+\emph{Linear algebra:}
+\begin{enumerate}
+\item[\cmark/\xmark] Every real symmetric matrix has an eigenvector.
+\item[\xmark/\xmark] Let~$X$ be a topological space and let~$A : X \to
+M^\text{sym}_n(\RR)$ be a continuous map to the space of symmetric~$(n \times
+n)$-matrices. Then there is an open covering~$X = \bigcup_{i \in I} U_i$ such
+that or all indices~$i \in I$, there is a continuous map~$v : U_i \to \RR^n$
+such that for each~$x \in U_i$, the vector~$v(x)$ is an eigenvector of~$A(x)$.
+\end{enumerate}
+
+\emph{Commutative algebra:}
+\begin{enumerate}
+\item[\cmark/\cmark] Let~$M$ be a finitely generated projective module over a local
+ring~$A$. Then~$M$ is finite free.
+\item[\cmark/\cmark] Let~$M$ be a finitely generated projective module over an
+arbitrary commutative ring~$A$. Then there is a partition~$1 = f_1 + \cdots +
+f_n \in A$ of unity such that, for each index~$i$, the localized
+module~$M[f_i^{-1}]$ is finite free over~$A[f_i^{-1}]$.
+\item[\cmark/\xmark] Let~$M$ be a finitely generated module over a (residue)
+field~$k$. Then~$M$ is finite free.
+\item[\xmark/\xmark] Let~$M$ be a finitely generated module over an
+arbitrary commutative ring~$A$. Then there is a partition~$1 = f_1 + \cdots +
+f_n \in A$ of unity such that, for each index~$i$, the localized
+module~$M[f_i^{-1}]$ is finite free over~$A[f_i^{-1}]$.
+\item[\cmark/\cmark] Let~$M$ be a finitely generated module over a (residue)
+field~$k$. Then~$M$ is \notnot finite free.
+\item[\cmark/\cmark] Let~$M$ be a finitely generated module over an
+arbitrary commutative ring~$A$. If~$f = 0$ is the only element of~$A$ such
+that~$M[f^{-1}]$ is finite free over~$A[f^{-1}]$, then~$1 = 0$ in~$A$.
+\item[\cmark/\cmark] If~$k$ is a (residue) field, there is no linear
+surjection~$k^n \to k^m$ with~$m > n$.
+\item[\cmark/\cmark] Let~$A$ be an arbitrary commutative ring. If there exists
+a linear surjection~$A^n \to A^m$ with~$m > n$, then~$1 = 0$ in~$A$.
+\end{enumerate}
 
@@
 distance~$1$ from the origin, namely at~$\pm i$ in the complex plane, and the disk
-of convergence is only so big that it merely touches the nearest singularity.
+of convergence is and can only be so big that it just touches the nearest singularity.
 
@@
 the imaginary unit obtruded its effects so convincingly that eventually we
-embraced it and broadened our notion of existence.
+embraced it and broadened our notion of existence~\cite{wraith:phantoms}.
 
@@
 \section{Ring-theoretic prelude: the method of indeterminates}
+\label{sect:indeterminates}
 
@@
 \section{The generic prime filter}
-\newpage
-\ \newpage
+
+A commutative ring~$A$ is often studied by its \emph{stalks}~$A_\ppp$ at prime
+filters~$\ppp \subseteq A$, those subsets of~$A$ which validate the conditions
+on the right:
+\begin{align*}
+  \top &\,\,\Longrightarrow\, 0 \in \ppp &
+  0 \in \ppp &\,\,\Longrightarrow\, \bot \\
+  x \in \ppp \wedge y \in \ppp &\,\,\Longrightarrow\, x + y \in \ppp &
+  x + y \in \ppp &\,\,\Longrightarrow\, x \in \ppp \vee y \in \ppp \\
+  1 \in \ppp &\,\,\Longrightarrow\, \bot &
+  \top &\,\,\Longrightarrow\, 1 \in\ppp \\
+  x \in \ppp \vee y \in \ppp &\,\Longleftrightarrow\, xy \in \ppp &
+  xy \in \ppp &\,\Longleftrightarrow\, x \in \ppp \wedge y \in \ppp
+\end{align*}
+The stalk~$A_\ppp$ is then the localization~$A[\ppp^{-1}]$, that is
+the ring of formal fractions~$x/s$ with~$x \in A$ and~$s \in \ppp$, where~$x/s
+= y/t$ iff there exists an element~$u \in \ppp$ such that~$utx = usy$.
+The set~$\ppp$ is multiplicatively closed (and saturated) by the last two
+conditions.\footnote{Textbooks on classical commutative algebra usually use
+\emph{prime ideals} instead of \emph{prime filters}, those subsets which
+validate the dual conditions on the left. In classical commutative algebra, we
+can freely pass between these two notions by taking complements, as the
+complement of a prime ideal is a prime filter and conversely. Constructivly,
+prime filters are more useful, and even classically they are slightly more
+convenient to work with.} Similarly, an~$A$-module~$M$ is often studied by its
+stalks~$M_\ppp = M[\ppp^{-1}]$.
+
+Stalks are interesting because they allow us to zoom in on ``local''
+situations---they tend to be simpler than the original ring or module, while
+still being intimately connected to the originals. For instance:
+\begin{enumerate}
+\item The stalks~$A_\ppp$ are always \emph{local
+rings},\label{page:local-rings} rings such that a finite sum is invertible only
+if one of its summands also is. (This amounts to~$1 \neq 0$ and that
+whenever~$x+y$ is invertible, $x$ is invertible or~$y$ is invertible; see
+Exercise~\ref{ex:local-rings}.)
+\item Classically, if~$A$ is reduced, the stalks~$A_\ppp$ at
+\emph{maximal} prime filters (equivalently minimal prime ideals) are even fields.
+\item Classically, an~$A$-module~$M$ is flat if and only if all of its stalks~$M_\ppp$ are.
+\end{enumerate}
 
@@
 
+\begin{exercise}[Verifying polynomial identities]
+
+\end{exercise}
+
 \begin{exercise}[Limitations of ring-theoretic generics]
@@
 
+\begin{exercise}[Local rings]\label{ex:local-rings}
+\begin{enumerate}
+\item Using the definition of local ring given on
+page~\pageref{page:local-rings}, verify that every geometric field in the sense
+of Footnote~\ref{fn:fields} is a local ring; that the (Cauchy or Dedekind)
+reals form a local ring; that for each prime number~$p$ the subring of~$\QQ$
+consisting of those fractions whose denominator is not divisible by~$p$ is a
+local ring; and that~$\ZZ$ and~$\QQ[X]$ are not local rings.
+
+{\noindent\scriptsize\emph{Note.} If~$a < b$ are real numbers, then for every
+real number~$x$, $x > a$ or~$x < b$.\par}
+
+\item Show that the stalk~$A_\ppp$ of a commutative ring~$A$ at a prime
+filter~$\ppp$ is a local ring. What goes wrong in case that~$\ppp$ is instead a
+prime ideal and one considers the localization~$A[(A\setminus\ppp)^{-1}]$?
+
+\item Verify in classical mathematics that a ring is local if and only if it
+has exactly one maximal ideal.
+
+{\noindent\scriptsize\emph{Hint.} Use that, in classical mathematics, every
+nonunit is contained in some maximal ideal.\par}
+\end{enumerate}
+\end{exercise}
+
+\begin{exercise}[Abstract existence of bounds from the method of
+indeterminates]
+\begin{enumerate}
+\item Using Krull's lemma in its unconstructive form, verify for every
+commutative ring~$A$ that if a polynomial~$f \in A[X]$ is nilpotent in~$A[X]$,
+all its coefficients are nilpotent in~$A$.
+\item Let~$f \in A[X]$ be a polynomial of degree~$d$ such that~$f^n = 0$.
+By the proof in~(a), there exists a number~$m$ such that the~$m$-th power of
+each of the coefficients of~$f$ vanishes. However, it is conceivable that~$m$
+depends not only on~$d$ and~$n$, but also on the ring~$A$ and the specific
+values of the coefficients. Explain how the method of indeterminates reviewed
+in Section~\ref{sect:indeterminates} can be used to establish that there exists
+some bound on~$m$ depending only on~$d$ and~$n$.
+\end{enumerate}
+\end{exercise}
+
 \begin{exercise}[Anonymously Noetherian rings]
-A commutative ring~$A$ is \emph{anonymously Noetherian} if for every
+A commutative ring~$A$ is \emph{anonymously Noetherian} iff for every
 ideal~$\aaa \subseteq A$ it is \notnot the case that~$\aaa$ is finitely

### 2022-09-18 slides-brixen2022b.tex
Slides for informal talk almost finished

--- 2021-11-23 slides-dagstuhl2021.tex
@@
 \usepackage[export]{adjustbox}
-%\setbeameroption{show notes on second screen}
-%\setbeamertemplate{note page}[plain]
-%\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}
-\newcommand{\jnote}[2]{}
 
@@
 
+\usepackage[normalem]{ulem}
 \usepackage{graphbox}
@@
 \usepackage{xstring}
+\usepackage{proof}
 \usepackage{ifthen}
@@
 
-\title{Bridging the foundational gap: Updating algebraic geometry in
-face of current challenges regarding formalizability, constructivity and
-predicativity}
+\title{Connecting inductive definitions, generic models and a modal multiverse
+for algebra and combinatorics}
+
 \author{Ingo Blechschmidt}
-\date{November 23th, 2021}
+\date{September ??th, 2022}
 
-\useinnertheme[shadow=true]{rounded}
+%\useinnertheme[shadow=true]
 \setbeamerfont{block title}{size={}}
@@
 \usecolortheme{seahorse}
-\definecolor{mypurple}{RGB}{150,0,255}
+\definecolor{mypurple}{RGB}{253,73,34}
 \setbeamercolor{structure}{fg=mypurple}
@@
   \begin{beamercolorbox}[dp=1ex,center]{}%
-    \usebeamercolor[fg]{item}{\textbf{{\Large\insertframetitle}}}
+    \begin{tikzpicture}
+      \def\R{8pt}
+      \node (title) {\hil{{\large\,\!\insertframetitle}}};
+      \begin{pgfonlayer}{background}
+        \draw[decorate, very thick, draw=mypurple!30]
+          ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
+          ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
+          ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
+          ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
+          cycle;
+      \end{pgfonlayer}
+    \end{tikzpicture}
   \end{beamercolorbox}%
@@
 }
+\newcommand{\normalnumber}[1]{%
+  {\renewcommand{\insertenumlabel}{#1}\!\usebeamertemplate{enumerate item}\!}
+}
 \newcommand{\bigheart}{\includegraphics{heart}}
@@
 \newcommand{\subhead}[1]{{\centering\textcolor{gray}{\hrulefill}\quad\textnormal{#1}\quad\textcolor{gray}{\hrulefill}\par}}
+
+\newcommand{\badbox}[1]{\colorbox{red!30}{#1}}
+
+% taken from JDH "The modal logic of arithmetic potentialism and the universal algorithm"
+\DeclareMathOperator{\possible}{\text{\tikz[scale=.6ex/1cm,baseline=-.6ex,rotate=45,line width=.1ex]{\draw (-1,-1) rectangle (1,1);}}}
+\DeclareMathOperator{\necessary}{\text{\tikz[scale=.6ex/1cm,baseline=-.6ex,line width=.1ex]{\draw (-1,-1) rectangle (1,1);}}}
+\DeclareMathOperator{\xpossible}{\text{\tikz[scale=.6ex/1cm,baseline=-.6ex,rotate=45,line width=.1ex]{\draw (-1,-1) rectangle (1,1); \draw[very thin] (-.6,-.6) rectangle (.6,.6);}}}
+\DeclareMathOperator{\xnecessary}{\text{\tikz[scale=.6ex/1cm,baseline=-.6ex,line width=.1ex]{\draw (-1,-1) rectangle (1,1); \draw[very thin] (-.6,-.6) rectangle (.6,.6);}}}
 
@@
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses}\end{minipage}}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{1.59cm}\includegraphics[width=\paperwidth]{forest-light}\end{minipage}}
 \begin{frame}[c]
@@
     \def\R{8pt}
-    \node (pretitle) {Bridging the foundational gap:};
-    \node (title) [below of=pretitle] {\vbox{\vspace*{-0.5em}Updating \hil{algebraic
-    geometry} in face of \hil{current challenges} regarding
-    \hil{formalizability}, \hil{constructivity} and \hil{predicativity}}};
+    \node (title) {\vbox{\vspace*{-0.0em}Connecting \hil{inductive definitions},
+    \hil{generic models} and \\ a \hil{modal multiverse} for algebra and combinatorics
+    \\[0.8em]}};
     \begin{pgfonlayer}{background}
@@
 
-  Dagstuhl Seminar 20202: \\
-  Geometric Logic, Constructivisation, and Automated Theorem Proving \\
+  REDCOM: \\
+  \emph{Reducing complexity in algebra, logic, combinatorics}
   \ \\
-  November 23th, 2021
+  September 19th, 2022
   \bigskip
@@
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
-\begin{frame}{Algebraic geometry in a nutshell}
-  Turn \hil{commutative rings} into \hil{spaces}, and \hil{glue} those
-  spaces.
-  \bigskip
+\definecolor{mypurple}{RGB}{150,0,255}
+\setbeamercolor{structure}{fg=mypurple}
 
-  \subhead{Examples}
+% \begin{document}
+
+\begin{frame}{Infinite data}
+  \begin{block}{}
+    \justifying
+    \textbf{Def.} A set~$X$ together with a binary relation~$R$ is
+    \hil{almost-full$_\infty$} iff every \bad{\uwave{infinite sequence}}~$\alpha : \NN \to X$ is
+    \hil{good} in that there exist numbers~$i < j$ such that~$\alpha(i)
+    \mathrel{R} \alpha(j)$.
+  \end{block}
+  \vspace*{-0.6em}
+  {\mbox{\small\emph{Examples.} $(\NN,{\leq})$,\ \ $X \times Y$ [Dickson],\ \
+  $X^*$ [Higman],\ \ $\mathrm{Tree}(X)$ [Kruskal]} \\
+  \bad{Only classically.}
+  \pause
+  Constructive reformulation:
+  \par}
+%  {\small\emph{Examples.}\vspace*{-0.8em}
+%  \begin{itemize}
+%    \item $(\NN,{\leq})$\vspace*{-0.6em}
+%    \item Cartesian products of almost-full sets (by \emph{Dickson's Lemma}) \vspace*{-0.6em}
+%    \item Finite lists over almost-full sets (by \emph{Higman's Lemma}) \vspace*{-0.6em}
+%    \item Finite trees over almost-full sets (by \emph{Kruskal's Tree Theorem})
+%  \end{itemize}}
+
+  \begin{block}{}
+    \justifying
+    \textbf{Def.} For a predicate~$P$ on finite lists over~$X$, inductively
+    define:
+    \vspace*{-0.6em}
+    \[
+      \infer{P \,|\, \sigma}{P(\sigma)}
+      \qquad
+      \infer{P \,|\, \sigma}{\forall x \in X\_ P \,|\, (\sigma :: x)}
+    \]
+  \end{block}
+  \pause
+
+  \begin{block}{}
+    \textbf{Def.} A set~$X$ together with a binary relation~$R$ is
+    \hil{almost-full\textsubscript{ind}} iff~$\mathsf{Good} \,|\, []$, where
+    $\mathsf{Good}(\sigma) \equiv (\exists i < j\_ \sigma[i] \mathrel{R}
+    \sigma[j])$.
+  \end{block}
+  \vspace*{-0.6em}
   \begin{enumerate}
-    \item $\Spec(k[X_1,\ldots,X_n]) = \AA^n$.
-    \item $\Spec(k[X,Y]/(Y-X^2)) = \text{standard parabola}$.
-    \item Gluing
-    \begin{tikzpicture}[scale=0.2,baseline={([yshift=-.6ex]current bounding box.center)}]
-      \draw (0,0) circle (1);
-      \draw [fill=white] (90:1) circle (0.3);
-    \end{tikzpicture}
-    with
-    \begin{tikzpicture}[scale=0.2,baseline={([yshift=-.6ex]current bounding box.center)}]
-      \draw (0,0) circle (1);
-      \draw [fill=white] (270:1) circle (0.3);
-    \end{tikzpicture}
-    along
-    \begin{tikzpicture}[scale=0.2,baseline={([yshift=-.6ex]current bounding box.center)}]
-      \draw (0,0) circle (1);
-      \draw [fill=white] (90:1) circle (0.3);
-      \draw [fill=white] (270:1) circle (0.3);
-    \end{tikzpicture}
-    yields
-    $\PP^1$.
+    \item Constructively, almost-full\textsubscript{ind} $\Rightarrow$
+    almost-full$_\infty$.
+    \item With \badbox{\textsc{lem}+\textsc{dc}}, almost-full\textsubscript{ind} $\Leftarrow$
+    almost-full$_\infty$.
+    %\pause
+    %\item How much stronger is almost-full\textsubscript{ind}?
+    %many-valued sequences, \ldots
   \end{enumerate}
-  \bigskip
+\end{frame}
 
-  \subhead{Concrete results}
-  Fermat's Last Theorem: For~$n \geq 3$, no positive integers satisfy
-  \[ a^n + b^n = c^n. \]
+\begin{frame}{Constructive red flags}
+  \emph{Well-foundedness:} A transitive relation~$<$ on a set~$X$ is \ldots
+  \begin{itemize}
+    \item \ \\[-1.2em]\mbox{\hil{well-founded$_\infty$} iff
+    there is no \bad{\uwave{infinite chain}} $x_0 > x_1 > \cdots$.}
+    \item \hil{well-founded'$_\infty$} iff
+    there is no \bad{\uwave{bad set}} (inhabited and for every member there is
+    a smaller member).
+
+    \item<4-> \hil{well-founded\textsubscript{ind}} iff for every~$x \in X$,
+    $\mathsf{Acc}(x)$: %, where~$\mathsf{Acc}$ is inductively defined by:
+    \[
+      \infer{\mathsf{Acc}(x)}{\forall y < x\_ \mathsf{Acc}(y)}
+    \]
+    \vspace*{-1.3em}
+  \end{itemize}
+
+  \pause
+  \emph{Krull's Lemma:} An element~$x$ of a ring~$A$ is nilpotent if \ldots
+  \begin{itemize}
+    \item it is contained in
+    every \bad{\uwave{prime ideal}}.
+    \item<5-> the theory of prime ideals of~$A$ proves~``$x \in \ppp$''.
+  \end{itemize}
+  \medskip
+
+  \pause
+  \justifying
+  \emph{Dependent choice:}
+  Let~$R$ be a binary relation on a set~$X$
+  such that~$\forall a \in X\_ \exists b \in X\_ a \mathrel{R} b$. Let~$x_0
+  \in X$. Then there is an \bad{\uwave{infinite chain}} $x_0 \mathrel{R}
+  x_1 \mathrel{R} x_2 \mathrel{R} \cdots$.
+\end{frame}
+
+%\begin{frame}{Inductive definitions}
+%  \begin{block}{}
+%    \justifying
+%    \textbf{Def.} A transitive relation~$<$ on a set~$X$ is \hil{well-founded} iff
+%    for all~$x \in X$, $\mathsf{Acc}(x)$, where the accessibility predicate is
+%    inductively generated by:
+%  \end{block}
+%  \vspace*{-0.6em}
+%  {\small\emph{Examples.} $(\NN,{<})$\par}
+%
+%  \textbf{Prop.} If~$<$ is well-founded, there are no infinite descending
+%  chains~$x_0 > x_1 > \ldots$.
+%
+%  \emph{Proof.} By induction on~$x_0$.
+%\end{frame}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.59cm}\includegraphics[width=\paperwidth]{staircase}\end{minipage}}
+\begin{frame}{The set-theoretic multiverse}
+  \begin{block}{}
+    \justifying
+    \textbf{Def.} A \hil{model of set theory} is a (perhaps class-sized) structure $(M,{\in})$
+    satisfying axioms such as those of~\textsc{zfc}.
+  \end{block}
+  \vspace*{-0.6em}
+  {\small\emph{Examples.}\vspace*{-0.8em}
+  \begin{itemize}
+    \item $V$, the class of all sets \vspace*{-0.6em}
+    \item $L$, Gödel's constructible universe \vspace*{-0.6em}
+    \item $V[G]$, a forcing extension containing a generic filter of~$G$ \vspace*{-0.6em}
+    \item Henkin/term models from consistency of (extensions of)~\textsc{zfc}
+  \end{itemize}}
+  \pause
+
+  {\centering\emph{Embracing all models of set theory:}\par}
+  \begin{block}{}
+    \justifying
+    \textbf{Def.} $\possible\varphi$ iff~$\varphi$ holds in \hil{some extension} of
+    the current universe. \\
+    \phantom{\textbf{Def.}} $\necessary\varphi$ iff~$\varphi$ holds in \hil{all extensions}
+    of the current universe.
+  \end{block}
+  \begin{itemize}
+    \item $\necessary(\possible\textsc{CH} \wedge \possible\neg\text{CH})$,
+    the continuum hypothesis is a \hil{switch}
+    \item \ \\[-1.2em]\mbox{$\necessary\possible\necessary(\text{$X$ is countable})$,
+    existence of an enumeration is a \hil{button}}
+  \end{itemize}
 \end{frame}}
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.95cm}\includegraphics[width=\paperwidth]{staircase}\end{minipage}}
-\begin{frame}{Transfinite methods}
-  The standard presentation of algebraic geometry hinges on:
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
+\begin{frame}{A topos-theoretic multiverse}
+  \begin{block}{}
+    \justifying
+    \textbf{Def.} A statement~$\varphi$ holds \ldots
+    \begin{enumerate}
+      \item\small \hil{everywhere} ($\necessary\varphi$) iff
+      it holds in every (Grothendieck) topos (over the current base topos).
+      \item \hil{somewhere} ($\possible\varphi$) iff
+      it holds in some positive topos.
+      \item \hil{subsequently} ($\xpossible\varphi$) iff
+      it holds in some positive ouvert topos.
+    \end{enumerate}
+  \end{block}
+  \pause
+
+  \newcommand{\seqbad}{\only<2>{infinite descending sequence}\only<3->{bad set}}
   \begin{itemize}
-    \item large structures: classes, large categories, universes, \ldots
-    \item powersets
-    \item law of excluded middle
-    \item axiom of choice
+  \small
+    \item\justifying
+    \only<2-4>{A relation is well-founded\textsubscript{ind} \\
+    iff \emph{everywhere} there are no \seqbad s \\
+    iff the \emph{generic \seqbad} validates~$\bot$ \\
+    iff the \emph{theory} of \seqbad s proves~$\bot$.}
+    \only<5->{A relation is well-founded\textsubscript{ind} iff
+    \emph{nowhere} there are \seqbad s.}
+    \pause
+    \pause
+
+    \item \only<1-4>{A ring element~$x$ is nilpotent \\
+    iff \emph{everywhere} it is contained in all prime ideals \\
+    iff it is contained in the \emph{generic prime ideal} \\
+    iff the \emph{theory} of prime ideals proves~$x \not\in \ppp$.}
+    \only<5->{A ring element is nilpotent iff it is contained in all prime
+    ideals \emph{everywhere}.}
+
+    \pause
+    \pause
+    \item Given a set~$X$, \emph{subsequently} there is a surjection~$\NN
+    \twoheadrightarrow X$. \\
+    \hil{NB:} $(\possible\varphi) \Rightarrow \varphi$, if~$\varphi$ is a
+    geometric implication. \\
+    \phantom{\hil{NB:}} $(\xpossible\varphi) \Rightarrow \varphi$, if~$\varphi$
+    is first-order.
+
+    \pause
+    \item Given~$(X,R,x_0)$ as in~\textsc{dc}, \emph{subsequently} there is an
+    infinite chain.
+
+    \pause
+    \item Somewhere, the law of excluded middle holds.
   \end{itemize}
-  \bigskip
-
-  \subhead{despite:}
-  \begin{enumerate}
-    \item subject matter (in part) very concrete
-    \item computer algebra systems for computations practical
-    \item constructive algebra well-established
-    \item high-level proofs often constructive
-  \end{enumerate}
 \end{frame}}
 
-\begin{frame}{Formalizing algebraic geometry}
-  \bigskip
-  \bigskip
-  \bigskip
-  \includegraphics[width=\textwidth]{schemes-in-lean2}
+\end{document}
 
-  \vspace*{-17em}
-  \includegraphics[frame=0.1em,bb=0 170 610 625,width=0.3\textwidth]{schemes-in-coq}\hfill
-  \includegraphics[frame=0.1em,bb=0 170 610 625,width=0.3\textwidth]{schemes-in-lean}
-\end{frame}
+DC
+Krull
+Countability
+Well
+Well-founded
 
-\begin{frame}{A trinitarian challenge}
-  \begin{center}\tiny $1^3 + \cdots + n^3 = (1 + \cdots + n)^2$\end{center}
-  \vspace*{-0.6em}
+Choice:
+erst abzählbar machen,
+dann DC
+nein, funktioniert nicht ganz
+aber für diskrete Quelle habe surjektive (nicht offene?) Theorie/Topos
 
-  \begin{columns}
-    \begin{column}{0.33\textwidth}
-      \centering
-      \includegraphics[height=7em]{lem}
+∀x ∈ X. ∃y ∈ Y. φ(x,y).
+(Toposwechsel für Abzählbarkeit von X)
+∀n ∈ ℕ. ∃y ∈ Y. φ(πn,y).
+Yn := { y ∈ Y | φ(πn,y) }
+(~) auf M = (coprod_n Y_n): (n,y) ~ (n',y') gdw. n' = n+1
+Dann habe: M bewohnt (durch (0,...)) und ∀p ∈ M. ∃q ∈ M. p ~ q.
+(Toposwechsel für DC für (~))
+Habe f : ℕ → M mit f(0) = (0,...) und f(n) ~ f(n+1) für alle n.
+Also ist π₂.f eine Auswahlfunktion, genauer: ∀n. φ(πn, π₂(f(n))).
 
-      \hil{constructivity}
-      \par
-    \end{column}
-    \begin{column}{0.33\textwidth}
-      \centering
-      \includegraphics[height=7em]{sum-of-cubics}
+Zorn für abzählbare Mengen?
+C_0 = {}
+C_{n+1} = C_n cup { x_n | C_n cup {x_n} Kette }
+C = bigcup_n C_n
+z = sup(C)
+Dann ist z maximal?
+Gelte z ≤ e.
+Dann ist C cup { e } Kette.
+Wenn e = x_n0, so insbesondere C_n0 cup { e } Kette.
+Also e ∈ C_{n+1} ⊆ C, also e ≤ z.
+Also e = z.
+Problem: Kommt eher selten vor; Supremum wird oft nicht zur Menge gehören.
+Beispiel: Supremum von Menge endlich erzeugter Ideale ist nicht endlich erzeugt.
 
-      \hil{elegance}
-      \par
-    \end{column}
-
-    \begin{column}{0.33\textwidth}
-      \centering
-      \includegraphics[height=7em]{proof-assistants}
-
-      \par
-      \hil{formalizability}
-    \end{column}
-  \end{columns}
-  \bigskip
-  \bigskip
-
-  \subhead{Thesis}
-  \centering
-  Elegant mathematics lends itself to rewarding formalization.
-\end{frame}
-
-\begin{frame}{Foundational possibilities}
-  \small
-
-  \subhead{Schemes and scheme morphisms}
-  \only<1-2>{
-  \begin{itemize}
-    \setlength\itemsep{-0.2em}
-    %\item locally ringed spaces (or locales, distributive lattices, sites,
-    %toposes, arithmetic universes), formal geometries
-    \item locally ringed \ldots \\
-    \ldots{} topological spaces \tabto{4cm}-- \textit{unconstructive, hard to formalize} \\
-    \ldots{} locales \tabto{4cm}-- \textit{impredicative, superfluous opens} \\
-    \ldots{} distributive lattices \tabto{4cm}-- \textit{need extension from basis} \\
-    \ldots{} sites \tabto{4cm}-- \textit{current favorite! morphisms intricate} \\
-    \ldots{} toposes \tabto{4cm}-- \textit{large structure} \\
-    \ldots{} arithmetic universes \tabto{4cm}\mbox{-- \textit{a bit better;
-    issue with relative spectrum}}
-    \item formal geometries \tabto{4cm}-- \textit{can be regarded as sites}
-    \item functor of points \tabto{4cm}-- \textit{large structure, issue with
-    sizes or \\ \qquad
-    with schemes not of finite presentation over the base ring}
-    \item formal gluing data \tabto{4cm}-- \textit{morphisms intricate}
-  \end{itemize}
-  \textbf{A surprise of uncertain import:}
-  \pause
-  Internally to the big Zariski topos of a base scheme, the Zariski spectrum of
-  a finitely presented algebra does have enough points! [Cherubini--Coquand]}
-  \bigskip
-  \pause
-
-  \subhead{Zariski cohomology}
-  \begin{itemize}
-    \setlength\itemsep{-0.2em}
-    \small
-    \item Čech cohomology \tabto{4.5cm}-- \textit{ad hoc, but fine for}
-    \tabto{4.5cm}\phantom{--} \textit{quasicompact separated schemes}
-    \item injective resolutions \tabto{4.5cm}-- \textit{hopelessly unconstructive}
-    \item dynamical injectives \tabto{4.5cm}-- \textit{??}
-    \item flabby resolutions \tabto{4.5cm}-- \textit{probably unconstructive}
-    \item pointwise Kan extensions \tabto{4.5cm}-- \textit{fine! partially
-    defined $\mathbb{R}\Gamma$;} \tabto{4.5cm}-- \textit{existence
-    verified for} \tabto{4.5cm}\phantom{--} \textit{quasicompact separated schemes;}
-    \tabto{4.5cm}\phantom{--} \textit{hyper coverings?}
-  \end{itemize}
-
-  \subhead{Étale cohomology}
-  \textcolor{red!90}{??}
-\end{frame}
-
-\end{document}
+Attribution:
+Blass(?)
+Alexander
+Henry

### 2022-09-18 slides-brixen2022b.tex
Slides almost finished

@@
 \begin{frame}{Infinite data}
+  \vspace*{-0.6em}
   \begin{block}{}
@@
   {\mbox{\small\emph{Examples.} $(\NN,{\leq})$,\ \ $X \times Y$ [Dickson],\ \
-  $X^*$ [Higman],\ \ $\mathrm{Tree}(X)$ [Kruskal]} \\
+  $X^*$ [Higman],\ \ $\mathrm{Tree}(X)$ [Kruskal]} \\[-0.1em]
   \bad{Only classically.}
@@
   \par}
-%  {\small\emph{Examples.}\vspace*{-0.8em}
-%  \begin{itemize}
-%    \item $(\NN,{\leq})$\vspace*{-0.6em}
-%    \item Cartesian products of almost-full sets (by \emph{Dickson's Lemma}) \vspace*{-0.6em}
-%    \item Finite lists over almost-full sets (by \emph{Higman's Lemma}) \vspace*{-0.6em}
-%    \item Finite trees over almost-full sets (by \emph{Kruskal's Tree Theorem})
-%  \end{itemize}}
 
@@
       \qquad
-      \infer{P \,|\, \sigma}{\forall x \in X\_ P \,|\, (\sigma :: x)}
+      \infer{P \,|\, \sigma}{\forall x \in X\_\ P \,|\, (\sigma :: x)}
     \]
   \end{block}
+  \vspace*{-0.6em}
+  {\small\justifying ``No matter how the finite approximation~$\sigma$ to an infinite
+  sequence will \\[-0.1em] evolve to a better approximation, eventually~$P$ will hold.''\par}
   \pause
@@
   \end{block}
-  \vspace*{-0.6em}
+  \vspace*{-0.8em}
   \begin{enumerate}
     \item Constructively, almost-full\textsubscript{ind} $\Rightarrow$
-    almost-full$_\infty$.
+    almost-full$_\infty$. \vspace*{-0.2em}
     \item With \badbox{\textsc{lem}+\textsc{dc}}, almost-full\textsubscript{ind} $\Leftarrow$
@@
 \begin{frame}{Constructive red flags}
-  \emph{Well-foundedness:} A transitive relation~$<$ on a set~$X$ is \ldots
+  A transitive relation~$<$ on a set~$X$ is \ldots
   \begin{itemize}
@@
     \item \hil{well-founded'$_\infty$} iff
-    there is no \bad{\uwave{bad set}} (inhabited and for every member there is
+    there is no \bad{\uwave{bad set}} (inhabited and such that for every member there is
     a smaller member).
@@
     \[
-      \infer{\mathsf{Acc}(x)}{\forall y < x\_ \mathsf{Acc}(y)}
+      \infer{\mathsf{Acc}(x)}{\forall y < x\_\ \mathsf{Acc}(y)}
     \]
@@
 
+% \begin{document}
+
 {\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.59cm}\includegraphics[width=\paperwidth]{staircase}\end{minipage}}
 \begin{frame}{The set-theoretic multiverse}
+  \begin{tikzpicture}[remember picture,overlay]
+    \node[xshift=-1.5cm,yshift=-4cm] at (current page.north east)
+    {\includegraphics[width=2cm]{branching}};
+  \end{tikzpicture}
+  \vspace*{-2em}
+
   \begin{block}{}
@@
     \item $L$, Gödel's constructible universe \vspace*{-0.6em}
-    \item $V[G]$, a forcing extension containing a generic filter of~$G$ \vspace*{-0.6em}
+    \item $V[G]$, a forcing extension containing a generic filter~$G$ of \\ some
+    poset of forcing conditions \vspace*{-0.6em}
     \item Henkin/term models from consistency of (extensions of)~\textsc{zfc}
@@
 
-  {\centering\emph{Embracing all models of set theory:}\par}
+  %{\centering\emph{Embracing all models of set theory:}\par}
   \begin{block}{}
@@
   \end{block}
+  \vspace*{-0.6em}
   \begin{itemize}
@@
       it holds in some positive topos.
-      \item \hil{subsequently} ($\xpossible\varphi$) iff
+      \item \hil{proximally} ($\xpossible\varphi$) iff
       it holds in some positive ouvert topos.
@@
 
-  \newcommand{\seqbad}{\only<2>{infinite descending sequence}\only<3->{bad set}}
+  \newcommand{\seqbad}{\only<3>{infinite descending sequence}\only<4->{bad set}}
+  \begin{changemargin}{-2em}{-1.5em}
   \begin{itemize}
   \small
+    \item \justifying
+    \only<1-4>{A relation is almost-full\textsubscript{ind} \\
+    iff every infinite sequence \emph{everywhere} is good \\
+    iff the \emph{generic infinite sequence} is good \\
+    iff the \emph{theory of infinite sequences} proves goodness.\medskip}
+    \only<5->{A relation is almost-full\textsubscript{ind} iff every infinite
+    sequence \emph{everywhere} is good.}
+    \pause
+
     \item\justifying
-    \only<2-4>{A relation is well-founded\textsubscript{ind} \\
+    \only<3-4>{A relation is well-founded\textsubscript{ind} \\
     iff \emph{everywhere} there are no \seqbad s \\
@@
     \pause
+    \pause
 
-    \item \only<1-4>{A ring element~$x$ is nilpotent \\
+    \item \only<2-6>{A ring element~$x$ is nilpotent \\
     iff \emph{everywhere} it is contained in all prime ideals \\
     iff it is contained in the \emph{generic prime ideal} \\
-    iff the \emph{theory} of prime ideals proves~$x \not\in \ppp$.}
-    \only<5->{A ring element is nilpotent iff it is contained in all prime
-    ideals \emph{everywhere}.}
+    iff the \emph{theory} of prime ideals proves~$x \in \ppp$.}
+    \only<7->{\mbox{A ring element is nilpotent iff all prime
+    ideals \emph{everywhere} contain it.}}
 
@@
     \pause
-    \item Given a set~$X$, \emph{subsequently} there is a surjection~$\NN
+    \item Given a set~$X$, \emph{proximally} there is a surjection~$\NN
     \twoheadrightarrow X$. \\
@@
     \phantom{\hil{NB:}} $(\xpossible\varphi) \Rightarrow \varphi$, if~$\varphi$
-    is first-order.
+    is first-order. \\
+    \phantom{\hil{NB:}} $\varphi \Rightarrow (\necessary\varphi)$,\, if~$\varphi$
+    is a geometric formula.
 
     \pause
-    \item Given~$(X,R,x_0)$ as in~\textsc{dc}, \emph{subsequently} there is an
+    \item Given~$(X,R,x_0)$ as in~\textsc{dc}, \emph{proximally} there is an
     infinite chain.
@@
     \pause
-    \item Somewhere, the law of excluded middle holds.
+    \item \emph{Somewhere,} the law of excluded middle holds.
   \end{itemize}
+  \end{changemargin}
 \end{frame}}
 
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.05cm}\includegraphics[width=\paperwidth]{agda-af-implies-wf}\end{minipage}}
+\begin{frame}{Extracting constructive content}
+  \begin{block}{}
+    \justifying
+    \textbf{Prop.} Let~$({\leq})$ be a transitive almost-full\textsubscript{ind} relation.
+    Then~$({<})$, where~$x < y \equiv (x \leq y \wedge \neg(y \leq x))$,
+    is well-founded\textsubscript{ind}.
+  \end{block}
+  \vspace*{-0.5em}
+
+  \emph{Proof.} Everywhere, there can be no infinite descending chain, as any
+  such would also be good. \qed
+
+  \mbox{Unrolling this proof gives a program~$(\textsf{Good} \,|\, []) \to
+  \prod_{x \? X} \textsf{Acc}(x)$.}
+  \vfill
+\end{frame}}
+
+% \begin{document}
+\addtocounter{framenumber}{-1}
+\begin{frame}{Extracting constructive content}
+  \begin{block}{}
+    \justifying
+    \textbf{Prop.} [Dickson] If~$X$ and~$Y$ are almost-full\textsubscript{ind},
+    so is~$X \times Y$.
+  \end{block}
+  \pause
+
+  \vspace*{-0.3em}
+  \emph{Proof.}\small
+  \vspace*{-0.6em}
+  \begin{enumerate}
+  \item\justifying It suffices to verify that the \emph{generic infinite
+  sequence}~$\gamma = (\alpha,\beta) : \NN \to X \times Y$ is good. \pause Since
+  being good can be put as a geometric implication (in fact, a geometric
+  formula) and since \textsc{lem} holds \emph{somewhere}, we may assume~\textsc{lem}.\pause
+  \item By \textsc{lem} and well-foundedness, there is a minimal value~$\alpha(i_0)$
+  among all values of~$\alpha$. \vspace*{-0.2em}\pause
+  Similarly, there is a minimal value~$\alpha(i_1)$ among~$(\alpha(n))_{n >
+  i_0}$, \pause a minimal value~$\alpha(i_2)$ among~$(\alpha(n))_{n > i_1}$, and so
+  on. \pause By \emph{proximal dependent choice}, we can proximally collect these indices
+  into a function~$i : \NN \to \NN$. \vspace*{-0.2em}\pause
+  \item Switching~\textsc{lem} on again, there is a minimal value~$\beta(i(k_0))$
+  among all values of~$\beta \circ i$. \pause Hence~$\gamma$ is good in view of
+  \begin{align*}
+    \alpha(i(k_0)) &\leq \alpha(i(k_0+1)), &
+    \beta(i(k_0)) &\leq \beta(i(k_0+1)).
+  \end{align*}
+  \end{enumerate}
+\end{frame}
+
 \end{document}
+
+Mit LEM hat jede Sequenz α ein Minimum:
+Induktion über einen Wert, der vorkommt (zum Beispiel α(0)).
+
+Mit LEM gilt: ∃n. α(n) < x  oder  nicht.
+Im ersten Fall fertig nach IV.
+Im zweiten Fall:
+  Sei n beliebig. Weiß ¬(α(n) < x).
+  Also weiß: ¬(α(n) ≤ x) und ¬¬(x ≤ α(n)).
+  Mit LEM also x ≤ α(n).
+
+Mit DC bekomme Folge:
+i₀ so, dass α(i₀) Minimum.
+i₁ so, dass α(i₁) Minimum unter α(i₀+1), α(i₀+2), ...
+
+Betrachte Folge (β(iṉ))ṉ.
+Mit LEM finde Minimum: β(ik).
+Dann β(ik) ≤ β(i(k+1)) und natürlich auch α(ik) ≤ α(i(k+1)).
+
+x < y  heißt:  x ≤ y und ¬(y ≤ x)
+Behauptung: ≤ af und transitiv, dann < wf.
+
+Betrachte generische absteigende Folge.
+Da gut, habe x[i] ≤ x[j].
+Wegen Transitivität auch x[i] > x[j],
+also x[j] < x[i], also x[j] ≤ x[i] und ¬(x[i] ≤ x[j]).
+Widerspruch.
 
@@
 Henry
+
+0: title
+* expectations: no new method for extracting constructive content, rather
+  point of view for putting previous work in a uniform context;
+  philosophy turned mathematics
+
+1: infinite data
+* how much more general? multivalued sequences, "¬¬ sequences", ...
+* where do these definitions come from? conceptual answer
+
+3: set-theoretic multiverse
+* axioms of ZFC massively underspecify the concept of sets;
+  models for exploring the range of set-theoretic possibility
+* to settle a question shown to be independent from ZFC is to
+  understand how it behaves in the multiverse
+
+explanation of generic models

### 2022-09-18 slides-brixen2022b.tex
Slides finished

@@
 
-\setbeamertemplate{headline}{}
+%\setbeamertemplate{headline}{}
+\setbeamertemplate{headline}{%
+  \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex]{}%
+    \insertsectionnavigationhorizontal{\paperwidth}{}{}%
+  \end{beamercolorbox}%
+  \vskip-7pt%
+}
 
@@
   \bigskip
-  \bigskip
-
-  \begin{tikzpicture}
-    \def\R{8pt}
-    \node (title) {\vbox{\vspace*{-0.0em}Connecting \hil{inductive definitions},
-    \hil{generic models} and \\ a \hil{modal multiverse} for algebra and combinatorics
-    \\[0.8em]}};
-    \begin{pgfonlayer}{background}
-      \draw[decorate, very thick, draw=mypurple]
-        ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
-        ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
-        ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
-        ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
-        cycle;
-    \end{pgfonlayer}
-  \end{tikzpicture}
 
@@
   \textit{-- an invitation --}
+
+  \setbeamercolor{block body}{bg=black!100}
+  \begin{block}{}
+  \centering\normalsize\color{white}
+  Connecting \hil{inductive definitions},
+  \hil{generic models} and \\ a \hil{modal multiverse} for algebra and combinatorics
+  \end{block}
+
+%  \begin{tikzpicture}
+%    \def\R{8pt}
+%    \node (title) {\vbox{\vspace*{-0.0em}Connecting \hil{inductive definitions},
+%    \hil{generic models} and \\ a \hil{modal multiverse} for algebra and combinatorics
+%    \\[0.8em]}};
+%    \begin{pgfonlayer}{background}
+%      \draw[decorate, very thick, draw=mypurple]
+%        ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
+%        ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
+%        ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
+%        ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
+%        cycle;
+%    \end{pgfonlayer}
+%  \end{tikzpicture}
+
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
   \bigskip
@@
   REDCOM: \\
-  \emph{Reducing complexity in algebra, logic, combinatorics}
+  \emph{Reducing complexity in algebra, logic, combinatorics} \\
   \ \\
+  Brixen \\
   September 19th, 2022
@@
 
-% \begin{document}
+\section{Infinite data}
 
@@
 
+\section{Constructive red flags}
+
 \begin{frame}{Constructive red flags}
@@
 
-% \begin{document}
+\section{The set-theoretic multiverse}
 
@@
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
-\begin{frame}{A topos-theoretic multiverse}
+\section{The topos-theoretic multiverse}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1-lighter}\end{minipage}}
+\begin{frame}{Toposes and generic models}
+  \vspace*{-0.2em}
+  \begin{changemargin}{-1.2em}{-1.2em}
+  \begin{enumerate}
+    \item\justifying A \hil{(Grothendieck) topos} is a category of sheaves over some
+    \hil{site}.
+
+    {\small\emph{Examples.} $\Set$, $\Sh(X)$, $\Set[\TT]$.\par}
+    \bigskip
+    \pause
+
+    \item The \hil{Kripke--Joyal semantics} defines what it means for a
+    statement~$\varphi$ to hold ``internally in a topos~$\E$'', written~``$\E
+    \models \varphi$''. This semantics is sound with respect to intuitionistic
+    logic.
+
+%    {\small
+%    \begin{itemize}
+%    \item\justifying A statement holds in~$\Set$ iff it is true.
+%    \item A statement holds in the ``topos of double-negation sheaves'' iff its
+%    double-negation translation is true.
+%    \item A statement holds in~$\Sh(X)$ iff a globalized version of it holds for
+%    continuous families on the
+%    space~$X$.\end{itemize}\par}
+    \bigskip
+    \pause
+
+%    \item Every topos supports \hil{mathematical reasoning}: \\
+%    If~$\ \E \models \varphi\ $ and if~$\ \varphi$ entails~$\psi$
+%    intuitionistically,\, then~$\E \models \psi$.
+
+    \item Let~$\TT$ be a geometric theory. The \hil{classifying
+    topos}~$\Set[\TT]$ contains the \hil{generic~$\TT$-model}~$U_\TT$. It is
+    \hil{conservative} in that for geometric implications~$\varphi$, the
+    following are equivalent:
+    \begin{enumerate}
+      \item The statement~$\varphi$ holds for~$U_\TT$ in~$\Set[\TT]$.
+      \item The statement~$\varphi$ holds for every~$\TT$-model in every
+      topos.
+      \item The statement~$\varphi$ is provable modulo~$\TT$.
+    \end{enumerate}
+
+    \begin{center}
+      \begin{tikzpicture}[ultra thick, node distance=7mm]
+        \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40] (a) {$\ZZ$};
+        \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40, right=of a] (b) {$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$};
+        \node[regular polygon, regular polygon sides=5, draw=orange!80,
+        fill=orange!20, right=of b, rounded corners=1pt, inner sep=0cm] (c) {$\O_X$};
+        \node[star, rounded corners=1pt, star points=10, inner sep=0cm, draw=purple!80, fill=purple!20, right=of c] {$U_\TT$};
+      \end{tikzpicture}
+    \end{center}
+  \end{enumerate}
+  \end{changemargin}
+\end{frame}}
+
+\begin{frame}{Modal algebra and combinatorics}
   \begin{block}{}
@@
   \begin{changemargin}{-2em}{-1.5em}
-  \begin{itemize}
+  \begin{enumerate}
   \small
@@
     iff the \emph{generic \seqbad} validates~$\bot$ \\
-    iff the \emph{theory} of \seqbad s proves~$\bot$.}
+    iff the \emph{theory of \seqbad s} proves~$\bot$.}
     \only<5->{A relation is well-founded\textsubscript{ind} iff
@@
     iff it is contained in the \emph{generic prime ideal} \\
-    iff the \emph{theory} of prime ideals proves~$x \in \ppp$.}
+    iff the \emph{theory of prime ideals} proves~$x \in \ppp$.}
     \only<7->{\mbox{A ring element is nilpotent iff all prime
@@
     \item \emph{Somewhere,} the law of excluded middle holds.
-  \end{itemize}
+  \end{enumerate}
   \end{changemargin}
-\end{frame}}
+\end{frame}
 
@@
 
-% \begin{document}
 \addtocounter{framenumber}{-1}
@@
   \emph{Proof.}\small
-  \vspace*{-0.6em}
+  \vspace*{-0.4em}
   \begin{enumerate}
@@
     formula) and since \textsc{lem} holds \emph{somewhere}, we may assume~\textsc{lem}.\pause
+
     \item By \textsc{lem} and well-foundedness, there is a minimal value~$\alpha(i_0)$
@@
     on. \pause By \emph{proximal dependent choice}, we can proximally collect these indices
-  into a function~$i : \NN \to \NN$. \vspace*{-0.2em}\pause
+    into a function~$i : \NN \to \NN$; this switches~\textsc{lem} off. \vspace*{-0.2em}\pause
+
     \item Switching~\textsc{lem} on again, there is a minimal value~$\beta(i(k_0))$
@@
 Henry
+oE countable
 
@@
 
-explanation of generic models
+4: topos and generic models
+* truth in Set, truth in double-negation sheaves, truth in Sh(X)
+* generic ring, generic group, ...

### 2022-09-19 slides-brixen2022b.tex
typo

@@
 \begin{frame}{Constructive red flags}
+  \vspace*{-0.5em}
   A transitive relation~$<$ on a set~$X$ is \ldots
@@
 
-  \newcommand{\seqbad}{\only<3>{infinite descending sequence}\only<4->{bad set}}
+  \newcommand{\seqbad}{\only<3>{infinite descending chain}\only<4->{bad set}}
   \begin{changemargin}{-2em}{-1.5em}
@@
     iff it is contained in the \emph{generic prime ideal} \\
-    iff the \emph{theory of prime ideals} proves~$x \in \ppp$.}
+    iff the \emph{theory of prime ideals} proves~``$x \in \ppp$''.}
     \only<7->{\mbox{A ring element is nilpotent iff all prime
@@
     \pause
-    \item Given a set~$X$, \emph{proximally} there is a surjection~$\NN
-    \twoheadrightarrow X$. \\
+    \item Given an inhabited set~$X$, \emph{proximally} there is a surjection~$\NN
+    \twoheadrightarrow X$ [J--T]. \\
     \hil{NB:} $(\possible\varphi) \Rightarrow \varphi$, if~$\varphi$ is a
@@
     \pause
-    \item \emph{Somewhere,} the law of excluded middle holds.
+    \item \emph{Somewhere,} the law of excluded middle holds. [Barr]
   \end{enumerate}
@@
   \mbox{Unrolling this proof gives a program~$(\textsf{Good} \,|\, []) \to
-  \prod_{x \? X} \textsf{Acc}(x)$.}
+  \prod_{x:X} \textsf{Acc}(x)$.}
   \vfill
@@
     \justifying
-    \textbf{Prop.} [Dickson] If~$X$ and~$Y$ are almost-full\textsubscript{ind},
+    \textbf{Thm.} [Dickson] If~$X$ and~$Y$ are almost-full\textsubscript{ind},
     so is~$X \times Y$.
@@
 
+% \begin{document}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
+%{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{1.59cm}\includegraphics[width=\paperwidth]{forest-light}\end{minipage}}
+\begin{frame}{Outlook}
+  I learned the idea to study a modal multiverse of toposes from
+  \emph{Alexander Oldenziel}, circa 2016. Foreshadowing results:
+  \begin{itemize}
+    \scriptsize
+    \item[1984] André Joyal, Miles Tierney. \emph{An extension of the Galois theory of
+    Grothendieck.}
+    \item[1987] Andreas Blass. \emph{Well-ordering and induction in intuitionistic
+    logic and topoi.}
+    \item[2013] Shawn Henry. \emph{Classifying topoi and preservation of higher
+    order logic by geometric morphisms.}
+  \end{itemize}
+  Work by Milly Maietti and Steve Vickers on \emph{arithmetic universes} is
+  also closely connected. In progress:
+
+  \begin{itemize}
+    \item Develop details and formalize.
+    \item Determine the precise list of valid modal principles.
+    \item Carry out case studies with program extraction.
+    \item Incorporate the right adjoints of geometric morphisms.
+  \end{itemize}
+\end{frame}}
+
+\addtocounter{framenumber}{-1}
+
 \end{document}
-
-Mit LEM hat jede Sequenz α ein Minimum:
-Induktion über einen Wert, der vorkommt (zum Beispiel α(0)).
-
-Mit LEM gilt: ∃n. α(n) < x  oder  nicht.
-Im ersten Fall fertig nach IV.
-Im zweiten Fall:
-  Sei n beliebig. Weiß ¬(α(n) < x).
-  Also weiß: ¬(α(n) ≤ x) und ¬¬(x ≤ α(n)).
-  Mit LEM also x ≤ α(n).
-
-Mit DC bekomme Folge:
-i₀ so, dass α(i₀) Minimum.
-i₁ so, dass α(i₁) Minimum unter α(i₀+1), α(i₀+2), ...
-
-Betrachte Folge (β(iṉ))ṉ.
-Mit LEM finde Minimum: β(ik).
-Dann β(ik) ≤ β(i(k+1)) und natürlich auch α(ik) ≤ α(i(k+1)).
-
-x < y  heißt:  x ≤ y und ¬(y ≤ x)
-Behauptung: ≤ af und transitiv, dann < wf.
-
-Betrachte generische absteigende Folge.
-Da gut, habe x[i] ≤ x[j].
-Wegen Transitivität auch x[i] > x[j],
-also x[j] < x[i], also x[j] ≤ x[i] und ¬(x[i] ≤ x[j]).
-Widerspruch.
-
-DC
-Krull
-Countability
-Well
-Well-founded
 
@@
 
-Attribution:
-Blass(?)
-Alexander
-Henry
-oE countable
-
 0: title

### 2022-09-23 paper-modal-multiverse.tex
Beginning of multiverse paper

\documentclass[oneside]{amsart}

\usepackage[utf8]{inputenc}
\usepackage{amsthm,mathtools,stmaryrd,amssymb}
\usepackage[all]{xy}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{hyperref}
\usepackage{xspace}
\usepackage{color}

\usepackage[natbib=true,style=numeric,maxnames=10]{biblatex}
\usepackage[babel]{csquotes}
\bibliography{paper-modal-multiverse.bib}

\title{The topos-theoretic multiverse}
\author{Ingo Blechschmidt}
\address{Universität Augsburg \\
Institut für Mathematik \\
Universitätsstr. 14 \\
86159 Augsburg, Germany}
\email{ingo.blechschmidt@math.uni-augsburg.de}

\theoremstyle{definition}
\newtheorem{defn}{Definition}[section]
\newtheorem{ex}[defn]{Example}

\theoremstyle{plain}
\newtheorem{prop}[defn]{Proposition}
\newtheorem{cor}[defn]{Corollary}
\newtheorem{lemma}[defn]{Lemma}
\newtheorem{thm}[defn]{Theorem}
\newtheorem{scholium}[defn]{Scholium}

\theoremstyle{remark}
\newtheorem{rem}[defn]{Remark}
\newtheorem{question}[defn]{Question}
\newtheorem{speculation}[defn]{Speculation}
\newtheorem{caveat}[defn]{Caveat}
\newtheorem{conjecture}[defn]{Conjecture}

\newcommand{\xra}[1]{\xrightarrow{#1}}
\newcommand{\XXX}[1]{\textbf{\textcolor{red}{XXX: #1}}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\nnn}{\mathfrak{n}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\QQ}{\mathbb{Q}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\PSh}{PSh}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\effective}{ef{}fective\xspace}
\newcommand{\notnot}{\emph{not~not}\xspace}

\newcommand{\stacksproject}[1]{\cite[{\href{https://stacks.math.columbia.edu/tag/#1}{Tag~#1}}]{stacks-project}}

\begin{document}

\begin{abstract}
  To help put earlier results in constructive algebra and constructive
  combinatorics into perspective, explain a possible background for certain
  inductive definitions and give a unified framework for certain techniques for
  program extraction from classical proofs, we propose a modal study of the
  topos-theoretic multiverse.
\end{abstract}

\maketitle
\thispagestyle{empty}

\noindent
Thanks to the finer distinctions constructive mathematics offers, there is a
host of principles which are available in classical mathematics but seem naive
from a constructive point of view. A non-exhaustive list is:
\begin{enumerate}
\item A transitive relation is well-founded iff there is no infinite descending
chain.
\item A relation is almost-full iff every infinite sequence is good (meaning
that some element is related to a later element in the sequence).
\item \emph{Krull's lemma}: A ring element is nilpotent iff every prime
ideal contains it.
\item Every ring has a maximal ideal.\footnote{Here and in the following, by
\emph{ring} we mean commutative ring with unit and by \emph{maximal ideal} we
mean an ideal~$\mmm$ which is \emph{proper} in the sense that~$1 \in \mmm
\Rightarrow 1 = 0$ and such that for every proper ideal~$\nnn$ with~$\mmm
\subseteq \nnn$, $\mmm = \nnn$.}
\item \emph{Markov's principle:} If a function~$\NN \to \NN$ does \notnot have
a zero, then it actually has a zero.
\item \emph{Dependent choice:} If every element of a set is related by some relation to some other
element, every element can be completed to an infinite chain of related
elements.
\item The law of excluded middle holds.
\end{enumerate}
Constructive theorems always have a computational and geometric
content---from every constructive proof, a corresponding algorithm can be
extracted, and every constructive proof holds also for continuous families of
the objects in question. But the listed classical principles have no
computational witness or fail in continuous families, hence are not
available in constructive mathematics.

In the modal topos-theoretic multiverse, we have the following constructive
replacements to these principles.
\begin{enumerate}
\item A transitive relation is well-founded iff \emph{everywhere} there is no infinite descending
chain.
\item A relation is almost-full iff every infinite sequence \emph{everywhere} is good (meaning
that some element is related to a later element in the sequence).
\item \emph{Krull's lemma}: A ring element is nilpotent iff all prime
ideals \emph{everywhere} contain it.
\item Every ring \emph{proximally} has a maximal ideal.
\item If a function~$\NN \to \NN$ does \emph{everywhere} \notnot have
a zero, then it actually has a zero.
\item If every element of a set is related by some relation to some other
element, every element can \emph{proximally} be completed to an infinite chain
of related elements.
\item \emph{Somewhere,} the law of excluded middle holds.
\end{enumerate}

Briefly, a statement is said to hold \emph{everywhere} iff it holds in every
Grothendieck topos over the current base topos; a statement holds
\emph{somewhere} iff it holds in some positive Grothendieck topos over the
current base; and a statement holds \emph{proximally} iff it holds in some
positive ouvert Grothendieck topos over the current base. More such
\emph{modalities} are also useful and merit study.

This modal language not only allows us to recover certain classical principles
as above, but also makes certain powerful theorems about the topos-theoretic
landscape easy to use:
\begin{enumerate}
\addtocounter{enumi}{7}
\item For every (perhaps uncountable) set~$X$, \emph{proximally} there is a
surjection~$\NN \twoheadrightarrow X$.
\end{enumerate}
We can also adopt the notion of \emph{switches} and \emph{buttons} from the
modal study of the set-theoretic multiverse:
\begin{enumerate}
\addtocounter{enumi}{8}
\item \emph{The law of excluded middle is a switch:} Everywhere it is the case
that somewhere~\textsc{lem} holds and somewhere it does not.
\item If Zorn's lemma holds, it is also everythere the case that it holds
somewhere.
\end{enumerate}

\textbf{Acknowledgments.} Alexander Oldenziel, \ldots

\end{document}

### 2022-09-24 paper-modal-multiverse.tex
More on the modal multiverse

@@
 \usepackage[protrusion=true,expansion=true]{microtype}
+\usepackage{tikz}
+\usepackage{proof}
 \usepackage{hyperref}
@@
 \newcommand{\E}{\mathcal{E}}
+\newcommand{\U}{\mathcal{U}}
 \newcommand{\F}{\mathcal{F}}
@@
 \newcommand{\NN}{\mathbb{N}}
+\newcommand{\TT}{\mathbb{T}}
 \newcommand{\ZZ}{\mathbb{Z}}
@@
 
+% taken from JDH "The modal logic of arithmetic potentialism and the universal algorithm"
+\DeclareMathOperator{\possible}{\text{\tikz[scale=.6ex/1cm,baseline=-.6ex,rotate=45,line width=.1ex]{\draw (-1,-1) rectangle (1,1);}}}
+\DeclareMathOperator{\necessary}{\text{\tikz[scale=.6ex/1cm,baseline=-.6ex,line width=.1ex]{\draw (-1,-1) rectangle (1,1);}}}
+\DeclareMathOperator{\xpossible}{\text{\tikz[scale=.6ex/1cm,baseline=-.6ex,rotate=45,line width=.1ex]{\draw (-1,-1) rectangle (1,1); \draw[very thin] (-.6,-.6) rectangle (.6,.6);}}}
+\DeclareMathOperator{\xnecessary}{\text{\tikz[scale=.6ex/1cm,baseline=-.6ex,line width=.1ex]{\draw (-1,-1) rectangle (1,1); \draw[very thin] (-.6,-.6) rectangle (.6,.6);}}}
+
+\newcommand{\?}{\,{:}\,}
+
 \begin{document}
@@
   To help put earlier results in constructive algebra and constructive
-  combinatorics into perspective, explain a possible background for certain
-  inductive definitions and give a unified framework for certain techniques for
-  program extraction from classical proofs, we propose a modal study of the
-  topos-theoretic multiverse.
+  combinatorics into perspective, give possible background for certain
+  inductive definitions and form a unified framework for certain techniques for
+  extracting programs from classical proofs, we propose a modal study of the
+  topos-theoretic multiverse. Our proposal is inspired by the corresponding study
+  of the set-theoretic multiverse and focusses on applications in other
+  parts of mathematics.
+  % than the intriguing endeavour of exploring the range of set-theoretic possibility.
 \end{abstract}
@@
 \begin{enumerate}
+\renewcommand{\theenumi}{\arabic{enumi}*}
 \item A transitive relation is well-founded iff there is no infinite descending
@@
 \item A relation is almost-full iff every infinite sequence is good (meaning
-that some element is related to a later element in the sequence).
+that some element is related to a later element in the sequence -- this
+notion has been studied in combinatorics~\cite{xxx} and found applications in
+termination checking~\cite{xxx}).
 \item \emph{Krull's lemma}: A ring element is nilpotent iff every prime
@@
 \end{enumerate}
-Constructive theorems always have a computational and geometric
+Constructive theorems always carry computational and geometric
 content---from every constructive proof, a corresponding algorithm can be
@@
 that some element is related to a later element in the sequence).
-\item \emph{Krull's lemma}: A ring element is nilpotent iff all prime
+\item A ring element is nilpotent iff all prime
 ideals \emph{everywhere} contain it.
@@
 of related elements.
-\item \emph{Somewhere,} the law of excluded middle holds.
+\item \emph{Barr's theorem, simple version:} \emph{Somewhere,} the law of
+excluded middle holds.
 \end{enumerate}
 
-Briefly, a statement is said to hold \emph{everywhere} iff it holds in every
-Grothendieck topos over the current base topos; a statement holds
-\emph{somewhere} iff it holds in some positive Grothendieck topos over the
-current base; and a statement holds \emph{proximally} iff it holds in some
-positive ouvert Grothendieck topos over the current base. More such
-\emph{modalities} are also useful and merit study.
+Briefly, a statement~$\varphi$ is said to hold \emph{everywhere}
+($\necessary\varphi$) iff it holds in every Grothendieck topos over the
+current base topos; a statement holds \emph{somewhere} ($\possible\varphi$)
+iff it holds in some positive Grothendieck topos over the current base; and a
+statement holds \emph{proximally} ($\xpossible\varphi$) iff it holds in
+some positive ouvert Grothendieck topos over the current base. More such
+\emph{modalities} are also useful and merit study; the precise definitions are
+given in Section~\ref{sect:defn-modalities}.
 
-This modal language not only allows us to recover certain classical principles
-as above, but also makes certain powerful theorems about the topos-theoretic
-landscape easy to use:
+The modal language not only allows us to recover classical principles
+as above, but also renders some powerful theorems about the topos-theoretic
+landscape smoothly accessible:
 \begin{enumerate}
 \addtocounter{enumi}{7}
-\item For every (perhaps uncountable) set~$X$, \emph{proximally} there is a
+\item \emph{Barr's theorem, full version:} If Zorn's lemma holds, it is
+everywhere the case that it (and even the full axiom of choice) hold somewhere.
+\item If a geometric implication holds \emph{somewhere}, then it holds already here.
+\item \label{item:prox-descends}
+If a first-order statement holds \emph{proximally}, then it holds already here.
+\item \label{item:prox-countable}
+For every (perhaps uncountable) set~$X$, \emph{proximally} there is a
 surjection~$\NN \twoheadrightarrow X$.
 \end{enumerate}
-We can also adopt the notion of \emph{switches} and \emph{buttons} from the
-modal study of the set-theoretic multiverse:
+An example application of the latter two principles
+has recently been studied in constructive
+commutative algebra~\cite{blechschmidt-schuster:constructive-maximal-ideals}.
+For countable rings, an explicit construction of an maximal ideal is available.
+By item~(\ref{item:prox-countable}), this construction can also be carried out
+for arbitrary rings, though the result is not a maximal ideal in the narrow
+sense; rather, the resulting maximal ideal exists \emph{proximally}, in some
+positive ouvert Grothendieck topos. However, the first-order consequences of
+its existence, pertaining for instance to concrete statements about polynomials
+or matrices, pass down to the base topos by item~(\ref{item:prox-descends}).
+In this sense the proximally-existing maximal ideal is like the mathematical
+phantoms of Gavin Wraith~\cite{xxx}, encouraging us to broaden our notion of
+existence because it promises us to work wonders.
+
+We can also adopt the notions of \emph{switches} and \emph{buttons} from the
+modal study of the set-theoretic multiverse~\cite[xxx]{xxx}. Switches are
+statements~$\varphi$ such that~$\necessary(\possible\varphi \wedge
+\possible\neg\varphi)$ and buttons are statements~$\varphi$ such
+that~$\necessary\possible\necessary\varphi$; switches can be toggled on and off
+like a light switch, while buttons once pressed cannot be unpressed:
 \begin{enumerate}
-\addtocounter{enumi}{8}
+\addtocounter{enumi}{10}
 \item \emph{The law of excluded middle is a switch:} Everywhere it is the case
 that somewhere~\textsc{lem} holds and somewhere it does not.
-\item If Zorn's lemma holds, it is also everythere the case that it holds
-somewhere.
+\item \emph{Being countable is a button:} For every set~$X$, everywhere it is
+the case that somewhere (even proximally) it is case that everywhere~$X$ is
+countable.
 \end{enumerate}
+
+We argue that the modal operators~$\necessary,\possible,\xpossible$ and more
+suggested in our proposal are natural extensions and refinements of the
+familiar double-negation modality~$\neg\neg$ in constructive mathematics.
 
@@
 
+xxx: idea of multiverse not new, but arguably basic to topos theory; new is systematic
+study of its modal nature with the focus on applications
+
+xxx: related work: Hamkins, Victoria Gitman, ... concentrates on exploring
+range of set-theoretic possibility, obtaining beautiful results such as ...
+Shawn Henry, Andreas Blass ...
+
+\section{Modal operators}
+\label{sect:defn-modalities}
+
+\begin{defn}A \emph{Grothendieck topos} is a category which is equivalent to
+the category of sheaves on a small site. A \emph{Grothendieck topos over a
+given (Grothendieck or elementary) topos}~$\B$ is ``a Grothendieck topos from
+the point of view of~$\B$''; using a sufficiently expressive form of the
+internal language of~$\B$, this amounts to a bounded geometric morphism~$\E \to
+\B$, and this can be taken as the definition.
+\end{defn}
+
+Henceworth, by the unqualified word ``topos'' we will mean ``Grothendieck topos
+over the current base'', and the current base topos will be denoted~``$\Set$''.
+The base topos might be the ``true category of sets'', assuming that this
+concept is available in one's ontology, or also some other elementary topos
+such as the free topos~\cite{xxx} or the ef{}fective topos~\cite{xxx}.
+
+The xxx (use word of Hamkins) of a base~$\B$ is the collection of Grothendieck
+toposes over~$\B$ (which one should formalize in set-theoretic foundations more
+precisely as the proper class of small sites in~$\B$).
+
+\begin{rem}Perhaps arbitrary elementary toposes over the base, corresponding to
+possibly unbounded geometric morphisms~$\E \to \B$ or even arbitrary
+fibrations/indexed categories over~$\B$ validating an appropriate form of the
+axioms of elementary toposes, should also be
+taken as part of the topos-theoretic multiverse. This first proposal of the
+modal multiverse sticks to Grothendieck toposes for ease of
+formalizability (``for every small site'' can be expressed also in the more
+standard flavors of the internal topos language, while ``for every elementary
+topos'' requires more elaborate versions) and because the restricted multiverse
+is already sufficiently rich for the intended applications.
+\end{rem}
+
+\subsection{Positive toposes}
+
+\begin{defn}A topos~$\E$ is \emph{positive} if and only if the geometric
+morphism~$f : \E \to \Set$ is surjective, that is iff~$f^*$ reflects
+isomorphisms..\end{defn}
+
+\begin{ex}The topos of sheaves over a topological space~$X$ is positive if and
+only if~$X$ has a point; more generally, the topos of sheaves over a locale~$X$
+is positive if and only if every open covering of the top element of the frame
+of~$X$ is inhabited (xxx:verify). As such, positivity is a more informative
+version of the weaker property of being nontrivial (that is, that the top open
+and the bottom open do not coincide).\end{ex}
+
+\begin{ex}The spectrum of a ring~$A$, that is the classifying topos of prime
+filters of~$A$ (or equivalently the topos of sheaves over the classifying
+locale of prime filters of~$A$) is positive iff~$1 \neq 0$ in~$A$.\end{ex}
+
+\begin{ex}A sufficient criterion for the classifying topos of a geometric
+theory~$\TT$ being positive is that xxx.\end{ex}
+
+\subsection{Ouvert toposes}
+
+\begin{defn}A topos~$\E$ is \emph{ouvert} if and only if the geometric
+morphism~$f : \E \to \Set$ is open, that is preserves the interpretation of
+first-order formulas.\end{defn}
+
+\begin{ex}The topos of sheaves over a topological space is always ouvert. More
+generally, the topos of sheaves over a locale~$X$ if ouvert iff there exists a
+\emph{positivity predicate} on its frame of opens in the sense of~\cite{xxx}.
+With~\textsc{lem}, every locale is ouvert~\cite{xxx}.\end{ex}
+
+\begin{ex}The spectrum of a ring~$A$ is ouvert iff every element of~$A$ is
+nilpotent or not~\cite[Proposition~12.51]{blechschmidt:phd}.\end{ex}
+
+\begin{ex}A sufficient criterion for the classifying topos of a geometric
+theory~$\TT$ being ouvert is that the indexing sets of all disjunctions
+appearing on the right hand side of turnstiles, in a normal form presentation
+of~$\TT$, are inhabited~\cite{xxx}.\end{ex}
+
+\subsection{Modal operators}
+
+xxx: properly define language
+
+\section{Generic models and inductive definitions}
+\label{sect:generic-models}
+
+In constructive mathematics, the classical definition of well-founded relations
+as those transitive relations for which there exist no infinite descending
+chains is not particularly useful; while the chain condition is satisfied for
+the intended examples, by its negativity the condition is too weak to
+facilitate the intended proofs.
+
+The established substitute is to define a transitive relation~$({<})$ on a set~$X$ to
+be well-founded if and only if for every subset~$M \subseteq X$,
+\begin{equation}\label{eq:ho-wf}\tag{$\star$}
+\bigl(\forall x\?X\_ (\forall y\?X\_ y < x \Rightarrow y \in M) \Rightarrow x \in
+M\bigr) \Longrightarrow \bigl(\forall x\?X\_ x \in M\bigr).
+\end{equation}
+More economically, and preferably in predicative contexts where there is no
+single set or class of ``all subsets of~$X$'' but for instance a
+hierarchy of subsets of increasing universe level, a transitive relation~$({<})$ is
+well-founded if and only if every element of~$X$ is \emph{accessible}, where
+the accessibility predicate~$\mathsf{Acc}$ is inductively generated~\cite{xxx}
+by the following clause:
+\[
+  \infer{\mathsf{Acc}(x)}{\forall y<x\_\ \mathsf{Acc}(y)}
+\]
+In impredicative settings, this inductive definition of well-foundedness
+coincides with the higher-order characterization~(\ref{eq:ho-wf}), and for the
+purposes of this paper we view the inductive definition as the official one.
+
+Similar inductive notions are used to reformulate other classical definitions
+in a constructively more sensible way. For instance, the classical definition
+of a binary relation~$R$ on a set~$X$ being \emph{almost-full} is ``every
+infinite sequence~$\alpha : \NN \to X$ is \emph{good} in the sense that there
+exist numbers~$i < j$ such that~$\alpha(i) \mathrel{R} \alpha(j)$''.
+For a constructive reformulation, we shift to finite approximations of
+infinite sequences (finite lists of elements of~$X$) and define when such an
+approximation is deemed good:\footnote{By~``$\sigma[n]$'', we mean the element
+at position~$n$ of the finite list~$\sigma$. This notation is only meaningful
+if the length of~$\sigma$ is at least~$n+1$. By~``$\sigma :: x$'' we mean the
+enlarged list which has~$x$ as an extra element at its tail end, and by~``$[]$'' we
+denote the empty list. In computer science practice, it is often more efficient
+to prepend (``$x :: \sigma$'') instead of append, but this detail shall not
+concern us here.}
+\[ \mathsf{Good}(\sigma) \defeqv (\exists i<j\_ \sigma[i] \mathrel{R} \sigma[j]). \]
+We then inductively generate a relation~``$P \,|\, \sigma$'' for monotonous
+predicates~$P$ on finite lists expressing that no matter how the given finite
+approximation~$\sigma$ evolves over time to a better approximation,
+eventually~$P$ will hold, by the clauses
+\[
+  \infer{P \,|\, \sigma}{P(\sigma)}
+  \qquad
+  \infer{P \,|\, \sigma}{\forall x \in X\_\ P \,|\, (\sigma :: x)}
+\]
+and finally define that~$R$ is almost-full iff~``$\mathsf{Good} \,|\, []$''.
+With this inductive definition, expected properties of the class of almost-full
+relations such as being stable under cartesian products (Dickman's lemma),
+finite lists (Higman's lemma) or finitely-branching trees (Kruskal's theorem)
+can all be constructively verified~\cite{xxx}.
+
+A similar such definition has been proposed by XXX and XXX in commutative
+algebra for expressing that a ring is Noetherian~\cite{xxx}; the classical
+definition ``every ascending chain of ideals stabilizes'' and also the more
+meaningful and classically equivalent characterization as ``every ascending
+chain of finitely generated ideals stalls''\footnote{A chain~$\aaa_0 \subseteq
+\aaa_1 \subseteq \cdots$ \emph{stalls} iff for some number~$n \in \NN$,
+$\aaa_{n+1} = \aaa_n$. We are grateful to Matthias Hutzler for proposing this
+terminology.} are constructively too weak; firstly, without the axiom of
+dependent choice we can often not construct such chains~\cite{richman:xxx} (but only
+``multi-valued chains'' as in~\cite[Section~3.9]{blechschmidt:phd}; but also
+see~\cite{richman:xxx}), and secondly, being able to inspect suitable inductive
+witnesses helps in proving the Hilbert basis theorem~\cite{xxx}. XXX and XXX
+hence propose to call a ring \emph{Noetherian} if and only if~$\mathsf{Stalls}
+\,|\, []$, where~$\mathsf{Stalls}$ is the predicate on finite lists of finitely
+generated ideals expressing xxx.
+
+Is there a deeper explanation where these inductive definitions come from,
+apart from working well and being motivated on general constructive
+considerations? Also, constructively the inductive definitions are much
+stronger than their classical counterparts, equivalent only in presence
+of~\textsc{lem} and~\textsc{dc}. For instance, if a relation is almost-full in
+the inductive sense, not only is every infinite sequence good, but also every
+infinite ``multi-valued sequence''\footnote{xxx} or every infinite
+partially-defined sequence~$\alpha$ for which for every number~$n \in \NN$ it
+is \notnot the case that~$\alpha(n)$ exists. Can we pinpoint how much more
+general the inductive definitions are?
+
+The answers to both questions are positive (xxx:language), and the modal
+perspective fruitfully clarifies their connection.
+
+Namely, the theories of an infinite sequence or of an infinite descending chain
+are geometric. As such, there exist their classifying toposes, containing the
+\emph{generic infinite sequence} respectively the \emph{generic infinite
+descending chain}. When is this good respectively validates~$\bot$?
+
+\begin{prop}\label{prop:gen-good}Let~$R$ be a relation on a set~$X$. The
+generic infinite sequence over~$X$ is good if and only if~$R$ is almost-full in
+the inductive sense.\end{prop}
+
+\begin{proof}The classifying topos of the theory of an infinite sequence
+over~$X$ can be presented as the topos of sheaves over the site given by the
+partially ordered set of finite lists of elements of~$X$ with coverage given
+by xxx (see Appendix xxx). The Kripke--Joyal semantics states that the
+statement ``$\alpha_0$ is good'', where~$\alpha_0$ is the generic infinite
+sequence, holds in the classifying topos if and only if there is a
+covering~$\U$ of~$[]$ such that for every open~$U \in \U$, xxx. This precisely
+amounts to~$R$ being almost-full in the inductive sense.
+\end{proof}
+
+\begin{prop}Let~$({<})$ be a transitive relation on a set~$X$. The generic
+infinite descending chain over~$X$ validates~$\bot$ (that is, the classifying
+topos of such chains is trivial) if and only if the relation is well-founded in
+the inductive sense.\end{prop}
+
+\begin{proof}Similar as the proof of Proposition~\ref{prop:gen-good}. Details
+for the variant of ``bad sets'' instead of ``infinite descending chains'' have
+been developed (xxx:language) by xxx~\cite{xxx}.\end{proof}
+
+\section{Extracting programs from multiverse proofs}
+\label{sect:program-extraction}
+
+xxx: examples from slides
+
+\section{Perspectives}
+
+xxx: determine modal laws
+
+xxx: more case studies
+
+xxx: don't forget right adjoints
+
+xxx: predicative
+
 \end{document}

### 2022-09-26 slides-fischbachau2022.tex
Slides for tomorrow almost finished

--- 2022-09-18 slides-brixen2022b.tex
@@
-\documentclass[12pt,utf8,notheorems,compress,t]{beamer}
+\documentclass[11pt,utf8,notheorems,compress,t]{beamer}
 \usepackage{etex}
@@
 
-\usepackage[normalem]{ulem}
 \usepackage{graphbox}
@@
 \usepackage{booktabs}
-\usepackage{stmaryrd}
+\usepackage{stmaryrd,wasysym}
+\usepackage{bussproofs}
 \usepackage{amssymb}
@@
 \usepackage{xstring}
-\usepackage{proof}
 \usepackage{ifthen}
@@
 
-\title{Connecting inductive definitions, generic models and a modal multiverse
-for algebra and combinatorics}
-
+\title{Revisiting divisible, injective and flabby abelian groups from a
+constructive point of view}
 \author{Ingo Blechschmidt}
-\date{September ??th, 2022}
+\date{September 21th, 2022}
 
@@
 \usecolortheme{seahorse}
-\definecolor{mypurple}{RGB}{253,73,34}
+\definecolor{mypurple}{RGB}{150,0,255}
 \setbeamercolor{structure}{fg=mypurple}
@@
 \newcommand{\bad}[1]{\textcolor{red!90}{\textnormal{#1}}}
+\newcommand{\good}[1]{\textcolor{mypurple}{\textnormal{#1}}}
 
@@
 \newcommand{\badbox}[1]{\colorbox{red!30}{#1}}
-
-% taken from JDH "The modal logic of arithmetic potentialism and the universal algorithm"
-\DeclareMathOperator{\possible}{\text{\tikz[scale=.6ex/1cm,baseline=-.6ex,rotate=45,line width=.1ex]{\draw (-1,-1) rectangle (1,1);}}}
-\DeclareMathOperator{\necessary}{\text{\tikz[scale=.6ex/1cm,baseline=-.6ex,line width=.1ex]{\draw (-1,-1) rectangle (1,1);}}}
-\DeclareMathOperator{\xpossible}{\text{\tikz[scale=.6ex/1cm,baseline=-.6ex,rotate=45,line width=.1ex]{\draw (-1,-1) rectangle (1,1); \draw[very thin] (-.6,-.6) rectangle (.6,.6);}}}
-\DeclareMathOperator{\xnecessary}{\text{\tikz[scale=.6ex/1cm,baseline=-.6ex,line width=.1ex]{\draw (-1,-1) rectangle (1,1); \draw[very thin] (-.6,-.6) rectangle (.6,.6);}}}
+\newcommand{\infobox}[1]{\colorbox{yellow!70}{\color{black}#1}}
+\newcommand{\grayline}{\textcolor{gray}{\hspace*{-3em}\hrulefill\hspace*{-3em}}}
 
@@
 {\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{1.59cm}\includegraphics[width=\paperwidth]{forest-light}\end{minipage}}
+\definecolor{mypurple}{RGB}{253,73,34}
 \begin{frame}[c]
@@
 
-  \bigskip
-  %\includegraphics[height=0.32\textwidth]{olivia-lattices}
   \bigskip
@@
 
-  \begin{tikzpicture}
-    \def\R{8pt}
-    \node (title) {\vbox{\vspace*{-0.0em}Connecting \hil{inductive definitions},
-    \hil{generic models} and \\ a \hil{modal multiverse} for algebra and combinatorics
-    \\[0.8em]}};
-    \begin{pgfonlayer}{background}
-      \draw[decorate, very thick, draw=mypurple]
-        ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
-        ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
-        ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
-        ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
-        cycle;
-    \end{pgfonlayer}
-  \end{tikzpicture}
-
   \scriptsize
   \textit{-- an invitation --}
+
+  \setbeamercolor{block body}{bg=black!100}
+  \begin{block}{}
+    \centering\normalsize\color{white}
+    \hil{Extraction} of \hil{programs} from \hil{proofs}
+  \end{block}
+
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
   \bigskip
@@
 
-  REDCOM: \\
-  \emph{Reducing complexity in algebra, logic, combinatorics}
+  Autumn school on \\
+  \emph{Proof and Computation} \\
+  in Fischbachau \\
   \ \\
-  September 19th, 2022
-  \bigskip
+  September 26th to October 1st, 2022
 
@@
 
-\definecolor{mypurple}{RGB}{150,0,255}
-\setbeamercolor{structure}{fg=mypurple}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1-lighter}\end{minipage}}
+\begin{frame}{}
+  \textbf{Thm.}
+  For every number~$n \in \NN$, there is a prime larger than~$n$.
 
-% \begin{document}
-
-\begin{frame}{Infinite data}
-  \vspace*{-0.6em}
-  \begin{block}{}
-    \justifying
-    \textbf{Def.} A set~$X$ together with a binary relation~$R$ is
-    \hil{almost-full$_\infty$} iff every \bad{\uwave{infinite sequence}}~$\alpha : \NN \to X$ is
-    \hil{good} in that there exist numbers~$i < j$ such that~$\alpha(i)
-    \mathrel{R} \alpha(j)$.
-  \end{block}
-  \vspace*{-0.6em}
-  {\mbox{\small\emph{Examples.} $(\NN,{\leq})$,\ \ $X \times Y$ [Dickson],\ \
-  $X^*$ [Higman],\ \ $\mathrm{Tree}(X)$ [Kruskal]} \\[-0.1em]
-  \bad{Only classically.}
-  \pause
-  Constructive reformulation:
-  \par}
-
-  \begin{block}{}
-    \justifying
-    \textbf{Def.} For a predicate~$P$ on finite lists over~$X$, inductively
-    define:
-    \vspace*{-0.6em}
-    \[
-      \infer{P \,|\, \sigma}{P(\sigma)}
-      \qquad
-      \infer{P \,|\, \sigma}{\forall x \in X\_\ P \,|\, (\sigma :: x)}
-    \]
-  \end{block}
-  \vspace*{-0.6em}
-  {\small\justifying ``No matter how the finite approximation~$\sigma$ to an infinite
-  sequence will \\[-0.1em] evolve to a better approximation, eventually~$P$ will hold.''\par}
+  {\emph{Proof.} Any prime factor of~$n! + 1$ will do.\par}
+  \medskip
   \pause
 
-  \begin{block}{}
-    \textbf{Def.} A set~$X$ together with a binary relation~$R$ is
-    \hil{almost-full\textsubscript{ind}} iff~$\mathsf{Good} \,|\, []$, where
-    $\mathsf{Good}(\sigma) \equiv (\exists i < j\_ \sigma[i] \mathrel{R}
-    \sigma[j])$.
-  \end{block}
-  \vspace*{-0.8em}
+  {\centering\emph{``Every constructive theorem has a computable witness.''}
+  \pause
+  \[
+    \begin{array}{c@{\qquad}c@{\qquad}c}
+      \mathrm{HA} \vdash \varphi &\Longrightarrow&
+      \exists e\_ e \Vdash \varphi \\
+      \text{\small constructive proof} &\longmapsto&
+      \text{\small realizer}
+    \end{array}
+  \]}
+  \pause
+  \vspace*{-2.2em}
+
+  \begin{columns}
+    \begin{column}{0.45\textwidth}
+      \begin{itemize}
+        \item Integrated developments \\ \emph{SAT checking, \ldots}
+        \item Computability theory \\ \emph{induction $\widehat{=}$ recursion, \ldots}
+      \end{itemize}
+    \end{column}
+
+    \hspace*{-1em}
+    \begin{column}{0.65\textwidth}
+      \begin{itemize}
+        \item Metatheory of constructive systems \\ \emph{provability results, \ldots}
+        \item Philosophy of proof and computation \\ \emph{realizability in the real world, \ldots}
+      \end{itemize}
+    \end{column}
+  \end{columns}
+
+  \grayline
+  \pause
+
+  \justifying
+  \textbf{Thm.} Every infinite sequence~$\alpha : \NN \to \NN$ is \emph{good}
+  in that there are numbers~$i < j$ such that~$\alpha(i) \leq \alpha(j)$.
+
+  {\emph{Proof.} By~\badbox{\textsc{lem}}, there is a minimal value~$\alpha(i)$.
+  Set~$j \defeq i + 1$.\par}
+
+  {\centering\emph{``Every theorem has a computable witness.''}\par}
+\end{frame}}
+
+\begin{frame}{Heyting arithmetic}
+  The \hil{language of arithmetic} has
+  \begin{itemize}
+    \item as its single sort: $N$
+    \item as function symbols: $0$, $S$, $+$, $\cdot$
+    \item as relation symbols: $=$
+  \end{itemize}
+
+  \hil{Heyting arithmetic} has as axioms (the universal closure of)
+  \begin{align*}
+    &&& \neg(0 = Sx) \\
+    &&& S(x) = S(y) \Rightarrow x = y \\
+    x + 0 &= x &&&
+    x \cdot 0 &= 0 \\
+    x + S(y) &= S(x+y) &&&
+    x \cdot S(y) &= (x \cdot y) + x
+  \end{align*}
+  together with the \hil{induction scheme} (one axiom for each
+  formula~$\varphi$)
+  \[
+    \varphi(0) \wedge \bigl(\forall x\?N\_ \varphi(x) \Rightarrow
+    \varphi(S(x))\bigr) \quad\Longrightarrow\quad \forall x\?N\_ \varphi(x)
+  \]
+  and the rules of \hil{sequence calculus}.
+\end{frame}
+
+\begin{frame}{Sequence calculus}
+  \begin{center}
+    \vspace{-0.5em}
+    \phantom{a}\hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}\UnaryInfC{$\varphi \seq{\vec x} \varphi$}\DisplayProof\hfill
+    \AxiomC{$\varphi \seq{\vec x} \psi$}\UnaryInfC{$\varphi[\vec s/\vec x]
+    \seq{\vec y} \psi[\vec s/\vec x]$}\DisplayProof\hfill
+    \AxiomC{$\varphi \seq{\vec x} \psi$}\AxiomC{$\psi \seq{\vec x}
+    \chi$}\BinaryInfC{$\varphi \seq{\vec x} \chi$}\DisplayProof
+    \phantom{a}\hfill
+    \bigskip\medskip
+
+    \phantom{a}\hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}\UnaryInfC{$\varphi \seq{\vec x} \top$}\DisplayProof\hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}\UnaryInfC{$\varphi \wedge \psi \seq{\vec x} \varphi$}\DisplayProof\hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}\UnaryInfC{$\varphi \wedge \psi \seq{\vec x} \psi$}\DisplayProof\hfill
+    \AxiomC{$\varphi \seq{\vec x} \psi$}\AxiomC{$\varphi \seq{\vec x} \chi$}\BinaryInfC{$\varphi \seq{\vec x} \psi \wedge \chi$}\DisplayProof
+    \phantom{a}\hfill
+    \bigskip\medskip
+
+    \phantom{a}\hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}\UnaryInfC{$\bot \seq{\vec x} \varphi$}\DisplayProof\hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}\UnaryInfC{$\varphi \seq{\vec x} \varphi \vee \psi$}\DisplayProof\hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}\UnaryInfC{$\psi \seq{\vec x} \varphi \vee \psi$}\DisplayProof\hfill
+    \AxiomC{$\varphi \seq{\vec x} \chi$}\AxiomC{$\psi \seq{\vec x} \chi$}\BinaryInfC{$\varphi \vee \psi \seq{\vec x} \chi$}\DisplayProof
+    \phantom{a}\hfill
+    \bigskip\medskip
+
+    \phantom{a}\hfill
+    \Axiom$\varphi \wedge \psi\ \fCenter\seq{\vec x} \chi$
+    \doubleLine
+    \UnaryInf$\varphi\ \fCenter\seq{\vec x} \psi \Rightarrow \chi$
+    \DisplayProof
+    \phantom{a}\hfill
+    \bigskip\medskip
+
+    \phantom{a}\hfill
+    \Axiom$\varphi\ \fCenter\seq{\vec x, y} \psi$
+    \doubleLine
+    \UnaryInf$\exists y\?Y\_\! \varphi\ \fCenter\seq{\vec x} \psi$
+    \DisplayProof
+    {\tiny ($y$ not occurring in~$\psi$)}
+    \hfill
+    \Axiom$\varphi\ \fCenter\seq{\vec x, y} \psi$
+    \doubleLine
+    \UnaryInf$\varphi\ \fCenter\seq{\vec x\phantom{, y}} \forall y\?Y\_\! \psi$
+    \DisplayProof
+    {\tiny ($y$ not occurring in~$\varphi$)}
+    \hfill\phantom{a}
+
+    \phantom{a}\hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}
+    \UnaryInfC{$\top \seq{x} x = x$}
+    \DisplayProof
+    \hfill
+    \AxiomC{$\phantom{\seq{\vec x}}$}
+    \UnaryInfC{$(\vec x = \vec y) \wedge \varphi \seq{\vec z} \varphi[\vec y/\vec x]$}
+    \DisplayProof
+    \hfill\phantom{a} \\[0.5em]
+    (``$\vec x = \vec y\,$'' is short for~``$x_1 = y_1 \wedge \cdots \wedge x_n =
+    y_n$''.)
+  \end{center}
+\end{frame}
+
+\newcommand{\realizes}{\Vdash}
+
+\begin{frame}{Number realizability}
+  \vspace*{-1em}
+  \begin{changemargin}{-1.5em}{-0.5em}
+  \begin{tabbing}
+    $e \models (\forall f\?\NN^\NN\_ \varphi(n))$ \= \kill
+    $e \realizes s = t$ \> iff $s = t$. \\
+    $e \realizes \top$ \> iff true. \\
+    $e \realizes \bot$ \> iff false. \\
+    $e \realizes (\varphi \wedge \psi)$ \> iff~$\pi_1 \cdot e \downarrow$
+    and~$\pi_2 \cdot e \downarrow$ and $\pi_1 \cdot e \realizes \varphi$
+    and~$\pi_2 \cdot e \realizes \psi$. \\
+    $e \realizes (\varphi \vee \psi)$ \> iff~$\pi_1 \cdot e \downarrow$
+    and~$\pi_2 \cdot e \downarrow$ and \\ \> \qquad if~$\pi_1 \cdot e = 0$
+    then~$\pi_2 \cdot e \realizes
+    \varphi$, and \\ \> \qquad if~$\pi_1 \cdot e \neq 0$ then~$\pi_2 \cdot e \realizes \psi$. \\
+    $e \realizes (\varphi \Rightarrow \psi)$ \> iff for every~$r \in \NN$
+    such that~$r \realizes \varphi$, $e \cdot r \downarrow$ and~$e \cdot r \realizes \psi$. \\
+    $e \realizes (\forall n\?\NN\_ \varphi(n))$ \> iff for every~$n_0
+    \in \NN$, $e \cdot n_0 \downarrow$ and~$e \cdot n_0 \realizes \varphi(n_0)$. \\
+    $e \realizes (\exists n\?\NN\_ \varphi(n))$ \> iff~$\pi_1 \cdot e
+    \downarrow$ and~$\pi_2 \cdot e \downarrow$
+    and~$\pi_2 \cdot e \realizes \varphi(\pi_1 \cdot e)$. \\
+    $e \realizes (\forall f\?\NN^\NN\_ \varphi(f))$ \> iff for every~$f_0
+    : \NN \to \NN$ and every~$r_0 \in \NN$ such that \\ \> \qquad $f_0$ is computed by the~$r_0$-th
+    machine, \\ \> \qquad
+    $e \cdot r_0 \downarrow$ and~$e \cdot r_0 \realizes \varphi(f_0)$. \\
+    $e \realizes (\exists f\?\NN^\NN\_ \varphi(f))$ \> iff~$\pi_1 \cdot e \downarrow$
+    and~$\pi_2 \cdot e \downarrow$ and
+    the $(\pi_1 \cdot e)$-th machine \\ \> \qquad computes a function~$f_0 : \NN \to \NN$
+    and $\pi_2 \cdot e \realizes \varphi(f_0)$.
+  \end{tabbing}\end{changemargin}
+
+  \mbox{\textbf{Thm.} If~$\text{HA} \vdash \varphi$, then there is a
+  number~$e \in \NN$ such that~\only<1>{$e \realizes
+  \varphi$}\only<2>{$\text{HA} \vdash (\underline{e} \realizes \varphi)$}.}
+\end{frame}
+
+\begin{frame}{Metatheory of Heyting arithmetic}
   \begin{enumerate}
-    \item Constructively, almost-full\textsubscript{ind} $\Rightarrow$
-    almost-full$_\infty$. \vspace*{-0.2em}
-    \item With \badbox{\textsc{lem}+\textsc{dc}}, almost-full\textsubscript{ind} $\Leftarrow$
-    almost-full$_\infty$.
-    %\pause
-    %\item How much stronger is almost-full\textsubscript{ind}?
-    %many-valued sequences, \ldots
+    \item \hil{Unprovability results:}
+    \medskip
+
+    There are instances of~\badbox{\textsc{lem}} which HA does not prove.
+    \bigskip
+
+    \item \hil{Disjunction property:}
+    \medskip
+
+    If HA proves~$\varphi \vee \psi$, then HA
+    proves~$\varphi$ or HA proves~$\psi$.
+    \bigskip
+
+    \item \hil{Existence property:}
+    \medskip
+
+    If HA proves~$\exists n\?N\_ \varphi(n)$, then there is a number~$n_0 \in
+    \NN$ such that HA proves~$\varphi(\underline{n_0})$.
+    \bigskip
+
+    \item \hil{Growth rate:}
+    \medskip\justifying
+
+    If HA proves~$\forall x\?N\_ \exists y\?N\_ \varphi(x,y)$,
+    then there exists a function $f_0 : \NN \to \NN$ which is \hil{higher
+    primitive recursive} such that for all~$x_0 \in \NN$, HA proves
+    $\varphi(\underline{x_0}, \underline{f_0(x_0)})$.
   \end{enumerate}
@@
 
-\begin{frame}{Constructive red flags}
-  A transitive relation~$<$ on a set~$X$ is \ldots
-  \begin{itemize}
-    \item \ \\[-1.2em]\mbox{\hil{well-founded$_\infty$} iff
-    there is no \bad{\uwave{infinite chain}} $x_0 > x_1 > \cdots$.}
-    \item \hil{well-founded'$_\infty$} iff
-    there is no \bad{\uwave{bad set}} (inhabited and such that for every member there is
-    a smaller member).
+\begin{frame}{Other machine models}
+  \begin{enumerate}
+    \item Turing machines
 
-    \item<4-> \hil{well-founded\textsubscript{ind}} iff for every~$x \in X$,
-    $\mathsf{Acc}(x)$: %, where~$\mathsf{Acc}$ is inductively defined by:
-    \[
-      \infer{\mathsf{Acc}(x)}{\forall y < x\_\ \mathsf{Acc}(y)}
-    \]
-    \vspace*{-1.3em}
-  \end{itemize}
+    \item Infinite-time Turing machines
 
-  \pause
-  \emph{Krull's Lemma:} An element~$x$ of a ring~$A$ is nilpotent if \ldots
-  \begin{itemize}
-    \item it is contained in
-    every \bad{\uwave{prime ideal}}.
-    \item<5-> the theory of prime ideals of~$A$ proves~``$x \in \ppp$''.
-  \end{itemize}
-  \medskip
+    \item Gödel's System T
 
-  \pause
-  \justifying
-  \emph{Dependent choice:}
-  Let~$R$ be a binary relation on a set~$X$
-  such that~$\forall a \in X\_ \exists b \in X\_ a \mathrel{R} b$. Let~$x_0
-  \in X$. Then there is an \bad{\uwave{infinite chain}} $x_0 \mathrel{R}
-  x_1 \mathrel{R} x_2 \mathrel{R} \cdots$.
-\end{frame}
+    \item Untyped lambda calculus
 
-%\begin{frame}{Inductive definitions}
-%  \begin{block}{}
-%    \justifying
-%    \textbf{Def.} A transitive relation~$<$ on a set~$X$ is \hil{well-founded} iff
-%    for all~$x \in X$, $\mathsf{Acc}(x)$, where the accessibility predicate is
-%    inductively generated by:
-%  \end{block}
-%  \vspace*{-0.6em}
-%  {\small\emph{Examples.} $(\NN,{<})$\par}
-%
-%  \textbf{Prop.} If~$<$ is well-founded, there are no infinite descending
-%  chains~$x_0 > x_1 > \ldots$.
-%
-%  \emph{Proof.} By induction on~$x_0$.
-%\end{frame}
-
-% \begin{document}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.59cm}\includegraphics[width=\paperwidth]{staircase}\end{minipage}}
-\begin{frame}{The set-theoretic multiverse}
-  \begin{tikzpicture}[remember picture,overlay]
-    \node[xshift=-1.5cm,yshift=-4cm] at (current page.north east)
-    {\includegraphics[width=2cm]{branching}};
-  \end{tikzpicture}
-  \vspace*{-2em}
-
-  \begin{block}{}
-    \justifying
-    \textbf{Def.} A \hil{model of set theory} is a (perhaps class-sized) structure $(M,{\in})$
-    satisfying axioms such as those of~\textsc{zfc}.
-  \end{block}
-  \vspace*{-0.6em}
-  {\small\emph{Examples.}\vspace*{-0.8em}
-  \begin{itemize}
-    \item $V$, the class of all sets \vspace*{-0.6em}
-    \item $L$, Gödel's constructible universe \vspace*{-0.6em}
-    \item $V[G]$, a forcing extension containing a generic filter~$G$ of \\ some
-    poset of forcing conditions \vspace*{-0.6em}
-    \item Henkin/term models from consistency of (extensions of)~\textsc{zfc}
-  \end{itemize}}
-  \pause
-
-  %{\centering\emph{Embracing all models of set theory:}\par}
-  \begin{block}{}
-    \justifying
-    \textbf{Def.} $\possible\varphi$ iff~$\varphi$ holds in \hil{some extension} of
-    the current universe. \\
-    \phantom{\textbf{Def.}} $\necessary\varphi$ iff~$\varphi$ holds in \hil{all extensions}
-    of the current universe.
-  \end{block}
-  \vspace*{-0.6em}
-  \begin{itemize}
-    \item $\necessary(\possible\textsc{CH} \wedge \possible\neg\text{CH})$,
-    the continuum hypothesis is a \hil{switch}
-    \item \ \\[-1.2em]\mbox{$\necessary\possible\necessary(\text{$X$ is countable})$,
-    existence of an enumeration is a \hil{button}}
-  \end{itemize}
-\end{frame}}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
-\begin{frame}{A topos-theoretic multiverse}
-  \begin{block}{}
-    \justifying
-    \textbf{Def.} A statement~$\varphi$ holds \ldots
-    \begin{enumerate}
-      \item\small \hil{everywhere} ($\necessary\varphi$) iff
-      it holds in every (Grothendieck) topos (over the current base topos).
-      \item \hil{somewhere} ($\possible\varphi$) iff
-      it holds in some positive topos.
-      \item \hil{proximally} ($\xpossible\varphi$) iff
-      it holds in some positive ouvert topos.
-    \end{enumerate}
-  \end{block}
-  \pause
-
-  \newcommand{\seqbad}{\only<3>{infinite descending sequence}\only<4->{bad set}}
-  \begin{changemargin}{-2em}{-1.5em}
-  \begin{itemize}
-  \small
-    \item \justifying
-    \only<1-4>{A relation is almost-full\textsubscript{ind} \\
-    iff every infinite sequence \emph{everywhere} is good \\
-    iff the \emph{generic infinite sequence} is good \\
-    iff the \emph{theory of infinite sequences} proves goodness.\medskip}
-    \only<5->{A relation is almost-full\textsubscript{ind} iff every infinite
-    sequence \emph{everywhere} is good.}
-    \pause
-
-    \item\justifying
-    \only<3-4>{A relation is well-founded\textsubscript{ind} \\
-    iff \emph{everywhere} there are no \seqbad s \\
-    iff the \emph{generic \seqbad} validates~$\bot$ \\
-    iff the \emph{theory} of \seqbad s proves~$\bot$.}
-    \only<5->{A relation is well-founded\textsubscript{ind} iff
-    \emph{nowhere} there are \seqbad s.}
-    \pause
-    \pause
-    \pause
-
-    \item \only<2-6>{A ring element~$x$ is nilpotent \\
-    iff \emph{everywhere} it is contained in all prime ideals \\
-    iff it is contained in the \emph{generic prime ideal} \\
-    iff the \emph{theory} of prime ideals proves~$x \in \ppp$.}
-    \only<7->{\mbox{A ring element is nilpotent iff all prime
-    ideals \emph{everywhere} contain it.}}
-
-    \pause
-    \pause
-    \item Given a set~$X$, \emph{proximally} there is a surjection~$\NN
-    \twoheadrightarrow X$. \\
-    \hil{NB:} $(\possible\varphi) \Rightarrow \varphi$, if~$\varphi$ is a
-    geometric implication. \\
-    \phantom{\hil{NB:}} $(\xpossible\varphi) \Rightarrow \varphi$, if~$\varphi$
-    is first-order. \\
-    \phantom{\hil{NB:}} $\varphi \Rightarrow (\necessary\varphi)$,\, if~$\varphi$
-    is a geometric formula.
-
-    \pause
-    \item Given~$(X,R,x_0)$ as in~\textsc{dc}, \emph{proximally} there is an
-    infinite chain.
-
-    \pause
-    \item \emph{Somewhere,} the law of excluded middle holds.
-  \end{itemize}
-  \end{changemargin}
-\end{frame}}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.05cm}\includegraphics[width=\paperwidth]{agda-af-implies-wf}\end{minipage}}
-\begin{frame}{Extracting constructive content}
-  \begin{block}{}
-    \justifying
-    \textbf{Prop.} Let~$({\leq})$ be a transitive almost-full\textsubscript{ind} relation.
-    Then~$({<})$, where~$x < y \equiv (x \leq y \wedge \neg(y \leq x))$,
-    is well-founded\textsubscript{ind}.
-  \end{block}
-  \vspace*{-0.5em}
-
-  \emph{Proof.} Everywhere, there can be no infinite descending chain, as any
-  such would also be good. \qed
-
-  \mbox{Unrolling this proof gives a program~$(\textsf{Good} \,|\, []) \to
-  \prod_{x \? X} \textsf{Acc}(x)$.}
-  \vfill
-\end{frame}}
-
-% \begin{document}
-\addtocounter{framenumber}{-1}
-\begin{frame}{Extracting constructive content}
-  \begin{block}{}
-    \justifying
-    \textbf{Prop.} [Dickson] If~$X$ and~$Y$ are almost-full\textsubscript{ind},
-    so is~$X \times Y$.
-  \end{block}
-  \pause
-
-  \vspace*{-0.3em}
-  \emph{Proof.}\small
-  \vspace*{-0.6em}
-  \begin{enumerate}
-  \item\justifying It suffices to verify that the \emph{generic infinite
-  sequence}~$\gamma = (\alpha,\beta) : \NN \to X \times Y$ is good. \pause Since
-  being good can be put as a geometric implication (in fact, a geometric
-  formula) and since \textsc{lem} holds \emph{somewhere}, we may assume~\textsc{lem}.\pause
-  \item By \textsc{lem} and well-foundedness, there is a minimal value~$\alpha(i_0)$
-  among all values of~$\alpha$. \vspace*{-0.2em}\pause
-  Similarly, there is a minimal value~$\alpha(i_1)$ among~$(\alpha(n))_{n >
-  i_0}$, \pause a minimal value~$\alpha(i_2)$ among~$(\alpha(n))_{n > i_1}$, and so
-  on. \pause By \emph{proximal dependent choice}, we can proximally collect these indices
-  into a function~$i : \NN \to \NN$. \vspace*{-0.2em}\pause
-  \item Switching~\textsc{lem} on again, there is a minimal value~$\beta(i(k_0))$
-  among all values of~$\beta \circ i$. \pause Hence~$\gamma$ is good in view of
-  \begin{align*}
-    \alpha(i(k_0)) &\leq \alpha(i(k_0+1)), &
-    \beta(i(k_0)) &\leq \beta(i(k_0+1)).
-  \end{align*}
+    \item Machines in the real world
   \end{enumerate}
@@
 
-Mit LEM hat jede Sequenz α ein Minimum:
-Induktion über einen Wert, der vorkommt (zum Beispiel α(0)).
+start:
+- infinitude of primes
+- Dickson's lemma
+- "Every constructive theorem has a computable witness."
+- induction = recursion, Markov = unbounded search, ...
+- side effects
+- applications:
+  - integrated developments (for instance SAT checking)
+  - metatheory of constructive systems
+  - philosophy
+  - exploring computability theory
 
-Mit LEM gilt: ∃n. α(n) < x  oder  nicht.
-Im ersten Fall fertig nach IV.
-Im zweiten Fall:
-  Sei n beliebig. Weiß ¬(α(n) < x).
-  Also weiß: ¬(α(n) ≤ x) und ¬¬(x ≤ α(n)).
-  Mit LEM also x ≤ α(n).
+definition:
+- HA
+- TM
+- realizability (as formalization of BKH)
+- proof
 
-Mit DC bekomme Folge:
-i₀ so, dass α(i₀) Minimum.
-i₁ so, dass α(i₁) Minimum unter α(i₀+1), α(i₀+2), ...
+examples:
+- as in filmat paper
+  - also with lambda terms!
 
-Betrachte Folge (β(iṉ))ṉ.
-Mit LEM finde Minimum: β(ik).
-Dann β(ik) ≤ β(i(k+1)) und natürlich auch α(ik) ≤ α(i(k+1)).
+special kinds of statements:
+- negated: uninformative
+- double negation
+- Π⁰₁
 
-x < y  heißt:  x ≤ y und ¬(y ≤ x)
-Behauptung: ≤ af und transitiv, dann < wf.
+metatheory of HA:
+- disjunction property
+- existence property
+- unprovability of (for instance) halting
+- growth rate
 
-Betrachte generische absteigende Folge.
-Da gut, habe x[i] ≤ x[j].
-Wegen Transitivität auch x[i] > x[j],
-also x[j] < x[i], also x[j] ≤ x[i] und ¬(x[i] ≤ x[j]).
-Widerspruch.
+variants:
+- ITTM
+- real world
+- completion to Eff
 
-DC
-Krull
-Countability
-Well
-Well-founded
+----
 
-Choice:
-erst abzählbar machen,
-dann DC
-nein, funktioniert nicht ganz
-aber für diskrete Quelle habe surjektive (nicht offene?) Theorie/Topos
+- play
+- double-negation embedding
+- Friedman's trick
+- perhaps in Agda
 
-∀x ∈ X. ∃y ∈ Y. φ(x,y).
-(Toposwechsel für Abzählbarkeit von X)
-∀n ∈ ℕ. ∃y ∈ Y. φ(πn,y).
-Yn := { y ∈ Y | φ(πn,y) }
-(~) auf M = (coprod_n Y_n): (n,y) ~ (n',y') gdw. n' = n+1
-Dann habe: M bewohnt (durch (0,...)) und ∀p ∈ M. ∃q ∈ M. p ~ q.
-(Toposwechsel für DC für (~))
-Habe f : ℕ → M mit f(0) = (0,...) und f(n) ~ f(n+1) für alle n.
-Also ist π₂.f eine Auswahlfunktion, genauer: ∀n. φ(πn, π₂(f(n))).
-
-Zorn für abzählbare Mengen?
-C_0 = {}
-C_{n+1} = C_n cup { x_n | C_n cup {x_n} Kette }
-C = bigcup_n C_n
-z = sup(C)
-Dann ist z maximal?
-Gelte z ≤ e.
-Dann ist C cup { e } Kette.
-Wenn e = x_n0, so insbesondere C_n0 cup { e } Kette.
-Also e ∈ C_{n+1} ⊆ C, also e ≤ z.
-Also e = z.
-Problem: Kommt eher selten vor; Supremum wird oft nicht zur Menge gehören.
-Beispiel: Supremum von Menge endlich erzeugter Ideale ist nicht endlich erzeugt.
-
-Attribution:
-Blass(?)
-Alexander
-Henry
-
-0: title
-* expectations: no new method for extracting constructive content, rather
-  point of view for putting previous work in a uniform context;
-  philosophy turned mathematics
-
-1: infinite data
-* how much more general? multivalued sequences, "¬¬ sequences", ...
-* where do these definitions come from? conceptual answer
-
-3: set-theoretic multiverse
-* axioms of ZFC massively underspecify the concept of sets;
-  models for exploring the range of set-theoretic possibility
-* to settle a question shown to be independent from ZFC is to
-  understand how it behaves in the multiverse
-
-explanation of generic models
+\begin{block}{}
+  \justifying
+  \textbf{Thm.}
+  Every pair of infinite sequences~$\alpha,\beta : \NN \to \NN$ is \emph{good} in that
+  there are~$i < j$ such that~$\alpha(i)
+  \leq \alpha(j)$ and~$\beta(i) \leq \beta(j)$.
+\end{block}
+\vspace*{-0.6em}
+{\small\emph{Proof.} By~\badbox{\textsc{lem}}+\badbox{\textsc{dc}}, there is a
+monotonic subsequence
+\[ \alpha(i_0) \leq \alpha(i_1) \leq \alpha(i_2) \leq \cdots \]
+with~$i_0 < i_1 < \cdots$. By~\badbox{\textsc{lem}}, there is a minimal
+value~$\beta(i_{k_0})$ among all values of~$(\beta(i_k))_k$. Then
+$\alpha(i_{k_0}) \leq \alpha(i_{k_0+1})$ and
+$\beta(i_{k_0}) \leq \beta(i_{k_0+1})$. \qed
+\par}

### 2022-09-26 slides-fischbachau2022.tex
Honouring Ada Lovelace
More work on slides

@@
 \begin{frame}{}
+  \medskip
   \textbf{Thm.}
@@
 
-  {\centering\emph{``Every theorem has a computable witness.''}\par}
+  {\centering\emph{``Every theorem has a computable* witness.''} \\ \scriptsize
+  * with monadic side effects\par}
+\end{frame}}
+
+% \includegraphics[width=3cm]{lovelace-babbage}
+% \includegraphics[width=3cm]{first-program}
+\addtocounter{framenumber}{-1}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[height=\paperheight,valign=t]{ada-lovelace}\includegraphics[width=8cm,valign=t]{first-program}\end{minipage}}
+\begin{frame}[plain]
+  \vspace*{6cm}\hspace*{6cm}\begin{minipage}{4.7cm}
+    \huge\hil{Ada Lovelace},
+
+    \Large
+    the world's first \\
+    computer programmer
+    \medskip
+
+    * 1815 \ \ † 1852
+  \end{minipage}
 \end{frame}}
@@
 
-    There are instances of~\badbox{\textsc{lem}} which HA does not prove.
+    There are instances of~\badbox{\textsc{lem}} which HA does not prove,
+    such as ``every Turing machine terminates or does not terminate''.
     \bigskip
@@
     If HA proves~$\forall x\?N\_ \exists y\?N\_ \varphi(x,y)$,
-    then there exists a function $f_0 : \NN \to \NN$ which is \hil{higher
-    primitive recursive} such that for all~$x_0 \in \NN$, HA proves
+    then there exists a \hil{higher primitive recursive} function $f_0 : \NN
+    \to \NN$ such that for all~$x_0 \in \NN$, HA proves
     $\varphi(\underline{x_0}, \underline{f_0(x_0)})$.

### 2022-09-26 slides-fischbachau2022.tex
Slides finished for today

@@
 \newcommand{\RelSpec}{\operatorname{Spec}}
-\renewcommand{\_}{\mathpunct{.}}
+\renewcommand{\_}{\mathpunct{.}\,}
 \newcommand{\?}{\,{:}\,}
@@
     \item as function symbols: $0$, $S$, $+$, $\cdot$
-    \item as relation symbols: $=$
+    \item as its single relation symbol: $=$
   \end{itemize}
@@
 
-\begin{frame}{Other machine models}
+\begin{frame}{Range of machine models}
   \begin{enumerate}
-    \item Turing machines
+    \item \hil{Turing machines}
 
-    \item Infinite-time Turing machines
+    \visible<2->{``Every function $\NN \to \NN$ is computable'' is realized by
+    \textsf{cat}.}
+    \medskip
 
-    \item Gödel's System T
+    \item \hil{Untyped lambda calculus}
 
-    \item Untyped lambda calculus
+    \visible<3->{``Every function $\NN \to \NN$ is computable'' is \emph{not}
+    realized.}
+    \medskip
 
-    \item Machines in the real world
+    \item \hil{Infinite-time Turing machines}
+
+    \visible<4->{``Every function $\NN \to \NN$ has a zero or not'' is realized by an
+    infinite search.}
+    \medskip
+
+    \item \hil{Gödel's System T}
+
+    \visible<5->{Markov's principle is not realized.}
+    \medskip
+
+    \item \hil{Machines in the real world} -- philosophical
+
+    \justifying
+    \visible<6->{``Every function $\RR \to \RR$ is continuous'' is realized if,
+    in the physical world, only finitely many computational
+    steps can be carried out in finite time and if it is possible to form
+    tamper-free private communication channels.}
   \end{enumerate}

### 2022-09-27 slides-fischbachau2022.tex
typo

@@
     such that~$r \realizes \varphi$, $e \cdot r \downarrow$ and~$e \cdot r \realizes \psi$. \\
-    $e \realizes (\forall n\?\NN\_ \varphi(n))$ \> iff for every~$n_0
+    $e \realizes (\forall n\?N\_ \varphi(n))$ \> iff for every~$n_0
     \in \NN$, $e \cdot n_0 \downarrow$ and~$e \cdot n_0 \realizes \varphi(n_0)$. \\
-    $e \realizes (\exists n\?\NN\_ \varphi(n))$ \> iff~$\pi_1 \cdot e
+    $e \realizes (\exists n\?N\_ \varphi(n))$ \> iff~$\pi_1 \cdot e
     \downarrow$ and~$\pi_2 \cdot e \downarrow$
     and~$\pi_2 \cdot e \realizes \varphi(\pi_1 \cdot e)$. \\
-    $e \realizes (\forall f\?\NN^\NN\_ \varphi(f))$ \> iff for every~$f_0
+    $e \realizes (\forall f\?N^N\_ \varphi(f))$ \> iff for every~$f_0
     : \NN \to \NN$ and every~$r_0 \in \NN$ such that \\ \> \qquad $f_0$ is computed by the~$r_0$-th
@@
     $e \cdot r_0 \downarrow$ and~$e \cdot r_0 \realizes \varphi(f_0)$. \\
-    $e \realizes (\exists f\?\NN^\NN\_ \varphi(f))$ \> iff~$\pi_1 \cdot e \downarrow$
+    $e \realizes (\exists f\?N^N\_ \varphi(f))$ \> iff~$\pi_1 \cdot e \downarrow$
     and~$\pi_2 \cdot e \downarrow$ and

### 2022-09-27 slides-fischbachau2022.tex
Slide with examples

@@
 \usepackage{bussproofs}
+\usepackage{xspace}
 \usepackage{amssymb}
@@
 \newcommand{\grayline}{\textcolor{gray}{\hspace*{-3em}\hrulefill\hspace*{-3em}}}
+\newcommand{\notnot}{\emph{not~not}\xspace}
 
@@
 
+% \begin{document}
+
+\newcommand{\expl}[2]{
+  \justifying
+  ``$\realizes \!\text{\normalnumber{#1}}$'' amounts to: #2
+}
+
+\newcommand{\qswitch}[3]{\only<1-#1>{\includegraphics[height=0.7em]{question-mark}}\only<#2->{#3}}
+\newcommand{\ccmark}{\good{\cmark}}
+\newcommand{\cxmark}{\bad{\xmark}}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{0.00cm}\hfill\mbox{\includegraphics[width=0.2\textwidth]{turing-machine}\;\;}\end{minipage}}
+\begin{frame}{Exploring the realizability model}
+  \medskip
+  \medskip
+  \medskip
+  \begin{tabular}{@{\!\!\!\!\!\!\!}l@{\,}llp{1.8cm}}
+    \toprule
+    & Statement & classical? & realizable? \\
+    \midrule
+    \normalnumber{1} & Every number is prime or not prime. & \ccmark{}
+    (trivially) & \ccmark \\
+    \normalnumber{2} & After every number there is a prime. & \ccmark & \ccmark \\
+    \normalnumber{3} & Every map $\NN \to \NN$ has a zero or not. & \ccmark{} (trivially) & \cxmark \\
+    \normalnumber{4} & Every map $\NN \to \NN$ is computable. & \cxmark &
+    \qswitch{4}{5}{\ccmark}\only<1-4>{\,} \visible<5->{(trivially)} \\
+    \normalnumber{5} & Every map $\RR \to \RR$ is continuous. & \cxmark &
+    \qswitch{5}{6}{\ccmark} \\
+    \normalnumber{6} & Markov's principle holds. & \cmark{} (trivially) &
+    \qswitch{6}{7}{\ccmark{} (if MP)} \\
+    \normalnumber{7} & Countable choice holds. & \xmark &
+    \qswitch{7}{8}{\ccmark{} (always!)} \\
+    \normalnumber{8} & Heyting arithmetic is categorical. & \xmark &
+    \qswitch{8}{9}{\ccmark{} (if MP)} \\
+    \bottomrule
+  \end{tabular}
+  \medskip
+
+  \only<1>{\ \\\ \\\ \\\ }
+  \only<2>{\expl{1}{There is a machine which determines of any given
+  number whether it is prime or not. \\\ }}
+  \only<3>{\expl{2}{There is a machine which, given a number~$n$, computes a
+  prime larger than~$n$. \\\ \\\ }}
+  \only<4>{\expl{3}{There is a machine which, given a machine
+  computing a map~$f : \NN \to \NN$, determines whether~$f$ is constantly
+  zero or not. \\\ }}
+  \only<5>{\expl{4}{There is a machine which, given a machine
+  computing a map~$f : \NN \to \NN$, outputs a machine
+  computing~$f$. \\\ }}
+  \only<6>{\ \\\ \\\ \\\ }
+  \only<7>{\expl{6}{There is a machine which, given a machine
+  computing a map~$f : \NN \to \NN$ and given the promise that it is \notnot
+  the case that~$f$ has a zero, determines a zero of~$f$.}}
+  \only<8->{\expl{7}{\justifying There is a machine which, given a machine
+  computing for every~$x \in \NN$ some~$y \in A$ together with a realizer
+  of~$\varphi(x,y)$, outputs a machine computing a suitable choice
+  function~$\NN \to A$.}}
+\end{frame}}
+
 \begin{frame}{Metatheory of Heyting arithmetic}

### 2022-09-27 paper-modal-multiverse.tex
Two basic examples

@@
 
-xxx: examples from slides
+\begin{prop}Let~$({\leq})$ be a transitive almost-full relation.
+Then~$({<})$, where~$x < y \equiv (x \leq y \wedge \neg(y \leq x))$,
+is well-founded.\end{prop}
+
+\begin{proof}Everywhere, there can be no infinite descending chain, as any
+such would also be good.\end{proof}
+
+Unrolling this proof gives a program of type~$(\textsf{Good} \,|\, []) \to
+\prod_{x:X} \textsf{Acc}(x)$.
+
+\begin{thm}[Dickson's lemma] If~$X$ and~$Y$ are almost-full,
+so is~$X \times Y$.\end{thm}
+
+\begin{proof}
+  \begin{enumerate}
+    \item It suffices to verify that the \emph{generic infinite
+    sequence}~$\gamma = (\alpha,\beta) : \NN \to X \times Y$ is good. Since
+    being good can be put as a geometric implication (in fact, a geometric
+    formula) and since \textsc{lem} holds \emph{somewhere}, we may assume~\textsc{lem}.
+
+    \item By \textsc{lem} and well-foundedness, there is a minimal value~$\alpha(i_0)$
+    among all values of~$\alpha$. \vspace*{-0.2em}
+    Similarly, there is a minimal value~$\alpha(i_1)$ among~$(\alpha(n))_{n >
+    i_0}$, a minimal value~$\alpha(i_2)$ among~$(\alpha(n))_{n > i_1}$, and so
+    on. By \emph{proximal dependent choice}, we can proximally collect these indices
+    into a function~$i : \NN \to \NN$; this switches~\textsc{lem} off. \vspace*{-0.2em}
+
+    \item Switching~\textsc{lem} on again, there is a minimal value~$\beta(i(k_0))$
+    among all values of~$\beta \circ i$. Hence~$\gamma$ is good in view of
+    \begin{align*}
+      \alpha(i(k_0)) &\leq \alpha(i(k_0+1)), &
+      \beta(i(k_0)) &\leq \beta(i(k_0+1)). \qedhere
+    \end{align*}
+  \end{enumerate}
+\end{proof}
 

### 2022-09-27 slides-fischbachau2022.tex
typo
Title slide
Introduction to ¬¬-embedding

@@
 \usepackage{ifthen}
+\usepackage{bbm}
 \usepackage[normalem]{ulem}
@@
 
+\section{Realizability theory}
+
+\begin{frame}
+  \centering\bigskip
+  \includegraphics[height=9.5em]{turing-machine}
+  \medskip
+
+  \large Lecture I: \\
+  \hil{Realizability theory}
+
+  \normalsize
+
+  \emph{for extracting programs from constructive proofs}
+\end{frame}
+
 \begin{frame}{Heyting arithmetic}
@@
     \qswitch{5}{6}{\ccmark} \\
-    \normalnumber{6} & Markov's principle holds. & \cmark{} (trivially) &
+    \normalnumber{6} & Markov's principle holds. & \ccmark{} (trivially) &
     \qswitch{6}{7}{\ccmark{} (if MP)} \\
-    \normalnumber{7} & Countable choice holds. & \xmark &
+    \normalnumber{7} & Countable choice holds. & \ccmark &
     \qswitch{7}{8}{\ccmark{} (always!)} \\
-    \normalnumber{8} & Heyting arithmetic is categorical. & \xmark &
+    \normalnumber{8} & Heyting arithmetic is categorical. & \cxmark &
     \qswitch{8}{9}{\ccmark{} (if MP)} \\
@@
 
-    \visible<4->{``Every function $\NN \to \NN$ has a zero or not'' is realized by an
+    \visible<4->{``Every function $\NN \to \NN$ has a zero or not'' is realized by
     infinite search.}
@@
   \end{enumerate}
+\end{frame}
+
+\section{Proof transformations}
+
+\begin{frame}
+  \centering\bigskip
+  \includegraphics[height=9.5em]{philosophers-stone}
+  \medskip
+
+  \large Lecture II: \\
+  \hil{Proof transformations}
+
+  \normalsize
+
+  \emph{for extracting constructive proofs from classical proofs}
+\end{frame}
+
+\begin{frame}{The double-negation embedding}
+  \textbf{Thm.} Every infinite sequence~$\alpha : \NN \to \NN$ is \emph{good}
+  in that there are numbers~$i < j$ such that~$\alpha(i) \leq \alpha(j)$.
+
+  {\emph{Proof.} By~\badbox{\textsc{lem}}, there is a minimal value~$\alpha(i)$.
+  Set~$j \defeq i + 1$.\par}
+
+  \textbf{Thm.} Every infinite binary sequence~$\alpha : \NN \to \mathbbm{2}$
+  contains repeating entries.
+
+  {\emph{Proof.} By the \bad{infinite box principle}, which follows
+  from~\badbox{\textsc{lem}}, the sequence contains infinitely many zeros or
+  infinitely many ones. In either case the claim follows.\par}
+
+  How to extract \good{constructive proofs} from these
+  \bad{classical proofs}?
+  \pause
+
+  \grayline
+
+  \textbf{Thm.} Let~$X$ be an inhabited set of natural numbers. Then~$X$
+  contains a minimal element.
+
+  {\emph{Proof.} By~\badbox{\textsc{lem}}, \ldots\par}
 \end{frame}

### 2022-09-27 slides-fischbachau2022.tex
¬¬-motivation finished

@@
 \begin{frame}{The double-negation embedding}
-  \textbf{Thm.} Every infinite sequence~$\alpha : \NN \to \NN$ is \emph{good}
+  \vspace*{-0.8em}
+  \mbox{How to extract \good{constructive proofs} from
+  \bad{classical proofs} as the following?}
+
+  \begin{enumerate}
+    \item \textbf{Thm.} Every infinite sequence~$\alpha : \NN \to \NN$ is \emph{good}
     in that there are numbers~$i < j$ such that~$\alpha(i) \leq \alpha(j)$.
+    \medskip
 
@@
     Set~$j \defeq i + 1$.\par}
+    \medskip
+    \pause
 
-  \textbf{Thm.} Every infinite binary sequence~$\alpha : \NN \to \mathbbm{2}$
-  contains repeating entries.
+    \item \textbf{Thm.} Every infinite binary sequence
+    contains repeated terms.
+    \medskip
 
-  {\emph{Proof.} By the \bad{infinite box principle}, which follows
-  from~\badbox{\textsc{lem}}, the sequence contains infinitely many zeros or
-  infinitely many ones. In either case the claim follows.\par}
-
-  How to extract \good{constructive proofs} from these
-  \bad{classical proofs}?
+    {\emph{Proof.} By the \bad{infinite box principle},
+    infinitely many terms are zeros or are ones. In either case the claim
+    follows.\par}
     \pause
+  \end{enumerate}
+  \vspace*{-0.7em}
 
@@
 
-  \textbf{Thm.} Let~$X$ be an inhabited set of natural numbers. Then~$X$
-  contains a minimal element.
+  \textbf{Thm.} Let~$X$ be an inhabited set of natural numbers.
+  \begin{enumerate}
+    \item If~\badbox{\textsc{lem}}, then~$X$ contains a minimal element.
+    \item<5-> If~$X$ is \hil{detachable}, then~$X$ contains a minimal element.
+    \item<7-> It is \notnot the case that~$X$ contains a minimal element.
+  \end{enumerate}
 
-  {\emph{Proof.} By~\badbox{\textsc{lem}}, \ldots\par}
+  \justifying
+  \only<1-3>{\ \\\ \\\ \\\ }
+  \only<4-5>{\emph{Proof of~\normalnumber{1}.} There is some~$n \in X$.
+  By~\badbox{\textsc{lem}}, either~$\exists k \in X\_ k < n$ \\ or not. In the
+  first case, we continue by induction. Else~$n$ is minimal.\\\ \par}
+  \only<6-7>{\mbox{\emph{Proof of~\normalnumber{2}.} There is some~$n \in X$.
+  By~\good{assumption}, either~$\exists k \in X\_ k < n$} or not. In the
+  first case, we continue by induction. Else~$n$ is minimal.\\\ \par}
+  \only<8>{\emph{Proof of~\normalnumber{3}.} There is some~$n \in X$. Assume that~$X$ does not
+  contain a minimum. Then it is not the case that~$\exists k \in X\_ k <
+  n$, as else~$\bot$ by induction. Hence~$n$ is minimal. This is a
+  contradiction.\par}
+\end{frame}
+
+\section{Proof transformations}
+
+\begin{frame}
+  \centering\bigskip
+  \includegraphics[height=9.5em]{phantoms}
+  \medskip
+
+  \large Lecture III: \\
+  \hil{Extracting constructive proofs from invalid* proofs}
+
+  \normalsize
 \end{frame}
@@
 \par}
+
+Claim: (∃n ∈ X) ⇒ X has a minimum.
+
+Classical proof:
+By LEM, ∃(k ∈ X). k < n or not.
+In the first case, we conclude by induction.
+In the second case, we conclude that n is minimal:
+  Given k ∈ X, we have k < n or n ≤ k.
+  In the first case, ⊥ and hence conclusion.
+  In the second case, good.
+
+Translated proof:
+Assume that X does not have a minimum, show ⊥.
+May conclude that ¬(∃(k ∈ X). k < n). For else, X inhabited by <n, so ⊥ by induction.
+Thus n is a minimum:
+  Given k ∈ X, we have k < n or n ≤ k.
+  In the first case, ⊥ and hence conclusion.
+  In the second case, good.

### 2022-09-27 slides-fischbachau2022.tex
Slides finished for tomorrow

@@
 \usepackage{ifthen}
-\usepackage{bbm}
 \usepackage[normalem]{ulem}
@@
 
-% \begin{document}
-
 \newcommand{\expl}[2]{
@@
 
-\begin{frame}{The double-negation embedding}
+\begin{frame}{A case study in double negation}
   \vspace*{-0.8em}
@@
 
-    \item \textbf{Thm.} Every infinite binary sequence
-    contains repeated terms.
+    \item \textbf{Thm.} Every infinite binary sequence contains repeated terms.
     \medskip
@@
   \begin{enumerate}
-    \item If~\badbox{\textsc{lem}}, then~$X$ contains a minimal element.
+    \item Assuming~\badbox{\textsc{lem}}, the set~$X$ contains a minimal element.
     \item<5-> If~$X$ is \hil{detachable}, then~$X$ contains a minimal element.
-    \item<7-> It is \notnot the case that~$X$ contains a minimal element.
+    \item<7-> It is \hil{not~not} the case that~$X$ contains a minimal element.
   \end{enumerate}
@@
 
-\section{Proof transformations}
+\newcommand{\hnegneg}{\hil{$\neg\neg$}}
+\newcommand{\negg}{{\neg}\!\!\!{\neg}}
+\newcommand{\neggv}{{\neg}\!\!\!\phantom{{\neg}}}
+\newcommand{\bott}{\bot\!\!\!\!\bot}
+\newcommand{\bottv}{\bot\!\!\!\!\phantom{\bot}}
+
+\begin{frame}{The double-negation embedding}
+  \justifying
+  \vspace*{-1em}
+  \small
+  \textbf{Def.} For formulas over a fixed first-order signature, the
+  \hil{$\neg\neg$-translation} $\varphi \mapsto \varphi^{\neg\neg}$ is defined
+  by the following clauses.
+  \begin{align*}
+    (\varphi_\text{atomic})^{\neg\neg} &\defeqv \hnegneg \varphi_\text{atomic} &
+    (\varphi \Rightarrow \psi)^{\neg\neg} &\defeqv (\varphi^{\neg\neg} \Rightarrow \psi^{\neg\neg}) \\
+    \bot^{\neg\neg} &\defeqv \hnegneg\bot &
+    \top^{\neg\neg} &\defeqv \top \\
+    (\varphi \vee \psi)^{\neg\neg} &\defeqv \hnegneg(\varphi^{\neg\neg} \vee \psi^{\neg\neg}) &
+    (\varphi \wedge \psi)^{\neg\neg} &\defeqv (\varphi^{\neg\neg} \wedge \psi^{\neg\neg}) \\
+    (\exists x\?X\_ \varphi)^{\neg\neg} &\defeqv \hnegneg{(\exists x\?X\_ \varphi^{\neg\neg})} &
+    (\forall x\?X\_ \varphi)^{\neg\neg} &\defeqv (\forall x\?X\_ \varphi^{\neg\neg}) \\
+  \end{align*}
+  \vspace*{-3em}
+
+  \mbox{\textbf{Ex.} $(\forall a{:}X\_\! \exists b{:}X\_\! a = b \vee \ldots)^{\neg\neg}
+  \equiv (\forall a{:}X\_\! \hnegneg\exists b{:}X\_\! \hnegneg(\hnegneg(a=b) \vee
+  (\ldots)^{\neg\neg})).$}
+  \pause
+
+  \textbf{Prop.} Classically, $\varphi \Leftrightarrow \varphi^{\neg\neg}$.
+  \pause
+  \vspace*{-0.5em}
+
+  \grayline
+
+  \textbf{Thm.} For every formula~$\varphi$ and set of formulas~$\Gamma$:
+  \vspace*{-0.6em}
+  \begin{enumerate}
+    \item Minimally, $\neg\neg(\varphi^{\neg\neg}) \Rightarrow
+    \varphi^{\neg\neg}$.
+    \item Minimally, $\varphi^{\neg\neg} \Leftrightarrow \neg\neg\varphi$ in
+    case that~$\varphi$ is geometric
+    ($R{\top}{\bot}{\wedge}{\vee}{\exists}{\bigvee}$).
+    \item Classical logic proves~$\varphi$ from~$\Gamma$
+    iff minimal logic proves~$\varphi^{\neg\neg}$ from~$\Gamma^{\neg\neg}$.
+  \end{enumerate}
+  \pause
+  \vspace*{-0.3em}
+
+  \textbf{Cor.} If PA proves~$\varphi$, then~HA proves~$\varphi^{\neg\neg}$.
+  \pause
+
+  \textbf{Rem.} Theorem and corollary hold for every \hil{local
+  operator}~$\nabla$ in place of~$\neg\neg$, in particular for~$\negg\negg\varphi \defeq
+  ((\varphi \Rightarrow \bott) \Rightarrow \bott)$ for some arbitrary
+  formula~$\bott$.
+\end{frame}
+
+\newcommand{\negx}{\hil{${\only<1>{\neggv}\only<2->{\negg}}$}}
+\newcommand{\botx}{\hil{${\only<1>{\bottv}\only<2>{\bott}\only<3->{\beta}}$}}
+
+\begin{frame}{Barr's theorem / Friedman's trick / A-translation}
+  \textbf{Thm.} Let~$\Gamma$ be a set of geometric sequents over a fixed
+  signature. Let~$\sigma$ be a geometric sequent. Then the following are
+  equivalent:
+  \begin{enumerate}
+    \addtocounter{enumi}{-1}
+    \item $\sigma$ holds for the \good{generic model} of~$\Gamma$ (in its
+    classifying topos).
+    \item $\sigma$ is provable from~$\Gamma$ in \good{geometric logic}.
+    \item $\sigma$ is provable from~$\Gamma$ in \good{intuitionistic logic}.
+    \item $\sigma$ is provable from~$\Gamma$ in \bad{classical logic}.
+    \item (Assuming \badbox{\textsc{ac}}) $\sigma$ is provable from~$\Gamma$ in
+    \bad{classical logic with~\textsc{ac}}.
+  \end{enumerate}
+
+  \emph{Proof of~``\,\,$\normalnumber{3} \Rightarrow \normalnumber{2}$''.}
+  Write~$\sigma \equiv (\alpha \seq{\vec x} \beta)$. Then intuitionistically,
+  \[ \alpha \Longrightarrow
+    \negx\negx\alpha \Longleftrightarrow
+    \alpha^{\negx\negx} \Longrightarrow
+    \beta^{\negx\negx} \Longleftrightarrow
+    \negx\negx\beta \equiv
+    ((\beta{\Longrightarrow}\botx){\Rightarrow}\botx)
+    \visible<4->{\Longrightarrow \beta.}
+  \]
+\end{frame}
+
+\section{Constructive proofs from invalid proofs}
 
@@
 \end{frame}
+
+\addtocounter{framenumber}{-1}
 
@@
   In the second case, good.
+
+\begin{tabular}{ll@{\qquad\qquad}ll}
+  \toprule
+  $\varphi$ & $\varphi^{\neg\neg}$ &
+  $\varphi$ & $\varphi^{\neg\neg}$ \\\midrule
+  $\alpha_\text{atomic}$ & $\hnegneg \alpha_\text{atomic}$ &
+  $\alpha \Rightarrow \beta$ & $\alpha^{\nabla} \Rightarrow \beta^{\nabla}$ \\
+  $\bot^{\neg\neg}$ & $\hnegneg\bot$ &
+  $\top^{\neg\neg}$ & $\top$ \\
+  $\alpha \vee \beta$ & $\hnegneg(\alpha^{\neg\neg} \vee \beta^{\neg\neg})$ &
+  $\alpha \wedge \beta$ & $\alpha^{\neg\neg} \wedge \beta^{\neg\neg}$ \\
+  \bottomrule
+\end{tabular}

### 2022-09-27 slides-fischbachau2022.tex
Final touches

@@
   \medskip
-  \begin{tabular}{@{\!\!\!\!\!\!\!}l@{\,}llp{1.8cm}}
+  \begin{tabular}{@{\!\!\!\!\!\!}l@{\,}llp{1.8cm}}
     \toprule
-    & Statement & classical? & realizable? \\
+    & statement & classical? & realizable? \\
     \midrule
@@
   \only<4>{\expl{3}{There is a machine which, given a machine
-  computing a map~$f : \NN \to \NN$, determines whether~$f$ is constantly
+  computing a map~$f : \NN \to \NN$, determines whether~$f$ has a
   zero or not. \\\ }}
@@
 
-    \visible<2->{``Every function $\NN \to \NN$ is computable'' is realized by
+    \visible<2->{``Every map $\NN \to \NN$ is computable'' is realized by
     \textsf{cat}.}
@@
 
-    \visible<3->{``Every function $\NN \to \NN$ is computable'' is \emph{not}
+    \visible<3->{``Every map $\NN \to \NN$ is computable'' is \emph{not}
     realized.}
@@
 
-    \visible<4->{``Every function $\NN \to \NN$ has a zero or not'' is realized by
-    infinite search.}
+    \visible<4->{\mbox{``Every map $\NN \to \NN$ has a zero or not'' is realized by
+    infinite search.}\\[0em]}
     \medskip
@@
     \justifying
-    \visible<6->{``Every function $\RR \to \RR$ is continuous'' is realized if,
+    \visible<6->{``Every map $\RR \to \RR$ is continuous'' is realized if,
     in the physical world, only finitely many computational
@@
 \section{Proof transformations}
+
+\addtocounter{framenumber}{-1}
+\begin{frame}[plain]\frametitle{A classical logic fairy tale}
+  \vspace*{-0.5em}
+  \begin{changemargin}{-1em}{-1em}
+    \setlength{\columnsep}{15pt}
+    \begin{multicols}{2}
+      \justifying
+      \scriptsize
+
+      \textbf{Narrator.} Once upon a time, in a kingdom far, far away, the
+      queen of the land and of all Möbius strips called for her royal
+      philosopher.
+
+      \textbf{Queen.} Philosopher! I ask you to carry out the following order.
+      Get me the Philosopher's Stone, or alternatively find out how one could
+      produce arbitrary amounts of gold with it!
+
+      \textbf{Philosopher.} But my queen! I haven't studied anything useful!
+      How could I fulfill this order?
+
+      \textbf{Queen.} That is not my concern. I'll see you again tomorrow.
+      Should you not accomplish the task, I will take your head off.
+
+      \textbf{Narrator.} After a long and wakeful night the philosopher was
+      called to the queen again.
+
+      \textbf{Queen.} Tell me! What do you have to report?
+
+      \textbf{Philosopher.} It was not easy and I needed to follow lots of
+      obscure references, but finally I actually found out how to use the
+      Philosopher's Stone to produce arbitrary amounts of gold. But only I can
+      conduct this procedure, your royal highness.
+
+      \textbf{Queen.}
+      Alright. So be it.
+
+      \textbf{Narrator.} And so years passed by, during which the philosopher
+      imagined herself to be safe. The queen searched for the stone on her own,
+      but as long as she hadn't found it, the philosopher didn't need to worry.
+      Yet one day the impossible happened: The queen has found the stone! And
+      promptly called for her philosopher.
+
+      \textbf{Queen.} Philosopher, look! I have found the Philosopher's Stone!
+      Now live up to your promise! \emph{[She hands over the stone.]}
+
+      \textbf{Philosopher.} Thank you. \emph{[She inspects the stone.]} This is
+      indeed the Philosopher's Stone. Many years ago you asked me to either
+      acquire the Philosopher's Stone or find out how to produce arbitrary
+      amounts of gold using it. Now it's my pleasure to present to you the
+      Philosopher's Stone. \emph{[She returns the stone.]}
+    \end{multicols}
+  \end{changemargin}
+\end{frame}
 
@@
     ($R{\top}{\bot}{\wedge}{\vee}{\exists}{\bigvee}$).
-    \item Classical logic proves~$\varphi$ from~$\Gamma$
-    iff minimal logic proves~$\varphi^{\neg\neg}$ from~$\Gamma^{\neg\neg}$.
+    \item If~$\Gamma$ entails~$\varphi$ classically,
+    then~$\Gamma^{\neg\neg}$ entails~$\varphi^{\neg\neg}$ minimally.
   \end{enumerate}

### 2022-09-29 slides-fischbachau2022.tex
Introducing the eventually quantifier

@@
 \usepackage{bussproofs}
+\usepackage{proof}
 \usepackage{xspace}
@@
   \normalsize
-\end{frame}
-
-\addtocounter{framenumber}{-1}
+  \pause
+
+  \emph{* higher-order proofs of first-order statements using the assumption \\
+  that a given (perhaps uncountable) set is countable}
+\end{frame}
+
+\begin{frame}{A quantifier for finite approximations}
+  By a \hil{finite approximation} to an infinite sequence~$\alpha : \NN \to X$,
+  we mean a \hil{finite prefix} of~$\alpha$. Notation:
+
+  \begin{itemize}
+    \item empty list $[]$
+    \item extension $[x_1,\ldots,x_n] ::^r x_{n+1} = [x_1,\ldots,x_n,x_{n+1}]$
+    \item we write~``$\tau \preceq \sigma$'' iff~$\tau$ is
+  a refinement of~$\sigma$: $[x_1,\ldots,x_n,x_{n+1},x_{n+m}] \preceq
+  [x_1,\ldots,x_n]$
+  \end{itemize}
+  \justifying
+
+  For monotone predicates~$P$ of finite lists, we introduce
+  a new quantifier~$\nabla$ such that~``$\nabla (\tau \preceq \sigma)\_
+  P(\tau)$'' expresses that \hil{no matter how~$\sigma$ evolves to a
+  better approximation~$\tau$, eventually~$P(\tau)$ will hold}.
+  \[
+    \infer[{\text{\scriptsize$(\sigma \in X^*)$}}]{\nabla (\tau \preceq \sigma)\_ P(\tau)}{P(\sigma)}
+    \qquad
+    \infer[{\text{\scriptsize$(\sigma \in X^*)$}}]{\nabla (\tau \preceq \sigma)\_ P(\tau)}{\forall(x \in X)\_\ \nabla(\tau
+    \preceq (\sigma ::^r x))\_ P(\tau)}
+  \]
+  \[
+    \infer[{\text{\scriptsize$(\sigma \in X^*, a \in X)$}}]{\nabla(\tau \preceq \sigma)\_ P(\tau)}{\forall(\tau \preceq \sigma)\_\
+    a \in \tau \Rightarrow \nabla(\upsilon \preceq \tau)\_ P(\upsilon)}
+  \]
+\end{frame}
 

### 2022-09-29 slides-fischbachau2022.tex
Slides finished

--- 2022-09-27 slides-fischbachau2022.tex
@@
   \normalsize
-\end{frame}
-
-\addtocounter{framenumber}{-1}
+
+  \visible<3>{\emph{* higher-order proofs of first-order statements using the assumption \\
+  that a given (perhaps uncountable) set is countable}}
+
+  \pause
+  \vspace*{1em}
+  \mbox{\begin{minipage}{0.15\textwidth}
+    \centering\small
+    \includegraphics[height=5em]{zeta-function} \\
+    $\mathbb{C}$
+  \end{minipage}\quad
+  \begin{minipage}{0.20\textwidth}
+    \centering\small
+    \includegraphics[height=5em]{3-adic-numbers} \\
+    $\mathbb{Q}_p$
+  \end{minipage}\quad
+  \begin{minipage}{0.20\textwidth}
+    \centering\small
+    \includegraphics[height=5em]{bruyn-pope} \\
+    $\mathbb{F}_1$
+  \end{minipage}\quad
+  \begin{minipage}{0.42\textwidth}
+    \centering\small
+    \includegraphics[height=5em]{hilbert-hotel} \\
+    $\infty$
+  \end{minipage}}
+\end{frame}
+
+\begin{frame}{A quantifier for finite approximations}
+  Let~$X$ be a (perhaps uncountable) set. By a \hil{finite approximation} to a
+  surjection~$\NN \twoheadrightarrow X$, we mean a \hil{finite list} of
+  elements of~$X$. Notation:
+
+  \begin{itemize}
+    \item empty list: $[]$
+    \item extension: $[x_1,\ldots,x_n] ::^r x_{n+1} = [x_1,\ldots,x_n,x_{n+1}]$
+    \item refinement relation: $[x_1,\ldots,x_n,x_{n+1},x_{n+m}] \preceq [x_1,\ldots,x_n]$
+    \item element access: $\sigma[i] = \text{element at position~$i$
+    in~$\sigma$}$
+  \end{itemize}
+  \justifying
+
+  For monotone predicates~$P$ of finite lists, we introduce
+  a \hil{quantifier}~$\nabla$ such that~``$\nabla^{\tau \preceq \sigma}\_
+  P(\tau)$'' expresses that \hil{no matter how~$\sigma$ evolves to a
+  better approximation~$\tau$, eventually~$P(\tau)$ will hold}.
+  \[
+    \infer[{\text{\scriptsize$(\sigma \in X^*)$}}]{\nabla^{\tau \preceq \sigma}\_ P(\tau)}{P(\sigma)}
+    \qquad
+    \infer[{\text{\scriptsize$(\sigma \in X^*)$}}]{\nabla^{\tau \preceq
+    \sigma}\_ P(\tau)}{\forall^{x \in X}\_\ \nabla^{\tau
+    \preceq (\sigma ::^r x)}\_ P(\tau)}
+  \]
+  \[
+    \infer[{\text{\scriptsize$(\sigma \in X^*, a \in X)$}}]{\nabla^{\tau
+    \preceq \sigma}\_ P(\tau)}{\forall^{\tau \preceq \sigma}\_\
+    a \in \tau \Rightarrow \nabla^{\upsilon \preceq \tau}\_ P(\upsilon)}
+  \]
+\end{frame}
+
+\newcommand{\hnabla}{\hil{$\nabla$}}
+\newcommand{\genalpha}{\mbox{$\hspace{0.12em}\shortmid\hspace{-0.62em}\alpha$}}
+
+\begin{frame}{The generic surjection}
+  \vspace*{-0.6em}
+  \justifying
+  \textbf{Def.} The \hil{$\nabla$-translation} $\varphi \mapsto
+  \varphi^{\nabla}$ into formulas with a free variable $\sigma : X^*$ (denoting
+  the \hil{current stage}) is defined by the following clauses.
+  {\small
+  \begin{align*}
+    (\varphi_\text{atomic})^{\nabla} &\defeqv \hnabla^{\tau \preceq \sigma}\_
+    \varphi_\text{atomic} \\
+    \bot^{\nabla} &\defeqv \hnabla^{\tau\preceq\sigma}\_ \bot \\
+    (\varphi \vee \psi)^{\nabla} &\defeqv \hnabla^{\tau\preceq\sigma}\_
+      (\varphi^{\nabla}[\tau/\sigma] \vee \psi^{\nabla}[\tau/\sigma]) \\
+    (\exists^{x\?X}\_ \varphi)^{\nabla} &\defeqv
+    \hnabla^{\tau\preceq\sigma}\_(\exists^{x\?X}\_ \varphi^{\nabla}[\tau/\sigma]) \\
+    (\varphi \Rightarrow \psi)^{\nabla} &\defeqv
+      \forall^{\tau\preceq\sigma}\_ (\varphi^{\nabla}[\tau/\sigma] \Rightarrow
+      \psi^{\nabla}[\tau/\sigma]) \\
+    \top^{\nabla} &\defeqv \top \\
+    (\varphi \wedge \psi)^{\nabla} &\defeqv (\varphi^{\nabla} \wedge \psi^{\nabla}) \\
+    (\forall^{x\?X}\_ \varphi)^{\nabla} &\defeqv
+      (\forall^{x\?X}\_
+      \varphi^{\nabla}[\sigma/\tau]) \\
+    (\genalpha(n){=}x)^\nabla &\defeqv
+      (\hnabla^{\tau\preceq\sigma}\_ (\textsf{len}(\tau) > n \wedge \tau[n] = x))
+  \end{align*}}
+
+  \textbf{Ex.} $(\forall^{x\?X}\_ \exists^{n\?\NN}\_ \genalpha(n){=}x)^\nabla
+  \equiv$ \\
+  $\qquad\qquad
+    (\forall^{x\?X}\_ \nabla^{\tau\preceq\sigma}\_
+    \exists^{n\?\NN}\_
+    \nabla^{\upsilon\preceq\tau}\_ (\textsf{len}(\upsilon) > n \wedge \upsilon[n] = x))$
+
+  \textbf{Thm.} [Joyal--Tierney 1984] For \hil{first-order} formulas~$\varphi$,
+  $\varphi^\nabla \Rightarrow \varphi$.
+\end{frame}
 

### 2022-09-30 slides-fischbachau2022.tex
typo

@@
 
-  \textbf{Thm.} Let~$X$ be an inhabited set of natural numbers.
+  \textbf{Lemma.} Let~$X$ be an inhabited set of natural numbers.
   \begin{enumerate}
@@
   \justifying
+  \pause
 
@@
 
+  \vspace*{-0.3em}
   \textbf{Ex.} $(\forall^{x\?X}\_ \exists^{n\?\NN}\_ \genalpha(n){=}x)^\nabla
@@
     \exists^{n\?\NN}\_
-    \nabla^{\upsilon\preceq\tau}\_ (\textsf{len}(\upsilon) > n \wedge \upsilon[n] = x))$
-
-  \textbf{Thm.} [Joyal--Tierney 1984] For \hil{first-order} formulas~$\varphi$,
-  $\varphi^\nabla \Rightarrow \varphi$.
+    \nabla^{\upsilon\preceq\tau}\_ (\textsf{len}(\upsilon) > n \wedge \upsilon[n] = x)).$
+
+  \vspace*{-0.3em}
+  \textbf{Thm.} [Joyal--Tierney 1984] For \hil{first-order} formulas~$\varphi$
+  not referring to~$\genalpha$, $\varphi^\nabla \Rightarrow \varphi$
+  intuitionistically.
 \end{frame}

### 2022-10-02 slides-fischbachau2022.tex
Notes almost finished

@@
 
-\title{Revisiting divisible, injective and flabby abelian groups from a
-constructive point of view}
+\title{Extraction of programs from proofs}
 \author{Ingo Blechschmidt}
 \date{September 21th, 2022}
+
+\setbeameroption{show notes on second screen}
+%\setbeamertemplate{note page}[plain, bg=white!90!black]
+\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\footnotesize\justifying#2\par}}}
 
@@
 \begin{frame}{}
+  \jnote{1-5}{
+    From the displayed proof of Euclid's theorem, we can read off an algorithm
+    for computing arbitrarily large primes. There is a deeper reason to that:
+    The proof is \emph{constructive}, and from \emph{every} constructive proof
+    we can extract a corresponding program. One way to formally state and prove
+    this meta-statement is by \emph{realizability theory}, the subject of the
+    first lecture.
+  }
+
+  \jnote{4-5}{
+    In addition to the displayed applications of realizability theory,
+    personally I'm intrigued by it for mostly the following reasons:
+    (1)~Realizability elucidates the interplay between constructive and
+    computable mathematics. (2)~Realizability is a useful guide for pursuing
+    the question whether two given proofs are ``secretly the same'': Do they
+    have the same computational content? (3)~Realizability theory provides us
+    with a host of tantalizing anti-classical models of constructive
+    mathematics.
+  }
+
+  \jnote{5-5}{
+    In the second lecture, we will turn to extracting programs from
+    \emph{classical} proofs. We will do so by transforming classical proofs
+    into constructive ones and then applying the tools of the first lecture.
+    Amazingly, the displayed classical proof and others like it do have
+    constructive content---even though it is constructively and computably
+    impossible to determine minimal values of infinite sequences.
+  }
+
+  \jnote{6-}{
+    Finally, in the third lecture we will learn how to extract constructive
+    content from certain kinds of \emph{invalid} proofs---those which use the
+    preposterous assumption that every set is countable.
+
+    The methods presented in the second and in the third lecture are deeply
+    related to the \emph{dynamical approach to algebra} reported on in Stefan
+    Neuwirth's course. Coherent (and geometric) logic as in Marc Bezem's course
+    also plays an important role in this toolbox. It is greatly informed by a
+    categorical analysis as provided in Steve Awodey's course, particularly so
+    for the third lecture. The first and the second lecture overlap with
+    Chuangjie Xu's course, particularly in employing variants of the
+    double-negation translation.
+
+    This course is set in an informal constructive metatheory. Formalization
+    would both be possible in type theory as in Fredrik Nordvall-Forsberg's
+    course or in constructive set theory as in Hajime Ishihare's course.
+    All invocations of the transfinite such as by the law of excluded middle or
+    by the axiom of choice will be explicitly pointed out. For primers to
+    constructive mathematics, enjoy
+    \fixedhref{https://video.ias.edu/members/1213/0318-AndrejBauer}{Andrej
+    Bauer's 2013 IAS talk}, its \fixedhref{xxx}{written version} or
+    \fixedhref{xxx}{these course notes}.
+  }
+
   \medskip
@@
 
-  {\centering\emph{``Every theorem has a computable* witness.''} \\ \scriptsize
+  {\centering\emph{``\textnormal{Every} theorem has a computable* witness.''} \\ \scriptsize
   * with monadic side effects\par}
@@
   \end{minipage}
+
+  \jnote{1}{
+    This year we will celebrate the 207th birthday of Ada Lovelace, pioneer in
+    computing.
+
+    It is astonishing what she started and which long way we have come!
+
+    Perhaps you would enjoy the graphic novel \emph{The Thrilling Adventures of
+    Lovelace and Babbage} in her honor.
+  }
 \end{frame}}
@@
   \emph{for extracting programs from constructive proofs}
+
+  \jnote{1}{
+    For a written primer to realizability theory, see
+    \fixedhref{http://math.andrej.com/asset/data/c2c.pdf}{Andrej Bauer's course
+    notes} and the notes by
+    \fixedhref{https://www2.mathematik.tu-darmstadt.de/~streicher/REAL/REAL.pdf}{Thomas
+    Streicher}.
+  }
 \end{frame}
@@
   and the rules of \hil{sequence calculus}.
+
+  \jnote{1}{
+    Heyting arithmetic is a convenient base theory for a constructive analysis
+    of arithmetic. It has exactly the same axioms as Peano arithmetic, only
+    that HA is set in intuitionistic logic while PA adds the law of excluded
+    middle.
+
+    As is common, we define negation~``$\neg\varphi$'' as a shorthand for the
+    implication~``$\varphi \Rightarrow \bot$''.
+
+    HA is often expanded to HA\textsuperscript{$\omega$}, \emph{higher-order
+    Heyting arithmetic}, which includes sorts, term constructors and
+    appropriate rules for function types such as~$N^N$ and $N^{(N^N)}$.
+
+    In its original form, realizability theory is only concernced with
+    extracting computational witnesses from HA-proofs; however it is fruitfully
+    extended to all of~HA\textsuperscript{$\omega$}, and we will also glimpse
+    into this higher-order extension. (NB: HA\textsuperscript{$\omega$} is
+    conservative over~HA, and one way to show this is by realizability.)
+  }
 \end{frame}
@@
 \begin{frame}{Number realizability}
+  \jnote{1}{
+    A statement~$\varphi$ is \emph{realizable}, written~``$\realizes\varphi$'',
+    iff it has a \emph{realizer}, a number~$e \in \NN$ such that~$e \realizes
+    \varphi$. The recursive rules governing which numbers~$e$ are deemed to be
+    a realizer of~$\varphi$ make use of \emph{Kleene's original partial combinatory
+    algebra}, the natural numbers equipped with the following partial binary
+    operation~$({\cdot})$: $e \cdot n$ is the result of applying the
+    input~$n$ to the~$e$-th Turing machine (in some effective enumeration of all
+    Turing machines). We write~``$e \cdot n \downarrow$'' to signify that this
+    computation terminates.
+
+    Instead of Turing machines, we can also study other deterministic models of
+    computation by using different partial combinatory algebras, or even
+    nondeterministic and stateful models by a recent
+    \fixedhref{https://rosstate.org/publications/effectful/effectful-mfps19.pdf}{tantalizing
+    generalization} due to Liron Cohen and her coauthors Sofia Abreu Faro and
+    Ross Tate.
+
+    Realizability theory provides one way of formalizing the informal
+    Brouwer--Heyting--Kolmogorov interpretation of constructive mathematics.
+    For instance, that interpretation states that a witness of an
+    implication~$\varphi \Rightarrow \psi$ is a ``method'' for transforming
+    witnesses for~$\varphi$ into witnesses for~$\psi$. Realizability spells out
+    what ``method'' should mean: Turing machine.
+  }
+
+  \jnote{2}{
+    The clauses for disjunction and existential quantification require pairing
+    and unpairing. Given two numbers~$a$ and~$b$, there is a Turing
+    machine~$p_{a,b}$ which outputs~$a$ or~$b$ depending on whether its input
+    is zero or not zero. By~$\pi_1$ and~$\pi_2$, we mean (indices of) Turing
+    machines which, when called on input (an index of)~$p_{a,b}$, extract~$a$
+    repectively~$b$ by simulating its input on the input~$0$ or~$1$.
+
+    The soundness theorem is proven by an instructive induction on the
+    structure of~HA-proofs, verifying that if HA proves a sequent~$\varphi
+    \seq{\vec x} \psi$, then there is a realizer for~$\forall x_1\_ \ldots
+    \forall x_n\_ (\varphi \Rightarrow \psi)$. The core idea of the proof is to
+    verify that every axiom and every rule of~HA is realized. In this way,
+    computational content of every axiom and every rule is explicated.
+  }
+
+  \jnote{3}{
+    In the form ``$(\text{HA} \vdash \varphi) \Rightarrow
+    (\realizes\varphi)$'', the soundness theorem can be stated and proved in
+    most contexts in which the natural numbers exist as a complete entity, such
+    as constructive or classical set or type theories.
+
+    But in a sense, this is misleading: The mapping from proofs to realizers is
+    a computationally simple syntactical transformation. As such, already PRA
+    can prove the soundness theorem if we formulate it as~``$(\text{HA} \vdash
+    \varphi) \Rightarrow (\exists e\_ \text{HA} \vdash (e\realizes\varphi))$''.
+
+    NB: Some formulations of realizability state the clauses for disjunction
+    and existential quantification in a slighter simpler way, directly using
+    pairing and unpairing functions on the naturals. The price for this
+    simplification is that then the soundness theorem has to be formulated
+    as``$(\text{HA} \vdash \varphi) \Rightarrow (\exists e\_ \text{HA} \vdash
+    (e \realizes(\top \Rightarrow \varphi)))$''.
+  }
+
   \vspace*{-1em}
@@
   \mbox{\textbf{Thm.} If~$\text{HA} \vdash \varphi$, then there is a
-  number~$e \in \NN$ such that~\only<1>{$e \realizes
-  \varphi$}\only<2>{$\text{HA} \vdash (\underline{e} \realizes \varphi)$}.}
+  number~$e \in \NN$ such that~\only<1-2>{$e \realizes
+  \varphi$}\only<3>{$\text{HA} \vdash (\underline{e} \realizes \varphi)$}.}
 \end{frame}
@@
 \begin{frame}{Exploring the realizability model}
-  \medskip
+  \jnote{5}{
+    The statement that every function~$\NN \to \NN$ is computable by a Turing
+    machine (or equivalently, by a lambda term) is known as the \emph{formal
+    Church--Turing thesis}. It is an example of a statement which is realizable
+    but not provable in~HA\textsuperscript{$\omega$}.
+
+    Many of the curious properties of the realizability model follow from the
+    formal Church--Turing thesis. In fact, a slight generalization called the
+    \emph{extended Church thesis} suffices to completely characterize the
+    (provably) realizable statements:
+    \[ \text{HA+ECT} \vdash \varphi \quad\text{iff}\quad
+      \text{HA} \vdash (\realizes \varphi). \]
+
+    In the realizability model built using lambda terms instead of Turing
+    machines, the formal Church--Turing thesis fails. This is because of
+    changed calling conventions: In the model built using lambda terms, a
+    realizer for a statement of the form~``$\forall f\?N^N\_ \varphi(f)$'' is a
+    lambda term~$e$ such that for every lambda term~$r$ computing a
+    function~$f : \NN \to \NN$, the term $er$ is a realizer for~$\varphi(f)$.
+    However, the term~$e$ cannot inspect the form (``source code'') of its
+    argument.
+  }
+
+  \jnote{6}{
+    Statement~5 is not a statement in the language of~HA or
+    of~HA\textsuperscript{$\omega$}, but in an extension in which we can also
+    support quotients (to make sense of the construction of the reals using
+    equivalence classes of Cauchy sequences) or powersets (to support the
+    construction using Dedekind cuts). A proper interpretation is possible in
+    the category of assemblies or in the ef{}fective topos.
+
+    An exposition of this continuity phenomenon is provided in
+    \fixedhref{https://arxiv.org/pdf/2204.00948.pdf}{this survey paper}
+    (Example~6 there).
+  }
+
+  \jnote{7}{
+    By \emph{Markov's principle}, we mean the statement
+    \[ \forall f\?N^N\_ (\neg\neg(\exists n\?N\_ f(n) = 0)) \Rightarrow
+    (\exists n\?N\_ f(n) = 0). \]
+    By the clauses for implication and negation, a number~$e$ is a realizer for
+    a negated statement~$\neg\psi$ iff there is no realizer for~$\psi$:
+    \begin{align*}
+      e \realizes \neg\psi &\quad\text{iff}\quad
+      \text{for every~$r \in \NN$ such that~$r \realizes \psi$, $e \cdot r
+      \downarrow$ and~$e \cdot r \realizes \bot$} \\
+      &\quad\text{iff}\quad \text{for every~$r \in \NN$ such that~$r \realizes \psi$, falsum holds} \\
+      &\quad\text{iff}\quad \text{there is no number~$r \in \NN$ such that~$r \realizes \psi$} \\
+      &\quad\text{iff}\quad \text{$\psi$ is not realized}
+    \end{align*}
+    In particular, if there exists a realizer for a negated statement at all,
+    every number whatsoever is a realizer. As a consequence, \emph{realizers for
+    negated statements are never informative;} and a number~$e$ is a realizer
+    for~$\neg\neg\varphi$ iff~$\varphi$ is \emph{not~not} realizable. Hence a
+    realizer for~$\neg\neg\varphi$ encodes the mere promise that somewhere,
+    there is a realizer for~$\varphi$, without giving any indication how to
+    find it.
+  }
+
+  \jnote{10}{
+    As the examples illustrate, realizability and truth do not at all coincide:
+    There are many statements which are realizable but not true from the point of
+    view of classical mathematics (such as the formal Church--Turing thesis)
+    and vice versa (such as the statement that every function~$\NN \to \NN$ has
+    a zero or not).
+
+    Within the realizability model, the situation is radically different.
+    For every statement~$\varphi$, the statement~``$\varphi \Leftrightarrow
+    (\realizes \varphi)$'' is realizable.
+
+    The multiverse of models of constructive mathematics can be explored from
+    the point of view of any base model, and from the point of view of the
+    realizability model it looks quite different that from the point of view of
+    classical mathematics.
+  }
+
   \medskip
@@
     \normalnumber{5} & Every map $\RR \to \RR$ is continuous. & \cxmark &
-    \qswitch{5}{6}{\ccmark} \\
+    \qswitch{5}{6}{\ccmark{} (if MP)} \\
     \normalnumber{6} & Markov's principle holds. & \ccmark{} (trivially) &
@@
     \qswitch{8}{9}{\ccmark{} (if MP)} \\
+    \normalnumber{9} & A statement holds iff it is realized. & \cxmark &
+    \qswitch{9}{10}{\ccmark} \\
     \bottomrule
@@
 
-  \only<1>{\ \\\ \\\ \\\ }
+  \only<1>{\color{white}There is a machine which determines of any given
+  number whether it is prime or not. \\\ \\\ }
   \only<2>{\expl{1}{There is a machine which determines of any given
-  number whether it is prime or not. \\\ }}
+  number whether it is prime or not. \\\ \\\ }}
   \only<3>{\expl{2}{There is a machine which, given a number~$n$, computes a
@@
   computing a map~$f : \NN \to \NN$, determines whether~$f$ has a
-  zero or not. \\\ }}
+  zero or not. \\\ \\\ }}
   \only<5>{\expl{4}{There is a machine which, given a machine
   computing a map~$f : \NN \to \NN$, outputs a machine
-  computing~$f$. \\\ }}
+  computing~$f$. \\\ \\\ }}
   \only<6>{\ \\\ \\\ \\\ }
@@
   the case that~$f$ has a zero, determines a zero of~$f$.}}
-  \only<8->{\expl{7}{\justifying There is a machine which, given a machine
+  \only<8>{\expl{7}{\justifying There is a machine which, given a machine
   computing for every~$x \in \NN$ some~$y \in A$ together with a realizer
@@
   function~$\NN \to A$.}}
+  \only<9>{\ \\\ \\\ \\\ }
 \end{frame}}
@@
 \begin{frame}{Metatheory of Heyting arithmetic}
+  \jnote{1}{
+    Variants of the realizability model can be used to establish several
+    metatheoretic properties of Heyting arithmetic. For the second and third
+    properties, the keyword is ``realizability with proof''; for the fourth,
+    using the variant of realizability built using System~T terms instead of
+    Turing machines.
+  }
+
   \begin{enumerate}
@@
 \begin{frame}{Range of machine models}
+  \jnote{1-}{
+    Infinite time Turing machines were introduced
+    \fixedhref{https://arxiv.org/abs/math/9808093}{by Joel David Hamkins and
+    Andy Lewis}. Unlike ordinary Turing machines, they can carry out ``more
+    than infinitely many computational steps''. Where an ordinary Turing
+    machine fails to terminate, an infinite time Turing machine is put on
+    day~$\omega$ into a special limit state and can then meaningfully continue.
+
+    All functions in Gödel's System~T are unconditionally total. Hence
+    unbounded search cannot be implemented in System~T.
+
+    The idea to apply realizability to machines in the real world is
+    \fixedhref{http://math.andrej.com/2014/03/04/intuitionistic-mathematics-and-realizability-in-the-physical-world/}{due to Andrej Bauer}.
+  }
+
   \begin{enumerate}
@@
 
-    \item \hil{Machines in the real world} -- philosophical
+    \item \hil{Machines in the real world} -- \emph{philosophical}
 
@@
   \end{changemargin}
+
+  \jnote{1}{
+    Adapted from
+    \fixedhref{http://blog.ezyang.com/2013/04/a-classical-logic-fairy-tale/}{Edward
+    Yang's blog}.
+  }
 \end{frame}
@@
 \begin{frame}{A case study in double negation}
+  \jnote{2-4}{
+    The two displayed proofs seem to be devoid of computational content:
+    Computationally, it is neither possible to determine the minimal element of
+    an infinite sequence nor to determine whether a binary sequence contains
+    infinitely many zeros or infinitely many ones. These claims have no
+    realizer.
+
+    However, contrary to expectations, the two proofs do contain an obscured
+    constructive core, and this core can be uncovered by the two logical
+    metatheorems presented in this lecture, the double-negation embedding and
+    \emph{Friedman's trick}. Details are in Thierry Coquand's
+    \fixedhref{https://www.lama.univ-savoie.fr/pagesmembres/ruyer/divers/lecturearticles/coquand/cor3.ps}{Computational
+    Content of Classical Logic}.
+  }
+
+  \jnote{3-4}{
+    To set the stage, we analyze the lemma used in the first proof. The
+    statement ``every inhabited set of natural numbers contains a minimal
+    element'' is not realizable and hence does not directly have
+    computational content. In fact, this statement is equivalent to the law of
+    excluded middle. Correspondingly, it seems that proofs using it cannot ever
+    be constructivized. However, as we will see, much more important for
+    extraction of constructive content is the form of the asserted claims than
+    the form of the used auxiliary lemmas.
+  }
+
+  \jnote{5-}{
+    A set~$X$ of natural numbers is \emph{detachable} if and only if for every
+    number~$n \in \NN$, either~$n \in X$ or~$n \not\in X$. In classical
+    mathematics, every set of natural numbers is detachable. Constructively,
+    detachability is a nontrivial and computationally meaningful condition. A
+    realizer for detachability of a set~$X$ is a machine which reads a
+    number~$n$ as input and determines whether~$n$ belongs to~$X$ or not
+    (together with corresponding realizers).
+  }
+
+  \jnote{7-}{
+    Instead of strengthening the assumption, we can also weaken the conclusion.
+    Statement~3 merely asserts that it is impossible that no minimum exists.
+    The \emph{double-negation embedding} reviewed on the next slide explains
+    why it can be expected a priori that this weakening allows us to give a
+    constructive proof.
+  }
+
   \vspace*{-0.8em}
@@
 \begin{frame}{The double-negation embedding}
+  \jnote{1-3}{
+    The starting point of the double-negation embedding is that, while the law
+    of excluded middle is not available in intuitionistic logic, the double
+    negation of every instance is: For every formula~$\varphi$, the formula
+    \[ \neg\neg(\varphi \vee \neg\varphi) \]
+    is an intuitionistic tautology:
+
+    \begin{quote}
+      In order to show~$\neg\neg(\varphi \vee \neg\varphi)$,
+      assume~$\neg(\varphi \vee \neg\varphi)$ and deduce~$\bot$.
+
+      As a preparatory step, we verify~$\neg\varphi$: If~$\varphi$, then in
+      particular~$\varphi \vee \neg\varphi$, hence~$\bot$ by assumption.
+
+      Having established~$\neg\varphi$, we notice that also~$\varphi \vee
+      \neg\varphi$, hence~$\bot$ by assumption.
+    \end{quote}
+
+%    This observation unlocks the following technique to transcribe a classical
+%    proof of a statement to an intuitionistic (and even minimal logic) proof of
+%    its translation. Most steps in the classical proof will also be acceptable
+%    in intuitionistic logic. We simply mimic these. When the classical proof
+%    appeals to~\textsc{lem}, we instead use the appropriate doubly negated
+%    instance. We
+%
+%    The difference between intuitionistic and classical logic is concentrated
+%    in their treatment of disjunction and existential quantification.
+%    Classically, to verify a disjunction~$\varphi \vee \psi$, it suffices to
+%    establish~$\neg(\neg\varphi \wedge \neg\psi)$--the intuitionistically
+%    weaker claim that it is not the case that neither~$\varphi$ nor~$\psi$
+%    hold; similarly, to verify an existential statement~$\exists x\?X\_
+%    \varphi(x)$, it suffices to establish~$\neg\forall x\?X\_
+%    \neg\varphi(x)$---the claim that nonexistence is impossible.
+  }
+
+  \jnote{3}{
+    The three parts of the theorem are each proven by induction, on the
+    structure of~$\varphi$ for the first two parts and on the structure of
+    derivations for the third part. Carrying this out is an instructive
+    exercise!
+  }
+
+  \jnote{4}{
+    The corollary follows from the theorem because HA proves the
+    double-negation translation of every axiom of PA. For instance, the
+    translation of the axiom
+    \[ \forall x\?N\_ x + 0 = x \]
+    is
+    \[ \forall x\?N\_ \neg\neg(x + 0 = x), \]
+    and this weaker statement is provable thanks to the intuitionistic (even
+    minimal) tautology~$\varphi \Rightarrow \neg\neg\varphi$.
+
+    Similarly, the translation of an instance of the induction scheme
+    \[
+      \varphi(0) \wedge \bigl(\forall x\?N\_ \varphi(x) \Rightarrow
+      \varphi(S(x))\bigr) \quad\Longrightarrow\quad \forall x\?N\_ \varphi(x)
+    \]
+    is again an instance of the induction scheme:
+    \[
+      \varphi(0)^{\neg\neg} \wedge \bigl(\forall x\?N\_ \varphi(x)^{\neg\neg} \Rightarrow
+      \varphi(S(x))^{\neg\neg}\bigr) \quad\Longrightarrow\quad \forall x\?N\_
+      \varphi(x)^{\neg\neg}
+    \]
+  }
+
+  % xxx importance of remark
+
   \justifying

### 2022-10-03 slides-fischbachau2022.tex
More notes

@@
     for the third lecture. The first and the second lecture overlap with
-    Chuangjie Xu's course, particularly in employing variants of the
+    Chuangjie Xu's course, particularly regarding the
     double-negation translation.
@@
     course or in constructive set theory as in Hajime Ishihare's course.
-    All invocations of the transfinite such as by the law of excluded middle or
+    All appeals to the transfinite such as by the law of excluded middle or
     by the axiom of choice will be explicitly pointed out. For primers to
@@
 
-    It is astonishing what she started and which long way we have come!
+    It is astonishing what she started and what long way we have come!
 
@@
   \jnote{1}{
+    Monika Seisenberger has
+    \fixedhref{https://www.proofsociety.org/wp-content/uploads/2018/09/ProgramExtraction_slides.pdf}{many
+    and more detailed} slides on this topic.
+
     For a written primer to realizability theory, see
@@
     computational content of every axiom and every rule is explicated.
+
+    For instance, the statement (with no free variables)
+    \[ \bot \Rightarrow \varphi \]
+    is realized by any number~$e \in \NN$ such that for every~$r \in \NN$
+    with~$r \realizes \varphi$ (this condition is never satisfied), $e \cdot r
+    \downarrow$ and~$e \cdot r \realizes \varphi$, so by any number whatsoever.
   }
@@
 
+  \jnote{8}{
+    By \emph{countable choice}, we mean the statement
+    \[ (\forall x\?N\_ \exists y\?A\_ \varphi(x,y)) \Longrightarrow
+      (\exists f\?A^N\_ \forall x\?N\_ \varphi(x,f(x))). \]
+    Up to some repackaging, this statement is realized by the identity machine
+    which simply outputs its input unchanged.
+
+    Choice for higher type fails in the realizability model. For instance, the
+    statement
+    \[ (\forall f\?N^N\_ \exists y\?A\_ \varphi(f,y)) \Longrightarrow
+      (\exists \theta\?A^{N^N}\_ \forall f\?N^N\_ \varphi(f,\theta(f))) \]
+    is not realized. A realizer for the antecedent would be a machine which,
+    given an index for a Turing machine computing a total function~$f : \NN \to
+    \NN$, produces a code for a suitable element~$y$. However, this element~$y$
+    might not only depend on the extensional input/output behavior of~$f$, but
+    also on the specific index (source code), hence wouldn't describe an actual
+    function on the set of computable functions~$\NN \to \NN$.
+
+    This issue does not arise with countable choice, as natural numbers have
+    unique codes.
+  }
+
+  \jnote{9}{
+    Similar to Statement~5, Statement~8 can only be formulated in an extension
+    of the formal language used here.
+
+    It expresses that, up to unique isomorphism, there is just one model of
+    Heyting arithmetic, namely the standard model. This is in stark contrast
+    with the situation in classical mathematics, where Gödel's completeness
+    theorem/Henkin term models can be used to concoct a host of nonstandard
+    models.
+
+    As a consequence, Peano arithmetic is ``quasi-inconsistent'' from the point
+    of view of the realizability model, as it is consistent (being
+    equiconsistent with HA) but does not admit a model (every model of PA is
+    also a model of HA, but HA only has one model, and this does not validate
+    the PA-theorem ``every Turing machine terminates or does not terminate'').
+
+    Pointers to relevant literature are in
+    \fixedhref{https://arxiv.org/pdf/2204.00948.pdf}{this survey paper}
+    (Example~8 there). Also see
+    \fixedhref{https://www.ps.uni-saarland.de/Publications/documents/HermesKirst_2022_An-Analysis.pdf}{the
+    2022 paper by Marc Hermes and Dominik Kirst}, particularly also the final
+    paragraph of their Section~8.1 which alludes to a result in a different
+    direction.
+  }
+
   \jnote{10}{
@@
     the point of view of any base model, and from the point of view of the
-    realizability model it looks quite different that from the point of view of
+    realizability model it looks quite different than from the point of view of
     classical mathematics.
@@
     be constructivized. However, as we will see, much more important for
-    extraction of constructive content is the form of the asserted claims than
-    the form of the used auxiliary lemmas.
+    extraction of constructive content is the form of the asserted end results than
+    the form of auxiliary lemmas used in a proof.
   }
@@
     number~$n$ as input and determines whether~$n$ belongs to~$X$ or not
-    (together with corresponding realizers).
+    (outputting also corresponding realizers).
   }
@@
   \jnote{1-3}{
-    The starting point of the double-negation embedding is that, while the law
-    of excluded middle is not available in intuitionistic logic, the double
-    negation of every instance is: For every formula~$\varphi$, the formula
-    \[ \neg\neg(\varphi \vee \neg\varphi) \]
+    Minimal logic is intuitionistic logic minus \emph{ex falsum quodlibet}
+    ($\bot \seq{\vec x} \varphi$).
+
+    The double-negation embedding builds on a fundamental
+    observation: While the law
+    of excluded middle is not available in minimal or intuitionistic logic, the double
+    negation of every instance is---for every formula~$\varphi$, the formula
+    $\neg\neg(\varphi \vee \neg\varphi)$
     is an intuitionistic tautology:
@@
 
-%    This observation unlocks the following technique to transcribe a classical
-%    proof of a statement to an intuitionistic (and even minimal logic) proof of
-%    its translation. Most steps in the classical proof will also be acceptable
-%    in intuitionistic logic. We simply mimic these. When the classical proof
-%    appeals to~\textsc{lem}, we instead use the appropriate doubly negated
-%    instance. We
-%
-%    The difference between intuitionistic and classical logic is concentrated
-%    in their treatment of disjunction and existential quantification.
-%    Classically, to verify a disjunction~$\varphi \vee \psi$, it suffices to
-%    establish~$\neg(\neg\varphi \wedge \neg\psi)$--the intuitionistically
-%    weaker claim that it is not the case that neither~$\varphi$ nor~$\psi$
-%    hold; similarly, to verify an existential statement~$\exists x\?X\_
-%    \varphi(x)$, it suffices to establish~$\neg\forall x\?X\_
-%    \neg\varphi(x)$---the claim that nonexistence is impossible.
+    \vspace*{-0.6em}
+    Also, a doubly negated statement is not a dead end. Instead, we can
+    continue reasoning, by the following tautology (``Kleisli extension'',
+    ``monadic bind'', ``nucleus axiom''):
+
+    \vspace*{-1.6em}
+    \[ (\neg\neg\varphi \wedge (\varphi \Rightarrow \neg\neg\psi))
+    \Longrightarrow \neg\neg\psi \]
   }
@@
   \jnote{3}{
+    \vspace*{-0.8em}
     The three parts of the theorem are each proven by induction, on the
@@
 
-  % xxx importance of remark
+  \jnote{5}{
+    Live demo with
+    \fixedhref{https://agdapad.quasicoherent.io/~AgdaPadova/html/Dickson-pc2022.html}{Agda
+    code for the theorem on all infinite sequences of natural numbers being
+    good}:
+
+    \begin{enumerate}
+    \item\justifying The first proof there implements the classical proof,
+    postulating the law of excluded middle as an unprovable axiom. Thereby Agda
+    can verify the classical proof to be correct, however trying to run the
+    proof will get stuck on the postulated \textsc{lem}-oracle.
+
+    \item The second proof rewrites the statement of the theorem and its proof
+    by the double-negation translation. The result is a constructive proof
+    which doesn't rely on postulates; however there is still no direct
+    computational content, as the asserted claim is a (doubly) negated
+    statement.
+
+    \item In a logical sleight of hand, the third proof imports the
+    second proof but specializes~$\bot$, which was an arbitrary constant there,
+    to the asserted claim. The tautology~$\negg\negg\bott \Rightarrow \bott$ of
+    minimal logic is then used to escape the double-negation monad and obtain a
+    constructive proof of the full result.
+    \end{enumerate}
+  }
 
@@
 \begin{frame}{Barr's theorem / Friedman's trick / A-translation}
+  \jnote{4}{
+    The proof transformation for~``$3 \Rightarrow 2$'' is explicit in
+    nature, feasible in practice and increases the proof length only
+    polynomially. There is also an
+    \fixedhref{https://drops.dagstuhl.de/opus/volltexte/2022/16776/pdf/LIPIcs-TYPES-2021-7.pdf}{intriguing
+    alternative approach by Giulio Fellin, Sara Negri and Eugenio Orlandelli}
+    employing cut elimination (thereby increasing the proof length
+    substantially, but conceptually beneficial in other aspects). Furthermore,
+    this metatheorem can also be cast in semantic terms:
+
+    Every Grothendieck topos admits a cover (a surjective geometric morphism from
+    another Grothendieck topos) by a Grothendieck topos which is boolean. This
+    can be obtained by first adjoining a \emph{generic proposition}~$\chi$ and then
+    taking sheaves with respect to the modality~$(({\cdot} \Rightarrow \chi)
+    \Rightarrow \chi)$.
+
+    Applied to the classifying topos of~$\Gamma$, this insight yields a proof
+    of~``$3 \Rightarrow 0$'', since pullback along surjective geometric
+    morphisms reflects validity of geometric sequents.
+
+    Furthermore, every Grothendieck topos admits a cover by a Grothendieck
+    topos over a complete Boolean algebra. Assuming Zorn's lemma in the
+    metatheory, this topos validates~\textsc{lem} and~\textsc{zorn},
+    hence~\textsc{ac}. This shows~``$4 \Rightarrow 0$''.
+
+    An introduction to topos-theoretic generic models with a view towards
+    applications in constructive commutative algebra is contained in
+    \fixedhref{https://arxiv.org/abs/2012.13850}{this survey}.
+    \vspace*{2em}
+  }
+
   \textbf{Thm.} Let~$\Gamma$ be a set of geometric sequents over a fixed
@@
     \item $\sigma$ is provable from~$\Gamma$ in \bad{classical logic}.
-    \item (Assuming \badbox{\textsc{ac}}) $\sigma$ is provable from~$\Gamma$ in
+    \item (Assuming \badbox{\textsc{zorn}}) $\sigma$ is provable from~$\Gamma$ in
     \bad{classical logic with~\textsc{ac}}.
@@
 \begin{frame}
+  \jnote{3}{
+    \textbf{Thm. 1.} From every proof~$p$ of a statement~$\varphi$ such that
+    \vspace*{-0.5em}
+    \begin{enumerate}
+      \item the assertion~$\varphi$ is a first-order statement,
+      \item the proof~$p$ is constructive,
+      \item the proof~$p$ is formulated in a certain higher-order system (in
+      particular, the proof may, unlike its end result, freely employ
+      higher-order notions) and
+      \item for a certain set~$X$ appearing in~$p$, the proof assumes that~$X$
+      is countable,
+    \end{enumerate}
+    \vspace*{-0.5em}
+    a constructive proof of the same statement~$\varphi$, set in the same
+    system, but \emph{without requiring the countability assumption} can be
+    extracted.
+  }
+
+  \jnote{4}{
+    \textbf{Thm. 2.} From every proof~$p$ of a statement~$\varphi$ such that
+    \vspace*{-0.5em}
+    \begin{enumerate}
+      \item the assertion~$\varphi$ is a geometric implication,
+      \item the proof~$p$ is classical (constructive + \textsc{lem}, but still no
+      \textsc{zorn}),
+      \item the proof~$p$ is formulated in a certain higher-order system (in
+      particular, the proof may, unlike its end result, freely employ
+      higher-order notions) and
+      \item for a certain set~$X$ appearing in~$p$, the proof assumes that~$X$
+      is countable,
+    \end{enumerate}
+    \vspace*{-0.5em}
+    a classical proof of the same statement~$\varphi$, set in the same
+    system, but \emph{without requiring the countability assumption} can be
+    extracted.
+  }
+
   \centering\bigskip
@@
 
-  \visible<3>{\emph{* higher-order proofs of first-order statements using the assumption \\
+  \visible<3->{\emph{* higher-order proofs of first-order statements using the assumption \\
   that a given (perhaps uncountable) set is countable}}

### 2022-10-16 paper-modal-multiverse.tex
typo

@@
 \begin{abstract}
-  To help put earlier results in constructive algebra and constructive
+  To help put established results in constructive algebra and constructive
   combinatorics into perspective, give possible background for certain
@@
   topos-theoretic multiverse. Our proposal is inspired by the corresponding study
-  of the set-theoretic multiverse and focusses on applications in other
-  parts of mathematics.
-  % than the intriguing endeavour of exploring the range of set-theoretic possibility.
+  of the set-theoretic multiverse, but focuses less on exploring the range of
+  set/topos-theoretic possibility and more on concrete applications in
+  constructive mathematics.
 \end{abstract}
@@
 \item A relation is almost-full iff every infinite sequence is good (meaning
-that some element is related to a later element in the sequence -- this
-notion has been studied in combinatorics~\cite{xxx} and found applications in
+that some term on the sequence is related to some later term -- the notion
+of almost-full relations
+has been studied in combinatorics~\cite{xxx} and found applications in
 termination checking~\cite{xxx}).
@@
 content---from every constructive proof, a corresponding algorithm can be
-extracted, and every constructive proof holds also for continuous families of
-the objects in question. But the listed classical principles have no
+extracted~\cite{bauer:c2c}, and every constructive proof holds also for continuous families of
+the objects in question~\cite[Section~4.3]{blechschmidt:filmat}. But the listed classical principles have no
 computational witness or fail in continuous families, hence are not
@@
 chain.
-\item A relation is almost-full iff every infinite sequence \emph{everywhere} is good (meaning
-that some element is related to a later element in the sequence).
+\item A relation is almost-full iff every infinite sequence \emph{everywhere} is good.
 \item A ring element is nilpotent iff all prime
@@
 The modal language not only allows us to recover classical principles
-as above, but also renders some powerful theorems about the topos-theoretic
+as above, but also makes some powerful theorems about the topos-theoretic
 landscape smoothly accessible:
@@
 commutative algebra~\cite{blechschmidt-schuster:constructive-maximal-ideals}.
-For countable rings, an explicit construction of an maximal ideal is available.
+For countable rings, an explicit construction of a maximal ideal is available.
 By item~(\ref{item:prox-countable}), this construction can also be carried out
@@
 the point of view of~$\B$''; using a sufficiently expressive form of the
-internal language of~$\B$, this amounts to a bounded geometric morphism~$\E \to
+internal language of~$\B$, which allows to make direct sense of the notion in
+scare quotes, this amounts to a bounded geometric morphism~$\E \to
 \B$, and this can be taken as the definition.
@@
 
-\begin{rem}Perhaps arbitrary elementary toposes over the base, corresponding to
+\begin{rem}Perhaps arbitrary \emph{elementary} toposes over the base, corresponding to
 possibly unbounded geometric morphisms~$\E \to \B$ or even arbitrary
@@
 axioms of elementary toposes, should also be
-taken as part of the topos-theoretic multiverse. This first proposal of the
-modal multiverse sticks to Grothendieck toposes for ease of
+taken as part of the topos-theoretic multiverse. The first foray into the
+modal topos-theoretic multiverse outlined in this note
+sticks to Grothendieck toposes for ease of
 formalizability (``for every small site'' can be expressed also in the more
 standard flavors of the internal topos language, while ``for every elementary
-topos'' requires more elaborate versions) and because the restricted multiverse
-is already sufficiently rich for the intended applications.
+topos'' requires more elaborate versions); because the restricted multiverse
+is already sufficiently rich for the intended applications; and because we have
+a generalization to the predicative setting~\cite{crosilla:xxx} with arithmetic
+universes in mind. Predicatively, not even the category of sets might be an
+elementary topos, but they do form an arithmetic universe and sheaves over
+sites are a meaningful concept.
 \end{rem}
@@
 morphism~$f : \E \to \Set$ is surjective, that is iff~$f^*$ reflects
-isomorphisms..\end{defn}
+isomorphisms.\end{defn}
 
@@
 of~$X$ is inhabited (xxx:verify). As such, positivity is a more informative
-version of the weaker property of being nontrivial (that is, that the top open
+version of the weaker property of being nontrivial (the property that the top open
 and the bottom open do not coincide).\end{ex}
@@
 
-The established substitute is to define a transitive relation~$({<})$ on a set~$X$ to
-be well-founded if and only if for every subset~$M \subseteq X$,
+The established substitute is to declare that a transitive relation~$({<})$ on a set~$X$
+is well-founded if and only if for every subset~$M \subseteq X$,
 \begin{equation}\label{eq:ho-wf}\tag{$\star$}
@@
 single set or class of ``all subsets of~$X$'' but for instance a
-hierarchy of subsets of increasing universe level, a transitive relation~$({<})$ is
+hierarchy of subsets of increasing universe levels, a transitive relation~$({<})$ is
 well-founded if and only if every element of~$X$ is \emph{accessible}, where
@@
   \qquad
-  \infer{P \,|\, \sigma}{\forall x \in X\_\ P \,|\, (\sigma :: x)}
+  \infer{P \,|\, \sigma}{\forall x \? X\_\ P \,|\, (\sigma :: x)}
 \]
@@
 With this inductive definition, expected properties of the class of almost-full
-relations such as being stable under cartesian products (Dickman's lemma),
+relations such as stability under cartesian products (Dickman's lemma),
 finite lists (Higman's lemma) or finitely-branching trees (Kruskal's theorem)
@@
 the inductive sense, not only is every infinite sequence good, but also every
-infinite ``multi-valued sequence''\footnote{xxx} or every infinite
+infinite ``multi-valued sequence''\footnote{xxx} and every infinite
 partially-defined sequence~$\alpha$ for which for every number~$n \in \NN$ it
-is \notnot the case that~$\alpha(n)$ exists. Can we pinpoint how much more
-general the inductive definitions are?
+is \notnot the case that~$\alpha(n)$ exists. Can we pinpoint how much stronger
+the inductive definitions are?
 
@@
 
-Namely, the theories of an infinite sequence or of an infinite descending chain
+Namely, the theories of an infinite sequence and of an infinite descending chain
 are geometric. As such, there exist their classifying toposes, containing the
 \emph{generic infinite sequence} respectively the \emph{generic infinite
-descending chain}. When is this good respectively validates~$\bot$?
+descending chain}, and we may ask: When is this sequence good respectively when
+does this chain validate~$\bot$?
 
@@
     \item By \textsc{lem} and well-foundedness, there is a minimal value~$\alpha(i_0)$
-    among all values of~$\alpha$. \vspace*{-0.2em}
+    among all values of~$\alpha$.
     Similarly, there is a minimal value~$\alpha(i_1)$ among~$(\alpha(n))_{n >
@@
     on. By \emph{proximal dependent choice}, we can proximally collect these indices
-    into a function~$i : \NN \to \NN$; this switches~\textsc{lem} off. \vspace*{-0.2em}
+    into a function~$i : \NN \to \NN$;\footnote{We could also avoid dependent
+    choice by picking not some suitable indices~$i_0, i_1, \ldots$, but the
+    least such.} this switches~\textsc{lem} off.
 

### 2022-11-03 slides-paris2022.tex
Structure of Paris talk

A modal logical multiverse for commutative algebra and combinatorics

In the spirit of the set-theoretic multiverse philosophy put forward by
Joel David Hamkins, we explore a related modal multiverse populated by
Kripke models and more general worlds. In this
multiverse—as the talk will explain—the law of excluded middle can be
switched on and off like a light bulb and countability is a button (for
every set X of every world, there is a larger world containing a
surjection ℕ → X). Our interest in this multiverse is because of
concrete applications in commutative algebra and combinatorics,
including the endeavor of extracting algorithms from proofs utilizing
transfinite techniques. The talk will be framed by several examples of
this kind.

=== Motivation

Infinitude of primes
- algorithm visible from the proof

Two motivating questions:
- Dickson's lemma
- Surjectivity of matrices
How to extract algorithms here?
Spoil elementary proof of surjective case

=== Set-theoretic multiverse

Quick sketch of history including dream solution to CH

Multiverse position: acknowledge that notion of set is underdetermined by ZFC
axioms, embrace multiverse of models

Switches and buttons

Analogy with field extensions

=== Relaxation to topos-theoretic case

"Parametric mathematics" (esp. visible with toposes over spaces)

Modal operators

Examples of modal statements

### 2022-11-04 slides-paris2022.tex
First two slides

@@
+\documentclass[12pt,utf8,notheorems,compress,t]{beamer}
+\usepackage{etex}
+
+\usepackage{pgfpages}
+\usepackage[export]{adjustbox}
+
+% Workaround for the issue described at
+% https://tex.stackexchange.com/questions/164406/beamer-using-href-in-notes.
+\newcommand{\fixedhref}[2]{\makebox[0pt][l]{\hspace*{\paperwidth}\href{#1}{#2}}\href{#1}{#2}}
+
+\usepackage[english]{babel}
+
+\usepackage[normalem]{ulem}
+\usepackage{graphbox}
+\usepackage{mathtools}
+\usepackage{booktabs}
+\usepackage{stmaryrd}
+\usepackage{amssymb}
+\usepackage{array}
+\usepackage{ragged2e}
+\usepackage{multicol}
+\usepackage{tabto}
+\usepackage{xstring}
+\usepackage{proof}
+\usepackage{ifthen}
+\usepackage[normalem]{ulem}
+\usepackage[all]{xy}
+\xyoption{rotate}
+\usepackage{tikz}
+\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing,positioning}
+\hypersetup{colorlinks=true}
+
+\newcommand*\circled[1]{\tikz[baseline=(char.base)]{%
+  \node[shape=circle,draw,inner sep=1pt] (char) {#1};}}
+
+\DeclareFontFamily{U}{bbm}{}
+\DeclareFontShape{U}{bbm}{m}{n}
+   {  <5> <6> <7> <8> <9> <10> <12> gen * bbm
+      <10.95> bbm10%
+      <14.4>  bbm12%
+      <17.28><20.74><24.88> bbm17}{}
+\DeclareFontShape{U}{bbm}{m}{sl}
+   {  <5> <6> <7> bbmsl8%
+      <8> <9> <10> <12> gen * bbmsl
+      <10.95> bbmsl10%
+      <14.4> <17.28> <20.74> <24.88> bbmsl12}{}
+\DeclareFontShape{U}{bbm}{bx}{n}
+   {  <5> <6> <7> <8> <9> <10> <12> gen * bbmbx
+      <10.95> bbmbx10%
+      <14.4> <17.28> <20.74> <24.88> bbmbx12}{}
+\DeclareFontShape{U}{bbm}{bx}{sl}
+   {  <5> <6> <7> <8> <9> <10> <10.95> <12> <14.4> <17.28>%
+      <20.74> <24.88> bbmbxsl10}{}
+\DeclareFontShape{U}{bbm}{b}{n}
+   {  <5> <6> <7> <8> <9> <10> <10.95> <12> <14.4> <17.28>%
+      <20.74> <24.88> bbmb10}{}
+\DeclareMathAlphabet{\mathbbm}{U}{bbm}{m}{n}
+\SetMathAlphabet\mathbbm{bold}{U}{bbm}{bx}{n}
+
+\usepackage{pifont}
+\newcommand{\cmark}{\ding{51}}
+\newcommand{\xmark}{\ding{55}}
+\DeclareSymbolFont{extraup}{U}{zavm}{m}{n}
+\DeclareMathSymbol{\varheart}{\mathalpha}{extraup}{86}
+
+\graphicspath{{images/}}
+
+\usepackage[protrusion=true,expansion=true]{microtype}
+
+\setlength\parskip{\medskipamount}
+\setlength\parindent{0pt}
+
+\title{A modal logical multiverse for commutative algebra and combinatorics}
+
+\author{Ingo Blechschmidt}
+\date{November 7th, 2022}
+
+%\useinnertheme[shadow=true]
+\setbeamerfont{block title}{size={}}
+
+\useinnertheme{rectangles}
+
+\usecolortheme{orchid}
+\usecolortheme{seahorse}
+\definecolor{mypurple}{RGB}{253,73,34}
+\setbeamercolor{structure}{fg=mypurple}
+\definecolor{myred}{RGB}{150,0,0}
+%\setbeamercolor*{title}{bg=myred,fg=white}
+%\setbeamercolor*{titlelike}{bg=myred,fg=white}
+\setbeamercolor*{title}{bg=mypurple,fg=white}
+\setbeamercolor*{titlelike}{bg=mypurple,fg=white}
+\setbeamercolor{frame}{bg=black}
+
+\usefonttheme{serif}
+\usepackage[T1]{fontenc}
+\usepackage{libertine}
+
+% lifted from https://arxiv.org/abs/1506.08870
+\DeclareFontFamily{U}{min}{}
+\DeclareFontShape{U}{min}{m}{n}{<-> udmj30}{}
+\newcommand\yon{\!\text{\usefont{U}{min}{m}{n}\symbol{'210}}\!}
+
+\newcommand{\A}{\mathcal{A}}
+\newcommand{\B}{\mathcal{B}}
+\newcommand{\C}{\mathcal{C}}
+\newcommand{\M}{\mathcal{M}}
+\renewcommand{\AA}{\mathbb{A}}
+\newcommand{\BB}{\mathbb{B}}
+\newcommand{\pp}{\mathbbm{p}}
+\newcommand{\MM}{\mathbb{M}}
+\newcommand{\E}{\mathcal{E}}
+\newcommand{\F}{\mathcal{F}}
+\newcommand{\FF}{\mathbb{F}}
+\newcommand{\G}{\mathcal{G}}
+\newcommand{\J}{\mathcal{J}}
+\newcommand{\GG}{\mathbb{G}}
+\renewcommand{\O}{\mathcal{O}}
+\newcommand{\K}{\mathcal{K}}
+\newcommand{\NN}{\mathbb{N}}
+\newcommand{\QQ}{\mathbb{Q}}
+\newcommand{\RR}{\mathbb{R}}
+\newcommand{\TT}{\mathbb{T}}
+\newcommand{\PP}{\mathbb{P}}
+\newcommand{\ZZ}{\mathbb{Z}}
+\newcommand{\CC}{\mathbb{C}}
+\renewcommand{\P}{\mathcal{P}}
+\newcommand{\aaa}{\mathfrak{a}}
+\newcommand{\ppp}{\mathfrak{p}}
+\newcommand{\fff}{\mathfrak{f}}
+\newcommand{\mmm}{\mathfrak{m}}
+\newcommand{\defeq}{\vcentcolon=}
+\newcommand{\defeqv}{\vcentcolon\equiv}
+\newcommand{\Sh}{\mathrm{Sh}}
+\newcommand{\GL}{\mathrm{GL}}
+\newcommand{\Zar}{\mathrm{Zar}}
+\newcommand{\op}{\mathrm{op}}
+\newcommand{\Set}{\mathrm{Set}}
+\newcommand{\Eff}{\mathrm{Ef{}f}}
+\newcommand{\Sch}{\mathrm{Sch}}
+\newcommand{\Aff}{\mathrm{Aff}}
+\newcommand{\Ring}{\mathrm{Ring}}
+\newcommand{\LocRing}{\mathrm{LocRing}}
+\newcommand{\LRS}{\mathrm{LRS}}
+\newcommand{\Hom}{\mathrm{Hom}}
+\newcommand{\Spec}{\mathrm{Spec}}
+\newcommand{\lra}{\longrightarrow}
+\newcommand{\RelSpec}{\operatorname{Spec}}
+\renewcommand{\_}{\mathpunct{.}}
+\newcommand{\?}{\,{:}\,}
+\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
+\newcommand{\ul}[1]{\underline{#1}}
+\newcommand{\affl}{\ensuremath{{\ul{\ensuremath{\AA}}^1}}}
+\newcommand{\Ll}{\text{iff}}
+\newcommand{\inv}{inv.\@}
+\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}
+\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates
+
+\setbeamertemplate{blocks}[rounded][shadow=false]
+
+\newenvironment{indentblock}{%
+  \list{}{\leftmargin\leftmargin}%
+  \item\relax
+}{%
+  \endlist
+}
+
+% Adapted from https://latex.org/forum/viewtopic.php?t=2251 (Stefan Kottwitz)
+\newenvironment<>{hilblock}{
+  \begin{center}
+    \begin{minipage}{9.05cm}
+      \setlength{\textwidth}{9.05cm}
+      \begin{actionenv}#1
+        \def\insertblocktitle{}
+        \par
+        \usebeamertemplate{block begin}}{
+        \par
+        \usebeamertemplate{block end}
+      \end{actionenv}
+    \end{minipage}
+  \end{center}}
+
+\newenvironment{changemargin}[2]{%
+  \begin{list}{}{%
+    \setlength{\topsep}{0pt}%
+    \setlength{\leftmargin}{#1}%
+    \setlength{\rightmargin}{#2}%
+    \setlength{\listparindent}{\parindent}%
+    \setlength{\itemindent}{\parindent}%
+    \setlength{\parsep}{\parskip}%
+  }%
+  \item[]}{\end{list}}
+
+\tikzset{
+  invisible/.style={opacity=0,text opacity=0},
+  visible on/.style={alt={#1{}{invisible}}},
+  alt/.code args={<#1>#2#3}{%
+    \alt<#1>{\pgfkeysalso{#2}}{\pgfkeysalso{#3}}}
+}
+
+\newcommand{\pointthis}[3]{%
+  \tikz[remember picture,baseline]{
+    \node[anchor=base,inner sep=0,outer sep=0] (#2) {#2};
+    \node[visible on=#1,overlay,rectangle callout,rounded corners,callout relative pointer={(0.3cm,0.5cm)},fill=blue!20] at ($(#2.north)+(-0.1cm,-1.1cm)$) {#3};
+  }%
+}
+
+\tikzset{
+  invisible/.style={opacity=0,text opacity=0},
+  visible on/.style={alt={#1{}{invisible}}},
+  alt/.code args={<#1>#2#3}{%
+    \alt<#1>{\pgfkeysalso{#2}}{\pgfkeysalso{#3}}}
+}
+
+\newcommand{\hcancel}[5]{%
+  \tikz[baseline=(tocancel.base)]{
+    \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
+    \draw[red!80, line width=0.4mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
+  }%
+}
+
+\newcommand{\explain}[7]{%
+  \tikz[remember picture,baseline]{
+    \node[anchor=base,inner sep=2pt,outer sep=0,fill=#3,rounded corners] (label) {#1};
+    \node[anchor=north,visible on=<#2>,overlay,rectangle callout,rounded corners,callout
+    relative pointer={(0.0cm,0.5cm)+(0.0cm,#6)},fill=#3] at ($(label.south)+(0,-0.3cm)+(#4,#5)$) {#7};
+  }%
+}
+
+\newcommand{\explainstub}[2]{%
+  \tikz[remember picture,baseline]{
+    \node[anchor=base,inner sep=2pt,outer sep=0,fill=#2,rounded corners] (label) {#1};
+  }%
+}
+
+\newcommand{\squiggly}[1]{%
+  \tikz[remember picture,baseline]{
+    \node[anchor=base,inner sep=0,outer sep=0] (label) {#1};
+    \draw[thick,color=red!80,decoration={snake,amplitude=0.5pt,segment
+    length=3pt},decorate] ($(label.south west) + (0,-2pt)$) -- ($(label.south east) + (0,-2pt)$);
+  }%
+}
+
+% Adapted from https://latex.org/forum/viewtopic.php?t=2251 (Stefan Kottwitz)
+\newenvironment<>{varblock}[2]{\begin{varblockextra}{#1}{#2}{}}{\end{varblockextra}}
+\newenvironment<>{varblockextra}[3]{
+  \begin{center}
+    \begin{minipage}{#1}
+      \begin{actionenv}#4
+        {\centering \hil{#2}\par}
+	\def\insertblocktitle{}%\centering #2}
+        \def\varblockextraend{#3}
+	\usebeamertemplate{block begin}}{
+        \par
+        \usebeamertemplate{block end}
+        \varblockextraend
+      \end{actionenv}
+    \end{minipage}
+  \end{center}}
+
+\setbeamertemplate{headline}{}
+
+\setbeamertemplate{frametitle}{%
+  \leavevmode%
+  \vskip-1.6em%
+  \begin{beamercolorbox}[dp=1ex,center,wd=\paperwidth,ht=2.25ex]{title}%
+    \bf\insertframetitle
+  \end{beamercolorbox}%
+  \vskip-0.2em%
+}
+
+\setbeamertemplate{navigation symbols}{}
+
+\newcounter{framenumberpreappendix}
+\newcommand{\backupstart}{
+  \setcounter{framenumberpreappendix}{\value{framenumber}}
+}
+\newcommand{\backupend}{
+  \addtocounter{framenumberpreappendix}{-\value{framenumber}}
+  \addtocounter{framenumber}{\value{framenumberpreappendix}}
+}
+
+\newcommand{\insertframeextra}{}
+\setbeamertemplate{footline}{%
+  \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex,dp=1ex,right,rightskip=1mm,leftskip=1mm]{}%
+    % \inserttitle
+    \hfill
+    \insertframenumber\insertframeextra\,/\,\inserttotalframenumber
+  \end{beamercolorbox}%
+  \vskip0pt%
+}
+
+\newcommand{\hil}[1]{{\usebeamercolor[fg]{item}{\textbf{#1}}}}
+\newcommand{\bad}[1]{\textcolor{red!90}{\textnormal{#1}}}
+
+\newcommand{\bignumber}[1]{%
+  \renewcommand{\insertenumlabel}{#1}\scalebox{1.2}{\!\usebeamertemplate{enumerate item}\!}
+}
+\newcommand{\normalnumber}[1]{%
+  {\renewcommand{\insertenumlabel}{#1}\!\usebeamertemplate{enumerate item}\!}
+}
+\newcommand{\bigheart}{\includegraphics{heart}}
+
+\newcommand{\subhead}[1]{{\centering\textcolor{gray}{\hrulefill}\quad\textnormal{#1}\quad\textcolor{gray}{\hrulefill}\par}}
+
+\newcommand{\badbox}[1]{\colorbox{red!30}{#1}}
+
+% taken from JDH "The modal logic of arithmetic potentialism and the universal algorithm"
+\DeclareMathOperator{\possible}{\text{\tikz[scale=.6ex/1cm,baseline=-.6ex,rotate=45,line width=.1ex]{\draw (-1,-1) rectangle (1,1);}}}
+\DeclareMathOperator{\necessary}{\text{\tikz[scale=.6ex/1cm,baseline=-.6ex,line width=.1ex]{\draw (-1,-1) rectangle (1,1);}}}
+\DeclareMathOperator{\xpossible}{\text{\tikz[scale=.6ex/1cm,baseline=-.6ex,rotate=45,line width=.1ex]{\draw (-1,-1) rectangle (1,1); \draw[very thin] (-.6,-.6) rectangle (.6,.6);}}}
+\DeclareMathOperator{\xnecessary}{\text{\tikz[scale=.6ex/1cm,baseline=-.6ex,line width=.1ex]{\draw (-1,-1) rectangle (1,1); \draw[very thin] (-.6,-.6) rectangle (.6,.6);}}}
+
+\begin{document}
+
+\addtocounter{framenumber}{-1}
+
+%\setbeamertemplate{headline}{\mynav{gray}{gray}{gray}}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{1.59cm}\includegraphics[width=\paperwidth]{forest-light}\end{minipage}}
+\begin{frame}[c]
+  \centering
+
+  \bigskip
+  %\includegraphics[height=0.32\textwidth]{olivia-lattices}
+  \bigskip
+  \bigskip
+  \bigskip
+
+  \scriptsize
+  \textit{-- an invitation --}
+
+  \setbeamercolor{block body}{bg=black!100}
+  \begin{block}{}
+  \centering\normalsize\color{white}
+  A \hil{modal logical multiverse} for \\ commutative algebra and combinatorics
+  \end{block}
+
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+
+  \emph{Séminaire Général de Logique} \\
+  IMJ-PRG
+  \bigskip
+
+  Paris \\
+  November 7th, 2022
+  \bigskip
+
+  Ingo Blechschmidt \\
+  University of Augsburg
+  \par
+\end{frame}}
+
+\definecolor{mypurple}{RGB}{150,0,255}
+\setbeamercolor{structure}{fg=mypurple}
+
+\section{Motivation}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1-lighter}\end{minipage}}
+\begin{frame}{Three proofs}
+  \begin{center}\bignumber{1}\end{center}
+  \vspace*{-1.3em}
+  \begin{varblock}{\textwidth}{}
+    \justifying
+    \textbf{Theorem.}
+    For every number~$n$, there is a prime larger than~$n$.
+  \end{varblock}
+  \vspace*{-0.4em}
+  {\emph{Proof.} Any prime factor of~$n! + 1$ will do. \qed}
+  \pause
+  \medskip
+
+  \begin{center}\bignumber{2}\end{center}
+  \vspace*{-1.3em}
+  \begin{varblock}{\textwidth}{}
+    \justifying
+    \textbf{Theorem.}
+    Every infinite sequence~$\alpha : \NN \to \NN$ is \emph{good}
+    in that there are numbers~$i < j$ such that~$\alpha(i) \leq \alpha(j)$.
+  \end{varblock}
+  \vspace*{-0.4em}
+  {\emph{Proof.} By~\badbox{\textsc{lem}}, there is a minimal value~$\alpha(i)$.
+  Set~$j \defeq i + 1$.\qed}
+  \pause
+  \medskip
+
+  \begin{center}\bignumber{3}\end{center}
+  \vspace*{-1.3em}
+  \begin{varblock}{\textwidth}{}
+    \justifying
+    \textbf{Theorem.}
+    Let~$M$ be a surjective matrix with more rows than columns over a
+    commutative ring~$A$.
+    Then~$1 = 0$ in~$A$.
+  \end{varblock}
+  \vspace*{-0.4em}
+  {\justifying\emph{Proof.}
+  \only<3>{\bad{Assume not.} Then there is~a \bad{maximal ideal} $\mmm$.
+  The matrix is surjective over~$A/\mmm$, contradicting
+  basic linear algebra.}\only<4>{Write~$M = \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right)$. By surjectivity,
+  have~$u, v$ with
+  $
+    u \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}1\\0\end{smallmatrix}\right)
+  $ and $
+    v \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}0\\1\end{smallmatrix}\right).
+  $
+  Hence
+  $
+    1 = (vy) (ux) = (uy) (vx) = 0
+  $.}
+  \qed\par}
+\end{frame}}
+% XXX: chart of ideals of ℤ
+
+\begin{frame}{Three take-aways}
+  \bigskip
+  \begin{columns}[c]
+    \begin{column}{0.25\textwidth}
+      \includegraphics[width=\textwidth,valign=c]{turing-machine}
+    \end{column}
+    \begin{column}{0.75\textwidth}
+      Abstract proofs should be blueprints for concrete ones.
+    \end{column}
+  \end{columns}
+  \pause
+  \bigskip
+  \bigskip
+
+  \begin{columns}[c]
+    \begin{column}{0.25\textwidth}
+      \includegraphics[width=\textwidth,valign=c]{3-adic-numbers}
+    \end{column}
+    \begin{column}{0.75\textwidth}
+      For every inhabited set~$X$,
+      there is a \hil{flavor~of~parametrized~mathematics} \\
+      which is home
+      to a gadget called \\
+      \hil{generic surjection}~$\NN \twoheadrightarrow X$.
+    \end{column}
+  \end{columns}
+  \pause
+  \bigskip
+  \bigskip
+
+  \begin{columns}[c]
+    \begin{column}{0.25\textwidth}
+      \includegraphics[width=\textwidth,valign=c]{multiverse}
+    \end{column}
+    \begin{column}{0.75\textwidth}
+      There is a \hil{rich multiverse} of \\
+      flavors of parametrized mathematics.
+    \end{column}
+  \end{columns}
+\end{frame}
+
+\end{document}
+
 A modal logical multiverse for commutative algebra and combinatorics

### 2022-11-04 slides-paris2022.tex
More graphics

@@
 
+\input{images/primes.tex}
+
 \begin{document}
@@
 \begin{frame}{Three proofs}
-  \begin{center}\bignumber{1}\end{center}
-  \vspace*{-1.3em}
+  \begin{columns}[c]
+    \begin{column}{0.15\textwidth}
+      \scalebox{0.23}{\sieve{6}{6}}
+    \end{column}
+    \begin{column}{0.85\textwidth}
       \begin{varblock}{\textwidth}{}
         \justifying
-    \textbf{Theorem.}
-    For every number~$n$, there is a prime larger than~$n$.
+        \textbf{Thm.}
+        For every~$n$, there is a prime larger than~$n$.
       \end{varblock}
       \vspace*{-0.4em}
-  {\emph{Proof.} Any prime factor of~$n! + 1$ will do. \qed}
+      \emph{Proof.} Any prime factor of~$n! + 1$ will do. \qed
+    \end{column}
+  \end{columns}
   \pause
-  \medskip
 
-  \begin{center}\bignumber{2}\end{center}
-  \vspace*{-1.3em}
+  \begin{columns}[c]
+    \begin{column}{0.15\textwidth}
+      \includegraphics[width=\textwidth,valign=t]{monad}
+    \end{column}
+    \begin{column}{0.85\textwidth}
       \begin{varblock}{\textwidth}{}
         \justifying
-    \textbf{Theorem.}
+        \textbf{Thm.}
         Every infinite sequence~$\alpha : \NN \to \NN$ is \emph{good}
@@
       \vspace*{-0.4em}
-  {\emph{Proof.} By~\badbox{\textsc{lem}}, there is a minimal value~$\alpha(i)$.
-  Set~$j \defeq i + 1$.\qed}
+      \mbox{\emph{Proof.} There is a \bad{minimal value}~$\alpha(i)$.
+      Set~$j \defeq i{+}1$.\qed}
+    \end{column}
+  \end{columns}
   \pause
-  \medskip
 
-  \begin{center}\bignumber{3}\end{center}
-  \vspace*{-1.3em}
+  \begin{columns}[c]
+    \begin{column}{0.15\textwidth}
+      \includegraphics[width=\textwidth,valign=t]{calabi-yau}
+    \end{column}
+    \begin{column}{0.85\textwidth}
       \begin{varblock}{\textwidth}{}
         \justifying
-    \textbf{Theorem.}
+        \textbf{Thm.}
         Let~$M$ be a surjective matrix with more rows than columns over a
-    commutative ring~$A$.
-    Then~$1 = 0$ in~$A$.
+        ring~$A$. Then~$1 = 0$ in~$A$.
       \end{varblock}
       \vspace*{-0.4em}
-  {\justifying\emph{Proof.}
+      \justifying\emph{Proof.}
       \only<3>{\bad{Assume not.} Then there is~a \bad{maximal ideal} $\mmm$.
-  The matrix is surjective over~$A/\mmm$, contradicting
-  basic linear algebra.}\only<4>{Write~$M = \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right)$. By surjectivity,
+      The matrix is surjective over~$A/\mmm$. Since~$A/\mmm$ is a field, this
+      is a contradiction to basic linear algebra. \\[1.3em]\ }\only<4>{Write~$M =
+      \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right)$. By surjectivity,
       have~$u, v$ with
-  $
+      \vspace*{-0.6em}
+      \[
         u \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}1\\0\end{smallmatrix}\right)
-  $ and $
+        \quad\text{and}\quad
         v \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}0\\1\end{smallmatrix}\right).
-  $
+      \]
+
+      \vspace*{-0.6em}
       Hence
@@
       $.}
-  \qed\par}
+      \qed
+    \end{column}
+  \end{columns}
 \end{frame}}
 % XXX: chart of ideals of ℤ
+
+% \begin{document}
 
@@
     \begin{column}{0.75\textwidth}
-      There is a \hil{rich multiverse} of \\
+      There is a \hil{rich modal multiverse} of \\
       flavors of parametrized mathematics.
@@
   \end{columns}
+
+  \visible<3->{\begin{tikzpicture}[overlay]
+    \node[anchor=south east,inner sep=0] (image) at (10.8,0.3) {
+      \reflectbox{\includegraphics[width=1em]{butterfly}}
+    };
+  \end{tikzpicture}}
 \end{frame}

### 2022-11-05 slides-paris2022.tex
Slides almost finished

@@
 \newcommand{\aaa}{\mathfrak{a}}
+\newcommand{\ccc}{\mathfrak{c}}
 \newcommand{\ppp}{\mathfrak{p}}
@@
 \newcommand{\badbox}[1]{\colorbox{red!30}{#1}}
+\newcommand{\infobox}[1]{\colorbox{yellow!70}{\color{black}#1}}
 
@@
 
+% \begin{document}
+
+\section{The set-theoretic multiverse}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.59cm}\includegraphics[width=\paperwidth]{staircase}\end{minipage}}
+\begin{frame}{A brief timeline}
+  \begin{itemize}
+    \item[18xx] Cantor studies the \hil{continuum hypothesis}, the claim
+    that~$2^{\aleph_0} = \aleph_1$.
+    \item[19xx] Zermelo and Fraenkel offer the axioms of~\textsc{zfc}.
+    \item[1938] Gödel proves: If~\textsc{zfc} is consistent, then so
+    is~\textsc{zfc}+\textsc{ch}.
+    \item[1963] Cohen proves: If~\textsc{zfc} is consistent, then so
+    is~\textsc{zfc}+$\neg$\textsc{ch}.
+    \item[since 1963 and ongoing] Set theorists pursue additional axioms to
+    settle~\textsc{ch} (one way or another).  % XXX persue?
+    \item[201x] Hamkins offers his paper on the \hil{multiverse position} in
+    the philosophy of set theory, arguing that the program of pursing
+    additional axioms (while successful in many ways) is doomed to fail in
+    settling~\textsc{ch}.
+  \end{itemize}
+\end{frame}
+\begin{frame}{The set-theoretic multiverse}
+  \begin{tikzpicture}[remember picture,overlay]
+    \node[xshift=-1.5cm,yshift=-3.5cm] at (current page.north east)
+    {\includegraphics[width=2cm]{branching}};
+  \end{tikzpicture}
+  \vspace*{-1.5em}
+
+  \begin{block}{}
+    \justifying
+    \textbf{Def.} A \hil{model of set theory} is a (perhaps class-sized) structure $(M,{\in})$
+    satisfying axioms such as those of~\textsc{zfc}.
+  \end{block}
+  \vspace*{-0.6em}
+  {\small\emph{Examples.}\vspace*{-0.9em}
+  \begin{itemize}
+    \item $V$, the class of all sets \vspace*{-0.6em}
+    \item $L$, Gödel's constructible universe \vspace*{-0.6em}
+    \item $V[G]$, a forcing extension containing a generic filter~$G$ of \\ some
+    poset of forcing conditions \vspace*{-0.6em}
+    \item Henkin/term models from consistency of (extensions of)~\textsc{zfc}
+  \end{itemize}}
+  \pause
+
+  %{\centering\emph{Embracing all models of set theory:}\par}
+  \begin{block}{}
+    \justifying
+    \textbf{Def.} $\possible\varphi$ iff~$\varphi$ holds in \hil{some extension} of
+    the current universe. \\
+    \phantom{\textbf{Def.}} $\necessary\varphi$ iff~$\varphi$ holds in \hil{all extensions}
+    of the current universe.
+  \end{block}
+  \vspace*{-0.6em}
+  \begin{itemize}
+    \item $\necessary(\possible\textsc{CH} \wedge \possible\neg\text{CH})$,
+    the continuum hypothesis is a \hil{switch}
+    \item \ \\[-1.2em]\mbox{$\necessary\possible\necessary(\text{$X$ is countable})$,
+    existence of an enumeration is a \hil{button}}
+  \end{itemize}
+  \pause
+  \medskip
+
+  \textbf{NB.} In the multiverse of extensions of a given field~$K$,
+
+  \vspace*{-0.5em}
+  \begin{itemize}
+    \item ``there is a square root of~$-1$'' is a button and
+    \item ``every polynomial splits into linear factors'' is a switch.
+  \end{itemize}
+\end{frame}}
+
+% \begin{document}
+
+\section{Parametrized mathematics}
+
+\renewcommand{\insertframeextra}{a}
+\begin{frame}{Parametrized mathematics}
+  \medskip
+  {\centering\includegraphics[width=0.7\textwidth]{eigenvector}\par}
+  \small
+
+  \begin{changemargin}{-0.5em}{-1.5em}
+  \begin{enumerate}
+    \item[\infobox{local}]
+    \ \\[-1.2em]\mbox{``Every real symmetric matrix
+    does \only<4>{\hil{not not} }have an eigenvector.''
+    \good{\cmark}}
+    \medskip
+
+    \item[\infobox{global}] ``For every continuous family of symmetric
+    matrices, \\\only<4>{\hil{on a dense open} }eigenvectors can locally
+    be picked continuously.''
+    \only<1-2>{\includegraphics[height=0.7em]{question-mark}}%
+    \only<3>{\bad{\xmark}}%
+    \only<4>{\good{\cmark}}
+
+    \smallskip
+    \pause
+    \justifying
+    \scriptsize
+    ``Let~$X$ be a topological space and let~$A : X \to
+    M^\text{sym}_n(\RR)$ be a continuous map to the space of symmetric~$(n \times
+    n)$-matrices. Then there is an open covering~$\bigcup_{i \in I} U_i$
+    \only<1-3>{of~$X$}\only<4>{\hil{of a dense open subset}~$U \subseteq X$} such
+    that or all indices~$i \in I$, there is a continuous map~$v : U_i \to \RR^n$
+    such that for all~$x \in U_i$, the vector~$v(x)$ is an eigenvector
+    of~$A(x)$.''
+  \end{enumerate}
+  \end{changemargin}
+\end{frame}
+
+\addtocounter{framenumber}{-1}
+\renewcommand{\insertframeextra}{b}
+\begin{frame}{Parametrized mathematics}
+  \bigskip
+  \bigskip
+  {\centering
+    \includegraphics[height=9em]{damian-gvirtz-moduli-1}
+    \quad
+    \includegraphics[height=9em]{damian-gvirtz-moduli-2}
+    \par}
+  \small
+  \bigskip
+
+  \begin{changemargin}{-0.5em}{-1.5em}
+  \begin{enumerate}
+    \item[\infobox{local}]
+    \ \\[-1.2em]``Let~$M$ be a finitely generated module over a
+    field~$k$. \\ Then~$M$ is \only<4>{\hil{not not} }finite free.''
+    \good{\cmark}
+    \medskip
+
+    \item[\infobox{global}] ``Let~$M$ be a finitely generated module over a
+    ring~$A$. \\ Then~$M^\sim$ is finite locally
+    free\only<4>{ \hil{on a dense open}}.''
+    \only<1-2>{\includegraphics[height=0.7em]{question-mark}}%
+    \only<3>{\bad{\xmark}}%
+    \only<4>{\good{\cmark}}
+
+    \smallskip
+    \pause
+    \justifying
+    \scriptsize
+    ``Let~$M$ be a finitely generated module over an
+    arbitrary commutative ring~$A$. \only<1-3>{Then there is a partition~$1 = f_1 + \cdots +
+    f_n \in A$ of unity such that, for each index~$i$, the localized
+    module~$M[f_i^{-1}]$ is finite free over~$A[f_i^{-1}]$.}\only<4>{\hil{If~$f = 0$
+    is the only element of~$A$} such that~$M[f^{-1}]$ is finite free
+    over~$A[f^{-1}]$, \hil{then~$1 = 0$ in~$A$}.}''
+  \end{enumerate}
+  \end{changemargin}
+\end{frame}
+
+\addtocounter{framenumber}{-1}
+\renewcommand{\insertframeextra}{c}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{-1em}\includegraphics[width=\paperwidth]{train-tracks-cropped}\end{minipage}}
+\begin{frame}{Parametrized mathematics}
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \small
+
+  \begin{changemargin}{-0.5em}{-1.5em}
+  \begin{enumerate}
+    \item[\infobox{local}]
+    \ \\[-1.2em]``Let $R$ be a ring.
+    Let $n \geq 0$ be an integer.
+    We have
+    $$
+    H^q(\mathbf{P}^n, \mathcal{O}_{\mathbf{P}^n_R}(d)) = \textit{(omitted)}
+    $$
+    as $R$-modules.''
+    \good{\cmark}
+    \medskip
+
+    \item[\infobox{global}] ``Let $S$ be a scheme. Let $n \geq 1$.
+    Let $\mathcal{E}$ be a finite locally
+    free $\mathcal{O}_S$-module of constant rank $n + 1$.
+    For the structure morphism
+    $
+    \pi : \mathbf{P}(\mathcal{E}) \longrightarrow S,
+    $
+    we have
+    $$
+    R^q\pi_*(\mathcal{O}_{\mathbf{P}(\mathcal{E})}(d)) =
+    \textit{(omitted)}
+    $$
+    as sheaves of~$\mathcal{O}_S$-modules.'' \good{\cmark}
+  \end{enumerate}
+  \end{changemargin}
+\end{frame}}
+
+% \begin{document}
+
+\renewcommand{\insertframeextra}{}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
+\begin{frame}{Sites and toposes}
+  For every \hil{site}~$\C$, the
+  \hil{sheaves} on~$\C$ form a \hil{topos}. \emph{Notable sites:}
+  \begin{enumerate}
+    \item The \hil{site of opens} of a topological space
+    \item The \hil{Zariski site} of a ring
+    \item The \hil{site of finite approximations} of surjections~$\NN
+    \twoheadrightarrow X$
+    \item The \hil{classifying site} of a geometric theory
+  \end{enumerate}
+  \pause
+
+  Every topos supports an \hil{internal language},
+  and this language is sound with respect to \hil{intuitionistic
+  logic}.
+  \pause
+
+  \begin{block}{}
+    \justifying
+    \textbf{Def.} A statement~$\varphi$ holds \ldots
+    \begin{itemize}
+      \item\small \hil{everywhere} ($\necessary\varphi$) iff
+      it holds in every (Grothendieck) topos (over the current base topos).
+      \item \hil{somewhere} ($\possible\varphi$) iff
+      it holds in some positive topos.
+      \item \hil{proximally} ($\xpossible\varphi$) iff
+      it holds in some positive ouvert topos.
+    \end{itemize}
+  \end{block}
+\end{frame}
+\begin{frame}{The multiverse of parametrized mathematics}
+  \begin{block}{}
+    \justifying
+    \textbf{Def.} A statement~$\varphi$ holds \ldots
+    \begin{itemize}
+      \item\small \hil{everywhere} ($\necessary\varphi$) iff
+      it holds in every (Grothendieck) topos (over the current base topos).
+      \item \hil{somewhere} ($\possible\varphi$) iff
+      it holds in some positive topos.
+      \item \hil{proximally} ($\xpossible\varphi$) iff
+      it holds in some positive ouvert topos.
+    \end{itemize}
+  \end{block}
+
+  \begin{changemargin}{-1.2em}{-1.2em}
+    \small
+    \begin{enumerate}
+      %\item If a function~$\NN \to \NN$ does \emph{everywhere} \emph{not~not} have
+      %a zero, then it actually has a zero.
+      \item \emph{Somewhere,} the law of
+      excluded middle holds. \emph{In fact, we even have:} \\
+      It is \emph{everywhere} the case that the law of
+      excluded middle holds \emph{somewhere}.
+      \smallskip
+      \item Assuming Zorn's lemma, it is
+      everywhere the case that it the axiom of choice holds
+      somewhere. [Barr]
+      \smallskip
+      \item If a geometric implication holds \emph{somewhere}, then it holds already here.
+      \smallskip
+      \item If a first-order statement holds \emph{proximally}, then it holds already here.
+      \smallskip
+      \item For every set~$X$, \emph{proximally} there is a
+      surjection~$\NN \twoheadrightarrow X$.
+      \smallskip
+      \item Every ring \emph{proximally} has a maximal ideal.
+    \end{enumerate}
+  \end{changemargin}
+\end{frame}}
+
 \end{document}
@@
 
+Menagerie of properties of the multiverse
+
 === Relaxation to topos-theoretic case

### 2022-11-06 slides-paris2022.tex
Fixed timeline

@@
 
-% \begin{document}
-
 \begin{frame}{Three take-aways}
@@
 {\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.59cm}\includegraphics[width=\paperwidth]{staircase}\end{minipage}}
+% https://tex.stackexchange.com/a/198372/32372
+\newcommand\ytl[2]{%
+  \parbox[b]{4.5em}{\hfill{#1}~$\cdots\cdots$~}%
+  \makebox[0pt][c]{\color{mypurple}$\bullet$}{\color{mypurple}\vrule}\quad%
+  \parbox[c]{8.4cm}{\vspace{7pt}\raggedright#2\\[7pt]}%
+  \\[-2.0pt]}
 \begin{frame}{A brief timeline}
-  \begin{itemize}
-    \item[18xx] Cantor studies the \hil{continuum hypothesis}, the claim
-    that~$2^{\aleph_0} = \aleph_1$.
-    \item[19xx] Zermelo and Fraenkel offer the axioms of~\textsc{zfc}.
-    \item[1938] Gödel proves: If~\textsc{zfc} is consistent, then so
-    is~\textsc{zfc}+\textsc{ch}.
-    \item[1963] Cohen proves: If~\textsc{zfc} is consistent, then so
-    is~\textsc{zfc}+$\neg$\textsc{ch}.
-    \item[since 1963 and ongoing] Set theorists pursue additional axioms to
-    settle~\textsc{ch} (one way or another).  % XXX persue?
-    \item[201x] Hamkins offers his paper on the \hil{multiverse position} in
-    the philosophy of set theory, arguing that the program of pursing
-    additional axioms (while successful in many ways) is doomed to fail in
-    settling~\textsc{ch}.
-  \end{itemize}
+  \vspace*{-1.0em}
+  \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-3.5pt]
+  \ytl{1878}{Cantor advances the \hil{continuum hypothesis}, the claim
+  that~$2^{\aleph_0} = \aleph_1$.}
+  \pause
+  \ytl{1910s}{Zermelo--Fraenkel set theory emerges.}
+  \pause
+  \ytl{1920s}{Set theorists pursue additional axioms to
+  settle~\textsc{ch} (one way or another).}
+  \pause
+  \ytl{1938}{Gödel proves: If~\textsc{zfc} is consistent, so
+  is~\textsc{zfc}+\textsc{ch}.}
+  \pause
+  \ytl{1963}{Cohen proves: If~\textsc{zfc} is consistent, so
+  is~\textsc{zfc}+$\neg$\textsc{ch}.}
+  \pause
+  \ytl{2011}{Hamkins offers his paper on the \hil{multiverse position} in
+  the philosophy of set theory.}
+  \pause
+  %arguing that the program of pursuing
+  %additional axioms (while successful in many ways) is doomed to fail
+  %to settle~\textsc{ch}.}
+  \ytl{2016}{Oldenziel proposes to study the modal multiverse of parametrized
+  mathematics.}
+  \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
+  \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
+  \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
 \end{frame}
@@
 \end{frame}}
-
-% \begin{document}
 
@@
 
-% \begin{document}
-
 \renewcommand{\insertframeextra}{}

### 2022-11-07 slides-paris2022.tex
typo
Final version

@@
 
-%\setbeamertemplate{headline}{\mynav{gray}{gray}{gray}}
-
 {\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{1.59cm}\includegraphics[width=\paperwidth]{forest-light}\end{minipage}}
@@
   \bigskip
-  %\includegraphics[height=0.32\textwidth]{olivia-lattices}
   \bigskip
@@
   \begin{columns}[c]
-    \begin{column}{0.25\textwidth}
+    \begin{column}{0.2\textwidth}
       \includegraphics[width=\textwidth,valign=c]{turing-machine}
@@
   \begin{columns}[c]
-    \begin{column}{0.25\textwidth}
+    \begin{column}{0.2\textwidth}
       \includegraphics[width=\textwidth,valign=c]{3-adic-numbers}
@@
   \begin{columns}[c]
-    \begin{column}{0.25\textwidth}
+    \begin{column}{0.2\textwidth}
       \includegraphics[width=\textwidth,valign=c]{multiverse}
@@
 
-% \begin{document}
-
 \section{The set-theoretic multiverse}
@@
 % https://tex.stackexchange.com/a/198372/32372
+% \includegraphics[width=3cm]{lovelace-babbage}
+% \includegraphics[width=3cm]{first-program}
+\addtocounter{framenumber}{-1}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[height=\paperheight,valign=t]{ada-lovelace}\includegraphics[width=8cm,valign=t]{first-program}\end{minipage}}
+\begin{frame}[plain]
+  \vspace*{6cm}\hspace*{6cm}\begin{minipage}{4.7cm}
+    \Large\hil{Ada Lovelace},
+
+    \large
+    the world's first \\
+    computer programmer
+    \medskip
+
+    * 1815 \ \ † 1852
+  \end{minipage}
+\end{frame}}
 \newcommand\ytl[2]{%
@@
     is the only element of~$A$} such that~$M[f^{-1}]$ is finite free
-    over~$A[f^{-1}]$, \hil{then~$1 = 0$ in~$A$}.}''
+    over~$A[f^{-1}]$, \hil{then~$1 = 0$ in~$A$}.}''\only<4>{\\\ }
   \end{enumerate}
@@
       \smallskip
-      \item For every set~$X$, \emph{proximally} there is a
+      \item For every inhabited set~$X$, \emph{proximally} there is a
       surjection~$\NN \twoheadrightarrow X$.
@@
 \end{frame}}
+
+\backupstart
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[width=\paperwidth]{eventually-monad}\end{minipage}}
+\begin{frame}
+  \vspace*{\paperheight}\vspace*{-3.4em}\centering
+  \emph{Agda formalization available.}
+\end{frame}}
+
+\backupend
 

### 2022-11-28 paper-modal-multiverse.tex
typo
to do

@@
 
-\title{The topos-theoretic multiverse}
+\title{The topos-theoretic multiverse: a modal approach}
 \author{Ingo Blechschmidt}
@@
   To help put established results in constructive algebra and constructive
-  combinatorics into perspective, give possible background for certain
+  combinatorics into perspective, construct an origin story for certain
   inductive definitions and form a unified framework for certain techniques for
@@
 \item \emph{Dependent choice:} If every element of a set is related by some relation to some other
-element, every element can be completed to an infinite chain of related
+element, then every element can be completed to an infinite chain of related
 elements.
@@
 extracted~\cite{bauer:c2c}, and every constructive proof holds also for continuous families of
-the objects in question~\cite[Section~4.3]{blechschmidt:filmat}. But the listed classical principles have no
-computational witness or fail in continuous families, hence are not
+the objects in question~\cite[Section~4.3]{blechschmidt:filmat}. In contrast, listed classical principles above have no
+computational witness and/or fail in continuous families, hence are not
 available in constructive mathematics.
@@
 \item If every element of a set is related by some relation to some other
-element, every element can \emph{proximally} be completed to an infinite chain
+element, then every element can \emph{proximally} be completed to an infinite chain
 of related elements.
@@
 ($\necessary\varphi$) iff it holds in every Grothendieck topos over the
-current base topos; a statement holds \emph{somewhere} ($\possible\varphi$)
+current base topos (makng use of the \emph{internal language} of
+toposes~\cite{xxx}, where required in the form of the \emph{sheaf semantics} in
+order to support unbounded quantification~\cite{xxx}); a statement holds \emph{somewhere} ($\possible\varphi$)
 iff it holds in some positive Grothendieck topos over the current base; and a
@@
 
-The modal language not only allows us to recover classical principles
+This modal language not only allows us to recover classical principles
 as above, but also makes some powerful theorems about the topos-theoretic
@@
 \item \label{item:prox-countable}
-For every (perhaps uncountable) set~$X$, \emph{proximally} there is a
+For every (perhaps uncountable) inhabited set~$X$, \emph{proximally} there is a
 surjection~$\NN \twoheadrightarrow X$.
@@
 For countable rings, an explicit construction of a maximal ideal is available.
-By item~(\ref{item:prox-countable}), this construction can also be carried out
+By Item~(\ref{item:prox-countable}), this construction can also be carried out
 for arbitrary rings, though the result is not a maximal ideal in the narrow
@@
 its existence, pertaining for instance to concrete statements about polynomials
-or matrices, pass down to the base topos by item~(\ref{item:prox-descends}).
-In this sense the proximally-existing maximal ideal is like the mathematical
-phantoms of Gavin Wraith~\cite{xxx}, encouraging us to broaden our notion of
+or matrices, pass down to the base topos by Item~(\ref{item:prox-descends}).
+The resulting maximal ideal can be regarded as a mathematical
+phantom in the sense of Gavin Wraith~\cite{xxx}, not existing in a narrow
+sense (in the base topos), but existing proximally and hence encouraging us to broaden our notion of
 existence because it promises us to work wonders.
@@
 We argue that the modal operators~$\necessary,\possible,\xpossible$ and more
-suggested in our proposal are natural extensions and refinements of the
+suggested are natural extensions and refinements of the
 familiar double-negation modality~$\neg\neg$ in constructive mathematics.
@@
 
+xxx: This text is set in the context of constructive mathematics.
+Our definitions and results can be formalized in~\textsc{izf} or in the
+kind of language supported by toposes.
+
 \section{Modal operators}
@@
 
-\begin{defn}A \emph{Grothendieck topos} is a category which is equivalent to
+\begin{defn}A \emph{Grothendieck topos} is a category equivalent to
 the category of sheaves on a small site. A \emph{Grothendieck topos over a
@@
 the point of view of~$\B$''; using a sufficiently expressive form of the
-internal language of~$\B$, which allows to make direct sense of the notion in
+internal language of~$\B$, which allows us to make direct sense of the notion in
 scare quotes, this amounts to a bounded geometric morphism~$\E \to
-\B$, and this can be taken as the definition.
+\B$, and this notion can be taken as the definition.
 \end{defn}
@@
 The xxx (use word of Hamkins) of a base~$\B$ is the collection of Grothendieck
-toposes over~$\B$ (which one should formalize in set-theoretic foundations more
-precisely as the proper class of small sites in~$\B$).
+toposes over~$\B$ (which in set-theoretic foundations should more precisely be
+formalized as the proper class of small sites in~$\B$).
 
 \begin{rem}Perhaps arbitrary \emph{elementary} toposes over the base, corresponding to
-possibly unbounded geometric morphisms~$\E \to \B$ or even arbitrary
-fibrations/indexed categories over~$\B$ validating an appropriate form of the
+possibly unbounded geometric morphisms~$\E \to \B$, or even arbitrary
+fibrations/indexed categories over~$\B$ (XXX: are these the same?) validating an appropriate form of the
 axioms of elementary toposes, should also be
@@
 universes in mind. Predicatively, not even the category of sets might be an
-elementary topos, but they do form an arithmetic universe and sheaves over
-sites are a meaningful concept.
+elementary topos, but the category of sets and categories of sheaves are still
+arithmetic universes.
 \end{rem}
@@
 \begin{defn}A topos~$\E$ is \emph{positive} if and only if the geometric
-morphism~$f : \E \to \Set$ is surjective, that is iff~$f^*$ reflects
-isomorphisms.\end{defn}
+morphism~$f : \E \to \Set$ is surjective (that is~$f^*$ reflects
+isomorphisms).\end{defn}
 
@@
 is positive if and only if every open covering of the top element of the frame
-of~$X$ is inhabited (xxx:verify). As such, positivity is a more informative
-version of the weaker property of being nontrivial (the property that the top open
+of~$X$ is inhabited. As such, positivity is a more informative
+version of the constructively weaker property of being nontrivial (the property that the top open
 and the bottom open do not coincide).\end{ex}
@@
 \begin{ex}A sufficient criterion for the classifying topos of a geometric
-theory~$\TT$ being positive is that xxx.\end{ex}
+theory~$\TT$ to be positive is that xxx.\end{ex}
 
@@
 \begin{defn}A topos~$\E$ is \emph{ouvert} if and only if the geometric
-morphism~$f : \E \to \Set$ is open, that is preserves the interpretation of
+morphism~$f : \E \to \Set$ is open, that is~$f^*$ preserves the interpretation of
 first-order formulas.\end{defn}
@@
 \begin{ex}The topos of sheaves over a topological space is always ouvert. More
-generally, the topos of sheaves over a locale~$X$ if ouvert iff there exists a
+generally, the topos of sheaves over a locale~$X$ is ouvert iff there exists a
 \emph{positivity predicate} on its frame of opens in the sense of~\cite{xxx}.
@@
 chains is not particularly useful; while the chain condition is satisfied for
-the intended examples, by its negativity the condition is too weak to
+the intended examples (such as~$(\NN,{<})$), by its negativity the condition is too weak to
 facilitate the intended proofs.
@@
 hierarchy of subsets of increasing universe levels, a transitive relation~$({<})$ is
-well-founded if and only if every element of~$X$ is \emph{accessible}, where
+declared well-founded if and only if every element of~$X$ is \emph{accessible}, where
 the accessibility predicate~$\mathsf{Acc}$ is inductively generated~\cite{xxx}
@@
 see~\cite{richman:xxx}), and secondly, being able to inspect suitable inductive
-witnesses helps in proving the Hilbert basis theorem~\cite{xxx}. XXX and XXX
+witnesses enables us to prove the Hilbert basis theorem~\cite{xxx}. XXX and XXX
 hence propose to call a ring \emph{Noetherian} if and only if~$\mathsf{Stalls}
@@
 apart from working well and being motivated on general constructive
-considerations? Also, constructively the inductive definitions are much
+considerations? Also: Constructively the inductive definitions are much
 stronger than their classical counterparts, equivalent only in presence
 of~\textsc{lem} and~\textsc{dc}. For instance, if a relation is almost-full in
-the inductive sense, not only is every infinite sequence good, but also every
+the inductive sense, not only is every infinite sequence good, but so is every
 infinite ``multi-valued sequence''\footnote{xxx} and every infinite
@@
 
-The answers to both questions are positive (xxx:language), and the modal
+Both questions have positive answers, and the modal
 perspective fruitfully clarifies their connection.
@@
 
+\begin{cor}\label{cor:modal-good}Let~$R$ be a relation on a set~$X$. Then~$R$
+is almost-full in the inductive sense if and only if everywhere, every infinite
+sequence is good.\end{cor}
+
+\begin{proof}For the ``if'' direction, if every infinite sequence everywhere is
+good, then in particular the generic infinite sequence is. By
+Proposition~\ref{prop:gen-good}, this statement amounts to~$R$ being
+almost-full in the inductive sense.
+
+In the converse direction, we can either argue that, since being good is
+expressible as a geometric formula, if the generic infinite sequence is good
+then so is every infinite sequence in every topos; or we argue, using
+Proposition~\ref{prop:gen-good} again, that
+the property of being almost-full in the inductive sense is stable
+under pullback along geometric morphisms and hence passes from the base topos
+to every topos. Hence (the pullback of)~$R$ is almost-full in every topos and
+hence every infinite sequence in every topos is good.\end{proof}
+
 \begin{prop}Let~$({<})$ be a transitive relation on a set~$X$. The generic
@@
 
+\begin{cor}A transitive relation is well-founded in the inductive sense
+if and only if everywhere, it is not the case that there exists an infinite
+descending chain.\end{cor}
+
+\begin{proof}Analogous to the proof of
+Corollary~\ref{cor:modal-good}.\end{proof}
+
 \section{Extracting programs from multiverse proofs}
@@
 
-\begin{prop}Let~$({\leq})$ be a transitive almost-full relation.
+\begin{prop}\label{prop:af-wf}Let~$({\leq})$ be a transitive almost-full relation.
 Then~$({<})$, where~$x < y \equiv (x \leq y \wedge \neg(y \leq x))$,
@@
 \prod_{x:X} \textsf{Acc}(x)$.
+
+\begin{rem}For the proof of Proposition~\ref{prop:af-wf}, it is not relevant
+that pullback along geometric morphisms typically fails to preserve the
+negation occuring in the definition of~$({<})$, basically because we still
+have~$f^*(\llbracket x > y \rrbracket) \wedge f^*(\llbracket x \leq y
+\rrbracket) \Rightarrow f^*(\llbracket\bot\rrbracket) = \bot$.\end{rem}
 
@@
 \end{document}
+
+Cite? https://arxiv.org/abs/2210.04838

### 2023-02-12 paper-modal-multiverse.tex
to do
Switch to llncs style for submission to CiE 2023

@@
-\documentclass[oneside,reqno]{amsart}
+\documentclass[envcountsect,envcountsame,runningheads]{llncs}
 
 \usepackage[utf8]{inputenc}
-\usepackage{amsthm,mathtools,stmaryrd,amssymb}
+\usepackage[english]{babel}
+\usepackage{amsmath,mathtools,stmaryrd,amssymb}
 \usepackage[all]{xy}
@@
 
-\usepackage[natbib=true,style=numeric,maxnames=10]{biblatex}
-\usepackage[babel]{csquotes}
-\bibliography{paper-modal-multiverse.bib}
-
-\title{The topos-theoretic multiverse: a modal approach}
-\author{Ingo Blechschmidt}
-\address{Universität Augsburg \\
-Institut für Mathematik \\
-Universitätsstr. 14 \\
-86159 Augsburg, Germany}
-\email{ingo.blechschmidt@math.uni-augsburg.de}
-
-\theoremstyle{definition}
-\newtheorem{defn}{Definition}[section]
-\newtheorem{ex}[defn]{Example}
-
-\theoremstyle{plain}
-\newtheorem{prop}[defn]{Proposition}
-\newtheorem{cor}[defn]{Corollary}
-\newtheorem{lemma}[defn]{Lemma}
-\newtheorem{thm}[defn]{Theorem}
-\newtheorem{scholium}[defn]{Scholium}
-
-\theoremstyle{remark}
-\newtheorem{rem}[defn]{Remark}
-\newtheorem{question}[defn]{Question}
-\newtheorem{speculation}[defn]{Speculation}
-\newtheorem{caveat}[defn]{Caveat}
-\newtheorem{conjecture}[defn]{Conjecture}
+\title{The topos-theoretic multiverse: \\ a modal approach for computation}
 
@@
 
+\maketitle
+
 \begin{abstract}
@@
 \end{abstract}
-
-\maketitle
-\thispagestyle{empty}
 
@@
 \textbf{Acknowledgments.} Alexander Oldenziel, \ldots
+% xxx: Iosif
 
@@
 
-\begin{defn}A \emph{Grothendieck topos} is a category equivalent to
+\begin{definition}A \emph{Grothendieck topos} is a category equivalent to
 the category of sheaves on a small site. A \emph{Grothendieck topos over a
@@
 \B$, and this notion can be taken as the definition.
-\end{defn}
+\end{definition}
 
@@
 
-\begin{rem}Perhaps arbitrary \emph{elementary} toposes over the base, corresponding to
+\begin{remark}Perhaps arbitrary \emph{elementary} toposes over the base, corresponding to
 possibly unbounded geometric morphisms~$\E \to \B$, or even arbitrary
@@
 arithmetic universes.
-\end{rem}
+\end{remark}
 
@@
 
-\begin{defn}A topos~$\E$ is \emph{positive} if and only if the geometric
+\begin{definition}A topos~$\E$ is \emph{positive} if and only if the geometric
 morphism~$f : \E \to \Set$ is surjective (that is~$f^*$ reflects
-isomorphisms).\end{defn}
+isomorphisms).\end{definition}
 
-\begin{ex}The topos of sheaves over a topological space~$X$ is positive if and
+\begin{example}The topos of sheaves over a topological space~$X$ is positive if and
 only if~$X$ has a point; more generally, the topos of sheaves over a locale~$X$
@@
 version of the constructively weaker property of being nontrivial (the property that the top open
-and the bottom open do not coincide).\end{ex}
+and the bottom open do not coincide).\end{example}
 
-\begin{ex}The spectrum of a ring~$A$, that is the classifying topos of prime
+\begin{example}The spectrum of a ring~$A$, that is the classifying topos of prime
 filters of~$A$ (or equivalently the topos of sheaves over the classifying
-locale of prime filters of~$A$) is positive iff~$1 \neq 0$ in~$A$.\end{ex}
+locale of prime filters of~$A$) is positive iff~$1 \neq 0$ in~$A$.\end{example}
 
-\begin{ex}A sufficient criterion for the classifying topos of a geometric
-theory~$\TT$ to be positive is that xxx.\end{ex}
+\begin{example}A sufficient criterion for the classifying topos of a geometric
+theory~$\TT$ to be positive is that xxx.\end{example}
 
@@
 
-\begin{defn}A topos~$\E$ is \emph{ouvert} if and only if the geometric
+\begin{definition}A topos~$\E$ is \emph{ouvert} if and only if the geometric
 morphism~$f : \E \to \Set$ is open, that is~$f^*$ preserves the interpretation of
-first-order formulas.\end{defn}
+first-order formulas.\end{definition}
 
-\begin{ex}The topos of sheaves over a topological space is always ouvert. More
+\begin{example}The topos of sheaves over a topological space is always ouvert. More
 generally, the topos of sheaves over a locale~$X$ is ouvert iff there exists a
 \emph{positivity predicate} on its frame of opens in the sense of~\cite{xxx}.
-With~\textsc{lem}, every locale is ouvert~\cite{xxx}.\end{ex}
+With~\textsc{lem}, every locale is ouvert~\cite{xxx}.\end{example}
 
-\begin{ex}The spectrum of a ring~$A$ is ouvert iff every element of~$A$ is
-nilpotent or not~\cite[Proposition~12.51]{blechschmidt:phd}.\end{ex}
+\begin{example}The spectrum of a ring~$A$ is ouvert iff every element of~$A$ is
+nilpotent or not~\cite[Proposition~12.51]{blechschmidt:phd}.\end{example}
 
-\begin{ex}A sufficient criterion for the classifying topos of a geometric
+\begin{example}A sufficient criterion for the classifying topos of a geometric
 theory~$\TT$ being ouvert is that the indexing sets of all disjunctions
 appearing on the right hand side of turnstiles, in a normal form presentation
-of~$\TT$, are inhabited~\cite{xxx}.\end{ex}
+of~$\TT$, are inhabited~\cite{xxx}.\end{example}
 
@@
 
-\begin{prop}\label{prop:gen-good}Let~$R$ be a relation on a set~$X$. The
+\begin{proposition}\label{prop:gen-good}Let~$R$ be a relation on a set~$X$. The
 generic infinite sequence over~$X$ is good if and only if~$R$ is almost-full in
-the inductive sense.\end{prop}
+the inductive sense.\end{proposition}
 
@@
 
-\begin{cor}\label{cor:modal-good}Let~$R$ be a relation on a set~$X$. Then~$R$
+\begin{corollary}\label{cor:modal-good}Let~$R$ be a relation on a set~$X$. Then~$R$
 is almost-full in the inductive sense if and only if everywhere, every infinite
-sequence is good.\end{cor}
+sequence is good.\end{corollary}
 
@@
 
-\begin{prop}Let~$({<})$ be a transitive relation on a set~$X$. The generic
+\begin{proposition}Let~$({<})$ be a transitive relation on a set~$X$. The generic
 infinite descending chain over~$X$ validates~$\bot$ (that is, the classifying
 topos of such chains is trivial) if and only if the relation is well-founded in
-the inductive sense.\end{prop}
+the inductive sense.\end{proposition}
 
@@
 
-\begin{cor}A transitive relation is well-founded in the inductive sense
+\begin{corollary}A transitive relation is well-founded in the inductive sense
 if and only if everywhere, it is not the case that there exists an infinite
-descending chain.\end{cor}
+descending chain.\end{corollary}
 
@@
 
-\begin{prop}\label{prop:af-wf}Let~$({\leq})$ be a transitive almost-full relation.
+\begin{proposition}\label{prop:af-wf}Let~$({\leq})$ be a transitive almost-full relation.
 Then~$({<})$, where~$x < y \equiv (x \leq y \wedge \neg(y \leq x))$,
-is well-founded.\end{prop}
+is well-founded.\end{proposition}
 
@@
 
-\begin{rem}For the proof of Proposition~\ref{prop:af-wf}, it is not relevant
+\begin{remark}For the proof of Proposition~\ref{prop:af-wf}, it is not relevant
 that pullback along geometric morphisms typically fails to preserve the
@@
 have~$f^*(\llbracket x > y \rrbracket) \wedge f^*(\llbracket x \leq y
-\rrbracket) \Rightarrow f^*(\llbracket\bot\rrbracket) = \bot$.\end{rem}
+\rrbracket) \Rightarrow f^*(\llbracket\bot\rrbracket) = \bot$.\end{remark}
 
-\begin{thm}[Dickson's lemma] If~$X$ and~$Y$ are almost-full,
-so is~$X \times Y$.\end{thm}
+\begin{theorem}[Dickson's lemma] If~$X$ and~$Y$ are almost-full,
+so is~$X \times Y$.\end{theorem}
 
@@
       \alpha(i(k_0)) &\leq \alpha(i(k_0+1)), &
-      \beta(i(k_0)) &\leq \beta(i(k_0+1)). \qedhere
+      \beta(i(k_0)) &\leq \beta(i(k_0+1)).
     \end{align*}
@@
 
+\bibliographystyle{splncs04}
+\bibliography{mapaper-modal-multiverse}
+
 \end{document}
@@
 Cite? https://arxiv.org/abs/2210.04838
+
+Noetherian set = every infinite sequence everywhere contains repeating terms
+https://arxiv.org/pdf/1604.01186.pdf
+
+Cite coquand-persson:groebner
+
+Trees generically absolute as in
+https://www.math.uci.edu/~twilson/slides/UCLA-LSC-2013.pdf#page=11
+
+cite https://math.andrej.com/2012/10/03/am-i-a-constructive-mathematician/

### 2023-04-11 algar-lecture-notes.tex
More work on the Brixen 2022a slides
Slides almost finished
to do

@@
 Comment on Zorn's lemma
+
+completion of ℕ
+
+Exercise: ∀∃ ⇒ Eventually
+
+History of "AC ⇒ LEM":
+1967 Bishop
+1975 Douglas Bridges
+1975 Diaconescu
+1975 Anna Michaelides Penk
+1978 Goodman and Myhill
+2021 Andrew Swan, https://ncatlab.org/nlab/show/well-ordering+theorem
+
+Cite: https://www2.mathematik.tu-darmstadt.de/~streicher/CLM/clm.pdf
+
+Cite: https://www.mscsnet.mu.edu/~wim/publica/8802hist_print.pdf
+
+https://www.math.ucla.edu/~joan/publicgothenburghandout.pdf
+Peculiar characteristics of intuitionistic logic:
+◮ (A → B) → ¬(A & ¬B) is provable, but the converse is not.
+◮ ∃x¬A(x) → ¬∀xA(x) is provable, but the converse is not.
+◮ ¬¬(A ∨ ¬A) is provable, but ¬¬∀x(A(x) ∨ ¬A(x)) is not.
+
+Cite:
+https://www.math.ucla.edu/~joan/newnotes.pdf
+
+Cite:
+https://arxiv.org/pdf/2102.01561.pdf

### 2023-05-01 paper-flabby-objects.tex
to do

@@
 Huishan WU, Revers mathematics and semisimple rings
+
+https://www.math.uni-duesseldorf.de/~schroeer/publications_pdf/FlasqueSheaves.pdf
+and reference 17 there
+
+Remark that our flabby hulls are not qcoh in general
+
+Bell, J. L. (1997). Zorn’s lemma and complete Boolean algebras in
+intuitionistic type theories. The Journal of Symbolic Logic, 62(04),
+1265–1279.
+
+Grothendieck's letter to Larry Breen in 1975: doing without injective
+resolutions

### 2023-05-01 slides-luminy2023.tex
Beginning of CIRM slides

--- 2022-11-05 slides-paris2022.tex
@@
 \usepackage{tikz}
-\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing,positioning}
+\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing,positioning,svg.path}
 \hypersetup{colorlinks=true}
@@
 
-\title{A modal logical multiverse for commutative algebra and combinatorics}
+\title{New modal operators for constructive mathematics}
 
 \author{Ingo Blechschmidt}
-\date{November 7th, 2022}
+\date{May 2nd, 2023}
 
@@
 \definecolor{mypurple}{RGB}{253,73,34}
+\definecolor{mypurpledark}{RGB}{100,0,150}
 \setbeamercolor{structure}{fg=mypurple}
@@
 
+% https://tex.stackexchange.com/questions/172336/drawing-roman-laurel-leaves-spqr-in-tikz
+\tikzset{
+  laurel-wreath/.pic = {
+    \fill svg{M14.4-24.6c-1.5-1.5-2.6-3.3-3.1-5.3l-.4-1.7c-.2-1.1-.2-4.1 .2-5.7 .2-.9 .3-1.3 .5-1.3l1.4 1.1 2.5 2.4c2.7 2.5 5.2 6 5.8 8 .2 .6-.5 .3-2.2-.9-1.6-1.3-3.3-2.6-5-3.8l.1 1.4c.2 1.4 .5 2.7 1.1 4.6s.8 2.5 .5 2.5l-1.4-1.3zm69.6 1.1 .3-1.2c.8-2.3 1.3-4.8 1.6-7.3l-1.5 1.1c-1.3 .9-2.6 1.9-3.7 3-1.6 1.1-2 1.3-2.1 1 .7-1.8 1.6-3.4 2.8-4.9 1.3-1.7 6.5-6.8 7-6.8 .2 0 .3 .2 .3 .5l.3 1.6c.3 2.2 .2 5.7-.5 7.4-.8 1.9-1.6 3.1-3 4.7-1.1 1.1-1.4 1.3-1.5 .9z};
+    \fill svg{M10-29.4c-.8-1.1-1.4-2.2-2-4.1l-.7-3.5c-.2-3 .2-4.4 1.4-8.3l.5-1.4c.2-1.3 .3-1.9 .6-1.9 .3-.2 .6 .3 .7 .8s.9 2.2 1.9 3.6c1.4 2.2 2.7 4.4 3.9 6.6l.9 2.7c0 .6 0 .6-.3 .6-.6 0-4.9-4.4-5.8-6l-.2-.6-.1 1.7-.3 2.8c-.3 2.7-.3 3.8 0 5.5 .6 2 .5 2.4-.5 1.5zm79.2 .3 .4-2.4c.2-1.3 .2-2.7-.1-4.9l-.3-2.8v-1.6l-.7 1c-.8 1.3-5 5.5-5.5 5.5s-.5-.3 .2-1.9c.5-1.7 1.4-3.3 3.3-6.5 2.4-3.6 2.7-3.9 2.8-4.7 .5-1.3 .5-1.4 .8-1.2 .3 0 .6 .8 .6 1.5l.7 2.4c.9 2.7 1.1 3.6 1.2 6 .2 3.1-.5 6-2 8.2-.8 1.3-1.3 1.7-1.4 1.5z};
+    \fill svg{M5-40c-.4-3.2-.1-6.5 .9-9.6 .5-1.1 1.6-2.8 2.2-3.4l1.3-1.6 2-2.7 .2 .6c.1 1.3 .4 2.6 .9 3.8l.3 1c.8 1.7 1.1 2.7 1.6 5.3 .6 2.5 .6 4.6 .2 4.6-.3 0-.9-.8-1-1.1l-.5-.8c-1.4-2-3-5.2-2.9-6.5-.9 2.7-2 5.4-3.5 7.9l-.3 .8-.3 .8c0 .5-.6 1.6-.8 1.6l-.3-.7zm89.2 .2-.2-.5-.3-.9-1.1-2.7-1.1-2.4c-.6-1.4-1.2-2.8-1.6-4.2l-.3 .9c-.3 1.3-1.6 3.9-3 6-1.3 2-1.6 2-1.5 0s1.1-6.3 2.2-9c.8-1.7 1.1-3.1 .9-4.1-.2-1.1 .5-.8 2.2 1.8 3.3 4.4 3.8 5.4 4.4 7.8 .6 2.4 .5 7.7-.3 7.8l-.3-.5z};
+    \fill svg{M13.9-50.1c-.5-1.9-.8-3.9-.9-5.8-.2-1.6-.1-3.3 .1-4.9-.3 .8-1.7 2.5-4.2 5.1l-3 4.9-.3 .1c-.3 0-.3-2.2 0-3.3 .8-3 1.4-4.6 2.5-6.1 .9-1.3 1.7-1.9 2.5-2.5 1.1-.6 2.7-1.9 3.5-2.7 .9-.9 1.9-1.4 2.2-1.4v1.1l-.3 6.6c0 6.8 .2 6.3-1 8.9-.5 1.1-.8 1.1-1.1 0zm70.8-.4c-.8-2.2-.8-2.5-.7-6.3-.1-2.7-.1-5.5-.2-8.2-.3-1.6-.3-1.9 .5-1.6l.6 .5c1.4 1.4 3 2.5 3.9 3.1 1.3 .9 1.9 1.6 2.7 2.6l.6 .7 .2 .4 .2 .3c.8 .9 2 4.9 2 6.9 .2 1.9-.2 1.9-.9 .5-.7-1.4-1.5-2.7-2.6-4-1.6-1.5-3-3.2-4.2-5 .4 3 .3 6-.5 9 0 .8-.5 2.2-.8 2.3-.2 0-.5-.3-.8-1.2z};
+    \fill svg{M16.4-58.5l.2-1.5 .3-3.7c.2-2.8 .3-3.5 1.1-5.4l.7-1.3-.5 .4-1 .7c-.5 .4-1.1 .8-1.5 1.3l-.5 .3-1.9 1.6c-2.2 1.6-2.7 2-3.9 3.6-.5 .8-1.1 1.3-1.3 1.3-.5 0 0-2.4 1.1-4.7 1.5-3.4 4.3-6 7.7-7.4l1.3-.4 1.9-.4 2-.5c1.4 0 1.4 0 1 1.1-.5 .8-.8 2-1.1 4.2-.3 2.3-1.1 4.5-2.2 6.5l-.4 .6c-.6 1.1-1.3 2.1-2 3.2-.5 .6-.8 .8-1 .5zm66.3-.2c-.8-.9-2.8-4.4-3.5-6.1-.6-1.3-.9-2.5-1.1-3.5-.2-2.1-.7-4.1-1.5-6 0-.3 0-.3 1.2-.3l2.1 .5 1.9 .4 1.2 .4 .6 .1 1 .6c3 1.4 5.7 4.6 6.8 8.5l.7 2.6c-.2 .6-.5 .5-1.4-.7-2.2-2.7-4.8-5-7.7-6.9l-1.7-1.3 .6 1.3c.3 .6 .6 1.2 .8 1.9l.3 2.5 .3 3.9c.3 2.4 .2 2.8-.6};
+    \fill svg{M21.6-66.1l.4-1.1 .9-3.2c.3-1.9 1.1-3.3 2.4-4.7l.4-.8-1.2 .2-2.2 .3c-2.7 .3-5.3 1.2-7.7 2.5-.6 .5-1.3 .6-1.3 .3 0-.5 .9-1.9 2-2.9 .8-.9 2-1.9 3.2-2.6l.9-.4 2.2-1c.3-.2 1.3-.3 3.2-.1 3 0 4.1 .2 6.3 .7l1.1 .4c.5 .2 .6 .6 .3 .6-.5 0-1.4 .9-1.9 1.7l-1.2 1.8c-1.7 2.8-2.2 3.5-4.6 5.9l-3 2.7-.2-.3zm53.9-2c-2.7-2.8-3.5-3.8-5.4-6.8-.9-1.6-1.4-2.4-1.9-2.5l-.8-.5c-.3 0-.2-.5 .4-.6l1.1-.4c1.9-.6 3-.8 5.6-.9l3.3 .2c2 .6 3.8 1.5 5.4 2.8 .3 0 1.9 1.6 2.5 2.4l.9 1.8c0 .3-.3 .2-1.9-.6-2.8-1.4-4.4-1.9-7.7-2.2l-2.2-.5c-.9-.2-.9-.2-.6 .2 .6 .5 1.7 2 2.1 2.8l.9 2.5c.3 1.5 .6 3 .9 4.6l-2.6-2.3z};
+    \fill svg{M34.1-78.7c-3.4-1.3-6.9-2.1-10.6-2.5-.9 0-1.4 0-2.3 .3-2 .5-2 0 0-1.3l2.8-1.2c1.4-.5 1.9-.5 3.8-.6 3.8-.2 6.1 .3 9.3 1.7l3.6 1.1 2.2 .3c1.3 0 1.7 0 2.7-.3 1.1-.3 2.8-1.1 2.8-1.3l-1.3-.9c-1.9-1.4-3.1-2.7-3.1-3.2l.8-.6c.9-.3 1.3-.2 2 .8 .5 .8 1.1 1.4 2.9 2.7 .2 .3 .3 .2 1.1-.3 .9-.8 2.4-2 2.6-2.7 .5-.6 .9-.8 1.8-.5l.8 .6c0 .5-1.4 1.7-3.2 3.2l-1.3 .9c0 .2 1.7 .9 2.9 1.3 .9 .3 1.4 .3 2.7 .3l2.2-.3c1.7-.4 3.4-1 5-1.7 2-.8 4.4-1.3 7.7-1.1 2 .2 2.5 .2 3.8 .6 .9 .3 2.2 .8 2.8 1.2 2 1.1 2 1.6 .2 1.3-1.6-.3-1.9-.3-4.4 0-2.4 .3-4.7 .8-7 1.6l-1.5 .6c-2.9 .3-5.9 .2-8.8-.3-1.7-.3-3.6-.9-6-2.1l-1.1-.4-1.3 .6c-4.5 2.2-9.6 3-14.6 2.2zm-6.3-9.1c};
+  }
+}
+
 \newcommand{\pointthis}[3]{%
@@
   \begin{beamercolorbox}[dp=1ex,center,wd=\paperwidth,ht=2.25ex]{title}%
+    \vskip0.5em%
     \bf\insertframetitle
   \end{beamercolorbox}%
-  \vskip-0.2em%
+
+  \vskip-0.77em\hspace*{-1.02em}%
+  \textcolor{mypurpledark}{\rule[0em]{\paperwidth}{2.4pt}}
+
+  \vskip-0.4em%
 }
@@
 \newcommand{\hil}[1]{{\usebeamercolor[fg]{item}{\textbf{#1}}}}
+\newcommand{\hill}[1]{{\usebeamercolor[fg]{item}{#1}}}
 \newcommand{\bad}[1]{\textcolor{red!90}{\textnormal{#1}}}
@@
 
+\newcommand{\triang}{\hil{$\blacktriangleright$}}
+\newcommand{\concat}{\mathbin{{+}\mspace{-8mu}{+}}}
+
+\newcommand{\astikznode}[2]{\tikz[baseline,remember picture]{\node[anchor=base,inner sep=0,outer sep=0.1em] (#1) {#2};}}
+\newcommand{\astikznodecircled}[3]{\tikz[baseline,remember picture]{\node[anchor=base,circle,draw=#2,thick,inner sep=0,outer sep=0.05em] (#1) {#3};}}
+
+\setbeamersize{text margin left=1.5em,text margin right=1.5em}
+
 \begin{document}
@@
 
-%\setbeamertemplate{headline}{\mynav{gray}{gray}{gray}}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{1.59cm}\includegraphics[width=\paperwidth]{forest-light}\end{minipage}}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[height=\paperheight]{night-sky}\end{minipage}}
 \begin{frame}[c]
   \centering
+  \color{white}
 
   \bigskip
-  %\includegraphics[height=0.32\textwidth]{olivia-lattices}
   \bigskip
@@
     \centering\normalsize\color{white}
-  A \hil{modal logical multiverse} for \\ commutative algebra and combinatorics
+    \hil{New modal operators for constructive mathematics}
   \end{block}
@@
 
-  \emph{Séminaire Général de Logique} \\
-  IMJ-PRG
+  \emph{Type Theory, Constructive Mathematics and Geometric Logic} \\
   \bigskip
 
-  Paris \\
-  November 7th, 2022
+  CIRM \\
+  May 2nd, 2023
   \bigskip
 
-  Ingo Blechschmidt \\
-  University of Augsburg
+  Ingo Blechschmidt
   \par
@@
 
+% \begin{document}
+
 \section{Motivation}
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1-lighter}\end{minipage}}
-\begin{frame}{Three proofs}
-  \begin{columns}[c]
-    \begin{column}{0.15\textwidth}
-      \scalebox{0.23}{\sieve{6}{6}}
-    \end{column}
-    \begin{column}{0.85\textwidth}
-      \begin{varblock}{\textwidth}{}
-        \justifying
-        \textbf{Thm.}
-        For every~$n$, there is a prime larger than~$n$.
-      \end{varblock}
-      \vspace*{-0.4em}
-      \emph{Proof.} Any prime factor of~$n! + 1$ will do. \qed
-    \end{column}
-  \end{columns}
+\begin{frame}{Infinite data}
+  \[ 7,
+    \quad \astikznodecircled{x0}{white}{4}\!,
+    \quad \only<1-2>{\astikznodecircled{t1}{white}{3}}\only<3->{\astikznodecircled{t1}{mypurple}{3}}\!,
+    \quad \only<1>{\ldots}\pause \astikznodecircled{x1}{white}{1}\!,
+    \quad \only<2>{\ldots}\pause \astikznodecircled{t2}{mypurple}{8}\!,
+    \quad \only<3>{\ldots} \visible<4->{\astikznodecircled{x2}{white}{2}\!,}
+    \quad \only<4->{\ldots} \]
+  {\centering\begin{tikzpicture}[remember picture,overlay]
+    \node[draw=mypurple, circle, thick, inner sep=0.1em] (t3) {\scriptsize$\leq$};
+    \path[draw=mypurple,thick]
+      (t1)
+      to [out=-90, in=180] (t3)
+      to [out=0, in=-90] (t2);
+  \end{tikzpicture}\par}
+  \bigskip
   \pause
 
-  \begin{columns}[c]
-    \begin{column}{0.15\textwidth}
-      \includegraphics[width=\textwidth,valign=t]{monad}
-    \end{column}
-    \begin{column}{0.85\textwidth}
-      \begin{varblock}{\textwidth}{}
-        \justifying
-        \textbf{Thm.}
-        Every infinite sequence~$\alpha : \NN \to \NN$ is \emph{good}
-        in that there are numbers~$i < j$ such that~$\alpha(i) \leq \alpha(j)$.
-      \end{varblock}
-      \vspace*{-0.4em}
-      \mbox{\emph{Proof.} There is a \bad{minimal value}~$\alpha(i)$.
-      Set~$j \defeq i{+}1$.\qed}
-    \end{column}
-  \end{columns}
+  \textbf{Thm.} Every infinite sequence~$\alpha : \NN \to \NN$ is \hil{good} in that
+  there exist numbers~$i < j$ such that~$\alpha(i) \leq \alpha(j)$.
   \pause
 
-  \begin{columns}[c]
-    \begin{column}{0.15\textwidth}
-      \includegraphics[width=\textwidth,valign=t]{calabi-yau}
-    \end{column}
-    \begin{column}{0.85\textwidth}
-      \begin{varblock}{\textwidth}{}
-        \justifying
-        \textbf{Thm.}
-        Let~$M$ be a surjective matrix with more rows than columns over a
-        ring~$A$. Then~$1 = 0$ in~$A$.
-      \end{varblock}
-      \vspace*{-0.4em}
-      \justifying\emph{Proof.}
-      \only<3>{\bad{Assume not.} Then there is~a \bad{maximal ideal} $\mmm$.
-      The matrix is surjective over~$A/\mmm$. Since~$A/\mmm$ is a field, this
-      is a contradiction to basic linear algebra. \\[1.3em]\ }\only<4>{Write~$M =
-      \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right)$. By surjectivity,
-      have~$u, v$ with
-      \vspace*{-0.6em}
-      \[
-        u \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}1\\0\end{smallmatrix}\right)
-        \quad\text{and}\quad
-        v \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}0\\1\end{smallmatrix}\right).
-      \]
-
-      \vspace*{-0.6em}
-      Hence
-      $
-        1 = (vy) (ux) = (uy) (vx) = 0
-      $.}
-      \qed
-    \end{column}
-  \end{columns}
-\end{frame}}
-% XXX: chart of ideals of ℤ
-
-% \begin{document}
-
-\begin{frame}{Three take-aways}
-  \bigskip
-  \begin{columns}[c]
-    \begin{column}{0.25\textwidth}
-      \includegraphics[width=\textwidth,valign=c]{turing-machine}
-    \end{column}
-    \begin{column}{0.75\textwidth}
-      Abstract proofs should be blueprints for concrete ones.
-    \end{column}
-  \end{columns}
+  \mbox{\emph{Proof.} \emph{(offensive?)} By~\badbox{\textsc{lem}}, there is a
+  minimum~$\alpha(i)$.
+  Set~$j \defeq i + 1$.\par}
   \pause
   \bigskip
+
+  \textbf{Def.} A preorder~$X$ is \hil{well} iff every infinite sequence~$\NN \to X$ is good.
+
+  \textbf{Examples.} $(\NN,{\leq}),\ \
+  \color{white}\only<7->{\color{red!90}}\underbrace{\color{black}X \times Y,\ \ X^*,\ \ \mathrm{Tree}(X)}_{\text{\visible<7->{\bad{only classically}}}}$.
+  \pause
+  \pause
   \bigskip
 
-  \begin{columns}[c]
-    \begin{column}{0.25\textwidth}
-      \includegraphics[width=\textwidth,valign=c]{3-adic-numbers}
-    \end{column}
-    \begin{column}{0.75\textwidth}
-      For every inhabited set~$X$,
-      there is a \hil{flavor~of~parametrized~mathematics} \\
-      which is home
-      to a gadget called \\
-      \hil{generic surjection}~$\NN \twoheadrightarrow X$.
-    \end{column}
-  \end{columns}
-  \pause
-  \bigskip
-  \bigskip
-
-  \begin{columns}[c]
-    \begin{column}{0.25\textwidth}
-      \includegraphics[width=\textwidth,valign=c]{multiverse}
-    \end{column}
-    \begin{column}{0.75\textwidth}
-      There is a \hil{rich modal multiverse} of \\
-      flavors of parametrized mathematics.
-    \end{column}
-  \end{columns}
-
-  \visible<3->{\begin{tikzpicture}[overlay]
-    \node[anchor=south east,inner sep=0] (image) at (10.8,0.3) {
-      \reflectbox{\includegraphics[width=1em]{butterfly}}
-    };
-  \end{tikzpicture}}
+  \centering
+  \textbf{Moral:} \\
+  \small{\it Don't quantify over points of spaces which might not have enough.}
 \end{frame}
 
-% \begin{document}
+\backupstart
 
-\section{The set-theoretic multiverse}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.59cm}\includegraphics[width=\paperwidth]{staircase}\end{minipage}}
-\begin{frame}{A brief timeline}
-  \begin{itemize}
-    \item[18xx] Cantor studies the \hil{continuum hypothesis}, the claim
-    that~$2^{\aleph_0} = \aleph_1$.
-    \item[19xx] Zermelo and Fraenkel offer the axioms of~\textsc{zfc}.
-    \item[1938] Gödel proves: If~\textsc{zfc} is consistent, then so
-    is~\textsc{zfc}+\textsc{ch}.
-    \item[1963] Cohen proves: If~\textsc{zfc} is consistent, then so
-    is~\textsc{zfc}+$\neg$\textsc{ch}.
-    \item[since 1963 and ongoing] Set theorists pursue additional axioms to
-    settle~\textsc{ch} (one way or another).  % XXX persue?
-    \item[201x] Hamkins offers his paper on the \hil{multiverse position} in
-    the philosophy of set theory, arguing that the program of pursing
-    additional axioms (while successful in many ways) is doomed to fail in
-    settling~\textsc{ch}.
-  \end{itemize}
-\end{frame}
-\begin{frame}{The set-theoretic multiverse}
-  \begin{tikzpicture}[remember picture,overlay]
-    \node[xshift=-1.5cm,yshift=-3.5cm] at (current page.north east)
-    {\includegraphics[width=2cm]{branching}};
-  \end{tikzpicture}
-  \vspace*{-1.5em}
-
-  \begin{block}{}
-    \justifying
-    \textbf{Def.} A \hil{model of set theory} is a (perhaps class-sized) structure $(M,{\in})$
-    satisfying axioms such as those of~\textsc{zfc}.
-  \end{block}
-  \vspace*{-0.6em}
-  {\small\emph{Examples.}\vspace*{-0.9em}
-  \begin{itemize}
-    \item $V$, the class of all sets \vspace*{-0.6em}
-    \item $L$, Gödel's constructible universe \vspace*{-0.6em}
-    \item $V[G]$, a forcing extension containing a generic filter~$G$ of \\ some
-    poset of forcing conditions \vspace*{-0.6em}
-    \item Henkin/term models from consistency of (extensions of)~\textsc{zfc}
-  \end{itemize}}
-  \pause
-
-  %{\centering\emph{Embracing all models of set theory:}\par}
-  \begin{block}{}
-    \justifying
-    \textbf{Def.} $\possible\varphi$ iff~$\varphi$ holds in \hil{some extension} of
-    the current universe. \\
-    \phantom{\textbf{Def.}} $\necessary\varphi$ iff~$\varphi$ holds in \hil{all extensions}
-    of the current universe.
-  \end{block}
-  \vspace*{-0.6em}
-  \begin{itemize}
-    \item $\necessary(\possible\textsc{CH} \wedge \possible\neg\text{CH})$,
-    the continuum hypothesis is a \hil{switch}
-    \item \ \\[-1.2em]\mbox{$\necessary\possible\necessary(\text{$X$ is countable})$,
-    existence of an enumeration is a \hil{button}}
-  \end{itemize}
-  \pause
-  \medskip
-
-  \textbf{NB.} In the multiverse of extensions of a given field~$K$,
-
-  \vspace*{-0.5em}
-  \begin{itemize}
-    \item ``there is a square root of~$-1$'' is a button and
-    \item ``every polynomial splits into linear factors'' is a switch.
-  \end{itemize}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[width=\paperwidth]{eventually-monad}\end{minipage}}
+\begin{frame}
+  \vspace*{\paperheight}\vspace*{-3.4em}\centering
+  \emph{Agda formalization available.}
 \end{frame}}
 
-% \begin{document}
-
-\section{Parametrized mathematics}
-
-\renewcommand{\insertframeextra}{a}
-\begin{frame}{Parametrized mathematics}
-  \medskip
-  {\centering\includegraphics[width=0.7\textwidth]{eigenvector}\par}
-  \small
-
-  \begin{changemargin}{-0.5em}{-1.5em}
-  \begin{enumerate}
-    \item[\infobox{local}]
-    \ \\[-1.2em]\mbox{``Every real symmetric matrix
-    does \only<4>{\hil{not not} }have an eigenvector.''
-    \good{\cmark}}
-    \medskip
-
-    \item[\infobox{global}] ``For every continuous family of symmetric
-    matrices, \\\only<4>{\hil{on a dense open} }eigenvectors can locally
-    be picked continuously.''
-    \only<1-2>{\includegraphics[height=0.7em]{question-mark}}%
-    \only<3>{\bad{\xmark}}%
-    \only<4>{\good{\cmark}}
-
-    \smallskip
-    \pause
-    \justifying
-    \scriptsize
-    ``Let~$X$ be a topological space and let~$A : X \to
-    M^\text{sym}_n(\RR)$ be a continuous map to the space of symmetric~$(n \times
-    n)$-matrices. Then there is an open covering~$\bigcup_{i \in I} U_i$
-    \only<1-3>{of~$X$}\only<4>{\hil{of a dense open subset}~$U \subseteq X$} such
-    that or all indices~$i \in I$, there is a continuous map~$v : U_i \to \RR^n$
-    such that for all~$x \in U_i$, the vector~$v(x)$ is an eigenvector
-    of~$A(x)$.''
-  \end{enumerate}
-  \end{changemargin}
-\end{frame}
-
-\addtocounter{framenumber}{-1}
-\renewcommand{\insertframeextra}{b}
-\begin{frame}{Parametrized mathematics}
-  \bigskip
-  \bigskip
-  {\centering
-    \includegraphics[height=9em]{damian-gvirtz-moduli-1}
-    \quad
-    \includegraphics[height=9em]{damian-gvirtz-moduli-2}
-    \par}
-  \small
-  \bigskip
-
-  \begin{changemargin}{-0.5em}{-1.5em}
-  \begin{enumerate}
-    \item[\infobox{local}]
-    \ \\[-1.2em]``Let~$M$ be a finitely generated module over a
-    field~$k$. \\ Then~$M$ is \only<4>{\hil{not not} }finite free.''
-    \good{\cmark}
-    \medskip
-
-    \item[\infobox{global}] ``Let~$M$ be a finitely generated module over a
-    ring~$A$. \\ Then~$M^\sim$ is finite locally
-    free\only<4>{ \hil{on a dense open}}.''
-    \only<1-2>{\includegraphics[height=0.7em]{question-mark}}%
-    \only<3>{\bad{\xmark}}%
-    \only<4>{\good{\cmark}}
-
-    \smallskip
-    \pause
-    \justifying
-    \scriptsize
-    ``Let~$M$ be a finitely generated module over an
-    arbitrary commutative ring~$A$. \only<1-3>{Then there is a partition~$1 = f_1 + \cdots +
-    f_n \in A$ of unity such that, for each index~$i$, the localized
-    module~$M[f_i^{-1}]$ is finite free over~$A[f_i^{-1}]$.}\only<4>{\hil{If~$f = 0$
-    is the only element of~$A$} such that~$M[f^{-1}]$ is finite free
-    over~$A[f^{-1}]$, \hil{then~$1 = 0$ in~$A$}.}''
-  \end{enumerate}
-  \end{changemargin}
-\end{frame}
-
-\addtocounter{framenumber}{-1}
-\renewcommand{\insertframeextra}{c}
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{-1em}\includegraphics[width=\paperwidth]{train-tracks-cropped}\end{minipage}}
-\begin{frame}{Parametrized mathematics}
-  \bigskip
-  \bigskip
-  \bigskip
-  \bigskip
-  \bigskip
-  \bigskip
-  \bigskip
-  \bigskip
-  \bigskip
-  \bigskip
-  \bigskip
-  \bigskip
-  \bigskip
-  \bigskip
-  \small
-
-  \begin{changemargin}{-0.5em}{-1.5em}
-  \begin{enumerate}
-    \item[\infobox{local}]
-    \ \\[-1.2em]``Let $R$ be a ring.
-    Let $n \geq 0$ be an integer.
-    We have
-    $$
-    H^q(\mathbf{P}^n, \mathcal{O}_{\mathbf{P}^n_R}(d)) = \textit{(omitted)}
-    $$
-    as $R$-modules.''
-    \good{\cmark}
-    \medskip
-
-    \item[\infobox{global}] ``Let $S$ be a scheme. Let $n \geq 1$.
-    Let $\mathcal{E}$ be a finite locally
-    free $\mathcal{O}_S$-module of constant rank $n + 1$.
-    For the structure morphism
-    $
-    \pi : \mathbf{P}(\mathcal{E}) \longrightarrow S,
-    $
-    we have
-    $$
-    R^q\pi_*(\mathcal{O}_{\mathbf{P}(\mathcal{E})}(d)) =
-    \textit{(omitted)}
-    $$
-    as sheaves of~$\mathcal{O}_S$-modules.'' \good{\cmark}
-  \end{enumerate}
-  \end{changemargin}
-\end{frame}}
-
-% \begin{document}
-
-\renewcommand{\insertframeextra}{}
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
-\begin{frame}{Sites and toposes}
-  For every \hil{site}~$\C$, the
-  \hil{sheaves} on~$\C$ form a \hil{topos}. \emph{Notable sites:}
-  \begin{enumerate}
-    \item The \hil{site of opens} of a topological space
-    \item The \hil{Zariski site} of a ring
-    \item The \hil{site of finite approximations} of surjections~$\NN
-    \twoheadrightarrow X$
-    \item The \hil{classifying site} of a geometric theory
-  \end{enumerate}
-  \pause
-
-  Every topos supports an \hil{internal language},
-  and this language is sound with respect to \hil{intuitionistic
-  logic}.
-  \pause
-
-  \begin{block}{}
-    \justifying
-    \textbf{Def.} A statement~$\varphi$ holds \ldots
-    \begin{itemize}
-      \item\small \hil{everywhere} ($\necessary\varphi$) iff
-      it holds in every (Grothendieck) topos (over the current base topos).
-      \item \hil{somewhere} ($\possible\varphi$) iff
-      it holds in some positive topos.
-      \item \hil{proximally} ($\xpossible\varphi$) iff
-      it holds in some positive ouvert topos.
-    \end{itemize}
-  \end{block}
-\end{frame}
-\begin{frame}{The multiverse of parametrized mathematics}
-  \begin{block}{}
-    \justifying
-    \textbf{Def.} A statement~$\varphi$ holds \ldots
-    \begin{itemize}
-      \item\small \hil{everywhere} ($\necessary\varphi$) iff
-      it holds in every (Grothendieck) topos (over the current base topos).
-      \item \hil{somewhere} ($\possible\varphi$) iff
-      it holds in some positive topos.
-      \item \hil{proximally} ($\xpossible\varphi$) iff
-      it holds in some positive ouvert topos.
-    \end{itemize}
-  \end{block}
-
-  \begin{changemargin}{-1.2em}{-1.2em}
-    \small
-    \begin{enumerate}
-      %\item If a function~$\NN \to \NN$ does \emph{everywhere} \emph{not~not} have
-      %a zero, then it actually has a zero.
-      \item \emph{Somewhere,} the law of
-      excluded middle holds. \emph{In fact, we even have:} \\
-      It is \emph{everywhere} the case that the law of
-      excluded middle holds \emph{somewhere}.
-      \smallskip
-      \item Assuming Zorn's lemma, it is
-      everywhere the case that it the axiom of choice holds
-      somewhere. [Barr]
-      \smallskip
-      \item If a geometric implication holds \emph{somewhere}, then it holds already here.
-      \smallskip
-      \item If a first-order statement holds \emph{proximally}, then it holds already here.
-      \smallskip
-      \item For every set~$X$, \emph{proximally} there is a
-      surjection~$\NN \twoheadrightarrow X$.
-      \smallskip
-      \item Every ring \emph{proximally} has a maximal ideal.
-    \end{enumerate}
-  \end{changemargin}
-\end{frame}}
+\backupend
 
 \end{document}
-
-A modal logical multiverse for commutative algebra and combinatorics
-
-In the spirit of the set-theoretic multiverse philosophy put forward by
-Joel David Hamkins, we explore a related modal multiverse populated by
-Kripke models and more general worlds. In this
-multiverse—as the talk will explain—the law of excluded middle can be
-switched on and off like a light bulb and countability is a button (for
-every set X of every world, there is a larger world containing a
-surjection ℕ → X). Our interest in this multiverse is because of
-concrete applications in commutative algebra and combinatorics,
-including the endeavor of extracting algorithms from proofs utilizing
-transfinite techniques. The talk will be framed by several examples of
-this kind.
-
-=== Motivation
-
-Infinitude of primes
-- algorithm visible from the proof
-
-Two motivating questions:
-- Dickson's lemma
-- Surjectivity of matrices
-How to extract algorithms here?
-Spoil elementary proof of surjective case
-
-=== Set-theoretic multiverse
-
-Quick sketch of history including dream solution to CH
-
-Multiverse position: acknowledge that notion of set is underdetermined by ZFC
-axioms, embrace multiverse of models
-
-Switches and buttons
-
-Analogy with field extensions
-
-Menagerie of properties of the multiverse
-
-=== Relaxation to topos-theoretic case
-
-"Parametric mathematics" (esp. visible with toposes over spaces)
-
-Modal operators
-
-Examples of modal statements

### 2023-05-01 slides-luminy2023.tex
Slides almost finished

@@
-\documentclass[12pt,utf8,notheorems,compress,t]{beamer}
+\documentclass[12pt,utf8,notheorems,compress,t,aspectratio=169]{beamer}
+%\documentclass[12pt,utf8,notheorems,compress,t]{beamer}
 \usepackage{etex}
@@
 \usepackage{amssymb}
+\usepackage{manfnt}
 \usepackage{array}
@@
 \usepackage{tikz}
-\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing,positioning,svg.path}
+\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing,positioning,svg.path,fadings}
 \hypersetup{colorlinks=true}
@@
 
-  \vskip-0.77em\hspace*{-1.02em}%
-  \textcolor{mypurpledark}{\rule[0em]{\paperwidth}{2.4pt}}
+  \vskip-0.77em\hspace*{-2em}%
+  \textcolor{mypurpledark}{\rule[0em]{1.1\paperwidth}{2.4pt}}
 
@@
 \newcommand{\astikznodecircled}[3]{\tikz[baseline,remember picture]{\node[anchor=base,circle,draw=#2,thick,inner sep=0,outer sep=0.05em] (#1) {#3};}}
+\newcommand{\astikznodetransparentlycircled}[2]{\tikz[baseline,remember picture]{\node[anchor=base,circle,opacity=0,draw=white,text opacity=1,thick,inner sep=0,outer sep=0.05em] (#1) {#2};}}
 
-\setbeamersize{text margin left=1.5em,text margin right=1.5em}
+\setbeamersize{text margin left=2.0em,text margin right=2.0em}
 
@@
   \setbeamercolor{block body}{bg=black!100}
+  \begin{minipage}{0.8\textwidth}
     \begin{block}{}
       \centering\normalsize\color{white}
-    \hil{New modal operators for constructive mathematics}
+      \hil{New modal operators for~constructive~mathematics}
     \end{block}
+  \end{minipage}
 
@@
 
+\begin{tikzfadingfrompicture}[name=fade bottom]
+  \shade[top color=transparent!0,
+    bottom color=transparent!50] (0,0) rectangle (2,2);
+    \end{tikzfadingfrompicture}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[height=\paperheight]{sea-of-clouds-2}\end{minipage}}
 \begin{frame}{Infinite data}
-  \[ 7,
-    \quad \astikznodecircled{x0}{white}{4}\!,
-    \quad \only<1-2>{\astikznodecircled{t1}{white}{3}}\only<3->{\astikznodecircled{t1}{mypurple}{3}}\!,
-    \quad \only<1>{\ldots}\pause \astikznodecircled{x1}{white}{1}\!,
+  \vspace*{-1em}
+  \[ \astikznodetransparentlycircled{xm}{7}\!,
+    \quad \astikznodetransparentlycircled{x0}{4}\!,
+    \quad \only<1-2>{\astikznodetransparentlycircled{t1}{3}}\only<3->{\astikznodecircled{t1}{mypurple}{3}}\!,
+    \quad \only<1>{\ldots}\pause \astikznodetransparentlycircled{x1}{1}\!,
     \quad \only<2>{\ldots}\pause \astikznodecircled{t2}{mypurple}{8}\!,
-    \quad \only<3>{\ldots} \visible<4->{\astikznodecircled{x2}{white}{2}\!,}
+    \quad \only<3>{\ldots} \visible<4->{\astikznodetransparentlycircled{x2}{2}\!,}
     \quad \only<4->{\ldots} \]
@@
   \end{tikzpicture}\par}
-  \bigskip
+  \medskip
   \pause
 
-  \textbf{Thm.} Every infinite sequence~$\alpha : \NN \to \NN$ is \hil{good} in that
-  there exist numbers~$i < j$ such that~$\alpha(i) \leq \alpha(j)$.
+  \textbf{Thm.} Every sequence~$\alpha : \NN \to \NN$ is \hil{good} in that
+  there exist~$i < j$ with~$\alpha(i) \leq \alpha(j)$.
   \pause
 
-  \mbox{\emph{Proof.} \emph{(offensive?)} By~\badbox{\textsc{lem}}, there is a
+  \emph{Proof.} \emph{(offensive?)} By~\badbox{\textsc{lem}}, there is a
   minimum~$\alpha(i)$.
-  Set~$j \defeq i + 1$.\par}
+  Set~$j \defeq i + 1$. \qed\par
   \pause
-  \bigskip
+  \medskip
 
-  \textbf{Def.} A preorder~$X$ is \hil{well} iff every infinite sequence~$\NN \to X$ is good.
+  \textbf{Def.} A preorder~$X$ is \hil{well\only<9->{$^\star$}} iff every sequence~$\NN \to X$ is good.
 
   \textbf{Examples.} $(\NN,{\leq}),\ \
-  \color{white}\only<7->{\color{red!90}}\underbrace{\color{black}X \times Y,\ \ X^*,\ \ \mathrm{Tree}(X)}_{\text{\visible<7->{\bad{only classically}}}}$.
+  \color{white}\only<7->{\color{red!90}}\astikznode{onlyclass}{$\underbrace{\color{black}X \times Y,\ \ X^*,\ \ \mathrm{Tree}(X)}_{\text{\visible<7->{\bad{only classically}}}}$}$.
   \pause
   \pause
-  \bigskip
+  \medskip
 
-  \centering
-  \textbf{Moral:} \\
-  \small{\it Don't quantify over points of spaces which might not have enough.}
+  \begin{tikzpicture}[remember picture,overlay]
+    \node[thick, fill=black, rectangle, inner sep=0.3em, right=2em of onlyclass] (moral) {
+      \begin{minipage}{6cm}
+        \begin{columns}
+          \begin{column}{0.15\textwidth}
+            \hspace*{1.2em}\color{white}\dbend
+          \end{column}
+          \begin{column}{0.95\textwidth}
+            \color{white}\footnotesize
+            \it Don't quantify over points of spaces which might not have enough.
+          \end{column}
+        \end{columns}
+      \end{minipage}
+    };
+    \path[draw=red!90,thick,-stealth]
+      (moral) to
+      [out=180, in=-00] (onlyclass.east);
+  \end{tikzpicture}
+  \pause
+
+  \textbf{Def.} For a predicate~$P$ on finite lists over a set~$X$, inductively
+  define:
+  \[
+    \infer{P \,|\, \sigma}{P(\sigma)}
+    \qquad
+    \infer{P \,|\, \sigma}{\forall(x \in X)\_\ P \,|\, \sigma x}
+  \]
+
+  \textbf{Def.} A preorder is \hil{well} iff~$\mathsf{Good} \,|\, [\,]$, where
+  $\mathsf{Good}(\sigma) \defeqv (\exists(i < j)\_ \sigma[i] \leq \sigma[j])$.
+\end{frame}}
+
+\begin{frame}{Computational content from classical proofs}
+  \textbf{Def.} A transitive relation~$({<})$ on a set~$X$ is \ldots
+  \begin{enumerate}
+    \item \hil{well-founded$^\star$} iff
+    there is no \bad{infinite chain} $x_0 > x_1 > \cdots$,
+    \item \hil{well-founded} iff for every~$x \in X$,
+    $\mathsf{Acc}(x)$,
+  \end{enumerate}
+  where~$\mathsf{Acc}$ is inductively defined by:
+  \[
+    \infer{\mathsf{Acc}(x)}{\forall(y < x)\_\ \mathsf{Acc}(y)}
+  \]
+  \vspace*{-1em}
+  \pause
+
+  \textbf{Prop.} Let~$(X,{\leq})$ be preorder. Let~$x < y$ mean~$x \leq y \wedge \neg(y \leq x)$. \\
+  Then: If~$X$ is well$^\star$, then~$({<})$ is well-founded$^\star$.
+
+  \emph{Proof.} An infinite strictly descending chain would also be good. \qed
+
+  \emph{Can we extract a constructive proof that well implies well-founded?}
+\end{frame}
+
+\begin{frame}{Maximal ideals}
+  \only<1-7>{\textbf{Thm.}
+  Let~$M$ be a surjective matrix with more rows than columns over a
+  ring~$A$. Then~$1 = 0$ in~$A$.
+
+  \visible<2->{\emph{Proof.} \bad{Assume not.}}
+  \visible<3->{Then there is~a \bad{maximal ideal} $\mmm$.}
+  \visible<5->{The matrix is surjective over~$A/\mmm$.}
+  \visible<6->{Since~$A/\mmm$ is a field, this is a contradiction to basic linear algebra.\qed}}
+
+  \only<4-7>{\bigskip\par\centering\scalebox{0.9}{\begin{tikzpicture}
+    \node (0) at (0,1) {$(0) = \{0\}$};
+    \node (1) at (0,5) {$(1) = \ZZ$};
+    \node (2) at (-2,4) {$(2)$};
+    \node [right of=2] (3) {$(3)$};
+    \node [below of=2] (4) {$(4)$};
+    \node [below of=2, xshift=0.7cm] (6) {$(6)$};
+    \node [right of=3] (5) {$(5)$};
+    \node [right of=5] (7) {$(7)$};
+    \node [right of=7] (7d) {$\ldots$\phantom{(}};
+    \node [right of=7d, xshift=3cm, yshift=-2cm] (max)
+    {\vbox{\small\it maximal among the proper ideals \\ \medskip $\neg(1 \in
+    \mmm)$ \\ \medskip $x \in \mmm \quad\vee\quad 1 \in \mmm + (x)$}};
+    \node [below of=4] (8) {$(8)$};
+    \node [right of=8, xshift=3cm] (8d) {$\ldots$};
+    \draw (0) -- (8);
+    \draw (0) -- (8d);
+    \draw (0) -- (6);
+    \draw (2) -- (1);
+    \draw (3) -- (1);
+    \draw (5) -- (1);
+    \draw (7) -- (1);
+    \draw (7d) -- (1);
+    \draw (4) -- (2);
+    \draw (8) -- (4);
+    \draw (6) -- (2);
+    \draw (6) -- (3);
+    \draw [mypurple!30, thick, shorten <=-2pt, shorten >=-2pt, ->] (max) to [out=120, in=-30] (7d);
+    \begin{pgfonlayer}{background}
+      \draw[decorate, very thick, draw=mypurple!30]
+        ($(2.south west) + (8pt, 0)$) arc(270:180:8pt) --
+        ($(2.north west) + (0, -8pt)$) arc(180:90:8pt) --
+        ($(7d.north east) + (-8pt, 0)$) arc(90:0:8pt) --
+        ($(7d.south east) + (0, 8pt)$) arc(0:-90:8pt) --
+        cycle;
+    \end{pgfonlayer}
+  \end{tikzpicture}}\par}
+
+  \pause
+  \pause
+  \pause
+  \pause
+  \pause
+  \pause
+  \medskip
+  \raggedright
+
+  Let~$A$ be a ring. \emph{Does there exist a maximal ideal~$\mmm \subseteq A$?}
+  \pause
+  \begin{enumerate}
+    \item \good{Yes}, if \bad{Zorn's lemma} is available.
+    \medskip
+    \pause
+
+    \item \good{Yes}, if~$A$ is countable and membership of finitely generated ideals is decidable:
+    {\footnotesize
+    \begin{align*}
+      \mmm_0 &= \{ 0 \}, &
+      \mmm_{n+1} &= \begin{cases}
+        \mmm_n + (x_n), & \text{if $1 \not\in \mmm_n + (x_n)$}, {\qquad\quad\quad\!\!}\\
+        \mmm_n, & \text{else.}
+      \end{cases}
+    \end{align*}}
+    \pause
+
+    \item \good{Yes}, if~$A$ is countable (irrespective of membership decidability):
+    {\footnotesize\begin{align*}
+      \mmm_0 &= \{ 0 \}, &
+      \mmm_{n+1} &= \mmm_n + (\underbrace{\{ x \in A \,|\, x = x_n \wedge
+        1 \not\in \mmm_n + (x_n) \}}_{\text{a certain subsingleton set}})
+    \end{align*}}
+    \vspace*{-0.5em}
+    \pause
+
+    \item In the general case: \bad{No}, but we \good{do have} the first-order
+    consequences of the existence of a maximal ideal.
+  \end{enumerate}
 \end{frame}

### 2023-05-02 slides-luminy2023.tex
Slides finished

@@
 \newcommand{\Ring}{\mathrm{Ring}}
+\newcommand{\Cov}{\mathrm{Cov}}
 \newcommand{\LocRing}{\mathrm{LocRing}}
@@
 
-\setbeamersize{text margin left=2.0em,text margin right=2.0em}
+\setbeamersize{text margin left=1.60em,text margin right=1.60em}
 
@@
 
-  Ingo Blechschmidt
+  Ingo Blechschmidt \\
+  j.w.w. Alexander Oldenziel
   \par
@@
 \setbeamercolor{structure}{fg=mypurple}
-
-% \begin{document}
 
@@
   \[
-    \infer{P \,|\, \sigma}{P(\sigma)}
+    \infer{P \,|\, \sigma}{P\sigma}
     \qquad
@@
   \textbf{Def.} A preorder is \hil{well} iff~$\mathsf{Good} \,|\, [\,]$, where
-  $\mathsf{Good}(\sigma) \defeqv (\exists(i < j)\_ \sigma[i] \leq \sigma[j])$.
+  $\mathsf{Good}\,\sigma \defeqv (\exists(i < j)\_ \sigma[i] \leq \sigma[j])$.
 \end{frame}}
@@
 
-  \textbf{Prop.} Let~$(X,{\leq})$ be preorder. Let~$x < y$ mean~$x \leq y \wedge \neg(y \leq x)$. \\
+  \textbf{Prop.} Let~$(X,{\leq})$ be preorder. Let~``$x < y$'' mean~$x \leq y \wedge \neg(y \leq x)$. \\
   Then: If~$X$ is well$^\star$, then~$({<})$ is well-founded$^\star$.
@@
   \emph{Proof.} An infinite strictly descending chain would also be good. \qed
+  \pause
 
@@
 
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
 \begin{frame}{Maximal ideals}
@@
 
-  \only<4-7>{\bigskip\par\centering\scalebox{0.9}{\begin{tikzpicture}
+  \only<4-7>{\bigskip\par\centering\scalebox{0.9}{\centering\begin{tikzpicture}
     \node (0) at (0,1) {$(0) = \{0\}$};
@@
     \end{pgfonlayer}
-  \end{tikzpicture}}\par}
+  \end{tikzpicture}\par}\par}
 
@@
     \item \good{Yes}, if \bad{Zorn's lemma} is available.
-    \medskip
+    \bigskip
     \pause
@@
     \end{align*}}
-    \vspace*{-0.5em}
+    \vspace*{4.5em}
+    \visible<11>{{
+      \centering
+      \raisebox{0pt}[0pt][0pt]{
+        \hspace*{1.5em}
+        \begin{tikzpicture}
+          \node (inner) at (17.3mm,-20mm) {\textit{``a bad joke''}};
+          \path (0,0) pic{laurel-wreath};
+        \end{tikzpicture}
+        \hspace*{-3em}
+        \begin{tikzpicture}
+          \node (inner) at (17.5mm,-18mm) {\vbox{\small\centering\textit{``non- \\informative''}}};
+          \path (0,0) pic{laurel-wreath};
+        \end{tikzpicture}
+      }
+      \par
+    }}
+    \pause
+    \pause
+    \vspace*{-4.5em}
+
+    \item In the general case: \bad{No}, but the \emph{first-order
+    consequences} of the existence of a maximal ideal \good{do hold}.
+  \end{enumerate}
+\end{frame}}
+
+\begin{frame}{Questions}
+  \begin{enumerate}
+    \item \!Why was the inductive revolution so powerful?
+    \bigskip
+
+    \item \!Why do proofs using Zorn's maximal ideals work so well in constructive algebra?
+    \bigskip
+
+    \item \!Why are elements of~$\bigcap_\ppp \ppp$ not necessarily nilpotent?
+    \bigskip
+
+    \item How can we extract computational content from classical proofs?
+  \end{enumerate}
+\end{frame}
+
+\renewcommand{\insertframeextra}{a}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.23cm}\includegraphics[width=\paperwidth]{topos-horses-lighter-lighter}\end{minipage}}
+\begin{frame}{Constructive forcing (= Grothendieck toposes)}
+  \justifying
+  Let~$L$ be a \hil{forcing notion}, a preorder equipped
+  with a \hil{covering system}.\only<1-2>{\footnote{\scriptsize A covering system consists of a
+  set~$\Cov(\sigma) \subseteq P({\downarrow}\sigma)$ of \emph{coverings} for
+  each element~$\sigma \in L$ subject only to the following simulation
+  condition: If~$\tau \preccurlyeq \sigma$ and~$R \in \Cov(\sigma)$, there
+  should be a covering~$S \in \Cov(\tau)$ such that~$S \subseteq
+  {\downarrow}R$.}}\only<3->{\textsuperscript{\phantom{1}}}
+  \hil{Filters}~$F \subseteq L$ are subsets which are upward-closed,
+  downward-directed and split the covering system.\only<1-2>{\footnote{\scriptsize
+  A subset~$F$ splits the covering system iff for every~$\sigma \in L$ and~$R \in \Cov(\sigma)$,
+  if~$\sigma \in F$, then~$\tau \in F$ for some~$\tau \in R$.}}\only<3>{\textsuperscript{\phantom{2}}}
   \pause
 
-    \item In the general case: \bad{No}, but we \good{do have} the first-order
-    consequences of the existence of a maximal ideal.
+  {\centering\footnotesize\begin{tabular}{lll}
+    \toprule
+    preorder~$L$ & coverings of an element~$\sigma \in L$ & filters of~$L$ \\
+    \midrule
+    $X^*$ & $\{ \sigma x \,|\, x \in X \}$ & maps~$\NN \to X$ \\
+    $X^*$ & $\{ \sigma x \,|\, x \in X \}$,\ \
+    $\{ \sigma\tau \,|\, \tau \in X^*, a \in \sigma\tau \}$ for each~$a \in X$ & surjections~$\NN \twoheadrightarrow X$ \\
+    f.g. ideals & --- & ideals \\
+    f.g. ideals & $\{ \sigma+(a), \sigma+(b) \}$ for each~$ab \in \sigma$,\ \
+    $\{\}$ if~$1 \in \sigma$ & prime ideals \\
+    opens & $\mathcal{U}$ such that~$\sigma = \bigcup \mathcal{U}$ & points \\
+    $\{\star\}$ & $\{ \star \,|\, \varphi \} \cup \{ \star \,|\, \neg\varphi \}$ &
+    witnesses of~\textsc{lem}
+    \\
+    \bottomrule
+  \end{tabular}\par}
+  \bigskip
+
+  \only<3>{\textbf{Def.} Given a monotone predicate~$P$
+  on~$L$, inductively define:
+  \[
+    \infer{P \,|\, \sigma}{P\sigma}
+    \qquad
+    \infer[{\small(R \in \Cov(\sigma))}]{P \,|\, \sigma}{\forall(\tau \in R)\_\ P \,|\, \tau}
+  \]
+  We use quantifier-like notation: ``$\nabla\sigma\_ P\sigma$'' means~$P \,|\, \sigma$.}
+\end{frame}}
+
+\addtocounter{framenumber}{-1}
+\renewcommand{\insertframeextra}{b}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.23cm}\includegraphics[width=\paperwidth]{topos-horses-lighter-lighter}\end{minipage}}
+\begin{frame}{Constructive forcing (= Grothendieck toposes)}
+  A forcing notion is a template for a \hil{forcing extension}~$V^\nabla$
+  of the base universe~$V$:
+
+  \begin{center}\begin{tikzpicture}[remember picture,overlay]
+    \node[thick, draw=mypurpledark, rectangle, inner sep=0.3em] (kj) {
+      \begin{minipage}{0.6\textwidth}
+        \centering
+        \emph{When we say that a statement \hil{holds in~$V^\nabla$},} \\
+        we mean that its~$\nabla$-translation is true.\par
+        \vspace*{0.3em}
+      \end{minipage}
+    };
+  \end{tikzpicture}\end{center}
+  \bigskip
+  \pause
+
+  \textbf{Examples.}
+  \begin{enumerate}
+  \justifying
+  \item For~$L = \{\star\}$, ``every number is zero or a successor'' holds
+  in~$V^\nabla$ because
+
+  \vspace*{-1em}
+  \[ \forall(n \in \NN)\_ \neg\neg\bigl(\neg\neg(n = 0) \vee
+    \neg\neg(\exists(m \in \NN)\_ \neg\neg(n = S(m)))\bigr). \]
+
+  \pause
+  \item For~$L$ associated to a space, ``every symmetric matrix has an
+  eigenvector'' holds in~$V^\nabla$ iff for every continuous family of
+  symmetric matrices, locally there is a continuous eigenvector-picking
+  function.
+  \pause
+  \medskip
+
+  \item For any~$L$, ``there is a filter of~$L$'' holds in~$V^\nabla$,
+  witnessed by the \hil{generic filter} of~$L$.
+  \pause
+  \medskip
+
+  \item Let~$X$ be a preorder. Then:
+
+  \hil{$\blacktriangleright$} $X$ is well iff the generic sequence~$\NN \to X$
+  is good.
+
+  \hil{$\blacktriangleright$} $({<})$ is well-founded iff the generic strictly
+  descending chain validates~$\bot$.
+  \end{enumerate}
+\end{frame}}
+
+\renewcommand{\insertframeextra}{}
+\begin{frame}{The modal multiverse}
+  \only<1-2>{In general, ``$\varphi$ holds in~$V^\nabla$'' and~``$\varphi$ holds in~$V$''
+  are \emph{not} equivalent.
+
+  \hil{$\blacktriangleright$}
+  For \hil{positive} extensions, they are equivalent for coherent implications.\\
+  %\footnote{bounded first-order formulas of the
+  %form~$\forall\ldots\forall\_(\ldots\Rightarrow\ldots)$,
+  %where antecedent and succedent do not contain~``$\forall$''
+  %and~``${\Rightarrow}$''}.
+  \phantom{\hil{$\blacktriangleright$}} {\footnotesize -- e.g. the ``Barr cover''.\par}
+
+  \hil{$\blacktriangleright$}
+  For \hil{positive open} extensions, they are equivalent for bounded first-order formulas. \\
+  \phantom{\hil{$\blacktriangleright$}} {\footnotesize -- e.g. $V^\nabla$ containing
+  the generic surjection~$\NN \twoheadrightarrow X$, if~$X$ is inhabited.\par}
+  \bigskip}
+  \pause
+
+  \textbf{Def.} A statement~$\varphi$ holds \ldots
+  \begin{enumerate}
+    \small
+    \item[$\blacktriangleright$] \hil{everywhere} ($\necessary\varphi$) iff
+    it holds in every extension.
+    \item[$\blacktriangleright$] \hil{somewhere} ($\possible\varphi$) iff
+    it holds in some positive extension.
+    \item[$\blacktriangleright$] \hil{proximally} ($\xpossible\varphi$) iff
+    it holds in some positive overt extension.
+  \end{enumerate}
+  \bigskip
+  \pause
+
+  \small
+  \begin{multicols}{2}
+    For every set~$X$, \emph{proximally} \\
+    there is an enumeration~$\NN \twoheadrightarrow X$.
+
+    A preorder is well iff \emph{everywhere}, \\
+    every sequence is good.
+
+    A ring element is nilpotent iff \\
+    all prime ideals \emph{everywhere} contain it.
+
+    For every ring, \emph{proximally} \\ there is a maximal ideal.
+
+    A relation is well-founded iff \emph{everywhere}, there is no descending chain.
+
+    \emph{Somewhere}, \\ the law of excluded middle holds.
+  \end{multicols}
+\end{frame}
+
+\begin{frame}{The multiverse perspective}
+  \begin{enumerate}
+    \item \!\emph{Why was the inductive revolution so powerful?}
+
+    Because the inductive conditions
+    are equivalent to truth in \emph{all} forcing extensions.
+    \bigskip
+
+    \item \!\emph{Why do proofs using Zorn's maximal ideals work so well in constructive algebra?}
+
+    Because every ring proximally has a maximal ideal.
+    \bigskip
+
+    \item \!\emph{Why are elements of~$\bigcap_\ppp \ppp$ not necessarily nilpotent?}
+
+    Because we forgot the prime ideals in forcing extensions.
+    \bigskip
+
+    \item \emph{How can we extract computational content from classical proofs?}
+
+    By traveling the multiverse (upwards, keeping ties to the base), exploiting
+    that \\
+    -- \textsc{lem} holds \emph{somewhere} and \\
+    -- \textsc{dc} holds \emph{proximally}.
   \end{enumerate}
@@
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[width=\paperwidth]{eventually-monad}\end{minipage}}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[width=\paperwidth]{eventually-monad-cropped}\end{minipage}}
 \begin{frame}
-  \vspace*{\paperheight}\vspace*{-3.4em}\centering
+  \vspace*{\paperheight}\vspace*{-2.05em}\centering
   \emph{Agda formalization available.}

### 2023-05-02 slides-luminy2023.tex
Final touches

@@
 
-\addtocounter{framenumber}{-1}
+\addtocounter{framenumber}{-2}
+
+{
+\definecolor{mypurple}{RGB}{150,0,255}
+\setbeamercolor{structure}{fg=mypurple}
+\begin{frame}{}
+  \begin{center}\includegraphics[width=0.2\textwidth]{eigenvector}\end{center}
+
+  Let a continuous family of symmetric matrices be given:
+  \[
+  \begin{pmatrix}a_{11}(t)&\cdots&a_{1n}(t)\\\vdots&&\vdots\\a_{n1}(t)&\cdots&a_{nn}(t)\end{pmatrix}
+  \]
+
+  There for every parameter value~$t$, classically there is
+
+  \hil{$\blacktriangleright$} a full list of eigenvalues~$\lambda_1(t),\ldots,\lambda_n(t)$ and \\
+  \hil{$\blacktriangleright$} an eigenvector basis~$(v_1(t),\ldots,v_n(t))$.
+  \bigskip
+
+  \begin{columns}[c]
+    \begin{column}{0.01\textwidth}
+      \includegraphics[height=2.4em]{question-mark}
+    \end{column}
+    \begin{column}{0.9\textwidth}
+      \mbox{Can locally the functions~$\lambda_i$ be chosen to be continuous?
+      \only<2->{\hil{Yes.}}} \\
+      How about the~$v_i$? \only<2->{\hil{No.}}
+    \end{column}
+  \end{columns}
+\end{frame}
+}
 
@@
           \begin{column}{0.15\textwidth}
-            \hspace*{1.2em}\color{white}\dbend
+            \hspace*{1.0em}\color{white}\dbend
           \end{column}
@@
 
-  \emph{Can we extract a constructive proof that well implies well-founded?}
+  Can we extract a constructive proof that well preorders are well-founded?
 \end{frame}
@@
     \end{align*}}
-    \vspace*{4.5em}
+    \vspace*{3.5em}
     \visible<11>{{
@@
     \pause
-    \vspace*{-4.5em}
+    \vspace*{-3.5em}
 
@@
   \begin{enumerate}
-    \item \!Why was the inductive revolution so powerful?
+    \item \!Why has the inductive revolution been so powerful?
     \bigskip
@@
   \hil{$\blacktriangleright$}
-  For \hil{positive open} extensions, they are equivalent for bounded first-order formulas. \\
+  For \hil{positive overt} extensions, they are equivalent for bounded first-order formulas. \\
   \phantom{\hil{$\blacktriangleright$}} {\footnotesize -- e.g. $V^\nabla$ containing
@@
   \bigskip
+
+  \visible<2>{\begin{tikzpicture}[overlay]
+    \node[anchor=south east,inner sep=0] (image) at (14.8,-0.8) {
+      \includegraphics[width=0.2\textwidth]{branching}
+    };
+  \end{tikzpicture}\vspace*{-1.6em}}
+  \visible<3>{\begin{tikzpicture}[overlay]
+    \node[anchor=south east,inner sep=0] (image) at (14.8,0.9) {
+      \includegraphics[width=0.2\textwidth]{multiverse}
+    };
+  \end{tikzpicture}\vspace*{-0.3em}}
+
+  \only<2>{\scriptsize Foreshadowed by:
+  \begin{itemize}
+    \scriptsize
+    \item[1984\phantom{s}] André Joyal, Miles Tierney. \emph{An extension of the Galois theory of
+    Grothendieck.}
+    \item[1987\phantom{s}] Andreas Blass. \emph{Well-ordering and induction in intuitionistic
+    logic and topoi.}
+    \item[2010s] Milly Maietti, Steve Vickers. Ongoing work on arithmetic
+    universes.
+    \item[2011\phantom{s}] Joel David Hamkins. \emph{The set-theoretic multiverse.}
+    \item[2013\phantom{s}] Shawn Henry. \emph{Classifying topoi and preservation of higher
+    order logic by geometric morphisms.}
+  \end{itemize}}
   \pause
@@
     there is an enumeration~$\NN \twoheadrightarrow X$.
+    \bigskip
 
@@
     every sequence is good.
+    \bigskip
 
@@
     all prime ideals \emph{everywhere} contain it.
+    \bigskip
 
     For every ring, \emph{proximally} \\ there is a maximal ideal.
+    \bigskip
 
     A relation is well-founded iff \emph{everywhere}, there is no descending chain.
+    \bigskip
 
@@
   \begin{enumerate}
-    \item \!\emph{Why was the inductive revolution so powerful?}
+    \item \!\emph{Why has the inductive revolution been so powerful?}
 
@@
   \vspace*{\paperheight}\vspace*{-2.05em}\centering
-  \emph{Agda formalization available.}
+  \emph{Partial Agda formalization available.}
 \end{frame}}

### 2023-05-02 slides-luminy2023.tex
Absolutely final version

@@
 \usepackage{tikz}
-\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing,positioning,svg.path,fadings}
+\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing,positioning,svg.path}
 \hypersetup{colorlinks=true}
@@
 
-\begin{tikzfadingfrompicture}[name=fade bottom]
-  \shade[top color=transparent!0,
-    bottom color=transparent!50] (0,0) rectangle (2,2);
-    \end{tikzfadingfrompicture}
+\begin{frame}{Questions}
+  \begin{enumerate}
+    \item \!Why has the inductive revolution been so powerful?
+    \bigskip
+
+    \item \!Why do proofs using Zorn's maximal ideals work so well in constructive algebra?
+    \bigskip
+
+    \item \!Why are elements of~$\bigcap_\ppp \ppp$ not necessarily nilpotent?
+    \bigskip
+
+    \item How can we extract computational content from classical proofs?
+  \end{enumerate}
+\end{frame}
 
@@
     \begin{align*}
-      \mmm_0 &= \{ 0 \}, &
-      \mmm_{n+1} &= \begin{cases}
+      \mmm_0 &\defeq \{ 0 \}, &
+      \mmm_{n+1} &\defeq \begin{cases}
         \mmm_n + (x_n), & \text{if $1 \not\in \mmm_n + (x_n)$}, {\qquad\quad\quad\!\!}\\
@@
     {\footnotesize\begin{align*}
-      \mmm_0 &= \{ 0 \}, &
-      \mmm_{n+1} &= \mmm_n + (\underbrace{\{ x \in A \,|\, x = x_n \wedge
+      \mmm_0 &\defeq \{ 0 \}, &
+      \mmm_{n+1} &\defeq \mmm_n + (\underbrace{\{ x \in A \,|\, x = x_n \wedge
         1 \not\in \mmm_n + (x_n) \}}_{\text{a certain subsingleton set}})
@@
 
-    \item In the general case: \bad{No}, but the \emph{first-order
+    \item In the general case: \bad{No}\pause, \\ but \emph{first-order
     consequences} of the existence of a maximal ideal \good{do hold}.
@@
 
-  {\centering\footnotesize\begin{tabular}{lll}
+  {\centering\footnotesize\begin{tabular}{llll}
     \toprule
-    preorder~$L$ & coverings of an element~$\sigma \in L$ & filters of~$L$ \\
+    & preorder~$L$ & coverings of an element~$\sigma \in L$ & filters of~$L$ \\
     \midrule
-    $X^*$ & $\{ \sigma x \,|\, x \in X \}$ & maps~$\NN \to X$ \\
-    $X^*$ & $\{ \sigma x \,|\, x \in X \}$,\ \
-    $\{ \sigma\tau \,|\, \tau \in X^*, a \in \sigma\tau \}$ for each~$a \in X$ & surjections~$\NN \twoheadrightarrow X$ \\
-    f.g. ideals & --- & ideals \\
-    f.g. ideals & $\{ \sigma+(a), \sigma+(b) \}$ for each~$ab \in \sigma$,\ \
-    $\{\}$ if~$1 \in \sigma$ & prime ideals \\
-    opens & $\mathcal{U}$ such that~$\sigma = \bigcup \mathcal{U}$ & points \\
-    $\{\star\}$ & $\{ \star \,|\, \varphi \} \cup \{ \star \,|\, \neg\varphi \}$ &
+    \normalnumber{1} & $X^*$ & $\{ \sigma x \,|\, x \in X \}$ & maps~$\NN \to X$ \\
+    \normalnumber{2} & $X^*$ & $\{ \sigma x \,|\, x \in X \}$,\ \ $\{ \sigma\tau \,|\, \tau \in X^*, a \in \sigma\tau \}$ for each~$a \in X$ & surjections~$\NN \twoheadrightarrow X$ \\
+    \normalnumber{3} & f.g. ideals & --- & ideals \\
+    \normalnumber{4} & f.g. ideals & $\{ \sigma+(a), \sigma+(b) \}$ for each~$ab \in \sigma$,\ \ $\{\}$ if~$1 \in \sigma$ & prime ideals \\
+    \normalnumber{5} & opens & $\mathcal{U}$ such that~$\sigma = \bigcup \mathcal{U}$ & points \\
+    \normalnumber{6} & $\{\star\}$ & $\{ \star \,|\, \varphi \} \cup \{ \star \,|\, \neg\varphi \}$ &
     witnesses of~\textsc{lem}
@@
   of the base universe~$V$:
+  \medskip
 
@@
   \begin{multicols}{2}
-    For every set~$X$, \emph{proximally} \\
+    For every inhabited set~$X$, \emph{proximally} \\
     there is an enumeration~$\NN \twoheadrightarrow X$.

### 2023-06-29 paper-qcoh.tex
typo (closing #4)

@@
 that~$\TT$ is the theory of commutative rings with unit and~$A$ is the
-ring~$\QQ$ of integers, we have~$\Spec(A^\sim) \cong \Spec(0^\sim)$, where~$0$
-is the zero ring, as~$\QQ$ allows ring homomorphisms only in those finitely
+ring~$\QQ$ of rational numbers, we have~$\Spec(A^\sim) \cong \Spec(0^\sim)$, where~$0$
+is the zero ring, as~$\QQ$ allows ring homomorphisms only to those finitely
 presented rings in which~$1 = 0$ holds. Hence~$A^\sim$

### 2023-09-11 slides-herrsching2023.tex
Beginning of Herrsching slides

--- 2023-05-02 slides-luminy2023.tex
@@
 
-{
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\includegraphics[height=\paperheight]{multiverse-faded}\end{minipage}}
+\begin{frame}[c]
+  \centering
+  \color{white}
+
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+
+  \scriptsize
+  \textit{-- an invitation --}
+
+  \setbeamercolor{block body}{bg=black!100}
+  \begin{minipage}{0.35\textwidth}
+    \begin{block}{}
+      \centering\normalsize\color{white}
+      \hil{Constructive forcing}
+    \end{block}
+  \end{minipage}
+
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+
+%  \colorbox{black!50}{Autumn school on} \\
+%  \colorbox{black!50}{\emph{Proof and Computation}} \\
+%  \colorbox{black!50}{in Herrsching} \\
+%  \ \\
+%  \colorbox{black!50}{September 20th to September 16th, 2023}
+%
+%  \colorbox{black!50}{Ingo Blechschmidt} \\
+%  \par
+\end{frame}}
+
 \definecolor{mypurple}{RGB}{150,0,255}
 \setbeamercolor{structure}{fg=mypurple}
+
 \begin{frame}{}
@@
 \end{frame}
-}
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[height=\paperheight]{night-sky}\end{minipage}}
-\begin{frame}[c]
-  \centering
-  \color{white}
+% \begin{document}
+
+\newcommand\ytl[2]{%
+  \parbox[b]{4.5em}{\hfill{#1}~$\cdots\cdots$~}%
+  \makebox[0pt][c]{\color{mypurple}$\bullet$}{\color{mypurple}\vrule}\quad%
+  \parbox[c]{9.3cm}{\vspace{7pt}\raggedright#2\\[7pt]}%
+  \\[-2.0pt]}
+\begin{frame}{A brief timeline}
+  \begin{columns}[t]
+    \begin{column}{0.86\textwidth}
+      \vspace*{-1.0em}
+      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-3.5pt]
+      \ytl{1878}{Cantor advances the \hil{continuum hypothesis}, the claim
+      that~$2^{\aleph_0} = \aleph_1$.}
+      \pause
+      \ytl{1910s}{Zermelo--Fraenkel set theory emerges.}
+      \pause
+      \ytl{1920s}{Set theorists pursue additional axioms to
+      settle~\textsc{ch} \\ (one way or another).}
+      \pause
+      \ytl{1938}{Gödel proves: If~\textsc{zfc} is consistent, so
+      is~\textsc{zfc}+\textsc{ch}.}
+      \pause
+      \ytl{1963}{Cohen proves: If~\textsc{zfc} is consistent, so
+      is~\textsc{zfc}+$\neg$\textsc{ch}.}
+      \pause
+      \ytl{2011}{Hamkins offers his paper on the \hil{multiverse position} in
+      the philosophy of set theory.}
+      \pause
+      %arguing that the program of pursuing
+      %additional axioms (while successful in many ways) is doomed to fail
+      %to settle~\textsc{ch}.}
+      \ytl{2016}{Oldenziel proposes to study the modal multiverse of parametrized
+      mathematics.}
+      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
+      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
+      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
+    \end{column}
+
+    \begin{column}{0.20\textwidth}
+      \pause
+      \centering\includegraphics[width=\textwidth,valign=t]{roitman}
+
+      \scriptsize Judith Roitman %(* 1945)
+      \medskip
+
+      \emph{Mainstream mathematics is beginning to see results using
+      modern set theoretic techniques.}
+    \end{column}
+  \end{columns}
+\end{frame}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
+\begin{frame}{Constructive forcing}
+  \medskip
+  \begin{columns}[t]
+    \begin{column}{0.40\textwidth}
+      \quad\includegraphics[width=0.8\textwidth,valign=t]{branching}
+    \end{column}
+
+    \begin{column}{0.75\textwidth}
+      \hil{Extending the universe} in various ways, \\[0.7em]
+      \pause
+      \quad similarly how we can extend groups or rings, \\[0.7em]
+      \pause
+      \quad\quad in and for \hil{constructive mathematics} \\[0.7em]
+      \pause
+      \quad\quad\quad \hil{without} presupposing familiarity with \\[0.7em]
+      \pause
+      \quad\quad\quad\quad set theory, topos theory, or sheaves.
+      \pause
+    \end{column}
+  \end{columns}
 
@@
   \bigskip
-  \bigskip
-  \bigskip
 
-  \scriptsize
-  \textit{-- an invitation --}
-
-  \setbeamercolor{block body}{bg=black!100}
-  \begin{minipage}{0.8\textwidth}
-    \begin{block}{}
-      \centering\normalsize\color{white}
-      \hil{New modal operators for~constructive~mathematics}
-    \end{block}
-  \end{minipage}
-
-  \bigskip
-  \bigskip
-  \bigskip
-  \bigskip
-  \bigskip
-  \bigskip
-  \bigskip
-
-  \emph{Type Theory, Constructive Mathematics and Geometric Logic} \\
-  \bigskip
-
-  CIRM \\
-  May 2nd, 2023
-  \bigskip
-
-  Ingo Blechschmidt \\
-  j.w.w. Alexander Oldenziel
-  \par
-\end{frame}}
-
-\definecolor{mypurple}{RGB}{150,0,255}
-\setbeamercolor{structure}{fg=mypurple}
-
-\section{Motivation}
-
-\begin{frame}{Questions}
+  \emph{Outline:}
+  \vspace*{-0.5em}
   \begin{enumerate}
-    \item \!Why has the inductive revolution been so powerful?
-    \bigskip
-
-    \item \!Why do proofs using Zorn's maximal ideals work so well in constructive algebra?
-    \bigskip
-
-    \item \!Why are elements of~$\bigcap_\ppp \ppp$ not necessarily nilpotent?
-    \bigskip
-
-    \item How can we extract computational content from classical proofs?
-  \end{enumerate}
-\end{frame}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[height=\paperheight]{sea-of-clouds-2}\end{minipage}}
-\begin{frame}{Infinite data}
-  \vspace*{-1em}
-  \[ \astikznodetransparentlycircled{xm}{7}\!,
-    \quad \astikznodetransparentlycircled{x0}{4}\!,
-    \quad \only<1-2>{\astikznodetransparentlycircled{t1}{3}}\only<3->{\astikznodecircled{t1}{mypurple}{3}}\!,
-    \quad \only<1>{\ldots}\pause \astikznodetransparentlycircled{x1}{1}\!,
-    \quad \only<2>{\ldots}\pause \astikznodecircled{t2}{mypurple}{8}\!,
-    \quad \only<3>{\ldots} \visible<4->{\astikznodetransparentlycircled{x2}{2}\!,}
-    \quad \only<4->{\ldots} \]
-  {\centering\begin{tikzpicture}[remember picture,overlay]
-    \node[draw=mypurple, circle, thick, inner sep=0.1em] (t3) {\scriptsize$\leq$};
-    \path[draw=mypurple,thick]
-      (t1)
-      to [out=-90, in=180] (t3)
-      to [out=0, in=-90] (t2);
-  \end{tikzpicture}\par}
-  \medskip
-  \pause
-
-  \textbf{Thm.} Every sequence~$\alpha : \NN \to \NN$ is \hil{good} in that
-  there exist~$i < j$ with~$\alpha(i) \leq \alpha(j)$.
-  \pause
-
-  \emph{Proof.} \emph{(offensive?)} By~\badbox{\textsc{lem}}, there is a
-  minimum~$\alpha(i)$.
-  Set~$j \defeq i + 1$. \qed\par
-  \pause
-  \medskip
-
-  \textbf{Def.} A preorder~$X$ is \hil{well\only<9->{$^\star$}} iff every sequence~$\NN \to X$ is good.
-
-  \textbf{Examples.} $(\NN,{\leq}),\ \
-  \color{white}\only<7->{\color{red!90}}\astikznode{onlyclass}{$\underbrace{\color{black}X \times Y,\ \ X^*,\ \ \mathrm{Tree}(X)}_{\text{\visible<7->{\bad{only classically}}}}$}$.
-  \pause
-  \pause
-  \medskip
-
-  \begin{tikzpicture}[remember picture,overlay]
-    \node[thick, fill=black, rectangle, inner sep=0.3em, right=2em of onlyclass] (moral) {
-      \begin{minipage}{6cm}
-        \begin{columns}
-          \begin{column}{0.15\textwidth}
-            \hspace*{1.0em}\color{white}\dbend
-          \end{column}
-          \begin{column}{0.95\textwidth}
-            \color{white}\footnotesize
-            \it Don't quantify over points of spaces which might not have enough.
-          \end{column}
-        \end{columns}
-      \end{minipage}
-    };
-    \path[draw=red!90,thick,-stealth]
-      (moral) to
-      [out=180, in=-00] (onlyclass.east);
-  \end{tikzpicture}
-  \pause
-
-  \textbf{Def.} For a predicate~$P$ on finite lists over a set~$X$, inductively
-  define:
-  \[
-    \infer{P \,|\, \sigma}{P\sigma}
-    \qquad
-    \infer{P \,|\, \sigma}{\forall(x \in X)\_\ P \,|\, \sigma x}
-  \]
-
-  \textbf{Def.} A preorder is \hil{well} iff~$\mathsf{Good} \,|\, [\,]$, where
-  $\mathsf{Good}\,\sigma \defeqv (\exists(i < j)\_ \sigma[i] \leq \sigma[j])$.
-\end{frame}}
-
-\begin{frame}{Computational content from classical proofs}
-  \textbf{Def.} A transitive relation~$({<})$ on a set~$X$ is \ldots
-  \begin{enumerate}
-    \item \hil{well-founded$^\star$} iff
-    there is no \bad{infinite chain} $x_0 > x_1 > \cdots$,
-    \item \hil{well-founded} iff for every~$x \in X$,
-    $\mathsf{Acc}(x)$,
-  \end{enumerate}
-  where~$\mathsf{Acc}$ is inductively defined by:
-  \[
-    \infer{\mathsf{Acc}(x)}{\forall(y < x)\_\ \mathsf{Acc}(y)}
-  \]
-  \vspace*{-1em}
-  \pause
-
-  \textbf{Prop.} Let~$(X,{\leq})$ be preorder. Let~``$x < y$'' mean~$x \leq y \wedge \neg(y \leq x)$. \\
-  Then: If~$X$ is well$^\star$, then~$({<})$ is well-founded$^\star$.
-
-  \emph{Proof.} An infinite strictly descending chain would also be good. \qed
-  \pause
-
-  Can we extract a constructive proof that well preorders are well-founded?
-\end{frame}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
-\begin{frame}{Maximal ideals}
-  \only<1-7>{\textbf{Thm.}
-  Let~$M$ be a surjective matrix with more rows than columns over a
-  ring~$A$. Then~$1 = 0$ in~$A$.
-
-  \visible<2->{\emph{Proof.} \bad{Assume not.}}
-  \visible<3->{Then there is~a \bad{maximal ideal} $\mmm$.}
-  \visible<5->{The matrix is surjective over~$A/\mmm$.}
-  \visible<6->{Since~$A/\mmm$ is a field, this is a contradiction to basic linear algebra.\qed}}
-
-  \only<4-7>{\bigskip\par\centering\scalebox{0.9}{\centering\begin{tikzpicture}
-    \node (0) at (0,1) {$(0) = \{0\}$};
-    \node (1) at (0,5) {$(1) = \ZZ$};
-    \node (2) at (-2,4) {$(2)$};
-    \node [right of=2] (3) {$(3)$};
-    \node [below of=2] (4) {$(4)$};
-    \node [below of=2, xshift=0.7cm] (6) {$(6)$};
-    \node [right of=3] (5) {$(5)$};
-    \node [right of=5] (7) {$(7)$};
-    \node [right of=7] (7d) {$\ldots$\phantom{(}};
-    \node [right of=7d, xshift=3cm, yshift=-2cm] (max)
-    {\vbox{\small\it maximal among the proper ideals \\ \medskip $\neg(1 \in
-    \mmm)$ \\ \medskip $x \in \mmm \quad\vee\quad 1 \in \mmm + (x)$}};
-    \node [below of=4] (8) {$(8)$};
-    \node [right of=8, xshift=3cm] (8d) {$\ldots$};
-    \draw (0) -- (8);
-    \draw (0) -- (8d);
-    \draw (0) -- (6);
-    \draw (2) -- (1);
-    \draw (3) -- (1);
-    \draw (5) -- (1);
-    \draw (7) -- (1);
-    \draw (7d) -- (1);
-    \draw (4) -- (2);
-    \draw (8) -- (4);
-    \draw (6) -- (2);
-    \draw (6) -- (3);
-    \draw [mypurple!30, thick, shorten <=-2pt, shorten >=-2pt, ->] (max) to [out=120, in=-30] (7d);
-    \begin{pgfonlayer}{background}
-      \draw[decorate, very thick, draw=mypurple!30]
-        ($(2.south west) + (8pt, 0)$) arc(270:180:8pt) --
-        ($(2.north west) + (0, -8pt)$) arc(180:90:8pt) --
-        ($(7d.north east) + (-8pt, 0)$) arc(90:0:8pt) --
-        ($(7d.south east) + (0, 8pt)$) arc(0:-90:8pt) --
-        cycle;
-    \end{pgfonlayer}
-  \end{tikzpicture}\par}\par}
-
-  \pause
-  \pause
-  \pause
-  \pause
-  \pause
-  \pause
-  \medskip
-  \raggedright
-
-  Let~$A$ be a ring. \emph{Does there exist a maximal ideal~$\mmm \subseteq A$?}
-  \pause
-  \begin{enumerate}
-    \item \good{Yes}, if \bad{Zorn's lemma} is available.
-    \bigskip
-    \pause
-
-    \item \good{Yes}, if~$A$ is countable and membership of finitely generated ideals is decidable:
-    {\footnotesize
-    \begin{align*}
-      \mmm_0 &\defeq \{ 0 \}, &
-      \mmm_{n+1} &\defeq \begin{cases}
-        \mmm_n + (x_n), & \text{if $1 \not\in \mmm_n + (x_n)$}, {\qquad\quad\quad\!\!}\\
-        \mmm_n, & \text{else.}
-      \end{cases}
-    \end{align*}}
-    \pause
-
-    \item \good{Yes}, if~$A$ is countable (irrespective of membership decidability):
-    {\footnotesize\begin{align*}
-      \mmm_0 &\defeq \{ 0 \}, &
-      \mmm_{n+1} &\defeq \mmm_n + (\underbrace{\{ x \in A \,|\, x = x_n \wedge
-        1 \not\in \mmm_n + (x_n) \}}_{\text{a certain subsingleton set}})
-    \end{align*}}
-    \vspace*{3.5em}
-    \visible<11>{{
-      \centering
-      \raisebox{0pt}[0pt][0pt]{
-        \hspace*{1.5em}
-        \begin{tikzpicture}
-          \node (inner) at (17.3mm,-20mm) {\textit{``a bad joke''}};
-          \path (0,0) pic{laurel-wreath};
-        \end{tikzpicture}
-        \hspace*{-3em}
-        \begin{tikzpicture}
-          \node (inner) at (17.5mm,-18mm) {\vbox{\small\centering\textit{``non- \\informative''}}};
-          \path (0,0) pic{laurel-wreath};
-        \end{tikzpicture}
-      }
-      \par
-    }}
-    \pause
-    \pause
-    \vspace*{-3.5em}
-
-    \item In the general case: \bad{No}\pause, \\ but \emph{first-order
-    consequences} of the existence of a maximal ideal \good{do hold}.
+    \item What can forcing do for you?
+    \item Forcing notions and Kripke--Joyal semantics
+    \item Case studies in constructive algebra and combinatorics
   \end{enumerate}
@@
 
-\begin{frame}{Questions}
-  \begin{enumerate}
-    \item \!Why has the inductive revolution been so powerful?
-    \bigskip
+\end{document}
 
-    \item \!Why do proofs using Zorn's maximal ideals work so well in constructive algebra?
-    \bigskip
+- Judith Roitman
+- Proof and Computation
 
-    \item \!Why are elements of~$\bigcap_\ppp \ppp$ not necessarily nilpotent?
-    \bigskip
+- What can forcing do for you?
+  - Exploring the range of foundational possibility: CH, ¬CH, FTA, ...
+  - Eigenvalues
+  - Generic sequence
+  - Generic surjection
+  - Matthias
+  - "Harness the flexibility of parametric mathematics in an accessible way."
 
-    \item How can we extract computational content from classical proofs?
-  \end{enumerate}
-\end{frame}
+Overloading
 
-\renewcommand{\insertframeextra}{a}
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.23cm}\includegraphics[width=\paperwidth]{topos-horses-lighter-lighter}\end{minipage}}
-\begin{frame}{Constructive forcing (= Grothendieck toposes)}
-  \justifying
-  Let~$L$ be a \hil{forcing notion}, a preorder equipped
-  with a \hil{covering system}.\only<1-2>{\footnote{\scriptsize A covering system consists of a
-  set~$\Cov(\sigma) \subseteq P({\downarrow}\sigma)$ of \emph{coverings} for
-  each element~$\sigma \in L$ subject only to the following simulation
-  condition: If~$\tau \preccurlyeq \sigma$ and~$R \in \Cov(\sigma)$, there
-  should be a covering~$S \in \Cov(\tau)$ such that~$S \subseteq
-  {\downarrow}R$.}}\only<3->{\textsuperscript{\phantom{1}}}
-  \hil{Filters}~$F \subseteq L$ are subsets which are upward-closed,
-  downward-directed and split the covering system.\only<1-2>{\footnote{\scriptsize
-  A subset~$F$ splits the covering system iff for every~$\sigma \in L$ and~$R \in \Cov(\sigma)$,
-  if~$\sigma \in F$, then~$\tau \in F$ for some~$\tau \in R$.}}\only<3>{\textsuperscript{\phantom{2}}}
-  \pause
+1. Eigenvalues
+2. Generic surjection
+3. Generic sequence
+4. FTA
 
-  {\centering\footnotesize\begin{tabular}{llll}
-    \toprule
-    & preorder~$L$ & coverings of an element~$\sigma \in L$ & filters of~$L$ \\
-    \midrule
-    \normalnumber{1} & $X^*$ & $\{ \sigma x \,|\, x \in X \}$ & maps~$\NN \to X$ \\
-    \normalnumber{2} & $X^*$ & $\{ \sigma x \,|\, x \in X \}$,\ \ $\{ \sigma\tau \,|\, \tau \in X^*, a \in \sigma\tau \}$ for each~$a \in X$ & surjections~$\NN \twoheadrightarrow X$ \\
-    \normalnumber{3} & f.g. ideals & --- & ideals \\
-    \normalnumber{4} & f.g. ideals & $\{ \sigma+(a), \sigma+(b) \}$ for each~$ab \in \sigma$,\ \ $\{\}$ if~$1 \in \sigma$ & prime ideals \\
-    \normalnumber{5} & opens & $\mathcal{U}$ such that~$\sigma = \bigcup \mathcal{U}$ & points \\
-    \normalnumber{6} & $\{\star\}$ & $\{ \star \,|\, \varphi \} \cup \{ \star \,|\, \neg\varphi \}$ &
-    witnesses of~\textsc{lem}
-    \\
-    \bottomrule
-  \end{tabular}\par}
-  \bigskip
+Extending the universe in various ways,
 
-  \only<3>{\textbf{Def.} Given a monotone predicate~$P$
-  on~$L$, inductively define:
-  \[
-    \infer{P \,|\, \sigma}{P\sigma}
-    \qquad
-    \infer[{\small(R \in \Cov(\sigma))}]{P \,|\, \sigma}{\forall(\tau \in R)\_\ P \,|\, \tau}
-  \]
-  We use quantifier-like notation: ``$\nabla\sigma\_ P\sigma$'' means~$P \,|\, \sigma$.}
-\end{frame}}
+similarly how we can extend groups or rings,
 
-\addtocounter{framenumber}{-1}
-\renewcommand{\insertframeextra}{b}
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.23cm}\includegraphics[width=\paperwidth]{topos-horses-lighter-lighter}\end{minipage}}
-\begin{frame}{Constructive forcing (= Grothendieck toposes)}
-  A forcing notion is a template for a \hil{forcing extension}~$V^\nabla$
-  of the base universe~$V$:
-  \medskip
+in the context of and for the purposes of constructive mathematics
 
-  \begin{center}\begin{tikzpicture}[remember picture,overlay]
-    \node[thick, draw=mypurpledark, rectangle, inner sep=0.3em] (kj) {
-      \begin{minipage}{0.6\textwidth}
-        \centering
-        \emph{When we say that a statement \hil{holds in~$V^\nabla$},} \\
-        we mean that its~$\nabla$-translation is true.\par
-        \vspace*{0.3em}
-      \end{minipage}
-    };
-  \end{tikzpicture}\end{center}
-  \bigskip
-  \pause
+without requiring knowledge of
 
-  \textbf{Examples.}
-  \begin{enumerate}
-  \justifying
-  \item For~$L = \{\star\}$, ``every number is zero or a successor'' holds
-  in~$V^\nabla$ because
-
-  \vspace*{-1em}
-  \[ \forall(n \in \NN)\_ \neg\neg\bigl(\neg\neg(n = 0) \vee
-    \neg\neg(\exists(m \in \NN)\_ \neg\neg(n = S(m)))\bigr). \]
-
-  \pause
-  \item For~$L$ associated to a space, ``every symmetric matrix has an
-  eigenvector'' holds in~$V^\nabla$ iff for every continuous family of
-  symmetric matrices, locally there is a continuous eigenvector-picking
-  function.
-  \pause
-  \medskip
-
-  \item For any~$L$, ``there is a filter of~$L$'' holds in~$V^\nabla$,
-  witnessed by the \hil{generic filter} of~$L$.
-  \pause
-  \medskip
-
-  \item Let~$X$ be a preorder. Then:
-
-  \hil{$\blacktriangleright$} $X$ is well iff the generic sequence~$\NN \to X$
-  is good.
-
-  \hil{$\blacktriangleright$} $({<})$ is well-founded iff the generic strictly
-  descending chain validates~$\bot$.
-  \end{enumerate}
-\end{frame}}
-
-\renewcommand{\insertframeextra}{}
-\begin{frame}{The modal multiverse}
-  \only<1-2>{In general, ``$\varphi$ holds in~$V^\nabla$'' and~``$\varphi$ holds in~$V$''
-  are \emph{not} equivalent.
-
-  \hil{$\blacktriangleright$}
-  For \hil{positive} extensions, they are equivalent for coherent implications.\\
-  %\footnote{bounded first-order formulas of the
-  %form~$\forall\ldots\forall\_(\ldots\Rightarrow\ldots)$,
-  %where antecedent and succedent do not contain~``$\forall$''
-  %and~``${\Rightarrow}$''}.
-  \phantom{\hil{$\blacktriangleright$}} {\footnotesize -- e.g. the ``Barr cover''.\par}
-
-  \hil{$\blacktriangleright$}
-  For \hil{positive overt} extensions, they are equivalent for bounded first-order formulas. \\
-  \phantom{\hil{$\blacktriangleright$}} {\footnotesize -- e.g. $V^\nabla$ containing
-  the generic surjection~$\NN \twoheadrightarrow X$, if~$X$ is inhabited.\par}
-  \bigskip}
-  \pause
-
-  \textbf{Def.} A statement~$\varphi$ holds \ldots
-  \begin{enumerate}
-    \small
-    \item[$\blacktriangleright$] \hil{everywhere} ($\necessary\varphi$) iff
-    it holds in every extension.
-    \item[$\blacktriangleright$] \hil{somewhere} ($\possible\varphi$) iff
-    it holds in some positive extension.
-    \item[$\blacktriangleright$] \hil{proximally} ($\xpossible\varphi$) iff
-    it holds in some positive overt extension.
-  \end{enumerate}
-  \bigskip
-
-  \visible<2>{\begin{tikzpicture}[overlay]
-    \node[anchor=south east,inner sep=0] (image) at (14.8,-0.8) {
-      \includegraphics[width=0.2\textwidth]{branching}
-    };
-  \end{tikzpicture}\vspace*{-1.6em}}
-  \visible<3>{\begin{tikzpicture}[overlay]
-    \node[anchor=south east,inner sep=0] (image) at (14.8,0.9) {
-      \includegraphics[width=0.2\textwidth]{multiverse}
-    };
-  \end{tikzpicture}\vspace*{-0.3em}}
-
-  \only<2>{\scriptsize Foreshadowed by:
-  \begin{itemize}
-    \scriptsize
-    \item[1984\phantom{s}] André Joyal, Miles Tierney. \emph{An extension of the Galois theory of
-    Grothendieck.}
-    \item[1987\phantom{s}] Andreas Blass. \emph{Well-ordering and induction in intuitionistic
-    logic and topoi.}
-    \item[2010s] Milly Maietti, Steve Vickers. Ongoing work on arithmetic
-    universes.
-    \item[2011\phantom{s}] Joel David Hamkins. \emph{The set-theoretic multiverse.}
-    \item[2013\phantom{s}] Shawn Henry. \emph{Classifying topoi and preservation of higher
-    order logic by geometric morphisms.}
-  \end{itemize}}
-  \pause
-
-  \small
-  \begin{multicols}{2}
-    For every inhabited set~$X$, \emph{proximally} \\
-    there is an enumeration~$\NN \twoheadrightarrow X$.
-    \bigskip
-
-    A preorder is well iff \emph{everywhere}, \\
-    every sequence is good.
-    \bigskip
-
-    A ring element is nilpotent iff \\
-    all prime ideals \emph{everywhere} contain it.
-    \bigskip
-
-    For every ring, \emph{proximally} \\ there is a maximal ideal.
-    \bigskip
-
-    A relation is well-founded iff \emph{everywhere}, there is no descending chain.
-    \bigskip
-
-    \emph{Somewhere}, \\ the law of excluded middle holds.
-  \end{multicols}
-\end{frame}
-
-\begin{frame}{The multiverse perspective}
-  \begin{enumerate}
-    \item \!\emph{Why has the inductive revolution been so powerful?}
-
-    Because the inductive conditions
-    are equivalent to truth in \emph{all} forcing extensions.
-    \bigskip
-
-    \item \!\emph{Why do proofs using Zorn's maximal ideals work so well in constructive algebra?}
-
-    Because every ring proximally has a maximal ideal.
-    \bigskip
-
-    \item \!\emph{Why are elements of~$\bigcap_\ppp \ppp$ not necessarily nilpotent?}
-
-    Because we forgot the prime ideals in forcing extensions.
-    \bigskip
-
-    \item \emph{How can we extract computational content from classical proofs?}
-
-    By traveling the multiverse (upwards, keeping ties to the base), exploiting
-    that \\
-    -- \textsc{lem} holds \emph{somewhere} and \\
-    -- \textsc{dc} holds \emph{proximally}.
-  \end{enumerate}
-\end{frame}
-
-\backupstart
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[width=\paperwidth]{eventually-monad-cropped}\end{minipage}}
-\begin{frame}
-  \vspace*{\paperheight}\vspace*{-2.05em}\centering
-  \emph{Partial Agda formalization available.}
-\end{frame}}
-
-\backupend
-
-\end{document}
+set theory, topos theory, category theory, sheaves.

### 2023-09-11 slides-herrsching2023.tex
What can constructive forcing do for you?

@@
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses}\end{minipage}}
 \begin{frame}{Constructive forcing}
@@
 
+\begin{frame}{What can forcing do for you?}
+  \vspace*{-0.5em}
+
+  \begin{columns}
+    \begin{column}{0.45\textwidth}
+      \begin{block}{1. Explore foundational possibility}
+        There are forcing extensions with~\textsc{ch}, $\neg$\textsc{ch},
+        \textsc{lem}, $\neg$\textsc{lem}, \ldots
+      \end{block}
+    \end{column}
+    \pause
+
+    \begin{column}{0.45\textwidth}
+      \begin{block}{2. Demonstrate unprovability}
+        The fundamental theorem of algebra is \hil{not constructively provable}
+        as there is a forcing extension where \hil{it is false}.
+      \end{block}
+    \end{column}
+  \end{columns}
+  \pause
+  \bigskip
+
+  \begin{columns}
+    \begin{column}{0.45\textwidth}
+      \begin{block}{3. Harness convenient fictions}
+        For every set, there is a forcing extension where it is
+        \hil{countable}.
+      \end{block}
+    \end{column}
+    \pause
+
+    \begin{column}{0.45\textwidth}
+      \begin{block}{4. Constructivize classical theories}
+        A preorder~$X$ is well iff the \hil{generic sequence} $\NN \to X$ is
+        good.
+      \end{block}
+    \end{column}
+  \end{columns}
+  \pause
+  \bigskip
+
+  \begin{columns}
+    \begin{column}{0.45\textwidth}
+      \begin{block}{5. Study parametric mathematics}
+        Eigenvectors depend continuously on the parameter
+        iff, in a suitable forcing extension, they merely exist.
+      \end{block}
+    \end{column}
+    \pause
+
+    \begin{column}{0.45\textwidth}
+      \begin{block}{6. Build synthetic accounts}
+        \emph{See lectures by Matthias Hutzler.}
+      \end{block}
+    \end{column}
+  \end{columns}
+\end{frame}
+
 \end{document}

### 2023-09-11 slides-herrsching2023.tex
Two concrete examples

@@
 
-\addtocounter{framenumber}{-2}
+\addtocounter{framenumber}{-1}
 
@@
 
-  There for every parameter value~$t$, classically there is
+  Then for every parameter value~$t$, classically there is
 
@@
   \begin{columns}[t]
-    \begin{column}{0.86\textwidth}
-      \vspace*{-1.0em}
-      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-3.5pt]
+    \begin{column}{0.80\textwidth}
       \ytl{1878}{Cantor advances the \hil{continuum hypothesis}, the claim
@@
     \begin{column}{0.45\textwidth}
-      \begin{block}{6. Build synthetic accounts}
-        \emph{See lectures by Matthias Hutzler.}
+      \begin{block}{6. Develop synthetic accounts}
+        \emph{As in the lectures by Matthias Hutzler.}
       \end{block}
@@
 \end{frame}
+
+% \begin{document}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
+\begin{frame}{Maximal ideals}
+  \only<1-7>{\textbf{Thm.}
+  Let~$M$ be a surjective matrix with more rows than columns over a
+  ring~$A$. Then~$1 = 0$ in~$A$.
+
+  \visible<2->{\emph{Proof.} \bad{Assume not.}}
+  \visible<3->{Then there is~a \bad{maximal ideal} $\mmm$.}
+  \visible<5->{The matrix is surjective over~$A/\mmm$.}
+  \visible<6->{Since~$A/\mmm$ is a field, this is a contradiction to basic linear algebra.\qed}}
+
+  \only<4-7>{\bigskip\par\centering\scalebox{0.9}{\centering\begin{tikzpicture}
+    \node (0) at (0,1) {$(0) = \{0\}$};
+    \node (1) at (0,5) {$(1) = \ZZ$};
+    \node (2) at (-2,4) {$(2)$};
+    \node [right of=2] (3) {$(3)$};
+    \node [below of=2] (4) {$(4)$};
+    \node [below of=2, xshift=0.7cm] (6) {$(6)$};
+    \node [right of=3] (5) {$(5)$};
+    \node [right of=5] (7) {$(7)$};
+    \node [right of=7] (7d) {$\ldots$\phantom{(}};
+    \node [right of=7d, xshift=3cm, yshift=-2cm] (max)
+    {\vbox{\small{\it maximal among the proper ideals} \\ \medskip \hspace*{-6.75em}\textbullet \quad $\neg(1 \in
+    \mmm)$ \\ \medskip \textbullet \quad $x \in \mmm \quad\vee\quad 1 \in \mmm + (x)$}};
+    \node [below of=4] (8) {$(8)$};
+    \node [right of=8, xshift=3cm] (8d) {$\ldots$};
+    \draw (0) -- (8);
+    \draw (0) -- (8d);
+    \draw (0) -- (6);
+    \draw (2) -- (1);
+    \draw (3) -- (1);
+    \draw (5) -- (1);
+    \draw (7) -- (1);
+    \draw (7d) -- (1);
+    \draw (4) -- (2);
+    \draw (8) -- (4);
+    \draw (6) -- (2);
+    \draw (6) -- (3);
+    \draw [mypurple!30, thick, shorten <=-2pt, shorten >=-2pt, ->] (max) to [out=120, in=-30] (7d);
+    \begin{pgfonlayer}{background}
+      \draw[decorate, very thick, draw=mypurple!30]
+        ($(2.south west) + (8pt, 0)$) arc(270:180:8pt) --
+        ($(2.north west) + (0, -8pt)$) arc(180:90:8pt) --
+        ($(7d.north east) + (-8pt, 0)$) arc(90:0:8pt) --
+        ($(7d.south east) + (0, 8pt)$) arc(0:-90:8pt) --
+        cycle;
+    \end{pgfonlayer}
+  \end{tikzpicture}\par}\par}
+
+  \pause
+  \pause
+  \pause
+  \pause
+  \pause
+  \pause
+  \medskip
+  \raggedright
+
+  Let~$A$ be a ring. \emph{Does there exist a maximal ideal~$\mmm \subseteq A$?}
+  \pause
+  \begin{enumerate}
+    \item \good{Yes}, if \bad{Zorn's lemma} is available.
+    \bigskip
+    \pause
+
+    \item \good{Yes}, if~$A$ is countable and membership of finitely generated ideals is decidable:
+    {\footnotesize
+    Let~$A = \{ x_0, x_1, \ldots \}$. Then set:
+    \begin{align*}
+      \mmm_0 &\defeq \{ 0 \}, &
+      \mmm_{n+1} &\defeq \begin{cases}
+        \mmm_n + (x_n), & \text{if $1 \not\in \mmm_n + (x_n)$}, {\qquad\quad\quad\!\!}\\
+        \mmm_n, & \text{else.}
+      \end{cases}
+    \end{align*}}
+    \pause
+
+    \item \good{Yes}, if~$A$ is countable (irrespective of membership decidability):
+    {\footnotesize\begin{align*}
+      \mmm_0 &\defeq \{ 0 \}, &
+      \mmm_{n+1} &\defeq \mmm_n + (\underbrace{\{ x \in A \,|\, x = x_n \wedge
+        1 \not\in \mmm_n + (x_n) \}}_{\text{a certain subsingleton set}})
+    \end{align*}}
+    \vspace*{3.5em}
+    \visible<11>{{
+      \centering
+      \raisebox{0pt}[0pt][0pt]{
+        \hspace*{2.5em}
+        \scalebox{0.9}{\begin{tikzpicture}
+          \node (inner) at (17.3mm,-20mm) {\textit{``a bad joke''}};
+          \path (0,0) pic{laurel-wreath};
+        \end{tikzpicture}}
+        \hspace*{-3em}
+        \scalebox{0.9}{\begin{tikzpicture}
+          \node (inner) at (17.5mm,-18mm) {\vbox{\small\centering\textit{``non- \\informative''}}};
+          \path (0,0) pic{laurel-wreath};
+        \end{tikzpicture}}
+      }
+      \par
+    }}
+    \pause
+    \pause
+    \vspace*{-3.5em}
+
+    \item In the general case: \bad{No}\pause, but \good{yes} in a \emph{suitable forcing extension}\pause, and \\
+    \emph{first-order consequences} of its existence there \good{do hold} in
+    the base universe.
+  \end{enumerate}
+\end{frame}}
+
+% \begin{document}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[height=\paperheight]{sea-of-clouds-2}\end{minipage}}
+\begin{frame}{Infinite data}
+  \vspace*{-1em}
+  \[ \astikznodetransparentlycircled{xm}{7}\!,
+    \quad \astikznodetransparentlycircled{x0}{4}\!,
+    \quad \only<1-2>{\astikznodetransparentlycircled{t1}{3}}\only<3->{\astikznodecircled{t1}{mypurple}{3}}\!,
+    \quad \only<1>{\ldots}\pause \astikznodetransparentlycircled{x1}{1}\!,
+    \quad \only<2>{\ldots}\pause \astikznodecircled{t2}{mypurple}{8}\!,
+    \quad \only<3>{\ldots} \visible<4->{\astikznodetransparentlycircled{x2}{2}\!,}
+    \quad \only<4->{\ldots} \]
+  {\centering\begin{tikzpicture}[remember picture,overlay]
+    \node[draw=mypurple, circle, thick, inner sep=0.1em] (t3) {\scriptsize$\leq$};
+    \path[draw=mypurple,thick]
+      (t1)
+      to [out=-90, in=180] (t3)
+      to [out=0, in=-90] (t2);
+  \end{tikzpicture}\par}
+  \medskip
+  \pause
+
+  \textbf{Thm.} Every sequence~$\alpha : \NN \to \NN$ is \hil{good} in that
+  there exist~$i < j$ with~$\alpha(i) \leq \alpha(j)$.
+  \pause
+
+  \emph{Proof.} \emph{(offensive?)} By~\badbox{\textsc{lem}}, there is a
+  minimum~$\alpha(i)$.
+  Set~$j \defeq i + 1$. \qed\par
+  \pause
+  \medskip
+
+  \textbf{Def.} A preorder~$X$ is \hil{well\only<9->{$^\star$}} iff every sequence~$\NN \to X$ is good.
+
+  \textbf{Examples.} $(\NN,{\leq}),\ \
+  \color{white}\only<7->{\color{red!90}}\astikznode{onlyclass}{$\underbrace{\color{black}X \times Y,\ \ X^*,\ \ \mathrm{Tree}(X)}_{\text{\visible<7->{\bad{only classically}}}}$}$.
+  \pause
+  \pause
+  \medskip
+
+  \begin{tikzpicture}[remember picture,overlay]
+    \node[thick, fill=black, rectangle, inner sep=0.3em, right=2em of onlyclass] (moral) {
+      \begin{minipage}{6cm}
+        \begin{columns}
+          \begin{column}{0.15\textwidth}
+            \hspace*{1.0em}\color{white}\dbend
+          \end{column}
+          \begin{column}{0.95\textwidth}
+            \color{white}\footnotesize
+            \it Don't quantify over points of spaces which might not have enough.
+          \end{column}
+        \end{columns}
+      \end{minipage}
+    };
+    \path[draw=red!90,thick,-stealth]
+      (moral) to
+      [out=180, in=-00] (onlyclass.east);
+  \end{tikzpicture}
+  \pause
+
+  \textbf{Def.} For a predicate~$P$ on finite lists over a set~$X$, inductively
+  define:
+  \[
+    \infer{P \,|\, \sigma}{P\sigma}
+    \qquad
+    \infer{P \,|\, \sigma}{\forall(x \in X)\_\ P \,|\, \sigma x}
+  \]
+
+  \textbf{Def.} A preorder is \hil{well} iff~$\mathsf{Good} \,|\, [\,]$, where
+  $\mathsf{Good}\,[\sigma_1,\ldots,\sigma_n] \defeqv (\exists(i < j)\_ \sigma_i \leq \sigma_j)$.
+\end{frame}}
 

### 2023-09-11 slides-herrsching2023.tex
Slides for first talk finished

@@
 
-  Then for every parameter value~$t$, classically there is
+  Then for every parameter value~$t \in \Omega$, classically there is
 
@@
       \only<2->{\hil{Yes.}}} \\
-      How about the~$v_i$? \only<2->{\hil{No.}}
+      How about the~$v_i$? \only<2->{\hil{No}\only<3->{, but \hil{yes} on a dense
+      open subset of~$\Omega$.}}
     \end{column}
@@
 
+\section{What can forcing do for you?}
+
 \begin{frame}{What can forcing do for you?}
@@
       \begin{block}{1. Explore foundational possibility}
-        There are forcing extensions with~\textsc{ch}, $\neg$\textsc{ch},
-        \textsc{lem}, $\neg$\textsc{lem}, \ldots
+        There are forcing extensions with \\
+        \textsc{ch}, $\neg$\textsc{ch}, \textsc{lem}, $\neg$\textsc{lem}, \ldots
       \end{block}
@@
 
-  \textbf{Def.} For a predicate~$P$ on finite lists over a set~$X$, inductively
-  define:
-  \[
-    \infer{P \,|\, \sigma}{P\sigma}
-    \qquad
-    \infer{P \,|\, \sigma}{\forall(x \in X)\_\ P \,|\, \sigma x}
-  \]
+  \textbf{Def.} A preorder is \hil{well} iff any of the following equivalent
+  conditions hold:
+  \begin{enumerate}
+    \item The \hil{generic sequence} $\NN \to X$ is good.
+    \pause
+    \item Every sequence $\NN \to X$ \hil{in every forcing extension} is good.
+    \pause
+    \item There is a \hil{well-founded tree} witnessing universal goodness.
+  \end{enumerate}
 
-  \textbf{Def.} A preorder is \hil{well} iff~$\mathsf{Good} \,|\, [\,]$, where
-  $\mathsf{Good}\,[\sigma_1,\ldots,\sigma_n] \defeqv (\exists(i < j)\_ \sigma_i \leq \sigma_j)$.
+%  $\mathsf{Good}\,[\sigma_1,\ldots,\sigma_n] \defeqv (\exists(i < j)\_ \sigma_i \leq \sigma_j)$.
+%  \textbf{Def.} For a predicate~$P$ on finite lists over a set~$X$, inductively
+%  define:
+%  \[
+%    \infer{P \,|\, \sigma}{P\sigma}
+%    \qquad
+%    \infer{P \,|\, \sigma}{\forall(x \in X)\_\ P \,|\, \sigma x}
+%  \]
+%
+%  \textbf{Def.} A preorder is \hil{well} iff~$\mathsf{Good} \,|\, [\,]$, where
+%  $\mathsf{Good}\,[\sigma_1,\ldots,\sigma_n] \defeqv (\exists(i < j)\_ \sigma_i \leq \sigma_j)$.
 \end{frame}}
+
+\section{Basics of forcing}
+
+\begin{frame}{Forcing notions}
+  \small
+  \textbf{Def.} A \hil{forcing notion} consists of
+  a preorder~$L$ of \hil{forcing conditions}, and
+  for every~$\sigma \in L$, a set~$\Cov(\sigma) \subseteq
+  P({\downarrow}\sigma)$ of \hil{coverings} of~$\sigma$
+  such that: If~$\tau \preccurlyeq \sigma$ and~$R \in \Cov(\sigma)$, there
+  should be a covering~$S \in \Cov(\tau)$ such that~$S \subseteq
+  {\downarrow}R$.
+  \bigskip
+
+  {\centering\footnotesize\begin{tabular}{llll}
+    \toprule
+    & preorder~$L$ & coverings of an element~$\sigma \in L$ & filters of~$L$ \\
+    \midrule
+    \normalnumber{1} & $X^*$ & $\{ \sigma x \,|\, x \in X \}$ & maps~$\NN \to X$ \\
+    \normalnumber{2} & $X^*$ & $\{ \sigma x \,|\, x \in X \}$,\ \ $\{ \sigma\tau \,|\, \tau \in X^*, a \in \sigma\tau \}$ for each~$a \in X$ & surjections~$\NN \twoheadrightarrow X$ \\
+    \normalnumber{3} & f.g. ideals & --- & ideals \\
+    \normalnumber{4} & f.g. ideals & $\{ \sigma+(a), \sigma+(b) \}$ for each~$ab \in \sigma$,\ \ $\{\}$ if~$1 \in \sigma$ & prime ideals \\
+    \normalnumber{5} & opens & $\mathcal{U}$ such that~$\sigma = \bigcup \mathcal{U}$ & points \\
+    \normalnumber{6} & $\{\star\}$ & $\{ \star \,|\, \varphi \} \cup \{ \star \,|\, \neg\varphi \}$ &
+    witnesses of~\textsc{lem}
+    \\
+    \bottomrule
+  \end{tabular}\par}
+  \bigskip
+
+  \textbf{Def.} A \emph{filter} of a forcing notion~$(L,\mathrm{Cov})$
+  is a subset~$F \subseteq L$ such that
+  \vspace*{-0.4em}
+  \begin{enumerate}
+    \scriptsize
+    \item $F$ is upward-closed: if~$\tau \preccurlyeq \sigma$ and if~$\tau \in F$, then~$\sigma \in F$; \\[-3.0em]
+    \item $F$ is downward-directed: $F$ is inhabited, and if~$\alpha,\beta \in F$,
+    then there is a common refinement~$\sigma \preccurlyeq \alpha,\beta$ such
+    that~$\sigma \in F$; and \\[-2.0em]
+    \item $F$ splits the covering system: if~$\sigma \in F$ and~$R \in
+    \Cov(\sigma)$, then~$\tau \in F$ for some~$\tau \in R$.
+  \end{enumerate}
+
+\end{frame}
 
@@
 set theory, topos theory, category theory, sheaves.
+
+Forschungsfrage: Klassische Basen sind ja schon ziemlich maximal in dem Sinn,
+als dass etwa der Raum der Funktionen ℕ → ℕ genügend Punkte hat. Wie wäre es,
+in einer Welt zu arbeiten, in der auch der Raum der Surjektionen ℕ → X genügend
+Punkte hat?

### 2023-09-11 slides-herrsching2023.tex
One more example
Black background for better look on the projector
Ingredients for forcing

@@
 \newcommand{\aaa}{\mathfrak{a}}
+\newcommand{\bbb}{\mathfrak{b}}
 \newcommand{\ccc}{\mathfrak{c}}
@@
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\includegraphics[height=\paperheight]{multiverse-faded}\end{minipage}}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\includegraphics[height=\paperheight]{multiverse-faded-169}\end{minipage}}
 \begin{frame}[c]
@@
   \end{enumerate}
+\end{frame}
+
+\begin{frame}{Maximal ideals}
+  \textbf{Thm.}
+  Let~$M$ be a surjective matrix with more rows than columns over a
+  ring~$A$. Then~$1 = 0$ in~$A$.
+
+  \emph{Proof.} (special case) Write~$M = \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right)$. By surjectivity,
+  have~$u, v$ with
+  \[
+    u \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}1\\0\end{smallmatrix}\right)
+    \quad\text{and}\quad
+    v \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}0\\1\end{smallmatrix}\right).
+  \]
+  Hence
+  $
+    1 = (vy) (ux) = (uy) (vx) = 0
+  $. \qed
 \end{frame}}
-
-% \begin{document}
 
@@
 
+% \begin{document}
+
+\begin{frame}{Ingredients for forcing}
+  To construct a forcing extension, we require:
+  \begin{enumerate}
+    \item a base universe~$V$
+    \item a preorder~$L$ of \hil{forcing conditions} in~$V$\!,
+    pictured as \hil{finite approximations}
+
+    (\emph{convention:} $\tau \preccurlyeq \sigma$ means that~$\tau$ is a
+    better finite approximation than~$\sigma$)
+    \item a \hil{covering system} governing how finite approximations evolve to
+    better ones
+
+    (for each~$\sigma \in L$, a set~$\Cov(\sigma) \subseteq
+    P({\downarrow}\sigma)$, with a simulation condition)
+  \end{enumerate}
+  In the forcing extension, there will then be a \hil{generic filter} (ideal
+  object).
+  \pause
+
+  \vspace*{-1em}
+  \begin{columns}
+    \begin{column}{0.49\textwidth}
+      \small
+      \begin{block}{For the generic surjection~$\NN \twoheadrightarrow X$}
+        \justifying
+        Use \hil{finite lists}~$\sigma \in X^*$ as forcing conditions,
+        where $\tau \preccurlyeq \sigma$ iff~$\sigma$ is an initial segment of~$\tau$,
+        and be prepared to grow~$\sigma$ to \ldots
+        \footnotesize
+        \begin{enumerate}
+          \item[(a)] one of~$\{ \sigma x \,|\, x \in X \}$, to make~$\sigma$ more defined
+          \item[(b)] one of~$\{ \sigma \tau \,|\, \tau \in X^*, a \in
+          \sigma\tau \}$, for any~$a \in X$, to make~$\sigma$ more surjective
+        \end{enumerate}
+      \end{block}
+    \end{column}
+    \pause
+
+    \begin{column}{0.45\textwidth}
+      \small
+      \begin{block}{For the generic prime ideal of a ring~$A$}
+        \justifying
+        Use \hil{f.g.\@ ideals} as forcing conditions, where $\bbb \preccurlyeq
+        \aaa$ iff~$\bbb \supseteq \aaa$, and be prepared to grow~$\aaa$ to \ldots
+        \footnotesize
+        \begin{enumerate}
+          \item[(a)] one of~$\emptyset$, if~$1 \in \aaa$, to make~$\aaa$ more proper
+          \item[(b)] one of~$\{ \aaa+(x), \aaa+(y) \}$, if~$xy \in \aaa$, to
+          make~$\aaa$ more prime
+        \end{enumerate}
+      \end{block}
+    \end{column}
+  \end{columns}
+\end{frame}
+
 \begin{frame}{Forcing notions}
@@
   \end{enumerate}
-
 \end{frame}
@@
 
+Dominique Duval
+
 Forschungsfrage: Klassische Basen sind ja schon ziemlich maximal in dem Sinn,

### 2023-09-11 slides-herrsching2023.tex
Eventually monad and proof translations

@@
 \setbeamersize{text margin left=1.60em,text margin right=1.60em}
+
+\newlength\stextwidth
+\newcommand\makesamewidth[3][c]{%
+  \settowidth{\stextwidth}{#2}%
+  \makebox[\stextwidth][#1]{#3}%
+}
 
@@
   \end{enumerate}
-  In the forcing extension, there will then be a \hil{generic filter} (ideal
+  In the forcing extension~$V^\nabla$, there will then be a \hil{generic filter} (ideal
   object).
@@
 
-\begin{frame}{Forcing notions}
+\begin{frame}{The eventually monad}
+  Let~$L$ be a forcing notion.
+
+  Let~$P$ be a monotone predicate on~$L$
+  (if $\tau \preccurlyeq \sigma$, then $P\sigma \Rightarrow P\tau$). \\
+  For instance, in the case~$L = X^*$:
+  \begin{itemize}
+    \item $\mathsf{Repeats}\, x_0\ldots x_{n-1} \defeqv \exists i\_ \exists j\_ i < j \wedge x_i = x_j$
+    \item $\mathsf{Good}\, \,\,\,\,\,\,x_0\ldots x_{n-1} \defeqv \exists i\_ \exists j\_ i < j \wedge x_i \leq x_j$
+  \end{itemize}
+
+  We then define~``\hil{$P \mid \sigma$}'' (``$P$ bars~$\sigma$'') inductively by the clauses:
+  \begin{enumerate}
+    \item If~$P\sigma$, then~$P \mid \sigma$.
+    \item If~$P \mid \tau$ for all~$\tau \in R$, where~$R$ is some covering
+    of~$\sigma$, then~$P \mid \sigma$.
+  \end{enumerate}
+  So~$P \mid \sigma$ expresses in a direct inductive fashion that no matter
+  how~$\sigma$ evolves to a better approximation~$\tau$, eventually~$P\tau$
+  will hold.
+
+  We use quantifier-like notation: ``$\nabla(\tau \preccurlyeq \sigma)\_
+  P\tau$'' means ``$P \mid \sigma$''.
+\end{frame}
+% BOARD:
+% - examples for P | σ
+% - abuse of notation
+
+\begin{frame}{Proof translations}
+  \textbf{Thm.} Every~\textsc{iqc}-proof remains correct, with at most a
+  polynomial increase in length, if throughout we
+  replace
+  \[\begin{array}{rcl@{\quad\text{where}\quad}rcl}
+    \exists & \leadsto & \exists_\mathrm{class},
+    & \exists_\mathrm{class} &\defeqv& \neg\neg\exists, \\
+    \vee & \leadsto & \vee_\mathrm{class},
+    & \alpha \vee_\mathrm{class} \beta &\defeqv& \neg\neg(\alpha \vee \beta), \\
+    = & \leadsto & =_\mathrm{class},
+    & s =_\mathrm{class} t &\defeqv& \neg\neg(s = t).
+  \end{array} \]
+  \pause
+
+  \begin{columns}[c]
+    \begin{column}{0.01\textwidth}
+      \includegraphics[height=2.4em]{sheafification-man-2}
+    \end{column}
+    \quad
+    \begin{column}{0.9\textwidth}
+      \hil{When we say:}\ \ some statement ``holds in~$V^{\neg\neg}$'', \\
+      \makesamewidth[l]{\hil{When we say:}}{\hil{we mean:}}\ \ its translation holds in~$V$.
+    \end{column}
+  \end{columns}
+  \bigskip
+
+  Similarly for arbitrary forcing extensions~$V^\nabla$, ``just with~$\nabla$
+  instead of~$\neg\neg$''.
+  \bigskip
+
+  \pause
+  \textbf{Ex.} As~$\neg\neg(\varphi \vee \neg\varphi)$ is a theorem
+  of~\textsc{iqc}, the law of excluded middle holds in~$V^{\neg\neg}$.
+\end{frame}
+
+\begin{frame}{Formalities}
   \small
@@
 Punkte hat?
+
+Fun:
+- Modale Dinge

### 2023-09-12 slides-herrsching2019.tex
Gelfand duality and several smaller items
typo

@@
     Olivia Caramello's paper
-    \fixedhref{https://arxiv.org/abs/1301.0300}{Topological galois theory} is
+    \fixedhref{https://arxiv.org/abs/1301.0300}{Topological Galois theory} is
     also relevant.
@@
       \item one sort: $R$
-      \item five constant symbols: $0$ and $1$ (nullary), $-$ (unary), $+$ and
+      \item five function symbols: $0$ and $1$ (nullary), $-$ (unary), $+$ and
       $\cdot$ (binary)

### 2023-09-12 slides-herrsching2023.tex
to do
The ∇-translation

@@
 \newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates
+\newcommand{\forces}{\vDash}
 
@@
     \item $\mathsf{Good}\, \,\,\,\,\,\,x_0\ldots x_{n-1} \defeqv \exists i\_ \exists j\_ i < j \wedge x_i \leq x_j$
+    \quad (for some preorder~$\leq$ on~$X$)
   \end{itemize}
+  \pause
 
@@
   \end{enumerate}
-  So~$P \mid \sigma$ expresses in a direct inductive fashion that no matter
+  So~$P \mid \sigma$ expresses in a \hil{direct inductive fashion}:
+  \[ \text{``No matter
   how~$\sigma$ evolves to a better approximation~$\tau$, eventually~$P\tau$
-  will hold.
+  will hold.''} \]
+  \pause
 
@@
   of~\textsc{iqc}, the law of excluded middle holds in~$V^{\neg\neg}$.
+\end{frame}
+
+\newcommand{\defeqvi}{\quad iff\quad}
+\begin{frame}{The $\nabla$-translation}
+  \small
+  For bounded first-order formulas over the (large) first-order signature which has
+  \begin{enumerate}
+    \scriptsize
+    \item one sort~$\underline{X}$ for each set~$X$ in the base universe,
+    \\[-1.2em]
+    \item one~$n$-ary function symbol~$\underline{f} : \underline{X_1} \times
+    \cdots \times \underline{X_n} \to \underline{Y}$ for each map~$f : X_1 \times
+    \cdots \times X_n \to Y$,
+    \\[-1.2em]
+    \item one~$n$-ary relation symbol~$\underline{R} \hookrightarrow
+    \underline{X_1} \times \cdots \times \underline{X_n}$ for each relation~$R
+    \subseteq X_1 \times \cdots \times X_n$, and
+    \\[-1.2em]
+    \item an additional unary relation symbol~$G \hookrightarrow \underline{L}$
+    (for the \emph{generic filter} of~$L$),
+  \end{enumerate}
+  we recursively define:
+  \scriptsize
+  \begin{tabbing}
+    \quad \= $\sigma \forces \forces \forall(x\?\underline{X})\_ \varphi$ \=
+    \defeqvi $\textcolor{gray}{\forall(\tau \preccurlyeq \sigma)\_}\
+    \forall(x_0 \in X)\_ \tau \forces \varphi[\underline{x_0}/x]$.\qquad\quad \=
+    $\sigma \forces \exists(x\?\underline{X})\_ \varphi$
+    \= $\sigma \forces \underline{R}(\underline{s_1},\ldots,\underline{s_n})$ \= \defeqvi $s = t$. \= \kill
+
+    \> $\sigma \forces s = t$
+    \> \defeqvi $\nabla \sigma\_ \llbracket s \rrbracket = \llbracket t \rrbracket$.
+    \> $\sigma \forces \underline{R}(s_1,\ldots,s_n)$
+    \> \defeqvi $\nabla\sigma\_ R(\llbracket s_1 \rrbracket,\ldots,\llbracket s_n \rrbracket)$. \\[0.3em]
+
+    \> $\sigma \forces \varphi \Rightarrow \psi$
+    \> \defeqvi $\textcolor{gray}{\forall(\tau \preccurlyeq \sigma)\_}\ (\tau \forces \varphi) \Rightarrow
+    (\tau \forces \psi)$.
+    \> $\sigma \forces G\tau$
+    \> \defeqvi $\nabla\sigma\_ \sigma \preccurlyeq \llbracket\tau\rrbracket$. \\[0.3em]
+
+    \> $\sigma \forces \top$ \> \defeqvi $\top$.
+    \> $\sigma \forces \bot$ \> \defeqvi $\hil{$\nabla\sigma\_$}\ \bot$ \\[0.3em]
+
+    \> $\sigma \forces \varphi \wedge \psi$
+    \> \defeqvi $(\sigma \forces \varphi) \wedge (\sigma \forces \psi)$.
+    \> $\sigma \forces \varphi \vee \psi$
+    \> \defeqvi $\hil{$\nabla\sigma\_$}\ (\sigma \forces \varphi) \vee (\sigma \forces \psi)$. \\[0.3em]
+
+    \> $\sigma \forces \forall(x\?\underline{X})\_ \varphi$
+    \> \defeqvi $\textcolor{gray}{\forall(\tau \preccurlyeq \sigma)\_}\ \forall(x_0 \in X)\_ \tau \forces
+    \varphi[\underline{x_0}/x]$.
+    \> $\sigma \forces \exists(x\?\underline{X})\_ \varphi$
+    \> \defeqvi $\hil{$\nabla\sigma\_$}\ \exists(x_0 \in X)\_ \sigma \forces \varphi[\underline{x_0}/x]$.
+  \end{tabbing}
+  \small
+  Finally, we say that~$\varphi$ ``holds in~$V^\nabla$'' iff for all~$\sigma
+  \in L$, $\sigma \forces \varphi$.
 \end{frame}
@@
 - Modale Dinge
+- structure sheaf A~, field property etc. (like in redcom slides), generic freeness
+- nongeometric surprises

### 2023-09-14 slides-herrsching2023.tex
Instructive(?) examples for KJ

@@
 
-\title{New modal operators for constructive mathematics}
+\title{Constructive forcing}
 
 \author{Ingo Blechschmidt}
-\date{May 2nd, 2023}
+\date{September 20th to September 16th, 2023}
 
@@
 
-% \begin{document}
-
 {\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
@@
   \emph{Proof.} (special case) Write~$M = \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right)$. By surjectivity,
-  have~$u, v$ with
+  have~$u, v \in A$ with
   \[
@@
 
-% \begin{document}
-
 \begin{frame}{Ingredients for forcing}
@@
 
-  We then define~``\hil{$P \mid \sigma$}'' (``$P$ bars~$\sigma$'') inductively by the clauses:
+  We then define~``\hil{$P \mid \sigma$}'' (``$P$ bars~$\sigma$'') inductively by the following clauses:
   \begin{enumerate}
@@
   will hold.''} \]
-  \pause
-
+
+  \pause
   We use quantifier-like notation: ``$\nabla(\tau \preccurlyeq \sigma)\_
@@
 
+% \begin{document}
+
 \newcommand{\defeqvi}{\quad iff\quad}
@@
   \small
-  For bounded first-order formulas over the (large) first-order signature which has
+  \only<1>{For bounded first-order formulas over the (large) first-order signature which has
   \begin{enumerate}
@@
   \end{enumerate}
-  we recursively define:
+  we recursively define:}
   \scriptsize
@@
   \small
-  Finally, we say that~$\varphi$ ``holds in~$V^\nabla$'' iff for all~$\sigma
-  \in L$, $\sigma \forces \varphi$.
+  \only<1>{Finally, we say that~$\varphi$ ``holds in~$V^\nabla$'' iff for all~$\sigma
+  \in L$, $\sigma \forces \varphi$.}
+
+  \footnotesize
+  \begin{tabular}{@{}lp{0.27\textwidth}p{0.48\textwidth}}
+    \toprule
+    forcing notion & statement about~$V^\nabla$ & external meaning \\
+    \midrule
+    surjection $\NN \twoheadrightarrow X$ &
+    ``the gen.\@ surj.\@ is surjective'' &
+    $\forall(a{\in}X)\_ \forall(\sigma{\in}X^*)\_ \nabla(\tau{\preccurlyeq}\sigma)\_ \exists(n{\in}\NN)\_ \tau[n] = a$. \\[0.5em]
+    \pause
+
+    map $\NN \to X$ &
+    ``the gen.\@ sequence is good'' &
+    $\mathsf{Good} \mid [\,]$. \\[0.5em]
+
+    frame of opens &
+    ``every complex number has a square root'' &
+    For every open~$U \subseteq X$ and every cont.\@
+    function $f : U \to \CC$, there is an open covering $U = \bigcup_i U_i$ such
+    that for each index~$i$, there is a cont.\@ function $g : U_i \to \CC$
+    such that~$g^2 = f$. \\[4.5em]
+
+    Zariski &
+    ``$x \neq 0 \Rightarrow \text{$x$ inv.}$'' &
+    if the only f.p.\@ $k$-algebra in which~$x = 0$ is the zero algebra,
+    then~$x$ is invertible in~$k$
+  \end{tabular}
 \end{frame}
@@
 - nongeometric surprises
+
+\newcommand{\Good}{\mathsf{Good}}
+\newcommand{\vinfer}[2]{\left.#1\quad\middle|\quad\begin{array}{l}#2\end{array}\right.}
+\begin{frame}[plain]
+  \scriptsize
+  \[
+    \renewcommand{\arraystretch}{1.3}
+    \vinfer{\Good \mid [\,]}{
+      \vinfer{\Good \mid 0}{
+        \vinfer{\Good \mid 00}{\Good\ 00} \\[1em]
+        \vinfer{\Good \mid 01}{\Good\ 01} \\[1em]
+        \vinfer{\Good \mid 02}{\Good\ 02} \\
+        \vdots
+      } \\[1em]
+      \ \\[0.5em]
+      \vinfer{\Good \mid 1}{
+        \vinfer{\Good \mid 10}{
+          \vinfer{\Good \mid 100}{\Good\ 100} \\[1em]
+          \vinfer{\Good \mid 101}{\Good\ 101} \\[1em]
+          \vinfer{\Good \mid 102}{\Good\ 102} \\
+          \vdots
+        } \\[1em]
+        \vinfer{\Good \mid 11}{\Good\ 11} \\[1em]
+        \vinfer{\Good \mid 12}{\Good\ 12} \\[1em]
+        \vdots
+      } \\[1em]
+      \ \\[0.5em]
+      \vinfer{\Good \mid 2}{} \\[1em]
+      \vdots
+    }
+  \]
+\end{frame}

### 2023-09-14 algar-lecture-notes.tex
typo

@@
 The base step~$a = 0$ is trivial, since if~$a = 0$ then also~$a = 0 \vee a \neq
-0$ (we can always weaken statements by addition additional disjuncts).
+0$ (we can always weaken statements by adding additional disjuncts).
 
@@
 \to \mathsf{Bool}$, computing~$\mathsf{true}$ or~$\mathsf{false}$ depending on
-whether its input agree or are distinct. Because the presented proof proceeded
+whether its two inputs agree or are distinct. Because the presented proof proceeded
 by induction, the resulting program will proceed by recursion.
@@
 
-An example of a result of type~(b) us the basic observation that every
+An example of a result of type~(b) uses the basic observation that every
 inhabited set of natural numbers contains a minimal element. This result is not
@@
 \begin{defn}A set~$X$ is \emph{inhabited} if and only if there exists an
-element of~$X$. (Constructively, this is stronger than~$X$ not being
-empty.)\end{defn}
+element of~$X$.\end{defn}
+
+Constructively, the condition that a set~$X$ is inhabited is stronger than~$X$
+merely not being empty. The latter is equivalent to the statement that it is
+\notnot the case that~$X$ is inhabited.
 
@@
 elucidate proof mining for constructive content. The author learned this
-material from diverse texts and slides of Thierry Coquand; the
+material mostly from diverse texts and slides of Thierry Coquand; the
 reference~\cite{coquand:classical} is a good start.
@@
 axiom of choice ever help in proving arithmetical statements, those first-order
-statements in which all quantifiers range over the natural numbers? Well, it
+statements in which all quantifiers range over the natural numbers? Well, it surely
 might. But a result of Gödel states that~\textsc{zfc} (Zermelo--Fraenkel set
@@
 elimination procedure is facilitated by the \emph{double-negation translation}
-reviewed below. In some cases, a refined translation even allows to preserve the asserted
+reviewed below. In some cases, a refined translation even allows us to preserve the asserted
 claim exactly; this technique is variously known as \emph{Friedman's trick},

### 2023-09-14 slides-luminy2023.tex
typo

@@
     -- \textsc{lem} holds \emph{somewhere} and \\
-    -- \textsc{dc} holds \emph{proximally}.
+    -- instances of \textsc{dc} holds \emph{proximally}.
   \end{enumerate}

### 2023-09-14 slides-herrsching2023.tex
typo

@@
   \[\begin{array}{rcl@{\quad\text{where}\quad}rcl}
-    \exists & \leadsto & \exists_\mathrm{class},
-    & \exists_\mathrm{class} &\defeqv& \neg\neg\exists, \\
-    \vee & \leadsto & \vee_\mathrm{class},
-    & \alpha \vee_\mathrm{class} \beta &\defeqv& \neg\neg(\alpha \vee \beta), \\
-    = & \leadsto & =_\mathrm{class},
-    & s =_\mathrm{class} t &\defeqv& \neg\neg(s = t).
+    \exists & \leadsto & \exists^\mathrm{cl},
+    & \exists^\mathrm{cl} &\defeqv& \neg\neg\exists, \\
+    \vee & \leadsto & \vee^\mathrm{cl},
+    & \alpha \vee^\mathrm{cl} \beta &\defeqv& \neg\neg(\alpha \vee \beta), \\
+    = & \leadsto & =^\mathrm{cl},
+    & s =^\mathrm{cl} t &\defeqv& \neg\neg(s = t).
   \end{array} \]
@@
   \footnotesize
-  \begin{tabular}{@{}lp{0.27\textwidth}p{0.48\textwidth}}
+  \begin{tabular}{@{}lp{0.27\textwidth}p{0.48\textwidth}@{}}
     \toprule
@@
     that for each index~$i$, there is a cont.\@ function $g : U_i \to \CC$
-    such that~$g^2 = f$. \\[4.5em]
-
-    Zariski &
+    such that~$g^2 = f$. \\[4.3em]
+
+    big Zariski &
     ``$x \neq 0 \Rightarrow \text{$x$ inv.}$'' &
-    if the only f.p.\@ $k$-algebra in which~$x = 0$ is the zero algebra,
-    then~$x$ is invertible in~$k$
+    If the only f.p.\@ $k$-algebra in which~$x = 0$ is the zero algebra,
+    then~$x$ is invertible in~$k$.
   \end{tabular}
@@
 - nongeometric surprises
+
+außerdem: Konservativität.
 

### 2023-09-14 slides-herrsching2023.tex
Examples and outlook

@@
 \documentclass[12pt,utf8,notheorems,compress,t,aspectratio=169]{beamer}
-%\documentclass[12pt,utf8,notheorems,compress,t]{beamer}
 \usepackage{etex}
@@
 \usepackage[normalem]{ulem}
-\usepackage{graphbox}
 \usepackage{mathtools}
@@
 \usepackage{proof}
-\usepackage{ifthen}
-\usepackage[normalem]{ulem}
 \usepackage[all]{xy}
@@
 \setbeamercolor{structure}{fg=mypurple}
-\definecolor{myred}{RGB}{150,0,0}
-%\setbeamercolor*{title}{bg=myred,fg=white}
-%\setbeamercolor*{titlelike}{bg=myred,fg=white}
 \setbeamercolor*{title}{bg=mypurple,fg=white}
@@
 \newcommand{\defeqv}{\vcentcolon\equiv}
-\newcommand{\Sh}{\mathrm{Sh}}
-\newcommand{\GL}{\mathrm{GL}}
-\newcommand{\Zar}{\mathrm{Zar}}
-\newcommand{\op}{\mathrm{op}}
-\newcommand{\Set}{\mathrm{Set}}
-\newcommand{\Eff}{\mathrm{Ef{}f}}
-\newcommand{\Sch}{\mathrm{Sch}}
-\newcommand{\Aff}{\mathrm{Aff}}
-\newcommand{\Ring}{\mathrm{Ring}}
 \newcommand{\Cov}{\mathrm{Cov}}
-\newcommand{\LocRing}{\mathrm{LocRing}}
-\newcommand{\LRS}{\mathrm{LRS}}
-\newcommand{\Hom}{\mathrm{Hom}}
-\newcommand{\Spec}{\mathrm{Spec}}
-\newcommand{\lra}{\longrightarrow}
-\newcommand{\RelSpec}{\operatorname{Spec}}
 \renewcommand{\_}{\mathpunct{.}}
@@
 \newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
-\newcommand{\ul}[1]{\underline{#1}}
-\newcommand{\affl}{\ensuremath{{\ul{\ensuremath{\AA}}^1}}}
-\newcommand{\Ll}{\text{iff}}
 \newcommand{\inv}{inv.\@}
-\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}
-\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates
 \newcommand{\forces}{\vDash}
@@
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses}\end{minipage}}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.35cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
 \begin{frame}{Constructive forcing}
@@
 
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{-1cm}\includegraphics[width=\paperwidth]{forest-light-colored}\end{minipage}}
 \begin{frame}{The eventually monad}
@@
   P\tau$'' means ``$P \mid \sigma$''.
-\end{frame}
+\end{frame}}
 % BOARD:
@@
   \scriptsize
+  \only<2->{\vspace*{-0.4em}}
   \begin{tabbing}
@@
     ``the gen.\@ surj.\@ is surjective'' &
-    $\forall(a{\in}X)\_ \forall(\sigma{\in}X^*)\_ \nabla(\tau{\preccurlyeq}\sigma)\_ \exists(n{\in}\NN)\_ \tau[n] = a$. \\[0.5em]
+    $\forall(a{\in}X)\_ \forall(\sigma{\in}X^*)\_ \nabla(\tau{\preccurlyeq}\sigma)\_ \exists(n{\in}\NN)\_ \tau[n] = a$. \\[0.4em]
     \pause
@@
     ``the gen.\@ sequence is good'' &
-    $\mathsf{Good} \mid [\,]$. \\[0.5em]
+    $\mathsf{Good} \mid [\,]$. \\[0.4em]
 
@@
     If the only f.p.\@ $k$-algebra in which~$x = 0$ is the zero algebra,
-    then~$x$ is invertible in~$k$.
+    then~$x$ is invertible in~$k$.\\[1.5em]
+    \pause
+
+    little Zariski &
+    ``every f.g. vector space does \emph{not not} have a basis'' &
+    \makesamewidth[l]{}{\phantom{x}}\hil{Grothendieck's generic freeness lemma}
   \end{tabular}
+\end{frame}
+
+\begin{frame}{Outlook}
+  \begin{block}{Passing to and from extensions}
+    \justifying\small
+    \textbf{Thm.} Let~$\varphi$ be a \hil{bounded first-order formula} not
+    mentioning~$G$. In each of the following situations, we have that
+    $\varphi$ holds in~$V^\nabla$ iff~$\varphi$
+    holds in $V$:
+
+    \vspace*{-0.5em}
+    \begin{enumerate}
+      \item $L$ and all coverings are inhabited (proximality). \\[-1em]
+      \item $L$ contains a top element, every covering of the
+      top element is inhabited, and~$\varphi$ is a coherent implication
+      (positivity).
+    \end{enumerate}
+  \end{block}
+
+  \vspace*{-1em}
+  \begin{columns}
+    \begin{column}{0.46\textwidth}
+      \begin{block}{The mystery of nongeometric sequents}
+        \justifying
+        The \hil{generic ideal} of a ring is maximal:
+        \vspace*{-1em}
+        \[ (x \in \aaa \Rightarrow 1 \in \aaa) \Longrightarrow 1 \in \aaa + (x). \]
+
+        The \hil{generic ring} is a field:
+        \vspace*{-0.7em}
+        \[ (x = 0 \Rightarrow 1 = 0) \Longrightarrow (\exists y\_ xy = 1). \]
+      \end{block}
+
+    \end{column}
+
+    \begin{column}{0.50\textwidth}
+      \begin{block}{Traveling the multiverse \ldots}
+        \textsc{lem} is a \hil{switch} and \hil{holds positively};
+        being countable is a \hil{button}.
+        \medskip
+
+        Every instance of \textsc{dc} \hil{holds proximally}.
+        \medskip
+
+        A geometric implication is provable iff it holds \hil{everywhere}.
+      \end{block}
+      \vspace*{-0.3em}
+      \hfill\footnotesize\ldots{} upwards, but always keeping ties to the
+      base.{\ }
+    \end{column}
+  \end{columns}
 \end{frame}

### 2023-09-15 slides-herrsching2023.tex
Beginning of notes slides

@@
 \date{September 20th to September 16th, 2023}
+
+\setbeameroption{show notes on second screen=bottom}
+\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\footnotesize\justifying#2\par}}}
 
@@
 \begin{frame}[c]
+  \jnote{1}{These are the slides for a three-part lecture series for the autumn school
+  \fixedhref{https://www.mathematik.uni-muenchen.de/~schwicht/pc23.php}{\emph{Proof and Computation}}
+  organized by Klaus Mainzer, Peter Schuster and Helmut Schwichtenberg, held in
+  Herrsching from September 20th to September 16th, 2023.}
+
   \centering
@@
   \bigskip
-
-%  \colorbox{black!50}{Autumn school on} \\
-%  \colorbox{black!50}{\emph{Proof and Computation}} \\
-%  \colorbox{black!50}{in Herrsching} \\
-%  \ \\
-%  \colorbox{black!50}{September 20th to September 16th, 2023}
-%
-%  \colorbox{black!50}{Ingo Blechschmidt} \\
-%  \par
 \end{frame}}
@@
 \begin{frame}{}
+  \jnote{3}{
+    The slide presents a question on calculus. Constructive forcing
+    establishes a connection with constructive linear algebra; we will understand
+    that:
+    \begin{enumerate}\justifying
+      \item The eigenvalues (defined as the zeros of the characteristic
+      polynomial) locally depend continuously on the parameter \hil{because}
+      it is a theorem of constructive linear algebra that symmetric matrices have
+      a full list of eigenvalues.
+
+      \item The eigenvectors can \hil{not} be expected to locally depend
+      continuously \hil{because} it is \hil{not} a theorem of constructive
+      linear algebra that symmetric matrices have an eigenvector basis.
+
+      \item There is a dense open subset of the parameter space which restores
+      continuous dependence \hil{because} it \hil{is} a theorem of constructive
+      linear algebra that every symmetric matrix does \emph{not~not} have an
+      eigenvector basis.
+    \end{enumerate}
+
+    A simple example where the eigenvectors cannot be chosen to locally depend
+    continuously on the parameter is
+    \fixedhref{https://mathoverflow.net/a/60563/31233}{recorded here}. Removing
+    the origin there yields a suitable dense open subset.
+  }
+
   \begin{center}\includegraphics[width=0.2\textwidth]{eigenvector}\end{center}
@@
 \begin{frame}{A brief timeline}
+  \jnote{4}{
+    Gödel's proof is by the \hil{$L$-translation}, where $L$ is the
+    ``constructible universe''. This translation ``relativizes quantification
+    to~$L$'', for instance the~$L$-translation of
+    \[ \varphi \defeqv \bigl(\forall x\_ \exists y\_ \ldots\bigr) \qquad\text{is}\qquad
+    \varphi^L \equiv \bigl(\forall(x \in L)\_ \exists(y \in L)\_ (\ldots)^L\bigr). \]
+    We can then verify, in a weak metatheory such as~\textsc{pra}, that for
+    every formula~$\varphi$ in the language of set theory,
+    \[ \text{if $\textsc{zfc} \vdash \varphi$, then $\textsc{zf} \vdash
+    \varphi^L$}. \]
+    Specializing to~$\varphi \defeqv \bot$ we obtain: If~\textsc{zfc} is
+    inconsistent, then so is~\textsc{zf}. The axiom of choice does not
+    introduce new inconsistencies.
+
+    In modern semantic language: While the axiom of choice might fail in the
+    base universe~$V$ (= the class of all sets), it always holds in~$L$.
+    Gödel's~$L$ was the first \emph{inner model} (= class-sized model of set
+    theory) explicitly studied, nowadays we know many.
+  }
+
+  \jnote{5}{
+    For his proof, Cohen invented the technique of \emph{forcing}, situated in
+    classical mathematics where the base universe~$V$ is assumed to validate
+    the axioms of~\textsc{zfc}.
+
+    Just as a given ring~$R$ or group can be extended in various ways, to include
+    ``generic elements'' as in~$R[X]$ or elements with prescribed relations as
+    in~$R[X]/(X^2+1) =\vcentcolon R[i]$. The idea of forcing is to construct
+    similar such extensions, but of universes (traditionally set-sized models
+    of~\textsc{zfc}, but also class-sized models of~\textsc{zfc}, or models
+    of intuitionistic set theories, or models of type theories, or even models
+    of arithmetic) instead of rings.
+
+    In semantic language, from a high level the idea is the following: Whether
+    the base universe~$V$ contains a cardinal number intermediate
+    between~$\aleph_0$ and~$2^{\aleph_0}$ is uncertain. But there is a certain
+    extension of the base universe---constructed by forcing---which does
+    contain such a number. Like the base~$V$, this forcing extension still
+    validates the axioms of~\textsc{zfc}. Hence there cannot be a
+    \textsc{zfc}-proof of~\textsc{ch}, as in Cohen's extension~$\neg\textsc{ch}$
+    holds.
+
+    Syntactically, Cohen's forcing provides us with an explicit formula
+    translation~$\varphi \mapsto \varphi^C$ such that~\textsc{pra} proves:
+    For every formula~$\varphi$, if~$\textsc{zfc}{\text{+}}\neg\textsc{ch} \vdash \varphi$
+    then~$\textsc{zfc} \vdash \varphi^C$.
+    \bigskip
+  }
+
   \begin{columns}[t]

### 2023-09-15 slides-herrsching2023.tex
More notes

@@
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\includegraphics[height=\paperheight]{multiverse-faded-169}\end{minipage}}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\includegraphics[height=\paperheight]{multiverse-faded}\end{minipage}}
 \begin{frame}[c]
-  \jnote{1}{These are the slides for a three-part lecture series for the autumn school
+  \jnote{1}{
+    These are the slides for a three-part lecture series for the autumn school
     \fixedhref{https://www.mathematik.uni-muenchen.de/~schwicht/pc23.php}{\emph{Proof and Computation}}
     organized by Klaus Mainzer, Peter Schuster and Helmut Schwichtenberg, held in
-  Herrsching from September 20th to September 16th, 2023.}
+    Herrsching from September 20th to September 16th, 2023.
+
+    We explore, in a constructive metatheory, the ``Kripke--Joyal semantics of
+    the internal language of Grothendieck toposes over sites given by
+    preorders'' together with applications in constructive algebra and
+    combinatorics, but without presupposing familiarity with or using language
+    from topos or category theory.
+  }
 
@@
     We can then verify, in a weak metatheory such as~\textsc{pra}, that for
-    every formula~$\varphi$ in the language of set theory,
-    \[ \text{if $\textsc{zfc} \vdash \varphi$, then $\textsc{zf} \vdash
-    \varphi^L$}. \]
-    Specializing to~$\varphi \defeqv \bot$ we obtain: If~\textsc{zfc} is
+    every formula~$\varphi$ in the language of set theory:
+    If $\textsc{zfc}\text{+}\textsc{ch} \vdash \varphi$, then $\textsc{zf} \vdash
+    \varphi^L$.
+
+    Specializing to~$\varphi \defeqv \bot$ we obtain in particular: If~\textsc{zfc} is
     inconsistent, then so is~\textsc{zf}. The axiom of choice does not
@@
 
-    In modern semantic language: While the axiom of choice might fail in the
-    base universe~$V$ (= the class of all sets), it always holds in~$L$.
+    In modern semantic language: While the axiom of choice and~\textsc{ch} might fail in the
+    base universe~$V$ (= the class of all sets), they always hold in~$L$.
     Gödel's~$L$ was the first \emph{inner model} (= class-sized model of set
@@
 
-    Just as a given ring~$R$ or group can be extended in various ways, to include
+    Recall that a given ring~$R$ or group can be extended in various ways, to include
     ``generic elements'' as in~$R[X]$ or elements with prescribed relations as
     in~$R[X]/(X^2+1) =\vcentcolon R[i]$. The idea of forcing is to construct
-    similar such extensions, but of universes (traditionally set-sized models
-    of~\textsc{zfc}, but also class-sized models of~\textsc{zfc}, or models
+    similar such extensions, but not of rings but of universes (traditionally set-sized models
+    of~\textsc{zf} or~\textsc{zfc}, but also class-sized models of~\textsc{zfc}, or models
     of intuitionistic set theories, or models of type theories, or even models
-    of arithmetic) instead of rings.
-
-    In semantic language, from a high level the idea is the following: Whether
+    of arithmetic).
+
+    In semantic language, from a high level the idea of Cohen's independency proof is the following: Whether
     the base universe~$V$ contains a cardinal number intermediate
@@
     translation~$\varphi \mapsto \varphi^C$ such that~\textsc{pra} proves:
-    For every formula~$\varphi$, if~$\textsc{zfc}{\text{+}}\neg\textsc{ch} \vdash \varphi$
+    For every formula~$\varphi$, if~$\textsc{zfc}{\text{+}}\neg\textsc{ch} \vdash \varphi$,
     then~$\textsc{zfc} \vdash \varphi^C$.
     \bigskip
+  }
+
+  \jnote{6}{
+    Joel David Hamkins argues: In view of our rich experience with worlds which
+    validate~\textsc{ch} and worlds which don't, we shouldn't be surprised that
+    no proposed new axiom for settling~\textsc{ch} is ultimately convincing.
+
+    Instead, we should embrace the multiverse of all models of set theory and
+    explore how the truth value of statements of interest change when we
+    travel the multiverse (for instance, by passing from some universe to one
+    of its forcing extensions).
+
+    In this generalized sense, the continuum hypothesis is settled: We have a
+    good understanding of the stability properties of~\textsc{ch} under
+    important constructions. In particular, for a certain precise meaning of
+    ``universe'' and ``extension'', we know that~\textsc{ch} is a
+    \emph{switch}: $\necessary(\possible\textsc{ch} \wedge
+    \possible\neg\textsc{ch})$; in words: Every universe can be extended both to a
+    universe in which~\textsc{ch} holds and to a universe in
+    which~$\neg\textsc{ch}$ holds.
+
+    An exposition and references for further reading about the multiverse
+    position can be found
+    \fixedhref{https://www.speicherleck.de/iblech/stuff/multiverse.pdf}{here}.
+  }
+
+  \jnote{8}{
+    J. Roitman, The uses of set theory, \emph{Math.\@ Intelligencer}
+    \textbf{14}(1) (1992), 63--69.
+
+    Forcing is useful not only to explore the range of foundational
+    possibility; it has many more applications across several subjects of
+    mathematics.
+
+    In particular, we will discuss applications of the constructive version of
+    classical set-theoretic forcing in constructive algebra and combinatorics.
   }
@@
 \begin{frame}{What can forcing do for you?}
+  \jnote{1}{
+    Forcing~$\neg\textsc{ch}$ was the historical use case for forcing.
+
+    With constructive forcing, where we do not blanketly assume that the base
+    universe validates~\textsc{lem}, we can inquire the status of~\textsc{lem};
+    a result in this direction is that~\textsc{lem} is, like~\textsc{ch} in
+    the context of~\textsc{zfc}, a \emph{switch}:
+
+    The base universe can always be extended in such a way as to
+    force~$\neg\textsc{lem}$ (in fact, most forcing extensions will
+    falsify~\textsc{lem} even in case the base universe validates it), and also
+    in such a way to force~\textsc{lem}. The latter provides us with a semantic
+    view of one of the techniques for extracting constructive content from
+    classical proofs, namely the
+    \fixedhref{https://www.speicherleck.de/iblech/stuff/slides-fischbachau2022.pdf}{double-negation
+    translation combined by the continuation trick}.
+  }
+
+  \jnote{2}{
+    We will discuss the particular example of the (naive formulation of the)
+    fundamental theorem of algebra below.
+
+    Forcing has been used to construct countermodels to various questions of
+    constructive (reverse) mathematics, too many to list here. To give just one
+    pointer, a countermodel for the classical implication ``if there is no
+    infinite descending chain, then the partial order is inductively
+    well-founded'' is presented here:
+    A. Blass, Well-ordering and induction in intuitionistic logic and topoi,
+    in: \emph{Mathematical Logic and Theoretical Computer Science}. Ed.\@ by
+    D.\@ Kueker, E.\@ Lopez-Escobar, and C.\@ Smith. Vol.\@ 106. Lect.\@ Notes
+    Pure Appl. Math. Marcel Dekker, 1987, pp.\@ 29--48.
+  }
+
+  \jnote{3}{
+    The real numbers don't contain a number~$i$ such that~$i^2 = 0$. Still, for
+    many results in real analysis, it is convenient to broaden our notion of
+    existence and pass to the complex numbers; the imaginary unit is a
+    \emph{mathematical phantom} in the
+    \fixedhref{http://www.wra1th.plus.com/gcw/math/MathPhant.html}{sense of
+    Gavin Wraith}, a useful tool helping us deduce results about real numbers.
+    Nowadays there are few ontological concerns about the imaginary unit: We
+    understand that, in the end, every statement about complex numbers can be
+    recast as a statement about pairs of real numbers.
+
+    In exactly the same fashion, the objects furnished by forcing can be
+    understood as useful fictions. We will discuss how statements about
+    the forcing extension can be recast as statements about the base universe.
+
+    Given an inhabited set~$X$ in the base universe, most pronouncedly a set
+    which is uncountable or for which no surjection~$\NN \twoheadrightarrow X$
+    can be efficiently evaluated, a particularly tantalizing fiction is the
+    \emph{generic surjection}~$\NN \twoheadrightarrow X$. It exists in a
+    custom-tailored forcing extension of the base universe and is useful to
+    apply tools made for the countable setting to the uncountable; we will
+    discuss an example on the next slide.
+  }
+
+  \jnote{4}{
+    Typically there are already plenty of maps~$\NN \to X$ in the base
+    universe; hence constructing a forcing extension which contains a ``fresh''
+    such map---the so-called \emph{generic sequence}---is not something which
+    would usually be contemplated in classical set-theoretic forcing.
+
+    In the context of constructive mathematics, however, the generic sequence
+    turns out to be quite useful. It can be used to cast in a familiar naive
+    language---the language of infinite sequences---definitions, results and
+    proofs which use constructively more appropriate inductively defined
+    notions. We will discuss an example below.
+  }
+
+  \jnote{6}{
+    Slides by Matthias Hutzler:
+    \fixedhref{https://matthias-hutzler.de/pc23/slides.pdf}{Introduction to
+    synthetic algebraic geometry}
+  }
+
   \vspace*{-0.5em}

### 2023-09-17 slides-herrsching2023.tex
More notes

@@
     similar such extensions, but not of rings but of universes (traditionally set-sized models
-    of~\textsc{zf} or~\textsc{zfc}, but also class-sized models of~\textsc{zfc}, or models
+    of~\textsc{zf} or~\textsc{zfc}, but also class-sized models, or models
     of intuitionistic set theories, or models of type theories, or even models
@@
     Instead, we should embrace the multiverse of all models of set theory and
-    explore how the truth value of statements of interest change when we
-    travel the multiverse (for instance, by passing from some universe to one
+    explore how the truth values of statements of interest change when we
+    travel the multiverse (for instance, by passing from a universe to one
     of its forcing extensions).
@@
     many results in real analysis, it is convenient to broaden our notion of
-    existence and pass to the complex numbers; the imaginary unit is a
+    existence and pass to the complex plane; the imaginary unit is a
     \emph{mathematical phantom} in the
@@
     language---the language of infinite sequences---definitions, results and
-    proofs which use constructively more appropriate inductively defined
-    notions. We will discuss an example below.
+    proofs from constructive combinatorics which use constructively more
+    appropriate inductively defined notions. We will discuss an example below.
   }
@@
 \begin{frame}{Maximal ideals}
+  \jnote{1-}{
+    The theorem on the slide is a generalization of a fact from undergraduate
+    linear algebra: Over a field, no surjective matrix can have more rows than
+    columns. (``Surjective'' here means that the induced linear map is
+    surjective.)
+  }
+
+  \jnote{6-}{
+    The slide presents a standard proof as offered by most textbooks on
+    commutative algebra. The proof is quite efficient from a viewpoint of
+    mathematical organization, as it quickly succeeds in reducing to the field
+    situation. As such, it is short and memorable.
+
+    However, the proof can also be critized for appealing to the transfinite
+    two times; the methods of the proof are at odds with the concreteness of
+    the statement of the theorem---from given equations witnessing
+    surjectivity,~$Mv_i = e_i$, we are asked to deduce the equation~$1 = 0$.
+
+    For this reason, the theorem and its classical proof are often used as case
+    studies for tools and techniques aiming to extract constructive content
+    from classical proofs. One such technique employs constructive forcing.
+    The first(?) constructive proof, found directly without using extraction
+    techniques, is laid out in a
+    \fixedhref{https://www.ams.org/journals/proc/1988-103-04/S0002-9939-1988-0954974-5/S0002-9939-1988-0954974-5.pdf}{beautiful short note by Richman}.
+  }
+
   \only<1-7>{\textbf{Thm.}
@@
     {\vbox{\small{\it maximal among the proper ideals} \\ \medskip \hspace*{-6.75em}\textbullet \quad $\neg(1 \in
-    \mmm)$ \\ \medskip \textbullet \quad $x \in \mmm \quad\vee\quad 1 \in \mmm + (x)$}};
+    \mmm)$ \\ \medskip \textbullet \quad $\neg\bigl(1 \in \mmm + (x)\bigr) \Rightarrow x \in \mmm$}};
     \node [below of=4] (8) {$(8)$};

### 2023-11-23 slides-herrsching2023.tex
More notes

@@
 
+% \begin{document}
+
 {\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
 \begin{frame}{Maximal ideals}
-  \jnote{1-}{
+  \jnote{1-7}{
     The theorem on the slide is a generalization of a fact from undergraduate
@@
 
-  \jnote{6-}{
+  \jnote{6-7}{
     The slide presents a standard proof as offered by most textbooks on
@@
     from classical proofs. One such technique employs constructive forcing.
-    The first(?) constructive proof, found directly without using extraction
+    The first(?)\@ constructive proof, found directly without using extraction
     techniques, is laid out in a
     \fixedhref{https://www.ams.org/journals/proc/1988-103-04/S0002-9939-1988-0954974-5/S0002-9939-1988-0954974-5.pdf}{beautiful short note by Richman}.
+  }
+
+  \jnote{11}{
+    The iterative construction in the countable case without decidability
+    assumptions is due to Krivine; it was later clarified by Berardi and
+    Valentini. It is a parlor trick, resulting in a subset which formally
+    verifies the axioms for a maximal ideal but without carrying out any actual
+    work. Indeed, the resulting ideal will in general not be a detachable
+    subset of the ring.
+
+    Surprisingly, there is still computational content in this construction, as
+    explored in \fixedhref{https://arxiv.org/abs/2207.03873}{this joint paper
+    with Peter Schuster}; one interpretation of our observation is that classical
+    proofs don't ``really'' require a maximal ideal; they just use that notion
+    for structuring hidden computations.
+  }
+
+  \jnote{13-}{
+    In a suitable forcing extension, the ring appears countable. Hence we can
+    carry out the iterative maximal ideal construction there. The resulting ideal
+    will not be part of the base universe (instead, from the point of view of
+    the base universe we will just have constructed a certain sheaf of ideals on a
+    certain pointfree space), but bounded first-order consequences of its
+    existence still pass down to the base.
   }
@@
     \item In the general case: \bad{No}\pause, but \good{yes} in a \emph{suitable forcing extension}\pause, and \\
-    \emph{first-order consequences} of its existence there \good{do hold} in
+    \emph{bounded first-order consequences} of its existence there \good{do hold} in
     the base universe.
@@
 \begin{frame}{Maximal ideals}
+  \jnote{1}{
+    Unwinding all the definitions from constructive forcing and from the
+    iterative maximal ideal construction, and eliminating the application
+    of~\textsc{lem} from the classical proof presented before, we mechanically
+    arrive at the constructive direct proof presented on the slide.
+  }
+
   \textbf{Thm.}
@@
   \emph{Proof.} (special case) Write~$M = \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right)$. By surjectivity,
-  have~$u, v \in A$ with
+  we have~$u, v \in A$ with
   \[
@@
 
+% \begin{document}
+
 {\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[height=\paperheight]{sea-of-clouds-2}\end{minipage}}
 \begin{frame}{Infinite data}
+  \jnote{5-}{
+    The presented proof rests on the law of excluded middle and hence cannot
+    immediately be interpreted as a program for finding suitable indices~$i < j$.
+    However, constructive proofs are also possible (for instance by
+    induction on the value of a given term of the sequence), and furthermore
+    constructive proofs can be extracted from the presented classical proof.
+  }
+
+  \jnote{7-}{
+    The class of well preorders is stable under cartesian products, lists and
+    trees, by Dickson's Lemma, Higson's Lemma and Kruskal's Theorem,
+    respectively. However, in their naive formulations, these are merely
+    theorems of classical mathematics. For general constructive results, the
+    definition of ``well'' needs to be improved.
+
+    Classical texts often employ needless negations in their definitions; this
+    is an XXX
+  }
+
   \vspace*{-1em}

### 2023-11-29 slides-hottest2023.tex
First slide of HoTTEST talk

--- 2023-09-15 slides-herrsching2023.tex
@@
-\documentclass[12pt,utf8,notheorems,compress,t,aspectratio=169]{beamer}
+\documentclass[11pt,utf8,notheorems,compress,t,aspectratio=169]{beamer}
 \usepackage{etex}
@@
 
-\title{Constructive forcing}
+\title{Towards multiversal modal operators for homotopy type theory}
 
 \author{Ingo Blechschmidt}
-\date{September 20th to September 16th, 2023}
+\date{October 30th, 2023}
 
-\setbeameroption{show notes on second screen=bottom}
+%\setbeameroption{show notes on second screen=bottom}
 \newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\footnotesize\justifying#2\par}}}
@@
 \begin{frame}[c]
-  \jnote{1}{
-    These are the slides for a three-part lecture series for the autumn school
-    \fixedhref{https://www.mathematik.uni-muenchen.de/~schwicht/pc23.php}{\emph{Proof and Computation}}
-    organized by Klaus Mainzer, Peter Schuster and Helmut Schwichtenberg, held in
-    Herrsching from September 20th to September 16th, 2023.
-
-    We explore, in a constructive metatheory, the ``Kripke--Joyal semantics of
-    the internal language of Grothendieck toposes over sites given by
-    preorders'' together with applications in constructive algebra and
-    combinatorics, but without presupposing familiarity with or using language
-    from topos or category theory.
-  }
-
   \centering
@@
   \setbeamercolor{block body}{bg=black!100}
-  \begin{minipage}{0.35\textwidth}
+  \begin{minipage}{0.49\textwidth}
     \begin{block}{}
       \centering\normalsize\color{white}
-      \hil{Constructive forcing}
+      \hil{Towards multiversal modal operators for homotopy type theory}
     \end{block}
@@
 
-\begin{frame}{}
-  \jnote{3}{
-    The slide presents a question on calculus. Constructive forcing
-    establishes a connection with constructive linear algebra; we will understand
-    that:
-    \begin{enumerate}\justifying
-      \item The eigenvalues (defined as the zeros of the characteristic
-      polynomial) locally depend continuously on the parameter \hil{because}
-      it is a theorem of constructive linear algebra that symmetric matrices have
-      a full list of eigenvalues.
-
-      \item The eigenvectors can \hil{not} be expected to locally depend
-      continuously \hil{because} it is \hil{not} a theorem of constructive
-      linear algebra that symmetric matrices have an eigenvector basis.
-
-      \item There is a dense open subset of the parameter space which restores
-      continuous dependence \hil{because} it \hil{is} a theorem of constructive
-      linear algebra that every symmetric matrix does \emph{not~not} have an
-      eigenvector basis.
-    \end{enumerate}
-
-    A simple example where the eigenvectors cannot be chosen to locally depend
-    continuously on the parameter is
-    \fixedhref{https://mathoverflow.net/a/60563/31233}{recorded here}. Removing
-    the origin there yields a suitable dense open subset.
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[height=\paperheight]{sea-of-clouds-2}\end{minipage}}
+\begin{frame}{How should this notion be formalized in HoTT?}
+  \jnote{5-}{
+    The presented proof rests on the law of excluded middle and hence cannot
+    immediately be interpreted as a program for finding suitable indices~$i < j$.
+    However, constructive proofs are also possible (for instance by
+    induction on the value of a given term of the sequence), and furthermore
+    constructive proofs can be extracted from the presented classical proof.
   }
 
-  \begin{center}\includegraphics[width=0.2\textwidth]{eigenvector}\end{center}
+  \jnote{7-}{
+    The class of well preorders is stable under cartesian products, lists and
+    trees, by Dickson's Lemma, Higson's Lemma and Kruskal's Theorem,
+    respectively. However, in their naive formulations, these are merely
+    theorems of classical mathematics. For general constructive results, the
+    definition of ``well'' needs to be improved.
 
-  Let a continuous family of symmetric matrices be given:
-  \[
-  \begin{pmatrix}a_{11}(t)&\cdots&a_{1n}(t)\\\vdots&&\vdots\\a_{n1}(t)&\cdots&a_{nn}(t)\end{pmatrix}
-  \]
-
-  Then for every parameter value~$t \in \Omega$, classically there is
-
-  \hil{$\blacktriangleright$} a full list of eigenvalues~$\lambda_1(t),\ldots,\lambda_n(t)$ and \\
-  \hil{$\blacktriangleright$} an eigenvector basis~$(v_1(t),\ldots,v_n(t))$.
-  \bigskip
-
-  \begin{columns}[c]
-    \begin{column}{0.01\textwidth}
-      \includegraphics[height=2.4em]{question-mark}
-    \end{column}
-    \begin{column}{0.9\textwidth}
-      \mbox{Can locally the functions~$\lambda_i$ be chosen to be continuous?
-      \only<2->{\hil{Yes.}}} \\
-      How about the~$v_i$? \only<2->{\hil{No}\only<3->{, but \hil{yes} on a dense
-      open subset of~$\Omega$.}}
-    \end{column}
-  \end{columns}
-\end{frame}
-
-\newcommand\ytl[2]{%
-  \parbox[b]{4.5em}{\hfill{#1}~$\cdots\cdots$~}%
-  \makebox[0pt][c]{\color{mypurple}$\bullet$}{\color{mypurple}\vrule}\quad%
-  \parbox[c]{9.3cm}{\vspace{7pt}\raggedright#2\\[7pt]}%
-  \\[-2.0pt]}
-\begin{frame}{A brief timeline}
-  \jnote{4}{
-    Gödel's proof is by the \hil{$L$-translation}, where $L$ is the
-    ``constructible universe''. This translation ``relativizes quantification
-    to~$L$'', for instance the~$L$-translation of
-    \[ \varphi \defeqv \bigl(\forall x\_ \exists y\_ \ldots\bigr) \qquad\text{is}\qquad
-    \varphi^L \equiv \bigl(\forall(x \in L)\_ \exists(y \in L)\_ (\ldots)^L\bigr). \]
-    We can then verify, in a weak metatheory such as~\textsc{pra}, that for
-    every formula~$\varphi$ in the language of set theory:
-    If $\textsc{zfc}\text{+}\textsc{ch} \vdash \varphi$, then $\textsc{zf} \vdash
-    \varphi^L$.
-
-    Specializing to~$\varphi \defeqv \bot$ we obtain in particular: If~\textsc{zfc} is
-    inconsistent, then so is~\textsc{zf}. The axiom of choice does not
-    introduce new inconsistencies.
-
-    In modern semantic language: While the axiom of choice and~\textsc{ch} might fail in the
-    base universe~$V$ (= the class of all sets), they always hold in~$L$.
-    Gödel's~$L$ was the first \emph{inner model} (= class-sized model of set
-    theory) explicitly studied, nowadays we know many.
+    Classical texts often employ needless negations in their definitions; this
+    is an XXX
   }
 
-  \jnote{5}{
-    For his proof, Cohen invented the technique of \emph{forcing}, situated in
-    classical mathematics where the base universe~$V$ is assumed to validate
-    the axioms of~\textsc{zfc}.
-
-    Recall that a given ring~$R$ or group can be extended in various ways, to include
-    ``generic elements'' as in~$R[X]$ or elements with prescribed relations as
-    in~$R[X]/(X^2+1) =\vcentcolon R[i]$. The idea of forcing is to construct
-    similar such extensions, but not of rings but of universes (traditionally set-sized models
-    of~\textsc{zf} or~\textsc{zfc}, but also class-sized models of~\textsc{zfc}, or models
-    of intuitionistic set theories, or models of type theories, or even models
-    of arithmetic).
-
-    In semantic language, from a high level the idea of Cohen's independency proof is the following: Whether
-    the base universe~$V$ contains a cardinal number intermediate
-    between~$\aleph_0$ and~$2^{\aleph_0}$ is uncertain. But there is a certain
-    extension of the base universe---constructed by forcing---which does
-    contain such a number. Like the base~$V$, this forcing extension still
-    validates the axioms of~\textsc{zfc}. Hence there cannot be a
-    \textsc{zfc}-proof of~\textsc{ch}, as in Cohen's extension~$\neg\textsc{ch}$
-    holds.
-
-    Syntactically, Cohen's forcing provides us with an explicit formula
-    translation~$\varphi \mapsto \varphi^C$ such that~\textsc{pra} proves:
-    For every formula~$\varphi$, if~$\textsc{zfc}{\text{+}}\neg\textsc{ch} \vdash \varphi$,
-    then~$\textsc{zfc} \vdash \varphi^C$.
-    \bigskip
-  }
-
-  \jnote{6}{
-    Joel David Hamkins argues: In view of our rich experience with worlds which
-    validate~\textsc{ch} and worlds which don't, we shouldn't be surprised that
-    no proposed new axiom for settling~\textsc{ch} is ultimately convincing.
-
-    Instead, we should embrace the multiverse of all models of set theory and
-    explore how the truth value of statements of interest change when we
-    travel the multiverse (for instance, by passing from some universe to one
-    of its forcing extensions).
-
-    In this generalized sense, the continuum hypothesis is settled: We have a
-    good understanding of the stability properties of~\textsc{ch} under
-    important constructions. In particular, for a certain precise meaning of
-    ``universe'' and ``extension'', we know that~\textsc{ch} is a
-    \emph{switch}: $\necessary(\possible\textsc{ch} \wedge
-    \possible\neg\textsc{ch})$; in words: Every universe can be extended both to a
-    universe in which~\textsc{ch} holds and to a universe in
-    which~$\neg\textsc{ch}$ holds.
-
-    An exposition and references for further reading about the multiverse
-    position can be found
-    \fixedhref{https://www.speicherleck.de/iblech/stuff/multiverse.pdf}{here}.
-  }
-
-  \jnote{8}{
-    J. Roitman, The uses of set theory, \emph{Math.\@ Intelligencer}
-    \textbf{14}(1) (1992), 63--69.
-
-    Forcing is useful not only to explore the range of foundational
-    possibility; it has many more applications across several subjects of
-    mathematics.
-
-    In particular, we will discuss applications of the constructive version of
-    classical set-theoretic forcing in constructive algebra and combinatorics.
-  }
-
-  \begin{columns}[t]
-    \begin{column}{0.80\textwidth}
-      \ytl{1878}{Cantor advances the \hil{continuum hypothesis}, the claim
-      that~$2^{\aleph_0} = \aleph_1$.}
-      \pause
-      \ytl{1910s}{Zermelo--Fraenkel set theory emerges.}
-      \pause
-      \ytl{1920s}{Set theorists pursue additional axioms to
-      settle~\textsc{ch} \\ (one way or another).}
-      \pause
-      \ytl{1938}{Gödel proves: If~\textsc{zfc} is consistent, so
-      is~\textsc{zfc}+\textsc{ch}.}
-      \pause
-      \ytl{1963}{Cohen proves: If~\textsc{zfc} is consistent, so
-      is~\textsc{zfc}+$\neg$\textsc{ch}.}
-      \pause
-      \ytl{2011}{Hamkins offers his paper on the \hil{multiverse position} in
-      the philosophy of set theory.}
-      \pause
-      %arguing that the program of pursuing
-      %additional axioms (while successful in many ways) is doomed to fail
-      %to settle~\textsc{ch}.}
-      \ytl{2016}{Oldenziel proposes to study the modal multiverse of parametrized
-      mathematics.}
-      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
-      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
-      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
-    \end{column}
-
-    \begin{column}{0.20\textwidth}
-      \pause
-      \centering\includegraphics[width=\textwidth,valign=t]{roitman}
-
-      \scriptsize Judith Roitman %(* 1945)
-      \medskip
-
-      \emph{Mainstream mathematics is beginning to see results using
-      modern set theoretic techniques.}
-    \end{column}
-  \end{columns}
-\end{frame}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.35cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
-\begin{frame}{Constructive forcing}
-  \medskip
-  \begin{columns}[t]
-    \begin{column}{0.40\textwidth}
-      \quad\includegraphics[width=0.8\textwidth,valign=t]{branching}
-    \end{column}
-
-    \begin{column}{0.75\textwidth}
-      \hil{Extending the universe} in various ways, \\[0.7em]
-      \pause
-      \quad similarly how we can extend groups or rings, \\[0.7em]
-      \pause
-      \quad\quad in and for \hil{constructive mathematics} \\[0.7em]
-      \pause
-      \quad\quad\quad \hil{without} presupposing familiarity with \\[0.7em]
-      \pause
-      \quad\quad\quad\quad set theory, topos theory, or sheaves.
-      \pause
-    \end{column}
-  \end{columns}
-
-  \bigskip
-  \bigskip
-
-  \emph{Outline:}
-  \vspace*{-0.5em}
-  \begin{enumerate}
-    \item What can forcing do for you?
-    \item Forcing notions and Kripke--Joyal semantics
-    \item Case studies in constructive algebra and combinatorics
-  \end{enumerate}
-\end{frame}}
-
-\section{What can forcing do for you?}
-
-\begin{frame}{What can forcing do for you?}
-  \jnote{1}{
-    Forcing~$\neg\textsc{ch}$ was the historical use case for forcing.
-
-    With constructive forcing, where we do not blanketly assume that the base
-    universe validates~\textsc{lem}, we can inquire the status of~\textsc{lem};
-    a result in this direction is that~\textsc{lem} is, like~\textsc{ch} in
-    the context of~\textsc{zfc}, a \emph{switch}:
-
-    The base universe can always be extended in such a way as to
-    force~$\neg\textsc{lem}$ (in fact, most forcing extensions will
-    falsify~\textsc{lem} even in case the base universe validates it), and also
-    in such a way to force~\textsc{lem}. The latter provides us with a semantic
-    view of one of the techniques for extracting constructive content from
-    classical proofs, namely the
-    \fixedhref{https://www.speicherleck.de/iblech/stuff/slides-fischbachau2022.pdf}{double-negation
-    translation combined by the continuation trick}.
-  }
-
-  \jnote{2}{
-    We will discuss the particular example of the (naive formulation of the)
-    fundamental theorem of algebra below.
-
-    Forcing has been used to construct countermodels to various questions of
-    constructive (reverse) mathematics, too many to list here. To give just one
-    pointer, a countermodel for the classical implication ``if there is no
-    infinite descending chain, then the partial order is inductively
-    well-founded'' is presented here:
-    A. Blass, Well-ordering and induction in intuitionistic logic and topoi,
-    in: \emph{Mathematical Logic and Theoretical Computer Science}. Ed.\@ by
-    D.\@ Kueker, E.\@ Lopez-Escobar, and C.\@ Smith. Vol.\@ 106. Lect.\@ Notes
-    Pure Appl. Math. Marcel Dekker, 1987, pp.\@ 29--48.
-  }
-
-  \jnote{3}{
-    The real numbers don't contain a number~$i$ such that~$i^2 = 0$. Still, for
-    many results in real analysis, it is convenient to broaden our notion of
-    existence and pass to the complex numbers; the imaginary unit is a
-    \emph{mathematical phantom} in the
-    \fixedhref{http://www.wra1th.plus.com/gcw/math/MathPhant.html}{sense of
-    Gavin Wraith}, a useful tool helping us deduce results about real numbers.
-    Nowadays there are few ontological concerns about the imaginary unit: We
-    understand that, in the end, every statement about complex numbers can be
-    recast as a statement about pairs of real numbers.
-
-    In exactly the same fashion, the objects furnished by forcing can be
-    understood as useful fictions. We will discuss how statements about
-    the forcing extension can be recast as statements about the base universe.
-
-    Given an inhabited set~$X$ in the base universe, most pronouncedly a set
-    which is uncountable or for which no surjection~$\NN \twoheadrightarrow X$
-    can be efficiently evaluated, a particularly tantalizing fiction is the
-    \emph{generic surjection}~$\NN \twoheadrightarrow X$. It exists in a
-    custom-tailored forcing extension of the base universe and is useful to
-    apply tools made for the countable setting to the uncountable; we will
-    discuss an example on the next slide.
-  }
-
-  \jnote{4}{
-    Typically there are already plenty of maps~$\NN \to X$ in the base
-    universe; hence constructing a forcing extension which contains a ``fresh''
-    such map---the so-called \emph{generic sequence}---is not something which
-    would usually be contemplated in classical set-theoretic forcing.
-
-    In the context of constructive mathematics, however, the generic sequence
-    turns out to be quite useful. It can be used to cast in a familiar naive
-    language---the language of infinite sequences---definitions, results and
-    proofs which use constructively more appropriate inductively defined
-    notions. We will discuss an example below.
-  }
-
-  \jnote{6}{
-    Slides by Matthias Hutzler:
-    \fixedhref{https://matthias-hutzler.de/pc23/slides.pdf}{Introduction to
-    synthetic algebraic geometry}
-  }
-
-  \vspace*{-0.5em}
-
-  \begin{columns}
-    \begin{column}{0.45\textwidth}
-      \begin{block}{1. Explore foundational possibility}
-        There are forcing extensions with \\
-        \textsc{ch}, $\neg$\textsc{ch}, \textsc{lem}, $\neg$\textsc{lem}, \ldots
-      \end{block}
-    \end{column}
-    \pause
-
-    \begin{column}{0.45\textwidth}
-      \begin{block}{2. Demonstrate unprovability}
-        The fundamental theorem of algebra is \hil{not constructively provable}
-        as there is a forcing extension where \hil{it is false}.
-      \end{block}
-    \end{column}
-  \end{columns}
-  \pause
-  \bigskip
-
-  \begin{columns}
-    \begin{column}{0.45\textwidth}
-      \begin{block}{3. Harness convenient fictions}
-        For every set, there is a forcing extension where it is
-        \hil{countable}.
-      \end{block}
-    \end{column}
-    \pause
-
-    \begin{column}{0.45\textwidth}
-      \begin{block}{4. Constructivize classical theories}
-        A preorder~$X$ is well iff the \hil{generic sequence} $\NN \to X$ is
-        good.
-      \end{block}
-    \end{column}
-  \end{columns}
-  \pause
-  \bigskip
-
-  \begin{columns}
-    \begin{column}{0.45\textwidth}
-      \begin{block}{5. Study parametric mathematics}
-        Eigenvectors depend continuously on the parameter
-        iff, in a suitable forcing extension, they merely exist.
-      \end{block}
-    \end{column}
-    \pause
-
-    \begin{column}{0.45\textwidth}
-      \begin{block}{6. Develop synthetic accounts}
-        \emph{As in the lectures by Matthias Hutzler.}
-      \end{block}
-    \end{column}
-  \end{columns}
-\end{frame}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
-\begin{frame}{Maximal ideals}
-  \only<1-7>{\textbf{Thm.}
-  Let~$M$ be a surjective matrix with more rows than columns over a
-  ring~$A$. Then~$1 = 0$ in~$A$.
-
-  \visible<2->{\emph{Proof.} \bad{Assume not.}}
-  \visible<3->{Then there is~a \bad{maximal ideal} $\mmm$.}
-  \visible<5->{The matrix is surjective over~$A/\mmm$.}
-  \visible<6->{Since~$A/\mmm$ is a field, this is a contradiction to basic linear algebra.\qed}}
-
-  \only<4-7>{\bigskip\par\centering\scalebox{0.9}{\centering\begin{tikzpicture}
-    \node (0) at (0,1) {$(0) = \{0\}$};
-    \node (1) at (0,5) {$(1) = \ZZ$};
-    \node (2) at (-2,4) {$(2)$};
-    \node [right of=2] (3) {$(3)$};
-    \node [below of=2] (4) {$(4)$};
-    \node [below of=2, xshift=0.7cm] (6) {$(6)$};
-    \node [right of=3] (5) {$(5)$};
-    \node [right of=5] (7) {$(7)$};
-    \node [right of=7] (7d) {$\ldots$\phantom{(}};
-    \node [right of=7d, xshift=3cm, yshift=-2cm] (max)
-    {\vbox{\small{\it maximal among the proper ideals} \\ \medskip \hspace*{-6.75em}\textbullet \quad $\neg(1 \in
-    \mmm)$ \\ \medskip \textbullet \quad $x \in \mmm \quad\vee\quad 1 \in \mmm + (x)$}};
-    \node [below of=4] (8) {$(8)$};
-    \node [right of=8, xshift=3cm] (8d) {$\ldots$};
-    \draw (0) -- (8);
-    \draw (0) -- (8d);
-    \draw (0) -- (6);
-    \draw (2) -- (1);
-    \draw (3) -- (1);
-    \draw (5) -- (1);
-    \draw (7) -- (1);
-    \draw (7d) -- (1);
-    \draw (4) -- (2);
-    \draw (8) -- (4);
-    \draw (6) -- (2);
-    \draw (6) -- (3);
-    \draw [mypurple!30, thick, shorten <=-2pt, shorten >=-2pt, ->] (max) to [out=120, in=-30] (7d);
-    \begin{pgfonlayer}{background}
-      \draw[decorate, very thick, draw=mypurple!30]
-        ($(2.south west) + (8pt, 0)$) arc(270:180:8pt) --
-        ($(2.north west) + (0, -8pt)$) arc(180:90:8pt) --
-        ($(7d.north east) + (-8pt, 0)$) arc(90:0:8pt) --
-        ($(7d.south east) + (0, 8pt)$) arc(0:-90:8pt) --
-        cycle;
-    \end{pgfonlayer}
-  \end{tikzpicture}\par}\par}
-
-  \pause
-  \pause
-  \pause
-  \pause
-  \pause
-  \pause
-  \medskip
-  \raggedright
-
-  Let~$A$ be a ring. \emph{Does there exist a maximal ideal~$\mmm \subseteq A$?}
-  \pause
-  \begin{enumerate}
-    \item \good{Yes}, if \bad{Zorn's lemma} is available.
-    \bigskip
-    \pause
-
-    \item \good{Yes}, if~$A$ is countable and membership of finitely generated ideals is decidable:
-    {\footnotesize
-    Let~$A = \{ x_0, x_1, \ldots \}$. Then set:
-    \begin{align*}
-      \mmm_0 &\defeq \{ 0 \}, &
-      \mmm_{n+1} &\defeq \begin{cases}
-        \mmm_n + (x_n), & \text{if $1 \not\in \mmm_n + (x_n)$}, {\qquad\quad\quad\!\!}\\
-        \mmm_n, & \text{else.}
-      \end{cases}
-    \end{align*}}
-    \pause
-
-    \item \good{Yes}, if~$A$ is countable (irrespective of membership decidability):
-    {\footnotesize\begin{align*}
-      \mmm_0 &\defeq \{ 0 \}, &
-      \mmm_{n+1} &\defeq \mmm_n + (\underbrace{\{ x \in A \,|\, x = x_n \wedge
-        1 \not\in \mmm_n + (x_n) \}}_{\text{a certain subsingleton set}})
-    \end{align*}}
-    \vspace*{3.5em}
-    \visible<11>{{
-      \centering
-      \raisebox{0pt}[0pt][0pt]{
-        \hspace*{2.5em}
-        \scalebox{0.9}{\begin{tikzpicture}
-          \node (inner) at (17.3mm,-20mm) {\textit{``a bad joke''}};
-          \path (0,0) pic{laurel-wreath};
-        \end{tikzpicture}}
-        \hspace*{-3em}
-        \scalebox{0.9}{\begin{tikzpicture}
-          \node (inner) at (17.5mm,-18mm) {\vbox{\small\centering\textit{``non- \\informative''}}};
-          \path (0,0) pic{laurel-wreath};
-        \end{tikzpicture}}
-      }
-      \par
-    }}
-    \pause
-    \pause
-    \vspace*{-3.5em}
-
-    \item In the general case: \bad{No}\pause, but \good{yes} in a \emph{suitable forcing extension}\pause, and \\
-    \emph{first-order consequences} of its existence there \good{do hold} in
-    the base universe.
-  \end{enumerate}
-\end{frame}
-
-\begin{frame}{Maximal ideals}
-  \textbf{Thm.}
-  Let~$M$ be a surjective matrix with more rows than columns over a
-  ring~$A$. Then~$1 = 0$ in~$A$.
-
-  \emph{Proof.} (special case) Write~$M = \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right)$. By surjectivity,
-  have~$u, v \in A$ with
-  \[
-    u \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}1\\0\end{smallmatrix}\right)
-    \quad\text{and}\quad
-    v \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}0\\1\end{smallmatrix}\right).
-  \]
-  Hence
-  $
-    1 = (vy) (ux) = (uy) (vx) = 0
-  $. \qed
-\end{frame}}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[height=\paperheight]{sea-of-clouds-2}\end{minipage}}
-\begin{frame}{Infinite data}
   \vspace*{-1em}
@@
   \pause
+
+  \textbf{Def.} (classically) A preorder~$X$ is \hil{well} iff every sequence~$\NN \to X$ is good.
+
+  \textbf{Examples.} (classically)\ $(\NN,{\leq}),\ \ X \times Y,\ \ X^*,\ \ \mathrm{Tree}(X).$
   \medskip
-
-  \textbf{Def.} A preorder~$X$ is \hil{well\only<9->{$^\star$}} iff every sequence~$\NN \to X$ is good.
-
-  \textbf{Examples.} $(\NN,{\leq}),\ \
-  \color{white}\only<7->{\color{red!90}}\astikznode{onlyclass}{$\underbrace{\color{black}X \times Y,\ \ X^*,\ \ \mathrm{Tree}(X)}_{\text{\visible<7->{\bad{only classically}}}}$}$.
-  \pause
-  \pause
-  \medskip
-
-  \begin{tikzpicture}[remember picture,overlay]
-    \node[thick, fill=black, rectangle, inner sep=0.3em, right=2em of onlyclass] (moral) {
-      \begin{minipage}{6cm}
-        \begin{columns}
-          \begin{column}{0.15\textwidth}
-            \hspace*{1.0em}\color{white}\dbend
-          \end{column}
-          \begin{column}{0.95\textwidth}
-            \color{white}\footnotesize
-            \it Don't quantify over points of spaces which might not have enough.
-          \end{column}
-        \end{columns}
-      \end{minipage}
-    };
-    \path[draw=red!90,thick,-stealth]
-      (moral) to
-      [out=180, in=-00] (onlyclass.east);
-  \end{tikzpicture}
   \pause
 
-  \textbf{Def.} A preorder is \hil{well} iff any of the following equivalent
-  conditions hold:
-  \begin{enumerate}
-    \item The \hil{generic sequence} $\NN \to X$ is good.
+  \subhead{\emph{A naive formalization attempt}}
+
+  \textbf{Def.} $\mathsf{Well}(X,{\leq}) \defeq
+    \prod\limits_{\alpha \? \NN \to X} \, \Bigl\|\sum\limits_{i \? \NN} \sum\limits_{j \? \NN}
+      i < j \times \alpha(i) \leq \alpha(j)\Bigr\|_{-1}$
   \pause
-    \item Every sequence $\NN \to X$ \hil{in every forcing extension} is good.
-    \pause
-    \item There is a \hil{well-founded tree} witnessing universal goodness.
-  \end{enumerate}
 
-%  $\mathsf{Good}\,[\sigma_1,\ldots,\sigma_n] \defeqv (\exists(i < j)\_ \sigma_i \leq \sigma_j)$.
-%  \textbf{Def.} For a predicate~$P$ on finite lists over a set~$X$, inductively
-%  define:
-%  \[
-%    \infer{P \,|\, \sigma}{P\sigma}
-%    \qquad
-%    \infer{P \,|\, \sigma}{\forall(x \in X)\_\ P \,|\, \sigma x}
-%  \]
-%
-%  \textbf{Def.} A preorder is \hil{well} iff~$\mathsf{Good} \,|\, [\,]$, where
-%  $\mathsf{Good}\,[\sigma_1,\ldots,\sigma_n] \defeqv (\exists(i < j)\_ \sigma_i \leq \sigma_j)$.
+  \begin{multicols}{2}
+  \begin{itemize}
+    \item[\cmark] short
+    \item[\cmark] sweet
+    \item[\cmark] immediate
+    \columnbreak
+    \item[\xmark] philosophically strenuous  %: needless reference of the transfinite
+    \item[\xmark] not practical  %: cannot verify closure properties
+    \item[\xmark] not faithful(?)  %: missing out on some sequences
+  \end{itemize}
+  \end{multicols}
 \end{frame}}
 
-\section{Basics of forcing}
-
-\begin{frame}{Ingredients for forcing}
-  To construct a forcing extension, we require:
-  \begin{enumerate}
-    \item a base universe~$V$
-    \item a preorder~$L$ of \hil{forcing conditions} in~$V$\!,
-    pictured as \hil{finite approximations}
-
-    (\emph{convention:} $\tau \preccurlyeq \sigma$ means that~$\tau$ is a
-    better finite approximation than~$\sigma$)
-    \item a \hil{covering system} governing how finite approximations evolve to
-    better ones
-
-    (for each~$\sigma \in L$, a set~$\Cov(\sigma) \subseteq
-    P({\downarrow}\sigma)$, with a simulation condition)
-  \end{enumerate}
-  In the forcing extension~$V^\nabla$, there will then be a \hil{generic filter} (ideal
-  object).
-  \pause
-
-  \vspace*{-1em}
-  \begin{columns}
-    \begin{column}{0.49\textwidth}
-      \small
-      \begin{block}{For the generic surjection~$\NN \twoheadrightarrow X$}
-        \justifying
-        Use \hil{finite lists}~$\sigma \in X^*$ as forcing conditions,
-        where $\tau \preccurlyeq \sigma$ iff~$\sigma$ is an initial segment of~$\tau$,
-        and be prepared to grow~$\sigma$ to \ldots
-        \footnotesize
-        \begin{enumerate}
-          \item[(a)] one of~$\{ \sigma x \,|\, x \in X \}$, to make~$\sigma$ more defined
-          \item[(b)] one of~$\{ \sigma \tau \,|\, \tau \in X^*, a \in
-          \sigma\tau \}$, for any~$a \in X$, to make~$\sigma$ more surjective
-        \end{enumerate}
-      \end{block}
-    \end{column}
-    \pause
-
-    \begin{column}{0.45\textwidth}
-      \small
-      \begin{block}{For the generic prime ideal of a ring~$A$}
-        \justifying
-        Use \hil{f.g.\@ ideals} as forcing conditions, where $\bbb \preccurlyeq
-        \aaa$ iff~$\bbb \supseteq \aaa$, and be prepared to grow~$\aaa$ to \ldots
-        \footnotesize
-        \begin{enumerate}
-          \item[(a)] one of~$\emptyset$, if~$1 \in \aaa$, to make~$\aaa$ more proper
-          \item[(b)] one of~$\{ \aaa+(x), \aaa+(y) \}$, if~$xy \in \aaa$, to
-          make~$\aaa$ more prime
-        \end{enumerate}
-      \end{block}
-    \end{column}
-  \end{columns}
-\end{frame}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{-1cm}\includegraphics[width=\paperwidth]{forest-light-colored}\end{minipage}}
-\begin{frame}{The eventually monad}
-  Let~$L$ be a forcing notion.
-
-  Let~$P$ be a monotone predicate on~$L$
-  (if $\tau \preccurlyeq \sigma$, then $P\sigma \Rightarrow P\tau$). \\
-  For instance, in the case~$L = X^*$:
-  \begin{itemize}
-    \item $\mathsf{Repeats}\, x_0\ldots x_{n-1} \defeqv \exists i\_ \exists j\_ i < j \wedge x_i = x_j$
-    \item $\mathsf{Good}\, \,\,\,\,\,\,x_0\ldots x_{n-1} \defeqv \exists i\_ \exists j\_ i < j \wedge x_i \leq x_j$
-    \quad (for some preorder~$\leq$ on~$X$)
-  \end{itemize}
-  \pause
-
-  We then define~``\hil{$P \mid \sigma$}'' (``$P$ bars~$\sigma$'') inductively by the following clauses:
-  \begin{enumerate}
-    \item If~$P\sigma$, then~$P \mid \sigma$.
-    \item If~$P \mid \tau$ for all~$\tau \in R$, where~$R$ is some covering
-    of~$\sigma$, then~$P \mid \sigma$.
-  \end{enumerate}
-  So~$P \mid \sigma$ expresses in a \hil{direct inductive fashion}:
-  \[ \text{``No matter
-  how~$\sigma$ evolves to a better approximation~$\tau$, eventually~$P\tau$
-  will hold.''} \]
-
-  \pause
-  We use quantifier-like notation: ``$\nabla(\tau \preccurlyeq \sigma)\_
-  P\tau$'' means ``$P \mid \sigma$''.
-\end{frame}}
-% BOARD:
-% - examples for P | σ
-% - abuse of notation
-
-\begin{frame}{Proof translations}
-  \textbf{Thm.} Every~\textsc{iqc}-proof remains correct, with at most a
-  polynomial increase in length, if throughout we
-  replace
-  \[\begin{array}{rcl@{\quad\text{where}\quad}rcl}
-    \exists & \leadsto & \exists^\mathrm{cl},
-    & \exists^\mathrm{cl} &\defeqv& \neg\neg\exists, \\
-    \vee & \leadsto & \vee^\mathrm{cl},
-    & \alpha \vee^\mathrm{cl} \beta &\defeqv& \neg\neg(\alpha \vee \beta), \\
-    = & \leadsto & =^\mathrm{cl},
-    & s =^\mathrm{cl} t &\defeqv& \neg\neg(s = t).
-  \end{array} \]
-  \pause
-
-  \begin{columns}[c]
-    \begin{column}{0.01\textwidth}
-      \includegraphics[height=2.4em]{sheafification-man-2}
-    \end{column}
-    \quad
-    \begin{column}{0.9\textwidth}
-      \hil{When we say:}\ \ some statement ``holds in~$V^{\neg\neg}$'', \\
-      \makesamewidth[l]{\hil{When we say:}}{\hil{we mean:}}\ \ its translation holds in~$V$.
-    \end{column}
-  \end{columns}
-  \bigskip
-
-  Similarly for arbitrary forcing extensions~$V^\nabla$, ``just with~$\nabla$
-  instead of~$\neg\neg$''.
-  \bigskip
-
-  \pause
-  \textbf{Ex.} As~$\neg\neg(\varphi \vee \neg\varphi)$ is a theorem
-  of~\textsc{iqc}, the law of excluded middle holds in~$V^{\neg\neg}$.
-\end{frame}
-
-\newcommand{\defeqvi}{\quad iff\quad}
-\begin{frame}{The $\nabla$-translation}
-  \small
-  \only<1>{For bounded first-order formulas over the (large) first-order signature which has
-  \begin{enumerate}
-    \scriptsize
-    \item one sort~$\underline{X}$ for each set~$X$ in the base universe,
-    \\[-1.2em]
-    \item one~$n$-ary function symbol~$\underline{f} : \underline{X_1} \times
-    \cdots \times \underline{X_n} \to \underline{Y}$ for each map~$f : X_1 \times
-    \cdots \times X_n \to Y$,
-    \\[-1.2em]
-    \item one~$n$-ary relation symbol~$\underline{R} \hookrightarrow
-    \underline{X_1} \times \cdots \times \underline{X_n}$ for each relation~$R
-    \subseteq X_1 \times \cdots \times X_n$, and
-    \\[-1.2em]
-    \item an additional unary relation symbol~$G \hookrightarrow \underline{L}$
-    (for the \emph{generic filter} of~$L$),
-  \end{enumerate}
-  we recursively define:}
-  \scriptsize
-  \only<2->{\vspace*{-0.4em}}
-  \begin{tabbing}
-    \quad \= $\sigma \forces \forces \forall(x\?\underline{X})\_ \varphi$ \=
-    \defeqvi $\textcolor{gray}{\forall(\tau \preccurlyeq \sigma)\_}\
-    \forall(x_0 \in X)\_ \tau \forces \varphi[\underline{x_0}/x]$.\qquad\quad \=
-    $\sigma \forces \exists(x\?\underline{X})\_ \varphi$
-    \= $\sigma \forces \underline{R}(\underline{s_1},\ldots,\underline{s_n})$ \= \defeqvi $s = t$. \= \kill
-
-    \> $\sigma \forces s = t$
-    \> \defeqvi $\nabla \sigma\_ \llbracket s \rrbracket = \llbracket t \rrbracket$.
-    \> $\sigma \forces \underline{R}(s_1,\ldots,s_n)$
-    \> \defeqvi $\nabla\sigma\_ R(\llbracket s_1 \rrbracket,\ldots,\llbracket s_n \rrbracket)$. \\[0.3em]
-
-    \> $\sigma \forces \varphi \Rightarrow \psi$
-    \> \defeqvi $\textcolor{gray}{\forall(\tau \preccurlyeq \sigma)\_}\ (\tau \forces \varphi) \Rightarrow
-    (\tau \forces \psi)$.
-    \> $\sigma \forces G\tau$
-    \> \defeqvi $\nabla\sigma\_ \sigma \preccurlyeq \llbracket\tau\rrbracket$. \\[0.3em]
-
-    \> $\sigma \forces \top$ \> \defeqvi $\top$.
-    \> $\sigma \forces \bot$ \> \defeqvi $\hil{$\nabla\sigma\_$}\ \bot$ \\[0.3em]
-
-    \> $\sigma \forces \varphi \wedge \psi$
-    \> \defeqvi $(\sigma \forces \varphi) \wedge (\sigma \forces \psi)$.
-    \> $\sigma \forces \varphi \vee \psi$
-    \> \defeqvi $\hil{$\nabla\sigma\_$}\ (\sigma \forces \varphi) \vee (\sigma \forces \psi)$. \\[0.3em]
-
-    \> $\sigma \forces \forall(x\?\underline{X})\_ \varphi$
-    \> \defeqvi $\textcolor{gray}{\forall(\tau \preccurlyeq \sigma)\_}\ \forall(x_0 \in X)\_ \tau \forces
-    \varphi[\underline{x_0}/x]$.
-    \> $\sigma \forces \exists(x\?\underline{X})\_ \varphi$
-    \> \defeqvi $\hil{$\nabla\sigma\_$}\ \exists(x_0 \in X)\_ \sigma \forces \varphi[\underline{x_0}/x]$.
-  \end{tabbing}
-  \small
-  \only<1>{Finally, we say that~$\varphi$ ``holds in~$V^\nabla$'' iff for all~$\sigma
-  \in L$, $\sigma \forces \varphi$.}
-
-  \footnotesize
-  \begin{tabular}{@{}lp{0.27\textwidth}p{0.48\textwidth}@{}}
-    \toprule
-    forcing notion & statement about~$V^\nabla$ & external meaning \\
-    \midrule
-    surjection $\NN \twoheadrightarrow X$ &
-    ``the gen.\@ surj.\@ is surjective'' &
-    $\forall(a{\in}X)\_ \forall(\sigma{\in}X^*)\_ \nabla(\tau{\preccurlyeq}\sigma)\_ \exists(n{\in}\NN)\_ \tau[n] = a$. \\[0.4em]
-    \pause
-
-    map $\NN \to X$ &
-    ``the gen.\@ sequence is good'' &
-    $\mathsf{Good} \mid [\,]$. \\[0.4em]
-
-    frame of opens &
-    ``every complex number has a square root'' &
-    For every open~$U \subseteq X$ and every cont.\@
-    function $f : U \to \CC$, there is an open covering $U = \bigcup_i U_i$ such
-    that for each index~$i$, there is a cont.\@ function $g : U_i \to \CC$
-    such that~$g^2 = f$. \\[4.3em]
-
-    big Zariski &
-    ``$x \neq 0 \Rightarrow \text{$x$ inv.}$'' &
-    If the only f.p.\@ $k$-algebra in which~$x = 0$ is the zero algebra,
-    then~$x$ is invertible in~$k$.\\[1.5em]
-    \pause
-
-    little Zariski &
-    ``every f.g. vector space does \emph{not not} have a basis'' &
-    \makesamewidth[l]{}{\phantom{x}}\hil{Grothendieck's generic freeness lemma}
-  \end{tabular}
-\end{frame}
-
-\begin{frame}{Outlook}
-  \begin{block}{Passing to and from extensions}
-    \justifying\small
-    \textbf{Thm.} Let~$\varphi$ be a \hil{bounded first-order formula} not
-    mentioning~$G$. In each of the following situations, we have that
-    $\varphi$ holds in~$V^\nabla$ iff~$\varphi$
-    holds in $V$:
-
-    \vspace*{-0.5em}
-    \begin{enumerate}
-      \item $L$ and all coverings are inhabited (proximality). \\[-1em]
-      \item $L$ contains a top element, every covering of the
-      top element is inhabited, and~$\varphi$ is a coherent implication
-      (positivity).
-    \end{enumerate}
-  \end{block}
-
-  \vspace*{-1em}
-  \begin{columns}
-    \begin{column}{0.46\textwidth}
-      \begin{block}{The mystery of nongeometric sequents}
-        \justifying
-        The \hil{generic ideal} of a ring is maximal:
-        \vspace*{-1em}
-        \[ (x \in \aaa \Rightarrow 1 \in \aaa) \Longrightarrow 1 \in \aaa + (x). \]
-
-        The \hil{generic ring} is a field:
-        \vspace*{-0.7em}
-        \[ (x = 0 \Rightarrow 1 = 0) \Longrightarrow (\exists y\_ xy = 1). \]
-      \end{block}
-
-    \end{column}
-
-    \begin{column}{0.50\textwidth}
-      \begin{block}{Traveling the multiverse \ldots}
-        \textsc{lem} is a \hil{switch} and \hil{holds positively};
-        being countable is a \hil{button}.
-        \medskip
-
-        Every instance of \textsc{dc} \hil{holds proximally}.
-        \medskip
-
-        A geometric implication is provable iff it holds \hil{everywhere}.
-      \end{block}
-      \vspace*{-0.3em}
-      \hfill\footnotesize\ldots{} upwards, but always keeping ties to the
-      base.{\ }
-    \end{column}
-  \end{columns}
-\end{frame}
-
-\begin{frame}{Formalities}
-  \small
-  \textbf{Def.} A \hil{forcing notion} consists of
-  a preorder~$L$ of \hil{forcing conditions}, and
-  for every~$\sigma \in L$, a set~$\Cov(\sigma) \subseteq
-  P({\downarrow}\sigma)$ of \hil{coverings} of~$\sigma$
-  such that: If~$\tau \preccurlyeq \sigma$ and~$R \in \Cov(\sigma)$, there
-  should be a covering~$S \in \Cov(\tau)$ such that~$S \subseteq
-  {\downarrow}R$.
-  \bigskip
-
-  {\centering\footnotesize\begin{tabular}{llll}
-    \toprule
-    & preorder~$L$ & coverings of an element~$\sigma \in L$ & filters of~$L$ \\
-    \midrule
-    \normalnumber{1} & $X^*$ & $\{ \sigma x \,|\, x \in X \}$ & maps~$\NN \to X$ \\
-    \normalnumber{2} & $X^*$ & $\{ \sigma x \,|\, x \in X \}$,\ \ $\{ \sigma\tau \,|\, \tau \in X^*, a \in \sigma\tau \}$ for each~$a \in X$ & surjections~$\NN \twoheadrightarrow X$ \\
-    \normalnumber{3} & f.g. ideals & --- & ideals \\
-    \normalnumber{4} & f.g. ideals & $\{ \sigma+(a), \sigma+(b) \}$ for each~$ab \in \sigma$,\ \ $\{\}$ if~$1 \in \sigma$ & prime ideals \\
-    \normalnumber{5} & opens & $\mathcal{U}$ such that~$\sigma = \bigcup \mathcal{U}$ & points \\
-    \normalnumber{6} & $\{\star\}$ & $\{ \star \,|\, \varphi \} \cup \{ \star \,|\, \neg\varphi \}$ &
-    witnesses of~\textsc{lem}
-    \\
-    \bottomrule
-  \end{tabular}\par}
-  \bigskip
-
-  \textbf{Def.} A \emph{filter} of a forcing notion~$(L,\mathrm{Cov})$
-  is a subset~$F \subseteq L$ such that
-  \vspace*{-0.4em}
-  \begin{enumerate}
-    \scriptsize
-    \item $F$ is upward-closed: if~$\tau \preccurlyeq \sigma$ and if~$\tau \in F$, then~$\sigma \in F$; \\[-3.0em]
-    \item $F$ is downward-directed: $F$ is inhabited, and if~$\alpha,\beta \in F$,
-    then there is a common refinement~$\sigma \preccurlyeq \alpha,\beta$ such
-    that~$\sigma \in F$; and \\[-2.0em]
-    \item $F$ splits the covering system: if~$\sigma \in F$ and~$R \in
-    \Cov(\sigma)$, then~$\tau \in F$ for some~$\tau \in R$.
-  \end{enumerate}
-\end{frame}
-
-\addtocounter{framenumber}{-1}
-
 \end{document}
-
-- Judith Roitman
-- Proof and Computation
-
-- What can forcing do for you?
-  - Exploring the range of foundational possibility: CH, ¬CH, FTA, ...
-  - Eigenvalues
-  - Generic sequence
-  - Generic surjection
-  - Matthias
-  - "Harness the flexibility of parametric mathematics in an accessible way."
-
-Overloading
-
-1. Eigenvalues
-2. Generic surjection
-3. Generic sequence
-4. FTA
-
-Extending the universe in various ways,
-
-similarly how we can extend groups or rings,
-
-in the context of and for the purposes of constructive mathematics
-
-without requiring knowledge of
-
-set theory, topos theory, category theory, sheaves.
-
-Dominique Duval
-
-Forschungsfrage: Klassische Basen sind ja schon ziemlich maximal in dem Sinn,
-als dass etwa der Raum der Funktionen ℕ → ℕ genügend Punkte hat. Wie wäre es,
-in einer Welt zu arbeiten, in der auch der Raum der Surjektionen ℕ → X genügend
-Punkte hat?
-
-Fun:
-- Modale Dinge
-- structure sheaf A~, field property etc. (like in redcom slides), generic freeness
-- nongeometric surprises
-
-außerdem: Konservativität.
-
-\newcommand{\Good}{\mathsf{Good}}
-\newcommand{\vinfer}[2]{\left.#1\quad\middle|\quad\begin{array}{l}#2\end{array}\right.}
-\begin{frame}[plain]
-  \scriptsize
-  \[
-    \renewcommand{\arraystretch}{1.3}
-    \vinfer{\Good \mid [\,]}{
-      \vinfer{\Good \mid 0}{
-        \vinfer{\Good \mid 00}{\Good\ 00} \\[1em]
-        \vinfer{\Good \mid 01}{\Good\ 01} \\[1em]
-        \vinfer{\Good \mid 02}{\Good\ 02} \\
-        \vdots
-      } \\[1em]
-      \ \\[0.5em]
-      \vinfer{\Good \mid 1}{
-        \vinfer{\Good \mid 10}{
-          \vinfer{\Good \mid 100}{\Good\ 100} \\[1em]
-          \vinfer{\Good \mid 101}{\Good\ 101} \\[1em]
-          \vinfer{\Good \mid 102}{\Good\ 102} \\
-          \vdots
-        } \\[1em]
-        \vinfer{\Good \mid 11}{\Good\ 11} \\[1em]
-        \vinfer{\Good \mid 12}{\Good\ 12} \\[1em]
-        \vdots
-      } \\[1em]
-      \ \\[0.5em]
-      \vinfer{\Good \mid 2}{} \\[1em]
-      \vdots
-    }
-  \]
-\end{frame}

### 2023-11-29 slides-hottest2023.tex
First slide finished

@@
   \textbf{Examples.} (classically)\ $(\NN,{\leq}),\ \ X \times Y,\ \ X^*,\ \ \mathrm{Tree}(X).$
-  \medskip
+  \bigskip
   \pause
 
-  \subhead{\emph{A naive formalization attempt}}
+  %\subhead{\emph{A naive formalization attempt}}
+  \emph{A naive formalization attempt:}
 
-  \textbf{Def.} $\mathsf{Well}(X,{\leq}) \defeq
+  \textbf{Def.} $\displaystyle\mathsf{Well}(X,{\leq}) \defeq
     \prod\limits_{\alpha \? \NN \to X} \, \Bigl\|\sum\limits_{i \? \NN} \sum\limits_{j \? \NN}
@@
   \pause
+  \qquad\scalebox{2}{\dbend}
+  \medskip
 
-  \begin{multicols}{2}
   \begin{itemize}
-    \item[\cmark] short
-    \item[\cmark] sweet
-    \item[\cmark] immediate
-    \columnbreak
-    \item[\xmark] philosophically strenuous  %: needless reference of the transfinite
-    \item[\xmark] not practical  %: cannot verify closure properties
-    \item[\xmark] not faithful(?)  %: missing out on some sequences
+    \item[\bad{\xmark}] philosophically~strenuous  %: needless reference of the transfinite
+    \item[\bad{\xmark}] not practical  %: cannot verify closure properties
+    \item[\bad{\xmark}] not faithful(?)  %: missing out on some sequences
   \end{itemize}
-  \end{multicols}
 \end{frame}}

### 2023-11-29 slides-hottest2023.tex
Sequences provided by LEM and sequences depending on an environment

@@
 \usepackage{proof}
+\usepackage{agda}
 \usepackage[all]{xy}
@@
 
+\newcommand{\dnote}[1]{%
+  \begin{tabular}{@{}m{2em}@{}m{0.80\textwidth}@{}}%
+    \textdbend &#1%
+  \end{tabular}%
+  \par
+}
+
 \begin{document}
@@
   \jnote{7-}{
-    The class of well preorders is stable under cartesian products, lists and
+    The class of well quasiorders is stable under cartesian products, lists and
     trees, by Dickson's Lemma, Higson's Lemma and Kruskal's Theorem,
@@
 
+  \begin{block}{}
+    \justifying
     \textbf{Thm.} Every sequence~$\alpha : \NN \to \NN$ is \hil{good} in that
-  there exist~$i < j$ with~$\alpha(i) \leq \alpha(j)$.
+    there exist~$i < j$ with~$\alpha\,i \leq \alpha\,j$.
+  \end{block}
   \pause
-
+  \vspace*{-0.6em}
   \emph{Proof.} \emph{(offensive?)} By~\badbox{\textsc{lem}}, there is a
-  minimum~$\alpha(i)$.
+  minimum~$\alpha\,i$.
   Set~$j \defeq i + 1$. \qed\par
@@
 
-  \textbf{Def.} (classically) A preorder~$X$ is \hil{well} iff every sequence~$\NN \to X$ is good.
+  \textbf{Def.} (classically) A quasiorder~$X$ is \hil{well} iff every sequence~$\NN \to X$ is good.
 
@@
     \prod\limits_{\alpha \? \NN \to X} \, \Bigl\|\sum\limits_{i \? \NN} \sum\limits_{j \? \NN}
-      i < j \times \alpha(i) \leq \alpha(j)\Bigr\|_{-1}$
+      i < j \times \alpha\,i \leq \alpha\,j\Bigr\|_{-1}$
   \pause
@@
 
+\begin{frame}{Sequences provided by \textsc{lem}}
+  \justifying
+  \begin{block}{}
+    \justifying
+    \textbf{Lemma.} Let~$X$ be a well quasiorder. Let~$\alpha : \NN \to X$ be a
+    sequence. Assuming~\bad{\textsc{lem}}, there merely is a monotonic
+    subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
+  \end{block}
+  \vspace*{-0.6em}
+  \emph{Proof.} The set
+  \[ I \defeq \{ i \in \NN \,|\, \neg\mathop{\exists}\limits_{j \in \NN} i < j \times \alpha\,i \leq \alpha\,j
+  \} \]
+  cannot be in bijection with~$\NN$, as else the~$I$-extracted subsequence
+  of~$\alpha$ would not be good. By~\bad{\textsc{lem}}, the set~$I$ is finite.
+  Any index~$i_0$ larger than all the indices in~$I$ is a suitable starting
+  point for a monotonic subsequence. \qed
+  \medskip
+  \pause
+
+  \begin{block}{}
+    \justifying
+    \textbf{Prop.} Let~$X$ and~$Y$ be well quasiorders.
+    Assuming~\bad{\textsc{lem}},~$X \times Y$ is well.
+  \end{block}
+  \vspace*{-0.6em}
+  \emph{Proof.} Let an infinite sequence~$\gamma : \NN \to X \times Y$ be
+  given. Write~$\gamma\,k = (\alpha\,k,\beta\,k)$. By the lemma, there is a
+  monotonic subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$. Because~$Y$
+  is well, there are indices~$n < m$ such that~$\beta\,i_n \leq \beta\,i_m$. As
+  also~$\alpha\,i_n \leq \alpha\,i_m$, the sequence~$\gamma$ is good.
+  \medskip
+  \pause
+
+  \dnote{\emph{We cannot trust~\bad{\textsc{lem}}-provided sequences to be
+  available in the type~$\NN \to \NN$. Similarly with~\bad{\textsc{dc}}.}}
+\end{frame}
+
+\begin{frame}{Sequences depending on an environment}
+  Let~$X$ be an hset such that there is no surjection~$\NN \twoheadrightarrow X$.
+  Then the type~$\NN \to X$ misses the \hil{generic enumeration} of~$X$.
+  \medskip
+  \pause
+
+  \emph{Idea:} Approximate a (fictitious) surjection~$\NN \twoheadrightarrow X$
+  by finite sequences~$\sigma = x_0 \ldots x_{n-1} \? X^*$. Over time, such an
+  approximation can grow to
+  \begin{enumerate}
+    \item one of $\sigma y$, where~$y \? X$, so it becomes \emph{more defined}; or
+    \item one of~$\sigma \tau$, where~$\tau \? X^*$ such that~$a \in
+    \sigma\tau$, so it becomes \emph{more surjective}.
+  \end{enumerate}
+  \medskip
+  \pause
+
+  Formally, we have the type~$\nabla\underline{X}$ of stage-dependent values of~$X$:
+\end{frame}
+
 \end{document}

### 2023-11-29 slides-hottest2023.lagda.tex
First chapter finished

--- 2023-11-29 slides-hottest2023.tex
@@
 
+\newcommand{\genalpha}{\mbox{$\hspace{0.12em}\shortmid\hspace{-0.62em}\alpha$}}
+
+\usepackage{newunicodechar}
+
+\newunicodechar{∇}{\ensuremath{\nabla}}
+\newunicodechar{λ}{\ensuremath{\lambda}}
+\newunicodechar{σ}{\ensuremath{\sigma}}
+\newunicodechar{τ}{\ensuremath{\tau}}
+\newunicodechar{∷}{\ensuremath{\!\!\!}}
+\newunicodechar{⧺}{\ensuremath{\!\!\!}}
+\newunicodechar{₁}{\ensuremath{_1}}
+\newunicodechar{₂}{\ensuremath{_2}}
+\newunicodechar{∈}{\ensuremath{\in}}
+
 \begin{document}
@@
 
-  %\subhead{\emph{A naive formalization attempt}}
   \emph{A naive formalization attempt:}
 
-  \textbf{Def.} $\displaystyle\mathsf{Well}(X,{\leq}) \defeq
+  \textbf{Def.} $\displaystyle\mathsf{Well}_0(X,{\leq}) \defeq
     \prod\limits_{\alpha \? \NN \to X} \, \Bigl\|\sum\limits_{i \? \NN} \sum\limits_{j \? \NN}
@@
   \pause
-  \qquad\scalebox{2}{\dbend}
+  \qquad\scalebox{1.5}{\dbend}
   \medskip
@@
   Let~$X$ be an hset such that there is no surjection~$\NN \twoheadrightarrow X$.
-  Then the type~$\NN \to X$ misses the \hil{generic enumeration} of~$X$.
+  Then the type~$\NN \to X$ misses the \hil{generic enumeration}~$\genalpha$ of~$X$.
   \medskip
@@
 
-  \emph{Idea:} Approximate a (fictitious) surjection~$\NN \twoheadrightarrow X$
-  by finite sequences~$\sigma = x_0 \ldots x_{n-1} \? X^*$. Over time, such an
+  \emph{Idea:} Approximate (fictitious) surjections~$\NN \twoheadrightarrow X$
+  by finite sequences~$\sigma = x_0 \ldots x_{n-1} \? X^*$. \\ Over time, such an
   approximation can grow to
@@
 
-  Formally, we have the type~$\nabla\underline{X}$ of stage-dependent values of~$X$:
+  For~$k : \NN$ and~$a : X$, the expression~``$\genalpha\,k = a$'' does not
+  denote a proposition but rather a \hil{stage-dependent proposition}~$X^* \to
+  \mathsf{Prop}$, namely $\lambda x_0 \ldots x_{n-1}.\ k < n \times x_k = a$.
+  \medskip
+  \pause
+
+  Given a stage-dependent proposition~$P$, $\hil{$\nabla P$}\,\sigma$ expresses that no
+  matter how~$\sigma$ evolves to a better approximation~$\tau$,
+  eventually~$P\,\tau$ will hold:
+  \AgdaHide{
+  \begin{code}
+    {-# OPTIONS --no-prop --no-import-sorts --allow-unsolved-metas #-}
+    open import Agda.Primitive using (Set)
+    Prop = Set
+    postulate
+      X : Set
+      X* : Set
+      [] : X*
+      _∷_ : X* → X → X*
+      _⧺_ : X* → X* → X*
+      _∈_ : X → X* → Prop
+    module Foo where
+  \end{code}
+  }
+  {\small\begin{code}
+      data ∇ (P : X* → Prop) (σ : X*) : Prop where
+        now : P σ → ∇ P σ
+        later₁ : ((y : X) → ∇ P (σ ∷ y)) → ∇ P σ
+        later₂ : (a : X) → ((τ : X*) → a ∈ (σ ⧺ τ) → ∇ P (σ ⧺ τ)) → ∇ P σ
+  \end{code}}
+
+  \vspace*{-0.8em}
+  For instance, for~$P\,x_0 \ldots x_{n-1} \defeq \exists_{k : \NN} (k < n
+  \times x_k = a)$, we have~$\nabla P \, \varepsilon$. \quad ``$\genalpha$ is surjective''
+\end{frame}
+
+\begin{frame}{Well quasiorders revisited}
+  \emph{A naive formalization attempt:}
+
+  \textbf{Def.} $\displaystyle\mathsf{Well}_0(X,{\leq}) \defeq
+    \prod\limits_{\alpha \? \NN \to X} \, \Bigl\|\sum\limits_{i \? \NN} \sum\limits_{j \? \NN}
+      i < j \times \alpha\,i \leq \alpha\,j\Bigr\|_{-1}$
+  \pause
+  \qquad\scalebox{1.0}{\dbend}
+  \bigskip
+
+  \emph{An inductive rephrasing:}
+
+  \textbf{Def.} $\displaystyle\mathsf{Well}(X,{\leq}) \defeq
+    \nabla \mathsf{Good} \, \varepsilon$, where
+  {\vspace*{-0.7em}\small\begin{code}
+    data ∇ (P : X* → Prop) (σ : X*) : Prop where
+      now : P σ → ∇ P σ
+      later : ((y : X) → ∇ P (σ ∷ y)) → ∇ P σ
+  \end{code}}
+
+  \vspace*{-0.8em}
+  In other words: \emph{A quasiorder~$X$ is well iff the generic sequence~$\NN \to X$
+  is good.}
+
+  \begin{block}{}
+    \justifying
+    \textbf{Prop.} Let~$X$ and~$Y$ be well quasiorders.
+    \good{Without} assuming~\bad{\textsc{lem}},~$X \times Y$ is well.
+  \end{block}
+
+  \dnote{We have~$\mathsf{Well}(X,{\leq}) \to
+  \mathsf{Well}_0(X,{\leq})$, but the converse only holds in presence of \hil{bar
+  induction}. \emph{How much stronger exactly is the inductive rephrasing?}}
 \end{frame}

### 2023-11-29 slides-hottest2023.lagda.tex
Most slides finished

@@
 
-  \textbf{Def.} $\displaystyle\mathsf{Well}_0(X,{\leq}) \defeq
+  \textbf{Def.} $\displaystyle\mathsf{Well}_\infty(X,{\leq}) \defeq
     \prod\limits_{\alpha \? \NN \to X} \, \Bigl\|\sum\limits_{i \? \NN} \sum\limits_{j \? \NN}
@@
   \textbf{Def.} $\displaystyle\mathsf{Well}(X,{\leq}) \defeq
-    \nabla \mathsf{Good} \, \varepsilon$, where
+    \nabla \mathsf{Good} \, \varepsilon$, where~$\mathsf{Good}\,x_0 \ldots
+    x_{n-1} \defeq \mathop{\exists}\limits_{i:\NN}\mathop{\exists}\limits_{j:\NN} (i < j \times x_i \leq
+    x_j)$ and
   {\vspace*{-0.7em}\small\begin{code}
@@
   \dnote{We have~$\mathsf{Well}(X,{\leq}) \to
-  \mathsf{Well}_0(X,{\leq})$, but the converse only holds in presence of \hil{bar
+  \mathsf{Well}_\infty(X,{\leq})$, but the converse only holds in presence of \hil{bar
   induction}. \emph{How much stronger exactly is the inductive rephrasing?}}
@@
 
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses}\end{minipage}}
+\begin{frame}{Slide about toposes}
+  (something about overloading of the logical connectives)
+\end{frame}}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
+\begin{frame}{Maximal ideals as convenient fictions}
+  \jnote{1}{
+    The theorem on the slide is a generalization of a fact from undergraduate
+    linear algebra: Over a field, no surjective matrix can have more rows than
+    columns. (``Surjective'' here means that the induced linear map is
+    surjective.)
+  }
+
+  \jnote{6-7}{
+    The slide presents a standard proof as offered by most textbooks on
+    commutative algebra. The proof is quite efficient from a viewpoint of
+    mathematical organization, as it quickly succeeds in reducing to the field
+    situation. As such, it is short and memorable.
+
+    However, the proof can also be critized for appealing to the transfinite
+    two times; the methods of the proof are at odds with the concreteness of
+    the statement of the theorem---from given equations witnessing
+    surjectivity,~$Mv_i = e_i$, we are asked to deduce the equation~$1 = 0$.
+
+    For this reason, the theorem and its classical proof are often used as case
+    studies for tools and techniques aiming to extract constructive content
+    from classical proofs. One such technique employs constructive forcing.
+    The first(?)\@ constructive proof, found directly without using extraction
+    techniques, is laid out in a
+    \fixedhref{https://www.ams.org/journals/proc/1988-103-04/S0002-9939-1988-0954974-5/S0002-9939-1988-0954974-5.pdf}{beautiful short note by Richman}.
+  }
+
+  \jnote{8-9}{
+    For the countable case, there is an iterative construction of a maximal
+    ideal, making do without any decidability assumptions (due to Krivine,
+    later later clarified by Berardi and Valentini). It is a parlor trick,
+    resulting in a subset which formally verifies the axioms for a maximal
+    ideal but without carrying out any actual work. Indeed, the resulting ideal
+    will in general not be a detachable subset of the ring.
+
+    Surprisingly, there is still computational content in this construction, as
+    explored in \fixedhref{https://arxiv.org/abs/2207.03873}{this joint paper
+    with Peter Schuster}; one interpretation of our observation is that classical
+    proofs don't ``really'' require a maximal ideal; they just use that notion
+    for structuring hidden computations.
+  }
+
+  \jnote{9}{
+    In a suitable topos, the ring appears countable. Hence we can
+    carry out the iterative maximal ideal construction there. The resulting ideal
+    will not be part of the base universe (instead, from the point of view of
+    the base universe we will just have constructed a certain sheaf of ideals on a
+    certain pointfree space), but bounded first-order consequences of its
+    existence still pass down to the base.
+  }
+
+  \jnote{10}{
+    Unwinding all the definitions from toposes and from the
+    iterative maximal ideal construction, and eliminating the application
+    of~\textsc{lem} from the classical proof presented before, we mechanically
+    arrive at the constructive direct proof presented on the slide.
+  }
+
+  \begin{block}{}
+    \textbf{Thm.}
+    Let~$M$ be a surjective matrix with more rows than columns over a
+    commutative ring~$A$. Then~$1 = 0$ in~$A$.
+  \end{block}
+
+  \only<1-9>{
+    \visible<2->{\emph{Proof.} (classical) \bad{Assume not.}}
+    \visible<3->{Then there is~a \bad{maximal ideal} $\mmm$.}
+    \visible<5->{The matrix is surjective over~$A/\mmm$.}
+    \visible<6->{Since~$A/\mmm$ is a field, this is a contradiction to basic linear algebra.\qed}
+  }
+
+  \only<4-9>{\medskip\par\centering\scalebox{0.9}{\centering\begin{tikzpicture}
+    \node (0) at (0,1) {$(0) = \{0\}$};
+    \node (1) at (0,5) {$(1) = \ZZ$};
+    \node (2) at (-2,4) {$(2)$};
+    \node [right of=2] (3) {$(3)$};
+    \node [below of=2] (4) {$(4)$};
+    \node [below of=2, xshift=0.7cm] (6) {$(6)$};
+    \node [right of=3] (5) {$(5)$};
+    \node [right of=5] (7) {$(7)$};
+    \node [right of=7] (7d) {$\ldots$\phantom{(}};
+    \node [right of=7d, xshift=3cm, yshift=-2cm] (max)
+    {\vbox{\small{\it maximal among the proper ideals} \\ \medskip \hspace*{-6.75em}\textbullet \quad $\neg(1 \in
+    \mmm)$ \\ \medskip \textbullet \quad $\neg\bigl(1 \in \mmm + (x)\bigr) \Rightarrow x \in \mmm$}};
+    \node [below of=4] (8) {$(8)$};
+    \node [right of=8, xshift=3cm] (8d) {$\ldots$};
+    \draw (0) -- (8);
+    \draw (0) -- (8d);
+    \draw (0) -- (6);
+    \draw (2) -- (1);
+    \draw (3) -- (1);
+    \draw (5) -- (1);
+    \draw (7) -- (1);
+    \draw (7d) -- (1);
+    \draw (4) -- (2);
+    \draw (8) -- (4);
+    \draw (6) -- (2);
+    \draw (6) -- (3);
+    \draw [mypurple!30, thick, shorten <=-2pt, shorten >=-2pt, ->] (max) to [out=120, in=-30] (7d);
+    \begin{pgfonlayer}{background}
+      \draw[decorate, very thick, draw=mypurple!30]
+        ($(2.south west) + (8pt, 0)$) arc(270:180:8pt) --
+        ($(2.north west) + (0, -8pt)$) arc(180:90:8pt) --
+        ($(7d.north east) + (-8pt, 0)$) arc(90:0:8pt) --
+        ($(7d.south east) + (0, 8pt)$) arc(0:-90:8pt) --
+        cycle;
+    \end{pgfonlayer}
+  \end{tikzpicture}\par}\par}
+
+  \pause
+  \pause
+  \pause
+  \pause
+  \pause
+  \pause
+  \raggedright
+
+  \only<7-9>{
+    \emph{Does there exist a maximal ideal?}
+    \pause
+    \good{Yes}, if~$A$ is countable.
+    In the general case: \bad{No}\pause, but \good{yes} in a \emph{suitable topos}, and
+    \emph{bounded first-order consequences} of its existence there \good{do hold} in
+    the base.
+    \pause
+  }
+
+  \only<10>{
+    \emph{Proof.} (constructive, special case) Write~$M =
+    \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right)$. By surjectivity,
+    we have~$u, v \? A$ with
+    \[
+      u \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}1\\0\end{smallmatrix}\right)
+      \quad\text{and}\quad
+      v \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}0\\1\end{smallmatrix}\right).
+    \]
+    Hence
+    $
+      1 = (vy) (ux) = (uy) (vx) = 0
+    $. \qed
+  }
+\end{frame}}
+
+\newcommand\ytl[2]{%
+  \parbox[b]{4.5em}{\hfill{#1}~$\cdots\cdots$~}%
+  \makebox[0pt][c]{\color{mypurple}$\bullet$}{\color{mypurple}\vrule}\quad%
+  \parbox[c]{9.3cm}{\vspace{7pt}\raggedright#2\\[7pt]}%
+  \\[-2.0pt]}
+\begin{frame}{A brief timeline}
+  \jnote{4}{
+    Gödel's proof is by the \hil{$L$-translation}, where $L$ is the
+    ``constructible universe''. This translation ``relativizes quantification
+    to~$L$'', for instance the~$L$-translation of
+    \[ \varphi \defeqv \bigl(\forall x\_ \exists y\_ \ldots\bigr) \qquad\text{is}\qquad
+    \varphi^L \equiv \bigl(\forall(x \in L)\_ \exists(y \in L)\_ (\ldots)^L\bigr). \]
+    We can then verify, in a weak metatheory such as~\textsc{pra}, that for
+    every formula~$\varphi$ in the language of set theory:
+    If $\textsc{zfc}\text{+}\textsc{ch} \vdash \varphi$, then $\textsc{zf} \vdash
+    \varphi^L$.
+
+    Specializing to~$\varphi \defeqv \bot$ we obtain in particular: If~\textsc{zfc} is
+    inconsistent, then so is~\textsc{zf}. The axiom of choice does not
+    introduce new inconsistencies.
+
+    In modern semantic language: While the axiom of choice and~\textsc{ch} might fail in the
+    base universe~$V$ (= the class of all sets), they always hold in~$L$.
+    Gödel's~$L$ was the first \emph{inner model} (= class-sized model of set
+    theory) explicitly studied, nowadays we know many.
+  }
+
+  \jnote{5}{
+    For his proof, Cohen invented the technique of \emph{forcing}, situated in
+    classical mathematics where the base universe~$V$ is assumed to validate
+    the axioms of~\textsc{zfc}.
+
+    Recall that a given ring~$R$ or group can be extended in various ways, to include
+    ``generic elements'' as in~$R[X]$ or elements with prescribed relations as
+    in~$R[X]/(X^2+1) =\vcentcolon R[i]$. The idea of forcing is to construct
+    similar such extensions, but not of rings but of universes (traditionally set-sized models
+    of~\textsc{zf} or~\textsc{zfc}, but also class-sized models, or models
+    of intuitionistic set theories, or models of type theories, or even models
+    of arithmetic).
+
+    In semantic language, from a high level the idea of Cohen's independency proof is the following: Whether
+    the base universe~$V$ contains a cardinal number intermediate
+    between~$\aleph_0$ and~$2^{\aleph_0}$ is uncertain. But there is a certain
+    extension of the base universe---constructed by forcing---which does
+    contain such a number. Like the base~$V$, this forcing extension still
+    validates the axioms of~\textsc{zfc}. Hence there cannot be a
+    \textsc{zfc}-proof of~\textsc{ch}, as in Cohen's extension~$\neg\textsc{ch}$
+    holds.
+
+    Syntactically, Cohen's forcing provides us with an explicit formula
+    translation~$\varphi \mapsto \varphi^C$ such that~\textsc{pra} proves:
+    For every formula~$\varphi$, if~$\textsc{zfc}{\text{+}}\neg\textsc{ch} \vdash \varphi$,
+    then~$\textsc{zfc} \vdash \varphi^C$.
+    \bigskip
+  }
+
+  \jnote{6}{
+    Joel David Hamkins argues: In view of our rich experience with worlds which
+    validate~\textsc{ch} and worlds which don't, we shouldn't be surprised that
+    no proposed new axiom for settling~\textsc{ch} is ultimately convincing.
+
+    Instead, we should embrace the multiverse of all models of set theory and
+    explore how the truth values of statements of interest change when we
+    travel the multiverse (for instance, by passing from a universe to one
+    of its forcing extensions).
+
+    In this generalized sense, the continuum hypothesis is settled: We have a
+    good understanding of the stability properties of~\textsc{ch} under
+    important constructions. In particular, for a certain precise meaning of
+    ``universe'' and ``extension'', we know that~\textsc{ch} is a
+    \emph{switch}: $\necessary(\possible\textsc{ch} \wedge
+    \possible\neg\textsc{ch})$; in words: Every universe can be extended both to a
+    universe in which~\textsc{ch} holds and to a universe in
+    which~$\neg\textsc{ch}$ holds.
+
+    An exposition and references for further reading about the multiverse
+    position can be found
+    \fixedhref{https://www.speicherleck.de/iblech/stuff/multiverse.pdf}{here}.
+  }
+
+  \jnote{8}{
+    J. Roitman, The uses of set theory, \emph{Math.\@ Intelligencer}
+    \textbf{14}(1) (1992), 63--69.
+
+    Forcing is useful not only to explore the range of foundational
+    possibility; it has many more applications across several subjects of
+    mathematics.
+  }
+
+  \begin{columns}[t]
+    \begin{column}{0.80\textwidth}
+      \ytl{1878}{Cantor advances the \hil{continuum hypothesis}, the claim
+      that~$2^{\aleph_0} = \aleph_1$.}
+      \pause
+      \ytl{1910s}{Zermelo--Fraenkel set theory emerges.}
+      \pause
+      \ytl{1920s}{Set theorists pursue additional axioms to
+      settle~\textsc{ch} \\ (one way or another).}
+      \pause
+      \ytl{1938}{Gödel proves: If~\textsc{zfc} is consistent, so
+      is~\textsc{zfc}+\textsc{ch}.}
+      \pause
+      \ytl{1963}{Cohen proves: If~\textsc{zfc} is consistent, so
+      is~\textsc{zfc}+$\neg$\textsc{ch}.}
+      \pause
+      \ytl{2011}{Hamkins offers his paper on the \hil{multiverse position} in
+      the philosophy of set theory.}
+      \pause
+      %arguing that the program of pursuing
+      %additional axioms (while successful in many ways) is doomed to fail
+      %to settle~\textsc{ch}.}
+      \ytl{2016}{Oldenziel proposes to study the modal multiverse of parametrized
+      mathematics.}
+      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
+      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
+      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
+    \end{column}
+
+    \begin{column}{0.20\textwidth}
+      \pause
+      \centering\includegraphics[width=\textwidth,valign=t]{roitman}
+
+      \scriptsize Judith Roitman %(* 1945)
+      \medskip
+
+      \emph{Mainstream mathematics is beginning to see results using
+      modern set theoretic techniques.}
+    \end{column}
+  \end{columns}
+\end{frame}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.59cm}\includegraphics[width=\paperwidth]{staircase}\end{minipage}}
+\begin{frame}{The modal set-theoretic multiverse}
+  \begin{tikzpicture}[remember picture,overlay]
+    \node[xshift=-2.5cm,yshift=-4cm] at (current page.north east)
+    {\includegraphics[width=3cm]{branching}};
+  \end{tikzpicture}
+  \justifying
+  \vspace*{-2em}
+
+  \textbf{Def.} A \hil{model of set theory} is a (perhaps class-sized) structure $(M,{\in})$
+  satisfying axioms such as those of~\textsc{zfc}.
+
+  {\small\emph{Examples.}\vspace*{-0.8em}
+  \begin{itemize}
+    \item $V$, the class of all sets \vspace*{-0.6em}
+    \item $L$, Gödel's constructible universe \vspace*{-0.6em}
+    \item $V[G]$, a forcing extension containing a generic filter~$G$ of \\ some
+    poset of forcing conditions \vspace*{-0.6em}
+    \item Henkin/term models from consistency of (extensions of)~\textsc{zfc}
+  \end{itemize}}
+  \pause
+
+  \emph{We embrace all models of set theory:}
+
+  \textbf{Def.} $\possible\varphi$ iff~$\varphi$ holds in \hil{some extension} of
+  the current universe. \\
+  \phantom{\textbf{Def.}} $\necessary\varphi$ iff~$\varphi$ holds in \hil{all extensions}
+  of the current universe.
+  \begin{itemize}
+    \item $\necessary(\possible\textsc{CH} \wedge \possible\neg\text{CH})$,
+    the continuum hypothesis is a \hil{switch}.
+    \item $\necessary\possible\necessary(\text{$X$ is countable})$,
+    existence of an enumeration is a \hil{button}.
+  \end{itemize}
+\end{frame}}
+
+\begin{frame}{The modal topos-theoretic multiverse}
+  \textbf{Def.} A statement~$\varphi$ holds \ldots
+  \begin{enumerate}
+    \small
+    \item \hil{everywhere} ($\necessary\varphi$) iff it holds in every topos
+    (over the current base).
+    \item \hil{somewhere} ($\possible\varphi$) iff it holds in some positive topos.
+    \item \hil{proximally} ($\xpossible\varphi$) iff it holds in some positive overt topos.
+  \end{enumerate}
+  \bigskip
+
+  \begin{tikzpicture}[overlay]
+    \node[anchor=south east,inner sep=0] (image) at (14.8,0.2) {
+      \includegraphics[width=0.2\textwidth]{multiverse}
+    };
+  \end{tikzpicture}\vspace*{-2.3em}
+
+  \pause
+
+  \small
+  \begin{multicols}{2}
+    For every inhabited set~$X$, \emph{proximally} \\
+    there is an enumeration~$\NN \twoheadrightarrow X$.
+    \medskip
+
+    A quasiorder is well iff \emph{everywhere}, \\
+    every sequence is good.
+    \medskip
+
+    A ring element is nilpotent iff \\
+    all prime ideals \emph{everywhere} contain it.
+    \medskip
+
+    For every ring, \emph{proximally} \\ there is a maximal ideal.
+    \medskip
+
+    A relation is well-founded iff \emph{everywhere}, \\ there is no descending chain.
+    \medskip
+
+    \emph{Somewhere}, \\ the law of excluded middle holds.
+  \end{multicols}
+  \pause
+
+  \begin{block}{}
+    \justifying
+    \textbf{Prop.} Let~$(X,{\leq})$ be a well quasiorder.
+    Then~$({<})$, where~$x < y \equiv (x \leq y \wedge \neg(y \leq x))$,
+    is well-founded.
+  \end{block}
+  \vspace*{-0.5em}
+
+  \emph{Proof.} Everywhere, there can be no infinite descending chain, as any
+  such would also be good. \qed
+
+  Unrolling this proof gives a program~$\nabla\mathsf{Good}\,\varepsilon \to
+  \prod_{x:X} \mathsf{Acc}\,x$.
+\end{frame}
+
+\begin{frame}{The modal type-theoretic multiverse?}
+  \emph{As I understand it, \ldots}
+
+  For extensional type theory, similar picture as with toposes:
+  Given a model of extensional type theory and a Grothendieck site~$\mathcal{C}$ in it, there
+  is a model of extensional type theory deserving the name~``sheaf model
+  over~$\mathcal{C}$''. This sheaf model is just a thin translation layer of
+  the base model.
+  \pause
+
+  For HoTT-like theories, given a ``presheaf model'' we know how to construct a
+  ``sheaf model'', but we do not know how to construct presheaf models in the
+  first place.
+  \pause
+
+  Still, can pragmatically use~$\nabla$ and feel philosophically inspired.
+\end{frame}
+
 \end{document}

### 2023-11-30 slides-hottest2023.lagda.tex
Slides finished

@@
 \newcommand{\dnote}[1]{%
-  \begin{tabular}{@{}m{2em}@{}m{0.80\textwidth}@{}}%
+  \begin{tabular}{@{}m{2em}@{}m{0.83\textwidth}@{}}%
     \textdbend &#1%
@@
 
-  \textbf{Examples.} (classically)\ $(\NN,{\leq}),\ \ X \times Y,\ \ X^*,\ \ \mathrm{Tree}(X).$
+  \textbf{Examples.} (classically)\ \ $(\NN,{\leq}),\ \ X \times Y,\ \ X^*,\ \ \mathrm{Tree}(X).$
   \bigskip
@@
   \vspace*{-0.6em}
-  \emph{Proof.} The set
-  \[ I \defeq \{ i \in \NN \,|\, \neg\mathop{\exists}\limits_{j \in \NN} i < j \times \alpha\,i \leq \alpha\,j
-  \} \]
+  \emph{Proof.} The type
+  \[ I \defeq \sum_{i \? \NN} \neg\sum_{j \? \NN} i < j \times \alpha\,i \leq \alpha\,j \]
   cannot be in bijection with~$\NN$, as else the~$I$-extracted subsequence
-  of~$\alpha$ would not be good. By~\bad{\textsc{lem}}, the set~$I$ is finite.
+  of~$\alpha$ would not be good. By~\bad{\textsc{lem}}, the type~$I$ is finite.
   Any index~$i_0$ larger than all the indices in~$I$ is a suitable starting
@@
   \emph{Idea:} Approximate (fictitious) surjections~$\NN \twoheadrightarrow X$
-  by finite sequences~$\sigma = x_0 \ldots x_{n-1} \? X^*$. \\ Over time, such an
+  by finite sequences~$\sigma = x_0 \ldots x_{n-1} \? X^*$. \\
+  Starting with the empty sequence~$\varepsilon$, over time, such an
   approximation can grow to
@@
     Prop = Set
+    infixl 5 _∈_
     postulate
@@
         later₁ : ((y : X) → ∇ P (σ ∷ y)) → ∇ P σ
-        later₂ : (a : X) → ((τ : X*) → a ∈ (σ ⧺ τ) → ∇ P (σ ⧺ τ)) → ∇ P σ
+        later₂ : (a : X) → ((τ : X*) → a ∈ σ ⧺ τ → ∇ P (σ ⧺ τ)) → ∇ P σ
   \end{code}}
@@
   \vspace*{-0.8em}
-  For instance, for~$P\,x_0 \ldots x_{n-1} \defeq \exists_{k : \NN} (k < n
-  \times x_k = a)$, we have~$\nabla P \, \varepsilon$. \quad ``$\genalpha$ is surjective''
+  For instance, for~$P_a\,\sigma \defeq (a \in \sigma)$,
+  we have~$\nabla P_a \, \varepsilon$. \quad ``$\genalpha$ is surjective''
 \end{frame}
@@
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses}\end{minipage}}
-\begin{frame}{Slide about toposes}
-  (something about overloading of the logical connectives)
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
+\begin{frame}{Topos for reifying generic models}
+  Grothendieck toposes (= categories of sheaves over sites) are mathematical universes:
+  \begin{itemize}
+    \item The \hil{generic sequence} is an honest function~$\NN \to X$
+    \hil{in} a certain topos~$\mathcal{E}$.
+    \item The \hil{generic surjection} is an honest surjection \hil{in} an
+    appropriate subtopos of~$\mathcal{E}$.
+    \item In the \emph{double-negation subtopos} of the base, a predicate~$Q : \NN
+    \to \mathsf{Prop}$ such that
+    \[ Q\,x \times Q\,y \longrightarrow x = y \quad\text{and}\quad
+      \neg\neg \sum_{n\?\NN} Q\,n \]
+    looks like an ordinary natural number.
+  \end{itemize}
+
+  A statement holds in a topos iff a certain topos-directed translation holds
+  in the base. For instance, for the double-negation subtopos the translation
+  substitutes
+  \[\begin{array}{rcl@{\quad\text{where}\quad}rcl}
+    \exists & \leadsto & \exists^\mathrm{cl},
+    & \exists^\mathrm{cl} &\defeqv& \neg\neg\exists, \\
+    \vee & \leadsto & \vee^\mathrm{cl},
+    & \alpha \vee^\mathrm{cl} \beta &\defeqv& \neg\neg(\alpha \vee \beta), \\
+    = & \leadsto & =^\mathrm{cl},
+    & s =^\mathrm{cl} t &\defeqv& \neg\neg(s = t).
+  \end{array} \]
+  In general, stage-dependent~$\nabla$ instead of~$\neg\neg$.
 \end{frame}}
@@
     In the general case: \bad{No}\pause, but \good{yes} in a \emph{suitable topos}, and
-    \emph{bounded first-order consequences} of its existence there \good{do hold} in
+    \emph{bounded first-order consequences} of its existence there \good{pass down} to
     the base.
@@
   \jnote{8}{
-    J. Roitman, The uses of set theory, \emph{Math.\@ Intelligencer}
+    Judith Roitman. ``The uses of set theory''. \emph{Math.\@ Intelligencer}
     \textbf{14}(1) (1992), 63--69.
@@
 
+  \emph{Traveling the multiverse:}
+  \begin{itemize}
+    \item For every inhabited set~$X$, \emph{proximally} there is an enumeration~$\NN \twoheadrightarrow X$.
+    \item A quasiorder is well iff \emph{everywhere}, every sequence is good.
+    \item A ring element is nilpotent iff all prime ideals \emph{everywhere} contain it.
+    \item For every ring, \emph{proximally} there is a maximal ideal.
+    \item A relation is well-founded iff \emph{everywhere}, there is no descending chain.
+    \item \emph{Somewhere}, the law of excluded middle holds.
+  \end{itemize}
+  \pause
+
+  \only<1-3>{
     \small
-  \begin{multicols}{2}
-    For every inhabited set~$X$, \emph{proximally} \\
-    there is an enumeration~$\NN \twoheadrightarrow X$.
-    \medskip
-
-    A quasiorder is well iff \emph{everywhere}, \\
-    every sequence is good.
-    \medskip
-
-    A ring element is nilpotent iff \\
-    all prime ideals \emph{everywhere} contain it.
-    \medskip
-
-    For every ring, \emph{proximally} \\ there is a maximal ideal.
-    \medskip
-
-    A relation is well-founded iff \emph{everywhere}, \\ there is no descending chain.
-    \medskip
-
-    \emph{Somewhere}, \\ the law of excluded middle holds.
-  \end{multicols}
-  \pause
-
     \begin{block}{}
@@
     \prod_{x:X} \mathsf{Acc}\,x$.
+  }
+  \pause
+
+  \scriptsize
+  \emph{Foreshadowed by:}
+  \begin{itemize}
+    \item[1984\phantom{s}] André Joyal, Miles Tierney. ``An extension of the Galois theory of Grothendieck''.
+    \item[1987\phantom{s}] Andreas Blass. ``Well-ordering and induction in intuitionistic logic and topoi''.
+    \item[2010s] Milly Maietti, Steve Vickers. Ongoing work on arithmetic universes.
+    \item[2011\phantom{s}] Joel David Hamkins. ``The set-theoretic multiverse''.
+    \item[2013\phantom{s}] Shawn Henry. ``Classifying topoi and preservation of higher order logic by geometric morphisms''.
+  \end{itemize}
 \end{frame}
 
-\begin{frame}{The modal type-theoretic multiverse?}
-  \emph{As I understand it, \ldots}
-
-  For extensional type theory, similar picture as with toposes:
-  Given a model of extensional type theory and a Grothendieck site~$\mathcal{C}$ in it, there
-  is a model of extensional type theory deserving the name~``sheaf model
-  over~$\mathcal{C}$''. This sheaf model is just a thin translation layer of
-  the base model.
-  \pause
-
-  For HoTT-like theories, given a ``presheaf model'' we know how to construct a
-  ``sheaf model'', but we do not know how to construct presheaf models in the
-  first place.
-  \pause
+\begin{frame}{Towards the modal type-theoretic multiverse}
+  \small
+  \good{\cmark{}} \emph{There are type-theoretic multiverses,} such as
+  \begin{itemize}
+    \item the collection of all $\mathrm{PSh}(\mathcal{C} \times \mathcal{B})$,
+    where~$\mathcal{B}$ ranges over cube categories and~$\mathcal{C}$ over
+    arbitrary small categories, and their corresponding sheaf models
+
+    {\scriptsize
+    Coquand. ``A survey of constructive presheaf models of univalence''. \emph{ACM SIGLOG News}, 5.3 (2018).
+    \par}
+  \end{itemize}
+  \medskip
+  \pause
+
+  \bad{\xmark} \emph{Accessing the multiverse from within type theory is tricky:}
+  \begin{itemize}
+    \item Given a model of $\mathfrak{s}$CIC and a category~$\mathcal{C}$ in
+    it, we have a syntactic presheaf model of CIC.
+
+    {\scriptsize
+    Coquand, Jaber. ``A note on forcing and type theory''. \emph{Fundamenta Informaticae 100} (2010).
+    Jaber, Lewertowski, Pédrot, Sozeau, Tabareau. ``The definitional side of the forcing''. \emph{Proceedings of LICS '16} (2016). \\
+    Pédrot. ``Russian constructivism in a prefascist theory''. \emph{Proceedings of LICS '20} (2020). \\
+    \par}
+    \item Given a suitable lex modality, we have a syntactic sheaf model (model of modal types).
+
+    {\scriptsize
+    Coquand, Ruch, Sattler. ``Constructive sheaf models of type theory.'' \emph{Math. Struct. Comput. Sci.} 31.9 (2021).
+    Escardó, Xu. ``Sheaf models of type theory in type theory''. Unpublished (2016). \\
+    Quirin. ``Lawvere--Tierney sheafification in Homotopy Type Theory''. PhD thesis (2016). \\
+    \par}
+    \item (I believe) we have syntactic sheaf models in certain special cases,
+    when no coherence issues arise in defining the notion of presheaves.
+  \end{itemize}
 

### 2023-12-31 algar-lecture-notes.tex
typo

@@
 don't have a basis or that the notion of measurable subsets is not trivial.
-Such results also tend to not influence concrete situations too much: How could
+On the other hand,
+such results also tend to not influence concrete situations too much: How could
 an infinite basis for which there is no hope that we ever learn any actual

### 2024-01-11 slides-hottest2023.lagda.tex
typo

@@
   is well, there are indices~$n < m$ such that~$\beta\,i_n \leq \beta\,i_m$. As
-  also~$\alpha\,i_n \leq \alpha\,i_m$, the sequence~$\gamma$ is good.
+  also~$\alpha\,i_n \leq \alpha\,i_m$, the sequence~$\gamma$ is good. \qed
   \medskip
@@
   \dnote{\emph{We cannot trust~\bad{\textsc{lem}}-provided sequences to be
-  available in the type~$\NN \to \NN$. Similarly with~\bad{\textsc{dc}}.}}
+  available in the type~$\NN \to X$. Similarly with~\bad{\textsc{dc}}.}}
 \end{frame}
@@
   For instance, for~$P_a\,\sigma \defeq (a \in \sigma)$,
-  we have~$\nabla P_a \, \varepsilon$. \quad ``$\genalpha$ is surjective''
+  we have~$\nabla P_a \, \varepsilon$. \quad ``$\genalpha$ is surjective'':
+  \AgdaSymbol{(}\AgdaBound{a}\AgdaSpace{}%
+  \AgdaSymbol{:}\AgdaSpace{}%
+  \AgdaPostulate{X}\AgdaSymbol{)}\AgdaSpace{}%
+  \AgdaSymbol{→}\AgdaSpace{}%
+  \AgdaDatatype{∇}\AgdaSpace{}%
+  \AgdaBound{P$_a$}\AgdaSpace{}%
+  \AgdaBound{$\varepsilon$}
 \end{frame}
@@
   \end{block}
+  \pause
 
@@
     \to \mathsf{Prop}$ such that
-    \[ Q\,x \times Q\,y \longrightarrow x = y \quad\text{and}\quad
+    \[ Q\,x \times Q\,y \,\longrightarrow\, x = y \quad\text{and}\quad
       \neg\neg \sum_{n\?\NN} Q\,n \]
@@
       %to settle~\textsc{ch}.}
-      \ytl{2016}{Oldenziel proposes to study the modal multiverse of parametrized
-      mathematics.}
+      \ytl{2016}{Oldenziel proposes to study the modal multiverse of toposes.}
+      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
+      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
       \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
@@
     {\scriptsize
-    Coquand, Jaber. ``A note on forcing and type theory''. \emph{Fundamenta Informaticae 100} (2010).
+    Coquand, Jaber. ``A note on forcing and type theory''. \emph{Fundamenta Informaticae 100} (2010). \\
     Jaber, Lewertowski, Pédrot, Sozeau, Tabareau. ``The definitional side of the forcing''. \emph{Proceedings of LICS '16} (2016). \\
-    Pédrot. ``Russian constructivism in a prefascist theory''. \emph{Proceedings of LICS '20} (2020). \\
+    Pédrot. ``Russian constructivism in a prefascist theory''. \emph{Proceedings of LICS '20} (2020).
     \par}
@@
     {\scriptsize
-    Coquand, Ruch, Sattler. ``Constructive sheaf models of type theory.'' \emph{Math. Struct. Comput. Sci.} 31.9 (2021).
+    Coquand, Ruch, Sattler. ``Constructive sheaf models of type theory.'' \emph{Math. Struct. Comput. Sci.} 31.9 (2021). \\
     Escardó, Xu. ``Sheaf models of type theory in type theory''. Unpublished (2016). \\
-    Quirin. ``Lawvere--Tierney sheafification in Homotopy Type Theory''. PhD thesis (2016). \\
+    Quirin. ``Lawvere--Tierney sheafification in Homotopy Type Theory''. PhD thesis (2016).
     \par}

### 2024-01-11 paper-flabby-objects.tex
to do
regen pdf
typo

@@
 
-Trivially, every group satisfying the Baer condition is also divisible.
+Every group satisfying the Baer condition is divisible, by specializing~$\aaa$
+to the ideal~$(n)$ and using multiplication by~$x$ as the map~$\aaa \to I$.
 
@@
 divisible group~$\QQ\langle A \rangle / K$, where~$\QQ\langle A \rangle$ is the
-underlying abelian group of the free~$\QQ$-vector space on~$A$ and~$K$ is the
+underlying abelian group of the free~$\QQ$-vector space on the underlying set of~$A$ and~$K$ is the
 kernel of the canonical map~$\ZZ\langle A \rangle \to A$.
@@
 resolutions
+
+https://arxiv.org/abs/1903.01211

### 2024-08-11 paper-reflection.tex
to do

@@
 \end{document}
+
+Cite: https://link.springer.com/chapter/10.1007/BFb0061825
+
+Cite: https://arxiv.org/pdf/1110.2430.pdf (page 10)
+
+https://www.youtube.com/watch?v=lpOiuvzbBtw&list=PL6tK2CutTH9_lN7A0nFul_gEJUDSPJ91i&index=3
+Timestamp 29:49
+Aczel showed that RRS is stable under passage to sheaves on complete Heyting
+algebras.

### 2024-09-27 paper-modal-multiverse.tex
to do
Sync with Overleaf version

@@
 \newcommand{\xra}[1]{\xrightarrow{#1}}
+\newcommand{\xxx}[1]{}
 \newcommand{\XXX}[1]{\textbf{\textcolor{red}{XXX: #1}}}
@@
 
+\author{Ingo Blechschmidt\inst{1} \and Alexander Gietelink Oldenziel\inst{2}}
+\institute{Arberstr. 5, 86159 Augsburg, Germany
+\email{iblech@speicherleck.de}
+\and University College London, Gower Street, London, WC1E 6BT
+\email{a.f.d.a.gietelinkoldenziel@gmail.com}}
+
 \maketitle
+
+\xxx{explain somewhere that just as $\neg\neg$ is visible constructively but not classically and that $\neg\neg$ enriches our understanding, so do the new modal operators.
+Especially in view of the fact that if LEM holds, then $\varphi \Leftrightarrow \necessary\varphi$ for bounded first-order formulas, as in the presence of LEM every topos is overt.}
+
+\xxx{state (and ponder more) the example that every field has a separable closure somewhere}
+
+\xxx{Alexander:
+- it is not the case that for every topos E and every field k in E there is a separable closure k_s / k i don't think. what is a counterexample?
+- it should be the case that for every (geometric field?)
+-> it is positive (since it the etale topos of a field is Galois over the base so the pullback reflects isomorphisms I think: the pullback is a constant sheaf functor which sends a set to the trivial G-set on that set.)}
 
@@
 chain.
-\item A relation is almost-full iff every infinite sequence is good (meaning
-that some term on the sequence is related to some later term -- the notion
+\item A relation is almost-full iff every infinite sequence is good.\footnote{An infinite sequence is \emph{good} iff some term of the sequence is related to some later term. The notion
 of almost-full relations
-has been studied in combinatorics~\cite{xxx} and found applications in
-termination checking~\cite{xxx}).
+has been studied in combinatorics~\cite{vytiniotis-coquand-wahlstedt:af,carroy-pequignot:well,lescanne:well} and found applications in
+termination checking~\cite{blass-gurevich:well}.}
 \item \emph{Krull's lemma}: A ring element is nilpotent iff every prime
@@
 extracted~\cite{bauer:c2c}, and every constructive proof holds also for continuous families of
-the objects in question~\cite[Section~4.3]{blechschmidt:filmat}. In contrast, listed classical principles above have no
+the objects in question~\cite[Section~4.3]{blechschmidt:filmat}. In contrast, the listed classical principles above have no
 computational witness and/or fail in continuous families, hence are not
@@
 
-In the modal topos-theoretic multiverse, we have the following constructive
-replacements to these principles.
+In the modal topos-theoretic multiverse, we have the following constructive replacements to these principles.
 \begin{enumerate}
@@
 ($\necessary\varphi$) iff it holds in every Grothendieck topos over the
-current base topos (makng use of the \emph{internal language} of
-toposes~\cite{xxx}, where required in the form of the \emph{sheaf semantics} in
-order to support unbounded quantification~\cite{xxx}); a statement holds \emph{somewhere} ($\possible\varphi$)
+current base topos (making use of the \emph{internal language} of
+toposes~\cite{borceux:handbook3,streicher:ctcl,maietti:modular});\footnote{In order to support unbounded quantification we occasionally make use of the extension of the usual Kripke--Joyal semantics in the form of \emph{stack semantics}~\cite{shulman:stack-semantics})} a statement holds \emph{somewhere} ($\possible\varphi$)
 iff it holds in some positive Grothendieck topos over the current base; and a
 statement holds \emph{proximally} ($\xpossible\varphi$) iff it holds in
-some positive ouvert Grothendieck topos over the current base. More such
+some positive overt Grothendieck topos over the current base. More such
 \emph{modalities} are also useful and merit study; the precise definitions are
@@
 everywhere the case that it (and even the full axiom of choice) hold somewhere.
-\item If a geometric implication holds \emph{somewhere}, then it holds already here.
+\item If a geometric sequent $\wedge_{i=1}^n \phi_i \vdash \bigvee_{j\in J}\psi_j$ holds \emph{somewhere}, then it holds already here.
 \item \label{item:prox-descends}
-If a first-order statement holds \emph{proximally}, then it holds already here.
+If a bounded first-order statement holds \emph{proximally}, then it holds already here.
 \item \label{item:prox-countable}
@@
 has recently been studied in constructive
-commutative algebra~\cite{blechschmidt-schuster:constructive-maximal-ideals}.
-For countable rings, an explicit construction of a maximal ideal is available.
-By Item~(\ref{item:prox-countable}), this construction can also be carried out
+commutative algebra~\cite{blechschmidt-schuster:constructive-maximal-ideals}:
+For countable rings, an explicit iterative construction of a maximal ideal is available.
+By Item~\ref{item:prox-countable}, this construction can also be carried out
 for arbitrary rings, though the result is not a maximal ideal in the narrow
 sense; rather, the resulting maximal ideal exists \emph{proximally}, in some
-positive ouvert Grothendieck topos. However, the first-order consequences of
+positive overt Grothendieck topos. However, the first-order consequences of
 its existence, pertaining for instance to concrete statements about polynomials
-or matrices, pass down to the base topos by Item~(\ref{item:prox-descends}).
+or matrices, pass down to the base topos by Item~\ref{item:prox-descends}.
 The resulting maximal ideal can be regarded as a mathematical
-phantom in the sense of Gavin Wraith~\cite{xxx}, not existing in a narrow
+phantom in the sense of Gavin Wraith~\cite{wraith:phantoms}, not existing in a narrow
 sense (in the base topos), but existing proximally and hence encouraging us to broaden our notion of
@@
 We can also adopt the notions of \emph{switches} and \emph{buttons} from the
-modal study of the set-theoretic multiverse~\cite[xxx]{xxx}. Switches are
+modal study of the set-theoretic multiverse~\cite{hamkins:modal}. Switches are
 statements~$\varphi$ such that~$\necessary(\possible\varphi \wedge
-\possible\neg\varphi)$ and buttons are statements~$\varphi$ such
+\possible\neg\varphi)$, while buttons are statements~$\varphi$ such
 that~$\necessary\possible\necessary\varphi$; switches can be toggled on and off
@@
 \item \emph{Being countable is a button:} For every set~$X$, everywhere it is
-the case that somewhere (even proximally) it is case that everywhere~$X$ is
+the case that somewhere (even proximally) it is the case that everywhere~$X$ is
 countable.
@@
 
-We argue that the modal operators~$\necessary,\possible,\xpossible$ and more
-suggested are natural extensions and refinements of the
+We argue that the proposed modal operators~$\necessary,\possible,\xpossible$ and more are natural extensions and refinements of the
 familiar double-negation modality~$\neg\neg$ in constructive mathematics.
 
-\textbf{Acknowledgments.} Alexander Oldenziel, \ldots
-% xxx: Iosif
+\paragraph{Related work.}
+The idea of a mathematical multiverse is arguably basic to topos theory.
+We aim to present a more systematic study of the modal nature of the multiverse with a focus on applications in constructive mathematics.
+Specific precursors to XXX
 
-xxx: idea of multiverse not new, but arguably basic to topos theory; new is systematic
-study of its modal nature with the focus on applications
+In the airy reaches of classical set theory a related philosophy is put forward, with a number of striking results, concentrating on exploring the range of set-theoretic possibility~\cite{hamkins:multiverse,gitman-hamkins:multiverse,hamkins:constructibility,maddy-meadows:reconstruction,barton:forcing}.
 
-xxx: related work: Hamkins, Victoria Gitman, ... concentrates on exploring
+\xxx{related work: Hamkins, Victoria Gitman, ... concentrates on exploring
 range of set-theoretic possibility, obtaining beautiful results such as ...
-Shawn Henry, Andreas Blass ...
+Shawn Henry, Andreas Blass ...}
 
-xxx: This text is set in the context of constructive mathematics.
-Our definitions and results can be formalized in~\textsc{izf} or in the
-kind of language supported by toposes.
+This text is set in the context of constructive mathematics.
+Our definitions and results can be formalized in~\textsc{izf} or in the kind of language supported by toposes.
+
+\paragraph{Acknowledgments.} We gratefully acknowledge numerous discussions and worthwhile input by Thierry Coquand, Matthias Hutzler, Ivan Di Liberti, Milly Maietti, Iosif Petrakis, Peter Schuster, and also the organizers and participants of the ItaCa Fest 2022 and the 2022 REDCOM workshop in Brixen, where this work was presented.
+XXX
 
@@
 concept is available in one's ontology, or also some other elementary topos
-such as the free topos~\cite{xxx} or the ef{}fective topos~\cite{xxx}.
+such as the free topos~\cite{lambek:1980intuitionist} or the ef{}fective topos~\cite{hyland:effective-topos}.
+The reader is reminded that the nature of the base topos can have profound consequences for the truths of the internal language.
 
-The xxx (use word of Hamkins) of a base~$\B$ is the collection of Grothendieck
+The topos-theoretic multiverse of a base~$\B$ is the collection of Grothendieck
 toposes over~$\B$ (which in set-theoretic foundations should more precisely be
@@
 possibly unbounded geometric morphisms~$\E \to \B$, or even arbitrary
-fibrations/indexed categories over~$\B$ (XXX: are these the same?) validating an appropriate form of the
+fibrations/indexed categories over~$\B$ validating an appropriate form of the
 axioms of elementary toposes, should also be
-taken as part of the topos-theoretic multiverse. The first foray into the
+taken as part of the topos-theoretic multiverse. [XXX check fibrational characterization] The first foray into the
 modal topos-theoretic multiverse outlined in this note
@@
 
-\begin{definition}A topos~$\E$ is \emph{positive} if and only if the geometric
+\begin{definition}A topos~$\E$ is \emph{positive} if and only if the unique geometric
 morphism~$f : \E \to \Set$ is surjective (that is~$f^*$ reflects
@@
 
-\begin{example}A sufficient criterion for the classifying topos of a geometric
-theory~$\TT$ to be positive is that xxx.\end{example}
+\begin{example}A necessary and sufficient criterion for the classifying topos of a geometric
+theory~$\TT$ to be positive is that whenever~$\top \vdash \bigvee_{i \in I} \varphi_i$ modulo~$\TT$, then~$I$ is inhabited.\end{example}
 
-\subsection{Ouvert toposes}
+\subsection{Overt toposes}
 
-\begin{definition}A topos~$\E$ is \emph{ouvert} if and only if the geometric
-morphism~$f : \E \to \Set$ is open, that is~$f^*$ preserves the interpretation of
-first-order formulas.\end{definition}
+\begin{definition}A topos~$\E$ is \emph{overt} if and only if the unique geometric
+morphism~$f : \E \to \Set$ is open (that is~$f^*$ preserves the interpretation of
+bounded first-order formulas).\end{definition}
 
-\begin{example}The topos of sheaves over a topological space is always ouvert. More
-generally, the topos of sheaves over a locale~$X$ is ouvert iff there exists a
+\begin{example}The topos of sheaves over a topological space is always overt. More
+generally, the topos of sheaves over a locale~$X$ is overt iff there exists a
 \emph{positivity predicate} on its frame of opens in the sense of~\cite{xxx}.
-With~\textsc{lem}, every locale is ouvert~\cite{xxx}.\end{example}
+With~\textsc{lem}, every locale and indeed every topos is overt~\cite{xxx}.\end{example}
 
-\begin{example}The spectrum of a ring~$A$ is ouvert iff every element of~$A$ is
+\begin{example}The spectrum of a ring~$A$ is overt iff every element of~$A$ is
 nilpotent or not~\cite[Proposition~12.51]{blechschmidt:phd}.\end{example}
@@
 \begin{example}A sufficient criterion for the classifying topos of a geometric
-theory~$\TT$ being ouvert is that the indexing sets of all disjunctions
+theory~$\TT$ being overt is that the indexing sets of all disjunctions
 appearing on the right hand side of turnstiles, in a normal form presentation
-of~$\TT$, are inhabited~\cite{xxx}.\end{example}
+of~$\TT$, are inhabited~\cite[Proposition~V.3.2]{joyal-tierney:grothendieck}.\end{example}
 
@@
 
-xxx: properly define language
+\xxx{properly define language}
 
@@
 infinite sequence~$\alpha : \NN \to X$ is \emph{good} in the sense that there
-exist numbers~$i < j$ such that~$\alpha(i) \mathrel{R} \alpha(j)$''.
+exist indices~$i < j$ such that~$\alpha(i) \mathrel{R} \alpha(j)$''.
 For a constructive reformulation, we shift to finite approximations of
@@
 at position~$n$ of the finite list~$\sigma$. This notation is only meaningful
-if the length of~$\sigma$ is at least~$n+1$. By~``$\sigma :: x$'' we mean the
-enlarged list which has~$x$ as an extra element at its tail end, and by~``$[]$'' we
+if the length of~$\sigma$ is at least~$n+1$. By~``$\sigma x$'' we mean the
+enlarged list which has~$x$ as an extra element at its tail end, and by~``$[\,]$'' we
 denote the empty list. In computer science practice, it is often more efficient
-to prepend (``$x :: \sigma$'') instead of append, but this detail shall not
+to prepend (``$x \sigma$'') instead of append, but this detail shall not
 concern us here.}
@@
   \qquad
-  \infer{P \,|\, \sigma}{\forall x \? X\_\ P \,|\, (\sigma :: x)}
+  \infer{P \,|\, \sigma}{\forall x \? X\_\ P \,|\, \sigma x}
 \]
@@
 
-A similar such definition has been proposed by XXX and XXX in commutative
-algebra for expressing that a ring is Noetherian~\cite{xxx}; the classical
+A similar such definition has been proposed by Thierry Coquand, Henry Lombardi and Henrik Persson in commutative
+algebra for expressing that a ring is Noetherian~\cite{coquand-persson:groebner,coquand-lombardi:krull,coquand:invariant}; the classical
 definition ``every ascending chain of ideals stabilizes'' and also the more
@@
 chain of finitely generated ideals stalls''\footnote{A chain~$\aaa_0 \subseteq
-\aaa_1 \subseteq \cdots$ \emph{stalls} iff for some number~$n \in \NN$,
+\aaa_1 \subseteq \cdots$ \emph{stalls} iff for some index~$n \in \NN$,
 $\aaa_{n+1} = \aaa_n$. We are grateful to Matthias Hutzler for proposing this
 terminology.} are constructively too weak; firstly, without the axiom of
-dependent choice we can often not construct such chains~\cite{richman:xxx} (but only
+dependent choice we can often not construct such chains~\cite{richman:noetherian} (but only
 ``multi-valued chains'' as in~\cite[Section~3.9]{blechschmidt:phd}; but also
-see~\cite{richman:xxx}), and secondly, being able to inspect suitable inductive
-witnesses enables us to prove the Hilbert basis theorem~\cite{xxx}. XXX and XXX
+see~\cite[Section~4]{richman:noetherian}), and secondly, being able to inspect suitable inductive
+witnesses enables us to prove the Hilbert basis theorem~\cite[Corollary~16]{coquand-persson:groebner}. Coquand, Lombardi and Persson
 hence propose to call a ring \emph{Noetherian} if and only if~$\mathsf{Stalls}
-\,|\, []$, where~$\mathsf{Stalls}$ is the predicate on finite lists of finitely
+\,|\, [\,]$, where~$\mathsf{Stalls}$ is the predicate on finite lists of finitely
 generated ideals expressing xxx.
@@
 partially ordered set of finite lists of elements of~$X$ with coverage given
-by xxx (see Appendix xxx). The Kripke--Joyal semantics states that the
+by xxx (see Appendix xxx)~\cite[Example~4.3]{blechschmidt-schuster:reifying}, \cite[xxx]{joyal-tierny:xxx}. The Kripke--Joyal semantics states that the
 statement ``$\alpha_0$ is good'', where~$\alpha_0$ is the generic infinite
@@
 for the variant of ``bad sets'' instead of ``infinite descending chains'' have
-been developed (xxx:language) by xxx~\cite{xxx}.\end{proof}
+been developed (xxx:language) by Blass~\cite{blass:induction}.\end{proof}
 
@@
 
-    \item By \textsc{lem} and well-foundedness, there is a minimal value~$\alpha(i_0)$
-    among all values of~$\alpha$.
-    Similarly, there is a minimal value~$\alpha(i_1)$ among~$(\alpha(n))_{n >
-    i_0}$, a minimal value~$\alpha(i_2)$ among~$(\alpha(n))_{n > i_1}$, and so
-    on. By \emph{proximal dependent choice}, we can proximally collect these indices
-    into a function~$i : \NN \to \NN$;\footnote{We could also avoid dependent
-    choice by picking not some suitable indices~$i_0, i_1, \ldots$, but the
-    least such.} this switches~\textsc{lem} off.
+    \item (A ``Ramsey-style argument''.) The set~$I \defeq \{ n \in \NN \,|\, \neg\exists m > n\_ \alpha(n) \leq \alpha(m) \}$ is not in bijection with~$\NN$, as else the~$I$-extracted subsequence of~$\alpha$ would be an $X$-sequence which is not good.
+Hence, by \textsc{lem}, the set~$I$ is finite.
+Every index larger than all the indices in~$I$ is a suitable starting point for an infinite ascending chain~$\alpha(i_0) \leq \alpha(i_1) \leq \ldots$.\footnote{We can avoid dependent
+    choice here by always picking the
+    least possible next index.}
 
-    \item Switching~\textsc{lem} on again, there is a minimal value~$\beta(i(k_0))$
-    among all values of~$\beta \circ i$. Hence~$\gamma$ is good in view of
-    \begin{align*}
-      \alpha(i(k_0)) &\leq \alpha(i(k_0+1)), &
-      \beta(i(k_0)) &\leq \beta(i(k_0+1)).
-    \end{align*}
+    \item Because~$Y$ is almost-full, the sequence~$\beta(i_0),\beta(i_1),\ldots$ is good, that is there exists a pair of indices~$n < m$ such that~$\beta(i_n) \leq \beta(i_m)$.
+As also~$\alpha(i_n) \leq \alpha(i_m)$, the sequence~$\gamma$ is good.
+    % XXX: this only holds in the transitive case
   \end{enumerate}
@@
 
-xxx: determine modal laws
+\xxx{determine modal laws}
 
-xxx: more case studies
+\xxx{more case studies}
 
-xxx: don't forget right adjoints
+\xxx{don't forget right adjoints}
 
-xxx: predicative
+\xxx{predicative}
 
 \bibliographystyle{splncs04}
-\bibliography{mapaper-modal-multiverse}
+\bibliography{paper-modal-multiverse}
 
@@
 cite https://math.andrej.com/2012/10/03/am-i-a-constructive-mathematician/
+
+simpler logical form
+(every prime ideal vs. nilpotent)
+
+cite https://www.cse.chalmers.se/~coquand/potential.pdf page 41
+
+cite https://webspace.science.uu.nl/~ooste110/veldmanabstract.pdf
+("perhaps" operator)
+
+cite:
+https://arxiv.org/pdf/2108.10259.pdf
+https://arxiv.org/abs/2311.00581
+
+cite:
+Slides-What-if-your-potentialism-is-implicitly-actualist-Oxford-March-2024.pdf, slide 74 (Joel David Hamkins)
+
+check:
+https://www.youtube.com/watch?v=TpPQHS_wUsg, Giraud topos
+
+cite:
+https://arxiv.org/abs/2206.03169
+
+because of representation theorem,
+to say "for all Grothendieck toposes",
+suffices to say "for all localic toposes".
+Similarly for overt and positive
+
+cite
+https://arxiv.org/pdf/1712.04864
+for internal language of toposes
+
+show that inductive-wqo implies Berardi-wqo
+show the theorem displayed at https://ncatlab.org/nlab/show/well-quasi-order
+
+https://math.stackexchange.com/questions/4552108/does-the-inverse-image-part-of-a-geometric-morphism-preserve-transitive-set-obj
+wf preserved
+
+https://mathoverflow.net/questions/288789/constructive-proofs-of-existence-in-analysis-using-locales/290421?noredirect=1#comment1248291_290421
+Simon suggests "proper and surjective" as an interesting class.
+
+Completeness: If sequent holds for all models everywhere, then provable.

### 2024-11-30 slides-abmv2024.tex
Talk for ABMV 2024

--- 2023-11-29 slides-hottest2023.lagda.tex
@@
-\documentclass[11pt,utf8,notheorems,compress,t,aspectratio=169]{beamer}
+\documentclass[12pt,utf8,notheorems,compress,t,aspectratio=169]{beamer}
 \usepackage{etex}
@@
 
-\usepackage[normalem]{ulem}
 \usepackage{mathtools}
+\usepackage{xspace}
 \usepackage{booktabs}
@@
 
-\title{Towards multiversal modal operators for homotopy type theory}
+\title{Modal operators for a constructive account of well quasi-orders}
 
 \author{Ingo Blechschmidt}
-\date{October 30th, 2023}
+\date{November 30th, 2024}
 
@@
 \newcommand{\forces}{\vDash}
+\newcommand{\ind}{\ensuremath{_\text{ind}}\xspace}
+\newcommand{\sinf}{\ensuremath{_\infty}}
+\newcommand{\impl}{\ensuremath{_\text{impl}}\xspace}
 
-\setbeamertemplate{blocks}[rounded][shadow=false]
+\setbeamertemplate{blocks}[shadow=false]
 
@@
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\includegraphics[height=\paperheight]{multiverse-faded}\end{minipage}}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\includegraphics[width=\paperwidth]{tower}\end{minipage}}
 \begin{frame}[c]
@@
   \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
 
   \scriptsize
-  \textit{-- an invitation --}
 
   \setbeamercolor{block body}{bg=black!100}
-  \begin{minipage}{0.49\textwidth}
-    \begin{block}{}
+  \begin{minipage}{0.62\textwidth}
+    \begin{block}{\centering Arbeitstagung Bern--München--Verona}
       \centering\normalsize\color{white}
-      \hil{Towards multiversal modal operators for homotopy type theory}
+      \hil{Modal operators for \\ a constructive account of
+      well quasi-orders} \\[-0.9em]
+      \
     \end{block}
@@
   \bigskip
+  \bf
+  \colorbox{black}{November 30th, 2024}%
   \bigskip
-  \bigskip
-  \bigskip
+
+  \colorbox{black}{\begin{minipage}{0.2\textwidth}
+    \centering
+    Ingo Blechschmidt \\
+    University of Antwerp
+  \end{minipage}}
 \end{frame}}
@@
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[height=\paperheight]{sea-of-clouds-2}\end{minipage}}
-\begin{frame}{How should this notion be formalized in HoTT?}
-  \jnote{5-}{
-    The presented proof rests on the law of excluded middle and hence cannot
-    immediately be interpreted as a program for finding suitable indices~$i < j$.
-    However, constructive proofs are also possible (for instance by
-    induction on the value of a given term of the sequence), and furthermore
-    constructive proofs can be extracted from the presented classical proof.
-  }
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{2.1em}\includegraphics[height=\paperheight]{sea-of-clouds-1}\end{minipage}}
+\begin{frame}{Well quasi-orders}
+  \textbf{Def.} Let~$(X,{\leq})$ be a quasi-order.
+  \vspace*{-0.3em}
+  \begin{itemize}
+    \item A sequence~$\alpha : \NN \to X$ is \hil{good} iff
+    there exist~$i < j$ with~$\alpha\,i \leq \alpha\,j$.
+    \item The quasi-order~$X$ is \hil{well\only<4->{\sinf}} iff every sequence~$\NN \to X$ is good.
+  \end{itemize}
+  \pause
+  \vspace*{-0.9em}
 
-  \jnote{7-}{
-    The class of well quasiorders is stable under cartesian products, lists and
-    trees, by Dickson's Lemma, Higson's Lemma and Kruskal's Theorem,
-    respectively. However, in their naive formulations, these are merely
-    theorems of classical mathematics. For general constructive results, the
-    definition of ``well'' needs to be improved.
+  \begin{columns}[t]
+    \begin{column}{0.46\textwidth}
+      \begin{block}{Natural numbers\phantom{y}}
+        \textbf{Prop.} $(\NN, {\leq})$ is well\only<4->{\sinf}.
+        \smallskip
 
-    Classical texts often employ needless negations in their definitions; this
-    is an XXX
-  }
+        \emph{Proof.} Let~$\alpha : \NN \to \NN$. By~\badbox{\textsc{lem}}, there is a
+        \bad{minimum}~$\alpha\,i$.
+        Set~$j \defeq i + 1$. \qed
 
-  \vspace*{-1em}
-  \[ \astikznodetransparentlycircled{xm}{7}\!,
-    \quad \astikznodetransparentlycircled{x0}{4}\!,
-    \quad \only<1-2>{\astikznodetransparentlycircled{t1}{3}}\only<3->{\astikznodecircled{t1}{mypurple}{3}}\!,
-    \quad \only<1>{\ldots}\pause \astikznodetransparentlycircled{x1}{1}\!,
-    \quad \only<2>{\ldots}\pause \astikznodecircled{t2}{mypurple}{8}\!,
-    \quad \only<3>{\ldots} \visible<4->{\astikznodetransparentlycircled{x2}{2}\!,}
-    \quad \only<4->{\ldots} \]
-  {\centering\begin{tikzpicture}[remember picture,overlay]
-    \node[draw=mypurple, circle, thick, inner sep=0.1em] (t3) {\scriptsize$\leq$};
-    \path[draw=mypurple,thick]
-      (t1)
-      to [out=-90, in=180] (t3)
-      to [out=0, in=-90] (t2);
-  \end{tikzpicture}\par}
-  \medskip
+        \hfill\color{gray}offensive?
+      \end{block}
+    \end{column}
     \pause
 
-  \begin{block}{}
+    \begin{column}{0.50\textwidth}
+      \begin{block}{Key stability results}
         \justifying
-    \textbf{Thm.} Every sequence~$\alpha : \NN \to \NN$ is \hil{good} in that
-    there exist~$i < j$ with~$\alpha\,i \leq \alpha\,j$.
+        \only<1-5>{Assuming~\badbox{\textsc{lem}}
+        and~\badbox{\textsc{dc}}}\only<6->{Constructively}, \ldots
+
+        \hil{Dickson:} \tabto{1.8cm} If~$X$ and~$Y$ are well\only<4-5>{\sinf}\only<6->{\ind}, so is~$X \times Y$. \\
+        \hil{Higman:}  \tabto{1.8cm} If~$X$ is well\only<4-5>{\sinf}\only<6->{\ind}, so is~$X^\star$. \\
+        \hil{Kruskal:} \tabto{1.8cm} If~$X$ is well\only<4-5>{\sinf}\only<6->{\ind}, so is~$\mathrm{Tree}(X)$.
       \end{block}
+    \end{column}
+  \end{columns}
   \pause
-  \vspace*{-0.6em}
-  \emph{Proof.} \emph{(offensive?)} By~\badbox{\textsc{lem}}, there is a
-  minimum~$\alpha\,i$.
-  Set~$j \defeq i + 1$. \qed\par
   \pause
 
-  \textbf{Def.} (classically) A quasiorder~$X$ is \hil{well} iff every sequence~$\NN \to X$ is good.
-
-  \textbf{Examples.} (classically)\ $(\NN,{\leq}),\ \ X \times Y,\ \ X^*,\ \ \mathrm{Tree}(X).$
-  \bigskip
+  \textbf{Def.} A quasi-order~$X$ is \hil{well\ind} iff~$\mathsf{Good} \mid
+  [\,]$, where~$\mathsf{Good}\,x_0 \ldots x_{n-1}$ iff~$\exists i < j\_ x_i \leq x_j$.
+  \pause
   \pause
 
-  \emph{A naive formalization attempt:}
+  \small
+  With \bad{bar induction},\tabto{3.10cm} $\text{well}\ind \Leftarrow \text{well}\sinf$. \\[-0.2em]
+  Constructively,\tabto{3.10cm} $\text{well}\ind \Rightarrow \text{well}\sinf$.
+  \only<8>{
+    \bigskip
+    \begin{columns}[c]
+      \begin{column}{0.01\textwidth}
+	\includegraphics[height=2.4em]{question-mark}
+      \end{column}
+      \begin{column}{0.9\textwidth}
+      Is there a procedure for reinterpreting \hil{classical proofs}
+      regarding well\sinf{} as \newline \hil{blueprints for constructive proofs} regarding well\ind?
+      \end{column}
+    \end{columns}
+  }
+  \pause
+  \pause
+  Moreover, if~$X$ is well\ind, then \ldots
+  \begin{itemize}
+  \vspace*{-0.7em}
+    \small
+    \item for every \emph{partial} function~$\alpha$,\tabto{5.25cm}
+    if~$\forall n\_ \neg\neg(\alpha\,n\,{\downarrow})$,
+    then~$\neg\neg \exists i < j\_ \alpha\,i\,{\downarrow} \wedge
+    \alpha\,j\,{\downarrow} \wedge \alpha\,i \leq \alpha\,j$. \\[-2.0em]\
+    \item for every \emph{multivalued} function~$\alpha$,\tabto{5.25cm}
+    $\exists i < j\_ \exists x \in \alpha\,i\_
+    \exists y \in \alpha\,j\_ x \leq y$.
+  \end{itemize}
+  \pause
 
-  \textbf{Def.} $\displaystyle\mathsf{Well}_\infty(X,{\leq}) \defeq
-    \prod\limits_{\alpha \? \NN \to X} \, \Bigl\|\sum\limits_{i \? \NN} \sum\limits_{j \? \NN}
-      i < j \times \alpha\,i \leq \alpha\,j\Bigr\|_{-1}$
-  \pause
-  \qquad\scalebox{1.5}{\dbend}
-  \medskip
-
-  \begin{itemize}
-    \item[\bad{\xmark}] philosophically~strenuous  %: needless reference of the transfinite
-    \item[\bad{\xmark}] not practical  %: cannot verify closure properties
-    \item[\bad{\xmark}] not faithful(?)  %: missing out on some sequences
-  \end{itemize}
+  \vspace*{-0.7em}
+  \hil{Central insight:} A quasi-order $X$ is well\ind iff $\hil{$\necessary$}\,\forall\alpha : \NN \to
+  X\_ \exists i < j\_ \alpha\,i \leq \alpha\,j$.
 \end{frame}}
 
-\begin{frame}{Sequences provided by \textsc{lem}}
-  \justifying
-  \begin{block}{}
-    \justifying
-    \textbf{Lemma.} Let~$X$ be a well quasiorder. Let~$\alpha : \NN \to X$ be a
-    sequence. Assuming~\bad{\textsc{lem}}, there merely is a monotonic
-    subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
-  \end{block}
-  \vspace*{-0.6em}
-  \emph{Proof.} The set
-  \[ I \defeq \{ i \in \NN \,|\, \neg\mathop{\exists}\limits_{j \in \NN} i < j \times \alpha\,i \leq \alpha\,j
-  \} \]
-  cannot be in bijection with~$\NN$, as else the~$I$-extracted subsequence
-  of~$\alpha$ would not be good. By~\bad{\textsc{lem}}, the set~$I$ is finite.
-  Any index~$i_0$ larger than all the indices in~$I$ is a suitable starting
-  point for a monotonic subsequence. \qed
-  \medskip
-  \pause
-
-  \begin{block}{}
-    \justifying
-    \textbf{Prop.} Let~$X$ and~$Y$ be well quasiorders.
-    Assuming~\bad{\textsc{lem}},~$X \times Y$ is well.
-  \end{block}
-  \vspace*{-0.6em}
-  \emph{Proof.} Let an infinite sequence~$\gamma : \NN \to X \times Y$ be
-  given. Write~$\gamma\,k = (\alpha\,k,\beta\,k)$. By the lemma, there is a
-  monotonic subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$. Because~$Y$
-  is well, there are indices~$n < m$ such that~$\beta\,i_n \leq \beta\,i_m$. As
-  also~$\alpha\,i_n \leq \alpha\,i_m$, the sequence~$\gamma$ is good.
-  \medskip
-  \pause
-
-  \dnote{\emph{We cannot trust~\bad{\textsc{lem}}-provided sequences to be
-  available in the type~$\NN \to \NN$. Similarly with~\bad{\textsc{dc}}.}}
-\end{frame}
-
-\begin{frame}{Sequences depending on an environment}
-  Let~$X$ be an hset such that there is no surjection~$\NN \twoheadrightarrow X$.
-  Then the type~$\NN \to X$ misses the \hil{generic enumeration}~$\genalpha$ of~$X$.
-  \medskip
-  \pause
-
-  \emph{Idea:} Approximate (fictitious) surjections~$\NN \twoheadrightarrow X$
-  by finite sequences~$\sigma = x_0 \ldots x_{n-1} \? X^*$. \\ Over time, such an
-  approximation can grow to
-  \begin{enumerate}
-    \item one of $\sigma y$, where~$y \? X$, so it becomes \emph{more defined}; or
-    \item one of~$\sigma \tau$, where~$\tau \? X^*$ such that~$a \in
-    \sigma\tau$, so it becomes \emph{more surjective}.
-  \end{enumerate}
-  \medskip
-  \pause
-
-  For~$k : \NN$ and~$a : X$, the expression~``$\genalpha\,k = a$'' does not
-  denote a proposition but rather a \hil{stage-dependent proposition}~$X^* \to
-  \mathsf{Prop}$, namely $\lambda x_0 \ldots x_{n-1}.\ k < n \times x_k = a$.
-  \medskip
-  \pause
-
-  Given a stage-dependent proposition~$P$, $\hil{$\nabla P$}\,\sigma$ expresses that no
-  matter how~$\sigma$ evolves to a better approximation~$\tau$,
-  eventually~$P\,\tau$ will hold:
-  \AgdaHide{
-  \begin{code}
-    {-# OPTIONS --no-prop --no-import-sorts --allow-unsolved-metas #-}
-    open import Agda.Primitive using (Set)
-    Prop = Set
-    postulate
-      X : Set
-      X* : Set
-      [] : X*
-      _∷_ : X* → X → X*
-      _⧺_ : X* → X* → X*
-      _∈_ : X → X* → Prop
-    module Foo where
-  \end{code}
-  }
-  {\small\begin{code}
-      data ∇ (P : X* → Prop) (σ : X*) : Prop where
-        now : P σ → ∇ P σ
-        later₁ : ((y : X) → ∇ P (σ ∷ y)) → ∇ P σ
-        later₂ : (a : X) → ((τ : X*) → a ∈ (σ ⧺ τ) → ∇ P (σ ⧺ τ)) → ∇ P σ
-  \end{code}}
-
-  \vspace*{-0.8em}
-  For instance, for~$P\,x_0 \ldots x_{n-1} \defeq \exists_{k : \NN} (k < n
-  \times x_k = a)$, we have~$\nabla P \, \varepsilon$. \quad ``$\genalpha$ is surjective''
-\end{frame}
-
-\begin{frame}{Well quasiorders revisited}
-  \emph{A naive formalization attempt:}
-
-  \textbf{Def.} $\displaystyle\mathsf{Well}_\infty(X,{\leq}) \defeq
-    \prod\limits_{\alpha \? \NN \to X} \, \Bigl\|\sum\limits_{i \? \NN} \sum\limits_{j \? \NN}
-      i < j \times \alpha\,i \leq \alpha\,j\Bigr\|_{-1}$
-  \pause
-  \qquad\scalebox{1.0}{\dbend}
-  \bigskip
-
-  \emph{An inductive rephrasing:}
-
-  \textbf{Def.} $\displaystyle\mathsf{Well}(X,{\leq}) \defeq
-    \nabla \mathsf{Good} \, \varepsilon$, where~$\mathsf{Good}\,x_0 \ldots
-    x_{n-1} \defeq \mathop{\exists}\limits_{i:\NN}\mathop{\exists}\limits_{j:\NN} (i < j \times x_i \leq
-    x_j)$ and
-  {\vspace*{-0.7em}\small\begin{code}
-    data ∇ (P : X* → Prop) (σ : X*) : Prop where
-      now : P σ → ∇ P σ
-      later : ((y : X) → ∇ P (σ ∷ y)) → ∇ P σ
-  \end{code}}
-
-  \vspace*{-0.8em}
-  In other words: \emph{A quasiorder~$X$ is well iff the generic sequence~$\NN \to X$
-  is good.}
-
-  \begin{block}{}
-    \justifying
-    \textbf{Prop.} Let~$X$ and~$Y$ be well quasiorders.
-    \good{Without} assuming~\bad{\textsc{lem}},~$X \times Y$ is well.
-  \end{block}
-
-  \dnote{We have~$\mathsf{Well}(X,{\leq}) \to
-  \mathsf{Well}_\infty(X,{\leq})$, but the converse only holds in presence of \hil{bar
-  induction}. \emph{How much stronger exactly is the inductive rephrasing?}}
-\end{frame}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses}\end{minipage}}
-\begin{frame}{Slide about toposes}
-  (something about overloading of the logical connectives)
-\end{frame}}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
-\begin{frame}{Maximal ideals as convenient fictions}
-  \jnote{1}{
-    The theorem on the slide is a generalization of a fact from undergraduate
-    linear algebra: Over a field, no surjective matrix can have more rows than
-    columns. (``Surjective'' here means that the induced linear map is
-    surjective.)
-  }
-
-  \jnote{6-7}{
-    The slide presents a standard proof as offered by most textbooks on
-    commutative algebra. The proof is quite efficient from a viewpoint of
-    mathematical organization, as it quickly succeeds in reducing to the field
-    situation. As such, it is short and memorable.
-
-    However, the proof can also be critized for appealing to the transfinite
-    two times; the methods of the proof are at odds with the concreteness of
-    the statement of the theorem---from given equations witnessing
-    surjectivity,~$Mv_i = e_i$, we are asked to deduce the equation~$1 = 0$.
-
-    For this reason, the theorem and its classical proof are often used as case
-    studies for tools and techniques aiming to extract constructive content
-    from classical proofs. One such technique employs constructive forcing.
-    The first(?)\@ constructive proof, found directly without using extraction
-    techniques, is laid out in a
-    \fixedhref{https://www.ams.org/journals/proc/1988-103-04/S0002-9939-1988-0954974-5/S0002-9939-1988-0954974-5.pdf}{beautiful short note by Richman}.
-  }
-
-  \jnote{8-9}{
-    For the countable case, there is an iterative construction of a maximal
-    ideal, making do without any decidability assumptions (due to Krivine,
-    later later clarified by Berardi and Valentini). It is a parlor trick,
-    resulting in a subset which formally verifies the axioms for a maximal
-    ideal but without carrying out any actual work. Indeed, the resulting ideal
-    will in general not be a detachable subset of the ring.
-
-    Surprisingly, there is still computational content in this construction, as
-    explored in \fixedhref{https://arxiv.org/abs/2207.03873}{this joint paper
-    with Peter Schuster}; one interpretation of our observation is that classical
-    proofs don't ``really'' require a maximal ideal; they just use that notion
-    for structuring hidden computations.
-  }
-
-  \jnote{9}{
-    In a suitable topos, the ring appears countable. Hence we can
-    carry out the iterative maximal ideal construction there. The resulting ideal
-    will not be part of the base universe (instead, from the point of view of
-    the base universe we will just have constructed a certain sheaf of ideals on a
-    certain pointfree space), but bounded first-order consequences of its
-    existence still pass down to the base.
-  }
-
-  \jnote{10}{
-    Unwinding all the definitions from toposes and from the
-    iterative maximal ideal construction, and eliminating the application
-    of~\textsc{lem} from the classical proof presented before, we mechanically
-    arrive at the constructive direct proof presented on the slide.
-  }
-
-  \begin{block}{}
-    \textbf{Thm.}
-    Let~$M$ be a surjective matrix with more rows than columns over a
-    commutative ring~$A$. Then~$1 = 0$ in~$A$.
-  \end{block}
-
-  \only<1-9>{
-    \visible<2->{\emph{Proof.} (classical) \bad{Assume not.}}
-    \visible<3->{Then there is~a \bad{maximal ideal} $\mmm$.}
-    \visible<5->{The matrix is surjective over~$A/\mmm$.}
-    \visible<6->{Since~$A/\mmm$ is a field, this is a contradiction to basic linear algebra.\qed}
-  }
-
-  \only<4-9>{\medskip\par\centering\scalebox{0.9}{\centering\begin{tikzpicture}
-    \node (0) at (0,1) {$(0) = \{0\}$};
-    \node (1) at (0,5) {$(1) = \ZZ$};
-    \node (2) at (-2,4) {$(2)$};
-    \node [right of=2] (3) {$(3)$};
-    \node [below of=2] (4) {$(4)$};
-    \node [below of=2, xshift=0.7cm] (6) {$(6)$};
-    \node [right of=3] (5) {$(5)$};
-    \node [right of=5] (7) {$(7)$};
-    \node [right of=7] (7d) {$\ldots$\phantom{(}};
-    \node [right of=7d, xshift=3cm, yshift=-2cm] (max)
-    {\vbox{\small{\it maximal among the proper ideals} \\ \medskip \hspace*{-6.75em}\textbullet \quad $\neg(1 \in
-    \mmm)$ \\ \medskip \textbullet \quad $\neg\bigl(1 \in \mmm + (x)\bigr) \Rightarrow x \in \mmm$}};
-    \node [below of=4] (8) {$(8)$};
-    \node [right of=8, xshift=3cm] (8d) {$\ldots$};
-    \draw (0) -- (8);
-    \draw (0) -- (8d);
-    \draw (0) -- (6);
-    \draw (2) -- (1);
-    \draw (3) -- (1);
-    \draw (5) -- (1);
-    \draw (7) -- (1);
-    \draw (7d) -- (1);
-    \draw (4) -- (2);
-    \draw (8) -- (4);
-    \draw (6) -- (2);
-    \draw (6) -- (3);
-    \draw [mypurple!30, thick, shorten <=-2pt, shorten >=-2pt, ->] (max) to [out=120, in=-30] (7d);
-    \begin{pgfonlayer}{background}
-      \draw[decorate, very thick, draw=mypurple!30]
-        ($(2.south west) + (8pt, 0)$) arc(270:180:8pt) --
-        ($(2.north west) + (0, -8pt)$) arc(180:90:8pt) --
-        ($(7d.north east) + (-8pt, 0)$) arc(90:0:8pt) --
-        ($(7d.south east) + (0, 8pt)$) arc(0:-90:8pt) --
-        cycle;
-    \end{pgfonlayer}
-  \end{tikzpicture}\par}\par}
-
-  \pause
-  \pause
-  \pause
-  \pause
-  \pause
-  \pause
-  \raggedright
-
-  \only<7-9>{
-    \emph{Does there exist a maximal ideal?}
-    \pause
-    \good{Yes}, if~$A$ is countable.
-    In the general case: \bad{No}\pause, but \good{yes} in a \emph{suitable topos}, and
-    \emph{bounded first-order consequences} of its existence there \good{do hold} in
-    the base.
-    \pause
-  }
-
-  \only<10>{
-    \emph{Proof.} (constructive, special case) Write~$M =
-    \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right)$. By surjectivity,
-    we have~$u, v \? A$ with
-    \[
-      u \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}1\\0\end{smallmatrix}\right)
-      \quad\text{and}\quad
-      v \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}0\\1\end{smallmatrix}\right).
-    \]
-    Hence
-    $
-      1 = (vy) (ux) = (uy) (vx) = 0
-    $. \qed
-  }
-\end{frame}}
-
-\newcommand\ytl[2]{%
-  \parbox[b]{4.5em}{\hfill{#1}~$\cdots\cdots$~}%
-  \makebox[0pt][c]{\color{mypurple}$\bullet$}{\color{mypurple}\vrule}\quad%
-  \parbox[c]{9.3cm}{\vspace{7pt}\raggedright#2\\[7pt]}%
-  \\[-2.0pt]}
-\begin{frame}{A brief timeline}
-  \jnote{4}{
-    Gödel's proof is by the \hil{$L$-translation}, where $L$ is the
-    ``constructible universe''. This translation ``relativizes quantification
-    to~$L$'', for instance the~$L$-translation of
-    \[ \varphi \defeqv \bigl(\forall x\_ \exists y\_ \ldots\bigr) \qquad\text{is}\qquad
-    \varphi^L \equiv \bigl(\forall(x \in L)\_ \exists(y \in L)\_ (\ldots)^L\bigr). \]
-    We can then verify, in a weak metatheory such as~\textsc{pra}, that for
-    every formula~$\varphi$ in the language of set theory:
-    If $\textsc{zfc}\text{+}\textsc{ch} \vdash \varphi$, then $\textsc{zf} \vdash
-    \varphi^L$.
-
-    Specializing to~$\varphi \defeqv \bot$ we obtain in particular: If~\textsc{zfc} is
-    inconsistent, then so is~\textsc{zf}. The axiom of choice does not
-    introduce new inconsistencies.
-
-    In modern semantic language: While the axiom of choice and~\textsc{ch} might fail in the
-    base universe~$V$ (= the class of all sets), they always hold in~$L$.
-    Gödel's~$L$ was the first \emph{inner model} (= class-sized model of set
-    theory) explicitly studied, nowadays we know many.
-  }
-
-  \jnote{5}{
-    For his proof, Cohen invented the technique of \emph{forcing}, situated in
-    classical mathematics where the base universe~$V$ is assumed to validate
-    the axioms of~\textsc{zfc}.
-
-    Recall that a given ring~$R$ or group can be extended in various ways, to include
-    ``generic elements'' as in~$R[X]$ or elements with prescribed relations as
-    in~$R[X]/(X^2+1) =\vcentcolon R[i]$. The idea of forcing is to construct
-    similar such extensions, but not of rings but of universes (traditionally set-sized models
-    of~\textsc{zf} or~\textsc{zfc}, but also class-sized models, or models
-    of intuitionistic set theories, or models of type theories, or even models
-    of arithmetic).
-
-    In semantic language, from a high level the idea of Cohen's independency proof is the following: Whether
-    the base universe~$V$ contains a cardinal number intermediate
-    between~$\aleph_0$ and~$2^{\aleph_0}$ is uncertain. But there is a certain
-    extension of the base universe---constructed by forcing---which does
-    contain such a number. Like the base~$V$, this forcing extension still
-    validates the axioms of~\textsc{zfc}. Hence there cannot be a
-    \textsc{zfc}-proof of~\textsc{ch}, as in Cohen's extension~$\neg\textsc{ch}$
-    holds.
-
-    Syntactically, Cohen's forcing provides us with an explicit formula
-    translation~$\varphi \mapsto \varphi^C$ such that~\textsc{pra} proves:
-    For every formula~$\varphi$, if~$\textsc{zfc}{\text{+}}\neg\textsc{ch} \vdash \varphi$,
-    then~$\textsc{zfc} \vdash \varphi^C$.
-    \bigskip
-  }
-
-  \jnote{6}{
-    Joel David Hamkins argues: In view of our rich experience with worlds which
-    validate~\textsc{ch} and worlds which don't, we shouldn't be surprised that
-    no proposed new axiom for settling~\textsc{ch} is ultimately convincing.
-
-    Instead, we should embrace the multiverse of all models of set theory and
-    explore how the truth values of statements of interest change when we
-    travel the multiverse (for instance, by passing from a universe to one
-    of its forcing extensions).
-
-    In this generalized sense, the continuum hypothesis is settled: We have a
-    good understanding of the stability properties of~\textsc{ch} under
-    important constructions. In particular, for a certain precise meaning of
-    ``universe'' and ``extension'', we know that~\textsc{ch} is a
-    \emph{switch}: $\necessary(\possible\textsc{ch} \wedge
-    \possible\neg\textsc{ch})$; in words: Every universe can be extended both to a
-    universe in which~\textsc{ch} holds and to a universe in
-    which~$\neg\textsc{ch}$ holds.
-
-    An exposition and references for further reading about the multiverse
-    position can be found
-    \fixedhref{https://www.speicherleck.de/iblech/stuff/multiverse.pdf}{here}.
-  }
-
-  \jnote{8}{
-    J. Roitman, The uses of set theory, \emph{Math.\@ Intelligencer}
-    \textbf{14}(1) (1992), 63--69.
-
-    Forcing is useful not only to explore the range of foundational
-    possibility; it has many more applications across several subjects of
-    mathematics.
-  }
-
-  \begin{columns}[t]
-    \begin{column}{0.80\textwidth}
-      \ytl{1878}{Cantor advances the \hil{continuum hypothesis}, the claim
-      that~$2^{\aleph_0} = \aleph_1$.}
-      \pause
-      \ytl{1910s}{Zermelo--Fraenkel set theory emerges.}
-      \pause
-      \ytl{1920s}{Set theorists pursue additional axioms to
-      settle~\textsc{ch} \\ (one way or another).}
-      \pause
-      \ytl{1938}{Gödel proves: If~\textsc{zfc} is consistent, so
-      is~\textsc{zfc}+\textsc{ch}.}
-      \pause
-      \ytl{1963}{Cohen proves: If~\textsc{zfc} is consistent, so
-      is~\textsc{zfc}+$\neg$\textsc{ch}.}
-      \pause
-      \ytl{2011}{Hamkins offers his paper on the \hil{multiverse position} in
-      the philosophy of set theory.}
-      \pause
-      %arguing that the program of pursuing
-      %additional axioms (while successful in many ways) is doomed to fail
-      %to settle~\textsc{ch}.}
-      \ytl{2016}{Oldenziel proposes to study the modal multiverse of parametrized
-      mathematics.}
-      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
-      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
-      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
-    \end{column}
-
-    \begin{column}{0.20\textwidth}
-      \pause
-      \centering\includegraphics[width=\textwidth,valign=t]{roitman}
-
-      \scriptsize Judith Roitman %(* 1945)
-      \medskip
-
-      \emph{Mainstream mathematics is beginning to see results using
-      modern set theoretic techniques.}
-    \end{column}
-  \end{columns}
-\end{frame}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.59cm}\includegraphics[width=\paperwidth]{staircase}\end{minipage}}
-\begin{frame}{The modal set-theoretic multiverse}
-  \begin{tikzpicture}[remember picture,overlay]
-    \node[xshift=-2.5cm,yshift=-4cm] at (current page.north east)
-    {\includegraphics[width=3cm]{branching}};
-  \end{tikzpicture}
-  \justifying
-  \vspace*{-2em}
-
-  \textbf{Def.} A \hil{model of set theory} is a (perhaps class-sized) structure $(M,{\in})$
-  satisfying axioms such as those of~\textsc{zfc}.
-
-  {\small\emph{Examples.}\vspace*{-0.8em}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[width=\paperwidth]{staircase-more-faded}\end{minipage}}
+\begin{frame}{The modal multiverse of constructive forcing}
+  \textbf{Def.} A statement~$\varphi$ holds \ldots
   \begin{itemize}
-    \item $V$, the class of all sets \vspace*{-0.6em}
-    \item $L$, Gödel's constructible universe \vspace*{-0.6em}
-    \item $V[G]$, a forcing extension containing a generic filter~$G$ of \\ some
-    poset of forcing conditions \vspace*{-0.6em}
-    \item Henkin/term models from consistency of (extensions of)~\textsc{zfc}
-  \end{itemize}}
-  \pause
-
-  \emph{We embrace all models of set theory:}
-
-  \textbf{Def.} $\possible\varphi$ iff~$\varphi$ holds in \hil{some extension} of
-  the current universe. \\
-  \phantom{\textbf{Def.}} $\necessary\varphi$ iff~$\varphi$ holds in \hil{all extensions}
-  of the current universe.
-  \begin{itemize}
-    \item $\necessary(\possible\textsc{CH} \wedge \possible\neg\text{CH})$,
-    the continuum hypothesis is a \hil{switch}.
-    \item $\necessary\possible\necessary(\text{$X$ is countable})$,
-    existence of an enumeration is a \hil{button}.
+    \small
+    \item \hil{everywhere} \tabto{2.15cm}($\necessary\varphi$)\tabto{3.08cm} iff it holds in every topos
+    (over the current base).
+    \item \hil{somewhere} \tabto{2.15cm}($\possible\varphi$)\tabto{3.08cm} iff it holds in some positive topos.
+    \item \hil{proximally} \tabto{2.15cm}($\xpossible\varphi$)\tabto{3.08cm} iff it holds in some positive overt topos.
   \end{itemize}
-\end{frame}}
-
-\begin{frame}{The modal topos-theoretic multiverse}
-  \textbf{Def.} A statement~$\varphi$ holds \ldots
-  \begin{enumerate}
-    \small
-    \item \hil{everywhere} ($\necessary\varphi$) iff it holds in every topos
-    (over the current base).
-    \item \hil{somewhere} ($\possible\varphi$) iff it holds in some positive topos.
-    \item \hil{proximally} ($\xpossible\varphi$) iff it holds in some positive overt topos.
-  \end{enumerate}
   \bigskip
@@
   \begin{tikzpicture}[overlay]
-    \node[anchor=south east,inner sep=0] (image) at (14.8,0.2) {
-      \includegraphics[width=0.2\textwidth]{multiverse}
+    \node[anchor=south east,inner sep=0] (image) at (14.8,0.6) {
+      \includegraphics[width=0.2\textwidth]{branching}
     };
   \end{tikzpicture}\vspace*{-2.3em}
-
   \pause
 
-  \small
-  \begin{multicols}{2}
-    For every inhabited set~$X$, \emph{proximally} \\
-    there is an enumeration~$\NN \twoheadrightarrow X$.
-    \medskip
+  \colorbox{white}{\bignumber{1} A quasiorder is well\ind iff \emph{everywhere}, every sequence is good.}
 
-    A quasiorder is well iff \emph{everywhere}, \\
-    every sequence is good.
-    \medskip
+  \mbox{\qquad
+  \colorbox{white}{\bignumber{2} A relation is well-founded iff \emph{everywhere}, there is
+  no infinite descending chain.}}
 
-    A ring element is nilpotent iff \\
-    all prime ideals \emph{everywhere} contain it.
-    \medskip
+  \qquad
+  \qquad
+  \colorbox{white}{\bignumber{3} A ring element is nilpotent iff all prime ideals \emph{everywhere} contain it.}
 
-    For every ring, \emph{proximally} \\ there is a maximal ideal.
-    \medskip
+  \mbox{\qquad
+  \qquad
+  \qquad
+  \colorbox{white}{\bignumber{4} For every inhabited set~$X$, \emph{proximally} there is an
+  enumeration~$\NN \twoheadrightarrow X$.}}
 
-    A relation is well-founded iff \emph{everywhere}, \\ there is no descending chain.
-    \medskip
+  \qquad
+  \qquad
+  \qquad
+  \qquad
+  \colorbox{white}{\bignumber{5} For every ring, \emph{proximally} there is a maximal ideal.}
 
-    \emph{Somewhere}, \\ the law of excluded middle holds.
-  \end{multicols}
+  \qquad
+  \qquad
+  \qquad
+  \qquad
+  \qquad
+  \colorbox{white}{\bignumber{6} \emph{Somewhere}, the law of excluded middle holds.}
+\end{frame}}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{-4.68em}\includegraphics[width=1.2\paperwidth]{wqo-faded}\end{minipage}}
+\begin{frame}{Answering a question by Berardi--Buriola--Schuster}
+  %Stefano Berardi, Gabriele Buriola and Peter Schuster established that the
+  %following notion is useful for structuring constructive arguments:
+
+  \textbf{Def.} A quasi-order~$X$ is \hil{well\impl} iff (approximately) for
+  every monotone predicate~$B$,
+  \[ \text{if $B \mid_{\text{incr}} [\,]$, then $B \mid [\,]$.} \]
+  \justifying
+  {\small``Assume that no matter how the empty list evolves to an \emph{increasing}
+  list~$\sigma$, eventually~$B\,\sigma$. Then no matter how the empty
+  list evolves to an \emph{arbitrary} list~$\tau$, eventually~$B\,\tau$.''\par}
   \pause
 
-  \begin{block}{}
-    \justifying
-    \textbf{Prop.} Let~$(X,{\leq})$ be a well quasiorder.
-    Then~$({<})$, where~$x < y \equiv (x \leq y \wedge \neg(y \leq x))$,
-    is well-founded.
-  \end{block}
-  \vspace*{-0.5em}
-
-  \emph{Proof.} Everywhere, there can be no infinite descending chain, as any
-  such would also be good. \qed
-
-  Unrolling this proof gives a program~$\nabla\mathsf{Good}\,\varepsilon \to
-  \prod_{x:X} \mathsf{Acc}\,x$.
-\end{frame}
-
-\begin{frame}{The modal type-theoretic multiverse?}
-  \emph{As I understand it, \ldots}
-
-  For extensional type theory, similar picture as with toposes:
-  Given a model of extensional type theory and a Grothendieck site~$\mathcal{C}$ in it, there
-  is a model of extensional type theory deserving the name~``sheaf model
-  over~$\mathcal{C}$''. This sheaf model is just a thin translation layer of
-  the base model.
+  \hil{Equivalently:} If everywhere every \emph{increasing} infinite sequence~$\alpha : \NN
+  \to X$ has a finite prefix validating~$B$, then so does every
+  \emph{arbitrary} infinite sequence everywhere.
   \pause
 
-  For HoTT-like theories, given a ``presheaf model'' we know how to construct a
-  ``sheaf model'', but we do not know how to construct presheaf models in the
-  first place.
+  \textbf{Prop.} $\text{well\impl} \Rightarrow \text{well\ind}$.
+
+  \emph{Proof.} Trivially~$\mathsf{Good} \mid_{\text{incr}} [\,]$,
+  hence~$\mathsf{Good} \mid [\,]$ by assumption. \qed
   \pause
 
-  Still, can pragmatically use~$\nabla$ and feel philosophically inspired.
-\end{frame}
+  \textbf{Thm.} $\text{well\impl} \Leftarrow \text{well\ind}$.
+
+  \small
+  \emph{Proof.} Let~$x_0,x_1,\ldots$ be an infinite sequence (in an arbitrary topos).
+  \emph{Somewhere,} \textsc{lem} holds and implies that there is an increasing infinite subsequence~$x_{i_0},
+  x_{i_1}, \ldots$. \emph{There} we have, by assumption, a finite prefix
+  with~$B\,x_{i_0}\ldots x_{i_n}$. As~$B$ is monotone, \emph{there}
+  we also have~$B\,x_0x_1\ldots x_{i_n}$. So \emph{somewhere} there is a finite
+  prefix validating~$B$. Hence there actually is a finite prefix
+  validating~$B$. \qed
+\end{frame}}
+
+\backupstart
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{-2.2em}\includegraphics[width=\paperwidth]{agda-wqo}\end{minipage}}
+\begin{frame}
+  \vspace*{\paperheight}\vspace*{-2.5em}\centering
+  \emph{Agda formalization in progress.}
+\end{frame}}
+\backupend
 

### 2024-11-30 slides-abmv2024.tex
Final slides

@@
 \hypersetup{colorlinks=true}
-
-\newcommand*\circled[1]{\tikz[baseline=(char.base)]{%
-  \node[shape=circle,draw,inner sep=1pt] (char) {#1};}}
 
@@
 \newcommand{\astikznode}[2]{\tikz[baseline,remember picture]{\node[anchor=base,inner sep=0,outer sep=0.1em] (#1) {#2};}}
-\newcommand{\astikznodecircled}[3]{\tikz[baseline,remember picture]{\node[anchor=base,circle,draw=#2,thick,inner sep=0,outer sep=0.05em] (#1) {#3};}}
-\newcommand{\astikznodetransparentlycircled}[2]{\tikz[baseline,remember picture]{\node[anchor=base,circle,opacity=0,draw=white,text opacity=1,thick,inner sep=0,outer sep=0.05em] (#1) {#2};}}
+\newcommand{\astikznodecircled}[3]{\tikz[baseline,remember picture]{\node[anchor=base,circle,draw=#2,thick,inner sep=0.05em,outer sep=0.05em] (#1) {#3};}}
+\newcommand{\astikznodetransparentlycircled}[2]{\tikz[baseline,remember picture]{\node[anchor=base,circle,opacity=0,draw=white,text opacity=1,thick,inner sep=0.05em,outer sep=0.05em] (#1) {#2};}}
 
@@
       \end{block}
+      \only<2-4>{
+        \vspace*{-2em}
+        \[ \astikznodetransparentlycircled{xm}{7}\!,
+        \quad \astikznodetransparentlycircled{x0}{4}\!,
+        \quad \astikznodecircled{t1}{mypurple}{3}\!,
+        \quad \astikznodetransparentlycircled{x1}{1}\!,
+        \quad \astikznodecircled{t2}{mypurple}{8}\!,
+        \quad \astikznodetransparentlycircled{x2}{2}\!,
+        \quad \ldots \]
+      {\centering\begin{tikzpicture}[remember picture,overlay]
+        \node[draw=mypurple, circle, thick, inner sep=0.1em] (t3) {\scriptsize$\leq$};
+        \path[draw=mypurple,thick]
+          (t1)
+          to [out=-90, in=180] (t3)
+          to [out=0, in=-90] (t2);
+      \end{tikzpicture}\par}}
     \end{column}
+
     \pause

### 2024-12-01 slides-abmv2024.tex
Notes for the ABMV talk

@@
 
-%\setbeameroption{show notes on second screen=bottom}
+\setbeameroption{show notes on second screen=bottom}
 \newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\footnotesize\justifying#2\par}}}
@@
 \begin{frame}{Well quasi-orders}
+  \jnote{2-3}{
+    The presented proof rests on the law of excluded middle and hence cannot
+    immediately be interpreted as a program for finding suitable indices~$i <
+    j$. However, constructive proofs are also possible (for instance by
+    induction on the value of a given term of the sequence, see
+    \href{https://www.math.lmu.de/~petrakis/Dickson.pdf}{Constructive
+    combinatorics of Dickson's Lemma} by Iosif Petrakis for several fine
+    quantitative results). And even more:
+    There is a procedure for regarding this proof---and many others in the
+    theory of well quasi-orders---as \emph{blueprints} for more informative
+    constructive proofs. This shall be our motto for today:
+
+    \emph{Do not take classical proofs literally, instead ask which
+    constructive proofs they are blueprints for.}
+  }
+
+  \jnote{3}{
+    The displayed stability results, along with several others, provide a
+    flexible toolbox for constructing new well quasi-orders from given ones.
+    However, with the classical formulation of \emph{well},
+    renamed~``well\sinf'' on the next slide, these results are inherently
+    classical.
+
+    In Higman's lemma, the set~$X^*$ of finite lists of elements of~$X$ is
+    equipped with the following ordering: We have~$x_0 \ldots x_{n-1} \leq y_0
+    \ldots y_{m-1}$ iff there is an increasing injection~$f : \{ 0,\ldots,n-1 \} \to \{
+    0,\ldots,m-1 \}$ such that~$x_i \leq y_{f(i)}$ for all~$i < n$.
+  }
+
+  \jnote{4}{
+    The dependence of the theory on well quasi-orders on classical transfinite
+    methods is already present in one of basic observations of this theory:
+
+    \textbf{Lemma.} Let~$X$ be well\sinf. Let~$\alpha : \NN \to X$. Then there
+    is an increasing subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq
+    \ldots$.
+
+    \emph{Proof.} Let~$K \defeq \{ n \in \NN \,|\, \neg \exists m > n\_
+    \alpha\,n \leq \alpha\,m \}$ the set of indices of those terms which cannot
+    appear as the first component of a good pair. If~$K$ is in bijection
+    with~$\NN$, there is a subsequence~$\alpha\,k_0 \leq \alpha\,k_1 \leq
+    \ldots$ with~$k_0, k_1, \ldots \in K$. As~$X$ is well\sinf, this sequence
+    is good, a contradiction.
+
+    Hence~$K$ is not in bijection with~$\NN$. Assuming~\bad{\textsc{lem}},
+    it is hence bounded by a number~$N$, and (again
+    with~\bad{\textsc{lem}}), for every index~$a > N$ there is an index~$b >
+    a$ such that~$\alpha\,a \leq \alpha\,b$. Thus,
+    assuming~\bad{\textsc{dc}}, every number~$i_0 > N$ is a suitable
+    starting index for an infinite increasing subsequence. \qed
+
+    The appeal to dependent choice can be removed by always picking the
+    smallest possible next index in~$\NN \setminus K$, doable by yet another
+    invocation of~\bad{\textsc{lem}}, but the result remains fundamentally
+    noneffective---in the special case~$X = (\{ 0, 1 \}, {=})$, the statement
+    of the lemma implies the infinite pigeonhole principle.
+  }
+
+  \jnote{5-7}{
+    Luckily, thanks to work by Thierry Coquand, Daniel Fridlender and Monika
+    Seisenberger, a constructive substitute is available, the notion well\ind.
+    In classical mathematics (where~\textsc{lem} and~\textsc{dc} and hence bar
+    induction are available), this notion is equivalent to well\sinf.
+
+    The assertion~``$\mathsf{Good} \mid [\,]$'' is pronounced ``$\mathsf{Good}$
+    bars the empty list'', and is defined as follows: Let~$B$ be a predicate
+    on~$X^\star$. Then~$B \mid
+    \sigma$ is inductively generated by the following two clauses.
+
+    \vspace*{-0.5em}
+    \begin{enumerate}
+      \item If~$B\sigma$, then~$B \mid \sigma$.
+      \\[-1.3em]
+      \item If~$B \mid \sigma x$ for all~$x \in X$, then~$B \mid \sigma$.
+    \end{enumerate}
+    \vspace*{-0.5em}
+
+    Here~$\sigma x$ denotes the concatenation of the list~$\sigma$ with the
+    element~$x$. The accompanying induction principle is the following: Let~$Q$
+    be a predicate on~$X^*$ such that, for all~$\sigma \in X^*$,~$B\sigma \Rightarrow Q\sigma$ and
+    $(\forall x \in X\_ Q(\sigma x)) \Rightarrow Q\sigma$. Then, for
+    all~$\sigma \in X^*$: If~$B \mid \sigma$, then~$Q\sigma$.
+
+    Intuitively, the assertion~``$B \mid \sigma$'' expresses (in a positive
+    direct way) that no matter how~$\sigma$ evolves to a longer finite
+    list~$\tau$, eventually~$B\tau$ will hold.
+  }
+
+  \jnote{8}{
+    \begin{columns}[t]
+      \begin{column}{0.60\textwidth}
+        The original notion well\sinf:
+        \begin{itemize}
+          \justifying
+          \item[\cmark] short and simple
+          \\[-1.3em]
+          \item[\cmark] constructively satisfied for the main examples (but
+          only because of the theory around well\ind)
+          \\[-1.3em]
+          \item[\cmark] concise abstract proofs (albeit employing transfinite methods)
+          \\[-1.3em]
+          \item[\xmark] main results not constructively attainable
+          \\[-1.3em]
+          \item[\xmark] philosophically strenuous by the quantification over all sequences
+          \\[-1.3em]
+          \item[\xmark] not stable under ``change of base''---a forcing extension
+          of the universe may well contain more sequences than the base universe
+          \\[-1.3em]
+          \item[\xmark] negative (universal) condition
+        \end{itemize}
+      \end{column}
+
+      \begin{column}{0.45\textwidth}
+        The constructive substitute well\ind:
+        \begin{itemize}
+          \justifying
+          \item[\cmark] main results constructive
+          \\[-1.3em]
+          \item[\cmark] stable under change of base
+          \\[-1.3em]
+          \item[\cmark] positive (existential) condition
+          \\[-1.3em]
+          \item[\xmark] proofs intriguing, but also somewhat alien, not just some
+          trivial reshuffling of the classical arguments, classical sequence
+          language cannot be used
+        \end{itemize}
+      \end{column}
+    \end{columns}
+  }
+
+  \jnote{9}{
+    Constructively, the notion~well\ind is much stronger than~well\sinf, as it
+    ensures goodness (in an appropriate sense) of sequence-like entities
+    which are not actually honest maps~$\NN \to X$.
+
+    For partial maps~$\alpha$, by~$\alpha\,n\,\downarrow$ we mean that~$\alpha$
+    is defined on the input~$n$. If~\bad{\textsc{lem}} is available, then a
+    partial map such that~$\neg\neg(\alpha\,n\,\downarrow)$ for all~$n \in \NN$
+    is already a total map, but without~\bad{\textsc{lem}} the hypothesis
+    well\sinf does not have anything to say about such a partially-defined
+    sequence.
+
+    If~\bad{\textsc{dc}} is available, then every multivalued map contains a
+    singlevalued map, but again without~\bad{\textsc{dc}} the hypothesis well\sinf
+    does not have anything to say about multivalued sequences.
+  }
+
+  \jnote{10}{
+    It turns out that these entities are, or give rise to, actual maps~$\NN \to
+    X$---but in a forcing extension of the universe.
+
+    Forcing originated in set theory to construct new models for set theory
+    from given ones, in order to explore the range of set-theoretic
+    possibility. For instance, by forcing we can construct models of~\text{zfc}
+    validating the continuum hypothesis and also models which falsify it.
+
+    We here refer to a simplification of original forcing which is useful in a
+    constructive metatheory. At its core, every forcing extension is just a
+    formula and proof translation of a certain form. For instance, there is a
+    forcing extension validating~\bad{\textsc{lem}} even if the base universe
+    does not; this forcing extension is not a deep mystery, for a statement holds
+    in that forcing extension iff its double negation translation holds in the
+    base universe and it is well-known that the double negation translation
+    of~\bad{\textsc{lem}} is an intuitionistic tautology.
+
+    \href{https://www.speicherleck.de/iblech/stuff/slides-herrsching2023.pdf}{Here
+    is a set of slides on constructive forcing}, and Section~4 of
+    \href{https://raw.githubusercontent.com/iblech/constructive-maximal-ideals/master/tex/extended.pdf}{this
+    joint paper with Peter Schuster} contains a written summary of constructive
+    forcing.
+  }
+
   \textbf{Def.} Let~$(X,{\leq})$ be a quasi-order.
@@
 \begin{frame}{The modal multiverse of constructive forcing}
+  \jnote{1}{
+    By \emph{topos}, we mean \emph{Grothendieck topos}. In constructive
+    forcing, a ``forcing extension of the base universe'' is exactly the same
+    thing as a Grothendieck topos.
+
+    A particular member of the rich and varied landscape of toposes is the
+    \emph{trivial topos}, in which every statement whatsoever holds. By
+    restricting to positive toposes, we exclude this special case.
+
+    For positive toposes~$\E$, a geometric implication holds in~$\E$ iff it
+    holds in the base universe. For positive overt toposes~$\E$, we even have
+    that a bounded first-order formula holds in~$\E$ iff it holds in the base.
+    Hence, for the purpose of verifying a bounded first-order assertion about
+    the base, we can freely pass to a positiv overt topos with problem-adapted
+    better higher-order properties (such as that some uncountable set from the
+    base now appears countable, or that an infinite sequence whose existence is
+    predicted by failing dependent choice now actually exists).
+
+    \href{https://www.speicherleck.de/iblech/stuff/early-draft-modal-multiverse.pdf}{Here
+    is a rough early draft of a preprint with more details about the modal
+    multiverse.}
+  }
+
+  \jnote{2}{
+    The idea to study the modal multiverse of toposes in a principled manner
+    was proposed by Alexander Oldenziel in 2016. \emph{Foreshadowed by:}
+    \begin{itemize}
+      \item[1984\phantom{s}] André Joyal, Miles Tierney. ``An extension of the Galois theory of Grothendieck''.
+      \\[-1em]
+      \item[1987\phantom{s}] Andreas Blass. ``Well-ordering and induction in intuitionistic logic and topoi''.
+      \\[-1em]
+      \item[2010s] Milly Maietti, Steve Vickers. Ongoing work on arithmetic universes.
+      \\[-1em]
+      \item[2011\phantom{s}] Joel David Hamkins. ``The set-theoretic multiverse''.
+      \\[-1em]
+      \item[2013\phantom{s}] Shawn Henry. ``Classifying topoi and preservation of higher order logic by geometric morphisms''.
+    \end{itemize}
+  }
+
+  \jnote{3}{
+    With the modal language we seek to provide an accessible and modular
+    framework for constructivization results.
+
+    For instance, conservativity of classical logic over intuitionistic logic
+    for geometric implications (known under various names such as Barr's
+    theorem, Friedman's trick, escaping the continuation monad, \ldots) is
+    packaged up by the observation that \emph{somewhere}, the law of excluded
+    middle holds.
+
+    Another example:
+    In the community around Krull's lemma, it is well-known that
+    we can constructively infer that a given ring element~$x \in A$ is nilpotent from
+    knowing that it is contained in the \emph{generic prime ideal} of~$A$. This entity
+    is not actually an honest prime ideal of the ring~$A$ in the base
+    universe, but a certain combinatorial notion (efficiently dealt with using
+    \emph{entailment relations}). Constructive forcing allows us to reify the
+    generic prime ideal as an actual prime ideal in a suitable forcing extension.
+  }
+
+  \jnote{4}{
+    Here is an example how the modal language helps us to give a constructive
+    proof (for well\ind) of Dickson's lemma. In a similar vein Higman's lemma
+    and Kruskal's theorem can be proven; with the exception of a preparatory
+    first paragraph the original proofs can be copied word for word.
+
+    \textbf{Prop.} Let~$X$ and~$Y$ be well\ind quasi-orders. Then~$X \times Y$
+    is well.
+
+    \emph{Lemma.} Let~$\alpha = (\beta,\gamma) : \NN \to X \times Y$ be a sequence in an
+    arbitrary topos. We need to show that~$\alpha$ is good. It suffices to
+    prove that \emph{somewhere}, $\alpha$ is good, as goodness is a geometric
+    implication (in fact even a geometric formula). Without loss of generality,
+    we may suppose~\textsc{lem} as it holds \emph{somewhere}.
+
+    Hence there is an infinite increasing subsequence
+    \[ \beta\,k_0 \leq \beta\,k_1 \leq \ldots. \]
+    As~$Y$ is well\ind, the sequence~$\gamma\,k_0, \gamma\,k_1, \ldots$
+    is good, so there exist~$i < j$ with~$\gamma\,k_i \leq \gamma\,k_j$. Since we
+    also have~$\beta\,k_i \leq \beta\,k_j$, we are done. \qed
+  }
+
   \textbf{Def.} A statement~$\varphi$ holds \ldots
@@
   \colorbox{white}{\bignumber{6} \emph{Somewhere}, the law of excluded middle holds.}
+
+  \pause
 \end{frame}}
@@
 \begin{frame}{Answering a question by Berardi--Buriola--Schuster}
-  %Stefano Berardi, Gabriele Buriola and Peter Schuster established that the
-  %following notion is useful for structuring constructive arguments:
+  \jnote{1-3}{
+    Stefano Berardi, Gabriele Buriola and Peter Schuster recognized in their
+    recent article
+    \href{https://drops.dagstuhl.de/entities/document/10.4230/LIPIcs.CSL.2024.16}{A
+    general constructive form of Higman's lemma} the displayed implication as an
+    efficient organizing principle for structuring a constructive account of
+    well quasi-orders, which they elevated to the status of a definition.
+
+    They recognized that their definition implies well\ind (and hence
+    well\sinf), and posed the converse as an open question.
+
+    The generating clauses for~$B \mid_\text{incr} \sigma$, for increasing
+    finite lists~$\sigma$, are the following:
+
+    \vspace*{-0.5em}
+    \begin{enumerate}
+      \item If~$B\sigma$, then~$B \mid_\text{incr} \sigma$.
+      \\[-1.3em]
+      \item If~$B \mid_\text{incr} \sigma x$ for all~$x \in X$ such that~$\sigma x$ is
+      increasing, then~$B \mid_\text{incr} \sigma$.
+    \end{enumerate}
+  }
+
+  \jnote{4}{
+    The modal language unlocked the displayed positive answer to their
+    question. As the modal framework is fully constructive, the proof can in
+    principle be unrolled to yield a direct (but likely longer) argument not
+    involving constructive forcing, giving an explicit algorithm of type
+    \[ \mathsf{Good} \mid [\,] \ \times\ B \mid_\text{incr} [\,]
+    \ \longrightarrow\ B \mid [\,] \]
+    for transforming given witnessing trees. This unwinding is work in
+    progress.
+  }
 

### 2025-02-25 slides-paris2025.tex
Slides for Logique à Paris 2025, almost finished

--- 2024-12-01 slides-abmv2024.tex
@@
 
-\setbeameroption{show notes on second screen=bottom}
+%\setbeameroption{show notes on second screen=bottom}
 \newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\footnotesize\justifying#2\par}}}
@@
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\includegraphics[width=\paperwidth]{tower}\end{minipage}}
+\definecolor{mypurpleblack}{RGB}{30,0,50}
+\setbeamercolor{structure}{fg=black}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\vspace*{-7em}\includegraphics[width=\paperwidth]{forest2}\end{minipage}}
 \begin{frame}[c]
@@
   \setbeamercolor{block body}{bg=black!100}
-  \begin{minipage}{0.62\textwidth}
-    \begin{block}{\centering Arbeitstagung Bern--München--Verona}
+  \begin{minipage}{0.52\textwidth}
+    \begin{block}{}
       \centering\normalsize\color{white}
-      \hil{Modal operators for \\ a constructive account of
-      well quasi-orders} \\[-0.9em]
+      \hil{\color{white}Computing an integer using the \\ modal toposophic multiverse} \\[-0.4em]
       \
@@
   \bf
-  \colorbox{black}{November 30th, 2024}%
+  \colorbox{black}{\begin{minipage}{0.2\textwidth}
+    \centering
+    Logique à Paris \\
+    February 26th, 2025
+  \end{minipage}}
   \bigskip
@@
 
-  \textbf{Def.} A quasi-order~$X$ is \hil{well\ind} iff~$\mathsf{Good} \mid
-  [\,]$, where~$\mathsf{Good}\,x_0 \ldots x_{n-1}$ iff~$\exists i < j\_ x_i \leq x_j$.
+  \textbf{Def.} A quasi-order~$X$ is \hil{well\ind} iff there exists a
+  \hil{modulus of wellness} for~$X$.
   \pause
@@
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[width=\paperwidth]{staircase-more-faded}\end{minipage}}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{0.29cm}\hspace*{-1cm}\includegraphics[height=\paperheight]{tower-cropped}\end{minipage}}
 \begin{frame}{The modal multiverse of constructive forcing}
@@
 
+  \begin{changemargin}{7.5em}{-0.5em}
     \textbf{Def.} A statement~$\varphi$ holds \ldots
+    \vspace*{-0.3em}
     \begin{itemize}
       \small
-    \item \hil{everywhere} \tabto{2.15cm}($\necessary\varphi$)\tabto{3.08cm} iff it holds in every topos
+      \item \hil{everywhere} \tabto{2.15cm}($\necessary\varphi$)\tabto{3.08cm}
+      iff it holds \hil{in every topos}
       (over the current base).
-    \item \hil{somewhere} \tabto{2.15cm}($\possible\varphi$)\tabto{3.08cm} iff it holds in some positive topos.
-    \item \hil{proximally} \tabto{2.15cm}($\xpossible\varphi$)\tabto{3.08cm} iff it holds in some positive overt topos.
+      \item \hil{somewhere} \tabto{2.15cm}($\possible\varphi$)\tabto{3.08cm} iff
+      it holds \hil{in some positive topos}.
+      \item \hil{proximally} \tabto{2.15cm}($\xpossible\varphi$)\tabto{3.08cm}
+      iff it holds \hil{in some positive overt topos}.
     \end{itemize}
-  \bigskip
 
-  \begin{tikzpicture}[overlay]
-    \node[anchor=south east,inner sep=0] (image) at (14.8,0.6) {
-      \includegraphics[width=0.2\textwidth]{branching}
-    };
-  \end{tikzpicture}\vspace*{-2.3em}
-  \pause
+    \textbf{Def.} A (Grothendieck) \hil{topos} is a category equivalent to the
+    category of sheaves over a small site.
 
-  \colorbox{white}{\bignumber{1} A quasiorder is well\ind iff \emph{everywhere}, every sequence is good.}
+    \textbf{Examples for toposes.}
+    \begin{itemize}
+      \item $\mathrm{Set}$, the category of sets and maps.
+      \item The category of sets and maps which are \hil{defined up to~$\boldsymbol\neg\neg$}.
+      \item $\mathrm{Set}[G]$, the extension obtained by adding a \hil{generic
+      filter} of a \hil{forcing notion} (a quasi-order equipped with a
+      coverage).
+    \end{itemize}
+    The following are \bad{not} toposes:
+    \begin{itemize}
+      \item The category of sets and partially defined maps.
+      \item The category of abelian groups.
+    \end{itemize}
 
-  \mbox{\qquad
-  \colorbox{white}{\bignumber{2} A relation is well-founded iff \emph{everywhere}, there is
-  no infinite descending chain.}}
+    \begin{enumerate}
+      \item A quasiorder is well\ind iff \emph{everywhere}, every sequence is good.
+      \medskip
+      \item A ring element is nilpotent iff all prime ideals \emph{everywhere} contain it.
+      \medskip
+      \item For every inhabited set~$X$, \emph{proximally} there exists an
+      enumeration~$\NN \twoheadrightarrow X$.
+      \medskip
 
-  \qquad
-  \qquad
-  \colorbox{white}{\bignumber{3} A ring element is nilpotent iff all prime ideals \emph{everywhere} contain it.}
+      \item For every ring, \emph{proximally} there exists a maximal ideal.
+      \medskip
 
-  \mbox{\qquad
-  \qquad
-  \qquad
-  \colorbox{white}{\bignumber{4} For every inhabited set~$X$, \emph{proximally} there is an
-  enumeration~$\NN \twoheadrightarrow X$.}}
-
-  \qquad
-  \qquad
-  \qquad
-  \qquad
-  \colorbox{white}{\bignumber{5} For every ring, \emph{proximally} there is a maximal ideal.}
-
-  \qquad
-  \qquad
-  \qquad
-  \qquad
-  \qquad
-  \colorbox{white}{\bignumber{6} \emph{Somewhere}, the law of excluded middle holds.}
-
-  \pause
+      \item \emph{Somewhere}, the law of excluded middle holds.
+    \end{enumerate}
+  \end{changemargin}
 \end{frame}}
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{-4.68em}\includegraphics[width=1.2\paperwidth]{wqo-faded}\end{minipage}}
-\begin{frame}{Answering a question by Berardi--Buriola--Schuster}
-  \jnote{1-3}{
-    Stefano Berardi, Gabriele Buriola and Peter Schuster recognized in their
-    recent article
-    \href{https://drops.dagstuhl.de/entities/document/10.4230/LIPIcs.CSL.2024.16}{A
-    general constructive form of Higman's lemma} the displayed implication as an
-    efficient organizing principle for structuring a constructive account of
-    well quasi-orders, which they elevated to the status of a definition.
+\end{document}
 
-    They recognized that their definition implies well\ind (and hence
-    well\sinf), and posed the converse as an open question.
+- Well quasi-orders
+  *abstract proofs as blueprints for concrete computations?*
+  - def. good, well
+  - ex. ℕ
+  - classical stability results
 
-    The generating clauses for~$B \mid_\text{incr} \sigma$, for increasing
-    finite lists~$\sigma$, are the following:
+- A better notion
+  - partially defined sequences
+    (not really a morphism in Set, but in Sh_¬¬(Set))
+  - multivalued sequences
+    (approximately morphisms in some category)
+  - modulus of wellness
 
-    \vspace*{-0.5em}
-    \begin{enumerate}
-      \item If~$B\sigma$, then~$B \mid_\text{incr} \sigma$.
-      \\[-1.3em]
-      \item If~$B \mid_\text{incr} \sigma x$ for all~$x \in X$ such that~$\sigma x$ is
-      increasing, then~$B \mid_\text{incr} \sigma$.
-    \end{enumerate}
-  }
+- The modal toposophic multiverse
+  - examples for toposes:
+    - Set
+    - Sh_¬¬(Set)
+    - Set[G], obtained by adding a generic filter of a forcing notion
+    - not quite toposes: arbitrary partially defined maps, abelian groups
+  - modal operators
+  - Diamond φ ⇒ φ etc.
+  - X has a modulus of wellness iff everywhere, every sequence is good
+  - Diamond (¬LEM)
+  - etc.
 
-  \jnote{4}{
-    The modal language unlocked the displayed positive answer to their
-    question. As the modal framework is fully constructive, the proof can in
-    principle be unrolled to yield a direct (but likely longer) argument not
-    involving constructive forcing, giving an explicit algorithm of type
-    \[ \mathsf{Good} \mid [\,] \ \times\ B \mid_\text{incr} [\,]
-    \ \longrightarrow\ B \mid [\,] \]
-    for transforming given witnessing trees. This unwinding is work in
-    progress.
-  }
+- Applications
+  - Dickson, constructively
+  - surjective matrices (with Agda)
 
-  \textbf{Def.} A quasi-order~$X$ is \hil{well\impl} iff (approximately) for
-  every monotone predicate~$B$,
-  \[ \text{if $B \mid_{\text{incr}} [\,]$, then $B \mid [\,]$.} \]
-  \justifying
-  {\small``Assume that no matter how the empty list evolves to an \emph{increasing}
-  list~$\sigma$, eventually~$B\,\sigma$. Then no matter how the empty
-  list evolves to an \emph{arbitrary} list~$\tau$, eventually~$B\,\tau$.''\par}
-  \pause
-
-  \hil{Equivalently:} If everywhere every \emph{increasing} infinite sequence~$\alpha : \NN
-  \to X$ has a finite prefix validating~$B$, then so does every
-  \emph{arbitrary} infinite sequence everywhere.
-  \pause
-
-  \textbf{Prop.} $\text{well\impl} \Rightarrow \text{well\ind}$.
-
-  \emph{Proof.} Trivially~$\mathsf{Good} \mid_{\text{incr}} [\,]$,
-  hence~$\mathsf{Good} \mid [\,]$ by assumption. \qed
-  \pause
-
-  \textbf{Thm.} $\text{well\impl} \Leftarrow \text{well\ind}$.
-
-  \small
-  \emph{Proof.} Let~$x_0,x_1,\ldots$ be an infinite sequence (in an arbitrary topos).
-  \emph{Somewhere,} \textsc{lem} holds and implies that there is an increasing infinite subsequence~$x_{i_0},
-  x_{i_1}, \ldots$. \emph{There} we have, by assumption, a finite prefix
-  with~$B\,x_{i_0}\ldots x_{i_n}$. As~$B$ is monotone, \emph{there}
-  we also have~$B\,x_0x_1\ldots x_{i_n}$. So \emph{somewhere} there is a finite
-  prefix validating~$B$. Hence there actually is a finite prefix
-  validating~$B$. \qed
-\end{frame}}
-
-\backupstart
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{-2.2em}\includegraphics[width=\paperwidth]{agda-wqo}\end{minipage}}
-\begin{frame}
-  \vspace*{\paperheight}\vspace*{-2.5em}\centering
-  \emph{Agda formalization in progress.}
-\end{frame}}
-\backupend
-
-\end{document}
+Somewhere: eigenvalues etc.

### 2025-02-25 slides-paris2025.tex
Slides finished

@@
     category of sheaves over a small site.
+    \pause
 
+    \only<2>{
       \textbf{Examples for toposes.}
@@
       \end{itemize}
+    }
 
+    \only<3->{
+      \hil{Multiversal yoga:}
       \begin{enumerate}
         \item A quasiorder is well\ind iff \emph{everywhere}, every sequence is good.
-      \medskip
+        \smallskip
         \item A ring element is nilpotent iff all prime ideals \emph{everywhere} contain it.
-      \medskip
+        \smallskip
         \item For every inhabited set~$X$, \emph{proximally} there exists an
         enumeration~$\NN \twoheadrightarrow X$.
-      \medskip
+        \smallskip
 
         \item For every ring, \emph{proximally} there exists a maximal ideal.
-      \medskip
+        \smallskip
 
@@
       \end{enumerate}
+    }
   \end{changemargin}
+\end{frame}}
+
+\begin{frame}{Toposophic constructive combinatorics}
+  \justifying
+  \textbf{Prop.} Let~$X$ and~$Y$ be well\ind quasi-orders. Then~$X \times Y$
+  is well$\ind$.
+
+  \emph{Proof.} Let~$\alpha = (\beta,\gamma) : \NN \to X \times Y$ be a sequence in an
+  arbitrary topos. We need to show that~$\alpha$ is good, so find indices~$n < m$ such that
+  \[ \beta\,n \leq \beta\,m \quad\text{and}\quad \gamma\,n \leq \gamma\,m. \]
+
+  It suffices to prove that \emph{somewhere}, $\alpha$ is good, as goodness is
+  a geometric implication (in fact even a geometric formula). Hence without loss of
+  generality, we may suppose~\textsc{lem}.
+
+  Thus there is an infinite increasing subsequence
+  \[ \beta\,k_0 \leq \beta\,k_1 \leq \ldots. \]
+  As~$Y$ is well\ind, the sequence~$(\gamma\,k_0, \gamma\,k_1, \ldots)$
+  is good, so there exist~$i < j$ with~$\gamma\,k_i \leq \gamma\,k_j$. Since we
+  also have~$\beta\,k_i \leq \beta\,k_j$, we are done. \qed
+\end{frame}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
+\begin{frame}{Toposophic constructive algebra}
+  \justifying
+  \textbf{Thm.}
+  Let~$M$ be a surjective matrix with more rows than columns over a
+  ring~$A$. Then~$1 = 0$ in~$A$.
+
+  \emph{Classical proof.}
+  \bad{Assume not.} Then there is~a \bad{maximal ideal} $\mmm$.
+  The matrix~$M$ is surjective over~$A/\mmm$. Since~$A/\mmm$ is a field, this
+  is a contradiction to basic linear algebra. \qed
+  \pause
+
+  \emph{Multiversal constructive proof.}
+  We may work \emph{somewhere} where \good{\textsc{lem}} holds. So \good{assume not}.
+  \emph{Proximally,} there is a \good{maximal ideal}~$\mmm$.
+  The matrix~$M$ is still surjective \emph{there}, and also over~$A/\mmm$. Since~$A/\mmm$ is a field, this
+  is a contradiction to basic linear algebra. \qed
+  \pause
+
+  \emph{Unrolled constructive proof (special case).}
+  Write~$M =
+  \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right)$. By surjectivity,
+  have~$u, v$ with
+  \vspace*{-0.6em}
+  \[
+    u \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}1\\0\end{smallmatrix}\right)
+    \quad\text{and}\quad
+    v \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}0\\1\end{smallmatrix}\right).
+  \]
+  Hence
+  $
+    1 = (vy) (ux) = (uy) (vx) = 0
+  $.
+  \qed
+  \pause
+
+  \centering
+  \emph{\href{https://iblech.github.io/constructive-maximal-ideals/}{Agda
+  formalization available.}}
+  \par
 \end{frame}}

### 2025-02-26 slides-abmv2024.tex
Slides finished
typo

@@
     \emph{Proof.} Let~$K \defeq \{ n \in \NN \,|\, \neg \exists m > n\_
-    \alpha\,n \leq \alpha\,m \}$ the set of indices of those terms which cannot
+    \alpha\,n \leq \alpha\,m \}$ be the set of indices of those terms which cannot
     appear as the first component of a good pair. If~$K$ is in bijection
@@
     is already a total map, but without~\bad{\textsc{lem}} the hypothesis
-    well\sinf does not have anything to say about such a partially-defined
+    well\sinf{} does not have anything to say about such a partially-defined
     sequence.
@@
     If~\bad{\textsc{dc}} is available, then every multivalued map contains a
-    singlevalued map, but again without~\bad{\textsc{dc}} the hypothesis well\sinf
+    singlevalued map, but again without~\bad{\textsc{dc}} the hypothesis well\sinf{}
     does not have anything to say about multivalued sequences.
@@
     \textbf{Prop.} Let~$X$ and~$Y$ be well\ind quasi-orders. Then~$X \times Y$
-    is well.
+    is well\ind.
 
-    \emph{Lemma.} Let~$\alpha = (\beta,\gamma) : \NN \to X \times Y$ be a sequence in an
+    \emph{Proof.} Let~$\alpha = (\beta,\gamma) : \NN \to X \times Y$ be a sequence in an
     arbitrary topos. We need to show that~$\alpha$ is good. It suffices to

### 2025-02-26 slides-paris2025.tex
Backup slides

@@
     \textbf{Lemma.} Let~$X$ be well\sinf. Let~$\alpha : \NN \to X$. Then there
-    is an increasing subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq
+    is an infinite increasing subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq
     \ldots$.
@@
         \only<1-5>{Assuming~\badbox{\textsc{lem}}
-        and~\badbox{\textsc{dc}}}\only<6->{Constructively}, \ldots
+        and~\badbox{\textsc{dc}}}\only<6->{\good{Constructively}}, \ldots
 
@@
 
-  \jnote{4}{
-    Here is an example how the modal language helps us to give a constructive
-    proof (for well\ind) of Dickson's lemma. In a similar vein Higman's lemma
-    and Kruskal's theorem can be proven; with the exception of a preparatory
-    first paragraph the original proofs can be copied word for word.
-
-    \textbf{Prop.} Let~$X$ and~$Y$ be well\ind quasi-orders. Then~$X \times Y$
-    is well.
-
-    \emph{Lemma.} Let~$\alpha = (\beta,\gamma) : \NN \to X \times Y$ be a sequence in an
-    arbitrary topos. We need to show that~$\alpha$ is good. It suffices to
-    prove that \emph{somewhere}, $\alpha$ is good, as goodness is a geometric
-    implication (in fact even a geometric formula). Without loss of generality,
-    we may suppose~\textsc{lem} as it holds \emph{somewhere}.
-
-    Hence there is an infinite increasing subsequence
-    \[ \beta\,k_0 \leq \beta\,k_1 \leq \ldots. \]
-    As~$Y$ is well\ind, the sequence~$\gamma\,k_0, \gamma\,k_1, \ldots$
-    is good, so there exist~$i < j$ with~$\gamma\,k_i \leq \gamma\,k_j$. Since we
-    also have~$\beta\,k_i \leq \beta\,k_j$, we are done. \qed
-  }
-
   \begin{changemargin}{7.5em}{-0.5em}
@@
         \item $\mathrm{Set}$, the category of sets and maps.
-        \item The category of sets and maps which are \hil{defined up to~$\boldsymbol\neg\neg$}.
+        \item The category of sets and maps which are \hil{defined up to~$\boldsymbol{\neg\neg}$}.
         \item $\mathrm{Set}[G]$, the extension obtained by adding a \hil{generic
@@
 
-    \only<3->{
+    \visible<3->{
       \hil{Multiversal yoga:}
@@
         \smallskip
-        \item A ring element is nilpotent iff all prime ideals \emph{everywhere} contain it.
+        \item A ring element is nilpotent iff \\ all prime ideals \emph{everywhere} contain it.
         \smallskip
-        \item For every inhabited set~$X$, \emph{proximally} there exists an
+        \item For every inhabited set~$X$, \\ \emph{proximally} there exists an
         enumeration~$\NN \twoheadrightarrow X$.
@@
 
-\begin{frame}{Toposophic constructive combinatorics}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{-4.68em}\includegraphics[width=1.2\paperwidth]{wqo-faded}\end{minipage}}
+\begin{frame}{Multiversal constructive combinatorics}
   \justifying
@@
 
-  \emph{Proof.} Let~$\alpha = (\beta,\gamma) : \NN \to X \times Y$ be a sequence in an
-  arbitrary topos. We need to show that~$\alpha$ is good, so find indices~$n < m$ such that
+  \emph{Multiversal constructive proof.} Let~$\alpha = (\beta,\gamma) : \NN \to X \times Y$ be a sequence in an
+  arbitrary topos. We need to show that~$\alpha$ is good, i.\,e.\@ find indices~$n < m$ such that
   \[ \beta\,n \leq \beta\,m \quad\text{and}\quad \gamma\,n \leq \gamma\,m. \]
@@
   also have~$\beta\,k_i \leq \beta\,k_j$, we are done. \qed
-\end{frame}
+\end{frame}}
 
 {\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
-\begin{frame}{Toposophic constructive algebra}
+\begin{frame}{Multiversal constructive algebra}
   \justifying
@@
   ring~$A$. Then~$1 = 0$ in~$A$.
+  \medskip
 
@@
   is a contradiction to basic linear algebra. \qed
+  \medskip
   \pause
@@
 
+\backupstart
+
+\section{Basics of forcing}
+
+\begin{frame}{Ingredients for forcing}
+  \vspace*{-0.3em}
+  To construct a forcing extension, we require:
+  \begin{enumerate}
+    \item a base universe~$V$
+    \item a preorder~$L$ of \hil{forcing conditions} in~$V$\!,
+    pictured as \hil{finite approximations}
+
+    (\emph{convention:} $\tau \preccurlyeq \sigma$ means that~$\tau$ is a
+    better finite approximation than~$\sigma$)
+    \item a \hil{covering system} governing how finite approximations evolve to
+    better ones
+
+    (for each~$\sigma \in L$, a set~$\Cov(\sigma) \subseteq
+    P({\downarrow}\sigma)$, with a simulation condition)
+  \end{enumerate}
+  In the forcing extension~$V^\nabla$, there will then be a \hil{generic filter} (ideal
+  object).
+  \pause
+
+  \vspace*{-1em}
+  \begin{columns}
+    \begin{column}{0.49\textwidth}
+      \small
+      \begin{block}{For the generic surjection~$\NN \twoheadrightarrow X$}
+        \justifying
+        Use \hil{finite lists}~$\sigma \in X^*$ as forcing conditions,
+        where $\tau \preccurlyeq \sigma$ iff~$\sigma$ is an initial segment of~$\tau$,
+        and be prepared to grow~$\sigma$ to \ldots
+        \footnotesize
+        \begin{enumerate}
+          \item[(a)] one of~$\{ \sigma x \,|\, x \in X \}$, to make~$\sigma$ more defined
+          \item[(b)] one of~$\{ \sigma \tau \,|\, \tau \in X^*, a \in
+          \sigma\tau \}$, for any~$a \in X$, to make~$\sigma$ more surjective
+        \end{enumerate}
+        \vspace*{-0.4em}
+      \end{block}
+    \end{column}
+    \pause
+
+    \begin{column}{0.45\textwidth}
+      \small
+      \begin{block}{For the generic prime ideal of a ring~$A$}
+        \justifying
+        Use \hil{f.g.\@ ideals} as forcing conditions, where $\bbb \preccurlyeq
+        \aaa$ iff~$\bbb \supseteq \aaa$, and be prepared to grow~$\aaa$ to \ldots
+        \footnotesize
+        \begin{enumerate}
+          \item[(a)] one of~$\emptyset$, if~$1 \in \aaa$, to make~$\aaa$ more proper
+          \item[(b)] one of~$\{ \aaa+(x), \aaa+(y) \}$, if~$xy \in \aaa$, to
+          make~$\aaa$ more prime
+        \end{enumerate}
+        \vspace*{-0.4em}
+      \end{block}
+    \end{column}
+  \end{columns}
+\end{frame}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{-1cm}\includegraphics[width=\paperwidth]{forest-light-colored}\end{minipage}}
+\begin{frame}{The eventually monad}
+  Let~$L$ be a forcing notion.
+
+  Let~$P$ be a monotone predicate on~$L$
+  (if $\tau \preccurlyeq \sigma$, then $P\sigma \Rightarrow P\tau$). \\
+  For instance, in the case~$L = X^*$:
+  \begin{itemize}
+    \item $\mathsf{Repeats}\, x_0\ldots x_{n-1} \defeqv \exists i\_ \exists j\_ i < j \wedge x_i = x_j$
+    \item $\mathsf{Good}\, \,\,\,\,\,\,x_0\ldots x_{n-1} \defeqv \exists i\_ \exists j\_ i < j \wedge x_i \leq x_j$
+    \quad (for some preorder~$\leq$ on~$X$)
+  \end{itemize}
+  \pause
+
+  We then define~``\hil{$P \mid \sigma$}'' (``$P$ bars~$\sigma$'') inductively by the following clauses:
+  \begin{enumerate}
+    \item If~$P\sigma$, then~$P \mid \sigma$.
+    \item If~$P \mid \tau$ for all~$\tau \in R$, where~$R$ is some covering
+    of~$\sigma$, then~$P \mid \sigma$.
+  \end{enumerate}
+  So~$P \mid \sigma$ expresses in a \hil{direct inductive fashion}:
+  \[ \text{``No matter
+  how~$\sigma$ evolves to a better approximation~$\tau$, eventually~$P\tau$
+  will hold.''} \]
+
+  \pause
+  We use quantifier-like notation: ``$\nabla(\tau \preccurlyeq \sigma)\_
+  P\tau$'' means ``$P \mid \sigma$''.
+\end{frame}}
+% BOARD:
+% - examples for P | σ
+% - abuse of notation
+
+\begin{frame}{Proof translations}
+  \textbf{Thm.} Every~\textsc{iqc}-proof remains correct, with at most a
+  polynomial increase in length, if throughout we
+  replace
+  \[\begin{array}{rcl@{\quad\text{where}\quad}rcl}
+    \exists & \leadsto & \exists^\mathrm{cl},
+    & \exists^\mathrm{cl} &\defeqv& \neg\neg\exists, \\
+    \vee & \leadsto & \vee^\mathrm{cl},
+    & \alpha \vee^\mathrm{cl} \beta &\defeqv& \neg\neg(\alpha \vee \beta), \\
+    = & \leadsto & =^\mathrm{cl},
+    & s =^\mathrm{cl} t &\defeqv& \neg\neg(s = t).
+  \end{array} \]
+  \pause
+
+  \begin{columns}[c]
+    \begin{column}{0.01\textwidth}
+      \includegraphics[height=2.4em]{sheafification-man-2}
+    \end{column}
+    \quad
+    \begin{column}{0.9\textwidth}
+      \hil{When we say:}\ \ some statement ``holds in~$V^{\neg\neg}$'', \\
+      \makesamewidth[l]{\hil{When we say:}}{\hil{we mean:}}\ \ its translation holds in~$V$.
+    \end{column}
+  \end{columns}
+  \bigskip
+
+  Similarly for arbitrary forcing extensions~$V^\nabla$, ``just with~$\nabla$
+  instead of~$\neg\neg$''.
+  \bigskip
+
+  \pause
+  \textbf{Ex.} As~$\neg\neg(\varphi \vee \neg\varphi)$ is a theorem
+  of~\textsc{iqc}, the law of excluded middle holds in~$V^{\neg\neg}$.
+\end{frame}
+
+\newcommand{\defeqvi}{\quad iff\quad}
+\begin{frame}{The $\nabla$-translation}
+  \small
+  \only<1>{For bounded first-order formulas over the (large) first-order signature which has
+  \begin{enumerate}
+    \scriptsize
+    \item one sort~$\underline{X}$ for each set~$X$ in the base universe,
+    \\[-1.2em]
+    \item one~$n$-ary function symbol~$\underline{f} : \underline{X_1} \times
+    \cdots \times \underline{X_n} \to \underline{Y}$ for each map~$f : X_1 \times
+    \cdots \times X_n \to Y$,
+    \\[-1.2em]
+    \item one~$n$-ary relation symbol~$\underline{R} \hookrightarrow
+    \underline{X_1} \times \cdots \times \underline{X_n}$ for each relation~$R
+    \subseteq X_1 \times \cdots \times X_n$, and
+    \\[-1.2em]
+    \item an additional unary relation symbol~$G \hookrightarrow \underline{L}$
+    (for the \emph{generic filter} of~$L$),
+  \end{enumerate}
+  we recursively define:}
+  \scriptsize
+  \only<2->{\vspace*{-0.4em}}
+  \begin{tabbing}
+    \quad \= $\sigma \forces \forces \forall(x\?\underline{X})\_ \varphi$ \=
+    \defeqvi $\textcolor{gray}{\forall(\tau \preccurlyeq \sigma)\_}\
+    \forall(x_0 \in X)\_ \tau \forces \varphi[\underline{x_0}/x]$.\qquad\quad \=
+    $\sigma \forces \exists(x\?\underline{X})\_ \varphi$
+    \= $\sigma \forces \underline{R}(\underline{s_1},\ldots,\underline{s_n})$ \= \defeqvi $s = t$. \= \kill
+
+    \> $\sigma \forces s = t$
+    \> \defeqvi $\nabla \sigma\_ \llbracket s \rrbracket = \llbracket t \rrbracket$.
+    \> $\sigma \forces \underline{R}(s_1,\ldots,s_n)$
+    \> \defeqvi $\nabla\sigma\_ R(\llbracket s_1 \rrbracket,\ldots,\llbracket s_n \rrbracket)$. \\[0.3em]
+
+    \> $\sigma \forces \varphi \Rightarrow \psi$
+    \> \defeqvi $\textcolor{gray}{\forall(\tau \preccurlyeq \sigma)\_}\ (\tau \forces \varphi) \Rightarrow
+    (\tau \forces \psi)$.
+    \> $\sigma \forces G\tau$
+    \> \defeqvi $\nabla\sigma\_ \sigma \preccurlyeq \llbracket\tau\rrbracket$. \\[0.3em]
+
+    \> $\sigma \forces \top$ \> \defeqvi $\top$.
+    \> $\sigma \forces \bot$ \> \defeqvi $\hil{$\nabla\sigma\_$}\ \bot$ \\[0.3em]
+
+    \> $\sigma \forces \varphi \wedge \psi$
+    \> \defeqvi $(\sigma \forces \varphi) \wedge (\sigma \forces \psi)$.
+    \> $\sigma \forces \varphi \vee \psi$
+    \> \defeqvi $\hil{$\nabla\sigma\_$}\ (\sigma \forces \varphi) \vee (\sigma \forces \psi)$. \\[0.3em]
+
+    \> $\sigma \forces \forall(x\?\underline{X})\_ \varphi$
+    \> \defeqvi $\textcolor{gray}{\forall(\tau \preccurlyeq \sigma)\_}\ \forall(x_0 \in X)\_ \tau \forces
+    \varphi[\underline{x_0}/x]$.
+    \> $\sigma \forces \exists(x\?\underline{X})\_ \varphi$
+    \> \defeqvi $\hil{$\nabla\sigma\_$}\ \exists(x_0 \in X)\_ \sigma \forces \varphi[\underline{x_0}/x]$.
+  \end{tabbing}
+  \small
+  \only<1>{Finally, we say that~$\varphi$ ``holds in~$V^\nabla$'' iff for all~$\sigma
+  \in L$, $\sigma \forces \varphi$.}
+
+  \footnotesize
+  \begin{tabular}{@{}lp{0.27\textwidth}p{0.48\textwidth}@{}}
+    \toprule
+    forcing notion & statement about~$V^\nabla$ & external meaning \\
+    \midrule
+    surjection $\NN \twoheadrightarrow X$ &
+    ``the gen.\@ surj.\@ is surjective'' &
+    $\forall(\sigma{\in}X^*)\_ \forall(a{\in}X)\_ \nabla(\tau{\preccurlyeq}\sigma)\_ \exists(n{\in}\NN)\_ \tau[n] = a$. \\[0.4em]
+    \pause
+
+    map $\NN \to X$ &
+    ``the gen.\@ sequence is good'' &
+    $\mathsf{Good} \mid [\,]$. \\[0.4em]
+
+    frame of opens &
+    ``every complex number has a square root'' &
+    For every open~$U \subseteq X$ and every cont.\@
+    function $f : U \to \CC$, there is an open covering $U = \bigcup_i U_i$ such
+    that for each index~$i$, there is a cont.\@ function $g : U_i \to \CC$
+    such that~$g^2 = f$. \\[4.3em]
+
+    big Zariski &
+    ``$x \neq 0 \Rightarrow \text{$x$ inv.}$'' &
+    If the only f.p.\@ $k$-algebra in which~$x = 0$ is the zero algebra,
+    then~$x$ is invertible in~$k$.\\[1.5em]
+    \pause
+
+    little Zariski &
+    ``every f.g. vector space does \emph{not not} have a basis'' &
+    \makesamewidth[l]{}{\phantom{x}}\hil{Grothendieck's generic freeness lemma}
+  \end{tabular}
+\end{frame}
+
+\begin{frame}{Outlook}
+  \begin{block}{Passing to and from extensions}
+    \justifying\small
+    \textbf{Thm.} Let~$\varphi$ be a \hil{bounded first-order formula} not
+    mentioning~$G$. In each of the following situations, we have that
+    $\varphi$ holds in~$V^\nabla$ iff~$\varphi$
+    holds in $V$:
+
+    \vspace*{-0.5em}
+    \begin{enumerate}
+      \item $L$ and all coverings are inhabited (proximality). \\[-1em]
+      \item $L$ contains a top element, every covering of the
+      top element is inhabited, and~$\varphi$ is a coherent implication
+      (positivity).
+    \end{enumerate}
+    \vspace*{-0.8em}
+  \end{block}
+
+  \vspace*{-1.5em}
+  \begin{columns}
+    \begin{column}{0.46\textwidth}
+      \begin{block}{The mystery of nongeometric sequents}
+        \justifying
+        The \hil{generic ideal} of a ring is maximal:
+        \vspace*{-1em}
+        \[ (x \in \aaa \Rightarrow 1 \in \aaa) \Longrightarrow 1 \in \aaa + (x). \]
+
+        The \hil{generic ring} is a field:
+        \vspace*{-0.7em}
+        \[ (x = 0 \Rightarrow 1 = 0) \Longrightarrow (\exists y\_ xy = 1). \]
+        \vspace*{-1.4em}
+      \end{block}
+
+    \end{column}
+
+    \begin{column}{0.50\textwidth}
+      \begin{block}{Traveling the multiverse \ldots}
+        \textsc{lem} is a \hil{switch} and \hil{holds positively};
+        being countable is a \hil{button}.
+        \medskip
+
+        Every instance of \textsc{dc} \hil{holds proximally}.
+        \medskip
+
+        A geometric implication is provable iff it holds \hil{everywhere}.
+        \vspace*{-0.2em}
+      \end{block}
+      \vspace*{-0.3em}
+      \hfill\footnotesize\ldots{} upwards, but always keeping ties to the
+      base.{\ }
+    \end{column}
+  \end{columns}
+\end{frame}
+
+\begin{frame}{More on forcing notions}
+  \small
+  \textbf{Def.} A \hil{forcing notion} consists of
+  a preorder~$L$ of \hil{forcing conditions}, and
+  for every~$\sigma \in L$, a set~$\Cov(\sigma) \subseteq
+  P({\downarrow}\sigma)$ of \hil{coverings} of~$\sigma$
+  such that: If~$\tau \preccurlyeq \sigma$ and~$R \in \Cov(\sigma)$, there
+  should be a covering~$S \in \Cov(\tau)$ such that~$S \subseteq
+  {\downarrow}R$.
+  \bigskip
+
+  {\centering\footnotesize\begin{tabular}{llll}
+    \toprule
+    & preorder~$L$ & coverings of an element~$\sigma \in L$ & filters of~$L$ \\
+    \midrule
+    \normalnumber{1} & $X^*$ & $\{ \sigma x \,|\, x \in X \}$ & maps~$\NN \to X$ \\
+    \normalnumber{2} & $X^*$ & $\{ \sigma x \,|\, x \in X \}$,\ \ $\{ \sigma\tau \,|\, \tau \in X^*, a \in \sigma\tau \}$ for each~$a \in X$ & surjections~$\NN \twoheadrightarrow X$ \\
+    \normalnumber{3} & f.g. ideals & --- & ideals \\
+    \normalnumber{4} & f.g. ideals & $\{ \sigma+(a), \sigma+(b) \}$ for each~$ab \in \sigma$,\ \ $\{\}$ if~$1 \in \sigma$ & prime ideals \\
+    \normalnumber{5} & opens & $\mathcal{U}$ such that~$\sigma = \bigcup \mathcal{U}$ & points \\
+    \normalnumber{6} & $\{\star\}$ & $\{ \star \,|\, \varphi \} \cup \{ \star \,|\, \neg\varphi \}$ &
+    witnesses of~\textsc{lem}
+    \\
+    \bottomrule
+  \end{tabular}\par}
+  \bigskip
+
+  \textbf{Def.} A \emph{filter} of a forcing notion~$(L,\mathrm{Cov})$
+  is a subset~$F \subseteq L$ such that
+  \vspace*{-0.4em}
+  \begin{enumerate}
+    \scriptsize
+    \item $F$ is upward-closed: if~$\tau \preccurlyeq \sigma$ and if~$\tau \in F$, then~$\sigma \in F$; \\[-3.0em]
+    \item $F$ is downward-directed: $F$ is inhabited, and if~$\alpha,\beta \in F$,
+    then there is a common refinement~$\sigma \preccurlyeq \alpha,\beta$ such
+    that~$\sigma \in F$; and \\[-2.0em]
+    \item $F$ splits the covering system: if~$\sigma \in F$ and~$R \in
+    \Cov(\sigma)$, then~$\tau \in F$ for some~$\tau \in R$.
+  \end{enumerate}
+\end{frame}
+
+\backupend
+
 \end{document}

### 2025-02-27 slides-abmv2024.tex
More notes

@@
     Hence, for the purpose of verifying a bounded first-order assertion about
-    the base, we can freely pass to a positiv overt topos with problem-adapted
+    the base, we can freely pass to a positive overt topos with problem-adapted
     better higher-order properties (such as that some uncountable set from the

### 2025-02-27 slides-paris2025.tex
More notes

@@
 \begin{frame}{Well quasi-orders}
+  \jnote{1}{
+    Well quasi-orders are an important notion in proof theory and termination
+    analysis.
+  }
+
   \jnote{2-3}{
@@
   \jnote{4}{
+    \vspace*{-1em}
     The dependence of the theory on well quasi-orders on classical transfinite
-    methods is already present in one of basic observations of this theory:
+    methods is already present in one of the first and central observations of this theory:
 
@@
     smallest possible next index in~$\NN \setminus K$, doable by yet another
-    invocation of~\bad{\textsc{lem}}, but the result remains fundamentally
+    invocation of~\bad{\textsc{lem}}. But the result remains fundamentally
     noneffective---in the special case~$X = (\{ 0, 1 \}, {=})$, the statement
@@
     \emph{entailment relations}). Constructive forcing allows us to reify the
-    generic prime ideal as an actual prime ideal in a suitable forcing extension.
+    generic prime ideal as an actual prime ideal in a suitable forcing extension,
+    so in a suitable topos (der little Zariski topos of the ring).
   }
@@
 \begin{frame}{Multiversal constructive combinatorics}
+  \jnote{1}{
+    The displayed multiversal proof closely mimics the classical proof (for well\sinf), but
+    is fully constructive (for well\ind). It would be possible to streamline
+    this proof and unroll the topos-theoretic machinery, to obtain an explicit
+    algorithm of type
+    \[ \mathsf{Good} \mid_X [\,] \times \mathsf{Good} \mid_Y [\,] \ \longrightarrow\
+      \mathsf{Good} \mid_{X \times Y} [\,].
+    \]
+
+    The modal language was recently used to answer a question by Stefano
+    Berardi, Gabriele Buriola and Peter Schuster, see
+    \href{https://www.speicherleck.de/iblech/stuff/slides-abmv2024.pdf\#page=16}{this
+    set of slides}.
+  }
+
   \justifying
@@
 \begin{frame}{Multiversal constructive algebra}
+  \jnote{1-}{
+    The displayed classical proof is quite efficient from the point of view of
+    organizing mathematical knowledge, as it quickly reduces the general
+    situation of dealing with an arbitrary ring to dealing with a field.
+    Alas, read literally, it is hopeless ineffective.
+  }
+
+  \jnote{2-}{
+    By employing modal language, we can closely mimic the original proof and be
+    fully constructive at the same time.
+  }
+
+  \jnote{3-}{
+    By unwinding all modal definitions, the modal proof can be unrolled to a
+    fully explicit computation.
+  }
+
   \justifying
@@
 \backupstart
+
+\begin{frame}
+  \emph{Backup slides.}
+\end{frame}
 

### 2025-06-08 slides-types2025.tex
Outline of slides for TYPES 2025

- General structure like in HoTTest / ABMV?

- Proof of Berardi–Buriola–Schuster conjecture
- Higher-order functions defined everywhere (in every topos): dialogue trees
- Generic sequence is good = inductively good
- Fun properties of generic surjection

- Let's play Agda

---

1. Hilbert's program
   - basics about wqo's
   - subsequence lemma
   - Dickson
   - inductive reimagination
   - questions: where from? why work? how much stronger?

2. Where do our most cherished definitions come from?
   - a familiar business: enlarging mathematical structures
   - adjoining generic function, surjection, prime ideal, ...
   - observations:
     - well iff generic sequence is good iff everywhere, every sequence is good
       (in particular, every multivalued and every up-to-¬¬-defined sequence is good)
     - wellfounded iff for the generic decreasing sequence, ⊥ iff ...
   - fun facts about generic gadgets

3. Modal language for harnessing the multiverse
   - definition
   - examples
   - implicational wellness
   - dialogue

4. Implementation
   - idea: realize generic gadget as some kind of limit of approximations from
     the base
   - reinterpret, in a mechanic fashion, assertions about generic objects as
     assertions about approximations
     - ex.: "the generic function is defined on input n" as "no matter how the
       empty list evolves over a time to a better approximation xs, eventually
       xs will have length at least succ n"
   - crucially, this interpretation is sound with respect to intuitionistic
     reasoning

---

Towards topological type theory for decrypting transfinite methods in classical mathematics

In constructive mathematics, an ongoing challenge is to extracting
computational content from infinitary arguments in classical
mathematics—especially those that yield concrete results via abstract tools
such as minimal bad sequences or maximal ideals. Today, a wide range of such
techniques exists, constituting a late partial fulfillment of Hilbert's
program.

This talk presents work in progress on porting the modal operators "everywhere"
and "somewhere", as developed in set theory by Joel David Hamkins, Victoria
Gitman and their collaborators, into type theory. This modal enrichment
provides a uniform language for expressing several established
constructivization techniques, and is at the core of a relatively recent new
technique that offers insights on the origins of some of our most cherished
inductive definitions. These developments are closely connected to recent
advances in type-theoretic presheaf and sheaf models, dialogue
continuity, synthetic algebraic geometry and well-quasi-order theory.

---

Which functions are we excluding when we form the type of all functions between
two types? Does every field have an algebraic closure?

Inspired by the modal approach to the set-theoretic multiverse of Joel David
Hamkins, Victoria Gitman and their collaborators, we aim to introduce the modal
operators "everywhere" and "somewhere" to homotopy type theory, proposing a
multiversal perspective on these motivating questions.

---

Well quasi-orders are a combinatorial notion with featureful applications in
graph theory, termination checking, commutative algebra and other subjects.
However, their classical definition, though concise and elegant, poses
significant challenges in constructive mathematics and frameworks that lack
function sets. In response, several constructive substitutes have been
developed, most recently an implicational definition by Stefano Berardi,
Gabriele Buriola and Peter Schuster. In this talk, we investigate how a modal
approach can reinterpret classical proofs involving transfinite methods as
blueprints for constructive proofs grounded in these alternatives, thereby
combining the best of both worlds: Short and abstract proofs, but with
constructive content. We also indicate how the modal approach allows us to
compare the implicational definition with an earlier inductive one.

---

Combinatorics and commutative algebra abound with situations where we prove
quite concrete results by quite abstract transfinite methods, such as minimal
bad sequences or maximal ideals. Amazingly, such infinitary arguments can often
be understood as blueprints for quite explicit computations—as called for by
Hilbert’s programme. In this talk, we will travel the modal toposophic
multiverse to facilitate this kind of mining abstract proofs for refined
quantitative results. We will use a celebrated theorem from order theory about
sequences N→N that everybody can relate with as a running example.

---

Where do some of our most cherished inductive definitions come from? Which
functions are we excluding when we form the type of all functions between two
types? Does every field have an algebraic closure?

Inspired by the modal approach to the set-theoretic multiverse of Joel David
Hamkins, Victoria Gitman and their collaborators, we aim to introduce the modal
operators "everywhere" and "somewhere" to homotopy type theory, proposing a
multiversal perspective on these motivating questions.

Unlike the set-theoretic role model, we focus less on exploring the range of
foundational possibility and more on concrete applications in constructive
mathematics, with the goal of porting results of classical mathematics to
homotopy type theory. It will turn out that every field has an algebraic
closure somewhere; that a transitive relation is well-founded iff nowhere there
is an infinite descending chain; and that somewhere, the law of excluded middle
holds.

This is ongoing joint work with Alexander Oldenziel and connected to the recent
advances with type-theoretic presheaf models and sheaf models.

### 2025-06-08 slides-types2025.tex
More work on the slides

@@
+\documentclass[11pt,utf8,notheorems,compress,t,aspectratio=169]{beamer}
+\usepackage{etex}
+
+\usepackage{pgfpages}
+\usepackage[export]{adjustbox}
+
+% Workaround for the issue described at
+% https://tex.stackexchange.com/questions/164406/beamer-using-href-in-notes.
+\newcommand{\fixedhref}[2]{\makebox[0pt][l]{\hspace*{\paperwidth}\href{#1}{#2}}\href{#1}{#2}}
+
+\usepackage[english]{babel}
+
+\usepackage{mathtools}
+\usepackage{xspace}
+\usepackage{booktabs}
+\usepackage{stmaryrd}
+\usepackage{amssymb}
+\usepackage{manfnt}
+\usepackage{array}
+\usepackage{ragged2e}
+\usepackage{multicol}
+\usepackage{tabto}
+\usepackage{xstring}
+\usepackage{proof}
+\usepackage{agda}
+\usepackage[all]{xy}
+\xyoption{rotate}
+\usepackage{tikz}
+\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing,positioning,svg.path}
+\hypersetup{colorlinks=true}
+
+\DeclareFontFamily{U}{bbm}{}
+\DeclareFontShape{U}{bbm}{m}{n}
+   {  <5> <6> <7> <8> <9> <10> <12> gen * bbm
+      <10.95> bbm10%
+      <14.4>  bbm12%
+      <17.28><20.74><24.88> bbm17}{}
+\DeclareFontShape{U}{bbm}{m}{sl}
+   {  <5> <6> <7> bbmsl8%
+      <8> <9> <10> <12> gen * bbmsl
+      <10.95> bbmsl10%
+      <14.4> <17.28> <20.74> <24.88> bbmsl12}{}
+\DeclareFontShape{U}{bbm}{bx}{n}
+   {  <5> <6> <7> <8> <9> <10> <12> gen * bbmbx
+      <10.95> bbmbx10%
+      <14.4> <17.28> <20.74> <24.88> bbmbx12}{}
+\DeclareFontShape{U}{bbm}{bx}{sl}
+   {  <5> <6> <7> <8> <9> <10> <10.95> <12> <14.4> <17.28>%
+      <20.74> <24.88> bbmbxsl10}{}
+\DeclareFontShape{U}{bbm}{b}{n}
+   {  <5> <6> <7> <8> <9> <10> <10.95> <12> <14.4> <17.28>%
+      <20.74> <24.88> bbmb10}{}
+\DeclareMathAlphabet{\mathbbm}{U}{bbm}{m}{n}
+\SetMathAlphabet\mathbbm{bold}{U}{bbm}{bx}{n}
+
+\usepackage{pifont}
+\newcommand{\cmark}{\ding{51}}
+\newcommand{\xmark}{\ding{55}}
+\DeclareSymbolFont{extraup}{U}{zavm}{m}{n}
+\DeclareMathSymbol{\varheart}{\mathalpha}{extraup}{86}
+
+\graphicspath{{images/}}
+
+\usepackage[protrusion=true,expansion=true]{microtype}
+
+\setlength\parskip{\medskipamount}
+\setlength\parindent{0pt}
+
+\title{Modal operators for a constructive account of well quasi-orders}
+
+\author{Ingo Blechschmidt}
+\date{November 30th, 2024}
+
+%\setbeameroption{show notes on second screen=bottom}
+\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\footnotesize\justifying#2\par}}}
+
+%\useinnertheme[shadow=true]
+\setbeamerfont{block title}{size={}}
+
+\useinnertheme{rectangles}
+
+\usecolortheme{orchid}
+\usecolortheme{seahorse}
+\definecolor{mypurple}{RGB}{253,73,34}
+\definecolor{mypurpledark}{RGB}{100,0,150}
+\setbeamercolor{structure}{fg=mypurple}
+\setbeamercolor*{title}{bg=mypurple,fg=white}
+\setbeamercolor*{titlelike}{bg=mypurple,fg=white}
+\setbeamercolor{frame}{bg=black}
+
+\usefonttheme{serif}
+\usepackage[T1]{fontenc}
+\usepackage{libertine}
+
+% lifted from https://arxiv.org/abs/1506.08870
+\DeclareFontFamily{U}{min}{}
+\DeclareFontShape{U}{min}{m}{n}{<-> udmj30}{}
+\newcommand\yon{\!\text{\usefont{U}{min}{m}{n}\symbol{'210}}\!}
+
+\newcommand{\A}{\mathcal{A}}
+\newcommand{\B}{\mathcal{B}}
+\newcommand{\C}{\mathcal{C}}
+\newcommand{\M}{\mathcal{M}}
+\renewcommand{\AA}{\mathbb{A}}
+\newcommand{\BB}{\mathbb{B}}
+\newcommand{\pp}{\mathbbm{p}}
+\newcommand{\MM}{\mathbb{M}}
+\newcommand{\E}{\mathcal{E}}
+\newcommand{\F}{\mathcal{F}}
+\newcommand{\FF}{\mathbb{F}}
+\newcommand{\G}{\mathcal{G}}
+\newcommand{\J}{\mathcal{J}}
+\newcommand{\GG}{\mathbb{G}}
+\renewcommand{\O}{\mathcal{O}}
+\newcommand{\K}{\mathcal{K}}
+\newcommand{\NN}{\mathbb{N}}
+\newcommand{\QQ}{\mathbb{Q}}
+\newcommand{\RR}{\mathbb{R}}
+\newcommand{\TT}{\mathbb{T}}
+\newcommand{\PP}{\mathbb{P}}
+\newcommand{\ZZ}{\mathbb{Z}}
+\newcommand{\CC}{\mathbb{C}}
+\renewcommand{\P}{\mathcal{P}}
+\newcommand{\aaa}{\mathfrak{a}}
+\newcommand{\bbb}{\mathfrak{b}}
+\newcommand{\ccc}{\mathfrak{c}}
+\newcommand{\ppp}{\mathfrak{p}}
+\newcommand{\fff}{\mathfrak{f}}
+\newcommand{\mmm}{\mathfrak{m}}
+\newcommand{\defeq}{\vcentcolon=}
+\newcommand{\defeqv}{\vcentcolon\equiv}
+\newcommand{\Cov}{\mathrm{Cov}}
+\renewcommand{\_}{\mathpunct{.}}
+\newcommand{\?}{\,{:}\,}
+\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
+\newcommand{\inv}{inv.\@}
+\newcommand{\forces}{\vDash}
+\newcommand{\ind}{\ensuremath{_\text{ind}}\xspace}
+\newcommand{\sinf}{\ensuremath{_\infty}\xspace}
+\newcommand{\impl}{\ensuremath{_\text{impl}}\xspace}
+
+\setbeamertemplate{blocks}[shadow=false]
+
+\newenvironment{indentblock}{%
+  \list{}{\leftmargin\leftmargin}%
+  \item\relax
+}{%
+  \endlist
+}
+
+% Adapted from https://latex.org/forum/viewtopic.php?t=2251 (Stefan Kottwitz)
+\newenvironment<>{hilblock}{
+  \begin{center}
+    \begin{minipage}{9.05cm}
+      \setlength{\textwidth}{9.05cm}
+      \begin{actionenv}#1
+        \def\insertblocktitle{}
+        \par
+        \usebeamertemplate{block begin}}{
+        \par
+        \usebeamertemplate{block end}
+      \end{actionenv}
+    \end{minipage}
+  \end{center}}
+
+\newenvironment{changemargin}[2]{%
+  \begin{list}{}{%
+    \setlength{\topsep}{0pt}%
+    \setlength{\leftmargin}{#1}%
+    \setlength{\rightmargin}{#2}%
+    \setlength{\listparindent}{\parindent}%
+    \setlength{\itemindent}{\parindent}%
+    \setlength{\parsep}{\parskip}%
+  }%
+  \item[]}{\end{list}}
+
+\tikzset{
+  invisible/.style={opacity=0,text opacity=0},
+  visible on/.style={alt={#1{}{invisible}}},
+  alt/.code args={<#1>#2#3}{%
+    \alt<#1>{\pgfkeysalso{#2}}{\pgfkeysalso{#3}}}
+}
+
+% https://tex.stackexchange.com/questions/172336/drawing-roman-laurel-leaves-spqr-in-tikz
+\tikzset{
+  laurel-wreath/.pic = {
+    \fill svg{M14.4-24.6c-1.5-1.5-2.6-3.3-3.1-5.3l-.4-1.7c-.2-1.1-.2-4.1 .2-5.7 .2-.9 .3-1.3 .5-1.3l1.4 1.1 2.5 2.4c2.7 2.5 5.2 6 5.8 8 .2 .6-.5 .3-2.2-.9-1.6-1.3-3.3-2.6-5-3.8l.1 1.4c.2 1.4 .5 2.7 1.1 4.6s.8 2.5 .5 2.5l-1.4-1.3zm69.6 1.1 .3-1.2c.8-2.3 1.3-4.8 1.6-7.3l-1.5 1.1c-1.3 .9-2.6 1.9-3.7 3-1.6 1.1-2 1.3-2.1 1 .7-1.8 1.6-3.4 2.8-4.9 1.3-1.7 6.5-6.8 7-6.8 .2 0 .3 .2 .3 .5l.3 1.6c.3 2.2 .2 5.7-.5 7.4-.8 1.9-1.6 3.1-3 4.7-1.1 1.1-1.4 1.3-1.5 .9z};
+    \fill svg{M10-29.4c-.8-1.1-1.4-2.2-2-4.1l-.7-3.5c-.2-3 .2-4.4 1.4-8.3l.5-1.4c.2-1.3 .3-1.9 .6-1.9 .3-.2 .6 .3 .7 .8s.9 2.2 1.9 3.6c1.4 2.2 2.7 4.4 3.9 6.6l.9 2.7c0 .6 0 .6-.3 .6-.6 0-4.9-4.4-5.8-6l-.2-.6-.1 1.7-.3 2.8c-.3 2.7-.3 3.8 0 5.5 .6 2 .5 2.4-.5 1.5zm79.2 .3 .4-2.4c.2-1.3 .2-2.7-.1-4.9l-.3-2.8v-1.6l-.7 1c-.8 1.3-5 5.5-5.5 5.5s-.5-.3 .2-1.9c.5-1.7 1.4-3.3 3.3-6.5 2.4-3.6 2.7-3.9 2.8-4.7 .5-1.3 .5-1.4 .8-1.2 .3 0 .6 .8 .6 1.5l.7 2.4c.9 2.7 1.1 3.6 1.2 6 .2 3.1-.5 6-2 8.2-.8 1.3-1.3 1.7-1.4 1.5z};
+    \fill svg{M5-40c-.4-3.2-.1-6.5 .9-9.6 .5-1.1 1.6-2.8 2.2-3.4l1.3-1.6 2-2.7 .2 .6c.1 1.3 .4 2.6 .9 3.8l.3 1c.8 1.7 1.1 2.7 1.6 5.3 .6 2.5 .6 4.6 .2 4.6-.3 0-.9-.8-1-1.1l-.5-.8c-1.4-2-3-5.2-2.9-6.5-.9 2.7-2 5.4-3.5 7.9l-.3 .8-.3 .8c0 .5-.6 1.6-.8 1.6l-.3-.7zm89.2 .2-.2-.5-.3-.9-1.1-2.7-1.1-2.4c-.6-1.4-1.2-2.8-1.6-4.2l-.3 .9c-.3 1.3-1.6 3.9-3 6-1.3 2-1.6 2-1.5 0s1.1-6.3 2.2-9c.8-1.7 1.1-3.1 .9-4.1-.2-1.1 .5-.8 2.2 1.8 3.3 4.4 3.8 5.4 4.4 7.8 .6 2.4 .5 7.7-.3 7.8l-.3-.5z};
+    \fill svg{M13.9-50.1c-.5-1.9-.8-3.9-.9-5.8-.2-1.6-.1-3.3 .1-4.9-.3 .8-1.7 2.5-4.2 5.1l-3 4.9-.3 .1c-.3 0-.3-2.2 0-3.3 .8-3 1.4-4.6 2.5-6.1 .9-1.3 1.7-1.9 2.5-2.5 1.1-.6 2.7-1.9 3.5-2.7 .9-.9 1.9-1.4 2.2-1.4v1.1l-.3 6.6c0 6.8 .2 6.3-1 8.9-.5 1.1-.8 1.1-1.1 0zm70.8-.4c-.8-2.2-.8-2.5-.7-6.3-.1-2.7-.1-5.5-.2-8.2-.3-1.6-.3-1.9 .5-1.6l.6 .5c1.4 1.4 3 2.5 3.9 3.1 1.3 .9 1.9 1.6 2.7 2.6l.6 .7 .2 .4 .2 .3c.8 .9 2 4.9 2 6.9 .2 1.9-.2 1.9-.9 .5-.7-1.4-1.5-2.7-2.6-4-1.6-1.5-3-3.2-4.2-5 .4 3 .3 6-.5 9 0 .8-.5 2.2-.8 2.3-.2 0-.5-.3-.8-1.2z};
+    \fill svg{M16.4-58.5l.2-1.5 .3-3.7c.2-2.8 .3-3.5 1.1-5.4l.7-1.3-.5 .4-1 .7c-.5 .4-1.1 .8-1.5 1.3l-.5 .3-1.9 1.6c-2.2 1.6-2.7 2-3.9 3.6-.5 .8-1.1 1.3-1.3 1.3-.5 0 0-2.4 1.1-4.7 1.5-3.4 4.3-6 7.7-7.4l1.3-.4 1.9-.4 2-.5c1.4 0 1.4 0 1 1.1-.5 .8-.8 2-1.1 4.2-.3 2.3-1.1 4.5-2.2 6.5l-.4 .6c-.6 1.1-1.3 2.1-2 3.2-.5 .6-.8 .8-1 .5zm66.3-.2c-.8-.9-2.8-4.4-3.5-6.1-.6-1.3-.9-2.5-1.1-3.5-.2-2.1-.7-4.1-1.5-6 0-.3 0-.3 1.2-.3l2.1 .5 1.9 .4 1.2 .4 .6 .1 1 .6c3 1.4 5.7 4.6 6.8 8.5l.7 2.6c-.2 .6-.5 .5-1.4-.7-2.2-2.7-4.8-5-7.7-6.9l-1.7-1.3 .6 1.3c.3 .6 .6 1.2 .8 1.9l.3 2.5 .3 3.9c.3 2.4 .2 2.8-.6};
+    \fill svg{M21.6-66.1l.4-1.1 .9-3.2c.3-1.9 1.1-3.3 2.4-4.7l.4-.8-1.2 .2-2.2 .3c-2.7 .3-5.3 1.2-7.7 2.5-.6 .5-1.3 .6-1.3 .3 0-.5 .9-1.9 2-2.9 .8-.9 2-1.9 3.2-2.6l.9-.4 2.2-1c.3-.2 1.3-.3 3.2-.1 3 0 4.1 .2 6.3 .7l1.1 .4c.5 .2 .6 .6 .3 .6-.5 0-1.4 .9-1.9 1.7l-1.2 1.8c-1.7 2.8-2.2 3.5-4.6 5.9l-3 2.7-.2-.3zm53.9-2c-2.7-2.8-3.5-3.8-5.4-6.8-.9-1.6-1.4-2.4-1.9-2.5l-.8-.5c-.3 0-.2-.5 .4-.6l1.1-.4c1.9-.6 3-.8 5.6-.9l3.3 .2c2 .6 3.8 1.5 5.4 2.8 .3 0 1.9 1.6 2.5 2.4l.9 1.8c0 .3-.3 .2-1.9-.6-2.8-1.4-4.4-1.9-7.7-2.2l-2.2-.5c-.9-.2-.9-.2-.6 .2 .6 .5 1.7 2 2.1 2.8l.9 2.5c.3 1.5 .6 3 .9 4.6l-2.6-2.3z};
+    \fill svg{M34.1-78.7c-3.4-1.3-6.9-2.1-10.6-2.5-.9 0-1.4 0-2.3 .3-2 .5-2 0 0-1.3l2.8-1.2c1.4-.5 1.9-.5 3.8-.6 3.8-.2 6.1 .3 9.3 1.7l3.6 1.1 2.2 .3c1.3 0 1.7 0 2.7-.3 1.1-.3 2.8-1.1 2.8-1.3l-1.3-.9c-1.9-1.4-3.1-2.7-3.1-3.2l.8-.6c.9-.3 1.3-.2 2 .8 .5 .8 1.1 1.4 2.9 2.7 .2 .3 .3 .2 1.1-.3 .9-.8 2.4-2 2.6-2.7 .5-.6 .9-.8 1.8-.5l.8 .6c0 .5-1.4 1.7-3.2 3.2l-1.3 .9c0 .2 1.7 .9 2.9 1.3 .9 .3 1.4 .3 2.7 .3l2.2-.3c1.7-.4 3.4-1 5-1.7 2-.8 4.4-1.3 7.7-1.1 2 .2 2.5 .2 3.8 .6 .9 .3 2.2 .8 2.8 1.2 2 1.1 2 1.6 .2 1.3-1.6-.3-1.9-.3-4.4 0-2.4 .3-4.7 .8-7 1.6l-1.5 .6c-2.9 .3-5.9 .2-8.8-.3-1.7-.3-3.6-.9-6-2.1l-1.1-.4-1.3 .6c-4.5 2.2-9.6 3-14.6 2.2zm-6.3-9.1c};
+  }
+}
+
+\newcommand{\pointthis}[3]{%
+  \tikz[remember picture,baseline]{
+    \node[anchor=base,inner sep=0,outer sep=0] (#2) {#2};
+    \node[visible on=#1,overlay,rectangle callout,rounded corners,callout relative pointer={(0.3cm,0.5cm)},fill=blue!20] at ($(#2.north)+(-0.1cm,-1.1cm)$) {#3};
+  }%
+}
+
+\tikzset{
+  invisible/.style={opacity=0,text opacity=0},
+  visible on/.style={alt={#1{}{invisible}}},
+  alt/.code args={<#1>#2#3}{%
+    \alt<#1>{\pgfkeysalso{#2}}{\pgfkeysalso{#3}}}
+}
+
+\newcommand{\hcancel}[5]{%
+  \tikz[baseline=(tocancel.base)]{
+    \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
+    \draw[red!80, line width=0.4mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
+  }%
+}
+
+\newcommand{\explain}[7]{%
+  \tikz[remember picture,baseline]{
+    \node[anchor=base,inner sep=2pt,outer sep=0,fill=#3,rounded corners] (label) {#1};
+    \node[anchor=north,visible on=<#2>,overlay,rectangle callout,rounded corners,callout
+    relative pointer={(0.0cm,0.5cm)+(0.0cm,#6)},fill=#3] at ($(label.south)+(0,-0.3cm)+(#4,#5)$) {#7};
+  }%
+}
+
+\newcommand{\explainstub}[2]{%
+  \tikz[remember picture,baseline]{
+    \node[anchor=base,inner sep=2pt,outer sep=0,fill=#2,rounded corners] (label) {#1};
+  }%
+}
+
+\newcommand{\squiggly}[1]{%
+  \tikz[remember picture,baseline]{
+    \node[anchor=base,inner sep=0,outer sep=0] (label) {#1};
+    \draw[thick,color=red!80,decoration={snake,amplitude=0.5pt,segment
+    length=3pt},decorate] ($(label.south west) + (0,-2pt)$) -- ($(label.south east) + (0,-2pt)$);
+  }%
+}
+
+% Adapted from https://latex.org/forum/viewtopic.php?t=2251 (Stefan Kottwitz)
+\newenvironment<>{varblock}[2]{\begin{varblockextra}{#1}{#2}{}}{\end{varblockextra}}
+\newenvironment<>{varblockextra}[3]{
+  \begin{center}
+    \begin{minipage}{#1}
+      \begin{actionenv}#4
+        {\centering \hil{#2}\par}
+	\def\insertblocktitle{}%\centering #2}
+        \def\varblockextraend{#3}
+	\usebeamertemplate{block begin}}{
+        \par
+        \usebeamertemplate{block end}
+        \varblockextraend
+      \end{actionenv}
+    \end{minipage}
+  \end{center}}
+
+\setbeamertemplate{headline}{}
+
+\setbeamertemplate{frametitle}{%
+  \leavevmode%
+  \vskip-1.6em%
+  \begin{beamercolorbox}[dp=1ex,center,wd=\paperwidth,ht=2.25ex]{title}%
+    \vskip0.5em%
+    \bf\insertframetitle
+  \end{beamercolorbox}%
+
+  \vskip-0.77em\hspace*{-2em}%
+  \textcolor{mypurpledark}{\rule[0em]{1.1\paperwidth}{2.4pt}}
+
+  \vskip-0.4em%
+}
+
+\setbeamertemplate{navigation symbols}{}
+
+\newcounter{framenumberpreappendix}
+\newcommand{\backupstart}{
+  \setcounter{framenumberpreappendix}{\value{framenumber}}
+}
+\newcommand{\backupend}{
+  \addtocounter{framenumberpreappendix}{-\value{framenumber}}
+  \addtocounter{framenumber}{\value{framenumberpreappendix}}
+}
+
+\newcommand{\insertframeextra}{}
+\setbeamertemplate{footline}{%
+  \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex,dp=1ex,right,rightskip=1mm,leftskip=1mm]{}%
+    % \inserttitle
+    \hfill
+    \insertframenumber\insertframeextra\,/\,\inserttotalframenumber
+  \end{beamercolorbox}%
+  \vskip0pt%
+}
+
+\newcommand{\hil}[1]{{\usebeamercolor[fg]{item}{\textbf{#1}}}}
+\newcommand{\hill}[1]{{\usebeamercolor[fg]{item}{#1}}}
+\newcommand{\bad}[1]{\textcolor{red!90}{\textnormal{#1}}}
+\newcommand{\good}[1]{\textcolor{mypurple}{\textnormal{#1}}}
+
+\newcommand{\bignumber}[1]{%
+  \renewcommand{\insertenumlabel}{#1}\scalebox{1.2}{\!\usebeamertemplate{enumerate item}\!}
+}
+\newcommand{\normalnumber}[1]{%
+  {\renewcommand{\insertenumlabel}{#1}\!\usebeamertemplate{enumerate item}\!}
+}
+\newcommand{\bigheart}{\includegraphics{heart}}
+
+\newcommand{\subhead}[1]{{\centering\textcolor{gray}{\hrulefill}\quad\textnormal{#1}\quad\textcolor{gray}{\hrulefill}\par}}
+
+\newcommand{\badbox}[1]{\colorbox{red!30}{#1}}
+\newcommand{\infobox}[1]{\colorbox{yellow!70}{\color{black}#1}}
+
+% taken from JDH "The modal logic of arithmetic potentialism and the universal algorithm"
+\DeclareMathOperator{\possible}{\text{\tikz[scale=.6ex/1cm,baseline=-.6ex,rotate=45,line width=.1ex]{\draw (-1,-1) rectangle (1,1);}}}
+\DeclareMathOperator{\necessary}{\text{\tikz[scale=.6ex/1cm,baseline=-.6ex,line width=.1ex]{\draw (-1,-1) rectangle (1,1);}}}
+\DeclareMathOperator{\xpossible}{\text{\tikz[scale=.6ex/1cm,baseline=-.6ex,rotate=45,line width=.1ex]{\draw (-1,-1) rectangle (1,1); \draw[very thin] (-.6,-.6) rectangle (.6,.6);}}}
+\DeclareMathOperator{\xnecessary}{\text{\tikz[scale=.6ex/1cm,baseline=-.6ex,line width=.1ex]{\draw (-1,-1) rectangle (1,1); \draw[very thin] (-.6,-.6) rectangle (.6,.6);}}}
+
+\input{images/primes.tex}
+
+\newcommand{\triang}{\hil{$\blacktriangleright$}}
+\newcommand{\concat}{\mathbin{{+}\mspace{-8mu}{+}}}
+
+\newcommand{\astikznode}[2]{\tikz[baseline,remember picture]{\node[anchor=base,inner sep=0,outer sep=0.1em] (#1) {#2};}}
+\newcommand{\astikznodecircled}[3]{\tikz[baseline,remember picture]{\node[anchor=base,circle,draw=#2,thick,inner sep=0.05em,outer sep=0.05em] (#1) {#3};}}
+\newcommand{\astikznodetransparentlycircled}[2]{\tikz[baseline,remember picture]{\node[anchor=base,circle,opacity=0,draw=white,text opacity=1,thick,inner sep=0.05em,outer sep=0.05em] (#1) {#2};}}
+
+\setbeamersize{text margin left=1.60em,text margin right=1.60em}
+
+\newlength\stextwidth
+\newcommand\makesamewidth[3][c]{%
+  \settowidth{\stextwidth}{#2}%
+  \makebox[\stextwidth][#1]{#3}%
+}
+
+\newcommand{\dnote}[1]{%
+  \begin{tabular}{@{}m{2em}@{}m{0.83\textwidth}@{}}%
+    \textdbend &#1%
+  \end{tabular}%
+  \par
+}
+
+\newcommand{\genalpha}{\mbox{$\hspace{0.12em}\shortmid\hspace{-0.62em}\alpha$}}
+
+\usepackage{newunicodechar}
+
+\newunicodechar{∇}{\ensuremath{\nabla}}
+\newunicodechar{λ}{\ensuremath{\lambda}}
+\newunicodechar{σ}{\ensuremath{\sigma}}
+\newunicodechar{τ}{\ensuremath{\tau}}
+\newunicodechar{∷}{\ensuremath{\!\!\!}}
+\newunicodechar{⧺}{\ensuremath{\!\!\!}}
+\newunicodechar{₁}{\ensuremath{_1}}
+\newunicodechar{₂}{\ensuremath{_2}}
+\newunicodechar{∈}{\ensuremath{\in}}
+
+\begin{document}
+
+\addtocounter{framenumber}{-1}
+
+\definecolor{mypurpleblack}{RGB}{30,0,50}
+\setbeamercolor{structure}{fg=black}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\vspace*{-7em}\includegraphics[width=\paperwidth]{forest2}\end{minipage}}
+\begin{frame}[c]
+  \centering
+  \color{white}
+
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+
+  \scriptsize
+
+  \setbeamercolor{block body}{bg=black!100}
+  \begin{minipage}{0.52\textwidth}
+    \begin{block}{}
+      \centering\normalsize\color{white}
+      \hil{\color{white}Towards topological type theory for decrypting
+      transfinite methods in classical~mathematics} \\[-0.4em]
+      \
+    \end{block}
+  \end{minipage}
+
+  \bigskip
+  \bigskip
+  \bigskip
+  \bigskip
+  \bf
+  \colorbox{black}{\begin{minipage}{0.2\textwidth}
+    \centering
+    TYPES 2025 \\
+    June 9th, 2025
+  \end{minipage}}
+  \bigskip
+
+  \colorbox{black}{\begin{minipage}{0.2\textwidth}
+    \centering
+    Ingo Blechschmidt \\
+    University of Antwerp
+  \end{minipage}}
+\end{frame}}
+
+\definecolor{mypurple}{RGB}{150,0,255}
+\setbeamercolor{structure}{fg=mypurple}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{1.7em}\includegraphics[height=\paperheight]{sea-of-clouds-1}\end{minipage}}
+\begin{frame}{A case study in Hilbert's program}
+  \textbf{Def.} Let~$(X,{\leq})$ be a quasi-order.
+  \vspace*{-0.3em}
+  \begin{itemize}
+    \item A sequence~$\alpha : \NN \to X$ is \hil{good} iff
+    there merely exist~$i < j$ with~$\alpha\,i \leq \alpha\,j$.
+    \item The quasi-order~$X$ is \hil{well\only<4->{\sinf}} iff every sequence~$\NN \to X$ is good.
+  \end{itemize}
+  \pause
+  \vspace*{-1.3em}
+
+  \begin{columns}[t]
+    \begin{column}{0.46\textwidth}
+      \begin{block}{Natural numbers\phantom{y}}
+        \textbf{Prop.} $(\NN, {\leq})$ is well\only<4->{\sinf}.
+        \smallskip
+
+        \emph{Proof.} Let~$\alpha : \NN \to \NN$. By~\badbox{\textsc{lem}}, there is a
+        \bad{minimum}~$\alpha\,i$.
+        Set~$j \defeq i + 1$. \qed
+
+        \hfill\color{gray}offensive?
+      \end{block}
+      \only<2-6>{
+        \vspace*{-2em}
+        \[ \astikznodetransparentlycircled{xm}{7}\!,
+        \quad \astikznodetransparentlycircled{x0}{4}\!,
+        \quad \astikznodecircled{t1}{mypurple}{3}\!,
+        \quad \astikznodetransparentlycircled{x1}{1}\!,
+        \quad \astikznodecircled{t2}{mypurple}{8}\!,
+        \quad \astikznodetransparentlycircled{x2}{2}\!,
+        \quad \ldots \]
+      {\centering\begin{tikzpicture}[remember picture,overlay]
+        \node[draw=mypurple, circle, thick, inner sep=0.1em] (t3) {\scriptsize$\leq$};
+        \path[draw=mypurple,thick]
+          (t1)
+          to [out=-90, in=180] (t3)
+          to [out=0, in=-90] (t2);
+      \end{tikzpicture}\par}}
+    \end{column}
+
+    \pause
+
+    \begin{column}{0.50\textwidth}
+      \begin{block}{Key stability results}
+        \justifying
+        \only<1-5>{Assuming~\badbox{\textsc{lem}}
+        and~\badbox{\textsc{dc}}}\only<6->{\good{Constructively}}, \ldots
+
+        \hil{Dickson:} \tabto{1.8cm} If~$X$ and~$Y$ are well\only<4-5>{\sinf}\only<6->{\ind}, so is~$X \times Y$. \\
+        \hil{Higman:}  \tabto{1.8cm} If~$X$ is well\only<4-5>{\sinf}\only<6->{\ind}, so is~$\mathrm{List}\,X$. \\
+        \hil{Kruskal:} \tabto{1.8cm} If~$X$ is well\only<4-5>{\sinf}\only<6->{\ind}, so is~$\mathrm{Tree}\,X$.
+      \end{block}
+    \end{column}
+  \end{columns}
+  \pause
+  \pause
+  \pause
+  \pause
+
+  \textbf{Def.} A quasi-order~$X$ is \hil{well\ind} iff~$G\,[\,]$, where $G$ is
+  the following inductively defined predicate on \hil{finite lists}.
+  (In presence of \bad{bar induction}, $\text{well}\ind \Leftrightarrow
+  \text{well}\sinf$.)
+  \[
+    \small
+    \infer{\mathsf{now}\,p : G\,\sigma}{p : \mathsf{Good}\,\sigma}
+    \qquad
+    \infer{\mathsf{later}\,f : G\,\sigma}{f : (x : X) \to G\,(\sigma ::^r x)}
+  \]
+  \vspace*{-1.3em}
+
+  \only<8>{
+    \begin{columns}[c]
+      \begin{column}{0.01\textwidth}
+	\includegraphics[height=2.4em]{question-mark}
+      \end{column}
+      \begin{column}{0.9\textwidth}
+      Is there a procedure for reinterpreting \hil{classical proofs}
+      regarding well\sinf{} as \newline \hil{blueprints for constructive proofs} regarding well\ind?
+      \end{column}
+    \end{columns}
+  }
+
+  \only<9>{
+    \vspace*{-0.0em}
+    \hil{Central insight:} A quasi-order $X$ is well\ind iff $\hil{$\necessary$}\,\forall\alpha : \NN \to
+    X\_ \exists i < j\_ \alpha\,i \leq \alpha\,j$.
+  }
+\end{frame}}
+
+\begin{frame}{Missing functions in the type of all functions?}
+  \begin{columns}[t]
+    \begin{column}{0.57\textwidth}
+      \begin{block}{Behold: A transfinite tool \ldots\phantom{g}}
+        \justifying
+        \textbf{Lemma.} \badbox{\textsc{lem}} Let~$X$ be well\sinf.
+        Let~$\alpha : \NN \to X$. Then there is an increasing
+        subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
+      \end{block}
+
+      \small
+      \emph{Proof.}
+      \begin{changemargin}{-0.8em}{0em}\begin{enumerate}\justifying
+      \item The type
+      $I \defeq \sum_{i \? \NN} \neg\sum_{j \? \NN} i < j \times \alpha\,i \leq
+      \alpha\,j$
+      is not in bijection with~$\NN$, as else the~$I$-extracted subsequence
+      of~$\alpha$ would not be good.\smallskip
+      \item By~\badbox{\textsc{lem}}, the type~$I$ is finite.\smallskip
+      \item Any index~$i_0$ larger than all the indices in~$I$ is a suitable
+      starting point for an increasing subsequence. \qed
+      %  \item Let $\mathsf{Sad}\,i \defeq
+      %  \neg \mathsf{Happy}\,i$ be the proposition
+      %  expressing that~$i$ does not appear as the first component
+      %  of a good pair, where
+      %  $\mathsf{Happy}\,i \defeq \sum_{j \? \NN} i < j \times \alpha\,i \leq
+      %  \alpha\,j$.
+      %  \item If~$\mathsf{Sad}$ is anonymously dense in~$\NN$, i.e. $\forall a
+      %  \? \NN\_ \neg\neg\exists b\?\NN\_ b \geq a \times \mathsf{Sad}\,b$, then
+      %  \badbox{\textsc{lem}} and \badbox{\textsc{dc}} provide us with a
+      %  subsequence~$\alpha\,k_0 \leq \alpha\,k_1 \leq \ldots$
+      %  with~$\mathsf{Sad}\,k_0, \mathsf{Sad}\,k_1, \ldots$. This is a contradiction to wellness.
+      %  \item
+      %  Hence there is~\bad{not~not} an index~$a \? \NN$ such that~$\mathsf{Sad}$ is
+      %  strictly bounded by~$a$; hence every index~$b \geq a$ does~\bad{not not} have
+      %  a partner: $\forall b\geq a\_ \neg\neg\exists c>b\_
+      %  \alpha\,b \leq \alpha\,c$. Thus~\bad{\textsc{lem}} and \bad{\textsc{dc}}
+      %  provide us with the desired subsequence. \qed
+      \end{enumerate}\end{changemargin}
+    \end{column}
+
+    \begin{column}{0.43\textwidth}
+      \begin{block}{\ldots{} implying a concrete consequence}
+        \textbf{Cor.} \badbox{\textsc{lem}}
+        Let~$X$ and~$Y$ be well\sinf. Then~$X \times Y$ is
+        well\sinf.\phantom{g}
+      \end{block}
+
+      \small
+      \emph{Proof.}
+      \begin{changemargin}{-0.8em}{0em}\begin{enumerate}\justifying
+      \item Let a sequence~$\gamma : \NN \to X \times Y$ be
+      given. Write~$\gamma\,k = (\alpha\,k,\beta\,k)$.\smallskip
+      \item By the lemma, there is an
+      increasing subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.\smallskip
+      \item Because~$Y$
+      is well, there are indices~$n < m$ such that~$\beta\,i_n \leq
+      \beta\,i_m$.\smallskip
+      \item As
+      also~$\alpha\,i_n \leq \alpha\,i_m$, the sequence~$\gamma$ is good. \qed
+      \end{enumerate}\end{changemargin}
+    \end{column}
+  \end{columns}
+
+  \dnote{\emph{We cannot trust~\bad{\textsc{lem}}-provided sequences to be
+  available in the type~$\NN \to X$. Similarly with~\bad{\textsc{dc}}.}}
+\end{frame}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{0.29cm}\hspace*{-1cm}\includegraphics[height=\paperheight]{tower-cropped}\end{minipage}}
+\begin{frame}{Where do many cherished inductive definitions come from?}
+  \begin{changemargin}{8.0em}{-0.5em}
+    In mathematics, we routinely enlarge structures:
+    \begin{itemize}
+      \small
+      \item Pass from~$\mathbb{Q}$ to~$\mathbb{R}$, to embrace irrationals.
+      \item Pass from~$\mathbb{R}$ to~$\mathbb{C}$, to obtain $\sqrt{-1}$.
+      \item Pass from~$\mathbb{C}$ to~$\mathbb{C}[X]$, to obtain a ``generic number''.
+    \end{itemize}
+
+    In set and type theory, we can also enlarge \hil{the mathematical universe}:
+    \begin{itemize}
+      \small
+      \item Force a \hil{generic sequence} $\NN \to X$.
+      \item Force a \hil{generic enumeration} $\NN \twoheadrightarrow X$
+            (even if $X$ is uncountable).
+      \item Force a \hil{generic prime ideal} of a given ring.
+    \end{itemize}
+    \pause
+
+    Central observations of the multiversal yoga:
+    \begin{itemize}
+      \small
+      \item A quasi-order~$X$ is well\ind iff the generic sequence~$\NN \to X$
+      is good. % iff \hil{everywhere,} every sequence~$\NN \to X$ is good.
+      \item A relation is well-founded iff for the generic descending chain, $\bot$.
+      \item A ring element is nilpotent iff it is contained in the generic
+      prime ideal.
+    \end{itemize}
+    \pause
+
+    The mystery of nongeometric sequents:
+    \begin{itemize}
+    \small
+      \item The generic ring is a field.
+      \item \ \\[-1.2em]\mbox{For the generic surjection $f : \NN \twoheadrightarrow \RR$,
+      $\neg\neg \exists n \? \NN\_ f(n) = \pi \wedge f(n+1) = e$.}
+    \end{itemize}
+  \end{changemargin}
+\end{frame}}
+
+\begin{frame}{A modal language for harnessing the multiverse}
+  \textbf{Def.} A statement~$\varphi$ holds \ldots
+  \vspace*{-0.3em}
+  \begin{itemize}
+    \small
+    \item \hil{everywhere} \tabto{2.15cm}($\necessary\varphi$)\tabto{3.08cm}
+    iff it holds \hil{in every topos}
+    (over the current base).
+    \item \hil{somewhere} \tabto{2.15cm}($\possible\varphi$)\tabto{3.08cm} iff
+    it holds \hil{in some positive topos}.
+    \item \hil{proximally} \tabto{2.15cm}($\xpossible\varphi$)\tabto{3.08cm}
+    iff it holds \hil{in some positive overt topos}.
+  \end{itemize}
+
+  We then have:
+  \begin{enumerate}
+    \item $\mathsf{Well}\ind(X,{\le}) \Longleftrightarrow
+    \necessary\mathsf{Well}\sinf(X,{\le})$.
+    \item $(\possible\varphi) \Longrightarrow \varphi$,
+    if $\varphi$ is a geometric implication (``$\forall\ldots\forall (\%
+    \Rightarrow \%)$'', with no~$\forall$ nor~$\Rightarrow$ in~$\%$).
+    \item $(\xpossible\varphi) \Longrightarrow \varphi$,
+    if~$\varphi$ is a bounded first-order statement.
+    \item For every inhabited type~$X$, $\xpossible\exists f : \NN
+    \to X\_ \mathsf{Surjective}(f)$.
+    \item $\possible \badbox{\textsc{lem}}$.
+    (Baby Barr / Friedman's trick / nontrivial exit continuation)
+  \end{enumerate}
+\end{frame}
+
+\begin{frame}{Increasing subsequences as convenient fictions}
+  \justifying
+  Let~$X$ be a quasi-order.
+  Let~$B : \mathsf{List}\,X \to \mathsf{Prop}$ be a monotone predicate.
+
+  \textbf{Thm.} \badbox{\textsc{lem}} If~$X$ is well\sinf and
+  if every increasing sequence~$\alpha : \NN \to X$ has a prefix validating~$B$,
+  then every sequence has a prefix validating~$B$.
+
+  \emph{Proof.} Let~$\alpha : \NN \to X$ be a sequence. By the lemma,
+  there is an increasing subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
+  By assumption, this subsequence has a prefix validating~$B$. This prefix is part
+  of a prefix of the original sequence~$\alpha$. Hence we conclude by monotonicity. \qed
+
+  \textbf{Thm.} If~$X$ is well\ind and if~$\nabla^\nnearrow\,B\,[\,]$,
+  then~$\nabla\,B\,[\,]$.
+
+  \emph{Proof.} Let~$\alpha : \NN \to X$ be a sequence (in an arbitrary topos).
+  \emph{Somewhere,} \textsc{lem} holds. \emph{There}~$X$
+  is still well\sinf, so that we have an increasing subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
+  By assumption, this subsequence has a finite prefix validating~$B$. This
+  prefix is part of a prefix of the original sequence~$\alpha$. Hence~$\alpha$
+  has a prefix validating~$B$ by monotonicity. So
+  \emph{somewhere} there is a finite prefix validating~$B$. Thus there
+  actually is a finite prefix validating~$B$. \qed
+\end{frame}
+
+\end{document}
+
 - General structure like in HoTTest / ABMV?
@@
 ---
+
+0. A play in classical logic
 
@@
      reasoning
+
+5. Let's play Agda, CIRM, AI
 

### 2025-06-08 slides-types2025.lagda.tex
Slides almost finished

--- 2025-02-26 slides-paris2025.tex
@@
 \hypersetup{colorlinks=true}
+\usepackage[dvipsnames]{xcolor}
+\usepackage{fontawesome5}
 
@@
 \newcommand{\impl}{\ensuremath{_\text{impl}}\xspace}
+\newcommand{\formalized}{{\color{NavyBlue!75!White}{\raisebox{-0.5pt}{\scalebox{0.8}{\faCog}}}}}
 
@@
 \newunicodechar{τ}{\ensuremath{\tau}}
-\newunicodechar{∷}{\ensuremath{\!\!\!}}
+\newunicodechar{∷}{\ensuremath{::}}
 \newunicodechar{⧺}{\ensuremath{\!\!\!}}
@@
 \newunicodechar{∈}{\ensuremath{\in}}
+\newunicodechar{ℕ}{\ensuremath{\NN}}
+\newunicodechar{↠}{\ensuremath{\twoheadrightarrow}}
+\newunicodechar{ʳ}{\ensuremath{^r}}
 
@@
       \centering\normalsize\color{white}
-      \hil{\color{white}Computing an integer using the \\ modal toposophic multiverse} \\[-0.4em]
+      \hil{\color{white}Towards topological type theory for decrypting
+      transfinite methods in classical~mathematics} \\[-0.4em]
       \
@@
     \centering
-    Logique à Paris \\
-    February 26th, 2025
+    TYPES 2025 \\
+    June 9th, 2025
   \end{minipage}}
@@
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{2.1em}\includegraphics[height=\paperheight]{sea-of-clouds-1}\end{minipage}}
-\begin{frame}{Well quasi-orders}
-  \jnote{2-3}{
-    The presented proof rests on the law of excluded middle and hence cannot
-    immediately be interpreted as a program for finding suitable indices~$i <
-    j$. However, constructive proofs are also possible (for instance by
-    induction on the value of a given term of the sequence, see
-    \href{https://www.math.lmu.de/~petrakis/Dickson.pdf}{Constructive
-    combinatorics of Dickson's Lemma} by Iosif Petrakis for several fine
-    quantitative results). And even more:
-    There is a procedure for regarding this proof---and many others in the
-    theory of well quasi-orders---as \emph{blueprints} for more informative
-    constructive proofs. This shall be our motto for today:
-
-    \emph{Do not take classical proofs literally, instead ask which
-    constructive proofs they are blueprints for.}
-  }
-
-  \jnote{3}{
-    The displayed stability results, along with several others, provide a
-    flexible toolbox for constructing new well quasi-orders from given ones.
-    However, with the classical formulation of \emph{well},
-    renamed~``well\sinf'' on the next slide, these results are inherently
-    classical.
-
-    In Higman's lemma, the set~$X^*$ of finite lists of elements of~$X$ is
-    equipped with the following ordering: We have~$x_0 \ldots x_{n-1} \leq y_0
-    \ldots y_{m-1}$ iff there is an increasing injection~$f : \{ 0,\ldots,n-1 \} \to \{
-    0,\ldots,m-1 \}$ such that~$x_i \leq y_{f(i)}$ for all~$i < n$.
-  }
-
-  \jnote{4}{
-    The dependence of the theory on well quasi-orders on classical transfinite
-    methods is already present in one of basic observations of this theory:
-
-    \textbf{Lemma.} Let~$X$ be well\sinf. Let~$\alpha : \NN \to X$. Then there
-    is an infinite increasing subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq
-    \ldots$.
-
-    \emph{Proof.} Let~$K \defeq \{ n \in \NN \,|\, \neg \exists m > n\_
-    \alpha\,n \leq \alpha\,m \}$ be the set of indices of those terms which cannot
-    appear as the first component of a good pair. If~$K$ is in bijection
-    with~$\NN$, there is a subsequence~$\alpha\,k_0 \leq \alpha\,k_1 \leq
-    \ldots$ with~$k_0, k_1, \ldots \in K$. As~$X$ is well\sinf, this sequence
-    is good, a contradiction.
-
-    Hence~$K$ is not in bijection with~$\NN$. Assuming~\bad{\textsc{lem}},
-    it is hence bounded by a number~$N$, and (again
-    with~\bad{\textsc{lem}}), for every index~$a > N$ there is an index~$b >
-    a$ such that~$\alpha\,a \leq \alpha\,b$. Thus,
-    assuming~\bad{\textsc{dc}}, every number~$i_0 > N$ is a suitable
-    starting index for an infinite increasing subsequence. \qed
-
-    The appeal to dependent choice can be removed by always picking the
-    smallest possible next index in~$\NN \setminus K$, doable by yet another
-    invocation of~\bad{\textsc{lem}}, but the result remains fundamentally
-    noneffective---in the special case~$X = (\{ 0, 1 \}, {=})$, the statement
-    of the lemma implies the infinite pigeonhole principle.
-  }
-
-  \jnote{5-7}{
-    Luckily, thanks to work by Thierry Coquand, Daniel Fridlender and Monika
-    Seisenberger, a constructive substitute is available, the notion well\ind.
-    In classical mathematics (where~\textsc{lem} and~\textsc{dc} and hence bar
-    induction are available), this notion is equivalent to well\sinf.
-
-    The assertion~``$\mathsf{Good} \mid [\,]$'' is pronounced ``$\mathsf{Good}$
-    bars the empty list'', and is defined as follows: Let~$B$ be a predicate
-    on~$X^\star$. Then~$B \mid
-    \sigma$ is inductively generated by the following two clauses.
-
-    \vspace*{-0.5em}
-    \begin{enumerate}
-      \item If~$B\sigma$, then~$B \mid \sigma$.
-      \\[-1.3em]
-      \item If~$B \mid \sigma x$ for all~$x \in X$, then~$B \mid \sigma$.
-    \end{enumerate}
-    \vspace*{-0.5em}
-
-    Here~$\sigma x$ denotes the concatenation of the list~$\sigma$ with the
-    element~$x$. The accompanying induction principle is the following: Let~$Q$
-    be a predicate on~$X^*$ such that, for all~$\sigma \in X^*$,~$B\sigma \Rightarrow Q\sigma$ and
-    $(\forall x \in X\_ Q(\sigma x)) \Rightarrow Q\sigma$. Then, for
-    all~$\sigma \in X^*$: If~$B \mid \sigma$, then~$Q\sigma$.
-
-    Intuitively, the assertion~``$B \mid \sigma$'' expresses (in a positive
-    direct way) that no matter how~$\sigma$ evolves to a longer finite
-    list~$\tau$, eventually~$B\tau$ will hold.
-  }
-
-  \jnote{8}{
-    \begin{columns}[t]
-      \begin{column}{0.60\textwidth}
-        The original notion well\sinf:
-        \begin{itemize}
-          \justifying
-          \item[\cmark] short and simple
-          \\[-1.3em]
-          \item[\cmark] constructively satisfied for the main examples (but
-          only because of the theory around well\ind)
-          \\[-1.3em]
-          \item[\cmark] concise abstract proofs (albeit employing transfinite methods)
-          \\[-1.3em]
-          \item[\xmark] main results not constructively attainable
-          \\[-1.3em]
-          \item[\xmark] philosophically strenuous by the quantification over all sequences
-          \\[-1.3em]
-          \item[\xmark] not stable under ``change of base''---a forcing extension
-          of the universe may well contain more sequences than the base universe
-          \\[-1.3em]
-          \item[\xmark] negative (universal) condition
-        \end{itemize}
-      \end{column}
-
-      \begin{column}{0.45\textwidth}
-        The constructive substitute well\ind:
-        \begin{itemize}
-          \justifying
-          \item[\cmark] main results constructive
-          \\[-1.3em]
-          \item[\cmark] stable under change of base
-          \\[-1.3em]
-          \item[\cmark] positive (existential) condition
-          \\[-1.3em]
-          \item[\xmark] proofs intriguing, but also somewhat alien, not just some
-          trivial reshuffling of the classical arguments, classical sequence
-          language cannot be used
-        \end{itemize}
-      \end{column}
-    \end{columns}
-  }
-
-  \jnote{9}{
-    Constructively, the notion~well\ind is much stronger than~well\sinf, as it
-    ensures goodness (in an appropriate sense) of sequence-like entities
-    which are not actually honest maps~$\NN \to X$.
-
-    For partial maps~$\alpha$, by~$\alpha\,n\,\downarrow$ we mean that~$\alpha$
-    is defined on the input~$n$. If~\bad{\textsc{lem}} is available, then a
-    partial map such that~$\neg\neg(\alpha\,n\,\downarrow)$ for all~$n \in \NN$
-    is already a total map, but without~\bad{\textsc{lem}} the hypothesis
-    well\sinf{} does not have anything to say about such a partially-defined
-    sequence.
-
-    If~\bad{\textsc{dc}} is available, then every multivalued map contains a
-    singlevalued map, but again without~\bad{\textsc{dc}} the hypothesis well\sinf{}
-    does not have anything to say about multivalued sequences.
-  }
-
-  \jnote{10}{
-    It turns out that these entities are, or give rise to, actual maps~$\NN \to
-    X$---but in a forcing extension of the universe.
-
-    Forcing originated in set theory to construct new models for set theory
-    from given ones, in order to explore the range of set-theoretic
-    possibility. For instance, by forcing we can construct models of~\text{zfc}
-    validating the continuum hypothesis and also models which falsify it.
-
-    We here refer to a simplification of original forcing which is useful in a
-    constructive metatheory. At its core, every forcing extension is just a
-    formula and proof translation of a certain form. For instance, there is a
-    forcing extension validating~\bad{\textsc{lem}} even if the base universe
-    does not; this forcing extension is not a deep mystery, for a statement holds
-    in that forcing extension iff its double negation translation holds in the
-    base universe and it is well-known that the double negation translation
-    of~\bad{\textsc{lem}} is an intuitionistic tautology.
-
-    \href{https://www.speicherleck.de/iblech/stuff/slides-herrsching2023.pdf}{Here
-    is a set of slides on constructive forcing}, and Section~4 of
-    \href{https://raw.githubusercontent.com/iblech/constructive-maximal-ideals/master/tex/extended.pdf}{this
-    joint paper with Peter Schuster} contains a written summary of constructive
-    forcing.
-  }
-
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{1.7em}\includegraphics[height=\paperheight]{sea-of-clouds-1}\end{minipage}}
+\begin{frame}{A case study in Hilbert's program}
   \textbf{Def.} Let~$(X,{\leq})$ be a quasi-order.
@@
     \item A sequence~$\alpha : \NN \to X$ is \hil{good} iff
-    there exist~$i < j$ with~$\alpha\,i \leq \alpha\,j$.
+    there merely exist~$i < j$ with~$\alpha\,i \leq \alpha\,j$.
     \item The quasi-order~$X$ is \hil{well\only<4->{\sinf}} iff every sequence~$\NN \to X$ is good.
@@
   \pause
-  \vspace*{-0.9em}
+  \vspace*{-1.3em}
 
@@
       \begin{block}{Natural numbers\phantom{y}}
-        \textbf{Prop.} $(\NN, {\leq})$ is well\only<4->{\sinf}.
+        \textbf{Prop.} $(\NN, {\leq})$ is well\only<4->{\sinf}. \formalized
         \smallskip
@@
       \end{block}
-      \only<2-4>{
+      \only<2-6>{
         \vspace*{-2em}
@@
 
-        \hil{Dickson:} \tabto{1.8cm} If~$X$ and~$Y$ are well\only<4-5>{\sinf}\only<6->{\ind}, so is~$X \times Y$. \\
-        \hil{Higman:}  \tabto{1.8cm} If~$X$ is well\only<4-5>{\sinf}\only<6->{\ind}, so is~$X^\star$. \\
-        \hil{Kruskal:} \tabto{1.8cm} If~$X$ is well\only<4-5>{\sinf}\only<6->{\ind}, so is~$\mathrm{Tree}(X)$.
+        \hil{Dickson:} \tabto{1.7cm} If~$X$ and~$Y$ are well\only<4-5>{\sinf}\only<6->{\ind}, so is~$X \times Y$. \\
+        \hil{Higman:}  \tabto{1.7cm} If~$X$ is well\only<4-5>{\sinf}\only<6->{\ind}, so is~$\mathrm{List}\,X$. \\
+        \hil{Kruskal:} \tabto{1.7cm} If~$X$ is well\only<4-5>{\sinf}\only<6->{\ind}, so is~$\mathrm{Tree}\,X$.
       \end{block}
@@
   \pause
-
-  \textbf{Def.} A quasi-order~$X$ is \hil{well\ind} iff there exists a
-  \hil{modulus of wellness} for~$X$.
-  \pause
-  \pause
-
-  \small
-  With \bad{bar induction},\tabto{3.10cm} $\text{well}\ind \Leftarrow \text{well}\sinf$. \\[-0.2em]
-  Constructively,\tabto{3.10cm} $\text{well}\ind \Rightarrow \text{well}\sinf$.
+  \pause
+  \pause
+
+  \textbf{Def.} A quasi-order~$X$ is \hil{well\ind} iff~$G\,[\,]$, where $G$ is
+  the following inductively defined predicate on \hil{finite lists}.~\formalized{}
+  (In presence of \bad{bar induction}, $\text{well}\ind \Leftrightarrow
+  \text{well}\sinf$.)
+  \[
+    \small
+    \infer{\mathsf{now}\,p : G\,\sigma}{p : \mathsf{Good}\,\sigma}
+    \qquad
+    \infer{\mathsf{later}\,f : G\,\sigma}{f : (x : X) \to G\,(\sigma ::^r x)}
+  \]
+  \vspace*{-1.3em}
+
   \only<8>{
-    \bigskip
     \begin{columns}[c]
@@
   }
-  \pause
-  \pause
-  Moreover, if~$X$ is well\ind, then \ldots
-  \begin{itemize}
-  \vspace*{-0.7em}
-    \small
-    \item for every \emph{partial} function~$\alpha$,\tabto{5.25cm}
-    if~$\forall n\_ \neg\neg(\alpha\,n\,{\downarrow})$,
-    then~$\neg\neg \exists i < j\_ \alpha\,i\,{\downarrow} \wedge
-    \alpha\,j\,{\downarrow} \wedge \alpha\,i \leq \alpha\,j$. \\[-2.0em]\
-    \item for every \emph{multivalued} function~$\alpha$,\tabto{5.25cm}
-    $\exists i < j\_ \exists x \in \alpha\,i\_
-    \exists y \in \alpha\,j\_ x \leq y$.
-  \end{itemize}
-  \pause
-
-  \vspace*{-0.7em}
+
+  \only<9>{
+    \vspace*{-0.0em}
     \hil{Central insight:} A quasi-order $X$ is well\ind iff $\hil{$\necessary$}\,\forall\alpha : \NN \to
     X\_ \exists i < j\_ \alpha\,i \leq \alpha\,j$.
+  }
 \end{frame}}
 
+\begin{frame}{Missing functions in the type of all functions?}
+  \begin{columns}[t]
+    \begin{column}{0.57\textwidth}
+      \begin{block}{Behold: A transfinite tool \ldots\phantom{g}}
+        \justifying
+        \textbf{Lemma.} \badbox{\textsc{lem}} Let~$X$ be well\sinf.
+        Let~$\alpha : \NN \to X$. Then there is an increasing
+        subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
+      \end{block}
+
+      \small
+      \emph{Proof.}
+      \begin{changemargin}{-0.8em}{0em}\begin{enumerate}\justifying
+      \item The type
+      $I \defeq \sum_{i \? \NN} \neg\sum_{j \? \NN} i < j \times \alpha\,i \leq
+      \alpha\,j$
+      is not in bijection with~$\NN$, as else the~$I$-extracted subsequence
+      of~$\alpha$ would not be good.\smallskip
+      \item By~\badbox{\textsc{lem}}, the type~$I$ is finite.\smallskip
+      \item Any index~$i_0$ larger than all the indices in~$I$ is a suitable
+      starting point for an increasing subsequence. \qed
+      %  \item Let $\mathsf{Sad}\,i \defeq
+      %  \neg \mathsf{Happy}\,i$ be the proposition
+      %  expressing that~$i$ does not appear as the first component
+      %  of a good pair, where
+      %  $\mathsf{Happy}\,i \defeq \sum_{j \? \NN} i < j \times \alpha\,i \leq
+      %  \alpha\,j$.
+      %  \item If~$\mathsf{Sad}$ is anonymously dense in~$\NN$, i.e. $\forall a
+      %  \? \NN\_ \neg\neg\exists b\?\NN\_ b \geq a \times \mathsf{Sad}\,b$, then
+      %  \badbox{\textsc{lem}} and \badbox{\textsc{dc}} provide us with a
+      %  subsequence~$\alpha\,k_0 \leq \alpha\,k_1 \leq \ldots$
+      %  with~$\mathsf{Sad}\,k_0, \mathsf{Sad}\,k_1, \ldots$. This is a contradiction to wellness.
+      %  \item
+      %  Hence there is~\bad{not~not} an index~$a \? \NN$ such that~$\mathsf{Sad}$ is
+      %  strictly bounded by~$a$; hence every index~$b \geq a$ does~\bad{not not} have
+      %  a partner: $\forall b\geq a\_ \neg\neg\exists c>b\_
+      %  \alpha\,b \leq \alpha\,c$. Thus~\bad{\textsc{lem}} and \bad{\textsc{dc}}
+      %  provide us with the desired subsequence. \qed
+      \end{enumerate}\end{changemargin}
+    \end{column}
+
+    \begin{column}{0.43\textwidth}
+      \begin{block}{\ldots{} implying a concrete consequence}
+        \textbf{Cor.} \badbox{\textsc{lem}}
+        Let~$X$ and~$Y$ be well\sinf. Then~$X \times Y$ is
+        well\sinf.\phantom{g}
+      \end{block}
+
+      \small
+      \emph{Proof.}
+      \begin{changemargin}{-0.8em}{0em}\begin{enumerate}\justifying
+      \item Let a sequence~$\gamma : \NN \to X \times Y$ be
+      given. Write~$\gamma\,k = (\alpha\,k,\beta\,k)$.\smallskip
+      \item By the lemma, there is an
+      increasing subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.\smallskip
+      \item Because~$Y$
+      is well, there are indices~$n < m$ such that~$\beta\,i_n \leq
+      \beta\,i_m$.\smallskip
+      \item As
+      also~$\alpha\,i_n \leq \alpha\,i_m$, the sequence~$\gamma$ is good. \qed
+      \end{enumerate}\end{changemargin}
+    \end{column}
+  \end{columns}
+
+  \dnote{\emph{We cannot trust~\bad{\textsc{lem}}-provided sequences to be
+  available in the type~$\NN \to X$. Similarly with~\bad{\textsc{dc}}.}}
+\end{frame}
+
 {\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{0.29cm}\hspace*{-1cm}\includegraphics[height=\paperheight]{tower-cropped}\end{minipage}}
-\begin{frame}{The modal multiverse of constructive forcing}
-  \jnote{1}{
-    By \emph{topos}, we mean \emph{Grothendieck topos}. In constructive
-    forcing, a ``forcing extension of the base universe'' is exactly the same
-    thing as a Grothendieck topos.
-
-    A particular member of the rich and varied landscape of toposes is the
-    \emph{trivial topos}, in which every statement whatsoever holds. By
-    restricting to positive toposes, we exclude this special case.
-
-    For positive toposes~$\E$, a geometric implication holds in~$\E$ iff it
-    holds in the base universe. For positive overt toposes~$\E$, we even have
-    that a bounded first-order formula holds in~$\E$ iff it holds in the base.
-    Hence, for the purpose of verifying a bounded first-order assertion about
-    the base, we can freely pass to a positiv overt topos with problem-adapted
-    better higher-order properties (such as that some uncountable set from the
-    base now appears countable, or that an infinite sequence whose existence is
-    predicted by failing dependent choice now actually exists).
-
-    \href{https://www.speicherleck.de/iblech/stuff/early-draft-modal-multiverse.pdf}{Here
-    is a rough early draft of a preprint with more details about the modal
-    multiverse.}
-  }
-
-  \jnote{2}{
-    The idea to study the modal multiverse of toposes in a principled manner
-    was proposed by Alexander Oldenziel in 2016. \emph{Foreshadowed by:}
+\begin{frame}{Where do many cherished inductive definitions come from?}
+  \begin{changemargin}{8.0em}{-0.5em}
+    In mathematics, we routinely enlarge structures:
     \begin{itemize}
-      \item[1984\phantom{s}] André Joyal, Miles Tierney. ``An extension of the Galois theory of Grothendieck''.
-      \\[-1em]
-      \item[1987\phantom{s}] Andreas Blass. ``Well-ordering and induction in intuitionistic logic and topoi''.
-      \\[-1em]
-      \item[2010s] Milly Maietti, Steve Vickers. Ongoing work on arithmetic universes.
-      \\[-1em]
-      \item[2011\phantom{s}] Joel David Hamkins. ``The set-theoretic multiverse''.
-      \\[-1em]
-      \item[2013\phantom{s}] Shawn Henry. ``Classifying topoi and preservation of higher order logic by geometric morphisms''.
+      \small
+      \item Pass from~$\mathbb{Q}$ to~$\mathbb{R}$, to embrace irrationals.
+      \item Pass from~$\mathbb{R}$ to~$\mathbb{C}$, to obtain $\sqrt{-1}$.
+      \item Pass from~$\mathbb{C}$ to~$\mathbb{C}[X]$, to obtain a ``generic number''.
     \end{itemize}
-  }
-
-  \jnote{3}{
-    With the modal language we seek to provide an accessible and modular
-    framework for constructivization results.
-
-    For instance, conservativity of classical logic over intuitionistic logic
-    for geometric implications (known under various names such as Barr's
-    theorem, Friedman's trick, escaping the continuation monad, \ldots) is
-    packaged up by the observation that \emph{somewhere}, the law of excluded
-    middle holds.
-
-    Another example:
-    In the community around Krull's lemma, it is well-known that
-    we can constructively infer that a given ring element~$x \in A$ is nilpotent from
-    knowing that it is contained in the \emph{generic prime ideal} of~$A$. This entity
-    is not actually an honest prime ideal of the ring~$A$ in the base
-    universe, but a certain combinatorial notion (efficiently dealt with using
-    \emph{entailment relations}). Constructive forcing allows us to reify the
-    generic prime ideal as an actual prime ideal in a suitable forcing extension.
-  }
-
-  \begin{changemargin}{7.5em}{-0.5em}
+
+    In set and type theory, we can also enlarge \hil{the mathematical universe}:
+    \begin{itemize}
+      \small
+      \item Force a \hil{generic sequence} $\NN \to X$. \formalized
+      \item Force a \hil{generic enumeration} $\NN \twoheadrightarrow X$
+            (even if $X$ is uncountable). \formalized
+      \item Force a \hil{generic prime ideal} of a given ring. \formalized
+    \end{itemize}
+    \pause
+
+    Central observations of the multiversal yoga:
+    \begin{itemize}
+      \small
+      \item A quasi-order~$X$ is well\ind iff the generic sequence~$\NN \to X$
+      is good. \formalized
+      \item A relation is well-founded iff for the generic descending chain,
+      $\bot$. \formalized
+      \item A ring element is nilpotent iff it is contained in the generic
+      prime ideal. \formalized
+    \end{itemize}
+    \pause
+
+    The mystery of nongeometric sequents:
+    \begin{itemize}
+    \small
+      \item The generic ring is a field.
+      \item \ \\[-1.2em]\mbox{For the generic surjection $f : \NN \twoheadrightarrow \RR$,
+      $\neg\neg \exists n \? \NN\_ f(n) = \pi \wedge f(n+1) = e$. \formalized}
+    \end{itemize}
+  \end{changemargin}
+\end{frame}}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
+\begin{frame}{A modal language for harnessing the multiverse}
   \textbf{Def.} A statement~$\varphi$ holds \ldots
@@
     \small
-      \item \hil{everywhere} \tabto{2.15cm}($\necessary\varphi$)\tabto{3.08cm}
+    \item \hil{everywhere} \tabto{2.05cm}($\necessary\varphi$)\tabto{2.98cm}
     iff it holds \hil{in every topos}
     (over the current base).
-      \item \hil{somewhere} \tabto{2.15cm}($\possible\varphi$)\tabto{3.08cm} iff
+    \item \hil{somewhere} \tabto{2.05cm}($\possible\varphi$)\tabto{2.98cm} iff
     it holds \hil{in some positive topos}.
-      \item \hil{proximally} \tabto{2.15cm}($\xpossible\varphi$)\tabto{3.08cm}
+    \item \hil{proximally} \tabto{2.05cm}($\xpossible\varphi$)\tabto{2.98cm}
     iff it holds \hil{in some positive overt topos}.
@@
 
-    \textbf{Def.} A (Grothendieck) \hil{topos} is a category equivalent to the
-    category of sheaves over a small site.
-    \pause
-
-    \only<2>{
-      \textbf{Examples for toposes.}
+  We then have:
+  \begin{enumerate}
+    \item $\mathsf{Well}\ind(X,{\le}) \Longleftrightarrow
+    \necessary\mathsf{Well}\sinf(X,{\le})$. \formalized
+    \item $(\possible\varphi) \Longrightarrow \varphi$,
+    if $\varphi$ is a geometric implication (``$\forall\ldots\forall (\%
+    \Rightarrow \%)$'', with no~$\forall$ nor~$\Rightarrow$
+    in~$\%$).~\formalized
+    \item $(\xpossible\varphi) \Longrightarrow \varphi$,
+    if~$\varphi$ is a bounded first-order statement. \formalized
+    \item For every inhabited type~$X$, $\xpossible\exists f : \NN
+    \to X\_ \mathsf{Surjective}(f)$. \formalized
+    \item $\possible \badbox{\textsc{lem}}$.
+    (Baby Barr / Friedman's trick / nontrivial exit continuation)
+  \end{enumerate}
+\end{frame}}
+
+\begin{frame}{The plain combinatorics of toposes}
+  \begin{enumerate}\justifying
+    \item Realize a generic gadget as some kind of limit of \hil{approximations}
+    from the base universe. For the generic function $\NN \to X$:
+
+    \AgdaHide{
+      \begin{code}
+open import Data.List
+open import Agda.Primitive using (Set)
+
+Prop = Set
+postulate
+  X : Set
+  _∈_ : X → List X → Prop
+      \end{code}
+    }
+
+    {\small\begin{code}
+L : Set
+L = List X
+    \end{code}}
+
+    \pause
+    \item \hil{Reinterpret}, in a mechanical fashion, assertions about the generic
+    gadget as assertions about its approximations~(\formalized). Be prepared to evolve
+    approximations.
+
+    {\small\begin{code}
+data ∇ (P : L → Prop) : L → Prop where
+  now   : {σ : L} → P σ → ∇ P σ
+  later : {σ : L} → ((x : X) → ∇ P (σ ∷ʳ x)) → ∇ P σ
+    \end{code}
+
     \begin{itemize}
-        \item $\mathrm{Set}$, the category of sets and maps.
-        \item The category of sets and maps which are \hil{defined up to~$\boldsymbol{\neg\neg}$}.
-        \item $\mathrm{Set}[G]$, the extension obtained by adding a \hil{generic
-        filter} of a \hil{forcing notion} (a quasi-order equipped with a
-        coverage).
-      \end{itemize}
-      The following are \bad{not} toposes:
-      \begin{itemize}
-        \item The category of sets and partially defined maps.
-        \item The category of abelian groups.
-      \end{itemize}
-    }
-
-    \visible<3->{
-      \hil{Multiversal yoga:}
-      \begin{enumerate}
-        \item A quasiorder is well\ind iff \emph{everywhere}, every sequence is good.
-        \smallskip
-        \item A ring element is nilpotent iff \\ all prime ideals \emph{everywhere} contain it.
-        \smallskip
-        \item For every inhabited set~$X$, \\ \emph{proximally} there exists an
-        enumeration~$\NN \twoheadrightarrow X$.
-        \smallskip
-
-        \item For every ring, \emph{proximally} there exists a maximal ideal.
-        \smallskip
-
-        \item \emph{Somewhere}, the law of excluded middle holds.
+    \item
+    For a \hil{stage-dependent proposition}~$P : L \to \mathsf{Prop}$, $\nabla\,P\,\sigma$ expresses
+    that no matter how~$\sigma$ will evolve over a time to a better
+    approximation~$\tau$, eventually~$P\,\tau$ will hold.
+  % later₂ : (a : X) → ((τ : X*) → a ∈ σ ⧺ τ → ∇ P (σ ⧺ τ)) → ∇ P σ
+    \item We have the stage-dependent proposition ``$f_0\,n = x$'',
+    where~$f_0$ is the generic function.
+    \item That~$f_0$ is defined on input~$n$ can be expressed
+    as~$\nabla\,P\,[\,]$ where~$P\,\sigma \defeqv (\mathsf{length}\,\sigma > n)$.\end{itemize}}
+    \medskip
+    \pause
+
+    \item Crucially, this interpretation is \hil{sound} with respect to constructive
+    reasoning.
   \end{enumerate}
-    }
-  \end{changemargin}
-\end{frame}}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{-4.68em}\includegraphics[width=1.2\paperwidth]{wqo-faded}\end{minipage}}
-\begin{frame}{Multiversal constructive combinatorics}
+\end{frame}
+
+\begin{frame}{Increasing subsequences as convenient fictions}
   \justifying
-  \textbf{Prop.} Let~$X$ and~$Y$ be well\ind quasi-orders. Then~$X \times Y$
-  is well$\ind$.
-
-  \emph{Multiversal constructive proof.} Let~$\alpha = (\beta,\gamma) : \NN \to X \times Y$ be a sequence in an
-  arbitrary topos. We need to show that~$\alpha$ is good, i.\,e.\@ find indices~$n < m$ such that
-  \[ \beta\,n \leq \beta\,m \quad\text{and}\quad \gamma\,n \leq \gamma\,m. \]
-
-  It suffices to prove that \emph{somewhere}, $\alpha$ is good, as goodness is
-  a geometric implication (in fact even a geometric formula). Hence without loss of
-  generality, we may suppose~\textsc{lem}.
-
-  Thus there is an infinite increasing subsequence
-  \[ \beta\,k_0 \leq \beta\,k_1 \leq \ldots. \]
-  As~$Y$ is well\ind, the sequence~$(\gamma\,k_0, \gamma\,k_1, \ldots)$
-  is good, so there exist~$i < j$ with~$\gamma\,k_i \leq \gamma\,k_j$. Since we
-  also have~$\beta\,k_i \leq \beta\,k_j$, we are done. \qed
-\end{frame}}
+  Let~$X$ be a quasi-order.
+  Let~$B : \mathsf{List}\,X \to \mathsf{Prop}$ be a monotone predicate.
+
+  \begin{block}{Classical blueprint}
+    \textbf{Thm.} \badbox{\textsc{lem}} If~$X$ is well\sinf{} and
+    if every increasing sequence~$\alpha : \NN \to X$ has a prefix validating~$B$,
+    then every sequence has a prefix validating~$B$.
+  \end{block}
+  \vspace*{-0.4em}
+  {\small\emph{Proof.} Let~$\alpha : \NN \to X$ be a sequence. By the lemma,
+  there is an increasing subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
+  By assumption, this subsequence has a prefix validating~$B$. This prefix is part
+  of a prefix of the original sequence~$\alpha$. Hence we conclude by
+  monotonicity. \qed\par}
+  \medskip
+
+  \begin{block}{Constructive reimagination}
+    \textbf{Thm.} If~$X$ is well\ind and if~$\nabla^\nnearrow\,B\,[\,]$,
+    then~$\nabla\,B\,[\,]$. \formalized
+  \end{block}
+  \vspace*{-0.4em}
+  {\small\emph{Proof.} Let~$\alpha : \NN \to X$ be a sequence (in an arbitrary topos).
+  \emph{Somewhere,} \textsc{lem} holds. \emph{There}~$X$
+  is still well\sinf, so that we have an increasing subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
+  By assumption, this subsequence has a finite prefix validating~$B$. This
+  prefix is part of a prefix of the original sequence~$\alpha$. Hence~$\alpha$
+  has a prefix validating~$B$ by monotonicity \emph{there}. So
+  \emph{somewhere} there is a finite prefix validating~$B$. Thus there
+  actually is a finite prefix validating~$B$. \qed\par}
+\end{frame}
 
 {\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
-\begin{frame}{Multiversal constructive algebra}
+\begin{frame}{Maximal ideals as convenient fictions}
   \justifying
+  Let~$A$ be a commutative ring with unit.
+
   \textbf{Thm.}
-  Let~$M$ be a surjective matrix with more rows than columns over a
-  ring~$A$. Then~$1 = 0$ in~$A$.
+  Let~$M$ be a surjective matrix with more rows than columns over~$A$. Then~$1 = 0$ in~$A$.
   \medskip
@@
   \medskip
+
+  \only<1>{\medskip\par\centering\scalebox{0.9}{\centering\begin{tikzpicture}
+    \node (0) at (0,1) {$(0) = \{0\}$};
+    \node (1) at (0,5) {$(1) = \ZZ$};
+    \node (2) at (-2,4) {$(2)$};
+    \node [right of=2] (3) {$(3)$};
+    \node [below of=2] (4) {$(4)$};
+    \node [below of=2, xshift=0.7cm] (6) {$(6)$};
+    \node [right of=3] (5) {$(5)$};
+    \node [right of=5] (7) {$(7)$};
+    \node [right of=7] (7d) {$\ldots$\phantom{(}};
+    \node [right of=7d, xshift=3cm, yshift=-2cm] (max)
+    {\vbox{\small{\it maximal among the proper ideals} \\ \medskip \hspace*{-6.75em}\textbullet \quad $\neg(1 \in
+    \mmm)$ \\ \medskip \textbullet \quad $\neg\bigl(1 \in \mmm + (x)\bigr) \Rightarrow x \in \mmm$}};
+    \node [below of=4] (8) {$(8)$};
+    \node [right of=8, xshift=3cm] (8d) {$\ldots$};
+    \draw (0) -- (8);
+    \draw (0) -- (8d);
+    \draw (0) -- (6);
+    \draw (2) -- (1);
+    \draw (3) -- (1);
+    \draw (5) -- (1);
+    \draw (7) -- (1);
+    \draw (7d) -- (1);
+    \draw (4) -- (2);
+    \draw (8) -- (4);
+    \draw (6) -- (2);
+    \draw (6) -- (3);
+    \draw [mypurple!30, thick, shorten <=-2pt, shorten >=-2pt, ->] (max) to [out=120, in=-30] (7d);
+    \begin{pgfonlayer}{background}
+      \draw[decorate, very thick, draw=mypurple!30]
+        ($(2.south west) + (8pt, 0)$) arc(270:180:8pt) --
+        ($(2.north west) + (0, -8pt)$) arc(180:90:8pt) --
+        ($(7d.north east) + (-8pt, 0)$) arc(90:0:8pt) --
+        ($(7d.south east) + (0, 8pt)$) arc(0:-90:8pt) --
+        cycle;
+    \end{pgfonlayer}
+  \end{tikzpicture}\par}\par}
   \pause
@@
   We may work \emph{somewhere} where \good{\textsc{lem}} holds. So \good{assume not}.
-  \emph{Proximally,} there is a \good{maximal ideal}~$\mmm$.
+  \emph{Proximally,} there is a \good{maximal ideal}~$\mmm$~(\formalized).
   The matrix~$M$ is still surjective \emph{there}, and also over~$A/\mmm$. Since~$A/\mmm$ is a field, this
@@
   \qed
-  \pause
-
-  \centering
-  \emph{\href{https://iblech.github.io/constructive-maximal-ideals/}{Agda
-  formalization available.}}
-  \par
 \end{frame}}
 
+\begin{frame}{Formal metatheory}
+  \small
+  \good{\cmark{}} \emph{There are type-theoretic multiverses,} such as
+  \begin{itemize}
+    \item the collection of all $\mathrm{PSh}(\mathcal{C} \times \mathcal{B})$,
+    where~$\mathcal{B}$ ranges over cube categories and~$\mathcal{C}$ over
+    arbitrary small categories, and their corresponding sheaf models
+
+    {\scriptsize
+    Coquand. ``A survey of constructive presheaf models of univalence''. \emph{ACM SIGLOG News}, 5.3 (2018).
+    \par}
+  \end{itemize}
+  \medskip
+  \pause
+
+  \bad{\xmark} \emph{Accessing the multiverse from within intensional type theory is tricky:}
+  \begin{itemize}
+    \item Given a model of $\mathfrak{s}$CIC and a category~$\mathcal{C}$ in
+    it, we have a syntactic presheaf model of CIC.
+
+    {\scriptsize
+    Coquand, Jaber. ``A note on forcing and type theory''. \emph{Fundamenta Informaticae 100} (2010). \\
+    Jaber, Lewertowski, Pédrot, Sozeau, Tabareau. ``The definitional side of the forcing''. \emph{Proceedings of LICS '16} (2016). \\
+    Pédrot. ``Russian constructivism in a prefascist theory''. \emph{Proceedings of LICS '20} (2020).
+    \par}
+    \item Given a suitable lex modality, we have a syntactic sheaf model (model of modal types).
+
+    {\scriptsize
+    Coquand, Ruch, Sattler. ``Constructive sheaf models of type theory.'' \emph{Math. Struct. Comput. Sci.} 31.9 (2021). \\
+    Escardó, Xu. ``Sheaf models of type theory in type theory''. Unpublished (2016). \\
+    Quirin. ``Lawvere--Tierney sheafification in Homotopy Type Theory''. PhD thesis (2016).
+    \par}
+    \item (I believe) we have syntactic sheaf models in certain special cases,
+    when no coherence issues arise in defining the notion of presheaves.
+  \end{itemize}
+
+  Note: We can use~$\nabla$ even without a proper metatheoretic backing.
+\end{frame}
 \backupstart
+
+\begin{frame}
+  \emph{Backup slides.}
+\end{frame}
 
@@
 
-  We then define~``\hil{$P \mid \sigma$}'' (``$P$ bars~$\sigma$'') inductively by the following clauses:
+  We then define~``\hil{$\nabla\,P\,\sigma$}'' (``$P$ bars~$\sigma$'') inductively by the following clauses:
   \begin{enumerate}
-    \item If~$P\sigma$, then~$P \mid \sigma$.
-    \item If~$P \mid \tau$ for all~$\tau \in R$, where~$R$ is some covering
-    of~$\sigma$, then~$P \mid \sigma$.
+    \item If~$P\sigma$, then~$\nabla\,P\,\sigma$.
+    \item If~$\nabla\,P\,\tau$ for all~$\tau \in R$, where~$R$ is some covering
+    of~$\sigma$, then~$\nabla\,P\,\sigma$.
   \end{enumerate}
-  So~$P \mid \sigma$ expresses in a \hil{direct inductive fashion}:
+  So~$\nabla\,P\,\sigma$ expresses in a \hil{direct inductive fashion}:
   \[ \text{``No matter
-  how~$\sigma$ evolves to a better approximation~$\tau$, eventually~$P\tau$
+  how~$\sigma$ evolves to a better approximation~$\tau$, eventually~$P\,\tau$
   will hold.''} \]
@@
   We use quantifier-like notation: ``$\nabla(\tau \preccurlyeq \sigma)\_
-  P\tau$'' means ``$P \mid \sigma$''.
+  P\tau$'' means ``$\nabla\,P\,\sigma$''.
 \end{frame}}
@@
     little Zariski &
-    ``every f.g. vector space does \emph{not not} have a basis'' &
+    ``every f.g. vector space does \emph{not~not} have a basis'' &
     \makesamewidth[l]{}{\phantom{x}}\hil{Grothendieck's generic freeness lemma}
@@
 
-- Well quasi-orders
-  *abstract proofs as blueprints for concrete computations?*
-  - def. good, well
-  - ex. ℕ
-  - classical stability results
-
-- A better notion
-  - partially defined sequences
-    (not really a morphism in Set, but in Sh_¬¬(Set))
-  - multivalued sequences
-    (approximately morphisms in some category)
-  - modulus of wellness
-
-- The modal toposophic multiverse
-  - examples for toposes:
-    - Set
-    - Sh_¬¬(Set)
-    - Set[G], obtained by adding a generic filter of a forcing notion
-    - not quite toposes: arbitrary partially defined maps, abelian groups
-  - modal operators
-  - Diamond φ ⇒ φ etc.
-  - X has a modulus of wellness iff everywhere, every sequence is good
-  - Diamond (¬LEM)
-  - etc.
-
-- Applications
-  - Dickson, constructively
-  - surjective matrices (with Agda)
-
-Somewhere: eigenvalues etc.
+0. A play in classical logic
+
+1. Hilbert's program
+   - basics about wqo's
+   - subsequence lemma
+   - Dickson
+   - inductive reimagination
+   - questions: where from? why work? how much stronger?
+
+2. Where do our most cherished definitions come from?
+   - a familiar business: enlarging mathematical structures
+   - adjoining generic function, surjection, prime ideal, ...
+   - observations:
+     - well iff generic sequence is good iff everywhere, every sequence is good
+       (in particular, every multivalued and every up-to-¬¬-defined sequence is good)
+     - wellfounded iff for the generic decreasing sequence, ⊥ iff ...
+   - fun facts about generic gadgets
+
+3. Modal language for harnessing the multiverse
+   - definition
+   - examples
+   - implicational wellness
+   - dialogue
+
+4. Implementation
+   - idea: realize generic gadget as some kind of limit of approximations from
+     the base
+   - reinterpret, in a mechanic fashion, assertions about generic objects as
+     assertions about approximations
+     - ex.: "the generic function is defined on input n" as "no matter how the
+       empty list evolves over a time to a better approximation xs, eventually
+       xs will have length at least succ n"
+   - crucially, this interpretation is sound with respect to intuitionistic
+     reasoning
+
+5. Let's play Agda, CIRM, AI
+
+---
+
+Towards topological type theory for decrypting transfinite methods in classical mathematics
+
+In constructive mathematics, an ongoing challenge is to extracting
+computational content from infinitary arguments in classical
+mathematics—especially those that yield concrete results via abstract tools
+such as minimal bad sequences or maximal ideals. Today, a wide range of such
+techniques exists, constituting a late partial fulfillment of Hilbert's
+program.
+
+This talk presents work in progress on porting the modal operators "everywhere"
+and "somewhere", as developed in set theory by Joel David Hamkins, Victoria
+Gitman and their collaborators, into type theory. This modal enrichment
+provides a uniform language for expressing several established
+constructivization techniques, and is at the core of a relatively recent new
+technique that offers insights on the origins of some of our most cherished
+inductive definitions. These developments are closely connected to recent
+advances in type-theoretic presheaf and sheaf models, dialogue
+continuity, synthetic algebraic geometry and well-quasi-order theory.
+
+---
+
+Which functions are we excluding when we form the type of all functions between
+two types? Does every field have an algebraic closure?
+
+Inspired by the modal approach to the set-theoretic multiverse of Joel David
+Hamkins, Victoria Gitman and their collaborators, we aim to introduce the modal
+operators "everywhere" and "somewhere" to homotopy type theory, proposing a
+multiversal perspective on these motivating questions.
+
+---
+
+Well quasi-orders are a combinatorial notion with featureful applications in
+graph theory, termination checking, commutative algebra and other subjects.
+However, their classical definition, though concise and elegant, poses
+significant challenges in constructive mathematics and frameworks that lack
+function sets. In response, several constructive substitutes have been
+developed, most recently an implicational definition by Stefano Berardi,
+Gabriele Buriola and Peter Schuster. In this talk, we investigate how a modal
+approach can reinterpret classical proofs involving transfinite methods as
+blueprints for constructive proofs grounded in these alternatives, thereby
+combining the best of both worlds: Short and abstract proofs, but with
+constructive content. We also indicate how the modal approach allows us to
+compare the implicational definition with an earlier inductive one.
+
+---
+
+Combinatorics and commutative algebra abound with situations where we prove
+quite concrete results by quite abstract transfinite methods, such as minimal
+bad sequences or maximal ideals. Amazingly, such infinitary arguments can often
+be understood as blueprints for quite explicit computations—as called for by
+Hilbert’s programme. In this talk, we will travel the modal toposophic
+multiverse to facilitate this kind of mining abstract proofs for refined
+quantitative results. We will use a celebrated theorem from order theory about
+sequences N→N that everybody can relate with as a running example.
+
+---
+
+Where do some of our most cherished inductive definitions come from? Which
+functions are we excluding when we form the type of all functions between two
+types? Does every field have an algebraic closure?
+
+Inspired by the modal approach to the set-theoretic multiverse of Joel David
+Hamkins, Victoria Gitman and their collaborators, we aim to introduce the modal
+operators "everywhere" and "somewhere" to homotopy type theory, proposing a
+multiversal perspective on these motivating questions.
+
+Unlike the set-theoretic role model, we focus less on exploring the range of
+foundational possibility and more on concrete applications in constructive
+mathematics, with the goal of porting results of classical mathematics to
+homotopy type theory. It will turn out that every field has an algebraic
+closure somewhere; that a transitive relation is well-founded iff nowhere there
+is an infinite descending chain; and that somewhere, the law of excluded middle
+holds.
+
+This is ongoing joint work with Alexander Oldenziel and connected to the recent
+advances with type-theoretic presheaf models and sheaf models.

### 2025-06-08 slides-types2025.lagda.tex
Final info slide
Some minor cleanups

@@
     \necessary\mathsf{Well}\sinf(X,{\le})$. \formalized
-    \item $(\possible\varphi) \Longrightarrow \varphi$,
+    \item $(\possible\varphi) \Longleftrightarrow \varphi$,
     if $\varphi$ is a geometric implication (``$\forall\ldots\forall (\%
@@
     in~$\%$).~\formalized
-    \item $(\xpossible\varphi) \Longrightarrow \varphi$,
+    \item $(\xpossible\varphi) \Longleftrightarrow \varphi$,
     if~$\varphi$ is a bounded first-order statement. \formalized
-    \item For every inhabited type~$X$, $\xpossible\exists f : \NN
-    \to X\_ \mathsf{Surjective}(f)$. \formalized
+    \item For every inhabited type~$X$, $\xpossible\necessary
+    \mathsf{Countable}(X)$, where~$\mathsf{Countable}(X) \defeqv (\exists f : \NN
+    \to X\_ \mathsf{Surjective}(f))$. \formalized
+
+    So being countable is a \hil{button}.
     \item $\possible \badbox{\textsc{lem}}$.
     (Baby Barr / Friedman's trick / nontrivial exit continuation)
+
+    In fact, \badbox{\textsc{lem}} is a \hil{switch}:
+    $\necessary\bigl((\possible\badbox{\textsc{lem}}) \wedge
+    (\possible\neg\badbox{\textsc{lem}})\bigr)$
+
+    \item $\badbox{\textsc{zorn}} \Rightarrow
+    \necessary\possible\badbox{\textsc{ac}}$. (Great Barr)
   \end{enumerate}
@@
     \item Realize a generic gadget as some kind of limit of \hil{approximations}
-    from the base universe. For the generic function $\NN \to X$:
+    from the base universe. For the generic function $f_0 : \NN \to X$:
 
@@
     \item \hil{Reinterpret}, in a mechanical fashion, assertions about the generic
-    gadget as assertions about its approximations~(\formalized). Be prepared to evolve
-    approximations.
+    gadget as assertions about its approximations~(\formalized).
+
+    {\small\begin{itemize}
+    \item We have the \hil{stage-dependent proposition} ``$f_0\,n = x$'',
+    a certain function $L \to \mathsf{Prop}$: $\lambda \sigma.\
+    (\mathsf{lookupMaybe}\,\sigma\,n = \mathsf{just}\,x)$\end{itemize}}
+    \medskip
+
+    \pause
+    \item Be prepared to evolve approximations (\formalized).
 
@@
     \item
-    For a \hil{stage-dependent proposition}~$P : L \to \mathsf{Prop}$, $\nabla\,P\,\sigma$ expresses
+    For a stage-dependent proposition~$P : L \to \mathsf{Prop}$, $\nabla\,P\,\sigma$ expresses
     that no matter how~$\sigma$ will evolve over a time to a better
@@
   % later₂ : (a : X) → ((τ : X*) → a ∈ σ ⧺ τ → ∇ P (σ ⧺ τ)) → ∇ P σ
-    \item We have the stage-dependent proposition ``$f_0\,n = x$'',
-    where~$f_0$ is the generic function.
     \item That~$f_0$ is defined on input~$n$ can be expressed
@@
 
+\begin{frame}{Formal metatheory}
+  \small
+  \good{\cmark{}} \emph{There are type-theoretic multiverses,} such as
+  \begin{itemize}
+    \item the collection of all $\mathrm{PSh}(\mathcal{C} \times \mathcal{B})$,
+    where~$\mathcal{B}$ ranges over cube categories and~$\mathcal{C}$ over
+    arbitrary small categories, and their corresponding sheaf models
+
+    {\scriptsize
+    Coquand. ``A survey of constructive presheaf models of univalence''. \emph{ACM SIGLOG News}, 5.3 (2018).
+    \par}
+  \end{itemize}
+  \medskip
+  \pause
+
+  \bad{\xmark} \emph{Accessing the multiverse from within intensional type theory is tricky:}
+  \begin{itemize}
+    \item Given a model of $\mathfrak{s}$CIC and a category~$\mathcal{C}$ in
+    it, we have a syntactic presheaf model of CIC.
+
+    {\scriptsize
+    Coquand, Jaber. ``A note on forcing and type theory''. \emph{Fundamenta Informaticae 100} (2010). \\
+    Jaber, Lewertowski, Pédrot, Sozeau, Tabareau. ``The definitional side of the forcing''. \emph{Proceedings of LICS '16} (2016). \\
+    Pédrot. ``Russian constructivism in a prefascist theory''. \emph{Proceedings of LICS '20} (2020).
+    \par}
+    \item Given a suitable lex modality, we have a syntactic sheaf model (model of modal types).
+
+    {\scriptsize
+    Coquand, Ruch, Sattler. ``Constructive sheaf models of type theory.'' \emph{Math. Struct. Comput. Sci.} 31.9 (2021). \\
+    Escardó, Xu. ``Sheaf models of type theory in type theory''. Unpublished (2016). \\
+    Quirin. ``Lawvere--Tierney sheafification in Homotopy Type Theory''. PhD thesis (2016).
+    \par}
+    \item (I believe) we have syntactic sheaf models in certain special cases,
+    when no coherence issues arise in defining the notion of presheaves.
+  \end{itemize}
+
+  Note: We can use~$\nabla$ even without a proper metatheoretic backing.
+\end{frame}
+
+{\renewcommand{\insertframeextra}{a}
 \begin{frame}{Increasing subsequences as convenient fictions}
@@
   actually is a finite prefix validating~$B$. \qed\par}
-\end{frame}
-
+\end{frame}}
+
+\addtocounter{framenumber}{-1}
 {\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
+\renewcommand{\insertframeextra}{b}
 \begin{frame}{Maximal ideals as convenient fictions}
@@
 
-\begin{frame}{Formal metatheory}
-  \small
-  \good{\cmark{}} \emph{There are type-theoretic multiverses,} such as
-  \begin{itemize}
-    \item the collection of all $\mathrm{PSh}(\mathcal{C} \times \mathcal{B})$,
-    where~$\mathcal{B}$ ranges over cube categories and~$\mathcal{C}$ over
-    arbitrary small categories, and their corresponding sheaf models
-
-    {\scriptsize
-    Coquand. ``A survey of constructive presheaf models of univalence''. \emph{ACM SIGLOG News}, 5.3 (2018).
-    \par}
-  \end{itemize}
-  \medskip
-  \pause
-
-  \bad{\xmark} \emph{Accessing the multiverse from within intensional type theory is tricky:}
-  \begin{itemize}
-    \item Given a model of $\mathfrak{s}$CIC and a category~$\mathcal{C}$ in
-    it, we have a syntactic presheaf model of CIC.
-
-    {\scriptsize
-    Coquand, Jaber. ``A note on forcing and type theory''. \emph{Fundamenta Informaticae 100} (2010). \\
-    Jaber, Lewertowski, Pédrot, Sozeau, Tabareau. ``The definitional side of the forcing''. \emph{Proceedings of LICS '16} (2016). \\
-    Pédrot. ``Russian constructivism in a prefascist theory''. \emph{Proceedings of LICS '20} (2020).
-    \par}
-    \item Given a suitable lex modality, we have a syntactic sheaf model (model of modal types).
-
-    {\scriptsize
-    Coquand, Ruch, Sattler. ``Constructive sheaf models of type theory.'' \emph{Math. Struct. Comput. Sci.} 31.9 (2021). \\
-    Escardó, Xu. ``Sheaf models of type theory in type theory''. Unpublished (2016). \\
-    Quirin. ``Lawvere--Tierney sheafification in Homotopy Type Theory''. PhD thesis (2016).
-    \par}
-    \item (I believe) we have syntactic sheaf models in certain special cases,
-    when no coherence issues arise in defining the notion of presheaves.
-  \end{itemize}
-
-  Note: We can use~$\nabla$ even without a proper metatheoretic backing.
+\backupstart
+
+\begin{frame}[plain]
+  \begin{columns}[c]
+    \begin{column}{0.45\textwidth}
+      \includegraphics[width=0.95\textwidth]{lets-play-agda}
+    \end{column}
+    \begin{column}{0.5\textwidth}
+      \includegraphics[width=0.8\textwidth]{ai-2025}
+
+      \vspace*{-1.5cm}\hspace*{2cm}
+      \includegraphics[width=0.8\textwidth]{cirm-2025}
+
+      \vspace*{1cm}
+    \end{column}
+  \end{columns}
 \end{frame}
-\backupstart
 

### 2025-06-09 slides-types2025.lagda.tex
Slides as presented this morning

@@
     \pause
-    \item Be prepared to evolve approximations (\formalized).
+    \item Be prepared to \hil{evolve approximations} (\formalized).
 
@@
 
+\addtocounter{framenumber}{-1}
+{\renewcommand{\insertframeextra}{c}
+\begin{frame}{Stabilization as convenient fiction}
+  A commutative ring~$A$ with unit is \ldots
+  \begin{itemize}
+    \item Noetherian\sinf{} iff for every sequence~$x_0,x_1,\ldots$ of ring elements,
+    there is a number~$n : \NN$ such that~$x_n,x_{n+1},x_{n+2},\ldots \in
+    (x_0,\ldots,x_{n-1})$.
+    \item Noetherian\ensuremath{_\text{RS}} iff for every sequence~$x_0,x_1,\ldots$ of ring elements,
+    there is a number~$n : \NN$ such that~$x_n \in
+    (x_0,\ldots,x_{n-1})$.
+    \item Noetherian\ind{} iff a certain inductively defined condition holds
+    [Coquand–Persson].
+  \end{itemize}
+
+  We then have:
+  \begin{align*}
+    \mathsf{Noetherian\ind}(A) &\Longleftrightarrow
+    \necessary\mathsf{Noetherian}_\text{RS}(A) \\
+    &\Longleftrightarrow
+    \necessary\bigl(\badbox{\textsc{lem}} \Rightarrow
+    \mathsf{Noetherian}\sinf(A)\bigr) \\
+    &\Longleftrightarrow
+    \necessary\possible\mathsf{Noetherian}\sinf(A).
+  \end{align*}
+
+  Hence, from the corresponding classical proofs, we can extract
+  constructive proofs of:
+  \begin{enumerate}
+    \item Hilbert's basis theorem: If $A$ is Noetherian\ind, then so is $A[X]$.
+    \item Locality: If~$1 = f_1+\ldots+f_n$ and if each
+    ring~$A[f_i^{-1}]$ is Noetherian\ind, then so is~$A$. \\
+    (As a consequence, $A$ is Noetherian\ind
+    iff~$\mathcal{O}_{\mathrm{Spec}(A)}$ is.)
+  \end{enumerate}
+\end{frame}
+
 \backupstart

### 2025-06-20 slides-paris2025.tex
typo

@@
     generic prime ideal as an actual prime ideal in a suitable forcing extension,
-    so in a suitable topos (der little Zariski topos of the ring).
+    so in a suitable topos (the little Zariski topos of the ring).
   }

### 2025-06-20 slides-types2025.lagda.tex
Notes slides

@@
+% compile using: agda --latex; result will then be in the latex/ subfolder
 \documentclass[11pt,utf8,notheorems,compress,t,aspectratio=169]{beamer}
@@
 % https://tex.stackexchange.com/questions/164406/beamer-using-href-in-notes.
-\newcommand{\fixedhref}[2]{\makebox[0pt][l]{\hspace*{\paperwidth}\href{#1}{#2}}\href{#1}{#2}}
+\newcommand{\fixedhref}[2]{\parbox[t][0pt][t]{0pt}{\vspace*{-\paperheight}\vspace*{-0.6em}\href{#1}{#2}}\href{#1}{#2}}
 
@@
 
-\title{Modal operators for a constructive account of well quasi-orders}
+\title{Towards topological type theory for decrypting transfinite methods in classical mathematics}
 
 \author{Ingo Blechschmidt}
-\date{November 30th, 2024}
-
-%\setbeameroption{show notes on second screen=bottom}
+\date{June 9th, 2025}
+
+\setbeameroption{show notes on second screen=bottom}
 \newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\footnotesize\justifying#2\par}}}
+\makeatletter
+\setbeamertemplate{note page}
+{%
+  {%
+    \scriptsize
+    \insertvrule{.25\paperheight}{white!90!black}
+    \vskip-.25\paperheight
+    \nointerlineskip
+    \vbox{
+      %\hfill\insertslideintonotes{0.25}\hskip-\Gm@rmargin\hskip0pt%
+      %\vskip-0.25\paperheight%
+      %\nointerlineskip
+      \begin{pgfpicture}{0cm}{0cm}{0cm}{0cm}
+        \begin{pgflowlevelscope}{\pgftransformrotate{90}}
+          {\pgftransformshift{\pgfpoint{-2cm}{0.2cm}}%
+          \pgftext[base,left]{\footnotesize\the\year-\ifnum\month<10\relax0\fi\the\month-\ifnum\day<10\relax0\fi\the\day}}
+        \end{pgflowlevelscope}
+      \end{pgfpicture}}
+    \nointerlineskip
+    \vbox to .25\paperheight{\vskip0.5em
+      \hbox{\insertshorttitle[width=8cm]}%
+      \setbox\beamer@tempbox=\hbox{\insertsection}%
+      \hbox{\ifdim\wd\beamer@tempbox>1pt{\hskip4pt\raise3pt\hbox{\vrule
+            width0.4pt height7pt\vrule width 9pt
+            height0.4pt}}\hskip1pt\hbox{\begin{minipage}[t]{7.5cm}\def\breakhere{}\insertsection\end{minipage}}\fi%
+      }%
+      \setbox\beamer@tempbox=\hbox{\insertsubsection}%
+      \hbox{\ifdim\wd\beamer@tempbox>1pt{\hskip17.4pt\raise3pt\hbox{\vrule
+            width0.4pt height7pt\vrule width 9pt
+            height0.4pt}}\hskip1pt\hbox{\begin{minipage}[t]{7.5cm}\def\breakhere{}\insertsubsection\end{minipage}}\fi%
+      }%
+      \setbox\beamer@tempbox=\hbox{\insertshortframetitle}%
+      \hbox{\ifdim\wd\beamer@tempbox>1pt{\hskip30.8pt\raise3pt\hbox{\vrule
+            width0.4pt height7pt\vrule width 9pt
+            height0.4pt}}\hskip1pt\hbox{\insertshortframetitle[width=7cm]}\fi%
+      }%
+      \vfil}%
+  }%
+  \vskip.25em
+  \nointerlineskip
+  \insertnote
+}
+\makeatother
 
@@
 \newunicodechar{τ}{\ensuremath{\tau}}
-\newunicodechar{∷}{\ensuremath{::}}
+\newunicodechar{∷}{\ensuremath{\dblcolon}}
 \newunicodechar{⧺}{\ensuremath{\!\!\!}}
@@
 
+\addtocounter{framenumber}{-1}
+\begin{frame}\frametitle{A classical logic fairy tale}
+  \centering\bigskip\bigskip\bigskip
+  \includegraphics[height=9.5em]{philosophers-stone}
+
+  \jnote{1}{
+    \vspace*{-1em}
+    \begin{changemargin}{-1.5em}{-1.5em}
+      \setlength{\columnsep}{15pt}
+      \begin{multicols}{2}
+        \justifying
+        \scriptsize
+
+        \textbf{Narrator.} Once upon a time, in a kingdom far, far away, the
+        queen of the country and of all Möbius strips called for her royal
+        philosopher.
+
+        \textbf{Queen.} Philosopher! I ask you to carry out the following order.
+        Get me the Philosopher's Stone, or alternatively find out how one could
+        produce arbitrary amounts of gold with it!
+
+        \textbf{Philosopher.} But my queen! I haven't studied anything useful!
+        How could I fulfill this order?
+
+        \textbf{Queen.} That is not my concern. I'll see you again tomorrow.
+        Should you not accomplish the task, I will take your head off.
+
+        \textbf{Narrator.} After a long and wakeful night the philosopher was
+        called to the queen again.
+
+        \textbf{Queen.} Tell me! What do you have to report?
+
+        \textbf{Philosopher.} It was not easy and I needed to follow lots of
+        obscure references, but finally I actually found out how to use the
+        Philosopher's Stone to produce arbitrary amounts of gold. But only I can
+        conduct this procedure, your royal highness.
+
+        \textbf{Queen.}
+        Alright. So be it.
+
+        \textbf{Narrator.} And so years passed by, during which the philosopher
+        imagined herself to be safe. The queen searched for the stone on her own,
+        but as long as she hadn't found it, the philosopher didn't need to worry.
+        Yet one day the impossible happened: The queen has found the stone! And
+        promptly called for her philosopher.
+
+        \textbf{Queen.} Philosopher, look! I have found the Philosopher's Stone!
+        Now live up to your promise! \emph{[She hands over the stone.]}
+
+        \textbf{Philosopher.} Thank you. \emph{[She inspects the stone.]} This is
+        indeed the Philosopher's Stone. Many years ago you asked me to either
+        acquire the Philosopher's Stone or find out how to produce arbitrary
+        amounts of gold using it. Now it's my pleasure to present to you the
+        Philosopher's Stone. \emph{[She returns the stone.]}
+      \end{multicols}
+    \end{changemargin}
+  }
+
+  \jnote{2}{
+    Adapted from
+    \fixedhref{http://blog.ezyang.com/2013/04/a-classical-logic-fairy-tale/}{Edward
+    Yang's blog}.
+
+    The idea conveyed in the fairy tale is at the core of the
+    \emph{constructive} proof of $\neg\neg(A \vee \neg A)$. While the law of
+    excluded middle (\textsc{lem}) is not available in constructive
+    mathematics, up to a double negation every instance of it is.
+
+    This observation gives rise to the double negation translation, a
+    transformation of (statements and) proofs in classical logic to proofs in
+    intuitionistic logic, at the only expense of introducing double negations;
+    by following this up with Friedman's trick with the nontrivial exit
+    continuation (``Baby Barr Theorem''), we can in many cases even get rid of
+    these double negations in a second step.
+
+    Getting rid of applications of the law of excluded middle is in many
+    contexts (an amazing mind-boggingly wonder at first, but then) a routine
+    matter---see \href{https://rt.quasicoherent.io/}{here} for more details and
+    pointers to the literature.
+  }
+\end{frame}
+
 {\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{1.7em}\includegraphics[height=\paperheight]{sea-of-clouds-1}\end{minipage}}
 \begin{frame}{A case study in Hilbert's program}
+  \jnote{1}{
+    Well quasi-orders are an important notion in proof theory and termination
+    analysis.
+  }
+
+  \jnote{2-3}{
+    \vspace*{-0.5em}
+    The presented proof rests on the law of excluded middle and hence cannot
+    immediately be interpreted as a program for finding suitable indices~$i <
+    j$. However, constructive proofs are also possible (for instance by
+    induction on the value of a given term of the sequence, see
+    \href{https://www.math.lmu.de/~petrakis/Dickson.pdf}{Constructive
+    combinatorics of Dickson's Lemma} by Iosif Petrakis for several fine
+    quantitative results). And even more:
+    There is a procedure for regarding this proof---and many others in the
+    theory of well quasi-orders---as \emph{blueprints} for more informative
+    constructive proofs. This shall be our motto for today:
+
+    \emph{Do not take classical proofs literally, instead ask which
+    constructive proofs they are blueprints for.}
+  }
+
+  \jnote{3}{
+    The displayed stability results, along with several others, provide a
+    flexible toolbox for constructing new well quasi-orders from given ones.
+    However, with the classical formulation of \emph{well},
+    shortly to be renamed~``well\sinf'', these results are
+    \href{https://lets-play-agda.quasicoherent.io/Padova2025.Explorations.Forcing.Orthogonal.html}{inherently
+    classical} (even the weaker claim ``the product of two streamless sets is
+    streamless'' admits countermodels).
+
+    (In Higman's lemma, the set~$X^*$ of finite lists of elements of~$X$ is
+    equipped with the following ordering: We have~$(x_0 \dblcolon \ldots
+    \dblcolon x_{n-1} \dblcolon [\,]) \leq (y_0 \dblcolon \ldots \dblcolon
+    y_{m-1} \dblcolon [\,])$ iff there is an increasing injection~$f : \{
+    0,\ldots,n-1 \} \to \{ 0,\ldots,m-1 \}$ such that~$x_i \leq y_{f(i)}$ for all~$i < n$.
+    Kruskal's theorem is the result behind the celebrated enormous number
+    $\mathsf{TREE}(3)$.)
+  }
+
+  \jnote{7}{
+    Thanks to work by Thierry Coquand, Daniel Fridlender and Monika
+    Seisenberger, a constructive substitute is available, the notion well\ind.
+    In classical mathematics (where bar induction is available), this notion is
+    equivalent to well\sinf.
+
+    The predicate~$\mathsf{Good}$ of finite lists~$\sigma = (x_0 \dblcolon
+    \ldots \dblcolon x_{n-1} \dblcolon [\,])$ appearing in the definition
+    of~$G$ expresses that some earlier term is at most some later term,
+    i.\,e.\@ that there merely exist~$i < j$ such that~$x_i \leq x_j$.
+
+    The operator $\dblcolon^r$ is adding an element at the end.
+  }
+
+  \jnote{8}{
+    \begin{columns}[t]
+      \begin{column}{0.65\textwidth}
+        The original notion well\sinf:
+        \begin{itemize}
+          \justifying
+          \item[\cmark] short and simple
+          \\[-1.3em]
+          \item[\cmark] constructively satisfied for the main examples (but
+          only because of the theory around well\ind)
+          \\[-1.3em]
+          \item[\cmark] concise abstract proofs (albeit employing transfinite methods)
+          \\[-1.3em]
+          \item[\xmark] main results not constructively attainable
+          \\[-1.3em]
+          \item[\xmark] philosophically strenuous by the quantification over all sequences
+          \\[-1.3em]
+          \item[\xmark] not stable under ``change of base''---a forcing extension
+          of the universe may well contain more sequences than the base universe
+          \\[-1.3em]
+          \item[\xmark] negative (universal) condition
+        \end{itemize}
+      \end{column}
+
+      \begin{column}{0.33\textwidth}
+        The constructive substitute well\ind:
+        \begin{itemize}
+          \justifying
+          \item[\cmark] main results constructive
+          \\[-1.3em]
+          \item[\cmark] stable under change of base
+          \\[-1.3em]
+          \item[\cmark] positive (existential) condition
+          \\[-1.3em]
+          \item[\xmark] proofs intriguing, but also somewhat alien, not just some
+          trivial reshuffling of the classical arguments; classical sequence
+          language cannot be used
+        \end{itemize}
+      \end{column}
+    \end{columns}
+  }
+
   \textbf{Def.} Let~$(X,{\leq})$ be a quasi-order.
@@
     \qquad
-    \infer{\mathsf{later}\,f : G\,\sigma}{f : (x : X) \to G\,(\sigma ::^r x)}
+    \infer{\mathsf{later}\,f : G\,\sigma}{f : (x : X) \to G\,(\sigma \dblcolon^r x)}
   \]
@@
   }
-
-  \only<9>{
-    \vspace*{-0.0em}
-    \hil{Central insight:} A quasi-order $X$ is well\ind iff $\hil{$\necessary$}\,\forall\alpha : \NN \to
-    X\_ \exists i < j\_ \alpha\,i \leq \alpha\,j$.
-  }
 \end{frame}}
@@
 \begin{frame}{Missing functions in the type of all functions?}
+  \jnote{1}{
+    The double negation translation technique fails to constructivize the two
+    proofs presented on the slide.
+
+    The reason is that, after applying the translation, the ``$I$-extracted
+    subsequence'' appearing in the proof on the left will no longer actually be
+    a function~$\NN \to X$. Instead, it will be a partially defined function
+    with the property that each number~$n$ does \emph{not~not} belong to the
+    domain. However, the assumption that~$X$ is well\sinf{} does not apply to
+    these kinds of generalized sequences.
+
+    In contrast, the notion well\ind{} is much stronger. If~$X$ is well\ind,
+    then not only will every actual sequence~$\NN \to X$ be good, but (in a
+    suitable sense) so will every up-to-$\neg\neg$-partially-defined sequence,
+    every multivalued sequence and many more kinds of generalized sequences.
+  }
+
   \begin{columns}[t]
@@
       \small
-      \item Force a \hil{generic sequence} $\NN \to X$. \formalized
+      \item Force a \hil{generic sequence} $\NN \to X$.
+            \fixedhref{https://lets-play-agda.quasicoherent.io/Padova2025.Explorations.Forcing.Cohen.html}{\formalized}
       \item Force a \hil{generic enumeration} $\NN \twoheadrightarrow X$
-            (even if $X$ is uncountable). \formalized
-      \item Force a \hil{generic prime ideal} of a given ring. \formalized
+            (even if $X$ is uncountable).
+            \fixedhref{https://lets-play-agda.quasicoherent.io/Padova2025.Explorations.Forcing.Levy.html}{\formalized}
+      \item Force a \hil{generic prime ideal} of a given ring.
+            \fixedhref{https://iblech.github.io/constructive-maximal-ideals/Forcing.Prime.html}{\formalized}
     \end{itemize}
@@
       \item A quasi-order~$X$ is well\ind iff the generic sequence~$\NN \to X$
-      is good. \formalized
+      is good.
+      \item A set is Noetherian iff the generic sequence~$\NN \to X$
+      has two repeated terms.
       \item A relation is well-founded iff for the generic descending chain,
-      $\bot$. \formalized
+      $\bot$.
       \item A ring element is nilpotent iff it is contained in the generic
-      prime ideal. \formalized
+      prime ideal.
+      \fixedhref{https://iblech.github.io/constructive-maximal-ideals/Forcing.Prime.html\#corollary}{\formalized}
     \end{itemize}
@@
       \item \ \\[-1.2em]\mbox{For the generic surjection $f : \NN \twoheadrightarrow \RR$,
-      $\neg\neg \exists n \? \NN\_ f(n) = \pi \wedge f(n+1) = e$. \formalized}
+      $\neg\neg \exists n \? \NN\_ f(n) = \pi \wedge f(n+1) = e$.
+      \fixedhref{https://lets-play-agda.quasicoherent.io/Padova2025.Explorations.Forcing.Levy.html}{\formalized}}
     \end{itemize}
   \end{changemargin}
+
+  \jnote{1}{
+    The technique for enlarging the mathematical universe is called
+    \emph{forcing} and was originally pioneered by Paul Cohen. Set theorists
+    use forcing to construct new models of set theory from given ones, in order
+    to explore the range of set-theoretic possibility. For instance, by forcing
+    we can construct models of~\textsc{zfc} validating the continuum hypothesis
+    and also models which falsify it.
+
+    We here refer to a simplification of original forcing which is useful in a
+    constructive metatheory. At its core, every forcing extension is just a
+    formula and proof translation of a certain form. For instance, there is a
+    forcing extension validating~\bad{\textsc{lem}} even if the base universe
+    does not; this forcing extension is not a deep mystery, for a statement holds
+    in that forcing extension iff its double negation translation holds in the
+    base universe and it is well-known that the double negation translation
+    of every instance of~\bad{\textsc{lem}} is an intuitionistic tautology.
+
+    \href{https://www.speicherleck.de/iblech/stuff/slides-herrsching2023.pdf}{Here
+    is a set of slides on constructive forcing}, and Section~4 of
+    \href{https://raw.githubusercontent.com/iblech/constructive-maximal-ideals/master/tex/extended.pdf}{this
+    joint paper with Peter Schuster} contains a written summary of constructive
+    forcing.
+  }
+
+  \jnote{2}{
+    Forcing the existence of a new function~$\NN \to X$ is also called
+    \emph{Cohen forcing}. Forcing the existence of a new surjection is called
+    \emph{Lévy collapse}. Forcing the existence of a prime ideal does not have
+    an established name, perhaps it should be called \emph{Zariski forcing}; it
+    has been explored by Thierry Coquand and others of the constructive algebra
+    crew.
+
+    The multiversal yoga allows us to reason with the good notion well\ind{}
+    using the simple language of well\sinf, and similarly with several other
+    notions.
+  }
+
+  \jnote{3}{
+    For more on the mystery of nongeometric sequents, see
+    \href{https://www.speicherleck.de/iblech/stuff/slides-birmingham2019.pdf}{this
+    set of slides}.
+  }
 \end{frame}}
@@
 \begin{frame}{A modal language for harnessing the multiverse}
+  \jnote{1}{
+    By \emph{topos}, we mean \emph{Grothendieck topos}. In constructive
+    forcing, a ``forcing extension of the base universe'' is exactly the same
+    thing as a Grothendieck topos.
+
+    A particular member of the rich and varied landscape of toposes is the
+    \emph{trivial topos}, in which every statement whatsoever holds. By
+    restricting to positive toposes, we exclude this special case.
+
+    For positive toposes~$\E$, a geometric implication holds in~$\E$ iff it
+    holds in the base universe. For positive overt toposes~$\E$, we even have
+    that a bounded first-order formula holds in~$\E$ iff it holds in the base.
+    Hence, for the purpose of verifying a bounded first-order assertion about
+    the base, we can freely pass to a positive overt topos with problem-adapted
+    better higher-order properties (such as that some uncountable set from the
+    base now appears countable, or that an infinite sequence whose existence is
+    predicted by failing dependent choice now actually exists).
+
+    \href{https://www.speicherleck.de/iblech/stuff/early-draft-modal-multiverse.pdf}{Here
+    is a rough early draft of a preprint with more details about the modal
+    multiverse.}
+  }
+
+  \jnote{2}{
+    The idea to study the modal multiverse of toposes in a principled manner
+    was proposed by Alexander Oldenziel in 2016. \emph{Foreshadowed by:}
+    \begin{itemize}
+      \item[1984\phantom{s}] André Joyal, Miles Tierney. ``An extension of the Galois theory of Grothendieck''.
+      \\[-1em]
+      \item[1987\phantom{s}] Andreas Blass. ``Well-ordering and induction in intuitionistic logic and topoi''.
+      \\[-1em]
+      \item[2010s] Milly Maietti, Steve Vickers. Ongoing work on arithmetic universes.
+      \\[-1em]
+      \item[2011\phantom{s}] Joel David Hamkins. ``The set-theoretic multiverse''.
+      \\[-1em]
+      \item[2013\phantom{s}] Shawn Henry. ``Classifying topoi and preservation of higher order logic by geometric morphisms''.
+    \end{itemize}
+  }
+
+  \jnote{3}{
+    With the modal language we seek to provide an accessible and modular
+    framework for constructivization results.
+
+    For instance, conservativity of classical logic over intuitionistic logic
+    for geometric implications (known under various names such as Barr's
+    theorem, Friedman's trick, escaping the continuation monad, \ldots) is
+    packaged up by the observation that \emph{somewhere}, the law of excluded
+    middle holds.
+
+    Another example:
+    In the community around Krull's lemma, it is well-known that
+    we can constructively infer that a given ring element~$x \in A$ is nilpotent from
+    knowing that it is contained in the \emph{generic prime ideal} of~$A$. This entity
+    is not actually an honest prime ideal of the ring~$A$ in the base
+    universe, but a certain combinatorial notion (efficiently dealt with using
+    \emph{entailment relations}). Constructive forcing allows us to reify the
+    generic prime ideal as an actual prime ideal in a suitable forcing extension,
+    so in a suitable topos (a version of the little Zariski topos of the ring).
+  }
+
   \textbf{Def.} A statement~$\varphi$ holds \ldots
@@
     \item $\mathsf{Well}\ind(X,{\le}) \Longleftrightarrow
-    \necessary\mathsf{Well}\sinf(X,{\le})$. \formalized
+    \necessary\mathsf{Well}\sinf(X,{\le})$.
     \item $(\possible\varphi) \Longleftrightarrow \varphi$,
@@
     \Rightarrow \%)$'', with no~$\forall$ nor~$\Rightarrow$
-    in~$\%$).~\formalized
+    in~$\%$).~\fixedhref{https://iblech.github.io/constructive-maximal-ideals/Forcing.Semantics.html\#CoherentEquivalence}{\formalized}
     \item $(\xpossible\varphi) \Longleftrightarrow \varphi$,
-    if~$\varphi$ is a bounded first-order statement. \formalized
+    if~$\varphi$ is a bounded first-order statement.
+    \fixedhref{https://iblech.github.io/constructive-maximal-ideals/Forcing.Semantics.html\#1\%CB\%A2\%E1\%B5\%97OrderEquivalence}{\formalized}
     \item For every inhabited type~$X$, $\xpossible\necessary
     \mathsf{Countable}(X)$, where~$\mathsf{Countable}(X) \defeqv (\exists f : \NN
-    \to X\_ \mathsf{Surjective}(f))$. \formalized
+    \to X\_ \mathsf{Surjective}(f))$.
+    \fixedhref{https://lets-play-agda.quasicoherent.io/Padova2025.Explorations.Forcing.Levy.html}{\formalized}
 
@@
 \begin{frame}{The plain combinatorics of toposes}
+  \jnote{3-}{
+    Proofs (about the base universe) are (pure) programs; proofs about a forcing
+    extension are effectful programs unfolding in the~$\nabla$ monad.
+
+    The $\nabla$ monad is a refinement of the (monotonic) state monad.
+  }
+
   \begin{enumerate}\justifying
@@
     \item \hil{Reinterpret}, in a mechanical fashion, assertions about the generic
-    gadget as assertions about its approximations~(\formalized).
+    gadget as assertions about its
+    approximations~(\fixedhref{https://iblech.github.io/constructive-maximal-ideals/Forcing.Semantics.html}{\formalized}).
 
@@
     \pause
-    \item Be prepared to \hil{evolve approximations} (\formalized).
+    \item Be prepared to \hil{evolve approximations}
+    (\fixedhref{https://iblech.github.io/constructive-maximal-ideals/Forcing.Base.html}{\formalized}).
 
@@
 \begin{frame}{Formal metatheory}
+  \jnote{2}{
+    \href{https://msp.cis.strath.ac.uk/types2025/slides/TYPES2025-slides57.pdf}{Exciting
+    ongoing work on split type theory} by Martin Baillon, Assia Mahboubi and
+    Pierre-Marie Pédrot!
+  }
+
   \small
@@
 \begin{frame}{Increasing subsequences as convenient fictions}
+  \jnote{1}{
+    On this slide~$\nabla$ is the monad described on slide 5,
+    and~$\nabla^\nnearrow$ is the variant where lists are only allowed to grow
+    in a monotonically increasing manner; so~$\nabla$ is the monad for the
+    generic sequence~$\NN \to X$, while~$\nabla^\nnearrow$ is the monad for the
+    generic increasing sequence~$\NN \to X$.
+
+    The displayed multiversal proof gives a positive answer to a question posed
+    by Stefano Berardi, Gabriele Buriola and Peter Schuster, see
+    \href{https://www.speicherleck.de/iblech/stuff/slides-abmv2024.pdf\#page=16}{this
+    set of slides}.
+
+    At the end of the day, this proof yields a concrete algorithm (see
+    \href{https://www.speicherleck.de/iblech/stuff/dickson/implicational.pl}{here}
+    for manually-extracted Perl code). Given an infinite sequence, this
+    algorithm seeks, in a certain systematic but not exhaustive manner, a
+    sufficiently long monotonic subsequence, making judicious use of
+    backtracking. This search terminates, though this fact is not obvious just
+    by looking at the algorithm.
+  }
+
   \justifying
@@
     \textbf{Thm.} If~$X$ is well\ind and if~$\nabla^\nnearrow\,B\,[\,]$,
-    then~$\nabla\,B\,[\,]$. \formalized
+    then~$\nabla\,B\,[\,]$.
+    \fixedhref{https://www.speicherleck.de/iblech/stuff/dickson//Dickson.html}{\formalized}
   \end{block}
@@
 \begin{frame}{Maximal ideals as convenient fictions}
+  \jnote{1-}{
+    The displayed classical proof is quite efficient from the point of view of
+    organizing mathematical knowledge, as it quickly reduces the general
+    situation of dealing with an arbitrary ring to dealing with a field.
+    Alas, read literally, it is hopeless ineffective.
+  }
+
+  \jnote{2-}{
+    By employing modal language, we can closely mimic the original proof and be
+    fully constructive at the same time.
+  }
+
+  \jnote{3-}{
+    By unwinding all modal definitions, the modal proof can be unrolled to a
+    fully explicit computation.
+  }
+
   \justifying
@@
   We may work \emph{somewhere} where \good{\textsc{lem}} holds. So \good{assume not}.
-  \emph{Proximally,} there is a \good{maximal ideal}~$\mmm$~(\formalized).
+  \emph{Proximally,} there is a \good{maximal
+  ideal}~$\mmm$~(\fixedhref{https://iblech.github.io/constructive-maximal-ideals/index.html\#§4.§4-4}{\formalized}).
   The matrix~$M$ is still surjective \emph{there}, and also over~$A/\mmm$. Since~$A/\mmm$ is a field, this
@@
 
-  \emph{Unrolled constructive proof (special case).}
+  \emph{Unrolled constructive proof (special case,
+  \fixedhref{https://iblech.github.io/constructive-maximal-ideals/index.html\#§4.TestCase}{\formalized}).}
   Write~$M =
@@
     \item Noetherian\ind{} iff a certain inductively defined condition holds
-    [Coquand–Persson].
+    \fixedhref{https://dl.acm.org/doi/10.5555/646538.696017}{[Coquand–Persson]}.
   \end{itemize}
@@
 
-  Hence, from the corresponding classical proofs, we can extract
+  With this observation, from their classical counterparts, we can extract
   constructive proofs of:
@@
     (As a consequence, $A$ is Noetherian\ind
-    iff~$\mathcal{O}_{\mathrm{Spec}(A)}$ is.)
+    iff algebraic geometry's~$\mathcal{O}_{\mathrm{Spec}(A)}$ is.)
   \end{enumerate}
@@
 \begin{frame}[plain]
+  \jnote{1}{
+    \href{https://lets-play-agda.quasicoherent.io/}{Let's play Agda}, an
+    interactive course on Agda, complementing the
+    \href{https://agdapad.quasicoherent.io/}{Agdapad}.
+
+    \href{https://www.uni-augsburg.de/en/fakultaet/mntf/math/prof/geom/workshop-aimath/}{AI
+    Transforms Maths Research}, a workshop in Augsburg.
+
+    \href{https://conferences.cirm-math.fr/3377.html}{Synthetic mathematics}, a
+    conference in Luminy.
+  }
+
   \begin{columns}[c]

### 2025-08-25 slides-ai2025.tex
Beginning of AI slides

--- 2025-06-08 slides-types2025.lagda.tex
@@
 \usepackage{xstring}
+\usepackage{environ}
 \usepackage{proof}
@@
 
-\title{Modal operators for a constructive account of well quasi-orders}
+\title{Towards topological type theory for decrypting transfinite methods in classical mathematics}
 
 \author{Ingo Blechschmidt}
-\date{November 30th, 2024}
+\date{June 9th, 2025}
 
@@
 \newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\footnotesize\justifying#2\par}}}
+\makeatletter
+\setbeamertemplate{note page}
+{%
+  {%
+    \scriptsize
+    \insertvrule{.25\paperheight}{white!90!black}
+    \vskip-.25\paperheight
+    \nointerlineskip
+    \vbox{
+      %\hfill\insertslideintonotes{0.25}\hskip-\Gm@rmargin\hskip0pt%
+      %\vskip-0.25\paperheight%
+      %\nointerlineskip
+      \begin{pgfpicture}{0cm}{0cm}{0cm}{0cm}
+        \begin{pgflowlevelscope}{\pgftransformrotate{90}}
+          {\pgftransformshift{\pgfpoint{-2cm}{0.2cm}}%
+          \pgftext[base,left]{\footnotesize\the\year-\ifnum\month<10\relax0\fi\the\month-\ifnum\day<10\relax0\fi\the\day}}
+        \end{pgflowlevelscope}
+      \end{pgfpicture}}
+    \nointerlineskip
+    \vbox to .25\paperheight{\vskip0.5em
+      \hbox{\insertshorttitle[width=8cm]}%
+      \setbox\beamer@tempbox=\hbox{\insertsection}%
+      \hbox{\ifdim\wd\beamer@tempbox>1pt{\hskip4pt\raise3pt\hbox{\vrule
+            width0.4pt height7pt\vrule width 9pt
+            height0.4pt}}\hskip1pt\hbox{\begin{minipage}[t]{7.5cm}\def\breakhere{}\insertsection\end{minipage}}\fi%
+      }%
+      \setbox\beamer@tempbox=\hbox{\insertsubsection}%
+      \hbox{\ifdim\wd\beamer@tempbox>1pt{\hskip17.4pt\raise3pt\hbox{\vrule
+            width0.4pt height7pt\vrule width 9pt
+            height0.4pt}}\hskip1pt\hbox{\begin{minipage}[t]{7.5cm}\def\breakhere{}\insertsubsection\end{minipage}}\fi%
+      }%
+      \setbox\beamer@tempbox=\hbox{\insertshortframetitle}%
+      \hbox{\ifdim\wd\beamer@tempbox>1pt{\hskip30.8pt\raise3pt\hbox{\vrule
+            width0.4pt height7pt\vrule width 9pt
+            height0.4pt}}\hskip1pt\hbox{\insertshortframetitle[width=7cm]}\fi%
+      }%
+      \vfil}%
+  }%
+  \vskip.25em
+  \nointerlineskip
+  \insertnote
+}
+\makeatother
 
@@
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\vspace*{-7em}\includegraphics[width=\paperwidth]{forest2}\end{minipage}}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\vspace*{-0em}\includegraphics[width=\paperwidth]{lost-melody-2}\end{minipage}}
 \begin{frame}[c]
@@
   \bigskip
-  \bigskip
-  \bigskip
-  \bigskip
 
@@
   \setbeamercolor{block body}{bg=black!100}
-  \begin{minipage}{0.52\textwidth}
+  \begin{minipage}{0.40\textwidth}
     \begin{block}{}
       \centering\normalsize\color{white}
-      \hil{\color{white}Towards topological type theory for decrypting
-      transfinite methods in classical~mathematics} \\[-0.4em]
+      \hil{\color{white} Proofs as programs?} \\[-0.4em]
       \
@@
   \bf
-  \colorbox{black}{\begin{minipage}{0.2\textwidth}
+  \colorbox{black}{\begin{minipage}{0.3\textwidth}
     \centering
-    TYPES 2025 \\
-    June 9th, 2025
+    AI Transforms Math Research \\
+    University of Augsburg \\
+    August 26th, 2025
   \end{minipage}}
@@
 
-  \colorbox{black}{\begin{minipage}{0.2\textwidth}
+  \colorbox{black}{\begin{minipage}{0.3\textwidth}
     \centering
@@
 
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{1.7em}\includegraphics[height=\paperheight]{sea-of-clouds-1}\end{minipage}}
-\begin{frame}{A case study in Hilbert's program}
-  \textbf{Def.} Let~$(X,{\leq})$ be a quasi-order.
-  \vspace*{-0.3em}
-  \begin{itemize}
-    \item A sequence~$\alpha : \NN \to X$ is \hil{good} iff
-    there merely exist~$i < j$ with~$\alpha\,i \leq \alpha\,j$.
-    \item The quasi-order~$X$ is \hil{well\only<4->{\sinf}} iff every sequence~$\NN \to X$ is good.
-  \end{itemize}
-  \pause
-  \vspace*{-1.3em}
-
-  \begin{columns}[t]
-    \begin{column}{0.46\textwidth}
-      \begin{block}{Natural numbers\phantom{y}}
-        \textbf{Prop.} $(\NN, {\leq})$ is well\only<4->{\sinf}. \formalized
-        \smallskip
-
-        \emph{Proof.} Let~$\alpha : \NN \to \NN$. By~\badbox{\textsc{lem}}, there is a
-        \bad{minimum}~$\alpha\,i$.
-        Set~$j \defeq i + 1$. \qed
-
-        \hfill\color{gray}offensive?
-      \end{block}
-      \only<2-6>{
-        \vspace*{-2em}
-        \[ \astikznodetransparentlycircled{xm}{7}\!,
-        \quad \astikznodetransparentlycircled{x0}{4}\!,
-        \quad \astikznodecircled{t1}{mypurple}{3}\!,
-        \quad \astikznodetransparentlycircled{x1}{1}\!,
-        \quad \astikznodecircled{t2}{mypurple}{8}\!,
-        \quad \astikznodetransparentlycircled{x2}{2}\!,
-        \quad \ldots \]
-      {\centering\begin{tikzpicture}[remember picture,overlay]
-        \node[draw=mypurple, circle, thick, inner sep=0.1em] (t3) {\scriptsize$\leq$};
-        \path[draw=mypurple,thick]
-          (t1)
-          to [out=-90, in=180] (t3)
-          to [out=0, in=-90] (t2);
-      \end{tikzpicture}\par}}
+\begin{frame}{In this talk}
+  \bigskip
+  \bigskip
+  \begin{columns}
+    \begin{column}{0.25\textwidth}
+      \centering\footnotesize
+      \includegraphics[height=2cm]{calabi-yau} \mbox{commutative algebra and}
+      \\
+      \mbox{number theory}
     \end{column}
-
-    \pause
-
-    \begin{column}{0.50\textwidth}
-      \begin{block}{Key stability results}
-        \justifying
-        \only<1-5>{Assuming~\badbox{\textsc{lem}}
-        and~\badbox{\textsc{dc}}}\only<6->{\good{Constructively}}, \ldots
-
-        \hil{Dickson:} \tabto{1.7cm} If~$X$ and~$Y$ are well\only<4-5>{\sinf}\only<6->{\ind}, so is~$X \times Y$. \\
-        \hil{Higman:}  \tabto{1.7cm} If~$X$ is well\only<4-5>{\sinf}\only<6->{\ind}, so is~$\mathrm{List}\,X$. \\
-        \hil{Kruskal:} \tabto{1.7cm} If~$X$ is well\only<4-5>{\sinf}\only<6->{\ind}, so is~$\mathrm{Tree}\,X$.
-      \end{block}
+    \begin{column}{0.2\textwidth}
+      \centering\footnotesize
+      \includegraphics[height=2cm]{turing-machine} \\ \mbox{numerical content}
     \end{column}
-  \end{columns}
-  \pause
-  \pause
-  \pause
-  \pause
-
-  \textbf{Def.} A quasi-order~$X$ is \hil{well\ind} iff~$G\,[\,]$, where $G$ is
-  the following inductively defined predicate on \hil{finite lists}.~\formalized{}
-  (In presence of \bad{bar induction}, $\text{well}\ind \Leftrightarrow
-  \text{well}\sinf$.)
-  \[
-    \small
-    \infer{\mathsf{now}\,p : G\,\sigma}{p : \mathsf{Good}\,\sigma}
-    \qquad
-    \infer{\mathsf{later}\,f : G\,\sigma}{f : (x : X) \to G\,(\sigma ::^r x)}
-  \]
-  \vspace*{-1.3em}
-
-  \only<8>{
-    \begin{columns}[c]
-      \begin{column}{0.01\textwidth}
-	\includegraphics[height=2.4em]{question-mark}
+    \begin{column}{0.4\textheight}
+      \centering\footnotesize
+      \includegraphics[height=2cm]{3-adic-numbers} \\
+      \mbox{a fractal} \mbox{without points}
     \end{column}
-      \begin{column}{0.9\textwidth}
-      Is there a procedure for reinterpreting \hil{classical proofs}
-      regarding well\sinf{} as \newline \hil{blueprints for constructive proofs} regarding well\ind?
-      \end{column}
-    \end{columns}
-  }
-
-  \only<9>{
-    \vspace*{-0.0em}
-    \hil{Central insight:} A quasi-order $X$ is well\ind iff $\hil{$\necessary$}\,\forall\alpha : \NN \to
-    X\_ \exists i < j\_ \alpha\,i \leq \alpha\,j$.
-  }
-\end{frame}}
-
-\begin{frame}{Missing functions in the type of all functions?}
-  \begin{columns}[t]
-    \begin{column}{0.57\textwidth}
-      \begin{block}{Behold: A transfinite tool \ldots\phantom{g}}
-        \justifying
-        \textbf{Lemma.} \badbox{\textsc{lem}} Let~$X$ be well\sinf.
-        Let~$\alpha : \NN \to X$. Then there is an increasing
-        subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
-      \end{block}
-
-      \small
-      \emph{Proof.}
-      \begin{changemargin}{-0.8em}{0em}\begin{enumerate}\justifying
-      \item The type
-      $I \defeq \sum_{i \? \NN} \neg\sum_{j \? \NN} i < j \times \alpha\,i \leq
-      \alpha\,j$
-      is not in bijection with~$\NN$, as else the~$I$-extracted subsequence
-      of~$\alpha$ would not be good.\smallskip
-      \item By~\badbox{\textsc{lem}}, the type~$I$ is finite.\smallskip
-      \item Any index~$i_0$ larger than all the indices in~$I$ is a suitable
-      starting point for an increasing subsequence. \qed
-      %  \item Let $\mathsf{Sad}\,i \defeq
-      %  \neg \mathsf{Happy}\,i$ be the proposition
-      %  expressing that~$i$ does not appear as the first component
-      %  of a good pair, where
-      %  $\mathsf{Happy}\,i \defeq \sum_{j \? \NN} i < j \times \alpha\,i \leq
-      %  \alpha\,j$.
-      %  \item If~$\mathsf{Sad}$ is anonymously dense in~$\NN$, i.e. $\forall a
-      %  \? \NN\_ \neg\neg\exists b\?\NN\_ b \geq a \times \mathsf{Sad}\,b$, then
-      %  \badbox{\textsc{lem}} and \badbox{\textsc{dc}} provide us with a
-      %  subsequence~$\alpha\,k_0 \leq \alpha\,k_1 \leq \ldots$
-      %  with~$\mathsf{Sad}\,k_0, \mathsf{Sad}\,k_1, \ldots$. This is a contradiction to wellness.
-      %  \item
-      %  Hence there is~\bad{not~not} an index~$a \? \NN$ such that~$\mathsf{Sad}$ is
-      %  strictly bounded by~$a$; hence every index~$b \geq a$ does~\bad{not not} have
-      %  a partner: $\forall b\geq a\_ \neg\neg\exists c>b\_
-      %  \alpha\,b \leq \alpha\,c$. Thus~\bad{\textsc{lem}} and \bad{\textsc{dc}}
-      %  provide us with the desired subsequence. \qed
-      \end{enumerate}\end{changemargin}
-    \end{column}
-
-    \begin{column}{0.43\textwidth}
-      \begin{block}{\ldots{} implying a concrete consequence}
-        \textbf{Cor.} \badbox{\textsc{lem}}
-        Let~$X$ and~$Y$ be well\sinf. Then~$X \times Y$ is
-        well\sinf.\phantom{g}
-      \end{block}
-
-      \small
-      \emph{Proof.}
-      \begin{changemargin}{-0.8em}{0em}\begin{enumerate}\justifying
-      \item Let a sequence~$\gamma : \NN \to X \times Y$ be
-      given. Write~$\gamma\,k = (\alpha\,k,\beta\,k)$.\smallskip
-      \item By the lemma, there is an
-      increasing subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.\smallskip
-      \item Because~$Y$
-      is well, there are indices~$n < m$ such that~$\beta\,i_n \leq
-      \beta\,i_m$.\smallskip
-      \item As
-      also~$\alpha\,i_n \leq \alpha\,i_m$, the sequence~$\gamma$ is good. \qed
-      \end{enumerate}\end{changemargin}
+    \begin{column}{0.45\textheight}
+      \centering\footnotesize
+      \includegraphics[height=2cm]{cracks} \mbox{foundational crisis}
     \end{column}
@@
 
-  \dnote{\emph{We cannot trust~\bad{\textsc{lem}}-provided sequences to be
-  available in the type~$\NN \to X$. Similarly with~\bad{\textsc{dc}}.}}
-\end{frame}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{0.29cm}\hspace*{-1cm}\includegraphics[height=\paperheight]{tower-cropped}\end{minipage}}
-\begin{frame}{Where do many cherished inductive definitions come from?}
-  \begin{changemargin}{8.0em}{-0.5em}
-    In mathematics, we routinely enlarge structures:
-    \begin{itemize}
-      \small
-      \item Pass from~$\mathbb{Q}$ to~$\mathbb{R}$, to embrace irrationals.
-      \item Pass from~$\mathbb{R}$ to~$\mathbb{C}$, to obtain $\sqrt{-1}$.
-      \item Pass from~$\mathbb{C}$ to~$\mathbb{C}[X]$, to obtain a ``generic number''.
-    \end{itemize}
-
-    In set and type theory, we can also enlarge \hil{the mathematical universe}:
-    \begin{itemize}
-      \small
-      \item Force a \hil{generic sequence} $\NN \to X$. \formalized
-      \item Force a \hil{generic enumeration} $\NN \twoheadrightarrow X$
-            (even if $X$ is uncountable). \formalized
-      \item Force a \hil{generic prime ideal} of a given ring. \formalized
-    \end{itemize}
-    \pause
-
-    Central observations of the multiversal yoga:
-    \begin{itemize}
-      \small
-      \item A quasi-order~$X$ is well\ind iff the generic sequence~$\NN \to X$
-      is good. \formalized
-      \item A relation is well-founded iff for the generic descending chain,
-      $\bot$. \formalized
-      \item A ring element is nilpotent iff it is contained in the generic
-      prime ideal. \formalized
-    \end{itemize}
-    \pause
-
-    The mystery of nongeometric sequents:
-    \begin{itemize}
-    \small
-      \item The generic ring is a field.
-      \item \ \\[-1.2em]\mbox{For the generic surjection $f : \NN \twoheadrightarrow \RR$,
-      $\neg\neg \exists n \? \NN\_ f(n) = \pi \wedge f(n+1) = e$. \formalized}
-    \end{itemize}
-  \end{changemargin}
-\end{frame}}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
-\begin{frame}{A modal language for harnessing the multiverse}
-  \textbf{Def.} A statement~$\varphi$ holds \ldots
-  \vspace*{-0.3em}
-  \begin{itemize}
-    \small
-    \item \hil{everywhere} \tabto{2.05cm}($\necessary\varphi$)\tabto{2.98cm}
-    iff it holds \hil{in every topos}
-    (over the current base).
-    \item \hil{somewhere} \tabto{2.05cm}($\possible\varphi$)\tabto{2.98cm} iff
-    it holds \hil{in some positive topos}.
-    \item \hil{proximally} \tabto{2.05cm}($\xpossible\varphi$)\tabto{2.98cm}
-    iff it holds \hil{in some positive overt topos}.
-  \end{itemize}
-
-  We then have:
-  \begin{enumerate}
-    \item $\mathsf{Well}\ind(X,{\le}) \Longleftrightarrow
-    \necessary\mathsf{Well}\sinf(X,{\le})$. \formalized
-    \item $(\possible\varphi) \Longleftrightarrow \varphi$,
-    if $\varphi$ is a geometric implication (``$\forall\ldots\forall (\%
-    \Rightarrow \%)$'', with no~$\forall$ nor~$\Rightarrow$
-    in~$\%$).~\formalized
-    \item $(\xpossible\varphi) \Longleftrightarrow \varphi$,
-    if~$\varphi$ is a bounded first-order statement. \formalized
-    \item For every inhabited type~$X$, $\xpossible\necessary
-    \mathsf{Countable}(X)$, where~$\mathsf{Countable}(X) \defeqv (\exists f : \NN
-    \to X\_ \mathsf{Surjective}(f))$. \formalized
-
-    So being countable is a \hil{button}.
-    \item $\possible \badbox{\textsc{lem}}$.
-    (Baby Barr / Friedman's trick / nontrivial exit continuation)
-
-    In fact, \badbox{\textsc{lem}} is a \hil{switch}:
-    $\necessary\bigl((\possible\badbox{\textsc{lem}}) \wedge
-    (\possible\neg\badbox{\textsc{lem}})\bigr)$
-
-    \item $\badbox{\textsc{zorn}} \Rightarrow
-    \necessary\possible\badbox{\textsc{ac}}$. (Great Barr)
-  \end{enumerate}
-\end{frame}}
-
-\begin{frame}{The plain combinatorics of toposes}
-  \begin{enumerate}\justifying
-    \item Realize a generic gadget as some kind of limit of \hil{approximations}
-    from the base universe. For the generic function $f_0 : \NN \to X$:
-
-    \AgdaHide{
-      \begin{code}
-open import Data.List
-open import Agda.Primitive using (Set)
-
-Prop = Set
-postulate
-  X : Set
-  _∈_ : X → List X → Prop
-      \end{code}
-    }
-
-    {\small\begin{code}
-L : Set
-L = List X
-    \end{code}}
-
-    \pause
-    \item \hil{Reinterpret}, in a mechanical fashion, assertions about the generic
-    gadget as assertions about its approximations~(\formalized).
-
-    {\small\begin{itemize}
-    \item We have the \hil{stage-dependent proposition} ``$f_0\,n = x$'',
-    a certain function $L \to \mathsf{Prop}$: $\lambda \sigma.\
-    (\mathsf{lookupMaybe}\,\sigma\,n = \mathsf{just}\,x)$\end{itemize}}
+  \medskip
+  \medskip
+  \medskip
   \medskip
 
-    \pause
-    \item Be prepared to evolve approximations (\formalized).
-
-    {\small\begin{code}
-data ∇ (P : L → Prop) : L → Prop where
-  now   : {σ : L} → P σ → ∇ P σ
-  later : {σ : L} → ((x : X) → ∇ P (σ ∷ʳ x)) → ∇ P σ
-    \end{code}
-
-    \begin{itemize}
-    \item
-    For a stage-dependent proposition~$P : L \to \mathsf{Prop}$, $\nabla\,P\,\sigma$ expresses
-    that no matter how~$\sigma$ will evolve over a time to a better
-    approximation~$\tau$, eventually~$P\,\tau$ will hold.
-  % later₂ : (a : X) → ((τ : X*) → a ∈ σ ⧺ τ → ∇ P (σ ⧺ τ)) → ∇ P σ
-    \item That~$f_0$ is defined on input~$n$ can be expressed
-    as~$\nabla\,P\,[\,]$ where~$P\,\sigma \defeqv (\mathsf{length}\,\sigma > n)$.\end{itemize}}
-    \medskip
-    \pause
-
-    \item Crucially, this interpretation is \hil{sound} with respect to constructive
-    reasoning.
-  \end{enumerate}
-\end{frame}
-
-\begin{frame}{Formal metatheory}
-  \small
-  \good{\cmark{}} \emph{There are type-theoretic multiverses,} such as
-  \begin{itemize}
-    \item the collection of all $\mathrm{PSh}(\mathcal{C} \times \mathcal{B})$,
-    where~$\mathcal{B}$ ranges over cube categories and~$\mathcal{C}$ over
-    arbitrary small categories, and their corresponding sheaf models
-
-    {\scriptsize
-    Coquand. ``A survey of constructive presheaf models of univalence''. \emph{ACM SIGLOG News}, 5.3 (2018).
-    \par}
-  \end{itemize}
-  \medskip
-  \pause
-
-  \bad{\xmark} \emph{Accessing the multiverse from within intensional type theory is tricky:}
-  \begin{itemize}
-    \item Given a model of $\mathfrak{s}$CIC and a category~$\mathcal{C}$ in
-    it, we have a syntactic presheaf model of CIC.
-
-    {\scriptsize
-    Coquand, Jaber. ``A note on forcing and type theory''. \emph{Fundamenta Informaticae 100} (2010). \\
-    Jaber, Lewertowski, Pédrot, Sozeau, Tabareau. ``The definitional side of the forcing''. \emph{Proceedings of LICS '16} (2016). \\
-    Pédrot. ``Russian constructivism in a prefascist theory''. \emph{Proceedings of LICS '20} (2020).
-    \par}
-    \item Given a suitable lex modality, we have a syntactic sheaf model (model of modal types).
-
-    {\scriptsize
-    Coquand, Ruch, Sattler. ``Constructive sheaf models of type theory.'' \emph{Math. Struct. Comput. Sci.} 31.9 (2021). \\
-    Escardó, Xu. ``Sheaf models of type theory in type theory''. Unpublished (2016). \\
-    Quirin. ``Lawvere--Tierney sheafification in Homotopy Type Theory''. PhD thesis (2016).
-    \par}
-    \item (I believe) we have syntactic sheaf models in certain special cases,
-    when no coherence issues arise in defining the notion of presheaves.
-  \end{itemize}
-
-  Note: We can use~$\nabla$ even without a proper metatheoretic backing.
-\end{frame}
-
-{\renewcommand{\insertframeextra}{a}
-\begin{frame}{Increasing subsequences as convenient fictions}
-  \justifying
-  Let~$X$ be a quasi-order.
-  Let~$B : \mathsf{List}\,X \to \mathsf{Prop}$ be a monotone predicate.
-
-  \begin{block}{Classical blueprint}
-    \textbf{Thm.} \badbox{\textsc{lem}} If~$X$ is well\sinf{} and
-    if every increasing sequence~$\alpha : \NN \to X$ has a prefix validating~$B$,
-    then every sequence has a prefix validating~$B$.
-  \end{block}
-  \vspace*{-0.4em}
-  {\small\emph{Proof.} Let~$\alpha : \NN \to X$ be a sequence. By the lemma,
-  there is an increasing subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
-  By assumption, this subsequence has a prefix validating~$B$. This prefix is part
-  of a prefix of the original sequence~$\alpha$. Hence we conclude by
-  monotonicity. \qed\par}
-  \medskip
-
-  \begin{block}{Constructive reimagination}
-    \textbf{Thm.} If~$X$ is well\ind and if~$\nabla^\nnearrow\,B\,[\,]$,
-    then~$\nabla\,B\,[\,]$. \formalized
-  \end{block}
-  \vspace*{-0.4em}
-  {\small\emph{Proof.} Let~$\alpha : \NN \to X$ be a sequence (in an arbitrary topos).
-  \emph{Somewhere,} \textsc{lem} holds. \emph{There}~$X$
-  is still well\sinf, so that we have an increasing subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
-  By assumption, this subsequence has a finite prefix validating~$B$. This
-  prefix is part of a prefix of the original sequence~$\alpha$. Hence~$\alpha$
-  has a prefix validating~$B$ by monotonicity \emph{there}. So
-  \emph{somewhere} there is a finite prefix validating~$B$. Thus there
-  actually is a finite prefix validating~$B$. \qed\par}
-\end{frame}}
-
-\addtocounter{framenumber}{-1}
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
-\renewcommand{\insertframeextra}{b}
-\begin{frame}{Maximal ideals as convenient fictions}
-  \justifying
-  Let~$A$ be a commutative ring with unit.
-
-  \textbf{Thm.}
-  Let~$M$ be a surjective matrix with more rows than columns over~$A$. Then~$1 = 0$ in~$A$.
-  \medskip
-
-  \emph{Classical proof.}
-  \bad{Assume not.} Then there is~a \bad{maximal ideal} $\mmm$.
-  The matrix~$M$ is surjective over~$A/\mmm$. Since~$A/\mmm$ is a field, this
-  is a contradiction to basic linear algebra. \qed
-  \medskip
-
-  \only<1>{\medskip\par\centering\scalebox{0.9}{\centering\begin{tikzpicture}
-    \node (0) at (0,1) {$(0) = \{0\}$};
-    \node (1) at (0,5) {$(1) = \ZZ$};
-    \node (2) at (-2,4) {$(2)$};
-    \node [right of=2] (3) {$(3)$};
-    \node [below of=2] (4) {$(4)$};
-    \node [below of=2, xshift=0.7cm] (6) {$(6)$};
-    \node [right of=3] (5) {$(5)$};
-    \node [right of=5] (7) {$(7)$};
-    \node [right of=7] (7d) {$\ldots$\phantom{(}};
-    \node [right of=7d, xshift=3cm, yshift=-2cm] (max)
-    {\vbox{\small{\it maximal among the proper ideals} \\ \medskip \hspace*{-6.75em}\textbullet \quad $\neg(1 \in
-    \mmm)$ \\ \medskip \textbullet \quad $\neg\bigl(1 \in \mmm + (x)\bigr) \Rightarrow x \in \mmm$}};
-    \node [below of=4] (8) {$(8)$};
-    \node [right of=8, xshift=3cm] (8d) {$\ldots$};
-    \draw (0) -- (8);
-    \draw (0) -- (8d);
-    \draw (0) -- (6);
-    \draw (2) -- (1);
-    \draw (3) -- (1);
-    \draw (5) -- (1);
-    \draw (7) -- (1);
-    \draw (7d) -- (1);
-    \draw (4) -- (2);
-    \draw (8) -- (4);
-    \draw (6) -- (2);
-    \draw (6) -- (3);
-    \draw [mypurple!30, thick, shorten <=-2pt, shorten >=-2pt, ->] (max) to [out=120, in=-30] (7d);
-    \begin{pgfonlayer}{background}
-      \draw[decorate, very thick, draw=mypurple!30]
-        ($(2.south west) + (8pt, 0)$) arc(270:180:8pt) --
-        ($(2.north west) + (0, -8pt)$) arc(180:90:8pt) --
-        ($(7d.north east) + (-8pt, 0)$) arc(90:0:8pt) --
-        ($(7d.south east) + (0, 8pt)$) arc(0:-90:8pt) --
-        cycle;
-    \end{pgfonlayer}
-  \end{tikzpicture}\par}\par}
-  \pause
-
-  \emph{Multiversal constructive proof.}
-  We may work \emph{somewhere} where \good{\textsc{lem}} holds. So \good{assume not}.
-  \emph{Proximally,} there is a \good{maximal ideal}~$\mmm$~(\formalized).
-  The matrix~$M$ is still surjective \emph{there}, and also over~$A/\mmm$. Since~$A/\mmm$ is a field, this
-  is a contradiction to basic linear algebra. \qed
-  \medskip
-  \pause
-
-  \emph{Unrolled constructive proof (special case).}
-  Write~$M =
-  \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right)$. By surjectivity,
-  have~$u, v$ with
-  \vspace*{-0.6em}
-  \[
-    u \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}1\\0\end{smallmatrix}\right)
-    \quad\text{and}\quad
-    v \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}0\\1\end{smallmatrix}\right).
-  \]
-  Hence
-  $
-    1 = (vy) (ux) = (uy) (vx) = 0
-  $.
-  \qed
-\end{frame}}
-
-\backupstart
-
-\begin{frame}[plain]
-  \begin{columns}[c]
-    \begin{column}{0.45\textwidth}
-      \includegraphics[width=0.95\textwidth]{lets-play-agda}
+  \begin{columns}
+    \begin{column}{0.4\textheight}
+      \centering\footnotesize
+      \includegraphics[height=2cm]{multiverse} \mbox{traveling} \\
+      the \mbox{multiverse}
     \end{column}
-    \begin{column}{0.5\textwidth}
-      \includegraphics[width=0.8\textwidth]{ai-2025}
-
-      \vspace*{-1.5cm}\hspace*{2cm}
-      \includegraphics[width=0.8\textwidth]{cirm-2025}
-
-      \vspace*{1cm}
+    \begin{column}{0.4\textheight}
+      \centering\footnotesize
+      \includegraphics[height=2cm]{monad} monadic side effects
+    \end{column}
+    \begin{column}{0.4\textheight}
+      \centering\footnotesize
+      \includegraphics[height=2cm]{hen-2} \\ proof assistants
+    \end{column}
+    \begin{column}{0.4\textheight}
+      \centering\footnotesize
+      \includegraphics[height=2cm]{alien} \\ alien algorithms
     \end{column}
@@
 
-\begin{frame}
-  \emph{Backup slides.}
+\NewEnviron{mining}[1][0.6\textwidth]{
+  \includegraphics[height=2em,valign=c]{hammer-and-pick}\quad
+  \colorbox{mypurple!10}{\begin{minipage}{#1}
+    \BODY
+  \end{minipage}}
+}
+
+\begin{frame}{A glimpse of proof mining}
+  \textbf{Theorem.} For every natural number~$n$, there is a prime number
+  larger than~$n$.~\fixedhref{https://unimath.github.io/agda-unimath/elementary-number-theory.infinitude-of-primes.html}{\formalized}
+
+  \emph{Proof (Euclid).} Every prime factor of~$n! + 1$ will do. \qed
+  \pause
+
+  \begin{mining}
+    Let~$p_0, p_1, p_2, \ldots$ be the sequence of prime numbers.
+
+    \textbf{Scholium.} $p_{n+1} \leq p_n! + 1$.
+
+    \textbf{Scholium.} $p_n \leq 2^{2^n}$.
+  \end{mining}
+  \pause
+  \bigskip
+
+  \emph{Proof (Euler).} If there were only finitely many primes,
+  the identity
+  \[ \prod_p \frac{1}{1-1/p} = \sum_{n \geq 1} \frac{1}{n} \]
+  would imply that the harmonic series converges. \qed
+
+  \begin{mining}
+    \textbf{Scholium.} $p_n \leq \lceil e^{n+1-\gamma} \rceil$.
+  \end{mining}
 \end{frame}
 
-\section{Basics of forcing}
+% 1995: conjecture
+% 2003: solution
+% 2018: rate of convergence
+\begin{frame}{Proof mining in convex analysis}
+  \textbf{Theorem (zero displacement conjecture).}
+  Let~$H$ be a Hilbert space. \newline
+  Let~\mbox{$C_1, \ldots, C_N \subseteq H$} be nonempty
+  closed convex subsets with orthogonal projections~$P_{C_i}$. Let~$T = P_{C_N}
+  \circ \ldots \circ P_{C_1}$. Then for every~$x \in H$,
+  \[ \| T^{n+1}x - T^nx \| \xrightarrow{n \to \infty} 0. \]
 
-\begin{frame}{Ingredients for forcing}
-  \vspace*{-0.3em}
-  To construct a forcing extension, we require:
-  \begin{enumerate}
-    \item a base universe~$V$
-    \item a preorder~$L$ of \hil{forcing conditions} in~$V$\!,
-    pictured as \hil{finite approximations}
-
-    (\emph{convention:} $\tau \preccurlyeq \sigma$ means that~$\tau$ is a
-    better finite approximation than~$\sigma$)
-    \item a \hil{covering system} governing how finite approximations evolve to
-    better ones
-
-    (for each~$\sigma \in L$, a set~$\Cov(\sigma) \subseteq
-    P({\downarrow}\sigma)$, with a simulation condition)
-  \end{enumerate}
-  In the forcing extension~$V^\nabla$, there will then be a \hil{generic filter} (ideal
-  object).
+  \emph{Proof.} See
+  \fixedhref{https://www.ams.org/journals/proc/2003-131-01/S0002-9939-02-06528-0/S0002-9939-02-06528-0.pdf}{[Bauschke
+  2003]}, employing Minty's theorem, the Brézis--Haraux
+  theorem, Rockafellar's maximal monotonicity and sum theorems, strongly
+  nonexpansive mappings, conjugate functions, normal cones, \ldots \qed
   \pause
 
-  \vspace*{-1em}
-  \begin{columns}
-    \begin{column}{0.49\textwidth}
-      \small
-      \begin{block}{For the generic surjection~$\NN \twoheadrightarrow X$}
-        \justifying
-        Use \hil{finite lists}~$\sigma \in X^*$ as forcing conditions,
-        where $\tau \preccurlyeq \sigma$ iff~$\sigma$ is an initial segment of~$\tau$,
-        and be prepared to grow~$\sigma$ to \ldots
-        \footnotesize
-        \begin{enumerate}
-          \item[(a)] one of~$\{ \sigma x \,|\, x \in X \}$, to make~$\sigma$ more defined
-          \item[(b)] one of~$\{ \sigma \tau \,|\, \tau \in X^*, a \in
-          \sigma\tau \}$, for any~$a \in X$, to make~$\sigma$ more surjective
-        \end{enumerate}
-        \vspace*{-0.4em}
-      \end{block}
-    \end{column}
+  \begin{mining}[0.85\textwidth]
+    \textbf{Scholium
+    \fixedhref{https://www2.mathematik.tu-darmstadt.de/~kohlenbach/inconsistentfeasibility.pdf}{[Kohlenbach
+    2018]}.} In this situation, let~$b$ be an upper bound on the norm of~$x$ and let~$K$ be an
+    upper bound for the norm of~$N$ arbitrary points $c_i \in C_i$. Then
+    \[ \forall \varepsilon > 0\_
+      \forall n \geq \phi(\varepsilon,N,b,K)\_
+      \|T^{n+1}x - T^nx\| < \varepsilon, \]
+    where~$\phi(\varepsilon,N,b,K)$ is given by a certain explicit formula.
+  \end{mining}
+\end{frame}
+
+\begin{frame}{Proof mining in approximation theory}
+  Let~$n \in \NN$. Let~$P_n$ be the space of polynomials of degree at most~$n$. \\
+  Let~$f \in C[0,1]$ be a continuous function.
+
+  \textbf{Theorem.} There is a \hil{unique} best~$L_1$-approximation of~$f$ in~$P_n$.
   \pause
 
-    \begin{column}{0.45\textwidth}
-      \small
-      \begin{block}{For the generic prime ideal of a ring~$A$}
-        \justifying
-        Use \hil{f.g.\@ ideals} as forcing conditions, where $\bbb \preccurlyeq
-        \aaa$ iff~$\bbb \supseteq \aaa$, and be prepared to grow~$\aaa$ to \ldots
-        \footnotesize
-        \begin{enumerate}
-          \item[(a)] one of~$\emptyset$, if~$1 \in \aaa$, to make~$\aaa$ more proper
-          \item[(b)] one of~$\{ \aaa+(x), \aaa+(y) \}$, if~$xy \in \aaa$, to
-          make~$\aaa$ more prime
-        \end{enumerate}
-        \vspace*{-0.4em}
-      \end{block}
-    \end{column}
-  \end{columns}
+  Let~$\omega$ be a \hil{modulus of uniform continuity} for~$f$,
+  i.e. a function~$\mathbb{R}^+ \to \mathbb{R}^+$ such that
+  \[ \forall x,\tilde x \in [0,1]\_ \forall \varepsilon > 0\_
+    \bigl(|x-\tilde x| < \omega(\varepsilon) \Longrightarrow |f(x)-f(\tilde x)| <
+    \varepsilon\bigr). \]
+
+  \begin{mining}[0.85\textwidth]
+    \textbf{Scholium
+    \fixedhref{https://www2.mathematik.tu-darmstadt.de/~kohlenbach/Keele.pdf}{[Kohlenbach
+    1990]}.} $\forall \varepsilon > 0\_ \forall p_1,p_2 \in P_n\_$
+    \[
+      \bigwedge_{i=1}^2 \bigl(\|f - p_i\|_1 - \operatorname{dist}_1(f, P_n) <
+      \phi(\omega,n,\varepsilon)\bigr) \Longrightarrow \|p_1 - p_2\|_1 \leq
+      \varepsilon, \]
+    where~$\phi(\omega,n,\varepsilon)$ is given by a certain explicit formula.
+  \end{mining}
 \end{frame}
-
-{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{-1cm}\includegraphics[width=\paperwidth]{forest-light-colored}\end{minipage}}
-\begin{frame}{The eventually monad}
-  Let~$L$ be a forcing notion.
-
-  Let~$P$ be a monotone predicate on~$L$
-  (if $\tau \preccurlyeq \sigma$, then $P\sigma \Rightarrow P\tau$). \\
-  For instance, in the case~$L = X^*$:
-  \begin{itemize}
-    \item $\mathsf{Repeats}\, x_0\ldots x_{n-1} \defeqv \exists i\_ \exists j\_ i < j \wedge x_i = x_j$
-    \item $\mathsf{Good}\, \,\,\,\,\,\,x_0\ldots x_{n-1} \defeqv \exists i\_ \exists j\_ i < j \wedge x_i \leq x_j$
-    \quad (for some preorder~$\leq$ on~$X$)
-  \end{itemize}
-  \pause
-
-  We then define~``\hil{$\nabla\,P\,\sigma$}'' (``$P$ bars~$\sigma$'') inductively by the following clauses:
-  \begin{enumerate}
-    \item If~$P\sigma$, then~$\nabla\,P\,\sigma$.
-    \item If~$\nabla\,P\,\tau$ for all~$\tau \in R$, where~$R$ is some covering
-    of~$\sigma$, then~$\nabla\,P\,\sigma$.
-  \end{enumerate}
-  So~$\nabla\,P\,\sigma$ expresses in a \hil{direct inductive fashion}:
-  \[ \text{``No matter
-  how~$\sigma$ evolves to a better approximation~$\tau$, eventually~$P\,\tau$
-  will hold.''} \]
-
-  \pause
-  We use quantifier-like notation: ``$\nabla(\tau \preccurlyeq \sigma)\_
-  P\tau$'' means ``$\nabla\,P\,\sigma$''.
-\end{frame}}
-% BOARD:
-% - examples for P | σ
-% - abuse of notation
-
-\begin{frame}{Proof translations}
-  \textbf{Thm.} Every~\textsc{iqc}-proof remains correct, with at most a
-  polynomial increase in length, if throughout we
-  replace
-  \[\begin{array}{rcl@{\quad\text{where}\quad}rcl}
-    \exists & \leadsto & \exists^\mathrm{cl},
-    & \exists^\mathrm{cl} &\defeqv& \neg\neg\exists, \\
-    \vee & \leadsto & \vee^\mathrm{cl},
-    & \alpha \vee^\mathrm{cl} \beta &\defeqv& \neg\neg(\alpha \vee \beta), \\
-    = & \leadsto & =^\mathrm{cl},
-    & s =^\mathrm{cl} t &\defeqv& \neg\neg(s = t).
-  \end{array} \]
-  \pause
-
-  \begin{columns}[c]
-    \begin{column}{0.01\textwidth}
-      \includegraphics[height=2.4em]{sheafification-man-2}
-    \end{column}
-    \quad
-    \begin{column}{0.9\textwidth}
-      \hil{When we say:}\ \ some statement ``holds in~$V^{\neg\neg}$'', \\
-      \makesamewidth[l]{\hil{When we say:}}{\hil{we mean:}}\ \ its translation holds in~$V$.
-    \end{column}
-  \end{columns}
-  \bigskip
-
-  Similarly for arbitrary forcing extensions~$V^\nabla$, ``just with~$\nabla$
-  instead of~$\neg\neg$''.
-  \bigskip
-
-  \pause
-  \textbf{Ex.} As~$\neg\neg(\varphi \vee \neg\varphi)$ is a theorem
-  of~\textsc{iqc}, the law of excluded middle holds in~$V^{\neg\neg}$.
-\end{frame}
-
-\newcommand{\defeqvi}{\quad iff\quad}
-\begin{frame}{The $\nabla$-translation}
-  \small
-  \only<1>{For bounded first-order formulas over the (large) first-order signature which has
-  \begin{enumerate}
-    \scriptsize
-    \item one sort~$\underline{X}$ for each set~$X$ in the base universe,
-    \\[-1.2em]
-    \item one~$n$-ary function symbol~$\underline{f} : \underline{X_1} \times
-    \cdots \times \underline{X_n} \to \underline{Y}$ for each map~$f : X_1 \times
-    \cdots \times X_n \to Y$,
-    \\[-1.2em]
-    \item one~$n$-ary relation symbol~$\underline{R} \hookrightarrow
-    \underline{X_1} \times \cdots \times \underline{X_n}$ for each relation~$R
-    \subseteq X_1 \times \cdots \times X_n$, and
-    \\[-1.2em]
-    \item an additional unary relation symbol~$G \hookrightarrow \underline{L}$
-    (for the \emph{generic filter} of~$L$),
-  \end{enumerate}
-  we recursively define:}
-  \scriptsize
-  \only<2->{\vspace*{-0.4em}}
-  \begin{tabbing}
-    \quad \= $\sigma \forces \forces \forall(x\?\underline{X})\_ \varphi$ \=
-    \defeqvi $\textcolor{gray}{\forall(\tau \preccurlyeq \sigma)\_}\
-    \forall(x_0 \in X)\_ \tau \forces \varphi[\underline{x_0}/x]$.\qquad\quad \=
-    $\sigma \forces \exists(x\?\underline{X})\_ \varphi$
-    \= $\sigma \forces \underline{R}(\underline{s_1},\ldots,\underline{s_n})$ \= \defeqvi $s = t$. \= \kill
-
-    \> $\sigma \forces s = t$
-    \> \defeqvi $\nabla \sigma\_ \llbracket s \rrbracket = \llbracket t \rrbracket$.
-    \> $\sigma \forces \underline{R}(s_1,\ldots,s_n)$
-    \> \defeqvi $\nabla\sigma\_ R(\llbracket s_1 \rrbracket,\ldots,\llbracket s_n \rrbracket)$. \\[0.3em]
-
-    \> $\sigma \forces \varphi \Rightarrow \psi$
-    \> \defeqvi $\textcolor{gray}{\forall(\tau \preccurlyeq \sigma)\_}\ (\tau \forces \varphi) \Rightarrow
-    (\tau \forces \psi)$.
-    \> $\sigma \forces G\tau$
-    \> \defeqvi $\nabla\sigma\_ \sigma \preccurlyeq \llbracket\tau\rrbracket$. \\[0.3em]
-
-    \> $\sigma \forces \top$ \> \defeqvi $\top$.
-    \> $\sigma \forces \bot$ \> \defeqvi $\hil{$\nabla\sigma\_$}\ \bot$ \\[0.3em]
-
-    \> $\sigma \forces \varphi \wedge \psi$
-    \> \defeqvi $(\sigma \forces \varphi) \wedge (\sigma \forces \psi)$.
-    \> $\sigma \forces \varphi \vee \psi$
-    \> \defeqvi $\hil{$\nabla\sigma\_$}\ (\sigma \forces \varphi) \vee (\sigma \forces \psi)$. \\[0.3em]
-
-    \> $\sigma \forces \forall(x\?\underline{X})\_ \varphi$
-    \> \defeqvi $\textcolor{gray}{\forall(\tau \preccurlyeq \sigma)\_}\ \forall(x_0 \in X)\_ \tau \forces
-    \varphi[\underline{x_0}/x]$.
-    \> $\sigma \forces \exists(x\?\underline{X})\_ \varphi$
-    \> \defeqvi $\hil{$\nabla\sigma\_$}\ \exists(x_0 \in X)\_ \sigma \forces \varphi[\underline{x_0}/x]$.
-  \end{tabbing}
-  \small
-  \only<1>{Finally, we say that~$\varphi$ ``holds in~$V^\nabla$'' iff for all~$\sigma
-  \in L$, $\sigma \forces \varphi$.}
-
-  \footnotesize
-  \begin{tabular}{@{}lp{0.27\textwidth}p{0.48\textwidth}@{}}
-    \toprule
-    forcing notion & statement about~$V^\nabla$ & external meaning \\
-    \midrule
-    surjection $\NN \twoheadrightarrow X$ &
-    ``the gen.\@ surj.\@ is surjective'' &
-    $\forall(\sigma{\in}X^*)\_ \forall(a{\in}X)\_ \nabla(\tau{\preccurlyeq}\sigma)\_ \exists(n{\in}\NN)\_ \tau[n] = a$. \\[0.4em]
-    \pause
-
-    map $\NN \to X$ &
-    ``the gen.\@ sequence is good'' &
-    $\mathsf{Good} \mid [\,]$. \\[0.4em]
-
-    frame of opens &
-    ``every complex number has a square root'' &
-    For every open~$U \subseteq X$ and every cont.\@
-    function $f : U \to \CC$, there is an open covering $U = \bigcup_i U_i$ such
-    that for each index~$i$, there is a cont.\@ function $g : U_i \to \CC$
-    such that~$g^2 = f$. \\[4.3em]
-
-    big Zariski &
-    ``$x \neq 0 \Rightarrow \text{$x$ inv.}$'' &
-    If the only f.p.\@ $k$-algebra in which~$x = 0$ is the zero algebra,
-    then~$x$ is invertible in~$k$.\\[1.5em]
-    \pause
-
-    little Zariski &
-    ``every f.g. vector space does \emph{not~not} have a basis'' &
-    \makesamewidth[l]{}{\phantom{x}}\hil{Grothendieck's generic freeness lemma}
-  \end{tabular}
-\end{frame}
-
-\begin{frame}{Outlook}
-  \begin{block}{Passing to and from extensions}
-    \justifying\small
-    \textbf{Thm.} Let~$\varphi$ be a \hil{bounded first-order formula} not
-    mentioning~$G$. In each of the following situations, we have that
-    $\varphi$ holds in~$V^\nabla$ iff~$\varphi$
-    holds in $V$:
-
-    \vspace*{-0.5em}
-    \begin{enumerate}
-      \item $L$ and all coverings are inhabited (proximality). \\[-1em]
-      \item $L$ contains a top element, every covering of the
-      top element is inhabited, and~$\varphi$ is a coherent implication
-      (positivity).
-    \end{enumerate}
-    \vspace*{-0.8em}
-  \end{block}
-
-  \vspace*{-1.5em}
-  \begin{columns}
-    \begin{column}{0.46\textwidth}
-      \begin{block}{The mystery of nongeometric sequents}
-        \justifying
-        The \hil{generic ideal} of a ring is maximal:
-        \vspace*{-1em}
-        \[ (x \in \aaa \Rightarrow 1 \in \aaa) \Longrightarrow 1 \in \aaa + (x). \]
-
-        The \hil{generic ring} is a field:
-        \vspace*{-0.7em}
-        \[ (x = 0 \Rightarrow 1 = 0) \Longrightarrow (\exists y\_ xy = 1). \]
-        \vspace*{-1.4em}
-      \end{block}
-
-    \end{column}
-
-    \begin{column}{0.50\textwidth}
-      \begin{block}{Traveling the multiverse \ldots}
-        \textsc{lem} is a \hil{switch} and \hil{holds positively};
-        being countable is a \hil{button}.
-        \medskip
-
-        Every instance of \textsc{dc} \hil{holds proximally}.
-        \medskip
-
-        A geometric implication is provable iff it holds \hil{everywhere}.
-        \vspace*{-0.2em}
-      \end{block}
-      \vspace*{-0.3em}
-      \hfill\footnotesize\ldots{} upwards, but always keeping ties to the
-      base.{\ }
-    \end{column}
-  \end{columns}
-\end{frame}
-
-\begin{frame}{More on forcing notions}
-  \small
-  \textbf{Def.} A \hil{forcing notion} consists of
-  a preorder~$L$ of \hil{forcing conditions}, and
-  for every~$\sigma \in L$, a set~$\Cov(\sigma) \subseteq
-  P({\downarrow}\sigma)$ of \hil{coverings} of~$\sigma$
-  such that: If~$\tau \preccurlyeq \sigma$ and~$R \in \Cov(\sigma)$, there
-  should be a covering~$S \in \Cov(\tau)$ such that~$S \subseteq
-  {\downarrow}R$.
-  \bigskip
-
-  {\centering\footnotesize\begin{tabular}{llll}
-    \toprule
-    & preorder~$L$ & coverings of an element~$\sigma \in L$ & filters of~$L$ \\
-    \midrule
-    \normalnumber{1} & $X^*$ & $\{ \sigma x \,|\, x \in X \}$ & maps~$\NN \to X$ \\
-    \normalnumber{2} & $X^*$ & $\{ \sigma x \,|\, x \in X \}$,\ \ $\{ \sigma\tau \,|\, \tau \in X^*, a \in \sigma\tau \}$ for each~$a \in X$ & surjections~$\NN \twoheadrightarrow X$ \\
-    \normalnumber{3} & f.g. ideals & --- & ideals \\
-    \normalnumber{4} & f.g. ideals & $\{ \sigma+(a), \sigma+(b) \}$ for each~$ab \in \sigma$,\ \ $\{\}$ if~$1 \in \sigma$ & prime ideals \\
-    \normalnumber{5} & opens & $\mathcal{U}$ such that~$\sigma = \bigcup \mathcal{U}$ & points \\
-    \normalnumber{6} & $\{\star\}$ & $\{ \star \,|\, \varphi \} \cup \{ \star \,|\, \neg\varphi \}$ &
-    witnesses of~\textsc{lem}
-    \\
-    \bottomrule
-  \end{tabular}\par}
-  \bigskip
-
-  \textbf{Def.} A \emph{filter} of a forcing notion~$(L,\mathrm{Cov})$
-  is a subset~$F \subseteq L$ such that
-  \vspace*{-0.4em}
-  \begin{enumerate}
-    \scriptsize
-    \item $F$ is upward-closed: if~$\tau \preccurlyeq \sigma$ and if~$\tau \in F$, then~$\sigma \in F$; \\[-3.0em]
-    \item $F$ is downward-directed: $F$ is inhabited, and if~$\alpha,\beta \in F$,
-    then there is a common refinement~$\sigma \preccurlyeq \alpha,\beta$ such
-    that~$\sigma \in F$; and \\[-2.0em]
-    \item $F$ splits the covering system: if~$\sigma \in F$ and~$R \in
-    \Cov(\sigma)$, then~$\tau \in F$ for some~$\tau \in R$.
-  \end{enumerate}
-\end{frame}
-
-\backupend
 
@@
 
-0. A play in classical logic
+Beyond verified correctness and collaborative proof engineering: Proofs as programs?
 
-1. Hilbert's program
-   - basics about wqo's
-   - subsequence lemma
-   - Dickson
-   - inductive reimagination
-   - questions: where from? why work? how much stronger?
+A century ago, Hilbert called upon the mathematical community to explain how to
+extract concrete numerical content from abstract proofs involving transfinite
+methods. For instance, from a qualitative proof of the existence of a limit, we
+might hope to extract a quantitative bound on the rate of convergence. Proof
+assistants based on modern type theory offer the tantalizing prospect of
+facilitating such extraction by running formalized proofs as programs—adding
+value to mechanization beyond verified correctness and collaborative proof
+engineering. In the talk, we will explore a case study from commutative algebra
+on this kind of proof mining, obtained in joint work with Peter Schuster.
 
-2. Where do our most cherished definitions come from?
-   - a familiar business: enlarging mathematical structures
-   - adjoining generic function, surjection, prime ideal, ...
-   - observations:
-     - well iff generic sequence is good iff everywhere, every sequence is good
-       (in particular, every multivalued and every up-to-¬¬-defined sequence is good)
-     - wellfounded iff for the generic decreasing sequence, ⊥ iff ...
-   - fun facts about generic gadgets
+- Start:
 
-3. Modal language for harnessing the multiverse
-   - definition
-   - examples
-   - implicational wellness
-   - dialogue
+  - classical logic fairy tale?
+  - overview as in Swansea talk
 
-4. Implementation
-   - idea: realize generic gadget as some kind of limit of approximations from
-     the base
-   - reinterpret, in a mechanic fashion, assertions about generic objects as
-     assertions about approximations
-     - ex.: "the generic function is defined on input n" as "no matter how the
-       empty list evolves over a time to a better approximation xs, eventually
-       xs will have length at least succ n"
-   - crucially, this interpretation is sound with respect to intuitionistic
-     reasoning
+- A glimpse of proof mining
+  - Euclid proof
+  - Gauß proof
+  - Successes of proof mining in diverse fields
+  - Example: running Euclid
 
-5. Let's play Agda, CIRM, AI
+- A case study in algebra
+  - Textbook proof
+  - Issue: Zorn's lemma
+  - Issue: LEM
 
----
+- Motto: Running proofs as effectful programs
 
-Towards topological type theory for decrypting transfinite methods in classical mathematics
+- Brunerie?
+- Hilbert & certificates
+- Extracted algorithms are nontrivial, cf. implicational wqo
 
-In constructive mathematics, an ongoing challenge is to extracting
-computational content from infinitary arguments in classical
-mathematics—especially those that yield concrete results via abstract tools
-such as minimal bad sequences or maximal ideals. Today, a wide range of such
-techniques exists, constituting a late partial fulfillment of Hilbert's
-program.
+- Applications of AI:
+  - automatic application of metalogical extraction techniques
+  - support for refactoring proofs (for instance for inserting ¬¬'s)
+  - resulting constructive proofs/algorithms can look "alien" (= artificial intelligence)
 
-This talk presents work in progress on porting the modal operators "everywhere"
-and "somewhere", as developed in set theory by Joel David Hamkins, Victoria
-Gitman and their collaborators, into type theory. This modal enrichment
-provides a uniform language for expressing several established
-constructivization techniques, and is at the core of a relatively recent new
-technique that offers insights on the origins of some of our most cherished
-inductive definitions. These developments are closely connected to recent
-advances in type-theoretic presheaf and sheaf models, dialogue
-continuity, synthetic algebraic geometry and well-quasi-order theory.
-
----
-
-Which functions are we excluding when we form the type of all functions between
-two types? Does every field have an algebraic closure?
-
-Inspired by the modal approach to the set-theoretic multiverse of Joel David
-Hamkins, Victoria Gitman and their collaborators, we aim to introduce the modal
-operators "everywhere" and "somewhere" to homotopy type theory, proposing a
-multiversal perspective on these motivating questions.
-
----
-
-Well quasi-orders are a combinatorial notion with featureful applications in
-graph theory, termination checking, commutative algebra and other subjects.
-However, their classical definition, though concise and elegant, poses
-significant challenges in constructive mathematics and frameworks that lack
-function sets. In response, several constructive substitutes have been
-developed, most recently an implicational definition by Stefano Berardi,
-Gabriele Buriola and Peter Schuster. In this talk, we investigate how a modal
-approach can reinterpret classical proofs involving transfinite methods as
-blueprints for constructive proofs grounded in these alternatives, thereby
-combining the best of both worlds: Short and abstract proofs, but with
-constructive content. We also indicate how the modal approach allows us to
-compare the implicational definition with an earlier inductive one.
-
----
-
-Combinatorics and commutative algebra abound with situations where we prove
-quite concrete results by quite abstract transfinite methods, such as minimal
-bad sequences or maximal ideals. Amazingly, such infinitary arguments can often
-be understood as blueprints for quite explicit computations—as called for by
-Hilbert’s programme. In this talk, we will travel the modal toposophic
-multiverse to facilitate this kind of mining abstract proofs for refined
-quantitative results. We will use a celebrated theorem from order theory about
-sequences N→N that everybody can relate with as a running example.
-
----
-
-Where do some of our most cherished inductive definitions come from? Which
-functions are we excluding when we form the type of all functions between two
-types? Does every field have an algebraic closure?
-
-Inspired by the modal approach to the set-theoretic multiverse of Joel David
-Hamkins, Victoria Gitman and their collaborators, we aim to introduce the modal
-operators "everywhere" and "somewhere" to homotopy type theory, proposing a
-multiversal perspective on these motivating questions.
-
-Unlike the set-theoretic role model, we focus less on exploring the range of
-foundational possibility and more on concrete applications in constructive
-mathematics, with the goal of porting results of classical mathematics to
-homotopy type theory. It will turn out that every field has an algebraic
-closure somewhere; that a transitive relation is well-founded iff nowhere there
-is an infinite descending chain; and that somewhere, the law of excluded middle
-holds.
-
-This is ongoing joint work with Alexander Oldenziel and connected to the recent
-advances with type-theoretic presheaf models and sheaf models.
+- Agda icons
+- convenient fictions

### 2025-08-25 slides-ai2025.tex
More work on the slides

@@
     2018]}.} In this situation, let~$b$ be an upper bound on the norm of~$x$ and let~$K$ be an
-    upper bound for the norm of~$N$ arbitrary points $c_i \in C_i$. Then
+    upper bound on the norm of~$N$ arbitrary points $c_i \in C_i$. Then
     \[ \forall \varepsilon > 0\_
@@
 
+\begin{frame}{Backed by logical metatheorems?}
+  \hil{Metatheorems} have been and are being developed which \ldots
+  \begin{enumerate}
+    \item guarantee the \hil{extractability} of suitable numerical information in principle
+    (bounds on norms, convergence rates, moduli of uniqueness, rates of
+    asymptotic regularity, \ldots),
+    \item describe an \hil{algorithm} for carrying out the extraction and
+    \item support \hil{modular treatments} of auxiliary lemmas,
+  \end{enumerate}
+  \bad{\textbf{provided}} the input proof is supplied in a
+  certain \hil{formal system}.
+
+  NB: The \hil{quality} of the extracted data depends on the
+  \hil{sophistication} of the \hil{logical~principles} used in the proof
+  (Heine--Borel, Bolzano--Weierstraß, \ldots).
+
+  \bad{\textbf{In practice (2025):}} Algorithms \bad{\textbf{not}} used, only followed
+  as \bad{\textbf{rough guidelines}}, combined with hand-rolled optimizations.
+\end{frame}
+
 \end{document}
@@
 - Start:
-
   - classical logic fairy tale?
@@
   - Euclid proof
-  - Gauß proof
+  - Euler proof
   - Successes of proof mining in diverse fields
@@
 - convenient fictions
+
+- formulation problem
+- infinite objects
+- proving = programming

### 2025-08-25 slides-ai2025.tex
Rosetta stone and test cases

@@
 
-\begin{frame}{A glimpse of proof mining}
+\NewEnviron{simpleblock}{
+  \colorbox{mypurple!10}{\begin{minipage}{\textwidth}
+    \BODY
+  \end{minipage}}
+}
+
+\renewcommand{\insertframeextra}{a}
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.3cm}\includegraphics[width=\paperwidth]{fr1-lighter}\end{minipage}}
+\begin{frame}{A first glimpse of proof mining}
   \textbf{Theorem.} For every natural number~$n$, there is a prime number
@@
   \end{mining}
-\end{frame}
+\end{frame}}
 
@@
 % 2018: rate of convergence
+\addtocounter{framenumber}{-1}
+\renewcommand{\insertframeextra}{b}
 \begin{frame}{Proof mining in convex analysis}
@@
 
+\addtocounter{framenumber}{-1}
+\renewcommand{\insertframeextra}{c}
 \begin{frame}{Proof mining in approximation theory}
@@
 \end{frame}
+\renewcommand{\insertframeextra}{}
 
@@
   certain \hil{formal system}.
+  \pause
 
@@
   (Heine--Borel, Bolzano--Weierstraß, \ldots).
+  \pause
 
@@
   as \bad{\textbf{rough guidelines}}, combined with hand-rolled optimizations.
+  \hil{Let us explore tool support!}
+\end{frame}
+
+{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\vspace*{-0em}\includegraphics[width=\paperwidth]{curry-howard-faded}\end{minipage}}
+\newcommand{\connection}[1]{$\cdots\includegraphics[height=1em]{#1}\cdots$}
+\begin{frame}{Curry--Howard's Rosetta stone}
+  \bigskip
+  \centering
+  \renewcommand{\arraystretch}{1.3}
+  \begin{tabular}{r@{}c@{}l}
+    \hil{mathematics} & & \hil{programming} \\
+    proving a claim & \connection{emoji-thinking} & implementing a function \\
+    stating a claim & \connection{emoji-megaphone} & writing down a type signature \\
+    using a lemma & \connection{emoji-link} & calling a function \\
+    {\ \ \ \ \,}coming up with an auxiliary lemma & \connection{emoji-lightbulb} & coming up with an auxiliary function \\
+    committing a logical fallacy & \connection{emoji-cross} & committing a type error \\
+    \pause
+    --- & \connection{emoji-explosion} & encountering a runtime bug
+  \end{tabular}
+  \pause
+
+  \begin{hilblock}
+    \centering
+    To extract numerical data from a proof, \\
+    \hil{run the proof}.
+  \end{hilblock}
+\end{frame}}
+
+\begin{frame}{Three case studies}
+  \bigskip
+  \begin{columns}[c]
+    \begin{column}{0.15\textwidth}
+      \scalebox{0.23}{\sieve{7}{7}}
+    \end{column}
+    \begin{column}{0.85\textwidth}
+      \begin{simpleblock}
+        \justifying
+        \textbf{Theorem.}
+        For every natural number~$n$, there is a prime larger than~$n$.
+      \end{simpleblock}
+      \\[0.3em]
+
+      \emph{Proof.} Every prime factor of~$n! + 1$ will do. \qed
+    \end{column}
+  \end{columns}
+  \bigskip
+
+  \begin{columns}[c]
+    \begin{column}{0.15\textwidth}
+      \includegraphics[width=\textwidth,valign=t]{monad}
+    \end{column}
+    \begin{column}{0.85\textwidth}
+      \begin{simpleblock}
+        \textbf{Theorem.}
+        Every infinite sequence~$f : \NN \to \NN$ is \emph{good}
+        in that there are numbers $i < j$ such that~$f(i) \leq f(j)$.
+      \end{simpleblock}
+      \\[0.3em]
+
+      \emph{Proof.} There is a \bad{minimal value}~$f(i)$. Set~$j \defeq i+1$. \qed
+    \end{column}
+  \end{columns}
+  \bigskip
+
+  \begin{columns}[c]
+    \begin{column}{0.15\textwidth}
+      \includegraphics[width=\textwidth,valign=t]{calabi-yau}
+    \end{column}
+    \begin{column}{0.85\textwidth}
+      \begin{simpleblock}
+        \justifying
+        \textbf{Theorem.}
+        Let~$M$ be a surjective matrix with more rows than columns over a
+        commutative ring~$A$. Then~$1 = 0$ in~$A$.
+      \end{simpleblock}
+      \\[0.3em]
+
+      \justifying\emph{Proof.}
+      \bad{Assume not.} Then there is~a \bad{maximal ideal} $\mmm$.
+      The matrix is still surjective over~$A/\mmm$. Since~$A/\mmm$ is a field, this
+      is a contradiction to basic linear algebra. \qed
+    \end{column}
+  \end{columns}
 \end{frame}

### 2025-08-25 slides-ai2025.tex
Final version

@@
 
-  \textbf{Theorem.} There is a \hil{unique} best~$L_1$-approximation of~$f$ in~$P_n$.
+  \textbf{Theorem.} There is a \hil{unique} best~$L^1$-approximation of~$f$ in~$P_n$.
   \pause
@@
 \begin{frame}{Backed by logical metatheorems?}
-  \hil{Metatheorems} have been and are being developed which \ldots
+  \hil{Metatheorems} for backing proof mining have been and are being developed which \ldots
   \begin{enumerate}
-    \item guarantee the \hil{extractability} of suitable numerical information in principle
-    (bounds on norms, convergence rates, moduli of uniqueness, rates of
-    asymptotic regularity, \ldots),
-    \item describe an \hil{algorithm} for carrying out the extraction and
-    \item support \hil{modular treatments} of auxiliary lemmas,
+    \small
+    \item[{\includegraphics[height=1em,valign=c]{emoji-abacus}}] guarantee the
+    \hil{extractability} of suitable numerical information in principle \\
+    (bounds, convergence rates, moduli of uniqueness, rates of asymptotic
+    regularity, \ldots),
+    \item[{\includegraphics[height=1em,valign=c]{emoji-robot}}] describe an \hil{algorithm} for carrying out the extraction and
+    \item[{\includegraphics[height=1em,valign=c]{emoji-package}}] support \hil{modular treatments} of auxiliary lemmas,
   \end{enumerate}
-  \bad{\textbf{provided}} the input proof is supplied in a
-  certain \hil{formal system}.
+  \bad{\textbf{provided}} the input proof is \hil{formally} supplied in a
+  \hil{certain system}.
+  \fixedhref{https://www2.mathematik.tu-darmstadt.de/~kohlenbach/novikov.pdf}{[Kohlenbach--Oliva 2002]}
+  \pause
+
+  \bad{\textbf{In practice (2025):}} Algorithms \bad{\textbf{not}} used, only followed
+  as \bad{\textbf{rough guidelines}}, combined with hand-rolled optimizations.
+  \hil{Let us explore tool support!}
   \pause
@@
   (Heine--Borel, Bolzano--Weierstraß, \ldots).
-  \pause
-
-  \bad{\textbf{In practice (2025):}} Algorithms \bad{\textbf{not}} used, only followed
-  as \bad{\textbf{rough guidelines}}, combined with hand-rolled optimizations.
-  \hil{Let us explore tool support!}
 \end{frame}
@@
     {\ \ \ \ \,}coming up with an auxiliary lemma & \connection{emoji-lightbulb} & coming up with an auxiliary function \\
-    committing a logical fallacy & \connection{emoji-cross} & committing a type error \\
+    committing a logical error & \connection{emoji-cross} & committing a type error \\
+    reasoning circularly & \connection{emoji-cyclone} & getting stuck in an infinite loop \\
     \pause
@@
 
+\newcommand{\laterbad}[1]{\only<1>{#1}\only<2->{\bad{#1}}}
 \begin{frame}{Three case studies}
@@
         For every natural number~$n$, there is a prime larger than~$n$.
+        \fixedhref{https://unimath.github.io/agda-unimath/elementary-number-theory.infinitude-of-primes.html}{\formalized}
       \end{simpleblock}
@@
         in that there are numbers $i < j$ such that~$f(i) \leq f(j)$.
+        \fixedhref{https://lets-play-agda.quasicoherent.io/Padova2025.ComputationalContent.Dickson.html}{\formalized}
       \end{simpleblock}
@@
 
-      \emph{Proof.} There is a \bad{minimal value}~$f(i)$. Set~$j \defeq i+1$. \qed
+      \emph{Proof.} There is a \laterbad{minimal value}~$f(i)$. Set~$j \defeq i+1$. \qed
     \end{column}
@@
         commutative ring~$A$. Then~$1 = 0$ in~$A$.
+        \fixedhref{https://iblech.github.io/constructive-maximal-ideals/Krull.Dynamical.html\#example}{\formalized}
       \end{simpleblock}
@@
       \justifying\emph{Proof.}
-      \bad{Assume not.} Then there is~a \bad{maximal ideal} $\mmm$.
+      \laterbad{Assume not.} Then there is a \laterbad{maximal ideal}~$\mmm$.
       The matrix is still surjective over~$A/\mmm$. Since~$A/\mmm$ is a field, this
@@
 \end{frame}
+
+\renewcommand{\insertframeextra}{a}
+\begin{frame}{A case study in commutative algebra}
+  \begin{simpleblock}
+    \textbf{Theorem.}
+    Let~$M$ be a surjective matrix with more rows than columns over a
+    commutative ring~$A$. Then~$1 = 0$ in~$A$.
+    \fixedhref{https://iblech.github.io/constructive-maximal-ideals/Krull.Dynamical.html\#example}{\formalized}
+  \end{simpleblock}
+  \\[0.3em]
+
+  \justifying\emph{Proof.}
+  \bad{Assume not.} Then there is a \bad{maximal ideal}~$\mmm$.
+  The matrix is still surjective over~$A/\mmm$. Since~$A/\mmm$ is a field, this
+  is a contradiction to basic linear algebra. \qed
+
+  {\centering\scalebox{0.9}{\begin{tikzpicture}
+    \node (0) at (0,1) {$(0) = \{0\}$};
+    \node (1) at (0,5) {$(1) = \ZZ$};
+    \node (2) at (-2,4) {$(2)$};
+    \node [right of=2] (3) {$(3)$};
+    \node [below of=2] (4) {$(4)$};
+    \node [below of=2, xshift=0.7cm] (6) {$(6)$};
+    \node [right of=3] (5) {$(5)$};
+    \node [right of=5] (7) {$(7)$};
+    \node [right of=7] (7d) {$\ldots$\phantom{(}};
+    \node [right of=7d, xshift=1cm, yshift=-1cm] (max)
+    {\small\it maximal among the proper ideals};
+    \node [below of=4] (8) {$(8)$};
+    \node [right of=8, xshift=3cm] (8d) {$\ldots$};
+    \draw (0) -- (8);
+    \draw (0) -- (8d);
+    \draw (0) -- (6);
+    \draw (2) -- (1);
+    \draw (3) -- (1);
+    \draw (5) -- (1);
+    \draw (7) -- (1);
+    \draw (7d) -- (1);
+    \draw (4) -- (2);
+    \draw (8) -- (4);
+    \draw (6) -- (2);
+    \draw (6) -- (3);
+    \draw [mypurple!30, thick, shorten <=-2pt, shorten >=-2pt, ->] (max) to [out=120, in=-30] (7d);
+    \begin{pgfonlayer}{background}
+      \draw[decorate, very thick, draw=mypurple!30]
+        ($(2.south west) + (8pt, 0)$) arc(270:180:8pt) --
+        ($(2.north west) + (0, -8pt)$) arc(180:90:8pt) --
+        ($(7d.north east) + (-8pt, 0)$) arc(90:0:8pt) --
+        ($(7d.south east) + (0, 8pt)$) arc(0:-90:8pt) --
+        cycle;
+    \end{pgfonlayer}
+  \end{tikzpicture}}\par}
+  \pause
+
+  \begin{mining}[0.8\textwidth]\emph{Proof.}
+  Write~$M = \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right)$. By surjectivity,
+  have~$u, v \in A$ with
+  $
+    u \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}1\\0\end{smallmatrix}\right)
+  $ and $
+    v \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}0\\1\end{smallmatrix}\right)
+  $.
+  Hence
+  $
+    1 = (vy)(ux) = (uy)(vx) = 0
+  $. \qed
+  \end{mining}
+\end{frame}
+
+\addtocounter{framenumber}{-1}
+\renewcommand{\insertframeextra}{b}
+\begin{frame}{Maximal ideals as convenient fictions?}
+  \begin{itemize}
+    \item[\triang] In \bad{classical mathematics}, every ring has a maximal ideal
+    by \bad{Zorn's lemma}.
+    \bigskip\pause
+
+    \item[\triang] Without \bad{Zorn}, at least every \bad{countable} ring
+    $A = \{ x_0, x_1, \ldots \}$ has a maximal ideal.
+
+    \small
+    -- Iterative construction given
+    an ideal membership test
+    \fixedhref{https://link.springer.com/article/10.1007/BF01454872}{[Krull
+    1929]}:
+    \begin{align*}
+      \mmm_0 &= \{ 0 \}, &
+      \mmm_{n+1} &= \begin{cases}
+        \mmm_n + (x_n), & \text{if $1 \not\in \mmm_n + (x_n)$}, \\
+        \mmm_n, & \text{else.}
+      \end{cases}
+      \intertext{\visible<3->{-- Also without membership test!
+      \fixedhref{https://citeseerx.ist.psu.edu/document?repid=rep1&type=pdf&doi=a7ff57cacdad6735040459dcf450b231948dbb2f}{[Krivine
+      1996]},
+      \fixedhref{https://www.sciencedirect.com/science/article/pii/S0168007204000181}{[Berardi--Valentini
+      2004]}}}
+      \action<3->{\mmm_0 &= \{ 0 \}, &
+      \mmm_{n+1} &= \mmm_n + (\underbrace{\{ x_n \,|\, 1 \not\in \mmm_n + (x_n)
+      \}}_{\text{a certain subsingleton set}})}
+    \end{align*}
+    \bigskip
+    \pause
+    \pause
+
+    \normalsize
+    \justifying
+    \item[\triang] Without \bad{Zorn}, \hil{every} ring has a maximal ideal in
+    a ``suitable \hil{forcing extension} of the universe''
+    \fixedhref{https://raw.githubusercontent.com/iblech/constructive-maximal-ideals/master/tex/extended.pdf}{[B.--Schuster
+    2024]}.
+    In plain terms:
+    \hil{Approximate} a (perhaps non-existing) surjection~$\NN \twoheadrightarrow A$
+    by partial functions which can \hil{grow on demand}.
+  \end{itemize}
+\end{frame}
+\renewcommand{\insertframeextra}{}
 
@@
   - resulting constructive proofs/algorithms can look "alien" (= artificial intelligence)
-
-- Agda icons
-- convenient fictions
-
-- formulation problem
-- infinite objects
-- proving = programming

### 2025-12-17 paper-flabby-objects.tex
typo
to do

@@
 https://arxiv.org/abs/1903.01211
+
+Cite:
+https://arxiv.org/abs/1802.00960
+
+https://ncatlab.org/nlab/show/algebraically+injective+object
+
+https://www.cse.chalmers.se/~coquand/course231.pdf#page=23
+Perhaps flabby objects could be called "contractible objects"?
+
+cite
+https://arxiv.org/pdf/1903.01211
+Thm. 19, 20
+for discussion of "functionally flabby ↔ injective"
+
+Cite general idea or injectivity aspects of:
+https://1lab.dev/Data.Partial.Properties.html
